### 2012-05-23 doc/hmcsa_adjoint_vs_direct/direct_vs_adjoint.tex
Adding HMCSA direct vs. adjoint solver study

%%---------------------------------------------------------------------------%%
%% direct_vs_adjoint.tex
%% Stuart Slattery
%% Wed May 23 11:25:38 2012
%% Copyright (C) 2008-2010 Oak Ridge National Laboratory, UT-Battelle, LLC.
%%---------------------------------------------------------------------------%%
\documentclass[note]{TechNote}
\usepackage[centertags]{amsmath}
\usepackage{amssymb,amsthm}
\usepackage[mathcal]{euscript}
\usepackage{tabularx}
\usepackage{cite}
\usepackage{c++}
\usepackage{tmadd,tmath}

%%---------------------------------------------------------------------------%%
%% DEFINE SPECIFIC ENVIRONMENTS HERE
%%---------------------------------------------------------------------------%%
%\newcommand{\elfit}{\ensuremath{\operatorname{Im}(-1/\epsilon(\vq,\omega)}}
%\newcommand{\vOmega}{\ensuremath{\ve{\Omega}}}
%\newcommand{\hOmega}{\ensuremath{\hat{\ve{\Omega}}}}

%%---------------------------------------------------------------------------%%
%% BEGIN DOCUMENT
%%---------------------------------------------------------------------------%%
\begin{document}

%%---------------------------------------------------------------------------%%
%% OPTIONS FOR NOTE
%%---------------------------------------------------------------------------%%

\refno{RNSD-01}
\subject{Stochastic Linear Solver Comparison for Monte Carlo Synthetic
Acceleration Method}

%-------HEADING
\TIname{Reactor and Nuclear Systems Division}
\groupname{Radiation Transport Group}
\from{uy7}
\date{\today}
%-------HEADING

\audience{\email{uy7}{uy7@ornl.gov}}

%%---------------------------------------------------------------------------%%
%% BEGIN NOTE
%%---------------------------------------------------------------------------%%

\opening

\begin{abstract}
  The Monte Carlo Synthetic Acceleration (MCSA) method utilizes a
  stochastic linear solver in its solution scheme. Two stochastic
  solvers are available in the form of a direct and an adjoint
  formulation. This work implements the direct and adjoint solvers
  within an MCSA method and compares their speed and convergengence
  properties to determine which, if either, is most suitable for use
  with MCSA. 
\end{abstract}

%%---------------------------------------------------------------------------%%
%% OPTIONAL TOC

% \memotoc

%%---------------------------------------------------------------------------%%
\section{Introduction}

%%---------------------------------------------------------------------------%%
\section{Model Problem}

%%---------------------------------------------------------------------------%%
\section{Timing Study Results}

%%---------------------------------------------------------------------------%%
\section{Convergence Study Results}

%%---------------------------------------------------------------------------%%
\section{Conclusion}

%%---------------------------------------------------------------------------%%
%% BIBLIOGRAPHY STUFF

% \bibliographystyle{notes}
% \bibliography{direct_vs_adjoint}

\closing
\caution
\end{document}

%%---------------------------------------------------------------------------%%
%% end of direct_vs_adjoint.tex
%%---------------------------------------------------------------------------%%

### 2012-05-23 doc/hmcsa_adjoint_vs_direct/direct_vs_adjoint.tex
Added results png files for adjoint vs. direct study

@@
 \groupname{Radiation Transport Group}
-\from{uy7}
+\from{Stuart R. Slattery}
 \date{\today}
@@
 
-\audience{\email{uy7}{uy7@ornl.gov}}
+\audience{\email{Stuart Slattery}{uy7@ornl.gov},
+          \email{Thomas Evans}{tme@ornl.gov},
+          \email{Greg Davidson}{gqe@ornl.gov}}
 
@@
   formulation. This work implements the direct and adjoint solvers
-  within an MCSA method and compares their speed and convergengence
+  within an MCSA solver and compares their speed and convergengence
   properties to determine which, if either, is most suitable for use
-  with MCSA. 
+  with MCSA. The studies show that the adjoint solver requires
+  significantly less CPU time for convergence than the direct
+  method. Therefore, it is recommended that the adjoint solver be used
+  as the linear solver within an MCSA solver implementation.
 \end{abstract}
@@
 \section{Introduction}
+The Monte Carlo Synthetic Acceleration (MCSA) method developed by
+Evans et. al. \cite{Evans_2009} provides an acceleration to standard
+stochastic linear solver methods that are bound by the central limit
+theorem, instead giving exponential convergence. To do this, we must
+first consider stochastic solutions for linear systems. For the
+following linear system
+
+\begin{equation}
+  A x = b
+  \label{eq:linear_system}
+\end{equation}
+
+we can rewrite using an iteration matrix formulation as follows.
+
+\begin{equation}
+  x = (I-A) x + b
+  \label{eq:iteration_form}
+\end{equation}
+
+We will define $H=I-A$ to be the iteration matrix. Using this
+definition we can then rewrite the original linear system using the
+iteration matrix.
+
+\begin{equation}
+  x = (I-H)^{-1} b
+  \label{eq:iteration_form2}
+\end{equation}
+
+If the spectral radius of the iteration matrix is less than unity,
+then we can rewrite the inverted operator as a Von Neumann series.
+
+\begin{equation}
+  (I-H)^{-1} = \sum_k H^k
+  \label{eq:von_neumann_series}
+\end{equation}
+
+We then expect the following series solution for x to converge if the
+spectral radius of the iteration matrix is less than unity.
+
+\begin{equation}
+  x = b + H b + H^2 b + H^3 b + ...
+  \label{eq:series_solution}
+\end{equation}
+
+Using this form, we can approximate the terms in the series with a
+random walk sequence as outlined in \cite{Hammersley_1964}. There are
+two primary ways of achieving this; through a direct sequence and an
+adjoint sequence. For the direct sequence, analagous (and
+inconveniently so) to an adjoint stochastic transport calculation,
+each term in the source is simulated with an equivalent number of
+random walk histories. Each step in the random walk sequence will
+contribute to the solution in the starting source state. For the
+adjoint sequence, analagous to a forward stochastic transport
+calculation, the source term is sampled with a set number of random
+walk histories, meaning that regions of larger source will be sampled
+as a starting state more regularly. In this case, each step in the
+random walk sequence will contribute to the state in which it
+currently resides, not the current source state. It is this set of
+primary differences between the two solution methods that will dictate
+solver performance with MCSA.
+
+With this knowledge, for the linear system in
+Eq. \ref{eq:linear_system} we can then define the following MCSA
+solution sequence
+
+\begin{equation}
+  x^{k+1/2} = (I-A)x^k + b
+  \label{eq:mcsa_step1}
+\end{equation}
+\begin{equation}
+  r = b - A x^{k+1/2}
+  \label{eq:mcsa_step2}
+\end{equation}
+\begin{equation}
+  A \delta x = r
+  \label{eq:mcsa_step3}
+\end{equation}
+\begin{equation}
+  x^{k+1} = x^{k+1/2} + \delta x
+  \label{eq:mcsa_step4}
+\end{equation}
+
+where k is the iteration index. In brief, in Eq. \ref{eq:mcsa_step1}
+we compute an initial relaxation for $x$ with a Richardson iteration,
+in Eq. \ref{eq:mcsa_step2} we compute the residual, in
+Eq. \ref{eq:mcsa_step3} we compute the solution update via a
+stochastic linear solve and finally we update the solution in
+Eq. \ref{eq:mcsa_step4}. Of interest to this work is the stochastic
+linear solution needed to compute the solution update in Eq.
+\ref{eq:mcsa_step3} which we will choose to use either the direct or
+adjoint method. It is important to note that the stochastic solver is
+not computing $x$, rather it is computing an error term using the
+residual that we will apply to $x$.
 
@@
 \section{Model Problem}
+To study the behaviour of the MCSA method with both direct and adjoint
+solutions, we choose the two-dimensional time-dependent heat equation
+as a simple model problem.
+
+\begin{equation}
+  \frac{\partial u}{\partial t} = \alpha \nabla^2 u
+  \label{eq:heat_equation}
+\end{equation}
+
+where $\alpha$ is a constant coefficient. For all comparisons, a
+single time step is computed with implicit Euler time integration. The
+Laplacian is differenced with a second-order five point stencil
+
+\begin{equation}
+  \nabla^2_5 = \frac{1}{h^2}[u_{i-1,j} + u_{i+1,j} + u_{i,j-1} +
+  u_{i,j+1} - 4 u_{i,j}]
+  \label{eq:five_point_stencil}
+\end{equation}
+
+and a fourth-order nine point stencil both assuming a grid size of $h$
+in both the $i$ and $j$ directions.
+
+\begin{equation}
+  \nabla^2_9 = \frac{1}{6h^2}[4 u_{i-1,j} + 4 u_{i+1,j} + 4 u_{i,j-1}
+    + 4 u_{i,j+1} + u_{i-1,j-1} + u_{i-1,j+1} + u_{i+1,j-1} +
+    u_{i+1,j+1} - 20 u_{i,j}]
+  \label{eq:nine_point_stencil}
+\end{equation}
+
+For a single time step solution, we then have the following linear
+system to be solved with the MCSA method.
+
+\begin{equation}
+  A u^{n+1} = u^n
+  \label{eq:heat_eq_lin_sys}
+\end{equation}
+
+Both the stencils will be used to vary the size and density of the
+sparse linear system in Eq. \ref{eq:heat_eq_lin_sys}.
 
@@
 \section{Timing Study Results}
+To assess both the CPU time and number of iterations required to
+converge to a solution, a problem of constant grid size and $\alpha$
+was used with varying values of problem size, fixing the spectral
+radius of the system at a constant value for each variation. Both the
+five point and nine point stencils were used with both the direct and
+adjoint solvers. For each case, 50 histories were simulated per cycle
+with a weight cutoff of $1.0E-4$ for the stochastic solver and a
+convergence tolerance of $1.0E-8$ for the MCSA solver. All
+computations were completed on a 2.66 GHz Intel Core i7 machine with 8
+GB 1067 MHz DDR3 memory.
+
+Fig. \ref{fig:cpu_time} gives the CPU time needed for each case to
+converge in seconds and Fig. \ref{fig:iterations} gives the number of
+iterations needed for each case to converge.
 

### 2012-05-23 doc/hmcsa_adjoint_vs_direct/direct_vs_adjoint.tex
Updated tech note for direct vs. adjoint study

@@
 converge in seconds and Fig. \ref{fig:iterations} gives the number of
-iterations needed for each case to converge.
+iterations needed for each case to converge. We see clearly in
+Fig. \ref{fig:cpu_time} that the using the adjoint solver with MCSA
+results in several orders of magnitude speedup over the direct solver
+while the number of iterations required to converge is of a similar
+scale. We expect this for several reasons. First, with an equivalent
+number of histories specified for both solvers and a system of size
+$N$, the direct solver will compute $N$ times the number of histories
+random walks while the adjoint solver will only compute the number of
+histories specified. Initially, this may not seem like a fair
+comparison, however, we see from Fig. \ref{fig:cpu_time} that the
+number of iterations requried to converge is approximately the
+same. This is necessary in the direct method to ensure a contribution
+from each state as the random walk sequence will only contribute to
+the starting state. For the adjoint method, because the random
+sequence contributes to the state in which it currently resides, fewer
+histories are neccessary to compute contributions from all the nonzero
+parts of the source.
 
@@
 \section{Convergence Study Results}
+Although CPU time was shown to be greatly decreased by using the
+adjoint solver in the preceeding section, the iteration behavior was
+approximately the same. To assess the convergence properties of MCSA
+using each solver and stencil, the infinity norm of the residual
+computed in Eq. \ref{eq:mcsa_step2} was computed at each iteration for
+a fixed problem size. Fig. \ref{fig:convergence} gives the results of
+these computuations. First, it is worthy to note on the semilog plot
+that we are indeed achieving the expected exponential convergence from
+MCSA. Second, we note that both the direct and adjoint methods have
+approximately equivalent convergence behavior. Finally, it is observed
+the the fourth-order accurate Laplacian stencil converges in fewer
+iterations than the second-order accurate stencil as expected.
 
@@
 \section{Conclusion}
+A parameter study has been completed to verify that an adjoint
+stochastic solver should be used with an MCSA implementation instead
+of a direct stochastic solver. Several orders of magnitude reduction
+in CPU time is observed when using the adjoint solver over the
+direct. In addition, it was observed that MCSA had similar exponential
+convergence properties with both the direct and adjoint solvers. Even
+for cases where fewer iterations were needed to converge to a solution
+using the direct solver, the large number of histories required caused
+the CPU time to become large compared to the adjoint method.
 

### 2012-05-23 doc/hmcsa_adjoint_vs_direct/direct_vs_adjoint.tex
Added results figures to adjoint vs direct tech note

@@
 
-\refno{RNSD-01}
-\subject{Stochastic Linear Solver Comparison for Monte Carlo Synthetic
-Acceleration Method}
+\refno{RNSD-01} \subject{Stochastic Linear Solver Comparison for the
+  Monte Carlo Synthetic Acceleration Method}
 
@@
 The Monte Carlo Synthetic Acceleration (MCSA) method developed by
-Evans et. al. \cite{Evans_2009} provides an acceleration to standard
+Evans et. al. \cite{evans_2003} provides an acceleration to standard
 stochastic linear solver methods that are bound by the central limit
@@
 Using this form, we can approximate the terms in the series with a
-random walk sequence as outlined in \cite{Hammersley_1964}. There are
+random walk sequence as outlined in \cite{hammersley_1964}. There are
 two primary ways of achieving this; through a direct sequence and an
@@
 
+\begin{figure}[htpb!]
+  \centering
+  \includegraphics[width=5in]{AdjointDirectCPUTime.png}
+  \caption{\sl CPU Time (s) to converge vs. Problem Size (N for an NxN
+    problem). Both the adjoint and direct solvers are used with the
+    five point and nine point stencils. A siginificant CPU time
+    speedup is noted with the adjoint method due to the reduced number
+    of histories.}
+  \label{fig:cpu_time}
+\end{figure}
+
+\begin{figure}[htbp!]
+  \centering
+  \includegraphics[width=5in]{AdjointDirectIterations.png}
+  \caption{\sl Iterations to converge vs. Problem Size (N for an NxN
+    problem). Both the adjoint and direct solvers are used with the
+    five point and nine point stencils. Both methods give similar
+    iteration behavior.}
+  \label{fig:iterations}
+\end{figure}
+
 %%---------------------------------------------------------------------------%%
@@
 iterations than the second-order accurate stencil as expected.
+
+\begin{figure}[htpb!]
+  \centering
+  \includegraphics[width=5in]{AdjointDirectConvergence.png}
+  \caption{\sl Infinity norm of the solution residual vs. iteration
+    number for a problem of fixed size. Both the adjoint and direct
+    solvers are used with the five point and nine point stencils. Both
+    methods give similar iteration behavior.}
+  \label{fig:convergence}
+\end{figure}
 
@@
 
-% \bibliographystyle{notes}
-% \bibliography{direct_vs_adjoint}
+\bibliographystyle{rnotes}
+\bibliography{direct_vs_adjoint}
 

### 2012-05-23 doc/hmcsa_adjoint_vs_direct/direct_vs_adjoint.tex
Fixed a few typos in the direct vs adjoint tech note

@@
 stochastic linear solver methods that are bound by the central limit
-theorem, instead giving exponential convergence. To do this, we must
-first consider stochastic solutions for linear systems. For the
-following linear system
+theorem by instead giving exponential convergence. However, a
+component of the MCSA algorithm requires a linear stochastic solve to
+still be performed. Both the direct and adjoint methods are available
+to achieve this solution and will be studied here to assess if either
+provides improved performance over the other.
+
+We first consider an overview of stochastic solutions for linear
+systems. For the following linear system
 
@@
 linear solution needed to compute the solution update in Eq.
-\ref{eq:mcsa_step3} which we will choose to use either the direct or
-adjoint method. It is important to note that the stochastic solver is
-not computing $x$, rather it is computing an error term using the
+\ref{eq:mcsa_step3} for which we will choose to use either the direct
+or adjoint method. It is important to note that the stochastic solver
+is not computing $x$, rather it is computing an error term using the
 residual that we will apply to $x$.
@@
 histories are neccessary to compute contributions from all the nonzero
-parts of the source.
+parts of the source. Therefore, although the total number of histories
+per cycle is in fact larger for the direct method, the solution
+computed at each iteration is no better than that computed by the
+adjoint method.
 
@@
     problem). Both the adjoint and direct solvers are used with the
-    five point and nine point stencils. A siginificant CPU time
-    speedup is noted with the adjoint method due to the reduced number
-    of histories.}
+    five point and nine point stencils. A significant CPU time speedup
+    is noted with the adjoint method due to the reduced number of
+    histories.}
   \label{fig:cpu_time}

### 2012-05-26 doc/jcomp2012/jcomp2012.tex
Adding bibtex file

%%---------------------------------------------------------------------------%%
%% Monte Carlo Synthetic Acceleration Paper
%% Submitted to Journal of Computational Physics 2012
%%---------------------------------------------------------------------------%%
%%
%% Copyright 2007, 2008, 2009 Elsevier Ltd
%%
%% This file is part of the 'Elsarticle Bundle'.
%% ---------------------------------------------
%%
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%%
%% The list of all files belonging to the 'Elsarticle Bundle' is
%% given in the file `manifest.txt'.
%%

%%---------------------------------------------------------------------------%%
%% Document Options
%%---------------------------------------------------------------------------%%
%% Template article for Elsevier's document class `elsarticle'
%% with numbered style bibliographic references
%% SP 2008/03/01
%%
%%
%%
%% $Id: elsarticle-template-num.tex 4 2009-10-24 08:22:58Z rishi $
%%
%%
\documentclass[preprint,12pt]{elsarticle}

%% Use the option review to obtain double line spacing
%% \documentclass[preprint,review,12pt]{elsarticle}

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
%% \documentclass[final,1p,times]{elsarticle}
%% \documentclass[final,1p,times,twocolumn]{elsarticle}
%% \documentclass[final,3p,times]{elsarticle}
%% \documentclass[final,3p,times,twocolumn]{elsarticle}
%% \documentclass[final,5p,times]{elsarticle}
%% \documentclass[final,5p,times,twocolumn]{elsarticle}

%% if you use PostScript figures in your article
%% use the graphics package for simple commands
%% \usepackage{graphics}
%% or use the graphicx package for more complicated commands
%% \usepackage{graphicx}
%% or use the epsfig package if you prefer to use the old commands
%% \usepackage{epsfig}

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers after \end{frontmatter}.
%% \usepackage{lineno}

%% natbib.sty is loaded by default. However, natbib options can be
%% provided with \biboptions{...} command. Following options are
%% valid:

%%   round  -  round parentheses are used (default)
%%   square -  square brackets are used   [option]
%%   curly  -  curly braces are used      {option}
%%   angle  -  angle brackets are used    <option>
%%   semicolon  -  multiple citations separated by semi-colon
%%   colon  - same as semicolon, an earlier confusion
%%   comma  -  separated by comma
%%   numbers-  selects numerical citations
%%   super  -  numerical citations as superscripts
%%   sort   -  sorts multiple citations according to order in ref. list
%%   sort&compress   -  like sort, but also compresses numerical citations
%%   compress - compresses without sorting
%%
%% \biboptions{comma,round}

% \biboptions{}

\journal{Computational Physics}

%%---------------------------------------------------------------------------%%
%% Math Variables
%%---------------------------------------------------------------------------%%
\newcommand{\email}[1]{$\langle$#1@lanl.gov$\rangle$}

\newcommand{\vA}{\ensuremath{\ve{A}}}
\newcommand{\vb}{\ensuremath{\ve{b}}}
\newcommand{\vx}{\ensuremath{\ve{x}}}
\newcommand{\vr}{\ensuremath{\ve{r}}}
\newcommand{\vI}{\ensuremath{\ve{I}}}
\newcommand{\vH}{\ensuremath{\ve{H}}}
\newcommand{\vP}{\ensuremath{\ve{P}}}

\newcommand{\Cv}{\ensuremath{C_{v}}}
\newcommand{\ros}{\ensuremath{\sigma_{\scriptscriptstyle\mathrm{R}}}}
\newcommand{\bOmega}{\ensuremath{\mathbf{\Omega}}}

\newcommand{\dt}{\ensuremath{\Delta t}}
\newcommand{\sign}{\ensuremath{\tilde{\sigma}^n}}
\newcommand{\qn}{\ensuremath{q^n}}
\newcommand{\Tn}{\ensuremath{T^n}}
\newcommand{\Dn}{\ensuremath{D^n}}
\newcommand{\phin}{\ensuremath{\phi^n}}
\newcommand{\Di}{\ensuremath{\Delta_i}}
\newcommand{\Dj}{\ensuremath{\Delta_j}}
\newcommand{\Dk}{\ensuremath{\Delta_k}}

\newcommand{\sigT}{\ensuremath{\sigma_{T\,ijk}}}
\newcommand{\sigm}{\ensuremath{\sigma^{-}}}
\newcommand{\sigp}{\ensuremath{\sigma^{+}}}

\newcommand{\bphi}{\ensuremath{\boldsymbol{\phi}}}

\DeclareMathOperator{\diag}{diag}

%%---------------------------------------------------------------------------%%
\begin{document}

\begin{frontmatter}

%% Title, authors and addresses

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for the associated footnote;
%% use the fnref command within \author or \address for footnotes;
%% use the fntext command for the associated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for the associated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
%%
%% \title{Title\tnoteref{label1}}
%% \tnotetext[label1]{}
%% \author{Name\corref{cor1}\fnref{label2}}
%% \ead{email address}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
%% \address{Address\fnref{label3}}
%% \fntext[label3]{}

\title{A Monte Carlo Synthetic Acceleration Method for Solving the
  Thermal Radiation Diffusion Equation}

%% use optional labels to link authors explicitly to addresses:
%% \author[label1,label2]{<author name>}
%% \address[label1]{<address>}
%% \address[label2]{<address>}

\author{}

\address{}

%%---------------------------------------------------------------------------%%
\begin{abstract}
%% Text of abstract

\end{abstract}

%%---------------------------------------------------------------------------%%
\begin{keyword}
%% keywords here, in the form: keyword \sep keyword

%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)

\end{keyword}

\end{frontmatter}

%%
%% Start line numbering here if you want
%%
% \linenumbers

%% main text
%%---------------------------------------------------------------------------%%
\section{}
\label{}

%%---------------------------------------------------------------------------%%
%% The Appendices part is started with the command \appendix;
%% appendix sections are then done as normal sections
%% \appendix

%% \section{}
%% \label{}

%% References
%%
%% Following citation commands can be used in the body text:
%% Usage of \cite is as follows:
%%   \cite{key}         ==>>  [#]
%%   \cite[chap. 2]{key} ==>> [#, chap. 2]
%%

%%---------------------------------------------------------------------------%%
%% References with bibTeX database:

\bibliographystyle{elsarticle-num}
\bibliography{jcomp2012.bib}

%% Authors are advised to submit their bibtex database files. They are
%% requested to list a bibtex style file in the manuscript if they do
%% not want to use elsarticle-num.bst.

%% References without bibTeX database:

% \begin{thebibliography}{00}

%% \bibitem must have the following form:
%%   \bibitem{key}...
%%

% \bibitem{}

% \end{thebibliography}

\end{document}

%%
%% End of file `elsarticle-template-num.tex'.

### 2012-05-26 doc/jcomp2012/jcomp2012.tex
Adding images from nse07 paper

@@
   \begin{abstract}
-%% Text of abstract
+
+    We present a synthetic-acceleration based Monte Carlo method for
+    solving the 1T thermal radiation diffusion equations.  We show
+    that this method can be an effective solver for sparse matrix
+    systems.  We also discuss its general applicability to broader
+    classes of problems.
 
@@
 %%---------------------------------------------------------------------------%%
-\section{}
-\label{}
+\section{Introduction}
+\label{sec:introduction}
+
+In Ref.~\cite{evans_2003}, we developed a residual Monte Carlo method
+that was an application of Halton's Sequential Monte Carlo
+method~\cite{halton_1962,halton_1994} to the 1D equilibrium (1T)
+thermal radiation diffusion equation.  At that time, Halton's method
+was not widely known in computational physics.  While extending our
+method to 3D, we have discovered that the Sequential Monte Carlo
+method is actually a variant of iterative refinement cast as a
+residual method.  Using Monte Carlo to accelerate fixed-point
+(Richardson) iteration, our new method surpasses our previous 1D
+method and allows for extension to more general solution techniques.
+
+The purpose of this study is to demonstrate an efficient Monte Carlo
+solution method for 3D, time-dependent discrete systems.  In this
+note, we extend our previous work in the following manner:
+\begin{itemize}
+\item We develop a rigorous synthetic-acceleration Monte Carlo method
+  for solving sparse matrix systems.
+\item We apply this solver to the 1T thermal radiation diffusion
+  equation in 3D.
+\end{itemize}
+In \S~\ref{sec:monte-carlo-matrix} we review Monte Carlo methods for
+solving matrix systems. In \S~\ref{sec:iter-refin-monte} we discuss
+our new synthetic-acceleration scheme and connect with previous work
+by Halton~\cite{halton_1994} and others~\cite{evans_2003}.  The 1T
+radiation diffusion equation and its discrete form will be derived in
+\S~\ref{sec:discr-form-radi}.  We discuss using Synthetic-Accleration
+Monte Carlo to solve the discrete radiation diffusion equation in
+\S~\ref{sec:solut-radi-diff}.  Results from two 3D test problems are
+given in \S~\ref{sec:results}.  We finish with conclusions and ideas
+for future work in \S~\ref{sec:conclusions}.
 

### 2012-05-26 doc/jcomp2012/jcomp2012.tex
Dump of nse07 paper into tex file

@@
 %%---------------------------------------------------------------------------%%
+\section{Monte Carlo Matrix Solution Methods}
+\label{sec:monte-carlo-matrix}
+
+Consider the following matrix equation:
+\begin{equation}
+  \vA\vx = \vb\:,
+  \label{eq:Ax=b}
+\end{equation}
+which can be written,
+\begin{equation}
+  \begin{split}
+    \vx &= (\vI - \vA)\vx + \vb\\
+    &= \vH\vx + \vb\:.
+  \end{split}
+  \label{eq:point_iteration}
+\end{equation}
+Here, $\vH = (\vI - \vA)$ is the \latin{iteration matrix}.  When the
+spectral radius, $\rho(\vH)$, of the iteration matrix satisfies
+\begin{equation}
+  \rho(\vH) < 1\:,
+\end{equation}
+we can expand $\vA^{-1}$ using the \latin{Neumann Series},
+\begin{equation}
+  \vA^{-1} = (\vI - \vH)^{-1} =  \sum_{k=0}^{\infty} \vH^k\:.
+\end{equation} 
+Thus, when $\rho(\vH) < 1$ we can recast the solution vector $\vx$ as
+a series,
+\begin{equation}
+  \begin{split}
+    \vx &= (\vI - \vH)^{-1}\vb\\
+    &= \vb + \vH\vb + \vH^2\vb + \vH^3\vb + \ldots\:.
+    \label{eq:neumann_series}
+  \end{split}
+\end{equation}
+With this knowledge in hand, we can write an iterative method that
+solves Eq.~(\ref{eq:Ax=b}) (Richardson's Iteration),
+\begin{equation}
+  \vx^{k+1} = \vH\vx^k + \vb\:.
+  \label{eq:richardson}
+\end{equation}
+Equation~(\ref{eq:richardson}) will converge when $\rho(\vH) < 1\ 
+\forall\ \vb \in \mathcal{R}^{N}$~\cite{kelley_1995}.
+
+\subsection{Direct Method}
+\label{sec:direct-method}
+
+Now we consider a Monte Carlo method that can be used to estimate the solution
+to Eq.~(\ref{eq:Ax=b})~\cite{hammersley_1964}.  First, rewrite
+Eq.~(\ref{eq:neumann_series}) in order to calculate a component of
+$\vx$,
+\begin{equation}
+  \begin{split}
+    x_i &= (\vb)_i + (\vH\vb)_i + (\vH^2\vb)_i + (\vH^3\vb)_i +
+    \ldots\\
+    &= \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
+    \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}\:.
+  \end{split}
+  \label{eq:neumann_series_i}
+\end{equation}
+This series can be interpreted as a series of transitions from
+$i_{m-1}\rightarrow i_m$ that can be simulated by a random walk.
+Let $X$ be a random variable sampled from a random walk with $k$
+events that initiates in state $i$,
+\begin{equation}
+  \begin{split}
+    X(i_0 = i) &= \sum_{m=0}^{k}W_m b_{i_m}\\
+    &= \sum_{m=0}^{k}w_{i,i_1}w_{i_1,i_2}\ldots
+    w_{i_{m-1},i_m}b_{i_m}\:.
+  \end{split}
+  \label{eq:estimator_xi}
+\end{equation}
+Here, the particle weight on the $m^\text{th}$ step is denoted $W_m$,
+and each particle starts with unit weight. When the particle
+transitions between states, for example $i\rightarrow j$, the weight
+is multiplied by the factor
+\begin{equation}
+  w_{ij} = \frac{h_{ij}}{p_{ij}}\:.
+  \label{eq:weight}
+\end{equation}
+Then, the expected value of $X$ is
+\begin{equation}
+  \begin{split}
+    E[X(i_0=i)] &= \sum_{\nu}P_\nu X_\nu\\
+    &= \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
+    \sum_{i_k}^{N}p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
+    w_{i,i_1}w_{i_1,i_2}\dots w_{i_{k-1},i_k}b_{i_k}\\
+    &= x_i\:,
+  \end{split}
+  \label{eq:expectation_xi}
+\end{equation}
+where $\nu$ denotes a particular random walk permutation.
+Therefore, the estimator in Eq.~(\ref{eq:estimator_xi}) is an unbiased
+estimator of the components of $\vx$ provided $\rho(\vH) < 1$.
+
+We are left to define the transition probabilities, $p_{ij}$. The most
+straightforward approach is to set
+\begin{equation}
+  p_{ij} = \frac{|h_{ij}|}{\sum_{j}|h_{ij}|}\:.
+  \label{eq:probability}
+\end{equation}
+Each row of the transition probability matrix, $\vP$, represents a
+discrete probability density function for selecting a new state $j$,
+given that the current state is $i$.  Random walks can be terminated
+in two ways: the matrix can be augmented with a terminating event
+equation that describes the probability that a particle ends its walk,
+or the random walk can be terminated by weight cutoff.  Generally, we
+choose to terminate random walks using a weight cutoff as opposed to
+augmenting the matrix.
+
+\subsection{Adjoint Method}
+\label{sec:adjoint-method}
+
+An alternative approach to the one just described is to calculate
+contributions to every component of $\vx$ during the random walk.  In
+the mathematical literature this is referred to as an adjoint method;
+however, to transport practitioners this approach more closely
+resembles a traditional forward method.  In this method, the weight
+change from state $i\rightarrow j$ is
+\begin{equation}
+  w_{ij} = \frac{h_{ji}}{p_{ij}}\:.
+  \label{eq:adjoint-weight}
+\end{equation}
+The transition probabilities may be calculated as
+\begin{equation}
+  p_{ij} = \frac{|h_{ji}|}{\sum_{j}|h_{ji}|}\:.
+  \label{eq:adjoint-probability}
+\end{equation}
+Note that the indices are reversed so that the probabilities are
+normalized over a column, as opposed to the forward method in which
+the probabilities are normalized over a row.  This is equivalent to
+forming the Neumann series in reverse order.  Correspondingly, the
+estimator for this method is
+\begin{equation}
+  \begin{split}
+    X &= \sum_{m=0}^{k}W_m\delta_{i_m,i}\\
+    &= \sum_{m=0}^{k}\hat{b}_{i_0}w_{i_0,i_1}w_{i_1,i_2}\ldots
+                                  w_{i_{m-1},i_m}\delta_{i_m,i}\:.
+  \end{split}
+  \label{eq:adjoint-tally}
+\end{equation}
+Here, $\hat{b}_{i_0}$ is the sampled source in cell/state $i_0$.  The
+Kronecker delta implies that tallies are only made in the cell/state
+where the random walk currently resides.  This is the common approach
+in standard Monte Carlo transport simulations.  We will investigate
+this approach as applied to the 1T thermal radiation diffusion
+problem in \S~\ref{sec:solut-radi-diff}.
+
+Similar to the Direct Method, the Adjoint Method random walk process
+requires a terminating condition.  In all of the work that follows we
+utilize a relative weight cutoff.  The weight cutoff is defined as a
+fraction of the starting weight such that
+\begin{equation}
+  W_f = W_c\hat{b}_{i_0}\:,
+  \label{eq:weight_cutoff}
+\end{equation}
+where $W_f$ is the terminating weight of the random walk, $W_c$ is the
+input relative weight cutoff, and $\hat{b}_{i_0}$ is the initial
+weight of the source particle.  The random walk ends on the
+$m^\text{th}$ step if $W_m < W_f$.
+
+%%---------------------------------------------------------------------------%%
+\section{Synthetic-Acceleration Monte Carlo}
+\label{sec:iter-refin-monte}
+
+The methods presented in \S~\ref{sec:monte-carlo-matrix} are
+characterized by slow convergence rates.
+Halton~\cite{halton_1962,halton_1994} proposed a staged residual
+scheme called Sequential Monte Carlo to speed up the convergence of
+these methods.  A variant of this scheme has been successfully applied
+to the 1D nonlinear thermal radiation diffusion equation in
+Ref.~\cite{evans_2003}.
+
+Concisely, the Sequential Monte Carlo method solves
+Eq.~(\ref{eq:Ax=b}) using the adjoint solution technique described in
+\S~\ref{sec:adjoint-method}.  The following iteration scheme is
+applied:
+\begin{subequations}
+  \begin{gather}
+    \vr^{l+1} = \vb - \vA\vx^{l}\:, \\
+    \vA\delta\vx^{l+1} = \vr^{l+1}\:, \label{eq:residual_solve}\\
+    \vx^{l+1} = \vx^{l} + \delta\vx^{l+1}\:.
+  \end{gather}
+\end{subequations}
+In this scheme, the adjoint method is used to estimate the solution to
+Eq.~(\ref{eq:residual_solve}), and the residual is iterated to
+convergence.  This iteration sequence is closely related to iterative
+refinement; the exception being that there is no update to to
+$x^{l+1}$ between residual iterations.
+
+We propose a modification of this scheme that uses Monte Carlo as a
+synthetic-acceleration scheme for fixed-point iteration. The
+Fixed-Point Monte Carlo Synthetic-Acceleration (MCSA) method is
+defined as follows:
+\begin{subequations}
+  \begin{gather}
+    \vx^{l+1/2} = (\vI - \vA)\vx^l + \vb\:, \\
+    %%
+    \vr^{l+1/2} = \vb - \vA\vx^{l+1/2}\:,\\
+    %%
+    \hat{\vA}\delta\vx^{l+1/2} = \vr^{l+1/2}\:,
+    \label{eq:MCSA-residual_solve}\\ 
+    %%
+    \vx^{l+1}=\vx^{l+1/2}+\delta\vx^{l+1/2}\:.
+  \end{gather}
+  \label{eq:general-MCSA}
+\end{subequations}
+The hat on $\vA$ in Eq.~(\ref{eq:MCSA-residual_solve}) indicates that
+the Monte Carlo solution only approximately inverts this operator.
+However, $\hat{\vA}$ is the same operator as $\vA$;
+$\hat{\vA}\equiv\vA$.  Thus, we have defined a scheme in which the
+initial estimate of $\vx$ in each iteration is updated using a single
+fixed-point iteration\footnote{Although, a Krylov method could be used
+  here.}.  The residual is calculated and is used as a source to
+estimate the error, $\delta\vx^{l+1/2}$, by solving
+Eq.~(\ref{eq:MCSA-residual_solve}) via the Monte Carlo adjoint method.
+The error is used to calculate an updated iterate of the solution
+vector, $\vx^{l+1}$.  The entire sequence is iterated to convergence
+based on the following stopping criterion~\cite{kelley_1995},
+\begin{equation}
+  \|\vr\|_{\infty} < \epsilon\cdot\|\vb\|_\infty\:.
+  \label{eq:stopping-criteria}
+\end{equation}
+
+%%---------------------------------------------------------------------------%%
+\section{Discrete Form of the Radiation Diffusion Equation}
+\label{sec:discr-form-radi}
+
+\subsection{Equilibrium Diffusion Model}
+
+The general form of the equilibrium diffusion equation is
+\cite{morel_1996}
+\begin{equation} 
+  (\rho\Cv + 4aT^{3})\pder{T}{t} -
+  \nabla\cdot{\Bigl(\frac{4acT^{3}}{3\ros}\Bigr)\grad{T}} = Q\:,
+  \label{eq:eq-diff-T}
+\end{equation}
+where $\Cv\equiv\Cv(\ve{r},T(\ve{r},t))$
+$[\text{Jerks}\cdot\text{g}^{-1}\cdot\text{keV}^{-1}]$ is the specific
+heat capacity of the material, $\rho\equiv\rho(\ve{r})$
+$[\text{g}\cdot\text{cm}^{-3}]$ is the density of the material,
+$a=0.01372$ $[\text{Jerks}\cdot\text{cm}^{-3}\cdot\text{keV}^{-4}]$ is
+the radiation constant, $c=299.79$ $[\text{cm}\cdot\text{sh}^{-1}]$ is
+the vacuum light speed, and $T\equiv T(\ve{r},t)$ $[\text{keV}]$ is
+the temperature that characterizes both the radiation and the
+material.  The opacity, $\ros\equiv\ros$(\ve{r},T(\ve{r},t))
+$[\text{cm}^{-1}]$, is the Rosseland Mean opacity.  The source,
+$Q\equiv Q(\ve{r},t)$
+$[\text{Jerks}\cdot\text{cm}^{-3}\cdot\text{sh}^{-1}]$, is assumed to
+be independent of $T$.
+
+We chose to use the angle-energy integrated radiation intensity,
+$\phi\equiv\phi(\ve{r},t)$
+$[\text{Jerks}\cdot\text{cm}^{-2}\cdot\text{sh}^{-1}]$, as the primary
+state variable.  The intensity is defined,
+\begin{equation}
+  \label{eq:lte-phi}
+  \begin{split}
+    \phi(\ve{r},t) &=
+    \int_{4\pi}d\bOmega\int_{\nu}d\nu\:\psi(\ve{r},\nu,\hat{\bOmega},t) \\
+    &= \int_{\nu}d\nu\: 4\pi B(\nu,T) \\
+    &= acT^{4}\:,
+  \end{split}
+\end{equation}
+where $\psi$ is the angular radiation intensity.  The Planck function
+(or Planckian) is defined as,
+\begin{equation}
+  B(\nu,T)=\frac{2h\nu^{3}}{c^{2}}\bigl(e^{\frac{h\nu}{kT}} -
+  1\bigr)^{-1}\:,
+\end{equation}
+where $h$ is Planck's constant and $k$ is the Boltzmann constant.
+Utilizing this definition of $\phi$ we can write
+\begin{equation}
+  \label{eq:partial-T}
+  \partial T = \frac{1}{4acT^{3}}\partial\phi\:.
+\end{equation}
+Inserting \eqref{eq:partial-T} into \eqref{eq:eq-diff-T} yields
+\begin{equation}
+  \label{eq:eq-diff}
+  \Bigl(\frac{\Cv}{4acT^{3}} + \frac{1}{c}\Bigr)\pder{\phi}{t} - 
+  \nabla\cdot D\grad{\phi} = Q\:,
+\end{equation}
+where $D$ is the diffusion coefficient, $D\equiv
+D(\ve{r},T(\ve{r},t))\ [\text{cm}]$, and is defined
+\begin{equation}
+  D = \frac{1}{3\ros}\:.
+\end{equation}
+Equation~\eqref{eq:eq-diff} is the model equation we will solve using
+the Monte Carlo techniques presented in
+\S~\ref{sec:monte-carlo-matrix} and \ref{sec:iter-refin-monte}.
+
+\subsection{Discrete Diffusion Equation}
+
+Applying backward-Euler time differencing to Eq.~(\ref{eq:eq-diff})
+and lagging the temperature-dependent coefficients at $t^n$ gives
+\begin{equation}
+  -\nabla\cdot \Dn\grad{\phi} + \sign\phi = \qn\:,
+\end{equation}
+where we have suppressed the $n+1$ indices on $\phi$ and 
+\begin{align}
+  \sign &= \frac{\rho \Cv^n}{4ac(\Tn)^3\dt} + \frac{1}{c\dt}\:,\\
+  \qn &= \sign\phin + Q^n\:,\\
+  \phin &= ac(\Tn)^4\:.
+\end{align}
+Applying Fick's Law,
+\begin{equation}
+  \ve{F} = -D\grad{\phi}\:,
+  \label{eq:fick}
+\end{equation}
+where $\ve{F}$ is the radiation flux, we define a flux-balance
+equation,
+\begin{equation}
+  \dive{F} + \sign\phi = \qn\:.
+  \label{eq:flux-balance}
+\end{equation}
+Integrating Eq.~(\ref{eq:flux-balance}) over the cell illustrated in
+Fig.~\ref{fig:cell} gives
+\begin{equation}
+  \iiint\dive{F}\,d\ve{r} +
+  \sign_{ijk}\phi_{ijk}V_{ijk}=\qn_{ijk}V_{ijk}\:,
+  \label{eq:flux-balance-vol-integral}
+\end{equation}
+and $V_{ijk} = \Delta_i\Delta_j\Delta_k$.  We now apply the divergence
+theorem,
+\begin{equation}
+  \begin{split}
+    \iiint\dive{F}\,d\ve{r} &= \oint\hat{\ve{n}}\cdot\ve{F}\,dA\\
+    &= \sum_{f=1}^{6}\hat{\ve{n}}_f\cdot\ve{F}_f A_f\:,
+  \end{split}
+\end{equation}
+to Eq.(\ref{eq:flux-balance-vol-integral}) yielding
+\begin{equation}
+  \begin{split}
+    (F_{i+1/2} - F_{i-1/2})\Dj\Dk + (F_{j+1/2} &- F_{j-1/2})\Di\Dk \\
+    &+ (F_{k+1/2} - F_{k-1/2})\Di\Dj
+    + \sign_{ijk}\phi_{ijk}V_{ijk} = \qn_{ijk}V_{ijk}\:.
+  \end{split}
+  \label{eq:flux-balance-difference}
+\end{equation}
+
+Using Fick's Law, Eq.~(\ref{eq:fick}), we can write $O(\Delta^2)$
+expressions for the face-centered fluxes.  Defining $l\in\{i,j,k\}$,
+the face-centered fluxes are
+\begin{align}
+  F_{l+1/2} &= -D_{l+1/2}\frac{\phi_{l+1} -
+    \phi_{l}}{\Delta_{l+1/2}}\:, \label{eq:F_l+1/2}\\
+  F_{l-1/2} &= -D_{l-1/2}\frac{\phi_{l} -
+    \phi_{l-1}}{\Delta_{l-1/2}}\:, \label{eq:F_l-1/2}
+\end{align}
+where
+\begin{align}
+  \Delta_{l+1/2} &= \frac{\Delta_{l+1} + \Delta_{l}}{2}\:,\\
+  \Delta_{l-1/2} &= \frac{\Delta_{l} + \Delta_{l-1}}{2}\:.
+\end{align}
+In order to be conservative, the flux must be continuous across a
+face.  Thus, the face-centered flux evaluated from either side of the
+face must be equivalent,
+\begin{equation}
+  \begin{aligned}
+    F_{l+1/2}^{+} &= F_{l+1/2}^{-}\:,\\
+    2D_{l+1}\frac{\phi_{l+1} - \phi_{l+1/2}}{\Delta_{l+1}} &=
+    2D_{l}\frac{\phi_{l+1/2} - \phi_{l}}{\Delta_l}\:,\\
+    \frac{1}{3\sigma_{l+1}}\frac{\phi_{l+1} - \phi_{l+1/2}}{\Delta_{l+1}} &=
+    \frac{1}{3\sigma_{l}}\frac{\phi_{l+1/2} - \phi_{l}}{\Delta_l}\:,
+  \end{aligned}
+\end{equation}
+where we have dropped the subscript $\mathrm{R}$ from $\sigma$ for
+clarity.  Solving this system for $\phi_{l+1/2}$ and plugging into
+either $F_{l+1/2}^{+}$ or $F_{l+1/2}^{-}$ and setting the resulting
+expression equal to Eq.~(\ref{eq:F_l+1/2}), one finds the
+face-centered diffusion coefficient,
+\begin{equation}
+  D_{l+1/2} = \frac{2\Delta_{l+1/2}}{3\sigma_l\Delta_l + 
+    3\sigma_{l+1}\Delta_{l+1}}\:.
+  \label{eq:D_l+1/2}
+\end{equation}
+Following the same procedure at the $l-1/2$ face gives the diffusion
+coefficient on the low-side face,
+\begin{equation}
+  D_{l-1/2} = \frac{2\Delta_{l-1/2}}{3\sigma_l\Delta_l + 
+    3\sigma_{l-1}\Delta_{l-1}}\:.
+  \label{eq:D_l-1/2} 
+\end{equation}
+Using Eqs.~(\ref{eq:F_l+1/2}), (\ref{eq:F_l-1/2}), (\ref{eq:D_l+1/2}),
+and (\ref{eq:D_l-1/2}) in the discrete flux-balance equation,
+(\ref{eq:flux-balance-difference}), we obtain
+\begin{equation}
+  \begin{aligned}
+    \sigT\phi_{ijk} 
+    - \sigm_{i+1}\frac{\Delta_{i+1}}{\Di}\phi_{i+1\,jk} &
+    - \sigp_{i-1}\frac{\Delta_{i-1}}{\Di}\phi_{i-1\,jk}
+    - \sigm_{j+1}\frac{\Delta_{j+1}}{\Dj}\phi_{i\,j+1\,k} \\
+    &- \sigp_{j-1}\frac{\Delta_{j-1}}{\Dj}\phi_{i\,j-1\,k} 
+    - \sigm_{k+1}\frac{\Delta_{k+1}}{\Dk}\phi_{ij\,k+1}
+    - \sigp_{k-1}\frac{\Delta_{k-1}}{\Dk}\phi_{ij\,k-1}
+    = \qn_{ijk}\:.
+  \end{aligned}
+  \label{eq:discrete-diffusion}
+\end{equation}
+Here,
+\begin{equation}
+  \sigT = \sigp_i + \sigm_i + \sigp_j + \sigm_j + \sigp_k + \sigm_k
+  + \sign_{ijk}\:,
+\end{equation}
+and
+\begin{subequations}
+  \begin{align}
+    \sigp_{i} &= \frac{1}{\Di}\frac{2}{3\sigma_{ijk}\Di
+      + 3\sigma_{i+1\,jk}\Delta_{i+1}}\:,\\
+    \sigm_{i} &= \frac{1}{\Di}\frac{2}{3\sigma_{ijk}\Di
+      + 3\sigma_{i-1\,jk}\Delta_{i-1}}\:,\\
+    \sigp_{j} &= \frac{1}{\Dj}\frac{2}{3\sigma_{ijk}\Dj
+      + 3\sigma_{i\,j+1\,k}\Delta_{j+1}}\:,\\
+    \sigm_{j} &= \frac{1}{\Dj}\frac{2}{3\sigma_{ijk}\Dj
+      + 3\sigma_{i\,j-1\,k}\Delta_{j-1}}\:,\\
+    \sigp_{k} &= \frac{1}{\Dk}\frac{2}{3\sigma_{ijk}\Dk
+      + 3\sigma_{ij\,k+1}\Delta_{k+1}}\:,\\
+    \sigm_{k} &= \frac{1}{\Dk}\frac{2}{3\sigma_{ijk}\Dk
+      + 3\sigma_{ij\,k-1}\Delta_{k-1}}\:. 
+  \end{align}
+  \label{eq:sigmas}
+\end{subequations}
+Also, note that
+\begin{align*}
+  \sigp_{i-1} &= \frac{1}{\Delta_{i-1}}\frac{2}{3\sigma_{ijk}\Di
+    + 3\sigma_{i-1\,jk}\Delta_{i-1}}\:,\\
+  \sigm_{i+1} &= \frac{1}{\Delta_{i+1}}\frac{2}{3\sigma_{ijk}\Di
+    + 3\sigma_{i+1\,jk}\Delta_{i+1}}\:,\\
+  \ldots\:.
+\end{align*}
+Equations~(\ref{eq:discrete-diffusion}) through (\ref{eq:sigmas})
+represent the discretized, 1T diffusion equation on an orthogonal
+3D grid.
+
+Equations~(\ref{eq:discrete-diffusion}) through (\ref{eq:sigmas}) have
+been written in a manner that is amenable to Monte Carlo
+interpretation.  One can envision the terms in Eq.~(\ref{eq:sigmas})
+representing leakage out of each face of a cell.  The off-diagonal
+terms then represent sources into the cell, or, equivalently, these
+terms are leakage out of the neighboring cells.  A limitation of this
+interpretation is that it requires a symmetric system to implement the
+Monte Carlo scheme.  More specifically, the operator $\ve{D}$ must be
+an H-matrix \cite{kelley_1995} such that the off-diagonal elements are
+negative and symmetric.  In other words, the leakage out of a cell
+must be equivalent to the source into the adjacent cell.
+
+The methods described in \S\S~\ref{sec:monte-carlo-matrix} and
+\ref{sec:iter-refin-monte} do not suffer this limitation.  Thus, we
+will interpret the solution via MCSA as a transport-like process, but
+we do not require any constraints on the system other than the
+spectral radius must be less than 1.
+
+Before completing the derivation of discrete equations required to
+solve Eq.~(\ref{eq:discrete-diffusion}), we must specify boundary
+conditions.  For this work, the Marshak Boundary condition suffices,
+\begin{equation}
+  f_b^\pm = \frac{1}{4}\phi - \frac{1}{2}\ve{F}\cdot\hat{\ve{n}}\:,\quad
+  \vr\in \partial\Gamma\:,
+\end{equation}
+where $f_b$ is the incoming partial current on the problem boundary,
+$\partial\Gamma$.  On any low boundary we can write the flux using a
+first-order discrete form of Fick's Law,
+\begin{equation}
+  F_{1/2} = -\frac{2}{3\sigma_1}\frac{\phi_1 - \phi_{1/2}}{\Delta_1}\:.
+  \label{eq:low-boundary-flux}
+\end{equation}
+As before, we enforce continuity of flux on the boundary using the
+Marshak Boundary condition,
+\begin{equation}
+  -\frac{2}{3\sigma_1}\frac{\phi_1 - \phi_{1/2}}{\Delta_1} = 2f_b^{+}
+  -\frac{1}{2}\phi_{1/2}\:. 
+\end{equation}
+Solving for $\phi_{1/2}$ and plugging into
+Eq.~(\ref{eq:low-boundary-flux}) yields an expression for the low-side
+boundary flux,
+\begin{equation}
+  F_{1/2} = \frac{2}{3\sigma_1\Delta_1 + 4}(4f_b^{+}-\phi_1)\:.
+\end{equation}
+Following an identical procedure on high-side problem boundaries gives
+\begin{equation}
+  F_{N_l+1/2} = \frac{2}{3\sigma_{N_l}\Delta_{N_l} +
+    4}(\phi_{N_l}-4f_b^{-})\:, 
+\end{equation}
+where $l\in\{i,j,k\}$ is defined over the range $[1,N_l]$.
+
+These boundary fluxes take the place of the interior face-centered
+fluxes (defined in Eqs.~(\ref{eq:F_l+1/2}) and (\ref{eq:F_l-1/2})) in
+Eq.~(\ref{eq:flux-balance-difference}) yielding the following
+modifications to Eqs.~(\ref{eq:sigmas}) on a boundary,
+\begin{align}
+  \text{low boundary:}\quad & \sigma_1^{-} = \frac{1}{\Delta_1}
+  \frac{2}{3\sigma_1\Delta_1 + 4}\:,\\
+  \text{high boundary:}\quad & \sigma_{N_l}^{+} = \frac{1}{\Delta_{N_l}}
+  \frac{2}{3\sigma_{N_l}\Delta_{N_l} + 4}\:.
+\end{align}
+Furthermore, the following contributions are added to the source,
+$\qn_{ijk}$ on a boundary cell,
+\begin{align}
+  \text{low boundary:}\quad & \frac{8}{\Delta_1}
+  \frac{1}{3\sigma_1\Delta_1 + 4}f_b^{+}\:,\\
+  \text{high boundary:}\quad & \frac{8}{\Delta_{N_l}}
+  \frac{1}{3\sigma_{N_l}\Delta_{N_l} + 4}f_b^{-}\:.
+\end{align}
+Note that for vacuum boundary conditions, $f_b^{\pm} = 0$.
+
+For reflective boundary conditions we have the simple modification
+that
+\begin{align}
+  F_{1/2} &= 0\:, \\
+  F_{N_l+1/2} &= 0\:.
+\end{align}
+Thus, there are no leakage or adjacent cell source terms resulting
+from this boundary face.
+
+\subsection{Preconditioning}
+\label{sec:preconditioning}
+
+The discrete diffusion equation in (\ref{eq:discrete-diffusion}) can
+be written in matrix-operator form as follows,
+\begin{equation}
+  \ve{D}\bphi = \ve{q}\:,
+  \label{eq:diffusion-matrix}
+\end{equation}
+where $\ve{D}$ is a symmetric, positive-definite (SPD) operator.
+Unfortunately, for many problems $\rho(\ve{D}) > 1$.  Therefore, we
+must precondition Eq.~(\ref{eq:diffusion-matrix}) in order to use the
+MCSA method.  Preconditioning accomplishes two purposes: it reduces
+the spectral radius of $\ve{D}$ so that Monte Carlo solution methods
+are applicable, and it reduces the number of iterations required for
+convergence.  We apply left-preconditioning and write
+\begin{equation}
+  \ve{M}^{-1}\ve{D}\bphi = \ve{M}^{-1}\ve{q}\:.
+  \label{eq:preconditioned-diffusion}
+\end{equation}
+A good choice for the preconditioner, $\ve{M}$, is
+\begin{equation}
+  \ve{M}=\diag(\ve{D})\:.
+  \label{eq:preconditioner}
+\end{equation}
+Setting $\ve{M}$ equal to the diagonal elements of $\ve{D}$
+accomplishes several objectives:
+\begin{itemize}
+\item It reduces the spectral radius such that
+  $\rho(\ve{M}^{-1}\ve{D}) < \rho(\ve{D})$ and
+  $\rho(\ve{M}^{-1}\ve{D}) < 1$.  Thus, convergence is guaranteed and
+  fewer iterations are required resulting in accelerated convergence.
+\item Makes the diagonal elements of the iteration matrix zero,
+  $\diag(\vI - \ve{M}^{-1}\ve{D}) = 0$.  Thus, there will be no time
+  wasted in transitions $i\rightarrow i$ during the random walk
+  process.
+\end{itemize}
+The second item is quite important for Monte Carlo solutions of linear
+systems, and this type of scaling should be performed regardless of
+any additional preconditioner choices.
+
+A side effect of this preconditioner choice is that the preconditioned
+system, $\ve{M}^{-1}\ve{D}$, is no longer SPD.  This situation can be
+rectified by doing right-left preconditioning and defining
+\begin{equation}
+  \ve{M} = \ve{L}\ve{L}^{T}\:,
+\end{equation}
+where $\ve{L} = \ve{L}^{T} = \sqrt{\ve{M}}$.  Applying right-left
+preconditioning gives
+\begin{equation}
+  \begin{gathered}
+    \ve{M}^{-1}\ve{D}\ve{M}^{-1}\ve{M}\bphi = \ve{M}^{-1}\ve{q}\:,\\
+    (\ve{LL}^T)^{-1}\ve{D}(\ve{LL}^T)^{-1}\ve{LL}^T\bphi =
+    (\ve{LL}^T)^{-1}\ve{q}\:, \\
+    \ve{L}^{-T}\ve{L}^{-1}\ve{D}\ve{L}^{-T}\ve{L}^T\phi =
+    \ve{L}^{-T}\ve{L}^{-1}\ve{q}\:,\\ 
+    (\ve{L}^{-1}\ve{D}\ve{L}^{-T})\ve{L}^T\bphi = \ve{L}^{-1}\ve{q}\:,\\
+    \hat{\ve{D}}\hat{\bphi} = \hat{\ve{q}}\:.
+  \end{gathered}
+  \label{eq:right-left-preconditioning}
+\end{equation}
+This type of preconditioning is also referred to as row-column
+scaling.  The resulting preconditioned matrix, $\hat{\ve{D}}$, is now
+SPD, and we can use Conjugate Gradient (CG) solvers.  Therefore, we
+will use this preconditioning scheme for deterministic solutions of
+Eq.~(\ref{eq:diffusion-matrix}).  However, given that the convergence
+properties of $\hat{\ve{D}}$ are similar to $\ve{M}^{-1}\ve{D}$ and
+the Monte Carlo solvers do not require SPD matrices, we will be
+content with solving Eq.~(\ref{eq:preconditioned-diffusion}) using
+MCSA.
+
+The preconditioned system we will solve is defined in
+Eq.~(\ref{eq:preconditioned-diffusion}).  Applying $\ve{M}^{-1}$ to
+Eq.~(\ref{eq:discrete-diffusion}) yields
+\begin{equation}
+  \begin{aligned}
+    \phi_{ijk} 
+    - \frac{\sigm_{i+1}}{\sigT}\frac{\Delta_{i+1}}{\Di}&\phi_{i+1\,jk}
+    - \frac{\sigp_{i-1}}{\sigT}\frac{\Delta_{i-1}}{\Di}\phi_{i-1\,jk}
+    - \frac{\sigm_{j+1}}{\sigT}\frac{\Delta_{j+1}}{\Dj}\phi_{i\,j+1\,k} \\
+    &- \frac{\sigp_{j-1}}{\sigT}\frac{\Delta_{j-1}}{\Dj}\phi_{i\,j-1\,k} 
+    - \frac{\sigm_{k+1}}{\sigT}\frac{\Delta_{k+1}}{\Dk}\phi_{ij\,k+1}
+    - \frac{\sigp_{k-1}}{\sigT}\frac{\Delta_{k-1}}{\Dk}\phi_{ij\,k-1}
+    = \frac{\qn_{ijk}}{\sigT}\:.
+  \end{aligned}
+  \label{eq:preconditioned-discrete-diffusion}
+\end{equation}
+In the next section we will derive an MCSA method to solve this
+equation.
+
+%%---------------------------------------------------------------------------%%
+\section{Solution of the Radiation Diffusion Equation}
+\label{sec:solut-radi-diff}
+
+Here we derive the equations required to solve
+Eq.~(\ref{eq:preconditioned-discrete-diffusion}) using the MCSA
+method.  As discussed in \S~\ref{sec:iter-refin-monte}, the MCSA
+method requires two solution steps:
+\begin{enumerate}
+\item A fixed-point iteration to estimate $\bphi^{l+1/2}$.  This is a
+  matrix-vector multiply operation.
+\item An adjoint Monte Carlo solve to estimate $\delta\bphi^{l+1/2}$.
+\end{enumerate}
+For clarity, we rewrite the iteration scheme in
+Eqs.~(\ref{eq:general-MCSA}) using the diffusion operator notation
+from Eq.~(\ref{eq:preconditioned-diffusion}),
+\begin{align*}
+  %%
+  \bphi^{l+1/2} = (\vI - \ve{M}^{-1}\ve{D})\bphi^{l} +
+  \ve{M}^{-1}\ve{q}\:, \quad & (\text{fixed-point iteration})\\
+  %%
+  \vr^{l+1/2} = \ve{M}^{-1}\ve{q} - \ve{M}^{-1}\ve{D}\bphi^{l+1/2}\:,
+  \quad & (\text{compute residual})\\
+  %%
+  \ve{M}^{-1}\ve{D}\delta\bphi^{l+1/2} = \vr^{l+1/2}\:, \quad&
+  (\text{estimate $\delta\phi$ using adjoint Monte Carlo method})\\
+  %%
+  \bphi^{l+1} = \bphi^{l+1/2} + \delta\bphi^{l+1/2}\:. \quad&
+  (\text{update}\ \bphi)
+\end{align*}
+As described in \S~\ref{sec:iter-refin-monte}, the adjoint Monte Carlo
+method only approximately inverts $\ve{M}^{-1}\ve{D}$.  We now
+describe each of these steps in turn.
+
+First, we use Eq.~(\ref{eq:preconditioned-discrete-diffusion}) to
+evaluate $\bphi^{l+1/2}$ via a single fixed-point iteration as
+follows:
+\begin{equation}
+  \begin{aligned}
+    \phi_{ijk}^{l+1/2} =  
+    \frac{\sigm_{i+1}}{\sigT}\frac{\Delta_{i+1}}{\Di}&\phi_{i+1\,jk}^l
+    + \frac{\sigp_{i-1}}{\sigT}\frac{\Delta_{i-1}}{\Di}\phi_{i-1\,jk}^l
+    +
+    \frac{\sigm_{j+1}}{\sigT}\frac{\Delta_{j+1}}{\Dj}\phi_{i\,j+1\,k}^l \\ 
+    &+ \frac{\sigp_{j-1}}{\sigT}\frac{\Delta_{j-1}}{\Dj}\phi_{i\,j-1\,k}^l 
+    + \frac{\sigm_{k+1}}{\sigT}\frac{\Delta_{k+1}}{\Dk}\phi_{ij\,k+1}^l
+    + \frac{\sigp_{k-1}}{\sigT}\frac{\Delta_{k-1}}{\Dk}\phi_{ij\,k-1}^l
+    + \frac{\qn_{ijk}}{\sigT}\:.
+  \end{aligned}
+\end{equation}
+This result is used to compute the residual of the system.  The
+residual then becomes the source for the Monte Carlo simulation.
+
+The adjoint Monte Carlo method described in
+\S~\ref{sec:adjoint-method} is used to estimate
+\begin{equation}
+  \ve{M}^{-1}\ve{D}\delta\bphi^{l+1/2} = \vr^{l+1/2}\:.
+\end{equation}
+As noted above, the residual acts as the source for this simulation,
+and the operator $\ve{M}^{-1}\ve{D}$ is implicitly defined in
+Eq.~(\ref{eq:preconditioned-discrete-diffusion}).  There are two Monte
+Carlo interpretations that can be applied to this system.  The first
+follows the mathematical presentation given in
+\S~\ref{sec:monte-carlo-matrix}.  Namely, we form probabilities from
+the iteration matrix and perform random walks to generate unbiased
+estimates of the solution vector, $\delta\bphi^{l+1/2}$, using the
+estimator in Eq.~(\ref{eq:adjoint-tally}).
+
+However, a more natural interpretation for Monte Carlo transport
+practitioners is to use the machinery described in
+\S~\ref{sec:monte-carlo-matrix} to define Probability Distribution
+Functions (PDF) that describe the transport of particles between
+cells.  We then do a Monte Carlo transport calculation using
+Eq.~(\ref{eq:adjoint-weight}) to calculate the weight change at each
+collision. Equation~(\ref{eq:adjoint-probability}) gives the
+probability of scattering to an adjacent cell. We tally the
+contribution of $\phi$ in each cell using
+Eq.~(\ref{eq:adjoint-tally}).  The transport is terminated using a
+relative weight cutoff that is defined in
+Eq.~(\ref{eq:weight_cutoff}).  We now examine each of these steps in
+more detail.
+
+The connectivity of an arbitrary cell is illustrated in the 3D mesh
+segment in Fig.~\ref{fig:mesh_segment}.  For a particle residing in
+the blue cell, any collision results in a scatter to one of the six
+adjacent cells since $p_{ii}=0$ in the preconditioned system.  We can
+easily calculate the probabilities for each scattering event from
+Eq.~(\ref{eq:preconditioned-discrete-diffusion}) without explicitly
+forming the iteration matrix.  In this mesh cell, the probabilities
+for leakage out of each face are derived from the equations from each
+adjacent cell.  Using Eq.~(\ref{eq:preconditioned-discrete-diffusion})
+we can explicitly write the equations from each adjacent cell,
+\begin{align*}
+  \phi_1 &= \frac{\sigm_{7\rightarrow
+      1}}{\sigma_{T\,1}}\frac{\Delta_7}{\Delta_1}\phi_7 + \ldots\:,\\
+  %%
+  \phi_2 &= \frac{\sigp_{7\rightarrow
+      2}}{\sigma_{T\,2}}\frac{\Delta_7}{\Delta_2}\phi_7 + \ldots\:,\\
+  %%
+  \phi_3 &= \frac{\sigm_{7\rightarrow
+      3}}{\sigma_{T\,3}}\frac{\Delta_7}{\Delta_3}\phi_7 + \ldots\:,\\
+  %%
+  \phi_4 &= \frac{\sigp_{7\rightarrow
+      4}}{\sigma_{T\,4}}\frac{\Delta_7}{\Delta_4}\phi_7 + \ldots\:,\\
+  %%
+  \phi_5 &= \frac{\sigm_{7\rightarrow
+      5}}{\sigma_{T\,5}}\frac{\Delta_7}{\Delta_5}\phi_7 + \ldots\:,\\
+  %%
+  \phi_6 &= \frac{\sigp_{7\rightarrow
+      6}}{\sigma_{T\,6}}\frac{\Delta_7}{\Delta_6}\phi_7 + \ldots\:,\\
+  %%
+  \ldots\:.
+\end{align*} 
+A quick view of these equations reveals that the only non-zero entries
+in column 7 of the iteration matrix are
+\begin{equation*}
+  \begin{pmatrix}
+    h_{1,7} &
+    h_{2,7} &
+    h_{3,7} &
+    h_{4,7} &
+    h_{5,7} &
+    h_{6,7} 
+  \end{pmatrix}^{T}
+  = 
+  \begin{pmatrix}
+    \frac{\sigm_{7\rightarrow 1}}{\sigma_{T\,1}}\frac{\Delta_7}{\Delta_1} \\
+    \frac{\sigp_{7\rightarrow 2}}{\sigma_{T\,2}}\frac{\Delta_7}{\Delta_2} \\
+    \frac{\sigm_{7\rightarrow 3}}{\sigma_{T\,3}}\frac{\Delta_7}{\Delta_3} \\
+    \frac{\sigp_{7\rightarrow 4}}{\sigma_{T\,4}}\frac{\Delta_7}{\Delta_4} \\
+    \frac{\sigm_{7\rightarrow 5}}{\sigma_{T\,5}}\frac{\Delta_7}{\Delta_5} \\
+    \frac{\sigp_{7\rightarrow 6}}{\sigma_{T\,6}}\frac{\Delta_7}{\Delta_6}
+  \end{pmatrix}\:,
+\end{equation*}
+where, following Eq.~(\ref{eq:point_iteration}), $\ve{H} = (\vI -
+\ve{M}^{-1}\ve{D})$ is the iteration matrix.  Now, using
+Eq.~(\ref{eq:adjoint-probability}), we define the scattering
+probabilities for cell 7 as
+\begin{equation*}
+  P_{7\rightarrow n} = p_{7,n} = \frac{|h_{n,7}|}{|h_{1,7} + 
+    h_{2,7} + h_{3,7} + h_{4,7} + h_{5,7} + h_{6,7}|}\:,\quad
+  n\in \{1,\ldots,6\}\:.
+\end{equation*}
+The weight change for each scattering event is calculated using
+Eq.~(\ref{eq:adjoint-weight}),
+\begin{equation*}
+  W_{m} = \frac{h_{n,7}}{p_{7,n}}W_{m-1}\:,\quad
+  n\in \{1,\ldots,6\}\:,
+\end{equation*}
+where $m$ is the step-index in the particle history.  We also note
+that, for the 3D Cartesian mesh considered in this work, the maximum
+number of non-zero entries in columns of the iteration matrix is 6.
+Cells along the problem boundary can have fewer than 6 entries.
+
+%%---------------------------------------------------------------------------%%
+\section{Results}
+\label{sec:results}
+
+Here we examine the effectiveness of the MCSA method and compare it to
+traditional deterministic solution methods.  The deterministic methods
+considered are preconditioned Conjugate Gradient (PCG) and
+preconditioned fixed-point (Richardson) iteration (PFIX).  The PCG
+method uses right-left preconditioning a shown in
+Eq.~(\ref{eq:right-left-preconditioning}).  The PFIX scheme is used to
+solve Eq.~(\ref{eq:preconditioned-diffusion}), which is identical to
+the system being solved with MCSA.  We examine each method on two 3D
+problems: a Marshak wave problem and a multi-material duct problem.
+
+\subsection{Marshak Wave}
+
+The Marshak wave problem under consideration in this study has an
+incoming 1~keV blackbody boundary flux at $x=0$ and propagates
+in the $x$-dimension in time.  In other words, this a 3D
+representation of a 1D problem.  The full set of problem conditions
+is:
+\begin{center}
+  \begin{tabular}{ll}\hline
+    %%
+    B.C. & $f_b^{+}(x=0,T=1.0) = \frac{ac}{4}T^4$ \\
+    & $\ve{F} = 0\quad \partial\Gamma \in \{x^{+}, y^{-}, y^{+},
+    z^{-}, z^{+}\}$ \\
+    %%
+    material properties & $\rho = 3.0\,\text{g}\cdot\text{cm}^{-3}$ \\
+    & $\Cv = 0.1\,\text{Jerks}\cdot\text{g}^{-1}\cdot\text{keV}^{-1}$\\
+    & $T(t=0) = \sn{1}{-4}\,\text{keV}$\\
+    %% 
+    opacity & $\sigma = 100T^{-3}\,\text{cm}^2\cdot\text{g}^{-1}$ \\
+    %%
+    mesh & $\Di = \Dj = \Dk = 0.01\,\text{cm}$ \\
+    & $N_i = 100\quad N_j = N_k = 4$ \\
+    \hline
+  \end{tabular}
+\end{center}
+The problem is run to $10$~shakes.
+
+Figure~\ref{fig:marshak_10} shows the analytic Marshak Wave solution
+compared to the MCSA solution.  As opposed to most Monte Carlo
+techniques, the MCSA method generates solutions that are numerically
+equivalent to the PCG and PFIX results.  Because the MCSA method
+converges the residual of the system, the numerical accuracy is
+determined by the stopping criterion shown in
+Eq.~(\ref{eq:stopping-criteria}).  Also, the MCSA method is not bound
+by the Central Limit Theorem so convergence is not restricted to
+$1/\sqrt{N}$~\cite{halton_1994,evans_2003}.
+
+As stated in \S~\ref{sec:monte-carlo-matrix}, the MCSA method will
+converge when $\rho(\ve{H})< 1$ where $\ve{H}$ is the iteration matrix
+and is defined $\ve{H} = (\vI - \ve{M}^{-1}\ve{D})$.  The spectral
+radius for the Marshak problem is plotted in
+Fig.~\ref{fig:marshak_spectral}.  As stated in
+\S~\ref{sec:preconditioning} the preconditioning effectively reduces
+the spectral radius of $\ve{H}$ to less than 1.
+
+There are two principal degrees of freedom when utilizing the MCSA
+method: (1) the number of particles per stage, and (2) the weight
+cutoff.  Figure~\ref{fig:CPU_Np} shows the CPU time as a function of
+the requested number of particles per stage for a relative weight
+cutoff of $\sn{1}{-4}$.  For less than 5 particles per stage teh
+method does not converge.  Figure~\ref{fig:CPU_Ns} shows the variation
+in the maximum number of stages (iterations) per cycle with number of
+particles.  Analysis of these two figures shows that increasing the
+number of particles per stage reduces the number of iterations
+required to converge the solution.  However, the cost of the Monte
+Carlo transport is high; thus, better performance is attained by
+running more stages with fewer particles per stage \cite{evans_2003}.
+
+Figure~\ref{fig:CPU_wc} shows the CPU time as a function of relative
+weight cutoff.  These results indicate that the performance of the
+MCSA method is relatively insensitive to the weight cutoff.  This is
+especially true when running small numbers of particles per stage.
+When larger numbers of particles are used the weight cutoff has a more
+pronounced effect on the efficiency of the method.
+
+Having analyzed the variation of the MCSA algorithm efficiency with
+number of particles and weight cutoff, we now compare to traditional
+deterministic algorithms.  Table~\ref{tab:marshak_comparision} shows
+timing results for the Marshak problem using preconditioned CG (PCG),
+preconditioned Richardson iteration (PFIX), and MCSA. The results show
+that the MCSA method is the most efficient method, although the PCG
+results are nearly as fast.  More importantly, the MCSA method clearly
+accelerates Richardson iteration, as indicated by the fact that MCSA
+is 42\% faster than Richards iteration alone.
+
+\subsection{Multi-material Problem}
+
+We now test the performance of the MCSA method on a large
+multi-material problem.  The multi-material problem has a dog-legged
+duct through a thick wall where the radiation flows into a thin region
+bound by a foil on one side.  The mesh and geometry are illustrated in
+Fig.~\ref{fig:multi-mat-mesh}.  A blackbody boundary flux is defined
+on the low-$x$ face at $T=0.5$~keV.  The full set of problem
+conditions is:
+\begin{center}
+  \begin{tabular}{ll}\hline
+    %%
+    B.C. & $f_b^{+}(x=0,T=0.5) = \frac{ac}{4}T^4$\\
+    & $\ve{F} = 0\quad \partial\Gamma \in \{x^{+}, y^{-}, y^{+},
+    z^{-}, z^{+}\}$ \\
+    %%
+    material properties 
+    & $\rho = 1.5\,\text{g}\cdot\text{cm}^{-3}$ (duct) \\
+    & $\rho = 8.0\,\text{g}\cdot\text{cm}^{-3}$ (wall) \\
+    & $\rho = 4.0\,\text{g}\cdot\text{cm}^{-3}$ (foil) \\
+    & $\Cv = 0.1\,\text{Jerks}\cdot\text{g}^{-1}\cdot\text{keV}^{-1}$\\
+    & $T(t=0) = \sn{1}{-4}\,\text{keV}$\\
+    %% 
+    opacity 
+    & $\sigma = T^{-3}\,\text{cm}^2\cdot\text{g}^{-1}$ (duct) \\
+    & $\sigma = 1000T^{-1}\,\text{cm}^2\cdot\text{g}^{-1}$ (wall) \\
+    & $\sigma = 5T^{-2}\,\text{cm}^2\cdot\text{g}^{-1}$ (foil) \\
+    %%
+    mesh & $\Di = \Dj = \Dk = 0.025\,\text{cm}$ \\
+    & $N_i =  N_j = N_k = 60$ \\
+    \hline
+  \end{tabular}
+\end{center}
+A 2D plot of the temperature at 100~sh elapsed time is illustrated in
+Fig.~\ref{fig:multi-mat_T}.  The time-evolution of the temperature is
+shown at 4 edit points in Fig.~\ref{fig:edit-points}.  All three
+methods, PCG, PFIX, and MCSA, give numerically identical answers when
+using a stopping criterion of $\sn{1}{-8}$.
+
+Table~\ref{tab:multimat_comparision} shows the timing results for the
+multi-material problem using PCG, PFIX, and MCSA.  The results
+correspond roughly with the results from the Marshak problem.  The
+MCSA is marginally faster than PCG and 43\% faster than PFIX.
+
+%%---------------------------------------------------------------------------%%
+\section{Conclusions}
+\label{sec:conclusions}
+
+We have presented a new Monte Carlo solution method for solving the
+discrete, time-dependent diffusion equation in 3D.  The MCSA method
+has been shown to match results using standard solution techniques to
+arbitrary precision.  Also, the new method is faster than
+preconditioned CG and Richardson iteration.
+
+While we have demonstrated marginal improvements over standard
+solution schemes in this study, significant improvements could be
+realized in fully nonlinear-consistent solutions.  In these cases, the
+MCSA method competes with GMRES, which is more costly in memory and
+time than CG.  Another area where the MCSA scheme may have advantages
+over traditional solution methods is on dentritic, or adaptive,
+meshes.  These meshes yield matrices with poor condition numbers
+because of the changing cell volumes at different refinement levels.
+A smart transport algorithm could be developed that more efficiently
+solves the residual on these types of meshes.  These topics will be
+the focus of future study.
+
+%%---------------------------------------------------------------------------%%
 %% The Appendices part is started with the command \appendix;

### 2012-05-26 doc/jcomp2012/jcomp2012.tex
Added nse07 figures to tex

@@
 %% The amsthm package provides extended theorem environments
-%% \usepackage{amsthm}
+\usepackage{amsthm}
+\usepackage{amsmath}
+\usepackage{tmadd,tmath}
+\usepackage[mathcal]{euscript}
 
@@
 
+\begin{figure}[htpb!]
+  \centerline{
+    \input{3D_mesh_segment.pstex_t}}
+  \caption{3D mesh-segment.  The particle's current cell (in blue) is
+    connected to 6 adjacent cells.  The cell numbers are plotted
+    alongside the mesh such that the blue cell has index 7, the high
+    $z$ cell has index 2, etc.}
+  \label{fig:mesh_segment}
+\end{figure}
+
 %%---------------------------------------------------------------------------%%
@@
 
+\begin{figure}[htpb!]
+  \centerline{
+    \includegraphics[width=5in,clip]{marshak_10.eps}}
+  \caption{
+    Solution of the Marshak wave problem at $t=10$~sh.  The analytic
+    solution is compared to the MCSA solution.  Note that the MCSA,
+    PCG, and PFIXED methods all give numerically identical answers.}
+  \label{fig:marshak_10}
+\end{figure}
+
 As stated in \S~\ref{sec:monte-carlo-matrix}, the MCSA method will
@@
 the spectral radius of $\ve{H}$ to less than 1.
+
+\begin{figure}[htpb!]
+  \centerline{
+    \includegraphics[width=5in,clip]{spectral_radius.eps}}
+  \caption{Spectral radius of the iteration matrix, $\ve{H}$, for the
+    Marshak Wave problem as a function of elapsed problem time.}
+  \label{fig:marshak_spectral}
+\end{figure}
 
@@
 
+\begin{figure}[htpb!]
+  \centerline{
+    \includegraphics[width=5in,clip]{mrshk_np_CPU.eps}}
+  \caption{CPU time versus number of particles per stage for the
+    Marshak problem.  The number of particles is the requested number.
+    In each stage the exact number can vary depending upon the number
+    of cells and source strength per cell.}
+  \label{fig:CPU_Np}
+\end{figure}
+
+\begin{figure}[htpb!]
+  \centerline{
+    \includegraphics[width=5in,clip]{mrshk_np_Ns.eps}}
+  \caption{
+    Maximum number of stages per cycle versus number of particles per
+    stage for the Marshak problem.  The number of particles is the
+    requested number.  In each stage the exact number can vary
+    depending upon the number of cells and source strength per cell.}
+  \label{fig:CPU_Ns}
+\end{figure}
+
 Figure~\ref{fig:CPU_wc} shows the CPU time as a function of relative
@@
 pronounced effect on the efficiency of the method.
+
+\begin{figure}[htpb!]
+  \centerline{
+    \includegraphics[width=5in,clip]{mrshk_wc_CPU.eps}}
+  \caption{
+    CPU time versus relative weight cutoff for the Marshak problem.
+    Results are shown for 10 and 50 requested particles per stage.
+    The numbers in parenthesis show the maximum number of particles
+    run in a stage during the course of the problem.}
+  \label{fig:CPU_wc}
+\end{figure}
 
@@
 is 42\% faster than Richards iteration alone.
+
+\begin{table}[htpb!]
+  \begin{center}
+    \begin{tabular}{lrr}\hline\hline
+      \multicolumn{1}{c}{Method} & 
+      \multicolumn{1}{c}{Max Iterations} & 
+      \multicolumn{1}{c}{Relative CPU Time}\\\hline\hline
+      %%
+      PCG & 11 & 1.11 \\
+      %%
+      PFIX & 11 & 1.42 \\
+      %%
+      MCSA & 11 & 1.0 \\
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{Comparison of solution methods for the Marshak problem. All
+    methods were run with a stopping criterion of $\sn{1}{-8}$.  The
+    MCSA method used $N_p=10$ and $W_c = \sn{1}{-4}$.  The CPU time is
+    measured relative to the MCSA method, which required 72.6 s on a
+    3.6~GHz Xeon.}
+  \label{tab:marshak_comparision}
+\end{table}
 
@@
 
+\begin{figure}[htpb!]
+  \centerline{
+    \includegraphics[width=5in,clip]{mesh.eps}}
+  \caption{Multi-material problem mesh.  The duct region is blue.  The
+    wall is red.  The foil is green.  The mesh is $60\times 60\times
+    60$.}
+  \label{fig:multi-mat-mesh}
+\end{figure}
+
+\begin{figure}[htpb!]
+  \centerline{
+    \includegraphics[width=5in,clip]{T_multi_mat.eps}}
+  \caption{
+    Two-dimensional plot of the temperature at 100~sh on a cut-plane
+    positioned at the mid-point of the $z$-axis.  The positions of the
+    edit points where the time-evolution of the temperature is plotted
+    in Fig.~\ref{fig:edit-points} are indicated by the black disks.}
+  \label{fig:multi-mat_T}
+\end{figure}
+
+\begin{figure}[htpb!]
+  \begin{center}
+    \begin{tabular}{m{2in}m{4in}}
+      \subfigure[Edit points]{\includegraphics[width=2in,clip]{points.eps}} &
+      \subfigure[Temperature]{\includegraphics[width=4in,clip]{multi_mat.eps}}
+    \end{tabular}
+  \end{center}
+  \caption{(a) Edit points and (b) time-evolution of the temperature
+    at each point.  All points are centered in the $z$-axis.}
+  \label{fig:edit-points}
+\end{figure}
+
 Table~\ref{tab:multimat_comparision} shows the timing results for the
@@
 MCSA is marginally faster than PCG and 43\% faster than PFIX.
+
+\begin{table}[htpb!]
+  \begin{center}
+    \begin{tabular}{lrr}\hline\hline
+      \multicolumn{1}{c}{Method} & 
+      \multicolumn{1}{c}{Max Iterations} & 
+      \multicolumn{1}{c}{Relative CPU Time}\\\hline\hline
+      %%
+      PCG & 18 & 1.03 \\
+      %%
+      PFIX & 38 & 1.43 \\
+      %%
+      MCSA & 20 & 1.0 \\
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{
+    Comparison of solution methods for the multi-material problem. All
+    methods were run with a stopping criterion of $\sn{1}{-8}$.  The
+    MCSA method used $N_p=1000$ and $W_c = \sn{1}{-3}$.  The problem
+    was run to an elapsed time of 10~sh.  The MCSA method required 3.4
+    hours on a 3.6~GHz Xeon.}
+  \label{tab:multimat_comparision}
+\end{table}
 

### 2012-05-26 doc/hmcsa_adjoint_vs_direct/direct_vs_adjoint.tex
fixed a few typos in the hmcsa tex note

@@
           \email{Thomas Evans}{tme@ornl.gov},
-          \email{Greg Davidson}{gqe@ornl.gov}}
+          \email{Greg Davidson}{gqe@ornl.gov},
+          \email{Scott Mosher}{moshersw@ornl.gov}}
 
@@
 random walk sequence will contribute to the state in which it
-currently resides, not the current source state. It is this set of
+currently resides, not the starting source state. It is this set of
 primary differences between the two solution methods that will dictate
-solver performance with MCSA.
+stochastic solver performance with MCSA.
 

### 2012-05-26 doc/jcomp2012/jcomp2012.tex
Adding adjoint solver justification to jcomp2012 paper

@@
 \end{equation}
+
+\subsection{Justification of Adjoint Error Estimation}
+
+The MCSA method defined in Eq.(~\ref{eq:general-MCSA}) uses the
+adjoint method in Eq.~(\ref{eq:MCSA-residual_solve}) to estimate the
+error instead of the direct method. To demonstrate the effectiveness
+of the adjoint method over the direct method, we choose the
+two-dimensional time-dependent heat equation as a simple model
+problem
+
+\begin{equation}
+  \frac{\partial u}{\partial t} = \alpha \nabla^2 u
+  \label{eq:heat_equation}
+\end{equation}
+
+where $\alpha$ is a constant coefficient. For all comparisons, a
+single time step is computed with implicit Euler time integration. The
+Laplacian is differenced on a square Cartesian grid with a
+second-order five point stencil
+
+\begin{equation}
+  \nabla^2_5 = \frac{1}{h^2}[u_{i-1,j} + u_{i+1,j} + u_{i,j-1} +
+    u_{i,j+1} - 4 u_{i,j}]
+  \label{eq:five_point_stencil}
+\end{equation}
+
+and a fourth-order nine point stencil both assuming a grid size of $h$
+in both the $i$ and $j$ directions.
+
+\begin{equation}
+  \nabla^2_9 = \frac{1}{6h^2}[4 u_{i-1,j} + 4 u_{i+1,j} + 4 u_{i,j-1}
+    + 4 u_{i,j+1} + u_{i-1,j-1} + u_{i-1,j+1} + u_{i+1,j-1} +
+    u_{i+1,j+1} - 20 u_{i,j}]
+  \label{eq:nine_point_stencil}
+\end{equation}
+
+For a single time step solution, we then have the following linear
+system to be solved with the MCSA method.
+
+\begin{equation}
+  A u^{n+1} = u^n
+  \label{eq:heat_eq_lin_sys}
+\end{equation}
+
+Both the stencils will be used to vary the size and density of the
+sparse linear system in Eq.~(\ref{eq:heat_eq_lin_sys}).
+
+A timing study and convergence study is used to demonstrate the
+effectiveness of the adjoint method as compared to the direct
+method. To assess both the CPU time and number of iterations required
+to converge to a solution, a problem of constant grid size and
+$\alpha$ was used with varying values of problem size, fixing the
+spectral radius of the system at a constant value for each
+variation. Both the five point and nine point stencils were used with
+both the direct and adjoint solvers. For each case, 50 histories were
+simulated per cycle with a weight cutoff of $1.0E-4$ for the
+stochastic solver and a convergence tolerance of $1.0E-8$ for the MCSA
+solver. All computations were completed on a 2.66 GHz Intel Core i7
+machine with 8 GB 1067 MHz DDR3 memory.
+
+Fig.~\ref{fig:hmcsa_cpu_time} gives the CPU time needed for each case
+to converge in seconds and Fig.~\ref{fig:hmcsa_iterations} gives the
+number of iterations needed for each case to converge. We see clearly
+in Fig.~\ref{fig:hmcsa_cpu_time} that the using the adjoint solver
+with MCSA results in several orders of magnitude speedup over the
+direct solver while the number of iterations required to converge is
+of a similar scale. We expect this for several reasons. First, with an
+equivalent number of histories specified for both solvers and a system
+of size $N$, the direct solver will compute $N$ times the number of
+histories random walks while the adjoint solver will only compute the
+number of histories specified. Initially, this may not seem like a
+fair comparison, however, we see from Fig.~\ref{fig:hmcsa_cpu_time}
+that the number of iterations requried to converge is approximately
+the same. This is necessary in the direct method to ensure a
+contribution from each state as the random walk sequence will only
+contribute to the starting state. For the adjoint method, because the
+random sequence contributes to the state in which it currently
+resides, fewer histories are neccessary to compute contributions from
+all the nonzero parts of the source. Therefore, although the total
+number of histories per cycle is in fact larger for the direct method,
+the solution computed at each iteration is no better than that
+computed by the adjoint method.
+
+\begin{figure}[htpb!]
+  \centering
+  \includegraphics[width=5in,clip]{Adjoint_Direct_CPU_Time.eps}
+  \caption{\sl CPU Time (s) to converge vs. Problem Size (N for an NxN
+    problem). Both the adjoint and direct solvers are used with the
+    five point and nine point stencils. A significant CPU time speedup
+    is noted with the adjoint method due to the reduced number of
+    histories.}
+  \label{fig:hmcsa_cpu_time}
+\end{figure}
+
+\begin{figure}[htbp!]
+  \centering
+  \includegraphics[width=5in,clip]{Adjoint_Direct_Iterations.eps}
+  \caption{\sl Iterations to converge vs. Problem Size (N for an NxN
+    problem). Both the adjoint and direct solvers are used with the
+    five point and nine point stencils. Both methods give similar
+    iteration behavior.}
+  \label{fig:hmcsa_iterations}
+\end{figure}
+
+Although CPU time is shown to be greatly decreased by using the
+adjoint solver in Fig.~\ref{fig:hmcsa_cpu_time}, the iteration
+behavior was approximately the same. To assess the convergence
+properties of MCSA using each solver and stencil, the infinity norm of
+the residual computed in Eq.~(\ref{eq:general-MCSA}) was computed at
+each iteration for a fixed problem
+size. Fig.~\ref{fig:hmcsa_convergence} gives the results of these
+computations. First, it is worthy to note on the semilog plot that we
+are indeed achieving the expected exponential convergence from
+MCSA. Second, we note that both the direct and adjoint methods have
+approximately equivalent convergence behavior.
+
+\begin{figure}[htpb!]
+  \centering
+  \includegraphics[width=5in,clip]{Adjoint_Direct_Convergence.eps}
+  \caption{\sl Infinity norm of the solution residual vs. iteration
+    number for a problem of fixed size. Both the adjoint and direct
+    solvers are used with the five point and nine point stencils. Both
+    methods give similar iteration behavior.}
+  \label{fig:hmcsa_convergence}
+\end{figure}
 

### 2012-05-26 solvers/test/tstAnasazi.cpp
adding solvers subpackage build system

@@ -1,0 +1,123 @@
+//---------------------------------------------------------------------------//
+/*!
+ * \file   mesh/test/tstAnasazi.cpp
+ * \author Stuart Slattery
+ * \brief  Anasazi class unit tests.
+ */
+//---------------------------------------------------------------------------//
+
+#include <iostream>
+#include <vector>
+#include <cmath>
+#include <sstream>
+#include <ostream>
+
+#include <Teuchos_UnitTestHarness.hpp>
+#include <Teuchos_ParameterList.hpp>
+#include <Teuchos_RCP.hpp>
+
+#include <Epetra_SerialComm.h>
+#include <Epetra_Map.h>
+#include <Epetra_CrsMatrix.h>
+#include <Epetra_MultiVector.h>
+#include <Epetra_Operator.h>
+
+#include "AnasaziMultiVecTraits.hpp"
+#include "AnasaziBasicEigenproblem.hpp"
+#include "AnasaziBlockKrylovSchurSolMgr.hpp"
+#include "AnasaziBasicOutputManager.hpp"
+#include "AnasaziEpetraAdapter.hpp"
+
+//---------------------------------------------------------------------------//
+// TESTS
+//---------------------------------------------------------------------------//
+
+TEUCHOS_UNIT_TEST( Anasazi, block_krylov_schur_test)
+{
+    int problem_size = 10;
+
+    Epetra_SerialComm comm;
+    Epetra_Map map( problem_size, 0, comm );
+
+    // Build A.
+    Teuchos::RCP<Epetra_CrsMatrix> A = 
+	Teuchos::rcp( new Epetra_CrsMatrix( Copy, map, problem_size ) );
+    double lower_diag = -1.0;
+    double diag = 2.0;
+    double upper_diag = -1.0;
+    int global_row = 0;
+    int lower_row = 0;
+    int upper_row = 0;
+    for ( int i = 0; i < problem_size; ++i )
+    {
+	global_row = A->GRID(i);
+	lower_row = i-1;
+	upper_row = i+1;
+	if ( lower_row > -1 )
+	{
+	    A->InsertGlobalValues( global_row, 1, &lower_diag, &lower_row );
+	}
+	A->InsertGlobalValues( global_row, 1, &diag, &global_row );
+	if ( upper_row < problem_size )
+	{
+	    A->InsertGlobalValues( global_row, 1, &upper_diag, &upper_row );
+	}
+    }
+    A->FillComplete();
+
[…]

### 2012-05-26 solvers/test/tstTpetra.cpp

@@ -1,0 +1,149 @@
+//----------------------------------*-C++-*----------------------------------//
+// This library is free software; you can redistribute it and/or
+// modify it under the terms of the GNU Lesser General Public
+// License as published by the Free Software Foundation; either
+// version 3.0 of the License, or (at your option) any later version.
+//
+/*!
+ * \file   mesh/test/tstEpetra.cpp
+ * \author Stuart Slattery
+ * \brief  Epetra class unit tests.
+ */
+//---------------------------------------------------------------------------//
+
+#include <iostream>
+#include <vector>
+#include <cmath>
+#include <sstream>
+
+#include <Teuchos_UnitTestHarness.hpp>
+#include <Teuchos_ArrayRCP.hpp>
+
+#include <Epetra_SerialComm.h>
+#include <Epetra_Map.h>
+#include <Epetra_Vector.h>
+#include <Epetra_CrsMatrix.h>
+
+//---------------------------------------------------------------------------//
+// TESTS
+//---------------------------------------------------------------------------//
+
+TEUCHOS_UNIT_TEST( Epetra_Vector, epetra_vector_test)
+{
+    int error = 0;
+
+    Teuchos::ArrayRCP<double> data(10, 1.3);
+    Epetra_SerialComm comm;
+    Epetra_Map map( (int) data.size(), 0, comm );
+    Epetra_Vector epetra_vector( View, map, data.get() );
+
+    Teuchos::ArrayRCP<double> update_array(10, 1.0);
+    Epetra_Vector update_vector( View, map, update_array.get() );
+
+    Teuchos::ArrayRCP<double>::const_iterator const_it;
+    for ( int i = 0; i < (int) data.size(); ++i )
+    {
+	TEST_ASSERT( epetra_vector[i] == 1.3 );
+    }
+
+    std::vector<double> values( 4, 2.5 );
+    std::vector<int> indices( 4 );
+    indices[0] = 3;
+    indices[1] = 4;
+    indices[2] = 5;
+    indices[3] = 6;
+    error = epetra_vector.ReplaceGlobalValues( 4, 0, &values[0], &indices[0] );
+    TEST_ASSERT( error == 0 );
[…]

### 2012-05-26 solvers/src/IterativeSolver.hpp
Adding iterative solvers protocol based on Belos defintions.

@@ -1,0 +1,85 @@
+//---------------------------------------------------------------------------//
+/*!
+ * \file IterativeSolver.hpp
+ * \author Stuart R. Slattery
+ * \brief Protocol declaration for iterative solvers.
+ */
+//---------------------------------------------------------------------------//
+
+#ifndef CHIMERA_ITERATIVESOLVER_HPP
+#define CHIMERA_ITERATIVESOLVER_HPP
+
+#include <Teuchos_Describable.hpp>
+#include <Teuchos_RCP.hpp>
+
+#include <BelosLinearProblem.hpp>
+
+namespace Chimera
+{
+
+/*!
+ * \brief Protocol definition for iterative solvers.
+ *
+ * Scalar must implement Teuchos::ScalarTraits, MV must implement
+ * Belos::MultiVecTraits, and OP must implement Belos::OperatorTraits.
+ */
+template<typename Scalar typename MV, typename OP>
+class IterativeSolver : public Teuchos::Describable
+{
+  public:
+
+    //@{
+    //! Typedefs.
+    typedef Scalar                                 scalar_type;
+    typedef MV                                     multivector_type;
+    typedef OP                                     operator_type;
+    typedef Belos::LinearProblem<Scalar,MV,OP>     LinearProblemType;
+    Teuchos::RCP<LinearProblemType>                RCP_LinearProblem;
+    //@}
+
+    /*!
+     * \brief Default constructor.
+     */
+    IterativeSolver()
+    { /* ... */ }
+
+    /*!
+     * \brief Destructor.
+     */
+    virtual ~IterativeSolver()
+    { /* ... */ }
+
+    /*!
+     * \brief Iterate to convergence.
+     */
+    virtual void iterate() = 0;
+
+    /*!
+     * \brief Get the iteration count from the last solve.
+     */
+    virtual std::size_t getNumIters() = 0;
+
+    /*!
+     * \brief Get the convergence tolerance.
+     */
+    virtual double getTolerance() = 0;
+
+    /*!
+     * \brief Get the iteration number limit.
+     */
+    virtual std::size_t getMaxIters() = 0;
+
+    /*! 
+     * \brief Get the linear problem.
+     */
+    const RCP_LinearProblem& getLinearProblem() = 0;
+};
+
+} // end namespace Chimera
+
+#end // CHIMERA_ITERATIVESOLVER_HPP
+
+//---------------------------------------------------------------------------//
+// end IterativeSolver.hpp
+//---------------------------------------------------------------------------//
+

### 2012-07-09 doc/jcomp2012/jcomp2012.tex
Adding makefile for jcomp radiation diffusion paper

@@
   %%
-  %% \title{Title\tnoteref{label1}}
-  %% \tnotetext[label1]{}
-  %% \author{Name\corref{cor1}\fnref{label2}}
-  %% \ead{email address}
-  %% \ead[url]{home page}
-  %% \fntext[label2]{}
-  %% \cortext[cor1]{}
-  %% \address{Address\fnref{label3}}
-  %% \fntext[label3]{}
+  \title{Title\tnoteref{label1}}
+  \tnotetext[label1]{}
+  \author{Name\corref{cor1}\fnref{label2}}
+  \ead{email address}
+  \ead[url]{home page}
+  \fntext[label2]{}
+  \cortext[cor1]{}
+  \address{Address\fnref{label3}}
+  \fntext[label3]{}
 
@@
 of the adjoint method over the direct method, we choose the
-two-dimensional time-dependent heat equation as a simple model
+two-dimensional time-dependent Poisson equation as a simple model
 problem
@@
   \frac{\partial u}{\partial t} = \alpha \nabla^2 u
-  \label{eq:heat_equation}
+  \label{eq:poisson_equation}
 \end{equation}
@@
 where $\alpha$ is a constant coefficient. For all comparisons, a
-single time step is computed with implicit Euler time integration. The
-Laplacian is differenced on a square Cartesian grid with a
-second-order five point stencil
+single time step is computed with backwards Euler time
+integration. The Laplacian is differenced on a square Cartesian grid
+with a second-order five point stencil
 
@@
   A u^{n+1} = u^n
-  \label{eq:heat_eq_lin_sys}
+  \label{eq:poisson_eq_lin_sys}
 \end{equation}
@@
 Both the stencils will be used to vary the size and density of the
-sparse linear system in Eq.~(\ref{eq:heat_eq_lin_sys}).
+sparse linear system in Eq.~(\ref{eq:poisson_eq_lin_sys}).
 
@@
 
-Fig.~\ref{fig:hmcsa_cpu_time} gives the CPU time needed for each case
-to converge in seconds and Fig.~\ref{fig:hmcsa_iterations} gives the
+Fig.~\ref{fig:poisson_cpu_time} gives the CPU time needed for each case
+to converge in seconds and Fig.~\ref{fig:poisson_iterations} gives the
 number of iterations needed for each case to converge. We see clearly
-in Fig.~\ref{fig:hmcsa_cpu_time} that the using the adjoint solver
+in Fig.~\ref{fig:poisson_cpu_time} that the using the adjoint solver
 with MCSA results in several orders of magnitude speedup over the
@@
 number of histories specified. Initially, this may not seem like a
-fair comparison, however, we see from Fig.~\ref{fig:hmcsa_cpu_time}
+fair comparison, however, we see from Fig.~\ref{fig:poisson_cpu_time}
 that the number of iterations requried to converge is approximately
@@
     histories.}
-  \label{fig:hmcsa_cpu_time}
+  \label{fig:poisson_cpu_time}
 \end{figure}
@@
     iteration behavior.}
-  \label{fig:hmcsa_iterations}
+  \label{fig:poisson_iterations}
 \end{figure}
@@
 Although CPU time is shown to be greatly decreased by using the
-adjoint solver in Fig.~\ref{fig:hmcsa_cpu_time}, the iteration
+adjoint solver in Fig.~\ref{fig:poisson_cpu_time}, the iteration
 behavior was approximately the same. To assess the convergence
@@
 each iteration for a fixed problem
-size. Fig.~\ref{fig:hmcsa_convergence} gives the results of these
+size. Fig.~\ref{fig:poisson_convergence} gives the results of these
 computations. First, it is worthy to note on the semilog plot that we
@@
     methods give similar iteration behavior.}
-  \label{fig:hmcsa_convergence}
+  \label{fig:poisson_convergence}
 \end{figure}
@@
 number of particles and weight cutoff, we now compare to traditional
-deterministic algorithms.  Table~\ref{tab:marshak_comparision} shows
+deterministic algorithms.  Table~\ref{tab:marshak_comparison} shows
 timing results for the Marshak problem using preconditioned CG (PCG),
@@
 \bibliographystyle{elsarticle-num}
-\bibliography{jcomp2012.bib}
+\bibliography{references}
 

### 2012-07-09 doc/jcomp2012/jcomp2012.tex
Fixed a few typos in the jcomp paper

@@
 
-  %% use the tnoteref command within \title for footnotes;
-  %% use the tnotetext command for the associated footnote;
-  %% use the fnref command within \author or \address for footnotes;
-  %% use the fntext command for the associated footnote;
-  %% use the corref command within \author for corresponding author footnotes;
-  %% use the cortext command for the associated footnote;
-  %% use the ead command for the email address,
-  %% and the form \ead[url] for the home page:
-  %%
-  \title{Title\tnoteref{label1}}
+  \title{A Monte Carlo Synthetic Acceleration Method for Solving the
+    Thermal Radiation Diffusion Equation\tnoteref{label1}}
   \tnotetext[label1]{}
-  \author{Name\corref{cor1}\fnref{label2}}
-  \ead{email address}
-  \ead[url]{home page}
-  \fntext[label2]{}
-  \cortext[cor1]{}
-  \address{Address\fnref{label3}}
-  \fntext[label3]{}
-
-  \title{A Monte Carlo Synthetic Acceleration Method for Solving the
-    Thermal Radiation Diffusion Equation}
-
-  %% use optional labels to link authors explicitly to addresses:
-  %% \author[label1,label2]{<author name>}
-  %% \address[label1]{<address>}
-  %% \address[label2]{<address>}
-
-  \author{}
-
-  \address{}
+
+  \author[ornl]{Thomas M. Evans\corref{cor1}\fnref{fn1}}
+  \ead{evanstm@ornlgov}
+  \ead[url]{www.ornl.gov}
+  \cortext[cor1]{Corresponding Author}
+
+  \author[ornl]{Scott W. Mosher\fnref{fn1}}
+  \ead{moshersw@ornl.gov}
+
+  \author[wisc]{Stuart R. Slattery\fnref{fn2}}
+  \ead{sslattery@wisc.edu}
+  \ead[url]{www.wisc.edu}
+
+  \fntext[fn1]{Radiation Transport Group, Reactor and Nuclear Systems
+    Division} 
+  \fntext[fn2]{Department of Engineering Physics, College of Engineering}
+
+  \address[ornl]{Oak Ridge National Laboratory, 
+    1 Bethel Valley Rd., Oak Ridge, TN 378131 U.S.A.}
+  \address[wisc]{University of Wisconsin-Madison, 1500 Engineering
+    Dr., Madison, WI 53716, U.S.A.}
 
@@
     %% keywords here, in the form: keyword \sep keyword
-
+    radiation diffusion \sep Monte Carlo \sep sparse matrix systems
     %% MSC codes here, in the form: \MSC code \sep code
@@
     3.6~GHz Xeon.}
-  \label{tab:marshak_comparision}
+  \label{tab:marshak_comparison}
 \end{table}
@@
 
-Table~\ref{tab:multimat_comparision} shows the timing results for the
+Table~\ref{tab:multimat_comparison} shows the timing results for the
 multi-material problem using PCG, PFIX, and MCSA.  The results
@@
     hours on a 3.6~GHz Xeon.}
-  \label{tab:multimat_comparision}
+  \label{tab:multimat_comparison}
 \end{table}

### 2012-07-13 doc/jcomp2012/jcomp2012.tex
Adding figure and data file adjoint vs. direct cpu time comparision

@@
 \begin{equation}
-  \frac{\partial u}{\partial t} = \alpha \nabla^2 u
+  \frac{\partial u}{\partial t} = \nabla^2 u
   \label{eq:poisson_equation}
@@
 
-where $\alpha$ is a constant coefficient. For all comparisons, a
-single time step is computed with backwards Euler time
-integration. The Laplacian is differenced on a square Cartesian grid
-with a second-order five point stencil
+For all comparisons, a single time step is computed with backwards
+Euler time integration. The Laplacian is differenced on a square
+Cartesian grid with a second-order five point stencil
 
@@
 method. To assess both the CPU time and number of iterations required
-to converge to a solution, a problem of constant grid size and
-$\alpha$ was used with varying values of problem size, fixing the
-spectral radius of the system at a constant value for each
-variation. Both the five point and nine point stencils were used with
-both the direct and adjoint solvers. For each case, 50 histories were
-simulated per cycle with a weight cutoff of $1.0E-4$ for the
-stochastic solver and a convergence tolerance of $1.0E-8$ for the MCSA
-solver. All computations were completed on a 2.66 GHz Intel Core i7
-machine with 8 GB 1067 MHz DDR3 memory.
+to converge to a solution, a problem of constant grid size and was
+used with varying values of problem size, fixing the spectral radius
+of the system at a constant value for each variation. Both the five
+point and nine point stencils were used with both the direct and
+adjoint solvers. For each case, 50 histories were simulated per cycle
+with a weight cutoff of $1.0E-4$ for the stochastic solver and a
+convergence tolerance of $1.0E-8$ for the MCSA solver. All
+computations were completed on a 2.66 GHz Intel Core i7 machine with 8
+GB 1067 MHz DDR3 memory.
 

### 2012-07-13 doc/jcomp2012/jcomp2012.tex
Adding adjoint vs. direct convergence data and figures
Updates to the jcomp2012 paper

@@
   \title{A Monte Carlo Synthetic Acceleration Method for Solving the
-    Thermal Radiation Diffusion Equation\tnoteref{label1}}
-  \tnotetext[label1]{}
+    Thermal Radiation Diffusion Equation}
 
@@
   \address[wisc]{University of Wisconsin-Madison, 1500 Engineering
-    Dr., Madison, WI 53716, U.S.A.}
+    Dr., Madison, WI 53716 U.S.A.}
 
@@
   \centerline{
-    \input{3D_mesh_segment.pstex_t}}
+    \input{3D_mesh_segment.pdftex_t}}
   \caption{3D mesh-segment.  The particle's current cell (in blue) is
@@
 
+%%---------------------------------------------------------------------------%%
+
 \end{document}
@@
 %%
-%% End of file `elsarticle-template-num.tex'.
+%% End of file `jcomp2012.tex'.

### 2012-07-16 doc/jcomp2012/jcomp2012.tex
updated latex and makefile for figures

@@
 %% or use the graphicx package for more complicated commands
-%% \usepackage{graphicx}
+\usepackage{graphicx}
+\usepackage{subfigure}
 %% or use the epsfig package if you prefer to use the old commands
@@
 \usepackage[mathcal]{euscript}
+\usepackage{color}
 
@@
 
-  %% Title, authors and addresses
-
   \title{A Monte Carlo Synthetic Acceleration Method for Solving the
@@
 
-  %%---------------------------------------------------------------------------%%
   \begin{abstract}
@@
 
-  %%---------------------------------------------------------------------------%%
   \begin{keyword}
-    %% keywords here, in the form: keyword \sep keyword
     radiation diffusion \sep Monte Carlo \sep sparse matrix systems
-    %% MSC codes here, in the form: \MSC code \sep code
-    %% or \MSC[2008] code \sep code (2000 is the default)
-
   \end{keyword}
@@
 \end{frontmatter}
-
-%%
-%% Start line numbering here if you want
-%%
-% \linenumbers
 
@@
 solution method for 3D, time-dependent discrete systems.  In this
-note, we extend our previous work in the following manner:
+paper, we extend our previous work in the following manner:
 \begin{itemize}
@@
 \end{subequations}
-In this scheme, the adjoint method is used to estimate the solution to
-Eq.~(\ref{eq:residual_solve}), and the residual is iterated to
-convergence.  This iteration sequence is closely related to iterative
-refinement; the exception being that there is no update to to
-$x^{l+1}$ between residual iterations.
+In this scheme, the Monte Carlo adjoint method is used to estimate the
+solution to Eq.~(\ref{eq:residual_solve}), and the residual is
+iterated to convergence.  This iteration sequence is closely related
+to iterative refinement; the exception being that there is no update
+to to $x^{l+1}$ between residual iterations.
 
@@
 
-\subsection{Justification of Adjoint Error Estimation}
+\subsection{Monte Carlo Method Selection and Convergence}
 
@@
 
-Fig.~\ref{fig:poisson_cpu_time} gives the CPU time needed for each case
-to converge in seconds and Fig.~\ref{fig:poisson_iterations} gives the
-number of iterations needed for each case to converge. We see clearly
-in Fig.~\ref{fig:poisson_cpu_time} that the using the adjoint solver
-with MCSA results in several orders of magnitude speedup over the
-direct solver while the number of iterations required to converge is
-of a similar scale. We expect this for several reasons. First, with an
+Fig.~\ref{fig:poisson_cpu_time} gives the CPU time needed for each
+case to converge in seconds and Fig.~\ref{fig:poisson_iterations}
+gives the number of iterations needed for each case to converge
+relative to the problem size. We see clearly in
+Fig.~\ref{fig:poisson_cpu_time} that the using the adjoint solver with
+MCSA results in several orders of magnitude speedup over the direct
+solver while the number of iterations required to converge is of a
+similar scale. We expect this for several reasons. First, with an
 equivalent number of histories specified for both solvers and a system
@@
   \caption{\sl CPU Time (s) to converge vs. Problem Size (N for an NxN
-    problem). Both the adjoint and direct solvers are used with the
-    five point and nine point stencils. A significant CPU time speedup
-    is noted with the adjoint method due to the reduced number of
+    mesh). Both the adjoint and direct solvers are used with the five
+    point and nine point stencils. A significant CPU time speedup is
+    noted with the adjoint method due to the reduced number of
     histories.}
@@
   \caption{\sl Iterations to converge vs. Problem Size (N for an NxN
-    problem). Both the adjoint and direct solvers are used with the
-    five point and nine point stencils. Both methods give similar
-    iteration behavior.}
+    mesh. Both the adjoint and direct solvers are used with the five
+    point and nine point stencils. Both methods give similar iteration
+    behavior.}
   \label{fig:poisson_iterations}
@@
 must precondition Eq.~(\ref{eq:diffusion-matrix}) in order to use the
-MCSA method.  Preconditioning accomplishes two purposes: it reduces
-the spectral radius of $\ve{D}$ so that Monte Carlo solution methods
-are applicable, and it reduces the number of iterations required for
-convergence.  We apply left-preconditioning and write
+MCSA method. We apply left-preconditioning and write
 \begin{equation}
@@
   fewer iterations are required resulting in accelerated convergence.
-\item Makes the diagonal elements of the iteration matrix zero,
-  $\diag(\vI - \ve{M}^{-1}\ve{D}) = 0$.  Thus, there will be no time
-  wasted in transitions $i\rightarrow i$ during the random walk
+\item The diagonal elements of the preconditioned iteration matrix are
+  zero, $\diag(\vI - \ve{M}^{-1}\ve{D}) = 0$.  Thus, there will be no
+  time wasted in transitions $i\rightarrow i$ during the random walk
   process.
@@
     - \frac{\sigm_{j+1}}{\sigT}\frac{\Delta_{j+1}}{\Dj}\phi_{i\,j+1\,k} \\
-    &- \frac{\sigp_{j-1}}{\sigT}\frac{\Delta_{j-1}}{\Dj}\phi_{i\,j-1\,k} 
+    &-
+    \frac{\sigp_{j-1}}{\sigT}\frac{\Delta_{j-1}}{\Dj}\phi_{i\,j-1\,k}
     - \frac{\sigm_{k+1}}{\sigT}\frac{\Delta_{k+1}}{\Dk}\phi_{ij\,k+1}
@@
 method requires two solution steps:
+
 \begin{enumerate}
@@
 \end{enumerate}
+
 For clarity, we rewrite the iteration scheme in
@@
 from Eq.~(\ref{eq:preconditioned-diffusion}),
+
 \begin{align*}
@@
 \end{align*}
+
 As described in \S~\ref{sec:iter-refin-monte}, the adjoint Monte Carlo
@@
 follows:
+
 \begin{equation}
@@
 \end{equation}
+
 This result is used to compute the residual of the system.  The
@@
   \begin{center}
-    \begin{tabular}{m{2in}m{4in}}
+    \begin{tabular}{cc}
       \subfigure[Edit points]{\includegraphics[width=2in,clip]{points.eps}} &

### 2012-07-17 doc/jcomp2012/jcomp2012.tex
updating makefile for jcomp

@@
 method is actually a variant of iterative refinement cast as a
-residual method.  Using Monte Carlo to accelerate fixed-point
+residual method.  Using Monte Carlo to accelerate a fixed-point
 (Richardson) iteration, our new method surpasses our previous 1D
@@
 \end{equation}
-and $V_{ijk} = \Delta_i\Delta_j\Delta_k$.  We now apply the divergence
-theorem,
+and $V_{ijk} = \Delta_i\Delta_j\Delta_k$.
+
+\begin{figure}[htpb!]
+  \centerline{
+    \input{mesh_cell.pdftex_t}}
+  \caption{Mesh Cell.}
+  \label{fig:cell}
+\end{figure}
+
+We now apply the divergence theorem,
 \begin{equation}
@@
     \begin{tabular}{cc}
-      \subfigure[Edit points]{\includegraphics[width=2in,clip]{points.eps}} &
-      \subfigure[Temperature]{\includegraphics[width=4in,clip]{multi_mat.eps}}
+      \subfigure[Edit points]{\includegraphics[width=1.5in,clip]{points.eps}} &
+      \subfigure[Temperature]{\includegraphics[width=3.5in,clip]{multi_mat.eps}}
     \end{tabular}

### 2012-07-18 doc/jcomp2012/jcomp2012.tex
Almost done with rough draft for jcomp paper

@@
 \newcommand{\vr}{\ensuremath{\ve{r}}}
+\newcommand{\vu}{\ensuremath{\ve{u}}}
 \newcommand{\vI}{\ensuremath{\ve{I}}}
@@
     We present a synthetic-acceleration based Monte Carlo method for
-    solving the 1T thermal radiation diffusion equations.  We show
-    that this method can be an effective solver for sparse matrix
-    systems.  We also discuss its general applicability to broader
-    classes of problems.
+    solving the 1T thermal radiation diffusion equation.  We show that
+    this method can be an effective solver for sparse matrix systems
+    and is competitive with deterministic methods.  We also discuss
+    its general applicability to broader classes of problems.
 
@@
   \begin{keyword}
-    radiation diffusion \sep Monte Carlo \sep sparse matrix systems
+    radiation diffusion \sep synthetic acceleration \sep Monte Carlo
+    \sep sparse matrix systems
   \end{keyword}
@@
 solution method for 3D, time-dependent discrete systems. In this
-paper, we extend our previous work in the following manner:
+work, we extend our previous work in the following manner:
+
 \begin{itemize}
@@
 \end{itemize}
+
 In \S~\ref{sec:monte-carlo-matrix} we review Monte Carlo methods for
@@
 Consider the following matrix equation:
+
 \begin{equation}
@@
 \end{equation}
+
 which can be written,
+
 \begin{equation}
@@
 \end{equation}
+
 Here, $\vH = (\vI - \vA)$ is the \latin{iteration matrix}.  When the
 spectral radius, $\rho(\vH)$, of the iteration matrix satisfies
+
 \begin{equation}
@@
 \end{equation}
+
 we can expand $\vA^{-1}$ using the \latin{Neumann Series},
+
 \begin{equation}
@@
 \end{equation}
+
 Thus, when $\rho(\vH) < 1$ we can recast the solution vector $\vx$ as
 a series,
+
 \begin{equation}
@@
 \end{equation}
+
 With this knowledge in hand, we can write an iterative method that
 solves Eq.~(\ref{eq:Ax=b}) (Richardson's Iteration),
+
 \begin{equation}
@@
 \end{equation}
+
 Equation~(\ref{eq:richardson}) will converge when $\rho(\vH) < 1\ 
@@
 
-Now we consider a Monte Carlo method that can be used to estimate the solution
-to Eq.~(\ref{eq:Ax=b})~\cite{hammersley_1964}.  First, rewrite
-Eq.~(\ref{eq:neumann_series}) in order to calculate a component of
-$\vx$,
+Now we consider a Monte Carlo method that can be used to estimate the
+solution to Eq.~(\ref{eq:Ax=b})~\cite{hammersley_1964}.  First,
+rewrite Eq.~(\ref{eq:neumann_series}) in order to calculate a
+component of $\vx$,
+
 \begin{equation}
@@
 \end{equation}
-This series can be interpreted as a series of transitions from
-$i_{m-1}\rightarrow i_m$ that can be simulated by a random walk.
-Let $X$ be a random variable sampled from a random walk with $k$
-events that initiates in state $i$,
+
+The terms in the Neumann series can be interpreted as a series of
+transitions from $i_{m-1}\rightarrow i_m$ that can be simulated by a
+random walk.  Let $X$ be a random variable sampled from a random walk
+with $k$ events (particles) that initiates in state $i$,
+
 \begin{equation}
@@
 \end{equation}
+
 Here, the particle weight on the $m^\text{th}$ step is denoted $W_m$,
-and each particle starts with unit weight. When the particle
-transitions between states, for example $i\rightarrow j$, the weight
-is multiplied by the factor
-\begin{equation}
-  w_{ij} = \frac{h_{ij}}{p_{ij}}\:.
+and each particle starts with unit weight. At every step,
+contributions to each component of $X$ are provided only by the
+starting state of the particle, $b_{i_m}$. Therefore, for a given
+random walk permutation, each state must be used as a starting value
+in order to calculated its contribution to the expectation value
+(i.e. a state of size $N$ will have $N$ particles per random walk
+permutation). When the particle transitions between states, for
+example $i\rightarrow j$, the particle weight is multiplied by the
+transition factor.
+
+\begin{equation}
+  w_{ij} = \frac{h_{ij}}{p_{ij}}\:,
   \label{eq:weight}
 \end{equation}
-Then, the expected value of $X$ is
+
+where $p_{ij}$ denotes the probability of transitioning from state $i$
+to $j$. Then, the expected value of $X$ is
+
 \begin{equation}
@@
 \end{equation}
+
 where $\nu$ denotes a particular random walk permutation.
@@
 straightforward approach is to set
+
 \begin{equation}
@@
 \end{equation}
+
 Each row of the transition probability matrix, $\vP$, represents a
-discrete probability density function for selecting a new state $j$,
-given that the current state is $i$.  Random walks can be terminated
-in two ways: the matrix can be augmented with a terminating event
-equation that describes the probability that a particle ends its walk,
-or the random walk can be terminated by weight cutoff.  Generally, we
-choose to terminate random walks using a weight cutoff as opposed to
-augmenting the matrix.
+discrete probability density function that can be sampled to select a
+new state $j$, given that the current state is $i$.  Random walks can
+be terminated in two ways: the matrix can be augmented with a
+terminating event equation that describes the probability that a
+particle ends its walk, or the random walk can be terminated by weight
+cutoff, $W_c$.  Generally, we choose to terminate random walks using a
+weight cutoff as opposed to augmenting the matrix sucth that the
+random walk terminates when $W_m < W_c$.
 
@@
 change from state $i\rightarrow j$ is
+
 \begin{equation}
@@
 \end{equation}
+
 The transition probabilities may be calculated as
+
 \begin{equation}
@@
 \end{equation}
+
 Note that the indices are reversed so that the probabilities are
@@
 estimator for this method is
+
 \begin{equation}
@@
 \end{equation}
-Here, $\hat{b}_{i_0}$ is the sampled source in cell/state $i_0$.  The
-Kronecker delta implies that tallies are only made in the cell/state
-where the random walk currently resides.  This is the common approach
-in standard Monte Carlo transport simulations.  We will investigate
-this approach as applied to the 1T thermal radiation diffusion
-problem in \S~\ref{sec:solut-radi-diff}.
+
+Here, $\hat{b}_{i_0}$ is the sampled source and initial particle
+weight in state $i_0$.  The Kronecker delta implies that tallies are
+only made in the state where the random walk currently resides.  This
+is the common approach in standard Monte Carlo transport
+simulations. Unlike the direct method, because tallies are made in the
+state in which the particle currently resides, we are no longer
+required to start a particle in each state for each random walk
+permutation. Instead, sampling the source is suffucient and requires
+only one particle per permutation.
 
@@
 fraction of the starting weight such that
+
 \begin{equation}
@@
 \end{equation}
+
 where $W_f$ is the terminating weight of the random walk, $W_c$ is the
 input relative weight cutoff, and $\hat{b}_{i_0}$ is the initial
-weight of the source particle.  The random walk ends on the
+weight of the source particle.  The random walk terminates on the
 $m^\text{th}$ step if $W_m < W_f$.
@@
 The methods presented in \S~\ref{sec:monte-carlo-matrix} are
-characterized by slow convergence rates.
-Halton~\cite{halton_1962,halton_1994} proposed a staged residual
-scheme called Sequential Monte Carlo to speed up the convergence of
-these methods.  A variant of this scheme has been successfully applied
-to the 1D nonlinear thermal radiation diffusion equation in
-Ref.~\cite{evans_2003}.
+characterized by slow convergence rates bound by the Central Limit
+Theorem.  Halton~\cite{halton_1962,halton_1994} proposed a staged
+residual scheme called Sequential Monte Carlo to speed up the
+convergence of these methods.  A variant of this scheme has been
+successfully applied to the 1D nonlinear thermal radiation diffusion
+equation in Ref.~\cite{evans_2003}.
 
@@
 applied:
+
 \begin{subequations}
@@
 \end{subequations}
+
 In this scheme, the Monte Carlo adjoint method is used to estimate the
@@
 We propose a modification of this scheme that uses Monte Carlo as a
-synthetic-acceleration scheme for fixed-point iteration. The
-Fixed-Point Monte Carlo Synthetic-Acceleration (MCSA) method is
-defined as follows:
+synthetic-acceleration scheme for fixed-point iteration. We begin by
+subtracting Eq.~(\ref{eq:richardson}) from Eq:~(\ref{eq:point_iteration}):
+
+\begin{equation}
+  \delta \vx^{l+1} = (\vI - \vA)\delta \vx^{l}\:,
+  \label{eq:synthetic_error}
+\end{equation}
+
+where
+
+\begin{equation}
+  \delta \vx^{l+1} = \vx - \vx^{l+1}
+  \label{eq:lplus1error}
+\end{equation}
+
+is defined as the error at iteration $l+1$. We then subtract
+$(\vI-\vA) \delta \vx^{l+1}$ from Eq.~(\ref{eq:synthetic_error}) giving:
+
+\begin{equation}
+  \vA \delta \vx^{l+1} = (\vI - \vA)(\vx^{l+1} - \vx^{l})\:.
+  \label{eq:synthetic_error2}
+\end{equation}
+
+The following iterative scheme will then converge in one iteration:
+
 \begin{subequations}
   \begin{gather}
-    \vx^{l+1/2} = (\vI - \vA)\vx^l + \vb\:, \\
-    %%
-    \vr^{l+1/2} = \vb - \vA\vx^{l+1/2}\:,\\
+    \vx^{l+1} = (\vI - \vA)\vx^l + \vb\:,\\
+    %%
+    \vA \delta \vx^{l+1} = (\vI - \vA)(\vx^{l+1} - \vx^{l})\:.
+    \label{eq:synthetic-residual_solve}\\ 
+    %% 
+    \vx=\vx^{l+1}+\delta\vx^{l+1}\:.
+  \end{gather}
+  \label{eq:general-synthetic}
+\end{subequations}
+
+To accelerate the fixed point method, we can instead create a new
+iterative scheme by approximating the operator $\vA$ using the Monte
+Carlo method to estimate the solution error. With this idea hand, the
+Fixed-Point Monte Carlo Synthetic-Acceleration (MCSA) method can be
+defined as follows:
+
+\begin{subequations}
+  \begin{gather}
+    \vx^{l+1/2} = (\vI - \vA)\vx^l + \vb\:,\\
+    %%
+    \vr^{l+1/2} = \vb - \vA\vx^{l+1/2}\:,
+    \label{eq:MCSA-residual}\\     
     %%
@@
 \end{subequations}
+
 The hat on $\vA$ in Eq.~(\ref{eq:MCSA-residual_solve}) indicates that
@@
 estimate the error, $\delta\vx^{l+1/2}$, by solving
-Eq.~(\ref{eq:MCSA-residual_solve}) via the Monte Carlo adjoint method.
-The error is used to calculate an updated iterate of the solution
-vector, $\vx^{l+1}$.  The entire sequence is iterated to convergence
-based on the following stopping criterion~\cite{kelley_1995},
+Eq.~(\ref{eq:MCSA-residual_solve}) via the Monte Carlo adjoint
+method. The error is then used to calculate an updated iterate of the
+solution vector, $\vx^{l+1}$.  The entire sequence is iterated to
+convergence based on the following stopping
+criterion~\cite{kelley_1995},
 \begin{equation}
@@
 
-\subsection{Monte Carlo Method Selection and Convergence}
+\subsection{Monte Carlo Method Selection}
 
@@
 adjoint method in Eq.~(\ref{eq:MCSA-residual_solve}) to estimate the
-error instead of the direct method. To demonstrate the effectiveness
-of the adjoint method over the direct method, we choose the
-two-dimensional time-dependent Poisson equation as a simple model
-problem
-
-\begin{equation}
-  \frac{\partial u}{\partial t} = \nabla^2 u
+error instead of the direct method outlined in
+\S~\ref{sec:direct-method}. To demonstrate the effectiveness of the
+adjoint method over the direct method, we choose the two-dimensional
+time-dependent Poisson equation as a simple model problem:
+
+\begin{equation}
+  \frac{\partial \vu}{\partial t} = \nabla^2 \vu\:.
   \label{eq:poisson_equation}
@@
 \begin{equation}
-  \nabla^2_5 = \frac{1}{h^2}[u_{i-1,j} + u_{i+1,j} + u_{i,j-1} +
+  \nabla^2_5 = \frac{1}{\Delta^2}[u_{i-1,j} + u_{i+1,j} + u_{i,j-1} +
     u_{i,j+1} - 4 u_{i,j}]
@@
 
-and a fourth-order nine point stencil both assuming a grid size of $h$
-in both the $i$ and $j$ directions.
-
-\begin{equation}
-  \nabla^2_9 = \frac{1}{6h^2}[4 u_{i-1,j} + 4 u_{i+1,j} + 4 u_{i,j-1}
-    + 4 u_{i,j+1} + u_{i-1,j-1} + u_{i-1,j+1} + u_{i+1,j-1} +
-    u_{i+1,j+1} - 20 u_{i,j}]
+and a fourth-order nine point stencil both assuming a grid size of
+$\Delta$ in both the $i$ and $j$ directions.
+
+\begin{multline}
+  \nabla^2_9 = \frac{1}{6\Delta^2}[4 u_{i-1,j} + 4 u_{i+1,j} + 4
+    u_{i,j-1} + 4 u_{i,j+1} + u_{i-1,j-1}\\
+    + u_{i-1,j+1} + u_{i+1,j-1} + u_{i+1,j+1} - 20 u_{i,j}]
   \label{eq:nine_point_stencil}
-\end{equation}
-
-For a single time step solution, we then have the following linear
-system to be solved with the MCSA method.
-
-\begin{equation}
-  A u^{n+1} = u^n
+\end{multline}
+
+For a single time step solution, we then have the following sparse
+linear system to be solved with the MCSA method.
+
+\begin{equation}
+  \vA \vu^{n+1} = \vu^n
   \label{eq:poisson_eq_lin_sys}
@@
 method. To assess both the CPU time and number of iterations required
-to converge to a solution, a problem of constant grid size and was
-used with varying values of problem size, fixing the spectral radius
-of the system at a constant value for each variation. Both the five
-point and nine point stencils were used with both the direct and
-adjoint solvers. For each case, 50 histories were simulated per cycle
-with a weight cutoff of $1.0E-4$ for the stochastic solver and a
-convergence tolerance of $1.0E-8$ for the MCSA solver. All
-computations were completed on a 2.66 GHz Intel Core i7 machine with 8
-GB 1067 MHz DDR3 memory.
-
-Fig.~\ref{fig:poisson_cpu_time} gives the CPU time needed for each
-case to converge in seconds and Fig.~\ref{fig:poisson_iterations}
-gives the number of iterations needed for each case to converge
-relative to the problem size. We see clearly in
-Fig.~\ref{fig:poisson_cpu_time} that the using the adjoint solver with
-MCSA results in several orders of magnitude speedup over the direct
-solver while the number of iterations required to converge is of a
-similar scale. We expect this for several reasons. First, with an
-equivalent number of histories specified for both solvers and a system
-of size $N$, the direct solver will compute $N$ times the number of
-histories random walks while the adjoint solver will only compute the
-number of histories specified. Initially, this may not seem like a
-fair comparison, however, we see from Fig.~\ref{fig:poisson_cpu_time}
-that the number of iterations requried to converge is approximately
-the same. This is necessary in the direct method to ensure a
-contribution from each state as the random walk sequence will only
-contribute to the starting state. For the adjoint method, because the
-random sequence contributes to the state in which it currently
-resides, fewer histories are neccessary to compute contributions from
-all the nonzero parts of the source. Therefore, although the total
-number of histories per cycle is in fact larger for the direct method,
-the solution computed at each iteration is no better than that
-computed by the adjoint method.
+to converge to a solution, a problem of constant $\Delta$ and was used
+with varying values of grid size, fixing the spectral radius of the
+system at a constant value for each variation. Both the five point and
+nine point stencils were used with both the direct and adjoint
+solvers. For each case, 50 random walk permuations were simulated per
+iteration with a weight cutoff of $1.0E-4$ for the stochastic linear
+solver and a convergence tolerance of $1.0E-8$ for the MCSA iterative
+solver. All computations were completed on a 2.66 GHz Intel Core i7
+machine with 8 GB 1067 MHz DDR3 memory used for all computations
+presented in this section. Fig.~\ref{fig:poisson_cpu_time} gives the
+CPU time needed for each case to converge in seconds and
+Fig.~\ref{fig:poisson_iterations} gives the number of iterations
+needed for each case to converge relative to the problem size.
 
@@
   \includegraphics[width=5in,clip]{Adjoint_Direct_CPU_Time.eps}
-  \caption{\sl CPU Time (s) to converge vs. Problem Size (N for an NxN
-    mesh). Both the adjoint and direct solvers are used with the five
-    point and nine point stencils. A significant CPU time speedup is
-    noted with the adjoint method due to the reduced number of
-    histories.}
+  \caption{\sl CPU Time (s) to converge vs. Problem Size ($N$ for an
+    $N \times N$ mesh). Both the adjoint and direct solvers are used
+    with the five point and nine point stencils. A significant CPU
+    time speedup is noted with the adjoint method due to the reduced
+    number of histories.}
   \label{fig:poisson_cpu_time}
@@
   \includegraphics[width=5in,clip]{Adjoint_Direct_Iterations.eps}
-  \caption{\sl Iterations to converge vs. Problem Size (N for an NxN
-    mesh. Both the adjoint and direct solvers are used with the five
-    point and nine point stencils. Both methods give similar iteration
-    behavior.}
+  \caption{\sl Iterations to converge vs. Problem Size ($N$ for an $N
+    \times N$ mesh. Both the adjoint and direct solvers are used with
+    the five point and nine point stencils. Both methods give similar
+    iteration behavior.}
   \label{fig:poisson_iterations}
 \end{figure}
+
+We see clearly in Fig.~\ref{fig:poisson_cpu_time} that the using the
+adjoint solver with MCSA results in several orders of magnitude
+speedup over the direct solver while the number of iterations required
+to converge is of a similar scale. We expect this for several
+reasons. First, with an equivalent number of histories specified for
+both solvers and a system of size $N \times N$, the direct solver will
+compute $N \times N$ random walks per permuation while the adjoint solver
+will only compute one. Initially, this may not seem like a fair
+comparison, however, we see from Fig.~\ref{fig:poisson_cpu_time} that
+the number of iterations requried to converge is approximately the
+same. This is necessary in the direct method to ensure a contribution
+from each state as the random walk sequence will only contribute to
+the starting state. For the adjoint method, because the random walk
+sequence contributes to the state in which it currently resides, fewer
+histories are neccessary to compute contributions from all the nonzero
+parts of the source. Therefore, although the total number of random
+walks per iteration is in fact larger for the direct method, the
+solution computed at each iteration is no better than that computed by
+the adjoint method.
 
@@
 properties of MCSA using each solver and stencil, the infinity norm of
-the residual computed in Eq.~(\ref{eq:general-MCSA}) was computed at
+the residual computed in Eq.~(\ref{eq:general-MCSA}) was collected at
 each iteration for a fixed problem
@@
 MCSA. Second, we note that both the direct and adjoint methods have
-approximately equivalent convergence behavior.
+approximately equivalent convergence behavior, leading to CPU time
+being the dominating factor in method selection.
 
@@
 \cite{morel_1996}
+
 \begin{equation} 
@@
 \end{equation}
+
 where $\Cv\equiv\Cv(\ve{r},T(\ve{r},t))$
@@
 state variable.  The intensity is defined,
+
 \begin{equation}
@@
 \end{equation}
-where $\psi$ is the angular radiation intensity.  The Planck function
-(or Planckian) is defined as,
+
+where $\psi$ is the angular radiation intensity and $\nu$ is the
+radiation frequency.  The Planck function (or Planckian) is defined
+as,
+
 \begin{equation}
@@
 \end{equation}
+
 where $h$ is Planck's constant and $k$ is the Boltzmann constant.
 Utilizing this definition of $\phi$ we can write
+
 \begin{equation}
@@
 \end{equation}
+
 Inserting \eqref{eq:partial-T} into \eqref{eq:eq-diff-T} yields
+
 \begin{equation}
   \label{eq:eq-diff}
-  \Bigl(\frac{\Cv}{4acT^{3}} + \frac{1}{c}\Bigr)\pder{\phi}{t} - 
+  \Bigl(\frac{\rho\Cv}{4acT^{3}} + \frac{1}{c}\Bigr)\pder{\phi}{t} - 
   \nabla\cdot D\grad{\phi} = Q\:,
 \end{equation}
+
 where $D$ is the diffusion coefficient, $D\equiv
 D(\ve{r},T(\ve{r},t))\ [\text{cm}]$, and is defined
+
 \begin{equation}
@@
 \end{equation}
+
 Equation~\eqref{eq:eq-diff} is the model equation we will solve using
@@
 and lagging the temperature-dependent coefficients at $t^n$ gives
+
 \begin{equation}
@@
 \end{equation}
+
 where we have suppressed the $n+1$ indices on $\phi$ and
+
 \begin{align}
@@
 \end{align}
+
 Applying Fick's Law,
+
 \begin{equation}
@@
 \end{equation}
+
 where $\ve{F}$ is the radiation flux, we define a flux-balance
 equation,
+
 \begin{equation}
@@
 \end{equation}
+
 Integrating Eq.~(\ref{eq:flux-balance}) over the cell illustrated in
 Fig.~\ref{fig:cell} gives
+
 \begin{equation}
@@
 \end{equation}
+
 and $V_{ijk} = \Delta_i\Delta_j\Delta_k$.
@@
     \input{mesh_cell.pdftex_t}}
-  \caption{Mesh Cell.}
+  \caption{Orthogonal 3D mesh Cell.}
   \label{fig:cell}
@@
 We now apply the divergence theorem,
+
 \begin{equation}
@@
 \end{equation}
+
 to Eq.(\ref{eq:flux-balance-vol-integral}) yielding
+
 \begin{equation}
   \begin{split}
-    (F_{i+1/2} - F_{i-1/2})\Dj\Dk + (F_{j+1/2} &- F_{j-1/2})\Di\Dk \\
-    &+ (F_{k+1/2} - F_{k-1/2})\Di\Dj
-    + \sign_{ijk}\phi_{ijk}V_{ijk} = \qn_{ijk}V_{ijk}\:.
+    (F_{i+1/2} - F_{i-1/2})\Dj\Dk + (F_{j+1/2} &- F_{j-1/2})\Di\Dk +
+    (F_{k+1/2}\\ &- F_{k-1/2})\Di\Dj + \sign_{ijk}\phi_{ijk}V_{ijk} =
+    \qn_{ijk}V_{ijk}\:.
   \end{split}
@@
 the face-centered fluxes are
+
 \begin{align}
@@
 \end{align}
+
 where
+
 \begin{align}
@@
 \end{align}
+
 In order to be conservative, the flux must be continuous across a
@@
 face must be equivalent,
+
 \begin{equation}
@@
 \end{equation}
+
 where we have dropped the subscript $\mathrm{R}$ from $\sigma$ for
@@
 face-centered diffusion coefficient,
+
 \begin{equation}
@@
 \end{equation}
+
 Following the same procedure at the $l-1/2$ face gives the diffusion
 coefficient on the low-side face,
+
 \begin{equation}
@@
 \end{equation}
+
 Using Eqs.~(\ref{eq:F_l+1/2}), (\ref{eq:F_l-1/2}), (\ref{eq:D_l+1/2}),
@@
 (\ref{eq:flux-balance-difference}), we obtain
+
 \begin{equation}
@@
     - \sigm_{i+1}\frac{\Delta_{i+1}}{\Di}\phi_{i+1\,jk} &
-    - \sigp_{i-1}\frac{\Delta_{i-1}}{\Di}\phi_{i-1\,jk}
-    - \sigm_{j+1}\frac{\Delta_{j+1}}{\Dj}\phi_{i\,j+1\,k} \\
-    &- \sigp_{j-1}\frac{\Delta_{j-1}}{\Dj}\phi_{i\,j-1\,k} 
-    - \sigm_{k+1}\frac{\Delta_{k+1}}{\Dk}\phi_{ij\,k+1}
+    - \sigp_{i-1}\frac{\Delta_{i-1}}{\Di}\phi_{i-1\,jk}\\
+    &- \sigm_{j+1}\frac{\Delta_{j+1}}{\Dj}\phi_{i\,j+1\,k}
+    - \sigp_{j-1}\frac{\Delta_{j-1}}{\Dj}\phi_{i\,j-1\,k}\\ 
+    &- \sigm_{k+1}\frac{\Delta_{k+1}}{\Dk}\phi_{ij\,k+1}
     - \sigp_{k-1}\frac{\Delta_{k-1}}{\Dk}\phi_{ij\,k-1}
@@
 \end{equation}
+
 Here,
+
 \begin{equation}
@@
 \end{equation}
+
 and
+
 \begin{subequations}
@@
 \end{subequations}
+
 Also, note that
+
 \begin{align*}
@@
 \end{align*}
+
 Equations~(\ref{eq:discrete-diffusion}) through (\ref{eq:sigmas})
@@
 conditions.  For this work, the Marshak Boundary condition suffices,
+
 \begin{equation}
@@
 \end{equation}
+
 where $f_b$ is the incoming partial current on the problem boundary,
+
 $\partial\Gamma$.  On any low boundary we can write the flux using a
 first-order discrete form of Fick's Law,
+
 \begin{equation}
@@
 \end{equation}
+
 As before, we enforce continuity of flux on the boundary using the
 Marshak Boundary condition,
+
 \begin{equation}
@@
 \end{equation}
+
 Solving for $\phi_{1/2}$ and plugging into
@@
 boundary flux,
+
 \begin{equation}
@@
 \end{equation}
+
 Following an identical procedure on high-side problem boundaries gives
+
 \begin{equation}
@@
 \end{equation}
+
 where $l\in\{i,j,k\}$ is defined over the range $[1,N_l]$.
@@
 modifications to Eqs.~(\ref{eq:sigmas}) on a boundary,
+
 \begin{align}
@@
 \end{align}
+
 Furthermore, the following contributions are added to the source,
+
 $\qn_{ijk}$ on a boundary cell,
@@
 \end{align}
+
 Note that for vacuum boundary conditions, $f_b^{\pm} = 0$.
@@
 that
+
 \begin{align}
@@
 \end{align}
+
 Thus, there are no leakage or adjacent cell source terms resulting
@@
 be written in matrix-operator form as follows,
+
 \begin{equation}
@@
 \end{equation}
+
 where $\ve{D}$ is a symmetric, positive-definite (SPD) operator.
@@
 MCSA method. We apply left-preconditioning and write
+
 \begin{equation}
@@
 \end{equation}
-A good choice for the preconditioner, $\ve{M}$, is
+
+A good choice for the preconditioner, $\ve{M}$, is to use Jacobi
+preconditioning:
+
 \begin{equation}
@@
 \end{equation}
+
 Setting $\ve{M}$ equal to the diagonal elements of $\ve{D}$
 accomplishes several objectives:
+
 \begin{itemize}
@@
   zero, $\diag(\vI - \ve{M}^{-1}\ve{D}) = 0$.  Thus, there will be no
-  time wasted in transitions $i\rightarrow i$ during the random walk
-  process.
+  computations wasted in transitions $i\rightarrow i$ during the
+  random walk process.
 \end{itemize}
+
 The second item is quite important for Monte Carlo solutions of linear
@@
 rectified by doing right-left preconditioning and defining
+
 \begin{equation}
@@
 \end{equation}
+
 where $\ve{L} = \ve{L}^{T} = \sqrt{\ve{M}}$.  Applying right-left
 preconditioning gives
+
 \begin{equation}
@@
 \end{equation}
+
 This type of preconditioning is also referred to as row-column
@@
 Eq.~(\ref{eq:discrete-diffusion}) yields
+
 \begin{equation}
@@
     - \frac{\sigm_{i+1}}{\sigT}\frac{\Delta_{i+1}}{\Di}&\phi_{i+1\,jk}
-    - \frac{\sigp_{i-1}}{\sigT}\frac{\Delta_{i-1}}{\Di}\phi_{i-1\,jk}
-    - \frac{\sigm_{j+1}}{\sigT}\frac{\Delta_{j+1}}{\Dj}\phi_{i\,j+1\,k} \\
-    &-
-    \frac{\sigp_{j-1}}{\sigT}\frac{\Delta_{j-1}}{\Dj}\phi_{i\,j-1\,k}
-    - \frac{\sigm_{k+1}}{\sigT}\frac{\Delta_{k+1}}{\Dk}\phi_{ij\,k+1}
+    - \frac{\sigp_{i-1}}{\sigT}\frac{\Delta_{i-1}}{\Di}\phi_{i-1\,jk}\\
+    &- \frac{\sigm_{j+1}}{\sigT}\frac{\Delta_{j+1}}{\Dj}\phi_{i\,j+1\,k}
+    -\frac{\sigp_{j-1}}{\sigT}\frac{\Delta_{j-1}}{\Dj}\phi_{i\,j-1\,k}\\
+    &- \frac{\sigm_{k+1}}{\sigT}\frac{\Delta_{k+1}}{\Dk}\phi_{ij\,k+1}
     - \frac{\sigp_{k-1}}{\sigT}\frac{\Delta_{k-1}}{\Dk}\phi_{ij\,k-1}
@@
 \end{equation}
+
 In the next section we will derive an MCSA method to solve this
@@
     \frac{\sigm_{i+1}}{\sigT}\frac{\Delta_{i+1}}{\Di}&\phi_{i+1\,jk}^l
-    + \frac{\sigp_{i-1}}{\sigT}\frac{\Delta_{i-1}}{\Di}\phi_{i-1\,jk}^l
     +
-    \frac{\sigm_{j+1}}{\sigT}\frac{\Delta_{j+1}}{\Dj}\phi_{i\,j+1\,k}^l \\ 
-    &+ \frac{\sigp_{j-1}}{\sigT}\frac{\Delta_{j-1}}{\Dj}\phi_{i\,j-1\,k}^l 
-    + \frac{\sigm_{k+1}}{\sigT}\frac{\Delta_{k+1}}{\Dk}\phi_{ij\,k+1}^l
+    \frac{\sigp_{i-1}}{\sigT}\frac{\Delta_{i-1}}{\Di}\phi_{i-1\,jk}^l \\
+    &+ \frac{\sigm_{j+1}}{\sigT}\frac{\Delta_{j+1}}{\Dj}\phi_{i\,j+1\,k}^l 
+    + \frac{\sigp_{j-1}}{\sigT}\frac{\Delta_{j-1}}{\Dj}\phi_{i\,j-1\,k}^l \\
+    &+ \frac{\sigm_{k+1}}{\sigT}\frac{\Delta_{k+1}}{\Dk}\phi_{ij\,k+1}^l
     + \frac{\sigp_{k-1}}{\sigT}\frac{\Delta_{k-1}}{\Dk}\phi_{ij\,k-1}^l
@@
 \S~\ref{sec:adjoint-method} is used to estimate
+
 \begin{equation}
@@
 \end{equation}
+
 As noted above, the residual acts as the source for this simulation,
@@
 adjacent cell.  Using Eq.~(\ref{eq:preconditioned-discrete-diffusion})
-we can explicitly write the equations from each adjacent cell,
+we can explicitly write the equations from each adjacent cell, here
+only noting the contribution from cell 7:
+
 \begin{align*}
@@
   %%
-  \ldots\:.
 \end{align*}
+
 A quick view of these equations reveals that the only non-zero entries
 in column 7 of the iteration matrix are
+
 \begin{equation*}
@@
 \end{equation*}
+
 where, following Eq.~(\ref{eq:point_iteration}), $\ve{H} = (\vI -
@@
 probabilities for cell 7 as
+
 \begin{equation*}
@@
 \end{equation*}
+
 The weight change for each scattering event is calculated using
 Eq.~(\ref{eq:adjoint-weight}),
+
 \begin{equation*}
@@
 \end{equation*}
+
 where $m$ is the step-index in the particle history.  We also note
 that, for the 3D Cartesian mesh considered in this work, the maximum
-number of non-zero entries in columns of the iteration matrix is 6.
-Cells along the problem boundary can have fewer than 6 entries.
+number of non-zero entries in columns of the iteration matrix is 6,
+generating a sparse system.  Cells along the problem boundary can have
+fewer than 6 entries.
 
@@
 The Marshak wave problem under consideration in this study has an
-incoming 1~keV blackbody boundary flux at $x=0$ and propagates
-in the $x$-dimension in time.  In other words, this a 3D
-representation of a 1D problem.  The full set of problem conditions
-is:
+incoming 1~keV blackbody boundary flux at $x=0$ and propagates in the
+$x$-dimension in time.  In other words, this a 3D representation of a
+1D problem.  The full set of problem conditions is:
+
 \begin{center}
@@
 \end{center}
+
 The problem is run to $10$~shakes.
@@
 the requested number of particles per stage for a relative weight
-cutoff of $\sn{1}{-4}$.  For less than 5 particles per stage teh
+cutoff of $\sn{1}{-4}$.  For less than 5 particles per stage the
 method does not converge.  Figure~\ref{fig:CPU_Ns} shows the variation
@@
 accelerates Richardson iteration, as indicated by the fact that MCSA
-is 42\% faster than Richards iteration alone.
+is 42\% faster than Richardson iteration alone.
 
@@
 conditions is:
+
 \begin{center}
@@
 \end{center}
+
 A 2D plot of the temperature at 100~sh elapsed time is illustrated in
@@
 We have presented a new Monte Carlo solution method for solving the
-discrete, time-dependent diffusion equation in 3D.  The MCSA method
-has been shown to match results using standard solution techniques to
-arbitrary precision.  Also, the new method is faster than
-preconditioned CG and Richardson iteration.
+discrete, time-dependent radiation diffusion equation in 3D.  The MCSA
+method has been shown to match results using standard solution
+techniques to arbitrary precision.  Also, the new method has been to
+shown to be faster than preconditioned CG and Richardson iteration for
+the model problems presented here.
 

### 2012-07-19 doc/jcomp2012/jcomp2012.tex
Converted eps files to pdf files for compatibility with linux pdflatex.

@@
 %% or use the graphicx package for more complicated commands
+\usepackage{epsfig}
 \usepackage{graphicx}
 \usepackage{subfigure}
-%% or use the epsfig package if you prefer to use the old commands
-%% \usepackage{epsfig}
+%% or use the.pdffig package if you prefer to use the old commands
+%% \usepackage.pdffig}
 
@@
 \begin{equation}
-  \|\vr\|_{\infty} < \epsilon\cdot\|\vb\|_\infty\:.
+  \|\vr\|_{\infty} < .pdfilon\cdot\|\vb\|_\infty\:.
   \label{eq:stopping-criteria}
@@
   \centering
-  \includegraphics[width=5in,clip]{Adjoint_Direct_CPU_Time.eps}
+  \includegraphics[width=5in,clip]{Adjoint_Direct_CPU_Time.pdf}
   \caption{\sl CPU Time (s) to converge vs. Problem Size ($N$ for an
@@
   \centering
-  \includegraphics[width=5in,clip]{Adjoint_Direct_Iterations.eps}
+  \includegraphics[width=5in,clip]{Adjoint_Direct_Iterations.pdf}
   \caption{\sl Iterations to converge vs. Problem Size ($N$ for an $N
@@
   \centering
-  \includegraphics[width=5in,clip]{Adjoint_Direct_Convergence.eps}
+  \includegraphics[width=5in,clip]{Adjoint_Direct_Convergence.pdf}
   \caption{\sl Infinity norm of the solution residual vs. iteration
@@
 method.  As discussed in \S~\ref{sec:iter-refin-monte}, the MCSA
-method requires two solution steps:
+method requires two solution s.pdf:
 
@@
 method only approximately inverts $\ve{M}^{-1}\ve{D}$.  We now
-describe each of these steps in turn.
+describe each of these s.pdf in turn.
 
@@
 relative weight cutoff that is defined in
-Eq.~(\ref{eq:weight_cutoff}).  We now examine each of these steps in
+Eq.~(\ref{eq:weight_cutoff}).  We now examine each of these s.pdf in
 more detail.
@@
   \centerline{
-    \includegraphics[width=5in,clip]{marshak_10.eps}}
+    \includegraphics[width=5in,clip]{marshak_10.pdf}}
   \caption{
@@
   \centerline{
-    \includegraphics[width=5in,clip]{spectral_radius.eps}}
+    \includegraphics[width=5in,clip]{spectral_radius.pdf}}
   \caption{Spectral radius of the iteration matrix, $\ve{H}$, for the
@@
   \centerline{
-    \includegraphics[width=5in,clip]{mrshk_np_CPU.eps}}
+    \includegraphics[width=5in,clip]{mrshk_np_CPU.pdf}}
   \caption{CPU time versus number of particles per stage for the
@@
   \centerline{
-    \includegraphics[width=5in,clip]{mrshk_np_Ns.eps}}
+    \includegraphics[width=5in,clip]{mrshk_np_Ns.pdf}}
   \caption{
@@
   \centerline{
-    \includegraphics[width=5in,clip]{mrshk_wc_CPU.eps}}
+    \includegraphics[width=5in,clip]{mrshk_wc_CPU.pdf}}
   \caption{
@@
   \centerline{
-    \includegraphics[width=5in,clip]{mesh.eps}}
+    \includegraphics[width=5in,clip]{mesh.pdf}}
   \caption{Multi-material problem mesh.  The duct region is blue.  The
@@
   \centerline{
-    \includegraphics[width=5in,clip]{T_multi_mat.eps}}
+    \includegraphics[width=5in,clip]{T_multi_mat.pdf}}
   \caption{
@@
     \begin{tabular}{cc}
-      \subfigure[Edit points]{\includegraphics[width=1.5in,clip]{points.eps}} &
-      \subfigure[Temperature]{\includegraphics[width=3.5in,clip]{multi_mat.eps}}
+      \subfigure[Edit points]{\includegraphics[width=1.5in,clip]{points.pdf}} &
+      \subfigure[Temperature]{\includegraphics[width=3.5in,clip]{multi_mat.pdf}}
     \end{tabular}

### 2012-07-19 doc/jcomp2012/jcomp2012.tex
first draft done for mcsa jcomp

@@
 %%---------------------------------------------------------------------------%%
-%% Monte Carlo Synthetic Acceleration Paper
+%% Monte Carlo Synthetic-Acceleration Paper
 %% Submitted to Journal of Computational Physics 2012
@@
 
-  \title{A Monte Carlo Synthetic Acceleration Method for Solving the
+  \title{A Monte Carlo Synthetic-Acceleration Method for Solving the
     Thermal Radiation Diffusion Equation}
@@
   \address[ornl]{Oak Ridge National Laboratory, 
-    1 Bethel Valley Rd., Oak Ridge, TN 378131 U.S.A.}
+    1 Bethel Valley Rd., Oak Ridge, TN 37831 U.S.A.}
   \address[wisc]{University of Wisconsin-Madison, 1500 Engineering
@@
     We present a synthetic-acceleration based Monte Carlo method for
-    solving the 1T thermal radiation diffusion equation.  We show that
-    this method can be an effective solver for sparse matrix systems
-    and is competitive with deterministic methods.  We also discuss
-    its general applicability to broader classes of problems.
+    solving the 1T thermal radiation diffusion equation in 3
+    dimensions. We show that this method can be an effective solver
+    for sparse matrix systems and is competitive with deterministic
+    methods including preconditioned Conjugate Gradient. We also
+    discuss various aspects of preconditioning and its general
+    applicability to broader classes of problems.
 
@@
 The purpose of this study is to demonstrate an efficient Monte Carlo
-solution method for 3D, time-dependent discrete systems. In this
+solution method for 3D, time-dependent, discrete systems. In this
 work, we extend our previous work in the following manner:
@@
   for solving sparse matrix systems.
-\item We apply this solver to the 1T thermal radiation diffusion
+\item We apply this method to the 1T thermal radiation diffusion
   equation in 3D.
@@
 radiation diffusion equation and its discrete form will be derived in
-\S~\ref{sec:discr-form-radi}.  We discuss using Synthetic-Accleration
-Monte Carlo to solve the discrete radiation diffusion equation in
-\S~\ref{sec:solut-radi-diff}.  Results from two 3D test problems are
-given in \S~\ref{sec:results}.  We finish with conclusions and ideas
-for future work in \S~\ref{sec:conclusions}.
+\S~\ref{sec:discr-form-radi}.  We discuss using Monte Carlo
+Synthetic-Accleration to solve the discrete radiation diffusion
+equation in \S~\ref{sec:solut-radi-diff}.  Results from two 3D test
+problems are given in \S~\ref{sec:results}.  We finish with
+conclusions and ideas for future work in \S~\ref{sec:conclusions}.
 
@@
 random walk.  Let $X$ be a random variable sampled from a random walk
-with $k$ events (particles) that initiates in state $i$,
+with $k$ events (particles in the context of radiative transfer) that
+initiates in state $i$,
 
@@
 and each particle starts with unit weight. At every step,
-contributions to each component of $X$ are provided only by the
-starting state of the particle, $b_{i_m}$. Therefore, for a given
-random walk permutation, each state must be used as a starting value
-in order to calculated its contribution to the expectation value
-(i.e. a state of size $N$ will have $N$ particles per random walk
-permutation). When the particle transitions between states, for
-example $i\rightarrow j$, the particle weight is multiplied by the
-transition factor.
+contributions to each component of $X$ are generated only in the
+starting state of the particle, $i_0$, by the source term in the
+current state, $b_{i_m}$. Therefore, for a given random walk
+permutation, each state must be used as a starting value in order to
+calculate its component of the expectation value (i.e. a state of size
+$N$ will have $N$ particles per random walk permutation). When the
+particle transitions between states, for example $i\rightarrow j$, the
+particle weight is multiplied by the transition factor.
 
@@
 
-We are left to define the transition probabilities, $p_{ij}$. The most
+We are left to define the transition probabilities. The most
 straightforward approach is to set
@@
 cutoff, $W_c$.  Generally, we choose to terminate random walks using a
-weight cutoff as opposed to augmenting the matrix sucth that the
-random walk terminates when $W_m < W_c$.
+weight cutoff as opposed to augmenting the matrix such that the random
+walk terminates when $W_m < W_c$.
 
@@
 
-Similar to the Direct Method, the Adjoint Method random walk process
+Similar to the direct method, the adjoint method random walk process
 requires a terminating condition.  In all of the work that follows we
@@
 
-where $W_f$ is the terminating weight of the random walk, $W_c$ is the
-input relative weight cutoff, and $\hat{b}_{i_0}$ is the initial
-weight of the source particle.  The random walk terminates on the
+where $W_f$ is the terminating weight of the random walk and $W_c$ is
+the input relative weight cutoff. The random walk terminates on the
 $m^\text{th}$ step if $W_m < W_f$.
@@
 %%---------------------------------------------------------------------------%%
-\section{Synthetic-Acceleration Monte Carlo}
+\section{Monte Carlo Synthetic-Acceleration}
 \label{sec:iter-refin-monte}
@@
 We propose a modification of this scheme that uses Monte Carlo as a
-synthetic-acceleration scheme for fixed-point iteration. We begin by
-subtracting Eq.~(\ref{eq:richardson}) from Eq:~(\ref{eq:point_iteration}):
+synthetic-acceleration for the fixed-point iteration given by
+Eq.~(\ref{eq:richardson}). We begin by subtracting
+Eq.~(\ref{eq:richardson}) from Eq:~(\ref{eq:point_iteration}):
 
@@
 is defined as the error at iteration $l+1$. We then subtract
-$(\vI-\vA) \delta \vx^{l+1}$ from Eq.~(\ref{eq:synthetic_error}) giving:
-
-\begin{equation}
-  \vA \delta \vx^{l+1} = (\vI - \vA)(\vx^{l+1} - \vx^{l})\:.
+$(\vI-\vA) \delta \vx^{l+1}$ from Eq.~(\ref{eq:synthetic_error}),
+giving a linear system to solve for the error:
+
+\begin{equation}
+  \begin{split}
+    \vA \delta \vx^{l+1} &= (\vI - \vA)(\vx^{l+1} - \vx^{l}) \\
+    &= \vr^{l+1}\:.
+  \end{split}
   \label{eq:synthetic_error2}
@@
 
-The following iterative scheme will then converge in one iteration:
+The following scheme will then converge in one iteration:
 
@@
     %%
-    \vA \delta \vx^{l+1} = (\vI - \vA)(\vx^{l+1} - \vx^{l})\:.
-    \label{eq:synthetic-residual_solve}\\ 
+    \vA \delta \vx^{l+1} = \vr^{l+1}
+    \label{eq:synthetic-residual-solve}\\ 
     %% 
@@
 
-To accelerate the fixed point method, we can instead create a new
-iterative scheme by approximating the operator $\vA$ using the Monte
-Carlo method to estimate the solution error. With this idea hand, the
-Fixed-Point Monte Carlo Synthetic-Acceleration (MCSA) method can be
-defined as follows:
+To accelerate the fixed-point method and avoid inverting $\vA$
+directly in Eq.~(\ref{eq:synthetic-residual-solve}), we can instead
+use the above ideas to create a new iterative scheme by using the
+Monte Carlo method to estimate the solution error. With this in hand,
+the Fixed-Point Monte Carlo Synthetic-Acceleration (MCSA) method can
+be defined as follows:
 
@@
 fixed-point iteration\footnote{Although, a Krylov method could be used
-  here.}.  The residual is calculated and is used as a source to
-estimate the error, $\delta\vx^{l+1/2}$, by solving
-Eq.~(\ref{eq:MCSA-residual_solve}) via the Monte Carlo adjoint
+  instead of a stationary method here.}.  The residual is calculated
+and is used as a source to estimate the error, $\delta\vx^{l+1/2}$, by
+solving Eq.~(\ref{eq:MCSA-residual_solve}) via the Monte Carlo adjoint
 method. The error is then used to calculate an updated iterate of the
@@
 \begin{equation}
-  \|\vr\|_{\infty} < .pdfilon\cdot\|\vb\|_\infty\:.
+  \|\vr\|_{\infty} < \epsilon\cdot\|\vb\|_\infty\:.
   \label{eq:stopping-criteria}
@@
 solver and a convergence tolerance of $1.0E-8$ for the MCSA iterative
-solver. All computations were completed on a 2.66 GHz Intel Core i7
-machine with 8 GB 1067 MHz DDR3 memory used for all computations
-presented in this section. Fig.~\ref{fig:poisson_cpu_time} gives the
-CPU time needed for each case to converge in seconds and
-Fig.~\ref{fig:poisson_iterations} gives the number of iterations
-needed for each case to converge relative to the problem size.
-
-\begin{figure}[htpb!]
+solver. All computations presented in this section were completed on a
+2.66 GHz Intel Core i7 machine with 8 GB 1067 MHz DDR3 memory
+. Fig.~\ref{fig:poisson_cpu_time} gives the CPU time needed for each
+case to converge in seconds and Fig.~\ref{fig:poisson_iterations}
+gives the number of iterations needed for each case to converge
+relative to the problem size.
+
+\begin{figure}[h!]
   \centering
@@
   \caption{\sl CPU Time (s) to converge vs. Problem Size ($N$ for an
-    $N \times N$ mesh). Both the adjoint and direct solvers are used
-    with the five point and nine point stencils. A significant CPU
-    time speedup is noted with the adjoint method due to the reduced
-    number of histories.}
+    $N \times N$ square mesh). Both the adjoint and direct solvers are
+    used with the five point and nine point stencils. A significant
+    CPU time speedup is noted with the adjoint method due to the
+    reduced number of random walk events.}
   \label{fig:poisson_cpu_time}
@@
 
-\begin{figure}[htbp!]
+\begin{figure}[h!]
   \centering
@@
   \caption{\sl Iterations to converge vs. Problem Size ($N$ for an $N
-    \times N$ mesh. Both the adjoint and direct solvers are used with
-    the five point and nine point stencils. Both methods give similar
-    iteration behavior.}
+    \times N$ square mesh. Both the adjoint and direct solvers are
+    used with the five point and nine point stencils. Both methods
+    give similar iteration behavior.}
   \label{fig:poisson_iterations}
@@
 both solvers and a system of size $N \times N$, the direct solver will
-compute $N \times N$ random walks per permuation while the adjoint solver
-will only compute one. Initially, this may not seem like a fair
-comparison, however, we see from Fig.~\ref{fig:poisson_cpu_time} that
-the number of iterations requried to converge is approximately the
-same. This is necessary in the direct method to ensure a contribution
-from each state as the random walk sequence will only contribute to
-the starting state. For the adjoint method, because the random walk
-sequence contributes to the state in which it currently resides, fewer
-histories are neccessary to compute contributions from all the nonzero
-parts of the source. Therefore, although the total number of random
-walks per iteration is in fact larger for the direct method, the
-solution computed at each iteration is no better than that computed by
-the adjoint method.
+compute $N \times N$ random walks per permuation while the adjoint
+solver will only compute one. Initially, this may not seem like a fair
+comparison, however, we see from Fig.~\ref{fig:poisson_iterations}
+that the number of iterations requried to converge is approximately
+the same. This is necessary in the direct method to ensure a
+contribution from each state as the random walk sequence will only
+contribute to the starting state. For the adjoint method, because the
+random walk sequence contributes to the state in which it currently
+resides, fewer histories are neccessary to compute contributions from
+all the nonzero parts of the source. Therefore, although the total
+number of random walks per iteration is in fact larger for the direct
+method, the solution computed at each iteration is no better than that
+computed by the adjoint method.
 
@@
 
-\begin{figure}[htpb!]
+\begin{figure}[h!]
   \centering
@@
 
-\begin{figure}[htpb!]
+\begin{figure}[h!]
   \centerline{
     \input{mesh_cell.pdftex_t}}
-  \caption{Orthogonal 3D mesh Cell.}
+  \caption{Orthogonal 3D mesh cell.}
   \label{fig:cell}
@@
 \end{equation}
-
-In the next section we will derive an MCSA method to solve this
-equation.
 
@@
 
-\begin{figure}[htpb!]
+\begin{figure}[h!]
   \centerline{
@@
 
-\begin{table}[htpb!]
+\begin{table}[h!]
   \begin{center}
@@
 
-\begin{figure}[htpb!]
+\begin{figure}[h!]
   \begin{center}
     \begin{tabular}{cc}
-      \subfigure[Edit points]{\includegraphics[width=1.5in,clip]{points.pdf}} &
-      \subfigure[Temperature]{\includegraphics[width=3.5in,clip]{multi_mat.pdf}}
+      \subfigure[Edit points]{\includegraphics[width=2in,clip]{points.pdf}} &
+      \subfigure[Temperature]{\includegraphics[width=4in,clip]{multi_mat.pdf}}
     \end{tabular}
@@
 
-\begin{table}[htpb!]
+\begin{table}[htbp!]
   \begin{center}

### 2012-07-19 doc/jcomp2012/jcomp2012.tex
updated jcomp abstract
updates to section 3

@@
     methods including preconditioned Conjugate Gradient. We also
-    discuss various aspects of preconditioning and its general
-    applicability to broader classes of problems.
+    discuss various aspects of preconditioning the method and its
+    general applicability to broader classes of problems.
 
@@
 \S~\ref{sec:direct-method}. To demonstrate the effectiveness of the
-adjoint method over the direct method, we choose the two-dimensional
-time-dependent Poisson equation as a simple model problem:
+adjoint method over the direct method within the context of MCSA, we
+choose the two-dimensional time-dependent Poisson equation as a simple
+model problem:
 
@@
 compute $N \times N$ random walks per permuation while the adjoint
-solver will only compute one. Initially, this may not seem like a fair
-comparison, however, we see from Fig.~\ref{fig:poisson_iterations}
-that the number of iterations requried to converge is approximately
-the same. This is necessary in the direct method to ensure a
-contribution from each state as the random walk sequence will only
-contribute to the starting state. For the adjoint method, because the
-random walk sequence contributes to the state in which it currently
-resides, fewer histories are neccessary to compute contributions from
-all the nonzero parts of the source. Therefore, although the total
-number of random walks per iteration is in fact larger for the direct
-method, the solution computed at each iteration is no better than that
-computed by the adjoint method.
-
-Although CPU time is shown to be greatly decreased by using the
-adjoint solver in Fig.~\ref{fig:poisson_cpu_time}, the iteration
-behavior was approximately the same. To assess the convergence
-properties of MCSA using each solver and stencil, the infinity norm of
-the residual computed in Eq.~(\ref{eq:general-MCSA}) was collected at
-each iteration for a fixed problem
-size. Fig.~\ref{fig:poisson_convergence} gives the results of these
-computations. First, it is worthy to note on the semilog plot that we
-are indeed achieving the expected exponential convergence from
-MCSA. Second, we note that both the direct and adjoint methods have
-approximately equivalent convergence behavior, leading to CPU time
-being the dominating factor in method selection.
+solver will only compute one. This is necessary in the direct method
+to ensure a contribution from each state as the random walk sequence
+will only contribute to the starting state. For the adjoint method,
+because the random walk sequence contributes to the state in which it
+currently resides, fewer histories are neccessary to compute
+contributions from all the nonzero parts of the source. Initially,
+this may not seem like a fair comparison, however, we see from
+Fig.~\ref{fig:poisson_iterations} that the number of iterations
+requried to converge is approximately the same and therefore the extra
+computations performed by the direct method do not improve the error
+estimation. Therefore, although the total number of random walks per
+iteration is in fact larger for the direct method, the solution
+computed at each iteration is no better than that computed by the
+adjoint method.
+
+As an additional comparision, the convergence behavior of MCSA can be
+analyzed using both solvers to detect any performance benefits. To
+assess the convergence properties of MCSA using each solver and
+stencil, the infinity norm of the residual computed in
+Eq.~(\ref{eq:MCSA-residual}) was collected at each iteration for a
+fixed problem size. Fig.~\ref{fig:poisson_convergence} gives the
+results of these computations. First, it is worthy to note on the
+semilog plot that we are indeed achieving the expected exponential
+convergence from MCSA. Second, we note that both the direct and
+adjoint methods have approximately equivalent convergence behavior,
+leading to CPU time being the dominating factor in method selection.
 

### 2012-07-20 doc/model_problems/model_problems.tex
working on model problem doc

%        File: model_problems.tex
%
\documentclass[letterpaper,12pt]{article}
\usepackage[top=1.0in,bottom=1.0in,left=1.0in,right=1.0in]{geometry}
\usepackage{verbatim}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage[usenames]{color}
\usepackage[
naturalnames = true, 
colorlinks = true, 
linkcolor = black,
anchorcolor = black,
citecolor = black,
menucolor = black,
urlcolor = blue
]{hyperref}

%%---------------------------------------------------------------------------%%
\author{Stuart R. Slattery
\\ \href{mailto:sslattery@wisc.edu}{\texttt{sslattery@wisc.edu}}
}

\date{\today}
\title{A Finite Volume Formulation for the 2D Incompressible
  Navier-Stokes and Conduction Equations}
\begin{document}
\maketitle

%%---------------------------------------------------------------------------%%
\section{Computational Grid}

%%---------------------------------------------------------------------------%%
\section{Conduction Equation}
We begin the derivation of the conduction equation by defining the
conserved quantity, energy, in terms of temperature:

\begin{equation}
  E(x,y,t) = C_p(x,y) T(x,y,t)\:,
  \label{eq:conduction_energy}
\end{equation}

where $C_P$ is the specific heat capacity of the material. We also
define the heat flux using Fick's law:

\begin{equation}
  F(T(x,y,t)) = -\beta(x,y) \nabla T(x,y,t)\:,
  \label{eq:conduction_ficks_law}
\end{equation}

where $\beta$ is the thermal conductivity of the material. With these
two relations in hand, we can then write a conservation law for energy
\cite{leveque_2002}:

\begin{equation}
  \frac{d}{dt} \int \int_{C_{ij}} E(x,y,t) dx dy = -F(T(x,y,t)) |_{B_{ij}}\:.
  \label{eq:conduction_conservation}
\end{equation}

Here we balance the change in energy in the unit cell over a unit of
time with the flux on the cell boundaries. Inserting
Eq.~(\ref{eq:conduction_energy}) int
Eq.~(\ref{eq:conduction_conservation}) and dividing by the cell volume
recovers the cell averaged temperature and specific heat on the left
hand side:

\begin{equation}
  \frac{d}{dt} C_{P_{ij}} T_{ij} = -\frac{1}{\Delta_i
    \Delta_j}F(T(x,y,t)) |_{B_{ij}}\:.
  \label{eq:conduction_cell_averaged}
\end{equation}

We then insert Eq.~(\ref{eq:conduction_ficks_law}), difference the
left hand side, and integrate the right hand side over a unit time
step, maintaining the conservation properties of the system:

\begin{equation}
  C_{P_{ij}} ( T^{n+1}_{ij} - T^n_{ij} ) = \frac{1}{\Delta_i \Delta_j}
  \int_{t^n}^{t^{n+1}} [ -\beta_{i-1/2,j} \nabla T_{i-1/2,j} +
    \beta_{i+1/2,j} \nabla T_{i+1/2,j} - \beta_{i,j-1/2} \nabla
    T_{i,j-1/2} +\beta_{i,j+1/2} \nabla T_{i,j+1/2} ] dt\:.
  \label{eq:conduction_time_integral}
\end{equation}

We choose to approximate the integral on the right hand side and use a
first order central difference for the derivative in the heat flux
terms giving a backwards Euler representation for the time
integration:

\begin{equation}
  C_{P_{ij}}( T^{n+1}_{ij} - T^n_{ij} ) = \frac{\Delta_t}{\Delta_i
    \Delta_j} -\beta_{i-1/2,j} \frac{T^{n+1}_{ij}
    -T^{n+1}_{i-1,j}}{\Delta_i} +\beta_{i+1/2,j} \frac{T^{n+1}_{i+1,j}
    -T^{n+1}_{ij}}{\Delta_i} -\beta_{i,j-1/2} \frac{T^{n+1}_{ij}
    -T^{n+1}_{i,j-1}}{\Delta_j} +\beta_{i,j+1/2} \frac{T^{n+1}_{i,j+1}
    -T^{n+1}_{i,j-1}}{\Delta_j}
  \label{eq:conduction_fv_equation}
\end{equation}

\subsection{Boundary Conditions}

%%---------------------------------------------------------------------------%%
\section{Navier-Stokes Equations}

\subsection{Boundary Conditions}

%%---------------------------------------------------------------------------%%
\pagebreak
\bibliographystyle{ieeetr} \bibliography{references}
\end{document}

### 2012-07-22 doc/model_problems/model_problems.tex
Adding thesis proposal document.

@@
 \documentclass[letterpaper,12pt]{article}
-\usepackage[top=1.0in,bottom=1.0in,left=1.0in,right=1.0in]{geometry}
+\usepackage[top=1.0in,bottom=1.0in,left=1.25in,right=1.25in]{geometry}
 \usepackage{verbatim}
-\usepackage{amssymb}
 \usepackage{graphicx}
@@
 \usepackage{amsmath}
+\usepackage{amsthm}
+\usepackage{amssymb}
+\usepackage{tmadd,tmath}
+\usepackage[mathcal]{euscript}
 \usepackage[usenames]{color}
@@
 where $C_P$ is the specific heat capacity of the material. We also
-define the heat flux using Fick's law:
+define the heat flux, $F$, in terms of temperature using Fick's law:
 
@@
 
-Here we balance the change in energy in the unit cell over a unit of
-time with the flux on the cell boundaries. Inserting
+Here we balance the time rate of change in energy in the unit cell
+with the flux on the cell boundaries. Inserting
 Eq.~(\ref{eq:conduction_energy}) int
@@
 We then insert Eq.~(\ref{eq:conduction_ficks_law}), difference the
-left hand side, and integrate the right hand side over a unit time
-step, maintaining the conservation properties of the system:
+left hand side, and integrate, maintaining the conservation properties
+of the system:
 
-\begin{equation}
+\begin{multline}
   C_{P_{ij}} ( T^{n+1}_{ij} - T^n_{ij} ) = \frac{1}{\Delta_i \Delta_j}
-  \int_{t^n}^{t^{n+1}} [ -\beta_{i-1/2,j} \nabla T_{i-1/2,j} +
-    \beta_{i+1/2,j} \nabla T_{i+1/2,j} - \beta_{i,j-1/2} \nabla
+  \int_{t^n}^{t^{n+1}} [ -\beta_{i-1/2,j} \nabla T_{i-1/2,j}\\
+    + \beta_{i+1/2,j} \nabla T_{i+1/2,j} - \beta_{i,j-1/2} \nabla
     T_{i,j-1/2} +\beta_{i,j+1/2} \nabla T_{i,j+1/2} ] dt\:.
   \label{eq:conduction_time_integral}
-\end{equation}
+\end{multline}
 
@@
 
-\begin{equation}
+\begin{multline}
   C_{P_{ij}}( T^{n+1}_{ij} - T^n_{ij} ) = \frac{\Delta_t}{\Delta_i
@@
     -T^{n+1}_{i-1,j}}{\Delta_i} +\beta_{i+1/2,j} \frac{T^{n+1}_{i+1,j}
-    -T^{n+1}_{ij}}{\Delta_i} -\beta_{i,j-1/2} \frac{T^{n+1}_{ij}
+    -T^{n+1}_{ij}}{\Delta_i}\\
+  -\beta_{i,j-1/2} \frac{T^{n+1}_{ij}
     -T^{n+1}_{i,j-1}}{\Delta_j} +\beta_{i,j+1/2} \frac{T^{n+1}_{i,j+1}
-    -T^{n+1}_{i,j-1}}{\Delta_j}
-  \label{eq:conduction_fv_equation}
+    -T^{n+1}_{i,j-1}}{\Delta_j}\:.
+  \label{eq:conduction_fv_multline}
+\end{multline}
+
+We will define $\alpha$ as:
+
+\begin{equation}
+  \alpha_{ij} = \frac{\Delta_t}{C_{P_{ij}} \Delta_i \Delta_j}
+  \label{eq:conduction_coeff}\:.
 \end{equation}
 
+Rearranging for temperature, we can then form a system of equations
+amenable to matrix solutions, not yet considering boundary conditions,
+arriving at the discrete, time-dependent conduction equation:
+
+\begin{multline}
+ \Big[1 + \frac{\alpha_{ij}}{\Delta_i}(\beta_{i-1/2,j} + \beta_{i+1/2,j})\Big]
+ T^{n+1}_{ij} + \Big[-\frac{\alpha_{ij}}{\Delta_i} \beta_{i-1/2,j}\Big]
+ T^{n+1}_{i-1,j} + \Big[\frac{-\alpha_{ij}}{\Delta_i} \beta_{i+1/2,j}\Big]
+ T^{n+1}_{i+1,j}\\
+ + \Big[-\frac{\alpha_{ij}}{\Delta_j} \beta_{i-1/2,j}\Big]
+ T^{n+1}_{i,j-1} + \Big[-\frac{\alpha_{ij}}{\Delta_j} \beta_{i+1/2,j}\Big]
+ T^{n+1}_{i,j+1} = T^n_{ij}\:.
+ \label{eq:conduction_inner_terms}
+\end{multline}
+
 \subsection{Boundary Conditions}
+
+We will consider both Neumann and Dirichlet type boundary conditions
+for the conduction equation. For the Neumann condition we have a
+constant temperature at the boundary:
+
+\begin{equation}
+  T_{ij} = T_0, \forall (i,j) \in \Gamma
+  \label{eq:conduction_neumann}
+\end{equation}
+
+For the Dirichlet condition, we have a constant heat flux at the boundary.
+
+\subsection{Linear System Form}
 
@@
 
+\subsection{Nonlinear System Form}
+
 %%---------------------------------------------------------------------------%%

### 2012-07-22 doc/thesis_proposal/thesis_proposal.tex
Adding thesis proposal document.

%        File: latexdoc.tex
%     Created: Friday March 4 12:24:15 2011
% Last Change: Friday March 4 12:24:20 2011
%
\documentclass[letterpaper,12pt]{article}
\usepackage[top=1.0in,bottom=1.0in,left=1.0in,right=1.0in]{geometry}
\usepackage{verbatim}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage[usenames]{color}
\usepackage[
naturalnames = true, 
colorlinks = true, 
linkcolor = black,
anchorcolor = black,
citecolor = black,
menucolor = black,
urlcolor = blue
]{hyperref}
\usepackage{listings}
\usepackage{textcomp}
\definecolor{listinggray}{gray}{0.9}
\definecolor{lbcolor}{rgb}{0.9,0.9,0.9}
\lstset{
  backgroundcolor=\color{lbcolor},
  tabsize=4,
  rulecolor=,
  language=c++,
  basicstyle=\scriptsize,
  upquote=true,
  aboveskip={1.5\baselineskip},
  columns=fixed,
  showstringspaces=false,
  extendedchars=true,
  breaklines=true,
  prebreak =
  \raisebox{0ex}[0ex][0ex]{\ensuremath{\hookleftarrow}},
  frame=single,
  showtabs=false,
  showspaces=false,
  showstringspaces=false,
  identifierstyle=\ttfamily,
  keywordstyle=\color[rgb]{0,0,1},
  commentstyle=\color[rgb]{0.133,0.545,0.133},
  stringstyle=\color[rgb]{0.627,0.126,0.941},
}

%%---------------------------------------------------------------------------%%
\author{Stuart R. Slattery
\\ \href{mailto:sslattery@wisc.edu}{\texttt{sslattery@wisc.edu}}
}

\date{\today}
\title{<+  +>:\\
<+  +>}
\begin{document}
\maketitle

%%---------------------------------------------------------------------------%%
\section{<+  +>}<+  +>

%%---------------------------------------------------------------------------%%
\pagebreak
\bibliographystyle{ieeetr}
\bibliography{paper}
\end{document}

### 2012-07-22 doc/thesis_proposal/thesis_proposal.tex
updated thesis proposal

@@
-%        File: latexdoc.tex
-%     Created: Friday March 4 12:24:15 2011
-% Last Change: Friday March 4 12:24:20 2011
+%        File: thesis_proposal.tex
 %
 \documentclass[letterpaper,12pt]{article}
-\usepackage[top=1.0in,bottom=1.0in,left=1.0in,right=1.0in]{geometry}
+\usepackage[top=1.0in,bottom=1.0in,left=1.25in,right=1.25in]{geometry}
 \usepackage{verbatim}
@@
 ]{hyperref}
-\usepackage{listings}
-\usepackage{textcomp}
-\definecolor{listinggray}{gray}{0.9}
-\definecolor{lbcolor}{rgb}{0.9,0.9,0.9}
-\lstset{
-  backgroundcolor=\color{lbcolor},
-  tabsize=4,
-  rulecolor=,
-  language=c++,
-  basicstyle=\scriptsize,
-  upquote=true,
-  aboveskip={1.5\baselineskip},
-  columns=fixed,
-  showstringspaces=false,
-  extendedchars=true,
-  breaklines=true,
-  prebreak =
-  \raisebox{0ex}[0ex][0ex]{\ensuremath{\hookleftarrow}},
-  frame=single,
-  showtabs=false,
-  showspaces=false,
-  showstringspaces=false,
-  identifierstyle=\ttfamily,
-  keywordstyle=\color[rgb]{0,0,1},
-  commentstyle=\color[rgb]{0.133,0.545,0.133},
-  stringstyle=\color[rgb]{0.627,0.126,0.941},
-}
 
@@
 \date{\today}
-\title{<+  +>:\\
-<+  +>}
+\title{Thesis Proposal:\\
+  A Massively Parallel Stochastic Method for Nonlinear Problems}
 \begin{document}
@@
 %%---------------------------------------------------------------------------%%
-\section{<+  +>}<+  +>
+\section{Introduction}
+In many fields of engineering and physics, nonlinear problems are a
+primary focus of study. Recent focus on multiple physics systems adds
+a new level of complexity to common nonlinear systems as solution
+strategies change when they are coupled to other
+problems. Furthermore, a desire for predictive simulations to enhance
+the safety and performance of engineered systems creates a need for
+extremely high fidelity computations to be performed for these coupled
+systems as a means to capture effects not modeled by coarser
+methods. In order to achieve this high fidelity, state-of-the-art
+computing facilities must be leveraged in a way that is both efficient
+and considerate of hardware-related issues. As scientific computing
+moves towards exascale facilities with machines of $O(1,000,000)$
+cores already coming online, new algorithms to solve these complex
+problems must be developed to leverage this new hardware. Issues such
+as resiliency to node failure and scaling to large numbers of cores
+will be pertinent to robust algorithms aimed at this new
+hardware. Considering these issues, this thesis proposes a novel
+stochastic method to advance solution techniques for nonlinear
+problems.
+
+For some time, the particle transport community has been utilizing
+Monte Carlo methods for the solution of transport problems
+\cite{Lewis_1993}. The partial differential equation (PDE) community
+has focused on various deterministic methods for solutions to linear
+problems \cite{Saad_2003}. In between these two areas are a
+not widely known small group of stochastic methods for solving sparse
+linear systems \cite{Hammersley_1964, Halton_1962, Halton_1994,
+  Evans_2003}. In recent years, these methods have been further
+developed for transport problems in the form of Monte Carlo
+Synthetic-Acceleration (MCSA) \cite{Evans_2009} but have yet to be
+applied to more general sparse linear systems. Compared to other
+methods in these regime, MCSA offers two attractive qualities; (1) the
+linear problem operator need not be symmetric or positive-definite,
+thereby reducing preconditioning complexity and (2) the stochastic
+nature of the solution method provides a natural solution to the issue
+of resiliency and is amenable to parallelization using modern methods
+developed by the transport community \cite{Wagner_2011}. The
+developement of MCSA as a general linear solver and the development of
+a parallel MCSA method will be new and unique features of this work.
+
+In addition to linear solver advancements, nonlinear solvers may also
+benefit from a general and parallel MCSA scheme. In the engineering
+community, nonlinear problems are often addressed by linearizing the
+problem and using traditionaly iterative or direct methods
+\cite{Tannehill_1997}. In the mathematics community, various Newton
+methods have been popular \cite{Kelley_1995}. Recently, Jacobian-Free
+Newton-Krylov (JFNK) schemes have been utilized in multiple physics
+architectures and advanced single physics codes \cite{Knoll_2004,
+  Gaston_2009}. The benefits of JFNK schemes are that the Jacobian is
+never formed, simplifying the implementation, and a Krylov solver is
+leveraged (typically GMRES or Conjugate Gradient) providing excellent
+convergence properties for well-conditioned systems. However, there
+are two potential drawbacks to these methods for high fidelity
+predictive simulations: (1) the Jacobian is approximated by a
+first-order differencing method on the order of machine precision that
+can cause large errors for fine-grained simulation and (2) for systems
+that are not symmetric positive-definite (which will be the case for
+most multiphysics systems and certainly for most preconditioned
+systems) the Krylov subspace generated by the GMRES solver my become
+prohibitively large. To address these issues, this thesis proposes a
+new and novel method for nonlinear systems based on the MCSA method.
+
+The Forward-Automated Newton-MCSA (FANM) method is proposed as new
+nonlinear solution method. The key features of FANM are: full Jacobian
+generation using modern Forward Automated Differencing (FAD) methods
+\cite{Bartlett_2006}, and MCSA as the inner linear solver. This method
+has several attractive properties. First, the first-order
+approximation to the Jacobian used in JFNK type methods is eliminated
+by generating the Jacobian explicitly with the model equations through
+FAD. Second, the Jacobian need not be explicitly formed by the user
+but is instead automated through FAD; this eleminates the complexity
+of hand-coding derivatives and has also been demonstrated to be more
+efficient computationally than hand-coded derivatives
+\cite{Bartlett_2006}. Third, unlike GMRES, MCSA does not build a
+subspace during iterations. Although the Jacobian must be explicitly
+formed to use MCSA, for problems that take more than a few GMRES
+iterations to converge the size of the Krylov subspace will grow
+beyond that of the Jacobian. Finally, using MCSA for the linear solve
+provides its benefits for preconditioning, potential resiliency, and
+parallelism.
+
+%%---------------------------------------------------------------------------%%
+\section{Deliverables}
+
+This work will deliver the following items:
+
+\begin{itemize}
+\item Develop a massively parallel strategy for the MCSA method for
+  sparse linear systems.
+\item Implement a massively parallel strategy for the MCSA method for
+  sparse linear systems.
+\item Fully assess the performance of parallel MCSA on leadership class
+  computing systems in order to determine its feasibility using
+  benchmark problems.
+\item Compare the performance of MCSA to other iterative methods for
+  sparse linear systems including GMRES.
+\item Develop the FANM method for nonlinear systems.
+\item Implement the FANM method for nonlinear systems.
+\item Fully assess the performance of the FANM method on leadership
+  class computing systems using model problems.
+\item Compare the performance of MCSA to other nonlinear methods
+  including JFNK.
+\end{itemize}
+
+This work will not address the resiliency of these algorithms.
+
+%%---------------------------------------------------------------------------%%
+\section{Dissertation Outline}
+
+Based on the above work, the dissertation can be outlined in five
+distinct parts:
+
+\begin{enumerate}
+\item Iterative methods for sparse linear systems (fixed-point,
+  subspace, and stochastic methods).
+\item MCSA and MSOD (development and implementation).
+\item Iterative methods for nonlinear systems (Newton methods).
+\item FANM (developement and implementation).
+\item Application to model problems (conduction, Navier-Stokes,
+  conjugate heat transfer).
+\end{enumerate}
+
+%%---------------------------------------------------------------------------%%
+\section{Committee}
+
+\begin{itemize}
+\item Paul Wilson (advisor): Associate Professor, Department of
+  Engineering Physics.
+\item Greg Moses: Professor, Department of Engineering Physics.
+\item Carl Sovinec: Profssor, Department of Engineering Physics.
+\item Chris Rutland: Professor, Deptartment of Mechanical Engineering.
+\item Tom Evans: Senior R\&D Staff, Reactor and Nuclear Systems
+  Division, Oak Ridge National Laboratory.
+
+\end{itemize}
+
+Potential alternates include Robert Witt and Leslie Smith.
+
+%%---------------------------------------------------------------------------%%
+\section{Timeline}
 
@@
 \bibliographystyle{ieeetr}
-\bibliography{paper}
+\bibliography{references}
 \end{document}

### 2012-07-22 doc/thesis_proposal/thesis_proposal.tex
proposal updates

@@
 has focused on various deterministic methods for solutions to linear
-problems \cite{Saad_2003}. In between these two areas are a
-not widely known small group of stochastic methods for solving sparse
-linear systems \cite{Hammersley_1964, Halton_1962, Halton_1994,
-  Evans_2003}. In recent years, these methods have been further
-developed for transport problems in the form of Monte Carlo
-Synthetic-Acceleration (MCSA) \cite{Evans_2009} but have yet to be
-applied to more general sparse linear systems. Compared to other
-methods in these regime, MCSA offers two attractive qualities; (1) the
-linear problem operator need not be symmetric or positive-definite,
-thereby reducing preconditioning complexity and (2) the stochastic
-nature of the solution method provides a natural solution to the issue
-of resiliency and is amenable to parallelization using modern methods
-developed by the transport community \cite{Wagner_2011}. The
-developement of MCSA as a general linear solver and the development of
-a parallel MCSA method will be new and unique features of this work.
+problems \cite{Saad_2003}. In between these two areas are a not widely
+known small group of stochastic methods for solving sparse linear
+systems \cite{Hammersley_1964, Halton_1962, Halton_1994}. In recent
+years, these methods have been further developed for transport
+problems in the form of Monte Carlo Synthetic-Acceleration (MCSA)
+\cite{Evans_2003, Evans_2009} but have yet to be applied to more
+general sparse linear systems. Compared to other methods in these
+regime, MCSA offers two attractive qualities; (1) the linear problem
+operator need not be symmetric or positive-definite, thereby reducing
+preconditioning complexity and (2) the stochastic nature of the
+solution method provides a natural solution to the issue of resiliency
+and is amenable to parallelization using modern methods developed by
+the transport community \cite{Wagner_2011}. The developement of MCSA
+as a general linear solver and the development of a parallel MCSA
+method will be new and unique features of this work. Resiliency will
+not be addressed.
 
@@
 methods have been popular \cite{Kelley_1995}. Recently, Jacobian-Free
-Newton-Krylov (JFNK) schemes have been utilized in multiple physics
-architectures and advanced single physics codes \cite{Knoll_2004,
-  Gaston_2009}. The benefits of JFNK schemes are that the Jacobian is
-never formed, simplifying the implementation, and a Krylov solver is
-leveraged (typically GMRES or Conjugate Gradient) providing excellent
-convergence properties for well-conditioned systems. However, there
-are two potential drawbacks to these methods for high fidelity
-predictive simulations: (1) the Jacobian is approximated by a
-first-order differencing method on the order of machine precision that
-can cause large errors for fine-grained simulation and (2) for systems
-that are not symmetric positive-definite (which will be the case for
-most multiphysics systems and certainly for most preconditioned
-systems) the Krylov subspace generated by the GMRES solver my become
-prohibitively large. To address these issues, this thesis proposes a
-new and novel method for nonlinear systems based on the MCSA method.
+Newton-Krylov (JFNK) schemes \cite{Knoll_2004} have been utilized in
+multiple physics architectures and advanced single physics codes
+\cite{Gaston_2009}. The benefits of JFNK schemes are that the Jacobian
+is never formed, simplifying the implementation, and a Krylov solver
+is leveraged (typically GMRES or Conjugate Gradient), providing
+excellent convergence properties for well-conditioned
+systems. However, there are two potential drawbacks to these methods
+for high fidelity predictive simulations: (1) the Jacobian is
+approximated by a first-order differencing method on the order of
+machine precision such this error can grow beyond that of those in a
+fine-grained system and (2) for systems that are not symmetric
+positive-definite (which will be the case for most multiphysics
+systems and certainly for most preconditioned systems) the Krylov
+subspace generated by the GMRES solver may become prohibitively
+large. To address these issues, this thesis proposes a new and novel
+method for nonlinear systems based on the MCSA method.
 
@@
 of hand-coding derivatives and has also been demonstrated to be more
-efficient computationally than hand-coded derivatives
+efficient computationally than evaluating hand-coded derivatives
 \cite{Bartlett_2006}. Third, unlike GMRES, MCSA does not build a
@@
   subspace, and stochastic methods).
-\item MCSA and MSOD (development and implementation).
+\item MCSA and Multiple Set Overlapping Domain Decomposition
+  (development and implementation).
 \item Iterative methods for nonlinear systems (Newton methods).
@@
 
-Potential alternates include Robert Witt and Leslie Smith.
-
 %%---------------------------------------------------------------------------%%

### 2012-07-23 doc/model_problems/model_problems.tex
fixed errors in conduction equation derivation in model problem doc

@@
     -T^{n+1}_{i,j-1}}{\Delta_j} +\beta_{i,j+1/2} \frac{T^{n+1}_{i,j+1}
-    -T^{n+1}_{i,j-1}}{\Delta_j}\:.
+    -T^{n+1}_{i,j}}{\Delta_j}\:.
   \label{eq:conduction_fv_multline}
@@
 Rearranging for temperature, we can then form a system of equations
-amenable to matrix solutions, not yet considering boundary conditions,
-arriving at the discrete, time-dependent conduction equation:
+amenable to matrix solutions, not yet considering boundary conditions:
 
 \begin{multline}
- \Big[1 + \frac{\alpha_{ij}}{\Delta_i}(\beta_{i-1/2,j} + \beta_{i+1/2,j})\Big]
- T^{n+1}_{ij} + \Big[-\frac{\alpha_{ij}}{\Delta_i} \beta_{i-1/2,j}\Big]
- T^{n+1}_{i-1,j} + \Big[\frac{-\alpha_{ij}}{\Delta_i} \beta_{i+1/2,j}\Big]
- T^{n+1}_{i+1,j}\\
- + \Big[-\frac{\alpha_{ij}}{\Delta_j} \beta_{i-1/2,j}\Big]
- T^{n+1}_{i,j-1} + \Big[-\frac{\alpha_{ij}}{\Delta_j} \beta_{i+1/2,j}\Big]
+ \Big[1 + \frac{\alpha_{ij}}{\Delta_i}(\beta_{i-1/2,j} +
+   \beta_{i+1/2,j} + \beta_{i,j-1/2} + \beta_{i,j+1/2})\Big]
+ T^{n+1}_{ij} \\
+ + \Big[-\frac{\alpha_{ij}}{\Delta_i}
+   \beta_{i-1/2,j}\Big] T^{n+1}_{i-1,j}
+ + \Big[-\frac{\alpha_{ij}}{\Delta_i} \beta_{i+1/2,j}\Big]
+ T^{n+1}_{i+1,j} \\
+ + \Big[-\frac{\alpha_{ij}}{\Delta_j}
+   \beta_{i,j-1/2}\Big] T^{n+1}_{i,j-1} +
+ \Big[-\frac{\alpha_{ij}}{\Delta_j} \beta_{i,j+1/2}\Big]
  T^{n+1}_{i,j+1} = T^n_{ij}\:.
  \label{eq:conduction_inner_terms}
+\end{multline}
+
+Finally, we consider cell-centered thermal conductivity values as a
+single material occupies an entire cell:
+
+\begin{equation}
+  \beta_{i-1/2,j} = \frac{\beta_{ij}+\beta_{i-1,j}}{2}
+  \label{eq:conduction_conductivity}
+\end{equation}
+
+Applying Eq.~(\ref{eq:conduction_conductivity}), we arrive at the discrete,
+time-dependent conduction equation:
+
+\begin{multline}
+ \Big[1 + \frac{\alpha_{ij}}{2\Delta_i}(\beta_{i-1,j} +
+   \beta_{i+1,j} + 4 \beta_{i,j} + \beta_{i,j-1} + \beta_{i,j+1})\Big]
+ T^{n+1}_{ij} \\
+ + \Big[-\frac{\alpha_{ij}}{2 \Delta_i}
+   (\beta_{i-1,j}+\beta_{ij})\Big] T^{n+1}_{i-1,j}
+ + \Big[-\frac{\alpha_{ij}}{2 \Delta_i}(\beta_{ij}+\beta_{i+1,j})\Big]
+ T^{n+1}_{i+1,j} \\
+ + \Big[-\frac{\alpha_{ij}}{2 \Delta_j}
+   (\beta_{i,j-1}+\beta_{ij})\Big] T^{n+1}_{i,j-1} +
+ \Big[-\frac{\alpha_{ij}}{2 \Delta_j}(\beta_{ij}+\beta_{i,j+1})\Big]
+ T^{n+1}_{i,j+1} = T^n_{ij}\:.
+ \label{eq:conduction_discrete_equation}
 \end{multline}

### 2012-08-01 doc/model_problems/model_problems.tex
model problem updates
thesis proposal and model problems update

@@
 %%---------------------------------------------------------------------------%%
-\section{Computational Grid}
+\section{Computational Domain}
 
@@
 \begin{equation}
-  \frac{d}{dt} \int \int_{C_{ij}} E(x,y,t) dx dy = -F(T(x,y,t)) |_{B_{ij}}\:.
+  \frac{d}{dt} \iint_{C_{ij}} E(x,y,t) dx dy = -F(T(x,y,t)) |_{B_{ij}}\:.
   \label{eq:conduction_conservation}
@@
 with the flux on the cell boundaries. Inserting
-Eq.~(\ref{eq:conduction_energy}) int
+Eq.~(\ref{eq:conduction_energy}) into
 Eq.~(\ref{eq:conduction_conservation}) and dividing by the cell volume
@@
 \begin{equation}
-  \alpha_{ij} = \frac{\Delta_t}{C_{P_{ij}} \Delta_i \Delta_j}
-  \label{eq:conduction_coeff}\:.
+  \alpha_{ij} = \frac{\Delta_t}{C_{P_{ij}} \Delta_i \Delta_j}\:.
+  \label{eq:conduction_coeff}
 \end{equation}
@@
 %%---------------------------------------------------------------------------%%
-\section{Navier-Stokes Equations}
+\section{Incompressible Navier-Stokes Equations}
 

### 2012-08-01 doc/thesis_proposal/thesis_proposal.tex
thesis proposal and model problems update

@@
 In many fields of engineering and physics, nonlinear problems are a
-primary focus of study. Recent focus on multiple physics systems adds
-a new level of complexity to common nonlinear systems as solution
-strategies change when they are coupled to other
-problems. Furthermore, a desire for predictive simulations to enhance
-the safety and performance of engineered systems creates a need for
-extremely high fidelity computations to be performed for these coupled
-systems as a means to capture effects not modeled by coarser
-methods. In order to achieve this high fidelity, state-of-the-art
-computing facilities must be leveraged in a way that is both efficient
-and considerate of hardware-related issues. As scientific computing
-moves towards exascale facilities with machines of $O(1,000,000)$
-cores already coming online, new algorithms to solve these complex
-problems must be developed to leverage this new hardware. Issues such
-as resiliency to node failure and scaling to large numbers of cores
-will be pertinent to robust algorithms aimed at this new
-hardware. Considering these issues, this thesis proposes a novel
-stochastic method to advance solution techniques for nonlinear
-problems.
+primary focus of study. Recent focus on multiple physics systems (and
+in particular nuclear systems) adds a new level of complexity to
+common nonlinear systems as solution strategies change when they are
+coupled to other problems. Furthermore, a desire for predictive
+simulations to enhance the safety and performance of engineered
+systems creates a need for extremely high fidelity computations to be
+performed for these coupled systems as a means to capture effects not
+modeled by coarser methods. In order to achieve this high fidelity,
+state-of-the-art computing facilities must be leveraged in a way that
+is both efficient and considerate of hardware-related issues. As
+scientific computing moves towards exascale facilities with machines
+of $O(1,000,000)$ cores already coming online, new algorithms to solve
+these complex problems must be developed to leverage this new
+hardware. Issues such as resiliency to node failure and scaling to
+large numbers of cores will be pertinent to robust algorithms aimed at
+this new hardware. Considering these issues, this thesis proposes a
+novel stochastic method to advance solution techniques for nonlinear
+problems with a focus on coupled physics systems.
 
@@
 general sparse linear systems. Compared to other methods in these
-regime, MCSA offers two attractive qualities; (1) the linear problem
+regime, MCSA offers three attractive qualities; (1) the linear problem
 operator need not be symmetric or positive-definite, thereby reducing
-preconditioning complexity and (2) the stochastic nature of the
-solution method provides a natural solution to the issue of resiliency
-and is amenable to parallelization using modern methods developed by
+preconditioning complexity, (2) the stochastic nature of the
+solution method provides a natural solution to the issue of resiliency,
+and (3) is amenable to parallelization using modern methods developed by
 the transport community \cite{Wagner_2011}. The developement of MCSA
@@
 is leveraged (typically GMRES or Conjugate Gradient), providing
-excellent convergence properties for well-conditioned
+excellent convergence properties for well-conditioned and well-scaled
 systems. However, there are two potential drawbacks to these methods
@@
 of hand-coding derivatives and has also been demonstrated to be more
-efficient computationally than evaluating hand-coded derivatives
+efficient computationally than evaluating differenced derivatives
 \cite{Bartlett_2006}. Third, unlike GMRES, MCSA does not build a

### 2012-08-03 doc/thesis_proposal/thesis_proposal.tex
updated thesis proposal
fixed thesis proposal title

@@
 approximated by a first-order differencing method on the order of
-machine precision such this error can grow beyond that of those in a
-fine-grained system and (2) for systems that are not symmetric
+machine precision such that this error can grow beyond that of those
+in a fine-grained system and (2) for systems that are not symmetric
 positive-definite (which will be the case for most multiphysics
@@
 \item Greg Moses: Professor, Department of Engineering Physics.
-\item Carl Sovinec: Profssor, Department of Engineering Physics.
+\item Carl Sovinec: Professor, Department of Engineering Physics.
 \item Chris Rutland: Professor, Deptartment of Mechanical Engineering.
-\item Tom Evans: Senior R\&D Staff, Reactor and Nuclear Systems
+\item Tom Evans: Distinguished R\&D Staff, Reactor and Nuclear Systems
   Division, Oak Ridge National Laboratory.

### 2012-08-06 doc/model_problems/model_problems.tex
fixed equation in model problem doc
model problem updates

--- 2012-07-22 doc/model_problems/model_problems.tex
@@
 \date{\today}
-\title{A Finite Volume Formulation for the 2D Incompressible
-  Navier-Stokes and Conduction Equations}
+\title{Chimera Model Problems and Solution Schemes}
 \begin{document}
 \maketitle
+
+%%---------------------------------------------------------------------------%%
+\section{Introduction}
+In this document, finite volume formulations for the 2D incompressible
+Navier-Stokes equations and conduction equation are generated. The
+computational grid, boundary conditions, time integrations, and
+solution methods are also formulated. Using these formulations, the
+conjugate heat transfer problem is posed. These model problems and
+solution schemes will be used within the Chimera code framework.
 
@@
 \section{Conduction Equation}
-We begin the derivation of the conduction equation by defining the
-conserved quantity, energy, in terms of temperature:
+The conduction equation describes the movement of thermal energy
+through a solid material. We begin the derivation of the conduction
+equation by defining the conserved quantity, energy, in terms of
+temperature:
 
@@
 where $C_P$ is the specific heat capacity of the material. We also
-define the heat flux, $F$, in terms of temperature using Fick's law:
+define the energy flux, $F$, in terms of temperature using Fick's law:
 
@@
 where $\beta$ is the thermal conductivity of the material. With these
-two relations in hand, we can then write a conservation law for energy
-\cite{leveque_2002}:
+two relations in hand, we can then write a conservation statement for
+energy over a grid cell \cite{leveque_2002}:
 
 \begin{equation}
-  \frac{d}{dt} \int \int_{C_{ij}} E(x,y,t) dx dy = -F(T(x,y,t)) |_{B_{ij}}\:.
+  \frac{d}{dt} \iint_{C_{ij}} E(x,y,t) dx dy = -F(T(x,y,t))
+  |_{B_{ij}}\:.
   \label{eq:conduction_conservation}
@@
 with the flux on the cell boundaries. Inserting
-Eq.~(\ref{eq:conduction_energy}) int
+Eq.~(\ref{eq:conduction_energy}) into
 Eq.~(\ref{eq:conduction_conservation}) and dividing by the cell volume
-recovers the cell averaged temperature and specific heat on the left
+recovers the cell-averaged temperature and specific heat on the left
 hand side:
@@
 \begin{equation}
-  \frac{d}{dt} C_{P_{ij}} T_{ij} = -\frac{1}{\Delta_i
-    \Delta_j}F(T(x,y,t)) |_{B_{ij}}\:.
+  \frac{d}{dt} C_{P_{ij}} T_{ij} = -\frac{1}{V_{ij}}F(T(x,y,t))
+  |_{B_{ij}}\:.
   \label{eq:conduction_cell_averaged}
@@
 
-We then insert Eq.~(\ref{eq:conduction_ficks_law}), difference the
-left hand side, and integrate, maintaining the conservation properties
-of the system:
+\subsection{Spatial Discretization}
 
-\begin{multline}
-  C_{P_{ij}} ( T^{n+1}_{ij} - T^n_{ij} ) = \frac{1}{\Delta_i \Delta_j}
-  \int_{t^n}^{t^{n+1}} [ -\beta_{i-1/2,j} \nabla T_{i-1/2,j}\\
-    + \beta_{i+1/2,j} \nabla T_{i+1/2,j} - \beta_{i,j-1/2} \nabla
-    T_{i,j-1/2} +\beta_{i,j+1/2} \nabla T_{i,j+1/2} ] dt\:.
+We begin the spatial discretization by inserting
+Eq.~(\ref{eq:conduction_ficks_law}) and differencing the left hand
+side, and integrate, maintaining the conservation properties of the
+system:
+
+\begin{equation}
+  C_{P_{ij}} ( T^{n+1}_{ij} - T^n_{ij} ) = \frac{1}{V_{ij}}
+  \int_{t^n}^{t^{n+1}} \Big[(F_{i+1/2,j}-F_{i-1/2,j})\Delta_j +
+    (F_{i,j+1/2}-F_{i,j-1/2})\Delta_i \Big] dt\:.
   \label{eq:conduction_time_integral}
-\end{multline}
+\end{equation}
 
@@
 first order central difference for the derivative in the heat flux
-terms giving a backwards Euler representation for the time
-integration:
-
-\begin{multline}
-  C_{P_{ij}}( T^{n+1}_{ij} - T^n_{ij} ) = \frac{\Delta_t}{\Delta_i
-    \Delta_j} -\beta_{i-1/2,j} \frac{T^{n+1}_{ij}
-    -T^{n+1}_{i-1,j}}{\Delta_i} +\beta_{i+1/2,j} \frac{T^{n+1}_{i+1,j}
-    -T^{n+1}_{ij}}{\Delta_i}\\
-  -\beta_{i,j-1/2} \frac{T^{n+1}_{ij}
-    -T^{n+1}_{i,j-1}}{\Delta_j} +\beta_{i,j+1/2} \frac{T^{n+1}_{i,j+1}
-    -T^{n+1}_{i,j-1}}{\Delta_j}\:.
-  \label{eq:conduction_fv_multline}
-\end{multline}
-
-We will define $\alpha$ as:
+terms with a backwards Euler representation for the time integration:
 
 \begin{equation}
-  \alpha_{ij} = \frac{\Delta_t}{C_{P_{ij}} \Delta_i \Delta_j}
-  \label{eq:conduction_coeff}\:.
+  C_{P_{ij}}( T^{n+1}_{ij} - T^n_{ij} ) = \frac{\Delta_t}{V_{ij}}
+  \Big[ (F^{n+1}_{i+1/2,j}-F^{n+1}_{i-1/2,j})\Delta_j +
+    (F^{n+1}_{i,j+1/2}-F^{n+1}_{i,j-1/2})\Delta_i \Big]
+  \label{eq:conduction_fv_equation}
 \end{equation}
 
-Rearranging for temperature, we can then form a system of equations
-amenable to matrix solutions, not yet considering boundary conditions,
-arriving at the discrete, time-dependent conduction equation:
+The energy fluxes on the boundaries will be approximated with a
+second-order accurate central differencing scheme with the thermal
+conductivities lagged:
 
-\begin{multline}
- \Big[1 + \frac{\alpha_{ij}}{\Delta_i}(\beta_{i-1/2,j} + \beta_{i+1/2,j})\Big]
- T^{n+1}_{ij} + \Big[-\frac{\alpha_{ij}}{\Delta_i} \beta_{i-1/2,j}\Big]
- T^{n+1}_{i-1,j} + \Big[\frac{-\alpha_{ij}}{\Delta_i} \beta_{i+1/2,j}\Big]
- T^{n+1}_{i+1,j}\\
- + \Big[-\frac{\alpha_{ij}}{\Delta_j} \beta_{i-1/2,j}\Big]
- T^{n+1}_{i,j-1} + \Big[-\frac{\alpha_{ij}}{\Delta_j} \beta_{i+1/2,j}\Big]
- T^{n+1}_{i,j+1} = T^n_{ij}\:.
+\begin{equation}
+  \begin{aligned}
+  F^{n+1}_{i-1/2,j} = -\beta^n_{i-1/2,j} \frac{T^{n+1}_{ij}
+    -T^{n+1}_{i-1,j}}{\Delta_{i-1/2}} \\ F^{n+1}_{i+1/2,j} =
+  -\beta^n_{i+1/2,j} \frac{T^{n+1}_{i+1,j}
+    -T^{n+1}_{ij}}{\Delta_{i+1/2}} \\ F^{n+1}_{i,j-1/2} =
+  -\beta^n_{i,j-1/2} \frac{T^{n+1}_{ij}
+    -T^{n+1}_{i,j-1}}{\Delta_{j-1/2}} \\ F^{n+1}_{i,j+1/2} =
+  -\beta^n_{i,j+1/2} \frac{T^{n+1}_{i,j+1}
+    -T^{n+1}_{i,j}}{\Delta_{j+1/2}}\:.
+    \label{eq:conduction_fluxes}
+  \end{aligned}
+\end{equation}
+
+Here, we define:
+
+\begin{equation}
+  \begin{aligned}
+    \Delta_{l-1/2} = \frac{\Delta_l + \Delta_{l-1}}{2}
+    \\ \Delta_{l+1/2} = \frac{\Delta_{l+1} + \Delta_l}{2}\:,
+    \label{eq:conduction_half_delta}
+  \end{aligned}
+\end{equation}
+
+where $l \in [i,j]$. Rearranging for temperature, we can then form a
+system of equations amenable to matrix solutions, not yet considering
+boundary conditions:
+
+\begin{equation}
+ a_{ij}T^{n+1}_{ij} + a_{i-1,j}T^{n+1}_{i-1,j} +
+ a_{i+1,j}T^{n+1}_{i+1,j} + a_{i,j-1}T^{n+1}_{i,j-1} +
+ a_{i,j+1}T^{n+1}_{i,j+1} = \frac{C_{P_{ij}}V_{ij}}{\Delta t}
+ T^n_{ij}\:,
  \label{eq:conduction_inner_terms}
-\end{multline}
+\end{equation}
+
+where:
+
+\begin{equation}
+  \begin{aligned}
+    a_{i+1,j} = -\frac{\Delta_j \beta^n_{i+1/2}}{\Delta_{i+1/2}}
+    \\ a_{i-1,j} = -\frac{\Delta_j \beta^n_{i-1/2}}{\Delta_{i-1/2}}
+    \\ a_{ij} = \frac{\Delta_j \beta^n_{i+1/2}}{\Delta_{i+1/2}} +
+    \frac{\Delta_j \beta^n_{i-1/2}}{\Delta_{i-1/2}} + \frac{\Delta_i
+      \beta^n_{j-1/2}}{\Delta_{j-1/2}} + \frac{\Delta_i
+      \beta^n_{j+1/2}}{\Delta_{j+1/2}} +
+    \frac{C_{P_{ij}}V_{ij}}{\Delta t} \\ a_{i,j-1} = -\frac{\Delta_i
+      \beta^n_{j-1/2}}{\Delta_{j-1/2}} \\ a_{i,j+1} = -\frac{\Delta_i
+      \beta^n_{j+1/2}}{\Delta_{j+1/2}} \:.
+    \label{eq:conduction_coeffs}
+  \end{aligned}
+\end{equation}
+
+We are then left to resolve the thermal conductivity values defined on
+the cell boundaries with values defined at the cell centers. We choose
+these values such that the energy flux across a cell boundary is
+continuous from cell to cell, even if two adjacent cells have
+different materials. We do this be equating the boundary flux computed
+from the perspective of each neighboring cell with the gradient
+generated using a one sided difference:
+
+\begin{equation}
+  \begin{aligned}
+    F^{l+1}_{l+1/2} = -2 \beta_{l+1}
+    \frac{T_{l+1}-T_{l+1/2}}{\Delta_l} \\ F^{l}_{l+1/2} = -2 \beta_{l}
+    \frac{T_{l+1/2}-T_l}{\Delta_l} \\
+    F^{l+1}_{l+1/2} = F^{l}_{l+1/2} \:,
+    \label{eq:conduction_tc_step1}
+  \end{aligned}
+\end{equation}
+
+with $F^{l}_{l+1/2}$ defined by the cell-centered values in cell $l$
+and $l \in [i,j]$. We can follow the same procedure for the other side
+of the cell:
+
+\begin{equation}
+  \begin{aligned}
+    F^{l}_{l-1/2} = -2 \beta_{l} \frac{T_{l}-T_{l-1/2}}{\Delta_l}
+    \\ F^{l-1}_{l-1/2} = -2 \beta_{l-1}
+    \frac{T_{l-1/2}-T_{l-1}}{\Delta_l} \\ F^{l}_{l-1/2} =
+    F^{l-1}_{l-1/2} \:,
+    \label{eq:conduction_tc_step2}
+  \end{aligned}
+\end{equation}
+
+We solve Eqs~(\ref{eq:conduction_tc_step1}) and
+(\ref{eq:conduction_tc_step2}) for $T_{l+1/2}$ and $T_{l-1/2}$ to
+eliminate the cell centered temperatures:
+
+\begin{equation}
+  \begin{aligned}
+    T_{l+1/2} = \frac{\Delta_l \beta_{l+1} T_{l} + \Delta_{l+1}
+      \beta_{l} T_{l+1}} {\Delta_l \beta_{l+1} + \Delta_{l+1} \beta_{l}}
+    \\ T_{l-1/2} = \frac{\Delta_{l-1} \beta_{l} T_{l} + \Delta_{l}
+      \beta_{l-1} T_{l-1}} {\Delta_{l} \beta_{l-1} + \Delta_{l}
+      \beta_{l-1}} \:.
+    \label{eq:conduction_tc_step3}
+  \end{aligned}
+\end{equation}
+
+We then substitute into either of the the equations in both
+Eqs~(\ref{eq:conduction_tc_step1}) and (\ref{eq:conduction_tc_step2})
+and equate the result with Eq~(\ref{eq:conduction_fluxes}) to arrive
+at the final cell-centered thermal conductivity values:
+
+\begin{equation}
+  \begin{aligned}
+    \beta^n_{i+1/2,j} = \\
+    \beta^n_{i-1/2,j} = \\
+    \beta^n_{i,j+1/2} = \\
+    \beta^n_{i,j-1/2} = \:.
+    \label{eq:conduction_tc_step_4}
+  \end{aligned}
+\end{equation}
 
@@
 
-For the Dirichlet condition, we have a constant heat flux at the boundary.
+For the Dirichlet condition, we have a constant heat flux at the
+boundary.
 

### 2012-08-06 doc/model_problems/model_problems.tex
model problem updates

@@
 In this document, finite volume formulations for the 2D incompressible
-Navier-Stokes equations and conduction equation are generated. The
-computational grid, boundary conditions, time integrations, and
-solution methods are also formulated. Using these formulations, the
-conjugate heat transfer problem is posed. These model problems and
-solution schemes will be used within the Chimera code framework.
+Navier-Stokes equations and conduction equation are generated using an
+Eulerian approach. The computational grid, boundary conditions, time
+integrations, and solution methods are also formulated. Using these
+formulations, the conjugate heat transfer problem is posed. These
+model problems and solution schemes will be used within the Chimera
+code framework to demonstrate nonlinear capabilities.
 
 %%---------------------------------------------------------------------------%%
-\section{Computational Grid}
+\section{Computational Domain}
 
@@
 \section{Incompressible Navier-Stokes Equations}
+The incompressible Navier-Stokes equations describe a fluid a constant
+density. The will generated through mass, momentum, and energy
+conservation statements with an equation of state to close the system.
+
+\subsection{Mass}
+
+\subsection{Momentum}
+
+\subsection{Energy}
+For energy conservation, we choose a temperature framework in the same
+manner as the conduction problem.
+
+\subsection{Equation of State}
 

### 2012-08-10 doc/jcomp2012/jcomp2012.tex
updates to jcomp draft

@@
 %% or use the graphicx package for more complicated commands
-\usepackage{epsfig}
-\usepackage{graphicx}
-\usepackage{subfigure}
+\usepackage{epsfig} \usepackage{graphicx} \usepackage{subfigure}
 %% or use the.pdffig package if you prefer to use the old commands
@@
 %% The amsthm package provides extended theorem environments
-\usepackage{amsthm}
-\usepackage{amsmath}
-\usepackage{tmadd,tmath}
-\usepackage[mathcal]{euscript}
-\usepackage{color}
+\usepackage{amsthm} \usepackage{amsmath} \usepackage{tmadd,tmath}
+\usepackage[mathcal]{euscript} \usepackage{color}
 
@@
 \newcommand{\sign}{\ensuremath{\tilde{\sigma}^n}}
-\newcommand{\qn}{\ensuremath{q^n}}
-\newcommand{\Tn}{\ensuremath{T^n}}
+\newcommand{\qn}{\ensuremath{q^n}} \newcommand{\Tn}{\ensuremath{T^n}}
 \newcommand{\Dn}{\ensuremath{D^n}}
@@
   \author[ornl]{Thomas M. Evans\corref{cor1}\fnref{fn1}}
-  \ead{evanstm@ornlgov}
-  \ead[url]{www.ornl.gov}
+  \ead{evanstm@ornlgov} \ead[url]{www.ornl.gov}
   \cortext[cor1]{Corresponding Author}
 
-  \author[ornl]{Scott W. Mosher\fnref{fn1}}
-  \ead{moshersw@ornl.gov}
+  \author[ornl]{Scott W. Mosher\fnref{fn1}} \ead{moshersw@ornl.gov}
 
   \author[wisc]{Stuart R. Slattery\fnref{fn2}}
-  \ead{sslattery@wisc.edu}
-  \ead[url]{www.wisc.edu}
+  \ead{sslattery@wisc.edu} \ead[url]{www.wisc.edu}
 
   \fntext[fn1]{Radiation Transport Group, Reactor and Nuclear Systems
-    Division} 
-  \fntext[fn2]{Department of Engineering Physics, College of Engineering}
-
-  \address[ornl]{Oak Ridge National Laboratory, 
-    1 Bethel Valley Rd., Oak Ridge, TN 37831 U.S.A.}
-  \address[wisc]{University of Wisconsin-Madison, 1500 Engineering
-    Dr., Madison, WI 53716 U.S.A.}
+    Division} \fntext[fn2]{Department of Engineering Physics, College
+    of Engineering}
+
+  \address[ornl]{Oak Ridge National Laboratory, 1 Bethel Valley Rd.,
+    Oak Ridge, TN 37831 U.S.A.}  \address[wisc]{University of
+    Wisconsin-Madison, 1500 Engineering Dr., Madison, WI 53716 U.S.A.}
 
@@
     We present a synthetic-acceleration based Monte Carlo method for
-    solving the 1T thermal radiation diffusion equation in 3
-    dimensions. We show that this method can be an effective solver
-    for sparse matrix systems and is competitive with deterministic
-    methods including preconditioned Conjugate Gradient. We also
+    solving the 1T thermal radiation diffusion equation in three
+    dimensions. The algorithm performance is compared against
+    traditional solution schemes using a Marshak benchmark problem and
+    a more sophisticated multiple material problem. Our results show
+    that not only can our Monte Carlo method be an effective
+    solver for sparse matrix systems, but also that it is competitive
+    with deterministic methods including preconditioned Conjugate
+    Gradient while producing numerically identical results. We also
     discuss various aspects of preconditioning the method and its
@@
 
-In Ref.~\cite{evans_2003}, we developed a residual Monte Carlo method
-that was an application of Halton's Sequential Monte Carlo
-method~\cite{halton_1962,halton_1994} to the 1D equilibrium (1T)
-thermal radiation diffusion equation.  At that time, Halton's method
-was not widely known in computational physics.  While extending our
-method to 3D, we have discovered that the Sequential Monte Carlo
-method is actually a variant of iterative refinement cast as a
-residual method.  Using Monte Carlo to accelerate a fixed-point
+In previous work~\cite{evans_2003}, we developed a residual Monte
+Carlo method that was an application of Halton's Sequential Monte
+Carlo method~\cite{halton_1962,halton_1994} to the one-dimensional
+equilibrium (1T) thermal radiation diffusion equation.  At that time,
+Halton's method was not widely known in computational physics.  While
+extending our method to three dimensions, we have discovered that the Sequential
+Monte Carlo method is actually a variant of iterative refinement cast
+as a residual method.  Using Monte Carlo to accelerate a fixed-point
 (Richardson) iteration, our new method surpasses our previous 1D
@@
 solving matrix systems. In \S~\ref{sec:iter-refin-monte} we discuss
-our new synthetic-acceleration scheme and connect with previous work
-by Halton~\cite{halton_1994} and others~\cite{evans_2003}.  The 1T
-radiation diffusion equation and its discrete form will be derived in
-\S~\ref{sec:discr-form-radi}.  We discuss using Monte Carlo
+our new synthetic-acceleration scheme and connect it with previous
+work by Halton~\cite{halton_1994} and others~\cite{evans_2003}.  The
+1T radiation diffusion equation and its discrete form will be derived
+in \S~\ref{sec:discr-form-radi}.  We discuss using Monte Carlo
 Synthetic-Accleration to solve the discrete radiation diffusion
@@
 %%---------------------------------------------------------------------------%%
-\section{Monte Carlo Matrix Solution Methods}
+\section{Mathematical Methods}
 \label{sec:monte-carlo-matrix}
 
-Consider the following matrix equation:
-
+The idea of using Monte Carlo methods (random walks) to invert linear
+systems is not new.  The earliest referenceable work dates to a 1950
+paper by Forsythe and Leibler \cite{forsythe}.  Their paper credits
+the idea to unpublished work by J.v. Neumann and S.M. Ulam dating back
+to the 1940's.  The basic principles of Monte Carlo matrix inversion
+were further elucidated in Hammersley and Handscomb's 1964 text
+\cite{hammersley_1964}.  These early methods are distinguished by very
+slow, statistically noisy convergence properties; thus, they have not
+made any significant impact in the physics community.
+
+In order to address the convergence issues plaguing Monte Carlo
+solvers, Halton \cite{halton_1962,halton_1994} proposed a Sequential
+Monte Carlo method.  This algorithm demonstrated dramatically improved
+convergence over regular Monte Carlo.  Nonetheless, this method has
+not gained widespread use in computational physics.  In the next
+sections, we will briefly describe the Monte Carlo methods that can be
+used to solve linear systems and expand them to present our new
+synthetic acceleration method.
+
+%%---------------------------------------------------------------------------%%
+\subsection{Monte Carlo Matrix Solution Methods}
+\label{sec:background}
+To establish the mathematical framework for Monte Carlo linear
+solvers, we consider the following matrix equation:
 \begin{equation}
@@
 \end{equation}
-
 which can be written,
-
 \begin{equation}
   \begin{split}
-    \vx &= (\vI - \vA)\vx + \vb\\
-    &= \vH\vx + \vb\:.
+    \vx &= (\vI - \vA)\vx + \vb\\ &= \vH\vx + \vb\:.
   \end{split}
@@
 \end{equation}
-
-Here, $\vH = (\vI - \vA)$ is the \latin{iteration matrix}.  When the
-spectral radius, $\rho(\vH)$, of the iteration matrix satisfies
-
-\begin{equation}
-  \rho(\vH) < 1\:,
-\end{equation}
-
-we can expand $\vA^{-1}$ using the \latin{Neumann Series},
-
+When the spectral radius of the iteration matrix ($\vH$) is less than
+1, we can expand $\vA^{-1}$ using the \latin{Neumann Series},
 \begin{equation}
@@
 \end{equation} 
-
 Thus, when $\rho(\vH) < 1$ we can recast the solution vector $\vx$ as
 a series,
-
 \begin{equation}
   \begin{split}
-    \vx &= (\vI - \vH)^{-1}\vb\\
-    &= \vb + \vH\vb + \vH^2\vb + \vH^3\vb + \ldots\:.
+    \vx &= (\vI - \vH)^{-1}\vb\\ &= \vb + \vH\vb + \vH^2\vb + \vH^3\vb
+    + \ldots\:.
     \label{eq:neumann_series}
@@
 \end{equation}
-
 With this knowledge in hand, we can write an iterative method that
 solves Eq.~(\ref{eq:Ax=b}) (Richardson's Iteration),
-
 \begin{equation}
@@
 \end{equation}
-
-Equation~(\ref{eq:richardson}) will converge when $\rho(\vH) < 1\ 
-\forall\ \vb \in \mathcal{R}^{N}$~\cite{kelley_1995}.
-
-\subsection{Direct Method}
+Equation~(\ref{eq:richardson}) will converge for all $\vb \in
+\mathcal{R}^{N}$ when $\rho(\vH) < 1$~\cite{kelley_1995}.  All of the
+Monte Carlo methods that are described in \S\S~\ref{sec:direct-method}
+through \ref{sec:iter-refin-monte} rely on estimating the terms in
+Eq.~(\ref{eq:neumann_series}) through random walks.
+
+%%---------------------------------------------------------------------------%%
+
+\subsubsection{Direct Method}
 \label{sec:direct-method}
@@
 component of $\vx$,
-
 \begin{equation}
@@
     x_i &= (\vb)_i + (\vH\vb)_i + (\vH^2\vb)_i + (\vH^3\vb)_i +
-    \ldots\\
-    &= \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
+    \ldots\\ &= \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
     \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}\:.
@@
 \end{equation}
-
 The terms in the Neumann series can be interpreted as a series of
@@
 random walk.  Let $X$ be a random variable sampled from a random walk
-with $k$ events (particles in the context of radiative transfer) that
-initiates in state $i$,
-
+with $k$ events that initiates in state $i$,
 \begin{equation}
   \begin{split}
-    X(i_0 = i) &= \sum_{m=0}^{k}W_m b_{i_m}\\
-    &= \sum_{m=0}^{k}w_{i,i_1}w_{i_1,i_2}\ldots
-    w_{i_{m-1},i_m}b_{i_m}\:.
+    X(i_0 = i) &= \sum_{m=0}^{k}W_m b_{i_m}\\ &=
+    \sum_{m=0}^{k}w_{i,i_1}w_{i_1,i_2}\ldots w_{i_{m-1},i_m}b_{i_m}\:.
   \end{split}
@@
 \end{equation}
-
-Here, the particle weight on the $m^\text{th}$ step is denoted $W_m$,
-and each particle starts with unit weight. At every step,
-contributions to each component of $X$ are generated only in the
-starting state of the particle, $i_0$, by the source term in the
-current state, $b_{i_m}$. Therefore, for a given random walk
-permutation, each state must be used as a starting value in order to
-calculate its component of the expectation value (i.e. a state of size
-$N$ will have $N$ particles per random walk permutation). When the
-particle transitions between states, for example $i\rightarrow j$, the
-particle weight is multiplied by the transition factor.
-
+Here, the weight on the $m^\text{th}$ step is denoted $W_m$, and each
+random walk starts with unit weight. At every step, contributions to
+each component of $X$ are generated only in the starting state of the
+history, $i_0$, by the source term in the current state,
+$b_{i_m}$. Therefore, for a given random walk permutation, each state
+must be used as a starting value in order to calculate its component
+of the expectation value (i.e. a state of size $N$ will have $N$
+histories per random walk permutation). When there is a transition
+between states, for example $i\rightarrow j$, the weight is multiplied
+by the transition factor.
 \begin{equation}
@@
 \end{equation}
-
 where $p_{ij}$ denotes the probability of transitioning from state $i$
 to $j$. Then, the expected value of $X$ is
-
 \begin{equation}
   \begin{split}
-    E[X(i_0=i)] &= \sum_{\nu}P_\nu X_\nu\\
-    &= \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
+    E[X(i_0=i)] &= \sum_{\nu}P_\nu X_\nu\\ &=
+    \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
     \sum_{i_k}^{N}p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
-    w_{i,i_1}w_{i_1,i_2}\dots w_{i_{k-1},i_k}b_{i_k}\\
-    &= x_i\:,
+    w_{i,i_1}w_{i_1,i_2}\dots w_{i_{k-1},i_k}b_{i_k}\\ &= x_i\:,
   \end{split}
@@
 \end{equation}
-
-where $\nu$ denotes a particular random walk permutation.
-Therefore, the estimator in Eq.~(\ref{eq:estimator_xi}) is an unbiased
-estimator of the components of $\vx$ provided $\rho(\vH) < 1$.
+where $\nu$ denotes a particular random walk permutation.  Therefore,
+the estimator in Eq.~(\ref{eq:estimator_xi}) is an unbiased estimator
+of the components of $\vx$ provided $\rho(\vH) < 1$.
 
@@
 straightforward approach is to set
-
 \begin{equation}
@@
 \end{equation}
-
 Each row of the transition probability matrix, $\vP$, represents a
@@
 terminating event equation that describes the probability that a
-particle ends its walk, or the random walk can be terminated by weight
+history ends its walk, or the random walk can be terminated by weight
 cutoff, $W_c$.  Generally, we choose to terminate random walks using a
@@
 
-\subsection{Adjoint Method}
+%%---------------------------------------------------------------------------%%
+
+\subsubsection{Adjoint Method}
 \label{sec:adjoint-method}
@@
 contributions to every component of $\vx$ during the random walk. In
-the mathematical literature this is referred to as an adjoint method;
-however, to transport practitioners this approach more closely
-resembles a traditional forward method.  In this method, the weight
-change from state $i\rightarrow j$ is
-
+this method, the weight change from state $i\rightarrow j$ is
 \begin{equation}
@@
 \end{equation}
-
 The transition probabilities may be calculated as
-
 \begin{equation}
@@
 \end{equation}
-
 Note that the indices are reversed so that the probabilities are
@@
 estimator for this method is
-
 \begin{equation}
   \begin{split}
-    X &= \sum_{m=0}^{k}W_m\delta_{i_m,i}\\
-    &= \sum_{m=0}^{k}\hat{b}_{i_0}w_{i_0,i_1}w_{i_1,i_2}\ldots
+    X &= \sum_{m=0}^{k}W_m\delta_{i_m,i}\\ &=
+    \sum_{m=0}^{k}\hat{b}_{i_0}w_{i_0,i_1}w_{i_1,i_2}\ldots
     w_{i_{m-1},i_m}\delta_{i_m,i}\:.
@@
 \end{equation}
-
-Here, $\hat{b}_{i_0}$ is the sampled source and initial particle
-weight in state $i_0$.  The Kronecker delta implies that tallies are
-only made in the state where the random walk currently resides.  This
-is the common approach in standard Monte Carlo transport
-simulations. Unlike the direct method, because tallies are made in the
-state in which the particle currently resides, we are no longer
-required to start a particle in each state for each random walk
-permutation. Instead, sampling the source is suffucient and requires
-only one particle per permutation.
+Here, $\hat{b}_{i_0}$ is the sampled source and initial weight in
+state $i_0$.  The Kronecker delta implies that tallies are only made
+in the state where the random walk currently resides.  This is the
+common approach in standard Monte Carlo transport simulations. Unlike
+the direct method, because tallies are made in the state in which the
+random walk currently resides, we are no longer required to start a
+history in each state for each random walk permutation. Instead,
+sampling the source is sufficient and requires only one random walk
+per permutation.
 
@@
 fraction of the starting weight such that
-
 \begin{equation}
@@
 \end{equation}
-
 where $W_f$ is the terminating weight of the random walk and $W_c$ is
@@
 %%---------------------------------------------------------------------------%%
-\section{Monte Carlo Synthetic-Acceleration}
+
+\subsection{Monte Carlo Synthetic-Acceleration}
 \label{sec:iter-refin-monte}
 
-The methods presented in \S~\ref{sec:monte-carlo-matrix} are
-characterized by slow convergence rates bound by the Central Limit
-Theorem.  Halton~\cite{halton_1962,halton_1994} proposed a staged
-residual scheme called Sequential Monte Carlo to speed up the
-convergence of these methods.  A variant of this scheme has been
-successfully applied to the 1D nonlinear thermal radiation diffusion
-equation in Ref.~\cite{evans_2003}.
+The methods presented in \S~\ref{sec:background} are characterized by
+slow convergence rates bound by the Central Limit Theorem.
+Halton~\cite{halton_1962,halton_1994} proposed a staged residual
+scheme called Sequential Monte Carlo to speed up the convergence of
+these methods.  A variant of this scheme has been successfully applied
+to the 1D non-linear thermal radiation diffusion equation in
+Ref.~\cite{evans_2003}.
 
@@
 applied:
-
 \begin{subequations}
   \begin{gather}
-    \vr^{l+1} = \vb - \vA\vx^{l}\:, \\
-    \vA\delta\vx^{l+1} = \vr^{l+1}\:, \label{eq:residual_solve}\\
-    \vx^{l+1} = \vx^{l} + \delta\vx^{l+1}\:.
+    \vr^{l} = \vb - \vA\vx^{l}\:, \\ \vA\delta\vx^{l+1} =
+    \vr^{l}\:, \label{eq:residual_solve}\\ \vx^{l+1} = \vx^{l} +
+    \delta\vx^{l+1}\:.
   \end{gather}
 \end{subequations}
-
 In this scheme, the Monte Carlo adjoint method is used to estimate the
@@
 to iterative refinement; the exception being that there is no update
-to to $x^{l+1}$ between residual iterations.
+to $x^{l+1}$ between residual iterations.
 
@@
 Eq.~(\ref{eq:richardson}) from Eq:~(\ref{eq:point_iteration}):
-
 \begin{equation}
@@
 \end{equation}
-
 where
-
-\begin{equation}
-  \delta \vx^{l+1} = \vx - \vx^{l+1}
-  \label{eq:lplus1error}
-\end{equation}
-
-is defined as the error at iteration $l+1$. We then subtract
-$(\vI-\vA) \delta \vx^{l+1}$ from Eq.~(\ref{eq:synthetic_error}),
-giving a linear system to solve for the error:
-
+\begin{equation}
+  \delta \vx^{l} = \vx - \vx^{l}
+\end{equation}
+is defined as the error at iteration $l$. We then subtract $(\vI-\vA)
+\delta \vx^{l+1}$ from Eq.~(\ref{eq:synthetic_error}), giving a linear
+system to solve for the error:
 \begin{equation}
   \begin{split}
-    \vA \delta \vx^{l+1} &= (\vI - \vA)(\vx^{l+1} - \vx^{l}) \\
-    &= \vr^{l+1}\:.
+    \vA \delta \vx^{l+1} &= (\vI - \vA)(\vx^{l+1} - \vx^{l}) \\ &=
+    \vr^{l+1}\:.
   \end{split}
@@
 The following scheme will then converge in one iteration:
-
 \begin{subequations}
   \begin{gather}
-    \vx^{l+1} = (\vI - \vA)\vx^l + \vb\:,\\
-    %%
-    \vA \delta \vx^{l+1} = \vr^{l+1}
+    \vx^{l+1} = (\vI - \vA)\vx^l + \vb\:,\\ \vA \delta \vx^{l+1} =
+    \vr^{l+1}\;,
     \label{eq:synthetic-residual-solve}\\ 
-    %% 
     \vx=\vx^{l+1}+\delta\vx^{l+1}\:.
   \end{gather}
-  \label{eq:general-synthetic}
 \end{subequations}
-
 To accelerate the fixed-point method and avoid inverting $\vA$
@@
 be defined as follows:
-
 \begin{subequations}
+  \label{eq:MCSA-iteration}
   \begin{gather}
-    \vx^{l+1/2} = (\vI - \vA)\vx^l + \vb\:,\\
-    %%
-    \vr^{l+1/2} = \vb - \vA\vx^{l+1/2}\:,
+    \vx^{l+1/2} = (\vI - \vA)\vx^l + \vb\:,\\  \vr^{l+1/2} = \vb -
+    \vA\vx^{l+1/2}\:,
     \label{eq:MCSA-residual}\\     
-    %%
      \hat{\vA}\delta\vx^{l+1/2} = \vr^{l+1/2}\:,
     \label{eq:MCSA-residual_solve}\\ 
-    %% 
      \vx^{l+1}=\vx^{l+1/2}+\delta\vx^{l+1/2}\:.
   \end{gather}
-  \label{eq:general-MCSA}
 \end{subequations}
-
 The hat on $\vA$ in Eq.~(\ref{eq:MCSA-residual_solve}) indicates that
 the Monte Carlo solution only approximately inverts this operator.
-However, $\hat{\vA}$ is the same operator as $\vA$;
-$\hat{\vA}\equiv\vA$.  Thus, we have defined a scheme in which the
-initial estimate of $\vx$ in each iteration is updated using a single
-fixed-point iteration\footnote{Although, a Krylov method could be used
-  instead of a stationary method here.}.  The residual is calculated
-and is used as a source to estimate the error, $\delta\vx^{l+1/2}$, by
-solving Eq.~(\ref{eq:MCSA-residual_solve}) via the Monte Carlo adjoint
+Thus, we have defined a scheme in which the initial estimate of $\vx$
+in each iteration is updated using a single fixed-point
+iteration\footnote{Although, a subspace method could be used instead of
+  a stationary method here.}.  The residual is calculated and is used
+as a source to estimate the error, $\delta\vx^{l+1/2}$, by solving
+Eq.~(\ref{eq:MCSA-residual_solve}) via the Monte Carlo adjoint
 method. The error is then used to calculate an updated iterate of the
@@
 
-The MCSA method defined in Eq.(~\ref{eq:general-MCSA}) uses the
-adjoint method in Eq.~(\ref{eq:MCSA-residual_solve}) to estimate the
-error instead of the direct method outlined in
-\S~\ref{sec:direct-method}. To demonstrate the effectiveness of the
-adjoint method over the direct method within the context of MCSA, we
-choose the two-dimensional time-dependent Poisson equation as a simple
-model problem:
+The MCSA method defined in Eq.(~\ref{eq:MCSA-residual_solve}) uses the
+adjoint method to estimate the error instead of the direct method
+outlined in \S~\ref{sec:direct-method}. To demonstrate the
+effectiveness of the adjoint method over the direct method within the
+context of MCSA, we choose the two-dimensional time-dependent Poisson
+equation as a simple model problem:
 
@@
   \nabla^2_5 = \frac{1}{\Delta^2}[u_{i-1,j} + u_{i+1,j} + u_{i,j-1} +
-    u_{i,j+1} - 4 u_{i,j}]
+    u_{i,j+1} - 4 u_{i,j}]\:,
   \label{eq:five_point_stencil}
@@
 and a fourth-order nine point stencil both assuming a grid size of
-$\Delta$ in both the $i$ and $j$ directions.
+$\Delta$ in both the $i$ and $j$ directions
 
@@
   \nabla^2_9 = \frac{1}{6\Delta^2}[4 u_{i-1,j} + 4 u_{i+1,j} + 4
-    u_{i,j-1} + 4 u_{i,j+1} + u_{i-1,j-1}\\
-    + u_{i-1,j+1} + u_{i+1,j-1} + u_{i+1,j+1} - 20 u_{i,j}]
+    u_{i,j-1} + 4 u_{i,j+1} + u_{i-1,j-1}\\ + u_{i-1,j+1} +
+    u_{i+1,j-1} + u_{i+1,j+1} - 20 u_{i,j}]\:.
   \label{eq:nine_point_stencil}
@@
 For a single time step solution, we then have the following sparse
-linear system to be solved with the MCSA method.
-
-\begin{equation}
-  \vA \vu^{n+1} = \vu^n
+linear system to be solved with the MCSA method:
+
+\begin{equation}
+  \vA \vu^{n+1} = \vu^n\:.
   \label{eq:poisson_eq_lin_sys}
@@
 method. To assess both the CPU time and number of iterations required
-to converge to a solution, a problem of constant $\Delta$ and was used
+to converge to a solution, a problem of constant $\Delta$ was used
 with varying values of grid size, fixing the spectral radius of the
@@
     \phi(\ve{r},t) &=
-    \int_{4\pi}d\bOmega\int_{\nu}d\nu\:\psi(\ve{r},\nu,\hat{\bOmega},t) \\
-    &= \int_{\nu}d\nu\: 4\pi B(\nu,T) \\
-    &= acT^{4}\:,
+    \int_{4\pi}d\bOmega\int_{\nu}d\nu\:\psi(\ve{r},\nu,\hat{\bOmega},t)
+    \\ &= \int_{\nu}d\nu\: 4\pi B(\nu,T) \\ &= acT^{4}\:,
   \end{split}
@@
 \begin{align}
-  \sign &= \frac{\rho \Cv^n}{4ac(\Tn)^3\dt} + \frac{1}{c\dt}\:,\\
-  \qn &= \sign\phin + Q^n\:,\\
-  \phin &= ac(\Tn)^4\:.
+  \sign &= \frac{\rho \Cv^n}{4ac(\Tn)^3\dt} + \frac{1}{c\dt}\:,\\ \qn
+  &= \sign\phin + Q^n\:,\\ \phin &= ac(\Tn)^4\:.
 \end{align}
@@
 \begin{figure}[h!]
-  \centerline{
-    \input{mesh_cell.pdftex_t}}
+  \centerline{ \input{mesh_cell.pdftex_t}}
   \caption{Orthogonal 3D mesh cell.}
@@
   \begin{split}
-    \iiint\dive{F}\,d\ve{r} &= \oint\hat{\ve{n}}\cdot\ve{F}\,dA\\
-    &= \sum_{f=1}^{6}\hat{\ve{n}}_f\cdot\ve{F}_f A_f\:,
+    \iiint\dive{F}\,d\ve{r} &= \oint\hat{\ve{n}}\cdot\ve{F}\,dA\\ &=
+    \sum_{f=1}^{6}\hat{\ve{n}}_f\cdot\ve{F}_f A_f\:,
   \end{split}
@@
   F_{l+1/2} &= -D_{l+1/2}\frac{\phi_{l+1} -
-    \phi_{l}}{\Delta_{l+1/2}}\:, \label{eq:F_l+1/2}\\
-  F_{l-1/2} &= -D_{l-1/2}\frac{\phi_{l} -
+    \phi_{l}}{\Delta_{l+1/2}}\:, \label{eq:F_l+1/2}\\ F_{l-1/2} &=
+  -D_{l-1/2}\frac{\phi_{l} -
     \phi_{l-1}}{\Delta_{l-1/2}}\:, \label{eq:F_l-1/2}
@@
 \begin{align}
-  \Delta_{l+1/2} &= \frac{\Delta_{l+1} + \Delta_{l}}{2}\:,\\
-  \Delta_{l-1/2} &= \frac{\Delta_{l} + \Delta_{l-1}}{2}\:.
+  \Delta_{l+1/2} &= \frac{\Delta_{l+1} +
+    \Delta_{l}}{2}\:,\\ \Delta_{l-1/2} &= \frac{\Delta_{l} +
+    \Delta_{l-1}}{2}\:.
 \end{align}
@@
   \begin{aligned}
-    F_{l+1/2}^{+} &= F_{l+1/2}^{-}\:,\\
-    2D_{l+1}\frac{\phi_{l+1} - \phi_{l+1/2}}{\Delta_{l+1}} &=
-    2D_{l}\frac{\phi_{l+1/2} - \phi_{l}}{\Delta_l}\:,\\
-    \frac{1}{3\sigma_{l+1}}\frac{\phi_{l+1} - \phi_{l+1/2}}{\Delta_{l+1}} &=
+    F_{l+1/2}^{+} &= F_{l+1/2}^{-}\:,\\ 2D_{l+1}\frac{\phi_{l+1} -
+      \phi_{l+1/2}}{\Delta_{l+1}} &= 2D_{l}\frac{\phi_{l+1/2} -
+      \phi_{l}}{\Delta_l}\:,\\ \frac{1}{3\sigma_{l+1}}\frac{\phi_{l+1}
+      - \phi_{l+1/2}}{\Delta_{l+1}} &=
     \frac{1}{3\sigma_{l}}\frac{\phi_{l+1/2} - \phi_{l}}{\Delta_l}\:,
@@
 where we have dropped the subscript $\mathrm{R}$ from $\sigma$ for
-clarity.  Solving this system for $\phi_{l+1/2}$ and plugging into
+clarity.  Solving this system for $\phi_{l+1/2}$ and applying to
 either $F_{l+1/2}^{+}$ or $F_{l+1/2}^{-}$ and setting the resulting
@@
   \begin{aligned}
-    \sigT\phi_{ijk} 
-    - \sigm_{i+1}\frac{\Delta_{i+1}}{\Di}\phi_{i+1\,jk} &
-    - \sigp_{i-1}\frac{\Delta_{i-1}}{\Di}\phi_{i-1\,jk}\\
-    &- \sigm_{j+1}\frac{\Delta_{j+1}}{\Dj}\phi_{i\,j+1\,k}
-    - \sigp_{j-1}\frac{\Delta_{j-1}}{\Dj}\phi_{i\,j-1\,k}\\ 
-    &- \sigm_{k+1}\frac{\Delta_{k+1}}{\Dk}\phi_{ij\,k+1}
-    - \sigp_{k-1}\frac{\Delta_{k-1}}{\Dk}\phi_{ij\,k-1}
-    = \qn_{ijk}\:.
+    \sigT\phi_{ijk} -
+    \sigm_{i+1}\frac{\Delta_{i+1}}{\Di}\phi_{i+1\,jk} & -
+    \sigp_{i-1}\frac{\Delta_{i-1}}{\Di}\phi_{i-1\,jk}\\ &-
+    \sigm_{j+1}\frac{\Delta_{j+1}}{\Dj}\phi_{i\,j+1\,k} -
+    \sigp_{j-1}\frac{\Delta_{j-1}}{\Dj}\phi_{i\,j-1\,k}\\ &-
+    \sigm_{k+1}\frac{\Delta_{k+1}}{\Dk}\phi_{ij\,k+1} -
+    \sigp_{k-1}\frac{\Delta_{k-1}}{\Dk}\phi_{ij\,k-1} = \qn_{ijk}\:.
   \end{aligned}
@@
 \begin{equation}
-  \sigT = \sigp_i + \sigm_i + \sigp_j + \sigm_j + \sigp_k + \sigm_k
-  + \sign_{ijk}\:,
+  \sigT = \sigp_i + \sigm_i + \sigp_j + \sigm_j + \sigp_k + \sigm_k +
+  \sign_{ijk}\:,
 \end{equation}
@@
   \begin{align}
-    \sigp_{i} &= \frac{1}{\Di}\frac{2}{3\sigma_{ijk}\Di
-      + 3\sigma_{i+1\,jk}\Delta_{i+1}}\:,\\
-    \sigm_{i} &= \frac{1}{\Di}\frac{2}{3\sigma_{ijk}\Di
-      + 3\sigma_{i-1\,jk}\Delta_{i-1}}\:,\\
-    \sigp_{j} &= \frac{1}{\Dj}\frac{2}{3\sigma_{ijk}\Dj
-      + 3\sigma_{i\,j+1\,k}\Delta_{j+1}}\:,\\
-    \sigm_{j} &= \frac{1}{\Dj}\frac{2}{3\sigma_{ijk}\Dj
-      + 3\sigma_{i\,j-1\,k}\Delta_{j-1}}\:,\\
-    \sigp_{k} &= \frac{1}{\Dk}\frac{2}{3\sigma_{ijk}\Dk
-      + 3\sigma_{ij\,k+1}\Delta_{k+1}}\:,\\
-    \sigm_{k} &= \frac{1}{\Dk}\frac{2}{3\sigma_{ijk}\Dk
-      + 3\sigma_{ij\,k-1}\Delta_{k-1}}\:. 
+    \sigp_{i} &= \frac{1}{\Di}\frac{2}{3\sigma_{ijk}\Di +
+      3\sigma_{i+1\,jk}\Delta_{i+1}}\:,\\ \sigm_{i} &=
+    \frac{1}{\Di}\frac{2}{3\sigma_{ijk}\Di +
+      3\sigma_{i-1\,jk}\Delta_{i-1}}\:,\\ \sigp_{j} &=
+    \frac{1}{\Dj}\frac{2}{3\sigma_{ijk}\Dj +
+      3\sigma_{i\,j+1\,k}\Delta_{j+1}}\:,\\ \sigm_{j} &=
+    \frac{1}{\Dj}\frac{2}{3\sigma_{ijk}\Dj +
+      3\sigma_{i\,j-1\,k}\Delta_{j-1}}\:,\\ \sigp_{k} &=
+    \frac{1}{\Dk}\frac{2}{3\sigma_{ijk}\Dk +
+      3\sigma_{ij\,k+1}\Delta_{k+1}}\:,\\ \sigm_{k} &=
+    \frac{1}{\Dk}\frac{2}{3\sigma_{ijk}\Dk +
+      3\sigma_{ij\,k-1}\Delta_{k-1}}\:.
   \end{align}
@@
 \begin{align*}
-  \sigp_{i-1} &= \frac{1}{\Delta_{i-1}}\frac{2}{3\sigma_{ijk}\Di
-    + 3\sigma_{i-1\,jk}\Delta_{i-1}}\:,\\
-  \sigm_{i+1} &= \frac{1}{\Delta_{i+1}}\frac{2}{3\sigma_{ijk}\Di
-    + 3\sigma_{i+1\,jk}\Delta_{i+1}}\:,\\
-  \ldots\:.
+  \sigp_{i-1} &= \frac{1}{\Delta_{i-1}}\frac{2}{3\sigma_{ijk}\Di +
+    3\sigma_{i-1\,jk}\Delta_{i-1}}\:,\\ \sigm_{i+1} &=
+  \frac{1}{\Delta_{i+1}}\frac{2}{3\sigma_{ijk}\Di +
+    3\sigma_{i+1\,jk}\Delta_{i+1}}\:,\\ \ldots\:.
 \end{align*}
@@
 Equations~(\ref{eq:discrete-diffusion}) through (\ref{eq:sigmas})
-represent the discretized, 1T diffusion equation on an orthogonal
-3D grid.
+represent the discretized, 1T diffusion equation on an orthogonal 3D
+grid.
 
@@
 \begin{equation}
-  f_b^\pm = \frac{1}{4}\phi - \frac{1}{2}\ve{F}\cdot\hat{\ve{n}}\:,\quad
-  \vr\in \partial\Gamma\:,
+  f_b^\pm = \frac{1}{4}\phi -
+  \frac{1}{2}\ve{F}\cdot\hat{\ve{n}}\:,\quad \vr\in \partial\Gamma\:,
 \end{equation}
@@
 \begin{equation}
-  F_{1/2} = -\frac{2}{3\sigma_1}\frac{\phi_1 - \phi_{1/2}}{\Delta_1}\:.
+  F_{1/2} = -\frac{2}{3\sigma_1}\frac{\phi_1 -
+    \phi_{1/2}}{\Delta_1}\:.
   \label{eq:low-boundary-flux}
@@
 
-Solving for $\phi_{1/2}$ and plugging into
+Solving for $\phi_{1/2}$ and applying to
 Eq.~(\ref{eq:low-boundary-flux}) yields an expression for the low-side
@@
   \text{low boundary:}\quad & \sigma_1^{-} = \frac{1}{\Delta_1}
-  \frac{2}{3\sigma_1\Delta_1 + 4}\:,\\
-  \text{high boundary:}\quad & \sigma_{N_l}^{+} = \frac{1}{\Delta_{N_l}}
+  \frac{2}{3\sigma_1\Delta_1 + 4}\:,\\ \text{high boundary:}\quad &
+  \sigma_{N_l}^{+} = \frac{1}{\Delta_{N_l}}
   \frac{2}{3\sigma_{N_l}\Delta_{N_l} + 4}\:.
@@
   \text{low boundary:}\quad & \frac{8}{\Delta_1}
-  \frac{1}{3\sigma_1\Delta_1 + 4}f_b^{+}\:,\\
-  \text{high boundary:}\quad & \frac{8}{\Delta_{N_l}}
+  \frac{1}{3\sigma_1\Delta_1 + 4}f_b^{+}\:,\\ \text{high
+    boundary:}\quad & \frac{8}{\Delta_{N_l}}
   \frac{1}{3\sigma_{N_l}\Delta_{N_l} + 4}f_b^{-}\:.
@@
 \begin{align}
-  F_{1/2} &= 0\:, \\
-  F_{N_l+1/2} &= 0\:.
+  F_{1/2} &= 0\:, \\ F_{N_l+1/2} &= 0\:.
 \end{align}
@@
   \begin{gathered}
-    \ve{M}^{-1}\ve{D}\ve{M}^{-1}\ve{M}\bphi = \ve{M}^{-1}\ve{q}\:,\\
-    (\ve{LL}^T)^{-1}\ve{D}(\ve{LL}^T)^{-1}\ve{LL}^T\bphi =
-    (\ve{LL}^T)^{-1}\ve{q}\:, \\
-    \ve{L}^{-T}\ve{L}^{-1}\ve{D}\ve{L}^{-T}\ve{L}^T\phi =
-    \ve{L}^{-T}\ve{L}^{-1}\ve{q}\:,\\ 
-    (\ve{L}^{-1}\ve{D}\ve{L}^{-T})\ve{L}^T\bphi = \ve{L}^{-1}\ve{q}\:,\\
-    \hat{\ve{D}}\hat{\bphi} = \hat{\ve{q}}\:.
+    \ve{M}^{-1}\ve{D}\ve{M}^{-1}\ve{M}\bphi =
+    \ve{M}^{-1}\ve{q}\:,\\ (\ve{LL}^T)^{-1}\ve{D}(\ve{LL}^T)^{-1}\ve{LL}^T\bphi
+    = (\ve{LL}^T)^{-1}\ve{q}\:,
+    \\ \ve{L}^{-T}\ve{L}^{-1}\ve{D}\ve{L}^{-T}\ve{L}^T\phi =
+    \ve{L}^{-T}\ve{L}^{-1}\ve{q}\:,\\ (\ve{L}^{-1}\ve{D}\ve{L}^{-T})\ve{L}^T\bphi
+    = \ve{L}^{-1}\ve{q}\:,\\ \hat{\ve{D}}\hat{\bphi} = \hat{\ve{q}}\:.
   \end{gathered}
@@
   \begin{aligned}
-    \phi_{ijk} 
-    - \frac{\sigm_{i+1}}{\sigT}\frac{\Delta_{i+1}}{\Di}&\phi_{i+1\,jk}
-    - \frac{\sigp_{i-1}}{\sigT}\frac{\Delta_{i-1}}{\Di}\phi_{i-1\,jk}\\
-    &- \frac{\sigm_{j+1}}{\sigT}\frac{\Delta_{j+1}}{\Dj}\phi_{i\,j+1\,k}
-    -\frac{\sigp_{j-1}}{\sigT}\frac{\Delta_{j-1}}{\Dj}\phi_{i\,j-1\,k}\\
-    &- \frac{\sigm_{k+1}}{\sigT}\frac{\Delta_{k+1}}{\Dk}\phi_{ij\,k+1}
-    - \frac{\sigp_{k-1}}{\sigT}\frac{\Delta_{k-1}}{\Dk}\phi_{ij\,k-1}
-    = \frac{\qn_{ijk}}{\sigT}\:.
+    \phi_{ijk} -
+    \frac{\sigm_{i+1}}{\sigT}\frac{\Delta_{i+1}}{\Di}&\phi_{i+1\,jk} -
+    \frac{\sigp_{i-1}}{\sigT}\frac{\Delta_{i-1}}{\Di}\phi_{i-1\,jk}\\ &-
+    \frac{\sigm_{j+1}}{\sigT}\frac{\Delta_{j+1}}{\Dj}\phi_{i\,j+1\,k}
+    -\frac{\sigp_{j-1}}{\sigT}\frac{\Delta_{j-1}}{\Dj}\phi_{i\,j-1\,k}\\ &-
+    \frac{\sigm_{k+1}}{\sigT}\frac{\Delta_{k+1}}{\Dk}\phi_{ij\,k+1} -
+    \frac{\sigp_{k-1}}{\sigT}\frac{\Delta_{k-1}}{\Dk}\phi_{ij\,k-1} =
+    \frac{\qn_{ijk}}{\sigT}\:.
   \end{aligned}
@@
 For clarity, we rewrite the iteration scheme in
-Eqs.~(\ref{eq:general-MCSA}) using the diffusion operator notation
+Eq.~(\ref{eq:MCSA-iteration}) using the diffusion operator notation
 from Eq.~(\ref{eq:preconditioned-diffusion}),
@@
 \begin{align*}
-  %%
    \bphi^{l+1/2} = (\vI - \ve{M}^{-1}\ve{D})\bphi^{l} +
   \ve{M}^{-1}\ve{q}\:, \quad & (\text{fixed-point iteration})\\ 
-  %%
   \vr^{l+1/2} = \ve{M}^{-1}\ve{q} - \ve{M}^{-1}\ve{D}\bphi^{l+1/2}\:,
   \quad & (\text{compute residual})\\ 
-  %%
   \ve{M}^{-1}\ve{D}\delta\bphi^{l+1/2} = \vr^{l+1/2}\:, \quad&
   (\text{estimate $\delta\phi$ using adjoint Monte Carlo method})\\ 
-  %%
   \bphi^{l+1} = \bphi^{l+1/2} + \delta\bphi^{l+1/2}\:. \quad&
@@
 method only approximately inverts $\ve{M}^{-1}\ve{D}$.  We now
-describe each of these s.pdf in turn.
+describe each of these in turn.
 
@@
     +
-    \frac{\sigp_{i-1}}{\sigT}\frac{\Delta_{i-1}}{\Di}\phi_{i-1\,jk}^l \\
-    &+ \frac{\sigm_{j+1}}{\sigT}\frac{\Delta_{j+1}}{\Dj}\phi_{i\,j+1\,k}^l 
-    + \frac{\sigp_{j-1}}{\sigT}\frac{\Delta_{j-1}}{\Dj}\phi_{i\,j-1\,k}^l \\
-    &+ \frac{\sigm_{k+1}}{\sigT}\frac{\Delta_{k+1}}{\Dk}\phi_{ij\,k+1}^l
-    + \frac{\sigp_{k-1}}{\sigT}\frac{\Delta_{k-1}}{\Dk}\phi_{ij\,k-1}^l
+    \frac{\sigp_{i-1}}{\sigT}\frac{\Delta_{i-1}}{\Di}\phi_{i-1\,jk}^l
+    \\ &+
+    \frac{\sigm_{j+1}}{\sigT}\frac{\Delta_{j+1}}{\Dj}\phi_{i\,j+1\,k}^l
+    +
+    \frac{\sigp_{j-1}}{\sigT}\frac{\Delta_{j-1}}{\Dj}\phi_{i\,j-1\,k}^l
+    \\ &+
+    \frac{\sigm_{k+1}}{\sigT}\frac{\Delta_{k+1}}{\Dk}\phi_{ij\,k+1}^l
+    +
+    \frac{\sigp_{k-1}}{\sigT}\frac{\Delta_{k-1}}{\Dk}\phi_{ij\,k-1}^l
     + \frac{\qn_{ijk}}{\sigT}\:.
@@
   \phi_1 &= \frac{\sigm_{7\rightarrow
-      1}}{\sigma_{T\,1}}\frac{\Delta_7}{\Delta_1}\phi_7 + \ldots\:,\\
-  %%
-  \phi_2 &= \frac{\sigp_{7\rightarrow
-      2}}{\sigma_{T\,2}}\frac{\Delta_7}{\Delta_2}\phi_7 + \ldots\:,\\
-  %%
+      1}}{\sigma_{T\,1}}\frac{\Delta_7}{\Delta_1}\phi_7 +
+  \ldots\:,\\ \phi_2 &= \frac{\sigp_{7\rightarrow 2}}
+          {\sigma_{T\,2}}\frac{\Delta_7}{\Delta_2}\phi_7 + \ldots\:,\\ 
       \phi_3 &= \frac{\sigm_{7\rightarrow
-      3}}{\sigma_{T\,3}}\frac{\Delta_7}{\Delta_3}\phi_7 + \ldots\:,\\
-  %%
-  \phi_4 &= \frac{\sigp_{7\rightarrow
-      4}}{\sigma_{T\,4}}\frac{\Delta_7}{\Delta_4}\phi_7 + \ldots\:,\\
-  %%
-  \phi_5 &= \frac{\sigm_{7\rightarrow
-      5}}{\sigma_{T\,5}}\frac{\Delta_7}{\Delta_5}\phi_7 + \ldots\:,\\
-  %%
-  \phi_6 &= \frac{\sigp_{7\rightarrow
-      6}}{\sigma_{T\,6}}\frac{\Delta_7}{\Delta_6}\phi_7 + \ldots\:,\\
-  %%
+          3}}{\sigma_{T\,3}}\frac{\Delta_7}{\Delta_3}\phi_7 +
+      \ldots\:,\\  \phi_4 &= \frac{\sigp_{7\rightarrow
+      4}}{\sigma_{T\,4}}\frac{\Delta_7}{\Delta_4}\phi_7 +
+          \ldots\:,\\  \phi_5 &= \frac{\sigm_{7\rightarrow
+          5}}{\sigma_{T\,5}}\frac{\Delta_7}{\Delta_5}\phi_7 +
+              \ldots\:,\\  \phi_6 &= \frac{\sigp_{7\rightarrow
+              6}}{\sigma_{T\,6}}\frac{\Delta_7}{\Delta_6}\phi_7 +
+                  \ldots\:,\\ 
 \end{align*}
@@
   \begin{pmatrix}
-    h_{1,7} &
-    h_{2,7} &
-    h_{3,7} &
-    h_{4,7} &
-    h_{5,7} &
-    h_{6,7} 
+    h_{1,7} & h_{2,7} & h_{3,7} & h_{4,7} & h_{5,7} & h_{6,7}
   \end{pmatrix}^{T}
@@
   \begin{pmatrix}
-    \frac{\sigm_{7\rightarrow 1}}{\sigma_{T\,1}}\frac{\Delta_7}{\Delta_1} \\
-    \frac{\sigp_{7\rightarrow 2}}{\sigma_{T\,2}}\frac{\Delta_7}{\Delta_2} \\
-    \frac{\sigm_{7\rightarrow 3}}{\sigma_{T\,3}}\frac{\Delta_7}{\Delta_3} \\
-    \frac{\sigp_{7\rightarrow 4}}{\sigma_{T\,4}}\frac{\Delta_7}{\Delta_4} \\
-    \frac{\sigm_{7\rightarrow 5}}{\sigma_{T\,5}}\frac{\Delta_7}{\Delta_5} \\
-    \frac{\sigp_{7\rightarrow 6}}{\sigma_{T\,6}}\frac{\Delta_7}{\Delta_6}
+    \frac{\sigm_{7\rightarrow
+        1}}{\sigma_{T\,1}}\frac{\Delta_7}{\Delta_1}
+    \\ \frac{\sigp_{7\rightarrow
+        2}}{\sigma_{T\,2}}\frac{\Delta_7}{\Delta_2}
+    \\ \frac{\sigm_{7\rightarrow
+        3}}{\sigma_{T\,3}}\frac{\Delta_7}{\Delta_3}
+    \\ \frac{\sigp_{7\rightarrow
+        4}}{\sigma_{T\,4}}\frac{\Delta_7}{\Delta_4}
+    \\ \frac{\sigm_{7\rightarrow
+        5}}{\sigma_{T\,5}}\frac{\Delta_7}{\Delta_5}
+    \\ \frac{\sigp_{7\rightarrow
+        6}}{\sigma_{T\,6}}\frac{\Delta_7}{\Delta_6}
   \end{pmatrix}\:,
@@
 \begin{equation*}
-  P_{7\rightarrow n} = p_{7,n} = \frac{|h_{n,7}|}{|h_{1,7} + 
-    h_{2,7} + h_{3,7} + h_{4,7} + h_{5,7} + h_{6,7}|}\:,\quad
-  n\in \{1,\ldots,6\}\:.
+  P_{7\rightarrow n} = p_{7,n} = \frac{|h_{n,7}|}{|h_{1,7} + h_{2,7} +
+    h_{3,7} + h_{4,7} + h_{5,7} + h_{6,7}|}\:,\quad n\in
+  \{1,\ldots,6\}\:.
 \end{equation*}
@@
 \begin{equation*}
-  W_{m} = \frac{h_{n,7}}{p_{7,n}}W_{m-1}\:,\quad
-  n\in \{1,\ldots,6\}\:,
+  W_{m} = \frac{h_{n,7}}{p_{7,n}}W_{m-1}\:,\quad n\in
+  \{1,\ldots,6\}\:,
 \end{equation*}
@@
 \begin{figure}[h!]
-  \centerline{
-    \input{3D_mesh_segment.pdftex_t}}
+  \centerline{ \input{3D_mesh_segment.pdftex_t}}
   \caption{3D mesh-segment.  The particle's current cell (in blue) is
@@
   \begin{tabular}{ll}\hline
-    %%
-    B.C. & $f_b^{+}(x=0,T=1.0) = \frac{ac}{4}T^4$ \\
-    & $\ve{F} = 0\quad \partial\Gamma \in \{x^{+}, y^{-}, y^{+},
-    z^{-}, z^{+}\}$ \\
-    %%
-    material properties & $\rho = 3.0\,\text{g}\cdot\text{cm}^{-3}$ \\
-    & $\Cv = 0.1\,\text{Jerks}\cdot\text{g}^{-1}\cdot\text{keV}^{-1}$\\
-    & $T(t=0) = \sn{1}{-4}\,\text{keV}$\\
-    %% 
-    opacity & $\sigma = 100T^{-3}\,\text{cm}^2\cdot\text{g}^{-1}$ \\
-    %%
-    mesh & $\Di = \Dj = \Dk = 0.01\,\text{cm}$ \\
-    & $N_i = 100\quad N_j = N_k = 4$ \\
-    \hline
+    B.C. & $f_b^{+}(x=0,T=1.0) = \frac{ac}{4}T^4$ \\ & $\ve{F} =
+    0\quad \partial\Gamma \in \{x^{+}, y^{-}, y^{+}, z^{-}, z^{+}\}$
+    \\ material properties & $\rho =
+    3.0\,\text{g}\cdot\text{cm}^{-3}$ \\ & $\Cv =
+    0.1\,\text{Jerks}\cdot\text{g}^{-1}\cdot\text{keV}^{-1}$\\ &
+    $T(t=0) = \sn{1}{-4}\,\text{keV}$\\  opacity & $\sigma =
+    100T^{-3}\,\text{cm}^2\cdot\text{g}^{-1}$ \\  mesh & $\Di = \Dj
+    = \Dk = 0.01\,\text{cm}$ \\ & $N_i = 100\quad N_j = N_k = 4$
+    \\ \hline
   \end{tabular}
@@
 \begin{figure}[h!]
-  \centerline{
-    \includegraphics[width=5in,clip]{marshak_10.pdf}}
-  \caption{
-    Solution of the Marshak wave problem at $t=10$~sh.  The analytic
-    solution is compared to the MCSA solution.  Note that the MCSA,
-    PCG, and PFIXED methods all give numerically identical answers.}
+  \centerline{ \includegraphics[width=5in,clip]{marshak_10.pdf}}
+  \caption{ Solution of the Marshak wave problem at $t=10$~sh.  The
+    analytic solution is compared to the MCSA solution.  Note that the
+    MCSA, PCG, and PFIXED methods all give numerically identical
+    answers.}
   \label{fig:marshak_10}
@@
 \begin{figure}[h!]
-  \centerline{
-    \includegraphics[width=5in,clip]{spectral_radius.pdf}}
+  \centerline{ \includegraphics[width=5in,clip]{spectral_radius.pdf}}
   \caption{Spectral radius of the iteration matrix, $\ve{H}$, for the
@@
 \begin{figure}[h!]
-  \centerline{
-    \includegraphics[width=5in,clip]{mrshk_np_CPU.pdf}}
+  \centerline{ \includegraphics[width=5in,clip]{mrshk_np_CPU.pdf}}
   \caption{CPU time versus number of particles per stage for the
@@
 \begin{figure}[h!]
-  \centerline{
-    \includegraphics[width=5in,clip]{mrshk_np_Ns.pdf}}
-  \caption{
-    Maximum number of stages per cycle versus number of particles per
-    stage for the Marshak problem.  The number of particles is the
-    requested number.  In each stage the exact number can vary
-    depending upon the number of cells and source strength per cell.}
+  \centerline{ \includegraphics[width=5in,clip]{mrshk_np_Ns.pdf}}
+  \caption{ Maximum number of stages per cycle versus number of
+    particles per stage for the Marshak problem.  The number of
+    particles is the requested number.  In each stage the exact number
+    can vary depending upon the number of cells and source strength
+    per cell.}
   \label{fig:CPU_Ns}
@@
 \begin{figure}[h!]
-  \centerline{
-    \includegraphics[width=5in,clip]{mrshk_wc_CPU.pdf}}
-  \caption{
-    CPU time versus relative weight cutoff for the Marshak problem.
-    Results are shown for 10 and 50 requested particles per stage.
-    The numbers in parenthesis show the maximum number of particles
-    run in a stage during the course of the problem.}
+  \centerline{ \includegraphics[width=5in,clip]{mrshk_wc_CPU.pdf}}
+  \caption{ CPU time versus relative weight cutoff for the Marshak
+    problem.  Results are shown for 10 and 50 requested particles per
+    stage.  The numbers in parenthesis show the maximum number of
+    particles run in a stage during the course of the problem.}
   \label{fig:CPU_wc}
@@
     \begin{tabular}{lrr}\hline\hline
-      \multicolumn{1}{c}{Method} & 
-      \multicolumn{1}{c}{Max Iterations} & 
-      \multicolumn{1}{c}{Relative CPU Time}\\\hline\hline
-      %%
-      PCG & 11 & 1.11 \\
-      %%
-      PFIX & 11 & 1.42 \\
-      %%
-      MCSA & 11 & 1.0 \\
-      \hline\hline
+      \multicolumn{1}{c}{Method} & \multicolumn{1}{c}{Max Iterations}
+      & \multicolumn{1}{c}{Relative CPU Time}\\\hline\hline  PCG &
+      11 & 1.11 \\  PFIX & 11 & 1.42 \\  MCSA & 11 & 1.0
+      \\ \hline\hline
     \end{tabular}
@@
   \begin{tabular}{ll}\hline
-    %%
-    B.C. & $f_b^{+}(x=0,T=0.5) = \frac{ac}{4}T^4$\\
-    & $\ve{F} = 0\quad \partial\Gamma \in \{x^{+}, y^{-}, y^{+},
-    z^{-}, z^{+}\}$ \\
-    %%
-    material properties 
-    & $\rho = 1.5\,\text{g}\cdot\text{cm}^{-3}$ (duct) \\
-    & $\rho = 8.0\,\text{g}\cdot\text{cm}^{-3}$ (wall) \\
-    & $\rho = 4.0\,\text{g}\cdot\text{cm}^{-3}$ (foil) \\
-    & $\Cv = 0.1\,\text{Jerks}\cdot\text{g}^{-1}\cdot\text{keV}^{-1}$\\
-    & $T(t=0) = \sn{1}{-4}\,\text{keV}$\\
-    %% 
-    opacity 
-    & $\sigma = T^{-3}\,\text{cm}^2\cdot\text{g}^{-1}$ (duct) \\
-    & $\sigma = 1000T^{-1}\,\text{cm}^2\cdot\text{g}^{-1}$ (wall) \\
-    & $\sigma = 5T^{-2}\,\text{cm}^2\cdot\text{g}^{-1}$ (foil) \\
-    %%
-    mesh & $\Di = \Dj = \Dk = 0.025\,\text{cm}$ \\
-    & $N_i =  N_j = N_k = 60$ \\
-    \hline
+    B.C. & $f_b^{+}(x=0,T=0.5) = \frac{ac}{4}T^4$\\ & $\ve{F} =
+    0\quad \partial\Gamma \in \{x^{+}, y^{-}, y^{+}, z^{-}, z^{+}\}$
+    \\  material properties & $\rho =
+    1.5\,\text{g}\cdot\text{cm}^{-3}$ (duct) \\ & $\rho =
+    8.0\,\text{g}\cdot\text{cm}^{-3}$ (wall) \\ & $\rho =
+    4.0\,\text{g}\cdot\text{cm}^{-3}$ (foil) \\ & $\Cv =
+    0.1\,\text{Jerks}\cdot\text{g}^{-1}\cdot\text{keV}^{-1}$\\ &
+    $T(t=0) = \sn{1}{-4}\,\text{keV}$\\  opacity & $\sigma =
+    T^{-3}\,\text{cm}^2\cdot\text{g}^{-1}$ (duct) \\ & $\sigma =
+    1000T^{-1}\,\text{cm}^2\cdot\text{g}^{-1}$ (wall) \\ & $\sigma =
+    5T^{-2}\,\text{cm}^2\cdot\text{g}^{-1}$ (foil) \\  mesh & $\Di =
+    \Dj = \Dk = 0.025\,\text{cm}$ \\ & $N_i = N_j = N_k = 60$
+    \\ \hline
   \end{tabular}
@@
 \begin{figure}[h!]
-  \centerline{
-    \includegraphics[width=5in,clip]{mesh.pdf}}
+  \centerline{ \includegraphics[width=5in,clip]{mesh.pdf}}
   \caption{Multi-material problem mesh.  The duct region is blue.  The
@@
 \begin{figure}[h!]
-  \centerline{
-    \includegraphics[width=5in,clip]{T_multi_mat.pdf}}
-  \caption{
-    Two-dimensional plot of the temperature at 100~sh on a cut-plane
-    positioned at the mid-point of the $z$-axis.  The positions of the
-    edit points where the time-evolution of the temperature is plotted
-    in Fig.~\ref{fig:edit-points} are indicated by the black disks.}
+  \centerline{ \includegraphics[width=5in,clip]{T_multi_mat.pdf}}
+  \caption{ Two-dimensional plot of the temperature at 100~sh on a
+    cut-plane positioned at the mid-point of the $z$-axis.  The
+    positions of the edit points where the time-evolution of the
+    temperature is plotted in Fig.~\ref{fig:edit-points} are indicated
+    by the black disks.}
   \label{fig:multi-mat_T}
@@
     \begin{tabular}{cc}
-      \subfigure[Edit points]{\includegraphics[width=2in,clip]{points.pdf}} &
+      \subfigure[Edit
+        points]{\includegraphics[width=2in,clip]{points.pdf}} &
       \subfigure[Temperature]{\includegraphics[width=4in,clip]{multi_mat.pdf}}
@@
     \begin{tabular}{lrr}\hline\hline
-      \multicolumn{1}{c}{Method} & 
-      \multicolumn{1}{c}{Max Iterations} & 
-      \multicolumn{1}{c}{Relative CPU Time}\\\hline\hline
-      %%
-      PCG & 18 & 1.03 \\
-      %%
-      PFIX & 38 & 1.43 \\
-      %%
-      MCSA & 20 & 1.0 \\
-      \hline\hline
+      \multicolumn{1}{c}{Method} & \multicolumn{1}{c}{Max Iterations}
+      & \multicolumn{1}{c}{Relative CPU Time}\\\hline\hline  PCG &
+      18 & 1.03 \\  PFIX & 38 & 1.43 \\  MCSA & 20 & 1.0
+      \\ \hline\hline
     \end{tabular}
   \end{center}
-  \caption{
-    Comparison of solution methods for the multi-material problem. All
-    methods were run with a stopping criterion of $\sn{1}{-8}$.  The
-    MCSA method used $N_p=1000$ and $W_c = \sn{1}{-3}$.  The problem
-    was run to an elapsed time of 10~sh.  The MCSA method required 3.4
-    hours on a 3.6~GHz Xeon.}
+  \caption{ Comparison of solution methods for the multi-material
+    problem. All methods were run with a stopping criterion of
+    $\sn{1}{-8}$.  The MCSA method used $N_p=1000$ and $W_c =
+    \sn{1}{-3}$.  The problem was run to an elapsed time of 10~sh.
+    The MCSA method required 3.4 hours on a 3.6~GHz Xeon.}
   \label{tab:multimat_comparison}
@@
 We have presented a new Monte Carlo solution method for solving the
-discrete, time-dependent radiation diffusion equation in 3D.  The MCSA
-method has been shown to match results using standard solution
-techniques to arbitrary precision.  Also, the new method has been to
-shown to be faster than preconditioned CG and Richardson iteration for
-the model problems presented here.
+discrete, time-dependent radiation diffusion equation in three
+dimensions.  The MCSA method has been shown to match results using
+standard solution techniques to arbitrary precision. Also, the new
+method has been to shown to be faster than preconditioned CG and
+Richardson iteration for the model problems presented here.
 
@@
 
-\bibliographystyle{elsarticle-num}
-\bibliography{references}
+\bibliographystyle{elsarticle-num} \bibliography{references}
 

### 2012-08-10 doc/jcomp2012/jcomp2012.tex
jcomp updates

@@
 
-    We present a synthetic-acceleration based Monte Carlo method for
-    solving the 1T thermal radiation diffusion equation in three
-    dimensions. The algorithm performance is compared against
-    traditional solution schemes using a Marshak benchmark problem and
-    a more sophisticated multiple material problem. Our results show
-    that not only can our Monte Carlo method be an effective
-    solver for sparse matrix systems, but also that it is competitive
+    We present a novel synthetic-acceleration based Monte Carlo method
+    for solving the equilibrium thermal radiation diffusion equation
+    in three dimensions. The algorithm performance is compared against
+    traditional solution techniques using a Marshak benchmark problem
+    and a more complex multiple material problem. Our results show
+    that not only can our Monte Carlo method be an effective solver
+    for sparse matrix systems, but also that it performs competitively
     with deterministic methods including preconditioned Conjugate
@@
 Halton's method was not widely known in computational physics.  While
-extending our method to three dimensions, we have discovered that the Sequential
-Monte Carlo method is actually a variant of iterative refinement cast
-as a residual method.  Using Monte Carlo to accelerate a fixed-point
-(Richardson) iteration, our new method surpasses our previous 1D
-method and allows for extension to more general solution techniques.
+extending our method to three dimensions, we have discovered that the
+Sequential Monte Carlo method is actually a variant of iterative
+refinement cast as a residual method.  Using Monte Carlo to accelerate
+a fixed-point (Richardson) iteration, our new method surpasses our
+previous one dimensional method and allows for extension to more
+general solution techniques.
 
 The purpose of this study is to demonstrate an efficient Monte Carlo
-solution method for 3D, time-dependent, discrete systems. In this
-work, we extend our previous work in the following manner:
+solution method for three dimensional (3D), time-dependent, discrete
+systems. In this work, we extend our previous work in the following
+manner:
 
@@
 \item We apply this method to the 1T thermal radiation diffusion
-  equation in 3D.
+  equation in three dimensions.
+\item We compare numerical and performance results of the Monte Carlo
+  method against traditional deterministic solution methods.
 \end{itemize}
@@
 Monte Carlo method.  This algorithm demonstrated dramatically improved
-convergence over regular Monte Carlo.  Nonetheless, this method has
-not gained widespread use in computational physics.  In the next
+convergence over regular Monte Carlo.  Nonetheless, Halton's method
+has not gained widespread use in computational physics.  In the next
 sections, we will briefly describe the Monte Carlo methods that can be
@@
 \end{equation}
-When the spectral radius of the iteration matrix ($\vH$) is less than
-1, we can expand $\vA^{-1}$ using the \latin{Neumann Series},
+When the spectral radius ($\rho$) of the iteration matrix ($\vH$) is
+less than 1, we can expand $\vA^{-1}$ using the \latin{Neumann
+  Series},
 \begin{equation}
@@
 requires a terminating condition.  In all of the work that follows we
-utilize a relative weight cutoff.  The weight cutoff is defined as a
-fraction of the starting weight such that
+utilize a relative weight cutoff.  The relative weight cutoff is
+defined as a fraction of the starting weight such that
 \begin{equation}
@@
 in each iteration is updated using a single fixed-point
-iteration\footnote{Although, a subspace method could be used instead of
-  a stationary method here.}.  The residual is calculated and is used
+iteration\footnote{Although, a subspace method could be used instead
+  of a stationary method here.}.  The residual is calculated and used
 as a source to estimate the error, $\delta\vx^{l+1/2}$, by solving
@@
 
-The MCSA method defined in Eq.(~\ref{eq:MCSA-residual_solve}) uses the
-adjoint method to estimate the error instead of the direct method
+The MCSA method defined in Eq.~(\ref{eq:MCSA-iteration}) uses the
+adjoint method to estimate the error in
+Eq.~(\ref{eq:MCSA-residual_solve}) instead of the direct method
 outlined in \S~\ref{sec:direct-method}. To demonstrate the
@@
 
+Using the mathematical preliminaries developed in the previous
+section, we can now generate a discrete form of the radiation
+diffusion equation. We will show that this discrete form is amenable
+to interpretation using Monte Carlo concepts and can be written in a
+way that a random walk sequence can be generated.
+
 \subsection{Equilibrium Diffusion Model}
@@
   zero, $\diag(\vI - \ve{M}^{-1}\ve{D}) = 0$.  Thus, there will be no
-  computations wasted in transitions $i\rightarrow i$ during the
-  random walk process.
+  computations wasted for in-state transitions $i\rightarrow i$ during
+  the random walk process.
 \end{itemize}
@@
 method.  As discussed in \S~\ref{sec:iter-refin-monte}, the MCSA
-method requires two solution s.pdf:
+method requires two solutions:
 
@@
 relative weight cutoff that is defined in
-Eq.~(\ref{eq:weight_cutoff}).  We now examine each of these s.pdf in
+Eq.~(\ref{eq:weight_cutoff}).  We now examine each of these in
 more detail.

### 2012-08-10 doc/jcomp2012/jcomp2012.tex
jcomp updates

@@
 
-and a fourth-order nine point stencil both assuming a grid size of
-$\Delta$ in both the $i$ and $j$ directions
+and a fourth-order nine point stencil
 
@@
     u_{i,j-1} + 4 u_{i,j+1} + u_{i-1,j-1}\\ + u_{i-1,j+1} +
-    u_{i+1,j-1} + u_{i+1,j+1} - 20 u_{i,j}]\:.
+    u_{i+1,j-1} + u_{i+1,j+1} - 20 u_{i,j}]\:,
   \label{eq:nine_point_stencil}
@@
 
-For a single time step solution, we then have the following sparse
-linear system to be solved with the MCSA method:
+both assuming a grid size of $\Delta$ in both the $i$ and $j$
+directions. For a single time step solution, we then have the
+following sparse linear system to be solved with the MCSA method:
 
@@
 nine point stencils were used with both the direct and adjoint
-solvers. For each case, 50 random walk permuations were simulated per
-iteration with a weight cutoff of $1.0E-4$ for the stochastic linear
-solver and a convergence tolerance of $1.0E-8$ for the MCSA iterative
-solver. All computations presented in this section were completed on a
-2.66 GHz Intel Core i7 machine with 8 GB 1067 MHz DDR3 memory
-. Fig.~\ref{fig:poisson_cpu_time} gives the CPU time needed for each
-case to converge in seconds and Fig.~\ref{fig:poisson_iterations}
+solvers. For each case, 50 random walk permutations were computed per
+MCSA iteration with a weight cutoff of $1.0E-4$ for the stochastic
+linear solver and a convergence tolerance of $1.0E-8$ for the MCSA
+iterative solver. All computations presented in this section were
+completed on a 2.66 GHz Intel Core i7 machine with 8 GB 1067 MHz DDR3
+memory. Fig.~\ref{fig:poisson_cpu_time} gives the CPU time needed for
+each case to converge in seconds and Fig.~\ref{fig:poisson_iterations}
 gives the number of iterations needed for each case to converge
@@
   \caption{\sl Iterations to converge vs. Problem Size ($N$ for an $N
-    \times N$ square mesh. Both the adjoint and direct solvers are
+    \times N$ square mesh). Both the adjoint and direct solvers are
     used with the five point and nine point stencils. Both methods
@@
 both solvers and a system of size $N \times N$, the direct solver will
-compute $N \times N$ random walks per permuation while the adjoint
+compute $N \times N$ random walks per permutation while the adjoint
 solver will only compute one. This is necessary in the direct method
@@
 Fig.~\ref{fig:poisson_iterations} that the number of iterations
-requried to converge is approximately the same and therefore the extra
+required to converge is approximately the same and therefore the extra
 computations performed by the direct method do not improve the error
@@
 
-As an additional comparision, the convergence behavior of MCSA can be
+As an additional comparison, the convergence behavior of MCSA can be
 analyzed using both solvers to detect any performance benefits. To
@@
     number for a problem of fixed size. Both the adjoint and direct
-    solvers are used with the five point and nine point stencils. Both
-    methods give similar iteration behavior.}
+    solvers are used with the five point and nine point stencils.}
   \label{fig:poisson_convergence}
@@
 
-We chose to use the angle-energy integrated radiation intensity,
+We choose to use the angle-energy integrated radiation intensity,
 $\phi\equiv\phi(\ve{r},t)$
 $[\text{Jerks}\cdot\text{cm}^{-2}\cdot\text{sh}^{-1}]$, as the primary
-state variable.  The intensity is defined,
+state variable.  The intensity is defined as:
 
@@
 where $D$ is the diffusion coefficient, $D\equiv
-D(\ve{r},T(\ve{r},t))\ [\text{cm}]$, and is defined
+D(\ve{r},T(\ve{r},t))\ [\text{cm}]$, and is defined as
 

### 2012-08-10 doc/jcomp2012/jcomp2012.tex
final edits for jcomp draft

@@
 equilibrium (1T) thermal radiation diffusion equation.  At that time,
-Halton's method was not widely known in computational physics.  While
-extending our method to three dimensions, we have discovered that the
-Sequential Monte Carlo method is actually a variant of iterative
-refinement cast as a residual method.  Using Monte Carlo to accelerate
-a fixed-point (Richardson) iteration, our new method surpasses our
+Halton's method was not widely known in the general computational
+physics community.  While extending our method to three dimensions, we
+have realized that the Sequential Monte Carlo method is actually a
+variant of iterative refinement cast as a residual method. We have
+developed a new method in which Monte Carlo is used to accelerate a
+fixed-point (Richardson) iteration. This algorithm surpasses our
 previous one dimensional method and allows for extension to more
@@
 convergence over regular Monte Carlo.  Nonetheless, Halton's method
-has not gained widespread use in computational physics.  In the next
-sections, we will briefly describe the Monte Carlo methods that can be
-used to solve linear systems and expand them to present our new
-synthetic acceleration method.
+has not gained widespread use in computational physics
+applications. In the next sections, we will briefly describe the Monte
+Carlo methods that can be used to solve linear systems and expand them
+to present our new synthetic acceleration method.
 
@@
 solvers. For each case, 50 random walk permutations were computed per
-MCSA iteration with a weight cutoff of $1.0E-4$ for the stochastic
-linear solver and a convergence tolerance of $1.0E-8$ for the MCSA
+MCSA iteration with a weight cutoff of \sn{1}{-4} for the stochastic
+linear solver and a convergence tolerance of \sn{1}{-8} for the MCSA
 iterative solver. All computations presented in this section were
@@
 where $\Cv\equiv\Cv(\ve{r},T(\ve{r},t))$
-$[\text{Jerks}\cdot\text{g}^{-1}\cdot\text{keV}^{-1}]$ is the specific
+$[\text{GJ}\cdot\text{g}^{-1}\cdot\text{keV}^{-1}]$ is the specific
 heat capacity of the material, $\rho\equiv\rho(\ve{r})$
 $[\text{g}\cdot\text{cm}^{-3}]$ is the density of the material,
-$a=0.01372$ $[\text{Jerks}\cdot\text{cm}^{-3}\cdot\text{keV}^{-4}]$ is
-the radiation constant, $c=299.79$ $[\text{cm}\cdot\text{sh}^{-1}]$ is
+$a=0.01372$ $[\text{GJ}\cdot\text{cm}^{-3}\cdot\text{keV}^{-4}]$ is
+the radiation constant, $c=29.979$ $[\text{cm}\cdot\text{ns}^{-1}]$ is
 the vacuum light speed, and $T\equiv T(\ve{r},t)$ $[\text{keV}]$ is
@@
 $Q\equiv Q(\ve{r},t)$
-$[\text{Jerks}\cdot\text{cm}^{-3}\cdot\text{sh}^{-1}]$, is assumed to
+$[\text{GJ}\cdot\text{cm}^{-3}\cdot\text{ns}^{-1}]$, is assumed to
 be independent of $T$.
@@
 $\phi\equiv\phi(\ve{r},t)$
-$[\text{Jerks}\cdot\text{cm}^{-2}\cdot\text{sh}^{-1}]$, as the primary
+$[\text{GJ}\cdot\text{cm}^{-2}\cdot\text{ns}^{-1}]$, as the primary
 state variable.  The intensity is defined as:
@@
 systems, and this type of scaling should be performed regardless of
-any additional preconditioner choices.
-
-A side effect of this preconditioner choice is that the preconditioned
-system, $\ve{M}^{-1}\ve{D}$, is no longer SPD.  This situation can be
-rectified by doing right-left preconditioning and defining
-
-\begin{equation}
-  \ve{M} = \ve{L}\ve{L}^{T}\:,
-\end{equation}
-
-where $\ve{L} = \ve{L}^{T} = \sqrt{\ve{M}}$.  Applying right-left
-preconditioning gives
-
-\begin{equation}
-  \begin{gathered}
-    \ve{M}^{-1}\ve{D}\ve{M}^{-1}\ve{M}\bphi =
-    \ve{M}^{-1}\ve{q}\:,\\ (\ve{LL}^T)^{-1}\ve{D}(\ve{LL}^T)^{-1}\ve{LL}^T\bphi
-    = (\ve{LL}^T)^{-1}\ve{q}\:,
-    \\ \ve{L}^{-T}\ve{L}^{-1}\ve{D}\ve{L}^{-T}\ve{L}^T\phi =
-    \ve{L}^{-T}\ve{L}^{-1}\ve{q}\:,\\ (\ve{L}^{-1}\ve{D}\ve{L}^{-T})\ve{L}^T\bphi
-    = \ve{L}^{-1}\ve{q}\:,\\ \hat{\ve{D}}\hat{\bphi} = \hat{\ve{q}}\:.
-  \end{gathered}
-  \label{eq:right-left-preconditioning}
-\end{equation}
-
-This type of preconditioning is also referred to as row-column
-scaling.  The resulting preconditioned matrix, $\hat{\ve{D}}$, is now
-SPD, and we can use Conjugate Gradient (CG) solvers.  Therefore, we
-will use this preconditioning scheme for deterministic solutions of
-Eq.~(\ref{eq:diffusion-matrix}).  However, given that the convergence
-properties of $\hat{\ve{D}}$ are similar to $\ve{M}^{-1}\ve{D}$ and
-the Monte Carlo solvers do not require SPD matrices, we will be
-content with solving Eq.~(\ref{eq:preconditioned-diffusion}) using
-MCSA.
-
-The preconditioned system we will solve is defined in
-Eq.~(\ref{eq:preconditioned-diffusion}).  Applying $\ve{M}^{-1}$ to
-Eq.~(\ref{eq:discrete-diffusion}) yields
+any additional preconditioner choices. We note that the resulting
+system does not have to be SPD for the MCSA algorithm, thus any
+suitable choice of $\ve{M}$ will suffice.  The preconditioned system
+we will solve is defined in Eq.~(\ref{eq:preconditioned-diffusion}).
+Applying $\ve{M}^{-1}$ to Eq.~(\ref{eq:discrete-diffusion}) yields
 
@@
 preconditioned fixed-point (Richardson) iteration (PFIX).  The PCG
-method uses right-left preconditioning a shown in
-Eq.~(\ref{eq:right-left-preconditioning}).  The PFIX scheme is used to
-solve Eq.~(\ref{eq:preconditioned-diffusion}), which is identical to
-the system being solved with MCSA.  We examine each method on two 3D
+method uses right-left preconditioning with
+$\ve{M}=\ve{L}\ve{L}^{-1}$.  The PFIX scheme is used to solve
+Eq.~(\ref{eq:preconditioned-diffusion}), which is identical to the
+system being solved with MCSA.  We examine each method on two 3D
 problems: a Marshak wave problem and a multi-material duct problem.
@@
 incoming 1~keV blackbody boundary flux at $x=0$ and propagates in the
-$x$-dimension in time.  In other words, this a 3D representation of a
-1D problem.  The full set of problem conditions is:
+$x$-dimension in time as outlined in Ref.~\cite{larsen_1980}.  In
+other words, this a 3D representation of a 1D problem.  The full set
+of problem conditions is:
 
@@
     3.0\,\text{g}\cdot\text{cm}^{-3}$ \\ & $\Cv =
-    0.1\,\text{Jerks}\cdot\text{g}^{-1}\cdot\text{keV}^{-1}$\\ &
+    0.1\,\text{GJ}\cdot\text{g}^{-1}\cdot\text{keV}^{-1}$\\ &
     $T(t=0) = \sn{1}{-4}\,\text{keV}$\\  opacity & $\sigma =
@@
 
-The problem is run to $10$~shakes.
+The problem is run to $100$~nanoseconds.
 
@@
   \centerline{ \includegraphics[width=5in,clip]{marshak_10.pdf}}
-  \caption{ Solution of the Marshak wave problem at $t=10$~sh.  The
+  \caption{ Solution of the Marshak wave problem at $t=100$~ns.  The
     analytic solution is compared to the MCSA solution.  Note that the
@@
     4.0\,\text{g}\cdot\text{cm}^{-3}$ (foil) \\ & $\Cv =
-    0.1\,\text{Jerks}\cdot\text{g}^{-1}\cdot\text{keV}^{-1}$\\ &
+    0.1\,\text{GJ}\cdot\text{g}^{-1}\cdot\text{keV}^{-1}$\\ &
     $T(t=0) = \sn{1}{-4}\,\text{keV}$\\  opacity & $\sigma =
@@
 
-A 2D plot of the temperature at 100~sh elapsed time is illustrated in
+A 2D plot of the temperature at 1000~ns elapsed time is illustrated in
 Fig.~\ref{fig:multi-mat_T}.  The time-evolution of the temperature is
@@
   \centerline{ \includegraphics[width=5in,clip]{T_multi_mat.pdf}}
-  \caption{ Two-dimensional plot of the temperature at 100~sh on a
+  \caption{ Two-dimensional plot of the temperature at 1000~ns on a
     cut-plane positioned at the mid-point of the $z$-axis.  The
@@
     $\sn{1}{-8}$.  The MCSA method used $N_p=1000$ and $W_c =
-    \sn{1}{-3}$.  The problem was run to an elapsed time of 10~sh.
+    \sn{1}{-3}$.  The problem was run to an elapsed time of 100~ns.
     The MCSA method required 3.4 hours on a 3.6~GHz Xeon.}

### 2012-08-21 solvers/src/IterativeSolver.hpp
Adding Will Benton dissertation template to the prelim document

@@ -23,7 +23,7 @@ namespace Chimera
  * Scalar must implement Teuchos::ScalarTraits, MV must implement
  * Belos::MultiVecTraits, and OP must implement Belos::OperatorTraits.
  */
-template<typename Scalar typename MV, typename OP>
+template<typename Scalar, typename MV, typename OP>
 class IterativeSolver : public Teuchos::Describable
 {
   public:

### 2012-08-21 doc/prelim/document/chapters/linear_problem/linear_problem.tex
adding makefile and linear and nonlinear problem chapters to prelim

\chapter{The Linear Problem}\label{ch:linear_problem}

### 2012-08-21 doc/prelim/document/chapters/nonlinear_problem/nonlinear_problem.tex
adding makefile and linear and nonlinear problem chapters to prelim

\chapter{The Nonlinear Problem}\label{ch:nonlinear_problem}

### 2012-08-21 doc/prelim/document/frontmatter/abstract.tex
frontmatter updates

--- 2012-08-03 doc/thesis_proposal/thesis_proposal.tex
@@
-%        File: thesis_proposal.tex
-%
-\documentclass[letterpaper,12pt]{article}
-\usepackage[top=1.0in,bottom=1.0in,left=1.25in,right=1.25in]{geometry}
-\usepackage{verbatim}
-\usepackage{amssymb}
-\usepackage{graphicx}
-\usepackage{longtable}
-\usepackage{amsfonts}
-\usepackage{amsmath}
-\usepackage[usenames]{color}
-\usepackage[
-naturalnames = true, 
-colorlinks = true, 
-linkcolor = black,
-anchorcolor = black,
-citecolor = black,
-menucolor = black,
-urlcolor = blue
-]{hyperref}
+\svnidlong{$LastChangedBy$}{$LastChangedRevision$}{$LastChangedDate$}{$HeadURL: http://freevariable.com/dissertation/branches/diss-template/frontmatter/abstract.tex $}
+\vcinfo{}
 
-%%---------------------------------------------------------------------------%%
-\author{Stuart R. Slattery
-\\ \href{mailto:sslattery@wisc.edu}{\texttt{sslattery@wisc.edu}}
-}
-
-\date{\today}
-\title{Thesis Proposal:\\
-  A Massively Parallel Stochastic Method for Nonlinear Problems}
-\begin{document}
-\maketitle
-
-%%---------------------------------------------------------------------------%%
-\section{Introduction}
 In many fields of engineering and physics, nonlinear problems are a
@@
 parallelism.
-
-%%---------------------------------------------------------------------------%%
-\section{Deliverables}
-
-This work will deliver the following items:
-
-\begin{itemize}
-\item Develop a massively parallel strategy for the MCSA method for
-  sparse linear systems.
-\item Implement a massively parallel strategy for the MCSA method for
-  sparse linear systems.
-\item Fully assess the performance of parallel MCSA on leadership class
-  computing systems in order to determine its feasibility using
-  benchmark problems.
-\item Compare the performance of MCSA to other iterative methods for
-  sparse linear systems including GMRES.
-\item Develop the FANM method for nonlinear systems.
-\item Implement the FANM method for nonlinear systems.
-\item Fully assess the performance of the FANM method on leadership
-  class computing systems using model problems.
-\item Compare the performance of MCSA to other nonlinear methods
-  including JFNK.
-\end{itemize}
-
-This work will not address the resiliency of these algorithms.
-
-%%---------------------------------------------------------------------------%%
-\section{Dissertation Outline}
-
-Based on the above work, the dissertation can be outlined in five
-distinct parts:
-
-\begin{enumerate}
-\item Iterative methods for sparse linear systems (fixed-point,
-  subspace, and stochastic methods).
-\item MCSA and Multiple Set Overlapping Domain Decomposition
-  (development and implementation).
-\item Iterative methods for nonlinear systems (Newton methods).
-\item FANM (developement and implementation).
-\item Application to model problems (conduction, Navier-Stokes,
-  conjugate heat transfer).
-\end{enumerate}
-
-%%---------------------------------------------------------------------------%%
-\section{Committee}
-
-\begin{itemize}
-\item Paul Wilson (advisor): Associate Professor, Department of
-  Engineering Physics.
-\item Greg Moses: Professor, Department of Engineering Physics.
-\item Carl Sovinec: Professor, Department of Engineering Physics.
-\item Chris Rutland: Professor, Deptartment of Mechanical Engineering.
-\item Tom Evans: Distinguished R\&D Staff, Reactor and Nuclear Systems
-  Division, Oak Ridge National Laboratory.
-
-\end{itemize}
-
-%%---------------------------------------------------------------------------%%
-\section{Timeline}
-
-%%---------------------------------------------------------------------------%%
-\pagebreak
-\bibliographystyle{ieeetr}
-\bibliography{references}
-\end{document}

### 2012-08-21 doc/prelim/document/chapters/fanm/fanm.tex
adding parallel mcsa and fanm chapters

\chapter{The FANM Method for Nonlinear Problems}
\label{ch:fanm}

### 2012-08-21 doc/prelim/document/chapters/parallel_mcsa/parallel_mcsa.tex
adding parallel mcsa and fanm chapters

\chapter{Parallelization of the MCSA Algorithm}
\label{ch:parallel_mcsa}

### 2012-08-21 doc/prelim/document/chapters/chimera_codebase/chimera_codebase.tex
updating references

\chapter{The Chimera Code Base}
\label{ch:chimera_codebase}

### 2012-08-21 doc/prelim/document/frontmatter/abstract.tex
updating references

@@
 Monte Carlo methods for the solution of transport problems
-\cite{Lewis_1993}. The partial differential equation (PDE) community
+\cite{lewis_1993}. The partial differential equation (PDE) community
 has focused on various deterministic methods for solutions to linear
-problems \cite{Saad_2003}. In between these two areas are a not widely
+problems \cite{saad_2003}. In between these two areas are a not widely
 known small group of stochastic methods for solving sparse linear
-systems \cite{Hammersley_1964, Halton_1962, Halton_1994}. In recent
+systems \cite{hammersley_1964, halton_1962, halton_1994}. In recent
 years, these methods have been further developed for transport
 problems in the form of Monte Carlo Synthetic-Acceleration (MCSA)
-\cite{Evans_2003, Evans_2009} but have yet to be applied to more
+\cite{evans_2003, evans_2009} but have yet to be applied to more
 general sparse linear systems. Compared to other methods in these
@@
 and (3) is amenable to parallelization using modern methods developed by
-the transport community \cite{Wagner_2011}. The developement of MCSA
+the transport community \cite{wagner_2011}. The developement of MCSA
 as a general linear solver and the development of a parallel MCSA
@@
 problem and using traditionaly iterative or direct methods
-\cite{Tannehill_1997}. In the mathematics community, various Newton
-methods have been popular \cite{Kelley_1995}. Recently, Jacobian-Free
-Newton-Krylov (JFNK) schemes \cite{Knoll_2004} have been utilized in
+\cite{tannehill_1997}. In the mathematics community, various Newton
+methods have been popular \cite{kelley_1995}. Recently, Jacobian-Free
+Newton-Krylov (JFNK) schemes \cite{knoll_2004} have been utilized in
 multiple physics architectures and advanced single physics codes
-\cite{Gaston_2009}. The benefits of JFNK schemes are that the Jacobian
+\cite{gaston_2009}. The benefits of JFNK schemes are that the Jacobian
 is never formed, simplifying the implementation, and a Krylov solver
@@
 generation using modern Forward Automated Differencing (FAD) methods
-\cite{Bartlett_2006}, and MCSA as the inner linear solver. This method
+\cite{bartlett_2006}, and MCSA as the inner linear solver. This method
 has several attractive properties. First, the first-order
@@
 efficient computationally than evaluating differenced derivatives
-\cite{Bartlett_2006}. Third, unlike GMRES, MCSA does not build a
+\cite{bartlett_2006}. Third, unlike GMRES, MCSA does not build a
 subspace during iterations. Although the Jacobian must be explicitly

### 2012-08-21 doc/prelim/document/chapters/introduction/introduction.tex
adding intro and research proposal chapter

--- 2012-08-21 doc/prelim/document/frontmatter/abstract.tex
@@
-\svnidlong{$LastChangedBy$}{$LastChangedRevision$}{$LastChangedDate$}{$HeadURL: http://freevariable.com/dissertation/branches/diss-template/frontmatter/abstract.tex $}
-\vcinfo{}
+\chapter{Introduction}
+\label{ch:introduction}
 
-In many fields of engineering and physics, nonlinear problems are a
-primary focus of study. Recent focus on multiple physics systems (and
-in particular nuclear systems) adds a new level of complexity to
-common nonlinear systems as solution strategies change when they are
-coupled to other problems. Furthermore, a desire for predictive
-simulations to enhance the safety and performance of engineered
-systems creates a need for extremely high fidelity computations to be
-performed for these coupled systems as a means to capture effects not
-modeled by coarser methods. In order to achieve this high fidelity,
-state-of-the-art computing facilities must be leveraged in a way that
-is both efficient and considerate of hardware-related issues. As
-scientific computing moves towards exascale facilities with machines
-of $O(1,000,000)$ cores already coming online, new algorithms to solve
-these complex problems must be developed to leverage this new
-hardware. Issues such as resiliency to node failure and scaling to
-large numbers of cores will be pertinent to robust algorithms aimed at
-this new hardware. Considering these issues, this thesis proposes a
-novel stochastic method to advance solution techniques for nonlinear
-problems with a focus on coupled physics systems.
+In many fields of engineering and physics, linear and nonlinear
+problems are a primary focus of study. Recent focus on multiple
+physics systems (and in particular nuclear systems) adds a new level
+of complexity to common nonlinear systems as solution strategies
+change when they are coupled to other problems. Furthermore, a desire
+for predictive simulations to enhance the safety and performance of
+engineered systems creates a need for extremely high fidelity
+computations to be performed for these coupled systems as a means to
+capture effects not modeled by coarser methods.
+
+In order to achieve this high fidelity, state-of-the-art computing
+facilities must be leveraged in a way that is both efficient and
+considerate of hardware-related issues. As scientific computing moves
+towards exascale facilities with machines of $O(1,000,000)$ cores
+already coming online, new algorithms to solve these complex problems
+must be developed to leverage this new hardware. Issues such as
+resiliency to node failure and scaling to large numbers of cores will
+be pertinent to robust algorithms aimed at this new
+hardware. Considering these issues, this thesis proposes a the
+development of a massively parallel stochastic method for linear
+problems and a novel stochastic method to advance solution techniques
+for nonlinear problems with a focus on coupled physics systems.
+
+This preliminary report oultines the conventional methods used in
+practice for solving linear and nonlinear problems to provide a
+mathematical basis upon which to build new algorithms that aim to
+solve some of the aforementioned issues. These new algorithms will be
+outlined in full with links made to past work and their potential for
+offering improvements to the computational physics community. Progress
+to date will be reported including the development of a new physics
+framework that will serve as a test bed for these new methods. A
+research plan will then be provided that continues the development and
+implementation of these new methods and tests them in practical
+settings.
+
+\section{Motivation}
+\label{sec:motivation}
 

### 2012-08-21 doc/prelim/document/chapters/research_proposal/research_proposal.tex
adding intro and research proposal chapter

\chapter{Research Proposal}
\label{ch:research_proposal}

\section{Model Problems}
\label{sec:model_problems}

\section{Method Comparisons}
\label{sec:method_comparisons}

\section{Performance Metrics}
\label{sec:performance_metrics}

### 2012-08-22 doc/prelim/document/chapters/chimera_codebase/chimera_codebase.tex
Finished initial prelim outline

@@
 \label{ch:chimera_codebase}
+
+The Chimera code base will be the primary research tool for answering
+the questions posed in the previous chapters of the document. We may
+not need a chapter for this, but it will likely be a good idea
+somewhere in this document to outline the design of the software and
+the instrumentation that it will provide.

### 2012-08-22 doc/prelim/document/chapters/fanm/fanm.tex
adding intro and research proposal chapter
Finished initial prelim outline

@@
 \label{ch:fanm}
+
+In this chapter we will devise the FANM method. Based on the
+background in the previous section, we first need to motivate the
+development of this work by pointing out some of the issues with
+conventional methods. This includes JFNK coarseness and prohibitively
+large Krylov subspace. We can than point out some of the attractive
+qualities of the FANM method. Much of this will rely on advanced
+concepts including automatic differentiation and on-the-fly residual
+and Jacobian generation from nonlinear function evaluations. In
+addition, we will want to show that the Newton method will converge
+using MCSA as the linear solver. I suspect I will have some
+preliminary results for a serial implementation here as I already have
+a serial MCSA implemented. Finally, we will want to design the model
+problems that we will test FANM with. These should be problems that
+have already been worked up in the literature with benchmark-type
+solutions.
+
+\section{Issues with Current Nonlinear Methods}
+\label{sec:nonlinear_issues}
+
+\section{Automatic Differentiation}
+\label{sec:automatic_differentiation}
+
+\section{Residual and Jacobian Generation}
+\label{sec:fanm_generation}
+
+\section{Algorithm Outline}
+\label{sec:fanm_algorithm_outline}
+
+\section{Model Problems}
+\label{sec:fanm_model_problems}

### 2012-08-22 doc/prelim/document/chapters/linear_problem/linear_problem.tex
added outline to the linear and nonlinear problem chapters
adding intro and research proposal chapter
Finished initial prelim outline

@@
-\chapter{The Linear Problem}\label{ch:linear_problem}
+\chapter{The Linear Problem}
+\label{ch:linear_problem}
+
+The purpose of this chapter is to provide a full background on solving
+linear systems as it relates to this work. It may seem like there is a
+lot here, but we really do need it to fully define our new methods and
+compare both qualitatively and quantitatively to convential
+methods. For example, we need a solid explanation of stationary
+methods because MCSA is in fact an acceleration of a stationary method
+and therefore shares many properties with them. Subspace methods are
+the current class of methods most widely used for sparse systems and
+are at the core of the Newton-Krylov methods that we will be comparing
+the FANM method with. The most rigorous piece of this chapter should
+of course be the stochastic solver definitions, also providing me a
+place to include some of my work in this area. Finally,
+preconditioning is important for this work, MCSA requires it and
+Newton-Krylov methods almost always need some type of
+preconditioning. Therefore, we must also discuss these aspects.
+
+\section{Formulation of the Linear System}
+\label{sec:linear_system}
+
+\section{Iterative Methods for Solving Sparse Linear Systems}
+\label{sec:linear_methods}
+
+\subsection{Stationary Methods}
+\label{subsec:stationary_methods}
+
+\subsection{Subspace Methods}
+\label{subsec:subspace_methods}
+
+\subsubsection{Conjugate Gradient}
+\label{subsubsec:conjugate_gradient}
+
+\subsubsection{GMRES}
+\label{subsubsec:gmres}
+
+\subsection{Stochastic Methods}
+\label{subsec:stochastic_methods}
+
+\subsubsection{Direct Method}
+\label{subsubsec:direct_mc}
+
+\subsubsection{Adjoint Method}
+\label{subsubsec:adjoint_mc}
+
+\subsubsection{Sequential Monte Carlo}
+\label{subsusbsec:sequential_mc}
+
+\subsection{Monte Carlo Synthetic-Acceleration}
+\label{subsec:mcsa}
+
+\section{Preconditioning the Linear System}
+\label{sec:linear_preconditioning}

### 2012-08-22 doc/prelim/document/chapters/nonlinear_problem/nonlinear_problem.tex
added outline to the linear and nonlinear problem chapters
updating references
adding intro and research proposal chapter
Finished initial prelim outline

@@
-\chapter{The Nonlinear Problem}\label{ch:nonlinear_problem}
+\chapter{The Nonlinear Problem}
+\label{ch:nonlinear_problem}
+
+This work will result in a new and novel nonlinear solver, FANM. To
+movitate the development of FANM and thoroughly understand both its
+mathematical properties and its qualities as related to other
+nonlinear methods, we need a full workup of current iterative methods
+for solving nonlinear systems. We should acknowledge operator split
+and linearization methods for dealing with nonlinear systems
+(e.g. think pressure correction strategies for Navier-Stokes) but the
+focus here should be on fully-implicit nonlinear solves that are
+amenable to solving tightly-coupled multi-physics systems in a way that
+is fully consistent.
+
+The key here will be a full understanding of Newton methods (and they
+actually are quite simple compared to some of the linear solver
+methods like subspace methods). It is important to understand how the
+correction is computed from the Jacobian and how we may improve this
+with globalization methods. Although I may not use globalization
+methods in my work, they are important enough in practice that we must
+be sure that a FANM method is not prohibited from using these. Current
+methods, primarily Newton-Krylov methods, will of primary concern here
+as these will serve as the baseline for assessing the qualities of a
+FANM solver.
+
+Finally, as we will choose transient problems for our models, time
+integration in the context of nonlinear systems will be consider and
+we will survey applications to model problems in the field ( there is
+a vast amount of literature on fully-implicit nonlinear solves for the
+Navier-Stokes equations) that will give good benchmarks to test (such
+as natural convection and driven cavity problems).
+
+\section{Formulation of the Nonlinear System}
+\label{sec:nonlinear_system}
+
+\section{Iterative Methods for Solving Nonlinear Systems}
+\label{sec:nonlinear_methods}
+
+\subsection{General Newton Methods}
+\label{subsec:newton_methods}
+
+\subsection{Newton-Krylov Methods}
+\label{subsec:newton_krylov_methods}
+
+\subsection{Globalization Methods}
+\label{subsec:globalization_methods}
+
+\section{Preconditioning the Nonlinear System}
+\label{sec:nonlinear_preconditioning}
+
+\section{Time Integration}
+\label{sec:time_integration}
+
+\section{Applications to Model Problems}
+\label{sec:nonlinear_applications}

### 2012-08-22 doc/prelim/document/chapters/parallel_mcsa/parallel_mcsa.tex
adding intro and research proposal chapter
Finished initial prelim outline

@@
 \label{ch:parallel_mcsa}
+
+For the prelim, I suspect that I will not have a parallel MCSA
+algorithm up and running as I will be focusing on generating the code
+base needed to build the physics operators and tools for linear and
+nonlinear systems analysis. Therefore, here we should focus on
+motivating a parallel scheme for this work. We should start by first
+looking at current parallel methods for solving linear systems using
+the conventional techniques from the previous chapter. It must be
+clearly pointed out that for the new hardware coming online, these
+methods (or at least their current implementations) will be lacking in
+scalability and the potential for resiliency, and therefore a parallel
+MCSA is attractive, even if it ultimately does not show performance
+gains from a wall time perspective.
+
+Once we have done this, MSOD will need to be defined. The literature
+is actually quite lacking here even for reactor physics problems as
+this is really bleeding edge. Therefore, I should strive here to more
+rigorously define this methodology in terms of solving general linear
+systems. Furthermore, we should discuss the aspects of being able to
+vary the overlap such that we can generate a spectrum between complete
+domain decomposition and complete domain replication. Finally, the
+beginnings of the algorithm should be sketched out.
+
+\section{Current Parallel Methodologies for Solving Linear Systems}
+\label{sec:current_parallel_methods}
+
+\section{Multiple-Set Overlapping-Domain Decomposition}
+\label{sec:msod}
+
+\section{Algorithm Outline}
+\label{sec:parallel_mcsa_algorithm_outline}

### 2012-08-24 doc/prelim/document/chapters/introduction/introduction.tex
adding more references

@@
 Monte Carlo methods for the solution of transport problems
-\cite{lewis_1993}. The partial differential equation (PDE) community
+\citep{lewis_1993}. The partial differential equation (PDE) community
 has focused on various deterministic methods for solutions to linear
-problems \cite{saad_2003}. In between these two areas are a not widely
+problems \citep{saad_2003}. In between these two areas are a not widely
 known small group of stochastic methods for solving sparse linear
-systems \cite{hammersley_1964, halton_1962, halton_1994}. In recent
+systems \citep{hammersley_1964, halton_1962, halton_1994}. In recent
 years, these methods have been further developed for transport
 problems in the form of Monte Carlo Synthetic-Acceleration (MCSA)
-\cite{evans_2003, evans_2009} but have yet to be applied to more
+\citep{evans_2003, evans_2009} but have yet to be applied to more
 general sparse linear systems. Compared to other methods in these
@@
 and (3) is amenable to parallelization using modern methods developed by
-the transport community \cite{wagner_2011}. The developement of MCSA
+the transport community \citep{wagner_2011}. The developement of MCSA
 as a general linear solver and the development of a parallel MCSA
@@
 problem and using traditionaly iterative or direct methods
-\cite{tannehill_1997}. In the mathematics community, various Newton
-methods have been popular \cite{kelley_1995}. Recently, Jacobian-Free
-Newton-Krylov (JFNK) schemes \cite{knoll_2004} have been utilized in
+\citep{tannehill_1997}. In the mathematics community, various Newton
+methods have been popular \citep{kelley_1995}. Recently, Jacobian-Free
+Newton-Krylov (JFNK) schemes \citep{knoll_2004} have been utilized in
 multiple physics architectures and advanced single physics codes
-\cite{gaston_2009}. The benefits of JFNK schemes are that the Jacobian
+\citep{gaston_2009}. The benefits of JFNK schemes are that the Jacobian
 is never formed, simplifying the implementation, and a Krylov solver
@@
 generation using modern Forward Automated Differencing (FAD) methods
-\cite{bartlett_2006}, and MCSA as the inner linear solver. This method
+\citep{bartlett_2006}, and MCSA as the inner linear solver. This method
 has several attractive properties. First, the first-order
@@
 efficient computationally than evaluating differenced derivatives
-\cite{bartlett_2006}. Third, unlike GMRES, MCSA does not build a
+\citep{bartlett_2006}. Third, unlike GMRES, MCSA does not build a
 subspace during iterations. Although the Jacobian must be explicitly

### 2012-08-24 doc/me699/proposal/me699_proposal.tex
updates to me699 proposal

%        File: me699_proposal.tex
%
\documentclass[letterpaper,12pt]{article}
\usepackage[top=1.0in,bottom=1.0in,left=1.25in,right=1.25in]{geometry}
\usepackage{verbatim}
\usepackage{amsthm} 
\usepackage{tmadd,tmath}
\usepackage[mathcal]{euscript}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage[usenames]{color}
\usepackage[
naturalnames = true, 
colorlinks = true, 
linkcolor = black,
anchorcolor = black,
citecolor = black,
menucolor = black,
urlcolor = blue
]{hyperref}

%%---------------------------------------------------------------------------%%
\author{Stuart R. Slattery
\\ \href{mailto:sslattery@wisc.edu}{\texttt{sslattery@wisc.edu}}
}

\date{\today} \title{ME699 Proposal: Fully Implicit Solutions to the
  Steady and Incompressible Navier-Stokes Equations}
\begin{document}
\maketitle

%%---------------------------------------------------------------------------%%
\section{Proposal}
Fully implicit formulations of the Navier-Stokes equations are
becoming increasingly popular due to their robustness over
conventional segregated methods that arises from a fully consistent
problem formulation. Solutions with inexact Newton methods have
proliferated due to advancements in both hardware and numerical
methods for solving sparse linear sytems and preconditioning
\cite{shadid_1997,mchugh_1994,gropp_2001,knoll_2004}. To study the
details of how fully implicit solutions are implemented in practice,
and independent study is posed to develop a piece of software aimed at
generating fully implicit solutions to the steady and incompressible
Navier-Stokes equations.

\subsection{Model Problem}
We choose the steady incompressible Navier-Stokes equations as model
problems in residual form to solve for the state vector
$\{\ve{u},P,T\}$ where $\ve{u}$ is the fluid velocity, $P$ is the
fluid pressure, and $T$ is the fluid temperature. As formulated in
\cite{shadid_1997}, we consider vector forms of momentum transport:

\begin{equation}
  \rho \ve{u} \cdot \nabla \ve{u} - \nabla \cdot \ve{T} - \rho \ve{g}
  = \ve{0} \:,
  \label{eq:momentum_transport}
\end{equation}

continuity:

\begin{equation}
  \nabla \cdot \ve{u} = \ve{0} \:,
  \label{eq:continuity}
\end{equation}

and energy transport with the effects of viscous dissipation neglected
for low Mach number flow:

\begin{equation}
  \rho C_p \ve{u} \cdot \nabla T + \nabla \ve{q} = \ve{0} \:,
  \label{eq:energy_transport}
\end{equation}

where $\rho$ is the fluid density, $\ve{g}$ the gravity vector and
$C_p$ the specific heat of the fluid at constant pressure. To close
the system we define the stress tensor:

\begin{equation}
  \ve{T} = -P \ve{I} + \mu [ \nabla \ve{u} \nabla \ve{u}^{T} ] \:,
  \label{eq:stress_tensor}
\end{equation}

and the heat flux:

\begin{equation}
  \ve{q} = -\kappa \nabla T
  \label{eq:heat_flux}
\end{equation}

where $\ve{I}$ is the identity matrix, $\mu$ is the dynamic
viscosityof the fluid, and $\kappa$ is the thermal conductivity of the
fluid. These fluid properties will be assumed constant and boundary
conditions will be chosen as a combination of Neumann and Dirichlet
conditions depending on the problem chosen.

\subsection{Discretization}
Reference \cite{shadid_1997} uses a pressure stabilized
Petrov-Galerkin finite element formulation of these equations
\cite{hughes_1986}. Depending on the software infrastructure
developed, these may or may not be as easy to formulate as the finite
volume discretizations in \cite{mchugh_1994} due to the availabilty of
finite element discretization libraries, finite element assembly
engines, and unstructured mesh databases available in the Trilinos
scientific computing libraries \cite{trilinos_2005}. It will likely be
easiest to start with the finite volume approach.

\subsection{Solution Method}
We will choose an inexact Newton method for the nonlinear scheme solve
these equations \cite{dembo_1982}. Of particular interest will be
investigating how the forcing term affects convergence of the
nonlinear resdiual \cite{eisenstat_1996}. In addition, if convergence
issues are noted, globalization methods may be investigated to improve
the Newton correction \cite{pawlowski_2006}. The linear solver for the
Newton method will be GMRES as its convergence properties for these
types of systems are well documented \cite{saad_1986}. We will seek to
solve these equations using parallel methods provided by well
established framework libraries provided by the Trilinos suite of
scientific computing libraries \cite{trilinos_2005}.

\subsection{Benchmark Problems}
To verify the implementation, three widely reference two-dimensional
benchmark problems will be assessed using the resulting software. The
natural convection cavity problem \cite{davis_1983}, the driven cavity
problem \cite{ghia_1982}, and flow over a backwards step
\cite{gartling_1990}.

%%---------------------------------------------------------------------------%%
\pagebreak
\bibliographystyle{ieeetr}
\bibliography{references}
\end{document}

### 2012-08-24 doc/me699/proposal/me699_proposal.tex
updates to me699 proposal
finished 699 proposal

@@
 where $\ve{I}$ is the identity matrix, $\mu$ is the dynamic
-viscosityof the fluid, and $\kappa$ is the thermal conductivity of the
+viscosity of the fluid, and $\kappa$ is the thermal conductivity of the
 fluid. These fluid properties will be assumed constant and boundary
@@
 finite element discretization libraries, finite element assembly
-engines, and unstructured mesh databases available in the Trilinos
-scientific computing libraries \cite{trilinos_2005}. It will likely be
-easiest to start with the finite volume approach.
+engines, and unstructured mesh databases in the Trilinos scientific
+computing libraries \cite{trilinos_2005}. It will likely be easiest to
+start with the finite volume approach.
 
 \subsection{Solution Method}
-We will choose an inexact Newton method for the nonlinear scheme solve
-these equations \cite{dembo_1982}. Of particular interest will be
-investigating how the forcing term affects convergence of the
+We will choose an inexact Newton method for the nonlinear scheme to
+solve these equations \cite{dembo_1982}. Of particular interest will
+be investigating how the forcing term affects convergence of the
 nonlinear resdiual \cite{eisenstat_1996}. In addition, if convergence
@@
 the Newton correction \cite{pawlowski_2006}. The linear solver for the
-Newton method will be GMRES as its convergence properties for these
-types of systems are well documented \cite{saad_1986}. We will seek to
-solve these equations using parallel methods provided by well
-established framework libraries provided by the Trilinos suite of
-scientific computing libraries \cite{trilinos_2005}.
+Newton method will be GMRES \cite{saad_1986} as its convergence
+properties for these types of systems are well documented and its long
+recurrence relation has been shown to improve robustness and preserve
+a monotonically decreasing residuals in nonlinear solves. We will seek
+to solve these equations using parallel methods provided by well
+established framework libraries in the Trilinos suite of scientific
+computing libraries \cite{trilinos_2005}.
 
 \subsection{Benchmark Problems}
-To verify the implementation, three widely reference two-dimensional
+To verify the implementation, three widely referenced two-dimensional
 benchmark problems will be assessed using the resulting software. The

### 2012-08-28 doc/jcomp2012/jcomp2012.tex
Updated paragraph spacing: Stuart, make sure you only use blank lines in the .tex file where you want new paragraphs to start.

@@
 manner:
-
 \begin{itemize}
@@
 %%---------------------------------------------------------------------------%%
-
 \subsubsection{Direct Method}
@@
 %%---------------------------------------------------------------------------%%
-
 \subsubsection{Adjoint Method}
@@
 \end{equation}
-Note that the indices are reversed so that the probabilities are
-normalized over a column, as opposed to the forward method in which
-the probabilities are normalized over a row.  This is equivalent to
-forming the Neumann series in reverse order.  Correspondingly, the
-estimator for this method is
+Note that the indices are reversed so that the probabilities are normalized
+over a column, as opposed to the forward method in which the probabilities are
+normalized over a row.  This is equivalent to forming the Neumann series in
+reverse order.  Correspondingly, the estimator for this method is
 \begin{equation}
@@
 \end{equation}
-Here, $\hat{b}_{i_0}$ is the sampled source and initial weight in
-state $i_0$.  The Kronecker delta implies that tallies are only made
-in the state where the random walk currently resides.  This is the
-common approach in standard Monte Carlo transport simulations. Unlike
-the direct method, because tallies are made in the state in which the
-random walk currently resides, we are no longer required to start a
-history in each state for each random walk permutation. Instead,
-sampling the source is sufficient and requires only one random walk
-per permutation.
+Here, $\hat{b}_{i_0}$ is the sampled source and initial weight in state $i_0$.
+The Kronecker delta implies that tallies are only made in the state where the
+random walk currently resides.  This is the common approach in standard Monte
+Carlo transport simulations. Unlike the direct method, because tallies are
+made in the state in which the random walk currently resides, we are no longer
+required to start a history in each state for each random walk
+permutation. Instead, sampling the source is sufficient and requires only one
+random walk per permutation.
 
@@
 %%---------------------------------------------------------------------------%%
-
 \subsection{Monte Carlo Synthetic-Acceleration}
@@
 
-The methods presented in \S~\ref{sec:background} are characterized by
-slow convergence rates bound by the Central Limit Theorem.
-Halton~\cite{halton_1962,halton_1994} proposed a staged residual
-scheme called Sequential Monte Carlo to speed up the convergence of
-these methods.  A variant of this scheme has been successfully applied
-to the 1D non-linear thermal radiation diffusion equation in
-Ref.~\cite{evans_2003}.
-
-Concisely, the Sequential Monte Carlo method solves
-Eq.~(\ref{eq:Ax=b}) using the adjoint solution technique described in
-\S~\ref{sec:adjoint-method}.  The following iteration scheme is
-applied:
+The methods presented in \S~\ref{sec:background} are characterized by slow
+convergence rates bound by the Central Limit Theorem.
+Halton~\cite{halton_1962,halton_1994} proposed a staged residual scheme called
+Sequential Monte Carlo to speed up the convergence of these methods.  A
+variant of this scheme has been successfully applied to the 1D non-linear
+thermal radiation diffusion equation in Ref.~\cite{evans_2003}.
+
+Concisely, the Sequential Monte Carlo method solves Eq.~(\ref{eq:Ax=b}) using
+the adjoint solution technique described in \S~\ref{sec:adjoint-method}.  The
+following iteration scheme is applied:
 \begin{subequations}
@@
 In this scheme, the Monte Carlo adjoint method is used to estimate the
-solution to Eq.~(\ref{eq:residual_solve}), and the residual is
-iterated to convergence.  This iteration sequence is closely related
-to iterative refinement; the exception being that there is no update
-to $x^{l+1}$ between residual iterations.
+solution to Eq.~(\ref{eq:residual_solve}), and the residual is iterated to
+convergence.  This iteration sequence is closely related to iterative
+refinement; the exception being that there is no update to $x^{l+1}$ between
+residual iterations.
 
@@
 synthetic-acceleration for the fixed-point iteration given by
-Eq.~(\ref{eq:richardson}). We begin by subtracting
-Eq.~(\ref{eq:richardson}) from Eq:~(\ref{eq:point_iteration}):
+Eq.~(\ref{eq:richardson}). We begin by subtracting Eq.~(\ref{eq:richardson})
+from Eq:~(\ref{eq:point_iteration}):
 \begin{equation}
@@
 \end{subequations}
-To accelerate the fixed-point method and avoid inverting $\vA$
-directly in Eq.~(\ref{eq:synthetic-residual-solve}), we can instead
-use the above ideas to create a new iterative scheme by using the
-Monte Carlo method to estimate the solution error. With this in hand,
-the Fixed-Point Monte Carlo Synthetic-Acceleration (MCSA) method can
-be defined as follows:
+To accelerate the fixed-point method and avoid inverting $\vA$ directly in
+Eq.~(\ref{eq:synthetic-residual-solve}), we can instead use the above ideas to
+create a new iterative scheme by using the Monte Carlo method to estimate the
+solution error. With this in hand, the Fixed-Point Monte Carlo
+Synthetic-Acceleration (MCSA) method can be defined as follows:
 \begin{subequations}
@@
 \end{subequations}
-The hat on $\vA$ in Eq.~(\ref{eq:MCSA-residual_solve}) indicates that
-the Monte Carlo solution only approximately inverts this operator.
-Thus, we have defined a scheme in which the initial estimate of $\vx$
-in each iteration is updated using a single fixed-point
-iteration\footnote{Although, a subspace method could be used instead
-  of a stationary method here.}.  The residual is calculated and used
-as a source to estimate the error, $\delta\vx^{l+1/2}$, by solving
-Eq.~(\ref{eq:MCSA-residual_solve}) via the Monte Carlo adjoint
-method. The error is then used to calculate an updated iterate of the
-solution vector, $\vx^{l+1}$.  The entire sequence is iterated to
-convergence based on the following stopping
-criterion~\cite{kelley_1995},
+The hat on $\vA$ in Eq.~(\ref{eq:MCSA-residual_solve}) indicates that the
+Monte Carlo solution only approximately inverts this operator.  Thus, we have
+defined a scheme in which the initial estimate of $\vx$ in each iteration is
+updated using a single fixed-point iteration\footnote{Although, a subspace
+  method could be used instead of a stationary method here.}.  The residual is
+calculated and used as a source to estimate the error, $\delta\vx^{l+1/2}$, by
+solving Eq.~(\ref{eq:MCSA-residual_solve}) via the Monte Carlo adjoint
+method. The error is then used to calculate an updated iterate of the solution
+vector, $\vx^{l+1}$.  The entire sequence is iterated to convergence based on
+the following stopping criterion~\cite{kelley_1995},
 \begin{equation}
@@
 
-The MCSA method defined in Eq.~(\ref{eq:MCSA-iteration}) uses the
-adjoint method to estimate the error in
-Eq.~(\ref{eq:MCSA-residual_solve}) instead of the direct method
-outlined in \S~\ref{sec:direct-method}. To demonstrate the
-effectiveness of the adjoint method over the direct method within the
-context of MCSA, we choose the two-dimensional time-dependent Poisson
-equation as a simple model problem:
-
+The MCSA method defined in Eq.~(\ref{eq:MCSA-iteration}) uses the adjoint
+method to estimate the error in Eq.~(\ref{eq:MCSA-residual_solve}) instead of
+the direct method outlined in \S~\ref{sec:direct-method}. To demonstrate the
+effectiveness of the adjoint method over the direct method within the context
+of MCSA, we choose the two-dimensional time-dependent Poisson equation as a
+simple model problem:
 \begin{equation}
@@
 \end{equation}
-
-For all comparisons, a single time step is computed with backwards
-Euler time integration. The Laplacian is differenced on a square
-Cartesian grid with a second-order five point stencil
-
+For all comparisons, a single time step is computed with backwards Euler time
+integration. The Laplacian is differenced on a square Cartesian grid with a
+second-order five point stencil
 \begin{equation}
@@
 \end{equation}
-
 and a fourth-order nine point stencil 
-
 \begin{multline}
@@
 \end{multline}
-
-both assuming a grid size of $\Delta$ in both the $i$ and $j$
-directions. For a single time step solution, we then have the
-following sparse linear system to be solved with the MCSA method:
-
+both assuming a grid size of $\Delta$ in both the $i$ and $j$ directions. For
+a single time step solution, we then have the following sparse linear system
+to be solved with the MCSA method:
 \begin{equation}
@@
 \end{equation}
-
-Both the stencils will be used to vary the size and density of the
-sparse linear system in Eq.~(\ref{eq:poisson_eq_lin_sys}).
-
-A timing study and convergence study is used to demonstrate the
-effectiveness of the adjoint method as compared to the direct
-method. To assess both the CPU time and number of iterations required
-to converge to a solution, a problem of constant $\Delta$ was used
-with varying values of grid size, fixing the spectral radius of the
-system at a constant value for each variation. Both the five point and
-nine point stencils were used with both the direct and adjoint
-solvers. For each case, 50 random walk permutations were computed per
-MCSA iteration with a weight cutoff of \sn{1}{-4} for the stochastic
-linear solver and a convergence tolerance of \sn{1}{-8} for the MCSA
-iterative solver. All computations presented in this section were
-completed on a 2.66 GHz Intel Core i7 machine with 8 GB 1067 MHz DDR3
-memory. Fig.~\ref{fig:poisson_cpu_time} gives the CPU time needed for
-each case to converge in seconds and Fig.~\ref{fig:poisson_iterations}
-gives the number of iterations needed for each case to converge
-relative to the problem size.
-
+Both the stencils will be used to vary the size and density of the sparse
+linear system in Eq.~(\ref{eq:poisson_eq_lin_sys}).
+
+A timing study and convergence study is used to demonstrate the effectiveness
+of the adjoint method as compared to the direct method. To assess both the CPU
+time and number of iterations required to converge to a solution, a problem of
+constant $\Delta$ was used with varying values of grid size, fixing the
+spectral radius of the system at a constant value for each variation. Both the
+five point and nine point stencils were used with both the direct and adjoint
+solvers. For each case, 50 random walk permutations were computed per MCSA
+iteration with a weight cutoff of \sn{1}{-4} for the stochastic linear solver
+and a convergence tolerance of \sn{1}{-8} for the MCSA iterative solver. All
+computations presented in this section were completed on a 2.66 GHz Intel Core
+i7 machine with 8 GB 1067 MHz DDR3 memory. Fig.~\ref{fig:poisson_cpu_time}
+gives the CPU time needed for each case to converge in seconds and
+Fig.~\ref{fig:poisson_iterations} gives the number of iterations needed for
+each case to converge relative to the problem size.
 \begin{figure}[h!]
@@
 
-We see clearly in Fig.~\ref{fig:poisson_cpu_time} that the using the
-adjoint solver with MCSA results in several orders of magnitude
-speedup over the direct solver while the number of iterations required
-to converge is of a similar scale. We expect this for several
-reasons. First, with an equivalent number of histories specified for
-both solvers and a system of size $N \times N$, the direct solver will
-compute $N \times N$ random walks per permutation while the adjoint
-solver will only compute one. This is necessary in the direct method
-to ensure a contribution from each state as the random walk sequence
-will only contribute to the starting state. For the adjoint method,
+We see clearly in Fig.~\ref{fig:poisson_cpu_time} that the using the adjoint
+solver with MCSA results in several orders of magnitude speedup over the
+direct solver while the number of iterations required to converge is of a
+similar scale. We expect this for several reasons. First, with an equivalent
+number of histories specified for both solvers and a system of size $N \times
+N$, the direct solver will compute $N \times N$ random walks per permutation
+while the adjoint solver will only compute one. This is necessary in the
+direct method to ensure a contribution from each state as the random walk
+sequence will only contribute to the starting state. For the adjoint method,
 because the random walk sequence contributes to the state in which it
-currently resides, fewer histories are neccessary to compute
-contributions from all the nonzero parts of the source. Initially,
-this may not seem like a fair comparison, however, we see from
-Fig.~\ref{fig:poisson_iterations} that the number of iterations
-required to converge is approximately the same and therefore the extra
-computations performed by the direct method do not improve the error
-estimation. Therefore, although the total number of random walks per
-iteration is in fact larger for the direct method, the solution
-computed at each iteration is no better than that computed by the
-adjoint method.
-
-As an additional comparison, the convergence behavior of MCSA can be
-analyzed using both solvers to detect any performance benefits. To
-assess the convergence properties of MCSA using each solver and
-stencil, the infinity norm of the residual computed in
-Eq.~(\ref{eq:MCSA-residual}) was collected at each iteration for a
-fixed problem size. Fig.~\ref{fig:poisson_convergence} gives the
-results of these computations. First, it is worthy to note on the
-semilog plot that we are indeed achieving the expected exponential
-convergence from MCSA. Second, we note that both the direct and
-adjoint methods have approximately equivalent convergence behavior,
-leading to CPU time being the dominating factor in method selection.
-
+currently resides, fewer histories are neccessary to compute contributions
+from all the nonzero parts of the source. Initially, this may not seem like a
+fair comparison, however, we see from Fig.~\ref{fig:poisson_iterations} that
+the number of iterations required to converge is approximately the same and
+therefore the extra computations performed by the direct method do not improve
+the error estimation. Therefore, although the total number of random walks per
+iteration is in fact larger for the direct method, the solution computed at
+each iteration is no better than that computed by the adjoint method.
+
+As an additional comparison, the convergence behavior of MCSA can be analyzed
+using both solvers to detect any performance benefits. To assess the
+convergence properties of MCSA using each solver and stencil, the infinity
+norm of the residual computed in Eq.~(\ref{eq:MCSA-residual}) was collected at
+each iteration for a fixed problem size. Fig.~\ref{fig:poisson_convergence}
+gives the results of these computations. First, it is worthy to note on the
+semilog plot that we are indeed achieving the expected exponential convergence
+from MCSA. Second, we note that both the direct and adjoint methods have
+approximately equivalent convergence behavior, leading to CPU time being the
+dominating factor in method selection.
 \begin{figure}[h!]
@@
 
-Using the mathematical preliminaries developed in the previous
-section, we can now generate a discrete form of the radiation
-diffusion equation. We will show that this discrete form is amenable
-to interpretation using Monte Carlo concepts and can be written in a
-way that a random walk sequence can be generated.
+Using the mathematical preliminaries developed in the previous section, we can
+now generate a discrete form of the radiation diffusion equation. We will show
+that this discrete form is amenable to interpretation using Monte Carlo
+concepts and can be written in a way that a random walk sequence can be
+generated.
 
@@
 
-The general form of the equilibrium diffusion equation is
-\cite{morel_1996}
-
+The general form of the equilibrium diffusion equation is \cite{morel_1996}
 \begin{equation} 
@@
 \end{equation}
-
 where $\Cv\equiv\Cv(\ve{r},T(\ve{r},t))$
-$[\text{GJ}\cdot\text{g}^{-1}\cdot\text{keV}^{-1}]$ is the specific
-heat capacity of the material, $\rho\equiv\rho(\ve{r})$
-$[\text{g}\cdot\text{cm}^{-3}]$ is the density of the material,
-$a=0.01372$ $[\text{GJ}\cdot\text{cm}^{-3}\cdot\text{keV}^{-4}]$ is
-the radiation constant, $c=29.979$ $[\text{cm}\cdot\text{ns}^{-1}]$ is
-the vacuum light speed, and $T\equiv T(\ve{r},t)$ $[\text{keV}]$ is
-the temperature that characterizes both the radiation and the
-material.  The opacity, $\ros\equiv\ros$(\ve{r},T(\ve{r},t))
-$[\text{cm}^{-1}]$, is the Rosseland Mean opacity.  The source,
-$Q\equiv Q(\ve{r},t)$
-$[\text{GJ}\cdot\text{cm}^{-3}\cdot\text{ns}^{-1}]$, is assumed to
-be independent of $T$.
+$[\text{GJ}\cdot\text{g}^{-1}\cdot\text{keV}^{-1}]$ is the specific heat
+capacity of the material, $\rho\equiv\rho(\ve{r})$
+$[\text{g}\cdot\text{cm}^{-3}]$ is the density of the material, $a=0.01372$
+$[\text{GJ}\cdot\text{cm}^{-3}\cdot\text{keV}^{-4}]$ is the radiation
+constant, $c=29.979$ $[\text{cm}\cdot\text{ns}^{-1}]$ is the vacuum light
+speed, and $T\equiv T(\ve{r},t)$ $[\text{keV}]$ is the temperature that
+characterizes both the radiation and the material.  The opacity,
+$\ros\equiv\ros$(\ve{r},T(\ve{r},t)) $[\text{cm}^{-1}]$, is the Rosseland Mean
+opacity.  The source, $Q\equiv Q(\ve{r},t)$
+$[\text{GJ}\cdot\text{cm}^{-3}\cdot\text{ns}^{-1}]$, is assumed to be
+independent of $T$.
 
@@
 $\phi\equiv\phi(\ve{r},t)$
-$[\text{GJ}\cdot\text{cm}^{-2}\cdot\text{ns}^{-1}]$, as the primary
-state variable.  The intensity is defined as:
-
+$[\text{GJ}\cdot\text{cm}^{-2}\cdot\text{ns}^{-1}]$, as the primary state
+variable.  The intensity is defined as:
 \begin{equation}
@@
 \end{equation}
-
-where $\psi$ is the angular radiation intensity and $\nu$ is the
-radiation frequency.  The Planck function (or Planckian) is defined
-as,
-
+where $\psi$ is the angular radiation intensity and $\nu$ is the radiation
+frequency.  The Planck function (or Planckian) is defined as,
 \begin{equation}
@@
 \end{equation}
-
-where $h$ is Planck's constant and $k$ is the Boltzmann constant.
-Utilizing this definition of $\phi$ we can write
-
+where $h$ is Planck's constant and $k$ is the Boltzmann constant.  Utilizing
+this definition of $\phi$ we can write
 \begin{equation}
@@
 \end{equation}
-
 Inserting \eqref{eq:partial-T} into \eqref{eq:eq-diff-T} yields
-
 \begin{equation}
@@
 \end{equation}
-
 where $D$ is the diffusion coefficient, $D\equiv
 D(\ve{r},T(\ve{r},t))\ [\text{cm}]$, and is defined as
-
 \begin{equation}
@@
 \end{equation}
-
-Equation~\eqref{eq:eq-diff} is the model equation we will solve using
-the Monte Carlo techniques presented in
-\S~\ref{sec:monte-carlo-matrix} and \ref{sec:iter-refin-monte}.
+Equation~\eqref{eq:eq-diff} is the model equation we will solve using the
+Monte Carlo techniques presented in \S~\ref{sec:monte-carlo-matrix} and
+\ref{sec:iter-refin-monte}.
 
@@
 
-Applying backward-Euler time differencing to Eq.~(\ref{eq:eq-diff})
-and lagging the temperature-dependent coefficients at $t^n$ gives
-
+Applying backward-Euler time differencing to Eq.~(\ref{eq:eq-diff}) and
+lagging the temperature-dependent coefficients at $t^n$ gives
 \begin{equation}
@@
 \end{equation}
-
 where we have suppressed the $n+1$ indices on $\phi$ and
-
 \begin{align}
@@
 \end{align}
-
 Applying Fick's Law,
-
 \begin{equation}
@@
 \end{equation}
-
-where $\ve{F}$ is the radiation flux, we define a flux-balance
-equation,
-
+where $\ve{F}$ is the radiation flux, we define a flux-balance equation,
 \begin{equation}
@@
 \end{equation}
-
 Integrating Eq.~(\ref{eq:flux-balance}) over the cell illustrated in
 Fig.~\ref{fig:cell} gives
-
 \begin{equation}
@@
 \end{equation}
-
 and $V_{ijk} = \Delta_i\Delta_j\Delta_k$.
-
 \begin{figure}[h!]
@@
 \end{figure}
-
 We now apply the divergence theorem,
-
 \begin{equation}
@@
 \end{equation}
-
 to Eq.(\ref{eq:flux-balance-vol-integral}) yielding
-
 \begin{equation}
@@
 
-Using Fick's Law, Eq.~(\ref{eq:fick}), we can write $O(\Delta^2)$
-expressions for the face-centered fluxes.  Defining $l\in\{i,j,k\}$,
-the face-centered fluxes are
-
+Using Fick's Law, Eq.~(\ref{eq:fick}), we can write $O(\Delta^2)$ expressions
+for the face-centered fluxes.  Defining $l\in\{i,j,k\}$, the face-centered
+fluxes are
 \begin{align}
@@
 \end{align}
-
 where
-
 \begin{align}
@@
 \end{align}
-
-In order to be conservative, the flux must be continuous across a
-face.  Thus, the face-centered flux evaluated from either side of the
-face must be equivalent,
-
+In order to be conservative, the flux must be continuous across a face.  Thus,
+the face-centered flux evaluated from either side of the face must be
+equivalent,
 \begin{equation}
@@
 \end{equation}
-
-where we have dropped the subscript $\mathrm{R}$ from $\sigma$ for
-clarity.  Solving this system for $\phi_{l+1/2}$ and applying to
-either $F_{l+1/2}^{+}$ or $F_{l+1/2}^{-}$ and setting the resulting
-expression equal to Eq.~(\ref{eq:F_l+1/2}), one finds the
-face-centered diffusion coefficient,
-
+where we have dropped the subscript $\mathrm{R}$ from $\sigma$ for clarity.
+Solving this system for $\phi_{l+1/2}$ and applying to either $F_{l+1/2}^{+}$
+or $F_{l+1/2}^{-}$ and setting the resulting expression equal to
+Eq.~(\ref{eq:F_l+1/2}), one finds the face-centered diffusion coefficient,
 \begin{equation}
@@
 \end{equation}
-
 Following the same procedure at the $l-1/2$ face gives the diffusion
 coefficient on the low-side face,
-
 \begin{equation}
@@
 (\ref{eq:flux-balance-difference}), we obtain
-
 \begin{equation}
@@
 \end{equation}
-
 Here,
-
 \begin{equation}
@@
 \end{equation}
-
 and
-
 \begin{subequations}
@@
 \end{subequations}
-
 Also, note that
-
 \begin{align*}
@@
 \end{align*}
-
-Equations~(\ref{eq:discrete-diffusion}) through (\ref{eq:sigmas})
-represent the discretized, 1T diffusion equation on an orthogonal 3D
-grid.
-
-Equations~(\ref{eq:discrete-diffusion}) through (\ref{eq:sigmas}) have
-been written in a manner that is amenable to Monte Carlo
-interpretation.  One can envision the terms in Eq.~(\ref{eq:sigmas})
-representing leakage out of each face of a cell.  The off-diagonal
-terms then represent sources into the cell, or, equivalently, these
-terms are leakage out of the neighboring cells.  A limitation of this
-interpretation is that it requires a symmetric system to implement the
-Monte Carlo scheme.  More specifically, the operator $\ve{D}$ must be
-an H-matrix \cite{kelley_1995} such that the off-diagonal elements are
-negative and symmetric.  In other words, the leakage out of a cell
-must be equivalent to the source into the adjacent cell.
+Equations~(\ref{eq:discrete-diffusion}) through (\ref{eq:sigmas}) represent
+the discretized, 1T diffusion equation on an orthogonal 3D grid.
+
+Equations~(\ref{eq:discrete-diffusion}) through (\ref{eq:sigmas}) have been
+written in a manner that is amenable to Monte Carlo interpretation.  One can
+envision the terms in Eq.~(\ref{eq:sigmas}) representing leakage out of each
+face of a cell.  The off-diagonal terms then represent sources into the cell,
+or, equivalently, these terms are leakage out of the neighboring cells.  A
+limitation of this interpretation is that it requires a symmetric system to
+implement the Monte Carlo scheme.  More specifically, the operator $\ve{D}$
+must be an H-matrix \cite{kelley_1995} such that the off-diagonal elements are
+negative and symmetric.  In other words, the leakage out of a cell must be
+equivalent to the source into the adjacent cell.
 
 The methods described in \S\S~\ref{sec:monte-carlo-matrix} and
-\ref{sec:iter-refin-monte} do not suffer this limitation.  Thus, we
-will interpret the solution via MCSA as a transport-like process, but
-we do not require any constraints on the system other than the
-spectral radius must be less than 1.
-
-Before completing the derivation of discrete equations required to
-solve Eq.~(\ref{eq:discrete-diffusion}), we must specify boundary
-conditions.  For this work, the Marshak Boundary condition suffices,
-
+\ref{sec:iter-refin-monte} do not suffer this limitation.  Thus, we will
+interpret the solution via MCSA as a transport-like process, but we do not
+require any constraints on the system other than the spectral radius must be
+less than 1.
+
+Before completing the derivation of discrete equations required to solve
+Eq.~(\ref{eq:discrete-diffusion}), we must specify boundary conditions.  For
+this work, the Marshak Boundary condition suffices,
 \begin{equation}
@@
 \end{equation}
-
 where $f_b$ is the incoming partial current on the problem boundary,
-
 $\partial\Gamma$.  On any low boundary we can write the flux using a
 first-order discrete form of Fick's Law,
-
 \begin{equation}
@@
 \end{equation}
-
 As before, we enforce continuity of flux on the boundary using the
 Marshak Boundary condition,
-
 \begin{equation}
@@
 \end{equation}
-
-Solving for $\phi_{1/2}$ and applying to
-Eq.~(\ref{eq:low-boundary-flux}) yields an expression for the low-side
-boundary flux,
-
+Solving for $\phi_{1/2}$ and applying to Eq.~(\ref{eq:low-boundary-flux})
+yields an expression for the low-side boundary flux,
 \begin{equation}
@@
 \end{equation}
-
 Following an identical procedure on high-side problem boundaries gives
-
 \begin{equation}
@@
 \end{equation}
-
 where $l\in\{i,j,k\}$ is defined over the range $[1,N_l]$.
 
-These boundary fluxes take the place of the interior face-centered
-fluxes (defined in Eqs.~(\ref{eq:F_l+1/2}) and (\ref{eq:F_l-1/2})) in
-Eq.~(\ref{eq:flux-balance-difference}) yielding the following
-modifications to Eqs.~(\ref{eq:sigmas}) on a boundary,
-
+These boundary fluxes take the place of the interior face-centered fluxes
+(defined in Eqs.~(\ref{eq:F_l+1/2}) and (\ref{eq:F_l-1/2})) in
+Eq.~(\ref{eq:flux-balance-difference}) yielding the following modifications to
+Eqs.~(\ref{eq:sigmas}) on a boundary,
 \begin{align}
@@
 \end{align}
-
-Furthermore, the following contributions are added to the source,
-
-$\qn_{ijk}$ on a boundary cell,
+Furthermore, the following contributions are added to the source, $\qn_{ijk}$
+on a boundary cell,
 \begin{align}
@@
 \end{align}
-
 Note that for vacuum boundary conditions, $f_b^{\pm} = 0$.
 
-For reflective boundary conditions we have the simple modification
-that
-
+For reflective boundary conditions we have the simple modification that
 \begin{align}
@@
 \end{align}
-
-Thus, there are no leakage or adjacent cell source terms resulting
-from this boundary face.
+Thus, there are no leakage or adjacent cell source terms resulting from this
+boundary face.
 
@@
 
-The discrete diffusion equation in (\ref{eq:discrete-diffusion}) can
-be written in matrix-operator form as follows,
-
+The discrete diffusion equation in (\ref{eq:discrete-diffusion}) can be
+written in matrix-operator form as follows,
 \begin{equation}
@@
 \end{equation}
-
 where $\ve{D}$ is a symmetric, positive-definite (SPD) operator.
-Unfortunately, for many problems $\rho(\ve{D}) > 1$.  Therefore, we
-must precondition Eq.~(\ref{eq:diffusion-matrix}) in order to use the
-MCSA method. We apply left-preconditioning and write
-
+Unfortunately, for many problems $\rho(\ve{D}) > 1$.  Therefore, we must
+precondition Eq.~(\ref{eq:diffusion-matrix}) in order to use the MCSA
+method. We apply left-preconditioning and write
 \begin{equation}
@@
 \end{equation}
-
 A good choice for the preconditioner, $\ve{M}$, is to use Jacobi
 preconditioning:
-
 \begin{equation}
@@
 \end{equation}
-
-Setting $\ve{M}$ equal to the diagonal elements of $\ve{D}$
-accomplishes several objectives:
-
+Setting $\ve{M}$ equal to the diagonal elements of $\ve{D}$ accomplishes
+several objectives:
 \begin{itemize}
@@
 \end{itemize}
-
 The second item is quite important for Monte Carlo solutions of linear
-systems, and this type of scaling should be performed regardless of
-any additional preconditioner choices. We note that the resulting
-system does not have to be SPD for the MCSA algorithm, thus any
-suitable choice of $\ve{M}$ will suffice.  The preconditioned system
-we will solve is defined in Eq.~(\ref{eq:preconditioned-diffusion}).
-Applying $\ve{M}^{-1}$ to Eq.~(\ref{eq:discrete-diffusion}) yields
-
+systems, and this type of scaling should be performed regardless of any
+additional preconditioner choices.
+
+An advantage of MCSA is that the resulting system does not have to be SPD,
+thus any suitable choice of $\ve{M}$ will suffice.  The preconditioned system
+we will solve is defined in Eq.~(\ref{eq:preconditioned-diffusion}).  Applying
+$\ve{M}^{-1}$ to Eq.~(\ref{eq:discrete-diffusion}) yields
 \begin{equation}
@@
 Here we derive the equations required to solve
-Eq.~(\ref{eq:preconditioned-discrete-diffusion}) using the MCSA
-method.  As discussed in \S~\ref{sec:iter-refin-monte}, the MCSA
-method requires two solutions:
-
+Eq.~(\ref{eq:preconditioned-discrete-diffusion}) using the MCSA method.  As
+discussed in \S~\ref{sec:iter-refin-monte}, the MCSA method requires two
+solutions:
 \begin{enumerate}
@@
 \end{enumerate}
-
 For clarity, we rewrite the iteration scheme in
@@
 from Eq.~(\ref{eq:preconditioned-diffusion}),
-
 \begin{align*}
@@
 \end{align*}
-
-As described in \S~\ref{sec:iter-refin-monte}, the adjoint Monte Carlo
-method only approximately inverts $\ve{M}^{-1}\ve{D}$.  We now
-describe each of these in turn.
-
-First, we use Eq.~(\ref{eq:preconditioned-discrete-diffusion}) to
-evaluate $\bphi^{l+1/2}$ via a single fixed-point iteration as
-follows:
-
+As described in \S~\ref{sec:iter-refin-monte}, the adjoint Monte Carlo method
+only approximately inverts $\ve{M}^{-1}\ve{D}$.  We now describe each of these
+in turn.
+
+First, we use Eq.~(\ref{eq:preconditioned-discrete-diffusion}) to evaluate
+$\bphi^{l+1/2}$ via a single fixed-point iteration as follows:
 \begin{equation}
@@
 \end{equation}
-
-This result is used to compute the residual of the system.  The
-residual then becomes the source for the Monte Carlo simulation.
-
-The adjoint Monte Carlo method described in
-\S~\ref{sec:adjoint-method} is used to estimate
-
+This result is used to compute the residual of the system.  The residual then
+becomes the source for the Monte Carlo simulation.
+
+The adjoint Monte Carlo method described in \S~\ref{sec:adjoint-method} is
+used to estimate
 \begin{equation}
@@
 \end{equation}
-
-As noted above, the residual acts as the source for this simulation,
-and the operator $\ve{M}^{-1}\ve{D}$ is implicitly defined in
-Eq.~(\ref{eq:preconditioned-discrete-diffusion}).  There are two Monte
-Carlo interpretations that can be applied to this system.  The first
-follows the mathematical presentation given in
-\S~\ref{sec:monte-carlo-matrix}.  Namely, we form probabilities from
-the iteration matrix and perform random walks to generate unbiased
-estimates of the solution vector, $\delta\bphi^{l+1/2}$, using the
-estimator in Eq.~(\ref{eq:adjoint-tally}).
-
-However, a more natural interpretation for Monte Carlo transport
-practitioners is to use the machinery described in
-\S~\ref{sec:monte-carlo-matrix} to define Probability Distribution
-Functions (PDF) that describe the transport of particles between
-cells.  We then do a Monte Carlo transport calculation using
+As noted above, the residual acts as the source for this simulation, and the
+operator $\ve{M}^{-1}\ve{D}$ is implicitly defined in
+Eq.~(\ref{eq:preconditioned-discrete-diffusion}).  There are two Monte Carlo
+interpretations that can be applied to this system.  The first follows the
+mathematical presentation given in \S~\ref{sec:monte-carlo-matrix}.  Namely,
+we form probabilities from the iteration matrix and perform random walks to
+generate unbiased estimates of the solution vector, $\delta\bphi^{l+1/2}$,
+using the estimator in Eq.~(\ref{eq:adjoint-tally}).
+
+However, a more natural interpretation for Monte Carlo transport practitioners
+is to use the machinery described in \S~\ref{sec:monte-carlo-matrix} to define
+Probability Distribution Functions (PDF) that describe the transport of
+particles between cells.  We then do a Monte Carlo transport calculation using
 Eq.~(\ref{eq:adjoint-weight}) to calculate the weight change at each
-collision. Equation~(\ref{eq:adjoint-probability}) gives the
-probability of scattering to an adjacent cell. We tally the
-contribution of $\phi$ in each cell using
-Eq.~(\ref{eq:adjoint-tally}).  The transport is terminated using a
-relative weight cutoff that is defined in
-Eq.~(\ref{eq:weight_cutoff}).  We now examine each of these in
-more detail.
-
-The connectivity of an arbitrary cell is illustrated in the 3D mesh
-segment in Fig.~\ref{fig:mesh_segment}.  For a particle residing in
-the blue cell, any collision results in a scatter to one of the six
-adjacent cells since $p_{ii}=0$ in the preconditioned system.  We can
-easily calculate the probabilities for each scattering event from
-Eq.~(\ref{eq:preconditioned-discrete-diffusion}) without explicitly
-forming the iteration matrix.  In this mesh cell, the probabilities
-for leakage out of each face are derived from the equations from each
-adjacent cell.  Using Eq.~(\ref{eq:preconditioned-discrete-diffusion})
-we can explicitly write the equations from each adjacent cell, here
-only noting the contribution from cell 7:
-
+collision. Equation~(\ref{eq:adjoint-probability}) gives the probability of
+scattering to an adjacent cell. We tally the contribution of $\phi$ in each
+cell using Eq.~(\ref{eq:adjoint-tally}).  The transport is terminated using a
+relative weight cutoff that is defined in Eq.~(\ref{eq:weight_cutoff}).  We
+now examine each of these in more detail.
+
+The connectivity of an arbitrary cell is illustrated in the 3D mesh segment in
+Fig.~\ref{fig:mesh_segment}.  For a particle residing in the blue cell, any
+collision results in a scatter to one of the six adjacent cells since
+$p_{ii}=0$ in the preconditioned system.  We can easily calculate the
+probabilities for each scattering event from
+Eq.~(\ref{eq:preconditioned-discrete-diffusion}) without explicitly forming
+the iteration matrix.  In this mesh cell, the probabilities for leakage out of
+each face are derived from the equations from each adjacent cell.  Using
+Eq.~(\ref{eq:preconditioned-discrete-diffusion}) we can explicitly write the
+equations from each adjacent cell, here only noting the contribution from cell
+7:
 \begin{align*}
@@
 \end{align*} 
-
-A quick view of these equations reveals that the only non-zero entries
-in column 7 of the iteration matrix are
-
+A quick view of these equations reveals that the only non-zero entries in
+column 7 of the iteration matrix are
 \begin{equation*}
@@
 \end{equation*}
-
 where, following Eq.~(\ref{eq:point_iteration}), $\ve{H} = (\vI -
 \ve{M}^{-1}\ve{D})$ is the iteration matrix.  Now, using
-Eq.~(\ref{eq:adjoint-probability}), we define the scattering
-probabilities for cell 7 as
-
+Eq.~(\ref{eq:adjoint-probability}), we define the scattering probabilities for
+cell 7 as
 \begin{equation*}
@@
 \end{equation*}
-
 The weight change for each scattering event is calculated using
 Eq.~(\ref{eq:adjoint-weight}),
-
 \begin{equation*}
@@
 \end{equation*}
-
-where $m$ is the step-index in the particle history.  We also note
-that, for the 3D Cartesian mesh considered in this work, the maximum
-number of non-zero entries in columns of the iteration matrix is 6,
-generating a sparse system.  Cells along the problem boundary can have
-fewer than 6 entries.
-
+where $m$ is the step-index in the particle history.  We also note that, for
+the 3D Cartesian mesh considered in this work, the maximum number of non-zero
+entries in columns of the iteration matrix is 6, generating a sparse system.
+Cells along the problem boundary can have fewer than 6 entries.
 \begin{figure}[h!]
@@
 traditional deterministic solution methods.  The deterministic methods
-considered are preconditioned Conjugate Gradient (PCG) and
-preconditioned fixed-point (Richardson) iteration (PFIX).  The PCG
-method uses right-left preconditioning with
-$\ve{M}=\ve{L}\ve{L}^{-1}$.  The PFIX scheme is used to solve
-Eq.~(\ref{eq:preconditioned-diffusion}), which is identical to the
-system being solved with MCSA.  We examine each method on two 3D
-problems: a Marshak wave problem and a multi-material duct problem.
+considered are preconditioned Conjugate Gradient (PCG) and preconditioned
+fixed-point (Richardson) iteration (PFIX).  The PCG method uses right-left
+preconditioning with $\ve{M}=\ve{L}\ve{L}^{-1}$.  The PFIX scheme is used to
+solve Eq.~(\ref{eq:preconditioned-diffusion}), which is identical to the
+system being solved with MCSA.  We examine each method on two 3D problems: a
+Marshak wave problem and a multi-material duct problem.
 
@@
 
-The Marshak wave problem under consideration in this study has an
-incoming 1~keV blackbody boundary flux at $x=0$ and propagates in the
-$x$-dimension in time as outlined in Ref.~\cite{larsen_1980}.  In
-other words, this a 3D representation of a 1D problem.  The full set
-of problem conditions is:
-
+The Marshak wave problem under consideration in this study has an incoming
+1~keV blackbody boundary flux at $x=0$ and propagates in the $x$-dimension in
+time as outlined in Ref.~\cite{larsen_1980}.  In other words, this a 3D
+representation of a 1D problem.  The full set of problem conditions is:
 \begin{center}
@@
 \end{center}
-
 The problem is run to $100$~nanoseconds.
 
-Figure~\ref{fig:marshak_10} shows the analytic Marshak Wave solution
-compared to the MCSA solution.  As opposed to most Monte Carlo
-techniques, the MCSA method generates solutions that are numerically
-equivalent to the PCG and PFIX results.  Because the MCSA method
-converges the residual of the system, the numerical accuracy is
-determined by the stopping criterion shown in
-Eq.~(\ref{eq:stopping-criteria}).  Also, the MCSA method is not bound
-by the Central Limit Theorem so convergence is not restricted to
+Figure~\ref{fig:marshak_10} shows the analytic Marshak Wave solution compared
+to the MCSA solution.  As opposed to most Monte Carlo techniques, the MCSA
+method generates solutions that are numerically equivalent to the PCG and PFIX
+results.  Because the MCSA method converges the residual of the system, the
+numerical accuracy is determined by the stopping criterion shown in
+Eq.~(\ref{eq:stopping-criteria}).  Also, the MCSA method is not bound by the
+Central Limit Theorem so convergence is not restricted to
 $1/\sqrt{N}$~\cite{halton_1994,evans_2003}.
-
 \begin{figure}[h!]
@@
 
-As stated in \S~\ref{sec:monte-carlo-matrix}, the MCSA method will
-converge when $\rho(\ve{H})< 1$ where $\ve{H}$ is the iteration matrix
-and is defined $\ve{H} = (\vI - \ve{M}^{-1}\ve{D})$.  The spectral
-radius for the Marshak problem is plotted in
-Fig.~\ref{fig:marshak_spectral}.  As stated in
-\S~\ref{sec:preconditioning} the preconditioning effectively reduces
-the spectral radius of $\ve{H}$ to less than 1.
-
+As stated in \S~\ref{sec:monte-carlo-matrix}, the MCSA method will converge
+when $\rho(\ve{H})< 1$ where $\ve{H}$ is the iteration matrix and is defined
+$\ve{H} = (\vI - \ve{M}^{-1}\ve{D})$.  The spectral radius for the Marshak
+problem is plotted in Fig.~\ref{fig:marshak_spectral}.  As stated in
+\S~\ref{sec:preconditioning} the preconditioning effectively reduces the
+spectral radius of $\ve{H}$ to less than 1.
 \begin{figure}[h!]
@@
 
-There are two principal degrees of freedom when utilizing the MCSA
-method: (1) the number of particles per stage, and (2) the weight
-cutoff.  Figure~\ref{fig:CPU_Np} shows the CPU time as a function of
-the requested number of particles per stage for a relative weight
-cutoff of $\sn{1}{-4}$.  For less than 5 particles per stage the
-method does not converge.  Figure~\ref{fig:CPU_Ns} shows the variation
-in the maximum number of stages (iterations) per cycle with number of
-particles.  Analysis of these two figures shows that increasing the
-number of particles per stage reduces the number of iterations
-required to converge the solution.  However, the cost of the Monte
-Carlo transport is high; thus, better performance is attained by
+There are two principal degrees of freedom when utilizing the MCSA method: (1)
+the number of particles per stage, and (2) the weight cutoff.
+Figure~\ref{fig:CPU_Np} shows the CPU time as a function of the requested
+number of particles per stage for a relative weight cutoff of $\sn{1}{-4}$.
+For less than 5 particles per stage the method does not converge.
+Figure~\ref{fig:CPU_Ns} shows the variation in the maximum number of stages
+(iterations) per cycle with number of particles.  Analysis of these two
+figures shows that increasing the number of particles per stage reduces the
+number of iterations required to converge the solution.  However, the cost of
+the Monte Carlo transport is high; thus, better performance is attained by
 running more stages with fewer particles per stage \cite{evans_2003}.
-
 \begin{figure}[h!]
@@
 
-Figure~\ref{fig:CPU_wc} shows the CPU time as a function of relative
-weight cutoff.  These results indicate that the performance of the
-MCSA method is relatively insensitive to the weight cutoff.  This is
-especially true when running small numbers of particles per stage.
-When larger numbers of particles are used the weight cutoff has a more
-pronounced effect on the efficiency of the method.
-
+Figure~\ref{fig:CPU_wc} shows the CPU time as a function of relative weight
+cutoff.  These results indicate that the performance of the MCSA method is
+relatively insensitive to the weight cutoff.  This is especially true when
+running small numbers of particles per stage.  When larger numbers of
+particles are used the weight cutoff has a more pronounced effect on the
+efficiency of the method.
 \begin{figure}[h!]
@@
 
-Having analyzed the variation of the MCSA algorithm efficiency with
-number of particles and weight cutoff, we now compare to traditional
-deterministic algorithms.  Table~\ref{tab:marshak_comparison} shows
-timing results for the Marshak problem using preconditioned CG (PCG),
-preconditioned Richardson iteration (PFIX), and MCSA. The results show
-that the MCSA method is the most efficient method, although the PCG
-results are nearly as fast.  More importantly, the MCSA method clearly
-accelerates Richardson iteration, as indicated by the fact that MCSA
-is 42\% faster than Richardson iteration alone.
-
+Having analyzed the variation of the MCSA algorithm efficiency with number of
+particles and weight cutoff, we now compare to traditional deterministic
+algorithms.  Table~\ref{tab:marshak_comparison} shows timing results for the
+Marshak problem using preconditioned CG (PCG), preconditioned Richardson
+iteration (PFIX), and MCSA. The results show that the MCSA method is the most
+efficient method, although the PCG results are nearly as fast.  More
+importantly, the MCSA method clearly accelerates Richardson iteration, as
+indicated by the fact that MCSA is 42\% faster than Richardson iteration
+alone.
 \begin{table}[h!]
@@
 
-We now test the performance of the MCSA method on a large
-multi-material problem.  The multi-material problem has a dog-legged
-duct through a thick wall where the radiation flows into a thin region
-bound by a foil on one side.  The mesh and geometry are illustrated in
-Fig.~\ref{fig:multi-mat-mesh}.  A blackbody boundary flux is defined
-on the low-$x$ face at $T=0.5$~keV.  The full set of problem
-conditions is:
-
+We now test the performance of the MCSA method on a large multi-material
+problem.  The multi-material problem has a dog-legged duct through a thick
+wall where the radiation flows into a thin region bound by a foil on one side.
+The mesh and geometry are illustrated in Fig.~\ref{fig:multi-mat-mesh}.  A
+blackbody boundary flux is defined on the low-$x$ face at $T=0.5$~keV.  The
+full set of problem conditions is:
 \begin{center}
@@
 A 2D plot of the temperature at 1000~ns elapsed time is illustrated in
-Fig.~\ref{fig:multi-mat_T}.  The time-evolution of the temperature is
-shown at 4 edit points in Fig.~\ref{fig:edit-points}.  All three
-methods, PCG, PFIX, and MCSA, give numerically identical answers when
-using a stopping criterion of $\sn{1}{-8}$.
-
+Fig.~\ref{fig:multi-mat_T}.  The time-evolution of the temperature is shown at
+4 edit points in Fig.~\ref{fig:edit-points}.  All three methods, PCG, PFIX,
+and MCSA, give numerically identical answers when using a stopping criterion
+of $\sn{1}{-8}$.
 \begin{figure}[h!]
@@
 Table~\ref{tab:multimat_comparison} shows the timing results for the
-multi-material problem using PCG, PFIX, and MCSA.  The results
-correspond roughly with the results from the Marshak problem.  The
-MCSA is marginally faster than PCG and 43\% faster than PFIX.
-
+multi-material problem using PCG, PFIX, and MCSA.  The results correspond
+roughly with the results from the Marshak problem.  The MCSA is marginally
+faster than PCG and 43\% faster than PFIX.
 \begin{table}[htbp!]
@@
 
-We have presented a new Monte Carlo solution method for solving the
-discrete, time-dependent radiation diffusion equation in three
-dimensions.  The MCSA method has been shown to match results using
-standard solution techniques to arbitrary precision. Also, the new
-method has been to shown to be faster than preconditioned CG and
-Richardson iteration for the model problems presented here.
-
-While we have demonstrated marginal improvements over standard
-solution schemes in this study, significant improvements could be
-realized in fully nonlinear-consistent solutions.  In these cases, the
-MCSA method competes with GMRES, which is more costly in memory and
-time than CG.  Another area where the MCSA scheme may have advantages
-over traditional solution methods is on dentritic, or adaptive,
-meshes.  These meshes yield matrices with poor condition numbers
-because of the changing cell volumes at different refinement levels.
-A smart transport algorithm could be developed that more efficiently
-solves the residual on these types of meshes.  These topics will be
-the focus of future study.
+We have presented a new Monte Carlo solution method for solving the discrete,
+time-dependent radiation diffusion equation in three dimensions.  The MCSA
+method has been shown to match results using standard solution techniques to
+arbitrary precision. Also, the new method has been to shown to be faster than
+preconditioned CG and Richardson iteration for the model problems presented
+here.
+
+While we have demonstrated marginal improvements over standard solution
+schemes in this study, significant improvements could be realized in fully
+nonlinear-consistent solutions.  In these cases, the MCSA method competes with
+GMRES, which is more costly in memory and time than CG.  Another area where
+the MCSA scheme may have advantages over traditional solution methods is on
+dentritic, or adaptive, meshes.  These meshes yield matrices with poor
+condition numbers because of the changing cell volumes at different refinement
+levels.  A smart transport algorithm could be developed that more efficiently
+solves the residual on these types of meshes.  These topics will be the focus
+of future study.
 
@@
 
-\bibliographystyle{elsarticle-num} \bibliography{references}
+\bibliographystyle{elsarticle-num} 
+\bibliography{references}
 

### 2012-08-28 drivers/NavierStokes/main_driver.cpp
starting work on transient Poisson model

@@ -1,309 +1,0 @@
-// @HEADER
-// ***********************************************************************
-//
-//           Panzer: A partial differential equation assembly
-//       engine for strongly coupled complex multiphysics systems
-//                 Copyright (2011) Sandia Corporation
-//
-// Under the terms of Contract DE-AC04-94AL85000 with Sandia Corporation,
-// the U.S. Government retains certain rights in this software.
-//
-// Redistribution and use in source and binary forms, with or without
-// modification, are permitted provided that the following conditions are
-// met:
-//
-// 1. Redistributions of source code must retain the above copyright
-// notice, this list of conditions and the following disclaimer.
-//
-// 2. Redistributions in binary form must reproduce the above copyright
-// notice, this list of conditions and the following disclaimer in the
-// documentation and/or other materials provided with the distribution.
-//
-// 3. Neither the name of the Corporation nor the names of the
-// contributors may be used to endorse or promote products derived from
-// this software without specific prior written permission.
-//
-// THIS SOFTWARE IS PROVIDED BY SANDIA CORPORATION "AS IS" AND ANY
-// EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED TO, THE
-// IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR
-// PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL SANDIA CORPORATION OR THE
-// CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL, SPECIAL,
-// EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT LIMITED TO,
-// PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE, DATA, OR
-// PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY THEORY OF
-// LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT (INCLUDING
-// NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE OF THIS
-// SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-//
-// Questions? Contact Roger P. Pawlowski (rppawlo@sandia.gov) and
[…]

### 2012-08-28 drivers/NavierStokes/user_app_ClosureModel_Factory_TemplateBuilder.hpp

@@ -1,69 +1,0 @@
-// @HEADER
-// ***********************************************************************
-//
-//           Panzer: A partial differential equation assembly
-//       engine for strongly coupled complex multiphysics systems
-//                 Copyright (2011) Sandia Corporation
-//
-// Under the terms of Contract DE-AC04-94AL85000 with Sandia Corporation,
-// the U.S. Government retains certain rights in this software.
-//
-// Redistribution and use in source and binary forms, with or without
-// modification, are permitted provided that the following conditions are
-// met:
-//
-// 1. Redistributions of source code must retain the above copyright
-// notice, this list of conditions and the following disclaimer.
-//
-// 2. Redistributions in binary form must reproduce the above copyright
-// notice, this list of conditions and the following disclaimer in the
-// documentation and/or other materials provided with the distribution.
-//
-// 3. Neither the name of the Corporation nor the names of the
-// contributors may be used to endorse or promote products derived from
-// this software without specific prior written permission.
-//
-// THIS SOFTWARE IS PROVIDED BY SANDIA CORPORATION "AS IS" AND ANY
-// EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED TO, THE
-// IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR
-// PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL SANDIA CORPORATION OR THE
-// CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL, SPECIAL,
-// EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT LIMITED TO,
-// PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE, DATA, OR
-// PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY THEORY OF
-// LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT (INCLUDING
-// NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE OF THIS
-// SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-//
[…]

### 2012-08-28 drivers/NavierStokes/user_app_EquationSetFactory.hpp

@@ -1,87 +1,0 @@
-// @HEADER
-// ***********************************************************************
-//
-//           Panzer: A partial differential equation assembly
-//       engine for strongly coupled complex multiphysics systems
-//                 Copyright (2011) Sandia Corporation
-//
-// Under the terms of Contract DE-AC04-94AL85000 with Sandia Corporation,
-// the U.S. Government retains certain rights in this software.
-//
-// Redistribution and use in source and binary forms, with or without
-// modification, are permitted provided that the following conditions are
-// met:
-//
-// 1. Redistributions of source code must retain the above copyright
-// notice, this list of conditions and the following disclaimer.
-//
-// 2. Redistributions in binary form must reproduce the above copyright
-// notice, this list of conditions and the following disclaimer in the
-// documentation and/or other materials provided with the distribution.
-//
-// 3. Neither the name of the Corporation nor the names of the
-// contributors may be used to endorse or promote products derived from
-// this software without specific prior written permission.
-//
-// THIS SOFTWARE IS PROVIDED BY SANDIA CORPORATION "AS IS" AND ANY
-// EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED TO, THE
-// IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR
-// PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL SANDIA CORPORATION OR THE
-// CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL, SPECIAL,
-// EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT LIMITED TO,
-// PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE, DATA, OR
-// PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY THEORY OF
-// LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT (INCLUDING
-// NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE OF THIS
-// SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-//
[…]

### 2012-08-28 drivers/NavierStokes/user_app_EquationSet_Energy.hpp

@@ -1,78 +1,0 @@
-// @HEADER
-// ***********************************************************************
-//
-//           Panzer: A partial differential equation assembly
-//       engine for strongly coupled complex multiphysics systems
-//                 Copyright (2011) Sandia Corporation
-//
-// Under the terms of Contract DE-AC04-94AL85000 with Sandia Corporation,
-// the U.S. Government retains certain rights in this software.
-//
-// Redistribution and use in source and binary forms, with or without
-// modification, are permitted provided that the following conditions are
-// met:
-//
-// 1. Redistributions of source code must retain the above copyright
-// notice, this list of conditions and the following disclaimer.
-//
-// 2. Redistributions in binary form must reproduce the above copyright
-// notice, this list of conditions and the following disclaimer in the
-// documentation and/or other materials provided with the distribution.
-//
-// 3. Neither the name of the Corporation nor the names of the
-// contributors may be used to endorse or promote products derived from
-// this software without specific prior written permission.
-//
-// THIS SOFTWARE IS PROVIDED BY SANDIA CORPORATION "AS IS" AND ANY
-// EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED TO, THE
-// IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR
-// PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL SANDIA CORPORATION OR THE
-// CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL, SPECIAL,
-// EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT LIMITED TO,
-// PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE, DATA, OR
-// PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY THEORY OF
-// LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT (INCLUDING
-// NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE OF THIS
-// SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-//
[…]

### 2012-08-28 drivers/NavierStokes/user_app_NOXObserver_EpetraToExodus.hpp

@@ -1,245 +1,0 @@
-// @HEADER
-// ***********************************************************************
-//
-//           Panzer: A partial differential equation assembly
-//       engine for strongly coupled complex multiphysics systems
-//                 Copyright (2011) Sandia Corporation
-//
-// Under the terms of Contract DE-AC04-94AL85000 with Sandia Corporation,
-// the U.S. Government retains certain rights in this software.
-//
-// Redistribution and use in source and binary forms, with or without
-// modification, are permitted provided that the following conditions are
-// met:
-//
-// 1. Redistributions of source code must retain the above copyright
-// notice, this list of conditions and the following disclaimer.
-//
-// 2. Redistributions in binary form must reproduce the above copyright
-// notice, this list of conditions and the following disclaimer in the
-// documentation and/or other materials provided with the distribution.
-//
-// 3. Neither the name of the Corporation nor the names of the
-// contributors may be used to endorse or promote products derived from
-// this software without specific prior written permission.
-//
-// THIS SOFTWARE IS PROVIDED BY SANDIA CORPORATION "AS IS" AND ANY
-// EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED TO, THE
-// IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR
-// PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL SANDIA CORPORATION OR THE
-// CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL, SPECIAL,
-// EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT LIMITED TO,
-// PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE, DATA, OR
-// PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY THEORY OF
-// LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT (INCLUDING
-// NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE OF THIS
-// SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-//
[…]

### 2012-08-28 drivers/TransientPoisson/TransientPoisson_ClosureModelFactory.hpp

@@ -1,0 +1,63 @@
+//---------------------------------------------------------------------------//
+/*!
+ * \file TransientPoisson_ClosureModelFactory.hpp
+ * \author Stuart R. Slattery
+ * \brief Closure model factory for the transient Poisson problem.
+ */
+//---------------------------------------------------------------------------//
+
+#ifndef TRANSIENTPOISSON_CLOSUREMODEL_HPP
+#define TRANSIENTPOISSON_CLOSUREMODEL_HPP
+
+#include <string>
+#include <vector>
+
+#include <Teuchos_RCP.hpp>
+#include <Teuchos_ParamterList.hpp>
+
+#include <Phalanx_Evaluator.hpp>
+#include <Phalanx_FieldManager.hpp>
+
+#include <Panzer_ClosureModel_Factory.hpp>
+#include <Panzer_InputEquationSet.hpp>
+#include <Panzer_Traits.hpp>
+
+//---------------------------------------------------------------------------//
+
+namespace Chimera
+{
+namespace TransientPoisson
+{
+
+template<typename EvaluationType>
+class ClosureModelFactory : public panzer::ClosureModelFactory<EvaluationType>
+{
+  public:
+
+    // Factory method.
+   Teuchos::RCP<std::vector<Teuchos::RCP<PHX::Evaluator<panzer::Traits> > > >
+   buildClosureModels( const std::string& model_id,
+		       const panzer::InputEquationSet& equation_set,
+		       const Teuchos::ParameterList& models,
+		       const Teuchos::ParameterList& default_params,
+		       const Teuchos::ParameterList& user_data,
+		       const Teuchos::RCP<panzer::GlobalData>& global_data,
+		       PHX::FieldManager<panzer::Traits>& field_manager ) const;
+
+};
+
+} // end namepsace TransientPoisson
+} // end namespace Chimera
+
+//---------------------------------------------------------------------------//
+// Template includes.
+//---------------------------------------------------------------------------//
+
+#include "TransientPoisson_ClosureModelFactory_impl.hpp"
+
+#endif // end TRANSIENTPOISSON_CLOSUREMODEL_HPP
+
+//---------------------------------------------------------------------------//
+// end TransientPoisson_ClosureModelFactory.hpp
+//---------------------------------------------------------------------------//
+

### 2012-08-28 drivers/TransientPoisson/TransientPoisson_ClosureModelFactory_TemplateBuilder.hpp

@@ -1,0 +1,45 @@
+//---------------------------------------------------------------------------//
+/*!
+ * \file TransientPoisson_ClosureModelFactory_TemplateBuilder.hpp
+ * \author Stuart R. Slattery
+ * \brief ClosureModelFactory builder.
+ */
+//---------------------------------------------------------------------------//
+
+#ifndef TRANSIENTPOISSON_CLOSUREMODEL_BUILDER_HPP
+#define TRANSIENTPOISSON_CLOSUREMODEL_BUILDER_HPP
+
+#include "TransientPoisson_ClosureModelFactory.hpp"
+
+#include <Teuchos_RCP.hpp>
+
+#include <Panzer_Base.hpp>
+
+namespace Chimera
+{
+namespace TransientPoisson
+{
+
+class ClosureModelFactory_TemplateBuilder
+{
+  public:
+
+    // Builder method.
+    template<typename EvaluationType>
+    Teuchos::RCP<panzer::Base> build() const
+    {
+	return Teuchos::rcp( static_cast<panzer::Base*>( 
+				 new ClosureModelFactory<EvaluationType> ) );
+    }
+
+};
+
+} // end namespace TransientPoisson
+} // end namespace Chimera
+
+#endif // end TRANSIENTPOISSON_CLOSUREMODEL_BUILDER_HPP
+
+//---------------------------------------------------------------------------//
+// end TransientPoisson_ClosureModelFactory_TemplateBuilder.hpp
+//---------------------------------------------------------------------------//
+

### 2012-08-28 drivers/TransientPoisson/TransientPoisson_ClosureModelFactory_impl.hpp

@@ -1,0 +1,114 @@
+//---------------------------------------------------------------------------//
+/*!
+ * \file TransientPoisson_ClosureModelFactory_impl.hpp
+ * \author Stuart R. Slattery
+ * \brief Closure model factory for the transient Poisson problem.
+ */
+//---------------------------------------------------------------------------//
+
+#ifndef TRANSIENTPOISSON_CLOSUREMODEL_IMPL_HPP
+#define TRANSIENTPOISSON_CLOSUREMODEL_IMPL_HPP
+
+#include <Panzer_Constant.hpp>
+
+namespace Chimera
+{
+namespace TransientPoisson
+{
+//---------------------------------------------------------------------------//
+/*!
+ * \brief Closure model factory method.
+ */
+template<typename EvaluationType>
+Teuchos::RCP<std::vector<Teuchos::RCP<PHX::Evaluator<panzer::Traits> > > > 
+ClosureModelFactory<EvaluationType>::buildClosureModels( 
+    const std::string& model_id,
+    const panzer::InputEquationSet& equation_set,
+    const Teuchos::ParameterList& models,
+    const Teuchos::ParameterList& default_params,
+    const Teuchos::ParameterList& user_data,
+    const Teuchos::RCP<panzer::GlobalData>& global_data,
+    PHX::FieldManager<panzer::Traits>& field_manager ) const
+{
+    // Setup a vector for the closure evaluators.
+    Teuchos::RCP<std::vector<Teuchos::RCP<PHX::Evaluator<panzer::Traits> > > >  =
+	Teuchos::rcp( 
+	    new std::vector<Teuchos::RCP<PHX::Evaluator<panzer::Traits> > > );
+
+    // Check that this model id is valid.
+    if ( !models.isSublist( model_id ) )
+    {
+	models.print( std::cout );
+	std::stringstream msg;
+	msg << "Falied to find requested model, \"" << model_id 
+	    << "\", for equation set:\n" << std::endl;
+	TEUCHOS_TEST_FOR_EXCEPTION( !models.isSublist(model_id), 
+				    std::logic_error, msg.str() );
+    }
+
+    // Loop through the models and build evaluators.
+    const Teuchos::ParameterList& my_models = models.sublist( model_id );
+
+    for ( Teuchos::ParameterList::ConstIterator model_it = my_models.begin();
+	  model_it != my_models.end(); ++model_it )
+    {
+	bool found = false;
+
+	const std::string key = model_it->first;
[…]

### 2012-08-28 doc/jcomp2012/jcomp2012.tex
Added reviewer comments to paper: Stuart, the Adjoint_Direct_Convergence.pdf figure needs axis labels.

@@
 
-    We present a novel synthetic-acceleration based Monte Carlo method
-    for solving the equilibrium thermal radiation diffusion equation
-    in three dimensions. The algorithm performance is compared against
-    traditional solution techniques using a Marshak benchmark problem
-    and a more complex multiple material problem. Our results show
-    that not only can our Monte Carlo method be an effective solver
-    for sparse matrix systems, but also that it performs competitively
-    with deterministic methods including preconditioned Conjugate
-    Gradient while producing numerically identical results. We also
-    discuss various aspects of preconditioning the method and its
-    general applicability to broader classes of problems.
+    We present a novel synthetic-acceleration based Monte Carlo method for
+    solving the equilibrium thermal radiation diffusion equation in three
+    dimensions. The algorithm performance is compared against traditional
+    solution techniques using a Marshak benchmark problem and a more complex
+    multiple material problem. Our results show that not only can our Monte
+    Carlo method be an effective solver for sparse matrix systems, but also
+    that it performs competitively with deterministic methods including
+    preconditioned conjugate gradient while producing numerically identical
+    results. We also discuss various aspects of preconditioning the method and
+    its general applicability to broader classes of problems.
 
@@
 
-In previous work~\cite{evans_2003}, we developed a residual Monte
-Carlo method that was an application of Halton's Sequential Monte
-Carlo method~\cite{halton_1962,halton_1994} to the one-dimensional
-equilibrium (1T) thermal radiation diffusion equation.  At that time,
-Halton's method was not widely known in the general computational
-physics community.  While extending our method to three dimensions, we
-have realized that the Sequential Monte Carlo method is actually a
-variant of iterative refinement cast as a residual method. We have
-developed a new method in which Monte Carlo is used to accelerate a
-fixed-point (Richardson) iteration. This algorithm surpasses our
-previous one dimensional method and allows for extension to more
-general solution techniques.
-
-The purpose of this study is to demonstrate an efficient Monte Carlo
-solution method for three dimensional (3D), time-dependent, discrete
-systems. In this work, we extend our previous work in the following
-manner:
+In previous work~\cite{evans_2003}, we developed a residual Monte Carlo method
+that was an application of Halton's Sequential Monte Carlo
+method~\cite{halton_1962,halton_1994} to the one-dimensional equilibrium (1T)
+thermal radiation diffusion equation.  At that time, Halton's method was not
+widely known in the general computational physics community.  While extending
+our method to three dimensions, we have realized that the Sequential Monte
+Carlo method is actually a variant of iterative refinement cast as a residual
+method. We have developed a new method in which Monte Carlo is used to
+accelerate a fixed-point (Richardson) iteration. This algorithm surpasses our
+previous one dimensional method and allows for extension to more general
+solution techniques.
+
+The purpose of this study is to demonstrate an efficient Monte Carlo solution
+method for three dimensional (3D), time-dependent, discrete systems. In this
+work, our previous investigations are extended in the following manner:
 \begin{itemize}
-\item We develop a rigorous synthetic-acceleration Monte Carlo method
-  for solving sparse matrix systems.
-\item We apply this method to the 1T thermal radiation diffusion
-  equation in three dimensions.
-\item We compare numerical and performance results of the Monte Carlo
+\item we develop a rigorous synthetic-acceleration Monte Carlo method
+  for solving sparse matrix systems;
+\item we apply this method to the 1T thermal radiation diffusion
+  equation in three dimensions;
+\item we compare numerical and performance results of the Monte Carlo
   method against traditional deterministic solution methods.
@@
 
-In \S~\ref{sec:monte-carlo-matrix} we review Monte Carlo methods for
-solving matrix systems. In \S~\ref{sec:iter-refin-monte} we discuss
-our new synthetic-acceleration scheme and connect it with previous
-work by Halton~\cite{halton_1994} and others~\cite{evans_2003}.  The
-1T radiation diffusion equation and its discrete form will be derived
-in \S~\ref{sec:discr-form-radi}.  We discuss using Monte Carlo
-Synthetic-Accleration to solve the discrete radiation diffusion
-equation in \S~\ref{sec:solut-radi-diff}.  Results from two 3D test
-problems are given in \S~\ref{sec:results}.  We finish with
-conclusions and ideas for future work in \S~\ref{sec:conclusions}.
+In \S~\ref{sec:monte-carlo-matrix} we review Monte Carlo methods for solving
+matrix systems, and we discuss our new synthetic-acceleration scheme and
+connect it with previous work by Halton~\cite{halton_1994} and
+others~\cite{evans_2003}.  The 1T radiation diffusion equation and its
+discrete form will be derived in \S~\ref{sec:discr-form-radi}.  We discuss
+using Monte Carlo Synthetic-Accleration to solve the discrete radiation
+diffusion equation in \S~\ref{sec:solut-radi-diff}.  Results from two 3D test
+problems are given in \S~\ref{sec:results}.  We finish with conclusions and
+ideas for future work in \S~\ref{sec:conclusions}.
 
@@
 \label{sec:background}
+
 To establish the mathematical framework for Monte Carlo linear
@@
 \end{equation}
-Equation~(\ref{eq:richardson}) will converge for all $\vb \in
-\mathcal{R}^{N}$ when $\rho(\vH) < 1$~\cite{kelley_1995}.  All of the
-Monte Carlo methods that are described in \S\S~\ref{sec:direct-method}
-through \ref{sec:iter-refin-monte} rely on estimating the terms in
+Equation~(\ref{eq:richardson}) will converge for all $\vb,\vx^0 \in
+\mathcal{R}^{N}$ when $\rho(\vH) < 1$~\cite{kelley_1995}.  All of the Monte
+Carlo methods that are described in \S\S~\ref{sec:direct-method} through
+\ref{sec:iter-refin-monte} rely on estimating the terms in
 Eq.~(\ref{eq:neumann_series}) through random walks.
@@
 between states, for example $i\rightarrow j$, the weight is multiplied
-by the transition factor.
+by the transition factor
 \begin{equation}
@@
 Eq.~(\ref{eq:synthetic-residual-solve}), we can instead use the above ideas to
-create a new iterative scheme by using the Monte Carlo method to estimate the
-solution error. With this in hand, the Fixed-Point Monte Carlo
+create a new iterative scheme that applies the Monte Carlo method to estimate
+the solution error. With this in hand, the Fixed-Point Monte Carlo
 Synthetic-Acceleration (MCSA) method can be defined as follows:
@@
   \label{eq:partial-T}
-  \partial T = \frac{1}{4acT^{3}}\partial\phi\:.
+  \pder{T}{t} = \frac{1}{4acT^{3}}\pder{\phi}{t}\:.
 \end{equation}
@@
   \begin{split}
-    (F_{i+1/2} - F_{i-1/2})\Dj\Dk + (F_{j+1/2} &- F_{j-1/2})\Di\Dk +
+    (F_{i+1/2} - F_{i-1/2})\Dj\Dk &+ (F_{j+1/2} - F_{j-1/2})\Di\Dk +
     (F_{k+1/2}\\ &- F_{k-1/2})\Di\Dj + \sign_{ijk}\phi_{ijk}V_{ijk} =
@@
 \begin{itemize}
-\item It reduces the spectral radius such that
-  $\rho(\ve{M}^{-1}\ve{D}) < \rho(\ve{D})$ and
-  $\rho(\ve{M}^{-1}\ve{D}) < 1$.  Thus, convergence is guaranteed and
-  fewer iterations are required resulting in accelerated convergence.
+\item For diagonally dominant systems like Eq.~(\ref{eq:diffusion-matrix}), it
+  reduces the spectral radius such that $\rho(\ve{M}^{-1}\ve{D}) <
+  \rho(\ve{D})$ and $\rho(\ve{M}^{-1}\ve{D}) < 1$ \cite{golub}.  Thus,
+  convergence is guaranteed and fewer iterations are required resulting in
+  accelerated convergence.
 \item The diagonal elements of the preconditioned iteration matrix are
@@
   \centerline{ \includegraphics[width=5in,clip]{mesh.pdf}}
-  \caption{Multi-material problem mesh.  The duct region is blue.  The
-    wall is red.  The foil is green.  The mesh is $60\times 60\times
-    60$.}
+  \caption{Multi-material problem mesh.  The duct region is blue.  The wall is
+    red.  The foil is green.  The mesh is $60\times 60\times 60$.}
   \label{fig:multi-mat-mesh}
@@
   \centerline{ \includegraphics[width=5in,clip]{T_multi_mat.pdf}}
-  \caption{ Two-dimensional plot of the temperature at 1000~ns on a
-    cut-plane positioned at the mid-point of the $z$-axis.  The
-    positions of the edit points where the time-evolution of the
-    temperature is plotted in Fig.~\ref{fig:edit-points} are indicated
-    by the black disks.}
+  \caption{ Two-dimensional plot of the temperature in keV at 1000~ns on a
+    cut-plane positioned at the mid-point of the $z$-axis.  The positions of
+    the edit points where the time-evolution of the temperature is plotted in
+    Fig.~\ref{fig:edit-points} are indicated by the black disks.}
   \label{fig:multi-mat_T}
@@
   \begin{center}
-    \begin{tabular}{cc}
-      \subfigure[Edit
-        points]{\includegraphics[width=2in,clip]{points.pdf}} &
-      \subfigure[Temperature]{\includegraphics[width=4in,clip]{multi_mat.pdf}}
-    \end{tabular}
+    \includegraphics[width=5in,clip]{edit_points}
   \end{center}
-  \caption{(a) Edit points and (b) time-evolution of the temperature
-    at each point.  All points are centered in the $z$-axis.}
+  \caption{(a) Edit points and (b) time-evolution of the temperature at each
+    point.  All points are centered in the $z$-axis.}
   \label{fig:edit-points}

### 2012-08-29 doc/jcomp2012/jcomp2012.tex
Updated paper.

@@
     solving the equilibrium thermal radiation diffusion equation in three
-    dimensions. The algorithm performance is compared against traditional
-    solution techniques using a Marshak benchmark problem and a more complex
-    multiple material problem. Our results show that not only can our Monte
-    Carlo method be an effective solver for sparse matrix systems, but also
-    that it performs competitively with deterministic methods including
-    preconditioned conjugate gradient while producing numerically identical
-    results. We also discuss various aspects of preconditioning the method and
-    its general applicability to broader classes of problems.
+    spatial dimensions. The algorithm performance is compared against
+    traditional solution techniques using a Marshak benchmark problem and a
+    more complex multiple material problem. Our results show that our Monte
+    Carlo method is an effective solver for sparse matrix systems. For
+    solutions converged to the same tolerance, it performs competitively with
+    deterministic methods including preconditioned conjugate gradient. We also
+    discuss various aspects of preconditioning the method and its general
+    applicability to broader classes of problems.
 
@@
 that was an application of Halton's Sequential Monte Carlo
-method~\cite{halton_1962,halton_1994} to the one-dimensional equilibrium (1T)
+method~\cite{halton_1962,halton_1994} to the one-dimensional equilibrium
 thermal radiation diffusion equation.  At that time, Halton's method was not
 widely known in the general computational physics community.  While extending
-our method to three dimensions, we have realized that the Sequential Monte
-Carlo method is actually a variant of iterative refinement cast as a residual
-method. We have developed a new method in which Monte Carlo is used to
-accelerate a fixed-point (Richardson) iteration. This algorithm surpasses our
-previous one dimensional method and allows for extension to more general
+our method to three spatial dimensions, we have realized that the Sequential
+Monte Carlo method is actually a variant of iterative refinement cast as a
+residual method. We have developed a new method in which Monte Carlo is used
+to accelerate a fixed-point (Richardson) iteration. This algorithm surpasses
+our previous one dimensional method and allows for extension to more general
 solution techniques.
@@
   for solving sparse matrix systems;
-\item we apply this method to the 1T thermal radiation diffusion
-  equation in three dimensions;
+\item we apply this method to the one-temperature (1T) thermal radiation
+  diffusion equation in three dimensions;
 \item we compare numerical and performance results of the Monte Carlo
@@
 \end{enumerate}
-For clarity, we rewrite the iteration scheme in
-Eq.~(\ref{eq:MCSA-iteration}) using the diffusion operator notation
-from Eq.~(\ref{eq:preconditioned-diffusion}),
+For clarity, we rewrite the iteration scheme in Eq.~(\ref{eq:MCSA-iteration})
+using the diffusion operator notation from
+Eq.~(\ref{eq:preconditioned-diffusion}),
 \begin{align*}
@@
   \ve{M}^{-1}\ve{q}\:, \quad & (\text{fixed-point iteration})\\
+  %%
   \vr^{l+1/2} = \ve{M}^{-1}\ve{q} - \ve{M}^{-1}\ve{D}\bphi^{l+1/2}\:,
   \quad & (\text{compute residual})\\ 
-  \ve{M}^{-1}\ve{D}\delta\bphi^{l+1/2} = \vr^{l+1/2}\:, \quad&
+  %%
+  \ve{M}^{-1}\hat{\ve{D}}\delta\bphi^{l+1/2} = \vr^{l+1/2}\:, \quad&
   (\text{estimate $\delta\phi$ using adjoint Monte Carlo method})\\
+  %%
   \bphi^{l+1} = \bphi^{l+1/2} + \delta\bphi^{l+1/2}\:. \quad&
@@
 As described in \S~\ref{sec:iter-refin-monte}, the adjoint Monte Carlo method
-only approximately inverts $\ve{M}^{-1}\ve{D}$.  We now describe each of these
-in turn.
+only approximately inverts $\ve{M}^{-1}\hat{\ve{D}}$ as indicated by the
+$\hat{\cdot}$ notation.  We now describe each of these in turn.
 
@@
 \begin{equation}
-  \ve{M}^{-1}\ve{D}\delta\bphi^{l+1/2} = \vr^{l+1/2}\:.
+  \ve{M}^{-1}\hat{\ve{D}}\delta\bphi^{l+1/2} = \vr^{l+1/2}\:.
 \end{equation}
 As noted above, the residual acts as the source for this simulation, and the
-operator $\ve{M}^{-1}\ve{D}$ is implicitly defined in
+operator $\ve{M}^{-1}\hat{\ve{D}}$ is implicitly defined in
 Eq.~(\ref{eq:preconditioned-discrete-diffusion}).  There are two Monte Carlo
@@
 \begin{align*}
-  \phi_1 &= \frac{\sigm_{7\rightarrow
-      1}}{\sigma_{T\,1}}\frac{\Delta_7}{\Delta_1}\phi_7 +
-  \ldots\:,\\ \phi_2 &= \frac{\sigp_{7\rightarrow 2}}
+  \phi_1 &= \frac{\sigm_{7\rightarrow 1}}{\sigma_{T\,1}}
+  \frac{\Delta_7}{\Delta_1}\phi_7 + \ldots\:,\\ 
+  %%
+  \phi_2 &= \frac{\sigp_{7\rightarrow 2}}
   {\sigma_{T\,2}}\frac{\Delta_7}{\Delta_2}\phi_7 + \ldots\:,\\
-      \phi_3 &= \frac{\sigm_{7\rightarrow
-          3}}{\sigma_{T\,3}}\frac{\Delta_7}{\Delta_3}\phi_7 +
-      \ldots\:,\\  \phi_4 &= \frac{\sigp_{7\rightarrow
-      4}}{\sigma_{T\,4}}\frac{\Delta_7}{\Delta_4}\phi_7 +
-          \ldots\:,\\  \phi_5 &= \frac{\sigm_{7\rightarrow
-          5}}{\sigma_{T\,5}}\frac{\Delta_7}{\Delta_5}\phi_7 +
-              \ldots\:,\\  \phi_6 &= \frac{\sigp_{7\rightarrow
-              6}}{\sigma_{T\,6}}\frac{\Delta_7}{\Delta_6}\phi_7 +
-                  \ldots\:,\\ 
+  %%
+  \phi_3 &= \frac{\sigm_{7\rightarrow 3}}{\sigma_{T\,3}}
+  \frac{\Delta_7}{\Delta_3}\phi_7 + \ldots\:,\\ 
+  %%
+  \phi_4 &= \frac{\sigp_{7\rightarrow 4}}{\sigma_{T\,4}}
+  \frac{\Delta_7}{\Delta_4}\phi_7 + \ldots\:,\\
+  %%
+  \phi_5 &= \frac{\sigm_{7\rightarrow 5}}{\sigma_{T\,5}}
+  \frac{\Delta_7}{\Delta_5}\phi_7 + \ldots\:,\\
+  %%
+  \phi_6 &= \frac{\sigp_{7\rightarrow 6}}{\sigma_{T\,6}}
+  \frac{\Delta_7}{\Delta_6}\phi_7 + \ldots\:,\\ 
 \end{align*} 
@@
 fixed-point (Richardson) iteration (PFIX).  The PCG method uses right-left
-preconditioning with $\ve{M}=\ve{L}\ve{L}^{-1}$.  The PFIX scheme is used to
-solve Eq.~(\ref{eq:preconditioned-diffusion}), which is identical to the
-system being solved with MCSA.  We examine each method on two 3D problems: a
-Marshak wave problem and a multi-material duct problem.
+preconditioning with $\ve{M}=\ve{L}\ve{L}^{-1}$, where $\ve{L} = \ve{L}^{T} =
+\sqrt{\ve{M}}$.  Naturally, more powerful preconditioning methods could be
+used; however, this provides a fair comparison with the preconditioning used
+in the MCSA method.  Different precondioning strategies within MCSA will be
+the focus of a future paper.  The PFIX scheme is used to solve
+Eq.~(\ref{eq:preconditioned-diffusion}), which is identical to the system
+being solved with MCSA.  We examine each method on two 3D problems: a Marshak
+wave problem and a multi-material duct problem.
 
@@
   \centerline{ \includegraphics[width=5in,clip]{marshak_10.pdf}}
-  \caption{ Solution of the Marshak wave problem at $t=100$~ns.  The
-    analytic solution is compared to the MCSA solution.  Note that the
-    MCSA, PCG, and PFIXED methods all give numerically identical
-    answers.}
+  \caption{ Solution of the Marshak wave problem at $t=100$~ns.  The analytic
+    solution is compared to the MCSA solution.  The MCSA, PCG, and PFIXED
+    methods are all converged to the same tolerance and therefore give
+    numerically equivalent answers.}
   \label{fig:marshak_10}

### 2012-09-04 drivers/TransientPoisson/TransientPoisson_BCStrategy_Dirichlet_Constant.hpp
finished core implementation of transient poisson problem. main driver is next.

@@ -9,13 +9,56 @@
 #ifndef TRANSIENTPOISSON_BCSTRATEGY_DIRICHLET_CONSTANT_HPP
 #define TRANSIENTPOISSON_BCSTRATEGY_DIRICHLET_CONSTANT_HPP
 
+#include <string>
+
+#include <Teuchos_RCP.hpp>
+#include <Teuchos_ParameterList.hpp>
+
+#include <Panzer_BCStrategy_Dirichlet_DefaultImpl.hpp>
+#include <Panzer_Traits.hpp>
+#include <Panzer_PureBasis.hpp>
+#include <Panzer_PhysicsBlock.hpp>
+
+#include "Phalanx_FieldManager.hpp"
+
 namespace Chimera
 {
 namespace TransientPoisson
 {
 
+template<typename EvaluationType>
+class BCStrategy_Dirichlet_Constant :
+	public panzer::BCStrategy_Dirichlet_DefaultImpl<EvaluationType>
+{
+  public:
+    
+    BCStrategy_Dirichlet_Constant( const panzer::BC& bc,
+				   const Teuchos::RCP<panzer::GlobalData>& global_data );
+
+    void setup( const panzer::PhysicsBlock& side_pb,
+		const Teuchos::ParameterList& user_data );
+
+    void buildAndRegisterEvaluators( PHX::FieldManager<panzer::Traits>& fm,
+				     cont panzer::PhysicsBlock& pb,
+				     const panzer::ClosureModelFactory_TemplateManager<panzer::Traits>& factory,
+				     const Teuchos::ParameterList& models,
+				     const Teuchos::ParameterList& user_data ) const;
+
+    // Data.
+    std::string residual_name;
+    Teuchos::RCP<panzer::PureBasis> basis;
+};
+
 } // end namespace TransientPoisson
 } // end namespace Chimera
+
+//---------------------------------------------------------------------------//
+// Template includes.
+//---------------------------------------------------------------------------//
+
+#include "TransientPoisson_BCStrategy_Dirichlet_Constant_impl.hpp"
+
+//---------------------------------------------------------------------------//
 
 #endif // end TRANSIENTPOISSON_BCSTRATEGY_DIRICHLET_CONSTANT_HPP
 

### 2012-09-04 drivers/TransientPoisson/TransientPoisson_BCStrategy_Dirichlet_Constant_impl.hpp

@@ -1,0 +1,111 @@
+//---------------------------------------------------------------------------//
+/*!
+ * \file TransientPoisson_BCStrategy_Dirichlet_Constant_impl.hpp
+ * \author Stuart R. Slattery
+ * \brief Constant Dirichlet boundary condition implementation.
+ */
+//---------------------------------------------------------------------------//
+
+#ifndef TRANSIENTPOISSON_BCSTRATEGY_DIRICHLET_CONSTANT_IMPL_HPP
+#define TRANSIENTPOISSON_BCSTRATEGY_DIRICHLET_CONSTANT_IMPL_HPP
+
+#include <vector>
+
+#include <Teuchos_Assert.hpp>
+
+namespace Chimera
+{
+namespace TransientPoisson
+{
+//---------------------------------------------------------------------------//
+/*!
+ * \brief Constructor.
+ */
+template<typename EvaluationType>
+BCStrategy_Dirichlet_Constant<EvaluationType>::BCStrategy_Dirichlet_Constant( 
+    const panzer::BC& bc, const Teuchos::RCP<panzer::GlobalData>& global_data )
+    : panzer::BCStrategy_Dirichlet_DefaultImpl<EvaluationType>(bc,global_data)
+{
+    TEUCHOS_ASSERT( this->m_bc.strategy() == "Constant" );
+}
+
+//---------------------------------------------------------------------------//
+/*!
+ * \brief Setup the base class data.
+ */
+template<typename EvaluationType>
+void BCStrategy_Dirichlet_Constant<EvaluationType>::setup( 
+    const panzer::PhysicsBlock& side_pb, 
+    const Teuchos::ParameterList& user_data )
+{
+    // Get the dofs for the residual.
+    this->required_dof_names.push_back( this->m_bc.equationSetName() );
+
+    // Get the name of the residual.
+    this->residual_name = "Residual_" + this->m_bc.identifier();
+
+    // Map the residual to the dofs.
+    this->residual_to_dof_names_map[residual_name] = this->m_bc.equationSetName();
+
+    // Map the residual to the target.
+    this->residual_to_target_field_map[residual_name] = 
+	"Constant_" + this->m_bc.equationSetName();
+
+    // Get the basis for the dofs.
+    const std::vector<std::pair<std::string,Teuchos::RCP<panzer::PureBasis> > >&
+	dofs = side_pb.getProvidedDOFs();
[…]

### 2012-09-04 drivers/TransientPoisson/TransientPoisson_ClosureModelFactory.hpp

@@ -43,7 +43,6 @@ class ClosureModelFactory : public panzer::ClosureModelFactory<EvaluationType>
 		       const Teuchos::ParameterList& user_data,
 		       const Teuchos::RCP<panzer::GlobalData>& global_data,
 		       PHX::FieldManager<panzer::Traits>& field_manager ) const;
-
 };
 
 } // end namepsace TransientPoisson

### 2012-09-04 drivers/TransientPoisson/TransientPoisson_ClosureModelFactory_impl.hpp

@@ -31,8 +31,8 @@ ClosureModelFactory<EvaluationType>::buildClosureModels(
     PHX::FieldManager<panzer::Traits>& field_manager ) const
 {
     // Setup a vector for the closure evaluators.
-    Teuchos::RCP<std::vector<Teuchos::RCP<PHX::Evaluator<panzer::Traits> > > >  =
-	Teuchos::rcp( 
+    Teuchos::RCP<std::vector<Teuchos::RCP<PHX::Evaluator<panzer::Traits> > > >
+	evaluators = Teuchos::rcp( 
 	    new std::vector<Teuchos::RCP<PHX::Evaluator<panzer::Traits> > > );
 
     // Check that this model id is valid.
@@ -48,7 +48,6 @@ ClosureModelFactory<EvaluationType>::buildClosureModels(
 
     // Loop through the models and build evaluators.
     const Teuchos::ParameterList& my_models = models.sublist( model_id );
-
     for ( Teuchos::ParameterList::ConstIterator model_it = my_models.begin();
 	  model_it != my_models.end(); ++model_it )
     {

### 2012-09-04 drivers/TransientPoisson/TransientPoisson_EquationSet_impl.hpp

@@ -76,7 +76,7 @@ buildAndRegisterEquationSetEvaluators(
 
 	Teuchos::RCP< PHX::Evaluator<panzer::Traits> > transient_op = 
 	    Teuchos::rcp( 
-		new panzer::Integrator_BasisTimesScalar<EvaluationType,panzer::Traits>( transient_op ) );
+		new panzer::Integrator_BasisTimesScalar<EvaluationType,panzer::Traits>( plist ) );
 
 	field_manager.registerEvaluator<EvaluationType>( transient_op );
     }
@@ -95,7 +95,7 @@ buildAndRegisterEquationSetEvaluators(
 
 	Teuchos::RCP< PHX::Evaluator<panzer::Traits> > diffusion_op =
 	    Teuchos::rcp(
-		new panzer::Integrator_GradBasisDotVector<EvaluationType,panzer::Traits( diffusion_op ) );
+		new panzer::Integrator_GradBasisDotVector<EvaluationType,panzer::Traits( plist ) );
 
 	field_manager.registerEvaluator<EvaluationType>( diffusion_op );
     }
@@ -112,23 +112,23 @@ buildAndRegisterEquationSetEvaluators(
 
 	Teuchos::RCP< PHX::Evaluator<panzer::Traits> > source_op =
 	    Teuchos::rcp(
-		new panzer::Integrator_BasisTimesScalar<EvaluationType,panzer::Traits>( source_op ) );
+		new panzer::Integrator_BasisTimesScalar<EvaluationType,panzer::Traits>( plist ) );
 
 	field_manager.registerEvaluator<EvaluationType>( source_op );
     }
 
-    // Build the over residual for the Poisson equation by summing the above
-    // residuals.
+    // Build the overall residual for the Poisson equation by summing the
+    // above residuals.
     {
 	Teuchos::RCP< std::vector<std::string> residuals_to_sum = 
 	    Teuchos::rcp( new std::vector<std::string> );
 
 	if ( this->m_build_transient_support )
 	{
-	    residuals_to_sum.push_back( "RESIDUAL_TEMPERATURE_TRANSIENT_OP" );
+	    residuals_to_sum->push_back( "RESIDUAL_TEMPERATURE_TRANSIENT_OP" );
 	}
-	residuals_to_sum.push_back( "RESIDUAL_TEMPERATURE_DIFFUSION_OP" );
[…]

### 2012-09-04 drivers/TransientPoisson/TransientPoisson_main.cpp

@@ -1,0 +1,12 @@
+//---------------------------------------------------------------------------//
+/*!
+ * \file TransientPoisson_main.cpp
+ * \author Stuart R. Slattery
+ * \brief Driver for transient poisson problem.
+ */
+//---------------------------------------------------------------------------//
+
+//---------------------------------------------------------------------------//
+// end TransientPoisson_main.cpp
+//---------------------------------------------------------------------------//
+

### 2012-09-04 drivers/TransientPoisson/main.cpp

@@ -1,357 +1,0 @@
-// @HEADER
-// ***********************************************************************
-//
-//           Panzer: A partial differential equation assembly
-//       engine for strongly coupled complex multiphysics systems
-//                 Copyright (2011) Sandia Corporation
-//
-// Under the terms of Contract DE-AC04-94AL85000 with Sandia Corporation,
-// the U.S. Government retains certain rights in this software.
-//
-// Redistribution and use in source and binary forms, with or without
-// modification, are permitted provided that the following conditions are
-// met:
-//
-// 1. Redistributions of source code must retain the above copyright
-// notice, this list of conditions and the following disclaimer.
-//
-// 2. Redistributions in binary form must reproduce the above copyright
-// notice, this list of conditions and the following disclaimer in the
-// documentation and/or other materials provided with the distribution.
-//
-// 3. Neither the name of the Corporation nor the names of the
-// contributors may be used to endorse or promote products derived from
-// this software without specific prior written permission.
-//
-// THIS SOFTWARE IS PROVIDED BY SANDIA CORPORATION "AS IS" AND ANY
-// EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT LIMITED TO, THE
-// IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A PARTICULAR
-// PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL SANDIA CORPORATION OR THE
-// CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL, SPECIAL,
-// EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT LIMITED TO,
-// PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE, DATA, OR
-// PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY THEORY OF
-// LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT (INCLUDING
-// NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE OF THIS
-// SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
-//
-// Questions? Contact Roger P. Pawlowski (rppawlo@sandia.gov) and
-// Eric C. Cyr (eccyr@sandia.gov)
[…]

### 2012-09-06 doc/jcomp2012/jcomp2012.tex
Fixed preconditioning matrix.
Added technical editors comments.

@@
 
-    We present a novel synthetic-acceleration based Monte Carlo method for
+    We present a novel synthetic-acceleration-based Monte Carlo method for
     solving the equilibrium thermal radiation diffusion equation in three
@@
 that was an application of Halton's Sequential Monte Carlo
-method~\cite{halton_1962,halton_1994} to the one-dimensional equilibrium
+method~\cite{halton_1962,halton_1994} to the one-dimensional (1D) equilibrium
 thermal radiation diffusion equation.  At that time, Halton's method was not
 widely known in the general computational physics community.  While extending
-our method to three spatial dimensions, we have realized that the Sequential
-Monte Carlo method is actually a variant of iterative refinement cast as a
-residual method. We have developed a new method in which Monte Carlo is used
-to accelerate a fixed-point (Richardson) iteration. This algorithm surpasses
-our previous one dimensional method and allows for extension to more general
-solution techniques.
+our method to three spatial dimensions, we realized that the Sequential Monte
+Carlo method is actually a variant of iterative refinement cast as a residual
+method. We have developed a new method in which Monte Carlo is used to
+accelerate a fixed-point (Richardson) iteration. This algorithm surpasses our
+previous 1D method and allows for extension to more general solution
+techniques.
 
 The purpose of this study is to demonstrate an efficient Monte Carlo solution
-method for three dimensional (3D), time-dependent, discrete systems. In this
+method for three-dimensional (3D), time-dependent, discrete systems. In this
 work, our previous investigations are extended in the following manner:
@@
 discrete form will be derived in \S~\ref{sec:discr-form-radi}.  We discuss
-using Monte Carlo Synthetic-Accleration to solve the discrete radiation
+using Monte Carlo Synthetic-Acceleration to solve the discrete radiation
 diffusion equation in \S~\ref{sec:solut-radi-diff}.  Results from two 3D test
@@
 
-The idea of using Monte Carlo methods (random walks) to invert linear
-systems is not new.  The earliest referenceable work dates to a 1950
-paper by Forsythe and Leibler \cite{forsythe}.  Their paper credits
-the idea to unpublished work by J.v. Neumann and S.M. Ulam dating back
-to the 1940's.  The basic principles of Monte Carlo matrix inversion
-were further elucidated in Hammersley and Handscomb's 1964 text
-\cite{hammersley_1964}.  These early methods are distinguished by very
-slow, statistically noisy convergence properties; thus, they have not
-made any significant impact in the physics community.
+The idea of using Monte Carlo methods (random walks) to invert linear systems
+is not new.  The earliest referenced work dates to a 1950 paper by Forsythe
+and Leibler \cite{forsythe}.  Their paper credits the idea to unpublished work
+by J.v. Neumann and S.M. Ulam dating back to the 1940's.  The basic principles
+of Monte Carlo matrix inversion were further elucidated in Hammersley and
+Handscomb's 1964 text \cite{hammersley_1964}.  These early methods are
+distinguished by very slow, statistically noisy convergence properties; thus,
+they have not made any significant impact in the physics community.
 
@@
 Carlo methods that can be used to solve linear systems and expand them
-to present our new synthetic acceleration method.
+to present our new synthetic-acceleration method.
 
@@
 \end{equation}
-which can be written,
+which can be written as
 \begin{equation}
@@
 less than 1, we can expand $\vA^{-1}$ using the \latin{Neumann
-  Series},
+  Series}:
 \begin{equation}
@@
 \end{equation} 
-Thus, when $\rho(\vH) < 1$ we can recast the solution vector $\vx$ as
-a series,
+Thus, when $\rho(\vH) < 1$, we can recast the solution vector $\vx$ as
+a series:
 \begin{equation}
@@
 With this knowledge in hand, we can write an iterative method that
-solves Eq.~(\ref{eq:Ax=b}) (Richardson's Iteration),
+solves Eq.~(\ref{eq:Ax=b}) (Richardson's Iteration):
 \begin{equation}
@@
 rewrite Eq.~(\ref{eq:neumann_series}) in order to calculate a
-component of $\vx$,
+component of $\vx$:
 \begin{equation}
@@
 random walk.  Let $X$ be a random variable sampled from a random walk
-with $k$ events that initiates in state $i$,
+with $k$ events that initiates in state $i$:
 \begin{equation}
@@
 %%---------------------------------------------------------------------------%%
-\subsection{Monte Carlo Synthetic-Acceleration}
+\subsection{Monte Carlo Synthetic Acceleration}
 \label{sec:iter-refin-monte}
@@
 Sequential Monte Carlo to speed up the convergence of these methods.  A
-variant of this scheme has been successfully applied to the 1D non-linear
+variant of this scheme has been successfully applied to the 1D nonlinear
 thermal radiation diffusion equation in Ref.~\cite{evans_2003}.
@@
 convergence.  This iteration sequence is closely related to iterative
-refinement; the exception being that there is no update to $x^{l+1}$ between
+refinement, the exception being that there is no update to $x^{l+1}$ between
 residual iterations.
@@
 We propose a modification of this scheme that uses Monte Carlo as a
-synthetic-acceleration for the fixed-point iteration given by
+synthetic acceleration for the fixed-point iteration given by
 Eq.~(\ref{eq:richardson}). We begin by subtracting Eq.~(\ref{eq:richardson})
-from Eq:~(\ref{eq:point_iteration}):
+from Eq.~(\ref{eq:point_iteration}):
 \begin{equation}
@@
 defined a scheme in which the initial estimate of $\vx$ in each iteration is
-updated using a single fixed-point iteration\footnote{Although, a subspace
-  method could be used instead of a stationary method here.}.  The residual is
-calculated and used as a source to estimate the error, $\delta\vx^{l+1/2}$, by
-solving Eq.~(\ref{eq:MCSA-residual_solve}) via the Monte Carlo adjoint
-method. The error is then used to calculate an updated iterate of the solution
-vector, $\vx^{l+1}$.  The entire sequence is iterated to convergence based on
-the following stopping criterion~\cite{kelley_1995},
+updated using a single fixed-point iteration\footnote{A subspace method could
+  be used instead of a stationary method here.}.  The residual is calculated
+and used as a source to estimate the error, $\delta\vx^{l+1/2}$, by solving
+Eq.~(\ref{eq:MCSA-residual_solve}) via the Monte Carlo adjoint method. The
+error is then used to calculate an updated iterate of the solution vector,
+$\vx^{l+1}$.  The entire sequence is iterated to convergence based on the
+following stopping criterion~\cite{kelley_1995}:
 \begin{equation}
@@
 effectiveness of the adjoint method over the direct method within the context
-of MCSA, we choose the two-dimensional time-dependent Poisson equation as a
+of MCSA, we choose the 2D time-dependent Poisson equation as a
 simple model problem:
@@
 integration. The Laplacian is differenced on a square Cartesian grid with a
-second-order five point stencil
+second-order five-point stencil,
 \begin{equation}
@@
 \end{equation}
-and a fourth-order nine point stencil 
+and a fourth-order nine-point stencil,
 \begin{multline}
@@
 
-A timing study and convergence study is used to demonstrate the effectiveness
-of the adjoint method as compared to the direct method. To assess both the CPU
-time and number of iterations required to converge to a solution, a problem of
+A timing and convergence study is used to demonstrate the effectiveness of the
+adjoint method as compared to the direct method. To assess both the CPU time
+and number of iterations required to converge to a solution, a problem of
 constant $\Delta$ was used with varying values of grid size, fixing the
 spectral radius of the system at a constant value for each variation. Both the
-five point and nine point stencils were used with both the direct and adjoint
+five-point and nine-point stencils were used with both the direct and adjoint
 solvers. For each case, 50 random walk permutations were computed per MCSA
@@
 computations presented in this section were completed on a 2.66 GHz Intel Core
-i7 machine with 8 GB 1067 MHz DDR3 memory. Fig.~\ref{fig:poisson_cpu_time}
-gives the CPU time needed for each case to converge in seconds and
+i7 machine with 8 GB 1067 MHz DDR3 memory. Figure~\ref{fig:poisson_cpu_time}
+gives the CPU time needed for each case to converge in seconds, and
 Fig.~\ref{fig:poisson_iterations} gives the number of iterations needed for
@@
   \includegraphics[width=5in,clip]{Adjoint_Direct_CPU_Time.pdf}
-  \caption{\sl CPU Time (s) to converge vs. Problem Size ($N$ for an
-    $N \times N$ square mesh). Both the adjoint and direct solvers are
-    used with the five point and nine point stencils. A significant
-    CPU time speedup is noted with the adjoint method due to the
-    reduced number of random walk events.}
+  \caption{\sl CPU Time (s) to converge vs. Problem Size ($N$ for an $N \times
+    N$ square mesh). Both the adjoint and direct solvers are used with the
+    five point and nine point stencils. A significant CPU time speedup is
+    noted with the adjoint method due to the reduced number of random walk
+    events.}
   \label{fig:poisson_cpu_time}
@@
   \includegraphics[width=5in,clip]{Adjoint_Direct_Iterations.pdf}
-  \caption{\sl Iterations to converge vs. Problem Size ($N$ for an $N
-    \times N$ square mesh). Both the adjoint and direct solvers are
-    used with the five point and nine point stencils. Both methods
-    give similar iteration behavior.}
+  \caption{\sl Iterations to converge vs. Problem Size ($N$ for an $N \times
+    N$ square mesh). Both the adjoint and direct solvers are used with the
+    five-point and nine-point stencils. Both methods give similar iteration
+    behavior.}
   \label{fig:poisson_iterations}
@@
 from all the nonzero parts of the source. Initially, this may not seem like a
-fair comparison, however, we see from Fig.~\ref{fig:poisson_iterations} that
+fair comparison; however, we see from Fig.~\ref{fig:poisson_iterations} that
 the number of iterations required to converge is approximately the same and
@@
 norm of the residual computed in Eq.~(\ref{eq:MCSA-residual}) was collected at
-each iteration for a fixed problem size. Fig.~\ref{fig:poisson_convergence}
+each iteration for a fixed problem size. Figure~\ref{fig:poisson_convergence}
 gives the results of these computations. First, it is worthy to note on the
@@
   \includegraphics[width=5in,clip]{Adjoint_Direct_Convergence.pdf}
-  \caption{\sl Infinity norm of the solution residual vs. iteration
-    number for a problem of fixed size. Both the adjoint and direct
-    solvers are used with the five point and nine point stencils.}
+  \caption{\sl Infinity norm of the solution residual vs. iteration number for
+    a problem of fixed size. Both the adjoint and direct solvers are used with
+    the five point and nine point stencils.}
   \label{fig:poisson_convergence}
@@
 $[\text{GJ}\cdot\text{cm}^{-2}\cdot\text{ns}^{-1}]$, as the primary state
-variable.  The intensity is defined as:
+variable.  The intensity is defined as
 \begin{equation}
@@
 where $\psi$ is the angular radiation intensity and $\nu$ is the radiation
-frequency.  The Planck function (or Planckian) is defined as,
+frequency.  The Planck function (or Planckian) is defined as
 \begin{equation}
@@
 the face-centered flux evaluated from either side of the face must be
-equivalent,
+equivalent:
 \begin{equation}
@@
 Following the same procedure at the $l-1/2$ face gives the diffusion
-coefficient on the low-side face,
+coefficient on the low-side face:
 \begin{equation}
@@
 Eq.~(\ref{eq:discrete-diffusion}), we must specify boundary conditions.  For
-this work, the Marshak Boundary condition suffices,
+this work, the Marshak Boundary condition suffices:
 \begin{equation}
@@
 $\partial\Gamma$.  On any low boundary we can write the flux using a
-first-order discrete form of Fick's Law,
+first-order discrete form of Fick's Law:
 \begin{equation}
@@
 As before, we enforce continuity of flux on the boundary using the
-Marshak Boundary condition,
+Marshak Boundary condition:
 \begin{equation}
@@
 Solving for $\phi_{1/2}$ and applying to Eq.~(\ref{eq:low-boundary-flux})
-yields an expression for the low-side boundary flux,
+yields an expression for the low-side boundary flux:
 \begin{equation}
@@
 Eq.~(\ref{eq:flux-balance-difference}) yielding the following modifications to
-Eqs.~(\ref{eq:sigmas}) on a boundary,
+Eqs.~(\ref{eq:sigmas}) on a boundary:
 \begin{align}
@@
 Furthermore, the following contributions are added to the source, $\qn_{ijk}$
-on a boundary cell,
+on a boundary cell:
 \begin{align}
@@
 The discrete diffusion equation in (\ref{eq:discrete-diffusion}) can be
-written in matrix-operator form as follows,
+written in matrix-operator form as follows:
 \begin{equation}
@@
   \rho(\ve{D})$ and $\rho(\ve{M}^{-1}\ve{D}) < 1$ \cite{golub}.  Thus,
-  convergence is guaranteed and fewer iterations are required resulting in
+  convergence is guaranteed and fewer iterations are required, resulting in
   accelerated convergence.
-\item The diagonal elements of the preconditioned iteration matrix are
-  zero, $\diag(\vI - \ve{M}^{-1}\ve{D}) = 0$.  Thus, there will be no
-  computations wasted for in-state transitions $i\rightarrow i$ during
-  the random walk process.
+\item The diagonal elements of the preconditioned iteration matrix are zero:
+  $\diag(\vI - \ve{M}^{-1}\ve{D}) = 0$.  Thus, there will be no computations
+  wasted for in-state transitions $i\rightarrow i$ during the random walk
+  process.
 \end{itemize}
@@
 
-An advantage of MCSA is that the resulting system does not have to be SPD,
-thus any suitable choice of $\ve{M}$ will suffice.  The preconditioned system
+An advantage of MCSA is that the resulting system does not have to be SPD;
+thus, any suitable choice of $\ve{M}$ will suffice.  The preconditioned system
 we will solve is defined in Eq.~(\ref{eq:preconditioned-diffusion}).  Applying
@@
 using the diffusion operator notation from
-Eq.~(\ref{eq:preconditioned-diffusion}),
+Eq.~(\ref{eq:preconditioned-diffusion}):
 \begin{align*}
@@
   \phi_6 &= \frac{\sigp_{7\rightarrow 6}}{\sigma_{T\,6}}
-  \frac{\Delta_7}{\Delta_6}\phi_7 + \ldots\:,\\ 
+  \frac{\Delta_7}{\Delta_6}\phi_7 + \ldots\:.\\ 
 \end{align*} 
@@
 The weight change for each scattering event is calculated using
-Eq.~(\ref{eq:adjoint-weight}),
+Eq.~(\ref{eq:adjoint-weight}):
 \begin{equation*}
@@
 entries in columns of the iteration matrix is 6, generating a sparse system.
-Cells along the problem boundary can have fewer than 6 entries.
+Cells along the problem boundary can have fewer than six entries.
 \begin{figure}[h!]
@@
 fixed-point (Richardson) iteration (PFIX).  The PCG method uses right-left
-preconditioning with $\ve{M}=\ve{L}\ve{L}^{-1}$, where $\ve{L} = \ve{L}^{T} =
-\sqrt{\ve{M}}$.  Naturally, more powerful preconditioning methods could be
+preconditioning with $\ve{M}=\ve{L}\ve{L}^{T}$, where $\ve{L} = \ve{L}^{T} =
+\sqrt{\diag(\ve{D})}$.  Naturally, more powerful preconditioning methods could be
 used; however, this provides a fair comparison with the preconditioning used
@@
 1~keV blackbody boundary flux at $x=0$ and propagates in the $x$-dimension in
-time as outlined in Ref.~\cite{larsen_1980}.  In other words, this a 3D
-representation of a 1D problem.  The full set of problem conditions is:
+time as outlined in Ref.~\cite{larsen_1980}.  In other words, this is a 3D
+representation of a 1D problem.  The full set of problem conditions is as
+follows:
 \begin{center}
@@
 As stated in \S~\ref{sec:monte-carlo-matrix}, the MCSA method will converge
-when $\rho(\ve{H})< 1$ where $\ve{H}$ is the iteration matrix and is defined
+when $\rho(\ve{H})< 1$, where $\ve{H}$ is the iteration matrix and is defined
 $\ve{H} = (\vI - \ve{M}^{-1}\ve{D})$.  The spectral radius for the Marshak
-problem is plotted in Fig.~\ref{fig:marshak_spectral}.  As stated in
-\S~\ref{sec:preconditioning} the preconditioning effectively reduces the
+problem is plotted in Fig.~\ref{fig:marshak_spectral}.  As described in
+\S~\ref{sec:preconditioning}, the preconditioning effectively reduces the
 spectral radius of $\ve{H}$ to less than 1.
@@
 There are two principal degrees of freedom when utilizing the MCSA method: (1)
-the number of particles per stage, and (2) the weight cutoff.
+the number of particles per stage and (2) the weight cutoff.
 Figure~\ref{fig:CPU_Np} shows the CPU time as a function of the requested
 number of particles per stage for a relative weight cutoff of $\sn{1}{-4}$.
-For less than 5 particles per stage the method does not converge.
+For less than five particles per stage, the method does not converge.
 Figure~\ref{fig:CPU_Ns} shows the variation in the maximum number of stages
@@
 relatively insensitive to the weight cutoff.  This is especially true when
-running small numbers of particles per stage.  When larger numbers of
-particles are used the weight cutoff has a more pronounced effect on the
-efficiency of the method.
+running small numbers of particles per stage.  The weight cutoff has a more
+pronounced effect on the efficiency of the method when larger numbers of
+particles are used.
 \begin{figure}[h!]
@@
     problem.  Results are shown for 10 and 50 requested particles per
-    stage.  The numbers in parenthesis show the maximum number of
+    stage.  The numbers in parentheses show the maximum number of
     particles run in a stage during the course of the problem.}
@@
 \begin{table}[h!]
+  \caption{Comparison of solution methods for the Marshak problem. All
+    methods were run with a stopping criterion of $\sn{1}{-8}$.  The
+    MCSA method used $N_p=10$ and $W_c = \sn{1}{-4}$.  The CPU time is
+    measured relative to the MCSA method, which required 72.6 s on a
+    3.6~GHz Xeon.}
+  \label{tab:marshak_comparison}
   \begin{center}
@@
   \end{center}
-  \caption{Comparison of solution methods for the Marshak problem. All
-    methods were run with a stopping criterion of $\sn{1}{-8}$.  The
-    MCSA method used $N_p=10$ and $W_c = \sn{1}{-4}$.  The CPU time is
-    measured relative to the MCSA method, which required 72.6 s on a
-    3.6~GHz Xeon.}
-  \label{tab:marshak_comparison}
 \end{table}
@@
 blackbody boundary flux is defined on the low-$x$ face at $T=0.5$~keV.  The
-full set of problem conditions is:
+full set of problem conditions is as follows:
 \begin{center}
@@
 A 2D plot of the temperature at 1000~ns elapsed time is illustrated in
-Fig.~\ref{fig:multi-mat_T}.  The time-evolution of the temperature is shown at
-4 edit points in Fig.~\ref{fig:edit-points}.  All three methods, PCG, PFIX,
+Fig.~\ref{fig:multi-mat_T}.  The time evolution of the temperature is shown at
+four edit points in Fig.~\ref{fig:edit-points}.  All three methods, PCG, PFIX,
 and MCSA, give numerically identical answers when using a stopping criterion
@@
     cut-plane positioned at the mid-point of the $z$-axis.  The positions of
-    the edit points where the time-evolution of the temperature is plotted in
+    the edit points where the time evolution of the temperature is plotted in
     Fig.~\ref{fig:edit-points} are indicated by the black disks.}
@@
 \begin{table}[htbp!]
+  \caption{ Comparison of solution methods for the multi-material
+    problem. All methods were run with a stopping criterion of
+    $\sn{1}{-8}$.  The MCSA method used $N_p=1000$ and $W_c =
+    \sn{1}{-3}$.  The problem was run to an elapsed time of 100~ns.
+    The MCSA method required 3.4 hours on a 3.6~GHz Xeon.}
+  \label{tab:multimat_comparison}
   \begin{center}
@@
   \end{center}
-  \caption{ Comparison of solution methods for the multi-material
-    problem. All methods were run with a stopping criterion of
-    $\sn{1}{-8}$.  The MCSA method used $N_p=1000$ and $W_c =
-    \sn{1}{-3}$.  The problem was run to an elapsed time of 100~ns.
-    The MCSA method required 3.4 hours on a 3.6~GHz Xeon.}
-  \label{tab:multimat_comparison}
 \end{table}

### 2012-09-06 doc/prelim/document/chapters/linear_problem/linear_problem.tex
updates to linear systems chapter

@@
 
-\section{Formulation of the Linear System}
+\section{Preliminaries}
 \label{sec:linear_system}
+
+We seek solutions of the general linear problem in the following form:
+\begin{equation}
+  \ve{A} \ve{x} = \ve{b}\:,
+  \label{eq:linear_problem}
+\end{equation}
+where $\ve{A} \in \mathbb{R}^{N \times N}$ is a matrix operator such
+that $\ve{A} : \mathbb{R}^{N} \rightarrow \mathbb{R}^{N}$, $\ve{x} \in
+\mathbb{R}^N$ is the solution vector, and $\ve{b} \in \mathbb{R}^N$ is
+the forcing term. The solutions to Eq~(\ref{eq:linear_problem}) will
+be generated by inverting $\ve{A}$ either directly or indirectly:
+\begin{equation}
+  \ve{x} = \ve{A}^{-1} \ve{b}
+  \label{eq:linear_problem_solution}\:.
+\end{equation}
+From the statement in Eq~(\ref{eq:linear_problem_solution}) we can
+already place a restriction on $\ve{A}$ by requiring that it be
+\textit{nonsingular}, meaning that we can in fact compute
+$\ve{A}^{-1}$. In this work we will focus our efforts on approximately
+inverting the operator through various means.
+
+In a discussion of methods for solving linear systems, several
+mathematical tools are useful in characterizing the qualities of the
+linear system. Among the most useful are the \textit{Eigenvalues} of
+the matrix, $\sigma(\ve{A})$. We find these by solving the Eigenvalue
+problem:
+\begin{equation}
+  \ve{A} \ve{x} = \lambda \ve{x},\ \lambda \in \sigma(\ve{A})\:.
+  \label{eq:eigenvalue_problem}
+\end{equation}
+By writing Eq~(\ref{eq:eigenvalue_problem}) in a different form,
+\begin{equation}
+  (\ve{A} - \lambda \ve{I})\ve{x} = 0 \:,
+  \label{eq:eigenvalue_problem_2}
+\end{equation}
+and demanding that non-trivial solutions for $\ve{x}$ exist, it is
+then required that $|\ve{A} - \lambda \ve{I}| = 0$. Expanding this
+determinant yields a characteristic polynomial in terms of $\lambda$
+with roots that form the set of Eigenvalues, $\sigma(\ve{A})$. A
+quantity of particular interest that is computatable from the
+eigenvalues of a matrix is the \textit{spectral radius},
+$\rho(\ve{A})$ defined by Saad \citep{saad_2003} as:
+\begin{equation}
+  \rho(\ve{A}) = \max_{\lambda \in \sigma(\ve{A})} |\lambda| \:.
+  \label{eq:spectral_radius}
+\end{equation}
+In addition, for problems that have a large scale over which the
+independent variables may exist (e.g. a problem with events on
+timescales ranging from nanoseconds to hours), a good measure of this
+range is supplied by the \textit{stiffness ratio}:
+\begin{equation}
+  Stiffness Ratio = \frac{\max_{\lambda \in \sigma(\ve{A})}
+    |\lambda|}{\min_{\lambda \in \sigma(\ve{A})} |\lambda|}
+\end{equation}
+Those problems that have a wide range of scales in their independent
+variables, which will then be reflected in the operator, will then
+have a large stiffness ratio. We will define such problems as
+\textit{stiff}.
+
+General to both matrices and vectors, \textit{norms} are a mechanism
+for collapsing objects of many elements to a single value. Per
+LeVeque's text \citep{leveque_2007}, the q-norm of a vector is defined
+as:
+\begin{equation}
+  ||\ve{v}||_q = \Bigg[ \sum_{i=1}^N |v_i|^q \Bigg]^{1/q},\ \ve{v} \in
+  \mathbb{R}^N\:,\ q \in \mathbb{Z}^+
+  \label{eq:q_norm}
+\end{equation}
+where ${v_i}$ is the $i^{th}$ component of the vector. Depending on
+the value chosen for $q$, local or global qualities of the vector may
+be obtained. For example, $q=2$ provides the root of a quadrature sum
+of all elements in the vector giving a global measure of the vector
+while $q=\infty$ gives the maximum value in the vector, a local
+quantity that does not give information regarding the other elements
+in the vector.
+
+We can also compute the norm of a matrix by inferring from the norm of
+the vector on which it is operating. Per LeVeque, we search for a
+constant that is equivalent to $||\ve{A}||$:
+\begin{equation}
+  ||\ve{A}\ve{x}|| \leq C ||\ve{x}||\:,
+  \label{eq:matrix_norm_inequality}
+\end{equation}
+where the minimum value of $C$ that satisfies
+Eq~(\ref{eq:matrix_norm_inequality}) is equivalent to $||\ve{A}||$
+that is valid $\forall \ve{x} \in \mathbb{R}^N$. The general
+definition in Eq~(\ref{eq:matrix_norm_inequality}) can be expanded in
+simple terms for common norms including the infinity norm:
+\begin{equation}
+  ||\ve{A}||_{\infty} = \max_{1 \leq i \leq N} \sum^N_{j=1}|a_{ij}|\:,
+  \label{eq:matrix_infinity_norm}
+\end{equation}
+and the 2-norm:
+\begin{equation}
+  ||\ve{A}||_{2} = \sqrt{\rho(\ve{A}^T\ve{A})}\:,
+  \label{eq:matrix_2_norm}
+\end{equation}
+where $\rho$ is the spectral radius as defined in
+Eq~(\ref{eq:spectral_radius}).
+
+Knowing this, we can then define several useful properties of matrices
+including the \textit{condition number} \citep{saad_2003}:
+\begin{equation}
+  \kappa(\ve{A}) = ||\ve{A}||\ ||\ve{A}^{-1}||\:,
+  \label{eq:condition_number}
+\end{equation}
+which gives as a metric on assessing how close to singular the system
+is. This is due to the fact $||\ve{A}^{-1}||$ is large near
+singularities (and undefined for a singular matrix) and thus a large
+condition number will be generated. We define such matrices as
+\textit{ill-conditioned}.
 
@@
 \label{sec:linear_methods}
+The discretization of partial differential equations (\textit{PDEs})
+through common methods such as finite differences
+\citep{leveque_2007}, finite volumes \citep{leveque_2002}, and finite
+elements \citep{zienkiewicz_1977} ultimately generate matrix problems
+as in Eq~(\ref{eq:linear_problem}).In addition, these matrices are
+typically sparse, meaning that the vast majority of their constituent
+elements are zero. This sparsity is due to the fact that the influence
+of a particular grid element only expands as far as a few of its
+nearest neighbors depending on the order of discretization used, and
+therefore coupling among variables in a particular discrete equation
+in the system leads to a few non-zero entries.
 

### 2012-09-07 doc/prelim/document/chapters/introduction/introduction.tex
more updates to linear problem chapter

@@
 be pertinent to robust algorithms aimed at this new
-hardware. Considering these issues, this thesis proposes a the
+hardware. Considering these issues, this thesis proposes the
 development of a massively parallel stochastic method for linear
@@
 benefit from a general and parallel MCSA scheme. In the engineering
-community, nonlinear problems are often addressed by linearizing the
-problem and using traditionaly iterative or direct methods
-\citep{tannehill_1997}. In the mathematics community, various Newton
-methods have been popular \citep{kelley_1995}. Recently, Jacobian-Free
-Newton-Krylov (JFNK) schemes \citep{knoll_2004} have been utilized in
-multiple physics architectures and advanced single physics codes
-\citep{gaston_2009}. The benefits of JFNK schemes are that the Jacobian
-is never formed, simplifying the implementation, and a Krylov solver
-is leveraged (typically GMRES or Conjugate Gradient), providing
-excellent convergence properties for well-conditioned and well-scaled
-systems. However, there are two potential drawbacks to these methods
-for high fidelity predictive simulations: (1) the Jacobian is
-approximated by a first-order differencing method on the order of
-machine precision such that this error can grow beyond that of those
-in a fine-grained system and (2) for systems that are not symmetric
-positive-definite (which will be the case for most multiphysics
-systems and certainly for most preconditioned systems) the Krylov
-subspace generated by the GMRES solver may become prohibitively
-large. To address these issues, this thesis proposes a new and novel
-method for nonlinear systems based on the MCSA method.
+community, nonlinear problems are often addressed by either
+linearizing the problem or building a segregated scheme and using
+traditionaly iterative or direct methods to solve the resulting
+system \citep{tannehill_1997}. In the mathematics community, various
+Newton methods have been popular \citep{kelley_1995}. Recently,
+Jacobian-Free Newton-Krylov (JFNK) schemes \citep{knoll_2004} have
+been utilized in multiple physics architectures and advanced single
+physics codes \citep{gaston_2009}. The benefits of JFNK schemes are
+that the Jacobian is never formed, simplifying the implementation, and
+a Krylov solver is leveraged (typically GMRES or Conjugate Gradient),
+providing excellent convergence properties for well-conditioned and
+well-scaled systems. However, there are two potential drawbacks to
+these methods for high fidelity predictive simulations: (1) the
+Jacobian is approximated by a first-order differencing method on the
+order of machine precision such that this error can grow beyond that
+of those in a fine-grained system and (2) for systems that are not
+symmetric positive-definite (which will be the case for most
+multiphysics systems and certainly for most preconditioned systems)
+the Krylov subspace generated by the GMRES solver may become
+prohibitively large. To address these issues, this thesis proposes a
+new and novel method for nonlinear systems based on the MCSA method.
 

### 2012-09-07 doc/prelim/document/chapters/linear_problem/linear_problem.tex
more updates to linear problem chapter

@@
 \section{Preliminaries}
-\label{sec:linear_system}
-
+\label{sec:linear_preliminaries}
 We seek solutions of the general linear problem in the following form:
@@
 \end{equation}
-From the statement in Eq~(\ref{eq:linear_problem_solution}) we can
-already place a restriction on $\ve{A}$ by requiring that it be
-\textit{nonsingular}, meaning that we can in fact compute
-$\ve{A}^{-1}$. In this work we will focus our efforts on approximately
-inverting the operator through various means.
+In addition we can define the residual:
+\begin{equation}
+  \ve{r} = \ve{b} - \ve{A}\ve{x}\:,
+  \label{eq:linear_residual}
+\end{equation}
+such that an exact solution to $\ve{x}$ has been found when
+$\ve{r}=\ve{0}$.  From the statement in
+Eq~(\ref{eq:linear_problem_solution}) we can already place a
+restriction on $\ve{A}$ by requiring that it be \textit{nonsingular},
+meaning that we can in fact compute $\ve{A}^{-1}$. In this work we
+will focus our efforts on approximately inverting the operator through
+various means.
 
@@
 determinant yields a characteristic polynomial in terms of $\lambda$
-with roots that form the set of Eigenvalues, $\sigma(\ve{A})$. A
-quantity of particular interest that is computatable from the
-eigenvalues of a matrix is the \textit{spectral radius},
+with roots that form the set of Eigenvalues, $\sigma(\ve{A})$. Each
+component of $\sigma(\ve{A})$ can then be used to solve
+Eq~(\ref{eq:eigenvalue_problem_2}) for a particular permutation of
+$\ve{x}$. The set of all permutations for the \textit{Eigenvectors} of
+$\ve{A}$. A quantity of particular interest that is computatable from
+the eigenvalues of a matrix is the \textit{spectral radius},
 $\rho(\ve{A})$ defined by Saad \citep{saad_2003} as:
@@
 elements \citep{zienkiewicz_1977} ultimately generate matrix problems
-as in Eq~(\ref{eq:linear_problem}).In addition, these matrices are
+as in Eq~(\ref{eq:linear_problem}).  In addition, these matrices are
 typically sparse, meaning that the vast majority of their constituent
@@
 therefore coupling among variables in a particular discrete equation
-in the system leads to a few non-zero entries.
+in the system leads to a few non-zero entries. Because of the natural
+occurrence of sparse matrices in common numerical methods many
+iterative techniques have been developed to solve such systems. We
+discuss here conventional stationary and subspace methods to provide
+background and then give detail on a stochastic class of methods.
 
@@
 \label{subsec:stationary_methods}
+Stationary methods arise from splitting the operator in
+Eq~(\ref{eq:linear_problem})
+\begin{equation}
+  \ve{A} = \ve{M} - \ve{N}\:,
+  \label{eq:split_linear_operator}
+\end{equation}
+where the choice of $\ve{M}$ and $\ve{N}$ will be dictated by the
+particular method chosen. Using this split definition of the operator
+we can then write:
+\begin{equation}
+  \ve{M}\ve{x} + \ve{N}\ve{x} = \ve{b}\:.
+  \label{eq:linear_split_equation1}
+\end{equation}
+By rearranging, we can generate a form more useful for analysis:
+\begin{equation}
+  \ve{x} = \ve{H}\ve{x} + \ve{c}\:,
+  \label{eq:linear_split_equation2}
+\end{equation}
+where $\ve{H}=\ve{M}^{-1}\ve{N}$ is defined as the \textit{iteration
+  matrix} and $\ve{c}=\ve{M}^{-1}\ve{b}$. With the solution vector on
+both the left and right hand sides, an iterative method can then be
+formed:
+\begin{equation}
+    \ve{x}^{k+1} = \ve{H}\ve{x}^k + \ve{c}\:,
+  \label{eq:linear_iterative_method}
+\end{equation}
+with $k \in \mathbb{Z}^+$ defined as the \textit{iteration index}. In
+general, we will define methods in the form of
+Eq~(\ref{eq:linear_iterative_method}) as \textit{stationary
+  methods}. Given this, we can then generate a few statements
+regarding the convergence of such stationary methods. Defining
+$\boldsymbol{\delta}^k = \ve{u}^k - \ve{u}$ as the solution error at the
+$k^{th}$ iterate, we can subtract Eq~(\ref{eq:linear_split_equation2})
+from Eq~(\ref{eq:linear_iterative_method}) to arrive at an error form
+of the linear problem:
+\begin{equation}
+  \boldsymbol{\delta}^{k+1} = \ve{H}\boldsymbol{\delta}^k\:. 
+  \label{eq:linear_iterative_error}
+\end{equation}
+Our error after $k$ iterations is then:
+\begin{equation}
+  \boldsymbol{\delta}^{k} = \ve{H}^k\boldsymbol{\delta}^0\:. 
+  \label{eq:linear_k_iter_error}
+\end{equation}
+In other words, successive application of the iteration matrix is the
+mechanism driving down the error in a stationary method. We can then
+place restrictions on the iteration matrix by using the tools
+developed in \S~(\ref{sec:linear_preliminaries}). By assuming $\ve{H}$
+is diagonalizable (although we may generalize this with the Jordan
+canonical form of $H$ \citep{saad_2003}), we then have:
+\begin{equation}
+  \boldsymbol{\delta}^{k} =
+  \ve{R}\boldsymbol{\Lambda}^k\ve{R}^{-1}\boldsymbol{\delta}^0\:,
+  \label{eq:linear_k_iter_error_diag}
+\end{equation}
+where $\boldsymbol{\Lambda}$ contains the Eigenvalues of $\ve{H}$ on
+its diagonal and the columns of $\ve{R}$ contain the Eigenvectors of
+$\ve{H}$. Computing the 2-norm of the above form then gives:
+\begin{equation}
+  ||\boldsymbol{\delta}^{k}||_2 \leq ||\boldsymbol{\Lambda}^k||_2\ 
+  ||\ve{R}||_2\ ||\ve{R}^{-1}||_2\ ||\boldsymbol{\delta}^0||_2\:,
+  \label{eq:linear_k_iter_norm1}
+\end{equation}
+which gives:
+\begin{equation}
+  ||\boldsymbol{\delta}^{k}||_2 \leq \rho(\ve{H})^k \kappa(\ve{R})
+  ||\boldsymbol{\delta}^0||_2\:.
+  \label{eq:linear_k_iter_norm2}
+\end{equation}
+For iteration matrices where the Eigenvectors are orthogonal,
+$\kappa(\ve{R})=0$ and the error bound reduces to:
+\begin{equation}
+  ||\boldsymbol{\delta}^{k}||_2 \leq \rho(\ve{H})^k
+  ||\boldsymbol{\delta}^0||_2\:.
+  \label{eq:linear_k_iter_norm3}
+\end{equation}
+We can now resrtict $\ve{H}$ by asserting that $\rho(\ve{H}) < 1$
+for a stationary method to converge such that $k$ applications of the
+iteration matrix will not cause the error to grow in
+Eq~(\ref{eq:linear_k_iter_norm3}).
 

### 2012-09-12 doc/prelim/document/chapters/linear_problem/linear_problem.tex
changing some section names in the linear problem chapter
adding projection method figure

@@
 regarding the convergence of such stationary methods. Defining
-$\boldsymbol{\delta}^k = \ve{u}^k - \ve{u}$ as the solution error at the
+$\ve{e}^k = \ve{u}^k - \ve{u}$ as the solution error at the
 $k^{th}$ iterate, we can subtract Eq~(\ref{eq:linear_split_equation2})
@@
 \begin{equation}
-  \boldsymbol{\delta}^{k+1} = \ve{H}\boldsymbol{\delta}^k\:. 
+  \ve{e}^{k+1} = \ve{H}\ve{e}^k\:. 
   \label{eq:linear_iterative_error}
@@
 \begin{equation}
-  \boldsymbol{\delta}^{k} = \ve{H}^k\boldsymbol{\delta}^0\:. 
+  \ve{e}^{k} = \ve{H}^k\ve{e}^0\:. 
   \label{eq:linear_k_iter_error}
@@
 \begin{equation}
-  \boldsymbol{\delta}^{k} =
-  \ve{R}\boldsymbol{\Lambda}^k\ve{R}^{-1}\boldsymbol{\delta}^0\:,
+  \ve{e}^{k} =
+  \ve{R}\boldsymbol{\Lambda}^k\ve{R}^{-1}\ve{e}^0\:,
   \label{eq:linear_k_iter_error_diag}
@@
 \begin{equation}
-  ||\boldsymbol{\delta}^{k}||_2 \leq ||\boldsymbol{\Lambda}^k||_2\ 
-  ||\ve{R}||_2\ ||\ve{R}^{-1}||_2\ ||\boldsymbol{\delta}^0||_2\:,
+  ||\ve{e}^{k}||_2 \leq ||\boldsymbol{\Lambda}^k||_2\ 
+  ||\ve{R}||_2\ ||\ve{R}^{-1}||_2\ ||\ve{e}^0||_2\:,
   \label{eq:linear_k_iter_norm1}
@@
 \begin{equation}
-  ||\boldsymbol{\delta}^{k}||_2 \leq \rho(\ve{H})^k \kappa(\ve{R})
-  ||\boldsymbol{\delta}^0||_2\:.
+  ||\ve{e}^{k}||_2 \leq \rho(\ve{H})^k \kappa(\ve{R})
+  ||\ve{e}^0||_2\:.
   \label{eq:linear_k_iter_norm2}
@@
 \begin{equation}
-  ||\boldsymbol{\delta}^{k}||_2 \leq \rho(\ve{H})^k
-  ||\boldsymbol{\delta}^0||_2\:.
+  ||\ve{e}^{k}||_2 \leq \rho(\ve{H})^k
+  ||\ve{e}^0||_2\:.
   \label{eq:linear_k_iter_norm3}
@@
 
-\subsection{Subspace Methods}
-\label{subsec:subspace_methods}
+\subsection{Projection Methods}
+\label{subsec:projection_methods}
+The most common iterative methods used in scientific computing today
+for sparse systems are of a broad class known as \textit{projection
+  methods}. These methods not only provide access to more powerful
+means of reaching a solution, but also a powerful means of
+encapsulating the majority of common iterative methods including the
+stationary methods just discussed. All projection methods are built
+around a core framework where the solution to
+Eq~(\ref{eq:linear_problem}) is extracted from a \textit{search
+  subspace} $\mathcal{K}$ and bound by the constraint subspace
+$\mathcal{L}$ that will vary in definition depending on the iterative
+method selected. We build the approximate solution $\tilde{\ve{x}}$ by
+starting with an initial guess $\ve{x}_0$ and extracting a correction
+$\boldsymbol{\delta}$ from $\mathcal{K}$ such that:
+\begin{equation}
+  \tilde{\ve{x}} = \ve{x}_0 +
+  \boldsymbol{\delta},\ \boldsymbol{\delta} \in \mathcal{K}\:.
+  \label{eq:linear_projection_step}
+\end{equation}
+We bound this correction by asserting that the new residual,
+$\tilde{\ve{r}}$, be orthogonal to $\mathcal{L}$:
+\begin{equation}
+  \langle \tilde{\ve{r}},\ve{w} \rangle = 0,\ \forall w \in
+  \mathcal{L}\:.
+  \label{eq:linear_projection_constraint}
+\end{equation}
+We can generate a more physical and geometric-based understanding of
+these constraints by writing the new residual as $\tilde{\ve{r}} =
+\ve{r}_0 - \ve{A}\boldsymbol{\delta}$ and again asserting the residual
+must be orthogonal to $\mathcal{L}$. This situation is geometrically
+presented in Figure~\ref{fig:projection_constraint}.
+\begin{figure}[htpb!]
+  \begin{center}
+    \scalebox{1.75}{
+      \input{chapters/linear_problem/orthogonal_residual.pdftex_t} }
+  \end{center}
+  \caption{ Orthogonality of the correction residual with respect to
+    $\mathcal{L}$.}
+  \label{fig:projection_constraint}
+\end{figure}
 
-\subsubsection{Conjugate Gradient}
-\label{subsubsec:conjugate_gradient}
+\subsubsection{The Krylov Subspace}
+\label{subsubsec:krylov_subspace}
 
@@
 
+\subsubsection{A Stationary Method Perspective}
+\label{subsubsec:mcsa_stationary}
+
+\subsubsection{A Projection Method Perspective}
+\label{subsubsec:mcsa_projection}
+
 \section{Preconditioning the Linear System}

### 2012-09-12 doc/prelim/document/chapters/linear_problem/linear_problem.tex
updating makefile for xfig files
ufinished projection methods section and updated research proposal outline

@@
 variables, which will then be reflected in the operator, will then
-have a large stiffness ratio. We will define such problems as
-\textit{stiff}.
+have a large stiffness ratio. We will define such problems with large
+stiffness ratios as \textit{stiff}.
 
@@
 \label{subsec:projection_methods}
-The most common iterative methods used in scientific computing today
-for sparse systems are of a broad class known as \textit{projection
+Among the most common iterative methods used in scientific computing
+today for sparse systems are a broad class known as \textit{projection
   methods}. These methods not only provide access to more powerful
@@
 encapsulating the majority of common iterative methods including the
-stationary methods just discussed. All projection methods are built
-around a core framework where the solution to
-Eq~(\ref{eq:linear_problem}) is extracted from a \textit{search
-  subspace} $\mathcal{K}$ and bound by the constraint subspace
-$\mathcal{L}$ that will vary in definition depending on the iterative
-method selected. We build the approximate solution $\tilde{\ve{x}}$ by
-starting with an initial guess $\ve{x}_0$ and extracting a correction
-$\boldsymbol{\delta}$ from $\mathcal{K}$ such that:
+stationary methods just discussed in a common mathematical
+framework. All projection methods are built around a core structure
+where the solution to Eq~(\ref{eq:linear_problem}) is extracted from a
+\textit{search subspace} $\mathcal{K}$ and bound by a
+\textit{constraint subspace} $\mathcal{L}$ that will vary in
+definition depending on the iterative method selected. We build the
+approximate solution $\tilde{\ve{x}}$ by starting with an initial
+guess $\ve{x}_0$ and extracting a correction $\boldsymbol{\delta}$
+from $\mathcal{K}$ such that:
 \begin{equation}
@@
 \begin{equation}
-  \langle \tilde{\ve{r}},\ve{w} \rangle = 0,\ \forall w \in
+  \langle \tilde{\ve{r}},\ve{w} \rangle = 0,\ \forall \ve{w} \in
   \mathcal{L}\:.
@@
 \end{equation}
+
 We can generate a more physical and geometric-based understanding of
@@
 \ve{r}_0 - \ve{A}\boldsymbol{\delta}$ and again asserting the residual
-must be orthogonal to $\mathcal{L}$. This situation is geometrically
-presented in Figure~\ref{fig:projection_constraint}.
+must be orthogonal to $\mathcal{L}$. If $\tilde{\ve{r}}$ is to be
+orthogonal to $\mathcal{L}$, then $\ve{A}\boldsymbol{\delta}$ must be
+the projection of $\ve{r}_0$ onto the subspace $\mathcal{L}$ that
+eliminates the components of the residual that exist in
+$\mathcal{L}$. This situation is geometrically presented in
+Figure~\ref{fig:linear_projection_constraint}.
+
 \begin{figure}[htpb!]
@@
   \end{center}
-  \caption{ Orthogonality of the correction residual with respect to
-    $\mathcal{L}$.}
-  \label{fig:projection_constraint}
+  \caption{Orthogonality constraint of the new residual with respect
+    to $\mathcal{L}$.}
+  \label{fig:linear_projection_constraint}
 \end{figure}
 
-\subsubsection{The Krylov Subspace}
-\label{subsubsec:krylov_subspace}
+From Figure~\ref{fig:linear_projection_constraint} we then note that
+the following geometric condition must hold:
+\begin{equation}
+  ||\tilde{\ve{r}}||_2 \leq ||\ve{r}_0||_2,\ \forall \ve{r}_0 \in
+  \mathbb{C}^N\:,
+  \label{eq:linear_min_res}
+\end{equation}
+meaning that the residual of the system will always be
+\textit{minimized} with respect to the constraints.
+
+Given this minimization condition for the residual, we can form the
+outline of an iterative projection method. Consider a matrix $\ve{V}$
+to form a basis of $\mathcal{K}$ and a matrix $\ve{W}$ to form a basis
+of $\mathcal{L}$. As $\boldsymbol{\delta} \in \mathcal{K}$ by
+definition in Eq~(\ref{eq:linear_projection_step}), then
+$\boldsymbol{\delta}$ can instead be rewritten as:
+\begin{equation}
+  \boldsymbol{\delta} = \ve{V}\ve{y},\ \forall \ve{y} \in \mathbb{C}^N:\,
+  \label{eq:linear_delta_projection}
+\end{equation}
+where $\ve{V}$ \textit{projects} $\ve{y}$ onto $\mathcal{K}$. From the
+orthogonality constraint in Eq~(\ref{eq:linear_projection_constraint})
+it then follows that:
+\begin{equation}
+  \ve{y} = (\ve{W}^T\ve{A}\ve{V})^{-1}\ve{W}^T\ve{r}_0\:,
+  \label{eq:linear_constraint_projection}
+\end{equation}
+where here the projection onto $\mathcal{K}$ is constrained by the
+projection onto $\mathcal{L}$. Knowing this, we can then outline the
+following iteration scheme for a projection method:
+\begin{subequations}
+  \begin{gather}
+    \ve{r}^k = \ve{b} - \ve{A}\ve{x}^k\:,\\  
+    \ve{y} = (\ve{W}^T\ve{A}\ve{V})^{-1}\ve{W}^T\ve{r}_0\:,\\
+    \ve{x}^{k+1} = \ve{x}^k + \ve{V}\ve{y}\:,
+  \end{gather}
+  \label{eq:linear_projection_iteration}
+\end{subequations}
+where $\ve{V}$ and $\ve{W}$ are generated from the definitions of
+$\mathcal{K}$ and $\mathcal{L}$ which are defined prior to each
+iteration.
+
+From an iteration standpoint, as we choose $\boldsymbol{\delta}$ from
+$\mathcal{K}$ and constrain it with $\mathcal{L}$, each iteration
+performs a projection that systematically anihilates the components of
+the residual that exists in $\mathcal{L}$. This then means that if our
+convergence criteria for an iterative method is bound the residual of
+the system, then Eq~(\ref{eq:linear_min_res}) tells us that each
+projection step gaurantees us that the new residual will never be
+worse than that of the previous step and will typically move us
+towards convergence. Depending on the qualities of the system in
+Eq~(\ref{eq:linear_problem}), the selection of the subspaces
+$\mathcal{K}$ and $\mathcal{L}$ can serve to both gaurantee
+convergence and optimize the rate at which the residual is decreased.
 
@@
 \label{subsubsec:gmres}
+Among the most common projection methods used in practice are class
+known as \textit{Krylov subspace methods}.
 

### 2012-09-12 doc/prelim/document/chapters/research_proposal/research_proposal.tex
Finished initial prelim outline
ufinished projection methods section and updated research proposal outline

@@
 
-\section{Model Problems}
-\label{sec:model_problems}
+After all of the background information and explanations of the new
+work, this section is the most important of all. Here I will clearly
+define the questions I aim to answer and the strategy that I will use
+to attempt to answer them.
 
-\section{Method Comparisons}
-\label{sec:method_comparisons}
+\section{Progress to Date}
+\label{sec:progress}
 
-\section{Performance Metrics}
-\label{sec:performance_metrics}
+\subsection{Generalization of MCSA for Linear Problems}
+\label{subsec:mcsa_generalization}
+
+\subsection{Direct vs. Adjoint Analysis}
+\label{subsec:mcsa_direct_vs_adjoint}
+
+\subsection{MCSA Application to Finite Element Problems}
+\label{subsec:mcsa_finite_element}
+
+\section{Key Questions}
+\label{sec:key_questions}
+
+\section{Solution Strategies}
+\label{seq:solution_strategies}

### 2012-09-13 doc/prelim/document/chapters/introduction/introduction.tex
updates to prelim, projection methods section is complete

@@
 problems and a novel stochastic method to advance solution techniques
-for nonlinear problems with a focus on coupled physics systems.
+for nonlinear problems with the aim of improving the modeling of
+coupled physics systems for nuclear reactor analysis.
 
@@
 to date will be reported including the development of a new physics
-framework that will serve as a test bed for these new methods. A
+framework that will serve as a test bed for these new methods that
+leverages a preliminary implementation of the stochastic method. A
 research plan will then be provided that continues the development and
-implementation of these new methods and tests them in practical
-settings.
+implementation of these new methods and verifies them with benchmark
+solutions, culminating in the analysis of a multiphysics simulation of
+a pressurized water reactor assembly.
 
@@
 has focused on various deterministic methods for solutions to linear
-problems \citep{saad_2003}. In between these two areas are a not widely
-known small group of stochastic methods for solving sparse linear
-systems \citep{hammersley_1964, halton_1962, halton_1994}. In recent
-years, these methods have been further developed for transport
-problems in the form of Monte Carlo Synthetic-Acceleration (MCSA)
-\citep{evans_2003, evans_2009} but have yet to be applied to more
-general sparse linear systems. Compared to other methods in these
-regime, MCSA offers three attractive qualities; (1) the linear problem
-operator need not be symmetric or positive-definite, thereby reducing
-preconditioning complexity, (2) the stochastic nature of the
-solution method provides a natural solution to the issue of resiliency,
-and (3) is amenable to parallelization using modern methods developed by
-the transport community \citep{wagner_2011}. The developement of MCSA
-as a general linear solver and the development of a parallel MCSA
-method will be new and unique features of this work. Resiliency will
-not be addressed.
+problems \citep{saad_2003, kelley_1995}. In between these two areas
+are a not widely known small group of stochastic methods for solving
+sparse linear systems \citep{hammersley_1964, halton_1962,
+  halton_1994}. In recent years, these methods have been further
+developed for transport problems in the form of Monte Carlo
+Synthetic-Acceleration (MCSA) \citep{evans_2003, evans_2009} but have
+yet to be applied to more general sparse linear systems commonly
+generated by the computational physics community. Compared to other
+methods in these regime, MCSA offers three attractive qualities; (1)
+the linear problem operator need not be symmetric or
+positive-definite, thereby reducing preconditioning complexity, (2)
+the stochastic nature of the solution method provides a natural
+solution to the issue of resiliency, and (3) is amenable to
+parallelization using modern methods developed by the transport
+community \citep{wagner_2011}. The developement of MCSA as a general
+linear solver and the development of a parallel MCSA method will be
+new and unique features of this work. Resiliency will not be
+addressed.
 

### 2012-09-13 doc/prelim/document/chapters/linear_problem/linear_problem.tex
updates to prelim, projection methods section is complete

@@
 iterative techniques have been developed to solve such systems. We
-discuss here conventional stationary and subspace methods to provide
+discuss here conventional stationary and projection methods to provide
 background and then give detail on a stochastic class of methods.
@@
 is diagonalizable (although we may generalize this with the Jordan
-canonical form of $H$ \citep{saad_2003}), we then have:
+canonical form of $H$ \citep{saad_2003}, we then have:
 \begin{equation}
@@
 the system, then Eq~(\ref{eq:linear_min_res}) tells us that each
-projection step gaurantees us that the new residual will never be
-worse than that of the previous step and will typically move us
-towards convergence. Depending on the qualities of the system in
+projection step gaurantees us that the norm of the new residual will
+never be worse than that of the previous step and will typically move
+us towards convergence. Depending on the qualities of the system in
 Eq~(\ref{eq:linear_problem}), the selection of the subspaces
@@
 
-\subsubsection{GMRES}
-\label{subsubsec:gmres}
-Among the most common projection methods used in practice are class
-known as \textit{Krylov subspace methods}.
+\subsubsection{Krylov Subspace Methods}
+\label{subsubsec:krylov_methods}
+Among the most common projection techniques used in practice are a
+class of methods known as \textit{Krylov subspace methods}. Here, the
+search subspace is defined as the \textit{Krylov subspace}:
+\begin{equation}
+  \mathcal{K}_m(\ve{A},\ve{r}_0) = span\{\ve{r}_0, \ve{A}\ve{r}_0,
+  \ve{A}^2\ve{r}_0, \dots, \ve{A}^{m-1}\ve{r}_0\}\:,
+  \label{eq:krylov_subspace}
+\end{equation}
+where $m$ denotes the dimensionality of the subspace. In order to
+accomodate a more general structure for the operator in
+Eq~(\ref{eq:linear_problem}), we often choose an \textit{oblique}
+projection method where $\mathcal{K} \neq \mathcal{L}$. If we choose
+$\mathcal{L} = \ve{A} \mathcal{K}_m(\ve{A},\ve{r}_0)$, then we are
+ultimately solving the normal system $\ve{A}^T\ve{A}\ve{x} =
+\ve{A}^T\ve{b}$ where $\ve{A}^T\ve{A}$ will be symmetric positive
+definite if $\ve{A}$ is nonsingular, thereby expanding the range of
+operators over which these methods are valid. This choice of
+constraint subspace also then gives us the result via
+Eq~(\ref{eq:linear_projection_constraint}) that the residual is
+minimized for all $\boldsymbol{\delta} \in \mathcal{K}$, forming the
+basis for the \textit{generalized minimum Residual method} (GMRES)
+\citep{saad_1986}.
+
+Choosing GMRES as our model Krylov method, we are first tasked with
+finding a projector onto the subspace. We seek an orthonormal basis
+for $\mathcal{K}_m(\ve{A},\ve{r}_0)$ by an orthogonalization procedure
+that is commonly based on, but not limited to, the \textit{Arnoldi}
+recurrence relation. The Arnoldi procedure will generate an
+orthonormal basis, $\ve{V}_m \in \mathbb{C}^{N \times m}$, via a
+variant of the Gram-Schmidt procedure that re-applies the operator for
+each consecutive vector, thus forming a basis that spans the subspace
+in Eq~(\ref{eq:krylov_subspace}). Due to its equivalent
+dimensionality, $m$, to that of the subspace, we will refer to such
+recurrence relations as \textit{long recurrence relations}. Those
+orthogonal projection procedures that have a dimensionality less than
+$m$ will be refferred to as \textit{short recurrence relations}.  Once
+$\ve{V}_m$ is found, per the constraint subspace definition it then
+follows that its basis is defined as $\ve{W}_m = \ve{A} \ve{V}_m$.
+
+Knowing the projections onto the search and constraint subspaces, the
+GMRES method may be formulated. We note here several properties of
+this formulation and how they may facilitate or hinder the solution of
+large-scale, sparse linear problems, also noting that these properties
+are common among many Krylov methods. First, from a memory perspective
+GMRES is efficient from the standpoint that the operator $\ve{A}$ need
+not be explicitly stored. Rather, only the ability to compute the
+action of that operator on a vector of valid size is
+required. However, these savings in memory are balanced by the fact
+that the long recurrence relations used in the Arnoldi procedure
+require all vectors that span the Krylov space to be stored. If the
+size of these vectors becomes prohibitive, the Arnoldi procedure can
+be restarted at the cost of losing information in the
+orthogonalization process, creating the potentitial to generate new
+search directions that are not orthogonal to all previous search
+directions (and therefore less than optimal). From an implementation
+perspective, because the operator is not required to be formed, GMRES
+is significantly more flexible in its usage in that there are many
+instances where various processes serve to provide the action of that
+operator (e.g. radiation transport sweeps) that may not be amenable to
+its construction.
 
@@
 
-\subsubsection{A Stationary Method Perspective}
-\label{subsubsec:mcsa_stationary}
-
-\subsubsection{A Projection Method Perspective}
-\label{subsubsec:mcsa_projection}
-
 \section{Preconditioning the Linear System}

### 2012-09-13 doc/prelim/document/chapters/research_proposal/research_proposal.tex
updates to prelim, projection methods section is complete

@@
 define the questions I aim to answer and the strategy that I will use
-to attempt to answer them.
+to attempt to answer them. I think that it will be important to pick a
+challenge problem here that FANM has a chance to show gains in either
+memory footprint or parallelism over current methods. In particular, I
+would like to have this challenge problem derive from nuclear reactor
+multiphysics. I think Drekar would be a great test bed for FANM once
+complete as it has all the tools to build a FANM solve (it is in fact
+based on the same core libraries as Chimera) and a complex reactor
+core assembly multiphysics system capable of scaling to large
+machines. In addition, all of the complex diagnostic tools necessary
+for analysis of FANM performance as compared to the other methods
+should be available.
 
@@
 \section{Solution Strategies}
-\label{seq:solution_strategies}
+\label{sec:solution_strategies}
+
+\subsection{Parallel MCSA Verification}
+\label{subsubsec:mcsa_verification}
+
+\subsection{FANM Verification}
+\label{subsubsec:fanm_verification}
+
+\section{Challenge Problem}
+\label{sec:challenge_problem}

### 2012-09-13 doc/prelim/document/chapters/linear_problem/linear_problem.tex
additions to the stochastic methods section

@@
 minimized for all $\boldsymbol{\delta} \in \mathcal{K}$, forming the
-basis for the \textit{generalized minimum Residual method} (GMRES)
+basis for the \textit{generalized minimum residual method} (GMRES)
 \citep{saad_1986}.
@@
 instances where various processes serve to provide the action of that
-operator (e.g. radiation transport sweeps) that may not be amenable to
-its construction.
+operator (e.g. radiation transport sweeps \citep{evans_2010}) that may
+not be amenable to its construction.
 
@@
 \label{subsec:stochastic_methods}
+An alternative approach to approximate matrix inversion is to employ
+Monte Carlo methods that sample a distribution with an expectation
+value equivalent to that of the inverted operator. Such methods have
+been in existence for decades with the earliest reference noted here
+an enjoyable 1950 manuscript by Forsythe and Leibler
+\citep{forsythe_1950}. In their outline, Forsythe and Liebler in fact
+credit the creation of this technique to J. Von Neumann and S.M. Ulam
+some years earlier than its publication. Hammersley and Handscomb's
+1964 monograph \citep{hammersley_1964} presents additional detail on
+this topic using a collection of references from the 1950's and early
+1960's.
+
+We begin our discussion of stochastic methods using these texts by
+seeking a solution to Eq~(\ref{eq:linear_problem}). For a given linear
+operator $\ve{A}$, we can define the following operator:
+\begin{equation}
+  \ve{H} = \ve{I} - \ve{A}\:.
+  \label{eq:linear_mc_iteration_matrix}
+\end{equation}
+We can then form an alternative representation for $\ve{A}^{-1}$ by
+generating the \textit{Neumann series}:
+\begin{equation}
+  \ve{A}^{-1} = (\ve{I}-\ve{H})^{-1} = \sum_{k=0}^{\infty} \ve{H}^k\:,
+  \label{eq:neumann_series}
+\end{equation}
+which will converge if the spectral radius of $\ve{H}$ is less than
+1. If we then apply this inverse sum to the right hand side of
+Eq~(\ref{eq:linear_problem}) we acquire the solution to the linear
+problem:
+\begin{equation}
+  \ve{A}^{-1}\ve{b} = \sum_{k=0}^{\infty} \ve{H}^k\ve{b} = \ve{x}\:.
+  \label{eq:neumann_solution}
+\end{equation}
+An approximation of this summation will therefore lead to an
+approximation of the solution. If we expand the summation with a
+succession of matrix-vector multiply operations, we arrive at an
+alternative perspective of this summation by considering its $i^{th}$
+component:
+\begin{equation}
+  x_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
+  \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}\:,
+  \label{eq:expanded_neumann_solution}
+\end{equation}
+which can interpreted as a series of transitions between states
+$i_{m-1}$ and $i_{m}$ in $\ve{H}$. We can generate these transitions
+through Monte Carlo random walks by assigning them both a probability
+and weight. Knowing this, we then form the \textit{Neumann-Ulam
+  decomposition} of \ve{H}:
+\begin{equation}
+  \ve{H} = \ve{P} \circ \ve{W}\:,
+  \label{eq:neumann_ulam_decomposition}
+\end{equation}
+where $\circ$ denotes the Schur product operation, $\ve{P}$ denotes
+the transition probability matrix, and $\ve{W}$ denotes the transition
+weight matrix. The formulation of $\ve{P}$ and $\ve{W}$ will be
+dependent on whether we choose a direct or adjoint Monte Carlo
+sequence to estimate the state transitions in
+Eq~(\ref{eq:expanded_neumann_solution}).
 
@@
 \label{subsubsec:direct_mc}
+In the context of matrix inversion, a direct method resembles an
+adjoint Monte Carlo method in the reactor physics community where the
+solution state is sampled and the source terms that contribute to it
+assembled. To achieve this, we build the direct Neumann-Ulam
+decomposition by first choosing a probability matrix that is a column
+scaling of $\ve{H}$ such that its components are:
+\begin{equation}
+  p_{ij} = \frac{|h_{ij}|}{\sum_j |h_{ij}|}\:.
+  \label{eq:direct_probability}
+\end{equation}
+From this, we then see that the probability of transitioning from a
+state $i$ to a state $j$ is implicitly linked to the original operator
+$\ve{A}$ in that those terms with large values, and therefore make the
+greatest contribution to the numerical solution, will be sampled with
+a higher probability than smaller terms. In addition, the column
+scaling provides a normalization over the state to which we are
+transitioning such that $\sum_j p_{ij} = 1$, meaning that we sample
+the probabilities over the columns of the matrix. The components of
+the weight matrix are then defined by
+Eq~(\ref{eq:neumann_ulam_decomposition}) as:
+\begin{equation}
+  w_{ij} = \frac{h_{ij}}{p_{ij}}\:.
+  \label{eq:direct_weight}
+\end{equation}
+It should be noted here that if $\ve{A}$ is sparse, then $\ve{H}$,
+$\ve{P}$, and $\ve{W}$ must be sparse as well by definition.
+
+Using these matrices, we can then form the expectation value of the
+solution.
 

### 2012-09-13 doc/jcomp2012/jcomp2012.tex

@@
 
-  \title{A Monte Carlo Synthetic-Acceleration Method for Solving the
-    Thermal Radiation Diffusion Equation}
-
-  \author[ornl]{Thomas M. Evans\corref{cor1}\fnref{fn1}}
-  \ead{evanstm@ornlgov} \ead[url]{www.ornl.gov}
+  \title{A Monte Carlo Synthetic-Acceleration Method for Solving the Thermal
+    Radiation Diffusion Equation\tnoteref{ornl-cr}}
+  \tnotetext[ornl-cr]{Notice: This manuscript has been authored by
+    UT-Battelle, LLC, under contract DE-AC05-00OR22725 with the
+    U.S. Department of Energy. The United States Government retains and the
+    publisher, by accepting the article for publication, acknowledges that the
+    United States Government retains a non-exclusive, paid-up, irrevocable,
+    world-wide license to publish or reproduce the published form of this
+    manuscript, or allow others to do so, for United States Government
+    purposes.}
+
+  \author[ornl]{Thomas M. Evans\corref{cor1}\fnref{fn1}} \ead{evanstm@ornlgov}
   \cortext[cor1]{Corresponding Author}
@@
 
-  \author[wisc]{Stuart R. Slattery\fnref{fn2}}
-  \ead{sslattery@wisc.edu} \ead[url]{www.wisc.edu}
+  \author[wisc]{Stuart R. Slattery\fnref{fn2}} \ead{sslattery@wisc.edu}
 
   \fntext[fn1]{Radiation Transport Group, Reactor and Nuclear Systems
-    Division} \fntext[fn2]{Department of Engineering Physics, College
-    of Engineering}
-
-  \address[ornl]{Oak Ridge National Laboratory, 1 Bethel Valley Rd.,
-    Oak Ridge, TN 37831 U.S.A.}  \address[wisc]{University of
-    Wisconsin-Madison, 1500 Engineering Dr., Madison, WI 53716 U.S.A.}
+    Division} \fntext[fn2]{Department of Engineering Physics, College of
+    Engineering}
+
+  \address[ornl]{Oak Ridge National Laboratory, 1 Bethel Valley Rd., Oak
+    Ridge, TN 37831 U.S.A.}  \address[wisc]{University of Wisconsin-Madison,
+    1500 Engineering Dr., Madison, WI 53716 U.S.A.}
 
@@
 %%---------------------------------------------------------------------------%%
+\section*{Acknowledgements}
+
+Work for this paper was supported by Oak Ridge National Laboratory, which is
+managed and operated by UT-Batelle, LLC, for the U.S. Department of Energy under
+Contract No. DEAC05-00OR22725.
+
+%%---------------------------------------------------------------------------%%
 %% The Appendices part is started with the command \appendix;

### 2012-09-14 doc/prelim/document/chapters/linear_problem/linear_problem.tex

@@
 \end{equation}
-where $\circ$ denotes the Schur product operation, $\ve{P}$ denotes
-the transition probability matrix, and $\ve{W}$ denotes the transition
-weight matrix. The formulation of $\ve{P}$ and $\ve{W}$ will be
-dependent on whether we choose a direct or adjoint Monte Carlo
-sequence to estimate the state transitions in
-Eq~(\ref{eq:expanded_neumann_solution}).
+where $\circ$ denotes the Schur product operation\footnote{The Schur
+  product $\ve{A} = \ve{B} \circ \ve{C}$ is defined element-wise as
+  $a_{ij} = b_{ij} c_{ij}$.}, $\ve{P}$ denotes the transition
+probability matrix, and $\ve{W}$ denotes the transition weight
+matrix. The formulation of $\ve{P}$ and $\ve{W}$ will be dependent on
+whether we choose a direct or adjoint Monte Carlo sequence to estimate
+the state transitions in Eq~(\ref{eq:expanded_neumann_solution}).
 

### 2012-09-14 doc/prelim/document/chapters/linear_problem/linear_problem.tex
updates to direct method section

--- 2012-09-13 doc/prelim/document/chapters/linear_problem/linear_problem.tex
@@
 Eq~(\ref{eq:eigenvalue_problem_2}) for a particular permutation of
-$\ve{x}$. The set of all permutations for the \textit{Eigenvectors} of
-$\ve{A}$. A quantity of particular interest that is computatable from
-the eigenvalues of a matrix is the \textit{spectral radius},
-$\rho(\ve{A})$ defined by Saad \citep{saad_2003} as:
+$\ve{x}$. The set of all permutations form the \textit{Eigenvectors}
+of $\ve{A}$. A quantity of particular interest that is computatable
+from the eigenvalues of a matrix $\ve{A}$ is the \textit{spectral
+  radius}, $\rho(\ve{A})$, defined by Saad \citep{saad_2003} as:
 \begin{equation}
@@
 Eq~(\ref{eq:matrix_norm_inequality}) is equivalent to $||\ve{A}||$
-that is valid $\forall \ve{x} \in \mathbb{R}^N$. The general
+and is valid $\forall \ve{x} \in \mathbb{R}^N$. The general
 definition in Eq~(\ref{eq:matrix_norm_inequality}) can be expanded in
@@
 \citep{leveque_2007}, finite volumes \citep{leveque_2002}, and finite
-elements \citep{zienkiewicz_1977} ultimately generate matrix problems
+elements \citep{zienkiewicz_1977} ultimately generates matrix problems
 as in Eq~(\ref{eq:linear_problem}).  In addition, these matrices are
@@
 of a particular grid element only expands as far as a few of its
-nearest neighbors depending on the order of discretization used, and
+nearest neighbors depending on the order of discretization used and
 therefore coupling among variables in a particular discrete equation
@@
 iterative techniques have been developed to solve such systems. We
-discuss here conventional stationary and projection methods to provide
-background and then give detail on a stochastic class of methods.
+discuss here conventional stationary and projection methods for
+solving sparse systems to provide background and then give detail on a
+stochastic class of methods.
 
@@
 developed in \S~(\ref{sec:linear_preliminaries}). By assuming $\ve{H}$
-is diagonalizable (although we may generalize this with the Jordan
-canonical form of $H$ \citep{saad_2003}, we then have:
+is diagonalizable\footnote{We may generalize this to
+  non-diagonalizable matrices with the Jordan canonical form of
+  $\ve{H}$.} \citep{saad_2003}, we then have:
 \begin{equation}
@@
 For iteration matrices where the Eigenvectors are orthogonal,
-$\kappa(\ve{R})=0$ and the error bound reduces to:
+$\kappa(\ve{R})=1$ and the error bound reduces to:
 \begin{equation}
@@
 Among the most common iterative methods used in scientific computing
-today for sparse systems are a broad class known as \textit{projection
-  methods}. These methods not only provide access to more powerful
-means of reaching a solution, but also a powerful means of
-encapsulating the majority of common iterative methods including the
-stationary methods just discussed in a common mathematical
+today for sparse systems are of a broad class known as
+\textit{projection methods}. These methods not only provide access to
+more powerful means of reaching a solution, but also a powerful means
+of encapsulating the majority of common iterative methods including
+the stationary methods just discussed in a common mathematical
 framework. All projection methods are built around a core structure
@@
 the residual that exists in $\mathcal{L}$. This then means that if our
-convergence criteria for an iterative method is bound the residual of
-the system, then Eq~(\ref{eq:linear_min_res}) tells us that each
+convergence criteria for an iterative method is bound to the residual
+of the system, then Eq~(\ref{eq:linear_min_res}) tells us that each
 projection step gaurantees us that the norm of the new residual will
@@
 are common among many Krylov methods. First, from a memory perspective
-GMRES is efficient from the standpoint that the operator $\ve{A}$ need
-not be explicitly stored. Rather, only the ability to compute the
-action of that operator on a vector of valid size is
-required. However, these savings in memory are balanced by the fact
-that the long recurrence relations used in the Arnoldi procedure
-require all vectors that span the Krylov space to be stored. If the
-size of these vectors becomes prohibitive, the Arnoldi procedure can
-be restarted at the cost of losing information in the
-orthogonalization process, creating the potentitial to generate new
-search directions that are not orthogonal to all previous search
-directions (and therefore less than optimal). From an implementation
-perspective, because the operator is not required to be formed, GMRES
-is significantly more flexible in its usage in that there are many
-instances where various processes serve to provide the action of that
-operator (e.g. radiation transport sweeps \citep{evans_2010}) that may
-not be amenable to its construction.
+GMRES is efficient in that the operator $\ve{A}$ need not be
+explicitly stored. Rather, only the ability to compute the action of
+that operator on a vector of valid size is required. However, these
+savings in memory are balanced by the fact that the long recurrence
+relations used in the Arnoldi procedure require all vectors that span
+the Krylov space to be stored. If the size of these vectors becomes
+prohibitive, the Arnoldi procedure can be restarted at the cost of
+losing information in the orthogonalization process, creating the
+potentitial to generate new search directions that are not orthogonal
+to all previous search directions (and therefore less than
+optimal). From an implementation perspective, because the operator is
+not required to be formed, GMRES is significantly more flexible in its
+usage in that there are many instances where various processes serve
+to provide the action of that operator (e.g. radiation transport
+sweeps \citep{evans_2010}) that may not be amenable to its
+construction.
 
@@
 been in existence for decades with the earliest reference noted here
-an enjoyable 1950 manuscript by Forsythe and Leibler
+an enjoyable manuscript published in 1950 by Forsythe and Leibler
 \citep{forsythe_1950}. In their outline, Forsythe and Liebler in fact
@@
 seeking a solution to Eq~(\ref{eq:linear_problem}). For a given linear
-operator $\ve{A}$, we can define the following operator:
+operator $\ve{A}$, we can split it in the same manner as
+Eq~(\ref{eq:linear_split_equation2}) to define the following operator:
 \begin{equation}
@@
 \end{equation}
-which can interpreted as a series of transitions between states
-$i_{m-1}$ and $i_{m}$ in $\ve{H}$. We can generate these transitions
-through Monte Carlo random walks by assigning them both a probability
-and weight. Knowing this, we then form the \textit{Neumann-Ulam
-  decomposition} of \ve{H}:
+which can interpreted as a series of transitions between states,
+\begin{equation}
+ \nu = i \rightarrow i_1 \rightarrow \cdots \rightarrow i_{k-1}
+ \rightarrow i_{k}\:,
+  \label{eq:mc_walk_permutation}
+\end{equation}
+in $\ve{H}$ where $\nu$ is a particular sequence permutation. We can
+generate these sequences of transitions through Monte Carlo random
+walks by assigning them both a probability and weight. As a
+reinterpretation of the iteration matrix, we then form the
+\textit{Neumann-Ulam decomposition} of \ve{H}:
 \begin{equation}
@@
 \end{equation}
-where $\circ$ denotes the Schur product operation, $\ve{P}$ denotes
-the transition probability matrix, and $\ve{W}$ denotes the transition
-weight matrix. The formulation of $\ve{P}$ and $\ve{W}$ will be
-dependent on whether we choose a direct or adjoint Monte Carlo
-sequence to estimate the state transitions in
-Eq~(\ref{eq:expanded_neumann_solution}).
+where $\circ$ denotes the Hadamard product operation\footnote{The Hadamard
+  product $\ve{A} = \ve{B} \circ \ve{C}$ is defined element-wise as
+  $a_{ij} = b_{ij} c_{ij}$.}, $\ve{P}$ denotes the transition
+probability matrix, and $\ve{W}$ denotes the transition weight
+matrix. The formulation of $\ve{P}$ and $\ve{W}$ will be dependent on
+whether we choose a direct or adjoint Monte Carlo sequence to estimate
+the state transitions in Eq~(\ref{eq:expanded_neumann_solution}).
 
@@
 solution state is sampled and the source terms that contribute to it
-assembled. To achieve this, we build the direct Neumann-Ulam
-decomposition by first choosing a probability matrix that is a column
-scaling of $\ve{H}$ such that its components are:
+are assembled. To achieve this, we build the direct method
+Neumann-Ulam decomposition by first choosing a probability matrix that
+is a column scaling of $\ve{H}$ such that its components are:
 \begin{equation}
@@
 state $i$ to a state $j$ is implicitly linked to the original operator
-$\ve{A}$ in that those terms with large values, and therefore make the
-greatest contribution to the numerical solution, will be sampled with
-a higher probability than smaller terms. In addition, the column
-scaling provides a normalization over the state to which we are
+$\ve{A}$ in that those terms with large values, and therefore those
+that make the greatest contribution to the numerical solution, will be
+sampled with a higher probability than smaller terms. In addition, the
+column scaling provides a normalization over the state to which we are
 transitioning such that $\sum_j p_{ij} = 1$, meaning that we sample
@@
 It should be noted here that if $\ve{A}$ is sparse, then $\ve{H}$,
-$\ve{P}$, and $\ve{W}$ must be sparse as well by definition.
+$\ve{P}$, and $\ve{W}$ must be sparse as well by
+definition. Additionally, we only compute $\ve{P}$ and $\ve{W}$ from
+the non-zero elements of $\ve{H}$ as those components that are zero
+will not participate in the random walk. Furthermore, it prevents an
+infinite weight from being generated in Eq~(\ref{eq:direct_weight}).
 
 Using these matrices, we can then form the expectation value of the
-solution.
+direct solution. For a given random walk permutation $\nu$ with $k$
+events, we define the weight of that permutation to be:
+\begin{equation}
+  W_{m} = \sum_{m=0}^k w_{i,i_1} w_{i_1,i_2} \cdots w_{i_{m-1},i_m}\:,
+  \label{eq:direct_permutation_weight}
+\end{equation}
+such that the weight of each transition event contributes to the
+total. The contribution to the solution from a particular random walk
+permutation is then:
+\begin{equation}
+  X_{\nu}(i_0 = i) = \sum_{m=0}^k W_{m} b_{i_m}\:,
+  \label{eq:direct_permutation_contribution}
+\end{equation}
+where $X_{\nu}(i_0 = i)$ signifies that the solution state in which we
+are tallying defines state $i$.  We can interpret this precisely as
+before in that during the random walk we collect the source in the
+states that are visited and apply them to the solution tally. We then
+define the probability that a particular random walk permuation of $k$
+events will occur:
+\begin{equation}
+  P_{\nu} = p_{i,i_1} p_{i_1,i_2} \cdots p_{i_{k-1},i_k}\:.
+  \label{eq:direct_permutation_probability}
+\end{equation}
+Finally, we define the expectation value of $X$ to be the collection
+of all random walk permutations and their probabilities:
+\begin{equation}
+  E\{X(i_0 = i)\} = \sum_{\nu} P_{\nu} X_{p\nu}\:,
+  \label{eq:direct_expectation_value}
+\end{equation}
+which, if expanded, directly recovers the exact solution:
+\begin{multline}
+  E\{X(i_0 = i)\}\\
+  =\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots \sum_{i_k}^{N}
+  p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
+  w_{i,i_1}w_{i_1,i_2}\ldots w_{i_{k-1},i_k} b_{i_k}\\ = x_i\:,
+  \label{eq:direct_expectation_expansion}
+\end{multline}
+therefore providing an unbiased Monte Carlo estimator.
+
+We can compute the variance of the estimator through traditional
+methods by defining the variance, $\sigma_i$, for each component in
+the solution:
+\begin{equation}
+  {\sigma_i}^2 = E\{X(i_0 = i) - (\ve{A}^{-1}\ve{b})_i\}^2 = E\{X(i_0
+  = i)^2\} - x_i^2\:,
+  \label{eq:direct_variance_1}
+\end{equation}
+where the vector exponentials are computed element-wise. Inserting
+Eq~(\ref{eq:direct_expectation_value}) gives:
+\begin{equation}
+  \sigma_i^2 = \sum_{\nu} P_{\nu} X_{\nu}^2 - x_i^2\:,
+  \label{eq:direct_variance_2}
+\end{equation}
+and applying Eq~(\ref{eq:direct_permutation_contribution}):
+\begin{equation}
+  \sigma_i^2 = \sum_{\nu} P_{\nu} \sum_{m=0}^k W_{m}^2 b_{i_m}^2 -
+  x_i^2\:.
+  \label{eq:direct_variance_3}
+\end{equation}
+Finally, expanding the transition probabilities yields:
+\begin{equation}
+  \sigma_i^2 = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
+  \sum_{i_k}^{N} p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
+  w^2_{i,i_1}w^2_{i_1,i_2}\ldots w^2_{i_{k-1},i_k} b_{i_m} - x_i^2\:.
+  \label{eq:direct_variance_4}
+\end{equation}
+Using this defintion for the variance, we can arrive at a more natural
+reason for enforcing $\rho(\ve{H}) < 1$. Per the Hadamard product, we can
+concatenate the summation in Eq~(\ref{eq:direct_variance_4}):
+\begin{equation}
+  (\ve{P} \circ \ve{W} \circ \ve{W})^k =
+  \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots \sum_{i_k}^{N}
+  p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
+  w^2_{i,i_1}w^2_{i_1,i_2}\ldots w^2_{i_{k-1},i_k}\:.
+  \label{eq:double_weighted_decomposition}
+\end{equation}
+If we assign $\ve{G} = \ve{P} \circ \ve{W} \circ \ve{W}$ as in
+Eq~(\ref{eq:neumann_ulam_decomposition}), we then have a new
+formulation for the variance:
+\begin{equation}
+  \sigma^2_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
+  \sum_{i_k}^{N}g_{i,i_1}g_{i_1,i_2}\ldots g_{i_{k-1},i_k} b_{i_k}^2 -
+  x_i^2\:,
+\end{equation}
+which contains the general Neumann series for $\ve{G}$,
+\begin{equation}
+  \ve{T} = \sum_{k=0}^{\infty} \ve{G}^k\:,
+  \label{eq:variance_neumann_series}
+\end{equation}
+where $\ve{T} = (\ve{I}-\ve{G})^{-1}$. We can then insert $T$ back
+into the variance formulation:
+\begin{equation}
+  \sigma^2_i = (\ve{T}\ve{b})_i - x_i^2
+  \label{eq:direct_variance_5}
+\end{equation}
+We can relate $\ve{G}$ to $\ve{H}$ by noting that $\ve{G}$ simply
+contains an additional Hadamard product with the weight matrix. The
+Hadamard product has the property that:
+\begin{equation}
+  |\ve{H} \circ \ve{W}| \geq |\ve{H}|\ |\ve{W}|\:,
+  \label{eq:hadamard_inequality}
+\end{equation}
+meaning that $\rho(\ve{G}) \geq \rho(\ve{H})$. Using these relations
+to analyze Eq~(\ref{eq:direct_variance_5}), we see that if
+$\rho(\ve{H}) > 1$, then the sum in
+Eq~(\ref{eq:variance_neumann_series}) will not converge and an
+infinite variance will arise as the elements of $\ve{T}$ become
+infinite. Therefore, we must restrict $\ve{H}$ with $\rho(\ve{H}) < 1$
+so that our expectation values for the solution may have a finite
+variance.
+
+As we seek only approximate solutions, we need only to perform a
+minimal number of random walks in order to generate an approximation
+for $\ve{x}$. Furthermore, we also need conditions by which we may
+terminate a random walk. We do this by noticing that the factors added
+to Eq~(\ref{eq:direct_permutation_weight}) will become diminishingly
+small due to their definition in Eq~(\ref{eq:direct_weight}) such that
+their contributions to the solution estimate will become
+negligble. Therefore, we choose terminate a random walk sequence with
+a \textit{weight cutoff}, $W_c$, that is enforced when $W_m < W_c$ for
+a particular random walk permutation.
 
@@
 
-\subsection{Monte Carlo Synthetic-Acceleration}
-\label{subsec:mcsa}
+\subsubsection{Monte Carlo Synthetic-Acceleration}
+\label{subsubsec:mcsa}
 

### 2012-09-15 doc/prelim/document/chapters/linear_problem/linear_problem.tex
finished stochastic solvers section. only preconditioning is left

@@
 Neumann-Ulam decomposition by first choosing a probability matrix that
-is a column scaling of $\ve{H}$ such that its components are:
+is a row scaling of $\ve{H}$ such that its components are:
 \begin{equation}
@@
 sampled with a higher probability than smaller terms. In addition, the
-column scaling provides a normalization over the state to which we are
+row scaling provides a normalization over the state to which we are
 transitioning such that $\sum_j p_{ij} = 1$, meaning that we sample
-the probabilities over the columns of the matrix. The components of
+the probabilities over the rows of the matrix. The components of
 the weight matrix are then defined by
@@
 \begin{equation}
-  E\{X(i_0 = i)\} = \sum_{\nu} P_{\nu} X_{p\nu}\:,
+  E\{X(i_0 = i)\} = \sum_{\nu} P_{\nu} X_{\nu}\:,
   \label{eq:direct_expectation_value}
@@
 which, if expanded, directly recovers the exact solution:
-\begin{multline}
-  E\{X(i_0 = i)\}\\
-  =\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots \sum_{i_k}^{N}
-  p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
-  w_{i,i_1}w_{i_1,i_2}\ldots w_{i_{k-1},i_k} b_{i_k}\\ = x_i\:,
+\begin{equation}
+  \begin{split}
+    E\{X(i_0 = i)\}
+    &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
+    \sum_{i_k}^{N} p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
+    w_{i,i_1}w_{i_1,i_2}\ldots w_{i_{k-1},i_k} b_{i_k}\\ &= x_i\:,
+  \end{split}
   \label{eq:direct_expectation_expansion}
-\end{multline}
+\end{equation}
 therefore providing an unbiased Monte Carlo estimator.
@@
 \label{subsubsec:adjoint_mc}
+An alternative formulation for Monte Carlo matrix inversion is the
+adjoint method. In this scheme, random walk histories are born in a
+state generated by source sampling and tallies are generated in the
+states in which the random walk currently resides. In terms of
+particle transport, this adjoint method is equivalent to a traditional
+forward method. To begin, we first form the \textit{adjoint
+  Neumann-Ulam decomposition} of $\ve{H}$:
+\begin{equation}
+  \ve{H}^T = \ve{P} \circ \ve{W}\:,
+  \label{eq:adjoint_neumann_ulam}
+\end{equation}
+where now we are forming the decomposition with respect to the
+transpose of $\ve{H}$. We then following the same procedure as the
+direct method for forming the probability and weight matrices in the
+decomposition. Using the transpose form, probabilities should instead
+by column-scaled:
+\begin{equation}
+  p_{ij} = \frac{|h_{ji}|}{\sum_j |h_{ji}|}\:,
+  \label{eq:adjoint_probability}
+\end{equation}
+such that we expect to select a new state $j$ from the current state
+in the random walk $j$ by sampling column-wise. Per
+Eq~(\ref{eq:adjoint_neumann_ulam}), the transition weight is then
+defined as:
+\begin{equation}
+  w_{ij} = \frac{h_{ji}}{p_{ij}}\:.
+  \label{eq:adjoint_weight}
+\end{equation}
+Using the decomposition we can then define an expectation value for
+the adjoint method. Given Eq~(\ref{eq:direct_permutation_weight}) as
+the weight generated for a particular random walk permutation as in
+Eq~(\ref{eq:mc_walk_permutation}) and the fact that in the adjoint
+method we are only tallying the solution in the state in which the
+walk currently resides, the contribution to the solution from a
+particular random walk permutation is then:
+\begin{equation}
+  X_{\nu} = \sum_{m=0}^k W_{m} b_{i_0} \delta_{i,i_m}\:,
+  \label{eq:adjoint_permutation_contribution}
+\end{equation}
+where the Kronecker delta indicates that the tally contributes only in
+the current state and $b_{i_0}$ is that sampled source starting
+state. Note here that the estimator in
+Eq~(\ref{eq:adjoint_permutation_contribution}) does not have a
+dependency on the source state as in
+Eq~(\ref{eq:adjoint_permutation_contribution}), providing a remedy for
+the situation in the direct method where we much start a random walk
+in each source state for every permutation such that we may compute a
+contribution for that state. In the adjoint method, we instead tally
+in all states and those of lesser importance will not be visited as
+frequently by the random walk. Finally, the estimator using all
+permutations is:
+\begin{equation}
+  E\{X\} = \sum_{\nu} P_{\nu} X_{\nu}\:
+  \label{eq:adjoint_expectation_value}
+\end{equation}
+which, if expanded in the same way as the direct method, directly
+recovers the exact solution:
+\begin{equation}
+  \begin{split}
+    E\{X\} &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
+    \sum_{i_k}^{N} p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
+    w_{i,i_1}w_{i_1,i_2}\ldots w_{i_{k-1},i_k} b_{i_k} \\ &= x_i\:,
+  \end{split}
+  \label{eq:adjoint_expectation_expansion}
+\end{equation}
+therefore also providing an unbiased Monte Carlo estimator.
+
+We also desire a criteria for random walk termination. For the adjoint
+method, we utilize a \textit{relative weight cutoff}:
+\begin{equation}
+  W_f = W_c b_{i_0}\:,
+  \label{eq:relative_weight_cutoff}
+\end{equation}
+where $W_c$ is defined as in the direct method. The adjoint random
+walk will then be terminated after $m$ steps if $W_m < W_f$ as tally
+contributions become increasingly small.
 
@@
 \label{subsusbsec:sequential_mc}
+The direct and adjoint methods described are limited by a convergence
+rate of $1/\sqrt{N}$ by the Central Limit Theorem where $N$ is the
+number of random walk permutations. In 1962, Halton presented a
+residual Monte Carlo method that moves towards exponential convergence
+rates \citep{halton_1962} and further refined his work some years
+later \citep{halton_1994}. Applications of his work by the transport
+community have confirmed convergence rates on the order of of
+$exp(-N)$ \citep{evans_2003}. In much the same way as projection
+methods, Halton's method, sequential Monte Carlo, utilizes the adjoint
+Monte Carlo solver as a means of driving down the residual vector. He
+proposed the following iterative scheme as a solution to
+Eq~(\ref{eq:linear_problem})\:
+\begin{subequations}
+  \begin{gather}
+    \ve{r}^k = \ve{b} - \ve{A}\ve{x}^k\:,\\  
+    \ve{A}\boldsymbol{\delta}^{k+1} = \ve{r}^{k}\:,\\
+    \ve{x}^{k+1} = \ve{x}^k + \boldsymbol{\delta}^{k+1}\:,
+  \end{gather}
+  \label{eq:sequential_monte_carlo}
+\end{subequations}
+where the correction $\boldsymbol{\delta}$ is computed by the adjoint
+Monte Carlo method. The merits of Halton's approach are immediately
+visible in that we have now broken the binding of the convergence rate
+to the Central Limit Thereom. Here, the Monte Carlo solve is used to
+produce a correction from the residual, analagous to using the
+residual to extract a correction from the search subspace in a
+projection method. By doing this, all of the Monte Carlo error is
+bound in the correction used to update the solution and therefore does
+not explicitly manifest itself in the solution. The downside of such a
+method is that if the solution guess is poor, then many iterations are
+required in order to reach exponential converge as the Monte Carlo
+error (and therefore the Central Limit Theorem) does dominate in this
+situation. Therefore, if a good initial guess is not available,
+Halton's method is still characterized by poor performance.
 
@@
 \label{subsubsec:mcsa}
+Using the ideas of Halton, Evans and Mosher recently developed a Monte
+Carlo solution that was not prohibited severely by the quality of the
+initial guess for the system \citep{evans_2009}. Their approach was
+instead to use residual Monte Carlo as a synthetic acceleration for a
+stationary method. To derive this method, we begin by splitting the
+operator in Eq~(\ref{eq:linear_problem}) 
+\begin{equation}
+  \ve{x} = (\ve{I} - \ve{A})\ve{x} + \ve{b}\:.
+  \label{eq:linear_split}
+\end{equation}
+With this we can then define the stationary method
+\textit{Richardson's iteration} as:
+\begin{equation}
+  \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,
+  \label{eq:richardsons_iteration}
+\end{equation}
+which will converge if $\rho(\ve{I} - \ve{A}) < 1$. We then define the
+solution error at the $k^{th}$ iterate relative to the true solution:
+\begin{equation}
+  \delta \ve{x}^k = \ve{x} - \ve{x}^k\:.
+  \label{eq:mcsa_error}
+\end{equation}
+Subtracting Eq~(\ref{eq:richardsons_iteration}) from
+Eq~(\ref{eq:linear_split}) we get:
+\begin{equation}
+  \delta \ve{x}^{k+1} = (\ve{I} - \ve{A})\delta \ve{x}^k\:.
+  \label{eq:mcsa_setup_1}
+\end{equation}
+Subtracting from this $(\ve{I} - \ve{A})\delta \ve{x}^{k+1}$ yields:
+\begin{equation}
+  \begin{split}
+    \ve{A}\delta \ve{x}^{k+1} &= (\ve{I} -
+    \ve{A})(\ve{x}^{k+1}-\ve{x}^{k}) \\ &= \ve{r}^{k+1}\:.
+    \label{eq:mcsa_setup_2}
+  \end{split}
+\end{equation}
+Using this, we define the following scheme that will converge in one
+iteration if $\ve{A}$ is inverted exactly.
+\begin{subequations}
+  \begin{gather}
+    \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,\\
+    \ve{A} \delta \ve{x}^{k+1} = \ve{r}^{k+1}\:,\\
+    \ve{x} = \ve{x}^{k+1} + \delta \ve{x}^{k+1}\:.
+  \end{gather}
+  \label{eq:mcsa_setup_3}
+\end{subequations}
+However, $\ve{A}$ is only approximately inverted by our numerical
+methods and therfore we instead pose an iterative scheme in which the
+Monte Carlo solvers are used to invert the operator. The
+\textit{Fixed-Point Monte Carlo Synthetic-Acceleration} (MCSA) method
+is defined as:
+\begin{subequations}
+  \begin{gather}
+    \ve{x}^{k+1/2} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,\\
+    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\
+    \ve{A}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:,\\
+    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,
+    \label{eq:mcsa}
+  \end{gather}
+\end{subequations}
+where the adjoint Monte Carlo method is used to generate the solution
+correction from the residual. In addition to the Monte Carlo solver
+parameters dictating the number of histories and weight cutoff, the
+outer MCSA iterations also have the following stopping criteria:
+\begin{equation}
+  ||\ve{r}||_\infty < \epsilon \ ||\ve{b}||_\infty\:,
+  \label{eq:mcsa_stopping_criteria}
+\end{equation}
+where $\epsilon$ is a user-defined parameter.
 

### 2012-09-16 doc/prelim/document/chapters/linear_problem/linear_problem.tex
added section on preconditioning stochastic methods

@@
 
-\section{Preconditioning the Linear System}
-\label{sec:linear_preconditioning}
+\subsubsection{Preconditioning Stochastic Methods}
+\label{subsubsec:stochastic_preconditioning}
+In most cases, at least a minimal amount of \textit{preconditioning}
+of the linear system will be required in order to use the class of
+stochastic methods described. Although these methods have no symmetry
+requirements for convergence, they do require that the spectral radius
+of the iteration matrix be less than one. To achieve this, a Jacobi
+precondtioner, a form of left preconditioning, is used such that the
+preconditioning matrix $\ve{D}$ is:
+\begin{equation}
+  \ve{M} = diag(\ve{A})\:,
+  \label{eq:jacobi_preconditioner}
+\end{equation}
+such that its application means we are instead solving the following
+linear system:
+\begin{equation}
+  \ve{M}^{-1}\ve{A}\ve{x} = \ve{M}^{-1}\ve{b}\:.
+  \label{eq:jacobi_precond_linear_problem}
+\end{equation}
+Next, we can apply MCSA to solve
+Eq~(\ref{eq:jacobi_precond_linear_problem}): 
+\begin{subequations}
+  \begin{gather}
+    \ve{x}^{k+1/2} = (\ve{I} - \ve{M}^{-1}\ve{A})\ve{x}^k +
+    \ve{M}^{-1}\ve{b}\:,\\ \ve{r}^{k+1/2} = \ve{M}^{-1}\ve{b} -
+    \ve{M}^{-1}\ve{A}\ve{x}^{k+1/2}\:,\\ \ve{M}^{-1}\ve{A}\delta\ve{x}^{k+1/2}
+    = \ve{r}^{k+1/2}\:,\\ \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta
+    \ve{x}^{k+1/2}\:.
+    \label{eq:jacobi_preconditioned_mcsa}
+  \end{gather}
+\end{subequations}
+Choosing Jacobi preconditioning with MCSA is advantageous for several
+reasons. First, $\rho(\ve{I} - \ve{M}^{-1}\ve{A}) < 1$ is true for
+all $\ve{A}$ and is easy to formulate as the inversion of $\ve{M}$ is
+trivial. Second, because the adjoint Monte Carlo method used within MCSA to
+compute the correction operates on a linear problem with the
+preconditioned operator, then $\ve{H}$ in the adjoint solver will have
+a zero term in each of its diagonal elements, thereby eliminating
+all in-state transitions during the random walk sequence. Because of
+this, Jacobi preconditioning should always be performed, regardless of
+any other preconditioning that is applied to the system.

### 2012-09-17 doc/prelim/document/chapters/linear_problem/linear_problem.tex
prelim updates

@@
 \end{equation}
-where $\ve{A} \in \mathbb{R}^{N \times N}$ is a matrix operator such
-that $\ve{A} : \mathbb{R}^{N} \rightarrow \mathbb{R}^{N}$, $\ve{x} \in
-\mathbb{R}^N$ is the solution vector, and $\ve{b} \in \mathbb{R}^N$ is
+where $\ve{A} \in \mathbb{C}^{N \times N}$ is a matrix operator such
+that $\ve{A} : \mathbb{C}^{N} \rightarrow \mathbb{C}^{N}$, $\ve{x} \in
+\mathbb{C}^N$ is the solution vector, and $\ve{b} \in \mathbb{C}^N$ is
 the forcing term. The solutions to Eq~(\ref{eq:linear_problem}) will
@@
   ||\ve{v}||_q = \Bigg[ \sum_{i=1}^N |v_i|^q \Bigg]^{1/q},\ \ve{v} \in
-  \mathbb{R}^N\:,\ q \in \mathbb{Z}^+
+  \mathbb{C}^N\:,\ q \in \mathbb{Z}^+
   \label{eq:q_norm}
@@
 Eq~(\ref{eq:matrix_norm_inequality}) is equivalent to $||\ve{A}||$
-and is valid $\forall \ve{x} \in \mathbb{R}^N$. The general
+and is valid $\forall \ve{x} \in \mathbb{C}^N$. The general
 definition in Eq~(\ref{eq:matrix_norm_inequality}) can be expanded in
@@
 precondtioner, a form of left preconditioning, is used such that the
-preconditioning matrix $\ve{D}$ is:
+preconditioning matrix $\ve{M}$ is:
 \begin{equation}

### 2012-09-17 doc/prelim/document/chapters/parallel_mcsa/parallel_mcsa.tex
updates to linear systems chapter
updates to prelim, projection methods section is complete
prelim updates

@@
-\chapter{Parallelization of the MCSA Algorithm}
+\chapter{Parallelization of the MCSA Algorithm for Linear Problems}
 \label{ch:parallel_mcsa}
@@
 \label{sec:current_parallel_methods}
+Modern parallel implementations of linear solver methods rely heavily
+on capabilities provided by general linear algebra frameworks. In
+particular, Krylov methods that are formulated through matrix-vector
+multiply operations, are easily formulated in parallel using these
+frameworks.
 

### 2012-09-17 doc/prelim/document/chapters/introduction/introduction.tex
cleaning up chapter structure in prelim

@@
 problems and a novel stochastic method to advance solution techniques
-for nonlinear problems with the aim of improving the modeling of
-coupled physics systems for nuclear reactor analysis.
+for nonlinear problems.
 
@@
 parallelism.
+
+\section{The Multiphysics System}
+\label{sec:multiphysics_system}
+
+\section{Research Outline}
+\label{sec:research_outline}

### 2012-09-17 doc/prelim/document/chapters/linear_problem/linear_problem.tex
cleaning up chapter structure in prelim

@@
 any other preconditioning that is applied to the system.
+
+\section{Parallel Methods for Solving Sparse Linear Systems}
+\label{sec:parallel_krylov_methods}
+Modern parallel implementations of Krylov methods on distributed
+memory architectures rely heavily on capabilities provided by general
+linear algebra frameworks. This arises from the fact that Krylov
+methods require only a handful of operations in their
+implementation. Per Saad's text \citep{saad_2003}, these operations
+are preconditioning, matrix-vector multiplications, vector updates,
+and dot products. For the last three items, linear algebra libraries
+such as PETSc \citep{petsc_2012} and Trilinos \citep{trilinos_2005}
+provide efficient parallel implementations for these
+operations. Depending on the type of preconditioning used, efficient
+parallel implementations may also be available for those
+operations. In all cases, a series of scatter/gather operations are
+required such that global communication operations must
+occur. Therefore, the performance of such operations is bound to the
+implementation. However, asymptotically perfomance should be the same
+across all implementations.
+
+\section{Parallelization of the MCSA Algorithm}
+\label{section:parallel_mcsa}
+For the prelim, I suspect that I will not have a parallel MCSA
+algorithm up and running as I will be focusing on generating the code
+base needed to build the physics operators and tools for linear and
+nonlinear systems analysis. Therefore, here we should focus on
+motivating a parallel scheme for this work. We should start by first
+looking at current parallel methods for solving linear systems using
+the conventional techniques from the previous chapter. It must be
+clearly pointed out that for the new hardware coming online, these
+methods (or at least their current implementations) will be lacking in
+scalability and the potential for resiliency, and therefore a parallel
+MCSA is attractive, even if it ultimately does not show performance
+gains from a wall time perspective.
+
+Once we have done this, MSOD will need to be defined. The literature
+is actually quite lacking here even for reactor physics problems as
+this is really bleeding edge. Therefore, I should strive here to more
+rigorously define this methodology in terms of solving general linear
+systems. Furthermore, we should discuss the aspects of being able to
+vary the overlap such that we can generate a spectrum between complete
+domain decomposition and complete domain replication. Finally, the
+beginnings of the algorithm should be sketched out.
+
+\subsection{Multiple-Set Overlapping-Domain Decomposition}
+\label{subsec:msod}

### 2012-09-17 doc/prelim/document/chapters/nonlinear_problem/nonlinear_problem.tex
adding references
updates to prelim, projection methods section is complete
cleaning up chapter structure in prelim

@@
 
-\section{Formulation of the Nonlinear System}
-\label{sec:nonlinear_system}
+\section{Preliminaries}
+\label{sec:nonlinear_preliminaries}
 
@@
 
-\subsection{General Newton Methods}
+\subsection{Inexact Newton Methods}
 \label{subsec:newton_methods}
@@
 
-\section{Preconditioning the Nonlinear System}
-\label{sec:nonlinear_preconditioning}
+\section{The FANM Method}
+\label{sec:fanm}
+In this section we will devise the FANM method. Based on the
+background in the previous section, we first need to motivate the
+development of this work by pointing out some of the issues with
+conventional methods. This includes JFNK coarseness and either
+prohibitively large Krylov subspaces or a loss of orthogonality
+information in the subspace due to restarts. We can than point out
+some of the attractive qualities of the FANM method. Much of this will
+rely on advanced concepts including automatic differentiation and
+on-the-fly residual and Jacobian generation from nonlinear function
+evaluations. In addition, we will want to show that the Newton method
+will converge using MCSA as the linear solver. I suspect I will have
+some preliminary results for a serial implementation here as I already
+have a serial MCSA implemented. Finally, we will want to design the
+model problems that we will test FANM with. These should be problems
+that have already been worked up in the literature with benchmark-type
+solutions.
 
-\section{Time Integration}
-\label{sec:time_integration}
+\subsection{Automatic Differentiation}
+\label{sec:automatic_differentiation}
 
-\section{Applications to Model Problems}
-\label{sec:nonlinear_applications}
+\subsection{Residual and Jacobian Generation}
+\label{sec:fanm_generation}
+
+\subsection{Jacobian Storage vs. Subspace Storage and Restarts}
+\label{sec:fanm_storage}

### 2012-09-18 doc/prelim/document/chapters/linear_problem/linear_problem.tex
updates to linear problem chapter

@@
 Newton-Krylov methods almost always need some type of
-preconditioning. Therefore, we must also discuss these aspects.
+preconditioning. Therefore, we must also discuss these
+aspects.\footnote{The contents of this chapter, particularly those
+  sections relating to projection methods, are heavily based on Saad's
+  text \citep{saad_2003}.}
 
@@
 $\ve{V}_m$ is found, per the constraint subspace definition it then
-follows that its basis is defined as $\ve{W}_m = \ve{A} \ve{V}_m$.
+follows that its basis is defined as $\ve{W}_m = \ve{A}
+\ve{V}_m$. Knowing the projections onto the search and constraint
+subspaces, the GMRES iteration may be formulated as follows:
+\begin{algorithm}[htpb!]
+  \caption{GMRES Iteration}
+  \label{alg:gmres}
+  \begin{algorithmic}
+    \State $\ve{r}_0 := \ve{b}-\ve{A}\ve{x}_0$
+    \State $\beta := ||\ve{r}_0||_2$
+    \State $\ve{v}_1 := \ve{r}_0 / \beta$
+    \Comment{Create the orthonormal basis for the Krylov subspace}
+    \For{$j = 1, 2, \cdots, m$}
+    \State $h_{ij} \leftarrow \langle w_j,v_j \rangle$
+    \State $w_j \leftarrow w_j - h_{ij}v_i$
+    \EndFor
+    \State $h_{j+1,j} \leftarrow ||w_j||_2$
+    \State $v_{j+1} \leftarrow w_j / h_{j+1,j}$
+    \Comment{Apply the orthogonality constraints}
+    \State $\ve{y}_m \leftarrow argmin_y ||\beta \ve{e}_1 - \ve{H}_m\
+    \ve{y}||_2 $
+    \State $\ve{x}_m \leftarrow \ve{x}_0 + \ve{V_m} \ve{y}_m$
+  \end{algorithmic}
+\end{algorithm}
 
-Knowing the projections onto the search and constraint subspaces, the
-GMRES method may be formulated. We note here several properties of
-this formulation and how they may facilitate or hinder the solution of
-large-scale, sparse linear problems, also noting that these properties
-are common among many Krylov methods. First, from a memory perspective
-GMRES is efficient in that the operator $\ve{A}$ need not be
-explicitly stored. Rather, only the ability to compute the action of
-that operator on a vector of valid size is required. However, these
-savings in memory are balanced by the fact that the long recurrence
-relations used in the Arnoldi procedure require all vectors that span
-the Krylov space to be stored. If the size of these vectors becomes
-prohibitive, the Arnoldi procedure can be restarted at the cost of
-losing information in the orthogonalization process, creating the
-potentitial to generate new search directions that are not orthogonal
-to all previous search directions (and therefore less than
-optimal). From an implementation perspective, because the operator is
-not required to be formed, GMRES is significantly more flexible in its
-usage in that there are many instances where various processes serve
-to provide the action of that operator (e.g. radiation transport
-sweeps \citep{evans_2010}) that may not be amenable to its
-construction.
+We note here several properties of this formulation and how they may
+facilitate or hinder the solution of large-scale, sparse linear
+problems, also noting that these properties are common among many
+Krylov methods. First, from a memory perspective GMRES is efficient in
+that the operator $\ve{A}$ need not be explicitly stored. Rather, only
+the ability to compute the action of that operator on a vector of
+valid size is required. However, these savings in memory are balanced
+by the fact that the long recurrence relations used in the Arnoldi
+procedure require all vectors that span the Krylov space to be
+stored. If the size of these vectors becomes prohibitive, the Arnoldi
+procedure can be restarted at the cost of losing information in the
+orthogonalization process, creating the potentitial to generate new
+search directions that are not orthogonal to all previous search
+directions (and therefore less than optimal). From an implementation
+perspective, because the operator is not required to be formed, GMRES
+is significantly more flexible in its usage in that there are many
+instances where various processes serve to provide the action of that
+operator (e.g. radiation transport sweeps \citep{evans_2010}) that may
+not be amenable to its construction.
 
@@
 Eq~(\ref{eq:adjoint_permutation_contribution}), providing a remedy for
-the situation in the direct method where we much start a random walk
+the situation in the direct method where we must start a random walk
 in each source state for every permutation such that we may compute a
@@
 contributions become increasingly small.
+
+Given these ideas, we can then generate an algorithm for implementing
+the adjoint method:
+\begin{algorithm}[htpb!]
+  \caption{Adjoint Monte Carlo}
+  \label{alg:adjoint_mc}
+  \begin{algorithmic}
+    \State $\ve{r}_0 := \ve{b}-\ve{A}\ve{x}_0$
+  \end{algorithmic}
+\end{algorithm}
 
@@
 \label{sec:parallel_krylov_methods}
-Modern parallel implementations of Krylov methods on distributed
-memory architectures rely heavily on capabilities provided by general
-linear algebra frameworks. This arises from the fact that Krylov
-methods require only a handful of operations in their
-implementation. Per Saad's text \citep{saad_2003}, these operations
-are preconditioning, matrix-vector multiplications, vector updates,
-and dot products. For the last three items, linear algebra libraries
-such as PETSc \citep{petsc_2012} and Trilinos \citep{trilinos_2005}
-provide efficient parallel implementations for these
-operations. Depending on the type of preconditioning used, efficient
-parallel implementations may also be available for those
-operations. In all cases, a series of scatter/gather operations are
-required such that global communication operations must
-occur. Therefore, the performance of such operations is bound to the
-implementation. However, asymptotically perfomance should be the same
-across all implementations.
+Modern parallel implementations of methods on distributed memory
+architectures rely heavily on capabilities provided by general linear
+algebra frameworks. For methods like GMRES, this arises from the fact
+that Krylov methods require only a handful of operation types in their
+implementation that can be efficiently programmed on these
+architectures. Per Saad's text \citep{saad_2003} and as noted in
+Algorithm~\ref{alg:gmres}, these operations are preconditioning,
+matrix-vector multiplications, vector updates, and inner products. For
+the last three items, linear algebra libraries such as PETSc
+\citep{petsc_2012} and Trilinos \citep{trilinos_2005} provide
+efficient parallel implementations for these operations. Depending on
+the type of preconditioning used, efficient parallel implementations
+may also be available for those operations. In all cases, a series of
+scatter/gather operations are required such that global communication
+operations must occur. Although the relative performance of such
+operations is bound to the implementation, asymptotically perfomance
+should be the same across all implementations.
 

### 2012-09-18 doc/prelim/document/chapters/nonlinear_problem/nonlinear_problem.tex
updates to linear problem chapter

@@
 focus here should be on fully-implicit nonlinear solves that are
-amenable to solving tightly-coupled multi-physics systems in a way that
+amenable to solving tightly-coupled multiphysics systems in a way that
 is fully consistent.
@@
 as these will serve as the baseline for assessing the qualities of a
-FANM solver.
-
-Finally, as we will choose transient problems for our models, time
-integration in the context of nonlinear systems will be consider and
-we will survey applications to model problems in the field ( there is
-a vast amount of literature on fully-implicit nonlinear solves for the
-Navier-Stokes equations) that will give good benchmarks to test (such
-as natural convection and driven cavity problems).
+FANM solver. Finally, we will survey applications to model problems in
+the field ( there is a vast amount of literature on fully-implicit
+nonlinear solves for the Navier-Stokes equations) that will give good
+benchmarks to test (such as natural convection and driven cavity
+problems).
 

### 2012-09-18 doc/prelim/document/chapters/linear_problem/linear_problem.tex
working on parallel krylov section

@@
 \label{ch:linear_problem}
-
-The purpose of this chapter is to provide a full background on solving
-linear systems as it relates to this work. It may seem like there is a
-lot here, but we really do need it to fully define our new methods and
-compare both qualitatively and quantitatively to convential
-methods. For example, we need a solid explanation of stationary
-methods because MCSA is in fact an acceleration of a stationary method
-and therefore shares many properties with them. Subspace methods are
-the current class of methods most widely used for sparse systems and
-are at the core of the Newton-Krylov methods that we will be comparing
-the FANM method with. The most rigorous piece of this chapter should
-of course be the stochastic solver definitions, also providing me a
-place to include some of my work in this area. Finally,
-preconditioning is important for this work, MCSA requires it and
-Newton-Krylov methods almost always need some type of
-preconditioning. Therefore, we must also discuss these
-aspects.\footnote{The contents of this chapter, particularly those
-  sections relating to projection methods, are heavily based on Saad's
-  text \citep{saad_2003}.}
+The discretization of partial differential equations (\textit{PDEs})
+through common methods such as finite differences
+\citep{leveque_2007}, finite volumes \citep{leveque_2002}, and finite
+elements \citep{zienkiewicz_1977} ultimately generates sets of coupled
+equations in the form of matrix problems.  In addition, these matrices
+are typically sparse, meaning that the vast majority of their
+constituent elements are zero. This sparsity is due to the fact that
+the influence of a particular grid element only expands as far as a
+few of its nearest neighbors depending on the order of discretization
+used and therefore coupling among variables in a particular discrete
+equation in the system leads to a few non-zero entries. Because of the
+natural occurrence of sparse matrices in common numerical methods many
+iterative techniques have been developed to solve such systems. We
+discuss here conventional stationary and projection methods for
+solving sparse systems to provide background and then give detail on a
+reemerging stochastic class of methods. Details on the parallelization
+of conventional methods are discussed and a novel proposal for the
+parallelization of stochastic methods presented.\footnote{The contents
+  of this chapter, particularly those sections relating to projection
+  methods and matrix analysis, are heavily based on Saad's text
+  \citep{saad_2003}.}
 
@@
 \end{equation}
-such that an exact solution to $\ve{x}$ has been found when
+such that an exact solution $\ve{x}$ has been found when
 $\ve{r}=\ve{0}$.  From the statement in
@@
 \begin{equation}
-  ||\ve{A}||_{2} = \sqrt{\rho(\ve{A}^T\ve{A})}\:,
+p  ||\ve{A}||_{2} = \sqrt{\rho(\ve{A}^T\ve{A})}\:,
   \label{eq:matrix_2_norm}
@@
 Knowing this, we can then define several useful properties of matrices
-including the \textit{condition number} \citep{saad_2003}:
+including the \textit{condition number}:
 \begin{equation}
@@
 
-\section{Iterative Methods for Solving Sparse Linear Systems}
-\label{sec:linear_methods}
-The discretization of partial differential equations (\textit{PDEs})
-through common methods such as finite differences
-\citep{leveque_2007}, finite volumes \citep{leveque_2002}, and finite
-elements \citep{zienkiewicz_1977} ultimately generates matrix problems
-as in Eq~(\ref{eq:linear_problem}).  In addition, these matrices are
-typically sparse, meaning that the vast majority of their constituent
-elements are zero. This sparsity is due to the fact that the influence
-of a particular grid element only expands as far as a few of its
-nearest neighbors depending on the order of discretization used and
-therefore coupling among variables in a particular discrete equation
-in the system leads to a few non-zero entries. Because of the natural
-occurrence of sparse matrices in common numerical methods many
-iterative techniques have been developed to solve such systems. We
-discuss here conventional stationary and projection methods for
-solving sparse systems to provide background and then give detail on a
-stochastic class of methods.
-
-\subsection{Stationary Methods}
-\label{subsec:stationary_methods}
+\section{Stationary Methods}
+\label{sec:stationary_methods}
 Stationary methods arise from splitting the operator in
@@
 
-\subsection{Projection Methods}
-\label{subsec:projection_methods}
+\section{Projection Methods}
+\label{sec:projection_methods}
 Among the most common iterative methods used in scientific computing
@@
     \ve{r}^k = \ve{b} - \ve{A}\ve{x}^k\:,\\  
-    \ve{y} = (\ve{W}^T\ve{A}\ve{V})^{-1}\ve{W}^T\ve{r}_0\:,\\
-    \ve{x}^{k+1} = \ve{x}^k + \ve{V}\ve{y}\:,
+    \ve{y}^k = (\ve{W}^T\ve{A}\ve{V})^{-1}\ve{W}^T\ve{r}_0\:,\\
+    \ve{x}^{k+1} = \ve{x}^k + \ve{V}\ve{y}^k\:,
   \end{gather}
@@
 
-\subsubsection{Krylov Subspace Methods}
-\label{subsubsec:krylov_methods}
+\subsection{Krylov Subspace Methods}
+\label{subsec:krylov_methods}
 Among the most common projection techniques used in practice are a
@@
 instances where various processes serve to provide the action of that
-operator (e.g. radiation transport sweeps \citep{evans_2010}) that may
-not be amenable to its construction.
+operator (e.g. radiation transport sweeps \citep{evans_2010}) that
+normally may not be amenable to its full construction.
 
-\subsection{Stochastic Methods}
-\label{subsec:stochastic_methods}
+\subsection{Parallel Projection Methods}
+\label{subsec:parallel_krylov_methods}
+Modern parallel implementations of projection methods on distributed
+memory architectures rely heavily on capabilities provided by general
+linear algebra frameworks. For methods like GMRES, this arises from
+the fact that Krylov methods require only a handful of operation types
+in their implementation that can be efficiently programmed on these
+architectures. Per Saad's text \citep{saad_2003} and as noted in
+Algorithm~\ref{alg:gmres}, these operations are preconditioning,
+matrix-vector multiplications, vector updates, and inner products. For
+the last three items, linear algebra libraries such as PETSc
+\citep{petsc_2012} and Trilinos \citep{trilinos_2005} provide
+efficient parallel implementations for these operations. Depending on
+the type of preconditioning used, efficient parallel implementations
+may also be available for those operations. In all cases, a series of
+scatter/gather operations are required such that global communication
+operations must occur. Although the relative performance of such
+operations is bound to the implementation, asymptotically perfomance
+should be the same across all implementations.
+
+We will look at the three primary parallel matrix/vector operations as
+preconditioning is not an immediate requirement for implementing the
+algorithms. Furthermore, variants are available that reduce the number
+of global communications required (consider \cite{sosonkina_1998} as
+an example of reducing global operation counts using a different
+orthogonalization procedure than Arnoldi), however, we will only
+consider the basic algorithms here as this handful operations can be
+generalized to fit more complicated algorithms. In all of these cases,
+we assume a general matrix/vector formulation that is distributed in
+parallel such that both local and global knowledge of their
+decomposition is available on request. Furthermore, it is assumed that
+these objects are partitioned in such a way that the parallel
+formulation of the operator and vectors in
+Eq~(\ref{eq:linear_problem}) will be such that each parallel process
+contains only a subset of the global problem and that subset forms a
+local set of complete equations.
+
+\subsubsection{Parallel Vector Update}
+\label{subsubsec:parallel_vec_update}
+Parallel vector update operations arise from the construction of the
+orthonormal basis and the application of the correction generated by
+the constraints to the solution vector. Vector update operations are
+embarassingly parallel in that they require no communication
+operations to be successfully completed; all data operated on is
+local. These operations are globally of the form:
+\begin{equation}
+  \ve{y}[n] \leftarrow \ve{y}[n] + a * \ve{x}[n],\ \forall n \in [1,N_g]
+  \:,
+  \label{eq:global_vector_update}
+\end{equation}
+and locally of the form:
+\begin{equation}
+  \ve{y}[n] \leftarrow \ve{y}[n] + a * \ve{x}[n],\ \forall n \in [1,N_l]
+  \:,
+  \label{eq:local_vector_update}
+\end{equation}
+where $\ve{y}$ and $\ve{x}$ are vectors of global size $N_g$, local
+size $N_l$, and $a \in \mathbb{C}^N$. In order to avoid communcation,
+the vectors $\ve{y}$ and $\ve{x}$ must have the same parallel
+decomposition where each parallel process owns the same pieces of each
+vector.
+
+\subsubsection{Parallel Vector Product}
+\label{subsubsec:parallel_vector_product}
+Vector product operations are used in several instances during a
+Krylov iteration including vector norm computations and the
+orthogonalization procedure. By definition, the vector product is a
+global operation that effectively collapses two vectors to a single
+value. Therefore, we cannot eliminate all global
+communications. Instead, vector product operations are formulated as
+\textit{global reduction operations} that are efficiently supported by
+modern message passing libraries. For the dot product of two vectors
+$\ve{y}$ and $\ve{x}$, a single reduction is required such that:
+\begin{equation}
+  d_l = \ve{y}_l \cdot \ve{x}_l,\ d_g = \sum_p d_l \:,  
+  \label{eq:parallel_dot_product}
+\end{equation}
+where the $l$ subscript denotes a local quantity, $d_l$ is the local
+vector dot product, and $d_g$ is the global dot product generated by
+summing the local dot products over all $p$ processes. Parallel norm
+operations can be conducted with the same single reduction. Consider
+the infinity norm operation:
+\begin{subequations}
+  \begin{gather}
+    ||x||_{\infty,l} = \max_n \ve{y}[n],\ \forall n \in [1,N_l]\:\\
+    ||x||_{\infty,g} = \max_p ||x||_{\infty,g}\:.
+  \end{gather}
+  \label{eq:parallel_infinity_norm}
+\end{subequations}
+In this form, the local infinity norm is computed over the local piece
+of the vector. The reduction operation is then formed over all $p$
+processes such that the global max of the vector is computed and
+distributed to all processes.
+
+\subsubsection{Parallel Matrix-Vector Multiplications}
+\label{subsubsec:parallel_mat_vec_mutliply}
+We finally consider matrix-vector multiplication operations by
+considering sparse matrices in a compressed storage format. For these
+operations, considerably more communications will be required given
+that the entire global vector is required in order to compute a single
+element of the product vector.
+
+\subsubsection{Parallel Perfomance Implications for Projection Methods}
+\label{subsubsec:projection_method_performance}
+For very large distributed machines, global reduction operations
+required at several levels of Krylov algorithms stand to reduce
+scalablitity and perfomance.
+
+\section{Stochastic Methods}
+\label{sec:stochastic_methods}
 An alternative approach to approximate matrix inversion is to employ
@@
 
-\subsubsection{Direct Method}
-\label{subsubsec:direct_mc}
+\subsection{Direct Method}
+\label{subsec:direct_mc}
 In the context of matrix inversion, a direct method resembles an
@@
 
-\subsubsection{Adjoint Method}
-\label{subsubsec:adjoint_mc}
+\subsection{Adjoint Method}
+\label{subsec:adjoint_mc}
 An alternative formulation for Monte Carlo matrix inversion is the
@@
 
-\subsubsection{Sequential Monte Carlo}
-\label{subsusbsec:sequential_mc}
+\subsection{Sequential Monte Carlo}
+\label{subsec:sequential_mc}
 The direct and adjoint methods described are limited by a convergence
@@
 residual to extract a correction from the search subspace in a
-projection method. By doing this, all of the Monte Carlo error is
-bound in the correction used to update the solution and therefore does
-not explicitly manifest itself in the solution. The downside of such a
+projection method. By doing this, the Monte Carlo error is bound in
+the correction used to update the solution and therefore does not
+explicitly manifest itself in the solution. The downside of such a
 method is that if the solution guess is poor, then many iterations are
@@
 
-\subsubsection{Monte Carlo Synthetic-Acceleration}
-\label{subsubsec:mcsa}
+\subsection{Monte Carlo Synthetic-Acceleration}
+\label{subsec:mcsa}
 Using the ideas of Halton, Evans and Mosher recently developed a Monte
@@
 
-\subsubsection{Preconditioning Stochastic Methods}
+\subsection{Preconditioning Stochastic Methods}
 \label{subsubsec:stochastic_preconditioning}
@@
 
-\section{Parallel Methods for Solving Sparse Linear Systems}
-\label{sec:parallel_krylov_methods}
-Modern parallel implementations of methods on distributed memory
-architectures rely heavily on capabilities provided by general linear
-algebra frameworks. For methods like GMRES, this arises from the fact
-that Krylov methods require only a handful of operation types in their
-implementation that can be efficiently programmed on these
-architectures. Per Saad's text \citep{saad_2003} and as noted in
-Algorithm~\ref{alg:gmres}, these operations are preconditioning,
-matrix-vector multiplications, vector updates, and inner products. For
-the last three items, linear algebra libraries such as PETSc
-\citep{petsc_2012} and Trilinos \citep{trilinos_2005} provide
-efficient parallel implementations for these operations. Depending on
-the type of preconditioning used, efficient parallel implementations
-may also be available for those operations. In all cases, a series of
-scatter/gather operations are required such that global communication
-operations must occur. Although the relative performance of such
-operations is bound to the implementation, asymptotically perfomance
-should be the same across all implementations.
-
-\section{Parallelization of the MCSA Algorithm}
-\label{section:parallel_mcsa}
+\subsection{Parallelization of Stochastic Methods}
+\label{subsec:parallel_stochastic_methods}
 For the prelim, I suspect that I will not have a parallel MCSA
@@
 
-\subsection{Multiple-Set Overlapping-Domain Decomposition}
-\label{subsec:msod}
+\subsubsection{Multiple-Set Overlapping-Domain Decomposition}
+\label{subsubsec:msod}
+
+\subsubsection{Parallel MCSA}
+\label{subsubsec:parallel_mcsa}
+It should be noted here that a lot of the parallel matrix/vector
+operations discussed in the parallel projection methods section also
+apply here. Three of the four MCSA steps are in fact these operations
+with convergence checks also using a norm reduction.

### 2012-09-18 doc/prelim/document/chapters/linear_problem/linear_problem.tex
aworking on parallel matrix multiplication

@@
 operator (e.g. radiation transport sweeps \citep{evans_2010}) that
-normally may not be amenable to its full construction.
+normally may not be amenable to its full construction. In addition,
+the minimization problem is a straight-forward least-squares problem
+where $\ve{H}$ is an upper-Hessenberg matrix.
 
@@
 contains only a subset of the global problem and that subset forms a
-local set of complete equations.
+local set of complete equations. The form of this partitioning is
+problem dependent and often has a geometric or graph-based aspect to
+its construction in order to optimize communication
+patterns. Libraries such as Zoltan \citep{zoltan_2007}, provide an
+implementations of such algorithms.
 
@@
     ||x||_{\infty,l} = \max_n \ve{y}[n],\ \forall n \in [1,N_l]\:\\
-    ||x||_{\infty,g} = \max_p ||x||_{\infty,g}\:.
+    ||x||_{\infty,g} = \max_p ||x||_{\infty,l}\:.
   \end{gather}
@@
 \label{subsubsec:parallel_mat_vec_mutliply}
-We finally consider matrix-vector multiplication operations by
-considering sparse matrices in a compressed storage format. For these
-operations, considerably more communications will be required given
+We finally consider parallel matrix-vector multiplication operations
+using sparse matrices in a compressed storage format. For these
+operations, more complex communication patterns will be required given
 that the entire global vector is required in order to compute a single
-element of the product vector.
+element of the local product vector. Fortunately, the vast majority of
+the global vector components will be multiplied by zero due to the
+sparse matrix and therefore much of the vector can be
+neglected. Instead we only require data from a handful of other
+processes.
+
+\begin{figure}[htpb!]
+  \begin{center}
+    \scalebox{1.5}{
+      \input{chapters/linear_problem/partitioned_matrix.pdftex_t} }
+  \end{center}
+  \caption{Partitioned linear problem.}
+  \label{fig:partitioned_linear_problem}
+\end{figure}
 
@@
 required at several levels of Krylov algorithms stand to reduce
-scalablitity and perfomance.
+scalablitity and perfomance. Furthermore, communication between
+adjacent domains may also cause a bottleneck as the number of domains
+used in simulation grows. The end result is that global data must be
+collected and communicated. For scaling improvement, we seek a
+reduction in these types of operations.
 

### 2012-09-18 doc/prelim/document/chapters/linear_problem/linear_problem.tex
linear problem updates

@@
 neglected. Instead we only require data from a handful of other
-processes.
-
+processes. Consider the sparse matrix-vector multiply in
+Figure~\ref{fig:partitioned_matvec_multiply} that is partitioned on 3
+processors. 
 \begin{figure}[htpb!]
@@
   \end{center}
-  \caption{Partitioned linear problem.}
-  \label{fig:partitioned_linear_problem}
+  \caption{Sparse matrix-vector multiply partitioned on 3 processors.}
+  \label{fig:partitioned_matvec_multiply}
 \end{figure}
+Each process owns a set of equations that correlate to the physical
+domain of which it has ownership. We can break down the equations
+owned by each process in order to devise an efficient scheme for the
+mulitplication. Consider the portion of the matrix-vector multiply
+problem owned by process 1 in
+Figure~\ref{fig:partitioned_mat_vec_multiply}. As shown in
+Figure~\ref{fig:mat_vec_proc_1}, the components of the matrix will be
+multiplied by pieces of the vector that are owned by all
+processors. For those pieces of the matrix that are owned by process
+1, we do these multiplications first as no communication is
+required. Next, process one gathers the components of the global
+vector that it requires to complete its part of the vector
+product. Conversely, processes 2 and 3 must scatter their vector
+components that are required by other processes (such as process 1) in
+order to complete their pieces of the product. This then implies that
+these domain connections for proper gather and scatter combinations
+must be constructed a priori. These data structures are typically
+generated by the partitioning software.
 

### 2012-09-18 doc/prelim/document/chapters/linear_problem/linear_problem.tex
ufinished parallel matvec section

--- 2012-09-18 doc/prelim/document/chapters/linear_problem/linear_problem.tex
@@
   \end{center}
-  \caption{Orthogonality constraint of the new residual with respect
-    to $\mathcal{L}$.}
+  \caption{\textbf{Orthogonality constraint of the new residual with
+      respect to $\mathcal{L}$.} \textit{By projecting $\ve{r}_0$ onto
+      the constraint subspace, we minimize the new residual by
+      removing those components.}}
   \label{fig:linear_projection_constraint}
@@
 the global vector components will be multiplied by zero due to the
-sparse matrix and therefore much of the vector can be
+sparsity of the matrix and therefore much of the vector can be
 neglected. Instead we only require data from a handful of other
-processes.
-
+processes. Consider the sparse matrix-vector multiply in
+Figure~\ref{fig:partitioned_matvec_multiply} that is partitioned on 3
+processors.
 \begin{figure}[htpb!]
@@
   \end{center}
-  \caption{Partitioned linear problem.}
-  \label{fig:partitioned_linear_problem}
+  \caption{\textbf{Sparse matrix-vector multiply operation partitioned
+      on 3 processors.}\textit{Each process owns a set of equations
+      that correlates to its physical domain.}}
+  \label{fig:partitioned_matvec_multiply}
 \end{figure}
-
-\subsubsection{Parallel Perfomance Implications for Projection Methods}
+Each process owns a set of equations that correlate to the physical
+domain of which it has ownership. We can break down the equations
+owned by each process in order to devise an efficient scheme for the
+multiplication. Consider the portion of the matrix-vector multiply
+problem owned by process 1 in
+Figure~\ref{fig:partitioned_matvec_multiply}. As shown in
+Figure~\ref{fig:matvec_proc_1}, the components of the matrix will be
+multiplied by pieces of the vector that are owned by all
+processors. 
+\begin{figure}[htpb!]
+  \begin{center}
+    \scalebox{1.5}{
+      \input{chapters/linear_problem/matvec_proc_1.pdftex_t} }
+  \end{center}
+  \caption{\textbf{Components of sparse matrix-vector multiply
+      operation owned by process 1.} \textit{The numbers above the
+      matrix columns indicate the process that owns the piece of the
+      global vector they are acting on. In order to compute its local
+      components of the matrix-vector product, process 1 needs its
+      matrix elements along with all elements of the global vector
+      owned by processes 2 and 3. The piece of the matrix shown is
+      $\ve{A}_1$ and it is acting locally on $\ve{x}_1$ to compute the
+      local piece of the product, $\ve{y}_1$.}}
+  \label{fig:matvec_proc_1}
+\end{figure}
+For those pieces of the matrix that are owned by process 1 that act on
+the vector owned by process 1, we do these multiplications first as no
+communication is required. Next, process one gathers the components of
+the global vector that it requires to complete its part of the vector
+product. For this example, the components of matrix that will operate
+on the global vector are zero, and therefore no vector elements are
+required to be scattered from process 3 to process 1. Those matrix
+elements that will act on the piece of the vector owned by process 2
+are not all non-zero, and therefore we must gather the entire process
+2 vector components onto process 1 to complete the multiplication.
+Conversely, processes 2 and 3 must scatter its vector components that
+are required by other processes (such as process 1) in order to
+complete their pieces of the product. This then implies that these
+domain connections for proper gather and scatter combinations must be
+constructed a priori. These data structures are typically generated by
+a data partitioning library. Mathematically, if we are performing a
+global matrix-vector multiply of the form $\ve{A}\ve{x} = \ve{y}$,
+then for this example on process 1 we have a sequence of local
+matrix-vector multiplications: $\ve{A}_1\ve{x}_1 + \ve{A}_1\ve{x}_2 =
+\ve{y}_1$. Here, some of the data is intrinisically local, and some
+must be gathered from other processes using the partitioning data
+structures.
+
+\subsubsection{Parallel Perfomance Implications for Krylov Methods}
 \label{subsubsec:projection_method_performance}
-For very large distributed machines, global reduction operations
-required at several levels of Krylov algorithms stand to reduce
-scalablitity and perfomance. Furthermore, communication between
-adjacent domains may also cause a bottleneck as the number of domains
-used in simulation grows. The end result is that global data must be
-collected and communicated. For scaling improvement, we seek a
-reduction in these types of operations.
+Knowing the parallel characteristics of the key operations we must
+perform in order to implement Krylov methods, we can make a few
+statements about parallel perfomance and implications for operation on
+machines of increasing size.  For very large distributed machines,
+global reduction operations required at several levels of Krylov
+algorithms stand to reduce scalablitity and perfomance. Furthermore,
+communication between adjacent domains may also cause a bottleneck as
+the number of domains used in simulation grows. The end result is that
+global data must be collected and communicated. For scaling
+improvement, we seek a reduction in these types of operations.
 

### 2012-09-19 doc/prelim/document/chapters/introduction/introduction.tex
updating references to zotero database

@@
 Monte Carlo methods for the solution of transport problems
-\citep{lewis_1993}. The partial differential equation (PDE) community
-has focused on various deterministic methods for solutions to linear
-problems \citep{saad_2003, kelley_1995}. In between these two areas
-are a not widely known small group of stochastic methods for solving
-sparse linear systems \citep{hammersley_1964, halton_1962,
-  halton_1994}. In recent years, these methods have been further
-developed for transport problems in the form of Monte Carlo
-Synthetic-Acceleration (MCSA) \citep{evans_2003, evans_2009} but have
-yet to be applied to more general sparse linear systems commonly
-generated by the computational physics community. Compared to other
-methods in these regime, MCSA offers three attractive qualities; (1)
-the linear problem operator need not be symmetric or
-positive-definite, thereby reducing preconditioning complexity, (2)
-the stochastic nature of the solution method provides a natural
-solution to the issue of resiliency, and (3) is amenable to
+\citep{lewis_computational_1993}. The partial differential equation
+(PDE) community has focused on various deterministic methods for
+solutions to linear problems \citep{saad_iterative_2003,
+  kelley_iterative_1987}. In between these two areas are a not widely
+known small group of stochastic methods for solving sparse linear
+systems \citep{hammersley_monte_1964, halton_sequential_1962,
+  halton_sequential__1994}. In recent years, these methods have been
+further developed for transport problems in the form of Monte Carlo
+Synthetic-Acceleration (MCSA)
+\citep{t.m._evans__t.j._urbatsch_h._lichtenstein_j.e._morel_residual_2003,
+  evans_monte_2009} but have yet to be applied to more general sparse
+linear systems commonly generated by the computational physics
+community. Compared to other methods in these regime, MCSA offers
+three attractive qualities; (1) the linear problem operator need not
+be symmetric or positive-definite, thereby reducing preconditioning
+complexity, (2) the stochastic nature of the solution method provides
+a natural solution to the issue of resiliency, and (3) is amenable to
 parallelization using modern methods developed by the transport
-community \citep{wagner_2011}. The developement of MCSA as a general
-linear solver and the development of a parallel MCSA method will be
-new and unique features of this work. Resiliency will not be
-addressed.
+community
+\citep{j.c._wagner**_s.w._mosher_t.m._evans_d.e._peplow_and_j.a._turner_hybrid_2010}. The
+developement of MCSA as a general linear solver and the development of
+a parallel MCSA method will be new and unique features of this
+work. Resiliency will not be addressed.
 
@@
 linearizing the problem or building a segregated scheme and using
-traditionaly iterative or direct methods to solve the resulting
-system \citep{tannehill_1997}. In the mathematics community, various
-Newton methods have been popular \citep{kelley_1995}. Recently,
-Jacobian-Free Newton-Krylov (JFNK) schemes \citep{knoll_2004} have
-been utilized in multiple physics architectures and advanced single
-physics codes \citep{gaston_2009}. The benefits of JFNK schemes are
-that the Jacobian is never formed, simplifying the implementation, and
-a Krylov solver is leveraged (typically GMRES or Conjugate Gradient),
-providing excellent convergence properties for well-conditioned and
-well-scaled systems. However, there are two potential drawbacks to
-these methods for high fidelity predictive simulations: (1) the
-Jacobian is approximated by a first-order differencing method on the
-order of machine precision such that this error can grow beyond that
-of those in a fine-grained system and (2) for systems that are not
-symmetric positive-definite (which will be the case for most
-multiphysics systems and certainly for most preconditioned systems)
-the Krylov subspace generated by the GMRES solver may become
-prohibitively large. To address these issues, this thesis proposes a
-new and novel method for nonlinear systems based on the MCSA method.
+traditionaly iterative or direct methods to solve the resulting system
+\citep{pletcher_computational_1997}. In the mathematics community,
+various Newton methods have been popular
+\citep{kelley_iterative_1987}. Recently, Jacobian-Free Newton-Krylov
+(JFNK) schemes \citep{knoll_jacobian-free_2004} have been utilized in
+multiple physics architectures and advanced single physics codes
+\citep{gaston_parallel_2009}. The benefits of JFNK schemes are that the
+Jacobian is never formed, simplifying the implementation, and a Krylov
+solver is leveraged (typically GMRES or Conjugate Gradient), providing
+excellent convergence properties for well-conditioned and well-scaled
+systems. However, there are two potential drawbacks to these methods
+for high fidelity predictive simulations: (1) the Jacobian is
+approximated by a first-order differencing method on the order of
+machine precision such that this error can grow beyond that of those
+in a fine-grained system and (2) for systems that are not symmetric
+positive-definite (which will be the case for most multiphysics
+systems and certainly for most preconditioned systems) the Krylov
+subspace generated by the GMRES solver may become prohibitively
+large. To address these issues, this thesis proposes a new and novel
+method for nonlinear systems based on the MCSA method.
 
@@
 generation using modern Forward Automated Differencing (FAD) methods
-\citep{bartlett_2006}, and MCSA as the inner linear solver. This method
-has several attractive properties. First, the first-order
-approximation to the Jacobian used in JFNK type methods is eliminated
-by generating the Jacobian explicitly with the model equations through
-FAD. Second, the Jacobian need not be explicitly formed by the user
-but is instead automated through FAD; this eleminates the complexity
-of hand-coding derivatives and has also been demonstrated to be more
-efficient computationally than evaluating differenced derivatives
-\citep{bartlett_2006}. Third, unlike GMRES, MCSA does not build a
-subspace during iterations. Although the Jacobian must be explicitly
-formed to use MCSA, for problems that take more than a few GMRES
-iterations to converge the size of the Krylov subspace will grow
-beyond that of the Jacobian. Finally, using MCSA for the linear solve
-provides its benefits for preconditioning, potential resiliency, and
-parallelism.
+\citep{hutchison_automatic_2006}, and MCSA as the inner linear
+solver. This method has several attractive properties. First, the
+first-order approximation to the Jacobian used in JFNK type methods is
+eliminated by generating the Jacobian explicitly with the model
+equations through FAD. Second, the Jacobian need not be explicitly
+formed by the user but is instead automated through FAD; this
+eleminates the complexity of hand-coding derivatives and has also been
+demonstrated to be more efficient computationally than evaluating
+differenced derivatives \citep{hutchison_automatic_2006}. Third,
+unlike GMRES, MCSA does not build a subspace during
+iterations. Although the Jacobian must be explicitly formed to use
+MCSA, for problems that take more than a few GMRES iterations to
+converge the size of the Krylov subspace will grow beyond that of the
+Jacobian. Finally, using MCSA for the linear solve provides its
+benefits for preconditioning, potential resiliency, and parallelism.
 

### 2012-09-19 doc/prelim/document/chapters/linear_problem/linear_problem.tex
updating references to zotero database

@@
 through common methods such as finite differences
-\citep{leveque_2007}, finite volumes \citep{leveque_2002}, and finite
-elements \citep{zienkiewicz_1977} ultimately generates sets of coupled
+\citep{leveque_finite_2007}, finite volumes
+\citep{leveque_finite_2002}, and finite elements
+\citep{zienkiewicz_finite_2005} ultimately generates sets of coupled
 equations in the form of matrix problems.  In addition, these matrices
@@
   methods and matrix analysis, are heavily based on Saad's text
-  \citep{saad_2003}.}
+  \citep{saad_iterative_2003}.}
 
@@
 from the eigenvalues of a matrix $\ve{A}$ is the \textit{spectral
-  radius}, $\rho(\ve{A})$, defined by Saad \citep{saad_2003} as:
+  radius}, $\rho(\ve{A})$, defined by Saad \citep{saad_iterative_2003}
+as:
 \begin{equation}
@@
 for collapsing objects of many elements to a single value. Per
-LeVeque's text \citep{leveque_2007}, the q-norm of a vector is defined
+LeVeque's text \citep{leveque_finite_2007}, the q-norm of a vector is defined
 as:
@@
   non-diagonalizable matrices with the Jordan canonical form of
-  $\ve{H}$.} \citep{saad_2003}, we then have:
+  $\ve{H}$.} \citep{saad_iterative_2003}, we then have:
 \begin{equation}
@@
 basis for the \textit{generalized minimum residual method} (GMRES)
-\citep{saad_1986}.
+\citep{saad_gmres:_1986}.
 
@@
 instances where various processes serve to provide the action of that
-operator (e.g. radiation transport sweeps \citep{evans_2010}) that
-normally may not be amenable to its full construction. In addition,
-the minimization problem is a straight-forward least-squares problem
-where $\ve{H}$ is an upper-Hessenberg matrix.
+operator (e.g. radiation transport sweeps \citep{evans_denovo:_2010})
+that normally may not be amenable to its full construction. In
+addition, the minimization problem is a straight-forward least-squares
+problem where $\ve{H}$ is an upper-Hessenberg matrix.
 
@@
 in their implementation that can be efficiently programmed on these
-architectures. Per Saad's text \citep{saad_2003} and as noted in
-Algorithm~\ref{alg:gmres}, these operations are preconditioning,
-matrix-vector multiplications, vector updates, and inner products. For
-the last three items, linear algebra libraries such as PETSc
-\citep{petsc_2012} and Trilinos \citep{trilinos_2005} provide
-efficient parallel implementations for these operations. Depending on
-the type of preconditioning used, efficient parallel implementations
-may also be available for those operations. In all cases, a series of
-scatter/gather operations are required such that global communication
-operations must occur. Although the relative performance of such
-operations is bound to the implementation, asymptotically perfomance
-should be the same across all implementations.
+architectures. Per Saad's text \citep{saad_iterative_2003} and as
+noted in Algorithm~\ref{alg:gmres}, these operations are
+preconditioning, matrix-vector multiplications, vector updates, and
+inner products. For the last three items, linear algebra libraries
+such as PETSc \citep{gropp_scalable_1993} and Trilinos
+\citep{heroux_overview_2005} provide efficient parallel
+implementations for these operations. Depending on the type of
+preconditioning used, efficient parallel implementations may also be
+available for those operations. Due to their prevalence in modern
+numerical methods, parallel formulations these operations have
+warranted intense study \citep{tuminaro_parallel_1998}. In all cases, a series
+of scatter/gather operations are required such that global
+communication operations must occur. Although the relative performance
+of such operations is bound to the implementation, asymptotically
+perfomance should be the same across all implementations.
 
@@
 algorithms. Furthermore, variants are available that reduce the number
-of global communications required (consider \cite{sosonkina_1998} as
-an example of reducing global operation counts using a different
-orthogonalization procedure than Arnoldi), however, we will only
-consider the basic algorithms here as this handful operations can be
-generalized to fit more complicated algorithms. In all of these cases,
-we assume a general matrix/vector formulation that is distributed in
-parallel such that both local and global knowledge of their
-decomposition is available on request. Furthermore, it is assumed that
-these objects are partitioned in such a way that the parallel
-formulation of the operator and vectors in
-Eq~(\ref{eq:linear_problem}) will be such that each parallel process
-contains only a subset of the global problem and that subset forms a
-local set of complete equations. The form of this partitioning is
-problem dependent and often has a geometric or graph-based aspect to
-its construction in order to optimize communication
-patterns. Libraries such as Zoltan \citep{zoltan_2007}, provide an
-implementations of such algorithms.
+of global communications required (consider
+\cite{sosonkina_scalable_1998} as an example of reducing global
+operation counts using a different orthogonalization procedure than
+Arnoldi), however, we will only consider the basic algorithms here as
+this handful operations can be generalized to fit more complicated
+algorithms. In all of these cases, we assume a general matrix/vector
+formulation that is distributed in parallel such that both local and
+global knowledge of their decomposition is available on
+request. Furthermore, it is assumed that these objects are partitioned
+in such a way that the parallel formulation of the operator and
+vectors in Eq~(\ref{eq:linear_problem}) will be such that each
+parallel process contains only a subset of the global problem and that
+subset forms a local set of complete equations. The form of this
+partitioning is problem dependent and often has a geometric or
+graph-based aspect to its construction in order to optimize
+communication patterns. Libraries such as Zoltan
+\citep{devine_zoltan_2002}, provide an implementations of such
+algorithms.
 
@@
 neglected. Instead we only require data from a handful of other
-processes. Consider the sparse matrix-vector multiply in
+processes that can be acquired through asynchronous/synchronous
+communications. Consider the sparse matrix-vector multiply in
 Figure~\ref{fig:partitioned_matvec_multiply} that is partitioned on 3
@@
   \end{center}
-  \caption{\textbf{Sparse matrix-vector multiply operation partitioned
-      on 3 processors.}\textit{Each process owns a set of equations
-      that correlates to its physical domain.}}
+  \caption{\textbf{Sparse matrix-vector multiply $\ve{A}\ve{x}=\ve{y}$
+      operation partitioned on 3 processors.} \textit{Each process
+      owns a set of equations that correlates to its physical
+      domain.}}
   \label{fig:partitioned_matvec_multiply}
@@
 an enjoyable manuscript published in 1950 by Forsythe and Leibler
-\citep{forsythe_1950}. In their outline, Forsythe and Liebler in fact
-credit the creation of this technique to J. Von Neumann and S.M. Ulam
-some years earlier than its publication. Hammersley and Handscomb's
-1964 monograph \citep{hammersley_1964} presents additional detail on
-this topic using a collection of references from the 1950's and early
-1960's.
+\citep{forsythe_matrix_1950}. In their outline, Forsythe and Liebler
+in fact credit the creation of this technique to J. Von Neumann and
+S.M. Ulam some years earlier than its publication. Hammersley and
+Handscomb's 1964 monograph \citep{hammersley_monte_1964} presents
+additional detail on this topic using a collection of references from
+the 1950's and early 1960's.
 
@@
 residual Monte Carlo method that moves towards exponential convergence
-rates \citep{halton_1962} and further refined his work some years
-later \citep{halton_1994}. Applications of his work by the transport
-community have confirmed convergence rates on the order of of
-$exp(-N)$ \citep{evans_2003}. In much the same way as projection
-methods, Halton's method, sequential Monte Carlo, utilizes the adjoint
-Monte Carlo solver as a means of driving down the residual vector. He
-proposed the following iterative scheme as a solution to
-Eq~(\ref{eq:linear_problem})\:
+rates \citep{halton_sequential_1962} and further refined his work some
+years later \citep{halton_sequential_1994}. Applications of his work
+by the transport community have confirmed convergence rates on the
+order of of $exp(-N)$
+\citep{t.m._evans__t.j._urbatsch_h._lichtenstein_j.e._morel_residual_2003}. In
+much the same way as projection methods, Halton's method, sequential
+Monte Carlo, utilizes the adjoint Monte Carlo solver as a means of
+driving down the residual vector. He proposed the following iterative
+scheme as a solution to Eq~(\ref{eq:linear_problem})\:
 \begin{subequations}
@@
 Carlo solution that was not prohibited severely by the quality of the
-initial guess for the system \citep{evans_2009}. Their approach was
-instead to use residual Monte Carlo as a synthetic acceleration for a
-stationary method. To derive this method, we begin by splitting the
-operator in Eq~(\ref{eq:linear_problem}) 
+initial guess for the system \citep{evans_monte_2009}. Their approach
+was instead to use residual Monte Carlo as a synthetic acceleration
+for a stationary method. To derive this method, we begin by splitting
+the operator in Eq~(\ref{eq:linear_problem})
 \begin{equation}

### 2012-09-19 doc/prelim/document/chapters/introduction/introduction.tex
including more references

@@
 systems \citep{hammersley_monte_1964, halton_sequential_1962,
-  halton_sequential__1994}. In recent years, these methods have been
+  halton_sequential_1994}. In recent years, these methods have been
 further developed for transport problems in the form of Monte Carlo
-Synthetic-Acceleration (MCSA)
-\citep{t.m._evans__t.j._urbatsch_h._lichtenstein_j.e._morel_residual_2003,
+Synthetic-Acceleration (MCSA) \citep{evans_residual_2003,
   evans_monte_2009} but have yet to be applied to more general sparse
@@
 parallelization using modern methods developed by the transport
-community
-\citep{j.c._wagner**_s.w._mosher_t.m._evans_d.e._peplow_and_j.a._turner_hybrid_2010}. The
-developement of MCSA as a general linear solver and the development of
-a parallel MCSA method will be new and unique features of this
-work. Resiliency will not be addressed.
+community \citep{wagner_hybrid_2010}. The developement of MCSA as a
+general linear solver and the development of a parallel MCSA method
+will be new and unique features of this work. Resiliency will not be
+addressed.
 

### 2012-09-19 doc/prelim/document/chapters/linear_problem/linear_problem.tex
including more references

@@
 We finally consider parallel matrix-vector multiplication operations
-using sparse matrices in a compressed storage format. For these
-operations, more complex communication patterns will be required given
-that the entire global vector is required in order to compute a single
-element of the local product vector. Fortunately, the vast majority of
-the global vector components will be multiplied by zero due to the
-sparsity of the matrix and therefore much of the vector can be
-neglected. Instead we only require data from a handful of other
-processes that can be acquired through asynchronous/synchronous
-communications. Consider the sparse matrix-vector multiply in
+using sparse matrices in a compressed storage format by considering
+Saad's outline as well as the more formal work of Tuminaro
+\citep{tuminaro_parallel_1998}. For these operations, more complex
+communication patterns will be required given that the entire global
+vector is required in order to compute a single element of the local
+product vector. Fortunately, the vast majority of the global vector
+components will be multiplied by zero due to the sparsity of the
+matrix and therefore much of the vector can be neglected. Instead we
+only require data from a handful of other processes that can be
+acquired through asynchronous/synchronous communications. Consider the
+sparse matrix-vector multiply in
 Figure~\ref{fig:partitioned_matvec_multiply} that is partitioned on 3
@@
 algorithms stand to reduce scalablitity and perfomance. Furthermore,
-communication between adjacent domains may also cause a bottleneck as
-the number of domains used in simulation grows. The end result is that
-global data must be collected and communicated. For scaling
-improvement, we seek a reduction in these types of operations.
+communication between adjacent domains in matrix-vector multiply
+operations may also cause a bottleneck as the number of domains used
+in simulation grows. The end result is that global data must be
+collected and communicated. For scaling improvement, we seek a
+reduction in these types of operations. In addition, these issues
+become more prominent as the Krylov iterations progress, causing the
+Krylov subspace to grow and the total number of operations need to
+orthogonalize that subspace to increase.
 
@@
 in fact credit the creation of this technique to J. Von Neumann and
-S.M. Ulam some years earlier than its publication. Hammersley and
-Handscomb's 1964 monograph \citep{hammersley_monte_1964} presents
-additional detail on this topic using a collection of references from
-the 1950's and early 1960's.
+S.M. Ulam some years earlier than its publication. In 1952 Wasow
+provided a more formal explanation of Von Neumann and Ulam's method
+\citep{wasow_note_1952} and Hammersley and Handscomb's 1964 monograph
+\citep{hammersley_monte_1964} presents additional detail on this topic
+using a collection of references from the 1950's and early 1960's.
 
@@
 \end{equation}
-in $\ve{H}$ where $\nu$ is a particular sequence permutation. We can
-generate these sequences of transitions through Monte Carlo random
-walks by assigning them both a probability and weight. As a
-reinterpretation of the iteration matrix, we then form the
-\textit{Neumann-Ulam decomposition} of \ve{H}:
+in $\ve{H}$ where $\nu$ is a particular random walk sequence
+permutation. We can generate these sequences of transitions through
+Monte Carlo random walks by assigning them both a probability and
+weight. As a reinterpretation of the iteration matrix, we then form
+the \textit{Neumann-Ulam decomposition} of \ve{H}:
 \begin{equation}
@@
 \end{equation}
-where $\circ$ denotes the Hadamard product operation\footnote{The Hadamard
-  product $\ve{A} = \ve{B} \circ \ve{C}$ is defined element-wise as
-  $a_{ij} = b_{ij} c_{ij}$.}, $\ve{P}$ denotes the transition
-probability matrix, and $\ve{W}$ denotes the transition weight
-matrix. The formulation of $\ve{P}$ and $\ve{W}$ will be dependent on
-whether we choose a direct or adjoint Monte Carlo sequence to estimate
-the state transitions in Eq~(\ref{eq:expanded_neumann_solution}).
+where $\circ$ denotes the Hadamard product operation\footnote{The
+  Hadamard product $\ve{A} = \ve{B} \circ \ve{C}$ is defined
+  element-wise as $a_{ij} = b_{ij} c_{ij}$.}, $\ve{P}$ denotes the
+transition probability matrix, and $\ve{W}$ denotes the transition
+weight matrix. This decomposition, a generalization of Dimov's work
+\citep{dimov_new_1998}, is an extension of the original Neumann-Ulam
+scheme in that now a weight cutoff can be used to terminate a random
+walk sequence. The formulation of $\ve{P}$ and $\ve{W}$ will be
+dependent on whether we choose a direct or adjoint Monte Carlo
+sequence to estimate the state transitions in
+Eq~(\ref{eq:expanded_neumann_solution}).
 
@@
 are assembled. To achieve this, we build the direct method
-Neumann-Ulam decomposition by first choosing a probability matrix that
-is a row scaling of $\ve{H}$ such that its components are:
+Neumann-Ulam decomposition per Dimov's approach by first choosing a
+probability matrix that is a row scaling of $\ve{H}$ such that its
+components are:
 \begin{equation}
@@
 order of of $exp(-N)$
-\citep{t.m._evans__t.j._urbatsch_h._lichtenstein_j.e._morel_residual_2003}. In
+\citep{evans_residual_2003}. In
 much the same way as projection methods, Halton's method, sequential

### 2012-09-20 doc/prelim/document/chapters/linear_problem/linear_problem.tex
updates to linear problem

@@
 become more prominent as the Krylov iterations progress, causing the
-Krylov subspace to grow and the total number of operations need to
+Krylov subspace to grow and the total number of operations needed to
 orthogonalize that subspace to increase.
@@
 seeking a solution to Eq~(\ref{eq:linear_problem}). For a given linear
-operator $\ve{A}$, we can split it in the same manner as
-Eq~(\ref{eq:linear_split_equation2}) to define the following operator:
+operator $\ve{A}$, we can using diagonal splitting in the same manner
+as the stationary method in Eq~(\ref{eq:linear_split_equation2}) to
+define the following operator\footnote{It should be noted that
+  non-diagonal splittings have been recently explored in
+  \citep{srinivasan_monte_2010} and have the potential to improve
+  efficiency.}:
 \begin{equation}
@@
 
+\subsubsection{Estimator Variance}
+\label{subsubsec:estimator_variance}
 We can compute the variance of the estimator through traditional
@@
 
+\subsubsection{Reusing Random Walks}
+\label{subsubsec:reusing_random_walks}
+Ji recently proposed a variation of the direct method in which the
+random walk permutations in Eq~(\ref{eq:mc_walk_permutation}) can be
+used to contribute to the solution tally in all visited states instead
+of the only starting state \citep{ji_reusing_2012}.
+
 \subsection{Adjoint Method}
@@
 decomposition. Using the transpose form, probabilities should instead
-by column-scaled:
+be column-scaled:
 \begin{equation}
@@
 where the adjoint Monte Carlo method is used to generate the solution
-correction from the residual. In addition to the Monte Carlo solver
-parameters dictating the number of histories and weight cutoff, the
-outer MCSA iterations also have the following stopping criteria:
+correction from the residual. Using Monte Carlo in this way achieves
+the same effect as Halton's method, decoupling its convergence rate
+from the overall convergence rate of the method. Here, the approximate
+Monte Carlo solution is not driven to a particular convergence as it
+merely supplies a correction for the initial guess generated by
+Richardson's iteration. Rather, only a set number of histories are
+required using the adjoint method to generate the correction.
+
+In addition to the Monte Carlo solver parameters dictating the number
+of histories and weight cutoff, the outer MCSA iterations also have
+the following stopping criteria:
 \begin{equation}
@@
 \end{equation}
-where $\epsilon$ is a user-defined parameter.
+where $\epsilon$ is a user-defined parameter. We therefore have 3
+parameters to tune in an MCSA implementation: the number of Monte
+Carlo histories computed in the adjoint solve during each MCSA
+iteration, the weight cutoff for those histories, and the total MCSA
+convergence tolerance as specified by $\epsilon$.
 
@@
 \label{subsec:parallel_stochastic_methods}
-For the prelim, I suspect that I will not have a parallel MCSA
-algorithm up and running as I will be focusing on generating the code
-base needed to build the physics operators and tools for linear and
-nonlinear systems analysis. Therefore, here we should focus on
-motivating a parallel scheme for this work. We should start by first
-looking at current parallel methods for solving linear systems using
-the conventional techniques from the previous chapter. It must be
-clearly pointed out that for the new hardware coming online, these
-methods (or at least their current implementations) will be lacking in
-scalability and the potential for resiliency, and therefore a parallel
-MCSA is attractive, even if it ultimately does not show performance
-gains from a wall time perspective.
-
-Once we have done this, MSOD will need to be defined. The literature
-is actually quite lacking here even for reactor physics problems as
-this is really bleeding edge. Therefore, I should strive here to more
-rigorously define this methodology in terms of solving general linear
-systems. Furthermore, we should discuss the aspects of being able to
-vary the overlap such that we can generate a spectrum between complete
-domain decomposition and complete domain replication. Finally, the
-beginnings of the algorithm should be sketched out.
+For Monte Carlo methods, and in particular MCSA, to be viable at the
+production scale, scalable parallel implementations are required. In a
+general linear algebra context, this has not yet been
+achieved. Therefore, we will develop and implement parallel algorithms
+for these methods levaraging both the knowledge gained from the
+general parallel implementations of Krylov methods and modern parallel
+strategies for Monte Carlo as developed by the reactor physics
+community. In order to formulate a parallel MCSA algorithm, we
+recognize that the algorithm occurs in two stages, an outer iteration
+performing Richardson's iteration and applying the correction, and an
+inner Monte Carlo solver that is generating the correction via the
+adjoint method. The parallel aspects of both these components must be
+considered.
 
@@
 \label{subsubsec:msod}
+As observed in the discussion on parallel Krylov methods, large-scale
+problems will surely have their data partitioned such that each
+parllel process owns a subset of the equations in the linear
+system. Given this convention, the adjoint Monte Carlo algorithm must
+perform random walks over a domain that is decomposed and must remain
+decomposed due to memory limitations. This naturally leads us to seek
+parallel algorithms that handle domain decomposition. In 2010, Wagner
+and colleagues developed the \textit{multiple-set overlapping-domain}
+(MSOD) decomposition for parallel Monte Carlo applications for
+full-core reactor analysis \citep{wagner_hybrid_2010}.
+
+\subsubsection{Parallel Adjoint Method}
+\label{subsubsec:parallel_adjoint}
 

### 2012-09-20 doc/prelim/document/chapters/linear_problem/linear_problem.tex
Working on monte carlo domain decomposition section

@@
 
-\subsubsection{Multiple-Set Overlapping-Domain Decomposition}
+\subsubsection{Domain Decomposition for Monte Carlo}
 \label{subsubsec:msod}
@@
 decomposed due to memory limitations. This naturally leads us to seek
-parallel algorithms that handle domain decomposition. In 2010, Wagner
-and colleagues developed the \textit{multiple-set overlapping-domain}
-(MSOD) decomposition for parallel Monte Carlo applications for
-full-core reactor analysis \citep{wagner_hybrid_2010}.
+parallel algorithms that handle domain decomposition. In the context
+of radiation transport, Brunner and colleagues provided one of the
+first surveys of algorithms for achieving this
+\citep{brunner_comparison_2006}. In their work they identity two data
+sets that are required to be communicated: the sharing of particles
+that are transported from one domain to another and therefore from one
+processor to another and a global communication that signals if
+particle transport has been completed on all processors. The
+algorithms presented include a fully-locking synchronous scheme, a
+fully-asynchronous pattern, a traditional master/slave scheme, and a
+modified master/slave scheme the implements a binary tree pattern for
+the global reduction type operations needed to communicate between the
+master and slave processes.
+
+In 2010, Wagner and colleagues developed the \textit{multiple-set
+  overlapping-domain} (MSOD) decomposition for parallel Monte Carlo
+applications for full-core reactor analysis
+\citep{wagner_hybrid_2010}.
 

### 2012-09-20 doc/prelim/document/chapters/linear_problem/linear_problem.tex
adding example from gentile paper

@@
 parallel algorithms that handle domain decomposition. In the context
-of radiation transport, Brunner and colleagues provided one of the
-first surveys of algorithms for achieving this
-\citep{brunner_comparison_2006}. In their work they identity two data
-sets that are required to be communicated: the sharing of particles
-that are transported from one domain to another and therefore from one
-processor to another and a global communication that signals if
-particle transport has been completed on all processors. The
-algorithms presented include a fully-locking synchronous scheme, a
-fully-asynchronous pattern, a traditional master/slave scheme, and a
-modified master/slave scheme the implements a binary tree pattern for
-the global reduction type operations needed to communicate between the
-master and slave processes.
+of radiation transport, Brunner and colleagues provided a survey of
+algorithms for achieving this as implemented in production implicit
+Monte Carlo codes \citep{brunner_comparison_2006}. In their work they
+identity two data sets that are required to be communicated: the
+sharing of particles that are transported from one domain to another
+and therefore from one processor to another and a global communication
+that signals if particle transport has been completed on all
+processors. The algorithms presented are a fully-locking synchronous
+scheme, a asynchronous-send/synchronous-receive pattern, a traditional
+master/slave scheme, and a modified master/slave scheme the implements
+a binary tree pattern for the global reduction type operations needed
+to communicate between the master and slave processes. They observed
+that the modified master/slave scheme performed best in that global
+communications were implemented more efficiently than those required
+by the asynchronous scheme. Furthermore, none of these schemes handled
+load-imbalanced cases efficiently. Such cases will be common if the
+source sampled in the Monte Carlo random walk is not isotropic and
+evenly distributed throughout the global domain. It was noted that
+efficiencies were improved by increasing the frequency by which
+particle data was communicated between domain-adjacent processors.
+However, this ultimately increases communication costs.
+
+The 2006 work of Brunner is notable in that the Monte Carlo codes used
+to implement and test the algorithms adhered to a strict policy of
+generated identical results, indepdent of domain decomposition or
+domain replication as derived from the work of Gentile and colleagues
+\citep{gentile_obtaining_2005}. In their work, a procedure is given for
+obtaining results reproducable to machine precision for an arbitrary
+number of processors and domains. Differences can arise from using a 
+different random number sequence in each domain and performing
+a sequence of floating point operations on identical data in a
+different, leading to variations in round-off error and ultimately a
+non-identical answer. They use a simple example, recreated below in
+Figure~\ref{fig:gentile_example}, that illustrates these issues.
+\begin{figure}[htpb!]
+  \begin{center}
+    \scalebox{1.5}{
+      \input{chapters/linear_problem/gentile_example.pdftex_t} }
+  \end{center}
+  \caption{\textbf{Gentile's example illustrating how domain
+      decomposition can create reproducability issues in Monte Carlo.}
+    \textit{Both particles A and B start in zone 1 on processor
+      1. Particle A moves to zone 2 on processor 2 and scatters back
+      to zone 1 while B scatters in zone 1 and remains there.}}
+  \label{fig:gentile_example}
+\end{figure}
 
@@
 applications for full-core reactor analysis
-\citep{wagner_hybrid_2010}.
+\citep{wagner_hybrid_2010}. In their work, a certain amount of overlap
+between domains was used to decrease the number of particles leaving
+the domain.
 

### 2012-09-20 doc/prelim/document/chapters/linear_problem/linear_problem.tex
adding reproducability section

@@
 decomposed due to memory limitations. This naturally leads us to seek
-parallel algorithms that handle domain decomposition. In the context
-of radiation transport, Brunner and colleagues provided a survey of
-algorithms for achieving this as implemented in production implicit
-Monte Carlo codes \citep{brunner_comparison_2006}. In their work they
-identity two data sets that are required to be communicated: the
-sharing of particles that are transported from one domain to another
-and therefore from one processor to another and a global communication
-that signals if particle transport has been completed on all
-processors. The algorithms presented are a fully-locking synchronous
-scheme, a asynchronous-send/synchronous-receive pattern, a traditional
-master/slave scheme, and a modified master/slave scheme the implements
-a binary tree pattern for the global reduction type operations needed
-to communicate between the master and slave processes. They observed
-that the modified master/slave scheme performed best in that global
-communications were implemented more efficiently than those required
-by the asynchronous scheme. Furthermore, none of these schemes handled
-load-imbalanced cases efficiently. Such cases will be common if the
-source sampled in the Monte Carlo random walk is not isotropic and
-evenly distributed throughout the global domain. It was noted that
-efficiencies were improved by increasing the frequency by which
-particle data was communicated between domain-adjacent processors.
-However, this ultimately increases communication costs.
-
+parallel algorithms that handle domain decomposition.
+
+In the context of radiation transport, Brunner and colleagues provided
+a survey of algorithms for achieving this as implemented in production
+implicit Monte Carlo codes \citep{brunner_comparison_2006}. In their
+work they identity two data sets that are required to be communicated:
+the sharing of particles that are transported from one domain to
+another and therefore from one processor to another and a global
+communication that signals if particle transport has been completed on
+all processors. The algorithms presented are a fully-locking
+synchronous scheme, a asynchronous-send/synchronous-receive pattern, a
+traditional master/slave scheme, and a modified master/slave scheme
+the implements a binary tree pattern for the global reduction type
+operations needed to communicate between the master and slave
+processes. They observed that the modified master/slave scheme
+performed best in that global communications were implemented more
+efficiently than those required by the asynchronous
+scheme. Furthermore, none of these schemes handled load-imbalanced
+cases efficiently. Such cases will be common if the source sampled in
+the Monte Carlo random walk is not isotropic and evenly distributed
+throughout the global domain. It was noted that efficiencies were
+improved by increasing the frequency by which particle data was
+communicated between domain-adjacent processors.  However, this
+ultimately increases communication costs.
+
+In 2010, Wagner and colleagues developed the \textit{multiple-set
+  overlapping-domain} (MSOD) decomposition for parallel Monte Carlo
+applications for full-core reactor analysis
+\citep{wagner_hybrid_2010}. In their work, a certain amount of overlap
+between domains was used to decrease the number of particles leaving
+the domain.
+
+\subsubsection{Reproducable Domain Decomposed Results}
 The 2006 work of Brunner is notable in that the Monte Carlo codes used
 to implement and test the algorithms adhered to a strict policy of
-generated identical results, indepdent of domain decomposition or
+generating identical results indepdent of domain decomposition or
 domain replication as derived from the work of Gentile and colleagues
@@
       1. Particle A moves to zone 2 on processor 2 and scatters back
-      to zone 1 while B scatters in zone 1 and remains there.}}
+      to zone 1 while B scatters in zone 1 and remains there. A1 and
+      A2 denote the track of particle A that is in zone 1 while B1 and
+      B2 denote the track of particle B that is in zone 1.}}
   \label{fig:gentile_example}
 \end{figure}
-
-In 2010, Wagner and colleagues developed the \textit{multiple-set
-  overlapping-domain} (MSOD) decomposition for parallel Monte Carlo
-applications for full-core reactor analysis
-\citep{wagner_hybrid_2010}. In their work, a certain amount of overlap
-between domains was used to decrease the number of particles leaving
-the domain.
+In this example, the domain is decomposed on two processors with each
+processor owning one of the two zones. Starting with particle A, it is
+born in zone 1 and is transported to zone 2 where a scattering event
+occurs. Concerning the first reproducability issue, if the same
+sequence of random numbers is not used to compute the trajectory from
+the new scattering event, we cannot expect to achieve the same result
+if the domain were not decomposed. If the numbers are different, the
+scattering event in zone 2 may keep the particle there or even eject
+it from the domain if the sequence were different. The second issue is
+demonstrated by adding another particle B that remains in the
+domain. In this case, an efficient algorithm will transport particle A
+on processor 1 until it leaves zone 1 and then transport particle
+B. Particle A will not renter the domain until it has been
+communicated to processor 2, processor 2 performs the transport, and
+it is communicated back to processor 1. If we are doing a track-length
+tally in zone 1, then we sum the tracks lengths observed in that
+zone. In the single processor, single zone case particle A would be
+transported in its entirety and then particle B transported. This
+would result in a tally sum with the following order of operations:
+$(((A1+A2)+B1)+B2)$. If we were instead to use 2 processors, we would
+instead have the following order: $(((A1+B1)+B2)+A2)$. In the context
+of floating point operations, we cannot expect these to have an
+identical result to machine precesion as round-off approximations will
+differ.
 

### 2012-09-20 doc/prelim/document/chapters/linear_problem/linear_problem.tex
linear problem updates

@@
 all processors. The algorithms presented are a fully-locking
-synchronous scheme, a asynchronous-send/synchronous-receive pattern, a
-traditional master/slave scheme, and a modified master/slave scheme
+synchronous scheme, an asynchronous-send/synchronous-receive pattern,
+a traditional master/slave scheme, and a modified master/slave scheme
 the implements a binary tree pattern for the global reduction type
@@
 communicated between domain-adjacent processors. However, this
-ultimately increases communication costs.
+ultimately increases communication costs. In 2009, Brunner extended
+his work by using a more load-balanced approach with a fully
+asynchronous communication pattern
+\citep{brunner_efficient_2009}. Although the extended implementation
+allowed was more robust and allowed for scaling to larger numbers of
+processors, performance issues were still noted, with efficiencies
+improvements needed in both the weak and strong scaling cases for
+unbalanced problems. These results led Brunner to conclude that a
+combination of domain decomposition and domain replication could be
+used to solve some of these issues.
 
@@
 to implement and test the algorithms adhered to a strict policy of
-generating identical results indepdent of domain decomposition or
+generating identical results independent of domain decomposition or
 domain replication as derived from the work of Gentile and colleagues
@@
 identical result to machine precesion as round-off approximations will
-differ.
+differ resulting in non-commutative addition.
+
+Their solutions to these problems are elegant in that they require a
+minimal amount of modification to be applied to the Monte Carlo
+algorithm. To solve the first issue, in order to ensure each particle
+maintains an invariant random number sequence that determines its
+behavior regardless of domain decomposition, each particle carries
+with it a random number seed that describes its current state. These
+seeds are derived from the actual geometric location of the particle
+such that it is decomposition invariant. Non-commutative floating
+point operations are solved by instead mapping floating point values
+to 64-bit integer values for which additions will always be
+commutative. Once the operations are complete, these integers are
+mapped back to floating point values.
 

### 2012-09-21 doc/prelim/document/chapters/linear_problem/linear_problem.tex
adding msod figure

--- 2012-09-20 doc/prelim/document/chapters/linear_problem/linear_problem.tex
@@
-\chapter{The Linear Problem}
+\chapter{Conventional Solution Methods for Linear Problems}
 \label{ch:linear_problem}
@@
 discuss here conventional stationary and projection methods for
-solving sparse systems to provide background and then give detail on a
-reemerging stochastic class of methods. Details on the parallelization
-of conventional methods are discussed and a novel proposal for the
-parallelization of stochastic methods presented.\footnote{The contents
-  of this chapter, particularly those sections relating to projection
-  methods and matrix analysis, are heavily based on Saad's text
+solving sparse systems to provide the necessary background for the
+remainder of this work. Details on the parallelization of conventional
+methods are discussed.\footnote{The contents of this chapter,
+  particularly those sections relating to projection methods and
+  matrix analysis, are heavily based on Saad's text
   \citep{saad_iterative_2003}.}
@@
 
-\subsection{Parallel Projection Methods}
-\label{subsec:parallel_krylov_methods}
+\section{Parallel Projection Methods}
+\label{sec:parallel_krylov_methods}
 Modern parallel implementations of projection methods on distributed
@@
 
-\subsubsection{Parallel Vector Update}
-\label{subsubsec:parallel_vec_update}
+\subsection{Parallel Vector Update}
+\label{subsec:parallel_vec_update}
 Parallel vector update operations arise from the construction of the
@@
 
-\subsubsection{Parallel Vector Product}
-\label{subsubsec:parallel_vector_product}
+\subsection{Parallel Vector Product}
+\label{subsec:parallel_vector_product}
 Vector product operations are used in several instances during a
@@
 
-\subsubsection{Parallel Matrix-Vector Multiplications}
-\label{subsubsec:parallel_mat_vec_mutliply}
+\subsection{Parallel Matrix-Vector Multiplications}
+\label{subsec:parallel_mat_vec_mutliply}
 We finally consider parallel matrix-vector multiplication operations
@@
 
-\subsubsection{Parallel Perfomance Implications for Krylov Methods}
-\label{subsubsec:projection_method_performance}
+\subsection{Parallel Perfomance Implications for Krylov Methods}
+\label{subsec:projection_method_performance}
 Knowing the parallel characteristics of the key operations we must
@@
 orthogonalize that subspace to increase.
-
-\section{Stochastic Methods}
-\label{sec:stochastic_methods}
-An alternative approach to approximate matrix inversion is to employ
-Monte Carlo methods that sample a distribution with an expectation
-value equivalent to that of the inverted operator. Such methods have
-been in existence for decades with the earliest reference noted here
-an enjoyable manuscript published in 1950 by Forsythe and Leibler
-\citep{forsythe_matrix_1950}. In their outline, Forsythe and Liebler
-in fact credit the creation of this technique to J. Von Neumann and
-S.M. Ulam some years earlier than its publication. In 1952 Wasow
-provided a more formal explanation of Von Neumann and Ulam's method
-\citep{wasow_note_1952} and Hammersley and Handscomb's 1964 monograph
-\citep{hammersley_monte_1964} presents additional detail on this topic
-using a collection of references from the 1950's and early 1960's.
-
-We begin our discussion of stochastic methods using these texts by
-seeking a solution to Eq~(\ref{eq:linear_problem}). For a given linear
-operator $\ve{A}$, we can using diagonal splitting in the same manner
-as the stationary method in Eq~(\ref{eq:linear_split_equation2}) to
-define the following operator\footnote{It should be noted that
-  non-diagonal splittings have been recently explored in
-  \citep{srinivasan_monte_2010} and have the potential to improve
-  efficiency.}:
-\begin{equation}
-  \ve{H} = \ve{I} - \ve{A}\:.
-  \label{eq:linear_mc_iteration_matrix}
-\end{equation}
-We can then form an alternative representation for $\ve{A}^{-1}$ by
-generating the \textit{Neumann series}:
-\begin{equation}
-  \ve{A}^{-1} = (\ve{I}-\ve{H})^{-1} = \sum_{k=0}^{\infty} \ve{H}^k\:,
-  \label{eq:neumann_series}
-\end{equation}
-which will converge if the spectral radius of $\ve{H}$ is less than
-1. If we then apply this inverse sum to the right hand side of
-Eq~(\ref{eq:linear_problem}) we acquire the solution to the linear
-problem:
-\begin{equation}
-  \ve{A}^{-1}\ve{b} = \sum_{k=0}^{\infty} \ve{H}^k\ve{b} = \ve{x}\:.
-  \label{eq:neumann_solution}
-\end{equation}
-An approximation of this summation will therefore lead to an
-approximation of the solution. If we expand the summation with a
-succession of matrix-vector multiply operations, we arrive at an
-alternative perspective of this summation by considering its $i^{th}$
-component:
-\begin{equation}
-  x_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-  \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}\:,
-  \label{eq:expanded_neumann_solution}
-\end{equation}
-which can interpreted as a series of transitions between states,
-\begin{equation}
- \nu = i \rightarrow i_1 \rightarrow \cdots \rightarrow i_{k-1}
- \rightarrow i_{k}\:,
-  \label{eq:mc_walk_permutation}
-\end{equation}
-in $\ve{H}$ where $\nu$ is a particular random walk sequence
-permutation. We can generate these sequences of transitions through
-Monte Carlo random walks by assigning them both a probability and
-weight. As a reinterpretation of the iteration matrix, we then form
-the \textit{Neumann-Ulam decomposition} of \ve{H}:
-\begin{equation}
-  \ve{H} = \ve{P} \circ \ve{W}\:,
-  \label{eq:neumann_ulam_decomposition}
-\end{equation}
-where $\circ$ denotes the Hadamard product operation\footnote{The
-  Hadamard product $\ve{A} = \ve{B} \circ \ve{C}$ is defined
-  element-wise as $a_{ij} = b_{ij} c_{ij}$.}, $\ve{P}$ denotes the
-transition probability matrix, and $\ve{W}$ denotes the transition
-weight matrix. This decomposition, a generalization of Dimov's work
-\citep{dimov_new_1998}, is an extension of the original Neumann-Ulam
-scheme in that now a weight cutoff can be used to terminate a random
-walk sequence. The formulation of $\ve{P}$ and $\ve{W}$ will be
-dependent on whether we choose a direct or adjoint Monte Carlo
-sequence to estimate the state transitions in
-Eq~(\ref{eq:expanded_neumann_solution}).
-
-\subsection{Direct Method}
-\label{subsec:direct_mc}
-In the context of matrix inversion, a direct method resembles an
-adjoint Monte Carlo method in the reactor physics community where the
-solution state is sampled and the source terms that contribute to it
-are assembled. To achieve this, we build the direct method
-Neumann-Ulam decomposition per Dimov's approach by first choosing a
-probability matrix that is a row scaling of $\ve{H}$ such that its
-components are:
-\begin{equation}
-  p_{ij} = \frac{|h_{ij}|}{\sum_j |h_{ij}|}\:.
-  \label{eq:direct_probability}
-\end{equation}
-From this, we then see that the probability of transitioning from a
-state $i$ to a state $j$ is implicitly linked to the original operator
-$\ve{A}$ in that those terms with large values, and therefore those
-that make the greatest contribution to the numerical solution, will be
-sampled with a higher probability than smaller terms. In addition, the
-row scaling provides a normalization over the state to which we are
-transitioning such that $\sum_j p_{ij} = 1$, meaning that we sample
-the probabilities over the rows of the matrix. The components of
-the weight matrix are then defined by
-Eq~(\ref{eq:neumann_ulam_decomposition}) as:
-\begin{equation}
-  w_{ij} = \frac{h_{ij}}{p_{ij}}\:.
-  \label{eq:direct_weight}
-\end{equation}
-It should be noted here that if $\ve{A}$ is sparse, then $\ve{H}$,
-$\ve{P}$, and $\ve{W}$ must be sparse as well by
-definition. Additionally, we only compute $\ve{P}$ and $\ve{W}$ from
-the non-zero elements of $\ve{H}$ as those components that are zero
-will not participate in the random walk. Furthermore, it prevents an
-infinite weight from being generated in Eq~(\ref{eq:direct_weight}).
-
-Using these matrices, we can then form the expectation value of the
-direct solution. For a given random walk permutation $\nu$ with $k$
-events, we define the weight of that permutation to be:
-\begin{equation}
-  W_{m} = \sum_{m=0}^k w_{i,i_1} w_{i_1,i_2} \cdots w_{i_{m-1},i_m}\:,
-  \label{eq:direct_permutation_weight}
-\end{equation}
-such that the weight of each transition event contributes to the
-total. The contribution to the solution from a particular random walk
-permutation is then:
-\begin{equation}
-  X_{\nu}(i_0 = i) = \sum_{m=0}^k W_{m} b_{i_m}\:,
-  \label{eq:direct_permutation_contribution}
-\end{equation}
-where $X_{\nu}(i_0 = i)$ signifies that the solution state in which we
-are tallying defines state $i$.  We can interpret this precisely as
-before in that during the random walk we collect the source in the
-states that are visited and apply them to the solution tally. We then
-define the probability that a particular random walk permuation of $k$
-events will occur:
-\begin{equation}
-  P_{\nu} = p_{i,i_1} p_{i_1,i_2} \cdots p_{i_{k-1},i_k}\:.
-  \label{eq:direct_permutation_probability}
-\end{equation}
-Finally, we define the expectation value of $X$ to be the collection
-of all random walk permutations and their probabilities:
-\begin{equation}
-  E\{X(i_0 = i)\} = \sum_{\nu} P_{\nu} X_{\nu}\:,
-  \label{eq:direct_expectation_value}
-\end{equation}
-which, if expanded, directly recovers the exact solution:
-\begin{equation}
-  \begin{split}
-    E\{X(i_0 = i)\}
-    &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-    \sum_{i_k}^{N} p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
-    w_{i,i_1}w_{i_1,i_2}\ldots w_{i_{k-1},i_k} b_{i_k}\\ &= x_i\:,
-  \end{split}
-  \label{eq:direct_expectation_expansion}
-\end{equation}
-therefore providing an unbiased Monte Carlo estimator.
-
-\subsubsection{Estimator Variance}
-\label{subsubsec:estimator_variance}
-We can compute the variance of the estimator through traditional
-methods by defining the variance, $\sigma_i$, for each component in
-the solution:
-\begin{equation}
-  {\sigma_i}^2 = E\{X(i_0 = i) - (\ve{A}^{-1}\ve{b})_i\}^2 = E\{X(i_0
-  = i)^2\} - x_i^2\:,
-  \label{eq:direct_variance_1}
-\end{equation}
-where the vector exponentials are computed element-wise. Inserting
-Eq~(\ref{eq:direct_expectation_value}) gives:
-\begin{equation}
-  \sigma_i^2 = \sum_{\nu} P_{\nu} X_{\nu}^2 - x_i^2\:,
-  \label{eq:direct_variance_2}
-\end{equation}
-and applying Eq~(\ref{eq:direct_permutation_contribution}):
-\begin{equation}
-  \sigma_i^2 = \sum_{\nu} P_{\nu} \sum_{m=0}^k W_{m}^2 b_{i_m}^2 -
-  x_i^2\:.
-  \label{eq:direct_variance_3}
-\end{equation}
-Finally, expanding the transition probabilities yields:
-\begin{equation}
-  \sigma_i^2 = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-  \sum_{i_k}^{N} p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
-  w^2_{i,i_1}w^2_{i_1,i_2}\ldots w^2_{i_{k-1},i_k} b_{i_m} - x_i^2\:.
-  \label{eq:direct_variance_4}
-\end{equation}
-Using this defintion for the variance, we can arrive at a more natural
-reason for enforcing $\rho(\ve{H}) < 1$. Per the Hadamard product, we can
-concatenate the summation in Eq~(\ref{eq:direct_variance_4}):
-\begin{equation}
-  (\ve{P} \circ \ve{W} \circ \ve{W})^k =
-  \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots \sum_{i_k}^{N}
-  p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
-  w^2_{i,i_1}w^2_{i_1,i_2}\ldots w^2_{i_{k-1},i_k}\:.
-  \label{eq:double_weighted_decomposition}
-\end{equation}
-If we assign $\ve{G} = \ve{P} \circ \ve{W} \circ \ve{W}$ as in
-Eq~(\ref{eq:neumann_ulam_decomposition}), we then have a new
-formulation for the variance:
-\begin{equation}
-  \sigma^2_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-  \sum_{i_k}^{N}g_{i,i_1}g_{i_1,i_2}\ldots g_{i_{k-1},i_k} b_{i_k}^2 -
-  x_i^2\:,
-\end{equation}
-which contains the general Neumann series for $\ve{G}$,
-\begin{equation}
-  \ve{T} = \sum_{k=0}^{\infty} \ve{G}^k\:,
-  \label{eq:variance_neumann_series}
-\end{equation}
-where $\ve{T} = (\ve{I}-\ve{G})^{-1}$. We can then insert $T$ back
-into the variance formulation:
-\begin{equation}
-  \sigma^2_i = (\ve{T}\ve{b})_i - x_i^2
-  \label{eq:direct_variance_5}
-\end{equation}
-We can relate $\ve{G}$ to $\ve{H}$ by noting that $\ve{G}$ simply
-contains an additional Hadamard product with the weight matrix. The
-Hadamard product has the property that:
-\begin{equation}
-  |\ve{H} \circ \ve{W}| \geq |\ve{H}|\ |\ve{W}|\:,
-  \label{eq:hadamard_inequality}
-\end{equation}
-meaning that $\rho(\ve{G}) \geq \rho(\ve{H})$. Using these relations
-to analyze Eq~(\ref{eq:direct_variance_5}), we see that if
-$\rho(\ve{H}) > 1$, then the sum in
-Eq~(\ref{eq:variance_neumann_series}) will not converge and an
-infinite variance will arise as the elements of $\ve{T}$ become
-infinite. Therefore, we must restrict $\ve{H}$ with $\rho(\ve{H}) < 1$
-so that our expectation values for the solution may have a finite
-variance.
-
-As we seek only approximate solutions, we need only to perform a
-minimal number of random walks in order to generate an approximation
-for $\ve{x}$. Furthermore, we also need conditions by which we may
-terminate a random walk. We do this by noticing that the factors added
-to Eq~(\ref{eq:direct_permutation_weight}) will become diminishingly
-small due to their definition in Eq~(\ref{eq:direct_weight}) such that
-their contributions to the solution estimate will become
-negligble. Therefore, we choose terminate a random walk sequence with
-a \textit{weight cutoff}, $W_c$, that is enforced when $W_m < W_c$ for
-a particular random walk permutation.
-
-\subsubsection{Reusing Random Walks}
-\label{subsubsec:reusing_random_walks}
-Ji recently proposed a variation of the direct method in which the
-random walk permutations in Eq~(\ref{eq:mc_walk_permutation}) can be
-used to contribute to the solution tally in all visited states instead
-of the only starting state \citep{ji_reusing_2012}.
-
-\subsection{Adjoint Method}
-\label{subsec:adjoint_mc}
-An alternative formulation for Monte Carlo matrix inversion is the
-adjoint method. In this scheme, random walk histories are born in a
-state generated by source sampling and tallies are generated in the
-states in which the random walk currently resides. In terms of
-particle transport, this adjoint method is equivalent to a traditional
-forward method. To begin, we first form the \textit{adjoint
-  Neumann-Ulam decomposition} of $\ve{H}$:
-\begin{equation}
-  \ve{H}^T = \ve{P} \circ \ve{W}\:,
-  \label{eq:adjoint_neumann_ulam}
-\end{equation}
-where now we are forming the decomposition with respect to the
-transpose of $\ve{H}$. We then following the same procedure as the
-direct method for forming the probability and weight matrices in the
-decomposition. Using the transpose form, probabilities should instead
-be column-scaled:
-\begin{equation}
-  p_{ij} = \frac{|h_{ji}|}{\sum_j |h_{ji}|}\:,
-  \label{eq:adjoint_probability}
-\end{equation}
-such that we expect to select a new state $j$ from the current state
-in the random walk $j$ by sampling column-wise. Per
-Eq~(\ref{eq:adjoint_neumann_ulam}), the transition weight is then
-defined as:
-\begin{equation}
-  w_{ij} = \frac{h_{ji}}{p_{ij}}\:.
-  \label{eq:adjoint_weight}
-\end{equation}
-Using the decomposition we can then define an expectation value for
-the adjoint method. Given Eq~(\ref{eq:direct_permutation_weight}) as
-the weight generated for a particular random walk permutation as in
-Eq~(\ref{eq:mc_walk_permutation}) and the fact that in the adjoint
-method we are only tallying the solution in the state in which the
-walk currently resides, the contribution to the solution from a
-particular random walk permutation is then:
-\begin{equation}
-  X_{\nu} = \sum_{m=0}^k W_{m} b_{i_0} \delta_{i,i_m}\:,
-  \label{eq:adjoint_permutation_contribution}
-\end{equation}
-where the Kronecker delta indicates that the tally contributes only in
-the current state and $b_{i_0}$ is that sampled source starting
-state. Note here that the estimator in
-Eq~(\ref{eq:adjoint_permutation_contribution}) does not have a
-dependency on the source state as in
-Eq~(\ref{eq:adjoint_permutation_contribution}), providing a remedy for
-the situation in the direct method where we must start a random walk
-in each source state for every permutation such that we may compute a
-contribution for that state. In the adjoint method, we instead tally
-in all states and those of lesser importance will not be visited as
-frequently by the random walk. Finally, the estimator using all
-permutations is:
-\begin{equation}
-  E\{X\} = \sum_{\nu} P_{\nu} X_{\nu}\:
-  \label{eq:adjoint_expectation_value}
-\end{equation}
-which, if expanded in the same way as the direct method, directly
-recovers the exact solution:
-\begin{equation}
-  \begin{split}
-    E\{X\} &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-    \sum_{i_k}^{N} p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
-    w_{i,i_1}w_{i_1,i_2}\ldots w_{i_{k-1},i_k} b_{i_k} \\ &= x_i\:,
-  \end{split}
-  \label{eq:adjoint_expectation_expansion}
-\end{equation}
-therefore also providing an unbiased Monte Carlo estimator.
-
-We also desire a criteria for random walk termination. For the adjoint
-method, we utilize a \textit{relative weight cutoff}:
-\begin{equation}
-  W_f = W_c b_{i_0}\:,
-  \label{eq:relative_weight_cutoff}
-\end{equation}
-where $W_c$ is defined as in the direct method. The adjoint random
-walk will then be terminated after $m$ steps if $W_m < W_f$ as tally
-contributions become increasingly small.
-
-Given these ideas, we can then generate an algorithm for implementing
-the adjoint method:
-\begin{algorithm}[htpb!]
-  \caption{Adjoint Monte Carlo}
-  \label{alg:adjoint_mc}
-  \begin{algorithmic}
-    \State $\ve{r}_0 := \ve{b}-\ve{A}\ve{x}_0$
-  \end{algorithmic}
-\end{algorithm}
-
-\subsection{Sequential Monte Carlo}
-\label{subsec:sequential_mc}
-The direct and adjoint methods described are limited by a convergence
-rate of $1/\sqrt{N}$ by the Central Limit Theorem where $N$ is the
-number of random walk permutations. In 1962, Halton presented a
-residual Monte Carlo method that moves towards exponential convergence
-rates \citep{halton_sequential_1962} and further refined his work some
-years later \citep{halton_sequential_1994}. Applications of his work
-by the transport community have confirmed convergence rates on the
-order of of $exp(-N)$
-\citep{evans_residual_2003}. In
-much the same way as projection methods, Halton's method, sequential
-Monte Carlo, utilizes the adjoint Monte Carlo solver as a means of
-driving down the residual vector. He proposed the following iterative
-scheme as a solution to Eq~(\ref{eq:linear_problem})\:
-\begin{subequations}
-  \begin{gather}
-    \ve{r}^k = \ve{b} - \ve{A}\ve{x}^k\:,\\  
-    \ve{A}\boldsymbol{\delta}^{k+1} = \ve{r}^{k}\:,\\
-    \ve{x}^{k+1} = \ve{x}^k + \boldsymbol{\delta}^{k+1}\:,
-  \end{gather}
-  \label{eq:sequential_monte_carlo}
-\end{subequations}
-where the correction $\boldsymbol{\delta}$ is computed by the adjoint
-Monte Carlo method. The merits of Halton's approach are immediately
-visible in that we have now broken the binding of the convergence rate
-to the Central Limit Thereom. Here, the Monte Carlo solve is used to
-produce a correction from the residual, analagous to using the
-residual to extract a correction from the search subspace in a
-projection method. By doing this, the Monte Carlo error is bound in
-the correction used to update the solution and therefore does not
-explicitly manifest itself in the solution. The downside of such a
-method is that if the solution guess is poor, then many iterations are
-required in order to reach exponential converge as the Monte Carlo
-error (and therefore the Central Limit Theorem) does dominate in this
-situation. Therefore, if a good initial guess is not available,
-Halton's method is still characterized by poor performance.
-
-\subsection{Monte Carlo Synthetic-Acceleration}
-\label{subsec:mcsa}
-Using the ideas of Halton, Evans and Mosher recently developed a Monte
-Carlo solution that was not prohibited severely by the quality of the
-initial guess for the system \citep{evans_monte_2009}. Their approach
-was instead to use residual Monte Carlo as a synthetic acceleration
-for a stationary method. To derive this method, we begin by splitting
-the operator in Eq~(\ref{eq:linear_problem})
-\begin{equation}
-  \ve{x} = (\ve{I} - \ve{A})\ve{x} + \ve{b}\:.
-  \label{eq:linear_split}
-\end{equation}
-With this we can then define the stationary method
-\textit{Richardson's iteration} as:
-\begin{equation}
-  \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,
-  \label{eq:richardsons_iteration}
-\end{equation}
-which will converge if $\rho(\ve{I} - \ve{A}) < 1$. We then define the
-solution error at the $k^{th}$ iterate relative to the true solution:
-\begin{equation}
-  \delta \ve{x}^k = \ve{x} - \ve{x}^k\:.
-  \label{eq:mcsa_error}
-\end{equation}
-Subtracting Eq~(\ref{eq:richardsons_iteration}) from
-Eq~(\ref{eq:linear_split}) we get:
-\begin{equation}
-  \delta \ve{x}^{k+1} = (\ve{I} - \ve{A})\delta \ve{x}^k\:.
-  \label{eq:mcsa_setup_1}
-\end{equation}
-Subtracting from this $(\ve{I} - \ve{A})\delta \ve{x}^{k+1}$ yields:
-\begin{equation}
-  \begin{split}
-    \ve{A}\delta \ve{x}^{k+1} &= (\ve{I} -
-    \ve{A})(\ve{x}^{k+1}-\ve{x}^{k}) \\ &= \ve{r}^{k+1}\:.
-    \label{eq:mcsa_setup_2}
-  \end{split}
-\end{equation}
-Using this, we define the following scheme that will converge in one
-iteration if $\ve{A}$ is inverted exactly.
-\begin{subequations}
-  \begin{gather}
-    \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,\\
-    \ve{A} \delta \ve{x}^{k+1} = \ve{r}^{k+1}\:,\\
-    \ve{x} = \ve{x}^{k+1} + \delta \ve{x}^{k+1}\:.
-  \end{gather}
-  \label{eq:mcsa_setup_3}
-\end{subequations}
-However, $\ve{A}$ is only approximately inverted by our numerical
-methods and therfore we instead pose an iterative scheme in which the
-Monte Carlo solvers are used to invert the operator. The
-\textit{Fixed-Point Monte Carlo Synthetic-Acceleration} (MCSA) method
-is defined as:
-\begin{subequations}
-  \begin{gather}
-    \ve{x}^{k+1/2} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,\\
-    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\
-    \ve{A}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:,\\
-    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,
-    \label{eq:mcsa}
-  \end{gather}
-\end{subequations}
-where the adjoint Monte Carlo method is used to generate the solution
-correction from the residual. Using Monte Carlo in this way achieves
-the same effect as Halton's method, decoupling its convergence rate
-from the overall convergence rate of the method. Here, the approximate
-Monte Carlo solution is not driven to a particular convergence as it
-merely supplies a correction for the initial guess generated by
-Richardson's iteration. Rather, only a set number of histories are
-required using the adjoint method to generate the correction.
-
-In addition to the Monte Carlo solver parameters dictating the number
-of histories and weight cutoff, the outer MCSA iterations also have
-the following stopping criteria:
-\begin{equation}
-  ||\ve{r}||_\infty < \epsilon \ ||\ve{b}||_\infty\:,
-  \label{eq:mcsa_stopping_criteria}
-\end{equation}
-where $\epsilon$ is a user-defined parameter. We therefore have 3
-parameters to tune in an MCSA implementation: the number of Monte
-Carlo histories computed in the adjoint solve during each MCSA
-iteration, the weight cutoff for those histories, and the total MCSA
-convergence tolerance as specified by $\epsilon$.
-
-\subsection{Preconditioning Stochastic Methods}
-\label{subsubsec:stochastic_preconditioning}
-In most cases, at least a minimal amount of \textit{preconditioning}
-of the linear system will be required in order to use the class of
-stochastic methods described. Although these methods have no symmetry
-requirements for convergence, they do require that the spectral radius
-of the iteration matrix be less than one. To achieve this, a Jacobi
-precondtioner, a form of left preconditioning, is used such that the
-preconditioning matrix $\ve{M}$ is:
-\begin{equation}
-  \ve{M} = diag(\ve{A})\:,
-  \label{eq:jacobi_preconditioner}
-\end{equation}
-such that its application means we are instead solving the following
-linear system:
-\begin{equation}
-  \ve{M}^{-1}\ve{A}\ve{x} = \ve{M}^{-1}\ve{b}\:.
-  \label{eq:jacobi_precond_linear_problem}
-\end{equation}
-Next, we can apply MCSA to solve
-Eq~(\ref{eq:jacobi_precond_linear_problem}): 
-\begin{subequations}
-  \begin{gather}
-    \ve{x}^{k+1/2} = (\ve{I} - \ve{M}^{-1}\ve{A})\ve{x}^k +
-    \ve{M}^{-1}\ve{b}\:,\\ \ve{r}^{k+1/2} = \ve{M}^{-1}\ve{b} -
-    \ve{M}^{-1}\ve{A}\ve{x}^{k+1/2}\:,\\ \ve{M}^{-1}\ve{A}\delta\ve{x}^{k+1/2}
-    = \ve{r}^{k+1/2}\:,\\ \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta
-    \ve{x}^{k+1/2}\:.
-    \label{eq:jacobi_preconditioned_mcsa}
-  \end{gather}
-\end{subequations}
-Choosing Jacobi preconditioning with MCSA is advantageous for several
-reasons. First, $\rho(\ve{I} - \ve{M}^{-1}\ve{A}) < 1$ is true for
-all $\ve{A}$ and is easy to formulate as the inversion of $\ve{M}$ is
-trivial. Second, because the adjoint Monte Carlo method used within MCSA to
-compute the correction operates on a linear problem with the
-preconditioned operator, then $\ve{H}$ in the adjoint solver will have
-a zero term in each of its diagonal elements, thereby eliminating
-all in-state transitions during the random walk sequence. Because of
-this, Jacobi preconditioning should always be performed, regardless of
-any other preconditioning that is applied to the system.
-
-\subsection{Parallelization of Stochastic Methods}
-\label{subsec:parallel_stochastic_methods}
-For Monte Carlo methods, and in particular MCSA, to be viable at the
-production scale, scalable parallel implementations are required. In a
-general linear algebra context, this has not yet been
-achieved. Therefore, we will develop and implement parallel algorithms
-for these methods levaraging both the knowledge gained from the
-general parallel implementations of Krylov methods and modern parallel
-strategies for Monte Carlo as developed by the reactor physics
-community. In order to formulate a parallel MCSA algorithm, we
-recognize that the algorithm occurs in two stages, an outer iteration
-performing Richardson's iteration and applying the correction, and an
-inner Monte Carlo solver that is generating the correction via the
-adjoint method. The parallel aspects of both these components must be
-considered.
-
-\subsubsection{Domain Decomposition for Monte Carlo}
-\label{subsubsec:msod}
-As observed in the discussion on parallel Krylov methods, large-scale
-problems will surely have their data partitioned such that each
-parllel process owns a subset of the equations in the linear
-system. Given this convention, the adjoint Monte Carlo algorithm must
-perform random walks over a domain that is decomposed and must remain
-decomposed due to memory limitations. This naturally leads us to seek
-parallel algorithms that handle domain decomposition. In the context
-of radiation transport, Brunner and colleagues provided a survey of
-algorithms for achieving this as implemented in production implicit
-Monte Carlo codes \citep{brunner_comparison_2006}. In their work they
-identity two data sets that are required to be communicated: the
-sharing of particles that are transported from one domain to another
-and therefore from one processor to another and a global communication
-that signals if particle transport has been completed on all
-processors. The algorithms presented are a fully-locking synchronous
-scheme, a asynchronous-send/synchronous-receive pattern, a traditional
-master/slave scheme, and a modified master/slave scheme the implements
-a binary tree pattern for the global reduction type operations needed
-to communicate between the master and slave processes. They observed
-that the modified master/slave scheme performed best in that global
-communications were implemented more efficiently than those required
-by the asynchronous scheme. Furthermore, none of these schemes handled
-load-imbalanced cases efficiently. Such cases will be common if the
-source sampled in the Monte Carlo random walk is not isotropic and
-evenly distributed throughout the global domain. It was noted that
-efficiencies were improved by increasing the frequency by which
-particle data was communicated between domain-adjacent processors.
-However, this ultimately increases communication costs.
-
-The 2006 work of Brunner is notable in that the Monte Carlo codes used
-to implement and test the algorithms adhered to a strict policy of
-generated identical results, indepdent of domain decomposition or
-domain replication as derived from the work of Gentile and colleagues
-\citep{gentile_obtaining_2005}. In their work, a procedure is given for
-obtaining results reproducable to machine precision for an arbitrary
-number of processors and domains. Differences can arise from using a 
-different random number sequence in each domain and performing
-a sequence of floating point operations on identical data in a
-different, leading to variations in round-off error and ultimately a
-non-identical answer. They use a simple example, recreated below in
-Figure~\ref{fig:gentile_example}, that illustrates these issues.
-\begin{figure}[htpb!]
-  \begin{center}
-    \scalebox{1.5}{
-      \input{chapters/linear_problem/gentile_example.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Gentile's example illustrating how domain
-      decomposition can create reproducability issues in Monte Carlo.}
-    \textit{Both particles A and B start in zone 1 on processor
-      1. Particle A moves to zone 2 on processor 2 and scatters back
-      to zone 1 while B scatters in zone 1 and remains there.}}
-  \label{fig:gentile_example}
-\end{figure}
-
-In 2010, Wagner and colleagues developed the \textit{multiple-set
-  overlapping-domain} (MSOD) decomposition for parallel Monte Carlo
-applications for full-core reactor analysis
-\citep{wagner_hybrid_2010}. In their work, a certain amount of overlap
-between domains was used to decrease the number of particles leaving
-the domain.
-
-\subsubsection{Parallel Adjoint Method}
-\label{subsubsec:parallel_adjoint}
-
-\subsubsection{Parallel MCSA}
-\label{subsubsec:parallel_mcsa}
-It should be noted here that a lot of the parallel matrix/vector
-operations discussed in the parallel projection methods section also
-apply here. Three of the four MCSA steps are in fact these operations
-with convergence checks also using a norm reduction.

### 2012-09-21 doc/prelim/document/chapters/parallel_mc/parallel_mc.tex
adding msod figure

\chapter{Stochastic Solution Methods for Linear Problems}
\label{chap:stochastic_methods}
An alternative approach to approximate matrix inversion is to employ
Monte Carlo methods that sample a distribution with an expectation
value equivalent to that of the inverted operator. Such methods have
been in existence for decades with the earliest reference noted here
an enjoyable manuscript published in 1950 by Forsythe and Leibler
\citep{forsythe_matrix_1950}. In their outline, Forsythe and Liebler
in fact credit the creation of this technique to J. Von Neumann and
S.M. Ulam some years earlier than its publication. In 1952 Wasow
provided a more formal explanation of Von Neumann and Ulam's method
\citep{wasow_note_1952} and Hammersley and Handscomb's 1964 monograph
\citep{hammersley_monte_1964} presents additional detail on this topic
using a collection of references from the 1950's and early 1960's.

We begin our discussion of stochastic methods using these texts by
seeking a solution to Eq~(\ref{eq:linear_problem}). For a given linear
operator $\ve{A}$, we can using diagonal splitting in the same manner
as the stationary method in Eq~(\ref{eq:linear_split_equation2}) to
define the following operator\footnote{It should be noted that
  non-diagonal splittings have been recently explored in
  \citep{srinivasan_monte_2010} and have the potential to improve
  efficiency.}:
\begin{equation}
  \ve{H} = \ve{I} - \ve{A}\:.
  \label{eq:linear_mc_iteration_matrix}
\end{equation}
We can then form an alternative representation for $\ve{A}^{-1}$ by
generating the \textit{Neumann series}:
\begin{equation}
  \ve{A}^{-1} = (\ve{I}-\ve{H})^{-1} = \sum_{k=0}^{\infty} \ve{H}^k\:,
  \label{eq:neumann_series}
\end{equation}
which will converge if the spectral radius of $\ve{H}$ is less than
1. If we then apply this inverse sum to the right hand side of
Eq~(\ref{eq:linear_problem}) we acquire the solution to the linear
problem:
\begin{equation}
  \ve{A}^{-1}\ve{b} = \sum_{k=0}^{\infty} \ve{H}^k\ve{b} = \ve{x}\:.
  \label{eq:neumann_solution}
\end{equation}
An approximation of this summation will therefore lead to an
approximation of the solution. If we expand the summation with a
succession of matrix-vector multiply operations, we arrive at an
alternative perspective of this summation by considering its $i^{th}$
component:
\begin{equation}
  x_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}\:,
  \label{eq:expanded_neumann_solution}
\end{equation}
which can interpreted as a series of transitions between states,
\begin{equation}
 \nu = i \rightarrow i_1 \rightarrow \cdots \rightarrow i_{k-1}
 \rightarrow i_{k}\:,
  \label{eq:mc_walk_permutation}
\end{equation}
in $\ve{H}$ where $\nu$ is a particular random walk sequence
permutation. We can generate these sequences of transitions through
Monte Carlo random walks by assigning them both a probability and
weight. As a reinterpretation of the iteration matrix, we then form
the \textit{Neumann-Ulam decomposition} of \ve{H}:
\begin{equation}
  \ve{H} = \ve{P} \circ \ve{W}\:,
  \label{eq:neumann_ulam_decomposition}
\end{equation}
where $\circ$ denotes the Hadamard product operation\footnote{The
  Hadamard product $\ve{A} = \ve{B} \circ \ve{C}$ is defined
  element-wise as $a_{ij} = b_{ij} c_{ij}$.}, $\ve{P}$ denotes the
transition probability matrix, and $\ve{W}$ denotes the transition
weight matrix. This decomposition, a generalization of Dimov's work
\citep{dimov_new_1998}, is an extension of the original Neumann-Ulam
scheme in that now a weight cutoff can be used to terminate a random
walk sequence. The formulation of $\ve{P}$ and $\ve{W}$ will be
dependent on whether we choose a direct or adjoint Monte Carlo
sequence to estimate the state transitions in
Eq~(\ref{eq:expanded_neumann_solution}).

\section{Direct Method}
\label{sec:direct_mc}
In the context of matrix inversion, a direct method resembles an
adjoint Monte Carlo method in the reactor physics community where the
solution state is sampled and the source terms that contribute to it
are assembled. To achieve this, we build the direct method
Neumann-Ulam decomposition per Dimov's approach by first choosing a
probability matrix that is a row scaling of $\ve{H}$ such that its
components are:
\begin{equation}
  p_{ij} = \frac{|h_{ij}|}{\sum_j |h_{ij}|}\:.
  \label{eq:direct_probability}
\end{equation}
From this, we then see that the probability of transitioning from a
state $i$ to a state $j$ is implicitly linked to the original operator
$\ve{A}$ in that those terms with large values, and therefore those
that make the greatest contribution to the numerical solution, will be
sampled with a higher probability than smaller terms. In addition, the
row scaling provides a normalization over the state to which we are
transitioning such that $\sum_j p_{ij} = 1$, meaning that we sample
the probabilities over the rows of the matrix. The components of
the weight matrix are then defined by
Eq~(\ref{eq:neumann_ulam_decomposition}) as:
\begin{equation}
  w_{ij} = \frac{h_{ij}}{p_{ij}}\:.
  \label{eq:direct_weight}
\end{equation}
It should be noted here that if $\ve{A}$ is sparse, then $\ve{H}$,
$\ve{P}$, and $\ve{W}$ must be sparse as well by
definition. Additionally, we only compute $\ve{P}$ and $\ve{W}$ from
the non-zero elements of $\ve{H}$ as those components that are zero
will not participate in the random walk. Furthermore, it prevents an
infinite weight from being generated in Eq~(\ref{eq:direct_weight}).

Using these matrices, we can then form the expectation value of the
direct solution. For a given random walk permutation $\nu$ with $k$
events, we define the weight of that permutation to be:
\begin{equation}
  W_{m} = \sum_{m=0}^k w_{i,i_1} w_{i_1,i_2} \cdots w_{i_{m-1},i_m}\:,
  \label{eq:direct_permutation_weight}
\end{equation}
such that the weight of each transition event contributes to the
total. The contribution to the solution from a particular random walk
permutation is then:
\begin{equation}
  X_{\nu}(i_0 = i) = \sum_{m=0}^k W_{m} b_{i_m}\:,
  \label{eq:direct_permutation_contribution}
\end{equation}
where $X_{\nu}(i_0 = i)$ signifies that the solution state in which we
are tallying defines state $i$.  We can interpret this precisely as
before in that during the random walk we collect the source in the
states that are visited and apply them to the solution tally. We then
define the probability that a particular random walk permutation of
$k$ events will occur:
\begin{equation}
  P_{\nu} = p_{i,i_1} p_{i_1,i_2} \cdots p_{i_{k-1},i_k}\:.
  \label{eq:direct_permutation_probability}
\end{equation}
Finally, we define the expectation value of $X$ to be the collection
of all random walk permutations and their probabilities:
\begin{equation}
  E\{X(i_0 = i)\} = \sum_{\nu} P_{\nu} X_{\nu}\:,
  \label{eq:direct_expectation_value}
\end{equation}
which, if expanded, directly recovers the exact solution:
\begin{equation}
  \begin{split}
    E\{X(i_0 = i)\}
    &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N} p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
    w_{i,i_1}w_{i_1,i_2}\ldots w_{i_{k-1},i_k} b_{i_k}\\ &= x_i\:,
  \end{split}
  \label{eq:direct_expectation_expansion}
\end{equation}
therefore providing an unbiased Monte Carlo estimator.

As we seek only approximate solutions, we need only to perform a
minimal number of random walks in order to generate an approximation
for $\ve{x}$. Furthermore, we also need conditions by which we may
terminate a random walk. We do this by noticing that the factors added
to Eq~(\ref{eq:direct_permutation_weight}) will become diminishingly
small due to their definition in Eq~(\ref{eq:direct_weight}) such that
their contributions to the solution estimate will become
negligble. Therefore, we choose terminate a random walk sequence with
a \textit{weight cutoff}, $W_c$, that is enforced when $W_m < W_c$ for
a particular random walk permutation.

\subsection{Estimator Variance}
\label{subsec:estimator_variance}
We can compute the variance of the estimator through traditional
methods by defining the variance, $\sigma_i$, for each component in
the solution:
\begin{equation}
  {\sigma_i}^2 = E\{X(i_0 = i) - (\ve{A}^{-1}\ve{b})_i\}^2 = E\{X(i_0
  = i)^2\} - x_i^2\:,
  \label{eq:direct_variance_1}
\end{equation}
where the vector exponentials are computed element-wise. Inserting
Eq~(\ref{eq:direct_expectation_value}) gives:
\begin{equation}
  \sigma_i^2 = \sum_{\nu} P_{\nu} X_{\nu}^2 - x_i^2\:,
  \label{eq:direct_variance_2}
\end{equation}
and applying Eq~(\ref{eq:direct_permutation_contribution}):
\begin{equation}
  \sigma_i^2 = \sum_{\nu} P_{\nu} \sum_{m=0}^k W_{m}^2 b_{i_m}^2 -
  x_i^2\:.
  \label{eq:direct_variance_3}
\end{equation}
Finally, expanding the transition probabilities yields:
\begin{equation}
  \sigma_i^2 = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N} p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
  w^2_{i,i_1}w^2_{i_1,i_2}\ldots w^2_{i_{k-1},i_k} b_{i_m} - x_i^2\:.
  \label{eq:direct_variance_4}
\end{equation}
Using this defintion for the variance, we can arrive at a more natural
reason for enforcing $\rho(\ve{H}) < 1$. Per the Hadamard product, we can
concatenate the summation in Eq~(\ref{eq:direct_variance_4}):
\begin{equation}
  (\ve{P} \circ \ve{W} \circ \ve{W})^k =
  \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots \sum_{i_k}^{N}
  p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
  w^2_{i,i_1}w^2_{i_1,i_2}\ldots w^2_{i_{k-1},i_k}\:.
  \label{eq:double_weighted_decomposition}
\end{equation}
If we assign $\ve{G} = \ve{P} \circ \ve{W} \circ \ve{W}$ as in
Eq~(\ref{eq:neumann_ulam_decomposition}), we then have a new
formulation for the variance:
\begin{equation}
  \sigma^2_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}g_{i,i_1}g_{i_1,i_2}\ldots g_{i_{k-1},i_k} b_{i_k}^2 -
  x_i^2\:,
\end{equation}
which contains the general Neumann series for $\ve{G}$,
\begin{equation}
  \ve{T} = \sum_{k=0}^{\infty} \ve{G}^k\:,
  \label{eq:variance_neumann_series}
\end{equation}
where $\ve{T} = (\ve{I}-\ve{G})^{-1}$. We can then insert $T$ back
into the variance formulation:
\begin{equation}
  \sigma^2_i = (\ve{T}\ve{b})_i - x_i^2
  \label{eq:direct_variance_5}
\end{equation}
We can relate $\ve{G}$ to $\ve{H}$ by noting that $\ve{G}$ simply
contains an additional Hadamard product with the weight matrix. The
Hadamard product has the property that:
\begin{equation}
  |\ve{H} \circ \ve{W}| \geq |\ve{H}|\ |\ve{W}|\:,
  \label{eq:hadamard_inequality}
\end{equation}
meaning that $\rho(\ve{G}) \geq \rho(\ve{H})$. Using these relations
to analyze Eq~(\ref{eq:direct_variance_5}), we see that if
$\rho(\ve{H}) > 1$, then the sum in
Eq~(\ref{eq:variance_neumann_series}) will not converge and an
infinite variance will arise as the elements of $\ve{T}$ become
infinite. Therefore, we must restrict $\ve{H}$ with $\rho(\ve{H}) < 1$
so that our expectation values for the solution may have a finite
variance.

\subsection{Reusing Random Walks}
\label{subsec:reusing_random_walks}
Ji recently proposed a variation of the direct method in which the
random walk permutations in Eq~(\ref{eq:mc_walk_permutation}) can be
used to contribute to the solution tally in all visited states instead
of the only starting state \citep{ji_reusing_2012}.

\section{Adjoint Method}
\label{sec:adjoint_mc}
An alternative formulation for Monte Carlo matrix inversion is the
adjoint method. In this scheme, random walk histories are born in a
state generated by source sampling and tallies are generated in the
states in which the random walk currently resides. In terms of
particle transport, this adjoint method is equivalent to a traditional
forward method. To begin, we first form the \textit{adjoint
  Neumann-Ulam decomposition} of $\ve{H}$:
\begin{equation}
  \ve{H}^{\dagger} = \ve{P} \circ \ve{W}\:,
  \label{eq:adjoint_neumann_ulam}
\end{equation}
where now we are forming the decomposition with respect to the
conjugate transpose of $\ve{H}$\footnote{This is sometimes referred to
  as the adjoint form of the matrix}. We then follow the same
procedure as the direct method for forming the probability and weight
matrices in the decomposition. Using the adjoint form, probabilities
should instead be column-scaled:
\begin{equation}
  p_{ij} = \frac{|h_{ji}|}{\sum_j |h_{ji}|}\:,
  \label{eq:adjoint_probability}
\end{equation}
such that we expect to select a new state $j$ from the current state
in the random walk $j$ by sampling column-wise. Per
Eq~(\ref{eq:adjoint_neumann_ulam}), the transition weight is then
defined as:
\begin{equation}
  w_{ij} = \frac{h_{ji}}{p_{ij}}\:.
  \label{eq:adjoint_weight}
\end{equation}
Using the decomposition we can then define an expectation value for
the adjoint method. Given Eq~(\ref{eq:direct_permutation_weight}) as
the weight generated for a particular random walk permutation as in
Eq~(\ref{eq:mc_walk_permutation}) and the fact that in the adjoint
method we are only tallying the solution in the state in which the
walk currently resides, the contribution to the solution from a
particular random walk permutation is then:
\begin{equation}
  X_{\nu} = \sum_{m=0}^k W_{m} b_{i_0} \delta_{i,i_m}\:,
  \label{eq:adjoint_permutation_contribution}
\end{equation}
where the Kronecker delta indicates that the tally contributes only in
the current state and $b_{i_0}$ is that sampled source starting
state. Note here that the estimator in
Eq~(\ref{eq:adjoint_permutation_contribution}) does not have a
dependency on the source state as in
Eq~(\ref{eq:adjoint_permutation_contribution}), providing a remedy for
the situation in the direct method where we must start a random walk
in each source state for every permutation such that we may compute a
contribution for that state. In the adjoint method, we instead tally
in all states and those of lesser importance will not be visited as
frequently by the random walk. Finally, the estimator using all
permutations is:
\begin{equation}
  E\{X\} = \sum_{\nu} P_{\nu} X_{\nu}\:
  \label{eq:adjoint_expectation_value}
\end{equation}
which, if expanded in the same way as the direct method, directly
recovers the exact solution:
\begin{equation}
  \begin{split}
    E\{X\} &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N} p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
    w_{i,i_1}w_{i_1,i_2}\ldots w_{i_{k-1},i_k} b_{i_k} \\ &= x_i\:,
  \end{split}
  \label{eq:adjoint_expectation_expansion}
\end{equation}
therefore also providing an unbiased Monte Carlo estimator.

We also desire a criteria for random walk termination. For the adjoint
method, we utilize a \textit{relative weight cutoff}:
\begin{equation}
  W_f = W_c b_{i_0}\:,
  \label{eq:relative_weight_cutoff}
\end{equation}
where $W_c$ is defined as in the direct method. The adjoint random
walk will then be terminated after $m$ steps if $W_m < W_f$ as tally
contributions become increasingly small.

Given these ideas, we can then generate an algorithm for implementing
the adjoint method:
\begin{algorithm}[htpb!]
  \caption{Adjoint Monte Carlo}
  \label{alg:adjoint_mc}
  \begin{algorithmic}
    \State $\ve{r}_0 := \ve{b}-\ve{A}\ve{x}_0$
  \end{algorithmic}
\end{algorithm}

\section{Sequential Monte Carlo}
\label{sec:sequential_mc}
The direct and adjoint methods described are limited by a convergence
rate of $1/\sqrt{N}$ by the Central Limit Theorem where $N$ is the
number of random walk permutations. In 1962, Halton presented a
residual Monte Carlo method that moves towards exponential convergence
rates \citep{halton_sequential_1962} and further refined his work some
years later \citep{halton_sequential_1994}. Applications of his work
by the transport community have confirmed convergence rates on the
order of of $exp(-N)$
\citep{evans_residual_2003}. In
much the same way as projection methods, Halton's method, sequential
Monte Carlo, utilizes the adjoint Monte Carlo solver as a means of
driving down the residual vector. He proposed the following iterative
scheme as a solution to Eq~(\ref{eq:linear_problem})\:
\begin{subequations}
  \begin{gather}
    \ve{r}^k = \ve{b} - \ve{A}\ve{x}^k\:,\\  
    \ve{A}\boldsymbol{\delta}^{k+1} = \ve{r}^{k}\:,\\
    \ve{x}^{k+1} = \ve{x}^k + \boldsymbol{\delta}^{k+1}\:,
  \end{gather}
  \label{eq:sequential_monte_carlo}
\end{subequations}
where the correction $\boldsymbol{\delta}$ is computed by the adjoint
Monte Carlo method. The merits of Halton's approach are immediately
visible in that we have now broken the binding of the convergence rate
to the Central Limit Thereom. Here, the Monte Carlo solve is used to
produce a correction from the residual, analagous to using the
residual to extract a correction from the search subspace in a
projection method. By doing this, the Monte Carlo error is bound in
the correction used to update the solution and therefore does not
explicitly manifest itself in the overall convergence of the
solution. The downside of such a method is that if the solution guess
is poor, then many iterations are required in order to reach
exponential converge as the Monte Carlo error (and therefore the
Central Limit Theorem) does dominate in this situation. Therefore, if
a good initial guess is not available, Halton's method is still
characterized by poor performance.

\section{Monte Carlo Synthetic-Acceleration}
\label{sec:mcsa}
Using the ideas of Halton, Evans and Mosher recently developed a Monte
Carlo solution that was not prohibited severely by the quality of the
initial guess for the system \citep{evans_monte_2009}. Their approach
was instead to use residual Monte Carlo as a synthetic acceleration
for a stationary method. To derive this method, we begin by splitting
the operator in Eq~(\ref{eq:linear_problem})
\begin{equation}
  \ve{x} = (\ve{I} - \ve{A})\ve{x} + \ve{b}\:.
  \label{eq:linear_split}
\end{equation}
With this we can then define the stationary method
\textit{Richardson's iteration} as:
\begin{equation}
  \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,
  \label{eq:richardsons_iteration}
\end{equation}
which will converge if $\rho(\ve{I} - \ve{A}) < 1$. We then define the
solution error at the $k^{th}$ iterate relative to the true solution:
\begin{equation}
  \delta \ve{x}^k = \ve{x} - \ve{x}^k\:.
  \label{eq:mcsa_error}
\end{equation}
Subtracting Eq~(\ref{eq:richardsons_iteration}) from
Eq~(\ref{eq:linear_split}) we get:
\begin{equation}
  \delta \ve{x}^{k+1} = (\ve{I} - \ve{A})\delta \ve{x}^k\:.
  \label{eq:mcsa_setup_1}
\end{equation}
Subtracting from this $(\ve{I} - \ve{A})\delta \ve{x}^{k+1}$ yields:
\begin{equation}
  \begin{split}
    \ve{A}\delta \ve{x}^{k+1} &= (\ve{I} -
    \ve{A})(\ve{x}^{k+1}-\ve{x}^{k}) \\ &= \ve{r}^{k+1}\:.
    \label{eq:mcsa_setup_2}
  \end{split}
\end{equation}
Using this, we define the following scheme that will converge in one
iteration if $\ve{A}$ is inverted exactly.
\begin{subequations}
  \begin{gather}
    \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,\\
    \ve{A} \delta \ve{x}^{k+1} = \ve{r}^{k+1}\:,\\
    \ve{x} = \ve{x}^{k+1} + \delta \ve{x}^{k+1}\:.
  \end{gather}
  \label{eq:mcsa_setup_3}
\end{subequations}
However, $\ve{A}$ is only approximately inverted by our numerical
methods and therfore we instead pose an iterative scheme in which the
Monte Carlo solvers are used to invert the operator. The
\textit{Fixed-Point Monte Carlo Synthetic-Acceleration} (MCSA) method
is defined as:
\begin{subequations}
  \begin{gather}
    \ve{x}^{k+1/2} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,\\
    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\
    \ve{A}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:,\\
    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,
    \label{eq:mcsa}
  \end{gather}
\end{subequations}
where the adjoint Monte Carlo method is used to generate the solution
correction from the residual. Using Monte Carlo in this way achieves
the same effect as Halton's method, decoupling its convergence rate
from the overall convergence rate of the method. Here, the approximate
Monte Carlo solution is not driven to a particular convergence as it
merely supplies a correction for the initial guess generated by
Richardson's iteration. Rather, only a set number of histories are
required using the adjoint method to generate the correction.

In addition to the Monte Carlo solver parameters dictating the number
of histories and weight cutoff, the outer MCSA iterations also have
the following stopping criteria:
\begin{equation}
  ||\ve{r}||_\infty < \epsilon \ ||\ve{b}||_\infty\:,
  \label{eq:mcsa_stopping_criteria}
\end{equation}
where $\epsilon$ is a user-defined parameter. We therefore have 3
parameters to tune in an MCSA implementation: the number of Monte
Carlo histories computed in the adjoint solve during each MCSA
iteration, the weight cutoff for those histories, and the total MCSA
convergence tolerance as specified by $\epsilon$.

\subsection{Preconditioning Stochastic Methods}
\label{sec:stochastic_preconditioning}
In most cases, at least a minimal amount of \textit{preconditioning}
of the linear system will be required in order to use the class of
stochastic methods described. Although these methods have no symmetry
requirements for convergence, they do require that the spectral radius
of the iteration matrix be less than one. To achieve this, a Jacobi
precondtioner, a form of left preconditioning, is used such that the
preconditioning matrix $\ve{M}$ is:
\begin{equation}
  \ve{M} = diag(\ve{A})\:,
  \label{eq:jacobi_preconditioner}
\end{equation}
such that its application means we are instead solving the following
linear system:
\begin{equation}
  \ve{M}^{-1}\ve{A}\ve{x} = \ve{M}^{-1}\ve{b}\:.
  \label{eq:jacobi_precond_linear_problem}
\end{equation}
Next, we can apply MCSA to solve
Eq~(\ref{eq:jacobi_precond_linear_problem}): 
\begin{subequations}
  \begin{gather}
    \ve{x}^{k+1/2} = (\ve{I} - \ve{M}^{-1}\ve{A})\ve{x}^k +
    \ve{M}^{-1}\ve{b}\:,\\ \ve{r}^{k+1/2} = \ve{M}^{-1}\ve{b} -
    \ve{M}^{-1}\ve{A}\ve{x}^{k+1/2}\:,\\ \ve{M}^{-1}\ve{A}\delta\ve{x}^{k+1/2}
    = \ve{r}^{k+1/2}\:,\\ \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta
    \ve{x}^{k+1/2}\:.
    \label{eq:jacobi_preconditioned_mcsa}
  \end{gather}
\end{subequations}
Choosing Jacobi preconditioning with MCSA is advantageous for several
reasons. First, $\rho(\ve{I} - \ve{M}^{-1}\ve{A}) < 1$ is true for
all $\ve{A}$ and is easy to formulate as the inversion of $\ve{M}$ is
trivial. Second, because the adjoint Monte Carlo method used within MCSA to
compute the correction operates on a linear problem with the
preconditioned operator, then $\ve{H}$ in the adjoint solver will have
a zero term in each of its diagonal elements, thereby eliminating
all in-state transitions during the random walk sequence. Because of
this, Jacobi preconditioning should always be performed, regardless of
any other preconditioning that is applied to the system.

\section{Parallelization of Stochastic Methods}
\label{sec:parallel_stochastic_methods}
For Monte Carlo methods, and in particular MCSA, to be viable at the
production scale, scalable parallel implementations are required. In a
general linear algebra context, this has not yet been
achieved. Therefore, we will develop and implement parallel algorithms
for these methods levaraging both the knowledge gained from the
general parallel implementations of Krylov methods and modern parallel
strategies for Monte Carlo as developed by the reactor physics
community. In order to formulate a parallel MCSA algorithm, we
recognize that the algorithm occurs in two stages, an outer iteration
performing Richardson's iteration and applying the correction, and an
inner Monte Carlo solver that is generating the correction via the
adjoint method. The parallel aspects of both these components must be
considered.

\subsection{Domain Decomposition for Monte Carlo}
\label{subsec:msod}
As observed in the discussion on parallel Krylov methods, large-scale
problems will surely have their data partitioned such that each
parallel process owns a subset of the equations in the linear
system. Given this convention, the adjoint Monte Carlo algorithm must
perform random walks over a domain that is decomposed and must remain
decomposed due to memory limitations. This naturally leads us to seek
parallel algorithms that handle domain decomposition
\citep{romano_towards_2010}.

In the context of radiation transport, Brunner and colleagues provided
a survey of algorithms for achieving this as implemented in production
implicit Monte Carlo codes \citep{brunner_comparison_2006}. In their
work they identify two data sets that are required to be communicated:
the sharing of particles that are transported from one domain to
another and therefore from one processor to another and a global
communication that signals if particle transport has been completed on
all processors. The algorithms presented are a fully-locking
synchronous scheme, an asynchronous-send/synchronous-receive pattern,
a traditional master/slave scheme, and a modified master/slave scheme
the implements a binary tree pattern for the global reduction type
operations needed to communicate between the master and slave
processes. They observed that the modified master/slave scheme
performed best in that global communications were implemented more
efficiently than those required by the asynchronous
scheme. Furthermore, none of these schemes handled load-imbalanced
cases efficiently. Such cases will be common if the source sampled in
the Monte Carlo random walk is not isotropic and evenly distributed
throughout the global domain. It was noted that efficiencies were
improved by increasing the frequency by which particle data was
communicated between domain-adjacent processors. However, this
ultimately increases communication costs. In 2009, Brunner extended
his work by using a more load-balanced approach with a fully
asynchronous communication pattern
\citep{brunner_efficient_2009}. Although the extended implementation
was more robust and allowed for scaling to larger numbers of
processors, performance issues were still noted with parallel
efficiency improvements needed in both the weak and strong scaling
cases for unbalanced problems. These results led Brunner to conclude
that a combination of domain decomposition and domain replication
could be used to solve some of these issues.

\subsubsection{Multiple-Set Overlapping-Domain Decomposition}
\label{subsubsec:msod}
In 2010, Wagner and colleagues developed the \textit{multiple-set
  overlapping-domain} (MSOD) decomposition for parallel Monte Carlo
applications for full-core light water reactor analysis
\citep{wagner_hybrid_2010}. In their work, an extension of Brunner's,
their scheme employed the similar parallel algorithms for particle
transport but a certain amount of overlap between domains was used to
decrease the number of particles leaving the domain. In addition,
Wagner utilized a level of replication of the domain such that the
domain was only decomposed on $O(100)$ processors and if replicated
$O(1000)$ times achieves simulation on $O(100,000)$ processors, thus
providing spatial and particle parellelism. Each collection of
processors that constitutes a representation of the entire domain is
referred to as a set, and within a set overlap occurs among its
subdomains. The original motivation was to decompose the domain in a
way that it remained in a physical cabinet in a large distributed
machine, thus reducing latency costs during communication. A multiple
set scheme is also motivated by the fact that communication only
occurs within a set, limiting communications during the transport
procedure to a group of $O(100)$ processors, a number that was shown
to have excellent parallel efficiencies in Brunner's work and
therefore will scale well in this algorithm. The overlapping domains
within each set also demonstrated to reduce communication costs. On
each processor, the source is sampled in the local domain that would
exist if no overlap was used while tallies can be made over the entire
overlapping domain. To demonstrate this, consider the example adapted
from Mervin's work with Wagner and others in the same area
\citep{mervin_variance_2012} and presented in
Figure~\ref{fig:msod_example}.
\begin{figure}[htpb!]
  \begin{center}
    \scalebox{1.5}{
      \input{chapters/parallel_mc/msod_example.pdftex_t} }
  \end{center}
  \caption{\textbf{Overlapping domain example illustrating how domain
      overlap can reduce communication costs.}
    \textit{All particles start in the blue region of interest. The
      dashed line represents 0.5 domain overlap between domains.}}
  \label{fig:msod_example}
\end{figure}
In this example, 3 particle histories are presented eminating from the
blue region of interest. Starting with particle A, if no domain
overlap is used then the only the blue domain exists on the starting
processor. Particle A is then transported through 3 other domains
before the history ends, therefore requiring three communications to
occur in Brunner's algorithm. If a 0.5 domain overlap is permitted as
shown by the dashed line, then the starting process owns enough of the
domain such that no communications must occur in order to complete the
particle A transport process. Using 0.5 domain overlap also easily
eliminates cases such as the represented by the path of particle C. In
this case, particle C is scattering between two adjacent domains,
incurring a large latency cost for a single particle. Finally, with
particle B we observe that 0.5 domain overlap will still not elminate
all communications. However, if 1 domain overlap were used, the entire
geometry shown in Figure~\ref{fig:msod_example} would be contained on
the source processor and therefore transport all 3 particles without
communication.

Wagner and colleagues used this methodology for a 2-dimensional
calculation of a pressurized water reactor core and varied the domain
overlap from 0 to 3 domain overlap (a $7 \times 7$ box in the context
of our example) where a domain constituted a fuel assembly. For the
fully domain decomposed case, they observed that 76.12\% of all source
particles leave the domain. At 1.5 domain overlap, that percentage of
source particles born in the center assembly leaving the processor
domain dropped to 1.05\% and even further for 0.02\% for the 3 domain
overlap. Based on these results, this overlap approach, coupled with
the multiple sets paradigm that will scale for existing parallel
transport algorithms, provides a scalable Monte Carlo algorithm for
today's modern machines.

Although domain decomposition was shown to be efficient in a load
balanced situation \citep{siegel_analysis_2012}, careful consideration
must be made for situations where this is not the case. Given the
stochastic nature of the problem and lack of a globally homogoneous
domain, parallel Monte Carlo simulations are inherently load
imbalanced. Procassini and others worked to alleviate some of the load
imbalances that are generated by both particle and spatial parallelism
and are therefore applicable to the MSOD algorithm
\citep{procassini_dynamic_2005}. They chose a dynamic balancing scheme
in which the number of times a particular domain was replicated was
dependent on the amount of work in that domain (i.e. domains with a
high particle flux and therefore more particle histories to
compute). In this variation, domains that require more work will be
replicated more frequently at reduced particle counts in each
replication. By doing this, load balance is improved as each domain
has approximately equal work although each domain may represent a
different spatial location and have a differing number of histories to
compute. Furthermore, they noted that as the simulation progressed and
particles were transported throughout the domain, the amount of
replication for each domain would vary as particle histories began to
diffuse, causing some regions to have higher work loads and some to
have smaller work loads than the initial conditions. Compared to
Wagner's work where the fission source was distributed relatively
evenly throughout the domain, fixed source problems (and especially
those that have a point-like source) like those presented in
Procassini's work will be more prone to changing load balance
requirements.

\subsubsection{Reproducable Domain Decomposed Results}
\label{subsubsec:reproducable_mc}
The 2006 work of Brunner is notable in that the Monte Carlo codes used
to implement and test the algorithms adhered to a strict policy of
generating identical results independent of domain decomposition or
domain replication as derived from the work of Gentile and colleagues
\citep{gentile_obtaining_2005}. In Gentile's work, a procedure is
given for obtaining results reproducable to machine precision for an
arbitrary number of processors and domains. Differences can arise from
using a different random number sequence in each domain and performing
a sequence of floating point operations on identical data in a
different order, leading to variations in round-off error and
ultimately a non-identical answer. They use a simple example,
recreated below in Figure~\ref{fig:gentile_example}, that illustrates
these issues.
\begin{figure}[htpb!]
  \begin{center}
    \scalebox{1.5}{
      \input{chapters/parallel_mc/gentile_example.pdftex_t} }
  \end{center}
  \caption{\textbf{Gentile's example illustrating how domain
      decomposition can create reproducability issues in Monte Carlo.}
    \textit{Both particles A and B start in zone 1 on processor
      1. Particle A moves to zone 2 on processor 2 and scatters back
      to zone 1 while B scatters in zone 1 and remains there. A1 and
      A2 denote the track of particle A that is in zone 1 while B1 and
      B2 denote the track of particle B that is in zone 1.}}
  \label{fig:gentile_example}
\end{figure}
In this example, the domain is decomposed on two processors with each
processor owning one of the two zones. Starting with particle A, it is
born in zone 1 and is transported to zone 2 where a scattering event
occurs. Concerning the first reproducability issue, if the same
sequence of random numbers is not used to compute the trajectory from
the new scattering event, we cannot expect to achieve the same result
if the domain were not decomposed. If the numbers are different, the
scattering event in zone 2 may keep the particle there or even eject
it from the domain if the sequence were different. The second issue is
demonstrated by adding another particle B that remains in the
domain. In this case, an efficient algorithm will transport particle A
on processor 1 until it leaves zone 1 and then transport particle
B. Particle A will not renter the domain until it has been
communicated to processor 2, processor 2 performs the transport, and
it is communicated back to processor 1. If we are doing a track-length
tally in zone 1, then we sum the tracks lengths observed in that
zone. In the single processor, single zone case particle A would be
transported in its entirety and then particle B transported. This
would result in a tally sum with the following order of operations:
$(((A1+A2)+B1)+B2)$. If we were instead to use 2 processors, we would
instead have the following order: $(((A1+B1)+B2)+A2)$. In the context
of floating point operations, we cannot expect these to have an
identical result to machine precesion as round-off approximations will
differ resulting in non-commutative addition.

Their solutions to these problems are elegant in that they require a
minimal amount of modification to be applied to the Monte Carlo
algorithm. To solve the first issue, in order to ensure each particle
maintains an invariant random number sequence that determines its
behavior regardless of domain decomposition, each particle carries
with it a random number seed that describes its current state. These
seeds are derived from the actual geometric location of the particle
such that it is decomposition invariant. Non-commutative floating
point operations are solved by instead mapping floating point values
to 64-bit integer values for which additions will always be
commutative. Once the operations are complete, these integers are
mapped back to floating point values.

\subsection{Parallel Adjoint Method}
\label{subsec:parallel_adjoint}
We can take much of what was learned from the survey of parllel Monte
Carlo methods for radiation transport and directly apply it to a
parallel formulation of our stochastic linear solvers.

\subsection{Parallel MCSA}
\label{subsec:parallel_mcsa}
It should be noted here that a lot of the parallel matrix/vector
operations discussed in the parallel projection methods section also
apply here. Three of the four MCSA steps are in fact these operations
with convergence checks also using a norm reduction.

### 2012-09-21 doc/prelim/document/chapters/research_proposal/research_proposal.tex
adding msod figure

@@
 
+\section{Experimental Framework}
+\label{sec:experimental_framework}
+
+The Chimera code base will be the primary research tool for answering
+the questions posed in the previous chapters of the document. We may
+not need a chapter for this, but it will likely be a good idea
+somewhere in this document to outline the design of the software and
+the instrumentation that it will provide.
+
+\subsection{Algorithm-Based Design}
+\label{subsec:chimera_design}
+
+\subsection{Graph-Based Multiphysics Formulation}
+\label{subsec:multiphysics_graph}
+
+\subsection{Finite Element Assembly}
+\label{subsec:fem_assembly}
+
+\subsection{Software Engineering}
+\label{subsec:software_engineering}
+
 \section{Progress to Date}
@@
 
-\section{Key Questions}
+\section{Numerical Experiments}
 \label{sec:key_questions}

### 2012-09-22 doc/prelim/document/chapters/parallel_mc/parallel_mc.tex
adding load balance example

@@
 therefore will scale well in this algorithm. The overlapping domains
-within each set also demonstrated to reduce communication costs. On
-each processor, the source is sampled in the local domain that would
-exist if no overlap was used while tallies can be made over the entire
+within each set also demonstrated reduced communication costs. On each
+processor, the source is sampled in the local domain that would exist
+if no overlap was used while tallies can be made over the entire
 overlapping domain. To demonstrate this, consider the example adapted
@@
 geometry shown in Figure~\ref{fig:msod_example} would be contained on
-the source processor and therefore transport all 3 particles without
-communication.
+the source processor and therefore transport of all 3 particles
+without communication would occur.
 
@@
 fully domain decomposed case, they observed that 76.12\% of all source
-particles leave the domain. At 1.5 domain overlap, that percentage of
+particles leave the domain. At 1.5 domain overlap, the percentage of
 source particles born in the center assembly leaving the processor
@@
 
+\subsubsection{Load Balancing Concerns}
+\label{subsubsec:mc_load_balancing}
 Although domain decomposition was shown to be efficient in a load
@@
 dependent on the amount of work in that domain (i.e. domains with a
-high particle flux and therefore more particle histories to
-compute). In this variation, domains that require more work will be
-replicated more frequently at reduced particle counts in each
-replication. By doing this, load balance is improved as each domain
-has approximately equal work although each domain may represent a
-different spatial location and have a differing number of histories to
-compute. Furthermore, they noted that as the simulation progressed and
-particles were transported throughout the domain, the amount of
-replication for each domain would vary as particle histories began to
-diffuse, causing some regions to have higher work loads and some to
-have smaller work loads than the initial conditions. Compared to
-Wagner's work where the fission source was distributed relatively
-evenly throughout the domain, fixed source problems (and especially
-those that have a point-like source) like those presented in
-Procassini's work will be more prone to changing load balance
-requirements.
+high particle flux and therefore more particle histories to compute
+require more work). In this variation, domains that require more work
+will be replicated more frequently at reduced particle counts in each
+replication. Consider the example in
+Figure~\ref{fig:procassini_example} adapted from Procassini's work.
+\begin{figure}[htpb!]
+  \begin{center}
+    \scalebox{1.5}{
+      \input{chapters/parallel_mc/procassini_example.pdftex_t} }
+  \end{center}
+  \caption{\textbf{Example illustrating how domain decomposition can
+      create load balance issues in Monte Carlo.}  \textit{A domain is
+      decomposed into 4 zones on 8 processors with a point source in
+      the lower left zone. As the particles diffuse from the source in
+      the random walk sequence as shown in the top row, their tracks
+      populate the entire domain. As given in the bottom row, as the
+      global percentage of particles increases in a zone, that zone's
+      replication count is increased.}}
+  \label{fig:procassini_example}
+\end{figure}
+In this example, a geometry is decomposed into 4 domains on 8
+processors with a point source in the bottom left domain. To begin,
+because the point source is concentrated in one domain, that domain is
+replicated 5 times such the amount of work it has to do per processor
+is roughly balanced with the others. As the particles begin to diffuse
+away from the point source, the amount of replication is adjusted to
+maintain load balance. Near the end of the simulation, the diffusion
+of particles is enough that all domains have equal replication.  By
+doing this, load balance is improved as each domain has approximately
+equal work although each domain may represent a different spatial
+location and have a differing number of histories to
+compute. Furthermore, Procassini and colleagues noted that as the
+simulation progressed and particles were transported throughout the
+domain, the amount of replication for each domain would vary as
+particle histories began to diffuse, causing some regions to have
+higher work loads and some to have smaller work loads than the initial
+conditions. Compared to Wagner's work where the fission source was
+distributed relatively evenly throughout the domain, fixed source
+problems (and especially those that have a point-like source) like
+those presented in Procassini's work will be more prone to changing
+load balance requirements.
 

### 2012-09-22 doc/prelim/document/chapters/parallel_mc/parallel_mc.tex
finised initial parallel mc content

@@
 Carlo solution that was not prohibited severely by the quality of the
-initial guess for the system \citep{evans_monte_2009}. Their approach
-was instead to use residual Monte Carlo as a synthetic acceleration
-for a stationary method. To derive this method, we begin by splitting
-the operator in Eq~(\ref{eq:linear_problem})
+initial guess for the system \citep{evans_monte_2009} and later
+applied it more rigorously as a solution mechanism for the radiation
+diffusion equation \citep{evans_monte_2012}, achieving identical
+numerical results as convential Krylov solvers. Their approach was
+instead to use residual Monte Carlo as a synthetic acceleration for a
+stationary method. To derive this method, we begin by splitting the
+operator in Eq~(\ref{eq:linear_problem})
 \begin{equation}
@@
     \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,
+  \end{gather}
   \label{eq:mcsa}
-  \end{gather}
 \end{subequations}
@@
 will be replicated more frequently at reduced particle counts in each
-replication. Consider the example in
+replication. Furthermore, Procassini and colleagues noted that as the
+simulation progressed and particles were transported throughout the
+domain, the amount of replication for each domain would vary as
+particle histories began to diffuse, causing some regions to have
+higher work loads and some to have smaller work loads than the initial
+conditions. Consider the example in
 Figure~\ref{fig:procassini_example} adapted from Procassini's work.
@@
 location and have a differing number of histories to
-compute. Furthermore, Procassini and colleagues noted that as the
-simulation progressed and particles were transported throughout the
-domain, the amount of replication for each domain would vary as
-particle histories began to diffuse, causing some regions to have
-higher work loads and some to have smaller work loads than the initial
-conditions. Compared to Wagner's work where the fission source was
+compute. Compared to Wagner's work where the fission source was
 distributed relatively evenly throughout the domain, fixed source
@@
 maintains an invariant random number sequence that determines its
-behavior regardless of domain decomposition, each particle carries
-with it a random number seed that describes its current state. These
-seeds are derived from the actual geometric location of the particle
-such that it is decomposition invariant. Non-commutative floating
-point operations are solved by instead mapping floating point values
-to 64-bit integer values for which additions will always be
-commutative. Once the operations are complete, these integers are
-mapped back to floating point values.
+behavior regardless of domain decomposition, each particle is a random
+number seed that describes its current state upon entering the domain
+of a new processor. These seeds are derived from the actual geometric
+location of the particle such that it is decomposition
+invariant. Non-commutative floating point operations are solved by
+instead mapping floating point values to 64-bit integer values for
+which additions will always be commutative. Once the operations are
+complete, these integers are mapped back to floating point values.
 
@@
 \label{subsec:parallel_adjoint}
-We can take much of what was learned from the survey of parllel Monte
+We can take much of what was learned from the survey of parallel Monte
 Carlo methods for radiation transport and directly apply it to a
-parallel formulation of our stochastic linear solvers.
+parallel formulation of our stochastic linear solvers. Direct analogs
+can be derived from these works by noting that the primary difference
+between solving a linear system with Monte Carlo methods and fixed
+source Monte Carlo transport problems is the content of Markov chains
+that are generated. These transitions represeneted by these chains are
+bound by probabilities and weights and are initiated by the sampling
+of a source. In the context of transport problems, those transitions
+represent events such as pariticle scattering and absorbition with
+probabilities that are determined by physical data in the form of
+cross sections. For stochastic matrix inversion, those transitions
+represent moving between the equations of the linear system (and
+therefore the physical domain which they represent) and their
+probabilites are defined by the coefficients of those
+equations. Ultimately, we tally the contributions to generate
+expectation values in the desired states as we progress through the
+chains.
+
+Therefore, parallel methods for Monte Carlo radiation transport and
+abstract those that apply to matrix inversion methods as an initial
+means of developing a parallel Neumann-Ulam-type solver. Based on the
+results observed in the last decade of parallel Monte Carlo
+development, we can generate many practical questions that this type
+of work could answer. Given a decomposed domain, it is clear that
+domain overlap significantly reduces the amount of communication
+required between adjacent domains as histories move to states that are
+not owned by the local processor. How much domain overlap is suitable
+for matrix inversion problems? Are we memory limited by large problems
+such that only so much overlap is feasible? Are there metrics related
+to the properties of the matrix such that we can provide guidelines
+for selecting the amount of overlap required? Furthermore, as Wagner
+and colleagues observed \citep{wagner_hybrid_2010}, with marginal
+domain overlap for reactor problems, the percentage of histories
+leaving the local domain can easily be reduced to less than 1\%. Evans
+and colleagues in their initial MCSA development typically used only
+50 particle histories in order to compute a correction in an MCSA
+iteration \citep{evans_monte_2012}. Per the Central Limit Theorem,
+that correlates to a statistical uncertainty of 14.1\% in the
+correction yet as compared to convential Krylov solvers, MCSA achieved
+the identical numerical results. If good convergence and numerically
+accurate solutions can be achieved with such a large uncertainty in
+the correction computation, then perhaps with enough domain overlap
+the minimal amount of histories that do transition to non-local states
+can be ignored and thus eliminate all communication in the parallel
+Monte Carlo adjoint solver and create an embarassingly parallel
+method. It will be important to determine if the expectation value
+bias in the MCSA correction generated by this approximation will
+remain within the bounds of the already high statistical uncertainty
+for unbiased estimates, thus providing numerically equivalent results.
+
+From a domain replication perspective, this may be difficult to
+achieve with a production scale linear solver. Typically, memory is at
+a premimium and therefore the more distinct domains available in the
+decomposition, the spatially finer and/or the numerically more
+accurate the discretization that can be implemented. How much
+replication is possible for large problems? How does replication
+facilitate parallel performance when coupled with domain overlap? How
+can we measure how much domain overlap is feasible? Replicating
+domains may therefore run into memory limitations for exceptionally
+large problems such that the operator, solution vector, and source
+vector must be copied in their entirety multiple times. From a
+resiliency standpoint, such an operation will be required, and
+therefore its performance and memory implications on conventional
+problems must be analyzed.
 
@@
 \label{subsec:parallel_mcsa}
-It should be noted here that a lot of the parallel matrix/vector
-operations discussed in the parallel projection methods section also
-apply here. Three of the four MCSA steps are in fact these operations
-with convergence checks also using a norm reduction.
+With a parallel adjoint Neumann-Ulam solver implementation, the
+parallel implementation of the MCSA method will be trivial. Recall the
+MCSA iteration procedure outlined in Eq~(\ref{eq:mcsa}). In
+\S~\ref{sec:parallel_krylov_methods} we discussed parallel matrix and
+vector operations as utilized in conventional Krylov methods. We
+utilize these here for the parallel MCSA implementation. In the first
+step, a parallel matrix-vector multiply is used to apply the split
+operator to the previous iterate's solution. A parallel vector update
+is then performed with the source vector to arrive at the initial
+iteration guess. In the next step, the residual is computed by the
+same operations where now the operator is applied to the solution
+guess with a parallel matrix-vector multiply and then a parallel
+vector update with the source vector is performed. Once the correction
+is computed with a parallel adjoint Neumann-Ulam solve, this
+correction is applied to the guess with a parallel vector update to
+get the new iteration solution. Additionally, as given by
+Eq~(\ref{eq:mcsa_stopping_criteria}), 2 parallel vector reductions
+will be required to check the stopping criteria: one initially to
+compute the infinity norm of the source vector, and another at every
+iteration to compute the infinity norm of the residual vector. For
+this implementation, all of the issues that will be potentially
+generated by the parallel adjoint solver implementation will manifest
+themselves here as the quality of the correction will be of intense
+study.

### 2012-09-22 doc/prelim/document/chapters/nonlinear_problem/nonlinear_problem.tex
working on parallel krylov section
adding msod figure
working on the nonlinear chapter

@@
-\chapter{The Nonlinear Problem}
+\chapter{Stochastic Solution Methods for Nonlinear Problems}
 \label{ch:nonlinear_problem}
-
-This work will result in a new and novel nonlinear solver, FANM. To
-movitate the development of FANM and thoroughly understand both its
-mathematical properties and its qualities as related to other
-nonlinear methods, we need a full workup of current iterative methods
-for solving nonlinear systems. We should acknowledge operator split
-and linearization methods for dealing with nonlinear systems
-(e.g. think pressure correction strategies for Navier-Stokes) but the
-focus here should be on fully-implicit nonlinear solves that are
-amenable to solving tightly-coupled multiphysics systems in a way that
-is fully consistent.
-
-The key here will be a full understanding of Newton methods (and they
-actually are quite simple compared to some of the linear solver
-methods like subspace methods). It is important to understand how the
-correction is computed from the Jacobian and how we may improve this
-with globalization methods. Although I may not use globalization
-methods in my work, they are important enough in practice that we must
-be sure that a FANM method is not prohibited from using these. Current
-methods, primarily Newton-Krylov methods, will of primary concern here
-as these will serve as the baseline for assessing the qualities of a
-FANM solver. Finally, we will survey applications to model problems in
-the field ( there is a vast amount of literature on fully-implicit
-nonlinear solves for the Navier-Stokes equations) that will give good
-benchmarks to test (such as natural convection and driven cavity
-problems).
+Nonlinear equation sets are a hallmark of multiphysics
+problems. Systems of partial differntial equations such as those that
+describe fluid flow or more general transport processes when
+discretized by conventional methods yield discrete sets of equations
+with nonlinearities present in the variables. Traditionally, such
+systems have been solved by linearizing them such that the
+nonlinearities in the variable are eliminated and more traditional
+linear methods can be used for solutions. Often characterized as
+segregated methods where physics operators are split and their action
+on the system approximated in steps, such methods lack consistency and
+accuracy in resolving the nonlinear component of the solution. In the
+last 30 years, fully consistent nonlinear methods based on Newton's
+method have become more popular and many advances have been made in
+the computational physics field to employ these methods. Currently,
+Krylov methods are used with Newton's method to form a robust solution
+sequence for these types of problems. However, given our goals of
+advancing stochastic solver technology for linear problems, we aim to
+apply them to these fully consistent schemes in order to gauge if some
+improvements can made. We therefore devise a new method using MCSA in
+conjuction with new methods, accomodating the fact that MCSA requires
+an explicitly formed linear operator in its implementation. In
+addition, we will review parallel strategies for a Newton-MCSA method.
 
@@
 \label{sec:nonlinear_preliminaries}
+We formulate the \textit{nonlinear problem} as follows
+\citep{knoll_jacobian-free_2004}:
+\begin{equation}
+  \ve{F}(\ve{u}) = \ve{0}\:,
+  \label{eq:nonlinear_problem}
+\end{equation}
+where $\ve{u}$ is the solution vector and $\ve{F}(\ve{u})$ is the
+function of nonlinear residuals. We write the nonlinear system in this
+form so that when an exact solution for $\ve{u}$ is achieved, all
+residuals evaluate to zero. \textit{Newton's method} is a root finding
+algorithm and therefore we can use it to solve
+Eq~(\ref{eq:nonlinear_problem}) if we interpret the exact solution
+$\ve{u}$ to be the roots of $\ve{F}(\ve{u})$. Newton's method is also
+an iterative scheme, and we can generate this scheme by generating the
+Taylor expansion of the $k+1$ iterate of the nonlinear residuals about
+the $k$ iterate:
+\begin{equation}
+  \ve{F}(\ve{u}^{k+1}) = \ve{F}(\ve{u}^{k}) +
+  \ve{F}'(\ve{u}^{k})(\ve{u}^{k+1}-\ve{u}^{k}) +
+  \frac{\ve{F}''(\ve{u}^{k})}{2}(\ve{u}^{k+1}-\ve{u}^{k})^2 + \cdots
+  \:.
+  \label{eq:newton_derivation_1}
+\end{equation}
+If we ignore the nonlinear terms in the expansion and assert that the
+$k+1$ iterate $\ve{u}^{k+1}$ is that exact solution such that
+$\ve{F}(\ve{u}^{k+1}) = \ve{0}$, then we are left with the following:
+\begin{equation}
+  -\ve{F}(\ve{u}^{k}) =
+  \ve{F}'(\ve{u}^{k})(\ve{u}^{k+1}-\ve{u}^{k})\:.
+  \label{eq:newton_derivation_2}
+\end{equation}
+We note two things in Eq~(\ref{eq:newton_derivation_2}). The first is
+that $\ve{F}'(\ve{u}^{k})$ is in fact the \textit{Jacobian},
+$\ve{J}(\ve{u})$, of the set of nonlinear residuals and is defined
+element-wise as:
+\begin{equation}
+  J_{ij} = \frac{\partial F_i(\ve{u})}{\partial u_j}\:.
+  \label{eq:jacobian_def}
+\end{equation}
+Second, we note that $(\ve{u}^{k+1}-\ve{u}^{k})$ is simply the
+\textit{Newton correction} at the $k$ iterate, $\delta \ve{u}^k$. We
+can then rerarrange Eq~(\ref{eq:newton_derivation_2}) to define the
+Newton iteration scheme for nonlinear problems:
+\begin{subequations}
+  \begin{gather}
+    \ve{J}(\ve{u}) \delta \ve{u}^k = -\ve{F}(\ve{u}^{k})\\
+    \ve{u}^{k+1} = \ve{u}^k + \delta \ve{u}^k\:.
+  \end{gather}
+  \label{eq:newton_iteration}
+\end{subequations}
+We then have two distinct steps to perform: the solution of a linear
+system to compute the Newton correction where the Jacobian matrix of
+the nonlinear equation set is the linear operator, and the application
+of the correction to the previous iterate's solution to arrive at the
+next iterate's solution. In the asymptotic lmiit, the iterations of
+Newton's method will converge the nonlinear residual quadratically
+\citep{kelley_iterative_1995}. Convergence criteria is set for
+stopping the iteration sequence based on the nonlinear
+residual. Commonly the following criteria is used:
+\begin{equation}
+  ||\ve{F}(\ve{u}^{k})|| < \epsilon ||\ve{F}(\ve{u}^{0})||\:,
+  \label{eq:newton_stopping_criteria}
+\end{equation}
+where $\epsilon$ is a user defined tolerance parameter. Newton's
+method is \textit{consistent} in that all components of the nonlinear
+functions that describe the physics we are modeling are updated
+simulataneously in the iteration sequence with respect to one
+another. This is in comparison to \textit{inconsistent} strategies,
+such as a pressure correction strategy for solving the Navier-Stokes
+equations \citep{pletcher_computational_1997}, where the components of
+$\ve{u}$ are updated in a staggered fashion depending on the
+particular equations that they are associated with.
 
-\section{Iterative Methods for Solving Nonlinear Systems}
-\label{sec:nonlinear_methods}
-
-\subsection{Inexact Newton Methods}
-\label{subsec:newton_methods}
+\section{Inexact Newton Methods}
+\label{sec:newton_methods}
+Inexact Newton methods arise when the Jacobian operator is not exactly
+inverted, resulting in an inexact Newton correction as initially
+described by Dembo and others \citep{dembo_inexact_1982}. For common
+sparse nonlinear systems, which in turn yield a sparse Jacobian
+matrix, this situation occurs when conventional iterative methods are
+applied.
 
@@
 \label{subsec:newton_krylov_methods}
+A form of inexact Newton methods, \textit{Newton-Krylov methods} are
+nonlinear iterative methods that use a Krylov subspace method as the
+linear solver for generating the Newton correction. As we investigated
+in Chapter~\ref{chap:linear_problem}, Krylov methods are robust and
+enjoy efficient parallel implementations. Furthermore, their lack of
+explicit dependence on the operator make them easier to implement than
+other methods.
 
-\subsection{Globalization Methods}
-\label{subsec:globalization_methods}
+In most cases, the Jacobian is difficult to form from the difference
+equations. For simple nonlinear cases such as the Navier Stokes
+equations, the derivatives can be computed and coded, but due to the
+complexity of those derivatives and the resulting difference equations
+this task can be tedious, error prone, and must be repeated for every
+equation set. By recognizing that Krylov methods only need the action
+of the operator on the vector instead of the operator itself, the
+Jacobian can instead be approximated through various numerical methods
+including a difference-based Jacobian-free formulation. Jacobian-Free
+methods, and in particular \textit{Jacobian-Free Newton-Krylov} (JFNK)
+methods \citep{knoll_jacobian-free_2004}, rely on forming the action
+of the Jacobian on a vector as required by the Krylov solver through a
+forward difference scheme. In this case, the action of the Jacobian on
+some vector $\ve{v}$ is given as:
+\begin{equation}
+  \ve{J}(\ve{u})\ve{v} = \frac{\ve{F}(\ve{u} + \epsilon \ve{v}) -
+    \ve{F}(\ve{u})}{\epsilon}\:,
+  \label{eq:jacobian_free_product}
+\end{equation}
+where $\epsilon$ is a small number typically on the order of machine
+precision. Kelley \citep{kelley_iterative_1995} points out a potential
+downfall of this formulation in that if the discretization error in
+$\ve{F}(\ve{u})$ is on the order of the perturbation parameter
+$\epsilon$, then the finite difference error from
+Eq~(\ref{eq:jacobian_free_product}) pollutes the solution.
 
@@
 that have already been worked up in the literature with benchmark-type
-solutions.
+solutions. Parallelism in the context of the linear solvers chapter
+should also be discussed.
 
-\subsection{Automatic Differentiation}
-\label{sec:automatic_differentiation}
+\subsubsection{Automatic Differentiation}
+\label{subsubsec:automatic_differentiation}
 
 \subsection{Residual and Jacobian Generation}
-\label{sec:fanm_generation}
+\label{subsec:fanm_generation}
 
 \subsection{Jacobian Storage vs. Subspace Storage and Restarts}
-\label{sec:fanm_storage}
+\label{subsec:fanm_storage}
+
+\subsection{Parallel FANM Implementation}
+\label{subsec:parallel_fanm}

### 2012-09-22 doc/prelim/document/chapters/parallel_mc/parallel_mc.tex
working on the nonlinear chapter

@@
 study.
+
+In addition to parallel implementation and performance, MCSA's
+potential for aiding advancement in non-symmetric matrix solutions
+leads to a natural comparison with the GMRES algorithm. As both
+solvers are aimed at the same class of problem, we desire a set of
+metrics that will allow us to quantitatively compare the two. Given
+that the Krylov subspace maintained by GMRES can become large, do we
+benifit from a memory standpoint with an MCSA scheme in that no
+subspace is required? Does this benefit outweight the fact that the
+linear operator must be explicitly formed in order to build the
+transition probabilities for the random walk sequence? For
+non-symmetric systems, does MCSA exhibit similar convergence
+properties to Krylov methods? If a Krylov methods build a subspace,
+can those memory savings in MCSA be used to implement domain
+replication in the adjoint solver? Such questions can be answered by a
+comparative study of the two solvers that controls the system size and
+the iterations required to converge.

### 2012-09-22 doc/prelim/document/chapters/nonlinear_problem/nonlinear_problem.tex
adding load balance example

@@
 \end{equation}
-where $\ve{u}$ is the solution vector and $\ve{F}(\ve{u})$ is the
-function of nonlinear residuals. We write the nonlinear system in this
-form so that when an exact solution for $\ve{u}$ is achieved, all
-residuals evaluate to zero. \textit{Newton's method} is a root finding
-algorithm and therefore we can use it to solve
-Eq~(\ref{eq:nonlinear_problem}) if we interpret the exact solution
-$\ve{u}$ to be the roots of $\ve{F}(\ve{u})$. Newton's method is also
-an iterative scheme, and we can generate this scheme by generating the
-Taylor expansion of the $k+1$ iterate of the nonlinear residuals about
-the $k$ iterate:
+where $\ve{u} \in \mathbb{C}^n$ is the solution vector and
+$\ve{F}:\mathbb{C}^N \rightarrow \mathbb{C}^N$ is the function of
+nonlinear residuals. We write the nonlinear system in this form so
+that when an exact solution for $\ve{u}$ is achieved, all residuals
+evaluate to zero. \textit{Newton's method} is a root finding algorithm
+and therefore we can use it to solve Eq~(\ref{eq:nonlinear_problem})
+if we interpret the exact solution $\ve{u}$ to be the roots of
+$\ve{F}(\ve{u})$. Newton's method is also an iterative scheme, and we
+can generate this scheme by generating the Taylor expansion of the
+$k+1$ iterate of the nonlinear residuals about the $k$ iterate:
 \begin{equation}
@@
 matrix, this situation occurs when conventional iterative methods are
-applied.
+applied. In their definition, Dembo formulated inexact methods such
+that they are independent of the linear method used to solve for the
+Newton correction and therefore are amenable to use with any linear
+solver.
 
@@
 linear solver for generating the Newton correction. As we investigated
-in Chapter~\ref{chap:linear_problem}, Krylov methods are robust and
+in Chapter~\ref{ch:linear_problem}, Krylov methods are robust and
 enjoy efficient parallel implementations. Furthermore, their lack of
@@
 
+\subsubsection{Choosing a Krylov Method}
+\label{subsubsec:krylov_choice}
+From their numerical studies, McHugh and Knoll found
+that the robustness enjoyed by the long recurrence relations used by
+GMRES for the orthogonalization procedure proved to be a better choice
+for use with Newton's method
+\citep{mchugh_inexact_1993,knoll_newton-krylov_1995}.
+
+\subsubsection{Jacobian-Free Approximation}
+\label{subsubsec:jacobian_free_approximation}
 In most cases, the Jacobian is difficult to form from the difference

### 2012-09-22 doc/prelim/document/chapters/nonlinear_problem/nonlinear_problem.tex
nonlinear updates

@@
 Newton correction and therefore are amenable to use with any linear
-solver.
+solver. Furthermore, they bind the convergence of the outer nonlinear
+iteration to the inner linear iteration such that:
+\begin{equation}
+  ||\ve{J}(\ve{u}^k)\delta \ve{u}^k + \ve{F}(\ve{u}^k)|| \leq \eta^k
+  ||\ve{F}(\ve{u}^k)||\:,
+  \label{eq:inexact_newton_forcing}
+\end{equation}
+where $\eta^k$ is defined as the \textit{forcing term} at the $k$
+iterate. Eq~(\ref{eq:inexact_newton_forcing}) then states that the
+residual generated by the linear residual is bound by the nonlinear
+residual and how tightly it is bound is defined by the forcing
+term. This is useful in that we can vary how tightly coupled the
+Newton correction is to the nonlinear iteration by relaxing or
+tightening the convergence properties on the linear iterative
+method. As a result strategies for determining the forcing term can
+vary depending on the problem type \citep{eisenstat_choosing_1996}.
 
@@
 A form of inexact Newton methods, \textit{Newton-Krylov methods} are
-nonlinear iterative methods that use a Krylov subspace method as the
-linear solver for generating the Newton correction. As we investigated
-in Chapter~\ref{ch:linear_problem}, Krylov methods are robust and
-enjoy efficient parallel implementations. Furthermore, their lack of
-explicit dependence on the operator make them easier to implement than
-other methods.
+nonlinear iterative methods that leverage a Krylov subspace method as
+the linear solver for generating the Newton correction. As we
+investigated in Chapter~\ref{ch:linear_problem}, Krylov methods are
+robust and enjoy efficient parallel implementations. Furthermore,
+their lack of explicit dependence on the operator make them easier to
+implement than other methods.
 
-\subsubsection{Choosing a Krylov Method}
-\label{subsubsec:krylov_choice}
-From their numerical studies, McHugh and Knoll found
+With many Krylov methods available, which to use with the Newton
+method is dependent on many factors including convergence rates and
+memory usage. From their numerical studies, McHugh and Knoll found
 that the robustness enjoyed by the long recurrence relations used by

### 2012-09-24 doc/prelim/document/chapters/linear_problem/linear_problem.tex
nonlinear updates

@@
 \end{equation}
-where $\ve{A} \in \mathbb{C}^{N \times N}$ is a matrix operator such
-that $\ve{A} : \mathbb{C}^{N} \rightarrow \mathbb{C}^{N}$, $\ve{x} \in
-\mathbb{C}^N$ is the solution vector, and $\ve{b} \in \mathbb{C}^N$ is
+where $\ve{A} \in \mathbb{R}^{N \times N}$ is a matrix operator such
+that $\ve{A} : \mathbb{R}^{N} \rightarrow \mathbb{R}^{N}$, $\ve{x} \in
+\mathbb{R}^N$ is the solution vector, and $\ve{b} \in \mathbb{R}^N$ is
 the forcing term. The solutions to Eq~(\ref{eq:linear_problem}) will
@@
   ||\ve{v}||_q = \Bigg[ \sum_{i=1}^N |v_i|^q \Bigg]^{1/q},\ \ve{v} \in
-  \mathbb{C}^N\:,\ q \in \mathbb{Z}^+
+  \mathbb{R}^N\:,\ q \in \mathbb{Z}^+
   \label{eq:q_norm}
@@
 Eq~(\ref{eq:matrix_norm_inequality}) is equivalent to $||\ve{A}||$
-and is valid $\forall \ve{x} \in \mathbb{C}^N$. The general
+and is valid $\forall \ve{x} \in \mathbb{R}^N$. The general
 definition in Eq~(\ref{eq:matrix_norm_inequality}) can be expanded in
@@
   ||\tilde{\ve{r}}||_2 \leq ||\ve{r}_0||_2,\ \forall \ve{r}_0 \in
-  \mathbb{C}^N\:,
+  \mathbb{R}^N\:,
   \label{eq:linear_min_res}
@@
 \begin{equation}
-  \boldsymbol{\delta} = \ve{V}\ve{y},\ \forall \ve{y} \in \mathbb{C}^N:\,
+  \boldsymbol{\delta} = \ve{V}\ve{y},\ \forall \ve{y} \in \mathbb{R}^N:\,
   \label{eq:linear_delta_projection}
@@
 recurrence relation. The Arnoldi procedure will generate an
-orthonormal basis, $\ve{V}_m \in \mathbb{C}^{N \times m}$, via a
+orthonormal basis, $\ve{V}_m \in \mathbb{R}^{N \times m}$, via a
 variant of the Gram-Schmidt procedure that re-applies the operator for
@@
 where $\ve{y}$ and $\ve{x}$ are vectors of global size $N_g$, local
-size $N_l$, and $a \in \mathbb{C}^N$. In order to avoid communcation,
+size $N_l$, and $a \in \mathbb{R}^N$. In order to avoid communcation,
 the vectors $\ve{y}$ and $\ve{x}$ must have the same parallel

### 2012-09-24 doc/prelim/document/chapters/nonlinear_problem/nonlinear_problem.tex
nonlinear updates

--- 2012-09-22 doc/prelim/document/chapters/nonlinear_problem/nonlinear_problem.tex
@@
 \label{ch:nonlinear_problem}
-Nonlinear equation sets are a hallmark of multiphysics
+Nonlinear equation sets are a common occurence in multiphysics
 problems. Systems of partial differntial equations such as those that
@@
 method have become more popular and many advances have been made in
-the computational physics field to employ these methods. Currently,
-Krylov methods are used with Newton's method to form a robust solution
-sequence for these types of problems. However, given our goals of
-advancing stochastic solver technology for linear problems, we aim to
-apply them to these fully consistent schemes in order to gauge if some
-improvements can made. We therefore devise a new method using MCSA in
-conjuction with new methods, accomodating the fact that MCSA requires
-an explicitly formed linear operator in its implementation. In
-addition, we will review parallel strategies for a Newton-MCSA method.
+the computational physics field to employ these methods.
+
+Currently, Krylov methods are used with Newton's method to form a
+robust solution sequence for these types of problems. However, given
+our goals of advancing stochastic solver technology for linear
+problems, we aim to apply them to these fully consistent schemes in
+order to gauge if some improvements can be made. We therefore devise a
+new method using MCSA in conjuction with new methods, accomodating the
+fact that MCSA requires an explicitly formed linear operator in its
+implementation as compared to Krylov methods. In addition to handling
+this dependency, we will review parallel strategies for a nonlinear
+MCSA method.
 
@@
 \end{equation}
-where $\ve{u}$ is the solution vector and $\ve{F}(\ve{u})$ is the
-function of nonlinear residuals. We write the nonlinear system in this
-form so that when an exact solution for $\ve{u}$ is achieved, all
-residuals evaluate to zero. \textit{Newton's method} is a root finding
-algorithm and therefore we can use it to solve
-Eq~(\ref{eq:nonlinear_problem}) if we interpret the exact solution
-$\ve{u}$ to be the roots of $\ve{F}(\ve{u})$. Newton's method is also
-an iterative scheme, and we can generate this scheme by generating the
-Taylor expansion of the $k+1$ iterate of the nonlinear residuals about
-the $k$ iterate:
+where $\ve{u} \in \mathbb{R}^n$ is the solution vector and
+$\ve{F}:\mathbb{R}^N \rightarrow \mathbb{R}^N$ is the function of
+nonlinear residuals. We write the nonlinear system in this form so
+that when an exact solution for $\ve{u}$ is achieved, all residuals
+evaluate to zero. \textit{Newton's method} is a root finding algorithm
+and therefore we can use it to solve Eq~(\ref{eq:nonlinear_problem})
+if we interpret the exact solution $\ve{u}$ to be the roots of
+$\ve{F}(\ve{u})$. Newton's method is also an iterative scheme, and we
+can generate this procedure by building the Taylor expansion of the
+$k+1$ iterate of the nonlinear residuals about the previous $k$
+iterate:
 \begin{equation}
@@
 \end{equation}
-If we ignore the nonlinear terms in the expansion and assert that the
-$k+1$ iterate $\ve{u}^{k+1}$ is that exact solution such that
-$\ve{F}(\ve{u}^{k+1}) = \ve{0}$, then we are left with the following:
+If we ignore the nonlinear terms in the expansion and assert that at
+the $k+1$ iterate $\ve{u}^{k+1}$ is that exact solution such that
+$\ve{F}(\ve{u}^{k+1}) = \ve{0}$, then we are left with the following
+equality:
 \begin{equation}
@@
 \end{equation}
-We note two things in Eq~(\ref{eq:newton_derivation_2}). The first is
-that $\ve{F}'(\ve{u}^{k})$ is in fact the \textit{Jacobian},
+We note two things of importance in
+Eq~(\ref{eq:newton_derivation_2}). The first is that
+$\ve{F}'(\ve{u}^{k})$ is in fact the \textit{Jacobian},
 $\ve{J}(\ve{u})$, of the set of nonlinear residuals and is defined
@@
 Second, we note that $(\ve{u}^{k+1}-\ve{u}^{k})$ is simply the
-\textit{Newton correction} at the $k$ iterate, $\delta \ve{u}^k$. We
-can then rerarrange Eq~(\ref{eq:newton_derivation_2}) to define the
-Newton iteration scheme for nonlinear problems:
+solution update from the $k$ iterate to the $k+1$ iterate. We will
+define this update as the \textit{Newton correction} at the $k$
+iterate, $\delta \ve{u}^k$. We can then rerarrange
+Eq~(\ref{eq:newton_derivation_2}) to define the Newton iteration
+scheme for nonlinear problems:
 \begin{subequations}
@@
 \end{subequations}
-We then have two distinct steps to perform: the solution of a linear
-system to compute the Newton correction where the Jacobian matrix of
-the nonlinear equation set is the linear operator, and the application
-of the correction to the previous iterate's solution to arrive at the
-next iterate's solution. In the asymptotic lmiit, the iterations of
-Newton's method will converge the nonlinear residual quadratically
-\citep{kelley_iterative_1995}. Convergence criteria is set for
-stopping the iteration sequence based on the nonlinear
-residual. Commonly the following criteria is used:
+We then have three distinct steps to perform: evaluation of the
+nonlinear residuals using the solution at the $k$ iterate, the
+solution of a linear system to compute the Newton correction where the
+Jacobian matrix of the nonlinear equation set is the linear operator,
+and the application of the correction to the previous iterate's
+solution to arrive at the next iterate's solution. In the asymptotic
+limit, the iterations of Newton's method will converge the nonlinear
+residual quadratically \citep{kelley_iterative_1995}. Convergence
+criteria is set for stopping the iteration sequence based on the
+nonlinear residual. Commonly, the following criteria is used:
 \begin{equation}
@@
 matrix, this situation occurs when conventional iterative methods are
-applied.
+applied. In their definition, Dembo formulated inexact methods such
+that they are independent of the linear method used to solve for the
+Newton correction and therefore are amenable to use with any linear
+solver. Furthermore, they bind the convergence of the outer nonlinear
+iteration to the inner linear iteration such that:
+\begin{equation}
+  ||\ve{J}(\ve{u}^k)\delta \ve{u}^k + \ve{F}(\ve{u}^k)|| \leq \eta^k
+  ||\ve{F}(\ve{u}^k)||\:,
+  \label{eq:inexact_newton_forcing}
+\end{equation}
+where $\eta^k \in [0,1)$ is defined as the \textit{forcing term} at
+  the $k$ iterate. Eq~(\ref{eq:inexact_newton_forcing}) then states
+  that the residual generated by the linear solver is bound by the
+  nonlinear residual and how tightly it is bound is defined by the
+  forcing term. This is useful in that we can vary how tightly coupled
+  the convergence of the linear iterations used to generate the Newton
+  correction is to the nonlinear iteration by relaxing or tightening
+  the convergence properties on the linear iterative method. As a
+  result, strategies for determining the forcing term can vary
+  depending on the problem type and can greatly affect the convergence
+  of the method or even prohibit convergence
+  \citep{eisenstat_choosing_1996}. In addition, globalization methods
+  may be used to modify the Newton correction in a more desireable
+  direction such that convergence properties can be improved when the
+  initial guess for $\ve{u}$ is poor
+  \citep{pawlowski_globalization_2006}.
 
@@
 A form of inexact Newton methods, \textit{Newton-Krylov methods} are
-nonlinear iterative methods that use a Krylov subspace method as the
-linear solver for generating the Newton correction. As we investigated
-in Chapter~\ref{chap:linear_problem}, Krylov methods are robust and
-enjoy efficient parallel implementations. Furthermore, their lack of
-explicit dependence on the operator make them easier to implement than
-other methods.
+nonlinear iterative methods that leverage a Krylov subspace method as
+the linear solver for generating the Newton correction
+\citep{kelley_iterative_1995}. As we investigated in
+Chapter~\ref{ch:linear_problem}, Krylov methods are robust and enjoy
+efficient parallel implementations on modern
+architectures. Furthermore, their lack of explicit dependence on the
+operator make them easier to implement than other
+methods. Additionally, although many iterations can become memory
+intensive due to the need to store the Krylov subspace for the
+orthogonalization procedure, at each nonlinear iteration this cost is
+reset as the Jacbobian matrix will change due to its dependence on the
+solution vector. This means that for every nonlinear iteration, a
+completely new linear system is formed for generating the Newton
+correction and we can modify the Krylov solver parameters accordingly
+to accomodate this. In most nonlinear problems, the Jacobian operator
+is generally nonsymmetric and therefore either Krylov methods with
+long recurrence relations that can handle nonsymmetric systems must be
+considered or Newton correction system must be preconditioned such
+that the operator is symmetric and short recurrence relations method
+can be potentially be used.
 
+With many Krylov methods available, which to use with the Newton
+method is dependent on many factors including convergence rates and
+memory usage. Several studies have been performed to investigate this
+\citep{mchugh_inexact_1993,knoll_newton-krylov_1995}. In their
+numerical studies in 1995, Knoll and McHugh used the set of highly
+nonlinear and stiff convection-diffusion-reaction equations to solve a
+set of tokamak plasma problems with the goal of measuring solver
+performance with Newton's method. They note several tradeoffs in using
+Krylov methods the a Newton solver. The first is that the optimization
+condition that results from the constraints (e.g. the minimization of
+the GMRES residual over the Krylov space) can be relaxed by
+restricting the size of the subspace such that only a fixed number of
+subspace vectors may be maintained, thus reducing memory
+requirements. We can also relax the optimization condition instead
+restarting the recurrence relation with a new set of vectors once a
+certain number of vectors have been generated, the optimization
+condition is maintain over that particular set of vectors, however,
+Knoll and McHugh note that this ultimately slows the convergence rate
+as compared to keeping all vectors as the new set of vectors is not
+necessarily orthogonal to the previous set, and therfore not optimal
+over the entire iteration procedure. The orthogonality condition can
+be relaxed by using a recurrence relation that does not generate a
+strictly orthonormal basis for the Krylov subspace such as the Lanzcos
+biorthogonalization procedure, resulting in memory savings due to the
+shorter Lanzcos recurrence relation.
+
+As a comparison, Knoll and McHugh chose an Arnoldi-based GMRES with a
+fixed vector basis approximately the size of the number of iterations
+required to converge as the long recurrence relation solver and
+conjugate gradients squared (CGS), bi-orthogonalized conjugate
+gradient stabilized (Bi-CGSTAB), and transpose-free quasiminimal
+residual (TFQMR) methdos as Lanzcos-based short recurrence relation
+solvers. All solvers were used to compute the right-preconditioned
+Newton correction system. For standard implementations of Newton's
+method where the Jacobian operator was explicitly formed using
+difference equations, all methods exhibited roughly equivalent
+iteration count performance for both the inner linear iterations and
+the outer nonlinear iterations in terms of iterations required to
+converge with Bi-CGSTAB typically performing the best for
+implementations where the Jacobian was explicitly formed and GMRES
+perfoming best for matrix-free implementations. However, upon
+investigating the convergence of the inner iterations, it was observed
+that the GMRES solver was significantly more robust, always generating
+a monotonically decreasing residual as compared to the Lanzcos-based
+methods which had the tendency to oscillate. Based on these results,
+in all of their future work Knoll and McHugh tended to use GMRES as
+the Krylov solver \citep{knoll_jacobian-free_2004}.
+
+\subsubsection{Jacobian-Free Approximation}
+\label{subsubsec:jacobian_free_approximation}
 In most cases, the Jacobian is difficult to form from the difference
-equations. For simple nonlinear cases such as the Navier Stokes
-equations, the derivatives can be computed and coded, but due to the
-complexity of those derivatives and the resulting difference equations
-this task can be tedious, error prone, and must be repeated for every
-equation set. By recognizing that Krylov methods only need the action
-of the operator on the vector instead of the operator itself, the
-Jacobian can instead be approximated through various numerical methods
-including a difference-based Jacobian-free formulation. Jacobian-Free
-methods, and in particular \textit{Jacobian-Free Newton-Krylov} (JFNK)
-methods \citep{knoll_jacobian-free_2004}, rely on forming the action
-of the Jacobian on a vector as required by the Krylov solver through a
+equations and costly to evaluate. For simple nonlinear cases such as
+the Navier-Stokes equations, the derivatives can be computed and
+coded, but due to the complexity of those derivatives and the
+resulting difference equations this task can be tedious, error prone,
+and must be repeated for every equation set. Further, in their 1995
+work, Knoll and McHugh also noted that a dominating part of their
+computation time was the evaluation of the difference equations for
+building the Jacobian \citep{knoll_newton-krylov_1995}. By recognizing
+that Krylov methods only need the action of the operator on the vector
+instead of the operator itself, the Jacobian can instead be
+approximated through various numerical methods including a
+difference-based Jacobian-free formulation. Jacobian-Free methods, and
+in particular \textit{Jacobian-Free Newton-Krylov} (JFNK) methods
+\citep{knoll_jacobian-free_2004}, rely on forming the action of the
+Jacobian on a vector as required by the Krylov solver through a
 forward difference scheme. In this case, the action of the Jacobian on
@@
 Eq~(\ref{eq:jacobian_free_product}) pollutes the solution.
+
+\subsubsection{Automatic Differentiation for Jacobian Generation}
+\label{subsubsec:automatic_differentiation}
 
@@
 
-\subsubsection{Automatic Differentiation}
-\label{subsubsec:automatic_differentiation}
-
 \subsection{Residual and Jacobian Generation}

### 2012-09-24 doc/prelim/document/chapters/parallel_mc/parallel_mc.tex
nonlinear updates

@@
 
+\section{Preliminaries}
+\label{sec:mc_preliminaries}
 We begin our discussion of stochastic methods using these texts by
@@
 \begin{equation}
-  |\ve{H} \circ \ve{W}| \geq |\ve{H}|\ |\ve{W}|\:,
+  |\ve{H} \circ \ve{W}| \geq |\ve{H}|\ |\ve{W}|\:.
   \label{eq:hadamard_inequality}
 \end{equation}
-meaning that $\rho(\ve{G}) \geq \rho(\ve{H})$. Using these relations
-to analyze Eq~(\ref{eq:direct_variance_5}), we see that if
-$\rho(\ve{H}) > 1$, then the sum in
+Using the norm property of the Hadamard product and
+Eq~(\ref{eq:neumann_ulam_decomposition}), we can define the norm of
+$\ve{W}$ as:
+\begin{equation}
+  \frac{|\ve{H}|}{|\ve{P}|} \geq |\ve{W}|\:.
+  \label{eq:weight_norm}
+\end{equation}
+We then have the following inequality for relating $\ve{G}$ and
+$\ve{H}$:
+\begin{equation}
+  |\ve{G}| \geq \frac{|\ve{H}|^2}{|\ve{P}|}
+  \label{eq:variance_norm_inequality}
+\end{equation}
+Using these relations to analyze Eq~(\ref{eq:direct_variance_5}), we
+see that if $\rho(\ve{H}) > 1$, then the sum in
 Eq~(\ref{eq:variance_neumann_series}) will not converge and an
 infinite variance will arise as the elements of $\ve{T}$ become
-infinite. Therefore, we must restrict $\ve{H}$ with $\rho(\ve{H}) < 1$
-so that our expectation values for the solution may have a finite
-variance.
+infinite in Eq~(\ref{eq:direct_variance_5}). Therefore, we must
+restrict $\ve{H}$ with $\rho(\ve{H}) < 1$ so that our expectation
+values for the solution may have a finite variance.
 
@@
 \begin{equation}
-  \ve{H}^{\dagger} = \ve{P} \circ \ve{W}\:,
+  \ve{H}^{T} = \ve{P} \circ \ve{W}\:,
   \label{eq:adjoint_neumann_ulam}
@@
 where now we are forming the decomposition with respect to the
-conjugate transpose of $\ve{H}$\footnote{This is sometimes referred to
-  as the adjoint form of the matrix}. We then follow the same
+transpose of $\ve{H}$\footnote{This is sometimes referred to as the
+  adjoint form for real valued matrices}. We then follow the same
 procedure as the direct method for forming the probability and weight

### 2012-09-24 doc/prelim/document/chapters/nonlinear_problem/nonlinear_problem.tex
nonlinear updates

@@
 describe fluid flow or more general transport processes when
-discretized by conventional methods yield discrete sets of equations
-with nonlinearities present in the variables. Traditionally, such
-systems have been solved by linearizing them such that the
+discretized by conventional methods yield discrete sets of stiff
+equations with nonlinearities present in the variables. Traditionally,
+such systems have been solved by linearizing them such that the
 nonlinearities in the variable are eliminated and more traditional
@@
 implementation as compared to Krylov methods. In addition to handling
-this dependency, we will review parallel strategies for a nonlinear
+this dependency, we will construct a parallel strategy for a nonlinear
 MCSA method.
@@
 If we ignore the nonlinear terms in the expansion and assert that at
-the $k+1$ iterate $\ve{u}^{k+1}$ is that exact solution such that
+the $k+1$ iterate $\ve{u}^{k+1}$ is the exact solution such that
 $\ve{F}(\ve{u}^{k+1}) = \ve{0}$, then we are left with the following
@@
 long recurrence relations that can handle nonsymmetric systems must be
-considered or Newton correction system must be preconditioned such
-that the operator is symmetric and short recurrence relations method
+considered or the Newton correction system must be preconditioned such
+that the operator is symmetric and short recurrence relation methods
 can be potentially be used.
@@
 performance with Newton's method. They note several tradeoffs in using
-Krylov methods the a Newton solver. The first is that the optimization
-condition that results from the constraints (e.g. the minimization of
-the GMRES residual over the Krylov space) can be relaxed by
-restricting the size of the subspace such that only a fixed number of
-subspace vectors may be maintained, thus reducing memory
-requirements. We can also relax the optimization condition instead
+Krylov methods with the Newton solver. The first is that the
+optimization condition that results from the constraints (e.g. the
+minimization of the GMRES residual over the Krylov space) can be
+relaxed by restricting the size of the subspace such that only a fixed
+number of subspace vectors may be maintained, thus reducing memory
+requirements. We can also relax the optimization condition by instead
 restarting the recurrence relation with a new set of vectors once a
@@
 the outer nonlinear iterations in terms of iterations required to
-converge with Bi-CGSTAB typically performing the best for
+converge. Bi-CGSTAB typically performed the best for
 implementations where the Jacobian was explicitly formed and GMRES
@@
 In most cases, the Jacobian is difficult to form from the difference
-equations and costly to evaluate. For simple nonlinear cases such as
-the Navier-Stokes equations, the derivatives can be computed and
-coded, but due to the complexity of those derivatives and the
-resulting difference equations this task can be tedious, error prone,
-and must be repeated for every equation set. Further, in their 1995
-work, Knoll and McHugh also noted that a dominating part of their
-computation time was the evaluation of the difference equations for
-building the Jacobian \citep{knoll_newton-krylov_1995}. By recognizing
-that Krylov methods only need the action of the operator on the vector
-instead of the operator itself, the Jacobian can instead be
-approximated through various numerical methods including a
-difference-based Jacobian-free formulation. Jacobian-Free methods, and
-in particular \textit{Jacobian-Free Newton-Krylov} (JFNK) methods
+equations and costly to evaluate for large equation sets. For simple
+nonlinear cases such as the Navier-Stokes equations, the derivatives
+can be computed and coded, but due to the complexity of those
+derivatives and the resulting difference equations this task can be
+tedious, error prone, and must be repeated for every equation
+set. Further, in their 1995 work, Knoll and McHugh also noted that a
+dominating part of their computation time was the evaluation of the
+difference equations for building the Jacobian
+\citep{knoll_newton-krylov_1995}. By recognizing that Krylov methods
+only need the action of the operator on the vector instead of the
+operator itself, the Jacobian can instead be approximated through
+various numerical methods including a difference-based Jacobian-free
+formulation. Jacobian-Free methods, and in particular
+\textit{Jacobian-Free Newton-Krylov} (JFNK) methods
 \citep{knoll_jacobian-free_2004}, rely on forming the action of the
@@
 $\epsilon$, then the finite difference error from
-Eq~(\ref{eq:jacobian_free_product}) pollutes the solution.
+Eq~(\ref{eq:jacobian_free_product}) pollutes the
+solution. Furthermore, Knoll and McHugh noted that for preconditioning
+purposes, part of the Jacobian must still explicitly be formed
+periodically and that linear solver robustness issues were magnified
+by the matrix-free approach due to the first-order approximation.
 
@@
 \label{subsubsec:automatic_differentiation}
+If it is acceptable to store the actual Jacobian matrix, other methods
+are available to construct it without requiring hand-coding and
+evaluating derivatives, thus eliminating the associated issues. In
+addition, if any additional equations are added to the system or a
+higher order functional approximation is desired, these derivatives
+must be regenerated and coded. Becoming more prominent in the 1990's,
+\textit{automatic differentation} is a mechanism by which the
+derivatives of a function can be generated automatically by evaluating
+it. Automatic differntiation is built on the concept that all
+functions discretely represented in a computer are ultimately
+represented by elementary mathematical operations. If the chain rule
+is applied to those elementary operations, then the derivative of that
+function can be computed to the order of accuracy of its original
+discretization in a completely automated way.
+
+In their 1994 work, Averick and others compare the qualities of sparse
+Jacobian approximation by differencing as in
+Eq~(\ref{eq:jacobian_free_product}), hand-coded derivatives, and
+automatic differentiation \citep{averick_computing_1994}. If code to
+perform the Jacobian matrix-vector multiply or an equivalent
+approximation as in Eq~(\ref{eq:jacobian_free_product}), then the
+Jacobian matrix can be explicitly formed by applying the code to the
+column vectors of the identity matrix at the cost of $N$ function
+evaluations for $\ve{J}(\ve{u}) \in \mathbb{R}^{N \times N}$.
 

### 2012-09-26 doc/prelim/document/chapters/introduction/introduction.tex
nonlinear updates

@@
 solutions to linear problems \citep{saad_iterative_2003,
-  kelley_iterative_1987}. In between these two areas are a not widely
+  kelley_iterative_1995}. In between these two areas are a not widely
 known small group of stochastic methods for solving sparse linear
@@
 various Newton methods have been popular
-\citep{kelley_iterative_1987}. Recently, Jacobian-Free Newton-Krylov
+\citep{kelley_iterative_1995}. Recently, Jacobian-Free Newton-Krylov
 (JFNK) schemes \citep{knoll_jacobian-free_2004} have been utilized in

### 2012-09-26 doc/prelim/document/chapters/nonlinear_problem/nonlinear_problem.tex
nonlinear updates

@@
 tedious, error prone, and must be repeated for every equation
-set. Further, in their 1995 work, Knoll and McHugh also noted that a
-dominating part of their computation time was the evaluation of the
+set. Furthermore, in their 1995 work, Knoll and McHugh also noted that
+a dominating part of their computation time was the evaluation of the
 difference equations for building the Jacobian
@@
 $\epsilon$, then the finite difference error from
-Eq~(\ref{eq:jacobian_free_product}) pollutes the
-solution. Furthermore, Knoll and McHugh noted that for preconditioning
-purposes, part of the Jacobian must still explicitly be formed
-periodically and that linear solver robustness issues were magnified
-by the matrix-free approach due to the first-order approximation.
+Eq~(\ref{eq:jacobian_free_product}) pollutes the solution. In
+addition, Knoll and McHugh noted that for preconditioning purposes,
+part of the Jacobian must still explicitly be formed periodically and
+that linear solver robustness issues were magnified by the matrix-free
+approach due to the first-order approximation. This formation
+frequency coupled with the numerous evaluations of the Jacobian
+approximation create a situation where after so many nonlinear
+iterations, it becomes cheaper to instead fully form the
+Jacobian. For simple equation sets, this may only take 5-10 Newton
+iterations to reach this point while over 30 may be required for
+larger equations sets and therefore larger Jacobians.
 
@@
 addition, if any additional equations are added to the system or a
-higher order functional approximation is desired, these derivatives
-must be regenerated and coded. Becoming more prominent in the 1990's,
-\textit{automatic differentation} is a mechanism by which the
-derivatives of a function can be generated automatically by evaluating
-it. Automatic differntiation is built on the concept that all
-functions discretely represented in a computer are ultimately
-represented by elementary mathematical operations. If the chain rule
-is applied to those elementary operations, then the derivative of that
-function can be computed to the order of accuracy of its original
-discretization in a completely automated way.
+higher order functional approximation is desired, it would be useful
+to avoid regenerating and coding these derivatives. Becoming more
+prominent in the 1990's, \textit{automatic differentiation} is a
+mechanism by which the derivatives of a function can be generated
+automatically by evaluating it. Automatic differntiation is built on
+the concept that all functions discretely represented in a computer
+are ultimately represented by elementary mathematical operations. If
+the chain rule is applied to those elementary operations, then the
+derivatives of those functions can be computed to the order of
+accuracy of their original discretization in a completely automated
+way \citep{averick_computing_1994}.
 
-In their 1994 work, Averick and others compare the qualities of sparse
-Jacobian approximation by differencing as in
-Eq~(\ref{eq:jacobian_free_product}), hand-coded derivatives, and
-automatic differentiation \citep{averick_computing_1994}. If code to
-perform the Jacobian matrix-vector multiply or an equivalent
-approximation as in Eq~(\ref{eq:jacobian_free_product}), then the
-Jacobian matrix can be explicitly formed by applying the code to the
-column vectors of the identity matrix at the cost of $N$ function
-evaluations for $\ve{J}(\ve{u}) \in \mathbb{R}^{N \times N}$.
+The work of Bartlett and others \citep{hutchison_automatic_2006}
+extended initial Fortran-based work in the area of automatic
+differentiation implementations to leverage the parametric type and
+operator overloading features of C++ \citep{stroustrup_c++_1997}. They
+formulate the differentiation problem from an element viewpoint by
+assuming that a global Jacobian can be assembled from local element
+function evaluations of $e_k : \mathbb{R}^{n_k} \rightarrow
+\mathbb{R}^{m_k}$, similar to the finite element assembly procedure
+as:
+\begin{equation}
+  \ve{J}(\ve{u}) = \sum_{i=1}^N \ve{Q}^T_i \ve{J}_k \ve{P}_i\:,
+  \label{eq:fad_global_jacobian}
+\end{equation}
+where $\ve{J}_{k_i} = \partial e_{k_i} / \partial P_i u$ is the
+$k^{th}$ element function Jacobian, $\ve{Q} \in \mathbb{R}^{n_{k_i}
+  \times N}$ is a projector onto the element domain and $\ve{P} \in
+\mathbb{R}^{m_{k_i} \times N}$ a projector onto the element range for
+$\ve{F}(\ve{u}) \in \mathbb{R}^{N \times N}$. The Jacobian matrix for
+each element will therefore have entirely local data in a dense
+structure, eliminating the need for parallel communication and sparse
+techniques during differentiation. Only when all local differentials
+are computed does communication of the Jacobian occur through
+gather/scatter operations in order to properly assembly it. Also of
+benefit is the fact that element-level computations generally consist
+of a smaller number of degrees of freedom, thus reducing memory
+requirements during evaluation as compared to a global formulation of
+the problem. Such a formulation is not limited to finite element
+formulations and is amenable to any scheme where the system is
+globally sparse with degrees of freedom coupled to local domains
+including finite volume representations. The templating capabilities
+of C++ were leveraged with the element-based evaluation and assembly
+scheme as in Eq~(\ref{eq:fad_global_jacobian}) by templating element
+function evaluation code on the evaluation type. If these functions
+are instantiated with standard floating point types then the residual
+is returned. If they are instead instantiated with the
+operator-overloaded automatic differentiation types, both the residual
+and Jacobian are returned.
+
+Of interest to Bartlett, Averick, and the many others that have
+researched automatic differentation are measuring its performance
+relative to hand-coded derivatives and capturing the Jacboian matrix
+from matrix-free approximations. Given their element-based function
+evaluation scheme, Bartlett's work varied the number of degrees of
+freedom per element and compared both the floating point operation
+count and CPU time for both the templated automatic differentiation
+method and hand-coded derivatives for Jacobian evaluations. Although
+they observed a 50\% increase in floating point operations in the
+templated method over the hand-coded method, run times were observed
+to be over 3 times faster for the templated method. They hypothesize
+that this is due to the fact that the element-based formulation of the
+templated method is causing better utilization of cache and therefore
+faster data access. Furthermore, they observed linear scaling behavior
+for automatic differentiation as the number of degrees of freedom per
+element were increased up to a few hundred. Based on these results,
+this type of automatic differentation formulation was deemed
+acceptable for use in large-scale production physics codes.
 
@@
 \label{sec:fanm}
-In this section we will devise the FANM method. Based on the
-background in the previous section, we first need to motivate the
-development of this work by pointing out some of the issues with
-conventional methods. This includes JFNK coarseness and either
-prohibitively large Krylov subspaces or a loss of orthogonality
-information in the subspace due to restarts. We can than point out
-some of the attractive qualities of the FANM method. Much of this will
-rely on advanced concepts including automatic differentiation and
-on-the-fly residual and Jacobian generation from nonlinear function
-evaluations. In addition, we will want to show that the Newton method
-will converge using MCSA as the linear solver. I suspect I will have
-some preliminary results for a serial implementation here as I already
-have a serial MCSA implemented. Finally, we will want to design the
-model problems that we will test FANM with. These should be problems
-that have already been worked up in the literature with benchmark-type
-solutions. Parallelism in the context of the linear solvers chapter
-should also be discussed.
+In production physics codes based on nonlinear equations sets,
+Newton-Krylov methods are the primary means of generating a fully
+consistent solution scheme
+\citep{evans_development_2006,evans_enhanced_2007,gaston_parallel_2009,godoy_parallel_2012}. Typically,
+for large scale simulations these problems are memory limited due to
+the subspaces generated by robust Krylov methods. Often a matrix-free
+approach is chosen to relax memory requirements over directly
+generating the Jacobian matrix. However, as we observed in previous
+sections, these matrix-free methods suffer from poorly scaled problems
+and the first order error introduced by the Jacobian approximation. In
+addition, it was observed that eventually, the savings induced by the
+matrix-free approach is eventually amortized over a number of
+nonlinear iterations where it becomes more efficient computationally
+to instead form the Jacobian.
+
+In Chapter~\ref{ch:stochastic_methods}, we focused our efforts on
+developing and improving Monte Carlo methods for inverting linear
+systems. These methods, when used to accelerate a stationary method in
+MCSA, enjoy a robust implementation and exponential convergence
+rates. Further, the only storage required is that of the full linear
+system including the linear operator so that we may generate
+transition probabilities for the random walk sequence. Although this
+requires more storage to represent the linear system than that of a
+Krylov method where the operator is not required, we do not incur any
+additional storage costs once the iteration sequence
+begins. Furthermore, in the context of nonlinear problems, the
+Jacobian matrix that we are required to generate for the Monte Carlo
+solvers may be generated at will from the nonlinear functions in the
+Newton system using automatic differentation. Not only do we then have
+a simple and automated way to generate the Jacobian, but we also enjoy
+a Jacobian of numerical precision equivalent to that of our function
+evaluations. We therefore propose the \textit{Forward-Automated
+  Newton-MCSA} (FANM) method that utilizes all of the above
+components. We hypothesize that such a method will be competetive with
+Newton-Krylov methods not only from a convergence and timing
+perspective, but also relax scaling requirements of matrix-free
+methods and memory costs of both matrix-free and fully formed Jacobian
+methods to allow the application developer to solve problems of finer
+discretization and higher order functional representations while
+maintaining an robust an efficient parallel implementation.
 
@@
 \label{subsec:fanm_storage}
+To gauge the memory benefits of a FANM method over a Krylov-based
+scheme, we must look at the storage requirements of the Jacobian
+matrix as compared to the Krylov subspace vectors for sparse linear
+problems. Saad's text provides us relations for both of these cases
+\citep{saad_iterative_2003}. Beginning with the Jacobian storage, for
+sparse problems production linear algebra library implementations
+utilize \textit{compressed row storage} to efficiently store the
+non-zero components of a sparse matrix while still allowing for all
+standard parallel matrix computations to be performed. Per
+\S~\ref{sec:parallel_krylov_methods}, recall that efficient parallel
+matrix-vector multiplications rely on processors knowing which other
+processors contain their neighboring matrix and vector
+elements. Typically this information is represented by a graph which
+must be stored along with the matrix elements themselves. To store the
+elements, consider the following sparse matrix:
+\[
+  \ve{A} =
+  \begin{bmatrix}
+    2 & 0 & 8 & 0 & 0 & 0 \\ 
+    4 & 5 & 0 & 1 & 0 & 0 \\ 
+    0 & 2 & 1 & 0 & 1 & 0 \\ 
+    0 & 0 & 3 & 7 & 0 & 2 \\ 
+    0 & 0 & 0 & 4 & 9 & 0 \\
+    0 & 0 & 0 & 0 & 9 & 1 \\
+  \end{bmatrix}\:.
+\]
+The compressed form of \ve{A} would then be:
+\[
+  \begin{array}{lccccccccccccccc}
+    \ve{A}\ values: & 2 & 8 & 4 & 5 & 1 & 2 & 1 & 1 & 3 & 7 & 2 & 4 & 9
+    & 9 & 1 \\ column: & 1 & 3 & 1 & 2 & 4 & 2 & 3 & 5 & 3 & 4 & 6 & 4
+    & 5 & 5 & 6 \\ row\ start: & 3 & 5 & 9 & 12 & 14 & 16 \\
+  \end{array}
+\]
+where the first row contains the non-zero values of the matrix
+$\ve{A}$ starts at, the second row contains the column index of those
+values, and the third row contains the index of the A value and column
+index rows that each matrix row starts at. For this simple example, we
+actually break even by storing 36 elements in the full matrix and 36
+pieces of data in the compressed format. However, for large, sparse
+matrices there is a significant storage savings using the compressed
+format.
 

### 2012-09-26 doc/prelim/document/chapters/parallel_mc/parallel_mc.tex
nonlinear updates

@@
 \chapter{Stochastic Solution Methods for Linear Problems}
-\label{chap:stochastic_methods}
+\label{ch:stochastic_methods}
 An alternative approach to approximate matrix inversion is to employ

### 2012-09-26 doc/prelim/document/chapters/nonlinear_problem/nonlinear_problem.tex
finished draft on nonlinear section

@@
 
-\subsection{Residual and Jacobian Generation}
-\label{subsec:fanm_generation}
-
 \subsection{Jacobian Storage vs. Subspace Storage and Restarts}
@@
     & 9 & 1 \\ column: & 1 & 3 & 1 & 2 & 4 & 2 & 3 & 5 & 3 & 4 & 6 & 4
-    & 5 & 5 & 6 \\ row\ start: & 3 & 5 & 9 & 12 & 14 & 16 \\
+    & 5 & 5 & 6 \\ row\ start: & 1 & 3 & 6 & 9 & 12 & 14 & 16 \\
   \end{array}
@@
 values, and the third row contains the index of the A value and column
-index rows that each matrix row starts at. For this simple example, we
-actually break even by storing 36 elements in the full matrix and 36
-pieces of data in the compressed format. However, for large, sparse
-matrices there is a significant storage savings using the compressed
-format.
+index rows that each matrix row starts at ending with the start of the
+next row. For this simple example, we actually break even by storing
+36 elements in the full matrix and 36 pieces of data in the compressed
+format. However, for large, sparse matrices there is a significant
+storage savings using the compressed format.
+
+In his discussion on orthogonalization schemes for Krylov subspaces,
+he provides us with space and time complexities for these
+operations. To find a solution vector $\ve{x} \in \mathbb{R}^{N \times
+  N}$ with $m$ Krylov iterations (and therefore $m+1$ subspace
+vectors) the storage requirement is then $(m+1) N$ for most common
+orthogonalization techniques. Using our simple example above (even
+though we did not see any gains in storage over the full matrix
+representation), if it takes 10 GMRES iterations using Arnoldi
+orthogonalization to converge to a solution, then a total of 66 pieces
+of data are required to be stored instead of 36 for the explicitly
+formed matrix case. For larger sparse matrices, this disparity in
+memory requirements will be even greater, especially for
+ill-conditioned systems that require many GMRES iterations to
+converge.
 
@@
 \label{subsec:parallel_fanm}
+Like the parallelization of the MCSA method described in
+\S~\ref{sec:parallel_stochastic_methods}, a parallel FANM method
+relies on a basic set of parallel matrix-vector operations as well as
+the global residual and Jacobian assembly procedure described in
+\S~\ref{subsubsec:automatic_differentiation}. Consider the Newton
+iteration scheme in Eq~(\ref{eq:newton_iteration}). We must first
+assemble the linear system in parallel through the element-wise
+function evaluations to generate both the global Jacobian operator and
+the global residual vector on the right hand side. Per Bartlett's
+work, efficient and automated parallel mechanisms are available to do
+this through a sequence of scatter/gather operations. With these tools
+available for residual and Jacobian generation, the remainder of the
+parallel procedure is simple, with the linear Newton correction system
+solved using the parallel MCSA method as previously described and the
+Newton correction applied to the previous iterate's solution through a
+vector update operation.
+
+As Newton methods are formulated independent of the inner linear
+solver generating the Newton corrections, the actual performance of
+the nonlinear iterations using MCSA is expected to be similar to that
+of traditional Newton-Krylov methods. Furthermore, we expect to
+achieve numerically identical answers with a Newton-MCSA method as
+other Newton methods and we should indeed verify this. The parallel
+performance of such a method will inherently be bound to the parallel
+Monte Carlo implementation of the linear solver as the parallel
+operations at the nonlinear iteration level are identical to those
+that you would perform with a Newton-Krylov method. Matrix-free
+formulation will have different parallel performance than these
+methods, and therefore we should compare FANM perfomance to JFNK-based
+schemes. More important than performance in the situation is the
+reduced memory pressure that a FANM implementation provides, as
+discussed in \S~\ref{subsec:fanm_storage}, because a FANM method will
+not generate a subspace in the linear solver and compressed storage
+for sparse matrices are utilized, we expect significant memory savings
+over Newton-Krylov methods. We must measure the memory utilization of
+both of these methods in order to quantify their differences and
+provide additional analysis of the FANM method's merits, or lack
+thereof.

### 2012-09-27 doc/prelim/document/chapters/nonlinear_problem/nonlinear_problem.tex
starting work on the research proposal

@@
 such systems have been solved by linearizing them such that the
-nonlinearities in the variable are eliminated and more traditional
+nonlinearities in the variables are eliminated and more traditional
 linear methods can be used for solutions. Often characterized as
@@
 this dependency, we will construct a parallel strategy for a nonlinear
-MCSA method.
+MCSA-based method.
 
@@
   of the method or even prohibit convergence
-  \citep{eisenstat_choosing_1996}. In addition, globalization methods
-  may be used to modify the Newton correction in a more desireable
-  direction such that convergence properties can be improved when the
-  initial guess for $\ve{u}$ is poor
+  \citep{eisenstat_choosing_1996}. In addition, \textit{globalization
+    methods} may be used to modify the Newton correction in a more
+  desireable direction such that convergence properties can be
+  improved when the initial guess for $\ve{u}$ is poor
   \citep{pawlowski_globalization_2006}.
@@
 restarting the recurrence relation with a new set of vectors once a
-certain number of vectors have been generated, the optimization
+certain number of vectors have been generated. The optimization
 condition is maintain over that particular set of vectors, however,
@@
 mechanism by which the derivatives of a function can be generated
-automatically by evaluating it. Automatic differntiation is built on
+automatically by evaluating it. Automatic differentiation is built on
 the concept that all functions discretely represented in a computer
@@
 Of interest to Bartlett, Averick, and the many others that have
-researched automatic differentation are measuring its performance
+researched automatic differentation are measures of its performance
 relative to hand-coded derivatives and capturing the Jacboian matrix
@@
 this type of automatic differentation formulation was deemed
-acceptable for use in large-scale production physics codes.
+acceptable for use in large-scale, production physics codes.
 
@@
 for large scale simulations these problems are memory limited due to
-the subspaces generated by robust Krylov methods. Often a matrix-free
+the subspaces generated by robust Krylov methods. Often, a matrix-free
 approach is chosen to relax memory requirements over directly
-generating the Jacobian matrix. However, as we observed in previous
-sections, these matrix-free methods suffer from poorly scaled problems
-and the first order error introduced by the Jacobian approximation. In
-addition, it was observed that eventually, the savings induced by the
-matrix-free approach is eventually amortized over a number of
-nonlinear iterations where it becomes more efficient computationally
-to instead form the Jacobian.
+generating the Jacobian matrix and facilitate the
+implementation. However, as we observed in previous sections, these
+matrix-free methods suffer from poorly scaled problems and the first
+order error introduced by the Jacobian approximation. In addition, it
+was observed that the savings induced by the matrix-free approach is
+eventually amortized over a number of nonlinear iterations where it
+becomes more efficient computationally to instead form the Jacobian.
 
@@
 methods to allow the application developer to solve problems of finer
-discretization and higher order functional representations while
-maintaining an robust an efficient parallel implementation.
+discretization and higher-order functional representations while
+maintaining a robust and efficient parallel implementation.
 
@@
 In his discussion on orthogonalization schemes for Krylov subspaces,
-he provides us with space and time complexities for these
+Saad provides us with space and time complexities for these
 operations. To find a solution vector $\ve{x} \in \mathbb{R}^{N \times
@@
 of data are required to be stored instead of 36 for the explicitly
-formed matrix case. For larger sparse matrices, this disparity in
-memory requirements will be even greater, especially for
-ill-conditioned systems that require many GMRES iterations to
-converge.
+formed matrix case (we do not consider graph storage here for
+comparison as both methods will require the graph for parallel
+operations). For larger sparse matrices, this disparity in memory
+requirements will be even greater, especially for ill-conditioned
+systems that require many GMRES iterations to converge.
 
@@
 Newton correction applied to the previous iterate's solution through a
-vector update operation.
+parallel vector update operation.
 
@@
 that you would perform with a Newton-Krylov method. Matrix-free
-formulation will have different parallel performance than these
+formulations will have different parallel performance than these
 methods, and therefore we should compare FANM perfomance to JFNK-based

### 2012-09-27 doc/prelim/document/chapters/research_proposal/research_proposal.tex
starting work on the research proposal

@@
 \label{ch:research_proposal}
-
-After all of the background information and explanations of the new
-work, this section is the most important of all. Here I will clearly
-define the questions I aim to answer and the strategy that I will use
-to attempt to answer them. I think that it will be important to pick a
-challenge problem here that FANM has a chance to show gains in either
-memory footprint or parallelism over current methods. In particular, I
-would like to have this challenge problem derive from nuclear reactor
-multiphysics. I think Drekar would be a great test bed for FANM once
-complete as it has all the tools to build a FANM solve (it is in fact
-based on the same core libraries as Chimera) and a complex reactor
-core assembly multiphysics system capable of scaling to large
-machines. In addition, all of the complex diagnostic tools necessary
-for analysis of FANM performance as compared to the other methods
-should be available.
+To develop the parallel MCSA and FANM methods, a research plan is
+proposed to drive their development and quantify their merits and
+feasiblity as compared to more conventional methods for linear and
+nonlinear problems. We seek to rigorously verify both of these methods
+and there implementation and will describe the means by which that
+will be accomplished. Further more, the numerical experiments
+performed will be constructed within a software framework that will
+permit an agile development strategy.
 

### 2012-09-27 doc/prelim/document/chapters/research_proposal/research_proposal.tex
adding jcomp work to proposal

@@
 and there implementation and will describe the means by which that
-will be accomplished. Further more, the numerical experiments
-performed will be constructed within a software framework that will
-permit an agile development strategy.
+will be accomplished. Furthermore, the numerical experiments performed
+will be constructed within a software framework that will permit an
+agile development strategy. In this chapter we describe that
+experimental framework and the progress achieved to date by its
+use. In addition, we will a formulate a plan for verifying the new
+stochastic methods and prepare a set of numerical experiments that
+will determine some of their properties. Finally, we select a
+challenge problem designed to potentially demsonstrate the
+effectiveness of our new methods in the context of real-world,
+production level applications.
 
@@
 \label{sec:experimental_framework}
+For all of our numerical experiments, we need to readily be able to do
+two things: generate linear and nonlinear systems in parallel. To do
+this effectively requires many software components to be implemented
+including linear algebra data structures, a parallel communication
+framework, mesh generation and partitioning, application of boundary
+and initial conditions, automatic differentiation technology, and
+other high-level framework technologies. These development of these
+components, however, are not the focus of this work but rather the
+necessary machinery needed for the experimental apparatus. We
+therefore choose an environment in which all of these tools are
+readily available in the form of production software implementations
+and will demonstrate scalability in high performance computing
+environments. This environment will be based on the mathematical
+framework of the finite element method and will permit an
+algorithm-oriented design for the parallel stochastic solvers we will
+develop.
 
-The Chimera code base will be the primary research tool for answering
-the questions posed in the previous chapters of the document. We may
-not need a chapter for this, but it will likely be a good idea
-somewhere in this document to outline the design of the software and
-the instrumentation that it will provide.
+\subsection{Finite Element Formulation}
+\label{subsec:fem_assembly}
+Choosing a finite element formulation for our physics problems permits
+us flexibility from both a mathematical and software development
+standpoint. We briefly review the \textit{finite element method} as
+presented in Zienkiewicz's text \citep{zienkiewicz_finite_2005}. In
+general, we are concerned with solving a set of differential
+equations:
+\begin{subequations}
+  \begin{gather}
+    \ve{A}(\ve{u}) = \ve{0},\ \forall \ve{u} \in \Omega \\
+    \ve{B}(\ve{u}) = \ve{0},\ \forall \ve{u} \in \Gamma\:,
+  \end{gather}
+  \label{eq:finite_element_de}
+\end{subequations}
+where $\Omega$ is the domain over which the set of differential
+equations is defined and $\Gamma$ is the boundary of that domain. We
+then have a set of differential equations, $\ve{A}(\ve{u})$, in
+residual form that represent the physics on the domain and the set
+$\ve{B}(\ve{u})$ that is representative of those physics on the
+boundary. As an alternative, we can instead state these differential
+equations in an equivalent integral form via a conservation statement:
+\begin{equation}
+  \int_{\Omega} \ve{v}^T \ve{A}(\ve{u}) d\Omega + \int_{\Gamma}
+  \bar{\ve{v}}^T \ve{B}(\ve{u}) d\Gamma = \ve{0} \:,
+  \label{eq:fe_integral_form}
+\end{equation}
+where $\ve{v}$ and $\bar{\ve{v}}$ are arbitrary, finite-valued vectors
+that satisfy the integral relation and the constraints in
+Eq~(\ref{eq:finite_element_de}).
 
-\subsection{Algorithm-Based Design}
-\label{subsec:chimera_design}
+If the $n^{th}$-order derivative of $\ve{u}$ contains discontinuities
+(i.e. is not smooth), then the integral form presented in
+Eq~(\ref{eq:fe_integral_form}) is restricted in that only derivatives
+up to order $n+1$ may appear in the functions $\ve{A}(\ve{u})$ and
+$\ve{B}(\ve{u})$. We can relax this restriction by integrating
+Eq~(\ref{eq:fe_integral_form}) by parts to arrive at the \textit{weak
+  form} of the differential equations:
+\begin{equation}
+  \int_{\Omega} \ve{C}(\ve{v}^T) \ve{D}(\ve{u}) d\Omega +
+  \int_{\Gamma} \ve{E}(\bar{\ve{v}}^T) \ve{F}(\ve{u}) d\Gamma = \ve{0}
+  \:,
+  \label{eq:fe_weak_form}
+\end{equation}
+where the new functional form now has lower-order derivatives in
+$\ve{D}(\ve{u})$ and $\ve{F}(\ve{u})$, thus relaxing the smoothness
+requirement.
 
-\subsection{Graph-Based Multiphysics Formulation}
-\label{subsec:multiphysics_graph}
-
-\subsection{Finite Element Assembly}
-\label{subsec:fem_assembly}
-
-\subsection{Software Engineering}
-\label{subsec:software_engineering}
+In our experimental framework, we will formulate all of our problems
+using the weak form of the differential equations. Doing so allows us
+to leverage modern finite element assembly engine software with
+embedded automatic differentiation technology. Using the Panzer
+library in the Trilinos sceintific computing framework
+\citep{heroux_overview_2005}, function evaluation routines in the form
+of Eq~(\ref{eq:fe_weak_form}) are generated to represent the finite
+element problem and boundary conditions. Per Bartlett's automatic
+differentiation work and an associated parallel linear algebra
+framework, the equation sets for all physics implemented are assembled
+into the nonlinear system. By leveraging the general linear algebra
+framework, we are able to abstract away those concepts in the
+stochastic solvers implementation and instead focus on an
+algorithm-oriented approach for the solver implementation
+\citep{musser_algorithm-oriented_1994}.
 
@@
 \label{sec:progress}
+To date, progress has been made on development of the experimental
+framework based on the Panzer library. Using this framework, several
+preliminary objectives have been achieved including an implementation
+of the MCSA algorithm using a general linear algebra framework and
+initial analysis of its performance using both the direct and adjoint
+Neumann-Ulam methods. Here we report the results of this work and its
+relationship with future work.
 
@@
 \label{subsec:mcsa_generalization}
+Previous implementations of MCSA had an entirely physics-based
+approach \citep{evans_monte_2009,evans_monte_2012}. In these works,
+the transition probabilities and weights were generated directly from
+the discretized differential equations instead of a general linear
+operator. With this type of approach, transition probabilities and
+weight formulations must be rederived for each type of physics
+implemented. To reach the broader physics community and facilitate
+deployment for more complicated physical systems and multiple physics
+problems, MCSA would benefit from an implementation that utilizes the
+general operator form of a linear system as in
+Eq~(\ref{eq:linear_problem}). Operating on the general form of the
+linear system will also greatly simplify the implementation of our
+numerical experiments as different equations sets and boundary
+conditions can be rapidly implemented within the Panzer library to
+generate a complex array of linear problems.
+
+Generalizing MCSA for all linear problems has been achieved with both
+the direct and adjoint Neumann-Ulam methods as the Monte Carlo solver
+by leveraging the Epetra library within Trilinos
+\citep{heroux_overview_2005}. Using Epetra (and its variants) we have
+access to several features that will enhance future development and
+experimentation. First, Epetra employs a compressed storage format for
+sparse matrices as outlined in \S~\ref{subsec:fanm_storage}. This will
+permit efficient memory use and permit the savings we expect by
+implementing the FANM method. Second, Epetra is fully parallelized in
+that the parallel graph structures and decomposed storage formats are
+utilized to domain decompose both the operator and vector in the
+linear system. These data structures provide access to the required
+parallel matrix-vector operations outlined in
+\S~\ref{sec:parallel_krylov_methods}. In addition, the parallel Monte
+Carlo methods we will implement for the Neumann-Ulam methods as
+outlined in \S~\ref{sec:parallel_stochastic_methods} will be greatly
+facilitated and perhaps even enabled by these data
+structures. Finally, the Epetra framework and its variants are
+commonplace in many production level physics codes and therefore such
+an implementation will provide an easy path to incorporating an MCSA
+or FANM solver into those codes for further experimentation and
+analysis.
 
@@
 \label{subsec:mcsa_direct_vs_adjoint}
+Preliminary work has been completed to characterize the behavior of
+the direct and adjoint Monte Carlo solvers with the MCSA
+method\footnote{The contents of this section have been taken directly
+  from my contributions to \citep{evans_monte_2012}}. The aim of this
+initial work was to determine which of the two Monte Carlo solvers
+provided the best MCSA performance. The MCSA method defined in
+Eq.~(\ref{eq:MCSA-iteration}) uses the adjoint method to estimate the
+error in Eq.~(\ref{eq:MCSA-residual_solve}) instead of the direct
+method outlined in \S~\ref{sec:direct-method}. To demonstrate the
+effectiveness of the adjoint method over the direct method within the
+context of MCSA, we choose the 2D time-dependent Poisson equation as a
+simple model problem:
+\begin{equation}
+  \frac{\partial \vu}{\partial t} = \nabla^2 \vu\:.
+  \label{eq:poisson_equation}
+\end{equation}
+For all comparisons, a single time step is computed with backwards
+Euler time integration. The Laplacian is differenced on a square
+Cartesian grid with a second-order five-point stencil,
+\begin{equation}
+  \nabla^2_5 = \frac{1}{\Delta^2}[u_{i-1,j} + u_{i+1,j} + u_{i,j-1} +
+    u_{i,j+1} - 4 u_{i,j}]\:,
+  \label{eq:five_point_stencil}
+\end{equation}
+and a fourth-order nine-point stencil,
+\begin{multline}
+  \nabla^2_9 = \frac{1}{6\Delta^2}[4 u_{i-1,j} + 4 u_{i+1,j} + 4
+    u_{i,j-1} + 4 u_{i,j+1} + u_{i-1,j-1}\\ + u_{i-1,j+1} +
+    u_{i+1,j-1} + u_{i+1,j+1} - 20 u_{i,j}]\:,
+  \label{eq:nine_point_stencil}
+\end{multline}
+both assuming a grid size of $\Delta$ in both the $i$ and $j$
+directions. For a single time step solution, we then have the
+following sparse linear system to be solved with the MCSA method:
+\begin{equation}
+  \vA \vu^{n+1} = \vu^n\:.
+  \label{eq:poisson_eq_lin_sys}
+\end{equation}
+Both the stencils will be used to vary the size and density of the
+sparse linear system in Eq.~(\ref{eq:poisson_eq_lin_sys}).
+
+A timing and convergence study is used to demonstrate the
+effectiveness of the adjoint method as compared to the direct
+method. To assess both the CPU time and number of iterations required
+to converge to a solution, a problem of constant $\Delta$ was used
+with varying values of grid size, fixing the spectral radius of the
+system at a constant value for each variation. Both the five-point and
+nine-point stencils were used with both the direct and adjoint
+solvers. For each case, 50 random walk permutations were computed per
+MCSA iteration with a weight cutoff of \sn{1}{-4} for the stochastic
+linear solver and a convergence tolerance of \sn{1}{-8} for the MCSA
+iterative solver. All computations presented in this section were
+completed on a 2.66 GHz Intel Core i7 machine with 8 GB 1067 MHz DDR3
+memory. Figure~\ref{fig:poisson_cpu_time} gives the CPU time needed
+for each case to converge in seconds, and
+Fig.~\ref{fig:poisson_iterations} gives the number of iterations
+needed for each case to converge relative to the problem size.
+\begin{figure}[h!]
+  \centering
+  \includegraphics[width=5in,clip]{Adjoint_Direct_CPU_Time.pdf}
+  \caption{\sl CPU Time (s) to converge vs. Problem Size ($N$ for an
+    $N \times N$ square mesh). Both the adjoint and direct solvers are
+    used with the five point and nine point stencils. A significant
+    CPU time speedup is noted with the adjoint method due to the
+    reduced number of random walk events.}
+  \label{fig:poisson_cpu_time}
+\end{figure}
+
+\begin{figure}[h!]
+  \centering
+  \includegraphics[width=5in,clip]{Adjoint_Direct_Iterations.pdf}
+  \caption{\sl Iterations to converge vs. Problem Size ($N$ for an $N
+    \times N$ square mesh). Both the adjoint and direct solvers are
+    used with the five-point and nine-point stencils. Both methods
+    give similar iteration behavior.}
+  \label{fig:poisson_iterations}
+\end{figure}
+
+We see clearly in Fig.~\ref{fig:poisson_cpu_time} that the using the
+adjoint solver with MCSA results in several orders of magnitude
+speedup over the direct solver while the number of iterations required
+to converge is of a similar scale. We expect this for several
+reasons. First, with an equivalent number of histories specified for
+both solvers and a system of size $N \times N$, the direct solver will
+compute $N \times N$ random walks per permutation while the adjoint
+solver will only compute one. This is necessary in the direct method
+to ensure a contribution from each state as the random walk sequence
+will only contribute to the starting state. For the adjoint method,
+because the random walk sequence contributes to the state in which it
+currently resides, fewer histories are neccessary to compute
+contributions from all the nonzero parts of the source. Initially,
+this may not seem like a fair comparison; however, we see from
+Fig.~\ref{fig:poisson_iterations} that the number of iterations
+required to converge is approximately the same and therefore the extra
+computations performed by the direct method do not improve the error
+estimation. Therefore, although the total number of random walks per
+iteration is in fact larger for the direct method, the solution
+computed at each iteration is no better than that computed by the
+adjoint method.
+
+As an additional comparison, the convergence behavior of MCSA can be
+analyzed using both solvers to detect any performance benefits. To
+assess the convergence properties of MCSA using each solver and
+stencil, the infinity norm of the residual computed in
+Eq.~(\ref{eq:MCSA-residual}) was collected at each iteration for a
+fixed problem size. Figure~\ref{fig:poisson_convergence} gives the
+results of these computations. First, it is worthy to note on the
+semilog plot that we are indeed achieving the expected exponential
+convergence from MCSA. Second, we note that both the direct and
+adjoint methods have approximately equivalent convergence behavior,
+leading to CPU time being the dominating factor in method selection.
+\begin{figure}[h!]
+  \centering
+  \includegraphics[width=5in,clip]{Adjoint_Direct_Convergence.pdf}
+  \caption{\sl Infinity norm of the solution residual vs. iteration
+    number for a problem of fixed size. Both the adjoint and direct
+    solvers are used with the five point and nine point stencils.}
+  \label{fig:poisson_convergence}
+\end{figure}
 
@@
 
+\section{Stochastic Methods Verification}
+\label{sec:method_verification}
+
 \section{Numerical Experiments}
-\label{sec:key_questions}
-
-\section{Solution Strategies}
-\label{sec:solution_strategies}
-
-\subsection{Parallel MCSA Verification}
-\label{subsubsec:mcsa_verification}
-
-\subsection{FANM Verification}
-\label{subsubsec:fanm_verification}
+\label{sec:numerical_experiments}
 

### 2012-09-27 doc/prelim/document/chapters/research_proposal/research_proposal.tex
incorporated jcomp work

@@
 provided the best MCSA performance. The MCSA method defined in
-Eq.~(\ref{eq:MCSA-iteration}) uses the adjoint method to estimate the
-error in Eq.~(\ref{eq:MCSA-residual_solve}) instead of the direct
-method outlined in \S~\ref{sec:direct-method}. To demonstrate the
-effectiveness of the adjoint method over the direct method within the
-context of MCSA, we choose the 2D time-dependent Poisson equation as a
-simple model problem:
+Eq.~(\ref{eq:mcsa}) uses the adjoint method to estimate the error
+instead of the direct method outlined in \S~\ref{sec:direct_mc}. To
+demonstrate the effectiveness of the adjoint method over the direct
+method within the context of MCSA, we choose the 2D time-dependent
+Poisson equation as a simple model problem:
 \begin{equation}
-  \frac{\partial \vu}{\partial t} = \nabla^2 \vu\:.
+  \frac{\partial \ve{u}}{\partial t} = \nabla^2 \ve{u}\:.
   \label{eq:poisson_equation}
@@
 \begin{equation}
-  \vA \vu^{n+1} = \vu^n\:.
+  \ve{A} \ve{u}^{n+1} = \ve{u}^n\:.
   \label{eq:poisson_eq_lin_sys}
@@
   \centering
-  \includegraphics[width=5in,clip]{Adjoint_Direct_CPU_Time.pdf}
-  \caption{\sl CPU Time (s) to converge vs. Problem Size ($N$ for an
-    $N \times N$ square mesh). Both the adjoint and direct solvers are
-    used with the five point and nine point stencils. A significant
-    CPU time speedup is noted with the adjoint method due to the
-    reduced number of random walk events.}
+  \includegraphics[width=5in,clip]{chapters/research_proposal/Adjoint_Direct_CPU_Time.pdf}
+  \caption{\textbf{CPU Time (s) to converge vs. Problem Size ($N$ for
+      an $N \times N$ square mesh).} \textit{Both the adjoint and
+      direct solvers are used with the five point and nine point
+      stencils. A significant CPU time speedup is noted with the
+      adjoint method due to the reduced number of random walk
+      events.}}
   \label{fig:poisson_cpu_time}
@@
   \centering
-  \includegraphics[width=5in,clip]{Adjoint_Direct_Iterations.pdf}
-  \caption{\sl Iterations to converge vs. Problem Size ($N$ for an $N
-    \times N$ square mesh). Both the adjoint and direct solvers are
-    used with the five-point and nine-point stencils. Both methods
-    give similar iteration behavior.}
+  \includegraphics[width=5in,clip]{chapters/research_proposal/Adjoint_Direct_Iterations.pdf}
+  \caption{\textbf{Iterations to converge vs. Problem Size ($N$ for an
+      $N \times N$ square mesh).} \textit{Both the adjoint and direct
+      solvers are used with the five-point and nine-point
+      stencils. Both methods give similar iteration behavior.}}
   \label{fig:poisson_iterations}
@@
 stencil, the infinity norm of the residual computed in
-Eq.~(\ref{eq:MCSA-residual}) was collected at each iteration for a
-fixed problem size. Figure~\ref{fig:poisson_convergence} gives the
-results of these computations. First, it is worthy to note on the
-semilog plot that we are indeed achieving the expected exponential
-convergence from MCSA. Second, we note that both the direct and
-adjoint methods have approximately equivalent convergence behavior,
-leading to CPU time being the dominating factor in method selection.
+Eq.~(\ref{eq:mcsa}) was collected at each iteration for a fixed
+problem size. Figure~\ref{fig:poisson_convergence} gives the results
+of these computations. First, it is worthy to note on the semilog plot
+that we are indeed achieving the expected exponential convergence from
+MCSA. Second, we note that both the direct and adjoint methods have
+approximately equivalent convergence behavior, leading to CPU time
+being the dominating factor in method selection.
 \begin{figure}[h!]
   \centering
-  \includegraphics[width=5in,clip]{Adjoint_Direct_Convergence.pdf}
-  \caption{\sl Infinity norm of the solution residual vs. iteration
-    number for a problem of fixed size. Both the adjoint and direct
-    solvers are used with the five point and nine point stencils.}
+  \includegraphics[width=5in,clip]{chapters/research_proposal/Adjoint_Direct_Convergence.pdf}
+  \caption{\textbf{Infinity norm of the solution residual
+      vs. iteration number for a problem of fixed size.} \textit{Both
+      the adjoint and direct solvers are used with the five point and
+      nine point stencils.}}
   \label{fig:poisson_convergence}

### 2012-09-27 doc/prelim/document/chapters/research_proposal/research_proposal.tex
adding heat equation setup figure

@@
 \label{subsec:mcsa_finite_element}
+The Epetra-based MCSA implementation has enabled the solution of
+finite element problems leveraging the Panzer finite element assembly
+engine, providing an initial basis for our experimental framework. To
+demonstrate this, consider the 2-dimensional steady-state heat
+equation:
+\begin{equation}
+  \boldsymbol{\nabla} \cdot k \boldsymbol{\nabla} T = Q\:,
+  \label{eq:heat_equation}
+\end{equation}
+where $k$ is the thermal conductivity, $T$ is the temperature, and $Q$
+is the thermal source. We can write the heat equation in weak form
+using Eq~(\ref{eq:fe_weak_form}):
+\begin{equation}
+  \int_{\Omega} \boldsymbol{\nabla}^T \ve{v} k \boldsymbol{\nabla} T
+  d\Omega - \int_{\Omega} \ve{v}Q d\Omega - \int_{\Gamma} \ve{v}
+  \bar{q} d\Gamma - \int_{\Gamma} \ve{v} k \frac{\partial T}{\partial
+    n} d\Gamma= 0\:,
+  \label{eq:weak_heat_equation}
+\end{equation}
+where the boundary conditions have now been incorporated. We can
+interpret the last two integral terms as the contributions from a
+fixed boundary flux source and the flux through the domain
+boundary. This weak formulation can then be directly implemented
+within the finite element assembly engine along with the appropriate
+boundary conditions and source terms. For this example, we choose a
+square Cartesian grid with boundary conditions as shown in
+Figure~\ref{fig:heat_eq_setup}.
 

### 2012-09-27 doc/prelim/document/chapters/research_proposal/research_proposal.tex
adding heat equation solution

@@
 Figure~\ref{fig:heat_eq_setup}.  
+\begin{figure}[htpb!]
+  \begin{center}
+    \scalebox{1.5}{
+      \input{chapters/research_proposal/heat_eq_setup.pdftex_t} }
+  \end{center}
+  \caption{\textbf{Problem setup for finite element solution to 2D
+      heat equation.} \textit{Dirichlet conditions are set for the
+      temperature on all 4 boundaries of the Cartesian grid with
+      thermal conductivity set to 2.0. No thermal source was
+      present.}}
+  \label{fig:heat_eq_setup}
+\end{figure}
+A value of 2.0 is used for $k$ with no thermal source and second-order
+basis functions are chosen for the finite element form. This problem
+was solved using Panzer to assemble the finite element system and
+Jacobi preconditioned MCSA as the linear solver with the adjoint
+Neumann-Ulam method as the Monte Carlo solver. The results of this
+calculation are presented in Figure~\ref{fig:heat_eq_solution}.
+\begin{figure}[htpb!]
+  \centering
+  \includegraphics[width=5in,clip]{chapters/research_proposal/heat_eq_solution.eps}
+  \caption{\textbf{Finite element solution to the 2D steady-state heat
+      equation using MCSA.} \text{A $100 \times 100$ Cartesian grid
+      was used with Jacobi preconditioning applied to the linear
+      system.}}
+  \label{fig:heat_eq_solution}
+\end{figure}
 

### 2012-09-27 drivers/TransientPoisson/TransientPoisson_main.cpp

@@ -78,8 +78,8 @@ int main( int argc, char * argv[] )
 	Teuchos::rcp( new Teuchos::ParameterList );
     plist->set( "X Blocks", 1 );
     plist->set( "Y Blocks", 1 );
-    plist->set( "X Elements", 10 );
-    plist->set( "Y Elements", 10 );
+    plist->set( "X Elements", 100 );
+    plist->set( "Y Elements", 100 );
     mesh_factory.setParameterList( plist );
 
     Teuchos::RCP<panzer_stk::STK_Interface> mesh = 
@@ -127,7 +127,39 @@ int main( int argc, char * argv[] )
 	    std::string element_block_id = "eblock-0_0";
 	    std::string dof_name = "TEMPERATURE";
 	    std::string strategy = "Constant";
-	    double value = -5.0;
+	    double value = 0.0;
+	    Teuchos::ParameterList bc_list;
+	    bc_list.set( "Value", value );
+	    panzer::BC bc( bc_id, bc_type, sideset_id, element_block_id,
+			   dof_name, strategy, bc_list );
+	    boundary_conditions.push_back( bc );
+	}
+
+	// Right side boundary condition.
+	{
+	    std::size_t bc_id = 2;
+	    panzer::BCType bc_type = panzer::BCT_Dirichlet;
+	    std::string sideset_id = "right";
+	    std::string element_block_id = "eblock-0_0";
+	    std::string dof_name = "TEMPERATURE";
+	    std::string strategy = "Constant";
+	    double value = 5.0;
+	    Teuchos::ParameterList bc_list;
+	    bc_list.set( "Value", value );
+	    panzer::BC bc( bc_id, bc_type, sideset_id, element_block_id,
+			   dof_name, strategy, bc_list );
+	    boundary_conditions.push_back( bc );
+	}
+
+	// Bottom side boundary condition.
+	{
+	    std::size_t bc_id = 3;
+	    panzer::BCType bc_type = panzer::BCT_Dirichlet;
+	    std::string sideset_id = "bottom";
+	    std::string element_block_id = "eblock-0_0";
+	    std::string dof_name = "TEMPERATURE";
+	    std::string strategy = "Constant";
+	    double value = 0.0;
 	    Teuchos::ParameterList bc_list;
 	    bc_list.set( "Value", value );
 	    panzer::BC bc( bc_id, bc_type, sideset_id, element_block_id,
[…]

### 2012-09-27 doc/prelim/document/chapters/research_proposal/research_proposal.tex
updates to research proposal

@@
   \centering
-  \includegraphics[width=5in,clip]{chapters/research_proposal/heat_eq_solution.eps}
+  \includegraphics[width=5in,clip]{chapters/research_proposal/heat_eq_solution.png}
   \caption{\textbf{Finite element solution to the 2D steady-state heat
-      equation using MCSA.} \text{A $100 \times 100$ Cartesian grid
+      equation using MCSA.} \textit{A $100 \times 100$ Cartesian grid
       was used with Jacobi preconditioning applied to the linear

### 2012-09-28 doc/prelim/document/chapters/parallel_mc/parallel_mc.tex
research proposal updates

@@
 diffusion equation \citep{evans_monte_2012}, achieving identical
-numerical results as convential Krylov solvers. Their approach was
-instead to use residual Monte Carlo as a synthetic acceleration for a
-stationary method. To derive this method, we begin by splitting the
-operator in Eq~(\ref{eq:linear_problem})
+numerical results as and marginally better performance than
+conventional Krylov solvers. Their approach was instead to use
+residual Monte Carlo as a synthetic acceleration for a stationary
+method. To derive this method, we begin by splitting the operator in
+Eq~(\ref{eq:linear_problem})
 \begin{equation}

### 2012-09-28 doc/prelim/document/chapters/research_proposal/research_proposal.tex
research proposal updates

@@
 will determine some of their properties. Finally, we select a
-challenge problem designed to potentially demsonstrate the
+large-scale challenge problem designed to potentially demonstrate the
 effectiveness of our new methods in the context of real-world,
@@
 sparse matrices as outlined in \S~\ref{subsec:fanm_storage}. This will
-permit efficient memory use and permit the savings we expect by
+permit efficient memory use and allow for the savings we expect by
 implementing the FANM method. Second, Epetra is fully parallelized in
 that the parallel graph structures and decomposed storage formats are
-utilized to domain decompose both the operator and vector in the
+utilized to domain decompose both the operator and vectors in the
 linear system. These data structures provide access to the required
@@
 the direct and adjoint Monte Carlo solvers with the MCSA
-method\footnote{The contents of this section have been taken directly
-  from my contributions to \citep{evans_monte_2012}}. The aim of this
+method.\footnote{The contents of this section have been taken directly
+  from my contributions to \citep{evans_monte_2012}} The aim of this
 initial work was to determine which of the two Monte Carlo solvers
@@
 demonstrate the effectiveness of the adjoint method over the direct
-method within the context of MCSA, we choose the 2D time-dependent
-Poisson equation as a simple model problem:
+method within the context of MCSA, we choose the 2-dimensional
+time-dependent Poisson equation as a simple model problem:
 \begin{equation}
@@
 \end{figure}
+In the results, we see the characteristic parabolic solution profiles
+of the heat equation corresponding corresponding to the correct
+Dirichlet condition. Based on this work, a finite element formulation
+generated by the Panzer framework can be readily solved with a general
+MCSA implementation. In addition, the above problem can already be
+solved in parallel with conventional Krylov methods. Therefore, once a
+parallel MCSA implementation is complete, this framework can be used
+to generate those experiments and immediately compare to Krylov
+methods. In addition, extension to nonlinear model problems requires
+only a few small additions to the framework, providing the remainder
+of the toolset we need for our analysis.
 
@@
 \label{sec:method_verification}
+As we perform our numerical experiments, the results from the new
+tools that we are using must be verified against benchmark
+results. These benchmark tests should allow use to verify that the new
+stochastic solvers that we develop generate the expected numerical
+results in parallel for linear and nonlinear problems that have a
+known, well-documented solution. In addition, as we explore parallel
+strategies for the Monte Carlo methods and introduce bias or
+non-reproducable results into our solution estimators, these
+verification problems will allow us to determine if that cost is
+acceptable. To verify the linear solvers, we will choose the above
+2-dimensional heat equation problem in Eq~(\ref{eq:heat_equation}) on
+a rectilinear grid for which analytical solutions are available. As a
+complement to comparison with the analytical solutions, we may also
+check those results against production implementations of Krylov
+solvers. For the nonlinear solvers, we employ a new set of equations
+with readily available benchmark solutions.
+
+\subsection{FANM Verification}
+\label{subsec:fanm_verification}
+To verify the FANM method for nonlinear problems, we choose benchmark
+solutions for the 2-dimensional, steady, incompressible Navier-Stokes
+equations on a recitilinear grid in much the same way as Shadid and
+Pawlowski's work on Newton-Krylov methods for the solution of these
+equations \citep{shadid_inexact_1997,pawlowski_globalization_2006}. We
+define these equations as follows :
+\begin{subequations}
+  \begin{gather}
+    \rho \ve{u} \cdot \nabla \ve{u} - \nabla \cdot \ve{T} - \rho
+    \ve{g} = \ve{0}
+    \label{eq:ns_momentum}\\
+    \nabla \cdot \ve{u} = 0
+    \label{eq:ns_continuity}\\
+    \rho C_p \ve{u} \cdot \nabla T + \nabla \cdot \ve{q} = 0\:,
+    \label{eq:ns_energy}
+  \end{gather}
+  \label{eq:navier_stokes}
+\end{subequations}
+where $\rho$ is the fluid density, $\ve{u}$ is the fluid velocity,
+$\ve{g}$ gravity, $C_p$ the specific heat capacity at constant
+pressure of the fluid and $T$ the temperature of the
+fluid. Eq~(\ref{eq:ns_momentum}) provides momentum transport,
+Eq~(\ref{eq:ns_continuity}) provides the mass balance, and
+Eq~(\ref{eq:ns_energy}) provides energy transport. In addition, we
+close the system with the following equations:
+\begin{subequations}
+  \begin{gather}
+    \ve{T} = -P \ve{I} + \mu[\nabla \ve{u} + \nabla \ve{u}^T]
+    \label{eq:ns_stress_tensor}\\
+    \ve{q} = - k \nabla T\:,
+    \label{eq:ns_heat_flux}
+  \end{gather}
+  \label{eq:ns_closure}
+\end{subequations}
+where $\ve{T}$ is the stress tensor, $P$ is the fluid pressure, $\mu$
+is the dynamic viscosity of the fluid, $\ve{q}$ is the heat flux in
+the fluid, and $k$ is the thermal conductivity of the fluid. We will
+then apply these equations to the following three benchmark problems.
+
+\subsubsection{Thermal Convection Cavity Problem}
+\label{subsubsec:natural_convection_cavity}
+
+\subsubsection{Lid Driven Cavity Problem}
+\label{subsubsec:lid_driven_cavity}
+
+\subsubsection{Backward-Facing Step Problem}
+\label{subsubsec:backward_facing_step}
 
@@
 \label{sec:numerical_experiments}
+Using the verification problems for the linear and nonlinear solvers
+as a testing environment, we propose a set of experiments that will be
+used to characterize several qualities of these stochastic methods.
 
@@
 \label{sec:challenge_problem}
+
+\section{Conclusion}
+\label{sec:conclusion}

### 2012-09-28 doc/prelim/document/chapters/parallel_mc/parallel_mc.tex
working on the numerical experiments section

@@
 between solving a linear system with Monte Carlo methods and fixed
-source Monte Carlo transport problems is the content of Markov chains
-that are generated. These transitions represeneted by these chains are
-bound by probabilities and weights and are initiated by the sampling
-of a source. In the context of transport problems, those transitions
-represent events such as pariticle scattering and absorbition with
-probabilities that are determined by physical data in the form of
-cross sections. For stochastic matrix inversion, those transitions
-represent moving between the equations of the linear system (and
-therefore the physical domain which they represent) and their
-probabilites are defined by the coefficients of those
+source Monte Carlo transport problems is the content of the Markov
+chains that are generated. These transitions represeneted by these
+chains are bound by probabilities and weights and are initiated by the
+sampling of a source. In the context of transport problems, those
+transitions represent events such as pariticle scattering and
+absorbition with probabilities that are determined by physical data in
+the form of cross sections. For stochastic matrix inversion, those
+transitions represent moving between the equations of the linear
+system (and therefore the physical domain which they represent) and
+their probabilites are defined by the coefficients of those
 equations. Ultimately, we tally the contributions to generate
 expectation values in the desired states as we progress through the
-chains.
+chains. Therefore, parallel methods for Monte Carlo radiation
+transport can be abstracted and we can use those concepts that apply
+to matrix inversion methods as an initial means of developing a
+parallel Neumann-Ulam-type solver. Based on the results observed in
+the last decade of parallel Monte Carlo development, we can generate
+many practical questions that this type of work could answer.
 
-Therefore, parallel methods for Monte Carlo radiation transport and
-abstract those that apply to matrix inversion methods as an initial
-means of developing a parallel Neumann-Ulam-type solver. Based on the
-results observed in the last decade of parallel Monte Carlo
-development, we can generate many practical questions that this type
-of work could answer. Given a decomposed domain, it is clear that
-domain overlap significantly reduces the amount of communication
-required between adjacent domains as histories move to states that are
-not owned by the local processor. How much domain overlap is suitable
-for matrix inversion problems? Are we memory limited by large problems
-such that only so much overlap is feasible? Are there metrics related
-to the properties of the matrix such that we can provide guidelines
-for selecting the amount of overlap required? Furthermore, as Wagner
-and colleagues observed \citep{wagner_hybrid_2010}, with marginal
-domain overlap for reactor problems, the percentage of histories
-leaving the local domain can easily be reduced to less than 1\%. Evans
-and colleagues in their initial MCSA development typically used only
-50 particle histories in order to compute a correction in an MCSA
+Given a decomposed domain, it is clear that domain overlap
+significantly reduces the amount of communication required between
+adjacent domains as histories move to states that are not owned by the
+local processor. How much domain overlap is suitable for matrix
+inversion problems? Are we memory limited by large problems such that
+only so much overlap is feasible? Are there metrics related to the
+properties of the matrix such that we can provide guidelines for
+selecting the amount of overlap required? Furthermore, as Wagner and
+colleagues observed \citep{wagner_hybrid_2010}, with marginal domain
+overlap for reactor problems, the percentage of histories leaving the
+local domain can easily be reduced to less than 1\%. Evans and
+colleagues in their initial MCSA development typically used only 50
+particle histories in order to compute a correction in an MCSA
 iteration \citep{evans_monte_2012}. Per the Central Limit Theorem,
 that correlates to a statistical uncertainty of 14.1\% in the
-correction yet as compared to convential Krylov solvers, MCSA achieved
-the identical numerical results. If good convergence and numerically
-accurate solutions can be achieved with such a large uncertainty in
-the correction computation, then perhaps with enough domain overlap
-the minimal amount of histories that do transition to non-local states
-can be ignored and thus eliminate all communication in the parallel
-Monte Carlo adjoint solver and create an embarassingly parallel
-method. It will be important to determine if the expectation value
-bias in the MCSA correction generated by this approximation will
+correction, yet as compared to convential Krylov solvers, MCSA
+achieved identical numerical results. If good convergence and
+numerically accurate solutions can be achieved with such a large
+uncertainty in the correction computation, then perhaps with enough
+domain overlap the minimal amount of histories that do transition to
+non-local states can be ignored and thus eliminate all communication
+in the parallel Monte Carlo adjoint solver and create an embarassingly
+parallel method. It will be important to determine if the expectation
+value bias in the MCSA correction generated by this approximation will
 remain within the bounds of the already high statistical uncertainty

### 2012-09-28 doc/prelim/document/chapters/research_proposal/research_proposal.tex
working on the numerical experiments section

@@
 library in the Trilinos sceintific computing framework
-\citep{heroux_overview_2005}, function evaluation routines in the form
-of Eq~(\ref{eq:fe_weak_form}) are generated to represent the finite
-element problem and boundary conditions. Per Bartlett's automatic
-differentiation work and an associated parallel linear algebra
-framework, the equation sets for all physics implemented are assembled
-into the nonlinear system. By leveraging the general linear algebra
-framework, we are able to abstract away those concepts in the
-stochastic solvers implementation and instead focus on an
-algorithm-oriented approach for the solver implementation
-\citep{musser_algorithm-oriented_1994}.
+\citep{notz_graph-based_2010,heroux_overview_2005}, function
+evaluation routines in the form of Eq~(\ref{eq:fe_weak_form}) are
+generated to represent the finite element problem and boundary
+conditions. Per Bartlett's automatic differentiation work and an
+associated parallel linear algebra framework, the equation sets for
+all physics implemented are assembled into the nonlinear system. By
+leveraging the general linear algebra framework, we are able to
+abstract away those concepts in the stochastic solvers implementation
+and instead focus on an algorithm-oriented approach for the solver
+implementation \citep{musser_algorithm-oriented_1994}.
 
@@
 Eq~(\ref{eq:ns_continuity}) provides the mass balance, and
-Eq~(\ref{eq:ns_energy}) provides energy transport. In addition, we
-close the system with the following equations:
+Eq~(\ref{eq:ns_energy}) provides energy transport with viscous
+dissipation effects neglected. In addition, we close the system with
+the following equations:
 \begin{subequations}
@@
 \end{subequations}
-where $\ve{T}$ is the stress tensor, $P$ is the fluid pressure, $\mu$
-is the dynamic viscosity of the fluid, $\ve{q}$ is the heat flux in
-the fluid, and $k$ is the thermal conductivity of the fluid. We will
-then apply these equations to the following three benchmark problems.
+where $\ve{T}$ is the stress tensor, $P$ is the hydrodynamic pressure,
+$\mu$ is the dynamic viscosity of the fluid, $\ve{q}$ is the heat flux
+in the fluid, and $k$ is the thermal conductivity of the fluid. This
+set of strongly coupled equations possesses both the nonlinearities
+and asymmetries that we are seeking for qualification of the FANM
+method. Further, physical parameters withing these equations can be
+tuned to enhance the nonlinearities. We will then apply these
+equations to the following three standard benchmark problems.
 
@@
 \label{subsubsec:natural_convection_cavity}
+In 1983 benchmark solution for the natural convection of air in a
+square cavity was published \citep{de_vahl_davis_natural_1983} as
+shown in Figure~\ref{fig:natural_convection_cavity} for the solution
+of the energy, mass, and momentum equations.
+\begin{figure}[htpb!]
+  \begin{center}
+    \scalebox{1.5}{
+      \input{chapters/research_proposal/natural_convection_cavity.pdftex_t} }
+  \end{center}
+  \caption{\textbf{Problem setup for the natural convection cavity
+      benchmark.} \textit{Dirichlet conditions are set for the
+      temperature on the left and right while Neumann conditions are
+      set on the top and bottom of the Cartesian grid. The temperature
+      gradients will cause bouyancy-driven flow. Zero velocity
+      Dirichlet conditions are set on each boundary. No thermal source
+      was present.}}
+  \label{fig:natural_convection_cavity}
+\end{figure}
+In this problem, a rectilinear grid is applied to the unit square. No
+heat flow is allowed out of the top and bottom of the square with a
+zero Neumann condition specified. Bouyancy driven flow is generated by
+the temperature gradient from the cold and hot Dirichlet conditions on
+the left and right boundaries of the box. By adjusting the Rayleigh
+number of the fluid (and therefore adjusting the ratio of convective
+to conductive heat transfer), we can adjust the influence of the
+nonlinear convection term in Eq~(\ref{eq:ns_momentum}). In Shadid's
+work, Rayleigh numbers of up to \sn{1}{6} were used for this benchmark
+on a $100 \times 100$ square mesh grid.
 
@@
 \label{subsubsec:lid_driven_cavity}
+As an extension of the convection problem, the second benchmark
+problem given by Ghia \citep{ghia_high-re_1982} adds a driver for the
+flow to introduce higher Reynolds numbers into the system, providing
+more intertial force to overcome the viscous forces in the fluid. The
+setup for this problem is equally simple, containing only the
+Dirichlet conditions as given in Figure~\ref{fig:lid_driven_cavity}
+and is only applied to the mass and momentum equations on the unit
+square.
+\begin{figure}[htpb!]
+  \begin{center}
+    \scalebox{1.5}{
+      \input{chapters/research_proposal/lid_driven_cavity.pdftex_t} }
+  \end{center}
+  \caption{\textbf{Problem setup for the lid driven cavity benchmark.}
+    \textit{Dirichlet conditions of zero are set for the velocity on
+      the left and right and bottom while the Dirichlet condition set
+      on the top provides a driving force on the fluid.}}
+  \label{fig:lid_driven_cavity}
+\end{figure}
+The top boundary condition will provide a driver for the flow and its
+variation will in turn vary the Reynolds number of the fluid. An
+increased velocity will generate more inertial forces in the fluid,
+which will overcome the viscous forces and again increase the
+influence of the nonlinear terms in Eq~(\ref{eq:ns_momentum}). Shadid
+also used a $100 \times 100$ grid for this benchmark problem with
+Reynolds numbers up to \sn{1}{4}.
 
@@
 \label{subsubsec:backward_facing_step}
+The third benchmark was generated by Gartling in 1990 and consists of
+both flow over a backward step and an outflow boundary condition
+\citep{gartling_test_1990}. Using the mass and momentum equations
+while neglecting the energy equation, this problem utilizes a longer
+domain with a 1/30 aspect ratio with the boundary conditions as shown
+in Figure~\ref{fig:backward_facing_step}.
+\begin{figure}[htpb!]
+  \begin{center}
+    \scalebox{1.3}{
+      \input{chapters/research_proposal/backward_facing_step.pdftex_t} }
+  \end{center}
+  \caption{\textbf{Problem setup for the backward facing step
+      benchmark.} \textit{Zero velocity boundary conditions are
+      applied at the top and bottom of the domain while the outflow
+      boundary condition on the right boundary is represented by zero
+      stress tensor components in the direction of the flow. For the
+      inlet conditions, the left boundary is split such that the top
+      half has a fully formed parabolic flow profile and the bottom
+      half has a zero velocity condition, simulating flow over a
+      step.}}
+  \label{fig:backward_facing_step}
+\end{figure}
+In this problem, the inflow condition is specified by a fully-formed
+parabolic flow profile over a zero velocity boundary representing a
+step. The flow over this step will generate a recirculating backflow
+under the inlet flow towards the step. As in the lid driven cavity
+problem, the nonlinear behavior of this benchmark and the difficulty
+in obtaining a solution is dictated by the Reynolds number of the
+fluid. In Shadid's work, a $20 \times 400$ non-square recitlinear grid
+was used to discretize the domain with Reynolds number up to
+\sn{5}{2}.
 
-\section{Numerical Experiments}
+\section{Proposed Numerical Experiments}
 \label{sec:numerical_experiments}
@@
 as a testing environment, we propose a set of experiments that will be
-used to characterize several qualities of these stochastic methods.
+used to characterize several qualities of these stochastic methods. At
+the end of Chapter~\ref{ch:stochastic_methods}, we posed several
+research questions related to the parallelization of the adjoint
+Neumann-Ulam method and the MCSA method while in
+Chapter~\ref{ch:nonlinear_problem} we posed questions related to the
+the FANM method. We will generate numerical experiments from the these
+questions by dividing them into 4 categories: domain overlap studies
+for the parallel Neumann-Ulam method, domain replication studies for
+the parallel Neumann-Ulam method, parallel performance and numerical
+accuracy studies for the MCSA method, and parallel performance and
+numerical accuracy studies for the FANM method.
+
+\subsection{Domain Overlap Studies for the Parallel Neumann-Ulam
+  Method}
+\label{subsec:domain_overlap_studies}
+Characterizing and understanding the domain overlap behavior of a
+parallel Neumann-Ulam method will be critical to efficient
+implementations of other solvers that will be built on top of these
+methods. In the reactor physics work that we reviewed, the amount
+domain overlap used could be directly correlated to physical
+parameters of the system, namely the dimension of repeated geometric
+structures and the mean free path of neutrons in the system. We
+therefore seek analgous measurements for the general linear
+system. These measurements will characterize how far from its starting
+state a random walk sequence moves as a function of properties of the
+linear system. Our experimental framework will allow us to fix various
+properties of the system while varying others including the size of
+the linear system, its spectral radius, and other parameters related
+to the Eigenvalues and conditioning of the linear operator. By
+varying the domain overlap and measuring parallel performance with
+respect to these properties, we can develop guidelines for setting the
+overlap with respect to measurable properties of the linear
+operator. In addition, varying the amount of overlap will allow us to
+measure the additional memory costs incurred.
+
+\subsection{Domain Replication Studies for the Parallel Neumann-Ulam
+  Method}
+\label{subsec:domain_replication_studies}
+Employing domain replication with the Neumann-Ulam method will be an
+exercise in batch statistics. By using replication, we also expect to
+incur an additional memory overhead. As noted in previous work, domain
+replication was coupled with domain overlap in order to improve load
+balancing and scalability in the implementations. As replication will
+likely be required to solve future resiliency issues and may even by
+required by this work in order to achieve scalability, it parallel
+performance must also be measured. We will therefore measure the
+memory and performance overhead incurred by varying the amount of
+domain replication leveraged by the Neumann-Ulam solver with respect
+to a fixed linear system.
+
+\subsection{Parallel Performance and Numerical Accuracy Studies for MCSA
+  Method}
+\label{subsec:mcsa_studies}
+A parallel MCSA implementation and our numerical test framework will
+allow us to compare its performance to Krylov methods and allow us to
+experiment with the way we combine it with the Neumann-Ulam solver. As
+disucussed in \S~\ref{subsec:parallel_adjoint}, previous work with
+MCSA used a correction generated by the Neumann-Ulam method with a
+fairly large uncertainty near 14\%. We hypothesized that if we can
+identify the amount of overlap in or Neumann-Ulam experiments required
+to eliminate most transitions between adjacent domains owned by
+different processors, then perhaps transport to adjacent domains can
+be ignored. This will result in a biased estimator for the MCSA
+correction that cannont be reproduced, however, these studies will aim
+to determine if that bias and lack of reproducablity affects the
+numerical results and reproducability of the MCSA method. Furthermore,
+we will use the experimental framework to measure the convergence
+rates for nonsymmetric systems and memory usage of the MCSA algorithm
+and compare them to conventional Krylov methods using the same
+framework.
+
+\subsection{Parallel Performance and Numerical Accuracy Studies for FANM
+  Method}
+\label{subsec:fanm_studies}
 

### 2012-09-28 doc/prelim/document/chapters/nonlinear_problem/nonlinear_problem.tex
finished draft of numerical experiments

@@
 methods, and therefore we should compare FANM perfomance to JFNK-based
-schemes. More important than performance in the situation is the
+schemes. More important than performance in this situation is the
 reduced memory pressure that a FANM implementation provides, as

### 2012-09-28 doc/prelim/document/chapters/research_proposal/research_proposal.tex
finished draft of numerical experiments

@@
 \label{subsec:fanm_studies}
+Aside from resiliency motivations, the primary driver behind the FANM
+method is reduced memory pressure due to the lack of additional
+storage beyond that of the linear system in the MCSA method. The
+nonlinear benchmark problems we have chosen will let us measure this
+memory usage and compare it to conventional Newton-Krylov
+methods. Furthermore, by varying the strength of the nonlinearities in
+those problems (effectively by increasing the amount of convective
+transport in the system), the benchmarks will become more difficult to
+solve and require even more iterations to converge. This will give us
+a means by which to vary the complexity of the system which we can
+then compare to memory usage. Additionally, with the finite element
+assembly framework we can choose functional discretizations of varying
+accuracy, allowing us to vary the degrees of freedom in the system and
+therefore also vary the memory required.
 

### 2012-09-28 doc/prelim/document/chapters/research_proposal/research_proposal.tex
working on challenge problem

@@
 to determine if that bias and lack of reproducablity affects the
-numerical results and reproducability of the MCSA method. Furthermore,
-we will use the experimental framework to measure the convergence
-rates for nonsymmetric systems and memory usage of the MCSA algorithm
-and compare them to conventional Krylov methods using the same
-framework.
+numerical results and reproducability of the MCSA method, yielding a
+critical outcome for this work. Furthermore, we will use the
+experimental framework to measure the convergence rates for
+nonsymmetric systems and memory usage of the MCSA algorithm and
+compare them to conventional Krylov methods using the same framework.
 
@@
 accuracy, allowing us to vary the degrees of freedom in the system and
-therefore also vary the memory required.
+therefore also vary the memory required for each problem.
 
@@
 \label{sec:challenge_problem}
+Our numerical experiments will provide us with a set of metrics giving
+best-performance solver parameters for problems of different types. As
+an additional bonus, these experiments will let us further develop new
+parallel implementations of the stochastic methods in order to improve
+their performance and scalability. With these advancements in hand, we
+look forward towards applying them to large-scale problems in reactor
+physics.
 

### 2012-09-28 doc/prelim/document/chapters/research_proposal/research_proposal.tex
finished research proposal draft

@@
 
-\section{Challenge Problem}
+\section{Proposed Challenge Problem}
 \label{sec:challenge_problem}
 Our numerical experiments will provide us with a set of metrics giving
-best-performance solver parameters for problems of different types. As
-an additional bonus, these experiments will let us further develop new
-parallel implementations of the stochastic methods in order to improve
-their performance and scalability. With these advancements in hand, we
-look forward towards applying them to large-scale problems in reactor
-physics.
+best-performance stochastic solver parameters for problems of
+different types. As an additional bonus, these experiments will let us
+further develop new parallel implementations of the stochastic methods
+in order to improve their performance and scalability. With these
+advancements in hand, we look forward towards applying them to
+large-scale problems in reactor physics. The Consortium for Advanced
+Simulation of Light Water Reactors (CASL) is a Department of Energy
+Modeling and Simulation Hub focused on predictive multiphysics
+modeling of nuclear reactors
+\citep{u.s._depertment_of_energy_casl:_2011}. Their effors are focused
+on enabling nuclear reactor analysis and development using today's
+supercomputing technology to achieve higher power uprates, higher fuel
+burnup, and predictive accident scenario analysis. As a means of
+driving their research and development, CASL has worked with the
+nuclear industry to identify challenge problems of critical importance
+to achieving these goals.
+
+To meet some of the high fidelity modeling requirements of these
+challenge problems, the Drekar multiphysics code is being developed at
+Sandia National Laboratories for fluid flow and conjugate heat
+transfer problems \citep{pawlowski_drekar::cfd_2012}. Drekar is being
+used to analyze complex problems including the grid-to-rod-fretting
+(GTRF) phenomena observed in light water reactor fuel rods. GTRF
+occurs due to the turbulent flow of water through the subchannels of
+the reactor core. This turbulence causes the nuclear fuel pins to
+vibrate and the fuel cladding to wear against structures in the
+reactor, causing localized hot spots and ultimately cladding
+failure. Reducing these instances of failure and understanding how
+they occur through predictive multiple physics simulation of fluid
+flow and heat transfer will permit more robust fuel assembly
+designs. In addition, the turbulence studies that Drekar performs will
+allow for improved flow patterns to be generated in the assembly
+subchannels, allowing for better heat transfer between the nuclear
+fuel and the water, ultimately allowing for a more efficient reactor.
+
+Drekar is a massively parallel physics framework built upon the same
+Panzer library as well as other open-source tools that we are
+leveraging in our own experimental framework. Furthermore, its
+multiphysics formulation utilizes Newton-Krylov methods to generate a
+fully consistent solution scheme for the Navier-Stokes equations with
+various turbulence models for fluid simulation as well as the heat
+conduction problem for simulation in solid materials. Performance
+metrics show scaling to $O(100,000)$ cores on leadership class
+machines with $O(\sn{1}{9})$ degrees of freedom in these
+simulations. Given these features, Drekar will be an excellent
+framework in which we can pose our own challenge problem for the FANM
+and newly parallelized MCSA methods proposed by this research. At the
+time our preliminary numerical experiments are completed for
+characterizing the parallel FANM and MCSA implementations, we propose
+implementing them within the Drekar framework and performing the
+equivalent of its largest GTRF mutliphysics problem to date, most
+likely to be a fuel-assembly sized problem. Doing so will allow us to
+not only implement our new methods in a production physics code and
+apply it to a problem of great interest to the nuclear community, but
+it will also allow us to compare its performance and scaling to the
+Newton-Krylov methods currently leveraged within Drekar. If
+successful, we expect to demonstrate improved computational
+performance from a memory usage and/or an algorithmic scaling
+perspective.
 
@@
 \label{sec:conclusion}
+We have proposed the research and development of novel, massively
+parallel stochastic methods for the solution of linear and nonlinear
+problems. These new methods aim to enhance the solution of multiple
+physics problems of great national interest. Current progress has
+shown that these methods can indeed be expanded into a general linear
+algebra framework that permits their application to a broad range of
+problems. Additional analysis has also been completed that begins to
+characterize their performance in different configurations. Leveraging
+this work, we have defined a multiphysics framework based on the
+finite element method and have demonstrated that these stochastic
+methods can be used effectively to solve the linear systems generated
+by these discretizations.
+
+To guide our development of parallel stochastic methods, we have
+devised several classes of numerical experiments using our
+multiphysics framework that will provide insight into their properties
+and performance for linear and nonlinear problems. Furthermore, these
+experiments aim to provide metrics that will give an indication of
+both solver performance for various problem types and the type of
+parallel decomposition required to effectively solve these
+problems. Domain decomposition, domain replication, and their
+combination will be of great interest for the analysis of parallel
+Monte Carlo methods while algorithm scalability and memory usage will
+be of interest for the MCSA and FANM methods that leverage them.
+
+We culminate our research and development with a challenge problem by
+applying these new stochastic methods in the Drekar multiphysics
+code. With Drekar, we look forward to providing a new and robust
+solution method at the production scale. With these tools and a
+leadership class computing platform, we will provide performance
+measurements for the stochastic solvers and compare them to
+conventional methods currently leveraged within Drekar. With the
+potential to improve computational performance in these environments,
+these new methods have the potential to help meet the high fidelty
+modeling and simulation needs of the nuclear community.

### 2012-09-29 doc/prelim/document/chapters/introduction/introduction.tex
working on intro

@@
 \label{ch:introduction}
-
 In many fields of engineering and physics, linear and nonlinear
 problems are a primary focus of study. Recent focus on multiple
-physics systems (and in particular nuclear systems) adds a new level
-of complexity to common nonlinear systems as solution strategies
-change when they are coupled to other problems. Furthermore, a desire
+physics systems in the nuclear reactor modeling and simulation
+community adds a new level of complexity to common nonlinear systems
+as solution strategies change when they are coupled to other problems
+\citep{u.s._depertment_of_energy_casl:_2011}. Furthermore, a desire
 for predictive simulations to enhance the safety and performance of
-engineered systems creates a need for extremely high fidelity
+nuclear systems creates a need for extremely high fidelity
 computations to be performed for these coupled systems as a means to
-capture effects not modeled by coarser methods.
+capture effects not modeled by coarser methods
+\citep{keyes_prospects_1997}.
 
@@
 already coming online, new algorithms to solve these complex problems
-must be developed to leverage this new hardware. Issues such as
-resiliency to node failure and scaling to large numbers of cores will
-be pertinent to robust algorithms aimed at this new
-hardware. Considering these issues, this thesis proposes the
-development of a massively parallel stochastic method for linear
-problems and a novel stochastic method to advance solution techniques
-for nonlinear problems.
+must be developed to leverage this new hardware
+\citep{kogge_using_2011}. Issues such as resiliency to node failure
+and scaling to large numbers of cores will be pertinent to robust
+algorithms aimed at this new hardware. Considering these issues, this
+thesis proposes the development of a massively parallel stochastic
+method for linear problems and a novel stochastic method to advance
+solution techniques for nonlinear problems.
 
+We discuss in this chapter our motivation developing these Monte Carlo
+solvers by providing background on the current state of solver
+techniques for multiphysics problems and the hardware that they
+leverage. In addition, multiphysics problems of interest to the
+nuclear engineering community are provided to further motivate this
+work. We then provide a formal presentation of the multiphysics
+system and the problems that arise from it.
+
+\section{Motivation}
+\label{sec:motivation}
+
+\subsection{Multiphysics Methods Advancement}
+\label{subsec:methods_advancement}
+For some time, the particle transport community has been utilizing
+Monte Carlo methods for the solution of transport problems
+\citep{lewis_computational_1993}. The partial differential equation
+(PDE) community has focused on various deterministic methods for
+solutions to linear problems \citep{saad_iterative_2003,
+  kelley_iterative_1995}. In between these two areas are a not widely
+known small group of stochastic methods for solving sparse linear
+systems \citep{forsythe_matrix_1950, hammersley_monte_1964,
+  halton_sequential_1962, halton_sequential_1994}. In recent years,
+these methods have been further developed for transport problems in
+the form of Monte Carlo Synthetic-Acceleration (MCSA)
+\citep{evans_residual_2003, evans_monte_2009, evans_monte_2012} but
+have yet to be applied to more general sparse linear systems commonly
+generated by the computational physics community. Compared to other
+methods in this regime, MCSA offers three attractive qualities; (1)
+the linear problem operator need not be symmetric or
+positive-definite, thereby reducing preconditioning complexity, (2)
+the stochastic nature of the solution method provides a natural
+solution to the issue of resiliency, and (3) is amenable to
+parallelization using modern methods developed by the transport
+community \citep{wagner_hybrid_2010}. The developement of MCSA as a
+general linear solver and the development of a parallel MCSA method
+will be new and unique features of this work, providing a framework
+with which other issues such as resiliency may be addressed.
+
+In addition to linear solver advancements, nonlinear solvers may also
+benefit from a general and parallel MCSA scheme. In the nuclear
+engineering community, nonlinear problems are often addressed by
+either linearizing the problem or building a segregated scheme and
+using traditionaly iterative or direct methods to solve the resulting
+system \citep{pletcher_computational_1997}. In the mathematics
+community, various Newton methods have been popular
+\citep{kelley_iterative_1995}. Recently, Jacobian-Free Newton-Krylov
+(JFNK) schemes \citep{knoll_jacobian-free_2004} have been utilized in
+multiple physics architectures and advanced single physics codes
+\citep{gaston_parallel_2009}. The benefits of JFNK schemes are that
+the Jacobian is never formed, simplifying the implementation, and a
+Krylov solver is leveraged (typically GMRES or Conjugate Gradient),
+providing excellent convergence properties for well-conditioned and
+well-scaled systems. However, there are two potential drawbacks to
+these methods for high fidelity predictive simulations: (1) the
+Jacobian is approximated by a first-order differencing method on the
+order of machine precision such that this error can grow beyond that
+of those in a fine-grained system \citep{kelly_iterative_1995} and (2)
+for systems that are not symmetric positive-definite (which will be
+the case for most multiphysics systems and certainly for most
+preconditioned systems) the Krylov subspace generated by the GMRES
+solver may become prohibitively large
+\citep{knoll_newton-krylov_1995}. To address these issues, this thesis
+proposes a new and novel method for nonlinear systems based on the
+MCSA method.
+
+The Forward-Automated Newton-MCSA (FANM) method is proposed as new
+nonlinear solution method. The key features of FANM are: full Jacobian
+generation using modern Forward Automated Differencing (FAD) methods
+\citep{bartlett_automatic_2006}, and MCSA as the inner linear
+solver. This method has several attractive properties. First, the
+first-order approximation to the Jacobian used in JFNK type methods is
+eliminated by generating the Jacobian explicitly with the model
+equations through FAD. Second, the Jacobian need not be explicitly
+formed by the user but is instead automated through FAD; this
+eleminates the complexity of hand-coding derivatives and has also been
+demonstrated to be more efficient computationally than evaluating
+differenced derivatives. Third, unlike GMRES, MCSA does not build a
+subspace during iterations. Although the Jacobian must be explicitly
+formed to use MCSA, for problems that take more than a few GMRES
+iterations to converge the size of the Krylov subspace will grow
+beyond that of the Jacobian. Finally, using MCSA for the linear solve
+provides its benefits for preconditioning, potential resiliency, and
+parallelism.
+
+\subsection{Problems of Interest}
+\label{subsec:problems of interest}
+
+\section{The Multiphysics System}
+\label{sec:multiphysics_system}
+
+\section{Research Outline}
+\label{sec:research_outline}
 This preliminary report oultines the conventional methods used in
@@
 a pressurized water reactor assembly.
-
-\section{Motivation}
-\label{sec:motivation}
-
-For some time, the particle transport community has been utilizing
-Monte Carlo methods for the solution of transport problems
-\citep{lewis_computational_1993}. The partial differential equation
-(PDE) community has focused on various deterministic methods for
-solutions to linear problems \citep{saad_iterative_2003,
-  kelley_iterative_1995}. In between these two areas are a not widely
-known small group of stochastic methods for solving sparse linear
-systems \citep{hammersley_monte_1964, halton_sequential_1962,
-  halton_sequential_1994}. In recent years, these methods have been
-further developed for transport problems in the form of Monte Carlo
-Synthetic-Acceleration (MCSA) \citep{evans_residual_2003,
-  evans_monte_2009} but have yet to be applied to more general sparse
-linear systems commonly generated by the computational physics
-community. Compared to other methods in these regime, MCSA offers
-three attractive qualities; (1) the linear problem operator need not
-be symmetric or positive-definite, thereby reducing preconditioning
-complexity, (2) the stochastic nature of the solution method provides
-a natural solution to the issue of resiliency, and (3) is amenable to
-parallelization using modern methods developed by the transport
-community \citep{wagner_hybrid_2010}. The developement of MCSA as a
-general linear solver and the development of a parallel MCSA method
-will be new and unique features of this work. Resiliency will not be
-addressed.
-
-In addition to linear solver advancements, nonlinear solvers may also
-benefit from a general and parallel MCSA scheme. In the engineering
-community, nonlinear problems are often addressed by either
-linearizing the problem or building a segregated scheme and using
-traditionaly iterative or direct methods to solve the resulting system
-\citep{pletcher_computational_1997}. In the mathematics community,
-various Newton methods have been popular
-\citep{kelley_iterative_1995}. Recently, Jacobian-Free Newton-Krylov
-(JFNK) schemes \citep{knoll_jacobian-free_2004} have been utilized in
-multiple physics architectures and advanced single physics codes
-\citep{gaston_parallel_2009}. The benefits of JFNK schemes are that the
-Jacobian is never formed, simplifying the implementation, and a Krylov
-solver is leveraged (typically GMRES or Conjugate Gradient), providing
-excellent convergence properties for well-conditioned and well-scaled
-systems. However, there are two potential drawbacks to these methods
-for high fidelity predictive simulations: (1) the Jacobian is
-approximated by a first-order differencing method on the order of
-machine precision such that this error can grow beyond that of those
-in a fine-grained system and (2) for systems that are not symmetric
-positive-definite (which will be the case for most multiphysics
-systems and certainly for most preconditioned systems) the Krylov
-subspace generated by the GMRES solver may become prohibitively
-large. To address these issues, this thesis proposes a new and novel
-method for nonlinear systems based on the MCSA method.
-
-The Forward-Automated Newton-MCSA (FANM) method is proposed as new
-nonlinear solution method. The key features of FANM are: full Jacobian
-generation using modern Forward Automated Differencing (FAD) methods
-\citep{hutchison_automatic_2006}, and MCSA as the inner linear
-solver. This method has several attractive properties. First, the
-first-order approximation to the Jacobian used in JFNK type methods is
-eliminated by generating the Jacobian explicitly with the model
-equations through FAD. Second, the Jacobian need not be explicitly
-formed by the user but is instead automated through FAD; this
-eleminates the complexity of hand-coding derivatives and has also been
-demonstrated to be more efficient computationally than evaluating
-differenced derivatives \citep{hutchison_automatic_2006}. Third,
-unlike GMRES, MCSA does not build a subspace during
-iterations. Although the Jacobian must be explicitly formed to use
-MCSA, for problems that take more than a few GMRES iterations to
-converge the size of the Krylov subspace will grow beyond that of the
-Jacobian. Finally, using MCSA for the linear solve provides its
-benefits for preconditioning, potential resiliency, and parallelism.
-
-\section{The Multiphysics System}
-\label{sec:multiphysics_system}
-
-\section{Research Outline}
-\label{sec:research_outline}

### 2012-09-29 doc/prelim/document/chapters/linear_problem/linear_problem.tex
working on intro

@@
 \end{equation}
-We can now resrtict $\ve{H}$ by asserting that $\rho(\ve{H}) < 1$
+We can now restrict $\ve{H}$ by asserting that $\rho(\ve{H}) < 1$
 for a stationary method to converge such that $k$ applications of the

### 2012-09-29 doc/prelim/document/chapters/introduction/introduction.tex
intro updates

@@
 as solution strategies change when they are coupled to other problems
-\citep{u.s._depertment_of_energy_casl:_2011}. Furthermore, a desire
+\citep{u.s._department_of_energy_casl:_2011}. Furthermore, a desire
 for predictive simulations to enhance the safety and performance of
@@
 computations to be performed for these coupled systems as a means to
-capture effects not modeled by coarser methods
-\citep{keyes_prospects_1997}.
+capture effects not modeled by coarser methods.
 
@@
 must be developed to leverage this new hardware
-\citep{kogge_using_2011}. Issues such as resiliency to node failure
-and scaling to large numbers of cores will be pertinent to robust
-algorithms aimed at this new hardware. Considering these issues, this
-thesis proposes the development of a massively parallel stochastic
-method for linear problems and a novel stochastic method to advance
-solution techniques for nonlinear problems.
+\citep{kogge_using_2011}. Issues such as resiliency to node failure,
+limited growth of memory available per node, and scaling to large
+numbers of cores will be pertinent to robust algorithms aimed at this
+new hardware. Considering these issues, this preliminary report
+proposes the development of a massively parallel stochastic method for
+linear problems and a novel stochastic method to advance solution
+techniques for nonlinear problems.
 
-We discuss in this chapter our motivation developing these Monte Carlo
-solvers by providing background on the current state of solver
+We discuss in this chapter our motivation for developing these Monte
+Carlo solvers by providing background on the current state of solver
 techniques for multiphysics problems and the hardware that they
@@
 nuclear engineering community are provided to further motivate this
-work. We then provide a formal presentation of the multiphysics
-system and the problems that arise from it.
+work. We then provide a formal presentation of the multiphysics system
+and the problems that arise from it.
 
@@
 \label{sec:motivation}
+
+\subsection{Problems of Interest}
+\label{subsec:problems of interest}
+
+\section{The Multiphysics System}
+\label{sec:multiphysics_system}
 
@@
 
-\subsection{Problems of Interest}
-\label{subsec:problems of interest}
-
-\section{The Multiphysics System}
-\label{sec:multiphysics_system}
-
 \section{Research Outline}

### 2012-09-29 doc/prelim/document/chapters/introduction/introduction.tex
updating section headers

@@
 as solution strategies change when they are coupled to other problems
-\citep{u.s._department_of_energy_casl:_2011}. Furthermore, a desire
+\citep{u.s._department_of_energy_casl_2011}. Furthermore, a desire
 for predictive simulations to enhance the safety and performance of
@@
 In order to achieve this high fidelity, state-of-the-art computing
-facilities must be leveraged in a way that is both efficient and
+\index{facilities} must be leveraged in a way that is both efficient and
 considerate of hardware-related issues. As scientific computing moves
@@
 order of machine precision such that this error can grow beyond that
-of those in a fine-grained system \citep{kelly_iterative_1995} and (2)
-for systems that are not symmetric positive-definite (which will be
-the case for most multiphysics systems and certainly for most
+of those in a fine-grained system \citep{kelley_iterative_1995} and
+(2) for systems that are not symmetric positive-definite (which will
+be the case for most multiphysics systems and certainly for most
 preconditioned systems) the Krylov subspace generated by the GMRES

### 2012-09-29 doc/prelim/document/chapters/research_proposal/research_proposal.tex
intro updates
updating section headers

@@
 modeling of nuclear reactors
-\citep{u.s._depertment_of_energy_casl:_2011}. Their effors are focused
+\citep{u.s._department_of_energy_casl_2011}. Their effors are focused
 on enabling nuclear reactor analysis and development using today's
@@
 Sandia National Laboratories for fluid flow and conjugate heat
-transfer problems \citep{pawlowski_drekar::cfd_2012}. Drekar is being
+transfer problems \citep{pawlowski_drekar_2012}. Drekar is being
 used to analyze complex problems including the grid-to-rod-fretting

### 2012-09-29 doc/prelim/document/chapters/introduction/introduction.tex
adding figure to introduction

@@
 \label{sec:motivation}
+Predictive modeling and simulation capability requires the combination
+of high fidelity models, high performance computing hardware that can
+handle the intense compute loads required by these models, and modern
+algorithms for solving these problems that leverage this high
+performance hardware. For nuclear reactor analysis, this predictive
+capability can enable tighter design tolerances for improved thermal
+performance and efficiency, higher fuel burnup and therefore reduction
+in generated waste, and high confidence in accident scenario
+models. The physics that dominate these types of analysis include
+neutronics, thermal hydraulics, computational fluid dynamics, and
+structural mechanics.
 
-\subsection{Problems of Interest}
-\label{subsec:problems of interest}
+Although solution techniques in each of these individual categories
+has advanced over the last few decades and in fact leveraged modern
+algorithms and computer architectures, true predictive capability for
+engineered systems can only be achieved through a coupled, multiple
+physics analysis where the effects of feedback between physics are
+modeled. For example, consider the safety analysis of a departure from
+nucleate boiling scenario in the subchannel of a nuclear fuel assembly
+as shown in Figure~\ref{fig:dnb_example}.
+\begin{figure}[htpb!]
+  \begin{center}
+    \scalebox{1.5}{
+      \input{chapters/introduction/dnb_example.pdftex_t} }
+  \end{center}
+  \caption{\textbf{Multiphysics analysis of departure from nuclear
+      boiling.} \textit{A neutronics solution is required to compute
+      power generation in the fuel pins, fluid dynamics is required to
+      characterize boiling and fluid temperature and density, heat
+      transfer is required to compute the fuel and cladding
+      temperature, and the nuclear data modified with the temperature
+      and density data. Strong coupling among the variables creates a
+      nonlinearities.}}
+  \label{fig:dnb_example}
+\end{figure}
+When this event occurs, heat transfer is greatly reduced between the
+fuel and the coolant, causing the fuel centerline temperature to
+rapidly rise. To characterize this boiling phenomena and how it
+affects fuel failure we must consider a neutronics analysis in order
+to compute power generation in the fuel pins, fluid dynamics analysis
+to characterize coolant boiling, temperature, and density, solid
+material heat transfer to characterize fuel and cladding temperature
+and heat transfer with the coolant, and nuclear data processing to
+characterize how changing material temperatures and densities changes
+the cross sections needed for the neutronics calculation. As shown in
+Figure~\ref{fig:dnb_example}, many couplings are required among
+individual physics components in order to accurately model this
+situation with each physics generating and receiving many
+responses. Those variables that are very tightly coupled, such as the
+temperatures generated by the fluid dynamics and heat transfer
+components, will have strong nonlinearities in their behavior and
+would therefore benefit from fully consistent nonlinear solution
+schemes instead of fixed-point type iterations between
+physics\footnote{Fixed-point iterations between physics are commonly
+  referred to as Picard iterations.}. Furthermore, the space and time
+scales over which these effects occur will also vary greatly.
 
@@
 will be new and unique features of this work, providing a framework
-with which other issues such as resiliency may be addressed.
+with which other issues such as resiliency may be addressed in the
+future.
 

### 2012-09-29 doc/prelim/document/chapters/introduction/introduction.tex
finished motiviation

@@
 We discuss in this chapter our motivation for developing these Monte
-Carlo solvers by providing background on the current state of solver
-techniques for multiphysics problems and the hardware that they
-leverage. In addition, multiphysics problems of interest to the
-nuclear engineering community are provided to further motivate this
-work. We then provide a formal presentation of the multiphysics system
-and the problems that arise from it.
+Carlo solvers by providing multiphysics problems of interest to the
+nuclear engineering community. In addition, background on the current
+state of solver techniques for multiphysics problems and the hardware
+that they leverage is provided to further motivate this work. We then
+provide a formal presentation of the multiphysics system and the
+problems that arise from it.
 
@@
 scales over which these effects occur will also vary greatly.
+
+The computational resources required to solve such problems are
+tremendous. Recent work in modeling the coupled fluid flow and solid
+model heat and mass transfer component of the departure from nucleate
+boiling model above was performed as part of analysis of the
+Department of Energy's Consortium for Advanced Simulation of Light
+Water Reactors (CASL) modeling and simulation hub using the Drekar
+multiphysics code \citep{pawlowski_drekar_2012}. Using Drekar,
+multiphysics simulations have been performed with fully consistent
+methods for the solution of nonlinear systems using meshes of
+$O(\sn{1}{9})$ elements leveraging $O(100,000)$ cores on leadership
+class machines. If we aim to advance multiphysics solution techniques,
+then we are motivated to tackle the solution of complex and general
+nonlinear problems exploiting leadership class levels of parallelism.
 

### 2012-09-29 doc/prelim/document/chapters/introduction/introduction.tex
adding hardware based motivation

@@
 In order to achieve this high fidelity, state-of-the-art computing
-\index{facilities} must be leveraged in a way that is both efficient and
-considerate of hardware-related issues. As scientific computing moves
-towards exascale facilities with machines of $O(1,000,000)$ cores
-already coming online, new algorithms to solve these complex problems
-must be developed to leverage this new hardware
+\index{facilities} must be leveraged in a way that is both efficient
+and considerate of hardware-related issues. As scientific computing
+moves towards exascale facilities with machines of $O(1,000,000)$
+cores already coming online, new algorithms to solve these complex
+problems must be developed to leverage this new hardware
 \citep{kogge_using_2011}. Issues such as resiliency to node failure,
@@
 Carlo solvers by providing multiphysics problems of interest to the
-nuclear engineering community. In addition, background on the current
-state of solver techniques for multiphysics problems and the hardware
-that they leverage is provided to further motivate this work. We then
-provide a formal presentation of the multiphysics system and the
-problems that arise from it.
+nuclear engineering community. Hardware-based motivations are also
+provided with the impact of forthcoming computing architectures
+considered. In addition, background on the current state of solver
+techniques for multiphysics problems and a brief comparision to the
+proposed methods is provided to further motivate this work.
 
-\section{Motivation}
-\label{sec:motivation}
+\section{Physics-Based Motivation}
+\label{sec:physics_motivation}
 Predictive modeling and simulation capability requires the combination
 of high fidelity models, high performance computing hardware that can
-handle the intense compute loads required by these models, and modern
-algorithms for solving these problems that leverage this high
+handle the intense computational loads required by these models, and
+modern algorithms for solving these problems that leverage this high
 performance hardware. For nuclear reactor analysis, this predictive
@@
 When this event occurs, heat transfer is greatly reduced between the
-fuel and the coolant, causing the fuel centerline temperature to
-rapidly rise. To characterize this boiling phenomena and how it
-affects fuel failure we must consider a neutronics analysis in order
-to compute power generation in the fuel pins, fluid dynamics analysis
-to characterize coolant boiling, temperature, and density, solid
-material heat transfer to characterize fuel and cladding temperature
-and heat transfer with the coolant, and nuclear data processing to
-characterize how changing material temperatures and densities changes
-the cross sections needed for the neutronics calculation. As shown in
+fuel and the coolant due to the vapor layer generated by boiling,
+causing the fuel centerline temperature to rapidly rise. To
+characterize this boiling phenomena and how it affects fuel failure we
+must consider a neutronics analysis in order to compute power
+generation in the fuel pins, fluid dynamics analysis to characterize
+coolant boiling, temperature, and density, solid material heat
+transfer to characterize fuel and cladding temperature and heat
+transfer with the coolant, and nuclear data processing to characterize
+how changing material temperatures and densities changes the cross
+sections needed for the neutronics calculation. As shown in
 Figure~\ref{fig:dnb_example}, many couplings are required among
@@
 model heat and mass transfer component of the departure from nucleate
-boiling model above was performed as part of analysis of the
+boiling example above was performed as part of analysis of the
 Department of Energy's Consortium for Advanced Simulation of Light
-Water Reactors (CASL) modeling and simulation hub using the Drekar
-multiphysics code \citep{pawlowski_drekar_2012}. Using Drekar,
-multiphysics simulations have been performed with fully consistent
-methods for the solution of nonlinear systems using meshes of
-$O(\sn{1}{9})$ elements leveraging $O(100,000)$ cores on leadership
-class machines. If we aim to advance multiphysics solution techniques,
-then we are motivated to tackle the solution of complex and general
-nonlinear problems exploiting leadership class levels of parallelism.
+Water Reactors (CASL) modeling and simulation hub used the Drekar
+multiphysics code developed at Sandia National Laboratories
+\citep{pawlowski_drekar_2012}. Using Drekar, multiphysics simulations
+have been performed with fully consistent methods for the solution of
+nonlinear systems using meshes of $O(\sn{1}{9})$ elements leveraging
+$O(100,000)$ cores on leadership class machines. Neutronics components
+to be implemented in CASL for multiphysics analysis, such as the
+Denovo radiation transport code developed at Oak Ridge National
+Laboratory \citep{evans_denovo:_2010}, computes trillions of unknowns
+for full core reactor analysis on $O(\sn{1}{9})$ as well. Given the
+large scale and complexity of these problems, if we aim to advance
+multiphysics solution techniques, then we are motivated to advance the
+solution of complex and general nonlinear problems exploiting
+leadership class levels of parallelism.
 
-\section{The Multiphysics System}
-\label{sec:multiphysics_system}
+\section{Hardware-Based Motivation}
+\label{sec:hardware_motivation}
+As leadership class machines move towards the exascale, new algorithms
+must developed that leverage their strengths and adapt to their
+shortcomings. Basic research is required now to advance methods in
+time for these new machines to become operational with the Department
+of Energy specifically allocating funding for the 2013 fiscal year to
+research resilient solver technologies for exascale facilities
+\citep{u.s._department_of_energy_resilient_2012}.
 
-\subsection{Multiphysics Methods Advancement}
-\label{subsec:methods_advancement}
+\section{Research Outline}
+\label{sec:research_outline}
 For some time, the particle transport community has been utilizing
@@
 
-\section{Research Outline}
-\label{sec:research_outline}
 This preliminary report oultines the conventional methods used in

### 2012-09-29 doc/prelim/document/chapters/introduction/introduction.tex
finished intro draft

@@
 
-We discuss in this chapter our motivation for developing these Monte
-Carlo solvers by providing multiphysics problems of interest to the
-nuclear engineering community. Hardware-based motivations are also
-provided with the impact of forthcoming computing architectures
-considered. In addition, background on the current state of solver
-techniques for multiphysics problems and a brief comparision to the
-proposed methods is provided to further motivate this work.
+We discuss in this chapter motivation for advancing Monte Carlo
+solvers by providing multiphysics problems of interest in nuclear
+reactor analysis. Hardware-based motivations are also provided by
+considering the impact of forthcoming computing architectures. In
+addition, background on the current state solver techniques for
+multiphysics problems and a brief comparision to the proposed methods
+is provided to further motivate this work.
 
@@
 time for these new machines to become operational with the Department
-of Energy specifically allocating funding for the 2013 fiscal year to
-research resilient solver technologies for exascale facilities
-\citep{u.s._department_of_energy_resilient_2012}.
+of Energy's Advanced Scientific Computing Reasarch office specifically
+allocating funding for the 2013 fiscal year to research resilient
+solver technologies for exascale facilities
+\citep{u.s._department_of_energy_resilient_2012}. Based on the
+language in this call for proposals, we can identify key issues that a
+set of robust, massively parallel stochastic solvers could provide a
+solution for. As machines begin to operate at hundreds of petaflops
+peak performance and beyond, restrictions on power avaiablity will
+require incredibly high levels of concurrency. Furthermore, this drop
+in power consumption will mean increased pressure on memory as memory
+per node is expected to stagnate while cores per node is expected to
+increase. As the number of cores increases, their clock speed is
+expected to stagnate or even decrease to further reduce power
+consumption and manufacturing costs.
+
+The end result of these hardware changes is that the larger numbers of
+low-powered processors will be prone to both soft failures such as bit
+errors in floating point operations and hard failures where the data
+owned by that processor cannot be recovered. Because these failures
+are predicted to be common, resilient solver technologies are required
+overcome these events. With linear and nonlinear solvers based on
+Monte Carlo techniques, such issues are alleviated by statistical
+arguments. In the case of soft failures, isolated floating point
+errors in Monte carlo simulation are absorbed within tally statistics
+while completely losing hardware during a hard failure is manifested
+as a high variance event where some portion of the Monte Carlo
+histories are lost. These stochastic methods are a paradigm shift from
+current determinstic solver techniques that will suffer greatly from
+the non-deterministic behavior expected from these exascale machines.
+
+In addition to resiliency concerns, the memory restrictions on future
+hardware will hinder modern solvers that derive their robustness from
+using large amounts of memory. Stochastic methods that are formulated
+to use less memory than conventional methods will serve to alleviate
+some of this pressure. In addition, new parallel strategies that may
+be implemented with stochastic methods could offer a new avenue for
+leveraging the expected levels of high concurrency in exascale
+machines.
 
@@
 nonlinear solution method. The key features of FANM are: full Jacobian
-generation using modern Forward Automated Differencing (FAD) methods
-\citep{bartlett_automatic_2006}, and MCSA as the inner linear
+generation using modern Forward Automated Differentiation (FAD)
+methods \citep{bartlett_automatic_2006}, and MCSA as the inner linear
 solver. This method has several attractive properties. First, the
@@
 
-This preliminary report oultines the conventional methods used in
-practice for solving linear and nonlinear problems to provide a
-mathematical basis upon which to build new algorithms that aim to
-solve some of the aforementioned issues. These new algorithms will be
-outlined in full with links made to past work and their potential for
-offering improvements to the computational physics community. Progress
-to date will be reported including the development of a new physics
-framework that will serve as a test bed for these new methods that
-leverages a preliminary implementation of the stochastic method. A
-research plan will then be provided that continues the development and
+This preliminary report oultines in Chapter~\ref{ch:linear_problem}
+the conventional methods used in practice for solving linear and
+nonlinear problems to provide a mathematical basis upon which to build
+new algorithms that aim to solve some of the aforementioned
+issues. Parallel schemes for conventional methods are also provided
+for background and understanding of how current methods may or may not
+map to future hardware. In addition some components of their parallel
+implementations may be applied to stochastic methods development. In
+Chapter~\ref{ch:stochastic_methods}, Monte Carlo algorithms for
+solving linear systems and new parallel strategies will be outlined in
+full with links made to past work and their potential for offering
+improvements to the multiphysics analysis community. From these
+stochastic methods, a new nonlinear method is proposed in
+Chapter~\ref{ch:nonlinear_problem} and compared to conventional
+methods for solving nonlinear problems. The research proposal for this
+work is presented in Chapter~\ref{ch:research_proposal} with progress
+to date reported including the development of a new finite
+element-based physics framework that leverages a preliminary general
+implementation of the stochastic method that will serve as a test bed
+for these new methods. An outline of a set of numerical experiments
+will then be provided that continues the development and
 implementation of these new methods and verifies them with benchmark
 solutions, culminating in the analysis of a multiphysics simulation of
-a pressurized water reactor assembly.
+a pressurized water reactor subsystem using a production code system.

### 2012-09-30 doc/prelim/document/chapters/introduction/introduction.tex
updates to intro

@@
 considering the impact of forthcoming computing architectures. In
-addition, background on the current state solver techniques for
-multiphysics problems and a brief comparision to the proposed methods
-is provided to further motivate this work.
+addition, background on the current solver techniques for multiphysics
+problems and a brief comparision to the proposed methods is provided
+to further motivate this work.
 
@@
 modeled. For example, consider the safety analysis of a departure from
-nucleate boiling scenario in the subchannel of a nuclear fuel assembly
-as shown in Figure~\ref{fig:dnb_example}.
+nucleate boiling scenario in the subchannel of a nuclear fuel
+assembly.
 \begin{figure}[htpb!]
@@
   \end{center}
-  \caption{\textbf{Multiphysics analysis of departure from nuclear
-      boiling.} \textit{A neutronics solution is required to compute
-      power generation in the fuel pins, fluid dynamics is required to
-      characterize boiling and fluid temperature and density, heat
-      transfer is required to compute the fuel and cladding
-      temperature, and the nuclear data modified with the temperature
-      and density data. Strong coupling among the variables creates a
-      nonlinearities.}}
+  \caption{\textbf{Multiphysics dependency analysis of departure from
+      nucleate boiling.} \textit{A neutronics solution is required to
+      compute power generation in the fuel pins, fluid dynamics is
+      required to characterize boiling and fluid temperature and
+      density, heat transfer is required to compute the fuel and
+      cladding temperature, and the nuclear data modified with the
+      temperature and density data. Strong coupling among the
+      variables creates strong nonlinearities.}}
   \label{fig:dnb_example}
@@
 The computational resources required to solve such problems are
-tremendous. Recent work in modeling the coupled fluid flow and solid
-model heat and mass transfer component of the departure from nucleate
-boiling example above was performed as part of analysis of the
-Department of Energy's Consortium for Advanced Simulation of Light
-Water Reactors (CASL) modeling and simulation hub used the Drekar
-multiphysics code developed at Sandia National Laboratories
+tremendous. Recent work in modeling coupled fluid flow and solid
+material heat and mass transfer in a reactor subsystem, similar to the
+same components of the departure from nucleate boiling example above,
+was performed as part of analysis of the Department of Energy's
+Consortium for Advanced Simulation of Light Water Reactors (CASL)
+modeling and simulation hub using the Drekar multiphysics code
+developed at Sandia National Laboratories
 \citep{pawlowski_drekar_2012}. Using Drekar, multiphysics simulations
@@
 Laboratory \citep{evans_denovo:_2010}, computes trillions of unknowns
-for full core reactor analysis on $O(\sn{1}{9})$ as well. Given the
-large scale and complexity of these problems, if we aim to advance
-multiphysics solution techniques, then we are motivated to advance the
-solution of complex and general nonlinear problems exploiting
-leadership class levels of parallelism.
+for full core reactor analysis on $O(\sn{1}{9})$ element meshes and
+$O(100,000)$ cores as well. Given the large scale and complexity of
+these problems, if we aim to advance multiphysics solution techniques,
+then we are motivated to advance the solution of complex and general
+nonlinear problems exploiting leadership class levels of parallelism.
 
@@
 shortcomings. Basic research is required now to advance methods in
-time for these new machines to become operational with the Department
-of Energy's Advanced Scientific Computing Reasarch office specifically
-allocating funding for the 2013 fiscal year to research resilient
-solver technologies for exascale facilities
+time for these new machines to become operational. Organized work is
+already moving forward in this area with the Department of Energy's
+Advanced Scientific Computing Reasarch office specifically allocating
+funding for the next several years to research resilient solver
+technologies for exascale facilities
 \citep{u.s._department_of_energy_resilient_2012}. Based on the
@@
   kelley_iterative_1995}. In between these two areas are a not widely
-known small group of stochastic methods for solving sparse linear
-systems \citep{forsythe_matrix_1950, hammersley_monte_1964,
+known group of stochastic methods for solving sparse linear systems
+\citep{forsythe_matrix_1950, hammersley_monte_1964,
   halton_sequential_1962, halton_sequential_1994}. In recent years,
@@
 for background and understanding of how current methods may or may not
-map to future hardware. In addition some components of their parallel
+map to future hardware. In addition, some components of their parallel
 implementations may be applied to stochastic methods development. In

### 2012-09-30 doc/prelim/document/chapters/nonlinear_problem/nonlinear_problem.tex
updates to intro

@@
 
-The work of Bartlett and others \citep{hutchison_automatic_2006}
+The work of Bartlett and others \citep{bartlett_automatic_2006}
 extended initial Fortran-based work in the area of automatic

### 2012-10-01 doc/prelim/document/chapters/linear_problem/linear_problem.tex
updating section headers
linear problem updates

@@
 \citep{zienkiewicz_finite_2005} ultimately generates sets of coupled
-equations in the form of matrix problems.  In addition, these matrices
-are typically sparse, meaning that the vast majority of their
+equations in the form of matrix problems. In many cases, these
+matrices are sparse, meaning that the vast majority of their
 constituent elements are zero. This sparsity is due to the fact that
@@
   meaning that we can in fact compute $\ve{A}^{-1}$. In this work we
-will focus our efforts on approximately inverting the operator through
-various means.
+  will focus our efforts on approximately inverting the operator
+  through various means.
 
@@
 \begin{equation}
-p  ||\ve{A}||_{2} = \sqrt{\rho(\ve{A}^T\ve{A})}\:,
+  ||\ve{A}||_{2} = \sqrt{\rho(\ve{A}^T\ve{A})}\:,
   \label{eq:matrix_2_norm}
@@
 \label{sec:stationary_methods}
-Stationary methods arise from splitting the operator in
-Eq~(\ref{eq:linear_problem})
+Stationary methods for linear systems arise from splitting the
+operator in Eq~(\ref{eq:linear_problem}):
 \begin{equation}
@@
 \begin{equation}
-  \ve{M}\ve{x} + \ve{N}\ve{x} = \ve{b}\:.
+  \ve{M}\ve{x} - \ve{N}\ve{x} = \ve{b}\:.
   \label{eq:linear_split_equation1}
@@
 preconditioning is not an immediate requirement for implementing the
-algorithms. Furthermore, variants are available that reduce the number
-of global communications required (consider
+algorithms. We note here that variants are available that reduce the
+number of global communications required (consider
 \cite{sosonkina_scalable_1998} as an example of reducing global
@@
 Arnoldi), however, we will only consider the basic algorithms here as
-this handful operations can be generalized to fit more complicated
+this handful of operations can be generalized to fit more complicated
 algorithms. In all of these cases, we assume a general matrix/vector
@@
 orthogonalization procedure. By definition, the vector product is a
-global operation that effectively collapses two vectors to a single
-value. Therefore, we cannot eliminate all global
+global operation that effectively collapses a set of vectors to a
+single value. Therefore, we cannot eliminate all global
 communications. Instead, vector product operations are formulated as

### 2012-10-01 doc/prelim/document/chapters/introduction/introduction.tex
spellcheck done. draft complete

@@
 moves towards exascale facilities with machines of $O(1,000,000)$
-cores already coming online, new algorithms to solve these complex
+cores already coming on-line, new algorithms to solve these complex
 problems must be developed to leverage this new hardware
@@
 addition, background on the current solver techniques for multiphysics
-problems and a brief comparision to the proposed methods is provided
+problems and a brief comparison to the proposed methods is provided
 to further motivate this work.
@@
 capability can enable tighter design tolerances for improved thermal
-performance and efficiency, higher fuel burnup and therefore reduction
+performance and efficiency, higher fuel burn-up and therefore reduction
 in generated waste, and high confidence in accident scenario
@@
 fuel and the coolant due to the vapor layer generated by boiling,
-causing the fuel centerline temperature to rapidly rise. To
+causing the fuel center-line temperature to rapidly rise. To
 characterize this boiling phenomena and how it affects fuel failure we
@@
 already moving forward in this area with the Department of Energy's
-Advanced Scientific Computing Reasarch office specifically allocating
+Advanced Scientific Computing Research office specifically allocating
 funding for the next several years to research resilient solver
@@
 solution for. As machines begin to operate at hundreds of petaflops
-peak performance and beyond, restrictions on power avaiablity will
+peak performance and beyond, restrictions on power availability will
 require incredibly high levels of concurrency. Furthermore, this drop
@@
 arguments. In the case of soft failures, isolated floating point
-errors in Monte carlo simulation are absorbed within tally statistics
+errors in Monte Carlo simulation are absorbed within tally statistics
 while completely losing hardware during a hard failure is manifested
@@
 histories are lost. These stochastic methods are a paradigm shift from
-current determinstic solver techniques that will suffer greatly from
+current deterministic solver techniques that will suffer greatly from
 the non-deterministic behavior expected from these exascale machines.
@@
 parallelization using modern methods developed by the transport
-community \citep{wagner_hybrid_2010}. The developement of MCSA as a
+community \citep{wagner_hybrid_2010}. The development of MCSA as a
 general linear solver and the development of a parallel MCSA method
@@
 either linearizing the problem or building a segregated scheme and
-using traditionaly iterative or direct methods to solve the resulting
+using traditionally iterative or direct methods to solve the resulting
 system \citep{pletcher_computational_1997}. In the mathematics
@@
 formed by the user but is instead automated through FAD; this
-eleminates the complexity of hand-coding derivatives and has also been
+eliminates the complexity of hand-coding derivatives and has also been
 demonstrated to be more efficient computationally than evaluating
-differenced derivatives. Third, unlike GMRES, MCSA does not build a
+difference derivatives. Third, unlike GMRES, MCSA does not build a
 subspace during iterations. Although the Jacobian must be explicitly
@@
 
-This preliminary report oultines in Chapter~\ref{ch:linear_problem}
+This preliminary report outlines in Chapter~\ref{ch:linear_problem}
 the conventional methods used in practice for solving linear and

### 2012-10-01 doc/prelim/document/chapters/linear_problem/linear_problem.tex
spellcheck done. draft complete

@@
 $\ve{x}$. The set of all permutations form the \textit{Eigenvectors}
-of $\ve{A}$. A quantity of particular interest that is computatable
+of $\ve{A}$. A quantity of particular interest that is computable
 from the eigenvalues of a matrix $\ve{A}$ is the \textit{spectral
@@
 $\mathcal{K}$ and constrain it with $\mathcal{L}$, each iteration
-performs a projection that systematically anihilates the components of
+performs a projection that systematically annihilates the components of
 the residual that exists in $\mathcal{L}$. This then means that if our
@@
 of the system, then Eq~(\ref{eq:linear_min_res}) tells us that each
-projection step gaurantees us that the norm of the new residual will
+projection step guarantees us that the norm of the new residual will
 never be worse than that of the previous step and will typically move
@@
 Eq~(\ref{eq:linear_problem}), the selection of the subspaces
-$\mathcal{K}$ and $\mathcal{L}$ can serve to both gaurantee
+$\mathcal{K}$ and $\mathcal{L}$ can serve to both guarantee
 convergence and optimize the rate at which the residual is decreased.
@@
 where $m$ denotes the dimensionality of the subspace. In order to
-accomodate a more general structure for the operator in
+accommodate a more general structure for the operator in
 Eq~(\ref{eq:linear_problem}), we often choose an \textit{oblique}
@@
 orthogonal projection procedures that have a dimensionality less than
-$m$ will be refferred to as \textit{short recurrence relations}.  Once
+$m$ will be referred to as \textit{short recurrence relations}.  Once
 $\ve{V}_m$ is found, per the constraint subspace definition it then
@@
 procedure can be restarted at the cost of losing information in the
-orthogonalization process, creating the potentitial to generate new
+orthogonalization process, creating the potential to generate new
 search directions that are not orthogonal to all previous search
@@
 of such operations is bound to the implementation, asymptotically
-perfomance should be the same across all implementations.
+performance should be the same across all implementations.
 
@@
 the constraints to the solution vector. Vector update operations are
-embarassingly parallel in that they require no communication
+embarrassingly parallel in that they require no communication
 operations to be successfully completed; all data operated on is
@@
 where $\ve{y}$ and $\ve{x}$ are vectors of global size $N_g$, local
-size $N_l$, and $a \in \mathbb{R}^N$. In order to avoid communcation,
+size $N_l$, and $a \in \mathbb{R}^N$. In order to avoid communication,
 the vectors $\ve{y}$ and $\ve{x}$ must have the same parallel
@@
 matrix-vector multiplications: $\ve{A}_1\ve{x}_1 + \ve{A}_1\ve{x}_2 =
-\ve{y}_1$. Here, some of the data is intrinisically local, and some
+\ve{y}_1$. Here, some of the data is intrinsically local, and some
 must be gathered from other processes using the partitioning data
@@
 
-\subsection{Parallel Perfomance Implications for Krylov Methods}
+\subsection{Parallel Performance Implications for Krylov Methods}
 \label{subsec:projection_method_performance}
@@
 perform in order to implement Krylov methods, we can make a few
-statements about parallel perfomance and implications for operation on
+statements about parallel performance and implications for operation on
 machines of increasing size.  For very large distributed machines,
 global reduction operations required at several levels of Krylov
-algorithms stand to reduce scalablitity and perfomance. Furthermore,
+algorithms stand to reduce scalability and performance. Furthermore,
 communication between adjacent domains in matrix-vector multiply

### 2012-10-01 doc/prelim/document/chapters/nonlinear_problem/nonlinear_problem.tex
spellcheck done. draft complete

@@
 \label{ch:nonlinear_problem}
-Nonlinear equation sets are a common occurence in multiphysics
-problems. Systems of partial differntial equations such as those that
+Nonlinear equation sets are a common occurrence in multiphysics
+problems. Systems of partial differential equations such as those that
 describe fluid flow or more general transport processes when
@@
 equations with nonlinearities present in the variables. Traditionally,
-such systems have been solved by linearizing them such that the
+such systems have been solved by linearizing them in a form where the
 nonlinearities in the variables are eliminated and more traditional
@@
 order to gauge if some improvements can be made. We therefore devise a
-new method using MCSA in conjuction with new methods, accomodating the
+new method using MCSA in conjunction with new methods, accommodating the
 fact that MCSA requires an explicitly formed linear operator in its
@@
 define this update as the \textit{Newton correction} at the $k$
-iterate, $\delta \ve{u}^k$. We can then rerarrange
+iterate, $\delta \ve{u}^k$. To finish, we can then rearrange
 Eq~(\ref{eq:newton_derivation_2}) to define the Newton iteration
@@
 \end{subequations}
-We then have three distinct steps to perform: evaluation of the
+There are then three distinct steps to perform: evaluation of the
 nonlinear residuals using the solution at the $k$ iterate, the
@@
 functions that describe the physics we are modeling are updated
-simulataneously in the iteration sequence with respect to one
+simultaneously in the iteration sequence with respect to one
 another. This is in comparison to \textit{inconsistent} strategies,
@@
     methods} may be used to modify the Newton correction in a more
-  desireable direction such that convergence properties can be
+  desire able direction such that convergence properties can be
   improved when the initial guess for $\ve{u}$ is poor
@@
 orthogonalization procedure, at each nonlinear iteration this cost is
-reset as the Jacbobian matrix will change due to its dependence on the
+reset as the Jacobian matrix will change due to its dependence on the
 solution vector. This means that for every nonlinear iteration, a
@@
 correction and we can modify the Krylov solver parameters accordingly
-to accomodate this. In most nonlinear problems, the Jacobian operator
-is generally nonsymmetric and therefore either Krylov methods with
-long recurrence relations that can handle nonsymmetric systems must be
+to accommodate this. In most nonlinear problems, the Jacobian operator
+is generally non-symmetric and therefore either Krylov methods with
+long recurrence relations that can handle non-symmetric systems must be
 considered or the Newton correction system must be preconditioned such
@@
 set of tokamak plasma problems with the goal of measuring solver
-performance with Newton's method. They note several tradeoffs in using
+performance with Newton's method. They note several trade offs in using
 Krylov methods with the Newton solver. The first is that the
@@
 as compared to keeping all vectors as the new set of vectors is not
-necessarily orthogonal to the previous set, and therfore not optimal
+necessarily orthogonal to the previous set, and therefore not optimal
 over the entire iteration procedure. The orthogonality condition can
@@
 gradient stabilized (Bi-CGSTAB), and transpose-free quasiminimal
-residual (TFQMR) methdos as Lanzcos-based short recurrence relation
+residual (TFQMR) methods as Lanzcos-based short recurrence relation
 solvers. All solvers were used to compute the right-preconditioned
@@
 implementations where the Jacobian was explicitly formed and GMRES
-perfoming best for matrix-free implementations. However, upon
+performing best for matrix-free implementations. However, upon
 investigating the convergence of the inner iterations, it was observed
@@
 various numerical methods including a difference-based Jacobian-free
-formulation. Jacobian-Free methods, and in particular
-\textit{Jacobian-Free Newton-Krylov} (JFNK) methods
-\citep{knoll_jacobian-free_2004}, rely on forming the action of the
-Jacobian on a vector as required by the Krylov solver through a
-forward difference scheme. In this case, the action of the Jacobian on
-some vector $\ve{v}$ is given as:
+formulation.
+
+Jacobian-Free methods, and in particular \textit{Jacobian-Free
+  Newton-Krylov} (JFNK) methods \citep{knoll_jacobian-free_2004}, rely
+on forming the action of the Jacobian on a vector as required by the
+Krylov solver through a forward difference scheme. In this case, the
+action of the Jacobian on some vector $\ve{v}$ is given as:
 \begin{equation}
@@
 Of interest to Bartlett, Averick, and the many others that have
-researched automatic differentation are measures of its performance
-relative to hand-coded derivatives and capturing the Jacboian matrix
+researched automatic differentiation are measures of its performance
+relative to hand-coded derivatives and capturing the Jacobian matrix
 from matrix-free approximations. Given their element-based function
@@
 element were increased up to a few hundred. Based on these results,
-this type of automatic differentation formulation was deemed
+this type of automatic differentiation formulation was deemed
 acceptable for use in large-scale, production physics codes.
@@
 Krylov method where the operator is not required, we do not incur any
-additional storage costs once the iteration sequence
-begins. Furthermore, in the context of nonlinear problems, the
-Jacobian matrix that we are required to generate for the Monte Carlo
-solvers may be generated at will from the nonlinear functions in the
-Newton system using automatic differentation. Not only do we then have
-a simple and automated way to generate the Jacobian, but we also enjoy
-a Jacobian of numerical precision equivalent to that of our function
-evaluations. We therefore propose the \textit{Forward-Automated
-  Newton-MCSA} (FANM) method that utilizes all of the above
-components. We hypothesize that such a method will be competetive with
-Newton-Krylov methods not only from a convergence and timing
-perspective, but also relax scaling requirements of matrix-free
-methods and memory costs of both matrix-free and fully formed Jacobian
-methods to allow the application developer to solve problems of finer
-discretization and higher-order functional representations while
-maintaining a robust and efficient parallel implementation.
+additional storage costs once the iteration sequence begins. In the
+context of nonlinear problems, the Jacobian matrix that we are
+required to generate for the Monte Carlo solvers may be generated at
+will from the nonlinear functions in the Newton system using automatic
+differentiation. Not only do we then have a simple and automated way to
+generate the Jacobian, but we also enjoy a Jacobian of numerical
+precision equivalent to that of our function evaluations. We therefore
+propose the \textit{Forward-Automated Newton-MCSA} (FANM) method that
+utilizes all of the above components. We hypothesize that such a
+method will be competitive with Newton-Krylov methods not only from a
+convergence and timing perspective, but also relax scaling
+requirements of matrix-free methods and memory costs of both
+matrix-free and fully formed Jacobian methods to allow the application
+developer to solve problems of finer discretization and higher-order
+functional representations while maintaining a robust and efficient
+parallel implementation.
 
@@
 formulations will have different parallel performance than these
-methods, and therefore we should compare FANM perfomance to JFNK-based
+methods, and therefore we should compare FANM performance to JFNK-based
 schemes. More important than performance in this situation is the

### 2012-10-01 doc/prelim/document/chapters/parallel_mc/parallel_mc.tex
spellcheck done. draft complete

@@
-\chapter{Stochastic Solution Methods for Linear Problems}
+\chapter{Monte Carlo Solution Methods for Linear Problems}
 \label{ch:stochastic_methods}
@@
 the non-zero elements of $\ve{H}$ as those components that are zero
-will not participate in the random walk. Furthermore, it prevents an
-infinite weight from being generated in Eq~(\ref{eq:direct_weight}).
+will not participate in the random walk. Doing so prevents an
+infinite weight from being generated in Eq~(\ref{eq:direct_weight})
+and eliminates unnecessary computations.
 
@@
 minimal number of random walks in order to generate an approximation
-for $\ve{x}$. Furthermore, we also need conditions by which we may
-terminate a random walk. We do this by noticing that the factors added
-to Eq~(\ref{eq:direct_permutation_weight}) will become diminishingly
-small due to their definition in Eq~(\ref{eq:direct_weight}) such that
-their contributions to the solution estimate will become
-negligble. Therefore, we choose terminate a random walk sequence with
+for $\ve{x}$. If we are only to approximate the solution, we also need
+conditions by which we may terminate a random walk. We do this by
+noticing that the factors added to
+Eq~(\ref{eq:direct_permutation_weight}) will become diminishingly
+small due to their definition in Eq~(\ref{eq:direct_weight}) and
+therefore their contributions to the solution estimate will become
+negligible. Using this, we choose terminate a random walk sequence with
 a \textit{weight cutoff}, $W_c$, that is enforced when $W_m < W_c$ for
@@
 \begin{equation}
-  \sigma_i^2 = \sum_{\nu} P_{\nu} X_{\nu}^2 - x_i^2\:,
+  \sigma_i^2 = \Big(\sum_{\nu} P_{\nu} X_{\nu}^2\Big)_i - x_i^2\:,
   \label{eq:direct_variance_2}
@@
 \begin{equation}
-  \sigma_i^2 = \sum_{\nu} P_{\nu} \sum_{m=0}^k W_{m}^2 b_{i_m}^2 -
-  x_i^2\:.
+  \sigma_i^2 = \Big(\sum_{\nu} P_{\nu} \sum_{m=0}^k W_{m}^2
+  b_{i_m}^2\Big)_i - x_i^2\:.
   \label{eq:direct_variance_3}
 \end{equation}
-Finally, expanding the transition probabilities yields:
+Finally, expanding the transition probabilities yields the variance:
 \begin{equation}
@@
 \end{equation}
-Using this defintion for the variance, we can arrive at a more natural
-reason for enforcing $\rho(\ve{H}) < 1$. Per the Hadamard product, we can
-concatenate the summation in Eq~(\ref{eq:direct_variance_4}):
+Using this definition for the variance, we can arrive at a more natural
+reason for enforcing $\rho(\ve{H}) < 1$ for our Monte Carlo method to
+converge. Per the Hadamard product, we can concatenate the summation
+in Eq~(\ref{eq:direct_variance_4}):
 \begin{equation}
@@
 where $\ve{T} = (\ve{I}-\ve{G})^{-1}$. We can then insert $T$ back
-into the variance formulation:
+into the variance formulation for a more concise definition:
 \begin{equation}
-  \sigma^2_i = (\ve{T}\ve{b})_i - x_i^2
+  \sigma^2_i = (\ve{T}\ve{b})_i - x_i^2\:.
   \label{eq:direct_variance_5}
@@
 \end{equation}
-We then have the following inequality for relating $\ve{G}$ and
-$\ve{H}$:
+Choosing the infinity norm of the operator as defined in
+Eq~(\ref{eq:matrix_infinity_norm}), the normalized probability matrix
+will yield a norm of 1 giving the following inequality for relating
+$\ve{G}$ and $\ve{H}$:
 \begin{equation}
-  |\ve{G}| \geq \frac{|\ve{H}|^2}{|\ve{P}|}
+  |\ve{G}| \geq |\ve{H}|^2
   \label{eq:variance_norm_inequality}
@@
 Using these relations to analyze Eq~(\ref{eq:direct_variance_5}), we
-see that if $\rho(\ve{H}) > 1$, then the sum in
+see that if $\rho(\ve{G}) > 1$, then the sum in
 Eq~(\ref{eq:variance_neumann_series}) will not converge and an
 infinite variance will arise as the elements of $\ve{T}$ become
-infinite in Eq~(\ref{eq:direct_variance_5}). Therefore, we must
-restrict $\ve{H}$ with $\rho(\ve{H}) < 1$ so that our expectation
-values for the solution may have a finite variance.
+infinite in Eq~(\ref{eq:direct_variance_5}). We must restrict $\ve{G}$
+to alleviate this and therefore restrict $\ve{H}$ due to
+Eq~(\ref{eq:variance_norm_inequality}) with $\rho(\ve{H}) < 1$ so that
+our expectation values for the solution may have a finite variance.
 
@@
 \label{subsec:reusing_random_walks}
-Ji recently proposed a variation of the direct method in which the
-random walk permutations in Eq~(\ref{eq:mc_walk_permutation}) can be
-used to contribute to the solution tally in all visited states instead
-of the only starting state \citep{ji_reusing_2012}.
+A downfall of the direct method is that a random walk must be
+initiated in each state of the solution vector in which the solution
+is desired due to the estimator definition in
+Eq~(\ref{eq:direct_expectation_expansion}). Because of this, a
+solution estimate in all states of a linear system may require a
+large number of random walks to be performed. Ji and Li recently
+proposed a variation of the direct method in which the random walk
+permutations in Eq~(\ref{eq:mc_walk_permutation}) can be used to
+contribute to the solution tally in all visited states instead of the
+only starting state \citep{ji_reusing_2012}. By doing this, they
+effectively reduce the number of random walks needed to sample all
+desired states in the system.
+
+Their method is based on the observation that while a random walk
+transitions between states in the system, if every state it visited
+could be considered that starting state, then a series a contributions
+to the solution could be generated by a single sequence of random
+numbers while maintaining statistical independence between the
+samples. To demonstrate this, consider the following random walk
+sequence starting in state 2: $2 \rightarrow 3 \rightarrow 1
+\rightarrow 2 \rightarrow 4 \rightarrow 5$. We can reuse these walks
+be recognizing that we also then have a random walk sequence starting
+in state 3 that can be applied to the state 3 estimator: $3
+\rightarrow 1 \rightarrow 2 \rightarrow 4 \rightarrow 5$. We also have
+the same situation for state 1: $1 \rightarrow 2 \rightarrow 4
+\rightarrow 5$, and all future states. Using this sampling scheme , Ji
+and Li observed 1 to 2 orders of magnitude reduction over the standard
+direct method in the number histories required to achieve a specified
+variance in their test problems.
 
@@
 transpose of $\ve{H}$\footnote{This is sometimes referred to as the
-  adjoint form for real valued matrices}. We then follow the same
+  adjoint form for real-valued matrices}. We then follow the same
 procedure as the direct method for forming the probability and weight
@@
 where the Kronecker delta indicates that the tally contributes only in
-the current state and $b_{i_0}$ is that sampled source starting
+the current state and $b_{i_0}$ is the sampled source starting
 state. Note here that the estimator in
@@
 
-Given these ideas, we can then generate an algorithm for implementing
-the adjoint method:
-\begin{algorithm}[htpb!]
-  \caption{Adjoint Monte Carlo}
-  \label{alg:adjoint_mc}
-  \begin{algorithmic}
-    \State $\ve{r}_0 := \ve{b}-\ve{A}\ve{x}_0$
-  \end{algorithmic}
-\end{algorithm}
-
 \section{Sequential Monte Carlo}
 \label{sec:sequential_mc}
-The direct and adjoint methods described are limited by a convergence
-rate of $1/\sqrt{N}$ by the Central Limit Theorem where $N$ is the
-number of random walk permutations. In 1962, Halton presented a
-residual Monte Carlo method that moves towards exponential convergence
-rates \citep{halton_sequential_1962} and further refined his work some
-years later \citep{halton_sequential_1994}. Applications of his work
-by the transport community have confirmed convergence rates on the
-order of of $exp(-N)$
-\citep{evans_residual_2003}. In
+The direct and adjoint Neumann-Ulam methods described are limited by a
+convergence rate of $1/\sqrt{N}$ by the Central Limit Theorem where
+$N$ is the number of random walk permutations. In 1962, Halton
+presented a residual Monte Carlo method that moves towards exponential
+convergence rates \citep{halton_sequential_1962} and further refined
+his work some years later \citep{halton_sequential_1994}. Applications
+of his work by the transport community have confirmed convergence
+rates on the order of of $exp(-N)$ \citep{evans_residual_2003}. In
 much the same way as projection methods, Halton's method, sequential
 Monte Carlo, utilizes the adjoint Monte Carlo solver as a means of
-driving down the residual vector. He proposed the following iterative
-scheme as a solution to Eq~(\ref{eq:linear_problem})\:
+directly reducing the residual vector. He proposed the following
+iterative scheme as a solution to Eq~(\ref{eq:linear_problem})\:
 \begin{subequations}
@@
     \ve{r}^k = \ve{b} - \ve{A}\ve{x}^k\:,\\  
-    \ve{A}\boldsymbol{\delta}^{k+1} = \ve{r}^{k}\:,\\
-    \ve{x}^{k+1} = \ve{x}^k + \boldsymbol{\delta}^{k+1}\:,
+    \ve{A}\boldsymbol{\delta}^{k} = \ve{r}^{k}\:,\\
+    \ve{x}^{k+1} = \ve{x}^k + \boldsymbol{\delta}^{k}\:,
   \end{gather}
@@
 visible in that we have now broken the binding of the convergence rate
-to the Central Limit Thereom. Here, the Monte Carlo solve is used to
-produce a correction from the residual, analagous to using the
+to the Central Limit Thereom. Here, the Monte Carlo solver is used to
+produce a correction from the residual, analogous to using the
 residual to extract a correction from the search subspace in a
@@
 Using the ideas of Halton, Evans and Mosher recently developed a Monte
-Carlo solution that was not prohibited severely by the quality of the
-initial guess for the system \citep{evans_monte_2009} and later
+Carlo solution method that was not prohibited severely by the quality
+of the initial guess for the system \citep{evans_monte_2009} and later
 applied it more rigorously as a solution mechanism for the radiation
-diffusion equation \citep{evans_monte_2012}, achieving identical
-numerical results as and marginally better performance than
-conventional Krylov solvers. Their approach was instead to use
-residual Monte Carlo as a synthetic acceleration for a stationary
-method. To derive this method, we begin by splitting the operator in
-Eq~(\ref{eq:linear_problem})
+diffusion equation \citep{evans_monte_2012}. With their new methods,
+they achieved identical numerical results as and marginally better
+performance than conventional Krylov solvers. Their approach was
+instead to use residual Monte Carlo as a synthetic acceleration for a
+stationary method. To derive this method, we begin by splitting the
+operator in Eq~(\ref{eq:linear_problem})
 \begin{equation}
@@
 However, $\ve{A}$ is only approximately inverted by our numerical
-methods and therfore we instead pose an iterative scheme in which the
+methods and therefore we instead pose an iterative scheme in which the
 Monte Carlo solvers are used to invert the operator. The
@@
 
-\subsection{Preconditioning Stochastic Methods}
+\subsection{Preconditioning MCSA}
 \label{sec:stochastic_preconditioning}
@@
 of the iteration matrix be less than one. To achieve this, a Jacobi
-precondtioner, a form of left preconditioning, is used such that the
+preconditioner, a form of left preconditioning, is used such that the
 preconditioning matrix $\ve{M}$ is:
@@
 achieved. Therefore, we will develop and implement parallel algorithms
-for these methods levaraging both the knowledge gained from the
+for these methods leveraging both the knowledge gained from the
 general parallel implementations of Krylov methods and modern parallel
@@
 their scheme employed the similar parallel algorithms for particle
-transport but a certain amount of overlap between domains was used to
-decrease the number of particles leaving the domain. In addition,
-Wagner utilized a level of replication of the domain such that the
-domain was only decomposed on $O(100)$ processors and if replicated
-$O(1000)$ times achieves simulation on $O(100,000)$ processors, thus
-providing spatial and particle parellelism. Each collection of
-processors that constitutes a representation of the entire domain is
-referred to as a set, and within a set overlap occurs among its
-subdomains. The original motivation was to decompose the domain in a
-way that it remained in a physical cabinet in a large distributed
-machine, thus reducing latency costs during communication. A multiple
-set scheme is also motivated by the fact that communication only
-occurs within a set, limiting communications during the transport
-procedure to a group of $O(100)$ processors, a number that was shown
-to have excellent parallel efficiencies in Brunner's work and
-therefore will scale well in this algorithm. The overlapping domains
-within each set also demonstrated reduced communication costs. On each
-processor, the source is sampled in the local domain that would exist
-if no overlap was used while tallies can be made over the entire
-overlapping domain. To demonstrate this, consider the example adapted
-from Mervin's work with Wagner and others in the same area
-\citep{mervin_variance_2012} and presented in
-Figure~\ref{fig:msod_example}.
+transport but a certain amount of overlap between adjacent domains was
+used to decrease the number of particles leaving the local domain. In
+addition, Wagner utilized a level of replication of the domain such
+that the domain was only decomposed on $O(100)$ processors and if
+replicated $O(1000)$ times achieves simulation on $O(100,000)$
+processors, thus providing spatial and particle parallelism. Each
+collection of processors that constitutes a representation of the
+entire domain is referred to as a set, and within a set overlap occurs
+among its subdomains. The original motivation was to decompose the
+domain in a way that it remained in a physical cabinet in a large
+distributed machine, thus reducing latency costs during
+communication. A multiple set scheme is also motivated by the fact
+that communication only occurs within a set, limiting communications
+during the transport procedure to a group of $O(100)$ processors, a
+number that was shown to have excellent parallel efficiencies in
+Brunner's work and therefore will scale well in this algorithm. The
+overlapping domains within each set also demonstrated reduced
+communication costs. On each processor, the source is sampled in the
+local domain that would exist if no overlap was used while tallies can
+be made over the entire overlapping domain.
+
+To demonstrate this, consider the example adapted from Mervin's work
+with Wagner and others in the same area \citep{mervin_variance_2012}
+and presented in Figure~\ref{fig:msod_example}.
 \begin{figure}[htpb!]
@@
 \end{figure}
-In this example, 3 particle histories are presented eminating from the
+In this example, 3 particle histories are presented emanating from the
 blue region of interest. Starting with particle A, if no domain
@@
 incurring a large latency cost for a single particle. Finally, with
-particle B we observe that 0.5 domain overlap will still not elminate
+particle B we observe that 0.5 domain overlap will still not eliminate
 all communications. However, if 1 domain overlap were used, the entire
@@
 must be made for situations where this is not the case. Given the
-stochastic nature of the problem and lack of a globally homogoneous
+stochastic nature of the problem and lack of a globally homogeneous
 domain, parallel Monte Carlo simulations are inherently load
@@
 higher work loads and some to have smaller work loads than the initial
-conditions. Consider the example in
-Figure~\ref{fig:procassini_example} adapted from Procassini's work.
+conditions.
+
+Consider the example in Figure~\ref{fig:procassini_example} adapted
+from Procassini's work.
 \begin{figure}[htpb!]
@@
 
-\subsubsection{Reproducable Domain Decomposed Results}
-\label{subsubsec:reproducable_mc}
+\subsubsection{Reproducible Domain Decomposed Results}
+\label{subsubsec:reproducible_mc}
 The 2006 work of Brunner is notable in that the Monte Carlo codes used
@@
 \citep{gentile_obtaining_2005}. In Gentile's work, a procedure is
-given for obtaining results reproducable to machine precision for an
+given for obtaining results reproducible to machine precision for an
 arbitrary number of processors and domains. Differences can arise from
@@
   \caption{\textbf{Gentile's example illustrating how domain
-      decomposition can create reproducability issues in Monte Carlo.}
+      decomposition can create reproducibility issues in Monte Carlo.}
     \textit{Both particles A and B start in zone 1 on processor
@@
 born in zone 1 and is transported to zone 2 where a scattering event
-occurs. Concerning the first reproducability issue, if the same
+occurs. Concerning the first reproducibility issue, if the same
 sequence of random numbers is not used to compute the trajectory from
@@
 of floating point operations, we cannot expect these to have an
-identical result to machine precesion as round-off approximations will
+identical result to machine precision as round-off approximations will
 differ resulting in non-commutative addition.
 
-Their solutions to these problems are elegant in that they require a
-minimal amount of modification to be applied to the Monte Carlo
-algorithm. To solve the first issue, in order to ensure each particle
-maintains an invariant random number sequence that determines its
-behavior regardless of domain decomposition, each particle is a random
-number seed that describes its current state upon entering the domain
-of a new processor. These seeds are derived from the actual geometric
-location of the particle such that it is decomposition
+Procassini's solutions to these problems are elegant in that they
+require a minimal amount of modification to be applied to the Monte
+Carlo algorithm. To solve the first issue, in order to ensure each
+particle maintains an invariant random number sequence that determines
+its behavior regardless of domain decomposition, each particle is a
+random number seed that describes its current state upon entering the
+domain of a new processor. These seeds are derived from the actual
+geometric location of the particle such that it is decomposition
 invariant. Non-commutative floating point operations are solved by
@@
 source Monte Carlo transport problems is the content of the Markov
-chains that are generated. These transitions represeneted by these
-chains are bound by probabilities and weights and are initiated by the
+chains that are generated. The transitions represented by these chains
+are bound by probabilities and weights and are initiated by the
 sampling of a source. In the context of transport problems, those
-transitions represent events such as pariticle scattering and
-absorbition with probabilities that are determined by physical data in
+transitions represent events such as particle scattering and
+absorption with probabilities that are determined by physical data in
 the form of cross sections. For stochastic matrix inversion, those
@@
 system (and therefore the physical domain which they represent) and
-their probabilites are defined by the coefficients of those
+their probabilities are defined by the coefficients of those
 equations. Ultimately, we tally the contributions to generate
@@
 that correlates to a statistical uncertainty of 14.1\% in the
-correction, yet as compared to convential Krylov solvers, MCSA
+correction, yet as compared to conventional Krylov solvers, MCSA
 achieved identical numerical results. If good convergence and
@@
 non-local states can be ignored and thus eliminate all communication
-in the parallel Monte Carlo adjoint solver and create an embarassingly
+in the parallel Monte Carlo adjoint solver and create an embarrassingly
 parallel method. It will be important to determine if the expectation
@@
 achieve with a production scale linear solver. Typically, memory is at
-a premimium and therefore the more distinct domains available in the
+a premium and therefore the more distinct domains available in the
 decomposition, the spatially finer and/or the numerically more
@@
 that the Krylov subspace maintained by GMRES can become large, do we
-benifit from a memory standpoint with an MCSA scheme in that no
-subspace is required? Does this benefit outweight the fact that the
+benefit from a memory standpoint with an MCSA scheme in that no
+subspace is required? Does this benefit outweigh the fact that the
 linear operator must be explicitly formed in order to build the

### 2012-10-01 doc/prelim/document/chapters/research_proposal/research_proposal.tex
spellcheck done. draft complete

@@
 proposed to drive their development and quantify their merits and
-feasiblity as compared to more conventional methods for linear and
+feasibility as compared to more conventional methods for linear and
 nonlinear problems. We seek to rigorously verify both of these methods
@@
 effectiveness of our new methods in the context of real-world,
-production level applications.
+production level applications for nuclear reactor analysis.
 
@@
 readily available in the form of production software implementations
-and will demonstrate scalability in high performance computing
+and that will demonstrate scalability in high performance computing
 environments. This environment will be based on the mathematical
@@
 In our experimental framework, we will formulate all of our problems
-using the weak form of the differential equations. Doing so allows us
-to leverage modern finite element assembly engine software with
-embedded automatic differentiation technology. Using the Panzer
-library in the Trilinos sceintific computing framework
+using the weak form of the differential equations as in
+Eq~(\ref{eq:fe_weak_form}). Doing so allows us to leverage modern
+finite element assembly engine software with embedded automatic
+differentiation technology. Using the Panzer library in the Trilinos
+scientific computing framework
 \citep{notz_graph-based_2010,heroux_overview_2005}, function
@@
 because the random walk sequence contributes to the state in which it
-currently resides, fewer histories are neccessary to compute
+currently resides, fewer histories are necessary to compute
 contributions from all the nonzero parts of the source. Initially,
@@
 problem size. Figure~\ref{fig:poisson_convergence} gives the results
-of these computations. First, it is worthy to note on the semilog plot
+of these computations. First, it is worthy to note on the semi-log plot
 that we are indeed achieving the expected exponential convergence from
@@
 In the results, we see the characteristic parabolic solution profiles
-of the heat equation corresponding corresponding to the correct
-Dirichlet condition. Based on this work, a finite element formulation
-generated by the Panzer framework can be readily solved with a general
-MCSA implementation. In addition, the above problem can already be
-solved in parallel with conventional Krylov methods. Therefore, once a
+of the heat equation corresponding to the correct Dirichlet
+condition. Based on this work, a finite element formulation generated
+by the Panzer framework can be readily solved with a general MCSA
+implementation. In addition, the above problem can already be solved
+in parallel with conventional Krylov methods. Therefore, once a
 parallel MCSA implementation is complete, this framework can be used
@@
 only a few small additions to the framework, providing the remainder
-of the toolset we need for our analysis.
+of the tool set we need for our analysis.
 
@@
 strategies for the Monte Carlo methods and introduce bias or
-non-reproducable results into our solution estimators, these
+non-reproducible results into our solution estimators, these
 verification problems will allow us to determine if that cost is
@@
 solutions for the 2-dimensional, steady, incompressible Navier-Stokes
-equations on a recitilinear grid in much the same way as Shadid and
+equations on a rectilinear grid in much the same way as Shadid and
 Pawlowski's work on Newton-Krylov methods for the solution of these
@@
 and asymmetries that we are seeking for qualification of the FANM
-method. Further, physical parameters withing these equations can be
+method. Further, physical parameters within these equations can be
 tuned to enhance the nonlinearities. We will then apply these
@@
 \label{subsubsec:natural_convection_cavity}
-In 1983 benchmark solution for the natural convection of air in a
+In 1983 a benchmark solution for the natural convection of air in a
 square cavity was published \citep{de_vahl_davis_natural_1983} as
@@
       set on the top and bottom of the Cartesian grid. The temperature
-      gradients will cause bouyancy-driven flow. Zero velocity
+      gradients will cause buoyancy-driven flow. Zero velocity
       Dirichlet conditions are set on each boundary. No thermal source
@@
 heat flow is allowed out of the top and bottom of the square with a
-zero Neumann condition specified. Bouyancy driven flow is generated by
+zero Neumann condition specified. Buoyancy driven flow is generated by
 the temperature gradient from the cold and hot Dirichlet conditions on
@@
 flow to introduce higher Reynolds numbers into the system, providing
-more intertial force to overcome the viscous forces in the fluid. The
+more inertial force to overcome the viscous forces in the fluid. The
 setup for this problem is equally simple, containing only the
@@
 in obtaining a solution is dictated by the Reynolds number of the
-fluid. In Shadid's work, a $20 \times 400$ non-square recitlinear grid
+fluid. In Shadid's work, a $20 \times 400$ non-square rectilinear grid
 was used to discretize the domain with Reynolds number up to
@@
 as a testing environment, we propose a set of experiments that will be
-used to characterize several qualities of these stochastic methods. At
-the end of Chapter~\ref{ch:stochastic_methods}, we posed several
-research questions related to the parallelization of the adjoint
-Neumann-Ulam method and the MCSA method while in
+used to characterize several qualities of parallel Monte Carlo
+solvers. At the end of Chapter~\ref{ch:stochastic_methods}, we posed
+several research questions related to the parallelization of the
+adjoint Neumann-Ulam method and the MCSA method while in
 Chapter~\ref{ch:nonlinear_problem} we posed questions related to the
-the FANM method. We will generate numerical experiments from the these
+the FANM method. We will generate numerical experiments from these
 questions by dividing them into 4 categories: domain overlap studies
@@
 structures and the mean free path of neutrons in the system. We
-therefore seek analgous measurements for the general linear
+therefore seek analogous measurements for the general linear
 system. These measurements will characterize how far from its starting
@@
 likely be required to solve future resiliency issues and may even by
-required by this work in order to achieve scalability, it parallel
+required by this work in order to achieve scalability, its parallel
 performance must also be measured. We will therefore measure the
@@
 
-\subsection{Parallel Performance and Numerical Accuracy Studies for MCSA
-  Method}
+\subsection{Parallel Performance and Numerical Accuracy Studies for
+  the MCSA Method}
 \label{subsec:mcsa_studies}
@@
 experiment with the way we combine it with the Neumann-Ulam solver. As
-disucussed in \S~\ref{subsec:parallel_adjoint}, previous work with
+discussed in \S~\ref{subsec:parallel_adjoint}, previous work with
 MCSA used a correction generated by the Neumann-Ulam method with a
@@
 be ignored. This will result in a biased estimator for the MCSA
-correction that cannont be reproduced, however, these studies will aim
-to determine if that bias and lack of reproducablity affects the
-numerical results and reproducability of the MCSA method, yielding a
+correction that cannot be reproduced, however, these studies will aim
+to determine if that bias and lack of reproducibility affects the
+numerical results and reproducibility of the MCSA method, yielding a
 critical outcome for this work. Furthermore, we will use the
 experimental framework to measure the convergence rates for
-nonsymmetric systems and memory usage of the MCSA algorithm and
+non-symmetric systems and memory usage of the MCSA algorithm and
 compare them to conventional Krylov methods using the same framework.
 
-\subsection{Parallel Performance and Numerical Accuracy Studies for FANM
-  Method}
+\subsection{Parallel Performance and Numerical Accuracy Studies for
+  the FANM Method}
 \label{subsec:fanm_studies}
@@
 modeling of nuclear reactors
-\citep{u.s._department_of_energy_casl_2011}. Their effors are focused
+\citep{u.s._department_of_energy_casl_2011}. Their efforts are focused
 on enabling nuclear reactor analysis and development using today's
 supercomputing technology to achieve higher power uprates, higher fuel
-burnup, and predictive accident scenario analysis. As a means of
+burn-up, and predictive accident scenario analysis. As a means of
 driving their research and development, CASL has worked with the
@@
 implementing them within the Drekar framework and performing the
-equivalent of its largest GTRF mutliphysics problem to date, most
+equivalent of its largest GTRF multiphysics problem to date, most
 likely to be a fuel-assembly sized problem. Doing so will allow us to
@@
 We have proposed the research and development of novel, massively
-parallel stochastic methods for the solution of linear and nonlinear
+parallel Monte Carlo methods for the solution of linear and nonlinear
 problems. These new methods aim to enhance the solution of multiple
@@
 this work, we have defined a multiphysics framework based on the
-finite element method and have demonstrated that these stochastic
+finite element method and have demonstrated that these Monte Carlo
 methods can be used effectively to solve the linear systems generated
@@
 
-To guide our development of parallel stochastic methods, we have
+To guide our development of parallel Monte Carlo methods, we have
 devised several classes of numerical experiments using our
@@
 conventional methods currently leveraged within Drekar. With the
-potential to improve computational performance in these environments,
-these new methods have the potential to help meet the high fidelty
-modeling and simulation needs of the nuclear community.
+potential to improve computational performance in production
+environments such as Drekar, these new methods have the potential to
+help meet the high fidelity modeling and simulation needs of the
+nuclear community.

### 2012-10-01 src/Chimera_RNGTraits.hpp
adding RNGTraits

@@ -1,0 +1,122 @@
+//---------------------------------------------------------------------------//
+/*
+  Copyright (c) 2012, Stuart R. Slattery
+  All rights reserved.
+
+  Redistribution and use in source and binary forms, with or without
+  modification, are permitted provided that the following conditions are
+  met:
+
+  *: Redistributions of source code must retain the above copyright
+  notice, this list of conditions and the following disclaimer.
+
+  *: Redistributions in binary form must reproduce the above copyright
+  notice, this list of conditions and the following disclaimer in the
+  documentation and/or other materials provided with the distribution.
+
+  *: Neither the name of the University of Wisconsin - Madison nor the
+  names of its contributors may be used to endorse or promote products
+  derived from this software without specific prior written permission.
+
+  THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
+  "AS IS" AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
+  LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
+  A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE COPYRIGHT
+  HOLDER OR CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
+  SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
+  LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
+  DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
+  THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
+  (INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
+  OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
+*/
+//---------------------------------------------------------------------------//
+// \file Chimera_RNGTraits.hpp
+// \author Stuart R. Slattery
+// \brief Traits definition for random number generators.
+//---------------------------------------------------------------------------//
+
+#ifndef Chimera_RNGTRAITS_HPP
[…]

### 2012-10-01 src/Chimera_BoostRNG.hpp
adding boost rng traits implementation

@@ -31,81 +31,42 @@
   OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
 */
 //---------------------------------------------------------------------------//
-// \file Chimera_RNGTraits.hpp
+// \file Chimera_BoostRNG.hpp
 // \author Stuart R. Slattery
-// \brief Traits definition for random number generators.
+// \brief RNGTraits implementation for Boost Psuedo Random Number Generators.
 //---------------------------------------------------------------------------//
 
-#ifndef Chimera_RNGTRAITS_HPP
-#define Chimera_RNGTRAITS_HPP
+#ifndef Chimera_BOOSTRNG_HPP
+#define Chimera_BOOSTRNG_HPP
+
+#include "Chimera_RNGTraits.hpp"
 
 namespace Chimera
 {
 namespace Solvers
 {
 //---------------------------------------------------------------------------//
-/*!
- * \brief Dummy struct. If a type does not create a specialization this will
- * not compile.
- */
-template<typename UndefinedRNGType>
-struct UndefinedRNGTraits
+template<>
+template<class BoostRNG>
+class RNGTraits<BoostRNG>
 {
-    static inline UndefinedRNGType notDefined() 
-    { return UndefinedRNGType::this_type_is_missing_a_specialization(); }
-};
+    typedef RNGType        BoostRNG;
+    typedef result_type    BoostRNG::result_type;
 
-//---------------------------------------------------------------------------//
-/*!
-  \class RNGTraits
-  \brief Random number generator traits definitions.
+    static inline result_type generate( const BoostRNG& boost_rng )
+    { return boost_rng(); }
 
-  RNGTraits provide access to random number generators.
-*/
-//---------------------------------------------------------------------------//
-template<typename RNGType>
-class RNGTraits
-{
-  public:
+    static inline void reset( const BoostRNG& boost_rng )
+    { boost_rng.reset(); }
 
-    //@{
-    //! Typedef for rng type.
-    typedef RNGType rng_type;
+    static inline void setSeed( const BoostRNG& boost_rng, const unsigned int seed )
+    { boost_rng.seed( seed ); }
 
-    //! Typedef for result type.
-    typedef RNGType::result_type result_type;
-    //@}
+    static inline result_type min( const BoostRNG& boost_rng )
+    { return boost_rng.min(); }
 
-    /*! 
-     * \brief Generate a random number.
-     */
[…]

### 2012-10-01 src/Chimera_BoostRNG.hpp
finished rng traits and boost implementations and tests

@@ -41,31 +41,65 @@
 
 #include "Chimera_RNGTraits.hpp"
 
+#include <boost/random/mersenne_twister.hpp>
+
 namespace Chimera
 {
 namespace Solvers
 {
 //---------------------------------------------------------------------------//
+// Specialization for mt19937
+//---------------------------------------------------------------------------//
 template<>
-template<class BoostRNG>
-class RNGTraits<BoostRNG>
+class RNGTraits<boost::mt19937>
 {
-    typedef RNGType        BoostRNG;
-    typedef result_type    BoostRNG::result_type;
+  public:
 
-    static inline result_type generate( const BoostRNG& boost_rng )
+    typedef boost::mt19937                  rng_type;
+    typedef boost::mt19937::result_type     result_type;
+
+    static inline boost::mt19937 create()
+    { return boost::mt19937(); }
+
+    static inline result_type generate( boost::mt19937& boost_rng )
     { return boost_rng(); }
 
-    static inline void reset( const BoostRNG& boost_rng )
-    { boost_rng.reset(); }
-
-    static inline void setSeed( const BoostRNG& boost_rng, const unsigned int seed )
+    static inline void setSeed( boost::mt19937& boost_rng, 
+				const result_type seed )
     { boost_rng.seed( seed ); }
 
-    static inline result_type min( const BoostRNG& boost_rng )
+    static inline result_type min( boost::mt19937& boost_rng )
     { return boost_rng.min(); }
 
-    static inline result_type max( const BoostRNG& boost_rng )
+    static inline result_type max( boost::mt19937& boost_rng )
+    { return boost_rng.max(); }
+};
+
+//---------------------------------------------------------------------------//
+// Specialization for mt11213b
+//---------------------------------------------------------------------------//
+template<>
+class RNGTraits<boost::mt11213b>
+{
+  public:
+
+    typedef boost::mt11213b                 rng_type;
+    typedef boost::mt11213b::result_type    result_type;
+
+    static inline boost::mt11213b create()
+    { return boost::mt11213b(); }
+
+    static inline result_type generate( boost::mt11213b& boost_rng )
+    { return boost_rng(); }
+
+    static inline void setSeed( boost::mt11213b& boost_rng, 
+				const result_type seed )
+    { boost_rng.seed( seed ); }
+
+    static inline result_type min( boost::mt11213b& boost_rng )
+    { return boost_rng.min(); }
+
+    static inline result_type max( boost::mt11213b& boost_rng )
     { return boost_rng.max(); }
 };
 

### 2012-10-01 src/Chimera_RNGTraits.hpp

@@ -63,7 +63,7 @@ struct UndefinedRNGTraits
   RNGTraits provide access to random number generators.
 */
 //---------------------------------------------------------------------------//
-template<typename RNGType>
+template<class RNGType>
 class RNGTraits
 {
   public:
@@ -73,38 +73,37 @@ class RNGTraits
     typedef RNGType rng_type;
 
     //! Typedef for result type.
-    typedef RNGType::result_type result_type;
+    typedef typename RNGType::result_type result_type;
     //@}
+
+    /*!
+     * \brief Create a random number generator in the base state.
+     */
+    static inline rng_type create()
+    { UndefinedRNGTraits<RNGType>::notDefined(); return 0; }
 
     /*! 
      * \brief Generate a random number.
      */
-    static inline result_type generate( const RNGType& rng )
+    static inline result_type generate( RNGType& rng )
     { UndefinedRNGTraits<RNGType>::notDefined(); return 0; }
-
-    /*!
-     * \brief Reset the state of the random number generator to the default
-     * value.
-     */
-    static inline void reset( const RNGType& rng )
-    { UndefinedRNGTraits<RNGType>::notDefined(); }
 
     /*!
      * \brief Set the current seed of the random number generator.
      */
-    static inline void setSeed( const RNGType& rng, const unsigned int seed )
+    static inline void setSeed( RNGType& rng, const result_type seed )
     { UndefinedRNGTraits<RNGType>::notDefined(); }
 
     /*!
      * \brief Lower bound of random range.
      */
-    static inline result_type min( const RNGType& rng )
+    static inline result_type min( RNGType& rng )
     { UndefinedRNGTraits<RNGType>::notDefined(); return 0; }
 
     /*!
      * \brief Upper bound of random range.
      */
-    static inline result_type max( const RNGType& rng )
+    static inline result_type max( RNGType& rng )
     { UndefinedRNGTraits<RNGType>::notDefined(); return 0; }
 };
 

### 2012-10-01 test/tstBoostRNG.cpp

@@ -1,0 +1,120 @@
+//---------------------------------------------------------------------------//
+/*!
+ * \file tstBoostRNG.cpp
+ * \author Stuart R. Slattery
+ * \brief Boost random number generator unit tests.
+ */
+//---------------------------------------------------------------------------//
+
+#include <iostream>
+#include <vector>
+#include <cmath>
+#include <cstdlib>
+#include <sstream>
+#include <algorithm>
+#include <cassert>
+
+#include <Chimera_RNGTraits.hpp>
+#include <Chimera_BoostRNG.hpp>
+
+#include <boost/random/mersenne_twister.hpp>
+
+#include <Teuchos_UnitTestHarness.hpp>
+#include <Teuchos_DefaultComm.hpp>
+#include <Teuchos_CommHelpers.hpp>
+#include <Teuchos_RCP.hpp>
+#include <Teuchos_ArrayRCP.hpp>
+#include <Teuchos_Array.hpp>
+#include <Teuchos_OpaqueWrapper.hpp>
+#include <Teuchos_TypeTraits.hpp>
+#include <Teuchos_Tuple.hpp>
+
+//---------------------------------------------------------------------------//
+// Tests
+//---------------------------------------------------------------------------//
+TEUCHOS_UNIT_TEST( BoostRNG, boost_mt11213b_test )
+{
+    using namespace Chimera::Solvers;
+
+    boost::mt11213b generator_1;
+    boost::mt11213b generator_2 = RNGTraits<boost::mt11213b>::create();
+
+    TEST_ASSERT( generator_1.min() == 
+		 RNGTraits<boost::mt11213b>::min( generator_2 ) );
+
+    TEST_ASSERT( generator_1.max() == 
+		 RNGTraits<boost::mt11213b>::max( generator_2 ) );
+
+    RNGTraits<boost::mt11213b>::result_type first_val_1 = generator_1();
+    RNGTraits<boost::mt11213b>::result_type first_val_2 = 
+	RNGTraits<boost::mt11213b>::generate( generator_2 );
+
+    for ( int i = 0; i < 10000; ++i )
+    {
+	TEST_ASSERT( generator_1() == 
+		     RNGTraits<boost::mt11213b>::generate( generator_2 ) );
+    }
+
+    RNGTraits<boost::mt11213b>::result_type new_seed = 184839;
+    generator_1.seed( new_seed );
[…]

### 2012-10-01 src/Chimera_BoostRNG.hpp
comments in boost rng

@@ -48,7 +48,7 @@ namespace Chimera
 namespace Solvers
 {
 //---------------------------------------------------------------------------//
-// Specialization for mt19937
+// Specialization for 64 bit mt19937
 //---------------------------------------------------------------------------//
 template<>
 class RNGTraits<boost::mt19937>
@@ -76,7 +76,7 @@ class RNGTraits<boost::mt19937>
 };
 
 //---------------------------------------------------------------------------//
-// Specialization for mt11213b
+// Specialization for 32 bit mt11213b
 //---------------------------------------------------------------------------//
 template<>
 class RNGTraits<boost::mt11213b>

### 2012-10-02 doc/prelim/document/chapters/introduction/introduction.tex
fixed typo in intro

@@
 Denovo radiation transport code developed at Oak Ridge National
-Laboratory \citep{evans_denovo:_2010}, computes trillions of unknowns
+Laboratory \citep{evans_denovo:_2010}, compute trillions of unknowns
 for full core reactor analysis on $O(\sn{1}{9})$ element meshes and

### 2012-10-02 doc/prelim/document/chapters/introduction/introduction.tex
grammar fix in intro

@@
 \citep{u.s._department_of_energy_resilient_2012}. Based on the
-language in this call for proposals, we can identify key issues that a
-set of robust, massively parallel stochastic solvers could provide a
-solution for. As machines begin to operate at hundreds of petaflops
-peak performance and beyond, restrictions on power availability will
-require incredibly high levels of concurrency. Furthermore, this drop
-in power consumption will mean increased pressure on memory as memory
-per node is expected to stagnate while cores per node is expected to
-increase. As the number of cores increases, their clock speed is
-expected to stagnate or even decrease to further reduce power
-consumption and manufacturing costs.
+language in this call for proposals, we can identify key issues for
+which a set of robust, massively parallel stochastic solvers could
+provide a solution. As machines begin to operate at hundreds of
+petaflops peak performance and beyond, restrictions on power
+availability will require incredibly high levels of
+concurrency. Furthermore, this drop in power consumption will mean
+increased pressure on memory as memory per node is expected to
+stagnate while cores per node is expected to increase. As the number
+of cores increases, their clock speed is expected to stagnate or even
+decrease to further reduce power consumption and manufacturing costs.
 

### 2012-10-02 doc/prelim/document/chapters/introduction/introduction.tex
prelim doc updates

--- 2012-10-02 doc/prelim/document/chapters/introduction/introduction.tex
@@
 new hardware. Considering these issues, this preliminary report
-proposes the development of a massively parallel stochastic method for
-linear problems and a novel stochastic method to advance solution
+proposes the development of a massively parallel Monte Carlo method
+for linear problems and a novel Monte Carlo method to advance solution
 techniques for nonlinear problems.
@@
 Consortium for Advanced Simulation of Light Water Reactors (CASL)
-modeling and simulation hub using the Drekar multiphysics code
+modeling and simulation hub. CASL used the Drekar multiphysics code
 developed at Sandia National Laboratories
-\citep{pawlowski_drekar_2012}. Using Drekar, multiphysics simulations
-have been performed with fully consistent methods for the solution of
-nonlinear systems using meshes of $O(\sn{1}{9})$ elements leveraging
-$O(100,000)$ cores on leadership class machines. Neutronics components
-to be implemented in CASL for multiphysics analysis, such as the
-Denovo radiation transport code developed at Oak Ridge National
+\citep{pawlowski_drekar_2012} for modeling goals in
+grid-to-rod-fretting analysis and will use a similar coupled physics
+structure for future departure from nucleate boiling analysis with
+comparison to experimental data. Using Drekar, multiphysics
+simulations have been performed with fully consistent methods for the
+solution of nonlinear systems using meshes of $O(\sn{1}{9})$ elements
+leveraging $O(100,000)$ cores on leadership class machines. Neutronics
+components to be implemented in CASL for multiphysics analysis, such
+as the Denovo radiation transport code developed at Oak Ridge National
 Laboratory \citep{evans_denovo:_2010}, compute trillions of unknowns
@@
 \citep{u.s._department_of_energy_resilient_2012}. Based on the
-language in this call for proposals, we can identify key issues that a
-set of robust, massively parallel stochastic solvers could provide a
-solution for. As machines begin to operate at hundreds of petaflops
-peak performance and beyond, restrictions on power availability will
-require incredibly high levels of concurrency. Furthermore, this drop
-in power consumption will mean increased pressure on memory as memory
-per node is expected to stagnate while cores per node is expected to
-increase. As the number of cores increases, their clock speed is
-expected to stagnate or even decrease to further reduce power
-consumption and manufacturing costs.
+language in this call for proposals, we can identify key issues for
+which a set of robust, massively parallel Monte Carlo solvers could
+provide a solution. As machines begin to operate at hundreds of
+petaflops peak performance and beyond, restrictions on power
+availability will require incredibly high levels of
+concurrency. Furthermore, this drop in power consumption will mean
+increased pressure on memory as memory per node is expected to
+stagnate while cores per node is expected to increase. As the number
+of cores increases, their clock speed is expected to stagnate or even
+decrease to further reduce power consumption and manufacturing costs.
 
@@
   kelley_iterative_1995}. In between these two areas are a not widely
-known group of stochastic methods for solving sparse linear systems
+known group of Monte Carlo methods for solving sparse linear systems
 \citep{forsythe_matrix_1950, hammersley_monte_1964,
   halton_sequential_1962, halton_sequential_1994}. In recent years,
-these methods have been further developed for transport problems in
-the form of Monte Carlo Synthetic-Acceleration (MCSA)
-\citep{evans_residual_2003, evans_monte_2009, evans_monte_2012} but
-have yet to be applied to more general sparse linear systems commonly
-generated by the computational physics community. Compared to other
-methods in this regime, MCSA offers three attractive qualities; (1)
-the linear problem operator need not be symmetric or
-positive-definite, thereby reducing preconditioning complexity, (2)
-the stochastic nature of the solution method provides a natural
-solution to the issue of resiliency, and (3) is amenable to
-parallelization using modern methods developed by the transport
-community \citep{wagner_hybrid_2010}. The development of MCSA as a
-general linear solver and the development of a parallel MCSA method
-will be new and unique features of this work, providing a framework
-with which other issues such as resiliency may be addressed in the
-future.
+these methods have been further developed for radiation transport
+problems in the form of Monte Carlo Synthetic-Acceleration (MCSA)
+\citep{evans_monte_2009, evans_monte_2012} but have yet to be applied
+to more general sparse linear systems commonly generated by the
+computational physics community. Compared to other methods in this
+regime, MCSA offers three attractive qualities; (1) the linear problem
+operator need not be symmetric or positive-definite, thereby reducing
+preconditioning complexity, (2) the stochastic nature of the solution
+method provides a natural solution to the issue of resiliency, and (3)
+is amenable to parallelization using modern methods developed by the
+transport community \citep{wagner_hybrid_2010}. The development of
+MCSA as a general linear solver and the development of a parallel MCSA
+method will be new and unique features of this work, providing a
+framework with which other issues such as resiliency may be addressed
+in the future.
 

### 2012-10-02 doc/prelim/document/chapters/research_proposal/research_proposal.tex
prelim doc updates

@@
 Our numerical experiments will provide us with a set of metrics giving
-best-performance stochastic solver parameters for problems of
+best-performance Monte Carlo solver parameters for problems of
 different types. As an additional bonus, these experiments will let us
-further develop new parallel implementations of the stochastic methods
+further develop new parallel implementations of the Monte Carlo methods
 in order to improve their performance and scalability. With these
@@
 We culminate our research and development with a challenge problem by
-applying these new stochastic methods in the Drekar multiphysics
+applying these new Monte Carlo methods in the Drekar multiphysics
 code. With Drekar, we look forward to providing a new and robust
-solution method at the production scale. With these tools and a
+solution method at the production scale. Using these tools and a
 leadership class computing platform, we will provide performance
-measurements for the stochastic solvers and compare them to
-conventional methods currently leveraged within Drekar. With the
+measurements for the Monte Carlo solvers and compare them to
+conventional methods currently leveraged within Drekar. Given the
 potential to improve computational performance in production
 environments such as Drekar, these new methods have the potential to
-help meet the high fidelity modeling and simulation needs of the
+help meet the high-fidelity modeling and simulation needs of the
 nuclear community.

### 2012-10-02 doc/prelim/presentation/prelim_presentation.tex
adding prelim exam slides

\documentclass{beamer}
\usetheme[white]{Wisconsin}
\usepackage{longtable}
\usepackage{listings}
\usepackage{color}
%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
\usepackage{amsthm} \usepackage{amsmath} \usepackage{tmadd,tmath}
\usepackage[mathcal]{euscript} \usepackage{color}
\usepackage{textcomp}
\definecolor{listinggray}{gray}{0.9}
\definecolor{lbcolor}{rgb}{0.9,0.9,0.9}
\lstset{
  backgroundcolor=\color{lbcolor},
  tabsize=4,
  rulecolor=,
  language=c++,
  basicstyle=\scriptsize,
  upquote=true,
  aboveskip={1.5\baselineskip},
  columns=fixed,
  showstringspaces=false,
  extendedchars=true,
  breaklines=true,
  prebreak =
  \raisebox{0ex}[0ex][0ex]{\ensuremath{\hookleftarrow}},
  frame=single,
  showtabs=false,
  showspaces=false,
  showstringspaces=false,
  identifierstyle=\ttfamily,
  keywordstyle=\color[rgb]{0,0,1},
  commentstyle=\color[rgb]{0.133,0.545,0.133},
  stringstyle=\color[rgb]{0.627,0.126,0.941},
}

%%---------------------------------------------------------------------------%%
\newcommand{\vA}{\ensuremath{\ve{A}}}
\newcommand{\vb}{\ensuremath{\ve{b}}}
\newcommand{\vx}{\ensuremath{\ve{x}}}
\newcommand{\vr}{\ensuremath{\ve{r}}}
\newcommand{\vu}{\ensuremath{\ve{u}}}
\newcommand{\vI}{\ensuremath{\ve{I}}}
\newcommand{\vH}{\ensuremath{\ve{H}}}
\newcommand{\vP}{\ensuremath{\ve{P}}}

\newcommand{\Cv}{\ensuremath{C_{v}}}
\newcommand{\ros}{\ensuremath{\sigma_{\scriptscriptstyle\mathrm{R}}}}
\newcommand{\bOmega}{\ensuremath{\mathbf{\Omega}}}

\newcommand{\dt}{\ensuremath{\Delta t}}
\newcommand{\sign}{\ensuremath{\tilde{\sigma}^n}}
\newcommand{\qn}{\ensuremath{q^n}} \newcommand{\Tn}{\ensuremath{T^n}}
\newcommand{\Dn}{\ensuremath{D^n}}
\newcommand{\phin}{\ensuremath{\phi^n}}
\newcommand{\Di}{\ensuremath{\Delta_i}}
\newcommand{\Dj}{\ensuremath{\Delta_j}}
\newcommand{\Dk}{\ensuremath{\Delta_k}}

\newcommand{\sigT}{\ensuremath{\sigma_{T\,ijk}}}
\newcommand{\sigm}{\ensuremath{\sigma^{-}}}
\newcommand{\sigp}{\ensuremath{\sigma^{+}}}

\newcommand{\bphi}{\ensuremath{\boldsymbol{\phi}}}

\DeclareMathOperator{\diag}{diag}
%%---------------------------------------------------------------------------%%

\author{Stuart R. Slattery
  \\ Engineering Physics Department
  \\ University of Wisconsin - Madison
  \\ \href{mailto:sslattery@wisc.edu}{\texttt{sslattery@wisc.edu}}
}

\date{\today}
\title{Massively Parallel Monte Carlo Methods for Linear and Nonlinear
Problems}
\begin{document}
\maketitle

%%---------------------------------------------------------------------------%%

\begin{frame}{My Research and Development Activities}

  \begin{itemize}
  \item Stochastic computational linear algebra.
  \item Multiphysics methods for reactor physics.
  \item High performance computing.
  \item Software Engineering.
  \end{itemize}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Data Transfer Kit (DTK)}

  DTK is a software component for generating and applying parallel
  topology map operators based on multiphysics parameters. Some
  requirements:

  \begin{itemize}
  \item A component based design not limited to specific mesh or
    field data structures
  \item Compatible with both mesh-based and mesh-free physics codes
  \item Based on load-balanced rendezvous algorithms of a desirable
    time complexity
  \item Operate on $O(1,000,000,000)$ element meshes
  \item Operate on $O(100,000)$ cores
  \item Integrate readily with current CASL physics components
  \item Open-source implementation (DTK has a BSD 3-clause license)
  \end{itemize}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Stochastic Linear and Nonlinear Solvers}

  Ph.D. Research: Massively Parallel Stochastic Methods for Linear and
  Nonlinear Problems.

  \begin{itemize}
  \item Moving to exascale facilities requires methods tolerant to
    both soft and hard errors
  \item Solver scalability of primary concern
  \item Difficult to form the Jacobian analytically for block operator
    problems, Jacobian-free methods are too coarse or do not perform
    well for poorly scaled problems
  \item Subspace (e.g. Krylov) methods are memory intensive
  \item Stochastic solvers are a novel research area that may bear
    fruit
  \item We can leverage bleeding-edge Monte Carlo methods in reactor
    physics for parallelism
  \end{itemize}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{MCSA and FANM}

  \textbf{Monte Carlo Synthetic Acceleration (MCSA)}: A stochastic
  solution method for sparse linear problems.

  \begin{subequations}
    \label{eq:MCSA-iteration}
    \begin{gather}
      \vx^{l+1/2} = (\vI - \vA)\vx^l + \vb\:,\\ \vr^{l+1/2} = \vb -
      \vA\vx^{l+1/2}\:,
      \label{eq:MCSA-residual}\\     
      \hat{\vA}\delta\vx^{l+1/2} = \vr^{l+1/2}\:,
      \label{eq:MCSA-residual_solve}\\ 
      \vx^{l+1}=\vx^{l+1/2}+\delta\vx^{l+1/2}\:.
    \end{gather}
  \end{subequations}

  \textbf{Forward-Automated Newton-MCSA (FANM)}: A stochastic solution
  method for nonlinear problems

  \begin{itemize}
  \item Form the Jacobian explicitly with forward automatic
    differentiation. (No nasty derivatives, no Jacobian
    approximations, compressed row storage cheaper than the subspace
    over many iterations)
  \item Use MCSA as the linear solver. (All the benefits of the
    stochastic solver)
  \end{itemize}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{General MCSA Preliminary Results}

  Transient Poisson equation

  \begin{figure}[htpb!]
    \begin{center}
      \includegraphics[width=4in]{basic_results.png}
    \end{center}
  \end{figure}

  \begin{itemize}
  \item MCSA implemented in C++ using Epetra linear algebra
    framework. General linear algebra implemenation possible with
    MCSA!
  \item Competitive and GMRES and Conjugate Gradient solvers.
  \end{itemize}

  An initial FANM implementation is close and will be based on the
  Epetra solver. Parallel MCSA will be later in the fall per
  development of the Chimera code base.

\end{frame}

%%---------------------------------------------------------------------------%%

\end{document}

### 2012-10-02 doc/prelim/document/chapters/research_proposal/research_proposal.tex
added outline of slides for prelim

@@
 use. In addition, we will a formulate a plan for verifying the new
-stochastic methods and prepare a set of numerical experiments that
+Monte Carlo methods and prepare a set of numerical experiments that
 will determine some of their properties. Finally, we select a
@@
 leveraging the general linear algebra framework, we are able to
-abstract away those concepts in the stochastic solvers implementation
+abstract away those concepts in the Monte Carlo solvers implementation
 and instead focus on an algorithm-oriented approach for the solver
@@
 
-\section{Stochastic Methods Verification}
+\section{Monte Carlo Methods Verification}
 \label{sec:method_verification}

### 2012-10-02 doc/prelim/presentation/prelim_presentation.tex
added outline of slides for prelim

@@
 
-\begin{frame}{My Research and Development Activities}
-
-  \begin{itemize}
-  \item Stochastic computational linear algebra.
-  \item Multiphysics methods for reactor physics.
-  \item High performance computing.
-  \item Software Engineering.
-  \end{itemize}
+\begin{frame}{Introduction}
 
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Data Transfer Kit (DTK)}
-
-  DTK is a software component for generating and applying parallel
-  topology map operators based on multiphysics parameters. Some
-  requirements:
-
-  \begin{itemize}
-  \item A component based design not limited to specific mesh or
-    field data structures
-  \item Compatible with both mesh-based and mesh-free physics codes
-  \item Based on load-balanced rendezvous algorithms of a desirable
-    time complexity
-  \item Operate on $O(1,000,000,000)$ element meshes
-  \item Operate on $O(100,000)$ cores
-  \item Integrate readily with current CASL physics components
-  \item Open-source implementation (DTK has a BSD 3-clause license)
-  \end{itemize}
+\begin{frame}{Physics-Based Motivation}
 
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Stochastic Linear and Nonlinear Solvers}
-
-  Ph.D. Research: Massively Parallel Stochastic Methods for Linear and
-  Nonlinear Problems.
-
-  \begin{itemize}
-  \item Moving to exascale facilities requires methods tolerant to
-    both soft and hard errors
-  \item Solver scalability of primary concern
-  \item Difficult to form the Jacobian analytically for block operator
-    problems, Jacobian-free methods are too coarse or do not perform
-    well for poorly scaled problems
-  \item Subspace (e.g. Krylov) methods are memory intensive
-  \item Stochastic solvers are a novel research area that may bear
-    fruit
-  \item We can leverage bleeding-edge Monte Carlo methods in reactor
-    physics for parallelism
-  \end{itemize}
+\begin{frame}{Hardware-Based Motivation}
 
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{MCSA and FANM}
-
-  \textbf{Monte Carlo Synthetic Acceleration (MCSA)}: A stochastic
-  solution method for sparse linear problems.
-
-  \begin{subequations}
-    \label{eq:MCSA-iteration}
-    \begin{gather}
-      \vx^{l+1/2} = (\vI - \vA)\vx^l + \vb\:,\\ \vr^{l+1/2} = \vb -
-      \vA\vx^{l+1/2}\:,
-      \label{eq:MCSA-residual}\\     
-      \hat{\vA}\delta\vx^{l+1/2} = \vr^{l+1/2}\:,
-      \label{eq:MCSA-residual_solve}\\ 
-      \vx^{l+1}=\vx^{l+1/2}+\delta\vx^{l+1/2}\:.
-    \end{gather}
-  \end{subequations}
-
-  \textbf{Forward-Automated Newton-MCSA (FANM)}: A stochastic solution
-  method for nonlinear problems
-
-  \begin{itemize}
-  \item Form the Jacobian explicitly with forward automatic
-    differentiation. (No nasty derivatives, no Jacobian
-    approximations, compressed row storage cheaper than the subspace
-    over many iterations)
-  \item Use MCSA as the linear solver. (All the benefits of the
-    stochastic solver)
-  \end{itemize}
+\begin{frame}{Research Outline}
 
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{General MCSA Preliminary Results}
+\begin{frame}{Conventional Solution Methods for Linear Problems}
 
-  Transient Poisson equation
+\end{frame}
 
-  \begin{figure}[htpb!]
-    \begin{center}
-      \includegraphics[width=4in]{basic_results.png}
-    \end{center}
-  \end{figure}
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Linear Problem Preliminaries}
 
-  \begin{itemize}
-  \item MCSA implemented in C++ using Epetra linear algebra
-    framework. General linear algebra implemenation possible with
-    MCSA!
-  \item Competitive and GMRES and Conjugate Gradient solvers.
-  \end{itemize}
+\end{frame}
 
-  An initial FANM implementation is close and will be based on the
-  Epetra solver. Parallel MCSA will be later in the fall per
-  development of the Chimera code base.
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Stationary Methods}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Projection Methods}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Parallel Projection Methods}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Monte Carlo Solution Methods for Linear Problems}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Monte Carlo Linear Solver Preliminaries}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Direct Method}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Adjoint Method}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Sequential Monte Carlo}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Monte Carlo Synthetic-Acceleration}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Parallelization of Stochastic Methods}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Monte Carlo Solution Methods for Nonlinear Problems}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Monte Carlo Nonlinear Solver Preliminaries}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Inexact Newton Methods}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{The FANM Method}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Research Proposal}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Experimental Framework}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Progress to Date}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Monte Carlo Methods Verification}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Proposed Numerical Experiments}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Proposed Challenge Problem}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Conclusion}
 

### 2012-10-02 doc/prelim/document/chapters/research_proposal/research_proposal.tex
working on intro slide drafts

@@
 environments such as Drekar, these new methods have the potential to
-help meet the high-fidelity modeling and simulation needs of the
+help meet the high fidelity modeling and simulation needs of the
 nuclear community.

### 2012-10-02 doc/prelim/presentation/prelim_presentation.tex
working on intro slide drafts

@@
 %%---------------------------------------------------------------------------%%
-
 \begin{frame}{Introduction}
-
+  \begin{itemize}
+    \item Predictive modeling and simulation capability yields nuclear
+      system designs that are safer and better performing.
+    \item Modern work focused on this task leverages multiple physics
+      simulation (CASL, NEAMS).
+    \item New hardware drives algorithm development (petascale and
+      exascale).
+    \item Monte Carlo methods have the potential to provide great
+      improvements that permit finer simulations and better mapping to
+      future hardware.
+    \item A set of massively parallel Monte Carlo methods is proposed
+      to advance multiple physics simulation on contemporary and
+      future leadership class machines.
+  \end{itemize}
 \end{frame}
@@
 \begin{frame}{Physics-Based Motivation}
+  \begin{itemize}
+    \item Predictive nuclear reactor analysis enables tighter design
+      tolerance for improved thermal performance and efficieny, higher
+      fuel burn-up, and high confidence in accident scenario models.
+    \item Neutronics, thermal hydraulics, computational fluid
+      dynamics, and structural mechanics among other physics
+      contribute to these predictive models.
+    \item Couplings are complicated and consistent models yield
+      nonlinearities in the variables through feedback effects.
+    \item  Capturing  these   effects  requires  tremendous  resources
+      with $O(\sn{1}{9})$ element meshes and $O(100,000)+$ cores used
+      in today's simulations.
+  \end{itemize}
+\end{frame}
 
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Physics-Based Motivation: DNB}
+  \begin{figure}[htpb!]
+    \begin{center}
+      \scalebox{1}{ \input{dnb_example.pdftex_t} }
+    \end{center}
+    \caption{\textbf{Multiphysics dependency analysis of departure from
+        nucleate boiling.} \textit{A neutronics solution is required to
+        compute power generation in the fuel pins, fluid dynamics is
+        required to characterize boiling and fluid temperature and
+        density, heat transfer is required to compute the fuel and
+        cladding temperature, and the nuclear data modified with the
+        temperature and density data. Strong coupling among the
+        variables creates strong nonlinearities.}}
+    \label{fig:dnb_example}
+  \end{figure}
 \end{frame}
@@
 \begin{frame}{Hardware-Based Motivation}
-
+  \begin{itemize}
+    \item
+  \end{itemize}
 \end{frame}

### 2012-10-02 doc/prelim/presentation/prelim_presentation.tex
adding Chimera linear problem container

@@
 %%---------------------------------------------------------------------------%%
-\newcommand{\vA}{\ensuremath{\ve{A}}}
-\newcommand{\vb}{\ensuremath{\ve{b}}}
-\newcommand{\vx}{\ensuremath{\ve{x}}}
-\newcommand{\vr}{\ensuremath{\ve{r}}}
-\newcommand{\vu}{\ensuremath{\ve{u}}}
-\newcommand{\vI}{\ensuremath{\ve{I}}}
-\newcommand{\vH}{\ensuremath{\ve{H}}}
-\newcommand{\vP}{\ensuremath{\ve{P}}}
-
-\newcommand{\Cv}{\ensuremath{C_{v}}}
-\newcommand{\ros}{\ensuremath{\sigma_{\scriptscriptstyle\mathrm{R}}}}
-\newcommand{\bOmega}{\ensuremath{\mathbf{\Omega}}}
-
-\newcommand{\dt}{\ensuremath{\Delta t}}
-\newcommand{\sign}{\ensuremath{\tilde{\sigma}^n}}
-\newcommand{\qn}{\ensuremath{q^n}} \newcommand{\Tn}{\ensuremath{T^n}}
-\newcommand{\Dn}{\ensuremath{D^n}}
-\newcommand{\phin}{\ensuremath{\phi^n}}
-\newcommand{\Di}{\ensuremath{\Delta_i}}
-\newcommand{\Dj}{\ensuremath{\Delta_j}}
-\newcommand{\Dk}{\ensuremath{\Delta_k}}
-
-\newcommand{\sigT}{\ensuremath{\sigma_{T\,ijk}}}
-\newcommand{\sigm}{\ensuremath{\sigma^{-}}}
-\newcommand{\sigp}{\ensuremath{\sigma^{+}}}
-
-\newcommand{\bphi}{\ensuremath{\boldsymbol{\phi}}}
-
-\DeclareMathOperator{\diag}{diag}
-%%---------------------------------------------------------------------------%%
-
 \author{Stuart R. Slattery
@@
   \begin{itemize}
-    \item
+    \item Modern hardware is moving in two directions: lightweight
+      machines and heterogeneous machines characterized by low power
+      and high concurrency.
+    \item High concurrency and low cost units means a higher potential
+      for both soft and hard failures.
+    \item Monte Carlo methods bury soft failures within the variance
+      of the tallies while hard failures are high variance events.
+    \item New machines will also be memory restricted with a continued
+      decrease of memory/FLOPS predicted.
+    \item Compared to conventional methods, we will show that Monte
+      Carlo methods offer a memory savings.
   \end{itemize}

### 2012-10-02 drivers/TransientPoisson/TransientPoisson_main.cpp

@@ -300,10 +300,10 @@ int main( int argc, char * argv[] )
 				  &*ep_container->get_x(),
 				  &*ep_container->get_f() ) );
 
-    Chimera::Solvers::JacobiPreconditioner preconditioner( problem );
+    Chimera::JacobiPreconditioner preconditioner( problem );
     preconditioner.precondition();
     
-    Chimera::Solvers::MCSA solver( problem );
+    Chimera::MCSA solver( problem );
     int max_iters = 1000;
     double tolerance = 1.0e-8;
     int num_histories = 50;

### 2012-10-02 src/Chimera_AdjointMC.cpp

@@ -50,8 +50,6 @@
 
 namespace Chimera
 {
-namespace Solvers
-{
 //---------------------------------------------------------------------------//
 /*!
  * \brief Constructor.
@@ -325,7 +323,6 @@ Epetra_CrsMatrix AdjointMC::buildC()
 
 //---------------------------------------------------------------------------//
 
-} // end namespace Solvers
 } // end namespace Chimera
 
 //---------------------------------------------------------------------------//

### 2012-10-02 src/Chimera_AdjointMC.hpp

@@ -46,8 +46,6 @@
 
 namespace Chimera
 {
-namespace Solvers
-{
 
 class AdjointMC
 {
@@ -92,7 +90,6 @@ class AdjointMC
     Epetra_CrsMatrix buildC();
 };
 
-} // end namespace Solvers
 } // end namespace Chimera
 
 #endif // end Chimera_ADJOINTMC_HPP

### 2012-10-02 src/Chimera_BoostRNG.hpp

@@ -45,8 +45,6 @@
 
 namespace Chimera
 {
-namespace Solvers
-{
 //---------------------------------------------------------------------------//
 // Specialization for 64 bit mt19937
 //---------------------------------------------------------------------------//
@@ -104,8 +102,6 @@ class RNGTraits<boost::mt11213b>
 };
 
 //---------------------------------------------------------------------------//
-
-} // end namespace Solvers
 
 } // end namespace Chimera
 

### 2012-10-02 src/Chimera_JacobiPreconditioner.cpp

@@ -45,9 +45,6 @@
 
 namespace Chimera
 {
-namespace Solvers
-{
-
 //---------------------------------------------------------------------------//
 /*!
  * \brief Constructor.
@@ -151,7 +148,6 @@ void JacobiPreconditioner::preconditionRHS()
 
 //---------------------------------------------------------------------------//
 
-} // end namespace Solvers
 } // end namespace Chimera
 
 //---------------------------------------------------------------------------//

### 2012-10-02 src/Chimera_JacobiPreconditioner.hpp

@@ -46,9 +46,9 @@
 
 namespace Chimera
 {
-namespace Solvers
-{
-
+//---------------------------------------------------------------------------//
+// 
+//---------------------------------------------------------------------------//
 class JacobiPreconditioner
 {
   private:
@@ -93,7 +93,6 @@ class JacobiPreconditioner
 
 //---------------------------------------------------------------------------//
 
-} // end namespace Solvers
 } // end namespace Chimera
 
 #endif // end Chimera_JACOBIPRECONDITIONER_HPP

### 2012-10-02 src/Chimera_LinearProblem.hpp

@@ -31,58 +31,103 @@
   OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
 */
 //---------------------------------------------------------------------------//
-// \file Chimera_MCSA.hpp
-// \author Stuart R. Slattery
-// \brief Monte Carlo Synthetic Acceleration solver declaration.
+/*!
+ * \file Chimera_LinearProblem.hpp
+ * \author Stuart Slattery
+ * \brief Linear problem declaration.
+ */
 //---------------------------------------------------------------------------//
 
-#ifndef Chimera_MCSA_HPP
-#define Chimera_MCSA_HPP
+#ifndef Chimera_LINEARPROBLEM_HPP
+#define Chimera_LINEARPROBLEM_HPP
 
 #include <Teuchos_RCP.hpp>
 
-#include <Epetra_CrsMatrix.h>
-#include <Epetra_LinearProblem.h>
+#include <Tpetra_CrsMatrix.hpp>
+#include <Tpetra_Vector.hpp>
 
 namespace Chimera
 {
 //---------------------------------------------------------------------------//
-// 
+// \class LinearProblem
+// \brief Linear problem container for Chimera.
 //---------------------------------------------------------------------------//
-class MCSA
+template<class Scalar, class LocalOrdinal=int, class GlobalOrdinal=LocalOrdinal>
+class LinearProblem
 {
-  private:
+  public:
 
-    // Linear problem.
-    Teuchos::RCP<Epetra_LinearProblem> d_linear_problem;
-
-    // Iteration count.
-    int d_num_iters;
+    //@{
+    //! Typedefs.
+    typedef Scalar                                               scalar_type;
+    typedef LocalOrdinal                                         local_ordinal_type;
+    typedef GlobalOrdinal                                        global_ordinal_type;
+    typedef Tpetra::Vector<Scalar,LocalOrdinal,GlobalOrdinal>    TpetraVector;
+    typedef Teuchos::RCP<TpetraVector>                           RCP_TpetraVector;
+    typedef Tpetra::CrsMatrix<Scalar,LocalOrdinal,GlobalOrdinal> TpetraCrsMatrix;
+    typedef Teuchos::RCP<TpetraCrsMatrix>                        RCP_TpetraCrsMatrix;
+    //@}
 
   public:
 
     // Constructor.
-    MCSA( Teuchos::RCP<Epetra_LinearProblem> &linear_problem );
[…]

### 2012-10-02 src/Chimera_MCSA.cpp

@@ -44,8 +44,6 @@
 
 namespace Chimera
 {
-namespace Solvers
-{
 //---------------------------------------------------------------------------//
 /*! 
  * \brief Constructor.
@@ -120,7 +118,6 @@ void MCSA::iterate( const int max_iters, const double tolerance,
 
 //---------------------------------------------------------------------------//
 
-} // end namespace Solvers
 } // end namespace Chimera
 
 //---------------------------------------------------------------------------//

### 2012-10-02 src/Chimera_OperatorTools.cpp

@@ -52,8 +52,6 @@
 
 namespace Chimera
 {
-namespace Solvers
-{
 //---------------------------------------------------------------------------//
 /*!
  * \brief Constructor.
@@ -127,7 +125,6 @@ OperatorTools::stiffnessRatio( const Teuchos::RCP<Epetra_Operator>& matrix )
 
 //---------------------------------------------------------------------------//
 
-} // end namespace Solvers
 } // end namespace Chimera
 
 //---------------------------------------------------------------------------//

### 2012-10-02 src/Chimera_OperatorTools.hpp

@@ -47,8 +47,6 @@
 
 namespace Chimera
 {
-namespace Solvers
-{
 
 class OperatorTools
 {
@@ -64,7 +62,6 @@ class OperatorTools
     double static stiffnessRatio( const Teuchos::RCP<Epetra_Operator>& matrix );
 };
 
-} // end namespace Solvers
 } // end namespace Chimera
 
 #endif // Chimera_OPERATORTOOLS_HPP

### 2012-10-02 src/Chimera_RNGTraits.hpp

@@ -41,8 +41,6 @@
 
 namespace Chimera
 {
-namespace Solvers
-{
 //---------------------------------------------------------------------------//
 /*!
  * \brief Dummy struct. If a type does not create a specialization this will
@@ -108,8 +106,6 @@ class RNGTraits
 };
 
 //---------------------------------------------------------------------------//
-
-} // end namespace Solvers
 
 } // end namespace Chimera
 

### 2012-10-02 test/tstBoostRNG.cpp

@@ -34,7 +34,7 @@
 //---------------------------------------------------------------------------//
 TEUCHOS_UNIT_TEST( BoostRNG, boost_mt11213b_test )
 {
-    using namespace Chimera::Solvers;
+    using namespace Chimera;
 
     boost::mt11213b generator_1;
     boost::mt11213b generator_2 = RNGTraits<boost::mt11213b>::create();
@@ -75,7 +75,7 @@ TEUCHOS_UNIT_TEST( BoostRNG, boost_mt11213b_test )
 
 TEUCHOS_UNIT_TEST( BoostRNG, boost_mt19937_test )
 {
-    using namespace Chimera::Solvers;
+    using namespace Chimera;
 
     boost::mt19937 generator_1;
     boost::mt19937 generator_2 = RNGTraits<boost::mt19937>::create();

### 2012-10-02 src/Chimera_LinearProblem.hpp
linear problem test passing

@@ -72,39 +72,41 @@ class LinearProblem
 
     // Constructor.
     LinearProblem( const RCP_TpetraCrsMatrix& A, const RCP_TpetraVector& x, 
-		   const RCP_TpetraVector& b )
-	: d_A( A )
-	, d_x( x )
-	, d_b( b )
-    { /* ... */ }
+		   const RCP_TpetraVector& b );
 
     // Destructor.
-    ~LinearProblem()
-    { /* ... */ }
+    ~LinearProblem();
 
-    // Set the operator of the linear problem.
+    //! Set the operator of the linear problem.
     void setOperator( const RCP_TpetraCrsMatrix& A )
     { d_A = A; }
 
-    // Set the solution vector of the linear problem.
+    //! Set the solution vector of the linear problem.
     void setLHS( const RCP_TpetraVector& x )
     { d_x = x; }
 
-    // Set the right-hand sideof the linear problem.
+    //! Set the right-hand sideof the linear problem.
     void setRHS( const RCP_TpetraVector& b )
     { d_b = b; }
 
-    // Get the operator of the linear problem.
+    //! Get the operator of the linear problem.
     RCP_TpetraCrsMatrix getOperator()
     { return d_A; }
 
-    // Get the solution vector of the linear problem.
+    //! Get the solution vector of the linear problem.
     RCP_TpetraVector getLHS()
     { return d_x; }
 
-    // Get the right-hand side of the linear problem.
+    //! Get the right-hand side of the linear problem.
     RCP_TpetraVector getRHS()
     { return d_b; }
+
+    // Compute the residual of the linear problem.
+    void computeResidual();
+
+    //! Get the residual of the linear problem.
+    RCP_TpetraVector getResidual()
+    { return d_r; }
 
   private:
     
@@ -121,9 +123,15 @@ class LinearProblem
     RCP_TpetraVector d_r;
 };
 
+} // end namespace Chimera
+
+//---------------------------------------------------------------------------//
+// Template includes.
 //---------------------------------------------------------------------------//
 
-} // end namespace Chimera
+#include "Chimera_LinearProblem_def.hpp"
+
+//---------------------------------------------------------------------------//
 
 #endif // end Chimera_LINEARPROBLEM_HPP
 

### 2012-10-02 src/Chimera_LinearProblem_def.hpp

@@ -1,0 +1,91 @@
+//---------------------------------------------------------------------------//
+/*
+  Copyright (c) 2012, Stuart R. Slattery
+  All rights reserved.
+
+  Redistribution and use in source and binary forms, with or without
+  modification, are permitted provided that the following conditions are
+  met:
+
+  *: Redistributions of source code must retain the above copyright
+  notice, this list of conditions and the following disclaimer.
+
+  *: Redistributions in binary form must reproduce the above copyright
+  notice, this list of conditions and the following disclaimer in the
+  documentation and/or other materials provided with the distribution.
+
+  *: Neither the name of the University of Wisconsin - Madison nor the
+  names of its contributors may be used to endorse or promote products
+  derived from this software without specific prior written permission.
+
+  THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND CONTRIBUTORS
+  "AS IS" AND ANY EXPRESS OR IMPLIED WARRANTIES, INCLUDING, BUT NOT
+  LIMITED TO, THE IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR
+  A PARTICULAR PURPOSE ARE DISCLAIMED. IN NO EVENT SHALL THE COPYRIGHT
+  HOLDER OR CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
+  SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT NOT
+  LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES; LOSS OF USE,
+  DATA, OR PROFITS; OR BUSINESS INTERRUPTION) HOWEVER CAUSED AND ON ANY
+  THEORY OF LIABILITY, WHETHER IN CONTRACT, STRICT LIABILITY, OR TORT
+  (INCLUDING NEGLIGENCE OR OTHERWISE) ARISING IN ANY WAY OUT OF THE USE
+  OF THIS SOFTWARE, EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
+*/
+//---------------------------------------------------------------------------//
+/*!
+ * \file Chimera_LinearProblem_def.hpp
+ * \author Stuart Slattery
+ * \brief Linear problem definition.
+ */
+//---------------------------------------------------------------------------//
+
+#ifndef Chimera_LINEARPROBLEM_DEF_HPP
[…]

### 2012-10-02 test/tstLinearProblem.cpp

@@ -1,0 +1,133 @@
+//---------------------------------------------------------------------------//
+/*!
+ * \file tstLinearProblem.cpp
+ * \author Stuart R. Slattery
+ * \brief Boost random number generator unit tests.
+ */
+//---------------------------------------------------------------------------//
+
+#include <iostream>
+#include <vector>
+#include <cmath>
+#include <cstdlib>
+#include <sstream>
+#include <algorithm>
+#include <cassert>
+
+#include <Chimera_LinearProblem.hpp>
+
+#include <Teuchos_UnitTestHarness.hpp>
+#include <Teuchos_DefaultComm.hpp>
+#include <Teuchos_CommHelpers.hpp>
+#include <Teuchos_RCP.hpp>
+#include <Teuchos_ArrayRCP.hpp>
+#include <Teuchos_Array.hpp>
+#include <Teuchos_OpaqueWrapper.hpp>
+#include <Teuchos_TypeTraits.hpp>
+#include <Teuchos_Tuple.hpp>
+
+#include <Tpetra_Map.hpp>
+#include <Tpetra_CrsMatrix.hpp>
+#include <Tpetra_Vector.hpp>
+
+//---------------------------------------------------------------------------//
+// Tests
+//---------------------------------------------------------------------------//
+TEUCHOS_UNIT_TEST( LinearProblem, linear_problem_test )
+{
+    // Setup parallel distribution.
+    Teuchos::RCP<const Teuchos::Comm<int> > comm = 
+	Teuchos::DefaultComm<int>::getComm();
+    int comm_rank = comm->getRank();
+
+    // Setup linear system distribution.
+    int local_num_rows = 10;
+    Teuchos::RCP<const Tpetra::Map<int> > row_map = 
+	Tpetra::createLocalMap<int,int>( local_num_rows, comm );
+
+    // Build the identity matrix operator.
+    Teuchos::RCP<Tpetra::CrsMatrix<double,int> > A = 
+	Tpetra::createCrsMatrix<double,int>( row_map, 1 );
+    Teuchos::Array<int> global_columns(1);
+    Teuchos::Array<double> matrix_values(1, 1.0);
+    int gid = 0;
+    for ( int i = 0; i < local_num_rows; ++i )
+    {
+	gid = i + local_num_rows*comm_rank;
+	global_columns[0] = gid;
+	A->insertGlobalValues( gid, global_columns(), matrix_values() );
+    }
[…]

### 2012-10-15 doc/prelim/document/chapters/parallel_mc/parallel_mc.tex
working on adjoint section per spaniers work

@@
-\chapter{Monte Carlo Solution Methods for Linear Problems}
+\chapter{Monte Carlo Solution Methods for Linear Systems}
 \label{ch:stochastic_methods}
@@
 \citep{wasow_note_1952} and Hammersley and Handscomb's 1964 monograph
-\citep{hammersley_monte_1964} presents additional detail on this topic
+\citep{hammersley_monte_1964} and Spanier and Gelbard's 1969 book
+\citep{spanier_monte_1969} present additional detail on this topic
 using a collection of references from the 1950's and early 1960's.
@@
 \label{sec:mc_preliminaries}
-We begin our discussion of stochastic methods using these texts by
+We begin our discussion of Monte Carlo methods using these texts by
 seeking a solution to Eq~(\ref{eq:linear_problem}). For a given linear
-operator $\ve{A}$, we can using diagonal splitting in the same manner
+operator $\ve{A}$, we can using diagonal splitting in a similar manner
 as the stationary method in Eq~(\ref{eq:linear_split_equation2}) to
@@
 \begin{equation}
-  \ve{H} = \ve{I} - \ve{A}\:.
+  \ve{H} = \ve{I} - \ve{A}\:,
   \label{eq:linear_mc_iteration_matrix}
+\end{equation}
+such that we are solving the system:
+\begin{equation}
+  \ve{x} = \ve{H} \ve{x} + \ve{b}\:.
+  \label{eq:richardson_split}
 \end{equation}
@@
 An alternative formulation for Monte Carlo matrix inversion is the
-adjoint method. In this scheme, random walk histories are born in a
-state generated by source sampling and tallies are generated in the
-states in which the random walk currently resides. In terms of
-particle transport, this adjoint method is equivalent to a traditional
-forward method. To begin, we first form the \textit{adjoint
+adjoint method. We begin with generating the linear system adjoint to
+Eq~(\ref{eq:richardson_split}) by defining the following inner product
+equivalence \citep{spanier_monte_1969}:
+\begin{equation}
+  \langle \ve{A}^T \ve{x}, \ve{y} \rangle = \langle \ve{x}, \ve{A}
+  \ve{y} \rangle\:,
+  \label{eq:adjoint_operator_product}
+\end{equation}
+where $\ve{y}$ is the adjoint solution vector. With this statement we
+can then define the \textit{adjoint linear system}:
+\begin{equation}
+  \ve{y} = \ve{H}^T \ve{y} + \ve{d}\:,
+  \label{eq:adjoint_linear_system}
+\end{equation}
+where $\ve{d}$ is the adjoint source. As was required for convergence
+with the direct method using Eq~(\ref{eq:richardson_split}), the
+spectral radius of $\ve{H}$ must remain less than 1 as $\ve{H}^T$
+contains the same eigenvalues and therefore has the same spectral
+radius. From this definition it follows that:
+\begin{equation}
+  \langle \ve{x}, \ve{d} \rangle = \langle \ve{y}, \ve{b} \rangle\:.
+  \label{eq:adjoint_vector_relation}
+\end{equation}
+In this scheme, random walk histories are born in a state generated by
+source sampling and tallies are generated in the states in which the
+random walk currently resides. By comparing the contents of their
+estimators, we can relate $\ve{y}$ and $\ve{d}$ to the direct method
+solution.
+
+In terms of particle transport, this adjoint method is equivalent to a
+traditional forward method. As a result of using the adjoint system,
+we modify our probabilities and weights using the \textit{adjoint
   Neumann-Ulam decomposition} of $\ve{H}$:
@@
   \begin{split}
-    E\{X\} &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-    \sum_{i_k}^{N} p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
-    w_{i,i_1}w_{i_1,i_2}\ldots w_{i_{k-1},i_k} b_{i_k} \\ &= x_i\:,
+    E\{X_j\} &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
+    \sum_{i_k}^{N} b_{i_0} h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}
+    \delta_{i_k,j} \\ &= x_{j}\:,
   \end{split}

### 2012-10-16 doc/prelim/document/chapters/introduction/introduction.tex
updates to parallel krylov performance section

@@
 As leadership class machines move towards the exascale, new algorithms
-must developed that leverage their strengths and adapt to their
+must be developed that leverage their strengths and adapt to their
 shortcomings. Basic research is required now to advance methods in
@@
 provide a solution. As machines begin to operate at hundreds of
-petaflops peak performance and beyond, restrictions on power
-availability will require incredibly high levels of
-concurrency. Furthermore, this drop in power consumption will mean
-increased pressure on memory as memory per node is expected to
-stagnate while cores per node is expected to increase. As the number
-of cores increases, their clock speed is expected to stagnate or even
-decrease to further reduce power consumption and manufacturing costs.
+petaflops peak performance and beyond, trends toward reduced energy
+consumption will require incredibly high levels of concurrency to
+achieve the desired computation rates. Furthermore, this drop in
+power consumption will mean increased pressure on memory as memory per
+node is expected to stagnate while cores per node is expected to
+increase. As the number of cores increases, their clock speed is
+expected to stagnate or even decrease to further reduce power
+consumption and manufacturing costs.
 
@@
 are predicted to be common, resilient solver technologies are required
-overcome these events. With linear and nonlinear solvers based on
+to overcome these events. With linear and nonlinear solvers based on
 Monte Carlo techniques, such issues are alleviated by statistical

### 2012-10-16 doc/prelim/document/chapters/linear_problem/linear_problem.tex
working on adjoint section per spaniers work
updates to parallel krylov performance section

@@
-\chapter{Conventional Solution Methods for Linear Problems}
+\chapter{Conventional Solution Methods for Linear Systems}
 \label{ch:linear_problem}
@@
     \ve{r}^k = \ve{b} - \ve{A}\ve{x}^k\:,\\  
-    \ve{y}^k = (\ve{W}^T\ve{A}\ve{V})^{-1}\ve{W}^T\ve{r}_0\:,\\
+    \ve{y}^k = (\ve{W}^T\ve{A}\ve{V})^{-1}\ve{W}^T\ve{r}^k\:,\\
     \ve{x}^{k+1} = \ve{x}^k + \ve{V}\ve{y}^k\:,
@@
 where $\ve{V}$ and $\ve{W}$ are generated from the definitions of
-$\mathcal{K}$ and $\mathcal{L}$ which are defined prior to each
+$\mathcal{K}$ and $\mathcal{L}$ and are updated prior to each
 iteration.
@@
 the vector owned by process 1, we do these multiplications first as no
-communication is required. Next, process one gathers the components of
-the global vector that it requires to complete its part of the vector
-product. For this example, the components of matrix that will operate
-on the global vector are zero, and therefore no vector elements are
-required to be scattered from process 3 to process 1. Those matrix
-elements that will act on the piece of the vector owned by process 2
-are not all non-zero, and therefore we must gather the entire process
-2 vector components onto process 1 to complete the multiplication.
-Conversely, processes 2 and 3 must scatter its vector components that
-are required by other processes (such as process 1) in order to
-complete their pieces of the product. This then implies that these
-domain connections for proper gather and scatter combinations must be
-constructed a priori. These data structures are typically generated by
-a data partitioning library. Mathematically, if we are performing a
-global matrix-vector multiply of the form $\ve{A}\ve{x} = \ve{y}$,
-then for this example on process 1 we have a sequence of local
-matrix-vector multiplications: $\ve{A}_1\ve{x}_1 + \ve{A}_1\ve{x}_2 =
-\ve{y}_1$. Here, some of the data is intrinsically local, and some
-must be gathered from other processes using the partitioning data
-structures.
+communication is required. Next, process 1 gathers the components of
+the global vector owned by the other two processes that it requires to
+complete its part of the vector product. For this example, the
+components of matrix owned by process 1 that will operate on the
+global vector components owned by process 3 are zero, and therefore no
+vector elements are required to be scattered from process 3 to process
+1. Those matrix elements owned by process 1 that will act on the piece
+of the vector owned by process 2 are not all non-zero, and therefore
+we must gather the entire process 2 vector components onto process 1
+to complete the multiplication.  Conversely, processes 2 and 3 must
+scatter their vector components that are required by other processes
+(such as process 1) in order to complete their pieces of the
+product. This then implies that these domain connections for proper
+gather and scatter combinations must be constructed a priori. These
+data structures are typically generated by a data partitioning
+library. Mathematically, if we are performing a global matrix-vector
+multiply of the form $\ve{A}\ve{x} = \ve{y}$, then for this example on
+process 1 we have a sequence of local matrix-vector multiplications:
+$\ve{A}_1\ve{x}_1 + \ve{A}_1\ve{x}_2 = \ve{y}_1$ as annotated by the
+notation in Figure~\ref{fig:matvec_proc_1}. Here, some of the data is
+intrinsically local, and some must be gathered from other processes
+using the partitioning data structures.
 
@@
 perform in order to implement Krylov methods, we can make a few
-statements about parallel performance and implications for operation on
-machines of increasing size.  For very large distributed machines,
-global reduction operations required at several levels of Krylov
-algorithms stand to reduce scalability and performance. Furthermore,
-communication between adjacent domains in matrix-vector multiply
-operations may also cause a bottleneck as the number of domains used
-in simulation grows. The end result is that global data must be
-collected and communicated. For scaling improvement, we seek a
+statements about parallel performance and implications for operation
+on machines of increasing size. Reconsider the matrix and vector
+operations required to implement Algorithm~\ref{alg:gmres}. For very
+large distributed machines, the global reduction operations required
+at several levels of Krylov algorithms stand to reduce scalability and
+performance. In the case of GMRES, these reductions include vector
+norm operations and the inner products required for basis
+orthogonalization. Furthermore, communication between adjacent domains
+in matrix-vector multiply operations may also cause a bottleneck as
+the number of domains used in a simulation grows and the number of
+processors participating in the gather/scatter sequence requires a
+large communication bandwidth. The end result is that global data must
+be collected and communicated. For scaling improvement, we seek a
 reduction in these types of operations. In addition, these issues
@@
 orthogonalize that subspace to increase.
+
+As an example of these performance implications in practice, in a 2001
+work, Gropp and colleagues presented results on fluid dynamic
+simulations that heavily leveraged Krylov methods in their solution
+schemes \citep{gropp_high-performance_2001}. In this follow-on to
+their 1997 Bell Prize-winning work, part of their analysis included
+identifying parallel scalability bottlenecks generated by solver
+implementations in a strong scaling exercise where the number of
+processors was increased with respect to a fixed global problem
+size. Gropp's observations show that for their particular hardware, a
+distributed memory machine similar to modern architectures, that
+global reduction operations did not impede scalability, meaning that
+the global reduction operation occupied approximately the same
+percentage of compute time independent of the number of processors
+used. Rather, it was the gather/scatter operations required to
+communicate data to neighboring processors that reduced performance
+with an increasing percentage of compute time consumed by these
+operations as processor count was increased. Furthermore, it was noted
+that this reduction in scaling was a product of poor algorithmic
+strong scaling rather than hardware or implementation related issues
+as the algorithm requires more data to be scattered/gathered as the
+number of processors and therefore computational domains increased. In
+the case of a weak scaling exercise, we would instead expect this
+percentage to exhibit a more desirable behavaior of remaining constant
+for gather/scatter operations as problem size would be scaled with the
+number of processors.

### 2012-10-16 doc/prelim/document/chapters/nonlinear_problem/nonlinear_problem.tex
finished rng traits and boost implementations and tests
working on adjoint section per spaniers work
incrporating more of Pauls changes. still a few more to go

@@
-\chapter{Stochastic Solution Methods for Nonlinear Problems}
+\chapter{Monte Carlo Solution Methods for Nonlinear Systems}
 \label{ch:nonlinear_problem}
@@
 
-Currently, Krylov methods are used with Newton's method to form a
-robust solution sequence for these types of problems. However, given
-our goals of advancing stochastic solver technology for linear
-problems, we aim to apply them to these fully consistent schemes in
-order to gauge if some improvements can be made. We therefore devise a
-new method using MCSA in conjunction with new methods, accommodating the
-fact that MCSA requires an explicitly formed linear operator in its
-implementation as compared to Krylov methods. In addition to handling
-this dependency, we will construct a parallel strategy for a nonlinear
-MCSA-based method.
+In the context of solving standalone linear systems, Monte Carlo
+methods do not provide significant merit over Krylov methods due to
+the fact that the linear operator must be explicitly formed. For many
+applications, such a requirement is prohibitive and perhaps not even
+feasible to implement. Therefore, a Monte Carlo solver is best suited
+for situations in which not only are Krylov methods applicable, but
+also in which the operator is readily, if not naturally,
+formed. Modern nonlinear methods meet both of these requirements with
+Newton methods used in conjuction with Krylov methods for a robust,
+fully implicit solution strategy. Furthermore, modern techniques exist
+that permit the automatic construction of the linear operator
+generated within a Newton method based on the nonlinear residual
+evaluations, providing all of the components necessary for a Monte
+Carlo solver to provide value. We therefore devise a new nonlinear
+method based on the MCSA algorithm and Newton's method and discuss its
+potential benefits.
 

### 2012-10-16 doc/prelim/document/chapters/parallel_mc/parallel_mc.tex
incrporating more of Pauls changes. still a few more to go

@@
 seeking a solution to Eq~(\ref{eq:linear_problem}). For a given linear
-operator $\ve{A}$, we can using diagonal splitting in a similar manner
+operator $\ve{A}$, we can use diagonal splitting in a similar manner
 as the stationary method in Eq~(\ref{eq:linear_split_equation2}) to
@@
 
-As we seek only approximate solutions, we need only to perform a
-minimal number of random walks in order to generate an approximation
-for $\ve{x}$. If we are only to approximate the solution, we also need
-conditions by which we may terminate a random walk. We do this by
-noticing that the factors added to
+In cases where we seek only approximate solutions, we need only to
+perform a minimal number of random walks in order to generate an
+approximation for $\ve{x}$. If we are only to approximate the
+solution, we also need conditions by which we may terminate a random
+walk. We do this by noticing that the factors added to
 Eq~(\ref{eq:direct_permutation_weight}) will become diminishingly
@@
 therefore their contributions to the solution estimate will become
-negligible. Using this, we choose terminate a random walk sequence with
-a \textit{weight cutoff}, $W_c$, that is enforced when $W_m < W_c$ for
-a particular random walk permutation.
+negligible. Using this, we choose terminate a random walk sequence
+with a \textit{weight cutoff}, $W_c$, that is enforced when $W_m <
+W_c$ for a particular random walk permutation.
 
@@
 Choosing Jacobi preconditioning with MCSA is advantageous for several
-reasons. First, $\rho(\ve{I} - \ve{M}^{-1}\ve{A}) < 1$ is true for
-all $\ve{A}$ and is easy to formulate as the inversion of $\ve{M}$ is
-trivial. Second, because the adjoint Monte Carlo method used within MCSA to
-compute the correction operates on a linear problem with the
+reasons. First, $\rho(\ve{I} - \ve{M}^{-1}\ve{A}) < 1$ is true for all
+$\ve{A}$ and is easy to formulate because the inversion of $\ve{M}$ is
+trivial. Second, because the adjoint Monte Carlo method used within
+MCSA to compute the correction operates on a linear problem with the
 preconditioned operator, then $\ve{H}$ in the adjoint solver will have
-a zero term in each of its diagonal elements, thereby eliminating
-all in-state transitions during the random walk sequence. Because of
-this, Jacobi preconditioning should always be performed, regardless of
-any other preconditioning that is applied to the system.
+a zero term in each of its diagonal elements, thereby eliminating all
+in-state transitions during the random walk sequence. Because of this,
+Jacobi preconditioning should always be performed, regardless of any
+other preconditioning that is applied to the system.
 
@@
 cases efficiently. Such cases will be common if the source sampled in
-the Monte Carlo random walk is not isotropic and evenly distributed
-throughout the global domain. It was noted that efficiencies were
-improved by increasing the frequency by which particle data was
-communicated between domain-adjacent processors. However, this
-ultimately increases communication costs. In 2009, Brunner extended
-his work by using a more load-balanced approach with a fully
-asynchronous communication pattern
+the Monte Carlo random walk is not isotropic and not evenly
+distributed throughout the global domain. It was noted that
+efficiencies were improved by increasing the frequency by which
+particle data was communicated between domain-adjacent
+processors. However, this ultimately increases communication costs. In
+2009, Brunner extended his work by using a more load-balanced approach
+with a fully asynchronous communication pattern
 \citep{brunner_efficient_2009}. Although the extended implementation
@@
 communication. A multiple set scheme is also motivated by the fact
-that communication only occurs within a set, limiting communications
-during the transport procedure to a group of $O(100)$ processors, a
-number that was shown to have excellent parallel efficiencies in
-Brunner's work and therefore will scale well in this algorithm. The
-overlapping domains within each set also demonstrated reduced
-communication costs. On each processor, the source is sampled in the
-local domain that would exist if no overlap was used while tallies can
-be made over the entire overlapping domain.
+that communication during particle transport only occurs within a set,
+limiting communications during the transport procedure to a group of
+$O(100)$ processors, a number that was shown to have excellent
+parallel efficiencies in Brunner's work and therefore will scale well
+in this algorithm. The overlapping domains within each set also
+demonstrated reduced communication costs. On each processor, the
+source is sampled in the local domain that would exist if no overlap
+was used while tallies can be made over the entire overlapping domain.
 
@@
 particle maintains an invariant random number sequence that determines
-its behavior regardless of domain decomposition, each particle is a
-random number seed that describes its current state upon entering the
-domain of a new processor. These seeds are derived from the actual
-geometric location of the particle such that it is decomposition
-invariant. Non-commutative floating point operations are solved by
-instead mapping floating point values to 64-bit integer values for
-which additions will always be commutative. Once the operations are
-complete, these integers are mapped back to floating point values.
+its behavior regardless of domain decomposition, each particle is
+assigned a random number seed that describes its current state upon
+entering the domain of a new processor. These seeds are derived from
+the actual geometric location of the particle such that it is
+decomposition invariant. Non-commutative floating point operations are
+overcome by instead mapping floating point values to 64-bit integer
+values for which additions will always be commutative. Once the
+operations are complete, these integers are mapped back to floating
+point values.
 

### 2012-10-16 doc/prelim/document/chapters/linear_problem/linear_problem.tex
updates to adjoint section

@@
 multiply of the form $\ve{A}\ve{x} = \ve{y}$, then for this example on
-process 1 we have a sequence of local matrix-vector multiplications:
-$\ve{A}_1\ve{x}_1 + \ve{A}_1\ve{x}_2 = \ve{y}_1$ as annotated by the
-notation in Figure~\ref{fig:matvec_proc_1}. Here, some of the data is
-intrinsically local, and some must be gathered from other processes
-using the partitioning data structures.
+process 1 we have a sequence of local matrix-vector multiplications:n
+$\ve{A}_1\ve{x}_1 + \ve{A}_1\ve{x}_2 = \ve{y}_1$, with the subscripts
+provided by notation in Figure~\ref{fig:matvec_proc_1}. Here, some of
+the data is intrinsically local, and some must be gathered from other
+processes using the partitioning data structures.
 

### 2012-10-16 doc/prelim/document/chapters/parallel_mc/parallel_mc.tex
updates to adjoint section

@@
 which will converge if the spectral radius of $\ve{H}$ is less than
-1. If we then apply this inverse sum to the right hand side of
+1. If we then apply this Neumann series to the right hand side of
 Eq~(\ref{eq:linear_problem}) we acquire the solution to the linear
@@
 \end{equation}
-An approximation of this summation will therefore lead to an
-approximation of the solution. If we expand the summation with a
+An approximation of this summation by truncation will therefore lead
+to an approximation of the solution. If we expand the summation with a
 succession of matrix-vector multiply operations, we arrive at an
@@
 \end{equation}
-in $\ve{H}$ where $\nu$ is a particular random walk sequence
-permutation. We can generate these sequences of transitions through
-Monte Carlo random walks by assigning them both a probability and
-weight. As a reinterpretation of the iteration matrix, we then form
-the \textit{Neumann-Ulam decomposition} of \ve{H}:
+in $\ve{H}$ where $\nu$ is interpreted as a particular random walk
+sequence permutation. We can generate these sequences of transitions
+through Monte Carlo random walks by assigning them both a probability
+and weight. As a reinterpretation of the iteration matrix, we then
+form the \textit{Neumann-Ulam decomposition} of \ve{H}:
 \begin{equation}
@@
 scheme in that now a weight cutoff can be used to terminate a random
-walk sequence. The formulation of $\ve{P}$ and $\ve{W}$ will be
+walk sequence and therefore truncate the Neumann series it is
+approximating. The formulation of $\ve{P}$ and $\ve{W}$ will be
 dependent on whether we choose a direct or adjoint Monte Carlo
@@
 Choosing the infinity norm of the operator as defined in
-Eq~(\ref{eq:matrix_infinity_norm}), the normalized probability matrix
-will yield a norm of 1 giving the following inequality for relating
-$\ve{G}$ and $\ve{H}$:
+Eq~(\ref{eq:matrix_infinity_norm}), the row normalized probability
+matrix will yield a norm of 1 giving the following inequality for
+relating $\ve{G}$ and $\ve{H}$:
 \begin{equation}
@@
 An alternative formulation for Monte Carlo matrix inversion is the
-adjoint method. We begin with generating the linear system adjoint to
+adjoint method. We begin with defining the linear system adjoint to
+Eq~(\ref{eq:linear_problem}):
+\begin{equation}
+  \ve{A}^T \ve{y} = \ve{d}\:,
+  \label{eq:adjoint_linear_problem}
+\end{equation}
+where $\ve{y}$ and $\ve{d}$ are the adjoint solution and source
+respectively. We can split this equation to mirror
 Eq~(\ref{eq:richardson_split}) by defining the following inner product
@@
   \langle \ve{A}^T \ve{x}, \ve{y} \rangle = \langle \ve{x}, \ve{A}
-  \ve{y} \rangle\:,
+  \ve{y} \rangle\:.
   \label{eq:adjoint_operator_product}
 \end{equation}
-where $\ve{y}$ is the adjoint solution vector. With this statement we
-can then define the \textit{adjoint linear system}:
+With this statement we can then define split equation.:
 \begin{equation}
-  \ve{y} = \ve{H}^T \ve{y} + \ve{d}\:,
+  \ve{y} = \ve{H}^T \ve{y} + \ve{d}\:.
   \label{eq:adjoint_linear_system}
 \end{equation}
-where $\ve{d}$ is the adjoint source. As was required for convergence
-with the direct method using Eq~(\ref{eq:richardson_split}), the
-spectral radius of $\ve{H}$ must remain less than 1 as $\ve{H}^T$
-contains the same eigenvalues and therefore has the same spectral
-radius. From this definition it follows that:
+As was required for convergence with the direct method using
+Eq~(\ref{eq:richardson_split}), the spectral radius of $\ve{H}$ must
+remain less than 1 as $\ve{H}^T$ contains the same eigenvalues and
+therefore has the same spectral radius. From this definition it
+follows that:
 \begin{equation}
@@
 \end{equation}
-In this scheme, random walk histories are born in a state generated by
-source sampling and tallies are generated in the states in which the
-random walk currently resides. By comparing the contents of their
-estimators, we can relate $\ve{y}$ and $\ve{d}$ to the direct method
-solution.
+By comparing the contents of their estimators, we can determine
+$\ve{y}$ and $\ve{d}$ from the direct method. In this scheme,
+random walk histories are born in a state generated by source sampling
+and tallies are generated in the states in which the random walk
+currently resides.
 
@@
 method we are only tallying the solution in the state in which the
-walk currently resides, the contribution to the solution from a
-particular random walk permutation is then:
+walk currently resides, the contribution to the solution in state $i$
+from a particular random walk permutation is then:
 \begin{equation}

### 2012-10-17 doc/prelim/document/chapters/parallel_mc/parallel_mc.tex
finished adjoint section updates and included more details on Seigels communication work

@@
 An alternative formulation for Monte Carlo matrix inversion is the
-adjoint method. We begin with defining the linear system adjoint to
+adjoint method. We begin by defining the linear system adjoint to
 Eq~(\ref{eq:linear_problem}):
@@
 where $\ve{y}$ and $\ve{d}$ are the adjoint solution and source
-respectively. We can split this equation to mirror
-Eq~(\ref{eq:richardson_split}) by defining the following inner product
-equivalence \citep{spanier_monte_1969}:
+respectively and $\ve{A}^T$ is the adjoint operator. We can split this
+equation to mirror Eq~(\ref{eq:richardson_split}) by defining the
+following inner product equivalence \citep{spanier_monte_1969}:
 \begin{equation}
@@
 \end{equation}
-With this statement we can then define split equation.:
+With this statement we can then define the split equation:
 \begin{equation}
   \ve{y} = \ve{H}^T \ve{y} + \ve{d}\:.
-  \label{eq:adjoint_linear_system}
+  \label{eq:adjoint_split_system}
 \end{equation}
@@
 \end{equation}
-By comparing the contents of their estimators, we can determine
-$\ve{y}$ and $\ve{d}$ from the direct method. In this scheme,
-random walk histories are born in a state generated by source sampling
-and tallies are generated in the states in which the random walk
-currently resides.
+Using these definitions, we can derive an estimator from the adjoint
+method that will also give the solution vector, $\ve{x}$. As with the
+direct method, we can acquire the adjoint solution by forming the
+Neumann series by writing Eq~(\ref{eq:adjoint_split_system}) as:
+\begin{equation}
+  \ve{y} = (\ve{I} - \ve{H}^T)^{-1} \ve{d}\:,
+  \label{eq:adjoint_split_system_2}
+\end{equation}
+which in turn yields the Neumann series using the adjoint operator:
+\begin{equation}
+  \ve{y} = \sum_{k=0}^{\infty} (\ve{H}^T)^k\ve{d}\:.
+  \label{eq:adjoint_neumann_series}
+\end{equation}
+We expand this summation to again yield a series of transitions that
+can be approximated by a Monte Carlo random walk sequence, this time
+forming the Neumann series in reverse order:
+\begin{equation}
+  y_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
+  \sum_{i_k}^{N}h_{i_k,i_{k-1}}\ldots h_{i_2,i_1} h_{i_1,i} d_{i_k}\:.
+  \label{eq:adjoint_neumann_solution}
+\end{equation}
+We can readily build an estimator for the adjoint solution from this
+series expansion, but we instead desire the direct solution. We
+achieve this by using Eq~(\ref{eq:adjoint_vector_relation}) as a
+constraint. Here we have 2 unkowns, $\ve{y}$ and $\ve{d}$ and
+therefore we require a second constraint to close the system. As a
+second constraint we select:
+\begin{equation}
+  \ve{d} = \boldsymbol{\delta}_i\:,
+  \label{eq:adjoint_second_constraint}
+\end{equation}
+where the $k^{th}$ component of the vector $\boldsymbol{\delta}_i$ is
+the Dirac delta function $\delta_{i_k,i}$. If we apply
+Eq~(\ref{eq:adjoint_second_constraint}) to our first constraint
+Eq~(\ref{eq:adjoint_vector_relation}), we get the following convenient
+outcome:
+\begin{equation}
+  \langle \ve{y}, \ve{b} \rangle = \langle \ve{x},
+  \boldsymbol{\delta}_i \rangle = x_i \:,
+  \label{eq:inner_product_constraint}
+\end{equation}
+meaning that if we compute the inner product of the direct source and
+the adjoint solution using a delta function source, we recover the
+direct solution we are looking for
 
@@
 the weight generated for a particular random walk permutation as in
-Eq~(\ref{eq:mc_walk_permutation}) and the fact that in the adjoint
-method we are only tallying the solution in the state in which the
-walk currently resides, the contribution to the solution in state $i$
+Eq~(\ref{eq:mc_walk_permutation}) and our result from
+Eq~(\ref{eq:inner_product_constraint}) generated by applying the
+adjoint constraints, the contribution to the solution in state $i$
 from a particular random walk permutation is then:
@@
 in all states and those of lesser importance will not be visited as
-frequently by the random walk. Finally, the estimator using all
-permutations is:
+frequently by the random walk. Finally, the expectation value using
+all permutations is:
 \begin{equation}
@@
 \end{equation}
-therefore also providing an unbiased Monte Carlo estimator.
+therefore also providing an unbiased Monte Carlo estimate of the
+solution.
 
-We also desire a criteria for random walk termination. For the adjoint
-method, we utilize a \textit{relative weight cutoff}:
+Like the direct method, we also desire a criteria for random walk
+termination for problems where only an approximate solution is
+necessary. For the adjoint method, we utilize a \textit{relative
+  weight cutoff}:
 \begin{equation}
@@
 production scale, scalable parallel implementations are required. In a
-general linear algebra context, this has not yet been
-achieved. Therefore, we will develop and implement parallel algorithms
-for these methods leveraging both the knowledge gained from the
-general parallel implementations of Krylov methods and modern parallel
-strategies for Monte Carlo as developed by the reactor physics
-community. In order to formulate a parallel MCSA algorithm, we
+general linear algebra context for discrete systems, this has not yet
+been achieved. Therefore, we will develop and implement parallel
+algorithms for these methods leveraging both the knowledge gained from
+the general parallel implementations of Krylov methods and modern
+parallel strategies for Monte Carlo as developed by the reactor
+physics community. In order to formulate a parallel MCSA algorithm, we
 recognize that the algorithm occurs in two stages, an outer iteration
@@
 decomposed due to memory limitations. This naturally leads us to seek
-parallel algorithms that handle domain decomposition
-\citep{romano_towards_2010}.
+parallel algorithms that handle domain decomposition.
 
@@
 that the domain was only decomposed on $O(100)$ processors and if
-replicated $O(1000)$ times achieves simulation on $O(100,000)$
+replicated $O(1,000)$ times achieves simulation on $O(100,000)$
 processors, thus providing spatial and particle parallelism. Each
@@
 
+\subsubsection{Domain-to-Domain Communication}
+\label{subsubsec:mc_comm_costs}
+With Wagner's work, we observed that domain overlap implemented with a
+domain decomposed Monte Carlo setting reduces the amount of nearest
+neighbor communication that must occur during transport. This
+communication, expected to be expensive, was analyzed by Siegel and
+colleagues for idealized, load balanced situations without domain
+overlap \citep{siegel_analysis_2012}. They arrive at a few empirical
+relationships and generate a numerical experiment that provides
+insight into the feasibility of large scale domain decomposed Monte
+Carlo as well as the relationship of communication to system
+parameters. In their work, they define a simple 3-dimensional system
+subdivded into Cartesian block domains with each processor containing
+a single domain. Given this geometry, a particle can either be
+absorbed in its current domain or be transported to one of the 6
+adjacent domains owned by another process.
+
+To quantify the number of particles that leak out of the local domain
+they define a leakage fraction, $\lambda$, as:
+\begin{equation}
+  \lambda =
+  \frac{average\ \#\ of\ particles\ leaving\ local\ domain}{total\ 
+    of\ \#\ of\ particles\ starting\ in\ local\ domain}\:.
+  \label{eq:leakage_fraction}
+\end{equation}
+For their studies, they assumed that the value of $\lambda$ was
+dependent on the total cross section of the system and the same for
+all processes for simplicity and load balancing. Using this
+relationship and their simple geometry, they were able to relate the
+amount of compute time required to transport all source particles in
+relation to the number of source particles, the leakage fraction, the
+latency and bandwidth of communication, the ratio of local to global
+geometry size, and the total cross section of the system.
+
+What they found was that there was indeed an optimum ratio of local
+domain to global domain size that results in a minimum total compute
+time. They also found that this ratio was only dependent on the
+latency and bandwidth constraints of the parallel machine. We expect
+this because nearest neighbor communication patterns scale with the
+volume to surface sizes of the local domains. The larger the volume,
+the more compute time is spent within the local volume and therefore
+at some point you are limited by the latency of the system, or the
+time it takes to receive the data to operate on. However, at a certain
+point, when the surface of those volumes becomes large enough, the
+amount of data that must be communicated grows large enough that it
+may overcome the bandwidth limitations of the system, resulting in a
+decrease in performance. In their numerical experiments, they used the
+above geometry and leakage rates based on that computed for a reactor
+system for a single material problem with a total cross section
+equivalent to that of homogonized light water reactor. Given the
+optimal local domain size, they observed for load balanced situations
+that nearest neighbor communication costs during transport are a small
+fraction of total compute time, leading to the conclusing that domain
+decomposition is feasible for large-scale Monte Carlo calculations.
+
 \subsubsection{Load Balancing Concerns}
 \label{subsubsec:mc_load_balancing}
-Although domain decomposition was shown to be efficient in a load
-balanced situation \citep{siegel_analysis_2012}, careful consideration
-must be made for situations where this is not the case. Given the
-stochastic nature of the problem and lack of a globally homogeneous
-domain, parallel Monte Carlo simulations are inherently load
-imbalanced. Procassini and others worked to alleviate some of the load
-imbalances that are generated by both particle and spatial parallelism
-and are therefore applicable to the MSOD algorithm
+Although domain decomposition was shown to be efficient in a prefectly
+load balanced situation in Siegel's work \citep{siegel_analysis_2012},
+careful consideration must be made for situations where this is not
+the case. Given the stochastic nature of the problem and lack of a
+globally homogeneous domain, parallel Monte Carlo simulations are
+inherently load imbalanced. Procassini and others worked to alleviate
+some of the load imbalances that are generated by both particle and
+spatial parallelism and are therefore applicable to the MSOD algorithm
 \citep{procassini_dynamic_2005}. They chose a dynamic balancing scheme
@@
 
-Given a decomposed domain, it is clear that domain overlap
-significantly reduces the amount of communication required between
-adjacent domains as histories move to states that are not owned by the
-local processor. How much domain overlap is suitable for matrix
-inversion problems? Are we memory limited by large problems such that
-only so much overlap is feasible? Are there metrics related to the
-properties of the matrix such that we can provide guidelines for
-selecting the amount of overlap required? Furthermore, as Wagner and
-colleagues observed \citep{wagner_hybrid_2010}, with marginal domain
-overlap for reactor problems, the percentage of histories leaving the
-local domain can easily be reduced to less than 1\%. Evans and
-colleagues in their initial MCSA development typically used only 50
-particle histories in order to compute a correction in an MCSA
-iteration \citep{evans_monte_2012}. Per the Central Limit Theorem,
-that correlates to a statistical uncertainty of 14.1\% in the
-correction, yet as compared to conventional Krylov solvers, MCSA
-achieved identical numerical results. If good convergence and
-numerically accurate solutions can be achieved with such a large
-uncertainty in the correction computation, then perhaps with enough
-domain overlap the minimal amount of histories that do transition to
-non-local states can be ignored and thus eliminate all communication
-in the parallel Monte Carlo adjoint solver and create an embarrassingly
-parallel method. It will be important to determine if the expectation
-value bias in the MCSA correction generated by this approximation will
-remain within the bounds of the already high statistical uncertainty
-for unbiased estimates, thus providing numerically equivalent results.
+Given a decomposed domain, per Wagner's work
+\citep{wagner_hybrid_2010} is clear that domain overlap significantly
+reduces the amount of communication required between adjacent domains
+as histories move to states that are not owned by the local
+processor. How much domain overlap is suitable for matrix inversion
+problems? Are we memory limited by large problems such that only so
+much overlap is feasible? Are there metrics related to the properties
+of the matrix such that we can provide guidelines for selecting the
+amount of overlap required? Furthermore, as Wagner and colleagues
+observed \citep{wagner_hybrid_2010}, with marginal domain overlap for
+reactor problems, the percentage of histories leaving the local domain
+can easily be reduced to less than 1\%. Evans and colleagues in their
+initial MCSA development typically used only 50 particle histories in
+order to compute a correction in an MCSA iteration
+\citep{evans_monte_2012}. Per the Central Limit Theorem, that
+correlates to a statistical uncertainty of 14.1\% in the correction,
+yet as compared to conventional Krylov solvers, MCSA achieved
+identical numerical results. If good convergence and numerically
+accurate solutions can be achieved with such a large uncertainty in
+the correction computation, then perhaps with enough domain overlap
+the minimal amount of histories that do transition to non-local states
+can be ignored and thus eliminate all communication in the parallel
+Monte Carlo adjoint solver transport sequence and create an
+embarrassingly parallel method. Given Gropp's work on parallel Krylov
+methods that we discussed in
+\S~\ref{subsec:projection_method_performance} and the empirical
+results of Siegel \citep{siegel_analysis_2012}, we know that these
+nearest neighbor computations between adjacent domains do not scale as
+well as global reduction operations and therefore we are improving
+scaling by eliminating them from the transport sequence. It will be
+important to determine if the expectation value bias in the MCSA
+correction generated by this approximation will remain within the
+bounds of the already high statistical uncertainty for unbiased
+estimates, thus providing numerically equivalent results.
 

### 2012-10-17 src/Chimera_History.hpp
Updated history class and tests.

@@ -41,8 +41,6 @@
 #ifndef Chimera_HISTORY_HPP
 #define Chimera_HISTORY_HPP
 
-#include "Chimera_RNGTraits.hpp"
-
 #include <Teuchos_RCP.hpp>
 #include <Teuchos_SerializationTraits.hpp>
 
@@ -54,7 +52,7 @@ namespace Chimera
  * \brief Encapsulation of a random walk history's state.
  */
 //---------------------------------------------------------------------------//
-template<class Scalar, class LO, class GO, class RNG>
+template<class Scalar, class LO, class GO>
 class History
 {
   public:
@@ -64,8 +62,6 @@ class History
     typedef Scalar                                    scalar_type;
     typedef LO                                        local_ordinal_type;
     typedef GO                                        global_ordinal_type;
-    typedef RNG                                       rng_type;
-    typedef Teuchos::RCP<RNG>                         RCP_RNG;
     //@}
 
     // Defaut constructor.
@@ -94,10 +90,6 @@ class History
     void setGlobalState( const GO global_state )
     { d_global_state = global_state; }
 
-    //! Set the RNG.
-    void setRNG( const RCP_RNG rng )
-    { d_rng = rng; }
-
     //! Terminate the history.
     void terminate()
     { d_active = false; }
@@ -107,16 +99,12 @@ class History
     { return d_weight; }
 
     //! Get the local history state.
-    LO LocalState() const
+    LO localState() const
     { return d_local_state; }
 
     //! Get the global history state.
     GO globalState() const 
     { return d_global_state; }
-
-    //! Get the RNG.
-    RCP_RNG rng() const
-    { return d_rng; }
 
     //! Get the active history state.
     bool active() const
@@ -133,9 +121,6 @@ class History
     // Global history state.
     GO d_global_state;
 
-    // RNG.
-    RCP_RNG d_rng;
-
     // Active history state.
     bool d_active;
 };
@@ -149,6 +134,86 @@ class History
 //---------------------------------------------------------------------------//
 
 #include "Chimera_History_def.hpp"
+
+//---------------------------------------------------------------------------//
+// SerializationTraits specializations.
+//---------------------------------------------------------------------------//
+namespace Teuchos
+{
[…]

### 2012-10-17 src/Chimera_History_def.hpp

@@ -47,12 +47,20 @@ namespace Chimera
 /*!
  * \brief Default constructor.
  */
-template<class Scalar, class LO, class GO, class RNG>
-History<Scalar,LO,GO,RNG>::History()
+template<class Scalar, class LO, class GO>
+History<Scalar,LO,GO>::History()
     : d_weight( 1.0 )
     , d_local_state( 0 )
     , d_global_state( 0 )
     , d_active( true )
+{ /* ... */ }
+
+//---------------------------------------------------------------------------//
+/*!
+ * \brief destructor.
+ */
+template<class Scalar, class LO, class GO>
+History<Scalar,LO,GO>::~History()
 { /* ... */ }
 
 //---------------------------------------------------------------------------//

### 2012-10-17 test/tstHistory.cpp

@@ -29,6 +29,7 @@
 #include <Teuchos_OpaqueWrapper.hpp>
 #include <Teuchos_TypeTraits.hpp>
 #include <Teuchos_Tuple.hpp>
+#include <Teuchos_Ptr.hpp>
 
 //---------------------------------------------------------------------------//
 // Tests
@@ -37,6 +38,81 @@ TEUCHOS_UNIT_TEST( History, history_test )
 {
     using namespace Chimera;
 
+    Teuchos::RCP<boost::mt19937> rng = RNGTraits<boost::mt19937>::create();
+
+    int num_rand = 10;
+    Teuchos::Array<double> randoms(num_rand);
+    for ( int i = 0; i < num_rand; ++i )
+    {
+	randoms[i] = 
+	    Teuchos::as<double>(RNGTraits<boost::mt19937>::generate(*rng)) /
+	    Teuchos::as<double>(RNGTraits<boost::mt19937>::max(*rng));
+    }
+
+    History<double,int,int> history;
+    TEST_ASSERT( history.active() );
+
+    double total = 2.0;    
+    history.setWeight( total );
+    TEST_ASSERT( history.weight() == total );
+
+    for ( int i = 0; i < num_rand; ++i )
+    {
+	history.addWeight( randoms[i] );
+	total += randoms[i];
+	TEST_ASSERT( history.weight() == total );
+    }
+
+    total = 2.0;
+    history.setWeight( total );
+    for ( int i = 0; i < num_rand; ++i )
+    {
+	history.multiplyWeight( randoms[i] );
+	total *= randoms[i];
+	TEST_ASSERT( history.weight() == total );
+    }
+
+    int local_state = 2929;
+    history.setLocalState( local_state );
+    TEST_ASSERT( history.localState() == local_state );
+
+    int global_state = 5943;
+    history.setGlobalState( global_state );
+    TEST_ASSERT( history.globalState() == global_state );
+
+    history.terminate();
+    TEST_ASSERT( !history.active() );
+}
+
+//---------------------------------------------------------------------------//
+TEUCHOS_UNIT_TEST( History, serialization_test )
+{
+    using namespace Chimera;
+
+    Teuchos::RCP<const Teuchos::Comm<int> > comm = 
+	Teuchos::DefaultComm<int>::getComm();
+    int comm_rank = comm->getRank();
+    
+    History<double,int,int> history;
+
+    double weight = 4.393939;
+    int local_state = 4941;
+    int global_state = 28390;
+
[…]

### 2012-10-17 src/Chimera_History.hpp
added a state constructor for the history class

@@ -67,6 +67,9 @@ class History
     // Defaut constructor.
     History();
 
+    // State constructor.
+    History( Scalar weight, LO local_state, GO global_state );
+
     // Destructor.
     ~History();
 
@@ -90,10 +93,6 @@ class History
     void setGlobalState( const GO global_state )
     { d_global_state = global_state; }
 
-    //! Terminate the history.
-    void terminate()
-    { d_active = false; }
-
     //! Get the history weight.
     Scalar weight() const
     { return d_weight; }
@@ -106,10 +105,6 @@ class History
     GO globalState() const 
     { return d_global_state; }
 
-    //! Get the active history state.
-    bool active() const
-    { return d_active; }
-
   private:
 
     // History weight.
@@ -120,9 +115,6 @@ class History
 
     // Global history state.
     GO d_global_state;
-
-    // Active history state.
-    bool d_active;
 };
 
 //---------------------------------------------------------------------------//

### 2012-10-17 src/Chimera_History_def.hpp

@@ -52,7 +52,17 @@ History<Scalar,LO,GO>::History()
     : d_weight( 1.0 )
     , d_local_state( 0 )
     , d_global_state( 0 )
-    , d_active( true )
+{ /* ... */ }
+
+//---------------------------------------------------------------------------//
+/*!
+ * \brief State constructor.
+ */
+template<class Scalar, class LO, class GO>
+History<Scalar,LO,GO>::History( Scalar weight, LO local_state, GO global_state )
+    : d_weight( weight )
+    , d_local_state( local_state )
+    , d_global_state( global_state )
 { /* ... */ }
 
 //---------------------------------------------------------------------------//

### 2012-10-17 test/tstHistory.cpp

@@ -34,7 +34,7 @@
 //---------------------------------------------------------------------------//
 // Tests
 //---------------------------------------------------------------------------//
-TEUCHOS_UNIT_TEST( History, history_test )
+TEUCHOS_UNIT_TEST( History, default_constructor_test )
 {
     using namespace Chimera;
 
@@ -50,7 +50,6 @@ TEUCHOS_UNIT_TEST( History, history_test )
     }
 
     History<double,int,int> history;
-    TEST_ASSERT( history.active() );
 
     double total = 2.0;    
     history.setWeight( total );
@@ -79,9 +78,54 @@ TEUCHOS_UNIT_TEST( History, history_test )
     int global_state = 5943;
     history.setGlobalState( global_state );
     TEST_ASSERT( history.globalState() == global_state );
+}
 
-    history.terminate();
-    TEST_ASSERT( !history.active() );
+//---------------------------------------------------------------------------//
+TEUCHOS_UNIT_TEST( History, state_constructor_test )
+{
+    using namespace Chimera;
+
+    Teuchos::RCP<boost::mt19937> rng = RNGTraits<boost::mt19937>::create();
+
+    int num_rand = 10;
+    Teuchos::Array<double> randoms(num_rand);
+    for ( int i = 0; i < num_rand; ++i )
+    {
+	randoms[i] = 
+	    Teuchos::as<double>(RNGTraits<boost::mt19937>::generate(*rng)) /
+	    Teuchos::as<double>(RNGTraits<boost::mt19937>::max(*rng));
+    }
+
+
+
+    double total = 2.0;    
+    int local_state = 2929;
+    int global_state = 5943;
+
+    History<double,int,int> history( total, local_state, global_state );
+    TEST_ASSERT( history.weight() == total );
+
+    for ( int i = 0; i < num_rand; ++i )
+    {
+	history.addWeight( randoms[i] );
+	total += randoms[i];
+	TEST_ASSERT( history.weight() == total );
+    }
+
+    total = 2.0;
+    history.setWeight( total );
+    for ( int i = 0; i < num_rand; ++i )
+    {
+	history.multiplyWeight( randoms[i] );
+	total *= randoms[i];
+	TEST_ASSERT( history.weight() == total );
+    }
+
+    history.setLocalState( local_state );
+    TEST_ASSERT( history.localState() == local_state );
+
+    history.setGlobalState( global_state );
+    TEST_ASSERT( history.globalState() == global_state );
 }
 
 //---------------------------------------------------------------------------//

### 2012-10-17 test/tstSamplingTools.cpp
adding local pdf sampling test. not finished yet

@@ -119,6 +119,12 @@ TEUCHOS_UNIT_TEST( SamplingTools, nonuniform_stratify_sample_test )
 }
 
 //---------------------------------------------------------------------------//
+TEUCHOS_UNIT_TEST( SamplingTools, uniform_local_pdf_sample_test )
+{
+
+}
+
+//---------------------------------------------------------------------------//
 // end tstBoostRNG.cpp
 //---------------------------------------------------------------------------//
 

### 2012-10-17 src/Chimera_SamplingTools.hpp
added a local pdf sampling test

@@ -76,7 +76,6 @@ class SamplingTools
     // Random sample a local discrete PDF for a new state index.
     template<class Scalar, class LO, class RNG>
     static LO sampleLocalDiscretePDF( 
-	const Scalar pdf_sum,
 	const Teuchos::ArrayView<Scalar>& pdf_values,
 	const Teuchos::ArrayView<LO>& pdf_indices,
 	const Teuchos::RCP<RNG>& rng );

### 2012-10-17 src/Chimera_SamplingTools_def.hpp

@@ -166,11 +166,13 @@ Teuchos::ArrayRCP<GO> SamplingTools::stratifySampleGlobalPDF(
  */
 template<class Scalar, class LO, class RNG>
 LO SamplingTools::sampleLocalDiscretePDF( 
-    const Scalar pdf_sum,
     const Teuchos::ArrayView<Scalar>& pdf_values,
     const Teuchos::ArrayView<LO>& pdf_indices,
     const Teuchos::RCP<RNG>& rng )
 {
+    Scalar pdf_sum = std::accumulate( 
+	pdf_values.begin(), pdf_values.end(), 0.0 );
+
     Scalar zeta = pdf_sum * 
 		  ( Teuchos::as<Scalar>(RNGTraits<RNG>::generate(*rng)) /
 		    Teuchos::as<Scalar>(RNGTraits<RNG>::max(*rng)) );

### 2012-10-17 test/tstSamplingTools.cpp

@@ -84,7 +84,7 @@ TEUCHOS_UNIT_TEST( SamplingTools, nonuniform_stratify_sample_test )
     int comm_rank = comm->getRank();
     int comm_size = comm->getSize();
 
-    // Setup linear operator distribution.
+    // Setup global PDF distribution.
     int local_num_rows = 10;
     int global_num_rows = local_num_rows*comm_size;
     Teuchos::RCP<const Tpetra::Map<int> > row_map = 
@@ -119,9 +119,33 @@ TEUCHOS_UNIT_TEST( SamplingTools, nonuniform_stratify_sample_test )
 }
 
 //---------------------------------------------------------------------------//
-TEUCHOS_UNIT_TEST( SamplingTools, uniform_local_pdf_sample_test )
+TEUCHOS_UNIT_TEST( SamplingTools, local_pdf_sample_test )
 {
+    using namespace Chimera;
 
+    // Setup a random number generator.
+    typedef boost::mt19937 RNG;
+    Teuchos::RCP<RNG> rng = RNGTraits<RNG>::create();
+
+    // Setup local PDF.
+    int pdf_size = 10;
+    Teuchos::Array<double> pdf_values( pdf_size, 0.0 );
+    pdf_values[ pdf_size-1 ] = 1.0;
+    Teuchos::Array<int> pdf_indices( pdf_size );
+    for ( int i = 0; i < pdf_size; ++i )
+    {
+	pdf_indices[i] = i+1;
+    }
+
+    // Sample the local PDF.
+    TEST_ASSERT( pdf_size == SamplingTools::sampleLocalDiscretePDF( 
+		     pdf_values(), pdf_indices(), rng ) );
+
+    // Reset the PDF to a different state and sample again.
+    std::fill( pdf_values.begin(), pdf_values.end(), 0.0 );
+    pdf_values[ pdf_size-3 ] = 1.0;
+    TEST_ASSERT( pdf_size-2 == SamplingTools::sampleLocalDiscretePDF( 
+		     pdf_values(), pdf_indices(), rng ) );
 }
 
 //---------------------------------------------------------------------------//

### 2012-10-18 doc/prelim/document/chapters/parallel_mc/parallel_mc.tex
finished adding Pauls changes

@@
 relationship and their simple geometry, they were able to relate the
-amount of compute time required to transport all source particles in
-relation to the number of source particles, the leakage fraction, the
-latency and bandwidth of communication, the ratio of local to global
-geometry size, and the total cross section of the system.
+amount of compute time required to transport all source particles to
+the number of source particles, the leakage fraction, the latency and
+bandwidth of communication, the ratio of local to global geometry
+size, and the total cross section of the system.
 
@@
 this because nearest neighbor communication patterns scale with the
-volume to surface sizes of the local domains. The larger the volume,
+volume and surface sizes of the local domains. The larger the volume,
 the more compute time is spent within the local volume and therefore
@@
 much overlap is feasible? Are there metrics related to the properties
-of the matrix such that we can provide guidelines for selecting the
-amount of overlap required? Furthermore, as Wagner and colleagues
-observed \citep{wagner_hybrid_2010}, with marginal domain overlap for
-reactor problems, the percentage of histories leaving the local domain
-can easily be reduced to less than 1\%. Evans and colleagues in their
+of the matrix including the eigenvalues and sparsity pattern such that
+we can provide guidelines for selecting the amount of overlap
+required? Furthermore, as Wagner and colleagues observed
+\citep{wagner_hybrid_2010}, with marginal domain overlap for reactor
+problems, the percentage of histories leaving the local domain can
+easily be reduced to less than 1\%. Evans and colleagues in their
 initial MCSA development typically used only 50 particle histories in

### 2012-10-18 doc/prelim/document/chapters/research_proposal/research_proposal.tex
incrporating more of Pauls changes. still a few more to go
finished adding Pauls changes

@@
 nonlinear problems. We seek to rigorously verify both of these methods
-and there implementation and will describe the means by which that
+and their implementation and will describe the means by which that
 will be accomplished. Furthermore, the numerical experiments performed
@@
 solvers. For the nonlinear solvers, we employ a new set of equations
-with readily available benchmark solutions.
+with readily available benchmark solutions. In both cases, we
+verification will be determined by comparing the locations of global
+minima and maxima in the solution as well as comparison of solution
+vector norms with respect to a specified tolerance.
 
@@
 Chapter~\ref{ch:nonlinear_problem} we posed questions related to the
-the FANM method. We will generate numerical experiments from these
+FANM method. We will generate numerical experiments from these
 questions by dividing them into 4 categories: domain overlap studies
@@
 properties of the system while varying others including the size of
-the linear system, its spectral radius, and other parameters related
-to the Eigenvalues and conditioning of the linear operator. By
-varying the domain overlap and measuring parallel performance with
-respect to these properties, we can develop guidelines for setting the
-overlap with respect to measurable properties of the linear
-operator. In addition, varying the amount of overlap will allow us to
-measure the additional memory costs incurred.
+the linear system, its spectral radius, its sparsity, its asymmetry,
+and other parameters related to the Eigenvalues, conditioning, and
+structure of the linear operator. By varying the domain overlap and
+measuring parallel performance with respect to these properties, we
+can develop guidelines for setting the overlap with respect to
+measurable properties of the linear operator. In addition, varying the
+amount of overlap will allow us to measure the additional memory costs
+incurred.
 

### 2012-10-24 doc/prelim/document/chapters/nonlinear_problem/nonlinear_problem.tex
spell check for prelim

@@
 formed. Modern nonlinear methods meet both of these requirements with
-Newton methods used in conjuction with Krylov methods for a robust,
+Newton methods used in conjunction with Krylov methods for a robust,
 fully implicit solution strategy. Furthermore, modern techniques exist

### 2012-10-24 doc/prelim/document/chapters/parallel_mc/parallel_mc.tex
spell check for prelim

@@
 achieve this by using Eq~(\ref{eq:adjoint_vector_relation}) as a
-constraint. Here we have 2 unkowns, $\ve{y}$ and $\ve{d}$ and
+constraint. Here we have 2 unknowns, $\ve{y}$ and $\ve{d}$ and
 therefore we require a second constraint to close the system. As a
@@
 visible in that we have now broken the binding of the convergence rate
-to the Central Limit Thereom. Here, the Monte Carlo solver is used to
+to the Central Limit Theorem. Here, the Monte Carlo solver is used to
 produce a correction from the residual, analogous to using the
@@
 entire domain is referred to as a set, and within a set overlap occurs
-among its subdomains. The original motivation was to decompose the
+among its sub-domains. The original motivation was to decompose the
 domain in a way that it remained in a physical cabinet in a large
@@
 system for a single material problem with a total cross section
-equivalent to that of homogonized light water reactor. Given the
+equivalent to that of homogenized light water reactor. Given the
 optimal local domain size, they observed for load balanced situations
 that nearest neighbor communication costs during transport are a small
-fraction of total compute time, leading to the conclusing that domain
+fraction of total compute time, leading to the conclusion that domain
 decomposition is feasible for large-scale Monte Carlo calculations.
@@
 \label{subsubsec:mc_load_balancing}
-Although domain decomposition was shown to be efficient in a prefectly
+Although domain decomposition was shown to be efficient in a perfectly
 load balanced situation in Siegel's work \citep{siegel_analysis_2012},

### 2012-10-24 doc/prelim/document/frontmatter/acks.tex
adding acks

Great thanks are owed to advisor and friend Paul Wilson for providing
me years of guidance and the opportunity to develop this work. Without
him, this work would have never been possible.

Tom Evans is responsible for presenting me with the seeds of this work
and for that I thank him. His mentoring over the past few years has
been invaluable.

Roger Pawlowski and other members of the Consortium for Advanced
Simulation of Light Water Reactors and Trilinos teams as well as many
staff members at Oak Ridge National Laboratory have provided
tremendous resources for my professional and technical development and
have greatly facilitated this work.

This work was performed under appointment to the Nuclear Regulatory
Commission Fellowship program at the University of Wisconsin - Madison
Department of Engineering Physics.

### 2012-10-24 doc/prelim/document/chapters/introduction/introduction.tex
updated intro

@@
 This preliminary report outlines in Chapter~\ref{ch:linear_problem}
-the conventional methods used in practice for solving linear and
-nonlinear problems to provide a mathematical basis upon which to build
-new algorithms that aim to solve some of the aforementioned
-issues. Parallel schemes for conventional methods are also provided
-for background and understanding of how current methods may or may not
-map to future hardware. In addition, some components of their parallel
+the conventional methods used in practice for solving linear problems
+to provide a mathematical basis upon which to build new algorithms
+that aim to solve some of the aforementioned issues. Parallel schemes
+for conventional methods are also provided for background and
+understanding of how current methods may or may not map to future
+hardware. In addition, some components of their parallel
 implementations may be applied to stochastic methods development. In

### 2012-10-24 doc/prelim/presentation/prelim_presentation.tex
presentation updates

@@
 \date{\today} 
-\title{Massively Parallel Monte Carlo Methods for Linear and Nonlinear
-Problems}
+\title{Massively Parallel Monte Carlo Methods for Discrete Linear and
+  Nonlinear Systems} 
 \begin{document}
@@
   \begin{itemize}
-    \item Predictive modeling and simulation capability yields nuclear
-      system designs that are safer and better performing.
+    \item Predictive modeling and simulation ehances engineering
+      capability to yield nuclear system designs that are safer and
+      better performing.
     \item Modern work focused on this task leverages multiple physics
@@
       decrease of memory/FLOPS predicted.
-    \item Compared to conventional methods, we will show that Monte
+    \item Compared to conventional methods, we aim show that Monte
       Carlo methods offer a memory savings.
@@
 \begin{frame}{Research Outline}
-
+  \begin{itemize}
+    \item Parallelization of classic Monte Carlo methods.
+    \item Parallel strategies taken from modern reactor physics
+      methods in Monte Carlo.
+    \item Research is required to explore how domain decomposition
+      patterns and the discrete system properties are related.
+    \item Research is required to explore how these methods perform
+      in modern multiple physics simulations where strong
+      nonlinearities are present.
+  \end{itemize}
 \end{frame}
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Conventional Solution Methods for Linear Problems}
+\begin{frame}{The Linear Problem}
 
-\end{frame}
+  We will seek solutions of the general linear problem:
 
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Linear Problem Preliminaries}
+  \[
+  \ve{A} \ve{x} = \ve{b}\:,
+  \]
+  \[
+  \ve{A} \in \mathbb{R}^{N \times N},\ \ve{A} : \mathbb{R}^{N}
+  \rightarrow \mathbb{R}^{N},\ \ve{x} \in \mathbb{R}^N,\ \ve{b} \in
+  \mathbb{R}^N\:.
+  \]
+
+  We will assert that $\ve{A}$ is \textit{nonsingular}. The solution
+  is then:
+  \[
+  \ve{x} = \ve{A}^{-1}\ve{b}\:.
+  \]
+
+  We can also define the residual of the system as:
+  \[
+  \ve{r} = \ve{b} - \ve{A}\ve{x}\:,
+  \]
+  where $\ve{r}=\ve{0}$ when an exact solution is found.
 
@@
 \begin{frame}{Stationary Methods}
+  General stationary methods are formed by splitting the linear
+  operator:
+  \[
+  \ve{A} = \ve{M} - \ve{N}\:.
+  \]
+
+  We can then reformulate the linear system as:
+  \[
+  \ve{x} = \ve{M}^{-1}\ve{N}\ve{x} + \ve{M}^{-1}\ve{b}\:.
+  \]
+
+  We identify $\ve{M}^{-1}\ve{N}$ as the \textit{iteration matrix},
+  $\ve{H}$, and apply iteration indices to the solution vectors to
+  give an iterative method.
+  \[
+  \ve{x}^{k+1} = \ve{H}\ve{x}^{k} + \ve{c}\:.
+  \]
+
+  The qualities of the iteration matrix dictate convergence. Defining
+  $\ve{e}^k = \ve{x}^k-\ve{x}$ as the error at the $k^{th}$ iterate
+  gives a recurrence relation:
+  \[
+  \ve{e}^{k+1} = \ve{H} \ve{e}^k\:,
+  \]
+
+  By diagonalizing $\ve{H}$ to extract its Eigenvalues, we find that 
+  \[
+  ||\ve{e}^{k}||_2 = \rho(\ve{H})^k ||\ve{e}^0||_2\:,
+  \]
+  and therefore we bound $\ve{H}$ by $\rho(\ve{H}) < 1$ for convergence.
 
@@
 \begin{frame}{Projection Methods}
+
+  We choose a search subspace $\mathcal{K}$ and a search subspace
+  $\mathcal{L}$ and determine the solution to the linear problem by
+  extracting the solution from the search subspace:
+  \[
+  \tilde{\ve{x}} = \ve{x}_0 +
+  \boldsymbol{\delta},\ \boldsymbol{\delta} \in \mathcal{K}\:,
+  \]
+  and by constraining it with the constraint subspace:
+  \[
+  \langle \tilde{\ve{r}},\ve{w} \rangle = 0,\ \forall \ve{w} \in
+  \mathcal{L}\:.
+  \]
 

### 2012-10-24 doc/prelim/presentation/prelim_presentation.tex
adding projection methods to presentation

@@
 
-  We choose a search subspace $\mathcal{K}$ and a search subspace
-  $\mathcal{L}$ and determine the solution to the linear problem by
-  extracting the solution from the search subspace:
+  We choose a \textit{search subspace} $\mathcal{K}$ and a
+  \textit{constraint subspace} $\mathcal{L}$ and determine the solution
+  to the linear problem by extracting the solution from the search
+  subspace:
   \[
@@
   \]
+
+  We can generate a more physical and geometric-based understanding of
+  these constraints by writing the new residual as:
+  \[
+  \tilde{\ve{r}} = \ve{r}_0 - \ve{A}\boldsymbol{\delta}
+  \]
+
+  If $\tilde{\ve{r}}$ is to be orthogonal to $\mathcal{L}$, then
+  $\ve{A}\boldsymbol{\delta}$ must be the projection of $\ve{r}_0$
+  onto the subspace $\mathcal{L}$ that eliminates the components of
+  the residual that exist in $\mathcal{L}$.
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{The Orthogonality Constraint}
+
+  \begin{figure}[htpb!]
+    \begin{center}
+      \scalebox{1.25}{
+        \input{orthogonal_residual.pdftex_t} }
+    \end{center}
+    \caption{\textbf{Orthogonality constraint of the new residual with
+        respect to $\mathcal{L}$.} \textit{By projecting $\ve{r}_0$ onto
+        the constraint subspace, we minimize the new residual by
+        removing those components.}}
+    \label{fig:linear_projection_constraint}
+  \end{figure}
+
+  From this we note:
+  \[
+  ||\tilde{\ve{r}}||_2 \leq ||\ve{r}_0||_2,\ \forall \ve{r}_0 \in
+  \mathbb{R}^N\:,
+  \]
+  meaning that the residual of the system will always be
+  \textit{minimized} with respect to the constraints.
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Projection Iteration}
+
+  Consider a matrix $\ve{V}$ to form a basis of $\mathcal{K}$ and a
+  matrix $\ve{W}$ to form a basis of $\mathcal{L}$.
+
+  \[
+  \boldsymbol{\delta} = \ve{V}\ve{y},\ \forall \ve{y} \in
+  \mathbb{R}^N\:.
+  \]
+
+  From the orthogonality constraint it then follows that:
+  \[
+  \ve{y} = (\ve{W}^T\ve{A}\ve{V})^{-1}\ve{W}^T\ve{r}_0\:,
+  \]
+
+  We can then form an iteration sequence:
+  \[
+  \ve{r}^k = \ve{b} - \ve{A}\ve{x}^k
+  \]
+  \[
+  \ve{y}^k = (\ve{W}^T\ve{A}\ve{V})^{-1}\ve{W}^T\ve{r}^k
+  \]
+  \[
+  \ve{x}^{k+1} = \ve{x}^k + \ve{V}\ve{y}^k\:,
+  \]
+
+  with $\ve{V}$ and $\ve{W}$ updated prior to each iteration.
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Krylov Subspace Methods}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{GMRES}
 

### 2012-10-24 doc/prelim/presentation/prelim_presentation.tex
adding pdf files for presentation

@@
 \begin{frame}{Stationary Methods}
+
   General stationary methods are formed by splitting the linear
@@
 
-  We can then reformulate the linear system as:
   \[
@@
 
-  We identify $\ve{M}^{-1}\ve{N}$ as the \textit{iteration matrix},
-  $\ve{H}$, and apply iteration indices to the solution vectors to
-  give an iterative method.
+  We identify $\ve{H} =\ve{M}^{-1}\ve{N}$ as the \textit{iteration
+    matrix}
+
   \[
@@
 
-  The qualities of the iteration matrix dictate convergence. Defining
-  $\ve{e}^k = \ve{x}^k-\ve{x}$ as the error at the $k^{th}$ iterate
-  gives a recurrence relation:
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Stationary Methods Convergence}
+
+  \begin{itemize}
+  \item The qualities of the iteration matrix dictate convergence 
+  \item Define $\ve{e}^k = \ve{x}^k-\ve{x}$ as the error at the
+    $k^{th}$ iterate
+  \item The iteration error is generated by a recurrence relation
+  \end{itemize}
+
   \[

### 2012-10-24 doc/prelim/presentation/prelim_presentation.tex
adding to presentation

@@
 
+%% colors
+\setbeamercolor{boxheadcolor}{fg=white,bg=UWRed}
+\setbeamercolor{boxbodycolor}{fg=black,bg=white}
+
 %%---------------------------------------------------------------------------%%
@@
   \\ University of Wisconsin - Madison
-  \\ \href{mailto:sslattery@wisc.edu}{\texttt{sslattery@wisc.edu}}
 }
@@
 \begin{frame}{Introduction}
+
   \begin{itemize}
     \item Predictive modeling and simulation ehances engineering
-      capability to yield nuclear system designs that are safer and
-      better performing.
+      capability
     \item Modern work focused on this task leverages multiple physics
-      simulation (CASL, NEAMS).
+      simulation (CASL, NEAMS)
     \item New hardware drives algorithm development (petascale and
-      exascale).
+      exascale)
     \item Monte Carlo methods have the potential to provide great
       improvements that permit finer simulations and better mapping to
-      future hardware.
+      future hardware
     \item A set of massively parallel Monte Carlo methods is proposed
       to advance multiple physics simulation on contemporary and
-      future leadership class machines.
+      future leadership class machines
   \end{itemize}
+
 \end{frame}
@@
 \begin{frame}{Physics-Based Motivation}
+
+  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
+    {Predictive nuclear reactor analysis enables...}
     \begin{itemize}
-    \item Predictive nuclear reactor analysis enables tighter design
-      tolerance for improved thermal performance and efficieny, higher
-      fuel burn-up, and high confidence in accident scenario models.
+    \item Tighter design tolerance for improved thermal performance
+      and efficieny
+    \item Higher fuel burn-up
+    \item High confidence in accident scenario models
+    \end{itemize}
+  \end{beamerboxesrounded}
+
+  \pause 
+  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
+  {Multiple physics simulations are complicated...}
+    \begin{itemize}
     \item Neutronics, thermal hydraulics, computational fluid
-      dynamics, and structural mechanics among other physics
-      contribute to these predictive models.
-    \item Couplings are complicated and consistent models yield
-      nonlinearities in the variables through feedback effects.
-    \item  Capturing  these   effects  requires  tremendous  resources
-      with $O(\sn{1}{9})$ element meshes and $O(100,000)+$ cores used
-      in today's simulations.
+      dynamics, structural mechanics, and many other physics
+    \item Consistent models yield nonlinearities in the variables
+      through feedback effects
+    \item Tremendous computational resources are required with
+      $O(\sn{1}{9})$ element meshes and $O(100,000)+$ cores used in
+      today's simulations.
     \end{itemize}
+  \end{beamerboxesrounded}
+
 \end{frame}
@@
     \end{center}
-    \caption{\textbf{Multiphysics dependency analysis of departure from
-        nucleate boiling.} \textit{A neutronics solution is required to
-        compute power generation in the fuel pins, fluid dynamics is
-        required to characterize boiling and fluid temperature and
-        density, heat transfer is required to compute the fuel and
-        cladding temperature, and the nuclear data modified with the
-        temperature and density data. Strong coupling among the
-        variables creates strong nonlinearities.}}
+    \caption{\textbf{Multiphysics dependency analysis of departure
+        from nucleate boiling.} }
     \label{fig:dnb_example}
@@
 
-  General stationary methods are formed by splitting the linear
-  operator:
+  \begin{itemize}
+  \item General stationary methods are formed by splitting the linear
+    operator
+  \end{itemize}
+
   \[
@@
   \[
-  \ve{e}^{k+1} = \ve{H} \ve{e}^k\:,
+  \ve{e}^{k+1} = \ve{H} \ve{e}^k\:
   \]
 
-  By diagonalizing $\ve{H}$ to extract its Eigenvalues, we find that 
+  \begin{itemize}
+  \item We diagonalize $\ve{H}$ to extract its Eigenvalues
+  \end{itemize}
+
   \[
@@
   \]
-  and therefore we bound $\ve{H}$ by $\rho(\ve{H}) < 1$ for convergence.
+
+  \begin{itemize}
+  \item We bound $\ve{H}$ by $\rho(\ve{H}) < 1$ for convergence
+  \end{itemize}
 
@@
     \caption{\textbf{Orthogonality constraint of the new residual with
-        respect to $\mathcal{L}$.} \textit{By projecting $\ve{r}_0$ onto
-        the constraint subspace, we minimize the new residual by
-        removing those components.}}
-    \label{fig:linear_projection_constraint}
+        respect to $\mathcal{L}$.} }
   \end{figure}
 
-  From this we note:
+  \pause
+  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
+    {Minimization Property}
+
+    The residual of the system will always be \textit{minimized} with
+    respect to the constraints
     \[
@@
     \]
-  meaning that the residual of the system will always be
-  \textit{minimized} with respect to the constraints.
+  \end{beamerboxesrounded}
 

### 2012-10-25 doc/prelim/presentation/prelim_presentation.tex
updating dnb example
finished general projection methods in presentation

@@
   \begin{itemize}
-    \item Predictive modeling and simulation ehances engineering
+    \item Predictive modeling and simulation enhances engineering
       capability
@@
     \item Tighter design tolerance for improved thermal performance
-      and efficieny
+      and efficiency
     \item Higher fuel burn-up
@@
 \begin{frame}{Physics-Based Motivation: DNB}
+
+  \begin{columns}
+
+    \begin{column}{0.35\textwidth}
       \begin{figure}[htpb!]
         \begin{center}
-      \scalebox{1}{ \input{dnb_example.pdftex_t} }
+          \scalebox{1}{ \input{dnb_schematic.pdftex_t} }
+        \end{center}
+        \caption{\textbf{Departure from nucleate boiling scenario.} }
+      \end{figure}
+    \end{column}
+
+    \pause
+
+    \begin{column}{0.65\textwidth}
+      \begin{figure}[htpb!]
+        \begin{center}
+          \scalebox{0.8}{ \input{dnb_example.pdftex_t} }
         \end{center}
@@
             from nucleate boiling.} }
-    \label{fig:dnb_example}
       \end{figure}
+    \end{column}
+
+  \end{columns}
+
 \end{frame}
@@
 \begin{frame}{Hardware-Based Motivation}
+
   \begin{itemize}
-    \item Modern hardware is moving in two directions: lightweight
-      machines and heterogeneous machines characterized by low power
-      and high concurrency.
-    \item High concurrency and low cost units means a higher potential
-      for both soft and hard failures.
-    \item Monte Carlo methods bury soft failures within the variance
-      of the tallies while hard failures are high variance events.
-    \item New machines will also be memory restricted with a continued
-      decrease of memory/FLOPS predicted.
-    \item Compared to conventional methods, we aim show that Monte
-      Carlo methods offer a memory savings.
+  \item Modern hardware is moving in two directions:
+    \begin{itemize}
+    \item Lightweight machines
+    \item Heterogeneous machines
+    \item Both characterized by low power and high concurrency
     \end{itemize}
+  \item Some issues:
+    \begin{itemize}
+    \item Higher potential for both soft and hard failures
+    \item Memory restrictions are expected with a continued decrease
+      in memory/FLOPS
+    \end{itemize}
+  \item Potential resolution from Monte Carlo:
+    \begin{itemize}
+    \item Soft failures buried within the tally variance
+    \item Hard failures are high variance events
+    \item Possible memory savings over conventional methods
+    \end{itemize}
+  \end{itemize}
+
 \end{frame}
@@
   \begin{itemize}
-    \item Parallelization of classic Monte Carlo methods.
+    \item Parallelization of classic Monte Carlo methods for discrete
+      systems
+      \begin{itemize}
       \item Parallel strategies taken from modern reactor physics
-      methods in Monte Carlo.
-    \item Research is required to explore how domain decomposition
-      patterns and the discrete system properties are related.
-    \item Research is required to explore how these methods perform
-      in modern multiple physics simulations where strong
-      nonlinearities are present.
+        methods
+      \item Research is required to explore varying parallel
+        strategies
+      \item Scalability is of concern
+      \end{itemize}
+    \item Development of a novel nonlinear solver leveraging Monte Carlo
+      \begin{itemize}
+        \item Application to nonlinear problems of interest
+        \item Memory benefits
+        \item Performance benefits
+      \end{itemize}
   \end{itemize}
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{The Linear Problem}
+\begin{frame}{Linear Operator Equations}
 
-  We will seek solutions of the general linear problem:
+  \begin{itemize}
+  \item We seek solutions of the general linear operator equation
+  \end{itemize}
 
   \[
-  \ve{A} \ve{x} = \ve{b}\:,
+  \ve{A} \ve{x} = \ve{b}\:
   \]
@@
   \rightarrow \mathbb{R}^{N},\ \ve{x} \in \mathbb{R}^N,\ \ve{b} \in
-  \mathbb{R}^N\:.
+  \mathbb{R}^N\:
   \]
 
-  We will assert that $\ve{A}$ is \textit{nonsingular}. The solution
-  is then:
   \[
-  \ve{x} = \ve{A}^{-1}\ve{b}\:.
+  \ve{r} = \ve{b} - \ve{A}\ve{x}\:
   \]
 
-  We can also define the residual of the system as:
+  \begin{itemize}
+  \item $\ve{r}=\ve{0}$ when an exact solution is found.
+  \end{itemize}
+
+  \pause
+  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
+    {A Requirement}
+    Assert that $\ve{A}$ is \textit{nonsingular}. The solution is then:
     \[
-  \ve{r} = \ve{b} - \ve{A}\ve{x}\:,
+    \ve{x} = \ve{A}^{-1}\ve{b}\:
     \]
-  where $\ve{r}=\ve{0}$ when an exact solution is found.
+  \end{beamerboxesrounded}
 
@@
 
-  We identify $\ve{H} =\ve{M}^{-1}\ve{N}$ as the \textit{iteration
-    matrix}
+  \begin{itemize}
+  \item We identify $\ve{H} =\ve{M}^{-1}\ve{N}$ as the
+    \textit{iteration matrix}
+  \end{itemize}
 
@@
     $k^{th}$ iterate
-  \item The iteration error is generated by a recurrence relation
   \end{itemize}
@@
 
-  We choose a \textit{search subspace} $\mathcal{K}$ and a
-  \textit{constraint subspace} $\mathcal{L}$ and determine the solution
-  to the linear problem by extracting the solution from the search
-  subspace:
+  \begin{itemize}
+    \item Powerful class of iterative methods
+    \item Provides theory that encapsulates most other iterative
+      methods 
+    \item Leveraged in many modern physics codes at the petascale
+  \end{itemize}
+
+  \pause
+  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
+    {Search Subspace $\mathcal{K}$} 
+    Extract the solution from the search subspace:
     \[
     \tilde{\ve{x}} = \ve{x}_0 +
-  \boldsymbol{\delta},\ \boldsymbol{\delta} \in \mathcal{K}\:,
+    \boldsymbol{\delta},\ \boldsymbol{\delta} \in \mathcal{K}\:
     \]
-  and by constraining it with the constraint subspace:
+  \end{beamerboxesrounded}
+
+  \pause
+  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
+    {Constraint Subspace $\mathcal{L}$} 
+    Constrain the extraction with the constraint subspace by asserting
+    orthogonality with the residual:
     \[
     \langle \tilde{\ve{r}},\ve{w} \rangle = 0,\ \forall \ve{w} \in
-  \mathcal{L}\:.
+    \mathcal{L}\:
     \]  
-
-  We can generate a more physical and geometric-based understanding of
-  these constraints by writing the new residual as:
-  \[
-  \tilde{\ve{r}} = \ve{r}_0 - \ve{A}\boldsymbol{\delta}
-  \]
-
-  If $\tilde{\ve{r}}$ is to be orthogonal to $\mathcal{L}$, then
-  $\ve{A}\boldsymbol{\delta}$ must be the projection of $\ve{r}_0$
-  onto the subspace $\mathcal{L}$ that eliminates the components of
-  the residual that exist in $\mathcal{L}$.
+  \end{beamerboxesrounded}
 
@@
 \begin{frame}{The Orthogonality Constraint}
+
+  \[
+  \tilde{\ve{r}} = \ve{r}_0 - \ve{A}\boldsymbol{\delta}
+  \]
 
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Projection Iteration}
+\begin{frame}{Putting it All Together}
 
-  Consider a matrix $\ve{V}$ to form a basis of $\mathcal{K}$ and a
-  matrix $\ve{W}$ to form a basis of $\mathcal{L}$.
+  \begin{itemize}
+  \item Choose $\ve{V}$ as a basis of $\mathcal{K}$ and $\ve{W}$ as a
+    basis of $\mathcal{L}$
+  \end{itemize}
 
@@
   \boldsymbol{\delta} = \ve{V}\ve{y},\ \forall \ve{y} \in
-  \mathbb{R}^N\:.
+  \mathbb{R}^N
   \]
 
-  From the orthogonality constraint it then follows that:
   \[
-  \ve{y} = (\ve{W}^T\ve{A}\ve{V})^{-1}\ve{W}^T\ve{r}_0\:,
+  \ve{y} = (\ve{W}^T\ve{A}\ve{V})^{-1}\ve{W}^T\ve{r}_0
   \]
 
-  We can then form an iteration sequence:
+  \pause
+  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
+    {Projection Method Iteration}
   \[
@@
   \[
-  \ve{x}^{k+1} = \ve{x}^k + \ve{V}\ve{y}^k\:,
+  \ve{x}^{k+1} = \ve{x}^k + \ve{V}\ve{y}^k\:
   \]
-
-  with $\ve{V}$ and $\ve{W}$ updated prior to each iteration.
+  \[
+  Update\ \ve{V}\ and\ \ve{W}
+  \]
+  \end{beamerboxesrounded}
 

### 2012-10-25 doc/prelim/presentation/prelim_presentation.tex
prelim updates

@@
     \item Hard failures are high variance events
-    \item Possible memory savings over conventional methods
+    \item Memory savings over conventional methods
     \end{itemize}
@@
   \begin{itemize}
-    \item Parallelization of classic Monte Carlo methods for discrete
-      systems
+    \item Parallelization of Monte Carlo methods for discrete systems
       \begin{itemize}
@@
       \end{itemize}
-    \item Development of a novel nonlinear solver leveraging Monte Carlo
+    \item Development of a nonlinear solver leveraging Monte Carlo
       \begin{itemize}
@@
 
+  \[
+  \mathcal{K}_m(\ve{A},\ve{r}_0) = span\{\ve{r}_0, \ve{A}\ve{r}_0,
+  \ve{A}^2\ve{r}_0, \dots, \ve{A}^{m-1}\ve{r}_0\}
+  \]
+  \[
+  \mathcal{L} = \ve{A} \mathcal{K}_m(\ve{A},\ve{r}_0)
+  \]
+
+  \begin{itemize}
+  \item Yields the normal system $\ve{A}^T\ve{A}\ve{x} =
+    \ve{A}^T\ve{b}$
+  \item Must generate an orthonormal basis $\ve{V}_m \in \mathbb{R}^{N
+    \times m}$ for $\mathcal{K}_m(\ve{A},\ve{r}_0)$
+  \item $\ve{W}_m = \ve{A}\ve{V}_m$
+  \item Typically choose a Gram-Schmidt-like procedure such as
+    Arnoldi or Lanzcos
+  \end{itemize}
+
 \end{frame}

### 2012-10-28 doc/prelim/presentation/prelim_presentation.tex
finished linear methods section

@@
 \usepackage{textcomp}
+\usepackage{algorithm,algorithmic}
 \definecolor{listinggray}{gray}{0.9}
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{GMRES}
+\begin{frame}[fragile]{GMRES}
+
+\begin{algorithm}[H]
+  \begin{algorithmic}[1]
+    \STATE $\ve{r}_0 := \ve{b}-\ve{A}\ve{x}_0$
+    \STATE $\beta := ||\ve{r}_0||_2$
+    \STATE $\ve{v}_1 := \ve{r}_0 / \beta$
+    \COMMENT{Create the orthonormal basis for the Krylov subspace}
+    \FOR{$j = 1, 2, \cdots, m$}
+    \STATE $h_{ij} \leftarrow \langle w_j,v_j \rangle$
+    \STATE $w_j \leftarrow w_j - h_{ij}v_i$
+    \ENDFOR
+    \STATE $h_{j+1,j} \leftarrow ||w_j||_2$
+    \STATE $v_{j+1} \leftarrow w_j / h_{j+1,j}$
+    \COMMENT{Apply the orthogonality constraints}
+    \STATE $\ve{y}_m \leftarrow argmin_y ||\beta \ve{e}_1 - \ve{H}_m\
+    \ve{y}||_2 $
+    \STATE $\ve{x}_m \leftarrow \ve{x}_0 + \ve{V_m} \ve{y}_m$
+  \end{algorithmic}
+  \caption{GMRES Iteration}
+\end{algorithm}
 
@@
 \begin{frame}{Parallel Projection Methods}
+
+  \begin{itemize}
+  \item Parallel vector update
+  \end{itemize}
+  \[
+  \ve{y}[n] \leftarrow \ve{y}[n] + a * \ve{x}[n],\ \forall n \in [1,N_g]
+  \]
+  \[
+  \ve{y}[n] \leftarrow \ve{y}[n] + a * \ve{x}[n],\ \forall n \in [1,N_l]
+  \]
+
+  \begin{itemize}
+  \item Parallel dot product
+  \end{itemize}
+  \[
+  d_l = \ve{y}_l \cdot \ve{x}_l,\ d_g = \sum_p d_l
+  \]
+
+  \begin{itemize}
+  \item Parallel vector norm
+  \end{itemize}
+  \[
+  ||x||_{\infty,l} = \max_n \ve{y}[n],\ \forall n \in [1,N_l]
+  \]
+  \[
+  ||x||_{\infty,g} = \max_p ||x||_{\infty,l}
+  \]
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Parallel Matrix-Vector Multiplication}
+
+\begin{figure}[htpb!]
+  \begin{center}
+    \scalebox{0.75}{
+      \input{partitioned_matrix.pdftex_t} }
+  \end{center}
+  \caption{\textbf{Matrix-vector multiply $\ve{A}\ve{x}=\ve{y}$
+      operation on 3 processors.}
+  \label{fig:partitioned_matvec_multiply} }
+\end{figure}
+
+\pause
+\begin{figure}[htpb!]
+  \begin{center}
+    \scalebox{0.75}{
+      \input{matvec_proc_1.pdftex_t} }
+  \end{center}
+  \caption{\textbf{Components of multiply operation owned by process
+      1.} }
+  \label{fig:matvec_proc_1}
+\end{figure}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Parallel Projection Method Notes}
+
+\begin{itemize}
+  \item Global reduction operations observed not to impede scalability
+    \begin{itemize}
+    \item Dot product
+    \item Vector norms
+    \end{itemize}
+  \item Nearest neighbor computations have poor algorithmic strong
+    scaling
+    \begin{itemize}
+      \item Matrix-vector multiply
+      \item Weak scaling is better
+    \end{itemize}
+\end{itemize}
 

### 2012-10-28 src/Chimera_AdjointNeumannUlamSolver_def.hpp

@@ -139,7 +139,7 @@ void AdjointNeumannUlamSolver<Scalar,LO,GO,RNG>::walk()
 	    // Get the current history state.
 	    global_state = bank.top().globalState();
 	    local_state = state_map->getLocalElement( global_state );
-
+	    
 	    // Update LHS.
 	    lhs_view[local_state] += bank.top().weight();
 

### 2012-10-28 src/Chimera_OperatorTools_def.hpp

@@ -171,9 +171,9 @@ Scalar OperatorTools::spectralRadius(
 
     const int nev = 1;
     const int block_size = 1;
-    const int num_blocks = 3;
+    const int num_blocks = 10;
     const int max_restarts = 100;
-    const Scalar tol = 1.0e-6;
+    const Scalar tol = 1.0e-4;
 
     Teuchos::ParameterList krylovschur_params;
     krylovschur_params.set( "Which", "LM" );

### 2012-10-28 test/tstAdjointNeumannUlamSolver.cpp

@@ -53,7 +53,7 @@ TEUCHOS_UNIT_TEST( AdjointNeumannUlamSolver, adjoint_neumannulam_test )
     int problem_size = N*N;
     double dx = 0.01;
     double dy = 0.01;
-    double dt = 0.1;
+    double dt = 0.5;
     double alpha = 0.01;
     Teuchos::RCP<const Tpetra::Map<int> > row_map = 
 	Tpetra::createUniformContigMap<int,int>( problem_size, comm );
@@ -223,7 +223,7 @@ TEUCHOS_UNIT_TEST( AdjointNeumannUlamSolver, adjoint_neumannulam_test )
     // Check the solution vector.
     Teuchos::ArrayRCP<const double> X_view = 
 	linear_problem->getLHS()->get1dView();
-//    std::cout << X_view() << std::endl;
+    std::cout << X_view() << std::endl;
 }
 
 //---------------------------------------------------------------------------//

### 2012-10-28 doc/prelim/presentation/prelim_presentation.tex
added mc preliminaries

@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Parallel Projection Method Notes}
+\begin{frame}{Projection Method Notes}
 
@@
     \end{itemize}
+
 \end{itemize}
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Monte Carlo Solution Methods for Linear Problems}
+\begin{frame}{Monte Carlo Solution Methods for Discrete Linear Systems}
+
+  \begin{itemize}
+    \item First proposed by J. Von Neumann and S.M. Ulam in the 1940's
+    \item Earliest published reference in 1950
+    \item General lack of published work
+    \item Modern work by Evans and others has yielded new applications
+    \item No indication of parallel methods in the literature
+  \end{itemize}
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Monte Carlo Linear Solver Preliminaries}
+
+  \begin{itemize}
+  \item Split the operator
+  \end{itemize}
+
+  \[
+  \ve{H} = \ve{I} - \ve{A}
+  \]
+
+  \[
+  \ve{x} = \ve{H} \ve{x} + \ve{b}
+  \]
+
+  \begin{itemize}
+  \item Generate the \textit{Neumann series}
+  \end{itemize}
+
+  \[
+  \ve{A}^{-1} = (\ve{I}-\ve{H})^{-1} = \sum_{k=0}^{\infty} \ve{H}^k
+  \]
+
+  \begin{itemize}
+  \item Require $\rho(\ve{H}) < 1$ for convergence
+  \end{itemize}
+
+  \[
+  \ve{A}^{-1}\ve{b} = \sum_{k=0}^{\infty} \ve{H}^k\ve{b} = \ve{x}
+  \]
 
@@
 \begin{frame}{Monte Carlo Linear Solver Preliminaries}
+
+  \begin{itemize}
+  \item Expand the Nuemann series
+  \end{itemize}
+
+  \[
+  x_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
+  \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}
+  \]
+
+  \begin{itemize}
+  \item Define a sequence of state transitions
+  \end{itemize}
+
+  \[
+  \nu = i \rightarrow i_1 \rightarrow \cdots \rightarrow i_{k-1}
+  \rightarrow i_{k}
+  \]
+
+  \begin{itemize}
+  \item Define the \textit{Neumann-Ulam decomposition}\footnote{The
+    Hadamard product $\ve{A} = \ve{B} \circ \ve{C}$ is defined
+    element-wise as $a_{ij} = b_{ij} c_{ij}$.}
+  \end{itemize}
+
+  \[
+  \ve{H} = \ve{P} \circ \ve{W}
+  \]
 

### 2012-10-28 doc/prelim/presentation/prelim_presentation.tex
updated presentation

@@
 
+  \pause 
   \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
@@
 
+  \begin{itemize}
+    \item Compute row-normalized transition probabilities and weights
+  \end{itemize}
+
+  \[
+  p_{ij} = \frac{|h_{ij}|}{\sum_j |h_{ij}|},\ w_{ij} =
+  \frac{h_{ij}}{p_{ij}}
+  \]
+
+  \begin{itemize}
+    \item Generate an expectation value for the solution
+  \end{itemize}
+
+  \[
+  W_{m} = \sum_{m=0}^k w_{i,i_1} w_{i_1,i_2} \cdots w_{i_{m-1},i_m}
+  \]
+  \[
+  X_{\nu}(i_0 = i) = \sum_{m=0}^k W_{m} b_{i_m}
+  \]
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Direct Method}
+
+  \begin{itemize}
+    \item Compute the probability of a particular random walk permutation
+  \end{itemize}
+
+  \[
+  P_{\nu} = p_{i,i_1} p_{i_1,i_2} \cdots p_{i_{k-1},i_k}
+  \]
+
+  \begin{itemize}
+  \item Generate the estimator
+  \end{itemize}
+
+  \[
+  E\{X(i_0 = i)\} = \sum_{\nu} P_{\nu} X_{\nu}
+  \]
+
+  \begin{itemize}
+  \item Check that we recover the exact solution
+  \end{itemize}
+
+  \[
+  \begin{split}
+    E\{X(i_0 = i)\}
+    &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
+    \sum_{i_k}^{N} p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
+    w_{i,i_1}w_{i_1,i_2}\ldots w_{i_{k-1},i_k} b_{i_k}\\ &= x_i\:,
+  \end{split}
+  \]
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Direct Method: Evolution of a Solution}
+
 \end{frame}
@@
 \begin{frame}{Adjoint Method}
+
+  \begin{itemize}
+  \item Solve the adjoint linear system
+  \end{itemize}
+
+  \[
+  \ve{A}^T \ve{y} = \ve{d}
+  \]
+
+  \[
+  \ve{y} = \ve{H}^T \ve{y} + \ve{d}
+  \]
+
+  \begin{itemize}
+  \item Set the adjoint constraint
+  \end{itemize}
+
+  \[
+  \langle \ve{A}^T \ve{x}, \ve{y} \rangle = \langle \ve{x}, \ve{A}
+  \ve{y} \rangle
+  \]
+
+  \[
+  \langle \ve{x}, \ve{d} \rangle = \langle \ve{y}, \ve{b} \rangle
+  \]
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Adjoint Method}
+
+  \begin{itemize}
+  \item Generate the Neumann series for the adjoint operator
+  \end{itemize}
+
+  \[
+  \ve{y} = (\ve{I} - \ve{H}^T)^{-1} \ve{d}
+  \]
+
+  \[
+  \ve{y} = \sum_{k=0}^{\infty} (\ve{H}^T)^k\ve{d}
+  \]
+
+  \begin{itemize}
+  \item Expand the series
+  \end{itemize}
+
+  \[
+  y_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
+  \sum_{i_k}^{N}h_{i_k,i_{k-1}}\ldots h_{i_2,i_1} h_{i_1,i} d_{i_k}
+  \]
+
+  \begin{itemize}
+  \item Pick another constraint to yield the original solution
+  \end{itemize}
+
+  \[
+  \ve{d} = \boldsymbol{\delta}_i,\ \langle \ve{y}, \ve{b} \rangle =
+  \langle \ve{x}, \boldsymbol{\delta}_i \rangle = x_i
+  \]
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Adjoint Method}
+
+  \begin{itemize}
+  \item Use the adjoint Neumann-Ulam decomposition
+  \end{itemize}
+
+  \[
+  \ve{H}^{T} = \ve{P} \circ \ve{W}
+  \]
+
+  \[
+  p_{ij} = \frac{|h_{ji}|}{\sum_j |h_{ji}|},\ w_{ij} =
+  \frac{h_{ji}}{p_{ij}}
+  \]
+
+  \begin{itemize}
+  \item Build the estimator and expectation value
+  \end{itemize}
+
+  \[
+  X_{\nu} = \sum_{m=0}^k W_{m} b_{i_0} \delta_{i,i_m}
+  \]
+
+  \[
+  \begin{split}
+    E\{X_j\} &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
+    \sum_{i_k}^{N} b_{i_0} h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}
+    \delta_{i_k,j} \\ &= x_{j}\:,
+  \end{split}
+  \]
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Adjoint Method: Evolution of a Solution}
 
@@
 
+  \begin{itemize}
+  \item Neumann-Ulam methods bound by the Central Limit Theorem
+  \item Halton proposed an iterative residual method
+  \item Iteration error decoupled from Monte Carlo error
+  \item Exponential convergence
+  \end{itemize}
+
+  \[
+  \ve{r}^k = \ve{b} - \ve{A}\ve{x}^k
+  \]
+  \[
+  \ve{A}\boldsymbol{\delta}^{k} = \ve{r}^{k}
+  \]
+  \[
+  \ve{x}^{k+1} = \ve{x}^k + \boldsymbol{\delta}^{k}
+  \]
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Monte Carlo Synthetic-Acceleration}
+  \begin{itemize}
+  \item Split the operator to yield Richardson's iteration
+  \end{itemize}
+
+  \[
+  \ve{x} = (\ve{I} - \ve{A})\ve{x} + \ve{b}
+  \]
+  \[
+  \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}
+  \]
+
+  \begin{itemize}
+  \item Define the iteration error
+  \end{itemize}
+
+  \[
+  \delta \ve{x}^k = \ve{x} - \ve{x}^k
+  \]
+  \[
+  \delta \ve{x}^{k+1} = (\ve{I} - \ve{A})\delta \ve{x}^k
+  \]
+
 \end{frame}
@@
 
+  \begin{itemize}
+  \item Subtract $(\ve{I} - \ve{A})\delta \ve{x}^{k+1}$
+  \end{itemize}
+
+  \[
+  \begin{split}
+    \ve{A}\delta \ve{x}^{k+1} &= (\ve{I} -
+    \ve{A})(\ve{x}^{k+1}-\ve{x}^{k}) \\ &= \ve{r}^{k+1}
+  \end{split}
+  \]
+
+  \begin{itemize}
+  \item The following converges in one iteration with exact inversion
+    of $\ve{A}$:
+  \end{itemize}
+
+  \[
+  \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}
+  \]
+  \[
+  \ve{A} \delta \ve{x}^{k+1} = \ve{r}^{k+1}
+  \]
+  \[
+  \ve{x} = \ve{x}^{k+1} + \delta \ve{x}^{k+1}
+  \]
+
 \end{frame}
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Parallelization of Stochastic Methods}
+\begin{frame}{Monte Carlo Synthetic-Acceleration}
+
+  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
+    {MCSA Iteration}
+
+    \[
+    \ve{x}^{k+1/2} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}
+    \]
+    \[
+    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}
+    \]
+    \[
+    \hat{\ve{A}}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}
+    \]
+    \[
+    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}
+    \]
+
+  \end{beamerboxesrounded}
+
+  \begin{itemize}
+  \item Adjoint Neumann-Ulam solver computes the correction
+  \item Decouples Monte Carlo error from solution error
+  \item Exponential convergence
+  \item Demonstrated by Evans and colleagues to be competitive with
+    Krylov methods
+  \end{itemize}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Preconditioning Monte Carlo Methods}
+
+  \begin{itemize}
+    \item No symmetry requirements
+    \item Require $\rho(\ve{H}) < 1$
+    \item Choose Jacobi preconditioning at a minimum
+  \end{itemize}
+
+  \[
+  \ve{M} = diag(\ve{A})
+  \]
+  \[
+  \ve{M}^{-1}\ve{A}\ve{x} = \ve{M}^{-1}\ve{b}
+  \]
+
+  \begin{itemize}
+  \item Yields a preconditioned MCSA iteration with no in-state
+    transitions
+  \end{itemize}
+
+  \[
+  \ve{x}^{k+1/2} = (\ve{I} - \ve{M}^{-1}\ve{A})\ve{x}^k + \ve{b}
+  \]
+  \[
+  \ve{r}^{k+1/2} = \ve{b} - \ve{M}^{-1}\ve{A}\ve{x}^{k+1/2}
+  \]
+  \[
+  \ve{M}^{-1}\ve{A}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}
+  \]
+  \[
+  \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}
+  \]
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Parallelization of Monte Methods}
+
+  \begin{itemize}
+  \item No literature observed for parallel Neumann-Ulam solvers
+  \item Numerous references for modern parallel Monte Carlo methods
+    in reactor physics
+  \item MCSA parallelism comes from parallel matrix/vector operations
+  \end{itemize}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Domain Decomposition}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Multiple-Set Overlapping-Domain Decomposition}
+
+  \begin{figure}[htpb!]
+    \begin{center}
+      \scalebox{1.0}{
+        \input{msod_example.pdftex_t} }
+    \end{center}
+    \caption{\textbf{Overlapping domain example illustrating how domain
+        overlap can reduce communication costs.}}
+  \end{figure}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Domain-to-Domain Communication}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Load Balancing}
+
+  \begin{figure}[htpb!]
+    \begin{center}
+      \scalebox{1.0}{
+        \input{procassini_example.pdftex_t} }
+    \end{center}
+    \caption{\textbf{Example illustrating how domain decomposition can
+        create load balance issues in Monte Carlo.}}
+    \label{fig:procassini_example}
+  \end{figure}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Reproducability}
+
+  \begin{figure}[htpb!]
+    \begin{center}
+      \scalebox{1.0}{
+        \input{gentile_example.pdftex_t} }
+    \end{center}
+    \caption{\textbf{Gentile's example illustrating how domain
+        decomposition can create reproducibility issues in Monte Carlo.}}
+    \label{fig:gentile_example}
+  \end{figure}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Parallel Adjoint Method}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Parallel MCSA}
 
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Monte Carlo Nonlinear Solver Preliminaries}
+\begin{frame}{Nonlinear Preliminaries}
 
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Inexact Newton Methods}
+\begin{frame}{Nonlinear Preliminaries}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Newton-Krylov Methods}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Newton-Krylov Methods}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Matrix-Free Approximation}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Automatic Differentiation}
 
@@
 \begin{frame}{The FANM Method}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Jacobian Storage vs. Subspace Storage}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Parallel FANM Method}
 
@@
 \begin{frame}{Experimental Framework}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Progress to Date}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Progress to Date}
 

### 2012-10-31 doc/prelim/presentation/prelim_presentation.tex
adding solution evolution pngs

@@
     \FOR{$j = 1, 2, \cdots, m$}
-    \STATE $h_{ij} \leftarrow \langle w_j,v_j \rangle$
-    \STATE $w_j \leftarrow w_j - h_{ij}v_i$
+    \STATE $\ve{w}_j := \ve{A}\ve{v}_j$
+    \FOR{$i = 1, 2, \cdots, j$}
+    \STATE $h_{ij} \leftarrow \langle \ve{w}_j,\ve{v}_j \rangle$
+    \STATE $\ve{w}_j \leftarrow \ve{w}_j - h_{ij}\ve{v}_i$
     \ENDFOR
-    \STATE $h_{j+1,j} \leftarrow ||w_j||_2$
-    \STATE $v_{j+1} \leftarrow w_j / h_{j+1,j}$
+    \STATE $h_{j+1,j} \leftarrow ||\ve{w}_j||_2$
+    \STATE $\ve{v}_{j+1} \leftarrow \ve{w}_j / h_{j+1,j}$
+    \ENDFOR
     \COMMENT{Apply the orthogonality constraints}
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Monte Carlo Solution Methods for Discrete Linear Systems}
+\begin{frame}{Monte Carlo Methods for Discrete Linear Systems}
 
@@
     \item Modern work by Evans and others has yielded new applications
-    \item No indication of parallel methods in the literature
   \end{itemize}
@@
 
+  \begin{figure}[h!]
+    \begin{center}
+      \includegraphics[width=4in]{direct_1.png}
+    \end{center}
+    \caption{\textbf{Direct solution to Poisson Equation.} \textit{1
+        history per state, 2500 total histories. 0.785 seconds CPU
+        time} }
+  \end{figure}
+
 \end{frame}
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Parallelization of Monte Methods}
+\begin{frame}{Direct vs. Adjoint Analysis}
+
+  \begin{itemize}
+  \item Analysis needed to select Monte Carlo method
+  \item Time-dependent 2-dimensional Poisson equation
+  \item Spectral radius fixed
+  \item Sparsity varied with 2 Laplacian stencils
+  \end{itemize}
+
+  \[
+  \nabla^2_5 = \frac{1}{\Delta^2}[u_{i-1,j} + u_{i+1,j} + u_{i,j-1} +
+    u_{i,j+1} - 4 u_{i,j}]
+  \]
+  \[
+  \begin{split}
+    \nabla^2_9 = \frac{1}{6\Delta^2}[4 u_{i-1,j} + 4 u_{i+1,j} + 4
+      u_{i,j-1} + 4 u_{i,j+1} + u_{i-1,j-1}\\ + u_{i-1,j+1} +
+      u_{i+1,j-1} + u_{i+1,j+1} - 20 u_{i,j}]
+  \end{split}
+  \]
+
+  \begin{itemize}
+  \item Implicit Euler time differencing
+  \end{itemize}
+
+  \[
+  \ve{A} \ve{u}^{n+1} = \ve{u}^n
+  \]
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Direct vs. Adjoint Analysis}
+
+  \begin{columns}
+
+    \begin{column}{0.5\textwidth}
+      \begin{figure}[h!]
+        \centering
+        \includegraphics[width=2.5in,clip]{Adjoint_Direct_CPU_Time.pdf}
+        \caption{\textbf{CPU Time (s) to converge vs. Problem Size ($N$ for
+            an $N \times N$ square mesh).} }
+      \end{figure}
+    \end{column}
+
+    \begin{column}{0.5\textwidth}
+      \begin{figure}[h!]
+        \centering
+        \includegraphics[width=2.5in,clip]{Adjoint_Direct_Iterations.pdf}
+        \caption{\textbf{Iterations to converge vs. Problem Size ($N$ for an
+            $N \times N$ square mesh).} }
+      \end{figure}
+    \end{column}
+
+  \end{columns}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Direct vs. Adjoint Analysis}
+
+  \begin{figure}[h!]
+    \centering
+    \includegraphics[width=2.5in,clip]{Adjoint_Direct_Convergence.pdf}
+    \caption{\textbf{Infinity norm of the solution residual
+        vs. iteration number for a problem of fixed size.} }
+  \end{figure}
+
+  \begin{itemize}
+  \item CPU time dominating factor in method selection
+  \item Significant speedup with adjoint method
+  \item Does not affect convergence behavior
+  \item Use adjoint with MCSA and Sequential Monte Carlo
+  \end{itemize}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Parallelization of Monte Carlo Methods}
 

### 2012-10-31 doc/prelim/presentation/prelim_presentation.tex
adding solution evolution slides

@@
     \caption{\textbf{Direct solution to Poisson Equation.} \textit{1
-        history per state, 2500 total histories. 0.785 seconds CPU
-        time} }
+        history per state, \sn{2.5}{3} total histories. 0.785 seconds
+        CPU time.} }
+  \end{figure}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Direct Method: Evolution of a Solution}
+
+  \begin{figure}[h!]
+    \begin{center}
+      \includegraphics[width=4in]{direct_10.png}
+    \end{center}
+    \caption{\textbf{Direct solution to Poisson Equation.} \textit{10
+        histories per state, \sn{2.5}{4} total histories. 5.9 seconds
+        CPU time.} }
+  \end{figure}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Direct Method: Evolution of a Solution}
+
+  \begin{figure}[h!]
+    \begin{center}
+      \includegraphics[width=4in]{direct_100.png}
+    \end{center}
+    \caption{\textbf{Direct solution to Poisson Equation.} \textit{100
+        histories per state, \sn{2.5}{5} total histories. 54.7 seconds
+        CPU time.} }
+  \end{figure}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Direct Method: Evolution of a Solution}
+
+  \begin{figure}[h!]
+    \begin{center}
+      \includegraphics[width=4in]{direct_1000.png}
+    \end{center}
+    \caption{\textbf{Direct solution to Poisson Equation.} \textit{100
+        histories per state, \sn{2.5}{6} total histories. 644 seconds
+        CPU time.} }
   \end{figure}
@@
 
+  \begin{figure}[h!]
+    \begin{center}
+      \includegraphics[width=4in]{adjoint_1.png}
+    \end{center}
+    \caption{\textbf{Adjoint solution to Poisson Equation.}
+      \textit{\sn{1}{0} total histories, 0.286 seconds CPU time.} }
+  \end{figure}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Adjoint Method: Evolution of a Solution}
+
+  \begin{figure}[h!]
+    \begin{center}
+      \includegraphics[width=4in]{adjoint_10.png}
+    \end{center}
+    \caption{\textbf{Adjoint solution to Poisson Equation.}
+      \textit{\sn{1}{1} total histories, 0.278 seconds CPU time.} }
+  \end{figure}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Adjoint Method: Evolution of a Solution}
+
+  \begin{figure}[h!]
+    \begin{center}
+      \includegraphics[width=4in]{adjoint_100.png}
+    \end{center}
+    \caption{\textbf{Adjoint solution to Poisson Equation.}
+      \textit{\sn{1}{2} total histories, 0.275 seconds CPU time.} }
+  \end{figure}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Adjoint Method: Evolution of a Solution}
+
+  \begin{figure}[h!]
+    \begin{center}
+      \includegraphics[width=4in]{adjoint_1000.png}
+    \end{center}
+    \caption{\textbf{Adjoint solution to Poisson Equation.}
+      \textit{\sn{1}{3} total histories, 0.291 seconds CPU time.} }
+  \end{figure}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Adjoint Method: Evolution of a Solution}
+
+  \begin{figure}[h!]
+    \begin{center}
+      \includegraphics[width=4in]{adjoint_10000.png}
+    \end{center}
+    \caption{\textbf{Adjoint solution to Poisson Equation.}
+      \textit{\sn{1}{4} total histories, 0.428 seconds CPU time.} }
+  \end{figure}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Adjoint Method: Evolution of a Solution}
+
+  \begin{figure}[h!]
+    \begin{center}
+      \includegraphics[width=4in]{adjoint_100000.png}
+    \end{center}
+    \caption{\textbf{Adjoint solution to Poisson Equation.}
+      \textit{\sn{1}{5} total histories, 1.76 seconds CPU time.} }
+  \end{figure}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Adjoint Method: Evolution of a Solution}
+
+  \begin{figure}[h!]
+    \begin{center}
+      \includegraphics[width=4in]{adjoint_1000000.png}
+    \end{center}
+    \caption{\textbf{Adjoint solution to Poisson Equation.}
+      \textit{\sn{1}{6} total histories, 15.1 seconds CPU time.} }
+  \end{figure}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Adjoint Method: Evolution of a Solution}
+
+  \begin{figure}[h!]
+    \begin{center}
+      \includegraphics[width=4in]{adjoint_10000000.png}
+    \end{center}
+    \caption{\textbf{Adjoint solution to Poisson Equation.}
+      \textit{\sn{1}{7} total histories, 149 seconds CPU time.} }
+  \end{figure}
+
 \end{frame}

### 2012-10-31 doc/prelim/presentation/prelim_presentation.tex
added nonlinear preliminaries

@@
 
+  \begin{itemize}
+  \item We seek solutions of the general nonlinear problem
+  \end{itemize}
+
+  \[
+  \ve{F}(\ve{u}) = \ve{0}
+  \]
+  \[
+  \ve{u} \in \mathbb{R}^n,\ \ve{F}:\mathbb{R}^N \rightarrow
+  \mathbb{R}^N
+  \]
+
+  \begin{itemize}
+  \item We interpret the exact solution $\ve{u}$ to be the roots of
+    $\ve{F}(\ve{u})$
+  \item Taylor expand the residuals at the $k+1$ iterate about the $k$
+    iterate
+  \end{itemize}
+
+  \[
+  \ve{F}(\ve{u}^{k+1}) = \ve{F}(\ve{u}^{k}) +
+  \ve{F}'(\ve{u}^{k})(\ve{u}^{k+1}-\ve{u}^{k}) +
+  \frac{\ve{F}''(\ve{u}^{k})}{2}(\ve{u}^{k+1}-\ve{u}^{k})^2 + \cdots
+  \]
+
+  \begin{itemize}
+  \item Assert $\ve{u}^{k+1}$ is the exact solution
+  \end{itemize}
+
+  \[
+  -\ve{F}(\ve{u}^{k}) = \ve{F}'(\ve{u}^{k})(\ve{u}^{k+1}-\ve{u}^{k})
+  \]
+
 \end{frame}
@@
 
+  \begin{itemize}
+  \item $\ve{F}'(\ve{u}^{k})$ is the \textit{Jacobian}
+    $\ve{J}(\ve{u})$
+  \end{itemize}
+
+  \[
+  J_{ij} = \frac{\partial F_i(\ve{u})}{\partial u_j}
+  \]
+
+  \begin{itemize}
+  \item $(\ve{u}^{k+1}-\ve{u}^{k})$ is the solution update from the
+    $k$ iterate to the $k+1$ iterate
+  \end{itemize}
+
+  \[
+  \delta \ve{u}^k = \ve{u}^{k+1}-\ve{u}^{k}
+  \]
+
+  \begin{itemize}
+  \item Form Newton's method
+  \end{itemize}
+  \[
+  \ve{J}(\ve{u}) \delta \ve{u}^k = -\ve{F}(\ve{u}^{k})
+  \]
+  \[
+  \ve{u}^{k+1} = \ve{u}^k + \delta \ve{u}^k
+  \]
+
 \end{frame}

### 2012-10-31 doc/prelim/presentation/prelim_presentation.tex
working through fanm slides

@@
         \includegraphics[width=2.5in,clip]{Adjoint_Direct_CPU_Time.pdf}
-        \caption{\textbf{CPU Time (s) to converge vs. Problem Size ($N$ for
-            an $N \times N$ square mesh).} }
+        \caption{\textbf{CPU Time (s) to converge vs. Problem Size
+            ($N$ for an $N \times N$ square mesh).} }
       \end{figure}
@@
         \includegraphics[width=2.5in,clip]{Adjoint_Direct_Iterations.pdf}
-        \caption{\textbf{Iterations to converge vs. Problem Size ($N$ for an
-            $N \times N$ square mesh).} }
+        \caption{\textbf{Iterations to converge vs. Problem Size ($N$
+            for an $N \times N$ square mesh).} }
       \end{figure}
@@
 
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Newton-Krylov Methods}
+  \begin{itemize}
+  \item A form of inexact Newton methods
+  \end{itemize}
+
+  \[
+  ||\ve{J}(\ve{u}^k)\delta \ve{u}^k + \ve{F}(\ve{u}^k)|| \leq \eta^k
+  ||\ve{F}(\ve{u}^k)||,\ \eta^k \in [0,1)
+  \]
+
+  \begin{itemize}
+  \item Choose a Krylov method to solve for the Newton correction
+  \item GMRES with a long recurrence relation obeserved as more robust
+  \item Generates a monotonically decreasing residual from maintaining
+    the optimization and orthogonality conditions
+  \end{itemize}
+
+  \pause
+  Where does the Jacobian come from?
+  \pause
+  \begin{itemize}
+  \item The Jacobian can come from hand-coded derivatives
+    \begin{itemize}
+    \item Tedious and error prone
+    \item Repeated for each equation set and hard to do for multiphysics
+    \end{itemize}
+  \end{itemize}
 
@@
 
+  \begin{itemize}
+  \item Krylov methods only need the action of the linear operator
+  \item The action of the Jacobian can be approximated using a forward
+    difference
+  \end{itemize}
+
+  \[
+  \ve{J}(\ve{u})\ve{v} = \frac{\ve{F}(\ve{u} + \epsilon \ve{v}) -
+    \ve{F}(\ve{u})}{\epsilon}
+  \]
+
+  \begin{itemize}
+  \item Forms the basis of Jacobian-Free Newton-Krylov (JFNK) methods
+    \begin{itemize}
+    \item Sensitive to scaling and discretization error
+    \item Still have to form part of Jacobian periodically for
+      preconditioning
+    \item Eventually break even with generating and storing the full
+      Jacobian
+    \end{itemize}
+  \end{itemize}
+
 \end{frame}
@@
 
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{The FANM Method}
+  \begin{itemize}
+  \item Automatically generate Jacobians from nonlinear function
+    evaluations
+    \begin{itemize}
+    \item Overload math operators and apply the chain rule (FAD)
+    \item Yields evaluations as accurate as function discretization
+    \end{itemize}
+  \item Modern packages take an element-level assembly approach
+    \begin{itemize}
+    \item Function evaluations are local, communication builds global
+      data
+    \end{itemize}
+  \end{itemize}
+
+  \[
+  \ve{J}(\ve{u}) = \sum_{i=1}^N \ve{Q}^T_i \ve{J}_k \ve{P}_i
+  \]
+  \[
+  e_k : \mathbb{R}^{n_k} \rightarrow \mathbb{R}^{m_k},\ \ve{J}_{k_i} =
+  \partial e_{k_i} / \partial P_i u,\ \ve{P} \in \mathbb{R}^{m_{k_i}
+    \times N}
+  \]
+
+  \begin{itemize}
+  \item Performance studies give acceptable results for use in
+    large-scale, production physics codes
+  \end{itemize}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}[fragile]{The FANM Method}
+
+  Forward-Automated Newton-MCSA
+
+  \pause
+  \begin{algorithm}[H]
+    \begin{algorithmic}[1]
+      \STATE $k := 0$
+      \WHILE{$||\ve{F}(\ve{u}^{k})|| > \epsilon ||\ve{F}(\ve{u}^{0})||$}
+      \STATE $\ve{J}(\ve{u}^{k}) \leftarrow AD(\ve{F}(\ve{u}^k))$
+      \COMMENT{Automatic differentation}
+      \STATE $\ve{J}(\ve{u}^k) \delta \ve{u}^k = -\ve{F}(\ve{u}^{k})$
+      \COMMENT{Solve for the Newton correction with MCSA}
+      \STATE $\ve{u}^{k+1} \leftarrow \ve{u}^k + \delta \ve{u}^k$
+      \STATE $k \leftarrow k+1$
+      \ENDWHILE
+    \end{algorithmic}
+    \caption{FANM}
+  \end{algorithm}
+
+  \pause
+  \begin{itemize}
+  \item Robustness of Newton's method (inexact)
+  \item Accuracy and convenience of FAD
+  \item Parallelism, memory, and resiliency benefits of MCSA
+  \item Requires only nonlinear function evaluations
+  \end{itemize}
 
@@
 
+  \begin{columns}
+
+    \begin{column}{0.5\textwidth}
+      \begin{itemize}
+      \item Jacobian will be in a compressed row storage format
+      \end{itemize}
+    \end{column}
+
+    \begin{column}{0.5\textwidth}
+      \small{\[ \ve{A} =
+        \begin{bmatrix}
+          2 & 0 & 8 & 0 & 0 & 0 \\ 4 & 5 & 0 & 1 & 0 & 0 \\ 0 & 2 & 1
+          & 0 & 1 & 0 \\ 0 & 0 & 3 & 7 & 0 & 2 \\ 0 & 0 & 0 & 4 & 9 &
+          0 \\ 0 & 0 & 0 & 0 & 9 & 1 \\
+        \end{bmatrix}      
+        \]}
+    \end{column}
+
+  \end{columns}
+
+  \pause \small{ \[
+    \begin{array}{lccccccccccccccc}
+      \ve{A}\ values: & 2 & 8 & 4 & 5 & 1 & 2 & 1 & 1 & 3 & 7 & 2 & 4
+      & 9 & 9 & 1 \\ column: & 1 & 3 & 1 & 2 & 4 & 2 & 3 & 5 & 3 & 4 &
+      6 & 4 & 5 & 5 & 6 \\ row\ start: & 1 & 3 & 6 & 9 & 12 & 14 & 16
+      \\
+    \end{array}
+    \]}
+
+  \pause
+  \begin{itemize}
+  \item $m$ Krylov iterations require $(m+1)$ subspace vectors
+  \item Storage requirement is $\big[(m+1) N\big]$ for $\ve{x} \in
+    \mathbb{R}^{N \times N}$
+  \end{itemize}
+
+  \pause
+  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
+    {If we need 10 Krylov iterations to converge...}
+    \begin{itemize}
+    \item 66 elements required for subspace vectors
+    \item 72 total elements required for Jacobian and probability matrix
+    \end{itemize}
+  \end{beamerboxesrounded}
+
 \end{frame}

### 2012-10-31 doc/prelim/presentation/prelim_presentation.tex
finished nonlinear section

@@
 %%---------------------------------------------------------------------------%%
+\begin{frame}{Generalization of MCSA for Linear Problems}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
 \begin{frame}{Parallelization of Monte Carlo Methods}
@@
 \begin{frame}{Parallel MCSA}
+
+  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
+    {MCSA Iteration}
+
+    \[
+    \ve{x}^{k+1/2} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}
+    \]
+    \[
+    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}
+    \]
+    \[
+    \hat{\ve{A}}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}
+    \]
+    \[
+    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}
+    \]
+
+  \end{beamerboxesrounded}
+
+  \begin{itemize}
+  \item Richardson iteration and residual computation require parallel
+    matrix-vector multiply and parallel vector update
+  \item This work will generate a parallel adjoint Neumann-Ulam solver 
+  \item Application of correction requires parallel vector update
+  \item Convergence checks through parallel norm computation
+  \end{itemize}
 
@@
 
+  \begin{algorithm}[H]
+    \begin{algorithmic}[1]
+      \STATE $k := 0$
+      \WHILE{$||\ve{F}(\ve{u}^{k})|| > \epsilon ||\ve{F}(\ve{u}^{0})||$}
+      \STATE $\ve{J}(\ve{u}^{k}) \leftarrow AD(\ve{F}(\ve{u}^k))$
+      \COMMENT{Automatic differentation}
+      \STATE $\ve{J}(\ve{u}^k) \delta \ve{u}^k = -\ve{F}(\ve{u}^{k})$
+      \COMMENT{Solve for the Newton correction with MCSA}
+      \STATE $\ve{u}^{k+1} \leftarrow \ve{u}^k + \delta \ve{u}^k$
+      \STATE $k \leftarrow k+1$
+      \ENDWHILE
+    \end{algorithmic}
+    \caption{FANM}
+  \end{algorithm}
+
+  \begin{itemize}
+  \item Modern FAD packages are parallelized using element-level assembly
+  \item This work will generate a parallel MCSA solver
+  \item Application of the Newton correction requires a parallel vector
+    update
+  \item Convergence checks through parallel norm comptutation
+  \end{itemize}
+
 \end{frame}
@@
 
+  \begin{itemize}
+  \item Experimental framework
+  \item Methods verification
+  \item Numerical experiments
+  \item Challenge problem
+  \end{itemize}
+
 \end{frame}
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Progress to Date}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Progress to Date}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Progress to Date}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
 \begin{frame}{Monte Carlo Methods Verification}

### 2012-10-31 doc/prelim/presentation/prelim_presentation.tex
working on research plan

@@
     future hardware
-    \item A set of massively parallel Monte Carlo methods is proposed
-      to advance multiple physics simulation on contemporary and
-      future leadership class machines
+  \item A set of massively parallel Monte Carlo methods is proposed to
+    advance multiple physics simulation on contemporary and future
+    leadership class machines
   \end{itemize}
@@
       methods
-      \item Research is required to explore varying parallel
-        strategies
+    \item Research is required to explore varying parallel strategies
     \item Scalability is of concern
@@
   \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
-    {A Requirement}
-    Assert that $\ve{A}$ is \textit{nonsingular}. The solution is then:
+    {A Requirement} Assert that $\ve{A}$ is \textit{nonsingular}. The
+    solution is then:
     \[
@@
   \item Powerful class of iterative methods
-    \item Provides theory that encapsulates most other iterative
-      methods 
+  \item Provides theory that encapsulates most other iterative methods
   \item Leveraged in many modern physics codes at the petascale
@@
   \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
-    {Search Subspace $\mathcal{K}$} 
-    Extract the solution from the search subspace:
+    {Search Subspace $\mathcal{K}$} Extract the solution from the
+    search subspace:
     \[
@@
   \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
-    {Constraint Subspace $\mathcal{L}$} 
-    Constrain the extraction with the constraint subspace by asserting
-    orthogonality with the residual:
+    {Constraint Subspace $\mathcal{L}$} Constrain the extraction with
+    the constraint subspace by asserting orthogonality with the
+    residual:
     \[
@@
     \begin{center}
-      \scalebox{1.25}{
-        \input{orthogonal_residual.pdftex_t} }
+      \scalebox{1.25}{ \input{orthogonal_residual.pdftex_t} }
     \end{center}
@@
   \[
-  \boldsymbol{\delta} = \ve{V}\ve{y},\ \forall \ve{y} \in
-  \mathbb{R}^N
+  \boldsymbol{\delta} = \ve{V}\ve{y},\ \forall \ve{y} \in \mathbb{R}^N
   \]
@@
   \item $\ve{W}_m = \ve{A}\ve{V}_m$
-  \item Typically choose a Gram-Schmidt-like procedure such as
-    Arnoldi or Lanzcos
+  \item Typically choose a Gram-Schmidt-like procedure such as Arnoldi
+    or Lanzcos
   \end{itemize}
@@
     \begin{algorithmic}[1]
-    \STATE $\ve{r}_0 := \ve{b}-\ve{A}\ve{x}_0$
-    \STATE $\beta := ||\ve{r}_0||_2$
-    \STATE $\ve{v}_1 := \ve{r}_0 / \beta$
+      \STATE $\ve{r}_0 := \ve{b}-\ve{A}\ve{x}_0$ \STATE $\beta :=
+      ||\ve{r}_0||_2$ \STATE $\ve{v}_1 := \ve{r}_0 / \beta$
       \COMMENT{Create the orthonormal basis for the Krylov subspace}
-    \FOR{$j = 1, 2, \cdots, m$}
-    \STATE $\ve{w}_j := \ve{A}\ve{v}_j$
-    \FOR{$i = 1, 2, \cdots, j$}
-    \STATE $h_{ij} \leftarrow \langle \ve{w}_j,\ve{v}_j \rangle$
-    \STATE $\ve{w}_j \leftarrow \ve{w}_j - h_{ij}\ve{v}_i$
-    \ENDFOR
-    \STATE $h_{j+1,j} \leftarrow ||\ve{w}_j||_2$
-    \STATE $\ve{v}_{j+1} \leftarrow \ve{w}_j / h_{j+1,j}$
-    \ENDFOR
-    \COMMENT{Apply the orthogonality constraints}
-    \STATE $\ve{y}_m \leftarrow argmin_y ||\beta \ve{e}_1 - \ve{H}_m\
-    \ve{y}||_2 $
-    \STATE $\ve{x}_m \leftarrow \ve{x}_0 + \ve{V_m} \ve{y}_m$
+      \FOR{$j = 1, 2, \cdots, m$} \STATE $\ve{w}_j := \ve{A}\ve{v}_j$
+      \FOR{$i = 1, 2, \cdots, j$} \STATE $h_{ij} \leftarrow \langle
+      \ve{w}_j,\ve{v}_j \rangle$ \STATE $\ve{w}_j \leftarrow \ve{w}_j
+      - h_{ij}\ve{v}_i$ \ENDFOR \STATE $h_{j+1,j} \leftarrow
+      ||\ve{w}_j||_2$ \STATE $\ve{v}_{j+1} \leftarrow \ve{w}_j /
+      h_{j+1,j}$ \ENDFOR \COMMENT{Apply the orthogonality constraints}
+      \STATE $\ve{y}_m \leftarrow argmin_y ||\beta \ve{e}_1 -
+      \ve{H}_m\ \ve{y}||_2 $ \STATE $\ve{x}_m \leftarrow \ve{x}_0 +
+      \ve{V_m} \ve{y}_m$
     \end{algorithmic}
@@
   \[
-  \ve{y}[n] \leftarrow \ve{y}[n] + a * \ve{x}[n],\ \forall n \in [1,N_g]
-  \]
-  \[
-  \ve{y}[n] \leftarrow \ve{y}[n] + a * \ve{x}[n],\ \forall n \in [1,N_l]
+  \ve{y}[n] \leftarrow \ve{y}[n] + a * \ve{x}[n],\ \forall n \in
+     [1,N_g]
+  \]
+  \[
+  \ve{y}[n] \leftarrow \ve{y}[n] + a * \ve{x}[n],\ \forall n \in
+     [1,N_l]
   \]
@@
     \begin{center}
-    \scalebox{0.75}{
-      \input{partitioned_matrix.pdftex_t} }
+      \scalebox{0.75}{ \input{partitioned_matrix.pdftex_t} }
     \end{center}
@@
     \begin{center}
-    \scalebox{0.75}{
-      \input{matvec_proc_1.pdftex_t} }
+      \scalebox{0.75}{ \input{matvec_proc_1.pdftex_t} }
     \end{center}
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Projection Method Notes}
-
-\begin{itemize}
+\begin{frame}{Projection Methods Summary}
+
+  \begin{itemize}
+  \item Used in practice everywhere
   \item Global reduction operations observed not to impede scalability
@@
     \end{itemize}
-
   \end{itemize}
@@
   \begin{itemize}
-    \item Compute the probability of a particular random walk permutation
+  \item Compute the probability of a particular random walk
+    permutation
   \end{itemize}
@@
   \item No literature observed for parallel Neumann-Ulam solvers
-  \item Numerous references for modern parallel Monte Carlo methods
-    in reactor physics
+  \item Numerous references for modern parallel Monte Carlo methods in
+    reactor physics
   \item MCSA parallelism comes from parallel matrix/vector operations
@@
     \begin{center}
-      \scalebox{1.0}{
-        \input{msod_example.pdftex_t} }
-    \end{center}
-    \caption{\textbf{Overlapping domain example illustrating how domain
-        overlap can reduce communication costs.}}
+      \scalebox{1.0}{ \input{msod_example.pdftex_t} }
+    \end{center}
+    \caption{\textbf{Overlapping domain example illustrating how
+        domain overlap can reduce communication costs.}}
   \end{figure}
@@
     \begin{center}
-      \scalebox{1.0}{
-        \input{procassini_example.pdftex_t} }
+      \scalebox{1.0}{ \input{procassini_example.pdftex_t} }
     \end{center}
@@
     \begin{center}
-      \scalebox{1.0}{
-        \input{gentile_example.pdftex_t} }
+      \scalebox{1.0}{ \input{gentile_example.pdftex_t} }
     \end{center}
     \caption{\textbf{Gentile's example illustrating how domain
-        decomposition can create reproducibility issues in Monte Carlo.}}
+        decomposition can create reproducibility issues in Monte
+        Carlo.}}
     \label{fig:gentile_example}
@@
     \item Choose a Krylov method to solve for the Newton correction
-  \item GMRES with a long recurrence relation obeserved as more robust
-  \item Generates a monotonically decreasing residual from maintaining
-    the optimization and orthogonality conditions
-  \end{itemize}
-
-  \pause
-  Where does the Jacobian come from?
-  \pause
+    \item GMRES with a long recurrence relation obeserved as more
+      robust
+    \item Generates a monotonically decreasing residual from
+      maintaining the optimization and orthogonality conditions
+    \end{itemize}
+
+    \pause Where does the Jacobian come from?  \pause
     \begin{itemize}
@@
       \item Tedious and error prone
-    \item Repeated for each equation set and hard to do for multiphysics
+      \item Repeated for each equation set and hard to do for
+        multiphysics
       \end{itemize}
@@
     \begin{algorithmic}[1]
-      \STATE $k := 0$
-      \WHILE{$||\ve{F}(\ve{u}^{k})|| > \epsilon ||\ve{F}(\ve{u}^{0})||$}
-      \STATE $\ve{J}(\ve{u}^{k}) \leftarrow AD(\ve{F}(\ve{u}^k))$
-      \COMMENT{Automatic differentation}
-      \STATE $\ve{J}(\ve{u}^k) \delta \ve{u}^k = -\ve{F}(\ve{u}^{k})$
-      \COMMENT{Solve for the Newton correction with MCSA}
-      \STATE $\ve{u}^{k+1} \leftarrow \ve{u}^k + \delta \ve{u}^k$
-      \STATE $k \leftarrow k+1$
-      \ENDWHILE
+      \STATE $k := 0$ \WHILE{$||\ve{F}(\ve{u}^{k})|| > \epsilon
+        ||\ve{F}(\ve{u}^{0})||$} \STATE $\ve{J}(\ve{u}^{k}) \leftarrow
+      AD(\ve{F}(\ve{u}^k))$ \COMMENT{Automatic differentation} \STATE
+      $\ve{J}(\ve{u}^k) \delta \ve{u}^k = -\ve{F}(\ve{u}^{k})$
+      \COMMENT{Solve for the Newton correction with MCSA} \STATE
+      $\ve{u}^{k+1} \leftarrow \ve{u}^k + \delta \ve{u}^k$ \STATE $k
+      \leftarrow k+1$ \ENDWHILE
     \end{algorithmic}
@@
     \item 66 elements required for subspace vectors
-    \item 72 total elements required for Jacobian and probability matrix
+    \item 72 total elements required for Jacobian and probability
+      matrix
     \end{itemize}
@@
   \begin{itemize}
-  \item Modern FAD packages are parallelized using element-level assembly
+  \item Modern FAD packages are parallelized using element-level
+    assembly
   \item This work will generate a parallel MCSA solver
-  \item Application of the Newton correction requires a parallel vector
-    update
+  \item Application of the Newton correction requires a parallel
+    vector update
   \item Convergence checks through parallel norm comptutation
@@
 
+  \begin{itemize}
+  \item Need a framework to generate linear and nonlinear systems in
+    parallel
+    \begin{itemize}
+    \item Linear algebra data structures
+    \item Parallel communication framework
+    \item Mesh generation and partitioning
+    \item Automatic differentiation and more...
+    \end{itemize}
+  \item Choose the finite element method to leverage a finite element
+    assembly engine
+  \end{itemize}
+
+  \[
+  \ve{A}(\ve{u}) = \ve{0},\ \forall \ve{u} \in \Omega
+  \]
+  \[
+  \ve{B}(\ve{u}) = \ve{0},\ \forall \ve{u} \in \Gamma
+  \]
+  \[
+  \int_{\Omega} \ve{v}^T \ve{A}(\ve{u}) d\Omega + \int_{\Gamma}
+  \bar{\ve{v}}^T \ve{B}(\ve{u}) d\Gamma = \ve{0}
+  \]
+  \[
+  \int_{\Omega} \ve{C}(\ve{v}^T) \ve{D}(\ve{u}) d\Omega +
+  \int_{\Gamma} \ve{E}(\bar{\ve{v}}^T) \ve{F}(\ve{u}) d\Gamma = \ve{0}
+  \]
+
 \end{frame}
@@
 
+  \begin{itemize}
+  \item Analytic solution to the heat equation for the linear methods
+  \item Sequence of Navier-Stokes benchmarks for the nonlinear methods
+    \begin{itemize}
+    \item Thermal convection cavity problem (De Vahl Davis, 1983)
+    \item Lid driven cavity problem (Ghia et al., 1982)
+    \item Backward-Facing step problem (Gartling, 1990)
+    \end{itemize}
+  \item Tuning benchmark parameters varies the strength of
+    nonlinearities
+  \end{itemize}
+
+  \begin{columns}
+    \begin{column}{0.5\textwidth}
+      \[
+      \rho \ve{u} \cdot \nabla \ve{u} - \nabla \cdot \ve{T} - \rho
+      \ve{g} = \ve{0}
+      \]
+      \[
+      \nabla \cdot \ve{u} = 0
+      \]
+      \[
+      \rho C_p \ve{u} \cdot \nabla T + \nabla \cdot \ve{q} = 0
+      \]
+    \end{column}
+
+    \begin{column}{0.5\textwidth}
+      \[
+      \ve{T} = -P \ve{I} + \mu[\nabla \ve{u} + \nabla \ve{u}^T]
+      \]
+      \[
+      \ve{q} = - k \nabla T
+      \]
+    \end{column}
+  \end{columns}
+
 \end{frame}
@@
 
+  Domain Overlap Studies for the Parallel Neumann-Ulam Method
+
+  Domain Replication Studies for the Parallel Neumann-Ulam Method
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Proposed Numerical Experiments}
+
+  Parallel Performance and Numerical Accuracy Studies for the MCSA
+  Method
+
+  Parallel Performance and Numerical Accuracy Studies for the FANM
+  Method
+
 \end{frame}

### 2012-10-31 doc/prelim/document/chapters/research_proposal/research_proposal.tex
spell check for prelim
adding drekar images

@@
 operator. With this type of approach, transition probabilities and
-weight formulations must be rederived for each type of physics
+weight formulations must be re-derived for each type of physics
 implemented. To reach the broader physics community and facilitate
@@
 equations \citep{shadid_inexact_1997,pawlowski_globalization_2006}. We
-define these equations as follows :
+define these equations as follows:
 \begin{subequations}
@@
 experiment with the way we combine it with the Neumann-Ulam solver. As
-discussed in \S~\ref{subsec:parallel_adjoint}, previous work with
-MCSA used a correction generated by the Neumann-Ulam method with a
-fairly large uncertainty near 14\%. We hypothesized that if we can
-identify the amount of overlap in or Neumann-Ulam experiments required
-to eliminate most transitions between adjacent domains owned by
-different processors, then perhaps transport to adjacent domains can
-be ignored. This will result in a biased estimator for the MCSA
+discussed in \S~\ref{subsec:parallel_adjoint}, previous work with MCSA
+used a correction generated by the Neumann-Ulam method with a fairly
+large uncertainty near 14\%. We hypothesized that if we can identify
+the amount of overlap in our Neumann-Ulam experiments required to
+eliminate most transitions between adjacent domains owned by different
+processors, then perhaps transport to adjacent domains can be
+ignored. This will result in a biased estimator for the MCSA
 correction that cannot be reproduced, however, these studies will aim

### 2012-10-31 doc/prelim/presentation/prelim_presentation.tex
adding drekar images

@@
     \begin{algorithmic}[1]
-      \STATE $\ve{r}_0 := \ve{b}-\ve{A}\ve{x}_0$ \STATE $\beta :=
-      ||\ve{r}_0||_2$ \STATE $\ve{v}_1 := \ve{r}_0 / \beta$
+      \STATE $\ve{r}_0 := \ve{b}-\ve{A}\ve{x}_0$ 
+      \STATE $\beta := ||\ve{r}_0||_2$ 
+      \STATE $\ve{v}_1 := \ve{r}_0 / \beta$
       \COMMENT{Create the orthonormal basis for the Krylov subspace}
-      \FOR{$j = 1, 2, \cdots, m$} \STATE $\ve{w}_j := \ve{A}\ve{v}_j$
-      \FOR{$i = 1, 2, \cdots, j$} \STATE $h_{ij} \leftarrow \langle
-      \ve{w}_j,\ve{v}_j \rangle$ \STATE $\ve{w}_j \leftarrow \ve{w}_j
-      - h_{ij}\ve{v}_i$ \ENDFOR \STATE $h_{j+1,j} \leftarrow
-      ||\ve{w}_j||_2$ \STATE $\ve{v}_{j+1} \leftarrow \ve{w}_j /
-      h_{j+1,j}$ \ENDFOR \COMMENT{Apply the orthogonality constraints}
+      \FOR{$j = 1, 2, \cdots, m$} 
+      \STATE $\ve{w}_j := \ve{A}\ve{v}_j$
+      \FOR{$i = 1, 2, \cdots, j$} 
+      \STATE $h_{ij} \leftarrow \langle \ve{w}_j,\ve{v}_j \rangle$ 
+      \STATE $\ve{w}_j \leftarrow \ve{w}_j - h_{ij}\ve{v}_i$ 
+      \ENDFOR 
+      \STATE $h_{j+1,j} \leftarrow ||\ve{w}_j||_2$ 
+      \STATE $\ve{v}_{j+1} \leftarrow \ve{w}_j / h_{j+1,j}$ 
+      \ENDFOR 
+      \COMMENT{Apply the orthogonality constraints}
       \STATE $\ve{y}_m \leftarrow argmin_y ||\beta \ve{e}_1 -
-      \ve{H}_m\ \ve{y}||_2 $ \STATE $\ve{x}_m \leftarrow \ve{x}_0 +
-      \ve{V_m} \ve{y}_m$
+      \ve{H}_m\ \ve{y}||_2 $ 
+      \STATE $\ve{x}_m \leftarrow \ve{x}_0 + \ve{V_m} \ve{y}_m$
     \end{algorithmic}
@@
 
+  \pause
   \begin{figure}[h!]
@@
 
+  \begin{itemize}
+  \item Work to date has used a physics-based formulation
+    \begin{itemize}
+    \item Radiation transport equations used as model system
+    \item Transition probabilities built from problem-specific
+      parameters
+    \item Probabilities and weights must be rederived for each new
+      equation set
+    \end{itemize}
+  \item Desire a generalization for all linear operator equations
+    \begin{itemize}
+    \item Requires a general parallel framework
+    \item Requires implementation with a general linear algebra
+      framework
+    \item Operator, vector, and graph abstractions
+    \end{itemize}
+  \item Neumann-Ulam solvers and MCSA implemented using the Trilinos
+    Petra frameworks
+  \item Can be leveraged in modern physics implementations
+  \end{itemize}
+
 \end{frame}
@@
     reactor physics
-  \item MCSA parallelism comes from parallel matrix/vector operations
+  \item Build a strategy for applying modern methods to the
+    Neumann-Ulam method
+  \item MCSA iteration-level parallelism comes from parallel
+    matrix/vector operations
   \end{itemize}
@@
 
+  \pause
+  \begin{itemize}
+  \item Many multiphysics problems of interest are nonlinear
+  \item Segregated methods lack the consitency of fully implicit
+    methods
+  \item Newton methods often leverage Krylov solvers
+    \begin{itemize}
+    \item Robust implementations
+    \item No operator required
+    \item Memory intensive
+    \end{itemize}
+  \item Monte Carlo methods need the full operator
+  \item Automatic construction of the linear operator available
+    \begin{itemize}
+    \item Ideal for Monte Carlo
+    \item Relaxes memory requirements
+    \item Potential scaling improvements
+    \item Resiliency benefits
+    \end{itemize}
+  \end{itemize}
+
 \end{frame}
@@
     \item Choose a Krylov method to solve for the Newton correction
-    \item GMRES with a long recurrence relation obeserved as more
+    \item GMRES with a long recurrence relation observed as more
       robust
@@
     \begin{algorithmic}[1]
-      \STATE $k := 0$ \WHILE{$||\ve{F}(\ve{u}^{k})|| > \epsilon
-        ||\ve{F}(\ve{u}^{0})||$} \STATE $\ve{J}(\ve{u}^{k}) \leftarrow
-      AD(\ve{F}(\ve{u}^k))$ \COMMENT{Automatic differentation} \STATE
-      $\ve{J}(\ve{u}^k) \delta \ve{u}^k = -\ve{F}(\ve{u}^{k})$
-      \COMMENT{Solve for the Newton correction with MCSA} \STATE
-      $\ve{u}^{k+1} \leftarrow \ve{u}^k + \delta \ve{u}^k$ \STATE $k
-      \leftarrow k+1$ \ENDWHILE
+      \STATE $k := 0$ 
+      \WHILE{$||\ve{F}(\ve{u}^{k})|| > \epsilon
+        ||\ve{F}(\ve{u}^{0})||$} 
+      \STATE $\ve{J}(\ve{u}^{k}) \leftarrow AD(\ve{F}(\ve{u}^k))$ 
+      \COMMENT{Automatic differentation} 
+      \STATE $\ve{J}(\ve{u}^k) \delta \ve{u}^k = -\ve{F}(\ve{u}^{k})$
+      \COMMENT{Solve for the Newton correction with MCSA} 
+      \STATE $\ve{u}^{k+1} \leftarrow \ve{u}^k + \delta \ve{u}^k$ 
+      \STATE $k \leftarrow k+1$ 
+      \ENDWHILE
     \end{algorithmic}
@@
 
+  \pause
   Domain Overlap Studies for the Parallel Neumann-Ulam Method
-
+  \begin{itemize}
+  \item Correlate domain overlap behavior to linear system properties
+  \item Analyze random walk transport with respect to the operator:
+    \begin{itemize}
+    \item Eigenvalues
+    \item Sparsity
+    \item Asymmetry
+    \end{itemize}
+  \item Evaluate communication cost
+  \item Evaluate memory cost
+  \end{itemize}
+
+  \pause
   Domain Replication Studies for the Parallel Neumann-Ulam Method
+  \begin{itemize}
+  \item Feasibility from a memory perspective
+  \item Impact on scalability
+  \item Impact on load balancing
+  \end{itemize}
 
@@
   Method
-
+  \begin{itemize}
+  \item Characterize Neumann-Ulam parameters required for good
+    correction
+  \item Feasibility of MCSA with full-clip Neumann-Ulam
+  \item Performance for assymmetric systems
+  \item Memory usage compared to Krylov methods
+  \end{itemize}
+
+  \pause
   Parallel Performance and Numerical Accuracy Studies for the FANM
   Method
+  \begin{itemize}
+  \item Feasibility for problems of interest
+  \item Memory usage vs. Newton-Krylov methods
+  \item Scalability vs. Newton-Krylov methods
+  \end{itemize}
 
@@
 
+  \begin{columns}
+    \begin{column}{0.5\textwidth}
+      \begin{figure}[htpb!]
+        \begin{center}
+          \includegraphics[width=2.5in]{drekar1.png}
+        \end{center}
+      \end{figure}
+    \end{column}
+
+    \begin{column}{0.5\textwidth}
+      \begin{figure}[htpb!]
+        \begin{center}
+          \includegraphics[width=2.5in]{drekar2.png}
+        \end{center}
+      \end{figure}
+    \end{column}
+  \end{columns}
+
+  \begin{itemize}
+  \item Problems of interest are large scale problems in reactor
+    physics
+  \item CASL has leveraged the Drekar multiphysics code
+  \item Coupled fluid flow and heat transfer is required to
+    characterize many phenomena
+    \begin{itemize}
+    \item Departure from nucleate boiling
+    \item Grid-to-rod-fretting
+    \end{itemize}
+  \end{itemize}
+
 \end{frame}

### 2012-10-31 doc/prelim/presentation/prelim_presentation.tex
finished research proposal

@@
   \item Many multiphysics problems of interest are nonlinear
-  \item Segregated methods lack the consitency of fully implicit
+  \item Segregated methods lack the consistency of fully implicit
     methods
@@
       \STATE $\ve{J}(\ve{u}^{k}) \leftarrow AD(\ve{F}(\ve{u}^k))$ 
-      \COMMENT{Automatic differentation} 
+      \COMMENT{Automatic differentiation} 
       \STATE $\ve{J}(\ve{u}^k) \delta \ve{u}^k = -\ve{F}(\ve{u}^{k})$
@@
 
+  \begin{itemize}
+  \item Problems of interest are coupled large scale problems in
+    reactor physics
+  \item The Consortium for Advanced Simulation of LWRs (CASL) is a
+    modeling and simulation program aimed at:
+    \begin{itemize}
+      \item Higher power uprates and efficiency
+      \item Higher burn-up
+      \item Predictive accident scenario analysis
+    \end{itemize}
+  \item CASL and industry partners identified challenge problems:
+    \begin{itemize}
+    \item Departure from nucleate boiling
+    \item Grid-to-rod-fretting
+    \end{itemize}
+  \item Our challenge problem should reflect how this work aids the
+    solution of these problems
+  \end{itemize}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Proposed Challenge Problem}
+
   \begin{columns}
@@
   \begin{itemize}
-  \item Problems of interest are large scale problems in reactor
-    physics
-  \item CASL has leveraged the Drekar multiphysics code
-  \item Coupled fluid flow and heat transfer is required to
-    characterize many phenomena
-    \begin{itemize}
-    \item Departure from nucleate boiling
-    \item Grid-to-rod-fretting
-    \end{itemize}
-  \end{itemize}
+  \item CASL has leveraged the Drekar multiphysics code (SNL)
+  \item Coupled fluid flow and heat transfer helps characterize many
+    phenomena
+  \item Drekar is massively parallel and leverages Newton-Krylov
+    methods
+  \item Propose using the largest Drekar problem to date as a
+    challenge problem for the new Monte Carlo methods
+  \end{itemize}
+
+  {\small \sl Images source: www.casl.gov}
 

### 2012-11-01 doc/prelim/presentation/prelim_presentation.tex
finsihed draft except for conclusion

@@
     \end{itemize}
-  \item Development of a nonlinear solver leveraging Monte Carlo
+  \item Development of a nonlinear solver for discrete systems
+    leveraging Monte Carlo
     \begin{itemize}
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Putting it All Together}
+\begin{frame}{Putting It All Together}
 
@@
   \begin{itemize}
-  \item Used in practice everywhere
+  \item Widely used in practice
+  \item Krylov methods require only the action of the operator
+  \item Parallelism achieved through a handful of operations
   \item Global reduction operations observed not to impede scalability
@@
     \end{itemize}
+  \item Short and long reccurrence relations available for
+    orthogonalization
   \end{itemize}
@@
   \[
-  W_{m} = \sum_{m=0}^k w_{i,i_1} w_{i_1,i_2} \cdots w_{i_{m-1},i_m}
+  W_{m} = w_{i,i_1} w_{i_1,i_2} \cdots w_{i_{m-1},i_m}
   \]
@@
 
-  \pause
+  \begin{figure}[h!]
+    \begin{center}
+      \scalebox{1.0}{ \input{heat_eq_setup.pdftex_t} }
+    \end{center}
+  \caption{\textbf{Problem setup for 2D heat equation.}
+    \textit{Dirichlet conditions are set for the temperature on all 4
+      boundaries of the Cartesian grid. No thermal source was
+      present.}}
+  \end{figure}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Direct Method: Evolution of a Solution}
+
   \begin{figure}[h!]
@@
 
-  \pause
   \begin{figure}[h!]
@@
   \begin{itemize}
-  \item Work to date has used a physics-based formulation
+  \item Published work to date has used a physics-based formulation
     \begin{itemize}
@@
 
+  \begin{itemize}
+  \item Each parallel process owns a piece of the domain
+  \item Random walks must be transported across domains through
+    communication
+  \end{itemize}
+
+  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
+    {Brunner's Work (2006 and 2009)}
+
+    \begin{itemize}
+    \item Looked at 4 communication patterns:
+      \begin{itemize}
+      \item Fully-locking synchronous
+      \item asynchronous-send/synchronous-receive
+      \item master/slave
+      \item binary tree master/slave
+      \end{itemize}
+    \item Binary tree master/slave performed best but race conditions observed
+    \item A later improvement to their work showed a fully asynchronous
+      pattern to work best
+    \item Poor scaling for unbalanced problems
+    \end{itemize}
+
+  \end{beamerboxesrounded}
+
 \end{frame}
@@
 
+  \begin{columns}
+
+    \begin{column}{0.5\textwidth}
       \begin{figure}[htpb!]
         \begin{center}
-      \scalebox{1.0}{ \input{msod_example.pdftex_t} }
+          \scalebox{0.75}{ \input{msod_example.pdftex_t} }
         \end{center}
@@
       \end{figure}
+    \end{column}
+
+    \begin{column}{0.5\textwidth}
+      \begin{itemize}
+      \item Developed by Wagner and colleagues in 2010
+      \item Each set contains the full domain
+      \item Multiple sets replicate the domain
+      \item Domains overlap within a set
+      \item Reduces communication by a significant fraction
+      \item Increases scalability through smaller processor sets
+      \item Redundancy for resiliency (and useful work)
+      \end{itemize}
+    \end{column}
+
+  \end{columns}
 
@@
 \begin{frame}{Domain-to-Domain Communication}
+
+  \begin{itemize}
+  \item The amount of domain leakage dictates communication
+  \item Per Siegel's 2012 work, define a leakage fraction
+  \end{itemize}
+
+  \[
+  \lambda =
+  \frac{average\ \#\ of\ particles\ leaving\ local\ domain}{total\ 
+    of\ \#\ of\ particles\ starting\ in\ local\ domain}
+  \]
+
+  \begin{itemize}
+  \item Sigel observed $\lambda$ to be bound empirically to the mean
+    free path of the system
+  \item Optimum ratio of local to global domain size exists
+  \item Experiments show feasibility for large-scale calculations
+    \begin{itemize}
+    \item Not bandwidth or latency limited in load-balanced case
+    \end{itemize}
+    \pause
+  \item How do we define $\lambda$ empirically for linear operator
+    equations? 
+  \item Full-clip roullette for the MCSA correction?
+  \end{itemize}
 
@@
 
+  \begin{itemize}
+  \item Procassini's 2005 work addressed some concerns
+  \item Dynamic balancing replicates domains independently
+  \item Linear operator equations may benefit - fixed source problems
+  \end{itemize}
+
 \end{frame}
@@
     \begin{center}
-      \scalebox{1.0}{ \input{gentile_example.pdftex_t} }
+      \scalebox{0.7}{ \input{gentile_example.pdftex_t} }
     \end{center}
@@
 
+  \begin{itemize}
+  \item 2005 work by Gentile outlines reproducability in parallel
+    Monte carlo
+  \item Maintaining random number sequences
+  \item Floating point addition does not commute
+  \item Useful for benchmarking and unit testing
+  \end{itemize}
+
 \end{frame}
@@
 \begin{frame}{Parallel Adjoint Method}
+
+Strategy
+  \begin{itemize}
+  \item Direct analogs between particle transport and Neumann-Ulam
+    solvers 
+  \item Aim for MSOD implementation and fully asynchronous
+    communication patterns
+  \end{itemize}
+
+Questions
+  \begin{itemize}
+  \item How much domain overlap is suitable for linear operator
+    equations? 
+  \item Is memory a limitation for overlap and replication?
+  \item Do the linear operator properties help select overlap?
+  \item Will full-clip roullette perturb the MCSA solution?
+  \item How much does MSOD facilitate scaling?
+  \end{itemize}
 
@@
   \begin{itemize}
-  \item CASL has leveraged the Drekar multiphysics code (SNL)
+  \item CASL has utilized the Drekar multiphysics code (SNL)
   \item Coupled fluid flow and heat transfer helps characterize many

### 2012-11-01 doc/prelim/presentation/prelim_presentation.tex
added conclusion
spell check on presentation

@@
     \end{itemize}
-  \item Short and long reccurrence relations available for
+  \item Short and long recurrence relations available for
     orthogonalization
@@
   \begin{itemize}
-  \item Expand the Nuemann series
+  \item Expand the Neumann series
   \end{itemize}
@@
       parameters
-    \item Probabilities and weights must be rederived for each new
+    \item Probabilities and weights must be re-derived for each new
       equation set
@@
   \begin{itemize}
-  \item Sigel observed $\lambda$ to be bound empirically to the mean
+  \item Siegel observed $\lambda$ to be bound empirically to the mean
     free path of the system
@@
     equations? 
-  \item Full-clip roullette for the MCSA correction?
+    \pause
+  \item Full-clip roulette for the MCSA correction?
   \end{itemize}
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Reproducability}
+\begin{frame}{Reproducibility}
 
@@
   \begin{itemize}
-  \item 2005 work by Gentile outlines reproducability in parallel
-    Monte carlo
+  \item 2005 work by Gentile outlines reproducibility in parallel
+    Monte Carlo
   \item Maintaining random number sequences
@@
   \item Do the linear operator properties help select overlap?
-  \item Will full-clip roullette perturb the MCSA solution?
+  \item Will full-clip roulette perturb the MCSA solution?
   \item How much does MSOD facilitate scaling?
@@
     vector update
-  \item Convergence checks through parallel norm comptutation
+  \item Convergence checks through parallel norm computation
   \end{itemize}
@@
   \item Feasibility of MCSA with full-clip Neumann-Ulam
-  \item Performance for assymmetric systems
+  \item Performance for asymmetric systems
   \item Memory usage compared to Krylov methods
@@
 
+  \begin{itemize}
+  \item Proposed research and development of new Monte Carlo methods
+    \begin{itemize}
+    \item Parallelization of Monte Carlo methods for linear systems
+    \item FANM for nonlinear systems
+    \end{itemize}
+  \item Verification through benchmarks
+  \item Numerical experiments for understanding
+  \item Challenge problem for application
+  \item Directed towards hard problems in nuclear reactor analysis:
+    \begin{itemize}
+    \item Application to multiphysics
+    \item Potential improvements for scalability
+    \item Potential improvements for memory consumption
+    \item Looking forward to exascale
+    \end{itemize}
+  \end{itemize}
+
 \end{frame}

### 2012-11-01 doc/prelim/document/chapters/parallel_mc/parallel_mc.tex
finsihed draft except for conclusion
updated presentation
fixed adjoint expectation value

@@
 \begin{equation}
-  W_{m} = \sum_{m=0}^k w_{i,i_1} w_{i_1,i_2} \cdots w_{i_{m-1},i_m}\:,
+  W_{m} = w_{i,i_1} w_{i_1,i_2} \cdots w_{i_{m-1},i_m}\:,
   \label{eq:direct_permutation_weight}
@@
 \begin{equation}
-  X_{\nu} = \sum_{m=0}^k W_{m} b_{i_0} \delta_{i,i_m}\:,
+  X_{\nu} = \sum_{m=0}^k W_{m} \delta_{i,i_m}\:,
   \label{eq:adjoint_permutation_contribution}
@@
 where the Kronecker delta indicates that the tally contributes only in
-the current state and $b_{i_0}$ is the sampled source starting
-state. Note here that the estimator in
+the current state and $b_{i_0}$ will be the sampled source starting
+weight. Note here that the estimator in
 Eq~(\ref{eq:adjoint_permutation_contribution}) does not have a
@@
     E\{X_j\} &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-    \sum_{i_k}^{N} b_{i_0} h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}
+    \sum_{i_k}^{N} b_{i_0} h_{i_0,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}
     \delta_{i_k,j} \\ &= x_{j}\:,

### 2012-11-01 doc/prelim/presentation/prelim_presentation.tex
updated presentation
fixed adjoint expectation value
presentation updates

@@
   \[
-  \ve{A} \ve{x} = \ve{b}\:
+  \ve{A} \ve{x} = \ve{b}
   \]
@@
   \rightarrow \mathbb{R}^{N},\ \ve{x} \in \mathbb{R}^N,\ \ve{b} \in
-  \mathbb{R}^N\:
-  \]
-
-  \[
-  \ve{r} = \ve{b} - \ve{A}\ve{x}\:
-  \]
-
-  \begin{itemize}
-  \item $\ve{r}=\ve{0}$ when an exact solution is found.
+  \mathbb{R}^N
+  \]
+
+  \[
+  \ve{r} = \ve{b} - \ve{A}\ve{x}
+  \]
+
+  \begin{itemize}
+  \item $\ve{r}=\ve{0}$ when an exact solution is found
   \end{itemize}
@@
     \[
-    \ve{x} = \ve{A}^{-1}\ve{b}\:
+    \ve{x} = \ve{A}^{-1}\ve{b}
     \]
@@
   \[
-  \ve{A} = \ve{M} - \ve{N}\:.
-  \]
-
-  \[
-  \ve{x} = \ve{M}^{-1}\ve{N}\ve{x} + \ve{M}^{-1}\ve{b}\:.
+  \ve{A} = \ve{M} - \ve{N}
+  \]
+
+  \[
+  \ve{x} = \ve{M}^{-1}\ve{N}\ve{x} + \ve{M}^{-1}\ve{b}
   \]
@@
   \[
-  \ve{x}^{k+1} = \ve{H}\ve{x}^{k} + \ve{c}\:.
+  \ve{x}^{k+1} = \ve{H}\ve{x}^{k} + \ve{c}
   \]
@@
   \[
-  \ve{e}^{k+1} = \ve{H} \ve{e}^k\:
+  \ve{e}^{k+1} = \ve{H} \ve{e}^k
   \]
@@
   \[
-  ||\ve{e}^{k}||_2 = \rho(\ve{H})^k ||\ve{e}^0||_2\:,
+  ||\ve{e}^{k}||_2 = \rho(\ve{H})^k ||\ve{e}^0||_2
   \]
@@
     \tilde{\ve{x}} = \ve{x}_0 +
-    \boldsymbol{\delta},\ \boldsymbol{\delta} \in \mathcal{K}\:
+    \boldsymbol{\delta},\ \boldsymbol{\delta} \in \mathcal{K}
     \]
@@
     \langle \tilde{\ve{r}},\ve{w} \rangle = 0,\ \forall \ve{w} \in
-    \mathcal{L}\:
+    \mathcal{L}
     \]  
@@
     ||\tilde{\ve{r}}||_2 \leq ||\ve{r}_0||_2,\ \forall \ve{r}_0 \in
-    \mathbb{R}^N\:,
+    \mathbb{R}^N
     \]
@@
     \[
-    \ve{x}^{k+1} = \ve{x}^k + \ve{V}\ve{y}^k\:
+    \ve{x}^{k+1} = \ve{x}^k + \ve{V}\ve{y}^k
     \]
@@
     \sum_{i_k}^{N} p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
-    w_{i,i_1}w_{i_1,i_2}\ldots w_{i_{k-1},i_k} b_{i_k}\\ &= x_i\:,
+    w_{i,i_1}w_{i_1,i_2}\ldots w_{i_{k-1},i_k} b_{i_k}\\ &= x_i
   \end{split}
@@
       boundaries of the Cartesian grid. No thermal source was
-      present.}}
+      present. $50 \times 50$ grid.}}
   \end{figure}
@@
     \end{center}
-    \caption{\textbf{Direct solution to Poisson Equation.} \textit{100
-        histories per state, \sn{2.5}{5} total histories. 54.7 seconds
-        CPU time.} }
+    \caption{\textbf{Direct solution to Poisson Equation.}
+      \textit{1000 histories per state, \sn{2.5}{5} total
+        histories. 54.7 seconds CPU time.} }
   \end{figure}
@@
   \[
-  X_{\nu} = \sum_{m=0}^k W_{m} b_{i_0} \delta_{i,i_m}
+  X_{\nu} = \sum_{m=0}^k W_{m} \delta_{i,i_m}
   \]
@@
     E\{X_j\} &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-    \sum_{i_k}^{N} b_{i_0} h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}
-    \delta_{i_k,j} \\ &= x_{j}\:,
+    \sum_{i_k}^{N} b_{i_0} h_{i_0,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}
+    \delta_{i_k,j} \\ &= x_{j}
   \end{split}
@@
   \item Neumann-Ulam methods bound by the Central Limit Theorem
-  \item Halton proposed an iterative residual method
+  \item Halton proposed an iterative residual method in 1962
   \item Iteration error decoupled from Monte Carlo error
@@
   \begin{itemize}
+  \item Evans and Mosher 2009 work
   \item Split the operator to yield Richardson's iteration
@@
     equations? 
-    \pause
-  \item Full-clip roulette for the MCSA correction?
   \end{itemize}

### 2012-11-02 doc/prelim/presentation/prelim_presentation.tex
Removing stationary method slides
pull out some projection method slides

@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Stationary Methods}
-
-  \begin{itemize}
-  \item General stationary methods are formed by splitting the linear
-    operator
-  \end{itemize}
-
-  \[
-  \ve{A} = \ve{M} - \ve{N}
-  \]
-
-  \[
-  \ve{x} = \ve{M}^{-1}\ve{N}\ve{x} + \ve{M}^{-1}\ve{b}
-  \]
-
-  \begin{itemize}
-  \item We identify $\ve{H} =\ve{M}^{-1}\ve{N}$ as the
-    \textit{iteration matrix}
-  \end{itemize}
-
-  \[
-  \ve{x}^{k+1} = \ve{H}\ve{x}^{k} + \ve{c}
-  \]
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Stationary Methods Convergence}
-
-  \begin{itemize}
-  \item The qualities of the iteration matrix dictate convergence
-  \item Define $\ve{e}^k = \ve{x}^k-\ve{x}$ as the error at the
-    $k^{th}$ iterate
-  \end{itemize}
-
-  \[
-  \ve{e}^{k+1} = \ve{H} \ve{e}^k
-  \]
-
-  \begin{itemize}
-  \item We diagonalize $\ve{H}$ to extract its Eigenvalues
-  \end{itemize}
-
-  \[
-  ||\ve{e}^{k}||_2 = \rho(\ve{H})^k ||\ve{e}^0||_2
-  \]
-
-  \begin{itemize}
-  \item We bound $\ve{H}$ by $\rho(\ve{H}) < 1$ for convergence
-  \end{itemize}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
 \begin{frame}{Projection Methods}
@@
 
-    The residual of the system will always be \textit{minimized} with
-    respect to the constraints
+    The residual of the system is \textit{minimized} with respect to
+    the constraints
     \[
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Putting It All Together}
-
-  \begin{itemize}
-  \item Choose $\ve{V}$ as a basis of $\mathcal{K}$ and $\ve{W}$ as a
-    basis of $\mathcal{L}$
-  \end{itemize}
-
-  \[
-  \boldsymbol{\delta} = \ve{V}\ve{y},\ \forall \ve{y} \in \mathbb{R}^N
-  \]
-
-  \[
-  \ve{y} = (\ve{W}^T\ve{A}\ve{V})^{-1}\ve{W}^T\ve{r}_0
-  \]
-
-  \pause
-  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
-    {Projection Method Iteration}
-    \[
-    \ve{r}^k = \ve{b} - \ve{A}\ve{x}^k
-    \]
-    \[
-    \ve{y}^k = (\ve{W}^T\ve{A}\ve{V})^{-1}\ve{W}^T\ve{r}^k
-    \]
-    \[
-    \ve{x}^{k+1} = \ve{x}^k + \ve{V}\ve{y}^k
-    \]
-    \[
-    Update\ \ve{V}\ and\ \ve{W}
-    \]
-  \end{beamerboxesrounded}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
 \begin{frame}{Krylov Subspace Methods}
@@
   \]
+
+  \begin{itemize}
+  \item For GMRES:
+  \end{itemize}
   \[
@@
     \times m}$ for $\mathcal{K}_m(\ve{A},\ve{r}_0)$
-  \item $\ve{W}_m = \ve{A}\ve{V}_m$
   \item Typically choose a Gram-Schmidt-like procedure such as Arnoldi
     or Lanzcos
-  \end{itemize}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}[fragile]{GMRES}
-
-  \begin{algorithm}[H]
-    \begin{algorithmic}[1]
-      \STATE $\ve{r}_0 := \ve{b}-\ve{A}\ve{x}_0$ 
-      \STATE $\beta := ||\ve{r}_0||_2$ 
-      \STATE $\ve{v}_1 := \ve{r}_0 / \beta$
-      \COMMENT{Create the orthonormal basis for the Krylov subspace}
-      \FOR{$j = 1, 2, \cdots, m$} 
-      \STATE $\ve{w}_j := \ve{A}\ve{v}_j$
-      \FOR{$i = 1, 2, \cdots, j$} 
-      \STATE $h_{ij} \leftarrow \langle \ve{w}_j,\ve{v}_j \rangle$ 
-      \STATE $\ve{w}_j \leftarrow \ve{w}_j - h_{ij}\ve{v}_i$ 
-      \ENDFOR 
-      \STATE $h_{j+1,j} \leftarrow ||\ve{w}_j||_2$ 
-      \STATE $\ve{v}_{j+1} \leftarrow \ve{w}_j / h_{j+1,j}$ 
-      \ENDFOR 
-      \COMMENT{Apply the orthogonality constraints}
-      \STATE $\ve{y}_m \leftarrow argmin_y ||\beta \ve{e}_1 -
-      \ve{H}_m\ \ve{y}||_2 $ 
-      \STATE $\ve{x}_m \leftarrow \ve{x}_0 + \ve{V_m} \ve{y}_m$
-    \end{algorithmic}
-    \caption{GMRES Iteration}
-  \end{algorithm}
+  \item Short and long recurrence relations available for
+    orthogonalization
+  \end{itemize}
 

### 2012-11-02 doc/prelim/presentation/prelim_presentation.tex
updated solution evolution captions

@@
     \begin{center}
-      \scalebox{1.25}{ \input{orthogonal_residual.pdftex_t} }
+      \scalebox{1.1}{ \input{orthogonal_residual.pdftex_t} }
     \end{center}
     \caption{\textbf{Orthogonality constraint of the new residual with
-        respect to $\mathcal{L}$.} }
+        respect to $\mathcal{L}$.}}
   \end{figure}
@@
   \begin{itemize}
-  \item For GMRES:
+  \item For GMRES (Saad, 1986):
   \end{itemize}
@@
     \begin{center}
-      \scalebox{0.75}{ \input{partitioned_matrix.pdftex_t} }
+      \scalebox{0.6}{ \input{partitioned_matrix.pdftex_t} }
     \end{center}
@@
     \begin{center}
-      \scalebox{0.75}{ \input{matvec_proc_1.pdftex_t} }
+      \scalebox{0.7}{ \input{matvec_proc_1.pdftex_t} }
     \end{center}
@@
     \caption{\textbf{Direct solution to Poisson Equation.}
-      \textit{1000 histories per state, \sn{2.5}{5} total
+      \textit{100 histories per state, \sn{2.5}{5} total
         histories. 54.7 seconds CPU time.} }
@@
     \end{center}
-    \caption{\textbf{Direct solution to Poisson Equation.} \textit{100
+    \caption{\textbf{Direct solution to Poisson Equation.} \textit{1000
         histories per state, \sn{2.5}{6} total histories. 644 seconds

### 2012-11-03 doc/prelim/presentation/prelim_presentation.tex
adding some references to the presentation
combined Matrix-Free and FAD slides

@@
 
-  \pause
   \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
@@
 
-    \pause
-
     \begin{column}{0.65\textwidth}
@@
   \begin{itemize}
-  \item Powerful class of iterative methods
+  \item Powerful class of iterative methods (Saad,2003)
   \item Provides theory that encapsulates most other iterative methods
@@
 
-  \pause
   \begin{figure}[htpb!]
@@
   \item Global reduction operations observed not to impede scalability
+    (Gropp,2001)
     \begin{itemize}
@@
   \item Iteration error decoupled from Monte Carlo error
-  \item Exponential convergence
+  \item Exponential convergence (Evans,2003)
   \end{itemize}
@@
       pattern to work best
-    \item Poor scaling for unbalanced problems
+    \item Poor scaling for unbalanced problems for all schemes
     \end{itemize}
@@
 
-  \pause
   \begin{itemize}
@@
   \begin{itemize}
-  \item A form of inexact Newton methods
-  \end{itemize}
-
-  \[
-  ||\ve{J}(\ve{u}^k)\delta \ve{u}^k + \ve{F}(\ve{u}^k)|| \leq \eta^k
-  ||\ve{F}(\ve{u}^k)||,\ \eta^k \in [0,1)
-    \]
-
-    \begin{itemize}
   \item Choose a Krylov method to solve for the Newton correction
   \item GMRES with a long recurrence relation observed as more
-      robust
+    robust (Knoll,2004)
   \item Generates a monotonically decreasing residual from
@@
 
-    \pause Where does the Jacobian come from?  \pause
+  \pause Where does the Jacobian come from?
   \begin{itemize}
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Matrix-Free Approximation}
-
+\begin{frame}{Jacobian Approximations}
+
+  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
+    {Matrix-Free Approximation}
     \begin{itemize}
     \item Krylov methods only need the action of the linear operator
-  \item The action of the Jacobian can be approximated using a forward
-    difference
     \end{itemize}
@@
       \begin{itemize}
-    \item Sensitive to scaling and discretization error
-    \item Still have to form part of Jacobian periodically for
-      preconditioning
+      \item Sensitive to scaling and discretization error (Kelly,1995)
       \item Eventually break even with generating and storing the full
-      Jacobian
-    \end{itemize}
-  \end{itemize}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Automatic Differentiation}
-
+        Jacobian (Knoll,1994)
+      \end{itemize}
+    \end{itemize}
+  \end{beamerboxesrounded}
+
+  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
+    {Automatic Differentiation}
     \begin{itemize}
@@
       \end{itemize}
-  \item Modern packages take an element-level assembly approach
-    \begin{itemize}
-    \item Function evaluations are local, communication builds global
-      data
-    \end{itemize}
-  \end{itemize}
-
-  \[
-  \ve{J}(\ve{u}) = \sum_{i=1}^N \ve{Q}^T_i \ve{J}_k \ve{P}_i
-  \]
-  \[
-  e_k : \mathbb{R}^{n_k} \rightarrow \mathbb{R}^{m_k},\ \ve{J}_{k_i} =
-  \partial e_{k_i} / \partial P_i u,\ \ve{P} \in \mathbb{R}^{m_{k_i}
-    \times N}
-  \]
-
-  \begin{itemize}
     \item Performance studies give acceptable results for use in
-    large-scale, production physics codes
-  \end{itemize}
+      large-scale, production physics codes (Bartlett,2006)
+    \end{itemize}
+  \end{beamerboxesrounded}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Jacobian Storage vs. Subspace Storage}
+
+  \begin{columns}
+
+    \begin{column}{0.5\textwidth}
+      \begin{itemize}
+      \item Jacobian will be in a compressed row storage format
+      \end{itemize}
+    \end{column}
+
+    \begin{column}{0.5\textwidth}
+      \small{\[ \ve{A} =
+        \begin{bmatrix}
+          2 & 0 & 8 & 0 & 0 & 0 \\ 4 & 5 & 0 & 1 & 0 & 0 \\ 0 & 2 & 1
+          & 0 & 1 & 0 \\ 0 & 0 & 3 & 7 & 0 & 2 \\ 0 & 0 & 0 & 4 & 9 &
+          0 \\ 0 & 0 & 0 & 0 & 9 & 1 \\
+        \end{bmatrix}      
+        \]}
+    \end{column}
+
+  \end{columns}
+
+  \small{ \[
+    \begin{array}{lccccccccccccccc}
+      \ve{A}\ values: & 2 & 8 & 4 & 5 & 1 & 2 & 1 & 1 & 3 & 7 & 2 & 4
+      & 9 & 9 & 1 \\ column: & 1 & 3 & 1 & 2 & 4 & 2 & 3 & 5 & 3 & 4 &
+      6 & 4 & 5 & 5 & 6 \\ row\ start: & 1 & 3 & 6 & 9 & 12 & 14 & 16
+      \\
+    \end{array}
+    \]}
+
+  \pause
+  \begin{itemize}
+  \item $m$ Krylov iterations require $(m+1)$ subspace vectors
+  \item Storage requirement is $\big[(m+1) N\big]$ for $\ve{x} \in
+    \mathbb{R}^{N \times N}$
+  \end{itemize}
+
+  \pause
+  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
+    {If we need 10 Krylov iterations to converge...}
+    \begin{itemize}
+    \item 66 elements required for subspace vectors
+    \item 72 total elements required for Jacobian and probability
+      matrix
+    \end{itemize}
+  \end{beamerboxesrounded}
 
@@
 
-  \pause
   \begin{algorithm}[H]
@@
   \end{itemize}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Jacobian Storage vs. Subspace Storage}
-
-  \begin{columns}
-
-    \begin{column}{0.5\textwidth}
-      \begin{itemize}
-      \item Jacobian will be in a compressed row storage format
-      \end{itemize}
-    \end{column}
-
-    \begin{column}{0.5\textwidth}
-      \small{\[ \ve{A} =
-        \begin{bmatrix}
-          2 & 0 & 8 & 0 & 0 & 0 \\ 4 & 5 & 0 & 1 & 0 & 0 \\ 0 & 2 & 1
-          & 0 & 1 & 0 \\ 0 & 0 & 3 & 7 & 0 & 2 \\ 0 & 0 & 0 & 4 & 9 &
-          0 \\ 0 & 0 & 0 & 0 & 9 & 1 \\
-        \end{bmatrix}      
-        \]}
-    \end{column}
-
-  \end{columns}
-
-  \pause \small{ \[
-    \begin{array}{lccccccccccccccc}
-      \ve{A}\ values: & 2 & 8 & 4 & 5 & 1 & 2 & 1 & 1 & 3 & 7 & 2 & 4
-      & 9 & 9 & 1 \\ column: & 1 & 3 & 1 & 2 & 4 & 2 & 3 & 5 & 3 & 4 &
-      6 & 4 & 5 & 5 & 6 \\ row\ start: & 1 & 3 & 6 & 9 & 12 & 14 & 16
-      \\
-    \end{array}
-    \]}
-
-  \pause
-  \begin{itemize}
-  \item $m$ Krylov iterations require $(m+1)$ subspace vectors
-  \item Storage requirement is $\big[(m+1) N\big]$ for $\ve{x} \in
-    \mathbb{R}^{N \times N}$
-  \end{itemize}
-
-  \pause
-  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
-    {If we need 10 Krylov iterations to converge...}
-    \begin{itemize}
-    \item 66 elements required for subspace vectors
-    \item 72 total elements required for Jacobian and probability
-      matrix
-    \end{itemize}
-  \end{beamerboxesrounded}
 

### 2012-11-03 doc/prelim/presentation/prelim_presentation.tex
removing experimental framework slide
cleaning up gmres slide
removing mcsa derivation slides

@@
     \ve{A}^T\ve{b}$
-  \item Must generate an orthonormal basis $\ve{V}_m \in \mathbb{R}^{N
-    \times m}$ for $\mathcal{K}_m(\ve{A},\ve{r}_0)$
+  \item Must generate an orthonormal basis for
+    $\mathcal{K}_m(\ve{A},\ve{r}_0)$
   \item Typically choose a Gram-Schmidt-like procedure such as Arnoldi
@@
 \begin{frame}{Monte Carlo Synthetic-Acceleration}
-  \begin{itemize}
-  \item Evans and Mosher 2009 work
-  \item Split the operator to yield Richardson's iteration
-  \end{itemize}
-
-  \[
-  \ve{x} = (\ve{I} - \ve{A})\ve{x} + \ve{b}
-  \]
-  \[
-  \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}
-  \]
-
-  \begin{itemize}
-  \item Define the iteration error
-  \end{itemize}
-
-  \[
-  \delta \ve{x}^k = \ve{x} - \ve{x}^k
-  \]
-  \[
-  \delta \ve{x}^{k+1} = (\ve{I} - \ve{A})\delta \ve{x}^k
-  \]
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Monte Carlo Synthetic-Acceleration}
-
-  \begin{itemize}
-  \item Subtract $(\ve{I} - \ve{A})\delta \ve{x}^{k+1}$
-  \end{itemize}
-
-  \[
-  \begin{split}
-    \ve{A}\delta \ve{x}^{k+1} &= (\ve{I} -
-    \ve{A})(\ve{x}^{k+1}-\ve{x}^{k}) \\ &= \ve{r}^{k+1}
-  \end{split}
-  \]
-
-  \begin{itemize}
-  \item The following converges in one iteration with exact inversion
-    of $\ve{A}$:
-  \end{itemize}
-
-  \[
-  \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}
-  \]
-  \[
-  \ve{A} \delta \ve{x}^{k+1} = \ve{r}^{k+1}
-  \]
-  \[
-  \ve{x} = \ve{x}^{k+1} + \delta \ve{x}^{k+1}
-  \]
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Monte Carlo Synthetic-Acceleration}
 
   \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
-    {MCSA Iteration}
+    {MCSA Iteration (Evans,2009)}
 
@@
   \begin{itemize}
-  \item Experimental framework
   \item Methods verification
@@
   \end{itemize}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Experimental Framework}
-
-  \begin{itemize}
-  \item Need a framework to generate linear and nonlinear systems in
-    parallel
-    \begin{itemize}
-    \item Linear algebra data structures
-    \item Parallel communication framework
-    \item Mesh generation and partitioning
-    \item Automatic differentiation and more...
-    \end{itemize}
-  \item Choose the finite element method to leverage a finite element
-    assembly engine
-  \end{itemize}
-
-  \[
-  \ve{A}(\ve{u}) = \ve{0},\ \forall \ve{u} \in \Omega
-  \]
-  \[
-  \ve{B}(\ve{u}) = \ve{0},\ \forall \ve{u} \in \Gamma
-  \]
-  \[
-  \int_{\Omega} \ve{v}^T \ve{A}(\ve{u}) d\Omega + \int_{\Gamma}
-  \bar{\ve{v}}^T \ve{B}(\ve{u}) d\Gamma = \ve{0}
-  \]
-  \[
-  \int_{\Omega} \ve{C}(\ve{v}^T) \ve{D}(\ve{u}) d\Omega +
-  \int_{\Gamma} \ve{E}(\bar{\ve{v}}^T) \ve{F}(\ve{u}) d\Gamma = \ve{0}
-  \]
 

### 2012-11-03 doc/prelim/presentation/prelim_presentation.tex
removing reproducibility slide
reworked storage slide. added asymptotic relationships

@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Reproducibility}
-
-  \begin{figure}[htpb!]
-    \begin{center}
-      \scalebox{0.7}{ \input{gentile_example.pdftex_t} }
-    \end{center}
-    \caption{\textbf{Gentile's example illustrating how domain
-        decomposition can create reproducibility issues in Monte
-        Carlo.}}
-    \label{fig:gentile_example}
-  \end{figure}
-
-  \begin{itemize}
-  \item 2005 work by Gentile outlines reproducibility in parallel
-    Monte Carlo
-  \item Maintaining random number sequences
-  \item Floating point addition does not commute
-  \item Useful for benchmarking and unit testing
-  \end{itemize}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
 \begin{frame}{Parallel Adjoint Method}
@@
 
+  \begin{itemize}
+  \item CRS matrix with $q$ bands storage: $\big[(2q+1)N\big]$ for $\ve{x} \in
+    \mathbb{R}^{N \times N}$ 
     \pause
-  \begin{itemize}
   \item $m$ Krylov iterations require $(m+1)$ subspace vectors
-  \item Storage requirement is $\big[(m+1) N\big]$ for $\ve{x} \in
+  \item Subspace storage: $\big[(m+1) N\big]$ for $\ve{x} \in
     \mathbb{R}^{N \times N}$
@@
   \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
-    {If we need 10 Krylov iterations to converge...}
-    \begin{itemize}
-    \item 66 elements required for subspace vectors
-    \item 72 total elements required for Jacobian and probability
-      matrix
+    {If we need 25 Krylov iterations to converge...}
+    \begin{itemize}
+    \item Break even scenario: $4q+1 = m$
+    \item Jacobian and probability matrix with 6 bands or less is
+      cheaper
     \end{itemize}

### 2012-11-05 src/Chimera_AdjointNeumannUlamSolver_def.hpp
bug fixes in mcsa implementation

@@ -142,7 +142,7 @@ void AdjointNeumannUlamSolver<Scalar,LO,GO,RNG>::walk()
 	    local_state = state_map->getLocalElement( global_state );
 
 	    // Update LHS tally.
-	    lhs_view[local_state] += bank.top().weight();
+	    lhs_view[local_state] += bank.top().weightAbs();
 
 	    // Sample the probability matrix to get the new state.
 	    d_probability_matrix->getLocalRowView( 
@@ -349,22 +349,17 @@ AdjointNeumannUlamSolver<Scalar,LO,GO,RNG>::sampleSource()
     Teuchos::RCP<const Tpetra::Map<LO,GO> > source_map = source->getMap();
     Teuchos::ArrayView<const GO> global_states = 
 	source_map->getNodeElementList();
-
-    // Precondition the RHS with the linear operator split.    
-    Teuchos::RCP<Tpetra::Vector<Scalar,LO,GO> > precond_source =
-	Tpetra::createVector<Scalar,LO,GO>( source_map );
-    this->b_linear_operator_split->applyInvM( source, precond_source );
     Teuchos::ArrayRCP<const Scalar> local_source_view =
-	precond_source->get1dView();
+	source->get1dView();
 
-    // Stratify sample the preconditioned source.
+    // Stratify sample the source.
     Teuchos::ArrayRCP<GO> starting_states = 
 	SamplingTools::stratifySampleGlobalPDF( 
-	    this->b_histories_per_stage, precond_source );
+	    this->b_histories_per_stage, source );
     testInvariant( local_source_view.size() == starting_states.size() );
 
     // Get the starting source weight.
-    Scalar source_weight = precond_source->norm1();
+    Scalar source_weight = source->norm1();
 
     // Set the relative weight cutoff.
     d_relative_weight_cutoff *= source_weight;

### 2012-11-05 src/Chimera_JacobiSplit_def.hpp

@@ -41,10 +41,13 @@
 #ifndef Chimera_JACOBISPLIT_DEF_HPP
 #define Chimera_JACOBISPLIT_DEF_HPP
 
+#include <algorithm>
+
 #include "Chimera_Assertion.hpp"
 
 #include <Teuchos_ArrayView.hpp>
 #include <Teuchos_Array.hpp>
+#include <Teuchos_ArrayRCP.hpp>
 
 #include <Tpetra_Map.hpp>
 
@@ -109,8 +112,7 @@ void JacobiSplit<Scalar,LO,GO>::split()
     Teuchos::RCP<const Tpetra::Map<LO,GO> > col_map =
         this->b_linear_operator->getColMap();
 
-    // Build the iteration matrix by extracting the diagonal and scaling by
-    // its inverse.
+    // Setup the iteration matrix.
     this->b_iteration_matrix = Teuchos::rcp(
         new Tpetra::CrsMatrix<Scalar,LO,GO>(
             row_map, col_map,
@@ -118,19 +120,44 @@ void JacobiSplit<Scalar,LO,GO>::split()
 
     Teuchos::ArrayView<const LO> local_col_indices;
     Teuchos::ArrayView<const Scalar> local_values;
+    typename Teuchos::ArrayView<const Scalar>::const_iterator local_values_it;
+
+    Teuchos::Array<Scalar> iteration_matrix_vals;
+    typename Teuchos::ArrayView<Scalar>::iterator iteration_matrix_vals_it;
 
     Teuchos::Array<LO> diag_col_index(1,0);
     Teuchos::Array<Scalar> diag_zero(1,0.0);
 
-    for ( LO row_index = row_map->getMinLocalIndex();
-          row_index < row_map->getMaxLocalIndex() + 1;
-          ++row_index )
+    // Get the inverse diagonal.
+    RCP_TpetraVector diagonal_inv =
+        Tpetra::createVector<Scalar,LO,GO>( row_map );
+    this->b_linear_operator->getLocalDiagCopy( *diagonal_inv );
+    diagonal_inv->scale( -1.0 );
+    diagonal_inv->reciprocal( *diagonal_inv );
+    Teuchos::ArrayRCP<const Scalar> diag_inv_view = diagonal_inv->get1dView();
+    typename Teuchos::ArrayRCP<const Scalar>::const_iterator diagonal_inv_it;
+
+    // Build the iteration matrix by extracting the diagonal and scaling by
[…]

### 2012-11-05 src/Chimera_LinearProblem.hpp

@@ -91,22 +91,30 @@ class LinearProblem
     { d_b = b; }
 
     //! Get the operator of the linear problem.
-    RCP_TpetraCrsMatrix getOperator()
+    RCP_TpetraCrsMatrix& getOperator()
+    { return d_A; }
+    const RCP_TpetraCrsMatrix& getOperator() const
     { return d_A; }
 
     //! Get the solution vector of the linear problem.
-    RCP_TpetraVector getLHS()
+    RCP_TpetraVector& getLHS()
+    { return d_x; }
+    const RCP_TpetraVector& getLHS() const
     { return d_x; }
 
     //! Get the right-hand side of the linear problem.
-    RCP_TpetraVector getRHS()
+    RCP_TpetraVector& getRHS()
+    { return d_b; }
+    const RCP_TpetraVector& getRHS() const
     { return d_b; }
 
     // Compute the residual of the linear problem.
     void computeResidual();
 
     //! Get the residual of the linear problem.
-    RCP_TpetraVector getResidual()
+    RCP_TpetraVector& getResidual()
+    { return d_r; }
+    const RCP_TpetraVector& getResidual() const
     { return d_r; }
 
   private:

### 2012-11-05 src/Chimera_LinearSolverFactory_def.hpp

@@ -44,7 +44,7 @@
 #include <string>
 
 #include "Chimera_Assertion.hpp"
-#include "Chimera_BoostRNGTraits.hpp"
+#include "Chimera_BoostRNG.hpp"
 #include "Chimera_StationarySolver.hpp"
 #include "Chimera_SequentialMC.hpp"
 #include "Chimera_MCSA.hpp"
@@ -77,7 +77,7 @@ LinearSolverFactory::create(
 	if ( plist->get<std::string>("RNG TYPE") == "MT19937" )
 	{
 	    linear_solver = Teuchos::rcp( 
-		new SequentialMC<Scalar,LO,GO,Boost::mt19937>( 
+		new SequentialMC<Scalar,LO,GO,boost::mt19937>( 
 		    linear_problem, plist ) );
 	}
     }
@@ -87,7 +87,7 @@ LinearSolverFactory::create(
 	if ( plist->get<std::string>("RNG TYPE") == "MT19937" )
 	{
 	    linear_solver = Teuchos::rcp( 
-		new MCSA<Scalar,LO,GO,Boost::mt19937>( 
+		new MCSA<Scalar,LO,GO,boost::mt19937>( 
 		    linear_problem, plist ) );
 	}
     }

### 2012-11-05 src/Chimera_MCSA_def.hpp

@@ -71,7 +71,7 @@ MCSA<Scalar,LO,GO,RNG>::MCSA(
     // Set the base data.
     this->b_linear_problem = linear_problem;
     this->b_linear_operator_split = LinearOperatorSplitFactory::create( 
-	plist, this->linear_problem->getOperator() );
+	plist, this->b_linear_problem->getOperator() );
     this->b_tolerance = plist->get<Scalar>("TOLERANCE");
     this->b_max_num_iters = plist->get<int>("MAX ITERS");
     this->b_num_iters = 0;
@@ -135,9 +135,7 @@ void MCSA<Scalar,LO,GO,RNG>::iterate()
     this->b_is_converged = false;
 
     // Iterate.
-    while ( residual_norm > 
-	    Teuchos::as<typename Teuchos::ScalarTraits<Scalar>::magnitudeType>(
-		convergence_criteria) &&
+    while ( residual_norm > convergence_criteria &&
 	    this->b_num_iters < this->b_max_num_iters )
     {
 	// Do a stationary iteration.
@@ -150,12 +148,12 @@ void MCSA<Scalar,LO,GO,RNG>::iterate()
 	    this->b_linear_problem->getResidual() );
 
 	// Neumann-Ulam solver for the correction.
-	d_nu_solver->linearProblem->getRHS()->putScalar( 0.0 );
+	d_nu_solver->linearProblem()->getLHS()->putScalar( 0.0 );
 	d_nu_solver->walk();
 
 	// Apply the correction.
-	this->b_linear_problem->getRHS()->update( 
-	    1.0, d_nu_solver->linearProblem->getRHS(), 1.0 );
+	this->b_linear_problem->getLHS()->update( 
+	    1.0, *(d_nu_solver->linearProblem()->getLHS()), 1.0 );
 
 	// Update residual norm for convergence.
 	this->b_linear_problem->computeResidual();

### 2012-11-05 src/Chimera_NeumannUlamSolverFactory.hpp

@@ -78,13 +78,13 @@ class NeumannUlamSolverFactory
 	const Teuchos::RCP<RNG>& rng );
 };
 
+} // end namepsace Chimera
+
 //---------------------------------------------------------------------------//
 // Template includes.
 //---------------------------------------------------------------------------//
 
 #include "Chimera_NeumannUlamSolverFactory_def.hpp"
-
-} // end namepsace Chimera
 
 //---------------------------------------------------------------------------//
 

### 2012-11-05 src/Chimera_SamplingTools_def.hpp

@@ -111,7 +111,6 @@ Teuchos::ArrayRCP<GO> SamplingTools::stratifySampleGlobalPDF(
 	Teuchos::reduceAll( *comm, Teuchos::REDUCE_SUM, local_num_histories, 
 			    Teuchos::Ptr<GO>(&global_sum_check) );
 	);
-
     testPostcondition( global_sum_check == global_num_histories );
 
     // Second, stratify sample the local pdf to get the number of histories to

### 2012-11-05 src/Chimera_SequentialMC.hpp

@@ -96,13 +96,13 @@ class SequentialMC : public LinearSolver<Scalar,LO,GO>
 // Template includes.
 //---------------------------------------------------------------------------//
 
-#include "Chimera_SEQUENTIALMC_def.hpp"
+#include "Chimera_SequentialMC_def.hpp"
 
 //---------------------------------------------------------------------------//
 
 #endif // end Chimera_SEQUENTIALMC_HPP
 
 //---------------------------------------------------------------------------//
-// end Chimera_SEQUENTIALMC.hpp
+// end Chimera_SequentialMC.hpp
 //---------------------------------------------------------------------------//
 

### 2012-11-05 src/Chimera_SequentialMC_def.hpp

@@ -71,7 +71,7 @@ SequentialMC<Scalar,LO,GO,RNG>::SequentialMC(
     // Set the base data.
     this->b_linear_problem = linear_problem;
     this->b_linear_operator_split = LinearOperatorSplitFactory::create( 
-	plist, this->linear_problem->getOperator() );
+	plist, this->b_linear_problem->getOperator() );
     this->b_tolerance = plist->get<Scalar>("TOLERANCE");
     this->b_max_num_iters = plist->get<int>("MAX ITERS");
     this->b_num_iters = 0;
@@ -140,12 +140,12 @@ void SequentialMC<Scalar,LO,GO,RNG>::iterate()
 	    this->b_linear_problem->getResidual() );
 
 	// Neumann-Ulam solver for the correction.
-	d_nu_solver->linearProblem->getRHS()->putScalar( 0.0 );
+	d_nu_solver->linearProblem()->getLHS()->putScalar( 0.0 );
 	d_nu_solver->walk();
 
 	// Apply the correction.
-	this->b_linear_problem->getRHS()->update( 
-	    1.0, d_nu_solver->linearProblem->getRHS(), 1.0 );
+	this->b_linear_problem->getLHS()->update( 
+	    1.0, *(d_nu_solver->linearProblem()->getLHS()), 1.0 );
 
 	// Update residual norm for convergence.
 	this->b_linear_problem->computeResidual();
@@ -180,5 +180,5 @@ void SequentialMC<Scalar,LO,GO,RNG>::iterate()
 #endif // end Chimera_SEQUENTIALMC_DEF_HPP
 
 //---------------------------------------------------------------------------//
-// end Chimera_SEQUENTIALMC_def.hpp
+// end Chimera_SequentialMC_def.hpp
 //---------------------------------------------------------------------------//

### 2012-11-05 src/Chimera_StationaryIteration_def.hpp

@@ -70,21 +70,23 @@ StationaryIteration<Scalar,LO,GO>::~StationaryIteration()
 
 //---------------------------------------------------------------------------//
 /*!
- * \brief Do one stationary iteration.
+ * \brief Do one stationary iteration. x = (M^-1)Nx + (M^-1)b
  */
 template<class Scalar, class LO, class GO>
 void StationaryIteration<Scalar,LO,GO>::doOneIteration()
 {
+    RCP_TpetraVector tmp_vec = Tpetra::createVector<Scalar,LO,GO>( 
+	d_linear_problem->getOperator()->getRowMap() );
+
     d_linear_operator_split->iterationMatrix()->apply( 
-	*(d_linear_problem->getLHS()),
-	*(d_linear_problem->getLHS()) );
+	*(d_linear_problem->getLHS()), *(tmp_vec) );
 
     RCP_TpetraVector m_inv_b = Tpetra::createVector<Scalar,LO,GO>( 
 	d_linear_problem->getOperator()->getRowMap() );
 
     d_linear_operator_split->applyInvM( d_linear_problem->getRHS(), m_inv_b );
 
-    d_linear_problem->getLHS()->update( 1.0, *(m_inv_b), 1.0 );
+    d_linear_problem->getLHS()->update( 1.0, *(m_inv_b), 1.0, *tmp_vec, 0.0 );
 }
 
 //---------------------------------------------------------------------------//

### 2012-11-05 test/tstAdjointNeumannUlamSolver.cpp

@@ -2,7 +2,7 @@
 /*!
  * \file tstAdjointNeumannUlamSolver.cpp
  * \author Stuart R. Slattery
- * \brief Stationary solver tests.
+ * \brief Adjoint Neumann-Ulam solver tests.
  */
 //---------------------------------------------------------------------------//
 
@@ -49,12 +49,12 @@ TEUCHOS_UNIT_TEST( AdjointNeumannUlamSolver, adjoint_neumannulam_test )
 	Teuchos::DefaultComm<int>::getComm();
 
     // Build the linear operator - this is a 2D Transient Diffusion operator.
-    int N = 50;
+    int N = 10;
     int problem_size = N*N;
     double dx = 0.01;
     double dy = 0.01;
-    double dt = 0.5;
-    double alpha = 0.01;
+    double dt = 0.05;
+    double alpha = 0.001;
     Teuchos::RCP<const Tpetra::Map<int> > row_map = 
 	Tpetra::createUniformContigMap<int,int>( problem_size, comm );
     Teuchos::RCP<Tpetra::CrsMatrix<double,int,int> > A = 
@@ -193,7 +193,7 @@ TEUCHOS_UNIT_TEST( AdjointNeumannUlamSolver, adjoint_neumannulam_test )
     // Build the Adjoint solver.
     std::string split_type = "JACOBI";
     double weight_cutoff = 1.0e-4;
-    int histories_per_stage = 1000;
+    int histories_per_stage = 10000;
     Teuchos::RCP<Teuchos::ParameterList> plist =
 	Teuchos::rcp( new Teuchos::ParameterList() );
     plist->set<std::string>("SPLIT TYPE", split_type);
@@ -214,8 +214,8 @@ TEUCHOS_UNIT_TEST( AdjointNeumannUlamSolver, adjoint_neumannulam_test )
     TEST_ASSERT( solver->weightCutoff() == weight_cutoff );
     TEST_ASSERT( solver->historiesPerStage() == histories_per_stage );
 
-    std::cout << "SPEC RAD: " << Chimera::OperatorTools::spectralRadius( 
-	lin_op_split->iterationMatrix() ) << std::endl;
+    // std::cout << "SPEC RAD: " << Chimera::OperatorTools::spectralRadius( 
+    // 	lin_op_split->iterationMatrix() ) << std::endl;
 
     // Solve.
     solver->walk();

### 2012-11-05 test/tstMCSA.cpp

@@ -1,6 +1,6 @@
 //---------------------------------------------------------------------------//
 /*!
- * \file tstAdjointNeumannUlamSolver.cpp
+ * \file tstMCSA.cpp
  * \author Stuart R. Slattery
  * \brief Stationary solver tests.
  */
@@ -17,12 +17,8 @@
 
 #include <Chimera_LinearSolver.hpp>
 #include <Chimera_LinearProblem.hpp>
-#include <Chimera_LinearOperatorSplit.hpp>
-#include <Chimera_LinearOperatorSplitFactory.hpp>
-#include <Chimera_NeumannUlamSolver.hpp>
-#include <Chimera_AdjointNeumannUlamSolver.hpp>
-#include <Chimera_BoostRNG.hpp>
-#include <Chimera_OperatorTools.hpp>
+#include <Chimera_LinearSolverFactory.hpp>
+#include <Chimera_MCSA.hpp>
 
 #include <Teuchos_UnitTestHarness.hpp>
 #include <Teuchos_DefaultComm.hpp>
@@ -42,19 +38,19 @@
 //---------------------------------------------------------------------------//
 // Tests
 //---------------------------------------------------------------------------//
-TEUCHOS_UNIT_TEST( AdjointNeumannUlamSolver, adjoint_neumannulam_test )
+TEUCHOS_UNIT_TEST( MCSA, mcsa_test )
 {
     // Setup parallel distribution.
     Teuchos::RCP<const Teuchos::Comm<int> > comm = 
 	Teuchos::DefaultComm<int>::getComm();
 
     // Build the linear operator - this is a 2D Transient Diffusion operator.
-    int N = 50;
+    int N = 10;
     int problem_size = N*N;
     double dx = 0.01;
     double dy = 0.01;
-    double dt = 0.5;
-    double alpha = 0.01;
+    double dt = 0.05;
+    double alpha = 0.001;
     Teuchos::RCP<const Tpetra::Map<int> > row_map = 
 	Tpetra::createUniformContigMap<int,int>( problem_size, comm );
     Teuchos::RCP<Tpetra::CrsMatrix<double,int,int> > A = 
@@ -186,39 +182,23 @@ TEUCHOS_UNIT_TEST( AdjointNeumannUlamSolver, adjoint_neumannulam_test )
     Teuchos::RCP<Chimera::LinearProblem<double,int,int> > linear_problem = 
 	Teuchos::rcp( new Chimera::LinearProblem<double,int,int>( A, X, B ) );
 
-    // Build the random number generator.
[…]

### 2012-11-05 test/tstMCSA.cpp
working on MCSA. still some issues

@@ -192,7 +192,7 @@ TEUCHOS_UNIT_TEST( MCSA, mcsa_test )
     plist->set<std::string>( "SPLIT TYPE",          "JACOBI"  );
     plist->set<std::string>( "MC TYPE",             "ADJOINT" );
     plist->set<double>(      "WEIGHT CUTOFF",       1.0e-4    );
-    plist->set<int>(         "HISTORIES PER STAGE", 10000     );
+    plist->set<int>(         "HISTORIES PER STAGE", 100       );
 
     Teuchos::RCP<Chimera::LinearSolver<double,int,int> > solver =
 	Chimera::LinearSolverFactory::create( plist, linear_problem );

### 2012-11-05 doc/prelim/presentation/prelim_presentation.tex
updated storage slide
reduced nonlinear preliminaries to 1 slide

@@
     $\ve{F}(\ve{u})$
-  \item Taylor expand the residuals at the $k+1$ iterate about the $k$
-    iterate
   \end{itemize}
@@
   \frac{\ve{F}''(\ve{u}^{k})}{2}(\ve{u}^{k+1}-\ve{u}^{k})^2 + \cdots
-  \]
-
-  \begin{itemize}
-  \item Assert $\ve{u}^{k+1}$ is the exact solution
-  \end{itemize}
-
-  \[
-  -\ve{F}(\ve{u}^{k}) = \ve{F}'(\ve{u}^{k})(\ve{u}^{k+1}-\ve{u}^{k})
-  \]
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Nonlinear Preliminaries}
-
-  \begin{itemize}
-  \item $\ve{F}'(\ve{u}^{k})$ is the \textit{Jacobian}
-    $\ve{J}(\ve{u})$
-  \end{itemize}
-
-  \[
-  J_{ij} = \frac{\partial F_i(\ve{u})}{\partial u_j}
-  \]
-
-  \begin{itemize}
-  \item $(\ve{u}^{k+1}-\ve{u}^{k})$ is the solution update from the
-    $k$ iterate to the $k+1$ iterate
-  \end{itemize}
-
-  \[
-  \delta \ve{u}^k = \ve{u}^{k+1}-\ve{u}^{k}
   \]
@@
     \begin{itemize}
-    \item Break even scenario: $4q+1 = m$
-    \item Jacobian and probability matrix with 6 bands or less is
-      cheaper
+    \item Break-even scenario: $4q+1 = m$
+    \item Jacobian and probability matrix storage with 6 bands or less
+      is cheaper
     \end{itemize}

### 2012-11-05 doc/prelim/presentation/prelim_presentation.tex
Removed departure from nucleate boiling example from slides
removing sequential monte carlo slide

@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Physics-Based Motivation: DNB}
-
-  \begin{columns}
-
-    \begin{column}{0.35\textwidth}
-      \begin{figure}[htpb!]
-        \begin{center}
-          \scalebox{1}{ \input{dnb_schematic.pdftex_t} }
-        \end{center}
-        \caption{\textbf{Departure from nucleate boiling scenario.} }
-      \end{figure}
-    \end{column}
-
-    \begin{column}{0.65\textwidth}
-      \begin{figure}[htpb!]
-        \begin{center}
-          \scalebox{0.8}{ \input{dnb_example.pdftex_t} }
-        \end{center}
-        \caption{\textbf{Multiphysics dependency analysis of departure
-            from nucleate boiling.} }
-      \end{figure}
-    \end{column}
-
-  \end{columns}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
 \begin{frame}{Hardware-Based Motivation}
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Sequential Monte Carlo}
-
-  \begin{itemize}
-  \item Neumann-Ulam methods bound by the Central Limit Theorem
-  \item Halton proposed an iterative residual method in 1962
-  \item Iteration error decoupled from Monte Carlo error
-  \item Exponential convergence (Evans,2003)
-  \end{itemize}
-
-  \[
-  \ve{r}^k = \ve{b} - \ve{A}\ve{x}^k
-  \]
-  \[
-  \ve{A}\boldsymbol{\delta}^{k} = \ve{r}^{k}
-  \]
-  \[
-  \ve{x}^{k+1} = \ve{x}^k + \boldsymbol{\delta}^{k}
-  \]
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
 \begin{frame}{Monte Carlo Synthetic-Acceleration}
@@
   \begin{itemize}
+  \item Neumann-Ulam methods bound by the Central Limit Theorem
+  \item Build on Halton's 1962 iterative residual method
   \item Adjoint Neumann-Ulam solver computes the correction
-  \item Decouples Monte Carlo error from solution error
-  \item Exponential convergence
-  \item Demonstrated by Evans and colleagues to be competitive with
-    Krylov methods
+  \item Decouples MC error from solution error, exponential convergence
+  \item Demonstrated by Evans to be competitive with Krylov methods
   \end{itemize}

### 2012-11-05 doc/prelim/presentation/prelim_presentation.tex
removing sequential monte carlo slide
adding acknowledgements slide
updating acks slide

@@
   \item Neumann-Ulam methods bound by the Central Limit Theorem
-  \item Build on Halton's 1962 iterative residual method
+  \item Build on Halton's 1962 Sequential Monte Carlo method
   \item Adjoint Neumann-Ulam solver computes the correction
@@
 %%---------------------------------------------------------------------------%%
+\begin{frame}{Acknowledgements}
+
+  Thanks to...\footnote{This work was performed under
+        appointment to the Nuclear Regulatory Commission Fellowship
+        program at the University of Wisconsin - Madison Department of
+        Engineering Physics.}
+
+  \begin{itemize}
+  \item Paul Wilson
+  \item Tom Evans
+  \item ORNL and SNL
+  \end{itemize}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
 

### 2012-11-05 doc/prelim/presentation/prelim_presentation.tex
cleaning up numerical experiments

@@
   \pause
-  Domain Overlap Studies for the Parallel Neumann-Ulam Method
+  \textbf{Domain Overlap Studies for Parallel Neumann-Ulam Method} 
   \begin{itemize}
@@
   \pause
-  Domain Replication Studies for the Parallel Neumann-Ulam Method
+  \textbf{Domain Replication Studies for Parallel Neumann-Ulam Method}
   \begin{itemize}
@@
 
-  Parallel Performance and Numerical Accuracy Studies for the MCSA
-  Method
+  \textbf{Parallel Performance and Numerical Accuracy Studies for MCSA
+    Method}
   \begin{itemize}
@@
   \pause
-  Parallel Performance and Numerical Accuracy Studies for the FANM
-  Method
+  \textbf{Parallel Performance and Numerical Accuracy Studies for FANM
+    Method}
   \begin{itemize}

### 2012-11-05 doc/prelim/presentation/prelim_presentation.tex
adding hardware motivation references
adding references to physics based motivation

@@
   \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
-    {Predictive nuclear reactor analysis enables...}
+    {Predictive nuclear reactor analysis enables...            (DOE,2011)}
     \begin{itemize}
@@
       $O(\sn{1}{9})$ element meshes and $O(100,000)+$ cores used in
-      today's simulations.
+      today's simulations (Evans,2010)(Pawlowski,2012)
     \end{itemize}
@@
   \begin{itemize}
-  \item Modern hardware is moving in two directions:
+  \item Modern hardware is moving in two directions (Kogge,2011):
     \begin{itemize}
@@
     \begin{itemize}
-    \item Higher potential for both soft and hard failures
+    \item Higher potential for both soft and hard failures (DOE,2012)
     \item Memory restrictions are expected with a continued decrease

### 2012-11-05 doc/prelim/presentation/prelim_presentation.tex
adding spacing to slides

@@
     capability
+    \medskip
   \item Modern work focused on this task leverages multiple physics
     simulation (CASL, NEAMS)
+    \medskip
   \item New hardware drives algorithm development (petascale and
     exascale)
+    \medskip
   \item Monte Carlo methods have the potential to provide great
@@
     future hardware
+    \medskip
   \item A set of massively parallel Monte Carlo methods is proposed to
@@
     \end{itemize}
+    \medskip \medskip
   \item Some issues:
@@
     \end{itemize}
+    \medskip \medskip
   \item Potential resolution from Monte Carlo:
@@
     \end{itemize}
+    \medskip \medskip \medskip
   \item Development of a nonlinear solver for discrete systems
@@
 
+  \medskip \medskip 
   \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
@@
 
+  \medskip \medskip
   \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
@@
   \begin{itemize}
-  \item For GMRES (Saad, 1986):
-  \end{itemize}
-  \[
-  \mathcal{L} = \ve{A} \mathcal{K}_m(\ve{A},\ve{r}_0)
-  \]
-
+  \item For GMRES (Saad, 1986): $\mathcal{L} = \ve{A}
+    \mathcal{K}_m(\ve{A},\ve{r}_0)$
+  \end{itemize}
+
+  \medskip \medskip \medskip
+  \small{
   \begin{itemize}
@@
     \ve{A}^T\ve{b}$
+    \medskip
   \item Must generate an orthonormal basis for
     $\mathcal{K}_m(\ve{A},\ve{r}_0)$
+    \medskip
   \item Typically choose a Gram-Schmidt-like procedure such as Arnoldi
     or Lanzcos
+    \medskip
   \item Short and long recurrence relations available for
@@
   \end{itemize}
+  }
 
@@
 
+  \medskip \medskip
   \begin{itemize}
@@
   \item Widely used in practice
+    \medskip
   \item Krylov methods require only the action of the operator
+    \medskip
   \item Parallelism achieved through a handful of operations
+    \medskip
   \item Global reduction operations observed not to impede scalability
@@
     \end{itemize}
+    \medskip
   \item Nearest neighbor computations have poor algorithmic strong
@@
     \end{itemize}
+    \medskip
   \item Short and long recurrence relations available for
@@
   \item First proposed by J. Von Neumann and S.M. Ulam in the 1940's
+    \medskip \medskip
   \item Earliest published reference in 1950
+    \medskip \medskip
   \item General lack of published work
+    \medskip \medskip
   \item Modern work by Evans and others has yielded new applications
@@
 
+  \medskip
   \begin{itemize}
@@
   \[
-  \ve{y} = (\ve{I} - \ve{H}^T)^{-1} \ve{d}
-  \]
-
-  \[
-  \ve{y} = \sum_{k=0}^{\infty} (\ve{H}^T)^k\ve{d}
-  \]
-
+  \ve{y} = (\ve{I} - \ve{H}^T)^{-1} \ve{d} = \sum_{k=0}^{\infty}
+  (\ve{H}^T)^k\ve{d}
+  \]
+
+  \medskip
   \begin{itemize}
@@
     \end{itemize}
+    \medskip \medskip
   \item Desire a generalization for all linear operator equations
@@
     \end{itemize}
+    \medskip \medskip
   \item Neumann-Ulam solvers and MCSA implemented using the Trilinos
     Petra frameworks
+    \medskip \medskip
   \item Can be leveraged in modern physics implementations
@@
   \item No literature observed for parallel Neumann-Ulam solvers
+    \medskip \medskip
   \item Numerous references for modern parallel Monte Carlo methods in
     reactor physics
+    \medskip \medskip
   \item Build a strategy for applying modern methods to the
     Neumann-Ulam method
+    \medskip \medskip
   \item MCSA iteration-level parallelism comes from parallel
@@
       \item Fully-locking synchronous
-      \item asynchronous-send/synchronous-receive
-      \item master/slave
-      \item binary tree master/slave
+      \item Asynchronous-send/synchronous-receive
+      \item Master/slave
+      \item Binary tree master/slave
       \end{itemize}
@@
   \item The amount of domain leakage dictates communication
+    \medskip
   \item Per Siegel's 2012 work, define a leakage fraction
@@
     free path of the system
+    \medskip
   \item Optimum ratio of local to global domain size exists
+    \medskip
   \item Experiments show feasibility for large-scale calculations
@@
     \pause
+    \medskip
   \item How do we define $\lambda$ empirically for linear operator
@@
 
+  \medskip \medskip \medskip
 Questions
@@
   \item Many multiphysics problems of interest are nonlinear
+    \medskip
   \item Segregated methods lack the consistency of fully implicit
     methods
+    \medskip
   \item Newton methods often leverage Krylov solvers
@@
     \end{itemize}
+    \medskip
   \item Monte Carlo methods need the full operator
+    \medskip
   \item Automatic construction of the linear operator available
@@
   \item Choose a Krylov method to solve for the Newton correction
+    \medskip
   \item GMRES with a long recurrence relation observed as more
     robust (Knoll,2004)
+    \medskip
   \item Generates a monotonically decreasing residual from
@@
 
+  \medskip \medskip \medskip
   \pause Where does the Jacobian come from?
@@
 
+  \large{
   \begin{itemize}
   \item Methods verification
+  \medskip \medskip \medskip \medskip \medskip
   \item Numerical experiments
+    \medskip \medskip \medskip \medskip \medskip
   \item Challenge problem
   \end{itemize}
+  }
 
@@
   \item Analytic solution to the heat equation for the linear methods
+    \medskip
   \item Sequence of Navier-Stokes benchmarks for the nonlinear methods
@@
     \end{itemize}
+    \medskip
   \item Tuning benchmark parameters varies the strength of
@@
 
+  \medskip  \medskip
   \begin{columns}
@@
 
+  \medskip  \medskip
   \pause
@@
     reactor physics
+    \medskip
   \item The Consortium for Advanced Simulation of LWRs (CASL) is a
@@
     \end{itemize}
+    \medskip
   \item CASL and industry partners identified challenge problems:
@@
     \end{itemize}
+    \medskip
   \item Our challenge problem should reflect how this work aids the
@@
     \end{itemize}
+    \medskip
   \item Verification through benchmarks
+    \medskip
   \item Numerical experiments for understanding
+    \medskip
   \item Challenge problem for application
+    \medskip
   \item Directed towards hard problems in nuclear reactor analysis:
@@
 
-  Thanks to...\footnote{This work was performed under
-        appointment to the Nuclear Regulatory Commission Fellowship
-        program at the University of Wisconsin - Madison Department of
-        Engineering Physics.}
+  Thanks to...
 
@@
 
+  \medskip \medskip \medskip \medskip \medskip \medskip \medskip
+
+  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
+    {Support} 
+    This work was performed under appointment to the Nuclear Regulatory
+    Commission Fellowship program at the University of Wisconsin - Madison
+    Department of Engineering Physics.
+  \end{beamerboxesrounded}
+
 \end{frame}

### 2012-11-05 doc/prelim/presentation/prelim_presentation.tex
removing linear operator equations slides
updating krylov slide

@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Linear Operator Equations}
-
-  \begin{itemize}
-  \item We seek solutions of the general linear operator equation
-  \end{itemize}
-
-  \[
-  \ve{A} \ve{x} = \ve{b}
-  \]
-  \[
-  \ve{A} \in \mathbb{R}^{N \times N},\ \ve{A} : \mathbb{R}^{N}
-  \rightarrow \mathbb{R}^{N},\ \ve{x} \in \mathbb{R}^N,\ \ve{b} \in
-  \mathbb{R}^N
-  \]
-
-  \[
-  \ve{r} = \ve{b} - \ve{A}\ve{x}
-  \]
-
-  \begin{itemize}
-  \item $\ve{r}=\ve{0}$ when an exact solution is found
-  \end{itemize}
-
-  \medskip \medskip 
-  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
-    {A Requirement} Assert that $\ve{A}$ is \textit{nonsingular}. The
-    solution is then:
-    \[
-    \ve{x} = \ve{A}^{-1}\ve{b}
-    \]
-  \end{beamerboxesrounded}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
 \begin{frame}{Projection Methods}
@@
     \ve{A}^T\ve{b}$
+    \medskip
+  \item Require only the action of the operator
     \medskip

### 2012-11-05 doc/prelim/presentation/prelim_presentation.tex
updating krylov slide
fixed typo
cleaning up krylov slides

@@
   \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
-    {Search Subspace $\mathcal{K}$} Extract the solution from the
-    search subspace:
+    {Search Subspace $\mathcal{K}$} Extract the solution to
+    $\ve{A}\ve{x}=\ve{b}$ from the search subspace:
     \[
@@
     the constraint subspace by asserting orthogonality with the
-    residual:
+    residual ($\ve{r} = \ve{b}-\ve{A}\ve{x}$):
     \[
@@
     $\mathcal{K}_m(\ve{A},\ve{r}_0)$
-    \medskip
+    \begin{itemize}
     \item Typically choose a Gram-Schmidt-like procedure such as Arnoldi
       or Lanzcos
-    \medskip
     \item Short and long recurrence relations available for
       orthogonalization
+    \end{itemize}
   \end{itemize}
@@
   \begin{itemize}
-  \item Widely used in practice
-    \medskip
-  \item Krylov methods require only the action of the operator
-    \medskip
-  \item Parallelism achieved through a handful of operations
-    \medskip
   \item Global reduction operations observed not to impede scalability
@@
     \end{itemize}
+    \medskip
+  \item Widely used in practice
+    \medskip
+  \item Krylov methods require only the action of the operator
+    \medskip
+  \item Parallelism achieved through a handful of operations
     \medskip
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Acknowledgements}
+\begin{frame}{Acknowledgments}
 

### 2012-11-05 src/Chimera_AdjointNeumannUlamSolver_def.hpp
fixed failing stationary iteration test

@@ -57,7 +57,6 @@
 #include <Teuchos_OrdinalTraits.hpp>
 
 #include <Tpetra_Map.hpp>
-#include <Tpetra_CrsMatrix.hpp>
 #include <Tpetra_Export.hpp>
 
 namespace Chimera
@@ -133,8 +132,8 @@ void AdjointNeumannUlamSolver<Scalar,LO,GO,RNG>::walk()
     while ( walk )
     {
 	// If the bank isn't empty, process the top history. We need this here
-	// so proccesses that have emptied their banks can wait until all
-	// banks are empty and buffer communication can occur.
+	// so processes that have emptied their banks can wait until all banks
+	// are empty and buffer communication can occur.
 	if ( !bank.empty() )
 	{
 	    // Get the current history state.
@@ -150,8 +149,13 @@ void AdjointNeumannUlamSolver<Scalar,LO,GO,RNG>::walk()
 	    new_local_state = SamplingTools::sampleLocalDiscretePDF(
 		local_values, local_indices, this->b_rng );
 
+	    // Invalid state. Terminate the history.
+	    if ( new_local_state == Teuchos::OrdinalTraits<LO>::invalid() )
+	    {
+		transition_weight = 0.0;
+	    }
 	    // The new state is valid.
-	    if ( new_local_state != Teuchos::OrdinalTraits<LO>::invalid() )
+	    else
 	    {
 		new_global_state = col_map->getGlobalElement( new_local_state );
 
@@ -190,13 +194,9 @@ void AdjointNeumannUlamSolver<Scalar,LO,GO,RNG>::walk()
 		    transition_weight = std::abs( transition_h / transition_p );
 		}
 	    }
-	    // Invalid state. Terminate the history.
-	    else
-	    {
-		transition_weight = 0.0;
-	    }
 
-	    // We want to check this to insure the weight is decreasing.
+	    // We want to check this to insure the weight is decreasing for
+	    // convergence.
 	    testInvariant( 1.0 > transition_weight );
 
 	    // Update the history for the transition.

### 2012-11-05 src/Chimera_JacobiSplit_def.hpp

@@ -123,7 +123,7 @@ void JacobiSplit<Scalar,LO,GO>::split()
     typename Teuchos::ArrayView<const Scalar>::const_iterator local_values_it;
 
     Teuchos::Array<Scalar> iteration_matrix_vals;
-    typename Teuchos::ArrayView<Scalar>::iterator iteration_matrix_vals_it;
+    typename Teuchos::Array<Scalar>::iterator iteration_matrix_vals_it;
 
     Teuchos::Array<LO> diag_col_index(1,0);
     Teuchos::Array<Scalar> diag_zero(1,0.0);
@@ -148,6 +148,7 @@ void JacobiSplit<Scalar,LO,GO>::split()
             row_index, local_col_indices, local_values );
 
 	iteration_matrix_vals.resize( local_values.size() );
+
 	for ( local_values_it = local_values.begin(),
      iteration_matrix_vals_it = iteration_matrix_vals.begin();
 	      local_values_it != local_values.end();

### 2012-11-05 test/tstAdjointNeumannUlamSolver.cpp

@@ -203,8 +203,8 @@ TEUCHOS_UNIT_TEST( AdjointNeumannUlamSolver, adjoint_neumannulam_test )
     Teuchos::RCP<Chimera::LinearOperatorSplit<double,int,int> > lin_op_split =
 	Chimera::LinearOperatorSplitFactory::create( plist, A );
 
-    Teuchos::RCP<Chimera::NeumannUlamSolver<double,int,int,boost::mt19937> > solver =
-	Teuchos::rcp( 
+    Teuchos::RCP<Chimera::NeumannUlamSolver<double,int,int,boost::mt19937> >
+	solver = Teuchos::rcp( 
 	    new Chimera::AdjointNeumannUlamSolver<double,int,int,boost::mt19937>(
 		linear_problem, lin_op_split, rng, plist ) );
 

### 2012-11-05 test/tstStationaryIteration.cpp

@@ -60,7 +60,7 @@ TEUCHOS_UNIT_TEST( StationaryIteration, jacobi_iteration_test )
     Teuchos::Array<double> matrix_values(3);
     matrix_values[0] = 1.0;
     matrix_values[1] = 2.0;
-    matrix_values[2] = 3.0;
+    matrix_values[2] = 4.0;
     int gid = 0;
     for ( int i = 1; i < local_num_rows-1; ++i )
     {
@@ -73,8 +73,8 @@ TEUCHOS_UNIT_TEST( StationaryIteration, jacobi_iteration_test )
 
     Teuchos::Array<int> edge_global_columns(2);
     Teuchos::Array<double> edge_matrix_values(2);
-    edge_matrix_values[0] = 2.0;
-    edge_matrix_values[1] = 3.0;
+    edge_matrix_values[0] = matrix_values[1];
+    edge_matrix_values[1] = matrix_values[2];
     gid = local_num_rows*comm_rank;
     edge_global_columns[0] = gid;
     edge_global_columns[1] = gid+1;
@@ -91,8 +91,8 @@ TEUCHOS_UNIT_TEST( StationaryIteration, jacobi_iteration_test )
     }
     comm->barrier();
 
-    edge_matrix_values[0] = 1.0;
-    edge_matrix_values[1] = 2.0;
+    edge_matrix_values[0] = matrix_values[0];
+    edge_matrix_values[1] = matrix_values[1];
     gid = local_num_rows - 1 + local_num_rows*comm_rank;
     edge_global_columns[0] = gid-1;
     edge_global_columns[1] = gid;
@@ -160,30 +160,35 @@ TEUCHOS_UNIT_TEST( StationaryIteration, jacobi_iteration_test )
     for ( int i = 1; i < local_num_rows-1; ++i )
     {
 	TEST_ASSERT( X_view[i] == 
-		     B_val*(1.0+matrix_values[0]+matrix_values[2])/matrix_values[1] );
+		     (1-(matrix_values[0]+matrix_values[2])/
+		      matrix_values[1])*B_val/matrix_values[1] );
     }
 
     if ( comm_rank == 0 )
     {
 	TEST_ASSERT( X_view[0] == 
-		     B_val*(1.0+matrix_values[2])/matrix_values[1] );
+		     (1-(matrix_values[2])/
+		      matrix_values[1])*B_val/matrix_values[1] );   
     }
     else
     {
 	TEST_ASSERT( X_view[0] == 
[…]

### 2012-11-05 test/tstStationarySolver.cpp

@@ -59,7 +59,7 @@ TEUCHOS_UNIT_TEST( StationarySolver, jacobi_solve_test )
     Teuchos::Array<double> matrix_values(3);
     matrix_values[0] = 1.0;
     matrix_values[1] = 2.0;
-    matrix_values[2] = 3.0;
+    matrix_values[2] = 4.0;
     int gid = 0;
     for ( int i = 1; i < local_num_rows-1; ++i )
     {
@@ -72,8 +72,8 @@ TEUCHOS_UNIT_TEST( StationarySolver, jacobi_solve_test )
 
     Teuchos::Array<int> edge_global_columns(2);
     Teuchos::Array<double> edge_matrix_values(2);
-    edge_matrix_values[0] = 2.0;
-    edge_matrix_values[1] = 3.0;
+    edge_matrix_values[0] = matrix_values[1];
+    edge_matrix_values[1] = matrix_values[2];
     gid = local_num_rows*comm_rank;
     edge_global_columns[0] = gid;
     edge_global_columns[1] = gid+1;
@@ -90,8 +90,8 @@ TEUCHOS_UNIT_TEST( StationarySolver, jacobi_solve_test )
     }
     comm->barrier();
 
-    edge_matrix_values[0] = 1.0;
-    edge_matrix_values[1] = 2.0;
+    edge_matrix_values[0] = matrix_values[0];
+    edge_matrix_values[1] = matrix_values[1];
     gid = local_num_rows - 1 + local_num_rows*comm_rank;
     edge_global_columns[0] = gid-1;
     edge_global_columns[1] = gid;
@@ -149,35 +149,39 @@ TEUCHOS_UNIT_TEST( StationarySolver, jacobi_solve_test )
     solver->iterate();
 
     // Check the solution vector.
-    Teuchos::ArrayRCP<const double> X_view = 
-	linear_problem->getLHS()->get1dView();
+    Teuchos::ArrayRCP<const double> X_view = linear_problem->getLHS()->get1dView();
     for ( int i = 1; i < local_num_rows-1; ++i )
     {
 	TEST_ASSERT( X_view[i] == 
-		     B_val*(1.0+matrix_values[0]+matrix_values[2])/matrix_values[1] );
+		     (1-(matrix_values[0]+matrix_values[2])/
+		      matrix_values[1])*B_val/matrix_values[1] );
     }
 
     if ( comm_rank == 0 )
     {
 	TEST_ASSERT( X_view[0] == 
[…]

### 2012-11-05 test/tstLinearSolverFactory.cpp
fixed all failing stationary tests

@@ -59,7 +59,7 @@ TEUCHOS_UNIT_TEST( LinearSolverFactory, linear_solve_test )
     Teuchos::Array<double> matrix_values(3);
     matrix_values[0] = 1.0;
     matrix_values[1] = 2.0;
-    matrix_values[2] = 3.0;
+    matrix_values[2] = 4.0;
     int gid = 0;
     for ( int i = 1; i < local_num_rows-1; ++i )
     {
@@ -72,8 +72,8 @@ TEUCHOS_UNIT_TEST( LinearSolverFactory, linear_solve_test )
 
     Teuchos::Array<int> edge_global_columns(2);
     Teuchos::Array<double> edge_matrix_values(2);
-    edge_matrix_values[0] = 2.0;
-    edge_matrix_values[1] = 3.0;
+    edge_matrix_values[0] = matrix_values[1];
+    edge_matrix_values[1] = matrix_values[2];
     gid = local_num_rows*comm_rank;
     edge_global_columns[0] = gid;
     edge_global_columns[1] = gid+1;
@@ -90,8 +90,8 @@ TEUCHOS_UNIT_TEST( LinearSolverFactory, linear_solve_test )
     }
     comm->barrier();
 
-    edge_matrix_values[0] = 1.0;
-    edge_matrix_values[1] = 2.0;
+    edge_matrix_values[0] = matrix_values[0];
+    edge_matrix_values[1] = matrix_values[1];
     gid = local_num_rows - 1 + local_num_rows*comm_rank;
     edge_global_columns[0] = gid-1;
     edge_global_columns[1] = gid;
@@ -153,30 +153,35 @@ TEUCHOS_UNIT_TEST( LinearSolverFactory, linear_solve_test )
     for ( int i = 1; i < local_num_rows-1; ++i )
     {
 	TEST_ASSERT( X_view[i] == 
-		     B_val*(1.0+matrix_values[0]+matrix_values[2])/matrix_values[1] );
+		     (1-(matrix_values[0]+matrix_values[2])/
+		      matrix_values[1])*B_val/matrix_values[1] );
     }
 
     if ( comm_rank == 0 )
     {
 	TEST_ASSERT( X_view[0] == 
-		     B_val*(1.0+matrix_values[2])/matrix_values[1] );
+		     (1-(matrix_values[2])/
+		      matrix_values[1])*B_val/matrix_values[1] );   
     }
     else
     {
 	TEST_ASSERT( X_view[0] == 
[…]

### 2012-11-05 src/Chimera_AdjointNeumannUlamSolver_def.hpp
fixed adjoint solver for weight bug. everything seems to be functioning in parallel. still a small scaling issue for mcsa and sequential in output

@@ -112,7 +112,7 @@ void AdjointNeumannUlamSolver<Scalar,LO,GO,RNG>::walk()
     this->b_linear_problem->getLHS()->putScalar( 0.0 );
     Teuchos::ArrayRCP<Scalar> lhs_view =     
 	this->b_linear_problem->getLHS()->get1dViewNonConst();
-   
+
     // Sample the source and populate the history bank.
     HistoryBank<HistoryType> bank = sampleSource();
 
@@ -141,7 +141,7 @@ void AdjointNeumannUlamSolver<Scalar,LO,GO,RNG>::walk()
 	    local_state = state_map->getLocalElement( global_state );
 
 	    // Update LHS tally.
-	    lhs_view[local_state] += bank.top().weightAbs();
+	    lhs_view[local_state] += bank.top().weight();
 
 	    // Sample the probability matrix to get the new state.
 	    d_probability_matrix->getLocalRowView( 
@@ -197,7 +197,7 @@ void AdjointNeumannUlamSolver<Scalar,LO,GO,RNG>::walk()
 
 	    // We want to check this to insure the weight is decreasing for
 	    // convergence.
-	    testInvariant( 1.0 > transition_weight );
+	    testInvariant( 1.0 > transition_weight && transition_weight >= 0.0 );
 
 	    // Update the history for the transition.
 	    bank.top().setGlobalState( new_global_state );
@@ -311,7 +311,8 @@ void AdjointNeumannUlamSolver<Scalar,LO,GO,RNG>::buildProbabilityMatrix()
 //---------------------------------------------------------------------------//
 /*!
  * \brief Build the ghost iteration matrix. This consists of the non-zero
- * off-process components that we will need for computing the adjoint weights.
+ * off-process components that we will need for computing the adjoint weights
+ * before the history leaves the local domain.
  */
 template<class Scalar, class LO, class GO, class RNG>
 void AdjointNeumannUlamSolver<Scalar,LO,GO,RNG>::buildGhostIterationMatrix()

### 2012-11-05 src/Chimera_MCSA_def.hpp

@@ -166,8 +166,12 @@ void MCSA<Scalar,LO,GO,RNG>::iterate()
 	++(this->b_num_iters);
 
 	// Print iteration data.
-	std::cout << "MCSA Iteration " << this->b_num_iters << ": Residual = "
-		  << residual_norm << std::endl;
+	if ( this->b_linear_problem->getOperator()->getComm()->getRank() == 0 )
+	{
+	    std::cout << "MCSA Iteration " << this->b_num_iters << ": Residual = "
+		      << residual_norm << std::endl;
+	}
+	this->b_linear_problem->getOperator()->getComm()->barrier();
     }
 
     // Check for convergence.

### 2012-11-05 src/Chimera_SequentialMC_def.hpp

@@ -158,8 +158,12 @@ void SequentialMC<Scalar,LO,GO,RNG>::iterate()
 	++(this->b_num_iters);
 
 	// Print iteration data.
-	std::cout << "Sequential MC Iteration " << this->b_num_iters 
-		  << ": Residual = " << residual_norm << std::endl;
+	if ( this->b_linear_problem->getOperator()->getComm()->getRank() == 0 )
+	{
+	    std::cout << "Sequential MC Iteration " << this->b_num_iters 
+		      << ": Residual = " << residual_norm << std::endl;
+	}
+	this->b_linear_problem->getOperator()->getComm()->barrier();
     }
 
     // Check for convergence.

### 2012-11-05 test/tstAdjointNeumannUlamSolver.cpp

@@ -147,7 +147,7 @@ TEUCHOS_UNIT_TEST( AdjointNeumannUlamSolver, adjoint_neumannulam_test )
 	row = i + j*N;
 	if ( row_map->isNodeGlobalElement( row ) )
 	{
-	    B->replaceGlobalValue( row, 5.0 );
+	    B->replaceGlobalValue( row, -5.0 );
 	}
     }
     // right
@@ -157,7 +157,7 @@ TEUCHOS_UNIT_TEST( AdjointNeumannUlamSolver, adjoint_neumannulam_test )
 	row = i + j*N;
 	if ( row_map->isNodeGlobalElement( row ) )
 	{
-	    B->replaceGlobalValue( row, 5.0 );
+	    B->replaceGlobalValue( row, -5.0 );
 	}
     }
     // bottom
@@ -193,7 +193,7 @@ TEUCHOS_UNIT_TEST( AdjointNeumannUlamSolver, adjoint_neumannulam_test )
     // Build the Adjoint solver.
     std::string split_type = "JACOBI";
     double weight_cutoff = 1.0e-4;
-    int histories_per_stage = 10000;
+    int histories_per_stage = 1000;
     Teuchos::RCP<Teuchos::ParameterList> plist =
 	Teuchos::rcp( new Teuchos::ParameterList() );
     plist->set<std::string>("SPLIT TYPE", split_type);

### 2012-11-05 test/tstMCSA.cpp

@@ -2,7 +2,7 @@
 /*!
  * \file tstMCSA.cpp
  * \author Stuart R. Slattery
- * \brief Stationary solver tests.
+ * \brief MCSA tests.
  */
 //---------------------------------------------------------------------------//
 
@@ -192,7 +192,7 @@ TEUCHOS_UNIT_TEST( MCSA, mcsa_test )
     plist->set<std::string>( "SPLIT TYPE",          "JACOBI"  );
     plist->set<std::string>( "MC TYPE",             "ADJOINT" );
     plist->set<double>(      "WEIGHT CUTOFF",       1.0e-4    );
-    plist->set<int>(         "HISTORIES PER STAGE", 100       );
+    plist->set<int>(         "HISTORIES PER STAGE", 1000      );
 
     Teuchos::RCP<Chimera::LinearSolver<double,int,int> > solver =
 	Chimera::LinearSolverFactory::create( plist, linear_problem );

### 2012-11-05 test/tstSequentialMC.cpp

@@ -1,8 +1,8 @@
 //---------------------------------------------------------------------------//
 /*!
- * \file tstAdjointNeumannUlamSolver.cpp
+ * \file tstSequentialMC.cpp
  * \author Stuart R. Slattery
- * \brief Stationary solver tests.
+ * \brief SequentialMC tests.
  */
 //---------------------------------------------------------------------------//
 
@@ -17,12 +17,8 @@
 
 #include <Chimera_LinearSolver.hpp>
 #include <Chimera_LinearProblem.hpp>
-#include <Chimera_LinearOperatorSplit.hpp>
-#include <Chimera_LinearOperatorSplitFactory.hpp>
-#include <Chimera_NeumannUlamSolver.hpp>
-#include <Chimera_AdjointNeumannUlamSolver.hpp>
-#include <Chimera_BoostRNG.hpp>
-#include <Chimera_OperatorTools.hpp>
+#include <Chimera_LinearSolverFactory.hpp>
+#include <Chimera_SequentialMC.hpp>
 
 #include <Teuchos_UnitTestHarness.hpp>
 #include <Teuchos_DefaultComm.hpp>
@@ -42,19 +38,19 @@
 //---------------------------------------------------------------------------//
 // Tests
 //---------------------------------------------------------------------------//
-TEUCHOS_UNIT_TEST( AdjointNeumannUlamSolver, adjoint_neumannulam_test )
+TEUCHOS_UNIT_TEST( SequentialMC, sequential_mc_test )
 {
     // Setup parallel distribution.
     Teuchos::RCP<const Teuchos::Comm<int> > comm = 
 	Teuchos::DefaultComm<int>::getComm();
 
     // Build the linear operator - this is a 2D Transient Diffusion operator.
-    int N = 50;
+    int N = 10;
     int problem_size = N*N;
     double dx = 0.01;
     double dy = 0.01;
-    double dt = 0.5;
-    double alpha = 0.01;
+    double dt = 0.05;
+    double alpha = 0.001;
     Teuchos::RCP<const Tpetra::Map<int> > row_map = 
 	Tpetra::createUniformContigMap<int,int>( problem_size, comm );
     Teuchos::RCP<Tpetra::CrsMatrix<double,int,int> > A = 
@@ -186,47 +182,31 @@ TEUCHOS_UNIT_TEST( AdjointNeumannUlamSolver, adjoint_neumannulam_test )
     Teuchos::RCP<Chimera::LinearProblem<double,int,int> > linear_problem = 
[…]

### 2012-11-20 doc/mandc2013/notes/spectral_theory.tex
adding spectral theory note

--- 2012-11-01 doc/prelim/document/chapters/parallel_mc/parallel_mc.tex
@@
-\chapter{Monte Carlo Solution Methods for Linear Systems}
-\label{ch:stochastic_methods}
-An alternative approach to approximate matrix inversion is to employ
-Monte Carlo methods that sample a distribution with an expectation
-value equivalent to that of the inverted operator. Such methods have
-been in existence for decades with the earliest reference noted here
-an enjoyable manuscript published in 1950 by Forsythe and Leibler
-\citep{forsythe_matrix_1950}. In their outline, Forsythe and Liebler
-in fact credit the creation of this technique to J. Von Neumann and
-S.M. Ulam some years earlier than its publication. In 1952 Wasow
-provided a more formal explanation of Von Neumann and Ulam's method
-\citep{wasow_note_1952} and Hammersley and Handscomb's 1964 monograph
-\citep{hammersley_monte_1964} and Spanier and Gelbard's 1969 book
-\citep{spanier_monte_1969} present additional detail on this topic
-using a collection of references from the 1950's and early 1960's.
+%        File: latexdoc.tex
+%     Created: Friday March 4 12:24:15 2011
+% Last Change: Friday March 4 12:24:20 2011
+%
+\documentclass[letterpaper,12pt]{article}
+\usepackage[top=1.0in,bottom=1.0in,left=1.25in,right=1.25in]{geometry}
+\usepackage{verbatim}
+\usepackage{amssymb}
+\usepackage{graphicx}
+\usepackage{longtable}
+\usepackage{amsfonts}
+\usepackage{amsmath}
+\usepackage{amsthm}
+\usepackage[mathcal]{euscript}
+\usepackage{tabularx}
+\usepackage{cite}
+\usepackage{c++}
+\usepackage{tmadd,tmath}
+\usepackage[usenames]{color}
+\usepackage[
+naturalnames = true, 
+colorlinks = true, 
+linkcolor = black,
+anchorcolor = black,
+citecolor = black,
+menucolor = black,
+urlcolor = blue
+]{hyperref}
+\usepackage{listings}
+\usepackage{textcomp}
+\definecolor{listinggray}{gray}{0.9}
+\definecolor{lbcolor}{rgb}{0.9,0.9,0.9}
+\lstset{
+  backgroundcolor=\color{lbcolor},
+  tabsize=4,
+  rulecolor=,
+  language=c++,
+  basicstyle=\scriptsize,
+  upquote=true,
+  aboveskip={1.5\baselineskip},
+  columns=fixed,
+  showstringspaces=false,
+  extendedchars=true,
+  breaklines=true,
+  prebreak =
+  \raisebox{0ex}[0ex][0ex]{\ensuremath{\hookleftarrow}},
+  frame=single,
+  showtabs=false,
+  showspaces=false,
+  showstringspaces=false,
+  identifierstyle=\ttfamily,
+  keywordstyle=\color[rgb]{0,0,1},
+  commentstyle=\color[rgb]{0.133,0.545,0.133},
+  stringstyle=\color[rgb]{0.627,0.126,0.941},
+}
 
-\section{Preliminaries}
-\label{sec:mc_preliminaries}
-We begin our discussion of Monte Carlo methods using these texts by
-seeking a solution to Eq~(\ref{eq:linear_problem}). For a given linear
-operator $\ve{A}$, we can use diagonal splitting in a similar manner
-as the stationary method in Eq~(\ref{eq:linear_split_equation2}) to
-define the following operator\footnote{It should be noted that
-  non-diagonal splittings have been recently explored in
-  \citep{srinivasan_monte_2010} and have the potential to improve
-  efficiency.}:
+%%---------------------------------------------------------------------------%%
+\author{Stuart R. Slattery
+\\ \href{mailto:sslattery@wisc.edu}{\texttt{sslattery@wisc.edu}}
+}
+
+\date{\today}
+\title{A Spectral Theory for the Domain-Decomposed Adjoint
+  Neumann-Ulam Method}
+\begin{document}
+\maketitle
+
+%%---------------------------------------------------------------------------%%
+\begin{abstract}
+
+  The serial and parallel behavior of the adjoint Neumann-Ulam method
+  for solving linear operator equations is analyzed using the spectral
+  properties of the linear operator. Relationships for the average
+  length of the Markov chains and the fraction of histories leaking
+  from a parallel domain are made with respect to the eigenvalues of
+  the linear operator. The one-speed, two-dimensional neutron
+  diffusion equation is used as a model problem to test the spectral
+  theory for symmetric operators. These relationships will serve as
+  guidelines for selecting an appropriate parallel decomposition
+  stratagy and provide a basis for performance models for future
+  implementations of a parallel adjoint Neumann-Ulam method.
+
+\end{abstract}
+
+%%---------------------------------------------------------------------------%%
+\section{The Adjoint Neumann-Ulam Method}
+We seek solutions of the general linear problem in the following form:
 \begin{equation}
-  \ve{H} = \ve{I} - \ve{A}\:,
-  \label{eq:linear_mc_iteration_matrix}
+  \ve{A} \ve{x} = \ve{b}\:,
+  \label{eq:linear_problem}
 \end{equation}
-such that we are solving the system:
-\begin{equation}
-  \ve{x} = \ve{H} \ve{x} + \ve{b}\:.
-  \label{eq:richardson_split}
-\end{equation}
-We can then form an alternative representation for $\ve{A}^{-1}$ by
-generating the \textit{Neumann series}:
-\begin{equation}
-  \ve{A}^{-1} = (\ve{I}-\ve{H})^{-1} = \sum_{k=0}^{\infty} \ve{H}^k\:,
-  \label{eq:neumann_series}
-\end{equation}
-which will converge if the spectral radius of $\ve{H}$ is less than
-1. If we then apply this Neumann series to the right hand side of
-Eq~(\ref{eq:linear_problem}) we acquire the solution to the linear
-problem:
-\begin{equation}
-  \ve{A}^{-1}\ve{b} = \sum_{k=0}^{\infty} \ve{H}^k\ve{b} = \ve{x}\:.
-  \label{eq:neumann_solution}
-\end{equation}
-An approximation of this summation by truncation will therefore lead
-to an approximation of the solution. If we expand the summation with a
-succession of matrix-vector multiply operations, we arrive at an
-alternative perspective of this summation by considering its $i^{th}$
-component:
-\begin{equation}
-  x_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-  \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}\:,
-  \label{eq:expanded_neumann_solution}
-\end{equation}
-which can interpreted as a series of transitions between states,
-\begin{equation}
- \nu = i \rightarrow i_1 \rightarrow \cdots \rightarrow i_{k-1}
- \rightarrow i_{k}\:,
-  \label{eq:mc_walk_permutation}
-\end{equation}
-in $\ve{H}$ where $\nu$ is interpreted as a particular random walk
-sequence permutation. We can generate these sequences of transitions
-through Monte Carlo random walks by assigning them both a probability
-and weight. As a reinterpretation of the iteration matrix, we then
-form the \textit{Neumann-Ulam decomposition} of \ve{H}:
-\begin{equation}
-  \ve{H} = \ve{P} \circ \ve{W}\:,
-  \label{eq:neumann_ulam_decomposition}
-\end{equation}
-where $\circ$ denotes the Hadamard product operation\footnote{The
-  Hadamard product $\ve{A} = \ve{B} \circ \ve{C}$ is defined
-  element-wise as $a_{ij} = b_{ij} c_{ij}$.}, $\ve{P}$ denotes the
-transition probability matrix, and $\ve{W}$ denotes the transition
-weight matrix. This decomposition, a generalization of Dimov's work
-\citep{dimov_new_1998}, is an extension of the original Neumann-Ulam
-scheme in that now a weight cutoff can be used to terminate a random
-walk sequence and therefore truncate the Neumann series it is
-approximating. The formulation of $\ve{P}$ and $\ve{W}$ will be
-dependent on whether we choose a direct or adjoint Monte Carlo
-sequence to estimate the state transitions in
-Eq~(\ref{eq:expanded_neumann_solution}).
-
-\section{Direct Method}
-\label{sec:direct_mc}
-In the context of matrix inversion, a direct method resembles an
-adjoint Monte Carlo method in the reactor physics community where the
-solution state is sampled and the source terms that contribute to it
-are assembled. To achieve this, we build the direct method
-Neumann-Ulam decomposition per Dimov's approach by first choosing a
-probability matrix that is a row scaling of $\ve{H}$ such that its
-components are:
-\begin{equation}
-  p_{ij} = \frac{|h_{ij}|}{\sum_j |h_{ij}|}\:.
-  \label{eq:direct_probability}
-\end{equation}
-From this, we then see that the probability of transitioning from a
-state $i$ to a state $j$ is implicitly linked to the original operator
-$\ve{A}$ in that those terms with large values, and therefore those
-that make the greatest contribution to the numerical solution, will be
-sampled with a higher probability than smaller terms. In addition, the
-row scaling provides a normalization over the state to which we are
-transitioning such that $\sum_j p_{ij} = 1$, meaning that we sample
-the probabilities over the rows of the matrix. The components of
-the weight matrix are then defined by
-Eq~(\ref{eq:neumann_ulam_decomposition}) as:
-\begin{equation}
-  w_{ij} = \frac{h_{ij}}{p_{ij}}\:.
-  \label{eq:direct_weight}
-\end{equation}
-It should be noted here that if $\ve{A}$ is sparse, then $\ve{H}$,
-$\ve{P}$, and $\ve{W}$ must be sparse as well by
-definition. Additionally, we only compute $\ve{P}$ and $\ve{W}$ from
-the non-zero elements of $\ve{H}$ as those components that are zero
-will not participate in the random walk. Doing so prevents an
-infinite weight from being generated in Eq~(\ref{eq:direct_weight})
-and eliminates unnecessary computations.
-
-Using these matrices, we can then form the expectation value of the
-direct solution. For a given random walk permutation $\nu$ with $k$
-events, we define the weight of that permutation to be:
-\begin{equation}
-  W_{m} = w_{i,i_1} w_{i_1,i_2} \cdots w_{i_{m-1},i_m}\:,
-  \label{eq:direct_permutation_weight}
-\end{equation}
-such that the weight of each transition event contributes to the
-total. The contribution to the solution from a particular random walk
-permutation is then:
-\begin{equation}
-  X_{\nu}(i_0 = i) = \sum_{m=0}^k W_{m} b_{i_m}\:,
-  \label{eq:direct_permutation_contribution}
-\end{equation}
-where $X_{\nu}(i_0 = i)$ signifies that the solution state in which we
-are tallying defines state $i$.  We can interpret this precisely as
-before in that during the random walk we collect the source in the
-states that are visited and apply them to the solution tally. We then
-define the probability that a particular random walk permutation of
-$k$ events will occur:
-\begin{equation}
-  P_{\nu} = p_{i,i_1} p_{i_1,i_2} \cdots p_{i_{k-1},i_k}\:.
-  \label{eq:direct_permutation_probability}
-\end{equation}
-Finally, we define the expectation value of $X$ to be the collection
-of all random walk permutations and their probabilities:
-\begin{equation}
-  E\{X(i_0 = i)\} = \sum_{\nu} P_{\nu} X_{\nu}\:,
-  \label{eq:direct_expectation_value}
-\end{equation}
-which, if expanded, directly recovers the exact solution:
-\begin{equation}
-  \begin{split}
-    E\{X(i_0 = i)\}
-    &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-    \sum_{i_k}^{N} p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
-    w_{i,i_1}w_{i_1,i_2}\ldots w_{i_{k-1},i_k} b_{i_k}\\ &= x_i\:,
-  \end{split}
-  \label{eq:direct_expectation_expansion}
-\end{equation}
-therefore providing an unbiased Monte Carlo estimator.
-
-In cases where we seek only approximate solutions, we need only to
-perform a minimal number of random walks in order to generate an
-approximation for $\ve{x}$. If we are only to approximate the
-solution, we also need conditions by which we may terminate a random
-walk. We do this by noticing that the factors added to
-Eq~(\ref{eq:direct_permutation_weight}) will become diminishingly
-small due to their definition in Eq~(\ref{eq:direct_weight}) and
-therefore their contributions to the solution estimate will become
-negligible. Using this, we choose terminate a random walk sequence
-with a \textit{weight cutoff}, $W_c$, that is enforced when $W_m <
-W_c$ for a particular random walk permutation.
-
-\subsection{Estimator Variance}
-\label{subsec:estimator_variance}
-We can compute the variance of the estimator through traditional
-methods by defining the variance, $\sigma_i$, for each component in
-the solution:
-\begin{equation}
-  {\sigma_i}^2 = E\{X(i_0 = i) - (\ve{A}^{-1}\ve{b})_i\}^2 = E\{X(i_0
-  = i)^2\} - x_i^2\:,
-  \label{eq:direct_variance_1}
-\end{equation}
-where the vector exponentials are computed element-wise. Inserting
-Eq~(\ref{eq:direct_expectation_value}) gives:
-\begin{equation}
-  \sigma_i^2 = \Big(\sum_{\nu} P_{\nu} X_{\nu}^2\Big)_i - x_i^2\:,
-  \label{eq:direct_variance_2}
-\end{equation}
-and applying Eq~(\ref{eq:direct_permutation_contribution}):
-\begin{equation}
-  \sigma_i^2 = \Big(\sum_{\nu} P_{\nu} \sum_{m=0}^k W_{m}^2
-  b_{i_m}^2\Big)_i - x_i^2\:.
-  \label{eq:direct_variance_3}
-\end{equation}
-Finally, expanding the transition probabilities yields the variance:
-\begin{equation}
-  \sigma_i^2 = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-  \sum_{i_k}^{N} p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
-  w^2_{i,i_1}w^2_{i_1,i_2}\ldots w^2_{i_{k-1},i_k} b_{i_m} - x_i^2\:.
-  \label{eq:direct_variance_4}
-\end{equation}
-Using this definition for the variance, we can arrive at a more natural
-reason for enforcing $\rho(\ve{H}) < 1$ for our Monte Carlo method to
-converge. Per the Hadamard product, we can concatenate the summation
-in Eq~(\ref{eq:direct_variance_4}):
-\begin{equation}
-  (\ve{P} \circ \ve{W} \circ \ve{W})^k =
-  \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots \sum_{i_k}^{N}
-  p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
-  w^2_{i,i_1}w^2_{i_1,i_2}\ldots w^2_{i_{k-1},i_k}\:.
-  \label{eq:double_weighted_decomposition}
-\end{equation}
-If we assign $\ve{G} = \ve{P} \circ \ve{W} \circ \ve{W}$ as in
-Eq~(\ref{eq:neumann_ulam_decomposition}), we then have a new
-formulation for the variance:
-\begin{equation}
-  \sigma^2_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-  \sum_{i_k}^{N}g_{i,i_1}g_{i_1,i_2}\ldots g_{i_{k-1},i_k} b_{i_k}^2 -
-  x_i^2\:,
-\end{equation}
-which contains the general Neumann series for $\ve{G}$,
-\begin{equation}
-  \ve{T} = \sum_{k=0}^{\infty} \ve{G}^k\:,
-  \label{eq:variance_neumann_series}
-\end{equation}
-where $\ve{T} = (\ve{I}-\ve{G})^{-1}$. We can then insert $T$ back
-into the variance formulation for a more concise definition:
-\begin{equation}
-  \sigma^2_i = (\ve{T}\ve{b})_i - x_i^2\:.
-  \label{eq:direct_variance_5}
-\end{equation}
-We can relate $\ve{G}$ to $\ve{H}$ by noting that $\ve{G}$ simply
-contains an additional Hadamard product with the weight matrix. The
-Hadamard product has the property that:
-\begin{equation}
-  |\ve{H} \circ \ve{W}| \geq |\ve{H}|\ |\ve{W}|\:.
-  \label{eq:hadamard_inequality}
-\end{equation}
-Using the norm property of the Hadamard product and
-Eq~(\ref{eq:neumann_ulam_decomposition}), we can define the norm of
-$\ve{W}$ as:
-\begin{equation}
-  \frac{|\ve{H}|}{|\ve{P}|} \geq |\ve{W}|\:.
-  \label{eq:weight_norm}
-\end{equation}
-Choosing the infinity norm of the operator as defined in
-Eq~(\ref{eq:matrix_infinity_norm}), the row normalized probability
-matrix will yield a norm of 1 giving the following inequality for
-relating $\ve{G}$ and $\ve{H}$:
-\begin{equation}
-  |\ve{G}| \geq |\ve{H}|^2
-  \label{eq:variance_norm_inequality}
-\end{equation}
-Using these relations to analyze Eq~(\ref{eq:direct_variance_5}), we
-see that if $\rho(\ve{G}) > 1$, then the sum in
-Eq~(\ref{eq:variance_neumann_series}) will not converge and an
-infinite variance will arise as the elements of $\ve{T}$ become
-infinite in Eq~(\ref{eq:direct_variance_5}). We must restrict $\ve{G}$
-to alleviate this and therefore restrict $\ve{H}$ due to
-Eq~(\ref{eq:variance_norm_inequality}) with $\rho(\ve{H}) < 1$ so that
-our expectation values for the solution may have a finite variance.
-
-\subsection{Reusing Random Walks}
-\label{subsec:reusing_random_walks}
-A downfall of the direct method is that a random walk must be
-initiated in each state of the solution vector in which the solution
-is desired due to the estimator definition in
-Eq~(\ref{eq:direct_expectation_expansion}). Because of this, a
-solution estimate in all states of a linear system may require a
-large number of random walks to be performed. Ji and Li recently
-proposed a variation of the direct method in which the random walk
-permutations in Eq~(\ref{eq:mc_walk_permutation}) can be used to
-contribute to the solution tally in all visited states instead of the
-only starting state \citep{ji_reusing_2012}. By doing this, they
-effectively reduce the number of random walks needed to sample all
-desired states in the system.
-
-Their method is based on the observation that while a random walk
-transitions between states in the system, if every state it visited
-could be considered that starting state, then a series a contributions
-to the solution could be generated by a single sequence of random
-numbers while maintaining statistical independence between the
-samples. To demonstrate this, consider the following random walk
-sequence starting in state 2: $2 \rightarrow 3 \rightarrow 1
-\rightarrow 2 \rightarrow 4 \rightarrow 5$. We can reuse these walks
-be recognizing that we also then have a random walk sequence starting
-in state 3 that can be applied to the state 3 estimator: $3
-\rightarrow 1 \rightarrow 2 \rightarrow 4 \rightarrow 5$. We also have
-the same situation for state 1: $1 \rightarrow 2 \rightarrow 4
-\rightarrow 5$, and all future states. Using this sampling scheme , Ji
-and Li observed 1 to 2 orders of magnitude reduction over the standard
-direct method in the number histories required to achieve a specified
-variance in their test problems.
-
-\section{Adjoint Method}
-\label{sec:adjoint_mc}
-An alternative formulation for Monte Carlo matrix inversion is the
-adjoint method. We begin by defining the linear system adjoint to
-Eq~(\ref{eq:linear_problem}):
+where $\ve{A} \in \mathbb{R}^{N \times N}$ is a matrix operator such
+that $\ve{A} : \mathbb{R}^{N} \rightarrow \mathbb{R}^{N}$, $\ve{x} \in
+\mathbb{R}^N$ is the solution vector, and $\ve{b} \in \mathbb{R}^N$ is
+the forcing term. We choose the adjoint method Monte Carlo method to
+invert the linear operator. We begin by defining the linear system
+adjoint to Eq~(\ref{eq:linear_problem}):
 \begin{equation}
@@
 respectively and $\ve{A}^T$ is the adjoint operator. We can split this
-equation to mirror Eq~(\ref{eq:richardson_split}) by defining the
-following inner product equivalence \citep{spanier_monte_1969}:
+equation by defining the following inner product equivalence
+\cite{spanier_monte_1969}:
 \begin{equation}
@@
 \end{equation}
-As was required for convergence with the direct method using
-Eq~(\ref{eq:richardson_split}), the spectral radius of $\ve{H}$ must
+For convergence, we require that the spectral radius of $\ve{H}$ must
 remain less than 1 as $\ve{H}^T$ contains the same eigenvalues and
@@
 where now we are forming the decomposition with respect to the
-transpose of $\ve{H}$\footnote{This is sometimes referred to as the
-  adjoint form for real-valued matrices}. We then follow the same
-procedure as the direct method for forming the probability and weight
-matrices in the decomposition. Using the adjoint form, probabilities
-should instead be column-scaled:
+transpose of $\ve{H}$. We then follow the same procedure as the direct
+method for forming the probability and weight matrices in the
+decomposition. Using the adjoint form, probabilities should instead be
+column-scaled:
 \begin{equation}
@@
 Using the decomposition we can then define an expectation value for
-the adjoint method. Given Eq~(\ref{eq:direct_permutation_weight}) as
-the weight generated for a particular random walk permutation as in
-Eq~(\ref{eq:mc_walk_permutation}) and our result from
+the adjoint method. Using our result from
 Eq~(\ref{eq:inner_product_constraint}) generated by applying the
@@
 
-\section{Sequential Monte Carlo}
-\label{sec:sequential_mc}
-The direct and adjoint Neumann-Ulam methods described are limited by a
-convergence rate of $1/\sqrt{N}$ by the Central Limit Theorem where
-$N$ is the number of random walk permutations. In 1962, Halton
-presented a residual Monte Carlo method that moves towards exponential
-convergence rates \citep{halton_sequential_1962} and further refined
-his work some years later \citep{halton_sequential_1994}. Applications
-of his work by the transport community have confirmed convergence
-rates on the order of of $exp(-N)$ \citep{evans_residual_2003}. In
-much the same way as projection methods, Halton's method, sequential
-Monte Carlo, utilizes the adjoint Monte Carlo solver as a means of
-directly reducing the residual vector. He proposed the following
-iterative scheme as a solution to Eq~(\ref{eq:linear_problem})\:
-\begin{subequations}
-  \begin{gather}
-    \ve{r}^k = \ve{b} - \ve{A}\ve{x}^k\:,\\  
-    \ve{A}\boldsymbol{\delta}^{k} = \ve{r}^{k}\:,\\
-    \ve{x}^{k+1} = \ve{x}^k + \boldsymbol{\delta}^{k}\:,
-  \end{gather}
-  \label{eq:sequential_monte_carlo}
-\end{subequations}
-where the correction $\boldsymbol{\delta}$ is computed by the adjoint
-Monte Carlo method. The merits of Halton's approach are immediately
-visible in that we have now broken the binding of the convergence rate
-to the Central Limit Theorem. Here, the Monte Carlo solver is used to
-produce a correction from the residual, analogous to using the
-residual to extract a correction from the search subspace in a
-projection method. By doing this, the Monte Carlo error is bound in
-the correction used to update the solution and therefore does not
-explicitly manifest itself in the overall convergence of the
-solution. The downside of such a method is that if the solution guess
-is poor, then many iterations are required in order to reach
-exponential converge as the Monte Carlo error (and therefore the
-Central Limit Theorem) does dominate in this situation. Therefore, if
-a good initial guess is not available, Halton's method is still
-characterized by poor performance.
+%%---------------------------------------------------------------------------%%
+\section{Model Problem}
+For our numerical experiments, we choose the one-speed,
+two-dimensional neutron diffusion equation as the model problem
+\cite{duderstadt_nuclear_1976}: 
+\begin{equation}
+  -\boldsymbol{\nabla} \cdot D \boldsymbol{\nabla} \phi + \Sigma_a
+  \phi = S\:,
+  \label{eq:diffusion_eq}
+\end{equation}
+where $\phi$ is the neutron flux, $\Sigma_a$ is the absorption cross
+section, $S$ is the source of neutrons and $D$ is the diffusion
+coefficient defined as:
+\begin{equation}
+  D = \frac{1}{3 ( \Sigma_t - \bar{\mu}\Sigma_s )}\:,
+  \label{eq:diffusion_coeff}
+\end{equation}
+where $\Sigma_s$ is the scattering cross section, $\Sigma_t = \Sigma_a
++ \Sigma_s$ is the total cross section, and $\bar{\mu}$ is the cosine
+of the average scattering angle. For simplification, we will take
+$\bar{\mu} = 0$ for our analysis giving $D=(3 \Sigma_t)^{-1}$. In
+addition, to further simplify we will assume a homogenous domain such
+that the cross sections remain constant throughout. Doing this permits
+us to simplify Eq~(\ref{eq:diffusion_eq}):
+\begin{equation}
+  -D \boldsymbol{\nabla}^2 \phi + \Sigma_a \phi = S\:.
+  \label{eq:diffusion_eq_simple}
+\end{equation}
 
-\section{Monte Carlo Synthetic-Acceleration}
-\label{sec:mcsa}
-Using the ideas of Halton, Evans and Mosher recently developed a Monte
-Carlo solution method that was not prohibited severely by the quality
-of the initial guess for the system \citep{evans_monte_2009} and later
-applied it more rigorously as a solution mechanism for the radiation
-diffusion equation \citep{evans_monte_2012}. With their new methods,
-they achieved identical numerical results as and marginally better
-performance than conventional Krylov solvers. Their approach was
-instead to use residual Monte Carlo as a synthetic acceleration for a
-stationary method. To derive this method, we begin by splitting the
-operator in Eq~(\ref{eq:linear_problem})
+We choose a finite difference scheme on a square Cartesian grid to
+discretize the problem. For the Laplacian, we choose the 9-point
+stencil shown in Figure~\ref{fig:stencil} \cite{leveque_finite_2007}:
+\begin{multline}
+  \nabla^2_9\phi = \frac{1}{6\Delta^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j}
+    + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ +
+    \phi_{i-1,j+1} + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20
+    \phi_{i,j}]\:.
+  \label{eq:nine_point_stencil}
+\end{multline}
+
+We then have the following linear system to solve:
+\begin{multline}
+  -\frac{1}{6\Delta^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j}
+    + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ +
+    \phi_{i-1,j+1} + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20
+    \phi_{i,j}] + \Sigma_a \phi_{i,j} = s_{i,j}\:,
+  \label{eq:fd_system}
+\end{multline}
+and in operator form:
 \begin{equation}
-  \ve{x} = (\ve{I} - \ve{A})\ve{x} + \ve{b}\:.
-  \label{eq:linear_split}
+  \ve{D}\boldsymbol{\phi}=\ve{S}\:,
+  \label{eq:operator_system}
 \end{equation}
-With this we can then define the stationary method
-\textit{Richardson's iteration} as:
-\begin{equation}
-  \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,
-  \label{eq:richardsons_iteration}
-\end{equation}
-which will converge if $\rho(\ve{I} - \ve{A}) < 1$. We then define the
-solution error at the $k^{th}$ iterate relative to the true solution:
-\begin{equation}
-  \delta \ve{x}^k = \ve{x} - \ve{x}^k\:.
-  \label{eq:mcsa_error}
-\end{equation}
-Subtracting Eq~(\ref{eq:richardsons_iteration}) from
-Eq~(\ref{eq:linear_split}) we get:
-\begin{equation}
-  \delta \ve{x}^{k+1} = (\ve{I} - \ve{A})\delta \ve{x}^k\:.
-  \label{eq:mcsa_setup_1}
-\end{equation}
-Subtracting from this $(\ve{I} - \ve{A})\delta \ve{x}^{k+1}$ yields:
-\begin{equation}
-  \begin{split}
-    \ve{A}\delta \ve{x}^{k+1} &= (\ve{I} -
-    \ve{A})(\ve{x}^{k+1}-\ve{x}^{k}) \\ &= \ve{r}^{k+1}\:.
-    \label{eq:mcsa_setup_2}
-  \end{split}
-\end{equation}
-Using this, we define the following scheme that will converge in one
-iteration if $\ve{A}$ is inverted exactly.
-\begin{subequations}
-  \begin{gather}
-    \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,\\
-    \ve{A} \delta \ve{x}^{k+1} = \ve{r}^{k+1}\:,\\
-    \ve{x} = \ve{x}^{k+1} + \delta \ve{x}^{k+1}\:.
-  \end{gather}
-  \label{eq:mcsa_setup_3}
-\end{subequations}
-However, $\ve{A}$ is only approximately inverted by our numerical
-methods and therefore we instead pose an iterative scheme in which the
-Monte Carlo solvers are used to invert the operator. The
-\textit{Fixed-Point Monte Carlo Synthetic-Acceleration} (MCSA) method
-is defined as:
-\begin{subequations}
-  \begin{gather}
-    \ve{x}^{k+1/2} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,\\
-    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\
-    \ve{A}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:,\\
-    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,
-  \end{gather}
-  \label{eq:mcsa}
-\end{subequations}
-where the adjoint Monte Carlo method is used to generate the solution
-correction from the residual. Using Monte Carlo in this way achieves
-the same effect as Halton's method, decoupling its convergence rate
-from the overall convergence rate of the method. Here, the approximate
-Monte Carlo solution is not driven to a particular convergence as it
-merely supplies a correction for the initial guess generated by
-Richardson's iteration. Rather, only a set number of histories are
-required using the adjoint method to generate the correction.
+where $\ve{D}$ is the diffusion operator, $\ve{S}$ is the source in
+vector form and $\boldsymbol{\phi}$ is the vector of unkown fluxes.
 
-In addition to the Monte Carlo solver parameters dictating the number
-of histories and weight cutoff, the outer MCSA iterations also have
-the following stopping criteria:
-\begin{equation}
-  ||\ve{r}||_\infty < \epsilon \ ||\ve{b}||_\infty\:,
-  \label{eq:mcsa_stopping_criteria}
-\end{equation}
-where $\epsilon$ is a user-defined parameter. We therefore have 3
-parameters to tune in an MCSA implementation: the number of Monte
-Carlo histories computed in the adjoint solve during each MCSA
-iteration, the weight cutoff for those histories, and the total MCSA
-convergence tolerance as specified by $\epsilon$.
+To close the system a non-reentrant current condition is applied
+to all boundaries of the domain.
 
-\subsection{Preconditioning MCSA}
-\label{sec:stochastic_preconditioning}
-In most cases, at least a minimal amount of \textit{preconditioning}
-of the linear system will be required in order to use the class of
-stochastic methods described. Although these methods have no symmetry
-requirements for convergence, they do require that the spectral radius
-of the iteration matrix be less than one. To achieve this, a Jacobi
-preconditioner, a form of left preconditioning, is used such that the
-preconditioning matrix $\ve{M}$ is:
-\begin{equation}
-  \ve{M} = diag(\ve{A})\:,
-  \label{eq:jacobi_preconditioner}
-\end{equation}
-such that its application means we are instead solving the following
-linear system:
-\begin{equation}
-  \ve{M}^{-1}\ve{A}\ve{x} = \ve{M}^{-1}\ve{b}\:.
-  \label{eq:jacobi_precond_linear_problem}
-\end{equation}
-Next, we can apply MCSA to solve
-Eq~(\ref{eq:jacobi_precond_linear_problem}): 
-\begin{subequations}
-  \begin{gather}
-    \ve{x}^{k+1/2} = (\ve{I} - \ve{M}^{-1}\ve{A})\ve{x}^k +
-    \ve{M}^{-1}\ve{b}\:,\\ \ve{r}^{k+1/2} = \ve{M}^{-1}\ve{b} -
-    \ve{M}^{-1}\ve{A}\ve{x}^{k+1/2}\:,\\ \ve{M}^{-1}\ve{A}\delta\ve{x}^{k+1/2}
-    = \ve{r}^{k+1/2}\:,\\ \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta
-    \ve{x}^{k+1/2}\:.
-    \label{eq:jacobi_preconditioned_mcsa}
-  \end{gather}
-\end{subequations}
-Choosing Jacobi preconditioning with MCSA is advantageous for several
-reasons. First, $\rho(\ve{I} - \ve{M}^{-1}\ve{A}) < 1$ is true for all
-$\ve{A}$ and is easy to formulate because the inversion of $\ve{M}$ is
-trivial. Second, because the adjoint Monte Carlo method used within
-MCSA to compute the correction operates on a linear problem with the
-preconditioned operator, then $\ve{H}$ in the adjoint solver will have
-a zero term in each of its diagonal elements, thereby eliminating all
-in-state transitions during the random walk sequence. Because of this,
-Jacobi preconditioning should always be performed, regardless of any
-other preconditioning that is applied to the system.
+%%---------------------------------------------------------------------------%%
+\section{Spectral Analysis}
 
-\section{Parallelization of Stochastic Methods}
-\label{sec:parallel_stochastic_methods}
-For Monte Carlo methods, and in particular MCSA, to be viable at the
-production scale, scalable parallel implementations are required. In a
-general linear algebra context for discrete systems, this has not yet
-been achieved. Therefore, we will develop and implement parallel
-algorithms for these methods leveraging both the knowledge gained from
-the general parallel implementations of Krylov methods and modern
-parallel strategies for Monte Carlo as developed by the reactor
-physics community. In order to formulate a parallel MCSA algorithm, we
-recognize that the algorithm occurs in two stages, an outer iteration
-performing Richardson's iteration and applying the correction, and an
-inner Monte Carlo solver that is generating the correction via the
-adjoint method. The parallel aspects of both these components must be
-considered.
+%%---------------------------------------------------------------------------%%
+\section{Length of Random Walks}
 
-\subsection{Domain Decomposition for Monte Carlo}
-\label{subsec:msod}
-As observed in the discussion on parallel Krylov methods, large-scale
-problems will surely have their data partitioned such that each
-parallel process owns a subset of the equations in the linear
-system. Given this convention, the adjoint Monte Carlo algorithm must
-perform random walks over a domain that is decomposed and must remain
-decomposed due to memory limitations. This naturally leads us to seek
-parallel algorithms that handle domain decomposition.
+%%---------------------------------------------------------------------------%%
+\section{Domain Leakage}
 
-In the context of radiation transport, Brunner and colleagues provided
-a survey of algorithms for achieving this as implemented in production
-implicit Monte Carlo codes \citep{brunner_comparison_2006}. In their
-work they identify two data sets that are required to be communicated:
-the sharing of particles that are transported from one domain to
-another and therefore from one processor to another and a global
-communication that signals if particle transport has been completed on
-all processors. The algorithms presented are a fully-locking
-synchronous scheme, an asynchronous-send/synchronous-receive pattern,
-a traditional master/slave scheme, and a modified master/slave scheme
-the implements a binary tree pattern for the global reduction type
-operations needed to communicate between the master and slave
-processes. They observed that the modified master/slave scheme
-performed best in that global communications were implemented more
-efficiently than those required by the asynchronous
-scheme. Furthermore, none of these schemes handled load-imbalanced
-cases efficiently. Such cases will be common if the source sampled in
-the Monte Carlo random walk is not isotropic and not evenly
-distributed throughout the global domain. It was noted that
-efficiencies were improved by increasing the frequency by which
-particle data was communicated between domain-adjacent
-processors. However, this ultimately increases communication costs. In
-2009, Brunner extended his work by using a more load-balanced approach
-with a fully asynchronous communication pattern
-\citep{brunner_efficient_2009}. Although the extended implementation
-was more robust and allowed for scaling to larger numbers of
-processors, performance issues were still noted with parallel
-efficiency improvements needed in both the weak and strong scaling
-cases for unbalanced problems. These results led Brunner to conclude
-that a combination of domain decomposition and domain replication
-could be used to solve some of these issues.
+%%---------------------------------------------------------------------------%%
+\section{Conclusion}
 
-\subsubsection{Multiple-Set Overlapping-Domain Decomposition}
-\label{subsubsec:msod}
-In 2010, Wagner and colleagues developed the \textit{multiple-set
-  overlapping-domain} (MSOD) decomposition for parallel Monte Carlo
-applications for full-core light water reactor analysis
-\citep{wagner_hybrid_2010}. In their work, an extension of Brunner's,
-their scheme employed the similar parallel algorithms for particle
-transport but a certain amount of overlap between adjacent domains was
-used to decrease the number of particles leaving the local domain. In
-addition, Wagner utilized a level of replication of the domain such
-that the domain was only decomposed on $O(100)$ processors and if
-replicated $O(1,000)$ times achieves simulation on $O(100,000)$
-processors, thus providing spatial and particle parallelism. Each
-collection of processors that constitutes a representation of the
-entire domain is referred to as a set, and within a set overlap occurs
-among its sub-domains. The original motivation was to decompose the
-domain in a way that it remained in a physical cabinet in a large
-distributed machine, thus reducing latency costs during
-communication. A multiple set scheme is also motivated by the fact
-that communication during particle transport only occurs within a set,
-limiting communications during the transport procedure to a group of
-$O(100)$ processors, a number that was shown to have excellent
-parallel efficiencies in Brunner's work and therefore will scale well
-in this algorithm. The overlapping domains within each set also
-demonstrated reduced communication costs. On each processor, the
-source is sampled in the local domain that would exist if no overlap
-was used while tallies can be made over the entire overlapping domain.
-
-To demonstrate this, consider the example adapted from Mervin's work
-with Wagner and others in the same area \citep{mervin_variance_2012}
-and presented in Figure~\ref{fig:msod_example}.
-\begin{figure}[htpb!]
-  \begin{center}
-    \scalebox{1.5}{
-      \input{chapters/parallel_mc/msod_example.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Overlapping domain example illustrating how domain
-      overlap can reduce communication costs.}
-    \textit{All particles start in the blue region of interest. The
-      dashed line represents 0.5 domain overlap between domains.}}
-  \label{fig:msod_example}
-\end{figure}
-In this example, 3 particle histories are presented emanating from the
-blue region of interest. Starting with particle A, if no domain
-overlap is used then the only the blue domain exists on the starting
-processor. Particle A is then transported through 3 other domains
-before the history ends, therefore requiring three communications to
-occur in Brunner's algorithm. If a 0.5 domain overlap is permitted as
-shown by the dashed line, then the starting process owns enough of the
-domain such that no communications must occur in order to complete the
-particle A transport process. Using 0.5 domain overlap also easily
-eliminates cases such as the represented by the path of particle C. In
-this case, particle C is scattering between two adjacent domains,
-incurring a large latency cost for a single particle. Finally, with
-particle B we observe that 0.5 domain overlap will still not eliminate
-all communications. However, if 1 domain overlap were used, the entire
-geometry shown in Figure~\ref{fig:msod_example} would be contained on
-the source processor and therefore transport of all 3 particles
-without communication would occur.
-
-Wagner and colleagues used this methodology for a 2-dimensional
-calculation of a pressurized water reactor core and varied the domain
-overlap from 0 to 3 domain overlap (a $7 \times 7$ box in the context
-of our example) where a domain constituted a fuel assembly. For the
-fully domain decomposed case, they observed that 76.12\% of all source
-particles leave the domain. At 1.5 domain overlap, the percentage of
-source particles born in the center assembly leaving the processor
-domain dropped to 1.05\% and even further for 0.02\% for the 3 domain
-overlap. Based on these results, this overlap approach, coupled with
-the multiple sets paradigm that will scale for existing parallel
-transport algorithms, provides a scalable Monte Carlo algorithm for
-today's modern machines.
-
-\subsubsection{Domain-to-Domain Communication}
-\label{subsubsec:mc_comm_costs}
-With Wagner's work, we observed that domain overlap implemented with a
-domain decomposed Monte Carlo setting reduces the amount of nearest
-neighbor communication that must occur during transport. This
-communication, expected to be expensive, was analyzed by Siegel and
-colleagues for idealized, load balanced situations without domain
-overlap \citep{siegel_analysis_2012}. They arrive at a few empirical
-relationships and generate a numerical experiment that provides
-insight into the feasibility of large scale domain decomposed Monte
-Carlo as well as the relationship of communication to system
-parameters. In their work, they define a simple 3-dimensional system
-subdivded into Cartesian block domains with each processor containing
-a single domain. Given this geometry, a particle can either be
-absorbed in its current domain or be transported to one of the 6
-adjacent domains owned by another process.
-
-To quantify the number of particles that leak out of the local domain
-they define a leakage fraction, $\lambda$, as:
-\begin{equation}
-  \lambda =
-  \frac{average\ \#\ of\ particles\ leaving\ local\ domain}{total\ 
-    of\ \#\ of\ particles\ starting\ in\ local\ domain}\:.
-  \label{eq:leakage_fraction}
-\end{equation}
-For their studies, they assumed that the value of $\lambda$ was
-dependent on the total cross section of the system and the same for
-all processes for simplicity and load balancing. Using this
-relationship and their simple geometry, they were able to relate the
-amount of compute time required to transport all source particles to
-the number of source particles, the leakage fraction, the latency and
-bandwidth of communication, the ratio of local to global geometry
-size, and the total cross section of the system.
-
-What they found was that there was indeed an optimum ratio of local
-domain to global domain size that results in a minimum total compute
-time. They also found that this ratio was only dependent on the
-latency and bandwidth constraints of the parallel machine. We expect
-this because nearest neighbor communication patterns scale with the
-volume and surface sizes of the local domains. The larger the volume,
-the more compute time is spent within the local volume and therefore
-at some point you are limited by the latency of the system, or the
-time it takes to receive the data to operate on. However, at a certain
-point, when the surface of those volumes becomes large enough, the
-amount of data that must be communicated grows large enough that it
-may overcome the bandwidth limitations of the system, resulting in a
-decrease in performance. In their numerical experiments, they used the
-above geometry and leakage rates based on that computed for a reactor
-system for a single material problem with a total cross section
-equivalent to that of homogenized light water reactor. Given the
-optimal local domain size, they observed for load balanced situations
-that nearest neighbor communication costs during transport are a small
-fraction of total compute time, leading to the conclusion that domain
-decomposition is feasible for large-scale Monte Carlo calculations.
-
-\subsubsection{Load Balancing Concerns}
-\label{subsubsec:mc_load_balancing}
-Although domain decomposition was shown to be efficient in a perfectly
-load balanced situation in Siegel's work \citep{siegel_analysis_2012},
-careful consideration must be made for situations where this is not
-the case. Given the stochastic nature of the problem and lack of a
-globally homogeneous domain, parallel Monte Carlo simulations are
-inherently load imbalanced. Procassini and others worked to alleviate
-some of the load imbalances that are generated by both particle and
-spatial parallelism and are therefore applicable to the MSOD algorithm
-\citep{procassini_dynamic_2005}. They chose a dynamic balancing scheme
-in which the number of times a particular domain was replicated was
-dependent on the amount of work in that domain (i.e. domains with a
-high particle flux and therefore more particle histories to compute
-require more work). In this variation, domains that require more work
-will be replicated more frequently at reduced particle counts in each
-replication. Furthermore, Procassini and colleagues noted that as the
-simulation progressed and particles were transported throughout the
-domain, the amount of replication for each domain would vary as
-particle histories began to diffuse, causing some regions to have
-higher work loads and some to have smaller work loads than the initial
-conditions.
-
-Consider the example in Figure~\ref{fig:procassini_example} adapted
-from Procassini's work.
-\begin{figure}[htpb!]
-  \begin{center}
-    \scalebox{1.5}{
-      \input{chapters/parallel_mc/procassini_example.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Example illustrating how domain decomposition can
-      create load balance issues in Monte Carlo.}  \textit{A domain is
-      decomposed into 4 zones on 8 processors with a point source in
-      the lower left zone. As the particles diffuse from the source in
-      the random walk sequence as shown in the top row, their tracks
-      populate the entire domain. As given in the bottom row, as the
-      global percentage of particles increases in a zone, that zone's
-      replication count is increased.}}
-  \label{fig:procassini_example}
-\end{figure}
-In this example, a geometry is decomposed into 4 domains on 8
-processors with a point source in the bottom left domain. To begin,
-because the point source is concentrated in one domain, that domain is
-replicated 5 times such the amount of work it has to do per processor
-is roughly balanced with the others. As the particles begin to diffuse
-away from the point source, the amount of replication is adjusted to
-maintain load balance. Near the end of the simulation, the diffusion
-of particles is enough that all domains have equal replication.  By
-doing this, load balance is improved as each domain has approximately
-equal work although each domain may represent a different spatial
-location and have a differing number of histories to
-compute. Compared to Wagner's work where the fission source was
-distributed relatively evenly throughout the domain, fixed source
-problems (and especially those that have a point-like source) like
-those presented in Procassini's work will be more prone to changing
-load balance requirements.
-
-\subsubsection{Reproducible Domain Decomposed Results}
-\label{subsubsec:reproducible_mc}
-The 2006 work of Brunner is notable in that the Monte Carlo codes used
-to implement and test the algorithms adhered to a strict policy of
-generating identical results independent of domain decomposition or
-domain replication as derived from the work of Gentile and colleagues
-\citep{gentile_obtaining_2005}. In Gentile's work, a procedure is
-given for obtaining results reproducible to machine precision for an
-arbitrary number of processors and domains. Differences can arise from
-using a different random number sequence in each domain and performing
-a sequence of floating point operations on identical data in a
-different order, leading to variations in round-off error and
-ultimately a non-identical answer. They use a simple example,
-recreated below in Figure~\ref{fig:gentile_example}, that illustrates
-these issues.
-\begin{figure}[htpb!]
-  \begin{center}
-    \scalebox{1.5}{
-      \input{chapters/parallel_mc/gentile_example.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Gentile's example illustrating how domain
-      decomposition can create reproducibility issues in Monte Carlo.}
-    \textit{Both particles A and B start in zone 1 on processor
-      1. Particle A moves to zone 2 on processor 2 and scatters back
-      to zone 1 while B scatters in zone 1 and remains there. A1 and
-      A2 denote the track of particle A that is in zone 1 while B1 and
-      B2 denote the track of particle B that is in zone 1.}}
-  \label{fig:gentile_example}
-\end{figure}
-In this example, the domain is decomposed on two processors with each
-processor owning one of the two zones. Starting with particle A, it is
-born in zone 1 and is transported to zone 2 where a scattering event
-occurs. Concerning the first reproducibility issue, if the same
-sequence of random numbers is not used to compute the trajectory from
-the new scattering event, we cannot expect to achieve the same result
-if the domain were not decomposed. If the numbers are different, the
-scattering event in zone 2 may keep the particle there or even eject
-it from the domain if the sequence were different. The second issue is
-demonstrated by adding another particle B that remains in the
-domain. In this case, an efficient algorithm will transport particle A
-on processor 1 until it leaves zone 1 and then transport particle
-B. Particle A will not renter the domain until it has been
-communicated to processor 2, processor 2 performs the transport, and
-it is communicated back to processor 1. If we are doing a track-length
-tally in zone 1, then we sum the tracks lengths observed in that
-zone. In the single processor, single zone case particle A would be
-transported in its entirety and then particle B transported. This
-would result in a tally sum with the following order of operations:
-$(((A1+A2)+B1)+B2)$. If we were instead to use 2 processors, we would
-instead have the following order: $(((A1+B1)+B2)+A2)$. In the context
-of floating point operations, we cannot expect these to have an
-identical result to machine precision as round-off approximations will
-differ resulting in non-commutative addition.
-
-Procassini's solutions to these problems are elegant in that they
-require a minimal amount of modification to be applied to the Monte
-Carlo algorithm. To solve the first issue, in order to ensure each
-particle maintains an invariant random number sequence that determines
-its behavior regardless of domain decomposition, each particle is
-assigned a random number seed that describes its current state upon
-entering the domain of a new processor. These seeds are derived from
-the actual geometric location of the particle such that it is
-decomposition invariant. Non-commutative floating point operations are
-overcome by instead mapping floating point values to 64-bit integer
-values for which additions will always be commutative. Once the
-operations are complete, these integers are mapped back to floating
-point values.
-
-\subsection{Parallel Adjoint Method}
-\label{subsec:parallel_adjoint}
-We can take much of what was learned from the survey of parallel Monte
-Carlo methods for radiation transport and directly apply it to a
-parallel formulation of our stochastic linear solvers. Direct analogs
-can be derived from these works by noting that the primary difference
-between solving a linear system with Monte Carlo methods and fixed
-source Monte Carlo transport problems is the content of the Markov
-chains that are generated. The transitions represented by these chains
-are bound by probabilities and weights and are initiated by the
-sampling of a source. In the context of transport problems, those
-transitions represent events such as particle scattering and
-absorption with probabilities that are determined by physical data in
-the form of cross sections. For stochastic matrix inversion, those
-transitions represent moving between the equations of the linear
-system (and therefore the physical domain which they represent) and
-their probabilities are defined by the coefficients of those
-equations. Ultimately, we tally the contributions to generate
-expectation values in the desired states as we progress through the
-chains. Therefore, parallel methods for Monte Carlo radiation
-transport can be abstracted and we can use those concepts that apply
-to matrix inversion methods as an initial means of developing a
-parallel Neumann-Ulam-type solver. Based on the results observed in
-the last decade of parallel Monte Carlo development, we can generate
-many practical questions that this type of work could answer.
-
-Given a decomposed domain, per Wagner's work
-\citep{wagner_hybrid_2010} is clear that domain overlap significantly
-reduces the amount of communication required between adjacent domains
-as histories move to states that are not owned by the local
-processor. How much domain overlap is suitable for matrix inversion
-problems? Are we memory limited by large problems such that only so
-much overlap is feasible? Are there metrics related to the properties
-of the matrix including the eigenvalues and sparsity pattern such that
-we can provide guidelines for selecting the amount of overlap
-required? Furthermore, as Wagner and colleagues observed
-\citep{wagner_hybrid_2010}, with marginal domain overlap for reactor
-problems, the percentage of histories leaving the local domain can
-easily be reduced to less than 1\%. Evans and colleagues in their
-initial MCSA development typically used only 50 particle histories in
-order to compute a correction in an MCSA iteration
-\citep{evans_monte_2012}. Per the Central Limit Theorem, that
-correlates to a statistical uncertainty of 14.1\% in the correction,
-yet as compared to conventional Krylov solvers, MCSA achieved
-identical numerical results. If good convergence and numerically
-accurate solutions can be achieved with such a large uncertainty in
-the correction computation, then perhaps with enough domain overlap
-the minimal amount of histories that do transition to non-local states
-can be ignored and thus eliminate all communication in the parallel
-Monte Carlo adjoint solver transport sequence and create an
-embarrassingly parallel method. Given Gropp's work on parallel Krylov
-methods that we discussed in
-\S~\ref{subsec:projection_method_performance} and the empirical
-results of Siegel \citep{siegel_analysis_2012}, we know that these
-nearest neighbor computations between adjacent domains do not scale as
-well as global reduction operations and therefore we are improving
-scaling by eliminating them from the transport sequence. It will be
-important to determine if the expectation value bias in the MCSA
-correction generated by this approximation will remain within the
-bounds of the already high statistical uncertainty for unbiased
-estimates, thus providing numerically equivalent results.
-
-From a domain replication perspective, this may be difficult to
-achieve with a production scale linear solver. Typically, memory is at
-a premium and therefore the more distinct domains available in the
-decomposition, the spatially finer and/or the numerically more
-accurate the discretization that can be implemented. How much
-replication is possible for large problems? How does replication
-facilitate parallel performance when coupled with domain overlap? How
-can we measure how much domain overlap is feasible? Replicating
-domains may therefore run into memory limitations for exceptionally
-large problems such that the operator, solution vector, and source
-vector must be copied in their entirety multiple times. From a
-resiliency standpoint, such an operation will be required, and
-therefore its performance and memory implications on conventional
-problems must be analyzed.
-
-\subsection{Parallel MCSA}
-\label{subsec:parallel_mcsa}
-With a parallel adjoint Neumann-Ulam solver implementation, the
-parallel implementation of the MCSA method will be trivial. Recall the
-MCSA iteration procedure outlined in Eq~(\ref{eq:mcsa}). In
-\S~\ref{sec:parallel_krylov_methods} we discussed parallel matrix and
-vector operations as utilized in conventional Krylov methods. We
-utilize these here for the parallel MCSA implementation. In the first
-step, a parallel matrix-vector multiply is used to apply the split
-operator to the previous iterate's solution. A parallel vector update
-is then performed with the source vector to arrive at the initial
-iteration guess. In the next step, the residual is computed by the
-same operations where now the operator is applied to the solution
-guess with a parallel matrix-vector multiply and then a parallel
-vector update with the source vector is performed. Once the correction
-is computed with a parallel adjoint Neumann-Ulam solve, this
-correction is applied to the guess with a parallel vector update to
-get the new iteration solution. Additionally, as given by
-Eq~(\ref{eq:mcsa_stopping_criteria}), 2 parallel vector reductions
-will be required to check the stopping criteria: one initially to
-compute the infinity norm of the source vector, and another at every
-iteration to compute the infinity norm of the residual vector. For
-this implementation, all of the issues that will be potentially
-generated by the parallel adjoint solver implementation will manifest
-themselves here as the quality of the correction will be of intense
-study.
-
-In addition to parallel implementation and performance, MCSA's
-potential for aiding advancement in non-symmetric matrix solutions
-leads to a natural comparison with the GMRES algorithm. As both
-solvers are aimed at the same class of problem, we desire a set of
-metrics that will allow us to quantitatively compare the two. Given
-that the Krylov subspace maintained by GMRES can become large, do we
-benefit from a memory standpoint with an MCSA scheme in that no
-subspace is required? Does this benefit outweigh the fact that the
-linear operator must be explicitly formed in order to build the
-transition probabilities for the random walk sequence? For
-non-symmetric systems, does MCSA exhibit similar convergence
-properties to Krylov methods? If a Krylov methods build a subspace,
-can those memory savings in MCSA be used to implement domain
-replication in the adjoint solver? Such questions can be answered by a
-comparative study of the two solvers that controls the system size and
-the iterations required to converge.
+%%---------------------------------------------------------------------------%%
+\pagebreak
+\bibliographystyle{ieeetr}
+\bibliography{references}
+\end{document}

### 2012-11-20 doc/mandc2013/notes/spectral_theory.tex
adding makefile and stencil fig to spectral theory doc
updates to spectral note

@@
 \date{\today}
-\title{A Spectral Theory for the Domain-Decomposed Adjoint
+\title{A Spectral Analysis of the Domain-Decomposed Adjoint
   Neumann-Ulam Method}
@@
   properties of the linear operator. Relationships for the average
-  length of the Markov chains and the fraction of histories leaking
+  length of the random walks and the fraction of histories leaking
   from a parallel domain are made with respect to the eigenvalues of
@@
 \end{multline}
-
+\begin{figure}[htpb!]
+  \begin{center}
+    \scalebox{1.25}{\input{stencil.pdftex_t}}
+  \end{center}
+  \caption{\textbf{Nine-point Laplacian stencil.}}
+  \label{fig:stencil}
+\end{figure}
 We then have the following linear system to solve:
@@
 where $\ve{D}$ is the diffusion operator, $\ve{S}$ is the source in
-vector form and $\boldsymbol{\phi}$ is the vector of unkown fluxes.
+vector form and $\boldsymbol{\phi}$ is the vector of unkown fluxes. We
+note here that the diffusion operator is symmetric.
 
-To close the system a non-reentrant current condition is applied
-to all boundaries of the domain.
+To close the system a non-reentrant current condition is applied to
+all boundaries of the domain. For our problems, we choose the
+formulation of Duderstadt by assuming the flux is zero at some ghost
+point beyond the grid. Consider for example the equations on the $i=0$
+boundary of the domain:
+\begin{multline}
+  -\frac{1}{6\Delta^2}[4 \phi_{-1,j} + 4 \phi_{1,j}
+    + 4 \phi_{0,j-1} + 4 \phi_{0,j+1} + \phi_{-1,j-1}\\ +
+    \phi_{-1,j+1} + \phi_{1,j-1} + \phi_{1,j+1} - 20
+    \phi_{0,j}] + \Sigma_a \phi_{0,j} = s_{0,j}\:.
+  \label{eq:x_min_bnd}
+\end{multline}
+Here we note some terms where $i=-1$ and therefore are representative
+of grid points beyond the boundary of the grid. We set the flux at
+these points to be zero, giving a valid set of equations for the $i=0$
+boundary:
+\begin{multline}
+  -\frac{1}{6\Delta^2}[4 \phi_{1,j} + 4 \phi_{0,j-1} + 4 \phi_{0,j+1}
+    \\ + \phi_{-1,j+1} + \phi_{1,j-1} + \phi_{1,j+1} - 20 \phi_{0,j}]
+  + \Sigma_a \phi_{0,j} = s_{0,j}\:.
+  \label{eq:x_min_bnd}
+\end{multline}
 
@@
 %%---------------------------------------------------------------------------%%
-\section{Length of Random Walks}
+\section{Random Walk Lengths}
 

### 2012-11-20 doc/mandc2013/notes/spectral_theory.tex
spectral note updates

@@
 \date{\today}
-\title{A Spectral Analysis of the Domain-Decomposed Adjoint
+\title{A Spectral Analysis of the Domain Decomposed Adjoint
   Neumann-Ulam Method}
@@
 
-  The serial and parallel behavior of the adjoint Neumann-Ulam method
+  The domain decomposed behavior of the adjoint Neumann-Ulam method
   for solving linear operator equations is analyzed using the spectral
   properties of the linear operator. Relationships for the average
-  length of the random walks and the fraction of histories leaking
-  from a parallel domain are made with respect to the eigenvalues of
+  length of the adjoint random walks and the fraction of histories
+  leaking from a domain are made with respect to the eigenvalues of
   the linear operator. The one-speed, two-dimensional neutron
@@
   theory for symmetric operators. These relationships will serve as
-  guidelines for selecting an appropriate parallel decomposition
-  stratagy and provide a basis for performance models for future
-  implementations of a parallel adjoint Neumann-Ulam method.
+  guidelines for selecting an appropriate parallel algorithm strategy
+  and provide a basis for performance models for future parallel
+  implementations of the adjoint Neumann-Ulam method.
 
@@
 \end{equation}
-where $\ve{A} \in \mathbb{R}^{N \times N}$ is a matrix operator such
+where $\ve{A} \in \mathbb{R}^{N \times N}$ is a linear operator such
 that $\ve{A} : \mathbb{R}^{N} \rightarrow \mathbb{R}^{N}$, $\ve{x} \in
@@
 %%---------------------------------------------------------------------------%%
+\section{Domain Decomposition}
+
+%%---------------------------------------------------------------------------%%
 \section{Model Problem}
@@
 where $\phi$ is the neutron flux, $\Sigma_a$ is the absorption cross
-section, $S$ is the source of neutrons and $D$ is the diffusion
+section, $S$ is the source of neutrons. In addition, $D$ is the diffusion
 coefficient defined as:
@@
 + \Sigma_s$ is the total cross section, and $\bar{\mu}$ is the cosine
-of the average scattering angle. For simplification, we will take
+of the average scattering angle. For simplicity, we will take
 $\bar{\mu} = 0$ for our analysis giving $D=(3 \Sigma_t)^{-1}$. In
@@
 that the cross sections remain constant throughout. Doing this permits
-us to simplify Eq~(\ref{eq:diffusion_eq}):
+us to rewrite Eq~(\ref{eq:diffusion_eq}) as:
 \begin{equation}
@@
 To close the system a non-reentrant current condition is applied to
-all boundaries of the domain. For our problems, we choose the
+all global boundaries of the domain. For our problems, we choose the
 formulation of Duderstadt by assuming the flux is zero at some ghost
@@
 Here we note some terms where $i=-1$ and therefore are representative
-of grid points beyond the boundary of the grid. We set the flux at
+of grid points beyond the boundary of the domain. We set the flux at
 these points to be zero, giving a valid set of equations for the $i=0$
@@
 \end{multline}
+We repeat this procedure for the other boundaries of the domain.
 

### 2012-11-20 doc/mandc2013/notes/spectral_theory.tex
updated adjoint section and started on decomposition section in spectral note

@@
 \mathbb{R}^N$ is the solution vector, and $\ve{b} \in \mathbb{R}^N$ is
-the forcing term. We choose the adjoint method Monte Carlo method to
-invert the linear operator. We begin by defining the linear system
+the forcing term. Choosing the adjoint method Monte Carlo method to
+invert the linear operator, we begin by defining the linear system
 adjoint to Eq~(\ref{eq:linear_problem}):
@@
 where $\ve{y}$ and $\ve{d}$ are the adjoint solution and source
-respectively and $\ve{A}^T$ is the adjoint operator. We can split this
-equation by defining the following inner product equivalence
-\cite{spanier_monte_1969}:
+respectively and $\ve{A}^T$ is the adjoint operator.  With this
+statement we can then define the split equation:
 \begin{equation}
-  \langle \ve{A}^T \ve{x}, \ve{y} \rangle = \langle \ve{x}, \ve{A}
-  \ve{y} \rangle\:.
-  \label{eq:adjoint_operator_product}
-\end{equation}
-With this statement we can then define the split equation:
-\begin{equation}
-  \ve{y} = \ve{H}^T \ve{y} + \ve{d}\:.
+  \ve{y} = \ve{H}^T \ve{y} + \ve{d}\:,
   \label{eq:adjoint_split_system}
 \end{equation}
-For convergence, we require that the spectral radius of $\ve{H}$ must
+where $\ve{H}$ is defined as the \textit{iteration matrix}.  For
+convergence, we require that the spectral radius of $\ve{H}$ must
 remain less than 1 as $\ve{H}^T$ contains the same eigenvalues and
-therefore has the same spectral radius. From this definition it
-follows that:
-\begin{equation}
-  \langle \ve{x}, \ve{d} \rangle = \langle \ve{y}, \ve{b} \rangle\:.
-  \label{eq:adjoint_vector_relation}
-\end{equation}
-Using these definitions, we can derive an estimator from the adjoint
-method that will also give the solution vector, $\ve{x}$. As with the
-direct method, we can acquire the adjoint solution by forming the
-Neumann series by writing Eq~(\ref{eq:adjoint_split_system}) as:
+therefore has the same spectral radius.  Using these definitions, we
+can derive an estimator from the adjoint method that will also give
+the solution vector, $\ve{x}$. As with the direct method, we can
+acquire the adjoint solution by forming the Neumann series by writing
+Eq~(\ref{eq:adjoint_split_system}) as:
 \begin{equation}
@@
 We can readily build an estimator for the adjoint solution from this
-series expansion, but we instead desire the direct solution. We
-achieve this by using Eq~(\ref{eq:adjoint_vector_relation}) as a
-constraint. Here we have 2 unknowns, $\ve{y}$ and $\ve{d}$ and
-therefore we require a second constraint to close the system. As a
-second constraint we select:
+series expansion, but we instead desire the direct solution. The
+adjoint estimator can be related to the solution by defining the
+following inner product equivalence \cite{spanier_monte_1969}:
+\begin{equation}
+  \langle \ve{A}^T \ve{x}, \ve{y} \rangle = \langle \ve{x}, \ve{A}
+  \ve{y} \rangle\:.
+  \label{eq:adjoint_operator_product}
+\end{equation}
+From this definition it follows that:
+\begin{equation}
+  \langle \ve{x}, \ve{d} \rangle = \langle \ve{y}, \ve{b} \rangle\:.
+  \label{eq:adjoint_vector_relation}
+\end{equation}
+ Here we have 2 unknowns, $\ve{y}$ and $\ve{d}$ and therefore we
+ require two constraints to close the system. We use
+ Eq~(\ref{eq:adjoint_vector_relation}) as the first constraint and as
+ a second constraint we select:
 \begin{equation}
@@
     E\{X_j\} &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-    \sum_{i_k}^{N} b_{i_0} h_{i_0,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}
-    \delta_{i_k,j} \\ &= x_{j}\:,
+    \sum_{i_k}^{N} b_{i_0} h_{i_0,i_1}h_{i_1,i_2}\ldots
+    h_{i_{k-1},i_k} \delta_{i_k,j} \\ &= x_{j}\:,
   \end{split}
@@
 \section{Domain Decomposition}
+For very large linear systems, two things are likely to occur: a
+single compute node does not contain enough memory to hold the entire
+problem or a single compute node takes a prohibitively long time to
+solve the system. Domain decomposition methods permit the global
+computational domain to be subdivided into smaller regions. These
+smaller regions require less memory to store and fewer computational
+resources to solve as compared to the global problem. Typically, these
+smaller regions are then represented separately in indiviudal compute
+nodes. Therefore, in order to solve large scale problems with the
+Neumann-Ulam method, we must explore its behavior in a domain
+decomposed evironment.
+
+For reactor physics Monte Carlo simulations, domain decomposition has
+been identified as a key principle in moving forward in high
+performance computing
+\cite{brunner_comparison_2006,siegel_analysis_2012}. To accomplish
+this, we recognize from the literature that stochastic histories must
+be transported from domain to domain as the simulation progress and
+they transition to states that are not in the local domain. Because we
+have chosen a domain decomposition strategy in a parallel environment,
+this means that communication of these histories must occur between
+compute nodes owning neighboring pieces of the global domain.
 
@@
 where $\phi$ is the neutron flux, $\Sigma_a$ is the absorption cross
-section, $S$ is the source of neutrons. In addition, $D$ is the diffusion
-coefficient defined as:
+section, $S$ is the source of neutrons. In addition, $D$ is the
+diffusion coefficient defined as:
 \begin{equation}
@@
 \begin{multline}
-  -\frac{1}{6\Delta^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j}
-    + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ +
-    \phi_{i-1,j+1} + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20
-    \phi_{i,j}] + \Sigma_a \phi_{i,j} = s_{i,j}\:,
+  -\frac{1}{6\Delta^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j} + 4
+    \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ + \phi_{i-1,j+1}
+    + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20 \phi_{i,j}] + \Sigma_a
+  \phi_{i,j} = s_{i,j}\:,
   \label{eq:fd_system}
@@
 \begin{multline}
-  -\frac{1}{6\Delta^2}[4 \phi_{-1,j} + 4 \phi_{1,j}
-    + 4 \phi_{0,j-1} + 4 \phi_{0,j+1} + \phi_{-1,j-1}\\ +
-    \phi_{-1,j+1} + \phi_{1,j-1} + \phi_{1,j+1} - 20
-    \phi_{0,j}] + \Sigma_a \phi_{0,j} = s_{0,j}\:.
+  -\frac{1}{6\Delta^2}[4 \phi_{-1,j} + 4 \phi_{1,j} + 4 \phi_{0,j-1} +
+    4 \phi_{0,j+1} + \phi_{-1,j-1}\\ + \phi_{-1,j+1} + \phi_{1,j-1} +
+    \phi_{1,j+1} - 20 \phi_{0,j}] + \Sigma_a \phi_{0,j} = s_{0,j}\:.
   \label{eq:x_min_bnd}

### 2012-11-20 doc/mandc2013/notes/spectral_theory.tex
updates to data and tex for spectral analysis

@@
 \end{abstract}
+
+%%---------------------------------------------------------------------------%%
+\section{Introduction}
+For very large linear systems, two things are likely to occur: a
+single compute node does not contain enough memory to hold the entire
+problem or a single compute node takes a prohibitively long time to
+solve the system. Domain decomposition methods permit the global
+computational domain to be subdivided into smaller regions. These
+smaller regions require less memory to store and fewer computational
+resources to solve as compared to the global problem. Typically, these
+smaller regions are then represented separately in indiviudal compute
+nodes. Therefore, in order to solve large scale problems with the
+adjoint Neumann-Ulam method, we must explore its behavior in a domain
+decomposed evironment.
+
+For reactor physics Monte Carlo simulations, domain decomposition has
+been identified as a key principle in moving forward in high
+performance computing
+\cite{brunner_comparison_2006,siegel_analysis_2012}. To accomplish
+this, we recognize from the literature that stochastic histories must
+be transported from domain to domain as the simulation progress and
+they transition to states that are not in the local domain. Because we
+have chosen a domain decomposition strategy in a parallel environment,
+this means that communication of these histories must occur between
+compute nodes owning neighboring pieces of the global domain. We wish
+to characterize this communication not only because communication is
+in general expensive, but also because nearest-neighbor communication
+sequences have poor algorithmic strong scaling \cite{gropp_cfd_2001}.
+
+For problems where the linear operator is symmetric, a host of
+analytic theories exist based on the eigenvalue spectrum of the
+operator that characterize their behavior in the context of
+deterministic linear solvers. Using past work, these theories are
+adapted to the domain decomposed adjoint Neumann-Ulam method using the
+one-speed, two-dimensional neutron diffusion equation. In this paper
+we describe the adjoint Neumann-Ulam method followed by a presentation
+and finite difference discretization of the model problem. Using the
+linear system generated by this discretization, we use a spectral
+theory to generate analytic relations for the eigenvalues of the
+operator. Using the Eigenvalue spectra, we then build relationships to
+characterize the transport of random walk histories in a decomposed
+domain and compare these analytic results to numerical experiments
+conducted with the model problem.
 
@@
 %%---------------------------------------------------------------------------%%
-\section{Domain Decomposition}
-For very large linear systems, two things are likely to occur: a
-single compute node does not contain enough memory to hold the entire
-problem or a single compute node takes a prohibitively long time to
-solve the system. Domain decomposition methods permit the global
-computational domain to be subdivided into smaller regions. These
-smaller regions require less memory to store and fewer computational
-resources to solve as compared to the global problem. Typically, these
-smaller regions are then represented separately in indiviudal compute
-nodes. Therefore, in order to solve large scale problems with the
-Neumann-Ulam method, we must explore its behavior in a domain
-decomposed evironment.
-
-For reactor physics Monte Carlo simulations, domain decomposition has
-been identified as a key principle in moving forward in high
-performance computing
-\cite{brunner_comparison_2006,siegel_analysis_2012}. To accomplish
-this, we recognize from the literature that stochastic histories must
-be transported from domain to domain as the simulation progress and
-they transition to states that are not in the local domain. Because we
-have chosen a domain decomposition strategy in a parallel environment,
-this means that communication of these histories must occur between
-compute nodes owning neighboring pieces of the global domain.
-
-%%---------------------------------------------------------------------------%%
 \section{Model Problem}
@@
 discretize the problem. For the Laplacian, we choose the 9-point
-stencil shown in Figure~\ref{fig:stencil} \cite{leveque_finite_2007}:
+stencil shown in Figure~\ref{fig:stencil} over a grid of size $h$
+\cite{leveque_finite_2007}:
 \begin{multline}
-  \nabla^2_9\phi = \frac{1}{6\Delta^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j}
+  \nabla^2_9\phi = \frac{1}{6h^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j}
     + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ +
@@
 \begin{multline}
-  -\frac{1}{6\Delta^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j} + 4
+  -\frac{1}{6h^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j} + 4
     \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ + \phi_{i-1,j+1}
@@
 \begin{equation}
-  \ve{D}\boldsymbol{\phi}=\ve{S}\:,
+  \ve{D}\boldsymbol{\phi}=\ve{s}\:,
   \label{eq:operator_system}
@@
 
-To close the system a non-reentrant current condition is applied to
-all global boundaries of the domain. For our problems, we choose the
-formulation of Duderstadt by assuming the flux is zero at some ghost
-point beyond the grid. Consider for example the equations on the $i=0$
-boundary of the domain:
+To close the system, a set of boundary conditions is required. In the
+case of a non-reentrant current condition applied to all global
+boundaries of the domain, we choose the formulation of Duderstadt by
+assuming the flux is zero at some ghost point beyond the
+grid. Consider for example the equations on the $i=0$ boundary of the
+domain:
 \begin{multline}
-  -\frac{1}{6\Delta^2}[4 \phi_{-1,j} + 4 \phi_{1,j} + 4 \phi_{0,j-1} +
+  -\frac{1}{6h^2}[4 \phi_{-1,j} + 4 \phi_{1,j} + 4 \phi_{0,j-1} +
     4 \phi_{0,j+1} + \phi_{-1,j-1}\\ + \phi_{-1,j+1} + \phi_{1,j-1} +
@@
 \begin{multline}
-  -\frac{1}{6\Delta^2}[4 \phi_{1,j} + 4 \phi_{0,j-1} + 4 \phi_{0,j+1}
+  -\frac{1}{6h^2}[4 \phi_{1,j} + 4 \phi_{0,j-1} + 4 \phi_{0,j+1}
     \\ + \phi_{-1,j+1} + \phi_{1,j-1} + \phi_{1,j+1} - 20 \phi_{0,j}]
   + \Sigma_a \phi_{0,j} = s_{0,j}\:.
-  \label{eq:x_min_bnd}
+  \label{eq:x_min_bnd_2}
 \end{multline}
@@
 
+For reflecting boundary conditions, the net current across a boundary
+is zero.
+
 %%---------------------------------------------------------------------------%%
 \section{Spectral Analysis}
+The convergence of the Neumann series approximated by the Monte Carlo
+solver is dependent on the Eigenvalues of the iteration matrix. In
+addition, the Eigenvalues of the linear operator from which the
+iteration matrix was derived allow for further analysis of convergence
+properties. We will compute the Eigenvalues of these matrices by
+assuming Eigenfunctions of the form \cite{leveque_finite_2007}:
+\begin{equation}
+  \Phi_{p,q}(x,y) = e^{2 \pi \imath p x} e^{2 \pi \imath q y}\:,
+  \label{eq:eigenfunction_form}
+\end{equation}
+where different combinations of $p$ and $q$ represent the different
+Eigenmodes of the solution. As these are valid forms of the solution,
+then the action of the linear operator on these Eigenfunctions should
+give the Eigenvalues of the matrix.
+
+For the model problem, we first compute the Eigenvalues for the
+diffusion operator $\ve{D}$ by applying the operator the
+Eigenfunctions and noting that $x=ih$ and $y=jh$:
+\begin{multline}
+  \ve{D}\Phi_{p,q}(x,y) = \lambda_{p,q}(\ve{D}) =\\ -\frac{D}{6h^2}[4
+    e^{-2 \pi \imath p h} + 4 e^{2 \pi \imath p h} + 4 e^{-2 \pi
+      \imath q h} + 4 e^{2 \pi \imath q h} + e^{-2 \pi \imath p h} e^{-2
+      \pi \imath q h} \\ + e^{-2 \pi \imath p h} e^{2 \pi \imath q q} +
+    e^{2 \pi \imath p h} e^{-2 \pi \imath q h} + e^{2 \pi \imath p h}
+    e^{2 \pi \imath q h} - 20] + \Sigma_a \:.
+  \label{eq:deriv_diff_1}
+\end{multline}
+Using Euler's formula, we can collapse the exponentials to
+trigonometric functions:
+\begin{equation}
+  \lambda_{p,q}(\ve{D}) = -\frac{D}{6h^2}[ 8 \cos(\pi p h) + 8
+    \cos(\pi q h) + 4 \cos(\pi p h) \cos(\pi q h) - 20] + \Sigma_a\:.
+  \label{eq:deriv_diff_2}
+\end{equation}
+
+As Eq~(\ref{eq:diffusion_eq}) is diagonally dominant, Jacobi
+preconditioning is sufficient to reduce the spectral radius of the
+operator below unity and therefore ensure convergence of the Neumann
+series. The preconditioner in this case is then $\ve{M} =
+diag(\ve{D})$ such that we are solving the following linear system:
+\begin{equation}
+  \ve{M}^{-1} \ve{D} \boldsymbol{\phi} = \ve{M}^{-1} \ve{s}\:.
+  \label{eq:precond_diffsion}
+\end{equation}
+The operator $\ve{M}^{-1} \ve{D}$ is merely the original diffusion
+operator with each row scaled by the diagonal component. As we have
+defined a homogenous domain, this scaling factor is the same for all
+rows in the operator. This scaling factor, $\alpha$, is then defined
+as the $\phi_{i,j}$ coefficient from Eq~(\ref{eq:fd_system}):
+\begin{equation}
+  \alpha = [\frac{10 D}{3 h^2} + \Sigma_a]^{-1}\:.
+  \label{eq:jacobi_scaling}
+\end{equation}
+Using this coefficient, we then have the following spectrum of
+preconditioned eigenvalues:
+\begin{equation}
+  \lambda_{p,q}(\ve{M}^{-1} \ve{D}) = \alpha \lambda_{p,q}(\ve{D})\:.
+  \label{eq:preconditioned_eigenvalues}
+\end{equation}
+
+A good measure of how well-conditioned a linear system is can be
+obtained by computing the condition number for the linear operator
+defined as:
+\begin{equation}
+  \kappa(\ve{A}) = ||\ve{A}||\ ||\ve{A}^{-1}||\:,
+  \label{eq:condition_number}
+\end{equation}
+which can be related to the Eigenvalues of the operator as:
+\begin{equation}
+  \kappa(\ve{A}) =
+  \frac{\max_{p,q}\lambda_{p,q}(\ve{A})}{\min_{p,q}\lambda_{p,q}(\ve{A})}\:.
+  \label{eq:condition_number_2}
+\end{equation}
+To compute the condition number for the preconditioned diffusion
+system, we must therefore compute the minimum and maximum Eigenvalues
+of the diffusion operator. Per Leveque's text
+\cite{leveque_finite_2007}, we expect to find these minimum and
+maximum eigenvalues when $p=q$. To find these values,
+Eq~(\ref{eq:deriv_diff_2}) is plotted as a function of $p$ with $p=q$
+in Figure~\ref{fig:diffusion_spectrum} without the scaling.
 
 %%---------------------------------------------------------------------------%%
-\section{Random Walk Lengths}
+\section{Numerical Experiments}
 
-%%---------------------------------------------------------------------------%%
-\section{Domain Leakage}
+\subsection{Random Walk Lengths}
+
+\subsection{Domain Leakage}
 

### 2012-11-20 doc/mandc2013/notes/spectral_theory.tex
fixed eigenvalue computation and finished linear operator section

@@
 \end{multline}
-\begin{figure}[htpb!]
+\begin{figure}[t!]
   \begin{center}
@@
 then the action of the linear operator on these Eigenfunctions should
-give the Eigenvalues of the matrix.
+give the Eigenvalues of the matrix as they lie on the unit circle in
+the complex plane.
 
+\subsection{Diffusion Operator Spectrum}
 For the model problem, we first compute the Eigenvalues for the
@@
 maximum eigenvalues when $p=q$. To find these values,
-Eq~(\ref{eq:deriv_diff_2}) is plotted as a function of $p$ with $p=q$
-in Figure~\ref{fig:diffusion_spectrum} without the scaling.
+Eq~(\ref{eq:preconditioned_eigenvalues}) is plotted as a function of
+$p$ with $p=q$ in Figure~\ref{fig:diffusion_spectrum} for various
+values of $\Sigma_a$.
+
+We see a minimum eigenvalue at $p=0$ and a maximum at $p=10$, letting
+us compute the condition number for the Jacobi preconditioned
+diffusion operator:
+\begin{equation}
+  \kappa(\ve{M}^{-1} \ve{D}) =
+  \frac{\frac{D}{6h^2}\Sigma_a}{\frac{D}{3h^2}[8 \cos(\pi 10 h) + 2
+      \cos^2(\pi 10 h) - 10] + \Sigma_a}\:.
+  \label{eq:diffusion_condition}
+\end{equation}
+Visually we can also see how increasing the absorption in the problem
+creates a more well conditioned system as less diffusion is
+expected. If we have a problem with only scattering, $\Sigma_a = 0$
+and we observe an infinite condition number as $\lambda_{min}
+\rightarrow 0$. We note also that the diffusion operator contains
+purely positive Eigenvalues and can therefore be classified as
+symmetric positive definite (SPD).
+
+\subsection{Iteration Matrix Spectrum}
+
+\subsection{Neumann Series Convergence}
+
+\subsection{The Wigner Rational Approximation}
 
@@
 
-\subsection{Random Walk Lengths}
+\subsection{Random Walk Length}
 

### 2012-11-21 doc/mandc2013/notes/spectral_theory.tex
uadded convergence section

@@
 this, we recognize from the literature that stochastic histories must
-be transported from domain to domain as the simulation progress and
+be transported from domain to domain as the simulation progresses and
 they transition to states that are not in the local domain. Because we
@@
 sequences have poor algorithmic strong scaling \cite{gropp_cfd_2001}.
+We observe this because nearest neighbor communication patterns scale
+with the volume and surface sizes of the local domains. The larger the
+volume, the more compute time is spent within the local volume and
+therefore at some point you are limited by the latency of the system,
+or the time it takes to receive the data to operate on. However, at a
+certain point, when the surface of those volumes becomes large enough,
+the amount of data that must be communicated grows large enough that
+it may overcome the bandwidth limitations of the system, resulting in
+a decrease in performance.
 
@@
 \end{equation}
-where $\ve{H}$ is defined as the \textit{iteration matrix}.  For
-convergence, we require that the spectral radius of $\ve{H}$ must
-remain less than 1 as $\ve{H}^T$ contains the same eigenvalues and
-therefore has the same spectral radius.  Using these definitions, we
-can derive an estimator from the adjoint method that will also give
-the solution vector, $\ve{x}$. As with the direct method, we can
-acquire the adjoint solution by forming the Neumann series by writing
+where $\ve{H}$ is defined as the \textit{iteration matrix}.  Using
+these definitions, we can derive an estimator from the adjoint method
+that will also give the solution vector, $\ve{x}$. We can acquire the
+adjoint solution by forming the Neumann series by writing
 Eq~(\ref{eq:adjoint_split_system}) as:
@@
 \end{equation}
-We expand this summation to again yield a series of transitions that
-can be approximated by a Monte Carlo random walk sequence, this time
-forming the Neumann series in reverse order:
+For this series to converge, we require that the spectral radius of
+$\ve{H}$ must remain less than 1 as $\ve{H}^T$ contains the same
+eigenvalues and therefore has the same spectral radius. We expand this
+summation to again yield a series of transitions that can be
+approximated by a Monte Carlo random walk sequence, this time forming
+the Neumann series in reverse order:
 \begin{equation}
@@
 
-Like the direct method, we also desire a criteria for random walk
-termination for problems where only an approximate solution is
-necessary. For the adjoint method, we utilize a \textit{relative
-  weight cutoff}:
+We also desire a criteria for random walk termination for problems
+where only an approximate solution is necessary. For the adjoint
+method, we utilize a \textit{relative weight cutoff}:
 \begin{equation}
@@
 For our numerical experiments, we choose the one-speed,
-two-dimensional neutron diffusion equation as the model problem
+two-dimensional neutron diffusion equation as a model problem
 \cite{duderstadt_nuclear_1976}:
@@
 where $\phi$ is the neutron flux, $\Sigma_a$ is the absorption cross
-section, $S$ is the source of neutrons. In addition, $D$ is the
+section, and $S$ is the source of neutrons. In addition, $D$ is the
 diffusion coefficient defined as:
@@
 \end{equation}
-where $\ve{D}$ is the diffusion operator, $\ve{S}$ is the source in
+where $\ve{D}$ is the diffusion operator, $\ve{s}$ is the source in
 vector form and $\boldsymbol{\phi}$ is the vector of unkown fluxes. We
@@
 \section{Spectral Analysis}
-The convergence of the Neumann series approximated by the Monte Carlo
+The convergence of the Neumann series in
+Eq~(\ref{eq:adjoint_neumann_series}) approximated by the Monte Carlo
 solver is dependent on the Eigenvalues of the iteration matrix. In
@@
 For the model problem, we first compute the Eigenvalues for the
-diffusion operator $\ve{D}$ by applying the operator the
+diffusion operator $\ve{D}$ by applying the operator to the
 Eigenfunctions and noting that $x=ih$ and $y=jh$:
@@
 preconditioning is sufficient to reduce the spectral radius of the
-operator below unity and therefore ensure convergence of the Neumann
-series. The preconditioner in this case is then $\ve{M} =
+iteration matrix below unity and therefore ensure convergence of the
+Neumann series. The preconditioner in this case is then $\ve{M} =
 diag(\ve{D})$ such that we are solving the following linear system:
@@
 operator with each row scaled by the diagonal component. As we have
-defined a homogenous domain, this scaling factor is the same for all
+defined a homogenous domain, the scaling factor is the same for all
 rows in the operator. This scaling factor, $\alpha$, is then defined
@@
 
-A good measure of how well-conditioned a linear system is can be
+A good measure of how well-conditioned is a linear system can be
 obtained by computing the condition number for the linear operator
@@
 values of $\Sigma_a$.
-
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=5in,clip]{diffusion_spectrum.png}
+  \end{center}
+  \caption{\textbf{Eigenvalue spectra for the diffusion equation.}}
+  \label{fig:diffusion_spectrum}
+\end{figure}
 We see a minimum eigenvalue at $p=0$ and a maximum at $p=10$, letting
@@
   \kappa(\ve{M}^{-1} \ve{D}) =
-  \frac{\frac{D}{6h^2}\Sigma_a}{\frac{D}{3h^2}[8 \cos(\pi 10 h) + 2
-      \cos^2(\pi 10 h) - 10] + \Sigma_a}\:.
+  \frac{\frac{D}{6h^2}\Sigma_a}{\frac{D}{3h^2}[8 \cos(10 \pi h) + 2
+      \cos^2(10 \pi h) - 10] + \Sigma_a}\:.
   \label{eq:diffusion_condition}
@@
 \subsection{Iteration Matrix Spectrum}
+The spectral radius of the iteration matrix is obtained by seeking its
+largest Eigenvalue. As with the diffusion operator, we can use the
+same analysis techniques to find the Eigenvalues for the iteration
+matrix. We use a few simplifications by noting that if the Jacobi
+preconditioned spectral radius is $\ve{H} = \ve{I} -
+\ve{M}^{-1}\ve{D}$, the we expect all terms on the diagonal of the
+iteration matrix to be zero such that we have the following stencil:
+\begin{equation}
+  \ve{H}\boldsymbol{\phi} = -\frac{\alpha D}{6h^2}[4 \phi_{i-1,j} + 4
+    \phi_{i+1,j} + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ +
+    \phi_{i-1,j+1} + \phi_{i+1,j-1} + \phi_{i+1,j+1}]\:.
+  \label{eq:iteration_stencil}
+\end{equation}
+Inserting the Eigenfunctions defined by Eq~\ref{eq:eigenfunction_form}
+we get:
+\begin{multline}
+  \lambda_{p,q}(\ve{H}) = -\frac{\alpha D}{6h^2}[4 e^{-2 \pi \imath p
+      h} + 4 e^{2 \pi \imath p h} + 4 e^{-2 \pi \imath q h} + 4 e^{2
+      \pi \imath q h} + e^{-2 \pi \imath p h} e^{-2 \pi \imath q h}
+    \\ + e^{-2 \pi \imath p h} e^{2 \pi \imath q q} + e^{2 \pi \imath
+      p h} e^{-2 \pi \imath q h} + e^{2 \pi \imath p h} e^{2 \pi
+      \imath q h}]\:,
+  \label{eq:iteration_deriv}
+\end{multline}
+which simplifies to:
+\begin{equation}
+  \lambda_{p,q}(\ve{H}) = -\frac{\alpha D}{6h^2}[ 8 \cos(\pi p h) + 8
+    \cos(\pi q h) + 4 \cos(\pi p h) \cos(\pi q h)]\:,
+  \label{eq:iteration_spectrum}
+\end{equation}
+giving the Eigenvalue spectrum for the Jacobi preconditioned iteration
+matrix. As with the linear operator, we expect the maximum Eigenvalue
+to exist when $p=q$. In addition, we note that we have the same cosine
+form of the spectrum as for the linear operator expect that the
+spectra will be shifted by 10 in the $-p$ direction. We therefore
+expect the maximum Eigenvalues to be found when $p=q=0$, giving the
+following for the spectral radius of the Jacobi precondition iteration
+matrix:
+\begin{equation}
+  \rho(\ve{H}) = \frac{10 \alpha D}{3 h^2}\:.
+  \label{eq:iteration_radius}
+\end{equation}
 
 \subsection{Neumann Series Convergence}
+The Neumann-Ulam method for matrix inversion is effectively an
+approximation to a stationary method (the Jacobi method in the case
+of the preconditioning above). Stationary methods for linear systems arise from splitting the
+operator in Eq~(\ref{eq:linear_problem}):
+\begin{equation}
+  \ve{A} = \ve{M} - \ve{N}\:,
+  \label{eq:split_linear_operator}
+\end{equation}
+where the choice of $\ve{M}$ and $\ve{N}$ will be dictated by the
+particular method chosen. Using this split definition of the operator
+we can then write:
+\begin{equation}
+  \ve{M}\ve{x} - \ve{N}\ve{x} = \ve{b}\:.
+  \label{eq:linear_split_equation1}
+\end{equation}
+By rearranging, we can generate a form more useful for analysis:
+\begin{equation}
+  \ve{x} = \ve{H}\ve{x} + \ve{c}\:,
+  \label{eq:linear_split_equation2}
+\end{equation}
+where $\ve{H}=\ve{M}^{-1}\ve{N}$ is defined as the \textit{iteration
+  matrix} and $\ve{c}=\ve{M}^{-1}\ve{b}$. With the solution vector on
+both the left and right hand sides, an iterative method can then be
+formed:
+\begin{equation}
+  \ve{x}^{k+1} = \ve{H}\ve{x}^k + \ve{c}\:,
+  \label{eq:linear_iterative_method}
+\end{equation}
+with $k \in \mathbb{Z}^+$ defined as the \textit{iteration
+  index}. Given this, we can then generate a few statements regarding
+the convergence of such stationary methods. Defining $\ve{e}^k =
+\ve{u}^k - \ve{u}$ as the solution error at the $k^{th}$ iterate, we
+can subtract Eq~(\ref{eq:linear_split_equation2}) from
+Eq~(\ref{eq:linear_iterative_method}) to arrive at an error form of
+the linear problem:
+\begin{equation}
+  \ve{e}^{k+1} = \ve{H}\ve{e}^k\:. 
+  \label{eq:linear_iterative_error}
+\end{equation}
+Our error after $k$ iterations is then:
+\begin{equation}
+  \ve{e}^{k} = \ve{H}^k\ve{e}^0\:. 
+  \label{eq:linear_k_iter_error}
+\end{equation}
+In other words, successive application of the iteration matrix is the
+mechanism driving down the error in a stationary method. We can then
+place restrictions on the iteration matrix by assuming $\ve{H}$ is
+diagonalizable \cite{saad_iterative_2003}, we then have:
+\begin{equation}
+  \ve{e}^{k} =
+  \ve{R}\boldsymbol{\Lambda}^k\ve{R}^{-1}\ve{e}^0\:,
+  \label{eq:linear_k_iter_error_diag}
+\end{equation}
+where $\boldsymbol{\Lambda}$ contains the Eigenvalues of $\ve{H}$ on
+its diagonal and the columns of $\ve{R}$ contain the Eigenvectors of
+$\ve{H}$. Computing the 2-norm of the above form then gives:
+\begin{equation}
+  ||\ve{e}^{k}||_2 \leq ||\boldsymbol{\Lambda}^k||_2\ 
+  ||\ve{R}||_2\ ||\ve{R}^{-1}||_2\ ||\ve{e}^0||_2\:,
+  \label{eq:linear_k_iter_norm1}
+\end{equation}
+which gives:
+\begin{equation}
+  ||\ve{e}^{k}||_2 \leq \rho(\ve{H})^k \kappa(\ve{R})
+  ||\ve{e}^0||_2\:.
+  \label{eq:linear_k_iter_norm2}
+\end{equation}
+For iteration matrices where the Eigenvectors are orthogonal,
+$\kappa(\ve{R})=1$ and the error bound reduces to:
+\begin{equation}
+  ||\ve{e}^{k}||_2 \leq \rho(\ve{H})^k ||\ve{e}^0||_2\:.
+  \label{eq:linear_k_iter_norm3}
+\end{equation}
+We can now restrict $\ve{H}$ by asserting that $\rho(\ve{H}) < 1$
+for a stationary method to converge such that $k$ applications of the
+iteration matrix will not cause the error to grow in
+Eq~(\ref{eq:linear_k_iter_norm3}).
+
+In the adjoint Neumann-Ulam method, $k$ iterations, equivalent to $k$
+applications of the iteration matrix, is effectively approximated by a
+random walk of length $k$ to yield the summation in
+Eq~(\ref{eq:adjoint_neumann_solution}). This random walk length, or
+the number of transitions before the termination of a history (either
+by the weight cutoff, absorption, or exiting the domain) is therefore
+bound to the number of stationary iterations required to converge to
+the specified tolerance. In the case of the adjoint Neumann-Ulam
+method, no such tolerance exists, however, we have specified a weight
+cutoff $W_c$ under which low-weight histories will be prematurely
+terminated as their contributions are deemed minute. After $k$
+iterations, a stationary method is terminated as the error has reached
+some fraction, $\epsilon$, of the initial error:
+\begin{equation}
+  ||\ve{e}^{k}||_2 = \epsilon ||\ve{e}^0||_2\:.
+  \label{eq:linear_k_iter_norm3}
+\end{equation}
+Per Eq~(\ref{eq:linear_k_iter_norm3}), we see that this fraction is
+equivalent to $\epsilon = \rho(\ve{H})$. For the adjoint Neumann-Ulam
+method, if we take this fraction to be the weight cutoff, a measure of
+how accurately the contributions of a particular history to the
+solution are tallied, we then have the following empirical relation
+for $k$:
+\begin{equation}
+  k = \frac{ \log(W_c) }{ \log( \rho(\ve{H}) ) }\:.
+  \label{eq:analytic_k}
+\end{equation}
+This then gives us a means to estimate the length of the random walks
+derived from a particular linear operator based on the Eigenvalues of
+its iteration matrix (independent of the linear operator splitting
+chosen) and based on the weight cutoff parameter used in the
+Neumann-Ulam method
 
 \subsection{The Wigner Rational Approximation}
+In a domain decomposed situtation, not all histories will remain
+within the domain they started in and must instead be
+communicated. This communication, expected to be expensive, was
+analyzed by Siegel and colleagues for idealized, load balanced
+situations for full-core Monte Carlo\cite{siegel_analysis_2012}.  To
+quantify the number of particles that leak out of the local domain
+they define a leakage fraction, $\lambda$, as:
+\begin{equation}
+  \lambda = \frac{average\ \#\ of\ particles\ leaving\ local\ domain}
+          {total\ of\ \#\ of\ particles\ starting\ in\ local\ domain}\:.
+  \label{eq:leakage_fraction}
+\end{equation}
+For their studies, they assumed that the value of $\lambda$ was
+dependent on the total cross section via the Wigner rational
+approximation. The approximation was developed to compute the neutron
+leakage probability from by Wigner as a simple interpolation between
+the small and large fuel lump cases.
 

### 2012-11-21 doc/mandc2013/notes/spectral_theory.tex
finished theory section. experiment results next

@@
   length of the adjoint random walks and the fraction of histories
-  leaking from a domain are made with respect to the eigenvalues of
-  the linear operator. The one-speed, two-dimensional neutron
-  diffusion equation is used as a model problem to test the spectral
-  theory for symmetric operators. These relationships will serve as
-  guidelines for selecting an appropriate parallel algorithm strategy
-  and provide a basis for performance models for future parallel
-  implementations of the adjoint Neumann-Ulam method.
+  leaking from a domain are made with respect to the Eigenvalues of
+  the linear operator. In addition, a formal definition of the optical
+  thickness of a linear operator is presented based in the spectral
+  analysis. The one-speed, two-dimensional neutron diffusion equation
+  is used as a model problem to test the spectral theory for symmetric
+  operators. In general, the spectral theory shows good agreement with
+  measured computational results.
 
@@
 in general expensive, but also because nearest-neighbor communication
-sequences have poor algorithmic strong scaling \cite{gropp_cfd_2001}.
-We observe this because nearest neighbor communication patterns scale
-with the volume and surface sizes of the local domains. The larger the
-volume, the more compute time is spent within the local volume and
-therefore at some point you are limited by the latency of the system,
-or the time it takes to receive the data to operate on. However, at a
-certain point, when the surface of those volumes becomes large enough,
-the amount of data that must be communicated grows large enough that
-it may overcome the bandwidth limitations of the system, resulting in
-a decrease in performance.
+sequences have poor algorithmic strong scaling
+\cite{gropp_high-performance_2001}.  We observe this because nearest
+neighbor communication patterns scale with the volume and surface
+sizes of the local domains. The larger the volume, the more compute
+time is spent within the local volume and therefore at some point you
+are limited by the latency of the system, or the time it takes to
+receive the data to operate on. However, at a certain point, when the
+surface of those volumes becomes large enough, the amount of data that
+must be communicated grows large enough that it may overcome the
+bandwidth limitations of the system, resulting in a decrease in
+performance.
 
@@
 In the adjoint Neumann-Ulam method, $k$ iterations, equivalent to $k$
-applications of the iteration matrix, is effectively approximated by a
+applications of the iteration matrix, are approximated by a
 random walk of length $k$ to yield the summation in
@@
   ||\ve{e}^{k}||_2 = \epsilon ||\ve{e}^0||_2\:.
-  \label{eq:linear_k_iter_norm3}
+  \label{eq:linear_k_iter_norm4}
 \end{equation}
 Per Eq~(\ref{eq:linear_k_iter_norm3}), we see that this fraction is
-equivalent to $\epsilon = \rho(\ve{H})$. For the adjoint Neumann-Ulam
+equivalent to $\epsilon = \rho(\ve{H})^k$. For the adjoint Neumann-Ulam
 method, if we take this fraction to be the weight cutoff, a measure of
@@
 This then gives us a means to estimate the length of the random walks
-derived from a particular linear operator based on the Eigenvalues of
-its iteration matrix (independent of the linear operator splitting
-chosen) and based on the weight cutoff parameter used in the
-Neumann-Ulam method
+that will be generated from a particular linear operator based on the
+Eigenvalues of its iteration matrix (independent of the linear
+operator splitting chosen) and based on the weight cutoff parameter
+used in the Neumann-Ulam method
 
-\subsection{The Wigner Rational Approximation}
+\subsection{Wigner Rational Approximation}
 In a domain decomposed situtation, not all histories will remain
@@
 For their studies, they assumed that the value of $\lambda$ was
-dependent on the total cross section via the Wigner rational
-approximation. The approximation was developed to compute the neutron
-leakage probability from by Wigner as a simple interpolation between
-the small and large fuel lump cases.
+dependent on the total cross section of the system via the Wigner
+rational approximation. This approximation was developed by Wigner to
+compute the neutron leakage probability from as a simple interpolation
+between the small and large fuel lump cases, approximately
+characterizing the leakage behavior of many common geometries.
+\cite{duderstadt_nuclear_1976}. Outlined more thoroughly by Hwang's
+chapter in \cite{azmy_nuclear_2010}, we begin the derivation of
+Wigner's approximation by defining $l$ to be the average chord length
+across the domain, and $\Sigma_t$ to be the total cross section in the
+domain. We then bound the behavior for the leakage fraction in terms
+of the optical thickness, $\Sigma_t l$, of the local domain. In the
+case of a small domain or a small cross section (and therefore small
+probabilty of absorption or scatter), $\Sigma_t l \rightarrow 0,
+\lambda \rightarrow 1$. For a large domain or large cross section,
+$\Sigma_t l \rightarrow \infty, \lambda \rightarrow (\Sigma_t
+l)^{-1}$, where $(\Sigma_t l)^{-1}$ is effectively the number of
+interactions that will occur over the length of the coord. Using
+Wigner's simple interpolation, we then have a simple formula for the
+domain leakage, $\lambda = 1/(1+\Sigma_t l)$. This provides a leakage
+estimate, valid over the entire range of $\Sigma_t l$, in terms of the
+optical thicknes of the system.
+
+In the case of domain decomposed linear operator equations, the chord
+length can be interpreted as the number of discrete states spanning a
+domain, $n_i$, along the typical length of that domain (e.g. the
+number of discrete states in the $i$ direction in a square domain
+using our model diffusion problem). We then seek the equivalent of the
+total cross section, $\Sigma_t$ in terms of the discrete random walk
+behavior. The total cross section can also be interpreted as the
+inverse of the mean free path of particles in the system. In the
+Neumann-Ulam method, this is analagous to the length of the Markov
+chains (and therefore the number of state-to-state transitions)
+generated while sampling the Neumann series. In the previous section,
+we determined this length with Eq~(\ref{eq:analytic_k}) to be
+dependent on the Eigenvalues of the operator. In addition, we must
+also consider the number of states that may be traversed in a given
+transition, $n_s$, along the path of the typical length of the
+system. For our diffusion model problem, $n_s$ would equate to the
+number of states in the $i$ (or $j$ as the problem is symmetric)
+direction that a history could potentially move and is dependent on
+the stencil used for the discretization. We then have the following
+relation for the \textit{optical thickness}, $\tau$, of a linear
+operator equation domain:
+\begin{equation}
+  \tau = \frac{n_i n_s}{k}\:,
+  \label{eq:matrix_optical_thickness}
+\end{equation}
+and inserting Eq~(\ref{eq:analytic_k}) gives:
+\begin{equation}
+  \tau = \frac{n_i n_s \log(\rho(\ve{H}))}{\log(W_c)}\:.
+  \label{eq:matrix_optical_thickness_2}
+\end{equation}
+Using this optical thickness, we can then complete the Wigner rational
+approximation by defining the bounds of $\tau \rightarrow 0, \lambda
+\rightarrow 1$ and $\tau \rightarrow \infty, \lambda \rightarrow
+\tau^{-1}$. We expect this behavior as optically thin domains will
+communicate most of their histories while optically thick domains will
+leak the fraction of histories that did not interact within. Using
+the Wigner approximation, we can then define the leakage fraction out
+of a domain for the adjoint Neumann-Ulam method:
+\begin{equation}
+  \lambda = \frac{1}{1+\tau}\:.
+  \label{eq:analytic_domain_leakage}
+\end{equation}
+Here the fraction is explictly bound to the Eigenvalues of the
+iteration matrix, the size of the domain, the content of the
+discretization stencil, and the weight cutoff selected to terminate
+low weight histories.  We also note here an interesting consequence of
+this approximation: no matter how optically thick a domain may be, the
+domain is still expected to leak some histories and thus communicate
+them. Therefore, no matter how large the domains are made, if they do
+not encapsulate the entire global domain they require some
+communication in order to implement a domain decomposed adjoint
+Neumann-Ulam method.
 
@@
 \section{Conclusion}
+These relationships will serve as guidelines for
+selecting an appropriate parallel algorithm strategy and provide a
+basis for performance models for future parallel implementations of
+the adjoint Neumann-Ulam method.
 

### 2012-11-21 doc/mandc2013/notes/spectral_theory.tex
finished spectral draft

@@
   the linear operator. In addition, a formal definition of the optical
-  thickness of a linear operator is presented based in the spectral
+  thickness of a linear operator is presented based on the spectral
   analysis. The one-speed, two-dimensional neutron diffusion equation
@@
 resources to solve as compared to the global problem. Typically, these
-smaller regions are then represented separately in indiviudal compute
+smaller regions are then represented separately in individual compute
 nodes. Therefore, in order to solve large scale problems with the
 adjoint Neumann-Ulam method, we must explore its behavior in a domain
-decomposed evironment.
+decomposed environment.
 
@@
 $\bar{\mu} = 0$ for our analysis giving $D=(3 \Sigma_t)^{-1}$. In
-addition, to further simplify we will assume a homogenous domain such
+addition, to further simplify we will assume a homogeneous domain such
 that the cross sections remain constant throughout. Doing this permits
@@
 where $\ve{D}$ is the diffusion operator, $\ve{s}$ is the source in
-vector form and $\boldsymbol{\phi}$ is the vector of unkown fluxes. We
+vector form and $\boldsymbol{\phi}$ is the vector of unknown fluxes. We
 note here that the diffusion operator is symmetric.
@@
 \begin{multline}
-  \ve{D}\Phi_{p,q}(x,y) = \lambda_{p,q}(\ve{D}) =\\ -\frac{D}{6h^2}[4
-    e^{-2 \pi \imath p h} + 4 e^{2 \pi \imath p h} + 4 e^{-2 \pi
-      \imath q h} + 4 e^{2 \pi \imath q h} + e^{-2 \pi \imath p h} e^{-2
-      \pi \imath q h} \\ + e^{-2 \pi \imath p h} e^{2 \pi \imath q q} +
-    e^{2 \pi \imath p h} e^{-2 \pi \imath q h} + e^{2 \pi \imath p h}
-    e^{2 \pi \imath q h} - 20] + \Sigma_a \:.
+  \ve{D}\Phi_{p,q}(x,y) = \lambda_{p,q}(\ve{D})
+  =\\ -\frac{D}{6h^2}\Big[4 e^{-2 \pi \imath p h} + 4 e^{2 \pi \imath
+      p h} + 4 e^{-2 \pi \imath q h} + 4 e^{2 \pi \imath q h} + e^{-2
+      \pi \imath p h} e^{-2 \pi \imath q h} \\ + e^{-2 \pi \imath p h}
+    e^{2 \pi \imath q h} + e^{2 \pi \imath p h} e^{-2 \pi \imath q h}
+    + e^{2 \pi \imath p h} e^{2 \pi \imath q h} - 20\Big] + \Sigma_a
+  \:.
   \label{eq:deriv_diff_1}
@@
 operator with each row scaled by the diagonal component. As we have
-defined a homogenous domain, the scaling factor is the same for all
+defined a homogeneous domain, the scaling factor is the same for all
 rows in the operator. This scaling factor, $\alpha$, is then defined
@@
 \begin{equation}
-  \alpha = [\frac{10 D}{3 h^2} + \Sigma_a]^{-1}\:.
+  \alpha = \Bigg[\frac{10 D}{3 h^2} + \Sigma_a\Bigg]^{-1}\:.
   \label{eq:jacobi_scaling}
@@
 \begin{equation}
-  \ve{H}\boldsymbol{\phi} = -\frac{\alpha D}{6h^2}[4 \phi_{i-1,j} + 4
-    \phi_{i+1,j} + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ +
-    \phi_{i-1,j+1} + \phi_{i+1,j-1} + \phi_{i+1,j+1}]\:.
+  \ve{H}\boldsymbol{\phi} = -\frac{\alpha D}{6h^2}[4 \phi_{i-1,j}
+    + 4 \phi_{i+1,j} + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} +
+    \phi_{i-1,j-1}\\ + \phi_{i-1,j+1} + \phi_{i+1,j-1} +
+    \phi_{i+1,j+1}]\:.
   \label{eq:iteration_stencil}
@@
 \begin{multline}
-  \lambda_{p,q}(\ve{H}) = -\frac{\alpha D}{6h^2}[4 e^{-2 \pi \imath p
+  \lambda_{p,q}(\ve{H}) = -\frac{\alpha D}{6h^2}\Big[4 e^{-2 \pi \imath p
       h} + 4 e^{2 \pi \imath p h} + 4 e^{-2 \pi \imath q h} + 4 e^{2
@@
       p h} e^{-2 \pi \imath q h} + e^{2 \pi \imath p h} e^{2 \pi
-      \imath q h}]\:,
+      \imath q h}\Big]\:,
   \label{eq:iteration_deriv}
@@
 \subsection{Wigner Rational Approximation}
-In a domain decomposed situtation, not all histories will remain
+In a domain decomposed situation, not all histories will remain
 within the domain they started in and must instead be
@@
 rational approximation. This approximation was developed by Wigner to
-compute the neutron leakage probability from as a simple interpolation
-between the small and large fuel lump cases, approximately
-characterizing the leakage behavior of many common geometries.
-\cite{duderstadt_nuclear_1976}. Outlined more thoroughly by Hwang's
-chapter in \cite{azmy_nuclear_2010}, we begin the derivation of
-Wigner's approximation by defining $l$ to be the average chord length
-across the domain, and $\Sigma_t$ to be the total cross section in the
-domain. We then bound the behavior for the leakage fraction in terms
-of the optical thickness, $\Sigma_t l$, of the local domain. In the
-case of a small domain or a small cross section (and therefore small
-probabilty of absorption or scatter), $\Sigma_t l \rightarrow 0,
-\lambda \rightarrow 1$. For a large domain or large cross section,
-$\Sigma_t l \rightarrow \infty, \lambda \rightarrow (\Sigma_t
-l)^{-1}$, where $(\Sigma_t l)^{-1}$ is effectively the number of
-interactions that will occur over the length of the coord. Using
-Wigner's simple interpolation, we then have a simple formula for the
-domain leakage, $\lambda = 1/(1+\Sigma_t l)$. This provides a leakage
-estimate, valid over the entire range of $\Sigma_t l$, in terms of the
-optical thicknes of the system.
+compute the neutron leakage probability from the fuel in a reactor as
+a simple interpolation between the small and large fuel lump cases,
+approximately characterizing the leakage behavior of many common
+geometries.  \cite{duderstadt_nuclear_1976}. Outlined more thoroughly
+by Hwang's chapter in \cite{azmy_nuclear_2010}, we begin the
+derivation of Wigner's approximation by defining $l$ to be the average
+chord length across the domain, and $\Sigma_t$ to be the total cross
+section in the domain. We then bound the behavior for the leakage
+fraction in terms of the optical thickness, $\Sigma_t l$, of the local
+domain. In the case of a small domain or a small cross section (and
+therefore small probability of absorption or scatter), $\Sigma_t l
+\rightarrow 0, \lambda \rightarrow 1$. For a large domain or large
+cross section, $\Sigma_t l \rightarrow \infty, \lambda \rightarrow
+(\Sigma_t l)^{-1}$, where $(\Sigma_t l)^{-1}$ is effectively the
+number of interactions that will occur over the length of the
+chord. Using Wigner's simple interpolation, we then have a simple
+formula for the domain leakage, $\lambda = 1/(1+\Sigma_t l)$. This
+provides a leakage estimate, valid over the entire range of $\Sigma_t
+l$, in terms of the optical thickness of the system.
 
@@
 inverse of the mean free path of particles in the system. In the
-Neumann-Ulam method, this is analagous to the length of the Markov
+Neumann-Ulam method, this is analogous to the length of the Markov
 chains (and therefore the number of state-to-state transitions)
@@
 \end{equation}
-Here the fraction is explictly bound to the Eigenvalues of the
-iteration matrix, the size of the domain, the content of the
+Here, the leakage fraction is explicitly bound to the Eigenvalues of
+the iteration matrix, the size of the domain, the content of the
 discretization stencil, and the weight cutoff selected to terminate
@@
 \section{Numerical Experiments}
+To test the relationships developed by the spectral analysis, we form
+two simple numerical experiments using the diffusion model problem:
+one to measure the length of the random walks as a function of the
+iteration matrix Eigenvalues, and one to measure the domain leakage
+fraction as a function of the iteration matrix Eigenvalues. Before
+doing this, we verify our spectral model for the spectral radius of
+the iteration matrix by numerically computing the largest Eigenvalue
+of the diffusion operator using an Eigenvalue solver. For this
+verification, a $100 \times 100$ square grid with $h=0.1$ and
+non-reentrant current boundary conditions was used and the absorption
+cross varied from 0 to 100 while the scattering cross section was
+fixed at unity. Figure~\ref{fig:measured_spec_rad} gives the measured
+spectral radius of the iteration matrix and the computed spectral
+radius for the preconditioned diffusion operator using
+Eq~(\ref{eq:iteration_radius}) as function of the absorption to
+scattering ratio $(\Sigma_a / \Sigma_s)$. We note here that although
+there is excellent agreement between the measured and analytic
+results, slight differences likely arise due to the non-reentrant
+current condition not being included in the spectral analysis. To
+verify this (and serve as a better testing environment), reflecting
+boundary conditions must be implemented.
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=5in,clip]{measured_spec_rad.png}
+  \end{center}
+  \caption{\textbf{Measured and analytic preconditioned diffusion
+      operator spectral radius as a function of the absorption cross
+      section to scattering cross section ratio.} \textit{The blue
+      data was computed numerically by an Eigensolver while the red
+      data was generated by Eq~(\ref{eq:iteration_radius}).}}
+  \label{fig:measured_spec_rad}
+\end{figure}
 
 \subsection{Random Walk Length}
+With the spectral data in hand, we can go about setting up an
+experiment to measure the length of the random walks generated by the
+adjoint Neumann-Ulam solver. To do this, we again use a $100 \times
+100$ square grid with $h=0.1$ and non-reentrant current boundary
+conditions was used and the absorption cross varied from 0 to 100
+while the scattering cross section was fixed at unity. A weight cutoff
+of \sn{1}{-4} was used with 10,000 histories generated by a point
+source source of strength 1 in the center of the domain such that the
+vacuum boundary conditions will have less of an effect on the length
+of the random walks. For each of the histories, the number of
+transitions made was tallied to provide an effective value of $k$ for
+each history. This value was then averaged over all histories to get a
+measured value of $k$ for the particular
+operator. Figure~\ref{fig:measured_length} gives these measurements as
+well as the analytic result computed by Eq~(\ref{eq:analytic_k}) as a
+function of the absorption to scattering ratio $(\Sigma_a /
+\Sigma_s)$. Again, we note here that although there is excellent
+agreement between the measured and analytic results, slight
+differences likely arise due to the non-reentrant current condition
+not being included in the spectral analysis.
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=5in,clip]{measured_length.png}
+  \end{center}
+  \caption{\textbf{Measured and analytic random walk length as a
+      function of the absorption cross section to scattering cross
+      section ratio.} \textit{The blue data was computed numerically
+      by an adjoint Neumann-Ulam implementation while the red data was
+      generated by Eq~(\ref{eq:analytic_k}).}}
+  \label{fig:measured_length}
+\end{figure}
 
 \subsection{Domain Leakage}
+Finally, we seek to measure the leakage from a domain in a domain
+decomposed Monte Carlo calculation and assess the quality of our
+analytic relation for the optical thickness of domain and the
+associated Wigner approximation. For this experiment, we again use a
+$100 \times 100$ square grid with $h=0.1$ and non-reentrant current
+boundary conditions. The grid is then decomposed into 100 square
+domains, 10 in each cardinal direction. For the cross sections, the
+absorption cross was varied from 1 to 100 while the scattering cross
+section was set to zero to create a purely absorbing environment. The
+optical thickness of these domains will vary as a function of the
+absorption cross section if the other parameters are fixed. To compute
+the optical thickness, along with the spectral radius as given by
+Eq~(\ref{eq:iteration_radius}), we also need the parameters $n_i$ and
+$n_s$ which respectively describe the typical domain length and the
+average number of states moved along that typical length per history
+transition. For our grid above, the domains are of size $10 \times 10$
+cells and therefore choosing the $i$ direction there are $n_i=11$ grid
+points or states along the typical length of the domain. Looking at
+the Laplacian stencil in Figure~\ref{fig:stencil}, we see that all
+history transitions will only move a single state in either the $i$ or
+$j$ directions due to the symmetry of the problem. We therefore will
+move on average 1 state along the typical length of the domain per
+history transition, giving $n_s=1$.Figure~\ref{fig:optical_thickness})
+gives the domain optical thickness as a function of $Sigma_a$ as
+specified by Eq~(\ref{eq:matrix_optical_thickness_2}) using the above
+parameters. Here we note that as the domains become more optically
+thick, this relationship becomes exponential as expected.
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=5in,clip]{optical_thickness.png}
+  \end{center}
+  \caption{\textbf{Analytic domain optical thickness as a function of
+      the absorption cross section.} \textit{The data was
+      generated by Eq~(\ref{eq:matrix_optical_thickness_2}).}}
+  \label{fig:optical_thickness}
+\end{figure}
+
+To compute the leakage fraction numerically, over the global domain,
+500,000 histories were sampled from a uniform source of unity. At the
+start of a stage of histories, the number of histories starting in the
+domain was computed and as the stage progressed, the number of
+histories that exited the domain were tallied with the ratio of the
+two providing a numerical measure for the leakage
+fraction. Figure~\ref{fig:measured_length} gives the domain leakage
+measurements for the domain in the center of the global grid as well
+as the analytic result computed by
+Eq~(\ref{eq:analytic_domain_leakage}) as a function of the absorption
+cross section. Again, we note good qualitative agreement between the
+measured and analytic quantities.
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=5in,clip]{measured_leakage.png}
+  \end{center}
+  \caption{\textbf{Measured and analytic domain leakage as a function
+      of the absorption cross section.} \textit{The blue data was
+      computed numerically by a domain-decomposed adjoint Neumann-Ulam
+      implementation while the red data was generated by
+      Eq~(\ref{eq:analytic_domain_leakage}).}}
+  \label{fig:measured_leakage.}
+\end{figure}
 
@@
 \section{Conclusion}
-These relationships will serve as guidelines for
-selecting an appropriate parallel algorithm strategy and provide a
-basis for performance models for future parallel implementations of
-the adjoint Neumann-Ulam method.
+These relationships will serve as guidelines for selecting an
+appropriate parallel algorithm strategy and provide a basis for
+performance models for future parallel implementations of the adjoint
+Neumann-Ulam method.
 

### 2012-11-24 doc/mandc2013/notes/spectral_theory.tex
added more data for leakage fraction and updated paper

@@
   the linear operator. In addition, a formal definition of the optical
-  thickness of a linear operator is presented based on the spectral
-  analysis. The one-speed, two-dimensional neutron diffusion equation
-  is used as a model problem to test the spectral theory for symmetric
-  operators. In general, the spectral theory shows good agreement with
-  measured computational results.
+  thickness of a domain in the decomposition is presented based on the
+  spectral analysis. The one-speed, two-dimensional neutron diffusion
+  equation is used as a model problem to test the spectral theory for
+  symmetric operators. In general, the spectral theory shows good
+  agreement with measured computational results.
 
@@
 \section{Introduction}
+An alternative approach to approximate matrix inversion is to employ
+Monte Carlo methods that sample a distribution with an expectation
+value equivalent to that of the inverted operator. Such methods have
+been in existence for decades with the earliest reference noted here
+an enjoyable manuscript published in 1950 by Forsythe and Leibler
+\cite{forsythe_matrix_1950}. In their outline, Forsythe and Liebler
+in fact credit the creation of this technique to J. Von Neumann and
+S.M. Ulam some years earlier than its publication. In 1952 Wasow
+provided a more formal explanation of Von Neumann and Ulam's method
+\cite{wasow_note_1952} and Hammersley and Handscomb's 1964 monograph
+\cite{hammersley_monte_1964} and Spanier and Gelbard's 1969 book
+\cite{spanier_monte_1969} present additional detail on a method
+adjoint to that described by Neumann and Ulam.
+
 For very large linear systems, two things are likely to occur: a
@@
 sequences have poor algorithmic strong scaling
-\cite{gropp_high-performance_2001}.  We observe this because nearest
-neighbor communication patterns scale with the volume and surface
-sizes of the local domains. The larger the volume, the more compute
-time is spent within the local volume and therefore at some point you
-are limited by the latency of the system, or the time it takes to
-receive the data to operate on. However, at a certain point, when the
-surface of those volumes becomes large enough, the amount of data that
-must be communicated grows large enough that it may overcome the
-bandwidth limitations of the system, resulting in a decrease in
-performance.
+\cite{gropp_high-performance_2001}.
 
@@
 theory to generate analytic relations for the eigenvalues of the
-operator. Using the Eigenvalue spectra, we then build relationships to
-characterize the transport of random walk histories in a decomposed
-domain and compare these analytic results to numerical experiments
-conducted with the model problem.
+operator based on system parameters. Using the Eigenvalue spectra, we
+then build relationships to characterize the transport of random walk
+histories in a decomposed domain and compare these analytic results to
+numerical experiments conducted with the model problem.
 
@@
 \end{equation}
-where $\ve{H}$ is defined as the \textit{iteration matrix}.  Using
-these definitions, we can derive an estimator from the adjoint method
-that will also give the solution vector, $\ve{x}$. We can acquire the
-adjoint solution by forming the Neumann series by writing
-Eq~(\ref{eq:adjoint_split_system}) as:
+where $\ve{H} = \ve{I} - \ve{A}$ and is defined as the
+\textit{iteration matrix}.  Using these definitions, we can derive an
+estimator from the adjoint method that will also give the solution
+vector, $\ve{x}$. We can acquire the adjoint solution by forming the
+Neumann series by writing Eq~(\ref{eq:adjoint_split_system}) as:
 \begin{equation}

### 2012-11-26 doc/mandc2013/notes/spectral_theory.tex
updates to first mandc paper

@@
   length of the adjoint random walks and the fraction of histories
-  leaking from a domain are made with respect to the Eigenvalues of
-  the linear operator. In addition, a formal definition of the optical
-  thickness of a domain in the decomposition is presented based on the
-  spectral analysis. The one-speed, two-dimensional neutron diffusion
-  equation is used as a model problem to test the spectral theory for
-  symmetric operators. In general, the spectral theory shows good
-  agreement with measured computational results.
+  leaking from a domain in the decomposition are made with respect to
+  the Eigenvalues and other properties of the linear operator. In
+  addition, a formal definition of the optical thickness of a domain
+  in the decomposition is presented based on the spectral
+  analysis. The one-speed, two-dimensional neutron diffusion equation
+  is used as a model problem to test the spectral theory for symmetric
+  operators. In general, the spectral theory shows good agreement with
+  measured computational results.
 
@@
 For problems where the linear operator is symmetric, a host of
-analytic theories exist based on the eigenvalue spectrum of the
+analytic theories exist based on the Eigenvalue spectrum of the
 operator that characterize their behavior in the context of
@@
 linear system generated by this discretization, we use a spectral
-theory to generate analytic relations for the eigenvalues of the
+theory to generate analytic relations for the Eigenvalues of the
 operator based on system parameters. Using the Eigenvalue spectra, we
@@
 $\ve{H}$ must remain less than 1 as $\ve{H}^T$ contains the same
-eigenvalues and therefore has the same spectral radius. We expand this
+Eigenvalues and therefore has the same spectral radius. We expand this
 summation to again yield a series of transitions that can be
-approximated by a Monte Carlo random walk sequence, this time forming
-the Neumann series in reverse order:
+approximated by a Monte Carlo random walk sequence, forming the
+Neumann series in reverse order:
 \begin{equation}
@@
 We can readily build an estimator for the adjoint solution from this
-series expansion, but we instead desire the direct solution. The
-adjoint estimator can be related to the solution by defining the
-following inner product equivalence \cite{spanier_monte_1969}:
+series expansion, but we instead desire the solution to
+Eq~(\ref{eq:linear_problem}). The adjoint estimator can be related to
+the solution by defining the following inner product equivalence
+\cite{spanier_monte_1969}:
 \begin{equation}
@@
 \end{equation}
-Here we have 2 unknowns, $\ve{y}$ and $\ve{d}$ and therefore we
+Here we have 2 unknowns, $\ve{y}$ and $\ve{d}$, and therefore we
 require two constraints to close the system. We use
-Eq~(\ref{eq:adjoint_vector_relation}) as the first constraint and as
-a second constraint we select:
+Eq~(\ref{eq:adjoint_vector_relation}) as the first constraint and as a
+second constraint we select:
 \begin{equation}
@@
 \end{equation}
-meaning that if we compute the inner product of the direct source and
+meaning that if we compute the inner product of the original source and
 the adjoint solution using a delta function source, we recover the
-direct solution we are looking for
+original solution we are looking for. In terms of particle transport, a
+delta function source means that this adjoint method is equivalent to
+a traditional forward method where tallies are made only in the state
+in which the random walk currently resides.
 
-In terms of particle transport, this adjoint method is equivalent to a
-traditional forward method. As a result of using the adjoint system,
-we modify our probabilities and weights using the \textit{adjoint
-  Neumann-Ulam decomposition} of $\ve{H}$:
+In order to perform the random walks required to approximate the
+summation in Eq~(\ref{eq:adjoint_neumann_solution}), probabilities and
+weights are generated using the \textit{adjoint Neumann-Ulam
+  decomposition} of $\ve{H}$:
 \begin{equation}
@@
 where now we are forming the decomposition with respect to the
-transpose of $\ve{H}$. We then follow the same procedure as the direct
-method for forming the probability and weight matrices in the
-decomposition. Using the adjoint form, probabilities should instead be
-column-scaled:
+transpose of $\ve{H}$. We then build the probability and weight
+matrices in the decomposition where probabilities are column-scaled:
 \begin{equation}
@@
 the current state and $b_{i_0}$ will be the sampled source starting
-weight. Note here that the estimator in
-Eq~(\ref{eq:adjoint_permutation_contribution}) does not have a
-dependency on the source state as in
-Eq~(\ref{eq:adjoint_permutation_contribution}), providing a remedy for
-the situation in the direct method where we must start a random walk
-in each source state for every permutation such that we may compute a
-contribution for that state. In the adjoint method, we instead tally
-in all states and those of lesser importance will not be visited as
-frequently by the random walk. Finally, the expectation value using
-all permutations is:
+weight. Finally, the expectation value using all permutations is:
 \begin{equation}
@@
 \end{equation}
-which, if expanded in the same way as the direct method, directly
-recovers the exact solution:
+which if expanded directly recovers the exact solution:
 \begin{equation}
@@
 where only an approximate solution is necessary. For the adjoint
-method, we utilize a \textit{relative weight cutoff}:
+method, we utilize a relative weight cutoff parameter:
 \begin{equation}
@@
 \end{equation}
-where $W_c$ is defined as in the direct method. The adjoint random
+where $W_c$ is the \textit{weight cutoff}. The adjoint random
 walk will then be terminated after $m$ steps if $W_m < W_f$ as tally
@@
 where $\ve{D}$ is the diffusion operator, $\ve{s}$ is the source in
-vector form and $\boldsymbol{\phi}$ is the vector of unknown fluxes. We
-note here that the diffusion operator is symmetric.
+vector form and $\boldsymbol{\phi}$ is the vector of unknown fluxes.
 
@@
 Using this coefficient, we then have the following spectrum of
-preconditioned eigenvalues:
+preconditioned Eigenvalues:
 \begin{equation}
   \lambda_{p,q}(\ve{M}^{-1} \ve{D}) = \alpha \lambda_{p,q}(\ve{D})\:.
-  \label{eq:preconditioned_eigenvalues}
+  \label{eq:preconditioned_Eigenvalues}
 \end{equation}
@@
 \cite{leveque_finite_2007}, we expect to find these minimum and
-maximum eigenvalues when $p=q$. To find these values,
-Eq~(\ref{eq:preconditioned_eigenvalues}) is plotted as a function of
+maximum Eigenvalues when $p=q$. To find these values,
+Eq~(\ref{eq:preconditioned_Eigenvalues}) is plotted as a function of
 $p$ with $p=q$ in Figure~\ref{fig:diffusion_spectrum} for various
@@
 \end{figure}
-We see a minimum eigenvalue at $p=0$ and a maximum at $p=10$, letting
+We see a minimum Eigenvalue at $p=0$ and a maximum at $p=10$, letting
 us compute the condition number for the Jacobi preconditioned
@@
 and we observe an infinite condition number as $\lambda_{min}
-\rightarrow 0$. We note also that the diffusion operator contains
-purely positive Eigenvalues and can therefore be classified as
-symmetric positive definite (SPD).
+\rightarrow 0$.
 
@@
 matrix. We use a few simplifications by noting that if the Jacobi
-preconditioned spectral radius is $\ve{H} = \ve{I} -
-\ve{M}^{-1}\ve{D}$, the we expect all terms on the diagonal of the
+preconditioned iteration matrix is $\ve{H} = \ve{I} -
+\ve{M}^{-1}\ve{D}$, the we except all terms on the diagonal of the
 iteration matrix to be zero such that we have the following stencil:
@@
 The Neumann-Ulam method for matrix inversion is effectively an
-approximation to a stationary method (the Jacobi method in the case
-of the preconditioning above). Stationary methods for linear systems arise from splitting the
-operator in Eq~(\ref{eq:linear_problem}):
+approximation to a stationary method (the Jacobi method in the case of
+the preconditioning above). Stationary methods for linear systems
+arise from splitting the operator in Eq~(\ref{eq:linear_problem}):
 \begin{equation}
@@
 \end{equation}
-where $\ve{H}=\ve{M}^{-1}\ve{N}$ is defined as the \textit{iteration
-  matrix} and $\ve{c}=\ve{M}^{-1}\ve{b}$. With the solution vector on
+where $\ve{H}=\ve{M}^{-1}\ve{N}$ is again defined as the iteration
+matrix and $\ve{c}=\ve{M}^{-1}\ve{b}$. With the solution vector on
 both the left and right hand sides, an iterative method can then be
@@
 In the adjoint Neumann-Ulam method, $k$ iterations, equivalent to $k$
-applications of the iteration matrix, are approximated by a
-random walk of length $k$ to yield the summation in
-Eq~(\ref{eq:adjoint_neumann_solution}). This random walk length, or
-the number of transitions before the termination of a history (either
-by the weight cutoff, absorption, or exiting the domain) is therefore
-bound to the number of stationary iterations required to converge to
-the specified tolerance. In the case of the adjoint Neumann-Ulam
-method, no such tolerance exists, however, we have specified a weight
-cutoff $W_c$ under which low-weight histories will be prematurely
-terminated as their contributions are deemed minute. After $k$
-iterations, a stationary method is terminated as the error has reached
-some fraction, $\epsilon$, of the initial error:
+applications of the iteration matrix, are approximated by a random
+walk of average length $k$ to yield the summation in
+Eq~(\ref{eq:adjoint_neumann_solution})
+\cite{dimov_new_1998,halton_sequential_1994,danilov_asymptotic_2000}. This
+random walk length, or the number of transitions before the
+termination of a history (either by the weight cutoff, absorption, or
+exiting the domain) is therefore bound to the number of stationary
+iterations required to converge to the specified tolerance. In the
+case of the adjoint Neumann-Ulam method, no such tolerance exists,
+however, we have specified a weight cutoff $W_c$ under which
+low-weight histories will be prematurely terminated as their
+contributions are deemed minute. After $k$ iterations, a stationary
+method is terminated as the error has reached some fraction,
+$\epsilon$, of the initial error:
 \begin{equation}
@@
 approximately characterizing the leakage behavior of many common
-geometries.  \cite{duderstadt_nuclear_1976}. Outlined more thoroughly
-by Hwang's chapter in \cite{azmy_nuclear_2010}, we begin the
-derivation of Wigner's approximation by defining $l$ to be the average
-chord length across the domain, and $\Sigma_t$ to be the total cross
-section in the domain. We then bound the behavior for the leakage
-fraction in terms of the optical thickness, $\Sigma_t l$, of the local
-domain. In the case of a small domain or a small cross section (and
-therefore small probability of absorption or scatter), $\Sigma_t l
-\rightarrow 0, \lambda \rightarrow 1$. For a large domain or large
-cross section, $\Sigma_t l \rightarrow \infty, \lambda \rightarrow
-(\Sigma_t l)^{-1}$, where $(\Sigma_t l)^{-1}$ is effectively the
-number of interactions that will occur over the length of the
-chord. Using Wigner's simple interpolation, we then have a simple
-formula for the domain leakage, $\lambda = 1/(1+\Sigma_t l)$. This
-provides a leakage estimate, valid over the entire range of $\Sigma_t
-l$, in terms of the optical thickness of the system.
+geometries \cite{duderstadt_nuclear_1976}.
+
+Outlined more thoroughly by Hwang's chapter in
+\cite{azmy_nuclear_2010}, we begin the derivation of Wigner's
+approximation by defining $l$ to be the average chord length across
+the domain, and $\Sigma_t$ to be the total cross section in the
+domain. We then bound the behavior for the leakage fraction in terms
+of the optical thickness, $\Sigma_t l$, of the local domain. In the
+case of a small domain or a small cross section (and therefore small
+probability of absorption or scatter), $\Sigma_t l \rightarrow 0,
+\lambda \rightarrow 1$. For a large domain or large cross section,
+$\Sigma_t l \rightarrow \infty, \lambda \rightarrow (\Sigma_t
+l)^{-1}$, where $(\Sigma_t l)^{-1}$ is effectively the number of
+interactions that will occur over the length of the chord. Using
+Wigner's simple interpolation, we then have a simple formula for the
+domain leakage, $\lambda = 1/(1+\Sigma_t l)$. This provides a leakage
+estimate, valid over the entire range of $\Sigma_t l$, in terms of the
+optical thickness of the system.
 
@@
 the stencil used for the discretization. We then have the following
-relation for the \textit{optical thickness}, $\tau$, of a linear
-operator equation domain:
+relation for the \textit{optical thickness}, $\tau$, of a domain in
+the decomposition:
 \begin{equation}
@@
 iteration matrix Eigenvalues, and one to measure the domain leakage
-fraction as a function of the iteration matrix Eigenvalues. Before
-doing this, we verify our spectral model for the spectral radius of
-the iteration matrix by numerically computing the largest Eigenvalue
-of the diffusion operator using an Eigenvalue solver. For this
-verification, a $100 \times 100$ square grid with $h=0.1$ and
-non-reentrant current boundary conditions was used and the absorption
-cross varied from 0 to 100 while the scattering cross section was
-fixed at unity. Figure~\ref{fig:measured_spec_rad} gives the measured
-spectral radius of the iteration matrix and the computed spectral
-radius for the preconditioned diffusion operator using
+fraction as a function of the iteration matrix Eigenvalues and the
+discretization properties. Before doing this, we verify our
+computation of the spectral radius of the iteration matrix by
+numerically computing the largest Eigenvalue of the diffusion operator
+using an iterative Eigenvalue solver. For this verification, a $100
+\times 100$ square grid with $h=0.1$ and non-reentrant current
+boundary conditions was used and the absorption cross varied from 0 to
+100 while the scattering cross section was fixed at
+unity. Figure~\ref{fig:measured_spec_rad} gives the measured spectral
+radius of the iteration matrix and the computed spectral radius for
+the preconditioned diffusion operator using
 Eq~(\ref{eq:iteration_radius}) as function of the absorption to
@@
 transition. For our grid above, the domains are of size $10 \times 10$
-cells and therefore choosing the $i$ direction there are $n_i=11$ grid
+cells and therefore choosing the $i$ direction there are $n_i=10$ grid
 points or states along the typical length of the domain. Looking at
@@
 move on average 1 state along the typical length of the domain per
-history transition, giving $n_s=1$.Figure~\ref{fig:optical_thickness})
-gives the domain optical thickness as a function of $Sigma_a$ as
+history transition, giving $n_s=1$. Figure~\ref{fig:optical_thickness}
+gives the domain optical thickness as a function of $\Sigma_a$ as
 specified by Eq~(\ref{eq:matrix_optical_thickness_2}) using the above
@@
 
-To compute the leakage fraction numerically, over the global domain,
-500,000 histories were sampled from a uniform source of unity. At the
-start of a stage of histories, the number of histories starting in the
-domain was computed and as the stage progressed, the number of
-histories that exited the domain were tallied with the ratio of the
-two providing a numerical measure for the leakage
+To compute the leakage fraction numerically, over the global domain
+\sn{1}{5} histories were sampled from a uniform source of strength
+unity. At the start of a stage of histories, the number of histories
+starting in the domain was computed and as the stage progressed, the
+number of histories that exited the domain was tallied with the ratio
+of the two numbers providing a numerical measure for the leakage
 fraction. Figure~\ref{fig:measured_length} gives the domain leakage

### 2012-11-26 doc/mandc2013/notes/spectral_theory.tex
report updates and adding data for spectral analysis

@@
 expect the maximum Eigenvalues to be found when $p=q=0$, giving the
-following for the spectral radius of the Jacobi precondition iteration
-matrix:
+following for the spectral radius of the Jacobi preconditioned
+iteration matrix:
 \begin{equation}
@@
 well as the analytic result computed by Eq~(\ref{eq:analytic_k}) as a
-function of the absorption to scattering ratio $(\Sigma_a /
-\Sigma_s)$. Again, we note here that although there is excellent
+function of the iteration matrix spectral radius,
+$\rho(\ve{H})$. Again, we note here that although there is excellent
 agreement between the measured and analytic results, slight
@@
   \caption{\textbf{Measured and analytic random walk length as a
-      function of the absorption cross section to scattering cross
-      section ratio.} \textit{The blue data was computed numerically
-      by an adjoint Neumann-Ulam implementation while the red data was
-      generated by Eq~(\ref{eq:analytic_k}).}}
+      function of the iteration matrix spectral radius.} \textit{The
+      blue data was computed numerically by an adjoint Neumann-Ulam
+      implementation while the red data was generated by
+      Eq~(\ref{eq:analytic_k}).}}
   \label{fig:measured_length}

### 2012-11-27 doc/mandc2013/notes/spectral_theory.tex
fixed optical thickness reltationships
adding additional spectral data

@@
 number of states in the $i$ (or $j$ as the problem is symmetric)
-direction that a history could potentially move and is dependent on
-the stencil used for the discretization. We then have the following
-relation for the \textit{optical thickness}, $\tau$, of a domain in
-the decomposition:
+direction that a history could potentially move in a single transition
+and is dependent on the stencil used for the discretization. We then
+have the following relation for the \textit{optical thickness},
+$\tau$, of a domain in the decomposition:
 \begin{equation}
-  \tau = \frac{n_i n_s}{k}\:,
+  \tau = \frac{n_i}{n_s k}\:,
   \label{eq:matrix_optical_thickness}
@@
 \begin{equation}
-  \tau = \frac{n_i n_s \log(\rho(\ve{H}))}{\log(W_c)}\:.
+  \tau = \frac{n_i \log(\rho(\ve{H}))}{n_s \log(W_c)}\:.
   \label{eq:matrix_optical_thickness_2}

### 2012-11-27 doc/mandc2013/notes/spectral_theory.tex
updates to data and plots for spectral analysis

@@
       operator spectral radius as a function of the absorption cross
-      section to scattering cross section ratio.} \textit{The blue
-      data was computed numerically by an Eigensolver while the red
-      data was generated by Eq~(\ref{eq:iteration_radius}).}}
+      section to scattering cross section ratio.} \textit{Values of
+      $h=1$, $h=0.1$, and $h=0.01$ were used to verify the
+      behavior. The red data was computed numerically by an
+      Eigensolver while the black dashed data was generated by
+      Eq~(\ref{eq:iteration_radius}).}}
   \label{fig:measured_spec_rad}
@@
       function of the iteration matrix spectral radius.} \textit{The
-      blue data was computed numerically by an adjoint Neumann-Ulam
-      implementation while the red data was generated by
-      Eq~(\ref{eq:analytic_k}).}}
+      weight cutoff was varied with \sn{1}{-2}, \sn{1}{-4}, and
+      \sn{1}{-8}, to verify the validity of that parameter in
+      Eq~(\ref{eq:analytic_k}). The red data was computed numerically
+      by an adjoint Neumann-Ulam implementation while the black dashed
+      data was generated by Eq~(\ref{eq:analytic_k}).}}
   \label{fig:measured_length}
@@
   \caption{\textbf{Measured and analytic domain leakage as a function
-      of the absorption cross section.} \textit{The blue data was
-      computed numerically by a domain-decomposed adjoint Neumann-Ulam
-      implementation while the red data was generated by
+      of the iteration matrix spectral radius.} \textit{To test the
+      behavior with respect to domain size, $n_i=5$, $n_i=10$,
+      $n_i=20$, and $n_i=100$ were used. The red data was computed
+      numerically by a domain-decomposed adjoint Neumann-Ulam
+      implementation while the black dashed data was generated by
       Eq~(\ref{eq:analytic_domain_leakage}).}}

### 2012-11-29 doc/mandc2013/notes/spectral_theory.tex
finished gathering spectral data

@@
 
-\subsection{Wigner Rational Approximation}
+\subsection{Domain Leakage Approximations}
 In a domain decomposed situation, not all histories will remain
@@
 quantify the number of particles that leak out of the local domain
-they define a leakage fraction, $\lambda$, as:
+they define a leakage fraction, $\Lambda$, as:
 \begin{equation}
-  \lambda = \frac{average\ \#\ of\ particles\ leaving\ local\ domain}
+  \Lambda = \frac{average\ \#\ of\ particles\ leaving\ local\ domain}
           {total\ of\ \#\ of\ particles\ starting\ in\ local\ domain}\:.
@@
 \end{equation}
-For their studies, they assumed that the value of $\lambda$ was
+For their studies, they assumed that the value of $\Lambda$ was
 dependent on the total cross section of the system via the Wigner
@@
 probability of absorption or scatter), $\Sigma_t l \rightarrow 0,
-\lambda \rightarrow 1$. For a large domain or large cross section,
-$\Sigma_t l \rightarrow \infty, \lambda \rightarrow (\Sigma_t
+\Lambda \rightarrow 1$. For a large domain or large cross section,
+$\Sigma_t l \rightarrow \infty, \Lambda \rightarrow (\Sigma_t
 l)^{-1}$, where $(\Sigma_t l)^{-1}$ is effectively the number of
@@
 Wigner's simple interpolation, we then have a simple formula for the
-domain leakage, $\lambda = 1/(1+\Sigma_t l)$. This provides a leakage
+domain leakage, $\Lambda = 1/(1+\Sigma_t l)$. This provides a leakage
 estimate, valid over the entire range of $\Sigma_t l$, in terms of the
-optical thickness of the system.
+optical thickness of the system. In addition to Wigner's
+approximation, Hwang also defines the mean-chord approximation based
+on the same numerical limits with $\Lambda = (1-e^{-\Sigma_t
+  l})/\Sigma_t l$.
 
-In the case of domain decomposed linear operator equations, the chord
-length can be interpreted as the number of discrete states spanning a
-domain, $n_i$, along the typical length of that domain (e.g. the
-number of discrete states in the $i$ direction in a square domain
-using our model diffusion problem). We then seek the equivalent of the
-total cross section, $\Sigma_t$ in terms of the discrete random walk
-behavior. The total cross section can also be interpreted as the
-inverse of the mean free path of particles in the system. In the
-Neumann-Ulam method, this is analogous to the length of the Markov
-chains (and therefore the number of state-to-state transitions)
-generated while sampling the Neumann series. In the previous section,
-we determined this length with Eq~(\ref{eq:analytic_k}) to be
-dependent on the Eigenvalues of the operator. In addition, we must
-also consider the number of states that may be traversed in a given
-transition, $n_s$, along the path of the typical length of the
-system. For our diffusion model problem, $n_s$ would equate to the
+In the case of domain decomposed linear operator equations, we can use
+diffusion theory to estimate the optical thickness of a region and the
+corresponding leakage fraction in terms of properties of the linear
+operator and the discretization. To begin we must first calculate the
+mean distance a Monte Carlo history will move in the grid by computing
+the mean squared distance of its movement along the chord defined
+across the domain of length $l$. After a single transition a history
+will have moved a mean squared distance of:
+\begin{equation}
+  \langle \bar{r_1^2} \rangle = (n_s h)^2\:,
+  \label{eq:step_1_length}
+\end{equation}
+where $h$ is the size of the grid along the chord and $n_s$ is the
+number of grid elements a history will move on average every
+transition. For our diffusion model problem, $n_s$ would equate to the
 number of states in the $i$ (or $j$ as the problem is symmetric)
 direction that a history could potentially move in a single transition
-and is dependent on the stencil used for the discretization. We then
-have the following relation for the \textit{optical thickness},
-$\tau$, of a domain in the decomposition:
+and is dependent on the stencil used for the discretization. After $k$
+transitions in the random walk, the history will have moved a mean
+squared distance of:
 \begin{equation}
-  \tau = \frac{n_i}{n_s k}\:,
-  \label{eq:matrix_optical_thickness}
+  \langle \bar{r_k^2} \rangle = k (n_s h)^2\:.
+  \label{eq:step_k_length}
 \end{equation}
-and inserting Eq~(\ref{eq:analytic_k}) gives:
+If our our chord is of length $l$ and there are $n_i$ grid points (or
+states to which a history may transition along that chord), then $h =
+l / n_i$ giving:
 \begin{equation}
-  \tau = \frac{n_i \log(\rho(\ve{H}))}{n_s \log(W_c)}\:.
-  \label{eq:matrix_optical_thickness_2}
+  \langle \bar{r_k^2} \rangle = k \Bigg(\frac{n_s l}{n_i}\Bigg)^2\:.
+  \label{eq:step_k_length_sub}
 \end{equation}
-Using this optical thickness, we can then complete the Wigner rational
-approximation by defining the bounds of $\tau \rightarrow 0, \lambda
-\rightarrow 1$ and $\tau \rightarrow \infty, \lambda \rightarrow
+From diffusion theory, we expect the average number of interactions
+along the chord to be:
+\begin{equation}
+  \tau = \frac{l}{2 d \sqrt{\langle \bar{r_k^2} \rangle}}\:,
+  \label{eq:optical_thickness_1}
+\end{equation}
+where $d$ is the dimension of the problem and $\sqrt{\langle
+  \bar{r_k^2} \rangle}$ is effectively the mean free path of the Monte
+Carlo history in the domain. We can readily interperet $\tau$ to be
+the \textit{effective optical thickness} of the domain of length
+$l$. Inserting Eq~(\ref{eq:step_k_length_sub}) we arrive at:
+\begin{equation}
+  \tau = \frac{n_i}{2 d n_s \sqrt{k}}\:,
+  \label{eq:optical_thickness_2}
+\end{equation}
+which if expanded with Eq~(\ref{eq:analytic_k}) gives us the final
+relation for the effective optical thickness:
+\begin{equation}
+  \tau = \frac{n_i}{2 d n_s}
+  \sqrt{\frac{\log(\rho(\ve{H}))}{\log(W_c)}}\:.
+  \label{eq:optical_thickness_3}
+\end{equation}
+
+Using this optical thickness, we can then complete the leakage
+approximations by defining the bounds of $\tau \rightarrow 0, \Lambda
+\rightarrow 1$ and $\tau \rightarrow \infty, \Lambda \rightarrow
 \tau^{-1}$. We expect this behavior as optically thin domains will
 communicate most of their histories while optically thick domains will
-leak the fraction of histories that did not interact within. Using
-the Wigner approximation, we can then define the leakage fraction out
-of a domain for the adjoint Neumann-Ulam method:
+leak the fraction of histories that did not interact within. With
+these bounds, we can then define the leakage fraction out of a domain
+for the adjoint Neumann-Ulam method using the Wigner rational
+approximation:
 \begin{equation}
-  \lambda = \frac{1}{1+\tau}\:.
-  \label{eq:analytic_domain_leakage}
+  \Lambda = \frac{1}{1+\tau}\:,
+  \label{eq:wigner_domain_leakage}
 \end{equation}
+and using the mean-chord approximation:
+\begin{equation}
+  \Lambda = \frac{1-e^{-\tau}}{\tau}\:.
+  \label{eq:wigner_domain_leakage}
+\end{equation}
+
 Here, the leakage fraction is explicitly bound to the Eigenvalues of
@@
 there is excellent agreement between the measured and analytic
-results, slight differences likely arise due to the non-reentrant
-current condition not being included in the spectral analysis. To
-verify this (and serve as a better testing environment), reflecting
-boundary conditions must be implemented.
+results, slight differences arise due to the non-reentrant
+current condition not being included in the spectral analysis.
 \begin{figure}[t!]
@@
 conditions was used and the absorption cross varied from 0 to 100
-while the scattering cross section was fixed at unity. A weight cutoff
-of \sn{1}{-4} was used with 10,000 histories generated by a point
-source source of strength 1 in the center of the domain such that the
-vacuum boundary conditions will have less of an effect on the length
-of the random walks. For each of the histories, the number of
-transitions made was tallied to provide an effective value of $k$ for
-each history. This value was then averaged over all histories to get a
-measured value of $k$ for the particular
-operator. Figure~\ref{fig:measured_length} gives these measurements as
-well as the analytic result computed by Eq~(\ref{eq:analytic_k}) as a
-function of the iteration matrix spectral radius,
-$\rho(\ve{H})$. Again, we note here that although there is excellent
-agreement between the measured and analytic results, slight
-differences likely arise due to the non-reentrant current condition
-not being included in the spectral analysis.
+while the scattering cross section was fixed at unity. Three weight
+cutoff values of \sn{1}{-2}, \sn{1}{-4}, and \sn{1}{-8} were used with
+10,000 histories generated by a point source source of strength 1 in
+the center of the domain such that the vacuum boundary conditions will
+have less of an effect on the length of the random walks. For each of
+the histories, the number of transitions made was tallied to provide
+an effective value of $k$ for each history. This value was then
+averaged over all histories to get a measured value of $k$ for the
+particular operator. Figure~\ref{fig:measured_length} gives these
+measurements as well as the analytic result computed by
+Eq~(\ref{eq:analytic_k}) as a function of the iteration matrix
+spectral radius, $\rho(\ve{H})$. Again, we note here that although
+there is excellent agreement between the measured and analytic
+results, slight differences arise due to the non-reentrant current
+condition not being included in the spectral analysis.
 \begin{figure}[t!]

### 2012-11-29 doc/mandc2013/notes/spectral_theory.tex
adding new data to paper

@@
 If our our chord is of length $l$ and there are $n_i$ grid points (or
-states to which a history may transition along that chord), then $h =
+states to which a history may transition) along that chord, then $h =
 l / n_i$ giving:
@@
   \Lambda = \frac{1-e^{-\tau}}{\tau}\:.
-  \label{eq:wigner_domain_leakage}
+  \label{eq:mean_chord_domain_leakage}
 \end{equation}
@@
   \begin{center}
-    \includegraphics[width=5in,clip]{measured_leakage.png}
+    \includegraphics[width=5in,clip]{leakage_variation.png}
   \end{center}
@@
       of the iteration matrix spectral radius.} \textit{To test the
-      behavior with respect to domain size, $n_i=5$, $n_i=10$,
-      $n_i=20$, and $n_i=100$ were used. The red data was computed
-      numerically by a domain-decomposed adjoint Neumann-Ulam
-      implementation while the black dashed data was generated by
-      Eq~(\ref{eq:analytic_domain_leakage}).}}
-  \label{fig:measured_leakage.}
+      behavior with respect to domain size, $n_i=50$, $n_i=100$,and
+      $n_i=200$ were used. The red data was computed numerically by a
+      domain-decomposed adjoint Neumann-Ulam implementation, the black
+      fine-dashed data was generated by
+      Eq~(\ref{eq:mean_chord_domain_leakage}), and the dashed-dotted
+      black data was generated by
+      Eq~(\ref{eq:wigner_domain_leakage}).}}
+  \label{fig:measured_leakage}
+\end{figure}
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=5in,clip]{leakage_error.png}
+  \end{center}
+  \caption{\textbf{Measured and analytic domain leakage absolute error
+      as a function of the iteration matrix spectral radius.}
+    \textit{To test the behavior with respect to domain size, $n_i=5$
+      (blue), $n_i=10$ (red), and $n_i=100$ (yellow) were used. The
+      dashed lines represent the error using the Wigner approximation
+      while the solid lines represent the error using the mean-chord
+      approximation.}}
+  \label{fig:measured_leakage}
 \end{figure}

### 2012-11-30 doc/mandc2013/notes/spectral_theory.tex
paper updates for mandc theory

@@
 \subsection{Domain Leakage Approximations}
-In a domain decomposed situation, not all histories will remain
-within the domain they started in and must instead be
-communicated. This communication, expected to be expensive, was
-analyzed by Siegel and colleagues for idealized, load balanced
-situations for full-core Monte Carlo\cite{siegel_analysis_2012}.  To
+In a domain decomposed situation, not all histories will remain within
+the domain they started in and must instead be communicated. This
+communication, expected to be expensive, was analyzed by Siegel and
+colleagues for idealized, load balanced situations for full nuclear
+reactor core Monte Carlo simulations \cite{siegel_analysis_2012}.  To
 quantify the number of particles that leak out of the local domain
@@
 In the case of domain decomposed linear operator equations, we can use
-diffusion theory to estimate the optical thickness of a region and the
-corresponding leakage fraction in terms of properties of the linear
-operator and the discretization. To begin we must first calculate the
-mean distance a Monte Carlo history will move in the grid by computing
-the mean squared distance of its movement along the chord defined
-across the domain of length $l$. After a single transition a history
-will have moved a mean squared distance of:
+diffusion theory to estimate the optical thickness of a domain in the
+decomposition and the corresponding leakage fraction in terms of
+properties of the linear operator and the discretization. To begin we
+must first calculate the mean distance a Monte Carlo history will move
+in the grid by computing the mean squared distance of its movement
+along the chord defined across the domain of length $l$. After a
+single transition a history will have moved a mean squared distance
+of:
 \begin{equation}
@@
 \end{equation}
-
 Here, the leakage fraction is explicitly bound to the Eigenvalues of
@@
 using an iterative Eigenvalue solver. For this verification, a $100
-\times 100$ square grid with $h=0.1$ and non-reentrant current
-boundary conditions was used and the absorption cross varied from 0 to
-100 while the scattering cross section was fixed at
-unity. Figure~\ref{fig:measured_spec_rad} gives the measured spectral
-radius of the iteration matrix and the computed spectral radius for
-the preconditioned diffusion operator using
+\times 100$ square grid with $h=0.01$, $h=0.1$, and $h=1.0$ and
+non-reentrant current boundary conditions was used and the absorption
+cross varied from 0 to 100 while the scattering cross section was
+fixed at unity. Figure~\ref{fig:measured_spec_rad} gives the measured
+spectral radius of the iteration matrix and the computed spectral
+radius for the preconditioned diffusion operator using
 Eq~(\ref{eq:iteration_radius}) as function of the absorption to
-scattering ratio $(\Sigma_a / \Sigma_s)$. We note here that although
-there is excellent agreement between the measured and analytic
-results, slight differences arise due to the non-reentrant
-current condition not being included in the spectral analysis.
+scattering ratio $(\Sigma_a / \Sigma_s)$. Excellent agreement was
+observed between the analytic and numerical results with all data
+points computed within the tolerance of the iterative Eigenvalue
+solver.
 \begin{figure}[t!]
@@
       section to scattering cross section ratio.} \textit{Values of
-      $h=1$, $h=0.1$, and $h=0.01$ were used to verify the
+      $h=0.01$, $h=0.1$, and $h=1.0$ were used to verify the
       behavior. The red data was computed numerically by an
@@
 Eq~(\ref{eq:analytic_k}) as a function of the iteration matrix
-spectral radius, $\rho(\ve{H})$. Again, we note here that although
-there is excellent agreement between the measured and analytic
-results, slight differences arise due to the non-reentrant current
-condition not being included in the spectral analysis.
+spectral radius, $\rho(\ve{H})$. Again, we note good agreement between
+the measured and analytic results.
 \begin{figure}[t!]
@@
 analytic relation for the optical thickness of domain and the
-associated Wigner approximation. For this experiment, we again use a
-$100 \times 100$ square grid with $h=0.1$ and non-reentrant current
-boundary conditions. The grid is then decomposed into 100 square
-domains, 10 in each cardinal direction. For the cross sections, the
-absorption cross was varied from 1 to 100 while the scattering cross
-section was set to zero to create a purely absorbing environment. The
-optical thickness of these domains will vary as a function of the
-absorption cross section if the other parameters are fixed. To compute
-the optical thickness, along with the spectral radius as given by
-Eq~(\ref{eq:iteration_radius}), we also need the parameters $n_i$ and
-$n_s$ which respectively describe the typical domain length and the
-average number of states moved along that typical length per history
-transition. For our grid above, the domains are of size $10 \times 10$
-cells and therefore choosing the $i$ direction there are $n_i=10$ grid
-points or states along the typical length of the domain. Looking at
-the Laplacian stencil in Figure~\ref{fig:stencil}, we see that all
-history transitions will only move a single state in either the $i$ or
-$j$ directions due to the symmetry of the problem. We therefore will
-move on average 1 state along the typical length of the domain per
-history transition, giving $n_s=1$. Figure~\ref{fig:optical_thickness}
-gives the domain optical thickness as a function of $\Sigma_a$ as
-specified by Eq~(\ref{eq:matrix_optical_thickness_2}) using the above
-parameters. Here we note that as the domains become more optically
-thick, this relationship becomes exponential as expected.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=5in,clip]{optical_thickness.png}
-  \end{center}
-  \caption{\textbf{Analytic domain optical thickness as a function of
-      the absorption cross section.} \textit{The data was
-      generated by Eq~(\ref{eq:matrix_optical_thickness_2}).}}
-  \label{fig:optical_thickness}
-\end{figure}
+associated leakage approximations. For this experiment, square grid
+with $h=0.1$ and non-reentrant current boundary conditions. The grid
+is then decomposed into 9 square domains, 3 in each cardinal
+direction. For the cross sections, the absorption cross was varied
+from 1 to 100 while the scattering cross section was set to zero to
+create a purely absorbing environment. The optical thickness of these
+domains will vary as a function of the absorption cross section if the
+other parameters are fixed. To compute the optical thickness, along
+with the spectral radius as given by Eq~(\ref{eq:iteration_radius}),
+we also need the parameters $n_i$ and $n_s$ which respectively
+describe the typical domain length and the average number of states
+moved along that typical length per history transition. For our grid
+above, the domains are varied in size with $50 \times 50$, $100 \times
+100$, and $200 \times 200$ cells giving $n_i=50$, $n_i=100$, and
+$n_i=200$ grid points or states along the typical length of the domain
+respectively. Looking at the Laplacian stencil in
+Figure~\ref{fig:stencil}, we see that all history transitions will
+only move a single state in either the $i$ or $j$ directions due to
+the symmetry of the problem. Furthermore, if we choose the $i$
+direction, not all states we will transition to will move the history
+in that direction. Therefore, we look to the definition of the
+iteration matrix in Eq~(\ref{eq:iteration_stencil}) and the definition
+of the adjoint probability matrix in Eq~(\ref{eq:adjoint_probability})
+to estimate the $n_s$ parameter. For a particular transition starting
+at state $(i,j)$, 6 of the 8 possible new states in the stencil move
+the history in $i$ direction with
+giving $n_i = \frac{3}{5}$.
 
@@
       fine-dashed data was generated by
-      Eq~(\ref{eq:mean_chord_domain_leakage}), and the dashed-dotted
-      black data was generated by
-      Eq~(\ref{eq:wigner_domain_leakage}).}}
+      Eq~(\ref{eq:mean_chord_domain_leakage}) using the mean-chord
+      approximation, and the dashed-dotted black data was generated by
+      Eq~(\ref{eq:wigner_domain_leakage}) using the Wigner rational
+      approximation.}}
   \label{fig:measured_leakage}
@@
       as a function of the iteration matrix spectral radius.}
-    \textit{To test the behavior with respect to domain size, $n_i=5$
-      (blue), $n_i=10$ (red), and $n_i=100$ (yellow) were used. The
-      dashed lines represent the error using the Wigner approximation
-      while the solid lines represent the error using the mean-chord
-      approximation.}}
+    \textit{To test the behavior with respect to domain size, $n_i=50$
+      (blue), $n_i=100$ (red), and $n_i=200$ (yellow) were used. The
+      dashed lines represent the error using the Wigner rational
+      approximation while the solid lines represent the error using
+      the mean-chord approximation.}}
   \label{fig:measured_leakage}

### 2012-11-30 doc/mandc2013/notes/spectral_theory.tex
adding mandc2013 template for spectral analysis paper

@@
   the Eigenvalues and other properties of the linear operator. In
-  addition, a formal definition of the optical thickness of a domain
-  in the decomposition is presented based on the spectral
-  analysis. The one-speed, two-dimensional neutron diffusion equation
-  is used as a model problem to test the spectral theory for symmetric
-  operators. In general, the spectral theory shows good agreement with
-  measured computational results.
+  addition, an estimate of the effective optical thickness of a domain
+  in the decomposition is presented based on the spectral analysis and
+  diffusion theory. Using the leakage fraction estimate, Wigner
+  rational approximation and the mean chord approximation are applied
+  to estimate the leakage fraction. The one-speed, two-dimensional
+  neutron diffusion equation is used as a model problem to test the
+  spectral theory for symmetric operators. In general, the spectral
+  theory and derived approximations show good agreement with measured
+  computational results.
 
@@
 at state $(i,j)$, 6 of the 8 possible new states in the stencil move
-the history in $i$ direction with
-giving $n_i = \frac{3}{5}$.
+the history in $i$ direction with relative coefficients of 4 for
+moving in the $(\pm i,0)$ direction and of 1 for moving in the $(\pm
+i,\pm j)$. These coefficients dictate the frequency those states are
+visited relative to the others. For those 6 states we can visit along
+the typical length their sum is 12 out of the total 20 for the
+coefficients for all possible states with their ratio giving $n_i =
+\frac{3}{5}$.
 
 To compute the leakage fraction numerically, over the global domain
-\sn{1}{5} histories were sampled from a uniform source of strength
+\sn{3}{5} histories were sampled from a uniform source of strength
 unity. At the start of a stage of histories, the number of histories
@@
 as the analytic result computed by
-Eq~(\ref{eq:analytic_domain_leakage}) as a function of the absorption
-cross section. Again, we note good qualitative agreement between the
-measured and analytic quantities.
+Eqs~(\ref{eq:wigner_domain_leakage}) and
+(\ref{eq:mean_chord_domain_leakage}) as a function of the iteration
+matrix spectral radius.
 \begin{figure}[t!]
@@
 \end{figure}
-
+ Again, we note good qualitative agreement between the measured and
+ analytic quantities but we begin to see the limits of these leakage
+ approximations. To compare the quality of the two approximations, the
+ absolute error between the computed leakage fraction and that
+ generated by the Wigner rational and mean chord approximations is
+ plotted in Figure~\ref{fig:leakage_error} for all domain sizes. From
+ these error results, the mean chord approximation is shown to have a
+ lower error for ill-conditioned systems as compared to the Wigner
+ approximation while the Wigner approximation produces less error for
+ more well-conditioned systems. In general, the mean chord
+ approximation is a better choice to estimate the leakage fraction in
+ a domain from the adjoint Neumann-Ulam method and except for a single
+ data point with $n_i=50$, the mean chord approximation yielded
+ leakage fractions within 0.05 of the measured results.
 \begin{figure}[t!]
@@
       the mean-chord approximation.}}
-  \label{fig:measured_leakage}
+  \label{fig:leakage_error}
 \end{figure}

### 2012-11-30 doc/mandc2013/theory_paper/slatterywilsonevans.tex
adding mandc2013 template for spectral analysis paper

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  This is a sample LaTeX input file for your contribution to 
%  the MC2013 conference. Modified by R.C. Martineau at INL from A. 
%  Sood at LANL, from J. Wagner ORNL who obtained the original class 
%  file by Jim Warsa, LANL, 16 July 2002}
%
%  Please use it as a template for your full paper 
%    Accompanying/related file(s) include: 
%       1. Document class/format file: mc2013.cls
%       2. Sample Postscript Figure:   figure.eps
%       3. A PDF file showing the desired appearance: template.pdf 
%    Direct questions about these files to: richard.martinea@inl.gov
%
%    Notes: 
%      (1) You can use the "dvips" utility to convert .dvi 
%          files to PostScript.  Then, use either Acrobat 
%          Distiller or "ps2pdf" to convert to PDF format. 
%      (2) Different versions of LaTeX have been observed to 
%          shift the page down, causing improper margins.
%          If this occurs, adjust the "topmargin" value in the
%          mc2013.cls file to achieve the proper margins. 
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass{mc2013}
%
%  various packages that you may wish to activate for usage 
\usepackage{graphicx}
\usepackage{tabls}
\usepackage{afterpage}
\usepackage{cites}
%\usepackage{epsf}
%
%
% Insert authors' names and short version of title in lines below
%
\newcommand{\authorHead}      % Author's names here
   {Authors' names, use et al. if more than 3}  
\newcommand{\shortTitle}      % Short title here
   {Short version of title as entered by author on web page}  
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%   BEGIN DOCUMENT
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

%
%      Headers and Footers
\afterpage{%
\fancyhf{}%
\fancyhead[CE]{              
{\scriptsize \authorHead}}                                                
\fancyhead[CO]{               
{\scriptsize \shortTitle}}                  
%\lfoot{\scriptsize{
%International Conference on Mathematics and Computational Methods
%Applied to Nuclear Science \& Engineering (M\&C 2013), 
%\\ Sun Valley, Idaho, USA, May 5-9, 2013.}}%
\rfoot{\thepage/\totalpages{}}%

\pagestyle{fancy}
%\setlength{\topmargin}{-20pt}
}

\normalsize

\setlength{\baselineskip}{16.8pt}
\vspace{-3pt}

% 
% TITLE
%

\begin{center}
\textbf{\large \\%
TITLE OF THE PAPER: CENTERED, UPPERCASE, 14 POINT TIMES
NEW ROMAN, ON SECOND LINE FROM THE TOP MARGIN,
PREFERABLY NOT MORE THAN 3 LINES LONG 
}
% 
% FIRST AUTHORS 
%
\\
\setlength{\baselineskip}{14pt}
\textbf{Author A and Author B}\footnote{Footnote, if necessary, in Times 
  New Roman font and font size 9} \\
Name of Institute  \\
Corresponding Address \\
A@institute.gov; B@institute.gov \\

% 
% SECOND AUTHORS (if not needed delete from here) 
%
\vspace{12pt}
\textbf{Double space and list Author C}\\
Department of Nuclear Engineering  \\
Name of University \\
Address \\
C@name.univ.edu\\ 
%
% SECOND AUTHORS (to here)
%

\end{center}

%
% SET RAGGED RIGHT MARGIN
%
\raggedright

\section*{ABSTRACT} 
\begin{quote}
\begin{small}
Use 8.5x11 paper size, with 1" margins on all sides.  A required 200-250 
word abstract starts on this line.  Leave two blank lines before ``ABSTRACT''
and one after.  Use 10 point Times New Roman here and single 
spacing. The abstract is a very brief summary highlighting main 
accomplishments, what is new, and how it relates to the state-of-the-art.

\emph{Key Words}: List of at most five key words
\end{small} 
\end{quote}

\setlength{\baselineskip}{14pt}
\normalsize

\Section{INTRODUCTION}

Paper starts here with two blank lines before first section title.  Use 
8.5 x 11 paper size, with 1" margins on all sides.  Double-space before and
after each subsequent section's title.  Most of the formatting is globally
set by the specifications given in the {\it mc2013.cls} file.  
First-level section titles must be all uppercase and centered, and must 
be numbered in Arabic numerals as shown above.  Unlike the Word and WordPerfect
templates, the body text and section titles for this {\LaTeX} template use 
11-point font.  A. Sood found that the 11-point font in {\LaTeX} more closely 
matched the 12-point font in Word and WordPerfect.  Introduce the topic of 
your work in this section.

The first line of a paragraph is not indented; rather double-space between 
paragraphs.  This is done automatically. If you are a {\LaTeX} expert you 
can likely define the established format characteristics in a more elegant 
manner than what is done here.  This is perfectly fine, so long as the 
established paper format/appearance is maintained.  This template is merely 
to assist with maintaining uniformity in paper appearance.

There are four types of reference styles: journal paper\cite{journal}, 
proceedings paper\cite{proc_paper}, book\cite{book}, and website\cite{website}.
References to websites are discouraged, but acceptable if absolutely necessary.
The paper will be included as a {\it .pdf} file for the proceedings. Note that 
it is the author's responsibility to review the final PDF version of the paper
to ensure proper translation into PDF.  Final PDF file size should be no more 
than 10 MB. Recommended paper length is 10-12 pages.

\Section{SECOND OR SUBSEQUENT MAJOR HEADING} 
\label{sec:first}

This is Section~\ref{sec:first}. It is followed by a subsection, that is, 
\ref{sec:second}. For this class file you must use the 
\texttt{$\backslash$Section\{\}} to denote sections (note the capital ``S'').  
The style for subsection titles and all text in this template is defined in 
the {\it mc2013.cls} file.  Make sure to avoid widow/orphan lines.

\Subsection{Subsection Title: First Character of Each Non-trivial Word is 
Uppercase} 
\label{sec:second}

Double-space before and after secondary titles is automatic.  Figures and 
tables should appear as close as possible to where they are first
cited, e.g., Fig. \ref{fig:amdahl}, in the text.  Figures are numbered in 
Arabic numerals, with the caption centered below the figure, in 
{\bf boldface}.

Triple-space before the figure, and after the figure caption.

%
\vspace{16pt}
\begin{figure}[!htb]
\begin{spacing}{1.0}
\centering
\includegraphics[scale=0.60]{./figure.eps}
\caption{\bf Speedup vs Number of Processors for Varying Parallelizable 
Fractions in Order to See a Multiple Line Caption} 
\label{fig:amdahl}
\end{spacing}
\end{figure}
\vspace{16pt}
%

When importing figures or any graphical image please verify two things:
\begin{itemize}
\item Any number, text or symbol is in Times font and is not smaller than 
    10-point after reduction to the actual window in your paper
\item That is can be translated into PDF
\end{itemize}

Equations, such as Eq. (\ref{sample_equation}), should be centered and 
sequentially numbered to the flush right of the formula.
\begin{equation}
\label{sample_equation}
Speedup\ =\ {1 \over{{f \over {p}} + (1-f)}}
\end{equation}
The continuation of a paragraph after an equation should not be indented.  
All paragraphs, as well as section or subsection headings, are separated by 
just one single empty line.

\Subsubsection{Sub-section level and lower: only first character uppercase}

See Table \ref{table:example} for a sample table.  The tabls package is
recommended for improved row and column spacing.  Notice the caption appears 
above the table by setting the \texttt{$\backslash$caption} command immediately 
after the \texttt{$\backslash$begin\{table\}}. Tables are numbered in Roman 
numerals, with the caption centered above the table, in {\bf boldface}.  
Triple-space before and after the table.

\vspace{16pt}
\begin{table}[!htb]
\centering
\caption{\bf Parallel Performance for the Sample Problem}
\label{table:example} 
\vspace{14pt}
\begin{tabular}{||r||c|c|c||} \hline \hline
 \multicolumn{1}{||c||}{Number of} &
 \multicolumn{1}{c|}{Wall-Clock} &
 \multicolumn{1}{c|}{Speedup} &
 \multicolumn{1}{c||}{Efficiency} \\
 \multicolumn{1}{||c||}{Processors} &
 \multicolumn{1}{c|}{Time$^{a}$ (min)} &
 \multicolumn{1}{c|}{(T$_{s}$/T$_{p}$)} &
 \multicolumn{1}{c||}{(\%)} \\ \hline\hline
\ 1 &  100.0 & \ ---    & ---  \\ \hline
\ 2 &   52.6 & \ 1.9    & 95.0 \\ \hline \hline
\end{tabular}
\end{table}
\vspace{16pt}

\Section{CONCLUSIONS}

Present your summary and conclusions here.

\section*{ACKNOWLEDGEMENTS}

Acknowledge the help of colleagues and sources of funding, as appropriate.

The format for this template was adapted from the template for PHYSOR 2002 
posted on the Internet.  Most of the {\LaTeX} format definitions contained
in this and the accompanying files were adapted from sample files supplied 
by R. Martineau (INL), who plagiarized (adapted) the files from A. Sood 
(LANL), J. Wagner (ORNL), and James Warsa (LANL).

%\Section*{REFERENCES}
\setlength{\baselineskip}{12pt}
\begin{thebibliography}{300}
\bibitem{journal} B. Author(s), ``Title,'' {\it Journal Name in Italic}, 
          {\bf Volume in Bold}, pp. 34-89 (19xx).
\bibitem{proc_paper} C. D. Author(s), ``Article Title,'' {\it Proceedings of
          Meeting in Italic}, Location, Dates of Meeting, Vol. n, pp. 134-156 
          (19xx).
\bibitem{book} E. F. Author, {\it Book Title in Italic}, Publisher, City \&
          Country (19xx). 
\bibitem{website} ``Spallation Neutron Source: The next-generation 
          neutron-scattering facility in the United States,'' 
          http://www.sns.gov/documentation/sns\_brochure.pdf (2002).
\end{thebibliography}

\end{document}

### 2012-11-30 doc/mandc2013/notes/spectral_theory.tex
updated mandc theory paper for section formatting. fig file not working with this template

@@
 %%---------------------------------------------------------------------------%%
-\section{Conclusion}
+\section{Conclusions}
 These relationships will serve as guidelines for selecting an

### 2012-11-30 doc/mandc2013/theory_paper/slattery_wilson_evans.tex
updated mandc theory paper for section formatting. fig file not working with this template

--- 2012-11-30 doc/mandc2013/notes/spectral_theory.tex
@@
-%        File: latexdoc.tex
-%     Created: Friday March 4 12:24:15 2011
-% Last Change: Friday March 4 12:24:20 2011
-%
-\documentclass[letterpaper,12pt]{article}
-\usepackage[top=1.0in,bottom=1.0in,left=1.25in,right=1.25in]{geometry}
-\usepackage{verbatim}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+%  This is a sample LaTeX input file for your contribution to 
+%  the MC2013 conference. Modified by R.C. Martineau at INL from A. 
+%  Sood at LANL, from J. Wagner ORNL who obtained the original class 
+%  file by Jim Warsa, LANL, 16 July 2002}
+%
+%  Please use it as a template for your full paper 
+%    Accompanying/related file(s) include: 
+%       1. Document class/format file: mc2013.cls
+%       2. Sample Postscript Figure:   figure.eps
+%       3. A PDF file showing the desired appearance: template.pdf 
+%    Direct questions about these files to: richard.martinea@inl.gov
+%
+%    Notes: 
+%      (1) You can use the "dvips" utility to convert .dvi 
+%          files to PostScript.  Then, use either Acrobat 
+%          Distiller or "ps2pdf" to convert to PDF format. 
+%      (2) Different versions of LaTeX have been observed to 
+%          shift the page down, causing improper margins.
+%          If this occurs, adjust the "topmargin" value in the
+%          mc2013.cls file to achieve the proper margins. 
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\documentclass{mc2013}
+%
+%  various packages that you may wish to activate for usage 
+\usepackage{graphicx}
+\usepackage{tabls}
+\usepackage{afterpage}
+\usepackage{cites}
 \usepackage{amssymb}
-\usepackage{graphicx}
-\usepackage{longtable}
 \usepackage{amsfonts}
@@
 \usepackage[mathcal]{euscript}
-\usepackage{tabularx}
-\usepackage{cite}
-\usepackage{c++}
 \usepackage{tmadd,tmath}
-\usepackage[usenames]{color}
-\usepackage[
-naturalnames = true, 
-colorlinks = true, 
-linkcolor = black,
-anchorcolor = black,
-citecolor = black,
-menucolor = black,
-urlcolor = blue
-]{hyperref}
-
-%%---------------------------------------------------------------------------%%
-\author{Stuart R. Slattery
-  \\ \href{mailto:sslattery@wisc.edu}{\texttt{sslattery@wisc.edu}}
+
+%\usepackage{epsf}
+%
+%
+% Insert authors' names and short version of title in lines below
+%
+\newcommand{\authorHead}      % Author's names here
+   {Stuart R. Slattery and Paul P.H. Wilson and Thomas M. Evans}  
+\newcommand{\shortTitle}      % Short title here
+   {Domain-Decomposed Neumann-Ulam Method Analysis}  
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+%   BEGIN DOCUMENT
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{document}
+
+%
+%      Headers and Footers
+\afterpage{%
+\fancyhf{}%
+\fancyhead[CE]{              
+{\scriptsize \authorHead}}                                                
+\fancyhead[CO]{               
+{\scriptsize \shortTitle}}                  
+%\lfoot{\scriptsize{
+%International Conference on Mathematics and Computational Methods
+%Applied to Nuclear Science \& Engineering (M\&C 2013), 
+%\\ Sun Valley, Idaho, USA, May 5-9, 2013.}}%
+\rfoot{\thepage/\totalpages{}}%
+
+\pagestyle{fancy}
+%\setlength{\topmargin}{-20pt}
 }
 
-\date{\today}
-\title{A Spectral Analysis of the Domain Decomposed Adjoint
-  Neumann-Ulam Method}
-\begin{document}
-\maketitle
-
-%%---------------------------------------------------------------------------%%
-\begin{abstract}
-
-  The domain decomposed behavior of the adjoint Neumann-Ulam method
-  for solving linear operator equations is analyzed using the spectral
+\normalsize
+
+\setlength{\baselineskip}{16.8pt}
+\vspace{-3pt}
+
+% 
+% TITLE
+%
+
+\begin{center}
+\textbf{\large \\% 
+  A Spectral Analysis of the Domain Decomposed
+  Adjoint Neumann-Ulam Monte Carlo Method for Linear Operator
+  Equations}
+% 
+% FIRST AUTHORS 
+%
+\\
+\setlength{\baselineskip}{14pt}
+\textbf{Stuart R. Slattery and Paul P.H. Wilson}\footnote{Please send
+  correspondence to sslattery@wisc.edu} \\
+Engineering Physics Department \\
+University of Wisconsin - Madison  \\
+1500 Engineering Dr., Madison, WI 53706 \\
+sslattery@wisc.edu; wilsonp@engr.wisc.edu \\
+
+% 
+% SECOND AUTHORS (if not needed delete from here) 
+%
+\vspace{12pt}
+\textbf{Thomas M. Evans} \\
+Radiation Transport Group  \\
+Oak Ridge National Laboratory \\
+1 Bethel Valley Road, Oak Ridge, TN 37830 \\
+evanstm@ornl.gov \\ 
+%
+% SECOND AUTHORS (to here)
+%
+
+\end{center}
+
+%
+% SET RAGGED RIGHT MARGIN
+%
+\raggedright
+
+\section*{ABSTRACT} 
+\begin{quote}
+\begin{small}
+The domain decomposed behavior of the adjoint Neumann-Ulam method for
+solving linear operator equations is analyzed using the spectral
 properties of the linear operator. Relationships for the average
@@
 in the decomposition is presented based on the spectral analysis and
-  diffusion theory. Using the leakage fraction estimate, Wigner
-  rational approximation and the mean chord approximation are applied
-  to estimate the leakage fraction. The one-speed, two-dimensional
-  neutron diffusion equation is used as a model problem to test the
-  spectral theory for symmetric operators. In general, the spectral
-  theory and derived approximations show good agreement with measured
-  computational results.
-
-\end{abstract}
+diffusion theory. Using the leakage fraction estimate, Wigner rational
+approximation and the mean chord approximation are applied to estimate
+the leakage fraction. The one-speed, two-dimensional neutron diffusion
+equation is used as a model problem to test the spectral theory for
+symmetric operators. In general, the spectral theory and derived
+approximations show good agreement with measured computational
+results.
+
+\emph{Key Words}: monte carlo domain decomposition adjoint
+\end{small} 
+\end{quote}
+
+\setlength{\baselineskip}{14pt}
+\normalsize
 
 %%---------------------------------------------------------------------------%%
-\section{Introduction}
+\Section{INTRODUCTION}
+\label{sec:intro}
+
 An alternative approach to approximate matrix inversion is to employ
@@
 
-For reactor physics Monte Carlo simulations, domain decomposition has
-been identified as a key principle in moving forward in high
-performance computing
+To enable higher fidelity reactor physics Monte Carlo simulations,
+domain decomposition has been identified as a key principle in moving
+forward in high performance computing
 \cite{brunner_comparison_2006,siegel_analysis_2012}. To accomplish
@@
 
-For problems where the linear operator is symmetric, a host of
-analytic theories exist based on the Eigenvalue spectrum of the
+When solving problems where the linear operator is symmetric, a host
+of analytic theories exist based on the Eigenvalue spectrum of the
 operator that characterize their behavior in the context of
@@
 %%---------------------------------------------------------------------------%%
-\section{The Adjoint Neumann-Ulam Method}
+\Section{THE ADJOINT NEUMANN-ULAM METHOD} 
+\label{sec:adjoint_nu}
+
 We seek solutions of the general linear problem in the following form:
@@
 %%---------------------------------------------------------------------------%%
-\section{Model Problem}
+\Section{MODEL PROBLEM}
+\label{sec:model_problem}
+
 For our numerical experiments, we choose the one-speed,
@@
 \end{multline}
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.25}{\input{stencil.pdftex_t}}
-  \end{center}
-  \caption{\textbf{Nine-point Laplacian stencil.}}
-  \label{fig:stencil}
-\end{figure}
+%% \vspace{16pt}
+%% \begin{figure}[ht!]
+%% \begin{spacing}{1.0}
+%%   \begin{center}
+%%     \scalebox{1.25}{\input{stencil.pdftex_t}}
+%%   \end{center}
+%%   \caption{\textbf{Nine-point Laplacian stencil.}}
+%%   \label{fig:stencil}
+%% \end{spacing}
+%% \end{figure}
+%% \vspace{16pt}
 We then have the following linear system to solve:
@@
 %%---------------------------------------------------------------------------%%
-\section{Spectral Analysis}
+\Section{SPECTRAL ANALYSIS}
+\label{sec:spectral_analysis}
+
 The convergence of the Neumann series in
@@
 \subsection{Diffusion Operator Spectrum}
+\label{subsec:diffusion_spectrum}
+
 For the model problem, we first compute the Eigenvalues for the
@@
 values of $\Sigma_a$.
-\begin{figure}[t!]
+\vspace{16pt}
+\begin{figure}[ht!]
+\begin{spacing}{1.0}
   \begin{center}
@@
   \label{fig:diffusion_spectrum}
+\end{spacing}
 \end{figure}
+\vspace{16pt}
 We see a minimum Eigenvalue at $p=0$ and a maximum at $p=10$, letting
@@
 \subsection{Iteration Matrix Spectrum}
+\label{subsec:iteration_spectrum}
+
 The spectral radius of the iteration matrix is obtained by seeking its
@@
     + 4 \phi_{i+1,j} + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} +
-    \phi_{i-1,j-1}\\ + \phi_{i-1,j+1} + \phi_{i+1,j-1} +
+    \phi_{i-1,j-1} + \phi_{i-1,j+1} + \phi_{i+1,j-1} +
     \phi_{i+1,j+1}]\:.
@@
 \subsection{Neumann Series Convergence}
+\label{subsec:neumann_convergence}
+
 The Neumann-Ulam method for matrix inversion is effectively an
@@
 \subsection{Domain Leakage Approximations}
+\label{subsec:domain_leak_approx}
+
 In a domain decomposed situation, not all histories will remain within
@@
 %%---------------------------------------------------------------------------%%
-\section{Numerical Experiments}
+\section{NUMERICAL EXPERIMENTS}
+\label{sec:numerical_experiments}
+
 To test the relationships developed by the spectral analysis, we form
@@
 solver.
-\begin{figure}[t!]
+\vspace{16pt}
+\begin{figure}[ht!]
+\begin{spacing}{1.0}
   \begin{center}
@@
   \label{fig:measured_spec_rad}
+\end{spacing}
 \end{figure}
+\vspace{16pt}
 
 \subsection{Random Walk Length}
+\label{subsec:walk_length}
+
 With the spectral data in hand, we can go about setting up an
@@
 the measured and analytic results.
-\begin{figure}[t!]
+\vspace{16pt}
+\begin{figure}[ht!]
+\begin{spacing}{1.0}
   \begin{center}
@@
   \label{fig:measured_length}
+\end{spacing}
 \end{figure}
+\vspace{16pt}
 
 \subsection{Domain Leakage}
+\label{subsec:domain_leakage}
+
 Finally, we seek to measure the leakage from a domain in a domain
@@
 matrix spectral radius.
-\begin{figure}[t!]
+\vspace{16pt}
+\begin{figure}[ht!]
+\begin{spacing}{1.0}
   \begin{center}
@@
   \label{fig:measured_leakage}
+\end{spacing}
 \end{figure}
+\vspace{16pt}
  Again, we note good qualitative agreement between the measured and
@@
  leakage fractions within 0.05 of the measured results.
-\begin{figure}[t!]
+\vspace{16pt}
+\begin{figure}[ht!]
+\begin{spacing}{1.0}
   \begin{center}
@@
   \label{fig:leakage_error}
+\end{spacing}
 \end{figure}
+\vspace{16pt}
 
 %%---------------------------------------------------------------------------%%
-\section{Conclusion}
+\section{CONCLUSION}
+
 These relationships will serve as guidelines for selecting an
@@
 %%---------------------------------------------------------------------------%%
-\pagebreak
+\section*{ACKNOWLEDGEMENTS}
+
+This work was performed under appointment to the Nuclear Regulatory
+Commision Fellowship program at the University of Wisconsin - Madison
+Engineering Physics Department.
+
+%%---------------------------------------------------------------------------%%
+%\Section*{REFERENCES}
+\setlength{\baselineskip}{12pt}
 \bibliographystyle{ieeetr}
 \bibliography{references}
+
+%% \begin{thebibliography}{300}
+%% \bibitem{journal} B. Author(s), ``Title,'' {\it Journal Name in Italic}, 
+%%           {\bf Volume in Bold}, pp. 34-89 (19xx).
+%% \bibitem{proc_paper} C. D. Author(s), ``Article Title,'' {\it Proceedings of
+%%           Meeting in Italic}, Location, Dates of Meeting, Vol. n, pp. 134-156 
+%%           (19xx).
+%% \bibitem{book} E. F. Author, {\it Book Title in Italic}, Publisher, City \&
+%%           Country (19xx). 
+%% \bibitem{website} ``Spallation Neutron Source: The next-generation 
+%%           neutron-scattering facility in the United States,'' 
+%%           http://www.sns.gov/documentation/sns\_brochure.pdf (2002).
+%% \end{thebibliography}
+
 \end{document}

### 2012-11-30 doc/mandc2013/theory_paper/slattery_wilson_evans.tex
mandc theory paper updates

@@
 \newcommand{\authorHead}      % Author's names here
-   {Stuart R. Slattery and Paul P.H. Wilson and Thomas M. Evans}  
+   {Slattery, Wilson, and Evans}  
 \newcommand{\shortTitle}      % Short title here
-   {Domain-Decomposed Neumann-Ulam Method Analysis}  
+   {Domain Decomposed Neumann-Ulam Method}  
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \textbf{\large \\% 
-  A Spectral Analysis of the Domain Decomposed
-  Adjoint Neumann-Ulam Monte Carlo Method for Linear Operator
-  Equations}
-% 
-% FIRST AUTHORS 
+A SPECTRAL ANALYSIS OF THE DOMAIN DECOMPOSED ADJOINT MONTE CARLO
+METHOD FOR LINEAR OPERATOR EQUATIONS}
+% 
+% First AUTHORS 
 %
@@
 \setlength{\baselineskip}{14pt}
-\textbf{Stuart R. Slattery and Paul P.H. Wilson}\footnote{Please send
-  correspondence to sslattery@wisc.edu} \\
+\textbf{S.R. Slattery and P.P.H. Wilson} \\
 Engineering Physics Department \\
@@
 \vspace{12pt}
-\textbf{Thomas M. Evans} \\
-Radiation Transport Group  \\
+\textbf{T.M. Evans} \\
 Oak Ridge National Laboratory \\
@@
 \begin{small}
-The domain decomposed behavior of the adjoint Neumann-Ulam method for
-solving linear operator equations is analyzed using the spectral
-properties of the linear operator. Relationships for the average
-length of the adjoint random walks and the fraction of histories
-leaking from a domain in the decomposition are made with respect to
-the Eigenvalues and other properties of the linear operator. In
-addition, an estimate of the effective optical thickness of a domain
-in the decomposition is presented based on the spectral analysis and
-diffusion theory. Using the leakage fraction estimate, Wigner rational
-approximation and the mean chord approximation are applied to estimate
-the leakage fraction. The one-speed, two-dimensional neutron diffusion
-equation is used as a model problem to test the spectral theory for
-symmetric operators. In general, the spectral theory and derived
-approximations show good agreement with measured computational
-results.
-
-\emph{Key Words}: monte carlo domain decomposition adjoint
+The domain decomposed behavior of the adjoint Neumann-Ulam Monte Carlo
+method for solving linear operator equations is analyzed using the
+spectral properties of the linear operator. Relationships for the
+average length of the adjoint random walks, a measure of convergence
+speed and serial performace, are made with respect to the Eigenvalues
+and other properties of the linear operator. In addition,
+relationships for the effective optical thickness of a domain in the
+decomposition are presented based on the spectral analysis and
+diffusion theory. Using the effective optical thickness, the Wigner
+rational approximation and the mean chord approximation are applied to
+estimate the leakage fraction of stochastic histories from a domain in
+the decomposition as a measure of parallel performance and potential
+communication costs. The one-speed, two-dimensional neutron diffusion
+equation is used as a model problem to test the models for symmetric
+operators. In general, the spectral theory and derived approximations
+show good agreement with measured computational results.
+
+\emph{Key Words}: Monte Carlo domain decomposition MCSA
 \end{small} 
@@
 an enjoyable manuscript published in 1950 by Forsythe and Leibler
-\cite{forsythe_matrix_1950}. In their outline, Forsythe and Liebler
-in fact credit the creation of this technique to J. Von Neumann and
+\cite{forsythe_matrix_1950}. In their outline, Forsythe and Liebler in
+fact credit the creation of this technique to J. Von Neumann and
 S.M. Ulam some years earlier than its publication. In 1952 Wasow
@@
 \cite{spanier_monte_1969} present additional detail on a method
-adjoint to that described by Neumann and Ulam.
-
-For very large linear systems, two things are likely to occur: a
-single compute node does not contain enough memory to hold the entire
-problem or a single compute node takes a prohibitively long time to
-solve the system. Domain decomposition methods permit the global
-computational domain to be subdivided into smaller regions. These
-smaller regions require less memory to store and fewer computational
-resources to solve as compared to the global problem. Typically, these
-smaller regions are then represented separately in individual compute
-nodes. Therefore, in order to solve large scale problems with the
-adjoint Neumann-Ulam method, we must explore its behavior in a domain
-decomposed environment.
+adjoint to that described by Neumann and Ulam. In recent years, the
+adjoint Neuman-Ulam method has been leveraged in Monte Carlo Synthetic
+Acceleration (MCSA) methods that are competetive with conventional
+methods including Krylov methods and are applicable to a wide variety
+of linear systems \cite{evans_monte_2009,evans_monte_2012}. To advance
+the applicability of these new MCSA methods, we therefore aim to apply
+the underlying Monte Carlo methods to large-scale problems in parallel
+computing environments.
 
@@
 sequences have poor algorithmic strong scaling
-\cite{gropp_high-performance_2001}.
+\cite{gropp_high-performance_2001}. Therefore, in order to assess the
+feasibility of solving large scale problems in parallel with the
+adjoint Neumann-Ulam method, we must explore its behavior in a domain
+decomposed environment.
 
@@
 one-speed, two-dimensional neutron diffusion equation. In this paper
-we describe the adjoint Neumann-Ulam method followed by a presentation
-and finite difference discretization of the model problem. Using the
-linear system generated by this discretization, we use a spectral
-theory to generate analytic relations for the Eigenvalues of the
-operator based on system parameters. Using the Eigenvalue spectra, we
-then build relationships to characterize the transport of random walk
-histories in a decomposed domain and compare these analytic results to
-numerical experiments conducted with the model problem.
+we describe the adjoint Neumann-Ulam Monte Carlo method followed by a
+presentation of the model problem. Using the linear system generated
+by this discretization, we use a spectral theory to generate analytic
+relations for the Eigenvalues of the operator based on system
+parameters. Using the Eigenvalue spectra, we then build relationships
+to characterize the transport of stochastic histories in a decomposed
+domain and the fraction of histories that leak from a domain and will
+therefore have to be communicated. Finally, we compare these analytic
+results to numerical experiments conducted with the model problem and
+draw conclusions looking towards future work.
 
 %%---------------------------------------------------------------------------%%
-\Section{THE ADJOINT NEUMANN-ULAM METHOD} 
+\Section{ADJOINT NEUMANN-ULAM METHOD} 
 \label{sec:adjoint_nu}
@@
 \begin{equation}
-  \ve{A} \ve{x} = \ve{b}\:,
+  \ve{A} \ve{x} = \ve{b}\:.
   \label{eq:linear_problem}
 \end{equation}
-where $\ve{A} \in \mathbb{R}^{N \times N}$ is a linear operator such
-that $\ve{A} : \mathbb{R}^{N} \rightarrow \mathbb{R}^{N}$, $\ve{x} \in
-\mathbb{R}^N$ is the solution vector, and $\ve{b} \in \mathbb{R}^N$ is
-the forcing term. Choosing the adjoint method Monte Carlo method to
-invert the linear operator, we begin by defining the linear system
-adjoint to Eq~(\ref{eq:linear_problem}):
+Choosing the adjoint method Monte Carlo method to invert the linear
+operator, we define the linear system adjoint to
+Eq~(\ref{eq:linear_problem}):
 \begin{equation}
@@
 where $\ve{y}$ and $\ve{d}$ are the adjoint solution and source
-respectively and $\ve{A}^T$ is the adjoint operator.  With this
-statement we can then define the split equation:
-\begin{equation}
-  \ve{y} = \ve{H}^T \ve{y} + \ve{d}\:,
-  \label{eq:adjoint_split_system}
-\end{equation}
-where $\ve{H} = \ve{I} - \ve{A}$ and is defined as the
-\textit{iteration matrix}.  Using these definitions, we can derive an
+respectively and $\ve{A}^T$ is the adjoint operator. We can derive an
 estimator from the adjoint method that will also give the solution
-vector, $\ve{x}$. We can acquire the adjoint solution by forming the
-Neumann series by writing Eq~(\ref{eq:adjoint_split_system}) as:
+vector, $\ve{x}$. We first rearrange
+Eq~(\ref{eq:adjoint_linear_problem}) as:
 \begin{equation}
@@
 \end{equation}
-which in turn yields the Neumann series using the adjoint operator:
+where $\ve{H} = \ve{I} - \ve{A}$ and is defined as the
+\textit{iteration matrix}.  This in turn yields the adjoint Neumann series
+using the adjoint operator:
 \begin{equation}
@@
 Eigenvalues and therefore has the same spectral radius. We expand this
-summation to again yield a series of transitions that can be
-approximated by a Monte Carlo random walk sequence, forming the
-Neumann series in reverse order:
+summation to yield a series of transitions that can be approximated by
+a Monte Carlo random walk sequence:
 \begin{equation}
@@
 \end{equation}
-We can readily build an estimator for the adjoint solution from this
-series expansion, but we instead desire the solution to
-Eq~(\ref{eq:linear_problem}). The adjoint estimator can be related to
-the solution by defining the following inner product equivalence
-\cite{spanier_monte_1969}:
+To find the solution to Eq~(\ref{eq:linear_problem}) we build an
+estimator for the adjoint solution from this series expansion. The
+adjoint estimator can be related to the solution by defining the
+following inner product equivalence \cite{spanier_monte_1969}:
 \begin{equation}
@@
 the adjoint solution using a delta function source, we recover the
-original solution we are looking for. In terms of particle transport, a
-delta function source means that this adjoint method is equivalent to
-a traditional forward method where tallies are made only in the state
-in which the random walk currently resides.
+original solution.
 
@@
 discretize the problem. For the Laplacian, we choose the 9-point
-stencil shown in Figure~\ref{fig:stencil} over a grid of size $h$
-\cite{leveque_finite_2007}:
+stencil over a grid of size $h$ \cite{leveque_finite_2007}:
 \begin{multline}
@@
 \end{multline}
-%% \vspace{16pt}
-%% \begin{figure}[ht!]
-%% \begin{spacing}{1.0}
-%%   \begin{center}
-%%     \scalebox{1.25}{\input{stencil.pdftex_t}}
-%%   \end{center}
-%%   \caption{\textbf{Nine-point Laplacian stencil.}}
-%%   \label{fig:stencil}
-%% \end{spacing}
-%% \end{figure}
-%% \vspace{16pt}
 We then have the following linear system to solve:
@@
 vector form and $\boldsymbol{\phi}$ is the vector of unknown fluxes.
-
-To close the system, a set of boundary conditions is required. In the
-case of a non-reentrant current condition applied to all global
-boundaries of the domain, we choose the formulation of Duderstadt by
-assuming the flux is zero at some ghost point beyond the
-grid. Consider for example the equations on the $i=0$ boundary of the
-domain:
-\begin{multline}
-  -\frac{1}{6h^2}[4 \phi_{-1,j} + 4 \phi_{1,j} + 4 \phi_{0,j-1} +
-    4 \phi_{0,j+1} + \phi_{-1,j-1}\\ + \phi_{-1,j+1} + \phi_{1,j-1} +
-    \phi_{1,j+1} - 20 \phi_{0,j}] + \Sigma_a \phi_{0,j} = s_{0,j}\:.
-  \label{eq:x_min_bnd}
-\end{multline}
-Here we note some terms where $i=-1$ and therefore are representative
-of grid points beyond the boundary of the domain. We set the flux at
-these points to be zero, giving a valid set of equations for the $i=0$
-boundary:
-\begin{multline}
-  -\frac{1}{6h^2}[4 \phi_{1,j} + 4 \phi_{0,j-1} + 4 \phi_{0,j+1}
-    \\ + \phi_{-1,j+1} + \phi_{1,j-1} + \phi_{1,j+1} - 20 \phi_{0,j}]
-  + \Sigma_a \phi_{0,j} = s_{0,j}\:.
-  \label{eq:x_min_bnd_2}
-\end{multline}
-We repeat this procedure for the other boundaries of the domain.
-
-For reflecting boundary conditions, the net current across a boundary
-is zero.
 
@@
 \cite{leveque_finite_2007}, we expect to find these minimum and
-maximum Eigenvalues when $p=q$. To find these values,
-Eq~(\ref{eq:preconditioned_Eigenvalues}) is plotted as a function of
-$p$ with $p=q$ in Figure~\ref{fig:diffusion_spectrum} for various
-values of $\Sigma_a$.
-\vspace{16pt}
-\begin{figure}[ht!]
-\begin{spacing}{1.0}
-  \begin{center}
-    \includegraphics[width=5in,clip]{diffusion_spectrum.png}
-  \end{center}
-  \caption{\textbf{Eigenvalue spectra for the diffusion equation.}}
-  \label{fig:diffusion_spectrum}
-\end{spacing}
-\end{figure}
-\vspace{16pt}
-We see a minimum Eigenvalue at $p=0$ and a maximum at $p=10$, letting
-us compute the condition number for the Jacobi preconditioned
-diffusion operator:
+maximum Eigenvalues when $p=q$. We find a minimum Eigenvalue at $p=0$
+and a maximum at $p=10$, letting us compute the condition number for
+the Jacobi preconditioned diffusion operator:
 \begin{equation}
@@
 \end{equation}
-Visually we can also see how increasing the absorption in the problem
-creates a more well conditioned system as less diffusion is
-expected. If we have a problem with only scattering, $\Sigma_a = 0$
-and we observe an infinite condition number as $\lambda_{min}
-\rightarrow 0$.
 
@@
 The Neumann-Ulam method for matrix inversion is effectively an
-approximation to a stationary method (the Jacobi method in the case of
-the preconditioning above). Stationary methods for linear systems
-arise from splitting the operator in Eq~(\ref{eq:linear_problem}):
-\begin{equation}
-  \ve{A} = \ve{M} - \ve{N}\:,
-  \label{eq:split_linear_operator}
-\end{equation}
-where the choice of $\ve{M}$ and $\ve{N}$ will be dictated by the
-particular method chosen. Using this split definition of the operator
-we can then write:
-\begin{equation}
-  \ve{M}\ve{x} - \ve{N}\ve{x} = \ve{b}\:.
-  \label{eq:linear_split_equation1}
-\end{equation}
-By rearranging, we can generate a form more useful for analysis:
-\begin{equation}
-  \ve{x} = \ve{H}\ve{x} + \ve{c}\:,
-  \label{eq:linear_split_equation2}
-\end{equation}
-where $\ve{H}=\ve{M}^{-1}\ve{N}$ is again defined as the iteration
-matrix and $\ve{c}=\ve{M}^{-1}\ve{b}$. With the solution vector on
-both the left and right hand sides, an iterative method can then be
-formed:
+approximation to a stationary method. Stationary methods for linear
+systems arise from splitting the operator in
+Eq~(\ref{eq:linear_problem}) and iterating:
 \begin{equation}
@@
 with $k \in \mathbb{Z}^+$ defined as the \textit{iteration
-  index}. Given this, we can then generate a few statements regarding
-the convergence of such stationary methods. Defining $\ve{e}^k =
-\ve{u}^k - \ve{u}$ as the solution error at the $k^{th}$ iterate, we
-can subtract Eq~(\ref{eq:linear_split_equation2}) from
-Eq~(\ref{eq:linear_iterative_method}) to arrive at an error form of
-the linear problem:
-\begin{equation}
-  \ve{e}^{k+1} = \ve{H}\ve{e}^k\:. 
-  \label{eq:linear_iterative_error}
-\end{equation}
-Our error after $k$ iterations is then:
+  index}. Defining $\ve{e}^k = \ve{u}^k - \ve{u}$ as the solution
+error at the $k^{th}$ iterate, the error after $k$ iterations is then:
 \begin{equation}
@@
 \end{equation}
-In other words, successive application of the iteration matrix is the
-mechanism driving down the error in a stationary method. We can then
-place restrictions on the iteration matrix by assuming $\ve{H}$ is
-diagonalizable \cite{saad_iterative_2003}, we then have:
-\begin{equation}
-  \ve{e}^{k} =
-  \ve{R}\boldsymbol{\Lambda}^k\ve{R}^{-1}\ve{e}^0\:,
-  \label{eq:linear_k_iter_error_diag}
-\end{equation}
-where $\boldsymbol{\Lambda}$ contains the Eigenvalues of $\ve{H}$ on
-its diagonal and the columns of $\ve{R}$ contain the Eigenvectors of
-$\ve{H}$. Computing the 2-norm of the above form then gives:
-\begin{equation}
-  ||\ve{e}^{k}||_2 \leq ||\boldsymbol{\Lambda}^k||_2\ 
-  ||\ve{R}||_2\ ||\ve{R}^{-1}||_2\ ||\ve{e}^0||_2\:,
-  \label{eq:linear_k_iter_norm1}
-\end{equation}
-which gives:
-\begin{equation}
-  ||\ve{e}^{k}||_2 \leq \rho(\ve{H})^k \kappa(\ve{R})
-  ||\ve{e}^0||_2\:.
-  \label{eq:linear_k_iter_norm2}
-\end{equation}
-For iteration matrices where the Eigenvectors are orthogonal,
-$\kappa(\ve{R})=1$ and the error bound reduces to:
+By assuming $\ve{H}$ is diagonalizable \cite{leveque_finite_2007}, we
+then have:
 \begin{equation}
@@
 termination of a history (either by the weight cutoff, absorption, or
-exiting the domain) is therefore bound to the number of stationary
-iterations required to converge to the specified tolerance. In the
-case of the adjoint Neumann-Ulam method, no such tolerance exists,
-however, we have specified a weight cutoff $W_c$ under which
-low-weight histories will be prematurely terminated as their
-contributions are deemed minute. After $k$ iterations, a stationary
-method is terminated as the error has reached some fraction,
-$\epsilon$, of the initial error:
+exiting the domain) is therefore approximately the number of
+stationary iterations required to converge to the specified
+tolerance. In the case of the adjoint Neumann-Ulam method, no such
+tolerance exists, however, we have specified a weight cutoff $W_c$
+under which low-weight histories will be prematurely terminated as
+their contributions are deemed minute. After $k$ iterations, a
+stationary method is terminated as the error has reached some
+fraction, $\epsilon$, of the initial error:
 \begin{equation}
@@
 operator splitting chosen) and based on the weight cutoff parameter
-used in the Neumann-Ulam method
+used in the Neumann-Ulam method.
 
@@
 dependent on the total cross section of the system via the Wigner
-rational approximation. This approximation was developed by Wigner to
-compute the neutron leakage probability from the fuel in a reactor as
-a simple interpolation between the small and large fuel lump cases,
-approximately characterizing the leakage behavior of many common
-geometries \cite{duderstadt_nuclear_1976}.
-
-Outlined more thoroughly by Hwang's chapter in
-\cite{azmy_nuclear_2010}, we begin the derivation of Wigner's
-approximation by defining $l$ to be the average chord length across
-the domain, and $\Sigma_t$ to be the total cross section in the
-domain. We then bound the behavior for the leakage fraction in terms
-of the optical thickness, $\Sigma_t l$, of the local domain. In the
-case of a small domain or a small cross section (and therefore small
-probability of absorption or scatter), $\Sigma_t l \rightarrow 0,
-\Lambda \rightarrow 1$. For a large domain or large cross section,
-$\Sigma_t l \rightarrow \infty, \Lambda \rightarrow (\Sigma_t
-l)^{-1}$, where $(\Sigma_t l)^{-1}$ is effectively the number of
-interactions that will occur over the length of the chord. Using
-Wigner's simple interpolation, we then have a simple formula for the
-domain leakage, $\Lambda = 1/(1+\Sigma_t l)$. This provides a leakage
-estimate, valid over the entire range of $\Sigma_t l$, in terms of the
-optical thickness of the system. In addition to Wigner's
-approximation, Hwang also defines the mean-chord approximation based
-on the same numerical limits with $\Lambda = (1-e^{-\Sigma_t
-  l})/\Sigma_t l$.
+rational approximation. Outlined more thoroughly by Hwang's chapter in
+\cite{azmy_nuclear_2010}, we will use both the Wigner rational
+approximation and the mean chord approximation as a means to estimate
+the leakage fraction.
 
@@
   \bar{r_k^2} \rangle}$ is effectively the mean free path of the Monte
-Carlo history in the domain. We can readily interperet $\tau$ to be
-the \textit{effective optical thickness} of the domain of length
+Carlo history in the domain. We can readily interpret $\tau$ to be the
+\textit{effective optical thickness} of the domain of length
 $l$. Inserting Eq~(\ref{eq:step_k_length_sub}) we arrive at:
@@
 leak the fraction of histories that did not interact within. With
-these bounds, we can then define the leakage fraction out of a domain
+these bounds we can then define the leakage fraction out of a domain
 for the adjoint Neumann-Ulam method using the Wigner rational
@@
 low weight histories.  We also note here an interesting consequence of
-this approximation: no matter how optically thick a domain may be, the
+this approximation; no matter how optically thick a domain may be, the
 domain is still expected to leak some histories and thus communicate
@@
 not encapsulate the entire global domain they require some
-communication in order to implement a domain decomposed adjoint
-Neumann-Ulam method.
+communication in order to implement a domain decomposed Neumann-Ulam
+method.
 
@@
 solver.
-\vspace{16pt}
 \begin{figure}[ht!]
@@
 \end{figure}
-\vspace{16pt}
 
@@
 the measured and analytic results.
-\vspace{16pt}
 \begin{figure}[ht!]
@@
 respectively. Looking at the Laplacian stencil in
-Figure~\ref{fig:stencil}, we see that all history transitions will
-only move a single state in either the $i$ or $j$ directions due to
-the symmetry of the problem. Furthermore, if we choose the $i$
+Eq~(\ref{eq:nine_point_stencil}), we see that all history transitions
+will only move a single state in either the $i$ or $j$ directions due
+to the symmetry of the problem. Furthermore, if we choose the $i$
 direction, not all states we will transition to will move the history
@@
 
-To compute the leakage fraction numerically, over the global domain
-\sn{3}{5} histories were sampled from a uniform source of strength
-unity. At the start of a stage of histories, the number of histories
+To compute the leakage fraction numerically, \sn{3}{5} histories were
+sampled from a uniform source of strength unity over the global
+domain. At the start of a stage of histories, the number of histories
 starting in the domain was computed and as the stage progressed, the
@@
 matrix spectral radius.
-\vspace{16pt}
 \begin{figure}[ht!]
@@
 \end{figure}
-\vspace{16pt}
 Again, we note good qualitative agreement between the measured and
- analytic quantities but we begin to see the limits of these leakage
+analytic quantities but we begin to see the limits of the leakage
 approximations. To compare the quality of the two approximations, the
@@
 generated by the Wigner rational and mean chord approximations is
- plotted in Figure~\ref{fig:leakage_error} for all domain sizes. From
- these error results, the mean chord approximation is shown to have a
- lower error for ill-conditioned systems as compared to the Wigner
- approximation while the Wigner approximation produces less error for
- more well-conditioned systems. In general, the mean chord
- approximation is a better choice to estimate the leakage fraction in
- a domain from the adjoint Neumann-Ulam method and except for a single
- data point with $n_i=50$, the mean chord approximation yielded
- leakage fractions within 0.05 of the measured results.
-\vspace{16pt}
+plotted in Figure~\ref{fig:leakage_error} for all domain sizes
+tested. 
 \begin{figure}[ht!]
@@
 \end{figure}
-\vspace{16pt}
+From these error results, the mean chord approximation is
+shown to have a lower error for ill-conditioned systems as compared
+to the Wigner approximation while the Wigner approximation produces
+less error for more well-conditioned systems. In general, the mean
+chord approximation is a better choice to estimate the leakage
+fraction in a domain from the adjoint Neumann-Ulam method and except
+for a single data point with $n_i=50$, the mean chord approximation
+yielded leakage fractions within 0.05 of the measured results.
 
 %%---------------------------------------------------------------------------%%
-\section{CONCLUSION}
+\section{CONCLUSIONS}
 

### 2012-11-30 doc/mandc2013/theory_paper/slattery_wilson_evans.tex
updating figures and adding gnumeric data file

@@
 \end{equation}
-For their studies, they assumed that the value of $\Lambda$ was
-dependent on the total cross section of the system via the Wigner
-rational approximation. Outlined more thoroughly by Hwang's chapter in
-\cite{azmy_nuclear_2010}, we will use both the Wigner rational
-approximation and the mean chord approximation as a means to estimate
-the leakage fraction.
+For their studies, Siegel and colleagues assumed that the value of
+$\Lambda$ was dependent on the total cross section of the system via
+the Wigner rational approximation. Outlined more thoroughly by Hwang's
+chapter in \cite{azmy_nuclear_2010}, we will use both the Wigner
+rational approximation and the mean chord approximation as a means to
+estimate the leakage fraction.
 
@@
     \end{center}
-    \caption{\textbf{Measured and analytic domain leakage as a function
-        of the iteration matrix spectral radius.} \textit{To test the
-        behavior with respect to domain size, $n_i=50$, $n_i=100$,and
-        $n_i=200$ were used. The red data was computed numerically by a
-        domain-decomposed adjoint Neumann-Ulam implementation, the black
-        fine-dashed data was generated by
+    \caption{\textbf{Measured and analytic domain leakage as a
+        function of the iteration matrix spectral radius.} \textit{To
+        test the behavior with respect to domain size, $n_i=50$,
+        $n_i=100$,and $n_i=200$ were used. The red data was computed
+        numerically by a domain-decomposed adjoint Neumann-Ulam
+        implementation, the black dashed data was generated by
         Eq~(\ref{eq:mean_chord_domain_leakage}) using the mean-chord
-        approximation, and the dashed-dotted black data was generated by
-        Eq~(\ref{eq:wigner_domain_leakage}) using the Wigner rational
-        approximation.}}
+        approximation, and the dashed-dotted black data was generated
+        by Eq~(\ref{eq:wigner_domain_leakage}) using the Wigner
+        rational approximation.}}
     \label{fig:measured_leakage}
@@
       \textit{To test the behavior with respect to domain size, $n_i=50$
-        (blue), $n_i=100$ (red), and $n_i=200$ (yellow) were used. The
+        (green), $n_i=100$ (blue), and $n_i=200$ (red) were used. The
         dashed lines represent the error using the Wigner rational

### 2012-11-30 doc/mandc2013/theory_paper/slattery_wilson_evans.tex
added random walk length relative error plot

@@
 equation is used as a model problem to test the models for symmetric
-operators. In general, the spectral theory and derived approximations
-show good agreement with measured computational results.
+operators. In general, the derived approximations show good agreement
+with measured computational results.
 
@@
 to characterize this communication not only because communication is
-in general expensive, but also because nearest-neighbor communication
-sequences have poor algorithmic strong scaling
-\cite{gropp_high-performance_2001}. Therefore, in order to assess the
-feasibility of solving large scale problems in parallel with the
-adjoint Neumann-Ulam method, we must explore its behavior in a domain
-decomposed environment.
+in general expensive, but also because these nearest-neighbor
+communication sequences have poor algorithmic strong scaling
+\cite{gropp_high-performance_2001}.
 
-When solving problems where the linear operator is symmetric, a host
-of analytic theories exist based on the Eigenvalue spectrum of the
-operator that characterize their behavior in the context of
-deterministic linear solvers. Using past work, these theories are
-adapted to the domain decomposed adjoint Neumann-Ulam method using the
-one-speed, two-dimensional neutron diffusion equation. In this paper
-we describe the adjoint Neumann-Ulam Monte Carlo method followed by a
-presentation of the model problem. Using the linear system generated
-by this discretization, we use a spectral theory to generate analytic
-relations for the Eigenvalues of the operator based on system
-parameters. Using the Eigenvalue spectra, we then build relationships
-to characterize the transport of stochastic histories in a decomposed
-domain and the fraction of histories that leak from a domain and will
-therefore have to be communicated. Finally, we compare these analytic
-results to numerical experiments conducted with the model problem and
-draw conclusions looking towards future work.
+The purpose of this study is to provide a simple, analytic theory
+based on the properties of the linear system the will allow for
+estimates of the domain decomposed behavior of the adjoint
+Neumann-Ulam method. When solving problems where the linear operator
+is symmetric, a host of analytic theories exist based on the
+Eigenvalue spectrum of the operator that characterize their behavior
+in the context of deterministic linear solvers. Using past work, these
+theories are adapted to the domain decomposed adjoint Neumann-Ulam
+method using the one-speed, two-dimensional neutron diffusion
+equation. In this paper we describe the adjoint Neumann-Ulam Monte
+Carlo method followed by a presentation of the model problem. Using
+the linear system generated by this discretization, we use a spectral
+theory to generate analytic relations for the Eigenvalues of the
+operator based on system parameters. Using the Eigenvalue spectra, we
+then build relationships to characterize the transport of stochastic
+histories in a decomposed domain and the fraction of histories that
+leak from a domain and will therefore have to be
+communicated. Finally, we compare these analytic results to numerical
+experiments conducted with the model problem and draw conclusions
+looking towards future work.
 
@@
 where $\ve{y}$ and $\ve{d}$ are the adjoint solution and source
-respectively and $\ve{A}^T$ is the adjoint operator. We can derive an
-estimator from the adjoint method that will also give the solution
-vector, $\ve{x}$. We first rearrange
+respectively and $\ve{A}^T$ is the adjoint operator. We can derive a
+Monte Carlo estimator from the adjoint method that will also give the
+solution vector, $\ve{x}$. We first rearrange
 Eq~(\ref{eq:adjoint_linear_problem}) as:
@@
 \end{equation}
-where $\ve{H} = \ve{I} - \ve{A}$ and is defined as the
-\textit{iteration matrix}.  This in turn yields the adjoint Neumann series
-using the adjoint operator:
+where $\ve{H} = \ve{I} - \ve{A}$ is the \textit{iteration matrix}.
+This yields the adjoint Neumann series:
 \begin{equation}
@@
 In order to perform the random walks required to approximate the
-summation in Eq~(\ref{eq:adjoint_neumann_solution}), probabilities and
-weights are generated using the \textit{adjoint Neumann-Ulam
-  decomposition} of $\ve{H}$:
+summation in Eq~(\ref{eq:adjoint_neumann_solution}), we build
+transition probability and weight matrices where probabilities are
+column-scaled and the transition weight is defined as:
 \begin{equation}
-  \ve{H}^{T} = \ve{P} \circ \ve{W}\:,
-  \label{eq:adjoint_neumann_ulam}
-\end{equation}
-where now we are forming the decomposition with respect to the
-transpose of $\ve{H}$. We then build the probability and weight
-matrices in the decomposition where probabilities are column-scaled:
-\begin{equation}
-  p_{ij} = \frac{|h_{ji}|}{\sum_j |h_{ji}|}\:,
+  p_{ij} = \frac{|h_{ji}|}{\sum_j |h_{ji}|}\:,\ w_{ij} =
+  \frac{h_{ji}}{p_{ij}},
   \label{eq:adjoint_probability}
@@
 such that we expect to select a new state $j$ from the current state
-in the random walk $j$ by sampling column-wise. Per
-Eq~(\ref{eq:adjoint_neumann_ulam}), the transition weight is then
-defined as:
-\begin{equation}
-  w_{ij} = \frac{h_{ji}}{p_{ij}}\:.
-  \label{eq:adjoint_weight}
-\end{equation}
-Using the decomposition we can then define an expectation value for
-the adjoint method. Using our result from
+in the random walk $j$ by sampling column-wise.  Using our result from
 Eq~(\ref{eq:inner_product_constraint}) generated by applying the
@@
 
-\subsection{Diffusion Operator Spectrum}
-\label{subsec:diffusion_spectrum}
+\subsection{Iteration Matrix Spectrum}
+\label{subsec:iteration_spectrum}
 
@@
 operator with each row scaled by the diagonal component. As we have
-defined a homogeneous domain, the scaling factor is the same for all
-rows in the operator. This scaling factor, $\alpha$, is then defined
-as the $\phi_{i,j}$ coefficient from Eq~(\ref{eq:fd_system}):
+defined a homogeneous domain, the scaling factor, $\alpha$, is the
+same for all rows in the operator and defined as the $\phi_{i,j}$
+coefficient from Eq~(\ref{eq:fd_system}):
 \begin{equation}
@@
 \end{equation}
-
-A good measure of how well-conditioned is a linear system can be
-obtained by computing the condition number for the linear operator
-defined as:
-\begin{equation}
-  \kappa(\ve{A}) = ||\ve{A}||\ ||\ve{A}^{-1}||\:,
-  \label{eq:condition_number}
-\end{equation}
-which can be related to the Eigenvalues of the operator as:
-\begin{equation}
-  \kappa(\ve{A}) =
-  \frac{\max_{p,q}\lambda_{p,q}(\ve{A})}{\min_{p,q}\lambda_{p,q}(\ve{A})}\:.
-  \label{eq:condition_number_2}
-\end{equation}
-To compute the condition number for the preconditioned diffusion
-system, we must therefore compute the minimum and maximum Eigenvalues
-of the diffusion operator. Per Leveque's text
-\cite{leveque_finite_2007}, we expect to find these minimum and
-maximum Eigenvalues when $p=q$. We find a minimum Eigenvalue at $p=0$
-and a maximum at $p=10$, letting us compute the condition number for
-the Jacobi preconditioned diffusion operator:
-\begin{equation}
-  \kappa(\ve{M}^{-1} \ve{D}) =
-  \frac{\frac{D}{6h^2}\Sigma_a}{\frac{D}{3h^2}[8 \cos(10 \pi h) + 2
-      \cos^2(10 \pi h) - 10] + \Sigma_a}\:.
-  \label{eq:diffusion_condition}
-\end{equation}
-
-\subsection{Iteration Matrix Spectrum}
-\label{subsec:iteration_spectrum}
 
@@
 \end{equation}
-Inserting the Eigenfunctions defined by Eq~\ref{eq:eigenfunction_form}
-we get:
+Inserting the Eigenfunctions defined by
+Eq~(\ref{eq:eigenfunction_form}) we get:
 \begin{multline}
@@
 giving the Eigenvalue spectrum for the Jacobi preconditioned iteration
-matrix. As with the linear operator, we expect the maximum Eigenvalue
-to exist when $p=q$. In addition, we note that we have the same cosine
-form of the spectrum as for the linear operator expect that the
-spectra will be shifted by 10 in the $-p$ direction. We therefore
-expect the maximum Eigenvalues to be found when $p=q=0$, giving the
-following for the spectral radius of the Jacobi preconditioned
+matrix. We expect the maximum Eigenvalue to exist when $p=q=0$, giving
+the following for the spectral radius of the Jacobi preconditioned
 iteration matrix:
@@
 
-The Neumann-Ulam method for matrix inversion is effectively an
-approximation to a stationary method. Stationary methods for linear
-systems arise from splitting the operator in
-Eq~(\ref{eq:linear_problem}) and iterating:
+The adjoint Monte Carlo method is effectively an approximation to a
+stationary method. Stationary methods for linear systems arise from
+splitting the operator in Eq~(\ref{eq:linear_problem}) and iterating:
 \begin{equation}
@@
     \begin{center}
-      \includegraphics[width=5in,clip]{measured_spec_rad.png}
+      \includegraphics[width=4in,clip]{measured_spec_rad.png}
     \end{center}
@@
         section to scattering cross section ratio.} \textit{Values of
-        $h=0.01$, $h=0.1$, and $h=1.0$ were used to verify the
-        behavior. The red data was computed numerically by an
-        Eigensolver while the black dashed data was generated by
-        Eq~(\ref{eq:iteration_radius}).}}
+        $h=0.01$, $h=0.1$, and $h=1.0$ were used. The red data was
+        computed numerically by an Eigensolver while the black dashed
+        data was generated by Eq~(\ref{eq:iteration_radius}).}}
     \label{fig:measured_spec_rad}
@@
 averaged over all histories to get a measured value of $k$ for the
-particular operator. Figure~\ref{fig:measured_length} gives these
-measurements as well as the analytic result computed by
+particular operator. On the left, figure~\ref{fig:measured_length}
+these measurements as well as the analytic result computed by
 Eq~(\ref{eq:analytic_k}) as a function of the iteration matrix
-spectral radius, $\rho(\ve{H})$. Again, we note good agreement between
-the measured and analytic results.
+spectral radius, $\rho(\ve{H})$. On the right,
+figure~\ref{fig:measured_length} gives the relative error between the
+predicted and observed results. We note good qualitative agreement
+between the measured and analytic results. However, we obeserve a
+larger relative error for both long and short random walks.
 \begin{figure}[ht!]
@@
     \begin{center}
-      \includegraphics[width=5in,clip]{measured_length.png}
+      \includegraphics[width=6.5in,clip]{measured_length.png}
     \end{center}
@@
         weight cutoff was varied with \sn{1}{-2}, \sn{1}{-4}, and
-        \sn{1}{-8}, to verify the validity of that parameter in
-        Eq~(\ref{eq:analytic_k}). The red data was computed numerically
-        by an adjoint Neumann-Ulam implementation while the black dashed
-        data was generated by Eq~(\ref{eq:analytic_k}).}}
+        \sn{1}{-8}. In the left plot, the red data was computed
+        numerically by an adjoint Neumann-Ulam implementation while
+        the black dashed data was generated by
+        Eq~(\ref{eq:analytic_k}). In the right plot, the relative
+        error between the predicted and measured results is
+        presented.}}
     \label{fig:measured_length}
@@
 $n_i=200$ grid points or states along the typical length of the domain
-respectively. Looking at the Laplacian stencil in
-Eq~(\ref{eq:nine_point_stencil}), we see that all history transitions
-will only move a single state in either the $i$ or $j$ directions due
-to the symmetry of the problem. Furthermore, if we choose the $i$
-direction, not all states we will transition to will move the history
-in that direction. Therefore, we look to the definition of the
-iteration matrix in Eq~(\ref{eq:iteration_stencil}) and the definition
-of the adjoint probability matrix in Eq~(\ref{eq:adjoint_probability})
-to estimate the $n_s$ parameter. For a particular transition starting
-at state $(i,j)$, 6 of the 8 possible new states in the stencil move
-the history in $i$ direction with relative coefficients of 4 for
-moving in the $(\pm i,0)$ direction and of 1 for moving in the $(\pm
-i,\pm j)$. These coefficients dictate the frequency those states are
-visited relative to the others. For those 6 states we can visit along
-the typical length their sum is 12 out of the total 20 for the
-coefficients for all possible states with their ratio giving $n_i =
-\frac{3}{5}$.
+respectively with weight cutoff of \sn{1}{-4}. Looking at the
+Laplacian stencil in Eq~(\ref{eq:nine_point_stencil}), we see that all
+history transitions will only move a single state in either the $i$ or
+$j$ directions due to the symmetry of the problem. Furthermore, if we
+choose the $i$ direction, not all states we will transition to will
+move the history in that direction. Therefore, we look to the
+definition of the iteration matrix in Eq~(\ref{eq:iteration_stencil})
+and the definition of the adjoint probability matrix in
+Eq~(\ref{eq:adjoint_probability}) to estimate the $n_s$ parameter. For
+a particular transition starting at state $(i,j)$, 6 of the 8 possible
+new states in the stencil move the history in $i$ direction with
+relative coefficients of 4 for moving in the $(\pm i,0)$ direction and
+of 1 for moving in the $(\pm i,\pm j)$. These coefficients dictate the
+frequency those states are visited relative to the others. For those 6
+states we can visit along the typical length their sum is 12 out of
+the total 20 for the coefficients for all possible states with their
+ratio giving $n_i = \frac{3}{5}$.
 
@@
 of the two numbers providing a numerical measure for the leakage
-fraction. Figure~\ref{fig:measured_length} gives the domain leakage
+fraction. Figure~\ref{fig:measured_leakage} gives the domain leakage
 measurements for the domain in the center of the global grid as well
@@
     \begin{center}
-      \includegraphics[width=5in,clip]{leakage_variation.png}
+      \includegraphics[width=4.5in,clip]{leakage_variation.png}
     \end{center}
@@
     \begin{center}
-      \includegraphics[width=5in,clip]{leakage_error.png}
+      \includegraphics[width=4.5in,clip]{leakage_error.png}
     \end{center}
@@
 \end{figure}
-From these error results, the mean chord approximation is
-shown to have a lower error for ill-conditioned systems as compared
-to the Wigner approximation while the Wigner approximation produces
-less error for more well-conditioned systems. In general, the mean
-chord approximation is a better choice to estimate the leakage
-fraction in a domain from the adjoint Neumann-Ulam method and except
-for a single data point with $n_i=50$, the mean chord approximation
-yielded leakage fractions within 0.05 of the measured results.
+From these error results, the mean chord approximation is shown to
+have a lower error for ill-conditioned systems as compared to the
+Wigner approximation while the Wigner approximation produces less
+error for more well-conditioned systems. We also note that for the
+optically thick domains, the error is likely corresponded to that
+observed in Figure~\ref{fig:measured_length} for the $k$ parameter
+while the large relative error in $k$ for optically thin domains does
+not affect the approximation. In general, the mean chord approximation
+is a better choice to estimate the leakage fraction in a domain from
+the adjoint Neumann-Ulam method and except for a single data point
+with $n_i=50$, the mean chord approximation yielded leakage fractions
+within 0.05 of the measured results. As the domain becomes more
+optically thick (with both increasing $n_i$ and decreasing
+$\rho(\ve{H})$), the approximations are more accurate.
 
@@
 
-These relationships will serve as guidelines for selecting an
-appropriate parallel algorithm strategy and provide a basis for
-performance models for future parallel implementations of the adjoint
-Neumann-Ulam method.
+We have presented an analytic analysis of the domain decomposed
+behavior of the adjoint Neumann Ulam method for linear operator
+equations. Good agreement was observed between the derived analytic
+relationships and numerical results generated by a domain decomposed
+implementation of the adjoint method.
+
+In future work, these relationships will serve as guidelines for
+selecting an appropriate parallel algorithm strategy and provide a
+basis for performance models for future parallel implementations of
+the adjoint Neumann-Ulam method and could potentially be extended to
+non-symmetric systems. In addition, these relations will be used to
+analyze parallel implementations of Monte Carlo synthetic acceleration
+methods that leverage the adjoint method and serve as an intial
+grounds for assessing their feasibility for large-scale problems.
 
@@
 %%---------------------------------------------------------------------------%%
-%\Section*{REFERENCES}
+%%\Section*{REFERENCES}
 \setlength{\baselineskip}{12pt}
@@
 
-%% \begin{thebibliography}{300}
-%% \bibitem{journal} B. Author(s), ``Title,'' {\it Journal Name in Italic}, 
-%%           {\bf Volume in Bold}, pp. 34-89 (19xx).
-%% \bibitem{proc_paper} C. D. Author(s), ``Article Title,'' {\it Proceedings of
-%%           Meeting in Italic}, Location, Dates of Meeting, Vol. n, pp. 134-156 
-%%           (19xx).
-%% \bibitem{book} E. F. Author, {\it Book Title in Italic}, Publisher, City \&
-%%           Country (19xx). 
-%% \bibitem{website} ``Spallation Neutron Source: The next-generation 
-%%           neutron-scattering facility in the United States,'' 
-%%           http://www.sns.gov/documentation/sns\_brochure.pdf (2002).
-%% \end{thebibliography}
-
 \end{document}

### 2012-11-30 doc/mandc2013/theory_paper/slattery_wilson_evans.tex
finished fist draft of mandc theory paper

@@
 average length of the adjoint random walks, a measure of convergence
-speed and serial performace, are made with respect to the Eigenvalues
+speed and serial performance, are made with respect to the Eigenvalues
 and other properties of the linear operator. In addition,
@@
 adjoint Neuman-Ulam method has been leveraged in Monte Carlo Synthetic
-Acceleration (MCSA) methods that are competetive with conventional
+Acceleration (MCSA) methods that are competitive with conventional
 methods including Krylov methods and are applicable to a wide variety
@@
 \end{equation}
-We can now restrict $\ve{H}$ by asserting that $\rho(\ve{H}) < 1$
-for a stationary method to converge such that $k$ applications of the
-iteration matrix will not cause the error to grow in
-Eq~(\ref{eq:linear_k_iter_norm3}).
 
@@
 Per Eq~(\ref{eq:linear_k_iter_norm3}), we see that this fraction is
-equivalent to $\epsilon = \rho(\ve{H})^k$. For the adjoint Neumann-Ulam
-method, if we take this fraction to be the weight cutoff, a measure of
-how accurately the contributions of a particular history to the
-solution are tallied, we then have the following empirical relation
+equivalent to $\epsilon = \rho(\ve{H})^k$. For the adjoint
+Neumann-Ulam method, if we take this fraction to be the weight cutoff,
+a measure of how accurately the contributions of a particular history
+to the solution are tallied, we then have the following relationship
 for $k$:
@@
 discretization stencil, and the weight cutoff selected to terminate
-low weight histories.  We also note here an interesting consequence of
-this approximation; no matter how optically thick a domain may be, the
-domain is still expected to leak some histories and thus communicate
-them. Therefore, no matter how large the domains are made, if they do
-not encapsulate the entire global domain they require some
-communication in order to implement a domain decomposed Neumann-Ulam
-method.
+low weight histories.
 
@@
 using an iterative Eigenvalue solver. For this verification, a $100
-\times 100$ square grid with $h=0.01$, $h=0.1$, and $h=1.0$ and
-non-reentrant current boundary conditions was used and the absorption
-cross varied from 0 to 100 while the scattering cross section was
-fixed at unity. Figure~\ref{fig:measured_spec_rad} gives the measured
-spectral radius of the iteration matrix and the computed spectral
-radius for the preconditioned diffusion operator using
+\times 100$ square grid with $h=0.01$, $h=0.1$, and $h=1.0$ and the
+absorption cross varied from 0 to 100 while the scattering cross
+section was fixed at unity. Figure~\ref{fig:measured_spec_rad} gives
+the measured spectral radius of the iteration matrix and the computed
+spectral radius for the preconditioned diffusion operator using
 Eq~(\ref{eq:iteration_radius}) as function of the absorption to
@@
 adjoint Neumann-Ulam solver. To do this, we again use a $100 \times
-100$ square grid with $h=0.1$ and non-reentrant current boundary
-conditions was used and the absorption cross varied from 0 to 100
-while the scattering cross section was fixed at unity. Three weight
-cutoff values of \sn{1}{-2}, \sn{1}{-4}, and \sn{1}{-8} were used with
-10,000 histories generated by a point source source of strength 1 in
-the center of the domain such that the vacuum boundary conditions will
-have less of an effect on the length of the random walks. For each of
-the histories, the number of transitions made was tallied to provide
-an effective value of $k$ for each history. This value was then
-averaged over all histories to get a measured value of $k$ for the
-particular operator. On the left, figure~\ref{fig:measured_length}
-these measurements as well as the analytic result computed by
-Eq~(\ref{eq:analytic_k}) as a function of the iteration matrix
-spectral radius, $\rho(\ve{H})$. On the right,
+100$ square grid with $h=0.1$ and the absorption cross varied from 0
+to 100 while the scattering cross section was fixed at unity. Three
+weight cutoff values of \sn{1}{-2}, \sn{1}{-4}, and \sn{1}{-8} were
+used with 10,000 histories generated by a point source source of
+strength 1 in the center of the domain such that the boundary
+conditions will have less of an effect on the length of the random
+walks. For each of the histories, the number of transitions made was
+tallied to provide an effective value of $k$ for each history. This
+value was then averaged over all histories to get a measured value of
+$k$ for the particular operator. On the left,
+figure~\ref{fig:measured_length} these measurements as well as the
+analytic result computed by Eq~(\ref{eq:analytic_k}) as a function of
+the iteration matrix spectral radius, $\rho(\ve{H})$. On the right,
 figure~\ref{fig:measured_length} gives the relative error between the
 predicted and observed results. We note good qualitative agreement
-between the measured and analytic results. However, we obeserve a
+between the measured and analytic results. However, we observe a
 larger relative error for both long and short random walks.
@@
 analytic relation for the optical thickness of domain and the
-associated leakage approximations. For this experiment, square grid
-with $h=0.1$ and non-reentrant current boundary conditions. The grid
-is then decomposed into 9 square domains, 3 in each cardinal
-direction. For the cross sections, the absorption cross was varied
-from 1 to 100 while the scattering cross section was set to zero to
-create a purely absorbing environment. The optical thickness of these
-domains will vary as a function of the absorption cross section if the
-other parameters are fixed. To compute the optical thickness, along
-with the spectral radius as given by Eq~(\ref{eq:iteration_radius}),
-we also need the parameters $n_i$ and $n_s$ which respectively
-describe the typical domain length and the average number of states
-moved along that typical length per history transition. For our grid
-above, the domains are varied in size with $50 \times 50$, $100 \times
-100$, and $200 \times 200$ cells giving $n_i=50$, $n_i=100$, and
-$n_i=200$ grid points or states along the typical length of the domain
-respectively with weight cutoff of \sn{1}{-4}. Looking at the
-Laplacian stencil in Eq~(\ref{eq:nine_point_stencil}), we see that all
-history transitions will only move a single state in either the $i$ or
-$j$ directions due to the symmetry of the problem. Furthermore, if we
-choose the $i$ direction, not all states we will transition to will
-move the history in that direction. Therefore, we look to the
-definition of the iteration matrix in Eq~(\ref{eq:iteration_stencil})
-and the definition of the adjoint probability matrix in
-Eq~(\ref{eq:adjoint_probability}) to estimate the $n_s$ parameter. For
-a particular transition starting at state $(i,j)$, 6 of the 8 possible
-new states in the stencil move the history in $i$ direction with
-relative coefficients of 4 for moving in the $(\pm i,0)$ direction and
-of 1 for moving in the $(\pm i,\pm j)$. These coefficients dictate the
-frequency those states are visited relative to the others. For those 6
-states we can visit along the typical length their sum is 12 out of
-the total 20 for the coefficients for all possible states with their
-ratio giving $n_i = \frac{3}{5}$.
+associated leakage approximations. For this experiment, a square grid
+with $h=0.1$ was decomposed into 9 square domains, 3 in each cardinal
+direction with measurements occurring in the central domain without
+boundary grid points. For the cross sections, the absorption cross was
+varied from 1 to 100 while the scattering cross section was set to
+zero to create a purely absorbing environment. The optical thickness
+of these domains will vary as a function of the absorption cross
+section if the other parameters are fixed. To compute the optical
+thickness, along with the spectral radius as given by
+Eq~(\ref{eq:iteration_radius}), we also need the parameters $n_i$ and
+$n_s$ which respectively describe the typical domain length and the
+average number of states moved along that typical length per history
+transition. For our grid above, the domains are varied in size with
+$50 \times 50$, $100 \times 100$, and $200 \times 200$ cells giving
+$n_i=50$, $n_i=100$, and $n_i=200$ grid points or states along the
+typical length of the domain respectively with weight cutoff of
+\sn{1}{-4}. Looking at the Laplacian stencil in
+Eq~(\ref{eq:nine_point_stencil}), we see that all history transitions
+will only move a single state in either the $i$ or $j$ directions due
+to the symmetry of the problem. Furthermore, if we choose the $i$
+direction, not all states we will transition to will move the history
+in that direction. Therefore, we look to the definition of the
+iteration matrix in Eq~(\ref{eq:iteration_stencil}) and the definition
+of the adjoint probability matrix in Eq~(\ref{eq:adjoint_probability})
+to estimate the $n_s$ parameter. For a particular transition starting
+at state $(i,j)$, 6 of the 8 possible new states in the stencil move
+the history in $i$ direction with relative coefficients of 4 for
+moving in the $(\pm i,0)$ direction and of 1 for moving in the $(\pm
+i,\pm j)$. These coefficients dictate the frequency those states are
+visited relative to the others. For those 6 states we can visit along
+the typical length, their sum is 12 out of the total 20 for the
+coefficients for all possible states with their ratio giving $n_i =
+\frac{3}{5}$.
 
@@
 domain. At the start of a stage of histories, the number of histories
-starting in the domain was computed and as the stage progressed, the
-number of histories that exited the domain was tallied with the ratio
-of the two numbers providing a numerical measure for the leakage
-fraction. Figure~\ref{fig:measured_leakage} gives the domain leakage
-measurements for the domain in the center of the global grid as well
-as the analytic result computed by
+starting in the center domain was computed and as the stage
+progressed, the number of histories that exited that domain was
+tallied with the ratio of the two numbers providing a numerical
+measure for the leakage fraction. Figure~\ref{fig:measured_leakage}
+gives the domain leakage measurements for the domain in the center of
+the global grid as well as the analytic result computed by
 Eqs~(\ref{eq:wigner_domain_leakage}) and
@@
 while the large relative error in $k$ for optically thin domains does
-not affect the approximation. In general, the mean chord approximation
-is a better choice to estimate the leakage fraction in a domain from
-the adjoint Neumann-Ulam method and except for a single data point
-with $n_i=50$, the mean chord approximation yielded leakage fractions
-within 0.05 of the measured results. As the domain becomes more
-optically thick (with both increasing $n_i$ and decreasing
+not affect the approximation significantly. In general, the mean chord
+approximation is a better choice to estimate the leakage fraction in a
+domain from the adjoint Neumann-Ulam method and except for a single
+data point with $n_i=50$, the mean chord approximation yielded leakage
+fractions within 0.05 of the measured results. As the domain becomes
+more optically thick (with both increasing $n_i$ and decreasing
 $\rho(\ve{H})$), the approximations are more accurate.
@@
 basis for performance models for future parallel implementations of
-the adjoint Neumann-Ulam method and could potentially be extended to
+the adjoint Neumann-Ulam method. To be applicable to a broader range
+of problems, these models could potentially be extended to
 non-symmetric systems. In addition, these relations will be used to
 analyze parallel implementations of Monte Carlo synthetic acceleration
-methods that leverage the adjoint method and serve as an intial
+methods that leverage the adjoint method and serve as an initial
 grounds for assessing their feasibility for large-scale problems.
@@
 %%---------------------------------------------------------------------------%%
-\section*{ACKNOWLEDGEMENTS}
+\section*{ACKNOWLEDGMENTS}
 
 This work was performed under appointment to the Nuclear Regulatory
-Commision Fellowship program at the University of Wisconsin - Madison
+Commission Fellowship program at the University of Wisconsin - Madison
 Engineering Physics Department.

### 2012-12-03 doc/mandc2013/theory_paper/slattery_wilson_evans.tex
updated paper title and keywords

@@
 \textbf{\large \\% 
-A SPECTRAL ANALYSIS OF THE DOMAIN DECOMPOSED ADJOINT MONTE CARLO
-METHOD FOR LINEAR OPERATOR EQUATIONS}
+A SPECTRAL ANALYSIS OF THE DOMAIN DECOMPOSED MONTE CARLO METHOD FOR
+LINEAR SYSTEMS} 
 % 
@@
 
-\emph{Key Words}: Monte Carlo domain decomposition MCSA
+\emph{Key Words}: Monte Carlo, domain decomposition, MCSA
 \end{small} 

### 2012-12-10 doc/sna2013/abstract/sna_abstract.tex
adding snamc2013 extended abstract

--- 2012-11-30 doc/mandc2013/notes/spectral_theory.tex
@@
-%        File: latexdoc.tex
-%     Created: Friday March 4 12:24:15 2011
-% Last Change: Friday March 4 12:24:20 2011
-%
 \documentclass[letterpaper,12pt]{article}
@@
 \usepackage{amsmath}
-\usepackage{amsthm}
-\usepackage[mathcal]{euscript}
-\usepackage{tabularx}
-\usepackage{cite}
-\usepackage{c++}
-\usepackage{tmadd,tmath}
 \usepackage[usenames]{color}
@@
 
-\date{\today}
-\title{A Spectral Analysis of the Domain Decomposed Adjoint
-  Neumann-Ulam Method}
+\date{\today} \title{An Overlapping-Domain Decomposed Monte Carlo
+  Synthetic Acceleration Method for Linear Systems}
 \begin{document}
@@
 %%---------------------------------------------------------------------------%%
-\begin{abstract}
-
-  The domain decomposed behavior of the adjoint Neumann-Ulam method
-  for solving linear operator equations is analyzed using the spectral
-  properties of the linear operator. Relationships for the average
-  length of the adjoint random walks and the fraction of histories
-  leaking from a domain in the decomposition are made with respect to
-  the Eigenvalues and other properties of the linear operator. In
-  addition, a formal definition of the optical thickness of a domain
-  in the decomposition is presented based on the spectral
-  analysis. The one-speed, two-dimensional neutron diffusion equation
-  is used as a model problem to test the spectral theory for symmetric
-  operators. In general, the spectral theory shows good agreement with
-  measured computational results.
-
-\end{abstract}
+\section{Introduction}
 
 %%---------------------------------------------------------------------------%%
-\section{Introduction}
-An alternative approach to approximate matrix inversion is to employ
-Monte Carlo methods that sample a distribution with an expectation
-value equivalent to that of the inverted operator. Such methods have
-been in existence for decades with the earliest reference noted here
-an enjoyable manuscript published in 1950 by Forsythe and Leibler
-\cite{forsythe_matrix_1950}. In their outline, Forsythe and Liebler
-in fact credit the creation of this technique to J. Von Neumann and
-S.M. Ulam some years earlier than its publication. In 1952 Wasow
-provided a more formal explanation of Von Neumann and Ulam's method
-\cite{wasow_note_1952} and Hammersley and Handscomb's 1964 monograph
-\cite{hammersley_monte_1964} and Spanier and Gelbard's 1969 book
-\cite{spanier_monte_1969} present additional detail on a method
-adjoint to that described by Neumann and Ulam.
-
-For very large linear systems, two things are likely to occur: a
-single compute node does not contain enough memory to hold the entire
-problem or a single compute node takes a prohibitively long time to
-solve the system. Domain decomposition methods permit the global
-computational domain to be subdivided into smaller regions. These
-smaller regions require less memory to store and fewer computational
-resources to solve as compared to the global problem. Typically, these
-smaller regions are then represented separately in individual compute
-nodes. Therefore, in order to solve large scale problems with the
-adjoint Neumann-Ulam method, we must explore its behavior in a domain
-decomposed environment.
-
-For reactor physics Monte Carlo simulations, domain decomposition has
-been identified as a key principle in moving forward in high
-performance computing
-\cite{brunner_comparison_2006,siegel_analysis_2012}. To accomplish
-this, we recognize from the literature that stochastic histories must
-be transported from domain to domain as the simulation progresses and
-they transition to states that are not in the local domain. Because we
-have chosen a domain decomposition strategy in a parallel environment,
-this means that communication of these histories must occur between
-compute nodes owning neighboring pieces of the global domain. We wish
-to characterize this communication not only because communication is
-in general expensive, but also because nearest-neighbor communication
-sequences have poor algorithmic strong scaling
-\cite{gropp_high-performance_2001}.
-
-For problems where the linear operator is symmetric, a host of
-analytic theories exist based on the Eigenvalue spectrum of the
-operator that characterize their behavior in the context of
-deterministic linear solvers. Using past work, these theories are
-adapted to the domain decomposed adjoint Neumann-Ulam method using the
-one-speed, two-dimensional neutron diffusion equation. In this paper
-we describe the adjoint Neumann-Ulam method followed by a presentation
-and finite difference discretization of the model problem. Using the
-linear system generated by this discretization, we use a spectral
-theory to generate analytic relations for the Eigenvalues of the
-operator based on system parameters. Using the Eigenvalue spectra, we
-then build relationships to characterize the transport of random walk
-histories in a decomposed domain and compare these analytic results to
-numerical experiments conducted with the model problem.
+\section{Monte Carlo Synthetic Acceleration}
 
 %%---------------------------------------------------------------------------%%
-\section{The Adjoint Neumann-Ulam Method}
-We seek solutions of the general linear problem in the following form:
-\begin{equation}
-  \ve{A} \ve{x} = \ve{b}\:,
-  \label{eq:linear_problem}
-\end{equation}
-where $\ve{A} \in \mathbb{R}^{N \times N}$ is a linear operator such
-that $\ve{A} : \mathbb{R}^{N} \rightarrow \mathbb{R}^{N}$, $\ve{x} \in
-\mathbb{R}^N$ is the solution vector, and $\ve{b} \in \mathbb{R}^N$ is
-the forcing term. Choosing the adjoint method Monte Carlo method to
-invert the linear operator, we begin by defining the linear system
-adjoint to Eq~(\ref{eq:linear_problem}):
-\begin{equation}
-  \ve{A}^T \ve{y} = \ve{d}\:,
-  \label{eq:adjoint_linear_problem}
-\end{equation}
-where $\ve{y}$ and $\ve{d}$ are the adjoint solution and source
-respectively and $\ve{A}^T$ is the adjoint operator.  With this
-statement we can then define the split equation:
-\begin{equation}
-  \ve{y} = \ve{H}^T \ve{y} + \ve{d}\:,
-  \label{eq:adjoint_split_system}
-\end{equation}
-where $\ve{H} = \ve{I} - \ve{A}$ and is defined as the
-\textit{iteration matrix}.  Using these definitions, we can derive an
-estimator from the adjoint method that will also give the solution
-vector, $\ve{x}$. We can acquire the adjoint solution by forming the
-Neumann series by writing Eq~(\ref{eq:adjoint_split_system}) as:
-\begin{equation}
-  \ve{y} = (\ve{I} - \ve{H}^T)^{-1} \ve{d}\:,
-  \label{eq:adjoint_split_system_2}
-\end{equation}
-which in turn yields the Neumann series using the adjoint operator:
-\begin{equation}
-  \ve{y} = \sum_{k=0}^{\infty} (\ve{H}^T)^k\ve{d}\:.
-  \label{eq:adjoint_neumann_series}
-\end{equation}
-For this series to converge, we require that the spectral radius of
-$\ve{H}$ must remain less than 1 as $\ve{H}^T$ contains the same
-Eigenvalues and therefore has the same spectral radius. We expand this
-summation to again yield a series of transitions that can be
-approximated by a Monte Carlo random walk sequence, forming the
-Neumann series in reverse order:
-\begin{equation}
-  y_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-  \sum_{i_k}^{N}h_{i_k,i_{k-1}}\ldots h_{i_2,i_1} h_{i_1,i} d_{i_k}\:.
-  \label{eq:adjoint_neumann_solution}
-\end{equation}
-We can readily build an estimator for the adjoint solution from this
-series expansion, but we instead desire the solution to
-Eq~(\ref{eq:linear_problem}). The adjoint estimator can be related to
-the solution by defining the following inner product equivalence
-\cite{spanier_monte_1969}:
-\begin{equation}
-  \langle \ve{A}^T \ve{x}, \ve{y} \rangle = \langle \ve{x}, \ve{A}
-  \ve{y} \rangle\:.
-  \label{eq:adjoint_operator_product}
-\end{equation}
-From this definition it follows that:
-\begin{equation}
-  \langle \ve{x}, \ve{d} \rangle = \langle \ve{y}, \ve{b} \rangle\:.
-  \label{eq:adjoint_vector_relation}
-\end{equation}
-Here we have 2 unknowns, $\ve{y}$ and $\ve{d}$, and therefore we
-require two constraints to close the system. We use
-Eq~(\ref{eq:adjoint_vector_relation}) as the first constraint and as a
-second constraint we select:
-\begin{equation}
-  \ve{d} = \boldsymbol{\delta}_i\:,
-  \label{eq:adjoint_second_constraint}
-\end{equation}
-where the $k^{th}$ component of the vector $\boldsymbol{\delta}_i$ is
-the Dirac delta function $\delta_{i_k,i}$. If we apply
-Eq~(\ref{eq:adjoint_second_constraint}) to our first constraint
-Eq~(\ref{eq:adjoint_vector_relation}), we get the following convenient
-outcome:
-\begin{equation}
-  \langle \ve{y}, \ve{b} \rangle = \langle \ve{x},
-  \boldsymbol{\delta}_i \rangle = x_i \:,
-  \label{eq:inner_product_constraint}
-\end{equation}
-meaning that if we compute the inner product of the original source and
-the adjoint solution using a delta function source, we recover the
-original solution we are looking for. In terms of particle transport, a
-delta function source means that this adjoint method is equivalent to
-a traditional forward method where tallies are made only in the state
-in which the random walk currently resides.
-
-In order to perform the random walks required to approximate the
-summation in Eq~(\ref{eq:adjoint_neumann_solution}), probabilities and
-weights are generated using the \textit{adjoint Neumann-Ulam
-  decomposition} of $\ve{H}$:
-\begin{equation}
-  \ve{H}^{T} = \ve{P} \circ \ve{W}\:,
-  \label{eq:adjoint_neumann_ulam}
-\end{equation}
-where now we are forming the decomposition with respect to the
-transpose of $\ve{H}$. We then build the probability and weight
-matrices in the decomposition where probabilities are column-scaled:
-\begin{equation}
-  p_{ij} = \frac{|h_{ji}|}{\sum_j |h_{ji}|}\:,
-  \label{eq:adjoint_probability}
-\end{equation}
-such that we expect to select a new state $j$ from the current state
-in the random walk $j$ by sampling column-wise. Per
-Eq~(\ref{eq:adjoint_neumann_ulam}), the transition weight is then
-defined as:
-\begin{equation}
-  w_{ij} = \frac{h_{ji}}{p_{ij}}\:.
-  \label{eq:adjoint_weight}
-\end{equation}
-Using the decomposition we can then define an expectation value for
-the adjoint method. Using our result from
-Eq~(\ref{eq:inner_product_constraint}) generated by applying the
-adjoint constraints, the contribution to the solution in state $i$
-from a particular random walk permutation is then:
-\begin{equation}
-  X_{\nu} = \sum_{m=0}^k W_{m} \delta_{i,i_m}\:,
-  \label{eq:adjoint_permutation_contribution}
-\end{equation}
-where the Kronecker delta indicates that the tally contributes only in
-the current state and $b_{i_0}$ will be the sampled source starting
-weight. Finally, the expectation value using all permutations is:
-\begin{equation}
-  E\{X\} = \sum_{\nu} P_{\nu} X_{\nu}\:
-  \label{eq:adjoint_expectation_value}
-\end{equation}
-which if expanded directly recovers the exact solution:
-\begin{equation}
-  \begin{split}
-    E\{X_j\} &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-    \sum_{i_k}^{N} b_{i_0} h_{i_0,i_1}h_{i_1,i_2}\ldots
-    h_{i_{k-1},i_k} \delta_{i_k,j} \\ &= x_{j}\:,
-  \end{split}
-  \label{eq:adjoint_expectation_expansion}
-\end{equation}
-therefore also providing an unbiased Monte Carlo estimate of the
-solution.
-
-We also desire a criteria for random walk termination for problems
-where only an approximate solution is necessary. For the adjoint
-method, we utilize a relative weight cutoff parameter:
-\begin{equation}
-  W_f = W_c b_{i_0}\:,
-  \label{eq:relative_weight_cutoff}
-\end{equation}
-where $W_c$ is the \textit{weight cutoff}. The adjoint random
-walk will then be terminated after $m$ steps if $W_m < W_f$ as tally
-contributions become increasingly small.
+\section{Domain Decomposition and Overlap}
 
 %%---------------------------------------------------------------------------%%
-\section{Model Problem}
-For our numerical experiments, we choose the one-speed,
-two-dimensional neutron diffusion equation as a model problem
-\cite{duderstadt_nuclear_1976}:
-\begin{equation}
-  -\boldsymbol{\nabla} \cdot D \boldsymbol{\nabla} \phi + \Sigma_a
-  \phi = S\:,
-  \label{eq:diffusion_eq}
-\end{equation}
-where $\phi$ is the neutron flux, $\Sigma_a$ is the absorption cross
-section, and $S$ is the source of neutrons. In addition, $D$ is the
-diffusion coefficient defined as:
-\begin{equation}
-  D = \frac{1}{3 ( \Sigma_t - \bar{\mu}\Sigma_s )}\:,
-  \label{eq:diffusion_coeff}
-\end{equation}
-where $\Sigma_s$ is the scattering cross section, $\Sigma_t = \Sigma_a
-+ \Sigma_s$ is the total cross section, and $\bar{\mu}$ is the cosine
-of the average scattering angle. For simplicity, we will take
-$\bar{\mu} = 0$ for our analysis giving $D=(3 \Sigma_t)^{-1}$. In
-addition, to further simplify we will assume a homogeneous domain such
-that the cross sections remain constant throughout. Doing this permits
-us to rewrite Eq~(\ref{eq:diffusion_eq}) as:
-\begin{equation}
-  -D \boldsymbol{\nabla}^2 \phi + \Sigma_a \phi = S\:.
-  \label{eq:diffusion_eq_simple}
-\end{equation}
-
-We choose a finite difference scheme on a square Cartesian grid to
-discretize the problem. For the Laplacian, we choose the 9-point
-stencil shown in Figure~\ref{fig:stencil} over a grid of size $h$
-\cite{leveque_finite_2007}:
-\begin{multline}
-  \nabla^2_9\phi = \frac{1}{6h^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j}
-    + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ +
-    \phi_{i-1,j+1} + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20
-    \phi_{i,j}]\:.
-  \label{eq:nine_point_stencil}
-\end{multline}
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.25}{\input{stencil.pdftex_t}}
-  \end{center}
-  \caption{\textbf{Nine-point Laplacian stencil.}}
-  \label{fig:stencil}
-\end{figure}
-We then have the following linear system to solve:
-\begin{multline}
-  -\frac{1}{6h^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j} + 4
-    \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ + \phi_{i-1,j+1}
-    + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20 \phi_{i,j}] + \Sigma_a
-  \phi_{i,j} = s_{i,j}\:,
-  \label{eq:fd_system}
-\end{multline}
-and in operator form:
-\begin{equation}
-  \ve{D}\boldsymbol{\phi}=\ve{s}\:,
-  \label{eq:operator_system}
-\end{equation}
-where $\ve{D}$ is the diffusion operator, $\ve{s}$ is the source in
-vector form and $\boldsymbol{\phi}$ is the vector of unknown fluxes.
-
-To close the system, a set of boundary conditions is required. In the
-case of a non-reentrant current condition applied to all global
-boundaries of the domain, we choose the formulation of Duderstadt by
-assuming the flux is zero at some ghost point beyond the
-grid. Consider for example the equations on the $i=0$ boundary of the
-domain:
-\begin{multline}
-  -\frac{1}{6h^2}[4 \phi_{-1,j} + 4 \phi_{1,j} + 4 \phi_{0,j-1} +
-    4 \phi_{0,j+1} + \phi_{-1,j-1}\\ + \phi_{-1,j+1} + \phi_{1,j-1} +
-    \phi_{1,j+1} - 20 \phi_{0,j}] + \Sigma_a \phi_{0,j} = s_{0,j}\:.
-  \label{eq:x_min_bnd}
-\end{multline}
-Here we note some terms where $i=-1$ and therefore are representative
-of grid points beyond the boundary of the domain. We set the flux at
-these points to be zero, giving a valid set of equations for the $i=0$
-boundary:
-\begin{multline}
-  -\frac{1}{6h^2}[4 \phi_{1,j} + 4 \phi_{0,j-1} + 4 \phi_{0,j+1}
-    \\ + \phi_{-1,j+1} + \phi_{1,j-1} + \phi_{1,j+1} - 20 \phi_{0,j}]
-  + \Sigma_a \phi_{0,j} = s_{0,j}\:.
-  \label{eq:x_min_bnd_2}
-\end{multline}
-We repeat this procedure for the other boundaries of the domain.
-
-For reflecting boundary conditions, the net current across a boundary
-is zero.
-
-%%---------------------------------------------------------------------------%%
-\section{Spectral Analysis}
-The convergence of the Neumann series in
-Eq~(\ref{eq:adjoint_neumann_series}) approximated by the Monte Carlo
-solver is dependent on the Eigenvalues of the iteration matrix. In
-addition, the Eigenvalues of the linear operator from which the
-iteration matrix was derived allow for further analysis of convergence
-properties. We will compute the Eigenvalues of these matrices by
-assuming Eigenfunctions of the form \cite{leveque_finite_2007}:
-\begin{equation}
-  \Phi_{p,q}(x,y) = e^{2 \pi \imath p x} e^{2 \pi \imath q y}\:,
-  \label{eq:eigenfunction_form}
-\end{equation}
-where different combinations of $p$ and $q$ represent the different
-Eigenmodes of the solution. As these are valid forms of the solution,
-then the action of the linear operator on these Eigenfunctions should
-give the Eigenvalues of the matrix as they lie on the unit circle in
-the complex plane.
-
-\subsection{Diffusion Operator Spectrum}
-For the model problem, we first compute the Eigenvalues for the
-diffusion operator $\ve{D}$ by applying the operator to the
-Eigenfunctions and noting that $x=ih$ and $y=jh$:
-\begin{multline}
-  \ve{D}\Phi_{p,q}(x,y) = \lambda_{p,q}(\ve{D})
-  =\\ -\frac{D}{6h^2}\Big[4 e^{-2 \pi \imath p h} + 4 e^{2 \pi \imath
-      p h} + 4 e^{-2 \pi \imath q h} + 4 e^{2 \pi \imath q h} + e^{-2
-      \pi \imath p h} e^{-2 \pi \imath q h} \\ + e^{-2 \pi \imath p h}
-    e^{2 \pi \imath q h} + e^{2 \pi \imath p h} e^{-2 \pi \imath q h}
-    + e^{2 \pi \imath p h} e^{2 \pi \imath q h} - 20\Big] + \Sigma_a
-  \:.
-  \label{eq:deriv_diff_1}
-\end{multline}
-Using Euler's formula, we can collapse the exponentials to
-trigonometric functions:
-\begin{equation}
-  \lambda_{p,q}(\ve{D}) = -\frac{D}{6h^2}[ 8 \cos(\pi p h) + 8
-    \cos(\pi q h) + 4 \cos(\pi p h) \cos(\pi q h) - 20] + \Sigma_a\:.
-  \label{eq:deriv_diff_2}
-\end{equation}
-
-As Eq~(\ref{eq:diffusion_eq}) is diagonally dominant, Jacobi
-preconditioning is sufficient to reduce the spectral radius of the
-iteration matrix below unity and therefore ensure convergence of the
-Neumann series. The preconditioner in this case is then $\ve{M} =
-diag(\ve{D})$ such that we are solving the following linear system:
-\begin{equation}
-  \ve{M}^{-1} \ve{D} \boldsymbol{\phi} = \ve{M}^{-1} \ve{s}\:.
-  \label{eq:precond_diffsion}
-\end{equation}
-The operator $\ve{M}^{-1} \ve{D}$ is merely the original diffusion
-operator with each row scaled by the diagonal component. As we have
-defined a homogeneous domain, the scaling factor is the same for all
-rows in the operator. This scaling factor, $\alpha$, is then defined
-as the $\phi_{i,j}$ coefficient from Eq~(\ref{eq:fd_system}):
-\begin{equation}
-  \alpha = \Bigg[\frac{10 D}{3 h^2} + \Sigma_a\Bigg]^{-1}\:.
-  \label{eq:jacobi_scaling}
-\end{equation}
-Using this coefficient, we then have the following spectrum of
-preconditioned Eigenvalues:
-\begin{equation}
-  \lambda_{p,q}(\ve{M}^{-1} \ve{D}) = \alpha \lambda_{p,q}(\ve{D})\:.
-  \label{eq:preconditioned_Eigenvalues}
-\end{equation}
-
-A good measure of how well-conditioned is a linear system can be
-obtained by computing the condition number for the linear operator
-defined as:
-\begin{equation}
-  \kappa(\ve{A}) = ||\ve{A}||\ ||\ve{A}^{-1}||\:,
-  \label{eq:condition_number}
-\end{equation}
-which can be related to the Eigenvalues of the operator as:
-\begin{equation}
-  \kappa(\ve{A}) =
-  \frac{\max_{p,q}\lambda_{p,q}(\ve{A})}{\min_{p,q}\lambda_{p,q}(\ve{A})}\:.
-  \label{eq:condition_number_2}
-\end{equation}
-To compute the condition number for the preconditioned diffusion
-system, we must therefore compute the minimum and maximum Eigenvalues
-of the diffusion operator. Per Leveque's text
-\cite{leveque_finite_2007}, we expect to find these minimum and
-maximum Eigenvalues when $p=q$. To find these values,
-Eq~(\ref{eq:preconditioned_Eigenvalues}) is plotted as a function of
-$p$ with $p=q$ in Figure~\ref{fig:diffusion_spectrum} for various
-values of $\Sigma_a$.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=5in,clip]{diffusion_spectrum.png}
-  \end{center}
-  \caption{\textbf{Eigenvalue spectra for the diffusion equation.}}
-  \label{fig:diffusion_spectrum}
-\end{figure}
-We see a minimum Eigenvalue at $p=0$ and a maximum at $p=10$, letting
-us compute the condition number for the Jacobi preconditioned
-diffusion operator:
-\begin{equation}
-  \kappa(\ve{M}^{-1} \ve{D}) =
-  \frac{\frac{D}{6h^2}\Sigma_a}{\frac{D}{3h^2}[8 \cos(10 \pi h) + 2
-      \cos^2(10 \pi h) - 10] + \Sigma_a}\:.
-  \label{eq:diffusion_condition}
-\end{equation}
-Visually we can also see how increasing the absorption in the problem
-creates a more well conditioned system as less diffusion is
-expected. If we have a problem with only scattering, $\Sigma_a = 0$
-and we observe an infinite condition number as $\lambda_{min}
-\rightarrow 0$.
-
-\subsection{Iteration Matrix Spectrum}
-The spectral radius of the iteration matrix is obtained by seeking its
-largest Eigenvalue. As with the diffusion operator, we can use the
-same analysis techniques to find the Eigenvalues for the iteration
-matrix. We use a few simplifications by noting that if the Jacobi
-preconditioned iteration matrix is $\ve{H} = \ve{I} -
-\ve{M}^{-1}\ve{D}$, the we except all terms on the diagonal of the
-iteration matrix to be zero such that we have the following stencil:
-\begin{equation}
-  \ve{H}\boldsymbol{\phi} = -\frac{\alpha D}{6h^2}[4 \phi_{i-1,j}
-    + 4 \phi_{i+1,j} + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} +
-    \phi_{i-1,j-1}\\ + \phi_{i-1,j+1} + \phi_{i+1,j-1} +
-    \phi_{i+1,j+1}]\:.
-  \label{eq:iteration_stencil}
-\end{equation}
-Inserting the Eigenfunctions defined by Eq~\ref{eq:eigenfunction_form}
-we get:
-\begin{multline}
-  \lambda_{p,q}(\ve{H}) = -\frac{\alpha D}{6h^2}\Big[4 e^{-2 \pi \imath p
-      h} + 4 e^{2 \pi \imath p h} + 4 e^{-2 \pi \imath q h} + 4 e^{2
-      \pi \imath q h} + e^{-2 \pi \imath p h} e^{-2 \pi \imath q h}
-    \\ + e^{-2 \pi \imath p h} e^{2 \pi \imath q q} + e^{2 \pi \imath
-      p h} e^{-2 \pi \imath q h} + e^{2 \pi \imath p h} e^{2 \pi
-      \imath q h}\Big]\:,
-  \label{eq:iteration_deriv}
-\end{multline}
-which simplifies to:
-\begin{equation}
-  \lambda_{p,q}(\ve{H}) = -\frac{\alpha D}{6h^2}[ 8 \cos(\pi p h) + 8
-    \cos(\pi q h) + 4 \cos(\pi p h) \cos(\pi q h)]\:,
-  \label{eq:iteration_spectrum}
-\end{equation}
-giving the Eigenvalue spectrum for the Jacobi preconditioned iteration
-matrix. As with the linear operator, we expect the maximum Eigenvalue
-to exist when $p=q$. In addition, we note that we have the same cosine
-form of the spectrum as for the linear operator expect that the
-spectra will be shifted by 10 in the $-p$ direction. We therefore
-expect the maximum Eigenvalues to be found when $p=q=0$, giving the
-following for the spectral radius of the Jacobi preconditioned
-iteration matrix:
-\begin{equation}
-  \rho(\ve{H}) = \frac{10 \alpha D}{3 h^2}\:.
-  \label{eq:iteration_radius}
-\end{equation}
-
-\subsection{Neumann Series Convergence}
-The Neumann-Ulam method for matrix inversion is effectively an
-approximation to a stationary method (the Jacobi method in the case of
-the preconditioning above). Stationary methods for linear systems
-arise from splitting the operator in Eq~(\ref{eq:linear_problem}):
-\begin{equation}
-  \ve{A} = \ve{M} - \ve{N}\:,
-  \label{eq:split_linear_operator}
-\end{equation}
-where the choice of $\ve{M}$ and $\ve{N}$ will be dictated by the
-particular method chosen. Using this split definition of the operator
-we can then write:
-\begin{equation}
-  \ve{M}\ve{x} - \ve{N}\ve{x} = \ve{b}\:.
-  \label{eq:linear_split_equation1}
-\end{equation}
-By rearranging, we can generate a form more useful for analysis:
-\begin{equation}
-  \ve{x} = \ve{H}\ve{x} + \ve{c}\:,
-  \label{eq:linear_split_equation2}
-\end{equation}
-where $\ve{H}=\ve{M}^{-1}\ve{N}$ is again defined as the iteration
-matrix and $\ve{c}=\ve{M}^{-1}\ve{b}$. With the solution vector on
-both the left and right hand sides, an iterative method can then be
-formed:
-\begin{equation}
-  \ve{x}^{k+1} = \ve{H}\ve{x}^k + \ve{c}\:,
-  \label{eq:linear_iterative_method}
-\end{equation}
-with $k \in \mathbb{Z}^+$ defined as the \textit{iteration
-  index}. Given this, we can then generate a few statements regarding
-the convergence of such stationary methods. Defining $\ve{e}^k =
-\ve{u}^k - \ve{u}$ as the solution error at the $k^{th}$ iterate, we
-can subtract Eq~(\ref{eq:linear_split_equation2}) from
-Eq~(\ref{eq:linear_iterative_method}) to arrive at an error form of
-the linear problem:
-\begin{equation}
-  \ve{e}^{k+1} = \ve{H}\ve{e}^k\:. 
-  \label{eq:linear_iterative_error}
-\end{equation}
-Our error after $k$ iterations is then:
-\begin{equation}
-  \ve{e}^{k} = \ve{H}^k\ve{e}^0\:. 
-  \label{eq:linear_k_iter_error}
-\end{equation}
-In other words, successive application of the iteration matrix is the
-mechanism driving down the error in a stationary method. We can then
-place restrictions on the iteration matrix by assuming $\ve{H}$ is
-diagonalizable \cite{saad_iterative_2003}, we then have:
-\begin{equation}
-  \ve{e}^{k} =
-  \ve{R}\boldsymbol{\Lambda}^k\ve{R}^{-1}\ve{e}^0\:,
-  \label{eq:linear_k_iter_error_diag}
-\end{equation}
-where $\boldsymbol{\Lambda}$ contains the Eigenvalues of $\ve{H}$ on
-its diagonal and the columns of $\ve{R}$ contain the Eigenvectors of
-$\ve{H}$. Computing the 2-norm of the above form then gives:
-\begin{equation}
-  ||\ve{e}^{k}||_2 \leq ||\boldsymbol{\Lambda}^k||_2\ 
-  ||\ve{R}||_2\ ||\ve{R}^{-1}||_2\ ||\ve{e}^0||_2\:,
-  \label{eq:linear_k_iter_norm1}
-\end{equation}
-which gives:
-\begin{equation}
-  ||\ve{e}^{k}||_2 \leq \rho(\ve{H})^k \kappa(\ve{R})
-  ||\ve{e}^0||_2\:.
-  \label{eq:linear_k_iter_norm2}
-\end{equation}
-For iteration matrices where the Eigenvectors are orthogonal,
-$\kappa(\ve{R})=1$ and the error bound reduces to:
-\begin{equation}
-  ||\ve{e}^{k}||_2 \leq \rho(\ve{H})^k ||\ve{e}^0||_2\:.
-  \label{eq:linear_k_iter_norm3}
-\end{equation}
-We can now restrict $\ve{H}$ by asserting that $\rho(\ve{H}) < 1$
-for a stationary method to converge such that $k$ applications of the
-iteration matrix will not cause the error to grow in
-Eq~(\ref{eq:linear_k_iter_norm3}).
-
-In the adjoint Neumann-Ulam method, $k$ iterations, equivalent to $k$
-applications of the iteration matrix, are approximated by a random
-walk of average length $k$ to yield the summation in
-Eq~(\ref{eq:adjoint_neumann_solution})
-\cite{dimov_new_1998,halton_sequential_1994,danilov_asymptotic_2000}. This
-random walk length, or the number of transitions before the
-termination of a history (either by the weight cutoff, absorption, or
-exiting the domain) is therefore bound to the number of stationary
-iterations required to converge to the specified tolerance. In the
-case of the adjoint Neumann-Ulam method, no such tolerance exists,
-however, we have specified a weight cutoff $W_c$ under which
-low-weight histories will be prematurely terminated as their
-contributions are deemed minute. After $k$ iterations, a stationary
-method is terminated as the error has reached some fraction,
-$\epsilon$, of the initial error:
-\begin{equation}
-  ||\ve{e}^{k}||_2 = \epsilon ||\ve{e}^0||_2\:.
-  \label{eq:linear_k_iter_norm4}
-\end{equation}
-Per Eq~(\ref{eq:linear_k_iter_norm3}), we see that this fraction is
-equivalent to $\epsilon = \rho(\ve{H})^k$. For the adjoint Neumann-Ulam
-method, if we take this fraction to be the weight cutoff, a measure of
-how accurately the contributions of a particular history to the
-solution are tallied, we then have the following empirical relation
-for $k$:
-\begin{equation}
-  k = \frac{ \log(W_c) }{ \log( \rho(\ve{H}) ) }\:.
-  \label{eq:analytic_k}
-\end{equation}
-This then gives us a means to estimate the length of the random walks
-that will be generated from a particular linear operator based on the
-Eigenvalues of its iteration matrix (independent of the linear
-operator splitting chosen) and based on the weight cutoff parameter
-used in the Neumann-Ulam method
-
-\subsection{Domain Leakage Approximations}
-In a domain decomposed situation, not all histories will remain within
-the domain they started in and must instead be communicated. This
-communication, expected to be expensive, was analyzed by Siegel and
-colleagues for idealized, load balanced situations for full nuclear
-reactor core Monte Carlo simulations \cite{siegel_analysis_2012}.  To
-quantify the number of particles that leak out of the local domain
-they define a leakage fraction, $\Lambda$, as:
-\begin{equation}
-  \Lambda = \frac{average\ \#\ of\ particles\ leaving\ local\ domain}
-          {total\ of\ \#\ of\ particles\ starting\ in\ local\ domain}\:.
-  \label{eq:leakage_fraction}
-\end{equation}
-For their studies, they assumed that the value of $\Lambda$ was
-dependent on the total cross section of the system via the Wigner
-rational approximation. This approximation was developed by Wigner to
-compute the neutron leakage probability from the fuel in a reactor as
-a simple interpolation between the small and large fuel lump cases,
-approximately characterizing the leakage behavior of many common
-geometries \cite{duderstadt_nuclear_1976}.
-
-Outlined more thoroughly by Hwang's chapter in
-\cite{azmy_nuclear_2010}, we begin the derivation of Wigner's
-approximation by defining $l$ to be the average chord length across
-the domain, and $\Sigma_t$ to be the total cross section in the
-domain. We then bound the behavior for the leakage fraction in terms
-of the optical thickness, $\Sigma_t l$, of the local domain. In the
-case of a small domain or a small cross section (and therefore small
-probability of absorption or scatter), $\Sigma_t l \rightarrow 0,
-\Lambda \rightarrow 1$. For a large domain or large cross section,
-$\Sigma_t l \rightarrow \infty, \Lambda \rightarrow (\Sigma_t
-l)^{-1}$, where $(\Sigma_t l)^{-1}$ is effectively the number of
-interactions that will occur over the length of the chord. Using
-Wigner's simple interpolation, we then have a simple formula for the
-domain leakage, $\Lambda = 1/(1+\Sigma_t l)$. This provides a leakage
-estimate, valid over the entire range of $\Sigma_t l$, in terms of the
-optical thickness of the system. In addition to Wigner's
-approximation, Hwang also defines the mean-chord approximation based
-on the same numerical limits with $\Lambda = (1-e^{-\Sigma_t
-  l})/\Sigma_t l$.
-
-In the case of domain decomposed linear operator equations, we can use
-diffusion theory to estimate the optical thickness of a domain in the
-decomposition and the corresponding leakage fraction in terms of
-properties of the linear operator and the discretization. To begin we
-must first calculate the mean distance a Monte Carlo history will move
-in the grid by computing the mean squared distance of its movement
-along the chord defined across the domain of length $l$. After a
-single transition a history will have moved a mean squared distance
-of:
-\begin{equation}
-  \langle \bar{r_1^2} \rangle = (n_s h)^2\:,
-  \label{eq:step_1_length}
-\end{equation}
-where $h$ is the size of the grid along the chord and $n_s$ is the
-number of grid elements a history will move on average every
-transition. For our diffusion model problem, $n_s$ would equate to the
-number of states in the $i$ (or $j$ as the problem is symmetric)
-direction that a history could potentially move in a single transition
-and is dependent on the stencil used for the discretization. After $k$
-transitions in the random walk, the history will have moved a mean
-squared distance of:
-\begin{equation}
-  \langle \bar{r_k^2} \rangle = k (n_s h)^2\:.
-  \label{eq:step_k_length}
-\end{equation}
-If our our chord is of length $l$ and there are $n_i$ grid points (or
-states to which a history may transition) along that chord, then $h =
-l / n_i$ giving:
-\begin{equation}
-  \langle \bar{r_k^2} \rangle = k \Bigg(\frac{n_s l}{n_i}\Bigg)^2\:.
-  \label{eq:step_k_length_sub}
-\end{equation}
-From diffusion theory, we expect the average number of interactions
-along the chord to be:
-\begin{equation}
-  \tau = \frac{l}{2 d \sqrt{\langle \bar{r_k^2} \rangle}}\:,
-  \label{eq:optical_thickness_1}
-\end{equation}
-where $d$ is the dimension of the problem and $\sqrt{\langle
-  \bar{r_k^2} \rangle}$ is effectively the mean free path of the Monte
-Carlo history in the domain. We can readily interperet $\tau$ to be
-the \textit{effective optical thickness} of the domain of length
-$l$. Inserting Eq~(\ref{eq:step_k_length_sub}) we arrive at:
-\begin{equation}
-  \tau = \frac{n_i}{2 d n_s \sqrt{k}}\:,
-  \label{eq:optical_thickness_2}
-\end{equation}
-which if expanded with Eq~(\ref{eq:analytic_k}) gives us the final
-relation for the effective optical thickness:
-\begin{equation}
-  \tau = \frac{n_i}{2 d n_s}
-  \sqrt{\frac{\log(\rho(\ve{H}))}{\log(W_c)}}\:.
-  \label{eq:optical_thickness_3}
-\end{equation}
-
-Using this optical thickness, we can then complete the leakage
-approximations by defining the bounds of $\tau \rightarrow 0, \Lambda
-\rightarrow 1$ and $\tau \rightarrow \infty, \Lambda \rightarrow
-\tau^{-1}$. We expect this behavior as optically thin domains will
-communicate most of their histories while optically thick domains will
-leak the fraction of histories that did not interact within. With
-these bounds, we can then define the leakage fraction out of a domain
-for the adjoint Neumann-Ulam method using the Wigner rational
-approximation:
-\begin{equation}
-  \Lambda = \frac{1}{1+\tau}\:,
-  \label{eq:wigner_domain_leakage}
-\end{equation}
-and using the mean-chord approximation:
-\begin{equation}
-  \Lambda = \frac{1-e^{-\tau}}{\tau}\:.
-  \label{eq:mean_chord_domain_leakage}
-\end{equation}
-Here, the leakage fraction is explicitly bound to the Eigenvalues of
-the iteration matrix, the size of the domain, the content of the
-discretization stencil, and the weight cutoff selected to terminate
-low weight histories.  We also note here an interesting consequence of
-this approximation: no matter how optically thick a domain may be, the
-domain is still expected to leak some histories and thus communicate
-them. Therefore, no matter how large the domains are made, if they do
-not encapsulate the entire global domain they require some
-communication in order to implement a domain decomposed adjoint
-Neumann-Ulam method.
-
-%%---------------------------------------------------------------------------%%
-\section{Numerical Experiments}
-To test the relationships developed by the spectral analysis, we form
-two simple numerical experiments using the diffusion model problem:
-one to measure the length of the random walks as a function of the
-iteration matrix Eigenvalues, and one to measure the domain leakage
-fraction as a function of the iteration matrix Eigenvalues and the
-discretization properties. Before doing this, we verify our
-computation of the spectral radius of the iteration matrix by
-numerically computing the largest Eigenvalue of the diffusion operator
-using an iterative Eigenvalue solver. For this verification, a $100
-\times 100$ square grid with $h=0.01$, $h=0.1$, and $h=1.0$ and
-non-reentrant current boundary conditions was used and the absorption
-cross varied from 0 to 100 while the scattering cross section was
-fixed at unity. Figure~\ref{fig:measured_spec_rad} gives the measured
-spectral radius of the iteration matrix and the computed spectral
-radius for the preconditioned diffusion operator using
-Eq~(\ref{eq:iteration_radius}) as function of the absorption to
-scattering ratio $(\Sigma_a / \Sigma_s)$. Excellent agreement was
-observed between the analytic and numerical results with all data
-points computed within the tolerance of the iterative Eigenvalue
-solver.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=5in,clip]{measured_spec_rad.png}
-  \end{center}
-  \caption{\textbf{Measured and analytic preconditioned diffusion
-      operator spectral radius as a function of the absorption cross
-      section to scattering cross section ratio.} \textit{Values of
-      $h=0.01$, $h=0.1$, and $h=1.0$ were used to verify the
-      behavior. The red data was computed numerically by an
-      Eigensolver while the black dashed data was generated by
-      Eq~(\ref{eq:iteration_radius}).}}
-  \label{fig:measured_spec_rad}
-\end{figure}
-
-\subsection{Random Walk Length}
-With the spectral data in hand, we can go about setting up an
-experiment to measure the length of the random walks generated by the
-adjoint Neumann-Ulam solver. To do this, we again use a $100 \times
-100$ square grid with $h=0.1$ and non-reentrant current boundary
-conditions was used and the absorption cross varied from 0 to 100
-while the scattering cross section was fixed at unity. Three weight
-cutoff values of \sn{1}{-2}, \sn{1}{-4}, and \sn{1}{-8} were used with
-10,000 histories generated by a point source source of strength 1 in
-the center of the domain such that the vacuum boundary conditions will
-have less of an effect on the length of the random walks. For each of
-the histories, the number of transitions made was tallied to provide
-an effective value of $k$ for each history. This value was then
-averaged over all histories to get a measured value of $k$ for the
-particular operator. Figure~\ref{fig:measured_length} gives these
-measurements as well as the analytic result computed by
-Eq~(\ref{eq:analytic_k}) as a function of the iteration matrix
-spectral radius, $\rho(\ve{H})$. Again, we note good agreement between
-the measured and analytic results.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=5in,clip]{measured_length.png}
-  \end{center}
-  \caption{\textbf{Measured and analytic random walk length as a
-      function of the iteration matrix spectral radius.} \textit{The
-      weight cutoff was varied with \sn{1}{-2}, \sn{1}{-4}, and
-      \sn{1}{-8}, to verify the validity of that parameter in
-      Eq~(\ref{eq:analytic_k}). The red data was computed numerically
-      by an adjoint Neumann-Ulam implementation while the black dashed
-      data was generated by Eq~(\ref{eq:analytic_k}).}}
-  \label{fig:measured_length}
-\end{figure}
-
-\subsection{Domain Leakage}
-Finally, we seek to measure the leakage from a domain in a domain
-decomposed Monte Carlo calculation and assess the quality of our
-analytic relation for the optical thickness of domain and the
-associated leakage approximations. For this experiment, square grid
-with $h=0.1$ and non-reentrant current boundary conditions. The grid
-is then decomposed into 9 square domains, 3 in each cardinal
-direction. For the cross sections, the absorption cross was varied
-from 1 to 100 while the scattering cross section was set to zero to
-create a purely absorbing environment. The optical thickness of these
-domains will vary as a function of the absorption cross section if the
-other parameters are fixed. To compute the optical thickness, along
-with the spectral radius as given by Eq~(\ref{eq:iteration_radius}),
-we also need the parameters $n_i$ and $n_s$ which respectively
-describe the typical domain length and the average number of states
-moved along that typical length per history transition. For our grid
-above, the domains are varied in size with $50 \times 50$, $100 \times
-100$, and $200 \times 200$ cells giving $n_i=50$, $n_i=100$, and
-$n_i=200$ grid points or states along the typical length of the domain
-respectively. Looking at the Laplacian stencil in
-Figure~\ref{fig:stencil}, we see that all history transitions will
-only move a single state in either the $i$ or $j$ directions due to
-the symmetry of the problem. Furthermore, if we choose the $i$
-direction, not all states we will transition to will move the history
-in that direction. Therefore, we look to the definition of the
-iteration matrix in Eq~(\ref{eq:iteration_stencil}) and the definition
-of the adjoint probability matrix in Eq~(\ref{eq:adjoint_probability})
-to estimate the $n_s$ parameter. For a particular transition starting
-at state $(i,j)$, 6 of the 8 possible new states in the stencil move
-the history in $i$ direction with
-giving $n_i = \frac{3}{5}$.
-
-To compute the leakage fraction numerically, over the global domain
-\sn{1}{5} histories were sampled from a uniform source of strength
-unity. At the start of a stage of histories, the number of histories
-starting in the domain was computed and as the stage progressed, the
-number of histories that exited the domain was tallied with the ratio
-of the two numbers providing a numerical measure for the leakage
-fraction. Figure~\ref{fig:measured_length} gives the domain leakage
-measurements for the domain in the center of the global grid as well
-as the analytic result computed by
-Eq~(\ref{eq:analytic_domain_leakage}) as a function of the absorption
-cross section. Again, we note good qualitative agreement between the
-measured and analytic quantities.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=5in,clip]{leakage_variation.png}
-  \end{center}
-  \caption{\textbf{Measured and analytic domain leakage as a function
-      of the iteration matrix spectral radius.} \textit{To test the
-      behavior with respect to domain size, $n_i=50$, $n_i=100$,and
-      $n_i=200$ were used. The red data was computed numerically by a
-      domain-decomposed adjoint Neumann-Ulam implementation, the black
-      fine-dashed data was generated by
-      Eq~(\ref{eq:mean_chord_domain_leakage}) using the mean-chord
-      approximation, and the dashed-dotted black data was generated by
-      Eq~(\ref{eq:wigner_domain_leakage}) using the Wigner rational
-      approximation.}}
-  \label{fig:measured_leakage}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=5in,clip]{leakage_error.png}
-  \end{center}
-  \caption{\textbf{Measured and analytic domain leakage absolute error
-      as a function of the iteration matrix spectral radius.}
-    \textit{To test the behavior with respect to domain size, $n_i=50$
-      (blue), $n_i=100$ (red), and $n_i=200$ (yellow) were used. The
-      dashed lines represent the error using the Wigner rational
-      approximation while the solid lines represent the error using
-      the mean-chord approximation.}}
-  \label{fig:measured_leakage}
-\end{figure}
+\section{Results}
 
@@
 \section{Conclusion}
-These relationships will serve as guidelines for selecting an
-appropriate parallel algorithm strategy and provide a basis for
-performance models for future parallel implementations of the adjoint
-Neumann-Ulam method.
 
@@
 \bibliographystyle{ieeetr}
-\bibliography{references}
+\bibliography{paper}
 \end{document}

### 2012-12-10 doc/sna2013/abstract/sna_abstract.tex
sna abstract update
adding makefile to mandc 2013

@@
 \usepackage{amsmath}
+\usepackage{tabls}
+\usepackage{afterpage}
+\usepackage{amssymb}
+\usepackage{amsthm}
+\usepackage[mathcal]{euscript}
 \usepackage[usenames]{color}
@@
 %%---------------------------------------------------------------------------%%
-\author{Stuart R. Slattery
+\author{Stuart R. Slattery, Thomas M. Evans, Paul P.H. Wilson
 \\ \href{mailto:sslattery@wisc.edu}{\texttt{sslattery@wisc.edu}}
@@
 
-\date{\today} \title{An Overlapping-Domain Decomposed Monte Carlo
+\date{\today} 
+\title{A Multiple-Set Overlapping-Domain Decomposed Monte Carlo
   Synthetic Acceleration Method for Linear Systems} 
@@
 
+\abstract
+{
+We present a novel multiple-set overlapping-domain decomposed strategy
+for parallelizing the Monte Carlo Synthetic Acceleration (MCSA)
+method. MCSA methods use the Neumann-Ulam class of Monte Carlo solvers
+for linear systems to accelerate the fixed point method upon which
+the Neumann-Ulam solvers are based. To effectively parallelize MCSA
+methods requires the parallelization of the Neumann-Ulam solvers. To
+do this in a domain decomposed environment, we borrow strategies
+traditionally implemented in reactor physics to parallelize the
+problem. The multiple-set overlapping-domain decomposition algorithm
+is presented along with preliminary scaling results.
+}
+
 %%---------------------------------------------------------------------------%%
 \section{Introduction}
+As leadership class machines move towards the exascale, new algorithms
+must be developed that leverage their strengths and adapt to their
+shortcomings. Basic research is required now to advance methods in
+time for these new machines to become operational. We can identify key
+issues for which a set of robust, massively parallel Monte Carlo
+solvers could provide a solution. As machines begin to operate at
+hundreds of petaflops peak performance and beyond, trends toward
+reduced energy consumption will require incredibly high levels of
+concurrency to achieve the desired computation rates. Furthermore,
+this drop in power consumption will mean increased pressure on memory
+as memory per node is expected to stagnate while cores per node is
+expected to increase. As the number of cores increases, their clock
+speed is expected to stagnate or even decrease to further reduce power
+consumption and manufacturing costs.
+
+The end result of these hardware changes is that the larger numbers of
+low-powered processors will be prone to both soft failures such as bit
+errors in floating point operations and hard failures where the data
+owned by that processor cannot be recovered. Because these failures
+are predicted to be common, resilient solver technologies are required
+to overcome these events. With linear solvers based on Monte Carlo
+techniques, such issues are alleviated by statistical arguments. In
+the case of soft failures, isolated floating point errors in Monte
+Carlo simulation are absorbed within tally statistics while completely
+losing hardware during a hard failure is manifested as a high variance
+event where some portion of the Monte Carlo histories are lost.
 
@@
 
+An alternative approach to approximate matrix inversion is to employ
+Monte Carlo methods that sample a distribution with an expectation
+value equivalent to that of the inverted operator. Such methods have
+been in existence for decades with the earliest reference noted here
+an enjoyable manuscript published in 1950 by Forsythe and Leibler
+\cite{forsythe_matrix_1950}. In their outline, Forsythe and Liebler
+in fact credit the creation of this technique to J. Von Neumann and
+S.M. Ulam some years earlier than its publication. In 1952 Wasow
+provided a more formal explanation of Von Neumann and Ulam's method
+\cite{wasow_note_1952} and Hammersley and Handscomb's 1964 monograph
+\cite{hammersley_monte_1964} and Spanier and Gelbard's 1969 book
+\cite{spanier_monte_1969} present additional detail on this topic
+using a collection of references from the 1950's and early 1960's.
+
+Using the ideas of Halton, Evans and Mosher recently developed a Monte
+Carlo solution method that was not prohibited severely by the quality
+of the initial guess for the system \cite{evans_monte_2009} and later
+applied it more rigorously as a solution mechanism for the radiation
+diffusion equation \cite{evans_monte_2012}. With their new methods,
+they achieved identical numerical results as and marginally better
+performance than conventional Krylov solvers. Their approach was
+instead to use residual Monte Carlo as a synthetic acceleration for a
+stationary method. The \textit{Fixed-Point Monte Carlo
+  Synthetic-Acceleration} (MCSA) method is defined as:
+\begin{subequations}
+  \begin{gather}
+    \ve{x}^{k+1/2} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,\\
+    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\
+    \ve{A}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:,\\
+    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,
+  \end{gather}
+  \label{eq:mcsa}
+\end{subequations}
+where the adjoint Monte Carlo method is used to generate the solution
+correction from the residual. Using Monte Carlo in this way achieves
+the same effect as Halton's method, decoupling its convergence rate
+from the overall convergence rate of the method. Here, the approximate
+Monte Carlo solution is not driven to a particular convergence as it
+merely supplies a correction for the initial guess generated by
+Richardson's iteration. Rather, only a set number of histories are
+required using the adjoint method to generate the correction.
+
 %%---------------------------------------------------------------------------%%
-\section{Domain Decomposition and Overlap}
+\section{Multiple-Set Overlapping-Domain Decomposition}
+
+In 2010, Wagner and colleagues developed the \textit{multiple-set
+  overlapping-domain} (MSOD) decomposition for parallel Monte Carlo
+applications for full-core light water reactor analysis
+\cite{wagner_hybrid_2010}. In their work, an extension of Brunner's,
+their scheme employed the similar parallel algorithms for particle
+transport but a certain amount of overlap between adjacent domains was
+used to decrease the number of particles leaving the local domain. In
+addition, Wagner utilized a level of replication of the domain such
+that the domain was only decomposed on $O(100)$ processors and if
+replicated $O(1,000)$ times achieves simulation on $O(100,000)$
+processors, thus providing spatial and particle parallelism. Each
+collection of processors that constitutes a representation of the
+entire domain is referred to as a set, and within a set overlap occurs
+among its sub-domains. The original motivation was to decompose the
+domain in a way that it remained in a physical cabinet in a large
+distributed machine, thus reducing latency costs during
+communication. A multiple set scheme is also motivated by the fact
+that communication during particle transport only occurs within a set,
+limiting communications during the transport procedure to a group of
+$O(100)$ processors, a number that was shown to have excellent
+parallel efficiencies in Brunner's work and therefore will scale well
+in this algorithm. The overlapping domains within each set also
+demonstrated reduced communication costs. On each processor, the
+source is sampled in the local domain that would exist if no overlap
+was used while tallies can be made over the entire overlapping domain.
 

### 2012-12-10 doc/sna2013/abstract/sna_abstract.tex
adding sna content

@@
-\documentclass[letterpaper,12pt]{article}
-\usepackage[top=1.0in,bottom=1.0in,left=1.25in,right=1.25in]{geometry}
+\documentclass[letterpaper,10pt]{article}
+\usepackage[top=1.0in,bottom=1.0in,left=0.5in,right=0.5in]{geometry}
 \usepackage{verbatim}
@@
 \usepackage[mathcal]{euscript}
+\usepackage{tmadd,tmath}
 \usepackage[usenames]{color}
@@
 \abstract
-{
 We present a novel multiple-set overlapping-domain decomposed strategy
@@
 problem. The multiple-set overlapping-domain decomposition algorithm
-is presented along with preliminary scaling results.
-}
+is presented along with preliminary parallel scaling results.
 
@@
 must be developed that leverage their strengths and adapt to their
-shortcomings. Basic research is required now to advance methods in
-time for these new machines to become operational. We can identify key
-issues for which a set of robust, massively parallel Monte Carlo
-solvers could provide a solution. As machines begin to operate at
-hundreds of petaflops peak performance and beyond, trends toward
-reduced energy consumption will require incredibly high levels of
-concurrency to achieve the desired computation rates. Furthermore,
-this drop in power consumption will mean increased pressure on memory
-as memory per node is expected to stagnate while cores per node is
-expected to increase. As the number of cores increases, their clock
-speed is expected to stagnate or even decrease to further reduce power
-consumption and manufacturing costs.
-
-The end result of these hardware changes is that the larger numbers of
-low-powered processors will be prone to both soft failures such as bit
-errors in floating point operations and hard failures where the data
-owned by that processor cannot be recovered. Because these failures
-are predicted to be common, resilient solver technologies are required
-to overcome these events. With linear solvers based on Monte Carlo
-techniques, such issues are alleviated by statistical arguments. In
-the case of soft failures, isolated floating point errors in Monte
-Carlo simulation are absorbed within tally statistics while completely
-losing hardware during a hard failure is manifested as a high variance
-event where some portion of the Monte Carlo histories are lost.
+shortcomings. Looking towards machines begin operating at hundreds of
+petaflops peak performance and beyond, trends toward reduced energy
+consumption will require incredibly high levels of concurrency to
+achieve the desired computation rates \cite{kogge_using_2011}. The end
+result of these hardware changes is that the larger numbers of
+low-powered processors will be prone to both soft and hard failures
+such as bit errors in floating point operations and hard failures with
+resilient solver technologies are required to overcome these
+events. With linear solvers based on Monte Carlo techniques, such
+issues are alleviated by statistical arguments. In the case of soft
+failures, isolated floating point errors in Monte Carlo simulation are
+absorbed within tally statistics while completely losing hardware
+during a hard failure is manifested as a high variance event where
+some portion of the Monte Carlo histories are lost.
 
@@
 an enjoyable manuscript published in 1950 by Forsythe and Leibler
-\cite{forsythe_matrix_1950}. In their outline, Forsythe and Liebler
-in fact credit the creation of this technique to J. Von Neumann and
-S.M. Ulam some years earlier than its publication. In 1952 Wasow
-provided a more formal explanation of Von Neumann and Ulam's method
-\cite{wasow_note_1952} and Hammersley and Handscomb's 1964 monograph
-\cite{hammersley_monte_1964} and Spanier and Gelbard's 1969 book
-\cite{spanier_monte_1969} present additional detail on this topic
-using a collection of references from the 1950's and early 1960's.
+\cite{forsythe_matrix_1950}. In their outline, Forsythe and Liebler in
+fact credit the creation of this technique to J. Von Neumann and
+S.M. Ulam some years earlier than its publication. Hammersley and
+Handscomb's 1964 monograph \cite{hammersley_monte_1964} and Spanier
+and Gelbard's 1969 book \cite{spanier_monte_1969} present additional
+detail on this topic using a collection of references from the 1950's
+and early 1960's.
 
-Using the ideas of Halton, Evans and Mosher recently developed a Monte
-Carlo solution method that was not prohibited severely by the quality
-of the initial guess for the system \cite{evans_monte_2009} and later
-applied it more rigorously as a solution mechanism for the radiation
-diffusion equation \cite{evans_monte_2012}. With their new methods,
-they achieved identical numerical results as and marginally better
-performance than conventional Krylov solvers. Their approach was
-instead to use residual Monte Carlo as a synthetic acceleration for a
-stationary method. The \textit{Fixed-Point Monte Carlo
+Using the ideas of Halton's residual method
+\cite{halton_sequential_1994}, Evans and Mosher recently developed a
+Monte Carlo solution method that was not prohibited severely by the
+quality of the initial guess for the system \cite{evans_monte_2009}
+and later applied it more rigorously as a solution mechanism for the
+radiation diffusion equation \cite{evans_monte_2012}. With their new
+methods, they achieved identical numerical results as and marginally
+better performance than conventional Krylov solvers. Their approach
+was instead to use residual Monte Carlo as a synthetic acceleration
+for a stationary method. The \textit{Fixed-Point Monte Carlo
   Synthetic-Acceleration} (MCSA) method is defined as:
@@
     \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\
-    \ve{A}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:,\\
+    \hat{\ve{A}}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:,
+    \label{eq:residual_correction}\\
     \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,
@@
 \end{subequations}
-where the adjoint Monte Carlo method is used to generate the solution
-correction from the residual. Using Monte Carlo in this way achieves
-the same effect as Halton's method, decoupling its convergence rate
-from the overall convergence rate of the method. Here, the approximate
-Monte Carlo solution is not driven to a particular convergence as it
-merely supplies a correction for the initial guess generated by
-Richardson's iteration. Rather, only a set number of histories are
-required using the adjoint method to generate the correction.
+where the adjoint Neumann-Ulam Monte Carlo method is used to generate
+the solution correction from the residual in
+Eq~(\ref{eq:residual_correction}). Using Monte Carlo in this way
+achieves the same effect as Halton's method, decoupling its
+convergence rate from the overall convergence rate of the
+method. Here, the approximate Monte Carlo solution is not driven to a
+particular convergence as it merely supplies a correction for the
+initial guess generated by Richardson's iteration. Rather, only a set
+number of histories are required using the adjoint method to generate
+the correction.
 
@@
 
-In 2010, Wagner and colleagues developed the \textit{multiple-set
-  overlapping-domain} (MSOD) decomposition for parallel Monte Carlo
-applications for full-core light water reactor analysis
-\cite{wagner_hybrid_2010}. In their work, an extension of Brunner's,
-their scheme employed the similar parallel algorithms for particle
-transport but a certain amount of overlap between adjacent domains was
-used to decrease the number of particles leaving the local domain. In
-addition, Wagner utilized a level of replication of the domain such
-that the domain was only decomposed on $O(100)$ processors and if
-replicated $O(1,000)$ times achieves simulation on $O(100,000)$
-processors, thus providing spatial and particle parallelism. Each
-collection of processors that constitutes a representation of the
-entire domain is referred to as a set, and within a set overlap occurs
-among its sub-domains. The original motivation was to decompose the
-domain in a way that it remained in a physical cabinet in a large
-distributed machine, thus reducing latency costs during
-communication. A multiple set scheme is also motivated by the fact
-that communication during particle transport only occurs within a set,
-limiting communications during the transport procedure to a group of
-$O(100)$ processors, a number that was shown to have excellent
-parallel efficiencies in Brunner's work and therefore will scale well
-in this algorithm. The overlapping domains within each set also
-demonstrated reduced communication costs. On each processor, the
-source is sampled in the local domain that would exist if no overlap
-was used while tallies can be made over the entire overlapping domain.
+At the 2010 SNA+MC meeting, Wagner and colleagues developed the
+\textit{multiple-set overlapping-domain} (MSOD) decomposition for
+parallel Monte Carlo applications for full-core light water reactor
+analysis \cite{wagner_hybrid_2010}. In their work, overlap between
+adjacent domains in the decomposition was used to decrease the number
+of particles leaving the local domain. In addition, Wagner utilized a
+level of replication of the domain such that the domain was only
+decomposed on $O(100)$ processors and if replicated $O(1,000)$ times
+achieves simulation on $O(100,000)$ processors, thus providing spatial
+and particle parallelism. Each collection of processors that
+constitutes a representation of the entire domain is referred to as a
+set, and within a set overlap occurs among its sub-domains. The
+original motivation was to decompose the domain in a way that it
+remained in a physical cabinet in a large distributed machine, thus
+reducing latency costs during communication. A multiple set scheme is
+also motivated by the fact that communication during particle
+transport only occurs within a set, limiting communications during the
+transport procedure to a group of $O(100)$ processors, a number that
+was shown to have excellent parallel efficiencies in Brunner's work
+and therefore will scale well in this algorithm. The overlapping
+domains within each set also demonstrated reduced communication costs.
 
@@
 %%---------------------------------------------------------------------------%%
+\section*{Acknowledgments}
+
+This work was performed under appointment to the Nuclear Regulatory
+Commission Fellowship program at the University of Wisconsin - Madison
+Engineering Physics Department.
+
+%%---------------------------------------------------------------------------%%
 \pagebreak
 \bibliographystyle{ieeetr}
-\bibliography{paper}
+\bibliography{references}
 \end{document}

### 2012-12-10 drivers/NeutronDiffusion/DiffusionProblem.cpp
adding periodic boundaries to the neutron diffusion problem

@@ -68,177 +68,240 @@ DiffusionProblem::DiffusionProblem( const RCP_Comm& comm,
     // Fill the operator from global data on proc 0.
     if ( comm->getRank() == 0 )
     {
-	// Min X boundary vacuum (nonreentrant current).
-	for ( int j = 1; j < N-1; ++j )
+
+	// Vacuum boundaries.
+	if ( plist->get<std::string>("BC TYPE") == "VACUUM" )
 	{
-	    int i = 0;
-
-	    idx[0]                = i + j*N;
-	    idx_iplus1[0]         = (i+1) + j*N;
-	    idx_jminus1[0]        = i + (j-1)*N;
-	    idx_jplus1[0]         = i + (j+1)*N;
-	    idx_iplus1jminus1[0]  = (i+1) + (j-1)*N;
-	    idx_iplus1jplus1[0]   = (i+1) + (j+1)*N;
-
-	    A->insertGlobalValues( idx[0], idx(),                diag() );
-	    A->insertGlobalValues( idx[0], idx_iplus1(),         iplus1  );
-	    A->insertGlobalValues( idx[0], idx_jminus1(),        jminus1()  );
-	    A->insertGlobalValues( idx[0], idx_jplus1(),         jplus1()  );
-	    A->insertGlobalValues( idx[0], idx_iplus1jminus1(),  iplus1jminus1() );
-	    A->insertGlobalValues( idx[0], idx_iplus1jplus1(),   iplus1jplus1() );
-	}
-
-	// Max X boundary vacuum (nonreentrant current).
-	for ( int j = 1; j < N-1; ++j )
-	{
-	    int i = N-1;
-
-	    idx[0]                = i + j*N;
-	    idx_iminus1[0]        = (i-1) + j*N;
-	    idx_jminus1[0]        = i + (j-1)*N;
-	    idx_jplus1[0]         = i + (j+1)*N;
-	    idx_iminus1jminus1[0] = (i-1) + (j-1)*N;
-	    idx_iminus1jplus1[0]  = (i-1) + (j+1)*N;
-
-	    A->insertGlobalValues( idx[0], idx(),                diag() );
-	    A->insertGlobalValues( idx[0], idx_iminus1(),        iminus1()  );
-	    A->insertGlobalValues( idx[0], idx_jminus1(),        jminus1()  );
-	    A->insertGlobalValues( idx[0], idx_jplus1(),         jplus1()  );
[…]

### 2012-12-10 drivers/NeutronDiffusion/DiffusionProblem.cpp
fixed index error in periodic conditions

@@ -245,7 +245,7 @@ DiffusionProblem::DiffusionProblem( const RCP_Comm& comm,
 	}
 
 	// Periodic boundaries.
-	if ( plist->get<std::string>("BC TYPE") == "PERIODIC" )
+	else if ( plist->get<std::string>("BC TYPE") == "PERIODIC" )
 	{
 	    for ( int i = 0; i < N; ++i )
 	    {
@@ -288,6 +288,26 @@ DiffusionProblem::DiffusionProblem( const RCP_Comm& comm,
 			idx_jplus1[0]         = i;
 			idx_iminus1jplus1[0]  = (i-1);
 			idx_iplus1jplus1[0]   = (i+1);
+		    }
+
+		    if ( i == 0 && j == 0 )
+		    {
+			idx_iminus1jminus1[0] = (N-1) + (N-1)*N;
+		    }
+
+		    if ( i == 0 && j == N-1 )
+		    {
+			idx_iminus1jplus1[0] = (N-1);			
+		    }
+
+		    if ( i == N-1 && j == 0 )
+		    {
+			idx_iplus1jminus1[0] = (N-1)*N;
+		    }
+
+		    if ( i == N-1 && j == N-1 )
+		    {
+			idx_iplus1jplus1[0] = 0;
 		    }
 
 		    A->insertGlobalValues( idx[0], idx(),                diag() );

### 2012-12-11 doc/mandc2013/theory_paper/slattery_wilson_evans.tex
fixed subsection headings
mandc updates
fixed typo in mandc paper

@@
 
-\emph{Key Words}: Monte Carlo, domain decomposition, MCSA
+\emph{Key Words}: Monte Carlo, domain decomposition, MCSA, parallel
+computing
 \end{small} 
@@
 \end{equation}
-such that we expect to select a new state $j$ from the current state
+such that we expect to select a new state $i$ from the current state
 in the random walk $j$ by sampling column-wise.  Using our result from
@@
 
-\subsection{Iteration Matrix Spectrum}
+\Subsection{Iteration Matrix Spectrum}
 \label{subsec:iteration_spectrum}
@@
 
-\subsection{Neumann Series Convergence}
+\Subsection{Neumann Series Convergence}
 \label{subsec:neumann_convergence}
@@
 
-\subsection{Domain Leakage Approximations}
+\Subsection{Domain Leakage Approximations}
 \label{subsec:domain_leak_approx}
@@
 
-\subsection{Random Walk Length}
+\Subsection{Random Walk Length}
 \label{subsec:walk_length}
@@
 
-\subsection{Domain Leakage}
+\Subsection{Domain Leakage}
 \label{subsec:domain_leakage}
@@
 %%---------------------------------------------------------------------------%%
-\section*{ACKNOWLEDGMENTS}
+\section*{ACKNOWLEDGEMENTS}
 

### 2012-12-11 doc/mandc2013/theory_paper/slattery_wilson_evans.tex
updates to mandc doc

@@
 Eq~(\ref{eq:adjoint_neumann_series}) approximated by the Monte Carlo
-solver is dependent on the Eigenvalues of the iteration matrix. In
-addition, the Eigenvalues of the linear operator from which the
-iteration matrix was derived allow for further analysis of convergence
-properties. We will compute the Eigenvalues of these matrices by
-assuming Eigenfunctions of the form \cite{leveque_finite_2007}:
+solver is dependent on the Eigenvalues of the iteration matrix. We
+will compute these Eigenvalues by assuming Eigenfunctions of the form
+\cite{leveque_finite_2007}:
 \begin{equation}
@@
 \end{equation}
-where $h$ is the size of the grid along the chord and $n_s$ is the
-number of grid elements a history will move on average every
-transition. For our diffusion model problem, $n_s$ would equate to the
-number of states in the $i$ (or $j$ as the problem is symmetric)
-direction that a history could potentially move in a single transition
-and is dependent on the stencil used for the discretization. After $k$
-transitions in the random walk, the history will have moved a mean
-squared distance of:
+where $h$ is the size of the discrete grid elements along the chord
+and $n_s$ is the number of grid elements a history will move on
+average every transition. For our diffusion model problem, $n_s$ would
+equate to the number of states in the $i$ (or $j$ as the problem is
+symmetric) direction that a history could potentially move in a single
+transition and is dependent on the stencil used for the
+discretization. After $k$ transitions in the random walk, the history
+will have moved a mean squared distance of:
 \begin{equation}
@@
 \end{equation}
-If our our chord is of length $l$ and there are $n_i$ grid points (or
-states to which a history may transition) along that chord, then $h =
-l / n_i$ giving:
+If our our chord is of length $l$ and there are $n_i$ grid elements
+(or states to which a history may transition) along that chord, then
+$h = l / n_i$ giving:
 \begin{equation}

### 2012-12-11 doc/mandc2013/theory_paper/slattery_wilson_evans.tex
updated mandc paper for typos

@@
 \newcommand{\shortTitle}      % Short title here
-   {Domain Decomposed Neumann-Ulam Method}  
+   {Domain Decomposed Neumann-Ulam Method Analysis}  
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 The domain decomposed behavior of the adjoint Neumann-Ulam Monte Carlo
-method for solving linear operator equations is analyzed using the
-spectral properties of the linear operator. Relationships for the
-average length of the adjoint random walks, a measure of convergence
-speed and serial performance, are made with respect to the Eigenvalues
-and other properties of the linear operator. In addition,
-relationships for the effective optical thickness of a domain in the
-decomposition are presented based on the spectral analysis and
-diffusion theory. Using the effective optical thickness, the Wigner
-rational approximation and the mean chord approximation are applied to
-estimate the leakage fraction of stochastic histories from a domain in
-the decomposition as a measure of parallel performance and potential
-communication costs. The one-speed, two-dimensional neutron diffusion
-equation is used as a model problem to test the models for symmetric
-operators. In general, the derived approximations show good agreement
-with measured computational results.
+method for solving linear systems is analyzed using the spectral
+properties of the linear operator. Relationships for the average
+length of the adjoint random walks, a measure of convergence speed and
+serial performance, are made with respect to the Eigenvalues of the
+linear operator. In addition, relationships for the effective optical
+thickness of a domain in the decomposition are presented based on the
+spectral analysis and diffusion theory. Using the effective optical
+thickness, the Wigner rational approximation and the mean chord
+approximation are applied to estimate the leakage fraction of
+stochastic histories from a domain in the decomposition as a measure
+of parallel performance and potential communication costs. The
+one-speed, two-dimensional neutron diffusion equation is used as a
+model problem to test the models for symmetric operators. In general,
+the derived approximations show good agreement with measured
+computational results.
 
 \emph{Key Words}: Monte Carlo, domain decomposition, MCSA, parallel
-computing
+computing, linear solvers
 \end{small} 
@@
 
-To enable higher fidelity reactor physics Monte Carlo simulations,
-domain decomposition has been identified as a key principle in moving
-forward in high performance computing
+To date, parallel Neumann-Ulam methods have been limited to full
+domain replication with parallelism exploited through individual
+histories \cite{alexandrov_efficient_1998}. In order to enable higher
+fidelity reactor physics Monte Carlo simulations, domain decomposition
+has been identified as a key principle in moving forward in high
+performance computing
 \cite{brunner_comparison_2006,siegel_analysis_2012}. To accomplish
@@
 The purpose of this study is to provide a simple, analytic theory
-based on the properties of the linear system the will allow for
+based on the properties of the linear system that will allow for
 estimates of the domain decomposed behavior of the adjoint
@@
 Carlo method followed by a presentation of the model problem. Using
-the linear system generated by this discretization, we use a spectral
-theory to generate analytic relations for the Eigenvalues of the
-operator based on system parameters. Using the Eigenvalue spectra, we
-then build relationships to characterize the transport of stochastic
-histories in a decomposed domain and the fraction of histories that
-leak from a domain and will therefore have to be
+the linear system generated by the discretization of this, we use a
+spectral analysis to generate analytic relations for the Eigenvalues
+of the operator based on system parameters. Using the Eigenvalue
+spectra, we then build relationships to characterize the transport of
+stochastic histories in a decomposed domain and the fraction of
+histories that leak from a domain and will therefore have to be
 communicated. Finally, we compare these analytic results to numerical
@@
 In order to perform the random walks required to approximate the
-summation in Eq~(\ref{eq:adjoint_neumann_solution}), we build
-transition probability and weight matrices where probabilities are
-column-scaled and the transition weight is defined as:
+summation in Eq~(\ref{eq:adjoint_neumann_solution}), we build the
+column-scaled transition probability matrix and associated transition
+weights as:
 \begin{equation}
@@
 \end{equation}
-such that we expect to select a new state $i$ from the current state
-in the random walk $j$ by sampling column-wise.  Using our result from
+such that we expect to select a new state, $j$, from the current
+state, $i$, in the random walk by sampling the $i^{th}$ row of the
+probablity matrix as a discrete distribution.  Using our result from
 Eq~(\ref{eq:inner_product_constraint}) generated by applying the
-adjoint constraints, the contribution to the solution in state $i$
-from a particular random walk permutation is then:
+adjoint constraints, the Monte Carlo estimator for the solution in
+state $i$ for a particular random walk permutation is then:
 \begin{equation}
@@
 where the Kronecker delta indicates that the tally contributes only in
-the current state and $b_{i_0}$ will be the sampled source starting
-weight. Finally, the expectation value using all permutations is:
+the current state and $W_m$ is the history weight on the $m^{th}$
+step. Finally, the expectation value using all permutations is:
 \begin{equation}
@@
 \end{equation}
-therefore also providing an unbiased Monte Carlo estimate of the
-solution.
+where $b_{i_0}$ will be the sampled source starting weight, therefore
+also providing an unbiased Monte Carlo estimate of the solution.
 
@@
 then the action of the linear operator on these Eigenfunctions should
-give the Eigenvalues of the matrix as they lie on the unit circle in
+give the Eigenvalues of the matrix as they exist on the unit circle in
 the complex plane.
@@
 \begin{equation}
-  \ve{H}\boldsymbol{\phi} = -\frac{\alpha D}{6h^2}[4 \phi_{i-1,j}
+  \ve{H}\boldsymbol{\phi} = \frac{\alpha D}{6h^2}[4 \phi_{i-1,j}
     + 4 \phi_{i+1,j} + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} +
@@
 \begin{multline}
-  \lambda_{p,q}(\ve{H}) = -\frac{\alpha D}{6h^2}\Big[4 e^{-2 \pi \imath p
+  \lambda_{p,q}(\ve{H}) = \frac{\alpha D}{6h^2}\Big[4 e^{-2 \pi \imath p
       h} + 4 e^{2 \pi \imath p h} + 4 e^{-2 \pi \imath q h} + 4 e^{2
@@
 \begin{equation}
-  \lambda_{p,q}(\ve{H}) = -\frac{\alpha D}{6h^2}[ 8 \cos(\pi p h) + 8
+  \lambda_{p,q}(\ve{H}) = \frac{\alpha D}{6h^2}[ 8 \cos(\pi p h) + 8
     \cos(\pi q h) + 4 \cos(\pi p h) \cos(\pi q h)]\:,
@@
 giving the Eigenvalue spectrum for the Jacobi preconditioned iteration
-matrix. We expect the maximum Eigenvalue to exist when $p=q=0$, giving
-the following for the spectral radius of the Jacobi preconditioned
-iteration matrix:
+matrix. We find that the maximum Eigenvalue exists when $p=q=0$,
+giving the following for the spectral radius of the Jacobi
+preconditioned iteration matrix:
 \begin{equation}
@@
 Eq~(\ref{eq:adjoint_neumann_solution})
-\cite{dimov_new_1998,halton_sequential_1994,danilov_asymptotic_2000}. This
-random walk length, or the number of transitions before the
-termination of a history (either by the weight cutoff, absorption, or
-exiting the domain) is therefore approximately the number of
-stationary iterations required to converge to the specified
-tolerance. In the case of the adjoint Neumann-Ulam method, no such
-tolerance exists, however, we have specified a weight cutoff $W_c$
-under which low-weight histories will be prematurely terminated as
-their contributions are deemed minute. After $k$ iterations, a
-stationary method is terminated as the error has reached some
-fraction, $\epsilon$, of the initial error:
+\cite{dimov_new_1998,danilov_asymptotic_2000}. This random walk
+length, or the number of transitions before the termination of a
+history (either by the weight cutoff, absorption, or exiting the
+global domain) is therefore approximately the number of stationary
+iterations required to converge to the specified tolerance. In the
+case of the adjoint Neumann-Ulam method, no such tolerance exists,
+however, we have specified a weight cutoff $W_c$ under which
+low-weight histories will be prematurely terminated as their
+contributions are deemed minute. After $k$ iterations, a stationary
+method is terminated as the error has reached some fraction,
+$\epsilon$, of the initial error:
 \begin{equation}
@@
 in the grid by computing the mean squared distance of its movement
-along the chord defined across the domain of length $l$. After a
+along the chord of length $l$ defined across the domain. After a
 single transition a history will have moved a mean squared distance
@@
 Carlo history in the domain. We can readily interpret $\tau$ to be the
-\textit{effective optical thickness} of the domain of length
+\textit{effective optical thickness} of a domain of length
 $l$. Inserting Eq~(\ref{eq:step_k_length_sub}) we arrive at:
@@
     \end{center}
-    \caption{\textbf{Measured and analytic preconditioned diffusion
-        operator spectral radius as a function of the absorption cross
-        section to scattering cross section ratio.} \textit{Values of
-        $h=0.01$, $h=0.1$, and $h=1.0$ were used. The red data was
-        computed numerically by an Eigensolver while the black dashed
-        data was generated by Eq~(\ref{eq:iteration_radius}).}}
+    \caption{Measured and analytic preconditioned diffusion operator
+      spectral radius as a function of the absorption cross section to
+      scattering cross section ratio. Values of $h=0.01$, $h=0.1$, and
+      $h=1.0$ were used. The red data was computed numerically by an
+      Eigensolver while the black dashed data was generated by
+      Eq~(\ref{eq:iteration_radius}).}
     \label{fig:measured_spec_rad}
@@
 weight cutoff values of \sn{1}{-2}, \sn{1}{-4}, and \sn{1}{-8} were
-used with 10,000 histories generated by a point source source of
-strength 1 in the center of the domain such that the boundary
-conditions will have less of an effect on the length of the random
-walks. For each of the histories, the number of transitions made was
-tallied to provide an effective value of $k$ for each history. This
-value was then averaged over all histories to get a measured value of
-$k$ for the particular operator. On the left,
-figure~\ref{fig:measured_length} these measurements as well as the
-analytic result computed by Eq~(\ref{eq:analytic_k}) as a function of
-the iteration matrix spectral radius, $\rho(\ve{H})$. On the right,
-figure~\ref{fig:measured_length} gives the relative error between the
-predicted and observed results. We note good qualitative agreement
-between the measured and analytic results. However, we observe a
-larger relative error for both long and short random walks.
+used with 10,000 histories generated by a point source of strength 1
+in the center of the domain. For each of the histories, the number of
+transitions made was tallied to provide an effective value of $k$ for
+each history. This value was then averaged over all histories to get a
+measured value of $k$ for the particular operator. On the left,
+Figure~\ref{fig:measured_length} presents these measurements as well
+as the analytic result computed by Eq~(\ref{eq:analytic_k}) as a
+function of the iteration matrix spectral radius, $\rho(\ve{H})$. On
+the right, Figure~\ref{fig:measured_length} gives the relative error
+between the predicted and observed results. We note good qualitative
+agreement between the measured and analytic results. However, we
+observe a larger relative error for both long and short random walks.
 \begin{figure}[ht!]
@@
     \end{center}
-    \caption{\textbf{Measured and analytic random walk length as a
-        function of the iteration matrix spectral radius.} \textit{The
-        weight cutoff was varied with \sn{1}{-2}, \sn{1}{-4}, and
-        \sn{1}{-8}. In the left plot, the red data was computed
-        numerically by an adjoint Neumann-Ulam implementation while
-        the black dashed data was generated by
-        Eq~(\ref{eq:analytic_k}). In the right plot, the relative
-        error between the predicted and measured results is
-        presented.}}
+    \caption{Measured and analytic random walk length as a function of
+      the iteration matrix spectral radius. The weight cutoff was
+      varied with \sn{1}{-2}, \sn{1}{-4}, and \sn{1}{-8}. In the left
+      plot, the red data was computed numerically by an adjoint
+      Neumann-Ulam implementation while the black dashed data was
+      generated by Eq~(\ref{eq:analytic_k}). In the right plot, the
+      relative error between the predicted and measured results is
+      presented for each weight cutoff.}
     \label{fig:measured_length}
@@
 varied from 1 to 100 while the scattering cross section was set to
-zero to create a purely absorbing environment. The optical thickness
-of these domains will vary as a function of the absorption cross
-section if the other parameters are fixed. To compute the optical
-thickness, along with the spectral radius as given by
-Eq~(\ref{eq:iteration_radius}), we also need the parameters $n_i$ and
-$n_s$ which respectively describe the typical domain length and the
-average number of states moved along that typical length per history
-transition. For our grid above, the domains are varied in size with
-$50 \times 50$, $100 \times 100$, and $200 \times 200$ cells giving
-$n_i=50$, $n_i=100$, and $n_i=200$ grid points or states along the
-typical length of the domain respectively with weight cutoff of
-\sn{1}{-4}. Looking at the Laplacian stencil in
+zero to create a purely absorbing environment with weight cutoff of
+\sn{1}{-4}. The optical thickness of these domains will vary as a
+function of the absorption cross section if the other parameters are
+fixed. To compute the optical thickness, along with the spectral
+radius as given by Eq~(\ref{eq:iteration_radius}), we also need the
+parameters $n_i$ and $n_s$ which respectively describe the typical
+domain length and the average number of states moved along that
+typical length per history transition. For our grid above, the domains
+are varied in size with $50 \times 50$, $100 \times 100$, and $200
+\times 200$ cells giving $n_i=50$, $n_i=100$, and $n_i=200$ grid
+points or states along the typical length of the domain
+respectively. Looking at the Laplacian stencil in
 Eq~(\ref{eq:nine_point_stencil}), we see that all history transitions
@@
     \end{center}
-    \caption{\textbf{Measured and analytic domain leakage as a
-        function of the iteration matrix spectral radius.} \textit{To
-        test the behavior with respect to domain size, $n_i=50$,
-        $n_i=100$,and $n_i=200$ were used. The red data was computed
-        numerically by a domain-decomposed adjoint Neumann-Ulam
-        implementation, the black dashed data was generated by
+    \caption{Measured and analytic domain leakage as a function of the
+      iteration matrix spectral radius. To test the behavior with
+      respect to domain size, $n_i=50$, $n_i=100$,and $n_i=200$ were
+      used. The red data was computed numerically by a
+      domain-decomposed adjoint Neumann-Ulam implementation, the black
+      dashed data was generated by
       Eq~(\ref{eq:mean_chord_domain_leakage}) using the mean-chord
-        approximation, and the dashed-dotted black data was generated
-        by Eq~(\ref{eq:wigner_domain_leakage}) using the Wigner
-        rational approximation.}}
+      approximation, and the dashed-dotted black data was generated by
+      Eq~(\ref{eq:wigner_domain_leakage}) using the Wigner rational
+      approximation.}
     \label{fig:measured_leakage}
@@
     \end{center}
-    \caption{\textbf{Measured and analytic domain leakage absolute error
-        as a function of the iteration matrix spectral radius.}
-      \textit{To test the behavior with respect to domain size, $n_i=50$
-        (green), $n_i=100$ (blue), and $n_i=200$ (red) were used. The
-        dashed lines represent the error using the Wigner rational
+    \caption{Measured and analytic domain leakage absolute error as a
+      function of the iteration matrix spectral radius.  To test the
+      behavior with respect to domain size, $n_i=50$ (green),
+      $n_i=100$ (blue), and $n_i=200$ (red) were used. The dashed
+      lines represent the error using the Wigner rational
       approximation while the solid lines represent the error using
-        the mean-chord approximation.}}
+      the mean-chord approximation.}
     \label{fig:leakage_error}
@@
 We have presented an analytic analysis of the domain decomposed
-behavior of the adjoint Neumann Ulam method for linear operator
-equations. Good agreement was observed between the derived analytic
-relationships and numerical results generated by a domain decomposed
-implementation of the adjoint method.
+behavior of the adjoint Neumann Ulam method for linear systems. Good
+agreement was observed for the derived analytic relationships for
+random walk length and leakage fraction when compared to numerical
+results generated by a domain decomposed implementation of the adjoint
+method.
 

### 2012-12-11 doc/sna2013/abstract/slattery_wilson_evans.tex
adding sna+mc template

--- 2012-12-10 doc/sna2013/abstract/sna_abstract.tex
@@
-\documentclass[letterpaper,10pt]{article}
-\usepackage[top=1.0in,bottom=1.0in,left=0.5in,right=0.5in]{geometry}
-\usepackage{verbatim}
-\usepackage{amssymb}
-\usepackage{graphicx}
-\usepackage{longtable}
-\usepackage{amsfonts}
-\usepackage{amsmath}
-\usepackage{tabls}
-\usepackage{afterpage}
-\usepackage{amssymb}
-\usepackage{amsthm}
-\usepackage[mathcal]{euscript}
-\usepackage{tmadd,tmath}
-\usepackage[usenames]{color}
-\usepackage[
-naturalnames = true, 
-colorlinks = true, 
-linkcolor = black,
-anchorcolor = black,
-citecolor = black,
-menucolor = black,
-urlcolor = blue
-]{hyperref}
+\documentclass{snamc2013}
 
-%%---------------------------------------------------------------------------%%
-\author{Stuart R. Slattery, Thomas M. Evans, Paul P.H. Wilson
-\\ \href{mailto:sslattery@wisc.edu}{\texttt{sslattery@wisc.edu}}
-}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\usepackage{graphicx}  % allows inclusion of graphics
+\usepackage{booktabs}  % nice rules (thick lines) for tables
+\usepackage{microtype} % improves typography for PDF
+\usepackage[breaklinks=true,colorlinks=true,linkcolor=black,citecolor=black]{hyperref}
 
-\date{\today} 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \title{A Multiple-Set Overlapping-Domain Decomposed Monte Carlo
   Synthetic Acceleration Method for Linear Systems}
+
+\author[1]{Stuart R. Slattery}
+\author[1]{Paul P.H. Wilson}
+\author[2]{Thomas M. Evans}
+
+\affil[1]{University of Wisconsin - Madison, Engineering Physics
+  Department, 1500 Engineering Drive, Madison, WI 53706}
+\affil[2]{Oak Ridge National Laboratory, One Bethel Valley Road, Oak
+  Ridge, TN 37830}
+
+\abstract{We present a novel multiple-set overlapping-domain
+  decomposed strategy for parallelizing the Monte Carlo Synthetic
+  Acceleration (MCSA) method. MCSA methods use the Neumann-Ulam class
+  of Monte Carlo solvers for linear systems to accelerate the fixed
+  point method upon which the Neumann-Ulam solvers are based. To
+  effectively parallelize MCSA methods requires the parallelization of
+  the Neumann-Ulam solvers. To do this in a domain decomposed
+  environment, we borrow strategies traditionally implemented in
+  reactor physics to parallelize the problem. The multiple-set
+  overlapping-domain decomposition algorithm will be presented in the
+  submission along with parallel scaling results.}
+
+\keywords{MCSA, Monte Carlo, domain decomposition, linear solvers,
+  parallel computing}
+
 \begin{document}
-\maketitle
-
-\abstract
-We present a novel multiple-set overlapping-domain decomposed strategy
-for parallelizing the Monte Carlo Synthetic Acceleration (MCSA)
-method. MCSA methods use the Neumann-Ulam class of Monte Carlo solvers
-for linear systems to accelerate the fixed point method upon which
-the Neumann-Ulam solvers are based. To effectively parallelize MCSA
-methods requires the parallelization of the Neumann-Ulam solvers. To
-do this in a domain decomposed environment, we borrow strategies
-traditionally implemented in reactor physics to parallelize the
-problem. The multiple-set overlapping-domain decomposition algorithm
-is presented along with preliminary parallel scaling results.
 
@@
 \section{Introduction}
+
 As leadership class machines move towards the exascale, new algorithms
 must be developed that leverage their strengths and adapt to their
-shortcomings. Looking towards machines begin operating at hundreds of
+shortcomings. Looking towards machines operating at hundreds of
 petaflops peak performance and beyond, trends toward reduced energy
@@
 was instead to use residual Monte Carlo as a synthetic acceleration
-for a stationary method. The \textit{Fixed-Point Monte Carlo
-  Synthetic-Acceleration} (MCSA) method is defined as:
-\begin{subequations}
-  \begin{gather}
-    \ve{x}^{k+1/2} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,\\
-    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\
-    \hat{\ve{A}}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:,
-    \label{eq:residual_correction}\\
-    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,
-  \end{gather}
-  \label{eq:mcsa}
-\end{subequations}
-where the adjoint Neumann-Ulam Monte Carlo method is used to generate
-the solution correction from the residual in
-Eq~(\ref{eq:residual_correction}). Using Monte Carlo in this way
-achieves the same effect as Halton's method, decoupling its
-convergence rate from the overall convergence rate of the
-method. Here, the approximate Monte Carlo solution is not driven to a
-particular convergence as it merely supplies a correction for the
-initial guess generated by Richardson's iteration. Rather, only a set
-number of histories are required using the adjoint method to generate
-the correction.
+for a stationary method. Using Monte Carlo in this way achieves the
+same effect as Halton's method, decoupling its convergence rate from
+the overall convergence rate of the method. Here, the approximate
+Monte Carlo solution is not driven to a particular convergence as it
+merely supplies a correction for the initial guess generated by
+Richardson's iteration. Rather, only a set number of histories are
+required using the adjoint method to generate the correction.
 
@@
 
-At the 2010 SNA+MC meeting, Wagner and colleagues developed the
+Current implementations of parallel Neumann-Ulam methods have been
+limited to full domain replication with parallelism exploited through
+individual histories \cite{alexandrov_efficient_1998}. At the 2010
+SNA+MC meeting, Wagner and colleagues developed the
 \textit{multiple-set overlapping-domain} (MSOD) decomposition for
@@
 was shown to have excellent parallel efficiencies in Brunner's work
-and therefore will scale well in this algorithm. The overlapping
-domains within each set also demonstrated reduced communication costs.
+\cite{brunner_efficient_2009} and therefore will scale well in this
+algorithm. The overlapping domains within each set also demonstrated
+reduced communication costs.
 
 %%---------------------------------------------------------------------------%%
-\section{Results}
+\section{Submission Outline}
+
+With our submission we will present the following: a presentation of
+the Neumann-Ulam and MCSA methods, a presentation of the MSOD scheme
+as implemented for MCSA, and the results of parallel scaling studies
+for the implementation.
 
 %%---------------------------------------------------------------------------%%
-\section{Conclusion}
-
-%%---------------------------------------------------------------------------%%
-\section*{Acknowledgments}
-
-This work was performed under appointment to the Nuclear Regulatory
-Commission Fellowship program at the University of Wisconsin - Madison
-Engineering Physics Department.
-
-%%---------------------------------------------------------------------------%%
-\pagebreak
-\bibliographystyle{ieeetr}
+\bibliographystyle{ans}
 \bibliography{references}

### 2012-12-11 doc/sna2013/abstract/slattery_wilson_evans.tex
finished first sna extended abstract draft

@@
   effectively parallelize MCSA methods requires the parallelization of
-  the Neumann-Ulam solvers. To do this in a domain decomposed
-  environment, we borrow strategies traditionally implemented in
-  reactor physics to parallelize the problem. The multiple-set
-  overlapping-domain decomposition algorithm will be presented in the
-  submission along with parallel scaling results.}
+  the underlying Neumann-Ulam solvers. To do this in a domain
+  decomposed environment, we borrow strategies traditionally
+  implemented in particle transport Monte Carlo to parallelize the
+  problem. The multiple-set overlapping-domain decomposition algorithm
+  will be presented in the submission along with parallel scaling data
+  for the resulting implementation.}
 
@@
 \section{Introduction}
+
+For some time, the particle transport community has been utilizing
+Monte Carlo methods for the solution of transport problems
+\cite{lewis_computational_1993}. The partial differential equation
+(PDE) community has focused on various deterministic methods for
+solutions to linear problems \cite{saad_iterative_2003}. In between
+these two areas are a not widely known group of Monte Carlo methods
+for solving sparse linear systems \cite{forsythe_matrix_1950,
+  hammersley_monte_1964, halton_sequential_1962,
+  halton_sequential_1994}. In recent years, these methods have been
+further developed for radiation transport problems in the form of
+Monte Carlo Synthetic-Acceleration (MCSA) \cite{evans_monte_2009,
+  evans_monte_2012} but have yet to be applied to more general sparse
+linear systems commonly generated by the computational physics
+community. Compared to other methods in this regime, MCSA offers three
+attractive qualities; (1) the linear problem operator need not be
+symmetric or positive-definite, thereby reducing preconditioning
+complexity, (2) the stochastic nature of the solution method provides
+a natural solution to the issue of resiliency, and (3) is amenable to
+parallelization using modern methods developed by the transport
+community \cite{wagner_hybrid_2010}. The development of MCSA as a
+general linear solver and the development of a parallel MCSA method
+will be new and unique features of this work, providing a framework
+with which other issues such as resiliency may be addressed in the
+future.
 
@@
 limited to full domain replication with parallelism exploited through
-individual histories \cite{alexandrov_efficient_1998}. At the 2010
-SNA+MC meeting, Wagner and colleagues developed the
+individual histories \cite{alexandrov_efficient_1998}. For high
+performance computing environments, domain decomposition strategies
+have been identified as important for Monte Carlo applications
+\cite{brunner_comparison_2006,siegel_analysis_2012}. In a domain
+decomposed Monte Carlo implementation, stochastic histories must be
+communicated between domains as the simulation progresses and they
+transition to locations outside of the local domain. In order to
+implement domain decomposition strategy for Monte Carlo solvers in a
+parallel environment, communication of these histories must occur
+between compute nodes owning neighboring pieces of the global domain.
+
+At the 2010 SNA+MC meeting, Wagner and colleagues developed the
 \textit{multiple-set overlapping-domain} (MSOD) decomposition for
@@
 
-With our submission we will present the following: a presentation of
-the Neumann-Ulam and MCSA methods, a presentation of the MSOD scheme
-as implemented for MCSA, and the results of parallel scaling studies
-for the implementation.
+With our submission we will present a multiple-set overlapping-domain
+decomposed implementation of the Monte Carlo Synthetic Acceleration
+method. This presentation will include an outline of the Neumann-Ulam
+and MCSA methods followed by a presentation of the MSOD scheme as
+implemented for MCSA. This implementation will be tested and the
+results of parallel scaling studies presented that demonstrate the
+achieved parallel performance.
 

### 2012-12-12 doc/sna2013/abstract/slattery_wilson_evans.tex
sna abstract updates

@@
 \author[1]{Stuart R. Slattery}
+\author[2]{Thomas M. Evans}
 \author[1]{Paul P.H. Wilson}
-\author[2]{Thomas M. Evans}
 
@@
   decomposed environment, we borrow strategies traditionally
-  implemented in particle transport Monte Carlo to parallelize the
+  implemented in Monte Carlo particle transport to parallelize the
   problem. The multiple-set overlapping-domain decomposition algorithm
@@
 symmetric or positive-definite, thereby reducing preconditioning
-complexity, (2) the stochastic nature of the solution method provides
-a natural solution to the issue of resiliency, and (3) is amenable to
-parallelization using modern methods developed by the transport
-community \cite{wagner_hybrid_2010}. The development of MCSA as a
-general linear solver and the development of a parallel MCSA method
-will be new and unique features of this work, providing a framework
-with which other issues such as resiliency may be addressed in the
-future.
+complexity and widening applicability, (2) the stochastic nature of
+the solution method provides a natural solution to the issue of
+resiliency, and (3) is amenable to parallelization using modern
+methods developed by the transport community
+\cite{wagner_hybrid_2010}. The development of MCSA as a general linear
+solver and the development of a parallel MCSA method will be new and
+unique features of this work, providing a framework with which other
+issues such as resiliency may be addressed in the future.
 
@@
 low-powered processors will be prone to both soft and hard failures
-such as bit errors in floating point operations and hard failures with
-resilient solver technologies are required to overcome these
+with resilient solver technologies are required to overcome these
 events. With linear solvers based on Monte Carlo techniques, such
@@
 \section{Monte Carlo Synthetic Acceleration}
-
-An alternative approach to approximate matrix inversion is to employ
-Monte Carlo methods that sample a distribution with an expectation
-value equivalent to that of the inverted operator. Such methods have
-been in existence for decades with the earliest reference noted here
-an enjoyable manuscript published in 1950 by Forsythe and Leibler
-\cite{forsythe_matrix_1950}. In their outline, Forsythe and Liebler in
-fact credit the creation of this technique to J. Von Neumann and
-S.M. Ulam some years earlier than its publication. Hammersley and
-Handscomb's 1964 monograph \cite{hammersley_monte_1964} and Spanier
-and Gelbard's 1969 book \cite{spanier_monte_1969} present additional
-detail on this topic using a collection of references from the 1950's
-and early 1960's.
 
@@
 
+Monte Carlo Synthetic Acceleration (MCSA) methods rely the
+Neumann-Ulam class of linear solver to accelerate the stationary
+method. These Monte Carlo methods work by sampling a distribution with
+an expectation value equivalent to that of the inverted
+operator. Neumann and Ulam's work was first published in 1950 by
+Forsythe and Leibler \cite{forsythe_matrix_1950} with a time frame for
+the creation of the method some years earlier than its publication. In
+the following decade, the books of Hammersley and Handscomb
+\cite{hammersley_monte_1964} and Spanier and Gelbard
+\cite{spanier_monte_1969} present additional detail on these solvers.
+
 %%---------------------------------------------------------------------------%%
@@
 
-Current implementations of parallel Neumann-Ulam methods have been
-limited to full domain replication with parallelism exploited through
-individual histories \cite{alexandrov_efficient_1998}. For high
-performance computing environments, domain decomposition strategies
-have been identified as important for Monte Carlo applications
+In the literature, current implementations of parallel Neumann-Ulam
+methods have been limited to full domain replication with parallelism
+exploited through individual histories
+\cite{alexandrov_efficient_1998}. For high performance computing
+environments, domain decomposition strategies have been identified as
+important for Monte Carlo applications
 \cite{brunner_comparison_2006,siegel_analysis_2012}. In a domain

### 2013-01-07 doc/mandc2013/theory_paper/slattery_wilson_evans.tex
Update based on Jan 7 comments from Wilson.  Still need to change capitalization of Eigenvalue/function.

@@
 domain replication with parallelism exploited through individual
-histories \cite{alexandrov_efficient_1998}. In order to enable higher
-fidelity reactor physics Monte Carlo simulations, domain decomposition
-has been identified as a key principle in moving forward in high
-performance computing
+histories \cite{alexandrov_efficient_1998}. In reactor physics Monte
+Carlo applications, however, domain decomposition has been identified
+as a key principle in moving forward in high performance computing to
+enable higher fidelity simulations
 \cite{brunner_comparison_2006,siegel_analysis_2012}. To accomplish
@@
 in general expensive, but also because these nearest-neighbor
-communication sequences have poor algorithmic strong scaling
-\cite{gropp_high-performance_2001}.
+communication sequences, specifically, have poor algorithmic strong
+scaling \cite{gropp_high-performance_2001}.
 
@@
 \begin{equation}
-  \ve{d} = \boldsymbol{\delta}_i\:,
+  \ve{d} = \boldsymbol{\delta}_j\:,
   \label{eq:adjoint_second_constraint}
 \end{equation}
-where the $k^{th}$ component of the vector $\boldsymbol{\delta}_i$ is
-the Dirac delta function $\delta_{i_k,i}$. If we apply
-Eq~(\ref{eq:adjoint_second_constraint}) to our first constraint
-Eq~(\ref{eq:adjoint_vector_relation}), we get the following convenient
-outcome:
+where $\boldsymbol{\delta}_j$ is one of a set of vectors in which the
+$i^{th}$ component is the Kronecker delta function $\delta_{i,j}$. If
+we apply Eq~(\ref{eq:adjoint_second_constraint}) to our first
+constraint Eq~(\ref{eq:adjoint_vector_relation}), we get the following
+convenient outcome:
 \begin{equation}
   \langle \ve{y}, \ve{b} \rangle = \langle \ve{x},
-  \boldsymbol{\delta}_i \rangle = x_i \:,
+  \boldsymbol{\delta}_j \rangle = x_j \:,
   \label{eq:inner_product_constraint}
 \end{equation}
-meaning that if we compute the inner product of the original source and
-the adjoint solution using a delta function source, we recover the
-original solution.
+meaning that if we compute the inner product of the original source
+and the adjoint solution using a delta function source, we recover one
+component of the original solution.
 
@@
   p_{ij} = \frac{|h_{ji}|}{\sum_j |h_{ji}|}\:,\ w_{ij} =
-  \frac{h_{ji}}{p_{ij}},
+  \frac{|h_{ji}|}{p_{ij}},
   \label{eq:adjoint_probability}
 \end{equation}
-such that we expect to select a new state, $j$, from the current
-state, $i$, in the random walk by sampling the $i^{th}$ row of the
-probablity matrix as a discrete distribution.  Using our result from
+such that for each step, $m$, in the random walk, we expect to select
+a new state, $j$, from the current state, $i_m$, by sampling the
+$i_m^{th}$ row of the probablity matrix as a discrete distribution,
+and updating the weight, $W_m$, accordingly.  Using our result from
 Eq~(\ref{eq:inner_product_constraint}) generated by applying the
 adjoint constraints, the Monte Carlo estimator for the solution in
-state $i$ for a particular random walk permutation is then:
+state $i$ for a particular random walk permutation, $\nu$, is then:
 \begin{equation}
@@
   \begin{split}
-    E\{X_j\} &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-    \sum_{i_k}^{N} b_{i_0} h_{i_0,i_1}h_{i_1,i_2}\ldots
-    h_{i_{k-1},i_k} \delta_{i_k,j} \\ &= x_{j}\:,
+    E\{X_j\} &=\sum_{m=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
+    \sum_{i_m}^{N} b_{i_0} h_{i_0,i_1}h_{i_1,i_2}\ldots
+    h_{i_{m-1},i_m} \delta_{i_m,j} \\ &= x_{j}\:,
   \end{split}
@@
 \end{equation}
-where $W_c$ is the \textit{weight cutoff}. The adjoint random
-walk will then be terminated after $m$ steps if $W_m < W_f$ as tally
-contributions become increasingly small.
+where $W_c$ is the \textit{weight cutoff}. The adjoint random walk
+will then be terminated after $m$ steps if $W_m < W_f$ as tally
+contributions become increasingly small, effectively establishing a
+measure for the degree of approximation of the solution.
 
@@
 case of the adjoint Neumann-Ulam method, no such tolerance exists,
-however, we have specified a weight cutoff $W_c$ under which
-low-weight histories will be prematurely terminated as their
+however, we have specified a weight cutoff, $W_c$, that determines
+when low-weight histories will be prematurely terminated as their
 contributions are deemed minute. After $k$ iterations, a stationary
@@
 average every transition. For our diffusion model problem, $n_s$ would
-equate to the number of states in the $i$ (or $j$ as the problem is
-symmetric) direction that a history could potentially move in a single
+equate to the expected number of states in the $i$ (or $j$ as the problem is
+symmetric) direction that a history will move in a single
 transition and is dependent on the stencil used for the
@@
 \end{equation}
-where $d$ is the dimension of the problem and $\sqrt{\langle
+where $d$ is the dimensionality of the problem and $\sqrt{\langle
   \bar{r_k^2} \rangle}$ is effectively the mean free path of the Monte
@@
 
-Using this optical thickness, we can then complete the leakage
-approximations by defining the bounds of $\tau \rightarrow 0, \Lambda
-\rightarrow 1$ and $\tau \rightarrow \infty, \Lambda \rightarrow
-\tau^{-1}$. We expect this behavior as optically thin domains will
-communicate most of their histories while optically thick domains will
-leak the fraction of histories that did not interact within. With
-these bounds we can then define the leakage fraction out of a domain
-for the adjoint Neumann-Ulam method using the Wigner rational
-approximation:
+For optically thin domains, we expect that most histories will be
+communicated, while optically thick domains will leak the fraction of
+histories that did not interact within. Using the optical thickness
+defined in Eq~(\ref{eq:optical_thickness_3}), we can then complete the
+leakage approximations by defining the bounds of $\tau \rightarrow 0,
+\Lambda \rightarrow 1$ and $\tau \rightarrow \infty, \Lambda
+\rightarrow \tau^{-1}$.  With these bounds we can then define the
+leakage fraction out of a domain for the adjoint Neumann-Ulam method
+using the Wigner rational approximation:
 \begin{equation}

### 2013-01-08 doc/mandc2013/theory_paper/slattery_wilson_evans.tex
chaning Eigen to eigen in mandc paper

@@
 length of the adjoint random walks, a measure of convergence speed and
-serial performance, are made with respect to the Eigenvalues of the
+serial performance, are made with respect to the eigenvalues of the
 linear operator. In addition, relationships for the effective optical
@@
 is symmetric, a host of analytic theories exist based on the
-Eigenvalue spectrum of the operator that characterize their behavior
+eigenvalue spectrum of the operator that characterize their behavior
 in the context of deterministic linear solvers. Using past work, these
@@
 the linear system generated by the discretization of this, we use a
-spectral analysis to generate analytic relations for the Eigenvalues
-of the operator based on system parameters. Using the Eigenvalue
+spectral analysis to generate analytic relations for the eigenvalues
+of the operator based on system parameters. Using the eigenvalue
 spectra, we then build relationships to characterize the transport of
@@
 $\ve{H}$ must remain less than 1 as $\ve{H}^T$ contains the same
-Eigenvalues and therefore has the same spectral radius. We expand this
+eigenvalues and therefore has the same spectral radius. We expand this
 summation to yield a series of transitions that can be approximated by
@@
 where $\boldsymbol{\delta}_j$ is one of a set of vectors in which the
-$i^{th}$ component is the Kronecker delta function $\delta_{i,j}$. If
+$j^{th}$ component is the Kronecker delta function $\delta_{i,j}$. If
 we apply Eq~(\ref{eq:adjoint_second_constraint}) to our first
@@
 Eq~(\ref{eq:adjoint_neumann_series}) approximated by the Monte Carlo
-solver is dependent on the Eigenvalues of the iteration matrix. We
-will compute these Eigenvalues by assuming Eigenfunctions of the form
+solver is dependent on the eigenvalues of the iteration matrix. We
+will compute these eigenvalues by assuming eigenfunctions of the form
 \cite{leveque_finite_2007}:
@@
 where different combinations of $p$ and $q$ represent the different
-Eigenmodes of the solution. As these are valid forms of the solution,
-then the action of the linear operator on these Eigenfunctions should
-give the Eigenvalues of the matrix as they exist on the unit circle in
+eigenmodes of the solution. As these are valid forms of the solution,
+then the action of the linear operator on these eigenfunctions should
+give the eigenvalues of the matrix as they exist on the unit circle in
 the complex plane.
@@
 
-For the model problem, we first compute the Eigenvalues for the
+For the model problem, we first compute the eigenvalues for the
 diffusion operator $\ve{D}$ by applying the operator to the
-Eigenfunctions and noting that $x=ih$ and $y=jh$:
+eigenfunctions and noting that $x=ih$ and $y=jh$:
 \begin{multline}
@@
 Using this coefficient, we then have the following spectrum of
-preconditioned Eigenvalues:
+preconditioned eigenvalues:
 \begin{equation}
   \lambda_{p,q}(\ve{M}^{-1} \ve{D}) = \alpha \lambda_{p,q}(\ve{D})\:.
-  \label{eq:preconditioned_Eigenvalues}
+  \label{eq:preconditioned_eigenvalues}
 \end{equation}
@@
 The spectral radius of the iteration matrix is obtained by seeking its
-largest Eigenvalue. As with the diffusion operator, we can use the
-same analysis techniques to find the Eigenvalues for the iteration
+largest eigenvalue. As with the diffusion operator, we can use the
+same analysis techniques to find the eigenvalues for the iteration
 matrix. We use a few simplifications by noting that if the Jacobi
@@
 \end{equation}
-Inserting the Eigenfunctions defined by
+Inserting the eigenfunctions defined by
 Eq~(\ref{eq:eigenfunction_form}) we get:
@@
 \end{equation}
-giving the Eigenvalue spectrum for the Jacobi preconditioned iteration
-matrix. We find that the maximum Eigenvalue exists when $p=q=0$,
+giving the eigenvalue spectrum for the Jacobi preconditioned iteration
+matrix. We find that the maximum eigenvalue exists when $p=q=0$,
 giving the following for the spectral radius of the Jacobi
@@
 that will be generated from a particular linear operator based on the
-Eigenvalues of its iteration matrix (independent of the linear
+eigenvalues of its iteration matrix (independent of the linear
 operator splitting chosen) and based on the weight cutoff parameter
@@
 \end{equation}
-Here, the leakage fraction is explicitly bound to the Eigenvalues of
+Here, the leakage fraction is explicitly bound to the eigenvalues of
 the iteration matrix, the size of the domain, the content of the
@@
 one to measure the length of the random walks as a function of the
-iteration matrix Eigenvalues, and one to measure the domain leakage
-fraction as a function of the iteration matrix Eigenvalues and the
+iteration matrix eigenvalues, and one to measure the domain leakage
+fraction as a function of the iteration matrix eigenvalues and the
 discretization properties. Before doing this, we verify our
 computation of the spectral radius of the iteration matrix by
-numerically computing the largest Eigenvalue of the diffusion operator
-using an iterative Eigenvalue solver. For this verification, a $100
+numerically computing the largest eigenvalue of the diffusion operator
+using an iterative eigenvalue solver. For this verification, a $100
 \times 100$ square grid with $h=0.01$, $h=0.1$, and $h=1.0$ and the
@@
 observed between the analytic and numerical results with all data
-points computed within the tolerance of the iterative Eigenvalue
+points computed within the tolerance of the iterative eigenvalue
 solver.
@@
       $h=1.0$ were used. The red data was computed numerically by an
-      Eigensolver while the black dashed data was generated by
+      eigensolver while the black dashed data was generated by
       Eq~(\ref{eq:iteration_radius}).}

### 2013-01-15 doc/mandc2013/theory_paper/slattery_wilson_evans.tex
fixed typo in mandc paper

@@
 direction with measurements occurring in the central domain without
-boundary grid points. For the cross sections, the absorption cross was
-varied from 1 to 100 while the scattering cross section was set to
-zero to create a purely absorbing environment with weight cutoff of
-\sn{1}{-4}. The optical thickness of these domains will vary as a
-function of the absorption cross section if the other parameters are
-fixed. To compute the optical thickness, along with the spectral
+boundary grid points. For the cross sections, the absorption cross
+section was varied from 1 to 100 while the scattering cross section
+was set to zero to create a purely absorbing environment with weight
+cutoff of \sn{1}{-4}. The optical thickness of these domains will vary
+as a function of the absorption cross section if the other parameters
+are fixed. To compute the optical thickness, along with the spectral
 radius as given by Eq~(\ref{eq:iteration_radius}), we also need the

### 2013-01-30 doc/mandc2013/theory_paper/slattery_wilson_evans.tex
Fix typo on variable notation.

--- 2013-01-07 doc/mandc2013/theory_paper/slattery_wilson_evans.tex
@@
 the typical length, their sum is 12 out of the total 20 for the
-coefficients for all possible states with their ratio giving $n_i =
+coefficients for all possible states with their ratio giving $n_s =
 \frac{3}{5}$.

### 2013-01-30 doc/sna2013/abstract/slattery_wilson_evans.tex
sna updates for submission

@@
 
-At the 2010 SNA+MC meeting, Wagner and colleagues developed the
-\textit{multiple-set overlapping-domain} (MSOD) decomposition for
-parallel Monte Carlo applications for full-core light water reactor
-analysis \cite{wagner_hybrid_2010}. In their work, overlap between
-adjacent domains in the decomposition was used to decrease the number
-of particles leaving the local domain. In addition, Wagner utilized a
-level of replication of the domain such that the domain was only
-decomposed on $O(100)$ processors and if replicated $O(1,000)$ times
-achieves simulation on $O(100,000)$ processors, thus providing spatial
-and particle parallelism. Each collection of processors that
+At the 2010 SNA+MC meeting, Wagner and colleagues presented the
+\textit{multiple-set overlapping-domain} (MSOD) decomposition method
+for parallel Monte Carlo applications for full-core light water
+reactor analysis \cite{wagner_hybrid_2010}. In their work, overlap
+between adjacent domains in the decomposition was used to decrease the
+number of particles leaving the local domain. In addition, Wagner
+utilized a level of replication of the domain such that the domain was
+only decomposed on $O(100)$ processors and if replicated $O(1,000)$
+times achieves simulation on $O(100,000)$ processors, thus providing
+spatial and particle parallelism. Each collection of processors that
 constitutes a representation of the entire domain is referred to as a

### 2013-01-30 drivers/NeutronDiffusion/DiffusionProblem.cpp
fixed jacobi scaling issue. MCLS works!

@@ -56,6 +56,22 @@ DiffusionProblem::DiffusionProblem( const RCP_Comm& comm,
     Teuchos::Array<double> iminus1jplus1( 1, -D/(6.0*dy*dy) );
     Teuchos::Array<double> iplus1jplus1( 1, -D/(6.0*dy*dy) );
 
+
+    // Apply preconditioning.
+    double jacobi_scale = diag[0];
+    if ( jacobi_precondition )
+    {
+	iminus1[0] /= jacobi_scale;
+	iplus1[0] /= jacobi_scale;
+	jminus1[0] /= jacobi_scale;
+	jplus1[0] /= jacobi_scale;
+	iminus1jminus1[0] /= jacobi_scale;
+	iplus1jminus1[0] /= jacobi_scale;
+	iminus1jplus1[0] /= jacobi_scale;
+	iplus1jplus1[0] /= jacobi_scale;
+	diag[0] /= jacobi_scale;
+    }
+
     Teuchos::Array<int> idx(1);
     Teuchos::Array<int> idx_iminus1(1);
     Teuchos::Array<int> idx_iplus1(1);
@@ -323,12 +339,6 @@ DiffusionProblem::DiffusionProblem( const RCP_Comm& comm,
 		}
 	    }
 	}
-
-	// Apply preconditioning.
-	if ( jacobi_precondition )
-	{
-	    A->scale( 1/diag[0] );
-	}
     }
     comm->barrier();
 
@@ -361,7 +371,7 @@ DiffusionProblem::DiffusionProblem( const RCP_Comm& comm,
     // Jacobi precondition if necessary.
     if ( jacobi_precondition )
     {
-	B->scale( 1/diag[0] );
+	B->scale( 1.0/jacobi_scale );
     }
 
     // Build the linear problem.

### 2013-01-30 drivers/NeutronDiffusion/neutron_diffusion.cpp
adding xml parameter for MCLS solver type

@@ -115,9 +115,11 @@ int main( int argc, char * argv[] )
 			  diffusion_problem->getProblem()->getLHS(),
 			  diffusion_problem->getProblem()->getRHS() ) );
 
+    std::string solver_type = plist->get<std::string>("Solver Type");
+
     MCLS::SolverFactory<Vector,Matrix> factory;
     Teuchos::RCP<MCLS::SolverManager<Vector,Matrix> > solver_manager =
- 	factory.create( "Adjoint MC", comm, plist );
+ 	factory.create( solver_type, comm, plist );
     solver_manager->setProblem( linear_problem );
     solver_manager->solve();
 

### 2013-02-06 doc/presentations/meeting_2_6_13/meeting_2_6_13.tex
adding pngs for results

--- 2012-11-05 doc/prelim/presentation/prelim_presentation.tex
@@
 \date{\today} 
-\title{Massively Parallel Monte Carlo Methods for Discrete Linear and
-  Nonlinear Systems} 
+\title{Domain Decomposed Monte Carlo} 
 \begin{document}
 \maketitle
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Introduction}
-
-  \begin{itemize}
-  \item Predictive modeling and simulation enhances engineering
-    capability
-    \medskip
-  \item Modern work focused on this task leverages multiple physics
-    simulation (CASL, NEAMS)
-    \medskip
-  \item New hardware drives algorithm development (petascale and
-    exascale)
-    \medskip
-  \item Monte Carlo methods have the potential to provide great
-    improvements that permit finer simulations and better mapping to
-    future hardware
-    \medskip
-  \item A set of massively parallel Monte Carlo methods is proposed to
-    advance multiple physics simulation on contemporary and future
-    leadership class machines
-  \end{itemize}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Physics-Based Motivation}
-
-  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
-    {Predictive nuclear reactor analysis enables...            (DOE,2011)}
-    \begin{itemize}
-    \item Tighter design tolerance for improved thermal performance
-      and efficiency
-    \item Higher fuel burn-up
-    \item High confidence in accident scenario models
-    \end{itemize}
-  \end{beamerboxesrounded}
-
-  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
-    {Multiple physics simulations are complicated...}
-    \begin{itemize}
-    \item Neutronics, thermal hydraulics, computational fluid
-      dynamics, structural mechanics, and many other physics
-    \item Consistent models yield nonlinearities in the variables
-      through feedback effects
-    \item Tremendous computational resources are required with
-      $O(\sn{1}{9})$ element meshes and $O(100,000)+$ cores used in
-      today's simulations (Evans,2010)(Pawlowski,2012)
-    \end{itemize}
-  \end{beamerboxesrounded}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Hardware-Based Motivation}
-
-  \begin{itemize}
-  \item Modern hardware is moving in two directions (Kogge,2011):
-    \begin{itemize}
-    \item Lightweight machines
-    \item Heterogeneous machines
-    \item Both characterized by low power and high concurrency
-    \end{itemize}
-    \medskip \medskip
-  \item Some issues:
-    \begin{itemize}
-    \item Higher potential for both soft and hard failures (DOE,2012)
-    \item Memory restrictions are expected with a continued decrease
-      in memory/FLOPS
-    \end{itemize}
-    \medskip \medskip
-  \item Potential resolution from Monte Carlo:
-    \begin{itemize}
-    \item Soft failures buried within the tally variance
-    \item Hard failures are high variance events
-    \item Memory savings over conventional methods
-    \end{itemize}
-  \end{itemize}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Research Outline}
-  \begin{itemize}
-  \item Parallelization of Monte Carlo methods for discrete systems
-    \begin{itemize}
-    \item Parallel strategies taken from modern reactor physics
-      methods
-    \item Research is required to explore varying parallel strategies
-    \item Scalability is of concern
-    \end{itemize}
-    \medskip \medskip \medskip
-  \item Development of a nonlinear solver for discrete systems
-    leveraging Monte Carlo
-    \begin{itemize}
-    \item Application to nonlinear problems of interest
-    \item Memory benefits
-    \item Performance benefits
-    \end{itemize}
-  \end{itemize}
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Projection Methods}
-
-  \begin{itemize}
-  \item Powerful class of iterative methods (Saad,2003)
-  \item Provides theory that encapsulates most other iterative methods
-  \item Leveraged in many modern physics codes at the petascale
-  \end{itemize}
-
-  \medskip \medskip
-  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
-    {Search Subspace $\mathcal{K}$} Extract the solution from the
-    search subspace:
-    \[
-    \tilde{\ve{x}} = \ve{x}_0 +
-    \boldsymbol{\delta},\ \boldsymbol{\delta} \in \mathcal{K}
-    \]
-  \end{beamerboxesrounded}
-
-  \medskip \medskip
-  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
-    {Constraint Subspace $\mathcal{L}$} Constrain the extraction with
-    the constraint subspace by asserting orthogonality with the
-    residual:
-    \[
-    \langle \tilde{\ve{r}},\ve{w} \rangle = 0,\ \forall \ve{w} \in
-    \mathcal{L}
-    \]  
-  \end{beamerboxesrounded}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{The Orthogonality Constraint}
-
-  \[
-  \tilde{\ve{r}} = \ve{r}_0 - \ve{A}\boldsymbol{\delta}
-  \]
-
-  \begin{figure}[htpb!]
-    \begin{center}
-      \scalebox{1.1}{ \input{orthogonal_residual.pdftex_t} }
-    \end{center}
-    \caption{\textbf{Orthogonality constraint of the new residual with
-        respect to $\mathcal{L}$.}}
-  \end{figure}
-
-  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
-    {Minimization Property}
-
-    The residual of the system is \textit{minimized} with respect to
-    the constraints
-    \[
-    ||\tilde{\ve{r}}||_2 \leq ||\ve{r}_0||_2,\ \forall \ve{r}_0 \in
-    \mathbb{R}^N
-    \]
-  \end{beamerboxesrounded}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Krylov Subspace Methods}
-
-  \[
-  \mathcal{K}_m(\ve{A},\ve{r}_0) = span\{\ve{r}_0, \ve{A}\ve{r}_0,
-  \ve{A}^2\ve{r}_0, \dots, \ve{A}^{m-1}\ve{r}_0\}
-  \]
-
-  \begin{itemize}
-  \item For GMRES (Saad, 1986): $\mathcal{L} = \ve{A}
-    \mathcal{K}_m(\ve{A},\ve{r}_0)$
-  \end{itemize}
-
-  \medskip \medskip \medskip
-  \small{
-  \begin{itemize}
-  \item Yields the normal system $\ve{A}^T\ve{A}\ve{x} =
-    \ve{A}^T\ve{b}$
-    \medskip
-  \item Require only the action of the operator
-    \medskip
-  \item Must generate an orthonormal basis for
-    $\mathcal{K}_m(\ve{A},\ve{r}_0)$
-    \medskip
-  \item Typically choose a Gram-Schmidt-like procedure such as Arnoldi
-    or Lanzcos
-    \medskip
-  \item Short and long recurrence relations available for
-    orthogonalization
-  \end{itemize}
-  }
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Parallel Projection Methods}
-
-  \begin{itemize}
-  \item Parallel vector update
-  \end{itemize}
-  \[
-  \ve{y}[n] \leftarrow \ve{y}[n] + a * \ve{x}[n],\ \forall n \in
-     [1,N_g]
-  \]
-  \[
-  \ve{y}[n] \leftarrow \ve{y}[n] + a * \ve{x}[n],\ \forall n \in
-     [1,N_l]
-  \]
-
-  \medskip \medskip
-  \begin{itemize}
-  \item Parallel dot product
-  \end{itemize}
-  \[
-  d_l = \ve{y}_l \cdot \ve{x}_l,\ d_g = \sum_p d_l
-  \]
-
-  \medskip \medskip
-  \begin{itemize}
-  \item Parallel vector norm
-  \end{itemize}
-  \[
-  ||x||_{\infty,l} = \max_n \ve{y}[n],\ \forall n \in [1,N_l]
-  \]
-  \[
-  ||x||_{\infty,g} = \max_p ||x||_{\infty,l}
-  \]
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Parallel Matrix-Vector Multiplication}
-
-  \begin{figure}[htpb!]
-    \begin{center}
-      \scalebox{0.6}{ \input{partitioned_matrix.pdftex_t} }
-    \end{center}
-    \caption{\textbf{Matrix-vector multiply $\ve{A}\ve{x}=\ve{y}$
-        operation on 3 processors.}
-      \label{fig:partitioned_matvec_multiply} }
-  \end{figure}
-
-  \begin{figure}[htpb!]
-    \begin{center}
-      \scalebox{0.7}{ \input{matvec_proc_1.pdftex_t} }
-    \end{center}
-    \caption{\textbf{Components of multiply operation owned by process
-        1.} }
-    \label{fig:matvec_proc_1}
-  \end{figure}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Projection Methods Summary}
-
-  \begin{itemize}
-  \item Widely used in practice
-    \medskip
-  \item Krylov methods require only the action of the operator
-    \medskip
-  \item Parallelism achieved through a handful of operations
-    \medskip
-  \item Global reduction operations observed not to impede scalability
-    (Gropp,2001)
-    \begin{itemize}
-    \item Dot product
-    \item Vector norms
-    \end{itemize}
-    \medskip
-  \item Nearest neighbor computations have poor algorithmic strong
-    scaling
-    \begin{itemize}
-    \item Matrix-vector multiply
-    \item Weak scaling is better
-    \end{itemize}
-    \medskip
-  \item Short and long recurrence relations available for
-    orthogonalization
-  \end{itemize}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Monte Carlo Methods for Discrete Linear Systems}
-
-  \begin{itemize}
-  \item First proposed by J. Von Neumann and S.M. Ulam in the 1940's
-    \medskip \medskip
-  \item Earliest published reference in 1950
-    \medskip \medskip
-  \item General lack of published work
-    \medskip \medskip
-  \item Modern work by Evans and others has yielded new applications
-  \end{itemize}
-\end{frame}
 
@@
   \begin{itemize}
-  \item Split the operator
+  \item Split the linear operator
   \end{itemize}
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Direct Method}
-
-  \begin{itemize}
-  \item Compute row-normalized transition probabilities and weights
-  \end{itemize}
-
-  \[
-  p_{ij} = \frac{|h_{ij}|}{\sum_j |h_{ij}|},\ w_{ij} =
-  \frac{h_{ij}}{p_{ij}}
-  \]
-
-  \medskip \medskip
-  \begin{itemize}
-  \item Generate an expectation value for the solution
-  \end{itemize}
-
-  \[
-  W_{m} = w_{i,i_1} w_{i_1,i_2} \cdots w_{i_{m-1},i_m}
-  \]
-  \[
-  X_{\nu}(i_0 = i) = \sum_{m=0}^k W_{m} b_{i_m}
-  \]
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Direct Method}
-
-  \begin{itemize}
-  \item Compute the probability of a particular random walk
-    permutation
-  \end{itemize}
-
-  \[
-  P_{\nu} = p_{i,i_1} p_{i_1,i_2} \cdots p_{i_{k-1},i_k}
-  \]
-
-  \begin{itemize}
-  \item Generate the estimator
-  \end{itemize}
-
-  \[
-  E\{X(i_0 = i)\} = \sum_{\nu} P_{\nu} X_{\nu}
-  \]
-
-  \begin{itemize}
-  \item Check that we recover the exact solution
-  \end{itemize}
-
-  \[
-  \begin{split}
-    E\{X(i_0 = i)\}
-    &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-    \sum_{i_k}^{N} p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
-    w_{i,i_1}w_{i_1,i_2}\ldots w_{i_{k-1},i_k} b_{i_k}\\ &= x_i
-  \end{split}
-  \]
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Direct Method: Evolution of a Solution}
-
-  \begin{figure}[h!]
-    \begin{center}
-      \scalebox{1.0}{ \input{heat_eq_setup.pdftex_t} }
-    \end{center}
-  \caption{\textbf{Problem setup for 2D heat equation.}
-    \textit{Dirichlet conditions are set for the temperature on all 4
-      boundaries of the Cartesian grid. No thermal source was
-      present. $50 \times 50$ grid.}}
-  \end{figure}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Direct Method: Evolution of a Solution}
-
-  \begin{figure}[h!]
-    \begin{center}
-      \includegraphics[width=4in]{direct_1.png}
-    \end{center}
-    \caption{\textbf{Direct solution to Poisson Equation.} \textit{1
-        history per state, \sn{2.5}{3} total histories. 0.785 seconds
-        CPU time.} }
-  \end{figure}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Direct Method: Evolution of a Solution}
-
-  \begin{figure}[h!]
-    \begin{center}
-      \includegraphics[width=4in]{direct_10.png}
-    \end{center}
-    \caption{\textbf{Direct solution to Poisson Equation.} \textit{10
-        histories per state, \sn{2.5}{4} total histories. 5.9 seconds
-        CPU time.} }
-  \end{figure}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Direct Method: Evolution of a Solution}
-
-  \begin{figure}[h!]
-    \begin{center}
-      \includegraphics[width=4in]{direct_100.png}
-    \end{center}
-    \caption{\textbf{Direct solution to Poisson Equation.}
-      \textit{100 histories per state, \sn{2.5}{5} total
-        histories. 54.7 seconds CPU time.} }
-  \end{figure}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Direct Method: Evolution of a Solution}
-
-  \begin{figure}[h!]
-    \begin{center}
-      \includegraphics[width=4in]{direct_1000.png}
-    \end{center}
-    \caption{\textbf{Direct solution to Poisson Equation.} \textit{1000
-        histories per state, \sn{2.5}{6} total histories. 644 seconds
-        CPU time.} }
-  \end{figure}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Adjoint Method}
-
-  \begin{itemize}
-  \item Solve the adjoint linear system
-  \end{itemize}
-
-  \[
-  \ve{A}^T \ve{y} = \ve{d}
-  \]
-
-  \[
-  \ve{y} = \ve{H}^T \ve{y} + \ve{d}
-  \]
-
-  \medskip
-  \begin{itemize}
-  \item Set the adjoint constraint
-  \end{itemize}
-
-  \[
-  \langle \ve{A}^T \ve{x}, \ve{y} \rangle = \langle \ve{x}, \ve{A}
-  \ve{y} \rangle
-  \]
-
-  \[
-  \langle \ve{x}, \ve{d} \rangle = \langle \ve{y}, \ve{b} \rangle
-  \]
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Adjoint Method}
-
-  \begin{itemize}
-  \item Generate the Neumann series for the adjoint operator
-  \end{itemize}
-
-  \[
-  \ve{y} = (\ve{I} - \ve{H}^T)^{-1} \ve{d} = \sum_{k=0}^{\infty}
-  (\ve{H}^T)^k\ve{d}
-  \]
-
-  \medskip
-  \begin{itemize}
-  \item Expand the series
-  \end{itemize}
-
-  \[
-  y_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-  \sum_{i_k}^{N}h_{i_k,i_{k-1}}\ldots h_{i_2,i_1} h_{i_1,i} d_{i_k}
-  \]
-
-  \medskip
-  \begin{itemize}
-  \item Pick another constraint to yield the original solution
-  \end{itemize}
-
-  \[
-  \ve{d} = \boldsymbol{\delta}_i,\ \langle \ve{y}, \ve{b} \rangle =
-  \langle \ve{x}, \boldsymbol{\delta}_i \rangle = x_i
-  \]
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Adjoint Method}
-
-  \begin{itemize}
-  \item Use the adjoint Neumann-Ulam decomposition
-  \end{itemize}
-
-  \[
-  \ve{H}^{T} = \ve{P} \circ \ve{W}
-  \]
-
-  \[
-  p_{ij} = \frac{|h_{ji}|}{\sum_j |h_{ji}|},\ w_{ij} =
-  \frac{h_{ji}}{p_{ij}}
-  \]
-
-  \medskip
-  \begin{itemize}
-  \item Build the estimator and expectation value
-  \end{itemize}
-
-  \[
-  X_{\nu} = \sum_{m=0}^k W_{m} \delta_{i,i_m}
-  \]
-
-  \[
-  \begin{split}
-    E\{X_j\} &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-    \sum_{i_k}^{N} b_{i_0} h_{i_0,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}
-    \delta_{i_k,j} \\ &= x_{j}
-  \end{split}
-  \]
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
 \begin{frame}{Adjoint Method: Evolution of a Solution}
@@
     \begin{center}
-      \includegraphics[width=4in]{adjoint_1.png}
+      \includegraphics[width=4in]{../../prelim/presentation/adjoint_1.png}
     \end{center}
@@
     \begin{center}
-      \includegraphics[width=4in]{adjoint_10.png}
+      \includegraphics[width=4in]{../../prelim/presentation/adjoint_10.png}
     \end{center}
@@
     \begin{center}
-      \includegraphics[width=4in]{adjoint_100.png}
+      \includegraphics[width=4in]{../../prelim/presentation/adjoint_100.png}
     \end{center}
@@
     \begin{center}
-      \includegraphics[width=4in]{adjoint_1000.png}
+      \includegraphics[width=4in]{../../prelim/presentation/adjoint_1000.png}
     \end{center}
@@
     \begin{center}
-      \includegraphics[width=4in]{adjoint_10000.png}
+      \includegraphics[width=4in]{../../prelim/presentation/adjoint_10000.png}
     \end{center}
@@
     \begin{center}
-      \includegraphics[width=4in]{adjoint_100000.png}
+      \includegraphics[width=4in]{../../prelim/presentation/adjoint_100000.png}
     \end{center}
@@
     \begin{center}
-      \includegraphics[width=4in]{adjoint_1000000.png}
+      \includegraphics[width=4in]{../../prelim/presentation/adjoint_1000000.png}
     \end{center}
@@
     \begin{center}
-      \includegraphics[width=4in]{adjoint_10000000.png}
+      \includegraphics[width=4in]{../../prelim/presentation/adjoint_10000000.png}
     \end{center}
@@
   \end{figure}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Monte Carlo Synthetic-Acceleration}
-
-  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
-    {MCSA Iteration (Evans,2009)}
-
-    \[
-    \ve{x}^{k+1/2} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}
-    \]
-    \[
-    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}
-    \]
-    \[
-    \hat{\ve{A}}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}
-    \]
-    \[
-    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}
-    \]
-
-  \end{beamerboxesrounded}
-
-  \medskip \medskip
-  \begin{itemize}
-  \item Neumann-Ulam methods bound by the Central Limit Theorem
-  \item Build on Halton's 1962 Sequential Monte Carlo method
-  \item Adjoint Neumann-Ulam solver computes the correction
-  \item Decouples MC error from solution error, exponential convergence
-  \item Demonstrated by Evans to be competitive with Krylov methods
-  \end{itemize}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Preconditioning Monte Carlo Methods}
-
-  \begin{itemize}
-  \item No symmetry requirements
-  \item Require $\rho(\ve{H}) < 1$
-  \item Choose Jacobi preconditioning at a minimum
-  \end{itemize}
-
-  \[
-  \ve{M} = diag(\ve{A})
-  \]
-  \[
-  \ve{M}^{-1}\ve{A}\ve{x} = \ve{M}^{-1}\ve{b}
-  \]
-
-  \medskip
-  \begin{itemize}
-  \item Yields a preconditioned MCSA iteration with no in-state
-    transitions
-  \end{itemize}
-
-  \[
-  \ve{x}^{k+1/2} = (\ve{I} - \ve{M}^{-1}\ve{A})\ve{x}^k + \ve{b}
-  \]
-  \[
-  \ve{r}^{k+1/2} = \ve{b} - \ve{M}^{-1}\ve{A}\ve{x}^{k+1/2}
-  \]
-  \[
-  \ve{M}^{-1}\ve{A}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}
-  \]
-  \[
-  \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}
-  \]
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Direct vs. Adjoint Analysis}
-
-  \begin{itemize}
-  \item Analysis needed to select Monte Carlo method
-  \item Time-dependent 2-dimensional Poisson equation
-  \item Spectral radius fixed
-  \item Sparsity varied with 2 Laplacian stencils
-  \end{itemize}
-
-  \[
-  \nabla^2_5 = \frac{1}{\Delta^2}[u_{i-1,j} + u_{i+1,j} + u_{i,j-1} +
-    u_{i,j+1} - 4 u_{i,j}]
-  \]
-  \[
-  \begin{split}
-    \nabla^2_9 = \frac{1}{6\Delta^2}[4 u_{i-1,j} + 4 u_{i+1,j} + 4
-      u_{i,j-1} + 4 u_{i,j+1} + u_{i-1,j-1}\\ + u_{i-1,j+1} +
-      u_{i+1,j-1} + u_{i+1,j+1} - 20 u_{i,j}]
-  \end{split}
-  \]
-
-  \begin{itemize}
-  \item Implicit Euler time differencing
-  \end{itemize}
-
-  \[
-  \ve{A} \ve{u}^{n+1} = \ve{u}^n
-  \]
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Direct vs. Adjoint Analysis}
-
-  \begin{columns}
-
-    \begin{column}{0.5\textwidth}
-      \begin{figure}[h!]
-        \centering
-        \includegraphics[width=2.5in,clip]{Adjoint_Direct_CPU_Time.pdf}
-        \caption{\textbf{CPU Time (s) to converge vs. Problem Size
-            ($N$ for an $N \times N$ square mesh).} }
-      \end{figure}
-    \end{column}
-
-    \begin{column}{0.5\textwidth}
-      \begin{figure}[h!]
-        \centering
-        \includegraphics[width=2.5in,clip]{Adjoint_Direct_Iterations.pdf}
-        \caption{\textbf{Iterations to converge vs. Problem Size ($N$
-            for an $N \times N$ square mesh).} }
-      \end{figure}
-    \end{column}
-
-  \end{columns}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Direct vs. Adjoint Analysis}
-
-  \begin{figure}[h!]
-    \centering
-    \includegraphics[width=2.5in,clip]{Adjoint_Direct_Convergence.pdf}
-    \caption{\textbf{Infinity norm of the solution residual
-        vs. iteration number for a problem of fixed size.} }
-  \end{figure}
-
-  \begin{itemize}
-  \item CPU time dominating factor in method selection
-  \item Significant speedup with adjoint method
-  \item Does not affect convergence behavior
-  \item Use adjoint with MCSA and Sequential Monte Carlo
-  \end{itemize}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Generalization of MCSA for Linear Problems}
-
-  \begin{itemize}
-  \item Published work to date has used a physics-based formulation
-    \begin{itemize}
-    \item Radiation transport equations used as model system
-    \item Transition probabilities built from problem-specific
-      parameters
-    \item Probabilities and weights must be re-derived for each new
-      equation set
-    \end{itemize}
-    \medskip \medskip
-  \item Desire a generalization for all linear operator equations
-    \begin{itemize}
-    \item Requires a general parallel framework
-    \item Requires implementation with a general linear algebra
-      framework
-    \item Operator, vector, and graph abstractions
-    \end{itemize}
-    \medskip \medskip
-  \item Neumann-Ulam solvers and MCSA implemented using the Trilinos
-    Petra frameworks
-    \medskip \medskip
-  \item Can be leveraged in modern physics implementations
-  \end{itemize}
 
@@
     Neumann-Ulam method
-    \medskip \medskip
-  \item MCSA iteration-level parallelism comes from parallel
-    matrix/vector operations
   \end{itemize}
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Multiple-Set Overlapping-Domain Decomposition}
+\begin{frame}{Domain Decomposition}
+\end{frame}
 
-  \begin{columns}
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Fully Synchronous Algorithm}
+\end{frame}
 
-    \begin{column}{0.5\textwidth}
-      \begin{figure}[htpb!]
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Fully Asynchronous Algorithm}
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Brunner's Communication Parameters}
+
+  \begin{figure}[h!]
     \begin{center}
-          \scalebox{0.75}{ \input{msod_example.pdftex_t} }
+      \includegraphics[width=5in]{brunner_params.png}
     \end{center}
-        \caption{\textbf{Overlapping domain example illustrating how
-            domain overlap can reduce communication costs.}}
   \end{figure}
-    \end{column}
-
-    \begin{column}{0.5\textwidth}
-      \begin{itemize}
-      \item Developed by Wagner and colleagues in 2010
-      \item Each set contains the full domain
-      \item Multiple sets replicate the domain
-      \item Domains overlap within a set
-      \item Reduces communication by a significant fraction
-      \item Increases scalability through smaller processor sets
-      \item Redundancy for resiliency (and useful work)
-      \end{itemize}
-    \end{column}
-
-  \end{columns}
 
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Domain-to-Domain Communication}
+\begin{frame}{My Communication Parameters}
 
-  \begin{itemize}
-  \item The amount of domain leakage dictates communication
-    \medskip
-  \item Per Siegel's 2012 work, define a leakage fraction
-  \end{itemize}
-
-  \[
-  \lambda =
-  \frac{average\ \#\ of\ particles\ leaving\ local\ domain}{total\ 
-    of\ \#\ of\ particles\ starting\ in\ local\ domain}
-  \]
-
-  \medskip
-  \begin{itemize}
-  \item Siegel observed $\lambda$ to be bound empirically to the mean
-    free path of the system
-    \medskip
-  \item Optimum ratio of local to global domain size exists
-    \medskip
-  \item Experiments show feasibility for large-scale calculations
-    \begin{itemize}
-    \item Not bandwidth or latency limited in load-balanced case
-    \end{itemize}
-    \pause
-    \medskip
-  \item How do we define $\lambda$ empirically for linear operator
-    equations? 
-  \end{itemize}
+  \begin{figure}[h!]
+    \begin{center}
+      \includegraphics[width=5in]{my_params.png}
+    \end{center}
+  \end{figure}
 
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Load Balancing}
+\begin{frame}{Brunner's Strong Scaling}
 
-  \begin{figure}[htpb!]
+  \begin{figure}[h!]
     \begin{center}
-      \scalebox{1.0}{ \input{procassini_example.pdftex_t} }
+      \includegraphics[width=4in]{brunner_strong.png}
     \end{center}
-    \caption{\textbf{Example illustrating how domain decomposition can
-        create load balance issues in Monte Carlo.}}
-    \label{fig:procassini_example}
   \end{figure}
-
-  \begin{itemize}
-  \item Procassini's 2005 work addressed some concerns
-  \item Dynamic balancing replicates domains independently
-  \item Linear operator equations may benefit - fixed source problems
-  \end{itemize}
 
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Parallel Adjoint Method}
+\begin{frame}{My Strong Scaling}
 
-Strategy
-  \begin{itemize}
-  \item Direct analogs between particle transport and Neumann-Ulam
-    solvers 
-  \item Aim for MSOD implementation and fully asynchronous
-    communication patterns
-  \end{itemize}
-
-  \medskip \medskip \medskip
-Questions
-  \begin{itemize}
-  \item How much domain overlap is suitable for linear operator
-    equations? 
-  \item Is memory a limitation for overlap and replication?
-  \item Do the linear operator properties help select overlap?
-  \item Will full-clip roulette perturb the MCSA solution?
-  \item How much does MSOD facilitate scaling?
-  \end{itemize}
+  \begin{figure}[h!]
+    \begin{center}
+      \includegraphics[width=4in]{my_strong.png}
+    \end{center}
+  \end{figure}
 
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Parallel MCSA}
+\begin{frame}{Brunner's Weak Scaling}
 
-  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
-    {MCSA Iteration}
-
-    \[
-    \ve{x}^{k+1/2} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}
-    \]
-    \[
-    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}
-    \]
-    \[
-    \hat{\ve{A}}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}
-    \]
-    \[
-    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}
-    \]
-
-  \end{beamerboxesrounded}
-
-  \medskip \medskip
-  \begin{itemize}
-  \item Richardson iteration and residual computation require parallel
-    matrix-vector multiply and parallel vector update
-  \item This work will generate a parallel adjoint Neumann-Ulam solver
-  \item Application of correction requires parallel vector update
-  \item Convergence checks through parallel norm computation
-  \end{itemize}
+  \begin{figure}[h!]
+    \begin{center}
+      \includegraphics[width=4in]{brunner_weak.png}
+    \end{center}
+  \end{figure}
 
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Monte Carlo Solution Methods for Nonlinear Problems}
+\begin{frame}{My Weak Scaling}
 
-  \begin{itemize}
-  \item Many multiphysics problems of interest are nonlinear
-    \medskip
-  \item Segregated methods lack the consistency of fully implicit
-    methods
-    \medskip
-  \item Newton methods often leverage Krylov solvers
-    \begin{itemize}
-    \item Robust implementations
-    \item No operator required
-    \item Memory intensive
-    \end{itemize}
-    \medskip
-  \item Monte Carlo methods need the full operator
-    \medskip
-  \item Automatic construction of the linear operator available
-    \begin{itemize}
-    \item Ideal for Monte Carlo
-    \item Relaxes memory requirements
-    \item Potential scaling improvements
-    \item Resiliency benefits
-    \end{itemize}
-  \end{itemize}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Nonlinear Preliminaries}
-
-  \begin{itemize}
-  \item We seek solutions of the general nonlinear problem
-  \end{itemize}
-
-  \[
-  \ve{F}(\ve{u}) = \ve{0}
-  \]
-  \[
-  \ve{u} \in \mathbb{R}^n,\ \ve{F}:\mathbb{R}^N \rightarrow
-  \mathbb{R}^N
-  \]
-
-  \begin{itemize}
-  \item We interpret the exact solution $\ve{u}$ to be the roots of
-    $\ve{F}(\ve{u})$
-  \end{itemize}
-
-  \[
-  \ve{F}(\ve{u}^{k+1}) = \ve{F}(\ve{u}^{k}) +
-  \ve{F}'(\ve{u}^{k})(\ve{u}^{k+1}-\ve{u}^{k}) +
-  \frac{\ve{F}''(\ve{u}^{k})}{2}(\ve{u}^{k+1}-\ve{u}^{k})^2 + \cdots
-  \]
-
-  \begin{itemize}
-  \item Form Newton's method
-  \end{itemize}
-  \[
-  \ve{J}(\ve{u}) \delta \ve{u}^k = -\ve{F}(\ve{u}^{k})
-  \]
-  \[
-  \ve{u}^{k+1} = \ve{u}^k + \delta \ve{u}^k
-  \]
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Newton-Krylov Methods}
-
-  \begin{itemize}
-  \item Choose a Krylov method to solve for the Newton correction
-    \medskip
-  \item GMRES with a long recurrence relation observed as more
-    robust (Knoll,2004)
-    \medskip
-  \item Generates a monotonically decreasing residual from
-    maintaining the optimization and orthogonality conditions
-  \end{itemize}
-
-  \medskip \medskip \medskip
-  \pause Where does the Jacobian come from?
-  \begin{itemize}
-  \item The Jacobian can come from hand-coded derivatives
-    \begin{itemize}
-    \item Tedious and error prone
-    \item Repeated for each equation set and hard to do for
-      multiphysics
-    \end{itemize}
-  \end{itemize}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Jacobian Approximations}
-
-  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
-    {Matrix-Free Approximation}
-    \begin{itemize}
-    \item Krylov methods only need the action of the linear operator
-    \end{itemize}
-
-    \[
-    \ve{J}(\ve{u})\ve{v} = \frac{\ve{F}(\ve{u} + \epsilon \ve{v}) -
-      \ve{F}(\ve{u})}{\epsilon}
-    \]
-
-    \begin{itemize}
-    \item Forms the basis of Jacobian-Free Newton-Krylov (JFNK) methods
-      \begin{itemize}
-      \item Sensitive to scaling and discretization error (Kelly,1995)
-      \item Eventually break even with generating and storing the full
-        Jacobian (Knoll,1994)
-      \end{itemize}
-    \end{itemize}
-  \end{beamerboxesrounded}
-
-  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
-    {Automatic Differentiation}
-    \begin{itemize}
-    \item Automatically generate Jacobians from nonlinear function
-      evaluations
-      \begin{itemize}
-      \item Overload math operators and apply the chain rule (FAD)
-      \item Yields evaluations as accurate as function discretization
-      \end{itemize}
-    \item Performance studies give acceptable results for use in
-      large-scale, production physics codes (Bartlett,2006)
-    \end{itemize}
-  \end{beamerboxesrounded}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Jacobian Storage vs. Subspace Storage}
-
-  \begin{columns}
-
-    \begin{column}{0.5\textwidth}
-      \begin{itemize}
-      \item Jacobian will be in a compressed row storage format
-      \end{itemize}
-    \end{column}
-
-    \begin{column}{0.5\textwidth}
-      \small{\[ \ve{A} =
-        \begin{bmatrix}
-          2 & 0 & 8 & 0 & 0 & 0 \\ 4 & 5 & 0 & 1 & 0 & 0 \\ 0 & 2 & 1
-          & 0 & 1 & 0 \\ 0 & 0 & 3 & 7 & 0 & 2 \\ 0 & 0 & 0 & 4 & 9 &
-          0 \\ 0 & 0 & 0 & 0 & 9 & 1 \\
-        \end{bmatrix}      
-        \]}
-    \end{column}
-
-  \end{columns}
-
-  \small{ \[
-    \begin{array}{lccccccccccccccc}
-      \ve{A}\ values: & 2 & 8 & 4 & 5 & 1 & 2 & 1 & 1 & 3 & 7 & 2 & 4
-      & 9 & 9 & 1 \\ column: & 1 & 3 & 1 & 2 & 4 & 2 & 3 & 5 & 3 & 4 &
-      6 & 4 & 5 & 5 & 6 \\ row\ start: & 1 & 3 & 6 & 9 & 12 & 14 & 16
-      \\
-    \end{array}
-    \]}
-
-  \begin{itemize}
-  \item CRS matrix with $q$ bands storage: $\big[(2q+1)N\big]$ for $\ve{x} \in
-    \mathbb{R}^{N \times N}$ 
-    \pause
-  \item $m$ Krylov iterations require $(m+1)$ subspace vectors
-  \item Subspace storage: $\big[(m+1) N\big]$ for $\ve{x} \in
-    \mathbb{R}^{N \times N}$
-  \end{itemize}
-
-  \pause 
-  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
-    {If we need 25 Krylov iterations to converge...}
-    \begin{itemize}
-    \item Break-even scenario: $4q+1 = m$
-    \item Jacobian and probability matrix storage with 6 bands or less
-      is cheaper
-    \end{itemize}
-  \end{beamerboxesrounded}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}[fragile]{The FANM Method}
-
-  Forward-Automated Newton-MCSA
-
-  \begin{algorithm}[H]
-    \begin{algorithmic}[1]
-      \STATE $k := 0$ 
-      \WHILE{$||\ve{F}(\ve{u}^{k})|| > \epsilon
-        ||\ve{F}(\ve{u}^{0})||$} 
-      \STATE $\ve{J}(\ve{u}^{k}) \leftarrow AD(\ve{F}(\ve{u}^k))$ 
-      \COMMENT{Automatic differentiation} 
-      \STATE $\ve{J}(\ve{u}^k) \delta \ve{u}^k = -\ve{F}(\ve{u}^{k})$
-      \COMMENT{Solve for the Newton correction with MCSA} 
-      \STATE $\ve{u}^{k+1} \leftarrow \ve{u}^k + \delta \ve{u}^k$ 
-      \STATE $k \leftarrow k+1$ 
-      \ENDWHILE
-    \end{algorithmic}
-    \caption{FANM}
-  \end{algorithm}
-
-  \begin{itemize}
-  \item Robustness of Newton's method (inexact)
-  \item Accuracy and convenience of FAD
-  \item Parallelism, memory, and resiliency benefits of MCSA
-  \item Requires only nonlinear function evaluations
-  \end{itemize}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Parallel FANM Method}
-
-  \begin{algorithm}[H]
-    \begin{algorithmic}[1]
-      \STATE $k := 0$ 
-      \WHILE{$||\ve{F}(\ve{u}^{k})|| > \epsilon
-        ||\ve{F}(\ve{u}^{0})||$} 
-      \STATE $\ve{J}(\ve{u}^{k}) \leftarrow AD(\ve{F}(\ve{u}^k))$ 
-      \COMMENT{Automatic differentiation} 
-      \STATE $\ve{J}(\ve{u}^k) \delta \ve{u}^k = -\ve{F}(\ve{u}^{k})$
-      \COMMENT{Solve for the Newton correction with MCSA} 
-      \STATE $\ve{u}^{k+1} \leftarrow \ve{u}^k + \delta \ve{u}^k$ 
-      \STATE $k \leftarrow k+1$ 
-      \ENDWHILE
-    \end{algorithmic}
-    \caption{FANM}
-  \end{algorithm}
-
-  \begin{itemize}
-  \item Modern FAD packages are parallelized using element-level
-    assembly
-  \item This work will generate a parallel MCSA solver
-  \item Application of the Newton correction requires a parallel
-    vector update
-  \item Convergence checks through parallel norm computation
-  \end{itemize}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Research Proposal}
-
-  \large{
-  \begin{itemize}
-  \item Methods verification
-  \medskip \medskip \medskip \medskip \medskip
-  \item Numerical experiments
-    \medskip \medskip \medskip \medskip \medskip
-  \item Challenge problem
-  \end{itemize}
-  }
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Monte Carlo Methods Verification}
-
-  \begin{itemize}
-  \item Analytic solution to the heat equation for the linear methods
-    \medskip
-  \item Sequence of Navier-Stokes benchmarks for the nonlinear methods
-    \begin{itemize}
-    \item Thermal convection cavity problem (De Vahl Davis, 1983)
-    \item Lid driven cavity problem (Ghia et al., 1982)
-    \item Backward-Facing step problem (Gartling, 1990)
-    \end{itemize}
-    \medskip
-  \item Tuning benchmark parameters varies the strength of
-    nonlinearities
-  \end{itemize}
-
-  \medskip  \medskip
-  \begin{columns}
-    \begin{column}{0.5\textwidth}
-      \[
-      \rho \ve{u} \cdot \nabla \ve{u} - \nabla \cdot \ve{T} - \rho
-      \ve{g} = \ve{0}
-      \]
-      \[
-      \nabla \cdot \ve{u} = 0
-      \]
-      \[
-      \rho C_p \ve{u} \cdot \nabla T + \nabla \cdot \ve{q} = 0
-      \]
-    \end{column}
-
-    \begin{column}{0.5\textwidth}
-      \[
-      \ve{T} = -P \ve{I} + \mu[\nabla \ve{u} + \nabla \ve{u}^T]
-      \]
-      \[
-      \ve{q} = - k \nabla T
-      \]
-    \end{column}
-  \end{columns}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Proposed Numerical Experiments}
-
-  \pause
-  \textbf{Domain Overlap Studies for Parallel Neumann-Ulam Method} 
-  \begin{itemize}
-  \item Correlate domain overlap behavior to linear system properties
-  \item Analyze random walk transport with respect to the operator:
-    \begin{itemize}
-    \item Eigenvalues
-    \item Sparsity
-    \item Asymmetry
-    \end{itemize}
-  \item Evaluate communication cost
-  \item Evaluate memory cost
-  \end{itemize}
-
-  \medskip  \medskip
-  \pause
-  \textbf{Domain Replication Studies for Parallel Neumann-Ulam Method}
-  \begin{itemize}
-  \item Feasibility from a memory perspective
-  \item Impact on scalability
-  \item Impact on load balancing
-  \end{itemize}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Proposed Numerical Experiments}
-
-  \textbf{Parallel Performance and Numerical Accuracy Studies for MCSA
-    Method}
-  \begin{itemize}
-  \item Characterize Neumann-Ulam parameters required for good
-    correction
-  \item Feasibility of MCSA with full-clip Neumann-Ulam
-  \item Performance for asymmetric systems
-  \item Memory usage compared to Krylov methods
-  \end{itemize}
-
-  \medskip  \medskip
-  \pause
-  \textbf{Parallel Performance and Numerical Accuracy Studies for FANM
-    Method}
-  \begin{itemize}
-  \item Feasibility for problems of interest
-  \item Memory usage vs. Newton-Krylov methods
-  \item Scalability vs. Newton-Krylov methods
-  \end{itemize}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Proposed Challenge Problem}
-
-  \begin{itemize}
-  \item Problems of interest are coupled large scale problems in
-    reactor physics
-    \medskip
-  \item The Consortium for Advanced Simulation of LWRs (CASL) is a
-    modeling and simulation program aimed at:
-    \begin{itemize}
-      \item Higher power uprates and efficiency
-      \item Higher burn-up
-      \item Predictive accident scenario analysis
-    \end{itemize}
-    \medskip
-  \item CASL and industry partners identified challenge problems:
-    \begin{itemize}
-    \item Departure from nucleate boiling
-    \item Grid-to-rod-fretting
-    \end{itemize}
-    \medskip
-  \item Our challenge problem should reflect how this work aids the
-    solution of these problems
-  \end{itemize}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Proposed Challenge Problem}
-
-  \begin{columns}
-    \begin{column}{0.5\textwidth}
-      \begin{figure}[htpb!]
+  \begin{figure}[h!]
     \begin{center}
-          \includegraphics[width=2.5in]{drekar1.png}
+      \includegraphics[width=4in]{my_weak.png}
     \end{center}
   \end{figure}
-    \end{column}
-
-    \begin{column}{0.5\textwidth}
-      \begin{figure}[htpb!]
-        \begin{center}
-          \includegraphics[width=2.5in]{drekar2.png}
-        \end{center}
-      \end{figure}
-    \end{column}
-  \end{columns}
-
-  \begin{itemize}
-  \item CASL has utilized the Drekar multiphysics code (SNL)
-  \item Coupled fluid flow and heat transfer helps characterize many
-    phenomena
-  \item Drekar is massively parallel and leverages Newton-Krylov
-    methods
-  \item Propose using the largest Drekar problem to date as a
-    challenge problem for the new Monte Carlo methods
-  \end{itemize}
-
-  {\small \sl Images source: www.casl.gov}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Conclusion}
-
-  \begin{itemize}
-  \item Proposed research and development of new Monte Carlo methods
-    \begin{itemize}
-    \item Parallelization of Monte Carlo methods for linear systems
-    \item FANM for nonlinear systems
-    \end{itemize}
-    \medskip
-  \item Verification through benchmarks
-    \medskip
-  \item Numerical experiments for understanding
-    \medskip
-  \item Challenge problem for application
-    \medskip
-  \item Directed towards hard problems in nuclear reactor analysis:
-    \begin{itemize}
-    \item Application to multiphysics
-    \item Potential improvements for scalability
-    \item Potential improvements for memory consumption
-    \item Looking forward to exascale
-    \end{itemize}
-  \end{itemize}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Acknowledgements}
-
-  Thanks to...
-
-  \begin{itemize}
-  \item Paul Wilson
-  \item Tom Evans
-  \item ORNL and SNL
-  \end{itemize}
-
-  \medskip \medskip \medskip \medskip \medskip \medskip \medskip
-
-  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
-    {Support} 
-    This work was performed under appointment to the Nuclear Regulatory
-    Commission Fellowship program at the University of Wisconsin - Madison
-    Department of Engineering Physics.
-  \end{beamerboxesrounded}
 

### 2013-02-06 doc/presentations/meeting_2_6_13/meeting_2_6_13.tex
updated my comm param png
adding communication figure

@@
   \begin{itemize}
-  \item No literature observed for parallel Neumann-Ulam solvers
+  \item No literature observed for domain decomposed Neumann-Ulam
+    solvers; parallelism over histories only (domain replication)
     \medskip \medskip
@@
     Neumann-Ulam method
+  \item New implementation based on the Shift Monte Carlo code
   \end{itemize}
@@
     communication
+  \item Brunner and Brantley, \textit{An efficient, robust, domain
+    decomposition algorithm for particle Monte Carlo}, Journal of
+    Computational Physics, 2009
   \end{itemize}
 
-  \medskip \medskip
-  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
-    {Brunner's Work (2006 and 2009)}
-
-    \begin{itemize}
-    \item Looked at 4 communication patterns:
-      \begin{itemize}
-      \item Fully-locking synchronous
-      \item Asynchronous-send/synchronous-receive
-      \item Master/slave
-      \item Binary tree master/slave
-      \end{itemize}
-    \item Binary tree master/slave performed best but race conditions observed
-    \item A later improvement to their work showed a fully asynchronous
-      pattern to work best
-    \item Poor scaling for unbalanced problems for all schemes
-    \end{itemize}
-
-  \end{beamerboxesrounded}
+  \begin{figure}[htpb!]
+    \begin{center}
+      \scalebox{1.0}{ \input{procassini_example.pdftex_t} }
+    \end{center}
+  \end{figure}
 
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Domain Decomposition}
+\begin{frame}{Domain Decomposition Parallelism}
+
+  \begin{figure}[h!]
+    \begin{center}
+      \includegraphics[width=3.75in]{domain_comm.png}
+    \end{center}
+  \end{figure}
+
 \end{frame}
@@
 \begin{frame}{Fully Synchronous Algorithm}
+
+  Do until all histories completed on all procs:
+
+  \begin{enumerate}
+  \item All local histories transported to local completion
+    \begin{itemize}
+    \item Histories are buffered in the send buffer
+    \item The send buffer has no size limit
+    \end{itemize}
+  \item Fully locking communication to determine the end of local
+    transport
+    \medskip
+    \begin{itemize}
+    \item If send buffers are not empty: fully locking communication
+      of send buffers to receive buffers
+    \item If they are empty, fully locking communication to end transport
+    \end{itemize}
+  \end{enumerate}
+
 \end{frame}
@@
 \begin{frame}{Fully Asynchronous Algorithm}
+
+  Do until all histories completed on all procs:
+
+  \begin{enumerate}
+  \item Specified number of histories transported to local completion
+    (check frequency)
+    \begin{itemize}
+    \item If at check frequency, non-locking receive buffers
+    \item Any locally terminated histories non-locking communicated to
+      the master
+    \item Histories are buffered in the send buffers as needed
+    \item The send buffer has a size limit (max buffer size)
+    \item If any buffer is full, non-locking send it
+    \end{itemize}
+    \medskip
+  \item Master non-locking check for completed histories
+    \begin{itemize}
+    \item If completed total on master sums to total source
+      histories, send non-locking complete message
+    \end{itemize}
+  \end{enumerate}
+
 \end{frame}
@@
     \begin{center}
-      \includegraphics[width=5in]{my_params.png}
+      \includegraphics[width=4.75in]{my_params.png}
     \end{center}
@@
     \begin{center}
-      \includegraphics[width=4in]{my_strong.png}
+      \includegraphics[width=3.5in]{my_strong.png}
     \end{center}
@@
     \begin{center}
-      \includegraphics[width=4in]{my_weak.png}
+      \includegraphics[width=3in]{my_weak.png}
     \end{center}

### 2013-02-19 doc/jcomp2012/jcomp2012.tex
Finalized all files for submission to JCOMP; paper submitted\!
updatting jcomp paper with new comparison results for reviewer response

@@
 
-A timing and convergence study is used to demonstrate the effectiveness of the
-adjoint method as compared to the direct method. To assess both the CPU time
-and number of iterations required to converge to a solution, a problem of
-constant $\Delta$ was used with varying values of grid size, fixing the
-spectral radius of the system at a constant value for each variation. Both the
-five-point and nine-point stencils were used with both the direct and adjoint
-solvers. For each case, 50 random walk permutations were computed per MCSA
-iteration with a weight cutoff of \sn{1}{-4} for the stochastic linear solver
-and a convergence tolerance of \sn{1}{-8} for the MCSA iterative solver. All
-computations presented in this section were completed on a 2.66 GHz Intel Core
-i7 machine with 8 GB 1067 MHz DDR3 memory. Figure~\ref{fig:poisson_cpu_time}
-gives the CPU time needed for each case to converge in seconds, and
-Fig.~\ref{fig:poisson_iterations} gives the number of iterations needed for
-each case to converge relative to the problem size.
+A timing and convergence study is used to demonstrate the
+effectiveness of the adjoint method as compared to the direct
+method. To assess both the CPU time and number of iterations required
+to converge to a solution, a problem of constant $\Delta$ was used
+with varying values of grid size, fixing the spectral radius of the
+system at a constant value for each variation. Both the five-point and
+nine-point stencils were used with both the direct and adjoint
+solvers. For each case, $N \times N$ total random walk permutations
+were computed per MCSA iteration where $N \times N$ is the number of
+discrete grid points in the system. Solver parameters were set to a
+weight cutoff of \sn{1}{-4} for the stochastic linear solver and a
+convergence tolerance of \sn{1}{-8} for the MCSA iterative solver. All
+computations presented in this section were completed on a 2.66 GHz
+Intel Core i7 machine with 8 GB 1067 MHz DDR3
+memory. Figure~\ref{fig:poisson_cpu_time} gives the CPU time needed
+for each case to converge in seconds, and
+Figure~\ref{fig:poisson_iterations} gives the number of iterations
+needed for each case to converge relative to the problem size.
 \begin{figure}[h!]
   \centering
-  \includegraphics[width=5in,clip]{Adjoint_Direct_CPU_Time.pdf}
-  \caption{\sl CPU Time (s) to converge vs. Problem Size ($N$ for an $N \times
-    N$ square mesh). Both the adjoint and direct solvers are used with the
-    five point and nine point stencils. A significant CPU time speedup is
-    noted with the adjoint method due to the reduced number of random walk
-    events.}
+  \includegraphics[width=5in,clip]{Adjoint_Direct_CPU_Time_2.pdf}
+  \caption{\sl CPU Time (s) to converge vs. Problem Size ($N$ for an
+    $N \times N$ square mesh). Both the adjoint and direct solvers are
+    used with the five point and nine point stencils. A CPU time
+    speedup is noted with the adjoint method due to the higher density
+    of random walk events in regions with a large residual.}
   \label{fig:poisson_cpu_time}
@@
   \centering
-  \includegraphics[width=5in,clip]{Adjoint_Direct_Iterations.pdf}
-  \caption{\sl Iterations to converge vs. Problem Size ($N$ for an $N \times
-    N$ square mesh). Both the adjoint and direct solvers are used with the
-    five-point and nine-point stencils. Both methods give similar iteration
-    behavior.}
+  \includegraphics[width=5in,clip]{Adjoint_Direct_Iterations_2.pdf}
+  \caption{\sl Iterations to converge vs. Problem Size ($N$ for an $N
+    \times N$ square mesh). Both the adjoint and direct solvers are
+    used with the five-point and nine-point stencils. }
   \label{fig:poisson_iterations}
@@
 
-We see clearly in Fig.~\ref{fig:poisson_cpu_time} that the using the adjoint
-solver with MCSA results in several orders of magnitude speedup over the
-direct solver while the number of iterations required to converge is of a
-similar scale. We expect this for several reasons. First, with an equivalent
-number of histories specified for both solvers and a system of size $N \times
-N$, the direct solver will compute $N \times N$ random walks per permutation
-while the adjoint solver will only compute one. This is necessary in the
-direct method to ensure a contribution from each state as the random walk
-sequence will only contribute to the starting state. For the adjoint method,
-because the random walk sequence contributes to the state in which it
-currently resides, fewer histories are neccessary to compute contributions
-from all the nonzero parts of the source. Initially, this may not seem like a
-fair comparison; however, we see from Fig.~\ref{fig:poisson_iterations} that
-the number of iterations required to converge is approximately the same and
-therefore the extra computations performed by the direct method do not improve
-the error estimation. Therefore, although the total number of random walks per
-iteration is in fact larger for the direct method, the solution computed at
-each iteration is no better than that computed by the adjoint method.
-
-As an additional comparison, the convergence behavior of MCSA can be analyzed
-using both solvers to detect any performance benefits. To assess the
-convergence properties of MCSA using each solver and stencil, the infinity
-norm of the residual computed in Eq.~(\ref{eq:MCSA-residual}) was collected at
-each iteration for a fixed problem size. Figure~\ref{fig:poisson_convergence}
-gives the results of these computations. First, it is worthy to note on the
-semilog plot that we are indeed achieving the expected exponential convergence
-from MCSA. Second, we note that both the direct and adjoint methods have
-approximately equivalent convergence behavior, leading to CPU time being the
-dominating factor in method selection.
+We see clearly in Figure~\ref{fig:poisson_cpu_time} that the using the
+adjoint solver with MCSA results in a speedup over the direct solver
+while the number of iterations required to converge is also reduced as
+shown in Figure~\ref{fig:poisson_iterations}. We expect this for several
+reasons. First, with an equivalent number of histories specified for
+both solvers per MCSA iteration and a system of size $N \times N$, the
+direct solver will compute a single random walk for each state in the
+system per iteration to acquire a solution in that state, regardless
+of the size of the residual in that state. This is necessary in the
+direct method to ensure a contribution from each state as the random
+walk sequence will only contribute to the starting state. For the
+adjoint method, a total of $N \times N$ random walk events will have
+their starting state determined by sampling the residual
+vector. Because the random walk sequence contributes to the state in
+which it currently resides, sampling the residual vector as the Monte
+Carlo source gives a higher density of random walk events in regions
+with a high residual, thus giving a more accurate correction in that
+region due to reduced statistical error. From an iteration
+perspective, Figure~\ref{fig:poisson_iterations} shows that using the
+direct method yields a roughly unchanging number of iterations
+required to converge as the problem size increases. Again, if we
+desire a correction value for all states in the problem, then we must
+start a random walk in each state in the system which does not reduce
+the number of iterations need as the problem size grows. Conversely,
+as the problem size grows in the adjoint method, the additional
+stochastic histories that will be computed are concentrated in regions
+with a large residual, further reducing the stochastic error in the
+correction in those regions and subsequently reducing the required
+number of iterations to converge.
+
+As an additional comparison, the convergence behavior of MCSA can be
+analyzed using both the adjoint and direct solvers to detect any
+performance benefits. To assess the convergence properties of MCSA
+using each solver and stencil, the infinity norm of the residual
+computed in Eq.~(\ref{eq:MCSA-residual}) was collected at each
+iteration for a fixed problem size of
+$N=500$. Figure~\ref{fig:poisson_convergence} gives the results of
+these computations. First, it is worthy to note on the semilog plot
+that we are indeed achieving the expected exponential convergence from
+MCSA with both Monte Carlo solvers. Second, we note that using the
+adjoint method with the same number of stochastic histores per MCSA
+iteration gives a faster rate of converge for the same reasons as
+above.
 \begin{figure}[h!]
   \centering
-  \includegraphics[width=5in,clip]{Adjoint_Direct_Convergence.pdf}
-  \caption{\sl Infinity norm of the solution residual vs. iteration number for
-    a problem of fixed size. Both the adjoint and direct solvers are used with
-    the five point and nine point stencils.}
+  \includegraphics[width=5in,clip]{Adjoint_Direct_Convergence_2.pdf}
+  \caption{\sl Infinity norm of the solution residual vs. iteration
+    number for a problem of size $N=500$. Both the adjoint and direct
+    solvers are used with the five point and nine point stencils. A
+    higher rate of convergence is observed for MCSA using the adjoint
+    Monte Carlo solver as compared to the direct method when both
+    solvers compute the same number of random walks per iteration.}
   \label{fig:poisson_convergence}
+\end{figure}
+
+\subsection{Comparison to Sequential Monte Carlo}
+\label{subsec:sequential comparison}
+To further motivate using Monte Carlo Synthetic Acceleration, we
+compare its performance to Halton's Sequential Monte Carlo method on
+which our previous work in this area was based. For this comparison,
+we use the same transient Poisson problem as described in the previous
+section and choose only the 5-point stencil to discretize the
+Laplacian operator as the previous results yielded little qualitative
+difference between the discretizations. Both MCSA and Halton's method
+are used with the adjoint Monte Carlo solver. In order to complete the
+same study as in the previous section, the number of histories
+computed by the Monte Carlo solver at each iteration had to be doubled
+to $2 \times N \times N$ in order to ensure convergence and a
+monotonically decreasing residual in Sequential Monte Carlo
+Method. Figure~\ref{fig:seq_cpu_time} gives the CPU time results for
+this comparison as a function of problem size while
+Figure~\ref{fig:seq_iterations} gives the number of iterations to
+converge as a function of problem size. In both cases, using the Monte
+Carlo solver as a synthetic acceleration rather than in a pure
+residual Monte Carlo scheme resulted in a reduction in both CPU time
+and iterations required to converge. The additional Richardson
+extrapolation between each Monte Carlo solve in the MCSA method gives
+a better converged residual source to use with the Monte Carlo
+calculation while the Sequential method requires more iterations to
+achieve the same level of convergence in the residual.
+
+\begin{figure}[h!]
+  \centering
+  \includegraphics[width=5in,clip]{Sequential_CPU_Time.pdf}
+  \caption{\sl CPU Time (s) to converge vs. Problem Size ($N$ for an
+    $N \times N$ square mesh). Both the Sequential Monte Carlo and
+    MCSA solvers are used with the five point stencils and the adjoint
+    Monte Carlo solver. The number of random walks was twice the
+    number of discrete states in the system in order to ensure
+    convergence in the Sequential Monte Carlo method.}
+  \label{fig:seq_cpu_time}
+\end{figure}
+
+\begin{figure}[h!]
+  \centering
+  \includegraphics[width=5in,clip]{Sequential_Iterations.pdf}
+  \caption{\sl Iterations to converge vs. Problem Size ($N$ for an $N
+    \times N$ square mesh). Both the Sequential Monte Carlo and MCSA
+    solvers are used with the five point stencils and the adjoint
+    Monte Carlo solver.}
+  \label{fig:seq_iterations}
+\end{figure}
+
+The benefits of using a synthetic acceleration scheme are also noted
+when the infinity norm of the residual computed computed at each
+iteration for both methods was collected at each iteration for a fixed
+problem sizes of $N=100$ and $N=500$ as shown in figures
+Figure~\ref{fig:seq_100} and \ref{fig:seq_500} respectively. In both
+cases, the Sequential method is subject to two regimes of exponential
+convergence with the later regime converging the slowest while the
+MCSA method exhibits a single rate of exponential convergence, much
+higher than that computed by Halton's method.
+
+\begin{figure}[h!]
+  \centering
+  \includegraphics[width=5in,clip]{Sequential_Convergence.pdf}
+  \caption{\sl Infinity norm of the solution residual vs. iteration
+    number for a problem of size $N=100$. Both the Sequential Monte
+    Carlo and MCSA solvers are used with the five point stencils and
+    the adjoint Monte Carlo solver.}
+  \label{fig:seq_100}
+\end{figure}
+
+\begin{figure}[h!]
+  \centering
+  \includegraphics[width=5in,clip]{Sequential_Convergence.pdf}
+  \caption{\sl Infinity norm of the solution residual vs. iteration
+    number for a problem of size $N=500$. Both the Sequential Monte
+    Carlo and MCSA solvers are used with the five point stencils and
+    the adjoint Monte Carlo solver.}
+  \label{fig:seq_500}
 \end{figure}
@@
 Integrating Eq.~(\ref{eq:flux-balance}) over the cell illustrated in
-Fig.~\ref{fig:cell} gives
+Figure~\ref{fig:cell} gives
 \begin{equation}
@@
 \begin{figure}[h!]
-  \centerline{ \input{mesh_cell.pdftex_t}}
+  \centerline{ \includegraphics[clip]{mesh_cell.pdf} }
   \caption{Orthogonal 3D mesh cell.}
@@
 The connectivity of an arbitrary cell is illustrated in the 3D mesh segment in
-Fig.~\ref{fig:mesh_segment}.  For a particle residing in the blue cell, any
+Figure~\ref{fig:mesh_segment}.  For a particle residing in the blue cell, any
 collision results in a scatter to one of the six adjacent cells since
@@
 \begin{figure}[h!]
-  \centerline{ \input{3D_mesh_segment.pdftex_t}}
+  \centerline{ \includegraphics[clip]{3D_mesh_segment.pdf}}
   \caption{3D mesh-segment.  The particle's current cell (in blue) is
@@
 traditional deterministic solution methods.  The deterministic methods
-considered are preconditioned Conjugate Gradient (PCG) and preconditioned
-fixed-point (Richardson) iteration (PFIX).  The PCG method uses right-left
-preconditioning with $\ve{M}=\ve{L}\ve{L}^{T}$, where $\ve{L} = \ve{L}^{T} =
-\sqrt{\diag(\ve{D})}$.  Naturally, more powerful preconditioning methods could be
-used; however, this provides a fair comparison with the preconditioning used
-in the MCSA method.  Different precondioning strategies within MCSA will be
-the focus of a future paper.  The PFIX scheme is used to solve
-Eq.~(\ref{eq:preconditioned-diffusion}), which is identical to the system
-being solved with MCSA.  We examine each method on two 3D problems: a Marshak
-wave problem and a multi-material duct problem.
+considered are preconditioned Conjugate Gradient (PCG) and
+preconditioned fixed-point (Richardson) iteration (PFIX).  The PCG
+method uses right-left preconditioning with $\ve{M}=\ve{L}\ve{L}^{T}$,
+where $\ve{L} = \ve{L}^{T} = \sqrt{\diag(\ve{D})}$.  Naturally, more
+powerful preconditioning methods could be used; however, this provides
+a fair comparison with the preconditioning used in the MCSA method.
+Different preconditioning strategies within MCSA will be the focus of
+a future paper.  The PFIX scheme is used to solve
+Eq.~(\ref{eq:preconditioned-diffusion}), which is identical to the
+system being solved with MCSA.  We examine each method on two 3D
+problems: a Marshak wave problem and a multi-material duct problem.
 
@@
 $\ve{H} = (\vI - \ve{M}^{-1}\ve{D})$.  The spectral radius for the Marshak
-problem is plotted in Fig.~\ref{fig:marshak_spectral}.  As described in
+problem is plotted in Figure~\ref{fig:marshak_spectral}.  As described in
 \S~\ref{sec:preconditioning}, the preconditioning effectively reduces the
@@
 wall where the radiation flows into a thin region bound by a foil on one side.
-The mesh and geometry are illustrated in Fig.~\ref{fig:multi-mat-mesh}.  A
+The mesh and geometry are illustrated in Figure~\ref{fig:multi-mat-mesh}.  A
 blackbody boundary flux is defined on the low-$x$ face at $T=0.5$~keV.  The
@@
 A 2D plot of the temperature at 1000~ns elapsed time is illustrated in
-Fig.~\ref{fig:multi-mat_T}.  The time evolution of the temperature is shown at
-four edit points in Fig.~\ref{fig:edit-points}.  All three methods, PCG, PFIX,
+Figure~\ref{fig:multi-mat_T}.  The time evolution of the temperature is shown at
+four edit points in Figure~\ref{fig:edit-points}.  All three methods, PCG, PFIX,
 and MCSA, give numerically identical answers when using a stopping criterion
@@
     the edit points where the time evolution of the temperature is plotted in
-    Fig.~\ref{fig:edit-points} are indicated by the black disks.}
+    Figure~\re{fig:edit-points} are indicated by the black disks.}
   \label{fig:multi-mat_T}
@@
   \begin{center}
-    \includegraphics[width=5in,clip]{edit_points}
+    \includegraphics[width=5in,clip]{edit_points.pdf}
   \end{center}

### 2013-02-19 doc/jcomp2012/jcomp2012.tex
adding grace plots for direct and adjoint convergence

@@
   \centering
-  \includegraphics[width=5in,clip]{Adjoint_Direct_CPU_Time_2.pdf}
+  \includegraphics[width=5in,clip]{dir_adj_cpu.pdf}
   \caption{\sl CPU Time (s) to converge vs. Problem Size ($N$ for an
@@
   \centering
-  \includegraphics[width=5in,clip]{Adjoint_Direct_Iterations_2.pdf}
+  \includegraphics[width=5in,clip]{dir_adj_iterations.pdf}
   \caption{\sl Iterations to converge vs. Problem Size ($N$ for an $N
@@
   \centering
-  \includegraphics[width=5in,clip]{Adjoint_Direct_Convergence_2.pdf}
+  \includegraphics[width=5in,clip]{dir_adj_convergence.pdf}
   \caption{\sl Infinity norm of the solution residual vs. iteration
@@
   \centering
-  \includegraphics[width=5in,clip]{Sequential_CPU_Time.pdf}
+  \includegraphics[width=5in,clip]{seq_cpu.pdf}
   \caption{\sl CPU Time (s) to converge vs. Problem Size ($N$ for an
@@
   \centering
-  \includegraphics[width=5in,clip]{Sequential_Iterations.pdf}
+  \includegraphics[width=5in,clip]{seq_iterations.pdf}
   \caption{\sl Iterations to converge vs. Problem Size ($N$ for an $N
@@
   \centering
-  \includegraphics[width=5in,clip]{Sequential_Convergence.pdf}
+  \includegraphics[width=5in,clip]{seq_conv_100.pdf}
   \caption{\sl Infinity norm of the solution residual vs. iteration
@@
   \centering
-  \includegraphics[width=5in,clip]{Sequential_Convergence.pdf}
+  \includegraphics[width=5in,clip]{seq_conv_500.pdf}
   \caption{\sl Infinity norm of the solution residual vs. iteration

### 2013-02-19 doc/jcomp2012/jcomp2012.tex
adding plots for sequential vs mcsa convergence.

@@
 computed by the Monte Carlo solver at each iteration had to be doubled
-to $2 \times N \times N$ in order to ensure convergence and a
-monotonically decreasing residual in Sequential Monte Carlo
-Method. Figure~\ref{fig:seq_cpu_time} gives the CPU time results for
-this comparison as a function of problem size while
+to $2 \times N \times N$ in order to ensure convergence in Sequential
+Monte Carlo Method. Figure~\ref{fig:seq_cpu_time} gives the CPU time
+results for this comparison as a function of problem size while
 Figure~\ref{fig:seq_iterations} gives the number of iterations to
@@
 MCSA method exhibits a single rate of exponential convergence, much
-higher than that computed by Halton's method.
+higher than that computed by Halton's method. Even with the doubling
+of the number of stochastic histories computed per time step in order
+to ensure convergence for the Sequential method, we still see
+robustness issues with a non-monotonically decreasing residual
+observed in both cases and more pronounced for the $N=100$ case. In
+both cases the MCSA solver is observed to be robust with a
+monotonically decreasing residual.
 

### 2013-02-19 doc/jcomp2012/jcomp2012.tex
fixing document build

@@
 needed for each case to converge relative to the problem size.
-\begin{figure}[h!]
+\begin{figure}[ht!]
   \centering
@@
 
-\begin{figure}[h!]
+\begin{figure}[ht!]
   \centering
@@
 above.
-\begin{figure}[h!]
+\begin{figure}[ht!]
   \centering
-  \includegraphics[width=5in,clip]{dir_adj_convergence.pdf}
+  \includegraphics[width=5in,clip]{dir_adj_conv.pdf}
   \caption{\sl Infinity norm of the solution residual vs. iteration
@@
 The benefits of using a synthetic acceleration scheme are also noted
-when the infinity norm of the residual computed computed at each
-iteration for both methods was collected at each iteration for a fixed
-problem sizes of $N=100$ and $N=500$ as shown in figures
-Figure~\ref{fig:seq_100} and \ref{fig:seq_500} respectively. In both
-cases, the Sequential method is subject to two regimes of exponential
-convergence with the later regime converging the slowest while the
-MCSA method exhibits a single rate of exponential convergence, much
-higher than that computed by Halton's method. Even with the doubling
-of the number of stochastic histories computed per time step in order
-to ensure convergence for the Sequential method, we still see
-robustness issues with a non-monotonically decreasing residual
-observed in both cases and more pronounced for the $N=100$ case. In
-both cases the MCSA solver is observed to be robust with a
+when the infinity norm of the residual computed at each iteration for
+both methods was collected at each iteration for a fixed problem sizes
+of $N=100$ and $N=500$ as shown in figures Figure~\ref{fig:seq_100}
+and \ref{fig:seq_500} respectively. In both cases, the Sequential
+method is subject to two regimes of exponential convergence with the
+later regime converging the slowest while the MCSA method exhibits a
+single rate of exponential convergence, much higher than that computed
+by Halton's method. Even with the doubling of the number of stochastic
+histories computed per time step in order to ensure convergence for
+the Sequential method, we still see robustness issues with a
+non-monotonically decreasing residual observed for the $N=100$
+case. In both cases the MCSA solver is observed to be robust with a
 monotonically decreasing residual.
 
-\begin{figure}[h!]
+\begin{figure}[ht!]
   \centering
@@
 and $V_{ijk} = \Delta_i\Delta_j\Delta_k$.
-\begin{figure}[h!]
+\begin{figure}[ht!]
   \centerline{ \includegraphics[clip]{mesh_cell.pdf} }
@@
 Cells along the problem boundary can have fewer than six entries.
-\begin{figure}[h!]
+\begin{figure}[ht!]
   \centerline{ \includegraphics[clip]{3D_mesh_segment.pdf}}
@@
 $1/\sqrt{N}$~\cite{halton_1994,evans_2003}.
-\begin{figure}[h!]
+\begin{figure}[ht!]
   \centerline{ \includegraphics[width=5in,clip]{marshak_10.pdf}}
@@
 spectral radius of $\ve{H}$ to less than 1.
-\begin{figure}[h!]
+\begin{figure}[ht!]
   \centerline{ \includegraphics[width=5in,clip]{spectral_radius.pdf}}
@@
 running more stages with fewer particles per stage \cite{evans_2003}.
-\begin{figure}[h!]
+\begin{figure}[ht!]
   \centerline{ \includegraphics[width=5in,clip]{mrshk_np_CPU.pdf}}
@@
 
-\begin{figure}[h!]
+\begin{figure}[ht!]
   \centerline{ \includegraphics[width=5in,clip]{mrshk_np_Ns.pdf}}
@@
 particles are used.
-\begin{figure}[h!]
+\begin{figure}[ht!]
   \centerline{ \includegraphics[width=5in,clip]{mrshk_wc_CPU.pdf}}
@@
 alone.
-\begin{table}[h!]
+\begin{table}[ht!]
   \caption{Comparison of solution methods for the Marshak problem. All
@@
 of $\sn{1}{-8}$.
-\begin{figure}[h!]
+\begin{figure}[ht!]
   \centerline{ \includegraphics[width=5in,clip]{mesh.pdf}}
@@
 
-\begin{figure}[h!]
+\begin{figure}[ht!]
   \centerline{ \includegraphics[width=5in,clip]{T_multi_mat.pdf}}
-  \caption{ Two-dimensional plot of the temperature in keV at 1000~ns on a
-    cut-plane positioned at the mid-point of the $z$-axis.  The positions of
-    the edit points where the time evolution of the temperature is plotted in
-    Figure~\re{fig:edit-points} are indicated by the black disks.}
+  \caption{ Two-dimensional plot of the temperature in keV at 1000~ns
+    on a cut-plane positioned at the mid-point of the $z$-axis.  The
+    positions of the edit points where the time evolution of the
+    temperature is plotted in Figure~\ref{fig:edit-points} are
+    indicated by the black disks.}
   \label{fig:multi-mat_T}
@@
 
-\begin{figure}[h!]
+\begin{figure}[ht!]
   \begin{center}

### 2013-02-19 doc/jcomp2012/jcomp2012.tex
added 9 point convergence explanation in response to jcomp reviewers comments

@@
 weight cutoff of \sn{1}{-4} for the stochastic linear solver and a
-convergence tolerance of \sn{1}{-8} for the MCSA iterative solver. All
-computations presented in this section were completed on a 2.66 GHz
-Intel Core i7 machine with 8 GB 1067 MHz DDR3
-memory. Figure~\ref{fig:poisson_cpu_time} gives the CPU time needed
-for each case to converge in seconds, and
-Figure~\ref{fig:poisson_iterations} gives the number of iterations
-needed for each case to converge relative to the problem size.
+convergence tolerance of \sn{1}{-8} for the MCSA iterative solver.
+Figure~\ref{fig:poisson_cpu_time} gives the CPU time needed for each
+case to converge in seconds and Figure~\ref{fig:poisson_iterations}
+gives the number of iterations needed for each case to converge to the
+specified tolerance as a function of the problem size. All
+computations presented in this section and
+\S~\ref{subsec:sequential_comparison} were completed on a 3.0 GHz
+Intel Core 2 Quad Q9650 CPU machine with 16 GB 1067 MHz DDR3 memory.
 \begin{figure}[ht!]
@@
 iteration gives a faster rate of converge for the same reasons as
-above.
+above. We also note here that fewer iterations are required for
+convergence when the 9-point stencil is used to discretize the
+Laplacian operator (although at no gain in speed as given by the
+results in Figure~\ref{fig:poisson_cpu_time}). This is due to the fact
+that the smaller discretization error directly corresponds to a more
+well defined residual source generated by the Richardson extrapolation
+for the Monte Carlo calculation. In addition, the better defined
+source is transported through a domain described more accurately by
+the 9-point stencil, thus yielding a more accuracte correction vector
+from the Monte Carlo calculation.
 \begin{figure}[ht!]
@@
 \subsection{Comparison to Sequential Monte Carlo}
-\label{subsec:sequential comparison}
+\label{subsec:sequential_comparison}
 To further motivate using Monte Carlo Synthetic Acceleration, we
@@
 later regime converging the slowest while the MCSA method exhibits a
-single rate of exponential convergence, much higher than that computed
-by Halton's method. Even with the doubling of the number of stochastic
-histories computed per time step in order to ensure convergence for
-the Sequential method, we still see robustness issues with a
-non-monotonically decreasing residual observed for the $N=100$
+single rate of exponential convergence observed to be much higher than
+that computed by Halton's method. Even with the doubling of the number
+of stochastic histories computed per time step in order to ensure
+convergence for the Sequential method, we still see robustness issues
+with a non-monotonically decreasing residual observed for the $N=100$
 case. In both cases the MCSA solver is observed to be robust with a

### 2013-02-27 doc/spn/spn.tex
adding SPn document

--- 2013-02-06 doc/presentations/meeting_2_6_13/meeting_2_6_13.tex
@@
-\documentclass{beamer}
-\usetheme[white]{Wisconsin}
+%        File: spn.tex
+\documentclass[letterpaper]{article}
+\usepackage[top=1.0in,bottom=1.0in,left=1.0in,right=1.0in]{geometry}
+\usepackage{verbatim}
+\usepackage{amssymb}
+\usepackage{graphicx}
 \usepackage{longtable}
+\usepackage{amsfonts}
+\usepackage{amsmath}
+\usepackage[usenames]{color}
+\usepackage[
+naturalnames = true, 
+colorlinks = true, 
+linkcolor = black,
+anchorcolor = black,
+citecolor = black,
+menucolor = black,
+urlcolor = blue
+]{hyperref}
 \usepackage{listings}
-\usepackage{color}
-%% The amssymb package provides various useful mathematical symbols
-\usepackage{amssymb}
-%% The amsthm package provides extended theorem environments
-\usepackage{amsthm} \usepackage{amsmath} \usepackage{tmadd,tmath}
-\usepackage[mathcal]{euscript} \usepackage{color}
 \usepackage{textcomp}
-\usepackage{algorithm,algorithmic}
 \definecolor{listinggray}{gray}{0.9}
@@
 
-%% colors
-\setbeamercolor{boxheadcolor}{fg=white,bg=UWRed}
-\setbeamercolor{boxbodycolor}{fg=black,bg=white}
-
-%%---------------------------------------------------------------------------%%
 \author{Stuart R. Slattery
-  \\ Engineering Physics Department
-  \\ University of Wisconsin - Madison
+\\ \href{mailto:sslattery@wisc.edu}{\texttt{sslattery@wisc.edu}}
 }
@@
 \date{\today}
-\title{Domain Decomposed Monte Carlo} 
+\title{Monte Carlo Solutions for the SPn Equations}
 \begin{document}
 \maketitle
+\section{Introduction}
+\section{Derivation of the SPn Equations}
+\section{Spectral Analysis of the SPn Equations}
+\section{Monte Carlo Solutions for the SPn Equations}
+\section{Conclusion}
 
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Monte Carlo Linear Solver Preliminaries}
-
-  \begin{itemize}
-  \item Split the linear operator
-  \end{itemize}
-
-  \[
-  \ve{H} = \ve{I} - \ve{A}
-  \]
-
-  \[
-  \ve{x} = \ve{H} \ve{x} + \ve{b}
-  \]
-
-  \medskip
-  \begin{itemize}
-  \item Generate the \textit{Neumann series}
-  \end{itemize}
-
-  \[
-  \ve{A}^{-1} = (\ve{I}-\ve{H})^{-1} = \sum_{k=0}^{\infty} \ve{H}^k
-  \]
-
-  \medskip
-  \begin{itemize}
-  \item Require $\rho(\ve{H}) < 1$ for convergence
-  \end{itemize}
-
-  \[
-  \ve{A}^{-1}\ve{b} = \sum_{k=0}^{\infty} \ve{H}^k\ve{b} = \ve{x}
-  \]
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Monte Carlo Linear Solver Preliminaries}
-
-  \begin{itemize}
-  \item Expand the Neumann series
-  \end{itemize}
-
-  \[
-  x_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-  \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}
-  \]
-
-  \medskip
-  \begin{itemize}
-  \item Define a sequence of state transitions
-  \end{itemize}
-
-  \[
-  \nu = i \rightarrow i_1 \rightarrow \cdots \rightarrow i_{k-1}
-  \rightarrow i_{k}
-  \]
-
-  \medskip
-  \begin{itemize}
-  \item Define the \textit{Neumann-Ulam decomposition}\footnote{The
-    Hadamard product $\ve{A} = \ve{B} \circ \ve{C}$ is defined
-    element-wise as $a_{ij} = b_{ij} c_{ij}$.}
-  \end{itemize}
-
-  \[
-  \ve{H} = \ve{P} \circ \ve{W}
-  \]
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Adjoint Method: Evolution of a Solution}
-
-  \begin{figure}[h!]
-    \begin{center}
-      \includegraphics[width=4in]{../../prelim/presentation/adjoint_1.png}
-    \end{center}
-    \caption{\textbf{Adjoint solution to Poisson Equation.}
-      \textit{\sn{1}{0} total histories, 0.286 seconds CPU time.} }
-  \end{figure}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Adjoint Method: Evolution of a Solution}
-
-  \begin{figure}[h!]
-    \begin{center}
-      \includegraphics[width=4in]{../../prelim/presentation/adjoint_10.png}
-    \end{center}
-    \caption{\textbf{Adjoint solution to Poisson Equation.}
-      \textit{\sn{1}{1} total histories, 0.278 seconds CPU time.} }
-  \end{figure}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Adjoint Method: Evolution of a Solution}
-
-  \begin{figure}[h!]
-    \begin{center}
-      \includegraphics[width=4in]{../../prelim/presentation/adjoint_100.png}
-    \end{center}
-    \caption{\textbf{Adjoint solution to Poisson Equation.}
-      \textit{\sn{1}{2} total histories, 0.275 seconds CPU time.} }
-  \end{figure}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Adjoint Method: Evolution of a Solution}
-
-  \begin{figure}[h!]
-    \begin{center}
-      \includegraphics[width=4in]{../../prelim/presentation/adjoint_1000.png}
-    \end{center}
-    \caption{\textbf{Adjoint solution to Poisson Equation.}
-      \textit{\sn{1}{3} total histories, 0.291 seconds CPU time.} }
-  \end{figure}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Adjoint Method: Evolution of a Solution}
-
-  \begin{figure}[h!]
-    \begin{center}
-      \includegraphics[width=4in]{../../prelim/presentation/adjoint_10000.png}
-    \end{center}
-    \caption{\textbf{Adjoint solution to Poisson Equation.}
-      \textit{\sn{1}{4} total histories, 0.428 seconds CPU time.} }
-  \end{figure}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Adjoint Method: Evolution of a Solution}
-
-  \begin{figure}[h!]
-    \begin{center}
-      \includegraphics[width=4in]{../../prelim/presentation/adjoint_100000.png}
-    \end{center}
-    \caption{\textbf{Adjoint solution to Poisson Equation.}
-      \textit{\sn{1}{5} total histories, 1.76 seconds CPU time.} }
-  \end{figure}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Adjoint Method: Evolution of a Solution}
-
-  \begin{figure}[h!]
-    \begin{center}
-      \includegraphics[width=4in]{../../prelim/presentation/adjoint_1000000.png}
-    \end{center}
-    \caption{\textbf{Adjoint solution to Poisson Equation.}
-      \textit{\sn{1}{6} total histories, 15.1 seconds CPU time.} }
-  \end{figure}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Adjoint Method: Evolution of a Solution}
-
-  \begin{figure}[h!]
-    \begin{center}
-      \includegraphics[width=4in]{../../prelim/presentation/adjoint_10000000.png}
-    \end{center}
-    \caption{\textbf{Adjoint solution to Poisson Equation.}
-      \textit{\sn{1}{7} total histories, 149 seconds CPU time.} }
-  \end{figure}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Parallelization of Monte Carlo Methods}
-
-  \begin{itemize}
-  \item No literature observed for parallel Neumann-Ulam solvers
-    \medskip \medskip
-  \item Numerous references for modern parallel Monte Carlo methods in
-    reactor physics
-    \medskip \medskip
-  \item Build a strategy for applying modern methods to the
-    Neumann-Ulam method
-  \end{itemize}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Domain Decomposition}
-
-  \begin{itemize}
-  \item Each parallel process owns a piece of the domain
-  \item Random walks must be transported across domains through
-    communication
-  \end{itemize}
-
-  \medskip \medskip
-  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
-    {Brunner's Work (2006 and 2009)}
-
-    \begin{itemize}
-    \item Looked at 4 communication patterns:
-      \begin{itemize}
-      \item Fully-locking synchronous
-      \item Asynchronous-send/synchronous-receive
-      \item Master/slave
-      \item Binary tree master/slave
-      \end{itemize}
-    \item Binary tree master/slave performed best but race conditions observed
-    \item A later improvement to their work showed a fully asynchronous
-      pattern to work best
-    \item Poor scaling for unbalanced problems for all schemes
-    \end{itemize}
-
-  \end{beamerboxesrounded}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Domain Decomposition}
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Fully Synchronous Algorithm}
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Fully Asynchronous Algorithm}
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Brunner's Communication Parameters}
-
-  \begin{figure}[h!]
-    \begin{center}
-      \includegraphics[width=5in]{brunner_params.png}
-    \end{center}
-  \end{figure}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{My Communication Parameters}
-
-  \begin{figure}[h!]
-    \begin{center}
-      \includegraphics[width=5in]{my_params.png}
-    \end{center}
-  \end{figure}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Brunner's Strong Scaling}
-
-  \begin{figure}[h!]
-    \begin{center}
-      \includegraphics[width=4in]{brunner_strong.png}
-    \end{center}
-  \end{figure}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{My Strong Scaling}
-
-  \begin{figure}[h!]
-    \begin{center}
-      \includegraphics[width=4in]{my_strong.png}
-    \end{center}
-  \end{figure}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Brunner's Weak Scaling}
-
-  \begin{figure}[h!]
-    \begin{center}
-      \includegraphics[width=4in]{brunner_weak.png}
-    \end{center}
-  \end{figure}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{My Weak Scaling}
-
-  \begin{figure}[h!]
-    \begin{center}
-      \includegraphics[width=4in]{my_weak.png}
-    \end{center}
-  \end{figure}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-
+\pagebreak
+\bibliographystyle{ieeetr}
+\bibliography{paper}
 \end{document}

### 2013-02-27 doc/spn/spn.tex
adding to the spn doc

@@
 \date{\today}
-\title{Monte Carlo Solutions for the SPn Equations}
+\title{Monte Carlo Solutions for the $SP_N$ Equations}
 \begin{document}
 \maketitle
+
+%%---------------------------------------------------------------------------%%
 \section{Introduction}
-\section{Derivation of the SPn Equations}
-\section{Spectral Analysis of the SPn Equations}
-\section{Monte Carlo Solutions for the SPn Equations}
+\label{sec:introduction}
+
+%%---------------------------------------------------------------------------%%
+\section{Derivation of the $SP_N$ Equations}
+\label{sec:derivation}
+In this section, we derive the $SP_N$ equations. We begin by stating the
+general time-independent neutron transport equation followed by a
+derivation of the $P_N$ equations in planar geometry for multiple energy
+groups. From these equations, we then apply a set of approximations to
+yield the multi-dimensional, multi-group $SP_N$ equations.
+
+\subsection{The Neutron Transport Equation}
+\label{subsec:transport_eq}
+As a starting point we define the time-independent neutron transport
+equation \cite{lewis_computational_1993}:
+\begin{equation}
+  \hat{\Omega} \cdot \vec{\nabla} \psi(\vec{r},\hat{\Omega},E) +
+  \sigma(\vec{r},E) \psi(\vec{r},\hat{\Omega},E) = \int dE' \int
+  d\Omega' \sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}' \rightarrow
+  \hat{\Omega}) \psi(\vec{r},\hat{\Omega}',E') +
+  q(\vec{r},\hat{\Omega},E)\:,
+  \label{eq:general_transport}
+\end{equation}
+with the variables defined as:
+\begin{itemize}
+\item $\vec{r}$ - neutron spatial position
+\item $\hat{\Omega}$ - neutron streaming direction with radial
+  component $\mu$ and azimuthal component $\omega$
+\item $E$ - neutron energy
+\item $\psi(\vec{r},\hat{\Omega},E)$ - angular flux
+\item $\sigma(\vec{r},E)$ - total interaction cross section
+\item $\sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}'$ - probability
+  of scattering from direction $\hat{\Omega}'$ into an angular domain
+  $d\hat{\Omega}'$ about the direction $\hat{\Omega}$ and from energy
+  $E'$ to an energy domain $dE'$ about energy $E$
+\item $q(\vec{r},\hat{\Omega},E)$ - external source of neutrons.
+\end{itemize}
+For this work, it is sufficient to formulate
+Eq~(\ref{eq:general_transport}) in 1-dimensonal Cartesian geometry:
+\begin{equation}
+  \mu \frac{\partial}{\partial x} \psi(x,\mu,E) + \sigma(x,E)
+  \psi(x,\mu,E) = \int dE' \int d\Omega' \sigma_s(x,E' \rightarrow
+  E,\hat{\Omega}' \rightarrow \hat{\Omega})
+  \psi(\vec{r},\hat{\Omega}',E') + \frac{q(x,E)}{4 \pi}\:,
+  \label{eq:general_transport}
+\end{equation}
+where the angular component of the solution is no longer dependent on
+the azimuthal direction of travel and an isotropic source of neutrons
+is assumed.
+
+\subsection{Derivation of the $P_N$ Equations}
+\label{subsec:pn_equations}
+Next, we derive the $P_N$ equations, a simplified form of the general
+transport equation where Legendre polynomials are used to expand the
+angular flux and scattering cross section variables as a means of
+capturing the angular structure of the solution. Before deriving this
+form of the transport equation, we briefly discuss a few properties of
+Legendre polynomials that we will find useful in the derivation.
+
+\subsubsection{Legendre Polynomials}
+\label{subsubsec:legendre_polys}
+
+\subsubsection{Planar $P_N$ Equations}
+\label{subsubsec:planar_pn}
+
+\subsubsection{Boundary Conditions for the $P_N$ Equations}
+
+%%---------------------------------------------------------------------------%%
+\section{Spectral Analysis of the $SP_N$ Equations}
+\label{sec:spectral_analysis}
+
+%%---------------------------------------------------------------------------%%
+\section{Monte Carlo Solutions for the $SP_N$ Equations}
+\label{sec:monte_carlo}
+
+%%---------------------------------------------------------------------------%%
 \section{Conclusion}
+\label{sec:conclusion}
 
+%%---------------------------------------------------------------------------%%
 \pagebreak
 \bibliographystyle{ieeetr}
-\bibliography{paper}
+\bibliography{references}
 \end{document}

### 2013-02-27 doc/spn/spn.tex
adding to the spn doc
working on Pn equation derivation in doc

@@
 \label{sec:derivation}
-In this section, we derive the $SP_N$ equations. We begin by stating the
-general time-independent neutron transport equation followed by a
-derivation of the $P_N$ equations in planar geometry for multiple energy
-groups. From these equations, we then apply a set of approximations to
-yield the multi-dimensional, multi-group $SP_N$ equations.
+In this section, we derive the $SP_N$ equations. We begin by stating
+the general time-independent neutron transport equation followed by a
+derivation of the $P_N$ equations in planar geometry for multiple
+energy groups. From these equations, we then apply a set of
+approximations to yield the multi-dimensional, multi-group $SP_N$
+equations for fixed source problemsd
 
@@
   \psi(\vec{r},\hat{\Omega}',E') + \frac{q(x,E)}{4 \pi}\:,
-  \label{eq:general_transport}
+  \label{eq:cart_1d_transport}
 \end{equation}
@@
 \label{subsubsec:legendre_polys}
+The Legendre polynomials are an orthogonal set of functions that
+are solutions to Legendre's differential equation. They have the
+following form \cite{lewis_computational_1993}:
+\begin{equation}
+  P_l(\mu) = \frac{1}{2^l l!}\frac{d^l}{d \mu^l}(\mu^2-1)^l\:.
+  \label{eq:general_legendre_poly}
+\end{equation}
+These functions have several nice properties including orthogonality:
+\begin{equation}
+  \int_{-1}^{1} P_l(\mu) P_{l'}(\mu) d\mu = \frac{2}{2l+1}\delta_{l l'}\:,
+  \label{eq:legendre_orthog}
+\end{equation}
+a recurrence relation:
+\begin{equation}
+  \mu P_l(\mu) = \frac{1}{2l+1}[(l+1)P_{l+1}(\mu) + l P_{l-1}(\mu)]\:,
+  \label{eq:legendre_recurrence}
+\end{equation}
+and an addition theorem:
+\begin{equation}
+  P_l(\hat{\Omega} \cdot \hat{\Omega}') = \frac{1}{2n+1}\sum_{m=-l}^l
+  Y_{lm}(\hat{\Omega})Y^*_{lm}(\hat{\Omega}')\:.
+  \label{eq:legendre_addition}
+\end{equation}
+where the functions $Y_{lm}(\hat{\Omega})$ are the spherical
+harmonics. We can form the addition theorem in this way because the
+spherical harmonics are in fact just harmonic multiples of the
+Legendre polynomials:
+\begin{equation}
+  Y_{lm}(\hat{\Omega}) =
+  \sqrt{\frac{(2l+1)(l-m)!}{(l+m)!}}P^m_l(\mu)e^{i \omega m}\:,
+  \label{eq:spherical_harmonic}
+\end{equation}
+where $\omega$ is the azimuthal component of the streaming
+direction. We can reduce Eq~(\ref{eq:legendre_addition}) for the
+planar geometry we are studying by ignoring the azimuthal components
+of the addition theorem. As shown in Eq~(\ref{eq:spherical_harmonic}),
+the azimuthal dependence is given by the harmonic component, $e^{i
+  \omega m}$, and therefore we choose to ignore all terms in
+Eq.~(\ref{eq:legendre_addition}) where $m \neq 0$. This gives:
+\begin{equation}
+  P_l(\hat{\Omega} \cdot \hat{\Omega}') = \frac{1}{2n+1}
+  Y_{l0}(\hat{\Omega})Y^*_{l0}(\hat{\Omega}')\:.
+  \label{eq:legendre_addition_2}
+\end{equation}
+Per Eq~(\ref{eq:spherical_harmonic}) we have:
+\begin{equation}
+  Y_{l0}(\hat{\Omega}) = \sqrt{2l+1}P_l^0(\mu)\:,
+  \label{eq:harmonic_0}
+\end{equation}
+where $P_l^0(\mu) = P_l(\mu)$ is the $0^{th}$ associated Legendre
+function. Finally, we can reduce the addition theorem from
+Eq~(\ref{eq:legendre_addition_2}) with Eq~(\ref{eq:harmonic_0}) to a
+simple product for planar geometry:
+\begin{equation}
+  P_l(\hat{\Omega} \cdot \hat{\Omega}') = P_l(\mu)P_l(\mu')\:.
+  \label{eq:legendre_addition_3}
+\end{equation}
 
@@
 \label{subsubsec:planar_pn}
+With the Legendre polynomial properties defined above, we can proceed
+by deriving the $P_N$ equations for planar geometry. For these
+equations, we will start by assuming a monoenergtic field of neutrons
+such that we are solving the following reduced form of the transport
+equation: 
+\begin{equation}
+  \mu \frac{\partial}{\partial x} \psi(x,\mu) + \sigma(x) \psi(x,\mu)
+  = \int d\Omega' \sigma_s(x,\hat{\Omega}' \rightarrow \hat{\Omega})
+  \psi(\vec{r},\hat{\Omega}') + \frac{q(x)}{4 \pi}\:.
+  \label{eq:mono_transport}
+\end{equation}
+The $P_N$ equations introduce the approximation that the angular
+dependence of the scattering cross section, $sigma_s$, and the angular
+flux $\psi$, can be discretized by expanding them in Legendre
+polynomials as follows:
+\begin{equation}
+  \psi(x,\mu) = \sum_{n=0}^\infty (2n+1)P_n(\mu)\phi_n(x)\:,
+  \label{eq:flux_expansion}
+\end{equation}
+\begin{equation}
+  \sigma_{sm}(x) = \sum_{m=0}^\infty (2m+1)P_m(\mu)\sigma_s(x)\:,
+  \label{eq:scattering_expansion}
+\end{equation}
+where we have supressed the $2\pi$ generated by integrating away the
+azimuthal angular component and $\phi_n(x)$ in
+Eq~(\ref{eq:flux_expansion}) is referred to as the $n^{th}$ Legendre
+moment of the neutron flux and is given by:
+\begin{equation}
+  \phi_n(x) = \int_{-1}^1 P_n(\mu)\psi(x,\mu)d\mu\:.
+  \label{eq:legendre_moments}
+\end{equation}
 
 \subsubsection{Boundary Conditions for the $P_N$ Equations}
+\label{subsbusec:bcs_pn}
+
+\subsection{Formulation of the  $SP_N$ Equations}
+\label{subsec:spn_equations}
 

### 2013-02-27 doc/spn/spn.tex
finished Pn derivation, boundary conditions are next

@@
   component $\mu$ and azimuthal component $\omega$
+\item $\hat{\Omega}' \cdot \hat{\Omega} = \mu_0$ is the angle of
+  scattering
 \item $E$ - neutron energy
@@
 \begin{equation}
-  \int_{-1}^{1} P_l(\mu) P_{l'}(\mu) d\mu = \frac{2}{2l+1}\delta_{l l'}\:,
+  \int_{-1}^{1} P_l(\mu) P_{l'}(\mu) d\mu = \frac{1}{2l+1}\delta_{l l'}\:,
   \label{eq:legendre_orthog}
@@
 \end{equation}
+We first insert the expansions given by Eq~(\ref{eq:flux_expansion})
+and (\ref{eq:scattering_expansion}) into the planar transport equation
+given by Eq~(\ref{eq:mono_transport}):
+\begin{equation}
+  \frac{\partial}{\partial x}\Big[\sum_{n=0}^\infty (2n+1) \phi_n \mu
+    P_n(\mu) \Big] + \sigma \sum_{n=0}^\infty (2n+1) \phi_n P_n(\mu) =
+  \int_{-1}^1 \sum_{m=0}^\infty (2m+1) \sigma_{sm} P_m(\mu_0)
+  \sum_{n=0}^\infty (2n+1) \phi_n P_n(\mu') d \mu' + q\:,
+  \label{eq:pn_deriv_1}
+\end{equation}
+where the dependence on the spatial variable, $x$, has been
+supressed. To arrive at the $P_N$ equations we multiply
+Eq~(\ref{eq:pn_deriv_1}) by $P_m(\mu)$ and integrate over the angular
+domain $\int_{-1}^1 d \mu$. We will look at each term in
+Eq~(\ref{eq:pn_deriv_1}) individually.
+
+\paragraph{Streaming Term}
+We first apply the multiplication and integration as prescribed above:
+\begin{equation}
+  \frac{\partial}{\partial x}\Bigg[\sum_{n=0}^\infty (2n+1) \mu \phi_n
+    P_n(\mu) \Bigg] \rightarrow \int_{-1}^1 \frac{\partial}{\partial
+    x}\Bigg[\sum_{n=0}^\infty (2n+1) \phi_n \mu P_n(\mu) P_m(\mu) \Bigg]
+  d \mu\:.
+  \label{eq:pn_deriv_2}
+\end{equation}
+The $\mu P_n(\mu)$ term can be eliminated via the recurrence relation
+given by Eq~(\ref{eq:legendre_recurrence}):
+\begin{equation}
+\int_{-1}^1 \frac{\partial}{\partial x}\Bigg[\sum_{n=0}^\infty (2n+1)
+  \frac{\phi_n}{2n+1}[(n+1)P_{n+1}(\mu) + n P_{n-1}(\mu)] P_m(\mu)
+  \Bigg] d\mu\:,
+\label{eq:pn_deriv_3}
+\end{equation}
+which expands to:
+\begin{equation}
+  \sum_{n=0}^\infty \frac{\partial}{\partial
+    x}\phi_n\Bigg[(n+1)\int_{-1}^1 P_{n+1}(\mu)P_m(\mu) d\mu + n
+    \int_{-1}^1 P_{n-1}(\mu) P_m(\mu) d\mu \Bigg] \:.
+  \label{eq:pn_deriv_4}
+\end{equation}
+This reveals the orthogonality relation given by
+Eq~(\ref{eq:legendre_orthog}) that when inserted into
+Eq~(\ref{eq:pn_deriv_3}):
+\begin{equation}
+  \sum_{n=0}^\infty \frac{\partial}{\partial
+    x}\phi_n\Bigg[(n+1)\frac{1}{2n+1}\delta_{n,n+1} +
+    n\frac{1}{2n+1}\delta_{n,n-1} \Bigg] \:.
+  \label{eq:pn_deriv_5}
+\end{equation}
+We can then distribute the Legendre moment to arrive at the final form
+of the streaming term:
+\begin{equation}
+  \sum_{n=0}^\infty \frac{\partial}{\partial x} \frac{1}{n+1} \Big[
+    (n+1) \phi_{n+1} + n \phi_{n-1} \Big] \:.
+  \label{eq:pn_deriv_6}
+\end{equation}
+
+\paragraph{Removal Term}
+To reduce the removal term, the orthogonality relation is again
+applied after the integration:
+\begin{equation}
+  \sigma \sum_{n=0}^\infty (2n+1) \phi_n P_n(\mu) \rightarrow
+  \int_{-1}^1 \sigma \sum_{n=0}^\infty (2n+1) \phi_n P_n(\mu) P_m(\mu)
+  d\mu\:,
+  \label{eq:pn_deriv_7}
+\end{equation}
+\begin{equation}
+  \sigma \sum_{n=0}^\infty (2n+1) \phi_n \frac{1}{2n+1}\:,
+  \label{eq:pn_deriv_8}
+\end{equation}
+giving for the final removal term:
+\begin{equation}
+  \sum_{n=0}^\infty \sigma \phi_n \:.
+  \label{eq:pn_deriv_9}
+\end{equation}
+
+\paragraph{Scattering Term}
+For the scattering term:
+\begin{equation}
+  \int_{-1}^1 \sum_{m=0}^\infty (2m+1) \sigma_{sm} P_m(\mu_0)
+  \sum_{n=0}^\infty (2n+1) \phi_n P_n(\mu') d\mu' \rightarrow
+  \int_{-1}^1 \int_{-1}^1 \sum_{m=0}^\infty (2m+1) \sigma_{sm} P_m(\mu)
+  P_m(\mu_0) \sum_{n=0}^\infty (2n+1) \phi_n P_n(\mu') d\mu' d\mu\:.
+  \label{eq:pn_deriv_10}
+\end{equation}
+The addition thereom from Eq~(\ref{eq:legendre_addition_3}) is applied
+to give:
+\begin{equation}
+  \int_{-1}^1 \int_{-1}^1 \sum_{m=0}^\infty (2m+1) \sigma_{sm} P_m(\mu)
+  P_m(\mu)P_m(\mu') \sum_{n=0}^\infty (2n+1) \phi_n P_n(\mu') d\mu' d\mu\:,
+  \label{eq:pn_deriv_11}
+\end{equation}
+which can be rearranged as:
+\begin{equation}
+  \sum_{m=0}^\infty \sum_{n=0}^\infty (2m+1) \sigma_{sm} (2n+1) \phi_n
+  \int_{-1}^1 P_m(\mu') P_n(\mu') d\mu' \int_{-1}^1 P_m(\mu) P_m(\mu)
+  d\mu\:.
+  \label{eq:pn_deriv_12}
+\end{equation}
+Again we can apply orthogonality to eliminate the Legendre polynomials:
+\begin{equation}
+  \sum_{m=0}^\infty \sum_{n=0}^\infty (2m+1) \sigma_{sm} (2n+1) \phi_n
+  \frac{1}{2n+1}\delta_{nm}\frac{1}{2m+1}\delta_{mm}\:,
+  \label{eq:pn_deriv_13}
+\end{equation}
+which is reduced to:
+\begin{equation}
+  \sum_{n=0}^\infty \sigma_{sn} \phi_n\:,
+  \label{eq:pn_deriv_14}
+\end{equation}
+with the dependence on the index $m$ eliminated.
+
+\paragraph{Source Term}
+The laste term we are concerned with in Eq~(\ref{eq:pn_deriv_1}) is
+the source term:
+\begin{equation}
+  q \rightarrow \int_{-1}^1 q P_n(\mu) d\mu \:.
+  \label{eq:pn_deriv_15}
+\end{equation}
+We can leverage orthogonality by multiplying by $P_0(\mu) = 1$:
+\begin{equation}
+  \int_{-1}^1 q P_0 P_n(\mu) d\mu = \frac{q}{2*0+1}\delta_{n0}\:,
+  \label{eq:pn_deriv_16}
+\end{equation}
+giving a final source term of:
+\begin{equation}
+  q\delta_{n0}\:.
+  \label{eq:pn_deriv_17}
+\end{equation}
+
+Now that we have expanded all angular dependent terms in
+Eq~(\ref{eq:pn_deriv_1}) and reduced them appropriately, we can
+combine them to generate the $P_N$ equations:
+\begin{equation}
+    \sum_{n=0}^\infty \frac{\partial}{\partial x} \frac{1}{n+1} \Big[
+      (n+1) \phi_{n+1} + n \phi_{n-1} \Big] + \sum_{n=0}^\infty \sigma
+    \phi_n = \sum_{n=0}^\infty \sigma_{sn} \phi_n + q\delta_{n0}\:.
+  \label{eq:pn_deriv_18}
+\end{equation}
+More formally, the $P_N$ equations are written as:
+\begin{equation}
+   \frac{1}{n+1} \frac{\partial}{\partial x}\Big[ (n+1) \phi_{n+1} + n
+     \phi_{n-1} \Big] + \Sigma_n \phi_n = q\delta_{n0} \:,
+  \label{eq:final_pn_equations}
+\end{equation}
+where $\Sigma_n = \sigma-\sigma_{sn}$ and the summations are truncated
+at some level of approximation such that $n = 0,1,\dotsc,N$. This
+yields a set of $N+1$ equations for $N+2$ flux moments. We therefore
+require an additional equation to close the system. In accordance with
+the series truncation as an approximation we choose:
+\begin{equation}
+  \phi_{N+1} = 0\:.
+  \label{eq:pn_closure}
+\end{equation}
+As an example, we will construct the P5 equations from
+Eq~(\ref{eq:final_pn_equations}) and the closure given by
+Eq~(\ref{eq:pn_closure}):
+\begin{subequations}
+  \begin{gather}
+   \frac{\partial}{\partial x}\phi_{1} + \Sigma_0 \phi_0 = q\:,\\ 
+   \frac{1}{2} \frac{\partial}{\partial x}\Big[ 2
+     \phi_{2} + \phi_{0} \Big] + \Sigma_1 \phi_1 = 0\:,\\
+   \frac{1}{3} \frac{\partial}{\partial x}\Big[ 3 \phi_{3} + 2
+     \phi_{1} \Big] + \Sigma_2 \phi_2 = 0 \:,\\
+   \frac{1}{4} \frac{\partial}{\partial x}\Big[ 4 \phi_{4} + 3
+     \phi_{2} \Big] + \Sigma_3 \phi_3 = 0 \:,\\
+   \frac{1}{5} \frac{\partial}{\partial x}\Big[ 5 \phi_{5} + 4
+     \phi_{3} \Big] + \Sigma_4 \phi_4 = 0 \:,\\
+   \frac{1}{n+1} \frac{\partial}{\partial x} 5 \phi_{4} + \Sigma_5
+   \phi_5 = 0 \:.
+  \end{gather}
+  \label{eq:p5_equations}
+\end{subequations}
+This gives us a set of 6 coupled equations for the six Legendre
+moments requested defined over the entire spatial domain for a single
+energy group.
 
@@
 \pagebreak
-\bibliographystyle{ieeetr}
-\bibliography{references}
+\bibliographystyle{ieeetr} \bibliography{references}
 \end{document}

### 2013-02-27 doc/spn/spn.tex
added boundary conditions to pn equation derivation

@@
 approximations to yield the multi-dimensional, multi-group $SP_N$
-equations for fixed source problemsd
+equations for fixed source problems.
 
@@
   P_l(\hat{\Omega} \cdot \hat{\Omega}') = \frac{1}{2n+1}\sum_{m=-l}^l
-  Y_{lm}(\hat{\Omega})Y^*_{lm}(\hat{\Omega}')\:.
+  Y_{lm}(\hat{\Omega})Y^*_{lm}(\hat{\Omega}')\:,
   \label{eq:legendre_addition}
@@
 \begin{equation}
-  \sum_{n=0}^\infty \frac{\partial}{\partial x} \frac{1}{n+1} \Big[
+  \sum_{n=0}^\infty \frac{\partial}{\partial x} \frac{1}{2n+1} \Big[
     (n+1) \phi_{n+1} + n \phi_{n-1} \Big] \:.
@@
 For the scattering term:
-\begin{equation}
+\begin{multline}
   \int_{-1}^1 \sum_{m=0}^\infty (2m+1) \sigma_{sm} P_m(\mu_0)
-  \sum_{n=0}^\infty (2n+1) \phi_n P_n(\mu') d\mu' \rightarrow
+  \sum_{n=0}^\infty (2n+1) \phi_n P_n(\mu') d\mu' \rightarrow\\
   \int_{-1}^1 \int_{-1}^1 \sum_{m=0}^\infty (2m+1) \sigma_{sm} P_m(\mu)
@@
   \label{eq:pn_deriv_10}
-\end{equation}
+\end{multline}
 The addition thereom from Eq~(\ref{eq:legendre_addition_3}) is applied
@@
 \paragraph{Source Term}
-The laste term we are concerned with in Eq~(\ref{eq:pn_deriv_1}) is
-the source term:
+The last term we are concerned with in Eq~(\ref{eq:pn_deriv_1}) is the
+source term:
 \begin{equation}
@@
 \begin{equation}
-    \sum_{n=0}^\infty \frac{\partial}{\partial x} \frac{1}{n+1} \Big[
+    \sum_{n=0}^\infty \frac{\partial}{\partial x} \frac{1}{2n+1} \Big[
       (n+1) \phi_{n+1} + n \phi_{n-1} \Big] + \sum_{n=0}^\infty \sigma
@@
 \begin{equation}
-   \frac{1}{n+1} \frac{\partial}{\partial x}\Big[ (n+1) \phi_{n+1} + n
+   \frac{1}{2n+1} \frac{\partial}{\partial x}\Big[ (n+1) \phi_{n+1} + n
      \phi_{n-1} \Big] + \Sigma_n \phi_n = q\delta_{n0} \:,
@@
 where $\Sigma_n = \sigma-\sigma_{sn}$ and the summations are truncated
-at some level of approximation such that $n = 0,1,\dotsc,N$. This
+at some level of approximation $N$ such that $n = 0,1,\dotsc,N$. This
 yields a set of $N+1$ equations for $N+2$ flux moments. We therefore
 require an additional equation to close the system. In accordance with
-the series truncation as an approximation we choose:
+the series truncation as an approximation we choose the last moment in
+the expansion to be zero:
 \begin{equation}
@@
    \frac{\partial}{\partial x}\phi_{1} + \Sigma_0 \phi_0 = q\:,\\ 
-   \frac{1}{2} \frac{\partial}{\partial x}\Big[ 2
+   \frac{1}{3} \frac{\partial}{\partial x}\Big[ 2
      \phi_{2} + \phi_{0} \Big] + \Sigma_1 \phi_1 = 0\:,\\
-   \frac{1}{3} \frac{\partial}{\partial x}\Big[ 3 \phi_{3} + 2
+   \frac{1}{5} \frac{\partial}{\partial x}\Big[ 3 \phi_{3} + 2
      \phi_{1} \Big] + \Sigma_2 \phi_2 = 0 \:,\\
-   \frac{1}{4} \frac{\partial}{\partial x}\Big[ 4 \phi_{4} + 3
+   \frac{1}{7} \frac{\partial}{\partial x}\Big[ 4 \phi_{4} + 3
      \phi_{2} \Big] + \Sigma_3 \phi_3 = 0 \:,\\
-   \frac{1}{5} \frac{\partial}{\partial x}\Big[ 5 \phi_{5} + 4
+   \frac{1}{9} \frac{\partial}{\partial x}\Big[ 5 \phi_{5} + 4
      \phi_{3} \Big] + \Sigma_4 \phi_4 = 0 \:,\\
-   \frac{1}{n+1} \frac{\partial}{\partial x} 5 \phi_{4} + \Sigma_5
+   \frac{1}{11} \frac{\partial}{\partial x} 5 \phi_{4} + \Sigma_5
    \phi_5 = 0 \:.
@@
 moments requested defined over the entire spatial domain for a single
-energy group.
+energy group. In practice, only odd-numbered $P_N$ orders are
+generally used \cite{lewis_computational_1993}. This is due to the
+fact that using odd $N$ yields an even number of $N+1$ equations which
+can be split evenly on the left and right boundaries of the problem to
+facilitate the description of boundary conditions. We will choose this
+convention when deriving the boundary conditions.
 
@@
 \label{subsbusec:bcs_pn}
+Per the analysis in \cite{lewis_computational_1993}, two types of
+boundary conditions will be discussed: reflecting and Marshak. Marshak
+boundary conditions can be used to specify both vacuum conditions and
+isotropic source conditions on the boundary.
+
+\paragraph{Reflecting Boundary Conditions}
+In this case, the incoming flux should be equivalent to the outgoing
+flux at the boundary point $x_b$:
+\begin{equation}
+  \psi(x_b,\mu) = \psi(x_b,-\mu)\:.
+  \label{eq:reflecting_condition}
+\end{equation}
+Given the legendre expansion for the flux defined in
+Eq~(\ref{eq:flux_expansion}) and the Legendre polynomial property that
+$P_n(\mu) = (-1)^n P_n(-\mu)$, the condition specified by
+Eq~(\ref{eq:reflecting_condition}) can be satisfied if
+\begin{equation}
+  \phi_n = 0,\ \forall \ \text{odd}\ n\:,
+  \label{eq:reflecting_condition_odd}
+\end{equation}
+as all even $n$ yield $P_n(\mu) = P_n(-\mu)$ and therefore an
+equivalent reflecting condition for the flux moments.
+
+\paragraph{Marshak Boundary Conditions}
+The Marshak conditions come directly from the Legendre moments of the
+flux:
+\begin{equation}
+  \int_{\mu_b} P_i(\mu) \psi(\mu) d\mu = \int_{\mu_b} P_i(\mu)
+  \psi_b(\mu) d\mu\ \ \ \ \ \ \ \ \ \ \text{for $i=1,3,...,N$}\:,
+  \label{eq:general_marshak}
+\end{equation}
+where $\psi_b(\mu)$ is the prescribed angular flux on the boundary of
+interest and $\mu_b$ the angular domain defined by the boundary. To
+discretize this condition, we again insert the angular flux expansions
+from Eq~(\ref{eq:flux_expansion}) for the fluxes defined in the
+domain:
+\begin{equation}
+  \int_{\mu_b} P_i(\mu) \sum_{n=0}^N (2n+1) \phi_n P_n(\mu) d\mu =
+  \int_{\mu_b} P_i(\mu) \psi_b(\mu) d\mu\:.
+  \label{eq:marshak_expanded}
+\end{equation}
+The boundary flux, $\phi_b(\mu)$ is assumed to known and therefore
+Eq~(\ref{eq:marshak_expanded}) defines a set of $(N+1)/2$ equations to
+be solved on each boundary in the planar case, closing the system in
+the spatial domain.
+
+As an example of applying the Marshak conditions, consider the $P_3$
+case with an isotropic boundary source $\phi_b$ on the left side of
+the domain. In this case, the angular domain over which the boundary
+flux is defined will be $\mu_b \in [0,1]$, giving the bounds of
+integration. We first expand the summation for $i=1$:
+\begin{equation}
+  \int_0^1 \mu \Bigg[ \phi_0 + 3\phi_1\mu +
+    \frac{5}{2}\phi_2(3\mu^2-1) + \frac{7}{2}\phi_3(5\mu^3-3\mu)
+    \Bigg] d\mu = \int_0^1 \mu \phi_b d\mu\:,
+  \label{eq:marshak_p1_deriv_1}
+\end{equation}
+and then $i=3$:
+\begin{equation}
+  \int_0^1 \frac{1}{2}(5\mu^3-3\mu) \Bigg[ \phi_0 + 3\phi_1\mu +
+    \frac{5}{2}\phi_2(3\mu^2-1) + \frac{7}{2}\phi_3(5\mu^3-3\mu)
+    \Bigg] d\mu = \int_0^1 \frac{1}{2}(5\mu^3-3\mu) \phi_b d\mu\:.
+  \label{eq:marshak_p3_deriv_1}
+\end{equation}
+Expanding the polynomials in $\mu$ and carrying out the simple
+integration then gives 2 equations for the left hand boundary:
+\begin{equation}
+  \phi_0 + 2\phi_1 + \frac{5}{4}\phi_2 = \phi_b\:,
+  \label{eq:marshak_p1_deriv_2}
+\end{equation}
+\begin{equation}
+  \phi_0 - 5\phi_1 - 8\phi_2 = \phi_b\:.
+  \label{eq:marshak_p1_deriv_3}
+\end{equation}
+The right hand side boundary condition will yield 2 complementary
+equations if Marshak conditions are used or 2 non-zero moments to be
+solved for if reflected conditions are used. The formulation above
+also holds for vacuum conditions where $\phi_b = 0$.
 
@@
 \label{subsec:spn_equations}
+
+\subsection{Formulation of the Multigroup $SP_N$ Equations}
+\label{subsec:mg_spn_equations}
 

### 2013-02-28 doc/spn/spn.tex
adding intro to spn doc

@@
 \date{\today}
-\title{Monte Carlo Solutions for the $SP_N$ Equations}
+\title{Monte Carlo Synthetic Acceleration Solutions for the $SP_N$ Equations}
 \begin{document}
@@
 \label{sec:introduction}
+The neutron tranpsort problem is complicated. Solutions cover a large
+phase space and the problems of interest are often geometry complex,
+very large, or both requiring tremendous computational resources to
+generate an adequate solution. Modern deterministic methods for large
+scale problems are commonly variants on the discrete ordinates ($S_N$)
+method \cite{evans_denovo:_2010}. For full reactor core neutronics
+simulations, the $S_N$ method requires potentially trillions of
+unknown angular flux moments to be computed to achieve good accuracy
+for the repsonses to be computed. Other forms of the transport
+problem, including the $P_N$ method, take on a simpler form than the
+more common $S_N$ methods but lack in accuracy when compared while
+still requiring considerable computational resources for solutions in
+multiple dimensions.
+
+In the 1960's, Gelbard developed an ad-hoc multidimensional extension
+of the simple single dimension planar $P_N$ equations that created a
+system of coupled, diffusion-like equations known as the $SP_N$
+equations. Up until around the 1990's, the $SP_N$ methods was either
+widely unknown, widely unused, or combination of both even though
+numerical studies showed promising results with better solutions than
+diffusion theory and a significant reduction in computational time
+over more accurate methods such as discrete ordinates. Why did this
+happen? A significant problem, pointed out by Larsen, was that little
+rigor had been applied to the formulation of the $SP_N$ equations
+since their derivation through primarily heuristic arguments. Instead,
+studies at that time focused on simply comparing the results of the
+method to other contemporary transport solution strategies. In
+addition, many problems of interest from the literature at the time
+were either solved using nodal-type methods for reactor-sized problems
+are $S_N$-type methods for benchmark problems with intricate material
+configurations and potentially large flux gradients over small spatial
+domains.
+
+So why reconsider the $SP_N$ equations? Starting in the 1990's and
+primarily due to Larsen and his colleagues, the $SP_N$ equations have
+been given a more rigorous treatment with both variational and
+asymptotic derivations performed as a means of verification. In
+addition, these equations have been more rigorously studied as
+solution methods to MOX fuel problems and have been shown to provide
+accurate solutions. With this mathematical literature to provide a
+solid numerical footing for the method, we look at its application to
+today's challenge problems in full reactor core transport. The
+reduction in numerical complexity of current full core deterministic
+solution methods using the $S_N$ approximation could mean significant
+savings in both compute time and memory required. In addition, the
+characteristics of the solution to the transport problem for a steady
+state reactor core permit diffusion theory to be used; a staple of the
+nuclear industry since its inception. Therefore, if diffusion theory
+is applicable, then finer grained solutions that capture more of the
+physics contained in the transport equation should be possible with
+the $SP_N$ method. In doing so, we also expect from the literature to
+obtain computed responses on the order of accuracy we would expect
+from an appropriately discretized $S_N$ method at a fraction of the
+cost.
+
+To further motivate moving in this direction, recent developments in
+the Exnihilo neutronics package at Oak Ridge National Laboratory have
+permitted generation of the $SP_N$ system of equations for detailed
+full core reactor models. By fully forming these equations and
+formulating them as a linear algebra problem, we now have access to
+all of the modern advancements in computational linear algebra
+including Krylov solvers for asymmetric systems and preconditioning
+methods such as algebraic multigrid. This leads us to then explore the
+applicability of our work in discrete Monte Carlo methods for linear
+systems as a possible solution method for the $SP_N$ equations. If
+formulated correctly, we hypothesis that signficant improvement in
+usage of computational resources may be observed compared to modern
+solution techniques such as those suggested.
+
+In this document we derive the multigroup $SP_N$ equations from the
+Boltzmann transport equation. We then perform a spectral analysis on
+the resulting system of equations to determine its potential
+performance with both modern Krylov methods and Monte Carlo methods
+for asymmetric systems. Following this, we devise a Monte Carlo
+synthetic acceleration method as a proposed solution scheme.
 

### 2013-02-28 doc/spn/spn.tex
updated spn document formatting

@@
 %        File: spn.tex
-\documentclass[letterpaper]{article}
-\usepackage[top=1.0in,bottom=1.0in,left=1.0in,right=1.0in]{geometry}
+\documentclass[letterpaper,12pt]{article}
+\usepackage[top=1.0in,bottom=1.0in,left=1.25in,right=1.25in]{geometry}
 \usepackage{verbatim}
@@
 \date{\today}
-\title{Monte Carlo Synthetic Acceleration Solutions for the $SP_N$ Equations}
+\title{A Monte Carlo Synthetic Acceleration Method for the $SP_N$ Equations}
 \begin{document}
@@
 \label{sec:introduction}
-The neutron tranpsort problem is complicated. Solutions cover a large
+The neutron transport problem is complicated. Solutions cover a large
 phase space and the problems of interest are often geometry complex,
@@
 were either solved using nodal-type methods for reactor-sized problems
-are $S_N$-type methods for benchmark problems with intricate material
+or $S_N$-type methods for benchmark problems with intricate material
 configurations and potentially large flux gradients over small spatial
@@
 equation \cite{lewis_computational_1993}:
-\begin{equation}
+\begin{multline}
   \hat{\Omega} \cdot \vec{\nabla} \psi(\vec{r},\hat{\Omega},E) +
-  \sigma(\vec{r},E) \psi(\vec{r},\hat{\Omega},E) = \int dE' \int
+  \sigma(\vec{r},E) \psi(\vec{r},\hat{\Omega},E) = \\ \int dE' \int
   d\Omega' \sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}' \rightarrow
@@
   \label{eq:general_transport}
-\end{equation}
+\end{multline}
 with the variables defined as:
@@
 \item $\sigma(\vec{r},E)$ - total interaction cross section
-\item $\sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}'$ - probability
+\item $\sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}')$ - probability
   of scattering from direction $\hat{\Omega}'$ into an angular domain
@@
 Eq~(\ref{eq:general_transport}) in 1-dimensonal Cartesian geometry:
-\begin{equation}
+\begin{multline}
   \mu \frac{\partial}{\partial x} \psi(x,\mu,E) + \sigma(x,E)
-  \psi(x,\mu,E) = \int dE' \int d\Omega' \sigma_s(x,E' \rightarrow
+  \psi(x,\mu,E) = \\ \int dE' \int d\Omega' \sigma_s(x,E' \rightarrow
   E,\hat{\Omega}' \rightarrow \hat{\Omega})
@@
   \label{eq:cart_1d_transport}
-\end{equation}
+\end{multline}
 where the angular component of the solution is no longer dependent on
@@
 given by Eq~(\ref{eq:mono_transport}):
-\begin{equation}
+\begin{multline}
   \frac{\partial}{\partial x}\Big[\sum_{n=0}^\infty (2n+1) \phi_n \mu
     P_n(\mu) \Big] + \sigma \sum_{n=0}^\infty (2n+1) \phi_n P_n(\mu) =
-  \int_{-1}^1 \sum_{m=0}^\infty (2m+1) \sigma_{sm} P_m(\mu_0)
+  \\ \int_{-1}^1 \sum_{m=0}^\infty (2m+1) \sigma_{sm} P_m(\mu_0)
   \sum_{n=0}^\infty (2n+1) \phi_n P_n(\mu') d \mu' + q\:,
   \label{eq:pn_deriv_1}
-\end{equation}
+\end{multline}
 where the dependence on the spatial variable, $x$, has been
@@
 %%---------------------------------------------------------------------------%%
-\section{Monte Carlo Solutions for the $SP_N$ Equations}
+\section{Monte Carlo Synthetic Acceleration for the $SP_N$ Equations}
 \label{sec:monte_carlo}

### 2013-02-28 doc/spn/spn.tex
updating pn derivation

@@
 usage of computational resources may be observed compared to modern
-solution techniques such as those suggested.
+solution techniques such as those suggested. In addition, solving the
+$SP_N$ equations in this way also breaks away from the $S_N$ forms of
+parallelism where spatial parallelism is achieved by an efficient
+parallel sweep, angular efficiency achieved by pipelining, and energy
+parallelism achieved by decoupling the groups. With the $SP_N$
+equations as a full matrix system, we now can parallelize the problem
+as prescribed by the linear solver, which may be significantly more
+scalable than current $S_N$ transport practices.
 
@@
 performance with both modern Krylov methods and Monte Carlo methods
-for asymmetric systems. Following this, we devise a Monte Carlo
+for asymmetric linear systems. Following this, we devise a Monte Carlo
 synthetic acceleration method as a proposed solution scheme.
@@
 The $P_N$ equations introduce the approximation that the angular
-dependence of the scattering cross section, $sigma_s$, and the angular
+dependence of the scattering cross section, $\sigma_s$, and the angular
 flux $\psi$, can be discretized by expanding them in Legendre
@@
 
-\paragraph{Removal Term}
-To reduce the removal term, the orthogonality relation is again
+\paragraph{Collision Term}
+To reduce the collision term, the orthogonality relation is again
 applied after the integration:
@@
 \end{equation}
-giving for the final removal term:
+giving for the final collision term:
 \begin{equation}

### 2013-03-01 doc/spn/spn.tex
starting spn derivation

--- 2013-02-28 doc/spn/spn.tex
@@
 \date{\today}
-\title{Monte Carlo Synthetic Acceleration Solutions for the $SP_N$ Equations}
+\title{Monte Carlo Solutions for the $SP_N$ Equations}
 \begin{document}
@@
 \label{sec:introduction}
-The neutron tranpsort problem is complicated. Solutions cover a large
-phase space and the problems of interest are often geometry complex,
-very large, or both requiring tremendous computational resources to
-generate an adequate solution. Modern deterministic methods for large
-scale problems are commonly variants on the discrete ordinates ($S_N$)
-method \cite{evans_denovo:_2010}. For full reactor core neutronics
-simulations, the $S_N$ method requires potentially trillions of
-unknown angular flux moments to be computed to achieve good accuracy
-for the repsonses to be computed. Other forms of the transport
-problem, including the $P_N$ method, take on a simpler form than the
-more common $S_N$ methods but lack in accuracy when compared while
-still requiring considerable computational resources for solutions in
-multiple dimensions.
-
-In the 1960's, Gelbard developed an ad-hoc multidimensional extension
-of the simple single dimension planar $P_N$ equations that created a
-system of coupled, diffusion-like equations known as the $SP_N$
-equations. Up until around the 1990's, the $SP_N$ methods was either
-widely unknown, widely unused, or combination of both even though
-numerical studies showed promising results with better solutions than
-diffusion theory and a significant reduction in computational time
-over more accurate methods such as discrete ordinates. Why did this
-happen? A significant problem, pointed out by Larsen, was that little
-rigor had been applied to the formulation of the $SP_N$ equations
-since their derivation through primarily heuristic arguments. Instead,
-studies at that time focused on simply comparing the results of the
-method to other contemporary transport solution strategies. In
-addition, many problems of interest from the literature at the time
-were either solved using nodal-type methods for reactor-sized problems
-are $S_N$-type methods for benchmark problems with intricate material
-configurations and potentially large flux gradients over small spatial
-domains.
-
-So why reconsider the $SP_N$ equations? Starting in the 1990's and
-primarily due to Larsen and his colleagues, the $SP_N$ equations have
-been given a more rigorous treatment with both variational and
-asymptotic derivations performed as a means of verification. In
-addition, these equations have been more rigorously studied as
-solution methods to MOX fuel problems and have been shown to provide
-accurate solutions. With this mathematical literature to provide a
-solid numerical footing for the method, we look at its application to
-today's challenge problems in full reactor core transport. The
-reduction in numerical complexity of current full core deterministic
-solution methods using the $S_N$ approximation could mean significant
-savings in both compute time and memory required. In addition, the
-characteristics of the solution to the transport problem for a steady
-state reactor core permit diffusion theory to be used; a staple of the
-nuclear industry since its inception. Therefore, if diffusion theory
-is applicable, then finer grained solutions that capture more of the
-physics contained in the transport equation should be possible with
-the $SP_N$ method. In doing so, we also expect from the literature to
-obtain computed responses on the order of accuracy we would expect
-from an appropriately discretized $S_N$ method at a fraction of the
-cost.
-
-To further motivate moving in this direction, recent developments in
-the Exnihilo neutronics package at Oak Ridge National Laboratory have
-permitted generation of the $SP_N$ system of equations for detailed
-full core reactor models. By fully forming these equations and
-formulating them as a linear algebra problem, we now have access to
-all of the modern advancements in computational linear algebra
-including Krylov solvers for asymmetric systems and preconditioning
-methods such as algebraic multigrid. This leads us to then explore the
-applicability of our work in discrete Monte Carlo methods for linear
-systems as a possible solution method for the $SP_N$ equations. If
-formulated correctly, we hypothesis that signficant improvement in
-usage of computational resources may be observed compared to modern
-solution techniques such as those suggested.
-
-In this document we derive the multigroup $SP_N$ equations from the
-Boltzmann transport equation. We then perform a spectral analysis on
-the resulting system of equations to determine its potential
-performance with both modern Krylov methods and Monte Carlo methods
-for asymmetric systems. Following this, we devise a Monte Carlo
-synthetic acceleration method as a proposed solution scheme.
 
@@
 The $P_N$ equations introduce the approximation that the angular
-dependence of the scattering cross section, $sigma_s$, and the angular
-flux $\psi$, can be discretized by expanding them in Legendre
+dependence of the scattering cross section, $\sigma_s$, and the
+angular flux $\psi$, can be discretized by expanding them in Legendre
 polynomials as follows:
@@
 \begin{equation}
-  \phi_0 - 5\phi_1 - 8\phi_2 = \phi_b\:.
+  \phi_0 - 5\phi_2 - 8\phi_3 = \phi_b\:.
   \label{eq:marshak_p1_deriv_3}
@@
 \label{subsec:spn_equations}
+The $SP_N$ equations give $N+1$ coupled first-order equations
+capturing the spatial and angular-dependence of the solution. In
+multiple dimensions, the equation set becomes large and coupled not
+only through angular moments but also through the spatial
+variables. As a simpler alternative to multidimensional $P_N$
+solutions, Gelbard recognized in 1960 that the planar $P_N$ equations
+could be simplified and applied an ad-hoc method to extend them to
+multiple dimensions, yielding the $SP_N$ equations. These equations
+are not only fewer in number, but also take on a diffusion-like form
+while maintaining the angular character of the flux, making them
+amenable to solutions with modern diffusion methods.
+
+First, the $P_N$ equations can be simplified to $(N+1)/2$ second-order
+equations by the following procedure:
 

### 2013-03-01 doc/spn/spn.tex
added sp7 equations

@@
 of the simple single dimension planar $P_N$ equations that created a
-system of coupled, diffusion-like equations known as the $SP_N$
-equations. Up until around the 1990's, the $SP_N$ methods was either
-widely unknown, widely unused, or combination of both even though
-numerical studies showed promising results with better solutions than
-diffusion theory and a significant reduction in computational time
-over more accurate methods such as discrete ordinates. Why did this
-happen? A significant problem, pointed out by Larsen, was that little
-rigor had been applied to the formulation of the $SP_N$ equations
-since their derivation through primarily heuristic arguments. Instead,
-studies at that time focused on simply comparing the results of the
-method to other contemporary transport solution strategies. In
-addition, many problems of interest from the literature at the time
-were either solved using nodal-type methods for reactor-sized problems
-or $S_N$-type methods for benchmark problems with intricate material
-configurations and potentially large flux gradients over small spatial
-domains.
+system of coupled, diffusion-like equations known as the simplified
+$P_N$ ($SP_N$) equations. Up until around the 1990's, the $SP_N$
+methods was either widely unknown, widely unused, or combination of
+both even though numerical studies showed promising results with
+better solutions than diffusion theory and a significant reduction in
+computational time over more accurate methods such as discrete
+ordinates. Why did this happen? A significant problem, pointed out by
+Larsen, was that little rigor had been applied to the formulation of
+the $SP_N$ equations since their derivation through primarily
+heuristic arguments. Instead, studies at that time focused on simply
+comparing the results of the method to other contemporary transport
+solution strategies. In addition, many problems of interest from the
+literature at the time were either solved using nodal-type methods for
+reactor-sized problems or $S_N$-type methods for benchmark problems
+with intricate material configurations and potentially large flux
+gradients over small spatial domains.
 
@@
 \label{subsec:spn_equations}
-The $SP_N$ equations give $N+1$ coupled first-order equations
+The $P_N$ equations give $N+1$ coupled first-order equations
 capturing the spatial and angular-dependence of the solution. In
@@
 First, the $P_N$ equations can be simplified to $(N+1)/2$ second-order
-equations by the following procedure:
+equations by solving for the $n^th$ Legendre flux moment in the
+odd-order equations:
+\begin{equation}
+  \phi_n = \frac{1}{\Sigma_n}\Bigg[ q \delta_{no} -
+    \frac{\partial}{\partial x}\Big(\frac{n}{2n+1}\phi_{n-1} +
+    \frac{n+1}{2n+1} \phi_{n+1} \Big) \Bigg]\:, 
+  \label{eq:odd_moments}
+\end{equation}
+for $n = 1,3,\cdots,N$ and $\delta_{no} = 0\ \forall n \neq 0$. We can
+insert the odd moments into Eq~(\ref{eq:final_pn_equations}) to get a
+reduced group of equations for the even moments:
+\begin{multline}
+  -\frac{\partial}{\partial x}
+  \Bigg[\frac{n}{2n+1}\frac{1}{\Sigma_{n-1}} \frac{\partial}{\partial
+      x} \Big(\frac{n-1}{2n-1} \phi_{n-2} + \frac{n}{2n-1}\phi_n \Big)
+    \\+ \frac{n+1}{2n+1}\frac{1}{\Sigma_{n+1}} \frac{\partial}{\partial
+      x} \Big(\frac{n+1}{2n+3}\phi_n + \frac{n+2}{2n+3}\phi_{n+2}\Big)
+    \Bigg] \\+ \Sigma_n \phi_n = q \delta_{n0}\ \ \ \ \ \ \ \ \ n =
+  0,2,4,\cdots,N\:.
+  \label{eq:reduced_pn}
+\end{multline}
+Immediately, we note the diffusion-like nature of
+Eq~(\ref{eq:reduced_pn}) as compared to the original $P_N$
+equations. To extend these equations to multiple dimensions, gelbard
+simple replaced the x-plane spatial derivatives in the reduced set of
+equations with general multidimensional gradient operators:
+\begin{multline}
+  -\nabla \cdot \Bigg[\frac{n}{2n+1}\frac{1}{\Sigma_{n-1}} \nabla
+    \Big(\frac{n-1}{2n-1} \phi_{n-2} + \frac{n}{2n-1}\phi_n \Big) \\+
+    \frac{n+1}{2n+1}\frac{1}{\Sigma_{n+1}} \nabla
+    \Big(\frac{n+1}{2n+3}\phi_n + \frac{n+2}{2n+3}\phi_{n+2}\Big)
+    \Bigg] \\+ \Sigma_n \phi_n = q \delta_{n0}\ \ \ \ \ \ \ \ \ n =
+  0,2,4,\cdots,N\:,
+  \label{eq:spn_equations}
+\end{multline}
+yielding a multidimensional set of $(N+1)/1$ angular coupled equations
+known as the $SP_N$. Again, we provide closure to this set of
+equations with $\phi_{N+1} = 0$. As a further modification to these
+equations, we can use a change of variables to create a new group of
+equations such that the gradients are operating on a single vector. As
+a concrete example, we will consider the $SP_7$ equations:
+\begin{subequations}
+  \begin{gather}
+    -\nabla \cdot \frac{1}{3 \Sigma_1} \nabla ( \phi_0 + \phi_2 ) +
+    \Sigma_0 \phi_0 = q \\ 
+    -\nabla \cdot \Bigg[ \frac{2}{15 \Sigma_1} \nabla ( \phi_0 + \phi_2
+      ) + \frac{3}{35 \Sigma_3}( 3\phi_2 + 4\phi_4)\Bigg] + \Sigma_2
+    \phi_2 = 0\\
+    -\nabla \cdot \Bigg[ \frac{4}{63 \Sigma_3} \nabla ( 3\phi_2 +
+      4\phi_4 ) + \frac{5}{99 \Sigma_5}( 5\phi_4 + 6\phi_6)\Bigg] +
+    \Sigma_4 \phi_4 = 0\\
+    -\nabla \cdot \Bigg[ \frac{6}{143 \Sigma_5} \nabla ( 5\phi_4 +
+      6\phi_6 ) + \frac{7}{195 \Sigma_7}( 7\phi_6)\Bigg] + \Sigma_6
+    \phi_6 = 0 
+  \end{gather}
+  \label{eq:sp7_equations}
+\end{subequations}
 

### 2013-03-01 doc/spn/spn.tex
working on derivation of SPn boundary conditions

--- 2013-03-01 doc/spn/spn.tex
@@
 \usepackage{amsmath}
+\usepackage{amsthm}
 \usepackage[usenames]{color}
@@
 First, the $P_N$ equations can be simplified to $(N+1)/2$ second-order
-equations by the following procedure:
+equations by solving for the $n^th$ Legendre flux moment in the
+odd-order equations:
+\begin{equation}
+  \phi_n = \frac{1}{\Sigma_n}\Bigg[ q \delta_{no} -
+    \frac{\partial}{\partial x}\Big(\frac{n}{2n+1}\phi_{n-1} +
+    \frac{n+1}{2n+1} \phi_{n+1} \Big) \Bigg]\:, 
+  \label{eq:odd_moments}
+\end{equation}
+for $n = 1,3,\cdots,N$ and $\delta_{no} = 0\ \forall n \neq 0$. We can
+insert the odd moments into Eq~(\ref{eq:final_pn_equations}) to get a
+reduced group of equations for the even moments:
+\begin{multline}
+  -\frac{\partial}{\partial x}
+  \Bigg[\frac{n}{2n+1}\frac{1}{\Sigma_{n-1}} \frac{\partial}{\partial
+      x} \Big(\frac{n-1}{2n-1} \phi_{n-2} + \frac{n}{2n-1}\phi_n \Big)
+    \\+ \frac{n+1}{2n+1}\frac{1}{\Sigma_{n+1}} \frac{\partial}{\partial
+      x} \Big(\frac{n+1}{2n+3}\phi_n + \frac{n+2}{2n+3}\phi_{n+2}\Big)
+    \Bigg] \\+ \Sigma_n \phi_n = q \delta_{n0}\ \ \ \ \ \ \ \ \ n =
+  0,2,4,\cdots,N\:.
+  \label{eq:reduced_pn}
+\end{multline}
+Immediately, we note the diffusion-like nature of
+Eq~(\ref{eq:reduced_pn}) as compared to the original $P_N$
+equations. To extend these equations to multiple dimensions, gelbard
+simple replaced the x-plane spatial derivatives in the reduced set of
+equations with general multidimensional gradient operators:
+\begin{multline}
+  -\nabla \cdot \Bigg[\frac{n}{2n+1}\frac{1}{\Sigma_{n-1}} \nabla
+    \Big(\frac{n-1}{2n-1} \phi_{n-2} + \frac{n}{2n-1}\phi_n \Big) \\+
+    \frac{n+1}{2n+1}\frac{1}{\Sigma_{n+1}} \nabla
+    \Big(\frac{n+1}{2n+3}\phi_n + \frac{n+2}{2n+3}\phi_{n+2}\Big)
+    \Bigg] \\+ \Sigma_n \phi_n = q \delta_{n0}\ \ \ \ \ \ \ \ \ n =
+  0,2,4,\cdots,N\:,
+  \label{eq:spn_equations}
+\end{multline}
+yielding a multidimensional set of $(N+1)/1$ angular coupled equations
+known as the $SP_N$. Again, we provide closure to this set of
+equations with $\phi_{N+1} = 0$. As a concrete example, we will
+consider the $SP_7$ equations:
+\begin{subequations}
+  \begin{gather}
+    -\nabla \cdot \frac{1}{3 \Sigma_1} \nabla ( \phi_0 + 2\phi_2 ) +
+    \Sigma_0 \phi_0 = q \\ 
+    -\nabla \cdot \Bigg[ \frac{2}{15 \Sigma_1} \nabla ( \phi_0 + 2\phi_2
+      ) + \frac{3}{35 \Sigma_3}\nabla( 3\phi_2 + 4\phi_4)\Bigg] +
+    \Sigma_2 \phi_2 = 0\\
+    -\nabla \cdot \Bigg[ \frac{4}{63 \Sigma_3} \nabla ( 3\phi_2 +
+      4\phi_4 ) + \frac{5}{99 \Sigma_5}\nabla( 5\phi_4 +
+      6\phi_6)\Bigg] + \Sigma_4 \phi_4 = 0\\
+    -\nabla \cdot \Bigg[ \frac{6}{143 \Sigma_5} \nabla ( 5\phi_4 +
+      6\phi_6 ) + \frac{7}{195 \Sigma_7}\nabla(7\phi_6)\Bigg] +
+    \Sigma_6 \phi_6 = 0 \:.
+  \end{gather}
+  \label{eq:sp7_equations}
+\end{subequations}
+To further modify these equations, we can use a change of variables to
+create a new group of equations such that the gradients are operating
+on a single vector:
+\begin{subequations}
+  \begin{gather}
+    u_1 = \phi_0 + 2\phi_2 \\
+    u_2 = 3\phi_2 + 4\phi_4 \\
+    u_3 = 5\phi_4 + 6\phi_6 \\
+    u_4 = 7\phi_6 \:,
+  \end{gather}
+  \label{eq:spn7_subs}
+\end{subequations}
+or equivalently:
+\begin{subequations}
+  \begin{gather}
+    \phi_0 = u_1 - \frac{2}{3}u_2 + \frac{8}{15}u_3 -
+    \frac{16}{35}u_4\\
+    \phi_2 = \frac{1}{3}u_2 - \frac{4}{15}u_3 + \frac{8}{35}u_4\\ 
+    \phi_4 = \frac{1}{5}u_3 - \frac{6}{35}u_4\\
+    \phi_6 = \frac{1}{7}u_4\:.
+  \end{gather}
+  \label{eq:spn7_subs_inverse}
+\end{subequations}
+When substituted into Eq~(\ref{eq:sp7_equations}), these terms give:
+\begin{subequations}
+  \begin{gather}
+    -\nabla \cdot \frac{1}{3 \Sigma_1} \nabla u_1 + \Sigma_0 \Bigg[
+    u_1 - \frac{2}{3}u_2 + \frac{8}{15}u_3 - \frac{16}{35}u_4 \Bigg]
+    = -q \\
+    -\nabla \cdot \Bigg[ \frac{2}{15 \Sigma_1} \nabla u_1 +
+    \frac{3}{35 \Sigma_3} \nabla u_2 \Bigg] + \Sigma_2 \Bigg[
+    \frac{1}{3}u_2 - \frac{4}{15}u_3 + \frac{8}{35}u_4 \Bigg] = 0 \\
+    -\nabla \cdot \Bigg[ \frac{4}{63 \Sigma_3} \nabla u_2 +
+    \frac{5}{99 \Sigma_5} \nabla u_3 \Bigg] + \Sigma_4 \Bigg[
+    \frac{1}{5}u_3 - \frac{6}{35}u_4 \Bigg] = 0 \\ 
+    -\nabla \cdot \Bigg[ \frac{6}{143 \Sigma_5} \nabla u_3 +
+    \frac{7}{195 \Sigma_7} \nabla u_4 \Bigg] + \Sigma_6 \Bigg[
+    \frac{1}{7}u_4 \Bigg] = 0 \:.
+  \end{gather}
+  \label{eq:spn7_subs_equations}
+\end{subequations}
+If we rearrange the Eq~(\ref{eq:spn7_subs_equations} such that only
+one divergence operation is present in each equation, we can formulate
+this as a matrix system:
+\begin{equation}
+  -\nabla \cdot \mathbf{D} \nabla \mathbf{u} + \mathbf{A} \mathbf{u} =
+  \mathbf{q}\:,
+  \label{eq:spn_matrix}
+\end{equation}
+with $\mathbf{u}$ the vector of solution variables:
+\begin{equation}
+  \mathbf{u} = ( u_1\ \ u_2\ \ u_3\ \ u_4 )^T \:,
+  \label{eq:spn7_solution_vector}
+\end{equation}
+$\mathbf{D}$ the vector of effective diffusion coefficients:
+\begin{equation}
+  \mathbf{D} = \Bigg( \frac{1}{3\Sigma_1}\ \ \frac{1}{7\Sigma_3}\ \
+  \frac{1}{11\Sigma_5}\ \ \frac{1}{15\Sigma_7} \Bigg)^T\:,
+  \label{eq:spn7_diffusion_coeffs}
+\end{equation}
+$\mathbf{q}$ the vector of source terms where the $0^{th}$ moment
+source has now been distributed through the system:
+\begin{equation}
+  \mathbf{q} = (
+  q\ \ -\frac{2}{3}q\ \ \frac{8}{15}q\ \ -\frac{16}{35}q )^T\:,
+  \label{eq:spn7_source_vector}
+\end{equation}
+and $\mathbf{A}$ a matrix of angular scattering terms:
+% NOTE: I copied the following matrix directly out of Tom's tech note
+% on the SPn equations which I am effectively following here because I
+% was feeling lazy. I have verified its correctness.
+\begin{equation}
+  \mathbf{A} = 
+  \begin{bmatrix}
+    (\Sigma_0) &
+    (-\frac{2}{3}\Sigma_0) &
+    (\frac{8}{15}\Sigma_0) &
+    (-\frac{16}{35}\Sigma_0) \\
+    %%
+    &&&\\
+    %%
+    (-\frac{2}{3}\Sigma_0) &
+    (\frac{4}{15}\Sigma_0 + \frac{1}{3}\Sigma_2) &
+    (-\frac{16}{45}\Sigma_0 - \frac{4}{9}\Sigma_2) &
+    (\frac{32}{105}\Sigma_0 + \frac{8}{21}\Sigma_2) \\
+    %%
+    &&&\\
+    %%
+    (\frac{8}{15}\Sigma_0) &
+    (-\frac{16}{45}\Sigma_0 - \frac{4}{9}\Sigma_2) &
+    (\frac{64}{225}\Sigma_0 + \frac{16}{45}\Sigma_2 + \frac{9}{25}\Sigma_4) &
+    (-\frac{128}{525}\Sigma_0 - \frac{32}{105}\Sigma_2 - \frac{54}{175}\Sigma_4)
+    \\ 
+    %%
+    &&&\\
+    %%
+    (-\frac{16}{35}\Sigma_0) &
+    (\frac{32}{105}\Sigma_0 + \frac{8}{21}\Sigma_2) &
+    (-\frac{128}{525}\Sigma_0 - \frac{32}{105}\Sigma_2 - \frac{54}{175}\Sigma_4)
+    & 
+    (\frac{256}{1225}\Sigma_0 + \frac{64}{245}\Sigma_2 +
+    \frac{324}{1225}\Sigma_4 + \frac{13}{49}\Sigma_6)
+  \end{bmatrix}\:.
+\end{equation}
+This completes the derivation of the $SP_7$ equations for a single
+energy group. As noted by Evans, lower order $SP_N$ approximations can
+be generated by setting higher order even moments in this system to
+zero (e.g. $\phi_6 = \phi_4 = 0$ yields the $SP_3$ equations).
+
+\subsubsection{Boundary Conditions for the $SP_N$ Equations}
+\label{subsubsec:spn_boundary_conditions}
+We approach the formulation of the boundary conditions in much the
+same way as we did for the $P_N$ equations with both reflecting and
+Marshak conditions possible. To begin, we perform the expansion in
+Eq~(\ref{eq:marshak_expanded}) for the left side of the planar system,
+this time for $N=7$ to correspond to our $SP_7$ system. For $i=1$:
+\begin{multline}
+  \int_0^1 \mu \Bigg[ \phi_0 + 3\phi_1\mu +
+    \frac{5}{2}\phi_2(3\mu^2-1) + \frac{7}{2}\phi_3(5\mu^3-3\mu)
+    +\\ \frac{1}{8}(63\mu^5-70\mu^3+15\mu) \Bigg] d\mu = \int_0^1 \mu
+  \phi_b d\mu\:,
+  \label{eq:spn_bnd_p1}
+\end{multline}
+for $i=3$,
+\begin{multline}
+  \int_0^1 \frac{1}{2}(5\mu^3-3\mu) \Bigg[ \phi_0 + 3\phi_1\mu +
+    \frac{5}{2}\phi_2(3\mu^2-1) + \frac{7}{2}\phi_3(5\mu^3-3\mu) +\\
+    \frac{1}{8}(63\mu^5-70\mu^3+15\mu) \Bigg] d\mu = \int_0^1
+  \frac{1}{2}(5\mu^3-3\mu) \phi_b d\mu\:,
+  \label{eq:spn_bnd_p3}
+\end{multline}
+for $i=5$:
+\begin{multline}
+  \int_0^1 \frac{1}{8}(63\mu^5-70\mu^3+15\mu) \Bigg[ \phi_0 + 3\phi_1\mu +
+    \frac{5}{2}\phi_2(3\mu^2-1) + \frac{7}{2}\phi_3(5\mu^3-3\mu) +\\
+    \frac{1}{8}(63\mu^5-70\mu^3+15\mu) \Bigg] d\mu = \int_0^1
+  \frac{1}{8}(63\mu^5-70\mu^3+15\mu) \phi_b d\mu\:,
+  \label{eq:spn_bnd_p5}
+\end{multline}
+and for $i=7$:
+\begin{multline}
+  \int_0^1 \frac{1}{16}(429\mu^7-693\mu^5+315\mu^3+35\mu) \Bigg[
+    \phi_0 + 3\phi_1\mu + \frac{5}{2}\phi_2(3\mu^2-1) +
+    \frac{7}{2}\phi_3(5\mu^3-3\mu) \\+
+    \frac{1}{8}(63\mu^5-70\mu^3+15\mu) +
+    \frac{1}{16}(429\mu^7-693\mu^5+315\mu^3+35\mu) \Bigg] d\mu =\\
+  \int_0^1 \frac{1}{16}(429\mu^7-693\mu^5+315\mu^3+35\mu) \phi_b
+  d\mu\:.
+  \label{eq:spn_bnd_p7}
+\end{multline}
+Carrying out the simple integrations yields the following system of
+equations:
+\begin{subequations}
+  \begin{gather}
+    \frac{1}{2}\phi_0 + \phi_1 + \frac{5}{8}\phi_2 -
+    \frac{3}{16}\phi_4 + \frac{13}{128}\phi_6 =
+    \frac{1}{2}\phi_{b}\\ -\frac{1}{8}\phi_0 + \frac{5}{8}\phi_2 +
+    \phi_3 + \frac{81}{128}\phi_4 - \frac{13}{64}\phi_6 =
+    -\frac{1}{8}\phi_{b}\\ \frac{1}{16}\phi_0 - \frac{25}{128}\phi_2 +
+    \frac{81}{128}\phi_4 + \phi_5 + \frac{325}{512}\phi_6 =
+    \frac{1}{16}\phi_{b}\\ -\frac{5}{128}\phi_0 + \frac{7}{64}\phi_2 -
+    \frac{105}{512}\phi_4 + \frac{325}{512}\phi_6 + \phi_7 =
+    -\frac{5}{128}\phi_{b}\:,\\
+  \end{gather}
+  \label{eq:spn_bnd_integrated}
+\end{subequations}
+where $\phi_b$ is again an isotropic source prescribed on the planar
+boundary. For the odd moment in each boundary equation, we insert
+Eq~(\ref{eq:odd_moments}) to remove them, leaving only the even moments:
 
@@
 \pagebreak
-\bibliographystyle{ieeetr} \bibliography{references}
+\bibliographystyle{ieeetr} 
+\bibliography{references}
 \end{document}

### 2013-03-02 doc/spn/spn.tex
working on multigroup formulation

@@
   \hat{\Omega} \cdot \vec{\nabla} \psi(\vec{r},\hat{\Omega},E) +
-  \sigma(\vec{r},E) \psi(\vec{r},\hat{\Omega},E) = \\ \int dE' \int
-  d\Omega' \sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}' \rightarrow
-  \hat{\Omega}) \psi(\vec{r},\hat{\Omega}',E') +
+  \sigma(\vec{r},E) \psi(\vec{r},\hat{\Omega},E) = \\ \int \int
+  \sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}' \rightarrow
+  \hat{\Omega}) \psi(\vec{r},\hat{\Omega}',E') d\Omega' dE' +
   q(\vec{r},\hat{\Omega},E)\:,
@@
   \mu \frac{\partial}{\partial x} \psi(x,\mu,E) + \sigma(x,E)
-  \psi(x,\mu,E) = \\ \int dE' \int d\Omega' \sigma_s(x,E' \rightarrow
-  E,\hat{\Omega}' \rightarrow \hat{\Omega})
-  \psi(\vec{r},\hat{\Omega}',E') + \frac{q(x,E)}{4 \pi}\:,
+  \psi(x,\mu,E) = \\ \int \int \sigma_s(x,E') \rightarrow
+  E,\hat{\Omega}' \rightarrow \hat{\Omega}) \psi(x,\hat{\Omega}',E')
+  d\Omega' dE' + \frac{q(x,E)}{4 \pi}\:,
   \label{eq:cart_1d_transport}
@@
 \end{equation}
-These functions have several nice properties including orthogonality:
+These functions have several useful properties including
+orthogonality:
 \begin{equation}
@@
     \phi_0 = u_1 - \frac{2}{3}u_2 + \frac{8}{15}u_3 -
-    \frac{16}{35}u_4\\
+    \frac{16}{35}u_4 \\
     \phi_2 = \frac{1}{3}u_2 - \frac{4}{15}u_3 + \frac{8}{35}u_4\\ 
@@
     \frac{105}{512}\phi_4 + \frac{325}{512}\phi_6 + \phi_7 =
-    -\frac{5}{128}\phi_{b}\:,\\
+    -\frac{5}{128}\phi_{b}\:,
   \end{gather}
@@
 boundary. For the odd moment in each boundary equation, we insert
-Eq~(\ref{eq:odd_moments}) to remove them, leaving only the even moments:
+Eq~(\ref{eq:odd_moments}) to remove them, leaving only the even
+moments and a set of differential equations:
+\begin{subequations}
+  \begin{gather}
+    \frac{1}{2}\phi_0 + \frac{1}{3\Sigma_1}\frac{\partial}{\partial
+      x}(\phi_0+2\phi_2) + \frac{5}{8}\phi_2 - \frac{3}{16}\phi_4 +
+    \frac{13}{128}\phi_6 = \frac{1}{2}\phi_{b}\\ -\frac{1}{8}\phi_0 +
+    \frac{5}{8}\phi_2 + \frac{1}{7\Sigma_3}\frac{\partial}{\partial
+      x}(3\phi_2 + 4\phi_4) + \frac{81}{128}\phi_4 -
+    \frac{13}{64}\phi_6 = -\frac{1}{8}\phi_{b}\\ \frac{1}{16}\phi_0 -
+    \frac{25}{128}\phi_2 + \frac{81}{128}\phi_4 +
+    \frac{1}{11\Sigma_5}\frac{\partial}{\partial x}(5\phi_4 + 6\phi_6)
+    + \frac{325}{512}\phi_6 =
+    \frac{1}{16}\phi_{b}\\ -\frac{5}{128}\phi_0 + \frac{7}{64}\phi_2 -
+    \frac{105}{512}\phi_4 + \frac{325}{512}\phi_6 +
+    \frac{1}{15\Sigma_7}\frac{\partial}{\partial x}(7\phi_6)
+    = -\frac{5}{128}\phi_{b}\:.
+  \end{gather}
+  \label{eq:spn_bnd_subs}
+\end{subequations}
+We can again make the substitution of variables given by
+Eqs~(\ref{eq:spn7_subs}) and (\ref{eq:spn7_subs_inverse}) for
+consitency with the equations defined on the domain. In addition, we
+apply the $SP_N$ approximation to the derivatives by assuming they are
+instead multidimensional gradients on the boundary such that
+$\frac{\partial}{\partial x} \rightarrow
+\hat{\mathbf{n}}\cdot\nabla$. Doing this gives:
+\begin{multline}
+    \frac{1}{2}\Big(u_1 - \frac{2}{3}u_2 + \frac{8}{15}u_3 -
+    \frac{16}{35}u_4\Big) +
+    \frac{1}{3\Sigma_1}\hat{\mathbf{n}}\cdot\nabla\Big(\Big(u_1 -
+    \frac{2}{3}u_2 + \frac{8}{15}u_3 -
+    \frac{16}{35}u_4\Big)+\\2\Big(\frac{1}{3}u_2 - \frac{4}{15}u_3 +
+    \frac{8}{35}u_4\Big)\Big) + \frac{5}{8}\Big(\frac{1}{3}u_2 -
+    \frac{4}{15}u_3 + \frac{8}{35}u_4\Big)
+    -\\ \frac{3}{16}\Big(\frac{1}{5}u_3 - \frac{6}{35}u_4\Big) +
+    \frac{13}{128}\Big(\frac{1}{7}u_4\Big) = \frac{1}{2}\phi_{b}\:,
+\end{multline}
+\begin{multline}
+    -\frac{1}{8}\Big(u_1 - \frac{2}{3}u_2 + \frac{8}{15}u_3 -
+    \frac{16}{35}u_4\Big) + \frac{5}{8}\Big(\frac{1}{3}u_2 -
+    \frac{4}{15}u_3 + \frac{8}{35}u_4\Big)
+    +\\ \frac{1}{7\Sigma_3}\hat{\mathbf{n}}\cdot\nabla\Big(3\Big(\frac{1}{3}u_2
+    - \frac{4}{15}u_3 + \frac{8}{35}u_4\Big) + 4\Big(\frac{1}{5}u_3 -
+    \frac{6}{35}u_4\Big)\Big) + \frac{81}{128}\Big(\frac{1}{5}u_3
+    -\\ \frac{6}{35}u_4\Big) - \frac{13}{64}\Big(\frac{1}{7}u_4\Big) =
+    -\frac{1}{8}\phi_{b}\:,
+\end{multline}
+\begin{multline}
+    \frac{1}{16}\Big(u_1 - \frac{2}{3}u_2 + \frac{8}{15}u_3 -
+    \frac{16}{35}u_4\Big) - \frac{25}{128}\Big(\frac{1}{3}u_2 -
+    \frac{4}{15}u_3 +\\ \frac{8}{35}u_4\Big) +
+    \frac{81}{128}\Big(\frac{1}{5}u_3 - \frac{6}{35}u_4\Big) +
+    \frac{1}{11\Sigma_5}\hat{\mathbf{n}}\cdot\nabla\Big(5\Big(\frac{1}{5}u_3
+    -\\ \frac{6}{35}u_4\Big) + 6\Big(\frac{1}{7}u_4\Big)\Big) +
+    \frac{325}{512}\Big(\frac{1}{7}u_4\Big) = \frac{1}{16}\phi_{b}\:,
+\end{multline}
+\begin{multline}
+    -\frac{5}{128}\Big(u_1 - \frac{2}{3}u_2 + \frac{8}{15}u_3 -
+    \frac{16}{35}u_4\Big) + \frac{7}{64}\Big(\frac{1}{3}u_2 -
+    \frac{4}{15}u_3 + \frac{8}{35}u_4\Big)
+    -\\ \frac{105}{512}\Big(\frac{1}{5}u_3 - \frac{6}{35}u_4\Big) +
+    \frac{325}{512}\Big(\frac{1}{7}u_4\Big)
+    +\\ \frac{1}{15\Sigma_7}\hat{\mathbf{n}}\cdot\nabla\Big(7\Big(\frac{1}{7}u_4\Big)\Big)
+    = -\frac{5}{128}\phi_{b}\:.
+\end{multline}
+By rearranging the system such that we have a single gradient operator
+in each equation, we again have a matrix system:
+\begin{equation}
+  \hat{\mathbf{n}} \cdot \mathbf{D} \nabla \mathbf{u} + \mathbf{B}
+  \mathbf{u} = \mathbf{s}\:,
+  \label{eq:spn_bnd_matrix}
+\end{equation}
+where $\mathbf{D}$ and $\mathbf{u}$ are defined as before and:
+\begin{equation}
+  \mathbf{s} = \Big(\frac{1}{2}\phi_b\ -\frac{1}{8}\phi_b\
+  \frac{1}{16}\phi_b\ -\frac{5}{128}\phi_b \Big)^T\:,
+  \label{eq:spn_bnd_source}
+\end{equation}
+is the source vector on the boundary and
+%% I grabbed this matrix from Tom's tech note again.
+\begin{equation}
+  \mathbf{B} = \begin{bmatrix}
+    \frac{1}{2} &
+    -\frac{1}{8} &
+    \frac{1}{16} &
+    -\frac{5}{128} \\
+    %%
+    &&&\\
+    %%
+    -\frac{1}{8} &
+    \frac{7}{24} &
+    -\frac{41}{384} &
+    \frac{1}{16} \\
+    %%
+    &&&\\
+    %%
+    \frac{1}{16} &
+    -\frac{41}{384} &
+    \frac{407}{1920} &
+    -\frac{233}{2560} \\
+    %%
+    &&&\\
+    %%
+    -\frac{5}{128} &
+    \frac{1}{16} &
+    -\frac{233}{2560} &
+    \frac{3023}{17920}
+  \end{bmatrix}\:,
+  \label{eq:B}
+\end{equation}
+is a dense matrix of coefficients. Again, as pointed out by Evans, the
+$SP_1$ approximation reduces this boundary condition to the standard
+Marshak diffusion boundary condition. For reflecting boundary
+conditions, we use the same procedure as the $P_N$ equations where the
+odd-moments are zero such that Eq~(\ref{eq:reflecting_condition}) is
+true. From setting Eq~(\ref{eq:odd_moments}) to zero for odd $\phi_n$
+and again substituting Eq~(\ref{eq:spn7_subs}) we immediately find
+that:
+\begin{equation}
+  \nabla \mathbf{u} = 0
+  \label{eq:spn_reflecting}
+\end{equation}
+for reflecting $SP_N$ boundaries, providing enough equations to close
+the system.
 
@@
 \label{subsec:mg_spn_equations}
+Up to this point, we have formulated the $P_N$ and subsequently the
+$SP_N$ equations for a single neutron energy. To expand these
+equations for multiple energies, we start by stating the multigroup
+neutron transport equation for a single dimension for planar geometry:
+\begin{multline}
+  \mu \frac{\partial}{\partial x} \psi^g(x,\mu) + \sigma^g(x)
+  \psi^g(x,\mu) = \\ \sum_{g'=0}^{G} \int
+  \sigma_s^{gg'}(x,\hat{\Omega}' \rightarrow \hat{\Omega})
+  \psi^{g'}(x,\hat{\Omega}') d\Omega' + \frac{q^g(x)}{4 \pi}\:,
+  \label{eq:cart_1d_multigroup}
+\end{multline}
+where $g$ denotes the group index of $0$ to $G$ groups where $G=N_g-1$
+and the integration of the scattering emission term of energy has been
+replaced by a discrete summation. For scattering, $\sigma_s^{gg'}$
+provides the probability of scattering at a particular angle from
+group $g$ to $g'$. The result is an equation nearly identical in form
+to Eq~(\ref{eq:cart_1d_transport}) where now instead of forming the
+$P_N$ and $SP_N$ equations for a single energy group, we form them for
+each of the energy groups with group coupling occuring through the
+scattering term. The multigroup $P_N$ equations are then:
+\begin{equation}
+   \frac{1}{2n+1} \frac{\partial}{\partial x}\Big[ (n+1) \phi^g_{n+1}
+     + n \phi^g_{n-1} \Big] +
+   \sum_{g'}(\sigma^g\delta_{gg'}-\sigma^{gg'}_{sn}) \phi^g_n =
+   q\delta_{n0} \:,
+  \label{eq:multigroup_pn_equations}
+\end{equation}
+for $n = 0,1,\dotsc,N$ where the flux and scattering moments are
+defined in a group. We observe that a $N_g \times N_g$ scattering
+matrix is formed:
+\begin{equation}
+  \mathbf{\Sigma}_n =
+  \sum_{g'}(\sigma^g\delta_{gg'}-\sigma^{gg'}_{sn})\:,
+  \label{eq:scattering_matrix}
+\end{equation}
+or expanded:
+%% Also grabbed this matrix from Tom's tech note.
+\begin{equation}
+  \mathbf{\Sigma}_n =
+  \begin{bmatrix}
+    (\sigma^0-\sigma_{sn}^{00}) & -\sigma_{sn}^{01} & \dots &
+    -\sigma_{sn}^{0G} \\ &&&\\ -\sigma_{sn}^{10} &
+    (\sigma^1-\sigma_{sn}^{11}) & \dots & -\sigma_{sn}^{1G}
+    \\ &&&\\ \vdots & \vdots & \ddots & \vdots
+    \\ &&&\\ -\sigma_{sn}^{G0} & -\sigma_{sn}^{G1} & \dots &
+    (\sigma^G-\sigma_{sn}^{GG})
+  \end{bmatrix}\:.
+\end{equation}
+It is also useful to combine the group flux moments and sources into a
+single vector for more compact notation:
+\begin{equation}
+  \mathbf{\Phi_n} = (\phi^0_n\ \phi^1_n\ \cdots \phi^G_n )^T\:, 
+  \label{eq:group_flux_vector}
+\end{equation}
+\begin{equation}
+  \mathbf{q} = (q^0\ q^1\ \cdots q^G )^T\:.
+  \label{eq:group_source_vector}
+\end{equation}
+Next, we apply the $SP_N$ approximation to
+Eq~(\ref{eq:multigroup_pn_equations}) in identical fashion to the
+monoenergetic case. This gives:
+\begin{multline}
+  -\nabla \cdot \Bigg[\frac{n}{2n+1}\mathbf{\Sigma_{n-1}}^{-1} \nabla
+    \Big(\frac{n-1}{2n-1} \mathbf{\Phi_{n-2}} +
+    \frac{n}{2n-1}\mathbf{\Phi_n} \Big) \\+
+    \frac{n+1}{2n+1}\mathbf{\Sigma_{n+1}}^{-1} \nabla
+    \Big(\frac{n+1}{2n+3}\mathbf{\Phi_n} +
+    \frac{n+2}{2n+3}\mathbf{\Phi_{n+2}}\Big) \Bigg] \\+
+  \mathbf{\Sigma_n} \mathbf{\Phi_n} = \mathbf{q}
+  \delta_{n0}\ \ \ \ \ \ \ \ \ n = 0,2,4,\cdots,N\:.
+  \label{eq:multigroup_spn_equations}
+\end{multline}
+This adds more complexity than the monoenergetic formulation in that
+all unknowns in this group of equations are now vector quantities and
+scattering relationships are contained in matrices rather than a
+scalar quantity.
 

### 2013-03-03 doc/spn/spn.tex
small changes to multigroup section of spn doc

@@
   \end{bmatrix}\:.
+  \label{eq:A_matrix}
 \end{equation}
@@
   \end{bmatrix}\:,
-  \label{eq:B}
+  \label{eq:B_matrix}
 \end{equation}
@@
 scattering relationships are contained in matrices rather than a
-scalar quantity.
+scalar quantity. Because of this, the same sequence of variable
+changes and algebra can be used to build a set of matrix equations,
+this time in a block formulation:
+\begin{equation}
+  -\nabla \cdot \mathbb{D} \nabla \mathbb{U} + \mathbb{A} \mathbb{U} =
+  \mathbb{Q}\:,
+  \label{eq:spn_multigroup_system}
+\end{equation}
+where the definition of all quantities are the same with internal
+scalar values replaced by the group-vector values. In addition,
+$\mathbb{A}$ is now a block matrix of $N_g \times N_g$ submatrices
+generated from the moment scattering matrices. Analogously, on the
+boundaries we have:
+\begin{equation}
+  \hat{\mathbf{n}} \cdot \mathbb{D} \nabla \mathbb{U} + \mathbb{B}
+  \mathbb{U} = \mathbb{S}\:.
+  \label{eq:spn_multigroup_bnd}
+\end{equation}
 

### 2013-03-04 doc/spn/spn.tex
small updates to spn doc

@@
 The neutron transport problem is complicated. Solutions cover a large
-phase space and the problems of interest are often geometry complex,
-very large, or both requiring tremendous computational resources to
-generate an adequate solution. Modern deterministic methods for large
-scale problems are commonly variants on the discrete ordinates ($S_N$)
-method \cite{evans_denovo:_2010}. For full reactor core neutronics
-simulations, the $S_N$ method requires potentially trillions of
-unknown angular flux moments to be computed to achieve good accuracy
-for the repsonses to be computed. Other forms of the transport
-problem, including the $P_N$ method, take on a simpler form than the
-more common $S_N$ methods but lack in accuracy when compared while
-still requiring considerable computational resources for solutions in
-multiple dimensions.
+phase space and the problems of interest are often geometrically
+complex, very large, or both, requiring tremendous computational
+resources to generate an adequate solution. Modern deterministic
+methods for large scale problems are commonly variants on the discrete
+ordinates ($S_N$) method \cite{evans_denovo:_2010}. For full reactor
+core neutronics simulations, the $S_N$ method requires potentially
+trillions of unknown angular flux moments to be computed to achieve
+good accuracy for the repsonses of interest. Other forms of the
+transport problem, including the $P_N$ method, take on a simpler form
+than the more common $S_N$ methods but lack in accuracy when compared
+while still requiring considerable computational resources for
+solutions in multiple dimensions.
 
@@
 systems as a possible solution method for the $SP_N$ equations. If
-formulated correctly, we hypothesis that signficant improvement in
+formulated correctly, we hypothesize that signficant improvement in
 usage of computational resources may be observed compared to modern
@@
 \label{sec:derivation}
-In this section, we derive the $SP_N$ equations. We begin by stating
-the general time-independent neutron transport equation followed by a
-derivation of the $P_N$ equations in planar geometry for multiple
-energy groups. From these equations, we then apply a set of
-approximations to yield the multi-dimensional, multi-group $SP_N$
-equations for fixed source problems.
+In this section, we derive the $SP_N$ equations, closely following the
+work of Evans. We begin by stating the general time-independent
+neutron transport equation followed by a derivation of the $P_N$
+equations in planar geometry for multiple energy groups. From these
+equations, we then apply a set of approximations to yield the
+multi-dimensional, multi-group $SP_N$ equations for fixed source
+problems.
 
@@
 equations for multiple energies, we start by stating the multigroup
-neutron transport equation for a single dimension for planar geometry:
+neutron transport equation for a single dimension in planar geometry:
 \begin{multline}
@@
 \end{multline}
-where $g$ denotes the group index of $0$ to $G$ groups where $G=N_g-1$
-and the integration of the scattering emission term of energy has been
+where $g$ denotes the group index of $0$ to $G$ groups, $G=N_g-1$, and
+the integration of the scattering emission term of energy has been
 replaced by a discrete summation. For scattering, $\sigma_s^{gg'}$
@@
 \end{equation}
-or expanded:
+and when expanded gives:
 %% Also grabbed this matrix from Tom's tech note.

### 2013-03-04 doc/spn/spn.tex
finished SPN derivation, on to spectral analysis

@@
 one divergence operation is present in each equation, we can formulate
-this as a matrix system:
-\begin{equation}
-  -\nabla \cdot \mathbf{D} \nabla \mathbf{u} + \mathbf{A} \mathbf{u} =
-  \mathbf{q}\:,
+this as a matrix system of 4 equations in the case of the $SP_7$
+approximation:
+\begin{equation}
+  -\nabla \cdot D_n \nabla u_n + \sum_{m=1}^4 A_{nm} u_m =
+  q_n\ \ \ \ \ \ \ n = 1,2,3,4\:,
   \label{eq:spn_matrix}
@@
 \end{equation}
-This completes the derivation of the $SP_7$ equations for a single
-energy group. As noted by Evans, lower order $SP_N$ approximations can
-be generated by setting higher order even moments in this system to
-zero (e.g. $\phi_6 = \phi_4 = 0$ yields the $SP_3$ equations).
+Note that the term $\sum_{m=1}^4 A_{nm} u_m$ in
+Eq~(\ref{eq:spn_matrix}) couples the moments in each equation while
+the diffusive term in each equation is only for a single
+'psuedo-moment' $u_n$. This completes the derivation of the $SP_7$
+equations for a single energy group. As noted by Evans, lower order
+$SP_N$ approximations can be generated by setting higher order even
+moments in this system to zero (e.g. $\phi_6 = \phi_4 = 0$ yields the
+$SP_3$ equations).
 
@@
     \frac{4}{15}u_3 + \frac{8}{35}u_4\Big)
-    +\\ \frac{1}{7\Sigma_3}\hat{\mathbf{n}}\cdot\nabla\Big(3\Big(\frac{1}{3}u_2
-    - \frac{4}{15}u_3 + \frac{8}{35}u_4\Big) + 4\Big(\frac{1}{5}u_3 -
-    \frac{6}{35}u_4\Big)\Big) + \frac{81}{128}\Big(\frac{1}{5}u_3
-    -\\ \frac{6}{35}u_4\Big) - \frac{13}{64}\Big(\frac{1}{7}u_4\Big) =
-    -\frac{1}{8}\phi_{b}\:,
+    +\\ \frac{1}{7\Sigma_3}\hat{\mathbf{n}}\cdot\nabla\Big
+    (3\Big(\frac{1}{3}u_2 - \frac{4}{15}u_3 + \frac{8}{35}u_4\Big) +
+    4\Big(\frac{1}{5}u_3 - \frac{6}{35}u_4\Big)\Big) +
+    \frac{81}{128}\Big(\frac{1}{5}u_3 -\\ \frac{6}{35}u_4\Big) -
+    \frac{13}{64}\Big(\frac{1}{7}u_4\Big) = -\frac{1}{8}\phi_{b}\:,
 \end{multline}
@@
     \frac{325}{512}\Big(\frac{1}{7}u_4\Big)
-    +\\ \frac{1}{15\Sigma_7}\hat{\mathbf{n}}\cdot\nabla\Big(7\Big(\frac{1}{7}u_4\Big)\Big)
-    = -\frac{5}{128}\phi_{b}\:.
+    +\\ \frac{1}{15\Sigma_7}\hat{\mathbf{n}}
+    \cdot\nabla\Big(7\Big(\frac{1}{7}u_4\Big)\Big) =
+    -\frac{5}{128}\phi_{b}\:.
 \end{multline}
@@
 \begin{equation}
-  \hat{\mathbf{n}} \cdot \mathbf{D} \nabla \mathbf{u} + \mathbf{B}
-  \mathbf{u} = \mathbf{s}\:,
+  \mathbf{n} \cdot D_n \nabla u_n + \sum_{m=1}^4 B_{nm} u_m =
+  s_n\ \ \ \ \ \ \ n = 1,2,3,4\:,
   \label{eq:spn_bnd_matrix}
@@
 \begin{equation}
-  -\nabla \cdot \mathbb{D} \nabla \mathbb{U} + \mathbb{A} \mathbb{U} =
-  \mathbb{Q}\:,
+  -\nabla \cdot \mathbb{D}_n \nabla \mathbb{U}_n + \sum_{m=1}^4
+  \mathbb{A}_{nm} \mathbb{U}_m = \mathbb{Q}_n\:,
   \label{eq:spn_multigroup_system}
@@
 $\mathbb{A}$ is now a block matrix of $N_g \times N_g$ submatrices
-generated from the moment scattering matrices. Analogously, on the
-boundaries we have:
-\begin{equation}
-  \hat{\mathbf{n}} \cdot \mathbb{D} \nabla \mathbb{U} + \mathbb{B}
-  \mathbb{U} = \mathbb{S}\:.
+generated from the moment scattering matrices:
+\begin{equation}
+  \mathbf{A} = 
+  \begin{bmatrix}
+    (\mathbf{\Sigma}_0) &
+    (-\frac{2}{3}\mathbf{\Sigma}_0) &
+    (\frac{8}{15}\mathbf{\Sigma}_0) &
+    (-\frac{16}{35}\mathbf{\Sigma}_0) \\
+    %%
+    &&&\\
+    %%
+    (-\frac{2}{3}\mathbf{\Sigma}_0) &
+    (\frac{4}{15}\mathbf{\Sigma}_0 + \frac{1}{3}\mathbf{\Sigma}_2) &
+    (-\frac{16}{45}\mathbf{\Sigma}_0 - \frac{4}{9}\mathbf{\Sigma}_2) &
+    (\frac{32}{105}\mathbf{\Sigma}_0 + \frac{8}{21}\mathbf{\Sigma}_2) \\
+    %%
+    &&&\\
+    %%
+    (\frac{8}{15}\mathbf{\Sigma}_0) &
+    (-\frac{16}{45}\mathbf{\Sigma}_0 - \frac{4}{9}\mathbf{\Sigma}_2) &
+    (\frac{64}{225}\mathbf{\Sigma}_0 + \frac{16}{45}\mathbf{\Sigma}_2 + \frac{9}{25}\mathbf{\Sigma}_4) &
+    (-\frac{128}{525}\mathbf{\Sigma}_0 - \frac{32}{105}\mathbf{\Sigma}_2 - \frac{54}{175}\mathbf{\Sigma}_4)
+    \\ 
+    %%
+    &&&\\
+    %%
+    (-\frac{16}{35}\mathbf{\Sigma}_0) &
+    (\frac{32}{105}\mathbf{\Sigma}_0 + \frac{8}{21}\mathbf{\Sigma}_2) &
+    (-\frac{128}{525}\mathbf{\Sigma}_0 - \frac{32}{105}\mathbf{\Sigma}_2 - \frac{54}{175}\mathbf{\Sigma}_4)
+    & 
+    (\frac{256}{1225}\mathbf{\Sigma}_0 + \frac{64}{245}\mathbf{\Sigma}_2 +
+    \frac{324}{1225}\mathbf{\Sigma}_4 + \frac{13}{49}\mathbf{\Sigma}_6)
+  \end{bmatrix}\:.
+  \label{eq:A_block_matrix}
+\end{equation}
+Analogously, for Marshak conditions on the boundaries we have:
+\begin{equation}
+  \hat{\mathbf{n}} \cdot \mathbb{D}_n \nabla \mathbb{U}_n +
+  \sum_{m=1}^4 \mathbb{B}_{nm} \mathbb{U}_m = \mathbb{S}_n\:,
   \label{eq:spn_multigroup_bnd}
 \end{equation}
+with $\mathbb{S}_n$ the vector of group-wise boundary source term on
+each boundary for each psuedo-moment and $\mathbb{B}_{nm}$ is an $N_g
+\times N_g$ diagonal matrix with the value $B_{nm}$ on the
+diagonal. For reflecting conditions, again we have $\nabla
+\mathbb{U}_n = 0$ for all pseudo-moments.
+
+\subsection{A Note on Spatial Discretization and Matrix Symmetry}
+\label{subsec:spatial_discretization}
+To this point, the formulation of the multigroup $SP_N$ equations
+presented have discretized the transport equation in angle and energy
+but have yet to consider the spatial component of phase space. We
+don't explicitly consider it in this document but instead we will
+briefly comment on possible means of discretization. Of primary
+interest here is the discretization of the diffusive term in
+Eq~(\ref{eq:spn_multigroup_system}) and the convective term on the
+boundaries in Eq~(\ref{eq:spn_multigroup_bnd}). Many popular choices
+for spatial discretization are available here and include finite
+difference and finite element formulations. In the Denovo package at
+ORNL, Evans has implemented a finite volume scheme for these
+equations. Although arbitrary grids could be handled effectively
+through a finite element scheme, the rectilinear grid used in Denovo
+is easily discretized through the finite volume method in a
+conservative form. This conservative form is ideal for the
+diffusion-like equations in the domain given by
+Eq~(\ref{eq:spn_multigroup_system}) and the effective boundary current
+conditions given by Eq~(\ref{eq:spn_multigroup_bnd}) as the neutron
+current is balanced from cell-to-cell, continuity of the flux is
+preserved across cell/material boundaries, and boundary conditions are
+naturally represented through cell-face currents. Furthermore, this
+spatial discretization is a natural extension of the balance
+principles used to arrive at the general transport equation. The
+result of this finite volume discretization is one that is symmetric
+for all equations in the domain with respect to the spatial components
+of the solution.
+
+Given the symmetry of the spatial discretization equations in the
+domain given by monoenergetic Eq~(\ref{eq:spn_matrix}) form a
+symmetric linear operator acting on the pseudo-moment vector to give:
+\begin{equation}
+  \mathbf{L}\mathbf{u}=\mathbf{Q}\:,
+  \label{eq:matrix_system}
+\end{equation}
+where $\mathbf{L} = -\nabla \cdot \mathbf{D} \nabla +
+\mathbf{A}$. Moving to the multigroup formulation in
+Eq~(\ref{eq:spn_multigroup_system}) we have:
+\begin{equation}
+  \mathbb{L}\mathbb{U}=\mathbb{Q}\:,
+  \label{eq:multigroup_matrix_system}
+\end{equation}
+where $\mathbb{L} = -\nabla \cdot \mathbb{D} \nabla +
+\mathbb{A}$. Whether or not the resulting matrix is symmetric is
+determined by the group scattering matrices, $\sigma_{sn}^{gg'}$, for
+each of the moments. In general, the cross sections in these matrices
+do not form a symmetric matrix and therefore the resulting linear
+operator in Eq~(\ref{eq:multigroup_matrix_system}) will always be
+asymmetric for multigroup problems. Solutions to this linear system
+will then require techniques specifically formulated for asymmetric
+systems.
 
@@
 \label{sec:spectral_analysis}
+Now that we have an understanding of the linear system generated by
+the $SP_N$ approximation to the neutron transport equation and the
+various parameters in the system that may be adjusted, we can consider
+effective means of solution. As the solution is asymmetric, modern iterative
+solvers should be considered for this task. The performance of modern
+subspace methods are bound to various properties of the eigenvalue
+spectrum of the linear operator while methods based on stationary
+iterations have performance limits imposed by the spectral radius of
+the sytem. In this section we will review the important spectral
+properties to study for common iterative methods for asymmetric
+systems. Based on these properties, we will then perform a
+parameter-based spectral analysis for the linear operator generated by
+the multigroup $SP_N$ equations.
+
+\subsection{Iterative Solvers for Asymmetric Linear Systems}
+\label{subsec:asymmetric_iterative_solvers}
+The performance of different iterative solvers can depend on different
+properties of the linear system. As a means to study these
+differences, we use a 1992 paper by Nachtigal and colleagues
+\cite{nachtigal_how_1992} the performs and in-depth analysis of 3
+Krylov subspace methods that are still used today in modern physics
+applications: conjugate gradient iteration for the normal equations
+(CGN), generalized minimum residual iteration (GMRES), and a
+biconjugate gradient-based methd (CGS).
 

### 2013-03-04 doc/spn/spn.tex
adding stationary methods brief

@@
 various parameters in the system that may be adjusted, we can consider
-effective means of solution. As the solution is asymmetric, modern iterative
-solvers should be considered for this task. The performance of modern
-subspace methods are bound to various properties of the eigenvalue
-spectrum of the linear operator while methods based on stationary
-iterations have performance limits imposed by the spectral radius of
-the sytem. In this section we will review the important spectral
-properties to study for common iterative methods for asymmetric
-systems. Based on these properties, we will then perform a
+effective means of solution. As the solution is asymmetric, modern
+iterative solvers should be considered for this task. The performance
+of modern subspace methods are bound to various properties of the
+eigenvalue spectrum of the linear operator while methods based on
+stationary iterations have performance limits imposed by the spectral
+radius of the sytem. In this section we will review the important
+spectral properties to study for common iterative methods for
+asymmetric systems. Based on these properties, we will then perform a
 parameter-based spectral analysis for the linear operator generated by
@@
 
-\subsection{Iterative Solvers for Asymmetric Linear Systems}
-\label{subsec:asymmetric_iterative_solvers}
-The performance of different iterative solvers can depend on different
-properties of the linear system. As a means to study these
+\subsection{Stationary Methods for Asymmetric Linear Systems}
+\label{subsec:stationary_solvers}
+Stationary methods for linear systems arise from splitting the
+operator in Eq~(\ref{eq:multigroup_matrix_system})
+\cite{leveque_finite_2007}:
+\begin{equation}
+  \mathbb{L} = \mathbf{M} - \mathbf{N}\:,
+  \label{eq:split_linear_operator}
+\end{equation}
+where the choice of $\mathbf{M}$ and $\mathbf{N}$ will be dictated by
+the particular method chosen. Using this split definition of the
+operator we can then write:
+\begin{equation}
+  \mathbf{M}\mathbb{U} - \mathbf{N}\mathbb{U} = \mathbb{Q}\:.
+  \label{eq:linear_split_equation1}
+\end{equation}
+By rearranging, we can generate a form more useful for analysis:
+\begin{equation}
+  \mathbb{U} = \mathbf{H}\mathbb{U} + \mathbf{c}\:,
+  \label{eq:linear_split_equation2}
+\end{equation}
+where $\mathbf{H}=\mathbf{M}^{-1}\mathbf{N}$ is defined as the
+\textit{iteration matrix} and
+$\mathbf{c}=\mathbf{M}^{-1}\mathbb{Q}$. With the solution vector on
+both the left and right hand sides, an iterative method can then be
+formed:
+\begin{equation}
+    \mathbb{U}^{k+1} = \mathbf{H}\mathbb{U}^k + \mathbf{c}\:,
+  \label{eq:linear_iterative_method}
+\end{equation}
+with $k \in \mathbf{Z}^+$ defined as the \textit{iteration
+  index}. Defining $\mathbf{e}^k = \mathbf{u}^k - \mathbf{u}$ as the
+solution error at the $k^{th}$ iterate, we can subtract
+Eq~(\ref{eq:linear_split_equation2}) from
+Eq~(\ref{eq:linear_iterative_method}):
+\begin{equation}
+  \mathbf{e}^{k+1} = \mathbf{H}\mathbf{e}^k\:.
+  \label{eq:linear_iterative_error}
+\end{equation}
+Our error after $k$ iterations is then:
+\begin{equation}
+  \mathbf{e}^{k} = \mathbf{H}^k\mathbf{e}^0\:.
+  \label{eq:linear_k_iter_error}
+\end{equation}
+In other words, successive application of the iteration matrix is the
+mechanism driving down the error in a stationary method. By assuming
+$\mathbf{H}$ is diagonalizable \cite{saad_iterative_2003}, we then
+have:
+\begin{equation}
+  \mathbf{e}^{k} =
+  \mathbf{R}\boldsymbol{\Lambda}^k\mathbf{R}^{-1}\mathbf{e}^0\:,
+  \label{eq:linear_k_iter_error_diag}
+\end{equation}
+where $\boldsymbol{\Lambda}$ contains the Eigenvalues of $\mathbf{H}$
+on its diagonal and the columns of $\mathbf{R}$ contain the
+Eigenvectors of $\mathbf{H}$. Computing the 2-norm of the above form
+then gives:
+\begin{equation}
+  ||\mathbf{e}^{k}||_2 \leq ||\boldsymbol{\Lambda}^k||_2\ 
+  ||\mathbf{R}||_2\ ||\mathbf{R}^{-1}||_2\ ||\mathbf{e}^0||_2\:,
+  \label{eq:linear_k_iter_norm1}
+\end{equation}
+which gives:
+\begin{equation}
+  ||\mathbf{e}^{k}||_2 \leq \rho(\mathbf{H})^k \kappa(\mathbf{R})
+  ||\mathbf{e}^0||_2\:.
+  \label{eq:linear_k_iter_norm2}
+\end{equation}
+For iteration matrices where the Eigenvectors are orthogonal,
+$\kappa(\mathbf{R})=1$ and the error bound reduces to:
+\begin{equation}
+  ||\mathbf{e}^{k}||_2 \leq \rho(\mathbf{H})^k
+  ||\mathbf{e}^0||_2\:.
+  \label{eq:linear_k_iter_norm3}
+\end{equation}
+We can now restrict $\mathbf{H}$ by asserting that $\rho(\mathbf{H}) <
+1$ for a stationary method to converge such that $k$ applications of
+the iteration matrix will not cause the error to grow in
+Eq~(\ref{eq:linear_k_iter_norm3}). In addition, the performance of the
+stationary method is dictated by $\rho(\mathbf{H})$ such that
+performance of the method decreases as it approaches to unity.
+
+\subsection{Subspace Methods for Asymmetric Linear Systems}
+\label{subsec:subspace_solvers}
+The performance of different Krylov subspace solvers can depend on
+different properties of the linear system. As a means to study these
 differences, we use a 1992 paper by Nachtigal and colleagues
-\cite{nachtigal_how_1992} the performs and in-depth analysis of 3
+\cite{nachtigal_how_1992} that performs and in-depth analysis of 3
 Krylov subspace methods that are still used today in modern physics
@@
 (CGN), generalized minimum residual iteration (GMRES), and a
-biconjugate gradient-based methd (CGS).
+biconjugate gradient-based method (CGS).
 

### 2013-03-04 doc/spn/spn.tex
working on spn spectral analysis plan

@@
 various parameters in the system that may be adjusted, we can consider
-effective means of solution. As the solution is asymmetric, modern
+various means of solution. As the solution is asymmetric, modern
 iterative solvers should be considered for this task. The performance
-of modern subspace methods are bound to various properties of the
+of Krylov subspace methods are bound to various properties of the
 eigenvalue spectrum of the linear operator while methods based on
@@
 parameter-based spectral analysis for the linear operator generated by
-the multigroup $SP_N$ equations.
-
-\subsection{Stationary Methods for Asymmetric Linear Systems}
+the multigroup $SP_N$ equations. Preconditioning strategies for these
+methods are not considered. Krylov subspace method restarts and
+truncation will also not be considered.
+
+\subsection{Stationary Methods for Linear Systems}
 \label{subsec:stationary_solvers}
@@
 
-\subsection{Subspace Methods for Asymmetric Linear Systems}
+\subsection{Krylov Subspace Methods for Linear Systems}
 \label{subsec:subspace_solvers}
@@
 differences, we use a 1992 paper by Nachtigal and colleagues
-\cite{nachtigal_how_1992} that performs and in-depth analysis of 3
+\cite{nachtigal_how_1992} that performs and in-depth analysis of three
 Krylov subspace methods that are still used today in modern physics
@@
 (CGN), generalized minimum residual iteration (GMRES), and a
-biconjugate gradient-based method (CGS).
+biconjugate gradient-based method (CGS). For our work, we will
+consider only their analysis of GMRES as it currently leveraged in
+Denovo for $SP_N$ solutions and as it also applies to CGS for most
+matrices.
+
+For GMRES, at each iteration we are extracting a correction for the
+solution from the Krylov subspace:
+\begin{equation}
+  \mathbb{U}^k \in \mathbb{U}^0 + \langle \mathbf{r_0}, \mathbb{L}
+  \mathbf{r_0}, \cdots, \mathbb{L}^{k-1} \mathbf{r_0} \rangle\:,
+  \label{eq:gmres_iteration}
+\end{equation}
+where the extraction is constrained by ensuring orthogonality of the new
+residual with the Krylov space:
+\begin{equation}
+  \mathbf{r}^k \bot \langle \mathbb{L} \mathbf{r_0}, \mathbb{L}^2
+  \mathbf{r_0}, \cdots, \mathbb{L}^{k} \mathbf{r_0} \rangle\:,
+  \label{eq:gmres_constraint}
+\end{equation}
+with the residual at the $k^{th}$ iteration defined as:
+\begin{equation}
+  \mathbf{r}^k = \mathbb{Q} - \mathbb{L}\mathbb{U}^k\:.
+  \label{eq:residual}
+\end{equation}
+From \cite{nachtigal_how_1992}, we then have the following error at
+each iteration:
+\begin{equation}
+  \mathbf{e}^k = p_k(\mathbb{L})\mathbf{e}^0\:,
+  \label{eq:gmres_iteration_error}
+\end{equation}
+where $p_k(\mathbb{L})$ is an arbitrary $k^{th}$ degree polynomial and
+$p_k(0)=1$. This is a very similar form for the iteration error as
+observed for stationary methods in Eq~(\ref{eq:linear_k_iter_norm3})
+where now as these $p_k$ polynomials are constructed as the iterations
+progress, a rapid decrease in $||p_k(\mathbb{L})||$ means a rapid rate
+of convergence. For matrices $\mathbb{L}$ that are approximately
+normal (which was exactly the assumption we made to arrive at
+Eq~(\ref{eq:linear_k_iter_norm3})), then we can write the polynomial
+norm as:
+\begin{equation}
+  ||p_k(\mathbb{L})|| = \sup_{z\in\Lambda(\mathbb{L})}|p_k(z)|\:,
+  \label{eq:gmres_polynomial}
+\end{equation}
+where $\Lambda(\mathbb{L})$ is the eigenvalue spectrum of the normal
+matrix $\mathbb{L}$. So if we could obtain the polynomials, the
+largest eigenvalue of $\mathbb{L}$ could be used to calculate
+convergence rates. Next we consider a choice of polynomials to use for
+the convergence rate approximation. They should of course be
+orthogonal to reflect the orthogonal projection mechanism by which the
+residual is reduced. In Nachtigal's work, Chebyshev polynomials were
+used and defined on the interval of Eigenvalues,
+$[\lambda_{min},\lambda_{max}]$, and then normalized via the
+constraint $p_k(0)=1$. Thus the rate of convergence for these
+polynomials is an approximation to the rate of convergence of the
+method.
+
+We can check the validity of the normal
+approximation by computing the condition number of $\mathbb{L}$:
+\begin{equation}
+  \kappa(\mathbb{L}) = ||\mathbb{L}||\ ||\mathbb{L}^{-1}||
+  \label{eq:operator_condition_number}
+\end{equation}
+The closer $\kappa(\mathbb{L})$ is to one, the more normal the matrix
+and the better the above approximation. The same check can also be
+performed to verify the assumption used to generate
+Eq~(\ref{eq:linear_k_iter_norm3}) where $\kappa(\mathbf{H})$ can be
+computed. For our analysis, we will compute $\kappa_2$ where the
+matrix norms are induced by the 2-norm of a vector such that:
+\begin{equation}
+  ||\mathbb{L}||_2 = \sqrt{\rho(\mathbb{L}^T\mathbb{L})}\:.
+  \label{eq:matrix_2_norm}
+\end{equation}
+
+\subsection{Organization of the Spectral Analysis}
+\label{subsec:spectral_organization}
+For the spectral analysis we will perform a numerical parameter study
+for the multigroup $SP_N$ equations. Based on the above information
+for iterative methods, several key quantities should be computed for
+the $SP_N$ system:
+\begin{itemize}
+\item Eigenvalue spectrum for the linear operator $\mathbb{L}$
+\item Largest eigenvalue for the linear operator $\mathbb{L}$
+\item Smallest eigenvalue for the linear operator $\mathbb{L}$
+\item Condition number for the linear operator $\mathbb{L}$
+\item For each stationary method:
+  \begin{itemize}
+  \item Eigenvalue spectrum for the iteration matrix $\mathbf{H}$
+  \item Largest eigenvalue for the iteration matrix $\mathbf{H}$
+  \item Smallest eigenvalue for the iteration matrix $\mathbf{H}$
+  \item Condition number for the iteration matrix $\mathbf{H}$
+  \end{itemize}
+\item For GMRES:
+  \begin{itemize}
+  \item Chebyshev polynomial norm using eigenvalues of $\mathbb{L}$
+    for a fixed number of iterations
+  \end{itemize}
+\end{itemize}
+In addition, the following parameters may be varied in the $SP_N$
+system in Denovo:
+\begin{itemize}
+  \item $SP_N$ order
+  \item Number of energy groups $N_g$
+  \item Total cross sections $\sigma^g$
+  \item Scattering cross sections $\sigma_{sn}^{gg'}$
+  \item Boundary conditions (Marshak or reflecting)
+  \item Finite volume mesh size ($\Delta$ in all directions)
+  \item Number of mesh elements
+  \item Source of neutrons (Uniform or Point)
+\end{itemize}
+We will restrict the problem description in Denovo to one similar
+Brantley and Larsen's $SP_N$ work where the spectral radius was
+computed for convergence analysis a several proposed solution
+methods. From this definition we choose a single material problem with
+a uniform source of unity. For the boundary conditions, we choose
+reflecting conditions on all six sides of the cubic domain for an
+effectively homogenous infinite medium problem which we will mesh with
+a $5\times 5\times 5$ grid. In Brantley and Larsen's work, however, a
+single energy group was used an therefore we may not use their cross
+sections in our computations.
 

### 2013-03-04 doc/spn/spn.tex
adding more detail to spn spectral analysis description

@@
   \item Chebyshev polynomial norm using eigenvalues of $\mathbb{L}$
-    for a fixed number of iterations
+    for a fixed number of iterations determined by stationary method
+    convergence
   \end{itemize}
@@
 \end{itemize}
-We will restrict the problem description in Denovo to one similar
+We will restrict the problem description in Denovo to one similar to
 Brantley and Larsen's $SP_N$ work where the spectral radius was
@@
 effectively homogenous infinite medium problem which we will mesh with
-a $5\times 5\times 5$ grid. In Brantley and Larsen's work, however, a
-single energy group was used an therefore we may not use their cross
-sections in our computations.
+a $5\times 5\times 5$ grid with $\Delta=0.1$. We are then left to vary
+the angular and energy compoenents of the system in our parameter
+studies: the $SP_N$ order, the number of energy groups, and their
+cross sections. In Brantley and Larsen's work, a single energy group
+was used and therefore we may not use their cross sections in our
+computations.
 

### 2013-03-05 doc/spn/spn.tex
updated spectral analysis plan
finished draft of spectral analysis plan

@@
 \begin{itemize}
-  \item $SP_N$ order
+  \item $SP_N$ (angular flux expansion) order
+  \item $P_N$ (scattering cross section expansion) order
   \item Number of energy groups $N_g$
   \item Total cross sections $\sigma^g$
-  \item Scattering cross sections $\sigma_{sn}^{gg'}$
+  \item Scattering cross section matrix $\sigma_{sn}^{gg'}$
   \item Boundary conditions (Marshak or reflecting)
@@
 Brantley and Larsen's $SP_N$ work where the spectral radius was
-computed for convergence analysis a several proposed solution
+computed for convergence analysis of several proposed solution
 methods. From this definition we choose a single material problem with
-a uniform source of unity. For the boundary conditions, we choose
-reflecting conditions on all six sides of the cubic domain for an
-effectively homogenous infinite medium problem which we will mesh with
-a $5\times 5\times 5$ grid with $\Delta=0.1$. We are then left to vary
-the angular and energy compoenents of the system in our parameter
-studies: the $SP_N$ order, the number of energy groups, and their
-cross sections. In Brantley and Larsen's work, a single energy group
-was used and therefore we may not use their cross sections in our
-computations.
+a uniform source of unity in all energy groups. For the boundary
+conditions, we choose reflecting conditions on all six sides of the
+cubic domain for an effectively homogenous infinite medium problem
+which we will mesh with a $5\times 5\times 5$ grid with
+$\Delta=0.1$. We are then left to vary the angular and energy
+components of the system in our parameter studies: the $SP_N$ order,
+the number of energy groups, and their cross sections of order
+$P_N$. In Brantley and Larsen's work, a single energy group was used
+and therefore we may not use their cross sections in our multigroup
+computations. For our work, we will always choose a total cross
+section of unity for all groups and all flux moments. For the
+scattering matrix, two classes will be chosen: one with upscatter and
+one without. For the downscatter only case, all groups may scatter
+within the group and to all groups of lower energy (yielding a lower
+triangular scattering matrix) with a cross section of 0.5. For the
+upscatter case, all groups may scatter within the group and to all
+groups of lower energy given a cross section of 0.5 and for upscatter,
+all groups may scatter up to two groups higher in energy with a cross
+section of 0.25. All cross section matrices will be equivalent for
+each scattering moment. For these cross sections, the $P_N$ order will
+be varied with $N=0,1,3$. Both 1 and 10 energy groups will be used
+with the 10 energy group case using the scattering matrices specified
+above. For the angular flux, the $SP_N$ order will be varied with
+$N=1,3,5,7$. This gives a total of 36 linear system variations for the
+spectral analysis.
+
+\subsection{Spectral Analysis Results}
+\label{subsec:analysis_results}
 

### 2013-03-06 doc/kaczmarz/python/kaczmarz.py
adding kaczmarz projection script and mc python script

@@ -1,0 +1,48 @@
+import numpy
+import math
+import random
+
+m = 3
+
+# Kaczmarz projection method
+def kaczmarz(A,x,b):
+    r = b - A*x
+    num_iters = 0
+    while numpy.linalg.norm(r,numpy.inf)/numpy.linalg.norm(b,numpy.inf) > 1.0e-12:
+        for i in xrange(m):
+            j = m-i-1
+            alpha = numpy.asscalar( numpy.dot(A[j],A[j].T) )
+            scalarb = numpy.asscalar(b[j])
+            scalarAjdotx = numpy.asscalar( numpy.dot(A[j],x) )
+            x = x + (1/alpha)*(scalarb - scalarAjdotx)*(A.T)[j].T
+        r = b - A*x
+        num_iters += 1
+        print r
+    print "Kaczmarz iterations to converge:", num_iters
+    return x
+
+# Solve a 3x3 linear system Ax=b with initial state x_i = 0
+A = numpy.matrix([ [1.0, -0.3, -0.4],[-0.2, 1.0, -0.4],[-0.5, -0.4, 1.0] ])
+b = numpy.matrix([ [10.0], [1.0], [1.0] ])
+x_kaczmarz = numpy.matrix([ [0.0], [0.0], [0.0] ])
+x_adjoint = numpy.matrix([ [0.0], [0.0], [0.0] ])
+
+# Spectral Radius of A
+eigenvalues_A, eigenvectors_A = numpy.linalg.eig(A)
+rho_A = max(eigenvalues_A)
+print 
+print "Spectral radius A =", rho_A
+print
+
+# Kaczmarz solution
+x_kaczmarz = kaczmarz(A,x_kaczmarz,b)
+print
+print "Kaczmarz"
+print x_kaczmarz
+
+# Reference solution
+x_numpy = numpy.linalg.solve( A, b )
+print
+print "Reference"
+print x_numpy
+

### 2013-03-06 doc/kaczmarz/python/mcsa_sketch.py

@@ -1,0 +1,411 @@
+import numpy
+import math
+import random
+import time
+
+##---------------------------------------------------------------------------##
+# compute the inf norm of a vector
+def inf_norm(x):
+    m = len(x)
+    error = []
+    for i in xrange(m):
+        error.append( x[i][0] )
+
+    return max(error)
+
+##---------------------------------------------------------------------------##
+# Sampling function for stochastic matrices ( O(n) algorithm )
+def sample(P,s):
+    cdf = 0.0
+    zeta = random.random()
+    inverted = 0
+    for i in xrange( len(P[s]) ):
+        cdf += P[s][i]
+        if zeta < cdf and not inverted:
+            ns = i
+            inverted = 1
+
+    return int(ns)
+
+##---------------------------------------------------------------------------##
+# Direct Monte Carlo solution method
+def direct_mc(np,wc,H,P,b,x):
+    m = len(x)
+    for i in xrange(m):
+        for n in xrange(np):
+            s = i
+            walk = 1
+            w = 1.0
+            while walk:
+                x[i][0] += w*b[s][0]
+                ns = sample(P,s)
+                if P[s][ns] == 0:
+                    w = 0
+                else:
+                    w *= H[s][ns]/P[s][ns]
+                if w < wc:
+                    walk = 0
+                s = ns
+            
+        x[i][0] /= np
+
+##---------------------------------------------------------------------------##
+# Adjoint Monte Carlo solution method.
+def adjoint_mc(np,wc,H,Q,b,x):
+    m = len(x)
+
+    b_norm = 0.0
+    for j in xrange(m):
+        b_norm += b[j][0]
+
+    for n in xrange(np):
+        init = 0
+        b_cdf = 0
+        zeta = random.random()
+        inverted = 0
+        for j in xrange(m):
+            b_cdf += b[j][0]/b_norm
+            if zeta < b_cdf and not inverted:
+                init = j
+                inverted = 1
+        walk = 1
[…]

### 2013-03-12 doc/mandc2013/theory_paper/slattery_wilson_evans.tex
fixed typos from reviewer 1 in mandc paper

@@
 a new state, $j$, from the current state, $i_m$, by sampling the
-$i_m^{th}$ row of the probablity matrix as a discrete distribution,
+$i_m^{th}$ row of the probability matrix as a discrete distribution,
 and updating the weight, $W_m$, accordingly.  Using our result from
@@
 \end{equation}
-If our our chord is of length $l$ and there are $n_i$ grid elements
-(or states to which a history may transition) along that chord, then
-$h = l / n_i$ giving:
+If our chord is of length $l$ and there are $n_i$ grid elements (or
+states to which a history may transition) along that chord, then $h =
+l / n_i$ giving:
 \begin{equation}

### 2013-03-12 doc/mandc2013/theory_paper/slattery_wilson_evans.tex
updates to mandc

@@
 Carlo method followed by a presentation of the model problem. Using
-the linear system generated by the discretization of this, we use a
-spectral analysis to generate analytic relations for the eigenvalues
-of the operator based on system parameters. Using the eigenvalue
-spectra, we then build relationships to characterize the transport of
-stochastic histories in a decomposed domain and the fraction of
-histories that leak from a domain and will therefore have to be
-communicated. Finally, we compare these analytic results to numerical
-experiments conducted with the model problem and draw conclusions
-looking towards future work.
+the linear system generated by the discretization of the model
+problem, we use a spectral analysis to generate analytic relations for
+the eigenvalues of the operator based on system parameters. Using the
+eigenvalue spectra, we then build relationships to characterize the
+transport of stochastic histories in a decomposed domain and the
+fraction of histories that leak from a domain and will therefore have
+to be communicated. Finally, we compare these analytic results to
+numerical experiments conducted with the model problem and draw
+conclusions looking towards future work.
 

### 2013-03-12 doc/spn/data/spn.py
adding mcls xml input file for spn solve

@@ -11,9 +11,9 @@ initialize(sys.argv)
 ##---------------------------------------------------------------------------##
 # SPECTRAL ANALYSIS PARAMETERS
 ##---------------------------------------------------------------------------##
-pn_order = 7
-spn_order = 7
-num_groups = 10
+pn_order = 3
+spn_order = 5
+num_groups = 5
 upscatter = True
 
 sigma_t = 5.0
@@ -40,6 +40,8 @@ entries = {
     }
 
 db = DB.from_dict(entries)
+
+#db.insert("linear_solver_xml_file","mcls.xml")
 
 ## DECOMPOSITION
 
@@ -145,6 +147,7 @@ for g in xrange(num_groups):
 ##---------------------------------------------------------------------------##
 
 manager.write_Matlab()
+manager.solve(source)
 
 ##---------------------------------------------------------------------------##
 

### 2013-03-12 doc/spn/data/spn.py
working on spn solves using mcls through the python interface

@@ -12,8 +12,8 @@ initialize(sys.argv)
 # SPECTRAL ANALYSIS PARAMETERS
 ##---------------------------------------------------------------------------##
 pn_order = 3
-spn_order = 5
-num_groups = 5
+spn_order = 1
+num_groups = 1
 upscatter = True
 
 sigma_t = 5.0
@@ -28,9 +28,9 @@ entries = {
     "delta_x"     : 1.0,
     "delta_y"     : 1.0,
     "delta_z"     : 1.0,
-    "num_cells_i" : 4,
-    "num_cells_j" : 4,
-    "num_cells_k" : 4,
+    "num_cells_i" : 5,
+    "num_cells_j" : 5,
+    "num_cells_k" : 5,
     "num_groups"  : num_groups,
     "SPN_order"   : spn_order,
     "boundary"    : "reflect",
@@ -41,7 +41,7 @@ entries = {
 
 db = DB.from_dict(entries)
 
-#db.insert("linear_solver_xml_file","mcls.xml")
+db.insert("linear_solver_xml_file","mcls.xml")
 
 ## DECOMPOSITION
 

### 2013-03-20 doc/spn/testing/spn.py
updated meeting slides

@@ -14,7 +14,7 @@ initialize(sys.argv)
 pn_order = 5
 spn_order = 7
 num_groups = 10
-upscatter = True
+upscatter = False
 
 sigma_t = 5.0
 sigma_ing = 0.25

### 2013-03-21 doc/spn/fuel_assembly/sp7/fs_mcls.py
adding mcls files for spn assembly problem

@@ -1,0 +1,580 @@
+###############################################################################
+## fs.py
+## 9te [angband.ornl.gov]
+## Wed Jan 12 10:37:50 2011
+###############################################################################
+## Copyright (C) 2008 Oak Ridge National Laboratory, UT-Battelle, LLC.
+##---------------------------------------------------------------------------##
+## generated by /data/denovo/production/head/setup/bin/pygen built on 20110112
+###############################################################################
+
+import os, sys, math, string
+
+# pyspn equation type
+from spn_fv import *
+
+print_it = False
+
+##---------------------------------------------------------------------------##
+## MAIN
+##---------------------------------------------------------------------------##
+
+initialize(sys.argv)
+
+if node() == 0:
+    print "Denovo - pyspn Python Front-End"
+    print "-------------------------------"
+    print "Release      : %16s" % (release())
+    print "Release Date : %16s" % (release_date())
+    print "Build Date   : %16s" % (build_date())
+    print
+
+timer = Timer()
+timer.start()
+
+##---------------------------------------------------------------------------##
+## XS DATA
+
+####### UO2 Fuel-Clad Macroscopic Cross Sections ##########
+## Transport-corrected Total Cross Sections
+T_UO2 = [0.0, 0.0, 0.0, 0.0, 0.0, 0.0, 0.0]
+T_UO2[0] = 1.77949e-1
+T_UO2[1] = 3.29805e-1
+T_UO2[2] = 4.80388e-1
+T_UO2[3] = 5.54367e-1
+T_UO2[4] = 3.11801e-1
+T_UO2[5] = 3.95168e-1
+T_UO2[6] = 5.64406e-1
+
+## Fission Cross Section
+F_UO2 = [0.0, 0.0, 0.0, 0.0, 0.0, 0.0, 0.0]
+F_UO2[0] = 7.21206e-3
+F_UO2[1] = 8.19301e-4
[…]

### 2013-03-21 doc/dissertation/document/chapters/introduction/introduction.tex
adding dissertation document. might as well start now

--- 2012-10-24 doc/prelim/document/chapters/introduction/introduction.tex
@@
 numbers of cores will be pertinent to robust algorithms aimed at this
-new hardware. Considering these issues, this preliminary report
-proposes the development of a massively parallel Monte Carlo method
-for linear problems and a novel Monte Carlo method to advance solution
-techniques for nonlinear problems.
+new hardware. Considering these issues, this dissertation develops a
+massively parallel Monte Carlo method for linear problems and a novel
+Monte Carlo method to advance solution techniques for nonlinear
+problems.
 
@@
 components to be implemented in CASL for multiphysics analysis, such
-as the Denovo radiation transport code developed at Oak Ridge National
-Laboratory \citep{evans_denovo:_2010}, compute trillions of unknowns
-for full core reactor analysis on $O(\sn{1}{9})$ element meshes and
-$O(100,000)$ cores as well. Given the large scale and complexity of
-these problems, if we aim to advance multiphysics solution techniques,
-then we are motivated to advance the solution of complex and general
-nonlinear problems exploiting leadership class levels of parallelism.
+as the Exnihilo radiation transport suite developed at Oak Ridge
+National Laboratory \citep{evans_denovo:_2010}, compute trillions of
+unknowns for full core reactor analysis on $O(\sn{1}{9})$ element
+meshes and $O(100,000)$ cores as well. Given the large scale and
+complexity of these problems, if we aim to advance multiphysics
+solution techniques, then we are motivated to advance the solution of
+complex and general nonlinear problems exploiting leadership class
+levels of parallelism.
 
@@
 \citep{knoll_newton-krylov_1995}. To address these issues, this thesis
-proposes a new and novel method for nonlinear systems based on the
+develops a new and novel method for nonlinear systems based on the
 MCSA method.
 
-The Forward-Automated Newton-MCSA (FANM) method is proposed as new
+The Forward-Automated Newton-MCSA (FANM) method is developed as new
 nonlinear solution method. The key features of FANM are: full Jacobian
@@
 improvements to the multiphysics analysis community. From these
-stochastic methods, a new nonlinear method is proposed in
+stochastic methods, a new nonlinear method is developed in
 Chapter~\ref{ch:nonlinear_problem} and compared to conventional
-methods for solving nonlinear problems. The research proposal for this
-work is presented in Chapter~\ref{ch:research_proposal} with progress
-to date reported including the development of a new finite
-element-based physics framework that leverages a preliminary general
-implementation of the stochastic method that will serve as a test bed
-for these new methods. An outline of a set of numerical experiments
-will then be provided that continues the development and
-implementation of these new methods and verifies them with benchmark
-solutions, culminating in the analysis of a multiphysics simulation of
-a pressurized water reactor subsystem using a production code system.
+methods for solving nonlinear problems.

### 2013-03-21 doc/dissertation/document/chapters/linear_problem/linear_problem.tex
adding dissertation document. might as well start now

--- 2012-10-16 doc/prelim/document/chapters/linear_problem/linear_problem.tex
@@
 the case of a weak scaling exercise, we would instead expect this
-percentage to exhibit a more desirable behavaior of remaining constant
+percentage to exhibit a more desirable behavior of remaining constant
 for gather/scatter operations as problem size would be scaled with the

### 2013-03-21 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
adding dissertation document. might as well start now

--- 2012-10-24 doc/prelim/document/chapters/nonlinear_problem/nonlinear_problem.tex
@@
 certain number of vectors have been generated. The optimization
-condition is maintain over that particular set of vectors, however,
+condition is maintained over that particular set of vectors, however,
 Knoll and McHugh note that this ultimately slows the convergence rate

### 2013-03-21 doc/dissertation/document/frontmatter/abstract.tex
adding dissertation document. might as well start now

--- 2012-08-21 doc/prelim/document/frontmatter/abstract.tex
@@
 \vcinfo{}
-
-In many fields of engineering and physics, nonlinear problems are a
-primary focus of study. Recent focus on multiple physics systems (and
-in particular nuclear systems) adds a new level of complexity to
-common nonlinear systems as solution strategies change when they are
-coupled to other problems. Furthermore, a desire for predictive
-simulations to enhance the safety and performance of engineered
-systems creates a need for extremely high fidelity computations to be
-performed for these coupled systems as a means to capture effects not
-modeled by coarser methods. In order to achieve this high fidelity,
-state-of-the-art computing facilities must be leveraged in a way that
-is both efficient and considerate of hardware-related issues. As
-scientific computing moves towards exascale facilities with machines
-of $O(1,000,000)$ cores already coming online, new algorithms to solve
-these complex problems must be developed to leverage this new
-hardware. Issues such as resiliency to node failure and scaling to
-large numbers of cores will be pertinent to robust algorithms aimed at
-this new hardware. Considering these issues, this thesis proposes a
-novel stochastic method to advance solution techniques for nonlinear
-problems with a focus on coupled physics systems.
-
-For some time, the particle transport community has been utilizing
-Monte Carlo methods for the solution of transport problems
-\cite{lewis_1993}. The partial differential equation (PDE) community
-has focused on various deterministic methods for solutions to linear
-problems \cite{saad_2003}. In between these two areas are a not widely
-known small group of stochastic methods for solving sparse linear
-systems \cite{hammersley_1964, halton_1962, halton_1994}. In recent
-years, these methods have been further developed for transport
-problems in the form of Monte Carlo Synthetic-Acceleration (MCSA)
-\cite{evans_2003, evans_2009} but have yet to be applied to more
-general sparse linear systems. Compared to other methods in these
-regime, MCSA offers three attractive qualities; (1) the linear problem
-operator need not be symmetric or positive-definite, thereby reducing
-preconditioning complexity, (2) the stochastic nature of the
-solution method provides a natural solution to the issue of resiliency,
-and (3) is amenable to parallelization using modern methods developed by
-the transport community \cite{wagner_2011}. The developement of MCSA
-as a general linear solver and the development of a parallel MCSA
-method will be new and unique features of this work. Resiliency will
-not be addressed.
-
-In addition to linear solver advancements, nonlinear solvers may also
-benefit from a general and parallel MCSA scheme. In the engineering
-community, nonlinear problems are often addressed by linearizing the
-problem and using traditionaly iterative or direct methods
-\cite{tannehill_1997}. In the mathematics community, various Newton
-methods have been popular \cite{kelley_1995}. Recently, Jacobian-Free
-Newton-Krylov (JFNK) schemes \cite{knoll_2004} have been utilized in
-multiple physics architectures and advanced single physics codes
-\cite{gaston_2009}. The benefits of JFNK schemes are that the Jacobian
-is never formed, simplifying the implementation, and a Krylov solver
-is leveraged (typically GMRES or Conjugate Gradient), providing
-excellent convergence properties for well-conditioned and well-scaled
-systems. However, there are two potential drawbacks to these methods
-for high fidelity predictive simulations: (1) the Jacobian is
-approximated by a first-order differencing method on the order of
-machine precision such that this error can grow beyond that of those
-in a fine-grained system and (2) for systems that are not symmetric
-positive-definite (which will be the case for most multiphysics
-systems and certainly for most preconditioned systems) the Krylov
-subspace generated by the GMRES solver may become prohibitively
-large. To address these issues, this thesis proposes a new and novel
-method for nonlinear systems based on the MCSA method.
-
-The Forward-Automated Newton-MCSA (FANM) method is proposed as new
-nonlinear solution method. The key features of FANM are: full Jacobian
-generation using modern Forward Automated Differencing (FAD) methods
-\cite{bartlett_2006}, and MCSA as the inner linear solver. This method
-has several attractive properties. First, the first-order
-approximation to the Jacobian used in JFNK type methods is eliminated
-by generating the Jacobian explicitly with the model equations through
-FAD. Second, the Jacobian need not be explicitly formed by the user
-but is instead automated through FAD; this eleminates the complexity
-of hand-coding derivatives and has also been demonstrated to be more
-efficient computationally than evaluating differenced derivatives
-\cite{bartlett_2006}. Third, unlike GMRES, MCSA does not build a
-subspace during iterations. Although the Jacobian must be explicitly
-formed to use MCSA, for problems that take more than a few GMRES
-iterations to converge the size of the Krylov subspace will grow
-beyond that of the Jacobian. Finally, using MCSA for the linear solve
-provides its benefits for preconditioning, potential resiliency, and
-parallelism.

### 2013-03-21 doc/dissertation/document/chapters/mc_background/mc_background.tex
spn derivation dump

--- 2012-11-01 doc/prelim/document/chapters/parallel_mc/parallel_mc.tex
@@
     \scalebox{1.5}{
-      \input{chapters/parallel_mc/msod_example.pdftex_t} }
+      \input{chapters/mc_background/msod_example.pdftex_t} }
   \end{center}
@@
     \scalebox{1.5}{
-      \input{chapters/parallel_mc/procassini_example.pdftex_t} }
+      \input{chapters/mc_background/procassini_example.pdftex_t} }
   \end{center}
@@
     \scalebox{1.5}{
-      \input{chapters/parallel_mc/gentile_example.pdftex_t} }
+      \input{chapters/mc_background/gentile_example.pdftex_t} }
   \end{center}

### 2013-03-21 doc/dissertation/document/chapters/navier_stokes_solutions/navier_stokes_solutions.tex
spn derivation dump

\chapter{Monte Carlo Solution Methods for the Navier-Stokes Equations}
\label{ch:mc_ns_solutions}

Put Navier Stokes work here.

### 2013-03-21 doc/dissertation/document/chapters/spn_solutions/spn_solutions.tex
spn derivation dump

--- 2013-03-05 doc/spn/spn.tex
@@
-%        File: spn.tex
-\documentclass[letterpaper,12pt]{article}
-\usepackage[top=1.0in,bottom=1.0in,left=1.25in,right=1.25in]{geometry}
-\usepackage{verbatim}
-\usepackage{amssymb}
-\usepackage{graphicx}
-\usepackage{longtable}
-\usepackage{amsfonts}
-\usepackage{amsmath}
-\usepackage{amsthm}
-\usepackage[usenames]{color}
-\usepackage[
-naturalnames = true, 
-colorlinks = true, 
-linkcolor = black,
-anchorcolor = black,
-citecolor = black,
-menucolor = black,
-urlcolor = blue
-]{hyperref}
-\usepackage{listings}
-\usepackage{textcomp}
-\definecolor{listinggray}{gray}{0.9}
-\definecolor{lbcolor}{rgb}{0.9,0.9,0.9}
-\lstset{
-  backgroundcolor=\color{lbcolor},
-  tabsize=4,
-  rulecolor=,
-  language=c++,
-  basicstyle=\scriptsize,
-  upquote=true,
-  aboveskip={1.5\baselineskip},
-  columns=fixed,
-  showstringspaces=false,
-  extendedchars=true,
-  breaklines=true,
-  prebreak =
-  \raisebox{0ex}[0ex][0ex]{\ensuremath{\hookleftarrow}},
-  frame=single,
-  showtabs=false,
-  showspaces=false,
-  showstringspaces=false,
-  identifierstyle=\ttfamily,
-  keywordstyle=\color[rgb]{0,0,1},
-  commentstyle=\color[rgb]{0.133,0.545,0.133},
-  stringstyle=\color[rgb]{0.627,0.126,0.941},
-}
-
-\author{Stuart R. Slattery
-  \\ \href{mailto:sslattery@wisc.edu}{\texttt{sslattery@wisc.edu}}
-}
-
-\date{\today}
-\title{A Monte Carlo Synthetic Acceleration Method for the $SP_N$ Equations}
-\begin{document}
-\maketitle
-
-%%---------------------------------------------------------------------------%%
-\section{Introduction}
-\label{sec:introduction}
+\chapter{Monte Carlo Solution Methods for the Simplified $P_N$ Equations}
+\label{ch:mc_spn_solutions}
+
 The neutron transport problem is complicated. Solutions cover a large
@@
 \label{sec:monte_carlo}
-
-%%---------------------------------------------------------------------------%%
-\section{Conclusion}
-\label{sec:conclusion}
-
-%%---------------------------------------------------------------------------%%
-\pagebreak
-\bibliographystyle{ieeetr} 
-\bibliography{references}
-\end{document}

### 2013-03-21 doc/dissertation/document/chapters/mc_background/mc_background.tex
added jcomp analysis

@@
 other preconditioning that is applied to the system.
+
+\section{Monte Carlo Method Selection}
+
+The MCSA method defined in Eq.~(\ref{eq:mcsa}) uses the adjoint method
+to estimate the error in residual Monte Carlo solve instead of the
+direct method outlined in \S~\ref{sec:direct_mc}. To demonstrate
+the effectiveness of the adjoint method over the direct method within
+the context of MCSA, we choose the 2D time-dependent Poisson equation
+as a simple model problem:
+\begin{equation}
+  \frac{\partial \ve{u}}{\partial t} = \nabla^2 \ve{u}\:.
+  \label{eq:poisson_equation}
+\end{equation}
+For all comparisons, a single time step is computed with backwards Euler time
+integration. The Laplacian is differenced on a square Cartesian grid with a
+second-order five-point stencil,
+\begin{equation}
+  \nabla^2_5 = \frac{1}{\Delta^2}[u_{i-1,j} + u_{i+1,j} + u_{i,j-1} +
+    u_{i,j+1} - 4 u_{i,j}]\:,
+  \label{eq:five_point_stencil}
+\end{equation}
+and a fourth-order nine-point stencil,
+\begin{multline}
+  \nabla^2_9 = \frac{1}{6\Delta^2}[4 u_{i-1,j} + 4 u_{i+1,j} + 4
+    u_{i,j-1} + 4 u_{i,j+1} + u_{i-1,j-1}\\ + u_{i-1,j+1} +
+    u_{i+1,j-1} + u_{i+1,j+1} - 20 u_{i,j}]\:,
+  \label{eq:nine_point_stencil}
+\end{multline}
+both assuming a grid size of $\Delta$ in both the $i$ and $j$ directions. For
+a single time step solution, we then have the following sparse linear system
+to be solved with the MCSA method:
+\begin{equation}
+  \ve{A} \ve{u}^{n+1} = \ve{u}^n\:.
+  \label{eq:poisson_eq_lin_sys}
+\end{equation}
+Both the stencils will be used to vary the size and density of the sparse
+linear system in Eq.~(\ref{eq:poisson_eq_lin_sys}).
+
+A timing and convergence study is used to demonstrate the
+effectiveness of the adjoint method as compared to the direct
+method. To assess both the CPU time and number of iterations required
+to converge to a solution, a problem of constant $\Delta$ was used
+with varying values of grid size, fixing the spectral radius of the
+system at a constant value for each variation. Both the five-point and
+nine-point stencils were used with both the direct and adjoint
+solvers. For each case, $N \times N$ total random walk permutations
+were computed per MCSA iteration where $N \times N$ is the number of
+discrete grid points in the system. Solver parameters were set to a
+weight cutoff of \sn{1}{-4} for the stochastic linear solver and a
+convergence tolerance of \sn{1}{-8} for the MCSA iterative solver.
+Figure~\ref{fig:poisson_cpu_time} gives the CPU time needed for each
+case to converge in seconds and Figure~\ref{fig:poisson_iterations}
+gives the number of iterations needed for each case to converge to the
+specified tolerance as a function of the problem size. All
+computations presented in this section and
+\S~\ref{subsec:sequential_comparison} were completed on a 3.0 GHz
+Intel Core 2 Quad Q9650 CPU machine with 16 GB 1067 MHz DDR3 memory.
+\begin{figure}[ht!]
+  \centering
+  \includegraphics[width=5in,clip]{chapters/mc_background/dir_adj_cpu.pdf}
+  \caption{\textbf{CPU Time (s) to converge vs. Problem Size ($N$ for
+      an $N \times N$ square mesh).} \textit{Both the adjoint and
+      direct solvers are used with the five point and nine point
+      stencils. A CPU time speedup is noted with the adjoint method
+      due to the higher density of random walk events in regions with
+      a large residual.}}
+  \label{fig:poisson_cpu_time}
+\end{figure}
+
+\begin{figure}[ht!]
+  \centering
+  \includegraphics[width=5in,clip]{chapters/mc_background/dir_adj_iterations.pdf}
+  \caption{\textbf{Iterations to converge vs. Problem Size ($N$ for an
+      $N \times N$ square mesh).} \textit{Both the adjoint and direct
+      solvers are used with the five-point and nine-point stencils.}}
+  \label{fig:poisson_iterations}
+\end{figure}
+
+We see clearly in Figure~\ref{fig:poisson_cpu_time} that the using the
+adjoint solver with MCSA results in a speedup over the direct solver
+while the number of iterations required to converge is also reduced as
+shown in Figure~\ref{fig:poisson_iterations}. We expect this for several
+reasons. First, with an equivalent number of histories specified for
+both solvers per MCSA iteration and a system of size $N \times N$, the
+direct solver will compute a single random walk for each state in the
+system per iteration to acquire a solution in that state, regardless
+of the size of the residual in that state. This is necessary in the
+direct method to ensure a contribution from each state as the random
+walk sequence will only contribute to the starting state. For the
+adjoint method, a total of $N \times N$ random walk events will have
+their starting state determined by sampling the residual
+vector. Because the random walk sequence contributes to the state in
+which it currently resides, sampling the residual vector as the Monte
+Carlo source gives a higher density of random walk events in regions
+with a high residual, thus giving a more accurate correction in that
+region due to reduced statistical error. From an iteration
+perspective, Figure~\ref{fig:poisson_iterations} shows that using the
+direct method yields a roughly unchanging number of iterations
+required to converge as the problem size increases. Again, if we
+desire a correction value for all states in the problem, then we must
+start a random walk in each state in the system which does not reduce
+the number of iterations need as the problem size grows. Conversely,
+as the problem size grows in the adjoint method, the additional
+stochastic histories that will be computed are concentrated in regions
+with a large residual, further reducing the stochastic error in the
+correction in those regions and subsequently reducing the required
+number of iterations to converge.
+
+As an additional comparison, the convergence behavior of MCSA can be
+analyzed using both the adjoint and direct solvers to detect any
+performance benefits. To assess the convergence properties of MCSA
+using each solver and stencil, the infinity norm of the residual
+computed in Eq.~(\ref{eq:mcsa}) was collected at each iteration for a
+fixed problem size of $N=500$. Figure~\ref{fig:poisson_convergence}
+gives the results of these computations. First, it is worthy to note
+on the semilog plot that we are indeed achieving the expected
+exponential convergence from MCSA with both Monte Carlo
+solvers. Second, we note that using the adjoint method with the same
+number of stochastic histores per MCSA iteration gives a faster rate
+of converge for the same reasons as above. We also note here that
+fewer iterations are required for convergence when the 9-point stencil
+is used to discretize the Laplacian operator (although at no gain in
+speed as given by the results in
+Figure~\ref{fig:poisson_cpu_time}). This is due to the fact that the
+smaller discretization error directly corresponds to a more well
+defined residual source generated by the Richardson extrapolation for
+the Monte Carlo calculation. In addition, the better defined source is
+transported through a domain described more accurately by the 9-point
+stencil, thus yielding a more accuracte correction vector from the
+Monte Carlo calculation.
+\begin{figure}[ht!]
+  \centering
+  \includegraphics[width=5in,clip]{chapters/mc_background/dir_adj_conv.pdf}
+  \caption{\textbf{Infinity norm of the solution residual
+      vs. iteration number for a problem of size $N=500$.}
+    \textit{Both the adjoint and direct solvers are used with the five
+      point and nine point stencils. A higher rate of convergence is
+      observed for MCSA using the adjoint Monte Carlo solver as
+      compared to the direct method when both solvers compute the same
+      number of random walks per iteration.}}
+  \label{fig:poisson_convergence}
+\end{figure}
+
+\section{Comparison to Sequential Monte Carlo}
+\label{subsec:sequential_comparison}
+To further motivate using Monte Carlo Synthetic Acceleration, we
+compare its performance to Halton's Sequential Monte Carlo method on
+which our previous work in this area was based. For this comparison,
+we use the same transient Poisson problem as described in the previous
+section and choose only the 5-point stencil to discretize the
+Laplacian operator as the previous results yielded little qualitative
+difference between the discretizations. Both MCSA and Halton's method
+are used with the adjoint Monte Carlo solver. In order to complete the
+same study as in the previous section, the number of histories
+computed by the Monte Carlo solver at each iteration had to be doubled
+to $2 \times N \times N$ in order to ensure convergence in Sequential
+Monte Carlo Method. Figure~\ref{fig:seq_cpu_time} gives the CPU time
+results for this comparison as a function of problem size while
+Figure~\ref{fig:seq_iterations} gives the number of iterations to
+converge as a function of problem size. In both cases, using the Monte
+Carlo solver as a synthetic acceleration rather than in a pure
+residual Monte Carlo scheme resulted in a reduction in both CPU time
+and iterations required to converge. The additional Richardson
+extrapolation between each Monte Carlo solve in the MCSA method gives
+a better converged residual source to use with the Monte Carlo
+calculation while the Sequential method requires more iterations to
+achieve the same level of convergence in the residual.
+
+\begin{figure}[ht!]
+  \centering
+  \includegraphics[width=5in,clip]{chapters/mc_background/seq_cpu.pdf}
+  \caption{\textbf{CPU Time (s) to converge vs. Problem Size ($N$ for
+      an $N \times N$ square mesh).} \textit{oth the Sequential Monte
+      Carlo and MCSA solvers are used with the five point stencils and
+      the adjoint Monte Carlo solver. The number of random walks was
+      twice the number of discrete states in the system in order to
+      ensure convergence in the Sequential Monte Carlo method.}}
+  \label{fig:seq_cpu_time}
+\end{figure}
+
+\begin{figure}[ht!]
+  \centering
+  \includegraphics[width=5in,clip]{chapters/mc_background/seq_iterations.pdf}
+  \caption{\textbf{Iterations to converge vs. Problem Size ($N$ for an
+      $N \times N$ square mesh).} \textit{Both the Sequential Monte
+      Carlo and MCSA solvers are used with the five point stencils and
+      the adjoint Monte Carlo solver.}}
+  \label{fig:seq_iterations}
+\end{figure}
+
+The benefits of using a synthetic acceleration scheme are also noted
+when the infinity norm of the residual computed at each iteration for
+both methods was collected at each iteration for a fixed problem sizes
+of $N=100$ and $N=500$ as shown in figures Figure~\ref{fig:seq_100}
+and \ref{fig:seq_500} respectively. In both cases, the Sequential
+method is subject to two regimes of exponential convergence with the
+later regime converging the slowest while the MCSA method exhibits a
+single rate of exponential convergence observed to be much higher than
+that computed by Halton's method. Even with the doubling of the number
+of stochastic histories computed per time step in order to ensure
+convergence for the Sequential method, we still see robustness issues
+with a non-monotonically decreasing residual observed for the $N=100$
+case. In both cases the MCSA solver is observed to be robust with a
+monotonically decreasing residual.
+
+\begin{figure}[ht!]
+  \centering
+  \includegraphics[width=5in,clip]{chapters/mc_background/seq_conv_100.pdf}
+  \caption{\textbf{Infinity norm of the solution residual
+      vs. iteration number for a problem of size $N=100$.}
+    \textit{Both the Sequential Monte Carlo and MCSA solvers are used
+      with the five point stencils and the adjoint Monte Carlo
+      solver.}}
+  \label{fig:seq_100}
+\end{figure}
+
+\begin{figure}[ht!]
+  \centering
+  \includegraphics[width=5in,clip]{chapters/mc_background/seq_conv_500.pdf}
+  \caption{\textbf{Infinity norm of the solution residual
+      vs. iteration number for a problem of size $N=500$.}
+    \textit{Both the Sequential Monte Carlo and MCSA solvers are used
+      with the five point stencils and the adjoint Monte Carlo
+      solver.}}
+  \label{fig:seq_500}
+\end{figure}
 

### 2013-03-21 doc/dissertation/document/chapters/navier_stokes_solutions/navier_stokes_solutions.tex
added jcomp analysis

@@
 
-Put Navier Stokes work here.
+\subsection{FANM Verification}
+\label{subsec:fanm_verification}
+To verify the FANM method for nonlinear problems, we choose benchmark
+solutions for the 2-dimensional, steady, incompressible Navier-Stokes
+equations on a rectilinear grid in much the same way as Shadid and
+Pawlowski's work on Newton-Krylov methods for the solution of these
+equations \citep{shadid_inexact_1997,pawlowski_globalization_2006}. We
+define these equations as follows:
+\begin{subequations}
+  \begin{gather}
+    \rho \ve{u} \cdot \nabla \ve{u} - \nabla \cdot \ve{T} - \rho
+    \ve{g} = \ve{0}
+    \label{eq:ns_momentum}\\
+    \nabla \cdot \ve{u} = 0
+    \label{eq:ns_continuity}\\
+    \rho C_p \ve{u} \cdot \nabla T + \nabla \cdot \ve{q} = 0\:,
+    \label{eq:ns_energy}
+  \end{gather}
+  \label{eq:navier_stokes}
+\end{subequations}
+where $\rho$ is the fluid density, $\ve{u}$ is the fluid velocity,
+$\ve{g}$ gravity, $C_p$ the specific heat capacity at constant
+pressure of the fluid and $T$ the temperature of the
+fluid. Eq~(\ref{eq:ns_momentum}) provides momentum transport,
+Eq~(\ref{eq:ns_continuity}) provides the mass balance, and
+Eq~(\ref{eq:ns_energy}) provides energy transport with viscous
+dissipation effects neglected. In addition, we close the system with
+the following equations:
+\begin{subequations}
+  \begin{gather}
+    \ve{T} = -P \ve{I} + \mu[\nabla \ve{u} + \nabla \ve{u}^T]
+    \label{eq:ns_stress_tensor}\\
+    \ve{q} = - k \nabla T\:,
+    \label{eq:ns_heat_flux}
+  \end{gather}
+  \label{eq:ns_closure}
+\end{subequations}
+where $\ve{T}$ is the stress tensor, $P$ is the hydrodynamic pressure,
+$\mu$ is the dynamic viscosity of the fluid, $\ve{q}$ is the heat flux
+in the fluid, and $k$ is the thermal conductivity of the fluid. This
+set of strongly coupled equations possesses both the nonlinearities
+and asymmetries that we are seeking for qualification of the FANM
+method. Further, physical parameters within these equations can be
+tuned to enhance the nonlinearities. We will then apply these
+equations to the following three standard benchmark problems.
+
+\subsubsection{Thermal Convection Cavity Problem}
+\label{subsubsec:natural_convection_cavity}
+In 1983 a benchmark solution for the natural convection of air in a
+square cavity was published \citep{de_vahl_davis_natural_1983} as
+shown in Figure~\ref{fig:natural_convection_cavity} for the solution
+of the energy, mass, and momentum equations.
+\begin{figure}[htpb!]
+  \begin{center}
+    \scalebox{1.5}{
+      \input{chapters/navier_stokes_solutions/natural_convection_cavity.pdftex_t} }
+  \end{center}
+  \caption{\textbf{Problem setup for the natural convection cavity
+      benchmark.} \textit{Dirichlet conditions are set for the
+      temperature on the left and right while Neumann conditions are
+      set on the top and bottom of the Cartesian grid. The temperature
+      gradients will cause buoyancy-driven flow. Zero velocity
+      Dirichlet conditions are set on each boundary. No thermal source
+      was present.}}
+  \label{fig:natural_convection_cavity}
+\end{figure}
+In this problem, a rectilinear grid is applied to the unit square. No
+heat flow is allowed out of the top and bottom of the square with a
+zero Neumann condition specified. Buoyancy driven flow is generated by
+the temperature gradient from the cold and hot Dirichlet conditions on
+the left and right boundaries of the box. By adjusting the Rayleigh
+number of the fluid (and therefore adjusting the ratio of convective
+to conductive heat transfer), we can adjust the influence of the
+nonlinear convection term in Eq~(\ref{eq:ns_momentum}). In Shadid's
+work, Rayleigh numbers of up to \sn{1}{6} were used for this benchmark
+on a $100 \times 100$ square mesh grid.
+
+\subsubsection{Lid Driven Cavity Problem}
+\label{subsubsec:lid_driven_cavity}
+As an extension of the convection problem, the second benchmark
+problem given by Ghia \citep{ghia_high-re_1982} adds a driver for the
+flow to introduce higher Reynolds numbers into the system, providing
+more inertial force to overcome the viscous forces in the fluid. The
+setup for this problem is equally simple, containing only the
+Dirichlet conditions as given in Figure~\ref{fig:lid_driven_cavity}
+and is only applied to the mass and momentum equations on the unit
+square.
+\begin{figure}[htpb!]
+  \begin{center}
+    \scalebox{1.5}{
+      \input{chapters/navier_stokes_solutions/lid_driven_cavity.pdftex_t} }
+  \end{center}
+  \caption{\textbf{Problem setup for the lid driven cavity benchmark.}
+    \textit{Dirichlet conditions of zero are set for the velocity on
+      the left and right and bottom while the Dirichlet condition set
+      on the top provides a driving force on the fluid.}}
+  \label{fig:lid_driven_cavity}
+\end{figure}
+The top boundary condition will provide a driver for the flow and its
+variation will in turn vary the Reynolds number of the fluid. An
+increased velocity will generate more inertial forces in the fluid,
+which will overcome the viscous forces and again increase the
+influence of the nonlinear terms in Eq~(\ref{eq:ns_momentum}). Shadid
+also used a $100 \times 100$ grid for this benchmark problem with
+Reynolds numbers up to \sn{1}{4}.
+
+\subsubsection{Backward-Facing Step Problem}
+\label{subsubsec:backward_facing_step}
+The third benchmark was generated by Gartling in 1990 and consists of
+both flow over a backward step and an outflow boundary condition
+\citep{gartling_test_1990}. Using the mass and momentum equations
+while neglecting the energy equation, this problem utilizes a longer
+domain with a 1/30 aspect ratio with the boundary conditions as shown
+in Figure~\ref{fig:backward_facing_step}.
+\begin{figure}[htpb!]
+  \begin{center}
+    \scalebox{1.3}{
+      \input{chapters/navier_stokes_solutions/backward_facing_step.pdftex_t} }
+  \end{center}
+  \caption{\textbf{Problem setup for the backward facing step
+      benchmark.} \textit{Zero velocity boundary conditions are
+      applied at the top and bottom of the domain while the outflow
+      boundary condition on the right boundary is represented by zero
+      stress tensor components in the direction of the flow. For the
+      inlet conditions, the left boundary is split such that the top
+      half has a fully formed parabolic flow profile and the bottom
+      half has a zero velocity condition, simulating flow over a
+      step.}}
+  \label{fig:backward_facing_step}
+\end{figure}
+In this problem, the inflow condition is specified by a fully-formed
+parabolic flow profile over a zero velocity boundary representing a
+step. The flow over this step will generate a recirculating backflow
+under the inlet flow towards the step. As in the lid driven cavity
+problem, the nonlinear behavior of this benchmark and the difficulty
+in obtaining a solution is dictated by the Reynolds number of the
+fluid. In Shadid's work, a $20 \times 400$ non-square rectilinear grid
+was used to discretize the domain with Reynolds number up to
+\sn{5}{2}.

### 2013-03-21 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
adding parallel mc chapter

--- 2013-03-21 doc/dissertation/document/chapters/mc_background/mc_background.tex
@@
 
+%%---------------------------------------------------------------------------%%
 \section{Preliminaries}
@@
 
+%%---------------------------------------------------------------------------%%
 \section{Direct Method}
@@
 
+%%---------------------------------------------------------------------------%%
 \section{Adjoint Method}
@@
 
+%%---------------------------------------------------------------------------%%
 \section{Sequential Monte Carlo}
@@
 
+%%---------------------------------------------------------------------------%%
 \section{Monte Carlo Synthetic-Acceleration}
@@
 
+%%---------------------------------------------------------------------------%%
 \section{Monte Carlo Method Selection}
-
 The MCSA method defined in Eq.~(\ref{eq:mcsa}) uses the adjoint method
@@
 
+%%---------------------------------------------------------------------------%%
 \section{Comparison to Sequential Monte Carlo}
@@
 
+%%---------------------------------------------------------------------------%%
 \section{Parallelization of Stochastic Methods}

### 2013-03-21 doc/dissertation/document/chapters/mc_background/mc_background.tex
adding parallel mc section

@@
 \end{figure}
-
-\section{Parallelization of Stochastic Methods}
-\label{sec:parallel_stochastic_methods}
-For Monte Carlo methods, and in particular MCSA, to be viable at the
-production scale, scalable parallel implementations are required. In a
-general linear algebra context for discrete systems, this has not yet
-been achieved. Therefore, we will develop and implement parallel
-algorithms for these methods leveraging both the knowledge gained from
-the general parallel implementations of Krylov methods and modern
-parallel strategies for Monte Carlo as developed by the reactor
-physics community. In order to formulate a parallel MCSA algorithm, we
-recognize that the algorithm occurs in two stages, an outer iteration
-performing Richardson's iteration and applying the correction, and an
-inner Monte Carlo solver that is generating the correction via the
-adjoint method. The parallel aspects of both these components must be
-considered.
-
-\subsection{Domain Decomposition for Monte Carlo}
-\label{subsec:msod}
-As observed in the discussion on parallel Krylov methods, large-scale
-problems will surely have their data partitioned such that each
-parallel process owns a subset of the equations in the linear
-system. Given this convention, the adjoint Monte Carlo algorithm must
-perform random walks over a domain that is decomposed and must remain
-decomposed due to memory limitations. This naturally leads us to seek
-parallel algorithms that handle domain decomposition.
-
-In the context of radiation transport, Brunner and colleagues provided
-a survey of algorithms for achieving this as implemented in production
-implicit Monte Carlo codes \citep{brunner_comparison_2006}. In their
-work they identify two data sets that are required to be communicated:
-the sharing of particles that are transported from one domain to
-another and therefore from one processor to another and a global
-communication that signals if particle transport has been completed on
-all processors. The algorithms presented are a fully-locking
-synchronous scheme, an asynchronous-send/synchronous-receive pattern,
-a traditional master/slave scheme, and a modified master/slave scheme
-the implements a binary tree pattern for the global reduction type
-operations needed to communicate between the master and slave
-processes. They observed that the modified master/slave scheme
-performed best in that global communications were implemented more
-efficiently than those required by the asynchronous
-scheme. Furthermore, none of these schemes handled load-imbalanced
-cases efficiently. Such cases will be common if the source sampled in
-the Monte Carlo random walk is not isotropic and not evenly
-distributed throughout the global domain. It was noted that
-efficiencies were improved by increasing the frequency by which
-particle data was communicated between domain-adjacent
-processors. However, this ultimately increases communication costs. In
-2009, Brunner extended his work by using a more load-balanced approach
-with a fully asynchronous communication pattern
-\citep{brunner_efficient_2009}. Although the extended implementation
-was more robust and allowed for scaling to larger numbers of
-processors, performance issues were still noted with parallel
-efficiency improvements needed in both the weak and strong scaling
-cases for unbalanced problems. These results led Brunner to conclude
-that a combination of domain decomposition and domain replication
-could be used to solve some of these issues.
-
-\subsubsection{Multiple-Set Overlapping-Domain Decomposition}
-\label{subsubsec:msod}
-In 2010, Wagner and colleagues developed the \textit{multiple-set
-  overlapping-domain} (MSOD) decomposition for parallel Monte Carlo
-applications for full-core light water reactor analysis
-\citep{wagner_hybrid_2010}. In their work, an extension of Brunner's,
-their scheme employed the similar parallel algorithms for particle
-transport but a certain amount of overlap between adjacent domains was
-used to decrease the number of particles leaving the local domain. In
-addition, Wagner utilized a level of replication of the domain such
-that the domain was only decomposed on $O(100)$ processors and if
-replicated $O(1,000)$ times achieves simulation on $O(100,000)$
-processors, thus providing spatial and particle parallelism. Each
-collection of processors that constitutes a representation of the
-entire domain is referred to as a set, and within a set overlap occurs
-among its sub-domains. The original motivation was to decompose the
-domain in a way that it remained in a physical cabinet in a large
-distributed machine, thus reducing latency costs during
-communication. A multiple set scheme is also motivated by the fact
-that communication during particle transport only occurs within a set,
-limiting communications during the transport procedure to a group of
-$O(100)$ processors, a number that was shown to have excellent
-parallel efficiencies in Brunner's work and therefore will scale well
-in this algorithm. The overlapping domains within each set also
-demonstrated reduced communication costs. On each processor, the
-source is sampled in the local domain that would exist if no overlap
-was used while tallies can be made over the entire overlapping domain.
-
-To demonstrate this, consider the example adapted from Mervin's work
-with Wagner and others in the same area \citep{mervin_variance_2012}
-and presented in Figure~\ref{fig:msod_example}.
-\begin{figure}[htpb!]
-  \begin{center}
-    \scalebox{1.5}{
-      \input{chapters/mc_background/msod_example.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Overlapping domain example illustrating how domain
-      overlap can reduce communication costs.}
-    \textit{All particles start in the blue region of interest. The
-      dashed line represents 0.5 domain overlap between domains.}}
-  \label{fig:msod_example}
-\end{figure}
-In this example, 3 particle histories are presented emanating from the
-blue region of interest. Starting with particle A, if no domain
-overlap is used then the only the blue domain exists on the starting
-processor. Particle A is then transported through 3 other domains
-before the history ends, therefore requiring three communications to
-occur in Brunner's algorithm. If a 0.5 domain overlap is permitted as
-shown by the dashed line, then the starting process owns enough of the
-domain such that no communications must occur in order to complete the
-particle A transport process. Using 0.5 domain overlap also easily
-eliminates cases such as the represented by the path of particle C. In
-this case, particle C is scattering between two adjacent domains,
-incurring a large latency cost for a single particle. Finally, with
-particle B we observe that 0.5 domain overlap will still not eliminate
-all communications. However, if 1 domain overlap were used, the entire
-geometry shown in Figure~\ref{fig:msod_example} would be contained on
-the source processor and therefore transport of all 3 particles
-without communication would occur.
-
-Wagner and colleagues used this methodology for a 2-dimensional
-calculation of a pressurized water reactor core and varied the domain
-overlap from 0 to 3 domain overlap (a $7 \times 7$ box in the context
-of our example) where a domain constituted a fuel assembly. For the
-fully domain decomposed case, they observed that 76.12\% of all source
-particles leave the domain. At 1.5 domain overlap, the percentage of
-source particles born in the center assembly leaving the processor
-domain dropped to 1.05\% and even further for 0.02\% for the 3 domain
-overlap. Based on these results, this overlap approach, coupled with
-the multiple sets paradigm that will scale for existing parallel
-transport algorithms, provides a scalable Monte Carlo algorithm for
-today's modern machines.
-
-\subsubsection{Domain-to-Domain Communication}
-\label{subsubsec:mc_comm_costs}
-With Wagner's work, we observed that domain overlap implemented with a
-domain decomposed Monte Carlo setting reduces the amount of nearest
-neighbor communication that must occur during transport. This
-communication, expected to be expensive, was analyzed by Siegel and
-colleagues for idealized, load balanced situations without domain
-overlap \citep{siegel_analysis_2012}. They arrive at a few empirical
-relationships and generate a numerical experiment that provides
-insight into the feasibility of large scale domain decomposed Monte
-Carlo as well as the relationship of communication to system
-parameters. In their work, they define a simple 3-dimensional system
-subdivded into Cartesian block domains with each processor containing
-a single domain. Given this geometry, a particle can either be
-absorbed in its current domain or be transported to one of the 6
-adjacent domains owned by another process.
-
-To quantify the number of particles that leak out of the local domain
-they define a leakage fraction, $\lambda$, as:
-\begin{equation}
-  \lambda =
-  \frac{average\ \#\ of\ particles\ leaving\ local\ domain}{total\ 
-    of\ \#\ of\ particles\ starting\ in\ local\ domain}\:.
-  \label{eq:leakage_fraction}
-\end{equation}
-For their studies, they assumed that the value of $\lambda$ was
-dependent on the total cross section of the system and the same for
-all processes for simplicity and load balancing. Using this
-relationship and their simple geometry, they were able to relate the
-amount of compute time required to transport all source particles to
-the number of source particles, the leakage fraction, the latency and
-bandwidth of communication, the ratio of local to global geometry
-size, and the total cross section of the system.
-
-What they found was that there was indeed an optimum ratio of local
-domain to global domain size that results in a minimum total compute
-time. They also found that this ratio was only dependent on the
-latency and bandwidth constraints of the parallel machine. We expect
-this because nearest neighbor communication patterns scale with the
-volume and surface sizes of the local domains. The larger the volume,
-the more compute time is spent within the local volume and therefore
-at some point you are limited by the latency of the system, or the
-time it takes to receive the data to operate on. However, at a certain
-point, when the surface of those volumes becomes large enough, the
-amount of data that must be communicated grows large enough that it
-may overcome the bandwidth limitations of the system, resulting in a
-decrease in performance. In their numerical experiments, they used the
-above geometry and leakage rates based on that computed for a reactor
-system for a single material problem with a total cross section
-equivalent to that of homogenized light water reactor. Given the
-optimal local domain size, they observed for load balanced situations
-that nearest neighbor communication costs during transport are a small
-fraction of total compute time, leading to the conclusion that domain
-decomposition is feasible for large-scale Monte Carlo calculations.
-
-\subsubsection{Load Balancing Concerns}
-\label{subsubsec:mc_load_balancing}
-Although domain decomposition was shown to be efficient in a perfectly
-load balanced situation in Siegel's work \citep{siegel_analysis_2012},
-careful consideration must be made for situations where this is not
-the case. Given the stochastic nature of the problem and lack of a
-globally homogeneous domain, parallel Monte Carlo simulations are
-inherently load imbalanced. Procassini and others worked to alleviate
-some of the load imbalances that are generated by both particle and
-spatial parallelism and are therefore applicable to the MSOD algorithm
-\citep{procassini_dynamic_2005}. They chose a dynamic balancing scheme
-in which the number of times a particular domain was replicated was
-dependent on the amount of work in that domain (i.e. domains with a
-high particle flux and therefore more particle histories to compute
-require more work). In this variation, domains that require more work
-will be replicated more frequently at reduced particle counts in each
-replication. Furthermore, Procassini and colleagues noted that as the
-simulation progressed and particles were transported throughout the
-domain, the amount of replication for each domain would vary as
-particle histories began to diffuse, causing some regions to have
-higher work loads and some to have smaller work loads than the initial
-conditions.
-
-Consider the example in Figure~\ref{fig:procassini_example} adapted
-from Procassini's work.
-\begin{figure}[htpb!]
-  \begin{center}
-    \scalebox{1.5}{
-      \input{chapters/mc_background/procassini_example.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Example illustrating how domain decomposition can
-      create load balance issues in Monte Carlo.}  \textit{A domain is
-      decomposed into 4 zones on 8 processors with a point source in
-      the lower left zone. As the particles diffuse from the source in
-      the random walk sequence as shown in the top row, their tracks
-      populate the entire domain. As given in the bottom row, as the
-      global percentage of particles increases in a zone, that zone's
-      replication count is increased.}}
-  \label{fig:procassini_example}
-\end{figure}
-In this example, a geometry is decomposed into 4 domains on 8
-processors with a point source in the bottom left domain. To begin,
-because the point source is concentrated in one domain, that domain is
-replicated 5 times such the amount of work it has to do per processor
-is roughly balanced with the others. As the particles begin to diffuse
-away from the point source, the amount of replication is adjusted to
-maintain load balance. Near the end of the simulation, the diffusion
-of particles is enough that all domains have equal replication.  By
-doing this, load balance is improved as each domain has approximately
-equal work although each domain may represent a different spatial
-location and have a differing number of histories to
-compute. Compared to Wagner's work where the fission source was
-distributed relatively evenly throughout the domain, fixed source
-problems (and especially those that have a point-like source) like
-those presented in Procassini's work will be more prone to changing
-load balance requirements.
-
-\subsubsection{Reproducible Domain Decomposed Results}
-\label{subsubsec:reproducible_mc}
-The 2006 work of Brunner is notable in that the Monte Carlo codes used
-to implement and test the algorithms adhered to a strict policy of
-generating identical results independent of domain decomposition or
-domain replication as derived from the work of Gentile and colleagues
-\citep{gentile_obtaining_2005}. In Gentile's work, a procedure is
-given for obtaining results reproducible to machine precision for an
-arbitrary number of processors and domains. Differences can arise from
-using a different random number sequence in each domain and performing
-a sequence of floating point operations on identical data in a
-different order, leading to variations in round-off error and
-ultimately a non-identical answer. They use a simple example,
-recreated below in Figure~\ref{fig:gentile_example}, that illustrates
-these issues.
-\begin{figure}[htpb!]
-  \begin{center}
-    \scalebox{1.5}{
-      \input{chapters/mc_background/gentile_example.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Gentile's example illustrating how domain
-      decomposition can create reproducibility issues in Monte Carlo.}
-    \textit{Both particles A and B start in zone 1 on processor
-      1. Particle A moves to zone 2 on processor 2 and scatters back
-      to zone 1 while B scatters in zone 1 and remains there. A1 and
-      A2 denote the track of particle A that is in zone 1 while B1 and
-      B2 denote the track of particle B that is in zone 1.}}
-  \label{fig:gentile_example}
-\end{figure}
-In this example, the domain is decomposed on two processors with each
-processor owning one of the two zones. Starting with particle A, it is
-born in zone 1 and is transported to zone 2 where a scattering event
-occurs. Concerning the first reproducibility issue, if the same
-sequence of random numbers is not used to compute the trajectory from
-the new scattering event, we cannot expect to achieve the same result
-if the domain were not decomposed. If the numbers are different, the
-scattering event in zone 2 may keep the particle there or even eject
-it from the domain if the sequence were different. The second issue is
-demonstrated by adding another particle B that remains in the
-domain. In this case, an efficient algorithm will transport particle A
-on processor 1 until it leaves zone 1 and then transport particle
-B. Particle A will not renter the domain until it has been
-communicated to processor 2, processor 2 performs the transport, and
-it is communicated back to processor 1. If we are doing a track-length
-tally in zone 1, then we sum the tracks lengths observed in that
-zone. In the single processor, single zone case particle A would be
-transported in its entirety and then particle B transported. This
-would result in a tally sum with the following order of operations:
-$(((A1+A2)+B1)+B2)$. If we were instead to use 2 processors, we would
-instead have the following order: $(((A1+B1)+B2)+A2)$. In the context
-of floating point operations, we cannot expect these to have an
-identical result to machine precision as round-off approximations will
-differ resulting in non-commutative addition.
-
-Procassini's solutions to these problems are elegant in that they
-require a minimal amount of modification to be applied to the Monte
-Carlo algorithm. To solve the first issue, in order to ensure each
-particle maintains an invariant random number sequence that determines
-its behavior regardless of domain decomposition, each particle is
-assigned a random number seed that describes its current state upon
-entering the domain of a new processor. These seeds are derived from
-the actual geometric location of the particle such that it is
-decomposition invariant. Non-commutative floating point operations are
-overcome by instead mapping floating point values to 64-bit integer
-values for which additions will always be commutative. Once the
-operations are complete, these integers are mapped back to floating
-point values.
-
-\subsection{Parallel Adjoint Method}
-\label{subsec:parallel_adjoint}
-We can take much of what was learned from the survey of parallel Monte
-Carlo methods for radiation transport and directly apply it to a
-parallel formulation of our stochastic linear solvers. Direct analogs
-can be derived from these works by noting that the primary difference
-between solving a linear system with Monte Carlo methods and fixed
-source Monte Carlo transport problems is the content of the Markov
-chains that are generated. The transitions represented by these chains
-are bound by probabilities and weights and are initiated by the
-sampling of a source. In the context of transport problems, those
-transitions represent events such as particle scattering and
-absorption with probabilities that are determined by physical data in
-the form of cross sections. For stochastic matrix inversion, those
-transitions represent moving between the equations of the linear
-system (and therefore the physical domain which they represent) and
-their probabilities are defined by the coefficients of those
-equations. Ultimately, we tally the contributions to generate
-expectation values in the desired states as we progress through the
-chains. Therefore, parallel methods for Monte Carlo radiation
-transport can be abstracted and we can use those concepts that apply
-to matrix inversion methods as an initial means of developing a
-parallel Neumann-Ulam-type solver. Based on the results observed in
-the last decade of parallel Monte Carlo development, we can generate
-many practical questions that this type of work could answer.
-
-Given a decomposed domain, per Wagner's work
-\citep{wagner_hybrid_2010} is clear that domain overlap significantly
-reduces the amount of communication required between adjacent domains
-as histories move to states that are not owned by the local
-processor. How much domain overlap is suitable for matrix inversion
-problems? Are we memory limited by large problems such that only so
-much overlap is feasible? Are there metrics related to the properties
-of the matrix including the eigenvalues and sparsity pattern such that
-we can provide guidelines for selecting the amount of overlap
-required? Furthermore, as Wagner and colleagues observed
-\citep{wagner_hybrid_2010}, with marginal domain overlap for reactor
-problems, the percentage of histories leaving the local domain can
-easily be reduced to less than 1\%. Evans and colleagues in their
-initial MCSA development typically used only 50 particle histories in
-order to compute a correction in an MCSA iteration
-\citep{evans_monte_2012}. Per the Central Limit Theorem, that
-correlates to a statistical uncertainty of 14.1\% in the correction,
-yet as compared to conventional Krylov solvers, MCSA achieved
-identical numerical results. If good convergence and numerically
-accurate solutions can be achieved with such a large uncertainty in
-the correction computation, then perhaps with enough domain overlap
-the minimal amount of histories that do transition to non-local states
-can be ignored and thus eliminate all communication in the parallel
-Monte Carlo adjoint solver transport sequence and create an
-embarrassingly parallel method. Given Gropp's work on parallel Krylov
-methods that we discussed in
-\S~\ref{subsec:projection_method_performance} and the empirical
-results of Siegel \citep{siegel_analysis_2012}, we know that these
-nearest neighbor computations between adjacent domains do not scale as
-well as global reduction operations and therefore we are improving
-scaling by eliminating them from the transport sequence. It will be
-important to determine if the expectation value bias in the MCSA
-correction generated by this approximation will remain within the
-bounds of the already high statistical uncertainty for unbiased
-estimates, thus providing numerically equivalent results.
-
-From a domain replication perspective, this may be difficult to
-achieve with a production scale linear solver. Typically, memory is at
-a premium and therefore the more distinct domains available in the
-decomposition, the spatially finer and/or the numerically more
-accurate the discretization that can be implemented. How much
-replication is possible for large problems? How does replication
-facilitate parallel performance when coupled with domain overlap? How
-can we measure how much domain overlap is feasible? Replicating
-domains may therefore run into memory limitations for exceptionally
-large problems such that the operator, solution vector, and source
-vector must be copied in their entirety multiple times. From a
-resiliency standpoint, such an operation will be required, and
-therefore its performance and memory implications on conventional
-problems must be analyzed.
-
-\subsection{Parallel MCSA}
-\label{subsec:parallel_mcsa}
-With a parallel adjoint Neumann-Ulam solver implementation, the
-parallel implementation of the MCSA method will be trivial. Recall the
-MCSA iteration procedure outlined in Eq~(\ref{eq:mcsa}). In
-\S~\ref{sec:parallel_krylov_methods} we discussed parallel matrix and
-vector operations as utilized in conventional Krylov methods. We
-utilize these here for the parallel MCSA implementation. In the first
-step, a parallel matrix-vector multiply is used to apply the split
-operator to the previous iterate's solution. A parallel vector update
-is then performed with the source vector to arrive at the initial
-iteration guess. In the next step, the residual is computed by the
-same operations where now the operator is applied to the solution
-guess with a parallel matrix-vector multiply and then a parallel
-vector update with the source vector is performed. Once the correction
-is computed with a parallel adjoint Neumann-Ulam solve, this
-correction is applied to the guess with a parallel vector update to
-get the new iteration solution. Additionally, as given by
-Eq~(\ref{eq:mcsa_stopping_criteria}), 2 parallel vector reductions
-will be required to check the stopping criteria: one initially to
-compute the infinity norm of the source vector, and another at every
-iteration to compute the infinity norm of the residual vector. For
-this implementation, all of the issues that will be potentially
-generated by the parallel adjoint solver implementation will manifest
-themselves here as the quality of the correction will be of intense
-study.
-
-In addition to parallel implementation and performance, MCSA's
-potential for aiding advancement in non-symmetric matrix solutions
-leads to a natural comparison with the GMRES algorithm. As both
-solvers are aimed at the same class of problem, we desire a set of
-metrics that will allow us to quantitatively compare the two. Given
-that the Krylov subspace maintained by GMRES can become large, do we
-benefit from a memory standpoint with an MCSA scheme in that no
-subspace is required? Does this benefit outweigh the fact that the
-linear operator must be explicitly formed in order to build the
-transition probabilities for the random walk sequence? For
-non-symmetric systems, does MCSA exhibit similar convergence
-properties to Krylov methods? If a Krylov methods build a subspace,
-can those memory savings in MCSA be used to implement domain
-replication in the adjoint solver? Such questions can be answered by a
-comparative study of the two solvers that controls the system size and
-the iterations required to converge.

### 2013-03-21 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
adding parallel mc section

--- 2013-03-21 doc/dissertation/document/chapters/mc_background/mc_background.tex
@@
-\chapter{Monte Carlo Solution Methods for Linear Systems}
-\label{ch:stochastic_methods}
-An alternative approach to approximate matrix inversion is to employ
-Monte Carlo methods that sample a distribution with an expectation
-value equivalent to that of the inverted operator. Such methods have
-been in existence for decades with the earliest reference noted here
-an enjoyable manuscript published in 1950 by Forsythe and Leibler
-\citep{forsythe_matrix_1950}. In their outline, Forsythe and Liebler
-in fact credit the creation of this technique to J. Von Neumann and
-S.M. Ulam some years earlier than its publication. In 1952 Wasow
-provided a more formal explanation of Von Neumann and Ulam's method
-\citep{wasow_note_1952} and Hammersley and Handscomb's 1964 monograph
-\citep{hammersley_monte_1964} and Spanier and Gelbard's 1969 book
-\citep{spanier_monte_1969} present additional detail on this topic
-using a collection of references from the 1950's and early 1960's.
+\chapter{Parallel Monte Carlo Solution Methods for Linear Systems}
+\label{ch:parallel_methods}
 
-\section{Preliminaries}
-\label{sec:mc_preliminaries}
-We begin our discussion of Monte Carlo methods using these texts by
-seeking a solution to Eq~(\ref{eq:linear_problem}). For a given linear
-operator $\ve{A}$, we can use diagonal splitting in a similar manner
-as the stationary method in Eq~(\ref{eq:linear_split_equation2}) to
-define the following operator\footnote{It should be noted that
-  non-diagonal splittings have been recently explored in
-  \citep{srinivasan_monte_2010} and have the potential to improve
-  efficiency.}:
-\begin{equation}
-  \ve{H} = \ve{I} - \ve{A}\:,
-  \label{eq:linear_mc_iteration_matrix}
-\end{equation}
-such that we are solving the system:
-\begin{equation}
-  \ve{x} = \ve{H} \ve{x} + \ve{b}\:.
-  \label{eq:richardson_split}
-\end{equation}
-We can then form an alternative representation for $\ve{A}^{-1}$ by
-generating the \textit{Neumann series}:
-\begin{equation}
-  \ve{A}^{-1} = (\ve{I}-\ve{H})^{-1} = \sum_{k=0}^{\infty} \ve{H}^k\:,
-  \label{eq:neumann_series}
-\end{equation}
-which will converge if the spectral radius of $\ve{H}$ is less than
-1. If we then apply this Neumann series to the right hand side of
-Eq~(\ref{eq:linear_problem}) we acquire the solution to the linear
-problem:
-\begin{equation}
-  \ve{A}^{-1}\ve{b} = \sum_{k=0}^{\infty} \ve{H}^k\ve{b} = \ve{x}\:.
-  \label{eq:neumann_solution}
-\end{equation}
-An approximation of this summation by truncation will therefore lead
-to an approximation of the solution. If we expand the summation with a
-succession of matrix-vector multiply operations, we arrive at an
-alternative perspective of this summation by considering its $i^{th}$
-component:
-\begin{equation}
-  x_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-  \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}\:,
-  \label{eq:expanded_neumann_solution}
-\end{equation}
-which can interpreted as a series of transitions between states,
-\begin{equation}
- \nu = i \rightarrow i_1 \rightarrow \cdots \rightarrow i_{k-1}
- \rightarrow i_{k}\:,
-  \label{eq:mc_walk_permutation}
-\end{equation}
-in $\ve{H}$ where $\nu$ is interpreted as a particular random walk
-sequence permutation. We can generate these sequences of transitions
-through Monte Carlo random walks by assigning them both a probability
-and weight. As a reinterpretation of the iteration matrix, we then
-form the \textit{Neumann-Ulam decomposition} of \ve{H}:
-\begin{equation}
-  \ve{H} = \ve{P} \circ \ve{W}\:,
-  \label{eq:neumann_ulam_decomposition}
-\end{equation}
-where $\circ$ denotes the Hadamard product operation\footnote{The
-  Hadamard product $\ve{A} = \ve{B} \circ \ve{C}$ is defined
-  element-wise as $a_{ij} = b_{ij} c_{ij}$.}, $\ve{P}$ denotes the
-transition probability matrix, and $\ve{W}$ denotes the transition
-weight matrix. This decomposition, a generalization of Dimov's work
-\citep{dimov_new_1998}, is an extension of the original Neumann-Ulam
-scheme in that now a weight cutoff can be used to terminate a random
-walk sequence and therefore truncate the Neumann series it is
-approximating. The formulation of $\ve{P}$ and $\ve{W}$ will be
-dependent on whether we choose a direct or adjoint Monte Carlo
-sequence to estimate the state transitions in
-Eq~(\ref{eq:expanded_neumann_solution}).
-
-\section{Direct Method}
-\label{sec:direct_mc}
-In the context of matrix inversion, a direct method resembles an
-adjoint Monte Carlo method in the reactor physics community where the
-solution state is sampled and the source terms that contribute to it
-are assembled. To achieve this, we build the direct method
-Neumann-Ulam decomposition per Dimov's approach by first choosing a
-probability matrix that is a row scaling of $\ve{H}$ such that its
-components are:
-\begin{equation}
-  p_{ij} = \frac{|h_{ij}|}{\sum_j |h_{ij}|}\:.
-  \label{eq:direct_probability}
-\end{equation}
-From this, we then see that the probability of transitioning from a
-state $i$ to a state $j$ is implicitly linked to the original operator
-$\ve{A}$ in that those terms with large values, and therefore those
-that make the greatest contribution to the numerical solution, will be
-sampled with a higher probability than smaller terms. In addition, the
-row scaling provides a normalization over the state to which we are
-transitioning such that $\sum_j p_{ij} = 1$, meaning that we sample
-the probabilities over the rows of the matrix. The components of
-the weight matrix are then defined by
-Eq~(\ref{eq:neumann_ulam_decomposition}) as:
-\begin{equation}
-  w_{ij} = \frac{h_{ij}}{p_{ij}}\:.
-  \label{eq:direct_weight}
-\end{equation}
-It should be noted here that if $\ve{A}$ is sparse, then $\ve{H}$,
-$\ve{P}$, and $\ve{W}$ must be sparse as well by
-definition. Additionally, we only compute $\ve{P}$ and $\ve{W}$ from
-the non-zero elements of $\ve{H}$ as those components that are zero
-will not participate in the random walk. Doing so prevents an
-infinite weight from being generated in Eq~(\ref{eq:direct_weight})
-and eliminates unnecessary computations.
-
-Using these matrices, we can then form the expectation value of the
-direct solution. For a given random walk permutation $\nu$ with $k$
-events, we define the weight of that permutation to be:
-\begin{equation}
-  W_{m} = w_{i,i_1} w_{i_1,i_2} \cdots w_{i_{m-1},i_m}\:,
-  \label{eq:direct_permutation_weight}
-\end{equation}
-such that the weight of each transition event contributes to the
-total. The contribution to the solution from a particular random walk
-permutation is then:
-\begin{equation}
-  X_{\nu}(i_0 = i) = \sum_{m=0}^k W_{m} b_{i_m}\:,
-  \label{eq:direct_permutation_contribution}
-\end{equation}
-where $X_{\nu}(i_0 = i)$ signifies that the solution state in which we
-are tallying defines state $i$.  We can interpret this precisely as
-before in that during the random walk we collect the source in the
-states that are visited and apply them to the solution tally. We then
-define the probability that a particular random walk permutation of
-$k$ events will occur:
-\begin{equation}
-  P_{\nu} = p_{i,i_1} p_{i_1,i_2} \cdots p_{i_{k-1},i_k}\:.
-  \label{eq:direct_permutation_probability}
-\end{equation}
-Finally, we define the expectation value of $X$ to be the collection
-of all random walk permutations and their probabilities:
-\begin{equation}
-  E\{X(i_0 = i)\} = \sum_{\nu} P_{\nu} X_{\nu}\:,
-  \label{eq:direct_expectation_value}
-\end{equation}
-which, if expanded, directly recovers the exact solution:
-\begin{equation}
-  \begin{split}
-    E\{X(i_0 = i)\}
-    &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-    \sum_{i_k}^{N} p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
-    w_{i,i_1}w_{i_1,i_2}\ldots w_{i_{k-1},i_k} b_{i_k}\\ &= x_i\:,
-  \end{split}
-  \label{eq:direct_expectation_expansion}
-\end{equation}
-therefore providing an unbiased Monte Carlo estimator.
-
-In cases where we seek only approximate solutions, we need only to
-perform a minimal number of random walks in order to generate an
-approximation for $\ve{x}$. If we are only to approximate the
-solution, we also need conditions by which we may terminate a random
-walk. We do this by noticing that the factors added to
-Eq~(\ref{eq:direct_permutation_weight}) will become diminishingly
-small due to their definition in Eq~(\ref{eq:direct_weight}) and
-therefore their contributions to the solution estimate will become
-negligible. Using this, we choose terminate a random walk sequence
-with a \textit{weight cutoff}, $W_c$, that is enforced when $W_m <
-W_c$ for a particular random walk permutation.
-
-\subsection{Estimator Variance}
-\label{subsec:estimator_variance}
-We can compute the variance of the estimator through traditional
-methods by defining the variance, $\sigma_i$, for each component in
-the solution:
-\begin{equation}
-  {\sigma_i}^2 = E\{X(i_0 = i) - (\ve{A}^{-1}\ve{b})_i\}^2 = E\{X(i_0
-  = i)^2\} - x_i^2\:,
-  \label{eq:direct_variance_1}
-\end{equation}
-where the vector exponentials are computed element-wise. Inserting
-Eq~(\ref{eq:direct_expectation_value}) gives:
-\begin{equation}
-  \sigma_i^2 = \Big(\sum_{\nu} P_{\nu} X_{\nu}^2\Big)_i - x_i^2\:,
-  \label{eq:direct_variance_2}
-\end{equation}
-and applying Eq~(\ref{eq:direct_permutation_contribution}):
-\begin{equation}
-  \sigma_i^2 = \Big(\sum_{\nu} P_{\nu} \sum_{m=0}^k W_{m}^2
-  b_{i_m}^2\Big)_i - x_i^2\:.
-  \label{eq:direct_variance_3}
-\end{equation}
-Finally, expanding the transition probabilities yields the variance:
-\begin{equation}
-  \sigma_i^2 = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-  \sum_{i_k}^{N} p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
-  w^2_{i,i_1}w^2_{i_1,i_2}\ldots w^2_{i_{k-1},i_k} b_{i_m} - x_i^2\:.
-  \label{eq:direct_variance_4}
-\end{equation}
-Using this definition for the variance, we can arrive at a more natural
-reason for enforcing $\rho(\ve{H}) < 1$ for our Monte Carlo method to
-converge. Per the Hadamard product, we can concatenate the summation
-in Eq~(\ref{eq:direct_variance_4}):
-\begin{equation}
-  (\ve{P} \circ \ve{W} \circ \ve{W})^k =
-  \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots \sum_{i_k}^{N}
-  p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
-  w^2_{i,i_1}w^2_{i_1,i_2}\ldots w^2_{i_{k-1},i_k}\:.
-  \label{eq:double_weighted_decomposition}
-\end{equation}
-If we assign $\ve{G} = \ve{P} \circ \ve{W} \circ \ve{W}$ as in
-Eq~(\ref{eq:neumann_ulam_decomposition}), we then have a new
-formulation for the variance:
-\begin{equation}
-  \sigma^2_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-  \sum_{i_k}^{N}g_{i,i_1}g_{i_1,i_2}\ldots g_{i_{k-1},i_k} b_{i_k}^2 -
-  x_i^2\:,
-\end{equation}
-which contains the general Neumann series for $\ve{G}$,
-\begin{equation}
-  \ve{T} = \sum_{k=0}^{\infty} \ve{G}^k\:,
-  \label{eq:variance_neumann_series}
-\end{equation}
-where $\ve{T} = (\ve{I}-\ve{G})^{-1}$. We can then insert $T$ back
-into the variance formulation for a more concise definition:
-\begin{equation}
-  \sigma^2_i = (\ve{T}\ve{b})_i - x_i^2\:.
-  \label{eq:direct_variance_5}
-\end{equation}
-We can relate $\ve{G}$ to $\ve{H}$ by noting that $\ve{G}$ simply
-contains an additional Hadamard product with the weight matrix. The
-Hadamard product has the property that:
-\begin{equation}
-  |\ve{H} \circ \ve{W}| \geq |\ve{H}|\ |\ve{W}|\:.
-  \label{eq:hadamard_inequality}
-\end{equation}
-Using the norm property of the Hadamard product and
-Eq~(\ref{eq:neumann_ulam_decomposition}), we can define the norm of
-$\ve{W}$ as:
-\begin{equation}
-  \frac{|\ve{H}|}{|\ve{P}|} \geq |\ve{W}|\:.
-  \label{eq:weight_norm}
-\end{equation}
-Choosing the infinity norm of the operator as defined in
-Eq~(\ref{eq:matrix_infinity_norm}), the row normalized probability
-matrix will yield a norm of 1 giving the following inequality for
-relating $\ve{G}$ and $\ve{H}$:
-\begin{equation}
-  |\ve{G}| \geq |\ve{H}|^2
-  \label{eq:variance_norm_inequality}
-\end{equation}
-Using these relations to analyze Eq~(\ref{eq:direct_variance_5}), we
-see that if $\rho(\ve{G}) > 1$, then the sum in
-Eq~(\ref{eq:variance_neumann_series}) will not converge and an
-infinite variance will arise as the elements of $\ve{T}$ become
-infinite in Eq~(\ref{eq:direct_variance_5}). We must restrict $\ve{G}$
-to alleviate this and therefore restrict $\ve{H}$ due to
-Eq~(\ref{eq:variance_norm_inequality}) with $\rho(\ve{H}) < 1$ so that
-our expectation values for the solution may have a finite variance.
-
-\subsection{Reusing Random Walks}
-\label{subsec:reusing_random_walks}
-A downfall of the direct method is that a random walk must be
-initiated in each state of the solution vector in which the solution
-is desired due to the estimator definition in
-Eq~(\ref{eq:direct_expectation_expansion}). Because of this, a
-solution estimate in all states of a linear system may require a
-large number of random walks to be performed. Ji and Li recently
-proposed a variation of the direct method in which the random walk
-permutations in Eq~(\ref{eq:mc_walk_permutation}) can be used to
-contribute to the solution tally in all visited states instead of the
-only starting state \citep{ji_reusing_2012}. By doing this, they
-effectively reduce the number of random walks needed to sample all
-desired states in the system.
-
-Their method is based on the observation that while a random walk
-transitions between states in the system, if every state it visited
-could be considered that starting state, then a series a contributions
-to the solution could be generated by a single sequence of random
-numbers while maintaining statistical independence between the
-samples. To demonstrate this, consider the following random walk
-sequence starting in state 2: $2 \rightarrow 3 \rightarrow 1
-\rightarrow 2 \rightarrow 4 \rightarrow 5$. We can reuse these walks
-be recognizing that we also then have a random walk sequence starting
-in state 3 that can be applied to the state 3 estimator: $3
-\rightarrow 1 \rightarrow 2 \rightarrow 4 \rightarrow 5$. We also have
-the same situation for state 1: $1 \rightarrow 2 \rightarrow 4
-\rightarrow 5$, and all future states. Using this sampling scheme , Ji
-and Li observed 1 to 2 orders of magnitude reduction over the standard
-direct method in the number histories required to achieve a specified
-variance in their test problems.
-
-\section{Adjoint Method}
-\label{sec:adjoint_mc}
-An alternative formulation for Monte Carlo matrix inversion is the
-adjoint method. We begin by defining the linear system adjoint to
-Eq~(\ref{eq:linear_problem}):
-\begin{equation}
-  \ve{A}^T \ve{y} = \ve{d}\:,
-  \label{eq:adjoint_linear_problem}
-\end{equation}
-where $\ve{y}$ and $\ve{d}$ are the adjoint solution and source
-respectively and $\ve{A}^T$ is the adjoint operator. We can split this
-equation to mirror Eq~(\ref{eq:richardson_split}) by defining the
-following inner product equivalence \citep{spanier_monte_1969}:
-\begin{equation}
-  \langle \ve{A}^T \ve{x}, \ve{y} \rangle = \langle \ve{x}, \ve{A}
-  \ve{y} \rangle\:.
-  \label{eq:adjoint_operator_product}
-\end{equation}
-With this statement we can then define the split equation:
-\begin{equation}
-  \ve{y} = \ve{H}^T \ve{y} + \ve{d}\:.
-  \label{eq:adjoint_split_system}
-\end{equation}
-As was required for convergence with the direct method using
-Eq~(\ref{eq:richardson_split}), the spectral radius of $\ve{H}$ must
-remain less than 1 as $\ve{H}^T$ contains the same eigenvalues and
-therefore has the same spectral radius. From this definition it
-follows that:
-\begin{equation}
-  \langle \ve{x}, \ve{d} \rangle = \langle \ve{y}, \ve{b} \rangle\:.
-  \label{eq:adjoint_vector_relation}
-\end{equation}
-Using these definitions, we can derive an estimator from the adjoint
-method that will also give the solution vector, $\ve{x}$. As with the
-direct method, we can acquire the adjoint solution by forming the
-Neumann series by writing Eq~(\ref{eq:adjoint_split_system}) as:
-\begin{equation}
-  \ve{y} = (\ve{I} - \ve{H}^T)^{-1} \ve{d}\:,
-  \label{eq:adjoint_split_system_2}
-\end{equation}
-which in turn yields the Neumann series using the adjoint operator:
-\begin{equation}
-  \ve{y} = \sum_{k=0}^{\infty} (\ve{H}^T)^k\ve{d}\:.
-  \label{eq:adjoint_neumann_series}
-\end{equation}
-We expand this summation to again yield a series of transitions that
-can be approximated by a Monte Carlo random walk sequence, this time
-forming the Neumann series in reverse order:
-\begin{equation}
-  y_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-  \sum_{i_k}^{N}h_{i_k,i_{k-1}}\ldots h_{i_2,i_1} h_{i_1,i} d_{i_k}\:.
-  \label{eq:adjoint_neumann_solution}
-\end{equation}
-We can readily build an estimator for the adjoint solution from this
-series expansion, but we instead desire the direct solution. We
-achieve this by using Eq~(\ref{eq:adjoint_vector_relation}) as a
-constraint. Here we have 2 unknowns, $\ve{y}$ and $\ve{d}$ and
-therefore we require a second constraint to close the system. As a
-second constraint we select:
-\begin{equation}
-  \ve{d} = \boldsymbol{\delta}_i\:,
-  \label{eq:adjoint_second_constraint}
-\end{equation}
-where the $k^{th}$ component of the vector $\boldsymbol{\delta}_i$ is
-the Dirac delta function $\delta_{i_k,i}$. If we apply
-Eq~(\ref{eq:adjoint_second_constraint}) to our first constraint
-Eq~(\ref{eq:adjoint_vector_relation}), we get the following convenient
-outcome:
-\begin{equation}
-  \langle \ve{y}, \ve{b} \rangle = \langle \ve{x},
-  \boldsymbol{\delta}_i \rangle = x_i \:,
-  \label{eq:inner_product_constraint}
-\end{equation}
-meaning that if we compute the inner product of the direct source and
-the adjoint solution using a delta function source, we recover the
-direct solution we are looking for
-
-In terms of particle transport, this adjoint method is equivalent to a
-traditional forward method. As a result of using the adjoint system,
-we modify our probabilities and weights using the \textit{adjoint
-  Neumann-Ulam decomposition} of $\ve{H}$:
-\begin{equation}
-  \ve{H}^{T} = \ve{P} \circ \ve{W}\:,
-  \label{eq:adjoint_neumann_ulam}
-\end{equation}
-where now we are forming the decomposition with respect to the
-transpose of $\ve{H}$\footnote{This is sometimes referred to as the
-  adjoint form for real-valued matrices}. We then follow the same
-procedure as the direct method for forming the probability and weight
-matrices in the decomposition. Using the adjoint form, probabilities
-should instead be column-scaled:
-\begin{equation}
-  p_{ij} = \frac{|h_{ji}|}{\sum_j |h_{ji}|}\:,
-  \label{eq:adjoint_probability}
-\end{equation}
-such that we expect to select a new state $j$ from the current state
-in the random walk $j$ by sampling column-wise. Per
-Eq~(\ref{eq:adjoint_neumann_ulam}), the transition weight is then
-defined as:
-\begin{equation}
-  w_{ij} = \frac{h_{ji}}{p_{ij}}\:.
-  \label{eq:adjoint_weight}
-\end{equation}
-Using the decomposition we can then define an expectation value for
-the adjoint method. Given Eq~(\ref{eq:direct_permutation_weight}) as
-the weight generated for a particular random walk permutation as in
-Eq~(\ref{eq:mc_walk_permutation}) and our result from
-Eq~(\ref{eq:inner_product_constraint}) generated by applying the
-adjoint constraints, the contribution to the solution in state $i$
-from a particular random walk permutation is then:
-\begin{equation}
-  X_{\nu} = \sum_{m=0}^k W_{m} \delta_{i,i_m}\:,
-  \label{eq:adjoint_permutation_contribution}
-\end{equation}
-where the Kronecker delta indicates that the tally contributes only in
-the current state and $b_{i_0}$ will be the sampled source starting
-weight. Note here that the estimator in
-Eq~(\ref{eq:adjoint_permutation_contribution}) does not have a
-dependency on the source state as in
-Eq~(\ref{eq:adjoint_permutation_contribution}), providing a remedy for
-the situation in the direct method where we must start a random walk
-in each source state for every permutation such that we may compute a
-contribution for that state. In the adjoint method, we instead tally
-in all states and those of lesser importance will not be visited as
-frequently by the random walk. Finally, the expectation value using
-all permutations is:
-\begin{equation}
-  E\{X\} = \sum_{\nu} P_{\nu} X_{\nu}\:
-  \label{eq:adjoint_expectation_value}
-\end{equation}
-which, if expanded in the same way as the direct method, directly
-recovers the exact solution:
-\begin{equation}
-  \begin{split}
-    E\{X_j\} &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-    \sum_{i_k}^{N} b_{i_0} h_{i_0,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}
-    \delta_{i_k,j} \\ &= x_{j}\:,
-  \end{split}
-  \label{eq:adjoint_expectation_expansion}
-\end{equation}
-therefore also providing an unbiased Monte Carlo estimate of the
-solution.
-
-Like the direct method, we also desire a criteria for random walk
-termination for problems where only an approximate solution is
-necessary. For the adjoint method, we utilize a \textit{relative
-  weight cutoff}:
-\begin{equation}
-  W_f = W_c b_{i_0}\:,
-  \label{eq:relative_weight_cutoff}
-\end{equation}
-where $W_c$ is defined as in the direct method. The adjoint random
-walk will then be terminated after $m$ steps if $W_m < W_f$ as tally
-contributions become increasingly small.
-
-\section{Sequential Monte Carlo}
-\label{sec:sequential_mc}
-The direct and adjoint Neumann-Ulam methods described are limited by a
-convergence rate of $1/\sqrt{N}$ by the Central Limit Theorem where
-$N$ is the number of random walk permutations. In 1962, Halton
-presented a residual Monte Carlo method that moves towards exponential
-convergence rates \citep{halton_sequential_1962} and further refined
-his work some years later \citep{halton_sequential_1994}. Applications
-of his work by the transport community have confirmed convergence
-rates on the order of of $exp(-N)$ \citep{evans_residual_2003}. In
-much the same way as projection methods, Halton's method, sequential
-Monte Carlo, utilizes the adjoint Monte Carlo solver as a means of
-directly reducing the residual vector. He proposed the following
-iterative scheme as a solution to Eq~(\ref{eq:linear_problem})\:
-\begin{subequations}
-  \begin{gather}
-    \ve{r}^k = \ve{b} - \ve{A}\ve{x}^k\:,\\  
-    \ve{A}\boldsymbol{\delta}^{k} = \ve{r}^{k}\:,\\
-    \ve{x}^{k+1} = \ve{x}^k + \boldsymbol{\delta}^{k}\:,
-  \end{gather}
-  \label{eq:sequential_monte_carlo}
-\end{subequations}
-where the correction $\boldsymbol{\delta}$ is computed by the adjoint
-Monte Carlo method. The merits of Halton's approach are immediately
-visible in that we have now broken the binding of the convergence rate
-to the Central Limit Theorem. Here, the Monte Carlo solver is used to
-produce a correction from the residual, analogous to using the
-residual to extract a correction from the search subspace in a
-projection method. By doing this, the Monte Carlo error is bound in
-the correction used to update the solution and therefore does not
-explicitly manifest itself in the overall convergence of the
-solution. The downside of such a method is that if the solution guess
-is poor, then many iterations are required in order to reach
-exponential converge as the Monte Carlo error (and therefore the
-Central Limit Theorem) does dominate in this situation. Therefore, if
-a good initial guess is not available, Halton's method is still
-characterized by poor performance.
-
-\section{Monte Carlo Synthetic-Acceleration}
-\label{sec:mcsa}
-Using the ideas of Halton, Evans and Mosher recently developed a Monte
-Carlo solution method that was not prohibited severely by the quality
-of the initial guess for the system \citep{evans_monte_2009} and later
-applied it more rigorously as a solution mechanism for the radiation
-diffusion equation \citep{evans_monte_2012}. With their new methods,
-they achieved identical numerical results as and marginally better
-performance than conventional Krylov solvers. Their approach was
-instead to use residual Monte Carlo as a synthetic acceleration for a
-stationary method. To derive this method, we begin by splitting the
-operator in Eq~(\ref{eq:linear_problem})
-\begin{equation}
-  \ve{x} = (\ve{I} - \ve{A})\ve{x} + \ve{b}\:.
-  \label{eq:linear_split}
-\end{equation}
-With this we can then define the stationary method
-\textit{Richardson's iteration} as:
-\begin{equation}
-  \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,
-  \label{eq:richardsons_iteration}
-\end{equation}
-which will converge if $\rho(\ve{I} - \ve{A}) < 1$. We then define the
-solution error at the $k^{th}$ iterate relative to the true solution:
-\begin{equation}
-  \delta \ve{x}^k = \ve{x} - \ve{x}^k\:.
-  \label{eq:mcsa_error}
-\end{equation}
-Subtracting Eq~(\ref{eq:richardsons_iteration}) from
-Eq~(\ref{eq:linear_split}) we get:
-\begin{equation}
-  \delta \ve{x}^{k+1} = (\ve{I} - \ve{A})\delta \ve{x}^k\:.
-  \label{eq:mcsa_setup_1}
-\end{equation}
-Subtracting from this $(\ve{I} - \ve{A})\delta \ve{x}^{k+1}$ yields:
-\begin{equation}
-  \begin{split}
-    \ve{A}\delta \ve{x}^{k+1} &= (\ve{I} -
-    \ve{A})(\ve{x}^{k+1}-\ve{x}^{k}) \\ &= \ve{r}^{k+1}\:.
-    \label{eq:mcsa_setup_2}
-  \end{split}
-\end{equation}
-Using this, we define the following scheme that will converge in one
-iteration if $\ve{A}$ is inverted exactly.
-\begin{subequations}
-  \begin{gather}
-    \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,\\
-    \ve{A} \delta \ve{x}^{k+1} = \ve{r}^{k+1}\:,\\
-    \ve{x} = \ve{x}^{k+1} + \delta \ve{x}^{k+1}\:.
-  \end{gather}
-  \label{eq:mcsa_setup_3}
-\end{subequations}
-However, $\ve{A}$ is only approximately inverted by our numerical
-methods and therefore we instead pose an iterative scheme in which the
-Monte Carlo solvers are used to invert the operator. The
-\textit{Fixed-Point Monte Carlo Synthetic-Acceleration} (MCSA) method
-is defined as:
-\begin{subequations}
-  \begin{gather}
-    \ve{x}^{k+1/2} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,\\
-    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\
-    \ve{A}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:,\\
-    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,
-  \end{gather}
-  \label{eq:mcsa}
-\end{subequations}
-where the adjoint Monte Carlo method is used to generate the solution
-correction from the residual. Using Monte Carlo in this way achieves
-the same effect as Halton's method, decoupling its convergence rate
-from the overall convergence rate of the method. Here, the approximate
-Monte Carlo solution is not driven to a particular convergence as it
-merely supplies a correction for the initial guess generated by
-Richardson's iteration. Rather, only a set number of histories are
-required using the adjoint method to generate the correction.
-
-In addition to the Monte Carlo solver parameters dictating the number
-of histories and weight cutoff, the outer MCSA iterations also have
-the following stopping criteria:
-\begin{equation}
-  ||\ve{r}||_\infty < \epsilon \ ||\ve{b}||_\infty\:,
-  \label{eq:mcsa_stopping_criteria}
-\end{equation}
-where $\epsilon$ is a user-defined parameter. We therefore have 3
-parameters to tune in an MCSA implementation: the number of Monte
-Carlo histories computed in the adjoint solve during each MCSA
-iteration, the weight cutoff for those histories, and the total MCSA
-convergence tolerance as specified by $\epsilon$.
-
-\subsection{Preconditioning MCSA}
-\label{sec:stochastic_preconditioning}
-In most cases, at least a minimal amount of \textit{preconditioning}
-of the linear system will be required in order to use the class of
-stochastic methods described. Although these methods have no symmetry
-requirements for convergence, they do require that the spectral radius
-of the iteration matrix be less than one. To achieve this, a Jacobi
-preconditioner, a form of left preconditioning, is used such that the
-preconditioning matrix $\ve{M}$ is:
-\begin{equation}
-  \ve{M} = diag(\ve{A})\:,
-  \label{eq:jacobi_preconditioner}
-\end{equation}
-such that its application means we are instead solving the following
-linear system:
-\begin{equation}
-  \ve{M}^{-1}\ve{A}\ve{x} = \ve{M}^{-1}\ve{b}\:.
-  \label{eq:jacobi_precond_linear_problem}
-\end{equation}
-Next, we can apply MCSA to solve
-Eq~(\ref{eq:jacobi_precond_linear_problem}): 
-\begin{subequations}
-  \begin{gather}
-    \ve{x}^{k+1/2} = (\ve{I} - \ve{M}^{-1}\ve{A})\ve{x}^k +
-    \ve{M}^{-1}\ve{b}\:,\\ \ve{r}^{k+1/2} = \ve{M}^{-1}\ve{b} -
-    \ve{M}^{-1}\ve{A}\ve{x}^{k+1/2}\:,\\ \ve{M}^{-1}\ve{A}\delta\ve{x}^{k+1/2}
-    = \ve{r}^{k+1/2}\:,\\ \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta
-    \ve{x}^{k+1/2}\:.
-    \label{eq:jacobi_preconditioned_mcsa}
-  \end{gather}
-\end{subequations}
-Choosing Jacobi preconditioning with MCSA is advantageous for several
-reasons. First, $\rho(\ve{I} - \ve{M}^{-1}\ve{A}) < 1$ is true for all
-$\ve{A}$ and is easy to formulate because the inversion of $\ve{M}$ is
-trivial. Second, because the adjoint Monte Carlo method used within
-MCSA to compute the correction operates on a linear problem with the
-preconditioned operator, then $\ve{H}$ in the adjoint solver will have
-a zero term in each of its diagonal elements, thereby eliminating all
-in-state transitions during the random walk sequence. Because of this,
-Jacobi preconditioning should always be performed, regardless of any
-other preconditioning that is applied to the system.
-
-\section{Parallelization of Stochastic Methods}
-\label{sec:parallel_stochastic_methods}
 For Monte Carlo methods, and in particular MCSA, to be viable at the
@@
 
-\subsection{Domain Decomposition for Monte Carlo}
-\label{subsec:msod}
+%%---------------------------------------------------------------------------%%
+\section{Domain Decomposition for Monte Carlo}
+\label{sec:msod}
 As observed in the discussion on parallel Krylov methods, large-scale
@@
 
-\subsubsection{Multiple-Set Overlapping-Domain Decomposition}
-\label{subsubsec:msod}
+\subsection{Multiple-Set Overlapping-Domain Decomposition}
+\label{subsec:msod}
 In 2010, Wagner and colleagues developed the \textit{multiple-set
@@
 
-\subsubsection{Domain-to-Domain Communication}
-\label{subsubsec:mc_comm_costs}
+%%---------------------------------------------------------------------------%%
+\section{Domain-to-Domain Communication}
+\label{sec:mc_comm_costs}
 With Wagner's work, we observed that domain overlap implemented with a
@@
 
-\subsubsection{Load Balancing Concerns}
-\label{subsubsec:mc_load_balancing}
+\subsection{Load Balancing Concerns}
+\label{subsec:mc_load_balancing}
 Although domain decomposition was shown to be efficient in a perfectly
@@
 
-\subsubsection{Reproducible Domain Decomposed Results}
-\label{subsubsec:reproducible_mc}
+\subsection{Reproducible Domain Decomposed Results}
+\label{subsec:reproducible_mc}
 The 2006 work of Brunner is notable in that the Monte Carlo codes used
@@
 
-\subsection{Parallel Adjoint Method}
-\label{subsec:parallel_adjoint}
+\section{Parallel Adjoint Method}
+\label{sec:parallel_adjoint}
 We can take much of what was learned from the survey of parallel Monte
@@
 
-\subsection{Parallel MCSA}
-\label{subsec:parallel_mcsa}
+\section{Parallel MCSA}
+\label{sec:parallel_mcsa}
 With a parallel adjoint Neumann-Ulam solver implementation, the

### 2013-03-21 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
adding mandc paper to dissertation

@@
     \scalebox{1.5}{
-      \input{chapters/mc_background/msod_example.pdftex_t} }
+      \input{chapters/parallel_mc/msod_example.pdftex_t} }
   \end{center}
@@
 domain decomposed Monte Carlo setting reduces the amount of nearest
-neighbor communication that must occur during transport. This
+neighbor communication that must occur during transport.
+
+To date, parallel Neumann-Ulam methods have been limited to full
+domain replication with parallelism exploited through individual
+histories \cite{alexandrov_efficient_1998}. In reactor physics Monte
+Carlo applications, however, domain decomposition has been identified
+as a key principle in moving forward in high performance computing to
+enable higher fidelity simulations
+\cite{brunner_comparison_2006,siegel_analysis_2012}. To accomplish
+this, we recognize from the literature that stochastic histories must
+be transported from domain to domain as the simulation progresses and
+they transition to states that are not in the local domain. Because we
+have chosen a domain decomposition strategy in a parallel environment,
+this means that communication of these histories must occur between
+compute nodes owning neighboring pieces of the global domain. We wish
+to characterize this communication not only because communication is
+in general expensive, but also because these nearest-neighbor
+communication sequences, specifically, have poor algorithmic strong
+scaling \cite{gropp_high-performance_2001}.
+
+The purpose of this study is to provide a simple, analytic theory
+based on the properties of the linear system that will allow for
+estimates of the domain decomposed behavior of the adjoint
+Neumann-Ulam method. When solving problems where the linear operator
+is symmetric, a host of analytic theories exist based on the
+eigenvalue spectrum of the operator that characterize their behavior
+in the context of deterministic linear solvers. Using past work, these
+theories are adapted to the domain decomposed adjoint Neumann-Ulam
+method using the one-speed, two-dimensional neutron diffusion
+equation. In this paper we describe the adjoint Neumann-Ulam Monte
+Carlo method followed by a presentation of the model problem. Using
+the linear system generated by the discretization of the model
+problem, we use a spectral analysis to generate analytic relations for
+the eigenvalues of the operator based on system parameters. Using the
+eigenvalue spectra, we then build relationships to characterize the
+transport of stochastic histories in a decomposed domain and the
+fraction of histories that leak from a domain and will therefore have
+to be communicated. Finally, we compare these analytic results to
+numerical experiments conducted with the model problem and draw
+conclusions looking towards future work.
+
+\subsection{Model Problem}
+\label{subsec:model_problem}
+For our numerical experiments, we choose the one-speed,
+two-dimensional neutron diffusion equation as a model problem
+\cite{duderstadt_nuclear_1976}:
+\begin{equation}
+  -\boldsymbol{\nabla} \cdot D \boldsymbol{\nabla} \phi + \Sigma_a
+  \phi = S\:,
+  \label{eq:diffusion_eq}
+\end{equation}
+where $\phi$ is the neutron flux, $\Sigma_a$ is the absorption cross
+section, and $S$ is the source of neutrons. In addition, $D$ is the
+diffusion coefficient defined as:
+\begin{equation}
+  D = \frac{1}{3 ( \Sigma_t - \bar{\mu}\Sigma_s )}\:,
+  \label{eq:diffusion_coeff}
+\end{equation}
+where $\Sigma_s$ is the scattering cross section, $\Sigma_t = \Sigma_a
++ \Sigma_s$ is the total cross section, and $\bar{\mu}$ is the cosine
+of the average scattering angle. For simplicity, we will take
+$\bar{\mu} = 0$ for our analysis giving $D=(3 \Sigma_t)^{-1}$. In
+addition, to further simplify we will assume a homogeneous domain such
+that the cross sections remain constant throughout. Doing this permits
+us to rewrite Eq~(\ref{eq:diffusion_eq}) as:
+\begin{equation}
+  -D \boldsymbol{\nabla}^2 \phi + \Sigma_a \phi = S\:.
+  \label{eq:diffusion_eq_simple}
+\end{equation}
+
+We choose a finite difference scheme on a square Cartesian grid to
+discretize the problem. For the Laplacian, we choose the 9-point
+stencil shown in Figure~\ref{fig:stencil} over a grid of size $h$
+\cite{leveque_finite_2007}:
+\begin{multline}
+  \nabla^2_9\phi = \frac{1}{6h^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j}
+    + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ +
+    \phi_{i-1,j+1} + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20
+    \phi_{i,j}]\:.
+  \label{eq:nine_point_stencil}
+\end{multline}
+\begin{figure}[t!]
+  \begin{center}
+    \scalebox{1.25}{\input{chapters/parallel_mc/stencil.pdftex_t}}
+  \end{center}
+  \caption{\textbf{Nine-point Laplacian stencil.}}
+  \label{fig:stencil}
+\end{figure}
+We then have the following linear system to solve:
+\begin{multline}
+  -\frac{1}{6h^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j} + 4
+    \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ + \phi_{i-1,j+1}
+    + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20 \phi_{i,j}] + \Sigma_a
+  \phi_{i,j} = s_{i,j}\:,
+  \label{eq:fd_system}
+\end{multline}
+and in operator form:
+\begin{equation}
+  \ve{D}\boldsymbol{\phi}=\ve{s}\:,
+  \label{eq:operator_system}
+\end{equation}
+where $\ve{D}$ is the diffusion operator, $\ve{s}$ is the source in
+vector form and $\boldsymbol{\phi}$ is the vector of unknown fluxes.
+
+To close the system, a set of boundary conditions is required. In the
+case of a non-reentrant current condition applied to all global
+boundaries of the domain, we choose the formulation of Duderstadt by
+assuming the flux is zero at some ghost point beyond the
+grid. Consider for example the equations on the $i=0$ boundary of the
+domain:
+\begin{multline}
+  -\frac{1}{6h^2}[4 \phi_{-1,j} + 4 \phi_{1,j} + 4 \phi_{0,j-1} +
+    4 \phi_{0,j+1} + \phi_{-1,j-1}\\ + \phi_{-1,j+1} + \phi_{1,j-1} +
+    \phi_{1,j+1} - 20 \phi_{0,j}] + \Sigma_a \phi_{0,j} = s_{0,j}\:.
+  \label{eq:x_min_bnd}
+\end{multline}
+Here we note some terms where $i=-1$ and therefore are representative
+of grid points beyond the boundary of the domain. We set the flux at
+these points to be zero, giving a valid set of equations for the $i=0$
+boundary:
+\begin{multline}
+  -\frac{1}{6h^2}[4 \phi_{1,j} + 4 \phi_{0,j-1} + 4 \phi_{0,j+1}
+    \\ + \phi_{-1,j+1} + \phi_{1,j-1} + \phi_{1,j+1} - 20 \phi_{0,j}]
+  + \Sigma_a \phi_{0,j} = s_{0,j}\:.
+  \label{eq:x_min_bnd_2}
+\end{multline}
+We repeat this procedure for the other boundaries of the domain. For
+reflecting boundary conditions, the net current across a boundary is
+zero.
+
+\subsection{Spectral Analysis}
+\label{subsec:spectral_analysis}
+The convergence of the Neumann series in
+Eq~(\ref{eq:adjoint_neumann_series}) approximated by the Monte Carlo
+solver is dependent on the eigenvalues of the iteration matrix. We
+will compute these eigenvalues by assuming eigenfunctions of the form
+\cite{leveque_finite_2007}:
+\begin{equation}
+  \Phi_{p,q}(x,y) = e^{2 \pi \imath p x} e^{2 \pi \imath q y}\:,
+  \label{eq:eigenfunction_form}
+\end{equation}
+where different combinations of $p$ and $q$ represent the different
+eigenmodes of the solution. As these are valid forms of the solution,
+then the action of the linear operator on these eigenfunctions should
+give the eigenvalues of the matrix as they exist on the unit circle in
+the complex plane.
+
+\subsection{Iteration Matrix Spectrum}
+\label{subsec:iteration_spectrum}
+For the model problem, we first compute the eigenvalues for the
+diffusion operator $\ve{D}$ by applying the operator to the
+eigenfunctions and noting that $x=ih$ and $y=jh$:
+\begin{multline}
+  \ve{D}\Phi_{p,q}(x,y) = \lambda_{p,q}(\ve{D})
+  =\\ -\frac{D}{6h^2}\Big[4 e^{-2 \pi \imath p h} + 4 e^{2 \pi \imath
+      p h} + 4 e^{-2 \pi \imath q h} + 4 e^{2 \pi \imath q h} + e^{-2
+      \pi \imath p h} e^{-2 \pi \imath q h} \\ + e^{-2 \pi \imath p h}
+    e^{2 \pi \imath q h} + e^{2 \pi \imath p h} e^{-2 \pi \imath q h}
+    + e^{2 \pi \imath p h} e^{2 \pi \imath q h} - 20\Big] + \Sigma_a
+  \:.
+  \label{eq:deriv_diff_1}
+\end{multline}
+Using Euler's formula, we can collapse the exponentials to
+trigonometric functions:
+\begin{equation}
+  \lambda_{p,q}(\ve{D}) = -\frac{D}{6h^2}[ 8 \cos(\pi p h) + 8
+    \cos(\pi q h) + 4 \cos(\pi p h) \cos(\pi q h) - 20] + \Sigma_a\:.
+  \label{eq:deriv_diff_2}
+\end{equation}
+
+As Eq~(\ref{eq:diffusion_eq}) is diagonally dominant, Jacobi
+preconditioning is sufficient to reduce the spectral radius of the
+iteration matrix below unity and therefore ensure convergence of the
+Neumann series. The preconditioner in this case is then $\ve{M} =
+diag(\ve{D})$ such that we are solving the following linear system:
+\begin{equation}
+  \ve{M}^{-1} \ve{D} \boldsymbol{\phi} = \ve{M}^{-1} \ve{s}\:.
+  \label{eq:precond_diffsion}
+\end{equation}
+The operator $\ve{M}^{-1} \ve{D}$ is merely the original diffusion
+operator with each row scaled by the diagonal component. As we have
+defined a homogeneous domain, the scaling factor, $\alpha$, is the
+same for all rows in the operator and defined as the $\phi_{i,j}$
+coefficient from Eq~(\ref{eq:fd_system}):
+\begin{equation}
+  \alpha = \Bigg[\frac{10 D}{3 h^2} + \Sigma_a\Bigg]^{-1}\:.
+  \label{eq:jacobi_scaling}
+\end{equation}
+Using this coefficient, we then have the following spectrum of
+preconditioned eigenvalues:
+\begin{equation}
+  \lambda_{p,q}(\ve{M}^{-1} \ve{D}) = \alpha \lambda_{p,q}(\ve{D})\:.
+  \label{eq:preconditioned_eigenvalues}
+\end{equation}
+
+The spectral radius of the iteration matrix is obtained by seeking its
+largest eigenvalue. As with the diffusion operator, we can use the
+same analysis techniques to find the eigenvalues for the iteration
+matrix. We use a few simplifications by noting that if the Jacobi
+preconditioned iteration matrix is $\ve{H} = \ve{I} -
+\ve{M}^{-1}\ve{D}$, the we except all terms on the diagonal of the
+iteration matrix to be zero such that we have the following stencil:
+\begin{equation}
+  \ve{H}\boldsymbol{\phi} = \frac{\alpha D}{6h^2}[4 \phi_{i-1,j}
+    + 4 \phi_{i+1,j} + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} +
+    \phi_{i-1,j-1} + \phi_{i-1,j+1} + \phi_{i+1,j-1} +
+    \phi_{i+1,j+1}]\:.
+  \label{eq:iteration_stencil}
+\end{equation}
+Inserting the eigenfunctions defined by
+Eq~(\ref{eq:eigenfunction_form}) we get:
+\begin{multline}
+  \lambda_{p,q}(\ve{H}) = \frac{\alpha D}{6h^2}\Big[4 e^{-2 \pi \imath p
+      h} + 4 e^{2 \pi \imath p h} + 4 e^{-2 \pi \imath q h} + 4 e^{2
+      \pi \imath q h} + e^{-2 \pi \imath p h} e^{-2 \pi \imath q h}
+    \\ + e^{-2 \pi \imath p h} e^{2 \pi \imath q q} + e^{2 \pi \imath
+      p h} e^{-2 \pi \imath q h} + e^{2 \pi \imath p h} e^{2 \pi
+      \imath q h}\Big]\:,
+  \label{eq:iteration_deriv}
+\end{multline}
+which simplifies to:
+\begin{equation}
+  \lambda_{p,q}(\ve{H}) = \frac{\alpha D}{6h^2}[ 8 \cos(\pi p h) + 8
+    \cos(\pi q h) + 4 \cos(\pi p h) \cos(\pi q h)]\:,
+  \label{eq:iteration_spectrum}
+\end{equation}
+giving the eigenvalue spectrum for the Jacobi preconditioned iteration
+matrix. To find these maxium eigenvalue,
+Eq~(\ref{eq:preconditioned_eigenvalues}) is plotted as a function of
+$p$ with $p=q$ in Figure~\ref{fig:diffusion_spectrum} for various
+values of $\Sigma_a$.
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=5in,clip]{chapters/parallel_mc/diffusion_spectrum.png}
+  \end{center}
+  \caption{\textbf{Eigenvalue spectra for the diffusion equation.}}
+  \label{fig:diffusion_spectrum}
+\end{figure}.
+ We find that the maximum eigenvalue exists when $p=q=0$,
+giving the following for the spectral radius of the Jacobi
+preconditioned iteration matrix:
+\begin{equation}
+  \rho(\ve{H}) = \frac{10 \alpha D}{3 h^2}\:.
+  \label{eq:iteration_radius}
+\end{equation}
+
+\subsection{Neumann Series Convergence}
+\label{subsec:neumann_convergence}
+The adjoint Monte Carlo method is effectively an approximation to a
+stationary method. Stationary methods for linear systems arise from
+splitting the operator in Eq~(\ref{eq:linear_problem}) and iterating:
+\begin{equation}
+  \ve{x}^{k+1} = \ve{H}\ve{x}^k + \ve{c}\:,
+  \label{eq:linear_iterative_method}
+\end{equation}
+with $k \in \mathbb{Z}^+$ defined as the \textit{iteration
+  index}. Defining $\ve{e}^k = \ve{u}^k - \ve{u}$ as the solution
+error at the $k^{th}$ iterate, the error after $k$ iterations is then:
+\begin{equation}
+  \ve{e}^{k} = \ve{H}^k\ve{e}^0\:. 
+  \label{eq:linear_k_iter_error}
+\end{equation}
+By assuming $\ve{H}$ is diagonalizable \cite{leveque_finite_2007}, we
+then have:
+\begin{equation}
+  ||\ve{e}^{k}||_2 \leq \rho(\ve{H})^k ||\ve{e}^0||_2\:.
+  \label{eq:linear_k_iter_norm3}
+\end{equation}
+
+In the adjoint Neumann-Ulam method, $k$ iterations, equivalent to $k$
+applications of the iteration matrix, are approximated by a random
+walk of average length $k$ to yield the summation in
+Eq~(\ref{eq:adjoint_neumann_solution})
+\cite{dimov_new_1998,danilov_asymptotic_2000}. This random walk
+length, or the number of transitions before the termination of a
+history (either by the weight cutoff, absorption, or exiting the
+global domain) is therefore approximately the number of stationary
+iterations required to converge to the specified tolerance. In the
+case of the adjoint Neumann-Ulam method, no such tolerance exists,
+however, we have specified a weight cutoff, $W_c$, that determines
+when low-weight histories will be prematurely terminated as their
+contributions are deemed minute. After $k$ iterations, a stationary
+method is terminated as the error has reached some fraction,
+$\epsilon$, of the initial error:
+\begin{equation}
+  ||\ve{e}^{k}||_2 = \epsilon ||\ve{e}^0||_2\:.
+  \label{eq:linear_k_iter_norm4}
+\end{equation}
+Per Eq~(\ref{eq:linear_k_iter_norm3}), we see that this fraction is
+equivalent to $\epsilon = \rho(\ve{H})^k$. For the adjoint
+Neumann-Ulam method, if we take this fraction to be the weight cutoff,
+a measure of how accurately the contributions of a particular history
+to the solution are tallied, we then have the following relationship
+for $k$:
+\begin{equation}
+  k = \frac{ \log(W_c) }{ \log( \rho(\ve{H}) ) }\:.
+  \label{eq:analytic_k}
+\end{equation}
+This then gives us a means to estimate the length of the random walks
+that will be generated from a particular linear operator based on the
+eigenvalues of its iteration matrix (independent of the linear
+operator splitting chosen) and based on the weight cutoff parameter
+used in the Neumann-Ulam method.
+
+\subsubsection{Domain Leakage Approximations}
+\label{subsubsec:domain_leak_approx}
+In a domain decomposed situation, not all histories will remain within
+the domain they started in and must instead be communicated. This
 communication, expected to be expensive, was analyzed by Siegel and
-colleagues for idealized, load balanced situations without domain
-overlap \citep{siegel_analysis_2012}. They arrive at a few empirical
-relationships and generate a numerical experiment that provides
-insight into the feasibility of large scale domain decomposed Monte
-Carlo as well as the relationship of communication to system
-parameters. In their work, they define a simple 3-dimensional system
-subdivded into Cartesian block domains with each processor containing
-a single domain. Given this geometry, a particle can either be
-absorbed in its current domain or be transported to one of the 6
-adjacent domains owned by another process.
-
-To quantify the number of particles that leak out of the local domain
-they define a leakage fraction, $\lambda$, as:
+colleagues for idealized, load balanced situations for full nuclear
+reactor core Monte Carlo simulations \cite{siegel_analysis_2012}.  To
+quantify the number of particles that leak out of the local domain
+they define a leakage fraction, $\Lambda$, as:
 \begin{equation}
-  \lambda =
-  \frac{average\ \#\ of\ particles\ leaving\ local\ domain}{total\ 
-    of\ \#\ of\ particles\ starting\ in\ local\ domain}\:.
+  \Lambda = \frac{average\ \#\ of\ particles\ leaving\ local\ domain}
+          {total\ of\ \#\ of\ particles\ starting\ in\ local\ domain}\:.
           \label{eq:leakage_fraction}
 \end{equation}
-For their studies, they assumed that the value of $\lambda$ was
-dependent on the total cross section of the system and the same for
-all processes for simplicity and load balancing. Using this
-relationship and their simple geometry, they were able to relate the
-amount of compute time required to transport all source particles to
-the number of source particles, the leakage fraction, the latency and
-bandwidth of communication, the ratio of local to global geometry
-size, and the total cross section of the system.
+For their studies, Siegel and colleagues assumed that the value of
+$\Lambda$ was dependent on the total cross section of the system via
+the Wigner rational approximation. Outlined more thoroughly by Hwang's
+chapter in \cite{azmy_nuclear_2010}, we will use both the Wigner
+rational approximation and the mean chord approximation as a means to
+estimate the leakage fraction.
 
-What they found was that there was indeed an optimum ratio of local
-domain to global domain size that results in a minimum total compute
-time. They also found that this ratio was only dependent on the
-latency and bandwidth constraints of the parallel machine. We expect
-this because nearest neighbor communication patterns scale with the
-volume and surface sizes of the local domains. The larger the volume,
-the more compute time is spent within the local volume and therefore
-at some point you are limited by the latency of the system, or the
-time it takes to receive the data to operate on. However, at a certain
-point, when the surface of those volumes becomes large enough, the
-amount of data that must be communicated grows large enough that it
-may overcome the bandwidth limitations of the system, resulting in a
-decrease in performance. In their numerical experiments, they used the
-above geometry and leakage rates based on that computed for a reactor
-system for a single material problem with a total cross section
-equivalent to that of homogenized light water reactor. Given the
-optimal local domain size, they observed for load balanced situations
-that nearest neighbor communication costs during transport are a small
-fraction of total compute time, leading to the conclusion that domain
-decomposition is feasible for large-scale Monte Carlo calculations.
+In the case of domain decomposed linear operator equations, we can use
+diffusion theory to estimate the optical thickness of a domain in the
+decomposition and the corresponding leakage fraction in terms of
+properties of the linear operator and the discretization. To begin we
+must first calculate the mean distance a Monte Carlo history will move
+in the grid by computing the mean squared distance of its movement
+along the chord of length $l$ defined across the domain. After a
+single transition a history will have moved a mean squared distance
+of:
+\begin{equation}
+  \langle \bar{r_1^2} \rangle = (n_s h)^2\:,
+  \label{eq:step_1_length}
+\end{equation}
+where $h$ is the size of the discrete grid elements along the chord
+and $n_s$ is the number of grid elements a history will move on
+average every transition. For our diffusion model problem, $n_s$ would
+equate to the expected number of states in the $i$ (or $j$ as the problem is
+symmetric) direction that a history will move in a single
+transition and is dependent on the stencil used for the
+discretization. After $k$ transitions in the random walk, the history
+will have moved a mean squared distance of:
+\begin{equation}
+  \langle \bar{r_k^2} \rangle = k (n_s h)^2\:.
+  \label{eq:step_k_length}
+\end{equation}
+If our chord is of length $l$ and there are $n_i$ grid elements (or
+states to which a history may transition) along that chord, then $h =
+l / n_i$ giving:
+\begin{equation}
+  \langle \bar{r_k^2} \rangle = k \Bigg(\frac{n_s l}{n_i}\Bigg)^2\:.
+  \label{eq:step_k_length_sub}
+\end{equation}
+From diffusion theory, we expect the average number of interactions
+along the chord to be:
+\begin{equation}
+  \tau = \frac{l}{2 d \sqrt{\langle \bar{r_k^2} \rangle}}\:,
+  \label{eq:optical_thickness_1}
+\end{equation}
+where $d$ is the dimensionality of the problem and $\sqrt{\langle
+  \bar{r_k^2} \rangle}$ is effectively the mean free path of the Monte
+Carlo history in the domain. We can readily interpret $\tau$ to be the
+\textit{effective optical thickness} of a domain of length
+$l$. Inserting Eq~(\ref{eq:step_k_length_sub}) we arrive at:
+\begin{equation}
+  \tau = \frac{n_i}{2 d n_s \sqrt{k}}\:,
+  \label{eq:optical_thickness_2}
+\end{equation}
+which if expanded with Eq~(\ref{eq:analytic_k}) gives us the final
+relation for the effective optical thickness:
+\begin{equation}
+  \tau = \frac{n_i}{2 d n_s}
+  \sqrt{\frac{\log(\rho(\ve{H}))}{\log(W_c)}}\:.
+  \label{eq:optical_thickness_3}
+\end{equation}
 
-\subsection{Load Balancing Concerns}
-\label{subsec:mc_load_balancing}
+For optically thin domains, we expect that most histories will be
+communicated, while optically thick domains will leak the fraction of
+histories that did not interact within. Using the optical thickness
+defined in Eq~(\ref{eq:optical_thickness_3}), we can then complete the
+leakage approximations by defining the bounds of $\tau \rightarrow 0,
+\Lambda \rightarrow 1$ and $\tau \rightarrow \infty, \Lambda
+\rightarrow \tau^{-1}$.  With these bounds we can then define the
+leakage fraction out of a domain for the adjoint Neumann-Ulam method
+using the Wigner rational approximation:
+\begin{equation}
+  \Lambda = \frac{1}{1+\tau}\:,
+  \label{eq:wigner_domain_leakage}
+\end{equation}
+and using the mean-chord approximation:
+\begin{equation}
+  \Lambda = \frac{1-e^{-\tau}}{\tau}\:.
+  \label{eq:mean_chord_domain_leakage}
+\end{equation}
+Here, the leakage fraction is explicitly bound to the eigenvalues of
+the iteration matrix, the size of the domain, the content of the
+discretization stencil, and the weight cutoff selected to terminate
+low weight histories.
+
+\subsection{Numerical Experiments}
+\label{subsec:numerical_experiments}
+To test the relationships developed by the spectral analysis, we form
+two simple numerical experiments using the diffusion model problem:
+one to measure the length of the random walks as a function of the
+iteration matrix eigenvalues, and one to measure the domain leakage
+fraction as a function of the iteration matrix eigenvalues and the
+discretization properties. Before doing this, we verify our
+computation of the spectral radius of the iteration matrix by
+numerically computing the largest eigenvalue of the diffusion operator
+using an iterative eigenvalue solver. For this verification, a $100
+\times 100$ square grid with $h=0.01$, $h=0.1$, and $h=1.0$ and the
+absorption cross varied from 0 to 100 while the scattering cross
+section was fixed at unity. Figure~\ref{fig:measured_spec_rad} gives
+the measured spectral radius of the iteration matrix and the computed
+spectral radius for the preconditioned diffusion operator using
+Eq~(\ref{eq:iteration_radius}) as function of the absorption to
+scattering ratio $(\Sigma_a / \Sigma_s)$. Excellent agreement was
+observed between the analytic and numerical results with all data
+points computed within the tolerance of the iterative eigenvalue
+solver.
+\begin{figure}[ht!]
+  \begin{spacing}{1.0}
+    \begin{center}
+      \includegraphics[width=4in,clip]{chapters/parallel_mc/measured_spec_rad.png}
+    \end{center}
+    \caption{Measured and analytic preconditioned diffusion operator
+      spectral radius as a function of the absorption cross section to
+      scattering cross section ratio. Values of $h=0.01$, $h=0.1$, and
+      $h=1.0$ were used. The red data was computed numerically by an
+      eigensolver while the black dashed data was generated by
+      Eq~(\ref{eq:iteration_radius}).}
+    \label{fig:measured_spec_rad}
+  \end{spacing}
+\end{figure}
+
+\subsubsection{Random Walk Length}
+\label{subsubsec:walk_length}
+With the spectral data in hand, we can go about setting up an
+experiment to measure the length of the random walks generated by the
+adjoint Neumann-Ulam solver. To do this, we again use a $100 \times
+100$ square grid with $h=0.1$ and the absorption cross varied from 0
+to 100 while the scattering cross section was fixed at unity. Three
+weight cutoff values of \sn{1}{-2}, \sn{1}{-4}, and \sn{1}{-8} were
+used with 10,000 histories generated by a point source of strength 1
+in the center of the domain. For each of the histories, the number of
+transitions made was tallied to provide an effective value of $k$ for
+each history. This value was then averaged over all histories to get a
+measured value of $k$ for the particular operator. On the left,
+Figure~\ref{fig:measured_length} presents these measurements as well
+as the analytic result computed by Eq~(\ref{eq:analytic_k}) as a
+function of the iteration matrix spectral radius, $\rho(\ve{H})$. On
+the right, Figure~\ref{fig:measured_length} gives the relative error
+between the predicted and observed results. We note good qualitative
+agreement between the measured and analytic results. However, we
+observe a larger relative error for both long and short random walks.
+\begin{figure}[ht!]
+  \begin{spacing}{1.0}
+    \begin{center}
+      \includegraphics[width=6.5in,clip]{chapters/parallel_mc/measured_length.png}
+    \end{center}
+    \caption{Measured and analytic random walk length as a function of
+      the iteration matrix spectral radius. The weight cutoff was
+      varied with \sn{1}{-2}, \sn{1}{-4}, and \sn{1}{-8}. In the left
+      plot, the red data was computed numerically by an adjoint
+      Neumann-Ulam implementation while the black dashed data was
+      generated by Eq~(\ref{eq:analytic_k}). In the right plot, the
+      relative error between the predicted and measured results is
+      presented for each weight cutoff.}
+    \label{fig:measured_length}
+  \end{spacing}
+\end{figure}
+
+\subsubsection{Domain Leakage}
+\label{subsubsec:domain_leakage}
+Finally, we seek to measure the leakage from a domain in a domain
+decomposed Monte Carlo calculation and assess the quality of our
+analytic relation for the optical thickness of domain and the
+associated leakage approximations. For this experiment, a square grid
+with $h=0.1$ was decomposed into 9 square domains, 3 in each cardinal
+direction with measurements occurring in the central domain without
+boundary grid points. For the cross sections, the absorption cross
+section was varied from 1 to 100 while the scattering cross section
+was set to zero to create a purely absorbing environment with weight
+cutoff of \sn{1}{-4}. The optical thickness of these domains will vary
+as a function of the absorption cross section if the other parameters
+are fixed. To compute the optical thickness, along with the spectral
+radius as given by Eq~(\ref{eq:iteration_radius}), we also need the
+parameters $n_i$ and $n_s$ which respectively describe the typical
+domain length and the average number of states moved along that
+typical length per history transition. For our grid above, the domains
+are varied in size with $50 \times 50$, $100 \times 100$, and $200
+\times 200$ cells giving $n_i=50$, $n_i=100$, and $n_i=200$ grid
+points or states along the typical length of the domain
+respectively. Looking at the Laplacian stencil in
+Eq~(\ref{eq:nine_point_stencil}), we see that all history transitions
+will only move a single state in either the $i$ or $j$ directions due
+to the symmetry of the problem. Furthermore, if we choose the $i$
+direction, not all states we will transition to will move the history
+in that direction. Therefore, we look to the definition of the
+iteration matrix in Eq~(\ref{eq:iteration_stencil}) and the definition
+of the adjoint probability matrix in Eq~(\ref{eq:adjoint_probability})
+to estimate the $n_s$ parameter. For a particular transition starting
+at state $(i,j)$, 6 of the 8 possible new states in the stencil move
+the history in $i$ direction with relative coefficients of 4 for
+moving in the $(\pm i,0)$ direction and of 1 for moving in the $(\pm
+i,\pm j)$. These coefficients dictate the frequency those states are
+visited relative to the others. For those 6 states we can visit along
+the typical length, their sum is 12 out of the total 20 for the
+coefficients for all possible states with their ratio giving $n_s =
+\frac{3}{5}$.
+
+To compute the leakage fraction numerically, \sn{3}{5} histories were
+sampled from a uniform source of strength unity over the global
+domain. At the start of a stage of histories, the number of histories
+starting in the center domain was computed and as the stage
+progressed, the number of histories that exited that domain was
+tallied with the ratio of the two numbers providing a numerical
+measure for the leakage fraction. Figure~\ref{fig:measured_leakage}
+gives the domain leakage measurements for the domain in the center of
+the global grid as well as the analytic result computed by
+Eqs~(\ref{eq:wigner_domain_leakage}) and
+(\ref{eq:mean_chord_domain_leakage}) as a function of the iteration
+matrix spectral radius.
+\begin{figure}[ht!]
+  \begin{spacing}{1.0}
+    \begin{center}
+      \includegraphics[width=4.5in,clip]{chapters/parallel_mc/leakage_variation.png}
+    \end{center}
+    \caption{Measured and analytic domain leakage as a function of the
+      iteration matrix spectral radius. To test the behavior with
+      respect to domain size, $n_i=50$, $n_i=100$,and $n_i=200$ were
+      used. The red data was computed numerically by a
+      domain-decomposed adjoint Neumann-Ulam implementation, the black
+      dashed data was generated by
+      Eq~(\ref{eq:mean_chord_domain_leakage}) using the mean-chord
+      approximation, and the dashed-dotted black data was generated by
+      Eq~(\ref{eq:wigner_domain_leakage}) using the Wigner rational
+      approximation.}
+    \label{fig:measured_leakage}
+  \end{spacing}
+\end{figure}
+Again, we note good qualitative agreement between the measured and
+analytic quantities but we begin to see the limits of the leakage
+approximations. To compare the quality of the two approximations, the
+absolute error between the computed leakage fraction and that
+generated by the Wigner rational and mean chord approximations is
+plotted in Figure~\ref{fig:leakage_error} for all domain sizes
+tested. 
+\begin{figure}[ht!]
+  \begin{spacing}{1.0}
+    \begin{center}
+      \includegraphics[width=4.5in,clip]{chapters/parallel_mc/leakage_error.png}
+    \end{center}
+    \caption{Measured and analytic domain leakage absolute error as a
+      function of the iteration matrix spectral radius.  To test the
+      behavior with respect to domain size, $n_i=50$ (green),
+      $n_i=100$ (blue), and $n_i=200$ (red) were used. The dashed
+      lines represent the error using the Wigner rational
+      approximation while the solid lines represent the error using
+      the mean-chord approximation.}
+    \label{fig:leakage_error}
+  \end{spacing}
+\end{figure}
+From these error results, the mean chord approximation is shown to
+have a lower error for ill-conditioned systems as compared to the
+Wigner approximation while the Wigner approximation produces less
+error for more well-conditioned systems. We also note that for the
+optically thick domains, the error is likely corresponded to that
+observed in Figure~\ref{fig:measured_length} for the $k$ parameter
+while the large relative error in $k$ for optically thin domains does
+not affect the approximation significantly. In general, the mean chord
+approximation is a better choice to estimate the leakage fraction in a
+domain from the adjoint Neumann-Ulam method and except for a single
+data point with $n_i=50$, the mean chord approximation yielded leakage
+fractions within 0.05 of the measured results. As the domain becomes
+more optically thick (with both increasing $n_i$ and decreasing
+$\rho(\ve{H})$), the approximations are more accurate.
+
+We have presented an analytic analysis of the domain decomposed
+behavior of the adjoint Neumann Ulam method for linear systems. Good
+agreement was observed for the derived analytic relationships for
+random walk length and leakage fraction when compared to numerical
+results generated by a domain decomposed implementation of the adjoint
+method.
+
+In future work, these relationships will serve as guidelines for
+selecting an appropriate parallel algorithm strategy and provide a
+basis for performance models for future parallel implementations of
+the adjoint Neumann-Ulam method. To be applicable to a broader range
+of problems, these models could potentially be extended to
+non-symmetric systems. In addition, these relations will be used to
+analyze parallel implementations of Monte Carlo synthetic acceleration
+methods that leverage the adjoint method and serve as an initial
+grounds for assessing their feasibility for large-scale problems.
+
+%%---------------------------------------------------------------------------%%
+\section{Load Balancing Concerns}
+\label{sec:mc_load_balancing}
 Although domain decomposition was shown to be efficient in a perfectly
@@
     \scalebox{1.5}{
-      \input{chapters/mc_background/procassini_example.pdftex_t} }
+      \input{chapters/parallel_mc/procassini_example.pdftex_t} }
   \end{center}
@@
 
-\subsection{Reproducible Domain Decomposed Results}
-\label{subsec:reproducible_mc}
+%%---------------------------------------------------------------------------%%
+\section{Reproducible Domain Decomposed Results}
+\label{sec:reproducible_mc}
 The 2006 work of Brunner is notable in that the Monte Carlo codes used
@@
     \scalebox{1.5}{
-      \input{chapters/mc_background/gentile_example.pdftex_t} }
+      \input{chapters/parallel_mc/gentile_example.pdftex_t} }
   \end{center}
@@
 
+%%---------------------------------------------------------------------------%%
 \section{Parallel Adjoint Method}
@@
 
+%%---------------------------------------------------------------------------%%
 \section{Parallel MCSA}

### 2013-03-24 doc/dissertation/document/chapters/conclusion/conclusion.tex
updating disseration chapter structure

\chapter{Conclusions and Analysis}
\label{ch:conclusion}
Put conclusions here.

### 2013-03-24 doc/dissertation/document/chapters/mc_theory/mc_theory.tex
updating disseration chapter structure

--- 2013-03-21 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
@@
-\chapter{Parallel Monte Carlo Solution Methods for Linear Systems}
-\label{ch:parallel_methods}
-
-For Monte Carlo methods, and in particular MCSA, to be viable at the
-production scale, scalable parallel implementations are required. In a
-general linear algebra context for discrete systems, this has not yet
-been achieved. Therefore, we will develop and implement parallel
-algorithms for these methods leveraging both the knowledge gained from
-the general parallel implementations of Krylov methods and modern
-parallel strategies for Monte Carlo as developed by the reactor
-physics community. In order to formulate a parallel MCSA algorithm, we
-recognize that the algorithm occurs in two stages, an outer iteration
-performing Richardson's iteration and applying the correction, and an
-inner Monte Carlo solver that is generating the correction via the
-adjoint method. The parallel aspects of both these components must be
-considered.
+\chapter{An Analytic Framework for Parallel Monte Carlo Methods}
+\label{ch:analytic_framework}
 
 %%---------------------------------------------------------------------------%%
-\section{Domain Decomposition for Monte Carlo}
-\label{sec:msod}
-As observed in the discussion on parallel Krylov methods, large-scale
-problems will surely have their data partitioned such that each
-parallel process owns a subset of the equations in the linear
-system. Given this convention, the adjoint Monte Carlo algorithm must
-perform random walks over a domain that is decomposed and must remain
-decomposed due to memory limitations. This naturally leads us to seek
-parallel algorithms that handle domain decomposition.
-
-In the context of radiation transport, Brunner and colleagues provided
-a survey of algorithms for achieving this as implemented in production
-implicit Monte Carlo codes \citep{brunner_comparison_2006}. In their
-work they identify two data sets that are required to be communicated:
-the sharing of particles that are transported from one domain to
-another and therefore from one processor to another and a global
-communication that signals if particle transport has been completed on
-all processors. The algorithms presented are a fully-locking
-synchronous scheme, an asynchronous-send/synchronous-receive pattern,
-a traditional master/slave scheme, and a modified master/slave scheme
-the implements a binary tree pattern for the global reduction type
-operations needed to communicate between the master and slave
-processes. They observed that the modified master/slave scheme
-performed best in that global communications were implemented more
-efficiently than those required by the asynchronous
-scheme. Furthermore, none of these schemes handled load-imbalanced
-cases efficiently. Such cases will be common if the source sampled in
-the Monte Carlo random walk is not isotropic and not evenly
-distributed throughout the global domain. It was noted that
-efficiencies were improved by increasing the frequency by which
-particle data was communicated between domain-adjacent
-processors. However, this ultimately increases communication costs. In
-2009, Brunner extended his work by using a more load-balanced approach
-with a fully asynchronous communication pattern
-\citep{brunner_efficient_2009}. Although the extended implementation
-was more robust and allowed for scaling to larger numbers of
-processors, performance issues were still noted with parallel
-efficiency improvements needed in both the weak and strong scaling
-cases for unbalanced problems. These results led Brunner to conclude
-that a combination of domain decomposition and domain replication
-could be used to solve some of these issues.
-
-\subsection{Multiple-Set Overlapping-Domain Decomposition}
-\label{subsec:msod}
-In 2010, Wagner and colleagues developed the \textit{multiple-set
-  overlapping-domain} (MSOD) decomposition for parallel Monte Carlo
-applications for full-core light water reactor analysis
-\citep{wagner_hybrid_2010}. In their work, an extension of Brunner's,
-their scheme employed the similar parallel algorithms for particle
-transport but a certain amount of overlap between adjacent domains was
-used to decrease the number of particles leaving the local domain. In
-addition, Wagner utilized a level of replication of the domain such
-that the domain was only decomposed on $O(100)$ processors and if
-replicated $O(1,000)$ times achieves simulation on $O(100,000)$
-processors, thus providing spatial and particle parallelism. Each
-collection of processors that constitutes a representation of the
-entire domain is referred to as a set, and within a set overlap occurs
-among its sub-domains. The original motivation was to decompose the
-domain in a way that it remained in a physical cabinet in a large
-distributed machine, thus reducing latency costs during
-communication. A multiple set scheme is also motivated by the fact
-that communication during particle transport only occurs within a set,
-limiting communications during the transport procedure to a group of
-$O(100)$ processors, a number that was shown to have excellent
-parallel efficiencies in Brunner's work and therefore will scale well
-in this algorithm. The overlapping domains within each set also
-demonstrated reduced communication costs. On each processor, the
-source is sampled in the local domain that would exist if no overlap
-was used while tallies can be made over the entire overlapping domain.
-
-To demonstrate this, consider the example adapted from Mervin's work
-with Wagner and others in the same area \citep{mervin_variance_2012}
-and presented in Figure~\ref{fig:msod_example}.
-\begin{figure}[htpb!]
-  \begin{center}
-    \scalebox{1.5}{
-      \input{chapters/parallel_mc/msod_example.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Overlapping domain example illustrating how domain
-      overlap can reduce communication costs.}
-    \textit{All particles start in the blue region of interest. The
-      dashed line represents 0.5 domain overlap between domains.}}
-  \label{fig:msod_example}
-\end{figure}
-In this example, 3 particle histories are presented emanating from the
-blue region of interest. Starting with particle A, if no domain
-overlap is used then the only the blue domain exists on the starting
-processor. Particle A is then transported through 3 other domains
-before the history ends, therefore requiring three communications to
-occur in Brunner's algorithm. If a 0.5 domain overlap is permitted as
-shown by the dashed line, then the starting process owns enough of the
-domain such that no communications must occur in order to complete the
-particle A transport process. Using 0.5 domain overlap also easily
-eliminates cases such as the represented by the path of particle C. In
-this case, particle C is scattering between two adjacent domains,
-incurring a large latency cost for a single particle. Finally, with
-particle B we observe that 0.5 domain overlap will still not eliminate
-all communications. However, if 1 domain overlap were used, the entire
-geometry shown in Figure~\ref{fig:msod_example} would be contained on
-the source processor and therefore transport of all 3 particles
-without communication would occur.
-
-Wagner and colleagues used this methodology for a 2-dimensional
-calculation of a pressurized water reactor core and varied the domain
-overlap from 0 to 3 domain overlap (a $7 \times 7$ box in the context
-of our example) where a domain constituted a fuel assembly. For the
-fully domain decomposed case, they observed that 76.12\% of all source
-particles leave the domain. At 1.5 domain overlap, the percentage of
-source particles born in the center assembly leaving the processor
-domain dropped to 1.05\% and even further for 0.02\% for the 3 domain
-overlap. Based on these results, this overlap approach, coupled with
-the multiple sets paradigm that will scale for existing parallel
-transport algorithms, provides a scalable Monte Carlo algorithm for
-today's modern machines.
-
-%%---------------------------------------------------------------------------%%
-\section{Domain-to-Domain Communication}
-\label{sec:mc_comm_costs}
+\section{Domain-to-Domain Communication for Symmetric Systems}
+\label{sec:mc_comm_costs_symmetric}
 With Wagner's work, we observed that domain overlap implemented with a
@@
   \begin{center}
-    \scalebox{1.25}{\input{chapters/parallel_mc/stencil.pdftex_t}}
+    \scalebox{1.25}{\input{chapters/mc_theory/stencil.pdftex_t}}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=5in,clip]{chapters/parallel_mc/diffusion_spectrum.png}
+    \includegraphics[width=5in,clip]{chapters/mc_theory/diffusion_spectrum.png}
   \end{center}
@@
     \begin{center}
-      \includegraphics[width=4in,clip]{chapters/parallel_mc/measured_spec_rad.png}
+      \includegraphics[width=4in,clip]{chapters/mc_theory/measured_spec_rad.png}
     \end{center}
-    \caption{Measured and analytic preconditioned diffusion operator
-      spectral radius as a function of the absorption cross section to
-      scattering cross section ratio. Values of $h=0.01$, $h=0.1$, and
-      $h=1.0$ were used. The red data was computed numerically by an
-      eigensolver while the black dashed data was generated by
-      Eq~(\ref{eq:iteration_radius}).}
+    \caption{\textbf{Measured and analytic preconditioned diffusion
+        operator spectral radius as a function of the absorption cross
+        section to scattering cross section ratio.} \textit{Values of
+        $h=0.01$, $h=0.1$, and $h=1.0$ were used. The red data was
+        computed numerically by an eigensolver while the black dashed
+        data was generated by Eq~(\ref{eq:iteration_radius}).}}
     \label{fig:measured_spec_rad}
@@
     \begin{center}
-      \includegraphics[width=6.5in,clip]{chapters/parallel_mc/measured_length.png}
+      \includegraphics[width=6.5in,clip]{chapters/mc_theory/measured_length.png}
     \end{center}
-    \caption{Measured and analytic random walk length as a function of
-      the iteration matrix spectral radius. The weight cutoff was
-      varied with \sn{1}{-2}, \sn{1}{-4}, and \sn{1}{-8}. In the left
-      plot, the red data was computed numerically by an adjoint
-      Neumann-Ulam implementation while the black dashed data was
-      generated by Eq~(\ref{eq:analytic_k}). In the right plot, the
-      relative error between the predicted and measured results is
-      presented for each weight cutoff.}
+    \caption{\textbf{Measured and analytic random walk length as a
+        function of the iteration matrix spectral radius.} \textit{The
+        weight cutoff was varied with \sn{1}{-2}, \sn{1}{-4}, and
+        \sn{1}{-8}. In the left plot, the red data was computed
+        numerically by an adjoint Neumann-Ulam implementation while
+        the black dashed data was generated by
+        Eq~(\ref{eq:analytic_k}). In the right plot, the relative
+        error between the predicted and measured results is presented
+        for each weight cutoff.}}
     \label{fig:measured_length}
@@
     \begin{center}
-      \includegraphics[width=4.5in,clip]{chapters/parallel_mc/leakage_variation.png}
+      \includegraphics[width=4.5in,clip]{chapters/mc_theory/leakage_variation.png}
     \end{center}
-    \caption{Measured and analytic domain leakage as a function of the
-      iteration matrix spectral radius. To test the behavior with
-      respect to domain size, $n_i=50$, $n_i=100$,and $n_i=200$ were
-      used. The red data was computed numerically by a
-      domain-decomposed adjoint Neumann-Ulam implementation, the black
-      dashed data was generated by
+    \caption{\textbf{Measured and analytic domain leakage as a
+        function of the iteration matrix spectral radius.} \textit{To
+        test the behavior with respect to domain size, $n_i=50$,
+        $n_i=100$,and $n_i=200$ were used. The red data was computed
+        numerically by a domain-decomposed adjoint Neumann-Ulam
+        implementation, the black dashed data was generated by
         Eq~(\ref{eq:mean_chord_domain_leakage}) using the mean-chord
-      approximation, and the dashed-dotted black data was generated by
-      Eq~(\ref{eq:wigner_domain_leakage}) using the Wigner rational
-      approximation.}
+        approximation, and the dashed-dotted black data was generated
+        by Eq~(\ref{eq:wigner_domain_leakage}) using the Wigner
+        rational approximation.}}
     \label{fig:measured_leakage}
@@
     \begin{center}
-      \includegraphics[width=4.5in,clip]{chapters/parallel_mc/leakage_error.png}
+      \includegraphics[width=4.5in,clip]{chapters/mc_theory/leakage_error.png}
     \end{center}
-    \caption{Measured and analytic domain leakage absolute error as a
-      function of the iteration matrix spectral radius.  To test the
-      behavior with respect to domain size, $n_i=50$ (green),
-      $n_i=100$ (blue), and $n_i=200$ (red) were used. The dashed
-      lines represent the error using the Wigner rational
-      approximation while the solid lines represent the error using
-      the mean-chord approximation.}
+    \caption{\textbf{Measured and analytic domain leakage absolute
+        error as a function of the iteration matrix spectral radius.}
+      \textit{To test the behavior with respect to domain size,
+        $n_i=50$ (green), $n_i=100$ (blue), and $n_i=200$ (red) were
+        used. The dashed lines represent the error using the Wigner
+        rational approximation while the solid lines represent the
+        error using the mean-chord approximation.}}
     \label{fig:leakage_error}
@@
 
-We have presented an analytic analysis of the domain decomposed
-behavior of the adjoint Neumann Ulam method for linear systems. Good
-agreement was observed for the derived analytic relationships for
-random walk length and leakage fraction when compared to numerical
-results generated by a domain decomposed implementation of the adjoint
-method.
-
-In future work, these relationships will serve as guidelines for
-selecting an appropriate parallel algorithm strategy and provide a
-basis for performance models for future parallel implementations of
-the adjoint Neumann-Ulam method. To be applicable to a broader range
-of problems, these models could potentially be extended to
-non-symmetric systems. In addition, these relations will be used to
-analyze parallel implementations of Monte Carlo synthetic acceleration
-methods that leverage the adjoint method and serve as an initial
-grounds for assessing their feasibility for large-scale problems.
-
 %%---------------------------------------------------------------------------%%
-\section{Load Balancing Concerns}
-\label{sec:mc_load_balancing}
-Although domain decomposition was shown to be efficient in a perfectly
-load balanced situation in Siegel's work \citep{siegel_analysis_2012},
-careful consideration must be made for situations where this is not
-the case. Given the stochastic nature of the problem and lack of a
-globally homogeneous domain, parallel Monte Carlo simulations are
-inherently load imbalanced. Procassini and others worked to alleviate
-some of the load imbalances that are generated by both particle and
-spatial parallelism and are therefore applicable to the MSOD algorithm
-\citep{procassini_dynamic_2005}. They chose a dynamic balancing scheme
-in which the number of times a particular domain was replicated was
-dependent on the amount of work in that domain (i.e. domains with a
-high particle flux and therefore more particle histories to compute
-require more work). In this variation, domains that require more work
-will be replicated more frequently at reduced particle counts in each
-replication. Furthermore, Procassini and colleagues noted that as the
-simulation progressed and particles were transported throughout the
-domain, the amount of replication for each domain would vary as
-particle histories began to diffuse, causing some regions to have
-higher work loads and some to have smaller work loads than the initial
-conditions.
-
-Consider the example in Figure~\ref{fig:procassini_example} adapted
-from Procassini's work.
-\begin{figure}[htpb!]
-  \begin{center}
-    \scalebox{1.5}{
-      \input{chapters/parallel_mc/procassini_example.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Example illustrating how domain decomposition can
-      create load balance issues in Monte Carlo.}  \textit{A domain is
-      decomposed into 4 zones on 8 processors with a point source in
-      the lower left zone. As the particles diffuse from the source in
-      the random walk sequence as shown in the top row, their tracks
-      populate the entire domain. As given in the bottom row, as the
-      global percentage of particles increases in a zone, that zone's
-      replication count is increased.}}
-  \label{fig:procassini_example}
-\end{figure}
-In this example, a geometry is decomposed into 4 domains on 8
-processors with a point source in the bottom left domain. To begin,
-because the point source is concentrated in one domain, that domain is
-replicated 5 times such the amount of work it has to do per processor
-is roughly balanced with the others. As the particles begin to diffuse
-away from the point source, the amount of replication is adjusted to
-maintain load balance. Near the end of the simulation, the diffusion
-of particles is enough that all domains have equal replication.  By
-doing this, load balance is improved as each domain has approximately
-equal work although each domain may represent a different spatial
-location and have a differing number of histories to
-compute. Compared to Wagner's work where the fission source was
-distributed relatively evenly throughout the domain, fixed source
-problems (and especially those that have a point-like source) like
-those presented in Procassini's work will be more prone to changing
-load balance requirements.
-
-%%---------------------------------------------------------------------------%%
-\section{Reproducible Domain Decomposed Results}
-\label{sec:reproducible_mc}
-The 2006 work of Brunner is notable in that the Monte Carlo codes used
-to implement and test the algorithms adhered to a strict policy of
-generating identical results independent of domain decomposition or
-domain replication as derived from the work of Gentile and colleagues
-\citep{gentile_obtaining_2005}. In Gentile's work, a procedure is
-given for obtaining results reproducible to machine precision for an
-arbitrary number of processors and domains. Differences can arise from
-using a different random number sequence in each domain and performing
-a sequence of floating point operations on identical data in a
-different order, leading to variations in round-off error and
-ultimately a non-identical answer. They use a simple example,
-recreated below in Figure~\ref{fig:gentile_example}, that illustrates
-these issues.
-\begin{figure}[htpb!]
-  \begin{center}
-    \scalebox{1.5}{
-      \input{chapters/parallel_mc/gentile_example.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Gentile's example illustrating how domain
-      decomposition can create reproducibility issues in Monte Carlo.}
-    \textit{Both particles A and B start in zone 1 on processor
-      1. Particle A moves to zone 2 on processor 2 and scatters back
-      to zone 1 while B scatters in zone 1 and remains there. A1 and
-      A2 denote the track of particle A that is in zone 1 while B1 and
-      B2 denote the track of particle B that is in zone 1.}}
-  \label{fig:gentile_example}
-\end{figure}
-In this example, the domain is decomposed on two processors with each
-processor owning one of the two zones. Starting with particle A, it is
-born in zone 1 and is transported to zone 2 where a scattering event
-occurs. Concerning the first reproducibility issue, if the same
-sequence of random numbers is not used to compute the trajectory from
-the new scattering event, we cannot expect to achieve the same result
-if the domain were not decomposed. If the numbers are different, the
-scattering event in zone 2 may keep the particle there or even eject
-it from the domain if the sequence were different. The second issue is
-demonstrated by adding another particle B that remains in the
-domain. In this case, an efficient algorithm will transport particle A
-on processor 1 until it leaves zone 1 and then transport particle
-B. Particle A will not renter the domain until it has been
-communicated to processor 2, processor 2 performs the transport, and
-it is communicated back to processor 1. If we are doing a track-length
-tally in zone 1, then we sum the tracks lengths observed in that
-zone. In the single processor, single zone case particle A would be
-transported in its entirety and then particle B transported. This
-would result in a tally sum with the following order of operations:
-$(((A1+A2)+B1)+B2)$. If we were instead to use 2 processors, we would
-instead have the following order: $(((A1+B1)+B2)+A2)$. In the context
-of floating point operations, we cannot expect these to have an
-identical result to machine precision as round-off approximations will
-differ resulting in non-commutative addition.
-
-Procassini's solutions to these problems are elegant in that they
-require a minimal amount of modification to be applied to the Monte
-Carlo algorithm. To solve the first issue, in order to ensure each
-particle maintains an invariant random number sequence that determines
-its behavior regardless of domain decomposition, each particle is
-assigned a random number seed that describes its current state upon
-entering the domain of a new processor. These seeds are derived from
-the actual geometric location of the particle such that it is
-decomposition invariant. Non-commutative floating point operations are
-overcome by instead mapping floating point values to 64-bit integer
-values for which additions will always be commutative. Once the
-operations are complete, these integers are mapped back to floating
-point values.
-
-%%---------------------------------------------------------------------------%%
-\section{Parallel Adjoint Method}
-\label{sec:parallel_adjoint}
-We can take much of what was learned from the survey of parallel Monte
-Carlo methods for radiation transport and directly apply it to a
-parallel formulation of our stochastic linear solvers. Direct analogs
-can be derived from these works by noting that the primary difference
-between solving a linear system with Monte Carlo methods and fixed
-source Monte Carlo transport problems is the content of the Markov
-chains that are generated. The transitions represented by these chains
-are bound by probabilities and weights and are initiated by the
-sampling of a source. In the context of transport problems, those
-transitions represent events such as particle scattering and
-absorption with probabilities that are determined by physical data in
-the form of cross sections. For stochastic matrix inversion, those
-transitions represent moving between the equations of the linear
-system (and therefore the physical domain which they represent) and
-their probabilities are defined by the coefficients of those
-equations. Ultimately, we tally the contributions to generate
-expectation values in the desired states as we progress through the
-chains. Therefore, parallel methods for Monte Carlo radiation
-transport can be abstracted and we can use those concepts that apply
-to matrix inversion methods as an initial means of developing a
-parallel Neumann-Ulam-type solver. Based on the results observed in
-the last decade of parallel Monte Carlo development, we can generate
-many practical questions that this type of work could answer.
-
-Given a decomposed domain, per Wagner's work
-\citep{wagner_hybrid_2010} is clear that domain overlap significantly
-reduces the amount of communication required between adjacent domains
-as histories move to states that are not owned by the local
-processor. How much domain overlap is suitable for matrix inversion
-problems? Are we memory limited by large problems such that only so
-much overlap is feasible? Are there metrics related to the properties
-of the matrix including the eigenvalues and sparsity pattern such that
-we can provide guidelines for selecting the amount of overlap
-required? Furthermore, as Wagner and colleagues observed
-\citep{wagner_hybrid_2010}, with marginal domain overlap for reactor
-problems, the percentage of histories leaving the local domain can
-easily be reduced to less than 1\%. Evans and colleagues in their
-initial MCSA development typically used only 50 particle histories in
-order to compute a correction in an MCSA iteration
-\citep{evans_monte_2012}. Per the Central Limit Theorem, that
-correlates to a statistical uncertainty of 14.1\% in the correction,
-yet as compared to conventional Krylov solvers, MCSA achieved
-identical numerical results. If good convergence and numerically
-accurate solutions can be achieved with such a large uncertainty in
-the correction computation, then perhaps with enough domain overlap
-the minimal amount of histories that do transition to non-local states
-can be ignored and thus eliminate all communication in the parallel
-Monte Carlo adjoint solver transport sequence and create an
-embarrassingly parallel method. Given Gropp's work on parallel Krylov
-methods that we discussed in
-\S~\ref{subsec:projection_method_performance} and the empirical
-results of Siegel \citep{siegel_analysis_2012}, we know that these
-nearest neighbor computations between adjacent domains do not scale as
-well as global reduction operations and therefore we are improving
-scaling by eliminating them from the transport sequence. It will be
-important to determine if the expectation value bias in the MCSA
-correction generated by this approximation will remain within the
-bounds of the already high statistical uncertainty for unbiased
-estimates, thus providing numerically equivalent results.
-
-From a domain replication perspective, this may be difficult to
-achieve with a production scale linear solver. Typically, memory is at
-a premium and therefore the more distinct domains available in the
-decomposition, the spatially finer and/or the numerically more
-accurate the discretization that can be implemented. How much
-replication is possible for large problems? How does replication
-facilitate parallel performance when coupled with domain overlap? How
-can we measure how much domain overlap is feasible? Replicating
-domains may therefore run into memory limitations for exceptionally
-large problems such that the operator, solution vector, and source
-vector must be copied in their entirety multiple times. From a
-resiliency standpoint, such an operation will be required, and
-therefore its performance and memory implications on conventional
-problems must be analyzed.
-
-%%---------------------------------------------------------------------------%%
-\section{Parallel MCSA}
-\label{sec:parallel_mcsa}
-With a parallel adjoint Neumann-Ulam solver implementation, the
-parallel implementation of the MCSA method will be trivial. Recall the
-MCSA iteration procedure outlined in Eq~(\ref{eq:mcsa}). In
-\S~\ref{sec:parallel_krylov_methods} we discussed parallel matrix and
-vector operations as utilized in conventional Krylov methods. We
-utilize these here for the parallel MCSA implementation. In the first
-step, a parallel matrix-vector multiply is used to apply the split
-operator to the previous iterate's solution. A parallel vector update
-is then performed with the source vector to arrive at the initial
-iteration guess. In the next step, the residual is computed by the
-same operations where now the operator is applied to the solution
-guess with a parallel matrix-vector multiply and then a parallel
-vector update with the source vector is performed. Once the correction
-is computed with a parallel adjoint Neumann-Ulam solve, this
-correction is applied to the guess with a parallel vector update to
-get the new iteration solution. Additionally, as given by
-Eq~(\ref{eq:mcsa_stopping_criteria}), 2 parallel vector reductions
-will be required to check the stopping criteria: one initially to
-compute the infinity norm of the source vector, and another at every
-iteration to compute the infinity norm of the residual vector. For
-this implementation, all of the issues that will be potentially
-generated by the parallel adjoint solver implementation will manifest
-themselves here as the quality of the correction will be of intense
-study.
-
-In addition to parallel implementation and performance, MCSA's
-potential for aiding advancement in non-symmetric matrix solutions
-leads to a natural comparison with the GMRES algorithm. As both
-solvers are aimed at the same class of problem, we desire a set of
-metrics that will allow us to quantitatively compare the two. Given
-that the Krylov subspace maintained by GMRES can become large, do we
-benefit from a memory standpoint with an MCSA scheme in that no
-subspace is required? Does this benefit outweigh the fact that the
-linear operator must be explicitly formed in order to build the
-transition probabilities for the random walk sequence? For
-non-symmetric systems, does MCSA exhibit similar convergence
-properties to Krylov methods? If a Krylov methods build a subspace,
-can those memory savings in MCSA be used to implement domain
-replication in the adjoint solver? Such questions can be answered by a
-comparative study of the two solvers that controls the system size and
-the iterations required to converge.
+\section{Domain-to-Domain Communication for Asymmetric Systems}
+\label{sec:mc_comm_costs_asymmetric}
+Extend the above results to asymmetric systems.

### 2013-03-24 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
updating disseration chapter structure

--- 2013-03-21 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
@@
 %%---------------------------------------------------------------------------%%
-\section{Domain-to-Domain Communication}
-\label{sec:mc_comm_costs}
-With Wagner's work, we observed that domain overlap implemented with a
-domain decomposed Monte Carlo setting reduces the amount of nearest
-neighbor communication that must occur during transport. This
-communication, expected to be expensive, was analyzed by Siegel and
-colleagues for idealized, load balanced situations without domain
-overlap \citep{siegel_analysis_2012}. They arrive at a few empirical
-relationships and generate a numerical experiment that provides
-insight into the feasibility of large scale domain decomposed Monte
-Carlo as well as the relationship of communication to system
-parameters. In their work, they define a simple 3-dimensional system
-subdivded into Cartesian block domains with each processor containing
-a single domain. Given this geometry, a particle can either be
-absorbed in its current domain or be transported to one of the 6
-adjacent domains owned by another process.
-
-To quantify the number of particles that leak out of the local domain
-they define a leakage fraction, $\lambda$, as:
-\begin{equation}
-  \lambda =
-  \frac{average\ \#\ of\ particles\ leaving\ local\ domain}{total\ 
-    of\ \#\ of\ particles\ starting\ in\ local\ domain}\:.
-  \label{eq:leakage_fraction}
-\end{equation}
-For their studies, they assumed that the value of $\lambda$ was
-dependent on the total cross section of the system and the same for
-all processes for simplicity and load balancing. Using this
-relationship and their simple geometry, they were able to relate the
-amount of compute time required to transport all source particles to
-the number of source particles, the leakage fraction, the latency and
-bandwidth of communication, the ratio of local to global geometry
-size, and the total cross section of the system.
-
-What they found was that there was indeed an optimum ratio of local
-domain to global domain size that results in a minimum total compute
-time. They also found that this ratio was only dependent on the
-latency and bandwidth constraints of the parallel machine. We expect
-this because nearest neighbor communication patterns scale with the
-volume and surface sizes of the local domains. The larger the volume,
-the more compute time is spent within the local volume and therefore
-at some point you are limited by the latency of the system, or the
-time it takes to receive the data to operate on. However, at a certain
-point, when the surface of those volumes becomes large enough, the
-amount of data that must be communicated grows large enough that it
-may overcome the bandwidth limitations of the system, resulting in a
-decrease in performance. In their numerical experiments, they used the
-above geometry and leakage rates based on that computed for a reactor
-system for a single material problem with a total cross section
-equivalent to that of homogenized light water reactor. Given the
-optimal local domain size, they observed for load balanced situations
-that nearest neighbor communication costs during transport are a small
-fraction of total compute time, leading to the conclusion that domain
-decomposition is feasible for large-scale Monte Carlo calculations.
-
-\subsection{Load Balancing Concerns}
-\label{subsec:mc_load_balancing}
+\section{Load Balancing Concerns}
+\label{sec:mc_load_balancing}
 Although domain decomposition was shown to be efficient in a perfectly

### 2013-03-24 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
updating disseration chapter structure

--- 2013-03-21 doc/dissertation/document/chapters/spn_solutions/spn_solutions.tex
@@
-\chapter{Monte Carlo Solution Methods for the Simplified $P_N$ Equations}
-\label{ch:mc_spn_solutions}
+\chapter{The Simplified $P_N$ Equations}
+\label{ch:spn_equations}
 
@@
 systems.
-
-%%---------------------------------------------------------------------------%%
-\section{Spectral Analysis of the $SP_N$ Equations}
-\label{sec:spectral_analysis}
-Now that we have an understanding of the linear system generated by
-the $SP_N$ approximation to the neutron transport equation and the
-various parameters in the system that may be adjusted, we can consider
-various means of solution. As the solution is asymmetric, modern
-iterative solvers should be considered for this task. The performance
-of Krylov subspace methods are bound to various properties of the
-eigenvalue spectrum of the linear operator while methods based on
-stationary iterations have performance limits imposed by the spectral
-radius of the sytem. In this section we will review the important
-spectral properties to study for common iterative methods for
-asymmetric systems. Based on these properties, we will then perform a
-parameter-based spectral analysis for the linear operator generated by
-the multigroup $SP_N$ equations. Preconditioning strategies for these
-methods are not considered. Krylov subspace method restarts and
-truncation will also not be considered.
-
-\subsection{Stationary Methods for Linear Systems}
-\label{subsec:stationary_solvers}
-Stationary methods for linear systems arise from splitting the
-operator in Eq~(\ref{eq:multigroup_matrix_system})
-\cite{leveque_finite_2007}:
-\begin{equation}
-  \mathbb{L} = \mathbf{M} - \mathbf{N}\:,
-  \label{eq:split_linear_operator}
-\end{equation}
-where the choice of $\mathbf{M}$ and $\mathbf{N}$ will be dictated by
-the particular method chosen. Using this split definition of the
-operator we can then write:
-\begin{equation}
-  \mathbf{M}\mathbb{U} - \mathbf{N}\mathbb{U} = \mathbb{Q}\:.
-  \label{eq:linear_split_equation1}
-\end{equation}
-By rearranging, we can generate a form more useful for analysis:
-\begin{equation}
-  \mathbb{U} = \mathbf{H}\mathbb{U} + \mathbf{c}\:,
-  \label{eq:linear_split_equation2}
-\end{equation}
-where $\mathbf{H}=\mathbf{M}^{-1}\mathbf{N}$ is defined as the
-\textit{iteration matrix} and
-$\mathbf{c}=\mathbf{M}^{-1}\mathbb{Q}$. With the solution vector on
-both the left and right hand sides, an iterative method can then be
-formed:
-\begin{equation}
-    \mathbb{U}^{k+1} = \mathbf{H}\mathbb{U}^k + \mathbf{c}\:,
-  \label{eq:linear_iterative_method}
-\end{equation}
-with $k \in \mathbf{Z}^+$ defined as the \textit{iteration
-  index}. Defining $\mathbf{e}^k = \mathbf{u}^k - \mathbf{u}$ as the
-solution error at the $k^{th}$ iterate, we can subtract
-Eq~(\ref{eq:linear_split_equation2}) from
-Eq~(\ref{eq:linear_iterative_method}):
-\begin{equation}
-  \mathbf{e}^{k+1} = \mathbf{H}\mathbf{e}^k\:.
-  \label{eq:linear_iterative_error}
-\end{equation}
-Our error after $k$ iterations is then:
-\begin{equation}
-  \mathbf{e}^{k} = \mathbf{H}^k\mathbf{e}^0\:.
-  \label{eq:linear_k_iter_error}
-\end{equation}
-In other words, successive application of the iteration matrix is the
-mechanism driving down the error in a stationary method. By assuming
-$\mathbf{H}$ is diagonalizable \cite{saad_iterative_2003}, we then
-have:
-\begin{equation}
-  \mathbf{e}^{k} =
-  \mathbf{R}\boldsymbol{\Lambda}^k\mathbf{R}^{-1}\mathbf{e}^0\:,
-  \label{eq:linear_k_iter_error_diag}
-\end{equation}
-where $\boldsymbol{\Lambda}$ contains the Eigenvalues of $\mathbf{H}$
-on its diagonal and the columns of $\mathbf{R}$ contain the
-Eigenvectors of $\mathbf{H}$. Computing the 2-norm of the above form
-then gives:
-\begin{equation}
-  ||\mathbf{e}^{k}||_2 \leq ||\boldsymbol{\Lambda}^k||_2\ 
-  ||\mathbf{R}||_2\ ||\mathbf{R}^{-1}||_2\ ||\mathbf{e}^0||_2\:,
-  \label{eq:linear_k_iter_norm1}
-\end{equation}
-which gives:
-\begin{equation}
-  ||\mathbf{e}^{k}||_2 \leq \rho(\mathbf{H})^k \kappa(\mathbf{R})
-  ||\mathbf{e}^0||_2\:.
-  \label{eq:linear_k_iter_norm2}
-\end{equation}
-For iteration matrices where the Eigenvectors are orthogonal,
-$\kappa(\mathbf{R})=1$ and the error bound reduces to:
-\begin{equation}
-  ||\mathbf{e}^{k}||_2 \leq \rho(\mathbf{H})^k
-  ||\mathbf{e}^0||_2\:.
-  \label{eq:linear_k_iter_norm3}
-\end{equation}
-We can now restrict $\mathbf{H}$ by asserting that $\rho(\mathbf{H}) <
-1$ for a stationary method to converge such that $k$ applications of
-the iteration matrix will not cause the error to grow in
-Eq~(\ref{eq:linear_k_iter_norm3}). In addition, the performance of the
-stationary method is dictated by $\rho(\mathbf{H})$ such that
-performance of the method decreases as it approaches to unity.
-
-\subsection{Krylov Subspace Methods for Linear Systems}
-\label{subsec:subspace_solvers}
-The performance of different Krylov subspace solvers can depend on
-different properties of the linear system. As a means to study these
-differences, we use a 1992 paper by Nachtigal and colleagues
-\cite{nachtigal_how_1992} that performs and in-depth analysis of three
-Krylov subspace methods that are still used today in modern physics
-applications: conjugate gradient iteration for the normal equations
-(CGN), generalized minimum residual iteration (GMRES), and a
-biconjugate gradient-based method (CGS). For our work, we will
-consider only their analysis of GMRES as it currently leveraged in
-Denovo for $SP_N$ solutions and as it also applies to CGS for most
-matrices.
-
-For GMRES, at each iteration we are extracting a correction for the
-solution from the Krylov subspace:
-\begin{equation}
-  \mathbb{U}^k \in \mathbb{U}^0 + \langle \mathbf{r_0}, \mathbb{L}
-  \mathbf{r_0}, \cdots, \mathbb{L}^{k-1} \mathbf{r_0} \rangle\:,
-  \label{eq:gmres_iteration}
-\end{equation}
-where the extraction is constrained by ensuring orthogonality of the new
-residual with the Krylov space:
-\begin{equation}
-  \mathbf{r}^k \bot \langle \mathbb{L} \mathbf{r_0}, \mathbb{L}^2
-  \mathbf{r_0}, \cdots, \mathbb{L}^{k} \mathbf{r_0} \rangle\:,
-  \label{eq:gmres_constraint}
-\end{equation}
-with the residual at the $k^{th}$ iteration defined as:
-\begin{equation}
-  \mathbf{r}^k = \mathbb{Q} - \mathbb{L}\mathbb{U}^k\:.
-  \label{eq:residual}
-\end{equation}
-From \cite{nachtigal_how_1992}, we then have the following error at
-each iteration:
-\begin{equation}
-  \mathbf{e}^k = p_k(\mathbb{L})\mathbf{e}^0\:,
-  \label{eq:gmres_iteration_error}
-\end{equation}
-where $p_k(\mathbb{L})$ is an arbitrary $k^{th}$ degree polynomial and
-$p_k(0)=1$. This is a very similar form for the iteration error as
-observed for stationary methods in Eq~(\ref{eq:linear_k_iter_norm3})
-where now as these $p_k$ polynomials are constructed as the iterations
-progress, a rapid decrease in $||p_k(\mathbb{L})||$ means a rapid rate
-of convergence. For matrices $\mathbb{L}$ that are approximately
-normal (which was exactly the assumption we made to arrive at
-Eq~(\ref{eq:linear_k_iter_norm3})), then we can write the polynomial
-norm as:
-\begin{equation}
-  ||p_k(\mathbb{L})|| = \sup_{z\in\Lambda(\mathbb{L})}|p_k(z)|\:,
-  \label{eq:gmres_polynomial}
-\end{equation}
-where $\Lambda(\mathbb{L})$ is the eigenvalue spectrum of the normal
-matrix $\mathbb{L}$. So if we could obtain the polynomials, the
-largest eigenvalue of $\mathbb{L}$ could be used to calculate
-convergence rates. Next we consider a choice of polynomials to use for
-the convergence rate approximation. They should of course be
-orthogonal to reflect the orthogonal projection mechanism by which the
-residual is reduced. In Nachtigal's work, Chebyshev polynomials were
-used and defined on the interval of Eigenvalues,
-$[\lambda_{min},\lambda_{max}]$, and then normalized via the
-constraint $p_k(0)=1$. Thus the rate of convergence for these
-polynomials is an approximation to the rate of convergence of the
-method.
-
-We can check the validity of the normal
-approximation by computing the condition number of $\mathbb{L}$:
-\begin{equation}
-  \kappa(\mathbb{L}) = ||\mathbb{L}||\ ||\mathbb{L}^{-1}||
-  \label{eq:operator_condition_number}
-\end{equation}
-The closer $\kappa(\mathbb{L})$ is to one, the more normal the matrix
-and the better the above approximation. The same check can also be
-performed to verify the assumption used to generate
-Eq~(\ref{eq:linear_k_iter_norm3}) where $\kappa(\mathbf{H})$ can be
-computed. For our analysis, we will compute $\kappa_2$ where the
-matrix norms are induced by the 2-norm of a vector such that:
-\begin{equation}
-  ||\mathbb{L}||_2 = \sqrt{\rho(\mathbb{L}^T\mathbb{L})}\:.
-  \label{eq:matrix_2_norm}
-\end{equation}
-
-\subsection{Organization of the Spectral Analysis}
-\label{subsec:spectral_organization}
-For the spectral analysis we will perform a numerical parameter study
-for the multigroup $SP_N$ equations. Based on the above information
-for iterative methods, several key quantities should be computed for
-the $SP_N$ system:
-\begin{itemize}
-\item Eigenvalue spectrum for the linear operator $\mathbb{L}$
-\item Largest eigenvalue for the linear operator $\mathbb{L}$
-\item Smallest eigenvalue for the linear operator $\mathbb{L}$
-\item Condition number for the linear operator $\mathbb{L}$
-\item For each stationary method:
-  \begin{itemize}
-  \item Eigenvalue spectrum for the iteration matrix $\mathbf{H}$
-  \item Largest eigenvalue for the iteration matrix $\mathbf{H}$
-  \item Smallest eigenvalue for the iteration matrix $\mathbf{H}$
-  \item Condition number for the iteration matrix $\mathbf{H}$
-  \end{itemize}
-\item For GMRES:
-  \begin{itemize}
-  \item Chebyshev polynomial norm using eigenvalues of $\mathbb{L}$
-    for a fixed number of iterations determined by stationary method
-    convergence
-  \end{itemize}
-\end{itemize}
-In addition, the following parameters may be varied in the $SP_N$
-system in Denovo:
-\begin{itemize}
-  \item $SP_N$ (angular flux expansion) order
-  \item $P_N$ (scattering cross section expansion) order
-  \item Number of energy groups $N_g$
-  \item Total cross sections $\sigma^g$
-  \item Scattering cross section matrix $\sigma_{sn}^{gg'}$
-  \item Boundary conditions (Marshak or reflecting)
-  \item Finite volume mesh size ($\Delta$ in all directions)
-  \item Number of mesh elements
-  \item Source of neutrons (Uniform or Point)
-\end{itemize}
-We will restrict the problem description in Denovo to one similar to
-Brantley and Larsen's $SP_N$ work where the spectral radius was
-computed for convergence analysis of several proposed solution
-methods. From this definition we choose a single material problem with
-a uniform source of unity in all energy groups. For the boundary
-conditions, we choose reflecting conditions on all six sides of the
-cubic domain for an effectively homogenous infinite medium problem
-which we will mesh with a $5\times 5\times 5$ grid with
-$\Delta=0.1$. We are then left to vary the angular and energy
-components of the system in our parameter studies: the $SP_N$ order,
-the number of energy groups, and their cross sections of order
-$P_N$. In Brantley and Larsen's work, a single energy group was used
-and therefore we may not use their cross sections in our multigroup
-computations. For our work, we will always choose a total cross
-section of unity for all groups and all flux moments. For the
-scattering matrix, two classes will be chosen: one with upscatter and
-one without. For the downscatter only case, all groups may scatter
-within the group and to all groups of lower energy (yielding a lower
-triangular scattering matrix) with a cross section of 0.5. For the
-upscatter case, all groups may scatter within the group and to all
-groups of lower energy given a cross section of 0.5 and for upscatter,
-all groups may scatter up to two groups higher in energy with a cross
-section of 0.25. All cross section matrices will be equivalent for
-each scattering moment. For these cross sections, the $P_N$ order will
-be varied with $N=0,1,3$. Both 1 and 10 energy groups will be used
-with the 10 energy group case using the scattering matrices specified
-above. For the angular flux, the $SP_N$ order will be varied with
-$N=1,3,5,7$. This gives a total of 36 linear system variations for the
-spectral analysis.
-
-\subsection{Spectral Analysis Results}
-\label{subsec:analysis_results}
-
-%%---------------------------------------------------------------------------%%
-\section{Monte Carlo Synthetic Acceleration for the $SP_N$ Equations}
-\label{sec:monte_carlo}

### 2013-03-24 doc/dissertation/document/chapters/spn_solutions/spn_solutions.tex
updating disseration chapter structure

@@
 \label{ch:mc_spn_solutions}
-
-The neutron transport problem is complicated. Solutions cover a large
-phase space and the problems of interest are often geometrically
-complex, very large, or both, requiring tremendous computational
-resources to generate an adequate solution. Modern deterministic
-methods for large scale problems are commonly variants on the discrete
-ordinates ($S_N$) method \cite{evans_denovo:_2010}. For full reactor
-core neutronics simulations, the $S_N$ method requires potentially
-trillions of unknown angular flux moments to be computed to achieve
-good accuracy for the repsonses of interest. Other forms of the
-transport problem, including the $P_N$ method, take on a simpler form
-than the more common $S_N$ methods but lack in accuracy when compared
-while still requiring considerable computational resources for
-solutions in multiple dimensions.
-
-In the 1960's, Gelbard developed an ad-hoc multidimensional extension
-of the simple single dimension planar $P_N$ equations that created a
-system of coupled, diffusion-like equations known as the simplified
-$P_N$ ($SP_N$) equations. Up until around the 1990's, the $SP_N$
-methods was either widely unknown, widely unused, or combination of
-both even though numerical studies showed promising results with
-better solutions than diffusion theory and a significant reduction in
-computational time over more accurate methods such as discrete
-ordinates. Why did this happen? A significant problem, pointed out by
-Larsen, was that little rigor had been applied to the formulation of
-the $SP_N$ equations since their derivation through primarily
-heuristic arguments. Instead, studies at that time focused on simply
-comparing the results of the method to other contemporary transport
-solution strategies. In addition, many problems of interest from the
-literature at the time were either solved using nodal-type methods for
-reactor-sized problems or $S_N$-type methods for benchmark problems
-with intricate material configurations and potentially large flux
-gradients over small spatial domains.
-
-So why reconsider the $SP_N$ equations? Starting in the 1990's and
-primarily due to Larsen and his colleagues, the $SP_N$ equations have
-been given a more rigorous treatment with both variational and
-asymptotic derivations performed as a means of verification. In
-addition, these equations have been more rigorously studied as
-solution methods to MOX fuel problems and have been shown to provide
-accurate solutions. With this mathematical literature to provide a
-solid numerical footing for the method, we look at its application to
-today's challenge problems in full reactor core transport. The
-reduction in numerical complexity of current full core deterministic
-solution methods using the $S_N$ approximation could mean significant
-savings in both compute time and memory required. In addition, the
-characteristics of the solution to the transport problem for a steady
-state reactor core permit diffusion theory to be used; a staple of the
-nuclear industry since its inception. Therefore, if diffusion theory
-is applicable, then finer grained solutions that capture more of the
-physics contained in the transport equation should be possible with
-the $SP_N$ method. In doing so, we also expect from the literature to
-obtain computed responses on the order of accuracy we would expect
-from an appropriately discretized $S_N$ method at a fraction of the
-cost.
-
-To further motivate moving in this direction, recent developments in
-the Exnihilo neutronics package at Oak Ridge National Laboratory have
-permitted generation of the $SP_N$ system of equations for detailed
-full core reactor models. By fully forming these equations and
-formulating them as a linear algebra problem, we now have access to
-all of the modern advancements in computational linear algebra
-including Krylov solvers for asymmetric systems and preconditioning
-methods such as algebraic multigrid. This leads us to then explore the
-applicability of our work in discrete Monte Carlo methods for linear
-systems as a possible solution method for the $SP_N$ equations. If
-formulated correctly, we hypothesize that signficant improvement in
-usage of computational resources may be observed compared to modern
-solution techniques such as those suggested. In addition, solving the
-$SP_N$ equations in this way also breaks away from the $S_N$ forms of
-parallelism where spatial parallelism is achieved by an efficient
-parallel sweep, angular efficiency achieved by pipelining, and energy
-parallelism achieved by decoupling the groups. With the $SP_N$
-equations as a full matrix system, we now can parallelize the problem
-as prescribed by the linear solver, which may be significantly more
-scalable than current $S_N$ transport practices.
-
-In this document we derive the multigroup $SP_N$ equations from the
-Boltzmann transport equation. We then perform a spectral analysis on
-the resulting system of equations to determine its potential
-performance with both modern Krylov methods and Monte Carlo methods
-for asymmetric linear systems. Following this, we devise a Monte Carlo
-synthetic acceleration method as a proposed solution scheme.
-
-%%---------------------------------------------------------------------------%%
-\section{Derivation of the $SP_N$ Equations}
-\label{sec:derivation}
-In this section, we derive the $SP_N$ equations, closely following the
-work of Evans. We begin by stating the general time-independent
-neutron transport equation followed by a derivation of the $P_N$
-equations in planar geometry for multiple energy groups. From these
-equations, we then apply a set of approximations to yield the
-multi-dimensional, multi-group $SP_N$ equations for fixed source
-problems.
-
-\subsection{The Neutron Transport Equation}
-\label{subsec:transport_eq}
-As a starting point we define the time-independent neutron transport
-equation \cite{lewis_computational_1993}:
-\begin{multline}
-  \hat{\Omega} \cdot \vec{\nabla} \psi(\vec{r},\hat{\Omega},E) +
-  \sigma(\vec{r},E) \psi(\vec{r},\hat{\Omega},E) = \\ \int \int
-  \sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}' \rightarrow
-  \hat{\Omega}) \psi(\vec{r},\hat{\Omega}',E') d\Omega' dE' +
-  q(\vec{r},\hat{\Omega},E)\:,
-  \label{eq:general_transport}
-\end{multline}
-with the variables defined as:
-\begin{itemize}
-\item $\vec{r}$ - neutron spatial position
-\item $\hat{\Omega}$ - neutron streaming direction with radial
-  component $\mu$ and azimuthal component $\omega$
-\item $\hat{\Omega}' \cdot \hat{\Omega} = \mu_0$ is the angle of
-  scattering
-\item $E$ - neutron energy
-\item $\psi(\vec{r},\hat{\Omega},E)$ - angular flux
-\item $\sigma(\vec{r},E)$ - total interaction cross section
-\item $\sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}')$ - probability
-  of scattering from direction $\hat{\Omega}'$ into an angular domain
-  $d\hat{\Omega}'$ about the direction $\hat{\Omega}$ and from energy
-  $E'$ to an energy domain $dE'$ about energy $E$
-\item $q(\vec{r},\hat{\Omega},E)$ - external source of neutrons.
-\end{itemize}
-For this work, it is sufficient to formulate
-Eq~(\ref{eq:general_transport}) in 1-dimensonal Cartesian geometry:
-\begin{multline}
-  \mu \frac{\partial}{\partial x} \psi(x,\mu,E) + \sigma(x,E)
-  \psi(x,\mu,E) = \\ \int \int \sigma_s(x,E') \rightarrow
-  E,\hat{\Omega}' \rightarrow \hat{\Omega}) \psi(x,\hat{\Omega}',E')
-  d\Omega' dE' + \frac{q(x,E)}{4 \pi}\:,
-  \label{eq:cart_1d_transport}
-\end{multline}
-where the angular component of the solution is no longer dependent on
-the azimuthal direction of travel and an isotropic source of neutrons
-is assumed.
-
-\subsection{Derivation of the $P_N$ Equations}
-\label{subsec:pn_equations}
-Next, we derive the $P_N$ equations, a simplified form of the general
-transport equation where Legendre polynomials are used to expand the
-angular flux and scattering cross section variables as a means of
-capturing the angular structure of the solution. Before deriving this
-form of the transport equation, we briefly discuss a few properties of
-Legendre polynomials that we will find useful in the derivation.
-
-\subsubsection{Legendre Polynomials}
-\label{subsubsec:legendre_polys}
-The Legendre polynomials are an orthogonal set of functions that
-are solutions to Legendre's differential equation. They have the
-following form \cite{lewis_computational_1993}:
-\begin{equation}
-  P_l(\mu) = \frac{1}{2^l l!}\frac{d^l}{d \mu^l}(\mu^2-1)^l\:.
-  \label{eq:general_legendre_poly}
-\end{equation}
-These functions have several useful properties including
-orthogonality:
-\begin{equation}
-  \int_{-1}^{1} P_l(\mu) P_{l'}(\mu) d\mu = \frac{1}{2l+1}\delta_{l l'}\:,
-  \label{eq:legendre_orthog}
-\end{equation}
-a recurrence relation:
-\begin{equation}
-  \mu P_l(\mu) = \frac{1}{2l+1}[(l+1)P_{l+1}(\mu) + l P_{l-1}(\mu)]\:,
-  \label{eq:legendre_recurrence}
-\end{equation}
-and an addition theorem:
-\begin{equation}
-  P_l(\hat{\Omega} \cdot \hat{\Omega}') = \frac{1}{2n+1}\sum_{m=-l}^l
-  Y_{lm}(\hat{\Omega})Y^*_{lm}(\hat{\Omega}')\:,
-  \label{eq:legendre_addition}
-\end{equation}
-where the functions $Y_{lm}(\hat{\Omega})$ are the spherical
-harmonics. We can form the addition theorem in this way because the
-spherical harmonics are in fact just harmonic multiples of the
-Legendre polynomials:
-\begin{equation}
-  Y_{lm}(\hat{\Omega}) =
-  \sqrt{\frac{(2l+1)(l-m)!}{(l+m)!}}P^m_l(\mu)e^{i \omega m}\:,
-  \label{eq:spherical_harmonic}
-\end{equation}
-where $\omega$ is the azimuthal component of the streaming
-direction. We can reduce Eq~(\ref{eq:legendre_addition}) for the
-planar geometry we are studying by ignoring the azimuthal components
-of the addition theorem. As shown in Eq~(\ref{eq:spherical_harmonic}),
-the azimuthal dependence is given by the harmonic component, $e^{i
-  \omega m}$, and therefore we choose to ignore all terms in
-Eq.~(\ref{eq:legendre_addition}) where $m \neq 0$. This gives:
-\begin{equation}
-  P_l(\hat{\Omega} \cdot \hat{\Omega}') = \frac{1}{2n+1}
-  Y_{l0}(\hat{\Omega})Y^*_{l0}(\hat{\Omega}')\:.
-  \label{eq:legendre_addition_2}
-\end{equation}
-Per Eq~(\ref{eq:spherical_harmonic}) we have:
-\begin{equation}
-  Y_{l0}(\hat{\Omega}) = \sqrt{2l+1}P_l^0(\mu)\:,
-  \label{eq:harmonic_0}
-\end{equation}
-where $P_l^0(\mu) = P_l(\mu)$ is the $0^{th}$ associated Legendre
-function. Finally, we can reduce the addition theorem from
-Eq~(\ref{eq:legendre_addition_2}) with Eq~(\ref{eq:harmonic_0}) to a
-simple product for planar geometry:
-\begin{equation}
-  P_l(\hat{\Omega} \cdot \hat{\Omega}') = P_l(\mu)P_l(\mu')\:.
-  \label{eq:legendre_addition_3}
-\end{equation}
-
-\subsubsection{Planar $P_N$ Equations}
-\label{subsubsec:planar_pn}
-With the Legendre polynomial properties defined above, we can proceed
-by deriving the $P_N$ equations for planar geometry. For these
-equations, we will start by assuming a monoenergtic field of neutrons
-such that we are solving the following reduced form of the transport
-equation: 
-\begin{equation}
-  \mu \frac{\partial}{\partial x} \psi(x,\mu) + \sigma(x) \psi(x,\mu)
-  = \int d\Omega' \sigma_s(x,\hat{\Omega}' \rightarrow \hat{\Omega})
-  \psi(\vec{r},\hat{\Omega}') + \frac{q(x)}{4 \pi}\:.
-  \label{eq:mono_transport}
-\end{equation}
-The $P_N$ equations introduce the approximation that the angular
-dependence of the scattering cross section, $\sigma_s$, and the angular
-flux $\psi$, can be discretized by expanding them in Legendre
-polynomials as follows:
-\begin{equation}
-  \psi(x,\mu) = \sum_{n=0}^\infty (2n+1)P_n(\mu)\phi_n(x)\:,
-  \label{eq:flux_expansion}
-\end{equation}
-\begin{equation}
-  \sigma_{sm}(x) = \sum_{m=0}^\infty (2m+1)P_m(\mu)\sigma_s(x)\:,
-  \label{eq:scattering_expansion}
-\end{equation}
-where we have supressed the $2\pi$ generated by integrating away the
-azimuthal angular component and $\phi_n(x)$ in
-Eq~(\ref{eq:flux_expansion}) is referred to as the $n^{th}$ Legendre
-moment of the neutron flux and is given by:
-\begin{equation}
-  \phi_n(x) = \int_{-1}^1 P_n(\mu)\psi(x,\mu)d\mu\:.
-  \label{eq:legendre_moments}
-\end{equation}
-We first insert the expansions given by Eq~(\ref{eq:flux_expansion})
-and (\ref{eq:scattering_expansion}) into the planar transport equation
-given by Eq~(\ref{eq:mono_transport}):
-\begin{multline}
-  \frac{\partial}{\partial x}\Big[\sum_{n=0}^\infty (2n+1) \phi_n \mu
-    P_n(\mu) \Big] + \sigma \sum_{n=0}^\infty (2n+1) \phi_n P_n(\mu) =
-  \\ \int_{-1}^1 \sum_{m=0}^\infty (2m+1) \sigma_{sm} P_m(\mu_0)
-  \sum_{n=0}^\infty (2n+1) \phi_n P_n(\mu') d \mu' + q\:,
-  \label{eq:pn_deriv_1}
-\end{multline}
-where the dependence on the spatial variable, $x$, has been
-supressed. To arrive at the $P_N$ equations we multiply
-Eq~(\ref{eq:pn_deriv_1}) by $P_m(\mu)$ and integrate over the angular
-domain $\int_{-1}^1 d \mu$. We will look at each term in
-Eq~(\ref{eq:pn_deriv_1}) individually.
-
-\paragraph{Streaming Term}
-We first apply the multiplication and integration as prescribed above:
-\begin{equation}
-  \frac{\partial}{\partial x}\Bigg[\sum_{n=0}^\infty (2n+1) \mu \phi_n
-    P_n(\mu) \Bigg] \rightarrow \int_{-1}^1 \frac{\partial}{\partial
-    x}\Bigg[\sum_{n=0}^\infty (2n+1) \phi_n \mu P_n(\mu) P_m(\mu) \Bigg]
-  d \mu\:.
-  \label{eq:pn_deriv_2}
-\end{equation}
-The $\mu P_n(\mu)$ term can be eliminated via the recurrence relation
-given by Eq~(\ref{eq:legendre_recurrence}):
-\begin{equation}
-\int_{-1}^1 \frac{\partial}{\partial x}\Bigg[\sum_{n=0}^\infty (2n+1)
-  \frac{\phi_n}{2n+1}[(n+1)P_{n+1}(\mu) + n P_{n-1}(\mu)] P_m(\mu)
-  \Bigg] d\mu\:,
-\label{eq:pn_deriv_3}
-\end{equation}
-which expands to:
-\begin{equation}
-  \sum_{n=0}^\infty \frac{\partial}{\partial
-    x}\phi_n\Bigg[(n+1)\int_{-1}^1 P_{n+1}(\mu)P_m(\mu) d\mu + n
-    \int_{-1}^1 P_{n-1}(\mu) P_m(\mu) d\mu \Bigg] \:.
-  \label{eq:pn_deriv_4}
-\end{equation}
-This reveals the orthogonality relation given by
-Eq~(\ref{eq:legendre_orthog}) that when inserted into
-Eq~(\ref{eq:pn_deriv_3}):
-\begin{equation}
-  \sum_{n=0}^\infty \frac{\partial}{\partial
-    x}\phi_n\Bigg[(n+1)\frac{1}{2n+1}\delta_{n,n+1} +
-    n\frac{1}{2n+1}\delta_{n,n-1} \Bigg] \:.
-  \label{eq:pn_deriv_5}
-\end{equation}
-We can then distribute the Legendre moment to arrive at the final form
-of the streaming term:
-\begin{equation}
-  \sum_{n=0}^\infty \frac{\partial}{\partial x} \frac{1}{2n+1} \Big[
-    (n+1) \phi_{n+1} + n \phi_{n-1} \Big] \:.
-  \label{eq:pn_deriv_6}
-\end{equation}
-
-\paragraph{Collision Term}
-To reduce the collision term, the orthogonality relation is again
-applied after the integration:
-\begin{equation}
-  \sigma \sum_{n=0}^\infty (2n+1) \phi_n P_n(\mu) \rightarrow
-  \int_{-1}^1 \sigma \sum_{n=0}^\infty (2n+1) \phi_n P_n(\mu) P_m(\mu)
-  d\mu\:,
-  \label{eq:pn_deriv_7}
-\end{equation}
-\begin{equation}
-  \sigma \sum_{n=0}^\infty (2n+1) \phi_n \frac{1}{2n+1}\:,
-  \label{eq:pn_deriv_8}
-\end{equation}
-giving for the final collision term:
-\begin{equation}
-  \sum_{n=0}^\infty \sigma \phi_n \:.
-  \label{eq:pn_deriv_9}
-\end{equation}
-
-\paragraph{Scattering Term}
-For the scattering term:
-\begin{multline}
-  \int_{-1}^1 \sum_{m=0}^\infty (2m+1) \sigma_{sm} P_m(\mu_0)
-  \sum_{n=0}^\infty (2n+1) \phi_n P_n(\mu') d\mu' \rightarrow\\
-  \int_{-1}^1 \int_{-1}^1 \sum_{m=0}^\infty (2m+1) \sigma_{sm} P_m(\mu)
-  P_m(\mu_0) \sum_{n=0}^\infty (2n+1) \phi_n P_n(\mu') d\mu' d\mu\:.
-  \label{eq:pn_deriv_10}
-\end{multline}
-The addition thereom from Eq~(\ref{eq:legendre_addition_3}) is applied
-to give:
-\begin{equation}
-  \int_{-1}^1 \int_{-1}^1 \sum_{m=0}^\infty (2m+1) \sigma_{sm} P_m(\mu)
-  P_m(\mu)P_m(\mu') \sum_{n=0}^\infty (2n+1) \phi_n P_n(\mu') d\mu' d\mu\:,
-  \label{eq:pn_deriv_11}
-\end{equation}
-which can be rearranged as:
-\begin{equation}
-  \sum_{m=0}^\infty \sum_{n=0}^\infty (2m+1) \sigma_{sm} (2n+1) \phi_n
-  \int_{-1}^1 P_m(\mu') P_n(\mu') d\mu' \int_{-1}^1 P_m(\mu) P_m(\mu)
-  d\mu\:.
-  \label{eq:pn_deriv_12}
-\end{equation}
-Again we can apply orthogonality to eliminate the Legendre polynomials:
-\begin{equation}
-  \sum_{m=0}^\infty \sum_{n=0}^\infty (2m+1) \sigma_{sm} (2n+1) \phi_n
-  \frac{1}{2n+1}\delta_{nm}\frac{1}{2m+1}\delta_{mm}\:,
-  \label{eq:pn_deriv_13}
-\end{equation}
-which is reduced to:
-\begin{equation}
-  \sum_{n=0}^\infty \sigma_{sn} \phi_n\:,
-  \label{eq:pn_deriv_14}
-\end{equation}
-with the dependence on the index $m$ eliminated.
-
-\paragraph{Source Term}
-The last term we are concerned with in Eq~(\ref{eq:pn_deriv_1}) is the
-source term:
-\begin{equation}
-  q \rightarrow \int_{-1}^1 q P_n(\mu) d\mu \:.
-  \label{eq:pn_deriv_15}
-\end{equation}
-We can leverage orthogonality by multiplying by $P_0(\mu) = 1$:
-\begin{equation}
-  \int_{-1}^1 q P_0 P_n(\mu) d\mu = \frac{q}{2*0+1}\delta_{n0}\:,
-  \label{eq:pn_deriv_16}
-\end{equation}
-giving a final source term of:
-\begin{equation}
-  q\delta_{n0}\:.
-  \label{eq:pn_deriv_17}
-\end{equation}
-
-Now that we have expanded all angular dependent terms in
-Eq~(\ref{eq:pn_deriv_1}) and reduced them appropriately, we can
-combine them to generate the $P_N$ equations:
-\begin{equation}
-    \sum_{n=0}^\infty \frac{\partial}{\partial x} \frac{1}{2n+1} \Big[
-      (n+1) \phi_{n+1} + n \phi_{n-1} \Big] + \sum_{n=0}^\infty \sigma
-    \phi_n = \sum_{n=0}^\infty \sigma_{sn} \phi_n + q\delta_{n0}\:.
-  \label{eq:pn_deriv_18}
-\end{equation}
-More formally, the $P_N$ equations are written as:
-\begin{equation}
-   \frac{1}{2n+1} \frac{\partial}{\partial x}\Big[ (n+1) \phi_{n+1} + n
-     \phi_{n-1} \Big] + \Sigma_n \phi_n = q\delta_{n0} \:,
-  \label{eq:final_pn_equations}
-\end{equation}
-where $\Sigma_n = \sigma-\sigma_{sn}$ and the summations are truncated
-at some level of approximation $N$ such that $n = 0,1,\dotsc,N$. This
-yields a set of $N+1$ equations for $N+2$ flux moments. We therefore
-require an additional equation to close the system. In accordance with
-the series truncation as an approximation we choose the last moment in
-the expansion to be zero:
-\begin{equation}
-  \phi_{N+1} = 0\:.
-  \label{eq:pn_closure}
-\end{equation}
-As an example, we will construct the P5 equations from
-Eq~(\ref{eq:final_pn_equations}) and the closure given by
-Eq~(\ref{eq:pn_closure}):
-\begin{subequations}
-  \begin{gather}
-   \frac{\partial}{\partial x}\phi_{1} + \Sigma_0 \phi_0 = q\:,\\ 
-   \frac{1}{3} \frac{\partial}{\partial x}\Big[ 2
-     \phi_{2} + \phi_{0} \Big] + \Sigma_1 \phi_1 = 0\:,\\
-   \frac{1}{5} \frac{\partial}{\partial x}\Big[ 3 \phi_{3} + 2
-     \phi_{1} \Big] + \Sigma_2 \phi_2 = 0 \:,\\
-   \frac{1}{7} \frac{\partial}{\partial x}\Big[ 4 \phi_{4} + 3
-     \phi_{2} \Big] + \Sigma_3 \phi_3 = 0 \:,\\
-   \frac{1}{9} \frac{\partial}{\partial x}\Big[ 5 \phi_{5} + 4
-     \phi_{3} \Big] + \Sigma_4 \phi_4 = 0 \:,\\
-   \frac{1}{11} \frac{\partial}{\partial x} 5 \phi_{4} + \Sigma_5
-   \phi_5 = 0 \:.
-  \end{gather}
-  \label{eq:p5_equations}
-\end{subequations}
-This gives us a set of 6 coupled equations for the six Legendre
-moments requested defined over the entire spatial domain for a single
-energy group. In practice, only odd-numbered $P_N$ orders are
-generally used \cite{lewis_computational_1993}. This is due to the
-fact that using odd $N$ yields an even number of $N+1$ equations which
-can be split evenly on the left and right boundaries of the problem to
-facilitate the description of boundary conditions. We will choose this
-convention when deriving the boundary conditions.
-
-\subsubsection{Boundary Conditions for the $P_N$ Equations}
-\label{subsbusec:bcs_pn}
-Per the analysis in \cite{lewis_computational_1993}, two types of
-boundary conditions will be discussed: reflecting and Marshak. Marshak
-boundary conditions can be used to specify both vacuum conditions and
-isotropic source conditions on the boundary.
-
-\paragraph{Reflecting Boundary Conditions}
-In this case, the incoming flux should be equivalent to the outgoing
-flux at the boundary point $x_b$:
-\begin{equation}
-  \psi(x_b,\mu) = \psi(x_b,-\mu)\:.
-  \label{eq:reflecting_condition}
-\end{equation}
-Given the legendre expansion for the flux defined in
-Eq~(\ref{eq:flux_expansion}) and the Legendre polynomial property that
-$P_n(\mu) = (-1)^n P_n(-\mu)$, the condition specified by
-Eq~(\ref{eq:reflecting_condition}) can be satisfied if
-\begin{equation}
-  \phi_n = 0,\ \forall \ \text{odd}\ n\:,
-  \label{eq:reflecting_condition_odd}
-\end{equation}
-as all even $n$ yield $P_n(\mu) = P_n(-\mu)$ and therefore an
-equivalent reflecting condition for the flux moments.
-
-\paragraph{Marshak Boundary Conditions}
-The Marshak conditions come directly from the Legendre moments of the
-flux:
-\begin{equation}
-  \int_{\mu_b} P_i(\mu) \psi(\mu) d\mu = \int_{\mu_b} P_i(\mu)
-  \psi_b(\mu) d\mu\ \ \ \ \ \ \ \ \ \ \text{for $i=1,3,...,N$}\:,
-  \label{eq:general_marshak}
-\end{equation}
-where $\psi_b(\mu)$ is the prescribed angular flux on the boundary of
-interest and $\mu_b$ the angular domain defined by the boundary. To
-discretize this condition, we again insert the angular flux expansions
-from Eq~(\ref{eq:flux_expansion}) for the fluxes defined in the
-domain:
-\begin{equation}
-  \int_{\mu_b} P_i(\mu) \sum_{n=0}^N (2n+1) \phi_n P_n(\mu) d\mu =
-  \int_{\mu_b} P_i(\mu) \psi_b(\mu) d\mu\:.
-  \label{eq:marshak_expanded}
-\end{equation}
-The boundary flux, $\phi_b(\mu)$ is assumed to known and therefore
-Eq~(\ref{eq:marshak_expanded}) defines a set of $(N+1)/2$ equations to
-be solved on each boundary in the planar case, closing the system in
-the spatial domain.
-
-As an example of applying the Marshak conditions, consider the $P_3$
-case with an isotropic boundary source $\phi_b$ on the left side of
-the domain. In this case, the angular domain over which the boundary
-flux is defined will be $\mu_b \in [0,1]$, giving the bounds of
-integration. We first expand the summation for $i=1$:
-\begin{equation}
-  \int_0^1 \mu \Bigg[ \phi_0 + 3\phi_1\mu +
-    \frac{5}{2}\phi_2(3\mu^2-1) + \frac{7}{2}\phi_3(5\mu^3-3\mu)
-    \Bigg] d\mu = \int_0^1 \mu \phi_b d\mu\:,
-  \label{eq:marshak_p1_deriv_1}
-\end{equation}
-and then $i=3$:
-\begin{equation}
-  \int_0^1 \frac{1}{2}(5\mu^3-3\mu) \Bigg[ \phi_0 + 3\phi_1\mu +
-    \frac{5}{2}\phi_2(3\mu^2-1) + \frac{7}{2}\phi_3(5\mu^3-3\mu)
-    \Bigg] d\mu = \int_0^1 \frac{1}{2}(5\mu^3-3\mu) \phi_b d\mu\:.
-  \label{eq:marshak_p3_deriv_1}
-\end{equation}
-Expanding the polynomials in $\mu$ and carrying out the simple
-integration then gives 2 equations for the left hand boundary:
-\begin{equation}
-  \phi_0 + 2\phi_1 + \frac{5}{4}\phi_2 = \phi_b\:,
-  \label{eq:marshak_p1_deriv_2}
-\end{equation}
-\begin{equation}
-  \phi_0 - 5\phi_2 - 8\phi_3 = \phi_b\:.
-  \label{eq:marshak_p1_deriv_3}
-\end{equation}
-The right hand side boundary condition will yield 2 complementary
-equations if Marshak conditions are used or 2 non-zero moments to be
-solved for if reflected conditions are used. The formulation above
-also holds for vacuum conditions where $\phi_b = 0$.
-
-\subsection{Formulation of the $SP_N$ Equations}
-\label{subsec:spn_equations}
-The $P_N$ equations give $N+1$ coupled first-order equations
-capturing the spatial and angular-dependence of the solution. In
-multiple dimensions, the equation set becomes large and coupled not
-only through angular moments but also through the spatial
-variables. As a simpler alternative to multidimensional $P_N$
-solutions, Gelbard recognized in 1960 that the planar $P_N$ equations
-could be simplified and applied an ad-hoc method to extend them to
-multiple dimensions, yielding the $SP_N$ equations. These equations
-are not only fewer in number, but also take on a diffusion-like form
-while maintaining the angular character of the flux, making them
-amenable to solutions with modern diffusion methods.
-
-First, the $P_N$ equations can be simplified to $(N+1)/2$ second-order
-equations by solving for the $n^th$ Legendre flux moment in the
-odd-order equations:
-\begin{equation}
-  \phi_n = \frac{1}{\Sigma_n}\Bigg[ q \delta_{no} -
-    \frac{\partial}{\partial x}\Big(\frac{n}{2n+1}\phi_{n-1} +
-    \frac{n+1}{2n+1} \phi_{n+1} \Big) \Bigg]\:, 
-  \label{eq:odd_moments}
-\end{equation}
-for $n = 1,3,\cdots,N$ and $\delta_{no} = 0\ \forall n \neq 0$. We can
-insert the odd moments into Eq~(\ref{eq:final_pn_equations}) to get a
-reduced group of equations for the even moments:
-\begin{multline}
-  -\frac{\partial}{\partial x}
-  \Bigg[\frac{n}{2n+1}\frac{1}{\Sigma_{n-1}} \frac{\partial}{\partial
-      x} \Big(\frac{n-1}{2n-1} \phi_{n-2} + \frac{n}{2n-1}\phi_n \Big)
-    \\+ \frac{n+1}{2n+1}\frac{1}{\Sigma_{n+1}} \frac{\partial}{\partial
-      x} \Big(\frac{n+1}{2n+3}\phi_n + \frac{n+2}{2n+3}\phi_{n+2}\Big)
-    \Bigg] \\+ \Sigma_n \phi_n = q \delta_{n0}\ \ \ \ \ \ \ \ \ n =
-  0,2,4,\cdots,N\:.
-  \label{eq:reduced_pn}
-\end{multline}
-Immediately, we note the diffusion-like nature of
-Eq~(\ref{eq:reduced_pn}) as compared to the original $P_N$
-equations. To extend these equations to multiple dimensions, gelbard
-simple replaced the x-plane spatial derivatives in the reduced set of
-equations with general multidimensional gradient operators:
-\begin{multline}
-  -\nabla \cdot \Bigg[\frac{n}{2n+1}\frac{1}{\Sigma_{n-1}} \nabla
-    \Big(\frac{n-1}{2n-1} \phi_{n-2} + \frac{n}{2n-1}\phi_n \Big) \\+
-    \frac{n+1}{2n+1}\frac{1}{\Sigma_{n+1}} \nabla
-    \Big(\frac{n+1}{2n+3}\phi_n + \frac{n+2}{2n+3}\phi_{n+2}\Big)
-    \Bigg] \\+ \Sigma_n \phi_n = q \delta_{n0}\ \ \ \ \ \ \ \ \ n =
-  0,2,4,\cdots,N\:,
-  \label{eq:spn_equations}
-\end{multline}
-yielding a multidimensional set of $(N+1)/1$ angular coupled equations
-known as the $SP_N$. Again, we provide closure to this set of
-equations with $\phi_{N+1} = 0$. As a concrete example, we will
-consider the $SP_7$ equations:
-\begin{subequations}
-  \begin{gather}
-    -\nabla \cdot \frac{1}{3 \Sigma_1} \nabla ( \phi_0 + 2\phi_2 ) +
-    \Sigma_0 \phi_0 = q \\ 
-    -\nabla \cdot \Bigg[ \frac{2}{15 \Sigma_1} \nabla ( \phi_0 + 2\phi_2
-      ) + \frac{3}{35 \Sigma_3}\nabla( 3\phi_2 + 4\phi_4)\Bigg] +
-    \Sigma_2 \phi_2 = 0\\
-    -\nabla \cdot \Bigg[ \frac{4}{63 \Sigma_3} \nabla ( 3\phi_2 +
-      4\phi_4 ) + \frac{5}{99 \Sigma_5}\nabla( 5\phi_4 +
-      6\phi_6)\Bigg] + \Sigma_4 \phi_4 = 0\\
-    -\nabla \cdot \Bigg[ \frac{6}{143 \Sigma_5} \nabla ( 5\phi_4 +
-      6\phi_6 ) + \frac{7}{195 \Sigma_7}\nabla(7\phi_6)\Bigg] +
-    \Sigma_6 \phi_6 = 0 \:.
-  \end{gather}
-  \label{eq:sp7_equations}
-\end{subequations}
-To further modify these equations, we can use a change of variables to
-create a new group of equations such that the gradients are operating
-on a single vector:
-\begin{subequations}
-  \begin{gather}
-    u_1 = \phi_0 + 2\phi_2 \\
-    u_2 = 3\phi_2 + 4\phi_4 \\
-    u_3 = 5\phi_4 + 6\phi_6 \\
-    u_4 = 7\phi_6 \:,
-  \end{gather}
-  \label{eq:spn7_subs}
-\end{subequations}
-or equivalently:
-\begin{subequations}
-  \begin{gather}
-    \phi_0 = u_1 - \frac{2}{3}u_2 + \frac{8}{15}u_3 -
-    \frac{16}{35}u_4 \\
-    \phi_2 = \frac{1}{3}u_2 - \frac{4}{15}u_3 + \frac{8}{35}u_4\\ 
-    \phi_4 = \frac{1}{5}u_3 - \frac{6}{35}u_4\\
-    \phi_6 = \frac{1}{7}u_4\:.
-  \end{gather}
-  \label{eq:spn7_subs_inverse}
-\end{subequations}
-When substituted into Eq~(\ref{eq:sp7_equations}), these terms give:
-\begin{subequations}
-  \begin{gather}
-    -\nabla \cdot \frac{1}{3 \Sigma_1} \nabla u_1 + \Sigma_0 \Bigg[
-    u_1 - \frac{2}{3}u_2 + \frac{8}{15}u_3 - \frac{16}{35}u_4 \Bigg]
-    = -q \\
-    -\nabla \cdot \Bigg[ \frac{2}{15 \Sigma_1} \nabla u_1 +
-    \frac{3}{35 \Sigma_3} \nabla u_2 \Bigg] + \Sigma_2 \Bigg[
-    \frac{1}{3}u_2 - \frac{4}{15}u_3 + \frac{8}{35}u_4 \Bigg] = 0 \\
-    -\nabla \cdot \Bigg[ \frac{4}{63 \Sigma_3} \nabla u_2 +
-    \frac{5}{99 \Sigma_5} \nabla u_3 \Bigg] + \Sigma_4 \Bigg[
-    \frac{1}{5}u_3 - \frac{6}{35}u_4 \Bigg] = 0 \\ 
-    -\nabla \cdot \Bigg[ \frac{6}{143 \Sigma_5} \nabla u_3 +
-    \frac{7}{195 \Sigma_7} \nabla u_4 \Bigg] + \Sigma_6 \Bigg[
-    \frac{1}{7}u_4 \Bigg] = 0 \:.
-  \end{gather}
-  \label{eq:spn7_subs_equations}
-\end{subequations}
-If we rearrange the Eq~(\ref{eq:spn7_subs_equations} such that only
-one divergence operation is present in each equation, we can formulate
-this as a matrix system of 4 equations in the case of the $SP_7$
-approximation:
-\begin{equation}
-  -\nabla \cdot D_n \nabla u_n + \sum_{m=1}^4 A_{nm} u_m =
-  q_n\ \ \ \ \ \ \ n = 1,2,3,4\:,
-  \label{eq:spn_matrix}
-\end{equation}
-with $\mathbf{u}$ the vector of solution variables:
-\begin{equation}
-  \mathbf{u} = ( u_1\ \ u_2\ \ u_3\ \ u_4 )^T \:,
-  \label{eq:spn7_solution_vector}
-\end{equation}
-$\mathbf{D}$ the vector of effective diffusion coefficients:
-\begin{equation}
-  \mathbf{D} = \Bigg( \frac{1}{3\Sigma_1}\ \ \frac{1}{7\Sigma_3}\ \
-  \frac{1}{11\Sigma_5}\ \ \frac{1}{15\Sigma_7} \Bigg)^T\:,
-  \label{eq:spn7_diffusion_coeffs}
-\end{equation}
-$\mathbf{q}$ the vector of source terms where the $0^{th}$ moment
-source has now been distributed through the system:
-\begin{equation}
-  \mathbf{q} = (
-  q\ \ -\frac{2}{3}q\ \ \frac{8}{15}q\ \ -\frac{16}{35}q )^T\:,
-  \label{eq:spn7_source_vector}
-\end{equation}
-and $\mathbf{A}$ a matrix of angular scattering terms:
-% NOTE: I copied the following matrix directly out of Tom's tech note
-% on the SPn equations which I am effectively following here because I
-% was feeling lazy. I have verified its correctness.
-\begin{equation}
-  \mathbf{A} = 
-  \begin{bmatrix}
-    (\Sigma_0) &
-    (-\frac{2}{3}\Sigma_0) &
-    (\frac{8}{15}\Sigma_0) &
-    (-\frac{16}{35}\Sigma_0) \\
-    %%
-    &&&\\
-    %%
-    (-\frac{2}{3}\Sigma_0) &
-    (\frac{4}{15}\Sigma_0 + \frac{1}{3}\Sigma_2) &
-    (-\frac{16}{45}\Sigma_0 - \frac{4}{9}\Sigma_2) &
-    (\frac{32}{105}\Sigma_0 + \frac{8}{21}\Sigma_2) \\
-    %%
-    &&&\\
-    %%
-    (\frac{8}{15}\Sigma_0) &
-    (-\frac{16}{45}\Sigma_0 - \frac{4}{9}\Sigma_2) &
-    (\frac{64}{225}\Sigma_0 + \frac{16}{45}\Sigma_2 + \frac{9}{25}\Sigma_4) &
-    (-\frac{128}{525}\Sigma_0 - \frac{32}{105}\Sigma_2 - \frac{54}{175}\Sigma_4)
-    \\ 
-    %%
-    &&&\\
-    %%
-    (-\frac{16}{35}\Sigma_0) &
-    (\frac{32}{105}\Sigma_0 + \frac{8}{21}\Sigma_2) &
-    (-\frac{128}{525}\Sigma_0 - \frac{32}{105}\Sigma_2 - \frac{54}{175}\Sigma_4)
-    & 
-    (\frac{256}{1225}\Sigma_0 + \frac{64}{245}\Sigma_2 +
-    \frac{324}{1225}\Sigma_4 + \frac{13}{49}\Sigma_6)
-  \end{bmatrix}\:.
-  \label{eq:A_matrix}
-\end{equation}
-Note that the term $\sum_{m=1}^4 A_{nm} u_m$ in
-Eq~(\ref{eq:spn_matrix}) couples the moments in each equation while
-the diffusive term in each equation is only for a single
-'psuedo-moment' $u_n$. This completes the derivation of the $SP_7$
-equations for a single energy group. As noted by Evans, lower order
-$SP_N$ approximations can be generated by setting higher order even
-moments in this system to zero (e.g. $\phi_6 = \phi_4 = 0$ yields the
-$SP_3$ equations).
-
-\subsubsection{Boundary Conditions for the $SP_N$ Equations}
-\label{subsubsec:spn_boundary_conditions}
-We approach the formulation of the boundary conditions in much the
-same way as we did for the $P_N$ equations with both reflecting and
-Marshak conditions possible. To begin, we perform the expansion in
-Eq~(\ref{eq:marshak_expanded}) for the left side of the planar system,
-this time for $N=7$ to correspond to our $SP_7$ system. For $i=1$:
-\begin{multline}
-  \int_0^1 \mu \Bigg[ \phi_0 + 3\phi_1\mu +
-    \frac{5}{2}\phi_2(3\mu^2-1) + \frac{7}{2}\phi_3(5\mu^3-3\mu)
-    +\\ \frac{1}{8}(63\mu^5-70\mu^3+15\mu) \Bigg] d\mu = \int_0^1 \mu
-  \phi_b d\mu\:,
-  \label{eq:spn_bnd_p1}
-\end{multline}
-for $i=3$,
-\begin{multline}
-  \int_0^1 \frac{1}{2}(5\mu^3-3\mu) \Bigg[ \phi_0 + 3\phi_1\mu +
-    \frac{5}{2}\phi_2(3\mu^2-1) + \frac{7}{2}\phi_3(5\mu^3-3\mu) +\\
-    \frac{1}{8}(63\mu^5-70\mu^3+15\mu) \Bigg] d\mu = \int_0^1
-  \frac{1}{2}(5\mu^3-3\mu) \phi_b d\mu\:,
-  \label{eq:spn_bnd_p3}
-\end{multline}
-for $i=5$:
-\begin{multline}
-  \int_0^1 \frac{1}{8}(63\mu^5-70\mu^3+15\mu) \Bigg[ \phi_0 + 3\phi_1\mu +
-    \frac{5}{2}\phi_2(3\mu^2-1) + \frac{7}{2}\phi_3(5\mu^3-3\mu) +\\
-    \frac{1}{8}(63\mu^5-70\mu^3+15\mu) \Bigg] d\mu = \int_0^1
-  \frac{1}{8}(63\mu^5-70\mu^3+15\mu) \phi_b d\mu\:,
-  \label{eq:spn_bnd_p5}
-\end{multline}
-and for $i=7$:
-\begin{multline}
-  \int_0^1 \frac{1}{16}(429\mu^7-693\mu^5+315\mu^3+35\mu) \Bigg[
-    \phi_0 + 3\phi_1\mu + \frac{5}{2}\phi_2(3\mu^2-1) +
-    \frac{7}{2}\phi_3(5\mu^3-3\mu) \\+
-    \frac{1}{8}(63\mu^5-70\mu^3+15\mu) +
-    \frac{1}{16}(429\mu^7-693\mu^5+315\mu^3+35\mu) \Bigg] d\mu =\\
-  \int_0^1 \frac{1}{16}(429\mu^7-693\mu^5+315\mu^3+35\mu) \phi_b
-  d\mu\:.
-  \label{eq:spn_bnd_p7}
-\end{multline}
-Carrying out the simple integrations yields the following system of
-equations:
-\begin{subequations}
-  \begin{gather}
-    \frac{1}{2}\phi_0 + \phi_1 + \frac{5}{8}\phi_2 -
-    \frac{3}{16}\phi_4 + \frac{13}{128}\phi_6 =
-    \frac{1}{2}\phi_{b}\\ -\frac{1}{8}\phi_0 + \frac{5}{8}\phi_2 +
-    \phi_3 + \frac{81}{128}\phi_4 - \frac{13}{64}\phi_6 =
-    -\frac{1}{8}\phi_{b}\\ \frac{1}{16}\phi_0 - \frac{25}{128}\phi_2 +
-    \frac{81}{128}\phi_4 + \phi_5 + \frac{325}{512}\phi_6 =
-    \frac{1}{16}\phi_{b}\\ -\frac{5}{128}\phi_0 + \frac{7}{64}\phi_2 -
-    \frac{105}{512}\phi_4 + \frac{325}{512}\phi_6 + \phi_7 =
-    -\frac{5}{128}\phi_{b}\:,
-  \end{gather}
-  \label{eq:spn_bnd_integrated}
-\end{subequations}
-where $\phi_b$ is again an isotropic source prescribed on the planar
-boundary. For the odd moment in each boundary equation, we insert
-Eq~(\ref{eq:odd_moments}) to remove them, leaving only the even
-moments and a set of differential equations:
-\begin{subequations}
-  \begin{gather}
-    \frac{1}{2}\phi_0 + \frac{1}{3\Sigma_1}\frac{\partial}{\partial
-      x}(\phi_0+2\phi_2) + \frac{5}{8}\phi_2 - \frac{3}{16}\phi_4 +
-    \frac{13}{128}\phi_6 = \frac{1}{2}\phi_{b}\\ -\frac{1}{8}\phi_0 +
-    \frac{5}{8}\phi_2 + \frac{1}{7\Sigma_3}\frac{\partial}{\partial
-      x}(3\phi_2 + 4\phi_4) + \frac{81}{128}\phi_4 -
-    \frac{13}{64}\phi_6 = -\frac{1}{8}\phi_{b}\\ \frac{1}{16}\phi_0 -
-    \frac{25}{128}\phi_2 + \frac{81}{128}\phi_4 +
-    \frac{1}{11\Sigma_5}\frac{\partial}{\partial x}(5\phi_4 + 6\phi_6)
-    + \frac{325}{512}\phi_6 =
-    \frac{1}{16}\phi_{b}\\ -\frac{5}{128}\phi_0 + \frac{7}{64}\phi_2 -
-    \frac{105}{512}\phi_4 + \frac{325}{512}\phi_6 +
-    \frac{1}{15\Sigma_7}\frac{\partial}{\partial x}(7\phi_6)
-    = -\frac{5}{128}\phi_{b}\:.
-  \end{gather}
-  \label{eq:spn_bnd_subs}
-\end{subequations}
-We can again make the substitution of variables given by
-Eqs~(\ref{eq:spn7_subs}) and (\ref{eq:spn7_subs_inverse}) for
-consitency with the equations defined on the domain. In addition, we
-apply the $SP_N$ approximation to the derivatives by assuming they are
-instead multidimensional gradients on the boundary such that
-$\frac{\partial}{\partial x} \rightarrow
-\hat{\mathbf{n}}\cdot\nabla$. Doing this gives:
-\begin{multline}
-    \frac{1}{2}\Big(u_1 - \frac{2}{3}u_2 + \frac{8}{15}u_3 -
-    \frac{16}{35}u_4\Big) +
-    \frac{1}{3\Sigma_1}\hat{\mathbf{n}}\cdot\nabla\Big(\Big(u_1 -
-    \frac{2}{3}u_2 + \frac{8}{15}u_3 -
-    \frac{16}{35}u_4\Big)+\\2\Big(\frac{1}{3}u_2 - \frac{4}{15}u_3 +
-    \frac{8}{35}u_4\Big)\Big) + \frac{5}{8}\Big(\frac{1}{3}u_2 -
-    \frac{4}{15}u_3 + \frac{8}{35}u_4\Big)
-    -\\ \frac{3}{16}\Big(\frac{1}{5}u_3 - \frac{6}{35}u_4\Big) +
-    \frac{13}{128}\Big(\frac{1}{7}u_4\Big) = \frac{1}{2}\phi_{b}\:,
-\end{multline}
-\begin{multline}
-    -\frac{1}{8}\Big(u_1 - \frac{2}{3}u_2 + \frac{8}{15}u_3 -
-    \frac{16}{35}u_4\Big) + \frac{5}{8}\Big(\frac{1}{3}u_2 -
-    \frac{4}{15}u_3 + \frac{8}{35}u_4\Big)
-    +\\ \frac{1}{7\Sigma_3}\hat{\mathbf{n}}\cdot\nabla\Big
-    (3\Big(\frac{1}{3}u_2 - \frac{4}{15}u_3 + \frac{8}{35}u_4\Big) +
-    4\Big(\frac{1}{5}u_3 - \frac{6}{35}u_4\Big)\Big) +
-    \frac{81}{128}\Big(\frac{1}{5}u_3 -\\ \frac{6}{35}u_4\Big) -
-    \frac{13}{64}\Big(\frac{1}{7}u_4\Big) = -\frac{1}{8}\phi_{b}\:,
-\end{multline}
-\begin{multline}
-    \frac{1}{16}\Big(u_1 - \frac{2}{3}u_2 + \frac{8}{15}u_3 -
-    \frac{16}{35}u_4\Big) - \frac{25}{128}\Big(\frac{1}{3}u_2 -
-    \frac{4}{15}u_3 +\\ \frac{8}{35}u_4\Big) +
-    \frac{81}{128}\Big(\frac{1}{5}u_3 - \frac{6}{35}u_4\Big) +
-    \frac{1}{11\Sigma_5}\hat{\mathbf{n}}\cdot\nabla\Big(5\Big(\frac{1}{5}u_3
-    -\\ \frac{6}{35}u_4\Big) + 6\Big(\frac{1}{7}u_4\Big)\Big) +
-    \frac{325}{512}\Big(\frac{1}{7}u_4\Big) = \frac{1}{16}\phi_{b}\:,
-\end{multline}
-\begin{multline}
-    -\frac{5}{128}\Big(u_1 - \frac{2}{3}u_2 + \frac{8}{15}u_3 -
-    \frac{16}{35}u_4\Big) + \frac{7}{64}\Big(\frac{1}{3}u_2 -
-    \frac{4}{15}u_3 + \frac{8}{35}u_4\Big)
-    -\\ \frac{105}{512}\Big(\frac{1}{5}u_3 - \frac{6}{35}u_4\Big) +
-    \frac{325}{512}\Big(\frac{1}{7}u_4\Big)
-    +\\ \frac{1}{15\Sigma_7}\hat{\mathbf{n}}
-    \cdot\nabla\Big(7\Big(\frac{1}{7}u_4\Big)\Big) =
-    -\frac{5}{128}\phi_{b}\:.
-\end{multline}
-By rearranging the system such that we have a single gradient operator
-in each equation, we again have a matrix system:
-\begin{equation}
-  \mathbf{n} \cdot D_n \nabla u_n + \sum_{m=1}^4 B_{nm} u_m =
-  s_n\ \ \ \ \ \ \ n = 1,2,3,4\:,
-  \label{eq:spn_bnd_matrix}
-\end{equation}
-where $\mathbf{D}$ and $\mathbf{u}$ are defined as before and:
-\begin{equation}
-  \mathbf{s} = \Big(\frac{1}{2}\phi_b\ -\frac{1}{8}\phi_b\
-  \frac{1}{16}\phi_b\ -\frac{5}{128}\phi_b \Big)^T\:,
-  \label{eq:spn_bnd_source}
-\end{equation}
-is the source vector on the boundary and
-%% I grabbed this matrix from Tom's tech note again.
-\begin{equation}
-  \mathbf{B} = \begin{bmatrix}
-    \frac{1}{2} &
-    -\frac{1}{8} &
-    \frac{1}{16} &
-    -\frac{5}{128} \\
-    %%
-    &&&\\
-    %%
-    -\frac{1}{8} &
-    \frac{7}{24} &
-    -\frac{41}{384} &
-    \frac{1}{16} \\
-    %%
-    &&&\\
-    %%
-    \frac{1}{16} &
-    -\frac{41}{384} &
-    \frac{407}{1920} &
-    -\frac{233}{2560} \\
-    %%
-    &&&\\
-    %%
-    -\frac{5}{128} &
-    \frac{1}{16} &
-    -\frac{233}{2560} &
-    \frac{3023}{17920}
-  \end{bmatrix}\:,
-  \label{eq:B_matrix}
-\end{equation}
-is a dense matrix of coefficients. Again, as pointed out by Evans, the
-$SP_1$ approximation reduces this boundary condition to the standard
-Marshak diffusion boundary condition. For reflecting boundary
-conditions, we use the same procedure as the $P_N$ equations where the
-odd-moments are zero such that Eq~(\ref{eq:reflecting_condition}) is
-true. From setting Eq~(\ref{eq:odd_moments}) to zero for odd $\phi_n$
-and again substituting Eq~(\ref{eq:spn7_subs}) we immediately find
-that:
-\begin{equation}
-  \nabla \mathbf{u} = 0
-  \label{eq:spn_reflecting}
-\end{equation}
-for reflecting $SP_N$ boundaries, providing enough equations to close
-the system.
-
-\subsection{Formulation of the Multigroup $SP_N$ Equations}
-\label{subsec:mg_spn_equations}
-Up to this point, we have formulated the $P_N$ and subsequently the
-$SP_N$ equations for a single neutron energy. To expand these
-equations for multiple energies, we start by stating the multigroup
-neutron transport equation for a single dimension in planar geometry:
-\begin{multline}
-  \mu \frac{\partial}{\partial x} \psi^g(x,\mu) + \sigma^g(x)
-  \psi^g(x,\mu) = \\ \sum_{g'=0}^{G} \int
-  \sigma_s^{gg'}(x,\hat{\Omega}' \rightarrow \hat{\Omega})
-  \psi^{g'}(x,\hat{\Omega}') d\Omega' + \frac{q^g(x)}{4 \pi}\:,
-  \label{eq:cart_1d_multigroup}
-\end{multline}
-where $g$ denotes the group index of $0$ to $G$ groups, $G=N_g-1$, and
-the integration of the scattering emission term of energy has been
-replaced by a discrete summation. For scattering, $\sigma_s^{gg'}$
-provides the probability of scattering at a particular angle from
-group $g$ to $g'$. The result is an equation nearly identical in form
-to Eq~(\ref{eq:cart_1d_transport}) where now instead of forming the
-$P_N$ and $SP_N$ equations for a single energy group, we form them for
-each of the energy groups with group coupling occuring through the
-scattering term. The multigroup $P_N$ equations are then:
-\begin{equation}
-   \frac{1}{2n+1} \frac{\partial}{\partial x}\Big[ (n+1) \phi^g_{n+1}
-     + n \phi^g_{n-1} \Big] +
-   \sum_{g'}(\sigma^g\delta_{gg'}-\sigma^{gg'}_{sn}) \phi^g_n =
-   q\delta_{n0} \:,
-  \label{eq:multigroup_pn_equations}
-\end{equation}
-for $n = 0,1,\dotsc,N$ where the flux and scattering moments are
-defined in a group. We observe that a $N_g \times N_g$ scattering
-matrix is formed:
-\begin{equation}
-  \mathbf{\Sigma}_n =
-  \sum_{g'}(\sigma^g\delta_{gg'}-\sigma^{gg'}_{sn})\:,
-  \label{eq:scattering_matrix}
-\end{equation}
-and when expanded gives:
-%% Also grabbed this matrix from Tom's tech note.
-\begin{equation}
-  \mathbf{\Sigma}_n =
-  \begin{bmatrix}
-    (\sigma^0-\sigma_{sn}^{00}) & -\sigma_{sn}^{01} & \dots &
-    -\sigma_{sn}^{0G} \\ &&&\\ -\sigma_{sn}^{10} &
-    (\sigma^1-\sigma_{sn}^{11}) & \dots & -\sigma_{sn}^{1G}
-    \\ &&&\\ \vdots & \vdots & \ddots & \vdots
-    \\ &&&\\ -\sigma_{sn}^{G0} & -\sigma_{sn}^{G1} & \dots &
-    (\sigma^G-\sigma_{sn}^{GG})
-  \end{bmatrix}\:.
-\end{equation}
-It is also useful to combine the group flux moments and sources into a
-single vector for more compact notation:
-\begin{equation}
-  \mathbf{\Phi_n} = (\phi^0_n\ \phi^1_n\ \cdots \phi^G_n )^T\:, 
-  \label{eq:group_flux_vector}
-\end{equation}
-\begin{equation}
-  \mathbf{q} = (q^0\ q^1\ \cdots q^G )^T\:.
-  \label{eq:group_source_vector}
-\end{equation}
-Next, we apply the $SP_N$ approximation to
-Eq~(\ref{eq:multigroup_pn_equations}) in identical fashion to the
-monoenergetic case. This gives:
-\begin{multline}
-  -\nabla \cdot \Bigg[\frac{n}{2n+1}\mathbf{\Sigma_{n-1}}^{-1} \nabla
-    \Big(\frac{n-1}{2n-1} \mathbf{\Phi_{n-2}} +
-    \frac{n}{2n-1}\mathbf{\Phi_n} \Big) \\+
-    \frac{n+1}{2n+1}\mathbf{\Sigma_{n+1}}^{-1} \nabla
-    \Big(\frac{n+1}{2n+3}\mathbf{\Phi_n} +
-    \frac{n+2}{2n+3}\mathbf{\Phi_{n+2}}\Big) \Bigg] \\+
-  \mathbf{\Sigma_n} \mathbf{\Phi_n} = \mathbf{q}
-  \delta_{n0}\ \ \ \ \ \ \ \ \ n = 0,2,4,\cdots,N\:.
-  \label{eq:multigroup_spn_equations}
-\end{multline}
-This adds more complexity than the monoenergetic formulation in that
-all unknowns in this group of equations are now vector quantities and
-scattering relationships are contained in matrices rather than a
-scalar quantity. Because of this, the same sequence of variable
-changes and algebra can be used to build a set of matrix equations,
-this time in a block formulation:
-\begin{equation}
-  -\nabla \cdot \mathbb{D}_n \nabla \mathbb{U}_n + \sum_{m=1}^4
-  \mathbb{A}_{nm} \mathbb{U}_m = \mathbb{Q}_n\:,
-  \label{eq:spn_multigroup_system}
-\end{equation}
-where the definition of all quantities are the same with internal
-scalar values replaced by the group-vector values. In addition,
-$\mathbb{A}$ is now a block matrix of $N_g \times N_g$ submatrices
-generated from the moment scattering matrices:
-\begin{equation}
-  \mathbf{A} = 
-  \begin{bmatrix}
-    (\mathbf{\Sigma}_0) &
-    (-\frac{2}{3}\mathbf{\Sigma}_0) &
-    (\frac{8}{15}\mathbf{\Sigma}_0) &
-    (-\frac{16}{35}\mathbf{\Sigma}_0) \\
-    %%
-    &&&\\
-    %%
-    (-\frac{2}{3}\mathbf{\Sigma}_0) &
-    (\frac{4}{15}\mathbf{\Sigma}_0 + \frac{1}{3}\mathbf{\Sigma}_2) &
-    (-\frac{16}{45}\mathbf{\Sigma}_0 - \frac{4}{9}\mathbf{\Sigma}_2) &
-    (\frac{32}{105}\mathbf{\Sigma}_0 + \frac{8}{21}\mathbf{\Sigma}_2) \\
-    %%
-    &&&\\
-    %%
-    (\frac{8}{15}\mathbf{\Sigma}_0) &
-    (-\frac{16}{45}\mathbf{\Sigma}_0 - \frac{4}{9}\mathbf{\Sigma}_2) &
-    (\frac{64}{225}\mathbf{\Sigma}_0 + \frac{16}{45}\mathbf{\Sigma}_2 + \frac{9}{25}\mathbf{\Sigma}_4) &
-    (-\frac{128}{525}\mathbf{\Sigma}_0 - \frac{32}{105}\mathbf{\Sigma}_2 - \frac{54}{175}\mathbf{\Sigma}_4)
-    \\ 
-    %%
-    &&&\\
-    %%
-    (-\frac{16}{35}\mathbf{\Sigma}_0) &
-    (\frac{32}{105}\mathbf{\Sigma}_0 + \frac{8}{21}\mathbf{\Sigma}_2) &
-    (-\frac{128}{525}\mathbf{\Sigma}_0 - \frac{32}{105}\mathbf{\Sigma}_2 - \frac{54}{175}\mathbf{\Sigma}_4)
-    & 
-    (\frac{256}{1225}\mathbf{\Sigma}_0 + \frac{64}{245}\mathbf{\Sigma}_2 +
-    \frac{324}{1225}\mathbf{\Sigma}_4 + \frac{13}{49}\mathbf{\Sigma}_6)
-  \end{bmatrix}\:.
-  \label{eq:A_block_matrix}
-\end{equation}
-Analogously, for Marshak conditions on the boundaries we have:
-\begin{equation}
-  \hat{\mathbf{n}} \cdot \mathbb{D}_n \nabla \mathbb{U}_n +
-  \sum_{m=1}^4 \mathbb{B}_{nm} \mathbb{U}_m = \mathbb{S}_n\:,
-  \label{eq:spn_multigroup_bnd}
-\end{equation}
-with $\mathbb{S}_n$ the vector of group-wise boundary source term on
-each boundary for each psuedo-moment and $\mathbb{B}_{nm}$ is an $N_g
-\times N_g$ diagonal matrix with the value $B_{nm}$ on the
-diagonal. For reflecting conditions, again we have $\nabla
-\mathbb{U}_n = 0$ for all pseudo-moments.
-
-\subsection{A Note on Spatial Discretization and Matrix Symmetry}
-\label{subsec:spatial_discretization}
-To this point, the formulation of the multigroup $SP_N$ equations
-presented have discretized the transport equation in angle and energy
-but have yet to consider the spatial component of phase space. We
-don't explicitly consider it in this document but instead we will
-briefly comment on possible means of discretization. Of primary
-interest here is the discretization of the diffusive term in
-Eq~(\ref{eq:spn_multigroup_system}) and the convective term on the
-boundaries in Eq~(\ref{eq:spn_multigroup_bnd}). Many popular choices
-for spatial discretization are available here and include finite
-difference and finite element formulations. In the Denovo package at
-ORNL, Evans has implemented a finite volume scheme for these
-equations. Although arbitrary grids could be handled effectively
-through a finite element scheme, the rectilinear grid used in Denovo
-is easily discretized through the finite volume method in a
-conservative form. This conservative form is ideal for the
-diffusion-like equations in the domain given by
-Eq~(\ref{eq:spn_multigroup_system}) and the effective boundary current
-conditions given by Eq~(\ref{eq:spn_multigroup_bnd}) as the neutron
-current is balanced from cell-to-cell, continuity of the flux is
-preserved across cell/material boundaries, and boundary conditions are
-naturally represented through cell-face currents. Furthermore, this
-spatial discretization is a natural extension of the balance
-principles used to arrive at the general transport equation. The
-result of this finite volume discretization is one that is symmetric
-for all equations in the domain with respect to the spatial components
-of the solution.
-
-Given the symmetry of the spatial discretization equations in the
-domain given by monoenergetic Eq~(\ref{eq:spn_matrix}) form a
-symmetric linear operator acting on the pseudo-moment vector to give:
-\begin{equation}
-  \mathbf{L}\mathbf{u}=\mathbf{Q}\:,
-  \label{eq:matrix_system}
-\end{equation}
-where $\mathbf{L} = -\nabla \cdot \mathbf{D} \nabla +
-\mathbf{A}$. Moving to the multigroup formulation in
-Eq~(\ref{eq:spn_multigroup_system}) we have:
-\begin{equation}
-  \mathbb{L}\mathbb{U}=\mathbb{Q}\:,
-  \label{eq:multigroup_matrix_system}
-\end{equation}
-where $\mathbb{L} = -\nabla \cdot \mathbb{D} \nabla +
-\mathbb{A}$. Whether or not the resulting matrix is symmetric is
-determined by the group scattering matrices, $\sigma_{sn}^{gg'}$, for
-each of the moments. In general, the cross sections in these matrices
-do not form a symmetric matrix and therefore the resulting linear
-operator in Eq~(\ref{eq:multigroup_matrix_system}) will always be
-asymmetric for multigroup problems. Solutions to this linear system
-will then require techniques specifically formulated for asymmetric
-systems.
+In this chapter we apply a preconditioned MCSA method as a solution
+scheme for the $SP_N$ equations.
 
@@
 \subsection{Spectral Analysis Results}
-\label{subsec:analysis_results}
+\label{subsec:spn_analysis_results}
+
+\subsection{Preconditioning Strategy}
+\label{subsec:spn_preconditioning}
 

### 2013-03-25 doc/dissertation/document/chapters/mc_theory/mc_theory.tex
updating sections in dissertation

@@
-\chapter{An Analytic Framework for Parallel Monte Carlo Methods}
+\chapter{An Analytic Performance Framework for Domain-Decomposed Monte Carlo}
 \label{ch:analytic_framework}
@@
 %%---------------------------------------------------------------------------%%
-\section{Domain-to-Domain Communication for Symmetric Systems}
-\label{sec:mc_comm_costs_symmetric}
+\section{Leakage Fractions for Symmetric Systems}
+\label{sec:symmetric_leakage_fractions}
 With Wagner's work, we observed that domain overlap implemented with a
@@
 %%---------------------------------------------------------------------------%%
-\section{Domain-to-Domain Communication for Asymmetric Systems}
-\label{sec:mc_comm_costs_asymmetric}
+\section{Communication Costs for Symmetric Systems}
+\label{sec:Symmetric_comm_costs}
+Use Seigel's work to get communication costs.
+
+%%---------------------------------------------------------------------------%%
+\section{Leakage Fractions for Asymmetric Systems}
+\label{sec:asymmetric_leakage_fractions}
 Extend the above results to asymmetric systems.
+
+%%---------------------------------------------------------------------------%%
+\section{Communication Costs for Asymmetric Systems}
+\label{sec:asymmetric_comm_costs}
+Use Seigel's work to get communication costs.

### 2013-03-25 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
updating sections in dissertation

@@
 
-%%---------------------------------------------------------------------------%%
-\section{Derivation of the $SP_N$ Equations}
-\label{sec:derivation}
-In this section, we derive the $SP_N$ equations, closely following the
+In this chapter, we derive the $SP_N$ equations, closely following the
 work of Evans. We begin by stating the general time-independent
@@
 
-\subsection{The Neutron Transport Equation}
-\label{subsec:transport_eq}
+%%---------------------------------------------------------------------------%%
+\section{The Neutron Transport Equation}
+\label{sec:transport_eq}
 As a starting point we define the time-independent neutron transport
@@
 
-\subsection{Derivation of the $P_N$ Equations}
-\label{subsec:pn_equations}
+\section{Derivation of the $P_N$ Equations}
+\label{sec:pn_equations}
 Next, we derive the $P_N$ equations, a simplified form of the general
@@
 
-\subsubsection{Legendre Polynomials}
-\label{subsubsec:legendre_polys}
+\subsection{Legendre Polynomials}
+\label{subsec:legendre_polys}
 The Legendre polynomials are an orthogonal set of functions that
@@
 
-\subsubsection{Planar $P_N$ Equations}
-\label{subsubsec:planar_pn}
+\subsection{Planar $P_N$ Equations}
+\label{subsec:planar_pn}
 With the Legendre polynomial properties defined above, we can proceed
@@
 
-\subsubsection{Boundary Conditions for the $P_N$ Equations}
+\subsection{Boundary Conditions for the $P_N$ Equations}
 \label{subsbusec:bcs_pn}
@@
 
-\subsection{Formulation of the $SP_N$ Equations}
-\label{subsec:spn_equations}
+\section{Derivation of the $SP_N$ Equations}
+\label{sec:spn_equations}
 The $P_N$ equations give $N+1$ coupled first-order equations
@@
 
-\subsubsection{Boundary Conditions for the $SP_N$ Equations}
-\label{subsubsec:spn_boundary_conditions}
+\subsection{Boundary Conditions for the $SP_N$ Equations}
+\label{subsec:spn_boundary_conditions}
 We approach the formulation of the boundary conditions in much the
@@
 
-\subsection{Formulation of the Multigroup $SP_N$ Equations}
-\label{subsec:mg_spn_equations}
+\section{Derivation of the Multigroup $SP_N$ Equations}
+\label{sec:mg_spn_equations}
 Up to this point, we have formulated the $P_N$ and subsequently the
@@
 
-\subsection{A Note on Spatial Discretization and Matrix Symmetry}
-\label{subsec:spatial_discretization}
+\section{A Note on Spatial Discretization and Matrix Symmetry}
+\label{sec:spatial_discretization}
 To this point, the formulation of the multigroup $SP_N$ equations

### 2013-03-25 doc/dissertation/document/chapters/spn_solutions/spn_solutions.tex
updating sections in dissertation

@@
 \section{Spectral Analysis of the $SP_N$ Equations}
-\label{sec:spectral_analysis}
+\label{sec:spn_spectral_analysis}
 Now that we have an understanding of the linear system generated by
@@
 %%---------------------------------------------------------------------------%%
-\section{Monte Carlo Synthetic Acceleration for the $SP_N$ Equations}
-\label{sec:monte_carlo}
+\section{Fuel Assembly Scaling Results for the $SP_N$ Equations}
+\label{sec:spn_assembly_scaling}
+
+%%---------------------------------------------------------------------------%%
+\section{Full Core Scaling Results for the $SP_N$ Equations}
+\label{sec:spn_assembly_scaling}

### 2013-03-25 doc/dissertation/document/chapters/mc_background/mc_background.tex
updating disseration chapter structure
updating sections in dissertation
fixing typo in linear problem chapter and removing resuing random walks subsection in mc background chapter

@@
 our expectation values for the solution may have a finite variance.
-
-\subsection{Reusing Random Walks}
-\label{subsec:reusing_random_walks}
-A downfall of the direct method is that a random walk must be
-initiated in each state of the solution vector in which the solution
-is desired due to the estimator definition in
-Eq~(\ref{eq:direct_expectation_expansion}). Because of this, a
-solution estimate in all states of a linear system may require a
-large number of random walks to be performed. Ji and Li recently
-proposed a variation of the direct method in which the random walk
-permutations in Eq~(\ref{eq:mc_walk_permutation}) can be used to
-contribute to the solution tally in all visited states instead of the
-only starting state \citep{ji_reusing_2012}. By doing this, they
-effectively reduce the number of random walks needed to sample all
-desired states in the system.
-
-Their method is based on the observation that while a random walk
-transitions between states in the system, if every state it visited
-could be considered that starting state, then a series a contributions
-to the solution could be generated by a single sequence of random
-numbers while maintaining statistical independence between the
-samples. To demonstrate this, consider the following random walk
-sequence starting in state 2: $2 \rightarrow 3 \rightarrow 1
-\rightarrow 2 \rightarrow 4 \rightarrow 5$. We can reuse these walks
-be recognizing that we also then have a random walk sequence starting
-in state 3 that can be applied to the state 3 estimator: $3
-\rightarrow 1 \rightarrow 2 \rightarrow 4 \rightarrow 5$. We also have
-the same situation for state 1: $1 \rightarrow 2 \rightarrow 4
-\rightarrow 5$, and all future states. Using this sampling scheme , Ji
-and Li observed 1 to 2 orders of magnitude reduction over the standard
-direct method in the number histories required to achieve a specified
-variance in their test problems.
 
@@
 Monte Carlo calculation.
+
 \begin{figure}[ht!]
@@
 %%---------------------------------------------------------------------------%%
-\section{Comparison to Sequential Monte Carlo}
+\section{MCSA Comparison to Sequential Monte Carlo}
 \label{subsec:sequential_comparison}

### 2013-03-25 doc/dissertation/document/chapters/mc_background/mc_background.tex
Updating adjoint constraint defintions in mc background

@@
 achieve this by using Eq~(\ref{eq:adjoint_vector_relation}) as a
-constraint. Here we have 2 unknowns, $\ve{y}$ and $\ve{d}$ and
-therefore we require a second constraint to close the system. As a
+constraint. Here we have 2 unknowns, $\ve{y}$ and $\ve{d}$, and therefore we
+require two constraints to close the system. We use
+Eq~(\ref{eq:adjoint_vector_relation}) as the first constraint and as a
 second constraint we select:
 \begin{equation}
-  \ve{d} = \boldsymbol{\delta}_i\:,
+  \ve{d} = \boldsymbol{\delta}_j\:,
   \label{eq:adjoint_second_constraint}
 \end{equation}
-where the $k^{th}$ component of the vector $\boldsymbol{\delta}_i$ is
-the Dirac delta function $\delta_{i_k,i}$. If we apply
-Eq~(\ref{eq:adjoint_second_constraint}) to our first constraint
-Eq~(\ref{eq:adjoint_vector_relation}), we get the following convenient
-outcome:
+where $\boldsymbol{\delta}_j$ is one of a set of vectors in which the
+$j^{th}$ component is the Kronecker delta function $\delta_{i,j}$. If
+we apply Eq~(\ref{eq:adjoint_second_constraint}) to our first
+constraint Eq~(\ref{eq:adjoint_vector_relation}), we get the following
+convenient outcome:
 \begin{equation}
   \langle \ve{y}, \ve{b} \rangle = \langle \ve{x},
-  \boldsymbol{\delta}_i \rangle = x_i \:,
+  \boldsymbol{\delta}_j \rangle = x_j \:,
   \label{eq:inner_product_constraint}
 \end{equation}
-meaning that if we compute the inner product of the direct source and
-the adjoint solution using a delta function source, we recover the
-direct solution we are looking for
+meaning that if we compute the inner product of the original source
+and the adjoint solution using a delta function source, we recover one
+component of the original solution.
 

### 2013-03-25 doc/dissertation/document/chapters/mc_background/mc_background.tex
adding evolution of a solution for the direct method in background

@@
 W_c$ for a particular random walk permutation.
+
+\subsection{Direct Method: Evolution of a Solution}
+\label{subsec:direct_evolution}
+
+As a means of visually demonstrating the direct Monte Carlo method,
+consider a 2-dimensional thermal diffusion problem with sources on the
+left and right hand sides of the domain and a smaller uniform source
+as shown in Figure~\ref{fig:heat_setup}.
+\begin{figure}[h!]
+  \begin{center}
+    \scalebox{1.0}{ \input{heat_eq_setup.pdftex_t} }
+  \end{center}
+  \caption{\textbf{Problem setup for 2D heat equation.}
+    \textit{Dirichlet conditions are set for the temperature on all 4
+      boundaries of the Cartesian grid. Background source of 1.0
+      present. $50 \times 50$ grid.}}
+  \label{fig:heat_setup}
+\end{figure}
+For this problem, the number of histories used to compute the solution
+at each state in the domain was increased from 1 to 1000 in order to
+show its effects on the solution and the statistical nature of the
+method. Figure~\ref{fig:direct_evolution} gives these results.
+\begin{figure}[h!]
+  \begin{center}
+    \includegraphics[width=4in]{direct_evolution.png}
+  \end{center}
+  \caption{\textbf{Direct Monte Carlo solution to the heat equation
+      with varying numbers of histories.} \textit{Top left: 1 history
+      per state. Top right: 10 histories per state. Bottom left: 100
+      histories per state. Bottom right: 1000 histories per state.}}
+  \label{fig:direct_evolution}
+\end{figure}
+As the number of histories used per state is increased, the
+statistical variance of the solutions is decreased as more tallies are
+made. At 1000 histories per state, enough tallies have been made to
+generate a reasonable estimate for the structure of the solution.
 
@@
 contributions become increasingly small.
+
+\subsection{Adjoint Method: Evolution of a Solution}
+\label{subsec:adjoint_evolution}
 

### 2013-03-25 doc/dissertation/document/chapters/mc_background/mc_background.tex
adding adjoint evolution of a solution to mc background

@@
   \begin{center}
-    \scalebox{1.0}{ \input{heat_eq_setup.pdftex_t} }
+    \scalebox{1.0}{ \input{chapters/mc_background/heat_eq_setup.pdftex_t} }
   \end{center}
@@
     \textit{Dirichlet conditions are set for the temperature on all 4
-      boundaries of the Cartesian grid. Background source of 1.0
-      present. $50 \times 50$ grid.}}
+      boundaries of the Cartesian grid. Background source of 1/5 the
+      value of the boundary sources present. $50 \times 50$ grid.}}
   \label{fig:heat_setup}
@@
   \begin{center}
-    \includegraphics[width=4in]{direct_evolution.png}
+    \includegraphics[width=5in]{chapters/mc_background/direct_evolution.png}
   \end{center}
@@
 \label{subsec:adjoint_evolution}
+
+As a means of visually demonstrating the adjoint Monte Carlo method,
+again considerthe a 2-dimensional thermal diffusion problem with
+sources on the left and right hand sides of the domain and a smaller
+uniform source as shown in Figure~\ref{fig:heat_setup}. For the
+adjoint method, the number of histories sampled from the source was
+increased from 10 to 10,000,000 in order to show its effects on the
+solution and the statistical nature of the
+method. Figure~\ref{fig:adjoint_evolution} gives these results.
+\begin{figure}[h!]
+  \begin{center}
+    \includegraphics[width=5in]{chapters/mc_background/adjoint_evolution.png}
+  \end{center}
+  \caption{\textbf{Adjoint Monte Carlo solution to the heat equation
+      with varying numbers of histories.} \textit{Top left: 10
+      histories per state. Top right: 1,000 histories per
+      state. Bottom left: 100,000 histories per state. Bottom right:
+      10,000,000 histories per state.}}
+  \label{fig:adjoint_evolution}
+\end{figure}
+As the number of histories used per state is increased, the
+statistical variance of the solutions is decreased as more tallies are
+made. At 10,000,000 histories per state, enough tallies have been made
+to generate a reasonable estimate for the structure of the
+solution. The visual difference between
+Figures~\ref{fig:direct_evolution} and \ref{fig:adjoint_evolution} is
+precisely that determined by their mathematics. As the adjoint
+solution evolves with the addition of histories, more histories
+emanate from the source with more penetrating the domain and making
+contributions to those tallies. As the direct method evolves, the
+solution simply comes into focus as histories are emanated from each
+state in the system and contribute only to the solution in that state
+regardless of their path.
 

### 2013-03-25 doc/dissertation/document/chapters/mc_background/mc_background.tex
updating MCSA equations and point jacobi preconditioning in mc background

@@
   \begin{gather}
-    \ve{x}^{k+1/2} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,\\
+    \ve{x}^{k+1/2} = \ve{x}^k + \ve{r}^k\:,\\
     \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\
@@
 requirements for convergence, they do require that the spectral radius
-of the iteration matrix be less than one. To achieve this, a Jacobi
-preconditioner, a form of left preconditioning, is used such that the
-preconditioning matrix $\ve{M}$ is:
+of the iteration matrix be less than one. To achieve this for
+diagonally dominant matrices, a Jacobi preconditioner, a form of left
+preconditioning, is used such that the preconditioning matrix $\ve{M}$
+is:
 \begin{equation}
@@
   \begin{gather}
-    \ve{x}^{k+1/2} = (\ve{I} - \ve{M}^{-1}\ve{A})\ve{x}^k +
-    \ve{M}^{-1}\ve{b}\:,\\ \ve{r}^{k+1/2} = \ve{M}^{-1}\ve{b} -
-    \ve{M}^{-1}\ve{A}\ve{x}^{k+1/2}\:,\\ \ve{M}^{-1}\ve{A}\delta\ve{x}^{k+1/2}
-    = \ve{r}^{k+1/2}\:,\\ \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta
-    \ve{x}^{k+1/2}\:.
+    \ve{x}^{k+1/2} = \ve{x}^k +
+    \ve{M}^{-1}\ve{r}^k\:,\\ \ve{r}^{k+1/2} =
+    \ve{b}-\ve{A}\ve{x}^{k+1/2}\:,\\ \ve{M}^{-1}\ve{A}\delta\ve{x}^{k+1/2}
+    = \ve{M}^{-1}\ve{r}^{k+1/2}\:,\\ \ve{x}^{k+1} = \ve{x}^{k+1/2} +
+    \delta \ve{x}^{k+1/2}\:.
     \label{eq:jacobi_preconditioned_mcsa}
@@
 reasons. First, $\rho(\ve{I} - \ve{M}^{-1}\ve{A}) < 1$ is true for all
-$\ve{A}$ and is easy to formulate because the inversion of $\ve{M}$ is
-trivial. Second, because the adjoint Monte Carlo method used within
-MCSA to compute the correction operates on a linear problem with the
-preconditioned operator, then $\ve{H}$ in the adjoint solver will have
-a zero term in each of its diagonal elements, thereby eliminating all
-in-state transitions during the random walk sequence. Because of this,
-Jacobi preconditioning should always be performed, regardless of any
-other preconditioning that is applied to the system.
+$\ve{A}$ that is diagonally dominant and is easy to formulate because
+the inversion of $\ve{M}$ is trivial. Second, because the adjoint
+Monte Carlo method used within MCSA to compute the correction operates
+on a linear problem with the preconditioned operator, then $\ve{H}$ in
+the adjoint solver will have a zero term in each of its diagonal
+elements, thereby eliminating all in-state transitions during the
+random walk sequence. Because of this, Jacobi preconditioning should
+always be performed, regardless of any other preconditioning that is
+applied to the system.
 

### 2013-03-25 doc/dissertation/document/chapters/mc_background/mc_background.tex
changed jacobi to point jacobi in mc background

@@
 of the iteration matrix be less than one. To achieve this for
-diagonally dominant matrices, a Jacobi preconditioner, a form of left
+diagonally dominant matrices, a point Jacobi preconditioner, a form of left
 preconditioning, is used such that the preconditioning matrix $\ve{M}$
@@
 \end{subequations}
-Choosing Jacobi preconditioning with MCSA is advantageous for several
-reasons. First, $\rho(\ve{I} - \ve{M}^{-1}\ve{A}) < 1$ is true for all
-$\ve{A}$ that is diagonally dominant and is easy to formulate because
-the inversion of $\ve{M}$ is trivial. Second, because the adjoint
-Monte Carlo method used within MCSA to compute the correction operates
-on a linear problem with the preconditioned operator, then $\ve{H}$ in
-the adjoint solver will have a zero term in each of its diagonal
-elements, thereby eliminating all in-state transitions during the
-random walk sequence. Because of this, Jacobi preconditioning should
-always be performed, regardless of any other preconditioning that is
-applied to the system.
+Choosing point Jacobi preconditioning with MCSA is advantageous for
+several reasons. First, $\rho(\ve{I} - \ve{M}^{-1}\ve{A}) < 1$ is true
+for all $\ve{A}$ that is diagonally dominant and is easy to formulate
+because the inversion of $\ve{M}$ is trivial. Second, because the
+adjoint Monte Carlo method used within MCSA to compute the correction
+operates on a linear problem with the preconditioned operator, then
+$\ve{H}$ in the adjoint solver will have a zero term in each of its
+diagonal elements, thereby eliminating all in-state transitions during
+the random walk sequence. Because of this, point Jacobi
+preconditioning should almost always be performed, regardless of any
+other preconditioning that is applied to the system.
 

### 2013-03-25 doc/dissertation/document/chapters/mc_background/mc_background.tex
updating section headers in mc background
getting rid of the crap default math font in the thesis template

@@
 %%---------------------------------------------------------------------------%%
-\section{Direct Method}
+\section{Direct Monte Carlo Method}
 \label{sec:direct_mc}
@@
 \label{subsec:direct_evolution}
-
 As a means of visually demonstrating the direct Monte Carlo method,
 consider a 2-dimensional thermal diffusion problem with sources on the
-left and right hand sides of the domain and a smaller uniform source
-as shown in Figure~\ref{fig:heat_setup}.
+left and right hand sides of the domain and a uniform source
+throughout the domain of 1/5 the strength of the boundary sources as
+shown in Figure~\ref{fig:heat_setup}.
 \begin{figure}[h!]
@@
 For this problem, the number of histories used to compute the solution
-at each state in the domain was increased from 1 to 1000 in order to
-show its effects on the solution and the statistical nature of the
-method. Figure~\ref{fig:direct_evolution} gives these results.
+at each grid point (state) in the domain was increased from 1 to 1000
+in order to demonstrate the effects on the solution and the
+statistical nature of the method. Figure~\ref{fig:direct_evolution}
+gives these results.
 \begin{figure}[h!]
@@
 statistical variance of the solutions is decreased as more tallies are
-made. At 1000 histories per state, enough tallies have been made to
-generate a reasonable estimate for the structure of the solution.
+made. Starting with single history at each state in the
+domain, the high variance prevents a precise solution from being
+obtained although we begin to see the solution take shape as
+expected. At 1000 histories per state, enough tallies have been made
+to generate a reasonable estimate for the structure of the
+solution. It is interesting to note here that as the statistical
+uncertainty is reduced at each grid point in the domain, the solution
+is resolved in a certain sense, analagous to the convergence of a
+traditional iterative method.
 
@@
 %%---------------------------------------------------------------------------%%
-\section{Adjoint Method}
+\section{Adjoint Monte Carlo Method}
 \label{sec:adjoint_mc}
@@
 where $\ve{y}$ and $\ve{d}$ are the adjoint solution and source
-respectively and $\ve{A}^T$ is the adjoint operator. We can split this
-equation to mirror Eq~(\ref{eq:richardson_split}) by defining the
-following inner product equivalence \citep{spanier_monte_1969}:
+respectively and $\ve{A}^T$ is the adjoint operator for $\ve{A} \in
+\mathbb{R}^{N \times N}$. We can split this equation to mirror
+Eq~(\ref{eq:richardson_split}) by defining the following inner product
+equivalence \citep{spanier_monte_1969}:
 \begin{equation}
@@
 We can readily build an estimator for the adjoint solution from this
-series expansion, but we instead desire the direct solution. We
-achieve this by using Eq~(\ref{eq:adjoint_vector_relation}) as a
-constraint. Here we have 2 unknowns, $\ve{y}$ and $\ve{d}$, and therefore we
-require two constraints to close the system. We use
-Eq~(\ref{eq:adjoint_vector_relation}) as the first constraint and as a
-second constraint we select:
+series expansion, but we instead desire the solution to
+Eq~(\ref{eq:linear_problem}). Here we have 2 unknowns, $\ve{y}$ and
+$\ve{d}$, and therefore we require two constraints to close the
+system. We use Eq~(\ref{eq:adjoint_vector_relation}) as the first
+constraint and as a second constraint we select:
 \begin{equation}
@@
 %%---------------------------------------------------------------------------%%
-\section{Monte Carlo Synthetic-Acceleration}
+\section{Monte Carlo Synthetic Acceleration}
 \label{sec:mcsa}

### 2013-03-25 doc/dissertation/document/chapters/mc_background/mc_background.tex
updated mcsa and preconditioning section in mc background

@@
 therefore also providing an unbiased Monte Carlo estimate of the
-solution.
+solution. It should be noted here that
+Eq~(\ref{eq:adjoint_expectation_expansion}) only computes a single
+component of our desired solution vector when really what we desire is
+the entire solution vector. In an adjoint Monte Carlo simulation using
+this estimator, the $w_{i,j}$ elements that are added into the tally
+for each state are only selected if the random walk currently resides
+in that state. Much like a mesh tally in a particle transport
+simulation, we have $N$ simultaneous tallies for $\ve{A} \in
+\mathbb{R}^{N \times N}$ that will yield the entire solution
+vector. Based on their current state in the system, the random walks
+will contribute tally $j$ in this group.
 
@@
 \label{subsec:adjoint_evolution}
-
 As a means of visually demonstrating the adjoint Monte Carlo method,
@@
 emanate from the source with more penetrating the domain and making
-contributions to those tallies. As the direct method evolves, the
-solution simply comes into focus as histories are emanated from each
-state in the system and contribute only to the solution in that state
-regardless of their path.
+contributions to those tallies.
 
@@
 Monte Carlo, utilizes the adjoint Monte Carlo solver as a means of
-directly reducing the residual vector. He proposed the following
-iterative scheme as a solution to Eq~(\ref{eq:linear_problem})\:
+directly reducing the elements residual vector. He proposed the
+following iterative scheme as a solution to
+Eq~(\ref{eq:linear_problem})\:
 \begin{subequations}
@@
 exponential converge as the Monte Carlo error (and therefore the
-Central Limit Theorem) does dominate in this situation. Therefore, if
-a good initial guess is not available, Halton's method is still
-characterized by poor performance.
+Central Limit Theorem) does dominate in this situation.
 
@@
     \ve{A}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:,\\
-    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,
+    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,\\
+    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:,
   \end{gather}
@@
 where the adjoint Monte Carlo method is used to generate the solution
-correction from the residual. Using Monte Carlo in this way achieves
-the same effect as Halton's method, decoupling its convergence rate
-from the overall convergence rate of the method. Here, the approximate
-Monte Carlo solution is not driven to a particular convergence as it
-merely supplies a correction for the initial guess generated by
-Richardson's iteration. Rather, only a set number of histories are
-required using the adjoint method to generate the correction.
+correction from the residual and Richardson's iteration in the first
+step has been rewritten as a residual correction. Using Monte Carlo in
+this way achieves the same effect as Halton's method, decoupling its
+convergence rate from the overall convergence rate of the
+method. Here, the approximate Monte Carlo solution is not driven to a
+particular convergence as it merely supplies a correction for the
+initial guess generated by Richardson's iteration. Rather, only a set
+number of histories are required using the adjoint method to generate
+the correction.
 
@@
 requirements for convergence, they do require that the spectral radius
-of the iteration matrix be less than one. To achieve this for
-diagonally dominant matrices, a point Jacobi preconditioner, a form of left
-preconditioning, is used such that the preconditioning matrix $\ve{M}$
-is:
+of the iteration matrix be less than one. As an example, to achieve
+this for diagonally dominant matrices point Jacobi left
+preconditioning can be used such that the preconditioning matrix
+$\ve{M}$ is:
 \begin{equation}
-  \ve{M} = diag(\ve{A})\:,
+  \ve{M} = diag(\ve{A})\:.
   \label{eq:jacobi_preconditioner}
 \end{equation}
-such that its application means we are instead solving the following
-linear system:
+With the application of this preconditioner we are instead solving the
+following linear system:
 \begin{equation}
@@
     = \ve{M}^{-1}\ve{r}^{k+1/2}\:,\\ \ve{x}^{k+1} = \ve{x}^{k+1/2} +
-    \delta \ve{x}^{k+1/2}\:.
+    \delta \ve{x}^{k+1/2}\:,\\
+    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:,
     \label{eq:jacobi_preconditioned_mcsa}
@@
 \end{subequations}
+where the adjoint Monte Carlo solve now has a preconditioned operator
+from which to build weights and probabilities for transport and a
+preconditioned source vector to sample.
+
 Choosing point Jacobi preconditioning with MCSA is advantageous for
@@
 the random walk sequence. Because of this, point Jacobi
-preconditioning should almost always be performed, regardless of any
-other preconditioning that is applied to the system.
+preconditioning should be considered for many classes of problems,
+regardless of any other preconditioning that is applied to the system.
 

### 2013-03-25 doc/dissertation/document/chapters/mc_background/mc_background.tex
some cleanup in method comparisions in mc background

@@
 The MCSA method defined in Eq.~(\ref{eq:mcsa}) uses the adjoint method
-to estimate the error in residual Monte Carlo solve instead of the
-direct method outlined in \S~\ref{sec:direct_mc}. To demonstrate
-the effectiveness of the adjoint method over the direct method within
-the context of MCSA, we choose the 2D time-dependent Poisson equation
-as a simple model problem:
+to estimate the error in a residual Monte Carlo solve instead of the
+direct method outlined in \S\ref{sec:direct_mc}. To demonstrate the
+effectiveness of the adjoint method over the direct method within the
+context of MCSA, we choose the two-dimensional time-dependent Poisson
+equation as a simple model problem:
 \begin{equation}
@@
 to converge to a solution, a problem of constant $\Delta$ was used
-with varying values of grid size, fixing the spectral radius of the
-system at a constant value for each variation. Both the five-point and
-nine-point stencils were used with both the direct and adjoint
-solvers. For each case, $N \times N$ total random walk permutations
-were computed per MCSA iteration where $N \times N$ is the number of
-discrete grid points in the system. Solver parameters were set to a
-weight cutoff of \sn{1}{-4} for the stochastic linear solver and a
-convergence tolerance of \sn{1}{-8} for the MCSA iterative solver.
-Figure~\ref{fig:poisson_cpu_time} gives the CPU time needed for each
-case to converge in seconds and Figure~\ref{fig:poisson_iterations}
-gives the number of iterations needed for each case to converge to the
-specified tolerance as a function of the problem size. All
-computations presented in this section and
-\S~\ref{subsec:sequential_comparison} were completed on a 3.0 GHz
-Intel Core 2 Quad Q9650 CPU machine with 16 GB 1067 MHz DDR3 memory.
+with varying values of the number of mesh elements, fixing the
+spectral radius of the system at a constant value for each
+variation. Both the five-point and nine-point stencils were used with
+both the direct and adjoint solvers. For each case, $N \times N$ total
+random walk permutations were computed per MCSA iteration where $N
+\times N$ is the number of discrete grid points in the system. Solver
+parameters were set to a weight cutoff of \sn{1}{-4} for the
+stochastic linear solver and a convergence tolerance of \sn{1}{-8} for
+the MCSA iterative solver.  Figure~\ref{fig:poisson_cpu_time} gives
+the CPU time needed for each case to converge in seconds and
+Figure~\ref{fig:poisson_iterations} gives the number of iterations
+needed for each case to converge to the specified tolerance as a
+function of the problem size. All computations presented in this
+section and \S\ref{subsec:sequential_comparison} were completed on a
+3.0 GHz Intel Core 2 Quad Q9650 CPU machine with 16 GB 1067 MHz DDR3
+memory.
 \begin{figure}[ht!]
@@
 compare its performance to Halton's Sequential Monte Carlo method on
-which our previous work in this area was based. For this comparison,
-we use the same transient Poisson problem as described in the previous
+which previous work in this area was based. For this comparison, we
+use the same transient Poisson problem as described in the previous
 section and choose only the 5-point stencil to discretize the
@@
 to $2 \times N \times N$ in order to ensure convergence in Sequential
-Monte Carlo Method. Figure~\ref{fig:seq_cpu_time} gives the CPU time
-results for this comparison as a function of problem size while
+Monte Carlo Method. For the majority of the problems in the previous
+section, the Sequential method used with $N \times N$ histories would
+not converge. Figure~\ref{fig:seq_cpu_time} gives the CPU time results
+for this comparison as a function of problem size while
 Figure~\ref{fig:seq_iterations} gives the number of iterations to

### 2013-03-25 doc/dissertation/document/chapters/mc_theory/mc_theory.tex
adding additional pngs to symmetric spectral analysis for communication

@@
 \label{ch:analytic_framework}
-
+With Wagner's work \citep{wagner_hybrid_2010}, we observed that domain
+overlap implemented with a domain decomposed Monte Carlo setting
+reduces the amount of nearest neighbor communication that must occur
+during transport. To date, parallel Neumann-Ulam methods have been
+limited to full domain replication with parallelism exploited through
+individual histories \citep{alexandrov_efficient_1998} and in the
+previous chapter we exploited Wagner's work to alleviate this. To
+accomplish this, we recognized from the literature that stochastic
+histories must be transported from domain to domain as the simulation
+progresses and they transition to states that are not in the local
+domain. Because we have chosen a domain decomposition strategy in a
+parallel environment, this means that communication of these histories
+must occur between compute nodes owning neighboring pieces of the
+global domain as outlined in our algorithms. We wish to characterize
+this communication not only because communication is in general
+expensive, but also because these nearest-neighbor communication
+sequences, specifically, have poor algorithmic strong scaling
+\citep{gropp_high-performance_2001} in much the same way as a
+matrix-vector multiply operation. Therefore, we desire a framework to
+provide a simple, analytic theory based on the properties of the
+linear system that will allow for estimates of the domain decomposed
+behavior of the adjoint Neumann-Ulam method in terms of communication
+costs.
 %%---------------------------------------------------------------------------%%
@@
 \label{sec:symmetric_leakage_fractions}
-With Wagner's work, we observed that domain overlap implemented with a
-domain decomposed Monte Carlo setting reduces the amount of nearest
-neighbor communication that must occur during transport.
+When solving problems where the linear operator is symmetric, a host
+of analytic theories exist based on the eigenvalue spectrum of the
+operator that characterize their behavior in the context of
+deterministic linear solvers. Using past work, these theories are
+adapted to the domain decomposed adjoint Neumann-Ulam method using the
+one-speed, two-dimensional neutron diffusion equation. In this section
+we use a one-speed two-dimensional neutron diffusion system as the
+symmetric model problem. Using the linear system generated by the
+discretization of the model problem, we use a spectral analysis to
+generate analytic relations for the eigenvalues of the operator based
+on system parameters. Using the eigenvalue spectra, we then build
+relationships to characterize the transport of stochastic histories in
+a decomposed domain and the fraction of histories that leak from a
+domain and will therefore have to be communicated. Finally, we compare
+these analytic results to numerical experiments conducted with the
+model problem.
 
-To date, parallel Neumann-Ulam methods have been limited to full
-domain replication with parallelism exploited through individual
-histories \cite{alexandrov_efficient_1998}. In reactor physics Monte
-Carlo applications, however, domain decomposition has been identified
-as a key principle in moving forward in high performance computing to
-enable higher fidelity simulations
-\cite{brunner_comparison_2006,siegel_analysis_2012}. To accomplish
-this, we recognize from the literature that stochastic histories must
-be transported from domain to domain as the simulation progresses and
-they transition to states that are not in the local domain. Because we
-have chosen a domain decomposition strategy in a parallel environment,
-this means that communication of these histories must occur between
-compute nodes owning neighboring pieces of the global domain. We wish
-to characterize this communication not only because communication is
-in general expensive, but also because these nearest-neighbor
-communication sequences, specifically, have poor algorithmic strong
-scaling \cite{gropp_high-performance_2001}.
-
-The purpose of this study is to provide a simple, analytic theory
-based on the properties of the linear system that will allow for
-estimates of the domain decomposed behavior of the adjoint
-Neumann-Ulam method. When solving problems where the linear operator
-is symmetric, a host of analytic theories exist based on the
-eigenvalue spectrum of the operator that characterize their behavior
-in the context of deterministic linear solvers. Using past work, these
-theories are adapted to the domain decomposed adjoint Neumann-Ulam
-method using the one-speed, two-dimensional neutron diffusion
-equation. In this paper we describe the adjoint Neumann-Ulam Monte
-Carlo method followed by a presentation of the model problem. Using
-the linear system generated by the discretization of the model
-problem, we use a spectral analysis to generate analytic relations for
-the eigenvalues of the operator based on system parameters. Using the
-eigenvalue spectra, we then build relationships to characterize the
-transport of stochastic histories in a decomposed domain and the
-fraction of histories that leak from a domain and will therefore have
-to be communicated. Finally, we compare these analytic results to
-numerical experiments conducted with the model problem and draw
-conclusions looking towards future work.
-
-\subsection{Model Problem}
+\subsection{Symmetric Model Problem}
 \label{subsec:model_problem}
@@
 two-dimensional neutron diffusion equation as a model problem
-\cite{duderstadt_nuclear_1976}:
+\citep{duderstadt_nuclear_1976}:
 \begin{equation}
@@
 stencil shown in Figure~\ref{fig:stencil} over a grid of size $h$
-\cite{leveque_finite_2007}:
+\citep{leveque_finite_2007}:
 \begin{multline}
@@
 will compute these eigenvalues by assuming eigenfunctions of the form
-\cite{leveque_finite_2007}:
+\citep{leveque_finite_2007}:
 \begin{equation}
@@
 
-As Eq~(\ref{eq:diffusion_eq}) is diagonally dominant, Jacobi
-preconditioning is sufficient to reduce the spectral radius of the
-iteration matrix below unity and therefore ensure convergence of the
-Neumann series. The preconditioner in this case is then $\ve{M} =
-diag(\ve{D})$ such that we are solving the following linear system:
+As Eq~(\ref{eq:diffusion_eq}) is diagonally dominant, point Jacobi
+preconditioning as outlined in \S\ref{sec:stochastic_preconditioning}
+is sufficient to reduce the spectral radius of the iteration matrix
+below unity and therefore ensure convergence of the Neumann
+series. Applying this preconditioner, we are then solving the
+following diffusion system:
 \begin{equation}
@@
 iteration matrix to be zero such that we have the following stencil:
-\begin{equation}
-  \ve{H}\boldsymbol{\phi} = \frac{\alpha D}{6h^2}[4 \phi_{i-1,j}
-    + 4 \phi_{i+1,j} + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} +
-    \phi_{i-1,j-1} + \phi_{i-1,j+1} + \phi_{i+1,j-1} +
-    \phi_{i+1,j+1}]\:.
+\begin{multline}
+  \ve{H}\boldsymbol{\phi} = \frac{\alpha D}{6h^2}\Big[4 \phi_{i-1,j} +
+    4 \phi_{i+1,j} + 4 \phi_{i,j-1} + 4 \phi_{i,j+1}
+    +\\ \phi_{i-1,j-1} + \phi_{i-1,j+1} + \phi_{i+1,j-1} +
+    \phi_{i+1,j+1}\Big]\:.
   \label{eq:iteration_stencil}
-\end{equation}
+\end{multline}
 Inserting the eigenfunctions defined by
@@
   \label{fig:diffusion_spectrum}
-\end{figure}.
- We find that the maximum eigenvalue exists when $p=q=0$,
-giving the following for the spectral radius of the Jacobi
-preconditioned iteration matrix:
+\end{figure}
+We find that the maximum eigenvalue exists when $p=q=0$, giving the
+following for the spectral radius of the Jacobi preconditioned
+iteration matrix:
 \begin{equation}
@@
 \label{subsec:neumann_convergence}
-The adjoint Monte Carlo method is effectively an approximation to a
-stationary method. Stationary methods for linear systems arise from
-splitting the operator in Eq~(\ref{eq:linear_problem}) and iterating:
-\begin{equation}
-  \ve{x}^{k+1} = \ve{H}\ve{x}^k + \ve{c}\:,
-  \label{eq:linear_iterative_method}
-\end{equation}
-with $k \in \mathbb{Z}^+$ defined as the \textit{iteration
-  index}. Defining $\ve{e}^k = \ve{u}^k - \ve{u}$ as the solution
-error at the $k^{th}$ iterate, the error after $k$ iterations is then:
-\begin{equation}
-  \ve{e}^{k} = \ve{H}^k\ve{e}^0\:. 
-  \label{eq:linear_k_iter_error}
-\end{equation}
-By assuming $\ve{H}$ is diagonalizable \cite{leveque_finite_2007}, we
-then have:
-\begin{equation}
-  ||\ve{e}^{k}||_2 \leq \rho(\ve{H})^k ||\ve{e}^0||_2\:.
-  \label{eq:linear_k_iter_norm3}
-\end{equation}
-
-In the adjoint Neumann-Ulam method, $k$ iterations, equivalent to $k$
+As outlined in \S\ref{sec:mc_preliminaries}, the adjoint Monte Carlo
+method is effectively an approximation to a stationary method. In the
+adjoint Neumann-Ulam method, $k$ iterations, equivalent to $k$
 applications of the iteration matrix, are approximated by a random
@@
 Eq~(\ref{eq:adjoint_neumann_solution})
-\cite{dimov_new_1998,danilov_asymptotic_2000}. This random walk
+\citep{dimov_new_1998,danilov_asymptotic_2000}. This random walk
 length, or the number of transitions before the termination of a
@@
 colleagues for idealized, load balanced situations for full nuclear
-reactor core Monte Carlo simulations \cite{siegel_analysis_2012}.  To
-quantify the number of particles that leak out of the local domain
-they define a leakage fraction, $\Lambda$, as:
+reactor core Monte Carlo neutral particle simulations
+\citep{siegel_analysis_2012}.  To quantify the number of particles
+that leak out of the local domain they define a leakage fraction,
+$\Lambda$, as:
 \begin{equation}
@@
 the Wigner rational approximation. Outlined more thoroughly by Hwang's
-chapter in \cite{azmy_nuclear_2010}, we will use both the Wigner
+chapter in \citep{azmy_nuclear_2010}, we will use both the Wigner
 rational approximation and the mean chord approximation as a means to
@@
 solver.
-\begin{figure}[ht!]
+\begin{figure}[t!]
   \begin{spacing}{1.0}
     \begin{center}
-      \includegraphics[width=4in,clip]{chapters/mc_theory/measured_spec_rad.png}
+      \includegraphics[width=5in,clip]{chapters/mc_theory/measured_spec_rad.png}
     \end{center}
@@
 \label{subsubsec:walk_length}
-With the spectral data in hand, we can go about setting up an
-experiment to measure the length of the random walks generated by the
-adjoint Neumann-Ulam solver. To do this, we again use a $100 \times
-100$ square grid with $h=0.1$ and the absorption cross varied from 0
-to 100 while the scattering cross section was fixed at unity. Three
-weight cutoff values of \sn{1}{-2}, \sn{1}{-4}, and \sn{1}{-8} were
-used with 10,000 histories generated by a point source of strength 1
-in the center of the domain. For each of the histories, the number of
-transitions made was tallied to provide an effective value of $k$ for
-each history. This value was then averaged over all histories to get a
-measured value of $k$ for the particular operator. On the left,
-Figure~\ref{fig:measured_length} presents these measurements as well
-as the analytic result computed by Eq~(\ref{eq:analytic_k}) as a
-function of the iteration matrix spectral radius, $\rho(\ve{H})$. On
-the right, Figure~\ref{fig:measured_length} gives the relative error
+With the eigenvalue derivations verified, we can go about setting up
+an experiment to measure the length of the random walks generated by
+the adjoint Neumann-Ulam solver. To do this, we again use a $100
+\times 100$ square grid with $h=0.1$ and the absorption cross varied
+from 0 to 100 while the scattering cross section was fixed at
+unity. Three weight cutoff values of \sn{1}{-2}, \sn{1}{-4}, and
+\sn{1}{-8} were used with 10,000 histories generated by a point source
+of strength 1 in the center of the domain. For each of the histories,
+the number of transitions made was tallied to provide an effective
+value of $k$ for each history. This value was then averaged over all
+histories to get a measured value of $k$ for the particular
+operator. Figure~\ref{fig:measured_length} presents these measurements
+as well as the analytic result computed by Eq~(\ref{eq:analytic_k}) as
+a function of the iteration matrix spectral radius,
+$\rho(\ve{H})$. FigureOn the right,
+Figure~\ref{fig:measured_length_error} gives the relative error
 between the predicted and observed results. We note good qualitative
@@
 observe a larger relative error for both long and short random walks.
-\begin{figure}[ht!]
+\begin{figure}[t!]
   \begin{spacing}{1.0}
     \begin{center}
-      \includegraphics[width=6.5in,clip]{chapters/mc_theory/measured_length.png}
+      \includegraphics[width=5in,clip]{chapters/mc_theory/walk_length.png}
     \end{center}
@@
         weight cutoff was varied with \sn{1}{-2}, \sn{1}{-4}, and
-        \sn{1}{-8}. In the left plot, the red data was computed
-        numerically by an adjoint Neumann-Ulam implementation while
-        the black dashed data was generated by
-        Eq~(\ref{eq:analytic_k}). In the right plot, the relative
-        error between the predicted and measured results is presented
-        for each weight cutoff.}}
+        \sn{1}{-8}. The red data was computed numerically by an
+        adjoint Neumann-Ulam implementation while the black dashed
+        data was generated by Eq~(\ref{eq:analytic_k}).}}
     \label{fig:measured_length}
+  \end{spacing}
+\end{figure}
+\begin{figure}[t!]
+  \begin{spacing}{1.0}
+    \begin{center}
+      \includegraphics[width=5in,clip]{chapters/mc_theory/walk_length_error.png}
+    \end{center}
+    \caption{\textbf{Measured and analytic random walk length relative
+        error as a function of the iteration matrix spectral radius.}
+      \textit{The weight cutoff was varied with \sn{1}{-2},
+        \sn{1}{-4}, and \sn{1}{-8}. The relative error between the
+        predicted and measured results is presented for each weight
+        cutoff.}}
+    \label{fig:measured_length_error}
   \end{spacing}
@@
 matrix spectral radius.
-\begin{figure}[ht!]
+\begin{figure}[t!]
   \begin{spacing}{1.0}
@@
 tested. 
-\begin{figure}[ht!]
+\begin{figure}[t!]
   \begin{spacing}{1.0}

### 2013-03-25 doc/dissertation/document/chapters/spn_solutions/spn_solutions.tex
adding additional pngs to symmetric spectral analysis for communication

@@
 \label{sec:spn_spectral_analysis}
-Now that we have an understanding of the linear system generated by
-the $SP_N$ approximation to the neutron transport equation and the
-various parameters in the system that may be adjusted, we can consider
-various means of solution. As the solution is asymmetric, modern
-iterative solvers should be considered for this task. The performance
-of Krylov subspace methods are bound to various properties of the
-eigenvalue spectrum of the linear operator while methods based on
-stationary iterations have performance limits imposed by the spectral
-radius of the sytem. In this section we will review the important
-spectral properties to study for common iterative methods for
-asymmetric systems. Based on these properties, we will then perform a
-parameter-based spectral analysis for the linear operator generated by
-the multigroup $SP_N$ equations. Preconditioning strategies for these
-methods are not considered. Krylov subspace method restarts and
-truncation will also not be considered.
-
-\subsection{Stationary Methods for Linear Systems}
-\label{subsec:stationary_solvers}
-Stationary methods for linear systems arise from splitting the
-operator in Eq~(\ref{eq:multigroup_matrix_system})
-\cite{leveque_finite_2007}:
-\begin{equation}
-  \mathbb{L} = \mathbf{M} - \mathbf{N}\:,
-  \label{eq:split_linear_operator}
-\end{equation}
-where the choice of $\mathbf{M}$ and $\mathbf{N}$ will be dictated by
-the particular method chosen. Using this split definition of the
-operator we can then write:
-\begin{equation}
-  \mathbf{M}\mathbb{U} - \mathbf{N}\mathbb{U} = \mathbb{Q}\:.
-  \label{eq:linear_split_equation1}
-\end{equation}
-By rearranging, we can generate a form more useful for analysis:
-\begin{equation}
-  \mathbb{U} = \mathbf{H}\mathbb{U} + \mathbf{c}\:,
-  \label{eq:linear_split_equation2}
-\end{equation}
-where $\mathbf{H}=\mathbf{M}^{-1}\mathbf{N}$ is defined as the
-\textit{iteration matrix} and
-$\mathbf{c}=\mathbf{M}^{-1}\mathbb{Q}$. With the solution vector on
-both the left and right hand sides, an iterative method can then be
-formed:
-\begin{equation}
-    \mathbb{U}^{k+1} = \mathbf{H}\mathbb{U}^k + \mathbf{c}\:,
-  \label{eq:linear_iterative_method}
-\end{equation}
-with $k \in \mathbf{Z}^+$ defined as the \textit{iteration
-  index}. Defining $\mathbf{e}^k = \mathbf{u}^k - \mathbf{u}$ as the
-solution error at the $k^{th}$ iterate, we can subtract
-Eq~(\ref{eq:linear_split_equation2}) from
-Eq~(\ref{eq:linear_iterative_method}):
-\begin{equation}
-  \mathbf{e}^{k+1} = \mathbf{H}\mathbf{e}^k\:.
-  \label{eq:linear_iterative_error}
-\end{equation}
-Our error after $k$ iterations is then:
-\begin{equation}
-  \mathbf{e}^{k} = \mathbf{H}^k\mathbf{e}^0\:.
-  \label{eq:linear_k_iter_error}
-\end{equation}
-In other words, successive application of the iteration matrix is the
-mechanism driving down the error in a stationary method. By assuming
-$\mathbf{H}$ is diagonalizable \cite{saad_iterative_2003}, we then
-have:
-\begin{equation}
-  \mathbf{e}^{k} =
-  \mathbf{R}\boldsymbol{\Lambda}^k\mathbf{R}^{-1}\mathbf{e}^0\:,
-  \label{eq:linear_k_iter_error_diag}
-\end{equation}
-where $\boldsymbol{\Lambda}$ contains the Eigenvalues of $\mathbf{H}$
-on its diagonal and the columns of $\mathbf{R}$ contain the
-Eigenvectors of $\mathbf{H}$. Computing the 2-norm of the above form
-then gives:
-\begin{equation}
-  ||\mathbf{e}^{k}||_2 \leq ||\boldsymbol{\Lambda}^k||_2\ 
-  ||\mathbf{R}||_2\ ||\mathbf{R}^{-1}||_2\ ||\mathbf{e}^0||_2\:,
-  \label{eq:linear_k_iter_norm1}
-\end{equation}
-which gives:
-\begin{equation}
-  ||\mathbf{e}^{k}||_2 \leq \rho(\mathbf{H})^k \kappa(\mathbf{R})
-  ||\mathbf{e}^0||_2\:.
-  \label{eq:linear_k_iter_norm2}
-\end{equation}
-For iteration matrices where the Eigenvectors are orthogonal,
-$\kappa(\mathbf{R})=1$ and the error bound reduces to:
-\begin{equation}
-  ||\mathbf{e}^{k}||_2 \leq \rho(\mathbf{H})^k
-  ||\mathbf{e}^0||_2\:.
-  \label{eq:linear_k_iter_norm3}
-\end{equation}
-We can now restrict $\mathbf{H}$ by asserting that $\rho(\mathbf{H}) <
-1$ for a stationary method to converge such that $k$ applications of
-the iteration matrix will not cause the error to grow in
-Eq~(\ref{eq:linear_k_iter_norm3}). In addition, the performance of the
-stationary method is dictated by $\rho(\mathbf{H})$ such that
-performance of the method decreases as it approaches to unity.
-
-\subsection{Krylov Subspace Methods for Linear Systems}
-\label{subsec:subspace_solvers}
-The performance of different Krylov subspace solvers can depend on
-different properties of the linear system. As a means to study these
-differences, we use a 1992 paper by Nachtigal and colleagues
-\cite{nachtigal_how_1992} that performs and in-depth analysis of three
-Krylov subspace methods that are still used today in modern physics
-applications: conjugate gradient iteration for the normal equations
-(CGN), generalized minimum residual iteration (GMRES), and a
-biconjugate gradient-based method (CGS). For our work, we will
-consider only their analysis of GMRES as it currently leveraged in
-Denovo for $SP_N$ solutions and as it also applies to CGS for most
-matrices.
-
-For GMRES, at each iteration we are extracting a correction for the
-solution from the Krylov subspace:
-\begin{equation}
-  \mathbb{U}^k \in \mathbb{U}^0 + \langle \mathbf{r_0}, \mathbb{L}
-  \mathbf{r_0}, \cdots, \mathbb{L}^{k-1} \mathbf{r_0} \rangle\:,
-  \label{eq:gmres_iteration}
-\end{equation}
-where the extraction is constrained by ensuring orthogonality of the new
-residual with the Krylov space:
-\begin{equation}
-  \mathbf{r}^k \bot \langle \mathbb{L} \mathbf{r_0}, \mathbb{L}^2
-  \mathbf{r_0}, \cdots, \mathbb{L}^{k} \mathbf{r_0} \rangle\:,
-  \label{eq:gmres_constraint}
-\end{equation}
-with the residual at the $k^{th}$ iteration defined as:
-\begin{equation}
-  \mathbf{r}^k = \mathbb{Q} - \mathbb{L}\mathbb{U}^k\:.
-  \label{eq:residual}
-\end{equation}
-From \cite{nachtigal_how_1992}, we then have the following error at
-each iteration:
-\begin{equation}
-  \mathbf{e}^k = p_k(\mathbb{L})\mathbf{e}^0\:,
-  \label{eq:gmres_iteration_error}
-\end{equation}
-where $p_k(\mathbb{L})$ is an arbitrary $k^{th}$ degree polynomial and
-$p_k(0)=1$. This is a very similar form for the iteration error as
-observed for stationary methods in Eq~(\ref{eq:linear_k_iter_norm3})
-where now as these $p_k$ polynomials are constructed as the iterations
-progress, a rapid decrease in $||p_k(\mathbb{L})||$ means a rapid rate
-of convergence. For matrices $\mathbb{L}$ that are approximately
-normal (which was exactly the assumption we made to arrive at
-Eq~(\ref{eq:linear_k_iter_norm3})), then we can write the polynomial
-norm as:
-\begin{equation}
-  ||p_k(\mathbb{L})|| = \sup_{z\in\Lambda(\mathbb{L})}|p_k(z)|\:,
-  \label{eq:gmres_polynomial}
-\end{equation}
-where $\Lambda(\mathbb{L})$ is the eigenvalue spectrum of the normal
-matrix $\mathbb{L}$. So if we could obtain the polynomials, the
-largest eigenvalue of $\mathbb{L}$ could be used to calculate
-convergence rates. Next we consider a choice of polynomials to use for
-the convergence rate approximation. They should of course be
-orthogonal to reflect the orthogonal projection mechanism by which the
-residual is reduced. In Nachtigal's work, Chebyshev polynomials were
-used and defined on the interval of Eigenvalues,
-$[\lambda_{min},\lambda_{max}]$, and then normalized via the
-constraint $p_k(0)=1$. Thus the rate of convergence for these
-polynomials is an approximation to the rate of convergence of the
-method.
-
-We can check the validity of the normal
-approximation by computing the condition number of $\mathbb{L}$:
-\begin{equation}
-  \kappa(\mathbb{L}) = ||\mathbb{L}||\ ||\mathbb{L}^{-1}||
-  \label{eq:operator_condition_number}
-\end{equation}
-The closer $\kappa(\mathbb{L})$ is to one, the more normal the matrix
-and the better the above approximation. The same check can also be
-performed to verify the assumption used to generate
-Eq~(\ref{eq:linear_k_iter_norm3}) where $\kappa(\mathbf{H})$ can be
-computed. For our analysis, we will compute $\kappa_2$ where the
-matrix norms are induced by the 2-norm of a vector such that:
-\begin{equation}
-  ||\mathbb{L}||_2 = \sqrt{\rho(\mathbb{L}^T\mathbb{L})}\:.
-  \label{eq:matrix_2_norm}
-\end{equation}
-
-\subsection{Organization of the Spectral Analysis}
-\label{subsec:spectral_organization}
-For the spectral analysis we will perform a numerical parameter study
-for the multigroup $SP_N$ equations. Based on the above information
-for iterative methods, several key quantities should be computed for
-the $SP_N$ system:
-\begin{itemize}
-\item Eigenvalue spectrum for the linear operator $\mathbb{L}$
-\item Largest eigenvalue for the linear operator $\mathbb{L}$
-\item Smallest eigenvalue for the linear operator $\mathbb{L}$
-\item Condition number for the linear operator $\mathbb{L}$
-\item For each stationary method:
-  \begin{itemize}
-  \item Eigenvalue spectrum for the iteration matrix $\mathbf{H}$
-  \item Largest eigenvalue for the iteration matrix $\mathbf{H}$
-  \item Smallest eigenvalue for the iteration matrix $\mathbf{H}$
-  \item Condition number for the iteration matrix $\mathbf{H}$
-  \end{itemize}
-\item For GMRES:
-  \begin{itemize}
-  \item Chebyshev polynomial norm using eigenvalues of $\mathbb{L}$
-    for a fixed number of iterations determined by stationary method
-    convergence
-  \end{itemize}
-\end{itemize}
-In addition, the following parameters may be varied in the $SP_N$
-system in Denovo:
-\begin{itemize}
-  \item $SP_N$ (angular flux expansion) order
-  \item $P_N$ (scattering cross section expansion) order
-  \item Number of energy groups $N_g$
-  \item Total cross sections $\sigma^g$
-  \item Scattering cross section matrix $\sigma_{sn}^{gg'}$
-  \item Boundary conditions (Marshak or reflecting)
-  \item Finite volume mesh size ($\Delta$ in all directions)
-  \item Number of mesh elements
-  \item Source of neutrons (Uniform or Point)
-\end{itemize}
-We will restrict the problem description in Denovo to one similar to
-Brantley and Larsen's $SP_N$ work where the spectral radius was
-computed for convergence analysis of several proposed solution
-methods. From this definition we choose a single material problem with
-a uniform source of unity in all energy groups. For the boundary
-conditions, we choose reflecting conditions on all six sides of the
-cubic domain for an effectively homogenous infinite medium problem
-which we will mesh with a $5\times 5\times 5$ grid with
-$\Delta=0.1$. We are then left to vary the angular and energy
-components of the system in our parameter studies: the $SP_N$ order,
-the number of energy groups, and their cross sections of order
-$P_N$. In Brantley and Larsen's work, a single energy group was used
-and therefore we may not use their cross sections in our multigroup
-computations. For our work, we will always choose a total cross
-section of unity for all groups and all flux moments. For the
-scattering matrix, two classes will be chosen: one with upscatter and
-one without. For the downscatter only case, all groups may scatter
-within the group and to all groups of lower energy (yielding a lower
-triangular scattering matrix) with a cross section of 0.5. For the
-upscatter case, all groups may scatter within the group and to all
-groups of lower energy given a cross section of 0.5 and for upscatter,
-all groups may scatter up to two groups higher in energy with a cross
-section of 0.25. All cross section matrices will be equivalent for
-each scattering moment. For these cross sections, the $P_N$ order will
-be varied with $N=0,1,3$. Both 1 and 10 energy groups will be used
-with the 10 energy group case using the scattering matrices specified
-above. For the angular flux, the $SP_N$ order will be varied with
-$N=1,3,5,7$. This gives a total of 36 linear system variations for the
-spectral analysis.
 

### 2013-03-25 doc/dissertation/document/chapters/mc_theory/mc_theory.tex
updating symmetric leakage section in mc theory

@@
 a function of the iteration matrix spectral radius,
-$\rho(\ve{H})$. FigureOn the right,
-Figure~\ref{fig:measured_length_error} gives the relative error
-between the predicted and observed results. We note good qualitative
-agreement between the measured and analytic results. However, we
-observe a larger relative error for both long and short random walks.
+$\rho(\ve{H})$. Figure~\ref{fig:measured_length_error} gives the
+relative error between the predicted and observed results. We note
+good qualitative agreement between the measured and analytic
+results. However, we observe a larger relative error for both long and
+short random walks.
 \begin{figure}[t!]
@@
     \begin{center}
-      \includegraphics[width=4.5in,clip]{chapters/mc_theory/leakage_variation.png}
+      \includegraphics[width=5.25in,clip]{chapters/mc_theory/leakage_variation.png}
     \end{center}
@@
     \begin{center}
-      \includegraphics[width=4.5in,clip]{chapters/mc_theory/leakage_error.png}
+      \includegraphics[width=5in,clip]{chapters/mc_theory/leakage_error.png}
     \end{center}

### 2013-03-25 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
updating emission kernel integrand in transport equation and citation syntax in spn equations

@@
 methods for large scale problems are commonly variants on the discrete
-ordinates ($S_N$) method \cite{evans_denovo:_2010}. For full reactor
+ordinates ($S_N$) method \citep{evans_denovo:_2010}. For full reactor
 core neutronics simulations, the $S_N$ method requires potentially
@@
 
-In this document we derive the multigroup $SP_N$ equations from the
-Boltzmann transport equation. We then perform a spectral analysis on
-the resulting system of equations to determine its potential
-performance with both modern Krylov methods and Monte Carlo methods
-for asymmetric linear systems. Following this, we devise a Monte Carlo
-synthetic acceleration method as a proposed solution scheme.
-
 In this chapter, we derive the $SP_N$ equations, closely following the
@@
 As a starting point we define the time-independent neutron transport
-equation \cite{lewis_computational_1993}:
+equation \citep{lewis_computational_1993}:
 \begin{multline}
@@
 are solutions to Legendre's differential equation. They have the
-following form \cite{lewis_computational_1993}:
+following form \citep{lewis_computational_1993}:
 \begin{equation}
@@
   \mu \frac{\partial}{\partial x} \psi(x,\mu) + \sigma(x) \psi(x,\mu)
-  = \int d\Omega' \sigma_s(x,\hat{\Omega}' \rightarrow \hat{\Omega})
-  \psi(\vec{r},\hat{\Omega}') + \frac{q(x)}{4 \pi}\:.
+  = \int \sigma_s(x,\hat{\Omega}' \rightarrow \hat{\Omega})
+  \psi(\vec{r},\hat{\Omega}') d\Omega' + \frac{q(x)}{4 \pi}\:.
   \label{eq:mono_transport}
@@
 energy group. In practice, only odd-numbered $P_N$ orders are
-generally used \cite{lewis_computational_1993}. This is due to the
+generally used \citep{lewis_computational_1993}. This is due to the
 fact that using odd $N$ yields an even number of $N+1$ equations which
@@
 \label{subsbusec:bcs_pn}
-Per the analysis in \cite{lewis_computational_1993}, two types of
+Per the analysis in \citep{lewis_computational_1993}, two types of
 boundary conditions will be discussed: reflecting and Marshak. Marshak

### 2013-03-26 doc/dissertation/document/chapters/mc_background/mc_background.tex
adding additional pngs to symmetric spectral analysis for communication
updating margins

@@
 memory.
-\begin{figure}[ht!]
+\begin{figure}[t!]
   \centering
@@
 
-\begin{figure}[ht!]
+\begin{figure}[t!]
   \centering
@@
 and \ref{fig:seq_500} respectively. In both cases, the Sequential
-method is subject to two regimes of exponential convergence with the
-later regime converging the slowest while the MCSA method exhibits a
-single rate of exponential convergence observed to be much higher than
-that computed by Halton's method. Even with the doubling of the number
-of stochastic histories computed per time step in order to ensure
-convergence for the Sequential method, we still see robustness issues
-with a non-monotonically decreasing residual observed for the $N=100$
-case. In both cases the MCSA solver is observed to be robust with a
-monotonically decreasing residual.
+method is subject to two regimes of exponential convergence with high
+frequency error modes removed in the first regime leaving lower
+frequency and slower converging error modes in the second. Using MCSA
+we a see a single rate of exponential convergence observed to be much
+higher than that computed by Halton's method due to the fact that the
+extra Richardson iteration is providing a smoothing effect to
+alleviate the error mode variations. Even with the doubling of the
+number of stochastic histories computed per time step in order to
+ensure convergence for the Sequential method, we still see robustness
+issues with a non-monotonically decreasing residual observed for the
+$N=100$ case. In both cases the MCSA solver is observed to be robust
+with a monotonically decreasing residual.
 
-\begin{figure}[ht!]
+\begin{figure}[t!]
   \centering

### 2013-03-26 doc/dissertation/document/chapters/linear_problem/linear_problem.tex
fixing typo in linear problem chapter and removing resuing random walks subsection in mc background chapter
adding additional pngs to symmetric spectral analysis for communication
updating citation format in linear problem

@@
 
-\begin{figure}[htpb!]
+\begin{figure}[t!]
   \begin{center}
@@
 subspaces, the GMRES iteration may be formulated as follows:
-\begin{algorithm}[htpb!]
+\begin{algorithm}[t!]
   \caption{GMRES Iteration}
@@
 number of global communications required (consider
-\cite{sosonkina_scalable_1998} as an example of reducing global
+\citep{sosonkina_scalable_1998} as an example of reducing global
 operation counts using a different orthogonalization procedure than
@@
 communication patterns. Libraries such as Zoltan
-\citep{devine_zoltan_2002}, provide an implementations of such
+\citep{devine_zoltan_2002}, provide implementations of such
 algorithms.
@@
 processors.
-\begin{figure}[htpb!]
+\begin{figure}[t!]
   \begin{center}
@@
 processors. 
-\begin{figure}[htpb!]
+\begin{figure}[t!]
   \begin{center}

### 2013-03-26 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
updates to spn equations introduction

@@
 trillions of unknown angular flux moments to be computed to achieve
-good accuracy for the repsonses of interest. Other forms of the
-transport problem, including the $P_N$ method, take on a simpler form
-than the more common $S_N$ methods but lack in accuracy when compared
-while still requiring considerable computational resources for
-solutions in multiple dimensions.
+good accuracy for the repsonses of interest
+\citep{slaybaugh_acceleration_2011}. Other forms of the transport
+problem, including the $P_N$ method, take on a simpler form than the
+more common $S_N$ methods but lack in accuracy when compared while
+still requiring considerable computational resources for solutions in
+multiple dimensions.
 
@@
 $P_N$ ($SP_N$) equations. Up until around the 1990's, the $SP_N$
-methods was either widely unknown, widely unused, or combination of
+method was either widely unknown, widely unused, or combination of
 both even though numerical studies showed promising results with
@@
 full core reactor models. By fully forming these equations and
-formulating them as a linear algebra problem, we now have access to
-all of the modern advancements in computational linear algebra
-including Krylov solvers for asymmetric systems and preconditioning
-methods such as algebraic multigrid. This leads us to then explore the
+formulating them as a linear algebra problem instead of using the
+explicit iterative methods of the past, we now have access to all of
+the modern advancements in computational linear algebra including
+Krylov solvers for asymmetric systems and preconditioning methods such
+as algebraic multigrid. This leads us to then explore the
 applicability of our work in discrete Monte Carlo methods for linear
@@
 usage of computational resources may be observed compared to modern
-solution techniques such as those suggested. In addition, solving the
+solution techniques such as those suggested due to the form of the
+matrices generated by the discretization. In addition, solving the
 $SP_N$ equations in this way also breaks away from the $S_N$ forms of
@@
 In this chapter, we derive the $SP_N$ equations, closely following the
-work of Evans. We begin by stating the general time-independent
-neutron transport equation followed by a derivation of the $P_N$
-equations in planar geometry for multiple energy groups. From these
-equations, we then apply a set of approximations to yield the
-multi-dimensional, multi-group $SP_N$ equations for fixed source
-problems.
+work of Evans, in order to gain full understanding of the underlying
+system and its behavior in a Monte Carlo context. We begin by stating
+the general time-independent neutron transport equation followed by a
+derivation of the $P_N$ equations in planar geometry for multiple
+energy groups. From these equations, we then apply a set of
+approximations to yield the multi-dimensional, multi-group $SP_N$
+equations for fixed source problems. Boundary conditions are also
+discussed.
 

### 2013-03-26 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
cleaning up spn derivation

@@
 First, the $P_N$ equations can be simplified to $(N+1)/2$ second-order
-equations by solving for the $n^th$ Legendre flux moment in the
+equations by solving for the $n^{th}$ Legendre flux moment in the
 odd-order equations:
@@
 Eq~(\ref{eq:reduced_pn}) as compared to the original $P_N$
-equations. To extend these equations to multiple dimensions, gelbard
-simple replaced the x-plane spatial derivatives in the reduced set of
+equations. To extend these equations to multiple dimensions, Gelbard
+simply replaced the planar spatial derivatives in the reduced set of
 equations with general multidimensional gradient operators:
@@
 yielding a multidimensional set of $(N+1)/1$ angular coupled equations
-known as the $SP_N$. Again, we provide closure to this set of
-equations with $\phi_{N+1} = 0$. As a concrete example, we will
+defined as the $SP_N$ equations. Again, we provide closure to this set
+of equations with $\phi_{N+1} = 0$. As a concrete example, we will
 consider the $SP_7$ equations:
@@
   \mathbf{A} = 
-  \begin{bmatrix}
+  {\tiny \begin{bmatrix}
     (\Sigma_0) &
@@
     (-\frac{2}{3}\Sigma_0) &
-    (\frac{4}{15}\Sigma_0 + \frac{1}{3}\Sigma_2) &
+    (\frac{4}{9}\Sigma_0 + \frac{5}{9}\Sigma_2) &
     (-\frac{16}{45}\Sigma_0 - \frac{4}{9}\Sigma_2) &
@@
     \frac{324}{1225}\Sigma_4 + \frac{13}{49}\Sigma_6)
-  \end{bmatrix}\:.
+  \end{bmatrix}}\:.
   \label{eq:A_matrix}
@@
 Eqs~(\ref{eq:spn7_subs}) and (\ref{eq:spn7_subs_inverse}) for
-consitency with the equations defined on the domain. In addition, we
+consistency with the equations defined on the domain. In addition, we
 apply the $SP_N$ approximation to the derivatives by assuming they are
@@
   \mathbf{A} = 
-  \begin{bmatrix}
+  {\tiny \begin{bmatrix}
     (\mathbf{\Sigma}_0) &
@@
     (-\frac{2}{3}\mathbf{\Sigma}_0) &
-    (\frac{4}{15}\mathbf{\Sigma}_0 + \frac{1}{3}\mathbf{\Sigma}_2) &
+    (\frac{4}{9}\mathbf{\Sigma}_0 + \frac{5}{9}\mathbf{\Sigma}_2) &
     (-\frac{16}{45}\mathbf{\Sigma}_0 - \frac{4}{9}\mathbf{\Sigma}_2) &
@@
     \frac{324}{1225}\mathbf{\Sigma}_4 + \frac{13}{49}\mathbf{\Sigma}_6)
-  \end{bmatrix}\:.
+  \end{bmatrix}}\:.
   \label{eq:A_block_matrix}

### 2013-03-26 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding richardson iteration matrix spectral radius function

@@
 
+%%---------------------------------------------------------------------------%%
 \section{Derivation of the $P_N$ Equations}
@@
 
+%%---------------------------------------------------------------------------%%
 \section{Derivation of the $SP_N$ Equations}
@@
 Eq~(\ref{eq:odd_moments}) to remove them, leaving only the even
-moments and a set of differential equations:
+moments and a set of first-order differential equations:
 \begin{subequations}
@@
 
+%%---------------------------------------------------------------------------%%
 \section{Derivation of the Multigroup $SP_N$ Equations}
@@
 
-\section{A Note on Spatial Discretization and Matrix Symmetry}
+%%---------------------------------------------------------------------------%%
+\section{Spatial Discretization}
 \label{sec:spatial_discretization}
@@
 spatial discretization is a natural extension of the balance
-principles used to arrive at the general transport equation. The
-result of this finite volume discretization is one that is symmetric
-for all equations in the domain with respect to the spatial components
-of the solution.
-
-Given the symmetry of the spatial discretization equations in the
-domain given by monoenergetic Eq~(\ref{eq:spn_matrix}) form a
-symmetric linear operator acting on the pseudo-moment vector to give:
-\begin{equation}
-  \mathbf{L}\mathbf{u}=\mathbf{Q}\:,
-  \label{eq:matrix_system}
-\end{equation}
-where $\mathbf{L} = -\nabla \cdot \mathbf{D} \nabla +
-\mathbf{A}$. Moving to the multigroup formulation in
-Eq~(\ref{eq:spn_multigroup_system}) we have:
-\begin{equation}
-  \mathbb{L}\mathbb{U}=\mathbb{Q}\:,
-  \label{eq:multigroup_matrix_system}
-\end{equation}
-where $\mathbb{L} = -\nabla \cdot \mathbb{D} \nabla +
-\mathbb{A}$. Whether or not the resulting matrix is symmetric is
-determined by the group scattering matrices, $\sigma_{sn}^{gg'}$, for
-each of the moments. In general, the cross sections in these matrices
-do not form a symmetric matrix and therefore the resulting linear
-operator in Eq~(\ref{eq:multigroup_matrix_system}) will always be
-asymmetric for multigroup problems. Solutions to this linear system
-will then require techniques specifically formulated for asymmetric
-systems.
+principles used to arrive at the general transport equation.
+
+%%---------------------------------------------------------------------------%%
+\section{Spectral Analysis of the $SP_N$ Equations}
+\label{sec:spn_spectral_analysis}
+Before we can move on to solving the $SP_N$ equations with both
+conventional and Monte Carlo methods, we must first understand their
+spectral character to ensure the MCSA will apply. As presented in
+Chapter~\ref{chap:mc_background}, the spectral radius of the iteration
+matrix must be less than unity to ensure convergence and a finite
+variance. Therefore, we will compute the spectral radius of the $SP_N$
+equations parametrically varying the $SP_N$ order, the $P_N$ order,
+the number of energy groups, and upscattering/downscattering between
+energy groups.
+
+\subsection{Point Jacobi Spectral Analysis Results}
+\label{subsec:spn_analysis_results}
+
+\subsection{Preconditioning Strategy}
+\label{subsec:spn_preconditioning}
+
+\subsection{Block Jacobi Spectral Analysis Results}
+\label{subsec:spn_analysis_results}

### 2013-03-26 doc/spn/data/richarson.m

@@ -1,0 +1,12 @@
+function [] = richardson( matname )
+% Richardson iteration matrix eigenvalues
+filename = strcat(matname,'.mat');
+A = load(filename,'-ascii');
+A = spconvert(A); 
+I = speye(size(A,1)); 
+H = I-A;
+opts.tol=1.0e-8;
+eigs(H,1,'lm',opts)
+
+end
+

### 2013-03-26 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding spn sparsity plots to spn equations chapter

@@
 spectral character to ensure the MCSA will apply. As presented in
-Chapter~\ref{chap:mc_background}, the spectral radius of the iteration
-matrix must be less than unity to ensure convergence and a finite
-variance. Therefore, we will compute the spectral radius of the $SP_N$
-equations parametrically varying the $SP_N$ order, the $P_N$ order,
-the number of energy groups, and upscattering/downscattering between
-energy groups.
+Chapter~\ref{ch:stochastic_methods}, the spectral radius of the
+iteration matrix must be less than unity to ensure convergence and a
+finite variance. Therefore, we will compute the spectral radius of the
+$SP_N$ equations while parametrically varying the $SP_N$ order, the
+$P_N$ order, the number of energy groups, and
+upscattering/downscattering between energy groups. For each problem in
+the parameter study, the values given in
+Table~\ref{tab:spn_fixed_parameters} were fixed for each spatial
+location and energy group.
+\begin{table}[t!]
+  \begin{center}
+    \begin{tabular}{ll}\hline\hline
+      \multicolumn{1}{c}{Parameter}& 
+      \multicolumn{1}{c}{Value} \\\hline\hline
+      Mesh Element Size & 1.0 \\
+      Mesh Elements & $4 \times 4 \times 4$ \\
+      Reflecting Boundaries & True \\
+      Materials & 1 \\
+      Source Strength & 1.0 \\
+      Total Cross Section & 5.0 \\
+      In-Group Cross Section & 0.25 \\
+      Downscatter Cross Section & 1.0 \\
+      Upscatter Cross Section & 0.1 \\
+      Eigenvalue Solver Tolerance & $1.0\times10^{-8}$ \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Fixed parameters for the $SP_N$ spectral radius
+      parameter study.} \textit{The parameters were fixed for each
+      spatial location and energy group.}}
+  \label{tab:spn_fixed_parameters}
+\end{table}
+Mesh element sizes were the same in all cardinal directions with
+reflecting conditions on each boundary. For the energy parameter, 1
+and 10 energy groups were used with upscatter/downscatter varied for
+the 10 group case. For the downscatter cases, all groups could scatter
+to all lower energy groups with the cross sections given in
+Table~\ref{tab:spn_fixed_parameters} while for the upscatter case all
+groups could scatter to all other energy groups with different cross
+sections for upscatter and downscatter used. A uniform isotropic
+source of neutrons is given in each group as well.
+
+The matrices generated by the discretization of this problem are
+sparse with a block-based form as dictated by
+Eq~(\ref{eq:spn_multigroup_system}). Figure~\ref{fig:group1} gives the
+sparsity pattern for a monoenergetic $SP_7$ discretization while
+Figures~\ref{fig:group10ds} and \ref{fig:group10us} give the sparsity
+patterns for the same problem for 10 energy groups without and with
+upscatter respectively. For these figures, the parameters in
+Table~\ref{tab:spn_fixed_parameters} with the number of spatial
+elements reduced to $2 \times 2 \times 2$ to show the blocks generated
+by the discretization.
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=3in]{chapters/spn_equations/group1.png}
+  \end{center}
+  \caption{\textbf{Sparsity pattern for 1-group $SP_7$
+      discretization.} \textit{A $2\times 2 \times 2$ element mesh was
+      used to show detil of the blocks formed by the discretization.}}
+  \label{fig:group1}
+\end{figure}
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=3in]{chapters/spn_equations/group10ds.png}
+  \end{center}
+  \caption{\textbf{Sparsity pattern for 10-group $SP_7$ discretization
+      with downscatter only.} \textit{A $2\times 2 \times 2$ element
+      mesh was used to show detil of the blocks formed by the
+      discretization.}}
+  \label{fig:group10ds}
+\end{figure}
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=3in]{chapters/spn_equations/group10us.png}
+  \end{center}
+  \caption{\textbf{Sparsity pattern for 10-group $SP_7$ discretization
+      with downscatter and upscatter.} \textit{A $2\times 2 \times 2$ element
+      mesh was used to show detil of the blocks formed by the
+      discretization.}}
+  \label{fig:group10us}
+\end{figure}
 

### 2013-03-26 doc/spn/data/spn.py

@@ -13,8 +13,8 @@ initialize(sys.argv)
 ##---------------------------------------------------------------------------##
 pn_order = 1
 spn_order = 7
-num_groups = 1
-upscatter = False
+num_groups = 10
+upscatter = True
 
 sigma_t = 5.0
 sigma_ing = 0.25
@@ -28,9 +28,9 @@ entries = {
     "delta_x"     : 1.0,
     "delta_y"     : 1.0,
     "delta_z"     : 1.0,
-    "num_cells_i" : 5,
-    "num_cells_j" : 5,
-    "num_cells_k" : 5,
+    "num_cells_i" : 2,
+    "num_cells_j" : 2,
+    "num_cells_k" : 2,
     "num_groups"  : num_groups,
     "SPN_order"   : spn_order,
     "boundary"    : "reflect",

### 2013-03-26 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding block jacobi preconditioning section to spn equations

@@
       \multicolumn{1}{c}{Parameter}& 
-      \multicolumn{1}{c}{Value} \\\hline\hline
+      \multicolumn{1}{c}{Value} \\\hline
       Mesh Element Size & 1.0 \\
@@
       discretization.} \textit{A $2\times 2 \times 2$ element mesh was
-      used to show detil of the blocks formed by the discretization.}}
+      used to show detail of the blocks formed by the discretization.}}
   \label{fig:group1}
@@
       with downscatter only.} \textit{A $2\times 2 \times 2$ element
-      mesh was used to show detil of the blocks formed by the
+      mesh was used to show detail of the blocks formed by the
       discretization.}}
@@
       with downscatter and upscatter.} \textit{A $2\times 2 \times 2$ element
-      mesh was used to show detil of the blocks formed by the
+      mesh was used to show detail of the blocks formed by the
       discretization.}}
@@
 \end{figure}
+We note a few key features of the sparsity plots. The first is that
+for multigroup problems without full upscatter and downscatter
+(i.e. Figure~\ref{fig:group10ds}), the resulting matrix is asymmetric
+and therefore a linear solver that can handle asymmetric linear
+systems is required. Nearly all problems of interest will not have
+full upscattering or downscattering. Second we note the largely
+diagonal character of these systems, although the blocks from
+Eq~(\ref{eq:A_block_matrix}) are readily apparent. Our first attempt
+at preconditioning this system will be to use the point Jacobi
+preconditioning from \S\ref{sec:stochastic_preconditioning} due to
+this diagonal form.
 
@@
 \label{subsec:spn_analysis_results}
+Spectral radius computations were performed for the cases described
+above for the point Jacobi preconditioned iteration matrix $\ve{H} =
+\ve{I}-\ve{M}^{-1}\ve{A}$ with $\ve{M} =
+diag(\ve{A})$. Table~\ref{tab:group1pj} gives the results for the
+1-group case, Table~\ref{tab:group10dspj} for the 10-group case with
+full downscatter only and Table~\ref{tab:group10uspj} for the 10-group
+case with full downscatter and full upscatter.
+\begin{table}[t!]
+  \begin{center}
+    \begin{tabular}{cccccc}\hline\hline
+      \multicolumn{1}{c}{}& 
+      \multicolumn{1}{c}{}& 
+      \multicolumn{1}{c}{}& 
+      \multicolumn{1}{c}{$SP_N$ Order}& 
+      \multicolumn{1}{c}{}& 
+      \multicolumn{1}{c}{} \\
+       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
+       & \textbf{0} & 0.0635 & 0.6722 & 1.3144 & 1.976 \\
+       & \textbf{1} & 0.0666 & 0.6728 & 1.3141 & 1.9755 \\
+      $P_N$ Order & \textbf{3} & 0.0666 & 0.6822 & 1.3141 & 1.9755 \\
+       & \textbf{5} & 0.0666 & 0.6822 & 1.3278 & 1.9847 \\
+       & \textbf{7} & 0.0666 & 0.6822 & 1.3278 & 1.9917 \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Spectral radius results for the point Jacobi
+      preconditioned iteration matrix with 1 energy group.}}
+  \label{tab:group1pj}
+\end{table}
+\begin{table}[t!]
+  \begin{center}
+    \begin{tabular}{cccccc}\hline\hline
+      \multicolumn{1}{c}{}& 
+      \multicolumn{1}{c}{}& 
+      \multicolumn{1}{c}{}& 
+      \multicolumn{1}{c}{$SP_N$ Order}& 
+      \multicolumn{1}{c}{}& 
+      \multicolumn{1}{c}{} \\
+       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
+       & \textbf{0} & 0.0655 & 0.677 & 1.32 & 1.982 \\
+       & \textbf{1} & 0.071 & 0.67777 & 1.319 & 1.982 \\
+      $P_N$ Order & \textbf{3} & 0.071 & 0.687 & 1.327 & 1.9872 \\
+       & \textbf{5} & 0.071 & 0.687 & 1.336 & 1.997 \\
+       & \textbf{7} & 0.071 & 0.687 & 1.336 & 1.9995 \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Spectral radius results for the point Jacobi
+      preconditioned iteration matrix with 10 energy groups and full
+      downscatter.}}
+  \label{tab:group10dspj}
+\end{table}
+\begin{table}[t!]
+  \begin{center}
+    \begin{tabular}{cccccc}\hline\hline
+      \multicolumn{1}{c}{}& 
+      \multicolumn{1}{c}{}& 
+      \multicolumn{1}{c}{}& 
+      \multicolumn{1}{c}{$SP_N$ Order}& 
+      \multicolumn{1}{c}{}& 
+      \multicolumn{1}{c}{} \\
+       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
+       & \textbf{0} & 0.7283 & 0.81 & 1.47 & 2.1446 \\
+       & \textbf{1} & 0.7317 & 0.8 & 1.46 & 2.1368 \\
+      $P_N$ Order & \textbf{3} & 0.7317 & 0.91 & 1.526 & 2.2274 \\
+       & \textbf{5} & 0.7317 & 0.91 & 1.5344 & 2.2562 \\
+       & \textbf{7} & 0.7317 & 0.91 & 1.5345 & 1.2842 \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Spectral radius results for the point Jacobi
+      preconditioned iteration matrix with 10 energy groups, full
+      downscatter and full upscatter.}}
+  \label{tab:group10dspj}
+\end{table}
+It is readily apparent from the tabulated data the point Jacobi
+preconditioning is insufficient as a preconditioning mechanism for
+problems of order $SP_5$ or higher. In addition, as upscatter is
+introduced into the system, a much larger spectral radius is observed
+at lower $SP_N$ orders, signaling a need for a better preconditioning
+strategy to both ensure and improve covergence for Monte Carlo
+methods.
 
@@
 \label{subsec:spn_preconditioning}
+If Monte Carlo methods are to be used to solve the $SP_N$ system of
+equations, a different preconditioning strategy is required in order
+to ensure convergence for systems of all $SP_N$ and $P_N$ orders with
+arbitrary energy group structures. To achieve this, we look back to
+the sparsity plots we generated in Figures~\ref{fig:group1},
+\ref{fig:group10ds} and \ref{fig:group10us} as well as the multigroup
+$SP_N$ equations. Initially, the diagonal character of the system led
+us to try point Jacobi preconditioning with only marginal
+results. From the sparsity plots we note the block structure that
+ultimately arises from the multigroup scattering matrices and their
+insertion into Eq~(\ref{eq:A_block_matrix}). When full upscatter and
+downscatter are used the resulting blocks are completely dense while
+only downscatter gives a lower triangular scattering matrix and the
+block structure shown in Figure~\ref{fig:group10ds}.
+
+Based on this both block and diagonally dominant structure for
+matrices formed by the general multigroup $SP_N$ equations, we instead
+choose \textit{block Jacobi} preconditioning as a left preconditioner
+for the system. Like point Jacobi preconditioning, block Jacobi
+preconditioning extracts the diagonal elements of the matrix as the
+preconditioner where now the elements extracted are the blocks on the
+diagonal as shown on the left side of
+Figure~\ref{fig:block_jacobi_ex}.
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=4in]{chapters/spn_equations/block_preconditioning}
+  \end{center}
+  \caption{\textbf{Block Jacobi preconditioning strategy used for the
+      $SP_N$ equations.} \textit{Left: The preconditioner is formed by
+      the diagonal blocks. Right: Inversion is trivial and decoupled
+      by block.}}
+  \label{fig:block_jacobi_ex}
+\end{figure}
+The inversion of this preconditioner is trivial as shown on the right
+side of Figure~\ref{fig:block_jacobi_ex}. Each block can be inverted
+individually and combined to form the inverse. For high performance
+implementations this has several attractive properties. First, the
+blocks in the matrix come from the energy/angle discretization of the
+transport equation. Each block on the diagonal is bound to a mesh
+element in the system (note there are 8 blocks in each of the sparsity
+patterns with a mesh of $2 \times 2 \times 2$) and therefore we expect
+the matrix elements forming the block to be entirely local. Second,
+these blocks are typically dense and nearly lower triangular for many
+transport problems meaning that established dense matrix methods based
+on LU-factorization and other schemes can be used for fast inversion.
 

### 2013-03-26 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
updated block jacobi strategy section in spn equations

@@
       downscatter and full upscatter.}}
-  \label{tab:group10dspj}
+  \label{tab:group10uspj}
 \end{table}
@@
   \begin{center}
-    \includegraphics[width=4in]{chapters/spn_equations/block_preconditioning}
+    \includegraphics[width=5in]{chapters/spn_equations/block_preconditioning.png}
   \end{center}
@@
       $SP_N$ equations.} \textit{Left: The preconditioner is formed by
-      the diagonal blocks. Right: Inversion is trivial and decoupled
-      by block.}}
+      the diagonal blocks of the matrix. Right: Inversion is trivial
+      and decoupled by block.}}
   \label{fig:block_jacobi_ex}
@@
 blocks in the matrix come from the energy/angle discretization of the
-transport equation. Each block on the diagonal is bound to a mesh
-element in the system (note there are 8 blocks in each of the sparsity
-patterns with a mesh of $2 \times 2 \times 2$) and therefore we expect
-the matrix elements forming the block to be entirely local. Second,
-these blocks are typically dense and nearly lower triangular for many
+transport equation as given by Eq~(\ref{eq:A_block_matrix}). Each
+block on the diagonal is bound to a mesh element in the system (note
+there are 8 blocks on the diagonal in each of the sparsity patterns
+with a mesh of $2 \times 2 \times 2$) and therefore we expect the
+matrix elements forming the block to be entirely local. Second, these
+blocks are typically dense and nearly lower triangular for many
 transport problems meaning that established dense matrix methods based
-on LU-factorization and other schemes can be used for fast inversion.
+on LU-factorization and other schemes can be used for fast inversion
+\citep{lapack_citation}.
 

### 2013-03-26 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
finshed add spn spectral analysis to spn equations

@@
        & \textbf{0} & 0.0655 & 0.677 & 1.32 & 1.982 \\
-       & \textbf{1} & 0.071 & 0.67777 & 1.319 & 1.982 \\
+       & \textbf{1} & 0.071 & 0.6777 & 1.319 & 1.982 \\
       $P_N$ Order & \textbf{3} & 0.071 & 0.687 & 1.327 & 1.9872 \\
@@
        & \textbf{5} & 0.7317 & 0.91 & 1.5344 & 2.2562 \\
-       & \textbf{7} & 0.7317 & 0.91 & 1.5345 & 1.2842 \\
+       & \textbf{7} & 0.7317 & 0.91 & 1.5345 & 2.2842 \\
       %%
@@
 diagonal as shown on the left side of
-Figure~\ref{fig:block_jacobi_ex}.
+Figure~\ref{fig:block_jacobi_ex}. For the $SP_N$ equations,
+Eq~(\ref{eq:A_block_matrix}) gives a block size of $N_g\times(N+1)/2$.
 \begin{figure}[t!]
@@
 \label{subsec:spn_analysis_results}
+Spectral radius computations for the block Jacobi preconditioned
+iteration matrix were performed for the same problems as the point
+Jacobi preconditioned case. Table~\ref{tab:group1bj} gives the results
+for the 1-group case, Table~\ref{tab:group10dsbj} for the 10-group
+case with full downscatter only and Table~\ref{tab:group10usbj} for
+the 10-group case with full downscatter and full upscatter. A block
+size of $N_g\times(N+1)/2$ was used for the preconditioner in all
+cases.
+\begin{table}[t!]
+  \begin{center}
+    \begin{tabular}{cccccc}\hline\hline
+      \multicolumn{1}{c}{}& 
+      \multicolumn{1}{c}{}& 
+      \multicolumn{1}{c}{}& 
+      \multicolumn{1}{c}{$SP_N$ Order}& 
+      \multicolumn{1}{c}{}& 
+      \multicolumn{1}{c}{} \\
+       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
+       & \textbf{0} & 0.0635 & 0.1269 & 0.1444 & 0.1513 \\
+       & \textbf{1} & 0.0666 & 0.1315 & 0.1474 & 0.1534 \\
+      $P_N$ Order & \textbf{3} & 0.0666 & 0.1365 & 0.154 & 0.1592 \\
+       & \textbf{5} & 0.0666 & 0.1365 & 0.1562 & 0.163 \\
+       & \textbf{7} & 0.0666 & 0.1365 & 0.1562 & 0.164 \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Spectral radius results for the block Jacobi
+      preconditioned iteration matrix with 1 energy group.}}
+  \label{tab:group1bj}
+\end{table}
+\begin{table}[t!]
+  \begin{center}
+    \begin{tabular}{cccccc}\hline\hline
+      \multicolumn{1}{c}{}& 
+      \multicolumn{1}{c}{}& 
+      \multicolumn{1}{c}{}& 
+      \multicolumn{1}{c}{$SP_N$ Order}& 
+      \multicolumn{1}{c}{}& 
+      \multicolumn{1}{c}{} \\
+       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
+       & \textbf{0} & 0.0647 & 0.1275 & 0.1449 & 0.1514 \\
+       & \textbf{1} & 0.0686 & 0.1338 & 0.1484 & 0.1547 \\
+      $P_N$ Order & \textbf{3} & 0.0687 & 0.1399 & 0.1582 & 0.1625 \\
+       & \textbf{5} & 0.0692 & 0.1399 & 0.1582 & 0.1657 \\
+       & \textbf{7} & 0.0678 & 0.1393 & 0.1624 & 0.166 \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Spectral radius results for the block Jacobi
+      preconditioned iteration matrix with 10 energy groups and full
+      downscatter.}}
+  \label{tab:group10dsbj}
+\end{table}
+\begin{table}[t!]
+  \begin{center}
+    \begin{tabular}{cccccc}\hline\hline
+      \multicolumn{1}{c}{}& 
+      \multicolumn{1}{c}{}& 
+      \multicolumn{1}{c}{}& 
+      \multicolumn{1}{c}{$SP_N$ Order}& 
+      \multicolumn{1}{c}{}& 
+      \multicolumn{1}{c}{} \\
+       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
+       & \textbf{0} & 0.1887 & 0.2267 & 0.2285 & 0.2286 \\
+       & \textbf{1} & 0.4535 & 0.5044 & 0.5045 & 0.5045 \\
+      $P_N$ Order & \textbf{3} & 0.4535 & 0.6453 & 0.6506 & 0.6506 \\
+       & \textbf{5} & 0.4535 & 0.6453 & 0.6802 & 0.6818 \\
+       & \textbf{7} & 0.4535 & 0.6453 & 0.6802 & 0.6927 \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Spectral radius results for the block Jacobi
+      preconditioned iteration matrix with 10 energy groups, full
+      downscatter and full upscatter.}}
+  \label{tab:group10usbj}
+\end{table}
+From the tabulated block Jacobi data it is clear that this is the
+preferred preconditioning choice for all $SP_N$ problems in term of
+Monte Carlo solution methods. All cases were observed to have a
+spectral radius below unity and often significantly smaller than that,
+greatly improving convergence rates over the point Jacobi preconditioned
+problem.

### 2013-03-26 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
adding additional pngs to symmetric spectral analysis for communication
reordering parallel mc chapter

@@
 adjoint method. The parallel aspects of both these components must be
-considered.
+considered. In this chapter we review particle transport methods for
+domain decomposed Monte Carlo and then devise an algorithm based on
+these methods for efficient parallel solutions using the adjoint
+Neumann-Ulam method and MCSA.
 
@@
 decomposed due to memory limitations. This naturally leads us to seek
-parallel algorithms that handle domain decomposition.
+parallel Monte Carlo algorithms that handle domain decomposition.
 
@@
 \citep{wagner_hybrid_2010}. In their work, an extension of Brunner's,
-their scheme employed the similar parallel algorithms for particle
+their scheme employed similar parallel algorithms for particle
 transport but a certain amount of overlap between adjacent domains was
@@
 and presented in Figure~\ref{fig:msod_example}.
-\begin{figure}[htpb!]
+\begin{figure}[t!]
   \begin{center}
@@
 %%---------------------------------------------------------------------------%%
+\section{Fully Asynchronous MSOD Neumann-Ulam Algorithm}
+\label{sec:asynchronous_algorithm}
+We can take much of what was learned from the survey of parallel Monte
+Carlo methods for radiation transport and directly apply it to a
+parallel formulation of our stochastic linear solvers. Direct analogs
+can be derived from these works by noting that the primary difference
+between solving a linear system with Monte Carlo methods and fixed
+source Monte Carlo transport problems is the content of the Markov
+chains that are generated. The transitions represented by these chains
+are bound by probabilities and weights and are initiated by the
+sampling of a source. In the context of transport problems, those
+transitions represent events such as particle scattering and
+absorption with probabilities that are determined by physical data in
+the form of cross sections. For stochastic matrix inversion, those
+transitions represent moving between the equations of the linear
+system (and therefore the physical domain which they represent) and
+their probabilities are defined by the coefficients of those
+equations. Ultimately, we tally the contributions to generate
+expectation values in the desired states as we progress through the
+chains. Therefore, parallel methods for Monte Carlo radiation
+transport can be abstracted and we can use those concepts that apply
+to matrix inversion methods as an initial means of developing a
+parallel Neumann-Ulam-type solver. Based on the results observed in
+the last decade of parallel Monte Carlo development, we can generate
+many practical questions that this type of work could answer.
+
+%%---------------------------------------------------------------------------%%
+\section{Parallel MCSA}
+\label{sec:parallel_mcsa}
+With a parallel adjoint Neumann-Ulam solver implementation, the
+parallel implementation of the MCSA method will be trivial. Recall the
+MCSA iteration procedure outlined in Eq~(\ref{eq:mcsa}). In
+\S~\ref{sec:parallel_krylov_methods} we discussed parallel matrix and
+vector operations as utilized in conventional Krylov methods. We
+utilize these here for the parallel MCSA implementation. In the first
+step, a parallel matrix-vector multiply is used to apply the split
+operator to the previous iterate's solution. A parallel vector update
+is then performed with the source vector to arrive at the initial
+iteration guess. In the next step, the residual is computed by the
+same operations where now the operator is applied to the solution
+guess with a parallel matrix-vector multiply and then a parallel
+vector update with the source vector is performed. Once the correction
+is computed with a parallel adjoint Neumann-Ulam solve, this
+correction is applied to the guess with a parallel vector update to
+get the new iteration solution. Additionally, as given by
+Eq~(\ref{eq:mcsa_stopping_criteria}), 2 parallel vector reductions
+will be required to check the stopping criteria: one initially to
+compute the infinity norm of the source vector, and another at every
+iteration to compute the infinity norm of the residual vector. For
+this implementation, all of the issues that will be potentially
+generated by the parallel adjoint solver implementation will manifest
+themselves here as the quality of the correction will be of intense
+study.
+
+In addition to parallel implementation and performance, MCSA's
+potential for aiding advancement in non-symmetric matrix solutions
+leads to a natural comparison with the GMRES algorithm. As both
+solvers are aimed at the same class of problem, we desire a set of
+metrics that will allow us to quantitatively compare the two. Given
+that the Krylov subspace maintained by GMRES can become large, do we
+benefit from a memory standpoint with an MCSA scheme in that no
+subspace is required? Does this benefit outweigh the fact that the
+linear operator must be explicitly formed in order to build the
+transition probabilities for the random walk sequence? For
+non-symmetric systems, does MCSA exhibit similar convergence
+properties to Krylov methods? If a Krylov methods build a subspace,
+can those memory savings in MCSA be used to implement domain
+replication in the adjoint solver? Such questions can be answered by a
+comparative study of the two solvers that controls the system size and
+the iterations required to converge.
+
+%%---------------------------------------------------------------------------%%
 \section{Load Balancing Concerns}
@@
 from Procassini's work.
-\begin{figure}[htpb!]
+\begin{figure}[t!]
   \begin{center}
@@
 these issues.
-\begin{figure}[htpb!]
+\begin{figure}[t!]
   \begin{center}
@@
 point values.
-
-%%---------------------------------------------------------------------------%%
-\section{Parallel Adjoint Method}
-\label{sec:parallel_adjoint}
-We can take much of what was learned from the survey of parallel Monte
-Carlo methods for radiation transport and directly apply it to a
-parallel formulation of our stochastic linear solvers. Direct analogs
-can be derived from these works by noting that the primary difference
-between solving a linear system with Monte Carlo methods and fixed
-source Monte Carlo transport problems is the content of the Markov
-chains that are generated. The transitions represented by these chains
-are bound by probabilities and weights and are initiated by the
-sampling of a source. In the context of transport problems, those
-transitions represent events such as particle scattering and
-absorption with probabilities that are determined by physical data in
-the form of cross sections. For stochastic matrix inversion, those
-transitions represent moving between the equations of the linear
-system (and therefore the physical domain which they represent) and
-their probabilities are defined by the coefficients of those
-equations. Ultimately, we tally the contributions to generate
-expectation values in the desired states as we progress through the
-chains. Therefore, parallel methods for Monte Carlo radiation
-transport can be abstracted and we can use those concepts that apply
-to matrix inversion methods as an initial means of developing a
-parallel Neumann-Ulam-type solver. Based on the results observed in
-the last decade of parallel Monte Carlo development, we can generate
-many practical questions that this type of work could answer.
-
-Given a decomposed domain, per Wagner's work
-\citep{wagner_hybrid_2010} is clear that domain overlap significantly
-reduces the amount of communication required between adjacent domains
-as histories move to states that are not owned by the local
-processor. How much domain overlap is suitable for matrix inversion
-problems? Are we memory limited by large problems such that only so
-much overlap is feasible? Are there metrics related to the properties
-of the matrix including the eigenvalues and sparsity pattern such that
-we can provide guidelines for selecting the amount of overlap
-required? Furthermore, as Wagner and colleagues observed
-\citep{wagner_hybrid_2010}, with marginal domain overlap for reactor
-problems, the percentage of histories leaving the local domain can
-easily be reduced to less than 1\%. Evans and colleagues in their
-initial MCSA development typically used only 50 particle histories in
-order to compute a correction in an MCSA iteration
-\citep{evans_monte_2012}. Per the Central Limit Theorem, that
-correlates to a statistical uncertainty of 14.1\% in the correction,
-yet as compared to conventional Krylov solvers, MCSA achieved
-identical numerical results. If good convergence and numerically
-accurate solutions can be achieved with such a large uncertainty in
-the correction computation, then perhaps with enough domain overlap
-the minimal amount of histories that do transition to non-local states
-can be ignored and thus eliminate all communication in the parallel
-Monte Carlo adjoint solver transport sequence and create an
-embarrassingly parallel method. Given Gropp's work on parallel Krylov
-methods that we discussed in
-\S~\ref{subsec:projection_method_performance} and the empirical
-results of Siegel \citep{siegel_analysis_2012}, we know that these
-nearest neighbor computations between adjacent domains do not scale as
-well as global reduction operations and therefore we are improving
-scaling by eliminating them from the transport sequence. It will be
-important to determine if the expectation value bias in the MCSA
-correction generated by this approximation will remain within the
-bounds of the already high statistical uncertainty for unbiased
-estimates, thus providing numerically equivalent results.
-
-From a domain replication perspective, this may be difficult to
-achieve with a production scale linear solver. Typically, memory is at
-a premium and therefore the more distinct domains available in the
-decomposition, the spatially finer and/or the numerically more
-accurate the discretization that can be implemented. How much
-replication is possible for large problems? How does replication
-facilitate parallel performance when coupled with domain overlap? How
-can we measure how much domain overlap is feasible? Replicating
-domains may therefore run into memory limitations for exceptionally
-large problems such that the operator, solution vector, and source
-vector must be copied in their entirety multiple times. From a
-resiliency standpoint, such an operation will be required, and
-therefore its performance and memory implications on conventional
-problems must be analyzed.
-
-%%---------------------------------------------------------------------------%%
-\section{Parallel MCSA}
-\label{sec:parallel_mcsa}
-With a parallel adjoint Neumann-Ulam solver implementation, the
-parallel implementation of the MCSA method will be trivial. Recall the
-MCSA iteration procedure outlined in Eq~(\ref{eq:mcsa}). In
-\S~\ref{sec:parallel_krylov_methods} we discussed parallel matrix and
-vector operations as utilized in conventional Krylov methods. We
-utilize these here for the parallel MCSA implementation. In the first
-step, a parallel matrix-vector multiply is used to apply the split
-operator to the previous iterate's solution. A parallel vector update
-is then performed with the source vector to arrive at the initial
-iteration guess. In the next step, the residual is computed by the
-same operations where now the operator is applied to the solution
-guess with a parallel matrix-vector multiply and then a parallel
-vector update with the source vector is performed. Once the correction
-is computed with a parallel adjoint Neumann-Ulam solve, this
-correction is applied to the guess with a parallel vector update to
-get the new iteration solution. Additionally, as given by
-Eq~(\ref{eq:mcsa_stopping_criteria}), 2 parallel vector reductions
-will be required to check the stopping criteria: one initially to
-compute the infinity norm of the source vector, and another at every
-iteration to compute the infinity norm of the residual vector. For
-this implementation, all of the issues that will be potentially
-generated by the parallel adjoint solver implementation will manifest
-themselves here as the quality of the correction will be of intense
-study.
-
-In addition to parallel implementation and performance, MCSA's
-potential for aiding advancement in non-symmetric matrix solutions
-leads to a natural comparison with the GMRES algorithm. As both
-solvers are aimed at the same class of problem, we desire a set of
-metrics that will allow us to quantitatively compare the two. Given
-that the Krylov subspace maintained by GMRES can become large, do we
-benefit from a memory standpoint with an MCSA scheme in that no
-subspace is required? Does this benefit outweigh the fact that the
-linear operator must be explicitly formed in order to build the
-transition probabilities for the random walk sequence? For
-non-symmetric systems, does MCSA exhibit similar convergence
-properties to Krylov methods? If a Krylov methods build a subspace,
-can those memory savings in MCSA be used to implement domain
-replication in the adjoint solver? Such questions can be answered by a
-comparative study of the two solvers that controls the system size and
-the iterations required to converge.

### 2013-03-26 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
outlining sections for explanation of asynchronous MSOD algorithm

@@
 adjoint method. The parallel aspects of both these components must be
-considered. In this chapter we review particle transport methods for
-domain decomposed Monte Carlo and then devise an algorithm based on
-these methods for efficient parallel solutions using the adjoint
-Neumann-Ulam method and MCSA.
+considered. In this chapter we briefly review particle transport
+methods for domain decomposed Monte Carlo and then devise an algorithm
+based on these methods for efficient parallel solutions using the
+adjoint Neumann-Ulam method and MCSA.
 
@@
 \label{sec:asynchronous_algorithm}
-We can take much of what was learned from the survey of parallel Monte
-Carlo methods for radiation transport and directly apply it to a
-parallel formulation of our stochastic linear solvers. Direct analogs
-can be derived from these works by noting that the primary difference
-between solving a linear system with Monte Carlo methods and fixed
-source Monte Carlo transport problems is the content of the Markov
-chains that are generated. The transitions represented by these chains
-are bound by probabilities and weights and are initiated by the
-sampling of a source. In the context of transport problems, those
-transitions represent events such as particle scattering and
-absorption with probabilities that are determined by physical data in
-the form of cross sections. For stochastic matrix inversion, those
-transitions represent moving between the equations of the linear
-system (and therefore the physical domain which they represent) and
-their probabilities are defined by the coefficients of those
-equations. Ultimately, we tally the contributions to generate
+We can take much of what was learned from Brunner and Wagner's
+parallel Monte Carlo methods for radiation transport and directly
+apply it to a parallel formulation of our stochastic linear
+solvers. Direct analogs can be derived from these works by noting that
+the primary difference between solving a linear system with Monte
+Carlo methods and fixed source Monte Carlo transport problems is the
+content of the Markov chains that are generated. The transitions
+represented by these chains are bound by probabilities and weights and
+are initiated by the sampling of a source. In the context of transport
+problems, those transitions represent events such as particle
+scattering and absorption with probabilities that are determined by
+physical data in the form of cross sections. For stochastic matrix
+inversion, those transitions represent moving between the equations of
+the linear system (and therefore the physical domain which they
+represent) and their probabilities are defined by the coefficients of
+those equations. Ultimately, we tally the contributions to generate
 expectation values in the desired states as we progress through the
@@
 to matrix inversion methods as an initial means of developing a
-parallel Neumann-Ulam-type solver. Based on the results observed in
-the last decade of parallel Monte Carlo development, we can generate
-many practical questions that this type of work could answer.
+parallel Neumann-Ulam-type solver.
+
+\subsection{Parallel Random Number Generation}
+\label{subsec:parallel_rng}
+
+\subsection{Generating the Transport Domain with Overlap and Preconditioning}
+\label{subsec:domain_generation}
+
+\subsection{Generating the Transport Source with Preconditioning}
+\label{subsec:source_generation}
+
+\subsection{Transition Processing Kernel}
+\label{subsec:transition_processing}
+
+\subsection{Local Domain Transport Kernel}
+\label{subsec:local_domain_transport}
+
+\subsection{Domain-to-Domain Communication Kernel}
+\label{subsec:domain_to_domain_kernel}
+
+\subsection{Asynchronous Transport Kernel}
+\label{subsec:async_transport_kernel}
+
+\subsection{Multiple Set Algorithm}
+\label{subsec:multiple_sets_algorithm}
 

### 2013-03-26 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
starting work on parallel rngs but I need a few references for this

@@
 to matrix inversion methods as an initial means of developing a
-parallel Neumann-Ulam-type solver.
+parallel Neumann-Ulam-type solver. In this section the fully
+asynchronous MSOD Neumann-Ulam algorithm developed for this work is
+presented step-by-step in detail.
 
@@
 \label{subsec:parallel_rng}
+Correct parallel Monte Carlo requires parallel random number
+generation. To ensure distinct Markov chains for each history in the
+problem, regardless of its state in phase space or parallel space, a
+unique stream of random numbers is necessary to achieve uncorrelated
+tally results for the solution.
 
@@
 \label{sec:parallel_mcsa}
-With a parallel adjoint Neumann-Ulam solver implementation, the
-parallel implementation of the MCSA method will be trivial. Recall the
-MCSA iteration procedure outlined in Eq~(\ref{eq:mcsa}). In
-\S~\ref{sec:parallel_krylov_methods} we discussed parallel matrix and
-vector operations as utilized in conventional Krylov methods. We
-utilize these here for the parallel MCSA implementation. In the first
-step, a parallel matrix-vector multiply is used to apply the split
-operator to the previous iterate's solution. A parallel vector update
-is then performed with the source vector to arrive at the initial
-iteration guess. In the next step, the residual is computed by the
-same operations where now the operator is applied to the solution
-guess with a parallel matrix-vector multiply and then a parallel
-vector update with the source vector is performed. Once the correction
-is computed with a parallel adjoint Neumann-Ulam solve, this
-correction is applied to the guess with a parallel vector update to
-get the new iteration solution. Additionally, as given by
+With the parallel adjoint Neumann-Ulam solver implementation described
+aboive, the parallel implementation of the MCSA method is
+trivial. Recall the MCSA iteration procedure outlined in
+Eq~(\ref{eq:mcsa}). In \S\ref{sec:parallel_krylov_methods} we
+discussed parallel matrix and vector operations as utilized in
+conventional Krylov methods. We utilize these here for the parallel
+MCSA implementation. In the first step, a parallel matrix-vector
+multiply is used to apply the split operator to the previous iterate's
+solution. A parallel vector update is then performed with the source
+vector to arrive at the initial iteration guess. In the next step, the
+residual is computed by the same operations where now the operator is
+applied to the solution guess with a parallel matrix-vector multiply
+and then a parallel vector update with the source vector is
+performed. Once the correction is computed with a parallel adjoint
+Neumann-Ulam solve, this correction is applied to the guess with a
+parallel vector update to get the new iteration
+solution. Additionally, as given by
 Eq~(\ref{eq:mcsa_stopping_criteria}), 2 parallel vector reductions
@@
 study.
-
-In addition to parallel implementation and performance, MCSA's
-potential for aiding advancement in non-symmetric matrix solutions
-leads to a natural comparison with the GMRES algorithm. As both
-solvers are aimed at the same class of problem, we desire a set of
-metrics that will allow us to quantitatively compare the two. Given
-that the Krylov subspace maintained by GMRES can become large, do we
-benefit from a memory standpoint with an MCSA scheme in that no
-subspace is required? Does this benefit outweigh the fact that the
-linear operator must be explicitly formed in order to build the
-transition probabilities for the random walk sequence? For
-non-symmetric systems, does MCSA exhibit similar convergence
-properties to Krylov methods? If a Krylov methods build a subspace,
-can those memory savings in MCSA be used to implement domain
-replication in the adjoint solver? Such questions can be answered by a
-comparative study of the two solvers that controls the system size and
-the iterations required to converge.
 

### 2013-04-27 doc/dissertation/document/chapters/mc_background/mc_background.tex
updates to spn equations introduction
refactoring the monte carlo chapters and adding stubs for new sections to be completed

@@
 using a collection of references from the 1950's and early 1960's.
+
+In this chapter, we will present the fundamentals of the Monte Carlo
+method for discrete linear systems. Both the forward and adjoint
+methods will be presented and analyzed including a variance analysis
+of several Monte Carlo estimators. The Sequential Monte Carlo and
+Monte Carlo Synthetic Acceleration methods will then be presented as a
+means of leveraging the basic Monte Carlo methods in a iterative
+refinement scheme that accelerates their convergence. Using these
+iterative schemes, a set of modest variance reduction schemes will be
+introduced and their effects on the Monte Carlo methods explored.
 
@@
 total. The contribution to the solution from a particular random walk
-permutation is then:
+permutation is then the \textit{terminal estimator}:
 \begin{equation}
@@
 shown in Figure~\ref{fig:heat_setup}.
-\begin{figure}[h!]
+\begin{figure}[t!]
   \begin{center}
@@
 gives these results.
-\begin{figure}[h!]
+\begin{figure}[t!]
   \begin{center}
@@
 traditional iterative method.
-
-\subsection{Estimator Variance}
-\label{subsec:estimator_variance}
-We can compute the variance of the estimator through traditional
-methods by defining the variance, $\sigma_i$, for each component in
-the solution:
-\begin{equation}
-  {\sigma_i}^2 = E\{X(i_0 = i) - (\ve{A}^{-1}\ve{b})_i\}^2 = E\{X(i_0
-  = i)^2\} - x_i^2\:,
-  \label{eq:direct_variance_1}
-\end{equation}
-where the vector exponentials are computed element-wise. Inserting
-Eq~(\ref{eq:direct_expectation_value}) gives:
-\begin{equation}
-  \sigma_i^2 = \Big(\sum_{\nu} P_{\nu} X_{\nu}^2\Big)_i - x_i^2\:,
-  \label{eq:direct_variance_2}
-\end{equation}
-and applying Eq~(\ref{eq:direct_permutation_contribution}):
-\begin{equation}
-  \sigma_i^2 = \Big(\sum_{\nu} P_{\nu} \sum_{m=0}^k W_{m}^2
-  b_{i_m}^2\Big)_i - x_i^2\:.
-  \label{eq:direct_variance_3}
-\end{equation}
-Finally, expanding the transition probabilities yields the variance:
-\begin{equation}
-  \sigma_i^2 = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-  \sum_{i_k}^{N} p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
-  w^2_{i,i_1}w^2_{i_1,i_2}\ldots w^2_{i_{k-1},i_k} b_{i_m} - x_i^2\:.
-  \label{eq:direct_variance_4}
-\end{equation}
-Using this definition for the variance, we can arrive at a more natural
-reason for enforcing $\rho(\ve{H}) < 1$ for our Monte Carlo method to
-converge. Per the Hadamard product, we can concatenate the summation
-in Eq~(\ref{eq:direct_variance_4}):
-\begin{equation}
-  (\ve{P} \circ \ve{W} \circ \ve{W})^k =
-  \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots \sum_{i_k}^{N}
-  p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
-  w^2_{i,i_1}w^2_{i_1,i_2}\ldots w^2_{i_{k-1},i_k}\:.
-  \label{eq:double_weighted_decomposition}
-\end{equation}
-If we assign $\ve{G} = \ve{P} \circ \ve{W} \circ \ve{W}$ as in
-Eq~(\ref{eq:neumann_ulam_decomposition}), we then have a new
-formulation for the variance:
-\begin{equation}
-  \sigma^2_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-  \sum_{i_k}^{N}g_{i,i_1}g_{i_1,i_2}\ldots g_{i_{k-1},i_k} b_{i_k}^2 -
-  x_i^2\:,
-\end{equation}
-which contains the general Neumann series for $\ve{G}$,
-\begin{equation}
-  \ve{T} = \sum_{k=0}^{\infty} \ve{G}^k\:,
-  \label{eq:variance_neumann_series}
-\end{equation}
-where $\ve{T} = (\ve{I}-\ve{G})^{-1}$. We can then insert $T$ back
-into the variance formulation for a more concise definition:
-\begin{equation}
-  \sigma^2_i = (\ve{T}\ve{b})_i - x_i^2\:.
-  \label{eq:direct_variance_5}
-\end{equation}
-We can relate $\ve{G}$ to $\ve{H}$ by noting that $\ve{G}$ simply
-contains an additional Hadamard product with the weight matrix. The
-Hadamard product has the property that:
-\begin{equation}
-  |\ve{H} \circ \ve{W}| \geq |\ve{H}|\ |\ve{W}|\:.
-  \label{eq:hadamard_inequality}
-\end{equation}
-Using the norm property of the Hadamard product and
-Eq~(\ref{eq:neumann_ulam_decomposition}), we can define the norm of
-$\ve{W}$ as:
-\begin{equation}
-  \frac{|\ve{H}|}{|\ve{P}|} \geq |\ve{W}|\:.
-  \label{eq:weight_norm}
-\end{equation}
-Choosing the infinity norm of the operator as defined in
-Eq~(\ref{eq:matrix_infinity_norm}), the row normalized probability
-matrix will yield a norm of 1 giving the following inequality for
-relating $\ve{G}$ and $\ve{H}$:
-\begin{equation}
-  |\ve{G}| \geq |\ve{H}|^2
-  \label{eq:variance_norm_inequality}
-\end{equation}
-Using these relations to analyze Eq~(\ref{eq:direct_variance_5}), we
-see that if $\rho(\ve{G}) > 1$, then the sum in
-Eq~(\ref{eq:variance_neumann_series}) will not converge and an
-infinite variance will arise as the elements of $\ve{T}$ become
-infinite in Eq~(\ref{eq:direct_variance_5}). We must restrict $\ve{G}$
-to alleviate this and therefore restrict $\ve{H}$ due to
-Eq~(\ref{eq:variance_norm_inequality}) with $\rho(\ve{H}) < 1$ so that
-our expectation values for the solution may have a finite variance.
 
@@
 adjoint constraints, the contribution to the solution in state $i$
-from a particular random walk permutation is then:
+from a particular random walk permutation is then the
+\textit{collision estimator}:
 \begin{equation}
@@
 
+\subsection{Expected Value Estimator}
+\label{subsec:expected_value_estimator}
+Here we introduce an estimator based on the method of expected values.
+
 \subsection{Adjoint Method: Evolution of a Solution}
@@
 method. Figure~\ref{fig:adjoint_evolution} gives these results.
-\begin{figure}[h!]
+\begin{figure}[t!]
   \begin{center}
@@
 contributions to those tallies.
+
+%%---------------------------------------------------------------------------%%
+\section{Estimator Variance}
+\label{subsec:estimator_variance}
+
+\subsection{Terminal Estimator Variance}
+\label{subsec:terminal_variance}
+We can compute the variance of the estimators through traditional
+methods by defining the variance, $\sigma_i$, for each component in
+the solution:
+\begin{equation}
+  {\sigma_i}^2 = E\{X(i_0 = i) - (\ve{A}^{-1}\ve{b})_i\}^2 = E\{X(i_0
+  = i)^2\} - x_i^2\:,
+  \label{eq:direct_variance_1}
+\end{equation}
+where the vector exponentials are computed element-wise. Inserting
+Eq~(\ref{eq:direct_expectation_value}) gives:
+\begin{equation}
+  \sigma_i^2 = \Big(\sum_{\nu} P_{\nu} X_{\nu}^2\Big)_i - x_i^2\:,
+  \label{eq:direct_variance_2}
+\end{equation}
+and applying Eq~(\ref{eq:direct_permutation_contribution}):
+\begin{equation}
+  \sigma_i^2 = \Big(\sum_{\nu} P_{\nu} \sum_{m=0}^k W_{m}^2
+  b_{i_m}^2\Big)_i - x_i^2\:.
+  \label{eq:direct_variance_3}
+\end{equation}
+Finally, expanding the transition probabilities yields the variance:
+\begin{equation}
+  \sigma_i^2 = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
+  \sum_{i_k}^{N} p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
+  w^2_{i,i_1}w^2_{i_1,i_2}\ldots w^2_{i_{k-1},i_k} b_{i_m} - x_i^2\:.
+  \label{eq:direct_variance_4}
+\end{equation}
+Using this definition for the variance, we can arrive at a more natural
+reason for enforcing $\rho(\ve{H}) < 1$ for our Monte Carlo method to
+converge. Per the Hadamard product, we can concatenate the summation
+in Eq~(\ref{eq:direct_variance_4}):
+\begin{equation}
+  (\ve{P} \circ \ve{W} \circ \ve{W})^k =
+  \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots \sum_{i_k}^{N}
+  p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
+  w^2_{i,i_1}w^2_{i_1,i_2}\ldots w^2_{i_{k-1},i_k}\:.
+  \label{eq:double_weighted_decomposition}
+\end{equation}
+If we assign $\ve{G} = \ve{P} \circ \ve{W} \circ \ve{W}$ as in
+Eq~(\ref{eq:neumann_ulam_decomposition}), we then have a new
+formulation for the variance:
+\begin{equation}
+  \sigma^2_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
+  \sum_{i_k}^{N}g_{i,i_1}g_{i_1,i_2}\ldots g_{i_{k-1},i_k} b_{i_k}^2 -
+  x_i^2\:,
+\end{equation}
+which contains the general Neumann series for $\ve{G}$,
+\begin{equation}
+  \ve{T} = \sum_{k=0}^{\infty} \ve{G}^k\:,
+  \label{eq:variance_neumann_series}
+\end{equation}
+where $\ve{T} = (\ve{I}-\ve{G})^{-1}$. We can then insert $T$ back
+into the variance formulation for a more concise definition:
+\begin{equation}
+  \sigma^2_i = (\ve{T}\ve{b})_i - x_i^2\:.
+  \label{eq:direct_variance_5}
+\end{equation}
+We can relate $\ve{G}$ to $\ve{H}$ by noting that $\ve{G}$ simply
+contains an additional Hadamard product with the weight matrix. The
+Hadamard product has the property that:
+\begin{equation}
+  |\ve{H} \circ \ve{W}| \geq |\ve{H}|\ |\ve{W}|\:.
+  \label{eq:hadamard_inequality}
+\end{equation}
+Using the norm property of the Hadamard product and
+Eq~(\ref{eq:neumann_ulam_decomposition}), we can define the norm of
+$\ve{W}$ as:
+\begin{equation}
+  \frac{|\ve{H}|}{|\ve{P}|} \geq |\ve{W}|\:.
+  \label{eq:weight_norm}
+\end{equation}
+Choosing the infinity norm of the operator as defined in
+Eq~(\ref{eq:matrix_infinity_norm}), the row normalized probability
+matrix will yield a norm of 1 giving the following inequality for
+relating $\ve{G}$ and $\ve{H}$:
+\begin{equation}
+  |\ve{G}| \geq |\ve{H}|^2
+  \label{eq:variance_norm_inequality}
+\end{equation}
+Using these relations to analyze Eq~(\ref{eq:direct_variance_5}), we
+see that if $\rho(\ve{G}) > 1$, then the sum in
+Eq~(\ref{eq:variance_neumann_series}) will not converge and an
+infinite variance will arise as the elements of $\ve{T}$ become
+infinite in Eq~(\ref{eq:direct_variance_5}). We must restrict $\ve{G}$
+to alleviate this and therefore restrict $\ve{H}$ due to
+Eq~(\ref{eq:variance_norm_inequality}) with $\rho(\ve{H}) < 1$ so that
+our expectation values for the solution may have a finite variance.
+
+\subsection{Collision Estimator Variance}
+\label{subsec:collision_variance}
+
+\subsection{Expected Value Estimator Variance}
+\label{subsec:expected_value_variance}
 
@@
 regardless of any other preconditioning that is applied to the system.
+
+\subsection{General Preconditioning Strategies}
+\label{subsec:general_mcsa_strategies}
+It is possible to use general left, right, and left/right
+preconditioning with MCSA by carefully considering the underly Monte
+Carlo that will be solved with the Neumann Ulam method.
 
@@
 \end{figure}
+
+%%---------------------------------------------------------------------------%%
+\section{Variance Reduction Techniques}
+\label{sec:variance_reduction}
+
+\subsection{Artificial Absorption}
+\label{subsec:artificial_absorption}
+
+\subsection{Reduced Domain Approximation}
+\label{subsec:reduced_domain_approximation}

### 2013-04-27 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
more dissertation refactoring

@@
-\chapter{Monte Carlo Solution Methods for Nonlinear Systems}
+\chapter{Monte Carlo Solution Methods Applied to Nonlinear Systems}
 \label{ch:nonlinear_problem}
@@
 thereof.
+
+%%---------------------------------------------------------------------------%%
+\section{Monte Carlo Solution Methods for the Navier-Stokes Equations}
+\label{sec:mc_ns_solutions}
+
+\subsection{FANM Verification}
+\label{subsec:fanm_verification}
+To verify the FANM method for nonlinear problems, we choose benchmark
+solutions for the 2-dimensional, steady, incompressible Navier-Stokes
+equations on a rectilinear grid in much the same way as Shadid and
+Pawlowski's work on Newton-Krylov methods for the solution of these
+equations \citep{shadid_inexact_1997,pawlowski_globalization_2006}. We
+define these equations as follows:
+\begin{subequations}
+  \begin{gather}
+    \rho \ve{u} \cdot \nabla \ve{u} - \nabla \cdot \ve{T} - \rho
+    \ve{g} = \ve{0}
+    \label{eq:ns_momentum}\\
+    \nabla \cdot \ve{u} = 0
+    \label{eq:ns_continuity}\\
+    \rho C_p \ve{u} \cdot \nabla T + \nabla \cdot \ve{q} = 0\:,
+    \label{eq:ns_energy}
+  \end{gather}
+  \label{eq:navier_stokes}
+\end{subequations}
+where $\rho$ is the fluid density, $\ve{u}$ is the fluid velocity,
+$\ve{g}$ gravity, $C_p$ the specific heat capacity at constant
+pressure of the fluid and $T$ the temperature of the
+fluid. Eq~(\ref{eq:ns_momentum}) provides momentum transport,
+Eq~(\ref{eq:ns_continuity}) provides the mass balance, and
+Eq~(\ref{eq:ns_energy}) provides energy transport with viscous
+dissipation effects neglected. In addition, we close the system with
+the following equations:
+\begin{subequations}
+  \begin{gather}
+    \ve{T} = -P \ve{I} + \mu[\nabla \ve{u} + \nabla \ve{u}^T]
+    \label{eq:ns_stress_tensor}\\
+    \ve{q} = - k \nabla T\:,
+    \label{eq:ns_heat_flux}
+  \end{gather}
+  \label{eq:ns_closure}
+\end{subequations}
+where $\ve{T}$ is the stress tensor, $P$ is the hydrodynamic pressure,
+$\mu$ is the dynamic viscosity of the fluid, $\ve{q}$ is the heat flux
+in the fluid, and $k$ is the thermal conductivity of the fluid. This
+set of strongly coupled equations possesses both the nonlinearities
+and asymmetries that we are seeking for qualification of the FANM
+method. Further, physical parameters within these equations can be
+tuned to enhance the nonlinearities. We will then apply these
+equations to the following three standard benchmark problems.
+
+\subsubsection{Thermal Convection Cavity Problem}
+\label{subsubsec:natural_convection_cavity}
+In 1983 a benchmark solution for the natural convection of air in a
+square cavity was published \citep{de_vahl_davis_natural_1983} as
+shown in Figure~\ref{fig:natural_convection_cavity} for the solution
+of the energy, mass, and momentum equations.
+\begin{figure}[t!]
+  \begin{center}
+    \scalebox{1.5}{
+      \input{chapters/navier_stokes_solutions/natural_convection_cavity.pdftex_t} }
+  \end{center}
+  \caption{\textbf{Problem setup for the natural convection cavity
+      benchmark.} \textit{Dirichlet conditions are set for the
+      temperature on the left and right while Neumann conditions are
+      set on the top and bottom of the Cartesian grid. The temperature
+      gradients will cause buoyancy-driven flow. Zero velocity
+      Dirichlet conditions are set on each boundary. No thermal source
+      was present.}}
+  \label{fig:natural_convection_cavity}
+\end{figure}
+In this problem, a rectilinear grid is applied to the unit square. No
+heat flow is allowed out of the top and bottom of the square with a
+zero Neumann condition specified. Buoyancy driven flow is generated by
+the temperature gradient from the cold and hot Dirichlet conditions on
+the left and right boundaries of the box. By adjusting the Rayleigh
+number of the fluid (and therefore adjusting the ratio of convective
+to conductive heat transfer), we can adjust the influence of the
+nonlinear convection term in Eq~(\ref{eq:ns_momentum}). In Shadid's
+work, Rayleigh numbers of up to \sn{1}{6} were used for this benchmark
+on a $100 \times 100$ square mesh grid.
+
+\subsubsection{Lid Driven Cavity Problem}
+\label{subsubsec:lid_driven_cavity}
+As an extension of the convection problem, the second benchmark
+problem given by Ghia \citep{ghia_high-re_1982} adds a driver for the
+flow to introduce higher Reynolds numbers into the system, providing
+more inertial force to overcome the viscous forces in the fluid. The
+setup for this problem is equally simple, containing only the
+Dirichlet conditions as given in Figure~\ref{fig:lid_driven_cavity}
+and is only applied to the mass and momentum equations on the unit
+square.
+\begin{figure}[t!]
+  \begin{center}
+    \scalebox{1.5}{
+      \input{chapters/navier_stokes_solutions/lid_driven_cavity.pdftex_t} }
+  \end{center}
+  \caption{\textbf{Problem setup for the lid driven cavity benchmark.}
+    \textit{Dirichlet conditions of zero are set for the velocity on
+      the left and right and bottom while the Dirichlet condition set
+      on the top provides a driving force on the fluid.}}
+  \label{fig:lid_driven_cavity}
+\end{figure}
+The top boundary condition will provide a driver for the flow and its
+variation will in turn vary the Reynolds number of the fluid. An
+increased velocity will generate more inertial forces in the fluid,
+which will overcome the viscous forces and again increase the
+influence of the nonlinear terms in Eq~(\ref{eq:ns_momentum}). Shadid
+also used a $100 \times 100$ grid for this benchmark problem with
+Reynolds numbers up to \sn{1}{4}.
+
+\subsubsection{Backward-Facing Step Problem}
+\label{subsubsec:backward_facing_step}
+The third benchmark was generated by Gartling in 1990 and consists of
+both flow over a backward step and an outflow boundary condition
+\citep{gartling_test_1990}. Using the mass and momentum equations
+while neglecting the energy equation, this problem utilizes a longer
+domain with a 1/30 aspect ratio with the boundary conditions as shown
+in Figure~\ref{fig:backward_facing_step}.
+\begin{figure}[t!]
+  \begin{center}
+    \scalebox{1.3}{
+      \input{chapters/navier_stokes_solutions/backward_facing_step.pdftex_t} }
+  \end{center}
+  \caption{\textbf{Problem setup for the backward facing step
+      benchmark.} \textit{Zero velocity boundary conditions are
+      applied at the top and bottom of the domain while the outflow
+      boundary condition on the right boundary is represented by zero
+      stress tensor components in the direction of the flow. For the
+      inlet conditions, the left boundary is split such that the top
+      half has a fully formed parabolic flow profile and the bottom
+      half has a zero velocity condition, simulating flow over a
+      step.}}
+  \label{fig:backward_facing_step}
+\end{figure}
+In this problem, the inflow condition is specified by a fully-formed
+parabolic flow profile over a zero velocity boundary representing a
+step. The flow over this step will generate a recirculating backflow
+under the inlet flow towards the step. As in the lid driven cavity
+problem, the nonlinear behavior of this benchmark and the difficulty
+in obtaining a solution is dictated by the Reynolds number of the
+fluid. In Shadid's work, a $20 \times 400$ non-square rectilinear grid
+was used to discretize the domain with Reynolds number up to
+\sn{5}{2}.

### 2013-04-27 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
refactoring the monte carlo chapters and adding stubs for new sections to be completed
more dissertation refactoring

@@
 \label{ch:parallel_methods}
-
 For Monte Carlo methods, and in particular MCSA, to be viable at the
@@
 adjoint method. The parallel aspects of both these components must be
-considered. In this chapter we briefly review particle transport
-methods for domain decomposed Monte Carlo and then devise an algorithm
-based on these methods for efficient parallel solutions using the
-adjoint Neumann-Ulam method and MCSA.
+considered.
+
+In this chapter we briefly review particle transport
+methods for domain decomposed Monte Carlo and then provide an analytic
+framework from which to estimate its performance. We then devise an
+algorithm based on these methods for efficient parallel solutions
+using the adjoint Neumann-Ulam method and MCSA and perform parallel
+scaling studings to test its performance and compare this performance
+to conventional methods.
 
@@
 today's modern machines.
+
+%%---------------------------------------------------------------------------%%
+\section{An Analytic Performance Framework for Domain-Decomposed Monte Carlo}
+\label{sec:analytic_framework}
+With Wagner's work \citep{wagner_hybrid_2010}, we observed that domain
+overlap implemented with a domain decomposed Monte Carlo setting
+reduces the amount of nearest neighbor communication that must occur
+during transport. To date, parallel Neumann-Ulam methods have been
+limited to full domain replication with parallelism exploited through
+individual histories \citep{alexandrov_efficient_1998} and in the
+previous chapter we exploited Wagner's work to alleviate this. To
+accomplish this, we recognized from the literature that stochastic
+histories must be transported from domain to domain as the simulation
+progresses and they transition to states that are not in the local
+domain. Because we have chosen a domain decomposition strategy in a
+parallel environment, this means that communication of these histories
+must occur between compute nodes owning neighboring pieces of the
+global domain as outlined in our algorithms. We wish to characterize
+this communication not only because communication is in general
+expensive, but also because these nearest-neighbor communication
+sequences, specifically, have poor algorithmic strong scaling
+\citep{gropp_high-performance_2001} in much the same way as a
+matrix-vector multiply operation. Therefore, we desire a framework to
+provide a simple, analytic theory based on the properties of the
+linear system that will allow for estimates of the domain decomposed
+behavior of the adjoint Neumann-Ulam method in terms of communication
+costs.
+
+\subsection{Leakage Fractions for Symmetric Systems}
+\label{sec:symmetric_leakage_fractions}
+When solving problems where the linear operator is symmetric, a host
+of analytic theories exist based on the eigenvalue spectrum of the
+operator that characterize their behavior in the context of
+deterministic linear solvers. Using past work, these theories are
+adapted to the domain decomposed adjoint Neumann-Ulam method using the
+one-speed, two-dimensional neutron diffusion equation. In this subsection
+we use a one-speed two-dimensional neutron diffusion system as the
+symmetric model problem. Using the linear system generated by the
+discretization of the model problem, we use a spectral analysis to
+generate analytic relations for the eigenvalues of the operator based
+on system parameters. Using the eigenvalue spectra, we then build
+relationships to characterize the transport of stochastic histories in
+a decomposed domain and the fraction of histories that leak from a
+domain and will therefore have to be communicated. Finally, we compare
+these analytic results to numerical experiments conducted with the
+model problem.
+
+\subsubsection{Symmetric Model Problem}
+\label{subsec:model_problem}
+For our numerical experiments, we choose the one-speed,
+two-dimensional neutron diffusion equation as a model problem
+\citep{duderstadt_nuclear_1976}:
+\begin{equation}
+  -\boldsymbol{\nabla} \cdot D \boldsymbol{\nabla} \phi + \Sigma_a
+  \phi = S\:,
+  \label{eq:diffusion_eq}
+\end{equation}
+where $\phi$ is the neutron flux, $\Sigma_a$ is the absorption cross
+subsection, and $S$ is the source of neutrons. In addition, $D$ is the
+diffusion coefficient defined as:
+\begin{equation}
+  D = \frac{1}{3 ( \Sigma_t - \bar{\mu}\Sigma_s )}\:,
+  \label{eq:diffusion_coeff}
+\end{equation}
+where $\Sigma_s$ is the scattering cross subsection, $\Sigma_t = \Sigma_a
++ \Sigma_s$ is the total cross subsection, and $\bar{\mu}$ is the cosine
+of the average scattering angle. For simplicity, we will take
+$\bar{\mu} = 0$ for our analysis giving $D=(3 \Sigma_t)^{-1}$. In
+addition, to further simplify we will assume a homogeneous domain such
+that the cross subsections remain constant throughout. Doing this permits
+us to rewrite Eq~(\ref{eq:diffusion_eq}) as:
+\begin{equation}
+  -D \boldsymbol{\nabla}^2 \phi + \Sigma_a \phi = S\:.
+  \label{eq:diffusion_eq_simple}
+\end{equation}
+
+We choose a finite difference scheme on a square Cartesian grid to
+discretize the problem. For the Laplacian, we choose the 9-point
+stencil shown in Figure~\ref{fig:stencil} over a grid of size $h$
+\citep{leveque_finite_2007}:
+\begin{multline}
+  \nabla^2_9\phi = \frac{1}{6h^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j}
+    + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ +
+    \phi_{i-1,j+1} + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20
+    \phi_{i,j}]\:.
+  \label{eq:nine_point_stencil}
+\end{multline}
+\begin{figure}[t!]
+  \begin{center}
+    \scalebox{1.25}{\input{chapters/mc_theory/stencil.pdftex_t}}
+  \end{center}
+  \caption{\textbf{Nine-point Laplacian stencil.}}
+  \label{fig:stencil}
+\end{figure}
+We then have the following linear system to solve:
+\begin{multline}
+  -\frac{1}{6h^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j} + 4
+    \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ + \phi_{i-1,j+1}
+    + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20 \phi_{i,j}] + \Sigma_a
+  \phi_{i,j} = s_{i,j}\:,
+  \label{eq:fd_system}
+\end{multline}
+and in operator form:
+\begin{equation}
+  \ve{D}\boldsymbol{\phi}=\ve{s}\:,
+  \label{eq:operator_system}
+\end{equation}
+where $\ve{D}$ is the diffusion operator, $\ve{s}$ is the source in
+vector form and $\boldsymbol{\phi}$ is the vector of unknown fluxes.
+
+To close the system, a set of boundary conditions is required. In the
+case of a non-reentrant current condition applied to all global
+boundaries of the domain, we choose the formulation of Duderstadt by
+assuming the flux is zero at some ghost point beyond the
+grid. Consider for example the equations on the $i=0$ boundary of the
+domain:
+\begin{multline}
+  -\frac{1}{6h^2}[4 \phi_{-1,j} + 4 \phi_{1,j} + 4 \phi_{0,j-1} +
+    4 \phi_{0,j+1} + \phi_{-1,j-1}\\ + \phi_{-1,j+1} + \phi_{1,j-1} +
+    \phi_{1,j+1} - 20 \phi_{0,j}] + \Sigma_a \phi_{0,j} = s_{0,j}\:.
+  \label{eq:x_min_bnd}
+\end{multline}
+Here we note some terms where $i=-1$ and therefore are representative
+of grid points beyond the boundary of the domain. We set the flux at
+these points to be zero, giving a valid set of equations for the $i=0$
+boundary:
+\begin{multline}
+  -\frac{1}{6h^2}[4 \phi_{1,j} + 4 \phi_{0,j-1} + 4 \phi_{0,j+1}
+    \\ + \phi_{-1,j+1} + \phi_{1,j-1} + \phi_{1,j+1} - 20 \phi_{0,j}]
+  + \Sigma_a \phi_{0,j} = s_{0,j}\:.
+  \label{eq:x_min_bnd_2}
+\end{multline}
+We repeat this procedure for the other boundaries of the domain. For
+reflecting boundary conditions, the net current across a boundary is
+zero.
+
+\subsubsection{Spectral Analysis}
+\label{subsec:spectral_analysis}
+The convergence of the Neumann series in
+Eq~(\ref{eq:adjoint_neumann_series}) approximated by the Monte Carlo
+solver is dependent on the eigenvalues of the iteration matrix. We
+will compute these eigenvalues by assuming eigenfunctions of the form
+\citep{leveque_finite_2007}:
+\begin{equation}
+  \Phi_{p,q}(x,y) = e^{2 \pi \imath p x} e^{2 \pi \imath q y}\:,
+  \label{eq:eigenfunction_form}
+\end{equation}
+where different combinations of $p$ and $q$ represent the different
+eigenmodes of the solution. As these are valid forms of the solution,
+then the action of the linear operator on these eigenfunctions should
+give the eigenvalues of the matrix as they exist on the unit circle in
+the complex plane.
+
+\subsubsection{Iteration Matrix Spectrum}
+\label{subsec:iteration_spectrum}
+For the model problem, we first compute the eigenvalues for the
+diffusion operator $\ve{D}$ by applying the operator to the
+eigenfunctions and noting that $x=ih$ and $y=jh$:
+\begin{multline}
+  \ve{D}\Phi_{p,q}(x,y) = \lambda_{p,q}(\ve{D})
+  =\\ -\frac{D}{6h^2}\Big[4 e^{-2 \pi \imath p h} + 4 e^{2 \pi \imath
+      p h} + 4 e^{-2 \pi \imath q h} + 4 e^{2 \pi \imath q h} + e^{-2
+      \pi \imath p h} e^{-2 \pi \imath q h} \\ + e^{-2 \pi \imath p h}
+    e^{2 \pi \imath q h} + e^{2 \pi \imath p h} e^{-2 \pi \imath q h}
+    + e^{2 \pi \imath p h} e^{2 \pi \imath q h} - 20\Big] + \Sigma_a
+  \:.
+  \label{eq:deriv_diff_1}
+\end{multline}
+Using Euler's formula, we can collapse the exponentials to
+trigonometric functions:
+\begin{equation}
+  \lambda_{p,q}(\ve{D}) = -\frac{D}{6h^2}[ 8 \cos(\pi p h) + 8
+    \cos(\pi q h) + 4 \cos(\pi p h) \cos(\pi q h) - 20] + \Sigma_a\:.
+  \label{eq:deriv_diff_2}
+\end{equation}
+
+As Eq~(\ref{eq:diffusion_eq}) is diagonally dominant, point Jacobi
+preconditioning as outlined in \S\ref{sec:stochastic_preconditioning}
+is sufficient to reduce the spectral radius of the iteration matrix
+below unity and therefore ensure convergence of the Neumann
+series. Applying this preconditioner, we are then solving the
+following diffusion system:
+\begin{equation}
+  \ve{M}^{-1} \ve{D} \boldsymbol{\phi} = \ve{M}^{-1} \ve{s}\:.
+  \label{eq:precond_diffsion}
+\end{equation}
+The operator $\ve{M}^{-1} \ve{D}$ is merely the original diffusion
+operator with each row scaled by the diagonal component. As we have
+defined a homogeneous domain, the scaling factor, $\alpha$, is the
+same for all rows in the operator and defined as the $\phi_{i,j}$
+coefficient from Eq~(\ref{eq:fd_system}):
+\begin{equation}
+  \alpha = \Bigg[\frac{10 D}{3 h^2} + \Sigma_a\Bigg]^{-1}\:.
+  \label{eq:jacobi_scaling}
+\end{equation}
+Using this coefficient, we then have the following spectrum of
+preconditioned eigenvalues:
+\begin{equation}
+  \lambda_{p,q}(\ve{M}^{-1} \ve{D}) = \alpha \lambda_{p,q}(\ve{D})\:.
+  \label{eq:preconditioned_eigenvalues}
+\end{equation}
+
+The spectral radius of the iteration matrix is obtained by seeking its
+largest eigenvalue. As with the diffusion operator, we can use the
+same analysis techniques to find the eigenvalues for the iteration
+matrix. We use a few simplifications by noting that if the Jacobi
+preconditioned iteration matrix is $\ve{H} = \ve{I} -
+\ve{M}^{-1}\ve{D}$, the we except all terms on the diagonal of the
+iteration matrix to be zero such that we have the following stencil:
+\begin{multline}
+  \ve{H}\boldsymbol{\phi} = \frac{\alpha D}{6h^2}\Big[4 \phi_{i-1,j} +
+    4 \phi_{i+1,j} + 4 \phi_{i,j-1} + 4 \phi_{i,j+1}
+    +\\ \phi_{i-1,j-1} + \phi_{i-1,j+1} + \phi_{i+1,j-1} +
+    \phi_{i+1,j+1}\Big]\:.
+  \label{eq:iteration_stencil}
+\end{multline}
+Inserting the eigenfunctions defined by
+Eq~(\ref{eq:eigenfunction_form}) we get:
+\begin{multline}
+  \lambda_{p,q}(\ve{H}) = \frac{\alpha D}{6h^2}\Big[4 e^{-2 \pi \imath p
+      h} + 4 e^{2 \pi \imath p h} + 4 e^{-2 \pi \imath q h} + 4 e^{2
+      \pi \imath q h} + e^{-2 \pi \imath p h} e^{-2 \pi \imath q h}
+    \\ + e^{-2 \pi \imath p h} e^{2 \pi \imath q q} + e^{2 \pi \imath
+      p h} e^{-2 \pi \imath q h} + e^{2 \pi \imath p h} e^{2 \pi
+      \imath q h}\Big]\:,
+  \label{eq:iteration_deriv}
+\end{multline}
+which simplifies to:
+\begin{equation}
+  \lambda_{p,q}(\ve{H}) = \frac{\alpha D}{6h^2}[ 8 \cos(\pi p h) + 8
+    \cos(\pi q h) + 4 \cos(\pi p h) \cos(\pi q h)]\:,
+  \label{eq:iteration_spectrum}
+\end{equation}
+giving the eigenvalue spectrum for the Jacobi preconditioned iteration
+matrix. To find these maxium eigenvalue,
+Eq~(\ref{eq:preconditioned_eigenvalues}) is plotted as a function of
+$p$ with $p=q$ in Figure~\ref{fig:diffusion_spectrum} for various
+values of $\Sigma_a$.
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=5in,clip]{chapters/mc_theory/diffusion_spectrum.png}
+  \end{center}
+  \caption{\textbf{Eigenvalue spectra for the diffusion equation.}}
+  \label{fig:diffusion_spectrum}
+\end{figure}
+We find that the maximum eigenvalue exists when $p=q=0$, giving the
+following for the spectral radius of the Jacobi preconditioned
+iteration matrix:
+\begin{equation}
+  \rho(\ve{H}) = \frac{10 \alpha D}{3 h^2}\:.
+  \label{eq:iteration_radius}
+\end{equation}
+
+\subsubsection{Neumann Series Convergence}
+\label{subsec:neumann_convergence}
+As outlined in \S\ref{sec:mc_preliminaries}, the adjoint Monte Carlo
+method is effectively an approximation to a stationary method. In the
+adjoint Neumann-Ulam method, $k$ iterations, equivalent to $k$
+applications of the iteration matrix, are approximated by a random
+walk of average length $k$ to yield the summation in
+Eq~(\ref{eq:adjoint_neumann_solution})
+\citep{dimov_new_1998,danilov_asymptotic_2000}. This random walk
+length, or the number of transitions before the termination of a
+history (either by the weight cutoff, absorption, or exiting the
+global domain) is therefore approximately the number of stationary
+iterations required to converge to the specified tolerance. In the
+case of the adjoint Neumann-Ulam method, no such tolerance exists,
+however, we have specified a weight cutoff, $W_c$, that determines
+when low-weight histories will be prematurely terminated as their
+contributions are deemed minute. After $k$ iterations, a stationary
+method is terminated as the error has reached some fraction,
+$\epsilon$, of the initial error:
+\begin{equation}
+  ||\ve{e}^{k}||_2 = \epsilon ||\ve{e}^0||_2\:.
+  \label{eq:linear_k_iter_norm4}
+\end{equation}
+Per Eq~(\ref{eq:linear_k_iter_norm3}), we see that this fraction is
+equivalent to $\epsilon = \rho(\ve{H})^k$. For the adjoint
+Neumann-Ulam method, if we take this fraction to be the weight cutoff,
+a measure of how accurately the contributions of a particular history
+to the solution are tallied, we then have the following relationship
+for $k$:
+\begin{equation}
+  k = \frac{ \log(W_c) }{ \log( \rho(\ve{H}) ) }\:.
+  \label{eq:analytic_k}
+\end{equation}
+This then gives us a means to estimate the length of the random walks
+that will be generated from a particular linear operator based on the
+eigenvalues of its iteration matrix (independent of the linear
+operator splitting chosen) and based on the weight cutoff parameter
+used in the Neumann-Ulam method.
+
+\subsubsection{Domain Leakage Approximations}
+\label{subsubsec:domain_leak_approx}
+In a domain decomposed situation, not all histories will remain within
+the domain they started in and must instead be communicated. This
+communication, expected to be expensive, was analyzed by Siegel and
+colleagues for idealized, load balanced situations for full nuclear
+reactor core Monte Carlo neutral particle simulations
+\citep{siegel_analysis_2012}.  To quantify the number of particles
+that leak out of the local domain they define a leakage fraction,
+$\Lambda$, as:
+\begin{equation}
+  \Lambda = \frac{average\ \#\ of\ particles\ leaving\ local\ domain}
+          {total\ of\ \#\ of\ particles\ starting\ in\ local\ domain}\:.
+          \label{eq:leakage_fraction}
+\end{equation}
+For their studies, Siegel and colleagues assumed that the value of
+$\Lambda$ was dependent on the total cross subsection of the system via
+the Wigner rational approximation. Outlined more thoroughly by Hwang's
+chapter in \citep{azmy_nuclear_2010}, we will use both the Wigner
+rational approximation and the mean chord approximation as a means to
+estimate the leakage fraction.
+
+In the case of domain decomposed linear operator equations, we can use
+diffusion theory to estimate the optical thickness of a domain in the
+decomposition and the corresponding leakage fraction in terms of
+properties of the linear operator and the discretization. To begin we
+must first calculate the mean distance a Monte Carlo history will move
+in the grid by computing the mean squared distance of its movement
+along the chord of length $l$ defined across the domain. After a
+single transition a history will have moved a mean squared distance
+of:
+\begin{equation}
+  \langle \bar{r_1^2} \rangle = (n_s h)^2\:,
+  \label{eq:step_1_length}
+\end{equation}
+where $h$ is the size of the discrete grid elements along the chord
+and $n_s$ is the number of grid elements a history will move on
+average every transition. For our diffusion model problem, $n_s$ would
+equate to the expected number of states in the $i$ (or $j$ as the problem is
+symmetric) direction that a history will move in a single
+transition and is dependent on the stencil used for the
+discretization. After $k$ transitions in the random walk, the history
+will have moved a mean squared distance of:
+\begin{equation}
+  \langle \bar{r_k^2} \rangle = k (n_s h)^2\:.
+  \label{eq:step_k_length}
+\end{equation}
+If our chord is of length $l$ and there are $n_i$ grid elements (or
+states to which a history may transition) along that chord, then $h =
+l / n_i$ giving:
+\begin{equation}
+  \langle \bar{r_k^2} \rangle = k \Bigg(\frac{n_s l}{n_i}\Bigg)^2\:.
+  \label{eq:step_k_length_sub}
+\end{equation}
+From diffusion theory, we expect the average number of interactions
+along the chord to be:
+\begin{equation}
+  \tau = \frac{l}{2 d \sqrt{\langle \bar{r_k^2} \rangle}}\:,
+  \label{eq:optical_thickness_1}
+\end{equation}
+where $d$ is the dimensionality of the problem and $\sqrt{\langle
+  \bar{r_k^2} \rangle}$ is effectively the mean free path of the Monte
+Carlo history in the domain. We can readily interpret $\tau$ to be the
+\textit{effective optical thickness} of a domain of length
+$l$. Inserting Eq~(\ref{eq:step_k_length_sub}) we arrive at:
+\begin{equation}
+  \tau = \frac{n_i}{2 d n_s \sqrt{k}}\:,
+  \label{eq:optical_thickness_2}
+\end{equation}
+which if expanded with Eq~(\ref{eq:analytic_k}) gives us the final
+relation for the effective optical thickness:
+\begin{equation}
+  \tau = \frac{n_i}{2 d n_s}
+  \sqrt{\frac{\log(\rho(\ve{H}))}{\log(W_c)}}\:.
+  \label{eq:optical_thickness_3}
+\end{equation}
+
+For optically thin domains, we expect that most histories will be
+communicated, while optically thick domains will leak the fraction of
+histories that did not interact within. Using the optical thickness
+defined in Eq~(\ref{eq:optical_thickness_3}), we can then complete the
+leakage approximations by defining the bounds of $\tau \rightarrow 0,
+\Lambda \rightarrow 1$ and $\tau \rightarrow \infty, \Lambda
+\rightarrow \tau^{-1}$.  With these bounds we can then define the
+leakage fraction out of a domain for the adjoint Neumann-Ulam method
+using the Wigner rational approximation:
+\begin{equation}
+  \Lambda = \frac{1}{1+\tau}\:,
+  \label{eq:wigner_domain_leakage}
+\end{equation}
+and using the mean-chord approximation:
+\begin{equation}
+  \Lambda = \frac{1-e^{-\tau}}{\tau}\:.
+  \label{eq:mean_chord_domain_leakage}
+\end{equation}
+Here, the leakage fraction is explicitly bound to the eigenvalues of
+the iteration matrix, the size of the domain, the content of the
+discretization stencil, and the weight cutoff selected to terminate
+low weight histories.
+
+\subsubsection{Numerical Experiments}
+\label{subsec:numerical_experiments}
+To test the relationships developed by the spectral analysis, we form
+two simple numerical experiments using the diffusion model problem:
+one to measure the length of the random walks as a function of the
+iteration matrix eigenvalues, and one to measure the domain leakage
+fraction as a function of the iteration matrix eigenvalues and the
+discretization properties. Before doing this, we verify our
+computation of the spectral radius of the iteration matrix by
+numerically computing the largest eigenvalue of the diffusion operator
+using an iterative eigenvalue solver. For this verification, a $100
+\times 100$ square grid with $h=0.01$, $h=0.1$, and $h=1.0$ and the
+absorption cross varied from 0 to 100 while the scattering cross
+subsection was fixed at unity. Figure~\ref{fig:measured_spec_rad} gives
+the measured spectral radius of the iteration matrix and the computed
+spectral radius for the preconditioned diffusion operator using
+Eq~(\ref{eq:iteration_radius}) as function of the absorption to
+scattering ratio $(\Sigma_a / \Sigma_s)$. Excellent agreement was
+observed between the analytic and numerical results with all data
+points computed within the tolerance of the iterative eigenvalue
+solver.
+\begin{figure}[t!]
+  \begin{spacing}{1.0}
+    \begin{center}
+      \includegraphics[width=5in,clip]{chapters/mc_theory/measured_spec_rad.png}
+    \end{center}
+    \caption{\textbf{Measured and analytic preconditioned diffusion
+        operator spectral radius as a function of the absorption cross
+        subsection to scattering cross subsection ratio.} \textit{Values of
+        $h=0.01$, $h=0.1$, and $h=1.0$ were used. The red data was
+        computed numerically by an eigensolver while the black dashed
+        data was generated by Eq~(\ref{eq:iteration_radius}).}}
+    \label{fig:measured_spec_rad}
+  \end{spacing}
+\end{figure}
+
+\subsubsection{Random Walk Length}
+\label{subsubsec:walk_length}
+With the eigenvalue derivations verified, we can go about setting up
+an experiment to measure the length of the random walks generated by
+the adjoint Neumann-Ulam solver. To do this, we again use a $100
+\times 100$ square grid with $h=0.1$ and the absorption cross varied
+from 0 to 100 while the scattering cross subsection was fixed at
+unity. Three weight cutoff values of \sn{1}{-2}, \sn{1}{-4}, and
+\sn{1}{-8} were used with 10,000 histories generated by a point source
+of strength 1 in the center of the domain. For each of the histories,
+the number of transitions made was tallied to provide an effective
+value of $k$ for each history. This value was then averaged over all
+histories to get a measured value of $k$ for the particular
+operator. Figure~\ref{fig:measured_length} presents these measurements
+as well as the analytic result computed by Eq~(\ref{eq:analytic_k}) as
+a function of the iteration matrix spectral radius,
+$\rho(\ve{H})$. Figure~\ref{fig:measured_length_error} gives the
+relative error between the predicted and observed results. We note
+good qualitative agreement between the measured and analytic
+results. However, we observe a larger relative error for both long and
+short random walks.
+\begin{figure}[t!]
+  \begin{spacing}{1.0}
+    \begin{center}
+      \includegraphics[width=5in,clip]{chapters/mc_theory/walk_length.png}
+    \end{center}
+    \caption{\textbf{Measured and analytic random walk length as a
+        function of the iteration matrix spectral radius.} \textit{The
+        weight cutoff was varied with \sn{1}{-2}, \sn{1}{-4}, and
+        \sn{1}{-8}. The red data was computed numerically by an
+        adjoint Neumann-Ulam implementation while the black dashed
+        data was generated by Eq~(\ref{eq:analytic_k}).}}
+    \label{fig:measured_length}
+  \end{spacing}
+\end{figure}
+\begin{figure}[t!]
+  \begin{spacing}{1.0}
+    \begin{center}
+      \includegraphics[width=5in,clip]{chapters/mc_theory/walk_length_error.png}
+    \end{center}
+    \caption{\textbf{Measured and analytic random walk length relative
+        error as a function of the iteration matrix spectral radius.}
+      \textit{The weight cutoff was varied with \sn{1}{-2},
+        \sn{1}{-4}, and \sn{1}{-8}. The relative error between the
+        predicted and measured results is presented for each weight
+        cutoff.}}
+    \label{fig:measured_length_error}
+  \end{spacing}
+\end{figure}
+
+\subsubsection{Domain Leakage}
+\label{subsubsec:domain_leakage}
+Finally, we seek to measure the leakage from a domain in a domain
+decomposed Monte Carlo calculation and assess the quality of our
+analytic relation for the optical thickness of domain and the
+associated leakage approximations. For this experiment, a square grid
+with $h=0.1$ was decomposed into 9 square domains, 3 in each cardinal
+direction with measurements occurring in the central domain without
+boundary grid points. For the cross subsections, the absorption cross
+subsection was varied from 1 to 100 while the scattering cross subsection
+was set to zero to create a purely absorbing environment with weight
+cutoff of \sn{1}{-4}. The optical thickness of these domains will vary
+as a function of the absorption cross subsection if the other parameters
+are fixed. To compute the optical thickness, along with the spectral
+radius as given by Eq~(\ref{eq:iteration_radius}), we also need the
+parameters $n_i$ and $n_s$ which respectively describe the typical
+domain length and the average number of states moved along that
+typical length per history transition. For our grid above, the domains
+are varied in size with $50 \times 50$, $100 \times 100$, and $200
+\times 200$ cells giving $n_i=50$, $n_i=100$, and $n_i=200$ grid
+points or states along the typical length of the domain
+respectively. Looking at the Laplacian stencil in
+Eq~(\ref{eq:nine_point_stencil}), we see that all history transitions
+will only move a single state in either the $i$ or $j$ directions due
+to the symmetry of the problem. Furthermore, if we choose the $i$
+direction, not all states we will transition to will move the history
+in that direction. Therefore, we look to the definition of the
+iteration matrix in Eq~(\ref{eq:iteration_stencil}) and the definition
+of the adjoint probability matrix in Eq~(\ref{eq:adjoint_probability})
+to estimate the $n_s$ parameter. For a particular transition starting
+at state $(i,j)$, 6 of the 8 possible new states in the stencil move
+the history in $i$ direction with relative coefficients of 4 for
+moving in the $(\pm i,0)$ direction and of 1 for moving in the $(\pm
+i,\pm j)$. These coefficients dictate the frequency those states are
+visited relative to the others. For those 6 states we can visit along
+the typical length, their sum is 12 out of the total 20 for the
+coefficients for all possible states with their ratio giving $n_s =
+\frac{3}{5}$.
+
+To compute the leakage fraction numerically, \sn{3}{5} histories were
+sampled from a uniform source of strength unity over the global
+domain. At the start of a stage of histories, the number of histories
+starting in the center domain was computed and as the stage
+progressed, the number of histories that exited that domain was
+tallied with the ratio of the two numbers providing a numerical
+measure for the leakage fraction. Figure~\ref{fig:measured_leakage}
+gives the domain leakage measurements for the domain in the center of
+the global grid as well as the analytic result computed by
+Eqs~(\ref{eq:wigner_domain_leakage}) and
+(\ref{eq:mean_chord_domain_leakage}) as a function of the iteration
+matrix spectral radius.
+\begin{figure}[t!]
+  \begin{spacing}{1.0}
+    \begin{center}
+      \includegraphics[width=5.25in,clip]{chapters/mc_theory/leakage_variation.png}
+    \end{center}
+    \caption{\textbf{Measured and analytic domain leakage as a
+        function of the iteration matrix spectral radius.} \textit{To
+        test the behavior with respect to domain size, $n_i=50$,
+        $n_i=100$,and $n_i=200$ were used. The red data was computed
+        numerically by a domain-decomposed adjoint Neumann-Ulam
+        implementation, the black dashed data was generated by
+        Eq~(\ref{eq:mean_chord_domain_leakage}) using the mean-chord
+        approximation, and the dashed-dotted black data was generated
+        by Eq~(\ref{eq:wigner_domain_leakage}) using the Wigner
+        rational approximation.}}
+    \label{fig:measured_leakage}
+  \end{spacing}
+\end{figure}
+Again, we note good qualitative agreement between the measured and
+analytic quantities but we begin to see the limits of the leakage
+approximations. To compare the quality of the two approximations, the
+absolute error between the computed leakage fraction and that
+generated by the Wigner rational and mean chord approximations is
+plotted in Figure~\ref{fig:leakage_error} for all domain sizes
+tested. 
+\begin{figure}[t!]
+  \begin{spacing}{1.0}
+    \begin{center}
+      \includegraphics[width=5in,clip]{chapters/mc_theory/leakage_error.png}
+    \end{center}
+    \caption{\textbf{Measured and analytic domain leakage absolute
+        error as a function of the iteration matrix spectral radius.}
+      \textit{To test the behavior with respect to domain size,
+        $n_i=50$ (green), $n_i=100$ (blue), and $n_i=200$ (red) were
+        used. The dashed lines represent the error using the Wigner
+        rational approximation while the solid lines represent the
+        error using the mean-chord approximation.}}
+    \label{fig:leakage_error}
+  \end{spacing}
+\end{figure}
+From these error results, the mean chord approximation is shown to
+have a lower error for ill-conditioned systems as compared to the
+Wigner approximation while the Wigner approximation produces less
+error for more well-conditioned systems. We also note that for the
+optically thick domains, the error is likely corresponded to that
+observed in Figure~\ref{fig:measured_length} for the $k$ parameter
+while the large relative error in $k$ for optically thin domains does
+not affect the approximation significantly. In general, the mean chord
+approximation is a better choice to estimate the leakage fraction in a
+domain from the adjoint Neumann-Ulam method and except for a single
+data point with $n_i=50$, the mean chord approximation yielded leakage
+fractions within 0.05 of the measured results. As the domain becomes
+more optically thick (with both increasing $n_i$ and decreasing
+$\rho(\ve{H})$), the approximations are more accurate.
+
+\subsection{Reduction in Domain Leakage from Overlap}
+\label{sec:overlap_reduction}
+
+\subsection{Communication Costs for Symmetric Systems}
+\label{sec:symmetric_comm_costs}
+Use Seigel's work to get communication costs.
 
@@
 point values.
+
+%%---------------------------------------------------------------------------%%
+\section{Parallel Scaling Studies}
+\label{sec:scaling_studies}
+For all cases, we explore scaling for just the Monte Carlo transport
+kernel and MCSA as a whole iterative scheme. This will be necessary to
+determine if the initial relaxation step required at each iteration is
+a limiting factor in convergence. For each case, it will be
+constructive to study the scaling performance as a function of
+spectral radius - dictated by the cross sections and grid size to
+diffusion length ratio. In addition, all of these studies should
+hopefully have data from both the cluster and Titan and should all be
+compared to Aztec GMRES and CG implementations with appropriate
+preconditioning.
+
+\subsection{Overlapping Domain Decomposition}
+\label{subsec:overlapping_domain_decomp}
+
+\subsubsection{Weak Scaling}
+\label{subsubsec:overlapping_weak}
+
+\subsubsection{Strong Scaling}
+\label{subsubsec:overlapping_strong}
+
+\subsection{Multiple Set Domain Decomposition}
+\label{subsec:multiple_set_decomposition}
+
+\subsubsection{Weak Scaling}
+\label{subsubsec:multiple_set_weak}
+
+\subsubsection{Strong Scaling}
+\label{subsubsec:multiple_set_strong}
+
+\subsection{MSOD Decomposition}
+\label{subsec:msod_decomposition}
+
+\subsubsection{Weak Scaling}
+\label{subsubsec:msod_weak}
+
+\subsubsection{Strong Scaling}
+\label{subsubsec:msod_strong}
+
+\subsection{Effects of Load Imbalance}
+\label{subsec:msod_load_imbalance}
+
+\subsubsection{Weak Scaling}
+\label{subsubsec:load_imbalance_weak}
+
+\subsubsection{Strong Scaling}
+\label{subsubsec:load_imbalance_strong}

### 2013-04-27 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
more dissertation refactoring

@@
-\chapter{The Simplified $P_N$ Equations}
+\chapter{Monte Carlo Solution Methods for the Simplified $P_N$ Equations}
 \label{ch:spn_equations}
@@
 
-\subsection{Preconditioning Strategy}
-\label{subsec:spn_preconditioning}
+\subsection{Block Jacobi Preconditioning}
+\label{sec:spn_preconditioning}
 If Monte Carlo methods are to be used to solve the $SP_N$ system of
@@
 problem.
+
+%%---------------------------------------------------------------------------%%
+\section{Fuel Assembly Criticality Calculations}
+\label{sec:fuel_assembly_calcs}
+
+\subsection{Jacobi Preconditioned Calculations}
+\label{subsec:jacobi_prec_assembly_calc}
+
+\subsection{MCSA Break Down}
+\label{subsec:mcsa_break_down}
+
+\subsection{Advanced Preconditioning Strategies}
+\label{subsec:spn_advanced_preconditioning}
+
+\subsubsection{ILUT Preconditioning}
+\label{subsubsec:spn_ilut_preconditioning}
+
+\subsubsection{Sparse Approximate Inverse Preconditioning}
+\label{subsubsec:spn_spainv_preconditioning}
+
+\subsubsection{SPAINV Improved ILUT Preconditioning}
+\label{subsubsec:spn_spainv_preconditioning}
+
+\subsection{MCSA Comparison to Conventional Methods}
+\label{subsec:spn_comparison}

### 2013-04-28 doc/dissertation/document/chapters/mc_background/mc_background.tex
more dissertation refactoring
adding expected value estimator

@@
 define the following operator\footnote{It should be noted that
-  non-diagonal splittings have been recently explored in
+  non-diagonal splittings have been recently explored
   \citep{srinivasan_monte_2010} and have the potential to improve
-  efficiency.}:
+  efficiency. However, it was observed in this work that this type of
+  splitting did not improve performance in the asymptotic limit of
+  $\rho(\ve{H}) \rightarrow 1$ for non-trivial problems.}:
 \begin{equation}
@@
 succession of matrix-vector multiply operations, we arrive at an
-alternative perspective of this summation by considering its $i^{th}$
-component:
+alternative perspective of this summation by considering the $i^{th}$
+component of the solution vector:
 \begin{equation}
@@
 sequence to estimate the state transitions in
-Eq~(\ref{eq:expanded_neumann_solution}).
+Eq~(\ref{eq:expanded_neumann_solution}). In the direct method, we will
+use the provided linear operator $\ve{A}$ to form the Neumann Ulam
+decomposition while the adjoint method will use the adjoint linear
+operator ($\ve{A}^T$ for real-valued systems) to form the
+decomposition.
 
@@
 \label{sec:direct_mc}
-In the context of matrix inversion, a direct method resembles an
+In the context of matrix inversion, a direct (forward) method resembles an
 adjoint Monte Carlo method in the reactor physics community where the
@@
 the non-zero elements of $\ve{H}$ as those components that are zero
-will not participate in the random walk. Doing so prevents an
-infinite weight from being generated in Eq~(\ref{eq:direct_weight})
-and eliminates unnecessary computations.
+will not participate in the random walk and produce identical sparsity
+patterns.
 
 Using these matrices, we can then form the expectation value of the
-direct solution. For a given random walk permutation $\nu$ with $k$
-events, we define the weight of that permutation to be:
+direct solution. For a given random walk permutation $\nu$, we define
+the weight of that permutation on the $m^{th}$ step to be:
 \begin{equation}
-  W_{m} = w_{i,i_1} w_{i_1,i_2} \cdots w_{i_{m-1},i_m}\:,
+  W_{m} = w_{i_0,i_1} w_{i_1,i_2} \cdots w_{i_{m-1},i_m}\:,
   \label{eq:direct_permutation_weight}
 \end{equation}
-such that the weight of each transition event contributes to the
-total. The contribution to the solution from a particular random walk
-permutation is then the \textit{terminal estimator}:
+such that the weight of each transition event contributes to the total
+through multiplication with $W_0 = 1$ as the starting weight. The
+contribution to the solution from a particular random walk permutation
+with $k$ total events is then the \textit{forward estimator}:
 \begin{equation}
-  X_{\nu}(i_0 = i) = \sum_{m=0}^k W_{m} b_{i_m}\:,
+  X_{i_0 = i}(\nu) = \sum_{m=0}^k W_{m} b_{i_m}\:,
   \label{eq:direct_permutation_contribution}
 \end{equation}
-where $X_{\nu}(i_0 = i)$ signifies that the solution state in which we
-are tallying defines state $i$.  We can interpret this precisely as
-before in that during the random walk we collect the source in the
-states that are visited and apply them to the solution tally. We then
-define the probability that a particular random walk permutation of
-$k$ events will occur:
+where $X_{i_0 = i}(\nu)$ signifies that the solution state, $i_0$, in
+which the random walk $\nu$ started is also the state, $i$, in which
+we are tallying. We then define the probability that a particular
+random walk permutation of $k$ events will occur:
 \begin{equation}
@@
 \begin{equation}
-  E\{X(i_0 = i)\} = \sum_{\nu} P_{\nu} X_{\nu}\:,
+  E\{X_i\} = \sum_{\nu} P_{\nu} X_{i}(\nu)\:,
   \label{eq:direct_expectation_value}
 \end{equation}
-which, if expanded, directly recovers the exact solution:
+which, if expanded, directly recovers the exact solution by forming
+the Neumann series:
 \begin{equation}
   \begin{split}
-    E\{X(i_0 = i)\}
+    E\{X_i\}
     &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
@@
 In cases where we seek only approximate solutions, we need only to
-perform a minimal number of random walks in order to generate an
+perform a predetermined number of random walks in order to generate an
 approximation for $\ve{x}$. If we are only to approximate the
 solution, we also need conditions by which we may terminate a random
-walk. We do this by noticing that the factors added to
-Eq~(\ref{eq:direct_permutation_weight}) will become diminishingly
-small due to their definition in Eq~(\ref{eq:direct_weight}) and
-therefore their contributions to the solution estimate will become
-negligible. Using this, we choose terminate a random walk sequence
-with a \textit{weight cutoff}, $W_c$, that is enforced when $W_m <
-W_c$ for a particular random walk permutation.
+walk as the Neumann Ulam decomposition defined by
+Eqs~(\ref{direct_probability}) and (\ref{eq:direct_weight}) will
+create a random walk weight in Eq~(\ref{eq:direct_permutation_weight})
+that approaches, but never reaches zero. We do this by noticing that
+the factors added to Eq~(\ref{eq:direct_permutation_weight}) will
+become diminishingly small due to their definition in
+Eq~(\ref{eq:direct_weight}) and therefore their contributions to the
+solution estimate will become negligible. Using this, we choose
+terminate a random walk sequence with a \textit{weight cutoff}, $W_c$,
+that is enforced when $W_m < W_c$ for a particular random walk
+permutation.
 
@@
 \label{sec:adjoint_mc}
-An alternative formulation for Monte Carlo matrix inversion is the
-adjoint method. We begin by defining the linear system adjoint to
-Eq~(\ref{eq:linear_problem}):
+Often a more useful form, an alternative for Monte Carlo matrix
+inversion is the adjoint method. We begin by defining the linear
+system adjoint to Eq~(\ref{eq:linear_problem}):
 \begin{equation}
@@
 such that we expect to select a new state $j$ from the current state
-in the random walk $j$ by sampling column-wise. Per
+in the random walk $j$ by sampling column-wise (or row-wise if an
+adjoint probability matrix is formed). Per
 Eq~(\ref{eq:adjoint_neumann_ulam}), the transition weight is then
@@
 adjoint constraints, the contribution to the solution in state $i$
-from a particular random walk permutation is then the
+from a particular random walk permutation of $k$ events is then the
 \textit{collision estimator}:
 \begin{equation}
-  X_{\nu} = \sum_{m=0}^k W_{m} \delta_{i,i_m}\:,
+  X_{j}(\nu) = \sum_{m=0}^k W_{m} \delta_{i_m,j}\:,
   \label{eq:adjoint_permutation_contribution}
@@
 where the Kronecker delta indicates that the tally contributes only in
-the current state and $b_{i_0}$ will be the sampled source starting
-weight. Note here that the estimator in
+the current state the random walk starting weight will be $W_0 =
+b_{i_0}$. Note here that the estimator in
 Eq~(\ref{eq:adjoint_permutation_contribution}) does not have a
@@
 the situation in the direct method where we must start a random walk
-in each source state for every permutation such that we may compute a
-contribution for that state. In the adjoint method, we instead tally
-in all states and those of lesser importance will not be visited as
-frequently by the random walk. Finally, the expectation value using
+in each source state for every permutation if we want to compute a
+solution estimate for that state. In the adjoint method, we instead
+tally in all states and those of lesser importance will not be visited
+as frequently by the random walk. Finally, the expectation value using
 all permutations is:
 \begin{equation}
-  E\{X\} = \sum_{\nu} P_{\nu} X_{\nu}\:
+  E\{X_j\} = \sum_{\nu} P_{\nu} X_{j}(\nu)\:
   \label{eq:adjoint_expectation_value}
@@
 \label{subsec:expected_value_estimator}
-Here we introduce an estimator based on the method of expected values.
+In addition to the collision estimator, an additional estimator is
+available is available due to the work of Okten
+\citep{okten_solving_2005} that uses the method of expected values as
+a means to improve the Monte Carlo estimate. As outlined by Spanier
+and Gelbard \citep{spanier_monte_1969}, the method of expected values
+is a deterministic averaging of events that may potentially occur in
+the Monte Carlo random walk sequence. Okten applied this principle
+directly to discrete Monte Carlo by forming the \textit{expected value
+  estimator} for a random walk of $k$ events:
+\begin{equation}
+  X_{j}(\nu) = b_j + \sum_{m=0}^k W_m h_{j,i_m}\,
+  \label{eq:expected_value_estimator}
+\end{equation}
+where now the contribution of the iteration matrix is
+deterministically averaged at step $m$ over all potentially states $j$
+that may be reached from the current state $i_m$. Via Okten, the
+estimator can be shown to be unbiased through a comparison to
+the expected value estimator. We can first rewrite the summation in
+Eq~(\ref{eq:expected_value_estimator}):
+\begin{equation}
+  X_{j}(\nu) = b_j + \sum_{m=0}^k \sum_{i=1}^N W_m
+  \delta_{i_m,i} h_{ji}\,
+  \label{eq:unbiased_eval_1}
+\end{equation}
+where $N$ is the number of states in the system. Immediately, we see
+the collision estimator as defined by
+Eq~(\ref{eq:adjoint_permutation_contribution}) and can therefore write
+the expectation value as:
+\begin{equation}
+  E\{X_{j}\} = b_j + \sum_{i=1}^N E\{X_{i}\} h_{ji}\,
+  \label{eq:unbiased_eval_2}
+\end{equation}
+which is equivalently is the $j^{th}$ component of
+Eq~(\ref{eq:richardson_split}):
+\begin{equation}
+  E\{X_{j}\} = b_j + \sum_{i=1}^N x_{i} h_{ji}\,
+  \label{eq:unbiased_eval_2}
+\end{equation}
+and is therefore and unbiased estimate. Compared to the collision
+estimator, provides additional information at every step of the random
+walk, yielding potentially better statistics with the same amount of
+transport work. Conveniently, even if no Monte Carlo histories are
+computed, the estimator still deterministically computes the first
+term of the Neumann Series, $\ve{H}^0\ve{b}$, whereas the collision
+estimator will provide no information.
 
@@
 As a means of visually demonstrating the adjoint Monte Carlo method,
-again considerthe a 2-dimensional thermal diffusion problem with
+again consider the 2-dimensional thermal diffusion problem with
 sources on the left and right hand sides of the domain and a smaller
-uniform source as shown in Figure~\ref{fig:heat_setup}. For the
-adjoint method, the number of histories sampled from the source was
-increased from 10 to 10,000,000 in order to show its effects on the
-solution and the statistical nature of the
+uniform source as shown in Figure~\ref{fig:heat_setup}. Using the
+adjoint method with the collision estimator, the number of histories
+sampled from the source was increased from 10 to 10,000,000 in order
+to show its effects on the solution and the statistical nature of the
 method. Figure~\ref{fig:adjoint_evolution} gives these results.
@@
 As the number of histories used per state is increased, the
-statistical variance of the solutions is decreased as more tallies are
-made. At 10,000,000 histories per state, enough tallies have been made
-to generate a reasonable estimate for the structure of the
-solution. The visual difference between
+statistical variance of the solutions is decreased as more tally
+contributions are made. At 10,000,000 histories per state, enough
+information has been tallied to generate a reasonable estimate for the
+structure of the solution. The visual difference between
 Figures~\ref{fig:direct_evolution} and \ref{fig:adjoint_evolution} is
@@
 solution evolves with the addition of histories, more histories
-emanate from the source with more penetrating the domain and making
-contributions to those tallies.
+emanate from the boundary and smaller uniform source with more
+penetrating from the boundary into the domain and making contributions
+to the tallies in those states.
 
@@
 
-\subsection{Terminal Estimator Variance}
-\label{subsec:terminal_variance}
-We can compute the variance of the estimators through traditional
+\subsection{Forward Estimator Variance}
+\label{subsec:forward_variance}
+We can compute the variance of the forward estimator through traditional
 methods by defining the variance, $\sigma_i$, for each component in
@@
 
+\subsection{Weight Capping}
+\label{subsec:weight_capping}
+
 \subsection{Reduced Domain Approximation}

### 2013-04-28 doc/dissertation/document/chapters/mc_background/mc_background.tex
adding to the mc background chapter

@@
 permutation.
+
+\subsection{Forward Estimator Variance}
+\label{subsec:forward_variance}
+We can compute the variance of the forward estimator through traditional
+methods by defining the variance, $\sigma_i$, for each component in
+the solution:
+\begin{equation}
+  {\sigma_i}^2 = E\{X_i - (\ve{A}^{-1}\ve{b})_i\}^2 = E\{X_i^2\} - x_i^2\:,
+  \label{eq:direct_variance_1}
+\end{equation}
+where the vector exponentials are computed element-wise. Inserting
+Eq~(\ref{eq:direct_expectation_value}) gives:
+\begin{equation}
+  \sigma_i^2 = \Big(\sum_{\nu} P_{\nu} X_{\nu}^2\Big)_i - x_i^2\:,
+  \label{eq:direct_variance_2}
+\end{equation}
+and applying Eq~(\ref{eq:direct_permutation_contribution}):
+\begin{equation}
+  \sigma_i^2 = \Big(\sum_{\nu} P_{\nu} \sum_{m=0}^k W_{m}^2
+  b_{i_m}^2\Big)_i - x_i^2\:.
+  \label{eq:direct_variance_3}
+\end{equation}
+Finally, expanding the transition probabilities yields the variance:
+\begin{equation}
+  \sigma_i^2 = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
+  \sum_{i_k}^{N} p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
+  w^2_{i,i_1}w^2_{i_1,i_2}\ldots w^2_{i_{k-1},i_k} b_{i_k}^2 - x_i^2\:.
+  \label{eq:direct_variance_4}
+\end{equation}
+Using this definition for the variance, we can arrive at a more natural
+reason for enforcing $\rho(\ve{H}) < 1$ for our Monte Carlo method to
+converge. Per the Hadamard product, we can concatenate the summation
+in Eq~(\ref{eq:direct_variance_4}):
+\begin{equation}
+  (\ve{P} \circ \ve{W} \circ \ve{W})^k =
+  \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots \sum_{i_k}^{N}
+  p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
+  w^2_{i,i_1}w^2_{i_1,i_2}\ldots w^2_{i_{k-1},i_k}\:.
+  \label{eq:double_weighted_decomposition}
+\end{equation}
+If we assign $\ve{G} = \ve{P} \circ \ve{W} \circ \ve{W}$ as in
+Eq~(\ref{eq:neumann_ulam_decomposition}), we then have a new
+formulation for the variance:
+\begin{equation}
+  \sigma^2_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
+  \sum_{i_k}^{N}g_{i,i_1}g_{i_1,i_2}\ldots g_{i_{k-1},i_k} b_{i_k}^2 -
+  x_i^2\:,
+\end{equation}
+which contains the general Neumann series for $\ve{G}$,
+\begin{equation}
+  \ve{T} = \sum_{k=0}^{\infty} \ve{G}^k\:,
+  \label{eq:variance_neumann_series}
+\end{equation}
+where $\ve{T} = (\ve{I}-\ve{G})^{-1}$. We can then insert $T$ back
+into the variance formulation for a more concise definition:
+\begin{equation}
+  \sigma^2_i = (\ve{T}\ve{b})_i - x_i^2\:.
+  \label{eq:direct_variance_5}
+\end{equation}
+We can relate $\ve{G}$ to $\ve{H}$ by noting that $\ve{G}$ simply
+contains an additional Hadamard product with the weight matrix. The
+Hadamard product has the property that:
+\begin{equation}
+  |\ve{H} \circ \ve{W}| \geq |\ve{H}|\ |\ve{W}|\:.
+  \label{eq:hadamard_inequality}
+\end{equation}
+Using the norm property of the Hadamard product and
+Eq~(\ref{eq:neumann_ulam_decomposition}), we can define the norm of
+$\ve{W}$ as:
+\begin{equation}
+  \frac{|\ve{H}|}{|\ve{P}|} \geq |\ve{W}|\:.
+  \label{eq:weight_norm}
+\end{equation}
+Choosing the infinity norm of the operator as defined in
+Eq~(\ref{eq:matrix_infinity_norm}), the row normalized probability
+matrix will yield a norm of 1 giving the following inequality for
+relating $\ve{G}$ and $\ve{H}$:
+\begin{equation}
+  |\ve{G}| \geq |\ve{H}|^2
+  \label{eq:variance_norm_inequality}
+\end{equation}
+Using these relations to analyze Eq~(\ref{eq:direct_variance_5}), we
+see that if $\rho(\ve{G}) > 1$, then the sum in
+Eq~(\ref{eq:variance_neumann_series}) will not converge and an
+infinite variance will arise as the elements of $\ve{T}$ become
+infinite in Eq~(\ref{eq:direct_variance_5}). We must restrict $\ve{G}$
+to alleviate this and therefore restrict $\ve{H}$ due to
+Eq~(\ref{eq:variance_norm_inequality}) with $\rho(\ve{H}) < 1$ so that
+our expectation values for the solution may have a finite variance.
 
@@
 
-In terms of particle transport, this adjoint method is equivalent to a
-traditional forward method. As a result of using the adjoint system,
-we modify our probabilities and weights using the \textit{adjoint
-  Neumann-Ulam decomposition} of $\ve{H}$:
+In terms of radiation transport, this adjoint method is equivalent to
+a traditional forward method where the initial state $i_0$ of the
+random walk is determined by sampling the source vector $\ve{b}$ with
+probabilities:
+\begin{equation}
+  P_{i_0=i}(\nu) = \frac{b_i}{||\ve{b}||_1}\:.
+  \label{eq:adjoint_source_probability}
+\end{equation}
+ As a result of using the adjoint system, we modify our probabilities
+ and weights using the \textit{adjoint Neumann-Ulam decomposition} of
+ $\ve{H}$:
 \begin{equation}
@@
 \end{equation}
-such that we expect to select a new state $j$ from the current state
-in the random walk $j$ by sampling column-wise (or row-wise if an
+such that we expect to select a new state, $j$, from the current state
+in the random walk, $i$, by sampling column-wise (or row-wise if an
 adjoint probability matrix is formed). Per
@@
 Eq~(\ref{eq:inner_product_constraint}) generated by applying the
-adjoint constraints, the contribution to the solution in state $i$
+adjoint constraints, the contribution to the solution in state $j$
 from a particular random walk permutation of $k$ events is then the
@@
 the entire solution vector. In an adjoint Monte Carlo simulation using
-this estimator, the $w_{i,j}$ elements that are added into the tally
+this estimator, the $w_{ij}$ elements that are added into the tally
 for each state are only selected if the random walk currently resides
@@
 
+\subsection{Collision Estimator Variance}
+\label{subsec:forward_variance}
+We can compute the variance of the collision estimator in the same way
+as the direct estimator for each component in the solution where now:
+\begin{equation}
+  \sigma_j^2 = \Big(\sum_{\nu} P_{\nu} \sum_{m=0}^k W_{m}^2
+  \delta_{i_m,j}\Big)_j - x_j^2\:.
+  \label{eq:collision_variance_1}
+\end{equation}
+Expanding the transition probabilities yields the same variance as the
+direct estimator, this time with the terms in reverse order and the
+introduction of the Kronecker delta:
+\begin{equation}
+  \sigma_j^2 = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
+  \sum_{i_k}^{N} p_{i_k,i_{k-1}}\ldots p_{i_2,i_1} p_{i_1,i_0}
+  w^2_{i_k,i_{k-1}}\ldots w^2_{i_2,i_1} w^2_{i_1,i_0}
+  b^2_{i_0}\delta_{i_k,j} - x_i^2\:.
+  \label{eq:collision_variance_2}
+\end{equation}
+If we again $\ve{G} = \ve{P} \circ \ve{W} \circ \ve{W}$, we then have:
+\begin{equation}
+  \sigma^2_j = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
+  \sum_{i_k}^{N}g_{i_{k},i_{k-1}} \ldots
+  g_{i_2,i_1}g_{i_1,i_0}b_{i_0}\delta_{i_k,j}^2 - x_i^2\:,
+\label{eq:collision_variance_3}
+\end{equation}
+which again contains the general Neumann series for $\ve{G}$ as given
+by Eq~(\ref{eq:variance_neumann_series}).  Inserting $T$ back into the
+variance formulation gives:
+\begin{equation}
+  \sigma^2_j = (\ve{T}\ve{b})_j - x_i^2\:.
+  \label{eq:direct_variance_5}
+\end{equation}
+We can relate $\ve{G}$ to $\ve{H}$ by noting that $\ve{G}$ simply
+contains an additional Hadamard product with the weight matrix. The
+Hadamard product has the property that:
+\begin{equation}
+  |\ve{H} \circ \ve{W}| \geq |\ve{H}|\ |\ve{W}|\:.
+  \label{eq:hadamard_inequality}
+\end{equation}
+Using the norm property of the Hadamard product and
+Eq~(\ref{eq:neumann_ulam_decomposition}), we can define the norm of
+$\ve{W}$ as:
+\begin{equation}
+  \frac{|\ve{H}|}{|\ve{P}|} \geq |\ve{W}|\:.
+  \label{eq:weight_norm}
+\end{equation}
+Choosing the infinity norm of the operator as defined in
+Eq~(\ref{eq:matrix_infinity_norm}), the row normalized probability
+matrix will yield a norm of 1 giving the following inequality for
+relating $\ve{G}$ and $\ve{H}$:
+\begin{equation}
+  |\ve{G}| \geq |\ve{H}|^2
+  \label{eq:variance_norm_inequality}
+\end{equation}
+Using these relations to analyze Eq~(\ref{eq:direct_variance_5}), we
+see that if $\rho(\ve{G}) > 1$, then the sum in
+Eq~(\ref{eq:variance_neumann_series}) will not converge and an
+infinite variance will arise as the elements of $\ve{T}$ become
+infinite in Eq~(\ref{eq:direct_variance_5}). We must restrict $\ve{G}$
+to alleviate this and therefore restrict $\ve{H}$ due to
+Eq~(\ref{eq:variance_norm_inequality}) with $\rho(\ve{H}) < 1$ so that
+our expectation values for the solution may have a finite variance.
+
 \subsection{Expected Value Estimator}
@@
 In addition to the collision estimator, an additional estimator is
-available is available due to the work of Okten
-\citep{okten_solving_2005} that uses the method of expected values as
-a means to improve the Monte Carlo estimate. As outlined by Spanier
-and Gelbard \citep{spanier_monte_1969}, the method of expected values
-is a deterministic averaging of events that may potentially occur in
-the Monte Carlo random walk sequence. Okten applied this principle
+available due to the work of Okten \citep{okten_solving_2005} that
+uses the method of expected values as a means to improve the Monte
+Carlo estimate. As outlined by Spanier and Gelbard
+\citep{spanier_monte_1969}, the method of expected values is a
+deterministic averaging of events that may potentially occur in the
+Monte Carlo random walk sequence. Okten applied this principle
 directly to discrete Monte Carlo by forming the \textit{expected value
@@
 and is therefore and unbiased estimate. Compared to the collision
-estimator, provides additional information at every step of the random
-walk, yielding potentially better statistics with the same amount of
-transport work. Conveniently, even if no Monte Carlo histories are
-computed, the estimator still deterministically computes the first
+estimator, the expected value estimator provides additional
+information at every step of the random walk, yielding potentially
+better statistics with the same amount of transport
+work. Conveniently, even if no Monte Carlo histories are computed, the
+expected value estimator still deterministically computes the first
 term of the Neumann Series, $\ve{H}^0\ve{b}$, whereas the collision
@@
 to the tallies in those states.
-
-%%---------------------------------------------------------------------------%%
-\section{Estimator Variance}
-\label{subsec:estimator_variance}
-
-\subsection{Forward Estimator Variance}
-\label{subsec:forward_variance}
-We can compute the variance of the forward estimator through traditional
-methods by defining the variance, $\sigma_i$, for each component in
-the solution:
-\begin{equation}
-  {\sigma_i}^2 = E\{X(i_0 = i) - (\ve{A}^{-1}\ve{b})_i\}^2 = E\{X(i_0
-  = i)^2\} - x_i^2\:,
-  \label{eq:direct_variance_1}
-\end{equation}
-where the vector exponentials are computed element-wise. Inserting
-Eq~(\ref{eq:direct_expectation_value}) gives:
-\begin{equation}
-  \sigma_i^2 = \Big(\sum_{\nu} P_{\nu} X_{\nu}^2\Big)_i - x_i^2\:,
-  \label{eq:direct_variance_2}
-\end{equation}
-and applying Eq~(\ref{eq:direct_permutation_contribution}):
-\begin{equation}
-  \sigma_i^2 = \Big(\sum_{\nu} P_{\nu} \sum_{m=0}^k W_{m}^2
-  b_{i_m}^2\Big)_i - x_i^2\:.
-  \label{eq:direct_variance_3}
-\end{equation}
-Finally, expanding the transition probabilities yields the variance:
-\begin{equation}
-  \sigma_i^2 = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-  \sum_{i_k}^{N} p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
-  w^2_{i,i_1}w^2_{i_1,i_2}\ldots w^2_{i_{k-1},i_k} b_{i_m} - x_i^2\:.
-  \label{eq:direct_variance_4}
-\end{equation}
-Using this definition for the variance, we can arrive at a more natural
-reason for enforcing $\rho(\ve{H}) < 1$ for our Monte Carlo method to
-converge. Per the Hadamard product, we can concatenate the summation
-in Eq~(\ref{eq:direct_variance_4}):
-\begin{equation}
-  (\ve{P} \circ \ve{W} \circ \ve{W})^k =
-  \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots \sum_{i_k}^{N}
-  p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
-  w^2_{i,i_1}w^2_{i_1,i_2}\ldots w^2_{i_{k-1},i_k}\:.
-  \label{eq:double_weighted_decomposition}
-\end{equation}
-If we assign $\ve{G} = \ve{P} \circ \ve{W} \circ \ve{W}$ as in
-Eq~(\ref{eq:neumann_ulam_decomposition}), we then have a new
-formulation for the variance:
-\begin{equation}
-  \sigma^2_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-  \sum_{i_k}^{N}g_{i,i_1}g_{i_1,i_2}\ldots g_{i_{k-1},i_k} b_{i_k}^2 -
-  x_i^2\:,
-\end{equation}
-which contains the general Neumann series for $\ve{G}$,
-\begin{equation}
-  \ve{T} = \sum_{k=0}^{\infty} \ve{G}^k\:,
-  \label{eq:variance_neumann_series}
-\end{equation}
-where $\ve{T} = (\ve{I}-\ve{G})^{-1}$. We can then insert $T$ back
-into the variance formulation for a more concise definition:
-\begin{equation}
-  \sigma^2_i = (\ve{T}\ve{b})_i - x_i^2\:.
-  \label{eq:direct_variance_5}
-\end{equation}
-We can relate $\ve{G}$ to $\ve{H}$ by noting that $\ve{G}$ simply
-contains an additional Hadamard product with the weight matrix. The
-Hadamard product has the property that:
-\begin{equation}
-  |\ve{H} \circ \ve{W}| \geq |\ve{H}|\ |\ve{W}|\:.
-  \label{eq:hadamard_inequality}
-\end{equation}
-Using the norm property of the Hadamard product and
-Eq~(\ref{eq:neumann_ulam_decomposition}), we can define the norm of
-$\ve{W}$ as:
-\begin{equation}
-  \frac{|\ve{H}|}{|\ve{P}|} \geq |\ve{W}|\:.
-  \label{eq:weight_norm}
-\end{equation}
-Choosing the infinity norm of the operator as defined in
-Eq~(\ref{eq:matrix_infinity_norm}), the row normalized probability
-matrix will yield a norm of 1 giving the following inequality for
-relating $\ve{G}$ and $\ve{H}$:
-\begin{equation}
-  |\ve{G}| \geq |\ve{H}|^2
-  \label{eq:variance_norm_inequality}
-\end{equation}
-Using these relations to analyze Eq~(\ref{eq:direct_variance_5}), we
-see that if $\rho(\ve{G}) > 1$, then the sum in
-Eq~(\ref{eq:variance_neumann_series}) will not converge and an
-infinite variance will arise as the elements of $\ve{T}$ become
-infinite in Eq~(\ref{eq:direct_variance_5}). We must restrict $\ve{G}$
-to alleviate this and therefore restrict $\ve{H}$ due to
-Eq~(\ref{eq:variance_norm_inequality}) with $\rho(\ve{H}) < 1$ so that
-our expectation values for the solution may have a finite variance.
-
-\subsection{Collision Estimator Variance}
-\label{subsec:collision_variance}
-
-\subsection{Expected Value Estimator Variance}
-\label{subsec:expected_value_variance}
 
@@
 \subsection{Preconditioning MCSA}
-\label{sec:stochastic_preconditioning}
+\label{subsec:stochastic_preconditioning}
 In most cases, at least a minimal amount of \textit{preconditioning}
@@
 requirements for convergence, they do require that the spectral radius
-of the iteration matrix be less than one. As an example, to achieve
-this for diagonally dominant matrices point Jacobi left
-preconditioning can be used such that the preconditioning matrix
-$\ve{M}$ is:
+of the iteration matrix be less than one. Preconditioning serves as a
+means of achieving this.
+
+\subsubsection{Basic Preconditioning}
+\label{subsubsec:basic_mcsa_preconditioning}
+As an example, to achieve this for diagonally dominant matrices point
+Jacobi left preconditioning can be used such that the preconditioning
+matrix $\ve{M}$ is:
 \begin{equation}
@@
 
-\subsection{General Preconditioning Strategies}
-\label{subsec:general_mcsa_strategies}
+\subsubsection{General Preconditioning Strategies}
+\label{subsubsec:general_mcsa_preconditioning}
 It is possible to use general left, right, and left/right
 preconditioning with MCSA by carefully considering the underly Monte
-Carlo that will be solved with the Neumann Ulam method.
+Carlo that will be solved with the Neumann Ulam method. We consider
+here the general left/right preconditioned method as the left or right
+preconditioned methods can be inferred from its formulation. We
+consider a left preconditioner $\ve{M_L}$ and a right preconditioner
+$\ve{M_R}$. The left/right preconditioned linear problem is then:
+\begin{equation}
+  \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\ve{M}_R\ve{x} = \ve{M}_L^{-1}\ve{b}\:.
+  \label{eq:left_right_linear_problem}
+\end{equation}
+To handle the right preconditioning, the system is written with a
+substitution of variables:
+\begin{equation}
+  \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\ve{u} = \ve{M}_L^{-1}\ve{b}\:,
+  \label{eq:left_right_subs_problem}
+\end{equation}
+with
+\begin{equation}
+  \ve{x} = \ve{M}_R^{-1}\ve{u}\:.
+  \label{eq:left_right_recover}
+\end{equation}
+To apply such a method to MCSA, we solve for the substituted variable
+$\ve{u}$ during the iteration sequence:
+\begin{subequations}
+  \begin{gather}
+    \ve{u}^{k+1/2} = \ve{u}^k + \ve{r}^k\:,\\
+    \ve{r}^{k+1/2} = \ve{M}_L^{-1}(\ve{b}-\ve{A}\ve{M}_R^{-1}\ve{u}^{k+1/2})\:,\\ 
+    \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\delta\ve{u}^{k+1/2} = \ve{r}^{k+1/2}\:,\\ 
+    \ve{u}^{k+1} = \ve{u}^{k+1/2} + \delta \ve{u}^{k+1/2}\:,\\
+    \ve{r}^{k+1} = \ve{M}_L^{-1}(\ve{b}-\ve{A}\ve{M}_R^{-1}\ve{u}^{k+1})\:,
+    \label{eq:left_right_mcsa}
+  \end{gather}
+\end{subequations}
+and then recover the original solution vector with
+Eq~(\ref{eq:left_right_recover}). For the Monte Carlo problem, we
+isolate the generation of the correction:
+\begin{equation}
+  \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\delta\ve{u}^{k+1/2} = \ve{r}^{k+1/2}\:,
+  \label{eq:left_right_correction}
+\end{equation}
+and note that the precondtioned residual of the substituted variable
+is now serving as the source and the new iteration matrix is:
+\begin{equation}
+  \ve{H} = \ve{I} - \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\:.
+  \label{eq:left_right_iteration_matrix}
+\end{equation}
+As we require $(i,j)$ element-wise access to the iteration matrix in
+order to construct probabilities and weights for the Monte Carlo
+procedure, the \textit{composite operator},
+$\ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}$, must be formed via matrix-matrix
+multiplication. Several possible shortcomings of this preconditioning
+approach are readily observed. First, the matrix-matrix multiplication
+operation for sparse, parallel distributed matrices is significantly
+more expensive than a matrix-vector multiplication operation. Second,
+each preconditioner must be explicitly inverted, an operation in
+itself that my be expensive prohibits the use of any preconditioners
+which prohibit the extraction of the inverse. Third, for many modern
+preconditioning methods, this inversion may yield dense matrices,
+destroying sparsity and further impeding the performance of a
+matrix-matrix multiplication operation. It is also interesting to note
+that the Monte Carlo problem in the general left/right preconditioned
+scheme given by Eq~(\ref{eq:left_right_correction}) is not fully
+left/right preconditioned (meaning that we do not recover $\ve{x}$),
+but instead part of a sequence for finding the substituted variable
+$\ve{u}$. We do, however, gain the benefits of this general
+preconditioning by building the iteration matrix in
+Eq~(\ref{eq:left_right_iteration_matrix}) from the fully
+preconditioned linear operator. In addition, for MCSA to apply to
+increasingly difficult problems, more advanced preconditioning
+techniques that require the generation of the composite operator may
+be necessary for convergence.
 
@@
 A timing and convergence study is used to demonstrate the
-effectiveness of the adjoint method as compared to the direct
-method. To assess both the CPU time and number of iterations required
-to converge to a solution, a problem of constant $\Delta$ was used
-with varying values of the number of mesh elements, fixing the
-spectral radius of the system at a constant value for each
-variation. Both the five-point and nine-point stencils were used with
-both the direct and adjoint solvers. For each case, $N \times N$ total
-random walk permutations were computed per MCSA iteration where $N
-\times N$ is the number of discrete grid points in the system. Solver
-parameters were set to a weight cutoff of \sn{1}{-4} for the
-stochastic linear solver and a convergence tolerance of \sn{1}{-8} for
-the MCSA iterative solver.  Figure~\ref{fig:poisson_cpu_time} gives
-the CPU time needed for each case to converge in seconds and
-Figure~\ref{fig:poisson_iterations} gives the number of iterations
-needed for each case to converge to the specified tolerance as a
-function of the problem size. All computations presented in this
-section and \S\ref{subsec:sequential_comparison} were completed on a
-3.0 GHz Intel Core 2 Quad Q9650 CPU machine with 16 GB 1067 MHz DDR3
-memory.
+effectiveness of the adjoint method with the collision estimator as
+compared to the direct method. To assess both the CPU time and number
+of iterations required to converge to a solution, a problem of
+constant $\Delta$ was used with varying values of the number of mesh
+elements, fixing the spectral radius of the system at a constant value
+for each variation. Both the five-point and nine-point stencils were
+used with both the direct and adjoint solvers. For each case, $N
+\times N$ total random walk permutations were computed per MCSA
+iteration where $N \times N$ is the number of discrete grid points in
+the system. Solver parameters were set to a weight cutoff of
+\sn{1}{-4} for the stochastic linear solver and a convergence
+tolerance of \sn{1}{-8} for the MCSA iterative solver.
+Figure~\ref{fig:poisson_cpu_time} gives the CPU time needed for each
+case to converge in seconds and Figure~\ref{fig:poisson_iterations}
+gives the number of iterations needed for each case to converge to the
+specified tolerance as a function of the problem size. All
+computations presented in this section and
+\S\ref{subsec:sequential_comparison} were completed on a 3.0 GHz Intel
+Core 2 Quad Q9650 CPU machine with 16 GB 1067 MHz DDR3 memory.
 \begin{figure}[t!]
@@
 
-\begin{figure}[t!]
+\begin{figure}[ht!]
   \centering
@@
   \caption{\textbf{CPU Time (s) to converge vs. Problem Size ($N$ for
-      an $N \times N$ square mesh).} \textit{oth the Sequential Monte
+      an $N \times N$ square mesh).} \textit{Both the Sequential Monte
       Carlo and MCSA solvers are used with the five point stencils and
@@
 \label{sec:variance_reduction}
+Variance reduction is a mechanism by which the probabilities and
+weights of the Monte Carlo problem are modified to improve the
+performance of a given estimator. In many cases, such modifications
+will introduce bias into the system. However, using Monte Carlo within
+an MCSA iterative scheme provides a potential buffer for the solution
+from this bias as the correction generated by the Monte Carlo solve
+will contain large statistical error even without variance reduction.
 
@@
 \label{subsec:artificial_absorption}
+The random walk sequences generated by the adjoint Neumann Ulam
+decomposition in Eqs~(\ref{eq:adjoint_probability}) and
+(\ref{eq:adjoint_weight}) will continue on indefinitely without the
+implementation of a weight cutoff procedure (which in itself is
+effectively a form of variance reduction). This is due to the fact
+that all states to which a stochastic history may move in a given
+transition exist within the system and have a non-zero weight. A
+preliminary variance reduction scheme is to introduce
+\textit{artifical absorption} into the system such that at each
+transition event, a stochastic history will now have some finite
+probability of being terminated through absorption as well as
+transition to other states in the system. This probability, $p_{abs}$,
+modifies the probabilities and weights in the adjoint random walk
+sequence as:
+\begin{equation}
+  p_{ij} = \frac{|h_{ji}|}{\sum_i|h_{ji}|}(1-p_{abs})\:,
+  \label{eq:absorption_probabilities}
+\end{equation}
+and
+\begin{equation}
+  w_{ij} = \frac{sign(h_{ji})\sum_i|h_{ji}|}{(1-p_{abs})}\:.
+  \label{eq:absorption_weights}
+\end{equation}
+At each transition step, a history at state $i$ will then sample the
+probability distribution function generated by
+Eq~(\ref{eq:absorption_probabilities}) with an additional absorption
+state added to the distribution. If the sampling procedure results in
+a transition to the absorption state, the history is immediately
+terminated. The advantages of this for residual Monte Carlo are that
+increasing the absorption probability decreases the length of each
+random walk (resulting in speedup), while the random walk has a higher
+weight contribution at every step. This higher weight means that each
+history is depositing more information near its birth site which will
+on average be located where the system residual is the largest. An
+additional advantage of this formulation is that the full Neumann Ulam
+decomposition is maintained, resulting in Monte Carlo estimates that
+are unbiased.
+
+To observe the effects of this variance reduction on MCSA solutions,
+the transient Poisson problem was again solved with the adjoint
+method, this time with varying values of artificial
+absorption. Compare CPU time and iterative performance with a fixed
+grid.
 

### 2013-04-29 doc/dissertation/document/chapters/mc_background/mc_background.tex
adding data for estimators from hmcsa

@@
 the non-zero elements of $\ve{H}$ as those components that are zero
-will not participate in the random walk and produce identical sparsity
-patterns.
+will not participate in the random walk and thus produce identical
+sparsity patterns for all matrices.
 
@@
 walk as the Neumann Ulam decomposition defined by
-Eqs~(\ref{direct_probability}) and (\ref{eq:direct_weight}) will
+Eqs~(\ref{eq:direct_probability}) and (\ref{eq:direct_weight}) will
 create a random walk weight in Eq~(\ref{eq:direct_permutation_weight})
@@
   x_i^2\:,
+  \label{eq:direct_variance_6}
 \end{equation}
@@
 shown in Figure~\ref{fig:heat_setup}.
-\begin{figure}[t!]
+\begin{figure}[p!]
   \begin{center}
-    \scalebox{1.0}{ \input{chapters/mc_background/heat_eq_setup.pdftex_t} }
+    \scalebox{1.2}{ \input{chapters/mc_background/heat_eq_setup.pdftex_t} }
   \end{center}
@@
 gives these results.
-\begin{figure}[t!]
+\begin{figure}[p!]
   \begin{center}
@@
 \label{sec:adjoint_mc}
-Often a more useful form, an alternative for Monte Carlo matrix
+Often a more useful form, an alternative to forward Monte Carlo matrix
 inversion is the adjoint method. We begin by defining the linear
@@
 where $\ve{y}$ and $\ve{d}$ are the adjoint solution and source
-respectively and $\ve{A}^T$ is the adjoint operator for $\ve{A} \in
-\mathbb{R}^{N \times N}$. We can split this equation to mirror
-Eq~(\ref{eq:richardson_split}) by defining the following inner product
-equivalence \citep{spanier_monte_1969}:
+respectively vectors and $\ve{A}^T$ is the adjoint operator for
+$\ve{A} \in \mathbb{R}^{N \times N}$. We can split this equation to
+mirror Eq~(\ref{eq:richardson_split}): 
+\begin{equation}
+  \ve{y} = \ve{H}^T \ve{y} + \ve{d}\:.
+  \label{eq:adjoint_split_system}
+\end{equation}
+As was required for convergence with the direct method using
+Eq~(\ref{eq:richardson_split}), the spectral radius of $\ve{H}$ must
+remain less than 1 as $\ve{H}^T$ contains the same eigenvalues and
+therefore has the same spectral radius. By defining the following
+inner product equivalence \citep{spanier_monte_1969}:
 \begin{equation}
@@
 \end{equation}
-With this statement we can then define the split equation:
-\begin{equation}
-  \ve{y} = \ve{H}^T \ve{y} + \ve{d}\:.
-  \label{eq:adjoint_split_system}
-\end{equation}
-As was required for convergence with the direct method using
-Eq~(\ref{eq:richardson_split}), the spectral radius of $\ve{H}$ must
-remain less than 1 as $\ve{H}^T$ contains the same eigenvalues and
-therefore has the same spectral radius. From this definition it
-follows that:
+it follows that:
 \begin{equation}
@@
 \begin{equation}
-  P_{i_0=i}(\nu) = \frac{b_i}{||\ve{b}||_1}\:.
+  P_{i_0=i}(\nu) = \frac{|b_i|}{||\ve{b}||_1}\:.
   \label{eq:adjoint_source_probability}
 \end{equation}
- As a result of using the adjoint system, we modify our probabilities
- and weights using the \textit{adjoint Neumann-Ulam decomposition} of
+The random walk starting weight will then be $W_0 = b_{i_0}$.  As a
+result of using the adjoint system, we modify our probabilities and
+weights using the \textit{adjoint Neumann-Ulam decomposition} of
 $\ve{H}$:
@@
 where the Kronecker delta indicates that the tally contributes only in
-the current state the random walk starting weight will be $W_0 =
-b_{i_0}$. Note here that the estimator in
-Eq~(\ref{eq:adjoint_permutation_contribution}) does not have a
-dependency on the source state as in
+the current state, $i_m$, of the random walk.  Note here that the
+estimator in Eq~(\ref{eq:adjoint_permutation_contribution}) does not
+have a dependency on the source state as in
 Eq~(\ref{eq:adjoint_permutation_contribution}), providing a remedy for
@@
 therefore also providing an unbiased Monte Carlo estimate of the
-solution. It should be noted here that
+solution. Note that this expansion produces the effective sequence of
+matrix-vector multiplications with the $b_{i_0}$ component of the
+source vector which will be selected at random for each walk
+permutation and the unbiasedness of the estimator relies on an
+unbiased sampling of the source. It should also be noted here that
 Eq~(\ref{eq:adjoint_expectation_expansion}) only computes a single
@@
 \begin{equation}
-  \sigma_j^2 = \Big(\sum_{\nu} P_{\nu} \sum_{m=0}^k W_{m}^2
-  \delta_{i_m,j}\Big)_j - x_j^2\:.
+  \sigma_j^2 = \sum_{\nu} P_{\nu} \sum_{m=0}^k W_{m}^2 \delta_{i_m,j}
+  - x_j^2\:.
   \label{eq:collision_variance_1}
@@
 \end{equation}
-If we again $\ve{G} = \ve{P} \circ \ve{W} \circ \ve{W}$, we then have:
+If we again define $\ve{G} = \ve{P} \circ \ve{W} \circ \ve{W}$, we
+then have:
 \begin{equation}
@@
   \sum_{i_k}^{N}g_{i_{k},i_{k-1}} \ldots
-  g_{i_2,i_1}g_{i_1,i_0}b_{i_0}\delta_{i_k,j}^2 - x_i^2\:,
+  g_{i_2,i_1}g_{i_1,i_0}b_{i_0}\delta_{i_k,j} - x_i^2\:,
 \label{eq:collision_variance_3}
 \end{equation}
-which again contains the general Neumann series for $\ve{G}$ as given
-by Eq~(\ref{eq:variance_neumann_series}).  Inserting $T$ back into the
-variance formulation gives:
-\begin{equation}
-  \sigma^2_j = (\ve{T}\ve{b})_j - x_i^2\:.
-  \label{eq:direct_variance_5}
-\end{equation}
-We can relate $\ve{G}$ to $\ve{H}$ by noting that $\ve{G}$ simply
-contains an additional Hadamard product with the weight matrix. The
-Hadamard product has the property that:
-\begin{equation}
-  |\ve{H} \circ \ve{W}| \geq |\ve{H}|\ |\ve{W}|\:.
-  \label{eq:hadamard_inequality}
-\end{equation}
-Using the norm property of the Hadamard product and
-Eq~(\ref{eq:neumann_ulam_decomposition}), we can define the norm of
-$\ve{W}$ as:
-\begin{equation}
-  \frac{|\ve{H}|}{|\ve{P}|} \geq |\ve{W}|\:.
-  \label{eq:weight_norm}
-\end{equation}
-Choosing the infinity norm of the operator as defined in
-Eq~(\ref{eq:matrix_infinity_norm}), the row normalized probability
-matrix will yield a norm of 1 giving the following inequality for
-relating $\ve{G}$ and $\ve{H}$:
-\begin{equation}
-  |\ve{G}| \geq |\ve{H}|^2
-  \label{eq:variance_norm_inequality}
-\end{equation}
-Using these relations to analyze Eq~(\ref{eq:direct_variance_5}), we
-see that if $\rho(\ve{G}) > 1$, then the sum in
-Eq~(\ref{eq:variance_neumann_series}) will not converge and an
-infinite variance will arise as the elements of $\ve{T}$ become
-infinite in Eq~(\ref{eq:direct_variance_5}). We must restrict $\ve{G}$
-to alleviate this and therefore restrict $\ve{H}$ due to
-Eq~(\ref{eq:variance_norm_inequality}) with $\rho(\ve{H}) < 1$ so that
-our expectation values for the solution may have a finite variance.
+where now $\ve{G}$ in this case is the transpose of that used in
+Eq~(\ref{eq:direct_variance_6}). The Kronecker delta implies that the
+variance contribution from each random walk will only be in its
+current state and for many random walks, many starting weights of
+$b_{i_0}$ will contribute to the variance in the same way as they
+contribute to the solution tally. As the transpose is formed and the
+Neumann series of $\ve{G}$ is now formed in reverse order in
+Eq~(\ref{eq:collision_variance_3}), then
+Eq~(\ref{eq:collision_variance_3}) and Eq~(\ref{eq:direct_variance_6})
+are equivalent and therefore the variance of the forward and collision
+estimators are equivalent as well and bound by the same restrictions
+on $\rho(\ve{G})$ to ensure that the variance is finite.
 
@@
 where now the contribution of the iteration matrix is
-deterministically averaged at step $m$ over all potentially states $j$
+deterministically averaged at step $m$ over all potential states $j$
 that may be reached from the current state $i_m$. Via Okten, the
@@
 method. Figure~\ref{fig:adjoint_evolution} gives these results.
-\begin{figure}[t!]
+\begin{figure}[p!]
   \begin{center}
@@
 \end{subequations}
-where the correction $\boldsymbol{\delta}$ is computed by the adjoint
-Monte Carlo method. The merits of Halton's approach are immediately
-visible in that we have now broken the binding of the convergence rate
-to the Central Limit Theorem. Here, the Monte Carlo solver is used to
-produce a correction from the residual, analogous to using the
-residual to extract a correction from the search subspace in a
-projection method. By doing this, the Monte Carlo error is bound in
-the correction used to update the solution and therefore does not
-explicitly manifest itself in the overall convergence of the
-solution. The downside of such a method is that if the solution guess
-is poor, then many iterations are required in order to reach
-exponential converge as the Monte Carlo error (and therefore the
-Central Limit Theorem) does dominate in this situation.
+where the correction $\boldsymbol{\delta}^k$ is computed by the
+adjoint Monte Carlo method at each iteration. The merits of Halton's
+approach are immediately visible in that we have now broken the
+binding of the convergence rate to the Central Limit Theorem. Here,
+the Monte Carlo solver is used to produce a correction from the
+residual, analogous to using the residual to extract a correction from
+the search subspace in a projection method. By doing this, the Monte
+Carlo error is bound in the correction used to update the solution and
+therefore does not explicitly manifest itself in the overall
+convergence of the solution. The downside of such a method is that if
+the solution guess is poor, then many iterations are required in order
+to reach exponential converge as the Monte Carlo error (and therefore
+the Central Limit Theorem) does dominate in this situation.
 
@@
 diffusion equation \citep{evans_monte_2012}. With their new methods,
-they achieved identical numerical results as and marginally better
-performance than conventional Krylov solvers. Their approach was
-instead to use residual Monte Carlo as a synthetic acceleration for a
-stationary method. To derive this method, we begin by splitting the
-operator in Eq~(\ref{eq:linear_problem})
+they achieved identical numerical results as conventional Krylov
+solvers well as comparable performance in both number of iterations
+and CPU time. Their approach was instead to use residual Monte Carlo
+as a synthetic acceleration for a stationary method. To derive this
+method, we begin by splitting the operator in
+Eq~(\ref{eq:linear_problem})
 \begin{equation}
@@
 Using this, we define the following scheme that will converge in one
-iteration if $\ve{A}$ is inverted exactly.
+iteration if $\ve{A}$ is inverted exactly:
 \begin{subequations}
@@
 number of histories are required using the adjoint method to generate
-the correction.
+the correction. In addition, the fact that the scheme in
+Eq~(\ref{eq:mcsa_setup_3}) will converge in one iteration if $\ve{A}$
+is inverted exactly means that as more and more stochastic histories
+are used to compute the correction and the error is reduced towards
+zero, the number of iterations required for MCSA to converge should
+decrease accordingly, thus accelerating the solution.
 
@@
 \end{equation}
-where $\epsilon$ is a user-defined parameter. We therefore have 3
-parameters to tune in an MCSA implementation: the number of Monte
-Carlo histories computed in the adjoint solve during each MCSA
-iteration, the weight cutoff for those histories, and the total MCSA
-convergence tolerance as specified by $\epsilon$.
+where $\epsilon$ is a user-defined parameter. As with any iterative
+method, other stopping criteria using other vector norms could be
+computed, however, for this work we will only use
+Eq~(\ref{eq:mcsa_stopping_criteria}). We therefore have 3 parameters
+to tune in an MCSA implementation: the number of Monte Carlo histories
+computed in the adjoint solve during each MCSA iteration, the weight
+cutoff for those histories, and the total MCSA convergence tolerance
+as specified by $\epsilon$.
 
@@
 of the iteration matrix be less than one. Preconditioning serves as a
-means of achieving this.
+means of achieving this by altering the eigenvalue spectrum of the
+iteration matrix.
 
@@
 \label{subsubsec:basic_mcsa_preconditioning}
-As an example, to achieve this for diagonally dominant matrices point
-Jacobi left preconditioning can be used such that the preconditioning
-matrix $\ve{M}$ is:
-\begin{equation}
-  \ve{M} = diag(\ve{A})\:.
+As an example, to achieve a spectral radius of less than one for
+diagonally dominant matrices point Jacobi preconditioning can be used
+such that the preconditioning matrix $\ve{M}$ is:
+\begin{equation}
+  \ve{M} = diag(\ve{A})\:,
   \label{eq:jacobi_preconditioner}
 \end{equation}
-With the application of this preconditioner we are instead solving the
-following linear system:
+which may be trivially inverted. With the application of this
+preconditioner we are instead solving the following scaled linear
+system:
 \begin{equation}
@@
 preconditioning should be considered for many classes of problems,
-regardless of any other preconditioning that is applied to the system.
+regardless of any other preconditioning that is applied to the
+system.
 
@@
 It is possible to use general left, right, and left/right
-preconditioning with MCSA by carefully considering the underly Monte
-Carlo that will be solved with the Neumann Ulam method. We consider
-here the general left/right preconditioned method as the left or right
-preconditioned methods can be inferred from its formulation. We
-consider a left preconditioner $\ve{M_L}$ and a right preconditioner
-$\ve{M_R}$. The left/right preconditioned linear problem is then:
+preconditioning with MCSA by carefully considering the underlying
+Monte Carlo problem that will be solved with the Neumann-Ulam
+method. We consider here the general left/right preconditioned method
+as the left or right preconditioned methods can be inferred from its
+formulation. We consider a left preconditioner $\ve{M_L}$ and a right
+preconditioner $\ve{M_R}$. The left/right preconditioned linear
+problem is then:
 \begin{equation}
@@
 order to construct probabilities and weights for the Monte Carlo
-procedure, the \textit{composite operator},
-$\ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}$, must be formed via matrix-matrix
-multiplication. Several possible shortcomings of this preconditioning
-approach are readily observed. First, the matrix-matrix multiplication
-operation for sparse, parallel distributed matrices is significantly
-more expensive than a matrix-vector multiplication operation. Second,
-each preconditioner must be explicitly inverted, an operation in
-itself that my be expensive prohibits the use of any preconditioners
-which prohibit the extraction of the inverse. Third, for many modern
+procedure from the Neumann-Ulam decomposition, the \textit{composite
+  operator}, $\ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}$, must be formed via
+matrix-matrix multiplication. Several possible shortcomings of this
+preconditioning approach are readily observed. First, the
+matrix-matrix multiplication operation for sparse, parallel
+distributed matrices is significantly more expensive than a
+matrix-vector multiplication operation. Second, each preconditioner
+must be explicitly inverted, an operation in itself that may be
+expensive and which prohibits the use of any preconditioners which
+provide no mechanism to extract their inverse. Third, for many modern
 preconditioning methods, this inversion may yield dense matrices,
@@
 Core 2 Quad Q9650 CPU machine with 16 GB 1067 MHz DDR3 memory.
-\begin{figure}[t!]
+\begin{figure}[p!]
   \centering
@@
 
-\begin{figure}[t!]
+\begin{figure}[p!]
   \centering
@@
 difference between the discretizations. Both MCSA and Halton's method
-are used with the adjoint Monte Carlo solver. In order to complete the
-same study as in the previous section, the number of histories
-computed by the Monte Carlo solver at each iteration had to be doubled
-to $2 \times N \times N$ in order to ensure convergence in Sequential
-Monte Carlo Method. For the majority of the problems in the previous
-section, the Sequential method used with $N \times N$ histories would
-not converge. Figure~\ref{fig:seq_cpu_time} gives the CPU time results
-for this comparison as a function of problem size while
+are used with the adjoint Monte Carlo solver and the collision
+estimator. In order to complete the same study as in the previous
+section, the number of histories computed by the Monte Carlo solver at
+each iteration had to be doubled to $2 \times N \times N$ in order to
+ensure convergence in Sequential Monte Carlo Method. For the majority
+of the problems in the previous section, the Sequential method used
+with $N \times N$ histories would not
+converge. Figure~\ref{fig:seq_cpu_time} gives the CPU time results for
+this comparison as a function of problem size while
 Figure~\ref{fig:seq_iterations} gives the number of iterations to
-converge as a function of problem size. In both cases, using the Monte
-Carlo solver as a synthetic acceleration rather than in a pure
-residual Monte Carlo scheme resulted in a reduction in both CPU time
-and iterations required to converge. The additional Richardson
-extrapolation between each Monte Carlo solve in the MCSA method gives
-a better converged residual source to use with the Monte Carlo
-calculation while the Sequential method requires more iterations to
-achieve the same level of convergence in the residual.
-
-\begin{figure}[ht!]
+converge as a function of problem size with a convergence tolerance of
+\sn{1}{-8}. In both cases, using the Monte Carlo solver as a synthetic
+acceleration rather than in a pure residual Monte Carlo scheme
+resulted in a reduction in both CPU time and iterations required to
+converge. The additional Richardson extrapolation between each Monte
+Carlo solve in the MCSA method gives a better converged residual
+source to use with the Monte Carlo calculation while the Sequential
+method requires more iterations to achieve the same level of
+convergence in the residual.
+
+\begin{figure}[p!]
   \centering
@@
 
-\begin{figure}[ht!]
+\begin{figure}[p!]
   \centering
@@
 \end{figure}
+
+%%---------------------------------------------------------------------------%%
+\section{Monte Carlo Parameter and Estimator Analysis}
+\label{sec:parameter_estimator_analysis}
+With the adjoint method shown to be more effective than the direct
+method and MCSA to have better iterative and timing performance than
+sequential Monte Carlo, we now aim to study the effects of the adjoint
+Monte Carlo parameters, weight cutoff and number of histories, on MCSA
+performance with both the collision and expected value estimators.
 

### 2013-04-29 doc/dissertation/document/chapters/mc_background/mc_background.tex
adding data pdf files

@@
 Monte Carlo parameters, weight cutoff and number of histories, on MCSA
-performance with both the collision and expected value estimators.
+performance with both the collision and expected value
+estimators. With the same Poisson problem, we will use a $200 \times
+200$ grid and a convergence tolerance of \sn{1}{-8} for all
+calculations. To study the effects of the number of Monte Carlo
+histories per MCSA iteration, the weight cutoff was fixed at
+\sn{1}{-4} while the number of histories per iteration was varied from
+6,000 to 100,000. It was observed that MCSA would not converge for
+this problem using the collision estimator with less than 6,000
+histories while the expected value estimator permitted convergence
+with only 2,000 histories. Figure~\ref{fig:estimator_nh_iters} gives
+the number of iterations required to converge for both estimators as a
+function of the number of histories per iteration. For smaller numbers
+of histories, the performance of the expected value estimator is
+significantly better than the collision estimator. This result is
+valuable in that less transport is required to achieve the same MCSA
+iterative performance with the expected value estimator, important for
+situations where transport is expensive (i.e. in domain decomposed
+calculations). Interstingly, as the number of histories per iteration
+are increased, the iterative performance with the collision estimator
+approaches that of the expected value estimator. However, given the
+performance of the expected value estimator at a fractional number of
+histories, one should strongly consider its use over using the collision
+estimator with more histories. The CPU time required to converge is
+presented in Figure~\ref{fig:estimator_nh_time} and reflects the
+results of the iterative performance. In general, using the collision
+estimator is slightly slower overall, but the time per iteration is
+faster due to the fact that the estimator does not have to cycle
+through multiple states during the tally procedure.
+
+\begin{figure}[p!]
+  \centering
+  \includegraphics[width=5in,clip]{chapters/mc_background/estimator_nh_iters.pdf}
+  \caption{\textbf{Iterations (s) to converge vs. Monte Carlo
+      histories per MCSA iteration for a $200 \times 200$ square mesh
+      and a weight cutoff of \sn{1}{-4}.} \textit{For low numbers of
+      histories, the expected value estimator performance is
+      significantly better than the collision estimator. At higher
+      numbers of histories, the estimators become roughly
+      equivalent.}}
+  \label{fig:estimator_nh_iters}
+\end{figure}
+
+\begin{figure}[p!]
+  \centering
+  \includegraphics[width=5in,clip]{chapters/mc_background/estimator_nh_time.pdf}
+  \caption{\textbf{CPU Time (s) to converge vs. Monte Carlo histories
+      per MCSA iteration for a $200 \times 200$ square mesh and a
+      weight cutoff of \sn{1}{-4}.} \textit{For low numbers of
+      histories, the expected value estimator performance is better
+      than the collision estimator due to a lower iteration count
+      while the actual compute time per iteration is higher.}}
+  \label{fig:estimator_nh_time}
+\end{figure}
+
+For the weight cutoff study, the number of histories per iteration was
+fixed at 40,000 and the weight cutoff varied from \sn{5}{-1} down to
+\sn{1}{-10}. Surprisingly, the number of iterations to converge given
+by Figure~\ref{fig:estimator_wc_iters} is effectively invariant to the
+weight cutoff, only seeing detrimental effects on the number of
+iterations at a very large weight cutoff. This suggests that a fairly
+large weight cutoff can be used in practice with both estimators and
+that the preliminary components of the random walk are more important
+to MCSA convergence than those that occur later and with lower weight
+contributions. To further motivate using a larger weight cutoff,
+Figure~\ref{fig:estimator_wc_time} gives the CPU time need to converge
+as a function of weight cutoff. As expected, lowering the weight
+cutoff lengthens the random walk lengths and increases CPU time at no
+gain of iterative performance. In general, these results suggest using
+the expected value estimator over the collision estimator with MCSA as
+well as using a larger weight cutoff.
+
+\begin{figure}[p!]
+  \centering
+  \includegraphics[width=5in,clip]{chapters/mc_background/estimator_wc_iters.pdf}
+  \caption{\textbf{Iterations (s) to converge vs. history weight
+      cutoff for a $200 \times 200$ square mesh and 40,000 histories.}
+    \textit{For low numbers of histories, the expected value estimator
+      performance is significantly better than the collision
+      estimator. At higher numbers of histories, the estimators become
+      roughly equivalent.}}
+  \label{fig:estimator_wc_iters}
+\end{figure}
+
+\begin{figure}[p!]
+  \centering
+  \includegraphics[width=5in,clip]{chapters/mc_background/estimator_wc_time.pdf}
+  \caption{\textbf{CPU Time (s) to converge vs. history weight cutoff
+      for a $200 \times 200$ square mesh and 40,000 histories.}
+    \textit{For low numbers of histories, the expected value estimator
+      performance is better than the collision estimator due to a
+      lower iteration count while the actual compute time per
+      iteration is higher.}}
+  \label{fig:estimator_wc_time}
+\end{figure}
 
@@
 implementation of a weight cutoff procedure (which in itself is
-effectively a form of variance reduction). This is due to the fact
-that all states to which a stochastic history may move in a given
+effectively a form of biased variance reduction). This is due to the
+fact that all states to which a stochastic history may move in a given
 transition exist within the system and have a non-zero weight. A
@@
 \begin{equation}
-  p_{ij} = \frac{|h_{ji}|}{\sum_i|h_{ji}|}(1-p_{abs})\:,
+  p_{ij} = \frac{|h_{ji}|}{\sum_j|h_{ji}|}(1-p_{abs})\:,
   \label{eq:absorption_probabilities}
@@
 \begin{equation}
-  w_{ij} = \frac{sign(h_{ji})\sum_i|h_{ji}|}{(1-p_{abs})}\:.
+  w_{ij} = \frac{sign(h_{ji})\sum_j|h_{ji}|}{(1-p_{abs})}\:.
   \label{eq:absorption_weights}
@@
 the transient Poisson problem was again solved with the adjoint
-method, this time with varying values of artificial
-absorption. Compare CPU time and iterative performance with a fixed
-grid.
+method, this time with varying values of artificial absorption. A $200
+\times 200$ grid was used with 40,000 histories at every iterations
+with the expected value estimator. To reduce the effect of the weight
+cutoff on this study, the weight cutoff was set to \sn{1}{-12} such
+that it is signicantly more likely that a history will be terminated
+by absorption rather than weight cutoff. Compare CPU time and
+iterative performance with a fixed grid.
 

### 2013-04-29 doc/dissertation/document/chapters/mc_background/mc_background.tex
updated artificial absorption section

@@
 additional advantage of this formulation is that the full Neumann Ulam
-decomposition is maintained, resulting in Monte Carlo estimates that
-are unbiased.
+decomposition is maintained.
 
@@
 method, this time with varying values of artificial absorption. A $200
-\times 200$ grid was used with 40,000 histories at every iterations
-with the expected value estimator. To reduce the effect of the weight
-cutoff on this study, the weight cutoff was set to \sn{1}{-12} such
-that it is signicantly more likely that a history will be terminated
-by absorption rather than weight cutoff. Compare CPU time and
-iterative performance with a fixed grid.
+\times 200$ grid was used with 40,000 histories at every iteration
+with the expected value estimator converged to a tolerance of
+\sn{1}{-8}. To reduce the effect of the weight cutoff on this study,
+the weight cutoff was set to \sn{1}{-12} such that it is signicantly
+more likely that a history will be terminated by absorption rather
+than weight cutoff. Figure~\ref{fig:absorption_iters} gives the number
+of iterations to converge as a function of the artificial absorption
+probability. Up to a probability of roughly 0.5, the iterative
+performance is not significanltly affected with the information lost
+due to shortened random walks causing a small increase in the number
+of iterations. At a probability of 0.6, the number of iterations
+required grows rapidly as the bias creates an MCSA correction that
+pushes the solution in the wrong direction. Convergence was observed
+to be lost at probabilities of 0.7 and higher. The random walk length
+has a strong effect on CPU time as
+well. Figure~\ref{fig:absorption_time} gives the CPU time in seconds
+required to converge as a function of artificial absorption
+probability. Initially, small absorption probabilities not only
+maintain iterative performance, but also drastically reduced the time
+required to converge as absorption was more frequent than even a large
+weight cutoff but the weights themselves were not significantly
+modified. As the number of iterations grow rapidly, so does the time
+required to converge to a solution. Based on these results, using a
+small amount of artificial absorption should provide some CPU speedup
+while maintaining iterative performance. For the problem presented
+here, an absorption probability of around 0.25 may be the best choice
+as it minimizes CPU time.
+
+\begin{figure}[p!]
+  \centering
+  \includegraphics[width=5in,clip]{chapters/mc_background/absorption_iters.pdf}
+  \caption{\textbf{Iterations (s) to converge vs. artificial
+      absorption probability for a $200 \times 200$ square mesh and
+      40,000 histories.}  \textit{The addition of absorption is
+      detrimental to iterative performance due to both the shortening
+      of the random walks as well as the modification of the
+      transition weight.}}
+  \label{fig:absorption_iters}
+\end{figure}
+
+\begin{figure}[p!]
+  \centering
+  \includegraphics[width=5in,clip]{chapters/mc_background/absorption_time.pdf}
+  \caption{\textbf{CPU Time (s) to converge vs. artificial absorption
+      probability for a $200 \times 200$ square mesh and 40,000
+      histories.}  \textit{The reduced random walk length speeds up
+      the calculation with a fixed number of histories. Too much
+      artificial absorption increases iterations rapidly giving an
+      increasing CPU time.}}
+  \label{fig:absorption_time}
+\end{figure}
 

### 2013-04-30 doc/dissertation/document/chapters/mc_background/mc_background.tex
adding to the estimator variance sections

@@
 value equivalent to that of the inverted operator. Such methods have
-been in existence for decades with the earliest reference noted here
-an enjoyable manuscript published in 1950 by Forsythe and Leibler
+been in existence for decades with the earliest reference noted here a
+manuscript published in 1950 by Forsythe and Leibler
 \citep{forsythe_matrix_1950}. In their outline, Forsythe and Liebler
@@
 Monte Carlo Synthetic Acceleration methods will then be presented as a
-means of leveraging the basic Monte Carlo methods in a iterative
+means of leveraging the basic Monte Carlo methods in an iterative
 refinement scheme that accelerates their convergence. Using these
-iterative schemes, a set of modest variance reduction schemes will be
-introduced and their effects on the Monte Carlo methods explored.
+iterative schemes, a set of modest variance reduction techniques will
+be introduced and their effects on the Monte Carlo methods explored.
 
@@
 \begin{equation}
-  \sigma_i^2 = \Big(\sum_{\nu} P_{\nu} X_{\nu}^2\Big)_i - x_i^2\:,
+  \sigma_i^2 = \sum_{\nu} P_{\nu} X_{i}(\nu)^2 - x_i^2\:,
   \label{eq:direct_variance_2}
@@
 \begin{equation}
-  \sigma_i^2 = \Big(\sum_{\nu} P_{\nu} \sum_{m=0}^k W_{m}^2
-  b_{i_m}^2\Big)_i - x_i^2\:.
+  \sigma_i^2 = \sum_{\nu} P_{\nu} \Big(\sum_{m=0}^k W_{m}^2 b_{i_m}^2
+  + 2 \sum_{\substack{ j=0 \\ m<j}}^k W_m W_j b_{i_m} b_{i_j} \Big)-
+  x_i^2\:.
   \label{eq:direct_variance_3}
 \end{equation}
-Finally, expanding the transition probabilities yields the variance:
-\begin{equation}
-  \sigma_i^2 = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-  \sum_{i_k}^{N} p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
-  w^2_{i,i_1}w^2_{i_1,i_2}\ldots w^2_{i_{k-1},i_k} b_{i_k}^2 - x_i^2\:.
+We can distribute the random walk probability to yield an expanded
+form of the variance:
+\begin{equation}
+  \sigma_i^2 = \sum_{\nu} P_{\nu} \sum_{m=0}^k W_{m}^2 b_{i_m}^2 + 2
+  \sum_{\nu} P_{\nu} \sum_{\substack{ j=0 \\ m<j}}^k W_m W_j b_{i_m}
+  b_{i_j} - x_i^2\:.
+  \label{eq:direct_variance_3_2}
+\end{equation}
+In particular, we will isolate the first summation of the variance by
+expanding it:
+\begin{equation}
+  \sum_{\nu} P_{\nu} \sum_{m=0}^k W_{m}^2 b_{i_m}^2 =
+  \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots \sum_{i_k}^{N}
+  p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
+  w^2_{i,i_1}w^2_{i_1,i_2}\ldots w^2_{i_{k-1},i_k} b_{i_k}^2\:.
   \label{eq:direct_variance_4}
 \end{equation}
-Using this definition for the variance, we can arrive at a more natural
+Using this expansion, we can arrive at a more natural
 reason for enforcing $\rho(\ve{H}) < 1$ for our Monte Carlo method to
@@
 If we assign $\ve{G} = \ve{P} \circ \ve{W} \circ \ve{W}$ as in
-Eq~(\ref{eq:neumann_ulam_decomposition}), we then have a new
-formulation for the variance:
-\begin{equation}
-  \sigma^2_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-  \sum_{i_k}^{N}g_{i,i_1}g_{i_1,i_2}\ldots g_{i_{k-1},i_k} b_{i_k}^2 -
-  x_i^2\:,
+Eq~(\ref{eq:neumann_ulam_decomposition}), we then have:
+\begin{equation}
+   (\ve{P} \circ \ve{W} \circ \ve{W})^k =
+  \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
+  \sum_{i_k}^{N}g_{i,i_1}g_{i_1,i_2}\ldots g_{i_{k-1},i_k}\:,
   \label{eq:direct_variance_6}
 \end{equation}
-which contains the general Neumann series for $\ve{G}$,
+which is the general Neumann series for $\ve{G}$,
 \begin{equation}
@@
 \begin{equation}
-  \sigma^2_i = (\ve{T}\ve{b})_i - x_i^2\:.
+  \sigma^2_i = (\ve{T}\ve{b}^2)_i + 2 \sum_{\nu} P_{\nu}
+  \sum_{\substack{ j=0 \\ m<j}}^k W_m W_j b_{i_m} b_{i_j} - x_i^2\:.
   \label{eq:direct_variance_5}
@@
 Using these relations to analyze Eq~(\ref{eq:direct_variance_5}), we
-see that if $\rho(\ve{G}) > 1$, then the sum in
+see that if $\rho(\ve{G}) > 1$, then the first sum in
 Eq~(\ref{eq:variance_neumann_series}) will not converge and an
@@
 \begin{equation}
-  \sigma_j^2 = \sum_{\nu} P_{\nu} \sum_{m=0}^k W_{m}^2 \delta_{i_m,j}
-  - x_j^2\:.
+  \sigma_j^2 = \sum_{\nu} P_{\nu} \sum_{m=0}^k W_{m}^2 \delta_{i_m,j} + 2
+  \sum_{\nu} P_{\nu} \sum_{\substack{ l=0 \\ m<l}}^k W_m W_l
+  \delta_{i_m,j} \delta_{i_l,j} - x_i^2\:.
   \label{eq:collision_variance_1}
 \end{equation}
-Expanding the transition probabilities yields the same variance as the
-direct estimator, this time with the terms in reverse order and the
+In this case, in the second summation $m \neq l$ for all states due to
+the form of the summation and therefore the delta functions,
+$\delta_{i_m,j}$ or $\delta_{i_l,j}$, will only be nonzero for random
+walks that are in the current solution state (when $i_m = i_l = j$)
+and other states visited do not contribute to the variance of the
+current state (when $i_m \neq i_l$). This is important as it shows a
+decoupling of the variance among the solution vector states, meaning
+that the variance in solution state $i$ does not depend on the
+variance in solution state $j$.
+
+Expanding the transition probabilities in the first sum again yields a
+Neumman summation in the same form as that explored for the forward
+estimator, this time with the terms in reverse order and the
 introduction of the Kronecker delta:
@@
 where now $\ve{G}$ in this case is the transpose of that used in
-Eq~(\ref{eq:direct_variance_6}). The Kronecker delta implies that the
-variance contribution from each random walk will only be in its
-current state and for many random walks, many starting weights of
+Eq~(\ref{eq:direct_variance_6}). The Kronecker delta again implies
+that the variance contribution from each random walk will only be in
+its current state and for many random walks, many starting weights of
 $b_{i_0}$ will contribute to the variance in the same way as they
 contribute to the solution tally. As the transpose is formed and the
-Neumann series of $\ve{G}$ is now formed in reverse order in
-Eq~(\ref{eq:collision_variance_3}), then
-Eq~(\ref{eq:collision_variance_3}) and Eq~(\ref{eq:direct_variance_6})
-are equivalent and therefore the variance of the forward and collision
-estimators are equivalent as well and bound by the same restrictions
-on $\rho(\ve{G})$ to ensure that the variance is finite.
+Neumann series of $\ve{G}$ is in reverse order in
+Eq~(\ref{eq:collision_variance_3}) relative to its formulation in the
+forward estimator, then Eq~(\ref{eq:collision_variance_3}) and
+Eq~(\ref{eq:direct_variance_6}) are equivalent and therefore the
+collision estimator is bound by the same restrictions on
+$\rho(\ve{G})$ and $\rho(\ve{H})$ to ensure that the variance is
+finite.
 
@@
 estimator will provide no information.
+
+\subsubsection{Expected Value Estimator Variance}
+\label{subsubsec:expected_value_estimator_variance}
 
@@
   operator}, $\ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}$, must be formed via
-matrix-matrix multiplication. Several possible shortcomings of this
-preconditioning approach are readily observed. First, the
-matrix-matrix multiplication operation for sparse, parallel
-distributed matrices is significantly more expensive than a
-matrix-vector multiplication operation. Second, each preconditioner
-must be explicitly inverted, an operation in itself that may be
-expensive and which prohibits the use of any preconditioners which
-provide no mechanism to extract their inverse. Third, for many modern
-preconditioning methods, this inversion may yield dense matrices,
-destroying sparsity and further impeding the performance of a
-matrix-matrix multiplication operation. It is also interesting to note
-that the Monte Carlo problem in the general left/right preconditioned
-scheme given by Eq~(\ref{eq:left_right_correction}) is not fully
-left/right preconditioned (meaning that we do not recover $\ve{x}$),
-but instead part of a sequence for finding the substituted variable
-$\ve{u}$. We do, however, gain the benefits of this general
-preconditioning by building the iteration matrix in
+matrix-matrix multiplication.
+
+Several possible shortcomings of this preconditioning approach are
+readily observed. First, the matrix-matrix multiplication operation
+for sparse, parallel distributed matrices is significantly more
+expensive than a matrix-vector multiplication operation. Second, each
+preconditioner must be explicitly inverted, an operation in itself
+that may be expensive and which prohibits the use of any
+preconditioners which provide no mechanism to extract their
+inverse. Third, for many modern preconditioning methods, this
+inversion may yield dense matrices, destroying sparsity and further
+impeding the performance of a matrix-matrix multiplication
+operation. It is also interesting to note that the Monte Carlo problem
+in the general left/right preconditioned scheme given by
+Eq~(\ref{eq:left_right_correction}) is not fully left/right
+preconditioned (meaning that we do not recover $\ve{x}$), but instead
+part of a sequence for finding the substituted variable $\ve{u}$. We
+do, however, gain the benefits of this general preconditioning by
+building the iteration matrix in
 Eq~(\ref{eq:left_right_iteration_matrix}) from the fully

### 2013-04-30 doc/dissertation/document/chapters/mc_background/mc_background.tex
adding expected value estimator variance

@@
   \sum_{\nu} P_{\nu} \sum_{\substack{ l=0 \\ m<l}}^k W_m W_l
-  \delta_{i_m,j} \delta_{i_l,j} - x_i^2\:.
+  \delta_{i_m,j} \delta_{i_l,j} - x_j^2\:.
   \label{eq:collision_variance_1}
@@
 Expanding the transition probabilities in the first sum again yields a
-Neumman summation in the same form as that explored for the forward
+Neumman series in the same form as that explored for the forward
 estimator, this time with the terms in reverse order and the
@@
   w^2_{i_k,i_{k-1}}\ldots w^2_{i_2,i_1} w^2_{i_1,i_0}
-  b^2_{i_0}\delta_{i_k,j} - x_i^2\:.
+  b^2_{i_0}\delta_{i_k,j} - x_j^2\:.
   \label{eq:collision_variance_2}
@@
   \sum_{i_k}^{N}g_{i_{k},i_{k-1}} \ldots
-  g_{i_2,i_1}g_{i_1,i_0}b_{i_0}\delta_{i_k,j} - x_i^2\:,
+  g_{i_2,i_1}g_{i_1,i_0}b_{i_0}\delta_{i_k,j} - x_j^2\:,
 \label{eq:collision_variance_3}
@@
 \label{subsubsec:expected_value_estimator_variance}
+Following the collision estimator variance, the expected value
+estimator variance is given as:
+\begin{multline}
+  \sigma_j^2 = \sum_{\nu} P_{\nu} b_j^2 + 2 \sum_{\nu} P_{\nu} b_j
+  \sum_{m=0}^k W_{m} h_{j,i_m} +\\ \sum_{\nu} P_{\nu} \sum_{m=0}^k
+  W_{m}^2 h_{j,i_m}^2 + 2 \sum_{\nu} P_{\nu} \sum_{\substack{ l=0
+      \\ m<l}}^k W_m W_l h_{j,i_m} h_{j,i_l} - x_j^2\:.
+  \label{eq:expected_value_variance_1}
+\end{multline}
+As the delta functions are no longer present, the final summation in
+Eq~(\ref{eq:expected_value_variance_1}) contains contributions for all
+valid state combinations in the same and therefore the variance of
+each state in the solution tally is coupled to a group of other states
+as defined by the sparsity pattern of the iteration. This coupling of
+variances is expected due to the deterministic averaging used to
+generate the expected value estimator.
 

### 2013-05-02 doc/dissertation/document/chapters/mc_background/mc_background.tex
some clean up in the mc background section

@@
 method for discrete linear systems. Both the forward and adjoint
-methods will be presented and analyzed including a variance analysis
-of several Monte Carlo estimators. The Sequential Monte Carlo and
-Monte Carlo Synthetic Acceleration methods will then be presented as a
-means of leveraging the basic Monte Carlo methods in an iterative
+methods will be presented and analyzed including a brief variance
+analysis of several Monte Carlo estimators. The Sequential Monte Carlo
+and Monte Carlo Synthetic Acceleration methods will then be presented
+as a means of leveraging the basic Monte Carlo methods in an iterative
 refinement scheme that accelerates their convergence. Using these
@@
 shown in Figure~\ref{fig:heat_setup}.
-\begin{figure}[p!]
+\begin{figure}[t!]
   \begin{center}
@@
 gives these results.
-\begin{figure}[p!]
+\begin{figure}[t!]
   \begin{center}
@@
 traditional iterative method.
+\clearpage
 
@@
 where $\ve{y}$ and $\ve{d}$ are the adjoint solution and source
-respectively vectors and $\ve{A}^T$ is the adjoint operator for
+vectors respectively and $\ve{A}^T$ is the adjoint operator for
 $\ve{A} \in \mathbb{R}^{N \times N}$. We can split this equation to
@@
 introduction of the Kronecker delta:
-\begin{equation}
+\begin{multline}
   \sigma_j^2 = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
@@
   w^2_{i_k,i_{k-1}}\ldots w^2_{i_2,i_1} w^2_{i_1,i_0}
-  b^2_{i_0}\delta_{i_k,j} - x_j^2\:.
+  b^2_{i_0}\delta_{i_k,j} + \\ 2 \sum_{\nu} P_{\nu} \sum_{\substack{
+      l=0 \\ m<l}}^k W_m W_l \delta_{i_m,j} \delta_{i_l,j} - x_j^2\:.
   \label{eq:collision_variance_2}
-\end{equation}
+\end{multline}
 If we again define $\ve{G} = \ve{P} \circ \ve{W} \circ \ve{W}$, we
 then have:
-\begin{equation}
+\begin{multline}
   \sigma^2_j = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
   \sum_{i_k}^{N}g_{i_{k},i_{k-1}} \ldots
-  g_{i_2,i_1}g_{i_1,i_0}b_{i_0}\delta_{i_k,j} - x_j^2\:,
+  g_{i_2,i_1}g_{i_1,i_0}b_{i_0}\delta_{i_k,j} + \\ 2 \sum_{\nu}
+  P_{\nu} \sum_{\substack{ l=0 \\ m<l}}^k W_m W_l \delta_{i_m,j}
+  \delta_{i_l,j} - x_j^2\:,
 \label{eq:collision_variance_3}
-\end{equation}
+\end{multline}
 where now $\ve{G}$ in this case is the transpose of that used in
@@
 Eq~(\ref{eq:expected_value_variance_1}) contains contributions for all
-valid state combinations in the same and therefore the variance of
-each state in the solution tally is coupled to a group of other states
-as defined by the sparsity pattern of the iteration. This coupling of
-variances is expected due to the deterministic averaging used to
-generate the expected value estimator.
+valid state combinations in the same row of the iteration matrix and
+therefore the variance of each state in the solution tally is coupled
+to a group of other states as defined by the sparsity pattern of the
+iteration matrix. This coupling of variances is expected due to the
+deterministic averaging used to generate the expected value estimator.
 
@@
 method. Figure~\ref{fig:adjoint_evolution} gives these results.
-\begin{figure}[p!]
+\begin{figure}[t!]
   \begin{center}
@@
 to the tallies in those states.
+\clearpage
 
@@
 of his work by the transport community have confirmed convergence
-rates on the order of of $exp(-N)$ \citep{evans_residual_2003}. In
+rates on the order of of $e^{-N}$ \citep{evans_residual_2003}. In
 much the same way as projection methods, Halton's method, sequential
@@
 Core 2 Quad Q9650 CPU machine with 16 GB 1067 MHz DDR3 memory.
-\begin{figure}[p!]
+\begin{figure}[t!]
   \centering
@@
 
-\begin{figure}[p!]
+\begin{figure}[t!]
   \centering
@@
 Monte Carlo calculation.
-
-\begin{figure}[p!]
+\begin{figure}[t!]
   \centering
@@
 \end{figure}
+
+\clearpage
 
@@
 
-\subsection{Weight Capping}
-\label{subsec:weight_capping}
-
 \subsection{Reduced Domain Approximation}
 \label{subsec:reduced_domain_approximation}
+
+\clearpage

### 2013-05-02 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
updating spn chapter

@@
 ordinates. Why did this happen? A significant problem, pointed out by
-Larsen, was that little rigor had been applied to the formulation of
-the $SP_N$ equations since their derivation through primarily
-heuristic arguments. Instead, studies at that time focused on simply
-comparing the results of the method to other contemporary transport
-solution strategies. In addition, many problems of interest from the
-literature at the time were either solved using nodal-type methods for
-reactor-sized problems or $S_N$-type methods for benchmark problems
-with intricate material configurations and potentially large flux
-gradients over small spatial domains.
+Brantley and Larsen \citep{brantley_simplified_2000}, was that little
+rigor had been applied to the formulation of the $SP_N$ equations
+since their derivation through primarily heuristic arguments. Instead,
+studies at that time focused on simply comparing the results of the
+method to other contemporary transport solution strategies. In
+addition, many problems of interest from the literature at the time
+were either solved using nodal-type methods for reactor-sized problems
+or $S_N$-type methods for benchmark problems with intricate material
+configurations and potentially large flux gradients over small spatial
+domains.
 
@@
 permitted generation of the $SP_N$ system of equations for detailed
-full core reactor models. By fully forming these equations and
-formulating them as a linear algebra problem instead of using the
-explicit iterative methods of the past, we now have access to all of
-the modern advancements in computational linear algebra including
-Krylov solvers for asymmetric systems and preconditioning methods such
-as algebraic multigrid. This leads us to then explore the
-applicability of our work in discrete Monte Carlo methods for linear
-systems as a possible solution method for the $SP_N$ equations. If
-formulated correctly, we hypothesize that signficant improvement in
-usage of computational resources may be observed compared to modern
-solution techniques such as those suggested due to the form of the
-matrices generated by the discretization. In addition, solving the
-$SP_N$ equations in this way also breaks away from the $S_N$ forms of
-parallelism where spatial parallelism is achieved by an efficient
-parallel sweep, angular efficiency achieved by pipelining, and energy
-parallelism achieved by decoupling the groups. With the $SP_N$
-equations as a full matrix system, we now can parallelize the problem
-as prescribed by the linear solver, which may be significantly more
-scalable than current $S_N$ transport practices.
+full core reactor models \citep{evans_simpli_2013}. By fully forming
+these equations and formulating them as a linear algebra problem
+instead of using the explicit iterative methods of the past, we now
+have access to all of the modern advancements in computational linear
+algebra including Krylov solvers for asymmetric systems and
+preconditioning methods such as algebraic multigrid. This leads us to
+then explore the applicability of our work in discrete Monte Carlo
+methods for linear systems as a possible solution method for the
+$SP_N$ equations. If formulated correctly, we hypothesize that
+signficant improvement in usage of computational resources may be
+observed compared to modern solution techniques such as those
+suggested due to the form of the matrices generated by the
+discretization. In addition, solving the $SP_N$ equations in this way
+also breaks away from the $S_N$ forms of parallelism where spatial
+parallelism is achieved by an efficient parallel sweep, angular
+efficiency achieved by pipelining, and energy parallelism achieved by
+decoupling the groups. With the $SP_N$ equations as a full matrix
+system, we now can parallelize the problem as prescribed by the linear
+solver, which may be significantly more scalable than current $S_N$
+transport practices.
 
@@
 work of Evans, in order to gain full understanding of the underlying
-system and its behavior in a Monte Carlo context. We begin by stating
-the general time-independent neutron transport equation followed by a
-derivation of the $P_N$ equations in planar geometry for multiple
-energy groups. From these equations, we then apply a set of
-approximations to yield the multi-dimensional, multi-group $SP_N$
-equations for fixed source problems. Boundary conditions are also
-discussed.
+system and its potential behavior in a Monte Carlo context. From the
+$P_N$ equations, we then apply a set of approximations to yield the
+multi-dimensional, multi-group $SP_N$ equations for fixed source
+problems with boundary conditions are also discussed\footnote{The
+  $SP_N$ derivations in this chapter are heavily based on those
+  presented by Evans in \citep{evans_simpli_2013}}. Using the
+fully-formed matrix for the transport problem, we explore solutions to
+the $SP_N$ equations using Monte Carlo Synthetic Acceleration using a
+light water reactor fuel assembly criticality calculation as the
+driving problem.
 
@@
 %%---------------------------------------------------------------------------%%
-\section{Derivation of the $P_N$ Equations}
-\label{sec:pn_equations}
-Next, we derive the $P_N$ equations, a simplified form of the general
-transport equation where Legendre polynomials are used to expand the
-angular flux and scattering cross section variables as a means of
-capturing the angular structure of the solution. Before deriving this
-form of the transport equation, we briefly discuss a few properties of
-Legendre polynomials that we will find useful in the derivation.
-
-\subsection{Legendre Polynomials}
-\label{subsec:legendre_polys}
-The Legendre polynomials are an orthogonal set of functions that
-are solutions to Legendre's differential equation. They have the
-following form \citep{lewis_computational_1993}:
-\begin{equation}
-  P_l(\mu) = \frac{1}{2^l l!}\frac{d^l}{d \mu^l}(\mu^2-1)^l\:.
-  \label{eq:general_legendre_poly}
-\end{equation}
-These functions have several useful properties including
-orthogonality:
-\begin{equation}
-  \int_{-1}^{1} P_l(\mu) P_{l'}(\mu) d\mu = \frac{1}{2l+1}\delta_{l l'}\:,
-  \label{eq:legendre_orthog}
-\end{equation}
-a recurrence relation:
-\begin{equation}
-  \mu P_l(\mu) = \frac{1}{2l+1}[(l+1)P_{l+1}(\mu) + l P_{l-1}(\mu)]\:,
-  \label{eq:legendre_recurrence}
-\end{equation}
-and an addition theorem:
-\begin{equation}
-  P_l(\hat{\Omega} \cdot \hat{\Omega}') = \frac{1}{2n+1}\sum_{m=-l}^l
-  Y_{lm}(\hat{\Omega})Y^*_{lm}(\hat{\Omega}')\:,
-  \label{eq:legendre_addition}
-\end{equation}
-where the functions $Y_{lm}(\hat{\Omega})$ are the spherical
-harmonics. We can form the addition theorem in this way because the
-spherical harmonics are in fact just harmonic multiples of the
-Legendre polynomials:
-\begin{equation}
-  Y_{lm}(\hat{\Omega}) =
-  \sqrt{\frac{(2l+1)(l-m)!}{(l+m)!}}P^m_l(\mu)e^{i \omega m}\:,
-  \label{eq:spherical_harmonic}
-\end{equation}
-where $\omega$ is the azimuthal component of the streaming
-direction. We can reduce Eq~(\ref{eq:legendre_addition}) for the
-planar geometry we are studying by ignoring the azimuthal components
-of the addition theorem. As shown in Eq~(\ref{eq:spherical_harmonic}),
-the azimuthal dependence is given by the harmonic component, $e^{i
-  \omega m}$, and therefore we choose to ignore all terms in
-Eq.~(\ref{eq:legendre_addition}) where $m \neq 0$. This gives:
-\begin{equation}
-  P_l(\hat{\Omega} \cdot \hat{\Omega}') = \frac{1}{2n+1}
-  Y_{l0}(\hat{\Omega})Y^*_{l0}(\hat{\Omega}')\:.
-  \label{eq:legendre_addition_2}
-\end{equation}
-Per Eq~(\ref{eq:spherical_harmonic}) we have:
-\begin{equation}
-  Y_{l0}(\hat{\Omega}) = \sqrt{2l+1}P_l^0(\mu)\:,
-  \label{eq:harmonic_0}
-\end{equation}
-where $P_l^0(\mu) = P_l(\mu)$ is the $0^{th}$ associated Legendre
-function. Finally, we can reduce the addition theorem from
-Eq~(\ref{eq:legendre_addition_2}) with Eq~(\ref{eq:harmonic_0}) to a
-simple product for planar geometry:
-\begin{equation}
-  P_l(\hat{\Omega} \cdot \hat{\Omega}') = P_l(\mu)P_l(\mu')\:.
-  \label{eq:legendre_addition_3}
-\end{equation}
-
-\subsection{Planar $P_N$ Equations}
-\label{subsec:planar_pn}
-With the Legendre polynomial properties defined above, we can proceed
-by deriving the $P_N$ equations for planar geometry. For these
-equations, we will start by assuming a monoenergtic field of neutrons
-such that we are solving the following reduced form of the transport
-equation: 
-\begin{equation}
-  \mu \frac{\partial}{\partial x} \psi(x,\mu) + \sigma(x) \psi(x,\mu)
-  = \int \sigma_s(x,\hat{\Omega}' \rightarrow \hat{\Omega})
-  \psi(\vec{r},\hat{\Omega}') d\Omega' + \frac{q(x)}{4 \pi}\:.
-  \label{eq:mono_transport}
-\end{equation}
-The $P_N$ equations introduce the approximation that the angular
-dependence of the scattering cross section, $\sigma_s$, and the angular
-flux $\psi$, can be discretized by expanding them in Legendre
-polynomials as follows:
-\begin{equation}
-  \psi(x,\mu) = \sum_{n=0}^\infty (2n+1)P_n(\mu)\phi_n(x)\:,
-  \label{eq:flux_expansion}
-\end{equation}
-\begin{equation}
-  \sigma_{sm}(x) = \sum_{m=0}^\infty (2m+1)P_m(\mu)\sigma_s(x)\:,
-  \label{eq:scattering_expansion}
-\end{equation}
-where we have supressed the $2\pi$ generated by integrating away the
-azimuthal angular component and $\phi_n(x)$ in
-Eq~(\ref{eq:flux_expansion}) is referred to as the $n^{th}$ Legendre
-moment of the neutron flux and is given by:
-\begin{equation}
-  \phi_n(x) = \int_{-1}^1 P_n(\mu)\psi(x,\mu)d\mu\:.
-  \label{eq:legendre_moments}
-\end{equation}
-We first insert the expansions given by Eq~(\ref{eq:flux_expansion})
-and (\ref{eq:scattering_expansion}) into the planar transport equation
-given by Eq~(\ref{eq:mono_transport}):
-\begin{multline}
-  \frac{\partial}{\partial x}\Big[\sum_{n=0}^\infty (2n+1) \phi_n \mu
-    P_n(\mu) \Big] + \sigma \sum_{n=0}^\infty (2n+1) \phi_n P_n(\mu) =
-  \\ \int_{-1}^1 \sum_{m=0}^\infty (2m+1) \sigma_{sm} P_m(\mu_0)
-  \sum_{n=0}^\infty (2n+1) \phi_n P_n(\mu') d \mu' + q\:,
-  \label{eq:pn_deriv_1}
-\end{multline}
-where the dependence on the spatial variable, $x$, has been
-supressed. To arrive at the $P_N$ equations we multiply
-Eq~(\ref{eq:pn_deriv_1}) by $P_m(\mu)$ and integrate over the angular
-domain $\int_{-1}^1 d \mu$. We will look at each term in
-Eq~(\ref{eq:pn_deriv_1}) individually.
-
-\paragraph{Streaming Term}
-We first apply the multiplication and integration as prescribed above:
-\begin{equation}
-  \frac{\partial}{\partial x}\Bigg[\sum_{n=0}^\infty (2n+1) \mu \phi_n
-    P_n(\mu) \Bigg] \rightarrow \int_{-1}^1 \frac{\partial}{\partial
-    x}\Bigg[\sum_{n=0}^\infty (2n+1) \phi_n \mu P_n(\mu) P_m(\mu) \Bigg]
-  d \mu\:.
-  \label{eq:pn_deriv_2}
-\end{equation}
-The $\mu P_n(\mu)$ term can be eliminated via the recurrence relation
-given by Eq~(\ref{eq:legendre_recurrence}):
-\begin{equation}
-\int_{-1}^1 \frac{\partial}{\partial x}\Bigg[\sum_{n=0}^\infty (2n+1)
-  \frac{\phi_n}{2n+1}[(n+1)P_{n+1}(\mu) + n P_{n-1}(\mu)] P_m(\mu)
-  \Bigg] d\mu\:,
-\label{eq:pn_deriv_3}
-\end{equation}
-which expands to:
-\begin{equation}
-  \sum_{n=0}^\infty \frac{\partial}{\partial
-    x}\phi_n\Bigg[(n+1)\int_{-1}^1 P_{n+1}(\mu)P_m(\mu) d\mu + n
-    \int_{-1}^1 P_{n-1}(\mu) P_m(\mu) d\mu \Bigg] \:.
-  \label{eq:pn_deriv_4}
-\end{equation}
-This reveals the orthogonality relation given by
-Eq~(\ref{eq:legendre_orthog}) that when inserted into
-Eq~(\ref{eq:pn_deriv_3}):
-\begin{equation}
-  \sum_{n=0}^\infty \frac{\partial}{\partial
-    x}\phi_n\Bigg[(n+1)\frac{1}{2n+1}\delta_{n,n+1} +
-    n\frac{1}{2n+1}\delta_{n,n-1} \Bigg] \:.
-  \label{eq:pn_deriv_5}
-\end{equation}
-We can then distribute the Legendre moment to arrive at the final form
-of the streaming term:
-\begin{equation}
-  \sum_{n=0}^\infty \frac{\partial}{\partial x} \frac{1}{2n+1} \Big[
-    (n+1) \phi_{n+1} + n \phi_{n-1} \Big] \:.
-  \label{eq:pn_deriv_6}
-\end{equation}
-
-\paragraph{Collision Term}
-To reduce the collision term, the orthogonality relation is again
-applied after the integration:
-\begin{equation}
-  \sigma \sum_{n=0}^\infty (2n+1) \phi_n P_n(\mu) \rightarrow
-  \int_{-1}^1 \sigma \sum_{n=0}^\infty (2n+1) \phi_n P_n(\mu) P_m(\mu)
-  d\mu\:,
-  \label{eq:pn_deriv_7}
-\end{equation}
-\begin{equation}
-  \sigma \sum_{n=0}^\infty (2n+1) \phi_n \frac{1}{2n+1}\:,
-  \label{eq:pn_deriv_8}
-\end{equation}
-giving for the final collision term:
-\begin{equation}
-  \sum_{n=0}^\infty \sigma \phi_n \:.
-  \label{eq:pn_deriv_9}
-\end{equation}
-
-\paragraph{Scattering Term}
-For the scattering term:
-\begin{multline}
-  \int_{-1}^1 \sum_{m=0}^\infty (2m+1) \sigma_{sm} P_m(\mu_0)
-  \sum_{n=0}^\infty (2n+1) \phi_n P_n(\mu') d\mu' \rightarrow\\
-  \int_{-1}^1 \int_{-1}^1 \sum_{m=0}^\infty (2m+1) \sigma_{sm} P_m(\mu)
-  P_m(\mu_0) \sum_{n=0}^\infty (2n+1) \phi_n P_n(\mu') d\mu' d\mu\:.
-  \label{eq:pn_deriv_10}
-\end{multline}
-The addition thereom from Eq~(\ref{eq:legendre_addition_3}) is applied
-to give:
-\begin{equation}
-  \int_{-1}^1 \int_{-1}^1 \sum_{m=0}^\infty (2m+1) \sigma_{sm} P_m(\mu)
-  P_m(\mu)P_m(\mu') \sum_{n=0}^\infty (2n+1) \phi_n P_n(\mu') d\mu' d\mu\:,
-  \label{eq:pn_deriv_11}
-\end{equation}
-which can be rearranged as:
-\begin{equation}
-  \sum_{m=0}^\infty \sum_{n=0}^\infty (2m+1) \sigma_{sm} (2n+1) \phi_n
-  \int_{-1}^1 P_m(\mu') P_n(\mu') d\mu' \int_{-1}^1 P_m(\mu) P_m(\mu)
-  d\mu\:.
-  \label{eq:pn_deriv_12}
-\end{equation}
-Again we can apply orthogonality to eliminate the Legendre polynomials:
-\begin{equation}
-  \sum_{m=0}^\infty \sum_{n=0}^\infty (2m+1) \sigma_{sm} (2n+1) \phi_n
-  \frac{1}{2n+1}\delta_{nm}\frac{1}{2m+1}\delta_{mm}\:,
-  \label{eq:pn_deriv_13}
-\end{equation}
-which is reduced to:
-\begin{equation}
-  \sum_{n=0}^\infty \sigma_{sn} \phi_n\:,
-  \label{eq:pn_deriv_14}
-\end{equation}
-with the dependence on the index $m$ eliminated.
-
-\paragraph{Source Term}
-The last term we are concerned with in Eq~(\ref{eq:pn_deriv_1}) is the
-source term:
-\begin{equation}
-  q \rightarrow \int_{-1}^1 q P_n(\mu) d\mu \:.
-  \label{eq:pn_deriv_15}
-\end{equation}
-We can leverage orthogonality by multiplying by $P_0(\mu) = 1$:
-\begin{equation}
-  \int_{-1}^1 q P_0 P_n(\mu) d\mu = \frac{q}{2*0+1}\delta_{n0}\:,
-  \label{eq:pn_deriv_16}
-\end{equation}
-giving a final source term of:
-\begin{equation}
-  q\delta_{n0}\:.
-  \label{eq:pn_deriv_17}
-\end{equation}
-
-Now that we have expanded all angular dependent terms in
-Eq~(\ref{eq:pn_deriv_1}) and reduced them appropriately, we can
-combine them to generate the $P_N$ equations:
-\begin{equation}
-    \sum_{n=0}^\infty \frac{\partial}{\partial x} \frac{1}{2n+1} \Big[
-      (n+1) \phi_{n+1} + n \phi_{n-1} \Big] + \sum_{n=0}^\infty \sigma
-    \phi_n = \sum_{n=0}^\infty \sigma_{sn} \phi_n + q\delta_{n0}\:.
-  \label{eq:pn_deriv_18}
-\end{equation}
-More formally, the $P_N$ equations are written as:
-\begin{equation}
-   \frac{1}{2n+1} \frac{\partial}{\partial x}\Big[ (n+1) \phi_{n+1} + n
-     \phi_{n-1} \Big] + \Sigma_n \phi_n = q\delta_{n0} \:,
-  \label{eq:final_pn_equations}
-\end{equation}
-where $\Sigma_n = \sigma-\sigma_{sn}$ and the summations are truncated
-at some level of approximation $N$ such that $n = 0,1,\dotsc,N$. This
-yields a set of $N+1$ equations for $N+2$ flux moments. We therefore
-require an additional equation to close the system. In accordance with
-the series truncation as an approximation we choose the last moment in
-the expansion to be zero:
-\begin{equation}
-  \phi_{N+1} = 0\:.
-  \label{eq:pn_closure}
-\end{equation}
-As an example, we will construct the P5 equations from
-Eq~(\ref{eq:final_pn_equations}) and the closure given by
-Eq~(\ref{eq:pn_closure}):
-\begin{subequations}
-  \begin{gather}
-   \frac{\partial}{\partial x}\phi_{1} + \Sigma_0 \phi_0 = q\:,\\ 
-   \frac{1}{3} \frac{\partial}{\partial x}\Big[ 2
-     \phi_{2} + \phi_{0} \Big] + \Sigma_1 \phi_1 = 0\:,\\
-   \frac{1}{5} \frac{\partial}{\partial x}\Big[ 3 \phi_{3} + 2
-     \phi_{1} \Big] + \Sigma_2 \phi_2 = 0 \:,\\
-   \frac{1}{7} \frac{\partial}{\partial x}\Big[ 4 \phi_{4} + 3
-     \phi_{2} \Big] + \Sigma_3 \phi_3 = 0 \:,\\
-   \frac{1}{9} \frac{\partial}{\partial x}\Big[ 5 \phi_{5} + 4
-     \phi_{3} \Big] + \Sigma_4 \phi_4 = 0 \:,\\
-   \frac{1}{11} \frac{\partial}{\partial x} 5 \phi_{4} + \Sigma_5
-   \phi_5 = 0 \:.
-  \end{gather}
-  \label{eq:p5_equations}
-\end{subequations}
-This gives us a set of 6 coupled equations for the six Legendre
-moments requested defined over the entire spatial domain for a single
-energy group. In practice, only odd-numbered $P_N$ orders are
-generally used \citep{lewis_computational_1993}. This is due to the
-fact that using odd $N$ yields an even number of $N+1$ equations which
-can be split evenly on the left and right boundaries of the problem to
-facilitate the description of boundary conditions. We will choose this
-convention when deriving the boundary conditions.
-
-\subsection{Boundary Conditions for the $P_N$ Equations}
-\label{subsbusec:bcs_pn}
-Per the analysis in \citep{lewis_computational_1993}, two types of
-boundary conditions will be discussed: reflecting and Marshak. Marshak
-boundary conditions can be used to specify both vacuum conditions and
-isotropic source conditions on the boundary.
-
-\paragraph{Reflecting Boundary Conditions}
-In this case, the incoming flux should be equivalent to the outgoing
-flux at the boundary point $x_b$:
-\begin{equation}
-  \psi(x_b,\mu) = \psi(x_b,-\mu)\:.
-  \label{eq:reflecting_condition}
-\end{equation}
-Given the legendre expansion for the flux defined in
-Eq~(\ref{eq:flux_expansion}) and the Legendre polynomial property that
-$P_n(\mu) = (-1)^n P_n(-\mu)$, the condition specified by
-Eq~(\ref{eq:reflecting_condition}) can be satisfied if
-\begin{equation}
-  \phi_n = 0,\ \forall \ \text{odd}\ n\:,
-  \label{eq:reflecting_condition_odd}
-\end{equation}
-as all even $n$ yield $P_n(\mu) = P_n(-\mu)$ and therefore an
-equivalent reflecting condition for the flux moments.
-
-\paragraph{Marshak Boundary Conditions}
-The Marshak conditions come directly from the Legendre moments of the
-flux:
-\begin{equation}
-  \int_{\mu_b} P_i(\mu) \psi(\mu) d\mu = \int_{\mu_b} P_i(\mu)
-  \psi_b(\mu) d\mu\ \ \ \ \ \ \ \ \ \ \text{for $i=1,3,...,N$}\:,
-  \label{eq:general_marshak}
-\end{equation}
-where $\psi_b(\mu)$ is the prescribed angular flux on the boundary of
-interest and $\mu_b$ the angular domain defined by the boundary. To
-discretize this condition, we again insert the angular flux expansions
-from Eq~(\ref{eq:flux_expansion}) for the fluxes defined in the
-domain:
-\begin{equation}
-  \int_{\mu_b} P_i(\mu) \sum_{n=0}^N (2n+1) \phi_n P_n(\mu) d\mu =
-  \int_{\mu_b} P_i(\mu) \psi_b(\mu) d\mu\:.
-  \label{eq:marshak_expanded}
-\end{equation}
-The boundary flux, $\phi_b(\mu)$ is assumed to known and therefore
-Eq~(\ref{eq:marshak_expanded}) defines a set of $(N+1)/2$ equations to
-be solved on each boundary in the planar case, closing the system in
-the spatial domain.
-
-As an example of applying the Marshak conditions, consider the $P_3$
-case with an isotropic boundary source $\phi_b$ on the left side of
-the domain. In this case, the angular domain over which the boundary
-flux is defined will be $\mu_b \in [0,1]$, giving the bounds of
-integration. We first expand the summation for $i=1$:
-\begin{equation}
-  \int_0^1 \mu \Bigg[ \phi_0 + 3\phi_1\mu +
-    \frac{5}{2}\phi_2(3\mu^2-1) + \frac{7}{2}\phi_3(5\mu^3-3\mu)
-    \Bigg] d\mu = \int_0^1 \mu \phi_b d\mu\:,
-  \label{eq:marshak_p1_deriv_1}
-\end{equation}
-and then $i=3$:
-\begin{equation}
-  \int_0^1 \frac{1}{2}(5\mu^3-3\mu) \Bigg[ \phi_0 + 3\phi_1\mu +
-    \frac{5}{2}\phi_2(3\mu^2-1) + \frac{7}{2}\phi_3(5\mu^3-3\mu)
-    \Bigg] d\mu = \int_0^1 \frac{1}{2}(5\mu^3-3\mu) \phi_b d\mu\:.
-  \label{eq:marshak_p3_deriv_1}
-\end{equation}
-Expanding the polynomials in $\mu$ and carrying out the simple
-integration then gives 2 equations for the left hand boundary:
-\begin{equation}
-  \phi_0 + 2\phi_1 + \frac{5}{4}\phi_2 = \phi_b\:,
-  \label{eq:marshak_p1_deriv_2}
-\end{equation}
-\begin{equation}
-  \phi_0 - 5\phi_2 - 8\phi_3 = \phi_b\:.
-  \label{eq:marshak_p1_deriv_3}
-\end{equation}
-The right hand side boundary condition will yield 2 complementary
-equations if Marshak conditions are used or 2 non-zero moments to be
-solved for if reflected conditions are used. The formulation above
-also holds for vacuum conditions where $\phi_b = 0$.
-
-%%---------------------------------------------------------------------------%%
-\section{Derivation of the $SP_N$ Equations}
+\section{Derivation of the Monoenergetic $SP_N$ Equations}
 \label{sec:spn_equations}
-The $P_N$ equations give $N+1$ coupled first-order equations
-capturing the spatial and angular-dependence of the solution. In
-multiple dimensions, the equation set becomes large and coupled not
-only through angular moments but also through the spatial
-variables. As a simpler alternative to multidimensional $P_N$
-solutions, Gelbard recognized in 1960 that the planar $P_N$ equations
-could be simplified and applied an ad-hoc method to extend them to
-multiple dimensions, yielding the $SP_N$ equations. These equations
-are not only fewer in number, but also take on a diffusion-like form
-while maintaining the angular character of the flux, making them
-amenable to solutions with modern diffusion methods.
+The $P_N$ equations as derived in Appendix~\ref{chap:pn_equations}
+give $N+1$ coupled first-order equations capturing the spatial and
+angular-dependence of the solution. In multiple dimensions, the
+equation set becomes large and coupled not only through angular
+moments but also through the spatial variables. As a simpler
+alternative to multidimensional $P_N$ solutions, Gelbard recognized in
+1960 that the planar $P_N$ equations could be simplified and applied
+an ad-hoc method to extend them to multiple dimensions, yielding the
+$SP_N$ equations. These equations are not only fewer in number, but
+also take on a diffusion-like form while maintaining the angular
+character of the flux, making them amenable to solutions with modern
+diffusion methods.
 
@@
 yielding a multidimensional set of $(N+1)/1$ angular coupled equations
-defined as the $SP_N$ equations. Again, we provide closure to this set
-of equations with $\phi_{N+1} = 0$. As a concrete example, we will
-consider the $SP_7$ equations:
+defined as the $SP_N$ equations. As with the $P_N$ equations, we
+provide closure to this set of equations with $\phi_{N+1} = 0$. As a
+concrete example, we will consider the $SP_7$ equations:
 \begin{subequations}
@@
 \end{subequations}
-If we rearrange the Eq~(\ref{eq:spn7_subs_equations} such that only
+If we rearrange the Eq~(\ref{eq:spn7_subs_equations}) such that only
 one divergence operation is present in each equation, we can formulate
@@
 'psuedo-moment' $u_n$. This completes the derivation of the $SP_7$
-equations for a single energy group. As noted by Evans, lower order
-$SP_N$ approximations can be generated by setting higher order even
-moments in this system to zero (e.g. $\phi_6 = \phi_4 = 0$ yields the
-$SP_3$ equations).
+equations for a single energy group. As noted by Evans
+\citep{evans_simpli_2013}, lower order $SP_N$ approximations can be
+generated by setting higher order even moments in this system to zero
+(e.g. $\phi_6 = \phi_4 = 0$ yields the $SP_3$ equations).
 
@@
 We approach the formulation of the boundary conditions in much the
-same way as we did for the $P_N$ equations with both reflecting and
-Marshak conditions possible. To begin, we perform the expansion in
+same way as for the $P_N$ equations with both reflecting and Marshak
+conditions possible. To begin, we perform the expansion in
 Eq~(\ref{eq:marshak_expanded}) for the left side of the planar system,

### 2013-05-02 doc/dissertation/document/backmatter/pn_equations.tex
adding pn and spn appendices

--- 2013-03-04 doc/spn/spn.tex
@@
-%        File: spn.tex
-\documentclass[letterpaper,12pt]{article}
-\usepackage[top=1.0in,bottom=1.0in,left=1.25in,right=1.25in]{geometry}
-\usepackage{verbatim}
-\usepackage{amssymb}
-\usepackage{graphicx}
-\usepackage{longtable}
-\usepackage{amsfonts}
-\usepackage{amsmath}
-\usepackage{amsthm}
-\usepackage[usenames]{color}
-\usepackage[
-naturalnames = true, 
-colorlinks = true, 
-linkcolor = black,
-anchorcolor = black,
-citecolor = black,
-menucolor = black,
-urlcolor = blue
-]{hyperref}
-\usepackage{listings}
-\usepackage{textcomp}
-\definecolor{listinggray}{gray}{0.9}
-\definecolor{lbcolor}{rgb}{0.9,0.9,0.9}
-\lstset{
-  backgroundcolor=\color{lbcolor},
-  tabsize=4,
-  rulecolor=,
-  language=c++,
-  basicstyle=\scriptsize,
-  upquote=true,
-  aboveskip={1.5\baselineskip},
-  columns=fixed,
-  showstringspaces=false,
-  extendedchars=true,
-  breaklines=true,
-  prebreak =
-  \raisebox{0ex}[0ex][0ex]{\ensuremath{\hookleftarrow}},
-  frame=single,
-  showtabs=false,
-  showspaces=false,
-  showstringspaces=false,
-  identifierstyle=\ttfamily,
-  keywordstyle=\color[rgb]{0,0,1},
-  commentstyle=\color[rgb]{0.133,0.545,0.133},
-  stringstyle=\color[rgb]{0.627,0.126,0.941},
-}
-
-\author{Stuart R. Slattery
-  \\ \href{mailto:sslattery@wisc.edu}{\texttt{sslattery@wisc.edu}}
-}
-
-\date{\today}
-\title{A Monte Carlo Synthetic Acceleration Method for the $SP_N$ Equations}
-\begin{document}
-\maketitle
-
 %%---------------------------------------------------------------------------%%
-\section{Introduction}
-\label{sec:introduction}
-The neutron transport problem is complicated. Solutions cover a large
-phase space and the problems of interest are often geometrically
-complex, very large, or both, requiring tremendous computational
-resources to generate an adequate solution. Modern deterministic
-methods for large scale problems are commonly variants on the discrete
-ordinates ($S_N$) method \cite{evans_denovo:_2010}. For full reactor
-core neutronics simulations, the $S_N$ method requires potentially
-trillions of unknown angular flux moments to be computed to achieve
-good accuracy for the repsonses of interest. Other forms of the
-transport problem, including the $P_N$ method, take on a simpler form
-than the more common $S_N$ methods but lack in accuracy when compared
-while still requiring considerable computational resources for
-solutions in multiple dimensions.
-
-In the 1960's, Gelbard developed an ad-hoc multidimensional extension
-of the simple single dimension planar $P_N$ equations that created a
-system of coupled, diffusion-like equations known as the simplified
-$P_N$ ($SP_N$) equations. Up until around the 1990's, the $SP_N$
-methods was either widely unknown, widely unused, or combination of
-both even though numerical studies showed promising results with
-better solutions than diffusion theory and a significant reduction in
-computational time over more accurate methods such as discrete
-ordinates. Why did this happen? A significant problem, pointed out by
-Larsen, was that little rigor had been applied to the formulation of
-the $SP_N$ equations since their derivation through primarily
-heuristic arguments. Instead, studies at that time focused on simply
-comparing the results of the method to other contemporary transport
-solution strategies. In addition, many problems of interest from the
-literature at the time were either solved using nodal-type methods for
-reactor-sized problems or $S_N$-type methods for benchmark problems
-with intricate material configurations and potentially large flux
-gradients over small spatial domains.
-
-So why reconsider the $SP_N$ equations? Starting in the 1990's and
-primarily due to Larsen and his colleagues, the $SP_N$ equations have
-been given a more rigorous treatment with both variational and
-asymptotic derivations performed as a means of verification. In
-addition, these equations have been more rigorously studied as
-solution methods to MOX fuel problems and have been shown to provide
-accurate solutions. With this mathematical literature to provide a
-solid numerical footing for the method, we look at its application to
-today's challenge problems in full reactor core transport. The
-reduction in numerical complexity of current full core deterministic
-solution methods using the $S_N$ approximation could mean significant
-savings in both compute time and memory required. In addition, the
-characteristics of the solution to the transport problem for a steady
-state reactor core permit diffusion theory to be used; a staple of the
-nuclear industry since its inception. Therefore, if diffusion theory
-is applicable, then finer grained solutions that capture more of the
-physics contained in the transport equation should be possible with
-the $SP_N$ method. In doing so, we also expect from the literature to
-obtain computed responses on the order of accuracy we would expect
-from an appropriately discretized $S_N$ method at a fraction of the
-cost.
-
-To further motivate moving in this direction, recent developments in
-the Exnihilo neutronics package at Oak Ridge National Laboratory have
-permitted generation of the $SP_N$ system of equations for detailed
-full core reactor models. By fully forming these equations and
-formulating them as a linear algebra problem, we now have access to
-all of the modern advancements in computational linear algebra
-including Krylov solvers for asymmetric systems and preconditioning
-methods such as algebraic multigrid. This leads us to then explore the
-applicability of our work in discrete Monte Carlo methods for linear
-systems as a possible solution method for the $SP_N$ equations. If
-formulated correctly, we hypothesize that signficant improvement in
-usage of computational resources may be observed compared to modern
-solution techniques such as those suggested. In addition, solving the
-$SP_N$ equations in this way also breaks away from the $S_N$ forms of
-parallelism where spatial parallelism is achieved by an efficient
-parallel sweep, angular efficiency achieved by pipelining, and energy
-parallelism achieved by decoupling the groups. With the $SP_N$
-equations as a full matrix system, we now can parallelize the problem
-as prescribed by the linear solver, which may be significantly more
-scalable than current $S_N$ transport practices.
-
-In this document we derive the multigroup $SP_N$ equations from the
-Boltzmann transport equation. We then perform a spectral analysis on
-the resulting system of equations to determine its potential
-performance with both modern Krylov methods and Monte Carlo methods
-for asymmetric linear systems. Following this, we devise a Monte Carlo
-synthetic acceleration method as a proposed solution scheme.
-
-%%---------------------------------------------------------------------------%%
-\section{Derivation of the $SP_N$ Equations}
-\label{sec:derivation}
-In this section, we derive the $SP_N$ equations, closely following the
-work of Evans. We begin by stating the general time-independent
-neutron transport equation followed by a derivation of the $P_N$
-equations in planar geometry for multiple energy groups. From these
-equations, we then apply a set of approximations to yield the
-multi-dimensional, multi-group $SP_N$ equations for fixed source
-problems.
-
-\subsection{The Neutron Transport Equation}
-\label{subsec:transport_eq}
-As a starting point we define the time-independent neutron transport
-equation \cite{lewis_computational_1993}:
-\begin{multline}
-  \hat{\Omega} \cdot \vec{\nabla} \psi(\vec{r},\hat{\Omega},E) +
-  \sigma(\vec{r},E) \psi(\vec{r},\hat{\Omega},E) = \\ \int \int
-  \sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}' \rightarrow
-  \hat{\Omega}) \psi(\vec{r},\hat{\Omega}',E') d\Omega' dE' +
-  q(\vec{r},\hat{\Omega},E)\:,
-  \label{eq:general_transport}
-\end{multline}
-with the variables defined as:
-\begin{itemize}
-\item $\vec{r}$ - neutron spatial position
-\item $\hat{\Omega}$ - neutron streaming direction with radial
-  component $\mu$ and azimuthal component $\omega$
-\item $\hat{\Omega}' \cdot \hat{\Omega} = \mu_0$ is the angle of
-  scattering
-\item $E$ - neutron energy
-\item $\psi(\vec{r},\hat{\Omega},E)$ - angular flux
-\item $\sigma(\vec{r},E)$ - total interaction cross section
-\item $\sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}')$ - probability
-  of scattering from direction $\hat{\Omega}'$ into an angular domain
-  $d\hat{\Omega}'$ about the direction $\hat{\Omega}$ and from energy
-  $E'$ to an energy domain $dE'$ about energy $E$
-\item $q(\vec{r},\hat{\Omega},E)$ - external source of neutrons.
-\end{itemize}
-For this work, it is sufficient to formulate
-Eq~(\ref{eq:general_transport}) in 1-dimensonal Cartesian geometry:
-\begin{multline}
-  \mu \frac{\partial}{\partial x} \psi(x,\mu,E) + \sigma(x,E)
-  \psi(x,\mu,E) = \\ \int \int \sigma_s(x,E') \rightarrow
-  E,\hat{\Omega}' \rightarrow \hat{\Omega}) \psi(x,\hat{\Omega}',E')
-  d\Omega' dE' + \frac{q(x,E)}{4 \pi}\:,
-  \label{eq:cart_1d_transport}
-\end{multline}
-where the angular component of the solution is no longer dependent on
-the azimuthal direction of travel and an isotropic source of neutrons
-is assumed.
-
-\subsection{Derivation of the $P_N$ Equations}
-\label{subsec:pn_equations}
-Next, we derive the $P_N$ equations, a simplified form of the general
+\chapter{Derivation of the $P_N$ Equations}
+\label{chap:pn_equations}
+Here we derive the $P_N$ equations, a simplified form of the general
 transport equation where Legendre polynomials are used to expand the
@@
 
-\subsubsection{Legendre Polynomials}
-\label{subsubsec:legendre_polys}
+\section{Legendre Polynomials}
+\label{sec:legendre_polys}
 The Legendre polynomials are an orthogonal set of functions that
 are solutions to Legendre's differential equation. They have the
-following form \cite{lewis_computational_1993}:
+following form \citep{lewis_computational_1993}:
 \begin{equation}
@@
 
-\subsubsection{Planar $P_N$ Equations}
-\label{subsubsec:planar_pn}
+\section{Planar $P_N$ Equations}
+\label{sec:planar_pn}
 With the Legendre polynomial properties defined above, we can proceed
@@
 
-\subsubsection{Boundary Conditions for the $P_N$ Equations}
+\section{Boundary Conditions for the $P_N$ Equations}
 \label{subsbusec:bcs_pn}
-Per the analysis in \cite{lewis_computational_1993}, two types of
+Per the analysis in \citep{lewis_computational_1993}, two types of
 boundary conditions will be discussed: reflecting and Marshak. Marshak
@@
 also holds for vacuum conditions where $\phi_b = 0$.
-
-\subsection{Formulation of the $SP_N$ Equations}
-\label{subsec:spn_equations}
-The $P_N$ equations give $N+1$ coupled first-order equations
-capturing the spatial and angular-dependence of the solution. In
-multiple dimensions, the equation set becomes large and coupled not
-only through angular moments but also through the spatial
-variables. As a simpler alternative to multidimensional $P_N$
-solutions, Gelbard recognized in 1960 that the planar $P_N$ equations
-could be simplified and applied an ad-hoc method to extend them to
-multiple dimensions, yielding the $SP_N$ equations. These equations
-are not only fewer in number, but also take on a diffusion-like form
-while maintaining the angular character of the flux, making them
-amenable to solutions with modern diffusion methods.
-
-First, the $P_N$ equations can be simplified to $(N+1)/2$ second-order
-equations by solving for the $n^th$ Legendre flux moment in the
-odd-order equations:
-\begin{equation}
-  \phi_n = \frac{1}{\Sigma_n}\Bigg[ q \delta_{no} -
-    \frac{\partial}{\partial x}\Big(\frac{n}{2n+1}\phi_{n-1} +
-    \frac{n+1}{2n+1} \phi_{n+1} \Big) \Bigg]\:, 
-  \label{eq:odd_moments}
-\end{equation}
-for $n = 1,3,\cdots,N$ and $\delta_{no} = 0\ \forall n \neq 0$. We can
-insert the odd moments into Eq~(\ref{eq:final_pn_equations}) to get a
-reduced group of equations for the even moments:
-\begin{multline}
-  -\frac{\partial}{\partial x}
-  \Bigg[\frac{n}{2n+1}\frac{1}{\Sigma_{n-1}} \frac{\partial}{\partial
-      x} \Big(\frac{n-1}{2n-1} \phi_{n-2} + \frac{n}{2n-1}\phi_n \Big)
-    \\+ \frac{n+1}{2n+1}\frac{1}{\Sigma_{n+1}} \frac{\partial}{\partial
-      x} \Big(\frac{n+1}{2n+3}\phi_n + \frac{n+2}{2n+3}\phi_{n+2}\Big)
-    \Bigg] \\+ \Sigma_n \phi_n = q \delta_{n0}\ \ \ \ \ \ \ \ \ n =
-  0,2,4,\cdots,N\:.
-  \label{eq:reduced_pn}
-\end{multline}
-Immediately, we note the diffusion-like nature of
-Eq~(\ref{eq:reduced_pn}) as compared to the original $P_N$
-equations. To extend these equations to multiple dimensions, gelbard
-simple replaced the x-plane spatial derivatives in the reduced set of
-equations with general multidimensional gradient operators:
-\begin{multline}
-  -\nabla \cdot \Bigg[\frac{n}{2n+1}\frac{1}{\Sigma_{n-1}} \nabla
-    \Big(\frac{n-1}{2n-1} \phi_{n-2} + \frac{n}{2n-1}\phi_n \Big) \\+
-    \frac{n+1}{2n+1}\frac{1}{\Sigma_{n+1}} \nabla
-    \Big(\frac{n+1}{2n+3}\phi_n + \frac{n+2}{2n+3}\phi_{n+2}\Big)
-    \Bigg] \\+ \Sigma_n \phi_n = q \delta_{n0}\ \ \ \ \ \ \ \ \ n =
-  0,2,4,\cdots,N\:,
-  \label{eq:spn_equations}
-\end{multline}
-yielding a multidimensional set of $(N+1)/1$ angular coupled equations
-known as the $SP_N$. Again, we provide closure to this set of
-equations with $\phi_{N+1} = 0$. As a concrete example, we will
-consider the $SP_7$ equations:
-\begin{subequations}
-  \begin{gather}
-    -\nabla \cdot \frac{1}{3 \Sigma_1} \nabla ( \phi_0 + 2\phi_2 ) +
-    \Sigma_0 \phi_0 = q \\ 
-    -\nabla \cdot \Bigg[ \frac{2}{15 \Sigma_1} \nabla ( \phi_0 + 2\phi_2
-      ) + \frac{3}{35 \Sigma_3}\nabla( 3\phi_2 + 4\phi_4)\Bigg] +
-    \Sigma_2 \phi_2 = 0\\
-    -\nabla \cdot \Bigg[ \frac{4}{63 \Sigma_3} \nabla ( 3\phi_2 +
-      4\phi_4 ) + \frac{5}{99 \Sigma_5}\nabla( 5\phi_4 +
-      6\phi_6)\Bigg] + \Sigma_4 \phi_4 = 0\\
-    -\nabla \cdot \Bigg[ \frac{6}{143 \Sigma_5} \nabla ( 5\phi_4 +
-      6\phi_6 ) + \frac{7}{195 \Sigma_7}\nabla(7\phi_6)\Bigg] +
-    \Sigma_6 \phi_6 = 0 \:.
-  \end{gather}
-  \label{eq:sp7_equations}
-\end{subequations}
-To further modify these equations, we can use a change of variables to
-create a new group of equations such that the gradients are operating
-on a single vector:
-\begin{subequations}
-  \begin{gather}
-    u_1 = \phi_0 + 2\phi_2 \\
-    u_2 = 3\phi_2 + 4\phi_4 \\
-    u_3 = 5\phi_4 + 6\phi_6 \\
-    u_4 = 7\phi_6 \:,
-  \end{gather}
-  \label{eq:spn7_subs}
-\end{subequations}
-or equivalently:
-\begin{subequations}
-  \begin{gather}
-    \phi_0 = u_1 - \frac{2}{3}u_2 + \frac{8}{15}u_3 -
-    \frac{16}{35}u_4 \\
-    \phi_2 = \frac{1}{3}u_2 - \frac{4}{15}u_3 + \frac{8}{35}u_4\\ 
-    \phi_4 = \frac{1}{5}u_3 - \frac{6}{35}u_4\\
-    \phi_6 = \frac{1}{7}u_4\:.
-  \end{gather}
-  \label{eq:spn7_subs_inverse}
-\end{subequations}
-When substituted into Eq~(\ref{eq:sp7_equations}), these terms give:
-\begin{subequations}
-  \begin{gather}
-    -\nabla \cdot \frac{1}{3 \Sigma_1} \nabla u_1 + \Sigma_0 \Bigg[
-    u_1 - \frac{2}{3}u_2 + \frac{8}{15}u_3 - \frac{16}{35}u_4 \Bigg]
-    = -q \\
-    -\nabla \cdot \Bigg[ \frac{2}{15 \Sigma_1} \nabla u_1 +
-    \frac{3}{35 \Sigma_3} \nabla u_2 \Bigg] + \Sigma_2 \Bigg[
-    \frac{1}{3}u_2 - \frac{4}{15}u_3 + \frac{8}{35}u_4 \Bigg] = 0 \\
-    -\nabla \cdot \Bigg[ \frac{4}{63 \Sigma_3} \nabla u_2 +
-    \frac{5}{99 \Sigma_5} \nabla u_3 \Bigg] + \Sigma_4 \Bigg[
-    \frac{1}{5}u_3 - \frac{6}{35}u_4 \Bigg] = 0 \\ 
-    -\nabla \cdot \Bigg[ \frac{6}{143 \Sigma_5} \nabla u_3 +
-    \frac{7}{195 \Sigma_7} \nabla u_4 \Bigg] + \Sigma_6 \Bigg[
-    \frac{1}{7}u_4 \Bigg] = 0 \:.
-  \end{gather}
-  \label{eq:spn7_subs_equations}
-\end{subequations}
-If we rearrange the Eq~(\ref{eq:spn7_subs_equations} such that only
-one divergence operation is present in each equation, we can formulate
-this as a matrix system of 4 equations in the case of the $SP_7$
-approximation:
-\begin{equation}
-  -\nabla \cdot D_n \nabla u_n + \sum_{m=1}^4 A_{nm} u_m =
-  q_n\ \ \ \ \ \ \ n = 1,2,3,4\:,
-  \label{eq:spn_matrix}
-\end{equation}
-with $\mathbf{u}$ the vector of solution variables:
-\begin{equation}
-  \mathbf{u} = ( u_1\ \ u_2\ \ u_3\ \ u_4 )^T \:,
-  \label{eq:spn7_solution_vector}
-\end{equation}
-$\mathbf{D}$ the vector of effective diffusion coefficients:
-\begin{equation}
-  \mathbf{D} = \Bigg( \frac{1}{3\Sigma_1}\ \ \frac{1}{7\Sigma_3}\ \
-  \frac{1}{11\Sigma_5}\ \ \frac{1}{15\Sigma_7} \Bigg)^T\:,
-  \label{eq:spn7_diffusion_coeffs}
-\end{equation}
-$\mathbf{q}$ the vector of source terms where the $0^{th}$ moment
-source has now been distributed through the system:
-\begin{equation}
-  \mathbf{q} = (
-  q\ \ -\frac{2}{3}q\ \ \frac{8}{15}q\ \ -\frac{16}{35}q )^T\:,
-  \label{eq:spn7_source_vector}
-\end{equation}
-and $\mathbf{A}$ a matrix of angular scattering terms:
-% NOTE: I copied the following matrix directly out of Tom's tech note
-% on the SPn equations which I am effectively following here because I
-% was feeling lazy. I have verified its correctness.
-\begin{equation}
-  \mathbf{A} = 
-  \begin{bmatrix}
-    (\Sigma_0) &
-    (-\frac{2}{3}\Sigma_0) &
-    (\frac{8}{15}\Sigma_0) &
-    (-\frac{16}{35}\Sigma_0) \\
-    %%
-    &&&\\
-    %%
-    (-\frac{2}{3}\Sigma_0) &
-    (\frac{4}{15}\Sigma_0 + \frac{1}{3}\Sigma_2) &
-    (-\frac{16}{45}\Sigma_0 - \frac{4}{9}\Sigma_2) &
-    (\frac{32}{105}\Sigma_0 + \frac{8}{21}\Sigma_2) \\
-    %%
-    &&&\\
-    %%
-    (\frac{8}{15}\Sigma_0) &
-    (-\frac{16}{45}\Sigma_0 - \frac{4}{9}\Sigma_2) &
-    (\frac{64}{225}\Sigma_0 + \frac{16}{45}\Sigma_2 + \frac{9}{25}\Sigma_4) &
-    (-\frac{128}{525}\Sigma_0 - \frac{32}{105}\Sigma_2 - \frac{54}{175}\Sigma_4)
-    \\ 
-    %%
-    &&&\\
-    %%
-    (-\frac{16}{35}\Sigma_0) &
-    (\frac{32}{105}\Sigma_0 + \frac{8}{21}\Sigma_2) &
-    (-\frac{128}{525}\Sigma_0 - \frac{32}{105}\Sigma_2 - \frac{54}{175}\Sigma_4)
-    & 
-    (\frac{256}{1225}\Sigma_0 + \frac{64}{245}\Sigma_2 +
-    \frac{324}{1225}\Sigma_4 + \frac{13}{49}\Sigma_6)
-  \end{bmatrix}\:.
-  \label{eq:A_matrix}
-\end{equation}
-Note that the term $\sum_{m=1}^4 A_{nm} u_m$ in
-Eq~(\ref{eq:spn_matrix}) couples the moments in each equation while
-the diffusive term in each equation is only for a single
-'psuedo-moment' $u_n$. This completes the derivation of the $SP_7$
-equations for a single energy group. As noted by Evans, lower order
-$SP_N$ approximations can be generated by setting higher order even
-moments in this system to zero (e.g. $\phi_6 = \phi_4 = 0$ yields the
-$SP_3$ equations).
-
-\subsubsection{Boundary Conditions for the $SP_N$ Equations}
-\label{subsubsec:spn_boundary_conditions}
-We approach the formulation of the boundary conditions in much the
-same way as we did for the $P_N$ equations with both reflecting and
-Marshak conditions possible. To begin, we perform the expansion in
-Eq~(\ref{eq:marshak_expanded}) for the left side of the planar system,
-this time for $N=7$ to correspond to our $SP_7$ system. For $i=1$:
-\begin{multline}
-  \int_0^1 \mu \Bigg[ \phi_0 + 3\phi_1\mu +
-    \frac{5}{2}\phi_2(3\mu^2-1) + \frac{7}{2}\phi_3(5\mu^3-3\mu)
-    +\\ \frac{1}{8}(63\mu^5-70\mu^3+15\mu) \Bigg] d\mu = \int_0^1 \mu
-  \phi_b d\mu\:,
-  \label{eq:spn_bnd_p1}
-\end{multline}
-for $i=3$,
-\begin{multline}
-  \int_0^1 \frac{1}{2}(5\mu^3-3\mu) \Bigg[ \phi_0 + 3\phi_1\mu +
-    \frac{5}{2}\phi_2(3\mu^2-1) + \frac{7}{2}\phi_3(5\mu^3-3\mu) +\\
-    \frac{1}{8}(63\mu^5-70\mu^3+15\mu) \Bigg] d\mu = \int_0^1
-  \frac{1}{2}(5\mu^3-3\mu) \phi_b d\mu\:,
-  \label{eq:spn_bnd_p3}
-\end{multline}
-for $i=5$:
-\begin{multline}
-  \int_0^1 \frac{1}{8}(63\mu^5-70\mu^3+15\mu) \Bigg[ \phi_0 + 3\phi_1\mu +
-    \frac{5}{2}\phi_2(3\mu^2-1) + \frac{7}{2}\phi_3(5\mu^3-3\mu) +\\
-    \frac{1}{8}(63\mu^5-70\mu^3+15\mu) \Bigg] d\mu = \int_0^1
-  \frac{1}{8}(63\mu^5-70\mu^3+15\mu) \phi_b d\mu\:,
-  \label{eq:spn_bnd_p5}
-\end{multline}
-and for $i=7$:
-\begin{multline}
-  \int_0^1 \frac{1}{16}(429\mu^7-693\mu^5+315\mu^3+35\mu) \Bigg[
-    \phi_0 + 3\phi_1\mu + \frac{5}{2}\phi_2(3\mu^2-1) +
-    \frac{7}{2}\phi_3(5\mu^3-3\mu) \\+
-    \frac{1}{8}(63\mu^5-70\mu^3+15\mu) +
-    \frac{1}{16}(429\mu^7-693\mu^5+315\mu^3+35\mu) \Bigg] d\mu =\\
-  \int_0^1 \frac{1}{16}(429\mu^7-693\mu^5+315\mu^3+35\mu) \phi_b
-  d\mu\:.
-  \label{eq:spn_bnd_p7}
-\end{multline}
-Carrying out the simple integrations yields the following system of
-equations:
-\begin{subequations}
-  \begin{gather}
-    \frac{1}{2}\phi_0 + \phi_1 + \frac{5}{8}\phi_2 -
-    \frac{3}{16}\phi_4 + \frac{13}{128}\phi_6 =
-    \frac{1}{2}\phi_{b}\\ -\frac{1}{8}\phi_0 + \frac{5}{8}\phi_2 +
-    \phi_3 + \frac{81}{128}\phi_4 - \frac{13}{64}\phi_6 =
-    -\frac{1}{8}\phi_{b}\\ \frac{1}{16}\phi_0 - \frac{25}{128}\phi_2 +
-    \frac{81}{128}\phi_4 + \phi_5 + \frac{325}{512}\phi_6 =
-    \frac{1}{16}\phi_{b}\\ -\frac{5}{128}\phi_0 + \frac{7}{64}\phi_2 -
-    \frac{105}{512}\phi_4 + \frac{325}{512}\phi_6 + \phi_7 =
-    -\frac{5}{128}\phi_{b}\:,
-  \end{gather}
-  \label{eq:spn_bnd_integrated}
-\end{subequations}
-where $\phi_b$ is again an isotropic source prescribed on the planar
-boundary. For the odd moment in each boundary equation, we insert
-Eq~(\ref{eq:odd_moments}) to remove them, leaving only the even
-moments and a set of differential equations:
-\begin{subequations}
-  \begin{gather}
-    \frac{1}{2}\phi_0 + \frac{1}{3\Sigma_1}\frac{\partial}{\partial
-      x}(\phi_0+2\phi_2) + \frac{5}{8}\phi_2 - \frac{3}{16}\phi_4 +
-    \frac{13}{128}\phi_6 = \frac{1}{2}\phi_{b}\\ -\frac{1}{8}\phi_0 +
-    \frac{5}{8}\phi_2 + \frac{1}{7\Sigma_3}\frac{\partial}{\partial
-      x}(3\phi_2 + 4\phi_4) + \frac{81}{128}\phi_4 -
-    \frac{13}{64}\phi_6 = -\frac{1}{8}\phi_{b}\\ \frac{1}{16}\phi_0 -
-    \frac{25}{128}\phi_2 + \frac{81}{128}\phi_4 +
-    \frac{1}{11\Sigma_5}\frac{\partial}{\partial x}(5\phi_4 + 6\phi_6)
-    + \frac{325}{512}\phi_6 =
-    \frac{1}{16}\phi_{b}\\ -\frac{5}{128}\phi_0 + \frac{7}{64}\phi_2 -
-    \frac{105}{512}\phi_4 + \frac{325}{512}\phi_6 +
-    \frac{1}{15\Sigma_7}\frac{\partial}{\partial x}(7\phi_6)
-    = -\frac{5}{128}\phi_{b}\:.
-  \end{gather}
-  \label{eq:spn_bnd_subs}
-\end{subequations}
-We can again make the substitution of variables given by
-Eqs~(\ref{eq:spn7_subs}) and (\ref{eq:spn7_subs_inverse}) for
-consitency with the equations defined on the domain. In addition, we
-apply the $SP_N$ approximation to the derivatives by assuming they are
-instead multidimensional gradients on the boundary such that
-$\frac{\partial}{\partial x} \rightarrow
-\hat{\mathbf{n}}\cdot\nabla$. Doing this gives:
-\begin{multline}
-    \frac{1}{2}\Big(u_1 - \frac{2}{3}u_2 + \frac{8}{15}u_3 -
-    \frac{16}{35}u_4\Big) +
-    \frac{1}{3\Sigma_1}\hat{\mathbf{n}}\cdot\nabla\Big(\Big(u_1 -
-    \frac{2}{3}u_2 + \frac{8}{15}u_3 -
-    \frac{16}{35}u_4\Big)+\\2\Big(\frac{1}{3}u_2 - \frac{4}{15}u_3 +
-    \frac{8}{35}u_4\Big)\Big) + \frac{5}{8}\Big(\frac{1}{3}u_2 -
-    \frac{4}{15}u_3 + \frac{8}{35}u_4\Big)
-    -\\ \frac{3}{16}\Big(\frac{1}{5}u_3 - \frac{6}{35}u_4\Big) +
-    \frac{13}{128}\Big(\frac{1}{7}u_4\Big) = \frac{1}{2}\phi_{b}\:,
-\end{multline}
-\begin{multline}
-    -\frac{1}{8}\Big(u_1 - \frac{2}{3}u_2 + \frac{8}{15}u_3 -
-    \frac{16}{35}u_4\Big) + \frac{5}{8}\Big(\frac{1}{3}u_2 -
-    \frac{4}{15}u_3 + \frac{8}{35}u_4\Big)
-    +\\ \frac{1}{7\Sigma_3}\hat{\mathbf{n}}\cdot\nabla\Big
-    (3\Big(\frac{1}{3}u_2 - \frac{4}{15}u_3 + \frac{8}{35}u_4\Big) +
-    4\Big(\frac{1}{5}u_3 - \frac{6}{35}u_4\Big)\Big) +
-    \frac{81}{128}\Big(\frac{1}{5}u_3 -\\ \frac{6}{35}u_4\Big) -
-    \frac{13}{64}\Big(\frac{1}{7}u_4\Big) = -\frac{1}{8}\phi_{b}\:,
-\end{multline}
-\begin{multline}
-    \frac{1}{16}\Big(u_1 - \frac{2}{3}u_2 + \frac{8}{15}u_3 -
-    \frac{16}{35}u_4\Big) - \frac{25}{128}\Big(\frac{1}{3}u_2 -
-    \frac{4}{15}u_3 +\\ \frac{8}{35}u_4\Big) +
-    \frac{81}{128}\Big(\frac{1}{5}u_3 - \frac{6}{35}u_4\Big) +
-    \frac{1}{11\Sigma_5}\hat{\mathbf{n}}\cdot\nabla\Big(5\Big(\frac{1}{5}u_3
-    -\\ \frac{6}{35}u_4\Big) + 6\Big(\frac{1}{7}u_4\Big)\Big) +
-    \frac{325}{512}\Big(\frac{1}{7}u_4\Big) = \frac{1}{16}\phi_{b}\:,
-\end{multline}
-\begin{multline}
-    -\frac{5}{128}\Big(u_1 - \frac{2}{3}u_2 + \frac{8}{15}u_3 -
-    \frac{16}{35}u_4\Big) + \frac{7}{64}\Big(\frac{1}{3}u_2 -
-    \frac{4}{15}u_3 + \frac{8}{35}u_4\Big)
-    -\\ \frac{105}{512}\Big(\frac{1}{5}u_3 - \frac{6}{35}u_4\Big) +
-    \frac{325}{512}\Big(\frac{1}{7}u_4\Big)
-    +\\ \frac{1}{15\Sigma_7}\hat{\mathbf{n}}
-    \cdot\nabla\Big(7\Big(\frac{1}{7}u_4\Big)\Big) =
-    -\frac{5}{128}\phi_{b}\:.
-\end{multline}
-By rearranging the system such that we have a single gradient operator
-in each equation, we again have a matrix system:
-\begin{equation}
-  \mathbf{n} \cdot D_n \nabla u_n + \sum_{m=1}^4 B_{nm} u_m =
-  s_n\ \ \ \ \ \ \ n = 1,2,3,4\:,
-  \label{eq:spn_bnd_matrix}
-\end{equation}
-where $\mathbf{D}$ and $\mathbf{u}$ are defined as before and:
-\begin{equation}
-  \mathbf{s} = \Big(\frac{1}{2}\phi_b\ -\frac{1}{8}\phi_b\
-  \frac{1}{16}\phi_b\ -\frac{5}{128}\phi_b \Big)^T\:,
-  \label{eq:spn_bnd_source}
-\end{equation}
-is the source vector on the boundary and
-%% I grabbed this matrix from Tom's tech note again.
-\begin{equation}
-  \mathbf{B} = \begin{bmatrix}
-    \frac{1}{2} &
-    -\frac{1}{8} &
-    \frac{1}{16} &
-    -\frac{5}{128} \\
-    %%
-    &&&\\
-    %%
-    -\frac{1}{8} &
-    \frac{7}{24} &
-    -\frac{41}{384} &
-    \frac{1}{16} \\
-    %%
-    &&&\\
-    %%
-    \frac{1}{16} &
-    -\frac{41}{384} &
-    \frac{407}{1920} &
-    -\frac{233}{2560} \\
-    %%
-    &&&\\
-    %%
-    -\frac{5}{128} &
-    \frac{1}{16} &
-    -\frac{233}{2560} &
-    \frac{3023}{17920}
-  \end{bmatrix}\:,
-  \label{eq:B_matrix}
-\end{equation}
-is a dense matrix of coefficients. Again, as pointed out by Evans, the
-$SP_1$ approximation reduces this boundary condition to the standard
-Marshak diffusion boundary condition. For reflecting boundary
-conditions, we use the same procedure as the $P_N$ equations where the
-odd-moments are zero such that Eq~(\ref{eq:reflecting_condition}) is
-true. From setting Eq~(\ref{eq:odd_moments}) to zero for odd $\phi_n$
-and again substituting Eq~(\ref{eq:spn7_subs}) we immediately find
-that:
-\begin{equation}
-  \nabla \mathbf{u} = 0
-  \label{eq:spn_reflecting}
-\end{equation}
-for reflecting $SP_N$ boundaries, providing enough equations to close
-the system.
-
-\subsection{Formulation of the Multigroup $SP_N$ Equations}
-\label{subsec:mg_spn_equations}
-Up to this point, we have formulated the $P_N$ and subsequently the
-$SP_N$ equations for a single neutron energy. To expand these
-equations for multiple energies, we start by stating the multigroup
-neutron transport equation for a single dimension in planar geometry:
-\begin{multline}
-  \mu \frac{\partial}{\partial x} \psi^g(x,\mu) + \sigma^g(x)
-  \psi^g(x,\mu) = \\ \sum_{g'=0}^{G} \int
-  \sigma_s^{gg'}(x,\hat{\Omega}' \rightarrow \hat{\Omega})
-  \psi^{g'}(x,\hat{\Omega}') d\Omega' + \frac{q^g(x)}{4 \pi}\:,
-  \label{eq:cart_1d_multigroup}
-\end{multline}
-where $g$ denotes the group index of $0$ to $G$ groups, $G=N_g-1$, and
-the integration of the scattering emission term of energy has been
-replaced by a discrete summation. For scattering, $\sigma_s^{gg'}$
-provides the probability of scattering at a particular angle from
-group $g$ to $g'$. The result is an equation nearly identical in form
-to Eq~(\ref{eq:cart_1d_transport}) where now instead of forming the
-$P_N$ and $SP_N$ equations for a single energy group, we form them for
-each of the energy groups with group coupling occuring through the
-scattering term. The multigroup $P_N$ equations are then:
-\begin{equation}
-   \frac{1}{2n+1} \frac{\partial}{\partial x}\Big[ (n+1) \phi^g_{n+1}
-     + n \phi^g_{n-1} \Big] +
-   \sum_{g'}(\sigma^g\delta_{gg'}-\sigma^{gg'}_{sn}) \phi^g_n =
-   q\delta_{n0} \:,
-  \label{eq:multigroup_pn_equations}
-\end{equation}
-for $n = 0,1,\dotsc,N$ where the flux and scattering moments are
-defined in a group. We observe that a $N_g \times N_g$ scattering
-matrix is formed:
-\begin{equation}
-  \mathbf{\Sigma}_n =
-  \sum_{g'}(\sigma^g\delta_{gg'}-\sigma^{gg'}_{sn})\:,
-  \label{eq:scattering_matrix}
-\end{equation}
-and when expanded gives:
-%% Also grabbed this matrix from Tom's tech note.
-\begin{equation}
-  \mathbf{\Sigma}_n =
-  \begin{bmatrix}
-    (\sigma^0-\sigma_{sn}^{00}) & -\sigma_{sn}^{01} & \dots &
-    -\sigma_{sn}^{0G} \\ &&&\\ -\sigma_{sn}^{10} &
-    (\sigma^1-\sigma_{sn}^{11}) & \dots & -\sigma_{sn}^{1G}
-    \\ &&&\\ \vdots & \vdots & \ddots & \vdots
-    \\ &&&\\ -\sigma_{sn}^{G0} & -\sigma_{sn}^{G1} & \dots &
-    (\sigma^G-\sigma_{sn}^{GG})
-  \end{bmatrix}\:.
-\end{equation}
-It is also useful to combine the group flux moments and sources into a
-single vector for more compact notation:
-\begin{equation}
-  \mathbf{\Phi_n} = (\phi^0_n\ \phi^1_n\ \cdots \phi^G_n )^T\:, 
-  \label{eq:group_flux_vector}
-\end{equation}
-\begin{equation}
-  \mathbf{q} = (q^0\ q^1\ \cdots q^G )^T\:.
-  \label{eq:group_source_vector}
-\end{equation}
-Next, we apply the $SP_N$ approximation to
-Eq~(\ref{eq:multigroup_pn_equations}) in identical fashion to the
-monoenergetic case. This gives:
-\begin{multline}
-  -\nabla \cdot \Bigg[\frac{n}{2n+1}\mathbf{\Sigma_{n-1}}^{-1} \nabla
-    \Big(\frac{n-1}{2n-1} \mathbf{\Phi_{n-2}} +
-    \frac{n}{2n-1}\mathbf{\Phi_n} \Big) \\+
-    \frac{n+1}{2n+1}\mathbf{\Sigma_{n+1}}^{-1} \nabla
-    \Big(\frac{n+1}{2n+3}\mathbf{\Phi_n} +
-    \frac{n+2}{2n+3}\mathbf{\Phi_{n+2}}\Big) \Bigg] \\+
-  \mathbf{\Sigma_n} \mathbf{\Phi_n} = \mathbf{q}
-  \delta_{n0}\ \ \ \ \ \ \ \ \ n = 0,2,4,\cdots,N\:.
-  \label{eq:multigroup_spn_equations}
-\end{multline}
-This adds more complexity than the monoenergetic formulation in that
-all unknowns in this group of equations are now vector quantities and
-scattering relationships are contained in matrices rather than a
-scalar quantity. Because of this, the same sequence of variable
-changes and algebra can be used to build a set of matrix equations,
-this time in a block formulation:
-\begin{equation}
-  -\nabla \cdot \mathbb{D}_n \nabla \mathbb{U}_n + \sum_{m=1}^4
-  \mathbb{A}_{nm} \mathbb{U}_m = \mathbb{Q}_n\:,
-  \label{eq:spn_multigroup_system}
-\end{equation}
-where the definition of all quantities are the same with internal
-scalar values replaced by the group-vector values. In addition,
-$\mathbb{A}$ is now a block matrix of $N_g \times N_g$ submatrices
-generated from the moment scattering matrices:
-\begin{equation}
-  \mathbf{A} = 
-  \begin{bmatrix}
-    (\mathbf{\Sigma}_0) &
-    (-\frac{2}{3}\mathbf{\Sigma}_0) &
-    (\frac{8}{15}\mathbf{\Sigma}_0) &
-    (-\frac{16}{35}\mathbf{\Sigma}_0) \\
-    %%
-    &&&\\
-    %%
-    (-\frac{2}{3}\mathbf{\Sigma}_0) &
-    (\frac{4}{15}\mathbf{\Sigma}_0 + \frac{1}{3}\mathbf{\Sigma}_2) &
-    (-\frac{16}{45}\mathbf{\Sigma}_0 - \frac{4}{9}\mathbf{\Sigma}_2) &
-    (\frac{32}{105}\mathbf{\Sigma}_0 + \frac{8}{21}\mathbf{\Sigma}_2) \\
-    %%
-    &&&\\
-    %%
-    (\frac{8}{15}\mathbf{\Sigma}_0) &
-    (-\frac{16}{45}\mathbf{\Sigma}_0 - \frac{4}{9}\mathbf{\Sigma}_2) &
-    (\frac{64}{225}\mathbf{\Sigma}_0 + \frac{16}{45}\mathbf{\Sigma}_2 + \frac{9}{25}\mathbf{\Sigma}_4) &
-    (-\frac{128}{525}\mathbf{\Sigma}_0 - \frac{32}{105}\mathbf{\Sigma}_2 - \frac{54}{175}\mathbf{\Sigma}_4)
-    \\ 
-    %%
-    &&&\\
-    %%
-    (-\frac{16}{35}\mathbf{\Sigma}_0) &
-    (\frac{32}{105}\mathbf{\Sigma}_0 + \frac{8}{21}\mathbf{\Sigma}_2) &
-    (-\frac{128}{525}\mathbf{\Sigma}_0 - \frac{32}{105}\mathbf{\Sigma}_2 - \frac{54}{175}\mathbf{\Sigma}_4)
-    & 
-    (\frac{256}{1225}\mathbf{\Sigma}_0 + \frac{64}{245}\mathbf{\Sigma}_2 +
-    \frac{324}{1225}\mathbf{\Sigma}_4 + \frac{13}{49}\mathbf{\Sigma}_6)
-  \end{bmatrix}\:.
-  \label{eq:A_block_matrix}
-\end{equation}
-Analogously, for Marshak conditions on the boundaries we have:
-\begin{equation}
-  \hat{\mathbf{n}} \cdot \mathbb{D}_n \nabla \mathbb{U}_n +
-  \sum_{m=1}^4 \mathbb{B}_{nm} \mathbb{U}_m = \mathbb{S}_n\:,
-  \label{eq:spn_multigroup_bnd}
-\end{equation}
-with $\mathbb{S}_n$ the vector of group-wise boundary source term on
-each boundary for each psuedo-moment and $\mathbb{B}_{nm}$ is an $N_g
-\times N_g$ diagonal matrix with the value $B_{nm}$ on the
-diagonal. For reflecting conditions, again we have $\nabla
-\mathbb{U}_n = 0$ for all pseudo-moments.
-
-\subsection{A Note on Spatial Discretization and Matrix Symmetry}
-\label{subsec:spatial_discretization}
-To this point, the formulation of the multigroup $SP_N$ equations
-presented have discretized the transport equation in angle and energy
-but have yet to consider the spatial component of phase space. We
-don't explicitly consider it in this document but instead we will
-briefly comment on possible means of discretization. Of primary
-interest here is the discretization of the diffusive term in
-Eq~(\ref{eq:spn_multigroup_system}) and the convective term on the
-boundaries in Eq~(\ref{eq:spn_multigroup_bnd}). Many popular choices
-for spatial discretization are available here and include finite
-difference and finite element formulations. In the Denovo package at
-ORNL, Evans has implemented a finite volume scheme for these
-equations. Although arbitrary grids could be handled effectively
-through a finite element scheme, the rectilinear grid used in Denovo
-is easily discretized through the finite volume method in a
-conservative form. This conservative form is ideal for the
-diffusion-like equations in the domain given by
-Eq~(\ref{eq:spn_multigroup_system}) and the effective boundary current
-conditions given by Eq~(\ref{eq:spn_multigroup_bnd}) as the neutron
-current is balanced from cell-to-cell, continuity of the flux is
-preserved across cell/material boundaries, and boundary conditions are
-naturally represented through cell-face currents. Furthermore, this
-spatial discretization is a natural extension of the balance
-principles used to arrive at the general transport equation. The
-result of this finite volume discretization is one that is symmetric
-for all equations in the domain with respect to the spatial components
-of the solution.
-
-Given the symmetry of the spatial discretization equations in the
-domain given by monoenergetic Eq~(\ref{eq:spn_matrix}) form a
-symmetric linear operator acting on the pseudo-moment vector to give:
-\begin{equation}
-  \mathbf{L}\mathbf{u}=\mathbf{Q}\:,
-  \label{eq:matrix_system}
-\end{equation}
-where $\mathbf{L} = -\nabla \cdot \mathbf{D} \nabla +
-\mathbf{A}$. Moving to the multigroup formulation in
-Eq~(\ref{eq:spn_multigroup_system}) we have:
-\begin{equation}
-  \mathbb{L}\mathbb{U}=\mathbb{Q}\:,
-  \label{eq:multigroup_matrix_system}
-\end{equation}
-where $\mathbb{L} = -\nabla \cdot \mathbb{D} \nabla +
-\mathbb{A}$. Whether or not the resulting matrix is symmetric is
-determined by the group scattering matrices, $\sigma_{sn}^{gg'}$, for
-each of the moments. In general, the cross sections in these matrices
-do not form a symmetric matrix and therefore the resulting linear
-operator in Eq~(\ref{eq:multigroup_matrix_system}) will always be
-asymmetric for multigroup problems. Solutions to this linear system
-will then require techniques specifically formulated for asymmetric
-systems.
-
-%%---------------------------------------------------------------------------%%
-\section{Spectral Analysis of the $SP_N$ Equations}
-\label{sec:spectral_analysis}
-Now that we have an understanding of the linear system generated by
-the $SP_N$ approximation to the neutron transport equation and the
-various parameters in the system that may be adjusted, we can consider
-various means of solution. As the solution is asymmetric, modern
-iterative solvers should be considered for this task. The performance
-of Krylov subspace methods are bound to various properties of the
-eigenvalue spectrum of the linear operator while methods based on
-stationary iterations have performance limits imposed by the spectral
-radius of the sytem. In this section we will review the important
-spectral properties to study for common iterative methods for
-asymmetric systems. Based on these properties, we will then perform a
-parameter-based spectral analysis for the linear operator generated by
-the multigroup $SP_N$ equations. Preconditioning strategies for these
-methods are not considered. Krylov subspace method restarts and
-truncation will also not be considered.
-
-\subsection{Stationary Methods for Linear Systems}
-\label{subsec:stationary_solvers}
-Stationary methods for linear systems arise from splitting the
-operator in Eq~(\ref{eq:multigroup_matrix_system})
-\cite{leveque_finite_2007}:
-\begin{equation}
-  \mathbb{L} = \mathbf{M} - \mathbf{N}\:,
-  \label{eq:split_linear_operator}
-\end{equation}
-where the choice of $\mathbf{M}$ and $\mathbf{N}$ will be dictated by
-the particular method chosen. Using this split definition of the
-operator we can then write:
-\begin{equation}
-  \mathbf{M}\mathbb{U} - \mathbf{N}\mathbb{U} = \mathbb{Q}\:.
-  \label{eq:linear_split_equation1}
-\end{equation}
-By rearranging, we can generate a form more useful for analysis:
-\begin{equation}
-  \mathbb{U} = \mathbf{H}\mathbb{U} + \mathbf{c}\:,
-  \label{eq:linear_split_equation2}
-\end{equation}
-where $\mathbf{H}=\mathbf{M}^{-1}\mathbf{N}$ is defined as the
-\textit{iteration matrix} and
-$\mathbf{c}=\mathbf{M}^{-1}\mathbb{Q}$. With the solution vector on
-both the left and right hand sides, an iterative method can then be
-formed:
-\begin{equation}
-    \mathbb{U}^{k+1} = \mathbf{H}\mathbb{U}^k + \mathbf{c}\:,
-  \label{eq:linear_iterative_method}
-\end{equation}
-with $k \in \mathbf{Z}^+$ defined as the \textit{iteration
-  index}. Defining $\mathbf{e}^k = \mathbf{u}^k - \mathbf{u}$ as the
-solution error at the $k^{th}$ iterate, we can subtract
-Eq~(\ref{eq:linear_split_equation2}) from
-Eq~(\ref{eq:linear_iterative_method}):
-\begin{equation}
-  \mathbf{e}^{k+1} = \mathbf{H}\mathbf{e}^k\:.
-  \label{eq:linear_iterative_error}
-\end{equation}
-Our error after $k$ iterations is then:
-\begin{equation}
-  \mathbf{e}^{k} = \mathbf{H}^k\mathbf{e}^0\:.
-  \label{eq:linear_k_iter_error}
-\end{equation}
-In other words, successive application of the iteration matrix is the
-mechanism driving down the error in a stationary method. By assuming
-$\mathbf{H}$ is diagonalizable \cite{saad_iterative_2003}, we then
-have:
-\begin{equation}
-  \mathbf{e}^{k} =
-  \mathbf{R}\boldsymbol{\Lambda}^k\mathbf{R}^{-1}\mathbf{e}^0\:,
-  \label{eq:linear_k_iter_error_diag}
-\end{equation}
-where $\boldsymbol{\Lambda}$ contains the Eigenvalues of $\mathbf{H}$
-on its diagonal and the columns of $\mathbf{R}$ contain the
-Eigenvectors of $\mathbf{H}$. Computing the 2-norm of the above form
-then gives:
-\begin{equation}
-  ||\mathbf{e}^{k}||_2 \leq ||\boldsymbol{\Lambda}^k||_2\ 
-  ||\mathbf{R}||_2\ ||\mathbf{R}^{-1}||_2\ ||\mathbf{e}^0||_2\:,
-  \label{eq:linear_k_iter_norm1}
-\end{equation}
-which gives:
-\begin{equation}
-  ||\mathbf{e}^{k}||_2 \leq \rho(\mathbf{H})^k \kappa(\mathbf{R})
-  ||\mathbf{e}^0||_2\:.
-  \label{eq:linear_k_iter_norm2}
-\end{equation}
-For iteration matrices where the Eigenvectors are orthogonal,
-$\kappa(\mathbf{R})=1$ and the error bound reduces to:
-\begin{equation}
-  ||\mathbf{e}^{k}||_2 \leq \rho(\mathbf{H})^k
-  ||\mathbf{e}^0||_2\:.
-  \label{eq:linear_k_iter_norm3}
-\end{equation}
-We can now restrict $\mathbf{H}$ by asserting that $\rho(\mathbf{H}) <
-1$ for a stationary method to converge such that $k$ applications of
-the iteration matrix will not cause the error to grow in
-Eq~(\ref{eq:linear_k_iter_norm3}). In addition, the performance of the
-stationary method is dictated by $\rho(\mathbf{H})$ such that
-performance of the method decreases as it approaches to unity.
-
-\subsection{Krylov Subspace Methods for Linear Systems}
-\label{subsec:subspace_solvers}
-The performance of different Krylov subspace solvers can depend on
-different properties of the linear system. As a means to study these
-differences, we use a 1992 paper by Nachtigal and colleagues
-\cite{nachtigal_how_1992} that performs and in-depth analysis of three
-Krylov subspace methods that are still used today in modern physics
-applications: conjugate gradient iteration for the normal equations
-(CGN), generalized minimum residual iteration (GMRES), and a
-biconjugate gradient-based method (CGS). For our work, we will
-consider only their analysis of GMRES as it currently leveraged in
-Denovo for $SP_N$ solutions and as it also applies to CGS for most
-matrices.
-
-For GMRES, at each iteration we are extracting a correction for the
-solution from the Krylov subspace:
-\begin{equation}
-  \mathbb{U}^k \in \mathbb{U}^0 + \langle \mathbf{r_0}, \mathbb{L}
-  \mathbf{r_0}, \cdots, \mathbb{L}^{k-1} \mathbf{r_0} \rangle\:,
-  \label{eq:gmres_iteration}
-\end{equation}
-where the extraction is constrained by ensuring orthogonality of the new
-residual with the Krylov space:
-\begin{equation}
-  \mathbf{r}^k \bot \langle \mathbb{L} \mathbf{r_0}, \mathbb{L}^2
-  \mathbf{r_0}, \cdots, \mathbb{L}^{k} \mathbf{r_0} \rangle\:,
-  \label{eq:gmres_constraint}
-\end{equation}
-with the residual at the $k^{th}$ iteration defined as:
-\begin{equation}
-  \mathbf{r}^k = \mathbb{Q} - \mathbb{L}\mathbb{U}^k\:.
-  \label{eq:residual}
-\end{equation}
-From \cite{nachtigal_how_1992}, we then have the following error at
-each iteration:
-\begin{equation}
-  \mathbf{e}^k = p_k(\mathbb{L})\mathbf{e}^0\:,
-  \label{eq:gmres_iteration_error}
-\end{equation}
-where $p_k(\mathbb{L})$ is an arbitrary $k^{th}$ degree polynomial and
-$p_k(0)=1$. This is a very similar form for the iteration error as
-observed for stationary methods in Eq~(\ref{eq:linear_k_iter_norm3})
-where now as these $p_k$ polynomials are constructed as the iterations
-progress, a rapid decrease in $||p_k(\mathbb{L})||$ means a rapid rate
-of convergence. For matrices $\mathbb{L}$ that are approximately
-normal (which was exactly the assumption we made to arrive at
-Eq~(\ref{eq:linear_k_iter_norm3})), then we can write the polynomial
-norm as:
-\begin{equation}
-  ||p_k(\mathbb{L})|| = \sup_{z\in\Lambda(\mathbb{L})}|p_k(z)|\:,
-  \label{eq:gmres_polynomial}
-\end{equation}
-where $\Lambda(\mathbb{L})$ is the eigenvalue spectrum of the normal
-matrix $\mathbb{L}$. So if we could obtain the polynomials, the
-largest eigenvalue of $\mathbb{L}$ could be used to calculate
-convergence rates. Next we consider a choice of polynomials to use for
-the convergence rate approximation. They should of course be
-orthogonal to reflect the orthogonal projection mechanism by which the
-residual is reduced. In Nachtigal's work, Chebyshev polynomials were
-used and defined on the interval of Eigenvalues,
-$[\lambda_{min},\lambda_{max}]$, and then normalized via the
-constraint $p_k(0)=1$. Thus the rate of convergence for these
-polynomials is an approximation to the rate of convergence of the
-method.
-
-We can check the validity of the normal
-approximation by computing the condition number of $\mathbb{L}$:
-\begin{equation}
-  \kappa(\mathbb{L}) = ||\mathbb{L}||\ ||\mathbb{L}^{-1}||
-  \label{eq:operator_condition_number}
-\end{equation}
-The closer $\kappa(\mathbb{L})$ is to one, the more normal the matrix
-and the better the above approximation. The same check can also be
-performed to verify the assumption used to generate
-Eq~(\ref{eq:linear_k_iter_norm3}) where $\kappa(\mathbf{H})$ can be
-computed. For our analysis, we will compute $\kappa_2$ where the
-matrix norms are induced by the 2-norm of a vector such that:
-\begin{equation}
-  ||\mathbb{L}||_2 = \sqrt{\rho(\mathbb{L}^T\mathbb{L})}\:.
-  \label{eq:matrix_2_norm}
-\end{equation}
-
-\subsection{Organization of the Spectral Analysis}
-\label{subsec:spectral_organization}
-For the spectral analysis we will perform a numerical parameter study
-for the multigroup $SP_N$ equations. Based on the above information
-for iterative methods, several key quantities should be computed for
-the $SP_N$ system:
-\begin{itemize}
-\item Eigenvalue spectrum for the linear operator $\mathbb{L}$
-\item Largest eigenvalue for the linear operator $\mathbb{L}$
-\item Smallest eigenvalue for the linear operator $\mathbb{L}$
-\item Condition number for the linear operator $\mathbb{L}$
-\item For each stationary method:
-  \begin{itemize}
-  \item Eigenvalue spectrum for the iteration matrix $\mathbf{H}$
-  \item Largest eigenvalue for the iteration matrix $\mathbf{H}$
-  \item Smallest eigenvalue for the iteration matrix $\mathbf{H}$
-  \item Condition number for the iteration matrix $\mathbf{H}$
-  \end{itemize}
-\item For GMRES:
-  \begin{itemize}
-  \item Chebyshev polynomial norm using eigenvalues of $\mathbb{L}$
-    for a fixed number of iterations
-  \end{itemize}
-\end{itemize}
-In addition, the following parameters may be varied in the $SP_N$
-system in Denovo:
-\begin{itemize}
-  \item $SP_N$ order
-  \item Number of energy groups $N_g$
-  \item Total cross sections $\sigma^g$
-  \item Scattering cross sections $\sigma_{sn}^{gg'}$
-  \item Boundary conditions (Marshak or reflecting)
-  \item Finite volume mesh size ($\Delta$ in all directions)
-  \item Number of mesh elements
-  \item Source of neutrons (Uniform or Point)
-\end{itemize}
-We will restrict the problem description in Denovo to one similar
-Brantley and Larsen's $SP_N$ work where the spectral radius was
-computed for convergence analysis a several proposed solution
-methods. From this definition we choose a single material problem with
-a uniform source of unity. For the boundary conditions, we choose
-reflecting conditions on all six sides of the cubic domain for an
-effectively homogenous infinite medium problem which we will mesh with
-a $5\times 5\times 5$ grid. In Brantley and Larsen's work, however, a
-single energy group was used an therefore we may not use their cross
-sections in our computations.
-
-%%---------------------------------------------------------------------------%%
-\section{Monte Carlo Synthetic Acceleration for the $SP_N$ Equations}
-\label{sec:monte_carlo}
-
-%%---------------------------------------------------------------------------%%
-\section{Conclusion}
-\label{sec:conclusion}
-
-%%---------------------------------------------------------------------------%%
-\pagebreak
-\bibliographystyle{ieeetr} 
-\bibliography{references}
-\end{document}

### 2013-05-02 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding pn and spn appendices

@@
 system and its potential behavior in a Monte Carlo context. From the
-$P_N$ equations, we then apply a set of approximations to yield the
+$P_N$ equations, we apply a set of approximations to yield the
 multi-dimensional, multi-group $SP_N$ equations for fixed source
-problems with boundary conditions are also discussed\footnote{The
-  $SP_N$ derivations in this chapter are heavily based on those
-  presented by Evans in \citep{evans_simpli_2013}}. Using the
-fully-formed matrix for the transport problem, we explore solutions to
-the $SP_N$ equations using Monte Carlo Synthetic Acceleration using a
-light water reactor fuel assembly criticality calculation as the
-driving problem.
+problems\footnote{The $SP_N$ derivations in this chapter are heavily
+  based on those presented by Evans in
+  \citep{evans_simpli_2013}}. Using the fully-formed matrix for the
+transport problem, we explore solutions to the $SP_N$ equations using
+Monte Carlo Synthetic Acceleration using a light water reactor fuel
+assembly criticality calculation as the driving problem.
 
@@
 
-\subsection{Boundary Conditions for the $SP_N$ Equations}
-\label{subsec:spn_boundary_conditions}
-We approach the formulation of the boundary conditions in much the
-same way as for the $P_N$ equations with both reflecting and Marshak
-conditions possible. To begin, we perform the expansion in
-Eq~(\ref{eq:marshak_expanded}) for the left side of the planar system,
-this time for $N=7$ to correspond to our $SP_7$ system. For $i=1$:
-\begin{multline}
-  \int_0^1 \mu \Bigg[ \phi_0 + 3\phi_1\mu +
-    \frac{5}{2}\phi_2(3\mu^2-1) + \frac{7}{2}\phi_3(5\mu^3-3\mu)
-    +\\ \frac{1}{8}(63\mu^5-70\mu^3+15\mu) \Bigg] d\mu = \int_0^1 \mu
-  \phi_b d\mu\:,
-  \label{eq:spn_bnd_p1}
-\end{multline}
-for $i=3$,
-\begin{multline}
-  \int_0^1 \frac{1}{2}(5\mu^3-3\mu) \Bigg[ \phi_0 + 3\phi_1\mu +
-    \frac{5}{2}\phi_2(3\mu^2-1) + \frac{7}{2}\phi_3(5\mu^3-3\mu) +\\
-    \frac{1}{8}(63\mu^5-70\mu^3+15\mu) \Bigg] d\mu = \int_0^1
-  \frac{1}{2}(5\mu^3-3\mu) \phi_b d\mu\:,
-  \label{eq:spn_bnd_p3}
-\end{multline}
-for $i=5$:
-\begin{multline}
-  \int_0^1 \frac{1}{8}(63\mu^5-70\mu^3+15\mu) \Bigg[ \phi_0 + 3\phi_1\mu +
-    \frac{5}{2}\phi_2(3\mu^2-1) + \frac{7}{2}\phi_3(5\mu^3-3\mu) +\\
-    \frac{1}{8}(63\mu^5-70\mu^3+15\mu) \Bigg] d\mu = \int_0^1
-  \frac{1}{8}(63\mu^5-70\mu^3+15\mu) \phi_b d\mu\:,
-  \label{eq:spn_bnd_p5}
-\end{multline}
-and for $i=7$:
-\begin{multline}
-  \int_0^1 \frac{1}{16}(429\mu^7-693\mu^5+315\mu^3+35\mu) \Bigg[
-    \phi_0 + 3\phi_1\mu + \frac{5}{2}\phi_2(3\mu^2-1) +
-    \frac{7}{2}\phi_3(5\mu^3-3\mu) \\+
-    \frac{1}{8}(63\mu^5-70\mu^3+15\mu) +
-    \frac{1}{16}(429\mu^7-693\mu^5+315\mu^3+35\mu) \Bigg] d\mu =\\
-  \int_0^1 \frac{1}{16}(429\mu^7-693\mu^5+315\mu^3+35\mu) \phi_b
-  d\mu\:.
-  \label{eq:spn_bnd_p7}
-\end{multline}
-Carrying out the simple integrations yields the following system of
-equations:
-\begin{subequations}
-  \begin{gather}
-    \frac{1}{2}\phi_0 + \phi_1 + \frac{5}{8}\phi_2 -
-    \frac{3}{16}\phi_4 + \frac{13}{128}\phi_6 =
-    \frac{1}{2}\phi_{b}\\ -\frac{1}{8}\phi_0 + \frac{5}{8}\phi_2 +
-    \phi_3 + \frac{81}{128}\phi_4 - \frac{13}{64}\phi_6 =
-    -\frac{1}{8}\phi_{b}\\ \frac{1}{16}\phi_0 - \frac{25}{128}\phi_2 +
-    \frac{81}{128}\phi_4 + \phi_5 + \frac{325}{512}\phi_6 =
-    \frac{1}{16}\phi_{b}\\ -\frac{5}{128}\phi_0 + \frac{7}{64}\phi_2 -
-    \frac{105}{512}\phi_4 + \frac{325}{512}\phi_6 + \phi_7 =
-    -\frac{5}{128}\phi_{b}\:,
-  \end{gather}
-  \label{eq:spn_bnd_integrated}
-\end{subequations}
-where $\phi_b$ is again an isotropic source prescribed on the planar
-boundary. For the odd moment in each boundary equation, we insert
-Eq~(\ref{eq:odd_moments}) to remove them, leaving only the even
-moments and a set of first-order differential equations:
-\begin{subequations}
-  \begin{gather}
-    \frac{1}{2}\phi_0 + \frac{1}{3\Sigma_1}\frac{\partial}{\partial
-      x}(\phi_0+2\phi_2) + \frac{5}{8}\phi_2 - \frac{3}{16}\phi_4 +
-    \frac{13}{128}\phi_6 = \frac{1}{2}\phi_{b}\\ -\frac{1}{8}\phi_0 +
-    \frac{5}{8}\phi_2 + \frac{1}{7\Sigma_3}\frac{\partial}{\partial
-      x}(3\phi_2 + 4\phi_4) + \frac{81}{128}\phi_4 -
-    \frac{13}{64}\phi_6 = -\frac{1}{8}\phi_{b}\\ \frac{1}{16}\phi_0 -
-    \frac{25}{128}\phi_2 + \frac{81}{128}\phi_4 +
-    \frac{1}{11\Sigma_5}\frac{\partial}{\partial x}(5\phi_4 + 6\phi_6)
-    + \frac{325}{512}\phi_6 =
-    \frac{1}{16}\phi_{b}\\ -\frac{5}{128}\phi_0 + \frac{7}{64}\phi_2 -
-    \frac{105}{512}\phi_4 + \frac{325}{512}\phi_6 +
-    \frac{1}{15\Sigma_7}\frac{\partial}{\partial x}(7\phi_6)
-    = -\frac{5}{128}\phi_{b}\:.
-  \end{gather}
-  \label{eq:spn_bnd_subs}
-\end{subequations}
-We can again make the substitution of variables given by
-Eqs~(\ref{eq:spn7_subs}) and (\ref{eq:spn7_subs_inverse}) for
-consistency with the equations defined on the domain. In addition, we
-apply the $SP_N$ approximation to the derivatives by assuming they are
-instead multidimensional gradients on the boundary such that
-$\frac{\partial}{\partial x} \rightarrow
-\hat{\mathbf{n}}\cdot\nabla$. Doing this gives:
-\begin{multline}
-    \frac{1}{2}\Big(u_1 - \frac{2}{3}u_2 + \frac{8}{15}u_3 -
-    \frac{16}{35}u_4\Big) +
-    \frac{1}{3\Sigma_1}\hat{\mathbf{n}}\cdot\nabla\Big(\Big(u_1 -
-    \frac{2}{3}u_2 + \frac{8}{15}u_3 -
-    \frac{16}{35}u_4\Big)+\\2\Big(\frac{1}{3}u_2 - \frac{4}{15}u_3 +
-    \frac{8}{35}u_4\Big)\Big) + \frac{5}{8}\Big(\frac{1}{3}u_2 -
-    \frac{4}{15}u_3 + \frac{8}{35}u_4\Big)
-    -\\ \frac{3}{16}\Big(\frac{1}{5}u_3 - \frac{6}{35}u_4\Big) +
-    \frac{13}{128}\Big(\frac{1}{7}u_4\Big) = \frac{1}{2}\phi_{b}\:,
-\end{multline}
-\begin{multline}
-    -\frac{1}{8}\Big(u_1 - \frac{2}{3}u_2 + \frac{8}{15}u_3 -
-    \frac{16}{35}u_4\Big) + \frac{5}{8}\Big(\frac{1}{3}u_2 -
-    \frac{4}{15}u_3 + \frac{8}{35}u_4\Big)
-    +\\ \frac{1}{7\Sigma_3}\hat{\mathbf{n}}\cdot\nabla\Big
-    (3\Big(\frac{1}{3}u_2 - \frac{4}{15}u_3 + \frac{8}{35}u_4\Big) +
-    4\Big(\frac{1}{5}u_3 - \frac{6}{35}u_4\Big)\Big) +
-    \frac{81}{128}\Big(\frac{1}{5}u_3 -\\ \frac{6}{35}u_4\Big) -
-    \frac{13}{64}\Big(\frac{1}{7}u_4\Big) = -\frac{1}{8}\phi_{b}\:,
-\end{multline}
-\begin{multline}
-    \frac{1}{16}\Big(u_1 - \frac{2}{3}u_2 + \frac{8}{15}u_3 -
-    \frac{16}{35}u_4\Big) - \frac{25}{128}\Big(\frac{1}{3}u_2 -
-    \frac{4}{15}u_3 +\\ \frac{8}{35}u_4\Big) +
-    \frac{81}{128}\Big(\frac{1}{5}u_3 - \frac{6}{35}u_4\Big) +
-    \frac{1}{11\Sigma_5}\hat{\mathbf{n}}\cdot\nabla\Big(5\Big(\frac{1}{5}u_3
-    -\\ \frac{6}{35}u_4\Big) + 6\Big(\frac{1}{7}u_4\Big)\Big) +
-    \frac{325}{512}\Big(\frac{1}{7}u_4\Big) = \frac{1}{16}\phi_{b}\:,
-\end{multline}
-\begin{multline}
-    -\frac{5}{128}\Big(u_1 - \frac{2}{3}u_2 + \frac{8}{15}u_3 -
-    \frac{16}{35}u_4\Big) + \frac{7}{64}\Big(\frac{1}{3}u_2 -
-    \frac{4}{15}u_3 + \frac{8}{35}u_4\Big)
-    -\\ \frac{105}{512}\Big(\frac{1}{5}u_3 - \frac{6}{35}u_4\Big) +
-    \frac{325}{512}\Big(\frac{1}{7}u_4\Big)
-    +\\ \frac{1}{15\Sigma_7}\hat{\mathbf{n}}
-    \cdot\nabla\Big(7\Big(\frac{1}{7}u_4\Big)\Big) =
-    -\frac{5}{128}\phi_{b}\:.
-\end{multline}
-By rearranging the system such that we have a single gradient operator
-in each equation, we again have a matrix system:
-\begin{equation}
-  \mathbf{n} \cdot D_n \nabla u_n + \sum_{m=1}^4 B_{nm} u_m =
-  s_n\ \ \ \ \ \ \ n = 1,2,3,4\:,
-  \label{eq:spn_bnd_matrix}
-\end{equation}
-where $\mathbf{D}$ and $\mathbf{u}$ are defined as before and:
-\begin{equation}
-  \mathbf{s} = \Big(\frac{1}{2}\phi_b\ -\frac{1}{8}\phi_b\
-  \frac{1}{16}\phi_b\ -\frac{5}{128}\phi_b \Big)^T\:,
-  \label{eq:spn_bnd_source}
-\end{equation}
-is the source vector on the boundary and
-%% I grabbed this matrix from Tom's tech note again.
-\begin{equation}
-  \mathbf{B} = \begin{bmatrix}
-    \frac{1}{2} &
-    -\frac{1}{8} &
-    \frac{1}{16} &
-    -\frac{5}{128} \\
-    %%
-    &&&\\
-    %%
-    -\frac{1}{8} &
-    \frac{7}{24} &
-    -\frac{41}{384} &
-    \frac{1}{16} \\
-    %%
-    &&&\\
-    %%
-    \frac{1}{16} &
-    -\frac{41}{384} &
-    \frac{407}{1920} &
-    -\frac{233}{2560} \\
-    %%
-    &&&\\
-    %%
-    -\frac{5}{128} &
-    \frac{1}{16} &
-    -\frac{233}{2560} &
-    \frac{3023}{17920}
-  \end{bmatrix}\:,
-  \label{eq:B_matrix}
-\end{equation}
-is a dense matrix of coefficients. Again, as pointed out by Evans, the
-$SP_1$ approximation reduces this boundary condition to the standard
-Marshak diffusion boundary condition. For reflecting boundary
-conditions, we use the same procedure as the $P_N$ equations where the
-odd-moments are zero such that Eq~(\ref{eq:reflecting_condition}) is
-true. From setting Eq~(\ref{eq:odd_moments}) to zero for odd $\phi_n$
-and again substituting Eq~(\ref{eq:spn7_subs}) we immediately find
-that:
-\begin{equation}
-  \nabla \mathbf{u} = 0
-  \label{eq:spn_reflecting}
-\end{equation}
-for reflecting $SP_N$ boundaries, providing enough equations to close
-the system.
-
 %%---------------------------------------------------------------------------%%
@@
 \end{equation}
-Analogously, for Marshak conditions on the boundaries we have:
-\begin{equation}
-  \hat{\mathbf{n}} \cdot \mathbb{D}_n \nabla \mathbb{U}_n +
-  \sum_{m=1}^4 \mathbb{B}_{nm} \mathbb{U}_m = \mathbb{S}_n\:,
-  \label{eq:spn_multigroup_bnd}
-\end{equation}
-with $\mathbb{S}_n$ the vector of group-wise boundary source term on
-each boundary for each psuedo-moment and $\mathbb{B}_{nm}$ is an $N_g
-\times N_g$ diagonal matrix with the value $B_{nm}$ on the
-diagonal. For reflecting conditions, again we have $\nabla
-\mathbb{U}_n = 0$ for all pseudo-moments.
 

### 2013-05-02 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
sttarting work on spn spatial discretization

@@
 generated by setting higher order even moments in this system to zero
-(e.g. $\phi_6 = \phi_4 = 0$ yields the $SP_3$ equations).
+(e.g. $\phi_6 = \phi_4 = 0$ yields the $SP_3$ equations). Boundary
+conditions for these equations are provided in
+Appendix~\ref{chap:spn_boundary_conditions}.
 
@@
 \end{equation}
+Boundary conditions for these equations are provided in
+Appendix~\ref{chap:spn_boundary_conditions}.
 
 %%---------------------------------------------------------------------------%%
-\section{Spatial Discretization}
-\label{sec:spatial_discretization}
+\subsection{Spatial Discretization}
+\label{subsec:spatial_discretization}
 To this point, the formulation of the multigroup $SP_N$ equations
@@
 interest here is the discretization of the diffusive term in
-Eq~(\ref{eq:spn_multigroup_system}) and the convective term on the
+Eq~(\ref{eq:spn_multigroup_system}) and the current term on the
 boundaries in Eq~(\ref{eq:spn_multigroup_bnd}). Many popular choices
@@
 ORNL, Evans has implemented a finite volume scheme for these
-equations. Although arbitrary grids could be handled effectively
+equations resulting in:
+\begin{equation}
+
+  \label{eq:spn_spatial_discretization}
+\end{equation}
+Although arbitrary grids could be handled effectively
 through a finite element scheme, the rectilinear grid used in Denovo

### 2013-05-03 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding block jacobi xfig schematic

@@
 \begin{equation}
-
+  discretization
   \label{eq:spn_spatial_discretization}
@@
 conventional and Monte Carlo methods, we must first understand their
-spectral character to ensure the MCSA will apply. As presented in
-Chapter~\ref{ch:stochastic_methods}, the spectral radius of the
-iteration matrix must be less than unity to ensure convergence and a
-finite variance. Therefore, we will compute the spectral radius of the
+spectral character to ensure the MCSA will be applicable. As presented
+in Chapter~\ref{ch:stochastic_methods}, the spectral radius of the
+iteration matrix must be less than unity to ensure
+convergence. Therefore, we will compute the spectral radius of the
 $SP_N$ equations while parametrically varying the $SP_N$ order, the
@@
 location and energy group.
-\begin{table}[t!]
+\begin{table}[h!]
   \begin{center}
@@
 at preconditioning this system will be to use the point Jacobi
-preconditioning from \S\ref{sec:stochastic_preconditioning} due to
-this diagonal form.
+preconditioning from \S\ref{subsubsec:basic_mcsa_preconditioning} due
+to this diagonal form.
 
@@
 to ensure convergence for systems of all $SP_N$ and $P_N$ orders with
-arbitrary energy group structures. To achieve this, we look back to
-the sparsity plots we generated in Figures~\ref{fig:group1},
-\ref{fig:group10ds} and \ref{fig:group10us} as well as the multigroup
-$SP_N$ equations. Initially, the diagonal character of the system led
-us to try point Jacobi preconditioning with only marginal
-results. From the sparsity plots we note the block structure that
-ultimately arises from the multigroup scattering matrices and their
-insertion into Eq~(\ref{eq:A_block_matrix}). When full upscatter and
-downscatter are used the resulting blocks are completely dense while
-only downscatter gives a lower triangular scattering matrix and the
-block structure shown in Figure~\ref{fig:group10ds}.
+arbitrary energy group structures. As another means of achieving this,
+we look back to the sparsity plots we generated in
+Figures~\ref{fig:group1}, \ref{fig:group10ds} and \ref{fig:group10us}
+as well as the multigroup $SP_N$ equations. Initially, the diagonal
+character of the system led us to try point Jacobi preconditioning
+with only marginal results. From the sparsity plots we note the block
+structure that ultimately arises from the multigroup scattering
+matrices and their insertion into Eq~(\ref{eq:A_block_matrix}). When
+full upscatter and downscatter are used the resulting blocks are
+completely dense while only downscatter gives a lower triangular
+scattering matrix and the block structure shown in
+Figure~\ref{fig:group10ds}.
 
@@
   \begin{center}
-    \includegraphics[width=5in]{chapters/spn_equations/block_preconditioning.png}
+    \scalebox{1.5}{
+    \input{chapters/spn_equations/block_jacobi.pdftex_t} }
   \end{center}

### 2013-05-03 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
starting section of fuel assembly calcs

@@
 
-%%---------------------------------------------------------------------------%%
-\subsection{Spatial Discretization}
+\subsection{Spatial Discretization of the $SP_N$ Equations}
 \label{subsec:spatial_discretization}
@@
 principles used to arrive at the general transport equation.
+
+\subsection{Eigenvalue Form of the $SP_N$ Equations}
+\label{subsec:eigenvalue_form}
 
@@
 case with full downscatter and full upscatter.
-\begin{table}[t!]
+\begin{table}[h!]
   \begin{center}
@@
 \end{table}
-\begin{table}[t!]
+\begin{table}[h!]
   \begin{center}
@@
       $SP_N$ equations.} \textit{Left: The preconditioner is formed by
-      the diagonal blocks of the matrix. Right: Inversion is trivial
-      and decoupled by block.}}
+      the diagonal blocks of the matrix. Right: Inversion of the
+      preconditioner is trivial and decoupled by block.}}
   \label{fig:block_jacobi_ex}
@@
 side of Figure~\ref{fig:block_jacobi_ex}. Each block can be inverted
-individually and combined to form the inverse. For high performance
-implementations this has several attractive properties. First, the
-blocks in the matrix come from the energy/angle discretization of the
-transport equation as given by Eq~(\ref{eq:A_block_matrix}). Each
-block on the diagonal is bound to a mesh element in the system (note
-there are 8 blocks on the diagonal in each of the sparsity patterns
-with a mesh of $2 \times 2 \times 2$) and therefore we expect the
-matrix elements forming the block to be entirely local. Second, these
-blocks are typically dense and nearly lower triangular for many
-transport problems meaning that established dense matrix methods based
-on LU-factorization and other schemes can be used for fast inversion
-\citep{lapack_citation}.
+individually and combined to form the inverse. In addition, in the
+limit of a block size of one, the block Jacobi method reduces to the
+point Jacobi method. For high performance implementations this has
+several attractive properties. First, the blocks in the matrix come
+from the energy/angle discretization of the transport equation as
+given by Eq~(\ref{eq:A_block_matrix}). Each block on the diagonal is
+bound to a mesh element in the system (note there are 8 blocks on the
+diagonal in each of the sparsity patterns with a mesh of $2 \times 2
+\times 2$) and therefore we expect the matrix elements forming the
+block to be entirely local. Second, these blocks are typically dense
+and nearly lower triangular for many transport problems meaning that
+established dense matrix methods can be used for fast inversion.
 
@@
 cases.
-\begin{table}[t!]
+\begin{table}[h!]
   \begin{center}
@@
 \end{table}
-From the tabulated block Jacobi data it is clear that this is the
-preferred preconditioning choice for all $SP_N$ problems in term of
-Monte Carlo solution methods. All cases were observed to have a
-spectral radius below unity and often significantly smaller than that,
-greatly improving convergence rates over the point Jacobi preconditioned
-problem.
+
+From the tabulated block Jacobi data it is clear that this is a viable
+preconditioning choice for all $SP_N$ problems in term of Monte Carlo
+solution methods. All cases were observed to have a spectral radius
+below unity and often significantly smaller than that, greatly
+improving convergence rates over the point Jacobi preconditioned
+problem. Based on these results, the block Jacobi method should be the
+first preconditioning strategy applied to more complicated problems
+representative of physical systems.
 
@@
 \label{sec:fuel_assembly_calcs}
+Fuel assembly calculations are a critical piece of nuclear engineering
+infrastructure for reactor core analysis and design. At this level,
+individual fuel pins may be resolved at fine resolution in a variety
+of configurations. As a sophisticated problem of interest to push the
+limits of MCSA, a $17 \times 17$ pin assembly will be used with
+varying energy group structure and $SP_N$ discretization in a
+criticality calculation. In this scheme, at every eigenvalue iteration
+MCSA is used to solve the resulting $SP_N$ problem using the provided
+fission source. This problem is significantly more complicated that
+the simple test problem used for the previous spectral
+analysis. Fission has been introduced into the set of equations and
+the addition of moderator into the system will increase the amount of
+scattering, creating a significantly more difficult problem.
 

### 2013-05-03 doc/dissertation/document/backmatter/spn_spatial_discretization.tex
adding appending for spn finite volume discretization

\chapter{Finite Volume Discretization for the $SP_N$ Equations}
\label{chap:spn_spatial_discretization}
Finite volume discretization for the $SP_N$ equations.

### 2013-05-03 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding problem 3 material figures
adding appending for spn finite volume discretization

@@
   \hat{\Omega} \cdot \vec{\nabla} \psi(\vec{r},\hat{\Omega},E) +
-  \sigma(\vec{r},E) \psi(\vec{r},\hat{\Omega},E) = \\ \int \int
-  \sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}' \rightarrow
-  \hat{\Omega}) \psi(\vec{r},\hat{\Omega}',E') d\Omega' dE' +
+  \sigma(\vec{r},E) \psi(\vec{r},\hat{\Omega},E) = \\ \iint
+  \sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}' \cdot \hat{\Omega})
+  \psi(\vec{r},\hat{\Omega}',E') d\Omega' dE' +
   q(\vec{r},\hat{\Omega},E)\:,
@@
   \mu \frac{\partial}{\partial x} \psi(x,\mu,E) + \sigma(x,E)
-  \psi(x,\mu,E) = \\ \int \int \sigma_s(x,E') \rightarrow
-  E,\hat{\Omega}' \rightarrow \hat{\Omega}) \psi(x,\hat{\Omega}',E')
-  d\Omega' dE' + \frac{q(x,E)}{4 \pi}\:,
+  \psi(x,\mu,E) = \\ \iint \sigma_s(x,E' \rightarrow E,\hat{\Omega}'
+  \cdot \hat{\Omega}) \psi(x,\hat{\Omega}',E') d\Omega' dE' +
+  \frac{q(x,E)}{4 \pi}\:,
   \label{eq:cart_1d_transport}
@@
 the azimuthal direction of travel and an isotropic source of neutrons
-is assumed.
+is assumed. In addition, for fission systems the eigenvalue form of the transport
+equation is:
+\begin{multline}
+  \hat{\Omega} \cdot \vec{\nabla} \psi(\vec{r},\hat{\Omega},E) +
+  \sigma(\vec{r},E) \psi(\vec{r},\hat{\Omega},E) = \\ \iint
+  \sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}' \cdot \hat{\Omega})
+  \psi(\vec{r},\hat{\Omega}',E') d\Omega' dE' + \\ \frac{1}{k} \chi(E)
+  \iint \nu \sigma_f(\hat{r},E') \psi(\vec{r},\hat{\Omega}',E')
+  d\Omega' dE' + q(\vec{r},\hat{\Omega},E) \:,
+  \label{eq:eigenvalue_transport}
+\end{multline}
+with the addition variables defined as
+\begin{itemize}
+\item $k$ - multiplication factor
+\item $\chi(E)$ - fission neutron energy spectrum
+\item $\nu$ - average number of neutrons per fission
+\item $\sigma_f(r,E')$ - fission cross section\:.
+\end{itemize}
+In 1-dimensional cartesian geometry,
+Eq~(\ref{eq:eigenvalue_transport}) becomes:
+\begin{multline}
+  \mu \frac{\partial}{\partial x} \psi(x,\mu,E) + \sigma(x,E)
+  \psi(x,\mu,E) = \\ \iint \sigma_s(x,E' \rightarrow
+  E,\hat{\Omega}' \cdot \hat{\Omega}) \psi(x,\hat{\Omega}',E')
+  d\Omega' dE' + \\ \frac{1}{k} \chi(E)
+  \iint \nu \sigma_f(x,E') \psi(x,\hat{\Omega}',E') d\Omega'
+  dE' + \frac{q(x,E)}{4 \pi}\:.
+  \label{eq:cart_1d_eigenvalue}
+\end{multline}
 
@@
 \label{sec:mg_spn_equations}
-Up to this point, we have formulated the $P_N$ and subsequently the
-$SP_N$ equations for a single neutron energy. To expand these
-equations for multiple energies, we start by stating the multigroup
-neutron transport equation for a single dimension in planar geometry:
+Up to this point, we have formulated the $SP_N$ equations for a single
+neutron energy. To expand these equations for multiple energies, we
+start by stating the multigroup neutron transport equation for a
+single dimension in planar geometry:
 \begin{multline}
@@
 to Eq~(\ref{eq:cart_1d_transport}) where now instead of forming the
-$P_N$ and $SP_N$ equations for a single energy group, we form them for
-each of the energy groups with group coupling occuring through the
-scattering term. The multigroup $P_N$ equations are then:
+$SP_N$ equations for a single energy group, we form them for each of
+the energy groups with group coupling occuring through the scattering
+term. The multigroup $P_N$ equations are then:
 \begin{equation}
@@
 of configurations. As a sophisticated problem of interest to push the
-limits of MCSA, a $17 \times 17$ pin assembly will be used with
-varying energy group structure and $SP_N$ discretization in a
-criticality calculation. In this scheme, at every eigenvalue iteration
-MCSA is used to solve the resulting $SP_N$ problem using the provided
-fission source. This problem is significantly more complicated that
-the simple test problem used for the previous spectral
-analysis. Fission has been introduced into the set of equations and
-the addition of moderator into the system will increase the amount of
-scattering, creating a significantly more difficult problem.
+limits of MCSA, a $17 \times 17$ hot zero-power pin assembly will be
+used with varying energy group structure and $SP_N$ discretization in
+a criticality calculation. A cross section along the vertical access
+showing homogenized fuel pin materials and the associated grid is
+given by Figure~\ref{fig:problem3_radial_mat} while a cross section of
+the materials configuration along the horizontal axis is given by
+Figure~\ref{fig:problem3_radial_mat}.
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=4in]{chapters/spn_equations/problem3_radial_mat.png}
+  \end{center}
+  \caption{\textbf{Fuel assembly mesh and geometry cross section.}
+    \textit{Reflecting boundaries are used on the left and lower
+      boundaries to create a complete $17 \times 17$ assembly
+      geometry. Each fuel pin is resolved by a $2 \times 2$ mesh.}}
+  \label{fig:problem3_radial_mat}
+\end{figure}
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6.25in]{chapters/spn_equations/problem3_axial_mat.png}
+  \end{center}
+  \caption{\textbf{Fuel assembly geometry cross section.}  \textit{The
+      geometry is subdivided along the axial direction into 50 zones
+      spaced to capture material boundaries. Important details include
+      spacer grids along the length of the fuel pins and core
+      structures on the top and bottom of the assembly.}}
+  \label{fig:problem3_axial_mat}
+\end{figure}
+Significant geometric details are contained in the model including
+spacer grids, fuel pins with homogenized cladding and gas gap, and
+core plenum. Table~\ref{tab:problem3_parameters} gives the primary
+design parameters for the fuel assembly calculations.
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{ll}\hline\hline
+      \multicolumn{1}{l}{\textbf{Parameter}} & 
+      \multicolumn{1}{l}{\textbf{Value}} \\
+      Power Level & 0 MW \\
+      Inlet Temperature & 326.85C \\
+      Fuel Temperature & 600C \\
+      Boron Concentration & 1300 ppm \\
+      Moderator Density & 0.743 g/cc \\
+      Helium Density & \sn{1.79}{-4} g/cc \\
+      Zirconium Density & 6.56 g/cc \\
+      Stainless Steel Density & 8.0 g/cc \\
+      Inconel Density & 8.19 g/cc \\
+      UO2 Density & 10.257 g/cc \\
+      Fuel Pin Radius (w/o clad) & 0.4096 cm \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Design parameters for the $17 \times 17$ pin fuel
+      assembly criticality calculation.}}
+  \label{tab:problem3_parameters}
+\end{table}
+
+To generate the multiplication factor and steady-state flux
+distribution for this problem, at every eigenvalue iteration MCSA is
+used to solve the resulting $SP_N$ problem using the provided fission
+source. Algorithm~\ref{alg:power_iteration} presents the use of MCSA
+within a power iteration strategy to find the multiplication factor.
+\begin{algorithm}[h!]
+  \caption{Power Iteration MCSA Scheme}
+  \label{alg:power_iteration}
+  \begin{algorithmic}
+    \State $k_0 =$ initial guess
+    \State $\ve{\phi_0} =$ initial guess
+    \State $n = 0$
+    \While{$|\frac{k^n - k^{n-1}}{k^n}| < \epsilon$}
+    \Comment{Iterate until convergence of the eigenvalue}
+    \State $\ve{M} \ve{\phi}^{n+1} = \frac{1}{k^n} \ve{F} \ve{\phi}^n$
+    \Comment{Solve for the new flux state with MCSA}
+    \State $k^{n+1} = k^n \frac{\int \ve{F} \ve{\phi}^{n+1} d\ve{r}}{\int
+      \ve{F} \ve{\phi}^n d\ve{r}}$
+    \Comment{Update the multiplication factor}
+    \State $n = n+1$
+    \EndWhile
+  \end{algorithmic}
+\end{algorithm}
+Here, $\ve{M}$ is the transport operator generated on the left-hand
+side of the $SP_N$ discretization and $\ve{F}$ is the fission matrix.
+This problem is significantly more complicated that the simple test
+problem used for the previous spectral analysis. Fission has been
+introduced into the set of equations and the addition of moderator
+into the system will increase the amount of scattering, creating a
+significantly more difficult problem.
 
@@
 
+\subsubsection{Monte Carlo Estimator Comparison}
+\label{subsubsec:spn_estimator_comparison}
+
 \subsection{MCSA Comparison to Conventional Methods}

### 2013-05-03 doc/dissertation/document/backmatter/spn_spatial_discretization.tex
adding cartesian mesh cell figure

@@
 \label{chap:spn_spatial_discretization}
-Finite volume discretization for the $SP_N$ equations.
+The moment-discretized form of the $SP_N$ equations given by
+Eq~(\ref{eq:spn_matrix}) for the monoenergetic case has yet to have
+the spatial components of the solution and the derivative operators
+discretized. To achieve this, we choose a conservation law approach,
+using Eq~(\ref{eq:spn_matrix}) as the balance equation, to yield a
+finite volume calculation \citep{leveque_finite_2002}. To do this, we
+first define a control volume about point $(i,j,k)$ in the domain as a
+cell in a Cartesian mesh as shown in Figure~\ref{fig:mesh_cell}).
+\begin{figure}[h!]
+  \begin{center}
+    \scalebox{1.5}{
+    \input{backmatter/mesh_cell.pdftex_t} }
+  \end{center}
+  \caption{\textbf{Cartesian mesh cell used for the spatial
+      discretization of the $SP_N$ equations.} \textit{The cell is
+      centered about $(i,j,k)$.}}
+  \label{fig:mesh_cell}
+\end{figure}

### 2013-05-03 doc/dissertation/document/backmatter/spn_spatial_discretization.tex
finishing up spn spatial discretization

@@
       discretization of the $SP_N$ equations.} \textit{The cell is
-      centered about $(i,j,k)$.}}
+      centered about $(i,j,k)$ and has a volume $V_{ijk} = \Delta_i
+      \Delta_j \Delta_k$.}}
   \label{fig:mesh_cell}
 \end{figure}
+
+To begin, we first define the neutron current in the $n^th$
+pseudo-moment, $u_n$ using Fick's law as:
+\begin{equation}
+  J_n = -D_n \nabla u_n\:,
+  \label{eq:neutron_current}
+\end{equation}
+where $J_n$ is the moment current and $D_n$ is the moment diffusion
+coefficient. Next, we substitue Eq~(\ref{eq:neutron_current}) into
+Eq~(\ref{eq:spn_matrix}) to get the balance equation:
+\begin{equation}
+    \nabla \cdot J_n + \sum_{m=1}^4 A_{nm} u_m = q_n\ \ \ \ \ \ \ n =
+    1,2,3,4\:,
+  \label{eq:neutron_balance}
+\end{equation}
+with the current term accounting for losses, $q_n$ accounting for
+neutron generation, and $\sum_{m=1}^4 A_{nm} u_m$ accounting for
+moment-to-moment transport. To enforce conservation, we integrate
+moments over the control volume of cell $ijk$ in Figure~\ref{fig:mesh_cell}):
+\begin{equation}
+    \int_{ijk} \Bigg[\nabla \cdot J_n + \sum_{m=1}^4 A_{nm} u_m - q_n
+      \Bigg] dV = 0 \ \ \ \ \ \ \ n = 1,2,3,4\:.
+  \label{eq:neutron_conservation}
+\end{equation}
+If for a given quantity $X$, we define the cell-averaged quantity
+$X_{ijk}$ as:
+\begin{equation}
+  X_{ijk} = \frac{\int_{ijk} X dV }{V_{ijk}}\:,
+  \label{eq:cell_averaged_quant}
+\end{equation}
+then Eq~(\ref{eq:neutron_conservation}) can be written as:
+\begin{equation}
+  \int_{ijk} \nabla \cdot J_n dV + \sum_{m=1}^4 A_{nm,ijk} u_{m,ijk} = q_{n,ijk}\:,
+  \label{eq:neutron_conservation_2}
+\end{equation}
+with $n$ now implied to be 1,2,3 or 4 and the moment matrix elements,
+moments, and sources now defined as volume averaged terms in the mesh
+cell. Using the divergence theorem, we can rewrite the remaining
+integral in terms of a discrete sum of area-weighted averages over the
+faces of the cell:
+\begin{equation}
+  \int_{ijk} \nabla \cdot J_n dV = \int_{\partial} J_n \cdot \hat{n}
+  dA = \sum_{f=1}^6 J_{n,f} \cdot \hat{n}_f A_f\:,
+\label{eq:face_currents}
+\end{equation}
+where $f$ is the face index, $\hat{n}_f$ is the unit normal vector for
+that face and $A_f$ is the area of the face. Inserting this into
+Eq~(\ref{eq:neutron_conservation_2}) and expanding the face current
+summation provides a discrete form of the $SP_N$ equations:
+\begin{multline}
+(J_{n,i+1/2} - J_{n,i-1/2}) \Delta_j \Delta_k + (J_{n,j+1/2} -
+  J_{n,j-1/2}) \Delta_i \Delta_k +\\ (J_{n,k+1/2} - J_{n,k-1/2})
+  \Delta_i \Delta_j + \sum_{m=1}^4 A_{nm,ijk} u_{m,ijk} V_{ijk} =
+  q_{n,ijk} V_{ijk}\:,
+  \label{eq:discrete_current}
+\end{multline}
+For the discretization we desire only cell-centered quanities and
+therefore we aim to elminate all $J_{n,l \pm 1/2}$ terms from the
+discretization where $l = \{i,j,k\}$. To do this, we enforce
+continuity of the moment flux and the moment currents across cell
+boundaries such that $J_{n,l \pm 1/2}^+ = J_{n,l \pm
+  1/2}^-$. Considering $J_{n,l+1/2}$, using
+Eq~(\ref{eq:neutron_current}) we then have:
+\begin{equation}
+  J_{n,l+1/2} = -2 D_{n,l+1/2}\frac{u_{n,l+1} - u_{n,l}}{\Delta_{l+1}+
+    \Delta_{l}}\:,
+  \label{eq:face_current}
+\end{equation}
+\begin{equation}
+  J_{n,l+1/2}^+ = -2 D_{n,l+1}\frac{u_{n,l+1} - u_{n,l+1/2}}{\Delta_{l+1}}\:,
+  \label{eq:face_current_plus}
+\end{equation}
+\begin{equation}
+  J_{n,l+1/2}^- = -2 D_{n,l}\frac{u_{n,l+1/2} - u_{n,l}}{\Delta_{l}}\:.
+  \label{eq:face_current_minus}
+\end{equation}
+To eliminate $D_{n,l+1/2}$ in Eq~(\ref{eq:face_current}), we equate
+Eqs~(\ref{eq:face_current_plus}) and (\ref{eq:face_current_minus}) to
+enforce continuity of the neutron current:
+\begin{equation}
+  -2 D_{n,l+1}\frac{u_{n,l+1} - u_{n,l+1/2}}{\Delta_{l+1}} = -2
+  D_{n,l}\frac{u_{n,l+1/2} - u_{n,l}}{\Delta_{l}}\:.
+  \label{eq:face_current_equality}
+\end{equation}
+Solving for $u_{n,l+1/2}$ gives:
+\begin{equation}
+  u_{n,l+1/2} = \frac{\Delta_l D_{n,l+1} u_{n,l+1} + \Delta_{l+1}
+    D_{n,l} u_{n,l}}{\Delta_l D_{n,l+1} + \Delta_{l+1} D_{n,l}}\:.
+  \label{eq:face_flux}
+\end{equation}
+Next, we equate Eq~(\ref{eq:face_current}) to
+Eq~(\ref{eq:face_current_plus}):
+\begin{equation}
+-2 D_{n,l+1/2}\frac{u_{n,l+1} - u_{n,l}}{\Delta_{l+1} + \Delta_{l}} =
+-2 D_{n,l+1}\frac{u_{n,l+1} - u_{n,l+1/2}}{\Delta_{l+1}}\:,
+  \label{eq:face_current_solve}
+\end{equation}
+and rearrange:
+\begin{equation}
+  \Delta_{l+1} D_{n,l+1/2} (u_{n,l+1} - u_{n,l}) = (\Delta_{l+1} +
+  \Delta_l) D_{n,l+1}u_{n,l+1} - (\Delta_{l+1} + \Delta_l)
+  D_{n,l+1}u_{n,l+1/2}\:.
+  \label{eq:face_current_solve_2}
+\end{equation}
+Inserting Eq~(\ref{eq:face_flux}) into
+Eq~(\ref{eq:face_current_solve_2}) gives:
+\begin{multline}
+  \Delta_{l+1} D_{n,l+1/2} (\Delta_l D_{n,l+1} + \Delta_{l+1} D_{n,l})
+  (u_{n,l+1} - u_{n,l}) = \\(\Delta_{l+1} + \Delta_l) D_{n,l+1}
+  (\Delta_l D_{n,l+1} + \Delta_{l+1} D_{n,l}) u_{n,l+1} -\\
+  (\Delta_{l+1} + \Delta_l) D_{n,l+1}(\Delta_l D_{n,l+1} u_{n,l+1} +
+  \Delta_{l+1} D_{n,l} u_{n,l})\:.
+  \label{eq:face_current_solve_3}
+\end{multline}
+Expanding gives 5 terms:
+\begin{multline}
+  \Delta_{l+1} D_{n,l+1/2} (\Delta_l D_{n,l+1} + \Delta_{l+1} D_{n,l})
+  (u_{n,l+1} - u_{n,l}) = \\(\Delta_{l+1} + \Delta_l) D_{n,l+1}
+  \Delta_l D_{n,l+1} u_{n,l+1} + (\Delta_{l+1} + \Delta_l) D_{n,l+1}
+  \Delta_{l+1} D_{n,l} u_{n,l+1} -\\ (\Delta_{l+1} + \Delta_l)
+  D_{n,l+1} \Delta_l D_{n,l+1} u_{n,l+1} - (\Delta_{l+1} + \Delta_l)
+  D_{n,l+1} \Delta_{l+1} D_{n,l} u_{n,l}\:,
+  \label{eq:face_current_solve_4}
+\end{multline}
+two of which cancel, giving:
+\begin{multline}
+  \Delta_{l+1} D_{n,l+1/2} (\Delta_l D_{n,l+1} + \Delta_{l+1} D_{n,l})
+  (u_{n,l+1} - u_{n,l}) = \\(\Delta_{l+1} + \Delta_l) \Delta_{l+1}
+  D_{n,l+1} D_{n,l} ( u_{n,l+1} - u_{n,l} ) \:.
+  \label{eq:face_current_solve_5}
+\end{multline}
+Solving Eq~(\ref{eq:face_current_solve_5}) for $D_{n,l+1/2}$ then
+gives the face diffusion coefficient in terms of cell-centered
+coefficients:
+\begin{equation}
+  D_{n,l+1/2} = \frac{\Delta_{l+1} D_{n,l+1} D_{n,l} + \Delta_l
+    D_{n,l+1} D_{n,l}}{\Delta_l D_{n,l+1} + \Delta_{l+1} D_{n,l}} \:.
+  \label{eq:face_current_solve_6}
+\end{equation}
+With this diffusion coefficient, we can then finish the derivation of
+the cell-face currents by inserting Eq~(\ref{eq:face_current_solve_6})
+into Eq~(\ref{eq:face_current}) giving:
+\begin{equation}
+  J_{n,l+1/2} = -2 \frac{D_{n,l+1} D_{n,l}}{\Delta_l D_{n,l+1} +
+    \Delta_{l+1} D_{n,l}} (u_{n,l+1} - u_{n,l})\:.
+  \label{eq:face_current_solve_7}
+\end{equation}
+Analagously, for cell face in the opposite direction, we have:
+\begin{equation}
+  J_{n,l-1/2} = -2 \frac{D_{n,l} D_{n,l-1}}{\Delta_{l-1} D_{n,l} +
+    \Delta_{l} D_{n,l-1}} (u_{n,l} - u_{n,l-1})\:.
+  \label{eq:face_current_solve_8}
+\end{equation}
+
+Before inserting the derived face fluxes into
+Eq~(\ref{eq:discrete_current}), we divide by the cell volume to get:
+\begin{multline}
+(J_{n,i+1/2} - J_{n,i-1/2}) \frac{1}{\Delta_i} + (J_{n,j+1/2} -
+  J_{n,j-1/2}) \frac{1}{\Delta_j} +\\ (J_{n,k+1/2} - J_{n,k-1/2})
+  \frac{1}{\Delta_k} + \sum_{m=1}^4 A_{nm,ijk} u_{m,ijk} =
+  q_{n,ijk}\:,
+  \label{eq:discrete_current_2}
+\end{multline}
+Based on the face currents computed, we identify the following
+coefficients that will appear in the moment equations:
+\begin{equation}
+  C_{n,l+1} = \frac{2}{\Delta_l}\Bigg( \frac{D_{n,l+1}
+    D_{n,l}}{\Delta_l D_{n,l+1} + \Delta_{l+1} D_{n,l}}\Bigg)\:,
+  \label{eq:moment_coeff_plus}
+\end{equation}
+\begin{equation}
+  C_{n,l-1} = \frac{2}{\Delta_l}\Bigg( \frac{D_{n,l}
+    D_{n,l-1}}{\Delta_{l-1} D_{n,l} + \Delta_{l} D_{n,l-1}}\Bigg)\:.
+  \label{eq:moment_coeff_minus}
+\end{equation}
+Inserting these coefficients into Eqs~(\ref{eq:face_current_solve_7})
+and (\ref{eq:face_current_solve_8}) gives:
+\begin{equation}
+  J_{n,l+1/2} = - \Delta_l C_{n_l+1} (u_{n,l+1} - u_{n,l})\:.
+  \label{eq:face_current_solve_9}
+\end{equation}
+\begin{equation}
+  J_{n,l-1/2} = - \Delta_l C_{n_l-1} (u_{n,l} - u_{n,l-1})\:.
+  \label{eq:face_current_solve_10}
+\end{equation}
+Applying Eqs~(\ref{eq:face_current_solve_9}) and
+(\ref{eq:face_current_solve_10}) Eq~(\ref{eq:discrete_current_2})
+gives the final spatially discretized $SP_N$ equations:
+\begin{multline}
+  - C_{n,i+1} u_{n_i+1} - C_{n,i-1} u_{n_i-1} - C_{n,j+1} u_{n_j+1} -
+  C_{n,j-1} u_{n_j-1} - C_{n,k+1} u_{n_k+1} - C_{n,k-1} u_{n_k-1} +\\
+  \sum_{m=1}^4 \Big[A_{nm,ijk} +
+    (C_{n,i+1}+C_{n,j+1}+C_{n,k+1}+C_{n,i-1}+C_{n,j-1}+C_{n,k-1})\delta_{nm}\Big]
+  u_{m,ijk} = q_{n,ijk}\:.
+  \label{eq:discrete_space_spn}
+\end{multline}
+
+On the boundaries of the domain, special treatment of the face
+currents must be made as their are no adjacent cells with which to
+close the system. Starting with Eq~(\ref{eq:spn_bnd_matrix}), we again
+insert the neutron current given by Eq~(\ref{eq:neutron_current}):
+\begin{equation}
+  \mathbf{n} \cdot J_n + \sum_{m=1}^4 B_{nm} u_m = s_n\:.
+  \label{eq:boundary_current_system}
+\end{equation}
+On the low boundary we then have:
+\begin{equation}
+  J_{n,1/2} = s_{n,1} - \sum_{m=1}^4 B_{nm,1/2} u_{m,1/2}\:,
+  \label{eq:low_bnd_current}
+\end{equation}
+or in terms of moments:
+\begin{equation}
+  J_{n,1/2} = -2 D_{n,1}\frac{u_{n,1} - u_{n,1/2}}{\Delta_1}\:.
+  \label{eq:low_bnd_current_2}
+\end{equation}
+Equating Eqs~(\ref{eq:low_bnd_current}) and
+(\ref{eq:low_bnd_current_2}) gives an extra equation for the boundary
+moments:
+\begin{equation}
+  s_{n,1} = \sum_{m=1}^4 B_{nm,1/2} u_{m,1/2} -2 D_{n,1}\frac{u_{n,1}
+    - u_{n,1/2}}{\Delta_1}\:,
+  \label{eq:low_bnd_current_3}
+\end{equation}
+or
+\begin{equation}
+  s_{n,1} = \sum_{m=1}^4 \Big[B_{nm,1/2} + \frac{2
+      D_{n,1}}{\Delta_1}\delta_{mn} \Big] u_{m,1/2} -2
+  \frac{D_{n,1}}{\Delta_1} u_{n,1}\:.
+  \label{eq:low_bnd_current_4}
+\end{equation}
+To use these boundary fluxes to close the system, we insert
+Eq~(\ref{eq:low_bnd_current_2}) into Eq~(\ref{eq:discrete_current_2})
+and then add Eq~(\ref{eq:low_bnd_current_4}) to the system to
+eliminate the boundary currents introduced by
+Eq~(\ref{eq:low_bnd_current_2}).
+
+For the multigroup $SP_N$ equations, the spatial discretization is
+identical, this time with the multigroup moment vectors where now:
+\begin{multline}
+  - \mathbb{C}_{n,i+1} \mathbb{U}_{n_i+1} - \mathbb{C}_{n,i-1}
+  \mathbb{U}_{n_i-1} - \mathbb{C}_{n,j+1} \mathbb{U}_{n_j+1} -
+  \mathbb{C}_{n,j-1} \mathbb{U}_{n_j-1} - \mathbb{C}_{n,k+1}
+  \mathbb{U}_{n_k+1} - \mathbb{C}_{n,k-1} \mathbb{U}_{n_k-1}
+  +\\ \mathbb{U}m_{m=1}^4 \Big[\mathbb{A}_{nm,ijk} +
+    (\mathbb{C}_{n,i+1}+\mathbb{C}_{n,j+1}+\mathbb{C}_{n,k+1}+
+    \mathbb{C}_{n,i-1}+\mathbb{C}_{n,j-1}+\mathbb{C}_{n,k-1})\delta_{nm}\Big]
+  \mathbb{U}_{m,ijk} = \mathbb{Q}_{n,ijk}\:,
+  \label{eq:discrete_mg_space_spn}
+\end{multline}
+and the coefficients are:
+\begin{equation}
+  \mathbb{C}_{n,l+1} = \frac{2}{\Delta_l}\Bigg( \frac{\mathbb{D}_{n,l+1}
+    \mathbb{D}_{n,l}}{\Delta_l \mathbb{D}_{n,l+1} + \Delta_{l+1} \mathbb{D}_{n,l}}\Bigg)\:,
+  \label{eq:mg_coeff_plus}
+\end{equation}
+\begin{equation}
+  \mathbb{C}_{n,l-1} = \frac{2}{\Delta_l}\Bigg( \frac{\mathbb{D}_{n,l}
+    \mathbb{D}_{n,l-1}}{\Delta_{l-1} \mathbb{D}_{n,l} + \Delta_{l}
+    \mathbb{D}_{n,l-1}}\Bigg)\:.
+  \label{eq:mg_coeff_minus}
+\end{equation}
+The boundary conditions are analagously formulated with the multigroup
+matrices/vectors.

### 2013-05-04 doc/dissertation/document/backmatter/pn_equations.tex
adding pn fission source derivation

@@
 also holds for vacuum conditions where $\phi_b = 0$.
+
+\section{Eigenvalue Form of the $P_N$ Equations}
+\label{sec:pn_eigenvalue}
+In addition to fixed source problems, the $P_N$ equations may be
+modified to solve eigenvalue problems by applying the same
+discretization techniques to the fission source in
+Eq~(\ref{eq:eigenvalue_transport}). Considering the multigroup
+case, we first replace the integration over energy with a summation
+over energy groups:
+\begin{multline}
+\frac{1}{k} \chi(E) \iint \nu \sigma_f(x,E') \psi(x,\hat{\Omega}',E')
+d\Omega' dE' =\\ \frac{1}{k} \sum_0^G \chi^g \int \nu \sigma^{g'}_f(x)
+\psi^{g'}(x,\hat{\Omega}') d\Omega'\:,
+  \label{eq:mg_fission_source}
+\end{multline}
+where $G = N_g-1$ and $N_g$ is the number of energy groups. Next, we
+expand the angular flux in the fission term by applying the expansion
+in Eq~(\ref{eq:flux_expansion}) and remove the dependence on the
+azimuthal angle:
+\begin{multline}
+\frac{1}{k} \sum_0^G \chi^g \int \nu \sigma^{g'}_f(x)
+\psi^{g'}(x,\hat{\Omega}') d\Omega' =\\ \frac{1}{k} \sum_0^G \chi^g
+\int_{-1}^{1} \nu \sigma^{g'}_f(x) \sum_{n=0}^\infty
+(2n+1)P_n(\mu)\phi^{g'}_n(x) d\mu'\:.
+  \label{eq:fission_flux_expansion}
+\end{multline}
+We can again use the the orthogonality property of the Legendre
+polynomials by multplying by $P_0(\mu) = 1$:
+\begin{equation}
+  \frac{1}{k} \sum_0^G \chi^g
+  \int_{-1}^{1} \nu \sigma^{g'}_f(x) \sum_{n=0}^\infty
+  (2n+1)P_n(\mu) P_0(\mu) \phi^{g'}_n(x) d\mu'\:,
+  \label{eq:fission_deriv_1}
+\end{equation}
+giving:
+\begin{equation}
+  \frac{1}{k} \sum_0^G \chi^g \nu \sigma^{g'}_f(x) \phi^{g'}_n(x)
+  \delta_{n0}\:,
+  \label{eq:fission_deriv_1}
+\end{equation}
+where $\chi^g$ and $\sigma^{g'}_f\phi^{g'}$ means that an effective
+inner product amongst group terms will be computed with the discrete
+$\chi$ spectrum and the fission reaction rate. In a multigoup form, if
+we choose to represent the energy dependent moments as vectors,
+$\mathbf{\Phi_n}$, composed of all energy groups, then this inner
+product takes the form of a spatially dependent fission matrix
+$\mathbf{F}$:
+\begin{equation}
+  \mathbf{F} = \nu
+  {\tiny \begin{bmatrix}
+    (\chi^0 \sigma_f^0) & (\chi^0 \sigma_f^1) & \cdots & (\chi^0
+      \sigma_f^G) \\
+    %%
+    &&&\\
+    %%
+    (\chi^1 \sigma_f^0) & (\chi^1 \sigma_f^1) & \cdots & (\chi^1
+      \sigma_f^G) \\
+    %%
+    &&&\\
+    %%
+      \vdots & \vdots & \ddots & \vdots \\ 
+    %%
+    &&&\\
+    %%
+    (\chi^G \sigma_f^0) & (\chi^G \sigma_f^1) & \cdots & (\chi^G
+      \sigma_f^G) \\
+  \end{bmatrix}}\:.
+  \label{eq:fission_matrix}
+\end{equation}
+This gives a final multigroup $P_N$ fission source of:
+\begin{equation}
+  \frac{1}{k} \mathbf{F} \mathbf{\Phi_n} \delta_{n0}
+  \label{eq:pn_fission_source}
+\end{equation}
+where the spatial dependence of the flux moments and fission matrix
+has been supressed.

### 2013-05-04 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding pn fission source derivation

@@
 where $g$ denotes the group index of $0$ to $G$ groups, $G=N_g-1$, and
-the integration of the scattering emission term of energy has been
+the integration of the scattering emission term over energy has been
 replaced by a discrete summation. For scattering, $\sigma_s^{gg'}$
@@
 presented have discretized the transport equation in angle and energy
-but have yet to consider the spatial component of phase space. We
-don't explicitly consider it in this document but instead we will
-briefly comment on possible means of discretization. Of primary
-interest here is the discretization of the diffusive term in
+but have yet to consider the spatial component of phase space. Of
+primary interest here is the discretization of the diffusive term in
 Eq~(\ref{eq:spn_multigroup_system}) and the current term on the
@@
 for spatial discretization are available here and include finite
-difference and finite element formulations. In the Denovo package at
-ORNL, Evans has implemented a finite volume scheme for these
-equations resulting in:
+difference and finite element formulations. In the Exnihilo package at
+ORNL, Evans has implemented a finite volume scheme over a Cartesian
+grid for these equations resulting in:
+\begin{multline}
+  - \mathbb{C}_{n,i+1} \mathbb{U}_{n_i+1} - \mathbb{C}_{n,i-1}
+  \mathbb{U}_{n_i-1} - \mathbb{C}_{n,j+1} \mathbb{U}_{n_j+1} -
+  \mathbb{C}_{n,j-1} \mathbb{U}_{n_j-1} - \mathbb{C}_{n,k+1}
+  \mathbb{U}_{n_k+1} - \mathbb{C}_{n,k-1} \mathbb{U}_{n_k-1}
+  +\\ \sum_{m=1}^4 \Big[\mathbb{A}_{nm,ijk} +
+    (\mathbb{C}_{n,i+1}+\mathbb{C}_{n,j+1}+\mathbb{C}_{n,k+1}+
+    \mathbb{C}_{n,i-1}+\mathbb{C}_{n,j-1}+\mathbb{C}_{n,k-1})\delta_{nm}\Big]
+  \mathbb{U}_{m,ijk} = \mathbb{Q}_{n,ijk}\:,
+  \label{eq:spn_space}
+\end{multline}
+with the coefficients defined as:
 \begin{equation}
-  discretization
-  \label{eq:spn_spatial_discretization}
+  \mathbb{C}_{n,l+1} = \frac{2}{\Delta_l}\Bigg( \frac{\mathbb{D}_{n,l+1}
+    \mathbb{D}_{n,l}}{\Delta_l \mathbb{D}_{n,l+1} + \Delta_{l+1} \mathbb{D}_{n,l}}\Bigg)\:,
+  \label{eq:spn_coeff_plus}
 \end{equation}
+\begin{equation}
+  \mathbb{C}_{n,l-1} = \frac{2}{\Delta_l}\Bigg( \frac{\mathbb{D}_{n,l}
+    \mathbb{D}_{n,l-1}}{\Delta_{l-1} \mathbb{D}_{n,l} + \Delta_{l}
+    \mathbb{D}_{n,l-1}}\Bigg)\:,
+  \label{eq:spn_coeff_minus}
+\end{equation}
+where $l={i,j,k}$ along the cardinal directions of the Cartesian grid
+cell in Figure~\ref{fig:mesh_cell}. The details of the derivation are
+provided in Appendix~\ref{chap:spn_spatial_discretization}.
+
 Although arbitrary grids could be handled effectively
@@
 \label{subsec:eigenvalue_form}
+For criticality problems, the space, angle, and energy discretizations
+for the $SP_N$ equations can be applied to the general eigenvalue from
+of the transport equation given by
+Eq~(\ref{eq:eigenvalue_transport}). We can readily replace the fixed
+source in Eq~(\ref{eq:multigroup_spn_equations}) with the fission
+source derived in \S~\ref{sec:pn_eigenvalue}:
+\begin{multline}
+  -\nabla \cdot \Bigg[\frac{n}{2n+1}\mathbf{\Sigma_{n-1}}^{-1} \nabla
+    \Big(\frac{n-1}{2n-1} \mathbf{\Phi_{n-2}} +
+    \frac{n}{2n-1}\mathbf{\Phi_n} \Big) \\+
+    \frac{n+1}{2n+1}\mathbf{\Sigma_{n+1}}^{-1} \nabla
+    \Big(\frac{n+1}{2n+3}\mathbf{\Phi_n} +
+    \frac{n+2}{2n+3}\mathbf{\Phi_{n+2}}\Big) \Bigg] \\+
+  \mathbf{\Sigma_n} \mathbf{\Phi_n} = \frac{1}{k} \mathbf{F}
+  \mathbf{\Phi_n} \delta_{n0} \ \ \ \ \ \ \ \ \ n = 0,2,4,\cdots,N\:.
+  \label{eq:multigroup_spn_eigenvalue}
+\end{multline}
+with the fission matrix $\mathbf{F}$ given by
+Eq~(\ref{eq:fission_matrix}).
 

### 2013-05-04 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding some to the spn fission section

@@
 with the fission matrix $\mathbf{F}$ given by
-Eq~(\ref{eq:fission_matrix}).
+Eq~(\ref{eq:fission_matrix}). We again apply the change of variables
+to yield the set of pseudo-moments $\mathbb{U}_n$ where now the
+application of the fission matrix to the moment vectors will be
+expanded into a set of block matrices in identical fashion to the
+scattering matrices.
+
+Finally, we can apply the spatial discretization to yield the general
+$SP_N$ eigenvalue problem.
+
+It should be noted here that with respect to a general MCSA scheme,
+the introduction of the fission in the system will only effect the
+source vector in the linear system. The linear operator, and therefore
+overall MCSA performance will be dictated by streaming and scattering
+as defined on the left-hand side.
 

### 2013-05-06 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding jacobi preconditioned fuel assembly data to doc

@@
 \end{multline}
-with the addition variables defined as
+with the additional variables defined as
 \begin{itemize}
@@
 conventional and Monte Carlo methods, we must first understand their
-spectral character to ensure the MCSA will be applicable. As presented
+spectral character to ensure that MCSA will be applicable. As presented
 in Chapter~\ref{ch:stochastic_methods}, the spectral radius of the
@@
 \end{table}
-It is readily apparent from the tabulated data the point Jacobi
-preconditioning is insufficient as a preconditioning mechanism for
-problems of order $SP_5$ or higher. In addition, as upscatter is
-introduced into the system, a much larger spectral radius is observed
-at lower $SP_N$ orders, signaling a need for a better preconditioning
-strategy to both ensure and improve covergence for Monte Carlo
-methods.
+It is readily apparent from the tabulated data that point Jacobi
+preconditioning is insufficient. For problems of order $SP_5$ and
+$SP_7$, the method will not converge at all and for the upscatter case
+with $SP_3$ the spectral radius is still quite large for large $P_N$
+orders. This data signals a need for a better preconditioning strategy
+to both ensure and improve covergence for Monte Carlo methods.
 
@@
 the materials configuration along the horizontal axis is given by
-Figure~\ref{fig:problem3_radial_mat}.
+Figure~\ref{fig:problem3_axial_mat}.
 \begin{figure}[t!]
@@
 core plenum. Table~\ref{tab:problem3_parameters} gives the primary
-design parameters for the fuel assembly calculations.
+design parameters for the fuel assembly calculations. Group cross
+sections and other discrete nuclear data are generated as needed by a
+cross-section processing module.
 \begin{table}[h!]
@@
     \State $k_0 =$ initial guess
-    \State $\ve{\phi_0} =$ initial guess
+    \State $\mathbf{\Phi}_0 =$ initial guess
     \State $n = 0$
@@
     \Comment{Iterate until convergence of the eigenvalue}
-    \State $\ve{M} \ve{\phi}^{n+1} = \frac{1}{k^n} \ve{F} \ve{\phi}^n$
+    \State $\mathbf{M} \mathbf{\Phi}^{n+1} = \frac{1}{k^n} \mathbf{F} \mathbf{\Phi}^n$
     \Comment{Solve for the new flux state with MCSA}
-    \State $k^{n+1} = k^n \frac{\int \ve{F} \ve{\phi}^{n+1} d\ve{r}}{\int
-      \ve{F} \ve{\phi}^n d\ve{r}}$
+    \State $k^{n+1} = k^n \frac{\int \mathbf{F} \mathbf{\Phi}^{n+1} d\mathbf{r}}{\int
+      \mathbf{F} \mathbf{\Phi}^n d\mathbf{r}}$
     \Comment{Update the multiplication factor}
@@
 \end{algorithm}
-Here, $\ve{M}$ is the transport operator generated on the left-hand
-side of the $SP_N$ discretization and $\ve{F}$ is the fission matrix.
-This problem is significantly more complicated that the simple test
-problem used for the previous spectral analysis. Fission has been
-introduced into the set of equations and the addition of moderator
-into the system will increase the amount of scattering, creating a
-significantly more difficult problem.
+Here, $\mathbf{M}$ is the transport operator generated on the
+left-hand side of the $SP_N$ discretization, $\mathbf{F}$ is the
+fission matrix, and $\mathbf{\Phi}$ the multigroup neutron flux. This
+problem is significantly more complicated than the simple test problem
+used for the previous spectral analysis. Fission has been introduced
+into the set of equations and the addition of moderator into the
+system will increase the amount of scattering, creating a
+significantly more difficult problem manifesting itself in an
+iteration matrix with a larger spectral radius. When using MCSA, the
+linear operator applied to $\mathbf{\Phi}^{n+1}$ at each eigenvalue
+iteration will dictate converge and remain unchanged throughout the
+computation while the addition of fission to the system will only
+modify the source of neutrons and not affect Monte Carlo transport.
 
@@
 \label{subsec:jacobi_prec_assembly_calc}
-
-\subsection{MCSA Break Down}
+Based on the success of block Jacobi preconditioning with the test
+problem used for the spectral radius parameter study, we use it first
+to solve the $17 \times 17$ fuel assembly problem. A single energy
+group problem was first solved with $SP_1$ discretization, effectively
+giving the one-speed neutron diffusion system for the fuel assembly
+resulting in 20,088 degrees of freedom in the
+problem. Figure~\ref{fig:block_jacobi_res_mcsa} gives the residual
+infinity norm as a function of iteration for the MCSA linear solve in
+the first eigenvalue iteration.
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=5in]{chapters/spn_equations/block_jacobi_res.pdf}
+  \end{center}
+  \caption{\textbf{Residual infinity norm vs. iteration for the block
+      Jacobi preconditioned MCSA solve during the first eigenvalue
+      iteration of the 1-group $17 \times 17$ fuel assembly problem.}
+    \textit{Convergence was not achieved with the block Jacobi
+      preconditioned method.}}
+  \label{fig:block_jacobi_res_mcsa}
+\end{figure}
+Convergence was not achieved as noted by the rapid rise in the
+residual over a few iterations. Based on the spectral radius
+computations performed, these results are not in line with
+expectations for this problem. Even if the problem may be
+ill-conditioned, we do expect convergence of MCSA. To investigate
+further, a block Jacobi preconditioned Richardson iteration was used
+to solve the same problem. Figure~\ref{fig:block_jacobi_res_richardson}
+gives the residual infinity norm as a function of iteration for the
+Richardson linear solve in the first eigenvalue iteration.
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=5in]{chapters/spn_equations/block_jacobi_rich_res.pdf}
+  \end{center}
+  \caption{\textbf{Residual infinity norm vs. iteration for the block
+      Jacobi preconditioned Richardson solve during the first
+      eigenvalue iteration of the 1-group $17 \times 17$ fuel assembly
+      problem.}  \textit{A spectral radius near 1 was observed for the
+      iteration matrix.}}
+  \label{fig:block_jacobi_res_richardson}
+\end{figure}
+Poor converge is observed for the Richardson iteration. However,
+convergence is achieved meaning that the criteria needed to satisfy
+MCSA convergence has also been met. The effective slope of the
+convergence curve in Figure~\ref{fig:block_jacobi_res_richardson} will
+give an approximation for the spectral radius via
+Eq~(\ref{eq:linear_k_iter_norm3}). Based on these results, it then
+appears that even if the simple criteria of a spectral radius of less
+than one is met and the Richardson iteration will converge with the
+same preconditioning, MCSA still may not converge. We then expect the
+issue to reside with the Neumann-Ulam solve providing the correction
+as the Richardson iteration is known to provide the correct
+result. Furthermore, the fact that the spectral radius is less than
+one means that the stochastic histories in the block Jacobi
+preconditioned Neumann-Ulam method are eventually being terminated by
+the weight cutoff as no artificial absorption was used. Based on this,
+the correction being generated by the Neumann-Ulam solve must be
+incorrect.
+
+\subsection{MCSA Breakdown}
 \label{subsec:mcsa_break_down}
+For the fuel assembly problem, the initial block Jacobi calculations
+used a one-speed $SP_1$ discretization, effectively giving a diffusion
+system. To study the breakdown of MCSA at iteration matrix spectral
+radii near one, we will use a simpler homogoneous 2-dimensional
+one-speed neutron diffusion system to isolate this behavior. In this
+system, we can vary the cross sections in order to achieve varying
+spectral radii.
 
@@
 
+\subsubsection{MCSA Relaxation Parameters}
+\label{subsubsec:spn_mcsa_relaxation}
+
 \subsubsection{Monte Carlo Estimator Comparison}

### 2013-05-06 doc/dissertation/document/chapters/mc_background/mc_background.tex
starting reduced domain approximation in mc background

@@
 \end{figure}
+\clearpage
 
@@
 \label{subsec:reduced_domain_approximation}
+For scalable parallel implementations and efficient Monte Carlo
+transport, the domain is required to be sparse. We may find through
+general preconditioning operations as outlined by the Monte Carlo
+problem in Eq~(\ref{eq:left_right_correction}) that this sparsity is
+lost. To alleviate this, we propose modifying the domain sampled by
+the Monte Carlo method to generate the correction within MCSA by
+removing terms in the iteration matrix through some predetermined
+criteria in order to recover sparsity. As the Neumann-Ulam scheme is a
+stochastic realization of multiple matrix-vector multiplies, the
+criteria we choose should be based on maintain those elements that
+will make the largest contributions to the multiplication. For each
+row in the preconditioned system, we can then apply the reduced domain
+approximation to eliminate those elements that either fall below a
+specified tolerance level, keep the $N$ largest elements in a row, or
+both giving:
+\begin{subequations}
+  \begin{gather}
+    \ve{x}^{k+1/2} = \ve{x}^k + \ve{r}^k\:,\\
+    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\
+    \hat{\ve{A}}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:, 
+    \label{eq:rda_correction}\\
+    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,\\
+    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:,
+  \end{gather}
+  \label{eq:rda_mcsa}
+\end{subequations}
+where $\hat{\ve{A}}$ in Eq~(\ref{eq:rda_correction}) means that the
+reduced domain approximation has been applied to the iteration matrix
+in the Monte Carlo method. If enough of the terms in the original
+iteration matrix have been maintained, the character of the original
+iteration matrix should be retained and continue to accelerate the
+iterative scheme.
 

### 2013-05-06 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
adding expected value estimator
adding history transport kernel

@@
 \label{subsec:transition_processing}
+Processing the transition of a single stochastic history from one
+state to another in the system is the fundamental kernel of the
+transport procedure. In this kernel, an incoming history has a state
+consisting of the current discrete state in the system in which it
+resides and its current weight and is processed as given by
+Algorithm~\ref{alg:transition_process}.
+\begin{algorithm}[h!]
+  \caption{Stochastic History Transition Processing Kernel}
+  \label{alg:power_iteration}
+  \begin{algorithmic}
+    \State processTransition( history ):
+    \State i = history.state
+    \State j = sampleRowCDF( C[i] )
+    \State history.state = j
+    \State history.weight *= w[i,j]
+    \EndWhile
+  \end{algorithmic}
+\end{algorithm}
+The core component here is the sampling of the CDF for a particular
+state generated by the probablities from the Neumann-Ulam
+decomposition. Upon exiting, the kernel has modified the state of the
+history with a new state and multiplied the weight with the
+corresponding transition weight to form the product in
+Eq~(\ref{eq:direct_permutation_weight}).
 
@@
 
-%%---------------------------------------------------------------------------%%
-\section{Parallel MCSA}
-\label{sec:parallel_mcsa}
-With the parallel adjoint Neumann-Ulam solver implementation described
-aboive, the parallel implementation of the MCSA method is
-trivial. Recall the MCSA iteration procedure outlined in
-Eq~(\ref{eq:mcsa}). In \S\ref{sec:parallel_krylov_methods} we
-discussed parallel matrix and vector operations as utilized in
-conventional Krylov methods. We utilize these here for the parallel
-MCSA implementation. In the first step, a parallel matrix-vector
-multiply is used to apply the split operator to the previous iterate's
-solution. A parallel vector update is then performed with the source
-vector to arrive at the initial iteration guess. In the next step, the
-residual is computed by the same operations where now the operator is
-applied to the solution guess with a parallel matrix-vector multiply
-and then a parallel vector update with the source vector is
-performed. Once the correction is computed with a parallel adjoint
-Neumann-Ulam solve, this correction is applied to the guess with a
-parallel vector update to get the new iteration
-solution. Additionally, as given by
-Eq~(\ref{eq:mcsa_stopping_criteria}), 2 parallel vector reductions
-will be required to check the stopping criteria: one initially to
-compute the infinity norm of the source vector, and another at every
-iteration to compute the infinity norm of the residual vector. For
-this implementation, all of the issues that will be potentially
-generated by the parallel adjoint solver implementation will manifest
-themselves here as the quality of the correction will be of intense
-study.
-
-%%---------------------------------------------------------------------------%%
-\section{Load Balancing Concerns}
+\subsection{Load Balancing Concerns}
 \label{sec:mc_load_balancing}
@@
 
-%%---------------------------------------------------------------------------%%
-\section{Reproducible Domain Decomposed Results}
+\subsection{Reproducible Domain Decomposed Results}
 \label{sec:reproducible_mc}
@@
 %%---------------------------------------------------------------------------%%
+\section{Parallel MCSA}
+\label{sec:parallel_mcsa}
+With the parallel adjoint Neumann-Ulam solver implementation described
+aboive, the parallel implementation of the MCSA method is
+trivial. Recall the MCSA iteration procedure outlined in
+Eq~(\ref{eq:mcsa}). In \S\ref{sec:parallel_krylov_methods} we
+discussed parallel matrix and vector operations as utilized in
+conventional Krylov methods. We utilize these here for the parallel
+MCSA implementation. In the first step, a parallel matrix-vector
+multiply is used to apply the split operator to the previous iterate's
+solution. A parallel vector update is then performed with the source
+vector to arrive at the initial iteration guess. In the next step, the
+residual is computed by the same operations where now the operator is
+applied to the solution guess with a parallel matrix-vector multiply
+and then a parallel vector update with the source vector is
+performed. Once the correction is computed with a parallel adjoint
+Neumann-Ulam solve, this correction is applied to the guess with a
+parallel vector update to get the new iteration
+solution. Additionally, as given by
+Eq~(\ref{eq:mcsa_stopping_criteria}), 2 parallel vector reductions
+will be required to check the stopping criteria: one initially to
+compute the infinity norm of the source vector, and another at every
+iteration to compute the infinity norm of the residual vector. For
+this implementation, all of the issues that will be potentially
+generated by the parallel adjoint solver implementation will manifest
+themselves here as the quality of the correction will be of intense
+study.
+
+%%---------------------------------------------------------------------------%%
 \section{Parallel Scaling Studies}

### 2013-05-13 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
updating parallel mc and spn chapters

@@
 adapted to the domain decomposed adjoint Neumann-Ulam method using the
-one-speed, two-dimensional neutron diffusion equation. In this subsection
-we use a one-speed two-dimensional neutron diffusion system as the
-symmetric model problem. Using the linear system generated by the
-discretization of the model problem, we use a spectral analysis to
-generate analytic relations for the eigenvalues of the operator based
-on system parameters. Using the eigenvalue spectra, we then build
-relationships to characterize the transport of stochastic histories in
-a decomposed domain and the fraction of histories that leak from a
-domain and will therefore have to be communicated. Finally, we compare
-these analytic results to numerical experiments conducted with the
-model problem.
-
-\subsubsection{Symmetric Model Problem}
-\label{subsec:model_problem}
-For our numerical experiments, we choose the one-speed,
-two-dimensional neutron diffusion equation as a model problem
-\citep{duderstadt_nuclear_1976}:
-\begin{equation}
-  -\boldsymbol{\nabla} \cdot D \boldsymbol{\nabla} \phi + \Sigma_a
-  \phi = S\:,
-  \label{eq:diffusion_eq}
-\end{equation}
-where $\phi$ is the neutron flux, $\Sigma_a$ is the absorption cross
-subsection, and $S$ is the source of neutrons. In addition, $D$ is the
-diffusion coefficient defined as:
-\begin{equation}
-  D = \frac{1}{3 ( \Sigma_t - \bar{\mu}\Sigma_s )}\:,
-  \label{eq:diffusion_coeff}
-\end{equation}
-where $\Sigma_s$ is the scattering cross subsection, $\Sigma_t = \Sigma_a
-+ \Sigma_s$ is the total cross subsection, and $\bar{\mu}$ is the cosine
-of the average scattering angle. For simplicity, we will take
-$\bar{\mu} = 0$ for our analysis giving $D=(3 \Sigma_t)^{-1}$. In
-addition, to further simplify we will assume a homogeneous domain such
-that the cross subsections remain constant throughout. Doing this permits
-us to rewrite Eq~(\ref{eq:diffusion_eq}) as:
-\begin{equation}
-  -D \boldsymbol{\nabla}^2 \phi + \Sigma_a \phi = S\:.
-  \label{eq:diffusion_eq_simple}
-\end{equation}
-
-We choose a finite difference scheme on a square Cartesian grid to
-discretize the problem. For the Laplacian, we choose the 9-point
-stencil shown in Figure~\ref{fig:stencil} over a grid of size $h$
-\citep{leveque_finite_2007}:
-\begin{multline}
-  \nabla^2_9\phi = \frac{1}{6h^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j}
-    + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ +
-    \phi_{i-1,j+1} + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20
-    \phi_{i,j}]\:.
-  \label{eq:nine_point_stencil}
-\end{multline}
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.25}{\input{chapters/mc_theory/stencil.pdftex_t}}
-  \end{center}
-  \caption{\textbf{Nine-point Laplacian stencil.}}
-  \label{fig:stencil}
-\end{figure}
-We then have the following linear system to solve:
-\begin{multline}
-  -\frac{1}{6h^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j} + 4
-    \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ + \phi_{i-1,j+1}
-    + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20 \phi_{i,j}] + \Sigma_a
-  \phi_{i,j} = s_{i,j}\:,
-  \label{eq:fd_system}
-\end{multline}
-and in operator form:
-\begin{equation}
-  \ve{D}\boldsymbol{\phi}=\ve{s}\:,
-  \label{eq:operator_system}
-\end{equation}
-where $\ve{D}$ is the diffusion operator, $\ve{s}$ is the source in
-vector form and $\boldsymbol{\phi}$ is the vector of unknown fluxes.
-
-To close the system, a set of boundary conditions is required. In the
-case of a non-reentrant current condition applied to all global
-boundaries of the domain, we choose the formulation of Duderstadt by
-assuming the flux is zero at some ghost point beyond the
-grid. Consider for example the equations on the $i=0$ boundary of the
-domain:
-\begin{multline}
-  -\frac{1}{6h^2}[4 \phi_{-1,j} + 4 \phi_{1,j} + 4 \phi_{0,j-1} +
-    4 \phi_{0,j+1} + \phi_{-1,j-1}\\ + \phi_{-1,j+1} + \phi_{1,j-1} +
-    \phi_{1,j+1} - 20 \phi_{0,j}] + \Sigma_a \phi_{0,j} = s_{0,j}\:.
-  \label{eq:x_min_bnd}
-\end{multline}
-Here we note some terms where $i=-1$ and therefore are representative
-of grid points beyond the boundary of the domain. We set the flux at
-these points to be zero, giving a valid set of equations for the $i=0$
-boundary:
-\begin{multline}
-  -\frac{1}{6h^2}[4 \phi_{1,j} + 4 \phi_{0,j-1} + 4 \phi_{0,j+1}
-    \\ + \phi_{-1,j+1} + \phi_{1,j-1} + \phi_{1,j+1} - 20 \phi_{0,j}]
-  + \Sigma_a \phi_{0,j} = s_{0,j}\:.
-  \label{eq:x_min_bnd_2}
-\end{multline}
-We repeat this procedure for the other boundaries of the domain. For
-reflecting boundary conditions, the net current across a boundary is
-zero.
+one-speed, two-dimensional neutron diffusion equation and spatial
+discretization presented in \S~\ref{subsec:model_problem}. Using the
+linear system generated by the discretization of the model problem, we
+use a spectral analysis to generate analytic relations for the
+eigenvalues of the operator based on system parameters. Using the
+eigenvalue spectra, we then build relationships to characterize the
+transport of stochastic histories in a decomposed domain and the
+fraction of histories that leak from a domain and will therefore have
+to be communicated. Finally, we compare these analytic results to
+numerical experiments conducted with the model problem.
 
@@
   \begin{center}
-    \includegraphics[width=5in,clip]{chapters/mc_theory/diffusion_spectrum.png}
+    \includegraphics[width=5in,clip]{chapters/parallel_mc/diffusion_spectrum.png}
   \end{center}
@@
     \begin{center}
-      \includegraphics[width=5in,clip]{chapters/mc_theory/measured_spec_rad.png}
+      \includegraphics[width=5in,clip]{chapters/parallel_mc/measured_spec_rad.png}
     \end{center}
@@
 histories to get a measured value of $k$ for the particular
-operator. Figure~\ref{fig:measured_length} presents these measurements
-as well as the analytic result computed by Eq~(\ref{eq:analytic_k}) as
-a function of the iteration matrix spectral radius,
-$\rho(\ve{H})$. Figure~\ref{fig:measured_length_error} gives the
-relative error between the predicted and observed results. We note
-good qualitative agreement between the measured and analytic
-results. However, we observe a larger relative error for both long and
-short random walks.
+operator. On the left, Figure~\ref{fig:measured_length} presents these
+measurements as well as the analytic result computed by
+Eq~(\ref{eq:analytic_k}) as a function of the iteration matrix
+spectral radius, $\rho(\ve{H})$. On the right,
+Figure~\ref{fig:measured_length} gives the relative difference between the
+predicted and observed results. We note good qualitative agreement
+between the measured and analytic results. However, we observe a
+larger relative difference for both long and short random walks.
 \begin{figure}[t!]
@@
     \begin{center}
-      \includegraphics[width=5in,clip]{chapters/mc_theory/walk_length.png}
+      \includegraphics[width=6.0in,clip]{chapters/parallel_mc/measured_length_2.pdf}
     \end{center}
@@
         weight cutoff was varied with \sn{1}{-2}, \sn{1}{-4}, and
-        \sn{1}{-8}. The red data was computed numerically by an
-        adjoint Neumann-Ulam implementation while the black dashed
-        data was generated by Eq~(\ref{eq:analytic_k}).}}
+        \sn{1}{-8}. In the left plot, the red data was computed
+        numerically by an adjoint Neumann-Ulam implementation while
+        the black dashed data was generated by
+        Eq~(\ref{eq:analytic_k}). In the right plot, the relative
+        difference between the predicted and measured results is
+        presented for each weight cutoff.}}
     \label{fig:measured_length}
-  \end{spacing}
-\end{figure}
-\begin{figure}[t!]
-  \begin{spacing}{1.0}
-    \begin{center}
-      \includegraphics[width=5in,clip]{chapters/mc_theory/walk_length_error.png}
-    \end{center}
-    \caption{\textbf{Measured and analytic random walk length relative
-        error as a function of the iteration matrix spectral radius.}
-      \textit{The weight cutoff was varied with \sn{1}{-2},
-        \sn{1}{-4}, and \sn{1}{-8}. The relative error between the
-        predicted and measured results is presented for each weight
-        cutoff.}}
-    \label{fig:measured_length_error}
   \end{spacing}
@@
     \begin{center}
-      \includegraphics[width=5.25in,clip]{chapters/mc_theory/leakage_variation.png}
+      \includegraphics[width=6.0in,clip]{chapters/parallel_mc/leakage_variation_2.pdf}
     \end{center}
@@
 approximations. To compare the quality of the two approximations, the
-absolute error between the computed leakage fraction and that
+absolute difference between the computed leakage fraction and that
 generated by the Wigner rational and mean chord approximations is
@@
     \begin{center}
-      \includegraphics[width=5in,clip]{chapters/mc_theory/leakage_error.png}
+      \includegraphics[width=6.0in,clip]{chapters/parallel_mc/leakage_error_2.pdf}
     \end{center}
     \caption{\textbf{Measured and analytic domain leakage absolute
-        error as a function of the iteration matrix spectral radius.}
+        difference as a function of the iteration matrix spectral radius.}
       \textit{To test the behavior with respect to domain size,
         $n_i=50$ (green), $n_i=100$ (blue), and $n_i=200$ (red) were
-        used. The dashed lines represent the error using the Wigner
+        used. The dashed lines represent the difference using the Wigner
         rational approximation while the solid lines represent the
-        error using the mean-chord approximation.}}
+        difference using the mean-chord approximation.}}
     \label{fig:leakage_error}
@@
 \end{figure}
-From these error results, the mean chord approximation is shown to
-have a lower error for ill-conditioned systems as compared to the
+From these difference results, the mean chord approximation is shown to
+have a lower difference for ill-conditioned systems as compared to the
 Wigner approximation while the Wigner approximation produces less
-error for more well-conditioned systems. We also note that for the
-optically thick domains, the error is likely corresponded to that
+difference for more well-conditioned systems. We also note that for the
+optically thick domains, the difference is likely corresponded to that
 observed in Figure~\ref{fig:measured_length} for the $k$ parameter
-while the large relative error in $k$ for optically thin domains does
+while the large relative difference in $k$ for optically thin domains does
 not affect the approximation significantly. In general, the mean chord
@@
     \State history.weight *= w[i,j]
-    \EndWhile
   \end{algorithmic}

### 2013-05-13 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
updating parallel mc and spn chapters

@@
 $SP_N$ equations. If formulated correctly, we hypothesize that
-signficant improvement in usage of computational resources may be
+significant improvement in usage of computational resources may be
 observed compared to modern solution techniques such as those
@@
 In this chapter, we derive the $SP_N$ equations, closely following the
-work of Evans, in order to gain full understanding of the underlying
-system and its potential behavior in a Monte Carlo context. From the
-$P_N$ equations, we apply a set of approximations to yield the
-multi-dimensional, multi-group $SP_N$ equations for fixed source
-problems\footnote{The $SP_N$ derivations in this chapter are heavily
-  based on those presented by Evans in
-  \citep{evans_simpli_2013}}. Using the fully-formed matrix for the
-transport problem, we explore solutions to the $SP_N$ equations using
-Monte Carlo Synthetic Acceleration using a light water reactor fuel
-assembly criticality calculation as the driving problem.
+work of Evans\footnote{The $SP_N$ derivations in this chapter are
+  heavily based on those presented by Evans in
+  \citep{evans_simpli_2013}}, in order to gain full understanding of
+the underlying system and its potential behavior in a Monte Carlo
+context. From the $P_N$ equations, we apply a set of approximations to
+yield the multi-dimensional, multi-group $SP_N$ equations for fixed
+source problems. Using the fully-formed matrix for the transport
+problem, we explore solutions to the $SP_N$ equations using Monte
+Carlo Synthetic Acceleration using a light water reactor fuel assembly
+criticality calculation as the driving problem.
 
@@
 \end{subequations}
-If we rearrange the Eq~(\ref{eq:spn7_subs_equations}) such that only
-one divergence operation is present in each equation, we can formulate
+If we rearrange Eq~(\ref{eq:spn7_subs_equations}) such that only one
+divergence operation is present in each equation, we can formulate
 this as a matrix system of 4 equations in the case of the $SP_7$
@@
 the diffusive term in each equation is only for a single
-'psuedo-moment' $u_n$. This completes the derivation of the $SP_7$
-equations for a single energy group. As noted by Evans
-\citep{evans_simpli_2013}, lower order $SP_N$ approximations can be
-generated by setting higher order even moments in this system to zero
-(e.g. $\phi_6 = \phi_4 = 0$ yields the $SP_3$ equations). Boundary
-conditions for these equations are provided in
-Appendix~\ref{chap:spn_boundary_conditions}.
+'psuedo-moment' $u_n$. As noted by Evans \citep{evans_simpli_2013},
+lower order $SP_N$ approximations can be generated by setting higher
+order even moments in this system to zero (e.g. $\phi_6 = \phi_4 = 0$
+yields the $SP_3$ equations). Boundary conditions for these equations
+are provided in Appendix~\ref{chap:spn_boundary_conditions}.
 
@@
 
-Finally, we can apply the spatial discretization to yield the general
-$SP_N$ eigenvalue problem.
-
 It should be noted here that with respect to a general MCSA scheme,
-the introduction of the fission in the system will only effect the
+the introduction of the fission in the system will only affect the
 source vector in the linear system. The linear operator, and therefore
@@
 conventional and Monte Carlo methods, we must first understand their
-spectral character to ensure that MCSA will be applicable. As presented
-in Chapter~\ref{ch:stochastic_methods}, the spectral radius of the
-iteration matrix must be less than unity to ensure
+spectral character to ensure that MCSA will be applicable. As
+presented in Chapter~\ref{ch:stochastic_methods}, the spectral radius
+of the iteration matrix must be less than unity to ensure
 convergence. Therefore, we will compute the spectral radius of the
-$SP_N$ equations while parametrically varying the $SP_N$ order, the
-$P_N$ order, the number of energy groups, and
-upscattering/downscattering between energy groups. For each problem in
-the parameter study, the values given in
+iteration matrix formed by the $SP_N$ equations while parametrically
+varying the $SP_N$ order, the $P_N$ order, the number of energy
+groups, and upscattering/downscattering between energy groups. For
+each problem in the parameter study, the values given in
 Table~\ref{tab:spn_fixed_parameters} were fixed for each spatial
@@
 with $SP_3$ the spectral radius is still quite large for large $P_N$
-orders. This data signals a need for a better preconditioning strategy
-to both ensure and improve covergence for Monte Carlo methods.
+orders and therefore convergence is expected to be slow. These
+eigenvalues signal a need for a better preconditioning strategy to
+both ensure and improve covergence for Monte Carlo methods.
 
@@
 of configurations. As a sophisticated problem of interest to push the
-limits of MCSA, a $17 \times 17$ hot zero-power pin assembly will be
+limits of MCSA, a hot zero-power $17 \times 17$ pin assembly will be
 used with varying energy group structure and $SP_N$ discretization in
-a criticality calculation. A cross section along the vertical access
+a criticality calculation. A cross section along the vertical axis
 showing homogenized fuel pin materials and the associated grid is
@@
       spaced to capture material boundaries. Important details include
-      spacer grids along the length of the fuel pins and core
+      spacer grids along the length of the fuel pins and reactor core
       structures on the top and bottom of the assembly.}}
@@
 Significant geometric details are contained in the model including
-spacer grids, fuel pins with homogenized cladding and gas gap, and
-core plenum. Table~\ref{tab:problem3_parameters} gives the primary
-design parameters for the fuel assembly calculations. Group cross
-sections and other discrete nuclear data are generated as needed by a
-cross-section processing module.
+spacer grids, fuel pins with homogenized cladding and gas gap, core
+plenum, and moderator with boron. Table~\ref{tab:problem3_parameters}
+gives the primary design parameters for the fuel assembly
+calculations. Group cross sections and other discrete nuclear data are
+generated as needed by a cross-section processing module dependent on
+the meshing parameters used to discretize the geometry.
 \begin{table}[h!]
@@
 computation while the addition of fission to the system will only
-modify the source of neutrons and not affect Monte Carlo transport.
+modify the source of neutrons and the multiplication factor while not
+affecting Monte Carlo transport.
 
@@
 give an approximation for the spectral radius via
-Eq~(\ref{eq:linear_k_iter_norm3}). Based on these results, it then
-appears that even if the simple criteria of a spectral radius of less
-than one is met and the Richardson iteration will converge with the
-same preconditioning, MCSA still may not converge. We then expect the
-issue to reside with the Neumann-Ulam solve providing the correction
-as the Richardson iteration is known to provide the correct
-result. Furthermore, the fact that the spectral radius is less than
-one means that the stochastic histories in the block Jacobi
-preconditioned Neumann-Ulam method are eventually being terminated by
-the weight cutoff as no artificial absorption was used. Based on this,
-the correction being generated by the Neumann-Ulam solve must be
-incorrect.
+Eq~(\ref{eq:linear_k_iter_norm3}). For the data in
+Figure~\ref{fig:block_jacobi_res_richardson}, $\rho(\mathbf{H})
+\approx 0.985$, nearing the limits of MCSA applicability and well
+beyond the spectral radii generated in the initial spectral analysis.
+
+Based on these results, it then appears that even if the simple
+criteria of a spectral radius of less than one is met and the
+Richardson iteration will converge with the same preconditioning, MCSA
+still may not converge. We then expect the issue to reside with the
+Neumann-Ulam solve providing the correction as the Richardson
+iteration is known to provide the correct result. Furthermore, the
+fact that the spectral radius is less than one means that the
+stochastic histories in the block Jacobi preconditioned Neumann-Ulam
+method are eventually being terminated by the weight cutoff as no
+artificial absorption was used. Based on this, the correction being
+generated by the Neumann-Ulam solve must be incorrect.
 
@@
 
+\subsubsection{Model Diffusion Problem}
+\label{subsec:model_problem}
+For our breakdown study, we choose the one-speed, two-dimensional
+neutron diffusion equation as a model problem
+\citep{duderstadt_nuclear_1976}:
+\begin{equation}
+  -\boldsymbol{\nabla} \cdot D \boldsymbol{\nabla} \phi + \Sigma_a
+  \phi = S\:,
+  \label{eq:diffusion_eq}
+\end{equation}
+where $\phi$ is the neutron flux, $\Sigma_a$ is the absorption cross
+subsection, and $S$ is the source of neutrons. In addition, $D$ is the
+diffusion coefficient defined as:
+\begin{equation}
+  D = \frac{1}{3 ( \Sigma_t - \bar{\mu}\Sigma_s )}\:,
+  \label{eq:diffusion_coeff}
+\end{equation}
+where $\Sigma_s$ is the scattering cross subsection, $\Sigma_t = \Sigma_a
++ \Sigma_s$ is the total cross subsection, and $\bar{\mu}$ is the cosine
+of the average scattering angle. For simplicity, we will take
+$\bar{\mu} = 0$ for our analysis giving $D=(3 \Sigma_t)^{-1}$. In
+addition, to further simplify we will assume a homogeneous domain such
+that the cross subsections remain constant throughout. Doing this permits
+us to rewrite Eq~(\ref{eq:diffusion_eq}) as:
+\begin{equation}
+  -D \boldsymbol{\nabla}^2 \phi + \Sigma_a \phi = S\:.
+  \label{eq:diffusion_eq_simple}
+\end{equation}
+
+We choose a finite difference scheme on a square Cartesian grid to
+discretize the problem. For the Laplacian, we choose the 9-point
+stencil shown in Figure~\ref{fig:stencil} over a grid of size $h$
+\citep{leveque_finite_2007}:
+\begin{multline}
+  \nabla^2_9\phi = \frac{1}{6h^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j}
+    + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ +
+    \phi_{i-1,j+1} + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20
+    \phi_{i,j}]\:.
+  \label{eq:nine_point_stencil}
+\end{multline}
+\begin{figure}[t!]
+  \begin{center}
+    \scalebox{1.25}{\input{chapters/spn_equations/stencil.pdftex_t}}
+  \end{center}
+  \caption{\textbf{Nine-point Laplacian stencil.}}
+  \label{fig:stencil}
+\end{figure}
+We then have the following linear system to solve:
+\begin{multline}
+  -\frac{1}{6h^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j} + 4
+    \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ + \phi_{i-1,j+1}
+    + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20 \phi_{i,j}] + \Sigma_a
+  \phi_{i,j} = s_{i,j}\:,
+  \label{eq:fd_system}
+\end{multline}
+and in operator form:
+\begin{equation}
+  \ve{D}\boldsymbol{\phi}=\ve{s}\:,
+  \label{eq:operator_system}
+\end{equation}
+where $\ve{D}$ is the diffusion operator, $\ve{s}$ is the source in
+vector form and $\boldsymbol{\phi}$ is the vector of unknown fluxes.
+
+To close the system, a set of boundary conditions is required. In the
+case of a non-reentrant current condition applied to all global
+boundaries of the domain, we choose the formulation of Duderstadt by
+assuming the flux is zero at some ghost point beyond the
+grid. Consider for example the equations on the $i=0$ boundary of the
+domain:
+\begin{multline}
+  -\frac{1}{6h^2}[4 \phi_{-1,j} + 4 \phi_{1,j} + 4 \phi_{0,j-1} +
+    4 \phi_{0,j+1} + \phi_{-1,j-1}\\ + \phi_{-1,j+1} + \phi_{1,j-1} +
+    \phi_{1,j+1} - 20 \phi_{0,j}] + \Sigma_a \phi_{0,j} = s_{0,j}\:.
+  \label{eq:x_min_bnd}
+\end{multline}
+Here we note some terms where $i=-1$ and therefore are representative
+of grid points beyond the boundary of the domain. We set the flux at
+these points to be zero, giving a valid set of equations for the $i=0$
+boundary:
+\begin{multline}
+  -\frac{1}{6h^2}[4 \phi_{1,j} + 4 \phi_{0,j-1} + 4 \phi_{0,j+1}
+    \\ + \phi_{-1,j+1} + \phi_{1,j-1} + \phi_{1,j+1} - 20 \phi_{0,j}]
+  + \Sigma_a \phi_{0,j} = s_{0,j}\:.
+  \label{eq:x_min_bnd_2}
+\end{multline}
+We repeat this procedure for the other boundaries of the domain. For
+reflecting boundary conditions, the net current across a boundary is
+zero.
+
+\subsubsection{MCSA Breakdown Analysis}
+\label{subsubsec:breakdown_analysis}
+
 \subsection{Advanced Preconditioning Strategies}

### 2013-05-14 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
adding mcsa breakdown data spreadsheet

@@
 decomposed Monte Carlo calculation and assess the quality of our
-analytic relation for the optical thickness of domain and the
+analytic relation for the optical thickness of a domain and the
 associated leakage approximations. For this experiment, a square grid
@@
 boundary grid points. For the cross subsections, the absorption cross
-subsection was varied from 1 to 100 while the scattering cross subsection
-was set to zero to create a purely absorbing environment with weight
-cutoff of \sn{1}{-4}. The optical thickness of these domains will vary
-as a function of the absorption cross subsection if the other parameters
-are fixed. To compute the optical thickness, along with the spectral
-radius as given by Eq~(\ref{eq:iteration_radius}), we also need the
-parameters $n_i$ and $n_s$ which respectively describe the typical
-domain length and the average number of states moved along that
-typical length per history transition. For our grid above, the domains
-are varied in size with $50 \times 50$, $100 \times 100$, and $200
-\times 200$ cells giving $n_i=50$, $n_i=100$, and $n_i=200$ grid
-points or states along the typical length of the domain
-respectively. Looking at the Laplacian stencil in
-Eq~(\ref{eq:nine_point_stencil}), we see that all history transitions
-will only move a single state in either the $i$ or $j$ directions due
-to the symmetry of the problem. Furthermore, if we choose the $i$
-direction, not all states we will transition to will move the history
-in that direction. Therefore, we look to the definition of the
-iteration matrix in Eq~(\ref{eq:iteration_stencil}) and the definition
-of the adjoint probability matrix in Eq~(\ref{eq:adjoint_probability})
-to estimate the $n_s$ parameter. For a particular transition starting
-at state $(i,j)$, 6 of the 8 possible new states in the stencil move
-the history in $i$ direction with relative coefficients of 4 for
-moving in the $(\pm i,0)$ direction and of 1 for moving in the $(\pm
-i,\pm j)$. These coefficients dictate the frequency those states are
-visited relative to the others. For those 6 states we can visit along
-the typical length, their sum is 12 out of the total 20 for the
-coefficients for all possible states with their ratio giving $n_s =
-\frac{3}{5}$.
+subsection was varied from 1 to 100 while the scattering cross
+subsection was set to zero to create a purely absorbing environment
+with weight cutoff of \sn{1}{-4}. The optical thickness of these
+domains will vary as a function of the absorption cross subsection if
+the other parameters are fixed. To compute the optical thickness,
+along with the spectral radius as given by
+Eq~(\ref{eq:iteration_radius}), we also need the parameters $n_i$ and
+$n_s$ which respectively describe the typical domain length and the
+average number of states moved along that typical length per history
+transition. For our grid above, the domains are varied in size with
+$50 \times 50$, $100 \times 100$, and $200 \times 200$ cells giving
+$n_i=50$, $n_i=100$, and $n_i=200$ grid points or states along the
+typical length of the domain respectively. Looking at the Laplacian
+stencil in Eq~(\ref{eq:nine_point_stencil}), we see that all history
+transitions will only move a single state in either the $i$ or $j$
+directions due to the symmetry of the problem. Furthermore, if we
+choose the $i$ direction, not all states we will transition to will
+move the history in that direction. Therefore, we look to the
+definition of the iteration matrix in Eq~(\ref{eq:iteration_stencil})
+and the definition of the adjoint probability matrix in
+Eq~(\ref{eq:adjoint_probability}) to estimate the $n_s$ parameter. For
+a particular transition starting at state $(i,j)$, 6 of the 8 possible
+new states in the stencil move the history in $i$ direction with
+relative coefficients of 4 for moving in the $(\pm i,0)$ direction and
+of 1 for moving in the $(\pm i,\pm j)$. These coefficients dictate the
+frequency those states are visited relative to the others. For those 6
+states we can visit along the typical length, their sum is 12 out of
+the total 20 for the coefficients for all possible states with their
+ratio giving $n_s = \frac{3}{5}$.
 

### 2013-05-14 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
refactoring spn chapter sections
adding mcsa breakdown data spreadsheet

@@
 eigenvalues signal a need for a better preconditioning strategy to
-both ensure and improve covergence for Monte Carlo methods.
+both ensure and improve convergence for Monte Carlo methods.
 
@@
 diagonal as shown on the left side of
-Figure~\ref{fig:block_jacobi_ex}. For the $SP_N$ equations,
-Eq~(\ref{eq:A_block_matrix}) gives a block size of $N_g\times(N+1)/2$.
+Figure~\ref{fig:block_jacobi_ex}. Shown on the right side of
+Figure~\ref{fig:block_jacobi_ex}, inversion of the preconditioner is
+trivial with each diagonal block inverted separately. For the $SP_N$
+equations, Eq~(\ref{eq:A_block_matrix}) gives a block size of
+$N_g\times(N+1)/2$.
 \begin{figure}[t!]
@@
 representative of physical systems.
+\clearpage
 
@@
 the materials configuration along the horizontal axis is given by
-Figure~\ref{fig:problem3_axial_mat}.
+Figure~\ref{fig:problem3_axial_mat}. On the top and bottom of the
+assembly, vacuum conditions are used as well as on the top and right
+boundaries in Figure~\ref{fig:problem3_radial_mat}. Reflecting
+conditions are used on the left and bottom boundaries of
+Figure~\ref{fig:problem3_radial_mat}, effectively giving a
+representation of one quarter of the assembly.
 \begin{figure}[t!]
@@
       boundaries to create a complete $17 \times 17$ assembly
-      geometry. Each fuel pin is resolved by a $2 \times 2$ mesh.}}
+      geometry. Gray regions are homogenized fuel and red regions are
+      homogenzied moderator. Each fuel pin is resolved by a $2 \times
+      2$ mesh.}}
   \label{fig:problem3_radial_mat}
@@
 spacer grids, fuel pins with homogenized cladding and gas gap, core
-plenum, and moderator with boron. Table~\ref{tab:problem3_parameters}
+plenum, and moderator with boron. Group cross sections and other
+discrete nuclear data are generated as needed by a cross-section
+processing module dependent on the meshing parameters used to
+discretize the geometry and single-dimension pin-cell calculations for
+intial flux spectrum generation. Table~\ref{tab:problem3_parameters}
 gives the primary design parameters for the fuel assembly
-calculations. Group cross sections and other discrete nuclear data are
-generated as needed by a cross-section processing module dependent on
-the meshing parameters used to discretize the geometry.
+calculations.
 \begin{table}[h!]
@@
 linear operator applied to $\mathbf{\Phi}^{n+1}$ at each eigenvalue
-iteration will dictate converge and remain unchanged throughout the
-computation while the addition of fission to the system will only
+iteration will dictate convergence and remain unchanged throughout the
+computation\footnote{The operator will change if, for example, physics
+  feedback through temperature or potentially burnup is
+  considered. These additional physics will modify the cross sections
+  used to assemble $\mathbf{M}$. The calculations presented here will
+  consist of a single eigenvalue calculation with a static
+  $\mathbf{M}$.} while the addition of fission to the system will only
 modify the source of neutrons and the multiplication factor while not
@@
 
-\subsection{Jacobi Preconditioned Calculations}
+\subsection{Preliminary Jacobi Preconditioned Calculations}
 \label{subsec:jacobi_prec_assembly_calc}
@@
 \end{figure}
-Poor converge is observed for the Richardson iteration. However,
+Poor converge is observed for the Richardson iteration, however,
 convergence is achieved meaning that the criteria needed to satisfy
@@
 method are eventually being terminated by the weight cutoff as no
-artificial absorption was used. Based on this, the correction being
-generated by the Neumann-Ulam solve must be incorrect.
+artificial absorption was used for these preliminary
+calculations. Based on this, the correction being generated by the
+Neumann-Ulam solve is the component of MCSA causing a divergent
+solution.
 
@@
 \label{subsec:mcsa_break_down}
-For the fuel assembly problem, the initial block Jacobi calculations
-used a one-speed $SP_1$ discretization, effectively giving a diffusion
+For the fuel assembly problem, the initial block Jacobi
+preconditioned\footnote{For 1-group $SP_1$ problems, the block size is
+  1, giving a point Jacobi preconditioner.}  calculations used a
+one-speed $SP_1$ discretization, effectively giving a diffusion
 system. To study the breakdown of MCSA at iteration matrix spectral
@@
 one-speed neutron diffusion system to isolate this behavior. In this
-system, we can vary the cross sections in order to achieve varying
-spectral radii.
+system, we can vary the cross sections while maintaining a fixed grid
+in order to achieve varying spectral radii. For these studies, we
+neglect fission as MCSA behavior is dictated by the transport operator
+$\mathbf{M}$ in an eigenvalue scheme with the fission matrix used to
+generate a fixed source.
 
@@
 
-\subsection{Advanced Preconditioning Strategies}
+%%---------------------------------------------------------------------------%%
+\section{Advanced Preconditioning Strategies}
 \label{subsec:spn_advanced_preconditioning}
 
-\subsubsection{ILUT Preconditioning}
-\label{subsubsec:spn_ilut_preconditioning}
-
-\subsubsection{Sparse Approximate Inverse Preconditioning}
-\label{subsubsec:spn_spainv_preconditioning}
-
-\subsubsection{SPAINV Improved ILUT Preconditioning}
-\label{subsubsec:spn_spainv_preconditioning}
-
-\subsubsection{MCSA Relaxation Parameters}
-\label{subsubsec:spn_mcsa_relaxation}
-
-\subsubsection{Monte Carlo Estimator Comparison}
-\label{subsubsec:spn_estimator_comparison}
-
-\subsection{MCSA Comparison to Conventional Methods}
-\label{subsec:spn_comparison}
+\subsection{ILUT Preconditioning}
+\label{subsec:spn_ilut_preconditioning}
+
+\subsection{Sparse Approximate Inverse Preconditioning}
+\label{subsec:spn_spainv_preconditioning}
+
+\subsection{SPAINV Improved ILUT Preconditioning}
+\label{subsec:spn_spainv_preconditioning}
+
+\subsection{MCSA Relaxation Parameters}
+\label{subsec:spn_mcsa_relaxation}
+
+\subsection{Monte Carlo Estimator Comparison}
+\label{subsec:spn_estimator_comparison}
+
+%%---------------------------------------------------------------------------%%
+\section{MCSA Comparison to Conventional Methods}
+\label{sec:spn_comparison}
+
+%%---------------------------------------------------------------------------%%
+\section{Multiple Fuel Assembly Challenge Problem}
+\label{sec:spn_challenge_problem}

### 2013-05-14 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
starting work on breakdown analysis and adding sections to the concluding chapter

@@
 infinity norm as a function of iteration for the MCSA linear solve in
-the first eigenvalue iteration.
+the first eigenvalue iteration using 25,000 stochastic histories at
+every iteration for the adjoint Neumann-Ulam solve.
 \begin{figure}[t!]
@@
 \label{subsubsec:breakdown_analysis}
+For each solver and estimator combination, the spectral radius of the
+iteration matrix was varied by changing the absorption cross section
+while fixing the grid size and scattering cross section. At each
+solve, a minimum of stochastic history per degree of freedom (DOF) in
+the problem was used to compute the Monte Carlo correction. If the
+solver could not converge in less than 100 iterations, the number of
+histories was increased by increments of 5,000 until convergence was
+achieved.
 
@@
 \label{sec:spn_comparison}
-
-%%---------------------------------------------------------------------------%%
-\section{Multiple Fuel Assembly Challenge Problem}
-\label{sec:spn_challenge_problem}

### 2013-05-15 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding figures to breakdown data

@@
 convergence is achieved meaning that the criteria needed to satisfy
-MCSA convergence has also been met. The effective slope of the
-convergence curve in Figure~\ref{fig:block_jacobi_res_richardson} will
-give an approximation for the spectral radius via
-Eq~(\ref{eq:linear_k_iter_norm3}). For the data in
-Figure~\ref{fig:block_jacobi_res_richardson}, $\rho(\mathbf{H})
-\approx 0.985$, nearing the limits of MCSA applicability and well
-beyond the spectral radii generated in the initial spectral analysis.
+MCSA convergence has also been met. The number of iterations required
+for the Richardson iteratio to converge will give an approximation for
+the spectral radius via Eq~(\ref{eq:linear_k_iter_norm3}). With 7291
+iterations require to converge to a tolerance of $\sn{1}{-6}$,
+$\rho(\mathbf{H}) \approx 0.998$, nearing the limits of MCSA
+applicability and well beyond the spectral radii generated in the
+initial spectral analysis.
 
@@
 while fixing the grid size and scattering cross section. At each
-solve, a minimum of stochastic history per degree of freedom (DOF) in
-the problem was used to compute the Monte Carlo correction. If the
+solve, a minimum of one stochastic history per degree of freedom (DOF)
+in the problem was used to compute the Monte Carlo correction. If the
 solver could not converge in less than 100 iterations, the number of

### 2013-05-15 doc/jcomp2012/jcomp2012.tex
Added more results for marshak and dog-leg problem.

@@
 \begin{equation}
-  \|\vr\|_{\infty} < \epsilon\cdot\|\vb\|_\infty\:.
+  \|\vr\|_{2} < \epsilon\cdot\|\vb\|_2\:.
   \label{eq:stopping-criteria}
@@
 Here we examine the effectiveness of the MCSA method and compare it to
-traditional deterministic solution methods.  The deterministic methods
-considered are preconditioned Conjugate Gradient (PCG) and
-preconditioned fixed-point (Richardson) iteration (PFIX).  The PCG
-method uses right-left preconditioning with $\ve{M}=\ve{L}\ve{L}^{T}$,
-where $\ve{L} = \ve{L}^{T} = \sqrt{\diag(\ve{D})}$.  Naturally, more
-powerful preconditioning methods could be used; however, this provides
-a fair comparison with the preconditioning used in the MCSA method.
-Different preconditioning strategies within MCSA will be the focus of
-a future paper.  The PFIX scheme is used to solve
-Eq.~(\ref{eq:preconditioned-diffusion}), which is identical to the
-system being solved with MCSA.  We examine each method on two 3D
-problems: a Marshak wave problem and a multi-material duct problem.
+traditional deterministic solution methods.  We implemented Krylov subspace
+methods to solve the radiation diffusion equation using the Trilinos solver
+library \cite{1089021}.  Conjugate Gradient (CG) is that standard method for
+symmetric positive-definite (SPD) systems like that defined in
+Eq.~(\ref{eq:discrete-diffusion}).  However, for completeness we include some
+comparisons using restarted GMRES($m$) so that asymmetric preconditioners can
+be tested. The stopping criterion in Eq.~(\ref{eq:stopping-criteria}) is used
+for all calculations. All methods are tested on two 3D problems: a
+uniform-medium Marshak wave problem and a multi-material duct problem.
 
@@
 to the MCSA solution.  As opposed to most Monte Carlo techniques, the MCSA
-method generates solutions that are numerically equivalent to the PCG and PFIX
-results.  Because the MCSA method converges the residual of the system, the
-numerical accuracy is determined by the stopping criterion shown in
-Eq.~(\ref{eq:stopping-criteria}).  Also, the MCSA method is not bound by the
-Central Limit Theorem so convergence is not restricted to
+method generates solutions that are numerically equivalent to the
+deterministic results.  Because the MCSA method converges the residual of the
+system, the numerical accuracy is determined by the stopping criterion shown
+in Eq.~(\ref{eq:stopping-criteria}).  Also, the MCSA method is not bound by
+the Central Limit Theorem so convergence is not restricted to
 $1/\sqrt{N}$~\cite{halton_1994,evans_2003}.
@@
 algorithms.  Table~\ref{tab:marshak_comparison} shows timing results for the
-Marshak problem using preconditioned CG (PCG), preconditioned Richardson
-iteration (PFIX), and MCSA. The results show that the MCSA method is the most
-efficient method, although the PCG results are nearly as fast.  More
-importantly, the MCSA method clearly accelerates Richardson iteration, as
-indicated by the fact that MCSA is 42\% faster than Richardson iteration
-alone.
+Marshak problem using CG with Jacobi and algebraic multigrid preconditioning
+and GMRES with incomplete LU (ILU) and ILU with threshold cutoff (ILUT).  The
+multigrid preconditioner is the smoothed aggregation algebraic multigrid
+method implemented in the Trilinos ML package \cite{Gee:2007uj}.  In all the
+problems that follow we use a $4/3$ damping factor in the aggregation and a
+symmetric Gauss-Seidel smoother with 2 sweeps per level.
+
+Analyzing Table~\ref{tab:marshak_comparison}, we see that the overall
+iterations are more than a factor of 2 less for the Krylov solvers compared to
+MCSA; however, the cost per iteration is signicantly more expensive.  For any
+given problem, the cost per iteration is roughly linear with particle count as
+illustrated in Fig.~\ref{fig:CPU_Np}.  Because this problem converges running
+only a maximum of 10 particles per iteration, 156,464 total particles over
+10,157 cycles resulting in an average of $\sim 1.6$ particles per iteration,
+the cost per iteration is exceptionally small.  Running more particles per
+iteration would reduce the number of iterations, but a greater cost per
+iteration.
 \begin{table}[ht!]
@@
     methods were run with a stopping criterion of $\sn{1}{-8}$.  The
-    MCSA method used $N_p=10$ and $W_c = \sn{1}{-4}$.  The CPU time is
-    measured relative to the MCSA method, which required 72.6 s on a
-    3.6~GHz Xeon.}
+    MCSA method used $N_p=10$ and $W_c = \sn{1}{-4}$.   Runtimes are relative
+    to the MCSA performance that required 57.72 seconds on
+    an 2.2 MHz AMD Opteron.}
   \label{tab:marshak_comparison}
   \begin{center}
-    \begin{tabular}{lrr}\hline\hline
-      \multicolumn{1}{c}{Method} & \multicolumn{1}{c}{Max Iterations}
-      & \multicolumn{1}{c}{Relative CPU Time}\\\hline\hline  PCG &
-      11 & 1.11 \\  PFIX & 11 & 1.42 \\  MCSA & 11 & 1.0
+    \small
+    \begin{tabular}{llrrrr}\hline\hline
+      \multicolumn{1}{c}{Method} &
+      \multicolumn{1}{c}{Preconditioner} &
+      \multicolumn{1}{c}{Total} &
+      \multicolumn{1}{c}{Average} &
+      \multicolumn{1}{c}{Relative Time} &
+      \multicolumn{1}{c}{Relative}\\
+      & & \multicolumn{1}{c}{Iterations} &
+      \multicolumn{1}{c}{Iterations} &
+      \multicolumn{1}{c}{per Iteration} &
+      \multicolumn{1}{c}{Runtime} \\\hline\hline
+      %%
+      CG & Jacobi & 46,746 & 4.60 & 25.91 & 3.29 \\
+      CG & ML & 12,150 & 1.20 & 74.72 & 2.69 \\
+      GMRES & ILU & 40,294 & 3.97 & 103.02 & 9.19 \\
+      GMRES & ILUT$^{\dagger}$ & 40,447 & 3.98 & 101.31 & 9.09 \\
+      MCSA & Jacobi & 98,863 & 9.73 & 1.0 & 1.0
+      %%
       \\ \hline\hline
+      \multicolumn{6}{l}{
+        $^{\dagger}$ The threshold cutoff was set to 0.01, and the fill was 1.}
     \end{tabular}
@@
 Table~\ref{tab:multimat_comparison} shows the timing results for the
-multi-material problem using PCG, PFIX, and MCSA.  The results correspond
-roughly with the results from the Marshak problem.  The MCSA is marginally
-faster than PCG and 43\% faster than PFIX.
+multi-material problem using CG with Jacobi and ML preconditioning versus MCSA
+run out to a time of 100~ns.  The MCSA method was set to use 1000 particles
+per iteration resulting in 74,237,525 total particles over 10,157 cycles or
+823.68 particles per iteration.  Because more particles are required to get a
+reasonable estimate of the Neumann series, the relative efficiency per
+iteration of MCSA versus CG is significanly reduced compared to the smaller
+Marshak wave problem.  Nonetheless, MCSA is still marginally more efficient
+than CG with Jacobi preconditioning and significantly more efficient than CG
+with algegraic multigrid.
 \begin{table}[htbp!]
@@
     \sn{1}{-3}$.  The problem was run to an elapsed time of 100~ns.
-    The MCSA method required 3.4 hours on a 3.6~GHz Xeon.}
+    Runtimes are relative to the MCSA performance that required 9,862.4
+    seconds on a 2.2 MHz AMD Opteron.}
   \label{tab:multimat_comparison}
   \begin{center}
-    \begin{tabular}{lrr}\hline\hline
-      \multicolumn{1}{c}{Method} & \multicolumn{1}{c}{Max Iterations}
-      & \multicolumn{1}{c}{Relative CPU Time}\\\hline\hline  PCG &
-      18 & 1.03 \\  PFIX & 38 & 1.43 \\  MCSA & 20 & 1.0
+    \small
+    \begin{tabular}{llrrrr}\hline\hline
+      \multicolumn{1}{c}{Method} &
+      \multicolumn{1}{c}{Preconditioner} &
+      \multicolumn{1}{c}{Total} &
+      \multicolumn{1}{c}{Average} &
+      \multicolumn{1}{c}{Relative Time} &
+      \multicolumn{1}{c}{Relative}\\
+      & & \multicolumn{1}{c}{Iterations} &
+      \multicolumn{1}{c}{Iterations} &
+      \multicolumn{1}{c}{per Iteration} &
+      \multicolumn{1}{c}{Runtime} \\\hline\hline
+      %%
+      CG & Jacobi & 42,881 & 4.22 & 2.93 & 1.02 \\
+      CG & ML & 14,097 & 1.39 & 26.60 & 1.94 \\
+      MCSA & Jacobi & 90,129 & 8.87 & 1.0 & 1.0
+      %%
       \\ \hline\hline

### 2013-05-16 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding breakdown figures

@@
 the materials configuration along the horizontal axis is given by
-Figure~\ref{fig:problem3_axial_mat}. On the top and bottom of the
-assembly, vacuum conditions are used as well as on the top and right
-boundaries in Figure~\ref{fig:problem3_radial_mat}. Reflecting
-conditions are used on the left and bottom boundaries of
+Figure~\ref{fig:problem3_axial_mat}. A detailed view of the assembly
+bottom is given in Figure~\ref{fig:problem3_end}. On the top and
+bottom of the assembly, vacuum conditions are used as well as on the
+top and right boundaries in
+Figure~\ref{fig:problem3_radial_mat}. Reflecting conditions are used
+on the left and bottom boundaries of
 Figure~\ref{fig:problem3_radial_mat}, effectively giving a
@@
   \begin{center}
-    \includegraphics[width=6.25in]{chapters/spn_equations/problem3_axial_mat.png}
+    \includegraphics[width=6.0in]{chapters/spn_equations/problem3_axial_mat.png}
   \end{center}
@@
       spacer grids along the length of the fuel pins and reactor core
-      structures on the top and bottom of the assembly.}}
+      structures on the top and bottom of the assembly. Lighter purple
+      material in the center of the assembly is moderator and darker
+      purple material fuel.}}
   \label{fig:problem3_axial_mat}
+\end{figure}
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6.0in]{chapters/spn_equations/problem3_end.png}
+  \end{center}
+  \caption{\textbf{Fuel assembly geometry end detail.}
+    \textit{Reactor core structure including spacer grids and plenum
+      has been included. Lighter purple material on the right of the
+      figure is moderator and darker purple material fuel.}}
+  \label{fig:problem3_end}
 \end{figure}
@@
 computations performed, these results are not in line with
-expectations for this problem. Even if the problem may be
-ill-conditioned, we do expect convergence of MCSA. To investigate
-further, a block Jacobi preconditioned Richardson iteration was used
-to solve the same problem. Figure~\ref{fig:block_jacobi_res_richardson}
-gives the residual infinity norm as a function of iteration for the
-Richardson linear solve in the first eigenvalue iteration.
+expectations for this problem. Additional computations performed with
+\sn{1}{6} histories per iteration exhibited the same divergent
+behavior at a significant computational cost and compute time. Even if
+the problem may be ill-conditioned, we do expect convergence of
+MCSA. To investigate further, a block Jacobi preconditioned Richardson
+iteration was used to solve the same
+problem. Figure~\ref{fig:block_jacobi_res_richardson} gives the
+residual infinity norm as a function of iteration for the Richardson
+linear solve in the first eigenvalue iteration.
 \begin{figure}[t!]
@@
 Poor converge is observed for the Richardson iteration, however,
-convergence is achieved meaning that the criteria needed to satisfy
-MCSA convergence has also been met. The number of iterations required
-for the Richardson iteratio to converge will give an approximation for
-the spectral radius via Eq~(\ref{eq:linear_k_iter_norm3}). With 7291
-iterations require to converge to a tolerance of $\sn{1}{-6}$,
-$\rho(\mathbf{H}) \approx 0.998$, nearing the limits of MCSA
-applicability and well beyond the spectral radii generated in the
-initial spectral analysis.
+convergence is achieved meaning that the preliminary criteria needed
+to satisfy MCSA convergence has also been met. The number of
+iterations required for the Richardson iteration to converge will give
+an approximation for the spectral radius via
+Eq~(\ref{eq:linear_k_iter_norm3}). With 7291 iterations required to
+converge to a tolerance of $\sn{1}{-6}$, $\rho(\mathbf{H}) \approx
+0.998$, nearing the limits of MCSA applicability and well beyond the
+spectral radii generated in the initial spectral analysis.
 
@@
 histories was increased by increments of 5,000 until convergence was
-achieved.
+achieved in less than 100 iterations. The number of iterations
+required to converge and the time to converge was recorded.
 

### 2013-05-16 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding breakdown analysis to spn chapter

@@
 achieved in less than 100 iterations. The number of iterations
-required to converge and the time to converge was recorded.
+required to converge MCSA and the time to converge was recorded as a
+means to capture the breakdown.
+
+Figure~\ref{fig:breakdown_iterations} gives the number of iterations
+required to converge for the chosen number of histories per iteration
+given by Figure~\ref{fig:breakdown_histories} using the adjoint solver
+with the collision and expected value estimators and the forward
+solver with the collision estimator. For spectral radii less than
+0.97, all MCSA problems converged with 1 history per DOF (10,000 for
+this problem) with the number of iterations required to converge
+increasing. Near a spectral radius of 0.97, the number of histories
+required to converge MCSA in less than 100 iterations takes a dramatic
+rise that exhibits neither exponential nor power law behavior. As the
+spectral radius approaches 1, the number of histories required becomes
+significant and effectively intractable. Even with this simple
+diffusion problem, the behavior is consistent with that oberseved for
+the fuel assembly problem with $SP_1$ discretization. In that case, we
+estimated a spectral radius of $\approx 0.998$, larger than any of the
+spectral radii that could be computed within even 90 minutes of
+compute time for this simple two dimensional problem. For that
+problem, even 1,000,000 histories ($\approx 50$ per DOF) was not
+enough to provide convergence.
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=5in]{chapters/spn_equations/breakdown_iterations.pdf}
+  \end{center}
+  \caption{\textbf{Iterations required to converge as a function of
+      spectral radius for the neutron diffusion problem.} \textit{The
+      number of histories was increased to achieve convergence in less
+      than 100 iterations. At least 10,000 histories were used for
+      each calculation.}}
+  \label{fig:breakdown_iterations}
+\end{figure}
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=5in]{chapters/spn_equations/breakdown_histories.pdf}
+  \end{center}
+  \caption{\textbf{Histories required to converge as a function of
+      spectral radius for the neutron diffusion problem.} \textit{The
+      number of histories was increased to achieve convergence in less
+      than 100 iterations. At least 10,000 histories were used for
+      each calculation.}}
+  \label{fig:breakdown_histories}
+\end{figure}
+In addition to the significantly larger number of histories required
+for to achieve convergence for ill-conditioned problems another
+penalty is paid due to histories that take longer to
+compute. Figure~\ref{fig:breakdown_time} gives the CPU time in
+seconds required to compute a single random walk averaged over the
+entire set of histories run in the calculation over all iterations. As
+the spectral radius increases (correlating to a higher ratio of
+scattering in the system) the random walk lengths increase, using more
+CPU time to finish the computation. Compared to spectral radii of 0.5,
+larger spectral radii over 0.97 have histories that require two orders
+of magnitude more computation time. This signficant increase in
+computation time per history coupled with the significant increase in
+the number of histories required to converge is evidence that for
+problems with spectral radii above $\approx 0.97$, using MCSA to solve
+any problems of interest is completely intractable. We therefore
+require a more expansive set of preconditioning techniques to reduce
+the eigenvalue spectrum of the $SP_N$ problem into a regime in which
+MCSA is more applicable.
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=5in]{chapters/spn_equations/breakdown_time.pdf}
+  \end{center}
+  \caption{\textbf{CPU time per history as a function of spectral
+      radius for the neutron diffusion problem.} \textit{As the
+      spectral radius grows, so do the length of the random
+      walks. Longer random walks require more CPU time to compute.}}
+  \label{fig:breakdown_time}
+\end{figure}
+
+\clearpage
 

### 2013-05-16 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
finished breakdown section

@@
 For each solver and estimator combination, the spectral radius of the
-iteration matrix was varied by changing the absorption cross section
-while fixing the grid size and scattering cross section. At each
-solve, a minimum of one stochastic history per degree of freedom (DOF)
-in the problem was used to compute the Monte Carlo correction. If the
-solver could not converge in less than 100 iterations, the number of
-histories was increased by increments of 5,000 until convergence was
-achieved in less than 100 iterations. The number of iterations
-required to converge MCSA and the time to converge was recorded as a
-means to capture the breakdown.
+iteration matrix generated by the diffusion problem was varied by
+changing the absorption cross section while fixing the grid size and
+scattering cross section. For each parameter variation, a minimum of
+one stochastic history per degree of freedom (DOF) in the problem was
+used to compute the Monte Carlo correction. If the solver could not
+converge in less than 100 iterations, the number of histories was
+increased by increments of 5,000 until convergence was achieved in
+less than 100 iterations. The number of iterations required to
+converge MCSA and the time to converge was recorded as a means to
+capture the breakdown.
 
@@
 this problem) with the number of iterations required to converge
-increasing. Near a spectral radius of 0.97, the number of histories
-required to converge MCSA in less than 100 iterations takes a dramatic
-rise that exhibits neither exponential nor power law behavior. As the
-spectral radius approaches 1, the number of histories required becomes
-significant and effectively intractable. Even with this simple
-diffusion problem, the behavior is consistent with that oberseved for
-the fuel assembly problem with $SP_1$ discretization. In that case, we
-estimated a spectral radius of $\approx 0.998$, larger than any of the
-spectral radii that could be computed within even 90 minutes of
-compute time for this simple two dimensional problem. For that
-problem, even 1,000,000 histories ($\approx 50$ per DOF) was not
-enough to provide convergence.
+increasing a function of spectral radius. Near a spectral radius of
+0.97, the number of histories required to converge MCSA in less than
+100 iterations takes a dramatic rise that exhibits neither exponential
+nor power law behavior. As the spectral radius approaches 1, the
+number of histories required becomes significant and effectively
+intractable to compute. Even with this simple diffusion problem, the
+behavior is consistent with that observed for the fuel assembly
+problem with $SP_1$ discretization. In that case, we estimated a
+spectral radius of $\approx 0.998$, larger than any of the spectral
+radii that could be computed within even 90 minutes of compute time
+for this simple two dimensional problem. For that problem, even
+1,000,000 histories ($\approx 50$ per DOF) was not enough to provide
+convergence. Even if single solve times of an hour can be tolerated,
+dozens of solves are typically required to compute the multiplication
+factor and flux spectrum within the $SP_N$ eigenvalue scheme, making
+the method unusable.
 \begin{figure}[t!]
@@
 \end{figure}
+
 In addition to the significantly larger number of histories required
@@
 penalty is paid due to histories that take longer to
-compute. Figure~\ref{fig:breakdown_time} gives the CPU time in
-seconds required to compute a single random walk averaged over the
-entire set of histories run in the calculation over all iterations. As
-the spectral radius increases (correlating to a higher ratio of
-scattering in the system) the random walk lengths increase, using more
-CPU time to finish the computation. Compared to spectral radii of 0.5,
-larger spectral radii over 0.97 have histories that require two orders
-of magnitude more computation time. This signficant increase in
+compute. Figure~\ref{fig:breakdown_time} gives the CPU time in seconds
+required to compute a single random walk averaged over the entire set
+of histories run in the calculation over all iterations. As the
+spectral radius increases (correlating to a higher ratio of scattering
+in the system) the random walk lengths increase, using more CPU time
+to finish the computation. Compared to spectral radii of 0.5, larger
+spectral radii over 0.97 have histories that require two orders of
+magnitude more computation time. This signficant increase in
 computation time per history coupled with the significant increase in
@@
 problems with spectral radii above $\approx 0.97$, using MCSA to solve
-any problems of interest is completely intractable. We therefore
-require a more expansive set of preconditioning techniques to reduce
-the eigenvalue spectrum of the $SP_N$ problem into a regime in which
-MCSA is more applicable.
+any problems of interest is completely unusable. We therefore require
+a more expansive set of preconditioning techniques to move the
+eigenvalue spectrum of the $SP_N$ problem into a regime in which MCSA
+is more applicable.
 \begin{figure}[t!]

### 2013-05-16 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
starting advanced spn preconditioning section

@@
       material in the center of the assembly is moderator and darker
-      purple material fuel.}}
+      purple/red material is fuel.}}
   \label{fig:problem3_axial_mat}
@@
       has been included. Lighter purple material on the right of the
-      figure is moderator and darker purple material fuel.}}
+      figure is moderator and darker purple/red material is fuel.}}
   \label{fig:problem3_end}
@@
 \label{subsec:spn_advanced_preconditioning}
+For the fuel assembly criticality problem presented in the previous
+section, the spectral radius was near one and in a regime in which
+MCSA breakdown is observed. In this regime the number of stochastic
+histories required to converge MCSA increases rapidly and the
+resulting poor performance is compounded by those histories being
+expensive to compute. To overcome this difficulty, advanced
+preconditioning strategies are required beyond simple Jacobi methods
+that can reduce the spectral radius into a region of better MCSA
+performance. Several modern algebraic preconditioning strategies will
+be presented here and used within the preconditioned framework given
+by Eq~(\ref{eq:left_right_mcsa}). Data showing their effects on MCSA
+solutions of the fuel assembly criticality problem will be presented.
 
@@
 \label{subsec:spn_estimator_comparison}
+
+\clearpage
 

### 2013-05-16 doc/jcomp2012/jcomp2012.tex
Nearly final edits.

@@
 In previous work~\cite{evans_2003}, we developed a residual Monte Carlo method
-that was an application of Halton's Sequential Monte Carlo
+that was an application of Halton's sequential Monte Carlo
 method~\cite{halton_1962,halton_1994} to the one-dimensional (1D) equilibrium
@@
 widely known in the general computational physics community.  While extending
-our method to three spatial dimensions, we realized that the Sequential Monte
+our method to three spatial dimensions, we realized that the sequential Monte
 Carlo method is actually a variant of iterative refinement cast as a residual
@@
 In order to address the convergence issues plaguing Monte Carlo
-solvers, Halton \cite{halton_1962,halton_1994} proposed a Sequential
+solvers, Halton \cite{halton_1962,halton_1994} proposed a sequential
 Monte Carlo method.  This algorithm demonstrated dramatically improved
@@
 \end{equation}
-Each row of the transition probability matrix, $\vP$, represents a
-discrete probability density function that can be sampled to select a
-new state $j$, given that the current state is $i$.  Random walks can
-be terminated in two ways: the matrix can be augmented with a
-terminating event equation that describes the probability that a
-history ends its walk, or the random walk can be terminated by weight
-cutoff, $W_c$.  Generally, we choose to terminate random walks using a
-weight cutoff as opposed to augmenting the matrix such that the random
-walk terminates when $W_m < W_c$.
+Each row of the transition probability matrix, $\vP$, represents a discrete
+probability density function that can be sampled to select a new state $j$,
+given that the current state is $i$.  Random walks can be terminated in two
+ways: the matrix can be augmented with a terminating event equation that
+describes the probability that a history ends its walk, or the random walk can
+be terminated by weight cutoff, $W_c$.  Generally, we choose to terminate
+random walks using a weight cutoff such that the random walk terminates when
+$W_m < W_c$.
 
@@
 random walk currently resides.  This is the common approach in standard Monte
-Carlo transport simulations. Unlike the direct method, because tallies are
-made in the state in which the random walk currently resides, we are no longer
-required to start a history in each state for each random walk
-permutation. Instead, sampling the source is sufficient and requires only one
-random walk per permutation.
+Carlo transport simulations. Thus, unlike the direct method, we are no longer
+required to start a history in each state for each random walk permutation
+because tallies are made in the state in which the random walk currently
+resides. Instead, sampling the source is sufficient, and one random walk per
+permutation is required.
 
@@
 Halton~\cite{halton_1962,halton_1994} proposed a staged residual scheme called
-Sequential Monte Carlo to speed up the convergence of these methods.  A
+sequential Monte Carlo to speed up the convergence of these methods.  A
 variant of this scheme has been successfully applied to the 1D nonlinear
@@
 
-Concisely, the Sequential Monte Carlo method solves Eq.~(\ref{eq:Ax=b}) using
+Concisely, the sequential Monte Carlo method solves Eq.~(\ref{eq:Ax=b}) using
 the adjoint solution technique described in \S~\ref{sec:adjoint-method}.  The
@@
   \begin{gather}
-    \vr^{l} = \vb - \vA\vx^{l}\:, \\ \vA\delta\vx^{l+1} =
+    \vr^{l} = \vb - \vA\vx^{l}\:, \\ \hat{\vA}\delta\vx^{l+1} =
     \vr^{l}\:, \label{eq:residual_solve}\\ \vx^{l+1} = \vx^{l} +
@@
   \end{gather}
+  \label{eq:sequential_mc}
 \end{subequations}
 In this scheme, the Monte Carlo adjoint method is used to estimate the
-solution to Eq.~(\ref{eq:residual_solve}), and the residual is iterated to
-convergence.  This iteration sequence is closely related to iterative
-refinement, the exception being that there is no update to $x^{l+1}$ between
-residual iterations.
+solution to Eq.~(\ref{eq:residual_solve}) as represented by the hat on $\vA$,
+and the residual is iterated to convergence.  This iteration sequence is
+closely related to iterative refinement, the exception being that there is no
+update to $x^{l+1}$ between residual iterations.
 
@@
   \end{gather}
+  \label{eq:MCSA}
 \end{subequations}
-The hat on $\vA$ in Eq.~(\ref{eq:MCSA-residual_solve}) indicates that the
-Monte Carlo solution only approximately inverts this operator.  Thus, we have
-defined a scheme in which the initial estimate of $\vx$ in each iteration is
-updated using a single fixed-point iteration\footnote{A subspace method could
-  be used instead of a stationary method here.}.  The residual is calculated
-and used as a source to estimate the error, $\delta\vx^{l+1/2}$, by solving
+As in the sequential method, the hat on $\vA$ in
+Eq.~(\ref{eq:MCSA-residual_solve}) indicates that the Monte Carlo solution
+only approximately inverts this operator.  Thus, we have defined a scheme in
+which the initial estimate of $\vx$ in each iteration is updated using a
+single fixed-point iteration\footnote{A subspace method could be used instead
+  of a stationary method here.}.  The residual is calculated and used as a
+source to estimate the error, $\delta\vx^{l+1/2}$, by solving
 Eq.~(\ref{eq:MCSA-residual_solve}) via the Monte Carlo adjoint method. The
@@
 \end{equation}
+
+A closer examination of the sequential method in Eq.~(\ref{eq:sequential_mc})
+shows that this scheme is equivalent to a single preconditioned Richardson
+iteration,
+\begin{equation}
+  \vx^{l+1} = \vx^{l} + \hat{\vA}^{-1}\vr^l\:,
+\end{equation}
+in which $\vA^{-1}$ indicates the approximate inversion of $\vA$ via the
+adjoint Monte Carlo method.  MCSA, on the other hand, is identical to a
+Richardson iteration followed by a preconditioned Richardson iteration,
+\begin{align}
+  \vx^{l+1/2} &= \vx^{l} + \vr^{l}\:,\\
+  \vx^{l+1} &= \vx^{l+1/2} + \hat{\vA}^{-1}\vr^{l+1/2}\:.
+\end{align}
+As above, the adjoint Monte Carlo process is used to approximately invert
+$\vA$.
 
@@
 
-A timing and convergence study is used to demonstrate the
-effectiveness of the adjoint method as compared to the direct
-method. To assess both the CPU time and number of iterations required
-to converge to a solution, a problem of constant $\Delta$ was used
-with varying values of grid size, fixing the spectral radius of the
-system at a constant value for each variation. Both the five-point and
-nine-point stencils were used with both the direct and adjoint
-solvers. For each case, $N \times N$ total random walk permutations
-were computed per MCSA iteration where $N \times N$ is the number of
-discrete grid points in the system. Solver parameters were set to a
-weight cutoff of \sn{1}{-4} for the stochastic linear solver and a
-convergence tolerance of \sn{1}{-8} for the MCSA iterative solver.
-Figure~\ref{fig:poisson_cpu_time} gives the CPU time needed for each
-case to converge in seconds and Figure~\ref{fig:poisson_iterations}
-gives the number of iterations needed for each case to converge to the
-specified tolerance as a function of the problem size. All
-computations presented in this section and
-\S~\ref{subsec:sequential_comparison} were completed on a 3.0 GHz
-Intel Core 2 Quad Q9650 CPU machine with 16 GB 1067 MHz DDR3 memory.
+A timing and convergence study is used to demonstrate the effectiveness of the
+adjoint method as compared to the direct method. To assess both the CPU time
+and number of iterations required to converge to a solution, a problem of
+constant $\Delta$ was used with varying number of grid points, fixing the
+spectral radius of the system at a constant value for each variation. Both the
+five-point and nine-point stencils were used with both the direct and adjoint
+solvers. For each case, $N \times N$ total random walk permutations were
+computed per MCSA iteration where $N \times N$ is the number of discrete grid
+points in the system. Solver parameters were set to a weight cutoff of
+\sn{1}{-4} for the stochastic linear solver and a convergence tolerance of
+\sn{1}{-8} for the MCSA iterative solver.  Figure~\ref{fig:poisson_cpu_time}
+gives the CPU time needed for each case to converge in seconds and
+Figure~\ref{fig:poisson_iterations} gives the number of iterations needed for
+each case to converge to the specified tolerance as a function of the problem
+size. All computations presented in this section and
+\S~\ref{subsec:sequential_comparison} were completed on a 3.0 GHz Intel Core 2
+Quad Q9650 CPU machine with 16 GB 1067 MHz DDR3 memory.
 \begin{figure}[ht!]
@@
 
-As an additional comparison, the convergence behavior of MCSA can be
-analyzed using both the adjoint and direct solvers to detect any
-performance benefits. To assess the convergence properties of MCSA
-using each solver and stencil, the infinity norm of the residual
-computed in Eq.~(\ref{eq:MCSA-residual}) was collected at each
-iteration for a fixed problem size of
-$N=500$. Figure~\ref{fig:poisson_convergence} gives the results of
-these computations. First, it is worthy to note on the semilog plot
-that we are indeed achieving the expected exponential convergence from
-MCSA with both Monte Carlo solvers. Second, we note that using the
-adjoint method with the same number of stochastic histores per MCSA
-iteration gives a faster rate of converge for the same reasons as
-above. We also note here that fewer iterations are required for
-convergence when the 9-point stencil is used to discretize the
-Laplacian operator (although at no gain in speed as given by the
-results in Figure~\ref{fig:poisson_cpu_time}). This is due to the fact
-that the smaller discretization error directly corresponds to a more
-well defined residual source generated by the Richardson extrapolation
-for the Monte Carlo calculation. In addition, the better defined
-source is transported through a domain described more accurately by
-the 9-point stencil, thus yielding a more accuracte correction vector
-from the Monte Carlo calculation.
+As an additional comparison, the convergence behavior of MCSA can be analyzed
+using both the adjoint and direct solvers to detect any performance
+benefits. To assess the convergence properties of MCSA using each solver and
+stencil, the infinity norm of the residual computed in
+Eq.~(\ref{eq:MCSA-residual}) was collected at each iteration for a fixed
+problem size of $N=500$. Figure~\ref{fig:poisson_convergence} gives the
+results of these computations. First, it is worthy to note on the semilog plot
+that we are indeed achieving the expected exponential convergence from MCSA
+with both Monte Carlo solvers. Second, we note that using the adjoint method
+with the same number of stochastic histories per MCSA iteration gives a faster
+rate of converge for the same reasons as above. We also note here that fewer
+iterations are required for convergence when the 9-point stencil is used to
+discretize the Laplacian operator (although at no gain in speed as given by
+the results in Figure~\ref{fig:poisson_cpu_time}). This is due to the fact
+that the smaller discretization error directly corresponds to a more well
+defined residual source generated by the Richardson iteration for the Monte
+Carlo calculation. In addition, the better defined source is transported
+through a domain described more accurately by the 9-point stencil, thus
+yielding a more accuracte correction vector from the Monte Carlo calculation.
 \begin{figure}[ht!]
@@
 \label{subsec:sequential_comparison}
-To further motivate using Monte Carlo Synthetic Acceleration, we
-compare its performance to Halton's Sequential Monte Carlo method on
-which our previous work in this area was based. For this comparison,
-we use the same transient Poisson problem as described in the previous
-section and choose only the 5-point stencil to discretize the
-Laplacian operator as the previous results yielded little qualitative
-difference between the discretizations. Both MCSA and Halton's method
-are used with the adjoint Monte Carlo solver. In order to complete the
-same study as in the previous section, the number of histories
-computed by the Monte Carlo solver at each iteration had to be doubled
-to $2 \times N \times N$ in order to ensure convergence in Sequential
-Monte Carlo Method. Figure~\ref{fig:seq_cpu_time} gives the CPU time
-results for this comparison as a function of problem size while
-Figure~\ref{fig:seq_iterations} gives the number of iterations to
-converge as a function of problem size. In both cases, using the Monte
-Carlo solver as a synthetic acceleration rather than in a pure
-residual Monte Carlo scheme resulted in a reduction in both CPU time
-and iterations required to converge. The additional Richardson
-extrapolation between each Monte Carlo solve in the MCSA method gives
-a better converged residual source to use with the Monte Carlo
-calculation while the Sequential method requires more iterations to
-achieve the same level of convergence in the residual.
+To further motivate using Monte Carlo Synthetic Acceleration, we compare its
+performance to Halton's sequential Monte Carlo method on which our previous
+work in this area was based. For this comparison, we use the same transient
+Poisson problem as described in the previous section and choose only the
+5-point stencil to discretize the Laplacian operator as the previous results
+yielded little qualitative difference between the discretizations. Both MCSA
+and Halton's method are used with the adjoint Monte Carlo solver. In order to
+complete the same study as in the previous section, the number of histories
+computed by the Monte Carlo solver at each iteration had to be doubled to $2
+\times N \times N$ in order to ensure convergence in sequential Monte Carlo
+Method. Figure~\ref{fig:seq_cpu_time} gives the CPU time results for this
+comparison as a function of problem size while Figure~\ref{fig:seq_iterations}
+gives the number of iterations to converge as a function of problem size. In
+both cases, using the Monte Carlo solver as a synthetic acceleration rather
+than in a sequential Monte Carlo scheme resulted in a reduction in both CPU
+time and iterations required to converge. The additional Richardson iteration
+between each Monte Carlo solve in the MCSA method gives a better converged
+residual source to use with the Monte Carlo calculation while the sequential
+method requires more iterations to achieve the same level of convergence in
+the residual.
 
@@
   \caption{\sl CPU Time (s) to converge vs. Problem Size ($N$ for an
-    $N \times N$ square mesh). Both the Sequential Monte Carlo and
+    $N \times N$ square mesh). Both the sequential Monte Carlo and
     MCSA solvers are used with the five point stencils and the adjoint
@@
     number of discrete states in the system in order to ensure
-    convergence in the Sequential Monte Carlo method.}
+    convergence in the sequential Monte Carlo method.}
   \label{fig:seq_cpu_time}
@@
   \caption{\sl Iterations to converge vs. Problem Size ($N$ for an $N
-    \times N$ square mesh). Both the Sequential Monte Carlo and MCSA
+    \times N$ square mesh). Both the sequential Monte Carlo and MCSA
     solvers are used with the five point stencils and the adjoint
@@
 
-The benefits of using a synthetic acceleration scheme are also noted
-when the infinity norm of the residual computed at each iteration for
-both methods was collected at each iteration for a fixed problem sizes
-of $N=100$ and $N=500$ as shown in figures Figure~\ref{fig:seq_100}
-and \ref{fig:seq_500} respectively. In both cases, the Sequential
-method is subject to two regimes of exponential convergence with the
-later regime converging the slowest while the MCSA method exhibits a
-single rate of exponential convergence observed to be much higher than
-that computed by Halton's method. Even with the doubling of the number
-of stochastic histories computed per time step in order to ensure
-convergence for the Sequential method, we still see robustness issues
-with a non-monotonically decreasing residual observed for the $N=100$
-case. In both cases the MCSA solver is observed to be robust with a
-monotonically decreasing residual.
+The benefits of using a synthetic acceleration scheme are also noted when the
+infinity norm of the residual computed at each iteration for both methods was
+collected at each iteration for a fixed problem sizes of $N=100$ and $N=500$
+as shown in figures Figure~\ref{fig:seq_100} and \ref{fig:seq_500}
+respectively. In both cases, the sequential method is characterized by a slow
+approach to the asymptotic rate of convergence.  Conversely, the MCSA method
+reaches its asymptotic convergence rate more quickly and is faster than
+Halton's method. Even with the doubling of the number of stochastic histories
+computed per time step in order to ensure convergence for the sequential
+method, we still see robustness issues with a non-monotonically decreasing
+residual observed for the $N=100$ case. In both cases the MCSA solver is
+observed to be robust with a monotonically decreasing residual.
 
@@
   \caption{\sl Infinity norm of the solution residual vs. iteration
-    number for a problem of size $N=100$. Both the Sequential Monte
+    number for a problem of size $N=100$. Both the sequential Monte
     Carlo and MCSA solvers are used with the five point stencils and
@@
   \caption{\sl Infinity norm of the solution residual vs. iteration
-    number for a problem of size $N=500$. Both the Sequential Monte
+    number for a problem of size $N=500$. Both the sequential Monte
     Carlo and MCSA solvers are used with the five point stencils and
@@
 
-The general form of the equilibrium diffusion equation is \cite{morel_1996}
+The general form of the equilibrium radiation diffusion equation is
+\cite{morel_1996}
 \begin{equation}
@@
 implement the Monte Carlo scheme.  More specifically, the operator $\ve{D}$
-must be an H-matrix \cite{kelley_1995} such that the off-diagonal elements are
+must be an M-matrix \cite{kelley_1995} such that the off-diagonal elements are
 negative and symmetric.  In other words, the leakage out of a cell must be
@@
 \begin{enumerate}
-\item A fixed-point iteration to estimate $\bphi^{l+1/2}$.  This is a
+\item A Richardson iteration to estimate $\bphi^{l+1/2}$.  This is a
   matrix-vector multiply operation.
@@
 \begin{align*}
-   \bphi^{l+1/2} = (\vI - \ve{M}^{-1}\ve{D})\bphi^{l} +
-  \ve{M}^{-1}\ve{q}\:, \quad & (\text{fixed-point iteration})\\
-  %%
-  \vr^{l+1/2} = \ve{M}^{-1}\ve{q} - \ve{M}^{-1}\ve{D}\bphi^{l+1/2}\:,
+   \bphi^{l+1/2} = (\vI - \ve{T})\bphi^{l} +
+  \ve{M}^{-1}\ve{q}\:, \quad & (\text{Richardson iteration})\\
+  %%
+  \vr^{l+1/2} = \ve{M}^{-1}\ve{q} - \ve{\ve{T}}\bphi^{l+1/2}\:,
   \quad & (\text{compute residual})\\
   %%
-  \ve{M}^{-1}\hat{\ve{D}}\delta\bphi^{l+1/2} = \vr^{l+1/2}\:, \quad&
+  \hat{\ve{T}}\delta\bphi^{l+1/2} = \vr^{l+1/2}\:, \quad&
   (\text{estimate $\delta\phi$ using adjoint Monte Carlo method})\\
   %%
-  \bphi^{l+1} = \bphi^{l+1/2} + \delta\bphi^{l+1/2}\:. \quad&
+  \bphi^{l+1} = \bphi^{l+1/2} + \delta\bphi^{l+1/2}\:, \quad&
   (\text{update}\ \bphi)
 \end{align*}
-As described in \S~\ref{sec:iter-refin-monte}, the adjoint Monte Carlo method
-only approximately inverts $\ve{M}^{-1}\hat{\ve{D}}$ as indicated by the
-$\hat{\cdot}$ notation.  We now describe each of these in turn.
+where $\ve{T} = \ve{M}^{-1}\ve{D}$.  As described in
+\S~\ref{sec:iter-refin-monte}, the adjoint Monte Carlo method only
+approximately inverts $\ve{T}$ as indicated by the $\hat{\cdot}$ notation.
+We now describe each of these steps in turn.
 
@@
 when $\rho(\ve{H})< 1$, where $\ve{H}$ is the iteration matrix and is defined
-$\ve{H} = (\vI - \ve{M}^{-1}\ve{D})$.  The spectral radius for the Marshak
-problem is plotted in Figure~\ref{fig:marshak_spectral}.  As described in
+$\ve{H} = (\vI - \ve{T})$.  The spectral radius for the Marshak problem is
+plotted in Figure~\ref{fig:marshak_spectral}.  As described in
 \S~\ref{sec:preconditioning}, the preconditioning effectively reduces the
@@
 
-There are two principal degrees of freedom when utilizing the MCSA method: (1)
-the number of particles per stage and (2) the weight cutoff.
+There are two principal parameters when utilizing the MCSA method: (1) the
+number of particles per iteration and (2) the weight cutoff.
 Figure~\ref{fig:CPU_Np} shows the CPU time as a function of the requested
-number of particles per stage for a relative weight cutoff of $\sn{1}{-4}$.
-For less than five particles per stage, the method does not converge.
-Figure~\ref{fig:CPU_Ns} shows the variation in the maximum number of stages
-(iterations) per cycle with number of particles.  Analysis of these two
-figures shows that increasing the number of particles per stage reduces the
-number of iterations required to converge the solution.  However, the cost of
-the Monte Carlo transport is high; thus, better performance is attained by
-running more stages with fewer particles per stage \cite{evans_2003}.
+number of particles per iteration for a relative weight cutoff of
+$\sn{1}{-4}$.  For less than five particles per iteration, the method does not
+converge.  Figure~\ref{fig:CPU_Ns} shows the variation in the maximum number
+of iterations per cycle with number of particles.  Analysis of these two
+figures shows that increasing the number of particles per iteration reduces
+the number of iterations required to converge the solution.  However, the cost
+of the Monte Carlo transport is high; thus, better performance is attained by
+running more iterations with fewer particles per iteration \cite{evans_2003}.
 \begin{figure}[ht!]
   \centerline{ \includegraphics[width=5in,clip]{mrshk_np_CPU.pdf}}
-  \caption{CPU time versus number of particles per stage for the
+  \caption{CPU time versus number of particles per iteration for the
     Marshak problem.  The number of particles is the requested number.
-    In each stage the exact number can vary depending upon the number
+    In each iteration the exact number can vary depending upon the number
     of cells and source strength per cell.}
@@
   \centerline{ \includegraphics[width=5in,clip]{mrshk_np_Ns.pdf}}
-  \caption{ Maximum number of stages per cycle versus number of
-    particles per stage for the Marshak problem.  The number of
-    particles is the requested number.  In each stage the exact number
+  \caption{ Maximum number of iterations per cycle versus number of
+    particles per iteration for the Marshak problem.  The number of
+    particles is the requested number.  In each iteration the exact number
     can vary depending upon the number of cells and source strength
@@
 relatively insensitive to the weight cutoff.  This is especially true when
-running small numbers of particles per stage.  The weight cutoff has a more
+running small numbers of particles per iteration.  The weight cutoff has a more
 pronounced effect on the efficiency of the method when larger numbers of
@@
     problem.  Results are shown for 10 and 50 requested particles per
-    stage.  The numbers in parentheses show the maximum number of
-    particles run in a stage during the course of the problem.}
+    iteration.  The numbers in parentheses show the maximum number of
+    particles run in a iteration during the course of the problem.}
   \label{fig:CPU_wc}

### 2013-05-16 doc/jcomp2012/jcomp2012.tex
Reviewer edits.

@@
 
+  \author[ornl]{Steven P. Hamilton\fnref{fn1}} \ead{hamiltonsp@ornl.gov}
+
   \fntext[fn1]{Radiation Transport Group, Reactor and Nuclear Systems
@@
 \label{sec:introduction}
+
+In nearly all computational engineering and physics fields, linear and
+non-linear solvers form the core components of modeling and simulation
+applications. Recent focus on multiphysics coupling adds additional complexity
+to common linear and non-linear systems as solution strategies change when
+physical models are coupled. Furthermore, a desire for predictive simulations
+to enhance the safety and performance of engineered systems creates a need for
+extremely high fidelity computations to be performed for these coupled systems
+as a means to capture effects not modeled by coarser methods. In order to
+achieve this high fidelity, state-of-the-art computing facilities must be
+leveraged in a way that is both efficient and considerate of hardware-related
+issues. As scientific computing moves towards exascale facilities, with
+machines of $O(1,000,000)$ cores already coming online, new algorithms to
+solve these complex problems must be developed to leverage this new
+hardware. Issues such as resiliency to node failures and scaling to large
+numbers of heterogeneous computing elements (CPUs and GPUs) will be pertinent
+to robust algorithms aimed at this new hardware.
 
@@
 \end{itemize}
+While parallel scaling and resiliency on Petascale systems and beyond provides
+strong motivation for pursuing this work, we do not directly address these
+issues in this study.  These topics will be the subject of future
+investigations.
 
@@
 \end{equation}
-With this knowledge in hand, we can write an iterative method that
-solves Eq.~(\ref{eq:Ax=b}) (Richardson's Iteration):
+Using this series, we can write an iterative method that solves
+Eq.~(\ref{eq:Ax=b}) (Richardson's Iteration):
 \begin{equation}
@@
 \end{equation}
-The terms in the Neumann series can be interpreted as a series of
-transitions from $i_{m-1}\rightarrow i_m$ that can be simulated by a
-random walk.  Let $X$ be a random variable sampled from a random walk
-with $k$ events that initiates in state $i$:
+Here, $m$ is the random walk step index and the $h$ terms are the matrix
+elements of $\vH$.  The terms in the Neumann series can be interpreted as a
+series of transitions from $i_{m-1}\rightarrow i_m$ that can be simulated by a
+random walk.  Let $X$ be a random variable sampled from a random walk with $k$
+events that initiates in state $i$:
 \begin{equation}
@@
 The Kronecker delta implies that tallies are only made in the state where the
-random walk currently resides.  This is the common approach in standard Monte
-Carlo transport simulations. Thus, unlike the direct method, we are no longer
-required to start a history in each state for each random walk permutation
-because tallies are made in the state in which the random walk currently
-resides. Instead, sampling the source is sufficient, and one random walk per
-permutation is required.
+random walk currently resides.  Thus, the adjoint method is equivalent to the
+common approach in standard Monte Carlo transport simulations. As opposed to
+the direct method, the adjoint method does not require starting a history in
+each state for each random walk permutation because tallies are made in the
+state in which the random walk currently resides. Instead, sampling the source
+is sufficient, and one random walk per permutation is required.
 
@@
 
-We propose a modification of this scheme that uses Monte Carlo as a
-synthetic acceleration for the fixed-point iteration given by
+We propose a modification of this scheme that uses Monte Carlo as a synthetic
+acceleration \cite{dsa} for the fixed-point iteration given by
 Eq.~(\ref{eq:richardson}). We begin by subtracting Eq.~(\ref{eq:richardson})
@@
 \end{subequations}
-To accelerate the fixed-point method and avoid inverting $\vA$ directly in
-Eq.~(\ref{eq:synthetic-residual-solve}), we can instead use the above ideas to
-create a new iterative scheme that applies the Monte Carlo method to estimate
-the solution error. With this in hand, the Fixed-Point Monte Carlo
-Synthetic-Acceleration (MCSA) method can be defined as follows:
+These steps can be written as an accelerated iteration scheme that applies the
+Monte Carlo method to estimate the error in each iteration. The Monte Carlo
+method is used to approximately invert $\vA$ in
+Eq.~(\ref{eq:synthetic-residual-solve}). This updated scheme, the Monte Carlo
+Synthetic-Acceleration (MCSA) method, can be defined as follows:
 \begin{subequations}
@@
 Eq.~(\ref{eq:discrete-diffusion}), we must specify boundary conditions.  For
-this work, the Marshak Boundary condition suffices:
+this work, the Marshak boundary condition suffices:
 \begin{equation}
@@
 As noted above, the residual acts as the source for this simulation, and the
-operator $\ve{M}^{-1}\hat{\ve{D}}$ is implicitly defined in
+operator $\ve{T}$ is implicitly defined in
 Eq.~(\ref{eq:preconditioned-discrete-diffusion}).  There are two Monte Carlo
@@
 \end{equation*}
-where, following Eq.~(\ref{eq:point_iteration}), $\ve{H} = (\vI -
-\ve{M}^{-1}\ve{D})$ is the iteration matrix.  Now, using
-Eq.~(\ref{eq:adjoint-probability}), we define the scattering probabilities for
-cell 7 as
+where, following Eq.~(\ref{eq:point_iteration}), $\ve{H} = (\vI - \ve{T})$ is
+the iteration matrix.  Now, using Eq.~(\ref{eq:adjoint-probability}), we
+define the scattering probabilities for cell 7 as
 \begin{equation*}
@@
   \caption{ Solution of the Marshak wave problem at $t=100$~ns.  The analytic
-    solution is compared to the MCSA solution.  The MCSA, PCG, and PFIXED
-    methods are all converged to the same tolerance and therefore give
-    numerically equivalent answers.}
+    solution is compared to the MCSA solution.  All methods are converged to
+    the same tolerance and therefore give numerically equivalent answers.}
   \label{fig:marshak_10}
@@
 A 2D plot of the temperature at 1000~ns elapsed time is illustrated in
-Figure~\ref{fig:multi-mat_T}.  The time evolution of the temperature is shown at
-four edit points in Figure~\ref{fig:edit-points}.  All three methods, PCG, PFIX,
-and MCSA, give numerically identical answers when using a stopping criterion
-of $\sn{1}{-8}$.
+Figure~\ref{fig:multi-mat_T}.  The time evolution of the temperature is shown
+at four edit points in Figure~\ref{fig:edit-points}.  All three methods give
+numerically identical answers when using a stopping criterion of $\sn{1}{-8}$.
 \begin{figure}[ht!]
@@
 run out to a time of 100~ns.  The MCSA method was set to use 1000 particles
-per iteration resulting in 74,237,525 total particles over 10,157 cycles or
-823.68 particles per iteration.  Because more particles are required to get a
-reasonable estimate of the Neumann series, the relative efficiency per
-iteration of MCSA versus CG is significanly reduced compared to the smaller
-Marshak wave problem.  Nonetheless, MCSA is still marginally more efficient
-than CG with Jacobi preconditioning and significantly more efficient than CG
-with algegraic multigrid.
+based on numerical experiments similar to those applied for the Marshak
+problem.  A value of 1000 particles per iteration yields 74,237,525 total
+particles over 10,157 cycles or 823.68 particles per iteration.  Because more
+particles are required to get a reasonable estimate of the Neumann series, the
+relative efficiency per iteration of MCSA versus CG is significanly reduced
+compared to the smaller Marshak wave problem.  Nonetheless, MCSA is still
+marginally more efficient than CG with Jacobi preconditioning and
+significantly more efficient than CG with algegraic multigrid.
 \begin{table}[htbp!]
@@
 \end{table}
+More work is required to investigate automatic techiniques for optimizing
+particle count versus number of iterations with regards to both efficiency and
+convergence.  This will be the topic of a future paper.
 
@@
 the MCSA scheme may have advantages over traditional solution methods is on
-dentritic, or adaptive, meshes.  These meshes yield matrices with poor
+dendritic, or adaptive, meshes.  These meshes yield matrices with poor
 condition numbers because of the changing cell volumes at different refinement

### 2013-05-16 doc/jcomp2012/response.txt
Final edits.

Dear Editor,

     Thank you for the thorough reviews to our paper.  We hope that we have
     successfully responded to each reviewer.  Our responses follow below:

     Response to reviewer 2.

     We have added an acknowledgement indicating that the early part of the
     work was performed at LANL.  We have added Section 2.4 that does detailed
     comparisons between sequential Monte Carlo and MCSA.  Furthermore, we
     have added text in the discussion that highlights the differences between
     the methods.

     This reviewer made a number of key editorial observations.  We have
     adapted the text for each of them.  Some key points:

     - we added a reference to synthetic acceleration
     - we defined the terms in Eq. 6 and added some text for clarity
     - we didn't add "adjoint" to the abstract because we feel it would be
       confusing to the traditional Monte Carlo transport community.  We feel
       the discussion in the text clarifies the differences sufficiently. We
       also clarified the discussion involving the Kronecker delta in Eq. 13.
     - we chose Np and Wc for the multimaterial problem using numerical
       experiments; we added text indicating this.

     Response to reviewer 3.

     This reviewer correctly observed that the solvers that we compared MCSA
     to in our original manuscript and the way the information was presented
     was deficient.  We have expanded the results section to include GMRES and
     CG results with a family of state-of-the-art preconditioners.  We also
     updated the tables to include average and total iterations along with
     other useful quantities that makes the comparisons more meaningful.
     Additional text has been added to the results section to clarify the
     comparisons.

     We added a sentence clarifying the use of +|- in Eq 51.

     We added a paragraph providing more motivation to the introduction.

     We presented Figure 11 showing the spectral radius as a function of
     problem time in order to address the issue of variable time-step size
     versus solver performance.  Because all solvers use the same time-steps,
     no additional parameterization studies should be required.  We do
     appreciate that different solvers can behave differently based on
     time-step size, we feel that for this model equation, we are capturing
     the comparison between each method fairly.

### 2013-05-16 doc/jcomp2012/jcomp2012.tex
Final edits.

@@
     solutions converged to the same tolerance, it performs competitively with
-    deterministic methods including preconditioned conjugate gradient. We also
-    discuss various aspects of preconditioning the method and its general
-    applicability to broader classes of problems.
+    deterministic methods including preconditioned conjugate gradient and
+    GMRES. We also discuss various aspects of preconditioning the method and
+    its general applicability to broader classes of problems.
 
@@
 computed by the Monte Carlo solver at each iteration had to be doubled to $2
-\times N \times N$ in order to ensure convergence in sequential Monte Carlo
-Method. Figure~\ref{fig:seq_cpu_time} gives the CPU time results for this
-comparison as a function of problem size while Figure~\ref{fig:seq_iterations}
-gives the number of iterations to converge as a function of problem size. In
-both cases, using the Monte Carlo solver as a synthetic acceleration rather
-than in a sequential Monte Carlo scheme resulted in a reduction in both CPU
-time and iterations required to converge. The additional Richardson iteration
-between each Monte Carlo solve in the MCSA method gives a better converged
-residual source to use with the Monte Carlo calculation while the sequential
-method requires more iterations to achieve the same level of convergence in
-the residual.
+\times N \times N$ in order to ensure convergence when using the sequential
+Monte Carlo Method. Figure~\ref{fig:seq_cpu_time} gives the CPU time results
+for this comparison as a function of problem size while
+Figure~\ref{fig:seq_iterations} gives the number of iterations to converge as
+a function of problem size. In both cases, using the Monte Carlo solver as a
+synthetic acceleration rather than in a sequential Monte Carlo scheme resulted
+in a reduction in both CPU time and iterations required to converge. The
+additional Richardson iteration between each Monte Carlo solve in the MCSA
+method gives a better converged residual source to use with the Monte Carlo
+calculation while the sequential method requires more iterations to achieve
+the same level of convergence in the residual.
 
@@
 respectively. In both cases, the sequential method is characterized by a slow
-approach to the asymptotic rate of convergence.  Conversely, the MCSA method
-reaches its asymptotic convergence rate more quickly and is faster than
-Halton's method. Even with the doubling of the number of stochastic histories
-computed per time step in order to ensure convergence for the sequential
-method, we still see robustness issues with a non-monotonically decreasing
-residual observed for the $N=100$ case. In both cases the MCSA solver is
-observed to be robust with a monotonically decreasing residual.
+approach to the asymptotic rate of convergence.  Conversely, MCSA reaches its
+asymptotic convergence rate more quickly and exhibits a faster final rate of
+convergence than Halton's method. Even with the doubling of the number of
+stochastic histories computed per time step in order to ensure convergence for
+the sequential method, we still see robustness issues with a non-monotonically
+decreasing residual observed for the $N=100$ case. In both cases the MCSA
+solver is observed to be robust with a monotonically decreasing residual.
 
@@
 \end{equation}
-where we have dropped the subscript $\mathrm{R}$ from $\sigma$ for clarity.
-Solving this system for $\phi_{l+1/2}$ and applying to either $F_{l+1/2}^{+}$
-or $F_{l+1/2}^{-}$ and setting the resulting expression equal to
-Eq.~(\ref{eq:F_l+1/2}), one finds the face-centered diffusion coefficient,
+where we have dropped the subscript $\mathrm{R}$ from $\sigma$ for clarity and
+the $\pm$ superscripts indicate positive and negative directions across the
+face.  Solving this system for $\phi_{l+1/2}$ and applying to either
+$F_{l+1/2}^{+}$ or $F_{l+1/2}^{-}$ and setting the resulting expression equal
+to Eq.~(\ref{eq:F_l+1/2}), one finds the face-centered diffusion coefficient,
 \begin{equation}
@@
 \end{equation}
-This result is used to compute the residual of the system.  The residual then
+This result is used to compute the residual of the system that, in turn,
 becomes the source for the Monte Carlo simulation.
@@
 \begin{equation}
-  \ve{M}^{-1}\hat{\ve{D}}\delta\bphi^{l+1/2} = \vr^{l+1/2}\:.
+  \hat{\ve{T}}\delta\bphi^{l+1/2} = \vr^{l+1/2}\:.
 \end{equation}
@@
 methods to solve the radiation diffusion equation using the Trilinos solver
-library \cite{1089021}.  Conjugate Gradient (CG) is that standard method for
+library \cite{1089021}.  Conjugate Gradient (CG) is the standard method for
 symmetric positive-definite (SPD) systems like that defined in
@@
 and GMRES with incomplete LU (ILU) and ILU with threshold cutoff (ILUT). The
-multigrid preconditioner is the smoothed aggregation algebraic multigrid
-method implemented in the Trilinos ML package \cite{Gee:2007uj}.  In all the
-problems that follow we use a $4/3$ damping factor in the aggregation and a
-symmetric Gauss-Seidel smoother with 2 sweeps per level.
+Krylov methods and preconditioners are implemented throught the Trilinos
+AztecOO package.  The multigrid preconditioner is the smoothed aggregation
+algebraic multigrid method implemented in the Trilinos ML package
+\cite{Gee:2007uj}.  In all the problems that follow we use a $4/3$ damping
+factor in the aggregation and a symmetric Gauss-Seidel smoother with 2 sweeps
+per level, which are the defaults for the ML smoothed aggregation option.
 
@@
 the cost per iteration is exceptionally small.  Running more particles per
-iteration would reduce the number of iterations, but a greater cost per
+iteration would reduce the number of iterations, but at a greater cost per
 iteration.
@@
 arbitrary precision. Also, the new method has been to shown to be faster than
-preconditioned CG and Richardson iteration for the model problems presented
-here.
+preconditioned CG and GMRES for the model problems presented here.
 
@@
 
+Finally, exascale hardware will require resiliency to both hard and soft
+errors.  Stochastic solver approaches present a natural framework for both
+soft and hard errors.  In the former case, soft errors can be presented as
+high variance events.  For hard errors, histories from failed nodes can be
+safely discarded in the final calculation. The MCSA method provides a starting
+framework towards analyzing both of these failure conditions.
+
 %%---------------------------------------------------------------------------%%
@@
 Work for this paper was supported by Oak Ridge National Laboratory, which is
-managed and operated by UT-Batelle, LLC, for the U.S. Department of Energy under
-Contract No. DEAC05-00OR22725.
+managed and operated by UT-Batelle, LLC, for the U.S. Department of Energy
+under Contract No. DEAC05-00OR22725.  The early part of this work was
+performed at Los Alamos National Laboratory under U.S. Government contract
+W-7405-ENG-36.
 

### 2013-05-16 doc/jcomp2012/response.txt
Final, final edits.

@@
 
-     Thank you for the thorough reviews to our paper.  We hope that we have
-     successfully responded to each reviewer.  Our responses follow below:
+     Thank you for the thorough reviews to our paper.  Both reviewers did a
+     very thorough job, and in general, we have no objections to any of their
+     specific comments.  We believe we have updated the paper to reflect that
+     main issues presented by the reviewers.  Other than editorial additions,
+     the main items added were:
+
+     - a new section directly comparing sequential MC with MCSA
+     - we implemented GMRES and Multigrid preconditioned CG along with ILU and
+       ILUT preconditioning and compared these to MCSA for the model problem
+
+     We feel that these improvements address each reviewers primary critique.
+     We have endeavered to address each of the specific comments by the
+     reviewers, these are addressed individually below:
 
@@
      was deficient.  We have expanded the results section to include GMRES and
-     CG results with a family of state-of-the-art preconditioners.  We also
-     updated the tables to include average and total iterations along with
-     other useful quantities that makes the comparisons more meaningful.
-     Additional text has been added to the results section to clarify the
-     comparisons.
+     CG results with a family of state-of-the-art preconditioners, including
+     algebraic multigrid.  We also updated the tables to include average and
+     total iterations along with other useful quantities that makes the
+     comparisons more meaningful.  Additional text has been added to the
+     results section to clarify the comparisons.
 

### 2013-05-17 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
starting spn ilut section

--- 2013-05-16 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
@@
 For the fuel assembly problem, the initial block Jacobi
-preconditioned\footnote{For 1-group $SP_1$ problems, the block size is
+preconditioned\footnote{For 1-group $SP_1$ problems the block size is
   1, giving a point Jacobi preconditioner.}  calculations used a
@@
 iteration matrix generated by the diffusion problem was varied by
-changing the absorption cross section while fixing the grid size and
-scattering cross section. For each parameter variation, a minimum of
-one stochastic history per degree of freedom (DOF) in the problem was
-used to compute the Monte Carlo correction. If the solver could not
-converge in less than 100 iterations, the number of histories was
-increased by increments of 5,000 until convergence was achieved in
-less than 100 iterations. The number of iterations required to
-converge MCSA and the time to converge was recorded as a means to
-capture the breakdown.
+changing the absorption cross section from 0.25 to 100 while fixing
+the grid size at $100 \times 100$ with $h = 0.1$ and a fixed
+scattering cross section of unity. For each parameter variation, a
+minimum of one stochastic history per degree of freedom (DOF) in the
+problem was used to compute the Monte Carlo correction. If the solver
+could not converge in less than 100 iterations, the number of
+histories was increased by increments of 5,000 until convergence was
+achieved in less than 100 iterations. The number of iterations
+required to converge MCSA and the time to converge was recorded as a
+means to capture the breakdown.
 
@@
 this problem) with the number of iterations required to converge
-increasing a function of spectral radius. Near a spectral radius of
+increasing as a function of spectral radius. Near a spectral radius of
 0.97, the number of histories required to converge MCSA in less than
@@
 number of histories required becomes significant and effectively
-intractable to compute. Even with this simple diffusion problem, the
+impractical to compute. Even with this simple diffusion problem, the
 behavior is consistent with that observed for the fuel assembly
@@
 dozens of solves are typically required to compute the multiplication
-factor and flux spectrum within the $SP_N$ eigenvalue scheme, making
-the method unusable.
+factor and flux spectrum within the $SP_N$ eigenvalue scheme for more
+difficult problems like the fuel assembly, making the method unusable.
 \begin{figure}[t!]
@@
   \end{center}
-  \caption{\textbf{Histories required to converge as a function of
-      spectral radius for the neutron diffusion problem.} \textit{The
-      number of histories was increased to achieve convergence in less
-      than 100 iterations. At least 10,000 histories were used for
-      each calculation.}}
+  \caption{\textbf{Histories per DOF required to converge as a
+      function of spectral radius for the neutron diffusion problem.}
+    \textit{The number of histories was increased to achieve
+      convergence in less than 100 iterations. At least 10,000
+      histories were used for each calculation with 10,000 DOFs in the
+      problem.}}
   \label{fig:breakdown_histories}
@@
 problems with spectral radii above $\approx 0.97$, using MCSA to solve
-any problems of interest is completely unusable. We therefore require
-a more expansive set of preconditioning techniques to move the
-eigenvalue spectrum of the $SP_N$ problem into a regime in which MCSA
-is more applicable.
+any problems of interest is entirely ineffective and not practical. We
+therefore require a more expansive set of preconditioning techniques
+to move the eigenvalue spectrum of the $SP_N$ problem into a regime in
+which MCSA is more applicable and in which performance is improved.
 \begin{figure}[t!]
@@
 \label{subsec:spn_advanced_preconditioning}
+For the fuel assembly criticality problem presented in the previous
+section, the spectral radius was near one and in a regime in which
+MCSA breakdown is observed. In this regime the number of stochastic
+histories required to converge MCSA increases rapidly and the
+resulting poor performance is compounded by those histories being
+increasingly expensive to compute. To overcome this difficulty,
+advanced preconditioning strategies for the $SP_N$ problem are
+required beyond simple Jacobi methods that can reduce the spectral
+radius into a region of better MCSA performance. Several modern
+algebraic preconditioning strategies will be presented here and used
+within the explicit preconditioning framework given by
+Eq~(\ref{eq:left_right_mcsa}). Data showing their effects on MCSA
+solutions of the fuel assembly criticality problem will be presented.
 
@@
 \label{subsec:spn_ilut_preconditioning}
+Incomplete lower-upper (ILU) factorizations of the linear operator can
+be used a simple mechanism to form an approximate inverse as a
+preconditioner. To build the factorization, the sparse upper and lower
+triangular factors, $\mathbf{L}$ and $\mathbf{U}$, are computed such
+that the residual matrix formed by the factorization
+\citep{saad_iterative_2003}:
+\begin{equation}
+  \mathbf{R} = \mathbf{L} \mathbf{U} - \mathbf{A} \:,
+  \label{eq:ilu_residual_matrix}
+\end{equation}
+has a specified sparsity sparsity pattern and element magnitude.
 

### 2013-05-19 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding to ilut section and updating ilut data

@@
 \end{equation}
-has a specified sparsity sparsity pattern and element magnitude.
+has a specified sparsity pattern and element magnitude. When a
+magnitude threshold is used, elements generated in the factors below
+that maginitude are dropped, resulting in ILU threshold (ILUT)
+preconditioning. The sparsity pattern in this case is determined from
+the input matrix to be preconditioned and the number of elements
+maintained in the factor is specified by a fill level parameter. A
+fill level of 1 will generate the same number of elements as the
+sparsity pattern of the input matrix while a fill level of 2 will
+contain twice as many elements in the factorization, resulting in a
+better representation of the true LU factorization. The inverse of the
+lower and upper triangular factors may then be easily inverted by
+means of simple elimination to produce the preconditioner. For MCSA,
+$\mathbf{L}^{-1}$ will be used on the left and $\mathbf{U}^{-1}$ on
+the right to precondition the system.
+
+For modern subspace methods, only the action to the preconditioner on
+a vector is required for efficient implementations and therefore a
+triangular solve can be used for this purpose. For the explicit
+preconditioning scheme presented in Eq~(\ref{eq:left_right_mcsa}), the
+fully inverted operator must be generated in order to build the set of
+probabilities and weights required for Monte Carlo
+sampling. Therefore, for a linear system of size $N$, $N$ triangular
+solves will be required in order to extract the inverse matrices from
+production ILU implementations. In addition, parallel implementations
+typically generate lower and upper factors that are only triangular
+locally, providing an easily parallelizable mechanism to generate the
+action of the inverse. A consequence of this choice for parallel
+scalability is a degradation of the preconditioning quality as the
+size of the parallel system is increased. For serial computations, the
+triangular factorization is potentially exact depending on the
+parameters chosen while at thousands of parallel tasks, the global
+triangular factors differ signficantly from the true factorization. As
+a result, more iterations are required at higher levels of parallelism
+to converge the system and will ultimately degrade overall scalability
+of the system with respect to total wall time to converge.
+
+MCSA preconditioned with ILUT was used to solve the fuel assembly
+problem presented in the previous section for a 1-group $SP_1$
+discretization. Unlike the Jacobi preconditioned strategy, convergence
+was achieved with ILUT preconditioning. To study convergence
+sensitivity to ILUT parameters, the fill level and drop tolerance were
+parametrically varied with the number of iterations to converge
+recorded along with the maximum number of non-zero entries observed in
+all matrix rows. To provide some sparsity to the factorization, the
+ILUT drop tolerance was used to drop elements in the extracted inverse
+factors. For each calculation, the number of iterations required to
+converge reported was for the first eigenvalue iteration with
+\sn{3}{4} histories at each MCSA iteration to compute the Monte Carlo
+correction.
 
@@
 \label{subsec:spn_spainv_preconditioning}
+Explicit formation of the preconditioner inverse is requirement of
+MCSA preconditioning.
 

### 2013-05-20 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding to preconditioning section for SPN.

@@
 trillions of unknown angular flux moments to be computed to achieve
-good accuracy for the repsonses of interest
+good accuracy for the responses of interest
 \citep{slaybaugh_acceleration_2011}. Other forms of the transport
@@
 parallelism is achieved by an efficient parallel sweep, angular
-efficiency achieved by pipelining, and energy parallelism achieved by
+efficiency achieved by pipe-lining, and energy parallelism achieved by
 decoupling the groups. With the $SP_N$ equations as a full matrix
@@
 For this work, it is sufficient to formulate
-Eq~(\ref{eq:general_transport}) in 1-dimensonal Cartesian geometry:
+Eq~(\ref{eq:general_transport}) in 1-dimensional Cartesian geometry:
 \begin{multline}
@@
 \end{itemize}
-In 1-dimensional cartesian geometry,
+In 1-dimensional Cartesian geometry,
 Eq~(\ref{eq:eigenvalue_transport}) becomes:
@@
 the diffusive term in each equation is only for a single
-'psuedo-moment' $u_n$. As noted by Evans \citep{evans_simpli_2013},
+'pseudo-moment' $u_n$. As noted by Evans \citep{evans_simpli_2013},
 lower order $SP_N$ approximations can be generated by setting higher
@@
 $SP_N$ equations for a single energy group, we form them for each of
-the energy groups with group coupling occuring through the scattering
+the energy groups with group coupling occurring through the scattering
 term. The multigroup $P_N$ equations are then:
@@
 scalar values replaced by the group-vector values. In addition,
-$\mathbb{A}$ is now a block matrix of $N_g \times N_g$ submatrices
+$\mathbb{A}$ is now a block matrix of $N_g \times N_g$ sub-matrices
 generated from the moment scattering matrices:
@@
       geometry. Gray regions are homogenized fuel and red regions are
-      homogenzied moderator. Each fuel pin is resolved by a $2 \times
+      homogenized moderator. Each fuel pin is resolved by a $2 \times
       2$ mesh.}}
@@
 discretize the geometry and single-dimension pin-cell calculations for
-intial flux spectrum generation. Table~\ref{tab:problem3_parameters}
+initial flux spectrum generation. Table~\ref{tab:problem3_parameters}
 gives the primary design parameters for the fuel assembly
@@
 system. To study the breakdown of MCSA at iteration matrix spectral
-radii near one, we will use a simpler homogoneous 2-dimensional
+radii near one, we will use a simpler homogeneous 2-dimensional
 one-speed neutron diffusion system to isolate this behavior. In this
@@
 spectral radii over 0.97 have histories that require two orders of
-magnitude more computation time. This signficant increase in
+magnitude more computation time. This significant increase in
 computation time per history coupled with the significant increase in
@@
 magnitude threshold is used, elements generated in the factors below
-that maginitude are dropped, resulting in ILU threshold (ILUT)
+that magnitude are dropped, resulting in ILU threshold (ILUT)
 preconditioning. The sparsity pattern in this case is determined from
@@
 locally, providing an easily parallelizable mechanism to generate the
-action of the inverse. A consequence of this choice for parallel
-scalability is a degradation of the preconditioning quality as the
-size of the parallel system is increased. For serial computations, the
-triangular factorization is potentially exact depending on the
-parameters chosen while at thousands of parallel tasks, the global
-triangular factors differ signficantly from the true factorization. As
-a result, more iterations are required at higher levels of parallelism
-to converge the system and will ultimately degrade overall scalability
-of the system with respect to total wall time to converge.
+action of the preconditioner inverse. A consequence of this choice for
+parallel scalability is a degradation of the preconditioning quality
+as the size of the parallel system is increased. For serial
+computations, the triangular factorization is potentially exact
+depending on the parameters chosen while at thousands of parallel
+tasks, the global triangular factors differ significantly from the true
+factorization. As a result, more iterations are required at higher
+levels of parallelism to converge the system and will ultimately
+degrade overall scalability of the system with respect to total wall
+time to converge.
 
@@
 \label{subsec:spn_spainv_preconditioning}
-Explicit formation of the preconditioner inverse is requirement of
-MCSA preconditioning.
+Explicit formation of the preconditioner inverse is requirement of the
+chosen MCSA preconditioning strategy. As a result, although ILUT
+preconditioning provided excellent iterative performance, a
+significant time penalty is paid to extract the inverse of the
+triangular factors through $N$ triangular solves. In addition, this
+explicit inversion yielded a composite linear operator that was dense,
+requiring modification of the system in order to achieve marginal
+scalability. Sparse approximate inverse (SPAINV) preconditioning is a
+technique that may potentially alleviate both of these constraints at
+the cost of reduced iterative performance. The method produces a
+sparse approximation of the inverse matrix directly by minimizing the
+Frobenius norm of the residual matrix \citep{saad_iterative_2003}:
+\begin{equation}
+  || \mathbf{I} - \mathbf{A} \mathbf{M} ||^2_F =
+  \sum_{j=1}^N ||\mathbf{e}_j - \mathbf{A} \mathbf{m}_j||^2_2 \:,
+  \label{eq:frobenius_norm_min}
+\end{equation}
+where the inverse preconditioning matrix $\mathbf{M}$ minimizes the
+norm on the right and $\mathbf{e}_j$ and $\mathbf{m}_j$ are the
+$j^{th}$ columns of the identity and preconditioning matrix
+respectively. Here, $\mathbf{M}$ is the actual inverse of the
+preconditioner and is formed directly by the algorithm. On the right
+hand side of Eq~(\ref{eq:frobenius_norm_min}), the Frobenius norm is
+represented as a effective sum of $N$ linear system residual norms. A
+few iterations of a subspace method or other iterative scheme can be
+used to effectively solve these systems and yield the columns of the
+approximate preconditioning matrix. At each iteration, tolerance
+values are used to eliminate the components of each $\mathbf{m}_j$
+column to maintain the desired level of sparsity. In addition, a
+sparsity pattern can be predetermined and enforced during this
+process.
 
@@
 \label{subsec:spn_spainv_preconditioning}
+For improved convergence over SPAINV preconditioning while maintaining
+sparsity, ILUT preconditioning may be improved through an additional
+application of SPAINV \citep{saad_iterative_2003}. For the ILUT
+preconditioning case we have the following linear problem to solve for
+the substituted variable:
+\begin{equation}
+  \mathbf{L}^{-1} \mathbf{A} \mathbf{U}^{-1} \mathbf{u} =
+  \mathbf{L}^{-1} \mathbf{b}\:.
+  \label{eq:ilut_standalone}
+\end{equation}
+We then apply SPAINV and minimize the following problem, this time on
+the left:
+\begin{equation}
+  || \mathbf{I} - \mathbf{M} \mathbf{L}^{-1} \mathbf{A}
+  \mathbf{U}^{-1} ||^2_F\:,
+  \label{eq:ilut_minimize}
+\end{equation}
+giving a new preconditioned system:
+\begin{equation}
+  \mathbf{M} \mathbf{L}^{-1} \mathbf{A} \mathbf{U}^{-1} \mathbf{u} =
+  \mathbf{M} \mathbf{L}^{-1} \mathbf{b}\:.
+  \label{eq:ilut_spainv}
+\end{equation}
+Doing this permits a smaller fill level to be used with ILUT and a
+lower-order sparsity pattern to be used with SPAINV, giving more
+sparsity in the resulting composite linear operator while maintaining
+good iterative performance.
 
@@
 \label{subsec:spn_mcsa_relaxation}
+The Richardson iteration upon which the Neumann-Ulam method is built
+may be implemented with a scalar relaxation parameter that can be
+adjusted to improve convergence:
+\begin{equation}
+  \mathbf{x} = \mathbf{x} + \omega \mathbf{r}\:,
+  \label{eq:richardson_relaxation}
+\end{equation}
+where $\omega$ is the relaxation parameter. Analogously, these
+relaxation parameter techniques can be applied to Monte Carlo to
+improve convergence as demonstrated by Dimov
+\citep{dimov_new_1998}. In this case, building an iteration matrix
+from Eq~(\ref{eq:richardson_relaxation} gives:
+\begin{equation}
+  \mathbf{H} = \mathbf{I} - \omega \mathbf{A}\:,
+  \label{eq:relaxed_iteration_matrix}
+\end{equation}
+with the probabilities and weights for the Monte Carlo game
+appropriately scaled. By inspection, such a scaling is a simple form
+of preconditioning on the left where all rows in the system are scaled
+by the same scalar parameter. For MCSA, we can stage this scheme with
+two separate relaxation parameters: one for the outer Richardson
+iteration and one for the inner Monte Carlo solve:
+\begin{subequations}
+  \begin{gather}
+    \ve{x}^{k+1/2} = \ve{x}^k + \omega_R \ve{r}^k\:,\\ \ve{r}^{k+1/2}
+    = \ve{b} -
+    \ve{A}\ve{x}^{k+1/2}\:,\\ \omega_N\ve{A}\delta\ve{x}^{k+1/2} =
+    \omega_N\ve{r}^{k+1/2}\:,\\ \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta
+    \ve{x}^{k+1/2}\:,\\ \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:,
+  \end{gather}
+  \label{eq:mcsa_relaxed}
+\end{subequations}
+where $\omega_R$ is the Richardson iteration relaxation parameter and
+$\omega_N$ is the Neumann-Ulam Monte Carlo solve relaxation parameter.
 
@@
 \label{subsec:spn_estimator_comparison}
+With convergence obtained for the fuel assembly problem, the advanced
+preconditioners will next be studied with each Monte Carlo
+estimator. Although the expected value estimator has achieved the best
+iterative performance in the previous comparison and analysis, it is
+possible that the other estimators may perform better from a timing
+perspective due to the sparsity lost by explicit preconditioning. The
+deterministic component of the expected value estimator that couples
+the current state to other local states through the iteration matrix
+stencil requires application of the estimator to potentially several
+orders of magnitude more states at every transition event.
 

### 2013-05-20 doc/dissertation/document/chapters/mc_background/mc_background.tex
starting spn ilut section
updates to ilut preconditioning

@@
   \begin{center}
-    \includegraphics[width=5in]{chapters/mc_background/direct_evolution.png}
+    \includegraphics[width=6in]{chapters/mc_background/direct_evolution.png}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=5in]{chapters/mc_background/adjoint_evolution.png}
+    \includegraphics[width=6in]{chapters/mc_background/adjoint_evolution.png}
   \end{center}
@@
     \ve{r}^{k+1} = \ve{M}_L^{-1}(\ve{b}-\ve{A}\ve{M}_R^{-1}\ve{u}^{k+1})\:,
+  \end{gather}
   \label{eq:left_right_mcsa}
-  \end{gather}
 \end{subequations}
@@
   \centering
-  \includegraphics[width=5in,clip]{chapters/mc_background/dir_adj_cpu.pdf}
+  \includegraphics[width=6in,clip]{chapters/mc_background/dir_adj_cpu.pdf}
   \caption{\textbf{CPU Time (s) to converge vs. Problem Size ($N$ for
@@
   \centering
-  \includegraphics[width=5in,clip]{chapters/mc_background/dir_adj_iterations.pdf}
+  \includegraphics[width=6in,clip]{chapters/mc_background/dir_adj_iterations.pdf}
   \caption{\textbf{Iterations to converge vs. Problem Size ($N$ for an
@@
   \centering
-  \includegraphics[width=5in,clip]{chapters/mc_background/dir_adj_conv.pdf}
+  \includegraphics[width=6in,clip]{chapters/mc_background/dir_adj_conv.pdf}
   \caption{\textbf{Infinity norm of the solution residual
@@
   \centering
-  \includegraphics[width=5in,clip]{chapters/mc_background/seq_cpu.pdf}
+  \includegraphics[width=6in,clip]{chapters/mc_background/seq_cpu.pdf}
   \caption{\textbf{CPU Time (s) to converge vs. Problem Size ($N$ for
@@
   \centering
-  \includegraphics[width=5in,clip]{chapters/mc_background/seq_iterations.pdf}
+  \includegraphics[width=6in,clip]{chapters/mc_background/seq_iterations.pdf}
   \caption{\textbf{Iterations to converge vs. Problem Size ($N$ for an
@@
   \centering
-  \includegraphics[width=5in,clip]{chapters/mc_background/seq_conv_100.pdf}
+  \includegraphics[width=6in,clip]{chapters/mc_background/seq_conv_100.pdf}
   \caption{\textbf{Infinity norm of the solution residual
@@
   \centering
-  \includegraphics[width=5in,clip]{chapters/mc_background/seq_conv_500.pdf}
+  \includegraphics[width=6in,clip]{chapters/mc_background/seq_conv_500.pdf}
   \caption{\textbf{Infinity norm of the solution residual
@@
   \centering
-  \includegraphics[width=5in,clip]{chapters/mc_background/estimator_nh_iters.pdf}
+  \includegraphics[width=6in,clip]{chapters/mc_background/estimator_nh_iters.pdf}
   \caption{\textbf{Iterations (s) to converge vs. Monte Carlo
@@
   \centering
-  \includegraphics[width=5in,clip]{chapters/mc_background/estimator_nh_time.pdf}
+  \includegraphics[width=6in,clip]{chapters/mc_background/estimator_nh_time.pdf}
   \caption{\textbf{CPU Time (s) to converge vs. Monte Carlo histories
@@
   \centering
-  \includegraphics[width=5in,clip]{chapters/mc_background/estimator_wc_iters.pdf}
+  \includegraphics[width=6in,clip]{chapters/mc_background/estimator_wc_iters.pdf}
   \caption{\textbf{Iterations (s) to converge vs. history weight
@@
   \centering
-  \includegraphics[width=5in,clip]{chapters/mc_background/estimator_wc_time.pdf}
+  \includegraphics[width=6in,clip]{chapters/mc_background/estimator_wc_time.pdf}
   \caption{\textbf{CPU Time (s) to converge vs. history weight cutoff
@@
   \centering
-  \includegraphics[width=5in,clip]{chapters/mc_background/absorption_iters.pdf}
+  \includegraphics[width=6in,clip]{chapters/mc_background/absorption_iters.pdf}
   \caption{\textbf{Iterations (s) to converge vs. artificial
@@
   \centering
-  \includegraphics[width=5in,clip]{chapters/mc_background/absorption_time.pdf}
+  \includegraphics[width=6in,clip]{chapters/mc_background/absorption_time.pdf}
   \caption{\textbf{CPU Time (s) to converge vs. artificial absorption

### 2013-05-20 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
updates to ilut preconditioning

@@
   \begin{center}
-    \includegraphics[width=3in]{chapters/spn_equations/group1.png}
+    \includegraphics[width=4in]{chapters/spn_equations/group1.png}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=3in]{chapters/spn_equations/group10ds.png}
+    \includegraphics[width=4in]{chapters/spn_equations/group10ds.png}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=3in]{chapters/spn_equations/group10us.png}
+    \includegraphics[width=4in]{chapters/spn_equations/group10us.png}
   \end{center}
@@
 to this diagonal form.
+
+\clearpage
 
@@
   \begin{center}
-    \includegraphics[width=5in]{chapters/spn_equations/block_jacobi_res.pdf}
+    \includegraphics[width=6in]{chapters/spn_equations/block_jacobi_res.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=5in]{chapters/spn_equations/block_jacobi_rich_res.pdf}
+    \includegraphics[width=6in]{chapters/spn_equations/block_jacobi_rich_res.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=5in]{chapters/spn_equations/breakdown_iterations.pdf}
+    \includegraphics[width=6in]{chapters/spn_equations/breakdown_iterations.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=5in]{chapters/spn_equations/breakdown_histories.pdf}
+    \includegraphics[width=6in]{chapters/spn_equations/breakdown_histories.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=5in]{chapters/spn_equations/breakdown_time.pdf}
+    \includegraphics[width=6in]{chapters/spn_equations/breakdown_time.pdf}
   \end{center}
@@
 correction.
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/spn_equations/ilut_iterations.pdf}
+  \end{center}
+  \caption{\textbf{Number of MCSA iterations required to converge the
+      second eigenvalue iteration for the fuel assembly problem as a
+      function of ILUT drop tolerance.} \textit{Each colored curve
+      represents the iteration behavior for a different ILUT fill
+      level. Fill levels of 1.5, 2.0, 2.5, and 3.0 were used.}}
+  \label{fig:ilut_iterations}
+\end{figure}
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/spn_equations/ilut_size.pdf}
+  \end{center}
+  \caption{\textbf{Maximum number of non-zero entries observed for all
+      rows in the composite linear operator for the fuel assembly
+      problem given as a function of ILUT drop tolerance.}
+    \textit{Each colored curve represents the row size for a different
+      ILUT fill level. Fill levels of 1.5, 2.0, 2.5, and 3.0 were
+      used.}}
+  \label{fig:ilut_size}
+\end{figure}
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/spn_equations/ilut_quality.pdf}
+  \end{center}
+  \caption{\textbf{ILUT preconditioning quality metric for the fuel
+      assembly problem given as a function of ILUT drop tolerance.}
+    \textit{Each colored curve represents the quality metric behavior
+      for a different ILUT fill level. Fill levels of 1.5, 2.0, 2.5,
+      and 3.0 were used.}}
+  \label{fig:ilut_quality}
+\end{figure}
+
+\clearpage
 

### 2013-05-20 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding psilut analysis data

@@
 sensitivity to ILUT parameters, the fill level and drop tolerance were
-parametrically varied with the number of iterations to converge
-recorded along with the maximum number of non-zero entries observed in
-all matrix rows. To provide some sparsity to the factorization, the
-ILUT drop tolerance was used to drop elements in the extracted inverse
-factors. For each calculation, the number of iterations required to
-converge reported was for the first eigenvalue iteration with
-\sn{3}{4} histories at each MCSA iteration to compute the Monte Carlo
-correction.
+parametrically varied with the number of iterations to converge the
+second eigenvalue iteration for the fuel assembly problem recorded
+along with the maximum number of non-zero entries observed in all
+matrix rows for the left/right preconditioned composite linear
+operator. To provide some sparsity to the factorization, the ILUT drop
+tolerance was used to drop elements in the extracted inverse
+triangular factors. For each calculation, the number of iterations
+required to converge reported was for the first eigenvalue iteration
+with \sn{3}{4} histories at each MCSA iteration to compute the Monte
+Carlo correction using the adjoint collision estimator. All ILUT
+calculations reported here were performed on a single CPU and
+therefore this data does not take into account the aforementioned
+effects of parallel decomposition on the quality of the ILUT
+preconditioning.
+
+Figure~\ref{fig:ilut_iterations} gives the number of iterations to
+converge the fixed source problem to a tolerance of \sn{1}{8}. As
+expected, the higher the fill level chosen for the ILUT factorization,
+the fewer iterations are required to converge the problem (only 9 MCSA
+iterations were required to converge the problem for the fill level of
+3 and drop tolerance of \sn{1}{-5} case). For higher levels of fill,
+the iterations needed to converge were not as sensitive, signaling
+that a larger drop tolerance can perhaps be used without a significant
+degradation in iterative performance. At smaller fill levels, the
+sensitivity to the ILUT drop tolerance is more significant. For all
+fill levels used, convergence was not achieved for the fuel assembly
+problem with a drop tolerance larger than \sn{1}{-3}.
 \begin{figure}[t!]
@@
 
+Unfortunately, gaining convergence (and excellent iterative
+performance) with MCSA for the $SP_N$ fuel assembly problem comes at
+an immediate cost. Figure~\ref{fig:ilut_size} gives the maximum number
+of non-zero entries in the composite linear operator generated by
+preconditioning as a function ILUT drop tolerance for varying values
+of fill level. For the 1-group $SP_1$ discretization, the original
+linear operator will contain only a maximum of 7 non-zero entries per
+row in the system. As observed in Figure~\ref{fig:ilut_size},
+performing the explicit preconditioning yields composite linear
+operators with $O(1,000)$ elements in a row for all combinations of
+fill level and drop tolerance, over 10\% of the row size for this
+particular. This massive number of row entires, observed for a
+significant fraction of rows in the system, creates several
+problems. First, sparsity is completely destroyed with each state in
+the system now coupled to over 10\% of the states in the system
+through the composite matrix. This means that Monte Carlo sampling
+tables will be large, requiring significant memory to store them and a
+substantial overhead in the sampling procedure during transport. In
+addition, the matrix-matrix multiply operations required to build the
+composite operator see significant performance losses due to the
+amount of data that must be handled. In parallel, losing sparsity
+severely inhibits performance where now parallel operations require
+communications amongst orders of magnitude more processors.
 \begin{figure}[t!]
@@
 
+As a means to assess the quality of the preconditioning, a simple
+metric is developed to address these concerns and allow comparison to
+future developments. For our studies, our core performance metric will
+be iterative performance with the minimum number of MCSA iterations
+required to converge the problem desired. For Monte Carlo
+calculations, this improved performance is balanced by the creation of
+a dense composite system and we seek to reduce the number of non-zero
+entries to a minimum value in order to lower the amount of coupling
+amongst states in the system. Given these two objectives, the
+following metric is proposed:
+\begin{equation}
+  \text{Quality Metric} = (\text{\# iterations})*(\text{maximum \# of
+    non-zero values})\:,
+\end{equation}
+where the highest quality preconditioning is one that minimizes this
+metric. For the ILUT preconditioned data provided,
+Figure~\ref{fig:ilut_quality} provides the computed metric as a
+function of ILUT drop tolerance for each fill level used. In general,
+the metric is similar to the data observed for the number of
+iterations required to converge as the non-zero row entries changed by
+a smaller fraction over drop tolerances tested.
 \begin{figure}[t!]
@@
 \end{figure}
+
+It should be noted here that although only the behavior of the linear
+solve during the second eigenvalue iteration is reported, the behavior
+of the linear solver was observed to be consistent throughout the
+eigenvalue iterations (25 were required to converge the 1-group $SP_1$
+problem). We expect this as the linear operator (which is unchanging
+in the fuel assembly criticality problem) dictates the convergence of
+MCSA. At each iteration the fission source provided to the linear
+problem is changing, however, we expect the same convergence behavior
+for all source vectors $\mathbf{b} \in \mathbb{R}^{N}$ where
+$||\mathbf{b}||_1 \neq 0$.
 
@@
 sparsity pattern can be predetermined and enforced during this
-process.
+process. By using the norm minimization strategy instead of a
+factorization such as ILUT, parallel results are reproduced regardless
+of parallel problem size, meaning that a serial computation will
+converge in the same number of iterations as a computation with
+thousands of cores.
+
+MCSA preconditioned with SPAINV was used to solve the fuel assembly
+problem as with ILUT for a 1-group $SP_1$ discretization. To study
+convergence sensitivity to SPAINV parameters, the number of levels in
+the sparsity pattern and threshold were parametrically varied with the
+number of iterations to converge the second eigenvalue iteration for
+the fuel assembly problem recorded along with the maximum number of
+non-zero entries observed in all matrix rows for the left/right
+preconditioned composite linear operator. For each calculation, the
+number of iterations required to converge reported was for the first
+eigenvalue iteration with \sn{3}{4} histories at each MCSA iteration
+to compute the Monte Carlo correction using the adjoint collision
+estimator. All SPAINV calculations reported here were performed on a
+single CPU.
+
+\clearpage
 
@@
 good iterative performance.
+
+\clearpage
 
@@
 
+\clearpage
+
 \subsection{Monte Carlo Estimator Comparison}

### 2013-05-21 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding gnumeric data files for spn preconditioner analysis

@@
 fill level and drop tolerance, over 10\% of the row size for this
-particular. This massive number of row entires, observed for a
+particular. This massive number of row entries, observed for a
 significant fraction of rows in the system, creates several
 problems. First, sparsity is completely destroyed with each state in
-the system now coupled to over 10\% of the states in the system
-through the composite matrix. This means that Monte Carlo sampling
-tables will be large, requiring significant memory to store them and a
-substantial overhead in the sampling procedure during transport. In
-addition, the matrix-matrix multiply operations required to build the
-composite operator see significant performance losses due to the
-amount of data that must be handled. In parallel, losing sparsity
-severely inhibits performance where now parallel operations require
-communications amongst orders of magnitude more processors.
+the system now coupled to over 10\% of the total states in the system
+through the composite iteration matrix. This means that Monte Carlo
+sampling tables will be large, requiring significant memory to store
+them and a substantial overhead in the sampling procedure during
+transport. In addition, the matrix-matrix multiply operations required
+to build the composite operator see significant performance losses due
+to the amount of data that must be handled. In parallel, losing
+sparsity severely inhibits performance where now parallel operations
+require communications amongst orders of magnitude more processors for
+nearest-neighbor type algorithms.
 \begin{figure}[t!]
@@
 entries to a minimum value in order to lower the amount of coupling
-amongst states in the system. Given these two objectives, the
-following metric is proposed:
+amongst states in the system and reduce the amount of memory used
+along with the potentially latency overhead. Given these two
+objectives, the following metric is proposed:
 \begin{equation}
@@
 \label{subsec:spn_spainv_preconditioning}
-Explicit formation of the preconditioner inverse is requirement of the
-chosen MCSA preconditioning strategy. As a result, although ILUT
+Explicit formation of the preconditioner inverse is a requirement of
+the chosen MCSA preconditioning strategy. As a result, although ILUT
 preconditioning provided excellent iterative performance, a
@@
 explicit inversion yielded a composite linear operator that was dense,
-requiring modification of the system in order to achieve marginal
+requiring modification of the system in order to achieve better
 scalability. Sparse approximate inverse (SPAINV) preconditioning is a
@@
 norm on the right and $\mathbf{e}_j$ and $\mathbf{m}_j$ are the
-$j^{th}$ columns of the identity and preconditioning matrix
+$j^{th}$ columns of the identity and preconditioning matrices
 respectively. Here, $\mathbf{M}$ is the actual inverse of the
@@
 few iterations of a subspace method or other iterative scheme can be
-used to effectively solve these systems and yield the columns of the
-approximate preconditioning matrix. At each iteration, tolerance
-values are used to eliminate the components of each $\mathbf{m}_j$
-column to maintain the desired level of sparsity. In addition, a
-sparsity pattern can be predetermined and enforced during this
-process. By using the norm minimization strategy instead of a
-factorization such as ILUT, parallel results are reproduced regardless
-of parallel problem size, meaning that a serial computation will
-converge in the same number of iterations as a computation with
-thousands of cores.
+used to effectively solve these systems to a loose tolerance and yield
+the columns of the approximate preconditioning matrix. At each
+iteration, threshold values are used to eliminate the components of
+each $\mathbf{m}_j$ column to maintain the desired level of
+sparsity. In addition, a sparsity pattern can be predetermined and
+enforced during this process. By using the norm minimization strategy
+instead of a factorization such as ILUT, parallel results are
+reproduced regardless of parallel problem size, meaning that a serial
+computation will converge in the same number of iterations as a
+computation with thousands of cores.
 

### 2013-05-21 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding preconditioning results

@@
   \caption{\textbf{Number of MCSA iterations required to converge the
-      second eigenvalue iteration for the fuel assembly problem as a
-      function of ILUT drop tolerance.} \textit{Each colored curve
-      represents the iteration behavior for a different ILUT fill
-      level. Fill levels of 1.5, 2.0, 2.5, and 3.0 were used.}}
+      second eigenvalue iteration for the fuel assembly problem with
+      ILUT preconditioning as a function of ILUT drop tolerance.}
+    \textit{Each colored curve represents the iteration behavior for a
+      different ILUT fill level. Fill levels of 1.5, 2.0, 2.5, and 3.0
+      were used.}}
   \label{fig:ilut_iterations}
@@
       rows in the composite linear operator for the fuel assembly
-      problem given as a function of ILUT drop tolerance.}
-    \textit{Each colored curve represents the row size for a different
-      ILUT fill level. Fill levels of 1.5, 2.0, 2.5, and 3.0 were
-      used.}}
+      problem with ILUT preconditioning given as a function of ILUT
+      drop tolerance.}  \textit{Each colored curve represents the row
+      size for a different ILUT fill level. Fill levels of 1.5, 2.0,
+      2.5, and 3.0 were used.}}
   \label{fig:ilut_size}
@@
 sparsity. In addition, a sparsity pattern can be predetermined and
-enforced during this process. By using the norm minimization strategy
-instead of a factorization such as ILUT, parallel results are
-reproduced regardless of parallel problem size, meaning that a serial
-computation will converge in the same number of iterations as a
-computation with thousands of cores.
+enforced during this process. For the SPAINV implementation used for
+this work, the sparsity pattern was defined by levels such that a
+level of $N$ for an input operator $\mathbf{A}$ will generate a
+preconditioning matrix $\mathbf{M}$ with the same sparsity pattern as
+$\mathbf{A}^{N+1}$. By using the norm minimization strategy instead of
+a factorization such as ILUT, parallel results are reproduced
+regardless of parallel problem size, meaning that a serial computation
+will converge in the same number of iterations as a computation with
+thousands of cores.
 
@@
 
+Figure~\ref{fig:spainv_iterations} gives the number of MCSA iterations
+needed to converge the fuel assembly problem and
+Figure~\ref{fig:spainv_size} the maximum number of non-zero entries
+per row obeserved in the composite linear operator as a function the
+SPAINV threshold. For this analysis, the number of levels in the
+sparsity pattern was varied from 3 to 7 with convergence of the fuel
+assembly problem not achieved for smaller level values. Compared to
+ILUT preconditioning, at higher levels we see comparable iterative
+performance with a relative invariance to the threshold value. The
+threshold did have a significant effect on the time required to
+generate the preconditioner with a threshold of 0.001 over an order of
+magnitude slower than a threshold of 0.01 due to the inclusion of a
+significant number of extra values in the input operator. In addition
+to comparable iterative performance, the sparsity of the composite
+operator is greatly improved with nearly an order of magnitude
+reduction in number of non-zero entries per row for lower level
+values.
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/spn_equations/spainv_iterations.pdf}
+  \end{center}
+  \caption{\textbf{Number of MCSA iterations required to converge the
+      second eigenvalue iteration for the fuel assembly problem with
+      SPAINV preconditioning as a function of SPAINV threshold.}
+    \textit{Each colored curve represents the iteration behavior for a
+      different SPAINV level pattern. Levels of 3, 4, 5, 6, and 7 were
+      used.}}
+  \label{fig:spainv_iterations}
+\end{figure}
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/spn_equations/spainv_size.pdf}
+  \end{center}
+  \caption{\textbf{Maximum number of non-zero entries observed for all
+      rows in the composite linear operator for the fuel assembly
+      problem with SPAINV preconditioning given as a function of
+      SPAINV threshold.}  \textit{Each colored curve represents the
+      row size for a different SPAINV level pattern. Levels of 3, 4,
+      5, 6, and 7 were used.}}
+  \label{fig:spainv_size}
+\end{figure}
+
+With the comparable iterative performance and improved sparsity,
+SPAINV preconditioning should then have a more favorable quality
+metric than ILUT preconditioning. Figure~\ref{fig:spainv_quality}
+gives the quality metric as a function of the threshold value for each
+of the sparsity levels computed. Not only do we see an improved
+quality metric over ILUT preconditioning (about an order of magnitude
+less), but we also note that the ideal sparsity level is not the largest
+nor the smallest. In fact, the largest and smallest levels (3 and 7)
+performed the worst in terms of the quality metric with a level of 4
+performing the best. When CPU time to generate the preconditioner is
+considered, 4 is the clear choice for this problem as more time is
+required to do the approximate inversion as the sparsity pattern of
+the preconditioner grows in size.
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/spn_equations/spainv_quality.pdf}
+  \end{center}
+  \caption{\textbf{SPAINV preconditioning quality metric for the fuel
+      assembly problem given as a function of SPAINV threshold.}
+    \textit{Each colored curve represents the quality metric behavior
+      for a different SPAINV level pattern. Levels of 3, 4, 5, 6, and
+      7 were used.}}
+  \label{fig:spainv_quality}
+\end{figure}
+
 \clearpage
@@
 sparsity, ILUT preconditioning may be improved through an additional
-application of SPAINV \citep{saad_iterative_2003}. For the ILUT
-preconditioning case we have the following linear problem to solve for
-the substituted variable:
+application of SPAINV (SPAINV-ILUT)\citep{saad_iterative_2003}. For
+the ILUT preconditioning case we have the following linear problem to
+solve for the substituted variable:
 \begin{equation}
@@
 good iterative performance.
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/spn_equations/psilut_iterations.pdf}
+  \end{center}
+  \caption{\textbf{Number of MCSA iterations required to converge the
+      second eigenvalue iteration for the fuel assembly problem with
+      SPAINV-ILUT preconditioning as a function of ILUT drop
+      tolerance.} \textit{Each colored curve represents the row size
+      for a different ILUT fill level. Fill levels of 1.5, 2.0, 3.0,
+      4.0, and 5.0 were used.}}
+  \label{fig:spainv_ilut_iterations}
+\end{figure}
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/spn_equations/psilut_size.pdf}
+  \end{center}
+  \caption{\textbf{Maximum number of non-zero entries observed for all
+      rows in the composite linear operator for the fuel assembly
+      problem with SPAINV-ILUT preconditioning given as a function of
+      ILUT drop tolerance.}  \textit{Each colored curve represents the
+      row size for a different ILUT fill level. Fill levels of 1.5,
+      2.0, 3.0, 4.0 and 5.0 were used.}}
+ \label{fig:spainv_ilut_size}
+\end{figure}
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/spn_equations/psilut_quality.pdf}
+  \end{center}
+  \caption{\textbf{SPAINV-ILUT preconditioning quality metric for the
+      fuel assembly problem given as a function of ILUT drop
+      tolerance.}  \textit{Each colored curve represents the row size
+      for a different ILUT fill level. Fill levels of 1.5, 2.0, 3.0,
+      4.0 and 5.0 were used.}}
+  \label{fig:spainv_ilut_quality}
+\end{figure}
 

### 2013-05-21 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding reduced domain approximation data for spainv and ilut

@@
 
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/spn_equations/ilut_rda_iterations.pdf}
+  \end{center}
+  \caption{\textbf{Number of MCSA iterations required to converge the
+      second eigenvalue iteration for the fuel assembly problem with
+      ILUT preconditioning as a function of reduced domain
+      approximation fill level.}  \textit{The largest fill level
+      presented is that using the ILUT parameters that gave the best
+      results with the approximation.}}
+  \label{fig:ilut_rda_iterations}
+\end{figure}
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/spn_equations/ilut_rda_quality.pdf}
+  \end{center}
+  \caption{\textbf{ILUT preconditioning quality metric for the fuel
+      assembly problem given as a function of reduced domain
+      approximation fill level.}  \textit{The largest fill level
+      presented is that using the ILUT parameters that gave the best
+      results with the approximation.}}
+  \label{fig:ilut_rda_quality}
+\end{figure}
+
 \clearpage
@@
 performance with a relative invariance to the threshold value. The
-threshold did have a significant effect on the time required to
-generate the preconditioner with a threshold of 0.001 over an order of
-magnitude slower than a threshold of 0.01 due to the inclusion of a
+threshold did have a significant effect, however, on the time required
+to generate the preconditioner with a threshold of 0.001 over an order
+of magnitude slower than a threshold of 0.01 due to the inclusion of a
 significant number of extra values in the input operator. In addition
@@
 
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/spn_equations/spainv_rda_iterations.pdf}
+  \end{center}
+  \caption{\textbf{Number of MCSA iterations required to converge the
+      second eigenvalue iteration for the fuel assembly problem with
+      SPAINV preconditioning as a function of reduced domain
+      approximation fill level.}  \textit{The largest fill level
+      presented is that using the SPAINV parameters that gave the best
+      results with the approximation.}}
+  \label{fig:spainv_rda_iterations}
+\end{figure}
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/spn_equations/spainv_rda_quality.pdf}
+  \end{center}
+  \caption{\textbf{SPAINV preconditioning quality metric for the fuel
+      assembly problem given as a function of reduced domain
+      approximation fill level.}  \textit{The largest fill level
+      presented is that using the SPAINV parameters that gave the best
+      results with the approximation.}}
+  \label{fig:spainv_rda_quality}
+\end{figure}
+
 \clearpage
@@
 
+MCSA preconditioned with SPAINV-ILUT was also used to solve the same
+fuel assembly problem as with the other preconditioners for the same
+1-group $SP_1$ discretization. To study convergence sensitivity to
+SPAINV parameters, the ILUT drop tolerance and ILUT fill level were
+parametrically varied with the number of iterations to converge the
+second eigenvalue iteration for the fuel assembly problem recorded
+along with the maximum number of non-zero entries observed in all
+matrix rows for the left/right preconditioned composite linear
+operator. For each calculation, the number of iterations required to
+converge reported was for the first eigenvalue iteration with
+\sn{3}{4} histories at each MCSA iteration to compute the Monte Carlo
+correction using the adjoint collision estimator. All SPAINV-ILUT
+calculations reported here were performed on a single
+CPU.
+
+Figure~\ref{fig:spainv_ilut_iterations} gives the number of iterations
+required to converge as a function of ILUT drop tolerance for ILUT
+fill levels of 1.5, 2.0, 3.0, 4.0 and 5.0 with a fixed SPAINV
+threshold of 1.0 and sparsity level of 1. Threshold values and higher
+resorted in loss of convergence while larger sparsity levels resulted
+in composite operator that was too dense. At smaller levels of fill,
+the benefits of using an addition step of SPAINV preconditioning are
+readily observed with the number of iterations required to converge
+cut in half for large drop tolerances. For smaller drop tolerances and
+higher ILUT levels of fill, the additional step of SPAINV
+preconditioning offers marginal improvement to iterative
+performance. As shown in Figure~\ref{fig:spainv_ilut_size},
+because ILUT is used in the preconditioning sequence, sparsity is
+again loss with only marginally improved non-zero entry numbers due to
+the application of the SPAINV threshold. As a result, the quality
+metrics given by Figure~\ref{fig:spainv_ilut_quality} are only
+marginally better than those observed for ILUT preconditioning and
+much larger than the SPAINV quality metric values.
 \begin{figure}[t!]

### 2013-05-22 doc/dissertation/document/chapters/mc_background/mc_background.tex
adding to rda section for sppn equations

@@
 stochastic realization of multiple matrix-vector multiplies, the
-criteria we choose should be based on maintain those elements that
-will make the largest contributions to the multiplication. For each
-row in the preconditioned system, we can then apply the reduced domain
-approximation to eliminate those elements that either fall below a
-specified tolerance level, keep the $N$ largest elements in a row, or
-both giving:
+criteria we choose should be based on maintaining those elements that
+will make the largest contributions to the multiplication and
+therefore recover as much of the original product as possible. For
+each row in the preconditioned system, we can then apply the reduced
+domain approximation to eliminate those elements that either fall
+below a specified tolerance level, keep the $N$ largest elements in
+each row, or both giving:
 \begin{subequations}
@@
 iteration matrix should be retained and continue to accelerate the
-iterative scheme.
+iterative scheme. The effects of the reduced domain approximation on
+systems where it is required due to the explicit preconditioning
+strategy chosen will be presented in Chapter~\ref{ch:spn_equations}.
 

### 2013-05-22 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding to rda section for sppn equations

@@
 
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/ilut_rda_iterations.pdf}
-  \end{center}
-  \caption{\textbf{Number of MCSA iterations required to converge the
-      second eigenvalue iteration for the fuel assembly problem with
-      ILUT preconditioning as a function of reduced domain
-      approximation fill level.}  \textit{The largest fill level
-      presented is that using the ILUT parameters that gave the best
-      results with the approximation.}}
-  \label{fig:ilut_rda_iterations}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/ilut_rda_quality.pdf}
-  \end{center}
-  \caption{\textbf{ILUT preconditioning quality metric for the fuel
-      assembly problem given as a function of reduced domain
-      approximation fill level.}  \textit{The largest fill level
-      presented is that using the ILUT parameters that gave the best
-      results with the approximation.}}
-  \label{fig:ilut_rda_quality}
-\end{figure}
-
 \clearpage
@@
 
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/spainv_rda_iterations.pdf}
-  \end{center}
-  \caption{\textbf{Number of MCSA iterations required to converge the
-      second eigenvalue iteration for the fuel assembly problem with
-      SPAINV preconditioning as a function of reduced domain
-      approximation fill level.}  \textit{The largest fill level
-      presented is that using the SPAINV parameters that gave the best
-      results with the approximation.}}
-  \label{fig:spainv_rda_iterations}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/spainv_rda_quality.pdf}
-  \end{center}
-  \caption{\textbf{SPAINV preconditioning quality metric for the fuel
-      assembly problem given as a function of reduced domain
-      approximation fill level.}  \textit{The largest fill level
-      presented is that using the SPAINV parameters that gave the best
-      results with the approximation.}}
-  \label{fig:spainv_rda_quality}
-\end{figure}
-
 \clearpage
@@
 
+\subsection{Applying the Reduced Domain Approximation}
+\label{subsec:spn_prec_rda}
+For each preconditioning technique presented, although convergence was
+achieved a primary concern is the number of non-zero states in each
+row of the system generated by the explicit preconditioning
+strategy. In many cases, orders of magnitude more matrix elements were
+generated resulting in poor scalability for domain decomposed
+algorithms. As outlined in
+\S~\ref{subsec:reduced_domain_approximation}, the reduced domain
+approximation may be used as a mechanism to potentially alleviate this
+problem by filtering elements of the composite matrix in each row that
+fall below a certain threshold value or by maintaining the largest $N$
+elements in each row where $N$ is a designated fill level.
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/spn_equations/rda_iterations.pdf}
+  \end{center}
+  \caption{\textbf{Number of MCSA iterations required to converge the
+      second eigenvalue iteration for the fuel assembly problem with
+      preconditioning as a function of reduced domain approximation
+      fill level.}  \textit{The largest fill level for each
+      preconditioning presented is that using the parameters that gave
+      the best results without the approximation.}}
+  \label{fig:rda_iterations}
+\end{figure}
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/spn_equations/rda_quality.pdf}
+  \end{center}
+  \caption{\textbf{Preconditioning quality metric for the fuel
+      assembly problem given as a function of reduced domain
+      approximation fill level.}  \textit{The largest fill level for
+      each preconditioning presented is that using the parameters that
+      gave the best results without the approximation.}}
+  \label{fig:rda_quality}
+\end{figure}
+
+\clearpage
+
 \subsection{MCSA Relaxation Parameters}

### 2013-05-23 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding more rda analysis

@@
 For the fuel assembly criticality problem presented in the previous
-section, the spectral radius was near one and in a regime in which
-MCSA breakdown is observed. In this regime the number of stochastic
-histories required to converge MCSA increases rapidly and the
-resulting poor performance is compounded by those histories being
-increasingly expensive to compute. To overcome this difficulty,
-advanced preconditioning strategies for the $SP_N$ problem are
-required beyond simple Jacobi methods that can reduce the spectral
+section, the spectral radius was near one using Jacobi preconditioning
+and in a regime in which MCSA breakdown is observed. In this regime
+the number of stochastic histories required to converge MCSA increases
+rapidly and the resulting poor performance is compounded by those
+histories being increasingly expensive to compute. To overcome this
+difficulty, advanced preconditioning strategies for the $SP_N$ problem
+are required beyond simple Jacobi methods that can reduce the spectral
 radius into a region of better MCSA performance. Several modern
@@
 triangular factors. For each calculation, the number of iterations
-required to converge reported was for the first eigenvalue iteration
+required to converge reported was for the secon deigenvalue iteration
 with \sn{3}{4} histories at each MCSA iteration to compute the Monte
@@
 operators with $O(1,000)$ elements in a row for all combinations of
-fill level and drop tolerance, over 10\% of the row size for this
-particular. This massive number of row entries, observed for a
-significant fraction of rows in the system, creates several
+fill level and drop tolerance, over 10\% of the total row size for
+this particular problem. This large number of row entries, observed
+for a significant fraction of rows in the system, creates several
 problems. First, sparsity is completely destroyed with each state in
@@
 \begin{equation}
-  \text{Quality Metric} = (\text{\# iterations})*(\text{maximum \# of
-    non-zero values})\:,
+  \text{Quality Metric} = (\text{\# iterations}) \times (\text{maximum
+    \# of non-zero values})\:,
 \end{equation}
@@
 requiring modification of the system in order to achieve better
-scalability. Sparse approximate inverse (SPAINV) preconditioning is a
-technique that may potentially alleviate both of these constraints at
-the cost of reduced iterative performance. The method produces a
-sparse approximation of the inverse matrix directly by minimizing the
-Frobenius norm of the residual matrix \citep{saad_iterative_2003}:
+scalability and Monte Carlo performance. Sparse approximate inverse
+(SPAINV) preconditioning is a technique that may potentially alleviate
+both of these constraints at the cost of reduced iterative
+performance. The method produces a sparse approximation of the inverse
+matrix directly by minimizing the Frobenius norm of the residual
+matrix \citep{saad_iterative_2003}:
 \begin{equation}
@@
 preconditioned composite linear operator. For each calculation, the
-number of iterations required to converge reported was for the first
-eigenvalue iteration with \sn{3}{4} histories at each MCSA iteration
-to compute the Monte Carlo correction using the adjoint collision
-estimator. All SPAINV calculations reported here were performed on a
-single CPU.
+number of iterations required to converge reported was for the second
+eigenvalue iteration also with \sn{3}{4} histories at each MCSA
+iteration to compute the Monte Carlo correction using the adjoint
+collision estimator. All SPAINV calculations reported here were
+performed on a single CPU.
 
@@
 \label{subsec:spn_spainv_preconditioning}
-For improved convergence over SPAINV preconditioning while maintaining
-sparsity, ILUT preconditioning may be improved through an additional
-application of SPAINV (SPAINV-ILUT)\citep{saad_iterative_2003}. For
-the ILUT preconditioning case we have the following linear problem to
-solve for the substituted variable:
+As a means to improve convergence over using only SPAINV
+preconditioning, ILUT preconditioning may be improved through an
+additional application of SPAINV
+(SPAINV-ILUT)\citep{saad_iterative_2003}. For the ILUT preconditioning
+case we have the following linear problem to solve for the substituted
+variable:
 \begin{equation}
@@
 Doing this permits a smaller fill level to be used with ILUT and a
-lower-order sparsity pattern to be used with SPAINV, giving more
-sparsity in the resulting composite linear operator while maintaining
-good iterative performance.
+lower-order sparsity pattern to be used with SPAINV, giving marginally
+more sparsity in the resulting composite linear operator while
+maintaining good iterative performance.
 
@@
 1-group $SP_1$ discretization. To study convergence sensitivity to
-SPAINV parameters, the ILUT drop tolerance and ILUT fill level were
-parametrically varied with the number of iterations to converge the
-second eigenvalue iteration for the fuel assembly problem recorded
+SPAINV-ILUT parameters, the ILUT drop tolerance and ILUT fill level
+were parametrically varied with the number of iterations to converge
+the second eigenvalue iteration for the fuel assembly problem recorded
 along with the maximum number of non-zero entries observed in all
 matrix rows for the left/right preconditioned composite linear
-operator. For each calculation, the number of iterations required to
-converge reported was for the first eigenvalue iteration with
-\sn{3}{4} histories at each MCSA iteration to compute the Monte Carlo
-correction using the adjoint collision estimator. All SPAINV-ILUT
-calculations reported here were performed on a single
-CPU.
+operator. SPAINV parameters for the data presented here were fixed as
+it was found that the preconditioning quality was largely insenstive
+to their variation. In addition, the fact that SPAINV is being used to
+preconditioned a problem already preconditioned with ILUT means that
+the input matrix sparsity pattern is already somewhat dense and large
+SPAINV levels would result in an even denser system. Therefore, a
+smaller sparsity pattern level will be used for this
+preconditioning. For each calculation, the number of iterations
+required to converge reported was for the second eigenvalue iteration.
+Again, \sn{3}{4} histories are used at each MCSA iteration to compute
+the Monte Carlo correction using the adjoint collision estimator. All
+SPAINV-ILUT calculations reported here were performed on a single CPU.
 
@@
 fill levels of 1.5, 2.0, 3.0, 4.0 and 5.0 with a fixed SPAINV
-threshold of 1.0 and sparsity level of 1. Threshold values and higher
-resorted in loss of convergence while larger sparsity levels resulted
-in composite operator that was too dense. At smaller levels of fill,
-the benefits of using an addition step of SPAINV preconditioning are
-readily observed with the number of iterations required to converge
-cut in half for large drop tolerances. For smaller drop tolerances and
-higher ILUT levels of fill, the additional step of SPAINV
-preconditioning offers marginal improvement to iterative
-performance. As shown in Figure~\ref{fig:spainv_ilut_size},
+threshold of 1.0 and sparsity level of 1. Threshold values higher than
+1.0 resulted in a loss of convergence and lower values did not
+significantly affect performance. Larger sparsity levels resulted in a
+composite operator that was too dense and a prohibitive amount of CPU
+time required to compute the preconditioner. At smaller levels of
+fill, the benefits of using an additional step of SPAINV
+preconditioning are readily observed with the number of iterations
+required to converge cut in half for large drop tolerances. For
+smaller drop tolerances and higher ILUT levels of fill, the additional
+step of SPAINV preconditioning offers marginal improvement to
+iterative performance. As shown in Figure~\ref{fig:spainv_ilut_size},
 because ILUT is used in the preconditioning sequence, sparsity is
-again loss with only marginally improved non-zero entry numbers due to
+again lost with only marginally improved non-zero entry numbers due to
 the application of the SPAINV threshold. As a result, the quality
@@
 \label{subsec:spn_prec_rda}
-For each preconditioning technique presented, although convergence was
-achieved a primary concern is the number of non-zero states in each
-row of the system generated by the explicit preconditioning
-strategy. In many cases, orders of magnitude more matrix elements were
-generated resulting in poor scalability for domain decomposed
-algorithms. As outlined in
+For each preconditioning technique presented convergence was achieved
+for the single fuel assembly problem. However, a primary concern is
+the number of non-zero states in each row of the system generated by
+the explicit preconditioning strategy. In many cases, orders of
+magnitude more matrix elements were generated resulting in poor
+scalability for domain decomposed algorithms and overall performance
+issues for Monte Carlo. As outlined in
 \S~\ref{subsec:reduced_domain_approximation}, the reduced domain
@@
 
+For the ILUT, SPAINV, and SPAINV-ILUT preconditioning strategies the
+reduced domain approximation will be applied to reduce the density of
+the composite linear operator to more manageable levels. For each
+preconditioner, the parameters that achieved the best quality metric
+results from the previous analysis were used with the reduced domain
+threshold was set to \sn{1}{-10} in order to eliminate any exceedingly
+small values from the matrix (generally this is simply removing
+non-zero elements within the floating point tolerance of zero). The
+reduced domain fill level was then varied, starting with the largest
+non-zero entries per row value observed for each of the
+preconditioning types in order to assess its effects relative to the
+case where no reduced domain approximation was applied.
+
+Figure~\ref{fig:rda_iterations} gives the number of iterations
+required to converge for each preconditioner type as a function of
+reduced domain fill level. Figure~\ref{fig:rda_quality} gives the
+corresponding quality metric for each data point where the number of
+non-zero entries used to compute the metric is equivalent to the
+reduced domain fill level. We first note the SPAINV preconditioning
+alone is signficantly more sensitive to the reduction in domain size
+over the ILUT-based methods, although the preconditioner was of
+$O(100)$ non-zero entries per row without any approximation
+applied. For the ILUT-based methods, performance was significantly
+better with converge achieved in less than 40 MCSA iterations with
+only 10 non-zero entries in each row (vs. 7 non-zero entries for the
+case with no preconditioning). SPAINV-ILUT iterative performance was
+marginally better than ILUT alone for all reduced domain fill
+levels. However, it should be noted that the ILUT level of fill used
+for the ILUT only calculations was set to 4 for this case while the
+SPAINV-ILUT preconditioner used an ILUT level of fill of 5 and
+therefore the marginally better performance is more likely a result of
+this addition of fill rather than the extra SPAINV
+preconditioning. Looking at the quality metric data, we see a nice
+power-law reduction in the quality metric as a function of the reduced
+domain fill level, achieving 2 orders of magnitude reduction in the
+quality metric for the ILUT-based preconditioning methods.
+
+Although applying the reduced domain approximation results in a
+successful recovery of sparsity for the Monte Carlo problem while
+maintaining good convergence properties, there is still an issue of
+forming the composite operator before applying the approximation and
+potentially generating the transpose of this operator in the case of
+the adjoint Monte Carlo method. Because of this, memory and scaling
+issues may still be observed when building the Monte Carlo problem. In
+addition, the expensive extraction of the inverse of the
+preconditioning operators for the explicit scheme creates a
+significant cost in overall performance. Future work in this area
+should be considerate of these important components of the algorithm.
+
 \begin{figure}[t!]
@@
       second eigenvalue iteration for the fuel assembly problem with
-      preconditioning as a function of reduced domain approximation
-      fill level.}  \textit{The largest fill level for each
-      preconditioning presented is that using the parameters that gave
-      the best results without the approximation.}}
+      each preconditioning as a function of reduced domain
+      approximation fill level.}  \textit{The largest fill level for
+      each preconditioning presented is that using the parameters that
+      gave the best results without the approximation.}}
   \label{fig:rda_iterations}
@@
 \end{equation}
-where $\omega$ is the relaxation parameter. Analogously, these
-relaxation parameter techniques can be applied to Monte Carlo to
-improve convergence as demonstrated by Dimov
-\citep{dimov_new_1998}. In this case, building an iteration matrix
-from Eq~(\ref{eq:richardson_relaxation} gives:
+where $\omega$ is the relaxation parameter. This is effectively a form
+of preconditioning where the system is being scaled on the left by a
+constant value in all rows. Analogously, these relaxation parameter
+techniques can be applied to Monte Carlo to improve convergence as
+demonstrated by Dimov \citep{dimov_new_1998}. In this case, building
+an iteration matrix from Eq~(\ref{eq:richardson_relaxation}) gives:
 \begin{equation}
@@
 by the same scalar parameter. For MCSA, we can stage this scheme with
-two separate relaxation parameters: one for the outer Richardson
+two separate relaxation parameters; one for the outer Richardson
 iteration and one for the inner Monte Carlo solve:

### 2013-05-23 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding solver comparison outputs for mcsa

@@
 preconditioner, the parameters that achieved the best quality metric
-results from the previous analysis were used with the reduced domain
-threshold was set to \sn{1}{-10} in order to eliminate any exceedingly
-small values from the matrix (generally this is simply removing
-non-zero elements within the floating point tolerance of zero). The
-reduced domain fill level was then varied, starting with the largest
-non-zero entries per row value observed for each of the
-preconditioning types in order to assess its effects relative to the
-case where no reduced domain approximation was applied.
+results from the previous analysis were used with ILUT parameters of a
+fill level of 5.0 and a drop tolerance of \sn{1}{-5}, SPAINV
+parameters of a sparsity level of 4 and a threshold of 0.1 and
+SPAINV-ILUT with ILUT parameters of a fill level of 4.0 and a drop
+tolerance of \sn{1}{-5} and SPAINV parameters of a sparsity level of 1
+and a threshold of 1.0. The reduced domain threshold was set to
+\sn{1}{-10} in order to eliminate any exceedingly small values from
+the matrix (generally this is simply removing non-zero elements within
+the floating point tolerance of zero). The reduced domain fill level
+was then varied, starting with the largest non-zero entries per row
+value observed for each of the preconditioning types in order to
+assess its effects relative to the case where no reduced domain
+approximation was applied.
 

### 2013-05-24 doc/dissertation/document/chapters/mc_background/mc_background.tex
adding solver comparison data to spn chapter

@@
 
+\subsection{Alternative Fixed Point Iterations}
+\label{subsec:mcsa_fp_alternatives}
+
 %%---------------------------------------------------------------------------%%
 \section{Monte Carlo Method Selection}
+\label{sec:mc_method_selection}p
 The MCSA method defined in Eq.~(\ref{eq:mcsa}) uses the adjoint method

### 2013-05-24 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding solver comparison data to spn chapter

@@
 \label{sec:spn_comparison}
+
+\subsection{Results}
+\label{subsec:spn_comparison_results}
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/spn_equations/solver_iters.pdf}
+  \end{center}
+  \caption{\textbf{Average number of iterations required to converge
+      each eigenvalue iteration for the fuel assembly problem as a
+      function of energy groups.} \textit{All methods were
+      preconditioned with identical ILUT paramters. For MCSA, several
+      variations of fixed point iterations and estimators were used.}}
+  \label{fig:comparison_iterations}
+\end{figure}
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/spn_equations/solver_time.pdf}
+  \end{center}
+  \caption{\textbf{Average CPU time per iteration in seconds for the
+      fuel assembly problem as a function of energy groups.}
+    \textit{All linear solver iterations over all eigenvalue
+      iterations were used to compute the averge. All methods were
+      preconditioned with identical ILUT paramters. For MCSA, several
+      variations of fixed point iterations and estimators were used.}}
+  \label{fig:comparison_time}
+\end{figure}
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/spn_equations/solver_p_time.pdf}
+  \end{center}
+  \caption{\textbf{Average CPU time per iteration in seconds for the
+      fuel assembly problem as a function of energy groups with
+      preconditioning time included for the MCSA methods.}
+    \textit{All linear solver iterations over all eigenvalue
+      iterations were used to compute the averge. All methods were
+      preconditioned with identical ILUT paramters. For MCSA, several
+      variations of fixed point iterations and estimators were used.}}
+  \label{fig:comparison_prec_time}
+\end{figure}

### 2013-05-28 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
small updates to spn chapter

@@
 the azimuthal direction of travel and an isotropic source of neutrons
-is assumed. In addition, for fission systems the eigenvalue form of the transport
-equation is:
+is assumed. In addition, for fission systems the eigenvalue form of
+the transport equation is:
 \begin{multline}
@@
 where $\phi$ is the neutron flux, $\Sigma_a$ is the absorption cross
-subsection, and $S$ is the source of neutrons. In addition, $D$ is the
+section, and $S$ is the source of neutrons. In addition, $D$ is the
 diffusion coefficient defined as:
@@
 \end{equation}
-where $\Sigma_s$ is the scattering cross subsection, $\Sigma_t = \Sigma_a
-+ \Sigma_s$ is the total cross subsection, and $\bar{\mu}$ is the cosine
+where $\Sigma_s$ is the scattering cross section, $\Sigma_t = \Sigma_a
++ \Sigma_s$ is the total cross section, and $\bar{\mu}$ is the cosine
 of the average scattering angle. For simplicity, we will take
@@
 addition, to further simplify we will assume a homogeneous domain such
-that the cross subsections remain constant throughout. Doing this permits
+that the cross sections remain constant throughout. Doing this permits
 us to rewrite Eq~(\ref{eq:diffusion_eq}) as:
@@
 Incomplete lower-upper (ILU) factorizations of the linear operator can
-be used a simple mechanism to form an approximate inverse as a
+be used a simple mechanism to form an approximate inverse of a
 preconditioner. To build the factorization, the sparse upper and lower
@@
 sensitivity to ILUT parameters, the fill level and drop tolerance were
-parametrically varied with the number of iterations to converge the
-second eigenvalue iteration for the fuel assembly problem recorded
-along with the maximum number of non-zero entries observed in all
-matrix rows for the left/right preconditioned composite linear
-operator. To provide some sparsity to the factorization, the ILUT drop
-tolerance was used to drop elements in the extracted inverse
-triangular factors. For each calculation, the number of iterations
-required to converge reported was for the secon deigenvalue iteration
-with \sn{3}{4} histories at each MCSA iteration to compute the Monte
-Carlo correction using the adjoint collision estimator. All ILUT
-calculations reported here were performed on a single CPU and
-therefore this data does not take into account the aforementioned
-effects of parallel decomposition on the quality of the ILUT
-preconditioning.
+parametrically varied with the number of iterations to converge an
+eigenvalue iteration for the fuel assembly problem recorded along with
+the maximum number of non-zero entries observed in all matrix rows for
+the left/right preconditioned composite linear operator. To provide
+some sparsity to the factorization, the ILUT drop tolerance was used
+to drop elements in the extracted inverse triangular factors. For each
+calculation, the number of iterations required to converge reported
+was for a single eigenvalue iteration with \sn{3}{4} histories at each
+MCSA iteration to compute the Monte Carlo correction using the adjoint
+collision estimator. All ILUT calculations reported here were
+performed on a single CPU and therefore this data does not take into
+account the aforementioned effects of parallel decomposition on the
+quality of the ILUT preconditioning.
 
@@
   \end{center}
-  \caption{\textbf{Number of MCSA iterations required to converge the
-      second eigenvalue iteration for the fuel assembly problem with
-      ILUT preconditioning as a function of ILUT drop tolerance.}
+  \caption{\textbf{Number of MCSA iterations required to converge an
+      eigenvalue iteration for the fuel assembly problem with ILUT
+      preconditioning as a function of ILUT drop tolerance.}
     \textit{Each colored curve represents the iteration behavior for a
@@
 amongst states in the system and reduce the amount of memory used
-along with the potentially latency overhead. Given these two
-objectives, the following metric is proposed:
+along with the potential latency overhead. Given these two objectives,
+the following metric is proposed:
 \begin{equation}
@@
 It should be noted here that although only the behavior of the linear
-solve during the second eigenvalue iteration is reported, the behavior
+solve during a single eigenvalue iteration is reported, the behavior
 of the linear solver was observed to be consistent throughout the
-eigenvalue iterations (25 were required to converge the 1-group $SP_1$
-problem). We expect this as the linear operator (which is unchanging
-in the fuel assembly criticality problem) dictates the convergence of
-MCSA. At each iteration the fission source provided to the linear
-problem is changing, however, we expect the same convergence behavior
-for all source vectors $\mathbf{b} \in \mathbb{R}^{N}$ where
-$||\mathbf{b}||_1 \neq 0$.
+eigenvalue iterations (25 total eigenvalue iterations were required to
+converge the 1-group $SP_1$ problem). We expect this as the linear
+operator (which is unchanging in the fuel assembly criticality
+problem) dictates the convergence of MCSA. At each iteration the
+fission source provided to the linear problem is changing, however, we
+expect the same convergence behavior for all source vectors
+$\mathbf{b} \in \mathbb{R}^{N}$ where $||\mathbf{b}||_1 \neq 0$.
 
@@
 Explicit formation of the preconditioner inverse is a requirement of
-the chosen MCSA preconditioning strategy. As a result, although ILUT
+the current MCSA preconditioning strategy. As a result, although ILUT
 preconditioning provided excellent iterative performance, a
@@
 the sparsity pattern and threshold were parametrically varied with the
-number of iterations to converge the second eigenvalue iteration for
-the fuel assembly problem recorded along with the maximum number of
+number of iterations to converge a single eigenvalue iteration for the
+fuel assembly problem recorded along with the maximum number of
 non-zero entries observed in all matrix rows for the left/right
 preconditioned composite linear operator. For each calculation, the
-number of iterations required to converge reported was for the second
-eigenvalue iteration also with \sn{3}{4} histories at each MCSA
-iteration to compute the Monte Carlo correction using the adjoint
-collision estimator. All SPAINV calculations reported here were
-performed on a single CPU.
+number of iterations required to converge reported was for a single
+eigenvalue iteration with \sn{3}{4} histories at each MCSA iteration
+to compute the Monte Carlo correction using the adjoint collision
+estimator. All SPAINV calculations reported here were performed on a
+single CPU.
 
@@
   \end{center}
-  \caption{\textbf{Number of MCSA iterations required to converge the
-      second eigenvalue iteration for the fuel assembly problem with
+  \caption{\textbf{Number of MCSA iterations required to converge a
+      single eigenvalue iteration for the fuel assembly problem with
       SPAINV preconditioning as a function of SPAINV threshold.}
@@
 As a means to improve convergence over using only SPAINV
-preconditioning, ILUT preconditioning may be improved through an
+preconditioning, ILUT preconditioning may be modified through an
 additional application of SPAINV
@@
 SPAINV-ILUT parameters, the ILUT drop tolerance and ILUT fill level
-were parametrically varied with the number of iterations to converge
-the second eigenvalue iteration for the fuel assembly problem recorded
+were parametrically varied with the number of iterations to converge a
+single eigenvalue iteration for the fuel assembly problem recorded
 along with the maximum number of non-zero entries observed in all
@@
 SPAINV levels would result in an even denser system. Therefore, a
-smaller sparsity pattern level will be used for this
+smaller SPAINV sparsity pattern level will be used for this
 preconditioning. For each calculation, the number of iterations
-required to converge reported was for the second eigenvalue iteration.
+required to converge reported was for a single eigenvalue iteration.
 Again, \sn{3}{4} histories are used at each MCSA iteration to compute
@@
 fill levels of 1.5, 2.0, 3.0, 4.0 and 5.0 with a fixed SPAINV
-threshold of 1.0 and sparsity level of 1. Threshold values higher than
-1.0 resulted in a loss of convergence and lower values did not
-significantly affect performance. Larger sparsity levels resulted in a
-composite operator that was too dense and a prohibitive amount of CPU
-time required to compute the preconditioner. At smaller levels of
+threshold of 1.0 and sparsity level of 1. SPAINV threshold values
+higher than 1.0 resulted in a loss of convergence and lower values did
+not significantly affect performance. Larger sparsity levels resulted
+in a composite operator that was too dense and a prohibitive amount of
+CPU time required to compute the preconditioner. At smaller levels of
 fill, the benefits of using an additional step of SPAINV
@@
   \end{center}
-  \caption{\textbf{Number of MCSA iterations required to converge the
-      second eigenvalue iteration for the fuel assembly problem with
+  \caption{\textbf{Number of MCSA iterations required to converge a
+      single eigenvalue iteration for the fuel assembly problem with
       SPAINV-ILUT preconditioning as a function of ILUT drop
@@
   \end{center}
-  \caption{\textbf{Number of MCSA iterations required to converge the
-      second eigenvalue iteration for the fuel assembly problem with
+  \caption{\textbf{Number of MCSA iterations required to converge a
+      single eigenvalue iteration for the fuel assembly problem with
       each preconditioning as a function of reduced domain

### 2013-05-31 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
updating spn relaxation section and data

@@
       problem with SPAINV preconditioning given as a function of
-      SPAINV threshold.}  \textit{Each colored curve represents the
-      row size for a different SPAINV level pattern. Levels of 3, 4,
-      5, 6, and 7 were used.}}
+      SPAINV threshold.} \textit{Each colored curve represents the row
+      size for a different SPAINV level pattern. Levels of 3, 4, 5, 6,
+      and 7 were used.}}
   \label{fig:spainv_size}
@@
 preconditioner, the parameters that achieved the best quality metric
-results from the previous analysis were used with ILUT parameters of a
-fill level of 5.0 and a drop tolerance of \sn{1}{-5}, SPAINV
-parameters of a sparsity level of 4 and a threshold of 0.1 and
+results from the previous analysis were used. These correspond to ILUT
+parameters of a fill level of 5.0 and a drop tolerance of \sn{1}{-5},
+SPAINV parameters of a sparsity level of 4 and a threshold of 0.1 and
 SPAINV-ILUT with ILUT parameters of a fill level of 4.0 and a drop
@@
 applied. For the ILUT-based methods, performance was significantly
-better with converge achieved in less than 40 MCSA iterations with
+better with convergence achieved in less than 40 MCSA iterations with
 only 10 non-zero entries in each row (vs. 7 non-zero entries for the
@@
 the adjoint Monte Carlo method. Because of this, memory and scaling
-issues may still be observed when building the Monte Carlo problem. In
-addition, the expensive extraction of the inverse of the
-preconditioning operators for the explicit scheme creates a
-significant cost in overall performance. Future work in this area
-should be considerate of these important components of the algorithm.
+issues may still be observed when building the probability and weight
+matrices for the Monte Carlo problem. In addition, the expensive
+extraction of the inverse of the preconditioning operators for the
+explicit scheme creates a significant cost in overall
+performance. Future work in this area should be considerate of these
+important components of the preconditioned MCSA algorithm.
 
@@
 
+We apply these relaxation parameters to the fuel assembly problem
+along with the reduced domain approximation as a means of studying
+their effects. For each calculation presented, the number of
+iterations required to converge reported was for a single eigenvalue
+iteration.  Again, \sn{3}{4} histories are used at each MCSA iteration
+to compute the Monte Carlo correction using the adjoint collision
+estimator. A reduced domain fill level of 100 is used with a threshold
+of \sn{1}{-10} to filter small values.
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/spn_equations/rich_relax_iters.pdf}
+  \end{center}
+  \caption{\textbf{Number of iterations to converge a single
+      eigenvalue iteration of the fuel assembly problem as a function
+      of the Richardson relaxation parameter.} \textit{The Neumann
+      relaxation parameter was fixed at 1.0 and \sn{3}{4} stochastic
+      histories were used to compute the MCSA correction at each
+      iteration.}}
+  \label{fig:rich_relax_iters}
+\end{figure}
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/spn_equations/rich_relax_time.pdf}
+  \end{center}
+  \caption{\textbf{CPU time in seconds to converge a single eigenvalue
+      iteration of the fuel assembly problem as a function of the
+      Richardson relaxation parameter.} \textit{The Neumann relaxation
+      parameter was fixed at 1.0 and \sn{3}{4} stochastic histories
+      were used to compute the MCSA correction at each iteration.}}
+  \label{fig:rich_relax_time}
+\end{figure}
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/spn_equations/neumann_relax_iters.pdf}
+  \end{center}
+  \caption{\textbf{Number of iterations to converge a single
+      eigenvalue iteration of the fuel assembly problem as a function
+      of the Neumann relaxation parameter.} \textit{The Richardson
+      relaxation parameter was fixed at 1.0 and \sn{3}{4} stochastic
+      histories were used to compute the MCSA correction at each
+      iteration.}}
+  \label{fig:neumann_relax_iters}
+\end{figure}
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/spn_equations/neumann_relax_time.pdf}
+  \end{center}
+  \caption{\textbf{CPU time in seconds to converge a single eigenvalue
+      iteration of the fuel assembly problem as a function of the
+      Neumann relaxation parameter.} \textit{The Richardson relaxation
+      parameter was fixed at 1.0 and \sn{3}{4} stochastic histories
+      were used to compute the MCSA correction at each iteration.}}
+  \label{fig:neumann_relax_time}
+\end{figure}
+
 \clearpage
@@
       function of energy groups.} \textit{All methods were
-      preconditioned with identical ILUT paramters. For MCSA, several
+      preconditioned with identical ILUT parameters. For MCSA, several
       variations of fixed point iterations and estimators were used.}}
@@
       iterations were used to compute the averge. All methods were
-      preconditioned with identical ILUT paramters. For MCSA, several
+      preconditioned with identical ILUT parameters. For MCSA, several
       variations of fixed point iterations and estimators were used.}}

### 2013-05-31 doc/dissertation/document/backmatter/mg_spn_equations.tex
moving multigroup spn derivation to an appendix

--- 2013-05-28 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
@@
-\chapter{Monte Carlo Solution Methods for the Simplified $P_N$ Equations}
-\label{ch:spn_equations}
-
-The neutron transport problem is complicated. Solutions cover a large
-phase space and the problems of interest are often geometrically
-complex, very large, or both, requiring tremendous computational
-resources to generate an adequate solution. Modern deterministic
-methods for large scale problems are commonly variants on the discrete
-ordinates ($S_N$) method \citep{evans_denovo:_2010}. For full reactor
-core neutronics simulations, the $S_N$ method requires potentially
-trillions of unknown angular flux moments to be computed to achieve
-good accuracy for the responses of interest
-\citep{slaybaugh_acceleration_2011}. Other forms of the transport
-problem, including the $P_N$ method, take on a simpler form than the
-more common $S_N$ methods but lack in accuracy when compared while
-still requiring considerable computational resources for solutions in
-multiple dimensions.
-
-In the 1960's, Gelbard developed an ad-hoc multidimensional extension
-of the simple single dimension planar $P_N$ equations that created a
-system of coupled, diffusion-like equations known as the simplified
-$P_N$ ($SP_N$) equations. Up until around the 1990's, the $SP_N$
-method was either widely unknown, widely unused, or combination of
-both even though numerical studies showed promising results with
-better solutions than diffusion theory and a significant reduction in
-computational time over more accurate methods such as discrete
-ordinates. Why did this happen? A significant problem, pointed out by
-Brantley and Larsen \citep{brantley_simplified_2000}, was that little
-rigor had been applied to the formulation of the $SP_N$ equations
-since their derivation through primarily heuristic arguments. Instead,
-studies at that time focused on simply comparing the results of the
-method to other contemporary transport solution strategies. In
-addition, many problems of interest from the literature at the time
-were either solved using nodal-type methods for reactor-sized problems
-or $S_N$-type methods for benchmark problems with intricate material
-configurations and potentially large flux gradients over small spatial
-domains.
-
-So why reconsider the $SP_N$ equations? Starting in the 1990's and
-primarily due to Larsen and his colleagues, the $SP_N$ equations have
-been given a more rigorous treatment with both variational and
-asymptotic derivations performed as a means of verification. In
-addition, these equations have been more rigorously studied as
-solution methods to MOX fuel problems and have been shown to provide
-accurate solutions. With this mathematical literature to provide a
-solid numerical footing for the method, we look at its application to
-today's challenge problems in full reactor core transport. The
-reduction in numerical complexity of current full core deterministic
-solution methods using the $S_N$ approximation could mean significant
-savings in both compute time and memory required. In addition, the
-characteristics of the solution to the transport problem for a steady
-state reactor core permit diffusion theory to be used; a staple of the
-nuclear industry since its inception. Therefore, if diffusion theory
-is applicable, then finer grained solutions that capture more of the
-physics contained in the transport equation should be possible with
-the $SP_N$ method. In doing so, we also expect from the literature to
-obtain computed responses on the order of accuracy we would expect
-from an appropriately discretized $S_N$ method at a fraction of the
-cost.
-
-To further motivate moving in this direction, recent developments in
-the Exnihilo neutronics package at Oak Ridge National Laboratory have
-permitted generation of the $SP_N$ system of equations for detailed
-full core reactor models \citep{evans_simpli_2013}. By fully forming
-these equations and formulating them as a linear algebra problem
-instead of using the explicit iterative methods of the past, we now
-have access to all of the modern advancements in computational linear
-algebra including Krylov solvers for asymmetric systems and
-preconditioning methods such as algebraic multigrid. This leads us to
-then explore the applicability of our work in discrete Monte Carlo
-methods for linear systems as a possible solution method for the
-$SP_N$ equations. If formulated correctly, we hypothesize that
-significant improvement in usage of computational resources may be
-observed compared to modern solution techniques such as those
-suggested due to the form of the matrices generated by the
-discretization. In addition, solving the $SP_N$ equations in this way
-also breaks away from the $S_N$ forms of parallelism where spatial
-parallelism is achieved by an efficient parallel sweep, angular
-efficiency achieved by pipe-lining, and energy parallelism achieved by
-decoupling the groups. With the $SP_N$ equations as a full matrix
-system, we now can parallelize the problem as prescribed by the linear
-solver, which may be significantly more scalable than current $S_N$
-transport practices.
-
-In this chapter, we derive the $SP_N$ equations, closely following the
-work of Evans\footnote{The $SP_N$ derivations in this chapter are
-  heavily based on those presented by Evans in
-  \citep{evans_simpli_2013}}, in order to gain full understanding of
-the underlying system and its potential behavior in a Monte Carlo
-context. From the $P_N$ equations, we apply a set of approximations to
-yield the multi-dimensional, multi-group $SP_N$ equations for fixed
-source problems. Using the fully-formed matrix for the transport
-problem, we explore solutions to the $SP_N$ equations using Monte
-Carlo Synthetic Acceleration using a light water reactor fuel assembly
-criticality calculation as the driving problem.
-
-%%---------------------------------------------------------------------------%%
-\section{The Neutron Transport Equation}
-\label{sec:transport_eq}
-As a starting point we define the time-independent neutron transport
-equation \citep{lewis_computational_1993}:
-\begin{multline}
-  \hat{\Omega} \cdot \vec{\nabla} \psi(\vec{r},\hat{\Omega},E) +
-  \sigma(\vec{r},E) \psi(\vec{r},\hat{\Omega},E) = \\ \iint
-  \sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}' \cdot \hat{\Omega})
-  \psi(\vec{r},\hat{\Omega}',E') d\Omega' dE' +
-  q(\vec{r},\hat{\Omega},E)\:,
-  \label{eq:general_transport}
-\end{multline}
-with the variables defined as:
-\begin{itemize}
-\item $\vec{r}$ - neutron spatial position
-\item $\hat{\Omega}$ - neutron streaming direction with radial
-  component $\mu$ and azimuthal component $\omega$
-\item $\hat{\Omega}' \cdot \hat{\Omega} = \mu_0$ is the angle of
-  scattering
-\item $E$ - neutron energy
-\item $\psi(\vec{r},\hat{\Omega},E)$ - angular flux
-\item $\sigma(\vec{r},E)$ - total interaction cross section
-\item $\sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}')$ - probability
-  of scattering from direction $\hat{\Omega}'$ into an angular domain
-  $d\hat{\Omega}'$ about the direction $\hat{\Omega}$ and from energy
-  $E'$ to an energy domain $dE'$ about energy $E$
-\item $q(\vec{r},\hat{\Omega},E)$ - external source of neutrons.
-\end{itemize}
-For this work, it is sufficient to formulate
-Eq~(\ref{eq:general_transport}) in 1-dimensional Cartesian geometry:
-\begin{multline}
-  \mu \frac{\partial}{\partial x} \psi(x,\mu,E) + \sigma(x,E)
-  \psi(x,\mu,E) = \\ \iint \sigma_s(x,E' \rightarrow E,\hat{\Omega}'
-  \cdot \hat{\Omega}) \psi(x,\hat{\Omega}',E') d\Omega' dE' +
-  \frac{q(x,E)}{4 \pi}\:,
-  \label{eq:cart_1d_transport}
-\end{multline}
-where the angular component of the solution is no longer dependent on
-the azimuthal direction of travel and an isotropic source of neutrons
-is assumed. In addition, for fission systems the eigenvalue form of
-the transport equation is:
-\begin{multline}
-  \hat{\Omega} \cdot \vec{\nabla} \psi(\vec{r},\hat{\Omega},E) +
-  \sigma(\vec{r},E) \psi(\vec{r},\hat{\Omega},E) = \\ \iint
-  \sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}' \cdot \hat{\Omega})
-  \psi(\vec{r},\hat{\Omega}',E') d\Omega' dE' + \\ \frac{1}{k} \chi(E)
-  \iint \nu \sigma_f(\hat{r},E') \psi(\vec{r},\hat{\Omega}',E')
-  d\Omega' dE' + q(\vec{r},\hat{\Omega},E) \:,
-  \label{eq:eigenvalue_transport}
-\end{multline}
-with the additional variables defined as
-\begin{itemize}
-\item $k$ - multiplication factor
-\item $\chi(E)$ - fission neutron energy spectrum
-\item $\nu$ - average number of neutrons per fission
-\item $\sigma_f(r,E')$ - fission cross section\:.
-\end{itemize}
-In 1-dimensional Cartesian geometry,
-Eq~(\ref{eq:eigenvalue_transport}) becomes:
-\begin{multline}
-  \mu \frac{\partial}{\partial x} \psi(x,\mu,E) + \sigma(x,E)
-  \psi(x,\mu,E) = \\ \iint \sigma_s(x,E' \rightarrow
-  E,\hat{\Omega}' \cdot \hat{\Omega}) \psi(x,\hat{\Omega}',E')
-  d\Omega' dE' + \\ \frac{1}{k} \chi(E)
-  \iint \nu \sigma_f(x,E') \psi(x,\hat{\Omega}',E') d\Omega'
-  dE' + \frac{q(x,E)}{4 \pi}\:.
-  \label{eq:cart_1d_eigenvalue}
-\end{multline}
-
-%%---------------------------------------------------------------------------%%
-\section{Derivation of the Monoenergetic $SP_N$ Equations}
-\label{sec:spn_equations}
-The $P_N$ equations as derived in Appendix~\ref{chap:pn_equations}
-give $N+1$ coupled first-order equations capturing the spatial and
-angular-dependence of the solution. In multiple dimensions, the
-equation set becomes large and coupled not only through angular
-moments but also through the spatial variables. As a simpler
-alternative to multidimensional $P_N$ solutions, Gelbard recognized in
-1960 that the planar $P_N$ equations could be simplified and applied
-an ad-hoc method to extend them to multiple dimensions, yielding the
-$SP_N$ equations. These equations are not only fewer in number, but
-also take on a diffusion-like form while maintaining the angular
-character of the flux, making them amenable to solutions with modern
-diffusion methods.
-
-First, the $P_N$ equations can be simplified to $(N+1)/2$ second-order
-equations by solving for the $n^{th}$ Legendre flux moment in the
-odd-order equations:
-\begin{equation}
-  \phi_n = \frac{1}{\Sigma_n}\Bigg[ q \delta_{no} -
-    \frac{\partial}{\partial x}\Big(\frac{n}{2n+1}\phi_{n-1} +
-    \frac{n+1}{2n+1} \phi_{n+1} \Big) \Bigg]\:, 
-  \label{eq:odd_moments}
-\end{equation}
-for $n = 1,3,\cdots,N$ and $\delta_{no} = 0\ \forall n \neq 0$. We can
-insert the odd moments into Eq~(\ref{eq:final_pn_equations}) to get a
-reduced group of equations for the even moments:
-\begin{multline}
-  -\frac{\partial}{\partial x}
-  \Bigg[\frac{n}{2n+1}\frac{1}{\Sigma_{n-1}} \frac{\partial}{\partial
-      x} \Big(\frac{n-1}{2n-1} \phi_{n-2} + \frac{n}{2n-1}\phi_n \Big)
-    \\+ \frac{n+1}{2n+1}\frac{1}{\Sigma_{n+1}} \frac{\partial}{\partial
-      x} \Big(\frac{n+1}{2n+3}\phi_n + \frac{n+2}{2n+3}\phi_{n+2}\Big)
-    \Bigg] \\+ \Sigma_n \phi_n = q \delta_{n0}\ \ \ \ \ \ \ \ \ n =
-  0,2,4,\cdots,N\:.
-  \label{eq:reduced_pn}
-\end{multline}
-Immediately, we note the diffusion-like nature of
-Eq~(\ref{eq:reduced_pn}) as compared to the original $P_N$
-equations. To extend these equations to multiple dimensions, Gelbard
-simply replaced the planar spatial derivatives in the reduced set of
-equations with general multidimensional gradient operators:
-\begin{multline}
-  -\nabla \cdot \Bigg[\frac{n}{2n+1}\frac{1}{\Sigma_{n-1}} \nabla
-    \Big(\frac{n-1}{2n-1} \phi_{n-2} + \frac{n}{2n-1}\phi_n \Big) \\+
-    \frac{n+1}{2n+1}\frac{1}{\Sigma_{n+1}} \nabla
-    \Big(\frac{n+1}{2n+3}\phi_n + \frac{n+2}{2n+3}\phi_{n+2}\Big)
-    \Bigg] \\+ \Sigma_n \phi_n = q \delta_{n0}\ \ \ \ \ \ \ \ \ n =
-  0,2,4,\cdots,N\:,
-  \label{eq:spn_equations}
-\end{multline}
-yielding a multidimensional set of $(N+1)/1$ angular coupled equations
-defined as the $SP_N$ equations. As with the $P_N$ equations, we
-provide closure to this set of equations with $\phi_{N+1} = 0$. As a
-concrete example, we will consider the $SP_7$ equations:
-\begin{subequations}
-  \begin{gather}
-    -\nabla \cdot \frac{1}{3 \Sigma_1} \nabla ( \phi_0 + 2\phi_2 ) +
-    \Sigma_0 \phi_0 = q \\ 
-    -\nabla \cdot \Bigg[ \frac{2}{15 \Sigma_1} \nabla ( \phi_0 + 2\phi_2
-      ) + \frac{3}{35 \Sigma_3}\nabla( 3\phi_2 + 4\phi_4)\Bigg] +
-    \Sigma_2 \phi_2 = 0\\
-    -\nabla \cdot \Bigg[ \frac{4}{63 \Sigma_3} \nabla ( 3\phi_2 +
-      4\phi_4 ) + \frac{5}{99 \Sigma_5}\nabla( 5\phi_4 +
-      6\phi_6)\Bigg] + \Sigma_4 \phi_4 = 0\\
-    -\nabla \cdot \Bigg[ \frac{6}{143 \Sigma_5} \nabla ( 5\phi_4 +
-      6\phi_6 ) + \frac{7}{195 \Sigma_7}\nabla(7\phi_6)\Bigg] +
-    \Sigma_6 \phi_6 = 0 \:.
-  \end{gather}
-  \label{eq:sp7_equations}
-\end{subequations}
-To further modify these equations, we can use a change of variables to
-create a new group of equations such that the gradients are operating
-on a single vector:
-\begin{subequations}
-  \begin{gather}
-    u_1 = \phi_0 + 2\phi_2 \\
-    u_2 = 3\phi_2 + 4\phi_4 \\
-    u_3 = 5\phi_4 + 6\phi_6 \\
-    u_4 = 7\phi_6 \:,
-  \end{gather}
-  \label{eq:spn7_subs}
-\end{subequations}
-or equivalently:
-\begin{subequations}
-  \begin{gather}
-    \phi_0 = u_1 - \frac{2}{3}u_2 + \frac{8}{15}u_3 -
-    \frac{16}{35}u_4 \\
-    \phi_2 = \frac{1}{3}u_2 - \frac{4}{15}u_3 + \frac{8}{35}u_4\\ 
-    \phi_4 = \frac{1}{5}u_3 - \frac{6}{35}u_4\\
-    \phi_6 = \frac{1}{7}u_4\:.
-  \end{gather}
-  \label{eq:spn7_subs_inverse}
-\end{subequations}
-When substituted into Eq~(\ref{eq:sp7_equations}), these terms give:
-\begin{subequations}
-  \begin{gather}
-    -\nabla \cdot \frac{1}{3 \Sigma_1} \nabla u_1 + \Sigma_0 \Bigg[
-    u_1 - \frac{2}{3}u_2 + \frac{8}{15}u_3 - \frac{16}{35}u_4 \Bigg]
-    = -q \\
-    -\nabla \cdot \Bigg[ \frac{2}{15 \Sigma_1} \nabla u_1 +
-    \frac{3}{35 \Sigma_3} \nabla u_2 \Bigg] + \Sigma_2 \Bigg[
-    \frac{1}{3}u_2 - \frac{4}{15}u_3 + \frac{8}{35}u_4 \Bigg] = 0 \\
-    -\nabla \cdot \Bigg[ \frac{4}{63 \Sigma_3} \nabla u_2 +
-    \frac{5}{99 \Sigma_5} \nabla u_3 \Bigg] + \Sigma_4 \Bigg[
-    \frac{1}{5}u_3 - \frac{6}{35}u_4 \Bigg] = 0 \\ 
-    -\nabla \cdot \Bigg[ \frac{6}{143 \Sigma_5} \nabla u_3 +
-    \frac{7}{195 \Sigma_7} \nabla u_4 \Bigg] + \Sigma_6 \Bigg[
-    \frac{1}{7}u_4 \Bigg] = 0 \:.
-  \end{gather}
-  \label{eq:spn7_subs_equations}
-\end{subequations}
-If we rearrange Eq~(\ref{eq:spn7_subs_equations}) such that only one
-divergence operation is present in each equation, we can formulate
-this as a matrix system of 4 equations in the case of the $SP_7$
-approximation:
-\begin{equation}
-  -\nabla \cdot D_n \nabla u_n + \sum_{m=1}^4 A_{nm} u_m =
-  q_n\ \ \ \ \ \ \ n = 1,2,3,4\:,
-  \label{eq:spn_matrix}
-\end{equation}
-with $\mathbf{u}$ the vector of solution variables:
-\begin{equation}
-  \mathbf{u} = ( u_1\ \ u_2\ \ u_3\ \ u_4 )^T \:,
-  \label{eq:spn7_solution_vector}
-\end{equation}
-$\mathbf{D}$ the vector of effective diffusion coefficients:
-\begin{equation}
-  \mathbf{D} = \Bigg( \frac{1}{3\Sigma_1}\ \ \frac{1}{7\Sigma_3}\ \
-  \frac{1}{11\Sigma_5}\ \ \frac{1}{15\Sigma_7} \Bigg)^T\:,
-  \label{eq:spn7_diffusion_coeffs}
-\end{equation}
-$\mathbf{q}$ the vector of source terms where the $0^{th}$ moment
-source has now been distributed through the system:
-\begin{equation}
-  \mathbf{q} = (
-  q\ \ -\frac{2}{3}q\ \ \frac{8}{15}q\ \ -\frac{16}{35}q )^T\:,
-  \label{eq:spn7_source_vector}
-\end{equation}
-and $\mathbf{A}$ a matrix of angular scattering terms:
-% NOTE: I copied the following matrix directly out of Tom's tech note
-% on the SPn equations which I am effectively following here because I
-% was feeling lazy. I have verified its correctness.
-\begin{equation}
-  \mathbf{A} = 
-  {\tiny \begin{bmatrix}
-    (\Sigma_0) &
-    (-\frac{2}{3}\Sigma_0) &
-    (\frac{8}{15}\Sigma_0) &
-    (-\frac{16}{35}\Sigma_0) \\
-    %%
-    &&&\\
-    %%
-    (-\frac{2}{3}\Sigma_0) &
-    (\frac{4}{9}\Sigma_0 + \frac{5}{9}\Sigma_2) &
-    (-\frac{16}{45}\Sigma_0 - \frac{4}{9}\Sigma_2) &
-    (\frac{32}{105}\Sigma_0 + \frac{8}{21}\Sigma_2) \\
-    %%
-    &&&\\
-    %%
-    (\frac{8}{15}\Sigma_0) &
-    (-\frac{16}{45}\Sigma_0 - \frac{4}{9}\Sigma_2) &
-    (\frac{64}{225}\Sigma_0 + \frac{16}{45}\Sigma_2 + \frac{9}{25}\Sigma_4) &
-    (-\frac{128}{525}\Sigma_0 - \frac{32}{105}\Sigma_2 - \frac{54}{175}\Sigma_4)
-    \\ 
-    %%
-    &&&\\
-    %%
-    (-\frac{16}{35}\Sigma_0) &
-    (\frac{32}{105}\Sigma_0 + \frac{8}{21}\Sigma_2) &
-    (-\frac{128}{525}\Sigma_0 - \frac{32}{105}\Sigma_2 - \frac{54}{175}\Sigma_4)
-    & 
-    (\frac{256}{1225}\Sigma_0 + \frac{64}{245}\Sigma_2 +
-    \frac{324}{1225}\Sigma_4 + \frac{13}{49}\Sigma_6)
-  \end{bmatrix}}\:.
-  \label{eq:A_matrix}
-\end{equation}
-Note that the term $\sum_{m=1}^4 A_{nm} u_m$ in
-Eq~(\ref{eq:spn_matrix}) couples the moments in each equation while
-the diffusive term in each equation is only for a single
-'pseudo-moment' $u_n$. As noted by Evans \citep{evans_simpli_2013},
-lower order $SP_N$ approximations can be generated by setting higher
-order even moments in this system to zero (e.g. $\phi_6 = \phi_4 = 0$
-yields the $SP_3$ equations). Boundary conditions for these equations
-are provided in Appendix~\ref{chap:spn_boundary_conditions}.
-
-%%---------------------------------------------------------------------------%%
-\section{Derivation of the Multigroup $SP_N$ Equations}
-\label{sec:mg_spn_equations}
-Up to this point, we have formulated the $SP_N$ equations for a single
-neutron energy. To expand these equations for multiple energies, we
-start by stating the multigroup neutron transport equation for a
-single dimension in planar geometry:
+\chapter{Derivation of the Multigroup $SP_N$ Equations}
+\label{chap:mg_spn_equations}
+In the text, we formulated the $SP_N$ equations for a single neutron
+energy. To expand these equations for multiple energies, we start by
+stating the multigroup neutron transport equation for a single
+dimension in planar geometry:
 \begin{multline}
@@
 \end{equation}
-Boundary conditions for these equations are provided in
-Appendix~\ref{chap:spn_boundary_conditions}.
-
-\subsection{Spatial Discretization of the $SP_N$ Equations}
-\label{subsec:spatial_discretization}
-To this point, the formulation of the multigroup $SP_N$ equations
-presented have discretized the transport equation in angle and energy
-but have yet to consider the spatial component of phase space. Of
-primary interest here is the discretization of the diffusive term in
-Eq~(\ref{eq:spn_multigroup_system}) and the current term on the
-boundaries in Eq~(\ref{eq:spn_multigroup_bnd}). Many popular choices
-for spatial discretization are available here and include finite
-difference and finite element formulations. In the Exnihilo package at
-ORNL, Evans has implemented a finite volume scheme over a Cartesian
-grid for these equations resulting in:
-\begin{multline}
-  - \mathbb{C}_{n,i+1} \mathbb{U}_{n_i+1} - \mathbb{C}_{n,i-1}
-  \mathbb{U}_{n_i-1} - \mathbb{C}_{n,j+1} \mathbb{U}_{n_j+1} -
-  \mathbb{C}_{n,j-1} \mathbb{U}_{n_j-1} - \mathbb{C}_{n,k+1}
-  \mathbb{U}_{n_k+1} - \mathbb{C}_{n,k-1} \mathbb{U}_{n_k-1}
-  +\\ \sum_{m=1}^4 \Big[\mathbb{A}_{nm,ijk} +
-    (\mathbb{C}_{n,i+1}+\mathbb{C}_{n,j+1}+\mathbb{C}_{n,k+1}+
-    \mathbb{C}_{n,i-1}+\mathbb{C}_{n,j-1}+\mathbb{C}_{n,k-1})\delta_{nm}\Big]
-  \mathbb{U}_{m,ijk} = \mathbb{Q}_{n,ijk}\:,
-  \label{eq:spn_space}
-\end{multline}
-with the coefficients defined as:
-\begin{equation}
-  \mathbb{C}_{n,l+1} = \frac{2}{\Delta_l}\Bigg( \frac{\mathbb{D}_{n,l+1}
-    \mathbb{D}_{n,l}}{\Delta_l \mathbb{D}_{n,l+1} + \Delta_{l+1} \mathbb{D}_{n,l}}\Bigg)\:,
-  \label{eq:spn_coeff_plus}
-\end{equation}
-\begin{equation}
-  \mathbb{C}_{n,l-1} = \frac{2}{\Delta_l}\Bigg( \frac{\mathbb{D}_{n,l}
-    \mathbb{D}_{n,l-1}}{\Delta_{l-1} \mathbb{D}_{n,l} + \Delta_{l}
-    \mathbb{D}_{n,l-1}}\Bigg)\:,
-  \label{eq:spn_coeff_minus}
-\end{equation}
-where $l={i,j,k}$ along the cardinal directions of the Cartesian grid
-cell in Figure~\ref{fig:mesh_cell}. The details of the derivation are
-provided in Appendix~\ref{chap:spn_spatial_discretization}.
-
-Although arbitrary grids could be handled effectively
-through a finite element scheme, the rectilinear grid used in Denovo
-is easily discretized through the finite volume method in a
-conservative form. This conservative form is ideal for the
-diffusion-like equations in the domain given by
-Eq~(\ref{eq:spn_multigroup_system}) and the effective boundary current
-conditions given by Eq~(\ref{eq:spn_multigroup_bnd}) as the neutron
-current is balanced from cell-to-cell, continuity of the flux is
-preserved across cell/material boundaries, and boundary conditions are
-naturally represented through cell-face currents. Furthermore, this
-spatial discretization is a natural extension of the balance
-principles used to arrive at the general transport equation.
-
-\subsection{Eigenvalue Form of the $SP_N$ Equations}
-\label{subsec:eigenvalue_form}
-For criticality problems, the space, angle, and energy discretizations
-for the $SP_N$ equations can be applied to the general eigenvalue from
-of the transport equation given by
-Eq~(\ref{eq:eigenvalue_transport}). We can readily replace the fixed
-source in Eq~(\ref{eq:multigroup_spn_equations}) with the fission
-source derived in \S~\ref{sec:pn_eigenvalue}:
-\begin{multline}
-  -\nabla \cdot \Bigg[\frac{n}{2n+1}\mathbf{\Sigma_{n-1}}^{-1} \nabla
-    \Big(\frac{n-1}{2n-1} \mathbf{\Phi_{n-2}} +
-    \frac{n}{2n-1}\mathbf{\Phi_n} \Big) \\+
-    \frac{n+1}{2n+1}\mathbf{\Sigma_{n+1}}^{-1} \nabla
-    \Big(\frac{n+1}{2n+3}\mathbf{\Phi_n} +
-    \frac{n+2}{2n+3}\mathbf{\Phi_{n+2}}\Big) \Bigg] \\+
-  \mathbf{\Sigma_n} \mathbf{\Phi_n} = \frac{1}{k} \mathbf{F}
-  \mathbf{\Phi_n} \delta_{n0} \ \ \ \ \ \ \ \ \ n = 0,2,4,\cdots,N\:.
-  \label{eq:multigroup_spn_eigenvalue}
-\end{multline}
-with the fission matrix $\mathbf{F}$ given by
-Eq~(\ref{eq:fission_matrix}). We again apply the change of variables
-to yield the set of pseudo-moments $\mathbb{U}_n$ where now the
-application of the fission matrix to the moment vectors will be
-expanded into a set of block matrices in identical fashion to the
-scattering matrices.
-
-It should be noted here that with respect to a general MCSA scheme,
-the introduction of the fission in the system will only affect the
-source vector in the linear system. The linear operator, and therefore
-overall MCSA performance will be dictated by streaming and scattering
-as defined on the left-hand side.
-
-%%---------------------------------------------------------------------------%%
-\section{Spectral Analysis of the $SP_N$ Equations}
-\label{sec:spn_spectral_analysis}
-Before we can move on to solving the $SP_N$ equations with both
-conventional and Monte Carlo methods, we must first understand their
-spectral character to ensure that MCSA will be applicable. As
-presented in Chapter~\ref{ch:stochastic_methods}, the spectral radius
-of the iteration matrix must be less than unity to ensure
-convergence. Therefore, we will compute the spectral radius of the
-iteration matrix formed by the $SP_N$ equations while parametrically
-varying the $SP_N$ order, the $P_N$ order, the number of energy
-groups, and upscattering/downscattering between energy groups. For
-each problem in the parameter study, the values given in
-Table~\ref{tab:spn_fixed_parameters} were fixed for each spatial
-location and energy group.
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{ll}\hline\hline
-      \multicolumn{1}{c}{Parameter}& 
-      \multicolumn{1}{c}{Value} \\\hline
-      Mesh Element Size & 1.0 \\
-      Mesh Elements & $4 \times 4 \times 4$ \\
-      Reflecting Boundaries & True \\
-      Materials & 1 \\
-      Source Strength & 1.0 \\
-      Total Cross Section & 5.0 \\
-      In-Group Cross Section & 0.25 \\
-      Downscatter Cross Section & 1.0 \\
-      Upscatter Cross Section & 0.1 \\
-      Eigenvalue Solver Tolerance & $1.0\times10^{-8}$ \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Fixed parameters for the $SP_N$ spectral radius
-      parameter study.} \textit{The parameters were fixed for each
-      spatial location and energy group.}}
-  \label{tab:spn_fixed_parameters}
-\end{table}
-Mesh element sizes were the same in all cardinal directions with
-reflecting conditions on each boundary. For the energy parameter, 1
-and 10 energy groups were used with upscatter/downscatter varied for
-the 10 group case. For the downscatter cases, all groups could scatter
-to all lower energy groups with the cross sections given in
-Table~\ref{tab:spn_fixed_parameters} while for the upscatter case all
-groups could scatter to all other energy groups with different cross
-sections for upscatter and downscatter used. A uniform isotropic
-source of neutrons is given in each group as well.
-
-The matrices generated by the discretization of this problem are
-sparse with a block-based form as dictated by
-Eq~(\ref{eq:spn_multigroup_system}). Figure~\ref{fig:group1} gives the
-sparsity pattern for a monoenergetic $SP_7$ discretization while
-Figures~\ref{fig:group10ds} and \ref{fig:group10us} give the sparsity
-patterns for the same problem for 10 energy groups without and with
-upscatter respectively. For these figures, the parameters in
-Table~\ref{tab:spn_fixed_parameters} with the number of spatial
-elements reduced to $2 \times 2 \times 2$ to show the blocks generated
-by the discretization.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4in]{chapters/spn_equations/group1.png}
-  \end{center}
-  \caption{\textbf{Sparsity pattern for 1-group $SP_7$
-      discretization.} \textit{A $2\times 2 \times 2$ element mesh was
-      used to show detail of the blocks formed by the discretization.}}
-  \label{fig:group1}
-\end{figure}
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4in]{chapters/spn_equations/group10ds.png}
-  \end{center}
-  \caption{\textbf{Sparsity pattern for 10-group $SP_7$ discretization
-      with downscatter only.} \textit{A $2\times 2 \times 2$ element
-      mesh was used to show detail of the blocks formed by the
-      discretization.}}
-  \label{fig:group10ds}
-\end{figure}
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4in]{chapters/spn_equations/group10us.png}
-  \end{center}
-  \caption{\textbf{Sparsity pattern for 10-group $SP_7$ discretization
-      with downscatter and upscatter.} \textit{A $2\times 2 \times 2$ element
-      mesh was used to show detail of the blocks formed by the
-      discretization.}}
-  \label{fig:group10us}
-\end{figure}
-We note a few key features of the sparsity plots. The first is that
-for multigroup problems without full upscatter and downscatter
-(i.e. Figure~\ref{fig:group10ds}), the resulting matrix is asymmetric
-and therefore a linear solver that can handle asymmetric linear
-systems is required. Nearly all problems of interest will not have
-full upscattering or downscattering. Second we note the largely
-diagonal character of these systems, although the blocks from
-Eq~(\ref{eq:A_block_matrix}) are readily apparent. Our first attempt
-at preconditioning this system will be to use the point Jacobi
-preconditioning from \S\ref{subsubsec:basic_mcsa_preconditioning} due
-to this diagonal form.
-
-\clearpage
-
-\subsection{Point Jacobi Spectral Analysis Results}
-\label{subsec:spn_analysis_results}
-Spectral radius computations were performed for the cases described
-above for the point Jacobi preconditioned iteration matrix $\ve{H} =
-\ve{I}-\ve{M}^{-1}\ve{A}$ with $\ve{M} =
-diag(\ve{A})$. Table~\ref{tab:group1pj} gives the results for the
-1-group case, Table~\ref{tab:group10dspj} for the 10-group case with
-full downscatter only and Table~\ref{tab:group10uspj} for the 10-group
-case with full downscatter and full upscatter.
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{cccccc}\hline\hline
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{$SP_N$ Order}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{} \\
-       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
-       & \textbf{0} & 0.0635 & 0.6722 & 1.3144 & 1.976 \\
-       & \textbf{1} & 0.0666 & 0.6728 & 1.3141 & 1.9755 \\
-      $P_N$ Order & \textbf{3} & 0.0666 & 0.6822 & 1.3141 & 1.9755 \\
-       & \textbf{5} & 0.0666 & 0.6822 & 1.3278 & 1.9847 \\
-       & \textbf{7} & 0.0666 & 0.6822 & 1.3278 & 1.9917 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Spectral radius results for the point Jacobi
-      preconditioned iteration matrix with 1 energy group.}}
-  \label{tab:group1pj}
-\end{table}
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{cccccc}\hline\hline
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{$SP_N$ Order}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{} \\
-       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
-       & \textbf{0} & 0.0655 & 0.677 & 1.32 & 1.982 \\
-       & \textbf{1} & 0.071 & 0.6777 & 1.319 & 1.982 \\
-      $P_N$ Order & \textbf{3} & 0.071 & 0.687 & 1.327 & 1.9872 \\
-       & \textbf{5} & 0.071 & 0.687 & 1.336 & 1.997 \\
-       & \textbf{7} & 0.071 & 0.687 & 1.336 & 1.9995 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Spectral radius results for the point Jacobi
-      preconditioned iteration matrix with 10 energy groups and full
-      downscatter.}}
-  \label{tab:group10dspj}
-\end{table}
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{cccccc}\hline\hline
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{$SP_N$ Order}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{} \\
-       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
-       & \textbf{0} & 0.7283 & 0.81 & 1.47 & 2.1446 \\
-       & \textbf{1} & 0.7317 & 0.8 & 1.46 & 2.1368 \\
-      $P_N$ Order & \textbf{3} & 0.7317 & 0.91 & 1.526 & 2.2274 \\
-       & \textbf{5} & 0.7317 & 0.91 & 1.5344 & 2.2562 \\
-       & \textbf{7} & 0.7317 & 0.91 & 1.5345 & 2.2842 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Spectral radius results for the point Jacobi
-      preconditioned iteration matrix with 10 energy groups, full
-      downscatter and full upscatter.}}
-  \label{tab:group10uspj}
-\end{table}
-It is readily apparent from the tabulated data that point Jacobi
-preconditioning is insufficient. For problems of order $SP_5$ and
-$SP_7$, the method will not converge at all and for the upscatter case
-with $SP_3$ the spectral radius is still quite large for large $P_N$
-orders and therefore convergence is expected to be slow. These
-eigenvalues signal a need for a better preconditioning strategy to
-both ensure and improve convergence for Monte Carlo methods.
-
-\subsection{Block Jacobi Preconditioning}
-\label{sec:spn_preconditioning}
-If Monte Carlo methods are to be used to solve the $SP_N$ system of
-equations, a different preconditioning strategy is required in order
-to ensure convergence for systems of all $SP_N$ and $P_N$ orders with
-arbitrary energy group structures. As another means of achieving this,
-we look back to the sparsity plots we generated in
-Figures~\ref{fig:group1}, \ref{fig:group10ds} and \ref{fig:group10us}
-as well as the multigroup $SP_N$ equations. Initially, the diagonal
-character of the system led us to try point Jacobi preconditioning
-with only marginal results. From the sparsity plots we note the block
-structure that ultimately arises from the multigroup scattering
-matrices and their insertion into Eq~(\ref{eq:A_block_matrix}). When
-full upscatter and downscatter are used the resulting blocks are
-completely dense while only downscatter gives a lower triangular
-scattering matrix and the block structure shown in
-Figure~\ref{fig:group10ds}.
-
-Based on this both block and diagonally dominant structure for
-matrices formed by the general multigroup $SP_N$ equations, we instead
-choose \textit{block Jacobi} preconditioning as a left preconditioner
-for the system. Like point Jacobi preconditioning, block Jacobi
-preconditioning extracts the diagonal elements of the matrix as the
-preconditioner where now the elements extracted are the blocks on the
-diagonal as shown on the left side of
-Figure~\ref{fig:block_jacobi_ex}. Shown on the right side of
-Figure~\ref{fig:block_jacobi_ex}, inversion of the preconditioner is
-trivial with each diagonal block inverted separately. For the $SP_N$
-equations, Eq~(\ref{eq:A_block_matrix}) gives a block size of
-$N_g\times(N+1)/2$.
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.5}{
-    \input{chapters/spn_equations/block_jacobi.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Block Jacobi preconditioning strategy used for the
-      $SP_N$ equations.} \textit{Left: The preconditioner is formed by
-      the diagonal blocks of the matrix. Right: Inversion of the
-      preconditioner is trivial and decoupled by block.}}
-  \label{fig:block_jacobi_ex}
-\end{figure}
-The inversion of this preconditioner is trivial as shown on the right
-side of Figure~\ref{fig:block_jacobi_ex}. Each block can be inverted
-individually and combined to form the inverse. In addition, in the
-limit of a block size of one, the block Jacobi method reduces to the
-point Jacobi method. For high performance implementations this has
-several attractive properties. First, the blocks in the matrix come
-from the energy/angle discretization of the transport equation as
-given by Eq~(\ref{eq:A_block_matrix}). Each block on the diagonal is
-bound to a mesh element in the system (note there are 8 blocks on the
-diagonal in each of the sparsity patterns with a mesh of $2 \times 2
-\times 2$) and therefore we expect the matrix elements forming the
-block to be entirely local. Second, these blocks are typically dense
-and nearly lower triangular for many transport problems meaning that
-established dense matrix methods can be used for fast inversion.
-
-\subsection{Block Jacobi Spectral Analysis Results}
-\label{subsec:spn_analysis_results}
-Spectral radius computations for the block Jacobi preconditioned
-iteration matrix were performed for the same problems as the point
-Jacobi preconditioned case. Table~\ref{tab:group1bj} gives the results
-for the 1-group case, Table~\ref{tab:group10dsbj} for the 10-group
-case with full downscatter only and Table~\ref{tab:group10usbj} for
-the 10-group case with full downscatter and full upscatter. A block
-size of $N_g\times(N+1)/2$ was used for the preconditioner in all
-cases.
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{cccccc}\hline\hline
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{$SP_N$ Order}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{} \\
-       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
-       & \textbf{0} & 0.0635 & 0.1269 & 0.1444 & 0.1513 \\
-       & \textbf{1} & 0.0666 & 0.1315 & 0.1474 & 0.1534 \\
-      $P_N$ Order & \textbf{3} & 0.0666 & 0.1365 & 0.154 & 0.1592 \\
-       & \textbf{5} & 0.0666 & 0.1365 & 0.1562 & 0.163 \\
-       & \textbf{7} & 0.0666 & 0.1365 & 0.1562 & 0.164 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Spectral radius results for the block Jacobi
-      preconditioned iteration matrix with 1 energy group.}}
-  \label{tab:group1bj}
-\end{table}
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{cccccc}\hline\hline
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{$SP_N$ Order}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{} \\
-       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
-       & \textbf{0} & 0.0647 & 0.1275 & 0.1449 & 0.1514 \\
-       & \textbf{1} & 0.0686 & 0.1338 & 0.1484 & 0.1547 \\
-      $P_N$ Order & \textbf{3} & 0.0687 & 0.1399 & 0.1582 & 0.1625 \\
-       & \textbf{5} & 0.0692 & 0.1399 & 0.1582 & 0.1657 \\
-       & \textbf{7} & 0.0678 & 0.1393 & 0.1624 & 0.166 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Spectral radius results for the block Jacobi
-      preconditioned iteration matrix with 10 energy groups and full
-      downscatter.}}
-  \label{tab:group10dsbj}
-\end{table}
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{cccccc}\hline\hline
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{$SP_N$ Order}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{} \\
-       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
-       & \textbf{0} & 0.1887 & 0.2267 & 0.2285 & 0.2286 \\
-       & \textbf{1} & 0.4535 & 0.5044 & 0.5045 & 0.5045 \\
-      $P_N$ Order & \textbf{3} & 0.4535 & 0.6453 & 0.6506 & 0.6506 \\
-       & \textbf{5} & 0.4535 & 0.6453 & 0.6802 & 0.6818 \\
-       & \textbf{7} & 0.4535 & 0.6453 & 0.6802 & 0.6927 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Spectral radius results for the block Jacobi
-      preconditioned iteration matrix with 10 energy groups, full
-      downscatter and full upscatter.}}
-  \label{tab:group10usbj}
-\end{table}
-
-From the tabulated block Jacobi data it is clear that this is a viable
-preconditioning choice for all $SP_N$ problems in term of Monte Carlo
-solution methods. All cases were observed to have a spectral radius
-below unity and often significantly smaller than that, greatly
-improving convergence rates over the point Jacobi preconditioned
-problem. Based on these results, the block Jacobi method should be the
-first preconditioning strategy applied to more complicated problems
-representative of physical systems.
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{Fuel Assembly Criticality Calculations}
-\label{sec:fuel_assembly_calcs}
-Fuel assembly calculations are a critical piece of nuclear engineering
-infrastructure for reactor core analysis and design. At this level,
-individual fuel pins may be resolved at fine resolution in a variety
-of configurations. As a sophisticated problem of interest to push the
-limits of MCSA, a hot zero-power $17 \times 17$ pin assembly will be
-used with varying energy group structure and $SP_N$ discretization in
-a criticality calculation. A cross section along the vertical axis
-showing homogenized fuel pin materials and the associated grid is
-given by Figure~\ref{fig:problem3_radial_mat} while a cross section of
-the materials configuration along the horizontal axis is given by
-Figure~\ref{fig:problem3_axial_mat}. A detailed view of the assembly
-bottom is given in Figure~\ref{fig:problem3_end}. On the top and
-bottom of the assembly, vacuum conditions are used as well as on the
-top and right boundaries in
-Figure~\ref{fig:problem3_radial_mat}. Reflecting conditions are used
-on the left and bottom boundaries of
-Figure~\ref{fig:problem3_radial_mat}, effectively giving a
-representation of one quarter of the assembly.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4in]{chapters/spn_equations/problem3_radial_mat.png}
-  \end{center}
-  \caption{\textbf{Fuel assembly mesh and geometry cross section.}
-    \textit{Reflecting boundaries are used on the left and lower
-      boundaries to create a complete $17 \times 17$ assembly
-      geometry. Gray regions are homogenized fuel and red regions are
-      homogenized moderator. Each fuel pin is resolved by a $2 \times
-      2$ mesh.}}
-  \label{fig:problem3_radial_mat}
-\end{figure}
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6.0in]{chapters/spn_equations/problem3_axial_mat.png}
-  \end{center}
-  \caption{\textbf{Fuel assembly geometry cross section.}  \textit{The
-      geometry is subdivided along the axial direction into 50 zones
-      spaced to capture material boundaries. Important details include
-      spacer grids along the length of the fuel pins and reactor core
-      structures on the top and bottom of the assembly. Lighter purple
-      material in the center of the assembly is moderator and darker
-      purple/red material is fuel.}}
-  \label{fig:problem3_axial_mat}
-\end{figure}
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6.0in]{chapters/spn_equations/problem3_end.png}
-  \end{center}
-  \caption{\textbf{Fuel assembly geometry end detail.}
-    \textit{Reactor core structure including spacer grids and plenum
-      has been included. Lighter purple material on the right of the
-      figure is moderator and darker purple/red material is fuel.}}
-  \label{fig:problem3_end}
-\end{figure}
-Significant geometric details are contained in the model including
-spacer grids, fuel pins with homogenized cladding and gas gap, core
-plenum, and moderator with boron. Group cross sections and other
-discrete nuclear data are generated as needed by a cross-section
-processing module dependent on the meshing parameters used to
-discretize the geometry and single-dimension pin-cell calculations for
-initial flux spectrum generation. Table~\ref{tab:problem3_parameters}
-gives the primary design parameters for the fuel assembly
-calculations.
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{ll}\hline\hline
-      \multicolumn{1}{l}{\textbf{Parameter}} & 
-      \multicolumn{1}{l}{\textbf{Value}} \\
-      Power Level & 0 MW \\
-      Inlet Temperature & 326.85C \\
-      Fuel Temperature & 600C \\
-      Boron Concentration & 1300 ppm \\
-      Moderator Density & 0.743 g/cc \\
-      Helium Density & \sn{1.79}{-4} g/cc \\
-      Zirconium Density & 6.56 g/cc \\
-      Stainless Steel Density & 8.0 g/cc \\
-      Inconel Density & 8.19 g/cc \\
-      UO2 Density & 10.257 g/cc \\
-      Fuel Pin Radius (w/o clad) & 0.4096 cm \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Design parameters for the $17 \times 17$ pin fuel
-      assembly criticality calculation.}}
-  \label{tab:problem3_parameters}
-\end{table}
-
-To generate the multiplication factor and steady-state flux
-distribution for this problem, at every eigenvalue iteration MCSA is
-used to solve the resulting $SP_N$ problem using the provided fission
-source. Algorithm~\ref{alg:power_iteration} presents the use of MCSA
-within a power iteration strategy to find the multiplication factor.
-\begin{algorithm}[h!]
-  \caption{Power Iteration MCSA Scheme}
-  \label{alg:power_iteration}
-  \begin{algorithmic}
-    \State $k_0 =$ initial guess
-    \State $\mathbf{\Phi}_0 =$ initial guess
-    \State $n = 0$
-    \While{$|\frac{k^n - k^{n-1}}{k^n}| < \epsilon$}
-    \Comment{Iterate until convergence of the eigenvalue}
-    \State $\mathbf{M} \mathbf{\Phi}^{n+1} = \frac{1}{k^n} \mathbf{F} \mathbf{\Phi}^n$
-    \Comment{Solve for the new flux state with MCSA}
-    \State $k^{n+1} = k^n \frac{\int \mathbf{F} \mathbf{\Phi}^{n+1} d\mathbf{r}}{\int
-      \mathbf{F} \mathbf{\Phi}^n d\mathbf{r}}$
-    \Comment{Update the multiplication factor}
-    \State $n = n+1$
-    \EndWhile
-  \end{algorithmic}
-\end{algorithm}
-Here, $\mathbf{M}$ is the transport operator generated on the
-left-hand side of the $SP_N$ discretization, $\mathbf{F}$ is the
-fission matrix, and $\mathbf{\Phi}$ the multigroup neutron flux. This
-problem is significantly more complicated than the simple test problem
-used for the previous spectral analysis. Fission has been introduced
-into the set of equations and the addition of moderator into the
-system will increase the amount of scattering, creating a
-significantly more difficult problem manifesting itself in an
-iteration matrix with a larger spectral radius. When using MCSA, the
-linear operator applied to $\mathbf{\Phi}^{n+1}$ at each eigenvalue
-iteration will dictate convergence and remain unchanged throughout the
-computation\footnote{The operator will change if, for example, physics
-  feedback through temperature or potentially burnup is
-  considered. These additional physics will modify the cross sections
-  used to assemble $\mathbf{M}$. The calculations presented here will
-  consist of a single eigenvalue calculation with a static
-  $\mathbf{M}$.} while the addition of fission to the system will only
-modify the source of neutrons and the multiplication factor while not
-affecting Monte Carlo transport.
-
-\subsection{Preliminary Jacobi Preconditioned Calculations}
-\label{subsec:jacobi_prec_assembly_calc}
-Based on the success of block Jacobi preconditioning with the test
-problem used for the spectral radius parameter study, we use it first
-to solve the $17 \times 17$ fuel assembly problem. A single energy
-group problem was first solved with $SP_1$ discretization, effectively
-giving the one-speed neutron diffusion system for the fuel assembly
-resulting in 20,088 degrees of freedom in the
-problem. Figure~\ref{fig:block_jacobi_res_mcsa} gives the residual
-infinity norm as a function of iteration for the MCSA linear solve in
-the first eigenvalue iteration using 25,000 stochastic histories at
-every iteration for the adjoint Neumann-Ulam solve.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/block_jacobi_res.pdf}
-  \end{center}
-  \caption{\textbf{Residual infinity norm vs. iteration for the block
-      Jacobi preconditioned MCSA solve during the first eigenvalue
-      iteration of the 1-group $17 \times 17$ fuel assembly problem.}
-    \textit{Convergence was not achieved with the block Jacobi
-      preconditioned method.}}
-  \label{fig:block_jacobi_res_mcsa}
-\end{figure}
-Convergence was not achieved as noted by the rapid rise in the
-residual over a few iterations. Based on the spectral radius
-computations performed, these results are not in line with
-expectations for this problem. Additional computations performed with
-\sn{1}{6} histories per iteration exhibited the same divergent
-behavior at a significant computational cost and compute time. Even if
-the problem may be ill-conditioned, we do expect convergence of
-MCSA. To investigate further, a block Jacobi preconditioned Richardson
-iteration was used to solve the same
-problem. Figure~\ref{fig:block_jacobi_res_richardson} gives the
-residual infinity norm as a function of iteration for the Richardson
-linear solve in the first eigenvalue iteration.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/block_jacobi_rich_res.pdf}
-  \end{center}
-  \caption{\textbf{Residual infinity norm vs. iteration for the block
-      Jacobi preconditioned Richardson solve during the first
-      eigenvalue iteration of the 1-group $17 \times 17$ fuel assembly
-      problem.}  \textit{A spectral radius near 1 was observed for the
-      iteration matrix.}}
-  \label{fig:block_jacobi_res_richardson}
-\end{figure}
-Poor converge is observed for the Richardson iteration, however,
-convergence is achieved meaning that the preliminary criteria needed
-to satisfy MCSA convergence has also been met. The number of
-iterations required for the Richardson iteration to converge will give
-an approximation for the spectral radius via
-Eq~(\ref{eq:linear_k_iter_norm3}). With 7291 iterations required to
-converge to a tolerance of $\sn{1}{-6}$, $\rho(\mathbf{H}) \approx
-0.998$, nearing the limits of MCSA applicability and well beyond the
-spectral radii generated in the initial spectral analysis.
-
-Based on these results, it then appears that even if the simple
-criteria of a spectral radius of less than one is met and the
-Richardson iteration will converge with the same preconditioning, MCSA
-still may not converge. We then expect the issue to reside with the
-Neumann-Ulam solve providing the correction as the Richardson
-iteration is known to provide the correct result. Furthermore, the
-fact that the spectral radius is less than one means that the
-stochastic histories in the block Jacobi preconditioned Neumann-Ulam
-method are eventually being terminated by the weight cutoff as no
-artificial absorption was used for these preliminary
-calculations. Based on this, the correction being generated by the
-Neumann-Ulam solve is the component of MCSA causing a divergent
-solution.
-
-\subsection{MCSA Breakdown}
-\label{subsec:mcsa_break_down}
-For the fuel assembly problem, the initial block Jacobi
-preconditioned\footnote{For 1-group $SP_1$ problems the block size is
-  1, giving a point Jacobi preconditioner.}  calculations used a
-one-speed $SP_1$ discretization, effectively giving a diffusion
-system. To study the breakdown of MCSA at iteration matrix spectral
-radii near one, we will use a simpler homogeneous 2-dimensional
-one-speed neutron diffusion system to isolate this behavior. In this
-system, we can vary the cross sections while maintaining a fixed grid
-in order to achieve varying spectral radii. For these studies, we
-neglect fission as MCSA behavior is dictated by the transport operator
-$\mathbf{M}$ in an eigenvalue scheme with the fission matrix used to
-generate a fixed source.
-
-\subsubsection{Model Diffusion Problem}
-\label{subsec:model_problem}
-For our breakdown study, we choose the one-speed, two-dimensional
-neutron diffusion equation as a model problem
-\citep{duderstadt_nuclear_1976}:
-\begin{equation}
-  -\boldsymbol{\nabla} \cdot D \boldsymbol{\nabla} \phi + \Sigma_a
-  \phi = S\:,
-  \label{eq:diffusion_eq}
-\end{equation}
-where $\phi$ is the neutron flux, $\Sigma_a$ is the absorption cross
-section, and $S$ is the source of neutrons. In addition, $D$ is the
-diffusion coefficient defined as:
-\begin{equation}
-  D = \frac{1}{3 ( \Sigma_t - \bar{\mu}\Sigma_s )}\:,
-  \label{eq:diffusion_coeff}
-\end{equation}
-where $\Sigma_s$ is the scattering cross section, $\Sigma_t = \Sigma_a
-+ \Sigma_s$ is the total cross section, and $\bar{\mu}$ is the cosine
-of the average scattering angle. For simplicity, we will take
-$\bar{\mu} = 0$ for our analysis giving $D=(3 \Sigma_t)^{-1}$. In
-addition, to further simplify we will assume a homogeneous domain such
-that the cross sections remain constant throughout. Doing this permits
-us to rewrite Eq~(\ref{eq:diffusion_eq}) as:
-\begin{equation}
-  -D \boldsymbol{\nabla}^2 \phi + \Sigma_a \phi = S\:.
-  \label{eq:diffusion_eq_simple}
-\end{equation}
-
-We choose a finite difference scheme on a square Cartesian grid to
-discretize the problem. For the Laplacian, we choose the 9-point
-stencil shown in Figure~\ref{fig:stencil} over a grid of size $h$
-\citep{leveque_finite_2007}:
-\begin{multline}
-  \nabla^2_9\phi = \frac{1}{6h^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j}
-    + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ +
-    \phi_{i-1,j+1} + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20
-    \phi_{i,j}]\:.
-  \label{eq:nine_point_stencil}
-\end{multline}
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.25}{\input{chapters/spn_equations/stencil.pdftex_t}}
-  \end{center}
-  \caption{\textbf{Nine-point Laplacian stencil.}}
-  \label{fig:stencil}
-\end{figure}
-We then have the following linear system to solve:
-\begin{multline}
-  -\frac{1}{6h^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j} + 4
-    \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ + \phi_{i-1,j+1}
-    + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20 \phi_{i,j}] + \Sigma_a
-  \phi_{i,j} = s_{i,j}\:,
-  \label{eq:fd_system}
-\end{multline}
-and in operator form:
-\begin{equation}
-  \ve{D}\boldsymbol{\phi}=\ve{s}\:,
-  \label{eq:operator_system}
-\end{equation}
-where $\ve{D}$ is the diffusion operator, $\ve{s}$ is the source in
-vector form and $\boldsymbol{\phi}$ is the vector of unknown fluxes.
-
-To close the system, a set of boundary conditions is required. In the
-case of a non-reentrant current condition applied to all global
-boundaries of the domain, we choose the formulation of Duderstadt by
-assuming the flux is zero at some ghost point beyond the
-grid. Consider for example the equations on the $i=0$ boundary of the
-domain:
-\begin{multline}
-  -\frac{1}{6h^2}[4 \phi_{-1,j} + 4 \phi_{1,j} + 4 \phi_{0,j-1} +
-    4 \phi_{0,j+1} + \phi_{-1,j-1}\\ + \phi_{-1,j+1} + \phi_{1,j-1} +
-    \phi_{1,j+1} - 20 \phi_{0,j}] + \Sigma_a \phi_{0,j} = s_{0,j}\:.
-  \label{eq:x_min_bnd}
-\end{multline}
-Here we note some terms where $i=-1$ and therefore are representative
-of grid points beyond the boundary of the domain. We set the flux at
-these points to be zero, giving a valid set of equations for the $i=0$
-boundary:
-\begin{multline}
-  -\frac{1}{6h^2}[4 \phi_{1,j} + 4 \phi_{0,j-1} + 4 \phi_{0,j+1}
-    \\ + \phi_{-1,j+1} + \phi_{1,j-1} + \phi_{1,j+1} - 20 \phi_{0,j}]
-  + \Sigma_a \phi_{0,j} = s_{0,j}\:.
-  \label{eq:x_min_bnd_2}
-\end{multline}
-We repeat this procedure for the other boundaries of the domain. For
-reflecting boundary conditions, the net current across a boundary is
-zero.
-
-\subsubsection{MCSA Breakdown Analysis}
-\label{subsubsec:breakdown_analysis}
-For each solver and estimator combination, the spectral radius of the
-iteration matrix generated by the diffusion problem was varied by
-changing the absorption cross section from 0.25 to 100 while fixing
-the grid size at $100 \times 100$ with $h = 0.1$ and a fixed
-scattering cross section of unity. For each parameter variation, a
-minimum of one stochastic history per degree of freedom (DOF) in the
-problem was used to compute the Monte Carlo correction. If the solver
-could not converge in less than 100 iterations, the number of
-histories was increased by increments of 5,000 until convergence was
-achieved in less than 100 iterations. The number of iterations
-required to converge MCSA and the time to converge was recorded as a
-means to capture the breakdown.
-
-Figure~\ref{fig:breakdown_iterations} gives the number of iterations
-required to converge for the chosen number of histories per iteration
-given by Figure~\ref{fig:breakdown_histories} using the adjoint solver
-with the collision and expected value estimators and the forward
-solver with the collision estimator. For spectral radii less than
-0.97, all MCSA problems converged with 1 history per DOF (10,000 for
-this problem) with the number of iterations required to converge
-increasing as a function of spectral radius. Near a spectral radius of
-0.97, the number of histories required to converge MCSA in less than
-100 iterations takes a dramatic rise that exhibits neither exponential
-nor power law behavior. As the spectral radius approaches 1, the
-number of histories required becomes significant and effectively
-impractical to compute. Even with this simple diffusion problem, the
-behavior is consistent with that observed for the fuel assembly
-problem with $SP_1$ discretization. In that case, we estimated a
-spectral radius of $\approx 0.998$, larger than any of the spectral
-radii that could be computed within even 90 minutes of compute time
-for this simple two dimensional problem. For that problem, even
-1,000,000 histories ($\approx 50$ per DOF) was not enough to provide
-convergence. Even if single solve times of an hour can be tolerated,
-dozens of solves are typically required to compute the multiplication
-factor and flux spectrum within the $SP_N$ eigenvalue scheme for more
-difficult problems like the fuel assembly, making the method unusable.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/breakdown_iterations.pdf}
-  \end{center}
-  \caption{\textbf{Iterations required to converge as a function of
-      spectral radius for the neutron diffusion problem.} \textit{The
-      number of histories was increased to achieve convergence in less
-      than 100 iterations. At least 10,000 histories were used for
-      each calculation.}}
-  \label{fig:breakdown_iterations}
-\end{figure}
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/breakdown_histories.pdf}
-  \end{center}
-  \caption{\textbf{Histories per DOF required to converge as a
-      function of spectral radius for the neutron diffusion problem.}
-    \textit{The number of histories was increased to achieve
-      convergence in less than 100 iterations. At least 10,000
-      histories were used for each calculation with 10,000 DOFs in the
-      problem.}}
-  \label{fig:breakdown_histories}
-\end{figure}
-
-In addition to the significantly larger number of histories required
-for to achieve convergence for ill-conditioned problems another
-penalty is paid due to histories that take longer to
-compute. Figure~\ref{fig:breakdown_time} gives the CPU time in seconds
-required to compute a single random walk averaged over the entire set
-of histories run in the calculation over all iterations. As the
-spectral radius increases (correlating to a higher ratio of scattering
-in the system) the random walk lengths increase, using more CPU time
-to finish the computation. Compared to spectral radii of 0.5, larger
-spectral radii over 0.97 have histories that require two orders of
-magnitude more computation time. This significant increase in
-computation time per history coupled with the significant increase in
-the number of histories required to converge is evidence that for
-problems with spectral radii above $\approx 0.97$, using MCSA to solve
-any problems of interest is entirely ineffective and not practical. We
-therefore require a more expansive set of preconditioning techniques
-to move the eigenvalue spectrum of the $SP_N$ problem into a regime in
-which MCSA is more applicable and in which performance is improved.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/breakdown_time.pdf}
-  \end{center}
-  \caption{\textbf{CPU time per history as a function of spectral
-      radius for the neutron diffusion problem.} \textit{As the
-      spectral radius grows, so do the length of the random
-      walks. Longer random walks require more CPU time to compute.}}
-  \label{fig:breakdown_time}
-\end{figure}
-
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{Advanced Preconditioning Strategies}
-\label{subsec:spn_advanced_preconditioning}
-For the fuel assembly criticality problem presented in the previous
-section, the spectral radius was near one using Jacobi preconditioning
-and in a regime in which MCSA breakdown is observed. In this regime
-the number of stochastic histories required to converge MCSA increases
-rapidly and the resulting poor performance is compounded by those
-histories being increasingly expensive to compute. To overcome this
-difficulty, advanced preconditioning strategies for the $SP_N$ problem
-are required beyond simple Jacobi methods that can reduce the spectral
-radius into a region of better MCSA performance. Several modern
-algebraic preconditioning strategies will be presented here and used
-within the explicit preconditioning framework given by
-Eq~(\ref{eq:left_right_mcsa}). Data showing their effects on MCSA
-solutions of the fuel assembly criticality problem will be presented.
-
-\subsection{ILUT Preconditioning}
-\label{subsec:spn_ilut_preconditioning}
-Incomplete lower-upper (ILU) factorizations of the linear operator can
-be used a simple mechanism to form an approximate inverse of a
-preconditioner. To build the factorization, the sparse upper and lower
-triangular factors, $\mathbf{L}$ and $\mathbf{U}$, are computed such
-that the residual matrix formed by the factorization
-\citep{saad_iterative_2003}:
-\begin{equation}
-  \mathbf{R} = \mathbf{L} \mathbf{U} - \mathbf{A} \:,
-  \label{eq:ilu_residual_matrix}
-\end{equation}
-has a specified sparsity pattern and element magnitude. When a
-magnitude threshold is used, elements generated in the factors below
-that magnitude are dropped, resulting in ILU threshold (ILUT)
-preconditioning. The sparsity pattern in this case is determined from
-the input matrix to be preconditioned and the number of elements
-maintained in the factor is specified by a fill level parameter. A
-fill level of 1 will generate the same number of elements as the
-sparsity pattern of the input matrix while a fill level of 2 will
-contain twice as many elements in the factorization, resulting in a
-better representation of the true LU factorization. The inverse of the
-lower and upper triangular factors may then be easily inverted by
-means of simple elimination to produce the preconditioner. For MCSA,
-$\mathbf{L}^{-1}$ will be used on the left and $\mathbf{U}^{-1}$ on
-the right to precondition the system.
-
-For modern subspace methods, only the action to the preconditioner on
-a vector is required for efficient implementations and therefore a
-triangular solve can be used for this purpose. For the explicit
-preconditioning scheme presented in Eq~(\ref{eq:left_right_mcsa}), the
-fully inverted operator must be generated in order to build the set of
-probabilities and weights required for Monte Carlo
-sampling. Therefore, for a linear system of size $N$, $N$ triangular
-solves will be required in order to extract the inverse matrices from
-production ILU implementations. In addition, parallel implementations
-typically generate lower and upper factors that are only triangular
-locally, providing an easily parallelizable mechanism to generate the
-action of the preconditioner inverse. A consequence of this choice for
-parallel scalability is a degradation of the preconditioning quality
-as the size of the parallel system is increased. For serial
-computations, the triangular factorization is potentially exact
-depending on the parameters chosen while at thousands of parallel
-tasks, the global triangular factors differ significantly from the true
-factorization. As a result, more iterations are required at higher
-levels of parallelism to converge the system and will ultimately
-degrade overall scalability of the system with respect to total wall
-time to converge.
-
-MCSA preconditioned with ILUT was used to solve the fuel assembly
-problem presented in the previous section for a 1-group $SP_1$
-discretization. Unlike the Jacobi preconditioned strategy, convergence
-was achieved with ILUT preconditioning. To study convergence
-sensitivity to ILUT parameters, the fill level and drop tolerance were
-parametrically varied with the number of iterations to converge an
-eigenvalue iteration for the fuel assembly problem recorded along with
-the maximum number of non-zero entries observed in all matrix rows for
-the left/right preconditioned composite linear operator. To provide
-some sparsity to the factorization, the ILUT drop tolerance was used
-to drop elements in the extracted inverse triangular factors. For each
-calculation, the number of iterations required to converge reported
-was for a single eigenvalue iteration with \sn{3}{4} histories at each
-MCSA iteration to compute the Monte Carlo correction using the adjoint
-collision estimator. All ILUT calculations reported here were
-performed on a single CPU and therefore this data does not take into
-account the aforementioned effects of parallel decomposition on the
-quality of the ILUT preconditioning.
-
-Figure~\ref{fig:ilut_iterations} gives the number of iterations to
-converge the fixed source problem to a tolerance of \sn{1}{8}. As
-expected, the higher the fill level chosen for the ILUT factorization,
-the fewer iterations are required to converge the problem (only 9 MCSA
-iterations were required to converge the problem for the fill level of
-3 and drop tolerance of \sn{1}{-5} case). For higher levels of fill,
-the iterations needed to converge were not as sensitive, signaling
-that a larger drop tolerance can perhaps be used without a significant
-degradation in iterative performance. At smaller fill levels, the
-sensitivity to the ILUT drop tolerance is more significant. For all
-fill levels used, convergence was not achieved for the fuel assembly
-problem with a drop tolerance larger than \sn{1}{-3}.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/ilut_iterations.pdf}
-  \end{center}
-  \caption{\textbf{Number of MCSA iterations required to converge an
-      eigenvalue iteration for the fuel assembly problem with ILUT
-      preconditioning as a function of ILUT drop tolerance.}
-    \textit{Each colored curve represents the iteration behavior for a
-      different ILUT fill level. Fill levels of 1.5, 2.0, 2.5, and 3.0
-      were used.}}
-  \label{fig:ilut_iterations}
-\end{figure}
-
-Unfortunately, gaining convergence (and excellent iterative
-performance) with MCSA for the $SP_N$ fuel assembly problem comes at
-an immediate cost. Figure~\ref{fig:ilut_size} gives the maximum number
-of non-zero entries in the composite linear operator generated by
-preconditioning as a function ILUT drop tolerance for varying values
-of fill level. For the 1-group $SP_1$ discretization, the original
-linear operator will contain only a maximum of 7 non-zero entries per
-row in the system. As observed in Figure~\ref{fig:ilut_size},
-performing the explicit preconditioning yields composite linear
-operators with $O(1,000)$ elements in a row for all combinations of
-fill level and drop tolerance, over 10\% of the total row size for
-this particular problem. This large number of row entries, observed
-for a significant fraction of rows in the system, creates several
-problems. First, sparsity is completely destroyed with each state in
-the system now coupled to over 10\% of the total states in the system
-through the composite iteration matrix. This means that Monte Carlo
-sampling tables will be large, requiring significant memory to store
-them and a substantial overhead in the sampling procedure during
-transport. In addition, the matrix-matrix multiply operations required
-to build the composite operator see significant performance losses due
-to the amount of data that must be handled. In parallel, losing
-sparsity severely inhibits performance where now parallel operations
-require communications amongst orders of magnitude more processors for
-nearest-neighbor type algorithms.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/ilut_size.pdf}
-  \end{center}
-  \caption{\textbf{Maximum number of non-zero entries observed for all
-      rows in the composite linear operator for the fuel assembly
-      problem with ILUT preconditioning given as a function of ILUT
-      drop tolerance.}  \textit{Each colored curve represents the row
-      size for a different ILUT fill level. Fill levels of 1.5, 2.0,
-      2.5, and 3.0 were used.}}
-  \label{fig:ilut_size}
-\end{figure}
-
-As a means to assess the quality of the preconditioning, a simple
-metric is developed to address these concerns and allow comparison to
-future developments. For our studies, our core performance metric will
-be iterative performance with the minimum number of MCSA iterations
-required to converge the problem desired. For Monte Carlo
-calculations, this improved performance is balanced by the creation of
-a dense composite system and we seek to reduce the number of non-zero
-entries to a minimum value in order to lower the amount of coupling
-amongst states in the system and reduce the amount of memory used
-along with the potential latency overhead. Given these two objectives,
-the following metric is proposed:
-\begin{equation}
-  \text{Quality Metric} = (\text{\# iterations}) \times (\text{maximum
-    \# of non-zero values})\:,
-\end{equation}
-where the highest quality preconditioning is one that minimizes this
-metric. For the ILUT preconditioned data provided,
-Figure~\ref{fig:ilut_quality} provides the computed metric as a
-function of ILUT drop tolerance for each fill level used. In general,
-the metric is similar to the data observed for the number of
-iterations required to converge as the non-zero row entries changed by
-a smaller fraction over drop tolerances tested.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/ilut_quality.pdf}
-  \end{center}
-  \caption{\textbf{ILUT preconditioning quality metric for the fuel
-      assembly problem given as a function of ILUT drop tolerance.}
-    \textit{Each colored curve represents the quality metric behavior
-      for a different ILUT fill level. Fill levels of 1.5, 2.0, 2.5,
-      and 3.0 were used.}}
-  \label{fig:ilut_quality}
-\end{figure}
-
-It should be noted here that although only the behavior of the linear
-solve during a single eigenvalue iteration is reported, the behavior
-of the linear solver was observed to be consistent throughout the
-eigenvalue iterations (25 total eigenvalue iterations were required to
-converge the 1-group $SP_1$ problem). We expect this as the linear
-operator (which is unchanging in the fuel assembly criticality
-problem) dictates the convergence of MCSA. At each iteration the
-fission source provided to the linear problem is changing, however, we
-expect the same convergence behavior for all source vectors
-$\mathbf{b} \in \mathbb{R}^{N}$ where $||\mathbf{b}||_1 \neq 0$.
-
-\clearpage
-
-\subsection{Sparse Approximate Inverse Preconditioning}
-\label{subsec:spn_spainv_preconditioning}
-Explicit formation of the preconditioner inverse is a requirement of
-the current MCSA preconditioning strategy. As a result, although ILUT
-preconditioning provided excellent iterative performance, a
-significant time penalty is paid to extract the inverse of the
-triangular factors through $N$ triangular solves. In addition, this
-explicit inversion yielded a composite linear operator that was dense,
-requiring modification of the system in order to achieve better
-scalability and Monte Carlo performance. Sparse approximate inverse
-(SPAINV) preconditioning is a technique that may potentially alleviate
-both of these constraints at the cost of reduced iterative
-performance. The method produces a sparse approximation of the inverse
-matrix directly by minimizing the Frobenius norm of the residual
-matrix \citep{saad_iterative_2003}:
-\begin{equation}
-  || \mathbf{I} - \mathbf{A} \mathbf{M} ||^2_F =
-  \sum_{j=1}^N ||\mathbf{e}_j - \mathbf{A} \mathbf{m}_j||^2_2 \:,
-  \label{eq:frobenius_norm_min}
-\end{equation}
-where the inverse preconditioning matrix $\mathbf{M}$ minimizes the
-norm on the right and $\mathbf{e}_j$ and $\mathbf{m}_j$ are the
-$j^{th}$ columns of the identity and preconditioning matrices
-respectively. Here, $\mathbf{M}$ is the actual inverse of the
-preconditioner and is formed directly by the algorithm. On the right
-hand side of Eq~(\ref{eq:frobenius_norm_min}), the Frobenius norm is
-represented as a effective sum of $N$ linear system residual norms. A
-few iterations of a subspace method or other iterative scheme can be
-used to effectively solve these systems to a loose tolerance and yield
-the columns of the approximate preconditioning matrix. At each
-iteration, threshold values are used to eliminate the components of
-each $\mathbf{m}_j$ column to maintain the desired level of
-sparsity. In addition, a sparsity pattern can be predetermined and
-enforced during this process. For the SPAINV implementation used for
-this work, the sparsity pattern was defined by levels such that a
-level of $N$ for an input operator $\mathbf{A}$ will generate a
-preconditioning matrix $\mathbf{M}$ with the same sparsity pattern as
-$\mathbf{A}^{N+1}$. By using the norm minimization strategy instead of
-a factorization such as ILUT, parallel results are reproduced
-regardless of parallel problem size, meaning that a serial computation
-will converge in the same number of iterations as a computation with
-thousands of cores.
-
-MCSA preconditioned with SPAINV was used to solve the fuel assembly
-problem as with ILUT for a 1-group $SP_1$ discretization. To study
-convergence sensitivity to SPAINV parameters, the number of levels in
-the sparsity pattern and threshold were parametrically varied with the
-number of iterations to converge a single eigenvalue iteration for the
-fuel assembly problem recorded along with the maximum number of
-non-zero entries observed in all matrix rows for the left/right
-preconditioned composite linear operator. For each calculation, the
-number of iterations required to converge reported was for a single
-eigenvalue iteration with \sn{3}{4} histories at each MCSA iteration
-to compute the Monte Carlo correction using the adjoint collision
-estimator. All SPAINV calculations reported here were performed on a
-single CPU.
-
-Figure~\ref{fig:spainv_iterations} gives the number of MCSA iterations
-needed to converge the fuel assembly problem and
-Figure~\ref{fig:spainv_size} the maximum number of non-zero entries
-per row obeserved in the composite linear operator as a function the
-SPAINV threshold. For this analysis, the number of levels in the
-sparsity pattern was varied from 3 to 7 with convergence of the fuel
-assembly problem not achieved for smaller level values. Compared to
-ILUT preconditioning, at higher levels we see comparable iterative
-performance with a relative invariance to the threshold value. The
-threshold did have a significant effect, however, on the time required
-to generate the preconditioner with a threshold of 0.001 over an order
-of magnitude slower than a threshold of 0.01 due to the inclusion of a
-significant number of extra values in the input operator. In addition
-to comparable iterative performance, the sparsity of the composite
-operator is greatly improved with nearly an order of magnitude
-reduction in number of non-zero entries per row for lower level
-values.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/spainv_iterations.pdf}
-  \end{center}
-  \caption{\textbf{Number of MCSA iterations required to converge a
-      single eigenvalue iteration for the fuel assembly problem with
-      SPAINV preconditioning as a function of SPAINV threshold.}
-    \textit{Each colored curve represents the iteration behavior for a
-      different SPAINV level pattern. Levels of 3, 4, 5, 6, and 7 were
-      used.}}
-  \label{fig:spainv_iterations}
-\end{figure}
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/spainv_size.pdf}
-  \end{center}
-  \caption{\textbf{Maximum number of non-zero entries observed for all
-      rows in the composite linear operator for the fuel assembly
-      problem with SPAINV preconditioning given as a function of
-      SPAINV threshold.}  \textit{Each colored curve represents the
-      row size for a different SPAINV level pattern. Levels of 3, 4,
-      5, 6, and 7 were used.}}
-  \label{fig:spainv_size}
-\end{figure}
-
-With the comparable iterative performance and improved sparsity,
-SPAINV preconditioning should then have a more favorable quality
-metric than ILUT preconditioning. Figure~\ref{fig:spainv_quality}
-gives the quality metric as a function of the threshold value for each
-of the sparsity levels computed. Not only do we see an improved
-quality metric over ILUT preconditioning (about an order of magnitude
-less), but we also note that the ideal sparsity level is not the largest
-nor the smallest. In fact, the largest and smallest levels (3 and 7)
-performed the worst in terms of the quality metric with a level of 4
-performing the best. When CPU time to generate the preconditioner is
-considered, 4 is the clear choice for this problem as more time is
-required to do the approximate inversion as the sparsity pattern of
-the preconditioner grows in size.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/spainv_quality.pdf}
-  \end{center}
-  \caption{\textbf{SPAINV preconditioning quality metric for the fuel
-      assembly problem given as a function of SPAINV threshold.}
-    \textit{Each colored curve represents the quality metric behavior
-      for a different SPAINV level pattern. Levels of 3, 4, 5, 6, and
-      7 were used.}}
-  \label{fig:spainv_quality}
-\end{figure}
-
-\clearpage
-
-\subsection{SPAINV Improved ILUT Preconditioning}
-\label{subsec:spn_spainv_preconditioning}
-As a means to improve convergence over using only SPAINV
-preconditioning, ILUT preconditioning may be modified through an
-additional application of SPAINV
-(SPAINV-ILUT)\citep{saad_iterative_2003}. For the ILUT preconditioning
-case we have the following linear problem to solve for the substituted
-variable:
-\begin{equation}
-  \mathbf{L}^{-1} \mathbf{A} \mathbf{U}^{-1} \mathbf{u} =
-  \mathbf{L}^{-1} \mathbf{b}\:.
-  \label{eq:ilut_standalone}
-\end{equation}
-We then apply SPAINV and minimize the following problem, this time on
-the left:
-\begin{equation}
-  || \mathbf{I} - \mathbf{M} \mathbf{L}^{-1} \mathbf{A}
-  \mathbf{U}^{-1} ||^2_F\:,
-  \label{eq:ilut_minimize}
-\end{equation}
-giving a new preconditioned system:
-\begin{equation}
-  \mathbf{M} \mathbf{L}^{-1} \mathbf{A} \mathbf{U}^{-1} \mathbf{u} =
-  \mathbf{M} \mathbf{L}^{-1} \mathbf{b}\:.
-  \label{eq:ilut_spainv}
-\end{equation}
-Doing this permits a smaller fill level to be used with ILUT and a
-lower-order sparsity pattern to be used with SPAINV, giving marginally
-more sparsity in the resulting composite linear operator while
-maintaining good iterative performance.
-
-MCSA preconditioned with SPAINV-ILUT was also used to solve the same
-fuel assembly problem as with the other preconditioners for the same
-1-group $SP_1$ discretization. To study convergence sensitivity to
-SPAINV-ILUT parameters, the ILUT drop tolerance and ILUT fill level
-were parametrically varied with the number of iterations to converge a
-single eigenvalue iteration for the fuel assembly problem recorded
-along with the maximum number of non-zero entries observed in all
-matrix rows for the left/right preconditioned composite linear
-operator. SPAINV parameters for the data presented here were fixed as
-it was found that the preconditioning quality was largely insenstive
-to their variation. In addition, the fact that SPAINV is being used to
-preconditioned a problem already preconditioned with ILUT means that
-the input matrix sparsity pattern is already somewhat dense and large
-SPAINV levels would result in an even denser system. Therefore, a
-smaller SPAINV sparsity pattern level will be used for this
-preconditioning. For each calculation, the number of iterations
-required to converge reported was for a single eigenvalue iteration.
-Again, \sn{3}{4} histories are used at each MCSA iteration to compute
-the Monte Carlo correction using the adjoint collision estimator. All
-SPAINV-ILUT calculations reported here were performed on a single CPU.
-
-Figure~\ref{fig:spainv_ilut_iterations} gives the number of iterations
-required to converge as a function of ILUT drop tolerance for ILUT
-fill levels of 1.5, 2.0, 3.0, 4.0 and 5.0 with a fixed SPAINV
-threshold of 1.0 and sparsity level of 1. SPAINV threshold values
-higher than 1.0 resulted in a loss of convergence and lower values did
-not significantly affect performance. Larger sparsity levels resulted
-in a composite operator that was too dense and a prohibitive amount of
-CPU time required to compute the preconditioner. At smaller levels of
-fill, the benefits of using an additional step of SPAINV
-preconditioning are readily observed with the number of iterations
-required to converge cut in half for large drop tolerances. For
-smaller drop tolerances and higher ILUT levels of fill, the additional
-step of SPAINV preconditioning offers marginal improvement to
-iterative performance. As shown in Figure~\ref{fig:spainv_ilut_size},
-because ILUT is used in the preconditioning sequence, sparsity is
-again lost with only marginally improved non-zero entry numbers due to
-the application of the SPAINV threshold. As a result, the quality
-metrics given by Figure~\ref{fig:spainv_ilut_quality} are only
-marginally better than those observed for ILUT preconditioning and
-much larger than the SPAINV quality metric values.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/psilut_iterations.pdf}
-  \end{center}
-  \caption{\textbf{Number of MCSA iterations required to converge a
-      single eigenvalue iteration for the fuel assembly problem with
-      SPAINV-ILUT preconditioning as a function of ILUT drop
-      tolerance.} \textit{Each colored curve represents the row size
-      for a different ILUT fill level. Fill levels of 1.5, 2.0, 3.0,
-      4.0, and 5.0 were used.}}
-  \label{fig:spainv_ilut_iterations}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/psilut_size.pdf}
-  \end{center}
-  \caption{\textbf{Maximum number of non-zero entries observed for all
-      rows in the composite linear operator for the fuel assembly
-      problem with SPAINV-ILUT preconditioning given as a function of
-      ILUT drop tolerance.}  \textit{Each colored curve represents the
-      row size for a different ILUT fill level. Fill levels of 1.5,
-      2.0, 3.0, 4.0 and 5.0 were used.}}
- \label{fig:spainv_ilut_size}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/psilut_quality.pdf}
-  \end{center}
-  \caption{\textbf{SPAINV-ILUT preconditioning quality metric for the
-      fuel assembly problem given as a function of ILUT drop
-      tolerance.}  \textit{Each colored curve represents the row size
-      for a different ILUT fill level. Fill levels of 1.5, 2.0, 3.0,
-      4.0 and 5.0 were used.}}
-  \label{fig:spainv_ilut_quality}
-\end{figure}
-
-\clearpage
-
-\subsection{Applying the Reduced Domain Approximation}
-\label{subsec:spn_prec_rda}
-For each preconditioning technique presented convergence was achieved
-for the single fuel assembly problem. However, a primary concern is
-the number of non-zero states in each row of the system generated by
-the explicit preconditioning strategy. In many cases, orders of
-magnitude more matrix elements were generated resulting in poor
-scalability for domain decomposed algorithms and overall performance
-issues for Monte Carlo. As outlined in
-\S~\ref{subsec:reduced_domain_approximation}, the reduced domain
-approximation may be used as a mechanism to potentially alleviate this
-problem by filtering elements of the composite matrix in each row that
-fall below a certain threshold value or by maintaining the largest $N$
-elements in each row where $N$ is a designated fill level.
-
-For the ILUT, SPAINV, and SPAINV-ILUT preconditioning strategies the
-reduced domain approximation will be applied to reduce the density of
-the composite linear operator to more manageable levels. For each
-preconditioner, the parameters that achieved the best quality metric
-results from the previous analysis were used with ILUT parameters of a
-fill level of 5.0 and a drop tolerance of \sn{1}{-5}, SPAINV
-parameters of a sparsity level of 4 and a threshold of 0.1 and
-SPAINV-ILUT with ILUT parameters of a fill level of 4.0 and a drop
-tolerance of \sn{1}{-5} and SPAINV parameters of a sparsity level of 1
-and a threshold of 1.0. The reduced domain threshold was set to
-\sn{1}{-10} in order to eliminate any exceedingly small values from
-the matrix (generally this is simply removing non-zero elements within
-the floating point tolerance of zero). The reduced domain fill level
-was then varied, starting with the largest non-zero entries per row
-value observed for each of the preconditioning types in order to
-assess its effects relative to the case where no reduced domain
-approximation was applied.
-
-Figure~\ref{fig:rda_iterations} gives the number of iterations
-required to converge for each preconditioner type as a function of
-reduced domain fill level. Figure~\ref{fig:rda_quality} gives the
-corresponding quality metric for each data point where the number of
-non-zero entries used to compute the metric is equivalent to the
-reduced domain fill level. We first note the SPAINV preconditioning
-alone is signficantly more sensitive to the reduction in domain size
-over the ILUT-based methods, although the preconditioner was of
-$O(100)$ non-zero entries per row without any approximation
-applied. For the ILUT-based methods, performance was significantly
-better with converge achieved in less than 40 MCSA iterations with
-only 10 non-zero entries in each row (vs. 7 non-zero entries for the
-case with no preconditioning). SPAINV-ILUT iterative performance was
-marginally better than ILUT alone for all reduced domain fill
-levels. However, it should be noted that the ILUT level of fill used
-for the ILUT only calculations was set to 4 for this case while the
-SPAINV-ILUT preconditioner used an ILUT level of fill of 5 and
-therefore the marginally better performance is more likely a result of
-this addition of fill rather than the extra SPAINV
-preconditioning. Looking at the quality metric data, we see a nice
-power-law reduction in the quality metric as a function of the reduced
-domain fill level, achieving 2 orders of magnitude reduction in the
-quality metric for the ILUT-based preconditioning methods.
-
-Although applying the reduced domain approximation results in a
-successful recovery of sparsity for the Monte Carlo problem while
-maintaining good convergence properties, there is still an issue of
-forming the composite operator before applying the approximation and
-potentially generating the transpose of this operator in the case of
-the adjoint Monte Carlo method. Because of this, memory and scaling
-issues may still be observed when building the Monte Carlo problem. In
-addition, the expensive extraction of the inverse of the
-preconditioning operators for the explicit scheme creates a
-significant cost in overall performance. Future work in this area
-should be considerate of these important components of the algorithm.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/rda_iterations.pdf}
-  \end{center}
-  \caption{\textbf{Number of MCSA iterations required to converge a
-      single eigenvalue iteration for the fuel assembly problem with
-      each preconditioning as a function of reduced domain
-      approximation fill level.}  \textit{The largest fill level for
-      each preconditioning presented is that using the parameters that
-      gave the best results without the approximation.}}
-  \label{fig:rda_iterations}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/rda_quality.pdf}
-  \end{center}
-  \caption{\textbf{Preconditioning quality metric for the fuel
-      assembly problem given as a function of reduced domain
-      approximation fill level.}  \textit{The largest fill level for
-      each preconditioning presented is that using the parameters that
-      gave the best results without the approximation.}}
-  \label{fig:rda_quality}
-\end{figure}
-
-\clearpage
-
-\subsection{MCSA Relaxation Parameters}
-\label{subsec:spn_mcsa_relaxation}
-The Richardson iteration upon which the Neumann-Ulam method is built
-may be implemented with a scalar relaxation parameter that can be
-adjusted to improve convergence:
-\begin{equation}
-  \mathbf{x} = \mathbf{x} + \omega \mathbf{r}\:,
-  \label{eq:richardson_relaxation}
-\end{equation}
-where $\omega$ is the relaxation parameter. This is effectively a form
-of preconditioning where the system is being scaled on the left by a
-constant value in all rows. Analogously, these relaxation parameter
-techniques can be applied to Monte Carlo to improve convergence as
-demonstrated by Dimov \citep{dimov_new_1998}. In this case, building
-an iteration matrix from Eq~(\ref{eq:richardson_relaxation}) gives:
-\begin{equation}
-  \mathbf{H} = \mathbf{I} - \omega \mathbf{A}\:,
-  \label{eq:relaxed_iteration_matrix}
-\end{equation}
-with the probabilities and weights for the Monte Carlo game
-appropriately scaled. By inspection, such a scaling is a simple form
-of preconditioning on the left where all rows in the system are scaled
-by the same scalar parameter. For MCSA, we can stage this scheme with
-two separate relaxation parameters; one for the outer Richardson
-iteration and one for the inner Monte Carlo solve:
-\begin{subequations}
-  \begin{gather}
-    \ve{x}^{k+1/2} = \ve{x}^k + \omega_R \ve{r}^k\:,\\ \ve{r}^{k+1/2}
-    = \ve{b} -
-    \ve{A}\ve{x}^{k+1/2}\:,\\ \omega_N\ve{A}\delta\ve{x}^{k+1/2} =
-    \omega_N\ve{r}^{k+1/2}\:,\\ \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta
-    \ve{x}^{k+1/2}\:,\\ \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:,
-  \end{gather}
-  \label{eq:mcsa_relaxed}
-\end{subequations}
-where $\omega_R$ is the Richardson iteration relaxation parameter and
-$\omega_N$ is the Neumann-Ulam Monte Carlo solve relaxation parameter.
-
-\clearpage
-
-\subsection{Monte Carlo Estimator Comparison}
-\label{subsec:spn_estimator_comparison}
-With convergence obtained for the fuel assembly problem, the advanced
-preconditioners will next be studied with each Monte Carlo
-estimator. Although the expected value estimator has achieved the best
-iterative performance in the previous comparison and analysis, it is
-possible that the other estimators may perform better from a timing
-perspective due to the sparsity lost by explicit preconditioning. The
-deterministic component of the expected value estimator that couples
-the current state to other local states through the iteration matrix
-stencil requires application of the estimator to potentially several
-orders of magnitude more states at every transition event.
-
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{MCSA Comparison to Conventional Methods}
-\label{sec:spn_comparison}
-
-\subsection{Results}
-\label{subsec:spn_comparison_results}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/solver_iters.pdf}
-  \end{center}
-  \caption{\textbf{Average number of iterations required to converge
-      each eigenvalue iteration for the fuel assembly problem as a
-      function of energy groups.} \textit{All methods were
-      preconditioned with identical ILUT paramters. For MCSA, several
-      variations of fixed point iterations and estimators were used.}}
-  \label{fig:comparison_iterations}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/solver_time.pdf}
-  \end{center}
-  \caption{\textbf{Average CPU time per iteration in seconds for the
-      fuel assembly problem as a function of energy groups.}
-    \textit{All linear solver iterations over all eigenvalue
-      iterations were used to compute the averge. All methods were
-      preconditioned with identical ILUT paramters. For MCSA, several
-      variations of fixed point iterations and estimators were used.}}
-  \label{fig:comparison_time}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/solver_p_time.pdf}
-  \end{center}
-  \caption{\textbf{Average CPU time per iteration in seconds for the
-      fuel assembly problem as a function of energy groups with
-      preconditioning time included for the MCSA methods.}
-    \textit{All linear solver iterations over all eigenvalue
-      iterations were used to compute the averge. All methods were
-      preconditioned with identical ILUT paramters. For MCSA, several
-      variations of fixed point iterations and estimators were used.}}
-  \label{fig:comparison_prec_time}
-\end{figure}

### 2013-05-31 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
moving multigroup spn derivation to an appendix

@@
 yields the $SP_3$ equations). Boundary conditions for these equations
-are provided in Appendix~\ref{chap:spn_boundary_conditions}.
-
-%%---------------------------------------------------------------------------%%
-\section{Derivation of the Multigroup $SP_N$ Equations}
-\label{sec:mg_spn_equations}
-Up to this point, we have formulated the $SP_N$ equations for a single
-neutron energy. To expand these equations for multiple energies, we
-start by stating the multigroup neutron transport equation for a
-single dimension in planar geometry:
-\begin{multline}
-  \mu \frac{\partial}{\partial x} \psi^g(x,\mu) + \sigma^g(x)
-  \psi^g(x,\mu) = \\ \sum_{g'=0}^{G} \int
-  \sigma_s^{gg'}(x,\hat{\Omega}' \rightarrow \hat{\Omega})
-  \psi^{g'}(x,\hat{\Omega}') d\Omega' + \frac{q^g(x)}{4 \pi}\:,
-  \label{eq:cart_1d_multigroup}
-\end{multline}
-where $g$ denotes the group index of $0$ to $G$ groups, $G=N_g-1$, and
-the integration of the scattering emission term over energy has been
-replaced by a discrete summation. For scattering, $\sigma_s^{gg'}$
-provides the probability of scattering at a particular angle from
-group $g$ to $g'$. The result is an equation nearly identical in form
-to Eq~(\ref{eq:cart_1d_transport}) where now instead of forming the
-$SP_N$ equations for a single energy group, we form them for each of
-the energy groups with group coupling occurring through the scattering
-term. The multigroup $P_N$ equations are then:
-\begin{equation}
-   \frac{1}{2n+1} \frac{\partial}{\partial x}\Big[ (n+1) \phi^g_{n+1}
-     + n \phi^g_{n-1} \Big] +
-   \sum_{g'}(\sigma^g\delta_{gg'}-\sigma^{gg'}_{sn}) \phi^g_n =
-   q\delta_{n0} \:,
-  \label{eq:multigroup_pn_equations}
-\end{equation}
-for $n = 0,1,\dotsc,N$ where the flux and scattering moments are
-defined in a group. We observe that a $N_g \times N_g$ scattering
-matrix is formed:
-\begin{equation}
-  \mathbf{\Sigma}_n =
-  \sum_{g'}(\sigma^g\delta_{gg'}-\sigma^{gg'}_{sn})\:,
-  \label{eq:scattering_matrix}
-\end{equation}
-and when expanded gives:
-%% Also grabbed this matrix from Tom's tech note.
-\begin{equation}
-  \mathbf{\Sigma}_n =
-  \begin{bmatrix}
-    (\sigma^0-\sigma_{sn}^{00}) & -\sigma_{sn}^{01} & \dots &
-    -\sigma_{sn}^{0G} \\ &&&\\ -\sigma_{sn}^{10} &
-    (\sigma^1-\sigma_{sn}^{11}) & \dots & -\sigma_{sn}^{1G}
-    \\ &&&\\ \vdots & \vdots & \ddots & \vdots
-    \\ &&&\\ -\sigma_{sn}^{G0} & -\sigma_{sn}^{G1} & \dots &
-    (\sigma^G-\sigma_{sn}^{GG})
-  \end{bmatrix}\:.
-\end{equation}
-It is also useful to combine the group flux moments and sources into a
-single vector for more compact notation:
-\begin{equation}
-  \mathbf{\Phi_n} = (\phi^0_n\ \phi^1_n\ \cdots \phi^G_n )^T\:, 
-  \label{eq:group_flux_vector}
-\end{equation}
-\begin{equation}
-  \mathbf{q} = (q^0\ q^1\ \cdots q^G )^T\:.
-  \label{eq:group_source_vector}
-\end{equation}
-Next, we apply the $SP_N$ approximation to
-Eq~(\ref{eq:multigroup_pn_equations}) in identical fashion to the
-monoenergetic case. This gives:
-\begin{multline}
-  -\nabla \cdot \Bigg[\frac{n}{2n+1}\mathbf{\Sigma_{n-1}}^{-1} \nabla
-    \Big(\frac{n-1}{2n-1} \mathbf{\Phi_{n-2}} +
-    \frac{n}{2n-1}\mathbf{\Phi_n} \Big) \\+
-    \frac{n+1}{2n+1}\mathbf{\Sigma_{n+1}}^{-1} \nabla
-    \Big(\frac{n+1}{2n+3}\mathbf{\Phi_n} +
-    \frac{n+2}{2n+3}\mathbf{\Phi_{n+2}}\Big) \Bigg] \\+
-  \mathbf{\Sigma_n} \mathbf{\Phi_n} = \mathbf{q}
-  \delta_{n0}\ \ \ \ \ \ \ \ \ n = 0,2,4,\cdots,N\:.
-  \label{eq:multigroup_spn_equations}
-\end{multline}
-This adds more complexity than the monoenergetic formulation in that
-all unknowns in this group of equations are now vector quantities and
-scattering relationships are contained in matrices rather than a
-scalar quantity. Because of this, the same sequence of variable
-changes and algebra can be used to build a set of matrix equations,
-this time in a block formulation:
-\begin{equation}
-  -\nabla \cdot \mathbb{D}_n \nabla \mathbb{U}_n + \sum_{m=1}^4
-  \mathbb{A}_{nm} \mathbb{U}_m = \mathbb{Q}_n\:,
-  \label{eq:spn_multigroup_system}
-\end{equation}
-where the definition of all quantities are the same with internal
-scalar values replaced by the group-vector values. In addition,
-$\mathbb{A}$ is now a block matrix of $N_g \times N_g$ sub-matrices
-generated from the moment scattering matrices:
-\begin{equation}
-  \mathbf{A} = 
-  {\tiny \begin{bmatrix}
-    (\mathbf{\Sigma}_0) &
-    (-\frac{2}{3}\mathbf{\Sigma}_0) &
-    (\frac{8}{15}\mathbf{\Sigma}_0) &
-    (-\frac{16}{35}\mathbf{\Sigma}_0) \\
-    %%
-    &&&\\
-    %%
-    (-\frac{2}{3}\mathbf{\Sigma}_0) &
-    (\frac{4}{9}\mathbf{\Sigma}_0 + \frac{5}{9}\mathbf{\Sigma}_2) &
-    (-\frac{16}{45}\mathbf{\Sigma}_0 - \frac{4}{9}\mathbf{\Sigma}_2) &
-    (\frac{32}{105}\mathbf{\Sigma}_0 + \frac{8}{21}\mathbf{\Sigma}_2) \\
-    %%
-    &&&\\
-    %%
-    (\frac{8}{15}\mathbf{\Sigma}_0) &
-    (-\frac{16}{45}\mathbf{\Sigma}_0 - \frac{4}{9}\mathbf{\Sigma}_2) &
-    (\frac{64}{225}\mathbf{\Sigma}_0 + \frac{16}{45}\mathbf{\Sigma}_2 + \frac{9}{25}\mathbf{\Sigma}_4) &
-    (-\frac{128}{525}\mathbf{\Sigma}_0 - \frac{32}{105}\mathbf{\Sigma}_2 - \frac{54}{175}\mathbf{\Sigma}_4)
-    \\ 
-    %%
-    &&&\\
-    %%
-    (-\frac{16}{35}\mathbf{\Sigma}_0) &
-    (\frac{32}{105}\mathbf{\Sigma}_0 + \frac{8}{21}\mathbf{\Sigma}_2) &
-    (-\frac{128}{525}\mathbf{\Sigma}_0 - \frac{32}{105}\mathbf{\Sigma}_2 - \frac{54}{175}\mathbf{\Sigma}_4)
-    & 
-    (\frac{256}{1225}\mathbf{\Sigma}_0 + \frac{64}{245}\mathbf{\Sigma}_2 +
-    \frac{324}{1225}\mathbf{\Sigma}_4 + \frac{13}{49}\mathbf{\Sigma}_6)
-  \end{bmatrix}}\:.
-  \label{eq:A_block_matrix}
-\end{equation}
-Boundary conditions for these equations are provided in
-Appendix~\ref{chap:spn_boundary_conditions}.
-
-\subsection{Spatial Discretization of the $SP_N$ Equations}
-\label{subsec:spatial_discretization}
-To this point, the formulation of the multigroup $SP_N$ equations
-presented have discretized the transport equation in angle and energy
-but have yet to consider the spatial component of phase space. Of
-primary interest here is the discretization of the diffusive term in
-Eq~(\ref{eq:spn_multigroup_system}) and the current term on the
-boundaries in Eq~(\ref{eq:spn_multigroup_bnd}). Many popular choices
-for spatial discretization are available here and include finite
-difference and finite element formulations. In the Exnihilo package at
-ORNL, Evans has implemented a finite volume scheme over a Cartesian
-grid for these equations resulting in:
-\begin{multline}
-  - \mathbb{C}_{n,i+1} \mathbb{U}_{n_i+1} - \mathbb{C}_{n,i-1}
-  \mathbb{U}_{n_i-1} - \mathbb{C}_{n,j+1} \mathbb{U}_{n_j+1} -
-  \mathbb{C}_{n,j-1} \mathbb{U}_{n_j-1} - \mathbb{C}_{n,k+1}
-  \mathbb{U}_{n_k+1} - \mathbb{C}_{n,k-1} \mathbb{U}_{n_k-1}
-  +\\ \sum_{m=1}^4 \Big[\mathbb{A}_{nm,ijk} +
-    (\mathbb{C}_{n,i+1}+\mathbb{C}_{n,j+1}+\mathbb{C}_{n,k+1}+
-    \mathbb{C}_{n,i-1}+\mathbb{C}_{n,j-1}+\mathbb{C}_{n,k-1})\delta_{nm}\Big]
-  \mathbb{U}_{m,ijk} = \mathbb{Q}_{n,ijk}\:,
-  \label{eq:spn_space}
-\end{multline}
-with the coefficients defined as:
-\begin{equation}
-  \mathbb{C}_{n,l+1} = \frac{2}{\Delta_l}\Bigg( \frac{\mathbb{D}_{n,l+1}
-    \mathbb{D}_{n,l}}{\Delta_l \mathbb{D}_{n,l+1} + \Delta_{l+1} \mathbb{D}_{n,l}}\Bigg)\:,
-  \label{eq:spn_coeff_plus}
-\end{equation}
-\begin{equation}
-  \mathbb{C}_{n,l-1} = \frac{2}{\Delta_l}\Bigg( \frac{\mathbb{D}_{n,l}
-    \mathbb{D}_{n,l-1}}{\Delta_{l-1} \mathbb{D}_{n,l} + \Delta_{l}
-    \mathbb{D}_{n,l-1}}\Bigg)\:,
-  \label{eq:spn_coeff_minus}
-\end{equation}
-where $l={i,j,k}$ along the cardinal directions of the Cartesian grid
-cell in Figure~\ref{fig:mesh_cell}. The details of the derivation are
-provided in Appendix~\ref{chap:spn_spatial_discretization}.
-
-Although arbitrary grids could be handled effectively
-through a finite element scheme, the rectilinear grid used in Denovo
-is easily discretized through the finite volume method in a
-conservative form. This conservative form is ideal for the
-diffusion-like equations in the domain given by
-Eq~(\ref{eq:spn_multigroup_system}) and the effective boundary current
-conditions given by Eq~(\ref{eq:spn_multigroup_bnd}) as the neutron
-current is balanced from cell-to-cell, continuity of the flux is
-preserved across cell/material boundaries, and boundary conditions are
-naturally represented through cell-face currents. Furthermore, this
-spatial discretization is a natural extension of the balance
-principles used to arrive at the general transport equation.
+are provided in Appendix~\ref{chap:spn_boundary_conditions}. The
+multigroup form is presented in Appendix~\ref{chap:spn_mg_equations}
+and the spatial discretization in
+Appendix~\ref{chap:spn_spatial_discretization}.
 

### 2013-05-31 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding relaxation data figures

@@
   \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/rich_relax_iters.pdf}
+    \includegraphics[width=6in]{chapters/spn_equations/relax_iters.pdf}
   \end{center}
@@
       eigenvalue iteration of the fuel assembly problem as a function
-      of the Richardson relaxation parameter.} \textit{The Neumann
-      relaxation parameter was fixed at 1.0 and \sn{3}{4} stochastic
-      histories were used to compute the MCSA correction at each
-      iteration.}}
+      of the relaxation parameters.} \textit{The Neumann relaxation
+      parameter was fixed at 1.0 and \sn{3}{4} stochastic histories
+      were used to compute the MCSA correction at each iteration.}}
   \label{fig:rich_relax_iters}
@@
   \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/rich_relax_time.pdf}
+    \includegraphics[width=6in]{chapters/spn_equations/relax_time.pdf}
   \end{center}
@@
       iteration of the fuel assembly problem as a function of the
-      Richardson relaxation parameter.} \textit{The Neumann relaxation
-      parameter was fixed at 1.0 and \sn{3}{4} stochastic histories
-      were used to compute the MCSA correction at each iteration.}}
+      relaxation parameters.} \textit{The Neumann relaxation parameter
+      was fixed at 1.0 and \sn{3}{4} stochastic histories were used to
+      compute the MCSA correction at each iteration.}}
   \label{fig:rich_relax_time}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/neumann_relax_iters.pdf}
-  \end{center}
-  \caption{\textbf{Number of iterations to converge a single
-      eigenvalue iteration of the fuel assembly problem as a function
-      of the Neumann relaxation parameter.} \textit{The Richardson
-      relaxation parameter was fixed at 1.0 and \sn{3}{4} stochastic
-      histories were used to compute the MCSA correction at each
-      iteration.}}
-  \label{fig:neumann_relax_iters}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/neumann_relax_time.pdf}
-  \end{center}
-  \caption{\textbf{CPU time in seconds to converge a single eigenvalue
-      iteration of the fuel assembly problem as a function of the
-      Neumann relaxation parameter.} \textit{The Richardson relaxation
-      parameter was fixed at 1.0 and \sn{3}{4} stochastic histories
-      were used to compute the MCSA correction at each iteration.}}
-  \label{fig:neumann_relax_time}
 \end{figure}

### 2013-06-01 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
additions to spn chapter

@@
 Figure~\ref{fig:problem3_radial_mat}, effectively giving a
-representation of one quarter of the assembly.
+representation of one quarter of the assembly. For the spatial
+discretization, each fuel pin (gray regions in
+Figure~\ref{fig:problem3_radial_mat}) is resolved by a $2 \times 2$
+mesh with materials and cross sections homgenized over this region.
 \begin{figure}[t!]
@@
 \end{figure}
+
 Significant geometric details are contained in the model including
@@
 
+\clearpage
+
 \subsection{Preliminary Jacobi Preconditioned Calculations}
@@
 $\mathbf{b} \in \mathbb{R}^{N}$ where $||\mathbf{b}||_1 \neq 0$.
-
-\clearpage
 
@@
 \end{equation}
-with the probabilities and weights for the Monte Carlo game
+with the probabilities and weights for the Monte Carlo procedure
 appropriately scaled. By inspection, such a scaling is a simple form
@@
 estimator. A reduced domain fill level of 100 is used with a threshold
-of \sn{1}{-10} to filter small values.
+of \sn{1}{-10} to filter small values. ILUT preconditioning with a
+drop tolerance of \sn{1}{-5} and fill level of 5 is used as
+well.
+
+Figure~\ref{fig:relax_iters} gives the number of iterations required
+to converge the fuel assembly problem for a 1-group SP1 discretization
+using varying combinations of the relaxation parameters. First, both
+parameters were fixed at the base case of 1 while the other parameter
+was varied as shown in the left-hand side plots of
+Figure~\ref{fig:relax_iters}. For the Richardson relaxation parameter,
+using a value larger than 1 gave better iterative performance up to a
+point, effectively providing a stronger extrapolation using the
+residual at each iteration. For the Neumann relaxation parameter, a
+value of less than 1 is observed to be ideal. Although initially
+counter-intuitive, the fact that the correction computed by the Monte
+Carlo solver has a stochastic error associated with it means that by
+using a Neumann relaxation parameter less than 1, the correction and
+its error are effectively dampened to improve iterative
+performance. Considering the CPU time required to converge a single
+eigenvalue iteration, similar results are also observed for the
+relaxation parameters as given by Figure~\ref{fig:relax_time}. For the
+base cases given by the plots on the left, it was found the a
+Richardson relaxation parameter of 1.1 and a Neumann relaxation
+parameter of 0.7 provided the fastest CPU time for convergence. Fixing
+each parameter at these new values, the calculations were repeated as
+shown for the plots on the right hand sides of both
+Figures~\ref{fig:relax_iters} and \ref{fig:relax_time}. For each
+repeated timing calculation, it was found that the same combination of
+relaxation parameters found in the base cases performed the best
+although they did not necessarily have the best iterative performance.
 
@@
       eigenvalue iteration of the fuel assembly problem as a function
-      of the relaxation parameters.} \textit{The Neumann relaxation
-      parameter was fixed at 1.0 and \sn{3}{4} stochastic histories
-      were used to compute the MCSA correction at each iteration.}}
-  \label{fig:rich_relax_iters}
+      of the relaxation parameters.} \textit{Starting in upper left
+      and moving counter-clockwise: Neumann relaxation parameter fixed
+      at 1.0, Richardson relaxation parameter fixed at 1.0, Neumann
+      relaxation parameter fixed at 0.7, Richardson relaxation
+      parameter fixed at 1.1. For each calculation \sn{3}{4}
+      stochastic histories were used to compute the MCSA correction at
+      each iteration.}}
+  \label{fig:relax_iters}
 \end{figure}
@@
       iteration of the fuel assembly problem as a function of the
-      relaxation parameters.} \textit{The Neumann relaxation parameter
-      was fixed at 1.0 and \sn{3}{4} stochastic histories were used to
-      compute the MCSA correction at each iteration.}}
-  \label{fig:rich_relax_time}
+      relaxation parameters.} \textit{Starting in upper left and
+      moving counter-clockwise: Neumann relaxation parameter fixed at
+      1.0, Richardson relaxation parameter fixed at 1.0, Neumann
+      relaxation parameter fixed at 0.7, Richardson relaxation
+      parameter fixed at 1.1. For each calculation \sn{3}{4}
+      stochastic histories were used to compute the MCSA correction at
+      each iteration.}}
+  \label{fig:relax_time}
 \end{figure}
@@
 
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/spn_equations/estimator_5000_convergence.pdf}
+  \end{center}
+  \caption{\textbf{Residual infinity norm as a function of MCSA
+      iteration with 5,000 stochastic histories per iteration.}
+    \textit{Both the collision and expected value estimators were used
+      with the adjoint Monte Carlo solver to compute the correction at
+      each iteration.}}
+  \label{fig:estimator_5000}
+\end{figure}
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/spn_equations/estimator_1000_convergence.pdf}
+  \end{center}
+  \caption{\textbf{Residual infinity norm as a function of MCSA
+      iteration with 1,000 stochastic histories per iteration.}
+    \textit{Both the collision and expected value estimators were used
+      with the adjoint Monte Carlo solver to compute the correction at
+      each iteration.}}
+  \label{fig:estimator_1000}
+\end{figure}
+
 \clearpage

### 2013-06-01 doc/dissertation/document/chapters/linear_problem/linear_problem.tex
adding data for estimators from hmcsa
updating spn chapter
making the linear problem chapter an appendix

@@
 remainder of this work. Details on the parallelization of conventional
-methods are discussed.\footnote{The contents of this chapter,
+methods are discussed.\footnote{The contents of this appendix,
   particularly those sections relating to projection methods and
@@
   \citep{saad_iterative_2003}.}
-
-\section{Preliminaries}
-\label{sec:linear_preliminaries}
-We seek solutions of the general linear problem in the following form:
-\begin{equation}
-  \ve{A} \ve{x} = \ve{b}\:,
-  \label{eq:linear_problem}
-\end{equation}
-where $\ve{A} \in \mathbb{R}^{N \times N}$ is a matrix operator such
-that $\ve{A} : \mathbb{R}^{N} \rightarrow \mathbb{R}^{N}$, $\ve{x} \in
-\mathbb{R}^N$ is the solution vector, and $\ve{b} \in \mathbb{R}^N$ is
-the forcing term. The solutions to Eq~(\ref{eq:linear_problem}) will
-be generated by inverting $\ve{A}$ either directly or indirectly:
-\begin{equation}
-  \ve{x} = \ve{A}^{-1} \ve{b}
-  \label{eq:linear_problem_solution}\:.
-\end{equation}
-In addition we can define the residual:
-\begin{equation}
-  \ve{r} = \ve{b} - \ve{A}\ve{x}\:,
-  \label{eq:linear_residual}
-\end{equation}
-such that an exact solution $\ve{x}$ has been found when
-$\ve{r}=\ve{0}$.  From the statement in
-Eq~(\ref{eq:linear_problem_solution}) we can already place a
-restriction on $\ve{A}$ by requiring that it be \textit{nonsingular},
-  meaning that we can in fact compute $\ve{A}^{-1}$. In this work we
-  will focus our efforts on approximately inverting the operator
-  through various means.
-
-In a discussion of methods for solving linear systems, several
-mathematical tools are useful in characterizing the qualities of the
-linear system. Among the most useful are the \textit{Eigenvalues} of
-the matrix, $\sigma(\ve{A})$. We find these by solving the Eigenvalue
-problem:
-\begin{equation}
-  \ve{A} \ve{x} = \lambda \ve{x},\ \lambda \in \sigma(\ve{A})\:.
-  \label{eq:eigenvalue_problem}
-\end{equation}
-By writing Eq~(\ref{eq:eigenvalue_problem}) in a different form,
-\begin{equation}
-  (\ve{A} - \lambda \ve{I})\ve{x} = 0 \:,
-  \label{eq:eigenvalue_problem_2}
-\end{equation}
-and demanding that non-trivial solutions for $\ve{x}$ exist, it is
-then required that $|\ve{A} - \lambda \ve{I}| = 0$. Expanding this
-determinant yields a characteristic polynomial in terms of $\lambda$
-with roots that form the set of Eigenvalues, $\sigma(\ve{A})$. Each
-component of $\sigma(\ve{A})$ can then be used to solve
-Eq~(\ref{eq:eigenvalue_problem_2}) for a particular permutation of
-$\ve{x}$. The set of all permutations form the \textit{Eigenvectors}
-of $\ve{A}$. A quantity of particular interest that is computable
-from the eigenvalues of a matrix $\ve{A}$ is the \textit{spectral
-  radius}, $\rho(\ve{A})$, defined by Saad \citep{saad_iterative_2003}
-as:
-\begin{equation}
-  \rho(\ve{A}) = \max_{\lambda \in \sigma(\ve{A})} |\lambda| \:.
-  \label{eq:spectral_radius}
-\end{equation}
-In addition, for problems that have a large scale over which the
-independent variables may exist (e.g. a problem with events on
-timescales ranging from nanoseconds to hours), a good measure of this
-range is supplied by the \textit{stiffness ratio}:
-\begin{equation}
-  Stiffness Ratio = \frac{\max_{\lambda \in \sigma(\ve{A})}
-    |\lambda|}{\min_{\lambda \in \sigma(\ve{A})} |\lambda|}
-\end{equation}
-Those problems that have a wide range of scales in their independent
-variables, which will then be reflected in the operator, will then
-have a large stiffness ratio. We will define such problems with large
-stiffness ratios as \textit{stiff}.
-
-General to both matrices and vectors, \textit{norms} are a mechanism
-for collapsing objects of many elements to a single value. Per
-LeVeque's text \citep{leveque_finite_2007}, the q-norm of a vector is defined
-as:
-\begin{equation}
-  ||\ve{v}||_q = \Bigg[ \sum_{i=1}^N |v_i|^q \Bigg]^{1/q},\ \ve{v} \in
-  \mathbb{R}^N\:,\ q \in \mathbb{Z}^+
-  \label{eq:q_norm}
-\end{equation}
-where ${v_i}$ is the $i^{th}$ component of the vector. Depending on
-the value chosen for $q$, local or global qualities of the vector may
-be obtained. For example, $q=2$ provides the root of a quadrature sum
-of all elements in the vector giving a global measure of the vector
-while $q=\infty$ gives the maximum value in the vector, a local
-quantity that does not give information regarding the other elements
-in the vector.
-
-We can also compute the norm of a matrix by inferring from the norm of
-the vector on which it is operating. Per LeVeque, we search for a
-constant that is equivalent to $||\ve{A}||$:
-\begin{equation}
-  ||\ve{A}\ve{x}|| \leq C ||\ve{x}||\:,
-  \label{eq:matrix_norm_inequality}
-\end{equation}
-where the minimum value of $C$ that satisfies
-Eq~(\ref{eq:matrix_norm_inequality}) is equivalent to $||\ve{A}||$
-and is valid $\forall \ve{x} \in \mathbb{R}^N$. The general
-definition in Eq~(\ref{eq:matrix_norm_inequality}) can be expanded in
-simple terms for common norms including the infinity norm:
-\begin{equation}
-  ||\ve{A}||_{\infty} = \max_{1 \leq i \leq N} \sum^N_{j=1}|a_{ij}|\:,
-  \label{eq:matrix_infinity_norm}
-\end{equation}
-and the 2-norm:
-\begin{equation}
-  ||\ve{A}||_{2} = \sqrt{\rho(\ve{A}^T\ve{A})}\:,
-  \label{eq:matrix_2_norm}
-\end{equation}
-where $\rho$ is the spectral radius as defined in
-Eq~(\ref{eq:spectral_radius}).
-
-Knowing this, we can then define several useful properties of matrices
-including the \textit{condition number}:
-\begin{equation}
-  \kappa(\ve{A}) = ||\ve{A}||\ ||\ve{A}^{-1}||\:,
-  \label{eq:condition_number}
-\end{equation}
-which gives as a metric on assessing how close to singular the system
-is. This is due to the fact $||\ve{A}^{-1}||$ is large near
-singularities (and undefined for a singular matrix) and thus a large
-condition number will be generated. We define such matrices as
-\textit{ill-conditioned}.
 
@@
 Figure~\ref{fig:linear_projection_constraint}.
-
 \begin{figure}[t!]
@@
 \textit{minimized} with respect to the constraints.
-
 Given this minimization condition for the residual, we can form the

### 2013-06-01 doc/dissertation/document/chapters/mc_background/mc_background.tex
additions to spn chapter
making the linear problem chapter an appendix

@@
 
+\section{Preliminaries}
+\label{sec:linear_preliminaries}
+We seek solutions of the general linear problem in the following form:
+\begin{equation}
+  \ve{A} \ve{x} = \ve{b}\:,
+  \label{eq:linear_problem}
+\end{equation}
+where $\ve{A} \in \mathbb{R}^{N \times N}$ is a matrix operator such
+that $\ve{A} : \mathbb{R}^{N} \rightarrow \mathbb{R}^{N}$, $\ve{x} \in
+\mathbb{R}^N$ is the solution vector, and $\ve{b} \in \mathbb{R}^N$ is
+the forcing term. The solutions to Eq~(\ref{eq:linear_problem}) will
+be generated by inverting $\ve{A}$ either directly or indirectly:
+\begin{equation}
+  \ve{x} = \ve{A}^{-1} \ve{b}
+  \label{eq:linear_problem_solution}\:.
+\end{equation}
+In addition we can define the residual:
+\begin{equation}
+  \ve{r} = \ve{b} - \ve{A}\ve{x}\:,
+  \label{eq:linear_residual}
+\end{equation}
+such that an exact solution $\ve{x}$ has been found when
+$\ve{r}=\ve{0}$.  From the statement in
+Eq~(\ref{eq:linear_problem_solution}) we can already place a
+restriction on $\ve{A}$ by requiring that it be \textit{nonsingular},
+meaning that we can in fact compute $\ve{A}^{-1}$. In this work we
+will focus our efforts on approximately inverting the operator through
+various means.
+
+In a discussion of methods for solving linear systems, several
+mathematical tools are useful in characterizing the qualities of the
+linear system. Among the most useful are the \textit{eigenvalues} of
+the matrix, $\sigma(\ve{A})$. We find these by solving the eigenvalue
+problem:
+\begin{equation}
+  \ve{A} \ve{x} = \lambda \ve{x},\ \lambda \in \sigma(\ve{A})\:.
+  \label{eq:eigenvalue_problem}
+\end{equation}
+By writing Eq~(\ref{eq:eigenvalue_problem}) in a different form,
+\begin{equation}
+  (\ve{A} - \lambda \ve{I})\ve{x} = 0 \:,
+  \label{eq:eigenvalue_problem_2}
+\end{equation}
+and demanding that non-trivial solutions for $\ve{x}$ exist, it is
+then required that $|\ve{A} - \lambda \ve{I}| = 0$. Expanding this
+determinant yields a characteristic polynomial in terms of $\lambda$
+with roots that form the set of eigenvalues, $\sigma(\ve{A})$. Each
+component of $\sigma(\ve{A})$ can then be used to solve
+Eq~(\ref{eq:eigenvalue_problem_2}) for a particular permutation of
+$\ve{x}$. The set of all permutations form the \textit{eigenvectors}
+of $\ve{A}$. A quantity of particular interest that is computable from
+the eigenvalues of a matrix $\ve{A}$ is the \textit{spectral radius},
+$\rho(\ve{A})$, defined by Saad \citep{saad_iterative_2003} as:
+\begin{equation}
+  \rho(\ve{A}) = \max_{\lambda \in \sigma(\ve{A})} |\lambda| \:.
+  \label{eq:spectral_radius}
+\end{equation}
+
 %%---------------------------------------------------------------------------%%
-\section{Preliminaries}
+\section{Neumann-Ulam Method}
 \label{sec:mc_preliminaries}
-We begin our discussion of Monte Carlo methods using these texts by
-seeking a solution to Eq~(\ref{eq:linear_problem}). For a given linear
-operator $\ve{A}$, we can use diagonal splitting in a similar manner
-as the stationary method in Eq~(\ref{eq:linear_split_equation2}) to
-define the following operator\footnote{It should be noted that
-  non-diagonal splittings have been recently explored
-  \citep{srinivasan_monte_2010} and have the potential to improve
-  efficiency. However, it was observed in this work that this type of
-  splitting did not improve performance in the asymptotic limit of
-  $\rho(\ve{H}) \rightarrow 1$ for non-trivial problems.}:
+We begin our discussion of Monte Carlo methods by seeking a solution
+to Eq~(\ref{eq:linear_problem}). For a given linear operator $\ve{A}$,
+we can use diagonal splitting in a similar manner as the stationary
+method in Eq~(\ref{eq:linear_split_equation2}) to define the following
+operator\footnote{It should be noted that non-diagonal splittings have
+  been recently explored \citep{srinivasan_monte_2010} and have the
+  potential to improve efficiency. However, it was observed in this
+  work that this type of splitting did not improve performance in the
+  asymptotic limit of $\rho(\ve{H}) \rightarrow 1$ for non-trivial
+  problems.}:
 \begin{equation}
@@
 to the tallies in those states.
-\clearpage
 
@@
 
+\subsubsection{Alternative Fixed Point Iterations}
+\label{subsubsec:alternative_fixed_point}
+
 \subsection{Preconditioning MCSA}
@@
 be necessary for convergence.
-
-\subsection{Alternative Fixed Point Iterations}
-\label{subsec:mcsa_fp_alternatives}
 
@@
 from this bias as the correction generated by the Monte Carlo solve
-will contain large statistical error even without variance reduction.
+will contain large statistical error even without variance
+reduction. For this work, any method that alters the weights and
+probabilities of the Monte Carlo method with the aim of improving
+either iterative performance or time to convergence will be defined as
+a variance reduction scheme.
 

### 2013-06-01 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
making the linear problem chapter an appendix

@@
 With convergence obtained for the fuel assembly problem, the advanced
-preconditioners will next be studied with each Monte Carlo
-estimator. Although the expected value estimator has achieved the best
-iterative performance in the previous comparison and analysis, it is
-possible that the other estimators may perform better from a timing
-perspective due to the sparsity lost by explicit preconditioning. The
-deterministic component of the expected value estimator that couples
-the current state to other local states through the iteration matrix
-stencil requires application of the estimator to potentially several
-orders of magnitude more states at every transition event.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/estimator_5000_convergence.pdf}
-  \end{center}
-  \caption{\textbf{Residual infinity norm as a function of MCSA
-      iteration with 5,000 stochastic histories per iteration.}
+preconditioners will next be studied with both the collision and
+expected value estimators using the best preconditioning and
+relaxation parameter combinations found by the previous
+analysis. Although the expected value estimator has achieved the best
+iterative performance in the previous comparison and analysis for the
+simple Poisson problem in \S~\ref{sec:parameter_estimator_analysis},
+it is possible that the collision estimator may perform better from a
+timing perspective due to the sparsity lost by explicit
+preconditioning. The deterministic component of the expected value
+estimator that couples the current state to other local states through
+the iteration matrix stencil requires application of the estimator to
+potentially several orders of magnitude more states at every
+transition event.
+
+For this study, the fuel assemby problem with a 1-group $SP_1$
+discretization was solved using MCSA with the adjoint Monte Carlo
+solver using both the collision and expected value estimators. Using
+the results from the previous section, a Neumann relaxation parameter
+of 0.7 was used with both estimators. For the collision estimator, a
+Richardson relaxation parameter of 1.1 was used while it was found the
+expected value estimator had the best performance when a a relaxation
+parameter of 1.0 was used instead. In addition, ILUT preconditioning
+with a drop tolerance of \sn{1}{-5} and a fill level of 5 was used
+with a reduced domain approximation fill level of 100 and a threshold
+of \sn{1}{-10} as in the previous analysis. For each estimator, the
+number of stochastic histories used to compute the MCSA correction was
+varied from 25 to 50,000 with the number of MCSA iterations required
+to converge to a tolerance of \sn{1}{-8} recorded for a single
+eigenvalue iteration along with the CPU time required to achieve
+convergence.
+
+Figure~\ref{fig:spn_estimator_iters} gives the iteration count results
+and Figure~\ref{fig:spn_estimator_iters} gives the timing
+results. Remarkably, although the fuel assembly problem is
+significantly more complicated than the simple Poisson problem studied
+in \S~\ref{sec:parameter_estimator_analysis}, the results here are
+largely the same with Figure~\ref{fig:spn_estimator_iters} and
+Figure~\ref{fig:estimator_nh_iters} having identical qualitative
+behavior. Using the expected value estimator creates a relative
+insensitivity to the number of histories used to compute the
+correction. With respect to CPU time, there is a more pronounced
+effect due to the larger density of states created by the explicit
+ILUT preconditioning. Using a reduced domain fill level of 100 creates
+a situation where the expected value estimator is significantly more
+expensive per history to compute that the collision estimator due to
+the coupling of states.
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/spn_equations/estimator_iters.pdf}
+  \end{center}
+  \caption{\textbf{MCSA iterations required to converge the fuel
+      assembly problem to a tolerance of \sn{1}{-8}.} \textit{Both the
+      collision and expected value estimators were used with the
+      adjoint Monte Carlo solver to compute the correction at each
+      iteration.}}
+  \label{fig:spn_estimator_iters}
+\end{figure}
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/spn_equations/estimator_time.pdf}
+  \end{center}
+  \caption{\textbf{Total CPU time in seconds required to converge the
+      fuel assembly problem to a tolerance of \sn{1}{-8}.}
     \textit{Both the collision and expected value estimators were used
@@
       each iteration.}}
-  \label{fig:estimator_5000}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/estimator_1000_convergence.pdf}
+  \label{fig:spn_estimator_time}
+\end{figure}
+
+From both Figure~\ref{fig:spn_estimator_iters} and
+Figure~\ref{fig:spn_estimator_iters} we see that both estimators have
+comparable iterative and timing performance when 1,000 histories are
+used at each MCSA iteration. Using 1,000 histories per iteration, the
+infinity norm of the residual vector is presented at every MCSA
+iteration for both estimators in
+Figure~\ref{fig:spn_estimator_convergence} as an additional means of
+comparison. At this number of histories, both estimators converge the
+fuel assembly problem monotonically at and approximately the same
+rate, giving little reason to choose one over the other. The end
+result is that both estimators provide a viable method for getting
+good MCSA iterative performance with reasonable timing
+performance. For the collision estimator, using around 20,000
+stochastic histories at every iteration gave the best iterative
+performance while 500 histories per iteration gave the best iterative
+performance when using the expected value estimator. For purely serial
+computational performance, there is not a distinguishing feature for
+the fuel assembly problem presented that would cause one to choose one
+of the estimators over the other.
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/spn_equations/estimator_convergence.pdf}
   \end{center}
@@
       each iteration.}}
-  \label{fig:estimator_1000}
+  \label{fig:spn_estimator_convergence}
 \end{figure}
@@
 
-\subsection{Results}
+\subsection{MCSA Verification}
+\label{subsec:spn_mcsa_verification}
+
+\subsection{Performance Results}
 \label{subsec:spn_comparison_results}

### 2013-06-01 doc/dissertation/document/chapters/mc_background/mc_background.tex
adding to spn estimator analysis section

@@
 \label{subsubsec:alternative_fixed_point}
+In addition to the basic Richardson iteration, the MCSA algorithm
+presented in Eq~(\ref{eq:mcsa}) can be used to accelerate any fixed
+point iteration which only depends on the state (typically the
+residual) of the last iteration. As outlined in
+Appendix~\ref{ch:linear_problem}, subspace methods take on a general
+form in one-dimension using the Petrov-Galerkin conditions as
+constraints for extracting a correction from the search subspace as
+given by Eq~(\ref{eq:linear_projection_iteration}) that may be used
+with MCSA. As an example, for positive-definite but not necessarily
+symmetric problems, the minimal residual (MINRES) iteration
+\citep{saad_iterative_2003} can be used in conjuction with MCSA where
+the residual vector, $\mathbf{r}$, defines the search subspace and the
+action of the linear operator on the residual vector,
+$\mathbf{A}\mathbf{r}$, defines the constraint subspace. Using this,
+we can then define an MCSA scheme that accelerates the MINRES
+iteration:
+\begin{subequations}
+  \begin{gather}
+    \alpha = \frac{\langle \mathbf{A}\mathbf{r}^k, \mathbf{r}^k
+      \rangle}{\langle \mathbf{A}\mathbf{r}^k, \mathbf{A}\mathbf{r}^k
+      \rangle}\:, \\
+    \ve{x}^{k+1/2} = \ve{x}^k + \alpha \ve{r}^k\:,
+    \label{eq:minres_correction}\\
+    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\ 
+    \ve{A}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:,\\ 
+    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta\ve{x}^{k+1/2}\:,\\ 
+    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:.
+  \end{gather}
+  \label{eq:mcsa_minres}
+\end{subequations}
+where $\alpha$ is an optimal extrapolation parameter generated the
+constraints. Interestingly, this scheme is nearly identical to the
+Richardson iteration version in Eq~(\ref{eq:mcsa}) except now the
+additional extrapolation parameter, $\alpha$, is applied to the
+residual correction in Eq~(\ref{eq:minres_correction}) as a means of
+selecting a more optimal search direction at each iteration,
+potentially further accelerating convergence.
 

### 2013-06-01 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
adding to spn estimator analysis section

@@
-\chapter{Monte Carlo Solution Methods Applied to Nonlinear Systems}
+\chapter{Monte Carlo Solution Methods Applied to the Navier-Stokes Equations}
 \label{ch:nonlinear_problem}
-Nonlinear equation sets are a common occurrence in multiphysics
-problems. Systems of partial differential equations such as those that
-describe fluid flow or more general transport processes when
-discretized by conventional methods yield discrete sets of stiff
-equations with nonlinearities present in the variables. Traditionally,
-such systems have been solved by linearizing them in a form where the
-nonlinearities in the variables are eliminated and more traditional
-linear methods can be used for solutions. Often characterized as
-segregated methods where physics operators are split and their action
-on the system approximated in steps, such methods lack consistency and
-accuracy in resolving the nonlinear component of the solution. In the
-last 30 years, fully consistent nonlinear methods based on Newton's
-method have become more popular and many advances have been made in
-the computational physics field to employ these methods.
+Nonlinear transport problems are a common occurrence in single and
+multiple physics problems. Systems of partial differential equations
+such as those that describe fluid flow or more general transport
+processes when discretized by conventional methods yield discrete sets
+of stiff equations with nonlinearities present in the
+variables. Traditionally, such systems have been solved by linearizing
+them in a form where the nonlinearities in the variables are
+eliminated and more traditional linear methods can be used for
+solutions. Often characterized as segregated methods where physics
+operators are split and their action on the system approximated in
+steps, such methods lack consistency and accuracy in resolving the
+nonlinear component of the solution. In the last 30 years, fully
+consistent nonlinear methods based on Newton's method have become more
+popular and many advances have been made in the computational physics
+field to employ these methods.
 

### 2013-06-01 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding to spn estimator analysis section

@@
 relaxation parameter combinations found by the previous
-analysis. Although the expected value estimator has achieved the best
-iterative performance in the previous comparison and analysis for the
-simple Poisson problem in \S~\ref{sec:parameter_estimator_analysis},
-it is possible that the collision estimator may perform better from a
-timing perspective due to the sparsity lost by explicit
-preconditioning. The deterministic component of the expected value
-estimator that couples the current state to other local states through
-the iteration matrix stencil requires application of the estimator to
-potentially several orders of magnitude more states at every
-transition event.
+analysis. Additionally, both the Richardson-based MCSA iteration and
+the MINRES-based MCSA iteration given by Eq~(\ref{eq:mcsa_minres})
+will be used. Although the expected value estimator has achieved the
+best iterative performance in the previous comparison and analysis for
+the simple Poisson problem in
+\S~\ref{sec:parameter_estimator_analysis}, it is possible that the
+collision estimator may perform better from a timing perspective due
+to the sparsity lost by explicit preconditioning. The deterministic
+component of the expected value estimator that couples the current
+state to other local states through the iteration matrix stencil
+requires application of the estimator to potentially several orders of
+magnitude more states at every transition event.
 
@@
 discretization was solved using MCSA with the adjoint Monte Carlo
-solver using both the collision and expected value estimators. Using
-the results from the previous section, a Neumann relaxation parameter
-of 0.7 was used with both estimators. For the collision estimator, a
-Richardson relaxation parameter of 1.1 was used while it was found the
-expected value estimator had the best performance when a a relaxation
-parameter of 1.0 was used instead. In addition, ILUT preconditioning
-with a drop tolerance of \sn{1}{-5} and a fill level of 5 was used
-with a reduced domain approximation fill level of 100 and a threshold
-of \sn{1}{-10} as in the previous analysis. For each estimator, the
-number of stochastic histories used to compute the MCSA correction was
-varied from 25 to 50,000 with the number of MCSA iterations required
-to converge to a tolerance of \sn{1}{-8} recorded for a single
-eigenvalue iteration along with the CPU time required to achieve
-convergence.
+solver using both the collision and expected value estimators and both
+the Richardson and MINRES fixed point iterations. Using the results
+from the previous section, a Neumann relaxation parameter of 0.7 was
+used with both estimators and fixed point iterations. For the
+collision estimator, a Richardson relaxation parameter of 1.1 was used
+with the Richardson iteration while it was found the expected value
+estimator had the best performance when a a relaxation parameter of
+1.0 was used instead. In addition, ILUT preconditioning with a drop
+tolerance of \sn{1}{-5} and a fill level of 5 was used with a reduced
+domain approximation fill level of 100 and a threshold of \sn{1}{-10}
+as in the previous analysis. For each estimator, the number of
+stochastic histories used to compute the MCSA correction was varied
+from 25 to 50,000 with the number of MCSA iterations required to
+converge to a tolerance of \sn{1}{-8} recorded for a single eigenvalue
+iteration along with the CPU time required to achieve convergence.
 
 Figure~\ref{fig:spn_estimator_iters} gives the iteration count results
-and Figure~\ref{fig:spn_estimator_iters} gives the timing
+and Figure~\ref{fig:spn_estimator_time} gives the timing
 results. Remarkably, although the fuel assembly problem is
@@
 Figure~\ref{fig:estimator_nh_iters} having identical qualitative
-behavior. Using the expected value estimator creates a relative
-insensitivity to the number of histories used to compute the
-correction. With respect to CPU time, there is a more pronounced
+behavior. Using the expected value estimator with both fixed point
+iterations creates a relative insensitivity to the number of histories
+used to compute the correction. For the collision estimator, the
+MINRES iteration permits fewer stochastic histories to be used with
+each MCSA iteration while still maintaining good iterative
+performance. With respect to CPU time, there is a more pronounced
 effect due to the larger density of states created by the explicit
@@
       adjoint Monte Carlo solver to compute the correction at each
-      iteration.}}
+      iteration. Each estimator was used with the Richardson and
+      MINRES fixed point iterations within an MCSA iteration.}}
   \label{fig:spn_estimator_iters}
@@
       with the adjoint Monte Carlo solver to compute the correction at
-      each iteration.}}
+      each iteration. Each estimator was used with the Richardson and
+      MINRES fixed point iterations within an MCSA iteration.}}
   \label{fig:spn_estimator_time}
@@
 From both Figure~\ref{fig:spn_estimator_iters} and
-Figure~\ref{fig:spn_estimator_iters} we see that both estimators have
+Figure~\ref{fig:spn_estimator_time} we see that both estimators have
 comparable iterative and timing performance when 1,000 histories are
@@
       with the adjoint Monte Carlo solver to compute the correction at
-      each iteration.}}
+      each iteration. Each estimator was used with the Richardson and
+      MINRES fixed point iterations within an MCSA iteration.}}
   \label{fig:spn_estimator_convergence}

### 2013-06-02 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
starting work on spn solvers comparison section

@@
 relaxation parameter combinations found by the previous
-analysis. Additionally, both the Richardson-based MCSA iteration and
-the MINRES-based MCSA iteration given by Eq~(\ref{eq:mcsa_minres})
-will be used. Although the expected value estimator has achieved the
-best iterative performance in the previous comparison and analysis for
-the simple Poisson problem in
+analysis. Additionally, both the Richardson-based MCSA iteration given
+by Eq~(\ref{eq:mcsa}) and the MINRES-based MCSA iteration given by
+Eq~(\ref{eq:mcsa_minres}) will be used. Although the expected value
+estimator achieved the best iterative performance in the previous
+comparison and analysis for the simple Poisson problem in
 \S~\ref{sec:parameter_estimator_analysis}, it is possible that the
@@
 with the Richardson iteration while it was found the expected value
-estimator had the best performance when a a relaxation parameter of
-1.0 was used instead. In addition, ILUT preconditioning with a drop
-tolerance of \sn{1}{-5} and a fill level of 5 was used with a reduced
-domain approximation fill level of 100 and a threshold of \sn{1}{-10}
-as in the previous analysis. For each estimator, the number of
-stochastic histories used to compute the MCSA correction was varied
-from 25 to 50,000 with the number of MCSA iterations required to
-converge to a tolerance of \sn{1}{-8} recorded for a single eigenvalue
-iteration along with the CPU time required to achieve convergence.
+estimator had the best performance when a Richardson relaxation
+parameter of 1.0 was used instead. In addition, ILUT preconditioning
+with a drop tolerance of \sn{1}{-5} and a fill level of 5 was used
+with a reduced domain approximation fill level of 100 and a threshold
+of \sn{1}{-10} as in the previous analysis. For each estimator, the
+number of stochastic histories used to compute the MCSA correction was
+varied from 25 to 50,000 with the number of MCSA iterations required
+to converge to a tolerance of \sn{1}{-8} recorded for a single
+eigenvalue iteration along with the CPU time required to achieve
+convergence.
 
@@
 iterations creates a relative insensitivity to the number of histories
-used to compute the correction. For the collision estimator, the
-MINRES iteration permits fewer stochastic histories to be used with
-each MCSA iteration while still maintaining good iterative
-performance. With respect to CPU time, there is a more pronounced
-effect due to the larger density of states created by the explicit
-ILUT preconditioning. Using a reduced domain fill level of 100 creates
-a situation where the expected value estimator is significantly more
-expensive per history to compute that the collision estimator due to
-the coupling of states.
+used to compute the correction. Marginally better iterative and timing
+performance was observed for the expected value estimator when using
+the Richardson iteration. Better timing performance is expected as
+computing the extrapolation parameter in the MINRES iteration requires
+several additional parallel operations. However, we expect that
+because the MINRES iteration will converge faster on its own when
+compared to the Richardson iteration that better iterative performance
+will be observed when it is accelerated with MCSA. The fact that this
+was not observed signals that an MCSA scheme leveraging the expected
+value estimator is also effectively insensitive to the fixed point
+iteration used. For the collision estimator, the MINRES iteration
+permits fewer stochastic histories to be used with each MCSA iteration
+while still maintaining good iterative performance. With respect to
+CPU time, there is a more pronounced effect due to the larger density
+of states created by the explicit ILUT preconditioning. Using a
+reduced domain fill level of 100 creates a situation where the
+expected value estimator is significantly more expensive per history
+to compute that the collision estimator due to the coupling of states.
 
@@
 From both Figure~\ref{fig:spn_estimator_iters} and
-Figure~\ref{fig:spn_estimator_time} we see that both estimators have
-comparable iterative and timing performance when 1,000 histories are
-used at each MCSA iteration. Using 1,000 histories per iteration, the
-infinity norm of the residual vector is presented at every MCSA
-iteration for both estimators in
+Figure~\ref{fig:spn_estimator_time} we see that both estimators and
+fixed point iterations have comparable iterative and timing
+performance when 1,000 histories are used at each MCSA
+iteration. Using 1,000 histories per MCSA iteration, the infinity norm
+of the residual vector is presented at every MCSA iteration for both
+estimators and both fixed point iterations in
 Figure~\ref{fig:spn_estimator_convergence} as an additional means of
-comparison. At this number of histories, both estimators converge the
-fuel assembly problem monotonically at and approximately the same
-rate, giving little reason to choose one over the other. The end
-result is that both estimators provide a viable method for getting
-good MCSA iterative performance with reasonable timing
+comparison. At this number of histories, all MCSA combinations
+converge the fuel assembly problem monotonically at and approximately
+the same rate, giving little reason to choose one over the other. The
+end result is that both estimators provide a viable method for getting
+good MCSA iterative performance with comparable timing
 performance. For the collision estimator, using around 20,000
@@
 the fuel assembly problem presented that would cause one to choose one
-of the estimators over the other.
+of the estimators and fixed point iterations over the other as they
+all have similar minimum CPU times over the range of values tested.
 
@@
 \label{sec:spn_comparison}
+Through the numerical studies presented up to this point in this
+chapter, we have developed a set of preconditioning techniques that
+permit MCSA to be used with the $SP_N$ form of the neutron transport
+equation and applied to a difficult fuel assembly criticality
+problem. Along with these preconditioning techniques, a set a varying
+MCSA parameters were analyzed to find those that yielded the best
+performance for this particular problem. In addition, several steps
+were taken to mitigate the dense composite linear operators that arise
+when performing explicit MCSA preconditioning in order to achieve
+convergence. Using these results, we compare MCSA directly to
+conventional linear solvers that would be typically used to solve the
+$SP_N$ form of the transport equation as presented here. In
+particular, we will use a pair of production Krylov solvers with which
+we will compare direct numerical results in order to verify MCSA
+solutions and iterative and CPU timing performance in order to gauge
+feasibility and identify areas that require further research.
 
@@
 \label{subsec:spn_mcsa_verification}
+In order to verify the correctness of the MCSA method and
+implementation used for this work, MCSA solutions to the fuel assembly
+problem with 1, 2, and 4 energy groups using $SP_1$ discretization
+will be directly compared over all group and cell fluxes computed by
+converging the criticality problem using both BiCGStab and GMRES
+production implementations from the Trilinos scientific computing
+library Aztec \citep{heroux_overview_2005}. In addition to comparing
+the cell and group fluxes, the number of eigenvalue iterations
+required to converge the problem along with the eigenvalue itself will
+be compared as a means of verification. For the verification, all
+solvers will be preconditioned with ILUT using a drop tolerance of
+\sn{1}{-5} and a fill level of 5. For GMRES, no restrictions were
+placed on the size of the subspace and therefore no restarts
+occurred. When the collision estimator was used with MCSA, 20,000
+stochastic histories were used to compute the correction for every
+energy group in the problem (40,000 and 80,000 histories total for the
+2 and 4 group calculations respectively) and 500 stochastic histories
+used for each energy group when the expected value estimator was used
+(1,000 and 1,500 histories total for the 2 and 4 group
+calculations). The reduced domain approximation was also applied in
+conjuction with the ILUT preconditioning using a fill level of 100 and
+a threshold value of \sn{1}{-10} to reduce the density of states in
+the Monte Carlo problem. For relaxation parameters, all MCSA
+compuations used a Neumann relaxation parameter of 0.7 while a
+Richardson relaxation of 1.1 was used with the collision estimator and
+1.0 with the expected value estimator as determined by the previous
+analysis. Table~\ref{tab:spn_solver_defs} gives definitions for the
+solvers used to generate the results in the remainder of this section.
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{ll}\hline\hline
+      \multicolumn{1}{l}{\textbf{Name}} & 
+      \multicolumn{1}{l}{\textbf{Definition}} \\
+      BiCGStab-ILUT & ILUT preconditioned BiCGStab \\
+      GMRES-ILUT & ILUT preconditioned GMRES \\
+      MCSA-ILUT-R-C & ILUT preconditioned MCSA with Richardson \\ 
+                    & fixed point iteration and collision estimator \\
+      MCSA-ILUT-MR-C & ILUT preconditioned MCSA with MINRES \\
+                     & fixed point iteration and collision estimator \\
+      MCSA-ILUT-R-EV & ILUT preconditioned MCSA with Richardson \\
+                     & fixed point iteration and expected value estimator \\
+      MCSA-ILUT-MR-EV & ILUT preconditioned MCSA with MINRES \\
+                      & fixed point iteration and expected value estimator \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Solver definitions used for MCSA verification and
+      performance analysis.}}
+  \label{tab:spn_solver_defs}
+\end{table}
+For the energy group structures,
+Table~\ref{tab:spn_group_structure} gives the lower bounds of each
+group in eV (an implicit upper bound of 20 MeV is assumed).
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{cl}\hline\hline
+      \multicolumn{1}{c}{\textbf{Number of Groups}} & 
+      \multicolumn{1}{l}{\textbf{Lower Bounds (eV)}} \\
+      1 & \{ \sn{1}{-5} \} \\
+      2 & \{ \sn{1}{-1}, \sn{1}{-5} \} \\
+      4 & \{ \sn{1}{1}, \sn{1}{0}, \sn{1}{-1}, \sn{1}{-5} \} \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Energy group lower bounds for the multigroup fuel
+      assembly criticality problem in electron volts.}}
+  \label{tab:spn_group_structure}
+\end{table}
 
@@
 \label{subsec:spn_comparison_results}
+Using the same set of solver parameters used for verification, MCSA
+will be directly compared to the production Krylov solvers in terms of
+both iterative performance and CPU timing. In addition to the
+parameters used for the verification analysis, artificial absorption
+was also introduced with a value of 0.2 for the MCSA calculations that
+used to collision esitmator as CPU timing performance was improved
+without a loss in iterative performance. Adding this absorption did
+not change the numerical results of the
+calculation. Figure~\ref{fig:spn_comparison_iterations} gives the
+average number of linear solver iterations required to converge a
+single eigenvalue iteration of the fuel assembly problem as a function
+of the number of energy groups.
 
@@
       variations of fixed point iterations and estimators were used.}}
-  \label{fig:comparison_iterations}
+  \label{fig:spn_comparison_iterations}
 \end{figure}
@@
       variations of fixed point iterations and estimators were used.}}
-  \label{fig:comparison_time}
+  \label{fig:spn_comparison_time}
 \end{figure}
@@
       variations of fixed point iterations and estimators were used.}}
-  \label{fig:comparison_prec_time}
-\end{figure}
+  \label{fig:spn_comparison_prec_time}
+\end{figure}

### 2013-06-02 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
finished draft of solver results section

@@
 was also introduced with a value of 0.2 for the MCSA calculations that
-used to collision esitmator as CPU timing performance was improved
+used the collision esitmator as CPU timing performance was improved
 without a loss in iterative performance. Adding this absorption did
 not change the numerical results of the
-calculation. Figure~\ref{fig:spn_comparison_iterations} gives the
-average number of linear solver iterations required to converge a
-single eigenvalue iteration of the fuel assembly problem as a function
-of the number of energy groups.
+calculation.
+
+Figure~\ref{fig:spn_comparison_iterations} gives the average number of
+linear solver iterations required to converge a single eigenvalue
+iteration of the fuel assembly problem as a function of the number of
+energy groups. In general, the iterative performance of all the
+solvers tested is comparable with BiCGStab performing the best. We
+expect this because not only are the multigroup $SP_N$ equations
+elliptic and positive-definite, but they are also nearly symmetric and
+therefore we expect the conjugate gradient-based methods to perform
+well. It also important to note that the iterative performance of each
+solver is not a strong function of the number of energy groups in the
+problem (and therefore the problem size). For MCSA, we are linearly
+increasing the number of stochastic histories used to compute the
+correction at each iteration as the number of energy groups are
+increased and therefore the work performed for the problem is of
+$O(N)$. It is also important to note the the reduced domain
+approximation parameters, particularly the fill level, were fixed as
+the number of energy groups was increased. By fixing the fill level,
+we are effectively fixing the amount of information contained in the
+composite linear operator available for the Monte Carlo
+problem. Limiting this information to 100 entries per row in the
+preconditioned system did not appear to have a significant effect on
+the iterative performance of the solver. For larger numbers of energy
+groups (and therefore DOFs) this may be an issue and this number may
+need to be increased to maintain iterative performance. Currently,
+memory restrictions arising from the explicit preconditioning strategy
+prevent larger energy groups to be analyzed using the appropriate
+preconditioning for the fuel assembly problem.
 
@@
 
+Although iterative performance for MCSA was comparable to that
+observed for production Krylov implementations using exactly the same
+preconditioning scheme, timing performance is not as
+comparable. Figure~\ref{fig:spn_comparison_time} gives the average CPU
+time per linear solver iteration required to converge the fuel
+assembly problem with all linear solver iterations over all eigenvalue
+iterations used to compute the average. The production Krylov solver
+implementations are approximately an order of magnitude faster than
+the MCSA implementations presented here. We expect these results for
+several reasons. First, even when the reduced domain approximation is
+used there are O(100) elements in each row of the system and each of
+those elements must be processed during a transition in the Monte
+Carlo random walk sequence. In addition, although the preconditioned
+fixed point iterations used in the MCSA sequence do not use the
+composite linear operator but rather a sequence of matrix vector
+multiplies to achieve the same effect, they do use the explicitly
+extracted inverses of the preconditioning matrices which themselves
+are dense, leading to exceedingly slow computation times even in the
+fixed point iteration. Second, the implementation presented here has
+not been optimized and it is likely several components of the
+algorithm can be implemented in a more desireable time complexity. The
+production Krylov solvers presented here have enjoyed decades of
+professional development and optimization. However, it is most
+important to note here that the CPU timing data presented here shows
+that all methods have the same time complexity; their differences are
+merely the manifestation of a large time constant due to the effects
+of the explicit preconditioning scheme. For transport systems where
+this is not a problem, the literature explcitly shows that MCSA can be
+competetive with these production Krylov methods using CPU time as a
+measure \citep{evans_monte_2012}.
+
 \begin{figure}[t!]
@@
   \end{center}
-  \caption{\textbf{Average CPU time per iteration in seconds for the
-      fuel assembly problem as a function of energy groups.}
-    \textit{All linear solver iterations over all eigenvalue
-      iterations were used to compute the averge. All methods were
-      preconditioned with identical ILUT parameters. For MCSA, several
-      variations of fixed point iterations and estimators were used.}}
+  \caption{\textbf{Average CPU time per linear solver iteration in
+      seconds for the fuel assembly problem as a function of energy
+      groups.}  \textit{All linear solver iterations over all
+      eigenvalue iterations were used to compute the averge. All
+      methods were preconditioned with identical ILUT parameters. For
+      MCSA, several variations of fixed point iterations and
+      estimators were used.}}
   \label{fig:spn_comparison_time}
 \end{figure}
+
+Finally, not considered in Figure~\ref{fig:spn_comparison_time} is the
+time required to actually form the explicit inverse preconditioner
+matrices and the composite operator through matrix-matrix
+multiplication. The timing numbers reported were simply to perform the
+CMSA iteration procedure with the composite operator already
+formed. Figure~\ref{fig:spn_comparison_prec_time} additionally
+presents the CPU times for MCSA convergence with the time to form the
+preconditioners and the composite linear operator through
+matrix-matrix multiplication amortized over all iterations. As is
+readily observed including the costs of these operations increases the
+MCSA computation time by another order of magnitude.
 
@@
 \end{figure}
+
+Based on the results in this section, MCSA shows promise as a
+competitive and perhaps even superior method for solutions to the
+neutron transport problem discretized with the $SP_N$
+approximation. Not only are the correct answers produced when compared
+to production linear solvers, but the iterative performance is
+comparable to production Krylov methods with identical preconditioning
+that would typically be used every-day calculations. From a CPU timing
+perspective, the methods yield the same time complexity as a function
+of the number of energy groups in the problem when compared to the
+production Krylov methods. Here, a large time constant is generated
+due to the explicit preconditioning strategy and the generation of
+dense linear operators as a result. To improve these results and put
+general MCSA schemes into a performance regime where they are
+competitive with Krylov methods for neutron transport problems,
+significant research will be required to improve upon the explicit
+preconditioning shceme presented here.

### 2013-06-03 doc/dissertation/document/chapters/mc_background/mc_background.tex
updating spn chapter

@@
 \end{subequations}
-where $\alpha$ is an optimal extrapolation parameter generated the
-constraints. Interestingly, this scheme is nearly identical to the
+where $\alpha$ is an optimal extrapolation parameter generated from
+the constraints. Interestingly, this scheme is nearly identical to the
 Richardson iteration version in Eq~(\ref{eq:mcsa}) except now the
@@
 iterative scheme. The effects of the reduced domain approximation on
-systems where it is required due to the explicit preconditioning
-strategy chosen will be presented in Chapter~\ref{ch:spn_equations}.
+neutron transport systems where it is required due to the explicit
+preconditioning strategy will be presented in
+Chapter~\ref{ch:spn_equations}.
 

### 2013-06-03 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
updating spn chapter

@@
 yields the $SP_3$ equations). Boundary conditions for these equations
-are provided in Appendix~\ref{chap:spn_boundary_conditions}. The
-multigroup form is presented in Appendix~\ref{chap:spn_mg_equations}
+are provided in Appendix~\ref{chap:spn_boundary_conditions}, the
+multigroup form is presented in Appendix~\ref{chap:mg_spn_equations},
 and the spatial discretization in
@@
 required to converge the problem along with the eigenvalue itself will
-be compared as a means of verification. For the verification, all
-solvers will be preconditioned with ILUT using a drop tolerance of
-\sn{1}{-5} and a fill level of 5. For GMRES, no restrictions were
-placed on the size of the subspace and therefore no restarts
-occurred. When the collision estimator was used with MCSA, 20,000
-stochastic histories were used to compute the correction for every
-energy group in the problem (40,000 and 80,000 histories total for the
-2 and 4 group calculations respectively) and 500 stochastic histories
-used for each energy group when the expected value estimator was used
-(1,000 and 1,500 histories total for the 2 and 4 group
+be compared as a means of verification.
+
+For the verification, all solvers will be preconditioned with ILUT
+using a drop tolerance of \sn{1}{-5} and a fill level of 5. For GMRES,
+no restrictions were placed on the size of the subspace and therefore
+no restarts occurred. When the collision estimator was used with MCSA,
+20,000 stochastic histories were used to compute the correction for
+every energy group in the problem (40,000 and 80,000 histories total
+for the 2 and 4 group calculations respectively) and 500 stochastic
+histories used for each energy group when the expected value estimator
+was used (1,000 and 1,500 histories total for the 2 and 4 group
 calculations). The reduced domain approximation was also applied in
@@
 analysis. Table~\ref{tab:spn_solver_defs} gives definitions for the
-solvers used to generate the results in the remainder of this section.
+solvers used to generate the results in the remainder of this
+section. For the energy group structures,
+Table~\ref{tab:spn_group_structure} gives the lower bounds of each
+group in eV (an implicit upper bound of 20 MeV is assumed).
 \begin{table}[h!]
@@
       \multicolumn{1}{l}{\textbf{Definition}} \\
-      BiCGStab-ILUT & ILUT preconditioned BiCGStab \\
-      GMRES-ILUT & ILUT preconditioned GMRES \\
-      MCSA-ILUT-R-C & ILUT preconditioned MCSA with Richardson \\ 
+      BiCGStab-ILUT & BiCGStab preconditioned with ILUT \\
+      GMRES-ILUT & GMRES preconditioned with ILUT \\
+      MCSA-ILUT-R-C & MCSA preconditioned with ILUT using Richardson \\ 
                     & fixed point iteration and collision estimator \\
-      MCSA-ILUT-MR-C & ILUT preconditioned MCSA with MINRES \\
+      MCSA-ILUT-MR-C & MCSA preconditioned with ILUT using MINRES \\
                      & fixed point iteration and collision estimator \\
-      MCSA-ILUT-R-EV & ILUT preconditioned MCSA with Richardson \\
+      MCSA-ILUT-R-EV & MCSA preconditioned with ILUT using Richardson \\
                      & fixed point iteration and expected value estimator \\
-      MCSA-ILUT-MR-EV & ILUT preconditioned MCSA with MINRES \\
+      MCSA-ILUT-MR-EV & MCSA preconditioned with ILUT using MINRES \\
                       & fixed point iteration and expected value estimator \\
@@
 \end{table}
-For the energy group structures,
-Table~\ref{tab:spn_group_structure} gives the lower bounds of each
-group in eV (an implicit upper bound of 20 MeV is assumed).
 \begin{table}[h!]
@@
 memory restrictions arising from the explicit preconditioning strategy
-prevent larger energy groups to be analyzed using the appropriate
-preconditioning for the fuel assembly problem.
+prevent larger energy groups to be analyzed using an appropriate
+preconditioner for the fuel assembly problem.
 
@@
       each eigenvalue iteration for the fuel assembly problem as a
-      function of energy groups.} \textit{All methods were
-      preconditioned with identical ILUT parameters. For MCSA, several
-      variations of fixed point iterations and estimators were used.}}
+      function of energy groups.}
+    \textit{Table~\ref{tab:spn_solver_defs} gives the description for
+      each solver type presented in the legend.}}
   \label{fig:spn_comparison_iterations}
@@
 several reasons. First, even when the reduced domain approximation is
-used there are O(100) elements in each row of the system and each of
+used there are $O(100)$ elements in each row of the system and each of
 those elements must be processed during a transition in the Monte
@@
 composite linear operator but rather a sequence of matrix vector
-multiplies to achieve the same effect, they do use the explicitly
-extracted inverses of the preconditioning matrices which themselves
-are dense, leading to exceedingly slow computation times even in the
-fixed point iteration. Second, the implementation presented here has
-not been optimized and it is likely several components of the
-algorithm can be implemented in a more desireable time complexity. The
+multiplies to achieve the same preconditioning effect, they do use the
+explicitly extracted inverse of the preconditioning matrices which
+themselves are dense, leading to exceedingly slow computation times
+even in the fixed point iteration. For the production Krylov methods,
+the ILUT preconditioners are applied to a vector with two triangular
+solves, one for each triangular factor. Also, the original linear
+operator has $O(10)$ elements in each row of the system, an order of
+magnitude less than that used in the reduced domain composite
+operator. Second, the implementation presented here has not been
+optimized and it is likely that several components of the algorithm
+can be implemented in a more desireable time complexity. The
 production Krylov solvers presented here have enjoyed decades of
 professional development and optimization. However, it is most
-important to note here that the CPU timing data presented here shows
-that all methods have the same time complexity; their differences are
-merely the manifestation of a large time constant due to the effects
-of the explicit preconditioning scheme. For transport systems where
-this is not a problem, the literature explcitly shows that MCSA can be
-competetive with these production Krylov methods using CPU time as a
-measure \citep{evans_monte_2012}.
+important to note here that the CPU timing data presented in
+Figure~\ref{fig:spn_comparison_time} shows that all methods have the
+same time complexity; their differences are merely the manifestation
+of a large time constant due to the effects of the explicit
+preconditioning scheme. For transport systems where this is not a
+problem, the literature explicitly shows that MCSA can be competetive
+with these production Krylov methods using CPU time as a measure
+\citep{evans_monte_2012}.
 
@@
       groups.}  \textit{All linear solver iterations over all
-      eigenvalue iterations were used to compute the averge. All
-      methods were preconditioned with identical ILUT parameters. For
-      MCSA, several variations of fixed point iterations and
-      estimators were used.}}
+      eigenvalue iterations were used to compute the
+      average. Table~\ref{tab:spn_solver_defs} gives the description
+      for each solver type presented in the legend.}}
   \label{fig:spn_comparison_time}
@@
 multiplication. The timing numbers reported were simply to perform the
-CMSA iteration procedure with the composite operator already
+MCSA iteration procedure with the composite operator already
 formed. Figure~\ref{fig:spn_comparison_prec_time} additionally
 presents the CPU times for MCSA convergence with the time to form the
-preconditioners and the composite linear operator through
-matrix-matrix multiplication amortized over all iterations. As is
-readily observed including the costs of these operations increases the
-MCSA computation time by another order of magnitude.
+inverse of the preconditioners and the composite linear operator
+through matrix-matrix multiplication amortized over all iterations. As
+is readily observed, including the costs of these operations increases
+the MCSA computation time by another order of magnitude.
 
@@
     \textit{All linear solver iterations over all eigenvalue
-      iterations were used to compute the averge. All methods were
-      preconditioned with identical ILUT parameters. For MCSA, several
-      variations of fixed point iterations and estimators were used.}}
+      iterations were used to compute the
+      average. Table~\ref{tab:spn_solver_defs} gives the description
+      for each solver type presented in the legend. Data labeled
+      starting with PMCSA is identical to those labeled with MCSA
+      except that they additionally include the cost of generating the
+      inverse of the preconditioners and composite linear operator.}}
   \label{fig:spn_comparison_prec_time}
@@
 
-Based on the results in this section, MCSA shows promise as a
-competitive and perhaps even superior method for solutions to the
+Based on the performance results in this section, MCSA shows promise
+as a competitive and perhaps even superior method for solutions to the
 neutron transport problem discretized with the $SP_N$
@@
 significant research will be required to improve upon the explicit
-preconditioning shceme presented here.
+preconditioning scheme presented here.

### 2013-06-03 doc/dissertation/document/chapters/introduction/introduction.tex
adding additional pngs to symmetric spectral analysis for communication
updates to introduction chapter

@@
 \label{ch:introduction}
-In many fields of engineering and physics, linear and nonlinear
-problems are a primary focus of study. Recent focus on multiple
-physics systems in the nuclear reactor modeling and simulation
-community adds a new level of complexity to common nonlinear systems
-as solution strategies change when they are coupled to other problems
-\citep{u.s._department_of_energy_casl_2011}. Furthermore, a desire
-for predictive simulations to enhance the safety and performance of
+In nearly all high-fidelity nuclear reactor simulations, linear and
+nonlinear transport problems are a primary focus of study. Recent
+focus on multiple physics systems in the nuclear reactor modeling and
+simulation community adds a new level of complexity to common linear
+and nonlinear systems as solution strategies change when they are
+coupled to other problems
+\citep{u.s._department_of_energy_casl_2011}. Furthermore, a desire for
+predictive simulations to enhance the safety and performance of
 nuclear systems creates a need for extremely high fidelity
-computations to be performed for these coupled systems as a means to
-capture effects not modeled by coarser methods.
+computations to be performed for these coupled transport systems as a
+means to capture effects not modeled by coarser methods.
 
@@
 
-We discuss in this chapter motivation for advancing Monte Carlo
-solvers by providing multiphysics problems of interest in nuclear
-reactor analysis. Hardware-based motivations are also provided by
-considering the impact of forthcoming computing architectures. In
-addition, background on the current solver techniques for multiphysics
-problems and a brief comparison to the proposed methods is provided
-to further motivate this work.
+We discuss in this chapter physics-based motivation for advancing
+Monte Carlo solvers and studying their application to transport
+systems by providing problems of interest in nuclear reactor
+analysis. Hardware-based motivations are also provided by considering
+the impact of forthcoming computing architectures. In addition,
+background on the current solver techniques for multiphysics problems
+and a brief comparison to the proposed methods is provided to further
+motivate this work. The organization of the document is then reviewed
+followed by an explanation of the key outcomes of this work and where
+in the document results and data supporting those outcomes can be
+located.
 
+%%---------------------------------------------------------------------------%%
 \section{Physics-Based Motivation}
@@
 assembly.
-\begin{figure}[htpb!]
+\begin{figure}[t!]
   \begin{center}
@@
 
+%%---------------------------------------------------------------------------%%
 \section{Hardware-Based Motivation}
@@
 
+%%---------------------------------------------------------------------------%%
 \section{Research Outline}
@@
 MCSA as a general linear solver and the development of a parallel MCSA
-method will be new and unique features of this work, providing a
+method are new and unique features of this work, providing a
 framework with which other issues such as resiliency may be addressed
@@
 solver may become prohibitively large
-\citep{knoll_newton-krylov_1995}. To address these issues, this thesis
+\citep{knoll_newton-krylov_1995}. To address these issues, this work
 develops a new and novel method for nonlinear systems based on the
@@
 
-This preliminary report outlines in Chapter~\ref{ch:linear_problem}
-the conventional methods used in practice for solving linear problems
-to provide a mathematical basis upon which to build new algorithms
-that aim to solve some of the aforementioned issues. Parallel schemes
-for conventional methods are also provided for background and
-understanding of how current methods may or may not map to future
-hardware. In addition, some components of their parallel
-implementations may be applied to stochastic methods development. In
-Chapter~\ref{ch:stochastic_methods}, Monte Carlo algorithms for
-solving linear systems and new parallel strategies will be outlined in
-full with links made to past work and their potential for offering
-improvements to the multiphysics analysis community. From these
-stochastic methods, a new nonlinear method is developed in
-Chapter~\ref{ch:nonlinear_problem} and compared to conventional
-methods for solving nonlinear problems.
+To present these new developments this document is arranged in the
+following manner. First, in Chapter~\ref{ch:stochastic_methods}, the
+fundamentals of the Monte Carlo method for linear systems are
+presented. Using this background on Monte Carlo, synthetic
+acceleration methods are then presented and analyzed using a simple
+model transport problem. Next, in Chapter~\ref{ch:spn_equations}, the
+Monte Carlo methods developed in Chapter~\ref{ch:stochastic_methods}
+are applied to the neutron transport problem. Specifically, the $SP_N$
+form of the Boltzmann neutron transport equation is developed and
+solved using the Monte Carlo methods. Using a difficult fission
+reactor fuel assembly criticality calculation to drive research and
+development for these Monte Carlo methods, several important issues
+regarding Monte Carlo solver applicability and performance will be
+discussed along with solutions to these problems. Using these
+solutions, the Monte Carlo methods are then verified against modern
+solution techniques for the $SP_N$ equations in order verify
+correctness for the fuel assembly criticality calculation. In
+Chapter~\ref{ch:nonlinear_problem}, the Monte Carlo methods are
+applied to the Navier-Stokes equations as a model nonlinear energy and
+momentum transport system. Additional difficulties that arise when
+using Monte Carlo methods in these types of systems will be presented
+along with the subsequently developed solutions. In
+Chapter~\ref{ch:parallel_methods}, the Monte Carlo methods are
+parallelized using modern reactor physics techniques for particle
+transport with a full parallel scaling analysis provided using a
+leadership-class computing facility. Finally, the work is summarized
+in Chapter~\ref{ch:conclusion} and topics of future work derived from
+the results of this research presented.

### 2013-06-03 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
updates to introduction chapter

@@
 %%---------------------------------------------------------------------------%%
-\section{MCSA Comparison to Conventional Methods}
-\label{sec:spn_comparison}
+\section{MCSA Verification}
+\label{sec:spn_mcsa_verification}
 Through the numerical studies presented up to this point in this
@@
 
-\subsection{MCSA Verification}
-\label{subsec:spn_mcsa_verification}
 In order to verify the correctness of the MCSA method and
@@
 
-\subsection{Performance Results}
-\label{subsec:spn_comparison_results}
+%%---------------------------------------------------------------------------%%
+\section{MCSA Performance Comparison to Conventional Methods}
+\label{sec:spn_comparison}
 Using the same set of solver parameters used for verification, MCSA
@@
 without a loss in iterative performance. Adding this absorption did
-not change the numerical results of the
-calculation.
+not change the numerical results of the calculation.
 

### 2013-06-03 doc/dissertation/document/chapters/mc_background/mc_background.tex
updating parallel chapter sections

@@
 \section{Monte Carlo Method Selection}
-\label{sec:mc_method_selection}p
+\label{sec:mc_method_selection}
 The MCSA method defined in Eq.~(\ref{eq:mcsa}) uses the adjoint method
@@
 context of MCSA, we choose the two-dimensional time-dependent Poisson
-equation as a simple model problem:
+equation as a simple model transport problem\footnote{The Poisson
+  equation is simple form of the diffusion equation and has a similar
+  elliptic character in its time-dependent form to the neutron
+  transport equations that will be solved in the next chapter.}:
 \begin{equation}
@@
 specified tolerance as a function of the problem size. All
-computations presented in this section and
-\S\ref{subsec:sequential_comparison} were completed on a 3.0 GHz Intel
-Core 2 Quad Q9650 CPU machine with 16 GB 1067 MHz DDR3 memory.
+computations presented in this section and the remaining sections of
+this chapter were completed on a 3.0 GHz Intel Core 2 Quad Q9650 CPU
+machine with 16 GB 1067 MHz DDR3 memory.
 \begin{figure}[t!]

### 2013-06-03 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
starting work on breakdown analysis and adding sections to the concluding chapter
adding to spn estimator analysis section
updating spn chapter
updating parallel chapter sections

@@
 recognize that the algorithm occurs in two stages, an outer iteration
-performing Richardson's iteration and applying the correction, and an
+performing fixed point iteration and applying the correction, and an
 inner Monte Carlo solver that is generating the correction via the
-adjoint method. The parallel aspects of both these components must be
-considered.
+adjoint or forward methods. The parallel aspects of both these
+components must be considered.
 
-In this chapter we briefly review particle transport
-methods for domain decomposed Monte Carlo and then provide an analytic
-framework from which to estimate its performance. We then devise an
-algorithm based on these methods for efficient parallel solutions
-using the adjoint Neumann-Ulam method and MCSA and perform parallel
-scaling studings to test its performance and compare this performance
-to conventional methods.
+In this chapter we briefly review particle transport methods for
+domain decomposed Monte Carlo and then provide an analytic framework
+from which to estimate its performance. We then devise an algorithm
+based on these methods for efficient parallel solutions using the
+adjoint Neumann-Ulam method and MCSA and perform parallel scaling
+studings to test its performance and compare this performance to
+conventional methods.
 
@@
 limited to full domain replication with parallelism exploited through
-individual histories \citep{alexandrov_efficient_1998} and in the
-previous chapter we exploited Wagner's work to alleviate this. To
-accomplish this, we recognized from the literature that stochastic
-histories must be transported from domain to domain as the simulation
-progresses and they transition to states that are not in the local
-domain. Because we have chosen a domain decomposition strategy in a
-parallel environment, this means that communication of these histories
-must occur between compute nodes owning neighboring pieces of the
-global domain as outlined in our algorithms. We wish to characterize
-this communication not only because communication is in general
-expensive, but also because these nearest-neighbor communication
-sequences, specifically, have poor algorithmic strong scaling
-\citep{gropp_high-performance_2001} in much the same way as a
+individual histories \citep{alexandrov_efficient_1998} and in this
+chapter we will exploit Wagner's work to alleviate this. To accomplish
+this, we recognize from the literature that stochastic histories must
+be transported from domain to domain as the simulation progresses and
+they transition to states that are not in the local domain. Because we
+have chosen a domain decomposition strategy in a parallel environment,
+this means that communication of these histories must occur between
+compute nodes owning neighboring pieces of the global domain. We wish
+to characterize this communication not only because communication is
+in general expensive, but also because these nearest-neighbor
+communication sequences, specifically, have poor algorithmic strong
+scaling \citep{gropp_high-performance_2001} in much the same way as a
 matrix-vector multiply operation. Therefore, we desire a framework to
@@
 
-\subsection{Reduction in Domain Leakage from Overlap}
-\label{sec:overlap_reduction}
-
 \subsection{Communication Costs for Symmetric Systems}
@@
 
-\subsection{Load Balancing Concerns}
-\label{sec:mc_load_balancing}
-Although domain decomposition was shown to be efficient in a perfectly
-load balanced situation in Siegel's work \citep{siegel_analysis_2012},
-careful consideration must be made for situations where this is not
-the case. Given the stochastic nature of the problem and lack of a
-globally homogeneous domain, parallel Monte Carlo simulations are
-inherently load imbalanced. Procassini and others worked to alleviate
-some of the load imbalances that are generated by both particle and
-spatial parallelism and are therefore applicable to the MSOD algorithm
-\citep{procassini_dynamic_2005}. They chose a dynamic balancing scheme
-in which the number of times a particular domain was replicated was
-dependent on the amount of work in that domain (i.e. domains with a
-high particle flux and therefore more particle histories to compute
-require more work). In this variation, domains that require more work
-will be replicated more frequently at reduced particle counts in each
-replication. Furthermore, Procassini and colleagues noted that as the
-simulation progressed and particles were transported throughout the
-domain, the amount of replication for each domain would vary as
-particle histories began to diffuse, causing some regions to have
-higher work loads and some to have smaller work loads than the initial
-conditions.
-
-Consider the example in Figure~\ref{fig:procassini_example} adapted
-from Procassini's work.
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.5}{
-      \input{chapters/parallel_mc/procassini_example.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Example illustrating how domain decomposition can
-      create load balance issues in Monte Carlo.}  \textit{A domain is
-      decomposed into 4 zones on 8 processors with a point source in
-      the lower left zone. As the particles diffuse from the source in
-      the random walk sequence as shown in the top row, their tracks
-      populate the entire domain. As given in the bottom row, as the
-      global percentage of particles increases in a zone, that zone's
-      replication count is increased.}}
-  \label{fig:procassini_example}
-\end{figure}
-In this example, a geometry is decomposed into 4 domains on 8
-processors with a point source in the bottom left domain. To begin,
-because the point source is concentrated in one domain, that domain is
-replicated 5 times such the amount of work it has to do per processor
-is roughly balanced with the others. As the particles begin to diffuse
-away from the point source, the amount of replication is adjusted to
-maintain load balance. Near the end of the simulation, the diffusion
-of particles is enough that all domains have equal replication.  By
-doing this, load balance is improved as each domain has approximately
-equal work although each domain may represent a different spatial
-location and have a differing number of histories to
-compute. Compared to Wagner's work where the fission source was
-distributed relatively evenly throughout the domain, fixed source
-problems (and especially those that have a point-like source) like
-those presented in Procassini's work will be more prone to changing
-load balance requirements.
-
-\subsection{Reproducible Domain Decomposed Results}
-\label{sec:reproducible_mc}
-The 2006 work of Brunner is notable in that the Monte Carlo codes used
-to implement and test the algorithms adhered to a strict policy of
-generating identical results independent of domain decomposition or
-domain replication as derived from the work of Gentile and colleagues
-\citep{gentile_obtaining_2005}. In Gentile's work, a procedure is
-given for obtaining results reproducible to machine precision for an
-arbitrary number of processors and domains. Differences can arise from
-using a different random number sequence in each domain and performing
-a sequence of floating point operations on identical data in a
-different order, leading to variations in round-off error and
-ultimately a non-identical answer. They use a simple example,
-recreated below in Figure~\ref{fig:gentile_example}, that illustrates
-these issues.
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.5}{
-      \input{chapters/parallel_mc/gentile_example.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Gentile's example illustrating how domain
-      decomposition can create reproducibility issues in Monte Carlo.}
-    \textit{Both particles A and B start in zone 1 on processor
-      1. Particle A moves to zone 2 on processor 2 and scatters back
-      to zone 1 while B scatters in zone 1 and remains there. A1 and
-      A2 denote the track of particle A that is in zone 1 while B1 and
-      B2 denote the track of particle B that is in zone 1.}}
-  \label{fig:gentile_example}
-\end{figure}
-In this example, the domain is decomposed on two processors with each
-processor owning one of the two zones. Starting with particle A, it is
-born in zone 1 and is transported to zone 2 where a scattering event
-occurs. Concerning the first reproducibility issue, if the same
-sequence of random numbers is not used to compute the trajectory from
-the new scattering event, we cannot expect to achieve the same result
-if the domain were not decomposed. If the numbers are different, the
-scattering event in zone 2 may keep the particle there or even eject
-it from the domain if the sequence were different. The second issue is
-demonstrated by adding another particle B that remains in the
-domain. In this case, an efficient algorithm will transport particle A
-on processor 1 until it leaves zone 1 and then transport particle
-B. Particle A will not renter the domain until it has been
-communicated to processor 2, processor 2 performs the transport, and
-it is communicated back to processor 1. If we are doing a track-length
-tally in zone 1, then we sum the tracks lengths observed in that
-zone. In the single processor, single zone case particle A would be
-transported in its entirety and then particle B transported. This
-would result in a tally sum with the following order of operations:
-$(((A1+A2)+B1)+B2)$. If we were instead to use 2 processors, we would
-instead have the following order: $(((A1+B1)+B2)+A2)$. In the context
-of floating point operations, we cannot expect these to have an
-identical result to machine precision as round-off approximations will
-differ resulting in non-commutative addition.
-
-Procassini's solutions to these problems are elegant in that they
-require a minimal amount of modification to be applied to the Monte
-Carlo algorithm. To solve the first issue, in order to ensure each
-particle maintains an invariant random number sequence that determines
-its behavior regardless of domain decomposition, each particle is
-assigned a random number seed that describes its current state upon
-entering the domain of a new processor. These seeds are derived from
-the actual geometric location of the particle such that it is
-decomposition invariant. Non-commutative floating point operations are
-overcome by instead mapping floating point values to 64-bit integer
-values for which additions will always be commutative. Once the
-operations are complete, these integers are mapped back to floating
-point values.
-
 %%---------------------------------------------------------------------------%%
@@
 %%---------------------------------------------------------------------------%%
-\section{Parallel Scaling Studies}
-\label{sec:scaling_studies}
+\section{Small Cluster Scaling Studies}
+\label{sec:small_scaling_studies}
+
+%%---------------------------------------------------------------------------%%
+\section{Leadership-Class Parallel Scaling Studies}
+\label{sec:leadership_scaling_studies}
 For all cases, we explore scaling for just the Monte Carlo transport
@@
 preconditioning.
+
+\subsection{Titan}
+\label{subsec:titan}
+
+\subsection{Communication Parameters}
+\label{subsec:comm_parameters}
+
+\subsection{Pure Domain Decomposition}
+\label{subsec:pure_domain_decomp}
+
+\subsubsection{Weak Scaling}
+\label{subsubsec:pure_weak}
+
+\subsubsection{Strong Scaling}
+\label{subsubsec:pure_strong}
 
@@
 
-\subsection{Effects of Load Imbalance}
-\label{subsec:msod_load_imbalance}
-
-\subsubsection{Weak Scaling}
-\label{subsubsec:load_imbalance_weak}
-
-\subsubsection{Strong Scaling}
-\label{subsubsec:load_imbalance_strong}
+%%---------------------------------------------------------------------------%%
+\section{Multiple Fuel Assembly $SP_N$ Challenge Problem}
+\label{sec:spn_challenge_problem}

### 2013-06-03 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
updates to introduction chapter
updating nonlinear chapter outline ;

@@
-\chapter{Monte Carlo Solution Methods Applied to the Navier-Stokes Equations}
+\chapter{Monte Carlo Solution Methods for the Navier-Stokes Equations}
 \label{ch:nonlinear_problem}
@@
 
-\section{Inexact Newton Methods}
-\label{sec:newton_methods}
+\subsection{Inexact Newton Methods}
+\label{subsec:newton_methods}
 Inexact Newton methods arise when the Jacobian operator is not exactly
@@
 
+%%---------------------------------------------------------------------------%%
 \section{The FANM Method}
@@
 %%---------------------------------------------------------------------------%%
-\section{Monte Carlo Solution Methods for the Navier-Stokes Equations}
-\label{sec:mc_ns_solutions}
-
-\subsection{FANM Verification}
-\label{subsec:fanm_verification}
+\section{Navier-Stokes Benchmark Problems}
+\label{sec:ns_benchmarks}
 To verify the FANM method for nonlinear problems, we choose benchmark
@@
 
-\subsubsection{Thermal Convection Cavity Problem}
-\label{subsubsec:natural_convection_cavity}
+\subsection{Thermal Convection Cavity Problem}
+\label{subsec:natural_convection_cavity}
 In 1983 a benchmark solution for the natural convection of air in a
@@
 
-\subsubsection{Lid Driven Cavity Problem}
-\label{subsubsec:lid_driven_cavity}
+\subsection{Lid Driven Cavity Problem}
+\label{subsec:lid_driven_cavity}
 As an extension of the convection problem, the second benchmark
@@
 
-\subsubsection{Backward-Facing Step Problem}
-\label{subsubsec:backward_facing_step}
+\subsection{Backward-Facing Step Problem}
+\label{subsec:backward_facing_step}
 The third benchmark was generated by Gartling in 1990 and consists of
@@
 \sn{5}{2}.
+
+%%---------------------------------------------------------------------------%%
+\section{Analysis of the Navier-Stokes Equations}
+\label{sec:ns_analysis}
+
+%%---------------------------------------------------------------------------%%
+\section{FANM Verification}
+\label{sec:fanm_verification}
+
+%%---------------------------------------------------------------------------%%
+\section{FANM Performance Comparison to Conventional Methods}
+\label{sec:fanm_comparison}

### 2013-06-06 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
updates to spn intro section

@@
 methods for large scale problems are commonly variants on the discrete
-ordinates ($S_N$) method \citep{evans_denovo:_2010}. For full reactor
-core neutronics simulations, the $S_N$ method requires potentially
+ordinates ($S_N$) method \citep{evans_denovo:_2010}. For fission
+reactor neutronics simulations, the $S_N$ method requires potentially
 trillions of unknown angular flux moments to be computed to achieve
@@
 solid numerical footing for the method, we look at its application to
-today's challenge problems in full reactor core transport. The
-reduction in numerical complexity of current full core deterministic
-solution methods using the $S_N$ approximation could mean significant
-savings in both compute time and memory required. In addition, the
-characteristics of the solution to the transport problem for a steady
-state reactor core permit diffusion theory to be used; a staple of the
-nuclear industry since its inception. Therefore, if diffusion theory
-is applicable, then finer grained solutions that capture more of the
-physics contained in the transport equation should be possible with
-the $SP_N$ method. In doing so, we also expect from the literature to
-obtain computed responses on the order of accuracy we would expect
-from an appropriately discretized $S_N$ method at a fraction of the
-cost.
+today's challenge problems in neutron transport for fission reactor
+analysis. The reduction in numerical complexity of current
+deterministic solution methods using the $S_N$ approximation could
+mean significant savings in both compute time and memory required. In
+addition, the characteristics of the solution to the transport problem
+for a steady state reactor core permit diffusion theory to be used; a
+staple of the nuclear industry since its inception. Therefore, if
+diffusion theory is applicable, then finer grained solutions that
+capture more of the physics contained in the transport equation should
+be possible with the $SP_N$ method. In doing so, we also expect from
+the literature to obtain computed responses on the order of accuracy
+we would expect from an appropriately discretized $S_N$ method at a
+fraction of the cost.
 
@@
 permitted generation of the $SP_N$ system of equations for detailed
-full core reactor models \citep{evans_simpli_2013}. By fully forming
-these equations and formulating them as a linear algebra problem
-instead of using the explicit iterative methods of the past, we now
-have access to all of the modern advancements in computational linear
-algebra including Krylov solvers for asymmetric systems and
-preconditioning methods such as algebraic multigrid. This leads us to
-then explore the applicability of our work in discrete Monte Carlo
-methods for linear systems as a possible solution method for the
-$SP_N$ equations. If formulated correctly, we hypothesize that
-significant improvement in usage of computational resources may be
-observed compared to modern solution techniques such as those
-suggested due to the form of the matrices generated by the
-discretization. In addition, solving the $SP_N$ equations in this way
-also breaks away from the $S_N$ forms of parallelism where spatial
-parallelism is achieved by an efficient parallel sweep, angular
-efficiency achieved by pipe-lining, and energy parallelism achieved by
-decoupling the groups. With the $SP_N$ equations as a full matrix
-system, we now can parallelize the problem as prescribed by the linear
-solver, which may be significantly more scalable than current $S_N$
-transport practices.
+neutronics models \citep{evans_simpli_2013}. By fully forming these
+equations and formulating them as a linear algebra problem instead of
+using the explicit iterative methods of the past, we now have access
+to all of the modern advancements in computational linear algebra
+including Krylov solvers for asymmetric systems and preconditioning
+methods such as algebraic multigrid. This leads us to then explore the
+applicability of our work in discrete Monte Carlo methods for linear
+systems as a possible solution method for the $SP_N$ equations. If
+formulated correctly, we hypothesize that significant improvement in
+usage of computational resources may be observed compared to modern
+solution techniques such as those suggested due to the form of the
+matrices generated by the discretization. In addition, solving the
+$SP_N$ equations in this way also breaks away from the $S_N$ forms of
+parallelism where spatial parallelism is achieved by an efficient
+parallel sweep, angular efficiency achieved by pipe-lining, and energy
+parallelism achieved by decoupling the groups. With the $SP_N$
+equations as a full matrix system, we now can parallelize the problem
+as prescribed by the linear solver, which may be significantly more
+scalable than current $S_N$ transport practices.
 
@@
 context. From the $P_N$ equations, we apply a set of approximations to
-yield the multi-dimensional, multi-group $SP_N$ equations for fixed
-source problems. Using the fully-formed matrix for the transport
+yield the $SP_N$ equations for fixed source and criticality
+problems. Using the fully-formed linear operator for the transport
 problem, we explore solutions to the $SP_N$ equations using Monte
 Carlo Synthetic Acceleration using a light water reactor fuel assembly
-criticality calculation as the driving problem.
+criticality calculation as the driving problem. In order to
+effectively solve this problem, a suite of preconditioners is
+developed and studied within the context of MCSA. Using these
+preconditioners, MCSA solutions for the fuel assembly problem are
+verified by comparing against production Krylov solvers and finally
+performance is analyzed through comparison with those same Krylov
+solvers.
 
@@
 \label{sec:transport_eq}
-As a starting point we define the time-independent neutron transport
-equation \citep{lewis_computational_1993}:
+As a starting point for the $SP_N$ equations we define the
+time-independent neutron transport equation
+\citep{lewis_computational_1993}:
 \begin{multline}

### 2013-06-08 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding first piece of verification data

@@
 
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/spn_equations/gmres_mcsa_1_1_diff.pdf}
+  \end{center}
+  \caption{\textbf{Absolute difference of the group 0 scalar flux
+      between the MCSA and Aztec GMRES calculations for the 1-group
+      calculation.} \textit{Over the entire domain a maximum absolute
+      difference of \sn{9}{-10} between the scalar flux computed by
+      MCSA and GMRES was observed. The solvers were converged to a
+      tolerance of \sn{1}{-8}.}}
+  \label{fig:gmres_mcsa_1_1_diff}
+\end{figure}
+
 %%---------------------------------------------------------------------------%%

### 2013-06-10 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding serial verification data for the eigenvalue

@@
 \label{sec:spn_mcsa_verification}
-Through the numerical studies presented up to this point in this
-chapter, we have developed a set of preconditioning techniques that
-permit MCSA to be used with the $SP_N$ form of the neutron transport
-equation and applied to a difficult fuel assembly criticality
-problem. Along with these preconditioning techniques, a set a varying
-MCSA parameters were analyzed to find those that yielded the best
-performance for this particular problem. In addition, several steps
-were taken to mitigate the dense composite linear operators that arise
-when performing explicit MCSA preconditioning in order to achieve
-convergence. Using these results, we compare MCSA directly to
-conventional linear solvers that would be typically used to solve the
-$SP_N$ form of the transport equation as presented here. In
-particular, we will use a pair of production Krylov solvers with which
-we will compare direct numerical results in order to verify MCSA
-solutions and iterative and CPU timing performance in order to gauge
+Through the numerical studies presented in this chapter, we have
+developed a set of preconditioning techniques that permit MCSA to be
+used with the $SP_N$ form of the neutron transport equation and
+applied to a difficult fuel assembly criticality problem. Along with
+these preconditioning techniques, a set a varying MCSA parameters were
+analyzed to find those that yielded the best performance for this
+particular problem. In addition, several steps were taken to mitigate
+the dense composite linear operators that arise when performing
+explicit MCSA preconditioning in order to achieve convergence. Using
+these results, we compare MCSA directly to conventional linear solvers
+that would be typically used to solve the $SP_N$ form of the transport
+equation as presented here. In particular, we will use a pair of
+production Krylov solvers with which we will compare direct numerical
+results in order to verify MCSA solutions in this section and
+iterative and CPU timing performance in the next order to gauge
 feasibility and identify areas that require further research.
@@
 required to converge the problem along with the eigenvalue itself will
-be compared as a means of verification.
-
-For the verification, all solvers will be preconditioned with ILUT
-using a drop tolerance of \sn{1}{-5} and a fill level of 5. For GMRES,
-no restrictions were placed on the size of the subspace and therefore
-no restarts occurred. When the collision estimator was used with MCSA,
-20,000 stochastic histories were used to compute the correction for
-every energy group in the problem (40,000 and 80,000 histories total
-for the 2 and 4 group calculations respectively) and 500 stochastic
-histories used for each energy group when the expected value estimator
-was used (1,000 and 1,500 histories total for the 2 and 4 group
-calculations). The reduced domain approximation was also applied in
-conjuction with the ILUT preconditioning using a fill level of 100 and
-a threshold value of \sn{1}{-10} to reduce the density of states in
-the Monte Carlo problem. For relaxation parameters, all MCSA
-compuations used a Neumann relaxation parameter of 0.7 while a
-Richardson relaxation of 1.1 was used with the collision estimator and
-1.0 with the expected value estimator as determined by the previous
-analysis. Table~\ref{tab:spn_solver_defs} gives definitions for the
-solvers used to generate the results in the remainder of this
-section. For the energy group structures,
+be compared as a means of verification. If all values computed using
+MCSA agree with the results from the production Krylov solvers within
+the solution tolerance then the MCSA solutions will be deemed verified
+for these problems. Additionaly, this verification is only applicable
+to serial results with parallel verification for this problem provided
+in Chapter~\ref{ch:parallel_methods}.
+
+For the verification calculations, all solvers will be preconditioned
+with ILUT using a drop tolerance of \sn{1}{-5} and a fill level of
+5. For GMRES, no restrictions were placed on the size of the subspace
+and therefore no restarts occurred. When the collision estimator was
+used with MCSA, \sn{2}{4} stochastic histories were used to compute
+the correction for every energy group in the problem (\sn{4}{4} and
+\sn{8}{4} histories total for the 2 and 4 group calculations
+respectively) corresponding to approximately 1 stochastic history per
+DOF. When the expected value estimator was used, \sn{5}{2} stochastic
+histories used for each energy group (\sn{1}{3} and \sn{1.5}{3}
+histories total for the 2 and 4 group calculations) giving
+approximately 1 stochastic history for every 5 DOFs. The reduced
+domain approximation was also applied in conjuction with the ILUT
+preconditioning using a fill level of 100 and a threshold value of
+\sn{1}{-10} to reduce the density of states in the Monte Carlo
+problem. For relaxation parameters, all MCSA compuations used a
+Neumann relaxation parameter of 0.7 while a Richardson relaxation of
+1.1 was used with the collision estimator and 1.0 with the expected
+value estimator as determined by the previous analysis of the
+relaxation parameters. Table~\ref{tab:spn_solver_defs} gives
+definitions for the solvers used to generate the results in the
+remainder of this section. For the energy group structures,
 Table~\ref{tab:spn_group_structure} gives the lower bounds of each
-group in eV (an implicit upper bound of 20 MeV is assumed).
+group in eV (an implicit upper bound of \sn{2}{6} eV is assumed for
+group 0).
 \begin{table}[h!]
@@
   \caption{\textbf{Solver definitions used for MCSA verification and
-      performance analysis.}}
+      performance analysis.} \textit{The ILUT preconditioner
+      parameters were identical for all calculations and solvers.}}
   \label{tab:spn_solver_defs}
@@
   \caption{\textbf{Energy group lower bounds for the multigroup fuel
-      assembly criticality problem in electron volts.}}
+      assembly criticality problem in electron volts.} \textit{An
+      implicit upper bound of \sn{2}{6} eV is assumedfor group zero.}}
   \label{tab:spn_group_structure}
 \end{table}
+
+For every combination of energy group structure and solver, the
+eigenvalue problem was converged with an eigensolver tolerance of
+\sn{1}{-6} and a linear solver tolerance of \sn{1}{-8} in a serial
+calculation. Table~\ref{tab:serial_ev_results} gives the eigenvalue
+computed by each solver for each problem as well as the number of
+eigenvalue iterations required to converge. For every problem, each
+solver converged to the same eigenvalue within the tolerance of the
+eigensolver in the same number of eigenvalue iterations.
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lccc}\hline\hline
+      \multicolumn{1}{c}{\textbf{Solver}} & 
+      \multicolumn{1}{c}{\textbf{Number of Groups}} & 
+      \multicolumn{1}{c}{\textbf{Iterations}} &
+      \multicolumn{1}{c}{\textbf{Eigenvalue}} \\
+      \hline
+      BiCGStab-ILUT & 1 & 25 & 1.1059870 \\
+      GMRES-ILUT & 1 & 25 & 1.1059870 \\
+      MCSA-ILUT-R-C & 1 & 25 & 1.1059870 \\
+      MCSA-ILUT-MR-C & 1 & 25 & 1.1059870 \\
+      MCSA-ILUT-R-EV & 1 & 25 & 1.1059870 \\
+      MCSA-ILUT-MR-EV & 1 & 25 & 1.1059870 \\
+      \hline
+      BiCGStab-ILUT & 2 & 35 & 1.1552390 \\
+      GMRES-ILUT & 2 & 35 & 1.1552390 \\
+      MCSA-ILUT-R-C & 2 & 35 & 1.1552390 \\
+      MCSA-ILUT-MR-C & 2 & 35 & 1.1552390 \\
+      MCSA-ILUT-R-EV & 2 & 35 & 1.1552390 \\
+      MCSA-ILUT-MR-EV & 2 & 35 & 1.1552390 \\
+      \hline
+      BiCGStab-ILUT & 4 & 31 & 1.1758500 \\
+      GMRES-ILUT & 4 & 31 & 1.1758500 \\
+      MCSA-ILUT-R-C & 4 & 31 & 1.1758500 \\
+      MCSA-ILUT-MR-C & 4 & 31 & 1.1758500 \\
+      MCSA-ILUT-R-EV & 4 & 31 & 1.1758500 \\
+      MCSA-ILUT-MR-EV & 4 & 31 & 1.1758500 \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Serial eigensolver verification results for the
+      fuel assembly problem.} \textit{Each calculation was converged
+      with the same eigensolver and linear solver tolerances.}}
+  \label{tab:serial_ev_results}
+\end{table}
+
+For each calculation, the scalar flux in each energy group was
+compared over the entire spatial with the GMRES solution used as the
+reference.
 

### 2013-06-10 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
adding serial verification data for the eigenvalue
restructuring verification results

@@
-\chapter{Parallel Monte Carlo Solution Methods for Linear Systems}
+\chapter{Parallel Monte Carlo Synthetic Acceleration}
 \label{ch:parallel_methods}
@@
 based on these methods for efficient parallel solutions using the
-adjoint Neumann-Ulam method and MCSA and perform parallel scaling
-studings to test its performance and compare this performance to
-conventional methods.
+Neumann-Ulam method and MCSA and verify its correctness by comparing
+the numerical results against production Krylov methods. We then
+perform parallel scaling studings to test its performance on both a
+small computing cluster and a leadership class machine and compare
+this performance to the production Krylov methods.
 
@@
 \label{sec:msod}
-As observed in the discussion on parallel Krylov methods, large-scale
-problems will surely have their data partitioned such that each
-parallel process owns a subset of the equations in the linear
-system. Given this convention, the adjoint Monte Carlo algorithm must
-perform random walks over a domain that is decomposed and must remain
-decomposed due to memory limitations. This naturally leads us to seek
-parallel Monte Carlo algorithms that handle domain decomposition.
+As observed in the discussion on parallel Krylov methods in
+Appendix~\ref{ch:linear_problem}, large-scale problems will surely
+have their data partitioned such that each parallel process owns a
+subset of the equations in the linear system. Given this convention,
+the adjoint Monte Carlo algorithm must perform random walks over a
+domain that is decomposed and must remain decomposed due to memory
+limitations. This naturally leads us to seek parallel Monte Carlo
+algorithms that handle domain decomposition.
 
@@
 %%---------------------------------------------------------------------------%%
+\section{Parallel MCSA Verification}
+\label{sec:parallel_verification}
+
+%%---------------------------------------------------------------------------%%
 \section{Small Cluster Scaling Studies}

### 2013-06-10 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
restructuring verification results

@@
-\chapter{Monte Carlo Solution Methods for the Simplified $P_N$ Equations}
+\chapter{Monte Carlo Synthetic Acceleration Methods for the $SP_N$ Equations}
 \label{ch:spn_equations}
@@
 
-For each calculation, the scalar flux in each energy group was
-compared over the entire spatial with the GMRES solution used as the
-reference.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/gmres_mcsa_1_1_diff.pdf}
+For each converged eigenvalue calculation, the scalar flux in each
+energy group was compared over the entire spatial with the
+BiCGStab-ILUT solution used as the reference to compute an absolute
+difference. For the 1 group calculations, these absolute differences
+are plotted spatially along the centerline of the fuel assembly in
+Figure~\ref{fig:serial_1_group_diff}. Interestingly, except for the
+MCSA-ILUT-MR-C calculation, all other 1-group calculations including
+GMRES had a maximum global flux difference of \sn{9.313}{-10} even
+though different spatial results were observed. For the MCSA-ILUT-MR-C
+calculation, a slightly larger maximum error of \sn{2.205}{-9} was
+observed for reasons that remain unexplained. All 1-group calculations
+have a minimum observed absolute difference of 0 with respect to the
+BiCGStab calculation.
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/spn_equations/serial_1_group.pdf}
   \end{center}
   \caption{\textbf{Absolute difference of the group 0 scalar flux
-      between the MCSA and Aztec GMRES calculations for the 1-group
-      calculation.} \textit{Over the entire domain a maximum absolute
-      difference of \sn{9}{-10} between the scalar flux computed by
-      MCSA and GMRES was observed. The solvers were converged to a
-      tolerance of \sn{1}{-8}.}}
-  \label{fig:gmres_mcsa_1_1_diff}
-\end{figure}
+      between the Aztec BiCGStab calculations and Aztec GMRES and the
+      MCSA variations for the 1-group calculation.} \textit{Left to
+      right: GMRES-ILUT, MCSA-ILUT-R-C, MCSA-ILUT-MR-C,
+      MCSA-ILUT-R-EV, MCSA-ILUT-MR-EV.}}
+  \label{fig:serial_1_group_diff}
+\end{figure}
+
+Table~\ref{tab:serial_differences} gives the maximum and minimum
+absolute differences for all calculations and all energy groups using
+the BiCGStab calculation as the reference. All comparisons were
+performed within a group such that the MCSA-ILUT-R-C results in group
+3 for the 4-group calculation were computed with the BiCGStab-ILUT
+solution in group 3 for the 4-group calculation as a reference.
 
@@
 will be directly compared to the production Krylov solvers in terms of
-both iterative performance and CPU timing. In addition to the
-parameters used for the verification analysis, artificial absorption
-was also introduced with a value of 0.2 for the MCSA calculations that
-used the collision esitmator as CPU timing performance was improved
-without a loss in iterative performance. Adding this absorption did
-not change the numerical results of the calculation.
+both iterative performance and CPU timing for solutions to the fuel
+assembly problem. In addition to the parameters used for the
+verification analysis, artificial absorption was also introduced with
+a value of 0.2 for the MCSA calculations that used the collision
+esitmator as CPU timing performance was improved without a loss in
+iterative performance. Adding this absorption did not change the
+numerical results of the calculation.
 

### 2013-06-10 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding solver verification data for serial flux calculations

@@
 
-Table~\ref{tab:serial_differences} gives the maximum and minimum
+Table~\ref{tab:serial_differences_g1},
+Table~\ref{tab:serial_differences_g2}, and
+Table~\ref{tab:serial_differences_g4} give the maximum and minimum
 absolute differences for all calculations and all energy groups using
@@
 solution in group 3 for the 4-group calculation as a reference.
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lccc}\hline\hline
+      \multicolumn{1}{c}{\textbf{Solver}} & 
+      \multicolumn{1}{c}{\textbf{Group}} &
+      \multicolumn{1}{c}{\textbf{Minimum Difference}} &
+      \multicolumn{1}{c}{\textbf{Maximum Difference}} \\
+      \hline
+      GMRES-ILUT & 0 & 0.0 & \sn{9.313}{-10} \\
+      MCSA-ILUT-R-C & 0 & 0.0 & \sn{9.313}{-10} \\
+      MCSA-ILUT-MR-C & 0 & 0.0 & \sn{2.205}{-9} \\
+      MCSA-ILUT-R-EV & 0 & 0.0 & \sn{9.313}{-10} \\
+      MCSA-ILUT-MR-EV & 0 & 0.0 & \sn{9.313}{-10} \\
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Serial scalar flux verification results for the
+      1-group fuel assembly problem.} \textit{The maximum and minimum
+      absolute differences over the entire spatial domain are
+      presented with BiCGStab used as the reference. Each calculation
+      was converged with the same eigensolver and linear solver
+      tolerances.}}
+  \label{tab:serial_differences_g1}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lccc}\hline\hline
+      \multicolumn{1}{c}{\textbf{Solver}} & 
+      \multicolumn{1}{c}{\textbf{Group}} &
+      \multicolumn{1}{c}{\textbf{Minimum Difference}} &
+      \multicolumn{1}{c}{\textbf{Maximum Difference}} \\
+      \hline
+      GMRES-ILUT & 0 & 0.0 & \sn{2.794}{-9} \\
+      MCSA-ILUT-R-C & 0 & 0.0 & \sn{9.313}{-10} \\
+      MCSA-ILUT-MR-C & 0 & 0.0 & \sn{9.313}{-10} \\
+      MCSA-ILUT-R-EV & 0 & 0.0 & \sn{9.313}{-10} \\
+      MCSA-ILUT-MR-EV & 0 & 0.0 & \sn{9.313}{-10} \\
+      \hline
+      GMRES-ILUT & 1 & 0.0 & \sn{4.657}{-10} \\
+      MCSA-ILUT-R-C & 1 & 0.0 & \sn{1.364}{-10} \\
+      MCSA-ILUT-MR-C & 1 & 0.0 & \sn{1.261}{-9} \\
+      MCSA-ILUT-R-EV & 1 & 0.0 & \sn{2.474}{-10} \\
+      MCSA-ILUT-MR-EV & 1 & 0.0 & \sn{5.857}{-10} \\
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Serial scalar flux verification results for the
+      2-group fuel assembly problem.} \textit{The maximum and minimum
+      absolute differences over the entire spatial domain are
+      presented with BiCGStab used as the reference. Each calculation
+      was converged with the same eigensolver and linear solver
+      tolerances.}}
+  \label{tab:serial_differences_g2}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lccc}\hline\hline
+      \multicolumn{1}{c}{\textbf{Solver}} & 
+      \multicolumn{1}{c}{\textbf{Group}} &
+      \multicolumn{1}{c}{\textbf{Minimum Difference}} &
+      \multicolumn{1}{c}{\textbf{Maximum Difference}} \\
+      \hline
+      GMRES-ILUT & 0 & 0.0 & \sn{9.313}{-10} \\
+      MCSA-ILUT-R-C & 0 & 0.0 & \sn{1.397}{-9} \\
+      MCSA-ILUT-MR-C & 0 & 0.0 & \sn{1.870}{-9} \\
+      MCSA-ILUT-R-EV & 0 & 0.0 & \sn{1.397}{-9} \\
+      MCSA-ILUT-MR-EV & 0 & 0.0 & \sn{1.869}{-9} \\
+      \hline
+      GMRES-ILUT & 1 & 0.0 & \sn{5.821}{-11} \\
+      MCSA-ILUT-R-C & 1 & 0.0 & \sn{1.164}{-10} \\
+      MCSA-ILUT-MR-C & 1 & 0.0 & \sn{1.164}{-10} \\
+      MCSA-ILUT-R-EV & 1 & 0.0 & \sn{1.164}{-10} \\
+      MCSA-ILUT-MR-EV & 1 & 0.0 & \sn{1.164}{-10} \\
+      \hline
+      GMRES-ILUT & 2 & 0.0 & \sn{1.164}{-10} \\
+      MCSA-ILUT-R-C & 2 & 0.0 & \sn{2.328}{-10} \\
+      MCSA-ILUT-MR-C & 2 & 0.0 & \sn{2.328}{-10} \\
+      MCSA-ILUT-R-EV & 2 & 0.0 & \sn{2.328}{-10} \\
+      MCSA-ILUT-MR-EV & 2 & 0.0 & \sn{2.328}{-10} \\
+      \hline
+      GMRES-ILUT & 3 & 0.0 & \sn{1.164}{-10} \\
+      MCSA-ILUT-R-C & 3 & 0.0 & \sn{1.746}{-10} \\
+      MCSA-ILUT-MR-C & 3 & 0.0 & \sn{2.256}{-9} \\
+      MCSA-ILUT-R-EV & 3 & 0.0 & \sn{2.801}{-10} \\
+      MCSA-ILUT-MR-EV & 3 & 0.0 & \sn{2.256}{-10} \\
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Serial scalar flux verification results for the
+      4-group fuel assembly problem.} \textit{The maximum and minimum
+      absolute differences over the entire spatial domain are
+      presented with BiCGStab used as the reference. Each calculation
+      was converged with the same eigensolver and linear solver
+      tolerances.}}
+  \label{tab:serial_differences_g4}
+\end{table}
 

### 2013-06-10 doc/dissertation/document/chapters/mc_background/mc_background.tex
restructuring verification results
mc background edits and spell check

@@
-\chapter{Monte Carlo Solution Methods for Linear Systems}
+\chapter{Monte Carlo Synthetic Acceleration Methods}
 \label{ch:stochastic_methods}
@@
 manuscript published in 1950 by Forsythe and Leibler
-\citep{forsythe_matrix_1950}. In their outline, Forsythe and Liebler
+\citep{forsythe_matrix_1950}. In their outline, Forsythe and Leibler
 in fact credit the creation of this technique to J. Von Neumann and
@@
 \citep{spanier_monte_1969} present additional detail on this topic
-using a collection of references from the 1950's and early 1960's.
+using a collection of references from the 1950's and early
+1960's.
+
+Following this work, Halton presented a residual Monte Carlo scheme
+that improved dramatically on the performance of these methods
+\citep{halton_sequential_1962}. Recent work in this area has utilized
+these residual Monte Carlo methods in particle transport
+\citep{evans_residual_2003}, giving accelerated convergence over
+traditional methods. As an expansion of Halton's work, Evans and
+others have leveraged these methods in a synthetic acceleration scheme
+that has been shown to be competitive with production Krylov methods
+for radiation transport problems and improve performance significantly
+over Halton's sequential method \citep{evans_monte_2012}.
 
@@
 analysis of several Monte Carlo estimators. The Sequential Monte Carlo
-and Monte Carlo Synthetic Acceleration methods will then be presented
-as a means of leveraging the basic Monte Carlo methods in an iterative
-refinement scheme that accelerates their convergence. Using these
-iterative schemes, a set of modest variance reduction techniques will
-be introduced and their effects on the Monte Carlo methods explored.
+method of Halton and Monte Carlo Synthetic Acceleration methods will
+then be presented as a means of leveraging the basic Monte Carlo
+methods devised by Neumann and Ulam in an iterative refinement scheme
+that accelerates their convergence. Using these iterative schemes, a
+set of modest variance reduction techniques will be introduced and
+their effects along with other parameters of the algorithms on the
+Monte Carlo solutions explored using a simple model transport problem.
 
@@
 \label{sec:direct_mc}
-In the context of matrix inversion, a direct (forward) method resembles an
-adjoint Monte Carlo method in the reactor physics community where the
-solution state is sampled and the source terms that contribute to it
-are assembled. To achieve this, we build the direct method
-Neumann-Ulam decomposition per Dimov's approach by first choosing a
-probability matrix that is a row scaling of $\ve{H}$ such that its
-components are:
+In the context of matrix inversion, a direct (forward) method
+resembles an adjoint Monte Carlo method in the reactor physics
+community where the solution state is sampled and the source terms
+that contribute to it are assembled. To achieve this, we build the
+forward Neumann-Ulam decomposition per Dimov's approach by first
+choosing a probability matrix that is a row scaling of $\ve{H}$ such
+that its components are:
 \begin{equation}
@@
 Using these matrices, we can then form the expectation value of the
-direct solution. For a given random walk permutation $\nu$, we define
+forward solution. For a given random walk permutation $\nu$, we define
 the weight of that permutation on the $m^{th}$ step to be:
@@
 statistical variance of the solutions is decreased as more tallies are
-made. Starting with single history at each state in the
-domain, the high variance prevents a precise solution from being
-obtained although we begin to see the solution take shape as
-expected. At 1000 histories per state, enough tallies have been made
-to generate a reasonable estimate for the structure of the
-solution. It is interesting to note here that as the statistical
-uncertainty is reduced at each grid point in the domain, the solution
-is resolved in a certain sense, analagous to the convergence of a
-traditional iterative method.
+made. Starting with a single history at each state in the domain, the
+high variance prevents a precise solution from being obtained although
+we begin to see the solution take shape as expected. At 1000 histories
+per state, enough tallies have been made to generate a reasonable
+estimate for the structure of the solution. It is interesting to note
+here that as the statistical uncertainty is reduced at each grid point
+in the domain, the solution is resolved in a certain sense, analogous
+to the convergence of a traditional iterative method.
 \clearpage
@@
 Expanding the transition probabilities in the first sum again yields a
-Neumman series in the same form as that explored for the forward
+Neumann series in the same form as that explored for the forward
 estimator, this time with the terms in reverse order and the
@@
 solution evolves with the addition of histories, more histories
-emanate from the boundary and smaller uniform source with more
-penetrating from the boundary into the domain and making contributions
-to the tallies in those states.
+emanate from the boundary the smaller uniform source in the domain
+with more penetrating from the boundary into the domain and making
+contributions to the tallies in those states as they are transported.
 
@@
 of his work by the transport community have confirmed convergence
-rates on the order of of $e^{-N}$ \citep{evans_residual_2003}. In
-much the same way as projection methods, Halton's method, sequential
-Monte Carlo, utilizes the adjoint Monte Carlo solver as a means of
-directly reducing the elements residual vector. He proposed the
-following iterative scheme as a solution to
-Eq~(\ref{eq:linear_problem})\:
+rates on the order of of $e^{-N}$ \citep{evans_residual_2003}. In much
+the same way as the projection methods outlined in
+Appendix~\ref{ch:linear_problem}, Halton's method, sequential Monte
+Carlo, utilizes the adjoint Monte Carlo solver as a means of directly
+reducing the elements residual vector. He proposed the following
+iterative scheme as a solution to Eq~(\ref{eq:linear_problem})\:
 \begin{subequations}
@@
 they achieved identical numerical results as conventional Krylov
-solvers well as comparable performance in both number of iterations
+solvers as well as comparable performance in both number of iterations
 and CPU time. Their approach was instead to use residual Monte Carlo
@@
 \end{subequations}
-where the adjoint Monte Carlo method is used to generate the solution
-correction from the residual and Richardson's iteration in the first
-step has been rewritten as a residual correction. Using Monte Carlo in
-this way achieves the same effect as Halton's method, decoupling its
-convergence rate from the overall convergence rate of the
-method. Here, the approximate Monte Carlo solution is not driven to a
-particular convergence as it merely supplies a correction for the
+where a Neumann-Ulam Monte Carlo method is used to generate the
+solution correction from the residual and Richardson's iteration in
+the first step has been rewritten as a residual correction. Using
+Monte Carlo in this way achieves the same effect as Halton's method,
+decoupling its convergence rate from the overall convergence rate of
+the method. Here, the approximate Monte Carlo solution is not driven
+to a particular convergence as it merely supplies a correction for the
 initial guess generated by Richardson's iteration. Rather, only a set
-number of histories are required using the adjoint method to generate
-the correction. In addition, the fact that the scheme in
+number of histories are required using the Neumann-Ulam method to
+generate the correction. In addition, the fact that the scheme in
 Eq~(\ref{eq:mcsa_setup_3}) will converge in one iteration if $\ve{A}$
@@
 to tune in an MCSA implementation: the number of Monte Carlo histories
-computed in the adjoint solve during each MCSA iteration, the weight
-cutoff for those histories, and the total MCSA convergence tolerance
-as specified by $\epsilon$.
+computed in the Neumann-Ulam solve during each MCSA iteration, the
+weight cutoff for those histories, and the total MCSA convergence
+tolerance as specified by $\epsilon$.
 
@@
 Appendix~\ref{ch:linear_problem}, subspace methods take on a general
-form in one-dimension using the Petrov-Galerkin conditions as
-constraints for extracting a correction from the search subspace as
-given by Eq~(\ref{eq:linear_projection_iteration}) that may be used
-with MCSA. As an example, for positive-definite but not necessarily
-symmetric problems, the minimal residual (MINRES) iteration
-\citep{saad_iterative_2003} can be used in conjuction with MCSA where
-the residual vector, $\mathbf{r}$, defines the search subspace and the
-action of the linear operator on the residual vector,
-$\mathbf{A}\mathbf{r}$, defines the constraint subspace. Using this,
-we can then define an MCSA scheme that accelerates the MINRES
-iteration:
+form using the Petrov-Galerkin conditions as constraints for
+extracting a correction from the search subspace as given by
+Eq~(\ref{eq:linear_projection_iteration}). Those that are
+one-dimensional may be used with fixed-point MCSA as they only depend
+the previous iteration for information. As an example, for
+positive-definite but not necessarily symmetric problems, the minimal
+residual (MINRES) iteration \citep{saad_iterative_2003} can be used in
+conjunction with MCSA where the residual vector, $\mathbf{r}$, defines
+the search subspace and the action of the linear operator on the
+residual vector, $\mathbf{A}\mathbf{r}$, defines the constraint
+subspace. Using this, we can then define an MCSA scheme that
+accelerates the MINRES iteration:
 \begin{subequations}
@@
 \label{subsubsec:basic_mcsa_preconditioning}
-As an example, to achieve a spectral radius of less than one for
-diagonally dominant matrices point Jacobi preconditioning can be used
-such that the preconditioning matrix $\ve{M}$ is:
+As an example of basic preconditioning, to achieve a spectral radius
+of less than one for diagonally dominant matrices point Jacobi
+preconditioning can be used such that the preconditioning matrix
+$\ve{M}$ is:
 \begin{equation}
@@
 \end{subequations}
-where the adjoint Monte Carlo solve now has a preconditioned operator
-from which to build weights and probabilities for transport and a
-preconditioned source vector to sample.
+where the Neumann-Ulam Monte Carlo solve now has a preconditioned
+operator from which to build weights and probabilities for transport
+and a preconditioned source vector to sample.
 
@@
 because the inversion of $\ve{M}$ is trivial. Second, because the
-adjoint Monte Carlo method used within MCSA to compute the correction
-operates on a linear problem with the preconditioned operator, then
-$\ve{H}$ in the adjoint solver will have a zero term in each of its
-diagonal elements, thereby eliminating all in-state transitions during
-the random walk sequence. Because of this, point Jacobi
-preconditioning should be considered for many classes of problems,
-regardless of any other preconditioning that is applied to the
-system.
+Monte Carlo method used within MCSA to compute the correction operates
+on a linear problem with the preconditioned operator, then $\ve{H}$ in
+the Neumann-Ulam solver will have a zero term in each of its diagonal
+elements, thereby eliminating all in-state transitions during the
+random walk sequence. Because of this, point Jacobi preconditioning
+should be considered for many classes of problems, regardless of any
+other preconditioning that is applied to the system.
+
+\clearpage
 
@@
 \end{equation}
-and note that the precondtioned residual of the substituted variable
+and note that the preconditioned residual of the substituted variable
 is now serving as the source and the new iteration matrix is:
@@
 gives the results of these computations. First, it is worthy to note
-on the semilog plot that we are indeed achieving the expected
+on the semi-log plot that we are indeed achieving the expected
 exponential convergence from MCSA with both Monte Carlo
 solvers. Second, we note that using the adjoint method with the same
-number of stochastic histores per MCSA iteration gives a faster rate
+number of stochastic histories per MCSA iteration gives a faster rate
 of converge for the same reasons as above. We also note here that
@@
 transported through a domain described more accurately by the 9-point
-stencil, thus yielding a more accuracte correction vector from the
+stencil, thus yielding a more accurate correction vector from the
 Monte Carlo calculation.
@@
 histories while the expected value estimator permitted convergence
-with only 2,000 histories. Figure~\ref{fig:estimator_nh_iters} gives
-the number of iterations required to converge for both estimators as a
-function of the number of histories per iteration. For smaller numbers
-of histories, the performance of the expected value estimator is
-significantly better than the collision estimator. This result is
-valuable in that less transport is required to achieve the same MCSA
-iterative performance with the expected value estimator, important for
-situations where transport is expensive (i.e. in domain decomposed
-calculations). Interstingly, as the number of histories per iteration
+with only 2,000 histories.
+
+Figure~\ref{fig:estimator_nh_iters} gives the number of iterations
+required to converge for both estimators as a function of the number
+of histories per iteration. For smaller numbers of histories, the
+performance of the expected value estimator is significantly better
+than the collision estimator. This result is valuable in that less
+transport is required to achieve the same MCSA iterative performance
+with the expected value estimator, important for situations where
+transport is expensive (i.e. in domain decomposed
+calculations). Interestingly, as the number of histories per iteration
 are increased, the iterative performance with the collision estimator
@@
 performance of the expected value estimator at a fractional number of
-histories, one should strongly consider its use over using the collision
-estimator with more histories. The CPU time required to converge is
-presented in Figure~\ref{fig:estimator_nh_time} and reflects the
-results of the iterative performance. In general, using the collision
-estimator is slightly slower overall, but the time per iteration is
-faster due to the fact that the estimator does not have to cycle
-through multiple states during the tally procedure.
+histories, one should strongly consider its use over using the
+collision estimator with more histories. The CPU time required to
+converge is presented in Figure~\ref{fig:estimator_nh_time} and
+reflects the results of the iterative performance. In general, using
+the collision estimator is slightly slower overall, but the time per
+iteration is faster due to the fact that the estimator does not have
+to cycle through multiple states during the tally procedure.
 
@@
 preliminary variance reduction scheme is to introduce
-\textit{artifical absorption} into the system such that at each
+\textit{artificial absorption} into the system such that at each
 transition event, a stochastic history will now have some finite
@@
 \begin{equation}
-  w_{ij} = \frac{sign(h_{ji})\sum_j|h_{ji}|}{(1-p_{abs})}\:.
+  w_{ij} = \frac{sign(h_{ji})}{(1-p_{abs})}\sum_j|h_{ji}|\:.
   \label{eq:absorption_weights}
@@
 on average be located where the system residual is the largest. An
-additional advantage of this formulation is that the full Neumann Ulam
+additional advantage of this formulation is that the full Neumann-Ulam
 decomposition is maintained.
@@
 \sn{1}{-8}. To reduce the effect of the weight cutoff on this study,
-the weight cutoff was set to \sn{1}{-12} such that it is signicantly
+the weight cutoff was set to \sn{1}{-12} such that it is significantly
 more likely that a history will be terminated by absorption rather
@@
 probability. Up to a probability of roughly 0.5, the iterative
-performance is not significanltly affected with the information lost
+performance is not significantly affected with the information lost
 due to shortened random walks causing a small increase in the number
@@
 pushes the solution in the wrong direction. Convergence was observed
-to be lost at probabilities of 0.7 and higher. The random walk length
-has a strong effect on CPU time as
+to be lost at probabilities of 0.7 and higher.
+
+The random walk length has a strong effect on CPU time as
 well. Figure~\ref{fig:absorption_time} gives the CPU time in seconds
@@
 will make the largest contributions to the multiplication and
-therefore recover as much of the original product as possible. For
-each row in the preconditioned system, we can then apply the reduced
-domain approximation to eliminate those elements that either fall
-below a specified tolerance level, keep the $N$ largest elements in
-each row, or both giving:
+therefore recover as much of the original product as possible.
+
+For each row in the preconditioned system, we can then apply the
+reduced domain approximation to eliminate those elements that either
+fall below a specified tolerance level, keep the $N$ largest elements
+in each row, or both giving:
 \begin{subequations}

### 2013-06-10 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
mc background edits and spell check

@@
 with a reduced domain approximation fill level of 100 and a threshold
-of \sn{1}{-10} as in the previous analysis. For each estimator, the
-number of stochastic histories used to compute the MCSA correction was
-varied from 25 to 50,000 with the number of MCSA iterations required
-to converge to a tolerance of \sn{1}{-8} recorded for a single
-eigenvalue iteration along with the CPU time required to achieve
-convergence.
+of \sn{1}{-10} as in the previous analysis. Artificial absorption was
+also introduced with a value of 0.2 for the MCSA calculations that
+used the collision estimator as CPU timing performance was improved
+without a loss in iterative performance while no absorption was used
+for calculations using the expected value estimator. For each
+estimator, the number of stochastic histories used to compute the MCSA
+correction was varied from 25 to 50,000 with the number of MCSA
+iterations required to converge to a tolerance of \sn{1}{-8} recorded
+for a single eigenvalue iteration along with the CPU time required to
+achieve convergence.
 
@@
 production Krylov solvers with which we will compare direct numerical
-results in order to verify MCSA solutions in this section and
-iterative and CPU timing performance in the next order to gauge
-feasibility and identify areas that require further research.
+results in order to verify MCSA solutions in this section.
 
@@
       fuel assembly problem.} \textit{Each calculation was converged
-      with the same eigensolver and linear solver tolerances.}}
+      with the same eigensolver and linear solver
+      tolerances. Table~\ref{tab:spn_solver_defs} gives the
+      description for each solver type presented.}}
   \label{tab:serial_ev_results}
@@
       right: GMRES-ILUT, MCSA-ILUT-R-C, MCSA-ILUT-MR-C,
-      MCSA-ILUT-R-EV, MCSA-ILUT-MR-EV.}}
+      MCSA-ILUT-R-EV, MCSA-ILUT-MR-EV. Table~\ref{tab:spn_solver_defs}
+      gives the description for each solver type presented.}}
   \label{fig:serial_1_group_diff}
@@
 3 for the 4-group calculation were computed with the BiCGStab-ILUT
-solution in group 3 for the 4-group calculation as a reference.
+solution in group 3 for the 4-group calculation as a reference. For
+all group flux results given in the table, the maximum differences for
+all cases was less than the linear solver tolerance used to converge
+the flux at each eigenvalue iteration meaning that all values over the
+entire energy and spatial domain are within these limits. In addition,
+the minimum error for all calculations was observed to be zero,
+meaning in at least one mesh cell in all calculations for all energy
+groups the scalar flux values computed by all solvers were
+identical. Based on the flux results and the eigenvalue results
+presented here, the MCSA results should be deemed correct for the fuel
+assembly problem for serial calculations.
 
@@
       was converged with the same eigensolver and linear solver
-      tolerances.}}
+      tolerances. Table~\ref{tab:spn_solver_defs} gives the
+      description for each solver type presented.}}
   \label{tab:serial_differences_g1}
@@
       was converged with the same eigensolver and linear solver
-      tolerances.}}
+      tolerances. Table~\ref{tab:spn_solver_defs} gives the
+      description for each solver type presented.}}
   \label{tab:serial_differences_g2}
@@
       was converged with the same eigensolver and linear solver
-      tolerances.}}
+      tolerances. Table~\ref{tab:spn_solver_defs} gives the
+      description for each solver type presented.}}
   \label{tab:serial_differences_g4}
 \end{table}
+
+\clearpage
 
@@
 \label{sec:spn_comparison}
-Using the same set of solver parameters used for verification, MCSA
-will be directly compared to the production Krylov solvers in terms of
-both iterative performance and CPU timing for solutions to the fuel
-assembly problem. In addition to the parameters used for the
-verification analysis, artificial absorption was also introduced with
-a value of 0.2 for the MCSA calculations that used the collision
-esitmator as CPU timing performance was improved without a loss in
-iterative performance. Adding this absorption did not change the
-numerical results of the calculation.
-
-Figure~\ref{fig:spn_comparison_iterations} gives the average number of
-linear solver iterations required to converge a single eigenvalue
-iteration of the fuel assembly problem as a function of the number of
-energy groups. In general, the iterative performance of all the
-solvers tested is comparable with BiCGStab performing the best. We
+In the previous section, numerical results using MCSA for the fuel
+assembly calculation were directly with production Krylov methods in
+order to verify their correctness. Deemed correct, with the same set
+of solver parameters used for verification MCSA will now be directly
+compared to the same production Krylov solvers in terms of both
+iterative performance and CPU timing for solutions to the fuel
+assembly problem.  This comparison of both iterative and CPU timing
+performance is necessary in order to gauge feasibility of the method
+and identify areas that require further research and development.
+
+To compare iterative performance, the number of linear solver
+iterations required to converge each eigenvalue iteration was recorded
+and then averaged over all eigenvalue iterations for each
+solver. Figure~\ref{fig:spn_comparison_iterations} gives the average
+number of linear solver iterations required to converge a single
+eigenvalue iteration of the fuel assembly problem as a function of the
+number of energy groups. In general, the iterative performance of all
+the solvers tested is comparable with BiCGStab performing the best. We
 expect this because not only are the multigroup $SP_N$ equations
@@
 increased and therefore the work performed for the problem is of
-$O(N)$. It is also important to note the the reduced domain
-approximation parameters, particularly the fill level, were fixed as
-the number of energy groups was increased. By fixing the fill level,
-we are effectively fixing the amount of information contained in the
+$O(N)$. It is also important to note the reduced domain approximation
+parameters, particularly the fill level, were fixed as the number of
+energy groups was increased. By fixing the fill level, we are
+effectively fixing the amount of information contained in the
 composite linear operator available for the Monte Carlo
@@
 observed for production Krylov implementations using exactly the same
-preconditioning scheme, timing performance is not as
-comparable. Figure~\ref{fig:spn_comparison_time} gives the average CPU
+preconditioning scheme, timing performance is not as competetive. For
+this comparison, the amount of time to perform each linear solver
+iteration in every eigenvalue iteration was recorded and then averaged
+over all linear solver
+iterations. Figure~\ref{fig:spn_comparison_time} gives the average CPU
 time per linear solver iteration required to converge the fuel

### 2013-06-11 doc/dissertation/document/backmatter/diffusion_problem.tex
adding some example figures for domain decomposed monte carlo

--- 2013-06-10 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
@@
-\chapter{Monte Carlo Synthetic Acceleration Methods for the $SP_N$ Equations}
-\label{ch:spn_equations}
-
-The neutron transport problem is complicated. Solutions cover a large
-phase space and the problems of interest are often geometrically
-complex, very large, or both, requiring tremendous computational
-resources to generate an adequate solution. Modern deterministic
-methods for large scale problems are commonly variants on the discrete
-ordinates ($S_N$) method \citep{evans_denovo:_2010}. For fission
-reactor neutronics simulations, the $S_N$ method requires potentially
-trillions of unknown angular flux moments to be computed to achieve
-good accuracy for the responses of interest
-\citep{slaybaugh_acceleration_2011}. Other forms of the transport
-problem, including the $P_N$ method, take on a simpler form than the
-more common $S_N$ methods but lack in accuracy when compared while
-still requiring considerable computational resources for solutions in
-multiple dimensions.
-
-In the 1960's, Gelbard developed an ad-hoc multidimensional extension
-of the simple single dimension planar $P_N$ equations that created a
-system of coupled, diffusion-like equations known as the simplified
-$P_N$ ($SP_N$) equations. Up until around the 1990's, the $SP_N$
-method was either widely unknown, widely unused, or combination of
-both even though numerical studies showed promising results with
-better solutions than diffusion theory and a significant reduction in
-computational time over more accurate methods such as discrete
-ordinates. Why did this happen? A significant problem, pointed out by
-Brantley and Larsen \citep{brantley_simplified_2000}, was that little
-rigor had been applied to the formulation of the $SP_N$ equations
-since their derivation through primarily heuristic arguments. Instead,
-studies at that time focused on simply comparing the results of the
-method to other contemporary transport solution strategies. In
-addition, many problems of interest from the literature at the time
-were either solved using nodal-type methods for reactor-sized problems
-or $S_N$-type methods for benchmark problems with intricate material
-configurations and potentially large flux gradients over small spatial
-domains.
-
-So why reconsider the $SP_N$ equations? Starting in the 1990's and
-primarily due to Larsen and his colleagues, the $SP_N$ equations have
-been given a more rigorous treatment with both variational and
-asymptotic derivations performed as a means of verification. In
-addition, these equations have been more rigorously studied as
-solution methods to MOX fuel problems and have been shown to provide
-accurate solutions. With this mathematical literature to provide a
-solid numerical footing for the method, we look at its application to
-today's challenge problems in neutron transport for fission reactor
-analysis. The reduction in numerical complexity of current
-deterministic solution methods using the $S_N$ approximation could
-mean significant savings in both compute time and memory required. In
-addition, the characteristics of the solution to the transport problem
-for a steady state reactor core permit diffusion theory to be used; a
-staple of the nuclear industry since its inception. Therefore, if
-diffusion theory is applicable, then finer grained solutions that
-capture more of the physics contained in the transport equation should
-be possible with the $SP_N$ method. In doing so, we also expect from
-the literature to obtain computed responses on the order of accuracy
-we would expect from an appropriately discretized $S_N$ method at a
-fraction of the cost.
-
-To further motivate moving in this direction, recent developments in
-the Exnihilo neutronics package at Oak Ridge National Laboratory have
-permitted generation of the $SP_N$ system of equations for detailed
-neutronics models \citep{evans_simpli_2013}. By fully forming these
-equations and formulating them as a linear algebra problem instead of
-using the explicit iterative methods of the past, we now have access
-to all of the modern advancements in computational linear algebra
-including Krylov solvers for asymmetric systems and preconditioning
-methods such as algebraic multigrid. This leads us to then explore the
-applicability of our work in discrete Monte Carlo methods for linear
-systems as a possible solution method for the $SP_N$ equations. If
-formulated correctly, we hypothesize that significant improvement in
-usage of computational resources may be observed compared to modern
-solution techniques such as those suggested due to the form of the
-matrices generated by the discretization. In addition, solving the
-$SP_N$ equations in this way also breaks away from the $S_N$ forms of
-parallelism where spatial parallelism is achieved by an efficient
-parallel sweep, angular efficiency achieved by pipe-lining, and energy
-parallelism achieved by decoupling the groups. With the $SP_N$
-equations as a full matrix system, we now can parallelize the problem
-as prescribed by the linear solver, which may be significantly more
-scalable than current $S_N$ transport practices.
-
-In this chapter, we derive the $SP_N$ equations, closely following the
-work of Evans\footnote{The $SP_N$ derivations in this chapter are
-  heavily based on those presented by Evans in
-  \citep{evans_simpli_2013}}, in order to gain full understanding of
-the underlying system and its potential behavior in a Monte Carlo
-context. From the $P_N$ equations, we apply a set of approximations to
-yield the $SP_N$ equations for fixed source and criticality
-problems. Using the fully-formed linear operator for the transport
-problem, we explore solutions to the $SP_N$ equations using Monte
-Carlo Synthetic Acceleration using a light water reactor fuel assembly
-criticality calculation as the driving problem. In order to
-effectively solve this problem, a suite of preconditioners is
-developed and studied within the context of MCSA. Using these
-preconditioners, MCSA solutions for the fuel assembly problem are
-verified by comparing against production Krylov solvers and finally
-performance is analyzed through comparison with those same Krylov
-solvers.
-
-%%---------------------------------------------------------------------------%%
-\section{The Neutron Transport Equation}
-\label{sec:transport_eq}
-As a starting point for the $SP_N$ equations we define the
-time-independent neutron transport equation
-\citep{lewis_computational_1993}:
-\begin{multline}
-  \hat{\Omega} \cdot \vec{\nabla} \psi(\vec{r},\hat{\Omega},E) +
-  \sigma(\vec{r},E) \psi(\vec{r},\hat{\Omega},E) = \\ \iint
-  \sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}' \cdot \hat{\Omega})
-  \psi(\vec{r},\hat{\Omega}',E') d\Omega' dE' +
-  q(\vec{r},\hat{\Omega},E)\:,
-  \label{eq:general_transport}
-\end{multline}
-with the variables defined as:
-\begin{itemize}
-\item $\vec{r}$ - neutron spatial position
-\item $\hat{\Omega}$ - neutron streaming direction with radial
-  component $\mu$ and azimuthal component $\omega$
-\item $\hat{\Omega}' \cdot \hat{\Omega} = \mu_0$ is the angle of
-  scattering
-\item $E$ - neutron energy
-\item $\psi(\vec{r},\hat{\Omega},E)$ - angular flux
-\item $\sigma(\vec{r},E)$ - total interaction cross section
-\item $\sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}')$ - probability
-  of scattering from direction $\hat{\Omega}'$ into an angular domain
-  $d\hat{\Omega}'$ about the direction $\hat{\Omega}$ and from energy
-  $E'$ to an energy domain $dE'$ about energy $E$
-\item $q(\vec{r},\hat{\Omega},E)$ - external source of neutrons.
-\end{itemize}
-For this work, it is sufficient to formulate
-Eq~(\ref{eq:general_transport}) in 1-dimensional Cartesian geometry:
-\begin{multline}
-  \mu \frac{\partial}{\partial x} \psi(x,\mu,E) + \sigma(x,E)
-  \psi(x,\mu,E) = \\ \iint \sigma_s(x,E' \rightarrow E,\hat{\Omega}'
-  \cdot \hat{\Omega}) \psi(x,\hat{\Omega}',E') d\Omega' dE' +
-  \frac{q(x,E)}{4 \pi}\:,
-  \label{eq:cart_1d_transport}
-\end{multline}
-where the angular component of the solution is no longer dependent on
-the azimuthal direction of travel and an isotropic source of neutrons
-is assumed. In addition, for fission systems the eigenvalue form of
-the transport equation is:
-\begin{multline}
-  \hat{\Omega} \cdot \vec{\nabla} \psi(\vec{r},\hat{\Omega},E) +
-  \sigma(\vec{r},E) \psi(\vec{r},\hat{\Omega},E) = \\ \iint
-  \sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}' \cdot \hat{\Omega})
-  \psi(\vec{r},\hat{\Omega}',E') d\Omega' dE' + \\ \frac{1}{k} \chi(E)
-  \iint \nu \sigma_f(\hat{r},E') \psi(\vec{r},\hat{\Omega}',E')
-  d\Omega' dE' + q(\vec{r},\hat{\Omega},E) \:,
-  \label{eq:eigenvalue_transport}
-\end{multline}
-with the additional variables defined as
-\begin{itemize}
-\item $k$ - multiplication factor
-\item $\chi(E)$ - fission neutron energy spectrum
-\item $\nu$ - average number of neutrons per fission
-\item $\sigma_f(r,E')$ - fission cross section\:.
-\end{itemize}
-In 1-dimensional Cartesian geometry,
-Eq~(\ref{eq:eigenvalue_transport}) becomes:
-\begin{multline}
-  \mu \frac{\partial}{\partial x} \psi(x,\mu,E) + \sigma(x,E)
-  \psi(x,\mu,E) = \\ \iint \sigma_s(x,E' \rightarrow
-  E,\hat{\Omega}' \cdot \hat{\Omega}) \psi(x,\hat{\Omega}',E')
-  d\Omega' dE' + \\ \frac{1}{k} \chi(E)
-  \iint \nu \sigma_f(x,E') \psi(x,\hat{\Omega}',E') d\Omega'
-  dE' + \frac{q(x,E)}{4 \pi}\:.
-  \label{eq:cart_1d_eigenvalue}
-\end{multline}
-
-%%---------------------------------------------------------------------------%%
-\section{Derivation of the Monoenergetic $SP_N$ Equations}
-\label{sec:spn_equations}
-The $P_N$ equations as derived in Appendix~\ref{chap:pn_equations}
-give $N+1$ coupled first-order equations capturing the spatial and
-angular-dependence of the solution. In multiple dimensions, the
-equation set becomes large and coupled not only through angular
-moments but also through the spatial variables. As a simpler
-alternative to multidimensional $P_N$ solutions, Gelbard recognized in
-1960 that the planar $P_N$ equations could be simplified and applied
-an ad-hoc method to extend them to multiple dimensions, yielding the
-$SP_N$ equations. These equations are not only fewer in number, but
-also take on a diffusion-like form while maintaining the angular
-character of the flux, making them amenable to solutions with modern
-diffusion methods.
-
-First, the $P_N$ equations can be simplified to $(N+1)/2$ second-order
-equations by solving for the $n^{th}$ Legendre flux moment in the
-odd-order equations:
-\begin{equation}
-  \phi_n = \frac{1}{\Sigma_n}\Bigg[ q \delta_{no} -
-    \frac{\partial}{\partial x}\Big(\frac{n}{2n+1}\phi_{n-1} +
-    \frac{n+1}{2n+1} \phi_{n+1} \Big) \Bigg]\:, 
-  \label{eq:odd_moments}
-\end{equation}
-for $n = 1,3,\cdots,N$ and $\delta_{no} = 0\ \forall n \neq 0$. We can
-insert the odd moments into Eq~(\ref{eq:final_pn_equations}) to get a
-reduced group of equations for the even moments:
-\begin{multline}
-  -\frac{\partial}{\partial x}
-  \Bigg[\frac{n}{2n+1}\frac{1}{\Sigma_{n-1}} \frac{\partial}{\partial
-      x} \Big(\frac{n-1}{2n-1} \phi_{n-2} + \frac{n}{2n-1}\phi_n \Big)
-    \\+ \frac{n+1}{2n+1}\frac{1}{\Sigma_{n+1}} \frac{\partial}{\partial
-      x} \Big(\frac{n+1}{2n+3}\phi_n + \frac{n+2}{2n+3}\phi_{n+2}\Big)
-    \Bigg] \\+ \Sigma_n \phi_n = q \delta_{n0}\ \ \ \ \ \ \ \ \ n =
-  0,2,4,\cdots,N\:.
-  \label{eq:reduced_pn}
-\end{multline}
-Immediately, we note the diffusion-like nature of
-Eq~(\ref{eq:reduced_pn}) as compared to the original $P_N$
-equations. To extend these equations to multiple dimensions, Gelbard
-simply replaced the planar spatial derivatives in the reduced set of
-equations with general multidimensional gradient operators:
-\begin{multline}
-  -\nabla \cdot \Bigg[\frac{n}{2n+1}\frac{1}{\Sigma_{n-1}} \nabla
-    \Big(\frac{n-1}{2n-1} \phi_{n-2} + \frac{n}{2n-1}\phi_n \Big) \\+
-    \frac{n+1}{2n+1}\frac{1}{\Sigma_{n+1}} \nabla
-    \Big(\frac{n+1}{2n+3}\phi_n + \frac{n+2}{2n+3}\phi_{n+2}\Big)
-    \Bigg] \\+ \Sigma_n \phi_n = q \delta_{n0}\ \ \ \ \ \ \ \ \ n =
-  0,2,4,\cdots,N\:,
-  \label{eq:spn_equations}
-\end{multline}
-yielding a multidimensional set of $(N+1)/1$ angular coupled equations
-defined as the $SP_N$ equations. As with the $P_N$ equations, we
-provide closure to this set of equations with $\phi_{N+1} = 0$. As a
-concrete example, we will consider the $SP_7$ equations:
-\begin{subequations}
-  \begin{gather}
-    -\nabla \cdot \frac{1}{3 \Sigma_1} \nabla ( \phi_0 + 2\phi_2 ) +
-    \Sigma_0 \phi_0 = q \\ 
-    -\nabla \cdot \Bigg[ \frac{2}{15 \Sigma_1} \nabla ( \phi_0 + 2\phi_2
-      ) + \frac{3}{35 \Sigma_3}\nabla( 3\phi_2 + 4\phi_4)\Bigg] +
-    \Sigma_2 \phi_2 = 0\\
-    -\nabla \cdot \Bigg[ \frac{4}{63 \Sigma_3} \nabla ( 3\phi_2 +
-      4\phi_4 ) + \frac{5}{99 \Sigma_5}\nabla( 5\phi_4 +
-      6\phi_6)\Bigg] + \Sigma_4 \phi_4 = 0\\
-    -\nabla \cdot \Bigg[ \frac{6}{143 \Sigma_5} \nabla ( 5\phi_4 +
-      6\phi_6 ) + \frac{7}{195 \Sigma_7}\nabla(7\phi_6)\Bigg] +
-    \Sigma_6 \phi_6 = 0 \:.
-  \end{gather}
-  \label{eq:sp7_equations}
-\end{subequations}
-To further modify these equations, we can use a change of variables to
-create a new group of equations such that the gradients are operating
-on a single vector:
-\begin{subequations}
-  \begin{gather}
-    u_1 = \phi_0 + 2\phi_2 \\
-    u_2 = 3\phi_2 + 4\phi_4 \\
-    u_3 = 5\phi_4 + 6\phi_6 \\
-    u_4 = 7\phi_6 \:,
-  \end{gather}
-  \label{eq:spn7_subs}
-\end{subequations}
-or equivalently:
-\begin{subequations}
-  \begin{gather}
-    \phi_0 = u_1 - \frac{2}{3}u_2 + \frac{8}{15}u_3 -
-    \frac{16}{35}u_4 \\
-    \phi_2 = \frac{1}{3}u_2 - \frac{4}{15}u_3 + \frac{8}{35}u_4\\ 
-    \phi_4 = \frac{1}{5}u_3 - \frac{6}{35}u_4\\
-    \phi_6 = \frac{1}{7}u_4\:.
-  \end{gather}
-  \label{eq:spn7_subs_inverse}
-\end{subequations}
-When substituted into Eq~(\ref{eq:sp7_equations}), these terms give:
-\begin{subequations}
-  \begin{gather}
-    -\nabla \cdot \frac{1}{3 \Sigma_1} \nabla u_1 + \Sigma_0 \Bigg[
-    u_1 - \frac{2}{3}u_2 + \frac{8}{15}u_3 - \frac{16}{35}u_4 \Bigg]
-    = -q \\
-    -\nabla \cdot \Bigg[ \frac{2}{15 \Sigma_1} \nabla u_1 +
-    \frac{3}{35 \Sigma_3} \nabla u_2 \Bigg] + \Sigma_2 \Bigg[
-    \frac{1}{3}u_2 - \frac{4}{15}u_3 + \frac{8}{35}u_4 \Bigg] = 0 \\
-    -\nabla \cdot \Bigg[ \frac{4}{63 \Sigma_3} \nabla u_2 +
-    \frac{5}{99 \Sigma_5} \nabla u_3 \Bigg] + \Sigma_4 \Bigg[
-    \frac{1}{5}u_3 - \frac{6}{35}u_4 \Bigg] = 0 \\ 
-    -\nabla \cdot \Bigg[ \frac{6}{143 \Sigma_5} \nabla u_3 +
-    \frac{7}{195 \Sigma_7} \nabla u_4 \Bigg] + \Sigma_6 \Bigg[
-    \frac{1}{7}u_4 \Bigg] = 0 \:.
-  \end{gather}
-  \label{eq:spn7_subs_equations}
-\end{subequations}
-If we rearrange Eq~(\ref{eq:spn7_subs_equations}) such that only one
-divergence operation is present in each equation, we can formulate
-this as a matrix system of 4 equations in the case of the $SP_7$
-approximation:
-\begin{equation}
-  -\nabla \cdot D_n \nabla u_n + \sum_{m=1}^4 A_{nm} u_m =
-  q_n\ \ \ \ \ \ \ n = 1,2,3,4\:,
-  \label{eq:spn_matrix}
-\end{equation}
-with $\mathbf{u}$ the vector of solution variables:
-\begin{equation}
-  \mathbf{u} = ( u_1\ \ u_2\ \ u_3\ \ u_4 )^T \:,
-  \label{eq:spn7_solution_vector}
-\end{equation}
-$\mathbf{D}$ the vector of effective diffusion coefficients:
-\begin{equation}
-  \mathbf{D} = \Bigg( \frac{1}{3\Sigma_1}\ \ \frac{1}{7\Sigma_3}\ \
-  \frac{1}{11\Sigma_5}\ \ \frac{1}{15\Sigma_7} \Bigg)^T\:,
-  \label{eq:spn7_diffusion_coeffs}
-\end{equation}
-$\mathbf{q}$ the vector of source terms where the $0^{th}$ moment
-source has now been distributed through the system:
-\begin{equation}
-  \mathbf{q} = (
-  q\ \ -\frac{2}{3}q\ \ \frac{8}{15}q\ \ -\frac{16}{35}q )^T\:,
-  \label{eq:spn7_source_vector}
-\end{equation}
-and $\mathbf{A}$ a matrix of angular scattering terms:
-% NOTE: I copied the following matrix directly out of Tom's tech note
-% on the SPn equations which I am effectively following here because I
-% was feeling lazy. I have verified its correctness.
-\begin{equation}
-  \mathbf{A} = 
-  {\tiny \begin{bmatrix}
-    (\Sigma_0) &
-    (-\frac{2}{3}\Sigma_0) &
-    (\frac{8}{15}\Sigma_0) &
-    (-\frac{16}{35}\Sigma_0) \\
-    %%
-    &&&\\
-    %%
-    (-\frac{2}{3}\Sigma_0) &
-    (\frac{4}{9}\Sigma_0 + \frac{5}{9}\Sigma_2) &
-    (-\frac{16}{45}\Sigma_0 - \frac{4}{9}\Sigma_2) &
-    (\frac{32}{105}\Sigma_0 + \frac{8}{21}\Sigma_2) \\
-    %%
-    &&&\\
-    %%
-    (\frac{8}{15}\Sigma_0) &
-    (-\frac{16}{45}\Sigma_0 - \frac{4}{9}\Sigma_2) &
-    (\frac{64}{225}\Sigma_0 + \frac{16}{45}\Sigma_2 + \frac{9}{25}\Sigma_4) &
-    (-\frac{128}{525}\Sigma_0 - \frac{32}{105}\Sigma_2 - \frac{54}{175}\Sigma_4)
-    \\ 
-    %%
-    &&&\\
-    %%
-    (-\frac{16}{35}\Sigma_0) &
-    (\frac{32}{105}\Sigma_0 + \frac{8}{21}\Sigma_2) &
-    (-\frac{128}{525}\Sigma_0 - \frac{32}{105}\Sigma_2 - \frac{54}{175}\Sigma_4)
-    & 
-    (\frac{256}{1225}\Sigma_0 + \frac{64}{245}\Sigma_2 +
-    \frac{324}{1225}\Sigma_4 + \frac{13}{49}\Sigma_6)
-  \end{bmatrix}}\:.
-  \label{eq:A_matrix}
-\end{equation}
-Note that the term $\sum_{m=1}^4 A_{nm} u_m$ in
-Eq~(\ref{eq:spn_matrix}) couples the moments in each equation while
-the diffusive term in each equation is only for a single
-'pseudo-moment' $u_n$. As noted by Evans \citep{evans_simpli_2013},
-lower order $SP_N$ approximations can be generated by setting higher
-order even moments in this system to zero (e.g. $\phi_6 = \phi_4 = 0$
-yields the $SP_3$ equations). Boundary conditions for these equations
-are provided in Appendix~\ref{chap:spn_boundary_conditions}, the
-multigroup form is presented in Appendix~\ref{chap:mg_spn_equations},
-and the spatial discretization in
-Appendix~\ref{chap:spn_spatial_discretization}.
-
-\subsection{Eigenvalue Form of the $SP_N$ Equations}
-\label{subsec:eigenvalue_form}
-For criticality problems, the space, angle, and energy discretizations
-for the $SP_N$ equations can be applied to the general eigenvalue from
-of the transport equation given by
-Eq~(\ref{eq:eigenvalue_transport}). We can readily replace the fixed
-source in Eq~(\ref{eq:multigroup_spn_equations}) with the fission
-source derived in \S~\ref{sec:pn_eigenvalue}:
-\begin{multline}
-  -\nabla \cdot \Bigg[\frac{n}{2n+1}\mathbf{\Sigma_{n-1}}^{-1} \nabla
-    \Big(\frac{n-1}{2n-1} \mathbf{\Phi_{n-2}} +
-    \frac{n}{2n-1}\mathbf{\Phi_n} \Big) \\+
-    \frac{n+1}{2n+1}\mathbf{\Sigma_{n+1}}^{-1} \nabla
-    \Big(\frac{n+1}{2n+3}\mathbf{\Phi_n} +
-    \frac{n+2}{2n+3}\mathbf{\Phi_{n+2}}\Big) \Bigg] \\+
-  \mathbf{\Sigma_n} \mathbf{\Phi_n} = \frac{1}{k} \mathbf{F}
-  \mathbf{\Phi_n} \delta_{n0} \ \ \ \ \ \ \ \ \ n = 0,2,4,\cdots,N\:.
-  \label{eq:multigroup_spn_eigenvalue}
-\end{multline}
-with the fission matrix $\mathbf{F}$ given by
-Eq~(\ref{eq:fission_matrix}). We again apply the change of variables
-to yield the set of pseudo-moments $\mathbb{U}_n$ where now the
-application of the fission matrix to the moment vectors will be
-expanded into a set of block matrices in identical fashion to the
-scattering matrices.
-
-It should be noted here that with respect to a general MCSA scheme,
-the introduction of the fission in the system will only affect the
-source vector in the linear system. The linear operator, and therefore
-overall MCSA performance will be dictated by streaming and scattering
-as defined on the left-hand side.
-
-%%---------------------------------------------------------------------------%%
-\section{Spectral Analysis of the $SP_N$ Equations}
-\label{sec:spn_spectral_analysis}
-Before we can move on to solving the $SP_N$ equations with both
-conventional and Monte Carlo methods, we must first understand their
-spectral character to ensure that MCSA will be applicable. As
-presented in Chapter~\ref{ch:stochastic_methods}, the spectral radius
-of the iteration matrix must be less than unity to ensure
-convergence. Therefore, we will compute the spectral radius of the
-iteration matrix formed by the $SP_N$ equations while parametrically
-varying the $SP_N$ order, the $P_N$ order, the number of energy
-groups, and upscattering/downscattering between energy groups. For
-each problem in the parameter study, the values given in
-Table~\ref{tab:spn_fixed_parameters} were fixed for each spatial
-location and energy group.
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{ll}\hline\hline
-      \multicolumn{1}{c}{Parameter}& 
-      \multicolumn{1}{c}{Value} \\\hline
-      Mesh Element Size & 1.0 \\
-      Mesh Elements & $4 \times 4 \times 4$ \\
-      Reflecting Boundaries & True \\
-      Materials & 1 \\
-      Source Strength & 1.0 \\
-      Total Cross Section & 5.0 \\
-      In-Group Cross Section & 0.25 \\
-      Downscatter Cross Section & 1.0 \\
-      Upscatter Cross Section & 0.1 \\
-      Eigenvalue Solver Tolerance & $1.0\times10^{-8}$ \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Fixed parameters for the $SP_N$ spectral radius
-      parameter study.} \textit{The parameters were fixed for each
-      spatial location and energy group.}}
-  \label{tab:spn_fixed_parameters}
-\end{table}
-Mesh element sizes were the same in all cardinal directions with
-reflecting conditions on each boundary. For the energy parameter, 1
-and 10 energy groups were used with upscatter/downscatter varied for
-the 10 group case. For the downscatter cases, all groups could scatter
-to all lower energy groups with the cross sections given in
-Table~\ref{tab:spn_fixed_parameters} while for the upscatter case all
-groups could scatter to all other energy groups with different cross
-sections for upscatter and downscatter used. A uniform isotropic
-source of neutrons is given in each group as well.
-
-The matrices generated by the discretization of this problem are
-sparse with a block-based form as dictated by
-Eq~(\ref{eq:spn_multigroup_system}). Figure~\ref{fig:group1} gives the
-sparsity pattern for a monoenergetic $SP_7$ discretization while
-Figures~\ref{fig:group10ds} and \ref{fig:group10us} give the sparsity
-patterns for the same problem for 10 energy groups without and with
-upscatter respectively. For these figures, the parameters in
-Table~\ref{tab:spn_fixed_parameters} with the number of spatial
-elements reduced to $2 \times 2 \times 2$ to show the blocks generated
-by the discretization.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4in]{chapters/spn_equations/group1.png}
-  \end{center}
-  \caption{\textbf{Sparsity pattern for 1-group $SP_7$
-      discretization.} \textit{A $2\times 2 \times 2$ element mesh was
-      used to show detail of the blocks formed by the discretization.}}
-  \label{fig:group1}
-\end{figure}
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4in]{chapters/spn_equations/group10ds.png}
-  \end{center}
-  \caption{\textbf{Sparsity pattern for 10-group $SP_7$ discretization
-      with downscatter only.} \textit{A $2\times 2 \times 2$ element
-      mesh was used to show detail of the blocks formed by the
-      discretization.}}
-  \label{fig:group10ds}
-\end{figure}
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4in]{chapters/spn_equations/group10us.png}
-  \end{center}
-  \caption{\textbf{Sparsity pattern for 10-group $SP_7$ discretization
-      with downscatter and upscatter.} \textit{A $2\times 2 \times 2$ element
-      mesh was used to show detail of the blocks formed by the
-      discretization.}}
-  \label{fig:group10us}
-\end{figure}
-We note a few key features of the sparsity plots. The first is that
-for multigroup problems without full upscatter and downscatter
-(i.e. Figure~\ref{fig:group10ds}), the resulting matrix is asymmetric
-and therefore a linear solver that can handle asymmetric linear
-systems is required. Nearly all problems of interest will not have
-full upscattering or downscattering. Second we note the largely
-diagonal character of these systems, although the blocks from
-Eq~(\ref{eq:A_block_matrix}) are readily apparent. Our first attempt
-at preconditioning this system will be to use the point Jacobi
-preconditioning from \S\ref{subsubsec:basic_mcsa_preconditioning} due
-to this diagonal form.
-
-\clearpage
-
-\subsection{Point Jacobi Spectral Analysis Results}
-\label{subsec:spn_analysis_results}
-Spectral radius computations were performed for the cases described
-above for the point Jacobi preconditioned iteration matrix $\ve{H} =
-\ve{I}-\ve{M}^{-1}\ve{A}$ with $\ve{M} =
-diag(\ve{A})$. Table~\ref{tab:group1pj} gives the results for the
-1-group case, Table~\ref{tab:group10dspj} for the 10-group case with
-full downscatter only and Table~\ref{tab:group10uspj} for the 10-group
-case with full downscatter and full upscatter.
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{cccccc}\hline\hline
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{$SP_N$ Order}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{} \\
-       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
-       & \textbf{0} & 0.0635 & 0.6722 & 1.3144 & 1.976 \\
-       & \textbf{1} & 0.0666 & 0.6728 & 1.3141 & 1.9755 \\
-      $P_N$ Order & \textbf{3} & 0.0666 & 0.6822 & 1.3141 & 1.9755 \\
-       & \textbf{5} & 0.0666 & 0.6822 & 1.3278 & 1.9847 \\
-       & \textbf{7} & 0.0666 & 0.6822 & 1.3278 & 1.9917 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Spectral radius results for the point Jacobi
-      preconditioned iteration matrix with 1 energy group.}}
-  \label{tab:group1pj}
-\end{table}
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{cccccc}\hline\hline
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{$SP_N$ Order}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{} \\
-       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
-       & \textbf{0} & 0.0655 & 0.677 & 1.32 & 1.982 \\
-       & \textbf{1} & 0.071 & 0.6777 & 1.319 & 1.982 \\
-      $P_N$ Order & \textbf{3} & 0.071 & 0.687 & 1.327 & 1.9872 \\
-       & \textbf{5} & 0.071 & 0.687 & 1.336 & 1.997 \\
-       & \textbf{7} & 0.071 & 0.687 & 1.336 & 1.9995 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Spectral radius results for the point Jacobi
-      preconditioned iteration matrix with 10 energy groups and full
-      downscatter.}}
-  \label{tab:group10dspj}
-\end{table}
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{cccccc}\hline\hline
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{$SP_N$ Order}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{} \\
-       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
-       & \textbf{0} & 0.7283 & 0.81 & 1.47 & 2.1446 \\
-       & \textbf{1} & 0.7317 & 0.8 & 1.46 & 2.1368 \\
-      $P_N$ Order & \textbf{3} & 0.7317 & 0.91 & 1.526 & 2.2274 \\
-       & \textbf{5} & 0.7317 & 0.91 & 1.5344 & 2.2562 \\
-       & \textbf{7} & 0.7317 & 0.91 & 1.5345 & 2.2842 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Spectral radius results for the point Jacobi
-      preconditioned iteration matrix with 10 energy groups, full
-      downscatter and full upscatter.}}
-  \label{tab:group10uspj}
-\end{table}
-It is readily apparent from the tabulated data that point Jacobi
-preconditioning is insufficient. For problems of order $SP_5$ and
-$SP_7$, the method will not converge at all and for the upscatter case
-with $SP_3$ the spectral radius is still quite large for large $P_N$
-orders and therefore convergence is expected to be slow. These
-eigenvalues signal a need for a better preconditioning strategy to
-both ensure and improve convergence for Monte Carlo methods.
-
-\subsection{Block Jacobi Preconditioning}
-\label{sec:spn_preconditioning}
-If Monte Carlo methods are to be used to solve the $SP_N$ system of
-equations, a different preconditioning strategy is required in order
-to ensure convergence for systems of all $SP_N$ and $P_N$ orders with
-arbitrary energy group structures. As another means of achieving this,
-we look back to the sparsity plots we generated in
-Figures~\ref{fig:group1}, \ref{fig:group10ds} and \ref{fig:group10us}
-as well as the multigroup $SP_N$ equations. Initially, the diagonal
-character of the system led us to try point Jacobi preconditioning
-with only marginal results. From the sparsity plots we note the block
-structure that ultimately arises from the multigroup scattering
-matrices and their insertion into Eq~(\ref{eq:A_block_matrix}). When
-full upscatter and downscatter are used the resulting blocks are
-completely dense while only downscatter gives a lower triangular
-scattering matrix and the block structure shown in
-Figure~\ref{fig:group10ds}.
-
-Based on this both block and diagonally dominant structure for
-matrices formed by the general multigroup $SP_N$ equations, we instead
-choose \textit{block Jacobi} preconditioning as a left preconditioner
-for the system. Like point Jacobi preconditioning, block Jacobi
-preconditioning extracts the diagonal elements of the matrix as the
-preconditioner where now the elements extracted are the blocks on the
-diagonal as shown on the left side of
-Figure~\ref{fig:block_jacobi_ex}. Shown on the right side of
-Figure~\ref{fig:block_jacobi_ex}, inversion of the preconditioner is
-trivial with each diagonal block inverted separately. For the $SP_N$
-equations, Eq~(\ref{eq:A_block_matrix}) gives a block size of
-$N_g\times(N+1)/2$.
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.5}{
-    \input{chapters/spn_equations/block_jacobi.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Block Jacobi preconditioning strategy used for the
-      $SP_N$ equations.} \textit{Left: The preconditioner is formed by
-      the diagonal blocks of the matrix. Right: Inversion of the
-      preconditioner is trivial and decoupled by block.}}
-  \label{fig:block_jacobi_ex}
-\end{figure}
-The inversion of this preconditioner is trivial as shown on the right
-side of Figure~\ref{fig:block_jacobi_ex}. Each block can be inverted
-individually and combined to form the inverse. In addition, in the
-limit of a block size of one, the block Jacobi method reduces to the
-point Jacobi method. For high performance implementations this has
-several attractive properties. First, the blocks in the matrix come
-from the energy/angle discretization of the transport equation as
-given by Eq~(\ref{eq:A_block_matrix}). Each block on the diagonal is
-bound to a mesh element in the system (note there are 8 blocks on the
-diagonal in each of the sparsity patterns with a mesh of $2 \times 2
-\times 2$) and therefore we expect the matrix elements forming the
-block to be entirely local. Second, these blocks are typically dense
-and nearly lower triangular for many transport problems meaning that
-established dense matrix methods can be used for fast inversion.
-
-\subsection{Block Jacobi Spectral Analysis Results}
-\label{subsec:spn_analysis_results}
-Spectral radius computations for the block Jacobi preconditioned
-iteration matrix were performed for the same problems as the point
-Jacobi preconditioned case. Table~\ref{tab:group1bj} gives the results
-for the 1-group case, Table~\ref{tab:group10dsbj} for the 10-group
-case with full downscatter only and Table~\ref{tab:group10usbj} for
-the 10-group case with full downscatter and full upscatter. A block
-size of $N_g\times(N+1)/2$ was used for the preconditioner in all
-cases.
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{cccccc}\hline\hline
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{$SP_N$ Order}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{} \\
-       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
-       & \textbf{0} & 0.0635 & 0.1269 & 0.1444 & 0.1513 \\
-       & \textbf{1} & 0.0666 & 0.1315 & 0.1474 & 0.1534 \\
-      $P_N$ Order & \textbf{3} & 0.0666 & 0.1365 & 0.154 & 0.1592 \\
-       & \textbf{5} & 0.0666 & 0.1365 & 0.1562 & 0.163 \\
-       & \textbf{7} & 0.0666 & 0.1365 & 0.1562 & 0.164 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Spectral radius results for the block Jacobi
-      preconditioned iteration matrix with 1 energy group.}}
-  \label{tab:group1bj}
-\end{table}
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{cccccc}\hline\hline
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{$SP_N$ Order}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{} \\
-       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
-       & \textbf{0} & 0.0647 & 0.1275 & 0.1449 & 0.1514 \\
-       & \textbf{1} & 0.0686 & 0.1338 & 0.1484 & 0.1547 \\
-      $P_N$ Order & \textbf{3} & 0.0687 & 0.1399 & 0.1582 & 0.1625 \\
-       & \textbf{5} & 0.0692 & 0.1399 & 0.1582 & 0.1657 \\
-       & \textbf{7} & 0.0678 & 0.1393 & 0.1624 & 0.166 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Spectral radius results for the block Jacobi
-      preconditioned iteration matrix with 10 energy groups and full
-      downscatter.}}
-  \label{tab:group10dsbj}
-\end{table}
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{cccccc}\hline\hline
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{$SP_N$ Order}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{} \\
-       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
-       & \textbf{0} & 0.1887 & 0.2267 & 0.2285 & 0.2286 \\
-       & \textbf{1} & 0.4535 & 0.5044 & 0.5045 & 0.5045 \\
-      $P_N$ Order & \textbf{3} & 0.4535 & 0.6453 & 0.6506 & 0.6506 \\
-       & \textbf{5} & 0.4535 & 0.6453 & 0.6802 & 0.6818 \\
-       & \textbf{7} & 0.4535 & 0.6453 & 0.6802 & 0.6927 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Spectral radius results for the block Jacobi
-      preconditioned iteration matrix with 10 energy groups, full
-      downscatter and full upscatter.}}
-  \label{tab:group10usbj}
-\end{table}
-
-From the tabulated block Jacobi data it is clear that this is a viable
-preconditioning choice for all $SP_N$ problems in term of Monte Carlo
-solution methods. All cases were observed to have a spectral radius
-below unity and often significantly smaller than that, greatly
-improving convergence rates over the point Jacobi preconditioned
-problem. Based on these results, the block Jacobi method should be the
-first preconditioning strategy applied to more complicated problems
-representative of physical systems.
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{Fuel Assembly Criticality Calculations}
-\label{sec:fuel_assembly_calcs}
-Fuel assembly calculations are a critical piece of nuclear engineering
-infrastructure for reactor core analysis and design. At this level,
-individual fuel pins may be resolved at fine resolution in a variety
-of configurations. As a sophisticated problem of interest to push the
-limits of MCSA, a hot zero-power $17 \times 17$ pin assembly will be
-used with varying energy group structure and $SP_N$ discretization in
-a criticality calculation. A cross section along the vertical axis
-showing homogenized fuel pin materials and the associated grid is
-given by Figure~\ref{fig:problem3_radial_mat} while a cross section of
-the materials configuration along the horizontal axis is given by
-Figure~\ref{fig:problem3_axial_mat}. A detailed view of the assembly
-bottom is given in Figure~\ref{fig:problem3_end}. On the top and
-bottom of the assembly, vacuum conditions are used as well as on the
-top and right boundaries in
-Figure~\ref{fig:problem3_radial_mat}. Reflecting conditions are used
-on the left and bottom boundaries of
-Figure~\ref{fig:problem3_radial_mat}, effectively giving a
-representation of one quarter of the assembly. For the spatial
-discretization, each fuel pin (gray regions in
-Figure~\ref{fig:problem3_radial_mat}) is resolved by a $2 \times 2$
-mesh with materials and cross sections homgenized over this region.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4in]{chapters/spn_equations/problem3_radial_mat.png}
-  \end{center}
-  \caption{\textbf{Fuel assembly mesh and geometry cross section.}
-    \textit{Reflecting boundaries are used on the left and lower
-      boundaries to create a complete $17 \times 17$ assembly
-      geometry. Gray regions are homogenized fuel and red regions are
-      homogenized moderator. Each fuel pin is resolved by a $2 \times
-      2$ mesh.}}
-  \label{fig:problem3_radial_mat}
-\end{figure}
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6.0in]{chapters/spn_equations/problem3_axial_mat.png}
-  \end{center}
-  \caption{\textbf{Fuel assembly geometry cross section.} \textit{The
-      geometry is subdivided along the axial direction into 50 zones
-      spaced to capture material boundaries. Important details include
-      spacer grids along the length of the fuel pins and reactor core
-      structures on the top and bottom of the assembly. Lighter purple
-      material in the center of the assembly is moderator and darker
-      purple/red material is fuel.}}
-  \label{fig:problem3_axial_mat}
-\end{figure}
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6.0in]{chapters/spn_equations/problem3_end.png}
-  \end{center}
-  \caption{\textbf{Fuel assembly geometry end detail.}
-    \textit{Reactor core structure including spacer grids and plenum
-      has been included. Lighter purple material on the right of the
-      figure is moderator and darker purple/red material is fuel.}}
-  \label{fig:problem3_end}
-\end{figure}
-
-Significant geometric details are contained in the model including
-spacer grids, fuel pins with homogenized cladding and gas gap, core
-plenum, and moderator with boron. Group cross sections and other
-discrete nuclear data are generated as needed by a cross-section
-processing module dependent on the meshing parameters used to
-discretize the geometry and single-dimension pin-cell calculations for
-initial flux spectrum generation. Table~\ref{tab:problem3_parameters}
-gives the primary design parameters for the fuel assembly
-calculations.
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{ll}\hline\hline
-      \multicolumn{1}{l}{\textbf{Parameter}} & 
-      \multicolumn{1}{l}{\textbf{Value}} \\
-      Power Level & 0 MW \\
-      Inlet Temperature & 326.85C \\
-      Fuel Temperature & 600C \\
-      Boron Concentration & 1300 ppm \\
-      Moderator Density & 0.743 g/cc \\
-      Helium Density & \sn{1.79}{-4} g/cc \\
-      Zirconium Density & 6.56 g/cc \\
-      Stainless Steel Density & 8.0 g/cc \\
-      Inconel Density & 8.19 g/cc \\
-      UO2 Density & 10.257 g/cc \\
-      Fuel Pin Radius (w/o clad) & 0.4096 cm \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Design parameters for the $17 \times 17$ pin fuel
-      assembly criticality calculation.}}
-  \label{tab:problem3_parameters}
-\end{table}
-
-To generate the multiplication factor and steady-state flux
-distribution for this problem, at every eigenvalue iteration MCSA is
-used to solve the resulting $SP_N$ problem using the provided fission
-source. Algorithm~\ref{alg:power_iteration} presents the use of MCSA
-within a power iteration strategy to find the multiplication factor.
-\begin{algorithm}[h!]
-  \caption{Power Iteration MCSA Scheme}
-  \label{alg:power_iteration}
-  \begin{algorithmic}
-    \State $k_0 =$ initial guess
-    \State $\mathbf{\Phi}_0 =$ initial guess
-    \State $n = 0$
-    \While{$|\frac{k^n - k^{n-1}}{k^n}| < \epsilon$}
-    \Comment{Iterate until convergence of the eigenvalue}
-    \State $\mathbf{M} \mathbf{\Phi}^{n+1} = \frac{1}{k^n} \mathbf{F} \mathbf{\Phi}^n$
-    \Comment{Solve for the new flux state with MCSA}
-    \State $k^{n+1} = k^n \frac{\int \mathbf{F} \mathbf{\Phi}^{n+1} d\mathbf{r}}{\int
-      \mathbf{F} \mathbf{\Phi}^n d\mathbf{r}}$
-    \Comment{Update the multiplication factor}
-    \State $n = n+1$
-    \EndWhile
-  \end{algorithmic}
-\end{algorithm}
-Here, $\mathbf{M}$ is the transport operator generated on the
-left-hand side of the $SP_N$ discretization, $\mathbf{F}$ is the
-fission matrix, and $\mathbf{\Phi}$ the multigroup neutron flux. This
-problem is significantly more complicated than the simple test problem
-used for the previous spectral analysis. Fission has been introduced
-into the set of equations and the addition of moderator into the
-system will increase the amount of scattering, creating a
-significantly more difficult problem manifesting itself in an
-iteration matrix with a larger spectral radius. When using MCSA, the
-linear operator applied to $\mathbf{\Phi}^{n+1}$ at each eigenvalue
-iteration will dictate convergence and remain unchanged throughout the
-computation\footnote{The operator will change if, for example, physics
-  feedback through temperature or potentially burnup is
-  considered. These additional physics will modify the cross sections
-  used to assemble $\mathbf{M}$. The calculations presented here will
-  consist of a single eigenvalue calculation with a static
-  $\mathbf{M}$.} while the addition of fission to the system will only
-modify the source of neutrons and the multiplication factor while not
-affecting Monte Carlo transport.
-
-\clearpage
-
-\subsection{Preliminary Jacobi Preconditioned Calculations}
-\label{subsec:jacobi_prec_assembly_calc}
-Based on the success of block Jacobi preconditioning with the test
-problem used for the spectral radius parameter study, we use it first
-to solve the $17 \times 17$ fuel assembly problem. A single energy
-group problem was first solved with $SP_1$ discretization, effectively
-giving the one-speed neutron diffusion system for the fuel assembly
-resulting in 20,088 degrees of freedom in the
-problem. Figure~\ref{fig:block_jacobi_res_mcsa} gives the residual
-infinity norm as a function of iteration for the MCSA linear solve in
-the first eigenvalue iteration using 25,000 stochastic histories at
-every iteration for the adjoint Neumann-Ulam solve.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/block_jacobi_res.pdf}
-  \end{center}
-  \caption{\textbf{Residual infinity norm vs. iteration for the block
-      Jacobi preconditioned MCSA solve during the first eigenvalue
-      iteration of the 1-group $17 \times 17$ fuel assembly problem.}
-    \textit{Convergence was not achieved with the block Jacobi
-      preconditioned method.}}
-  \label{fig:block_jacobi_res_mcsa}
-\end{figure}
-Convergence was not achieved as noted by the rapid rise in the
-residual over a few iterations. Based on the spectral radius
-computations performed, these results are not in line with
-expectations for this problem. Additional computations performed with
-\sn{1}{6} histories per iteration exhibited the same divergent
-behavior at a significant computational cost and compute time. Even if
-the problem may be ill-conditioned, we do expect convergence of
-MCSA. To investigate further, a block Jacobi preconditioned Richardson
-iteration was used to solve the same
-problem. Figure~\ref{fig:block_jacobi_res_richardson} gives the
-residual infinity norm as a function of iteration for the Richardson
-linear solve in the first eigenvalue iteration.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/block_jacobi_rich_res.pdf}
-  \end{center}
-  \caption{\textbf{Residual infinity norm vs. iteration for the block
-      Jacobi preconditioned Richardson solve during the first
-      eigenvalue iteration of the 1-group $17 \times 17$ fuel assembly
-      problem.} \textit{A spectral radius near 1 was observed for the
-      iteration matrix.}}
-  \label{fig:block_jacobi_res_richardson}
-\end{figure}
-Poor converge is observed for the Richardson iteration, however,
-convergence is achieved meaning that the preliminary criteria needed
-to satisfy MCSA convergence has also been met. The number of
-iterations required for the Richardson iteration to converge will give
-an approximation for the spectral radius via
-Eq~(\ref{eq:linear_k_iter_norm3}). With 7291 iterations required to
-converge to a tolerance of $\sn{1}{-6}$, $\rho(\mathbf{H}) \approx
-0.998$, nearing the limits of MCSA applicability and well beyond the
-spectral radii generated in the initial spectral analysis.
-
-Based on these results, it then appears that even if the simple
-criteria of a spectral radius of less than one is met and the
-Richardson iteration will converge with the same preconditioning, MCSA
-still may not converge. We then expect the issue to reside with the
-Neumann-Ulam solve providing the correction as the Richardson
-iteration is known to provide the correct result. Furthermore, the
-fact that the spectral radius is less than one means that the
-stochastic histories in the block Jacobi preconditioned Neumann-Ulam
-method are eventually being terminated by the weight cutoff as no
-artificial absorption was used for these preliminary
-calculations. Based on this, the correction being generated by the
-Neumann-Ulam solve is the component of MCSA causing a divergent
-solution.
-
-\subsection{MCSA Breakdown}
-\label{subsec:mcsa_break_down}
-For the fuel assembly problem, the initial block Jacobi
-preconditioned\footnote{For 1-group $SP_1$ problems the block size is
-  1, giving a point Jacobi preconditioner.}  calculations used a
-one-speed $SP_1$ discretization, effectively giving a diffusion
-system. To study the breakdown of MCSA at iteration matrix spectral
-radii near one, we will use a simpler homogeneous 2-dimensional
-one-speed neutron diffusion system to isolate this behavior. In this
-system, we can vary the cross sections while maintaining a fixed grid
-in order to achieve varying spectral radii. For these studies, we
-neglect fission as MCSA behavior is dictated by the transport operator
-$\mathbf{M}$ in an eigenvalue scheme with the fission matrix used to
-generate a fixed source.
-
-\subsubsection{Model Diffusion Problem}
-\label{subsec:model_problem}
-For our breakdown study, we choose the one-speed, two-dimensional
-neutron diffusion equation as a model problem
-\citep{duderstadt_nuclear_1976}:
+\chapter{Two-Dimensional One-Speed Neutron Diffusion Model Problem}
+\label{chap:diffusion_problem}
+For many numerical experiments in this work, we choose the one-speed,
+two-dimensional neutron diffusion equation as a model transport
+problem \citep{duderstadt_nuclear_1976}:
 \begin{equation}
@@
 zero.
-
-\subsubsection{MCSA Breakdown Analysis}
-\label{subsubsec:breakdown_analysis}
-For each solver and estimator combination, the spectral radius of the
-iteration matrix generated by the diffusion problem was varied by
-changing the absorption cross section from 0.25 to 100 while fixing
-the grid size at $100 \times 100$ with $h = 0.1$ and a fixed
-scattering cross section of unity. For each parameter variation, a
-minimum of one stochastic history per degree of freedom (DOF) in the
-problem was used to compute the Monte Carlo correction. If the solver
-could not converge in less than 100 iterations, the number of
-histories was increased by increments of 5,000 until convergence was
-achieved in less than 100 iterations. The number of iterations
-required to converge MCSA and the time to converge was recorded as a
-means to capture the breakdown.
-
-Figure~\ref{fig:breakdown_iterations} gives the number of iterations
-required to converge for the chosen number of histories per iteration
-given by Figure~\ref{fig:breakdown_histories} using the adjoint solver
-with the collision and expected value estimators and the forward
-solver with the collision estimator. For spectral radii less than
-0.97, all MCSA problems converged with 1 history per DOF (10,000 for
-this problem) with the number of iterations required to converge
-increasing as a function of spectral radius. Near a spectral radius of
-0.97, the number of histories required to converge MCSA in less than
-100 iterations takes a dramatic rise that exhibits neither exponential
-nor power law behavior. As the spectral radius approaches 1, the
-number of histories required becomes significant and effectively
-impractical to compute. Even with this simple diffusion problem, the
-behavior is consistent with that observed for the fuel assembly
-problem with $SP_1$ discretization. In that case, we estimated a
-spectral radius of $\approx 0.998$, larger than any of the spectral
-radii that could be computed within even 90 minutes of compute time
-for this simple two dimensional problem. For that problem, even
-1,000,000 histories ($\approx 50$ per DOF) was not enough to provide
-convergence. Even if single solve times of an hour can be tolerated,
-dozens of solves are typically required to compute the multiplication
-factor and flux spectrum within the $SP_N$ eigenvalue scheme for more
-difficult problems like the fuel assembly, making the method unusable.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/breakdown_iterations.pdf}
-  \end{center}
-  \caption{\textbf{Iterations required to converge as a function of
-      spectral radius for the neutron diffusion problem.} \textit{The
-      number of histories was increased to achieve convergence in less
-      than 100 iterations. At least 10,000 histories were used for
-      each calculation.}}
-  \label{fig:breakdown_iterations}
-\end{figure}
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/breakdown_histories.pdf}
-  \end{center}
-  \caption{\textbf{Histories per DOF required to converge as a
-      function of spectral radius for the neutron diffusion problem.}
-    \textit{The number of histories was increased to achieve
-      convergence in less than 100 iterations. At least 10,000
-      histories were used for each calculation with 10,000 DOFs in the
-      problem.}}
-  \label{fig:breakdown_histories}
-\end{figure}
-
-In addition to the significantly larger number of histories required
-for to achieve convergence for ill-conditioned problems another
-penalty is paid due to histories that take longer to
-compute. Figure~\ref{fig:breakdown_time} gives the CPU time in seconds
-required to compute a single random walk averaged over the entire set
-of histories run in the calculation over all iterations. As the
-spectral radius increases (correlating to a higher ratio of scattering
-in the system) the random walk lengths increase, using more CPU time
-to finish the computation. Compared to spectral radii of 0.5, larger
-spectral radii over 0.97 have histories that require two orders of
-magnitude more computation time. This significant increase in
-computation time per history coupled with the significant increase in
-the number of histories required to converge is evidence that for
-problems with spectral radii above $\approx 0.97$, using MCSA to solve
-any problems of interest is entirely ineffective and not practical. We
-therefore require a more expansive set of preconditioning techniques
-to move the eigenvalue spectrum of the $SP_N$ problem into a regime in
-which MCSA is more applicable and in which performance is improved.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/breakdown_time.pdf}
-  \end{center}
-  \caption{\textbf{CPU time per history as a function of spectral
-      radius for the neutron diffusion problem.} \textit{As the
-      spectral radius grows, so do the length of the random
-      walks. Longer random walks require more CPU time to compute.}}
-  \label{fig:breakdown_time}
-\end{figure}
-
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{Advanced Preconditioning Strategies}
-\label{subsec:spn_advanced_preconditioning}
-For the fuel assembly criticality problem presented in the previous
-section, the spectral radius was near one using Jacobi preconditioning
-and in a regime in which MCSA breakdown is observed. In this regime
-the number of stochastic histories required to converge MCSA increases
-rapidly and the resulting poor performance is compounded by those
-histories being increasingly expensive to compute. To overcome this
-difficulty, advanced preconditioning strategies for the $SP_N$ problem
-are required beyond simple Jacobi methods that can reduce the spectral
-radius into a region of better MCSA performance. Several modern
-algebraic preconditioning strategies will be presented here and used
-within the explicit preconditioning framework given by
-Eq~(\ref{eq:left_right_mcsa}). Data showing their effects on MCSA
-solutions of the fuel assembly criticality problem will be presented.
-
-\subsection{ILUT Preconditioning}
-\label{subsec:spn_ilut_preconditioning}
-Incomplete lower-upper (ILU) factorizations of the linear operator can
-be used a simple mechanism to form an approximate inverse of a
-preconditioner. To build the factorization, the sparse upper and lower
-triangular factors, $\mathbf{L}$ and $\mathbf{U}$, are computed such
-that the residual matrix formed by the factorization
-\citep{saad_iterative_2003}:
-\begin{equation}
-  \mathbf{R} = \mathbf{L} \mathbf{U} - \mathbf{A} \:,
-  \label{eq:ilu_residual_matrix}
-\end{equation}
-has a specified sparsity pattern and element magnitude. When a
-magnitude threshold is used, elements generated in the factors below
-that magnitude are dropped, resulting in ILU threshold (ILUT)
-preconditioning. The sparsity pattern in this case is determined from
-the input matrix to be preconditioned and the number of elements
-maintained in the factor is specified by a fill level parameter. A
-fill level of 1 will generate the same number of elements as the
-sparsity pattern of the input matrix while a fill level of 2 will
-contain twice as many elements in the factorization, resulting in a
-better representation of the true LU factorization. The inverse of the
-lower and upper triangular factors may then be easily inverted by
-means of simple elimination to produce the preconditioner. For MCSA,
-$\mathbf{L}^{-1}$ will be used on the left and $\mathbf{U}^{-1}$ on
-the right to precondition the system.
-
-For modern subspace methods, only the action to the preconditioner on
-a vector is required for efficient implementations and therefore a
-triangular solve can be used for this purpose. For the explicit
-preconditioning scheme presented in Eq~(\ref{eq:left_right_mcsa}), the
-fully inverted operator must be generated in order to build the set of
-probabilities and weights required for Monte Carlo
-sampling. Therefore, for a linear system of size $N$, $N$ triangular
-solves will be required in order to extract the inverse matrices from
-production ILU implementations. In addition, parallel implementations
-typically generate lower and upper factors that are only triangular
-locally, providing an easily parallelizable mechanism to generate the
-action of the preconditioner inverse. A consequence of this choice for
-parallel scalability is a degradation of the preconditioning quality
-as the size of the parallel system is increased. For serial
-computations, the triangular factorization is potentially exact
-depending on the parameters chosen while at thousands of parallel
-tasks, the global triangular factors differ significantly from the true
-factorization. As a result, more iterations are required at higher
-levels of parallelism to converge the system and will ultimately
-degrade overall scalability of the system with respect to total wall
-time to converge.
-
-MCSA preconditioned with ILUT was used to solve the fuel assembly
-problem presented in the previous section for a 1-group $SP_1$
-discretization. Unlike the Jacobi preconditioned strategy, convergence
-was achieved with ILUT preconditioning. To study convergence
-sensitivity to ILUT parameters, the fill level and drop tolerance were
-parametrically varied with the number of iterations to converge an
-eigenvalue iteration for the fuel assembly problem recorded along with
-the maximum number of non-zero entries observed in all matrix rows for
-the left/right preconditioned composite linear operator. To provide
-some sparsity to the factorization, the ILUT drop tolerance was used
-to drop elements in the extracted inverse triangular factors. For each
-calculation, the number of iterations required to converge reported
-was for a single eigenvalue iteration with \sn{3}{4} histories at each
-MCSA iteration to compute the Monte Carlo correction using the adjoint
-collision estimator. All ILUT calculations reported here were
-performed on a single CPU and therefore this data does not take into
-account the aforementioned effects of parallel decomposition on the
-quality of the ILUT preconditioning.
-
-Figure~\ref{fig:ilut_iterations} gives the number of iterations to
-converge the fixed source problem to a tolerance of \sn{1}{8}. As
-expected, the higher the fill level chosen for the ILUT factorization,
-the fewer iterations are required to converge the problem (only 9 MCSA
-iterations were required to converge the problem for the fill level of
-3 and drop tolerance of \sn{1}{-5} case). For higher levels of fill,
-the iterations needed to converge were not as sensitive, signaling
-that a larger drop tolerance can perhaps be used without a significant
-degradation in iterative performance. At smaller fill levels, the
-sensitivity to the ILUT drop tolerance is more significant. For all
-fill levels used, convergence was not achieved for the fuel assembly
-problem with a drop tolerance larger than \sn{1}{-3}.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/ilut_iterations.pdf}
-  \end{center}
-  \caption{\textbf{Number of MCSA iterations required to converge an
-      eigenvalue iteration for the fuel assembly problem with ILUT
-      preconditioning as a function of ILUT drop tolerance.}
-    \textit{Each colored curve represents the iteration behavior for a
-      different ILUT fill level. Fill levels of 1.5, 2.0, 2.5, and 3.0
-      were used.}}
-  \label{fig:ilut_iterations}
-\end{figure}
-
-Unfortunately, gaining convergence (and excellent iterative
-performance) with MCSA for the $SP_N$ fuel assembly problem comes at
-an immediate cost. Figure~\ref{fig:ilut_size} gives the maximum number
-of non-zero entries in the composite linear operator generated by
-preconditioning as a function ILUT drop tolerance for varying values
-of fill level. For the 1-group $SP_1$ discretization, the original
-linear operator will contain only a maximum of 7 non-zero entries per
-row in the system. As observed in Figure~\ref{fig:ilut_size},
-performing the explicit preconditioning yields composite linear
-operators with $O(1,000)$ elements in a row for all combinations of
-fill level and drop tolerance, over 10\% of the total row size for
-this particular problem. This large number of row entries, observed
-for a significant fraction of rows in the system, creates several
-problems. First, sparsity is completely destroyed with each state in
-the system now coupled to over 10\% of the total states in the system
-through the composite iteration matrix. This means that Monte Carlo
-sampling tables will be large, requiring significant memory to store
-them and a substantial overhead in the sampling procedure during
-transport. In addition, the matrix-matrix multiply operations required
-to build the composite operator see significant performance losses due
-to the amount of data that must be handled. In parallel, losing
-sparsity severely inhibits performance where now parallel operations
-require communications amongst orders of magnitude more processors for
-nearest-neighbor type algorithms.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/ilut_size.pdf}
-  \end{center}
-  \caption{\textbf{Maximum number of non-zero entries observed for all
-      rows in the composite linear operator for the fuel assembly
-      problem with ILUT preconditioning given as a function of ILUT
-      drop tolerance.} \textit{Each colored curve represents the row
-      size for a different ILUT fill level. Fill levels of 1.5, 2.0,
-      2.5, and 3.0 were used.}}
-  \label{fig:ilut_size}
-\end{figure}
-
-As a means to assess the quality of the preconditioning, a simple
-metric is developed to address these concerns and allow comparison to
-future developments. For our studies, our core performance metric will
-be iterative performance with the minimum number of MCSA iterations
-required to converge the problem desired. For Monte Carlo
-calculations, this improved performance is balanced by the creation of
-a dense composite system and we seek to reduce the number of non-zero
-entries to a minimum value in order to lower the amount of coupling
-amongst states in the system and reduce the amount of memory used
-along with the potential latency overhead. Given these two objectives,
-the following metric is proposed:
-\begin{equation}
-  \text{Quality Metric} = (\text{\# iterations}) \times (\text{maximum
-    \# of non-zero values})\:,
-\end{equation}
-where the highest quality preconditioning is one that minimizes this
-metric. For the ILUT preconditioned data provided,
-Figure~\ref{fig:ilut_quality} provides the computed metric as a
-function of ILUT drop tolerance for each fill level used. In general,
-the metric is similar to the data observed for the number of
-iterations required to converge as the non-zero row entries changed by
-a smaller fraction over drop tolerances tested.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/ilut_quality.pdf}
-  \end{center}
-  \caption{\textbf{ILUT preconditioning quality metric for the fuel
-      assembly problem given as a function of ILUT drop tolerance.}
-    \textit{Each colored curve represents the quality metric behavior
-      for a different ILUT fill level. Fill levels of 1.5, 2.0, 2.5,
-      and 3.0 were used.}}
-  \label{fig:ilut_quality}
-\end{figure}
-
-It should be noted here that although only the behavior of the linear
-solve during a single eigenvalue iteration is reported, the behavior
-of the linear solver was observed to be consistent throughout the
-eigenvalue iterations (25 total eigenvalue iterations were required to
-converge the 1-group $SP_1$ problem). We expect this as the linear
-operator (which is unchanging in the fuel assembly criticality
-problem) dictates the convergence of MCSA. At each iteration the
-fission source provided to the linear problem is changing, however, we
-expect the same convergence behavior for all source vectors
-$\mathbf{b} \in \mathbb{R}^{N}$ where $||\mathbf{b}||_1 \neq 0$.
-
-\subsection{Sparse Approximate Inverse Preconditioning}
-\label{subsec:spn_spainv_preconditioning}
-Explicit formation of the preconditioner inverse is a requirement of
-the current MCSA preconditioning strategy. As a result, although ILUT
-preconditioning provided excellent iterative performance, a
-significant time penalty is paid to extract the inverse of the
-triangular factors through $N$ triangular solves. In addition, this
-explicit inversion yielded a composite linear operator that was dense,
-requiring modification of the system in order to achieve better
-scalability and Monte Carlo performance. Sparse approximate inverse
-(SPAINV) preconditioning is a technique that may potentially alleviate
-both of these constraints at the cost of reduced iterative
-performance. The method produces a sparse approximation of the inverse
-matrix directly by minimizing the Frobenius norm of the residual
-matrix \citep{saad_iterative_2003}:
-\begin{equation}
-  || \mathbf{I} - \mathbf{A} \mathbf{M} ||^2_F =
-  \sum_{j=1}^N ||\mathbf{e}_j - \mathbf{A} \mathbf{m}_j||^2_2 \:,
-  \label{eq:frobenius_norm_min}
-\end{equation}
-where the inverse preconditioning matrix $\mathbf{M}$ minimizes the
-norm on the right and $\mathbf{e}_j$ and $\mathbf{m}_j$ are the
-$j^{th}$ columns of the identity and preconditioning matrices
-respectively. Here, $\mathbf{M}$ is the actual inverse of the
-preconditioner and is formed directly by the algorithm. On the right
-hand side of Eq~(\ref{eq:frobenius_norm_min}), the Frobenius norm is
-represented as a effective sum of $N$ linear system residual norms. A
-few iterations of a subspace method or other iterative scheme can be
-used to effectively solve these systems to a loose tolerance and yield
-the columns of the approximate preconditioning matrix. At each
-iteration, threshold values are used to eliminate the components of
-each $\mathbf{m}_j$ column to maintain the desired level of
-sparsity. In addition, a sparsity pattern can be predetermined and
-enforced during this process. For the SPAINV implementation used for
-this work, the sparsity pattern was defined by levels such that a
-level of $N$ for an input operator $\mathbf{A}$ will generate a
-preconditioning matrix $\mathbf{M}$ with the same sparsity pattern as
-$\mathbf{A}^{N+1}$. By using the norm minimization strategy instead of
-a factorization such as ILUT, parallel results are reproduced
-regardless of parallel problem size, meaning that a serial computation
-will converge in the same number of iterations as a computation with
-thousands of cores.
-
-MCSA preconditioned with SPAINV was used to solve the fuel assembly
-problem as with ILUT for a 1-group $SP_1$ discretization. To study
-convergence sensitivity to SPAINV parameters, the number of levels in
-the sparsity pattern and threshold were parametrically varied with the
-number of iterations to converge a single eigenvalue iteration for the
-fuel assembly problem recorded along with the maximum number of
-non-zero entries observed in all matrix rows for the left/right
-preconditioned composite linear operator. For each calculation, the
-number of iterations required to converge reported was for a single
-eigenvalue iteration with \sn{3}{4} histories at each MCSA iteration
-to compute the Monte Carlo correction using the adjoint collision
-estimator. All SPAINV calculations reported here were performed on a
-single CPU.
-
-Figure~\ref{fig:spainv_iterations} gives the number of MCSA iterations
-needed to converge the fuel assembly problem and
-Figure~\ref{fig:spainv_size} the maximum number of non-zero entries
-per row obeserved in the composite linear operator as a function the
-SPAINV threshold. For this analysis, the number of levels in the
-sparsity pattern was varied from 3 to 7 with convergence of the fuel
-assembly problem not achieved for smaller level values. Compared to
-ILUT preconditioning, at higher levels we see comparable iterative
-performance with a relative invariance to the threshold value. The
-threshold did have a significant effect, however, on the time required
-to generate the preconditioner with a threshold of 0.001 over an order
-of magnitude slower than a threshold of 0.01 due to the inclusion of a
-significant number of extra values in the input operator. In addition
-to comparable iterative performance, the sparsity of the composite
-operator is greatly improved with nearly an order of magnitude
-reduction in number of non-zero entries per row for lower level
-values.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/spainv_iterations.pdf}
-  \end{center}
-  \caption{\textbf{Number of MCSA iterations required to converge a
-      single eigenvalue iteration for the fuel assembly problem with
-      SPAINV preconditioning as a function of SPAINV threshold.}
-    \textit{Each colored curve represents the iteration behavior for a
-      different SPAINV level pattern. Levels of 3, 4, 5, 6, and 7 were
-      used.}}
-  \label{fig:spainv_iterations}
-\end{figure}
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/spainv_size.pdf}
-  \end{center}
-  \caption{\textbf{Maximum number of non-zero entries observed for all
-      rows in the composite linear operator for the fuel assembly
-      problem with SPAINV preconditioning given as a function of
-      SPAINV threshold.} \textit{Each colored curve represents the row
-      size for a different SPAINV level pattern. Levels of 3, 4, 5, 6,
-      and 7 were used.}}
-  \label{fig:spainv_size}
-\end{figure}
-
-With the comparable iterative performance and improved sparsity,
-SPAINV preconditioning should then have a more favorable quality
-metric than ILUT preconditioning. Figure~\ref{fig:spainv_quality}
-gives the quality metric as a function of the threshold value for each
-of the sparsity levels computed. Not only do we see an improved
-quality metric over ILUT preconditioning (about an order of magnitude
-less), but we also note that the ideal sparsity level is not the largest
-nor the smallest. In fact, the largest and smallest levels (3 and 7)
-performed the worst in terms of the quality metric with a level of 4
-performing the best. When CPU time to generate the preconditioner is
-considered, 4 is the clear choice for this problem as more time is
-required to do the approximate inversion as the sparsity pattern of
-the preconditioner grows in size.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/spainv_quality.pdf}
-  \end{center}
-  \caption{\textbf{SPAINV preconditioning quality metric for the fuel
-      assembly problem given as a function of SPAINV threshold.}
-    \textit{Each colored curve represents the quality metric behavior
-      for a different SPAINV level pattern. Levels of 3, 4, 5, 6, and
-      7 were used.}}
-  \label{fig:spainv_quality}
-\end{figure}
-
-\clearpage
-
-\subsection{SPAINV Improved ILUT Preconditioning}
-\label{subsec:spn_spainv_preconditioning}
-As a means to improve convergence over using only SPAINV
-preconditioning, ILUT preconditioning may be modified through an
-additional application of SPAINV
-(SPAINV-ILUT)\citep{saad_iterative_2003}. For the ILUT preconditioning
-case we have the following linear problem to solve for the substituted
-variable:
-\begin{equation}
-  \mathbf{L}^{-1} \mathbf{A} \mathbf{U}^{-1} \mathbf{u} =
-  \mathbf{L}^{-1} \mathbf{b}\:.
-  \label{eq:ilut_standalone}
-\end{equation}
-We then apply SPAINV and minimize the following problem, this time on
-the left:
-\begin{equation}
-  || \mathbf{I} - \mathbf{M} \mathbf{L}^{-1} \mathbf{A}
-  \mathbf{U}^{-1} ||^2_F\:,
-  \label{eq:ilut_minimize}
-\end{equation}
-giving a new preconditioned system:
-\begin{equation}
-  \mathbf{M} \mathbf{L}^{-1} \mathbf{A} \mathbf{U}^{-1} \mathbf{u} =
-  \mathbf{M} \mathbf{L}^{-1} \mathbf{b}\:.
-  \label{eq:ilut_spainv}
-\end{equation}
-Doing this permits a smaller fill level to be used with ILUT and a
-lower-order sparsity pattern to be used with SPAINV, giving marginally
-more sparsity in the resulting composite linear operator while
-maintaining good iterative performance.
-
-MCSA preconditioned with SPAINV-ILUT was also used to solve the same
-fuel assembly problem as with the other preconditioners for the same
-1-group $SP_1$ discretization. To study convergence sensitivity to
-SPAINV-ILUT parameters, the ILUT drop tolerance and ILUT fill level
-were parametrically varied with the number of iterations to converge a
-single eigenvalue iteration for the fuel assembly problem recorded
-along with the maximum number of non-zero entries observed in all
-matrix rows for the left/right preconditioned composite linear
-operator. SPAINV parameters for the data presented here were fixed as
-it was found that the preconditioning quality was largely insenstive
-to their variation. In addition, the fact that SPAINV is being used to
-preconditioned a problem already preconditioned with ILUT means that
-the input matrix sparsity pattern is already somewhat dense and large
-SPAINV levels would result in an even denser system. Therefore, a
-smaller SPAINV sparsity pattern level will be used for this
-preconditioning. For each calculation, the number of iterations
-required to converge reported was for a single eigenvalue iteration.
-Again, \sn{3}{4} histories are used at each MCSA iteration to compute
-the Monte Carlo correction using the adjoint collision estimator. All
-SPAINV-ILUT calculations reported here were performed on a single CPU.
-
-Figure~\ref{fig:spainv_ilut_iterations} gives the number of iterations
-required to converge as a function of ILUT drop tolerance for ILUT
-fill levels of 1.5, 2.0, 3.0, 4.0 and 5.0 with a fixed SPAINV
-threshold of 1.0 and sparsity level of 1. SPAINV threshold values
-higher than 1.0 resulted in a loss of convergence and lower values did
-not significantly affect performance. Larger sparsity levels resulted
-in a composite operator that was too dense and a prohibitive amount of
-CPU time required to compute the preconditioner. At smaller levels of
-fill, the benefits of using an additional step of SPAINV
-preconditioning are readily observed with the number of iterations
-required to converge cut in half for large drop tolerances. For
-smaller drop tolerances and higher ILUT levels of fill, the additional
-step of SPAINV preconditioning offers marginal improvement to
-iterative performance. As shown in Figure~\ref{fig:spainv_ilut_size},
-because ILUT is used in the preconditioning sequence, sparsity is
-again lost with only marginally improved non-zero entry numbers due to
-the application of the SPAINV threshold. As a result, the quality
-metrics given by Figure~\ref{fig:spainv_ilut_quality} are only
-marginally better than those observed for ILUT preconditioning and
-much larger than the SPAINV quality metric values.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/psilut_iterations.pdf}
-  \end{center}
-  \caption{\textbf{Number of MCSA iterations required to converge a
-      single eigenvalue iteration for the fuel assembly problem with
-      SPAINV-ILUT preconditioning as a function of ILUT drop
-      tolerance.} \textit{Each colored curve represents the row size
-      for a different ILUT fill level. Fill levels of 1.5, 2.0, 3.0,
-      4.0, and 5.0 were used.}}
-  \label{fig:spainv_ilut_iterations}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/psilut_size.pdf}
-  \end{center}
-  \caption{\textbf{Maximum number of non-zero entries observed for all
-      rows in the composite linear operator for the fuel assembly
-      problem with SPAINV-ILUT preconditioning given as a function of
-      ILUT drop tolerance.} \textit{Each colored curve represents the
-      row size for a different ILUT fill level. Fill levels of 1.5,
-      2.0, 3.0, 4.0 and 5.0 were used.}}
- \label{fig:spainv_ilut_size}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/psilut_quality.pdf}
-  \end{center}
-  \caption{\textbf{SPAINV-ILUT preconditioning quality metric for the
-      fuel assembly problem given as a function of ILUT drop
-      tolerance.} \textit{Each colored curve represents the row size
-      for a different ILUT fill level. Fill levels of 1.5, 2.0, 3.0,
-      4.0 and 5.0 were used.}}
-  \label{fig:spainv_ilut_quality}
-\end{figure}
-
-\clearpage
-
-\subsection{Applying the Reduced Domain Approximation}
-\label{subsec:spn_prec_rda}
-For each preconditioning technique presented convergence was achieved
-for the single fuel assembly problem. However, a primary concern is
-the number of non-zero states in each row of the system generated by
-the explicit preconditioning strategy. In many cases, orders of
-magnitude more matrix elements were generated resulting in poor
-scalability for domain decomposed algorithms and overall performance
-issues for Monte Carlo. As outlined in
-\S~\ref{subsec:reduced_domain_approximation}, the reduced domain
-approximation may be used as a mechanism to potentially alleviate this
-problem by filtering elements of the composite matrix in each row that
-fall below a certain threshold value or by maintaining the largest $N$
-elements in each row where $N$ is a designated fill level.
-
-For the ILUT, SPAINV, and SPAINV-ILUT preconditioning strategies the
-reduced domain approximation will be applied to reduce the density of
-the composite linear operator to more manageable levels. For each
-preconditioner, the parameters that achieved the best quality metric
-results from the previous analysis were used. These correspond to ILUT
-parameters of a fill level of 5.0 and a drop tolerance of \sn{1}{-5},
-SPAINV parameters of a sparsity level of 4 and a threshold of 0.1 and
-SPAINV-ILUT with ILUT parameters of a fill level of 4.0 and a drop
-tolerance of \sn{1}{-5} and SPAINV parameters of a sparsity level of 1
-and a threshold of 1.0. The reduced domain threshold was set to
-\sn{1}{-10} in order to eliminate any exceedingly small values from
-the matrix (generally this is simply removing non-zero elements within
-the floating point tolerance of zero). The reduced domain fill level
-was then varied, starting with the largest non-zero entries per row
-value observed for each of the preconditioning types in order to
-assess its effects relative to the case where no reduced domain
-approximation was applied.
-
-Figure~\ref{fig:rda_iterations} gives the number of iterations
-required to converge for each preconditioner type as a function of
-reduced domain fill level. Figure~\ref{fig:rda_quality} gives the
-corresponding quality metric for each data point where the number of
-non-zero entries used to compute the metric is equivalent to the
-reduced domain fill level. We first note the SPAINV preconditioning
-alone is signficantly more sensitive to the reduction in domain size
-over the ILUT-based methods, although the preconditioner was of
-$O(100)$ non-zero entries per row without any approximation
-applied. For the ILUT-based methods, performance was significantly
-better with convergence achieved in less than 40 MCSA iterations with
-only 10 non-zero entries in each row (vs. 7 non-zero entries for the
-case with no preconditioning). SPAINV-ILUT iterative performance was
-marginally better than ILUT alone for all reduced domain fill
-levels. However, it should be noted that the ILUT level of fill used
-for the ILUT only calculations was set to 4 for this case while the
-SPAINV-ILUT preconditioner used an ILUT level of fill of 5 and
-therefore the marginally better performance is more likely a result of
-this addition of fill rather than the extra SPAINV
-preconditioning. Looking at the quality metric data, we see a nice
-power-law reduction in the quality metric as a function of the reduced
-domain fill level, achieving 2 orders of magnitude reduction in the
-quality metric for the ILUT-based preconditioning methods.
-
-Although applying the reduced domain approximation results in a
-successful recovery of sparsity for the Monte Carlo problem while
-maintaining good convergence properties, there is still an issue of
-forming the composite operator before applying the approximation and
-potentially generating the transpose of this operator in the case of
-the adjoint Monte Carlo method. Because of this, memory and scaling
-issues may still be observed when building the probability and weight
-matrices for the Monte Carlo problem. In addition, the expensive
-extraction of the inverse of the preconditioning operators for the
-explicit scheme creates a significant cost in overall
-performance. Future work in this area should be considerate of these
-important components of the preconditioned MCSA algorithm.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/rda_iterations.pdf}
-  \end{center}
-  \caption{\textbf{Number of MCSA iterations required to converge a
-      single eigenvalue iteration for the fuel assembly problem with
-      each preconditioning as a function of reduced domain
-      approximation fill level.} \textit{The largest fill level for
-      each preconditioning presented is that using the parameters that
-      gave the best results without the approximation.}}
-  \label{fig:rda_iterations}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/rda_quality.pdf}
-  \end{center}
-  \caption{\textbf{Preconditioning quality metric for the fuel
-      assembly problem given as a function of reduced domain
-      approximation fill level.} \textit{The largest fill level for
-      each preconditioning presented is that using the parameters that
-      gave the best results without the approximation.}}
-  \label{fig:rda_quality}
-\end{figure}
-
-\clearpage
-
-\subsection{MCSA Relaxation Parameters}
-\label{subsec:spn_mcsa_relaxation}
-The Richardson iteration upon which the Neumann-Ulam method is built
-may be implemented with a scalar relaxation parameter that can be
-adjusted to improve convergence:
-\begin{equation}
-  \mathbf{x} = \mathbf{x} + \omega \mathbf{r}\:,
-  \label{eq:richardson_relaxation}
-\end{equation}
-where $\omega$ is the relaxation parameter. This is effectively a form
-of preconditioning where the system is being scaled on the left by a
-constant value in all rows. Analogously, these relaxation parameter
-techniques can be applied to Monte Carlo to improve convergence as
-demonstrated by Dimov \citep{dimov_new_1998}. In this case, building
-an iteration matrix from Eq~(\ref{eq:richardson_relaxation}) gives:
-\begin{equation}
-  \mathbf{H} = \mathbf{I} - \omega \mathbf{A}\:,
-  \label{eq:relaxed_iteration_matrix}
-\end{equation}
-with the probabilities and weights for the Monte Carlo procedure
-appropriately scaled. By inspection, such a scaling is a simple form
-of preconditioning on the left where all rows in the system are scaled
-by the same scalar parameter. For MCSA, we can stage this scheme with
-two separate relaxation parameters; one for the outer Richardson
-iteration and one for the inner Monte Carlo solve:
-\begin{subequations}
-  \begin{gather}
-    \ve{x}^{k+1/2} = \ve{x}^k + \omega_R \ve{r}^k\:,\\ \ve{r}^{k+1/2}
-    = \ve{b} -
-    \ve{A}\ve{x}^{k+1/2}\:,\\ \omega_N\ve{A}\delta\ve{x}^{k+1/2} =
-    \omega_N\ve{r}^{k+1/2}\:,\\ \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta
-    \ve{x}^{k+1/2}\:,\\ \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:,
-  \end{gather}
-  \label{eq:mcsa_relaxed}
-\end{subequations}
-where $\omega_R$ is the Richardson iteration relaxation parameter and
-$\omega_N$ is the Neumann-Ulam Monte Carlo solve relaxation parameter.
-
-We apply these relaxation parameters to the fuel assembly problem
-along with the reduced domain approximation as a means of studying
-their effects. For each calculation presented, the number of
-iterations required to converge reported was for a single eigenvalue
-iteration.  Again, \sn{3}{4} histories are used at each MCSA iteration
-to compute the Monte Carlo correction using the adjoint collision
-estimator. A reduced domain fill level of 100 is used with a threshold
-of \sn{1}{-10} to filter small values. ILUT preconditioning with a
-drop tolerance of \sn{1}{-5} and fill level of 5 is used as
-well.
-
-Figure~\ref{fig:relax_iters} gives the number of iterations required
-to converge the fuel assembly problem for a 1-group SP1 discretization
-using varying combinations of the relaxation parameters. First, both
-parameters were fixed at the base case of 1 while the other parameter
-was varied as shown in the left-hand side plots of
-Figure~\ref{fig:relax_iters}. For the Richardson relaxation parameter,
-using a value larger than 1 gave better iterative performance up to a
-point, effectively providing a stronger extrapolation using the
-residual at each iteration. For the Neumann relaxation parameter, a
-value of less than 1 is observed to be ideal. Although initially
-counter-intuitive, the fact that the correction computed by the Monte
-Carlo solver has a stochastic error associated with it means that by
-using a Neumann relaxation parameter less than 1, the correction and
-its error are effectively dampened to improve iterative
-performance. Considering the CPU time required to converge a single
-eigenvalue iteration, similar results are also observed for the
-relaxation parameters as given by Figure~\ref{fig:relax_time}. For the
-base cases given by the plots on the left, it was found the a
-Richardson relaxation parameter of 1.1 and a Neumann relaxation
-parameter of 0.7 provided the fastest CPU time for convergence. Fixing
-each parameter at these new values, the calculations were repeated as
-shown for the plots on the right hand sides of both
-Figures~\ref{fig:relax_iters} and \ref{fig:relax_time}. For each
-repeated timing calculation, it was found that the same combination of
-relaxation parameters found in the base cases performed the best
-although they did not necessarily have the best iterative performance.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/relax_iters.pdf}
-  \end{center}
-  \caption{\textbf{Number of iterations to converge a single
-      eigenvalue iteration of the fuel assembly problem as a function
-      of the relaxation parameters.} \textit{Starting in upper left
-      and moving counter-clockwise: Neumann relaxation parameter fixed
-      at 1.0, Richardson relaxation parameter fixed at 1.0, Neumann
-      relaxation parameter fixed at 0.7, Richardson relaxation
-      parameter fixed at 1.1. For each calculation \sn{3}{4}
-      stochastic histories were used to compute the MCSA correction at
-      each iteration.}}
-  \label{fig:relax_iters}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/relax_time.pdf}
-  \end{center}
-  \caption{\textbf{CPU time in seconds to converge a single eigenvalue
-      iteration of the fuel assembly problem as a function of the
-      relaxation parameters.} \textit{Starting in upper left and
-      moving counter-clockwise: Neumann relaxation parameter fixed at
-      1.0, Richardson relaxation parameter fixed at 1.0, Neumann
-      relaxation parameter fixed at 0.7, Richardson relaxation
-      parameter fixed at 1.1. For each calculation \sn{3}{4}
-      stochastic histories were used to compute the MCSA correction at
-      each iteration.}}
-  \label{fig:relax_time}
-\end{figure}
-
-\clearpage
-
-\subsection{Monte Carlo Estimator Comparison}
-\label{subsec:spn_estimator_comparison}
-With convergence obtained for the fuel assembly problem, the advanced
-preconditioners will next be studied with both the collision and
-expected value estimators using the best preconditioning and
-relaxation parameter combinations found by the previous
-analysis. Additionally, both the Richardson-based MCSA iteration given
-by Eq~(\ref{eq:mcsa}) and the MINRES-based MCSA iteration given by
-Eq~(\ref{eq:mcsa_minres}) will be used. Although the expected value
-estimator achieved the best iterative performance in the previous
-comparison and analysis for the simple Poisson problem in
-\S~\ref{sec:parameter_estimator_analysis}, it is possible that the
-collision estimator may perform better from a timing perspective due
-to the sparsity lost by explicit preconditioning. The deterministic
-component of the expected value estimator that couples the current
-state to other local states through the iteration matrix stencil
-requires application of the estimator to potentially several orders of
-magnitude more states at every transition event.
-
-For this study, the fuel assemby problem with a 1-group $SP_1$
-discretization was solved using MCSA with the adjoint Monte Carlo
-solver using both the collision and expected value estimators and both
-the Richardson and MINRES fixed point iterations. Using the results
-from the previous section, a Neumann relaxation parameter of 0.7 was
-used with both estimators and fixed point iterations. For the
-collision estimator, a Richardson relaxation parameter of 1.1 was used
-with the Richardson iteration while it was found the expected value
-estimator had the best performance when a Richardson relaxation
-parameter of 1.0 was used instead. In addition, ILUT preconditioning
-with a drop tolerance of \sn{1}{-5} and a fill level of 5 was used
-with a reduced domain approximation fill level of 100 and a threshold
-of \sn{1}{-10} as in the previous analysis. For each estimator, the
-number of stochastic histories used to compute the MCSA correction was
-varied from 25 to 50,000 with the number of MCSA iterations required
-to converge to a tolerance of \sn{1}{-8} recorded for a single
-eigenvalue iteration along with the CPU time required to achieve
-convergence.
-
-Figure~\ref{fig:spn_estimator_iters} gives the iteration count results
-and Figure~\ref{fig:spn_estimator_time} gives the timing
-results. Remarkably, although the fuel assembly problem is
-significantly more complicated than the simple Poisson problem studied
-in \S~\ref{sec:parameter_estimator_analysis}, the results here are
-largely the same with Figure~\ref{fig:spn_estimator_iters} and
-Figure~\ref{fig:estimator_nh_iters} having identical qualitative
-behavior. Using the expected value estimator with both fixed point
-iterations creates a relative insensitivity to the number of histories
-used to compute the correction. Marginally better iterative and timing
-performance was observed for the expected value estimator when using
-the Richardson iteration. Better timing performance is expected as
-computing the extrapolation parameter in the MINRES iteration requires
-several additional parallel operations. However, we expect that
-because the MINRES iteration will converge faster on its own when
-compared to the Richardson iteration that better iterative performance
-will be observed when it is accelerated with MCSA. The fact that this
-was not observed signals that an MCSA scheme leveraging the expected
-value estimator is also effectively insensitive to the fixed point
-iteration used. For the collision estimator, the MINRES iteration
-permits fewer stochastic histories to be used with each MCSA iteration
-while still maintaining good iterative performance. With respect to
-CPU time, there is a more pronounced effect due to the larger density
-of states created by the explicit ILUT preconditioning. Using a
-reduced domain fill level of 100 creates a situation where the
-expected value estimator is significantly more expensive per history
-to compute that the collision estimator due to the coupling of states.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/estimator_iters.pdf}
-  \end{center}
-  \caption{\textbf{MCSA iterations required to converge the fuel
-      assembly problem to a tolerance of \sn{1}{-8}.} \textit{Both the
-      collision and expected value estimators were used with the
-      adjoint Monte Carlo solver to compute the correction at each
-      iteration. Each estimator was used with the Richardson and
-      MINRES fixed point iterations within an MCSA iteration.}}
-  \label{fig:spn_estimator_iters}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/estimator_time.pdf}
-  \end{center}
-  \caption{\textbf{Total CPU time in seconds required to converge the
-      fuel assembly problem to a tolerance of \sn{1}{-8}.}
-    \textit{Both the collision and expected value estimators were used
-      with the adjoint Monte Carlo solver to compute the correction at
-      each iteration. Each estimator was used with the Richardson and
-      MINRES fixed point iterations within an MCSA iteration.}}
-  \label{fig:spn_estimator_time}
-\end{figure}
-
-From both Figure~\ref{fig:spn_estimator_iters} and
-Figure~\ref{fig:spn_estimator_time} we see that both estimators and
-fixed point iterations have comparable iterative and timing
-performance when 1,000 histories are used at each MCSA
-iteration. Using 1,000 histories per MCSA iteration, the infinity norm
-of the residual vector is presented at every MCSA iteration for both
-estimators and both fixed point iterations in
-Figure~\ref{fig:spn_estimator_convergence} as an additional means of
-comparison. At this number of histories, all MCSA combinations
-converge the fuel assembly problem monotonically at and approximately
-the same rate, giving little reason to choose one over the other. The
-end result is that both estimators provide a viable method for getting
-good MCSA iterative performance with comparable timing
-performance. For the collision estimator, using around 20,000
-stochastic histories at every iteration gave the best iterative
-performance while 500 histories per iteration gave the best iterative
-performance when using the expected value estimator. For purely serial
-computational performance, there is not a distinguishing feature for
-the fuel assembly problem presented that would cause one to choose one
-of the estimators and fixed point iterations over the other as they
-all have similar minimum CPU times over the range of values tested.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/estimator_convergence.pdf}
-  \end{center}
-  \caption{\textbf{Residual infinity norm as a function of MCSA
-      iteration with 1,000 stochastic histories per iteration.}
-    \textit{Both the collision and expected value estimators were used
-      with the adjoint Monte Carlo solver to compute the correction at
-      each iteration. Each estimator was used with the Richardson and
-      MINRES fixed point iterations within an MCSA iteration.}}
-  \label{fig:spn_estimator_convergence}
-\end{figure}
-
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{MCSA Verification}
-\label{sec:spn_mcsa_verification}
-Through the numerical studies presented in this chapter, we have
-developed a set of preconditioning techniques that permit MCSA to be
-used with the $SP_N$ form of the neutron transport equation and
-applied to a difficult fuel assembly criticality problem. Along with
-these preconditioning techniques, a set a varying MCSA parameters were
-analyzed to find those that yielded the best performance for this
-particular problem. In addition, several steps were taken to mitigate
-the dense composite linear operators that arise when performing
-explicit MCSA preconditioning in order to achieve convergence. Using
-these results, we compare MCSA directly to conventional linear solvers
-that would be typically used to solve the $SP_N$ form of the transport
-equation as presented here. In particular, we will use a pair of
-production Krylov solvers with which we will compare direct numerical
-results in order to verify MCSA solutions in this section and
-iterative and CPU timing performance in the next order to gauge
-feasibility and identify areas that require further research.
-
-In order to verify the correctness of the MCSA method and
-implementation used for this work, MCSA solutions to the fuel assembly
-problem with 1, 2, and 4 energy groups using $SP_1$ discretization
-will be directly compared over all group and cell fluxes computed by
-converging the criticality problem using both BiCGStab and GMRES
-production implementations from the Trilinos scientific computing
-library Aztec \citep{heroux_overview_2005}. In addition to comparing
-the cell and group fluxes, the number of eigenvalue iterations
-required to converge the problem along with the eigenvalue itself will
-be compared as a means of verification. If all values computed using
-MCSA agree with the results from the production Krylov solvers within
-the solution tolerance then the MCSA solutions will be deemed verified
-for these problems. Additionaly, this verification is only applicable
-to serial results with parallel verification for this problem provided
-in Chapter~\ref{ch:parallel_methods}.
-
-For the verification calculations, all solvers will be preconditioned
-with ILUT using a drop tolerance of \sn{1}{-5} and a fill level of
-5. For GMRES, no restrictions were placed on the size of the subspace
-and therefore no restarts occurred. When the collision estimator was
-used with MCSA, \sn{2}{4} stochastic histories were used to compute
-the correction for every energy group in the problem (\sn{4}{4} and
-\sn{8}{4} histories total for the 2 and 4 group calculations
-respectively) corresponding to approximately 1 stochastic history per
-DOF. When the expected value estimator was used, \sn{5}{2} stochastic
-histories used for each energy group (\sn{1}{3} and \sn{1.5}{3}
-histories total for the 2 and 4 group calculations) giving
-approximately 1 stochastic history for every 5 DOFs. The reduced
-domain approximation was also applied in conjuction with the ILUT
-preconditioning using a fill level of 100 and a threshold value of
-\sn{1}{-10} to reduce the density of states in the Monte Carlo
-problem. For relaxation parameters, all MCSA compuations used a
-Neumann relaxation parameter of 0.7 while a Richardson relaxation of
-1.1 was used with the collision estimator and 1.0 with the expected
-value estimator as determined by the previous analysis of the
-relaxation parameters. Table~\ref{tab:spn_solver_defs} gives
-definitions for the solvers used to generate the results in the
-remainder of this section. For the energy group structures,
-Table~\ref{tab:spn_group_structure} gives the lower bounds of each
-group in eV (an implicit upper bound of \sn{2}{6} eV is assumed for
-group 0).
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{ll}\hline\hline
-      \multicolumn{1}{l}{\textbf{Name}} & 
-      \multicolumn{1}{l}{\textbf{Definition}} \\
-      BiCGStab-ILUT & BiCGStab preconditioned with ILUT \\
-      GMRES-ILUT & GMRES preconditioned with ILUT \\
-      MCSA-ILUT-R-C & MCSA preconditioned with ILUT using Richardson \\ 
-                    & fixed point iteration and collision estimator \\
-      MCSA-ILUT-MR-C & MCSA preconditioned with ILUT using MINRES \\
-                     & fixed point iteration and collision estimator \\
-      MCSA-ILUT-R-EV & MCSA preconditioned with ILUT using Richardson \\
-                     & fixed point iteration and expected value estimator \\
-      MCSA-ILUT-MR-EV & MCSA preconditioned with ILUT using MINRES \\
-                      & fixed point iteration and expected value estimator \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Solver definitions used for MCSA verification and
-      performance analysis.} \textit{The ILUT preconditioner
-      parameters were identical for all calculations and solvers.}}
-  \label{tab:spn_solver_defs}
-\end{table}
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{cl}\hline\hline
-      \multicolumn{1}{c}{\textbf{Number of Groups}} & 
-      \multicolumn{1}{l}{\textbf{Lower Bounds (eV)}} \\
-      1 & \{ \sn{1}{-5} \} \\
-      2 & \{ \sn{1}{-1}, \sn{1}{-5} \} \\
-      4 & \{ \sn{1}{1}, \sn{1}{0}, \sn{1}{-1}, \sn{1}{-5} \} \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Energy group lower bounds for the multigroup fuel
-      assembly criticality problem in electron volts.} \textit{An
-      implicit upper bound of \sn{2}{6} eV is assumedfor group zero.}}
-  \label{tab:spn_group_structure}
-\end{table}
-
-For every combination of energy group structure and solver, the
-eigenvalue problem was converged with an eigensolver tolerance of
-\sn{1}{-6} and a linear solver tolerance of \sn{1}{-8} in a serial
-calculation. Table~\ref{tab:serial_ev_results} gives the eigenvalue
-computed by each solver for each problem as well as the number of
-eigenvalue iterations required to converge. For every problem, each
-solver converged to the same eigenvalue within the tolerance of the
-eigensolver in the same number of eigenvalue iterations.
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lccc}\hline\hline
-      \multicolumn{1}{c}{\textbf{Solver}} & 
-      \multicolumn{1}{c}{\textbf{Number of Groups}} & 
-      \multicolumn{1}{c}{\textbf{Iterations}} &
-      \multicolumn{1}{c}{\textbf{Eigenvalue}} \\
-      \hline
-      BiCGStab-ILUT & 1 & 25 & 1.1059870 \\
-      GMRES-ILUT & 1 & 25 & 1.1059870 \\
-      MCSA-ILUT-R-C & 1 & 25 & 1.1059870 \\
-      MCSA-ILUT-MR-C & 1 & 25 & 1.1059870 \\
-      MCSA-ILUT-R-EV & 1 & 25 & 1.1059870 \\
-      MCSA-ILUT-MR-EV & 1 & 25 & 1.1059870 \\
-      \hline
-      BiCGStab-ILUT & 2 & 35 & 1.1552390 \\
-      GMRES-ILUT & 2 & 35 & 1.1552390 \\
-      MCSA-ILUT-R-C & 2 & 35 & 1.1552390 \\
-      MCSA-ILUT-MR-C & 2 & 35 & 1.1552390 \\
-      MCSA-ILUT-R-EV & 2 & 35 & 1.1552390 \\
-      MCSA-ILUT-MR-EV & 2 & 35 & 1.1552390 \\
-      \hline
-      BiCGStab-ILUT & 4 & 31 & 1.1758500 \\
-      GMRES-ILUT & 4 & 31 & 1.1758500 \\
-      MCSA-ILUT-R-C & 4 & 31 & 1.1758500 \\
-      MCSA-ILUT-MR-C & 4 & 31 & 1.1758500 \\
-      MCSA-ILUT-R-EV & 4 & 31 & 1.1758500 \\
-      MCSA-ILUT-MR-EV & 4 & 31 & 1.1758500 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Serial eigensolver verification results for the
-      fuel assembly problem.} \textit{Each calculation was converged
-      with the same eigensolver and linear solver tolerances.}}
-  \label{tab:serial_ev_results}
-\end{table}
-
-For each converged eigenvalue calculation, the scalar flux in each
-energy group was compared over the entire spatial with the
-BiCGStab-ILUT solution used as the reference to compute an absolute
-difference. For the 1 group calculations, these absolute differences
-are plotted spatially along the centerline of the fuel assembly in
-Figure~\ref{fig:serial_1_group_diff}. Interestingly, except for the
-MCSA-ILUT-MR-C calculation, all other 1-group calculations including
-GMRES had a maximum global flux difference of \sn{9.313}{-10} even
-though different spatial results were observed. For the MCSA-ILUT-MR-C
-calculation, a slightly larger maximum error of \sn{2.205}{-9} was
-observed for reasons that remain unexplained. All 1-group calculations
-have a minimum observed absolute difference of 0 with respect to the
-BiCGStab calculation.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/serial_1_group.pdf}
-  \end{center}
-  \caption{\textbf{Absolute difference of the group 0 scalar flux
-      between the Aztec BiCGStab calculations and Aztec GMRES and the
-      MCSA variations for the 1-group calculation.} \textit{Left to
-      right: GMRES-ILUT, MCSA-ILUT-R-C, MCSA-ILUT-MR-C,
-      MCSA-ILUT-R-EV, MCSA-ILUT-MR-EV.}}
-  \label{fig:serial_1_group_diff}
-\end{figure}
-
-Table~\ref{tab:serial_differences} gives the maximum and minimum
-absolute differences for all calculations and all energy groups using
-the BiCGStab calculation as the reference. All comparisons were
-performed within a group such that the MCSA-ILUT-R-C results in group
-3 for the 4-group calculation were computed with the BiCGStab-ILUT
-solution in group 3 for the 4-group calculation as a reference.
-
-%%---------------------------------------------------------------------------%%
-\section{MCSA Performance Comparison to Conventional Methods}
-\label{sec:spn_comparison}
-Using the same set of solver parameters used for verification, MCSA
-will be directly compared to the production Krylov solvers in terms of
-both iterative performance and CPU timing for solutions to the fuel
-assembly problem. In addition to the parameters used for the
-verification analysis, artificial absorption was also introduced with
-a value of 0.2 for the MCSA calculations that used the collision
-esitmator as CPU timing performance was improved without a loss in
-iterative performance. Adding this absorption did not change the
-numerical results of the calculation.
-
-Figure~\ref{fig:spn_comparison_iterations} gives the average number of
-linear solver iterations required to converge a single eigenvalue
-iteration of the fuel assembly problem as a function of the number of
-energy groups. In general, the iterative performance of all the
-solvers tested is comparable with BiCGStab performing the best. We
-expect this because not only are the multigroup $SP_N$ equations
-elliptic and positive-definite, but they are also nearly symmetric and
-therefore we expect the conjugate gradient-based methods to perform
-well. It also important to note that the iterative performance of each
-solver is not a strong function of the number of energy groups in the
-problem (and therefore the problem size). For MCSA, we are linearly
-increasing the number of stochastic histories used to compute the
-correction at each iteration as the number of energy groups are
-increased and therefore the work performed for the problem is of
-$O(N)$. It is also important to note the the reduced domain
-approximation parameters, particularly the fill level, were fixed as
-the number of energy groups was increased. By fixing the fill level,
-we are effectively fixing the amount of information contained in the
-composite linear operator available for the Monte Carlo
-problem. Limiting this information to 100 entries per row in the
-preconditioned system did not appear to have a significant effect on
-the iterative performance of the solver. For larger numbers of energy
-groups (and therefore DOFs) this may be an issue and this number may
-need to be increased to maintain iterative performance. Currently,
-memory restrictions arising from the explicit preconditioning strategy
-prevent larger energy groups to be analyzed using an appropriate
-preconditioner for the fuel assembly problem.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/solver_iters.pdf}
-  \end{center}
-  \caption{\textbf{Average number of iterations required to converge
-      each eigenvalue iteration for the fuel assembly problem as a
-      function of energy groups.}
-    \textit{Table~\ref{tab:spn_solver_defs} gives the description for
-      each solver type presented in the legend.}}
-  \label{fig:spn_comparison_iterations}
-\end{figure}
-
-Although iterative performance for MCSA was comparable to that
-observed for production Krylov implementations using exactly the same
-preconditioning scheme, timing performance is not as
-comparable. Figure~\ref{fig:spn_comparison_time} gives the average CPU
-time per linear solver iteration required to converge the fuel
-assembly problem with all linear solver iterations over all eigenvalue
-iterations used to compute the average. The production Krylov solver
-implementations are approximately an order of magnitude faster than
-the MCSA implementations presented here. We expect these results for
-several reasons. First, even when the reduced domain approximation is
-used there are $O(100)$ elements in each row of the system and each of
-those elements must be processed during a transition in the Monte
-Carlo random walk sequence. In addition, although the preconditioned
-fixed point iterations used in the MCSA sequence do not use the
-composite linear operator but rather a sequence of matrix vector
-multiplies to achieve the same preconditioning effect, they do use the
-explicitly extracted inverse of the preconditioning matrices which
-themselves are dense, leading to exceedingly slow computation times
-even in the fixed point iteration. For the production Krylov methods,
-the ILUT preconditioners are applied to a vector with two triangular
-solves, one for each triangular factor. Also, the original linear
-operator has $O(10)$ elements in each row of the system, an order of
-magnitude less than that used in the reduced domain composite
-operator. Second, the implementation presented here has not been
-optimized and it is likely that several components of the algorithm
-can be implemented in a more desireable time complexity. The
-production Krylov solvers presented here have enjoyed decades of
-professional development and optimization. However, it is most
-important to note here that the CPU timing data presented in
-Figure~\ref{fig:spn_comparison_time} shows that all methods have the
-same time complexity; their differences are merely the manifestation
-of a large time constant due to the effects of the explicit
-preconditioning scheme. For transport systems where this is not a
-problem, the literature explicitly shows that MCSA can be competetive
-with these production Krylov methods using CPU time as a measure
-\citep{evans_monte_2012}.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/solver_time.pdf}
-  \end{center}
-  \caption{\textbf{Average CPU time per linear solver iteration in
-      seconds for the fuel assembly problem as a function of energy
-      groups.}  \textit{All linear solver iterations over all
-      eigenvalue iterations were used to compute the
-      average. Table~\ref{tab:spn_solver_defs} gives the description
-      for each solver type presented in the legend.}}
-  \label{fig:spn_comparison_time}
-\end{figure}
-
-Finally, not considered in Figure~\ref{fig:spn_comparison_time} is the
-time required to actually form the explicit inverse preconditioner
-matrices and the composite operator through matrix-matrix
-multiplication. The timing numbers reported were simply to perform the
-MCSA iteration procedure with the composite operator already
-formed. Figure~\ref{fig:spn_comparison_prec_time} additionally
-presents the CPU times for MCSA convergence with the time to form the
-inverse of the preconditioners and the composite linear operator
-through matrix-matrix multiplication amortized over all iterations. As
-is readily observed, including the costs of these operations increases
-the MCSA computation time by another order of magnitude.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/solver_p_time.pdf}
-  \end{center}
-  \caption{\textbf{Average CPU time per iteration in seconds for the
-      fuel assembly problem as a function of energy groups with
-      preconditioning time included for the MCSA methods.}
-    \textit{All linear solver iterations over all eigenvalue
-      iterations were used to compute the
-      average. Table~\ref{tab:spn_solver_defs} gives the description
-      for each solver type presented in the legend. Data labeled
-      starting with PMCSA is identical to those labeled with MCSA
-      except that they additionally include the cost of generating the
-      inverse of the preconditioners and composite linear operator.}}
-  \label{fig:spn_comparison_prec_time}
-\end{figure}
-
-Based on the performance results in this section, MCSA shows promise
-as a competitive and perhaps even superior method for solutions to the
-neutron transport problem discretized with the $SP_N$
-approximation. Not only are the correct answers produced when compared
-to production linear solvers, but the iterative performance is
-comparable to production Krylov methods with identical preconditioning
-that would typically be used every-day calculations. From a CPU timing
-perspective, the methods yield the same time complexity as a function
-of the number of energy groups in the problem when compared to the
-production Krylov methods. Here, a large time constant is generated
-due to the explicit preconditioning strategy and the generation of
-dense linear operators as a result. To improve these results and put
-general MCSA schemes into a performance regime where they are
-competitive with Krylov methods for neutron transport problems,
-significant research will be required to improve upon the explicit
-preconditioning scheme presented here.

### 2013-06-11 doc/dissertation/document/chapters/introduction/introduction.tex
adding some example figures for domain decomposed monte carlo

@@
 levels of parallelism.
+
+\subsection{Solutions for the $SP_N$ Equations}
+\label{subsec:spn_motiviation}
+The neutron transport problem is complicated. Solutions cover a large
+phase space and the problems of interest are often geometrically
+complex, very large, or both, requiring tremendous computational
+resources to generate an adequate solution. Modern deterministic
+methods for large scale problems are commonly variants on the discrete
+ordinates ($S_N$) method \citep{evans_denovo:_2010}. For fission
+reactor neutronics simulations, the $S_N$ method requires potentially
+trillions of unknown angular flux moments to be computed to achieve
+good accuracy for the responses of interest
+\citep{slaybaugh_acceleration_2011}. Other forms of the transport
+problem, including the $P_N$ method, take on a simpler form than the
+more common $S_N$ methods but lack in accuracy when compared while
+still requiring considerable computational resources for solutions in
+multiple dimensions.
+
+In the 1960's, Gelbard developed an ad-hoc multidimensional extension
+of the simple single dimension planar $P_N$ equations that created a
+system of coupled, diffusion-like equations known as the simplified
+$P_N$ ($SP_N$) equations. Up until around the 1990's, the $SP_N$
+method was either widely unknown, widely unused, or combination of
+both even though numerical studies showed promising results with
+better solutions than diffusion theory and a significant reduction in
+computational time over more accurate methods such as discrete
+ordinates. Why did this happen? A significant problem, pointed out by
+Brantley and Larsen \citep{brantley_simplified_2000}, was that little
+rigor had been applied to the formulation of the $SP_N$ equations
+since their derivation through primarily heuristic arguments. Instead,
+studies at that time focused on simply comparing the results of the
+method to other contemporary transport solution strategies. In
+addition, many problems of interest from the literature at the time
+were either solved using nodal-type methods for reactor-sized problems
+or $S_N$-type methods for benchmark problems with intricate material
+configurations and potentially large flux gradients over small spatial
+domains.
+
+So why reconsider the $SP_N$ equations? Starting in the 1990's and
+primarily due to Larsen and his colleagues, the $SP_N$ equations have
+been given a more rigorous treatment with both variational and
+asymptotic derivations performed as a means of verification. In
+addition, these equations have been more rigorously studied as
+solution methods to MOX fuel problems and have been shown to provide
+accurate solutions. With this mathematical literature to provide a
+solid numerical footing for the method, we look at its application to
+today's challenge problems in neutron transport for fission reactor
+analysis. The reduction in numerical complexity of current
+deterministic solution methods using the $S_N$ approximation could
+mean significant savings in both compute time and memory required. In
+addition, the characteristics of the solution to the transport problem
+for a steady state reactor core permit diffusion theory to be used; a
+staple of the nuclear industry since its inception. Therefore, if
+diffusion theory is applicable, then finer grained solutions that
+capture more of the physics contained in the transport equation should
+be possible with the $SP_N$ method. In doing so, we also expect from
+the literature to obtain computed responses on the order of accuracy
+we would expect from an appropriately discretized $S_N$ method at a
+fraction of the cost.
+
+In order to leverage MCSA as a solution technique for physics
+problems, its current formulation requires the linear operator and all
+preconditioners to be explicitly formed. Recent developments in the
+Exnihilo neutronics package at Oak Ridge National Laboratory have
+permitted generation of the $SP_N$ system of equations for detailed
+neutronics models of fission reactor-based
+systems\citep{evans_simpli_2013}. By fully forming these equations and
+formulating them as a linear algebra problem instead of using the
+explicit iterative methods of the past, we now have access to all of
+the modern advancements in computational linear algebra including
+Krylov solvers for asymmetric systems and algebraic preconditioning
+methods. This leads us to then explore the applicability of our work
+in discrete Monte Carlo methods for linear systems as a possible
+solution method for the $SP_N$ equations. In addition, solving the
+$SP_N$ equations in this way also breaks away from the $S_N$ forms of
+parallelism where spatial parallelism is achieved by an efficient
+parallel sweep, angular efficiency achieved by pipe-lining, and energy
+parallelism achieved by decoupling the groups. With the $SP_N$
+equations as a full matrix system, we now can parallelize the problem
+as prescribed by the linear solver, which may be significantly more
+scalable than current $S_N$ transport practices.
+
+\subsection{Solutions for the Navier-Stokes Equations}
+\label{subsec:ns_motiviation}
 

### 2013-06-11 doc/dissertation/document/chapters/mc_background/mc_background.tex
adding some example figures for domain decomposed monte carlo

@@
 \end{equation}
-Choosing the infinity norm of the operator as defined in
-Eq~(\ref{eq:matrix_infinity_norm}), the row normalized probability
-matrix will yield a norm of 1 giving the following inequality for
-relating $\ve{G}$ and $\ve{H}$:
+Choosing the infinity norm of the operator, the row normalized
+probability matrix will yield a norm of 1 giving the following
+inequality for relating $\ve{G}$ and $\ve{H}$:
 \begin{equation}

### 2013-06-11 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
restructuring verification results
adding some example figures for domain decomposed monte carlo

@@
-\chapter{Monte Carlo Solution Methods for the Navier-Stokes Equations}
+\chapter{Monte Carlo Synthetic Acceleration Methods for the Navier-Stokes Equations}
 \label{ch:nonlinear_problem}
@@
 \label{subsec:parallel_fanm}
-Like the parallelization of the MCSA method described in
-\S~\ref{sec:parallel_stochastic_methods}, a parallel FANM method
-relies on a basic set of parallel matrix-vector operations as well as
-the global residual and Jacobian assembly procedure described in
+A parallel FANM method relies on a basic set of parallel matrix-vector
+operations as well as the global residual and Jacobian assembly
+procedure described in
 \S~\ref{subsubsec:automatic_differentiation}. Consider the Newton

### 2013-06-11 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
mc background edits and spell check
adding some example figures for domain decomposed monte carlo

@@
 \label{ch:parallel_methods}
-For Monte Carlo methods, and in particular MCSA, to be viable at the
-production scale, scalable parallel implementations are required. In a
-general linear algebra context for discrete systems, this has not yet
-been achieved. Therefore, we will develop and implement parallel
-algorithms for these methods leveraging both the knowledge gained from
-the general parallel implementations of Krylov methods and modern
-parallel strategies for Monte Carlo as developed by the reactor
-physics community. In order to formulate a parallel MCSA algorithm, we
-recognize that the algorithm occurs in two stages, an outer iteration
-performing fixed point iteration and applying the correction, and an
-inner Monte Carlo solver that is generating the correction via the
-adjoint or forward methods. The parallel aspects of both these
-components must be considered.
+For MCSA to be viable at the production scale, scalable parallel
+implementations of the algorithm are required. Reviewing the
+literature, MCSA has yet to be parallelized and the Neumann-Ulam Monte
+Carlo method on which it is built has only been parallelized through
+history-level parallelism with full domain replication. In order to
+solve large linear systems with MCSA, a domain-decomposed parallel
+strategy is required. In the formulation of a parallel MCSA algorithm,
+we recognize that the algorithm occurs in two stages, an outer
+iteration performing fixed point iteration and applying the
+correction, and an inner Monte Carlo solver that is generating the
+correction via the adjoint or forward methods. The parallel aspects of
+both these components must be considered. Therefore, we will develop
+and implement parallel algorithms for the Neumann-Ulam and MCSA
+methods leveraging both the knowledge gained from the general parallel
+implementations of Krylov methods reviewed in
+Appendix~\ref{ch:linear_problem} and modern parallel strategies for
+domain decomposed Monte Carlo as developed by the reactor physics
+community.
 
 In this chapter we briefly review particle transport methods for
-domain decomposed Monte Carlo and then provide an analytic framework
-from which to estimate its performance. We then devise an algorithm
-based on these methods for efficient parallel solutions using the
-Neumann-Ulam method and MCSA and verify its correctness by comparing
-the numerical results against production Krylov methods. We then
-perform parallel scaling studings to test its performance on both a
-small computing cluster and a leadership class machine and compare
-this performance to the production Krylov methods.
+domain decomposed Monte Carlo. Considering a domain decomposed
+Neumann-Ulam method, we derive an analytic framework based on
+algebraic quantities from which to estimate its performance when
+applied to linear systems. We then devise a parallel algorithm for the
+Neumann-Ulam method based on the multiple-set overlapping-domain
+decomposition algorithm and a parallel algorithm for the MCSA that
+leverages the parallel Monte Carlo algorithm and general parallel
+matrix-vector operations. In identical fashion to the serial
+computations presented in Chapter~\ref{ch:spn_equations}, we use
+parallel solutions for the $SP_N$ equations for the fuel assembly
+criticality problem to verify the correctness of the parallel
+algorithm and implementation by comparing the numerical results
+against production Krylov methods. Using the verified implementation
+of the algorithm, we perform parallel scaling studings to test its
+performance on both a small computing cluster and a leadership class
+machine and compare this performance to the production Krylov methods.
 
 %%---------------------------------------------------------------------------%%
-\section{Domain Decomposition for Monte Carlo}
+\section{Domain Decomposed Monte Carlo}
 \label{sec:msod}
@@
 subset of the equations in the linear system. Given this convention,
-the adjoint Monte Carlo algorithm must perform random walks over a
-domain that is decomposed and must remain decomposed due to memory
+the Neumann-Ulam Monte Carlo algorithm must perform random walks over
+a domain that is decomposed and must remain decomposed due to memory
 limitations. This naturally leads us to seek parallel Monte Carlo
@@
 
-In the context of radiation transport, Brunner and colleagues provided
-a survey of algorithms for achieving this as implemented in production
-implicit Monte Carlo codes \citep{brunner_comparison_2006}. In their
-work they identify two data sets that are required to be communicated:
-the sharing of particles that are transported from one domain to
-another and therefore from one processor to another and a global
-communication that signals if particle transport has been completed on
-all processors. The algorithms presented are a fully-locking
-synchronous scheme, an asynchronous-send/synchronous-receive pattern,
-a traditional master/slave scheme, and a modified master/slave scheme
-the implements a binary tree pattern for the global reduction type
-operations needed to communicate between the master and slave
-processes. They observed that the modified master/slave scheme
-performed best in that global communications were implemented more
-efficiently than those required by the asynchronous
-scheme. Furthermore, none of these schemes handled load-imbalanced
-cases efficiently. Such cases will be common if the source sampled in
-the Monte Carlo random walk is not isotropic and not evenly
-distributed throughout the global domain. It was noted that
-efficiencies were improved by increasing the frequency by which
-particle data was communicated between domain-adjacent
-processors. However, this ultimately increases communication costs. In
-2009, Brunner extended his work by using a more load-balanced approach
-with a fully asynchronous communication pattern
-\citep{brunner_efficient_2009}. Although the extended implementation
-was more robust and allowed for scaling to larger numbers of
-processors, performance issues were still noted with parallel
-efficiency improvements needed in both the weak and strong scaling
-cases for unbalanced problems. These results led Brunner to conclude
-that a combination of domain decomposition and domain replication
-could be used to solve some of these issues.
-
-\subsection{Multiple-Set Overlapping-Domain Decomposition}
-\label{subsec:msod}
-In 2010, Wagner and colleagues developed the \textit{multiple-set
-  overlapping-domain} (MSOD) decomposition for parallel Monte Carlo
-applications for full-core light water reactor analysis
-\citep{wagner_hybrid_2010}. In their work, an extension of Brunner's,
-their scheme employed similar parallel algorithms for particle
-transport but a certain amount of overlap between adjacent domains was
-used to decrease the number of particles leaving the local domain. In
-addition, Wagner utilized a level of replication of the domain such
-that the domain was only decomposed on $O(100)$ processors and if
-replicated $O(1,000)$ times achieves simulation on $O(100,000)$
-processors, thus providing spatial and particle parallelism. Each
-collection of processors that constitutes a representation of the
-entire domain is referred to as a set, and within a set overlap occurs
-among its sub-domains. The original motivation was to decompose the
-domain in a way that it remained in a physical cabinet in a large
-distributed machine, thus reducing latency costs during
-communication. A multiple set scheme is also motivated by the fact
-that communication during particle transport only occurs within a set,
-limiting communications during the transport procedure to a group of
-$O(100)$ processors, a number that was shown to have excellent
-parallel efficiencies in Brunner's work and therefore will scale well
-in this algorithm. The overlapping domains within each set also
-demonstrated reduced communication costs. On each processor, the
-source is sampled in the local domain that would exist if no overlap
-was used while tallies can be made over the entire overlapping domain.
-
-To demonstrate this, consider the example adapted from Mervin's work
-with Wagner and others in the same area \citep{mervin_variance_2012}
-and presented in Figure~\ref{fig:msod_example}.
 \begin{figure}[t!]
@@
     \scalebox{1.5}{
-      \input{chapters/parallel_mc/msod_example.pdftex_t} }
+      \input{chapters/parallel_mc/ddmc_example.pdftex_t} }
   \end{center}
-  \caption{\textbf{Overlapping domain example illustrating how domain
-      overlap can reduce communication costs.}
-    \textit{All particles start in the blue region of interest. The
-      dashed line represents 0.5 domain overlap between domains.}}
-  \label{fig:msod_example}
+  \caption{\textbf{Domain decomposed Monte Carlo transport example
+      illustrating how domain decomposition requires parallel
+      communication of particles.}}
+  \label{fig:ddmc_example}
 \end{figure}
-In this example, 3 particle histories are presented emanating from the
-blue region of interest. Starting with particle A, if no domain
-overlap is used then the only the blue domain exists on the starting
-processor. Particle A is then transported through 3 other domains
-before the history ends, therefore requiring three communications to
-occur in Brunner's algorithm. If a 0.5 domain overlap is permitted as
-shown by the dashed line, then the starting process owns enough of the
-domain such that no communications must occur in order to complete the
-particle A transport process. Using 0.5 domain overlap also easily
-eliminates cases such as the represented by the path of particle C. In
-this case, particle C is scattering between two adjacent domains,
-incurring a large latency cost for a single particle. Finally, with
-particle B we observe that 0.5 domain overlap will still not eliminate
-all communications. However, if 1 domain overlap were used, the entire
-geometry shown in Figure~\ref{fig:msod_example} would be contained on
-the source processor and therefore transport of all 3 particles
-without communication would occur.
 
-Wagner and colleagues used this methodology for a 2-dimensional
-calculation of a pressurized water reactor core and varied the domain
-overlap from 0 to 3 domain overlap (a $7 \times 7$ box in the context
-of our example) where a domain constituted a fuel assembly. For the
-fully domain decomposed case, they observed that 76.12\% of all source
-particles leave the domain. At 1.5 domain overlap, the percentage of
-source particles born in the center assembly leaving the processor
-domain dropped to 1.05\% and even further for 0.02\% for the 3 domain
-overlap. Based on these results, this overlap approach, coupled with
-the multiple sets paradigm that will scale for existing parallel
-transport algorithms, provides a scalable Monte Carlo algorithm for
-today's modern machines.
+\begin{figure}[t!]
+  \begin{center}
+    \scalebox{1.5}{
+      \input{chapters/parallel_mc/ddnu_example.pdftex_t} }
+  \end{center}
+  \caption{\textbf{Domain decomposed Neumann-Ulam example illustrating
+      how domain decomposition requires parallel communication of
+      histories.} \textit{Each mesh point corresponds to an equation
+      in the linear system and the coupling among equations is
+      described by the discretization of the problem.}}
+  \label{fig:ddnu_example}
+\end{figure}
+
+\clearpage
 
 %%---------------------------------------------------------------------------%%
-\section{An Analytic Performance Framework for Domain-Decomposed Monte Carlo}
+\section{Analytic Performance Framework for Domain-Decomposed Monte Carlo}
 \label{sec:analytic_framework}
-With Wagner's work \citep{wagner_hybrid_2010}, we observed that domain
-overlap implemented with a domain decomposed Monte Carlo setting
-reduces the amount of nearest neighbor communication that must occur
-during transport. To date, parallel Neumann-Ulam methods have been
-limited to full domain replication with parallelism exploited through
-individual histories \citep{alexandrov_efficient_1998} and in this
-chapter we will exploit Wagner's work to alleviate this. To accomplish
-this, we recognize from the literature that stochastic histories must
-be transported from domain to domain as the simulation progresses and
-they transition to states that are not in the local domain. Because we
-have chosen a domain decomposition strategy in a parallel environment,
-this means that communication of these histories must occur between
-compute nodes owning neighboring pieces of the global domain. We wish
-to characterize this communication not only because communication is
-in general expensive, but also because these nearest-neighbor
-communication sequences, specifically, have poor algorithmic strong
-scaling \citep{gropp_high-performance_2001} in much the same way as a
-matrix-vector multiply operation. Therefore, we desire a framework to
-provide a simple, analytic theory based on the properties of the
-linear system that will allow for estimates of the domain decomposed
-behavior of the adjoint Neumann-Ulam method in terms of communication
-costs.
+To date, parallel Neumann-Ulam methods have been limited to full
+domain replication with parallelism exploited through individual
+histories \citep{alexandrov_efficient_1998} and in this chapter we
+will exploit particle transport algorithms to alleviate this. To
+accomplish this, we recognize from the literature that stochastic
+histories must be transported from domain to domain as the simulation
+progresses and they transition to states that are not in the local
+domain. Because we have chosen a domain decomposition strategy in a
+parallel environment, this means that communication of these histories
+must occur between compute nodes owning neighboring pieces of the
+global domain. We wish to characterize this communication not only
+because communication is in general expensive, but also because these
+nearest-neighbor communication sequences, specifically, have poor
+algorithmic strong scaling \citep{gropp_high-performance_2001} in much
+the same way as a matrix-vector multiply operation. Therefore, we
+desire a framework to provide a simple, analytic theory based on the
+properties of the linear system that will allow for estimates of the
+domain decomposed behavior of the Neumann-Ulam method in terms of the
+amount of information that must be communicated.
 
-\subsection{Leakage Fractions for Symmetric Systems}
-\label{sec:symmetric_leakage_fractions}
 When solving problems where the linear operator is symmetric, a host
@@
 deterministic linear solvers. Using past work, these theories are
-adapted to the domain decomposed adjoint Neumann-Ulam method using the
+adapted to the domain decomposed Neumann-Ulam method using the
 one-speed, two-dimensional neutron diffusion equation and spatial
-discretization presented in \S~\ref{subsec:model_problem}. Using the
-linear system generated by the discretization of the model problem, we
-use a spectral analysis to generate analytic relations for the
-eigenvalues of the operator based on system parameters. Using the
-eigenvalue spectra, we then build relationships to characterize the
-transport of stochastic histories in a decomposed domain and the
-fraction of histories that leak from a domain and will therefore have
-to be communicated. Finally, we compare these analytic results to
-numerical experiments conducted with the model problem.
+discretization presented in
+Appendix~\ref{chap:diffusion_problem}. Using the linear system
+generated by the discretization of the model problem, we use a
+spectral analysis to generate analytic relations for the eigenvalues
+of the operator based on system parameters. Using the eigenvalue
+spectra, we then build relationships to characterize the transport of
+stochastic histories in a decomposed domain and the fraction of
+histories that leak from a domain and will therefore have to be
+communicated. Finally, we compare these analytic results to numerical
+experiments conducted with the model problem.
 
-\subsubsection{Spectral Analysis}
+\subsection{Spectral Analysis}
 \label{subsec:spectral_analysis}
@@
 
-\subsubsection{Iteration Matrix Spectrum}
-\label{subsec:iteration_spectrum}
 For the model problem, we first compute the eigenvalues for the
@@
 As Eq~(\ref{eq:diffusion_eq}) is diagonally dominant, point Jacobi
-preconditioning as outlined in \S\ref{sec:stochastic_preconditioning}
-is sufficient to reduce the spectral radius of the iteration matrix
-below unity and therefore ensure convergence of the Neumann
-series. Applying this preconditioner, we are then solving the
-following diffusion system:
+preconditioning as outlined in
+\S~\ref{subsec:stochastic_preconditioning} is sufficient to reduce the
+spectral radius of the iteration matrix below unity and therefore
+ensure convergence of the Neumann series. Applying this
+preconditioner, we are then solving the following diffusion system:
 \begin{equation}
@@
 
-\subsubsection{Neumann Series Convergence}
+\subsection{Neumann Series Convergence}
 \label{subsec:neumann_convergence}
@@
 
-\subsubsection{Domain Leakage Approximations}
-\label{subsubsec:domain_leak_approx}
+\subsection{Domain Leakage Approximations}
+\label{subsec:domain_leak_approx}
 In a domain decomposed situation, not all histories will remain within
@@
 
-\subsubsection{Numerical Experiments}
+\subsection{Numerical Experiments}
 \label{subsec:numerical_experiments}
@@
 
-\subsection{Communication Costs for Symmetric Systems}
-\label{sec:symmetric_comm_costs}
-Use Seigel's work to get communication costs.
+%%---------------------------------------------------------------------------%%
+\section{Fully Asynchronous Neumann-Ulam Algorithm}
+\label{sec:asynchronous_algorithm}
 
-%%---------------------------------------------------------------------------%%
-\section{Fully Asynchronous MSOD Neumann-Ulam Algorithm}
-\label{sec:asynchronous_algorithm}
-We can take much of what was learned from Brunner and Wagner's
-parallel Monte Carlo methods for radiation transport and directly
-apply it to a parallel formulation of our stochastic linear
-solvers. Direct analogs can be derived from these works by noting that
-the primary difference between solving a linear system with Monte
-Carlo methods and fixed source Monte Carlo transport problems is the
-content of the Markov chains that are generated. The transitions
-represented by these chains are bound by probabilities and weights and
-are initiated by the sampling of a source. In the context of transport
-problems, those transitions represent events such as particle
-scattering and absorption with probabilities that are determined by
-physical data in the form of cross sections. For stochastic matrix
-inversion, those transitions represent moving between the equations of
-the linear system (and therefore the physical domain which they
-represent) and their probabilities are defined by the coefficients of
-those equations. Ultimately, we tally the contributions to generate
+In the context of radiation transport, in 2009 Brunner and colleagues
+provided a fully asynchronous domain decomposed parallel aglorithm as
+implemented in production implicit Monte Carlo codes
+\citep{brunner_efficient_2009}. In their work they identify two data
+sets that are required to be communicated: the sharing of particles
+that are transported from one domain to another and therefore from one
+processor to another and a global communication that signals if
+particle transport has been completed on all processors. Although the
+implementation was robust and allowed for scaling to large numbers of
+processors, performance issues were still noted with parallel
+efficiency improvements needed in both the weak and strong scaling
+cases for unbalanced problems. These results led Brunner to conclude
+that a combination of domain decomposition and domain replication
+could be used to solve some of these issues.
+
+We can take much of what was learned from Brunner's parallel Monte
+Carlo method for radiation transport and directly apply it to a
+parallel formulation of the Neumann-Ulam method. Direct analogs can be
+derived from these works by noting that the primary difference between
+solving a linear system with Monte Carlo methods and fixed source
+Monte Carlo transport problems is the content of the Markov chains
+that are generated. The transitions represented by these chains are
+bound by probabilities and weights and are initiated by the sampling
+of a source. In the context of transport problems, those transitions
+represent events such as particle scattering and absorption with
+probabilities that are determined by physical data in the form of
+cross sections. For stochastic matrix inversion, those transitions
+represent moving between the equations of the linear system (and
+therefore the physical domain which they represent) and their
+probabilities are defined by the coefficients of those
+equations. Ultimately, we tally the contributions to generate
 expectation values in the desired states as we progress through the
@@
 
-\subsection{Parallel Random Number Generation}
-\label{subsec:parallel_rng}
-Correct parallel Monte Carlo requires parallel random number
-generation. To ensure distinct Markov chains for each history in the
-problem, regardless of its state in phase space or parallel space, a
-unique stream of random numbers is necessary to achieve uncorrelated
-tally results for the solution.
-
-\subsection{Generating the Transport Domain with Overlap and Preconditioning}
+\subsection{Generating the Transport Domain with Preconditioning}
 \label{subsec:domain_generation}
@@
   \end{algorithmic}
+  \label{alg:transition_process}
 \end{algorithm}
@@
 
-\subsection{Multiple Set Algorithm}
-\label{subsec:multiple_sets_algorithm}
+%%---------------------------------------------------------------------------%%
+\subsection{Multiple-Set Overlapping-Domain Neumann-Ulam Algorithm}
+\label{subsec:msod}
+In 2010, Wagner and colleagues developed the \textit{multiple-set
+  overlapping-domain} (MSOD) decomposition for parallel Monte Carlo
+applications for full-core light water reactor analysis
+\citep{wagner_hybrid_2010}. In their work, an extension of Brunner's,
+their scheme employed similar parallel algorithms for particle
+transport but a certain amount of overlap between adjacent domains was
+used to decrease the number of particles leaving the local domain. In
+addition, Wagner utilized a level of replication of the domain such
+that the domain was only decomposed on $O(100)$ processors and if
+replicated $O(1,000)$ times achieves simulation on $O(100,000)$
+processors, thus providing spatial and particle parallelism. Each
+collection of processors that constitutes a representation of the
+entire domain is referred to as a set, and within a set overlap occurs
+among its sub-domains. The original motivation was to decompose the
+domain in a way that it remained in a physical cabinet in a large
+distributed machine, thus reducing latency costs during
+communication. A multiple set scheme is also motivated by the fact
+that communication during particle transport only occurs within a set,
+limiting communications during the transport procedure to a group of
+$O(100)$ processors, a number that was shown to have excellent
+parallel efficiencies in Brunner's work and therefore will scale well
+in this algorithm. The overlapping domains within each set also
+demonstrated reduced communication costs. On each processor, the
+source is sampled in the local domain that would exist if no overlap
+was used while tallies can be made over the entire overlapping domain.
+
+To demonstrate this, consider the example adapted from Mervin's work
+with Wagner and others in the same area \citep{mervin_variance_2012}
+and presented in Figure~\ref{fig:msod_example}.
+\begin{figure}[t!]
+  \begin{center}
+    \scalebox{1.5}{
+      \input{chapters/parallel_mc/msod_example.pdftex_t} }
+  \end{center}
+  \caption{\textbf{Overlapping domain example illustrating how domain
+      overlap can reduce communication costs.}
+    \textit{All particles start in the blue region of interest. The
+      dashed line represents 0.5 domain overlap between domains.}}
+  \label{fig:msod_example}
+\end{figure}
+In this example, 3 particle histories are presented emanating from the
+blue region of interest. Starting with particle A, if no domain
+overlap is used then the only the blue domain exists on the starting
+processor. Particle A is then transported through 3 other domains
+before the history ends, therefore requiring three communications to
+occur in Brunner's algorithm. If a 0.5 domain overlap is permitted as
+shown by the dashed line, then the starting process owns enough of the
+domain such that no communications must occur in order to complete the
+particle A transport process. Using 0.5 domain overlap also easily
+eliminates cases such as the represented by the path of particle C. In
+this case, particle C is scattering between two adjacent domains,
+incurring a large latency cost for a single particle. Finally, with
+particle B we observe that 0.5 domain overlap will still not eliminate
+all communications. However, if 1 domain overlap were used, the entire
+geometry shown in Figure~\ref{fig:msod_example} would be contained on
+the source processor and therefore transport of all 3 particles
+without communication would occur.
+
+Wagner and colleagues used this methodology for a 2-dimensional
+calculation of a pressurized water reactor core and varied the domain
+overlap from 0 to 3 domain overlap (a $7 \times 7$ box in the context
+of our example) where a domain constituted a fuel assembly. For the
+fully domain decomposed case, they observed that 76.12\% of all source
+particles leave the domain. At 1.5 domain overlap, the percentage of
+source particles born in the center assembly leaving the processor
+domain dropped to 1.05\% and even further for 0.02\% for the 3 domain
+overlap. Based on these results, this overlap approach, coupled with
+the multiple sets paradigm that will scale for existing parallel
+transport algorithms, provides a scalable Monte Carlo algorithm for
+today's modern machines.
+
+\subsection{Generating Overlap}
+\label{subsec:msod_overlap}
+
+\subsection{Generating Multiple Sets}
+\label{subsec:msod_sets}
+
+\subsection{MSOD Neumann Ulam Algorithm}
+\label{subsec:msod_algorithm}
 
@@
 
-\subsection{Multiple Set Domain Decomposition}
-\label{subsec:multiple_set_decomposition}
-
-\subsubsection{Weak Scaling}
-\label{subsubsec:multiple_set_weak}
-
-\subsubsection{Strong Scaling}
-\label{subsubsec:multiple_set_strong}
-
 \subsection{MSOD Decomposition}

### 2013-06-11 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding some example figures for domain decomposed monte carlo

@@
 
-The neutron transport problem is complicated. Solutions cover a large
-phase space and the problems of interest are often geometrically
-complex, very large, or both, requiring tremendous computational
-resources to generate an adequate solution. Modern deterministic
-methods for large scale problems are commonly variants on the discrete
-ordinates ($S_N$) method \citep{evans_denovo:_2010}. For fission
-reactor neutronics simulations, the $S_N$ method requires potentially
-trillions of unknown angular flux moments to be computed to achieve
-good accuracy for the responses of interest
-\citep{slaybaugh_acceleration_2011}. Other forms of the transport
-problem, including the $P_N$ method, take on a simpler form than the
-more common $S_N$ methods but lack in accuracy when compared while
-still requiring considerable computational resources for solutions in
-multiple dimensions.
-
-In the 1960's, Gelbard developed an ad-hoc multidimensional extension
-of the simple single dimension planar $P_N$ equations that created a
-system of coupled, diffusion-like equations known as the simplified
-$P_N$ ($SP_N$) equations. Up until around the 1990's, the $SP_N$
-method was either widely unknown, widely unused, or combination of
-both even though numerical studies showed promising results with
-better solutions than diffusion theory and a significant reduction in
-computational time over more accurate methods such as discrete
-ordinates. Why did this happen? A significant problem, pointed out by
-Brantley and Larsen \citep{brantley_simplified_2000}, was that little
-rigor had been applied to the formulation of the $SP_N$ equations
-since their derivation through primarily heuristic arguments. Instead,
-studies at that time focused on simply comparing the results of the
-method to other contemporary transport solution strategies. In
-addition, many problems of interest from the literature at the time
-were either solved using nodal-type methods for reactor-sized problems
-or $S_N$-type methods for benchmark problems with intricate material
-configurations and potentially large flux gradients over small spatial
-domains.
-
-So why reconsider the $SP_N$ equations? Starting in the 1990's and
-primarily due to Larsen and his colleagues, the $SP_N$ equations have
-been given a more rigorous treatment with both variational and
-asymptotic derivations performed as a means of verification. In
-addition, these equations have been more rigorously studied as
-solution methods to MOX fuel problems and have been shown to provide
-accurate solutions. With this mathematical literature to provide a
-solid numerical footing for the method, we look at its application to
-today's challenge problems in neutron transport for fission reactor
-analysis. The reduction in numerical complexity of current
-deterministic solution methods using the $S_N$ approximation could
-mean significant savings in both compute time and memory required. In
-addition, the characteristics of the solution to the transport problem
-for a steady state reactor core permit diffusion theory to be used; a
-staple of the nuclear industry since its inception. Therefore, if
-diffusion theory is applicable, then finer grained solutions that
-capture more of the physics contained in the transport equation should
-be possible with the $SP_N$ method. In doing so, we also expect from
-the literature to obtain computed responses on the order of accuracy
-we would expect from an appropriately discretized $S_N$ method at a
-fraction of the cost.
-
-To further motivate moving in this direction, recent developments in
-the Exnihilo neutronics package at Oak Ridge National Laboratory have
-permitted generation of the $SP_N$ system of equations for detailed
-neutronics models \citep{evans_simpli_2013}. By fully forming these
-equations and formulating them as a linear algebra problem instead of
-using the explicit iterative methods of the past, we now have access
-to all of the modern advancements in computational linear algebra
-including Krylov solvers for asymmetric systems and preconditioning
-methods such as algebraic multigrid. This leads us to then explore the
-applicability of our work in discrete Monte Carlo methods for linear
-systems as a possible solution method for the $SP_N$ equations. If
-formulated correctly, we hypothesize that significant improvement in
-usage of computational resources may be observed compared to modern
-solution techniques such as those suggested due to the form of the
-matrices generated by the discretization. In addition, solving the
-$SP_N$ equations in this way also breaks away from the $S_N$ forms of
-parallelism where spatial parallelism is achieved by an efficient
-parallel sweep, angular efficiency achieved by pipe-lining, and energy
-parallelism achieved by decoupling the groups. With the $SP_N$
-equations as a full matrix system, we now can parallelize the problem
-as prescribed by the linear solver, which may be significantly more
-scalable than current $S_N$ transport practices.
-
-In this chapter, we derive the $SP_N$ equations, closely following the
-work of Evans\footnote{The $SP_N$ derivations in this chapter are
-  heavily based on those presented by Evans in
-  \citep{evans_simpli_2013}}, in order to gain full understanding of
-the underlying system and its potential behavior in a Monte Carlo
-context. From the $P_N$ equations, we apply a set of approximations to
-yield the $SP_N$ equations for fixed source and criticality
-problems. Using the fully-formed linear operator for the transport
-problem, we explore solutions to the $SP_N$ equations using Monte
-Carlo Synthetic Acceleration using a light water reactor fuel assembly
-criticality calculation as the driving problem. In order to
-effectively solve this problem, a suite of preconditioners is
-developed and studied within the context of MCSA. Using these
-preconditioners, MCSA solutions for the fuel assembly problem are
-verified by comparing against production Krylov solvers and finally
-performance is analyzed through comparison with those same Krylov
-solvers.
+In this chapter, we briefly derive the simplified $P_N$ ($SP_N$)
+equations, closely following the work of Evans
+\citep{evans_simpli_2013}\footnote{The $SP_N$ derivations in this
+  chapter and the associated appendices are heavily based on those
+  presented by Evans in \citep{evans_simpli_2013}}, in order to gain
+full understanding of the underlying system and its potential behavior
+in a Monte Carlo context. From the $P_N$ equations, we apply a set of
+approximations to yield the $SP_N$ equations for fixed source and
+criticality problems. Using the fully-formed linear operator for the
+transport problem, we explore solutions to the $SP_N$ equations with
+Monte Carlo Synthetic Acceleration using a light water reactor fuel
+assembly criticality calculation as the driving problem. Several
+difficulties arise when applying MCSA to these problems that were not
+observed when investigating the simpler transport systems.
+
+For Jacobi-based preconditioners, convergence with MCSA is difficult
+if not impossible for ill-conditioned systems with this behavior
+demonstrated using a simple neutron diffusion problem. In order to
+effectively solve the $SP_N$ equations for the fuel assembly problem,
+a suite of preconditioners along with a relaxation scheme is developed
+and studied within the context of MCSA. Using these preconditioners,
+several additional issues are observed and alleviated to a certain
+extent by applying the reduced domain approximation. MCSA solutions
+for the fuel assembly problem are then verified by comparing the
+solutions against production Krylov solvers for problems with varying
+energy groups. Finally performance is analyzed through comparison with
+those same Krylov solvers in terms of both iterative performance and
+CPU timing using the same set of problems utilized for the
+verification.
 
@@
     u_3 = 5\phi_4 + 6\phi_6 \\
-    u_4 = 7\phi_6 \:,
+    u_4 = 7\phi_6 \:.
   \end{gather}
   \label{eq:spn7_subs}
-\end{subequations}
-or equivalently:
-\begin{subequations}
-  \begin{gather}
-    \phi_0 = u_1 - \frac{2}{3}u_2 + \frac{8}{15}u_3 -
-    \frac{16}{35}u_4 \\
-    \phi_2 = \frac{1}{3}u_2 - \frac{4}{15}u_3 + \frac{8}{35}u_4\\ 
-    \phi_4 = \frac{1}{5}u_3 - \frac{6}{35}u_4\\
-    \phi_6 = \frac{1}{7}u_4\:.
-  \end{gather}
-  \label{eq:spn7_subs_inverse}
 \end{subequations}
@@
 For criticality problems, the space, angle, and energy discretizations
-for the $SP_N$ equations can be applied to the general eigenvalue from
+for the $SP_N$ equations can be applied to the general eigenvalue form
 of the transport equation given by
@@
 Eq~(\ref{eq:fission_matrix}). We again apply the change of variables
-to yield the set of pseudo-moments $\mathbb{U}_n$ where now the
-application of the fission matrix to the moment vectors will be
+to yield the set of multigroup pseudo-moments $\mathbb{U}_n$ where now
+the application of the fission matrix to the moment vectors will be
 expanded into a set of block matrices in identical fashion to the
-scattering matrices.
-
+scattering matrices:
+\begin{equation}
+  -\nabla \cdot \mathbb{D}_n \nabla \mathbb{U}_n + \sum_{m=1}^4
+  \mathbb{A}_{nm} \mathbb{U}_m = \frac{1}{k} \sum_{m=1}^4
+  \mathbb{F}_{nm} \mathbb{U}_n\ \ \ \ \ \ \ n = 1,2,3,4\:.
+  \label{eq:spn_fission_matrix}
+\end{equation}
 It should be noted here that with respect to a general MCSA scheme,
@@
 both ensure and improve convergence for Monte Carlo methods.
+
+\clearpage
 
@@
 Figure~\ref{fig:problem3_radial_mat}) is resolved by a $2 \times 2$
-mesh with materials and cross sections homgenized over this region.
+mesh with materials and cross sections homogenized over this region.
 \begin{figure}[t!]
@@
 Poor converge is observed for the Richardson iteration, however,
-convergence is achieved meaning that the preliminary criteria needed
-to satisfy MCSA convergence has also been met. The number of
-iterations required for the Richardson iteration to converge will give
-an approximation for the spectral radius via
+convergence is achieved meaning that the preliminary eigenvalue
+criteria needed to satisfy MCSA convergence has also been met. The
+number of iterations required for the Richardson iteration to converge
+will give an approximation for the spectral radius via
 Eq~(\ref{eq:linear_k_iter_norm3}). With 7291 iterations required to
@@
 system. To study the breakdown of MCSA at iteration matrix spectral
-radii near one, we will use a simpler homogeneous 2-dimensional
-one-speed neutron diffusion system to isolate this behavior. In this
-system, we can vary the cross sections while maintaining a fixed grid
-in order to achieve varying spectral radii. For these studies, we
+radii near one, we will use the simpler homogeneous 2-dimensional
+one-speed neutron diffusion system presented in
+Appendix~\ref{chap:diffusion_problem} to isolate this behavior. In
+this system, we can vary the cross sections while maintaining a fixed
+grid in order to achieve varying spectral radii. For these studies, we
 neglect fission as MCSA behavior is dictated by the transport operator
@@
 
-\subsubsection{Model Diffusion Problem}
-\label{subsec:model_problem}
-For our breakdown study, we choose the one-speed, two-dimensional
-neutron diffusion equation as a model problem
-\citep{duderstadt_nuclear_1976}:
-\begin{equation}
-  -\boldsymbol{\nabla} \cdot D \boldsymbol{\nabla} \phi + \Sigma_a
-  \phi = S\:,
-  \label{eq:diffusion_eq}
-\end{equation}
-where $\phi$ is the neutron flux, $\Sigma_a$ is the absorption cross
-section, and $S$ is the source of neutrons. In addition, $D$ is the
-diffusion coefficient defined as:
-\begin{equation}
-  D = \frac{1}{3 ( \Sigma_t - \bar{\mu}\Sigma_s )}\:,
-  \label{eq:diffusion_coeff}
-\end{equation}
-where $\Sigma_s$ is the scattering cross section, $\Sigma_t = \Sigma_a
-+ \Sigma_s$ is the total cross section, and $\bar{\mu}$ is the cosine
-of the average scattering angle. For simplicity, we will take
-$\bar{\mu} = 0$ for our analysis giving $D=(3 \Sigma_t)^{-1}$. In
-addition, to further simplify we will assume a homogeneous domain such
-that the cross sections remain constant throughout. Doing this permits
-us to rewrite Eq~(\ref{eq:diffusion_eq}) as:
-\begin{equation}
-  -D \boldsymbol{\nabla}^2 \phi + \Sigma_a \phi = S\:.
-  \label{eq:diffusion_eq_simple}
-\end{equation}
-
-We choose a finite difference scheme on a square Cartesian grid to
-discretize the problem. For the Laplacian, we choose the 9-point
-stencil shown in Figure~\ref{fig:stencil} over a grid of size $h$
-\citep{leveque_finite_2007}:
-\begin{multline}
-  \nabla^2_9\phi = \frac{1}{6h^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j}
-    + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ +
-    \phi_{i-1,j+1} + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20
-    \phi_{i,j}]\:.
-  \label{eq:nine_point_stencil}
-\end{multline}
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.25}{\input{chapters/spn_equations/stencil.pdftex_t}}
-  \end{center}
-  \caption{\textbf{Nine-point Laplacian stencil.}}
-  \label{fig:stencil}
-\end{figure}
-We then have the following linear system to solve:
-\begin{multline}
-  -\frac{1}{6h^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j} + 4
-    \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ + \phi_{i-1,j+1}
-    + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20 \phi_{i,j}] + \Sigma_a
-  \phi_{i,j} = s_{i,j}\:,
-  \label{eq:fd_system}
-\end{multline}
-and in operator form:
-\begin{equation}
-  \ve{D}\boldsymbol{\phi}=\ve{s}\:,
-  \label{eq:operator_system}
-\end{equation}
-where $\ve{D}$ is the diffusion operator, $\ve{s}$ is the source in
-vector form and $\boldsymbol{\phi}$ is the vector of unknown fluxes.
-
-To close the system, a set of boundary conditions is required. In the
-case of a non-reentrant current condition applied to all global
-boundaries of the domain, we choose the formulation of Duderstadt by
-assuming the flux is zero at some ghost point beyond the
-grid. Consider for example the equations on the $i=0$ boundary of the
-domain:
-\begin{multline}
-  -\frac{1}{6h^2}[4 \phi_{-1,j} + 4 \phi_{1,j} + 4 \phi_{0,j-1} +
-    4 \phi_{0,j+1} + \phi_{-1,j-1}\\ + \phi_{-1,j+1} + \phi_{1,j-1} +
-    \phi_{1,j+1} - 20 \phi_{0,j}] + \Sigma_a \phi_{0,j} = s_{0,j}\:.
-  \label{eq:x_min_bnd}
-\end{multline}
-Here we note some terms where $i=-1$ and therefore are representative
-of grid points beyond the boundary of the domain. We set the flux at
-these points to be zero, giving a valid set of equations for the $i=0$
-boundary:
-\begin{multline}
-  -\frac{1}{6h^2}[4 \phi_{1,j} + 4 \phi_{0,j-1} + 4 \phi_{0,j+1}
-    \\ + \phi_{-1,j+1} + \phi_{1,j-1} + \phi_{1,j+1} - 20 \phi_{0,j}]
-  + \Sigma_a \phi_{0,j} = s_{0,j}\:.
-  \label{eq:x_min_bnd_2}
-\end{multline}
-We repeat this procedure for the other boundaries of the domain. For
-reflecting boundary conditions, the net current across a boundary is
-zero.
-
-\subsubsection{MCSA Breakdown Analysis}
-\label{subsubsec:breakdown_analysis}
 For each solver and estimator combination, the spectral radius of the
@@
 In addition to the significantly larger number of histories required
-for to achieve convergence for ill-conditioned problems another
-penalty is paid due to histories that take longer to
+to achieve convergence for ill-conditioned problems another penalty is
+paid due to histories that take longer to
 compute. Figure~\ref{fig:breakdown_time} gives the CPU time in seconds
@@
 them and a substantial overhead in the sampling procedure during
-transport. In addition, the matrix-matrix multiply operations required
-to build the composite operator see significant performance losses due
-to the amount of data that must be handled. In parallel, losing
-sparsity severely inhibits performance where now parallel operations
-require communications amongst orders of magnitude more processors for
+transport. Second, the matrix-matrix multiply operations required to
+build the composite operator see significant performance losses due to
+the amount of data that must be handled. In parallel, losing sparsity
+severely inhibits performance where now parallel operations require
+communications amongst orders of magnitude more processors for
 nearest-neighbor type algorithms.
@@
 hand side of Eq~(\ref{eq:frobenius_norm_min}), the Frobenius norm is
-represented as a effective sum of $N$ linear system residual norms. A
-few iterations of a subspace method or other iterative scheme can be
+represented as a effective sum of $N$ linear system residual norms.
+
+A few iterations of a subspace method or other iterative scheme can be
 used to effectively solve these systems to a loose tolerance and yield
@@
 Figure~\ref{fig:spainv_size} the maximum number of non-zero entries
-per row obeserved in the composite linear operator as a function the
+per row observed in the composite linear operator as a function the
 SPAINV threshold. For this analysis, the number of levels in the
@@
 operator. SPAINV parameters for the data presented here were fixed as
-it was found that the preconditioning quality was largely insenstive
+it was found that the preconditioning quality was largely insensitive
 to their variation. In addition, the fact that SPAINV is being used to
@@
 non-zero entries used to compute the metric is equivalent to the
-reduced domain fill level. We first note the SPAINV preconditioning
-alone is signficantly more sensitive to the reduction in domain size
+reduced domain fill level. We first note that SPAINV preconditioning
+alone is significantly more sensitive to the reduction in domain size
 over the ILUT-based methods, although the preconditioner was of
@@
 \label{subsec:spn_mcsa_relaxation}
-The Richardson iteration upon which the Neumann-Ulam method is built
-may be implemented with a scalar relaxation parameter that can be
-adjusted to improve convergence:
+As another means of preconditioning (or variance reduction), the
+Richardson iteration upon which the Neumann-Ulam method is built may
+be implemented with a scalar relaxation parameter that can be adjusted
+to improve convergence:
 \begin{equation}
@@
 \end{equation}
-where $\omega$ is the relaxation parameter. This is effectively a form
-of preconditioning where the system is being scaled on the left by a
-constant value in all rows. Analogously, these relaxation parameter
-techniques can be applied to Monte Carlo to improve convergence as
-demonstrated by Dimov \citep{dimov_new_1998}. In this case, building
-an iteration matrix from Eq~(\ref{eq:richardson_relaxation}) gives:
+where $\omega$ is the relaxation parameter. This is very similar to
+point Jacobi preconditioning where the system is being scaled on the
+left by a constant value in all rows. Analogously, these relaxation
+parameter techniques can be applied to Monte Carlo to improve
+convergence as demonstrated by Dimov \citep{dimov_new_1998}. In this
+case, building an iteration matrix from
+Eq~(\ref{eq:richardson_relaxation}) gives:
 \begin{equation}
@@
 
-For this study, the fuel assemby problem with a 1-group $SP_1$
+For this study, the fuel assembly problem with a 1-group $SP_1$
 discretization was solved using MCSA with the adjoint Monte Carlo
@@
 MCSA agree with the results from the production Krylov solvers within
-the solution tolerance then the MCSA solutions will be deemed verified
-for these problems. Additionaly, this verification is only applicable
-to serial results with parallel verification for this problem provided
-in Chapter~\ref{ch:parallel_methods}.
+the solution tolerance, then the MCSA solutions will be deemed correct
+for these problems. Additionally, this verification is only applicable
+to serial results with parallel MCSA verification for this problem
+provided in Chapter~\ref{ch:parallel_methods}.
 
@@
 approximately 1 stochastic history for every 5 DOFs. The reduced
-domain approximation was also applied in conjuction with the ILUT
+domain approximation was also applied in conjunction with the ILUT
 preconditioning using a fill level of 100 and a threshold value of
 \sn{1}{-10} to reduce the density of states in the Monte Carlo
-problem. For relaxation parameters, all MCSA compuations used a
+problem. For relaxation parameters, all MCSA computations used a
 Neumann relaxation parameter of 0.7 while a Richardson relaxation of
@@
       assembly criticality problem in electron volts.} \textit{An
-      implicit upper bound of \sn{2}{6} eV is assumedfor group zero.}}
+      implicit upper bound of \sn{2}{6} eV is assumed for group zero.}}
   \label{tab:spn_group_structure}
@@
 For each converged eigenvalue calculation, the scalar flux in each
-energy group was compared over the entire spatial with the
+energy group was compared over the entire spatial domain with the
 BiCGStab-ILUT solution used as the reference to compute an absolute
-difference. For the 1 group calculations, these absolute differences
-are plotted spatially along the centerline of the fuel assembly in
+difference of the flux in each mesh cell. For the 1 group
+calculations, these absolute differences are plotted spatially along
+the center-line of the fuel assembly in
 Figure~\ref{fig:serial_1_group_diff}. Interestingly, except for the
@@
 GMRES had a maximum global flux difference of \sn{9.313}{-10} even
-though different spatial results were observed. For the MCSA-ILUT-MR-C
-calculation, a slightly larger maximum error of \sn{2.205}{-9} was
-observed for reasons that remain unexplained. All 1-group calculations
-have a minimum observed absolute difference of 0 with respect to the
-BiCGStab calculation.
+though different spatial results were observed for all cases. For the
+MCSA-ILUT-MR-C calculation, a slightly larger maximum error of
+\sn{2.205}{-9} was observed for reasons that remain unexplained. All
+1-group calculations have a minimum observed absolute difference of 0
+with respect to the BiCGStab calculation.
 
@@
       between the Aztec BiCGStab calculations and Aztec GMRES and the
-      MCSA variations for the 1-group calculation.} \textit{Left to
-      right: GMRES-ILUT, MCSA-ILUT-R-C, MCSA-ILUT-MR-C,
-      MCSA-ILUT-R-EV, MCSA-ILUT-MR-EV. Table~\ref{tab:spn_solver_defs}
-      gives the description for each solver type presented.}}
+      MCSA variations for the 1-group fuel assembly criticality
+      calculation.} \textit{Left to right: GMRES-ILUT, MCSA-ILUT-R-C,
+      MCSA-ILUT-MR-C, MCSA-ILUT-R-EV, MCSA-ILUT-MR-EV. The fluxes are
+      plotted on a plane intersecting the center-line of the
+      assembly. Table~\ref{tab:spn_solver_defs} gives the description
+      for each solver type presented.}}
   \label{fig:serial_1_group_diff}
@@
 Table~\ref{tab:serial_differences_g4} give the maximum and minimum
-absolute differences for all calculations and all energy groups using
-the BiCGStab calculation as the reference. All comparisons were
-performed within a group such that the MCSA-ILUT-R-C results in group
-3 for the 4-group calculation were computed with the BiCGStab-ILUT
-solution in group 3 for the 4-group calculation as a reference. For
-all group flux results given in the table, the maximum differences for
-all cases was less than the linear solver tolerance used to converge
-the flux at each eigenvalue iteration meaning that all values over the
-entire energy and spatial domain are within these limits. In addition,
-the minimum error for all calculations was observed to be zero,
-meaning in at least one mesh cell in all calculations for all energy
-groups the scalar flux values computed by all solvers were
-identical. Based on the flux results and the eigenvalue results
-presented here, the MCSA results should be deemed correct for the fuel
-assembly problem for serial calculations.
+absolute differences of the cell scalar fluxes for all calculations
+and all energy groups using the BiCGStab calculation as the
+reference. All comparisons were performed within a group such that the
+differences for the MCSA-ILUT-R-C results in group 3 for the 4-group
+calculation were computed with the BiCGStab-ILUT solution in group 3
+for the 4-group calculation as a reference. For all group flux results
+given in the table, the maximum differences for all cases was less
+than the linear solver tolerance used to converge the flux at each
+eigenvalue iteration meaning that all values over the entire energy
+and spatial domain are within these limits. In addition, the minimum
+error for all calculations was observed to be zero, meaning that at
+least one mesh cell in all calculations for all energy groups the
+scalar flux values computed by all solvers were identical. Based on
+the flux results and the eigenvalue results presented here, the MCSA
+results should be deemed correct for the fuel assembly problem for
+serial calculations with multiple energy groups.
 
@@
 In the previous section, numerical results using MCSA for the fuel
-assembly calculation were directly with production Krylov methods in
-order to verify their correctness. Deemed correct, with the same set
-of solver parameters used for verification MCSA will now be directly
-compared to the same production Krylov solvers in terms of both
-iterative performance and CPU timing for solutions to the fuel
-assembly problem.  This comparison of both iterative and CPU timing
-performance is necessary in order to gauge feasibility of the method
-and identify areas that require further research and development.
+assembly calculation were directly compared with production Krylov
+methods in order to verify their correctness. Deemed correct,
+performance of the algorithm for serial computations will now be
+addressed. With the same set of solver parameters used for
+verification, MCSA will now be directly compared to the same
+production Krylov solvers in terms of both iterative performance and
+CPU timing for solutions to the fuel assembly problem.  This
+performance comparison is necessary in order to assess the feasibility
+of the method and identify areas that require further research and
+development.
 
@@
 solver is not a strong function of the number of energy groups in the
-problem (and therefore the problem size). For MCSA, we are linearly
-increasing the number of stochastic histories used to compute the
-correction at each iteration as the number of energy groups are
-increased and therefore the work performed for the problem is of
-$O(N)$. It is also important to note the reduced domain approximation
-parameters, particularly the fill level, were fixed as the number of
-energy groups was increased. By fixing the fill level, we are
-effectively fixing the amount of information contained in the
-composite linear operator available for the Monte Carlo
+problem (and therefore the problem size).
+
+For MCSA, we are linearly increasing the number of stochastic
+histories used to compute the correction at each iteration as the
+number of energy groups are increased and therefore the work performed
+for the problem is of $O(N)$. It is also important to note the reduced
+domain approximation parameters, particularly the fill level, were
+fixed as the number of energy groups was increased. By fixing the fill
+level, we are effectively fixing the amount of information contained
+in the composite linear operator available for the Monte Carlo
 problem. Limiting this information to 100 entries per row in the
@@
 observed for production Krylov implementations using exactly the same
-preconditioning scheme, timing performance is not as competetive. For
+preconditioning scheme, timing performance is not as competitive. For
 this comparison, the amount of time to perform each linear solver
@@
 optimized and it is likely that several components of the algorithm
-can be implemented in a more desireable time complexity. The
+can be implemented in a more desirable time complexity. The
 production Krylov solvers presented here have enjoyed decades of
@@
 preconditioning scheme. For transport systems where this is not a
-problem, the literature explicitly shows that MCSA can be competetive
+problem, the literature explicitly shows that MCSA can be competitive
 with these production Krylov methods using CPU time as a measure

### 2013-06-11 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
updating section numbering

@@
 \label{sec:msod}
-As observed in the discussion on parallel Krylov methods in
-Appendix~\ref{ch:linear_problem}, large-scale problems will surely
-have their data partitioned such that each parallel process owns a
-subset of the equations in the linear system. Given this convention,
-the Neumann-Ulam Monte Carlo algorithm must perform random walks over
-a domain that is decomposed and must remain decomposed due to memory
-limitations. This naturally leads us to seek parallel Monte Carlo
-algorithms that handle domain decomposition.
+Large-scale problems will for reasons typically related to memory
+restrictions or performance have their data partitioned such that each
+parallel process owns a subset of the equations in the linear
+system. Given this convention, the Neumann-Ulam Monte Carlo algorithm
+must perform random walks over a domain that is decomposed and must
+remain decomposed to avoid the same performance and memory
+restrictions that required the domain to be decomposed. To parallelize
+this algorithm, we then seek parallel Monte Carlo algorithms that
+handle domain decomposition.
 
+To motivate this problem, consider the square domain presented in
+Figure~\ref{fig:ddmc_example}. Here, we imagine a Monte Carlo particle
+transport problem. If the domain were decomposed into 9 pieces as
+shown by the domain, each of those pieces could be parallel process in
+the computation. In Figure~\ref{fig:ddmc_example}, the tracks of 3
+particles that are born in the center domain are shown. For example,
+particle A is first transported from the center domain to the domain
+directly to the left. Before the scattering event in the new domain
+may be processed, particle A must be communicated between between the
+two parallel processes that own those domains. For any given particle
+in the transport simulation, this communication event may hardly occur
+during the lifetime of the particle (particle B), or may occur many
+times (particle C) depending on the problem parameters. Scalable
+parallel algorithms for domain Monte Carlo are those that handle this
+domain-to-domain communication of particles effectively and balance it
+with the physics computations.
 \begin{figure}[t!]
@@
 \end{figure}
+
+For a domain decomposed Neumann-Ulam method, the situation is very
+much the same where instead the transport process is now discrete and
+the "physics" of the transport process is the Monte Carlo game with
+probabilities and weights described by
+Eq~(\ref{eq:neumann_ulam_decomposition}) for the given linear
+problem. Figure~\ref{fig:ddnu_example} gives the domain decomposed
+Neumann-Ulam analog of particle transport problem in
+Figure~\ref{fig:ddmc_example}. Imagine in this problem that we are
+solving for the monoenergetic scalar neutron flux as in the model
+diffusion problem presented in Appendix~\ref{chap:diffusion_problem}
+such that the solution has only a spatial dependence. In
+Figure~\ref{fig:ddnu_example}, the domain has now been discretized by
+a mesh and again decomposed into 9 pieces. As we play the Monte Carlo
+game presented in Chapter~\ref{ch:stochastic_methods}, the
+discretization and resulting Neumann-Ulam decomposition of the problem
+describes how each state in the system is coupled to the other states
+in the system through the set of equations that forms the linear
+system. As discrete states (or rows in the linear system) are reached
+by stochastic histories that do not exist in the local domain, the
+histories must be computed in an analogous fashion to the particles in
+the previous example. Again consider history A in
+Figure~\ref{fig:ddnu_example}. This history begins in the center
+domain and through the course of transitioning through states in the
+system arrives at a state that exists in the domain directly to the
+left of the center domain. History A must then be communicated from
+the center domain to the domain immediately to the left. As with the
+particle transport system, we may find that histories in a domain
+decomposed Neumann-Ulam method are only communicated a few times
+(history B) if at all, or they may be communicated many times (history
+C) depending on the discretization and the outcome of the Monte Carlo
+game. The amount of communication of histories from domain to domain
+will be the primary factor in the parallel scalability of the
+algorithm. In the next section, the number of histories communicated in a given
+problem will be quantified analytically for a simple model problem.
 
@@
 When solving problems where the linear operator is symmetric, a host
-of analytic theories exist based on the eigenvalue spectrum of the
+of analytic techniques exist based on the eigenvalue spectrum of the
 operator that characterize their behavior in the context of
-deterministic linear solvers. Using past work, these theories are
+deterministic linear solvers. Using past work, these techniques are
 adapted to the domain decomposed Neumann-Ulam method using the
@@
 giving the eigenvalue spectrum for the Jacobi preconditioned iteration
-matrix. To find these maxium eigenvalue,
+matrix. To find the maxium eigenvalue,
 Eq~(\ref{eq:preconditioned_eigenvalues}) is plotted as a function of
@@
 
-In the case of domain decomposed linear operator equations, we can use
-diffusion theory to estimate the optical thickness of a domain in the
+In the case of domain decomposed linear systems, we can use diffusion
+theory to estimate the optical thickness of a domain in the
 decomposition and the corresponding leakage fraction in terms of
@@
 domains will vary as a function of the absorption cross subsection if
-the other parameters are fixed. To compute the optical thickness,
-along with the spectral radius as given by
-Eq~(\ref{eq:iteration_radius}), we also need the parameters $n_i$ and
-$n_s$ which respectively describe the typical domain length and the
-average number of states moved along that typical length per history
-transition. For our grid above, the domains are varied in size with
-$50 \times 50$, $100 \times 100$, and $200 \times 200$ cells giving
-$n_i=50$, $n_i=100$, and $n_i=200$ grid points or states along the
-typical length of the domain respectively. Looking at the Laplacian
-stencil in Eq~(\ref{eq:nine_point_stencil}), we see that all history
-transitions will only move a single state in either the $i$ or $j$
-directions due to the symmetry of the problem. Furthermore, if we
+the other parameters are fixed.
+
+To compute the optical thickness, along with the spectral radius as
+given by Eq~(\ref{eq:iteration_radius}), we also need the parameters
+$n_i$ and $n_s$ which respectively describe the typical domain length
+and the average number of states moved along that typical length per
+history transition. For our grid above, the domains are varied in size
+with $50 \times 50$, $100 \times 100$, and $200 \times 200$ cells
+giving $n_i=50$, $n_i=100$, and $n_i=200$ grid points or states along
+the typical length of the domain respectively. Looking at the
+Laplacian stencil in Eq~(\ref{eq:nine_point_stencil}), we see that all
+history transitions will only move a single state in either the $i$ or
+$j$ directions due to the symmetry of the problem. Furthermore, if we
 choose the $i$ direction, not all states we will transition to will
@@
 analytic quantities but we begin to see the limits of the leakage
-approximations. To compare the quality of the two approximations, the
-absolute difference between the computed leakage fraction and that
-generated by the Wigner rational and mean chord approximations is
-plotted in Figure~\ref{fig:leakage_error} for all domain sizes
-tested. 
+approximations.
+
+To compare the quality of the two approximations, the absolute
+difference between the computed leakage fraction and that generated by
+the Wigner rational and mean chord approximations is plotted in
+Figure~\ref{fig:leakage_error} for all domain sizes tested.
 \begin{figure}[t!]
@@
 
+\clearpage
+
 %%---------------------------------------------------------------------------%%
-\section{Fully Asynchronous Neumann-Ulam Algorithm}
+\section{Domain Decomposed Neumann-Ulam Algorithm}
 \label{sec:asynchronous_algorithm}
-
 In the context of radiation transport, in 2009 Brunner and colleagues
@@
 
+\clearpage
+
 %%---------------------------------------------------------------------------%%
-\subsection{Multiple-Set Overlapping-Domain Neumann-Ulam Algorithm}
+\section{Multiple-Set Overlapping-Domain Neumann-Ulam Algorithm}
 \label{subsec:msod}

### 2013-06-20 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
adding serial performance data

@@
 For their studies, Siegel and colleagues assumed that the value of
-$\Lambda$ was dependent on the total cross subsection of the system via
+$\Lambda$ was dependent on the total cross section of the system via
 the Wigner rational approximation. Outlined more thoroughly by Hwang's
@@
 absorption cross varied from 0 to 100 while the scattering cross
-subsection was fixed at unity. Figure~\ref{fig:measured_spec_rad} gives
+section was fixed at unity. Figure~\ref{fig:measured_spec_rad} gives
 the measured spectral radius of the iteration matrix and the computed
@@
         operator spectral radius as a function of the absorption cross
-        subsection to scattering cross subsection ratio.} \textit{Values of
+        section to scattering cross section ratio.} \textit{Values of
         $h=0.01$, $h=0.1$, and $h=1.0$ were used. The red data was
@@
 \times 100$ square grid with $h=0.1$ and the absorption cross varied
-from 0 to 100 while the scattering cross subsection was fixed at
+from 0 to 100 while the scattering cross section was fixed at
 unity. Three weight cutoff values of \sn{1}{-2}, \sn{1}{-4}, and
@@
 direction with measurements occurring in the central domain without
-boundary grid points. For the cross subsections, the absorption cross
-subsection was varied from 1 to 100 while the scattering cross
-subsection was set to zero to create a purely absorbing environment
+boundary grid points. For the cross sections, the absorption cross
+section was varied from 1 to 100 while the scattering cross
+section was set to zero to create a purely absorbing environment
 with weight cutoff of \sn{1}{-4}. The optical thickness of these
-domains will vary as a function of the absorption cross subsection if
+domains will vary as a function of the absorption cross section if
 the other parameters are fixed.

### 2013-06-30 doc/dissertation/document/backmatter/diffusion_problem.tex
cleaning up first sections of parallel chapter

@@
 For many numerical experiments in this work, we choose the one-speed,
-two-dimensional neutron diffusion equation as a model transport
-problem \citep{duderstadt_nuclear_1976}:
+two-dimensional neutron diffusion equation as a model neutron
+transport problem \citep{duderstadt_nuclear_1976}:
 \begin{equation}

### 2013-06-30 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
finished adding replication data for titan
cleaning up first sections of parallel chapter

@@
-\chapter{Parallel Monte Carlo Synthetic Acceleration}
+\chapter{Parallel Monte Carlo Synthetic \\ Acceleration Methods}
 \label{ch:parallel_methods}
-For MCSA to be viable at the production scale, scalable parallel
-implementations of the algorithm are required. Reviewing the
-literature, MCSA has yet to be parallelized and the Neumann-Ulam Monte
-Carlo method on which it is built has only been parallelized through
-history-level parallelism with full domain replication. In order to
-solve large linear systems with MCSA, a domain-decomposed parallel
+For MCSA to be viable at the production scale for nuclear engineering
+applications, scalable parallel implementations of the algorithm are
+required. Reviewing the literature, MCSA has yet to be parallelized
+and the Neumann-Ulam Monte Carlo method on which it is built has only
+been parallelized through history-level parallelism with full domain
+replication \citep{alexandrov_efficient_1998}. In order to solve large
+linear transport systems with MCSA, a domain-decomposed parallel
 strategy is required. In the formulation of a parallel MCSA algorithm,
@@
 algebraic quantities from which to estimate its performance when
-applied to linear systems. We then devise a parallel algorithm for the
-Neumann-Ulam method based on the multiple-set overlapping-domain
-decomposition algorithm and a parallel algorithm for the MCSA that
-leverages the parallel Monte Carlo algorithm and general parallel
-matrix-vector operations. In identical fashion to the serial
-computations presented in Chapter~\ref{ch:spn_equations}, we use
-parallel solutions for the $SP_N$ equations for the fuel assembly
-criticality problem to verify the correctness of the parallel
-algorithm and implementation by comparing the numerical results
-against production Krylov methods. Using the verified implementation
-of the algorithm, we perform parallel scaling studings to test its
-performance on both a small computing cluster and a leadership class
-machine and compare this performance to the production Krylov methods.
+applied to neutron transport systems. We then devise a parallel
+algorithm for the Neumann-Ulam method based on the multiple-set
+overlapping-domain decomposition algorithm and a parallel algorithm
+for the MCSA iteration that leverages the parallel Monte Carlo
+algorithm and general parallel matrix-vector operations. In identical
+fashion to the serial computations presented in
+Chapter~\ref{ch:spn_equations}, we use parallel solutions for the
+$SP_N$ equations for the fuel assembly criticality problem to verify
+the correctness of the parallel algorithm and implementation by
+comparing the numerical results against production Krylov
+methods. Using the verified implementation of the algorithm, we
+perform parallel scaling studings to test its performance on a
+leadership class machine and compare this performance to the
+production Krylov methods.
 
@@
 To motivate this problem, consider the square domain presented in
-Figure~\ref{fig:ddmc_example}. Here, we imagine a Monte Carlo particle
-transport problem. If the domain were decomposed into 9 pieces as
-shown by the domain, each of those pieces could be parallel process in
-the computation. In Figure~\ref{fig:ddmc_example}, the tracks of 3
-particles that are born in the center domain are shown. For example,
-particle A is first transported from the center domain to the domain
-directly to the left. Before the scattering event in the new domain
-may be processed, particle A must be communicated between between the
-two parallel processes that own those domains. For any given particle
-in the transport simulation, this communication event may hardly occur
-during the lifetime of the particle (particle B), or may occur many
-times (particle C) depending on the problem parameters. Scalable
-parallel algorithms for domain Monte Carlo are those that handle this
-domain-to-domain communication of particles effectively and balance it
-with the physics computations.
+Figure~\ref{fig:ddmc_example} and on this domain we imagine a Monte
+Carlo particle transport problem. If the domain were decomposed into 9
+subdomains as shown, each of those subdomains and their associated
+data (i.e. cross sections) could be owned by a different parallel
+process in the computation. In Figure~\ref{fig:ddmc_example}, the
+tracks of 3 particles that are born in the center subdomain are
+shown. In this example, particle A is first transported from the
+center domain to the domain directly to the left. Before the
+scattering event in the new domain may be processed, particle A must
+vbe communicated between between the two parallel processes that own
+those domains. For any given particle in the transport simulation,
+this communication event may hardly occur during the lifetime of the
+particle (particle B), or may occur many times (particle C) depending
+on the problem parameters and the random path in phase space taken by
+the particle. Scalable parallel algorithms for domain decomposed Monte
+Carlo are those that handle this domain-to-domain communication of
+particles effectively and balance it with the on-process computations
+for particle interactions, tallies, response calculations, and other
+simulation requirements.
 \begin{figure}[t!]
@@
 solving for the monoenergetic scalar neutron flux as in the model
-diffusion problem presented in Appendix~\ref{chap:diffusion_problem}
-such that the solution has only a spatial dependence. In
-Figure~\ref{fig:ddnu_example}, the domain has now been discretized by
-a mesh and again decomposed into 9 pieces. As we play the Monte Carlo
-game presented in Chapter~\ref{ch:stochastic_methods}, the
+diffusion problem presented in Appendix~\ref{chap:diffusion_problem}.
+In this problem, the solution has only a spatial dependence and
+therefore each point in the mesh corresponds to an equation in the
+resulting linear system. In Figure~\ref{fig:ddnu_example}, the domain
+has now been discretized by this mesh and again decomposed into 9
+subdomains, each owning a piece of the mesh and therefore the
+corresponding equations in the linear system. As we play the Monte
+Carlo game presented in Chapter~\ref{ch:stochastic_methods}, the
 discretization and resulting Neumann-Ulam decomposition of the problem
 describes how each state in the system is coupled to the other states
-in the system through the set of equations that forms the linear
-system. As discrete states (or rows in the linear system) are reached
-by stochastic histories that do not exist in the local domain, the
-histories must be computed in an analogous fashion to the particles in
-the previous example. Again consider history A in
-Figure~\ref{fig:ddnu_example}. This history begins in the center
-domain and through the course of transitioning through states in the
-system arrives at a state that exists in the domain directly to the
-left of the center domain. History A must then be communicated from
-the center domain to the domain immediately to the left. As with the
-particle transport system, we may find that histories in a domain
-decomposed Neumann-Ulam method are only communicated a few times
-(history B) if at all, or they may be communicated many times (history
-C) depending on the discretization and the outcome of the Monte Carlo
-game. The amount of communication of histories from domain to domain
-will be the primary factor in the parallel scalability of the
-algorithm. In the next section, the number of histories communicated in a given
-problem will be quantified analytically for a simple model problem.
-
+in the system through the set of equations that form the linear
+system. This coupling is then responsible for the discrete random walk
+that each history takes as shown in this example. As discrete states
+(or rows in the linear system) that do not exist in the local domain
+are reached during the random walk by stochastic histories, the
+histories must be communicated to the subdomain that does own the
+discrete state in an analogous fashion to the particles in the
+previous example.
 \begin{figure}[t!]
@@
 \end{figure}
+
+To further motivate using a parallel algorithm similar to particle
+transport methods, again consider history A in
+Figure~\ref{fig:ddnu_example}. This history begins in the center
+subdomain and through the course of transitioning through states in
+the system arrives at a state that exists in the subdomain directly to
+the left of the center subdomain. History A must then be communicated
+from the center subdomain to the subdomain immediately to the left to
+continue the random walk. As with the particle transport system, we
+may find that histories in a domain decomposed Neumann-Ulam method are
+only communicated a few times (history B) if at all, or they may be
+communicated many times (history C) depending on the discretization
+and the outcome of the Monte Carlo game. In identical fashion to the
+particle transport problem, a domain decomposed Neumann-Ulam method
+has the same communcation requirements with all histories in both
+examples requiring the same parallel operations. The amount of
+communication of histories from domain to domain will be the primary
+factor in the parallel scalability of the algorithm. In the next
+section, the number of histories communicated in a given problem will
+be quantified analytically for a simple model problem.
 
@@
 algorithmic strong scaling \citep{gropp_high-performance_2001} in much
-the same way as a matrix-vector multiply operation. Therefore, we
-desire a framework to provide a simple, analytic theory based on the
-properties of the linear system that will allow for estimates of the
-domain decomposed behavior of the Neumann-Ulam method in terms of the
-amount of information that must be communicated.
+the same way as a parallel matrix-vector multiply
+operation. Therefore, we desire a framework to provide a simple,
+analytic theory based on the properties of the transport system that
+will allow for estimates of the domain decomposed behavior of the
+Neumann-Ulam method in terms of the amount of information that must be
+communicated.
 
@@
 one-speed, two-dimensional neutron diffusion equation and spatial
-discretization presented in
-Appendix~\ref{chap:diffusion_problem}. Using the linear system
-generated by the discretization of the model problem, we use a
-spectral analysis to generate analytic relations for the eigenvalues
-of the operator based on system parameters. Using the eigenvalue
-spectra, we then build relationships to characterize the transport of
-stochastic histories in a decomposed domain and the fraction of
-histories that leak from a domain and will therefore have to be
-communicated. Finally, we compare these analytic results to numerical
-experiments conducted with the model problem.
+discretization presented in Appendix~\ref{chap:diffusion_problem} as a
+model transport problem. Using the linear system generated by the
+discretization of the model problem, we use a spectral analysis to
+generate analytic relations for the eigenvalues of the operator based
+on system parameters. Using the eigenvalue spectra, we then build
+relationships to characterize the transport of stochastic histories in
+a decomposed domain and the fraction of histories that leak from a
+domain and will therefore have to be communicated. Finally, we compare
+these analytic results to numerical experiments conducted with the
+model transport problem.
+
+\clearpage
 
@@
 eigenmodes of the solution. As these are valid forms of the solution,
-then the action of the linear operator on these eigenfunctions should
-give the eigenvalues of the matrix as they exist on the unit circle in
-the complex plane.
+then the action of the linear operator on these eigenfunctions will
+yield the eigenvalues of the matrix as they exist on the unit circle
+in the complex plane.
 
@@
 \end{equation}
+
+\clearpage
 
@@
 implemented in production implicit Monte Carlo codes
-\citep{brunner_efficient_2009}. In their work they identify two data
-sets that are required to be communicated: the sharing of particles
-that are transported from one domain to another and therefore from one
-processor to another and a global communication that signals if
-particle transport has been completed on all processors. Although the
-implementation was robust and allowed for scaling to large numbers of
-processors, performance issues were still noted with parallel
-efficiency improvements needed in both the weak and strong scaling
-cases for unbalanced problems. These results led Brunner to conclude
-that a combination of domain decomposition and domain replication
-could be used to solve some of these issues.
+\citep{brunner_efficient_2009}. We will adapt their algorithm and
+directly apply it to a parallel formulation of the Neumann-Ulam
+method. Direct analogs can be derived from their works by noting that
+the primary difference between solving a linear system with Monte
+Carlo methods and fixed source Monte Carlo transport problems is the
+content of the Markov chains that are generated. The transitions
+represented by these chains are bound by probabilities and weights and
+are initiated by the sampling of a discrete source. In the context of
+transport problems, those transitions represent events such as
+particle scattering and absorption with probabilities that are
+determined by physical data in the form of cross sections.
 
-We can take much of what was learned from Brunner's parallel Monte
-Carlo method for radiation transport and directly apply it to a
-parallel formulation of the Neumann-Ulam method. Direct analogs can be
-derived from these works by noting that the primary difference between
-solving a linear system with Monte Carlo methods and fixed source
-Monte Carlo transport problems is the content of the Markov chains
-that are generated. The transitions represented by these chains are
-bound by probabilities and weights and are initiated by the sampling
-of a source. In the context of transport problems, those transitions
-represent events such as particle scattering and absorption with
-probabilities that are determined by physical data in the form of
-cross sections. For stochastic matrix inversion, those transitions
-represent moving between the equations of the linear system (and
-therefore the physical domain which they represent) and their
+For stochastic matrix inversion, those transitions represent moving
+between the equations of the linear system (and therefore the
+components of phase space which they represent) and their
 probabilities are defined by the coefficients of those
@@
 to matrix inversion methods as an initial means of developing a
-parallel Neumann-Ulam-type solver. In this section the fully
-asynchronous MSOD Neumann-Ulam algorithm developed for this work is
-presented step-by-step in detail.
+parallel Neumann-Ulam-type solver.
 
-\subsection{Generating the Transport Domain with Preconditioning}
+In this section Brunner and Brantley's fully asynchronous algorithm is
+briefly presented along with its application to the Neumann-Ulam
+method. Their algorithm is presented in detail in the reference and
+was effectively implemented verbatim for this work. In their work they
+identify two data sets that are required to be communicated: the
+sharing of particles that are transported from one domain to another
+and therefore from one processor to another and a global communication
+that signals if particle transport has been completed on all
+processors. Both of these processes will be addressed at a high level
+along with how the parallel Monte Carlo data structures are
+constructed.
+
+\subsection{Parallel Transport Domain and Source}
 \label{subsec:domain_generation}
-
-\subsection{Generating the Transport Source with Preconditioning}
-\label{subsec:source_generation}
-
-\subsection{Transition Processing Kernel}
-\label{subsec:transition_processing}
-Processing the transition of a single stochastic history from one
-state to another in the system is the fundamental kernel of the
-transport procedure. In this kernel, an incoming history has a state
-consisting of the current discrete state in the system in which it
-resides and its current weight and is processed as given by
-Algorithm~\ref{alg:transition_process}.
-\begin{algorithm}[h!]
-  \caption{Stochastic History Transition Processing Kernel}
-  \label{alg:power_iteration}
-  \begin{algorithmic}
-    \State processTransition( history ):
-    \State i = history.state
-    \State j = sampleRowCDF( C[i] )
-    \State history.state = j
-    \State history.weight *= w[i,j]
-  \end{algorithmic}
-  \label{alg:transition_process}
-\end{algorithm}
-The core component here is the sampling of the CDF for a particular
-state generated by the probablities from the Neumann-Ulam
-decomposition. Upon exiting, the kernel has modified the state of the
-history with a new state and multiplied the weight with the
-corresponding transition weight to form the product in
-Eq~(\ref{eq:direct_permutation_weight}).
-
-\subsection{Local Domain Transport Kernel}
-\label{subsec:local_domain_transport}
+To utilize the parallel transport algorithm, we must generate the
+parallel transport domain. For the Neumann-Ulam method, the domain
+consists of all states in the system that are local, and the
+probabilities and weights for all state transitions possible in the
+local domain. At the core of this representation is the Neumann-Ulam
+decomposition of the linear operator. Given an input decomposition for
+the linear operator, by definition the Neumann-Ulam decomposition will
+have the same parallel decomposition. In addition, any modification
+made to the linear operator through preconditioning or relaxation
+parameters will also modify the Neumann-Ulam decomposition. All
+possible transitions are described by the local graph of the sparse
+input matrix. In addition, any left preconditioning or relaxation
+parameters will modify the right hand side of the linear system and
+therefore the fixed source in the Monte Carlo calculation. This source
+vector will have the same parallel decomposition as the input matrix
+and therefore all of the birth states for the histories will exist for
+at least a single transition event within the local domain.
 
@@
 
-\subsection{Asynchronous Transport Kernel}
-\label{subsec:async_transport_kernel}
+\begin{figure}[htpb!]
+  \begin{center}
+    \scalebox{1.5}{ \input{chapters/parallel_mc/domain_to_domain.pdftex_t} }
+  \end{center}
+  \caption{\textbf{Nearest neighbor history communication sequence.}
+    \textit{Each subdomain in the system has a set of nearest
+      neighbors determined by the parallel adjacency graph of the
+      input matrix. The subdomains are indexed by an integer.}}
+  \label{fig:nearest_neighbor_comm}
+\end{figure}
+
+\subsection{Ending the Transport Sequence}
+\label{subsec:ending_transport}
+
+\begin{figure}[htpb!]
+  \begin{center}
+    \scalebox{1.25}{
+      \input{chapters/parallel_mc/binary_comm_tree.pdftex_t} }
+    \caption{\textbf{Binary communication tree for coordinating the
+        end of a parallel Neumann-Ulam solve.} \textit{Each child
+        process reports to its parents how many histories completed
+        within its domain. When the root process sums all complete
+        histories, it forwards the stop signal to its children which
+        in turn forward the message up the tree. The subdomains are
+        indexed by an integer.}}
+  \end{center}
+  \label{fig:binary_comm_tree}
+\end{figure}
 
@@
 \label{subsec:msod}
-In 2010, Wagner and colleagues developed the \textit{multiple-set
+Although the implementation presented in the previous section was
+observed by Brunner and Brantley to be robust and allowed for scaling
+to large numbers of processors, performance issues were still noted
+with parallel efficiency improvements needed in both the weak and
+strong scaling cases for unbalanced problems. These results led them
+to conclude that a combination of domain decomposition and domain
+replication could be used to solve some of these issues. In 2010,
+Wagner and colleagues developed the \textit{multiple-set
   overlapping-domain} (MSOD) decomposition for parallel Monte Carlo
@@
 that the domain was only decomposed on $O(100)$ processors and if
-replicated $O(1,000)$ times achieves simulation on $O(100,000)$
-processors, thus providing spatial and particle parallelism. Each
-collection of processors that constitutes a representation of the
-entire domain is referred to as a set, and within a set overlap occurs
-among its sub-domains. The original motivation was to decompose the
-domain in a way that it remained in a physical cabinet in a large
-distributed machine, thus reducing latency costs during
-communication. A multiple set scheme is also motivated by the fact
-that communication during particle transport only occurs within a set,
-limiting communications during the transport procedure to a group of
-$O(100)$ processors, a number that was shown to have excellent
-parallel efficiencies in Brunner's work and therefore will scale well
-in this algorithm. The overlapping domains within each set also
-demonstrated reduced communication costs. On each processor, the
+replicated $O(1,000)$ times potentially achieves simulation on
+$O(100,000)$ processors, thus providing both spatial and particle
+parallelism. Each collection of processors that constitutes a
+representation of the entire domain is referred to as a set, and
+within a set overlap occurs among its sub-domains. The original
+motivation was to decompose the domain in a way that it remained in a
+physical cabinet in a large distributed machine, thus reducing latency
+costs during communication. A multiple set scheme is also motivated by
+the fact that communication during particle transport only occurs
+within a set, limiting communications during the transport procedure
+to a group of $O(100)$ processors, a number that was shown to have
+excellent parallel efficiencies in Brunner's work and therefore will
+scale well in this algorithm. The overlapping domains within each set
+also demonstrated reduced communication costs. On each processor, the
 source is sampled in the local domain that would exist if no overlap
@@
 
+\begin{figure}[htpb!]
+  \begin{center}
+    \scalebox{0.6}{ \input{chapters/parallel_mc/msod_construction.pdftex_t} }
+  \end{center}
+  \caption{\textbf{MSOD construction.} \textit{The linear system is
+      used to construct the Monte Carlo transport domain on the
+      primary set. The Monte Carlo data structures are then broadcast
+      amongst the blocks to all sets in the system.}}
+  \label{fig:msod_construction}
+\end{figure}
+
 \subsection{MSOD Neumann Ulam Algorithm}
 \label{subsec:msod_algorithm}
+
+\begin{figure}[htpb!]
+  \begin{center}
+    \scalebox{0.6}{ \input{chapters/parallel_mc/msod_tally.pdftex_t} }
+  \end{center}
+  \caption{\textbf{MSOD tally reduction.} \textit{The tally computed
+      independently in each set is reduced across the blocks to the
+      primary set. The linear system vector is updated on the primary
+      set. In this example the tally vector in the block containing
+      upper-left subdomains is combined. Identical reduction
+      operations occur for all other blocks simultaneously.}}
+  \label{fig:msod_tally_reduction}
+\end{figure}
 
@@
 %%---------------------------------------------------------------------------%%
-\section{Small Cluster Scaling Studies}
-\label{sec:small_scaling_studies}
+\section{Parallel Performance Metrics}
+\label{sec:parallel_performance_metrics}
 
@@
 
-\subsection{MSOD Decomposition}
+\subsection{Multiple Set Decomposition}
 \label{subsec:msod_decomposition}

### 2013-06-30 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
cleaning up first sections of parallel chapter
adding first titan scaling figures to parallel chapter

@@
-\chapter{Monte Carlo Synthetic Acceleration Methods for the Navier-Stokes Equations}
+\chapter{Monte Carlo Synthetic Acceleration Methods for the
+  Navier-Stokes \\Equations\ }
 \label{ch:nonlinear_problem}
@@
 %%---------------------------------------------------------------------------%%
-\section{Preliminaries}
+\section{Preliminaries\ }
 \label{sec:nonlinear_preliminaries}
@@
 %%---------------------------------------------------------------------------%%
-\section{The FANM Method}
+\section{The FANM Method\ }
 \label{sec:fanm}
@@
 %%---------------------------------------------------------------------------%%
-\section{Navier-Stokes Benchmark Problems}
+\section{Navier-Stokes Benchmark Problems\ }
 \label{sec:ns_benchmarks}
@@
 %%---------------------------------------------------------------------------%%
-\section{Analysis of the Navier-Stokes Equations}
+\section{Analysis of the Navier-Stokes Equations\ }
 \label{sec:ns_analysis}
@@
 %%---------------------------------------------------------------------------%%
-\section{FANM Verification}
+\section{FANM Verification\ }
 \label{sec:fanm_verification}
@@
 %%---------------------------------------------------------------------------%%
-\section{FANM Performance Comparison to Conventional Methods}
+\section{FANM Performance Comparison to Conventional Methods\ }
 \label{sec:fanm_comparison}

### 2013-06-30 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
adding first titan scaling figures to parallel chapter

@@
-\chapter{Parallel Monte Carlo Synthetic \\ Acceleration Methods}
+\chapter{Parallel Monte Carlo Synthetic \\ Acceleration Methods\ }
 \label{ch:parallel_methods}
@@
 %%---------------------------------------------------------------------------%%
-\section{Domain Decomposed Monte Carlo}
+\section{Domain Decomposed Monte Carlo\ }
 \label{sec:msod}
@@
 %%---------------------------------------------------------------------------%%
-\section{Analytic Performance Framework for Domain-Decomposed Monte Carlo}
+\section{Analytic Framework for Domain-Decomposed Monte Carlo\ }
 \label{sec:analytic_framework}
@@
 
-\subsection{Spectral Analysis}
+\subsection{Spectral Analysis }
 \label{subsec:spectral_analysis}
@@
 
-\subsection{Neumann Series Convergence}
+\subsection{Neumann Series Convergence }
 \label{subsec:neumann_convergence}
@@
 
-\subsection{Domain Leakage Approximations}
+\subsection{Domain Leakage Approximations }
 \label{subsec:domain_leak_approx}
@@
 
-\subsection{Numerical Experiments}
+\subsection{Numerical Experiments }
 \label{subsec:numerical_experiments}
@@
 
-\subsubsection{Random Walk Length}
+\subsubsection{Random Walk Length }
 \label{subsubsec:walk_length}
@@
 
-\subsubsection{Domain Leakage}
+\subsubsection{Domain Leakage }
 \label{subsubsec:domain_leakage}
@@
 %%---------------------------------------------------------------------------%%
-\section{Domain Decomposed Neumann-Ulam Algorithm}
+\section{Domain Decomposed Neumann-Ulam Algorithm\ }
 \label{sec:asynchronous_algorithm}
@@
 
-\subsection{Parallel Transport Domain and Source}
+\subsection{Parallel Transport Domain and Source }
 \label{subsec:domain_generation}
@@
 
-\subsection{Domain-to-Domain Communication Kernel}
+\subsection{Domain-to-Domain Communication Kernel }
 \label{subsec:domain_to_domain_kernel}
@@
 
-\subsection{Ending the Transport Sequence}
+\subsection{Ending the Transport Sequence }
 \label{subsec:ending_transport}
@@
 %%---------------------------------------------------------------------------%%
-\section{Multiple-Set Overlapping-Domain Neumann-Ulam Algorithm}
+\section{Multiple-Set Overlapping-Domain Algorithm\ }
 \label{subsec:msod}
@@
 
-\subsection{Generating Overlap}
+\subsection{Generating Overlap }
 \label{subsec:msod_overlap}
 
-\subsection{Generating Multiple Sets}
+\subsection{Generating Multiple Sets }
 \label{subsec:msod_sets}
@@
 
-\subsection{MSOD Neumann Ulam Algorithm}
+\subsection{MSOD Neumann Ulam Algorithm }
 \label{subsec:msod_algorithm}
@@
 %%---------------------------------------------------------------------------%%
-\section{Parallel MCSA}
+\section{Parallel MCSA\ }
 \label{sec:parallel_mcsa}
@@
 %%---------------------------------------------------------------------------%%
-\section{Parallel MCSA Verification}
+\section{Parallel MCSA Verification\ }
 \label{sec:parallel_verification}
@@
 %%---------------------------------------------------------------------------%%
-\section{Parallel Performance Metrics}
+\section{Parallel Performance Metrics\ }
 \label{sec:parallel_performance_metrics}
@@
 %%---------------------------------------------------------------------------%%
-\section{Leadership-Class Parallel Scaling Studies}
+\section{Leadership-Class Parallel Scaling Studies\ }
 \label{sec:leadership_scaling_studies}
@@
 
-\subsection{Titan}
+\subsection{Titan }
 \label{subsec:titan}
@@
 
+\subsubsection{Strong Scaling}
+\label{subsubsec:pure_strong}
+
+\begin{figure}[htpb!]
+  \begin{center}
+    \includegraphics[width=6in]{titan_pure_strong.pdf}
+  \end{center}
+  \caption{\textbf{Pure domain decomposition.} \textit{Super-linear
+      speed-up from memory thrashing in base case. MCLS is an order
+      of magnitude slower arithmetically, causing the higher
+      efficiency.}}
+\end{figure}
+
 \subsubsection{Weak Scaling}
@@
 
-\subsubsection{Strong Scaling}
-\label{subsubsec:pure_strong}
+\begin{figure}[htpb!]
+  \begin{center}
+    \includegraphics[width=6in]{titan_weak_absolute.pdf}
+  \end{center}
+\end{figure}
+
+\begin{figure}[htpb!]
+  \begin{center}
+    \includegraphics[width=6in]{titan_weak_implementation.pdf}
+  \end{center}
+\end{figure}
 
@@
 
+\subsubsection{Strong Scaling}
+\label{subsubsec:overlapping_strong}
+
 \subsubsection{Weak Scaling}
 \label{subsubsec:overlapping_weak}
-
-\subsubsection{Strong Scaling}
-\label{subsubsec:overlapping_strong}
 
@@
 
+\subsubsection{Strong Scaling}
+\label{subsubsec:msod_strong}
+
+\begin{figure}[htpb!]
+  \begin{center}
+    \includegraphics[width=6in]{titan_strong_ms_eff.pdf}
+  \end{center}
+  \caption{\small{Absolute parallel efficiency relative to
+      16-core 1-set base case.}}
+\end{figure}
+
+\begin{figure}[htpb!]
+  \begin{center}
+    \includegraphics[width=6in]{titan_strong_ms_time.pdf}
+  \end{center}
+  \caption{\small{Wall time in seconds to solution for each
+      case.}}
+\end{figure}
+
 \subsubsection{Weak Scaling}
@@
 
-\subsubsection{Strong Scaling}
-\label{subsubsec:msod_strong}
+\begin{figure}[htpb!]
+  \begin{center}
+    \includegraphics[width=6in]{titan_weak_ms_eff.pdf}
+  \end{center}
+  \caption{Absolute parallel efficiency relative to 16-core
+    1-set base case.}
+\end{figure}
+
+\begin{figure}[htpb!]
+  \begin{center}
+    \includegraphics[width=6in]{titan_weak_ms_time.pdf}
+  \end{center}
+  \caption{Wall time in seconds to solution for each case.}
+\end{figure}
 
 %%---------------------------------------------------------------------------%%
-\section{Multiple Fuel Assembly $SP_N$ Challenge Problem}
+\section{Multiple Fuel Assembly $SP_N$ Challenge Problem\ }
 \label{sec:spn_challenge_problem}

### 2013-06-30 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
adding scaling figure  captions

@@
   \begin{center}
-    \includegraphics[width=6in]{titan_pure_strong.pdf}
+    \includegraphics[width=6in]{chapters/parallel_mc/titan_pure_strong.pdf}
   \end{center}
-  \caption{\textbf{Pure domain decomposition.} \textit{Super-linear
-      speed-up from memory thrashing in base case. MCLS is an order
-      of magnitude slower arithmetically, causing the higher
-      efficiency.}}
+  \caption{\textbf{Pure domain decomposition absolute strong scaling.}
+    \textit{Super-linear speed-up from memory thrashing in base
+      case. MCLS is an order of magnitude slower arithmetically,
+      causing the higher efficiency.}}
+  \label{fig:titan_pure_strong}
 \end{figure}
+
+\clearpage
 
@@
   \begin{center}
-    \includegraphics[width=6in]{titan_weak_absolute.pdf}
+    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_absolute.pdf}
   \end{center}
+  \caption{\textbf{Pure domain decomposition absolute weak scaling}}
+  \label{fig:titan_weak_absolute}
 \end{figure}
@@
   \begin{center}
-    \includegraphics[width=6in]{titan_weak_implementation.pdf}
+    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_implementation.pdf}
   \end{center}
+  \caption{\textbf{Pure domain decomposition implementation weak scaling}}
+  \label{fig:titan_weak_implementation}
 \end{figure}
+
+\clearpage
 
@@
 
+\clearpage
+
 \subsubsection{Weak Scaling}
 \label{subsubsec:overlapping_weak}
+
+\clearpage
 
@@
   \begin{center}
-    \includegraphics[width=6in]{titan_strong_ms_eff.pdf}
+    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_ms_eff.pdf}
   \end{center}
-  \caption{\small{Absolute parallel efficiency relative to
-      16-core 1-set base case.}}
+  \caption{\textbf{Absolute parallel efficiency relative to 16-core
+      1-set base case.}}
+  \label{fig:titan_strong_ms_eff}
 \end{figure}
@@
   \begin{center}
-    \includegraphics[width=6in]{titan_strong_ms_time.pdf}
+    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_ms_time.pdf}
   \end{center}
-  \caption{\small{Wall time in seconds to solution for each
-      case.}}
+  \caption{\textbf{Wall time in seconds to solution for each case.}}
+  \label{fig:titan_strong_ms_time}
 \end{figure}
+
+\clearpage
 
@@
   \begin{center}
-    \includegraphics[width=6in]{titan_weak_ms_eff.pdf}
+    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_ms_eff.pdf}
   \end{center}
-  \caption{Absolute parallel efficiency relative to 16-core
-    1-set base case.}
+  \caption{\textbf{Absolute parallel efficiency relative to 16-core
+      1-set base case.}}
+  \label{fig:titan_weak_ms_eff}
 \end{figure}
@@
   \begin{center}
-    \includegraphics[width=6in]{titan_weak_ms_time.pdf}
+    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_ms_time.pdf}
   \end{center}
-  \caption{Wall time in seconds to solution for each case.}
+  \caption{\textbf{Wall time in seconds to solution for each case.}}
+  \label{fig:titan_weak_ms_time}
 \end{figure}
+
+\clearpage
 

### 2013-06-30 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
adding full clip scaling section
adding asynchronous algorithms for neumann ulam transport

@@
 In this section Brunner and Brantley's fully asynchronous algorithm is
-briefly presented along with its application to the Neumann-Ulam
-method. Their algorithm is presented in detail in the reference and
-was effectively implemented verbatim for this work. In their work they
+presented along with its application to the Neumann-Ulam method. Their
+algorithm is presented in further detail in the reference and was
+effectively implemented verbatim for this work. In their work they
 identify two data sets that are required to be communicated: the
@@
 processors. Both of these processes will be addressed at a high level
-along with how the parallel Monte Carlo data structures are
-constructed.
+along with how the Monte Carlo data structures are
+constructed in parallel.
 
@@
 
-\subsection{Domain-to-Domain Communication Kernel }
-\label{subsec:domain_to_domain_kernel}
+\clearpage
+
+\subsection{The Algorithm}
+\label{subsec:parallel_mc_algorithm}
+
+\begin{algorithm}[h!]
+  \caption{Parallel Neumann-Ulam Algorithm}
+  \label{alg:parallel_mc_algorithm}
+  \begin{algorithmic}[1]
+    \State get list of neighbor processors 
+    \Comment{Each neighbor owns an adjacent subdomain}
+    \ForAll{neighbors}
+    \State post nonblocking receive for maximum history buffer size
+    \State allocate history buffer
+    \EndFor
+    \State \textit{historiesCompleted} = 0
+    \Comment{local+child finished histories}
+    \State \textit{localProcessed} = 0
+    \Comment{local trajectories computed (not necessarily finished)}
+    \State calculate parent and children processor ID numbers in
+    binary tree
+    \ForAll{child processes}
+    \State post nonblocking receive for \textit{historiesCompleted} history tally
+    \EndFor
+    \State post nonblocking receive for stop message from parent
+    \While{stop flag not set}
+    \If{any local histories in source or stack}
+    \State \textbf{LocalHistoryTransport()}
+    \State ++\textit{localProcessed}
+    \EndIf
+    \If{checkPeriod == \textit{localProcessed} || no local histories}
+    \State \textbf{ProcessMessages()}
+    \State \textit{localProcessed} = 0
+    \EndIf
+    \If{no local histories}
+    \State \textbf{ControlTermination()}
+    \EndIf
+    \EndWhile
+    \State cancel outstanding nonblocking requests
+    \State free all history buffers
+  \end{algorithmic}
+\end{algorithm}
+
+\clearpage
+
+\subsubsection{Domain-to-Domain Communication Kernels}
+\label{subsubsec:domain_to_domain_kernel}
+
+\begin{algorithm}[h!]
+  \caption{\textbf{LocalHistoryTransport()}}
+  \label{alg:local_history_transport}
+  \begin{algorithmic}[1]
+    \State transport history through the domain until termination
+    \If{history is in neighbor boundary state}
+    \State add history to neighbor buffer
+    \If{neighbor buffer is full}
+    \State nonblocking send history buffer to neighbor
+    \State allocate new history buffer for neighbor
+    \EndIf
+    \Else
+    \If{history terminated by weight cutoff}
+    \State post-process history
+    \State ++\textit{historiesCompleted}
+    \EndIf
+    \EndIf
+  \end{algorithmic}
+\end{algorithm}
+
+\begin{algorithm}[h!]
+  \caption{\textbf{ProcessMessages()}}
+  \label{alg:process_messages}
+  \begin{algorithmic}[1]
+    \ForAll{received history buffers from neighbor}
+    \State unpack number of histories in buffer
+    \State add the histories to the running stack
+    \State repost nonblocking receive with neighbor
+    \EndFor
+    \ForAll{\textit{historiesCompleted} messages from children}
+    \State \textit{historiesCompleted} += message value
+    \State repost nonblocking receive with child
+    \EndFor
+  \end{algorithmic}
+\end{algorithm}
 
@@
 
-\subsection{Ending the Transport Sequence }
-\label{subsec:ending_transport}
+\clearpage
+
+\subsubsection{Ending the Transport Sequence}
+\label{subsubsec:ending_transport}
+
+\begin{algorithm}[h!]
+  \caption{\textbf{ControlTermination()}}
+  \label{alg:control_termination}
+  \begin{algorithmic}[1]
+    \State nonblocking send an partially full history buffers
+    \ForAll{\textit{historiesCompleted} messages from children}
+    \State \textit{historiesCompleted} += message \textit{historiesCompleted}
+    \State repost nonblocking receive with child
+    \EndFor
+    \If{MASTER processor}
+    \If{\textit{historiesCompleted} == global \# of source histories}
+    \State set stop flag
+    \ForAll{children}
+    \State nonblocking send stop message to child
+    \EndFor
+    \EndIf
+    \Else
+    \State nonblocking send \textit{historiesCompleted} to parent
+    \State \textit{historiesCompleted} = 0
+    \State check for stop signal from parent
+    \If{stop signal from parent}
+    \ForAll{children}
+    \State nonblocking send stop message to child
+    \EndFor
+    \EndIf
+    \EndIf
+  \end{algorithmic}
+\end{algorithm}
 
@@
 $O(100,000)$ processors, thus providing both spatial and particle
-parallelism. Each collection of processors that constitutes a
-representation of the entire domain is referred to as a set, and
-within a set overlap occurs among its sub-domains. The original
-motivation was to decompose the domain in a way that it remained in a
-physical cabinet in a large distributed machine, thus reducing latency
-costs during communication. A multiple set scheme is also motivated by
-the fact that communication during particle transport only occurs
-within a set, limiting communications during the transport procedure
-to a group of $O(100)$ processors, a number that was shown to have
-excellent parallel efficiencies in Brunner's work and therefore will
-scale well in this algorithm. The overlapping domains within each set
-also demonstrated reduced communication costs. On each processor, the
+parallelism.
+
+Each collection of processors that constitutes a representation of the
+entire domain is referred to as a set, and within a set overlap occurs
+among its sub-domains. The original motivation was to decompose the
+domain in a way that it remained in a physical cabinet in a large
+distributed machine, thus reducing latency costs during
+communication. A multiple set scheme is also motivated by the fact
+that communication during particle transport only occurs within a set,
+limiting communications during the transport procedure to a group of
+$O(100)$ processors, a number that was shown to have excellent
+parallel efficiencies in Brunner's work and therefore will scale well
+in this algorithm. The overlapping domains within each set also
+demonstrated reduced communication costs. On each processor, the
 source is sampled in the local domain that would exist if no overlap
@@
 
+\clearpage
+
 %%---------------------------------------------------------------------------%%
@@
 trivial. Recall the MCSA iteration procedure outlined in
-Eq~(\ref{eq:mcsa}). In \S\ref{sec:parallel_krylov_methods} we
-discussed parallel matrix and vector operations as utilized in
-conventional Krylov methods. We utilize these here for the parallel
-MCSA implementation. In the first step, a parallel matrix-vector
-multiply is used to apply the split operator to the previous iterate's
+Eq~(\ref{eq:mcsa}). In \S\ref{sec:parallel_krylov_methods} we discuss
+parallel matrix and vector operations as utilized in conventional
+Krylov methods. We utilize these here for the parallel MCSA
+implementation. In the first step, a parallel matrix-vector multiply
+is used to apply the split operator to the previous iterate's
 solution. A parallel vector update is then performed with the source
@@
 
+How effectively the presented algorithms parallelize MCSA will be
+determined by several performance measures. For this work, these
+measures come directly from Keyes' primer on the scalability of domain
+decomposed methods and specifically his discussion on how to measure
+their parallel performance \citep{keyes_how_1999}. All of these
+metrics are derived from two measured quantities: the number of
+iterations required for convergence and the total wall time required
+for convergence. Each of these quantities are then measured in varying
+parameter studies by changing the number of processes in a simulation
+and by varying the local or global problem size. Using the notation of
+Keyes, we will define \textit{parallel efficiency} as $\eta(N,P)$
+where the efficiency is a function of the global problem size, $N$,
+and the number of processes in the calculation, $P$. There are several
+types of efficiencies that will be useful in our analysis of
+scalability with different definitions depending on the type of
+parametric scaling study performed. For all efficiencies, a value of 1
+is considered perfect while a value of 0 signals the worst possible
+performance. Often, as was observed in this work, it is the case that
+efficiencies above 1 are computed in certain scaling studies. For the
+data presented in the following sections, an explanation of each
+observed case of superunitary efficiencies will be provided.
+
+Of primary interest to the nuclear engineer is the \textit{strong
+  scaling} performance of an algorithm
+\citep{siegel_analysis_2012}. In this type of scaling study, the
+global problem size is fixed and the local problem size changes with
+the number of processes in the calculation. In this case, as $P$
+increases, $N$ decreases, thus reducing performance by having a larger
+ratio of communication to computational work. This type of scaling is
+important because typically the global problem size is fixed due to
+the problem of interest. Consider the desire for a high fidelity
+neutronics simulation of the entire core of a nuclear reactor. In this
+case, the global problem is known from the geometry definition
+including the number of fuel assemblies and the geometry of those
+assemblies all the way down to the resolution of the space, angle, and
+energy discretizations required in order to capture physical phenomena
+of interest. Because the global problem size is known a priori, strong
+scaling is an important measure of how useful a solution technique is
+when larger and larger values of $P$ are used to solve the
+problem. Ideally, the time to solution for the the problem will
+decrease linearly as a function of $P$, permitting a faster time to
+solution using the maximum available resources thus making sure all
+$P$ are positively affecting the runtime of the problem.
+
+For strong scaling, the objective is to then decrease the runtime of a
+fixed size global problem as a linear function of $P$. We can then
+define the strong scaling \textit{absolute efficiency} as:
+\begin{equation}
+  \eta_{strong}(N,P) = \frac{1}{P} \frac{T(N,1)}{T(N,P)}\:,
+  \label{eq:strong_scaling_absolute}
+\end{equation}
+where $T(N,P)$ is the wall time for a computation of global size $N$
+using $P$ processes. Note here that this particular definition is
+measured with respect to a serial computation. On leadership class
+machines, it is often the case that a problem size large enough to
+effectively execute on $O(10,000)$ cores or larger is significantly
+larger than any problem that may be solved by a single core due to
+memory restrictions. Therefore, we consider absolute scaling for a
+base case where $Q$ processes are used instead of a serial
+computation:
+\begin{equation}
+  \eta_{strong}(N,P|Q) = \frac{Q}{P} \frac{T(N,Q)}{T(N,P)}\:.
+  \label{eq:strong_scaling_absolute_ref}
+\end{equation}
+The absolute efficiency measure presented here only considers the
+total time to solution. Often, that total time to solution may be
+changing for an iterative method due to the fact that the number of
+iterations required to converge may be changing as a function of
+$P$. This will often happen in cases where modifying $N$ changes the
+spectral character of the problem, often making it stiffer and
+therefore more difficult to solve when $N$ increases. Therefore we
+also consider \textit{algorithmic efficiency} from Keyes' work which
+considers this potential change in iteration count:
+\begin{equation}
+  \eta_{alg}(N,P|Q) = \frac{I(N,Q)}{I(N,P)}\:,
+  \label{eq:algorithmic_efficiency}
+\end{equation}
+where $I(N,P)$ is the number of iterations required to converge a
+problem of global size $N$ computed using $P$ processes. We can then
+adjust the absolute efficiency using this new algorithmic efficiency
+in order to compute the parallel scalability of a single iteration
+such that algorithmic effects may be neglected. We define the
+\textit{implementation efficiency} as:
+\begin{equation}
+  \eta_{impl}(N,P|Q) = \frac{\eta}{\eta_{alg}}\:,
+  \label{eq:implementation efficiency}
+\end{equation}
+using the absolute efficiency in the numerator, giving an effective
+measure of scalability of a single iteration of the algorithm.
+
+In addition to strong scaling, in certain cases it is useful to
+consider the \textit{weak scaling} of an iterative method. In this
+case, the ratio $N/P$ is fixed such that the local problem size
+remains the same for all values of $P$. In this case, the objective is
+to maintain a constant wall time while both $N$ and $P$ grow as a
+fixed ratio. In this case, we define the absolute efficiency as:
+\begin{equation}
+  \eta_{weak}(M|N,P|Q) = \frac{T(M,Q)}{T(N,P)}\:,
+  \label{eq:weak_scaling_absolute}
+\end{equation}
+subject to the constraint $M/Q = N/P$. The scaling is perfect if the
+runtime is static for all $P$ when $N/P$ is fixed. Often, weak scaling
+is of interest to the nuclear engineer due to the fact that the
+computational resources available are fixed and may not be able to
+contain the entire global problem of interest. In this case, an
+algorithm with good weak scaling performance permits all of the
+computational resources available to be used effectively to allow for
+the highest fidelity problem possible to be solved. In addition, for
+the resources available, the memory to processor ratio on a
+distributed machine is often fixed, giving an exercise in fixed $N/P$
+to solve the highest fidelity problem possible. Therefore, good weak
+scaling should not be viewed as an enhancement of time to solution for
+the engineer but rather permitting a higher fidelity problem to be
+solved within a given time constraint. For problems where increasing
+fidelity increases the stiffness of the system, there is often a
+trade-off between the absolute efficiencies computed for a given
+method and the degrading algorithmic efficiencies that arise when
+increasing the fidelity. Both time and hardware constraints should be
+considered when preparing a parallel calculation.
+
 %%---------------------------------------------------------------------------%%
@@
   \end{center}
-  \caption{\textbf{Pure domain decomposition implementation weak scaling}}
+  \caption{\textbf{Pure domain decomposition implementation weak
+      scaling}}
   \label{fig:titan_weak_implementation}
@@
 \label{subsubsec:msod_strong}
+
+\begin{figure}[htpb!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_ms_time.pdf}
+  \end{center}
+  \caption{\textbf{Wall time in seconds to solution for each case.}}
+  \label{fig:titan_strong_ms_time}
+\end{figure}
 
@@
 
-\begin{figure}[htpb!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_ms_time.pdf}
-  \end{center}
-  \caption{\textbf{Wall time in seconds to solution for each case.}}
-  \label{fig:titan_strong_ms_time}
-\end{figure}
-
 \clearpage
@@
 \label{subsubsec:msod_weak}
+
+\begin{figure}[htpb!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_ms_time.pdf}
+  \end{center}
+  \caption{\textbf{Wall time in seconds to solution for each case.}}
+  \label{fig:titan_weak_ms_time}
+\end{figure}
 
@@
 
-\begin{figure}[htpb!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_ms_time.pdf}
-  \end{center}
-  \caption{\textbf{Wall time in seconds to solution for each case.}}
-  \label{fig:titan_weak_ms_time}
-\end{figure}
+\clearpage
+
+\subsection{Full-Clip Neumann-Ulam}
+\label{subsec:full_clip}
+
+\subsubsection{Weak Scaling}
+\label{subsubsec:full_clip_weak}
+
+\clearpage
+
+\subsubsection{Strong Scaling}
+\label{subsubsec:full_clip_strong}
 

### 2013-06-30 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
updating titan comm parameter data

@@
     \ForAll{child processes}
-    \State post nonblocking receive for \textit{historiesCompleted} history tally
+    \State post nonblocking receive for \textit{historiesCompleted} tally
     \EndFor
@@
 
+\begin{figure}[htpb!]
+  \begin{center}
+    \scalebox{0.75}{
+      \input{chapters/parallel_mc/master_comm_tree.pdftex_t} }
+    \caption{\textbf{Master/slave scheme for coordinating the end of a
+        parallel Neumann-Ulam solve.} \textit{Each slave process
+        reports to the master how many histories completed within its
+        domain. When the master process sums all complete histories,
+        it sends the stop signal to all slave processes. The
+        subdomains are indexed by an integer.}}
+  \end{center}
+  \label{fig:master_comm_tree}
+\end{figure}
+
 \clearpage

### 2013-07-01 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
adding efficiency plots to scaling section. updating multiple set weak scaling definition

@@
 method. Direct analogs can be derived from their works by noting that
-the primary difference between solving a linear system with Monte
-Carlo methods and fixed source Monte Carlo transport problems is the
-content of the Markov chains that are generated. The transitions
-represented by these chains are bound by probabilities and weights and
-are initiated by the sampling of a discrete source. In the context of
-transport problems, those transitions represent events such as
-particle scattering and absorption with probabilities that are
+the primary difference between solving a linear transport system with
+Monte Carlo methods and traditional fixed source Monte Carlo transport
+problems is the content of the Markov chains that are generated. The
+transitions represented by these chains are bound by probabilities and
+weights and are initiated by the sampling of a discrete source. In the
+context of transport problems, those transitions represent events such
+as particle scattering and absorption with probabilities that are
 determined by physical data in the form of cross sections.
@@
 
+\begin{figure}[htpb!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/parallel_mc/titan_comm_parameters.pdf}
+  \end{center}
+  \caption{\textbf{Neumann Ulam parallel algorithm free parameter
+      sensitivity.}  \textit{All runtimes were within 6.5\% of the
+      fastest observed case.}}
+  \label{fig:titan_comm_parameters}
+\end{figure}
+
+\clearpage
+
 \subsection{Pure Domain Decomposition}
@@
   \end{center}
-  \caption{\textbf{Pure domain decomposition absolute strong scaling.}
-    \textit{Super-linear speed-up from memory thrashing in base
-      case. MCLS is an order of magnitude slower arithmetically,
-      causing the higher efficiency.}}
+  \caption{\textbf{Pure domain decomposition strong scaling absolute
+      efficiency.}  \textit{Super-linear speed-up from memory
+      thrashing in base case. MCLS is an order of magnitude slower
+      arithmetically, causing the higher efficiency.}}
   \label{fig:titan_pure_strong}
+\end{figure}
+
+\begin{figure}[htpb!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_bvsm.pdf}
+  \end{center}
+  \caption{\textbf{Binary tree vs. master/slave communication scheme
+      absolute efficiency.} \textit{Pure domain decomposition absolute
+      strong scaling. The master/slave scheme will not scale beyond
+      $O(1,000)$ cores due to load imbalance.}}
+  \label{fig:titan_strong_bvsm}
 \end{figure}
@@
   \end{center}
-  \caption{\textbf{Pure domain decomposition absolute weak scaling}}
+  \caption{\textbf{Pure domain decomposition weak scaling absolute
+      efficiency.}}
   \label{fig:titan_weak_absolute}
@@
   \begin{center}
+    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_alg_eff.pdf}
+  \end{center}
+  \caption{\textbf{Pure domain decomposition weak scaling algorithmic
+      efficiency.}}
+  \label{fig:titan_weak_algorithmic}
+\end{figure}
+
+\begin{figure}[htpb!]
+  \begin{center}
     \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_implementation.pdf}
   \end{center}
-  \caption{\textbf{Pure domain decomposition implementation weak
-      scaling}}
+  \caption{\textbf{Pure domain decomposition weak scaling
+      implementation efficiency.}}
   \label{fig:titan_weak_implementation}
+\end{figure}
+
+\begin{figure}[htpb!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_bvsm.pdf}
+  \end{center}
+  \caption{\textbf{Binary tree vs. master/slave communication scheme
+      absolute efficiency.}  \textit{Pure domain decomposition
+      absolute weak scaling. The master/slave scheme will not scale
+      beyond $O(1,000)$ cores due to load imbalance.}}
+  \label{fig:titan_weak_bvsm}
 \end{figure}
@@
 
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lcc}\hline\hline
+      \multicolumn{1}{c}{Case}& 
+      \multicolumn{1}{c}{Iterations}&
+      \multicolumn{1}{c}{$\eta_{alg}$} \\\hline
+      1 & 22 & 1.0 \\
+      2-split & 22 & 1.0 \\
+      2-replicated & 16 & 1.375 \\
+      4-split & 22 & 1.0 \\
+      4-replicated & 13 & 1.692 \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Strong scaling algorithmic efficiency using
+      multiple sets.}}
+  \label{tab:ms_strong_alg_eff}
+\end{table}
+
+\begin{figure}[htpb!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_ms_impeff.pdf}
+  \end{center}
+  \caption{\textbf{Strong scaling implementation efficiency relative
+      to 16-core 1-set base case.}}
+  \label{fig:titan_strong_ms_impeff}
+\end{figure}
+
 \clearpage
@@
 
+For a multiple set analysis of the weak scaling results, we have to
+consider that adding sets to the problem is actually a form of strong
+scaling. Consider a single set problem ($S=1$) for a given $P$ and
+$N$. When that problem is now solved using 2 sets we have $S=2$, $N$
+and $2P$. For any number of sets we then have a global problem size of
+$N$ while the number of processors is $SP$ where $P$ is the number of
+processors in the single set case. Therefore, $N/P$ is no longer fixed
+in the weak scaling study but rather $N/(SP)$ is fixed. Therefore, we
+modify the weak scaling absolute efficiency relationship to account
+for this fact:
+\begin{equation}
+\eta_{weak}(M|N,P|Q,S) = \frac{T(M,Q)}{S \times T(N,P)}\:,
+  \label{eq:ms_weak_efficiency}
+\end{equation}
+where now the absolute efficiency is a function of the number of sets
+in the problem. For problems where adding sets improves time to
+solution, this new weak scaling formulation accounts for how
+efficiently those extra resources are incorporated into the
+problem. Without such a modification, the largest case run for the
+split 4 set case would have a computed weak scaling efficiency of
+nearly 100\% using the standard formula for absolute
+efficiency. However, using 4 times as many cores as the single set
+case increases the runtime for a fixed global problem size and
+therefore we actually expect to compute a very low efficiency that
+shows this very poor use of resources.
+
 \begin{figure}[htpb!]
@@
   \end{center}
-  \caption{\textbf{Absolute parallel efficiency relative to 16-core
+  \caption{\textbf{Weak scaling absolute efficiency relative to
+      16-core 1-set base case.}}
+  \label{fig:titan_weak_ms_eff}
+\end{figure}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lcc}\hline\hline
+      \multicolumn{1}{c}{Case}& 
+      \multicolumn{1}{c}{Iterations}&
+      \multicolumn{1}{c}{$\eta_{alg}$} \\\hline
+      1 & 22 & 1.0 \\
+      2-split & 22 & 1.0 \\
+      2-replicated & 16 & 1.375 \\
+      4-split & 22 & 1.0 \\
+      4-replicated & 13 & 1.692 \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Weak scaling algorithmic efficiency using multiple
+      sets.}}
+  \label{tab:ms_weak_alg_eff}
+\end{table}
+
+\begin{figure}[htpb!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_ms_impeff.pdf}
+  \end{center}
+  \caption{\textbf{Implementation parallel efficiency relative to 16-core
       1-set base case.}}
-  \label{fig:titan_weak_ms_eff}
+  \label{fig:titan_weak_ms_impeff}
 \end{figure}
@@
 \label{subsubsec:full_clip_strong}
-
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{Multiple Fuel Assembly $SP_N$ Challenge Problem\ }
-\label{sec:spn_challenge_problem}

### 2013-07-01 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
updating parallel algorithm explanation

@@
 In this section Brunner and Brantley's fully asynchronous algorithm is
-presented along with its application to the Neumann-Ulam method. Their
-algorithm is presented in further detail in the reference and was
-effectively implemented verbatim for this work. In their work they
-identify two data sets that are required to be communicated: the
-sharing of particles that are transported from one domain to another
-and therefore from one processor to another and a global communication
-that signals if particle transport has been completed on all
-processors. Both of these processes will be addressed at a high level
-along with how the Monte Carlo data structures are
-constructed in parallel.
+presented along with its application to the Neumann-Ulam method. For
+considerably more detail, the algorithm presented can be found in
+\citep{brunner_efficient_2009} and was effectively implemented
+verbatim for this work. In their work they identify two data sets that
+are required to be communicated: the sharing of particles that are
+transported from one domain to another and therefore from one
+processor to another and a global communication that signals if
+particle transport has been completed on all processors. Both of these
+processes will be addressed at a high level along with how the Monte
+Carlo data structures they require are constructed in parallel.
 
@@
 \label{subsec:domain_generation}
-To utilize the parallel transport algorithm, we must generate the
-parallel transport domain. For the Neumann-Ulam method, the domain
-consists of all states in the system that are local, and the
-probabilities and weights for all state transitions possible in the
-local domain. At the core of this representation is the Neumann-Ulam
-decomposition of the linear operator. Given an input decomposition for
-the linear operator, by definition the Neumann-Ulam decomposition will
-have the same parallel decomposition. In addition, any modification
-made to the linear operator through preconditioning or relaxation
-parameters will also modify the Neumann-Ulam decomposition. All
-possible transitions are described by the local graph of the sparse
-input matrix. In addition, any left preconditioning or relaxation
-parameters will modify the right hand side of the linear system and
-therefore the fixed source in the Monte Carlo calculation. This source
-vector will have the same parallel decomposition as the input matrix
-and therefore all of the birth states for the histories will exist for
-at least a single transition event within the local domain.
+To utilize a parallel transport algorithm, we must generate the
+required data with the correct parallel decomposition. For the
+Neumann-Ulam method, the transport domain consists of all states in
+the system that are local, and the probabilities and weights for all
+state transitions possible in the local domain. At the core of this
+representation is the Neumann-Ulam decomposition of the linear
+operator. Given an input decomposition for the linear operator, by
+definition the Neumann-Ulam decomposition will have the same parallel
+decomposition. In addition, any modification made to the linear
+operator through preconditioning or relaxation parameters will also
+modify the Neumann-Ulam decomposition. All possible transitions are
+described by the local graph of the sparse input matrix. In addition,
+any left preconditioning or relaxation parameters will modify the
+right hand side of the linear system and therefore the fixed source in
+the Monte Carlo calculation. This source vector will have the same
+parallel decomposition as the input matrix and therefore all of the
+birth states for the histories will exist for at least a single
+transition event within the local domain.
+
+Compared to the serial construction of the Neumann-Ulam decomposition,
+data for all states that is required to be on process must be
+collected in parallel. For the pure domain decomposition case, given
+an input parallel decomposition for the Neumann-Ulam decomposition,
+for all local states $m$ we require access to $P_{mn}$ and $W_{mn}$
+(or $P^T_{mn}$ and $W^T_{mn}$) for the adjoint method for all possible
+states $n$. Given by Figure~\ref{fig:diffusion_graph}, consider the
+adjacency graph of a single state in the neutron diffusion matrix for
+the model problem presented in Appendix~\ref{chap:diffusion_problem}.
+\begin{figure}[htpb!]
+  \begin{center}
+    \scalebox{1.5}{ \input{chapters/parallel_mc/stencil_graph.pdftex_t} }
+  \end{center}
+  \caption{\textbf{Neutron diffusion equation state adjacency graph.}
+    \textit{Adjacent boundary states are collected by traversing the
+      graph one step for all local boundary states and collecting the
+      equations for those states that are not local. For a given
+      transition in the random walk for the diffusion problem, a
+      history at mesh point $(i,j)$ may transition to all adjacent
+      mesh points including itself, corresponding to a global state
+      transition from $m$ to $n$ in the linear system.}}
+  \label{fig:diffusion_graph}
+\end{figure}
+
+In this example, we currently reside in state $m$ of the system,
+directly correlating to physical location $(i,j)$ in the mesh (the
+center node). The discretization stencil of the diffusion problem
+dictates the structure of this graph and the states to which a history
+may transition. If we play the Monte Carlo game to move to a new state
+$n$, then we may move to any of the nodes in this graph, including the
+node at which we started. If grid point $(i,j)$ in this example is on
+the right of the local domain boundary and we transition to the right
+to state $n$ corresponding to mesh point $(i+1,j)$, we are now in a
+state that is owned by the adjacent domain. We must then gather the
+data in the Neumann-Ulam decomposition from the neighboring process
+that owns grid point $(i+1,j)$. Doing this provides us with the data
+required by the estimators permitting $P_{mn}$ and $W_{mn}$ to be
+computed locally for this particular boundary transition. For the
+parallel source, these adjacent states are not required as histories
+may only be born in states in the local domain. Once the local
+Neumann-Ulam decomposition has been generated with the proper data
+collected from adjacent domains, Monte Carlo transport may proceed in
+parallel.
 
@@
 \label{subsec:parallel_mc_algorithm}
+
+Here we present in detail Brunner and Brantley's 2009 algorithm in
+detail and discuss how it is adapted to parallelize the Neumann Ulam
+method. Presented in Algorithm~\ref{alg:parallel_mc_algorithm}, the
+top level sequence performs the task of managing history transport
+through the local domain, communication of histories to adjacent
+domains, and the completion of transport. For each of these tasks,
+additional algorithms, shown in bold in
+Algorithm~\ref{alg:parallel_mc_algorithm}
+(e.g. \textbf{LocalHistoryTransport}), are presented for additional
+detail.
 
@@
     \EndIf
-    \If{checkPeriod == \textit{localProcessed} || no local histories}
+    \If{message check period == \textit{localProcessed} || no local histories}
     \State \textbf{ProcessMessages()}
@@
 \end{algorithm}
-
-\clearpage
 
@@
 \end{figure}
-
-\clearpage
 
@@
 
-\subsection{Full-Clip Neumann-Ulam}
+\subsection{Subdomain Neumann-Ulam}
 \label{subsec:full_clip}
+
+Full-clip actually uses the solver as a true subdomain solver, just
+like you would in a Schwarz method. In fact, this is a stochastic
+realization of a Richardson iteration - they are very similar in those
+regards.
+
+Overlap is not effective because histories apparently dump most of
+their useful information within the first few steps (which is why we
+can use such a large weight cutoff and there is an insensitivity to
+weight cutoff). So keeping them around to do extra work is not
+improving iterative performance - thus increasing compute times. This
+is true even for ill-conditioned cases where adding overlap might keep
+a number of histories within the domain for significantly longer.
+
+\begin{figure}[htpb!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_subdomain.pdf}
+  \end{center}
+  \caption{\textbf{Strong scaling absolute efficiency for pure domain
+      decomposition.} \textit{The subdomain Monte Carlo solver
+      improves the parallel efficiency and therefore time to solution
+      by eliminating all parallel communication during the Monte Carlo
+      solve.}}
+  \label{fig:titan_strong_subdomain}
+\end{figure}
+
+\begin{figure}[htpb!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_subdomain.pdf}
+  \end{center}
+  \caption{\textbf{Weak scaling absolute efficiency for pure domain
+      decomposition.} \textit{The subdomain Monte Carlo solver
+      improves the parallel efficiency and therefore time to solution
+      by eliminating all parallel communication during the Monte Carlo
+      solve.}}
+  \label{fig:titan_weak_subdomain}
+\end{figure}
+
+\subsubsection{Strong Scaling}
+\label{subsubsec:full_clip_strong}
+
+\clearpage
 
@@
 \label{subsubsec:full_clip_weak}
-
-\clearpage
-
-\subsubsection{Strong Scaling}
-\label{subsubsec:full_clip_strong}

### 2013-07-02 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
finished single set algorithm description

@@
 particle transport has been completed on all processors. Both of these
-processes will be addressed at a high level along with how the Monte
-Carlo data structures they require are constructed in parallel.
+communication sequences will be addressed at a high level along with
+how the Monte Carlo data structures they require are constructed in
+parallel.
 
@@
 representation is the Neumann-Ulam decomposition of the linear
-operator. Given an input decomposition for the linear operator, by
-definition the Neumann-Ulam decomposition will have the same parallel
+operator as given by Eq~\ref{eq:neumann_ulam_decomposition}. Given an
+input decomposition for the linear operator, by definition the
+Neumann-Ulam decomposition will have the same parallel
 decomposition. In addition, any modification made to the linear
@@
 required by the estimators permitting $P_{mn}$ and $W_{mn}$ to be
-computed locally for this particular boundary transition. For the
-parallel source, these adjacent states are not required as histories
-may only be born in states in the local domain. Once the local
-Neumann-Ulam decomposition has been generated with the proper data
-collected from adjacent domains, Monte Carlo transport may proceed in
-parallel.
+computed locally for this particular boundary transition. In addition,
+we also collect the identification number for the process that owns
+this state such that we have an address to send all histories that
+leave the local domain boundary for this state. For the parallel
+source, these adjacent states are not required as histories may only
+be born in states in the local domain. Once the local Neumann-Ulam
+decomposition has been generated with the proper data collected from
+adjacent domains, Monte Carlo transport may proceed in parallel.
 
@@
 through the local domain, communication of histories to adjacent
-domains, and the completion of transport. For each of these tasks,
-additional algorithms, shown in bold in
+domains, and the completion of transport. For each of these specific
+tasks, additional algorithms, shown in bold in
 Algorithm~\ref{alg:parallel_mc_algorithm}
-(e.g. \textbf{LocalHistoryTransport}), are presented for additional
-detail.
+(e.g. \textbf{LocalHistoryTransport()}), are presented for additional
+detail in the same manner as Brunner and Brantley.
 
@@
 
-\subsubsection{Domain-to-Domain Communication Kernels}
-\label{subsubsec:domain_to_domain_kernel}
+Successful execution of this algorithm requires construction of the
+parallel Neumann-Ulam decomposition as described in the preceeding
+section. This data is used to begin
+Algorithm~\ref{alg:parallel_mc_algorithm} where lines 1-5 use the
+collected list of neighboring processors generated while building the
+Neumann-Ulam decompostion to setup the set of asynchronous messages
+required for domain-to-domain communication. Again, consider this
+communication patter for the 9 subdomain example given by
+Figure~\ref{fig:nearest_neighbor_comm}. In this pattern, each domain
+has two neighbors with which they will communicate parallel histories
+and this communication goes both ways as represented by the adjacent
+arrows in the figure. For each set of neighbors, a non-blocking send
+and receive is required with data buffers allocated with a
+user-defined size prepared for incoming and outgoing histories. This
+nonblocking structure is critical to the performance of the algorithm
+in that it permits local history transport to receive while new work
+to do is being collected. When a given process is ready to do more
+work, it can check these data buffers for incoming histories requiring
+further transport. In this way there is a maximum amount of overlap
+between communication and transport of histories.
+
+\begin{figure}[htpb!]
+  \begin{center}
+    \scalebox{1.5}{ \input{chapters/parallel_mc/domain_to_domain.pdftex_t} }
+  \end{center}
+  \caption{\textbf{Nearest neighbor history communication sequence.}
+    \textit{Each subdomain in the system has a set of nearest
+      neighbors determined by the parallel adjacency graph of the
+      input matrix. The subdomains are indexed by an integer.}}
+  \label{fig:nearest_neighbor_comm}
+\end{figure}
+
+Once the nearest-neighbor communication sequence has been prepared,
+the completion of transport sequence is readied in lines 6 and 8-12 in
+Algorithm~\ref{alg:parallel_mc_algorithm}. In these lines we are
+setting up an asynchronous binary communication tree as presented for
+the same 9 subdomain example in Figure~\ref{fig:binary_comm_tree}. In
+this communication pattern, each process has one parent process
+(except for the MASTER process 0) to which it will nonblocking send
+the number of histories that terminated their transport procedure by
+weight cutoff in the local domain. Equivalently, each process has up
+to two child processes from which it will receive their completed
+number of histories with a nonblocking receive operation. Setting up a
+tree in this manner lets the completed history tally
+(\textit{historiesCompleted} in the algorithm) be updated
+incrementally and funnelled to the root process. Because we are
+solving fixed source problems with the Neumann-Ulam algorithm without
+any type of variance reduction that may generate more histories that
+we started with, once the root process completion sum tallies to the
+number of histories in the source, transport is complete. Once this
+occurs, the root process nonblocking sends a process to its children
+and each process nonblocking receives a stop message from its parent
+as shown in Figure~\ref{fig:binary_comm_tree}. The stop message is
+then propagated up the tree in the same manner.
+
+\begin{figure}[htpb!]
+  \begin{center}
+    \scalebox{1.25}{
+      \input{chapters/parallel_mc/binary_comm_tree.pdftex_t} }
+    \caption{\textbf{Binary communication tree for coordinating the
+        end of a parallel Neumann-Ulam solve.} \textit{Each child
+        process reports to its parents how many histories completed
+        within its domain. When the root process sums all complete
+        histories, it forwards the stop signal to its children which
+        in turn forward the message up the tree. The subdomains are
+        indexed by an integer.}}
+  \end{center}
+  \label{fig:binary_comm_tree}
+\end{figure}
+
+With these communication structures prepared, we can now enter the
+main transport loop at line 13 of
+Algorithm~\ref{alg:parallel_mc_algorithm}. In this loop, a set of
+mutually exclusive tasks enabled by the fully asynchronous
+communication patterns are executed until the stop signal is received
+from the parent process in the binary tree. For this algorithm, being
+mutually exclusive means that all local operations occur independently
+of other processes in the system and their current state. For each
+process, there are two data structures from which histories may be
+obtained for the transport procedure. The first is the local source
+and the second is a LIFO (last in first out) stack of histories
+transported to the local domain from the adjacent domain. In lines
+14-17, if histories exist in either of these data structures, then
+they are transported through the local domain using
+Algorithm~\ref{alg:local_history_transport}. If the history is
+terminated by the weight cutoff the \textit{historiesCompleted} tally
+is updated and if it hits the boundary of the domain it is added to
+the outgoing history buffer for the neighboring dokmain. If that
+buffer is full, it is sent to the neighboring domain with a
+nonblocking operation and the memory associated with that buffer
+reallocated. In all instances that
+Algorithm~\ref{alg:local_history_transport} is executed,
+\textit{localProcessed} is incremented as a measure of the number of
+histories that have been processed locally.
 
@@
 
+Continuing in the transport loop, if there are no local histories to
+transport or the \textit{localProcessed} count has reached some
+user-define check frequency, lines 18-21 in
+Algorithm~\ref{alg:local_history_transport} check for more histories
+to transport in the incoming data buffers by calling
+Algorithm~\ref{alg:process_messages}. For each neighboring domain in
+the problem, if there are histories in those data buffers then they
+are added to the stack from processing and the nonblocking receive
+operation reinstantiated. In addition, the terminated histories count
+is updated with those values received from the child processes.
+
 \begin{algorithm}[h!]
@@
 
-\begin{figure}[htpb!]
-  \begin{center}
-    \scalebox{1.5}{ \input{chapters/parallel_mc/domain_to_domain.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Nearest neighbor history communication sequence.}
-    \textit{Each subdomain in the system has a set of nearest
-      neighbors determined by the parallel adjacency graph of the
-      input matrix. The subdomains are indexed by an integer.}}
-  \label{fig:nearest_neighbor_comm}
-\end{figure}
-
-\subsubsection{Ending the Transport Sequence}
-\label{subsubsec:ending_transport}
+Finally, the transport loop finishes in lines 22-24 of
+Algorithm~\ref{alg:local_history_transport} by calling
+Algorithm~\ref{alg:control_termination} if there are no local
+histories to transport in the stack or the source. In
+Algorithm~\ref{alg:control_termination}, we continue to forward the
+terminated history tally down to the parent in the binary tree and
+send all history buffers to our neighbors, even if they are not
+full. If Algorithm~\ref{alg:control_termination} is called by the
+MASTER process, it checks for completion of the problem and if it is
+complete, forwards to stop flag onto its children as in
+Figure~\ref{fig:binary_comm_tree}. If the process is not the MASTER,
+then we check for a stop signal from the parent and forward to the
+child processes if transport is complete.
 
@@
 
-\begin{figure}[htpb!]
-  \begin{center}
-    \scalebox{1.25}{
-      \input{chapters/parallel_mc/binary_comm_tree.pdftex_t} }
-    \caption{\textbf{Binary communication tree for coordinating the
-        end of a parallel Neumann-Ulam solve.} \textit{Each child
-        process reports to its parents how many histories completed
-        within its domain. When the root process sums all complete
-        histories, it forwards the stop signal to its children which
-        in turn forward the message up the tree. The subdomains are
-        indexed by an integer.}}
-  \end{center}
-  \label{fig:binary_comm_tree}
-\end{figure}
+In this manner the transport loop continues until all process have
+received the stop signal from their parents. Older versions of this
+algorithm, particularly some of those presented in
+\citep{brunner_comparison_2006}, use a master/slave approach as given
+by Figure~\ref{fig:master_comm_tree} for the completion of a transport
+stage instead of the binary tree scheme. In this approach, the root
+process still manages the final summation of the completion tally, it
+directly receives completed tally results from all other processes in
+the problem instead of from just its children. Although this
+implementation is potentially simpler to understand than the binary
+tree approach, the authors of \citep{brunner_comparison_2006} observed
+that this type of termination pattern, even when implemented in a
+fully asynchronous manner, was a major scalability bottleneck. This
+bottleneck is due to the fact that the master process falls behind the
+others, creating a load imbalance even with the addition of a few
+integer addition operations. We will explicitly demonstrate in scaling
+studies how this bottleneck also occurs within a Neumann-Ulam
+implementation of this algorithm thus requiring the implementation of
+a binary communication tree.
 

### 2013-07-02 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
adding explanation for msod generation

@@
 communication patterns are executed until the stop signal is received
-from the parent process in the binary tree. For this algorithm, being
-mutually exclusive means that all local operations occur independently
-of other processes in the system and their current state. For each
+from the parent process in the binary tree. In this case, mutually
+exclusivivity permits all local operations to occur independently of
+other processes in the system and their current state. For each
 process, there are two data structures from which histories may be
@@
 Algorithm~\ref{alg:local_history_transport}. If the history is
-terminated by the weight cutoff the \textit{historiesCompleted} tally
-is updated and if it hits the boundary of the domain it is added to
-the outgoing history buffer for the neighboring dokmain. If that
-buffer is full, it is sent to the neighboring domain with a
-nonblocking operation and the memory associated with that buffer
-reallocated. In all instances that
-Algorithm~\ref{alg:local_history_transport} is executed,
-\textit{localProcessed} is incremented as a measure of the number of
+terminated by weight cutoff the \textit{historiesCompleted} tally is
+updated. If it hits the boundary of the domain it is added to the
+outgoing history buffer for the neighboring domain and if the buffer
+is full, it is sent to the neighboring domain with a nonblocking
+operation and the memory associated with that buffer reallocated. In
+all instances that Algorithm~\ref{alg:local_history_transport} is
+executed, \textit{localProcessed} is incremented to account for
 histories that have been processed locally.
@@
 
+There is an additional advantage to Brunner and Brantley's work that
+although not immediately applicable to this work has the potential to
+provide value in the future. In addition to a robust fully
+asynchronous communication pattern, this algorith may also be modified
+to account for situations where the total number of histories in a
+given stage are not known before starting transport. From a physics
+perspective, we might expect this for situations where perhaps an
+(n,2n) interaction occurs in a neutronics problem. In this case, the
+algorithm is modified to account for both histories terminated and
+created and several mechanisms are introduced to determine completion
+of the transport stage. For future variations of this work, certain
+variance reduction techniques which create histories, such as
+splitting, have the potential to be successfully employed as a means
+of accelerating the time to solution for a given problem using
+MCSA. The parallel Neumann-Ulam algorithm presented here may be
+adapted to account for these additional techniques.
+
 \clearpage
@@
 that the domain was only decomposed on $O(100)$ processors and if
-replicated $O(1,000)$ times potentially achieves simulation on
-$O(100,000)$ processors, thus providing both spatial and particle
+replicated $O(1,000)$ times potentially achieves efficient simulation
+on $O(100,000)$ processors, thus providing both spatial and particle
 parallelism.
@@
 particle A transport process. Using 0.5 domain overlap also easily
-eliminates cases such as the represented by the path of particle C. In
-this case, particle C is scattering between two adjacent domains,
-incurring a large latency cost for a single particle. Finally, with
-particle B we observe that 0.5 domain overlap will still not eliminate
-all communications. However, if 1 domain overlap were used, the entire
-geometry shown in Figure~\ref{fig:msod_example} would be contained on
-the source processor and therefore transport of all 3 particles
-without communication would occur.
+eliminates cases such as that represented by the path of particle
+C. In this case, particle C is scattering between two adjacent
+domains, incurring a large latency cost for a single
+particle. Finally, with particle B we observe that 0.5 domain overlap
+will still not eliminate all communications. However, if 1 domain
+overlap were used, the entire geometry shown in
+Figure~\ref{fig:msod_example} would be contained on the source
+processor and therefore transport of all 3 particles without
+communication would occur.
 
@@
 overlap from 0 to 3 domain overlap (a $7 \times 7$ box in the context
-of our example) where a domain constituted a fuel assembly. For the
-fully domain decomposed case, they observed that 76.12\% of all source
-particles leave the domain. At 1.5 domain overlap, the percentage of
-source particles born in the center assembly leaving the processor
-domain dropped to 1.05\% and even further for 0.02\% for the 3 domain
-overlap. Based on these results, this overlap approach, coupled with
-the multiple sets paradigm that will scale for existing parallel
-transport algorithms, provides a scalable Monte Carlo algorithm for
-today's modern machines.
-
-\subsection{Generating Overlap }
-\label{subsec:msod_overlap}
-
-\subsection{Generating Multiple Sets }
-\label{subsec:msod_sets}
+of our example) where a domain contained an entire fuel assembly. For
+the fully domain decomposed case, they observed that 76.12\% of all
+source particles leave the domain. At 1.5 domain overlap, the
+percentage of source particles born in the center assembly leaving the
+processor domain dropped to 1.05\% and even further for 0.02\% for the
+3 domain overlap. Based on their results, we hypothesize that the
+overlap approach coupled with the multiple sets paradigm that will
+enhance the scalability of the pure domain-decomposition algorithm for
+Neumann-Ulam presented in the previous section.
+
+\subsection{Generating the MSOD Decomposition }
+\label{subsec:msod_generation}
+In \S~\ref{subsec:domain_generation} we discussed how we generated the
+parallel transport domain from the Neumann-Ulam decomposition of a
+domain decomposed linear operator. We can readily adapt those data
+structures to account for the extra information required by an
+implementation of the MSOD algorithm. First, we consider the
+generation of overlap. Conveniently, this is identical to the
+neighboring states discovery problem discussed in conjuction with
+Figure~\ref{fig:diffusion_graph}. To generate the boundary for the
+local transport domain, we needed to gather all data from the
+immediately adjacent states in the system that did not already exist
+in the local domain. We solved this problem by travsersing the graph
+of the matrix for each of the boundary states and determining which
+processes owned those adjacent states.
+
+To generate overlap, we use the identical algorithm but in this case
+we perform as many graph traversals as the specified amount of
+overlap. For example, consider the analytic relations derived in
+\S~\ref{sec:analytic_framework} for the length of random walks in a
+Neumann-Ulam method and the amount of leakage from a domain
+considering a neutron diffusion problem. From these relations we might
+determine that for our particular problem, if we can grow the domain
+by 10 additional discrete state transitions, we can reduce the number
+of histories that must be comunicated by a significant fraction. We
+determine the information we must gather from neighboring domains to
+build the overlap by traversing the graph of the matrix outward from
+the boundary 10 times. At each step, we find the data in the adjacent
+domain that we require and make that data local, incrementing the size
+of the overap. Once the overlap has been gathered, one extra graph
+traversal on the boundary states is performed to get the new
+neighboring states and their owning process.
 

### 2013-07-02 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
adding parallel mcsa section

@@
 transport or the \textit{localProcessed} count has reached some
-user-define check frequency, lines 18-21 in
+user-defined check frequency, lines 18-21 in
 Algorithm~\ref{alg:local_history_transport} check for more histories
@@
 
+Once overlap has been generated, the transport domain for a single set
+is complete. However, if multiple sets are to be used, an additional
+set of parallel procedures is required to generate the necessary data
+structures. Figure~\ref{fig:msod_construction} gives a schematic
+representation of the MSOD construction process using 4 sets. For a
+problem where $P$ parallel processors are available and $S$ sets are
+to be used, each set is allocated $P/S$ processors for computation. On
+the first $P/S$ processors, the linear problem is generated. From the
+linear problem, the linear operator is used to construct the transport
+domain, the solution vector used to construct the tallies, and the
+forcing term vector used to construct the fixed source. Once these
+Monte Carlo data structures have been generated, they are broadcast to
+the remaining sets in the problem as shown in
+Figure~\ref{fig:msod_construction} such that we have $S$ replications
+of the Monte Carlo problem for a single instance of the original
+linear problem.
+
 \begin{figure}[htpb!]
@@
   \end{center}
-  \caption{\textbf{MSOD construction.} \textit{The linear system is
-      used to construct the Monte Carlo transport domain on the
-      primary set. The Monte Carlo data structures are then broadcast
-      amongst the blocks to all sets in the system.}}
+  \caption{\textbf{MSOD construction for 4 sets with overlap.}
+    \textit{The linear system is used to construct the Monte Carlo
+      transport domain on the primary set. The Monte Carlo data
+      structures are then broadcast amongst the blocks to all sets in
+      the system.}}
   \label{fig:msod_construction}
 \end{figure}
+
+\clearpage
 
@@
 
+With the ability to generate the transport domain within an MSOD
+decomposition as well as the source and tally structures, we can now
+define how to perform a Neumann-Ulam solve using MSOD. Given by
+Algorithm~\ref{alg:msod_transport}, we begin by constructing the Monte
+Carlo data structures for each set including the transport domain, the
+source, and the tallies. Next, in line 4 we perform the asynchronous
+transport algorithm presented in \S~\ref{sec:asynchronous_algorithm}
+for each of the individual sets. Adding overlap to
+Algorithm~\ref{alg:parallel_mc_algorithm} in this case simply modifies
+the set of local states and the set of neighboring states to
+additional states gathered during overlap generation.
+
+\begin{algorithm}[h!]
+  \caption{\textbf{MSOD Transport Sequence}}
+  \label{alg:msod_transport}
+  \begin{algorithmic}[1]
+    \State build MSOD domain
+    \State build Monte Carlo source
+    \State build Monte Carlo tally
+    \State perform parallel Neumann-Ulam transport in each set
+    \State combine set tallies
+    \State combine block tallies
+  \end{algorithmic}
+\end{algorithm}
+
+Once complete, the tallies for each individual computation must be
+gathered to the original set in order to build the correction vector
+within an MCSA sequence. To do this, we perform two parallel vector
+operations. First, overlap not only creates additional states in the
+local transport domain but also additional states in the local tally
+vector such that any events that occur in the local domain may be
+tallied in a local vector. Because of this, the tally vector in a
+single set will share local states with other domains and a parallel
+sum reduction operation is required to transform the tally into the
+original Neumann-Ulam parallel decomposition. Second, the tallies in
+individual sets must be combined and applied to the primary set in the
+problem that shares the processor space with the original linear
+problem. For this operation, we employ the concept of blocks as a
+subdivision of parallel space complementary to the concept of sets.
+
+Consider the schematic representation of the final parallel reduction
+presented in Figure~\ref{fig:msod_tally_reduction}. For this
+operation, the tally vectors in each set must be summed together and
+applied to the primary set. As each set is an exact replica off the
+others, the parallel decomposition of the tally vector is the same
+within all sets. We can take advantage of this by building a processor
+space that encapsulates each identical domain in each set. We can then
+use this processor space for the reduction. For the schematic in
+Figure~\ref{fig:msod_tally_reduction}, the upper-left domain in each
+set creates a single block, giving 9 total blocks for this example
+problem. The processor space the encapsulates each of the upper-left
+domains is then used for the parallel reduction operation. This
+reduction will happen in a mutually exclusive manner for each of the
+blocks in the system. One benefit of this approach is that this
+reduction occurs only over a subset of the processors in the problem,
+providing better scalability than a reduction operation over all
+processors in the problem.
+
 \begin{figure}[htpb!]
@@
   \end{center}
-  \caption{\textbf{MSOD tally reduction.} \textit{The tally computed
-      independently in each set is reduced across the blocks to the
-      primary set. The linear system vector is updated on the primary
-      set. In this example the tally vector in the block containing
-      upper-left subdomains is combined. Identical reduction
-      operations occur for all other blocks simultaneously.}}
+  \caption{\textbf{MSOD tally reduction over blocks.} \textit{The
+      tally computed independently in each set is reduced across the
+      blocks to the primary set. The linear system vector is updated
+      on the primary set. In this example the tally vector in the
+      block containing upper-left subdomains is combined. Identical
+      reduction operations occur for all other blocks
+      simultaneously.}}
   \label{fig:msod_tally_reduction}
@@
 With the parallel adjoint Neumann-Ulam solver implementation described
-aboive, the parallel implementation of the MCSA method is
-trivial. Recall the MCSA iteration procedure outlined in
-Eq~(\ref{eq:mcsa}). In \S\ref{sec:parallel_krylov_methods} we discuss
-parallel matrix and vector operations as utilized in conventional
-Krylov methods. We utilize these here for the parallel MCSA
-implementation. In the first step, a parallel matrix-vector multiply
-is used to apply the split operator to the previous iterate's
-solution. A parallel vector update is then performed with the source
-vector to arrive at the initial iteration guess. In the next step, the
-residual is computed by the same operations where now the operator is
-applied to the solution guess with a parallel matrix-vector multiply
-and then a parallel vector update with the source vector is
-performed. Once the correction is computed with a parallel adjoint
-Neumann-Ulam solve, this correction is applied to the guess with a
-parallel vector update to get the new iteration
-solution. Additionally, as given by
+aboive, the parallel implementation of an MCSA iteration is
+trivial. Recall the MCSA iteration procedure presented again here for
+clarity.
+\begin{subequations}
+  \begin{gather}
+    \label{eq:mcsa_r1}
+    \ve{x}^{k+1/2} = \ve{x}^k + \ve{r}^k\:,\\
+    \label{eq:mcsa_r2}
+    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\
+    \label{eq:mcsa_r3}
+    \ve{A}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:,\\
+    \label{eq:mcsa_r4}
+    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,\\
+    \label{eq:mcsa_r5}
+    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:,
+  \end{gather}
+\end{subequations}
+In \S\ref{sec:parallel_krylov_methods} we discuss parallel matrix and
+vector operations as utilized in conventional Krylov methods. We
+utilize these here for the parallel MCSA implementation for the
+computations required in Eqs~(\ref{eq:mcsa_r1}), (\ref{eq:mcsa_r2}),
+(\ref{eq:mcsa_r4}), and (\ref{eq:mcsa_r5}). In the first step, a
+parallel matrix-vector multiply is used to apply the split operator to
+the previous iterate's solution. A parallel vector update is then
+performed with the source vector to arrive at the initial iteration
+guess. In the next step, the residual is computed by the same
+operations where now the operator is applied to the solution guess
+with a parallel matrix-vector multiply and then a parallel vector
+update with the source vector is performed. Once the correction is
+computed with a parallel adjoint Neumann-Ulam solve, this correction
+is applied to the guess with a parallel vector update to get the new
+iteration solution. Additionally, as given by
 Eq~(\ref{eq:mcsa_stopping_criteria}), 2 parallel vector reductions
@@
 compute the infinity norm of the source vector, and another at every
-iteration to compute the infinity norm of the residual vector. For
-this implementation, all of the issues that will be potentially
-generated by the parallel adjoint solver implementation will manifest
-themselves here as the quality of the correction will be of intense
-study.
+iteration to compute the infinity norm of the residual vector.
+
+To parallelize Eq~(\ref{eq:mcsa_r3}), we apply the MSOD Neumann-Ulam
+algorithm presented in the previous section. The problem is
+potentially replicated with overlap and the Monte Carlo procedure
+returns a correction vector, $\delta \ve{x}$, with the same parallel
+decomposition as the input linear problem. As was the case for
+generating the replications required for a multiple set
+implementation, the explict matrix and vector operations required for
+the rest of an MCSA iteration occur only on the subset of processors
+containing the primary set. Only the Monte Carlo data structures exist
+on the rest of the processors in the problem as replication of the
+other operations does not generate any more information that could be
+used to accelerate the time to solution.
 
@@
 \label{sec:parallel_verification}
+Before we consider the parallel performance of the MCSA algorithm, we
+must first verify that the algorithm generates the correct solution in
+parallel with respect to reference solutions provided by production
+Krylov methods.
 

### 2013-07-02 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
adding titan info and communication parameter explanation

@@
 \label{subsec:titan}
+All parallel scaling studies presented in this work were performed on
+the Titan Cray XK7 machine at the Oak Ridge Leadership Computing
+Facility located at Oak Ridge National
+Laboratory\footnote{Specifications presented for Titan in the section
+  were gathered from the Oak Ridge Leadership Computing Facility at
+  \url{http://www.olcf.ornl.gov/computing-resources/titan-cray-xk7/}}. Titan
+is a heterogeneous architecture machine combining both CPU and GPU
+components to achieve a theoretical peak performance of more than 20
+petaflops. The technical specifications for Titan are given in
+Table~\ref{tab:titan_hardware} and spefically for only the CPU portion
+of the machine.
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lc}\hline\hline
+      Processor & 16 core AMD \\
+      Nodes & 18,688 AMD Opterons \\
+      Cores/Node & 16 \\
+      Total Cores & 299,008 \\
+      Memory/Node & 32 GB \\
+      Memory/Core & 2 GB \\
+      Interconnect & Gemini \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Titan Cray XK7 hardware specifications.}}
+  \label{tab:titan_hardware}
+\end{table}
+
+For the scaling studies presented in this work, only the CPU
+components specified in of Titan were utilized. This is due to the
+fact that the implementations of the parallel MCSA algorithm utilized
+code written only with the Message Passing Interface (MPI) and
+therefore could not take advantage of the GPU accelerators. Even
+considering just the CPU portion of titan, nearly 300,000 CPUs are
+available to test the parallel performance of the algorithm at high
+levels of concurrency.
+
+For each parallel performance metric presented in the previous
+section, a reference case is required to compute efficiencies for all
+other cases. For smaller machines, often a serial computation can be
+performed and used as the base case to get a true measure of speed-up
+for a particular problem. For our scaling studies, we will always use
+as a reference case a calculation which uses 16 cores, filling an
+entire node of the machine. We fill an entire node so that we may
+ignore the effects on scalibility measures that arise from using only
+a subset of cores in a node. Typically this results in a large drop in
+efficiency from a serial computation to a calculation that fills the
+entire node. Once a node is filled, adding subsequent nodes has less
+of an effect on the scaling, providing a more stable region of
+analysis.
 
@@
 
+In our description of Brunner and Brantley's algorithm as applied to
+the Neumann-Ulam method, we noted two free parameters that are
+user-defined at runtime. The first is the static data buffer size for
+communicating histories between neighboring domains and the second is
+the frequency by which a domain checks for incoming information from
+neighboring domain, both of which may potentially affect latency and
+load balancing. Changing the static history buffer size parameter
+affects the scalability of the algorithm in two ways. If the buffer
+size is increased, the buffer takes longer to fill and is therefore
+sent less frequently, potentially causing scalability bottlenecks
+where a neighboring domain is idle and waiting for more work to do but
+is waiting for another domain to complete work. In addition, if the
+buffer size is too large, the network may become saturated with these
+very large data packets, slowing down performance. If the buffer size
+is too small, too many history messages are sent, not providing enough
+data for a neighboring process to work on to make up for the time
+spent communicating. Considering the second parameter, when the check
+frequency period is increased, more histories are processed locally
+before looking to do work elsewhere. If this value becomes too large,
+incoming data buffers may not be cleared frequently enough and cause a
+bottle neck in the domain-to-domain communications. If this value is
+too small, too much time is spent looking for work instead of actually
+performing it.
+
+To assess how these parameters affect the performance of the parallel
+Monte Carlo algorithm on the Titan machine, a parameter study was
+performed using the neutron diffusion problem on 64 cores (4
+nodes). Each parameter was varied between 16 and 4096 and the time to
+solve the problem recorded. The results of this study are presented in
+Figure~\ref{fig:titan_comm_parameters}. Compared to the results of
+Brunner and Brantley for the same type of study, we note qualitatively
+similar results with the general observation that a larger history
+buffer check period is better than a shorter one. The larger the check
+period, the more on-process work is done before checking
+messages. Second, there was not a significant variation observed in
+the history buffer size parameter for a given check
+frequency.
+
 \begin{figure}[htpb!]
@@
   \caption{\textbf{Neumann Ulam parallel algorithm free parameter
-      sensitivity.}  \textit{All runtimes were within 6.5\% of the
-      fastest observed case.}}
+      sensitivity.} \textit{The color scale represents the ratio of
+      the runtime of a particular parameter variation with the runtime
+      from the faster parameter variation. A value of 1.02 on the
+      color scale corresponds to a runtime 2\% longer than the faster
+      time. All runtimes were within 6.5\% of the fastest observed
+      case.}}
   \label{fig:titan_comm_parameters}
 \end{figure}
+
+Compared to the reference work, there was also not a significant
+change in the runtime over the values of parameters tested with all
+runtimes reported within 6.5\% of the fastest case observed. For this
+work, a history buffer size of 4096 and a buffer check period of 128
+performed best for the intitial calculations. However, most values in
+the range tested will give good performance. Brunner and Brantley
+found that a value of 1024 was best for both parameters. For the
+results observed here, these parameters were only 0.7\% slower than
+the fastest case. It may be best practice to fix these parameters
+internally at 1024 as both this work and the reference found these to
+perform well several years apart on both different parallel machines
+and for different transport sequences.
 

### 2013-07-02 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
updating figure locations in parallel chapter

@@
 the model problem presented in Appendix~\ref{chap:diffusion_problem}.
-\begin{figure}[htpb!]
+\begin{figure}[t!]
   \begin{center}
@@
 \label{sec:leadership_scaling_studies}
-For all cases, we explore scaling for just the Monte Carlo transport
-kernel and MCSA as a whole iterative scheme. This will be necessary to
-determine if the initial relaxation step required at each iteration is
-a limiting factor in convergence. For each case, it will be
-constructive to study the scaling performance as a function of
-spectral radius - dictated by the cross sections and grid size to
-diffusion length ratio. In addition, all of these studies should
-hopefully have data from both the cluster and Titan and should all be
-compared to Aztec GMRES and CG implementations with appropriate
-preconditioning.
+Using the parallel performance metrics presented in the previous
+section, we will next performance various scaling studies using an
+implementation of the parallel MCSA algorithm to assess the quality of
+the algorithm. For these scaling studies, large problems at high
+levels of concurrency will be solved at a leadership-class computing
+facility. In this section, we will outline the parallel machine on
+which the calculations were performed and then perform a series a
+studies that identify both strengths and weaknesses of the parallel
+algorithm. Both pure domain decomposition and MSOD parallel
+performance will be explored and compared to performance results for
+the same problems using a set of production parallel Krylov solvers.
 
@@
   \end{center}
-  \caption{\textbf{Neumann Ulam parallel algorithm free parameter
-      sensitivity.} \textit{The color scale represents the ratio of
-      the runtime of a particular parameter variation with the runtime
-      from the faster parameter variation. A value of 1.02 on the
-      color scale corresponds to a runtime 2\% longer than the faster
-      time. All runtimes were within 6.5\% of the fastest observed
-      case.}}
+  \caption{\textbf{Neumann Ulam parallel algorithm history buffer
+      check period vs. history buffer size sensitivity.} \textit{The
+      color scale represents the ratio of the runtime of a particular
+      parameter variation with the runtime from the faster parameter
+      variation. A value of 1.02 on the color scale corresponds to a
+      runtime 2\% longer than the faster time. All runtimes were
+      within 6.5\% of the fastest observed case.}}
   \label{fig:titan_comm_parameters}
@@
 perform well several years apart on both different parallel machines
-and for different transport sequences.
+and for different transport sequences. Therefore, for the scaling
+studies presented in the following sections, values of 1024 for both
+the history buffer size and message check frequency will be used.
 

### 2013-07-03 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
adding pure domain decomposition write up

@@
 In \S\ref{sec:parallel_krylov_methods} we discuss parallel matrix and
-vector operations as utilized in conventional Krylov methods. We
+vector operations as utilized in conventional projection methods. We
 utilize these here for the parallel MCSA implementation for the
@@
 
-To parallelize Eq~(\ref{eq:mcsa_r3}), we apply the MSOD Neumann-Ulam
-algorithm presented in the previous section. The problem is
-potentially replicated with overlap and the Monte Carlo procedure
-returns a correction vector, $\delta \ve{x}$, with the same parallel
-decomposition as the input linear problem. As was the case for
-generating the replications required for a multiple set
+To parallelize the solution of Eq~(\ref{eq:mcsa_r3}), we apply the
+MSOD Neumann-Ulam algorithm presented in the previous section. The
+problem is potentially replicated with overlap and the Monte Carlo
+procedure returns a correction vector, $\delta \ve{x}$, with the same
+parallel decomposition as the input linear problem. As was the case
+for generating the replications required for a multiple set
 implementation, the explict matrix and vector operations required for
@@
 the number of processes in the calculation. In this case, as $P$
-increases, $N$ decreases, thus reducing performance by having a larger
-ratio of communication to computational work. This type of scaling is
-important because typically the global problem size is fixed due to
-the problem of interest. Consider the desire for a high fidelity
-neutronics simulation of the entire core of a nuclear reactor. In this
-case, the global problem is known from the geometry definition
-including the number of fuel assemblies and the geometry of those
-assemblies all the way down to the resolution of the space, angle, and
-energy discretizations required in order to capture physical phenomena
-of interest. Because the global problem size is known a priori, strong
-scaling is an important measure of how useful a solution technique is
-when larger and larger values of $P$ are used to solve the
-problem. Ideally, the time to solution for the the problem will
-decrease linearly as a function of $P$, permitting a faster time to
-solution using the maximum available resources thus making sure all
-$P$ are positively affecting the runtime of the problem.
+increases, $N$ is fixed, thus reducing the local problem size and
+performance by having a larger ratio of communication to computational
+work. This type of scaling is important because typically the global
+problem size is fixed due to the problem of interest. Consider the
+desire for a high fidelity neutronics simulation of the entire core of
+a nuclear reactor. In this case, the global problem is known from the
+geometry definition including the number of fuel assemblies and the
+geometry of those assemblies all the way down to the resolution of the
+space, angle, and energy discretizations required in order to capture
+physical phenomena of interest. Because the global problem size is
+known a priori, strong scaling is an important measure of how useful a
+solution technique is when larger and larger values of $P$ are used to
+solve the problem. Ideally, the time to solution for the the problem
+will decrease linearly as a function of $P$, permitting a faster time
+to solution using the maximum available resources and thus making sure
+all $P$ are positively affecting the runtime of the problem.
 
 For strong scaling, the objective is to then decrease the runtime of a
-fixed size global problem as a linear function of $P$. We can then
-define the strong scaling \textit{absolute efficiency} as:
+fixed size global problem as a linear function of $P$. Based on this,
+we can define the strong scaling \textit{absolute efficiency} as:
 \begin{equation}
@@
 implementation of the parallel MCSA algorithm to assess the quality of
-the algorithm. For these scaling studies, large problems at high
-levels of concurrency will be solved at a leadership-class computing
-facility. In this section, we will outline the parallel machine on
-which the calculations were performed and then perform a series a
-studies that identify both strengths and weaknesses of the parallel
-algorithm. Both pure domain decomposition and MSOD parallel
-performance will be explored and compared to performance results for
-the same problems using a set of production parallel Krylov solvers.
+the algorithm using the neutron diffusion problem presented in
+Appendix~\ref{chap:diffusion_problem}. For these scaling studies,
+large problems at high levels of concurrency will be solved at a
+leadership-class computing facility. In this section, we will outline
+the parallel machine on which the calculations were performed and then
+present a series of numerical studies that identify both strengths and
+weaknesses of the parallel algorithm. Both pure domain decomposition
+and MSOD parallel performance will be explored and compared to
+performance results for the same problems using a set of production
+parallel Krylov solvers. For the scaling results presented here, all
+parallel Monte Carlo results were computed using the Monte Carlo
+Linear Solvers\footnote{MCLS is an open source library and can be
+  downloaded at \url{https://github.com/sslattery/MCLS}} (MCLS)
+library generated specifically for this work while the conjugate
+gradient and GMRES Krylov solvers used are part of the Belos linear
+solver package within the Trilinos framework
+\citep{heroux_overview_2005}. In general, the Belos package produces
+solutions to the neutron diffusion problem over an order of magnitude
+faster than the MCLS library for serial computations. Therefore, when
+we directly compare these two codes, the MCLS parallel efficiencies
+will appear artificially inflated relative to the Belos efficiencies
+due to the fact that the MCLS ratio of work to communication will be
+much larger than that for the Belos calculations.
 
@@
 the Titan Cray XK7 machine at the Oak Ridge Leadership Computing
-Facility located at Oak Ridge National
-Laboratory\footnote{Specifications presented for Titan in the section
-  were gathered from the Oak Ridge Leadership Computing Facility at
-  \url{http://www.olcf.ornl.gov/computing-resources/titan-cray-xk7/}}. Titan
-is a heterogeneous architecture machine combining both CPU and GPU
+Facility located at Oak Ridge National Laboratory. Titan is a
+heterogeneous architecture machine combining both CPU and GPU
 components to achieve a theoretical peak performance of more than 20
 petaflops. The technical specifications for Titan are given in
-Table~\ref{tab:titan_hardware} and spefically for only the CPU portion
-of the machine.
+Table~\ref{tab:titan_hardware} and specifically for only the CPU
+portion of the machine.
+
 \begin{table}[h!]
@@
   \end{center}
-  \caption{\textbf{Titan Cray XK7 hardware specifications.}}
+  \caption{\textbf{Titan Cray XK7 hardware specifications.}
+    \textit{Specifications presented for Titan in this table were
+      gathered from the Oak Ridge Leadership Computing Facility at}
+    \url{http://www.olcf.ornl.gov/computing-resources/titan-cray-xk7/}}
   \label{tab:titan_hardware}
@@
 For the scaling studies presented in this work, only the CPU
-components specified in of Titan were utilized. This is due to the
-fact that the implementations of the parallel MCSA algorithm utilized
-code written only with the Message Passing Interface (MPI) and
-therefore could not take advantage of the GPU accelerators. Even
-considering just the CPU portion of titan, nearly 300,000 CPUs are
-available to test the parallel performance of the algorithm at high
-levels of concurrency.
+components of Titan specified in Table~\ref{tab:titan_hardware} were
+utilized. This is due to the fact that the implementations of the
+parallel MCSA algorithm utilized code written only with the Message
+Passing Interface (MPI) and therefore could not take advantage of the
+GPU accelerators. Even considering just the CPU portion of Titan,
+nearly 300,000 CPUs are available to test the parallel performance of
+the MCSA algorithm at high levels of concurrency.
 
@@
 size is increased, the buffer takes longer to fill and is therefore
-sent less frequently, potentially causing scalability bottlenecks
+sent less frequently, potentially causing load balancing bottlenecks
 where a neighboring domain is idle and waiting for more work to do but
@@
 data for a neighboring process to work on to make up for the time
-spent communicating. Considering the second parameter, when the check
-frequency period is increased, more histories are processed locally
-before looking to do work elsewhere. If this value becomes too large,
-incoming data buffers may not be cleared frequently enough and cause a
-bottle neck in the domain-to-domain communications. If this value is
-too small, too much time is spent looking for work instead of actually
-performing it.
+spent communicating. Considering the second parameter, when the
+message check frequency period is increased, more histories are
+processed locally before looking to do work elsewhere. If this value
+becomes too large, incoming data buffers may not be cleared frequently
+enough and cause decreased performance in the domain-to-domain
+communications. If this value is too small, too much time is spent
+looking for work instead of actually performing it.
 
@@
 performed using the neutron diffusion problem on 64 cores (4
-nodes). Each parameter was varied between 16 and 4096 and the time to
-solve the problem recorded. The results of this study are presented in
+nodes). This problem had a reasonably well conditioned spectral radius
+of 0.78 with a global problem size of \sn{1.6}{7} DOFs with one
+stochastic history used per DOF in the MCSA Neumann-Ulam used to
+compute the correction for the acceleration. Each parameter was varied
+between 16 and 4096 and the time to solve the problem recorded. The
+results of this study are presented in
 Figure~\ref{fig:titan_comm_parameters}. Compared to the results of
@@
 messages. Second, there was not a significant variation observed in
-the history buffer size parameter for a given check
-frequency.
+the history buffer size parameter for a given check frequency.
 
@@
 
-\subsection{Pure Domain Decomposition}
+\subsection{Pure Domain Decomposition Scaling}
 \label{subsec:pure_domain_decomp}
 
-\subsubsection{Strong Scaling}
-\label{subsubsec:pure_strong}
+For the first set of scaling studies, we consider the parallel MCSA
+algorithm where MSOD has not been used in the parallel Neumann-Ulam
+algorithm. We refer to this case as pure domain decomposition where no
+overlap or replication has been leveraged. Such a test will isolate
+specifically the performance of Brunner and Brantley's algorithm as
+applied to the Neumann-Ulam algorithm. For the strong scaling study,
+the global problem size was fixed at \sn{1.6}{7} DOFs for all solvers
+and the 16 core run used as the base case. MCSA was used again with
+one stochastic history used per DOF in the Neumann-Ulam solve to
+compute the correction for the
+acceleration. Figure~\ref{fig:titan_pure_strong} gives the results of
+the this study with the given absolute efficiencies computed using
+Eq~(\ref{eq:strong_scaling_absolute}).
 
@@
   \caption{\textbf{Pure domain decomposition strong scaling absolute
-      efficiency.}  \textit{Super-linear speed-up from memory
-      thrashing in base case. MCLS is an order of magnitude slower
+      efficiency.}  \textit{Super-linear speed-up is from memory
+      thrashing in the base case. MCLS is an order of magnitude slower
       arithmetically, causing the higher efficiency.}}
@@
 \end{figure}
+
+There are several key features of Figure~\ref{fig:titan_pure_strong}
+worth noting. First, up to around 5,000 cores, all methods experience
+a super-linear improvement in parallel efficiencies. For all methods,
+this increase is not a function of the algorithmic efficiency as those
+computed using Eq~(\ref{eq:algorithmic_efficiency}) remained at unity,
+meaning that all computations converged in the same number of
+iterations relative to the base case\footnote{When the algorithmic
+  efficiency remains at unity for all calculations, the implementation
+  efficiency is equivalent to the absolute efficiency and therefore is
+  not presented here.}. Instead, what is happening here and was
+observed by Brunner and Brantley in their work is that the large
+problem used to scale to O(10,000) cores was large enough that memory
+thrashing was occuring in the 16 core base case. As the number of
+cores in the strong scaling study was increased and the local problem
+size decreased, more of the local problem fit into the cache of the
+system, thus causing improved serial performance and therefore
+enhanced runtimes. In this region of super-linear efficiencies, we
+also observe larger efficiencies for GMRES as compared to conjugate
+gradient. This occurs due to the fact that conjugate gradient has a
+static set of operations performed at every iteration while GMRES
+operates on a continually growing subspace. In addition, the
+orthogonalization operations required to generate that subspace
+consume much more time than the basic conjugate gradient operations,
+causing a larger serial runtime and subsequently higher
+efficiencies. For the same reasons, we also observe higher
+efficiencies for the MCSA implementation due to the serial performance
+of the implementation.
+
+Beyond around 5,000 cores, the global problem size selected for this
+strong scaling study causes the local problem size to shrink enough
+that parallel communication costs begin to overtake arithmetic costs
+on process. This takeover causes this effective strong scaling wall
+that is observed in all methods. Using any more cores to solve this
+particular global problem is not an efficient use of those parallel
+resources. Furthermore, the fact that we see a very similar trends in
+efficiencies for all methods means that they are bound by the same
+parallel performance limitations in a strong scaling environment. This
+is important as we can expect the same qualitative scaling performance
+from MCSA onc arithmetic optimization has been performed to improve to
+serial performance of the implementation. In addition, these results
+are valuable in that the show MCSA, when properly parallelized, has
+the ability to be competetive on leadership-class computing platforms
+with conventional subspace methods with good strong scaling observed
+in the pure domain decomposition case to O(10,000) cores.
+
+Next, we consider the weak scaling performance of the algorithm. In
+this case, we fix the local problem size at \sn{4}{4} DOFs and
+increase the number of processors used in the
+simulation. Figure~\ref{fig:titan_weak_absolute} gives the weak
+scaling absolute efficiencies for this problem as computed by
+Eq~(\ref{eq:weak_scaling_absolute}). In general, the relative trends
+when comparing MCSA to the Krylov methods are the same with MCSA and
+GMRES scaling better than the conjugate gradient method due to the
+increased amount of arithmetic work on-process. In addition, it is
+important to note that the weak scaling of MCSA when properly
+parallelized remains competetive with the Krylov methods keeping in
+mind the artificial inflation of efficiencies.
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_absolute.pdf}
+  \end{center}
+  \caption{\textbf{Pure domain decomposition weak scaling absolute
+      efficiency.}}
+  \label{fig:titan_weak_absolute}
+\end{figure}
+
+As an additional means of comparison between MCSA and the Krylov
+methods, we compute the algorithmic efficiencies of the calculations
+using Eq~(\ref{eq:algorithmic_efficiency}). As shown in
+Figure~\ref{fig:titan_weak_algorithmic}, the algorithm efficiency of
+the Krylov algorithm improves as a function of the number of cores in
+the problem due to the fact that the number of iterations required to
+converge decreased from 20 in the base case to 19 and then to 18 for
+the final set of computations. When this reduction in iterations is
+used to account for the observed runtimes and computed absolute
+efficiencies for weak scaling we end up with the implementation
+efficiencies computed using Eq~(\ref{eq:implementation_efficiency})
+given in Figure~\ref{fig:titan_weak_implementation}. Because MCSA
+maintains the same iterative performance as a function of global
+problem size in the weak scaling study, the implementation
+efficiencies compare more favorably to the Krylov methods. In effect,
+we are measuring the parallel performance of a single iteration for
+each method by looking at these implementation efficieinces. As the
+global problem size increases, MCSA is less prone to a drop in
+parallel efficiency. When optimization has been completed and runtimes
+potentially approach those of the Krylov methods, this comparison may
+not be as favorable. Overall, weak scaling performance the parallel
+MCSA algorithm is excellent for the pure domain decomposition case
+with absolute parallel efficiencies of over 80\% observed at over
+32,000 cores.
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_alg_eff.pdf}
+  \end{center}
+  \caption{\textbf{Pure domain decomposition weak scaling algorithmic
+      efficiency.}}
+  \label{fig:titan_weak_algorithmic}
+\end{figure}
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_implementation.pdf}
+  \end{center}
+  \caption{\textbf{Pure domain decomposition weak scaling
+      implementation efficiency.}}
+  \label{fig:titan_weak_implementation}
+\end{figure}
+
+In addition to comparing MCSA scaling to conventional methods, we also
+motivate avoiding any type of master/slave algorithms at these high
+levels of concurrency by comparing the binary communication tree
+implementation used in the previous results with a parallel MCSA
+implementation that uses the master/slave scheme to determine the end
+of a history transport stage as shown in
+Figure~\ref{fig:master_comm_tree}
+instead. Figure~\ref{fig:titan_strong_bvsm} provides this comparison
+using the strong scaling study while Figure~\ref{fig:titan_weak_bvsm}
+gives the results of the comparison for the weak scaling study. In
+both cases, the deficiencies caused by using the master/slave scheme
+for transport completion are obvious. At around 1,000 cores, both the
+strong and weak scaling performance seriously degrades while the
+binary tree pattern maintains execellent performance. At this number
+of cores, any extra work given to the master process, even if it
+amounts to summing a handful of integers and checking for a few extra
+messages as in this algorithm, this extra work grows enough to create
+serious load imbalance and reduce performance.
 
@@
   \label{fig:titan_strong_bvsm}
-\end{figure}
-
-\clearpage
-
-\subsubsection{Weak Scaling}
-\label{subsubsec:pure_weak}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_absolute.pdf}
-  \end{center}
-  \caption{\textbf{Pure domain decomposition weak scaling absolute
-      efficiency.}}
-  \label{fig:titan_weak_absolute}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_alg_eff.pdf}
-  \end{center}
-  \caption{\textbf{Pure domain decomposition weak scaling algorithmic
-      efficiency.}}
-  \label{fig:titan_weak_algorithmic}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_implementation.pdf}
-  \end{center}
-  \caption{\textbf{Pure domain decomposition weak scaling
-      implementation efficiency.}}
-  \label{fig:titan_weak_implementation}
 \end{figure}

### 2013-07-03 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
adding write up for multiple set calculations

@@
 scattering event in the new domain may be processed, particle A must
-vbe communicated between between the two parallel processes that own
+be communicated between between the two parallel processes that own
 those domains. For any given particle in the transport simulation,
@@
   \eta_{impl}(N,P|Q) = \frac{\eta}{\eta_{alg}}\:,
-  \label{eq:implementation efficiency}
+  \label{eq:implementation_efficiency}
 \end{equation}
@@
 entire node of the machine. We fill an entire node so that we may
-ignore the effects on scalibility measures that arise from using only
+ignore the effects on scalability measures that arise from using only
 a subset of cores in a node. Typically this results in a large drop in
@@
 
-\subsubsection{Strong Scaling}
-\label{subsubsec:overlapping_strong}
-
-\clearpage
-
-\subsubsection{Weak Scaling}
-\label{subsubsec:overlapping_weak}
-
 \clearpage
@@
 
-\subsubsection{Strong Scaling}
-\label{subsubsec:msod_strong}
+Given that overlap was not successful in boosting parallel
+efficiencies, we next look to apply replication to the problem in the
+form of multiple sets. Before performing parallel scaling studies
+using multiple sets, several aspects of the replicated parallel
+problem must be considered in order to effectively craft the studies
+and interpret their results. First is the question of how many
+stochastic histories to use to generate the correction vector in the
+MCSA iteration. For this work we consider two choices, the split case
+and the fully replicated case. As an example, consider a single set
+problem where 20,000 histories are used in the Monte Carlo solve. If
+we choose the split case, then solving the same problem with 2 sets
+would give 10,000 histories in each set and 4 sets would give 5,000
+histories in each set. Splitting the histories in this way provides
+the same global number of histories used to compute the correction in
+the single set problem. We expect to reduce the time to solution by
+doing this as each individual set solve will occur faster in an almost
+linear fashion with the reduced number of histories. For the
+replicated history case, each additional set in the problem will use
+just as many histories as the single set case. Using 2 sets with fully
+replicated histories for our example problem then has each set
+computing 20,000 histories with 40,000 global histories in the
+computation. For the 4 set case, the number of global histories is
+increased to 80,000. We also expect this type of replication to reduce
+the time to solution as the extra global histories should reduce the
+stochastic uncertainty in the correction vector and reduce the overall
+number of MCSA iterations required to converge.
+
+For the strong scaling study, we apply the additional sets as
+follows. Using the same problems as the pure domain decomposition
+scaling study, the multiple sets are used to replicate those problems
+and boost the core count. For example, for the 4 set case, the data
+point at 64 cores has the same $N/P$ ratio as the single set problem
+with 16 cores. This is due to the fact that $N$ is fixed for the
+strong scaling study and 4 replications gives a set size of 16
+cores. For the 2 set case, the 32 core data point is solving the same
+$N/P$ problem as the 16 core base case. In addition to considering the
+replicated domain, we must also consider how to modify the
+efficiencies we will compute. For the strong scaling case, recall that
+the objective is to decrease the time to solution for a given global
+problem size $N$ by increasing $P$. Therefore, we may use
+Eqs~(\ref{eq:strong_scaling_absolute_ref}),
+(\ref{eq:algorithmic_efficiency}), and
+(\ref{eq:implementation_efficiency}) to compute the various strong
+scaling efficiencies, but we must modify the reference computation in
+each of these formulas to reflect our objective. For all efficiency
+computations, the 16 core single set computation will be used as the
+reference case. By adding more cores to the problem, either by
+replicating with multiple sets, or decreasing the local problem size,
+any computational resources added to the problem should aim to result
+in speedup. If this speedup for one technique is less than another,
+then that technique is deemed a less efficient use of those
+computational resources.
+
+For the multiple sets problem, the single set strong scaling study
+from the pure domain decomposition base case was repeated for multiple
+sets with both splitting of histories across sets and full replication
+of histories across sets. Figure~\ref{fig:titan_strong_ms_time} gives
+the actual wall times for these computations. There are two important
+features to note in the wall time data. First, splitting the histories
+amongst sets instead of replicating them is faster due to the fact
+that cutting the number of histories in half reduces the time of an
+iteration by a factor that is nearly linear while increasing linearly
+the number of histories through replication does not decrease the
+iterations required to converge by a linear factor. Second, until the
+strong scaling wall is hit starting at around 10,000 cores, the single
+set case actually perfoms the fastest with only marginal improvement in
+runtimes at higher numbers of cores with more sets. At these larger
+numbers of cores, however, the improvements are so modest that one is
+better off running the same problem using a single set at lower core
+counts to conserve computational resource while maintaining a good tim
+to solution.
 
@@
   \end{center}
-  \caption{\textbf{Wall time in seconds to solution for each case.}}
+  \caption{\textbf{Wall time in seconds to solution for each case for
+      the strong scaling study with multiple sets.}}
   \label{fig:titan_strong_ms_time}
 \end{figure}
+
+From this timing data we then compute the absolute efficiencies for
+strong scaling as given in Figure~\ref{fig:titan_strong_ms_eff}. As
+observed in the timing data, because the single set case had the
+fastest run times up to the strong scaling wall it also has the best
+observed efficiencies. For the 2 set case with history splitting,
+although the time to perform an iteration was nearly cut in half, the
+extra parallel overhead associated with reducing the tally across the
+blocks to combine the set results increases run times and reduces
+efficiencies.This reduction across blocks is even more obvious for the
+4 set case. For the replicated cases, we do observe a flat region of
+parallel efficiencies at larger core counts (around 32,000 for the 4
+set case with history replication), but the additional cores do not
+reduce the runtime of the global problem efficiently. Ultimately, all
+cases effectively hit the strong scaling wall at the same time. We
+expect this if we consider the fact that adding sets to the problem is
+in itself effectively a form of strong scaling. For any given problem
+of fixed size $N$, adding extra sets to the computation decreases the
+ratio of $N/P$ as $P$ will grow with the addition of sets. For the
+split history case, the splitting further exacerbates this situation
+by further decreasing the amount of histories that will be computed in
+the local domain, giving a larger ratio of communication to work in
+the Monte Carlo solve. As with with single set results, we also
+observe some efficiencies greater than unity for the multiple set
+cases also due to improved cache utilization.
 
@@
 
+Also of interest here is the effect of splitting histories amongst
+sets versus replicating them on the iterative performance of the
+method. Table~\ref{tab:ms_strong_alg_eff} gives the iterations
+required to converge and computed algorithmic efficiences for each of
+the cases presented in the scaling data. As expected, splitting
+histories maintains a fixed global number of histories and therefore
+maintained the number of iterations required to converge relative to
+the single set base case. When histories were replicated, adding sets
+decreased the number of iterations required to converge in a nonlinear
+fashion. Using these algorithmic efficiencies, the implementation
+efficiencies for the multiple set cases were computed and are given in
+Figure~\ref{fig:titan_strong_ms_impeff}. From these results, we see
+that the replicated history cases are even less efficient than in the
+absolute case. Not only do we have the additional parallel overhead of
+doing the block level reduction, but we also have additional histories
+that increase the time to solution for a single iteration. Given that
+the time to solution is not efficiently improved using multiple sets
+in both the split and replicated history cases, multiple set domain
+decomposition is not an effective means to boost the strong scaling
+efficiencies of MCSA.
+
 \begin{table}[h!]
@@
     \begin{tabular}{lcc}\hline\hline
-      \multicolumn{1}{c}{Case}& 
+      \multicolumn{1}{l}{Case}& 
       \multicolumn{1}{c}{Iterations}&
@@
   \caption{\textbf{Strong scaling implementation efficiency relative
-      to 16-core 1-set base case.}}
+      to the 16-core 1-set base case.}}
   \label{fig:titan_strong_ms_impeff}
@@
 
-\clearpage
-
-\subsubsection{Weak Scaling}
-\label{subsubsec:msod_weak}
+Next, we continue to a weak scaling study of the multiple sets
+problem. As with the strong scaling study, we will modify the pure
+domain decomposition study to account for the additional
+replication. For each calculation in the single set study, that
+problem will be replicated either 2 or 4 times with both split and
+replicated histories to assess the effects on scalability and
+runtime. Figure~\ref{fig:titan_weak_ms_time} gives the wall times from
+the multiple sets weak scaling study. Of primary importance here is
+the fact that adding sets to the problem with both split and
+replicated histories actually decreases the time to solution. This was
+not observed in the strong scaling case due to the larger decrease in
+computational work as the number of cores was increased. This result
+is significant in that it shows a solution technique in which the
+linear problem can actually be replicated and enhance time to
+solution, something that cannot be acheived with conventional
+methods. At larger core counts, however, this reduction in compute
+time is not as great as at lower core counts due to the fact that the
+tally reduction across blocks requires more resources. In addition, at
+over 130,000 cores for the largest 4 set computation, load balance is
+likely becoming an issue due to the linear system and subsequent
+parallel matrix and vector operations only occuring on a subset of the
+entire processor space of the problem.
 
@@
   \end{center}
-  \caption{\textbf{Wall time in seconds to solution for each case.}}
+  \caption{\textbf{Wall time in seconds to solution for each case for
+      the weak scaling study with multiple sets.}}
   \label{fig:titan_weak_ms_time}
@@
 
-For a multiple set analysis of the weak scaling results, we have to
-consider that adding sets to the problem is actually a form of strong
-scaling. Consider a single set problem ($S=1$) for a given $P$ and
-$N$. When that problem is now solved using 2 sets we have $S=2$, $N$
-and $2P$. For any number of sets we then have a global problem size of
-$N$ while the number of processors is $SP$ where $P$ is the number of
-processors in the single set case. Therefore, $N/P$ is no longer fixed
-in the weak scaling study but rather $N/(SP)$ is fixed. Therefore, we
-modify the weak scaling absolute efficiency relationship to account
-for this fact:
+For a multiple set analysis of the weak scaling timing results, we
+again have to consider that adding sets to the problem is actually a
+form of strong scaling. Consider a single set problem ($S=1$) for a
+given $P$ and $N$. When that problem is now solved using 2 sets we
+have $S=2$, $N$ and $2P$. For any number of sets we then have a global
+problem size of $N$ while the number of processors is $SP$ where $P$
+is the number of processors in the single set case. Therefore, $N/P$
+is no longer fixed in the weak scaling study but rather $N/(SP)$ is
+fixed. As a result, we modify the weak scaling absolute efficiency
+relationship to account for this fact:
 \begin{equation}
@@
 where now the absolute efficiency is a function of the number of sets
-in the problem. For problems where adding sets improves time to
-solution, this new weak scaling formulation accounts for how
-efficiently those extra resources are incorporated into the
-problem. Without such a modification, the largest case run for the
-split 4 set case would have a computed weak scaling efficiency of
-nearly 100\% using the standard formula for absolute
+in the problem. Interestingly, Eq~(\ref{eq:ms_weak_efficiency}) has a
+very similar form to the strong scaling efficiency given by
+Eq~(\ref{eq:strong_scaling_absolute}), representative of the strong
+scaling nature of adding additional sets. For problems where adding
+sets improves time to solution, this new weak scaling formulation
+accounts for how efficiently those extra resources are incorporated
+into the problem. Without such a modification, the largest case run
+for the split 4 set case would have a computed weak scaling efficiency
+of nearly 100\% using the standard formula for absolute
 efficiency. However, using 4 times as many cores as the single set
@@
 
+Figure~\ref{fig:titan_weak_ms_eff} gives the results of the weak
+scaling efficiencies using the multiple set metric given by
+Eq~(\ref{eq:ms_weak_efficiency}). Although we do reduce the time to
+solution over the base cases for nearly all data points, the
+observation here is that the reduction in time to solution is still
+not large enough to efficiently use all resources. For example, for
+the 2 set case to be a perfect use of resources with either split or
+replicated histories, the runtime of that calculation must be half of
+that for the single set case. For the 4 set case, this run time should
+be a quarter of the single set case for perfect scaling. This was not
+the observation for the runtimes in
+Figure~\ref{fig:titan_weak_ms_time} and therefore the single set case
+is still the most efficient use of computational resources in this
+scaling study.
+
 \begin{figure}[t!]
@@
 
+As in the strong scaling case, splitting and replicating histories
+across sets had effectively the same results for the weak scaling
+study with the iterations to converge and compute algorithmic
+efficienies given by Table~\ref{tab:ms_weak_alg_eff}. Again, when the
+implementation efficiencies are computed, we see in
+Figure~\ref{fig:titan_weak_ms_impeff} that splitting histories amongst
+sets is still a more efficient manner of introducing multiple sets
+rather than replicating histories in an attempt to reduce the number
+of iterations required to converge.  Here we also note that this is
+not a true weak scaling study as indicated by
+Eq~(\ref{eq:ms_weak_efficiency}) but rather an odd combination of both
+weak and strong scaling. Regardless of this, although multiple sets
+are not the most efficient use of resources in this analysis, they do
+enhance the time to solution through replication and do so
+significantly at core counts of O(1,000) and lower. In both the weak
+and strong scaling cases, due to the additional computational and
+communication overhead of adding sets to the system, one should never
+expect to boost parallel efficiencies for MCSA using this
+technique. Further, in a purely strong scaling environment, no
+performance gains should be expected as adding sets creates an even
+more parasitic ratio of $N/P$. In a weak scaling environment, although
+no efficiency increase will be noted, one can expect an enhancement of
+time to convergence by replicating.
+
 \begin{table}[h!]
@@
 \end{figure}
-
-\subsubsection{Strong Scaling}
-\label{subsubsec:full_clip_strong}
-
-\clearpage
-
-\subsubsection{Weak Scaling}
-\label{subsubsec:full_clip_weak}

### 2013-07-03 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
adding more to the multiple sets analysis

@@
       thrashing in the base case. MCLS is an order of magnitude slower
-      arithmetically, causing the higher efficiency.}}
+      arithmetically, causing the higher efficiency. GMRES executes
+      more operations than conjugate gradient, also creating a higher
+      efficiency.}}
   \label{fig:titan_pure_strong}
@@
   \caption{\textbf{Pure domain decomposition weak scaling absolute
+      efficiency.} \textit{MCLS is an order of magnitude slower
+      arithmetically, causing the higher efficiency. GMRES executes
+      more operations than conjugate gradient, also creating a higher
       efficiency.}}
@@
   \caption{\textbf{Pure domain decomposition weak scaling algorithmic
+      efficiency.} \textit{As the global problem size was increased in
+      the weak scaling study, both conjugate gradient and GMRES
+      converged in fewer iterations, causing the rise in algorithmic
       efficiency.}}
@@
   \caption{\textbf{Pure domain decomposition weak scaling
-      implementation efficiency.}}
+      implementation efficiency.} \textit{Considering the scalability
+      of a single iteration of the implementation, the inflated
+      efficiencies of the MCLS MCSA implementation become more
+      obvious. Implementation optimization will likely reduce the MCSA
+      values to a level more comparable to the Belos values.}}
   \label{fig:titan_weak_implementation}
@@
 stochastic histories to use to generate the correction vector in the
-MCSA iteration. For this work we consider two choices, the split case
-and the fully replicated case. As an example, consider a single set
-problem where 20,000 histories are used in the Monte Carlo solve. If
-we choose the split case, then solving the same problem with 2 sets
-would give 10,000 histories in each set and 4 sets would give 5,000
-histories in each set. Splitting the histories in this way provides
-the same global number of histories used to compute the correction in
-the single set problem. We expect to reduce the time to solution by
-doing this as each individual set solve will occur faster in an almost
-linear fashion with the reduced number of histories. For the
-replicated history case, each additional set in the problem will use
-just as many histories as the single set case. Using 2 sets with fully
-replicated histories for our example problem then has each set
+MCSA iteration. For this work we consider two choices; the split case
+and the fully replicated case. As an example these cases, consider a
+single set problem where 20,000 histories are used in the Monte Carlo
+solve. If we choose the split case, then solving the same problem with
+2 sets would give 10,000 histories in each set and 4 sets would give
+5,000 histories in each set. Splitting the histories in this way
+provides the same global number of histories used to compute the
+correction in the single set problem. We expect to reduce the time to
+solution by doing this as each individual set solve will occur faster
+in an almost linear fashion with the reduced number of histories. For
+the replicated history case, each additional set in the problem will
+use just as many histories as the single set case. Using 2 sets with
+fully replicated histories for our example problem then has each set
 computing 20,000 histories with 40,000 global histories in the
@@
 
+\subsubsection{Strong Scaling with Multiple Sets}
+\label{subsubsec:ms_strong}
+
 For the strong scaling study, we apply the additional sets as
@@
 and boost the core count. For example, for the 4 set case, the data
-point at 64 cores has the same $N/P$ ratio as the single set problem
-with 16 cores. This is due to the fact that $N$ is fixed for the
-strong scaling study and 4 replications gives a set size of 16
+point at 64 cores has the same $N/P$ ratio in one set as the single
+set problem with 16 cores. This is due to the fact that $N$ is fixed
+for the strong scaling study and 4 replications gives a set size of 16
 cores. For the 2 set case, the 32 core data point is solving the same
-$N/P$ problem as the 16 core base case. In addition to considering the
-replicated domain, we must also consider how to modify the
-efficiencies we will compute. For the strong scaling case, recall that
-the objective is to decrease the time to solution for a given global
-problem size $N$ by increasing $P$. Therefore, we may use
-Eqs~(\ref{eq:strong_scaling_absolute_ref}),
+$N/P$ problem in a set as the 16 core base case.
+
+In addition to considering the replicated domain, we must also
+consider how to modify the efficiencies we will compute. For the
+strong scaling case, recall that the objective is to decrease the time
+to solution for a given global problem size $N$ by increasing
+$P$. Therefore, we may use Eqs~(\ref{eq:strong_scaling_absolute_ref}),
 (\ref{eq:algorithmic_efficiency}), and
@@
 any computational resources added to the problem should aim to result
-in speedup. If this speedup for one technique is less than another,
-then that technique is deemed a less efficient use of those
-computational resources.
+in speedup of this reference case. If this speedup for one technique
+of adding cores is less than another, then that technique is deemed a
+less efficient use of those computational resources.
 
@@
 sets with both splitting of histories across sets and full replication
-of histories across sets. Figure~\ref{fig:titan_strong_ms_time} gives
-the actual wall times for these computations. There are two important
-features to note in the wall time data. First, splitting the histories
-amongst sets instead of replicating them is faster due to the fact
-that cutting the number of histories in half reduces the time of an
-iteration by a factor that is nearly linear while increasing linearly
-the number of histories through replication does not decrease the
-iterations required to converge by a linear factor. Second, until the
-strong scaling wall is hit starting at around 10,000 cores, the single
-set case actually perfoms the fastest with only marginal improvement in
+of histories across sets using both 2 and 4
+sets. Figure~\ref{fig:titan_strong_ms_time} gives the actual wall
+times for these computations. There are two important features to note
+in the wall time data. First, splitting the histories amongst sets
+instead of replicating them is faster due to the fact that cutting the
+number of histories in half reduces the time of an iteration by a
+factor that is nearly linear while increasing linearly the number of
+histories through replication does not decrease the iterations
+required to converge by a linear factor. Second, until the strong
+scaling wall is hit starting at around 10,000 cores, the single set
+case actually perfoms the fastest with only marginal improvement in
 runtimes at higher numbers of cores with more sets. At these larger
@@
 better off running the same problem using a single set at lower core
-counts to conserve computational resource while maintaining a good tim
-to solution.
+counts to conserve computational resources while maintaining a good
+time to solution.
 
@@
   \caption{\textbf{Wall time in seconds to solution for each case for
-      the strong scaling study with multiple sets.}}
+      the strong scaling study with multiple sets.} \textit{Until the
+      strong scaling wall is reached at O(10,000) cores, the single
+      set case performs the best. For the multiple set cases, it is
+      faster to split histories amongst sets and maintain iterative
+      performance rather than replicate histories to improve iterative
+      performance.}}
   \label{fig:titan_strong_ms_time}
@@
 From this timing data we then compute the absolute efficiencies for
-strong scaling as given in Figure~\ref{fig:titan_strong_ms_eff}. As
-observed in the timing data, because the single set case had the
-fastest run times up to the strong scaling wall it also has the best
-observed efficiencies. For the 2 set case with history splitting,
-although the time to perform an iteration was nearly cut in half, the
-extra parallel overhead associated with reducing the tally across the
-blocks to combine the set results increases run times and reduces
-efficiencies.This reduction across blocks is even more obvious for the
-4 set case. For the replicated cases, we do observe a flat region of
-parallel efficiencies at larger core counts (around 32,000 for the 4
-set case with history replication), but the additional cores do not
-reduce the runtime of the global problem efficiently. Ultimately, all
-cases effectively hit the strong scaling wall at the same time. We
-expect this if we consider the fact that adding sets to the problem is
-in itself effectively a form of strong scaling. For any given problem
-of fixed size $N$, adding extra sets to the computation decreases the
-ratio of $N/P$ as $P$ will grow with the addition of sets. For the
-split history case, the splitting further exacerbates this situation
-by further decreasing the amount of histories that will be computed in
+strong scaling as given in Figure~\ref{fig:titan_strong_ms_eff} using
+the 16-core single set calculation as the reference case. As observed
+in the timing data, because the single set case had the fastest run
+times up to the strong scaling wall it also has the best observed
+efficiencies. For the 2 set case with history splitting, although the
+time to perform an iteration was nearly cut in half, the extra
+parallel overhead associated with reducing the tally across the blocks
+to combine the set results increases run times and reduces
+efficiencies. This reduction across blocks is even more obvious for
+the 4 set case. For the replicated cases, we do observe a flat region
+of parallel efficiencies at larger core counts (up to around 32,000
+for the 4 set case with history replication) showing improved scaling
+within context of not considering the single set
+calculation. Unfortunately, the additional cores do not reduce the
+runtime of the global problem as efficiently as expected. Ultimately,
+all cases effectively hit the strong scaling wall at the same time at
+around O(10,000) cores.
+
+We expect this if we consider the fact that adding sets to the problem
+is in itself effectively a form of strong scaling. For any given
+problem of fixed size $N$, adding extra sets to the computation
+decreases the global ratio of $N/P$ as $P$ will grow with the addition
+of sets (although $N/P$ is constant within a set). For the split
+history case, the splitting further exacerbates this situation by
+further decreasing the amount of histories that will be computed in
 the local domain, giving a larger ratio of communication to work in
@@
 
+Although not as dramatic as one might originally expect from
+introducing replication, there are some modest gains in efficiencies
+at high core counts by using multiple sets. Of interest here is where
+the multiple set cases perform better than the single set case after
+the single set case has hit the strong scaling wall. In
+Figure~\ref{fig:titan_strong_ms_eff}, at 16,384 cores the single set
+case operates at 38\% efficiency while at the same number of cores,
+the 2 set case with history replication operates at 62\%
+efficiency. This is reasonable improvement in the use of those
+computational resources, gained by replicating the problem. Beyond
+16,384 cores, there is not a compelling reason to use more
+computational resources for the this problem.
+
 \begin{figure}[t!]
@@
   \end{center}
-  \caption{\textbf{Absolute parallel efficiency relative to 16-core
-      1-set base case.}}
+  \caption{\textbf{Multiple set absolute parallel efficiency.}
+    \textit{Computed relative to the 16-core 1-set base case. Using a
+      single set is the best means by which to improve strong scaling
+      up to the strong scaling wall. At higher core counts past where
+      the single set case hits the strong scaling wall adding sets can
+      provide some improvements in efficiencies.}}
   \label{fig:titan_strong_ms_eff}
@@
 required to converge and computed algorithmic efficiences for each of
-the cases presented in the scaling data. As expected, splitting
+the cases presented in the strong scaling data. As expected, splitting
 histories maintains a fixed global number of histories and therefore
@@
 doing the block level reduction, but we also have additional histories
-that increase the time to solution for a single iteration. Given that
-the time to solution is not efficiently improved using multiple sets
-in both the split and replicated history cases, multiple set domain
-decomposition is not an effective means to boost the strong scaling
-efficiencies of MCSA.
+that increase the time to solution for a single iteration.
 
@@
   \end{center}
-  \caption{\textbf{Strong scaling implementation efficiency relative
-      to the 16-core 1-set base case.}}
+  \caption{\textbf{Multiple set strong scaling implementation
+      efficiency.} \textit{Computed relative to the 16-core 1-set base
+      case. Considering iterations required to converge, the
+      performance of the split history cases at much better than the
+      cases with history replication.}}
   \label{fig:titan_strong_ms_impeff}
@@
 
+\clearpage
+
+\subsubsection{Weak Scaling with Multiple Sets}
+\label{subsubsec:ms_weak}
+
 Next, we continue to a weak scaling study of the multiple sets
-problem. As with the strong scaling study, we will modify the pure
-domain decomposition study to account for the additional
+problem. In the same way as the strong scaling study, we will modify
+the pure domain decomposition study to account for the additional
 replication. For each calculation in the single set study, that
@@
 replicated histories actually decreases the time to solution. This was
-not observed in the strong scaling case due to the larger decrease in
-computational work as the number of cores was increased. This result
-is significant in that it shows a solution technique in which the
-linear problem can actually be replicated and enhance time to
-solution, something that cannot be acheived with conventional
+not observed in the strong scaling case until much larger core counts
+due as the multiple set computations hit the strong scaling wall after
+the single set computation.
+
+This result is significant in that it shows a solution technique in
+which the linear problem can actually be replicated and enhance time
+to solution, something that cannot be acheived with conventional
 methods. At larger core counts, however, this reduction in compute
-time is not as great as at lower core counts due to the fact that the
-tally reduction across blocks requires more resources. In addition, at
-over 130,000 cores for the largest 4 set computation, load balance is
-likely becoming an issue due to the linear system and subsequent
-parallel matrix and vector operations only occuring on a subset of the
-entire processor space of the problem.
+time is not as large as at lower core counts due to the fact that the
+tally reduction across blocks requires more resources and those
+resources are clearly growing with core count (most obviously in the 4
+set case with split histories). In addition, at 131,072 cores for the
+largest 4 set computation, load balance is likely becoming an issue
+due to the linear system and subsequent parallel matrix and vector
+operations only occuring on a subset of the entire processor space of
+the problem while the Monte Carlo calculation is occuring on all
+processors.
 
@@
 where now the absolute efficiency is a function of the number of sets
-in the problem. Interestingly, Eq~(\ref{eq:ms_weak_efficiency}) has a
-very similar form to the strong scaling efficiency given by
-Eq~(\ref{eq:strong_scaling_absolute}), representative of the strong
-scaling nature of adding additional sets. For problems where adding
-sets improves time to solution, this new weak scaling formulation
-accounts for how efficiently those extra resources are incorporated
-into the problem. Without such a modification, the largest case run
-for the split 4 set case would have a computed weak scaling efficiency
-of nearly 100\% using the standard formula for absolute
-efficiency. However, using 4 times as many cores as the single set
-case increases the runtime for a fixed global problem size and
-therefore we actually expect to compute a very low efficiency that
-shows this very poor use of resources.
+in the problem and again $M/Q = N/P$. As with the strong scaling
+study, we will use the 16-core single set case as the reference
+computation for efficiencies. Interestingly,
+Eq~(\ref{eq:ms_weak_efficiency}) has a very similar form to the strong
+scaling efficiency given by Eq~(\ref{eq:strong_scaling_absolute}),
+representative of the strong scaling nature of adding additional
+sets. For problems where adding sets improves time to solution, this
+new weak scaling formulation accounts for how efficiently those extra
+resources are incorporated into the problem. Without such a
+modification, the largest case run for the split 4 set case would have
+a computed weak scaling efficiency of nearly 100\% using the standard
+formula for absolute efficiency. However, using 4 times as many cores
+as the single set case increased the runtime for a fixed global
+problem size and therefore we would actually expect to compute a very
+low efficiency that shows this very poor use of resources.
 
@@
 across sets had effectively the same results for the weak scaling
-study with the iterations to converge and compute algorithmic
-efficienies given by Table~\ref{tab:ms_weak_alg_eff}. Again, when the
+study with the iterations to converge and computed algorithmic
+efficiencies given by Table~\ref{tab:ms_weak_alg_eff}. Again, when the
 implementation efficiencies are computed, we see in
@@
 communication overhead of adding sets to the system, one should never
-expect to boost parallel efficiencies for MCSA using this
-technique. Further, in a purely strong scaling environment, no
-performance gains should be expected as adding sets creates an even
-more parasitic ratio of $N/P$. In a weak scaling environment, although
-no efficiency increase will be noted, one can expect an enhancement of
-time to convergence by replicating.
+expect to seriously boost parallel efficiencies for MCSA using this
+technique accept for certain instances past the strong scaling wall in
+a purely strong scaling environment. In a weak scaling environment,
+although no efficiency increase will be noted, one can expect an
+enhancement of time to convergence by replicating.
 
@@
 
+Based on the analysis of the pure domain decomposed MCSA algorithm and
+subsequently the MSOD implementation, one may ask then ask if there is
+any need to commincate histories from domain-to-domain at all. Overlap
+was not successful because so few histories needed to be communicated
+in the first place that keeping them on-process resulted in more work
+than the communications savings that occurred. If that is true, then
+perhaps without any overlap or communication, just doing local Monte
+Carlo may be satisfactory for accelerating the MCSA iteration.
+
 Full-clip actually uses the solver as a true subdomain solver, just

### 2013-07-03 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
adding first part of subdomain section

@@
 strong and weak scaling performance seriously degrades while the
-binary tree pattern maintains execellent performance. At this number
-of cores, any extra work given to the master process, even if it
-amounts to summing a handful of integers and checking for a few extra
-messages as in this algorithm, this extra work grows enough to create
-serious load imbalance and reduce performance.
+binary tree pattern maintains excellent performance. At this number of
+cores, any extra work given to the master process, even if it amounts
+to summing a handful of integers and checking for a few extra messages
+as in this algorithm, grows enough to create serious load imbalance
+and reduce performance.
 
@@
 perhaps without any overlap or communication, just doing local Monte
-Carlo may be satisfactory for accelerating the MCSA iteration.
-
-Full-clip actually uses the solver as a true subdomain solver, just
-like you would in a Schwarz method. In fact, this is a stochastic
-realization of a Richardson iteration - they are very similar in those
-regards.
-
-Overlap is not effective because histories apparently dump most of
-their useful information within the first few steps (which is why we
-can use such a large weight cutoff and there is an insensitivity to
-weight cutoff). So keeping them around to do extra work is not
-improving iterative performance - thus increasing compute times. This
-is true even for ill-conditioned cases where adding overlap might keep
-a number of histories within the domain for significantly longer.
+Carlo may be satisfactory for accelerating the MCSA iteration. In this
+case, histories near the boundaries of the domain may be truncated
+before reaching the weight cutoff if the reach the domain boundary
+first. This truncation has the potential to increase the stochastic
+uncertainty of the Neumann-Ulam result and therefore increase the
+number of MCSA iterations required to converge. In addition, the
+scalability of the method will then be purely limited by the parallel
+matrix and vector operations required to implement the remainder of
+the algorithm.
+
+Considering using MCSA in this way such that the Monte Carlo solve
+occurs only in the subdomains casts the method as more of a true
+domain decomposition as outlined by Chapter 14 of Saad's text on
+iterative methods \citep{saad_iterative_2003}. Compared to traditional
+domain decomposition methods we see many similarities including the
+potential to overlap the subdomains to improve the results and the
+approximate subdomain solves effectively through a relaxation
+scheme. By building the boundary data as outlined in
+\S~\ref{subsec:domain_generation}, we are effectively gathering one
+level of the adjacent domains, albeit in this case to satisfy the data
+requirements of the Monte Carlo estimators. Furthermore, the fact that
+we are using residual Monte Carlo to compute the correction vector
+casts MCSA as a stochastic realization of a additive Schwarz method
+with a fixed point iterations used as a relaxation step\footnote{See
+  section 14.3.3 in \citep{saad_iterative_2003} for a full definition
+  of the additive Schwarz method.}. Both the fixed point iteration and
+the computation of the residual propagate the solution from individual
+subdomains to the others through the matrix-vector multiply operation
+after all subdomain Monte Carlo solves have been completed. Perhaps an
+even more important consequence of viewing MCSA as a stochastic
+additive Schwarz procedure is that it has the potential to more
+readily fit as an accelerator into other linear solver techniques
+including GMRES or potentially as a preconditioner.
+
+Using the Neumann-Ulam solver to compute the MCSA correction only in
+the subdomains, the strong and weak scaling studies were repeated to
+assess the performance gains made by eliminating communication in the
+Monte Carlo solver. For each scaling study, the 16-core base case with
+Monte Carlo communication was used as the reference point for
+efficiency computations in order to measure how eliminating
+communication boosts
+efficiencies. Figure~\ref{fig:titan_strong_subdomain} gives the
+results of the strong scaling study and
+Figure~\ref{fig:titan_weak_subdomain} gives the results of the the
+weak scaling study in terms of absolute efficiencies. In both cases,
+the benefits to parallel performance of using MCSA as a subdomain
+solver are obvious. In the strong scaling case we make the important
+observation that we hit the strong scaling wall at the same time with
+and without the additional Monte Carlo communication. Although the
+Monte Carlo communication did contribute to the efficiency decrease,
+it is clear that the remaining matrix and vector operations are the
+primary impediment to scalability. Furthermore, we see an even greater
+rise above unity in efficiencies because there is less communication
+overhead hiding the better cache utilization as the local problem
+shrinks. At 16,384 cores, performing the subdomain solves boost the
+absolute efficiency from 38\% to 67\% relative to the reference case.
+
+In the weak scaling study, we do not have the scaling wall to contend
+with and eliminating Monte Carlo communication through subdomain
+solves provides an effectively linear boost in the weak scaling
+absolute efficiency. We still see a small reduction in weak scaling
+efficiency as a function of core count due to the communication
+overhead for the parallel matrix and vector operations. Here we
+observe super unitary efficiencies for the subdomain case as there was
+a substantial improvement in runtimes relative to the 16-core base
+case with Monte Carlo communication.
 
@@
 \end{figure}
+
+Of additional concern for the subdomain method is the potential
+decrease in iterative performance due to the fact that histories near
+the boundary will potentially have their random walk truncated, thus
+decreasing the information available to compute the correction
+vector. For both the weak and strong scaling study, all calculations
+converged in 22 or in only a few cases 21 iterations for both the
+subdomain solve and the solve with communication giving an algorithmic
+efficiency of 1.048 for all cases. This small improvement in iteration
+count was not observed to be a function of problem size, number of
+processors, or whether the subdomain method was used. Therefore, there
+is no reason to expect overlap to improve the iterative or scaling
+performance of the method as iterative performance is maintained
+without it. Overlap is not effective because histories relinquish most
+of their useful information within the first few steps (which is why
+we can use such a large weight cutoff and there is an insensitivity to
+weight cutoff as shown in
+Figure~\ref{fig:estimator_wc_iters}). Keeping them around longer with
+overlap to do extra work in the local domain will not improve
+iterative performance and thus increasing compute times.
+
+\clearpage
+
+\subsubsection{Strong Scaling with Multiple Sets}
+\label{subsubsec:subdomain_ms_strong}
+
+\clearpage
+
+\subsubsection{Weak Scaling with Multiple Sets}
+\label{subsubsec:subdomain_ms_weak}

### 2013-07-03 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
spell check on parallel chapter

@@
 methods. Using the verified implementation of the algorithm, we
-perform parallel scaling studings to test its performance on a
+perform parallel scaling studies to test its performance on a
 leadership class machine and compare this performance to the
@@
 particle transport problem, a domain decomposed Neumann-Ulam method
-has the same communcation requirements with all histories in both
+has the same communication requirements with all histories in both
 examples requiring the same parallel operations. The amount of
@@
 giving the eigenvalue spectrum for the Jacobi preconditioned iteration
-matrix. To find the maxium eigenvalue,
+matrix. To find the maximum eigenvalue,
 Eq~(\ref{eq:preconditioned_eigenvalues}) is plotted as a function of
@@
 In the context of radiation transport, in 2009 Brunner and colleagues
-provided a fully asynchronous domain decomposed parallel aglorithm as
+provided a fully asynchronous domain decomposed parallel algorithm as
 implemented in production implicit Monte Carlo codes
@@
 Successful execution of this algorithm requires construction of the
-parallel Neumann-Ulam decomposition as described in the preceeding
+parallel Neumann-Ulam decomposition as described in the preceding
 section. This data is used to begin
@@
 collected list of neighboring processors generated while building the
-Neumann-Ulam decompostion to setup the set of asynchronous messages
+Neumann-Ulam decomposition to setup the set of asynchronous messages
 required for domain-to-domain communication. Again, consider this
@@
 (\textit{historiesCompleted} in the algorithm) be updated
-incrementally and funnelled to the root process. Because we are
+incrementally and funneled to the root process. Because we are
 solving fixed source problems with the Neumann-Ulam algorithm without
@@
 from the parent process in the binary tree. In this case, mutually
-exclusivivity permits all local operations to occur independently of
+exclusivity permits all local operations to occur independently of
 other processes in the system and their current state. For each
@@
 are added to the stack from processing and the nonblocking receive
-operation reinstantiated. In addition, the terminated histories count
+operation re-instantiated. In addition, the terminated histories count
 is updated with those values received from the child processes.
@@
 provide value in the future. In addition to a robust fully
-asynchronous communication pattern, this algorith may also be modified
+asynchronous communication pattern, this algorithm may also be modified
 to account for situations where the total number of histories in a
@@
 generation of overlap. Conveniently, this is identical to the
-neighboring states discovery problem discussed in conjuction with
+neighboring states discovery problem discussed in conjunction with
 Figure~\ref{fig:diffusion_graph}. To generate the boundary for the
@@
 immediately adjacent states in the system that did not already exist
-in the local domain. We solved this problem by travsersing the graph
+in the local domain. We solved this problem by traversing the graph
 of the matrix for each of the boundary states and determining which
@@
 by 10 additional discrete state transitions, we can reduce the number
-of histories that must be comunicated by a significant fraction. We
+of histories that must be communicated by a significant fraction. We
 determine the information we must gather from neighboring domains to
@@
 domain that we require and make that data local, incrementing the size
-of the overap. Once the overlap has been gathered, one extra graph
+of the overlap. Once the overlap has been gathered, one extra graph
 traversal on the boundary states is performed to get the new
@@
       transport domain on the primary set. The Monte Carlo data
-      structures are then broadcast amongst the blocks to all sets in
+      structures are then broadcast among the blocks to all sets in
       the system.}}
@@
 With the parallel adjoint Neumann-Ulam solver implementation described
-aboive, the parallel implementation of an MCSA iteration is
+above, the parallel implementation of an MCSA iteration is
 trivial. Recall the MCSA iteration procedure presented again here for
@@
 for generating the replications required for a multiple set
-implementation, the explict matrix and vector operations required for
+implementation, the explicit matrix and vector operations required for
 the rest of an MCSA iteration occur only on the subset of processors
@@
 data presented in the following sections, an explanation of each
-observed case of superunitary efficiencies will be provided.
+observed case of super-unitary efficiencies will be provided.
 
@@
 work, a history buffer size of 4096 and a buffer check period of 128
-performed best for the intitial calculations. However, most values in
+performed best for the initial calculations. However, most values in
 the range tested will give good performance. Brunner and Brantley
@@
 problem used to scale to O(10,000) cores was large enough that memory
-thrashing was occuring in the 16 core base case. As the number of
+thrashing was occurring in the 16 core base case. As the number of
 cores in the strong scaling study was increased and the local problem
@@
 is important as we can expect the same qualitative scaling performance
-from MCSA onc arithmetic optimization has been performed to improve to
+from MCSA once arithmetic optimization has been performed to improve to
 serial performance of the implementation. In addition, these results
 are valuable in that the show MCSA, when properly parallelized, has
-the ability to be competetive on leadership-class computing platforms
+the ability to be competitive on leadership-class computing platforms
 with conventional subspace methods with good strong scaling observed
@@
 important to note that the weak scaling of MCSA when properly
-parallelized remains competetive with the Krylov methods keeping in
+parallelized remains competitive with the Krylov methods keeping in
 mind the artificial inflation of efficiencies.
@@
 we are measuring the parallel performance of a single iteration for
-each method by looking at these implementation efficieinces. As the
+each method by looking at these implementation efficiencies. As the
 global problem size increases, MCSA is less prone to a drop in
@@
 times for these computations. There are two important features to note
-in the wall time data. First, splitting the histories amongst sets
+in the wall time data. First, splitting the histories among sets
 instead of replicating them is faster due to the fact that cutting the
@@
 scaling wall is hit starting at around 10,000 cores, the single set
-case actually perfoms the fastest with only marginal improvement in
+case actually performs the fastest with only marginal improvement in
 runtimes at higher numbers of cores with more sets. At these larger
@@
       set case performs the best. For the multiple set cases, it is
-      faster to split histories amongst sets and maintain iterative
+      faster to split histories among sets and maintain iterative
       performance rather than replicate histories to improve iterative
@@
 
-Also of interest here is the effect of splitting histories amongst
+Also of interest here is the effect of splitting histories among
 sets versus replicating them on the iterative performance of the
 method. Table~\ref{tab:ms_strong_alg_eff} gives the iterations
-required to converge and computed algorithmic efficiences for each of
+required to converge and computed algorithmic efficiencies for each of
 the cases presented in the strong scaling data. As expected, splitting
@@
 which the linear problem can actually be replicated and enhance time
-to solution, something that cannot be acheived with conventional
+to solution, something that cannot be achieved with conventional
 methods. At larger core counts, however, this reduction in compute
@@
 due to the linear system and subsequent parallel matrix and vector
-operations only occuring on a subset of the entire processor space of
-the problem while the Monte Carlo calculation is occuring on all
+operations only occurring on a subset of the entire processor space of
+the problem while the Monte Carlo calculation is occurring on all
 processors.
@@
 implementation efficiencies are computed, we see in
-Figure~\ref{fig:titan_weak_ms_impeff} that splitting histories amongst
+Figure~\ref{fig:titan_weak_ms_impeff} that splitting histories among
 sets is still a more efficient manner of introducing multiple sets
@@
 subsequently the MSOD implementation, one may ask then ask if there is
-any need to commincate histories from domain-to-domain at all. Overlap
+any need to communicate histories from domain-to-domain at all. Overlap
 was not successful because so few histories needed to be communicated

### 2013-07-04 doc/dissertation/document/backmatter/nonlinear_background.tex
adding to overlap scaling section and preparing spreadsheet for more titan data

--- 2013-06-11 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
@@
-\chapter{Monte Carlo Synthetic Acceleration Methods for the Navier-Stokes Equations}
-\label{ch:nonlinear_problem}
-Nonlinear transport problems are a common occurrence in single and
-multiple physics problems. Systems of partial differential equations
-such as those that describe fluid flow or more general transport
-processes when discretized by conventional methods yield discrete sets
-of stiff equations with nonlinearities present in the
-variables. Traditionally, such systems have been solved by linearizing
-them in a form where the nonlinearities in the variables are
-eliminated and more traditional linear methods can be used for
-solutions. Often characterized as segregated methods where physics
-operators are split and their action on the system approximated in
-steps, such methods lack consistency and accuracy in resolving the
-nonlinear component of the solution. In the last 30 years, fully
-consistent nonlinear methods based on Newton's method have become more
-popular and many advances have been made in the computational physics
-field to employ these methods.
+\chapter{Conventional Solution Methods for Nonlinear Systems\ }
+\label{chap:nonlinear_background}
 
-In the context of solving standalone linear systems, Monte Carlo
-methods do not provide significant merit over Krylov methods due to
-the fact that the linear operator must be explicitly formed. For many
-applications, such a requirement is prohibitive and perhaps not even
-feasible to implement. Therefore, a Monte Carlo solver is best suited
-for situations in which not only are Krylov methods applicable, but
-also in which the operator is readily, if not naturally,
-formed. Modern nonlinear methods meet both of these requirements with
-Newton methods used in conjunction with Krylov methods for a robust,
-fully implicit solution strategy. Furthermore, modern techniques exist
-that permit the automatic construction of the linear operator
-generated within a Newton method based on the nonlinear residual
-evaluations, providing all of the components necessary for a Monte
-Carlo solver to provide value. We therefore devise a new nonlinear
-method based on the MCSA algorithm and Newton's method and discuss its
-potential benefits.
-
-%%---------------------------------------------------------------------------%%
-\section{Preliminaries}
-\label{sec:nonlinear_preliminaries}
-We formulate the \textit{nonlinear problem} as follows
-\citep{knoll_jacobian-free_2004}:
-\begin{equation}
-  \ve{F}(\ve{u}) = \ve{0}\:,
-  \label{eq:nonlinear_problem}
-\end{equation}
-where $\ve{u} \in \mathbb{R}^n$ is the solution vector and
-$\ve{F}:\mathbb{R}^N \rightarrow \mathbb{R}^N$ is the function of
-nonlinear residuals. We write the nonlinear system in this form so
-that when an exact solution for $\ve{u}$ is achieved, all residuals
-evaluate to zero. \textit{Newton's method} is a root finding algorithm
-and therefore we can use it to solve Eq~(\ref{eq:nonlinear_problem})
-if we interpret the exact solution $\ve{u}$ to be the roots of
-$\ve{F}(\ve{u})$. Newton's method is also an iterative scheme, and we
-can generate this procedure by building the Taylor expansion of the
-$k+1$ iterate of the nonlinear residuals about the previous $k$
-iterate:
-\begin{equation}
-  \ve{F}(\ve{u}^{k+1}) = \ve{F}(\ve{u}^{k}) +
-  \ve{F}'(\ve{u}^{k})(\ve{u}^{k+1}-\ve{u}^{k}) +
-  \frac{\ve{F}''(\ve{u}^{k})}{2}(\ve{u}^{k+1}-\ve{u}^{k})^2 + \cdots
-  \:.
-  \label{eq:newton_derivation_1}
-\end{equation}
-If we ignore the nonlinear terms in the expansion and assert that at
-the $k+1$ iterate $\ve{u}^{k+1}$ is the exact solution such that
-$\ve{F}(\ve{u}^{k+1}) = \ve{0}$, then we are left with the following
-equality:
-\begin{equation}
-  -\ve{F}(\ve{u}^{k}) =
-  \ve{F}'(\ve{u}^{k})(\ve{u}^{k+1}-\ve{u}^{k})\:.
-  \label{eq:newton_derivation_2}
-\end{equation}
-We note two things of importance in
-Eq~(\ref{eq:newton_derivation_2}). The first is that
-$\ve{F}'(\ve{u}^{k})$ is in fact the \textit{Jacobian},
-$\ve{J}(\ve{u})$, of the set of nonlinear residuals and is defined
-element-wise as:
-\begin{equation}
-  J_{ij} = \frac{\partial F_i(\ve{u})}{\partial u_j}\:.
-  \label{eq:jacobian_def}
-\end{equation}
-Second, we note that $(\ve{u}^{k+1}-\ve{u}^{k})$ is simply the
-solution update from the $k$ iterate to the $k+1$ iterate. We will
-define this update as the \textit{Newton correction} at the $k$
-iterate, $\delta \ve{u}^k$. To finish, we can then rearrange
-Eq~(\ref{eq:newton_derivation_2}) to define the Newton iteration
-scheme for nonlinear problems:
-\begin{subequations}
-  \begin{gather}
-    \ve{J}(\ve{u}) \delta \ve{u}^k = -\ve{F}(\ve{u}^{k})\\
-    \ve{u}^{k+1} = \ve{u}^k + \delta \ve{u}^k\:.
-  \end{gather}
-  \label{eq:newton_iteration}
-\end{subequations}
-There are then three distinct steps to perform: evaluation of the
-nonlinear residuals using the solution at the $k$ iterate, the
-solution of a linear system to compute the Newton correction where the
-Jacobian matrix of the nonlinear equation set is the linear operator,
-and the application of the correction to the previous iterate's
-solution to arrive at the next iterate's solution. In the asymptotic
-limit, the iterations of Newton's method will converge the nonlinear
-residual quadratically \citep{kelley_iterative_1995}. Convergence
-criteria is set for stopping the iteration sequence based on the
-nonlinear residual. Commonly, the following criteria is used:
-\begin{equation}
-  ||\ve{F}(\ve{u}^{k})|| < \epsilon ||\ve{F}(\ve{u}^{0})||\:,
-  \label{eq:newton_stopping_criteria}
-\end{equation}
-where $\epsilon$ is a user defined tolerance parameter. Newton's
-method is \textit{consistent} in that all components of the nonlinear
-functions that describe the physics we are modeling are updated
-simultaneously in the iteration sequence with respect to one
-another. This is in comparison to \textit{inconsistent} strategies,
-such as a pressure correction strategy for solving the Navier-Stokes
-equations \citep{pletcher_computational_1997}, where the components of
-$\ve{u}$ are updated in a staggered fashion depending on the
-particular equations that they are associated with.
-
-\subsection{Inexact Newton Methods}
-\label{subsec:newton_methods}
+\section{Inexact Newton Methods}
+\label{sec:newton_methods}
 Inexact Newton methods arise when the Jacobian operator is not exactly
@@
 
-\subsection{Newton-Krylov Methods}
-\label{subsec:newton_krylov_methods}
+\section{Newton-Krylov Methods}
+\label{sec:newton_krylov_methods}
 A form of inexact Newton methods, \textit{Newton-Krylov methods} are
@@
 
-\subsubsection{Jacobian-Free Approximation}
-\label{subsubsec:jacobian_free_approximation}
+\subsection{Jacobian-Free Approximation}
+\label{subsec:jacobian_free_approximation}
 In most cases, the Jacobian is difficult to form from the difference
@@
 
-\subsubsection{Automatic Differentiation for Jacobian Generation}
-\label{subsubsec:automatic_differentiation}
+\subsection{Automatic Differentiation for Jacobian Generation}
+\label{subsec:automatic_differentiation}
 If it is acceptable to store the actual Jacobian matrix, other methods
@@
 acceptable for use in large-scale, production physics codes.
-
-%%---------------------------------------------------------------------------%%
-\section{The FANM Method}
-\label{sec:fanm}
-In production physics codes based on nonlinear equations sets,
-Newton-Krylov methods are the primary means of generating a fully
-consistent solution scheme
-\citep{evans_development_2006,evans_enhanced_2007,gaston_parallel_2009,godoy_parallel_2012}. Typically,
-for large scale simulations these problems are memory limited due to
-the subspaces generated by robust Krylov methods. Often, a matrix-free
-approach is chosen to relax memory requirements over directly
-generating the Jacobian matrix and facilitate the
-implementation. However, as we observed in previous sections, these
-matrix-free methods suffer from poorly scaled problems and the first
-order error introduced by the Jacobian approximation. In addition, it
-was observed that the savings induced by the matrix-free approach is
-eventually amortized over a number of nonlinear iterations where it
-becomes more efficient computationally to instead form the Jacobian.
-
-In Chapter~\ref{ch:stochastic_methods}, we focused our efforts on
-developing and improving Monte Carlo methods for inverting linear
-systems. These methods, when used to accelerate a stationary method in
-MCSA, enjoy a robust implementation and exponential convergence
-rates. Further, the only storage required is that of the full linear
-system including the linear operator so that we may generate
-transition probabilities for the random walk sequence. Although this
-requires more storage to represent the linear system than that of a
-Krylov method where the operator is not required, we do not incur any
-additional storage costs once the iteration sequence begins. In the
-context of nonlinear problems, the Jacobian matrix that we are
-required to generate for the Monte Carlo solvers may be generated at
-will from the nonlinear functions in the Newton system using automatic
-differentiation. Not only do we then have a simple and automated way to
-generate the Jacobian, but we also enjoy a Jacobian of numerical
-precision equivalent to that of our function evaluations. We therefore
-propose the \textit{Forward-Automated Newton-MCSA} (FANM) method that
-utilizes all of the above components. We hypothesize that such a
-method will be competitive with Newton-Krylov methods not only from a
-convergence and timing perspective, but also relax scaling
-requirements of matrix-free methods and memory costs of both
-matrix-free and fully formed Jacobian methods to allow the application
-developer to solve problems of finer discretization and higher-order
-functional representations while maintaining a robust and efficient
-parallel implementation.
-
-\subsection{Jacobian Storage vs. Subspace Storage and Restarts}
-\label{subsec:fanm_storage}
-To gauge the memory benefits of a FANM method over a Krylov-based
-scheme, we must look at the storage requirements of the Jacobian
-matrix as compared to the Krylov subspace vectors for sparse linear
-problems. Saad's text provides us relations for both of these cases
-\citep{saad_iterative_2003}. Beginning with the Jacobian storage, for
-sparse problems production linear algebra library implementations
-utilize \textit{compressed row storage} to efficiently store the
-non-zero components of a sparse matrix while still allowing for all
-standard parallel matrix computations to be performed. Per
-\S~\ref{sec:parallel_krylov_methods}, recall that efficient parallel
-matrix-vector multiplications rely on processors knowing which other
-processors contain their neighboring matrix and vector
-elements. Typically this information is represented by a graph which
-must be stored along with the matrix elements themselves. To store the
-elements, consider the following sparse matrix:
-\[
-  \ve{A} =
-  \begin{bmatrix}
-    2 & 0 & 8 & 0 & 0 & 0 \\ 
-    4 & 5 & 0 & 1 & 0 & 0 \\ 
-    0 & 2 & 1 & 0 & 1 & 0 \\ 
-    0 & 0 & 3 & 7 & 0 & 2 \\ 
-    0 & 0 & 0 & 4 & 9 & 0 \\
-    0 & 0 & 0 & 0 & 9 & 1 \\
-  \end{bmatrix}\:.
-\]
-The compressed form of \ve{A} would then be:
-\[
-  \begin{array}{lccccccccccccccc}
-    \ve{A}\ values: & 2 & 8 & 4 & 5 & 1 & 2 & 1 & 1 & 3 & 7 & 2 & 4 & 9
-    & 9 & 1 \\ column: & 1 & 3 & 1 & 2 & 4 & 2 & 3 & 5 & 3 & 4 & 6 & 4
-    & 5 & 5 & 6 \\ row\ start: & 1 & 3 & 6 & 9 & 12 & 14 & 16 \\
-  \end{array}
-\]
-where the first row contains the non-zero values of the matrix
-$\ve{A}$ starts at, the second row contains the column index of those
-values, and the third row contains the index of the A value and column
-index rows that each matrix row starts at ending with the start of the
-next row. For this simple example, we actually break even by storing
-36 elements in the full matrix and 36 pieces of data in the compressed
-format. However, for large, sparse matrices there is a significant
-storage savings using the compressed format.
-
-In his discussion on orthogonalization schemes for Krylov subspaces,
-Saad provides us with space and time complexities for these
-operations. To find a solution vector $\ve{x} \in \mathbb{R}^{N \times
-  N}$ with $m$ Krylov iterations (and therefore $m+1$ subspace
-vectors) the storage requirement is then $(m+1) N$ for most common
-orthogonalization techniques. Using our simple example above (even
-though we did not see any gains in storage over the full matrix
-representation), if it takes 10 GMRES iterations using Arnoldi
-orthogonalization to converge to a solution, then a total of 66 pieces
-of data are required to be stored instead of 36 for the explicitly
-formed matrix case (we do not consider graph storage here for
-comparison as both methods will require the graph for parallel
-operations). For larger sparse matrices, this disparity in memory
-requirements will be even greater, especially for ill-conditioned
-systems that require many GMRES iterations to converge.
-
-\subsection{Parallel FANM Implementation}
-\label{subsec:parallel_fanm}
-A parallel FANM method relies on a basic set of parallel matrix-vector
-operations as well as the global residual and Jacobian assembly
-procedure described in
-\S~\ref{subsubsec:automatic_differentiation}. Consider the Newton
-iteration scheme in Eq~(\ref{eq:newton_iteration}). We must first
-assemble the linear system in parallel through the element-wise
-function evaluations to generate both the global Jacobian operator and
-the global residual vector on the right hand side. Per Bartlett's
-work, efficient and automated parallel mechanisms are available to do
-this through a sequence of scatter/gather operations. With these tools
-available for residual and Jacobian generation, the remainder of the
-parallel procedure is simple, with the linear Newton correction system
-solved using the parallel MCSA method as previously described and the
-Newton correction applied to the previous iterate's solution through a
-parallel vector update operation.
-
-As Newton methods are formulated independent of the inner linear
-solver generating the Newton corrections, the actual performance of
-the nonlinear iterations using MCSA is expected to be similar to that
-of traditional Newton-Krylov methods. Furthermore, we expect to
-achieve numerically identical answers with a Newton-MCSA method as
-other Newton methods and we should indeed verify this. The parallel
-performance of such a method will inherently be bound to the parallel
-Monte Carlo implementation of the linear solver as the parallel
-operations at the nonlinear iteration level are identical to those
-that you would perform with a Newton-Krylov method. Matrix-free
-formulations will have different parallel performance than these
-methods, and therefore we should compare FANM performance to JFNK-based
-schemes. More important than performance in this situation is the
-reduced memory pressure that a FANM implementation provides, as
-discussed in \S~\ref{subsec:fanm_storage}, because a FANM method will
-not generate a subspace in the linear solver and compressed storage
-for sparse matrices are utilized, we expect significant memory savings
-over Newton-Krylov methods. We must measure the memory utilization of
-both of these methods in order to quantify their differences and
-provide additional analysis of the FANM method's merits, or lack
-thereof.
-
-%%---------------------------------------------------------------------------%%
-\section{Navier-Stokes Benchmark Problems}
-\label{sec:ns_benchmarks}
-To verify the FANM method for nonlinear problems, we choose benchmark
-solutions for the 2-dimensional, steady, incompressible Navier-Stokes
-equations on a rectilinear grid in much the same way as Shadid and
-Pawlowski's work on Newton-Krylov methods for the solution of these
-equations \citep{shadid_inexact_1997,pawlowski_globalization_2006}. We
-define these equations as follows:
-\begin{subequations}
-  \begin{gather}
-    \rho \ve{u} \cdot \nabla \ve{u} - \nabla \cdot \ve{T} - \rho
-    \ve{g} = \ve{0}
-    \label{eq:ns_momentum}\\
-    \nabla \cdot \ve{u} = 0
-    \label{eq:ns_continuity}\\
-    \rho C_p \ve{u} \cdot \nabla T + \nabla \cdot \ve{q} = 0\:,
-    \label{eq:ns_energy}
-  \end{gather}
-  \label{eq:navier_stokes}
-\end{subequations}
-where $\rho$ is the fluid density, $\ve{u}$ is the fluid velocity,
-$\ve{g}$ gravity, $C_p$ the specific heat capacity at constant
-pressure of the fluid and $T$ the temperature of the
-fluid. Eq~(\ref{eq:ns_momentum}) provides momentum transport,
-Eq~(\ref{eq:ns_continuity}) provides the mass balance, and
-Eq~(\ref{eq:ns_energy}) provides energy transport with viscous
-dissipation effects neglected. In addition, we close the system with
-the following equations:
-\begin{subequations}
-  \begin{gather}
-    \ve{T} = -P \ve{I} + \mu[\nabla \ve{u} + \nabla \ve{u}^T]
-    \label{eq:ns_stress_tensor}\\
-    \ve{q} = - k \nabla T\:,
-    \label{eq:ns_heat_flux}
-  \end{gather}
-  \label{eq:ns_closure}
-\end{subequations}
-where $\ve{T}$ is the stress tensor, $P$ is the hydrodynamic pressure,
-$\mu$ is the dynamic viscosity of the fluid, $\ve{q}$ is the heat flux
-in the fluid, and $k$ is the thermal conductivity of the fluid. This
-set of strongly coupled equations possesses both the nonlinearities
-and asymmetries that we are seeking for qualification of the FANM
-method. Further, physical parameters within these equations can be
-tuned to enhance the nonlinearities. We will then apply these
-equations to the following three standard benchmark problems.
-
-\subsection{Thermal Convection Cavity Problem}
-\label{subsec:natural_convection_cavity}
-In 1983 a benchmark solution for the natural convection of air in a
-square cavity was published \citep{de_vahl_davis_natural_1983} as
-shown in Figure~\ref{fig:natural_convection_cavity} for the solution
-of the energy, mass, and momentum equations.
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.5}{
-      \input{chapters/navier_stokes_solutions/natural_convection_cavity.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Problem setup for the natural convection cavity
-      benchmark.} \textit{Dirichlet conditions are set for the
-      temperature on the left and right while Neumann conditions are
-      set on the top and bottom of the Cartesian grid. The temperature
-      gradients will cause buoyancy-driven flow. Zero velocity
-      Dirichlet conditions are set on each boundary. No thermal source
-      was present.}}
-  \label{fig:natural_convection_cavity}
-\end{figure}
-In this problem, a rectilinear grid is applied to the unit square. No
-heat flow is allowed out of the top and bottom of the square with a
-zero Neumann condition specified. Buoyancy driven flow is generated by
-the temperature gradient from the cold and hot Dirichlet conditions on
-the left and right boundaries of the box. By adjusting the Rayleigh
-number of the fluid (and therefore adjusting the ratio of convective
-to conductive heat transfer), we can adjust the influence of the
-nonlinear convection term in Eq~(\ref{eq:ns_momentum}). In Shadid's
-work, Rayleigh numbers of up to \sn{1}{6} were used for this benchmark
-on a $100 \times 100$ square mesh grid.
-
-\subsection{Lid Driven Cavity Problem}
-\label{subsec:lid_driven_cavity}
-As an extension of the convection problem, the second benchmark
-problem given by Ghia \citep{ghia_high-re_1982} adds a driver for the
-flow to introduce higher Reynolds numbers into the system, providing
-more inertial force to overcome the viscous forces in the fluid. The
-setup for this problem is equally simple, containing only the
-Dirichlet conditions as given in Figure~\ref{fig:lid_driven_cavity}
-and is only applied to the mass and momentum equations on the unit
-square.
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.5}{
-      \input{chapters/navier_stokes_solutions/lid_driven_cavity.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Problem setup for the lid driven cavity benchmark.}
-    \textit{Dirichlet conditions of zero are set for the velocity on
-      the left and right and bottom while the Dirichlet condition set
-      on the top provides a driving force on the fluid.}}
-  \label{fig:lid_driven_cavity}
-\end{figure}
-The top boundary condition will provide a driver for the flow and its
-variation will in turn vary the Reynolds number of the fluid. An
-increased velocity will generate more inertial forces in the fluid,
-which will overcome the viscous forces and again increase the
-influence of the nonlinear terms in Eq~(\ref{eq:ns_momentum}). Shadid
-also used a $100 \times 100$ grid for this benchmark problem with
-Reynolds numbers up to \sn{1}{4}.
-
-\subsection{Backward-Facing Step Problem}
-\label{subsec:backward_facing_step}
-The third benchmark was generated by Gartling in 1990 and consists of
-both flow over a backward step and an outflow boundary condition
-\citep{gartling_test_1990}. Using the mass and momentum equations
-while neglecting the energy equation, this problem utilizes a longer
-domain with a 1/30 aspect ratio with the boundary conditions as shown
-in Figure~\ref{fig:backward_facing_step}.
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.3}{
-      \input{chapters/navier_stokes_solutions/backward_facing_step.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Problem setup for the backward facing step
-      benchmark.} \textit{Zero velocity boundary conditions are
-      applied at the top and bottom of the domain while the outflow
-      boundary condition on the right boundary is represented by zero
-      stress tensor components in the direction of the flow. For the
-      inlet conditions, the left boundary is split such that the top
-      half has a fully formed parabolic flow profile and the bottom
-      half has a zero velocity condition, simulating flow over a
-      step.}}
-  \label{fig:backward_facing_step}
-\end{figure}
-In this problem, the inflow condition is specified by a fully-formed
-parabolic flow profile over a zero velocity boundary representing a
-step. The flow over this step will generate a recirculating backflow
-under the inlet flow towards the step. As in the lid driven cavity
-problem, the nonlinear behavior of this benchmark and the difficulty
-in obtaining a solution is dictated by the Reynolds number of the
-fluid. In Shadid's work, a $20 \times 400$ non-square rectilinear grid
-was used to discretize the domain with Reynolds number up to
-\sn{5}{2}.
-
-%%---------------------------------------------------------------------------%%
-\section{Analysis of the Navier-Stokes Equations}
-\label{sec:ns_analysis}
-
-%%---------------------------------------------------------------------------%%
-\section{FANM Verification}
-\label{sec:fanm_verification}
-
-%%---------------------------------------------------------------------------%%
-\section{FANM Performance Comparison to Conventional Methods}
-\label{sec:fanm_comparison}

### 2013-07-04 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
adding to overlap scaling section and preparing spreadsheet for more titan data

@@
 formed. Modern nonlinear methods meet both of these requirements with
-Newton methods used in conjunction with Krylov methods for a robust,
-fully implicit solution strategy. Furthermore, modern techniques exist
-that permit the automatic construction of the linear operator
-generated within a Newton method based on the nonlinear residual
-evaluations, providing all of the components necessary for a Monte
-Carlo solver to provide value. We therefore devise a new nonlinear
-method based on the MCSA algorithm and Newton's method and discuss its
-potential benefits.
+Newton methods used in conjunction with Krylov methods for a robust
+solution strategy. Furthermore, modern techniques exist that permit
+the automatic construction of the linear operator generated within a
+Newton method based on the nonlinear residual evaluations, providing
+all of the components necessary for a Monte Carlo solver to provide
+value. We therefore devise a new nonlinear method based on the MCSA
+algorithm and Newton's method and discuss its potential benefits.
 
@@
 
-\subsection{Inexact Newton Methods}
-\label{subsec:newton_methods}
-Inexact Newton methods arise when the Jacobian operator is not exactly
-inverted, resulting in an inexact Newton correction as initially
-described by Dembo and others \citep{dembo_inexact_1982}. For common
-sparse nonlinear systems, which in turn yield a sparse Jacobian
-matrix, this situation occurs when conventional iterative methods are
-applied. In their definition, Dembo formulated inexact methods such
-that they are independent of the linear method used to solve for the
-Newton correction and therefore are amenable to use with any linear
-solver. Furthermore, they bind the convergence of the outer nonlinear
-iteration to the inner linear iteration such that:
-\begin{equation}
-  ||\ve{J}(\ve{u}^k)\delta \ve{u}^k + \ve{F}(\ve{u}^k)|| \leq \eta^k
-  ||\ve{F}(\ve{u}^k)||\:,
-  \label{eq:inexact_newton_forcing}
-\end{equation}
-where $\eta^k \in [0,1)$ is defined as the \textit{forcing term} at
-  the $k$ iterate. Eq~(\ref{eq:inexact_newton_forcing}) then states
-  that the residual generated by the linear solver is bound by the
-  nonlinear residual and how tightly it is bound is defined by the
-  forcing term. This is useful in that we can vary how tightly coupled
-  the convergence of the linear iterations used to generate the Newton
-  correction is to the nonlinear iteration by relaxing or tightening
-  the convergence properties on the linear iterative method. As a
-  result, strategies for determining the forcing term can vary
-  depending on the problem type and can greatly affect the convergence
-  of the method or even prohibit convergence
-  \citep{eisenstat_choosing_1996}. In addition, \textit{globalization
-    methods} may be used to modify the Newton correction in a more
-  desire able direction such that convergence properties can be
-  improved when the initial guess for $\ve{u}$ is poor
-  \citep{pawlowski_globalization_2006}.
-
-\subsection{Newton-Krylov Methods}
-\label{subsec:newton_krylov_methods}
-A form of inexact Newton methods, \textit{Newton-Krylov methods} are
-nonlinear iterative methods that leverage a Krylov subspace method as
-the linear solver for generating the Newton correction
-\citep{kelley_iterative_1995}. As we investigated in
-Chapter~\ref{ch:linear_problem}, Krylov methods are robust and enjoy
-efficient parallel implementations on modern
-architectures. Furthermore, their lack of explicit dependence on the
-operator make them easier to implement than other
-methods. Additionally, although many iterations can become memory
-intensive due to the need to store the Krylov subspace for the
-orthogonalization procedure, at each nonlinear iteration this cost is
-reset as the Jacobian matrix will change due to its dependence on the
-solution vector. This means that for every nonlinear iteration, a
-completely new linear system is formed for generating the Newton
-correction and we can modify the Krylov solver parameters accordingly
-to accommodate this. In most nonlinear problems, the Jacobian operator
-is generally non-symmetric and therefore either Krylov methods with
-long recurrence relations that can handle non-symmetric systems must be
-considered or the Newton correction system must be preconditioned such
-that the operator is symmetric and short recurrence relation methods
-can be potentially be used.
-
-With many Krylov methods available, which to use with the Newton
-method is dependent on many factors including convergence rates and
-memory usage. Several studies have been performed to investigate this
-\citep{mchugh_inexact_1993,knoll_newton-krylov_1995}. In their
-numerical studies in 1995, Knoll and McHugh used the set of highly
-nonlinear and stiff convection-diffusion-reaction equations to solve a
-set of tokamak plasma problems with the goal of measuring solver
-performance with Newton's method. They note several trade offs in using
-Krylov methods with the Newton solver. The first is that the
-optimization condition that results from the constraints (e.g. the
-minimization of the GMRES residual over the Krylov space) can be
-relaxed by restricting the size of the subspace such that only a fixed
-number of subspace vectors may be maintained, thus reducing memory
-requirements. We can also relax the optimization condition by instead
-restarting the recurrence relation with a new set of vectors once a
-certain number of vectors have been generated. The optimization
-condition is maintained over that particular set of vectors, however,
-Knoll and McHugh note that this ultimately slows the convergence rate
-as compared to keeping all vectors as the new set of vectors is not
-necessarily orthogonal to the previous set, and therefore not optimal
-over the entire iteration procedure. The orthogonality condition can
-be relaxed by using a recurrence relation that does not generate a
-strictly orthonormal basis for the Krylov subspace such as the Lanzcos
-biorthogonalization procedure, resulting in memory savings due to the
-shorter Lanzcos recurrence relation.
-
-As a comparison, Knoll and McHugh chose an Arnoldi-based GMRES with a
-fixed vector basis approximately the size of the number of iterations
-required to converge as the long recurrence relation solver and
-conjugate gradients squared (CGS), bi-orthogonalized conjugate
-gradient stabilized (Bi-CGSTAB), and transpose-free quasiminimal
-residual (TFQMR) methods as Lanzcos-based short recurrence relation
-solvers. All solvers were used to compute the right-preconditioned
-Newton correction system. For standard implementations of Newton's
-method where the Jacobian operator was explicitly formed using
-difference equations, all methods exhibited roughly equivalent
-iteration count performance for both the inner linear iterations and
-the outer nonlinear iterations in terms of iterations required to
-converge. Bi-CGSTAB typically performed the best for
-implementations where the Jacobian was explicitly formed and GMRES
-performing best for matrix-free implementations. However, upon
-investigating the convergence of the inner iterations, it was observed
-that the GMRES solver was significantly more robust, always generating
-a monotonically decreasing residual as compared to the Lanzcos-based
-methods which had the tendency to oscillate. Based on these results,
-in all of their future work Knoll and McHugh tended to use GMRES as
-the Krylov solver \citep{knoll_jacobian-free_2004}.
-
-\subsubsection{Jacobian-Free Approximation}
-\label{subsubsec:jacobian_free_approximation}
-In most cases, the Jacobian is difficult to form from the difference
-equations and costly to evaluate for large equation sets. For simple
-nonlinear cases such as the Navier-Stokes equations, the derivatives
-can be computed and coded, but due to the complexity of those
-derivatives and the resulting difference equations this task can be
-tedious, error prone, and must be repeated for every equation
-set. Furthermore, in their 1995 work, Knoll and McHugh also noted that
-a dominating part of their computation time was the evaluation of the
-difference equations for building the Jacobian
-\citep{knoll_newton-krylov_1995}. By recognizing that Krylov methods
-only need the action of the operator on the vector instead of the
-operator itself, the Jacobian can instead be approximated through
-various numerical methods including a difference-based Jacobian-free
-formulation.
-
-Jacobian-Free methods, and in particular \textit{Jacobian-Free
-  Newton-Krylov} (JFNK) methods \citep{knoll_jacobian-free_2004}, rely
-on forming the action of the Jacobian on a vector as required by the
-Krylov solver through a forward difference scheme. In this case, the
-action of the Jacobian on some vector $\ve{v}$ is given as:
-\begin{equation}
-  \ve{J}(\ve{u})\ve{v} = \frac{\ve{F}(\ve{u} + \epsilon \ve{v}) -
-    \ve{F}(\ve{u})}{\epsilon}\:,
-  \label{eq:jacobian_free_product}
-\end{equation}
-where $\epsilon$ is a small number typically on the order of machine
-precision. Kelley \citep{kelley_iterative_1995} points out a potential
-downfall of this formulation in that if the discretization error in
-$\ve{F}(\ve{u})$ is on the order of the perturbation parameter
-$\epsilon$, then the finite difference error from
-Eq~(\ref{eq:jacobian_free_product}) pollutes the solution. In
-addition, Knoll and McHugh noted that for preconditioning purposes,
-part of the Jacobian must still explicitly be formed periodically and
-that linear solver robustness issues were magnified by the matrix-free
-approach due to the first-order approximation. This formation
-frequency coupled with the numerous evaluations of the Jacobian
-approximation create a situation where after so many nonlinear
-iterations, it becomes cheaper to instead fully form the
-Jacobian. For simple equation sets, this may only take 5-10 Newton
-iterations to reach this point while over 30 may be required for
-larger equations sets and therefore larger Jacobians.
-
-\subsubsection{Automatic Differentiation for Jacobian Generation}
-\label{subsubsec:automatic_differentiation}
-If it is acceptable to store the actual Jacobian matrix, other methods
-are available to construct it without requiring hand-coding and
-evaluating derivatives, thus eliminating the associated issues. In
-addition, if any additional equations are added to the system or a
-higher order functional approximation is desired, it would be useful
-to avoid regenerating and coding these derivatives. Becoming more
-prominent in the 1990's, \textit{automatic differentiation} is a
-mechanism by which the derivatives of a function can be generated
-automatically by evaluating it. Automatic differentiation is built on
-the concept that all functions discretely represented in a computer
-are ultimately represented by elementary mathematical operations. If
-the chain rule is applied to those elementary operations, then the
-derivatives of those functions can be computed to the order of
-accuracy of their original discretization in a completely automated
-way \citep{averick_computing_1994}.
-
-The work of Bartlett and others \citep{bartlett_automatic_2006}
-extended initial Fortran-based work in the area of automatic
-differentiation implementations to leverage the parametric type and
-operator overloading features of C++ \citep{stroustrup_c++_1997}. They
-formulate the differentiation problem from an element viewpoint by
-assuming that a global Jacobian can be assembled from local element
-function evaluations of $e_k : \mathbb{R}^{n_k} \rightarrow
-\mathbb{R}^{m_k}$, similar to the finite element assembly procedure
-as:
-\begin{equation}
-  \ve{J}(\ve{u}) = \sum_{i=1}^N \ve{Q}^T_i \ve{J}_k \ve{P}_i\:,
-  \label{eq:fad_global_jacobian}
-\end{equation}
-where $\ve{J}_{k_i} = \partial e_{k_i} / \partial P_i u$ is the
-$k^{th}$ element function Jacobian, $\ve{Q} \in \mathbb{R}^{n_{k_i}
-  \times N}$ is a projector onto the element domain and $\ve{P} \in
-\mathbb{R}^{m_{k_i} \times N}$ a projector onto the element range for
-$\ve{F}(\ve{u}) \in \mathbb{R}^{N \times N}$. The Jacobian matrix for
-each element will therefore have entirely local data in a dense
-structure, eliminating the need for parallel communication and sparse
-techniques during differentiation. Only when all local differentials
-are computed does communication of the Jacobian occur through
-gather/scatter operations in order to properly assembly it. Also of
-benefit is the fact that element-level computations generally consist
-of a smaller number of degrees of freedom, thus reducing memory
-requirements during evaluation as compared to a global formulation of
-the problem. Such a formulation is not limited to finite element
-formulations and is amenable to any scheme where the system is
-globally sparse with degrees of freedom coupled to local domains
-including finite volume representations. The templating capabilities
-of C++ were leveraged with the element-based evaluation and assembly
-scheme as in Eq~(\ref{eq:fad_global_jacobian}) by templating element
-function evaluation code on the evaluation type. If these functions
-are instantiated with standard floating point types then the residual
-is returned. If they are instead instantiated with the
-operator-overloaded automatic differentiation types, both the residual
-and Jacobian are returned.
-
-Of interest to Bartlett, Averick, and the many others that have
-researched automatic differentiation are measures of its performance
-relative to hand-coded derivatives and capturing the Jacobian matrix
-from matrix-free approximations. Given their element-based function
-evaluation scheme, Bartlett's work varied the number of degrees of
-freedom per element and compared both the floating point operation
-count and CPU time for both the templated automatic differentiation
-method and hand-coded derivatives for Jacobian evaluations. Although
-they observed a 50\% increase in floating point operations in the
-templated method over the hand-coded method, run times were observed
-to be over 3 times faster for the templated method. They hypothesize
-that this is due to the fact that the element-based formulation of the
-templated method is causing better utilization of cache and therefore
-faster data access. Furthermore, they observed linear scaling behavior
-for automatic differentiation as the number of degrees of freedom per
-element were increased up to a few hundred. Based on these results,
-this type of automatic differentiation formulation was deemed
-acceptable for use in large-scale, production physics codes.
-
 %%---------------------------------------------------------------------------%%
@@
 parallel implementation.
-
-\subsection{Jacobian Storage vs. Subspace Storage and Restarts}
-\label{subsec:fanm_storage}
-To gauge the memory benefits of a FANM method over a Krylov-based
-scheme, we must look at the storage requirements of the Jacobian
-matrix as compared to the Krylov subspace vectors for sparse linear
-problems. Saad's text provides us relations for both of these cases
-\citep{saad_iterative_2003}. Beginning with the Jacobian storage, for
-sparse problems production linear algebra library implementations
-utilize \textit{compressed row storage} to efficiently store the
-non-zero components of a sparse matrix while still allowing for all
-standard parallel matrix computations to be performed. Per
-\S~\ref{sec:parallel_krylov_methods}, recall that efficient parallel
-matrix-vector multiplications rely on processors knowing which other
-processors contain their neighboring matrix and vector
-elements. Typically this information is represented by a graph which
-must be stored along with the matrix elements themselves. To store the
-elements, consider the following sparse matrix:
-\[
-  \ve{A} =
-  \begin{bmatrix}
-    2 & 0 & 8 & 0 & 0 & 0 \\ 
-    4 & 5 & 0 & 1 & 0 & 0 \\ 
-    0 & 2 & 1 & 0 & 1 & 0 \\ 
-    0 & 0 & 3 & 7 & 0 & 2 \\ 
-    0 & 0 & 0 & 4 & 9 & 0 \\
-    0 & 0 & 0 & 0 & 9 & 1 \\
-  \end{bmatrix}\:.
-\]
-The compressed form of \ve{A} would then be:
-\[
-  \begin{array}{lccccccccccccccc}
-    \ve{A}\ values: & 2 & 8 & 4 & 5 & 1 & 2 & 1 & 1 & 3 & 7 & 2 & 4 & 9
-    & 9 & 1 \\ column: & 1 & 3 & 1 & 2 & 4 & 2 & 3 & 5 & 3 & 4 & 6 & 4
-    & 5 & 5 & 6 \\ row\ start: & 1 & 3 & 6 & 9 & 12 & 14 & 16 \\
-  \end{array}
-\]
-where the first row contains the non-zero values of the matrix
-$\ve{A}$ starts at, the second row contains the column index of those
-values, and the third row contains the index of the A value and column
-index rows that each matrix row starts at ending with the start of the
-next row. For this simple example, we actually break even by storing
-36 elements in the full matrix and 36 pieces of data in the compressed
-format. However, for large, sparse matrices there is a significant
-storage savings using the compressed format.
-
-In his discussion on orthogonalization schemes for Krylov subspaces,
-Saad provides us with space and time complexities for these
-operations. To find a solution vector $\ve{x} \in \mathbb{R}^{N \times
-  N}$ with $m$ Krylov iterations (and therefore $m+1$ subspace
-vectors) the storage requirement is then $(m+1) N$ for most common
-orthogonalization techniques. Using our simple example above (even
-though we did not see any gains in storage over the full matrix
-representation), if it takes 10 GMRES iterations using Arnoldi
-orthogonalization to converge to a solution, then a total of 66 pieces
-of data are required to be stored instead of 36 for the explicitly
-formed matrix case (we do not consider graph storage here for
-comparison as both methods will require the graph for parallel
-operations). For larger sparse matrices, this disparity in memory
-requirements will be even greater, especially for ill-conditioned
-systems that require many GMRES iterations to converge.
 
@@
 %%---------------------------------------------------------------------------%%
-\section{Analysis of the Navier-Stokes Equations\ }
-\label{sec:ns_analysis}
-
-%%---------------------------------------------------------------------------%%
 \section{FANM Verification\ }

### 2013-07-04 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
fixed mcsa equation in parallel chapter
adding to overlap scaling section and preparing spreadsheet for more titan data

@@
 trivial. Recall the MCSA iteration procedure presented again here for
-clarity.
+clarity:
 \begin{subequations}
@@
     \label{eq:mcsa_r5}
-    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:,
+    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:.
   \end{gather}
@@
 
+In the next set of scaling studies, we will explore how adding overlap
+to the subdomains in the problem affects runtimes and parallel
+performance as compared to the case with pure domain decomposition. To
+do this, we utilize the set of analytic relations developed in
+\S~\ref{sec:analytic_framework} to design these numerical
+experiments. For the neutron diffusion problem used in these scaling
+studies, the linear operator had a spectral radius of 0.787 and a
+weight cutoff of \sn{1}{-2}. Using Eq~(\ref{eq:analytic_k}) gives an
+expected random walk length of 19.22 transitions for any given history
+in the problem. Using the mean-chord approximation given by
+Eq~(\ref{eq:mean_chord_domain_leakage}), we then expect approximately
+5.3\% of the histories born in a particular domain to leak out after
+the first transport stage. By adding overlap, we aim to reduce this
+fraction of communicated histories and therefore boost the parallel
+performance of the algorithm. To determine the amount of overlap to
+test for the scaling studies, we use a variation of
+Eq~(\ref{eq:step_k_length}) to determine how many discrete states a
+history will move on average from its birth site. We modify
+Eq~(\ref{eq:step_k_length}) to only account for discrete movements:
+\begin{equation}
+  \sqrt{\bar{n^2_k}} = n_s \sqrt{k}\:,
+  \label{eq:discrete_distance}
+\end{equation}
+where $\sqrt{\bar{n^2_k}}$ is now the root-mean-squared number of
+discrete states a history will move from its birth site. For our
+diffusion problem used in these scaling studies, we compute a value of
+2.63 states per history for any of the dimensions of the system. For
+this diffusion problem, this means that we can expect to end up on
+average 2.63 states in both the $i$ and $j$ dimensions away from the
+starting state. Therefore, we will use the overlap generation
+algorithm outlined in \S~\ref{subsec:msod_generation} such that we
+parameterically grow the local domain using 1, 2, 3, and 5 levels to
+measure how scalability is affected when this root-mean-squared number
+of states is considered. For example, using 3 levels of overlap should
+keep over 50\% of the histories that would be communicated in the pure
+domain decomposition case on-process as this is greater than the
+root-mean-squared number of states we would expect a history to
+travel.
+
+For the performance metrics, we may use the same weak and strong
+scaling efficiency metrics that we used for the pure domain
+decomposition case. As we are comparing to the case without overlap
+and are looking to boost performance relative to the case without
+overlap, for both weak and strong scaling calculations we will use the
+16-core case without overlap as the reference computation for all
+efficiency measures reported. Doing this, permits an analysis of well
+the computational resources are utilized with the overlap relative to
+the case without.
+
 \clearpage
 
-\subsection{Multiple Set Decomposition}
+\subsection{Multiple Set Domain Decomposition}
 \label{subsec:msod_decomposition}
@@
 we are using residual Monte Carlo to compute the correction vector
-casts MCSA as a stochastic realization of a additive Schwarz method
+casts MCSA as a stochastic realization of an additive Schwarz method
 with a fixed point iterations used as a relaxation step\footnote{See

### 2013-07-05 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
reworking titan scaling data and adding addtional runs for averaging

@@
 diffusion problem used in these scaling studies, we compute a value of
-2.63 states per history for any of the dimensions of the system. For
-this diffusion problem, this means that we can expect to end up on
-average 2.63 states in both the $i$ and $j$ dimensions away from the
-starting state. Therefore, we will use the overlap generation
-algorithm outlined in \S~\ref{subsec:msod_generation} such that we
-parameterically grow the local domain using 1, 2, 3, and 5 levels to
-measure how scalability is affected when this root-mean-squared number
-of states is considered. For example, using 3 levels of overlap should
-keep over 50\% of the histories that would be communicated in the pure
-domain decomposition case on-process as this is greater than the
+2.63 states per history for any of the dimensions of the system. This
+means that we can expect to end up on average 2.63 states in both the
+$i$ and $j$ dimensions away from the starting state. Therefore, we
+will use the overlap generation algorithm outlined in
+\S~\ref{subsec:msod_generation} such that we parameterically grow the
+local domain using 1, 2, 3, and 5 levels to measure how scalability is
+affected when this root-mean-squared number of states is
+considered. For example, using 3 levels of overlap should keep over
+50\% of the histories that would be communicated in the pure domain
+decomposition case on-process as this is greater than the
 root-mean-squared number of states we would expect a history to
@@
 16-core case without overlap as the reference computation for all
-efficiency measures reported. Doing this, permits an analysis of well
-the computational resources are utilized with the overlap relative to
-the case without.
+efficiency measures reported.
 

### 2013-07-06 doc/parallel_verficiation/compare_flux.cpp
adding parallel verification tool for linear problems

@@ -1,0 +1,49 @@
+#include <iostream>
+#include <iomanip>
+#include <vector>
+#include <cmath>
+#include <cstdlib>
+#include <fstream>
+#include <algorithm>
+#include <string>
+#include <cassert>
+
+int main( int argc, char * argv[] )
+{
+    std::string file_1 = "neutron.dat";
+    std::string file_2 = "neutron.dat";
+
+    int global_length;
+
+    std::ifstream data_file_1;
+    data_file_1.open( file_1.c_str() );
+    data_file_1 >> global_length;
+    std::vector<double> global_vector_1( global_length );
+    for ( int i = 0; i < global_length; ++i )
+    {
+        data_file_1 >> global_vector_1[i];
+    }
+    data_file_1.close();
+
+    std::ifstream data_file_2;
+    data_file_2.open( file_2.c_str() );
+    data_file_2 >> global_length;
+    std::vector<double> global_vector_2( global_length );
+    for ( int i = 0; i < global_length; ++i )
+    {
+        data_file_2 >> global_vector_2[i];
+    }
+    data_file_2.close();
+
+    std::vector<double> difference( global_length );
+    for ( int i = 0; i < global_length; ++i )
+    {
+        difference[i] = std::abs( global_vector_1[i] - global_vector_2[i] );
+    }
+
+    std::cout << "Max Difference: " 
+              << *std::max_element( difference.begin(), difference.end() );
+
+    std::cout << "Min Difference: " 
+              << *std::min_element( difference.begin(), difference.end() );
+}

### 2013-07-06 doc/parallel_verficiation/compare_flux.cpp
adding verification data

@@ -10,8 +10,8 @@
 
 int main( int argc, char * argv[] )
 {
-    std::string file_1 = "neutron.dat";
-    std::string file_2 = "neutron.dat";
+    std::string file_1 = "cg_reference.dat";
+    std::string file_2 = "cg.dat";
 
     int global_length;
 
@@ -42,8 +42,10 @@ int main( int argc, char * argv[] )
     }
 
     std::cout << "Max Difference: " 
-              << *std::max_element( difference.begin(), difference.end() );
+              << *std::max_element( difference.begin(), difference.end() )
+              << std::endl;
 
     std::cout << "Min Difference: " 
-              << *std::min_element( difference.begin(), difference.end() );
+              << *std::min_element( difference.begin(), difference.end() )
+              << std::endl;
 }

### 2013-07-06 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
adding parallel verification data

@@
 parallel with respect to reference solutions provided by production
-Krylov methods.
+Krylov methods. To verify correctness, we use solutions the neutron
+diffusion system presented in
+Appendix~\ref{chap:diffusion_problem}. As the reference computation,
+serial results will be used from the conjugate gradient solver in
+Belos, a linear solvers package in the Trilinos scientific computing
+libraries \citep{heroux_overview_2005}. In identical fashion to the
+verification for the $SP_N$ equations performed in
+\S~\ref{sec:spn_mcsa_verification}, we will solve the neutron
+diffusion problem in parallel using multiple cores on a desktop
+machine. For each parallel solution, the solution at each grid point
+in the system will be compared to the solution at the same grid point
+in the serial reference case with the minimum and maxiumum absolute
+differences of these values reported.
+
+GMRES and conjugate gradient solvers from the Belos package were used
+on 4 cores as a means of both assuring the Krylov solvers produce the
+same results on multiple cores and as a means of providing an
+acceptable range of minimum and maximum values in the comparison. For
+MCSA, a range of MSOD parameters will be used such that problems with
+multiple sets and multiple blocks may be tested along with
+overlap. When multiple sets were leveraged, a single history was used
+to compute the MCSA correction for every DOF in the problem. For all
+solvers, the flux was converged to a tolerance of \sn{1}{-8} and
+therefore if the comparisons agree within this tolerance the solutions
+will be deemed
+correct. Table~\ref{tab:parallel_verification_parameters} gives the
+parameters for the verification problem and
+Table~\ref{tab:parallel_mcsa_verification} gives the results of the
+calculations. Solver parameters used for MCSA are given in
+Table~\ref{tab:parallel_mcsa_parameters}.
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lc}\hline\hline
+      \multicolumn{1}{l}{Parameter}& 
+      \multicolumn{1}{c}{Value}\\\hline
+      $dx$ & 0.1 \\
+      $dy$ & 0.1 \\
+      $N_x$ & 400 \\
+      $N_y$ & 400 \\
+      $N$ & 160,000 \\
+      $\Sigma_a$ & 5.0 \\
+      $\Sigma_s$ & 1.0 \\
+      Boundary Type & Vacuum \\
+      Linear Solver Tolerance & \sn{1}{-8} \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Parallel MCSA verification model problem parameters.}
+    \textit{The neutron diffusion equation in 2 dimensions is used for
+      the model problem.}}
+  \label{tab:parallel_verification_parameters}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lc}\hline\hline
+      \multicolumn{1}{l}{Parameter}& 
+      \multicolumn{1}{c}{Value}\\\hline
+      Histories & 1 per DOF \\
+      Weight Cutoff & \sn{1}{-2} \\
+      Fixed Point & Richardson \\
+      Estimator & Collision \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Parallel MCSA verification solver parameters.}}
+  \label{tab:parallel_mcsa_parameters}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lccccll}\hline\hline
+      \multicolumn{1}{l}{Solver}& 
+      \multicolumn{1}{c}{Blocks}& 
+      \multicolumn{1}{c}{Sets}& 
+      \multicolumn{1}{c}{Cores}& 
+      \multicolumn{1}{c}{Overlap}& 
+      \multicolumn{1}{l}{Min Difference}& 
+      \multicolumn{1}{l}{Max Difference}\\\hline
+      Conjugate Gradient & - & - & 4 & - & 0.0 & \sn{1.11022}{-16} \\
+      GMRES & - & - & 4 & - & \sn{5.64215}{-13} & \sn{3.83348}{-9} \\
+      MCSA & 2 & 1 & 2 & 0 & \sn{8.60423}{-16} & \sn{7.6806}{-9} \\
+      MCSA & 3 & 1 & 3 & 0 & \sn{6.41154}{-15} & \sn{7.65397}{-9} \\
+      MCSA & 4 & 1 & 4 & 0 & \sn{2.66454}{-15} & \sn{7.62623}{-9} \\
+      MCSA & 2 & 2 & 4 & 0 & \sn{1.52656}{-15} & \sn{7.57108}{-9} \\
+      MCSA & 1 & 4 & 4 & 0 & \sn{5.55112}{-16} & \sn{7.61631}{-9} \\
+      MCSA & 8 & 1 & 8 & 0 & \sn{6.13398}{-15} & \sn{7.51744}{-9} \\
+      MCSA & 4 & 2 & 8 & 0 & \sn{1.03251}{-14} & \sn{7.62601}{-9} \\
+      MCSA & 2 & 4 & 8 & 0 & \sn{5.13478}{-15} & \sn{7.60777}{-9} \\
+      MCSA & 1 & 8 & 8 & 0 & \sn{1.80411}{-15} & \sn{7.60651}{-9} \\
+      MCSA & 2 & 1 & 2 & 5 & \sn{2.47025}{-15} & \sn{7.58882}{-9} \\
+      MCSA & 3 & 1 & 3 & 5 & \sn{2.44249}{-15} & \sn{7.6326}{-9} \\
+      MCSA & 4 & 1 & 4 & 5 & \sn{7.49401}{-16} & \sn{7.49014}{-9} \\
+      MCSA & 2 & 2 & 4 & 5 & \sn{1.19349}{-15} & \sn{7.61814}{-9} \\
+      MCSA & 1 & 4 & 4 & 5 & \sn{5.55112}{-16} & \sn{7.61631}{-9} \\
+      MCSA & 8 & 1 & 8 & 5 & \sn{2.05391}{-15} & \sn{7.63955}{-9} \\
+      MCSA & 4 & 2 & 8 & 5 & \sn{9.57567}{-15} & \sn{7.61313}{-9} \\
+      MCSA & 2 & 4 & 8 & 5 & \sn{5.7454}{-15} & \sn{7.60767}{-9} \\
+      MCSA & 1 & 8 & 8 & 5 & \sn{1.80411}{-15} & \sn{7.60651}{-9} \\
+      MCSA & 2 & 1 & 2 & 10 & \sn{1.08247}{-15} & \sn{7.63231}{-9} \\
+      MCSA & 3 & 1 & 3 & 10 & \sn{3.27516}{-15} & \sn{7.60211}{-9} \\
+      MCSA & 4 & 1 & 4 & 10 & \sn{4.996}{-16} & \sn{7.64756}{-9} \\
+      MCSA & 2 & 2 & 4 & 10 & \sn{4.13558}{-15} & \sn{7.60259}{-9} \\
+      MCSA & 1 & 4 & 4 & 10 & \sn{5.55112}{-16} & \sn{7.61631}{-9} \\
+      MCSA & 8 & 1 & 8 & 10 & \sn{9.71445}{-16} & \sn{7.58535}{-9} \\
+      MCSA & 4 & 2 & 8 & 10 & \sn{7.77156}{-16} & \sn{7.65978}{-9} \\
+      MCSA & 2 & 4 & 8 & 10 & \sn{4.16334}{-16} & \sn{7.63191}{-9} \\
+      MCSA & 1 & 8 & 8 & 10 & \sn{1.80411}{-15} & \sn{7.60651}{-9} \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Parallel MCSA verification results.}
+    \textit{Difference values given are absolute values. The neutron
+      diffusion equation in 2 dimensions is used for the model problem
+      with the parameters given in
+      Table~\ref{tab:parallel_verification_parameters}. A serial
+      conjugate gradient calculation is used as the reference
+      computation and neutron fluxes are compared
+      element-wise. Parameters used for the MCSA solver are given in
+      Table~\ref{tab:parallel_mcsa_parameters}. All solvers including
+      the reference computation were converged to tolerance of
+      \sn{1}{-8}.}}
+  \label{tab:parallel_mcsa_verification}
+\end{table}
+
+From the data we see that for all solvers, the parallel results agree
+within the solution tolerance of the reference serial results for all
+DOF in the system. In addition, we do not notice any systematic
+differences in the results as a function of number of blocks, number
+of sets, or amount of overlap. In addition, the parallel GMRES results
+have maximum absolute differences of the same magnitude as all MCSA
+results and a larger minimum absolute difference than all MCSA results
+by three orders of magnitude. Based on these results, MCSA parallized
+with an MSOD, and in particular the implementation used here,
+generates the correct results with respect to the serial conjugate
+gradient reference computation.
+
+\clearpage
 

### 2013-07-06 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
adding msod scaling section

@@
 \subsection{Multiple Set Domain Decomposition}
-\label{subsec:msod_decomposition}
-
-Given that overlap was not successful in boosting parallel
-efficiencies, we next look to apply replication to the problem in the
-form of multiple sets. Before performing parallel scaling studies
-using multiple sets, several aspects of the replicated parallel
-problem must be considered in order to effectively craft the studies
-and interpret their results. First is the question of how many
-stochastic histories to use to generate the correction vector in the
-MCSA iteration. For this work we consider two choices; the split case
-and the fully replicated case. As an example these cases, consider a
-single set problem where 20,000 histories are used in the Monte Carlo
-solve. If we choose the split case, then solving the same problem with
-2 sets would give 10,000 histories in each set and 4 sets would give
-5,000 histories in each set. Splitting the histories in this way
-provides the same global number of histories used to compute the
-correction in the single set problem. We expect to reduce the time to
-solution by doing this as each individual set solve will occur faster
-in an almost linear fashion with the reduced number of histories. For
-the replicated history case, each additional set in the problem will
-use just as many histories as the single set case. Using 2 sets with
-fully replicated histories for our example problem then has each set
+\label{subsec:ms_decomposition}
+
+We next look to apply replication to the problem in the form of
+multiple sets. Before performing parallel scaling studies using
+multiple sets, several aspects of the replicated parallel problem must
+be considered in order to effectively craft the studies and interpret
+their results. First is the question of how many stochastic histories
+to use to generate the correction vector in the MCSA iteration. For
+this work we consider two choices; the split case and the fully
+replicated case. As an example these cases, consider a single set
+problem where 20,000 histories are used in the Monte Carlo solve. If
+we choose the split case, then solving the same problem with 2 sets
+would give 10,000 histories in each set and 4 sets would give 5,000
+histories in each set. Splitting the histories in this way provides
+the same global number of histories used to compute the correction in
+the single set problem. We expect to reduce the time to solution by
+doing this as each individual set solve will occur faster in an almost
+linear fashion with the reduced number of histories. For the
+replicated history case, each additional set in the problem will use
+just as many histories as the single set case. Using 2 sets with fully
+replicated histories for our example problem then has each set
 computing 20,000 histories with 40,000 global histories in the
@@
 
+\subsection{Multiple-Set Overlapping-Domain Decomposition}
+\label{subsec:msod_decomposition}
+
+\clearpage
+
 \subsection{Subdomain Neumann-Ulam}
@@
 subsequently the MSOD implementation, one may ask then ask if there is
-any need to communicate histories from domain-to-domain at all. Overlap
-was not successful because so few histories needed to be communicated
-in the first place that keeping them on-process resulted in more work
-than the communications savings that occurred. If that is true, then
-perhaps without any overlap or communication, just doing local Monte
-Carlo may be satisfactory for accelerating the MCSA iteration. In this
-case, histories near the boundaries of the domain may be truncated
-before reaching the weight cutoff if the reach the domain boundary
-first. This truncation has the potential to increase the stochastic
-uncertainty of the Neumann-Ulam result and therefore increase the
-number of MCSA iterations required to converge. In addition, the
-scalability of the method will then be purely limited by the parallel
-matrix and vector operations required to implement the remainder of
-the algorithm.
+any need to communicate histories from domain-to-domain at
+all. Overlap was not as successful because so few histories needed to
+be communicated in the first place that keeping them on-process
+resulted in more work than the communications savings that
+occurred. If that is true, then perhaps without any overlap or
+communication, just doing local Monte Carlo may be satisfactory for
+accelerating the MCSA iteration. In this case, histories near the
+boundaries of the domain may be truncated before reaching the weight
+cutoff if the reach the domain boundary first. This truncation has the
+potential to increase the stochastic uncertainty of the Neumann-Ulam
+result and therefore increase the number of MCSA iterations required
+to converge. In addition, the scalability of the method will then be
+purely limited by the parallel matrix and vector operations required
+to implement the remainder of the algorithm.
 
@@
 performance of the method as iterative performance is maintained
-without it. Overlap is not effective because histories relinquish most
-of their useful information within the first few steps (which is why
-we can use such a large weight cutoff and there is an insensitivity to
-weight cutoff as shown in
+without it. Overlap is not as effective because histories relinquish
+most of their useful information within the first few steps (which is
+why we can use such a large weight cutoff and there is an
+insensitivity to weight cutoff as shown in
 Figure~\ref{fig:estimator_wc_iters}). Keeping them around longer with

### 2013-07-06 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
adding overlap data to document

@@
 parallel with respect to reference solutions provided by production
-Krylov methods. To verify correctness, we use solutions the neutron
+Krylov methods. To verify correctness, we use solutions to the neutron
 diffusion system presented in
@@
 \citep{heroux_overview_2005}. In general, the Belos package produces
-solutions to the neutron diffusion problem over an order of magnitude
-faster than the MCLS library for serial computations. Therefore, when
-we directly compare these two codes, the MCLS parallel efficiencies
-will appear artificially inflated relative to the Belos efficiencies
-due to the fact that the MCLS ratio of work to communication will be
-much larger than that for the Belos calculations.
+solutions to the neutron diffusion problem nearly two orders of
+magnitude faster than the MCLS library for serial computations using
+the conjugate gradient solver. Therefore, when we directly compare
+these two codes, the MCLS parallel efficiencies will appear
+artificially inflated relative to the Belos efficiencies due to the
+fact that the MCLS ratio of work to communication will be much larger
+than that for the Belos calculations. For every scaling study
+presented, the calculations were performed 3 times with the average
+values reported for wall time and used for the efficiency
+computations.
 
@@
 one stochastic history used per DOF in the Neumann-Ulam solve to
-compute the correction for the
-acceleration. Figure~\ref{fig:titan_pure_strong} gives the results of
-the this study with the given absolute efficiencies computed using
-Eq~(\ref{eq:strong_scaling_absolute}).
+compute the correction for the acceleration with the remainder of the
+problems and solver parameters the same as those for the verification
+calculation given in Table~\ref{tab:parallel_verification_parameters}
+and Table~\ref{tab:parallel_mcsa_parameters}. As with all of the
+scaling studies, 3 calculations for all data points were performed
+with the average wall time reported and used for the efficiency
+calculations. Figure~\ref{fig:titan_pure_strong_time} gives the wall
+time per iteration for these calculations while
+Figure~\ref{fig:titan_pure_strong} gives the results of the this study
+with the given absolute efficiencies computed using
+Eq~(\ref{eq:strong_scaling_absolute}). We report wall time per
+iteration due to the fact that MCSA converged on the solution in 22
+iterations while the Krylov solvers converged in 19 iterations.
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/parallel_mc/titan_pure_strong_time.pdf}
+  \end{center}
+  \caption{\textbf{Pure domain decomposition strong scaling wall time
+      per iteration.} \textit{Wall time is reported per iteration
+      because the methods converged in different numbers of
+      iterations. MCLS is over an order of magnitude slower
+      arithmetically than the Krylov solvers. GMRES executes more
+      operations than conjugate gradient and is therefore slower as
+      well.}}
+  \label{fig:titan_pure_strong_time}
+\end{figure}
 
@@
       efficiency.}  \textit{Super-linear speed-up is from memory
-      thrashing in the base case. MCLS is an order of magnitude slower
-      arithmetically, causing the higher efficiency. GMRES executes
-      more operations than conjugate gradient, also creating a higher
-      efficiency.}}
+      thrashing in the base case. MCLS is over an order of magnitude
+      slower arithmetically, causing the higher efficiency. GMRES
+      executes more operations than conjugate gradient, also creating
+      a higher efficiency.}}
   \label{fig:titan_pure_strong}
@@
 
-There are several key features of Figure~\ref{fig:titan_pure_strong}
-worth noting. First, up to around 5,000 cores, all methods experience
-a super-linear improvement in parallel efficiencies. For all methods,
-this increase is not a function of the algorithmic efficiency as those
-computed using Eq~(\ref{eq:algorithmic_efficiency}) remained at unity,
-meaning that all computations converged in the same number of
-iterations relative to the base case\footnote{When the algorithmic
-  efficiency remains at unity for all calculations, the implementation
-  efficiency is equivalent to the absolute efficiency and therefore is
-  not presented here.}. Instead, what is happening here and was
-observed by Brunner and Brantley in their work is that the large
-problem used to scale to O(10,000) cores was large enough that memory
-thrashing was occurring in the 16 core base case. As the number of
-cores in the strong scaling study was increased and the local problem
-size decreased, more of the local problem fit into the cache of the
-system, thus causing improved serial performance and therefore
-enhanced runtimes. In this region of super-linear efficiencies, we
-also observe larger efficiencies for GMRES as compared to conjugate
-gradient. This occurs due to the fact that conjugate gradient has a
-static set of operations performed at every iteration while GMRES
-operates on a continually growing subspace. In addition, the
-orthogonalization operations required to generate that subspace
-consume much more time than the basic conjugate gradient operations,
-causing a larger serial runtime and subsequently higher
-efficiencies. For the same reasons, we also observe higher
+There are several key features of
+Figure~\ref{fig:titan_pure_strong_time} and
+Figure~\ref{fig:titan_pure_strong} worth noting. First, up to around
+5,000 cores, all methods experience a super-linear improvement in
+parallel efficiencies. For all methods, this increase is not a
+function of the algorithmic efficiency as those computed using
+Eq~(\ref{eq:algorithmic_efficiency}) remained at unity, meaning that
+all computations converged in the same number of iterations relative
+to the base case\footnote{When the algorithmic efficiency remains at
+  unity for all calculations, the implementation efficiency is
+  equivalent to the absolute efficiency and therefore is not presented
+  here.}. Instead, what is happening here and was observed by Brunner
+and Brantley in their work is that the large problem used to scale to
+O(10,000) cores was large enough that memory thrashing was occurring
+in the 16 core base case. As the number of cores in the strong scaling
+study was increased and the local problem size decreased, more of the
+local problem fit into the cache of the system, thus causing improved
+serial performance and therefore enhanced runtimes. In this region of
+super-linear efficiencies, we also observe larger efficiencies for
+GMRES as compared to conjugate gradient. This occurs due to the fact
+that conjugate gradient has a static set of operations performed at
+every iteration while GMRES operates on a continually growing
+subspace. In addition, the orthogonalization operations required to
+generate that subspace consume much more time than the basic conjugate
+gradient operations, causing a larger serial runtime and subsequently
+higher efficiencies. For the same reasons, we also observe higher
 efficiencies for the MCSA implementation due to the serial performance
@@
 increase the number of processors used in the
-simulation. Figure~\ref{fig:titan_weak_absolute} gives the weak
-scaling absolute efficiencies for this problem as computed by
+simulation. Figure~\ref{fig:titan_pure_weak_time} gives the wall
+time per iteration and Figure~\ref{fig:titan_weak_absolute} gives the
+weak scaling absolute efficiencies for this problem as computed by
 Eq~(\ref{eq:weak_scaling_absolute}). In general, the relative trends
@@
   \begin{center}
+    \includegraphics[width=6in]{chapters/parallel_mc/titan_pure_weak_time.pdf}
+  \end{center}
+  \caption{\textbf{Pure domain decomposition weak scaling wall time
+      per iteration.} \textit{Wall time is reported per iteration
+      because the methods converged in different numbers of
+      iterations. MCLS is over an order of magnitude slower
+      arithmetically than the Krylov solvers. GMRES executes more
+      operations than conjugate gradient and is therefore slower as
+      well.}}
+  \label{fig:titan_pure_weak_time}
+\end{figure}
+
+\begin{figure}[t!]
+  \begin{center}
     \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_absolute.pdf}
@@
   \caption{\textbf{Pure domain decomposition weak scaling absolute
-      efficiency.} \textit{MCLS is an order of magnitude slower
+      efficiency.} \textit{MCLS is an over order of magnitude slower
       arithmetically, causing the higher efficiency. GMRES executes
@@
 converge decreased from 20 in the base case to 19 and then to 18 for
-the final set of computations. When this reduction in iterations is
-used to account for the observed runtimes and computed absolute
-efficiencies for weak scaling we end up with the implementation
-efficiencies computed using Eq~(\ref{eq:implementation_efficiency})
-given in Figure~\ref{fig:titan_weak_implementation}. Because MCSA
-maintains the same iterative performance as a function of global
-problem size in the weak scaling study, the implementation
-efficiencies compare more favorably to the Krylov methods. In effect,
-we are measuring the parallel performance of a single iteration for
-each method by looking at these implementation efficiencies. As the
-global problem size increases, MCSA is less prone to a drop in
-parallel efficiency. When optimization has been completed and runtimes
-potentially approach those of the Krylov methods, this comparison may
-not be as favorable. Overall, weak scaling performance the parallel
-MCSA algorithm is excellent for the pure domain decomposition case
-with absolute parallel efficiencies of over 80\% observed at over
-32,000 cores.
+the final set of weak scaling computations. When this reduction in
+iterations is used to account for the observed runtimes and computed
+absolute efficiencies for weak scaling we end up with the
+implementation efficiencies computed using
+Eq~(\ref{eq:implementation_efficiency}) given in
+Figure~\ref{fig:titan_weak_implementation}. Because MCSA maintains the
+same iterative performance as a function of global problem size in the
+weak scaling study, the implementation efficiencies compare more
+favorably to the Krylov methods. Using the implementation efficiency,
+we are effectively measuring the parallel performance of a single
+iteration for each method by looking at these implementation
+efficiencies. As the global problem size increases, MCSA is less prone
+to a drop in parallel efficiency. When optimization has been completed
+and runtimes potentially approach those of the Krylov methods, this
+comparison may not be as favorable. Overall, weak scaling performance
+the parallel MCSA algorithm is excellent for the pure domain
+decomposition case with absolute parallel efficiencies of over 80\%
+observed at over 32,000 cores.
 
@@
   \caption{\textbf{Binary tree vs. master/slave communication scheme
-      absolute efficiency.} \textit{Pure domain decomposition absolute
-      strong scaling. The master/slave scheme will not scale beyond
-      $O(1,000)$ cores due to load imbalance.}}
+      strong scaling absolute efficiency.} \textit{The master/slave
+      scheme will not scale beyond $O(1,000)$ cores due to load
+      imbalance.}}
   \label{fig:titan_strong_bvsm}
@@
   \caption{\textbf{Binary tree vs. master/slave communication scheme
-      absolute efficiency.}  \textit{Pure domain decomposition
-      absolute weak scaling. The master/slave scheme will not scale
-      beyond $O(1,000)$ cores due to load imbalance.}}
+      weak scaling absolute efficiency.}  \textit{The master/slave
+      scheme will not scale beyond $O(1,000)$ cores due to load
+      imbalance.}}
   \label{fig:titan_weak_bvsm}
@@
 
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_overlap.pdf}
+  \end{center}
+  \caption{\textbf{Strong scaling absolute efficiency with varying
+      levels of overlap.} \textit{Values of 0, 1, 2, 3, and 5 levels
+      of overlap were used as the problem had a root-mean-squared
+      number of states of 2.63 for each history.}}
+  \label{fig:titan_strong_overlap}
+\end{figure}
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_overlap_diff.pdf}
+  \end{center}
+  \caption{\textbf{Strong scaling absolute efficiency difference with
+      respect to the 0 overlap base case.} \textit{Values of 1, 2, 3,
+      and 5 levels of overlap are compared here with the absolute
+      difference of the absolute efficiencies presented.}}
+  \label{fig:titan_strong_overlap_diff}
+\end{figure}
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_overlap.pdf}
+  \end{center}
+  \caption{\textbf{Weak scaling absolute efficiency with varying
+      levels of overlap.} \textit{Values of 0, 1, 2, 3, and 5 levels
+      of overlap were used as the problem had a root-mean-squared
+      number of states of 2.63 for each history.}}
+  \label{fig:titan_weak_overlap}
+\end{figure}
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_overlap_diff.pdf}
+  \end{center}
+  \caption{\textbf{Weak scaling absolute efficiency difference with
+      respect to the 0 overlap base case.} \textit{Values of 1, 2, 3,
+      and 5 levels of overlap are compared here with the absolute
+      difference of the absolute efficiencies presented.}}
+  \label{fig:titan_weak_overlap_diff}
+\end{figure}
+
 \clearpage
@@
 
-\subsection{Multiple-Set Overlapping-Domain Decomposition}
-\label{subsec:msod_decomposition}
-
-\clearpage
-
 \subsection{Subdomain Neumann-Ulam}

### 2013-07-07 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
finished preliminary draft of parallel chapter. editing is next

@@
 Using the parallel performance metrics presented in the previous
-section, we will next performance various scaling studies using an
-implementation of the parallel MCSA algorithm to assess the quality of
-the algorithm using the neutron diffusion problem presented in
+section, we will next measure performance in various scaling studies
+using an implementation of the parallel MCSA algorithm to assess its 
+quality using the neutron diffusion problem presented in
 Appendix~\ref{chap:diffusion_problem}. For these scaling studies,
@@
 performance results for the same problems using a set of production
-parallel Krylov solvers. For the scaling results presented here, all
-parallel Monte Carlo results were computed using the Monte Carlo
-Linear Solvers\footnote{MCLS is an open source library and can be
-  downloaded at \url{https://github.com/sslattery/MCLS}} (MCLS)
-library generated specifically for this work while the conjugate
-gradient and GMRES Krylov solvers used are part of the Belos linear
-solver package within the Trilinos framework
-\citep{heroux_overview_2005}. In general, the Belos package produces
-solutions to the neutron diffusion problem nearly two orders of
-magnitude faster than the MCLS library for serial computations using
-the conjugate gradient solver. Therefore, when we directly compare
-these two codes, the MCLS parallel efficiencies will appear
-artificially inflated relative to the Belos efficiencies due to the
-fact that the MCLS ratio of work to communication will be much larger
-than that for the Belos calculations. For every scaling study
+parallel Krylov solvers.
+
+For the scaling results presented here, all parallel Monte Carlo
+results were computed using the Monte Carlo Linear
+Solvers\footnote{MCLS is an open source library and can be downloaded
+  at \url{https://github.com/sslattery/MCLS}} (MCLS) library generated
+specifically for this work while the conjugate gradient and GMRES
+Krylov solvers used are part of the Belos linear solver package within
+the Trilinos framework \citep{heroux_overview_2005}. In general, the
+Belos package produces solutions to the neutron diffusion problem
+nearly two orders of magnitude faster than the MCLS library for serial
+computations using the conjugate gradient solver. Therefore, when we
+directly compare these two codes, the MCLS parallel efficiencies will
+appear artificially inflated relative to the Belos efficiencies due to
+the fact that the MCLS ratio of work to communication will be much
+larger than that for the Belos calculations. For every scaling study
 presented, the calculations were performed 3 times with the average
 values reported for wall time and used for the efficiency
-computations.
+computations. In addition, it should be noted that in practice,
+symmetric positive-definite problems like the neutron diffusion
+problem solved here would never be solved using GMRES. Instead, a
+method such as conjugate gradient would typically be used due to the
+enhanced performance. We use GMRES here to simply as a means of
+comparison to MCSA performance.
 
@@
 components to achieve a theoretical peak performance of more than 20
-petaflops. The technical specifications for Titan are given in
-Table~\ref{tab:titan_hardware} and specifically for only the CPU
-portion of the machine.
+petaflops. The technical specifications for Titan for only the CPU
+portion of the machine are given in Table~\ref{tab:titan_hardware}.
 
@@
 applied to the Neumann-Ulam algorithm. For the strong scaling study,
-the global problem size was fixed at \sn{1.6}{7} DOFs for all solvers
-and the 16 core run used as the base case. MCSA was used again with
-one stochastic history used per DOF in the Neumann-Ulam solve to
-compute the correction for the acceleration with the remainder of the
-problems and solver parameters the same as those for the verification
-calculation given in Table~\ref{tab:parallel_verification_parameters}
-and Table~\ref{tab:parallel_mcsa_parameters}. As with all of the
-scaling studies, 3 calculations for all data points were performed
-with the average wall time reported and used for the efficiency
+the global diffusion problem size was fixed at \sn{1.6}{7} DOFs for
+all solvers and the 16 core run used as the base case. MCSA was used
+again with one stochastic history used per DOF in the Neumann-Ulam
+solve to compute the correction for the acceleration with the
+remainder of the problems and solver parameters the same as those for
+the verification calculation given in
+Table~\ref{tab:parallel_verification_parameters} and
+Table~\ref{tab:parallel_mcsa_parameters}. As with all of the scaling
+studies, 3 calculations for all data points were performed with the
+average wall time reported and used for the efficiency
 calculations. Figure~\ref{fig:titan_pure_strong_time} gives the wall
@@
 is important as we can expect the same qualitative scaling performance
-from MCSA once arithmetic optimization has been performed to improve to
+from MCSA if arithmetic optimization has been performed to improve the
 serial performance of the implementation. In addition, these results
-are valuable in that the show MCSA, when properly parallelized, has
+are valuable in that they show MCSA, when properly parallelized, has
 the ability to be competitive on leadership-class computing platforms
@@
 efficiencies. As the global problem size increases, MCSA is less prone
-to a drop in parallel efficiency. When optimization has been completed
+to a drop in parallel efficiency. If optimization has been performed
 and runtimes potentially approach those of the Krylov methods, this
 comparison may not be as favorable. Overall, weak scaling performance
-the parallel MCSA algorithm is excellent for the pure domain
+of the parallel MCSA algorithm is excellent for the pure domain
 decomposition case with absolute parallel efficiencies of over 80\%
@@
 studies, the linear operator had a spectral radius of 0.787 and a
-weight cutoff of \sn{1}{-2}. Using Eq~(\ref{eq:analytic_k}) gives an
-expected random walk length of 19.22 transitions for any given history
-in the problem. Using the mean-chord approximation given by
+weight cutoff of \sn{1}{-2} was used with the Neumann-Ulam
+solver. Using Eq~(\ref{eq:analytic_k}) gives an expected random walk
+length of 19.22 transitions for any given history in the
+problem. Using the mean-chord approximation given by
 Eq~(\ref{eq:mean_chord_domain_leakage}), we then expect approximately
@@
 fraction of communicated histories and therefore boost the parallel
-performance of the algorithm. To determine the amount of overlap to
-test for the scaling studies, we use a variation of
-Eq~(\ref{eq:step_k_length}) to determine how many discrete states a
-history will move on average from its birth site. We modify
-Eq~(\ref{eq:step_k_length}) to only account for discrete movements:
+performance of the algorithm.
+
+To determine the amount of overlap to test for the scaling studies, we
+use a variation of Eq~(\ref{eq:step_k_length}) to determine how many
+discrete states a history will move on average from its birth site. We
+modify Eq~(\ref{eq:step_k_length}) to only account for discrete
+movements:
 \begin{equation}
@@
 
+The absolute efficiency results from the strong scaling exercise are
+given in Figure~\ref{fig:titan_strong_overlap}. Aside from the
+boosting of efficiencies at 2,048 cores, little evidence of strong
+scaling improvement is evident for all levels of overlap tested. To
+further explore any potential benefits of adding overlap into the
+system, Figure~\ref{fig:titan_strong_overlap_diff} gives the
+difference in the efficiencies between all overlap cases and the
+reference case without overlap at each core count. At smaller numbers
+of cores, there is no observed benefit to using overlap to improve
+strong scaling. The improvements noted at 2,048 cores are also present
+here and account for the best efficiency improvements for all core
+counts tested. at 16,384 cores, using 2 levels of overlap did give a
+6.4\% boost in efficiency while at 1,024 cores this same amount of
+overlap actually reduced the efficiency. Beyond 2 levels of overlap,
+efficiency levels begin to fall with 5 levels of overlap performing
+the worst for all cases. From this, we don't expect adding any more
+overlap will improve the scalability of the algorithm and the best
+strong scaling performance was observed when the number of overlap
+levels is approximately equal to the root-mean-squared number of
+states we expect a history to move from its starting point as computed
+by Eq~(\ref{eq:discrete_distance}). We will defer an explanation of
+the drop in efficiencies for levels of overlap beyond this value until
+we have analyzed the weak scaling data.
+
 \begin{figure}[t!]
@@
 
+For weak scaling, the same study as in the pure domain decomposition
+case was performed again with 1, 2, 3, and 5 levels of overlap. The
+absolute parallel efficiencies from these calculations are given in
+Figure~\ref{fig:titan_weak_overlap} and the differences in
+efficiencies from the calculations without overlap. The results here
+are much the same in that there is not a dramatic improvement in
+scalability relative to the case without overlap. Opposite the strong
+scaling calculation, overlap enhances weak scaling efficiencies at
+lower core counts and degrades the efficiencies at higher core
+counts. In addition, 1 level of overlap provided the best enhancement
+of scalability when overlap was beneficial.
+
 \begin{figure}[t!]
@@
 
+For both the strong and weak scaling studies, the parallel performance
+of the algorithm was either marginally improved or not improved at
+all. We would always expect an improvement in performance for using
+overlap to prevent the transport of histories from domain to domain
+during execution of the Neumann-Ulam algorithm. The reduction in
+tranport means a reduction in parallel communication and therefore
+should never degrade the performance of the algorithm. It is true that
+doing so in fact enhances Monte Carlo performance. However, the
+Neumann-Ulam algorithm is not representative of the entire MCSA
+algorithm. When overlap is used, an overlapping tally vector is
+constructed in parallel such that multiple cores may own multiple
+states in the system. When this occurs, the tally data associated with
+these shared states must be communicated after the Neumann-Ulam solve
+as outlined in \S~\ref{subsec:msod_algorithm}. This reduction
+operation then builds the full MCSA correction vector in the
+decomposition of the solution vector.
+
+Therefore, by building overlap we are in fact reducing parallel
+communication during the Neumann-Ulam solve but we are simply
+deferring that communication until after Monte Carlo transport when
+the overlapping tally vector must be reduced. When overlap becomes
+large enough (i.e. larger than the root-mean-squared number of
+discrete states a history will travel), then the savings in
+communication during the Neumann-Ulam solve start to become smaller
+than the communication created in the overlapping tally vector
+reduction. We see this in both the strong and weak scaling cases at
+larger levels of overlap. For the strong scaling case, we see better
+performance improvement at higher core counts because the total number
+of overlapping states on each core is smaller from the local problem
+size shrinking. For weak scaling, we see better performance
+improvement at lower numbers of cores because although the local
+overlap size is fixed, the global amount of overlap is growing with
+core count and therefore reducing the performance of the tally
+reduction operation.
+
+Based on this data, a few levels of overlap within the expected
+distance of travel for a history may offer some marginal improvement
+in scalability, however, it is not as effective as one might think
+given its improvement to reactor physics calculations as observed in
+the literature. This is largely due to the density of states in the
+system that creates the described performance bottlenecks in the
+overlapping tally vector reduction operation. In a Monte Carlo
+particle simulation, these overlapping tally reduction operations are
+not expected to be peformed at the speed they are required in
+MCSA. For example, in a criticality calculation with overlapping
+domains, we would expect to perform as many parallel tally vector
+reduction in a particle transport simulation as there are particle
+stages required for convergence of the eigenvalue. However, these
+particle stages will likely be much more costly in terms of wall time
+relative to a Neumann-Ulam solve in an MCSA iteration due to the more
+extensive particle physics and logic structure that must be
+evaluated. Therefore, we can expect this overlapping tally reduction
+to be less signficant to scalability in those operations. We do not
+have as complicated a logic structure in a history random walk for the
+Neumann-Ulam method and therefore overlap can inhibit scalability if
+too large.
+
 \clearpage
@@
 We next look to apply replication to the problem in the form of
-multiple sets. Before performing parallel scaling studies using
+multiple sets. As the previous analysis showed overlap to be of little
+assistance in improving scalability, we only apply multiple sets to
+the problem with pure domain decomposition within the sets for these
+scaling studies. Before performing parallel scaling studies using
 multiple sets, several aspects of the replicated parallel problem must
@@
 this work we consider two choices; the split case and the fully
-replicated case. As an example these cases, consider a single set
+replicated case. As an example of these cases, consider a single set
 problem where 20,000 histories are used in the Monte Carlo solve. If
@@
 
-Also of interest here is the effect of splitting histories among
-sets versus replicating them on the iterative performance of the
+Also of interest here is the effect of splitting histories among sets
+versus replicating them on the iterative performance of the
 method. Table~\ref{tab:ms_strong_alg_eff} gives the iterations
@@
 decreased the number of iterations required to converge in a nonlinear
-fashion. Using these algorithmic efficiencies, the implementation
-efficiencies for the multiple set cases were computed and are given in
+fashion. Therefore, increasing the number of histories per MCSA
+iteration improves algorithmic efficiency. Using these algorithmic
+efficiencies, the implementation efficiencies for the multiple set
+cases were computed and are given in
 Figure~\ref{fig:titan_strong_ms_impeff}. From these results, we see
 that the replicated history cases are even less efficient than in the
-absolute case. Not only do we have the additional parallel overhead of
+absolute case with respect to the performance of a single
+iteration. Not only do we have the additional parallel overhead of
 doing the block level reduction, but we also have additional histories
@@
 not observed in the strong scaling case until much larger core counts
-due as the multiple set computations hit the strong scaling wall after
-the single set computation.
+as the multiple set computations hit the strong scaling wall after the
+single set computation.
 
@@
 Figure~\ref{fig:titan_weak_ms_eff} gives the results of the weak
-scaling efficiencies using the multiple set metric given by
+scaling efficiency computations using the multiple set metric given by
 Eq~(\ref{eq:ms_weak_efficiency}). Although we do reduce the time to
@@
 expect to seriously boost parallel efficiencies for MCSA using this
-technique accept for certain instances past the strong scaling wall in
+technique except for certain instances past the strong scaling wall in
 a purely strong scaling environment. In a weak scaling environment,
@@
 Based on the analysis of the pure domain decomposed MCSA algorithm and
-subsequently the MSOD implementation, one may ask then ask if there is
-any need to communicate histories from domain-to-domain at
-all. Overlap was not as successful because so few histories needed to
-be communicated in the first place that keeping them on-process
-resulted in more work than the communications savings that
-occurred. If that is true, then perhaps without any overlap or
-communication, just doing local Monte Carlo may be satisfactory for
-accelerating the MCSA iteration. In this case, histories near the
-boundaries of the domain may be truncated before reaching the weight
-cutoff if the reach the domain boundary first. This truncation has the
-potential to increase the stochastic uncertainty of the Neumann-Ulam
-result and therefore increase the number of MCSA iterations required
-to converge. In addition, the scalability of the method will then be
-purely limited by the parallel matrix and vector operations required
-to implement the remainder of the algorithm.
+subsequently the multiple sets and overlapping domain implementation,
+one may ask then ask if there is any need to communicate histories
+from domain-to-domain at all. Overlap was not as successful because
+because the additional communication required to reduce the
+overlapping tally vector eventually becomes larger than the
+communication savings during Monte Carlo transport. In addition, for
+local domains that are large enough to get good performance
+(e.g. those used in the weak scaling studies), then our analytic
+relationships for domain leakage, and in particular
+Eq~(\ref{eq:mean_chord_domain_leakage}), tell us that only about 5\%
+of the histories in this diffusion problem will leave the domain in
+the first place.  If that is true, then perhaps without any overlap or
+communication, just doing local Monte Carlo transport such that
+histories are terminated upon reaching the domain boundary may be
+satisfactory for accelerating the MCSA iteration. In this case,
+histories near the boundaries of the domain may be truncated before
+reaching the weight cutoff if they reach the domain boundary
+first. This truncation has the potential to increase the stochastic
+uncertainty of the Neumann-Ulam result and therefore increase the
+number of MCSA iterations required to converge. In addition, the
+scalability of the method will then be purely limited by the parallel
+matrix and vector operations required to implement the remainder of
+the algorithm on the subset of processors containing the linear
+problem.
 
@@
 domain decomposition methods we see many similarities including the
-potential to overlap the subdomains to improve the results and the
-approximate subdomain solves effectively through a relaxation
-scheme. By building the boundary data as outlined in
+potential to overlap the subdomains to improve the results. By
+building the boundary data as outlined in
 \S~\ref{subsec:domain_generation}, we are effectively gathering one
@@
 casts MCSA as a stochastic realization of an additive Schwarz method
-with a fixed point iterations used as a relaxation step\footnote{See
+with a fixed point iteration used as a relaxation step\footnote{See
   section 14.3.3 in \citep{saad_iterative_2003} for a full definition
@@
 rise above unity in efficiencies because there is less communication
-overhead hiding the better cache utilization as the local problem
+overhead hiding the better cache utilization as the local problem size
 shrinks. At 16,384 cores, performing the subdomain solves boost the
@@
 vector. For both the weak and strong scaling study, all calculations
-converged in 22 or in only a few cases 21 iterations for both the
-subdomain solve and the solve with communication giving an algorithmic
-efficiency of 1.048 for all cases. This small improvement in iteration
-count was not observed to be a function of problem size, number of
-processors, or whether the subdomain method was used. Therefore, there
-is no reason to expect overlap to improve the iterative or scaling
-performance of the method as iterative performance is maintained
-without it. Overlap is not as effective because histories relinquish
-most of their useful information within the first few steps (which is
-why we can use such a large weight cutoff and there is an
-insensitivity to weight cutoff as shown in
+with split histories converged in 22 or in only a few cases 21
+iterations for both the subdomain solve and the solve with
+communication giving an algorithmic efficiency of 1.048 for all
+cases. This small improvement in iteration count was not observed to
+be a function of problem size, number of processors, or whether the
+subdomain method was used. For the replicated history cases, the
+algorithmic efficiency improvement was not as good as when
+communication was allowed for the Monte Carlo algorithm. For these
+calculations, the number of iterations to converge was reduced to 18
+or 19 in both the strong and weak scaling studies with lower core
+counts converging in fewer iterations.
+
+Therefore, there is no reason to expect overlap to improve the
+iterative or scaling performance of the method as iterative
+performance is maintained without it for this particular
+problem. Overlap is not as effective because histories relinquish most
+of their useful information within the first few steps (which is why
+we can use such a large weight cutoff and there is an insensitivity to
+weight cutoff as shown in
 Figure~\ref{fig:estimator_wc_iters}). Keeping them around longer with
 overlap to do extra work in the local domain will not improve
-iterative performance and thus increasing compute times.
-
-\clearpage
-
-\subsubsection{Strong Scaling with Multiple Sets}
-\label{subsubsec:subdomain_ms_strong}
-
-\clearpage
-
-\subsubsection{Weak Scaling with Multiple Sets}
-\label{subsubsec:subdomain_ms_weak}
+iterative performance and thus increase compute times.
+
+As a final analysis of MCSA using subdomain Monte Carlo, the multiple
+sets scaling studies are repeated to determine if any performance
+benefits are made from replication by eliminating Monte Carlo
+communication. Figure~\ref{fig:titan_strong_subdomain_ms_time} gives
+the runtimes for the strong scaling calculation and
+Figure~\ref{fig:titan_strong_subdomain_ms} the computed absolute
+efficiencies. For the weak scaling studies,
+Figure~\ref{fig:titan_weak_subdomain_ms_time} gives the wall times and
+Figure~\ref{fig:titan_weak_subdomain_ms} the efficiencies. For both
+scaling studies, the results are effectively the same as for the case
+with Monte Carlo communication. This is again due to the fact that
+adding sets to the problem is not only an exercise in strong scaling
+but it also adds additional parallel overhead in the form of the tally
+vector reduction across blocks.
+
+From a parallel performance perspective, MCSA may then be best used in
+a mode where the Monte Carlo is performed over only the subdomains
+with no replication or overlap added to the system. In this mode,
+iterative perforamance is maintained relative to the solutions with
+communication even for more ill-conditioned problems with larger
+spectral radii. Furthermore, no overlap was required to maintain
+iterative performance be letting histories near the boundary take a
+few more steps in their random walks. If in the future there are
+ill-conditioned problems where using subdomain Monte Carlo reduces the
+algorithmic efficiency of MCSA due to the information lost in the
+correction vector, overlap would be a viable mechanism to potentially
+maintain that iterative performance, keeping in mind the potential
+impacts on scalability that were observed in this work.
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_subdomain_ms_time.pdf}
+  \end{center}
+  \caption{\textbf{Strong scaling total wall time with multiple sets
+      and subdomain Monte Carlo.} \textit{The subdomain Monte Carlo
+      solver improves the parallel efficiency and therefore time to
+      solution by eliminating all parallel communication during the
+      Monte Carlo solve.}}
+  \label{fig:titan_strong_subdomain_ms_time}
+\end{figure}
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_subdomain_ms.pdf}
+  \end{center}
+  \caption{\textbf{Strong scaling absolute efficiency for with
+      multiple sets and subdomain Monte Carlo.} \textit{The subdomain
+      Monte Carlo solver improves the parallel efficiency and
+      therefore time to solution by eliminating all parallel
+      communication during the Monte Carlo solve.}}
+  \label{fig:titan_strong_subdomain_ms}
+\end{figure}
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_subdomain_ms_time.pdf}
+  \end{center}
+  \caption{\textbf{Weak scaling total wall time with multiple sets
+      and subdomain Monte Carlo.} \textit{The subdomain Monte Carlo
+      solver improves the parallel efficiency and therefore time to
+      solution by eliminating all parallel communication during the
+      Monte Carlo solve.}}
+  \label{fig:titan_weak_subdomain_ms_time}
+\end{figure}
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_subdomain_ms.pdf}
+  \end{center}
+  \caption{\textbf{Weak scaling absolute efficiency for with
+      multiple sets and subdomain Monte Carlo.} \textit{The subdomain
+      Monte Carlo solver improves the parallel efficiency and
+      therefore time to solution by eliminating all parallel
+      communication during the Monte Carlo solve.}}
+  \label{fig:titan_weak_subdomain_ms}
+\end{figure}

### 2013-07-07 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
spell check on parallel chapter

@@
 Neumann-Ulam method, we derive an analytic framework based on
-algebraic quantities from which to estimate its performance when
-applied to neutron transport systems. We then devise a parallel
+algebraic quantities from which to estimate aspects of its performance
+when applied to neutron transport systems. We then devise a parallel
 algorithm for the Neumann-Ulam method based on the multiple-set
@@
 for the MCSA iteration that leverages the parallel Monte Carlo
-algorithm and general parallel matrix-vector operations. In identical
-fashion to the serial computations presented in
-Chapter~\ref{ch:spn_equations}, we use parallel solutions for the
-$SP_N$ equations for the fuel assembly criticality problem to verify
-the correctness of the parallel algorithm and implementation by
-comparing the numerical results against production Krylov
-methods. Using the verified implementation of the algorithm, we
-perform parallel scaling studies to test its performance on a
-leadership class machine and compare this performance to the
-production Krylov methods.
+algorithm and general parallel matrix-vector operations. Using
+parallel solutions for the neutron diffusion problem, we verify the
+correctness of the parallel algorithm and implementation by comparing
+the numerical results against production Krylov methods. With the
+verified implementation of the algorithm, we perform parallel scaling
+studies to test its performance on a leadership class machine and
+compare this performance to the production Krylov methods.
 
@@
 in the system will be compared to the solution at the same grid point
-in the serial reference case with the minimum and maxiumum absolute
+in the serial reference case with the minimum and maximum absolute
 differences of these values reported.
@@
 results and a larger minimum absolute difference than all MCSA results
-by three orders of magnitude. Based on these results, MCSA parallized
+by three orders of magnitude. Based on these results, MCSA parallelized
 with an MSOD, and in particular the implementation used here,
@@
 will use the overlap generation algorithm outlined in
-\S~\ref{subsec:msod_generation} such that we parameterically grow the
+\S~\ref{subsec:msod_generation} such that we parametrically grow the
 local domain using 1, 2, 3, and 5 levels to measure how scalability is
@@
 during execution of the Neumann-Ulam algorithm. The reduction in
-tranport means a reduction in parallel communication and therefore
+transport means a reduction in parallel communication and therefore
 should never degrade the performance of the algorithm. It is true that
@@
 particle simulation, these overlapping tally reduction operations are
-not expected to be peformed at the speed they are required in
+not expected to be performed at the speed they are required in
 MCSA. For example, in a criticality calculation with overlapping
@@
 evaluated. Therefore, we can expect this overlapping tally reduction
-to be less signficant to scalability in those operations. We do not
+to be less significant to scalability in those operations. We do not
 have as complicated a logic structure in a history random walk for the
@@
 with no replication or overlap added to the system. In this mode,
-iterative perforamance is maintained relative to the solutions with
+iterative performance is maintained relative to the solutions with
 communication even for more ill-conditioned problems with larger

### 2013-07-07 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
finished parallel chapter draft

@@
       illustrating how domain decomposition requires parallel
-      communication of particles.}}
+      communication of particles.} \textit{If a particle crosses a
+      domain boundary it must be communicated to the next domain in
+      its path.}}
   \label{fig:ddmc_example}
@@
 Monte Carlo methods and traditional fixed source Monte Carlo transport
-problems is the content of the Markov chains that are generated. The
-transitions represented by these chains are bound by probabilities and
-weights and are initiated by the sampling of a discrete source. In the
-context of transport problems, those transitions represent events such
-as particle scattering and absorption with probabilities that are
-determined by physical data in the form of cross sections.
-
-For stochastic matrix inversion, those transitions represent moving
+problems is the content of the Markov chains that are generated.
+
+The transitions represented by these chains are bound by probabilities
+and weights and are initiated by the sampling of a discrete source. In
+the context of transport problems, those transitions represent events
+such as particle scattering and absorption with probabilities that are
+determined by physical data in the form of cross sections. For
+stochastic matrix inversion, those transitions represent moving
 between the equations of the linear system (and therefore the
@@
 
-In this section Brunner and Brantley's fully asynchronous algorithm is
-presented along with its application to the Neumann-Ulam method. For
+In this section Brunner and Brantley's fully asynchronous algorithm,
+which was effectively implemented verbatim for this work, is presented
+along with its application to the Neumann-Ulam method. For
 considerably more detail, the algorithm presented can be found in
-\citep{brunner_efficient_2009} and was effectively implemented
-verbatim for this work. In their work they identify two data sets that
-are required to be communicated: the sharing of particles that are
-transported from one domain to another and therefore from one
+\citep{brunner_efficient_2009}. In their work they identify two data
+sets that are required to be communicated: the sharing of particles
+that are transported from one domain to another and therefore from one
 processor to another and a global communication that signals if
@@
   \caption{\textbf{Neutron diffusion equation state adjacency graph.}
-    \textit{Adjacent boundary states are collected by traversing the
-      graph one step for all local boundary states and collecting the
-      equations for those states that are not local. For a given
+    \textit{The structure of the graph comes from the discretization
+      of the Laplacian operator describing the diffusion
+      physics. Adjacent boundary states are collected by traversing
+      the graph one step for all local boundary states and collecting
+      the equations for those states that are not local. For a given
       transition in the random walk for the diffusion problem, a
@@
 
-\subsection{The Algorithm}
+\subsection{Domain Decomposed Algorithm}
 \label{subsec:parallel_mc_algorithm}
@@
 required for domain-to-domain communication. Again, consider this
-communication patter for the 9 subdomain example given by
-Figure~\ref{fig:nearest_neighbor_comm}. In this pattern, each domain
-has two neighbors with which they will communicate parallel histories
-and this communication goes both ways as represented by the adjacent
-arrows in the figure. For each set of neighbors, a non-blocking send
-and receive is required with data buffers allocated with a
-user-defined size prepared for incoming and outgoing histories. This
-nonblocking structure is critical to the performance of the algorithm
-in that it permits local history transport to receive while new work
-to do is being collected. When a given process is ready to do more
-work, it can check these data buffers for incoming histories requiring
-further transport. In this way there is a maximum amount of overlap
-between communication and transport of histories.
+communication pattern for the 9 subdomain example given by
+Figure~\ref{fig:nearest_neighbor_comm}. In this pattern, each boundary
+domain has two neighbors and the center domain four neighbors with
+which they will communicate parallel histories and this communication
+goes both ways as represented by the adjacent arrows in the
+figure. For each set of neighbors, a non-blocking send and receive is
+required with data buffers allocated with a user-defined size prepared
+for incoming and outgoing histories. This nonblocking structure is
+critical to the performance of the algorithm in that it permits local
+history transport to continue while new histories to transport are
+being collected in the buffers. When a given process is ready to do
+more work, it can check these data buffers for incoming histories
+requiring further transport. In this way there is a maximum amount of
+overlap between communication and transport of histories.
 
@@
 To generate overlap, we use the identical algorithm but in this case
-we perform as many graph traversals as the specified amount of
+we perform as many graph level traversals as the specified amount of
 overlap. For example, consider the analytic relations derived in
@@
 considering a neutron diffusion problem. From these relations we might
-determine that for our particular problem, if we can grow the domain
+determine that, for our particular problem, if we can grow the domain
 by 10 additional discrete state transitions, we can reduce the number
@@
 build the overlap by traversing the graph of the matrix outward from
-the boundary 10 times. At each step, we find the data in the adjacent
+the boundary 10 levels. At each step, we find the data in the adjacent
 domain that we require and make that data local, incrementing the size
@@
 traversal on the boundary states is performed to get the new
-neighboring states and their owning process.
+neighboring states and their owning processes.
 
@@
 
-Once complete, the tallies for each individual computation must be
-gathered to the original set in order to build the correction vector
-within an MCSA sequence. To do this, we perform two parallel vector
-operations. First, overlap not only creates additional states in the
-local transport domain but also additional states in the local tally
-vector such that any events that occur in the local domain may be
-tallied in a local vector. Because of this, the tally vector in a
-single set will share local states with other domains and a parallel
-sum reduction operation is required to transform the tally into the
-original Neumann-Ulam parallel decomposition. Second, the tallies in
-individual sets must be combined and applied to the primary set in the
-problem that shares the processor space with the original linear
-problem. For this operation, we employ the concept of blocks as a
-subdivision of parallel space complementary to the concept of sets.
+Once Monte Carlo transport is complete complete, the tallies for each
+individual computation must be gathered to the original set in order
+to build the correction vector within an MCSA sequence. To do this, we
+perform two parallel vector operations. First, overlap not only
+creates additional states in the local transport domain but also
+additional states in the local tally vector such that any events that
+occur in the local domain may be tallied in a local vector. Because of
+this, the tally vector in a single set will share local states with
+other domains and a parallel sum reduction operation is required to
+transform the tally into the original Neumann-Ulam parallel
+decomposition. Second, the tallies in individual sets must be combined
+and applied to the primary set in the problem that shares the
+processor space with the original linear problem. For this operation,
+we employ the concept of blocks as a subdivision of parallel space
+complementary to the concept of sets.
 
@@
 operation, the tally vectors in each set must be summed together and
-applied to the primary set. As each set is an exact replica off the
+applied to the primary set. As each set is an exact replica of the
 others, the parallel decomposition of the tally vector is the same
@@
 set creates a single block, giving 9 total blocks for this example
-problem. The processor space the encapsulates each of the upper-left
+problem. The processor space that encapsulates each of the upper-left
 domains is then used for the parallel reduction operation. This
@@
 diffusion problem in parallel using multiple cores on a desktop
-machine. For each parallel solution, the solution at each grid point
-in the system will be compared to the solution at the same grid point
-in the serial reference case with the minimum and maximum absolute
-differences of these values reported.
+machine. For each parallel solution, then the solution at each grid
+point in the system will be compared to the solution at the same grid
+point in the serial reference case with the minimum and maximum
+absolute differences of these values reported.
 
@@
 multiple sets and multiple blocks may be tested along with
-overlap. When multiple sets were leveraged, a single history was used
-to compute the MCSA correction for every DOF in the problem. For all
+overlap. For all MCSA calculations, a single history was used for
+every DOF in the problem to compute the correction vector. For all
 solvers, the flux was converged to a tolerance of \sn{1}{-8} and
@@
 From the data we see that for all solvers, the parallel results agree
-within the solution tolerance of the reference serial results for all
-DOF in the system. In addition, we do not notice any systematic
+with the reference serial results within the solution tolerance for
+all DOF in the system. In addition, we do not notice any systematic
 differences in the results as a function of number of blocks, number
@@
 results and a larger minimum absolute difference than all MCSA results
-by three orders of magnitude. Based on these results, MCSA parallelized
-with an MSOD, and in particular the implementation used here,
-generates the correct results with respect to the serial conjugate
-gradient reference computation.
+by three orders of magnitude. Based on these results, MCSA
+parallelized with the MSOD algorithm, and in particular the
+implementation used here, generates the correct results with respect
+to the serial conjugate gradient reference computation.
 
@@
 determined by several performance measures. For this work, these
-measures come directly from Keyes' primer on the scalability of domain
-decomposed methods and specifically his discussion on how to measure
-their parallel performance \citep{keyes_how_1999}. All of these
-metrics are derived from two measured quantities: the number of
+measures will come directly from Keyes' primer on the scalability of
+domain decomposed methods and specifically his discussion on how to
+measure their parallel performance \citep{keyes_how_1999}. All of
+these metrics are derived from two measured quantities: the number of
 iterations required for convergence and the total wall time required
@@
 scalability with different definitions depending on the type of
-parametric scaling study performed. For all efficiencies, a value of 1
-is considered perfect while a value of 0 signals the worst possible
-performance. Often, as was observed in this work, it is the case that
-efficiencies above 1 are computed in certain scaling studies. For the
-data presented in the following sections, an explanation of each
-observed case of super-unitary efficiencies will be provided.
+parametric scaling study performed. For all efficiencies, a value of
+100\% is considered perfect while a value of 0\% signals the worst
+possible performance. Often, as was observed in this work, it is the
+case that efficiencies above 100\% are computed in certain scaling
+studies. For the data presented in the following sections, an
+explanation of each observed case of super-unitary efficiencies will
+be provided.
 
@@
 subject to the constraint $M/Q = N/P$. The scaling is perfect if the
-runtime is static for all $P$ when $N/P$ is fixed. Often, weak scaling
-is of interest to the nuclear engineer due to the fact that the
-computational resources available are fixed and may not be able to
-contain the entire global problem of interest. In this case, an
-algorithm with good weak scaling performance permits all of the
-computational resources available to be used effectively to allow for
-the highest fidelity problem possible to be solved. In addition, for
-the resources available, the memory to processor ratio on a
+runtime is static for all $P$ when $N/P$ is fixed.
+
+Often, weak scaling is of interest to the nuclear engineer due to the
+fact that the computational resources available are fixed and may not
+be able to contain the entire global problem of interest. In this
+case, an algorithm with good weak scaling performance permits all of
+the computational resources available to be used effectively to allow
+for the highest fidelity problem possible to be solved. In addition,
+for the resources available, the memory to processor ratio on a
 distributed machine is often fixed, giving an exercise in fixed $N/P$
@@
 
-For the scaling results presented here, all parallel Monte Carlo
-results were computed using the Monte Carlo Linear
-Solvers\footnote{MCLS is an open source library and can be downloaded
-  at \url{https://github.com/sslattery/MCLS}} (MCLS) library generated
-specifically for this work while the conjugate gradient and GMRES
-Krylov solvers used are part of the Belos linear solver package within
-the Trilinos framework \citep{heroux_overview_2005}. In general, the
-Belos package produces solutions to the neutron diffusion problem
-nearly two orders of magnitude faster than the MCLS library for serial
-computations using the conjugate gradient solver. Therefore, when we
-directly compare these two codes, the MCLS parallel efficiencies will
-appear artificially inflated relative to the Belos efficiencies due to
-the fact that the MCLS ratio of work to communication will be much
-larger than that for the Belos calculations. For every scaling study
+For the scaling results presented here, all parallel MCSA results were
+computed using the Monte Carlo Linear Solvers\footnote{MCLS is an open
+  source C++ library built on the Trilinos framework and can be
+  downloaded at \url{https://github.com/sslattery/MCLS}} (MCLS)
+library generated specifically for this work while the conjugate
+gradient and GMRES Krylov solvers used are part of the Belos linear
+solver package within the Trilinos framework
+\citep{heroux_overview_2005}. In general, the Belos package produces
+solutions to the neutron diffusion problem nearly two orders of
+magnitude faster than the MCLS library for serial computations using
+the conjugate gradient solver. Therefore, when we directly compare
+these two codes, the MCLS parallel efficiencies will appear
+artificially inflated relative to the Belos efficiencies due to the
+fact that the MCLS ratio of work to communication will be much larger
+than that for the Belos calculations. For every scaling study
 presented, the calculations were performed 3 times with the average
-values reported for wall time and used for the efficiency
-computations. In addition, it should be noted that in practice,
-symmetric positive-definite problems like the neutron diffusion
-problem solved here would never be solved using GMRES. Instead, a
-method such as conjugate gradient would typically be used due to the
-enhanced performance. We use GMRES here to simply as a means of
-comparison to MCSA performance.
+values reported for wall time and those average values used for the
+efficiency computations. In addition, it should be noted that in
+practice, symmetric positive-definite problems like the neutron
+diffusion problem solved here would never be solved using
+GMRES. Instead, a method such as conjugate gradient would typically be
+used due to the enhanced performance. We use GMRES here simply as a
+means of providing an additional comparison for MCSA performance.
 
@@
 utilized. This is due to the fact that the implementations of the
-parallel MCSA algorithm utilized code written only with the Message
-Passing Interface (MPI) and therefore could not take advantage of the
-GPU accelerators. Even considering just the CPU portion of Titan,
-nearly 300,000 CPUs are available to test the parallel performance of
-the MCSA algorithm at high levels of concurrency.
+parallel MCSA algorithm in the MCLS library utilized code written only
+with the Message Passing Interface (MPI) and therefore could not take
+advantage of the GPU accelerators. Even considering just the CPU
+portion of Titan, nearly 300,000 CPUs are available to test the
+parallel performance of the MCSA algorithm at high levels of
+concurrency.
 
@@
   \caption{\textbf{Strong scaling algorithmic efficiency using
-      multiple sets.}}
+      multiple sets.} \textit{Replicating histories improves
+      algorithmic efficiency by converging in fewer iterations.}}
   \label{tab:ms_strong_alg_eff}
@@
   \caption{\textbf{Weak scaling algorithmic efficiency using multiple
-      sets.}}
+      sets.} \textit{Replicating histories improves algorithmic
+      efficiency by converging in fewer iterations.}}
   \label{tab:ms_weak_alg_eff}

### 2013-07-08 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
updating makefile for nonlinear figures

@@
 
+\begin{algorithm}[h!]
+  \caption{FANM Algorithm}
+  \label{alg:fanm}
+  \begin{algorithmic}[1]
+    \State $k := 0$ 
+    \While{$||\ve{F}(\ve{u}^{k})|| > \epsilon
+      ||\ve{F}(\ve{u}^{0})||$} 
+    \State $\ve{J}(\ve{u}^{k}) \leftarrow AD(\ve{F}(\ve{u}^k))$ 
+    \Comment{Automatic differentiation} 
+    \State $\ve{J}(\ve{u}^k) \delta \ve{u}^k = -\ve{F}(\ve{u}^{k})$
+    \Comment{Solve for the Newton correction with MCSA} 
+    \State $\ve{u}^{k+1} \leftarrow \ve{u}^k + \delta \ve{u}^k$ 
+    \State $k \leftarrow k+1$ 
+    \EndWhile
+  \end{algorithmic}
+\end{algorithm}
+
 \subsection{Parallel FANM Implementation}
@@
     \scalebox{1.5}{
-      \input{chapters/navier_stokes_solutions/natural_convection_cavity.pdftex_t} }
+      \input{chapters/nonlinear_problem/natural_convection_cavity.pdftex_t} }
   \end{center}
@@
     \scalebox{1.5}{
-      \input{chapters/navier_stokes_solutions/lid_driven_cavity.pdftex_t} }
+      \input{chapters/nonlinear_problem/lid_driven_cavity.pdftex_t} }
   \end{center}
@@
     \scalebox{1.3}{
-      \input{chapters/navier_stokes_solutions/backward_facing_step.pdftex_t} }
+      \input{chapters/nonlinear_problem/backward_facing_step.pdftex_t} }
   \end{center}

### 2013-07-08 doc/parallel_verficiation/compare_flux.cpp
adding flux io to neutron diffusion

@@ -1,51 +1,0 @@
-#include <iostream>
-#include <iomanip>
-#include <vector>
-#include <cmath>
-#include <cstdlib>
-#include <fstream>
-#include <algorithm>
-#include <string>
-#include <cassert>
-
-int main( int argc, char * argv[] )
-{
-    std::string file_1 = "cg_reference.dat";
-    std::string file_2 = "cg.dat";
-
-    int global_length;
-
-    std::ifstream data_file_1;
-    data_file_1.open( file_1.c_str() );
-    data_file_1 >> global_length;
-    std::vector<double> global_vector_1( global_length );
-    for ( int i = 0; i < global_length; ++i )
-    {
-        data_file_1 >> global_vector_1[i];
-    }
-    data_file_1.close();
-
-    std::ifstream data_file_2;
-    data_file_2.open( file_2.c_str() );
-    data_file_2 >> global_length;
-    std::vector<double> global_vector_2( global_length );
-    for ( int i = 0; i < global_length; ++i )
-    {
-        data_file_2 >> global_vector_2[i];
-    }
-    data_file_2.close();
-
-    std::vector<double> difference( global_length );
-    for ( int i = 0; i < global_length; ++i )
-    {
-        difference[i] = std::abs( global_vector_1[i] - global_vector_2[i] );
-    }
-
-    std::cout << "Max Difference: " 
-              << *std::max_element( difference.begin(), difference.end() )
-              << std::endl;
-
-    std::cout << "Min Difference: " 
-              << *std::min_element( difference.begin(), difference.end() )
-              << std::endl;
-}

### 2013-07-08 drivers/NeutronDiffusion/neutron_diffusion.cpp

@@ -7,6 +7,7 @@
 //---------------------------------------------------------------------------//
 
 #include <iostream>
+#include <iomanip>
 #include <vector>
 #include <cmath>
 #include <cstdlib>
@@ -212,6 +213,44 @@ int main( int argc, char * argv[] )
 //                          "NEUTRON_FLUX" );
 //     vtk_output.write();
 
+    int global_length = 0;
+    Teuchos::ArrayRCP<double> global_vector;
+    if ( 0 == set_id )
+    {
+        // Collapse the flux to a single vector.
+        Teuchos::RCP<Vector> lhs = diffusion_problem->getProblem()->getLHS();
+        Teuchos::ArrayRCP<const double> data = lhs->getData();
+        global_length = lhs->getGlobalLength();
+        Teuchos::ArrayRCP<double> local_vector( global_length, 0.0 );
+        global_vector = Teuchos::ArrayRCP<double>( global_length, 0.0 );
+        for ( int i = 0; i < global_length; ++i )
+        {
+            if ( lhs->getMap()->isNodeGlobalElement(i) )
+            {
+                local_vector[i] = data[ lhs->getMap()->getLocalElement(i) ];
+            }
+        }
+
+        Teuchos::reduceAll<int,double>( *set_comm,
+                                        Teuchos::REDUCE_SUM,
+                                        global_length,
+                                        local_vector.getRawPtr(),
+                                        global_vector.getRawPtr() );
+    }
+    comm->barrier();
+    // Write the flux to a file.
+    if ( comm->getRank() == 0 )
+    {
+        std::ofstream ofile;
+        ofile.open( "neutron.dat" );
+        ofile << global_length << std::endl;
+        for ( int i = 0; i < global_length; ++i )
+        {
+            ofile << std::setprecision(18) << global_vector[i] << std::endl;
+        }
+        ofile.close();
+    }
+    comm->barrier();
     return 0;
 }
 

### 2013-07-08 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
adding nonlinear benchmarks data

@@
 where $\rho$ is the fluid density, $\ve{u}$ is the fluid velocity,
-$\ve{g}$ gravity, $C_p$ the specific heat capacity at constant
-pressure of the fluid and $T$ the temperature of the
-fluid. Eq~(\ref{eq:ns_momentum}) provides momentum transport,
-Eq~(\ref{eq:ns_continuity}) provides the mass balance, and
-Eq~(\ref{eq:ns_energy}) provides energy transport with viscous
-dissipation effects neglected. In addition, we close the system with
-the following equations:
+$C_p$ the specific heat capacity at constant pressure of the fluid,
+$T$ the temperature of the fluid, and $\ve{g}$ gravity resulting from
+the Boussinesq approxmiation. Eq~(\ref{eq:ns_momentum}) provides
+momentum transport, Eq~(\ref{eq:ns_continuity}) provides the mass
+balance, and Eq~(\ref{eq:ns_energy}) provides energy transport with
+viscous dissipation effects neglected. In addition, we close the
+system with the following equations:
 \begin{subequations}

### 2013-07-09 doc/dissertation/document/backmatter/nonlinear_background.tex
updating makefile for nonlinear figures
updating nonlinear section

@@
 
-\section{Inexact Newton Methods}
-\label{sec:newton_methods}
-Inexact Newton methods arise when the Jacobian operator is not exactly
-inverted, resulting in an inexact Newton correction as initially
-described by Dembo and others \citep{dembo_inexact_1982}. For common
-sparse nonlinear systems, which in turn yield a sparse Jacobian
-matrix, this situation occurs when conventional iterative methods are
-applied. In their definition, Dembo formulated inexact methods such
-that they are independent of the linear method used to solve for the
-Newton correction and therefore are amenable to use with any linear
-solver. Furthermore, they bind the convergence of the outer nonlinear
-iteration to the inner linear iteration such that:
-\begin{equation}
-  ||\ve{J}(\ve{u}^k)\delta \ve{u}^k + \ve{F}(\ve{u}^k)|| \leq \eta^k
-  ||\ve{F}(\ve{u}^k)||\:,
-  \label{eq:inexact_newton_forcing}
-\end{equation}
-where $\eta^k \in [0,1)$ is defined as the \textit{forcing term} at
-  the $k$ iterate. Eq~(\ref{eq:inexact_newton_forcing}) then states
-  that the residual generated by the linear solver is bound by the
-  nonlinear residual and how tightly it is bound is defined by the
-  forcing term. This is useful in that we can vary how tightly coupled
-  the convergence of the linear iterations used to generate the Newton
-  correction is to the nonlinear iteration by relaxing or tightening
-  the convergence properties on the linear iterative method. As a
-  result, strategies for determining the forcing term can vary
-  depending on the problem type and can greatly affect the convergence
-  of the method or even prohibit convergence
-  \citep{eisenstat_choosing_1996}. In addition, \textit{globalization
-    methods} may be used to modify the Newton correction in a more
-  desire able direction such that convergence properties can be
-  improved when the initial guess for $\ve{u}$ is poor
-  \citep{pawlowski_globalization_2006}.
-
-\section{Newton-Krylov Methods}
+\section{Newton-Krylov Methods\ }
 \label{sec:newton_krylov_methods}

### 2013-07-09 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
updating nonlinear section

--- 2013-07-08 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
@@
 
+\subsection{Inexact Newton Methods}
+\label{subsec:inexact_newton_methods}
+Inexact Newton methods arise when the Jacobian operator is not exactly
+inverted, resulting in an inexact Newton correction as initially
+described by Dembo and others \citep{dembo_inexact_1982}. For common
+sparse nonlinear systems, which in turn yield a sparse Jacobian
+matrix, this situation occurs when conventional iterative methods are
+applied. In their definition, Dembo formulated inexact methods such
+that they are independent of the linear method used to solve for the
+Newton correction and therefore are amenable to use with any linear
+solver. Furthermore, they bind the convergence of the outer nonlinear
+iteration to the inner linear iteration such that:
+\begin{equation}
+  ||\ve{J}(\ve{u}^k)\delta \ve{u}^k + \ve{F}(\ve{u}^k)|| \leq \eta^k
+  ||\ve{F}(\ve{u}^k)||\:,
+  \label{eq:inexact_newton_forcing}
+\end{equation}
+where $\eta^k \in [0,1)$ is defined as the \textit{forcing term} at
+  the $k$ iterate. Eq~(\ref{eq:inexact_newton_forcing}) then states
+  that the residual generated by the linear solver is bound by the
+  nonlinear residual and how tightly it is bound is defined by the
+  forcing term. This is useful in that we can vary how tightly coupled
+  the convergence of the linear iterations used to generate the Newton
+  correction is to the nonlinear iteration by relaxing or tightening
+  the convergence properties on the linear iterative method. As a
+  result, strategies for determining the forcing term can vary
+  depending on the problem type and can greatly affect the convergence
+  of the method or even prohibit convergence
+  \citep{eisenstat_choosing_1996}. In addition, \textit{globalization
+    methods} may be used to modify the Newton correction in a more
+  desireable direction such that convergence properties can be
+  improved when the initial guess for $\ve{u}$ is poor
+  \citep{pawlowski_globalization_2006}.
+
 %%---------------------------------------------------------------------------%%
@@
 generate the Jacobian, but we also enjoy a Jacobian of numerical
-precision equivalent to that of our function evaluations. We therefore
-propose the \textit{Forward-Automated Newton-MCSA} (FANM) method that
-utilizes all of the above components. We hypothesize that such a
-method will be competitive with Newton-Krylov methods not only from a
-convergence and timing perspective, but also relax scaling
+precision equivalent to that of our function evaluations.
+
+We therefore propose the \textit{Forward-Automated Newton-MCSA} (FANM)
+method that utilizes all of the above components. We hypothesize that
+such a method will be competitive with Newton-Krylov methods not only
+from a convergence and timing perspective, but also relax scaling
 requirements of matrix-free methods and memory costs of both
@@
 where $\rho$ is the fluid density, $\ve{u}$ is the fluid velocity,
-$\ve{g}$ gravity, $C_p$ the specific heat capacity at constant
-pressure of the fluid and $T$ the temperature of the
-fluid. Eq~(\ref{eq:ns_momentum}) provides momentum transport,
+$C_p$ the specific heat capacity at constant pressure of the fluid,
+$T$ the temperature of the fluid, and $\ve{g}$ the acceleration due to
+gravity. Eq~(\ref{eq:ns_momentum}) provides momentum transport,
 Eq~(\ref{eq:ns_continuity}) provides the mass balance, and

### 2013-07-11 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding first titan scaling figures to parallel chapter
adding parallel verification data
fixed paragraph spacing in spn chapter

@@
 case with full downscatter and full upscatter.
+
 \begin{table}[h!]
@@
 \end{table}
+
 It is readily apparent from the tabulated data that point Jacobi
@@
 limit of a block size of one, the block Jacobi method reduces to the
-point Jacobi method. For high performance implementations this has
-several attractive properties. First, the blocks in the matrix come
-from the energy/angle discretization of the transport equation as
-given by Eq~(\ref{eq:A_block_matrix}). Each block on the diagonal is
-bound to a mesh element in the system (note there are 8 blocks on the
-diagonal in each of the sparsity patterns with a mesh of $2 \times 2
-\times 2$) and therefore we expect the matrix elements forming the
-block to be entirely local. Second, these blocks are typically dense
-and nearly lower triangular for many transport problems meaning that
-established dense matrix methods can be used for fast inversion.
+point Jacobi method.
+
+For high performance implementations this has several attractive
+properties. First, the blocks in the matrix come from the energy/angle
+discretization of the transport equation as given by
+Eq~(\ref{eq:A_block_matrix}). Each block on the diagonal is bound to a
+mesh element in the system (note there are 8 blocks on the diagonal in
+each of the sparsity patterns with a mesh of $2 \times 2 \times 2$)
+and therefore we expect the matrix elements forming the block to be
+entirely local. Second, these blocks are typically dense and nearly
+lower triangular for many transport problems meaning that established
+dense matrix methods can be used for fast inversion.
 
@@
 %%---------------------------------------------------------------------------%%
-\section{MCSA Verification}
+\section{MCSA Verification\ }
 \label{sec:spn_mcsa_verification}
@@
 the solution tolerance, then the MCSA solutions will be deemed correct
-for these problems. Additionally, this verification is only applicable
-to serial results with parallel MCSA verification for this problem
-provided in Chapter~\ref{ch:parallel_methods}.
+for these problems.
 
@@
 %%---------------------------------------------------------------------------%%
-\section{MCSA Performance Comparison to Conventional Methods}
+\section{MCSA Performance Comparison to Conventional Methods\ }
 \label{sec:spn_comparison}

### 2013-07-11 doc/dissertation/document/chapters/conclusion/conclusion.tex
starting work on breakdown analysis and adding sections to the concluding chapter
updates to introduction chapter
updating section numbering
adding first titan scaling figures to parallel chapter
updating parallel algorithm explanation
updating preliminaries of nonlinear chapter

@@
-\chapter{Conclusions and Analysis}
+\chapter{Conclusions and Analysis\ }
 \label{ch:conclusion}
 Put conclusions here.
+
+%%---------------------------------------------------------------------------%%
+\section{Monte Carlo Synthetic Acceleration Solutions for the $SP_N$ Equations\ }
+\label{sec:spn_conclusion}
+
+%%---------------------------------------------------------------------------%%
+\section{Monte Carlo Synthetic Acceleration Solutions for Navier-Stokes Equations\ }
+\label{sec:nonlinear_conclusions}
+
+%%---------------------------------------------------------------------------%%
+\section{Parallel Monte Carlo Synthetic Acceleration\ }
+\label{sec:parallel_mc_conclusions}
+
+%%---------------------------------------------------------------------------%%
+\section{Future Work\ }
+\label{sec:future_work}
+
+Breaking away from $\rho(H) < 1$
+
+Monte Carlo methods of the 2nd Degree?
+
+Stochastic projection methods?
+
+Stochastic variants of conventional preconditioners?
+
+Variance reduction for preconditioning?
+
+Controling MCSA histories as a function of nonlinear iteration/forcing
+term?
+
+Implementation optimization - can we approach Krylov performance?
+
+Reduction in memory footprint
+
+Multiple sets parallel overhead reduction
+
+%%---------------------------------------------------------------------------%%
+\section{Closing Remarks\ }
+\label{sec:closing}

### 2013-07-11 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
updating preliminaries of nonlinear chapter

@@
 Nonlinear transport problems are a common occurrence in single and
-multiple physics problems. Systems of partial differential equations
-such as those that describe fluid flow or more general transport
-processes when discretized by conventional methods yield discrete sets
-of stiff equations with nonlinearities present in the
-variables. Traditionally, such systems have been solved by linearizing
-them in a form where the nonlinearities in the variables are
-eliminated and more traditional linear methods can be used for
+multiple physics problems in nuclear engineering. Systems of partial
+differential equations such as those that describe fluid flow or more
+general transport processes when discretized by conventional methods
+yield discrete sets of stiff equations with nonlinearities present in
+the variables. Traditionally, such systems have been solved by
+linearizing them in a form where the nonlinearities in the variables
+are eliminated and more traditional linear methods can be used for
 solutions. Often characterized as segregated methods where physics
@@
 consistent nonlinear methods based on Newton's method have become more
-popular and many advances have been made in the computational physics
+popular and many advances have been made in the nuclear engineering
 field to employ these methods.
@@
 In the context of solving standalone linear systems, Monte Carlo
-methods do not provide significant merit over Krylov methods due to
+methods do not provide significant merit over subspace methods due to
 the fact that the linear operator must be explicitly formed. For many
-applications, such a requirement is prohibitive and perhaps not even
-feasible to implement. Therefore, a Monte Carlo solver is best suited
-for situations in which not only are Krylov methods applicable, but
-also in which the operator is readily, if not naturally,
-formed. Modern nonlinear methods meet both of these requirements with
-Newton methods used in conjunction with Krylov methods for a robust
-solution strategy. Furthermore, modern techniques exist that permit
-the automatic construction of the linear operator generated within a
+applications including higher fidelity neutron transport
+discretizations beyond the $SP_N$ equations solved in the previous
+chapter, such a requirement is prohibitive and perhaps not even
+feasible to implement. Therefore, a discrete Monte Carlo solution
+method is best suited for situations in which the operator is readily,
+if not naturally, formed. Modern nonlinear methods meet this
+requirement with Newton methods used in conjunction with Krylov
+methods operating on a fully formed matrix for a robust solution
+strategy. Furthermore, modern techniques exist that permit the
+automatic construction of the Jacobian operator generated within a
 Newton method based on the nonlinear residual evaluations, providing
 all of the components necessary for a Monte Carlo solver to provide
-value. We therefore devise a new nonlinear method based on the MCSA
-algorithm and Newton's method and discuss its potential benefits.
+value.
+
+We therefore devise a new nonlinear method based on the MCSA algorithm
+and Newton's method. In this chapter the fundamentals of Newton's
+method will be presented along with a brief discussion of inexact
+Newton methods. Newton-Krylov methods, a variant of inexact Newton
+methods, will be outlined along with two common techniques for
+Jacobian generation. Following this, the new algorithm will be
+presented and implementation details will be discussed. Using the new
+algorithm and a production fluid physics code, a group of nonlinear
+benchmark problems will be used to verify the correctness of the
+method relative to solutions obtained using conventional
+methods. Finally, will compare the performance of the new method
+against contemporary Newton-Krylov methods for the benchmark problems.
 
@@
 
+\subsection{Newton-Krylov Methods\ }
+\label{subsec:newton_krylov_methods}
+A form of inexact Newton methods, \textit{Newton-Krylov methods} are
+nonlinear iterative methods that leverage a Krylov subspace method as
+the linear solver for generating the Newton correction
+\citep{kelley_iterative_1995}. As we outlined in
+Appendix~\ref{ch:linear_problem}, Krylov methods are robust and enjoy
+efficient parallel implementations on modern
+architectures. Furthermore, their lack of explicit dependence on the
+operator make them easier to implement than other
+methods. Additionally, although many iterations can become memory
+intensive due to the need to store the Krylov subspace for the
+orthogonalization procedure, at each nonlinear iteration this cost is
+reset as the Jacobian matrix will change due to its dependence on the
+solution vector. This means that for every nonlinear iteration, a
+completely new linear system is formed for generating the Newton
+correction and we can modify the Krylov solver parameters accordingly
+to accommodate this. In most nonlinear problems, the Jacobian operator
+is generally non-symmetric and therefore either Krylov methods with
+long recurrence relations that can handle non-symmetric systems must
+be considered or the Newton correction system must be preconditioned
+such that the operator is symmetric and short recurrence relation
+methods can be potentially be used.
+
+With many Krylov methods available, which to use with the Newton
+method is dependent on many factors including convergence rates and
+memory usage. Several studies have been performed to investigate this
+\citep{mchugh_inexact_1993,knoll_newton-krylov_1995}. In their
+numerical studies in 1995, Knoll and McHugh used the set of highly
+nonlinear and stiff convection-diffusion-reaction equations to solve a
+set of tokamak plasma problems with the goal of measuring solver
+performance with Newton's method. They note several trade-offs in
+using Krylov methods with the Newton solver. The first is that the
+optimization condition that results from the constraints (e.g. the
+minimization of the GMRES residual over the Krylov space) can be
+relaxed by restricting the size of the subspace such that only a fixed
+number of subspace vectors may be maintained, thus reducing memory
+requirements.
+
+We can also relax the optimization condition by instead restarting the
+recurrence relation with a new set of vectors once a certain number of
+vectors have been generated. The optimization condition is maintained
+over that particular set of vectors, however, Knoll and McHugh note
+that this ultimately slows the convergence rate as compared to keeping
+all vectors as the new set of vectors is not necessarily orthogonal to
+the previous set, and therefore not optimal over the entire iteration
+procedure. The orthogonality condition can be relaxed by using a
+recurrence relation that does not generate a strictly orthonormal
+basis for the Krylov subspace such as the Lanzcos biorthogonalization
+procedure, resulting in memory savings due to the shorter Lanzcos
+recurrence relation. Based on their numerical analysis, they observed
+GMRES to be the most robust for any given nonlinear problem and
+therefore used it for subsequent work in this area.
+
+\subsection{Jacobian-Free Approximation}
+\label{subsec:jacobian_free_approximation}
+In many cases, the Jacobian is difficult to form from the difference
+equations and costly to evaluate for large equation sets. For simple
+nonlinear cases such as the Navier-Stokes equations, the derivatives
+can be computed and coded, but due to the complexity of those
+derivatives and the resulting difference equations this task can be
+tedious, error prone, and must be repeated for every equation
+set. Furthermore, in their 1995 work, Knoll and McHugh also noted that
+a dominating part of their computation time was the evaluation of the
+difference equations for building the Jacobian
+\citep{knoll_newton-krylov_1995}. By recognizing that Krylov methods
+only need the action of the operator on the vector instead of the
+operator itself, the Jacobian can instead be approximated through
+various numerical methods including a difference-based Jacobian-free
+formulation.
+
+Jacobian-Free methods, and in particular \textit{Jacobian-Free
+  Newton-Krylov} (JFNK) methods \citep{knoll_jacobian-free_2004}, rely
+on forming the action of the Jacobian on a vector as required by the
+Krylov solver through a forward difference scheme. In this case, the
+action of the Jacobian on some vector $\ve{v}$ is given as:
+\begin{equation}
+  \ve{J}(\ve{u})\ve{v} = \frac{\ve{F}(\ve{u} + \epsilon \ve{v}) -
+    \ve{F}(\ve{u})}{\epsilon}\:,
+  \label{eq:jacobian_free_product}
+\end{equation}
+where $\epsilon$ is a small number typically on the order of machine
+precision. Kelley \citep{kelley_iterative_1995} points out a potential
+downfall of this formulation in that if the discretization error in
+$\ve{F}(\ve{u})$ is on the order of the perturbation parameter
+$\epsilon$, then the finite difference error from
+Eq~(\ref{eq:jacobian_free_product}) pollutes the solution. In
+addition, Knoll and McHugh noted that for preconditioning purposes,
+part of the Jacobian must still explicitly be formed periodically and
+that linear solver robustness issues were magnified by the matrix-free
+approach due to the first-order approximation. This formation
+frequency coupled with the numerous evaluations of the Jacobian
+approximation create a situation where after so many nonlinear
+iterations, it becomes cheaper to instead fully form the
+Jacobian. For simple equation sets, this may only take 5-10 Newton
+iterations to reach this point while over 30 may be required for
+larger equations sets and therefore larger Jacobians.
+
+\subsection{Automatic Differentiation for Jacobian Generation}
+\label{subsec:automatic_differentiation}
+If it is acceptable to store the actual Jacobian matrix, other methods
+are available to construct it without requiring hand-coding and
+evaluating derivatives, thus eliminating the associated issues. In
+addition, if any additional equations are added to the system or a
+higher order functional approximation is desired, it would be useful
+to avoid regenerating and coding these derivatives. Becoming more
+prominent in the 1990's, \textit{automatic differentiation} is a
+mechanism by which the derivatives of a function can be generated
+automatically by evaluating it. Automatic differentiation is built on
+the concept that all functions discretely represented in a computer
+are ultimately represented by elementary mathematical operations. If
+the chain rule is applied to those elementary operations, then the
+derivatives of those functions can be computed to the order of
+accuracy of their original discretization in a completely automated
+way \citep{averick_computing_1994}.
+
+The work of Bartlett and others \citep{bartlett_automatic_2006}
+extended initial Fortran-based work in the area of automatic
+differentiation implementations to leverage the parametric type and
+operator overloading features of C++ \citep{stroustrup_c++_1997}. They
+formulate the differentiation problem from an element viewpoint by
+assuming that a global Jacobian can be assembled from local element
+function evaluations of $e_k : \mathbb{R}^{n_k} \rightarrow
+\mathbb{R}^{m_k}$, similar to the finite element assembly procedure
+as:
+\begin{equation}
+  \ve{J}(\ve{u}) = \sum_{i=1}^N \ve{Q}^T_i \ve{J}_k \ve{P}_i\:,
+  \label{eq:fad_global_jacobian}
+\end{equation}
+where $\ve{J}_{k_i} = \partial e_{k_i} / \partial P_i u$ is the
+$k^{th}$ element function Jacobian, $\ve{Q} \in \mathbb{R}^{n_{k_i}
+  \times N}$ is a projector onto the element domain and $\ve{P} \in
+\mathbb{R}^{m_{k_i} \times N}$ a projector onto the element range for
+$\ve{F}(\ve{u}) \in \mathbb{R}^{N \times N}$. The Jacobian matrix for
+each element will therefore have entirely local data in a dense
+structure, eliminating the need for parallel communication and sparse
+techniques during differentiation. Only when all local differentials
+are computed does communication of the Jacobian occur through
+gather/scatter operations in order to properly assembly it.
+
+Also of benefit is the fact that element-level computations generally
+consist of a smaller number of degrees of freedom, thus reducing
+memory requirements during evaluation as compared to a global
+formulation of the problem. Such a formulation is not limited to
+finite element formulations and is amenable to any scheme where the
+system is globally sparse with degrees of freedom coupled to local
+domains including finite volume representations. The templating
+capabilities of C++ were leveraged with the element-based evaluation
+and assembly scheme as in Eq~(\ref{eq:fad_global_jacobian}) by
+templating element function evaluation code on the evaluation type. If
+these functions are instantiated with standard floating point types
+then the residual is returned. If they are instead instantiated with
+the operator-overloaded automatic differentiation types, both the
+residual and Jacobian are returned.
+
+Of interest to Bartlett, Averick, and the many others that have
+researched automatic differentiation are measures of its performance
+relative to hand-coded derivatives and capturing the Jacobian matrix
+from matrix-free approximations. Given their element-based function
+evaluation scheme, Bartlett's work varied the number of degrees of
+freedom per element and compared both the floating point operation
+count and CPU time for both the templated automatic differentiation
+method and hand-coded derivatives for Jacobian evaluations. Although
+they observed a 50\% increase in floating point operations in the
+templated method over the hand-coded method, run times were observed
+to be over 3 times faster for the templated method due to the fact
+that the element-based formulation of the templated method is causing
+better utilization of cache and therefore faster data
+access. Furthermore, they observed linear scaling behavior for
+automatic differentiation as the number of degrees of freedom per
+element were increased. Based on these results, this type of automatic
+differentiation formulation was deemed acceptable for use in
+large-scale, production physics codes.
+
 %%---------------------------------------------------------------------------%%
@@
 consistent solution scheme
-\citep{evans_development_2006,evans_enhanced_2007,gaston_parallel_2009,godoy_parallel_2012}. Typically,
-for large scale simulations these problems are memory limited due to
-the subspaces generated by robust Krylov methods. Often, a matrix-free
-approach is chosen to relax memory requirements over directly
-generating the Jacobian matrix and facilitate the
+\citep{evans_development_2006,evans_enhanced_2007,gaston_parallel_2009,godoy_parallel_2012}. It
+is common that for large scale simulations these problems are memory
+limited due to the subspaces generated by robust Krylov methods which
+may often build hundreds of vectors during a Newton step. Often, a
+matrix-free approach is chosen to relax memory requirements over
+directly generating the Jacobian matrix and facilitate the
 implementation. However, as we observed in previous sections, these
@@
 systems. These methods, when used to accelerate a stationary method in
-MCSA, enjoy a robust implementation and exponential convergence
-rates. Further, the only storage required is that of the full linear
-system including the linear operator so that we may generate
-transition probabilities for the random walk sequence. Although this
-requires more storage to represent the linear system than that of a
-Krylov method where the operator is not required, we do not incur any
-additional storage costs once the iteration sequence begins. In the
-context of nonlinear problems, the Jacobian matrix that we are
-required to generate for the Monte Carlo solvers may be generated at
-will from the nonlinear functions in the Newton system using automatic
-differentiation. Not only do we then have a simple and automated way to
-generate the Jacobian, but we also enjoy a Jacobian of numerical
+MCSA, enjoy exponential convergence rates. Although this requires more
+storage to represent the linear system than that of a Krylov method
+where the operator is not required, we do not incur any additional
+storage costs once the iteration sequence begins. In the context of
+nonlinear problems, the Jacobian matrix that we are required to
+generate for the Monte Carlo solvers may be generated at will from the
+nonlinear functions in the Newton system using automatic
+differentiation. Not only do we then have a simple and automated way
+to generate the Jacobian, but we also enjoy a Jacobian of numerical
 precision equivalent to that of our function evaluations.
@@
 We therefore propose the \textit{Forward-Automated Newton-MCSA} (FANM)
-method that utilizes all of the above components. We hypothesize that
-such a method will be competitive with Newton-Krylov methods not only
-from a convergence and timing perspective, but also relax scaling
-requirements of matrix-free methods and memory costs of both
-matrix-free and fully formed Jacobian methods to allow the application
-developer to solve problems of finer discretization and higher-order
-functional representations while maintaining a robust and efficient
-parallel implementation.
+method that utilizes all of the above components. Presented in
+Algorithm~\ref{alg:fanm}, the FANM method is an inexact Newton method
+where MCSA is used to compute the Newton correction. In line 3,
+automatic differentiation is used at each iteration to build the
+Jacobian operator which can in turn be used to build weights and
+probabilities for the Monte Carlo game. In line 4, MCSA is used to
+solve the linear problem for the Newton correction. As with other
+inexact methods, any given forcing term can be used to control the
+convergence of the MCSA iteration at every Newton step. Finally, in
+line 5 the Newton correction is applied and the iteration proceeds
+until convergence. In addition to forcing term adjustments, any other
+techniques for improving performance of the nonlinear iteration may be
+used with the FANM method including backtracking methods.
 
@@
 \label{subsec:parallel_fanm}
-A parallel FANM method relies on a basic set of parallel matrix-vector
-operations as well as the global residual and Jacobian assembly
-procedure described in
-\S~\ref{subsubsec:automatic_differentiation}. Consider the Newton
-iteration scheme in Eq~(\ref{eq:newton_iteration}). We must first
-assemble the linear system in parallel through the element-wise
-function evaluations to generate both the global Jacobian operator and
-the global residual vector on the right hand side. Per Bartlett's
-work, efficient and automated parallel mechanisms are available to do
-this through a sequence of scatter/gather operations. With these tools
-available for residual and Jacobian generation, the remainder of the
-parallel procedure is simple, with the linear Newton correction system
-solved using the parallel MCSA method as previously described and the
-Newton correction applied to the previous iterate's solution through a
-parallel vector update operation.
+Here we briefly comment on how a FANM method may be implemented in
+parallel, defering the bulk of the discussion on parallel methods
+Monte Carlo to Chapter~\ref{ch:parallel_methods}. A parallel FANM
+method relies on a basic set of parallel matrix-vector operations
+outlined in \S~\ref{sec:parallel_krylov_methods} as well as the global
+residual and Jacobian assembly procedure described in
+\S~\ref{subsec:automatic_differentiation}. Consider the FANM iteration
+scheme in Algorithm~\ref{alg:fanm}. We must first assemble the linear
+system in parallel through the element-wise function evaluations to
+generate both the global Jacobian operator and the global residual
+vector on the right hand side. Per Bartlett's work, efficient and
+automated parallel mechanisms are available to do this through a
+sequence of scatter/gather operations. With these tools available for
+residual and Jacobian generation, the remainder of the parallel
+procedure is simple. The linear Newton correction system is solved
+using the parallel MCSA method that will be described in
+Chapter~\ref{ch:parallel_methods} and the Newton correction applied to
+the previous iterate's solution through a parallel vector update.
 
-As Newton methods are formulated independent of the inner linear
-solver generating the Newton corrections, the actual performance of
-the nonlinear iterations using MCSA is expected to be similar to that
-of traditional Newton-Krylov methods. Furthermore, we expect to
-achieve numerically identical answers with a Newton-MCSA method as
-other Newton methods and we should indeed verify this. The parallel
-performance of such a method will inherently be bound to the parallel
-Monte Carlo implementation of the linear solver as the parallel
-operations at the nonlinear iteration level are identical to those
-that you would perform with a Newton-Krylov method. Matrix-free
-formulations will have different parallel performance than these
-methods, and therefore we should compare FANM performance to JFNK-based
-schemes. More important than performance in this situation is the
-reduced memory pressure that a FANM implementation provides, as
-discussed in \S~\ref{subsec:fanm_storage}, because a FANM method will
-not generate a subspace in the linear solver and compressed storage
-for sparse matrices are utilized, we expect significant memory savings
-over Newton-Krylov methods. We must measure the memory utilization of
-both of these methods in order to quantify their differences and
-provide additional analysis of the FANM method's merits, or lack
-thereof.
+\clearpage
 
@@
 nonlinear convection term in Eq~(\ref{eq:ns_momentum}). In Shadid's
-work, Rayleigh numbers of up to \sn{1}{6} were used for this benchmark
-on a $100 \times 100$ square mesh grid.
+work, Rayleigh numbers of up to \sn{1}{6} were used for this benchmark.
 
@@
 influence of the nonlinear terms in Eq~(\ref{eq:ns_momentum}). Shadid
-also used a $100 \times 100$ grid for this benchmark problem with
-Reynolds numbers up to \sn{1}{4}.
+used Reynolds numbers up to \sn{1}{4} for this benchmark problem.
 
@@
 was used to discretize the domain with Reynolds number up to
-\sn{5}{2}.
+\sn{8}{2}.
 

### 2013-07-11 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
spell check on nonlinear chapter

@@
 benchmark problems will be used to verify the correctness of the
-method relative to solutions obtained using conventional
-methods. Finally, will compare the performance of the new method
-against contemporary Newton-Krylov methods for the benchmark problems.
+method relative to solutions obtained using conventional methods in
+various flow regimes. Finally, we will compare the performance of the
+new method against contemporary Newton-Krylov methods for the same
+benchmark problems.
 
@@
     methods} may be used to modify the Newton correction in a more
-  desireable direction such that convergence properties can be
+  desirable direction such that convergence properties can be
   improved when the initial guess for $\ve{u}$ is poor
@@
 Here we briefly comment on how a FANM method may be implemented in
-parallel, defering the bulk of the discussion on parallel methods
+parallel, deferring the bulk of the discussion on parallel methods
 Monte Carlo to Chapter~\ref{ch:parallel_methods}. A parallel FANM
@@
 parabolic flow profile over a zero velocity boundary representing a
-step. The flow over this step will generate a recirculating backflow
+step. The flow over this step will generate a recirculating back-flow
 under the inlet flow towards the step. As in the lid driven cavity

### 2013-07-11 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
spell check on nonlinear chapter

@@
 maintain that iterative performance, keeping in mind the potential
-impacts on scalability that were observed in this work.
+impacts on scalability that were observed in this work. However,
+although eliminating all Monte Carlo communication was observed to be
+the best mode of use for MCSA, the parallel Monte Carlo methods
+developed for this work still provide value for cases where only a
+Monte Carlo solution is desired. In those cases, full transport of the
+Monte Carlo histories are required in order to generate the correct
+solution.
 

### 2013-07-15 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
fixed grammar in nonlinear intro
adding more to nonlinear benchmarks section

@@
 yield discrete sets of stiff equations with nonlinearities present in
-the variables. Traditionally, such systems have been solved by
+the variables. These sets of equations characterize momentum and
+energy transport through diffusion, viscous forces, and inertial
+forces. Traditionally, such systems have been solved by
 linearizing them in a form where the nonlinearities in the variables
@@
 the linear solver for generating the Newton correction
-\citep{kelley_iterative_1995}. As we outlined in
+\citep{kelley_iterative_1995}. As outlined in
 Appendix~\ref{ch:linear_problem}, Krylov methods are robust and enjoy
@@
 of the energy, mass, and momentum equations.
+
 \begin{figure}[t!]
@@
 \end{figure}
+
 In this problem, a rectilinear grid is applied to the unit square. No
@@
 zero Neumann condition specified. Buoyancy driven flow is generated by
-the temperature gradient from the cold and hot Dirichlet conditions on
+the temperature gradient from the hot and cold Dirichlet conditions on
 the left and right boundaries of the box. By adjusting the Rayleigh
@@
 work, Rayleigh numbers of up to \sn{1}{6} were used for this benchmark.
+
+Temperature solutions for these equations are given in
+Figure~\ref{fig:convection_isotherms} for Rayleigh numbers of
+\sn{1}{3}, \sn{1}{4}, \sn{1}{5} and \sn{1}{6}. Note that as the
+Rayleigh number is increased, the fluid begins to rotate in the
+clockwise direction. This causes the temperature gradients to rotate
+as well with the fluid, given the increased rotation observed for the
+isotherms in the figure.
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/nonlinear_problem/convection_isotherms.png}
+  \end{center}
+  \caption{\textbf{Isotherms from solution to the thermal convection
+        cavity problem.} \textit{Top left: Ra = \sn{1}{3}; Top right:
+        Ra = \sn{1}{4}; Bottom left: Ra = \sn{1}{5}; Bottom left: Ra =
+        \sn{1}{6}. Isotherms on unit temperature scale from 0 to 1
+        with divisions of 0.1 for each calculation.}}
+  \label{fig:convection_isotherms}
+\end{figure}
+
+\clearpage
 
@@
 square.
+
 \begin{figure}[t!]
@@
 \end{figure}
+
 The top boundary condition will provide a driver for the flow and its
@@
 used Reynolds numbers up to \sn{1}{4} for this benchmark problem.
+
+\clearpage
 
@@
 
+For each of the benchmark problems, we will present the results of
+computations using FANM and directly compare them to results using a
+Newton-Krylov method in order to verify the correctness of FANM and
+its applicability to fluid problems. For each case, we will compare
+the global minimum, maximum, and average values for fluid velocity,
+pressure, and temperature where applicable. In addition, we will
+report the tolerance to which the nonlinear residual was converged.
+
+Given the difficulty of these benchmark problems, the linear solve at
+each Newton step was preconditioned with an algebraic multigrid
+method. For FANM, this preconditioning was required to reduce the
+spectral radius of the Jacobian operators generated at each Newton
+step and for the Newton-Krylov solutions this preconditioning was
+required to obtain good convergence properties although not necessary
+for convergence. For every benchmark calculation, both FANM and
+Newton-Krylov were preconditioned with the same multigrid parameters
+such that the conditioning of the system would not be a factor in the
+verification of the method or the perfomance comparison in the
+following section.
+
+\subsection{Thermal Convection Cavity Results}
+\label{subsec:thermal_convection_verification}
+
+\clearpage
+
+\subsection{Lid Driven Cavity Results}
+\label{subsec:lid_driven_verification}
+
+\clearpage
+
 %%---------------------------------------------------------------------------%%
@@
 \label{sec:fanm_comparison}
+
+\subsection{Thermal Convection Cavity Results}
+\label{subsec:thermal_convection_comparison}
+
+\clearpage
+
+\subsection{Lid Driven Cavity Results}
+\label{subsec:lid_driven_comparison}

### 2013-07-15 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
adding nonlinear comparison data to document

@@
 tuned to enhance the nonlinearities. We will then apply these
-equations to the following three standard benchmark problems.
+equations to the following two standard benchmark problems.
 
@@
 as well with the fluid, given the increased rotation observed for the
-isotherms in the figure.
+isotherms in the figure. As the Rayleigh number increases, thermal
+energy and momentum transport in the system become more dominated by
+convection rather than diffusion, causing the nonlinearities in the
+advective derivatives to grow.
 
@@
 
-\subsection{Backward-Facing Step Problem}
-\label{subsec:backward_facing_step}
-The third benchmark was generated by Gartling in 1990 and consists of
-both flow over a backward step and an outflow boundary condition
-\citep{gartling_test_1990}. Using the mass and momentum equations
-while neglecting the energy equation, this problem utilizes a longer
-domain with a 1/30 aspect ratio with the boundary conditions as shown
-in Figure~\ref{fig:backward_facing_step}.
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.3}{
-      \input{chapters/nonlinear_problem/backward_facing_step.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Problem setup for the backward facing step
-      benchmark.} \textit{Zero velocity boundary conditions are
-      applied at the top and bottom of the domain while the outflow
-      boundary condition on the right boundary is represented by zero
-      stress tensor components in the direction of the flow. For the
-      inlet conditions, the left boundary is split such that the top
-      half has a fully formed parabolic flow profile and the bottom
-      half has a zero velocity condition, simulating flow over a
-      step.}}
-  \label{fig:backward_facing_step}
-\end{figure}
-In this problem, the inflow condition is specified by a fully-formed
-parabolic flow profile over a zero velocity boundary representing a
-step. The flow over this step will generate a recirculating back-flow
-under the inlet flow towards the step. As in the lid driven cavity
-problem, the nonlinear behavior of this benchmark and the difficulty
-in obtaining a solution is dictated by the Reynolds number of the
-fluid. In Shadid's work, a $20 \times 400$ non-square rectilinear grid
-was used to discretize the domain with Reynolds number up to
-\sn{8}{2}.
-
-\clearpage
-
 %%---------------------------------------------------------------------------%%
@@
 
+As with the radiation transport calculations in the previous chapter,
+algrebraic preconditioning with the multigrid method results in dense
+linear systems. To mitigate this, the reduced domain approximation was
+again applied to the Monte Carlo solution within the MCSA solve
+occuring at each nonlinear iteration. The values used for the
+approximation will be given for each benchmark problem.
+
 \subsection{Thermal Convection Cavity Results}
 \label{subsec:thermal_convection_verification}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lcc}\hline\hline
+      \multicolumn{1}{l}{Parameter}& 
+      \multicolumn{1}{c}{Value}&
+      \multicolumn{1}{c}{Units}\\\hline
+      $x_{min}$ & 0.0 & $m$ \\
+      $x_{max}$ & 1.0 & $m$ \\
+      $y_{min}$ & 0.0 & $m$ \\
+      $y_{max}$ & 1.0 & $m$ \\
+      $N_x$ & 50 & - \\
+      $N_y$ & 50 & - \\
+      $DOFs$ & 10404 & - \\
+      Initial Temperature & 0.0 & $K$ \\
+      Density & 1.0 & $kg / m^3$ \\
+      Specific Heat & 1.0 & J / $(kg \times K)$ \\
+      Dynamic Viscosity & 0.71 & $N \times s / m^2$ \\
+      Thermal Conductivity & 1.0 & $W / (m \times K)$ \\
+      Thermal Expansion Coefficient & 1000 & $1 / K$ \\
+      Thermal Diffusivity & 1.0 & $m^2 / s$ \\
+      Kinematic Viscosity & 0.71 & $m^2 / s$ \\
+      Gravitational Acceleration & 10 & $m / s^2$ \\
+      Cavity Characteristic Length & 1.0 & $m$ \\
+      Prandtl Number & 0.71 & - \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Thermal convection cavity FANM verification model
+      problem parameters.}  \textit{The Navier-Stokes equations in 2
+      dimensions are used for the model problem. The Rayleigh number
+      is varied by modifying the hot boundary temperature to induce
+      larger temperature gradients and bouyancy-driven flow.}}
+  \label{tab:thermal_convection_parameters}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lc}\hline\hline
+      \multicolumn{1}{l}{Parameter}& 
+      \multicolumn{1}{c}{Value}\\\hline
+      Histories & 500,000 per iteration \\
+      Weight Cutoff & \sn{1}{-2} \\
+      Fixed Point Iteration & Richardson \\
+      Estimator & Adjoint Collision \\
+      Reduced Domain Fill Level & 200 \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Thermal convection cavity FANM verification MCSA
+      solver parameters.} \textit{No relaxation parameters or variance
+      reduction techniques were used with MCSA other than the reduced
+      domain approximation.}}
+  \label{tab:convection_mcsa_parameters}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lll}\hline\hline
+      \multicolumn{1}{l}{Parameter}& 
+      \multicolumn{1}{l}{Newton-Krylov Result}&
+      \multicolumn{1}{l}{FANM Result}\\
+      \hline
+      Average $U_x$ & -1.288835730E-05 & -1.288835730E-05 \\
+      Minimum $U_x$ & -3.677243960E+00 & -3.677243960E+00 \\
+      Maximum $U_x$ & 3.688908550E+00 & 3.688908550E+00 \\
+      \hline
+      Average $U_y$ & 3.403670530E-02 & 3.403670530E-02 \\
+      Minimum $U_y$ & -3.721178580E+00 & -3.721178580E+00 \\
+      Maximum $U_y$ & 3.791483420E+00 & 3.791483420E+00 \\
+      \hline
+      Average $P$ & 2.345554770E+02 & 2.345554770E+02 \\
+      Minimum $P$ & -5.509779180E+00 & -5.509779180E+00 \\
+      Maximum $P$ & 4.994885560E+02 & 4.994885560E+02 \\
+      \hline
+      Average $T$ & 3.595148700E-02 & 3.595148700E-02 \\
+      Minimum $T$ & 2.092448340E-04 & 2.092448340E-04 \\
+      Maximum $T$ & 7.178874510E-02 & 7.178874510E-02 \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Thermal convection cavity FANM verification for a
+      Rayleigh number of \sn{1}{3}.} \textit{The values reported are
+      global minima, maxima, and averages for each variable in the
+      problem. The velocity is given in components in the $x$ and $y$
+      directions.}}
+  \label{tab:convection_ra1e3_results}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lll}\hline\hline
+      \multicolumn{1}{l}{Parameter}& 
+      \multicolumn{1}{l}{Newton-Krylov Result}&
+      \multicolumn{1}{l}{FANM Result}\\
+      \hline
+      Average $U_x$ & -7.489201240E-04 & -7.489201240E-04 \\
+      Minimum $U_x$ & -1.607318000E+01 & -1.607318000E+01 \\
+      Maximum $U_x$ & 1.669098270E+01 & 1.669098270E+01 \\
+      \hline
+      Average $U_y$ & 3.377624420E-01 & 3.377624420E-01 \\
+      Minimum $U_y$ & -1.932174800E+01 & -1.932174800E+01 \\
+      Maximum $U_y$ & 2.047326820E+01 & 2.047326820E+01 \\
+      \hline
+      Average $P$ & 1.861894950E+03 & 1.861894950E+03 \\
+      Minimum $P$ & -3.353416240E+01 & -3.353416240E+01 \\
+      Maximum $P$ & 4.440642810E+03 & 4.440642810E+03 \\
+      \hline
+      Average $T$ & 3.510568440E-01 & 3.510568440E-01 \\
+      Minimum $T$ & 2.092448340E-04 & 2.092448340E-04 \\
+      Maximum $T$ & 7.178874510E-02 & 7.178874510E-02 \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Thermal convection cavity FANM verification for a
+      Rayleigh number of \sn{1}{4}.} \textit{The values reported are
+      global minima, maxima, and averages for each variable in the
+      problem. The velocity is given in components in the $x$ and $y$
+      directions.}}
+  \label{tab:convection_ra1e4_results}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lll}\hline\hline
+      \multicolumn{1}{l}{Parameter}& 
+      \multicolumn{1}{l}{Newton-Krylov Result}&
+      \multicolumn{1}{l}{FANM Result}\\
+      \hline
+      Average $U_x$ & -2.043639370E-02 & -2.043639370E-02 \\
+      Minimum $U_x$ & -3.814922470E+01 & -3.814922470E+01 \\
+      Maximum $U_x$ & -3.814922470E+01 & -3.814922470E+01 \\
+      \hline
+      Average $U_y$ & 2.774864300E+00 & 2.774864300E+00 \\
+      Minimum $U_y$ & -6.038682990E+01 & -6.038682990E+01 \\
+      Maximum $U_y$ & 7.873873560E+01 & 7.873873560E+01 \\
+      \hline
+      Average $P$ & 1.324208380E+04 & 1.324208380E+04 \\
+      Minimum $P$ & -1.390235490E+02 & -1.390235490E+02 \\
+      Maximum $P$ & 3.559215530E+04 & 3.559215530E+04 \\
+      \hline
+      Average $T$ & 3.027827770E+00 & 3.027827770E+00 \\
+      Minimum $T$ & 1.626405740E-02 & 1.626405740E-02 \\
+      Maximum $T$ & 7.162571130E+00 & 7.162571130E+00 \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Thermal convection cavity FANM verification for a
+      Rayleigh number of \sn{1}{5}.} \textit{The values reported are
+      global minima, maxima, and averages for each variable in the
+      problem. The velocity is given in components in the $x$ and $y$
+      directions.}}
+  \label{tab:convection_ra1e5_results}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lll}\hline\hline
+      \multicolumn{1}{l}{Parameter}& 
+      \multicolumn{1}{l}{Newton-Krylov Result}&
+      \multicolumn{1}{l}{FANM Result}\\
+      \hline
+      Average $U_x$ & -1.802569840E-01 & -1.802569840E-01 \\
+      Minimum $U_x$ & -6.292084130E+01 & -6.292084130E+01 \\
+      Maximum $U_x$ & 2.951113240E+02 & 2.951113240E+02 \\
+      \hline
+      Average $U_y$ & 1.336398550E+01 & 1.336398550E+01 \\
+      Minimum $U_y$ & -1.345663290E+02 & -1.345663290E+02 \\
+      Maximum $U_y$ & 3.227873890E+02 & 3.227873890E+02 \\
+      \hline
+      Average $P$ & 6.354313840E+04 & 6.354313840E+04 \\
+      Minimum $P$ & -1.356309630E+02 & -1.356309630E+02 \\
+      Maximum $P$ & 2.316308730E+05 & 2.316308730E+05 \\
+      \hline
+      Average $T$ & 1.780727610E+01 & 1.780727610E+01 \\
+      Minimum $T$ & 8.265788890E-02 & 8.265788890E-02 \\
+      Maximum $T$ & 7.098873170E+01 & 7.098873170E+01 \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Thermal convection cavity FANM verification for a
+      Rayleigh number of \sn{1}{6}.} \textit{The values reported are
+      global minima, maxima, and averages for each variable in the
+      problem. The velocity is given in components in the $x$ and $y$
+      directions.}}
+  \label{tab:convection_ra1e6_results}
+\end{table}
 
@@
 \label{subsec:lid_driven_verification}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lcc}\hline\hline
+      \multicolumn{1}{l}{Parameter}& 
+      \multicolumn{1}{c}{Value}&
+      \multicolumn{1}{c}{Units}\\\hline
+      $x_{min}$ & 0.0 & $m$ \\
+      $x_{max}$ & 1.0 & $m$ \\
+      $y_{min}$ & 0.0 & $m$ \\
+      $y_{max}$ & 1.0 & $m$ \\
+      $N_x$ & 50 & - \\
+      $N_y$ & 50 & - \\
+      $DOFs$ & 7803 & - \\
+      Density & 1.0 & $kg / m^3$ \\
+      Dynamic Viscosity & 0.71 & $N \times s / m^2$ \\
+      Kinematic Viscosity & 0.71 & $m^2 / s$ \\
+      Gravitational Acceleration & 10 & $m / s^2$ \\
+      Cavity Characteristic Length & 1.0 & $m$ \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Lid driven cavity FANM verification model
+      problem parameters.}  \textit{The Navier-Stokes equations in 2
+      dimensions are used for the model problem. The Reynolds number
+      is varied by modifying the velocity magnitude on the upper
+      boundary to induce flow driven by inertial forces.}}
+  \label{tab:lid_driven_parameters}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lc}\hline\hline
+      \multicolumn{1}{l}{Parameter}& 
+      \multicolumn{1}{c}{Value}\\\hline
+      Weight Cutoff & \sn{1}{-2} \\
+      Fixed Point Iteration & Richardson \\
+      Estimator & Adjoint Collision \\
+      Histories, Re=100 & 500,000 per iteration \\
+      Histories, Re=300 & 500,000 per iteration \\
+      Histories, Re=500 & 500,000 per iteration \\
+      Histories, Re=700 & 1,000,000 per iteration \\
+      Reduced Domain Fill Level, Re=100 & 200 \\
+      Reduced Domain Fill Level, Re=300 & 200 \\
+      Reduced Domain Fill Level, Re=500 & 200 \\
+      Reduced Domain Fill Level, Re=700 & 300 \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Lid driven cavity FANM verification MCSA solver
+      parameters.} \textit{No relaxation parameters or variance
+      reduction techniques were used with MCSA other than the reduced
+      domain approximation. Different values of fill level for the
+      reduced domain approximation and histories per iteration were
+      used at different Reynolds numbers due to convergence
+      requirements.}}
+  \label{tab:driven_mcsa_parameters}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lll}\hline\hline
+      \multicolumn{1}{l}{Parameter}& 
+      \multicolumn{1}{l}{Newton-Krylov Result}&
+      \multicolumn{1}{l}{FANM Result}\\
+      \hline
+      Average $U_x$ & 1.165632320E-01 & 1.165632320E-01 \\
+      Minimum $U_x$ & -2.423642920E+01 & -2.423642920E+01 \\
+      Maximum $U_x$ & 9.742670880E+01 & 9.742670880E+01 \\
+      \hline
+      Average $U_y$ & 2.038480420E-02 & 2.038480420E-02 \\
+      Minimum $U_y$ & -5.266315430E+01 & -5.266315430E+01 \\
+      Maximum $U_y$ & 2.720044160E+01 & 2.720044160E+01 \\
+      \hline
+      Average $P$ & -1.656193900E+02 & -1.656193900E+02 \\
+      Minimum $P$ & -9.635479370E+03 & -9.635479370E+03 \\
+      Maximum $P$ & 1.854517020E+04 & 1.854517020E+04 \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Lid driven cavity FANM verification for a Reynolds
+      number of 100.} \textit{The values reported are global minima,
+      maxima, and averages for each variable in the problem. The
+      velocity is given in components in the $x$ and $y$ directions.}}
+  \label{tab:driven_re100_results}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lll}\hline\hline
+      \multicolumn{1}{l}{Parameter}& 
+      \multicolumn{1}{l}{Newton-Krylov Result}&
+      \multicolumn{1}{l}{FANM Result}\\
+      \hline
+      Average $U_x$ & 3.382078040E-01 & 3.382078040E-01 \\
+      Minimum $U_x$ & -9.467140360E+01 & -9.467140360E+01 \\
+      Maximum $U_x$ & 2.895885380E+02 & 2.895885380E+02 \\
+      \hline
+      Average $U_y$ & 8.780242440E-02 & 8.780242440E-02 \\
+      Minimum $U_y$ & -1.830156060E+02 & -1.830156060E+02 \\
+      Maximum $U_y$ & 8.827373340E+01 & 8.827373340E+01 \\
+      \hline
+      Average $P$ & -2.830380160E+03 & -2.830380160E+03 \\
+      Minimum $P$ & -2.352177540E+04 & -2.352177540E+04 \\
+      Maximum $P$ & 9.254997120E+04 & 9.254997120E+04 \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Lid driven cavity FANM verification for a Reynolds
+      number of 300.} \textit{The values reported are global minima,
+      maxima, and averages for each variable in the problem. The
+      velocity is given in components in the $x$ and $y$ directions.}}
+  \label{tab:driven_re300_results}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lll}\hline\hline
+      \multicolumn{1}{l}{Parameter}& 
+      \multicolumn{1}{l}{Newton-Krylov Result}&
+      \multicolumn{1}{l}{FANM Result}\\
+      \hline
+      Average $U_x$ & 5.098921320E-01 & 5.098921320E-01 \\
+      Minimum $U_x$ & -1.731861690E+02 & -1.731861690E+02 \\
+      Maximum $U_x$ & 4.789146040E+02 & 4.789146040E+02 \\
+      \hline
+      Average $U_y$ & 1.365493140E-01 & 1.365493140E-01 \\
+      Minimum $U_y$ & -3.170930360E+02 & -3.170930360E+02 \\
+      Maximum $U_y$ & 1.655401440E+02 & 1.655401440E+02 \\
+      \hline
+      Average $P$ & -9.378317550E+03 & -9.378317550E+03 \\
+      Minimum $P$ & -4.155215280E+04 & -4.155215280E+04 \\
+      Maximum $P$ & 2.058742010E+05 & 2.058742010E+05 \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Lid driven cavity FANM verification for a
+      Reynolds number of 500.} \textit{The values reported are
+      global minima, maxima, and averages for each variable in the
+      problem. The velocity is given in components in the $x$ and $y$
+      directions.}}
+  \label{tab:driven_re500_results}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lll}\hline\hline
+      \multicolumn{1}{l}{Parameter}& 
+      \multicolumn{1}{l}{Newton-Krylov Result}&
+      \multicolumn{1}{l}{FANM Result}\\
+      \hline
+      Average $U_x$ & 6.479541170E-01 & 6.479541170E-01 \\
+      Minimum $U_x$ & -2.550685400E+02 & -2.550685400E+02 \\
+      Maximum $U_x$ & 6.663054900E+02 & 6.663054900E+02 \\
+      \hline
+      Average $U_y$ & 1.833690840E-01 & 1.833690840E-01 \\
+      Minimum $U_y$ & -4.491052980E+02 & -4.491052980E+02 \\
+      Maximum $U_y$ & 2.474376340E+02 & 2.474376340E+02 \\
+      \hline
+      Average $P$ & 1.833690840E-01 & 1.833690840E-01 \\
+      Minimum $P$ & -6.241704500E+04 & -6.241704500E+04 \\
+      Maximum $P$ & 3.518174810E+05 & 3.518174810E+05 \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Lid driven cavity FANM verification for a Reynolds
+      number of 700.} \textit{The values reported are global minima,
+      maxima, and averages for each variable in the problem. The
+      velocity is given in components in the $x$ and $y$ directions.}}
+  \label{tab:driven_re700_results}
+\end{table}
 
@@
 
+Using the benchmarks from the verification, we now compare the
+performance of the FANM method against a conventional Krylov
+method. For each benchmark, we will analyze the iterative performance
+of both the Newton solver and the linear solver used to compute the
+correction vector at each step. In addition, timing results will be discussed
+
 \subsection{Thermal Convection Cavity Results}
 \label{subsec:thermal_convection_comparison}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lcc}\hline\hline
+      \multicolumn{1}{l}{Rayleigh Number}& 
+      \multicolumn{1}{c}{Newton-Krylov Iterations}&
+      \multicolumn{1}{c}{FANM Iterations}\\
+      \hline
+      \sn{1}{3} & 5 & 5 \\
+      \sn{1}{4} & 7 & 7 \\
+      \sn{1}{5} & 9 & 10 \\
+      \sn{1}{6} & 11 & 11 \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Thermal convection cavity nonlinear iteration
+      results for the performance comparison.} \textit{The results
+      presented here were obtained from the benchmark verification
+      calculations.}}
+  \label{tab:convection_nonlinear_iter_comparison}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lcc}\hline\hline
+      \multicolumn{1}{l}{Rayleigh Number}& 
+      \multicolumn{1}{c}{GMRES Iterations}&
+      \multicolumn{1}{c}{MCSA Iterations}\\
+      \hline
+      \sn{1}{3} & 23 & 18 \\
+      \sn{1}{4} & 23 & 17 \\
+      \sn{1}{5} & 25 & 20 \\
+      \sn{1}{6} & 39 & 25 \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Thermal convection cavity total sum of linear
+      iterations for all nonlinear iterations.} \textit{The results
+      presented here were obtained from the benchmark verification
+      calculations.}}
+  \label{tab:convection_linear_iter_comparison}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lcc}\hline\hline
+      \multicolumn{1}{l}{Rayleigh Number}& 
+      \multicolumn{1}{c}{Newton Krylov $||\ve{F(\ve{u})}||$}&
+      \multicolumn{1}{c}{FANM $||\ve{F(\ve{u})}||$}\\
+      \hline
+      \sn{1}{3} & \sn{4.542}{-14} & \sn{1.208}{-14} \\
+      \sn{1}{4} & \sn{1.045}{-12} & \sn{7.012}{-13} \\
+      \sn{1}{5} & \sn{1.784}{-12} & \sn{1.059}{-12} \\
+      \sn{1}{6} & \sn{3.404}{-12} & \sn{3.479}{-12} \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Thermal convection cavity nonlinear residual norm
+      achieved convergence.} \textit{The results presented here were
+      obtained from the benchmark verification calculations.}}
+  \label{tab:convection_residual_norm_comparison}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lcc}\hline\hline
+      \multicolumn{1}{l}{Rayleigh Number}& 
+      \multicolumn{1}{c}{Newton-Krylov Speedup}\\
+      \hline
+      \sn{1}{3} & 338 \\
+      \sn{1}{4} & 336 \\
+      \sn{1}{5} & 346 \\
+      \sn{1}{6} & 465 \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Thermal convection cavity Newton-Krylov speedup
+      over FANM method.} \textit{Speedup values are rounded to the
+      nearest integer. The results presented here were obtained from
+      the benchmark verification calculations.}}
+  \label{tab:convection_speedup_comparison}
+\end{table}
 
@@
 \label{subsec:lid_driven_comparison}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lcc}\hline\hline
+      \multicolumn{1}{l}{Reynolds Number}& 
+      \multicolumn{1}{c}{Newton-Krylov Iterations}&
+      \multicolumn{1}{c}{FANM Iterations}\\
+      \hline
+      100 & 6 & 6 \\
+      300 & 9 & 9 \\
+      500 & 11 & 11 \\
+      700 & 14 & 10 \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Lid driven cavity nonlinear iteration
+      results for the performance comparison.} \textit{The results
+      presented here were obtained from the benchmark verification
+      calculations.}}
+  \label{tab:driven_nonlinear_iter_comparison}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lcc}\hline\hline
+      \multicolumn{1}{l}{Reynolds Number}& 
+      \multicolumn{1}{c}{GMRES Iterations}&
+      \multicolumn{1}{c}{MCSA Iterations}\\
+      \hline
+      100 & 27 & 42 \\
+      300 & 35 & 52 \\
+      500 & 41 & 56 \\
+      700 & 21 & 14 \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Lid driven cavity total sum of linear
+      iterations for all nonlinear iterations.} \textit{The results
+      presented here were obtained from the benchmark verification
+      calculations.}}
+  \label{tab:driven_linear_iter_comparison}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lcc}\hline\hline
+      \multicolumn{1}{l}{Reynolds Number}& 
+      \multicolumn{1}{c}{Newton Krylov $||\ve{F(\ve{u})}||$}&
+      \multicolumn{1}{c}{FANM $||\ve{F(\ve{u})}||$}\\
+      \hline
+      100 & \sn{5.453}{-14} & \sn{6.577}{-14} \\
+      300 & \sn{2.537}{-13} & \sn{2.779}{-13} \\
+      500 & \sn{6.367}{-13} & \sn{6.252}{-13} \\
+      700 & \sn{9.159}{-13} & \sn{1.282}{-12} \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Lid driven cavity nonlinear residual norm
+      achieved convergence.} \textit{The results presented here were
+      obtained from the benchmark verification calculations.}}
+  \label{tab:driven_residual_norm_comparison}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lcc}\hline\hline
+      \multicolumn{1}{l}{Reynolds Number}& 
+      \multicolumn{1}{c}{Newton-Krylov Speedup}\\
+      \hline
+      100 & 299 \\
+      300 & 322 \\
+      500 & 288 \\
+      700 & 488 \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Lid driven cavity Newton-Krylov speedup
+      over FANM method.} \textit{Speedup values are rounded to the
+      nearest integer. The results presented here were obtained from
+      the benchmark verification calculations.}}
+  \label{tab:driven_speedup_comparison}
+\end{table}

### 2013-07-15 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
adding nonlinear comparison data to document

@@
       Weight Cutoff & \sn{1}{-2} \\
-      Fixed Point & Richardson \\
-      Estimator & Collision \\
+      Fixed Point Iteration & Richardson \\
+      Estimator & Adjoint Collision \\
       %%
@@
   \end{center}
-  \caption{\textbf{Parallel MCSA verification solver parameters.}}
+  \caption{\textbf{Parallel MCSA verification solver parameters.}
+    \textit{No relaxation parameters or variance reduction techniques
+      were used with MCSA.}}
   \label{tab:parallel_mcsa_parameters}

### 2013-07-15 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
adding driven cavity solution figures

@@
 
+Velocity magnitudes are given in
+Figure~\ref{fig:driven_velocity_isocurves} for Reynolds numbers of
+100, 300, 500 and 700. Increasing the velocity on the top boundary of
+the problem drives up the Reynolds number of the system making it more
+difficult to solve.  As the Reynolds number grows, momentum and energy
+transport in the system is dominated by inertial forces rather than
+viscous forces. More rotation of the fluid is induced by intertial
+forces and we begin to see some additional vortices form besides the
+primary vortex in the center of the box.
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/nonlinear_problem/driven_velocity_isocurves.png}
+  \end{center}
+  \caption{\textbf{Velocity magnitude isocurves from solution to the
+      thermal convection cavity problem.} \textit{Top left: Re = 100;
+      Top right: Re= 300; Bottom left: Re = 500; Bottom left: Re =
+      700. Isocurves on unit velocity scale from 0 to 1 with divisions
+      of 0.05 for each calculation.}}
+  \label{fig:driven_velocity_isocurves}
+\end{figure}
+
 \clearpage

### 2013-07-16 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
adding to fanm verification section

@@
 \end{equation}
-where $\epsilon$ is a user defined tolerance parameter. Newton's
-method is \textit{consistent} in that all components of the nonlinear
-functions that describe the physics we are modeling are updated
-simultaneously in the iteration sequence with respect to one
+where $\epsilon$ is a user defined tolerance parameter.
+
+Newton's method is \textit{consistent} in that all components of the
+nonlinear functions that describe the physics we are modeling are
+updated simultaneously in the iteration sequence with respect to one
 another. This is in comparison to \textit{inconsistent} strategies,
@@
 line 5 the Newton correction is applied and the iteration proceeds
-until convergence. In addition to forcing term adjustments, any other
+until convergence. In addition to forcing term adjustments, other
 techniques for improving performance of the nonlinear iteration may be
-used with the FANM method including backtracking methods.
+used with FANM including backtracking methods.
 
@@
 \end{algorithm}
-
-\subsection{Parallel FANM Implementation}
-\label{subsec:parallel_fanm}
-Here we briefly comment on how a FANM method may be implemented in
-parallel, deferring the bulk of the discussion on parallel methods
-Monte Carlo to Chapter~\ref{ch:parallel_methods}. A parallel FANM
-method relies on a basic set of parallel matrix-vector operations
-outlined in \S~\ref{sec:parallel_krylov_methods} as well as the global
-residual and Jacobian assembly procedure described in
-\S~\ref{subsec:automatic_differentiation}. Consider the FANM iteration
-scheme in Algorithm~\ref{alg:fanm}. We must first assemble the linear
-system in parallel through the element-wise function evaluations to
-generate both the global Jacobian operator and the global residual
-vector on the right hand side. Per Bartlett's work, efficient and
-automated parallel mechanisms are available to do this through a
-sequence of scatter/gather operations. With these tools available for
-residual and Jacobian generation, the remainder of the parallel
-procedure is simple. The linear Newton correction system is solved
-using the parallel MCSA method that will be described in
-Chapter~\ref{ch:parallel_methods} and the Newton correction applied to
-the previous iterate's solution through a parallel vector update.
-
-\clearpage
 
@@
 
-Temperature solutions for these equations are given in
+Isotherms of the temperature solution for these equations are given in
 Figure~\ref{fig:convection_isotherms} for Rayleigh numbers of
@@
 
-Velocity magnitudes are given in
-Figure~\ref{fig:driven_velocity_isocurves} for Reynolds numbers of
-100, 300, 500 and 700. Increasing the velocity on the top boundary of
-the problem drives up the Reynolds number of the system making it more
-difficult to solve.  As the Reynolds number grows, momentum and energy
-transport in the system is dominated by inertial forces rather than
-viscous forces. More rotation of the fluid is induced by intertial
-forces and we begin to see some additional vortices form besides the
-primary vortex in the center of the box.
+Isocurves of the velocity magnitudes are given in
+Figure~\ref{fig:driven_velocity_isocurves} for solutions to the lid
+driven cavity problem at Reynolds numbers of 100, 300, 500 and
+700. Increasing the velocity on the top boundary of the problem drives
+up the Reynolds number of the system making it more difficult to
+solve. More rotation of the fluid is induced by intertial forces and
+we begin to see some additional vortices form besides the primary
+vortex in the center of the box.
 
@@
 each Newton step was preconditioned with an algebraic multigrid
-method. For FANM, this preconditioning was required to reduce the
-spectral radius of the Jacobian operators generated at each Newton
-step and for the Newton-Krylov solutions this preconditioning was
-required to obtain good convergence properties although not necessary
-for convergence. For every benchmark calculation, both FANM and
+method, a common technique for preconditioning these equations
+\citep{ghia_high-re_1982,evans_enhanced_2007}. For FANM, this
+preconditioning was required to reduce the spectral radius of the
+Jacobian operators generated at each Newton step and for the
+Newton-Krylov solutions this preconditioning was required to obtain
+good convergence properties although not necessary for
+convergence. For every benchmark calculation, both FANM and
 Newton-Krylov were preconditioned with the same multigrid parameters
@@
 occuring at each nonlinear iteration. The values used for the
-approximation will be given for each benchmark problem.
+approximation will be given for each benchmark problem. Without
+multigrid preconditioning or too small of a reduced domain fill level,
+the residual in MCSA was observed to stagnate. Typically this happens
+because the Richardson iteration will damp the higher order error
+modes on the scale of the computational grid used to solve the problem
+but will miss lower order error modes. In terms of Monte Carlo, this
+stagnation manifests itself in histories that simply do no terminate
+and therefore no solution is acquired.
+
+Results in this and the following section were produced by the Drekar
+multiphysics code base being developed at Sandia National Laboratories
+for fluid flow and other multiple physics simulations
+\citep{pawlowski_drekar_2012}. Drekar is a massively parallel physics
+framework that utilizes Newton-Krylov methods to generate a fully
+consistent solution scheme for the Navier-Stokes equations among other
+systems with Jacobian generation through automatic differentiation. To
+achieve these results, a Monte Carlo solver generated for this work
+was injected into the nonlinear toolchain within Drekar to form an
+implementation of the FANM method. For the Newton-Krylov solutions,
+the same production implementation of GMRES from the Trilinos
+scientific computing library Aztec \citep{heroux_overview_2005} used
+for the verification of MCSA for the $SP_N$ solutions in the previous
+chapter was also used here as a means of comparison.
 
@@
 \label{subsec:thermal_convection_verification}
+
+A Drekar model of thermal convection cavity problem was solved using
+both the Newton-Krylov solver and a FANM solver for Rayleigh numbers
+of \sn{1}{3}, \sn{1}{4}, \sn{1}{5} and
+\sn{1}{6}. Table~\ref{tab:thermal_convection_parameters} gives the
+physical parameters of the problem and
+Table~\ref{tab:convection_mcsa_parameters} gives the MCSA solver
+parameters used within the FANM method. Variation in the Rayleigh
+number of the system was achieved by keeping the parameters given in
+Table~\ref{tab:thermal_convection_parameters} and varying temperature
+of the hot side of the cavity. In addition to multigrid
+preconditioning for the linear solution at each Newton step,
+backtracking as outlined in Shadid's work \citep{shadid_inexact_1997}
+was used as a means of globalization assisting the Newton correction
+computation while the forcing term was computed using method 2 from
+Shadid's work.
 
@@
 
+For the verification, at each Rayleigh number variation the global
+minimum, maximum, and average values were required for fluid velocity
+in the x and y directions, fluid pressure, and fluid
+temperature. Tables~\ref{tab:convection_ra1e3_results},
+\ref{tab:convection_ra1e4_results},
+\ref{tab:convection_ra1e5_results}, and
+\ref{tab:convection_ra1e6_results} give the results of these
+computations. It should be noted that the negative pressures computed
+are due to a relative pressure being reported. For each of the data
+points presented, both Newton-Krylov solutions and FANM solutions
+computed identical results for all of these global values.
+
 \begin{table}[h!]
@@
 
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lcc}\hline\hline
+      \multicolumn{1}{l}{Rayleigh Number}& 
+      \multicolumn{1}{c}{Newton Krylov $||\ve{F(\ve{u})}||$}&
+      \multicolumn{1}{c}{FANM $||\ve{F(\ve{u})}||$}\\
+      \hline
+      \sn{1}{3} & \sn{4.542}{-14} & \sn{1.208}{-14} \\
+      \sn{1}{4} & \sn{1.045}{-12} & \sn{7.012}{-13} \\
+      \sn{1}{5} & \sn{1.784}{-12} & \sn{1.059}{-12} \\
+      \sn{1}{6} & \sn{3.404}{-12} & \sn{3.479}{-12} \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Thermal convection cavity nonlinear residual norm
+      achieved convergence.} \textit{The results presented here were
+      obtained from the benchmark verification calculations.}}
+  \label{tab:convection_residual_norm_comparison}
+\end{table}
+
 \clearpage
@@
 \label{subsec:lid_driven_verification}
+
+A Drekar model of thermal convection cavity problem was solved using
+both the Newton-Krylov solver and a FANM solver for Reynolds numbers
+of 100, 300, 500, and
+700. Table~\ref{tab:thermal_convection_parameters} gives the physical
+parameters of the problem and
+Table~\ref{tab:convection_mcsa_parameters} gives the MCSA solver
+parameters used within the FANM method. Variation in the Reynolds
+number of the system was achieved by keeping the parameters given in
+Table~\ref{tab:thermal_convection_parameters} and varying the fluid
+velocity in the $x$ direction on the top boundary. In addition to
+multigrid preconditioning for the linear solution at each Newton step,
+a polynomial line search method was used as a means of globalization
+assisting the Newton correction computation as outlined by Pawlowski
+in \citep{pawlowski_globalization_2006} while the forcing term was
+computed using method 2 from Shadid's work
+\citep{shadid_inexact_1997}.
 
@@
 
+In the literature on this benchmark, solutions at Reynolds numbers of
+up to \sn{1}{4} were achieved using a variety of nonlinear solution
+techniques. With the FANM method, at Reynolds numbers above 700,
+convergence of the linear iterations could not be achieved regardless
+of how agressive the multigrid preconditioner was set. In these cases,
+the Jacobian operators are ill-conditioned enough that even GMRES
+converges slowly relative to the thermal convection cavity
+benchmark. Therefore, the spectral radius limitation on MCSA is
+extremely prohibitive in this case and others were transport driven by
+intertial forces is dominant. Because of this, only solutions to the
+lid driven cavity benchmark at Reynolds numbers of up to 700 are
+presented.
+
+For the verification, at each Reynolds number variation the global
+minimum, maximum, and average values were required for fluid velocity
+in the x and y directions, fluid pressure, and fluid
+temperature. Tables~\ref{tab:driven_re100_results},
+\ref{tab:driven_re300_results}, \ref{tab:driven_re500_results}, and
+\ref{tab:driven_re700_results} give the results of these
+computations. It should be noted that the negative pressures computed
+are due to a relative pressure being reported. For each of the data
+points presented, both Newton-Krylov solutions and FANM solutions
+computed identical results for all of these global values.
+
 \begin{table}[h!]
@@
 
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lcc}\hline\hline
+      \multicolumn{1}{l}{Reynolds Number}& 
+      \multicolumn{1}{c}{Newton Krylov $||\ve{F(\ve{u})}||$}&
+      \multicolumn{1}{c}{FANM $||\ve{F(\ve{u})}||$}\\
+      \hline
+      100 & \sn{5.453}{-14} & \sn{6.577}{-14} \\
+      300 & \sn{2.537}{-13} & \sn{2.779}{-13} \\
+      500 & \sn{6.367}{-13} & \sn{6.252}{-13} \\
+      700 & \sn{9.159}{-13} & \sn{1.282}{-12} \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Lid driven cavity nonlinear residual norm
+      achieved convergence.} \textit{The results presented here were
+      obtained from the benchmark verification calculations.}}
+  \label{tab:driven_residual_norm_comparison}
+\end{table}
+
 \clearpage
@@
       \multicolumn{1}{l}{Rayleigh Number}& 
-      \multicolumn{1}{c}{Newton Krylov $||\ve{F(\ve{u})}||$}&
-      \multicolumn{1}{c}{FANM $||\ve{F(\ve{u})}||$}\\
-      \hline
-      \sn{1}{3} & \sn{4.542}{-14} & \sn{1.208}{-14} \\
-      \sn{1}{4} & \sn{1.045}{-12} & \sn{7.012}{-13} \\
-      \sn{1}{5} & \sn{1.784}{-12} & \sn{1.059}{-12} \\
-      \sn{1}{6} & \sn{3.404}{-12} & \sn{3.479}{-12} \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Thermal convection cavity nonlinear residual norm
-      achieved convergence.} \textit{The results presented here were
-      obtained from the benchmark verification calculations.}}
-  \label{tab:convection_residual_norm_comparison}
-\end{table}
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lcc}\hline\hline
-      \multicolumn{1}{l}{Rayleigh Number}& 
       \multicolumn{1}{c}{Newton-Krylov Speedup}\\
@@
       \multicolumn{1}{l}{Reynolds Number}& 
-      \multicolumn{1}{c}{Newton Krylov $||\ve{F(\ve{u})}||$}&
-      \multicolumn{1}{c}{FANM $||\ve{F(\ve{u})}||$}\\
-      \hline
-      100 & \sn{5.453}{-14} & \sn{6.577}{-14} \\
-      300 & \sn{2.537}{-13} & \sn{2.779}{-13} \\
-      500 & \sn{6.367}{-13} & \sn{6.252}{-13} \\
-      700 & \sn{9.159}{-13} & \sn{1.282}{-12} \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Lid driven cavity nonlinear residual norm
-      achieved convergence.} \textit{The results presented here were
-      obtained from the benchmark verification calculations.}}
-  \label{tab:driven_residual_norm_comparison}
-\end{table}
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lcc}\hline\hline
-      \multicolumn{1}{l}{Reynolds Number}& 
       \multicolumn{1}{c}{Newton-Krylov Speedup}\\

### 2013-07-16 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
adding to fanm verification section

@@
 \label{ch:parallel_methods}
-For MCSA to be viable at the production scale for nuclear engineering
-applications, scalable parallel implementations of the algorithm are
-required. Reviewing the literature, MCSA has yet to be parallelized
-and the Neumann-Ulam Monte Carlo method on which it is built has only
-been parallelized through history-level parallelism with full domain
-replication \citep{alexandrov_efficient_1998}. In order to solve large
-linear transport systems with MCSA, a domain-decomposed parallel
-strategy is required. In the formulation of a parallel MCSA algorithm,
-we recognize that the algorithm occurs in two stages, an outer
-iteration performing fixed point iteration and applying the
-correction, and an inner Monte Carlo solver that is generating the
-correction via the adjoint or forward methods. The parallel aspects of
-both these components must be considered. Therefore, we will develop
-and implement parallel algorithms for the Neumann-Ulam and MCSA
-methods leveraging both the knowledge gained from the general parallel
-implementations of Krylov methods reviewed in
-Appendix~\ref{ch:linear_problem} and modern parallel strategies for
-domain decomposed Monte Carlo as developed by the reactor physics
-community.
+For MCSA and FANM to be viable at the production scale for nuclear
+engineering applications, scalable parallel implementations of the
+algorithm are required. Reviewing the literature, MCSA has yet to be
+parallelized and the Neumann-Ulam Monte Carlo method on which it is
+built has only been parallelized through history-level parallelism
+with full domain replication \citep{alexandrov_efficient_1998}. In
+order to solve large linear transport systems with MCSA, a
+domain-decomposed parallel strategy is required. In the formulation of
+a parallel MCSA algorithm, we recognize that the algorithm occurs in
+two stages, an outer iteration performing fixed point iteration and
+applying the correction, and an inner Monte Carlo solver that is
+generating the correction via the adjoint or forward methods. The
+parallel aspects of both these components must be
+considered. Therefore, we will develop and implement parallel
+algorithms for the Neumann-Ulam and MCSA methods leveraging both the
+knowledge gained from the general parallel implementations of Krylov
+methods reviewed in Appendix~\ref{ch:linear_problem} and modern
+parallel strategies for domain decomposed Monte Carlo as developed by
+the reactor physics community.
 
@@
 
+\subsection{Parallel FANM Implementation}
+\label{subsec:parallel_fanm}
+Here we briefly comment on how a FANM method may be implemented in
+parallel. A parallel FANM method relies on a basic set of parallel
+matrix-vector operations outlined in
+\S~\ref{sec:parallel_krylov_methods} as well as the global residual
+and Jacobian assembly procedure described in
+\S~\ref{subsec:automatic_differentiation}. Consider the FANM iteration
+scheme in Algorithm~\ref{alg:fanm}. We must first assemble the linear
+system in parallel through the element-wise function evaluations to
+generate both the global Jacobian operator and the global residual
+vector on the right hand side. Per Bartlett's work on automatic
+differentiation, efficient and automated parallel mechanisms are
+available to do this through a sequence of scatter/gather
+operations. With these tools available for residual and Jacobian
+generation, the remainder of the parallel procedure is simple. The
+linear Newton correction system is solved using the parallel MCSA
+method as described in the previous section and the Newton correction
+applied to the previous iterate's solution through a parallel vector
+update.
+
 %%---------------------------------------------------------------------------%%

### 2013-07-16 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
cleaning up the first part of the nonlinear chapter

@@
 multiple physics problems in nuclear engineering. Systems of partial
-differential equations such as those that describe fluid flow or more
-general transport processes when discretized by conventional methods
-yield discrete sets of stiff equations with nonlinearities present in
-the variables. These sets of equations characterize momentum and
-energy transport through diffusion, viscous forces, and inertial
-forces. Traditionally, such systems have been solved by
-linearizing them in a form where the nonlinearities in the variables
-are eliminated and more traditional linear methods can be used for
+differential equations, such as the Navier-Stokes equations that
+describe fluid flow, yield discrete sets of stiff equations with
+nonlinearities present in the variables when discretized by
+conventional methods. These sets of equations characterize momentum
+and energy transport through diffusion, viscous forces, and inertial
+forces. Traditionally, such systems have been solved by linearizing
+them in a form where the nonlinearities in the variables are
+eliminated and more traditional linear methods can be used for
 solutions. Often characterized as segregated methods where physics
@@
 methods, will be outlined along with two common techniques for
-Jacobian generation. Following this, the new algorithm will be
-presented and implementation details will be discussed. Using the new
-algorithm and a production fluid physics code, a group of nonlinear
-benchmark problems will be used to verify the correctness of the
-method relative to solutions obtained using conventional methods in
-various flow regimes. Finally, we will compare the performance of the
-new method against contemporary Newton-Krylov methods for the same
-benchmark problems.
+Jacobian generation as a means of defining current
+practices. Following this, the new algorithm will be presented and
+implementation details will be discussed. Using the new algorithm and
+a production fluid physics code, a group of challenging nonlinear
+benchmark problems for the Navier-Stokes equations will be used to
+verify the correctness of the method relative to solutions obtained
+using conventional methods in various flow regimes. Finally, we will
+compare the performance of the new method against contemporary
+Newton-Krylov methods for the same benchmark problems.
 
@@
 \citep{knoll_newton-krylov_1995}. By recognizing that Krylov methods
-only need the action of the operator on the vector instead of the
+only need the action of the operator on a vector instead of the
 operator itself, the Jacobian can instead be approximated through
@@
 $\epsilon$, then the finite difference error from
-Eq~(\ref{eq:jacobian_free_product}) pollutes the solution. In
-addition, Knoll and McHugh noted that for preconditioning purposes,
+Eq~(\ref{eq:jacobian_free_product}) pollutes the solution.
+
+In addition, Knoll and McHugh noted that for preconditioning purposes,
 part of the Jacobian must still explicitly be formed periodically and
@@
 approximation create a situation where after so many nonlinear
-iterations, it becomes cheaper to instead fully form the
-Jacobian. For simple equation sets, this may only take 5-10 Newton
-iterations to reach this point while over 30 may be required for
-larger equations sets and therefore larger Jacobians.
+iterations, it becomes cheaper to instead fully form the Jacobian. For
+simple equation sets, this may only take 5-10 Newton iterations to
+reach this point while over 30 may be required for larger equations
+sets and therefore larger Jacobians.
 
@@
 are computed does communication of the Jacobian occur through
-gather/scatter operations in order to properly assembly it.
+gather/scatter operations in order to properly assemble it.
 
@@
 \end{algorithm}
+
+At first glance, FANM is simply a variant of an inexact Newton method
+with a specific requirement of a fully formed Jacobian. However, FANM
+has the potential to provide value in several areas. First, if
+iterative or parallel performance of MCSA can be demonstrated to
+better than that of GMRES or similar subspace methods for asymmetric
+systems, FANM may outperform a Newton-Krylov method. Second, for
+situations where memory is a constraint, especially in situations
+where the Jacobian is generated often if not at every Newton
+iteration, not building a subspace for ill-conditioned problems will
+alleviate the associated cost. However, until the preconditioning
+issues associated with MCSA as presented in the previous chapter using
+the $SP_N$ equations can be resolved, this hypothesis can
+unfortunately not be researched. Explicit algebraic preconditioning
+used in conventional solution methods will almost certainly be
+required to reduce the spectral radius of most Jacobian operators for
+MCSA to apply and therefore the current preconditioning strategy will
+destroy all potential memory benefits of FANM.
+
+\clearpage
 
@@
 \label{subsec:natural_convection_cavity}
-In 1983 a benchmark solution for the natural convection of air in a
-square cavity was published \citep{de_vahl_davis_natural_1983} as
+In 1983 a set of benchmark solutions for the natural convection of air
+in a square cavity was published \citep{de_vahl_davis_natural_1983} as
 shown in Figure~\ref{fig:natural_convection_cavity} for the solution

### 2013-07-16 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
finished initial draft of nonlinear chapter

@@
 the global minimum, maximum, and average values for fluid velocity,
-pressure, and temperature where applicable. In addition, we will
-report the tolerance to which the nonlinear residual was converged.
+pressure, and temperature where applicable. If the values for these
+quantities computed using FANM match those computed using a production
+Newton-Krylov method, the implementation will be deemed correct for
+the purposes of this work. In addition, we will report the tolerance
+to which the nonlinear residual was converged as an additional means
+of comparison for correctness.
 
@@
 number of the system was achieved by keeping the parameters given in
-Table~\ref{tab:thermal_convection_parameters} and varying temperature
-of the hot side of the cavity. In addition to multigrid
+Table~\ref{tab:thermal_convection_parameters} constant and varying
+temperature of the hot side of the cavity. In addition to multigrid
 preconditioning for the linear solution at each Newton step,
 backtracking as outlined in Shadid's work \citep{shadid_inexact_1997}
-was used as a means of globalization assisting the Newton correction
-computation while the forcing term was computed using method 2 from
-Shadid's work.
+was used as a means of globalization to assist the Newton correction
+computation while the forcing term was computed at each Newton step
+using method 2 from Shadid's work. In this forcing term method,
+nonlinear residuals from the two previous Newton steps are used to
+inform the choice of the forcing term for the next step.
 
@@
 
-For the verification, at each Rayleigh number variation the global
-minimum, maximum, and average values were required for fluid velocity
-in the x and y directions, fluid pressure, and fluid
+For the verification calculations, at each Rayleigh number variation
+the global minimum, maximum, and average values were recorded for
+fluid velocity in the x and y directions, fluid pressure, and fluid
 temperature. Tables~\ref{tab:convection_ra1e3_results},
@@
 points presented, both Newton-Krylov solutions and FANM solutions
-computed identical results for all of these global values.
+computed identical numerical results for all of these global
+values. No variation in solution was noted for each of the
+computations within the tolerance of the converged nonlinear residual
+norms given in
+Table~\ref{tab:convection_residual_norm_comparison}. Variations in the
+convergence behavior and differences in performance for these
+solutions will be analyzed in the following section.
 
@@
 number of the system was achieved by keeping the parameters given in
-Table~\ref{tab:thermal_convection_parameters} and varying the fluid
-velocity in the $x$ direction on the top boundary. In addition to
-multigrid preconditioning for the linear solution at each Newton step,
-a polynomial line search method was used as a means of globalization
-assisting the Newton correction computation as outlined by Pawlowski
-in \citep{pawlowski_globalization_2006} while the forcing term was
-computed using method 2 from Shadid's work
+Table~\ref{tab:thermal_convection_parameters} fixed and varying the
+fluid velocity in the $x$ direction on the top boundary. In addition
+to multigrid preconditioning for the linear solution at each Newton
+step, a polynomial line search method was used as a means of
+globalization assisting the Newton correction computation as outlined
+by Pawlowski in \citep{pawlowski_globalization_2006} while the forcing
+term was again computed using method 2 from Shadid's work
 \citep{shadid_inexact_1997}.
@@
 up to \sn{1}{4} were achieved using a variety of nonlinear solution
-techniques. With the FANM method, at Reynolds numbers above 700,
+techniques. With the FANM method, at Reynolds numbers above 700
 convergence of the linear iterations could not be achieved regardless
@@
 benchmark. Therefore, the spectral radius limitation on MCSA is
-extremely prohibitive in this case and others were transport driven by
-intertial forces is dominant. Because of this, only solutions to the
-lid driven cavity benchmark at Reynolds numbers of up to 700 are
+extremely prohibitive in this case and others where transport driven
+by intertial forces is dominant. Because of this, only solutions to
+the lid driven cavity benchmark at Reynolds numbers of up to 700 are
 presented.
 
-For the verification, at each Reynolds number variation the global
-minimum, maximum, and average values were required for fluid velocity
-in the x and y directions, fluid pressure, and fluid
-temperature. Tables~\ref{tab:driven_re100_results},
+For the verification calculations, at each Reynolds number variation
+the global minimum, maximum, and average values were recorded for the
+fluid velocity in the x and y directions and the fluid
+pressure. Tables~\ref{tab:driven_re100_results},
 \ref{tab:driven_re300_results}, \ref{tab:driven_re500_results}, and
 \ref{tab:driven_re700_results} give the results of these
-computations. It should be noted that the negative pressures computed
-are due to a relative pressure being reported. For each of the data
-points presented, both Newton-Krylov solutions and FANM solutions
-computed identical results for all of these global values.
+computations. As with the thermal convection problem, the negative
+pressures computed are due to a relative pressure being reported. For
+each of the data points given, both Newton-Krylov solutions and FANM
+solutions computed identical numerical results for all of these global
+values. The norm of the converged nonlinear residual for the
+calculations is given in
+Table~\ref{tab:driven_residual_norm_comparison}. All computed global
+values were within these tolerances.
 
@@
 of both the Newton solver and the linear solver used to compute the
-correction vector at each step. In addition, timing results will be discussed
+correction vector at each step. In addition, timing results will be
+discussed.
 
@@
 \label{subsec:thermal_convection_comparison}
+
+At each Rayleigh number used for the thermal convection cavity
+problem, Table~\ref{tab:convection_nonlinear_iter_comparison} gives
+the number of Newton iterations required for convergence for each
+method. Limiting the number of nonlinear iterations required for
+solution is critical for efficiency as nonlinear function evaluations,
+Jacobian generation, and preconditioner construction are typically
+required at every Newton iteration. For all Rayleigh numbers except
+\sn{1}{5} the number of Newton iterations required to converge is the
+same while the \sn{1}{5} result required an extra iteration. This is
+due to the fact that the convergence of the linear residual has a
+strong effect on the convergence of the nonlinear residual, much
+stronger that the effect on the convergence of the eigenvalue in the
+$SP_N$ computations.
+
+As outlined in our discussion on inexact Newton methods, depending on
+how converged the linear residual is at each step, it is possible to
+throw of the nonlinear iteration and converge on the wrong nonlinear
+solution. Although the same preconditioner was used with both the
+Newton-Krylov and FANM calculations, the convergence properties of
+MCSA and GMRES are different and we therefore expect them to produce
+different results at each nonlinear iteration which will in turn
+effect the forcing term calculation. If they are different enough, it
+is possible that another Newton iteration may be required. Even
+considering the extra Newton iteration, the solutions are converged to
+effectivly the same residual norm tolerance and produce the same
+results as provided by the
+verification.
+
+Figure~\ref{fig:ra1e5_convergence} gives the nonlinear residual
+behavior for the \sn{1}{5} Rayleigh number case. We see that the extra
+iteration comes from slightly slower quadratic convergence with FANM
+for this particular case. The difference in convergence of the linear
+problem comes from the fact that the same multigrid preconditioner
+parameters are used for this case as for the \sn{1}{3} and \sn{1}{4}
+Rayleigh number cases. As the Rayleigh number increases, the problem
+becomes more difficult to solve and therefore a more agressive
+preconditioner should maintain iterative performance of the nonlinear
+method. This was observed to be the case as for the \sn{1}{6}
+Rayleigh number case the number of multigrid cycle applications was
+bumped up to 3 from 1 used for the other cases and FANM converged in
+the same number of nonlinear iterations as the Newton-Krylov method.
 
@@
 
+\begin{figure}[htpb!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/nonlinear_problem/ra_1e5_convergence.pdf}
+  \end{center}
+  \caption{\textbf{Nonlinear residual norm as a function of Newton
+      iteration for the \sn{1}{5} Rayleigh number case.} \textit{The
+      FANM calculation required an additional nonlinear iteration to
+      converge.}}
+  \label{fig:ra1e5_convergence}
+\end{figure}
+
+In addition to nonlinear iterative performance, performance of the
+inner linear solves is also interest. Here, we seek to maintain a
+minimum number of total linear solver iterations over the course of
+all nonlinear
+iterations. Table~\ref{tab:convection_linear_iter_comparison} gives
+the total number of GMRES iterations required to converge the problem
+for the Newton-Krylov method and the total number of MCSA iterations
+for the FANM method at each Rayleigh number used. For all cases, fewer
+MCSA iterations were required than GMRES iterations to converge the
+problem even with both solver preconditioned with the same multigrid
+package and the same parameters. For the \sn{1}{6} Rayleigh number
+case this difference is most drastic with FANM requiring only 64\% of
+the linear iterations required by the Newton-Krylov scheme. Even the
+\sn{1}{5} Rayleigh number case that required more Newton iterations
+required 5 fewer MCSA iterations to converge than GMRES iterations.
+
 \begin{table}[h!]
@@
 
+Although the iterative performance of FANM is excellent for the
+thermal convection cavity problem, the explicit preconditioning
+strategy causes the same issues as was observed for the $SP_N$
+equations in the previous
+chapter. Table~\ref{tab:convection_speedup_comparison} gives the
+speedup of the Newton-Krylov method over FANM. For all Rayleigh
+numbers, we see a fairly constant speedup of O(400) when using
+Newton-Krylov over FANM meaning that the implementation of the FANM
+method used here is significantly slower. As the inexact Newton scheme
+in Drekar was modified by simply swapping out the linear solver used
+at each Newton step, this increased compute time is coming directly
+from using MCSA as the solver and the overhead due to
+preconditioning. When preconditioning overhead for the $SP_N$
+equations was considered as shown by the data in
+Figure~\ref{fig:spn_comparison_prec_time}, a similar speedup of
+approximately two orders of magnitude was noted between the Krylov
+methods and MCSA showing consistency between those results and the
+fluid benchmark results. In addition, the fairly constant speedup
+values show that FANM has effectively the same time complexity as a
+Newton-Krylov method and the implementation timings simply differ by a
+constant value.
+
 \begin{table}[h!]
@@
 
+For the lid driven cavity benchmark, performance results were again
+compared between the FANM and Newton-Krylov calculations. In this
+case, the nonlinear iterative performance results as given by
+Table~\ref{tab:driven_nonlinear_iter_comparison} show the same number
+of Newton iterations were required to converge the 100, 300, and 500
+Reynolds number cases while the FANM calculation with a Reynolds
+number of 700 actually converged in 4 fewer iterations. The nonlinear
+residual norm as a function of Newton iteration for this case is given
+in Figure~\ref{fig:re700_convergence}.
+
+This result can be explained by the fact that as given by
+Table~\ref{tab:driven_mcsa_parameters}, although the preconditioner
+parameters were the same for both calculations, the MCSA iteration was
+pushed harder by doubling the number of histories used for the
+computation at each nonlinear iteration. These extra histories
+provided better convergence of the linear model at each iteration and
+did not result in oversolving, thereby increasing the convergence rate
+of the nonlinear model as well.
+
 \begin{table}[h!]
@@
 
+\begin{figure}[htpb!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/nonlinear_problem/re700_convergence.pdf}
+  \end{center}
+  \caption{\textbf{Nonlinear residual norm as a function of Newton
+      iteration for the 700 Reynolds number case.} \textit{The FANM
+      calculation converged in 4 fewer iterations than the
+      Newton-Krylov calculation.}}
+  \label{fig:re700_convergence}
+\end{figure}
+
+The difficulty of this benchmark problem as compared to the thermal
+convection cavity problem is readily apparent in the linear iteration
+counts as given by Table~\ref{driven_linear_iter_comparison}. As the
+opposite of what was observed for that benchmark, lid driven cavity
+results show that the linear models were significantly more
+ill-conditioned, requiring many more MCSA and GMRES iterations. For
+all cases except the 700 Reynolds number case many more MCSA
+iterations were required to converge with 64\% as many GMRES
+iterations required as MCSA iterations for the 100 Reynolds number
+case. For the 700 Reynolds number case, fewer MCSA iterations were
+observed due to the fact that 4 fewer nonlinear iterations were
+performed. The average number of linear iterations per nonlinear
+iterations was still much larger for the FANM calculations.
+
+These iterative results again point out the fundamental flaw of
+requiring a spectral radius of less than unity for the MCSA iteration
+to converge at each Newton step. As the problem becomes more dominated
+by inertial forces, the resulting linear models become increasingly
+stiff and difficult to precondition. As a result, convergence requires
+many more iterations that need for the flow problem driven by
+convection in the previous benchmark if convergence is achieved at
+all.
+
 \begin{table}[h!]
@@
 
+As a final comparison, the timing of both methods was again compared
+with the speedup of using the Newton-Krylov solver over FANM reported
+in Table~\ref{tab:driven_speedup_comparison}. The speedup values
+reported here are of the same magnitude as those reported for the
+thermal convection cavity problem and in agreement with those observed
+for the $SP_N$ calculations. In all cases, the preconditioning
+strategy must be signficantly improved to avoid these large overheads.
+
 \begin{table}[h!]

### 2013-07-17 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
cleaning up nonlinear chapter

@@
 up the Reynolds number of the system making it more difficult to
-solve. More rotation of the fluid is induced by intertial forces and
+solve. More rotation of the fluid is induced by inertial forces and
 we begin to see some additional vortices form besides the primary
@@
 method, a common technique for preconditioning these equations
-\citep{ghia_high-re_1982,evans_enhanced_2007}. For FANM, this
-preconditioning was required to reduce the spectral radius of the
-Jacobian operators generated at each Newton step and for the
-Newton-Krylov solutions this preconditioning was required to obtain
-good convergence properties although not necessary for
+\citep{ghia_high-re_1982,evans_enhanced_2007}. Without multigrid
+preconditioning, the residual in MCSA was observed to
+stagnate. Typically this happens because the Richardson iteration will
+damp the higher order error modes on the scale of the computational
+grid used to solve the problem but will miss lower order error
+modes. In terms of Monte Carlo, this stagnation manifests itself in
+histories that simply do no terminate and therefore no solution is
+acquired.
+
+For FANM, this preconditioning was required to reduce the spectral
+radius of the Jacobian operators generated at each Newton step and for
+the Newton-Krylov solutions this preconditioning was required to
+obtain good convergence properties although it was not necessary for
 convergence. For every benchmark calculation, both FANM and
@@
 such that the conditioning of the system would not be a factor in the
-verification of the method or the perfomance comparison in the
-following section.
-
-As with the radiation transport calculations in the previous chapter,
-algrebraic preconditioning with the multigrid method results in dense
-linear systems. To mitigate this, the reduced domain approximation was
-again applied to the Monte Carlo solution within the MCSA solve
-occuring at each nonlinear iteration. The values used for the
-approximation will be given for each benchmark problem. Without
-multigrid preconditioning or too small of a reduced domain fill level,
-the residual in MCSA was observed to stagnate. Typically this happens
-because the Richardson iteration will damp the higher order error
-modes on the scale of the computational grid used to solve the problem
-but will miss lower order error modes. In terms of Monte Carlo, this
-stagnation manifests itself in histories that simply do no terminate
-and therefore no solution is acquired.
+verification of the method or the performance comparison in the
+following section. As with the radiation transport calculations in the
+previous chapter, algrebraic preconditioning with the multigrid method
+results in dense linear systems. To mitigate this, the reduced domain
+approximation was again applied to the Monte Carlo solution within the
+MCSA solve occuring at each nonlinear iteration. The values used for
+the approximation will be given for each benchmark problem.
 
@@
 for the verification of MCSA for the $SP_N$ solutions in the previous
-chapter was also used here as a means of comparison.
+chapter was also used here as a means of comparison and FANM
+benchmarking.
 
@@
 both the Newton-Krylov solver and a FANM solver for Rayleigh numbers
-of \sn{1}{3}, \sn{1}{4}, \sn{1}{5} and
-\sn{1}{6}. Table~\ref{tab:thermal_convection_parameters} gives the
+of \sn{1}{3}, \sn{1}{4}, \sn{1}{5} and \sn{1}{6} on a $50 \times 50$
+square grid. Table~\ref{tab:thermal_convection_parameters} gives the
 physical parameters of the problem and
@@
 values. No variation in solution was noted for each of the
-computations within the tolerance of the converged nonlinear residual
-norms given in
+computations larger than the tolerance of the converged nonlinear
+residual norms given in
 Table~\ref{tab:convection_residual_norm_comparison}. Variations in the
@@
 both the Newton-Krylov solver and a FANM solver for Reynolds numbers
-of 100, 300, 500, and
-700. Table~\ref{tab:thermal_convection_parameters} gives the physical
+of 100, 300, 500, and 700 on a $50 \times 50$ square
+grid. Table~\ref{tab:thermal_convection_parameters} gives the physical
 parameters of the problem and
@@
 extremely prohibitive in this case and others where transport driven
-by intertial forces is dominant. Because of this, only solutions to
+by inertial forces is dominant. Because of this, only solutions to
 the lid driven cavity benchmark at Reynolds numbers of up to 700 are
@@
 Using the benchmarks from the verification, we now compare the
-performance of the FANM method against a conventional Krylov
+performance of the FANM method against a conventional Newton-Krylov
 method. For each benchmark, we will analyze the iterative performance
@@
 \sn{1}{5} the number of Newton iterations required to converge is the
-same while the \sn{1}{5} result required an extra iteration. This is
-due to the fact that the convergence of the linear residual has a
+same while the \sn{1}{5} result required an extra FANM iteration. This
+is due to the fact that the convergence of the linear residual has a
 strong effect on the convergence of the nonlinear residual, much
-stronger that the effect on the convergence of the eigenvalue in the
-$SP_N$ computations.
+stronger than the effect on the convergence of the eigenvalue in the
+$SP_N$ computations where we observed that identical numbers of
+eigenvalue iterations were required to converge each variation of the
+probem.
 
@@
 how converged the linear residual is at each step, it is possible to
-throw of the nonlinear iteration and converge on the wrong nonlinear
-solution. Although the same preconditioner was used with both the
+throw off the nonlinear iteration and decrease iterative
+performance. Although the same preconditioner was used with both the
 Newton-Krylov and FANM calculations, the convergence properties of
@@
 different results at each nonlinear iteration which will in turn
-effect the forcing term calculation. If they are different enough, it
+affect the forcing term calculation. If they are different enough, it
 is possible that another Newton iteration may be required. Even
 considering the extra Newton iteration, the solutions are converged to
-effectivly the same residual norm tolerance and produce the same
-results as provided by the
-verification.
+effectively the same residual norm tolerance and produce the same
+results as provided by the verification.
 
 Figure~\ref{fig:ra1e5_convergence} gives the nonlinear residual
-behavior for the \sn{1}{5} Rayleigh number case. We see that the extra
-iteration comes from slightly slower quadratic convergence with FANM
-for this particular case. The difference in convergence of the linear
-problem comes from the fact that the same multigrid preconditioner
-parameters are used for this case as for the \sn{1}{3} and \sn{1}{4}
-Rayleigh number cases. As the Rayleigh number increases, the problem
-becomes more difficult to solve and therefore a more agressive
-preconditioner should maintain iterative performance of the nonlinear
-method. This was observed to be the case as for the \sn{1}{6}
-Rayleigh number case the number of multigrid cycle applications was
-bumped up to 3 from 1 used for the other cases and FANM converged in
-the same number of nonlinear iterations as the Newton-Krylov method.
+behavior for the \sn{1}{5} Rayleigh number case that required the
+extra FANM iteration. We see that the extra iteration comes from
+slightly slower quadratic convergence with FANM for this particular
+case. The difference in convergence of the nonlinear problem comes
+from the fact that the same multigrid preconditioner parameters are
+used for this case as for the \sn{1}{3} and \sn{1}{4} Rayleigh number
+cases for the linear solutions. Using the same preconditioner for the
+harder \sn{1}{5} case means that the problem is still growing in
+difficulty even though it is preconditioned. The more ill-conditioned
+the linear model is, the closer the linear residual is converged to
+the requested tolerance whereas a better conditioned problem may
+produce a linear solution that overshoots the specified tolerance
+provided by the forcing term. When oversolving is not an issue, this
+results in faster convergence of the nonlinear problem.
+
+For the \sn{1}{5} Rayleigh number case, this means that the GMRES
+solver at each Newton step is converging the linear residual to a
+smaller tolerance than the MCSA solver, even though they have
+approximately the same forcing terms to begin with. As the Rayleigh
+number increases, the problem becomes more difficult to solve and
+therefore a more agressive preconditioner should maintain iterative
+performance of the nonlinear method. This was observed to be true for
+the \sn{1}{6} Rayleigh number case. For this calculation, the number
+of multigrid cycle applications was bumped up to 3 from 1 and FANM
+converged in the same number of nonlinear iterations as the
+Newton-Krylov method.
 
@@
 MCSA iterations were required than GMRES iterations to converge the
-problem even with both solver preconditioned with the same multigrid
+problem even with both solvers preconditioned with the same multigrid
 package and the same parameters. For the \sn{1}{6} Rayleigh number
@@
 fluid benchmark results. In addition, the fairly constant speedup
-values show that FANM has effectively the same time complexity as a
-Newton-Krylov method and the implementation timings simply differ by a
-constant value.
+values as a function of the primary problem parameter show that FANM
+has effectively the same time complexity as a Newton-Krylov method and
+the implementation timings simply differ by a constant value.
 
@@
 case, the nonlinear iterative performance results as given by
-Table~\ref{tab:driven_nonlinear_iter_comparison} show the same number
-of Newton iterations were required to converge the 100, 300, and 500
-Reynolds number cases while the FANM calculation with a Reynolds
-number of 700 actually converged in 4 fewer iterations. The nonlinear
-residual norm as a function of Newton iteration for this case is given
-in Figure~\ref{fig:re700_convergence}.
+Table~\ref{tab:driven_nonlinear_iter_comparison} show that the same
+number of Newton iterations were required to converge the 100, 300,
+and 500 Reynolds number cases while the FANM calculation with a
+Reynolds number of 700 actually converged in 4 fewer iterations. The
+nonlinear residual norm as a function of Newton iteration for this
+case is given in Figure~\ref{fig:re700_convergence}.
 
@@
 convection cavity problem is readily apparent in the linear iteration
-counts as given by Table~\ref{driven_linear_iter_comparison}. As the
-opposite of what was observed for that benchmark, lid driven cavity
-results show that the linear models were significantly more
+counts as given by Table~\ref{tab:driven_linear_iter_comparison}. As
+the opposite of what was observed for that benchmark, lid driven
+cavity results show that the linear models were significantly more
 ill-conditioned, requiring many more MCSA and GMRES iterations. For
-all cases except the 700 Reynolds number case many more MCSA
-iterations were required to converge with 64\% as many GMRES
-iterations required as MCSA iterations for the 100 Reynolds number
-case. For the 700 Reynolds number case, fewer MCSA iterations were
-observed due to the fact that 4 fewer nonlinear iterations were
+all cases except the 700 Reynolds number case, many more MCSA
+iterations were required to converge with the observed GMRES iteration
+count 64\% of that observed for MCSA when looking at the 100 Reynolds
+number case. For the 700 Reynolds number case, fewer MCSA iterations
+were observed due to the fact that 4 fewer nonlinear iterations were
 performed. The average number of linear iterations per nonlinear
-iterations was still much larger for the FANM calculations.
+iteration was still much larger for the FANM calculations.
 
@@
 stiff and difficult to precondition. As a result, convergence requires
-many more iterations that need for the flow problem driven by
+many more iterations than needed for the flow problem driven by
 convection in the previous benchmark if convergence is achieved at

### 2013-07-18 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
updating strong scaling figures for new multiple set reduction

@@
 case operates at 38\% efficiency while at the same number of cores,
-the 2 set case with history replication operates at 62\%
+the 2 set case with history replication operates at 58\%
 efficiency. This is reasonable improvement in the use of those

### 2013-07-18 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
updating figure captions in subdomain monte carlo section

@@
   \caption{\textbf{Strong scaling total wall time with multiple sets
-      and subdomain Monte Carlo.} \textit{The subdomain Monte Carlo
-      solver improves the parallel efficiency and therefore time to
-      solution by eliminating all parallel communication during the
-      Monte Carlo solve.}}
+      and subdomain Monte Carlo.} \textit{Adding multiple sets has in
+      general a negative effect on the wall time as was observed for
+      the case with full Monte Carlo transport.}}
   \label{fig:titan_strong_subdomain_ms_time}
@@
   \caption{\textbf{Strong scaling absolute efficiency for with
-      multiple sets and subdomain Monte Carlo.} \textit{The subdomain
-      Monte Carlo solver improves the parallel efficiency and
-      therefore time to solution by eliminating all parallel
-      communication during the Monte Carlo solve.}}
+      multiple sets and subdomain Monte Carlo.} \textit{Increased wall
+      times from adding parallel reduction operations for multiple
+      sets reduces parallel scalability at lower core counts but gives
+      a higher efficiency at the strong scaling wall.}}
   \label{fig:titan_strong_subdomain_ms}
@@
   \end{center}
-  \caption{\textbf{Weak scaling total wall time with multiple sets
-      and subdomain Monte Carlo.} \textit{The subdomain Monte Carlo
-      solver improves the parallel efficiency and therefore time to
-      solution by eliminating all parallel communication during the
-      Monte Carlo solve.}}
+  \caption{\textbf{Weak scaling total wall time with multiple sets and
+      subdomain Monte Carlo.} \textit{The addition of the set tally
+      reduction is significant with cost of the reduction observed as
+      core count grows.}}
   \label{fig:titan_weak_subdomain_ms_time}
@@
   \end{center}
-  \caption{\textbf{Weak scaling absolute efficiency for with
-      multiple sets and subdomain Monte Carlo.} \textit{The subdomain
-      Monte Carlo solver improves the parallel efficiency and
-      therefore time to solution by eliminating all parallel
-      communication during the Monte Carlo solve.}}
+  \caption{\textbf{Weak scaling absolute efficiency for with multiple
+      sets and subdomain Monte Carlo.} \textit{Weak scaling
+      efficiencies are not improved at large core counts when multiple
+      sets are added.}}
   \label{fig:titan_weak_subdomain_ms}

### 2013-07-18 doc/dissertation/document/chapters/mc_background/mc_background.tex
updating section numbering
adding a summary to each chapter

@@
 
-\subsubsection{Alternative Fixed Point Iterations}
+\subsection{Alternative Fixed Point Iterations}
 \label{subsubsec:alternative_fixed_point}
@@
 \clearpage
+
+%%---------------------------------------------------------------------------%%
+\section{Summary}
+\label{sec:mc_summary}
+
+In this chapter, the fundamentals of Monte Carlo synthetic
+acceleration methods were presented and explored in the context of a
+simple model transport system. The following are the signficant
+observations and findings.
+
+\begin{itemize}
+\item MCSA has been generalized for all discrete linear transport
+  problems
+\item MCSA can be preconditioned using a general left/right scheme but
+  requires the explicit inversion of the preconditioners and
+  construction of the composite linear operator
+\item The adjoint Neumann-Ulam method performs better when used with
+  MCSA compared to the forward method
+\item MCSA has superior performance to Halton's residual Monte Carlo
+  method
+\item The Richardson iteration in MCSA behaves as a smoother for the
+  stochastic Monte Carlo error
+\item MCSA iterative performance when used with the expected value
+  estimator is insensitive to histories per iteration
+\item MCSA iterative performance is insensitive to the stochastic
+  history weight cutoff
+\item Variance reduction with artificial absorption can enhance time
+  to solution
+\item The reduced domain approximation was introduced to mitigate the
+  consequences of composite linear operator generation
+\end{itemize}

### 2013-07-18 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
adding a summary to each chapter

@@
 conventional methods. These sets of equations characterize momentum
-and energy transport through diffusion, viscous forces, and inertial
-forces. Traditionally, such systems have been solved by linearizing
-them in a form where the nonlinearities in the variables are
-eliminated and more traditional linear methods can be used for
+and energy transport through diffusion, viscous forces and convection,
+and inertial forces. Traditionally, such systems have been solved by
+linearizing them in a form where the nonlinearities in the variables
+are eliminated and more traditional linear methods can be used for
 solutions. Often characterized as segregated methods where physics
@@
 \end{table}
+
+%%---------------------------------------------------------------------------%%
+\section{Summary}
+\label{sec:nonlinear_summary}
+
+In this chapter we have developed and explored Monte Carlo synthetic
+acceleration methods in the context of solutions to the nonlinear
+Navier-Stokes equations. The following are the signficant observations
+and findings.
+
+\begin{itemize}
+\item Forward-Automated Newton-MCSA (FANM), a new inexact Newton
+  method based on Monte Carlo synthetic acceleration, has been
+  developed
+\item The FANM method has been incorporated into the Drekar production
+  multiphysics code
+\item The FANM method has been verified to produce the same solutions
+  as a production Newton-Krylov method for two benchmark problems for
+  the Navier-Stokes equations in different flow regimes
+\item The FANM method has better iterative performance than the
+  Newton-Krylov method for convection dominated problems, converging
+  in fewer linear solver iterations with the same preconditioning for
+  high and lower Rayleigh numbers
+\item The Newton-Krylov method has better iterative performance than
+  the FANM method for flow dominated by interial forces, converging in
+  fewer linear solver iterations with the same preconditioning
+\item The spectral radius convergence restriction on MCSA was observed
+  to be a signficant hinderance by preventing solutions to forced flow
+  problems at high Reynolds numbers
+\item Explict algebraic preconditioning of MCSA creates FANM run times
+  $O(100)$ slower than the Newton-Krylov solver
+\item More Monte Carlo histories at every FANM iteration can reduce
+  the number of linear and nonlinear iterations required to converge
+  the problem
+\end{itemize}

### 2013-07-18 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
adding a summary to each chapter

@@
 \end{figure}
+
+\clearpage
+
+%%---------------------------------------------------------------------------%%
+\section{Summary}
+\label{sec:parallel_summary}
+
+In this chapter we have developed a domain-decomposed parallel
+algorithm for Monte Carlo synthetic acceleration and explored the
+qualities of the algorithm through parallel scaling studies. The
+following are the signficant observations and findings.
+
+\begin{itemize}
+\item MCSA can be effectively parallelized using domain decomposition
+\item The leakage fraction of stochastic histories from a parallel
+  subdomain can be quantified analytically using algebraic quantities
+\item The multiple-set overlapping-domain (MSOD) parallel algorithm
+  for particle transport has been adapted to parallelize MCSA
+\item MCSA parallelized with MSOD was verified to produce the same
+  results for solutions to the neutron diffusion equation as
+  production Krylov methods
+\item Strong scaling studies were performed with up to 65,536 cores
+  and weak scaling studies were performed with up to 131,072 cores
+\item MCSA scales comparably to production Krylov methods on the Titan
+  Cray XK7 leadership class machine
+\item Overlap in small quantities determined analytically by random
+  walk length can provide parallel efficiency boosts of a few percent
+  in strong scaling cases but is ineffective in weak scaling cases
+\item Multiple sets will never boost parallel efficienicies in both
+  weak and strong scaling cases due to additional communication
+  overhead
+\item Multiple sets offers a means to reduce time to solution by
+  solving a multiple copies of the original problem and combining the
+  solutions using superposition
+\item MCSA is most efficient used in parallel as a stochastic
+  realization of an additive Schwarz method
+\end{itemize}

### 2013-07-18 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding a summary to each chapter

@@
 preconditioning scheme presented here.
+
+\clearpage
+
+%%---------------------------------------------------------------------------%%
+\section{Summary}
+\label{sec:mc_summary}
+
+In this chapter, the Monte Carlo synthetic acceleration method has
+been applied to the $SP_N$ discretization of the neutron transport
+equation and explored in the context of a difficult nuclear fuel
+assembly criticality problem. The following are the signficant
+observations and findings.
+
+\begin{itemize}
+\item MCSA can be used to solve the asymmetric system generated by the
+  multigroup $SP_N$ equations
+\item Simple Jacobi-based preconditioning reduces the spectral radius
+  of the $SP_N$ system below unity, however, this is not sufficient
+  for MCSA convergence
+\item As the spectral radius approaches unity, MCSA was demonstrated
+  to break down with more stochastic histories required for
+  convergence and more CPU time required per history
+\item Advanced algebraic preconditioning strategies were applied to
+  the $SP_N$ equations with ILUT chosen for subsequent investigations
+\item The reduced domain approximation was required to mitigate the
+  memory constraints of the explicit preconditioning strategy
+\item MCSA relaxation parameters were developed an observed to enhance
+  the time to solution for the fuel assembly problem
+\item MCSA was verified to produce the same flux distribution and
+  k-eigenvalue for the fuel assembly as production Krylov methods
+\item MCSA was observed to converge in fewer iterations per eigenvalue
+  iteration than GMRES for the fuel assembly problem and more then
+  Bi-CGStab
+\item The explicit preconditioning strategy creates MCSA runtimes
+  $O(100)$ larger than the production Krylov methods
+\end{itemize}

### 2013-07-18 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
adding more to chapter summaries

@@
   developed
-\item The FANM method has been incorporated into the Drekar production
-  multiphysics code
+\item The FANM method has been incorporated into the Drekar
+  multiphysics production code base developed at Sandia National
+  Laboratories
 \item The FANM method has been verified to produce the same solutions
@@
   in fewer linear solver iterations with the same preconditioning for
-  high and lower Rayleigh numbers
+  high and low Rayleigh numbers
 \item The Newton-Krylov method has better iterative performance than
@@
   the problem
+\item With future optimization, reactor accident scenario analyses
+  that rely on convection dominated transport could experience
+  improved time to solution with FANM given the improved iterative
+  performance for problems of this type
 \end{itemize}

### 2013-07-18 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
adding more to chapter summaries

@@
 \item Strong scaling studies were performed with up to 65,536 cores
-  and weak scaling studies were performed with up to 131,072 cores
-\item MCSA scales comparably to production Krylov methods on the Titan
-  Cray XK7 leadership class machine
+  and weak scaling studies were performed with up to 131,072 cores on
+  the Titan Cray XK7 leadership class machine at Oak Ridge National
+  Laboratory
+\item MCSA scales favorably compared to production Krylov methods for
+  both strong and weak scaling cases
 \item Overlap in small quantities determined analytically by random
@@
   weak and strong scaling cases due to additional communication
-  overhead
+  overhead from added reduction operations
 \item Multiple sets offers a means to reduce time to solution by
@@
   realization of an additive Schwarz method
+\item With future optimization, large scale reactor analyses in which
+  the transport operator is fully formed (e.g. the $SP_N$ equations)
+  could experience improved time to solution on leadership-class
+  machines with MCSA given the improved scaling performance
 \end{itemize}

### 2013-07-18 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding more to chapter summaries

@@
   multigroup $SP_N$ equations
+\item MCSA has been incorporated into the Exnihilo neutronics
+  production code base developed at Oak Ridge National Laboratory
 \item Simple Jacobi-based preconditioning reduces the spectral radius
@@
 \item Advanced algebraic preconditioning strategies were applied to
-  the $SP_N$ equations with ILUT chosen for subsequent investigations
+  the $SP_N$ equations to obtain convergence with ILUT chosen for
+  subsequent investigations
 \item The reduced domain approximation was required to mitigate the
-  memory constraints of the explicit preconditioning strategy
+  memory and performance constraints of the explicit preconditioning
+  strategy although both remain significant obstacles
 \item MCSA relaxation parameters were developed an observed to enhance
@@
 \item The explicit preconditioning strategy creates MCSA runtimes
-  $O(100)$ larger than the production Krylov methods
+  $O(100)$ larger than the production Krylov methods for the fuel
+  assembly problem
 \end{itemize}

### 2013-07-18 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
editing nonlinear chapter

@@
 all of the components necessary for a Monte Carlo solver to provide
-value.
-
-We therefore devise a new nonlinear method based on the MCSA algorithm
-and Newton's method. In this chapter the fundamentals of Newton's
-method will be presented along with a brief discussion of inexact
-Newton methods. Newton-Krylov methods, a variant of inexact Newton
-methods, will be outlined along with two common techniques for
-Jacobian generation as a means of defining current
-practices. Following this, the new algorithm will be presented and
-implementation details will be discussed. Using the new algorithm and
-a production fluid physics code, a group of challenging nonlinear
-benchmark problems for the Navier-Stokes equations will be used to
-verify the correctness of the method relative to solutions obtained
-using conventional methods in various flow regimes. Finally, we will
-compare the performance of the new method against contemporary
+value. We therefore devise a new nonlinear method based on the MCSA algorithm
+and Newton's method.
+
+In this chapter the fundamentals of Newton's method will be presented
+along with a brief discussion of inexact Newton methods. Newton-Krylov
+methods, a variant of inexact Newton methods, will be outlined along
+with two common techniques for Jacobian generation as a means of
+defining current practices. Following this, the new algorithm will be
+presented and implementation details will be discussed. Using the new
+algorithm and a production fluid physics code, a group of challenging
+nonlinear benchmark problems for the Navier-Stokes equations will be
+used to verify the correctness of the method relative to solutions
+obtained using conventional methods in various flow regimes. Finally,
+we will compare the performance of the new method against contemporary
 Newton-Krylov methods for the same benchmark problems.
@@
 can generate this procedure by building the Taylor expansion of the
-$k+1$ iterate of the nonlinear residuals about the previous $k$
-iterate:
+$k+1$ iterate, $\ve{F}(\ve{u}^k)$, of the nonlinear residuals about
+the previous $k$ iterate, $\ve{F}(\ve{u}^{k+1})$:
 \begin{equation}
@@
 \end{subequations}
+
 There are then three distinct steps to perform: evaluation of the
@@
 techniques for improving performance of the nonlinear iteration may be
-used with FANM including backtracking methods.
+used with FANM including globalization methods.
 
@@
 tuned to enhance the nonlinearities. We will then apply these
-equations to the following two standard benchmark problems.
+equations to the following two standard benchmark problems that cover
+convection dominated flow and flow dominated by intertial forces.
 
@@
 \label{subsec:lid_driven_cavity}
-As an extension of the convection problem, the second benchmark
-problem given by Ghia \citep{ghia_high-re_1982} adds a driver for the
-flow to introduce higher Reynolds numbers into the system, providing
-more inertial force to overcome the viscous forces in the fluid. The
-setup for this problem is equally simple, containing only the
-Dirichlet conditions as given in Figure~\ref{fig:lid_driven_cavity}
-and is only applied to the mass and momentum equations on the unit
-square.
+To study performance for flow driven by intertial forces, the second
+benchmark problem given by Ghia \citep{ghia_high-re_1982} incorporates
+a driver for the flow to introduce higher Reynolds numbers into the
+system, providing more inertial force to overcome the viscous forces
+in the fluid. The setup for this problem is equally simple, containing
+only the Dirichlet conditions as given in
+Figure~\ref{fig:lid_driven_cavity} and is only applied to the mass and
+momentum equations on the unit square.
 
@@
 up the Reynolds number of the system making it more difficult to
-solve. More rotation of the fluid is induced by inertial forces and
-we begin to see some additional vortices form besides the primary
-vortex in the center of the box.
+solve. More rotation of the fluid is induced by inertial forces and we
+begin to see some additional vortices form besides the primary vortex
+near the center of the box.
 
@@
 verification of the method or the performance comparison in the
-following section. As with the radiation transport calculations in the
+following section. In addition, the same nonlinear solver parameters
+including globalization and forcing term strategies were used with
+both methods. As with the radiation transport calculations in the
 previous chapter, algrebraic preconditioning with the multigrid method
@@
       Specific Heat & 1.0 & J / $(kg \times K)$ \\
-      Dynamic Viscosity & 0.71 & $N \times s / m^2$ \\
+      Dynamic Viscosity & 0.71 & $(N \times s) / m^2$ \\
       Thermal Conductivity & 1.0 & $W / (m \times K)$ \\
@@
       Density & 1.0 & $kg / m^3$ \\
-      Dynamic Viscosity & 0.71 & $N \times s / m^2$ \\
+      Dynamic Viscosity & 0.71 & $(N \times s) / m^2$ \\
       Kinematic Viscosity & 0.71 & $m^2 / s$ \\
@@
 is due to the fact that the convergence of the linear residual has a
-strong effect on the convergence of the nonlinear residual, much
-stronger than the effect on the convergence of the eigenvalue in the
-$SP_N$ computations where we observed that identical numbers of
-eigenvalue iterations were required to converge each variation of the
-probem.
+strong effect on the convergence of the nonlinear residual. For
+nonlinear calculations, this is much stronger than the effect on the
+convergence of the eigenvalue in the $SP_N$ computations where we
+always observed that identical numbers of k-eigenvalue iterations were
+required to converge each variation of the probem.
 
@@
 harder \sn{1}{5} case means that the problem is still growing in
-difficulty even though it is preconditioned. The more ill-conditioned
-the linear model is, the closer the linear residual is converged to
-the requested tolerance whereas a better conditioned problem may
-produce a linear solution that overshoots the specified tolerance
-provided by the forcing term. When oversolving is not an issue, this
-results in faster convergence of the nonlinear problem.
+difficulty even though it is preconditioned in an identical
+manner. The more ill-conditioned the linear model is, the closer the
+linear residual is converged to the requested tolerance whereas a
+better conditioned problem may produce a linear solution that
+overshoots the specified tolerance provided by the forcing term. When
+oversolving the linear model is not an issue, this results in faster
+convergence of the nonlinear problem.
 
@@
 In addition to nonlinear iterative performance, performance of the
-inner linear solves is also interest. Here, we seek to maintain a
+inner linear solves is also of interest. Here, we seek to maintain a
 minimum number of total linear solver iterations over the course of
-all nonlinear
-iterations. Table~\ref{tab:convection_linear_iter_comparison} gives
+all nonlinear iterations in order to limit matrix-vector multiplies
+and other costly
+operations. Table~\ref{tab:convection_linear_iter_comparison} gives
 the total number of GMRES iterations required to converge the problem
@@
 \sn{1}{5} Rayleigh number case that required more Newton iterations
-required 5 fewer MCSA iterations to converge than GMRES iterations.
+required 5 fewer MCSA iterations to converge than GMRES
+iterations. If the performance of the MCSA implementation used here
+were to be potentially optimized in the future, this improved
+iterative performance could also mean improved time to solution.
 
@@
 at each Newton step, this increased compute time is coming directly
-from using MCSA as the solver and the overhead due to
+from using MCSA as the linear solver and the overhead due to
 preconditioning. When preconditioning overhead for the $SP_N$
@@
 methods and MCSA showing consistency between those results and the
-fluid benchmark results. In addition, the fairly constant speedup
-values as a function of the primary problem parameter show that FANM
-has effectively the same time complexity as a Newton-Krylov method and
-the implementation timings simply differ by a constant value.
+convection cavity benchmark results presented. In addition, the fairly
+constant speedup values as a function of the primary problem parameter
+show that FANM has effectively the same time complexity as a
+Newton-Krylov method and the implementation timings simply differ by a
+constant value, also consistent with the $SP_N$ observations.
 
@@
 did not result in oversolving, thereby increasing the convergence rate
-of the nonlinear model as well.
+of the nonlinear model as well. This result is significant in that it
+shows increasing the number of histories used for each MCSA iteration
+can improve the performance of both linear and nonlinear convergence.
 
@@
 convection in the previous benchmark if convergence is achieved at
-all.
+all. Much of this can be attributed to the physics at play in each of
+these benchmarks. For the thermal convection cavity benchmark, the
+forces at work in that problem were more effectively resolved by the
+Monte Carlo method as they are more diffusive in character than the
+flow driven by intertial forces in the lid driven cavity problem. This
+physics-based observation was also true for the $SP_N$ equations where
+the diffusion dominated physics in that case gave better convergence
+results for MCSA than GMRES.
 
@@
 reported here are of the same magnitude as those reported for the
-thermal convection cavity problem and in agreement with those observed
-for the $SP_N$ calculations. In all cases, the preconditioning
-strategy must be signficantly improved to avoid these large overheads.
+thermal convection cavity problem and again in agreement with those
+observed for the $SP_N$ calculations. Lack of signficant variation in
+these timing results also point to a qualititively similar time
+complexity for the FANM method as a function of primary problem
+parameter relative to the Newton-Krylov method. In all cases, the
+preconditioning strategy must be signficantly improved along with
+implementation optimization to avoid these large overheads.
 
@@
   the FANM method for flow dominated by interial forces, converging in
-  fewer linear solver iterations with the same preconditioning
+  fewer linear solver iterations with the same preconditioning for
+  high and low Reynolds numbers
 \item The spectral radius convergence restriction on MCSA was observed

### 2013-07-18 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
editing nonlinear chapter

@@
 \item Multiple sets offers a means to reduce time to solution by
-  solving a multiple copies of the original problem and combining the
+  solving multiple copies of the original problem and combining the
   solutions using superposition
-\item MCSA is most efficient used in parallel as a stochastic
+\item MCSA is most efficiently used in parallel as a stochastic
   realization of an additive Schwarz method

### 2013-07-18 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
editing nonlinear chapter

@@
   convergence and more CPU time required per history
+\item Light water reactor problems are difficult to solve with MCSA as
+  they have large spectral radii due to the neutron scattering in the
+  moderator
 \item Advanced algebraic preconditioning strategies were applied to
@@
   strategy although both remain significant obstacles
-\item MCSA relaxation parameters were developed an observed to enhance
-  the time to solution for the fuel assembly problem
+\item MCSA relaxation parameters were developed and observed to
+  enhance the time to solution for the fuel assembly criticality
+  problem
 \item MCSA was verified to produce the same flux distribution and
@@
 \item MCSA was observed to converge in fewer iterations per eigenvalue
-  iteration than GMRES for the fuel assembly problem and more then
-  Bi-CGStab
+  iteration than GMRES for the fuel assembly criticality problem and
+  more then Bi-CGStab
 \item The explicit preconditioning strategy creates MCSA runtimes
   $O(100)$ larger than the production Krylov methods for the fuel
-  assembly problem
+  assembly criticality problem
 \end{itemize}

### 2013-07-19 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
adding fanm output for step at re200

@@
 tuned to enhance the nonlinearities. We will then apply these
-equations to the following two standard benchmark problems that cover
-convection dominated flow and flow dominated by intertial forces.
+equations to the following three standard benchmark problems that
+cover convection dominated flow and flow dominated by intertial
+forces in two different geometries.
 
@@
       of 0.05 for each calculation.}}
+  \label{fig:driven_velocity_isocurves}
+\end{figure}
+
+\clearpage
+
+\subsubsection{Backward-Facing Step Problem}
+\label{subsubsec:backward_facing_step}
+The third benchmark was generated by Gartling in 1990 and consists of
+both flow over a backward step and an outflow boundary condition
+\citep{gartling_test_1990}. Using the mass and momentum equations
+while neglecting the energy equation, this problem utilizes a longer
+domain with a 1/25 aspect ratio used for this work\footnote{The
+  literature on this benchmark suggests using a domain with a 1/30
+  aspect ratio to reduce the effect of the outflow boundary condition
+  on the upstream recirculation zones. Using a 1/25 aspect ratio for
+  this work was not observed to disturb the recirculation zones and
+  was therefore used to reduce memory pressure on the FANM solver.}
+with the boundary conditions as shown in
+Figure~\ref{fig:backward_facing_step}.
+
+\begin{figure}[htpb!]
+  \begin{center}
+    \scalebox{1.2}{
+      \input{chapters/nonlinear_problem/backward_facing_step.pdftex_t} }
+  \end{center}
+  \caption{\textbf{Problem setup for the backward facing step
+      benchmark.} \textit{Zero velocity boundary conditions are
+      applied at the top and bottom of the domain while the outflow
+      boundary condition on the right boundary is represented by zero
+      stress tensor components in the direction of the flow. For the
+      inlet conditions, the left boundary is split such that the top
+      half has a fully formed parabolic flow profile and the bottom
+      half has a zero velocity condition, simulating flow over a
+      step.}}
+  \label{fig:backward_facing_step}
+\end{figure}
+
+In this problem, the inflow condition is specified by a fully-formed
+parabolic flow profile over a zero velocity boundary representing a
+step. The flow over this step will generate a recirculating backflow
+under the inlet flow towards the step and at higher Reynolds numbers a
+recirculating zone will form farther downstream from the step. As in
+the lid driven cavity problem, the nonlinear behavior of this
+benchmark and the difficulty in obtaining a solution is dictated by
+the Reynolds number of the fluid. In Shadid's work, a Reynolds numbers
+up to 800 were used.
+
+Compared to the lid driven cavity benchmark, the flow over a backward
+step is much more representative of flow driven by inertial forces in
+a nuclear reactor that may encounter several obstacles along the flow
+path. For example, spacer grids in fuel assemblies or mixing vanes
+designed to induce turbulence and improve heat transfer will create
+recirculation zones much like those observed for this benchmark
+problem. Figure~\ref{fig:step_velocity_isocurves} gives the velocity
+magnitude contours for solutions to this problem with Reynolds numbers
+of 200, 400, 600, and 800.
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/nonlinear_problem/driven_velocity_isocurves.png}
+  \end{center}
+  \caption{\textbf{Velocity magnitude isocurves from solution to the
+      thermal convection cavity problem.} \textit{From top to bottom:
+      Re = 200, Re = 400, Re = 600, Re = 800. Isocurves on unit
+      velocity scale from 0 to 1 with divisions of 0.05 for each
+      calculation.}}
   \label{fig:driven_velocity_isocurves}
@@
 of 100, 300, 500, and 700 on a $50 \times 50$ square
-grid. Table~\ref{tab:thermal_convection_parameters} gives the physical
+grid. Table~\ref{tab:lid_driven_parameters} gives the physical
 parameters of the problem and
-Table~\ref{tab:convection_mcsa_parameters} gives the MCSA solver
+Table~\ref{tab:driven_mcsa_parameters} gives the MCSA solver
 parameters used within the FANM method. Variation in the Reynolds
 number of the system was achieved by keeping the parameters given in
-Table~\ref{tab:thermal_convection_parameters} fixed and varying the
+Table~\ref{tab:lid_driven_parameters} fixed and varying the
 fluid velocity in the $x$ direction on the top boundary. In addition
@@
       Kinematic Viscosity & 0.71 & $m^2 / s$ \\
-      Gravitational Acceleration & 10 & $m / s^2$ \\
       Cavity Characteristic Length & 1.0 & $m$ \\
@@
   \end{center}
-  \caption{\textbf{Lid driven cavity FANM verification for a
-      Reynolds number of 500.} \textit{The values reported are
-      global minima, maxima, and averages for each variable in the
-      problem. The velocity is given in components in the $x$ and $y$
-      directions.}}
+  \caption{\textbf{Lid driven cavity FANM verification for a Reynolds
+      number of 500.} \textit{The values reported are global minima,
+      maxima, and averages for each variable in the problem. The
+      velocity is given in components in the $x$ and $y$ directions.}}
   \label{tab:driven_re500_results}
@@
   \end{center}
-  \caption{\textbf{Lid driven cavity nonlinear residual norm
+  \caption{\textbf{Lid driven cavity nonlinear residual norm achieved
+      convergence.} \textit{The results presented here were obtained
+      from the benchmark verification calculations.}}
+  \label{tab:driven_residual_norm_comparison}
+\end{table}
+
+\clearpage
+
+\subsection{Backward Facing Step Results}
+\label{subsec:backward_step_verification}
+
+A Drekar model of the backward facing step problem was solved using
+both the Newton-Krylov solver and a FANM solver for Reynolds numbers
+of 200, 400, 600, and 800 on a $20 \times 250$ rectangular
+grid. Table~\ref{tab:backward_step_parameters} gives the physical
+parameters of the problem and Table~\ref{tab:step_mcsa_parameters}
+gives the MCSA solver parameters used within the FANM
+method. Variation in the Reynolds number of the system was achieved by
+keeping the parameters given in
+Table~\ref{tab:backward_step_parameters} fixed and varying the maximum
+fluid velocity in the parabolic inlet condition. In addition to
+multigrid preconditioning for the linear solution at each Newton step,
+a polynomial line search method was used as a means of globalization
+assisting the Newton correction computation as outlined by Pawlowski
+in \citep{pawlowski_globalization_2006} while the forcing term was
+again computed using method 2 from Shadid's work
+\citep{shadid_inexact_1997}.
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lcc}\hline\hline
+      \multicolumn{1}{l}{Parameter}& 
+      \multicolumn{1}{c}{Value}&
+      \multicolumn{1}{c}{Units}\\\hline
+      $x_{min}$ & 0.0 & $m$ \\
+      $x_{max}$ & 1.0 & $m$ \\
+      $y_{min}$ & 0.0 & $m$ \\
+      $y_{max}$ & 1.0 & $m$ \\
+      $N_x$ & 50 & - \\
+      $N_y$ & 50 & - \\
+      $DOFs$ & 7803 & - \\
+      Density & 1.0 & $kg / m^3$ \\
+      Dynamic Viscosity & 0.71 & $(N \times s) / m^2$ \\
+      Kinematic Viscosity & 0.71 & $m^2 / s$ \\
+      Channel Width & 1.0 & $m$ \\
+      Step Height & 0.5 & $m$ \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Backward facing step FANM verification model
+      problem parameters.}  \textit{The Navier-Stokes equations in 2
+      dimensions are used for the model problem. The Reynolds number
+      is varied by modifying the velocity magnitude on the upper
+      boundary to induce flow step by inertial forces.}}
+  \label{tab:backward_step_parameters}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lc}\hline\hline
+      \multicolumn{1}{l}{Parameter}& 
+      \multicolumn{1}{c}{Value}\\\hline
+      Weight Cutoff & \sn{1}{-2} \\
+      Fixed Point Iteration & Richardson \\
+      Estimator & Adjoint Collision \\
+      Histories, Re=200 & 600,000 per iteration \\
+      Histories, Re=400 & 600,000 per iteration \\
+      Histories, Re=600 & 600,000 per iteration \\
+      Histories, Re=800 & 1,000,000 per iteration \\
+      Reduced Domain Fill Level, Re=200 & 200 \\
+      Reduced Domain Fill Level, Re=400 & 200 \\
+      Reduced Domain Fill Level, Re=600 & 200 \\
+      Reduced Domain Fill Level, Re=800 & 300 \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Backward facing step FANM verification MCSA solver
+      parameters.} \textit{No relaxation parameters or variance
+      reduction techniques were used with MCSA other than the reduced
+      domain approximation. Different values of fill level for the
+      reduced domain approximation and histories per iteration were
+      used at different Reynolds numbers due to convergence
+      requirements.}}
+  \label{tab:step_mcsa_parameters}
+\end{table}
+
+For the verification calculations, at each Reynolds number variation
+the global minimum, maximum, and average values were recorded for the
+fluid velocity in the x and y directions and the fluid
+pressure. Tables~\ref{tab:step_re200_results},
+\ref{tab:step_re400_results}, \ref{tab:step_re600_results}, and
+\ref{tab:step_re800_results} give the results of these
+computations. As with the thermal convection problem, the negative
+pressures computed are due to a relative pressure being reported. For
+each of the data points given, both Newton-Krylov solutions and FANM
+solutions computed identical numerical results for all of these global
+values. The norm of the converged nonlinear residual for the
+calculations is given in
+Table~\ref{tab:step_residual_norm_comparison}. All computed global
+values were within these tolerances.
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lll}\hline\hline
+      \multicolumn{1}{l}{Parameter}& 
+      \multicolumn{1}{l}{Newton-Krylov Result}&
+      \multicolumn{1}{l}{FANM Result}\\
+      \hline
+      Average $U_x$ & \\
+      Minimum $U_x$ & \\ 
+      Maximum $U_x$ & \\
+      \hline
+      Average $U_y$ & \\
+      Minimum $U_y$ & \\ 
+      Maximum $U_y$ & \\ 
+      \hline
+      Average $P$ & \\ 
+      Minimum $P$ & \\ 
+      Maximum $P$ & \\ 
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Backward facing step cavity FANM verification for a
+      Reynolds number of 200.} \textit{The values reported are global
+      minima, maxima, and averages for each variable in the
+      problem. The velocity is given in components in the $x$ and $y$
+      directions.}}
+  \label{tab:step_re200_results}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lll}\hline\hline
+      \multicolumn{1}{l}{Parameter}& 
+      \multicolumn{1}{l}{Newton-Krylov Result}&
+      \multicolumn{1}{l}{FANM Result}\\
+      \hline
+      Average $U_x$ & \\
+      Minimum $U_x$ & \\
+      Maximum $U_x$ & \\
+      \hline
+      Average $U_y$ & \\
+      Minimum $U_y$ & \\
+      Maximum $U_y$ & \\
+      \hline
+      Average $P$ & \\
+      Minimum $P$ & \\
+      Maximum $P$ & \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Backward facing step cavity FANM verification for a
+      Reynolds number of 400.} \textit{The values reported are global
+      minima, maxima, and averages for each variable in the
+      problem. The velocity is given in components in the $x$ and $y$
+      directions.}}
+  \label{tab:step_re400_results}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lll}\hline\hline
+      \multicolumn{1}{l}{Parameter}& 
+      \multicolumn{1}{l}{Newton-Krylov Result}&
+      \multicolumn{1}{l}{FANM Result}\\
+      \hline
+      Average $U_x$ & \\
+      Minimum $U_x$ & \\
+      Maximum $U_x$ & \\
+      \hline
+      Average $U_y$ & \\
+      Minimum $U_y$ & \\
+      Maximum $U_y$ & \\
+      \hline
+      Average $P$ & \\
+      Minimum $P$ & \\
+      Maximum $P$ & \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Backward facing step cavity FANM verification for a
+      Reynolds number of 600.} \textit{The values reported are global
+      minima, maxima, and averages for each variable in the
+      problem. The velocity is given in components in the $x$ and $y$
+      directions.}}
+  \label{tab:step_re600_results}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lll}\hline\hline
+      \multicolumn{1}{l}{Parameter}& 
+      \multicolumn{1}{l}{Newton-Krylov Result}&
+      \multicolumn{1}{l}{FANM Result}\\
+      \hline
+      Average $U_x$ & \\
+      Minimum $U_x$ & \\
+      Maximum $U_x$ & \\
+      \hline
+      Average $U_y$ & \\
+      Minimum $U_y$ & \\
+      Maximum $U_y$ & \\
+      \hline
+      Average $P$ & \\
+      Minimum $P$ & \\
+      Maximum $P$ & \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Backward facing step cavity FANM verification for a
+      Reynolds number of 800.} \textit{The values reported are global
+      minima, maxima, and averages for each variable in the
+      problem. The velocity is given in components in the $x$ and $y$
+      directions.}}
+  \label{tab:step_re800_results}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lcc}\hline\hline
+      \multicolumn{1}{l}{Reynolds Number}& 
+      \multicolumn{1}{c}{Newton Krylov $||\ve{F(\ve{u})}||$}&
+      \multicolumn{1}{c}{FANM $||\ve{F(\ve{u})}||$}\\
+      \hline
+      200 & \\
+      400 & \\
+      600 & \\
+      800 & \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Backward facing step cavity nonlinear residual norm
       achieved convergence.} \textit{The results presented here were
       obtained from the benchmark verification calculations.}}
-  \label{tab:driven_residual_norm_comparison}
+  \label{tab:step_residual_norm_comparison}
 \end{table}
@@
   \begin{center}
-    \begin{tabular}{lcc}\hline\hline
+    \begin{tabular}{lc}\hline\hline
       \multicolumn{1}{l}{Rayleigh Number}& 
@@
   \begin{center}
-    \begin{tabular}{lcc}\hline\hline
+    \begin{tabular}{lc}\hline\hline
       \multicolumn{1}{l}{Reynolds Number}& 
@@
   \end{center}
-  \caption{\textbf{Lid driven cavity Newton-Krylov speedup
+  \caption{\textbf{Lid driven cavity Newton-Krylov speedup over FANM
+      method.} \textit{Speedup values are rounded to the nearest
+      integer. The results presented here were obtained from the
+      benchmark verification calculations.}}
+  \label{tab:driven_speedup_comparison}
+\end{table}
+
+\clearpage
+
+\subsection{Backward Facing Step Results}
+\label{subsec:backward_step_comparison}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lcc}\hline\hline
+      \multicolumn{1}{l}{Reynolds Number}& 
+      \multicolumn{1}{c}{Newton-Krylov Iterations}&
+      \multicolumn{1}{c}{FANM Iterations}\\
+      \hline
+      200 & \\
+      400 & \\
+      600 & \\
+      800 & \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Backward facing step cavity nonlinear iteration
+      results for the performance comparison.} \textit{The results
+      presented here were obtained from the benchmark verification
+      calculations.}}
+  \label{tab:step_nonlinear_iter_comparison}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lcc}\hline\hline
+      \multicolumn{1}{l}{Reynolds Number}& 
+      \multicolumn{1}{c}{GMRES Iterations}&
+      \multicolumn{1}{c}{MCSA Iterations}\\
+      \hline
+      200 & \\
+      400 & \\
+      600 & \\
+      800 & \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Backward facing step cavity total sum of linear
+      iterations for all nonlinear iterations.} \textit{The results
+      presented here were obtained from the benchmark verification
+      calculations.}}
+  \label{tab:step_linear_iter_comparison}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lc}\hline\hline
+      \multicolumn{1}{l}{Reynolds Number}& 
+      \multicolumn{1}{c}{Newton-Krylov Speedup}\\
+      \hline
+      200 & \\
+      400 & \\
+      600 & \\
+      800 & \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Backward facing step cavity Newton-Krylov speedup
       over FANM method.} \textit{Speedup values are rounded to the
@@
       the benchmark verification calculations.}}
-  \label{tab:driven_speedup_comparison}
+  \label{tab:step_speedup_comparison}
 \end{table}

### 2013-07-19 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
adding backward step output

@@
       $x_{min}$ & 0.0 & $m$ \\
-      $x_{max}$ & 1.0 & $m$ \\
+      $x_{max}$ & 25.0 & $m$ \\
       $y_{min}$ & 0.0 & $m$ \\
       $y_{max}$ & 1.0 & $m$ \\
-      $N_x$ & 50 & - \\
-      $N_y$ & 50 & - \\
-      $DOFs$ & 7803 & - \\
+      $N_x$ & 20 & - \\
+      $N_y$ & 250 & - \\
+      $DOFs$ & 15813 & - \\
       Density & 1.0 & $kg / m^3$ \\
-      Dynamic Viscosity & 0.71 & $(N \times s) / m^2$ \\
-      Kinematic Viscosity & 0.71 & $m^2 / s$ \\
+      Dynamic Viscosity & 1.0 & $(N \times s) / m^2$ \\
+      Kinematic Viscosity & 1.0 & $m^2 / s$ \\
       Channel Width & 1.0 & $m$ \\

### 2013-07-19 doc/dissertation/document/chapters/introduction/introduction.tex
adding backward step fanm data for re 400 case

@@
 In nearly all high-fidelity nuclear reactor simulations, linear and
-nonlinear transport problems are a primary focus of study. Recent
-focus on multiple physics systems in the nuclear reactor modeling and
-simulation community adds a new level of complexity to common linear
-and nonlinear systems as solution strategies change when they are
-coupled to other problems
+nonlinear transport problems are a primary focus of study. Neutronics,
+fluid flow, and heat transfer are among the predominate physics that
+fall into this category and simulation technologies to solve these
+problems make up much of the analysis suite used for modern
+analysis. Recent focus on multiple physics systems in the nuclear
+reactor modeling and simulation community adds a new level of
+complexity to common linear and nonlinear systems as solution
+strategies change when they are coupled to other problems
 \citep{u.s._department_of_energy_casl_2011}. Furthermore, a desire for
@@
 nuclear systems creates a need for extremely high fidelity
-computations to be performed for these coupled transport systems as a
-means to capture effects not modeled by coarser methods.
+computations to be performed for these transport systems as a means to
+capture effects not modeled by coarser methods.
 
@@
 and considerate of hardware-related issues. As scientific computing
-moves towards exascale facilities with machines of $O(1,000,000)$
+moves beyond petascale facilities with machines of $O(1,000,000)$
 cores already coming on-line, new algorithms to solve these complex
@@
 
+This goal of this work is to research and develop parallel solutions
+for discrete linear and nonlinear transport problems that improve
+performance and scalabilty using a novel set of Monte Carlo synthetic
+acceleration methods.
+
 We discuss in this chapter physics-based motivation for advancing
-Monte Carlo solvers and studying their application to transport
-systems by providing problems of interest in nuclear reactor
-analysis. Hardware-based motivations are also provided by considering
-the impact of forthcoming computing architectures. In addition,
-background on the current solver techniques for multiphysics problems
-and a brief comparison to the proposed methods is provided to further
-motivate this work. The organization of the document is then reviewed
-followed by an explanation of the key outcomes of this work and where
-in the document results and data supporting those outcomes can be
-located.
+Monte Carlo synthetic acceleration methods and studying their
+application to transport systems by providing problems of interest in
+nuclear reactor analysis. Hardware-based motivations are also provided
+by considering the impact of forthcoming computing architectures. In
+addition, background on the current solver techniques for these
+physics problems and a brief comparison to the proposed methods are
+provided to further motivate this work. The organization of the
+document is then reviewed.
 
@@
 \label{sec:hardware_motivation}
-As leadership class machines move towards the exascale, new algorithms
+As leadership class machines move beyond the petascale, new algorithms
 must be developed that leverage their strengths and adapt to their
@@
 funding for the next several years to research resilient solver
-technologies for exascale facilities
+technologies for extreme scale facilities
 \citep{u.s._department_of_energy_resilient_2012}. Based on the
@@
 consumption will require incredibly high levels of concurrency to
-achieve the desired computation rates. Furthermore, this drop in
-power consumption will mean increased pressure on memory as memory per
-node is expected to stagnate while cores per node is expected to
+achieve the desired computation rates. Furthermore, this drop in power
+consumption will mean increased pressure on memory as memory per node
+is expected to stagnate while cores per node is expected to
 increase. As the number of cores increases, their clock speed is
@@
 current deterministic solver techniques that will suffer greatly from
-the non-deterministic behavior expected from these exascale machines.
+the non-deterministic behavior expected from the next generation of
+machines.
 
@@
 be implemented with stochastic methods could offer a new avenue for
-leveraging the expected levels of high concurrency in exascale
+leveraging the expected levels of high concurrency in extreme scale
 machines.
@@
 
+%%---------------------------------------------------------------------------%%
+\section{Outline}
+
 To present these new developments this document is arranged in the

### 2013-07-21 drivers/NeutronDiffusion/neutron_diffusion.cpp
adding re 500 results for backward facing step

@@ -96,32 +96,17 @@ int main( int argc, char * argv[] )
     Teuchos::RCP<const Teuchos::Comm<int> > set_comm = 
         comm->createSubcommunicator( comm_ranks() );
     Teuchos::RCP<Chimera::DiffusionProblem> diffusion_problem;
+    Teuchos::RCP<Chimera::Partitioner> partitioner;
     if ( 0 == set_id )
     {
         // Build and partition the mesh.
-        Teuchos::RCP<Chimera::Partitioner> partitioner = Teuchos::rcp(
-            new Chimera::Partitioner( set_comm, plist ) );
+        partitioner = Teuchos::rcp( new Chimera::Partitioner(set_comm, plist) );
 
         // Build the diffusion problem.
         diffusion_problem = Teuchos::rcp(
-            new Chimera::DiffusionProblem( set_comm, partitioner, plist, true ) );
+            new Chimera::DiffusionProblem(set_comm, partitioner, plist, true) );
     }
     comm->barrier();
-
-//     // // CHIMERA SOLVE
-//     // Build the solver.
-//     Teuchos::RCP<Chimera::LinearSolver<double,int,int> > csolver =
-//         Chimera::LinearSolverFactory::create( 
-//             plist, diffusion_problem->getProblem() );
-
-//     // Compute the iteration matrix spectral radius.
-//     double spec_rad = Chimera::OperatorTools::spectralRadius(
-//         csolver->linearOperatorSplit()->iterationMatrix() );
-//     if ( comm->getRank() == 0 )
-//     {
-//         std::cout << "SPECTRAL RADIUS: " << spec_rad << std::endl; 
-//     }
-//     comm->barrier();
 
     typedef Tpetra::Vector<double,int,int> Vector;
     typedef Tpetra::CrsMatrix<double,int,int> Matrix;
@@ -206,12 +191,15 @@ int main( int argc, char * argv[] )
         }
     }
 
-//     // Write the solution to VTK.
-//     Chimera::VTKOutput vtk_output( comm, partitioner, plist );
-//     vtk_output.addField( Chimera::VTKOutput::VERTEX_FIELD,
-//                          diffusion_problem->getProblem()->getLHS(),
-//                          "NEUTRON_FLUX" );
-//     vtk_output.write();
+    // Write the solution to VTK.
+    Chimera::VTKOutput vtk_output( comm, partitioner, plist );
+    vtk_output.addField( Chimera::VTKOutput::VERTEX_FIELD,
[…]

### 2013-07-21 src/Chimera_OperatorTools.hpp

@@ -41,6 +41,7 @@
 
 #include <Teuchos_RCP.hpp>
 
+#include <Tpetra_Vector.hpp>
 #include <Tpetra_CrsMatrix.hpp>
 
 namespace Chimera
@@ -85,8 +86,10 @@ class OperatorTools
 
     // Compute the spectral radius of an operator.
     template<class Scalar, class LO, class GO>
-    static Scalar spectralRadius( 
-	const Teuchos::RCP<Tpetra::CrsMatrix<Scalar,LO,GO> >& matrix );
+    static void spectralRadius( 
+	const Teuchos::RCP<Tpetra::CrsMatrix<Scalar,LO,GO> >& matrix,
+        double& spec_rad,
+        Teuchos::RCP<Tpetra::Vector<Scalar,LO,GO> >& eigenvector );
 
     // Compute the condition number of an operator.
     template<class Scalar, class LO, class GO>

### 2013-07-21 src/Chimera_OperatorTools_def.hpp

@@ -163,8 +163,10 @@ Teuchos::Array<GO> OperatorTools::getOffProcColumns(
  * \brief Compute the spectral radius of an operator.
  */
 template<class Scalar,class LO, class GO>
-Scalar OperatorTools::spectralRadius( 
-    const Teuchos::RCP<Tpetra::CrsMatrix<Scalar,LO,GO> >& matrix )
+void OperatorTools::spectralRadius( 
+    const Teuchos::RCP<Tpetra::CrsMatrix<Scalar,LO,GO> >& matrix,
+    double& spec_rad,
+    Teuchos::RCP<Tpetra::Vector<Scalar,LO,GO> >& eigenvector )
 {
     typedef Tpetra::MultiVector<Scalar,LO,GO> MV;
     typedef Tpetra::Operator<Scalar,LO,GO> OP;
@@ -173,9 +175,13 @@ Scalar OperatorTools::spectralRadius(
     const int block_size = 1;
     const int num_blocks = 10;
     const int max_restarts = 100;
-    const Scalar tol = 1.0e-4;
+    const Scalar tol = 1.0e-8;
+
+    int verbosity = Anasazi::Errors + Anasazi::Warnings + 
+                    Anasazi::FinalSummary + Anasazi::TimingDetails;
 
     Teuchos::ParameterList krylovschur_params;
+    krylovschur_params.set( "Verbosity", verbosity );
     krylovschur_params.set( "Which", "LM" );
     krylovschur_params.set( "Block Size", block_size );
     krylovschur_params.set( "Num Blocks", num_blocks );
@@ -201,9 +207,11 @@ Scalar OperatorTools::spectralRadius(
     Teuchos::RCP<MV> evecs = sol.Evecs;
 
     testPostcondition( sol.numVecs > 0 );
-    
-    return std::pow( evals[0].realpart*evals[0].realpart +
-		     evals[0].imagpart*evals[0].imagpart, 0.5 );
+
+    spec_rad = std::pow( evals[0].realpart*evals[0].realpart +
+                         evals[0].imagpart*evals[0].imagpart, 0.5 );
+
+    eigenvector = evecs->getVectorNonConst(0);
 }
 
 //---------------------------------------------------------------------------//

### 2013-07-22 doc/dissertation/document/chapters/mc_background/mc_background.tex
adding more to chapter summaries
fixing minres -> minimal residual

@@
 positive-definite but not necessarily symmetric problems, the minimal
-residual (MINRES) iteration \citep{saad_iterative_2003} can be used in
+residual iteration \citep{saad_iterative_2003} can be used in
 conjunction with MCSA where the residual vector, $\mathbf{r}$, defines
@@
 subspace. Using this, we can then define an MCSA scheme that
-accelerates the MINRES iteration:
+accelerates the minimal residual iteration:
 \begin{subequations}
@@
     \ve{x}^{k+1/2} = \ve{x}^k + \alpha \ve{r}^k\:,
-    \label{eq:minres_correction}\\
+    \label{eq:min_res_correction}\\
     \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\ 
@@
   \end{gather}
-  \label{eq:mcsa_minres}
+  \label{eq:mcsa_min_res}
 \end{subequations}
@@
 additional extrapolation parameter, $\alpha$, is applied to the
-residual correction in Eq~(\ref{eq:minres_correction}) as a means of
+residual correction in Eq~(\ref{eq:min_res_correction}) as a means of
 selecting a more optimal search direction at each iteration,
@@
 \begin{itemize}
-\item MCSA has been generalized for all discrete linear transport
-  problems
+\item Monte Carlo synthetic acceleration (MCSA) has been generalized
+  for all discrete linear transport problems
 \item MCSA can be preconditioned using a general left/right scheme but

### 2013-07-22 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
adding backward step fanm data for re 400 case
fixing minres -> minimal residual

@@
 all of the components necessary for a Monte Carlo solver to provide
-value. We therefore devise a new nonlinear method based on the MCSA algorithm
-and Newton's method.
+value. We therefore devise a new nonlinear method based on the MCSA
+algorithm and Newton's method.
 
@@
         \sn{1}{6}. Isotherms on unit temperature scale from 0 to 1
-        with divisions of 0.1 for each calculation.}}
+        with 10 uniformly spaced curves for each calculation.}}
   \label{fig:convection_isotherms}
@@
   \caption{\textbf{Velocity magnitude isocurves from solution to the
-      thermal convection cavity problem.} \textit{Top left: Re = 100;
-      Top right: Re= 300; Bottom left: Re = 500; Bottom left: Re =
-      700. Isocurves on unit velocity scale from 0 to 1 with divisions
-      of 0.05 for each calculation.}}
+      lid driven cavity problem.} \textit{Top left: Re = 100; Top
+      right: Re= 300; Bottom left: Re = 500; Bottom left: Re =
+      700. Isocurves on unit velocity scale from 0 to 1 with 20
+      uniformly spaced curves for each calculation.}}
   \label{fig:driven_velocity_isocurves}
@@
 with the boundary conditions as shown in
-Figure~\ref{fig:backward_facing_step}.
+Figure~\ref{fig:backward_facing_step}. For this work, a slightly
+different parabolic formulation was used than in the benchmark. In the
+benchmark solutions, the average velocity in the parabola was 2/3 that
+of the maximum velocity while formula given in
+Figure~\ref{fig:backward_facing_step} produces a more plug-like flow
+with an average velocity of 3/4 that of the maximum velocity.
 
@@
       half has a zero velocity condition, simulating flow over a
-      step.}}
+      step. The $H$ parameter given in the parabolic inlet condition
+      is the width of the inlet.}}
   \label{fig:backward_facing_step}
@@
 magnitude contours for solutions to this problem with Reynolds numbers
-of 200, 400, 600, and 800.
+of 200, 300, 400, and 500.
 
@@
   \caption{\textbf{Velocity magnitude isocurves from solution to the
-      thermal convection cavity problem.} \textit{From top to bottom:
-      Re = 200, Re = 400, Re = 600, Re = 800. Isocurves on unit
-      velocity scale from 0 to 1 with divisions of 0.05 for each
+      backward facing step problem.} \textit{From top to bottom: Re =
+      200, Re = 300, Re = 400, Re = 500. Isocurves on unit velocity
+      scale from 0 to 1 with 20 uniformly spaced curves for each
       calculation.}}
-  \label{fig:driven_velocity_isocurves}
+  \label{fig:step_velocity_isocurves}
 \end{figure}
@@
 both the Newton-Krylov solver and a FANM solver for Reynolds numbers
-of 200, 400, 600, and 800 on a $20 \times 250$ rectangular
+of 200, 300, 400, and 500 on a $20 \times 250$ rectangular
 grid. Table~\ref{tab:backward_step_parameters} gives the physical
@@
       dimensions are used for the model problem. The Reynolds number
-      is varied by modifying the velocity magnitude on the upper
-      boundary to induce flow step by inertial forces.}}
+      is varied by modifying the maximum velocity magnitude in the
+      parabolic inlet condition.}}
   \label{tab:backward_step_parameters}
@@
       Estimator & Adjoint Collision \\
-      Histories, Re=200 & 600,000 per iteration \\
-      Histories, Re=400 & 600,000 per iteration \\
-      Histories, Re=600 & 600,000 per iteration \\
-      Histories, Re=800 & 1,000,000 per iteration \\
+      Histories, Re=200 & 500,000 per iteration \\
+      Histories, Re=300 & 500,000 per iteration \\
+      Histories, Re=400 & 500,000 per iteration \\
+      Histories, Re=500 & 1,000,000 per iteration \\
       Reduced Domain Fill Level, Re=200 & 200 \\
@@
 
+In the literature on this benchmark, solutions at Reynolds numbers of
+up to 800 were achieved using a variety of nonlinear solution
+techniques. In identical fashion to the lid driven cavity problem, at
+Reynolds numbers above 500 convergence of the linear iterations could
+not be achieved with FANM regardless of how agressive the multigrid
+preconditioner was set. In these cases, the Jacobian operators are
+again ill-conditioned enough that a spectral radius of less than one
+could not be achieved with multigrid preconditioning and several cycle
+applications. Therefore, the spectral radius limitation on MCSA is
+extremely prohibitive in this case as well and even more so than the
+lid driven cavity problem where solutions at higher Reynolds numbers
+were achieved. Because of this, only solutions to the backward facing
+step benchmark at Reynolds numbers of up to 500 are presented.
+
 For the verification calculations, at each Reynolds number variation
@@
 pressure. Tables~\ref{tab:step_re200_results},
-\ref{tab:step_re400_results}, \ref{tab:step_re600_results}, and
-\ref{tab:step_re800_results} give the results of these
+\ref{tab:step_re300_results}, \ref{tab:step_re400_results}, and
+\ref{tab:step_re500_results} give the results of these
 computations. As with the thermal convection problem, the negative
@@
   \end{center}
-  \caption{\textbf{Backward facing step cavity FANM verification for a
+  \caption{\textbf{Backward facing step FANM verification for a
       Reynolds number of 200.} \textit{The values reported are global
@@
   \end{center}
-  \caption{\textbf{Backward facing step cavity FANM verification for a
+  \caption{\textbf{Backward facing step FANM verification for a
+      Reynolds number of 300.} \textit{The values reported are global
+      minima, maxima, and averages for each variable in the
+      problem. The velocity is given in components in the $x$ and $y$
+      directions.}}
+  \label{tab:step_re300_results}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lll}\hline\hline
+      \multicolumn{1}{l}{Parameter}& 
+      \multicolumn{1}{l}{Newton-Krylov Result}&
+      \multicolumn{1}{l}{FANM Result}\\
+      \hline
+      Average $U_x$ & \\
+      Minimum $U_x$ & \\
+      Maximum $U_x$ & \\
+      \hline
+      Average $U_y$ & \\
+      Minimum $U_y$ & \\
+      Maximum $U_y$ & \\
+      \hline
+      Average $P$ & \\
+      Minimum $P$ & \\
+      Maximum $P$ & \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Backward facing step FANM verification for a
       Reynolds number of 400.} \textit{The values reported are global
@@
   \end{center}
-  \caption{\textbf{Backward facing step cavity FANM verification for a
-      Reynolds number of 600.} \textit{The values reported are global
+  \caption{\textbf{Backward facing step FANM verification for a
+      Reynolds number of 500.} \textit{The values reported are global
       minima, maxima, and averages for each variable in the
@@
       directions.}}
-  \label{tab:step_re600_results}
-\end{table}
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lll}\hline\hline
-      \multicolumn{1}{l}{Parameter}& 
-      \multicolumn{1}{l}{Newton-Krylov Result}&
-      \multicolumn{1}{l}{FANM Result}\\
-      \hline
-      Average $U_x$ & \\
-      Minimum $U_x$ & \\
-      Maximum $U_x$ & \\
-      \hline
-      Average $U_y$ & \\
-      Minimum $U_y$ & \\
-      Maximum $U_y$ & \\
-      \hline
-      Average $P$ & \\
-      Minimum $P$ & \\
-      Maximum $P$ & \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Backward facing step cavity FANM verification for a
-      Reynolds number of 800.} \textit{The values reported are global
-      minima, maxima, and averages for each variable in the
-      problem. The velocity is given in components in the $x$ and $y$
-      directions.}}
-  \label{tab:step_re800_results}
+  \label{tab:step_re500_results}
 \end{table}
@@
       200 & \\
+      300 & \\
       400 & \\
-      600 & \\
-      800 & \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Backward facing step cavity nonlinear residual norm
+      500 & \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Backward facing step nonlinear residual norm
       achieved convergence.} \textit{The results presented here were
@@
   \end{center}
-  \caption{\textbf{Backward facing step cavity nonlinear iteration
+  \caption{\textbf{Backward facing step nonlinear iteration
       results for the performance comparison.} \textit{The results
@@
   \end{center}
-  \caption{\textbf{Backward facing step cavity total sum of linear
+  \caption{\textbf{Backward facing step total sum of linear
       iterations for all nonlinear iterations.} \textit{The results
@@
   \end{center}
-  \caption{\textbf{Backward facing step cavity Newton-Krylov speedup
+  \caption{\textbf{Backward facing step Newton-Krylov speedup
       over FANM method.} \textit{Speedup values are rounded to the
@@
 \item Forward-Automated Newton-MCSA (FANM), a new inexact Newton
-  method based on Monte Carlo synthetic acceleration, has been
+  method leveraging Monte Carlo synthetic acceleration, has been
   developed
@@
 \item The FANM method has been verified to produce the same solutions
-  as a production Newton-Krylov method for two benchmark problems for
-  the Navier-Stokes equations in different flow regimes
+  as a production Newton-Krylov method for three difficult benchmark
+  problems for the Navier-Stokes equations in different flow regimes
+  and geometries
 \item The FANM method has better iterative performance than the
@@
   problems at high Reynolds numbers
-\item Explict algebraic preconditioning of MCSA creates FANM run times
-  $O(100)$ slower than the Newton-Krylov solver
+\item Explict algebraic preconditioning used to overcome the spectral
+  radius restriction of MCSA creates FANM run times
+  $O(100)$-$O(1,000)$ slower than the Newton-Krylov solver
 \item More Monte Carlo histories at every FANM iteration can reduce
@@
   the problem
-\item With future optimization, reactor accident scenario analyses
-  that rely on convection dominated transport could experience
-  improved time to solution with FANM given the improved iterative
-  performance for problems of this type
+\item With future optimization, reactor accident analyses that rely on
+  convection dominated transport or steady state calculations
+  requiring flow at low Reynolds numbers could experience improved
+  time to solution with FANM given the improved iterative performance
+  observed for problems of these types
 \end{itemize}

### 2013-07-22 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
fixing minres -> minimal residual

@@
 transport problem, we explore solutions to the $SP_N$ equations with
-Monte Carlo Synthetic Acceleration using a light water reactor fuel
-assembly criticality calculation as the driving problem. Several
-difficulties arise when applying MCSA to these problems that were not
-observed when investigating the simpler transport systems.
+Monte Carlo Synthetic Acceleration using a challenging light water
+reactor fuel assembly criticality calculation as the driving
+problem. Several difficulties arise when applying MCSA to these
+problems that were not observed when investigating the simpler
+transport systems.
 
@@
 analysis. Additionally, both the Richardson-based MCSA iteration given
-by Eq~(\ref{eq:mcsa}) and the MINRES-based MCSA iteration given by
-Eq~(\ref{eq:mcsa_minres}) will be used. Although the expected value
+by Eq~(\ref{eq:mcsa}) and the minimal residual-based MCSA iteration given by
+Eq~(\ref{eq:mcsa_min_res}) will be used. Although the expected value
 estimator achieved the best iterative performance in the previous
@@
 solver using both the collision and expected value estimators and both
-the Richardson and MINRES fixed point iterations. Using the results
-from the previous section, a Neumann relaxation parameter of 0.7 was
-used with both estimators and fixed point iterations. For the
+the Richardson and minimal residual fixed point iterations. Using the
+results from the previous section, a Neumann relaxation parameter of
+0.7 was used with both estimators and fixed point iterations. For the
 collision estimator, a Richardson relaxation parameter of 1.1 was used
@@
 the Richardson iteration. Better timing performance is expected as
-computing the extrapolation parameter in the MINRES iteration requires
-several additional parallel operations. However, we expect that
-because the MINRES iteration will converge faster on its own when
-compared to the Richardson iteration that better iterative performance
-will be observed when it is accelerated with MCSA. The fact that this
-was not observed signals that an MCSA scheme leveraging the expected
-value estimator is also effectively insensitive to the fixed point
-iteration used. For the collision estimator, the MINRES iteration
-permits fewer stochastic histories to be used with each MCSA iteration
-while still maintaining good iterative performance. With respect to
-CPU time, there is a more pronounced effect due to the larger density
-of states created by the explicit ILUT preconditioning. Using a
-reduced domain fill level of 100 creates a situation where the
-expected value estimator is significantly more expensive per history
-to compute that the collision estimator due to the coupling of states.
+computing the extrapolation parameter in the minimal residual
+iteration requires several additional parallel operations. However, we
+expect that because the minimal residual iteration will converge
+faster on its own when compared to the Richardson iteration that
+better iterative performance will be observed when it is accelerated
+with MCSA. The fact that this was not observed signals that an MCSA
+scheme leveraging the expected value estimator is also effectively
+insensitive to the fixed point iteration used. For the collision
+estimator, the minimal residual iteration permits fewer stochastic
+histories to be used with each MCSA iteration while still maintaining
+good iterative performance. With respect to CPU time, there is a more
+pronounced effect due to the larger density of states created by the
+explicit ILUT preconditioning. Using a reduced domain fill level of
+100 creates a situation where the expected value estimator is
+significantly more expensive per history to compute that the collision
+estimator due to the coupling of states.
 
@@
       iteration. Each estimator was used with the Richardson and
-      MINRES fixed point iterations within an MCSA iteration.}}
+      minimal residual fixed point iterations within an MCSA
+      iteration.}}
   \label{fig:spn_estimator_iters}
@@
       each iteration. Each estimator was used with the Richardson and
-      MINRES fixed point iterations within an MCSA iteration.}}
+      minimal residual fixed point iterations within an MCSA
+      iteration.}}
   \label{fig:spn_estimator_time}
@@
       each iteration. Each estimator was used with the Richardson and
-      MINRES fixed point iterations within an MCSA iteration.}}
+      minimal residual fixed point iterations within an MCSA
+      iteration.}}
   \label{fig:spn_estimator_convergence}
@@
                     & fixed point iteration and collision estimator \\
-      MCSA-ILUT-MR-C & MCSA preconditioned with ILUT using MINRES \\
+      MCSA-ILUT-MR-C & MCSA preconditioned with ILUT using minimal residual \\
                      & fixed point iteration and collision estimator \\
@@
                      & fixed point iteration and expected value estimator \\
-      MCSA-ILUT-MR-EV & MCSA preconditioned with ILUT using MINRES \\
+      MCSA-ILUT-MR-EV & MCSA preconditioned with ILUT using minimal residual \\
                       & fixed point iteration and expected value estimator \\
@@
   more then Bi-CGStab
-\item The explicit preconditioning strategy creates MCSA runtimes
-  $O(100)$ larger than the production Krylov methods for the fuel
-  assembly criticality problem
+\item The explicit preconditioning strategy required to overcome the
+  MCSA spectral radius restriction creates runtimes $O(100)$ larger
+  than the production Krylov methods for the fuel assembly criticality
+  problem
 \end{itemize}

### 2013-07-22 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
working on frontmatter

@@
 along with a brief discussion of inexact Newton methods. Newton-Krylov
-methods, a variant of inexact Newton methods, will be outlined along
-with two common techniques for Jacobian generation as a means of
-defining current practices. Following this, the new algorithm will be
-presented and implementation details will be discussed. Using the new
-algorithm and a production fluid physics code, a group of challenging
-nonlinear benchmark problems for the Navier-Stokes equations will be
-used to verify the correctness of the method relative to solutions
-obtained using conventional methods in various flow regimes. Finally,
-we will compare the performance of the new method against contemporary
-Newton-Krylov methods for the same benchmark problems.
+methods, a variant of inexact Newton methods, will be outlined
+including with two common techniques for numerically representing the
+Jacobian most commonly used in today's applications. Following this,
+the new algorithm will be presented and implementation details will be
+discussed. Using the new algorithm and a production fluid physics
+code, a group of challenging nonlinear benchmark problems for the
+Navier-Stokes equations will be used to verify the correctness of the
+method relative to solutions obtained using conventional methods in
+various flow regimes and geometries. Finally, we will compare the
+performance of the new method against contemporary Newton-Krylov
+methods for the same benchmark problems.
 
@@
 with the boundary conditions as shown in
-Figure~\ref{fig:backward_facing_step}. For this work, a slightly
-different parabolic formulation was used than in the benchmark. In the
-benchmark solutions, the average velocity in the parabola was 2/3 that
-of the maximum velocity while formula given in
-Figure~\ref{fig:backward_facing_step} produces a more plug-like flow
-with an average velocity of 3/4 that of the maximum velocity.
+Figure~\ref{fig:backward_facing_step}.
 
@@
 parabolic flow profile over a zero velocity boundary representing a
-step. The flow over this step will generate a recirculating backflow
-under the inlet flow towards the step and at higher Reynolds numbers a
-recirculating zone will form farther downstream from the step. As in
-the lid driven cavity problem, the nonlinear behavior of this
-benchmark and the difficulty in obtaining a solution is dictated by
-the Reynolds number of the fluid. In Shadid's work, a Reynolds numbers
-up to 800 were used.
+step. For this work, a slightly different parabolic formulation for
+the inlet velocities was used than in the benchmark. In the benchmark
+solutions, the average velocity in the parabola was 2/3 that of the
+maximum velocity while the formula given in
+Figure~\ref{fig:backward_facing_step} produces a more plug-like flow
+with an average velocity 3/4 that of the maximum velocity. The flow
+over this step will generate a recirculating backflow under the inlet
+flow towards the step and at higher Reynolds numbers a recirculating
+zone will form farther downstream from the step. As in the lid driven
+cavity problem, the nonlinear behavior of this benchmark and the
+difficulty in obtaining a solution is dictated by the Reynolds number
+of the fluid. In Shadid's work, a Reynolds numbers up to 800 were
+used.
 
@@
       \hline
-      Average $U_x$ & -1.288835730E-05 & -1.288835730E-05 \\
-      Minimum $U_x$ & -3.677243960E+00 & -3.677243960E+00 \\
-      Maximum $U_x$ & 3.688908550E+00 & 3.688908550E+00 \\
-      \hline
-      Average $U_y$ & 3.403670530E-02 & 3.403670530E-02 \\
-      Minimum $U_y$ & -3.721178580E+00 & -3.721178580E+00 \\
-      Maximum $U_y$ & 3.791483420E+00 & 3.791483420E+00 \\
+      Average $U$ & -1.288835730E-05 & -1.288835730E-05 \\
+      Minimum $U$ & -3.677243960E+00 & -3.677243960E+00 \\
+      Maximum $U$ & 3.688908550E+00 & 3.688908550E+00 \\
+      \hline
+      Average $V$ & 3.403670530E-02 & 3.403670530E-02 \\
+      Minimum $V$ & -3.721178580E+00 & -3.721178580E+00 \\
+      Maximum $V$ & 3.791483420E+00 & 3.791483420E+00 \\
       \hline
@@
       \hline
-      Average $U_x$ & -7.489201240E-04 & -7.489201240E-04 \\
-      Minimum $U_x$ & -1.607318000E+01 & -1.607318000E+01 \\
-      Maximum $U_x$ & 1.669098270E+01 & 1.669098270E+01 \\
-      \hline
-      Average $U_y$ & 3.377624420E-01 & 3.377624420E-01 \\
-      Minimum $U_y$ & -1.932174800E+01 & -1.932174800E+01 \\
-      Maximum $U_y$ & 2.047326820E+01 & 2.047326820E+01 \\
+      Average $U$ & -7.489201240E-04 & -7.489201240E-04 \\
+      Minimum $U$ & -1.607318000E+01 & -1.607318000E+01 \\
+      Maximum $U$ & 1.669098270E+01 & 1.669098270E+01 \\
+      \hline
+      Average $V$ & 3.377624420E-01 & 3.377624420E-01 \\
+      Minimum $V$ & -1.932174800E+01 & -1.932174800E+01 \\
+      Maximum $V$ & 2.047326820E+01 & 2.047326820E+01 \\
       \hline
@@
       \hline
-      Average $U_x$ & -2.043639370E-02 & -2.043639370E-02 \\
-      Minimum $U_x$ & -3.814922470E+01 & -3.814922470E+01 \\
-      Maximum $U_x$ & -3.814922470E+01 & -3.814922470E+01 \\
-      \hline
-      Average $U_y$ & 2.774864300E+00 & 2.774864300E+00 \\
-      Minimum $U_y$ & -6.038682990E+01 & -6.038682990E+01 \\
-      Maximum $U_y$ & 7.873873560E+01 & 7.873873560E+01 \\
+      Average $U$ & -2.043639370E-02 & -2.043639370E-02 \\
+      Minimum $U$ & -3.814922470E+01 & -3.814922470E+01 \\
+      Maximum $U$ & -3.814922470E+01 & -3.814922470E+01 \\
+      \hline
+      Average $V$ & 2.774864300E+00 & 2.774864300E+00 \\
+      Minimum $V$ & -6.038682990E+01 & -6.038682990E+01 \\
+      Maximum $V$ & 7.873873560E+01 & 7.873873560E+01 \\
       \hline
@@
       \hline
-      Average $U_x$ & -1.802569840E-01 & -1.802569840E-01 \\
-      Minimum $U_x$ & -6.292084130E+01 & -6.292084130E+01 \\
-      Maximum $U_x$ & 2.951113240E+02 & 2.951113240E+02 \\
-      \hline
-      Average $U_y$ & 1.336398550E+01 & 1.336398550E+01 \\
-      Minimum $U_y$ & -1.345663290E+02 & -1.345663290E+02 \\
-      Maximum $U_y$ & 3.227873890E+02 & 3.227873890E+02 \\
+      Average $U$ & -1.802569840E-01 & -1.802569840E-01 \\
+      Minimum $U$ & -6.292084130E+01 & -6.292084130E+01 \\
+      Maximum $U$ & 2.951113240E+02 & 2.951113240E+02 \\
+      \hline
+      Average $V$ & 1.336398550E+01 & 1.336398550E+01 \\
+      Minimum $V$ & -1.345663290E+02 & -1.345663290E+02 \\
+      Maximum $V$ & 3.227873890E+02 & 3.227873890E+02 \\
       \hline
@@
       \hline
-      Average $U_x$ & 1.165632320E-01 & 1.165632320E-01 \\
-      Minimum $U_x$ & -2.423642920E+01 & -2.423642920E+01 \\
-      Maximum $U_x$ & 9.742670880E+01 & 9.742670880E+01 \\
-      \hline
-      Average $U_y$ & 2.038480420E-02 & 2.038480420E-02 \\
-      Minimum $U_y$ & -5.266315430E+01 & -5.266315430E+01 \\
-      Maximum $U_y$ & 2.720044160E+01 & 2.720044160E+01 \\
+      Average $U$ & 1.165632320E-01 & 1.165632320E-01 \\
+      Minimum $U$ & -2.423642920E+01 & -2.423642920E+01 \\
+      Maximum $U$ & 9.742670880E+01 & 9.742670880E+01 \\
+      \hline
+      Average $V$ & 2.038480420E-02 & 2.038480420E-02 \\
+      Minimum $V$ & -5.266315430E+01 & -5.266315430E+01 \\
+      Maximum $V$ & 2.720044160E+01 & 2.720044160E+01 \\
       \hline
@@
       \hline
-      Average $U_x$ & 3.382078040E-01 & 3.382078040E-01 \\
-      Minimum $U_x$ & -9.467140360E+01 & -9.467140360E+01 \\
-      Maximum $U_x$ & 2.895885380E+02 & 2.895885380E+02 \\
-      \hline
-      Average $U_y$ & 8.780242440E-02 & 8.780242440E-02 \\
-      Minimum $U_y$ & -1.830156060E+02 & -1.830156060E+02 \\
-      Maximum $U_y$ & 8.827373340E+01 & 8.827373340E+01 \\
+      Average $U$ & 3.382078040E-01 & 3.382078040E-01 \\
+      Minimum $U$ & -9.467140360E+01 & -9.467140360E+01 \\
+      Maximum $U$ & 2.895885380E+02 & 2.895885380E+02 \\
+      \hline
+      Average $V$ & 8.780242440E-02 & 8.780242440E-02 \\
+      Minimum $V$ & -1.830156060E+02 & -1.830156060E+02 \\
+      Maximum $V$ & 8.827373340E+01 & 8.827373340E+01 \\
       \hline
@@
       \hline
-      Average $U_x$ & 5.098921320E-01 & 5.098921320E-01 \\
-      Minimum $U_x$ & -1.731861690E+02 & -1.731861690E+02 \\
-      Maximum $U_x$ & 4.789146040E+02 & 4.789146040E+02 \\
-      \hline
-      Average $U_y$ & 1.365493140E-01 & 1.365493140E-01 \\
-      Minimum $U_y$ & -3.170930360E+02 & -3.170930360E+02 \\
-      Maximum $U_y$ & 1.655401440E+02 & 1.655401440E+02 \\
+      Average $U$ & 5.098921320E-01 & 5.098921320E-01 \\
+      Minimum $U$ & -1.731861690E+02 & -1.731861690E+02 \\
+      Maximum $U$ & 4.789146040E+02 & 4.789146040E+02 \\
+      \hline
+      Average $V$ & 1.365493140E-01 & 1.365493140E-01 \\
+      Minimum $V$ & -3.170930360E+02 & -3.170930360E+02 \\
+      Maximum $V$ & 1.655401440E+02 & 1.655401440E+02 \\
       \hline
@@
       \hline
-      Average $U_x$ & 6.479541170E-01 & 6.479541170E-01 \\
-      Minimum $U_x$ & -2.550685400E+02 & -2.550685400E+02 \\
-      Maximum $U_x$ & 6.663054900E+02 & 6.663054900E+02 \\
-      \hline
-      Average $U_y$ & 1.833690840E-01 & 1.833690840E-01 \\
-      Minimum $U_y$ & -4.491052980E+02 & -4.491052980E+02 \\
-      Maximum $U_y$ & 2.474376340E+02 & 2.474376340E+02 \\
+      Average $U$ & 6.479541170E-01 & 6.479541170E-01 \\
+      Minimum $U$ & -2.550685400E+02 & -2.550685400E+02 \\
+      Maximum $U$ & 6.663054900E+02 & 6.663054900E+02 \\
+      \hline
+      Average $V$ & 1.833690840E-01 & 1.833690840E-01 \\
+      Minimum $V$ & -4.491052980E+02 & -4.491052980E+02 \\
+      Maximum $V$ & 2.474376340E+02 & 2.474376340E+02 \\
       \hline
@@
       \hline
-      Average $U_x$ & \\
-      Minimum $U_x$ & \\ 
-      Maximum $U_x$ & \\
-      \hline
-      Average $U_y$ & \\
-      Minimum $U_y$ & \\ 
-      Maximum $U_y$ & \\ 
+      Average $U$ & \\
+      Minimum $U$ & \\
+      Maximum $U$ & \\
+      \hline
+      Average $V$ & \\
+      Minimum $V$ & \\
+      Maximum $V$ & \\
       \hline
@@
       \hline
-      Average $U_x$ & \\
-      Minimum $U_x$ & \\
-      Maximum $U_x$ & \\
-      \hline
-      Average $U_y$ & \\
-      Minimum $U_y$ & \\
-      Maximum $U_y$ & \\
+      Average $U$ & \\
+      Minimum $U$ & \\
+      Maximum $U$ & \\
+      \hline
+      Average $V$ & \\
+      Minimum $V$ & \\
+      Maximum $V$ & \\
       \hline
@@
 %%---------------------------------------------------------------------------%%
-\section{Summary}
+\section{Summary\ }
 \label{sec:nonlinear_summary}

### 2013-07-22 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
working on frontmatter

@@
 %%---------------------------------------------------------------------------%%
-\section{Summary}
+\section{Summary\ }
 \label{sec:mc_summary}
@@
 \begin{itemize}
-\item MCSA can be used to solve the asymmetric system generated by the
-  multigroup $SP_N$ equations
+\item MCSA can solve the asymmetric system generated by the $SP_N$
+  equations
 \item MCSA has been incorporated into the Exnihilo neutronics
@@
 \item Simple Jacobi-based preconditioning reduces the spectral radius
-  of the $SP_N$ system below unity, however, this is not sufficient
-  for MCSA convergence
+  of the $SP_N$ system below unity, however, this alone is not
+  sufficient for MCSA convergence
 \item As the spectral radius approaches unity, MCSA was demonstrated
@@
   iteration than GMRES for the fuel assembly criticality problem and
-  more then Bi-CGStab
+  more than Bi-CGStab using the same preconditioning
 \item The explicit preconditioning strategy required to overcome the

### 2013-07-23 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
adding to nonlinear analysis

@@
 its applicability to fluid problems. For each case, we will compare
-the global minimum, maximum, and average values for fluid velocity,
-pressure, and temperature where applicable. If the values for these
-quantities computed using FANM match those computed using a production
-Newton-Krylov method, the implementation will be deemed correct for
-the purposes of this work. In addition, we will report the tolerance
-to which the nonlinear residual was converged as an additional means
-of comparison for correctness.
+the global minimum, maximum, and average values for fluid velocity and
+pressure for all benchmarks and additionally temperature for the
+thermal convection cavity problem. If the values for these quantities
+computed using FANM match those computed using a production
+Newton-Krylov method, the implementation will be deemed correct and
+the method deemed valid for the purposes of this work. In addition, we
+will report the tolerance to which the nonlinear residual was
+converged as an additional means of comparison for correctness.
 
@@
 each Newton step was preconditioned with an algebraic multigrid
-method, a common technique for preconditioning these equations
-\citep{ghia_high-re_1982,evans_enhanced_2007}. Without multigrid
-preconditioning, the residual in MCSA was observed to
-stagnate. Typically this happens because the Richardson iteration will
+method\footnote{The ML algebraic multigrid library in the Trilinos
+  scientific computing libraries \citep{heroux_overview_2005} was used
+  for all problems.}, a common technique for preconditioning these
+equations \citep{ghia_high-re_1982,evans_enhanced_2007}. Without
+multigrid preconditioning, the residual in MCSA was observed to
+stagnate. Typically this happens because the Richardson iteration and
+Neumann series on which the Monte Carlo method is based will quickly
 damp the higher order error modes on the scale of the computational
-grid used to solve the problem but will miss lower order error
-modes. In terms of Monte Carlo, this stagnation manifests itself in
-histories that simply do no terminate and therefore no solution is
-acquired.
-
-For FANM, this preconditioning was required to reduce the spectral
-radius of the Jacobian operators generated at each Newton step and for
-the Newton-Krylov solutions this preconditioning was required to
-obtain good convergence properties although it was not necessary for
-convergence. For every benchmark calculation, both FANM and
-Newton-Krylov were preconditioned with the same multigrid parameters
-such that the conditioning of the system would not be a factor in the
-verification of the method or the performance comparison in the
-following section. In addition, the same nonlinear solver parameters
-including globalization and forcing term strategies were used with
-both methods. As with the radiation transport calculations in the
-previous chapter, algrebraic preconditioning with the multigrid method
-results in dense linear systems. To mitigate this, the reduced domain
-approximation was again applied to the Monte Carlo solution within the
-MCSA solve occuring at each nonlinear iteration. The values used for
-the approximation will be given for each benchmark problem.
+grid used to solve the problem but will reduce lower order error modes
+at a considerably lower rate. In terms of Monte Carlo, this stagnation
+manifests itself in histories that simply do no terminate and
+therefore no solution is acquired.
+
+For every benchmark calculation, both FANM and Newton-Krylov were
+preconditioned with the same multigrid parameters such that the
+conditioning of the system would not be a factor in the verification
+of the method or the performance comparison in the following
+section. With FANM, this preconditioning was required to reduce the
+spectral radius of the Jacobian operators generated at each Newton
+step. This preconditioning was required to obtain good convergence
+properties for the Newton-Krylov solutions although it was not
+necessary for their convergence. In addition, the same nonlinear
+solver parameters including globalization and forcing term strategies
+were used with both methods. As with the radiation transport
+calculations in the previous chapter, algrebraic preconditioning with
+the multigrid method results in dense linear systems. To mitigate
+this, the reduced domain approximation was again applied to the Monte
+Carlo solution within the MCSA solve occuring at each nonlinear
+iteration. The values used for the approximation will be given for
+each benchmark problem.
 
@@
 A Drekar model of thermal convection cavity problem was solved using
-both the Newton-Krylov solver and a FANM solver for Rayleigh numbers
+both the Newton-Krylov solver and the FANM solver for Rayleigh numbers
 of \sn{1}{3}, \sn{1}{4}, \sn{1}{5} and \sn{1}{6} on a $50 \times 50$
@@
 was used as a means of globalization to assist the Newton correction
-computation while the forcing term was computed at each Newton step
-using method 2 from Shadid's work. In this forcing term method,
-nonlinear residuals from the two previous Newton steps are used to
-inform the choice of the forcing term for the next step.
+computation while the forcing term at all Rayleigh numbers but
+\sn{1}{6} was computed at each Newton step using method 2 from
+Shadid's work. In this forcing term method, nonlinear residuals from
+the two previous Newton steps are used to inform the choice of the
+forcing term for the next step. At a Rayleigh number of \sn{1}{6}, a
+constant forcing term of \sn{1}{-4} was use to give better convergence
+properties.
 
@@
 
-For the verification calculations, at each Rayleigh number variation
-the global minimum, maximum, and average values were recorded for
-fluid velocity in the x and y directions, fluid pressure, and fluid
+For the verification calculations, at each Rayleigh number the global
+minimum, maximum, and average values were recorded for fluid velocity
+in the x and y directions, fluid pressure, and fluid
 temperature. Tables~\ref{tab:convection_ra1e3_results},
@@
   \begin{center}
-    \begin{tabular}{lcc}\hline\hline
-      \multicolumn{1}{l}{Rayleigh Number}& 
+    \begin{tabular}{ccc}\hline\hline
+      \multicolumn{1}{c}{Rayleigh Number}& 
       \multicolumn{1}{c}{Newton Krylov $||\ve{F(\ve{u})}||$}&
@@
 A Drekar model of thermal convection cavity problem was solved using
-both the Newton-Krylov solver and a FANM solver for Reynolds numbers
+both the Newton-Krylov solver and the FANM solver for Reynolds numbers
 of 100, 300, 500, and 700 on a $50 \times 50$ square
 grid. Table~\ref{tab:lid_driven_parameters} gives the physical
-parameters of the problem and
-Table~\ref{tab:driven_mcsa_parameters} gives the MCSA solver
-parameters used within the FANM method. Variation in the Reynolds
-number of the system was achieved by keeping the parameters given in
-Table~\ref{tab:lid_driven_parameters} fixed and varying the
-fluid velocity in the $x$ direction on the top boundary. In addition
-to multigrid preconditioning for the linear solution at each Newton
-step, a polynomial line search method was used as a means of
-globalization assisting the Newton correction computation as outlined
-by Pawlowski in \citep{pawlowski_globalization_2006} while the forcing
-term was again computed using method 2 from Shadid's work
-\citep{shadid_inexact_1997}.
+parameters of the problem and Table~\ref{tab:driven_mcsa_parameters}
+gives the MCSA solver parameters used within the FANM
+method. Variation in the Reynolds number of the system was achieved by
+keeping the parameters given in Table~\ref{tab:lid_driven_parameters}
+fixed and varying the fluid velocity in the $x$ direction on the top
+boundary. In addition to multigrid preconditioning for the linear
+solution at each Newton step, a polynomial line search method was used
+as a means of globalization assisting the Newton correction
+computation as outlined by Pawlowski in
+\citep{pawlowski_globalization_2006} while the forcing term was again
+computed using method 2 from Shadid's work \citep{shadid_inexact_1997}
+at all Reynolds numbers.
 
@@
 
-For the verification calculations, at each Reynolds number variation
-the global minimum, maximum, and average values were recorded for the
-fluid velocity in the x and y directions and the fluid
+For the verification calculations, at each Reynolds number the global
+minimum, maximum, and average values were recorded for the fluid
+velocity in the x and y directions and the fluid
 pressure. Tables~\ref{tab:driven_re100_results},
@@
 A Drekar model of the backward facing step problem was solved using
-both the Newton-Krylov solver and a FANM solver for Reynolds numbers
+both the Newton-Krylov solver and the FANM solver for Reynolds numbers
 of 200, 300, 400, and 500 on a $20 \times 250$ rectangular
@@
       Histories, Re=500 & 1,000,000 per iteration \\
-      Reduced Domain Fill Level, Re=200 & 200 \\
-      Reduced Domain Fill Level, Re=400 & 200 \\
-      Reduced Domain Fill Level, Re=600 & 200 \\
-      Reduced Domain Fill Level, Re=800 & 300 \\
+      Reduced Domain Fill Level, All Re & 200 \\
       %%
@@
       reduction techniques were used with MCSA other than the reduced
-      domain approximation. Different values of fill level for the
-      reduced domain approximation and histories per iteration were
-      used at different Reynolds numbers due to convergence
-      requirements.}}
+      domain approximation. The same fill level for the reduced domain
+      approximation was used at every Reynold numbers and different
+      histories per iteration were used at different Reynolds numbers
+      due to convergence requirements.}}
   \label{tab:step_mcsa_parameters}
@@
 
-For the verification calculations, at each Reynolds number variation
-the global minimum, maximum, and average values were recorded for the
-fluid velocity in the x and y directions and the fluid
+For the verification calculations, at each Reynolds number the global
+minimum, maximum, and average values were recorded for the fluid
+velocity in the x and y directions and the fluid
 pressure. Tables~\ref{tab:step_re200_results},
 \ref{tab:step_re300_results}, \ref{tab:step_re400_results}, and
-\ref{tab:step_re500_results} give the results of these
-computations. As with the thermal convection problem, the negative
-pressures computed are due to a relative pressure being reported. For
-each of the data points given, both Newton-Krylov solutions and FANM
-solutions computed identical numerical results for all of these global
-values. The norm of the converged nonlinear residual for the
-calculations is given in
-Table~\ref{tab:step_residual_norm_comparison}. All computed global
-values were within these tolerances.
+\ref{tab:step_re500_results} give the results of these computations.
+The norm of the converged nonlinear residual for the calculations is
+given in Table~\ref{tab:step_residual_norm_comparison}. The
+Newton-Krylov solutions and FANM solutions computed identical
+numerical results for all of these global values within converged
+tolerances.
 
@@
       \hline
-      Average $U$ & \\
-      Minimum $U$ & \\
-      Maximum $U$ & \\
-      \hline
-      Average $V$ & \\
-      Minimum $V$ & \\
-      Maximum $V$ & \\
-      \hline
-      Average $P$ & \\ 
-      Minimum $P$ & \\ 
-      Maximum $P$ & \\ 
+      Average $U$ & 8.776187720E+01 & 8.776187720E+01 \\
+      Minimum $U$ & -2.796828970E+01 & -2.796828970E+01 \\
+      Maximum $U$ & 2.639993990E+02 & 2.639993990E+02 \\
+      \hline
+      Average $V$ & -8.870766180E-01 & -8.870766180E-01 \\
+      Minimum $V$ & -2.636357380E+01 & -2.636357380E+01 \\
+      Maximum $V$ & 1.330345700E+01 & 1.330345700E+01 \\
+      \hline
+      Average $P$ & 1.226960220E+04 & 1.226960220E+04 \\ 
+      Minimum $P$ & 2.953808400E+01 & 2.953808400E+01 \\ 
+      Maximum $P$ & 2.202121190E+04 & 2.202121190E+04 \\ 
       %%
@@
       \hline
-      Average $U$ & \\
-      Minimum $U$ & \\
-      Maximum $U$ & \\
-      \hline
-      Average $V$ & \\
-      Minimum $V$ & \\
-      Maximum $V$ & \\
-      \hline
-      Average $P$ & \\
-      Minimum $P$ & \\
-      Maximum $P$ & \\
+      Average $U$ & 1.316701890E+02 & 1.316701890E+02 \\
+      Minimum $U$ & -4.357431070E+01 & -4.357431070E+01 \\
+      Maximum $U$ & 3.958553050E+02 & 3.958553050E+02 \\
+      \hline
+      Average $V$ & -1.330267460E+00 & -1.330267460E+00 \\
+      Minimum $V$ & -3.467114630E+01 & -3.467114630E+01 \\
+      Maximum $V$ & 1.694243270E+01 & 1.694243270E+01 \\
+      \hline
+      Average $P$ & 1.681552980E+04 & 1.681552980E+04 \\
+      Minimum $P$ & 4.129311640E+01 & 4.129311640E+01 \\
+      Maximum $P$ & 2.940820620E+04 & 2.940820620E+04 \\
       %%
@@
       \hline
-      Average $U$ & \\
-      Minimum $U$ & \\
-      Maximum $U$ & \\
-      \hline
-      Average $V$ & \\
-      Minimum $V$ & \\
-      Maximum $V$ & \\
-      \hline
-      Average $P$ & \\
-      Minimum $P$ & \\
-      Maximum $P$ & \\
+      Average $U$ & 1.759720450E+02 & 1.759720450E+02 \\
+      Minimum $U$ & -5.941731260E+01 & -5.941731260E+01 \\
+      Maximum $U$ & 5.287146290E+02 & 5.287146290E+02 \\
+      \hline
+      Average $V$ & -1.776706740E+00 & -1.776706740E+00 \\
+      Minimum $V$ & -4.497362670E+01 & -4.497362670E+01 \\
+      Maximum $V$ & 1.973372840E+01 & 1.973372840E+01 \\
+      \hline
+      Average $P$ & 1.954352940E+04 & 1.954352940E+04 \\
+      Minimum $P$ & 5.294095080E+01 & 5.294095080E+01 \\
+      Maximum $P$ & 3.457969480E+04 & 3.457969480E+04 \\
       %%
@@
       \hline
-      Average $U$ & \\
-      Minimum $U$ & \\
-      Maximum $U$ & \\
-      \hline
-      Average $V$ & \\
-      Minimum $V$ & \\
-      Maximum $V$ & \\
-      \hline
-      Average $P$ & \\
-      Minimum $P$ & \\
-      Maximum $P$ & \\
+      Average $U$ & 2.199858290E+02 & 2.199858290E+02 \\
+      Minimum $U$ & -7.549221730E+01 & -7.549221730E+01 \\
+      Maximum $U$ & 6.605863330E+02 & 6.605863330E+02 \\
+      \hline
+      Average $V$ & -2.220108250E+00 & -2.220108250E+00 \\
+      Minimum $V$ & -5.888473080E+01 & -5.888473080E+01 \\
+      Maximum $V$ & 2.201938360E+01 & 2.201938360E+01 \\
+      \hline
+      Average $P$ & 1.983947940E+04 & 1.983947940E+04 \\
+      Minimum $P$ & -9.707913370E+03 & -9.707913370E+03 \\
+      Maximum $P$ & 3.771193860E+04 & 3.771193860E+04 \\
       %%
@@
   \begin{center}
-    \begin{tabular}{lcc}\hline\hline
-      \multicolumn{1}{l}{Reynolds Number}& 
+    \begin{tabular}{ccc}\hline\hline
+      \multicolumn{1}{c}{Reynolds Number}& 
       \multicolumn{1}{c}{Newton Krylov $||\ve{F(\ve{u})}||$}&
@@
       \hline
-      200 & \\
-      300 & \\
-      400 & \\
-      500 & \\
+      200 & 1.939E-11 & 1.975E-11 \\
+      300 & 2.786E-11 & 2.751E-11 \\
+      400 & 3.360E-11 & 3.597E-11 \\
+      500 & 5.824E-11 & 4.062E-11 \\
       %%
@@
   \begin{center}
-    \begin{tabular}{lcc}\hline\hline
-      \multicolumn{1}{l}{Rayleigh Number}& 
+    \begin{tabular}{ccc}\hline\hline
+      \multicolumn{1}{c}{Rayleigh Number}& 
       \multicolumn{1}{c}{Newton-Krylov Iterations}&
@@
   \begin{center}
-    \begin{tabular}{lcc}\hline\hline
-      \multicolumn{1}{l}{Rayleigh Number}& 
+    \begin{tabular}{ccc}\hline\hline
+      \multicolumn{1}{c}{Rayleigh Number}& 
       \multicolumn{1}{c}{GMRES Iterations}&
@@
   \begin{center}
-    \begin{tabular}{lc}\hline\hline
-      \multicolumn{1}{l}{Rayleigh Number}& 
+    \begin{tabular}{cc}\hline\hline
+      \multicolumn{1}{c}{Rayleigh Number}& 
       \multicolumn{1}{c}{Newton-Krylov Speedup}\\
@@
   \begin{center}
-    \begin{tabular}{lcc}\hline\hline
-      \multicolumn{1}{l}{Reynolds Number}& 
+    \begin{tabular}{ccc}\hline\hline
+      \multicolumn{1}{c}{Reynolds Number}& 
       \multicolumn{1}{c}{Newton-Krylov Iterations}&
@@
   \begin{center}
-    \begin{tabular}{lcc}\hline\hline
-      \multicolumn{1}{l}{Reynolds Number}& 
+    \begin{tabular}{ccc}\hline\hline
+      \multicolumn{1}{c}{Reynolds Number}& 
       \multicolumn{1}{c}{GMRES Iterations}&
@@
   \begin{center}
-    \begin{tabular}{lc}\hline\hline
-      \multicolumn{1}{l}{Reynolds Number}& 
+    \begin{tabular}{cc}\hline\hline
+      \multicolumn{1}{c}{Reynolds Number}& 
       \multicolumn{1}{c}{Newton-Krylov Speedup}\\
@@
       \hline
-      200 & \\
-      400 & \\
-      600 & \\
-      800 & \\
+      200 & 10 & 9 \\
+      400 & 15 & 14 \\
+      600 & 10 & 10 \\
+      800 & 19 & 20 \\
       %%
@@
       \hline
-      200 & \\
-      400 & \\
-      600 & \\
-      800 & \\
+      200 & 24 & 13 \\
+      400 & 23 & 17 \\
+      600 & 18 & 12 \\
+      800 & 30 & 52 \\
       %%
@@
   \begin{center}
-    \begin{tabular}{lc}\hline\hline
-      \multicolumn{1}{l}{Reynolds Number}& 
+    \begin{tabular}{cc}\hline\hline
+      \multicolumn{1}{c}{Reynolds Number}& 
       \multicolumn{1}{c}{Newton-Krylov Speedup}\\
       \hline
-      200 & \\
-      400 & \\
-      600 & \\
-      800 & \\
+      200 & 400 \\
+      400 & 593 \\
+      600 & 825 \\
+      800 & 1057 \\
       %%
@@
   \label{tab:step_speedup_comparison}
+\end{table}
+
+\clearpage
+
+\subsection{Overall Performance Comparison}
+\label{subsec:nonlinear_overall_comparison}
+
+As a final overall comparison, we tabulate the results from the
+benchmark calculations over the three performance metrics: nonlinear
+iterations, linear iterations, and CPU
+time. Tables~\ref{tab:benchmark_nonlinear_comparison},
+\ref{tab:benchmark_linear_comparison}, and
+\ref{tab:benchmark_time_comparison} give the results. In these tables,
+the calculations that procduced the best performing results are
+tallied. In the case of a tie, both are tallied. Looking at the
+tabulated results, we see marginally better performance for FANM in
+terms of nonlinear iterations with 1 extra case outperforming the
+Newton-Krylov method although this is not conclusive enough to claim
+superior performance converging the nonlinear residual with FANM. As
+observed for each benchmark, the timing results were always in favor
+of the Newton-Krylov method due to the effects of the explict
+preconditioning strategy.
+
+For the total linear solver iterations given in
+Table~\ref{tab:benchmark_linear_comparison}, the difference is more
+dramatic with FANM converging in fewer the nonlinear residual in fewer
+linear solver iterations for twice as many cases as the Newton-Krylov
+method keeping in mind that for all cases both GMRES and MCSA were
+preconditioned with identical multigrid parameters using the same
+multigrid preconditioning library. This shows that when properly
+preconditioned, we can expect performance improvements in terms of
+total linear solver iterations required to converge the nonliner
+residual. With future optimization of MCSA implementations, this may
+also mean improved time to solution for FANM when compared to
+Newton-Krylov methods.
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lcc}\hline\hline
+      \multicolumn{1}{l}{Benchmark}& 
+      \multicolumn{1}{c}{Newton-Krylov}&
+      \multicolumn{1}{c}{FANM}\\
+      \hline
+      Convection, Ra=\sn{1}{3} & $\times$ & $\times$ \\
+      Convection, Ra=\sn{1}{4} & $\times$ & $\times$ \\
+      Convection, Ra=\sn{1}{5} & $\times$ & \\
+      Convection, Ra=\sn{1}{6} & $\times$ & $\times$ \\
+      Lid Driven, Re=100 & $\times$ & $\times$ \\
+      Lid Driven, Re=300 & $\times$ & $\times$ \\
+      Lid Driven, Re=500 & $\times$ & $\times$ \\
+      Lid Driven, Re=700 & & $\times$ \\
+      Backward Step, Re=200 & & $\times$ \\
+      Backward Step, Re=300 & & $\times$ \\
+      Backward Step, Re=400 & $\times$ & $\times$ \\
+      Backward Step, Re=500 & $\times$ & \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Navier-Stokes benchmark comparison for nonlinear
+      iterations.} \textit{Over all benchmarks, FANM performed better
+      in terms of nonlinear iterations for 1 more case than the
+      Newton-Krylov method.}}
+  \label{tab:benchmark_nonlinear_comparison}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lcc}\hline\hline
+      \multicolumn{1}{l}{Benchmark}& 
+      \multicolumn{1}{c}{Newton-Krylov}&
+      \multicolumn{1}{c}{FANM}\\
+      \hline
+      Convection, Ra=\sn{1}{3} & & $\times$ \\
+      Convection, Ra=\sn{1}{4} & & $\times$ \\
+      Convection, Ra=\sn{1}{5} & & $\times$ \\
+      Convection, Ra=\sn{1}{6} & & $\times$ \\
+      Lid Driven, Re=100 & $\times$ & \\
+      Lid Driven, Re=300 & $\times$ & \\
+      Lid Driven, Re=500 & $\times$ & \\
+      Lid Driven, Re=700 & & $\times$ \\
+      Backward Step, Re=200 & & $\times$ \\
+      Backward Step, Re=300 & & $\times$ \\
+      Backward Step, Re=400 & & $\times$ \\
+      Backward Step, Re=500 & $\times$ & \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Navier-Stokes benchmark comparison for total linear
+      solver iterations.} \textit{Over all benchmarks, FANM performed
+      better in terms of linear solver iterations for twice as many
+      cases as the Newton-Krylov method.}}
+  \label{tab:benchmark_linear_comparison}
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lcc}\hline\hline
+      \multicolumn{1}{l}{Benchmark}& 
+      \multicolumn{1}{c}{Newton-Krylov}&
+      \multicolumn{1}{c}{FANM}\\
+      \hline
+      Convection, Ra=\sn{1}{3} & $\times$ & \\
+      Convection, Ra=\sn{1}{4} & $\times$ & \\
+      Convection, Ra=\sn{1}{5} & $\times$ & \\
+      Convection, Ra=\sn{1}{6} & $\times$ & \\
+      Lid Driven, Re=100 & $\times$ & \\
+      Lid Driven, Re=300 & $\times$ & \\
+      Lid Driven, Re=500 & $\times$ & \\
+      Lid Driven, Re=700 & $\times$ & \\
+      Backward Step, Re=200 & $\times$ & \\
+      Backward Step, Re=300 & $\times$ & \\
+      Backward Step, Re=400 & $\times$ & \\
+      Backward Step, Re=500 & $\times$ & \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Navier-Stokes benchmark comparison for CPU time.}
+    \textit{For all benchmarks the explicit MCSA preconditioning
+      strategy caused signifcantly larger CPU times for FANM when
+      compared to the Newton-Krylov solutions.}}
+  \label{tab:benchmark_time_comparison}
 \end{table}
@@
   the problem
-\item With future optimization, reactor accident analyses that rely on
-  convection dominated transport or steady state calculations
-  requiring flow at low Reynolds numbers could experience improved
-  time to solution with FANM given the improved iterative performance
-  observed for problems of these types
+\item With future optimization and improved preconditioning, reactor
+  accident analyses that rely on convection dominated transport or
+  steady state calculations requiring driven flow could experience
+  improved time to solution with FANM given the improved iterative
+  performance observed for problems of these types
 \end{itemize}

### 2013-07-23 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
adding figures for nonlinear annalysis

@@
   for all problems.}, a common technique for preconditioning these
-equations \citep{ghia_high-re_1982,evans_enhanced_2007}. Without
-multigrid preconditioning, the residual in MCSA was observed to
-stagnate. Typically this happens because the Richardson iteration and
-Neumann series on which the Monte Carlo method is based will quickly
-damp the higher order error modes on the scale of the computational
-grid used to solve the problem but will reduce lower order error modes
-at a considerably lower rate. In terms of Monte Carlo, this stagnation
-manifests itself in histories that simply do no terminate and
-therefore no solution is acquired.
+equations \citep{ghia_high-re_1982,evans_enhanced_2007} with the
+preconditioner reconstructed at every Newton step for the new
+Jacobian. Without multigrid preconditioning, the residual in MCSA was
+observed to stagnate. Typically this happens because the Richardson
+iteration and Neumann series on which the Monte Carlo method is based
+will quickly damp the higher order error modes on the scale of the
+computational grid used to solve the problem but will reduce lower
+order error modes at a considerably lower rate. In terms of Monte
+Carlo, this stagnation manifests itself in histories that simply do no
+terminate and therefore no solution is acquired.
 
@@
 method. For each benchmark, we will analyze the iterative performance
-of both the Newton solver and the linear solver used to compute the
-correction vector at each step. In addition, timing results will be
-discussed.
+of both the nonlinear Newton solver and the linear solver used to
+compute the Newton correction vector at each step. Limiting the number
+of nonlinear iterations required for solution is critical for
+efficiency as nonlinear function evaluations, Jacobian generation, and
+preconditioner construction are typically required at every Newton
+iteration. In addition, timing results will also be discussed.
 
@@
 the number of Newton iterations required for convergence for each
-method. Limiting the number of nonlinear iterations required for
-solution is critical for efficiency as nonlinear function evaluations,
-Jacobian generation, and preconditioner construction are typically
-required at every Newton iteration. For all Rayleigh numbers except
-\sn{1}{5} the number of Newton iterations required to converge is the
-same while the \sn{1}{5} result required an extra FANM iteration. This
-is due to the fact that the convergence of the linear residual has a
-strong effect on the convergence of the nonlinear residual. For
-nonlinear calculations, this is much stronger than the effect on the
-convergence of the eigenvalue in the $SP_N$ computations where we
-always observed that identical numbers of k-eigenvalue iterations were
-required to converge each variation of the probem.
+method.  For all Rayleigh numbers except \sn{1}{5} the number of
+Newton iterations required to converge is the same while the \sn{1}{5}
+result required an extra FANM iteration. This is due to the fact that
+the convergence of the linear residual has a strong effect on the
+convergence of the nonlinear residual. For nonlinear calculations,
+this is much stronger than the effect on the convergence of the
+eigenvalue in the $SP_N$ computations where we always observed that
+identical numbers of k-eigenvalue iterations were required to converge
+each variation of the probem.
 
@@
 difficulty even though it is preconditioned in an identical
-manner. The more ill-conditioned the linear model is, the closer the
-linear residual is converged to the requested tolerance whereas a
-better conditioned problem may produce a linear solution that
-overshoots the specified tolerance provided by the forcing term. When
-oversolving the linear model is not an issue, this results in faster
-convergence of the nonlinear problem.
+manner.
 
@@
 solver at each Newton step is converging the linear residual to a
-smaller tolerance than the MCSA solver, even though they have
-approximately the same forcing terms to begin with. As the Rayleigh
-number increases, the problem becomes more difficult to solve and
-therefore a more agressive preconditioner should maintain iterative
-performance of the nonlinear method. This was observed to be true for
-the \sn{1}{6} Rayleigh number case. For this calculation, the number
-of multigrid cycle applications was bumped up to 3 from 1 and FANM
-converged in the same number of nonlinear iterations as the
-Newton-Krylov method.
+smaller tolerance than the MCSA solver as the forcing terms adapt to
+the convergence of the method over the course of nonlinear
+iterations. As the Rayleigh number increases, the problem becomes more
+difficult to solve and therefore a more agressive preconditioner
+should maintain iterative performance of the nonlinear method. This
+was observed to be true for the \sn{1}{6} Rayleigh number case. For
+this calculation, the number of multigrid cycle applications was
+bumped up to 3 from 1 and FANM converged in the same number of
+nonlinear iterations as the Newton-Krylov method.
 
@@
 package and the same parameters. For the \sn{1}{6} Rayleigh number
-case this difference is most drastic with FANM requiring only 64\% of
-the linear iterations required by the Newton-Krylov scheme. Even the
-\sn{1}{5} Rayleigh number case that required more Newton iterations
-required 5 fewer MCSA iterations to converge than GMRES
-iterations. If the performance of the MCSA implementation used here
-were to be potentially optimized in the future, this improved
-iterative performance could also mean improved time to solution.
+case where a constant forcing term of \sn{1}{-4} was kept, this
+difference is most drastic with FANM requiring only 64\% of the linear
+iterations required by the Newton-Krylov scheme with the linear solver
+per nonlinear iteration given for this case in
+Figure~\ref{fig:ra1e6_linear_iters}. Even the \sn{1}{5} Rayleigh
+number case that required more Newton iterations required 5 fewer MCSA
+iterations to converge than GMRES iterations. If the performance of
+the MCSA implementation used here were to be potentially optimized in
+the future, this improved iterative performance could also mean
+improved time to solution.
 
@@
 \end{table}
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/nonlinear_problem/convection_ra1e6_iters.pdf}
+  \end{center}
+  \caption{\textbf{Linear solver iterations as a function of Newton
+      iteration for the \sn{1}{6} Rayleigh number case.} \textit{MCSA
+      shows better iterative performance than GMRES at all but the
+      second nonlinear iteration. The forcing term was fixed at
+      \sn{1}{-4} at all nonlinear iterations.}}
+  \label{fig:ra1e6_linear_iters}
+\end{figure}
 
@@
 
-These iterative results again point out the fundamental flaw of
-requiring a spectral radius of less than unity for the MCSA iteration
-to converge at each Newton step. As the problem becomes more dominated
-by inertial forces, the resulting linear models become increasingly
-stiff and difficult to precondition. As a result, convergence requires
-many more iterations than needed for the flow problem driven by
-convection in the previous benchmark if convergence is achieved at
-all. Much of this can be attributed to the physics at play in each of
-these benchmarks. For the thermal convection cavity benchmark, the
-forces at work in that problem were more effectively resolved by the
-Monte Carlo method as they are more diffusive in character than the
-flow driven by intertial forces in the lid driven cavity problem. This
-physics-based observation was also true for the $SP_N$ equations where
-the diffusion dominated physics in that case gave better convergence
-results for MCSA than GMRES.
+As an additional means of analysis for the linear solver iteration
+results, we will look at a case where MCSA performed poorly compared
+to GMRES. Consider the linear iterations as a function of nonlinear
+iteration given in Figure~\ref{fig:driven_re500_iters} and the forcing
+term as a function of nonlinear iteration given in
+Figure~\ref{fig:driven_re500_forcing} for the 500 Reynolds number
+case. At the later Newton iterations where the region of fast
+quadratic convergence is reached, we see that MCSA required many more
+iterations to converge than GMRES, accounting for the majority of the
+extra iterations. However, we see in the forcing term data that
+forcing terms at later Newton iterations for MCSA were much smaller
+than those for GMRES. This means that MCSA was converged to a tighter
+tolerance at these iterations and therfore required the extra
+iterations.
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/nonlinear_problem/driven_re500_iters.pdf}
+  \end{center}
+  \caption{\textbf{Linear solver iterations as a function of Newton
+      iteration for the 500 Reynolds number case.} \textit{More MCSA
+      iterations were required to converge the linear models at later
+      Newton iterations due to smaller forcing terms.}}
+  \label{fig:driven_re500_iters}
+\end{figure}
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/nonlinear_problem/driven_re500_forcing.pdf}
+  \end{center}
+  \caption{\textbf{Forcing terms as a function of Newton
+      iteration for the 500 Reynolds number case.} \textit{Once the
+      region of fast quadratic convergence is reached, MCSA forcing
+      terms were much larger.}}
+  \label{fig:driven_re500_forcing}
+\end{figure}
 
@@
 
+The backward step problem exhibited different performance
+characteristics than the other two benchmark problems, likely due to
+the different geometry and boundary
+conditions. Table~\ref{tab:step_nonlinear_iter_comparison} gives the
+nonlinear iteration results for this benchmark. At low Reynolds
+numbers, FANM converges in fewer nonlinear iterations than the
+Newton-Krylov solver. However, as the problem becomes more difficult
+to solver at higher rates of flow, the Newton-Krylov solver eventually
+converges in fewer nonlinear iterations. We see the same behavior as
+well in the linear solver iteration data given by
+Table~\ref{tab:step_linear_iter_comparison}. At the 3 lowest Reynolds
+numbers, FANM required significantly fewer linear solver iterations to
+converge the problem while the ill-conditioned case at a Reynolds
+number of 500 is observed to use many more MCSA iterations.
+
 \begin{table}[h!]
@@
       200 & 10 & 9 \\
-      400 & 15 & 14 \\
-      600 & 10 & 10 \\
-      800 & 19 & 20 \\
+      300 & 15 & 14 \\
+      400 & 10 & 10 \\
+      500 & 19 & 20 \\
       %%
@@
       200 & 24 & 13 \\
-      400 & 23 & 17 \\
-      600 & 18 & 12 \\
-      800 & 30 & 52 \\
+      300 & 23 & 17 \\
+      400 & 18 & 12 \\
+      500 & 30 & 52 \\
       %%
@@
 
+Unlike the lid driven cavity benchmark, the extra linear solver
+iterations are not due to the smaller forcing terms but are rather
+truly a function of the ill-conditioning of the Jacobian matrices
+produced at each Newton step.
+
 \begin{table}[h!]
@@
       200 & 400 \\
-      400 & 593 \\
-      600 & 825 \\
-      800 & 1057 \\
+      300 & 593 \\
+      400 & 825 \\
+      500 & 1057 \\
       %%
@@
 terms of nonlinear iterations with 1 extra case outperforming the
-Newton-Krylov method although this is not conclusive enough to claim
-superior performance converging the nonlinear residual with FANM. As
-observed for each benchmark, the timing results were always in favor
-of the Newton-Krylov method due to the effects of the explict
-preconditioning strategy.
+Newton-Krylov method. As observed for each benchmark, the timing
+results were always in favor of the Newton-Krylov method due to the
+effects of the explict preconditioning strategy.
 

### 2013-07-23 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
finished add step problem to nonlinear chapter

@@
 equations to the following three standard benchmark problems that
-cover convection dominated flow and flow dominated by intertial
+cover convection dominated flow and flow dominated by inertial
 forces in two different geometries.
@@
 \label{subsec:lid_driven_cavity}
-To study performance for flow driven by intertial forces, the second
+To study performance for flow driven by inertial forces, the second
 benchmark problem given by Ghia \citep{ghia_high-re_1982} incorporates
@@
 solve. More rotation of the fluid is induced by inertial forces and we
-begin to see some additional vortices form besides the primary vortex
+begin to see some additional vortexes form besides the primary vortex
 near the center of the box.
@@
   \begin{center}
-    \includegraphics[width=6in]{chapters/nonlinear_problem/driven_velocity_isocurves.png}
+    \includegraphics[width=6in]{chapters/nonlinear_problem/step_velocity_isocurves.png}
   \end{center}
@@
       scale from 0 to 1 with 20 uniformly spaced curves for each
-      calculation.}}
+      calculation. The depression forming in the upper right portion
+      of visible flow is the second vortex.}}
   \label{fig:step_velocity_isocurves}
@@
 were used with both methods. As with the radiation transport
-calculations in the previous chapter, algrebraic preconditioning with
+calculations in the previous chapter, algebraic preconditioning with
 the multigrid method results in dense linear systems. To mitigate
 this, the reduced domain approximation was again applied to the Monte
-Carlo solution within the MCSA solve occuring at each nonlinear
+Carlo solution within the MCSA solve occurring at each nonlinear
 iteration. The values used for the approximation will be given for
@@
 achieve these results, a Monte Carlo solver generated for this work
-was injected into the nonlinear toolchain within Drekar to form an
+was injected into the nonlinear tool chain within Drekar to form an
 implementation of the FANM method. For the Newton-Krylov solutions,
@@
       is varied by modifying the hot boundary temperature to induce
-      larger temperature gradients and bouyancy-driven flow.}}
+      larger temperature gradients and buoyancy-driven flow.}}
   \label{tab:thermal_convection_parameters}
@@
 convergence of the linear iterations could not be achieved regardless
-of how agressive the multigrid preconditioner was set. In these cases,
+of how aggressive the multigrid preconditioner was set. In these cases,
 the Jacobian operators are ill-conditioned enough that even GMRES
@@
       domain approximation. The same fill level for the reduced domain
-      approximation was used at every Reynold numbers and different
+      approximation was used at every Reynolds numbers and different
       histories per iteration were used at different Reynolds numbers
@@
 Reynolds numbers above 500 convergence of the linear iterations could
-not be achieved with FANM regardless of how agressive the multigrid
+not be achieved with FANM regardless of how aggressive the multigrid
 preconditioner was set. In these cases, the Jacobian operators are
@@
 identical numbers of k-eigenvalue iterations were required to converge
-each variation of the probem.
+each variation of the problem.
 
@@
 iterations. As the Rayleigh number increases, the problem becomes more
-difficult to solve and therefore a more agressive preconditioner
+difficult to solve and therefore a more aggressive preconditioner
 should maintain iterative performance of the nonlinear method. This
@@
   \end{center}
-  \caption{\textbf{Nonlinear residual norm as a function of Newton
-      iteration for the \sn{1}{5} Rayleigh number case.} \textit{The
-      FANM calculation required an additional nonlinear iteration to
-      converge.}}
+  \caption{\textbf{Thermal convection cavity nonlinear residual norm
+      as a function of Newton iteration for the \sn{1}{5} Rayleigh
+      number case.} \textit{The FANM calculation required an
+      additional nonlinear iteration to converge.}}
   \label{fig:ra1e5_convergence}
@@
   \end{center}
-  \caption{\textbf{Linear solver iterations as a function of Newton
-      iteration for the \sn{1}{6} Rayleigh number case.} \textit{MCSA
-      shows better iterative performance than GMRES at all but the
-      second nonlinear iteration. The forcing term was fixed at
-      \sn{1}{-4} at all nonlinear iterations.}}
+  \caption{\textbf{Thermal convection cavity linear solver iterations
+      as a function of Newton iteration for the \sn{1}{6} Rayleigh
+      number case.} \textit{MCSA shows better iterative performance
+      than GMRES at all but the second nonlinear iteration. The
+      forcing term was fixed at \sn{1}{-4} at all nonlinear
+      iterations.}}
   \label{fig:ra1e6_linear_iters}
@@
   \end{center}
-  \caption{\textbf{Nonlinear residual norm as a function of Newton
-      iteration for the 700 Reynolds number case.} \textit{The FANM
-      calculation converged in 4 fewer iterations than the
-      Newton-Krylov calculation.}}
+  \caption{\textbf{Lid driven cavity nonlinear residual norm as a
+      function of Newton iteration for the 700 Reynolds number case.}
+    \textit{The FANM calculation converged in 4 fewer iterations than
+      the Newton-Krylov calculation.}}
   \label{fig:re700_convergence}
@@
 than those for GMRES. This means that MCSA was converged to a tighter
-tolerance at these iterations and therfore required the extra
+tolerance at these iterations and therefore required the extra
 iterations.
@@
   \end{center}
-  \caption{\textbf{Linear solver iterations as a function of Newton
-      iteration for the 500 Reynolds number case.} \textit{More MCSA
-      iterations were required to converge the linear models at later
-      Newton iterations due to smaller forcing terms.}}
+  \caption{\textbf{Lid driven cavity linear solver iterations as a
+      function of Newton iteration for the 500 Reynolds number case.}
+    \textit{More MCSA iterations were required to converge the linear
+      models at later Newton iterations due to smaller forcing
+      terms.}}
   \label{fig:driven_re500_iters}
@@
   \end{center}
-  \caption{\textbf{Forcing terms as a function of Newton
-      iteration for the 500 Reynolds number case.} \textit{Once the
-      region of fast quadratic convergence is reached, MCSA forcing
-      terms were much larger.}}
+  \caption{\textbf{Lid driven cavity forcing terms as a function of
+      Newton iteration for the 500 Reynolds number case.} \textit{Once
+      the region of fast quadratic convergence is reached, MCSA
+      forcing terms were much larger.}}
   \label{fig:driven_re500_forcing}
@@
 thermal convection cavity problem and again in agreement with those
-observed for the $SP_N$ calculations. Lack of signficant variation in
-these timing results also point to a qualititively similar time
+observed for the $SP_N$ calculations. Lack of significant variation in
+these timing results also point to a qualitatively similar time
 complexity for the FANM method as a function of primary problem
 parameter relative to the Newton-Krylov method. In all cases, the
-preconditioning strategy must be signficantly improved along with
+preconditioning strategy must be significantly improved along with
 implementation optimization to avoid these large overheads.
@@
 truly a function of the ill-conditioning of the Jacobian matrices
-produced at each Newton step.
+produced at each Newton step. We see this by comparing the results at
+Reynolds numbers of 400 where FANM outperformed Newton-Krylov and 500
+where Newton-Krylov outperformed FANM. Linear iterations as a function
+of nonlinear iteration given in Figure~\ref{fig:step_re400_iters} and
+the forcing term as a function of nonlinear iteration given in
+Figure~\ref{fig:step_re400_forcing} for the 400 Reynolds number case
+and the same data is given in Figure~\ref{fig:step_re500_iters} and
+Figure~\ref{fig:step_re500_forcing} for the 500 Reynolds number
+case. For both cases, FANM always has larger forcing terms and should
+therefore use fewer MCSA iterations at each Newton step to converge
+the linear model. This is indeed the case for the 400 Reynolds number
+case where the linear model is better conditioned and FANM used fewer
+total linear solver iterations to converge the nonlinear
+residual. However, at a Reynolds number of 500 FANM has larger forcing
+terms but requires many more MCSA iterations to converge the last 2
+Newton steps. Therefore, these extra iterations in this case must come
+from ill-conditioning.
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/nonlinear_problem/step_re400_iters.pdf}
+  \end{center}
+  \caption{\textbf{Backward facing step linear solver iterations as a
+      function of Newton iteration for the 400 Reynolds number case.}
+    \textit{More MCSA iterations were required to converge the linear
+      models at later Newton iterations due to ill-conditioning of the
+      Jacobian matrices.}}
+  \label{fig:step_re400_iters}
+\end{figure}
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/nonlinear_problem/step_re400_forcing.pdf}
+  \end{center}
+  \caption{\textbf{Backward facing step forcing terms as a function of
+      Newton iteration for the 400 Reynolds number case.} \textit{FANM
+      forcing terms were always larger.}}
+  \label{fig:step_re400_forcing}
+\end{figure}
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/nonlinear_problem/step_re500_iters.pdf}
+  \end{center}
+  \caption{\textbf{Backward facing step linear solver iterations as a
+      function of Newton iteration for the 500 Reynolds number case.}
+    \textit{More MCSA iterations were required to converge the linear
+      models at later Newton iterations due to ill-conditioning of the
+      Jacobian matrices.}}
+  \label{fig:step_re500_iters}
+\end{figure}
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/nonlinear_problem/step_re500_forcing.pdf}
+  \end{center}
+  \caption{\textbf{Backward facing step forcing terms as a function of
+      Newton iteration for the 500 Reynolds number case.} \textit{More
+      MCSA iterations were required to converge the linear models at
+      later Newton iterations due to ill-conditioning of the Jacobian
+      matrices.}}
+  \label{fig:step_re500_forcing}
+\end{figure}
+
+The ill-conditioning of the 500 Reynolds number case is also apparent
+in the timing data given in
+Table~\ref{tab:step_speedup_comparison}. For the first two cases, we
+see qualitatively similar speedup values for the Newton-Krylov solver
+relative to the observations for the first two benchmark problems of
+O(100). However, the we expect the ill-conditioning to cause the Monte
+Carlo histories in the MCSA solver to have longer random walk lengths
+based on our MCSA breakdown analysis in the previous chapter. In
+addition, we observed that the linear model required many more MCSA
+iterations to converge. These longer random walks and more iterations
+manifest themselves in the longer CPU time values and therefore the
+larger Newton Krylov speedup value of O(1,000) for the 500 Reynolds
+number case.
 
@@
 \ref{tab:benchmark_time_comparison} give the results. In these tables,
-the calculations that procduced the best performing results are
+the calculations that produced the best performing results are
 tallied. In the case of a tie, both are tallied. Looking at the
@@
 results were always in favor of the Newton-Krylov method due to the
-effects of the explict preconditioning strategy.
+effects of the explicit preconditioning strategy.
 
@@
 preconditioned, we can expect performance improvements in terms of
-total linear solver iterations required to converge the nonliner
+total linear solver iterations required to converge the nonlinear
 residual. With future optimization of MCSA implementations, this may
@@
     \textit{For all benchmarks the explicit MCSA preconditioning
-      strategy caused signifcantly larger CPU times for FANM when
+      strategy caused significantly larger CPU times for FANM when
       compared to the Newton-Krylov solutions.}}
@@
 In this chapter we have developed and explored Monte Carlo synthetic
-acceleration methods in the context of solutions to the nonlinear
-Navier-Stokes equations. The following are the signficant observations
-and findings.
+acceleration methods in the context of solutions to the Navier-Stokes
+equations. The following are the significant observations and findings.
 
@@
 \item The Newton-Krylov method has better iterative performance than
-  the FANM method for flow dominated by interial forces, converging in
+  the FANM method for flow dominated by inertial forces, converging in
   fewer linear solver iterations with the same preconditioning for
-  high and low Reynolds numbers
+  high Reynolds numbers
 \item The spectral radius convergence restriction on MCSA was observed
-  to be a signficant hinderance by preventing solutions to forced flow
+  to be a significant hindrance by preventing solutions to forced flow
   problems at high Reynolds numbers
-\item Explict algebraic preconditioning used to overcome the spectral
+\item Explicit algebraic preconditioning used to overcome the spectral
   radius restriction of MCSA creates FANM run times

### 2013-07-23 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
working on frontmatter
finished add step problem to nonlinear chapter

@@
 %%---------------------------------------------------------------------------%%
-\section{Summary}
+\section{Summary\ }
 \label{sec:parallel_summary}
@@
 qualities of the algorithm through parallel scaling studies. The
-following are the signficant observations and findings.
+following are the significant observations and findings.
 
@@
   in strong scaling cases but is ineffective in weak scaling cases
-\item Multiple sets will never boost parallel efficienicies in both
+\item Multiple sets will never boost parallel efficiencies in both
   weak and strong scaling cases due to additional communication

### 2013-07-23 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
finished add step problem to nonlinear chapter

@@
 severely inhibits performance where now parallel operations require
-communications amongst orders of magnitude more processors for
+communications among orders of magnitude more processors for
 nearest-neighbor type algorithms.
@@
 entries to a minimum value in order to lower the amount of coupling
-amongst states in the system and reduce the amount of memory used
+among states in the system and reduce the amount of memory used
 along with the potential latency overhead. Given these two objectives,
@@
 equation and explored in the context of a difficult nuclear fuel
-assembly criticality problem. The following are the signficant
+assembly criticality problem. The following are the significant
 observations and findings.

### 2013-07-23 doc/dissertation/document/chapters/introduction/introduction.tex
working on refactoring the introduction chapter

@@
 problems make up much of the analysis suite used for modern
-analysis. Recent focus on multiple physics systems in the nuclear
-reactor modeling and simulation community adds a new level of
-complexity to common linear and nonlinear systems as solution
+engineering calculations. Recent focus on multiple physics systems in
+the nuclear reactor modeling and simulation community adds a new level
+of complexity to common linear and nonlinear systems as solution
 strategies change when they are coupled to other problems
@@
 numbers of cores will be pertinent to robust algorithms aimed at this
-new hardware. Considering these issues, this dissertation develops a
-massively parallel Monte Carlo method for linear problems and a novel
-Monte Carlo method to advance solution techniques for nonlinear
-problems.
-
-This goal of this work is to research and develop parallel solutions
-for discrete linear and nonlinear transport problems that improve
-performance and scalabilty using a novel set of Monte Carlo synthetic
-acceleration methods.
+new hardware. Considering these issues, this work researches and
+develops parallel Monte Carlo Synthetic Acceleration methods for
+linear problems and a new method based on the synthetic acceleration
+techniques to advance solution algorithms for nonlinear problems.
 
 We discuss in this chapter physics-based motivation for advancing
-Monte Carlo synthetic acceleration methods and studying their
+Monte Carlo Synthetic Acceleration methods and studying their
 application to transport systems by providing problems of interest in
 nuclear reactor analysis. Hardware-based motivations are also provided
-by considering the impact of forthcoming computing architectures. In
-addition, background on the current solver techniques for these
-physics problems and a brief comparison to the proposed methods are
-provided to further motivate this work. The organization of the
-document is then reviewed.
+by considering the potential impact of these methods on forthcoming
+computing architectures. In addition, background on the current solver
+techniques for these physics problems and a brief comparison to the
+proposed methods are provided to further motivate this work. Finally,
+the statement of work is given and the organization of the document is
+then reviewed.
 
@@
 \end{figure}
+
 When this event occurs, heat transfer is greatly reduced between the
@@
 how changing material temperatures and densities changes the cross
-sections needed for the neutronics calculation. As shown in
-Figure~\ref{fig:dnb_example}, many couplings are required among
-individual physics components in order to accurately model this
+sections needed for the neutronics calculation.
+
+As shown in Figure~\ref{fig:dnb_example}, many couplings are required
+among individual physics components in order to accurately model this
 situation with each physics generating and receiving many
@@
 would therefore benefit from fully consistent nonlinear solution
-schemes instead of fixed-point type iterations between
-physics\footnote{Fixed-point iterations between physics are commonly
-  referred to as Picard iterations.}. Furthermore, the space and time
-scales over which these effects occur will also vary greatly.
-
-The computational resources required to solve such problems are
-tremendous. Recent work in modeling coupled fluid flow and solid
-material heat and mass transfer in a reactor subsystem, similar to the
-same components of the departure from nucleate boiling example above,
-was performed as part of analysis of the Department of Energy's
-Consortium for Advanced Simulation of Light Water Reactors (CASL)
-modeling and simulation hub. CASL used the Drekar multiphysics code
-developed at Sandia National Laboratories
-\citep{pawlowski_drekar_2012} for modeling goals in
-grid-to-rod-fretting analysis and will use a similar coupled physics
-structure for future departure from nucleate boiling analysis with
-comparison to experimental data. Using Drekar, multiphysics
-simulations have been performed with fully consistent methods for the
-solution of nonlinear systems using meshes of $O(\sn{1}{9})$ elements
-leveraging $O(100,000)$ cores on leadership class machines. Neutronics
-components to be implemented in CASL for multiphysics analysis, such
-as the Exnihilo radiation transport suite developed at Oak Ridge
-National Laboratory \citep{evans_denovo:_2010}, compute trillions of
-unknowns for full core reactor analysis on $O(\sn{1}{9})$ element
-meshes and $O(100,000)$ cores as well. Given the large scale and
-complexity of these problems, if we aim to advance multiphysics
-solution techniques, then we are motivated to advance the solution of
-complex and general nonlinear problems exploiting leadership class
-levels of parallelism.
+schemes such as that developed by this work instead of fixed-point
+type iterations between physics\footnote{Fixed-point iterations
+  between physics are commonly referred to as Picard iterations.}. To
+leverage such a method for analysis such as a departure from nucleate
+boiling calculation, we also require effective solutions for each of
+the physics involved including the neutronics problem and the fluid
+problem, both of which we discuss next.
 
@@
 
-In order to leverage MCSA as a solution technique for physics
-problems, its current formulation requires the linear operator and all
-preconditioners to be explicitly formed. Recent developments in the
-Exnihilo neutronics package at Oak Ridge National Laboratory have
-permitted generation of the $SP_N$ system of equations for detailed
-neutronics models of fission reactor-based
-systems\citep{evans_simpli_2013}. By fully forming these equations and
-formulating them as a linear algebra problem instead of using the
-explicit iterative methods of the past, we now have access to all of
-the modern advancements in computational linear algebra including
-Krylov solvers for asymmetric systems and algebraic preconditioning
-methods. This leads us to then explore the applicability of our work
-in discrete Monte Carlo methods for linear systems as a possible
-solution method for the $SP_N$ equations. In addition, solving the
-$SP_N$ equations in this way also breaks away from the $S_N$ forms of
-parallelism where spatial parallelism is achieved by an efficient
-parallel sweep, angular efficiency achieved by pipe-lining, and energy
-parallelism achieved by decoupling the groups. With the $SP_N$
-equations as a full matrix system, we now can parallelize the problem
-as prescribed by the linear solver, which may be significantly more
-scalable than current $S_N$ transport practices.
+In order to leverage Monte Carlo Synthetic Acceleration as a solution
+technique for physics problems, its current formulation requires the
+physics operator and all preconditioners to be explicitly
+formed. Recent developments in the Exnihilo neutronics package at Oak
+Ridge National Laboratory have permitted generation of the $SP_N$
+system of equations for detailed neutronics models of fission
+reactor-based systems\citep{evans_simpli_2013}. By fully forming these
+equations and formulating them as a linear algebra problem instead of
+using the explicit iterative methods of the past, we now have access
+to all of the modern advancements in computational linear algebra
+including Krylov solvers for asymmetric systems and algebraic
+preconditioning methods. This leads us to then explore the
+applicability of our work in discrete Monte Carlo methods for linear
+systems as a possible solution method for the $SP_N$ equations. In
+addition, solving the $SP_N$ equations in this way also breaks away
+from the $S_N$ forms of parallelism where spatial parallelism is
+achieved by an efficient parallel sweep, angular efficiency achieved
+by pipe-lining, and energy parallelism achieved by decoupling the
+groups. With the $SP_N$ equations as a full matrix system, we now can
+parallelize the problem as prescribed by the linear solver, which may
+be significantly more scalable than current $S_N$ transport practices.
 
@@
 \label{subsec:ns_motiviation}
+
+Nonlinear transport problems are a common occurrence in single and
+multiple physics problems in nuclear engineering. Systems of partial
+differential equations, such as the Navier-Stokes equations that
+describe fluid flow, yield discrete sets of stiff equations with
+nonlinearities present in the variables when discretized by
+conventional methods. These sets of equations characterize momentum
+and energy transport through diffusion, viscous forces and convection,
+and inertial forces. Traditionally, such systems have been solved by
+linearizing them in a form where the nonlinearities in the variables
+are eliminated and more traditional linear methods can be used for
+solutions. Often characterized as segregated methods where physics
+operators are split and their action on the system approximated in
+steps, such methods lack consistency and accuracy in resolving the
+nonlinear component of the solution. In the last 30 years, fully
+consistent nonlinear methods based on Newton's method have become more
+popular and many advances have been made in the nuclear engineering
+field to employ these methods.
+
+In the context of solving standalone linear systems, Monte Carlo
+methods do not provide significant merit over subspace methods due to
+the fact that the physics operator must be explicitly formed. For many
+applications including higher fidelity neutron transport
+discretizations beyond the $SP_N$ equations discussed in the previous
+section, such a requirement is prohibitive and perhaps not even
+feasible to implement. Therefore, a discrete Monte Carlo solution
+method is best suited for situations in which the operator is readily,
+if not naturally, formed. Modern nonlinear methods meet this
+requirement with Newton methods used in conjunction with Krylov
+methods operating on a fully formed matrix for a robust solution
+strategy. Furthermore, modern techniques exist that permit the
+automatic construction of the Jacobian operator generated within a
+Newton method based on the nonlinear residual evaluations, providing
+all of the components necessary for a Monte Carlo solver to provide
+value. We therefore devise a new nonlinear method in this work based
+on the Monte Carlo Synthetic Acceleration algorithm and Newton's
+method.
+
+\subsection{Domain Decomposed Monte Carlo Synthetic Acceleration}
+\label{subsec:parallel_motivation}
+
+The computational resources required to solve the neutronics and fluid
+dynamics problems are tremendous. Recent work in modeling coupled
+fluid flow and solid material heat and mass transfer in a reactor
+subsystem, similar to the same components of the departure from
+nucleate boiling example above, was performed as part of analysis of
+the Department of Energy's Consortium for Advanced Simulation of Light
+Water Reactors (CASL) modeling and simulation hub. CASL used the
+Drekar multiphysics code developed at Sandia National Laboratories
+\citep{pawlowski_drekar_2012} for modeling goals in
+grid-to-rod-fretting analysis and will use a similar coupled physics
+structure for future departure from nucleate boiling analysis with
+comparison to experimental data. Using Drekar, multiphysics
+simulations have been performed with fully consistent methods for the
+solution of nonlinear systems using meshes of $O(\sn{1}{9})$ elements
+leveraging $O(100,000)$ cores on leadership class machines. Neutronics
+components to be implemented in CASL for multiphysics analysis, such
+as the Exnihilo radiation transport suite developed at Oak Ridge
+National Laboratory \citep{evans_denovo:_2010}, compute trillions of
+unknowns for full core reactor analysis on $O(\sn{1}{9})$ element
+meshes and $O(100,000)$ cores as well. Given the large scale and
+complexity of these problems, if we aim to advance multiphysics
+solution techniques, then we are motivated to advance the solution of
+complex physics problems exploiting leadership class levels of
+parallelism.
+
+For Monte Carlo Synthetic Acceleration methods to be viable at the
+production scale for nuclear engineering applications, scalable
+parallel implementations of the algorithm are required. Reviewing the
+literature, Monte Carlo Synthetic Acceleration has yet to be
+parallelized and the Neumann-Ulam Monte Carlo method on which it is
+built has only been parallelized through history-level parallelism
+with full domain replication \citep{alexandrov_efficient_1998}. In
+order to solve large linear transport systems with Monte Carlo
+Synthetic Acceleration, a domain-decomposed parallel strategy is
+required. In the formulation of a parallel Monte Carlo Synthetic
+Acceleration algorithm, we recognize that the algorithm occurs in two
+stages, an outer iteration performing fixed point iteration and
+applying the correction, and an inner Monte Carlo solver that is
+generating the correction via the adjoint or forward methods. The
+parallel aspects of both these components must be
+considered. Therefore, we will develop and implement parallel
+algorithms for the fundmental random walk sequence and Monte Carlo
+Synthetic Acceleration methods leveraging both the knowledge gained
+from the general parallel implementations of subspace methods reviewed
+in Appendix~\ref{ch:linear_problem} and modern parallel strategies for
+domain decomposed Monte Carlo as developed by the reactor physics
+community.
 
@@
 %%---------------------------------------------------------------------------%%
-\section{Research Outline}
+\section{Research and Development Outline}
 \label{sec:research_outline}
@@
 these methods have been further developed for radiation transport
-problems in the form of Monte Carlo Synthetic-Acceleration (MCSA)
+problems in the form of Monte Carlo Synthetic Acceleration (MCSA)
 \citep{evans_monte_2009, evans_monte_2012} but have yet to be applied
@@
 computational physics community. Compared to other methods in this
-regime, MCSA offers three attractive qualities; (1) the linear problem
+regime, MCSA offers three attractive qualities; (1) the physics
 operator need not be symmetric or positive-definite, thereby reducing
@@
 MCSA as a general linear solver and the development of a parallel MCSA
-method are new and unique features of this work, providing a
-framework with which other issues such as resiliency may be addressed
-in the future.
+method are new and unique features of this work, providing a framework
+with which other issues such as resiliency may be addressed in the
+future.
 
@@
 iterations to converge the size of the Krylov subspace will grow
-beyond that of the Jacobian. Finally, using MCSA for the linear solve
-provides its benefits for preconditioning, potential resiliency, and
-parallelism.
+beyond that of the Jacobian. Finally, using MCSA to solve the linear
+model provides its benefits for preconditioning, potential resiliency,
+and parallelism.
 
 %%---------------------------------------------------------------------------%%
-\section{Outline}
+\section{Statement of Work}
+\label{sec:statement_of_work}
+
+This work will improve the iterative performance and parallel
+scalability of solutions to discrete linear and nonlinear transport
+problems by researching and developing a new set of domain decomposed
+Monte Carlo Synthetic Acceleration methods. Improvements in
+performance and scalability will be demonstrated in a wide variety of
+numerical experiments using both simple model neutron transport
+problems and neutron transport and fluid flow problems generated by
+production physics codes.
+
+%%---------------------------------------------------------------------------%%
+\section{Outline of the Document}
+\label{sec:doc_outline}
 
@@
 presented. Using this background on Monte Carlo, synthetic
-acceleration methods are then presented and analyzed using a simple
-model transport problem. Next, in Chapter~\ref{ch:spn_equations}, the
-Monte Carlo methods developed in Chapter~\ref{ch:stochastic_methods}
-are applied to the neutron transport problem. Specifically, the $SP_N$
-form of the Boltzmann neutron transport equation is developed and
-solved using the Monte Carlo methods. Using a difficult fission
-reactor fuel assembly criticality calculation to drive research and
-development for these Monte Carlo methods, several important issues
-regarding Monte Carlo solver applicability and performance will be
+acceleration methods are then presented and new analysis is provided
+using a simple model transport problem. Next, in
+Chapter~\ref{ch:spn_equations}, the Monte Carlo methods developed in
+Chapter~\ref{ch:stochastic_methods} are applied to the neutron
+transport problem. Specifically, the $SP_N$ form of the Boltzmann
+neutron transport equation is developed and solved using the Monte
+Carlo methods. Using a difficult fission reactor fuel assembly
+criticality calculation to drive research and development for these
+Monte Carlo methods, several important issues regarding Monte Carlo
+solver applicability, performance, and preconditioning will be
 discussed along with solutions to these problems. Using these
@@
 Chapter~\ref{ch:nonlinear_problem}, the Monte Carlo methods are
-applied to the Navier-Stokes equations as a model nonlinear energy and
-momentum transport system. Additional difficulties that arise when
-using Monte Carlo methods in these types of systems will be presented
-along with the subsequently developed solutions. In
-Chapter~\ref{ch:parallel_methods}, the Monte Carlo methods are
-parallelized using modern reactor physics techniques for particle
-transport with a full parallel scaling analysis provided using a
-leadership-class computing facility. Finally, the work is summarized
-in Chapter~\ref{ch:conclusion} and topics of future work derived from
-the results of this research presented.
+applied to the Navier-Stokes equations as a model nonlinear fluid flow
+system. A new nonlinear solution technique is developed based on
+Newton's method and both verified against and compared to conventional
+solution techniques. In Chapter~\ref{ch:parallel_methods}, the Monte
+Carlo methods are parallelized using modern reactor physics techniques
+for domain decomposed particle transport with a full parallel scaling
+analysis provided using a leadership-class computing
+facility. Finally, the work is summarized in
+Chapter~\ref{ch:conclusion} and topics of future work derived from the
+results of this research presented.

### 2013-07-23 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
working on refactoring the introduction chapter

@@
 \label{ch:nonlinear_problem}
-Nonlinear transport problems are a common occurrence in single and
-multiple physics problems in nuclear engineering. Systems of partial
-differential equations, such as the Navier-Stokes equations that
-describe fluid flow, yield discrete sets of stiff equations with
-nonlinearities present in the variables when discretized by
-conventional methods. These sets of equations characterize momentum
-and energy transport through diffusion, viscous forces and convection,
-and inertial forces. Traditionally, such systems have been solved by
-linearizing them in a form where the nonlinearities in the variables
-are eliminated and more traditional linear methods can be used for
-solutions. Often characterized as segregated methods where physics
-operators are split and their action on the system approximated in
-steps, such methods lack consistency and accuracy in resolving the
-nonlinear component of the solution. In the last 30 years, fully
-consistent nonlinear methods based on Newton's method have become more
-popular and many advances have been made in the nuclear engineering
-field to employ these methods.
-
-In the context of solving standalone linear systems, Monte Carlo
-methods do not provide significant merit over subspace methods due to
-the fact that the linear operator must be explicitly formed. For many
-applications including higher fidelity neutron transport
-discretizations beyond the $SP_N$ equations solved in the previous
-chapter, such a requirement is prohibitive and perhaps not even
-feasible to implement. Therefore, a discrete Monte Carlo solution
-method is best suited for situations in which the operator is readily,
-if not naturally, formed. Modern nonlinear methods meet this
-requirement with Newton methods used in conjunction with Krylov
-methods operating on a fully formed matrix for a robust solution
-strategy. Furthermore, modern techniques exist that permit the
-automatic construction of the Jacobian operator generated within a
-Newton method based on the nonlinear residual evaluations, providing
-all of the components necessary for a Monte Carlo solver to provide
-value. We therefore devise a new nonlinear method based on the MCSA
-algorithm and Newton's method.
 
@@
 including with two common techniques for numerically representing the
-Jacobian most commonly used in today's applications. Following this,
-the new algorithm will be presented and implementation details will be
-discussed. Using the new algorithm and a production fluid physics
-code, a group of challenging nonlinear benchmark problems for the
-Navier-Stokes equations will be used to verify the correctness of the
-method relative to solutions obtained using conventional methods in
-various flow regimes and geometries. Finally, we will compare the
-performance of the new method against contemporary Newton-Krylov
-methods for the same benchmark problems.
+Jacobian most commonly used in today's applications. Following this, a
+new algorithm utilizing Monte Carlo Synthetic Acceleration will be
+presented and implementation details will be discussed. Using the new
+algorithm and a production fluid physics code, a group of challenging
+nonlinear benchmark problems for the Navier-Stokes equations will be
+used to verify the correctness of the method relative to solutions
+obtained using conventional methods in various flow regimes and
+geometries. Finally, we will compare the performance of the new method
+against contemporary Newton-Krylov methods for the same benchmark
+problems.
 

### 2013-07-23 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
working on refactoring the introduction chapter

@@
 \label{ch:parallel_methods}
-For MCSA and FANM to be viable at the production scale for nuclear
-engineering applications, scalable parallel implementations of the
-algorithm are required. Reviewing the literature, MCSA has yet to be
-parallelized and the Neumann-Ulam Monte Carlo method on which it is
-built has only been parallelized through history-level parallelism
-with full domain replication \citep{alexandrov_efficient_1998}. In
-order to solve large linear transport systems with MCSA, a
-domain-decomposed parallel strategy is required. In the formulation of
-a parallel MCSA algorithm, we recognize that the algorithm occurs in
-two stages, an outer iteration performing fixed point iteration and
-applying the correction, and an inner Monte Carlo solver that is
-generating the correction via the adjoint or forward methods. The
-parallel aspects of both these components must be
-considered. Therefore, we will develop and implement parallel
-algorithms for the Neumann-Ulam and MCSA methods leveraging both the
-knowledge gained from the general parallel implementations of Krylov
-methods reviewed in Appendix~\ref{ch:linear_problem} and modern
-parallel strategies for domain decomposed Monte Carlo as developed by
-the reactor physics community.
 

### 2013-07-23 doc/dissertation/document/chapters/mc_background/mc_background.tex
finished add step problem to nonlinear chapter
document formatting for appearance

@@
 Monte Carlo solvers are used to invert the operator. The
-\textit{Fixed-Point Monte Carlo Synthetic-Acceleration} (MCSA) method
+\textit{Fixed-Point Monte Carlo Synthetic Acceleration} (MCSA) method
 is defined as:
@@
   \centering
-  \includegraphics[width=6in,clip]{chapters/mc_background/dir_adj_cpu.pdf}
+  \includegraphics[width=5in,clip]{chapters/mc_background/dir_adj_cpu.pdf}
   \caption{\textbf{CPU Time (s) to converge vs. Problem Size ($N$ for
@@
   \centering
-  \includegraphics[width=6in,clip]{chapters/mc_background/dir_adj_iterations.pdf}
+  \includegraphics[width=5in,clip]{chapters/mc_background/dir_adj_iterations.pdf}
   \caption{\textbf{Iterations to converge vs. Problem Size ($N$ for an
@@
   \centering
-  \includegraphics[width=6in,clip]{chapters/mc_background/dir_adj_conv.pdf}
+  \includegraphics[width=5in,clip]{chapters/mc_background/dir_adj_conv.pdf}
   \caption{\textbf{Infinity norm of the solution residual
@@
   \centering
-  \includegraphics[width=6in,clip]{chapters/mc_background/seq_cpu.pdf}
+  \includegraphics[width=4.5in,clip]{chapters/mc_background/seq_cpu.pdf}
   \caption{\textbf{CPU Time (s) to converge vs. Problem Size ($N$ for
@@
   \centering
-  \includegraphics[width=6in,clip]{chapters/mc_background/seq_iterations.pdf}
+  \includegraphics[width=4.5in,clip]{chapters/mc_background/seq_iterations.pdf}
   \caption{\textbf{Iterations to converge vs. Problem Size ($N$ for an
@@
   \centering
-  \includegraphics[width=6in,clip]{chapters/mc_background/seq_conv_100.pdf}
+  \includegraphics[width=4.5in,clip]{chapters/mc_background/seq_conv_100.pdf}
   \caption{\textbf{Infinity norm of the solution residual
@@
   \centering
-  \includegraphics[width=6in,clip]{chapters/mc_background/seq_conv_500.pdf}
+  \includegraphics[width=4.5in,clip]{chapters/mc_background/seq_conv_500.pdf}
   \caption{\textbf{Infinity norm of the solution residual
@@
   \centering
-  \includegraphics[width=6in,clip]{chapters/mc_background/estimator_nh_iters.pdf}
+  \includegraphics[width=4.75in,clip]{chapters/mc_background/estimator_nh_iters.pdf}
   \caption{\textbf{Iterations (s) to converge vs. Monte Carlo
@@
   \centering
-  \includegraphics[width=6in,clip]{chapters/mc_background/estimator_nh_time.pdf}
+  \includegraphics[width=4.75in,clip]{chapters/mc_background/estimator_nh_time.pdf}
   \caption{\textbf{CPU Time (s) to converge vs. Monte Carlo histories
@@
   \centering
-  \includegraphics[width=6in,clip]{chapters/mc_background/estimator_wc_iters.pdf}
+  \includegraphics[width=4.75in,clip]{chapters/mc_background/estimator_wc_iters.pdf}
   \caption{\textbf{Iterations (s) to converge vs. history weight
@@
   \centering
-  \includegraphics[width=6in,clip]{chapters/mc_background/estimator_wc_time.pdf}
+  \includegraphics[width=4.75in,clip]{chapters/mc_background/estimator_wc_time.pdf}
   \caption{\textbf{CPU Time (s) to converge vs. history weight cutoff
@@
   \centering
-  \includegraphics[width=6in,clip]{chapters/mc_background/absorption_iters.pdf}
+  \includegraphics[width=4.5in,clip]{chapters/mc_background/absorption_iters.pdf}
   \caption{\textbf{Iterations (s) to converge vs. artificial
@@
   \centering
-  \includegraphics[width=6in,clip]{chapters/mc_background/absorption_time.pdf}
+  \includegraphics[width=4.5in,clip]{chapters/mc_background/absorption_time.pdf}
   \caption{\textbf{CPU Time (s) to converge vs. artificial absorption
@@
 
-In this chapter, the fundamentals of Monte Carlo synthetic
-acceleration methods were presented and explored in the context of a
-simple model transport system. The following are the signficant
+In this chapter, the fundamentals of Monte Carlo Synthetic
+Acceleration methods were presented and explored in the context of a
+simple model transport system. The following are the significant
 observations and findings.
@@
 \begin{itemize}
-\item Monte Carlo synthetic acceleration (MCSA) has been generalized
+\item Monte Carlo Synthetic Acceleration (MCSA) has been generalized
   for all discrete linear transport problems

### 2013-07-23 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
document formatting for appearance

@@
 
-In this chapter we have developed and explored Monte Carlo synthetic
-acceleration methods in the context of solutions to the Navier-Stokes
+In this chapter we have developed and explored Monte Carlo Synthetic
+Acceleration methods in the context of solutions to the Navier-Stokes
 equations. The following are the significant observations and findings.
@@
 \item Forward-Automated Newton-MCSA (FANM), a new inexact Newton
-  method leveraging Monte Carlo synthetic acceleration, has been
+  method leveraging Monte Carlo Synthetic Acceleration, has been
   developed

### 2013-07-23 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
document formatting for appearance

@@
 giving the eigenvalue spectrum for the Jacobi preconditioned iteration
-matrix. To find the maximum eigenvalue,
-Eq~(\ref{eq:preconditioned_eigenvalues}) is plotted as a function of
-$p$ with $p=q$ in Figure~\ref{fig:diffusion_spectrum} for various
-values of $\Sigma_a$.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=5in,clip]{chapters/parallel_mc/diffusion_spectrum.png}
-  \end{center}
-  \caption{\textbf{Eigenvalue spectra for the diffusion equation.}}
-  \label{fig:diffusion_spectrum}
-\end{figure}
-We find that the maximum eigenvalue exists when $p=q=0$, giving the
-following for the spectral radius of the Jacobi preconditioned
-iteration matrix:
+matrix. We find that the maximum eigenvalue exists when $p=q=0$,
+giving the following for the spectral radius of the Jacobi
+preconditioned iteration matrix:
 \begin{equation}
@@
     \begin{center}
-      \includegraphics[width=5in,clip]{chapters/parallel_mc/measured_spec_rad.png}
+      \includegraphics[width=4in,clip]{chapters/parallel_mc/measured_spec_rad.png}
     \end{center}
@@
     \begin{center}
-      \includegraphics[width=6.0in,clip]{chapters/parallel_mc/leakage_variation_2.pdf}
+      \includegraphics[width=4.75in,clip]{chapters/parallel_mc/leakage_variation_2.pdf}
     \end{center}
@@
     \begin{center}
-      \includegraphics[width=6.0in,clip]{chapters/parallel_mc/leakage_error_2.pdf}
+      \includegraphics[width=4.75in,clip]{chapters/parallel_mc/leakage_error_2.pdf}
     \end{center}
@@
   \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_ms_impeff.pdf}
+    \includegraphics[width=5in]{chapters/parallel_mc/titan_strong_ms_impeff.pdf}
   \end{center}
@@
 In this chapter we have developed a domain-decomposed parallel
-algorithm for Monte Carlo synthetic acceleration and explored the
+algorithm for Monte Carlo Synthetic Acceleration and explored the
 qualities of the algorithm through parallel scaling studies. The

### 2013-07-23 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
document formatting for appearance

@@
 \label{sec:spn_equations}
-The $P_N$ equations as derived in Appendix~\ref{chap:pn_equations}
+The $P_N$ equations as derived in \citep{lewis_computational_1993}
 give $N+1$ coupled first-order equations capturing the spatial and
-angular-dependence of the solution. In multiple dimensions, the
-equation set becomes large and coupled not only through angular
-moments but also through the spatial variables. As a simpler
-alternative to multidimensional $P_N$ solutions, Gelbard recognized in
-1960 that the planar $P_N$ equations could be simplified and applied
-an ad-hoc method to extend them to multiple dimensions, yielding the
-$SP_N$ equations. These equations are not only fewer in number, but
-also take on a diffusion-like form while maintaining the angular
-character of the flux, making them amenable to solutions with modern
-diffusion methods.
+angular-dependence of the solution:
+\begin{equation}
+   \frac{1}{2n+1} \frac{\partial}{\partial x}\Big[ (n+1) \phi_{n+1} + n
+     \phi_{n-1} \Big] + \Sigma_n \phi_n = q\delta_{n0} \:,
+  \label{eq:final_pn_equations}
+\end{equation}
+where $\Sigma_n = \sigma-\sigma_{sn}$ and the summations are truncated
+at some level of approximation $N$ such that $n = 0,1,\dotsc,N$. This
+yields a set of $N+1$ equations for $N+2$ flux moments. In multiple
+dimensions, the equation set becomes large and coupled not only
+through angular moments but also through the spatial variables. As a
+simpler alternative to multidimensional $P_N$ solutions, Gelbard
+recognized in 1960 that the planar $P_N$ equations could be simplified
+and applied an ad-hoc method to extend them to multiple dimensions,
+yielding the $SP_N$ equations. These equations are not only fewer in
+number, but also take on a diffusion-like form while maintaining the
+angular character of the flux, making them amenable to solutions with
+modern diffusion methods.
 
@@
 order even moments in this system to zero (e.g. $\phi_6 = \phi_4 = 0$
-yields the $SP_3$ equations). Boundary conditions for these equations
-are provided in Appendix~\ref{chap:spn_boundary_conditions}, the
-multigroup form is presented in Appendix~\ref{chap:mg_spn_equations},
-and the spatial discretization in
-Appendix~\ref{chap:spn_spatial_discretization}.
+yields the $SP_3$ equations).
 
@@
 Eq~(\ref{eq:eigenvalue_transport}). We can readily replace the fixed
-source in Eq~(\ref{eq:multigroup_spn_equations}) with the fission
-source derived in \S~\ref{sec:pn_eigenvalue}:
+source in Eq~(\ref{eq:spn_matrix}) with a fission source this time
+considering the multigroup form of the $SP_N$ equations derived in
+Appendix~\ref{chap:mg_spn_equations}:
 \begin{multline}
@@
 with the fission matrix $\mathbf{F}$ given by
-Eq~(\ref{eq:fission_matrix}). We again apply the change of variables
-to yield the set of multigroup pseudo-moments $\mathbb{U}_n$ where now
+\citep{evans_simpli_2013}. We again apply the change of variables to
+yield the set of multigroup pseudo-moments $\mathbb{U}_n$ where now
 the application of the fission matrix to the moment vectors will be
@@
   \begin{center}
-    \includegraphics[width=4in]{chapters/spn_equations/group1.png}
+    \includegraphics[width=3.25in]{chapters/spn_equations/group1.png}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=4in]{chapters/spn_equations/group10ds.png}
+    \includegraphics[width=3.25in]{chapters/spn_equations/group10ds.png}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=4in]{chapters/spn_equations/group10us.png}
+    \includegraphics[width=3.25in]{chapters/spn_equations/group10us.png}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/block_jacobi_res.pdf}
+    \includegraphics[width=5in]{chapters/spn_equations/block_jacobi_res.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/block_jacobi_rich_res.pdf}
+    \includegraphics[width=5in]{chapters/spn_equations/block_jacobi_rich_res.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/breakdown_iterations.pdf}
+    \includegraphics[width=4.25in]{chapters/spn_equations/breakdown_iterations.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/breakdown_histories.pdf}
+    \includegraphics[width=4.25in]{chapters/spn_equations/breakdown_histories.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/breakdown_time.pdf}
+    \includegraphics[width=4.25in]{chapters/spn_equations/breakdown_time.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/ilut_iterations.pdf}
+    \includegraphics[width=4.25in]{chapters/spn_equations/ilut_iterations.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/ilut_size.pdf}
+    \includegraphics[width=4.25in]{chapters/spn_equations/ilut_size.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/ilut_quality.pdf}
+    \includegraphics[width=4.25in]{chapters/spn_equations/ilut_quality.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/spainv_iterations.pdf}
+    \includegraphics[width=4.25in]{chapters/spn_equations/spainv_iterations.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/spainv_size.pdf}
+    \includegraphics[width=4.25in]{chapters/spn_equations/spainv_size.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/spainv_quality.pdf}
+    \includegraphics[width=4.25in]{chapters/spn_equations/spainv_quality.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/psilut_iterations.pdf}
+    \includegraphics[width=4.25in]{chapters/spn_equations/psilut_iterations.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/psilut_size.pdf}
+    \includegraphics[width=4.25in]{chapters/spn_equations/psilut_size.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/psilut_quality.pdf}
+    \includegraphics[width=4.25in]{chapters/spn_equations/psilut_quality.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/rda_iterations.pdf}
+    \includegraphics[width=4.25in]{chapters/spn_equations/rda_iterations.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/rda_quality.pdf}
+    \includegraphics[width=4.25in]{chapters/spn_equations/rda_quality.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/solver_iters.pdf}
+    \includegraphics[width=5in]{chapters/spn_equations/solver_iters.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/solver_time.pdf}
+    \includegraphics[width=5in]{chapters/spn_equations/solver_time.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/solver_p_time.pdf}
+    \includegraphics[width=5in]{chapters/spn_equations/solver_p_time.pdf}
   \end{center}
@@
 
-In this chapter, the Monte Carlo synthetic acceleration method has
+In this chapter, the Monte Carlo Synthetic Acceleration method has
 been applied to the $SP_N$ discretization of the neutron transport

### 2013-07-24 doc/dissertation/document/backmatter/diffusion_problem.tex
adding first titan scaling figures to parallel chapter
adding summary sections in cconclusion chapter

--- 2013-06-11 doc/dissertation/document/backmatter/diffusion_problem.tex
@@
-\chapter{Two-Dimensional One-Speed Neutron Diffusion Model Problem}
+\chapter{Two-Dimensional One-Speed Neutron Diffusion Model Problem\ }
 \label{chap:diffusion_problem}
 For many numerical experiments in this work, we choose the one-speed,
-two-dimensional neutron diffusion equation as a model transport
-problem \citep{duderstadt_nuclear_1976}:
+two-dimensional neutron diffusion equation as a model neutron
+transport problem \cite{duderstadt_nuclear_1976}:
 \begin{equation}
@@
 stencil shown in Figure~\ref{fig:stencil} over a grid of size $h$
-\citep{leveque_finite_2007}:
+\cite{leveque_finite_2007}:
 \begin{multline}

### 2013-07-24 doc/dissertation/document/chapters/conclusion/conclusion.tex
spell check on nonlinear chapter
adding summary sections in cconclusion chapter

@@
 \label{ch:conclusion}
-Put conclusions here.
+
+For high fidelity simulations of nuclear reactors, physics solutions
+are required with speed and accuracy. When the problems become large
+enough, such as the simulation of an entire reactor core,
+leadership-class computing facilities must be leveraged on the grounds
+of requirements for both time to solution and the amount of memory
+required to contain the description of the entire problem and the
+solution. Looking forward to the next generation of machines,
+concurrency will go up and available memory will go down, putting
+pressure on physics application developers to research and develop
+solution techniques that can efficiently leverage this hardware. In
+nuclear reactor simulation, both neutron transport and fluid flow
+calculations consume a vast amount of computational time in order to
+properly characterize both steady state operational characteristics
+and transient accident scenarios. In both cases, we desire solution
+techniques that are aware of coming changes in hardware.
+
+In this work, we have presented Monte Carlo Synthetic Acceleration as
+a viable solution technique for both neutron transport and fluid flow
+problems on future hardware. To do this, we carried out three research
+and developement activities. First, we applied MCSA to the $SP_N$ form
+of the neutron transport equation and analyzed the preconditioning
+requirements and subsequent performance of the method when compared to
+conventional practices. Second, we used the knowledge gained from the
+research on neutron transport to develop the FANM method for nonlinear
+problems and studied its performance on three different benchmarks for
+the Navier-Stokes equations to demonstrate its applicability. Finally,
+we parallelized MCSA such that it may be applied to physics problems
+in a leadership-class computing environment.
+
+In this chapter, we review the work presented in this document and
+describe how this work met the goals of the research. We then
+reiterate the important issues discovered during the course of this
+work and develop a strategy for future work to potentially alleviate
+them. Finally, we close with some final remarks regarding Monte Carlo
+Synthetic Acceleration.
 
@@
 
+This work demonstrates the first application of MCSA to the neutron
+transport problem. To do this, we used the $SP_N$ form of the
+Boltzmann transport equation. In this form, the transport equation
+takes on a diffusion-like form where the angular character of the flux
+is accounted for in moment terms much like those found in the $P_N$
+equations. To meet the goals of this work and drive the research and
+development of MCSA for application to light water reactor problems, a
+difficult nuclear fuel assembly criticality calculation was used in a
+set of numerical experiments. These calculations were enabled by
+incorporating MCSA as a solution scheme into the k-eigenvalue solver
+in the Exnihilo production neutronics code base at Oak Ridge National
+Laboratory.
+
+When applying the new technique to the fuel assembly problem, it was
+initially found that MCSA could not converge the problem with basic
+Jacobi-based preconditioning. This was discovered to be due to the
+fact that the transport operator generated in the eigenvalue
+calculation was very ill-conditioned due to the large amount of
+neutron scattering in the light water moderator. The more scattering
+present in the system, the longer a random walk will take and thus a
+spectral radius approaching unity was observed. A simpler neutron
+diffusion problem was used to demonstrate the breakdown of MCSA when
+spectral radii this large are generated.
+
+As a result of the breakdown observation, a suite of advanced
+preconditioning techniques was applied to MCSA for the fuel assembly
+problem in order to reduce the spectral radius to a level below where
+breakdown occurs. From the results of this analysis, ILUT
+preconditioning was chosen for subsequent investigations. Using this
+preconditioning, it was verified that MCSA is indeed general enough to
+solve the asymmetric system generated by the $SP_N$
+equations. However, the explicit MCSA preconditioning strategy
+developed for this work generated very dense systems that consumed
+tremendous amounts of memory to store and compute time to
+generate. The memory problem was partially alleviated by applying the
+reduced domain approximation, however a significant memory overhead
+remained along with prohibitive time for construction.
+
+Using the developed preconditioners and the reduced domain
+approximation, a set of $SP_N$ calculations using up to 4 energy
+groups were enabled for the fuel assembly problem. To verify MCSA, the
+fuel assembly problem was solved with different energy groups using
+both MCSA and two production Krylov methods. MCSA was observed to
+produce the same scalar flux in all groups at all spatial locations
+and generate the same k-eigenvalue in the same number of eigenvalue
+iterations.
+
+For the same calculations, MCSA had qualitatively similar iterative
+performance to the production Krylov methods, showing improved
+performance over GMRES while a Conjugate Gradient-based method
+performed better using the same preconditioining. Improved performance
+over GMRES meets the goal of improving iterative performance to a
+certain extent but this was not the true for the Conjugate
+Gradient-based method. However, although the $SP_N$ equations are
+asymmetric for multigroup calculations, they are symmetric at the
+block level and therefore we expect the Conjugate Gradient-based
+method to perform better as this structure can be more readily
+utilized than the more general GMRES and MCSA methods. In terms of CPU
+time, MCSA was observed to perform $O(100)$ times slower than the
+Krylov methods due to both lack of optimization in the random walk
+sequence and the explicit preconditioning strategy.
+
 %%---------------------------------------------------------------------------%%
@@
 \label{sec:nonlinear_conclusions}
+
+This work presents the FANM method, a new nonlinear solution scheme
+based on an inexact Newton method. To meet the goals of this work the
+FANM method was characterized by applying it to the Navier-Stokes
+equations. This research was enabled by implementing the FANM method
+within the nonlinear solver sequence leveraged by the Drekar production
+multiphysics code base being developed at Sandia National
+Laboratories. In both convection and driven flow regimes we solved
+three difficult benchmark problems for the Navier-Stokes
+equations. Using the work in preconditioning from the research on the
+$SP_N$ equations, it was found that the same preconditioning strategy
+could be used to achieve convergence of the linear models generated at
+each FANM iteration by applying an algebraic multigrid method.
+
+For each benchmark problem, FANM solutions were verified against
+Newton-Krylov solutions leveraging GMRES as the linear solver. The
+FANM solutions were observed to be numerically identical to those
+generated by the Newton-Krylov method. In addition, the benchmark
+problems also served as a vehicle to compare the performance of FANM
+with the Newton-Krylov method. For the thermal convection cavity
+problem, FANM was observed to converge in fewer linear solver
+iterations than Newton-Krylov at all Rayleigh numbers tested when the
+same preconditioning was applied to both methods. This means that FANM
+demonstrated superior iterative performance for problems dominated by
+natural convection, meeting the goal of improved iterative performance
+in this case.
+
+Iterative performance for the problems dominated by intertial driven
+flow varied depending on the situation. For the lid driven cavity
+problem, FANM converged in more linear solver iterations for 3 out of
+4 cases. However, for each of these cases it was observed that the
+extra MCSA iterations were generated by a smaller forcing term at each
+Newton iteration when compared to the Newton-Krylov solver. Given that
+a smaller forcing term is equivalent to requesting convergence of the
+linear solver to a smaller residual, we expect these extra
+iterations. For the lid driven cavity case where FANM converged in
+fewer iterations, it was observed that increasing the number of
+stochastic histories used at every MCSA iteration enabled convergence
+in fewer iterations, again meeting the goal of improved iterative
+performance.
+
+The backward facing step problem at low Reynolds number gave better
+performance with FANM convergining in fewer linear solver
+iterations. However, as the Reynolds number was increased so did the
+number of MCSA iterations required to converge the linear model at
+each FANM iteration. Unlike the lid driven cavity problem, this was
+discovered to occur due to the inability of MCSA to quickly converge
+the ill-conditioned linear model rather than the introduction of
+smaller forcing terms.
+
+Timing performance for all benchmarks favored the Newton-Krylov solver
+with FANM observed to be $O(100)$ slower for the thermal convection
+cavity and lid driven cavity problems and $O(1,000)$ times for the
+backward facing step problem. For the first two cases, the timing
+differences were observed to be qualititavely the same as those
+observed for the $SP_N$ performance analysis. Again, this comes from
+both lack of optimization of the Monte Carlo sequence and the
+introduction of explicit preconditioning. For the backward facing step
+problem, the ill-conditioning of the system adds the extra order of
+magnitude slow down.
 
@@
 
+A new parallel MCSA algorithm was developed for this work to meet the
+goal of demonstrating the improved scalability of MCSA on
+leadership-class hardware. It was found the MCSA can indeed be
+effectively parallelized using the multiple-set overlapping-domain
+decomposition algorithm borrowed from the reactor physics
+community. Using a neutron diffusion problem, the parallel algorithm
+was verified to produce the same results as two production Krylov
+methods.
+
+The new algorithm was tested in a wide variety of parallel scaling
+studies on the Titan Cray XK7 machine at the Oak Ridge Leadership
+Computing Facility. To test the algorithm at high levels of
+concurrency, up to 65,356 cores were used in strong scaling exercises
+and 131,072 cores used in weak scaling exercises using the neutron
+diffusion problem. In general, the new parallel MCSA algorithm was
+observed to produce better parallel scaling results when compared to a
+production GMRES and Conjugate Gradient method. We note here that
+these scaling studies should be reconsidered if arithmetic
+optimization of the code has been completed.
+
+It was found that the leakage of histories from domain to domain in
+the parallel Monte Carlo algorithm could in fact be quantified
+analytically using the algebraic properties of the system. These
+relationships were then used to determine the amount of overlap one
+may require in the parallel algorithm to reduce communication costs
+and increase parallel efficiencies. Scaling studies showed that in the
+strong case overlap in small quanities on the order of the
+mean-free-path of a stochastic history in the simulation could boost
+parallel efficiencies by a few percent. However, it was found that
+this additional overal was not effective in boosting strong scaling
+efficiencies. It was found that overall overlap was not very effective
+due to the fact that the parallel communication saved during the Monte
+Carlo transport sequence is simply deferred until after transport is
+complete when it manifests itself as an overlapping parallel vector
+reduction operation.
+
+Applying multiple sets in the parallel algorithm was found to not
+enhance the weak scaling of the problem as an additional parallel
+overhead is introduced when the calculations from the set are combined
+in superposition. For the strong scaling case, improvements were not
+noted until after the strong scaling wall was hit. At this point,
+multiple sets were observed to increase parallel efficiencies from
+38\% to 58\% at 16,384 cores. Perhaps more important here is the fact
+that although the parallel efficiency was reduced, multiple sets were
+observed to actually improve the time to solution. Unlike a
+traditional Krylov method that we might apply to solve a neutron
+transport problem, using MCSA means that we can actually make a
+physical copy of the problem on the machine and can combine seperate
+Monte Carlo solutions for each copy through superposition to enhance
+time to solution. Time to solution is then improved because fewer
+histories were run in each copy and therefore each MCSA iteration is
+faster or more global histories are computed and fewer MCSA iterations
+are required to converge.
+
+Finally, given that overlap was not very constructive in boosting
+parallel efficiencies, it was postulated that very little
+domain-to-domain communication of histories was occuring in the first
+place and then subsequently verified using the analytic relationships
+for domain leakage. Using this idea, we implemented a subdomain
+Neumann-Ulam method with MCSA such that MCSA now takes the form of a
+stochastic realization of an additive Schwarz method. Scaling studies
+using this technique showed significant improvements in parallel
+scalability even when compared to the MCSA results when
+domain-to-domain communication was present, further enhancing the
+method and meeting the goal of improved scalability.
+
 %%---------------------------------------------------------------------------%%
@@
 
-Breaking away from $\rho(H) < 1$
+\subsection{Preconditioning}
+\label{subsec:future_preconditioning}
+
+Standalone Monte Carlo as a preconditioner? (e.g. for flex-GMRES)
+
+Stochastic variants of conventional preconditioners?
+
+Variance reduction for preconditioning?
+
+\subsection{Breaking away from $\rho(H) < 1$}
+\label{subsec:future_spec_rad}
 
@@
 
-Stochastic variants of conventional preconditioners?
-
-Variance reduction for preconditioning?
-
-Controling MCSA histories as a function of nonlinear iteration/forcing
-term?
+\subsection{Performance}
+\label{subsec:future_performance}
 
@@
 
+Controling MCSA histories as a function of nonlinear iteration/forcing
+term?
+
 %%---------------------------------------------------------------------------%%

### 2013-07-24 doc/dissertation/document/chapters/introduction/introduction.tex
adding summary sections in cconclusion chapter

@@
 strategies change when they are coupled to other problems
-\citep{u.s._department_of_energy_casl_2011}. Furthermore, a desire for
+\cite{u.s._department_of_energy_casl_2011}. Furthermore, a desire for
 predictive simulations to enhance the safety and performance of
@@
 problems must be developed to leverage this new hardware
-\citep{kogge_using_2011}. Issues such as resiliency to node failure,
+\cite{kogge_using_2011}. Issues such as resiliency to node failure,
 limited growth of memory available per node, and scaling to large
 numbers of cores will be pertinent to robust algorithms aimed at this
-new hardware. Considering these issues, this work researches and
-develops parallel Monte Carlo Synthetic Acceleration methods for
-linear problems and a new method based on the synthetic acceleration
-techniques to advance solution algorithms for nonlinear problems.
+new hardware.
+
+Considering these physics and hardware issues, this goal of this work
+is to research and develop Monte Carlo Synthetic Acceleration methods
+for neutron transport problems and a new method leveraging the
+synthetic acceleration techniques to advance solution algorithms for
+nonlinear problems in fluid flow. These methods are based on using the
+Monte Carlo method to directly accelerate the convergence of
+traditional fixed-point iterative schemes for these problems. We aim
+to address both their benefits and their shortcomings in the context
+of the physics of interest and desire to identify areas in which
+improvements can be made. Once the linear and nonlinear methods have
+been developed, we look forward to parallelizing them with the goal
+that they may be leveraged competitively on leadership class computing
+platforms.
+
+To achieve these goals, three complementary research and development
+activities are performed. First, Monte Carlo Sythetic Acceleration is
+applied to the neutron transport problem in a light water reactor
+calculation to develop the necessary preconditioning techniques and
+understanding required to solve difficult problems. Second, using
+these new techniques, a series of benchmarks for the Navier-Stokes
+equations are solved in different flow regimes to research the
+performance of the Monte Carlo algorithm for nonlinear problems using
+the new Forward-Automated Newton-MCSA algorithm developed by this
+work. Third, a novel parallel algorithm based on domain decomposed
+Monte Carlo methods for particle transport is developed to fully
+parallelize the solution techniques developed to solve the neutron
+transport and fluid flow problems on leadership class hardware with
+scaling studies performed using a neutron diffusion problem.
 
@@
 capability can enable tighter design tolerances for improved thermal
-performance and efficiency, higher fuel burn-up and therefore reduction
-in generated waste, and high confidence in accident scenario
+performance and efficiency, higher fuel burn-up and therefore
+reduction in generated waste, and high confidence in accident scenario
 models. The physics that dominate these types of analysis include
@@
 nucleate boiling scenario in the subchannel of a nuclear fuel
-assembly.
+assembly as shown in Figure~\ref{fig:dnb_example}.
+
 \begin{figure}[t!]
@@
 type iterations between physics\footnote{Fixed-point iterations
-  between physics are commonly referred to as Picard iterations.}. To
-leverage such a method for analysis such as a departure from nucleate
-boiling calculation, we also require effective solutions for each of
-the physics involved including the neutronics problem and the fluid
-problem, both of which we discuss next.
+  between physics are commonly referred to as Picard iterations in the
+  multiphysics community.}. To leverage such a method for analysis
+such as a departure from nucleate boiling calculation, we also require
+effective solutions for each of the physics involved including the
+neutronics problem and the fluid problem, both of which we discuss
+next.
 
@@
 methods for large scale problems are commonly variants on the discrete
-ordinates ($S_N$) method \citep{evans_denovo:_2010}. For fission
+ordinates ($S_N$) method \cite{evans_denovo:_2010}. For fission
 reactor neutronics simulations, the $S_N$ method requires potentially
@@
 good accuracy for the responses of interest
-\citep{slaybaugh_acceleration_2011}. Other forms of the transport
+\cite{slaybaugh_acceleration_2011}. Other forms of the transport
 problem, including the $P_N$ method, take on a simpler form than the
@@
 ordinates. Why did this happen? A significant problem, pointed out by
-Brantley and Larsen \citep{brantley_simplified_2000}, was that little
+Brantley and Larsen \cite{brantley_simplified_2000}, was that little
 rigor had been applied to the formulation of the $SP_N$ equations
@@
 In order to leverage Monte Carlo Synthetic Acceleration as a solution
-technique for physics problems, its current formulation requires the
-physics operator and all preconditioners to be explicitly
+technique for neutron transport problems, its current formulation
+requires the physics operator and all preconditioners to be explicitly
 formed. Recent developments in the Exnihilo neutronics package at Oak
@@
 system of equations for detailed neutronics models of fission
-reactor-based systems\citep{evans_simpli_2013}. By fully forming these
+reactor-based systems\cite{evans_simpli_2013}. By fully forming these
 equations and formulating them as a linear algebra problem instead of
@@
 including Krylov solvers for asymmetric systems and algebraic
-preconditioning methods. This leads us to then explore the
-applicability of our work in discrete Monte Carlo methods for linear
-systems as a possible solution method for the $SP_N$ equations. In
-addition, solving the $SP_N$ equations in this way also breaks away
-from the $S_N$ forms of parallelism where spatial parallelism is
-achieved by an efficient parallel sweep, angular efficiency achieved
-by pipe-lining, and energy parallelism achieved by decoupling the
-groups. With the $SP_N$ equations as a full matrix system, we now can
-parallelize the problem as prescribed by the linear solver, which may
-be significantly more scalable than current $S_N$ transport practices.
+preconditioning methods.
+
+This leads us to then research the applicability of Monte Carlo
+Synthetic Acceleration methods as a possible solution method for the
+$SP_N$ equations. In addition, solving the $SP_N$ equations in this
+way also breaks away from the $S_N$ forms of parallelism where spatial
+parallelism is achieved by an efficient parallel sweep, angular
+efficiency achieved by pipe-lining, and energy parallelism achieved by
+decoupling the groups. With the $SP_N$ equations as a full matrix
+system, we now can parallelize the problem as prescribed by the Monte
+Carlo algorithm, which may be significantly more scalable than current
+$S_N$ transport practices. In this work we therefore apply Monte Carlo
+Synthetic Acceleration to the $SP_N$ equations and research its
+applicability, performance, and convergence properties in the context
+of a light water reactor system.
 
@@
 automatic construction of the Jacobian operator generated within a
-Newton method based on the nonlinear residual evaluations, providing
-all of the components necessary for a Monte Carlo solver to provide
+Newton method based on the nonlinear residual evaluations, giving all
+of the components necessary for a Monte Carlo solver to provide
 value. We therefore devise a new nonlinear method in this work based
 on the Monte Carlo Synthetic Acceleration algorithm and Newton's
-method.
+method and research its properties, applicability, and performance in
+the context of fluid flow problems.
 
@@
 Drekar multiphysics code developed at Sandia National Laboratories
-\citep{pawlowski_drekar_2012} for modeling goals in
+\cite{pawlowski_drekar_2012} for modeling goals in
 grid-to-rod-fretting analysis and will use a similar coupled physics
@@
 as the Exnihilo radiation transport suite developed at Oak Ridge
-National Laboratory \citep{evans_denovo:_2010}, compute trillions of
+National Laboratory \cite{evans_denovo:_2010}, compute trillions of
 unknowns for full core reactor analysis on $O(\sn{1}{9})$ element
 meshes and $O(100,000)$ cores as well. Given the large scale and
-complexity of these problems, if we aim to advance multiphysics
-solution techniques, then we are motivated to advance the solution of
-complex physics problems exploiting leadership class levels of
-parallelism.
+complexity of these problems, if we aim to advance solution techniques
+in neutronics and fluid flow, then we are motivated to advance the
+solution of complex physics problems exploiting leadership class
+levels of parallelism.
 
@@
 built has only been parallelized through history-level parallelism
-with full domain replication \citep{alexandrov_efficient_1998}. In
-order to solve large linear transport systems with Monte Carlo
-Synthetic Acceleration, a domain-decomposed parallel strategy is
-required. In the formulation of a parallel Monte Carlo Synthetic
-Acceleration algorithm, we recognize that the algorithm occurs in two
-stages, an outer iteration performing fixed point iteration and
-applying the correction, and an inner Monte Carlo solver that is
-generating the correction via the adjoint or forward methods. The
-parallel aspects of both these components must be
-considered. Therefore, we will develop and implement parallel
-algorithms for the fundmental random walk sequence and Monte Carlo
-Synthetic Acceleration methods leveraging both the knowledge gained
-from the general parallel implementations of subspace methods reviewed
-in Appendix~\ref{ch:linear_problem} and modern parallel strategies for
-domain decomposed Monte Carlo as developed by the reactor physics
-community.
+with full domain replication \cite{alexandrov_efficient_1998}. In
+order to solve large linear physics systems with Monte Carlo Synthetic
+Acceleration, a domain decomposed parallel strategy is required. In
+the formulation of a parallel Monte Carlo Synthetic Acceleration
+algorithm, we recognize that the algorithm occurs in two stages, an
+initial step performing a fixed point iteration and a secondary step
+leveraging a Monte Carlo solver that is providing the
+acceleration. The parallel aspects of both these components must be
+considered. Therefore, in this work we will develop and implement
+parallel algorithms for the fundamental random walk sequence and Monte
+Carlo Synthetic Acceleration methods leveraging both the knowledge
+gained from the general parallel implementations of subspace methods
+reviewed in Appendix~\ref{ch:linear_problem} and modern parallel
+strategies for domain decomposed Monte Carlo as developed by the
+reactor physics community.
 
@@
 funding for the next several years to research resilient solver
-technologies for extreme scale facilities
-\citep{u.s._department_of_energy_resilient_2012}. Based on the
-language in this call for proposals, we can identify key issues for
-which a set of robust, massively parallel Monte Carlo solvers could
-provide a solution. As machines begin to operate at hundreds of
-petaflops peak performance and beyond, trends toward reduced energy
-consumption will require incredibly high levels of concurrency to
-achieve the desired computation rates. Furthermore, this drop in power
-consumption will mean increased pressure on memory as memory per node
-is expected to stagnate while cores per node is expected to
-increase. As the number of cores increases, their clock speed is
-expected to stagnate or even decrease to further reduce power
-consumption and manufacturing costs.
+technologies for next generation facilities
+\cite{u.s._department_of_energy_resilient_2012}. Based on the language
+in this call for proposals, we can identify key issues for which a set
+of robust, massively parallel Monte Carlo solvers could provide a
+solution. As machines begin to operate at hundreds of petaflops peak
+performance and beyond, trends toward reduced energy consumption will
+require incredibly high levels of concurrency to achieve the desired
+computation rates. Furthermore, this drop in power consumption will
+mean increased pressure on memory as memory per node is expected to
+stagnate while cores per node is expected to increase. As the number
+of cores increases, their clock speed is expected to stagnate or even
+decrease to further reduce power consumption and manufacturing costs.
 
@@
 to use less memory than conventional methods will serve to alleviate
-some of this pressure. In addition, new parallel strategies that may
+some of this pressure. New parallel strategies that may
 be implemented with stochastic methods could offer a new avenue for
-leveraging the expected levels of high concurrency in extreme scale
-machines.
+leveraging the expected levels of high concurrency in machines
+operating at hundreds of petaflops and beyond.
 
@@
 Monte Carlo methods for the solution of transport problems
-\citep{lewis_computational_1993}. The partial differential equation
+\cite{lewis_computational_1993}. The partial differential equation
 (PDE) community has focused on various deterministic methods for
-solutions to linear problems \citep{saad_iterative_2003,
+solutions to linear problems \cite{saad_iterative_2003,
   kelley_iterative_1995}. In between these two areas are a not widely
 known group of Monte Carlo methods for solving sparse linear systems
-\citep{forsythe_matrix_1950, hammersley_monte_1964,
+\cite{forsythe_matrix_1950, hammersley_monte_1964,
   halton_sequential_1962, halton_sequential_1994}. In recent years,
@@
 problems in the form of Monte Carlo Synthetic Acceleration (MCSA)
-\citep{evans_monte_2009, evans_monte_2012} but have yet to be applied
+\cite{evans_monte_2009, evans_monte_2012} but have yet to be applied
 to more general sparse linear systems commonly generated by the
@@
 regime, MCSA offers three attractive qualities; (1) the physics
-operator need not be symmetric or positive-definite, thereby reducing
-preconditioning complexity, (2) the stochastic nature of the solution
-method provides a natural solution to the issue of resiliency, and (3)
-is amenable to parallelization using modern methods developed by the
-transport community \citep{wagner_hybrid_2010}. The development of
-MCSA as a general linear solver and the development of a parallel MCSA
-method are new and unique features of this work, providing a framework
-with which other issues such as resiliency may be addressed in the
+operator need not be symmetric or positive-definite, (2) the
+stochastic nature of the solution method provides a natural solution
+to the issue of resiliency, and (3) is amenable to parallelization
+using modern methods developed by the transport community
+\cite{wagner_hybrid_2010}. The development of MCSA as a general
+solution technique and the development of a parallel MCSA method are
+new and unique features of this work, providing a framework with which
+other issues such as resiliency or performance may be addressed in the
 future.
 
-In addition to linear solver advancements, nonlinear solvers may also
-benefit from a general and parallel MCSA scheme. In the nuclear
-engineering community, nonlinear problems are often addressed by
-either linearizing the problem or building a segregated scheme and
-using traditionally iterative or direct methods to solve the resulting
-system \citep{pletcher_computational_1997}. In the mathematics
-community, various Newton methods have been popular
-\citep{kelley_iterative_1995}. Recently, Jacobian-Free Newton-Krylov
-(JFNK) schemes \citep{knoll_jacobian-free_2004} have been utilized in
+In addition to advancements in the solution of linear problems,
+nonlinear problems may also benefit from a general and parallel MCSA
+scheme. In the nuclear engineering community, nonlinear problems are
+often addressed by either linearizing the problem or building a
+segregated scheme and using traditionally iterative or direct methods
+to solve the resulting system \cite{pletcher_computational_1997}. In
+the mathematics community, various Newton methods have been popular
+\cite{kelley_iterative_1995}. Recently, Jacobian-Free Newton-Krylov
+(JFNK) schemes \cite{knoll_jacobian-free_2004} have been utilized in
 multiple physics architectures and advanced single physics codes
-\citep{gaston_parallel_2009}. The benefits of JFNK schemes are that
-the Jacobian is never formed, simplifying the implementation, and a
-Krylov solver is leveraged (typically GMRES or Conjugate Gradient),
-providing excellent convergence properties for well-conditioned and
-well-scaled systems. However, there are two potential drawbacks to
-these methods for high fidelity predictive simulations: (1) the
-Jacobian is approximated by a first-order differencing method on the
-order of machine precision such that this error can grow beyond that
-of those in a fine-grained system \citep{kelley_iterative_1995} and
-(2) for systems that are not symmetric positive-definite (which will
-be the case for most multiphysics systems and certainly for most
+\cite{gaston_parallel_2009}. The benefits of JFNK schemes are that the
+Jacobian is never formed, simplifying the implementation, and a Krylov
+solver is leveraged (typically GMRES or Conjugate Gradient), providing
+excellent convergence properties for well-conditioned and well-scaled
+systems. However, there are two potential drawbacks to these methods
+for high fidelity predictive simulations: (1) the Jacobian is
+approximated by a first-order differencing method on the order of
+machine precision such that this error can grow beyond that of those
+in a fine-grained system \cite{kelley_iterative_1995} and (2) for
+systems that are not symmetric positive-definite (which will be the
+case for most multiphysics systems and certainly for most
 preconditioned systems) the Krylov subspace generated by the GMRES
 solver may become prohibitively large
-\citep{knoll_newton-krylov_1995}. To address these issues, this work
-develops a new and novel method for nonlinear systems based on the
-MCSA method.
+\cite{knoll_newton-krylov_1995}.
 
-The Forward-Automated Newton-MCSA (FANM) method is developed as new
-nonlinear solution method. The key features of FANM are: full Jacobian
-generation using modern Forward Automated Differentiation (FAD)
-methods \citep{bartlett_automatic_2006}, and MCSA as the inner linear
-solver. This method has several attractive properties. First, the
-first-order approximation to the Jacobian used in JFNK type methods is
-eliminated by generating the Jacobian explicitly with the model
-equations through FAD. Second, the Jacobian need not be explicitly
-formed by the user but is instead automated through FAD; this
-eliminates the complexity of hand-coding derivatives and has also been
-demonstrated to be more efficient computationally than evaluating
-difference derivatives. Third, unlike GMRES, MCSA does not build a
-subspace during iterations. Although the Jacobian must be explicitly
-formed to use MCSA, for problems that take more than a few GMRES
-iterations to converge the size of the Krylov subspace will grow
+To address these issues, the Forward-Automated Newton-MCSA (FANM)
+method is developed as new nonlinear solution method. The key features
+of FANM are: full Jacobian generation using modern Forward Automated
+Differentiation (FAD) methods \cite{bartlett_automatic_2006}, and MCSA
+as the inner linear solver. This method has several attractive
+properties. First, the first-order approximation to the Jacobian used
+in JFNK type methods is eliminated by generating the Jacobian
+explicitly with the model equations through FAD. Second, the Jacobian
+need not be explicitly formed by the user but is instead automated
+through FAD; this eliminates the complexity of hand-coding derivatives
+and has also been demonstrated to be more efficient computationally
+than evaluating difference derivatives. Third, unlike GMRES, MCSA does
+not build a subspace during iterations. Although the Jacobian must be
+explicitly formed to use MCSA, for problems that take more than a few
+GMRES iterations to converge the size of the Krylov subspace will grow
 beyond that of the Jacobian. Finally, using MCSA to solve the linear
-model provides its benefits for preconditioning, potential resiliency,
-and parallelism.
+model provides its benefits for preconditioning, potential parallelism
+and performance.
 
@@
 
-This work will improve the iterative performance and parallel
-scalability of solutions to discrete linear and nonlinear transport
-problems by researching and developing a new set of domain decomposed
-Monte Carlo Synthetic Acceleration methods. Improvements in
+The goal of this work is to improve the iterative performance and
+parallel scalability of solutions to discrete linear and nonlinear
+transport problems by researching and developing a new set of domain
+decomposed Monte Carlo Synthetic Acceleration methods. Improvements in
 performance and scalability will be demonstrated in a wide variety of
@@
 problems and neutron transport and fluid flow problems generated by
-production physics codes.
+production physics codes in both simple and complex geometries.
 
@@
 
-To present these new developments this document is arranged in the
-following manner. First, in Chapter~\ref{ch:stochastic_methods}, the
-fundamentals of the Monte Carlo method for linear systems are
-presented. Using this background on Monte Carlo, synthetic
-acceleration methods are then presented and new analysis is provided
-using a simple model transport problem. Next, in
-Chapter~\ref{ch:spn_equations}, the Monte Carlo methods developed in
-Chapter~\ref{ch:stochastic_methods} are applied to the neutron
+To present the research and development required to meet the goals of
+this work this document is arranged in the following manner. First, in
+Chapter~\ref{ch:stochastic_methods}, the fundamentals of the Monte
+Carlo method for linear systems are presented. Using this background
+on Monte Carlo, synthetic acceleration methods are then presented and
+new analysis is provided using a simple model transport problem. Next,
+in Chapter~\ref{ch:spn_equations}, the Monte Carlo methods developed
+in Chapter~\ref{ch:stochastic_methods} are applied to the neutron
 transport problem. Specifically, the $SP_N$ form of the Boltzmann
 neutron transport equation is developed and solved using the Monte
-Carlo methods. Using a difficult fission reactor fuel assembly
-criticality calculation to drive research and development for these
-Monte Carlo methods, several important issues regarding Monte Carlo
-solver applicability, performance, and preconditioning will be
+Carlo methods. Using a difficult light water fission reactor fuel
+assembly criticality calculation to drive research and development for
+these Monte Carlo methods, several important issues regarding Monte
+Carlo solver applicability, performance, and preconditioning will be
 discussed along with solutions to these problems. Using these

### 2013-07-24 doc/dissertation/document/chapters/linear_problem/linear_problem.tex
adding first titan scaling figures to parallel chapter
adding summary sections in cconclusion chapter

@@
-\chapter{Conventional Solution Methods for Linear Systems}
+\chapter{Conventional Solution Methods for Linear Systems\ }
 \label{ch:linear_problem}
@@
 through common methods such as finite differences
-\citep{leveque_finite_2007}, finite volumes
-\citep{leveque_finite_2002}, and finite elements
-\citep{zienkiewicz_finite_2005} ultimately generates sets of coupled
+\cite{leveque_finite_2007}, finite volumes
+\cite{leveque_finite_2002}, and finite elements
+\cite{zienkiewicz_finite_2005} ultimately generates sets of coupled
 equations in the form of matrix problems. In many cases, these
@@
   matrix analysis, are heavily based on Saad's text
-  \citep{saad_iterative_2003}.}
+  \cite{saad_iterative_2003}.}
 
-\section{Stationary Methods}
+\section{Stationary Methods\ }
 \label{sec:stationary_methods}
@@
   non-diagonalizable matrices with the Jordan canonical form of
-  $\ve{H}$.} \citep{saad_iterative_2003}, we then have:
+  $\ve{H}$.} \cite{saad_iterative_2003}, we then have:
 \begin{equation}
@@
 
-\section{Projection Methods}
+\section{Projection Methods\ }
 \label{sec:projection_methods}
@@
 basis for the \textit{generalized minimum residual method} (GMRES)
-\citep{saad_gmres:_1986}.
+\cite{saad_gmres:_1986}.
 
@@
 instances where various processes serve to provide the action of that
-operator (e.g. radiation transport sweeps \citep{evans_denovo:_2010})
+operator (e.g. radiation transport sweeps \cite{evans_denovo:_2010})
 that normally may not be amenable to its full construction. In
@@
 
-\section{Parallel Projection Methods}
+\section{Parallel Projection Methods\ }
 \label{sec:parallel_krylov_methods}
@@
 in their implementation that can be efficiently programmed on these
-architectures. Per Saad's text \citep{saad_iterative_2003} and as
+architectures. Per Saad's text \cite{saad_iterative_2003} and as
 noted in Algorithm~\ref{alg:gmres}, these operations are
@@
 inner products. For the last three items, linear algebra libraries
-such as PETSc \citep{gropp_scalable_1993} and Trilinos
-\citep{heroux_overview_2005} provide efficient parallel
+such as PETSc \cite{gropp_scalable_1993} and Trilinos
+\cite{heroux_overview_2005} provide efficient parallel
 implementations for these operations. Depending on the type of
@@
 numerical methods, parallel formulations these operations have
-warranted intense study \citep{tuminaro_parallel_1998}. In all cases, a series
+warranted intense study \cite{tuminaro_parallel_1998}. In all cases, a series
 of scatter/gather operations are required such that global
@@
 number of global communications required (consider
-\citep{sosonkina_scalable_1998} as an example of reducing global
+\cite{sosonkina_scalable_1998} as an example of reducing global
 operation counts using a different orthogonalization procedure than
@@
 communication patterns. Libraries such as Zoltan
-\citep{devine_zoltan_2002}, provide implementations of such
+\cite{devine_zoltan_2002}, provide implementations of such
 algorithms.
@@
 Saad's outline as well as the more formal work of Tuminaro
-\citep{tuminaro_parallel_1998}. For these operations, more complex
+\cite{tuminaro_parallel_1998}. For these operations, more complex
 communication patterns will be required given that the entire global
@@
 simulations that heavily leveraged Krylov methods in their solution
-schemes \citep{gropp_high-performance_2001}. In this follow-on to
+schemes \cite{gropp_high-performance_2001}. In this follow-on to
 their 1997 Bell Prize-winning work, part of their analysis included

### 2013-07-24 doc/dissertation/document/chapters/mc_background/mc_background.tex
adding summary sections in cconclusion chapter

@@
 manuscript published in 1950 by Forsythe and Leibler
-\citep{forsythe_matrix_1950}. In their outline, Forsythe and Leibler
+\cite{forsythe_matrix_1950}. In their outline, Forsythe and Leibler
 in fact credit the creation of this technique to J. Von Neumann and
@@
 provided a more formal explanation of Von Neumann and Ulam's method
-\citep{wasow_note_1952} and Hammersley and Handscomb's 1964 monograph
-\citep{hammersley_monte_1964} and Spanier and Gelbard's 1969 book
-\citep{spanier_monte_1969} present additional detail on this topic
+\cite{wasow_note_1952} and Hammersley and Handscomb's 1964 monograph
+\cite{hammersley_monte_1964} and Spanier and Gelbard's 1969 book
+\cite{spanier_monte_1969} present additional detail on this topic
 using a collection of references from the 1950's and early
@@
 that improved dramatically on the performance of these methods
-\citep{halton_sequential_1962}. Recent work in this area has utilized
+\cite{halton_sequential_1962}. Recent work in this area has utilized
 these residual Monte Carlo methods in particle transport
-\citep{evans_residual_2003}, giving accelerated convergence over
+\cite{evans_residual_2003}, giving accelerated convergence over
 traditional methods. As an expansion of Halton's work, Evans and
@@
 for radiation transport problems and improve performance significantly
-over Halton's sequential method \citep{evans_monte_2012}.
+over Halton's sequential method \cite{evans_monte_2012}.
 
@@
 the eigenvalues of a matrix $\ve{A}$ is the \textit{spectral radius},
-$\rho(\ve{A})$, defined by Saad \citep{saad_iterative_2003} as:
+$\rho(\ve{A})$, defined by Saad \cite{saad_iterative_2003} as:
 \begin{equation}
@@
 operator\footnote{It should be noted that non-diagonal splittings have
-  been recently explored \citep{srinivasan_monte_2010} and have the
+  been recently explored \cite{srinivasan_monte_2010} and have the
   potential to improve efficiency. However, it was observed in this
@@
 weight matrix. This decomposition, a generalization of Dimov's work
-\citep{dimov_new_1998}, is an extension of the original Neumann-Ulam
+\cite{dimov_new_1998}, is an extension of the original Neumann-Ulam
 scheme in that now a weight cutoff can be used to terminate a random
@@
 therefore has the same spectral radius. By defining the following
-inner product equivalence \citep{spanier_monte_1969}:
+inner product equivalence \cite{spanier_monte_1969}:
 \begin{equation}
@@
 In addition to the collision estimator, an additional estimator is
-available due to the work of Okten \citep{okten_solving_2005} that
+available due to the work of Okten \cite{okten_solving_2005} that
 uses the method of expected values as a means to improve the Monte
 Carlo estimate. As outlined by Spanier and Gelbard
-\citep{spanier_monte_1969}, the method of expected values is a
+\cite{spanier_monte_1969}, the method of expected values is a
 deterministic averaging of events that may potentially occur in the
@@
 presented a residual Monte Carlo method that moves towards exponential
-convergence rates \citep{halton_sequential_1962} and further refined
-his work some years later \citep{halton_sequential_1994}. Applications
+convergence rates \cite{halton_sequential_1962} and further refined
+his work some years later \cite{halton_sequential_1994}. Applications
 of his work by the transport community have confirmed convergence
-rates on the order of of $e^{-N}$ \citep{evans_residual_2003}. In much
+rates on the order of of $e^{-N}$ \cite{evans_residual_2003}. In much
 the same way as the projection methods outlined in
@@
 Carlo solution method that was not prohibited severely by the quality
-of the initial guess for the system \citep{evans_monte_2009} and later
+of the initial guess for the system \cite{evans_monte_2009} and later
 applied it more rigorously as a solution mechanism for the radiation
-diffusion equation \citep{evans_monte_2012}. With their new methods,
+diffusion equation \cite{evans_monte_2012}. With their new methods,
 they achieved identical numerical results as conventional Krylov
@@
 positive-definite but not necessarily symmetric problems, the minimal
-residual iteration \citep{saad_iterative_2003} can be used in
+residual iteration \cite{saad_iterative_2003} can be used in
 conjunction with MCSA where the residual vector, $\mathbf{r}$, defines

### 2013-07-24 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
adding summary sections in cconclusion chapter

@@
 We formulate the \textit{nonlinear problem} as follows
-\citep{knoll_jacobian-free_2004}:
+\cite{knoll_jacobian-free_2004}:
 \begin{equation}
@@
 limit, the iterations of Newton's method will converge the nonlinear
-residual quadratically \citep{kelley_iterative_1995}. Convergence
+residual quadratically \cite{kelley_iterative_1995}. Convergence
 criteria is set for stopping the iteration sequence based on the
@@
 such as a pressure correction strategy for solving the Navier-Stokes
-equations \citep{pletcher_computational_1997}, where the components of
+equations \cite{pletcher_computational_1997}, where the components of
 $\ve{u}$ are updated in a staggered fashion depending on the
@@
 inverted, resulting in an inexact Newton correction as initially
-described by Dembo and others \citep{dembo_inexact_1982}. For common
+described by Dembo and others \cite{dembo_inexact_1982}. For common
 sparse nonlinear systems, which in turn yield a sparse Jacobian
@@
   of the method or even prohibit convergence
-  \citep{eisenstat_choosing_1996}. In addition, \textit{globalization
+  \cite{eisenstat_choosing_1996}. In addition, \textit{globalization
     methods} may be used to modify the Newton correction in a more
@@
   improved when the initial guess for $\ve{u}$ is poor
-  \citep{pawlowski_globalization_2006}.
+  \cite{pawlowski_globalization_2006}.
 
@@
 the linear solver for generating the Newton correction
-\citep{kelley_iterative_1995}. As outlined in
+\cite{kelley_iterative_1995}. As outlined in
 Appendix~\ref{ch:linear_problem}, Krylov methods are robust and enjoy
@@
 memory usage. Several studies have been performed to investigate this
-\citep{mchugh_inexact_1993,knoll_newton-krylov_1995}. In their
+\cite{mchugh_inexact_1993,knoll_newton-krylov_1995}. In their
 numerical studies in 1995, Knoll and McHugh used the set of highly
@@
 difference equations for building the Jacobian
-\citep{knoll_newton-krylov_1995}. By recognizing that Krylov methods
+\cite{knoll_newton-krylov_1995}. By recognizing that Krylov methods
 only need the action of the operator on a vector instead of the
@@
 Jacobian-Free methods, and in particular \textit{Jacobian-Free
-  Newton-Krylov} (JFNK) methods \citep{knoll_jacobian-free_2004}, rely
+  Newton-Krylov} (JFNK) methods \cite{knoll_jacobian-free_2004}, rely
 on forming the action of the Jacobian on a vector as required by the
@@
 where $\epsilon$ is a small number typically on the order of machine
-precision. Kelley \citep{kelley_iterative_1995} points out a potential
+precision. Kelley \cite{kelley_iterative_1995} points out a potential
 downfall of this formulation in that if the discretization error in
@@
 accuracy of their original discretization in a completely automated
-way \citep{averick_computing_1994}.
-
-The work of Bartlett and others \citep{bartlett_automatic_2006}
+way \cite{averick_computing_1994}.
+
+The work of Bartlett and others \cite{bartlett_automatic_2006}
 extended initial Fortran-based work in the area of automatic
 differentiation implementations to leverage the parametric type and
-operator overloading features of C++ \citep{stroustrup_c++_1997}. They
+operator overloading features of C++ \cite{stroustrup_c++_1997}. They
 formulate the differentiation problem from an element viewpoint by
@@
 consistent solution scheme
-\citep{evans_development_2006,evans_enhanced_2007,gaston_parallel_2009,godoy_parallel_2012}. It
+\cite{evans_development_2006,evans_enhanced_2007,gaston_parallel_2009,godoy_parallel_2012}. It
 is common that for large scale simulations these problems are memory
@@
 Pawlowski's work on Newton-Krylov methods for the solution of these
-equations \citep{shadid_inexact_1997,pawlowski_globalization_2006}. We
+equations \cite{shadid_inexact_1997,pawlowski_globalization_2006}. We
 define these equations as follows:
@@
 In 1983 a set of benchmark solutions for the natural convection of air
-in a square cavity was published \citep{de_vahl_davis_natural_1983} as
+in a square cavity was published \cite{de_vahl_davis_natural_1983} as
 shown in Figure~\ref{fig:natural_convection_cavity} for the solution
@@
 To study performance for flow driven by inertial forces, the second
-benchmark problem given by Ghia \citep{ghia_high-re_1982} incorporates
+benchmark problem given by Ghia \cite{ghia_high-re_1982} incorporates
 a driver for the flow to introduce higher Reynolds numbers into the
@@
 both flow over a backward step and an outflow boundary condition
-\citep{gartling_test_1990}. Using the mass and momentum equations
+\cite{gartling_test_1990}. Using the mass and momentum equations
 while neglecting the energy equation, this problem utilizes a longer
@@
 method\footnote{The ML algebraic multigrid library in the Trilinos
-  scientific computing libraries \citep{heroux_overview_2005} was used
+  scientific computing libraries \cite{heroux_overview_2005} was used
   for all problems.}, a common technique for preconditioning these
-equations \citep{ghia_high-re_1982,evans_enhanced_2007} with the
+equations \cite{ghia_high-re_1982,evans_enhanced_2007} with the
 preconditioner reconstructed at every Newton step for the new
@@
 for fluid flow and other multiple physics simulations
-\citep{pawlowski_drekar_2012}. Drekar is a massively parallel physics
+\cite{pawlowski_drekar_2012}. Drekar is a massively parallel physics
 framework that utilizes Newton-Krylov methods to generate a fully
@@
 the same production implementation of GMRES from the Trilinos
-scientific computing library Aztec \citep{heroux_overview_2005} used
+scientific computing library Aztec \cite{heroux_overview_2005} used
 for the verification of MCSA for the $SP_N$ solutions in the previous
@@
 preconditioning for the linear solution at each Newton step,
-backtracking as outlined in Shadid's work \citep{shadid_inexact_1997}
+backtracking as outlined in Shadid's work \cite{shadid_inexact_1997}
 was used as a means of globalization to assist the Newton correction
@@
 computation as outlined by Pawlowski in
-\citep{pawlowski_globalization_2006} while the forcing term was again
-computed using method 2 from Shadid's work \citep{shadid_inexact_1997}
+\cite{pawlowski_globalization_2006} while the forcing term was again
+computed using method 2 from Shadid's work \cite{shadid_inexact_1997}
 at all Reynolds numbers.
@@
 assisting the Newton correction computation as outlined by Pawlowski
-in \citep{pawlowski_globalization_2006} while the forcing term was
+in \cite{pawlowski_globalization_2006} while the forcing term was
 again computed using method 2 from Shadid's work
-\citep{shadid_inexact_1997}.
+\cite{shadid_inexact_1997}.
 

### 2013-07-24 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
adding summary sections in cconclusion chapter

@@
 domain replication with parallelism exploited through individual
-histories \citep{alexandrov_efficient_1998} and in this chapter we
+histories \cite{alexandrov_efficient_1998} and in this chapter we
 will exploit particle transport algorithms to alleviate this. To
@@
 nearest-neighbor communication sequences, specifically, have poor
-algorithmic strong scaling \citep{gropp_high-performance_2001} in much
+algorithmic strong scaling \cite{gropp_high-performance_2001} in much
 the same way as a parallel matrix-vector multiply
@@
 will compute these eigenvalues by assuming eigenfunctions of the form
-\citep{leveque_finite_2007}:
+\cite{leveque_finite_2007}:
 \begin{equation}
@@
 Eq~(\ref{eq:adjoint_neumann_solution})
-\citep{dimov_new_1998,danilov_asymptotic_2000}. This random walk
+\cite{dimov_new_1998,danilov_asymptotic_2000}. This random walk
 length, or the number of transitions before the termination of a
@@
 reactor core Monte Carlo neutral particle simulations
-\citep{siegel_analysis_2012}.  To quantify the number of particles
+\cite{siegel_analysis_2012}.  To quantify the number of particles
 that leak out of the local domain they define a leakage fraction,
@@
 the Wigner rational approximation. Outlined more thoroughly by Hwang's
-chapter in \citep{azmy_nuclear_2010}, we will use both the Wigner
+chapter in \cite{azmy_nuclear_2010}, we will use both the Wigner
 rational approximation and the mean chord approximation as a means to
@@
 implemented in production implicit Monte Carlo codes
-\citep{brunner_efficient_2009}. We will adapt their algorithm and
+\cite{brunner_efficient_2009}. We will adapt their algorithm and
 directly apply it to a parallel formulation of the Neumann-Ulam
@@
 considerably more detail, the algorithm presented can be found in
-\citep{brunner_efficient_2009}. In their work they identify two data
+\cite{brunner_efficient_2009}. In their work they identify two data
 sets that are required to be communicated: the sharing of particles
@@
 algorithm, particularly some of those presented in
-\citep{brunner_comparison_2006}, use a master/slave approach as given
+\cite{brunner_comparison_2006}, use a master/slave approach as given
 by Figure~\ref{fig:master_comm_tree} for the completion of a transport
@@
 implementation is potentially simpler to understand than the binary
-tree approach, the authors of \citep{brunner_comparison_2006} observed
+tree approach, the authors of \cite{brunner_comparison_2006} observed
 that this type of termination pattern, even when implemented in a
@@
 applications for full-core light water reactor analysis
-\citep{wagner_hybrid_2010}. In their work, an extension of Brunner's,
+\cite{wagner_hybrid_2010}. In their work, an extension of Brunner's,
 their scheme employed similar parallel algorithms for particle
@@
 To demonstrate this, consider the example adapted from Mervin's work
-with Wagner and others in the same area \citep{mervin_variance_2012}
+with Wagner and others in the same area \cite{mervin_variance_2012}
 and presented in Figure~\ref{fig:msod_example}.
@@
 Belos, a linear solvers package in the Trilinos scientific computing
-libraries \citep{heroux_overview_2005}. In identical fashion to the
+libraries \cite{heroux_overview_2005}. In identical fashion to the
 verification for the $SP_N$ equations performed in
@@
 domain decomposed methods and specifically his discussion on how to
-measure their parallel performance \citep{keyes_how_1999}. All of
+measure their parallel performance \cite{keyes_how_1999}. All of
 these metrics are derived from two measured quantities: the number of
@@
   scaling} performance of an algorithm
-\citep{siegel_analysis_2012}. In this type of scaling study, the
+\cite{siegel_analysis_2012}. In this type of scaling study, the
 global problem size is fixed and the local problem size changes with
@@
 solver package within the Trilinos framework
-\citep{heroux_overview_2005}. In general, the Belos package produces
+\cite{heroux_overview_2005}. In general, the Belos package produces
 solutions to the neutron diffusion problem nearly two orders of
@@
 domain decomposition as outlined by Chapter 14 of Saad's text on
-iterative methods \citep{saad_iterative_2003}. Compared to traditional
+iterative methods \cite{saad_iterative_2003}. Compared to traditional
 domain decomposition methods we see many similarities including the
@@
 with a fixed point iteration used as a relaxation step\footnote{See
-  section 14.3.3 in \citep{saad_iterative_2003} for a full definition
+  section 14.3.3 in \cite{saad_iterative_2003} for a full definition
   of the additive Schwarz method.}. Both the fixed point iteration and
@@
   in strong scaling cases but is ineffective in weak scaling cases
-\item Multiple sets will never boost parallel efficiencies in both
-  weak and strong scaling cases due to additional communication
-  overhead from added reduction operations
+\item Multiple sets will never boost parallel efficiencies in the weak
+  scaling case due to additional communication overhead from added
+  reduction operations but will boost efficiencies beyond the strong
+  scaling wall
 \item Multiple sets offers a means to reduce time to solution by

### 2013-07-24 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding summary sections in cconclusion chapter

@@
 equations, closely following the work of Evans
-\citep{evans_simpli_2013}\footnote{The $SP_N$ derivations in this
+\cite{evans_simpli_2013}\footnote{The $SP_N$ derivations in this
   chapter and the associated appendices are heavily based on those
-  presented by Evans in \citep{evans_simpli_2013}}, in order to gain
+  presented by Evans in \cite{evans_simpli_2013}}, in order to gain
 full understanding of the underlying system and its potential behavior
@@
 time-independent neutron transport equation
-\citep{lewis_computational_1993}:
+\cite{lewis_computational_1993}:
 \begin{multline}
@@
 \label{sec:spn_equations}
-The $P_N$ equations as derived in \citep{lewis_computational_1993}
+The $P_N$ equations as derived in \cite{lewis_computational_1993}
 give $N+1$ coupled first-order equations capturing the spatial and
@@
 the diffusive term in each equation is only for a single
-'pseudo-moment' $u_n$. As noted by Evans \citep{evans_simpli_2013},
+'pseudo-moment' $u_n$. As noted by Evans \cite{evans_simpli_2013},
 lower order $SP_N$ approximations can be generated by setting higher
@@
 with the fission matrix $\mathbf{F}$ given by
-\citep{evans_simpli_2013}. We again apply the change of variables to
+\cite{evans_simpli_2013}. We again apply the change of variables to
 yield the set of multigroup pseudo-moments $\mathbb{U}_n$ where now
@@
 that the residual matrix formed by the factorization
-\citep{saad_iterative_2003}:
+\cite{saad_iterative_2003}:
 \begin{equation}
@@
 matrix directly by minimizing the Frobenius norm of the residual
-matrix \citep{saad_iterative_2003}:
+matrix \cite{saad_iterative_2003}:
 \begin{equation}
@@
 additional application of SPAINV
-(SPAINV-ILUT)\citep{saad_iterative_2003}. For the ILUT preconditioning
+(SPAINV-ILUT)\cite{saad_iterative_2003}. For the ILUT preconditioning
 case we have the following linear problem to solve for the substituted
@@
 parameter techniques can be applied to Monte Carlo to improve
-convergence as demonstrated by Dimov \citep{dimov_new_1998}. In this
+convergence as demonstrated by Dimov \cite{dimov_new_1998}. In this
 case, building an iteration matrix from
@@
 production implementations from the Trilinos scientific computing
-library Aztec \citep{heroux_overview_2005}. In addition to comparing
+library Aztec \cite{heroux_overview_2005}. In addition to comparing
 the cell and group fluxes, the number of eigenvalue iterations
@@
 with these production Krylov methods using CPU time as a measure
-\citep{evans_monte_2012}.
+\cite{evans_monte_2012}.
 

### 2013-07-24 doc/dissertation/document/frontmatter/acks.tex
adding summary sections in cconclusion chapter

--- 2012-10-24 doc/prelim/document/frontmatter/acks.tex
@@
-Great thanks are owed to advisor and friend Paul Wilson for providing
-me years of guidance and the opportunity to develop this work. Without
-him, this work would have never been possible.
-
-Tom Evans is responsible for presenting me with the seeds of this work
-and for that I thank him. His mentoring over the past few years has
-been invaluable.
-
-Roger Pawlowski and other members of the Consortium for Advanced
-Simulation of Light Water Reactors and Trilinos teams as well as many
-staff members at Oak Ridge National Laboratory have provided
-tremendous resources for my professional and technical development and
-have greatly facilitated this work.
-
 This work was performed under appointment to the Nuclear Regulatory
@@
 Department of Engineering Physics.
+
+Thank you to the Oak Ridge Leadership Computing Facility (OLCF) for
+providing computational resources and support.

### 2013-07-24 doc/dissertation/document/chapters/conclusion/conclusion.tex
adding most of the future work section

@@
 
-\subsection{Preconditioning}
-\label{subsec:future_preconditioning}
-
-Standalone Monte Carlo as a preconditioner? (e.g. for flex-GMRES)
-
-Stochastic variants of conventional preconditioners?
-
-Variance reduction for preconditioning?
-
-\subsection{Breaking away from $\rho(H) < 1$}
-\label{subsec:future_spec_rad}
-
-Monte Carlo methods of the 2nd Degree?
-
-Stochastic projection methods?
+Throughout this work, three salient issues were observed over all
+others. First, the performance of the MCSA implementation developed
+for this work was severely lacking in terms of time to solution when
+compared to production Krylov methods. In all cases, this difference
+was several orders of magnitude. Second, the preconditioning strategy
+developed by this work, although often successful in achieving
+convergence for difficult problems, creates many obstacles to a
+production implementation. These obstacles all primarily stem from the
+dense composite operators that for the probabilities and weights for
+the Monte Carlo game. Third, perhaps the most restrictive and
+unattractive piece of MCSA as a whole is the simple fact that the
+spectral radius of the physics operator iteration matrix be less than
+one. This was demonstrated to prohibit convergence for cases where the
+preconditioning was considered agressive and Krylov solvers exhibited
+excellent convergence properties. We will address each one of these
+problems in turn in this section and provide suggestions for their
+solution.
 
@@
 
-Implementation optimization - can we approach Krylov performance?
+When issues arising from preconditioning are not considered, there are
+two primary issues that can be resolved in the MCSA implementation
+developed by this work that will likely result in performance gains
+for serial computations. First, all random walks in the implementation
+have a state which changes as the random walk proceeds. These states
+are those given by Eq~(\ref{eq:mc_walk_permutation}). For random walks
+that are traversing a domain which has been decomposed in a parallel
+environment, there are two sets of these states: one that represents
+the states over the entire global domain and one that represents them
+only over the local subdomain. Each state in the system gets a unique
+index in both of these sets that act as absolute coordinates in the
+random walk and permit the construction of the Monte Carlo estimates
+and other data needed for the solution.
 
-Reduction in memory footprint
+A fundamental flaw of the implementation developed for this work is
+that those random walk states given by
+Eq~(\ref{eq:mc_walk_permutation}) were formulated such that they were
+always represented using the indices from the global set. However, all
+data in the system (e.g. weights, probabilities, tallies, etc.) are
+stored such that they are accessed using the local indexing
+scheme. Because of this, each time a random walk wants to sample its
+current probability distribution function, get a weight for its
+transition, or add its estimate to a tally, a call must first be made
+to translate the current global state of the random walk into its
+local state\footnote{This is a lookup in a hash table.}. What this
+then means is that we are performing orders of magnitude more
+operations during the Monte Carlo sequence than if the random walk
+sequence operated in the local index space rather than the global
+index space. Preliminary profiling of the implementation has shown
+that indeed a vast majority of the time in the Monte Carlo sequence is
+spent doing this index translation. Therefore, future work should
+perform the significant refactoring of the implementation required to
+achieve this.
 
-Multiple sets parallel overhead reduction
+A second mode of optimization to improve performance regards how
+discrete probability distribution functions are sampled in the
+implementation. At each transition in the random walk sequence, the
+discrete probability distribution function generated by the current
+row of the physics operator in which a random walk resides is sampled
+to get the next state in the sequence. This operation is also used
+when the right hand side of the system is sampled to acquire the
+starting state for each random walk in the problem through
+Eq~(\ref{eq:adjoint_source_probability}). The current implementation
+uses the inverse transform method for sampling where the a random
+number is generated and the cumulative distribution function is search
+to find the next discrete state. This algorithm has $O(log(N))$ time
+complexity for sampling and $O(N)$ time complexity for
+construction. The alias sampling method is suggested instead as it
+produces $O(1)$ complexity for sampling and $O(N)$ time complexity for
+construction. Profiling of the implementation shows that this
+improvement could also provide quality reductions in run time.
 
-Controling MCSA histories as a function of nonlinear iteration/forcing
-term?
+For the parallel MCSA algorithm, overall parallel scaling results
+indicated that the implementation was of good quality in both the
+strong and weak scaling cases. However, for the multiple set cases in
+the weak scaling analysis, it was noted that there was a growing
+parallel overhead as a function of core count. This was counter
+intuitive in that the size of the parallel group over which the set
+reductions were occuring was fixed at all core counts and therefore we
+do not expect the addition of such an operation to increase parallel
+run times. It did, however, and therefore future research should
+explore why the implementation for multiple sets generated parallel
+overhead that increased with core count instead of remaining constant.
+
+Nonlinear performance may also be improved by a closer relationship
+between MCSA and the Newton iteration. We saw in our reseaerch how the
+forcing term selected very tightly controls the convergence of the
+linear model in FANM. For MCSA, in addition to the residual tolerance
+that is dictated by this forcing term, convergence may also be
+controlled by adjusting the number of stochastic histories computed at
+each MCSA iteration. As the nonlinear iteration evolves, we may find
+it useful to adjust the number of histories used at each MCSA
+iteration in the same way as we adjust the convergence tolerance of
+the linear model through the forcing term. Future research is
+suggested to analyze the effects of the number of stochastic histories
+on the convergence of the linear and nonlinear models and how this may
+be controlled by the forcing term or another parameter.
+
+With these implementation improvements perhaps MCSA and FANM timing
+may approach that of production Krylov methods for equally conditioned
+problems. Next we discuss potential research in the area of
+preconditioning which may enable performance research for MCSA in the
+area of memory consumption.
+
+\subsection{Preconditioning}
+\label{subsec:future_preconditioning}
+
+In general, preconditioning developments in this work enabled much of
+the desired research to be performed but also prevented some of it. Of
+primary concern is the large memory footprint induced by the explicit
+algebraic preconditioning strategy outlined in
+\S~\ref{subsubsec:general_mcsa_preconditioning}. By preconditioning
+in this way, we were able to leverage all modern algebraic
+preconditioning strategies used in current physics practice. The
+downside is that in order to generate weights and probabilities with
+which to play the Monte Carlo game, me must first invert each of the
+preconditioners which in many cases means applying their inverse over
+each row in the system, and then we must perform matrix-matrix
+multiply operations to build the composite physics operator.
+
+These operations were observed to substantially increase runtimes of
+MCSA and consume unacceptable amounts of memory, mostly due to the
+fact that inverting the preconditioners create dense
+matrices. Although a good fraction of the elements in each row of the
+composite operator could be eliminated through the reduced domain
+approximation developed by this work, memory costs still prevented
+higher fidelity simulations from being performed. In addition, the
+reduced domain approximation does no prevent the large run times
+incurred for inverting the preconditioners. We therefore propose the
+following three developments:
+
+\begin{itemize}
+\item \textbf{Research and develop alternative sampling schemes for
+  the general left/right preconditioned Monte Carlo method not relying
+  on composite matrix generation.} Developments that break away from
+  Matrix-Matrix multiplications within the preconditioned solver
+  framework improve scalability. This means avoiding the generation of
+  the composite operator and instead playing the Monte Carlo game with
+  weights and probabilities formed by considering the original linear
+  operator and the preconditioners individually. This individual
+  consideration may come from an alternative algebraic formulation of
+  the general preconditioned MCSA method in
+  Eq~(\ref{eq:left_right_mcsa}) or from an alternative random sampling
+  strategy for the Monte Carlo game.
+\item \textbf{Develop Monte Carlo variance reduction based
+  preconditioning strategies that do not rely on an algebraic
+  preconditioner.}  Constructing the complete inverse of the
+  preconditioner is the core time and scalability constraint when
+  modern algebraic preconditioners are used. Building the inverse for
+  preconditioners where a sparsity pattern is not enforced typically
+  results in a dense matrix that destroys scalability for domain
+  decomposed parallel computations. Alleviating both of these issues
+  by stochastic preconditioning methods will be studied. Decades of
+  research and development in Monte Carlo variance reduction for
+  particle transport \cite{booth_1994} may be leveraged to form
+  similar strategies for linear systems. In the same way that this
+  work has been able to adapt modern particle transport strategies for
+  parallelism and apply them to linear systems, the same strategy may
+  be applied to variance reduction techniques to improve MCSA
+  convergence by modifying the weights and probabilities of the Monte
+  Carlo game as an effective means of preconditioning.
+\item \textbf{Study physics or PDE-based preconditioners for linear
+  systems that produce a reduced-order model of the original system to
+  which the Monte Carlo method may be applied.} As the Monte Carlo
+  method serves to construct the solution correction vector in the
+  MCSA method, the Monte Carlo method may be applied to a
+  reduced-order physics or PDE-based system that captures the essence
+  of the original system and still allows for acceleration of the
+  iterative procedure. By reducing the order of the system, we are
+  effectively generating a Monte Carlo problem within MCSA in which
+  the linear operator is either more sparse, has a better-conditioned
+  eigenvalue spectrum, or both. The reduced domain approximation has
+  shown that a purely algebraic strategy for reducing the order of the
+  Monte Carlo problem can result in improved overall timing
+  performance due to increased sparsity at the cost of iterative
+  performance as a result of loss of information. A more well informed
+  strategy for order reduction via either an intelligent
+  simplification of the PDEs that describe the system or a
+  simplification of the physics involved may prove to be more
+  effective strategy.
+\end{itemize}
+
+\subsection{Breaking Away from $\rho(H) < 1$}
+\label{subsec:future_spec_rad}
+
+All methods required quality preconditioning. Even a Krylov method may
+be effectively useful without it. However, these methods typically
+converge eventually whereas quality preconditioning may not be enough
+to gaurantee convergence of MCSA. We saw in our work that in many
+cases very agressive preconditioning was not enough to condition the
+eigenvalues of the system into a regime where MCSA would apply. The
+limitation that the spectral radius of the physics operator iteration
+matrix must be less than one is by far the most prohibitive component
+of MCSA. The fact that the Monte Carlo sequence is bsed on the
+Richardson iteration means that without sufficient preconditioning,
+even if this condition is achieved convergence may be incredibly
+slow. Therefore, we propose two elements of future work that aim to
+move away from this restriction, both fundamentally based on sampling
+a sequence not based on the Neumann series.
+
+\subsection{Monte Carlo Methods of the Second Degree}
+\label{subsubsec:2_degree_mc}
+
+\subsection{Stochastic Projection Methods}
+\label{subsubsec:stochastic_projection_methods}
 

### 2013-07-24 doc/dissertation/document/chapters/conclusion/conclusion.tex
finished conclusion draft

@@
 production implementation. These obstacles all primarily stem from the
-dense composite operators that for the probabilities and weights for
-the Monte Carlo game. Third, perhaps the most restrictive and
-unattractive piece of MCSA as a whole is the simple fact that the
-spectral radius of the physics operator iteration matrix be less than
-one. This was demonstrated to prohibit convergence for cases where the
-preconditioning was considered agressive and Krylov solvers exhibited
-excellent convergence properties. We will address each one of these
-problems in turn in this section and provide suggestions for their
-solution.
+dense composite operators that are required to form the probabilities
+and weights for the Monte Carlo game. Third, perhaps the most
+restrictive and unattractive piece of MCSA as a whole is the simple
+fact that the spectral radius of the physics operator iteration matrix
+be less than one. This was demonstrated to prohibit convergence for
+cases where the preconditioning was considered agressive and Krylov
+solvers exhibited excellent convergence properties. We will address
+each one of these problems in turn in this section and provide
+suggestions for their solution.
 
@@
 uses the inverse transform method for sampling where the a random
-number is generated and the cumulative distribution function is search
-to find the next discrete state. This algorithm has $O(log(N))$ time
-complexity for sampling and $O(N)$ time complexity for
-construction. The alias sampling method is suggested instead as it
-produces $O(1)$ complexity for sampling and $O(N)$ time complexity for
-construction. Profiling of the implementation shows that this
-improvement could also provide quality reductions in run time.
+number is generated and the cumulative distribution function is
+queried with a binary seach to find the next discrete state. This
+algorithm has $O(log(N))$ time complexity for sampling and $O(N)$ time
+complexity for construction. The alias sampling method is suggested
+instead as it produces $O(1)$ complexity for sampling and $O(N)$ time
+complexity for construction \cite{smith_analysis_2005}. Profiling of
+the implementation shows that this improvement could also provide
+quality reductions in run time.
 
@@
 algebraic preconditioning strategy outlined in
-\S~\ref{subsubsec:general_mcsa_preconditioning}. By preconditioning
-in this way, we were able to leverage all modern algebraic
-preconditioning strategies used in current physics practice. The
-downside is that in order to generate weights and probabilities with
-which to play the Monte Carlo game, me must first invert each of the
-preconditioners which in many cases means applying their inverse over
-each row in the system, and then we must perform matrix-matrix
-multiply operations to build the composite physics operator.
+\S~\ref{subsubsec:general_mcsa_preconditioning}. By preconditioning in
+this way, we were able to leverage modern algebraic preconditioning
+strategies used in current physics applications and enable solutions
+for both neutronics and fluid flow simulations. The downside is that
+in order to generate weights and probabilities with which to play the
+Monte Carlo game, we must first invert each of the preconditioners
+which in many cases means applying their inverse over each row in the
+system, and then we must perform matrix-matrix multiply operations to
+build the composite physics operator.
 
@@
 MCSA and consume unacceptable amounts of memory, mostly due to the
-fact that inverting the preconditioners create dense
+fact that inverting the preconditioners created dense
 matrices. Although a good fraction of the elements in each row of the
@@
 higher fidelity simulations from being performed. In addition, the
-reduced domain approximation does no prevent the large run times
+reduced domain approximation does not prevent the large run times
 incurred for inverting the preconditioners. We therefore propose the
-following three developments:
+following three developments in preconditioning for Monte Carlo
+methods:
 
@@
   on composite matrix generation.} Developments that break away from
-  Matrix-Matrix multiplications within the preconditioned solver
+  matrix-matrix multiplications within the preconditioned solver
   framework improve scalability. This means avoiding the generation of
@@
   decomposed parallel computations. Alleviating both of these issues
-  by stochastic preconditioning methods will be studied. Decades of
+  by stochastic preconditioning methods should be studied. Decades of
   research and development in Monte Carlo variance reduction for
@@
   parallelism and apply them to linear systems, the same strategy may
-  be applied to variance reduction techniques to improve MCSA
+  possibly be applied to variance reduction techniques to improve MCSA
   convergence by modifying the weights and probabilities of the Monte
@@
   simplification of the PDEs that describe the system or a
-  simplification of the physics involved may prove to be more
+  simplification of the physics involved may prove to be a more
   effective strategy.
@@
 
-\subsection{Monte Carlo Methods of the Second Degree}
+\subsubsection{Monte Carlo Methods of the Second Degree}
 \label{subsubsec:2_degree_mc}
 
-\subsection{Stochastic Projection Methods}
+Iterative methods of the $n^{th}$ degree produce solutions at each
+iteration of the following form:
+\begin{equation}
+  \ve{x}^{k+1} = F(\ve{A},\ve{b},\ve{x}^{k},\ve{x}^{k-1},\dots,\ve{x}^{k-n+1})\:,
+\label{eq:nth_degree_iter}
+\end{equation}
+where $\ve{A}$ is the linear operator $\ve{b}$ is the right hand side
+as defined by Eq~(\ref{eq:linear_problem}). Richardson's iteration is
+an iterative method of the first degree such that $\ve{x}^{k+1} =
+F(\ve{A},\ve{b},\ve{x}^{k})$. To move the Monte Carlo method away from
+the spectral radius limitation and therefore we seek higher order
+methods as outlined by Hallett \cite{hughes_hallett_second-order_1984}
+that have garanteed convergence at all eigenvalue spectra. In his
+work, Hallett builds a second order method such that $\ve{x}^{k+1} =
+F(\ve{A},\ve{b},\ve{x}^{k-1})$. Such a method results from mapping the
+original Richardson iteration into a different of eigenvalues in the
+complex plane that have a spectral radius less than one, thus
+garanteeing convergence. Derived as expansions of the iteration
+sequence in Chebyshev polynomials, Hallett arrives at simple
+formulation for the second degree method that requires knowledge of
+the bounding curve in the complex plane of the eigenvalue spectra for
+good convergence properties.
+
+As it turns out, Dimov has already applied such a technique to the
+Monte Carlo method but arrived at the same conclusions in a different
+way \cite{dimov_parallel_2001,dimov_new_1998}. In his work, Dimov
+instead formulated the random walk sequence by instead building the
+Neumann series from the \textit{resolvent} of the operator. The
+resolvent operator is defined as that which spans all of the complex
+plane but has singularities at the points that are eigenvalues of the
+original operator. By building the Monte Carlo sequence from this
+operator, Dimov arrived at a Monte Carlo method which in fact was
+equivalent to a fixed point method of the second degree and utilized
+effectively the same spectral mapping procedure as Hallett. Just like
+the deterministic case, some knowledge of the eigenvalue spectrum of
+the system was required to ensure convergence. These methods should be
+researched and developed further in the context of MCSA as a potential
+means of always gauranteeing convergence of the algorithm regardless
+of the eigenvalue spectrum.
+
+\subsubsection{Stochastic Projection Methods}
 \label{subsubsec:stochastic_projection_methods}
+
+As a final means of improvement we consider stochastic projection
+methods as developed by Sabelfeld \cite{hutchison_stochastic_2011}. In
+his work, Sabelfeld claims his method is 'beyond Markov chains and von
+Neumann-Ulam scheme'. To a certain extent this is true, his method is
+based on a projection method and should always converge. In his work,
+he utilizes a Kaczmarz iteration:
+\begin{equation}
+  \ve{x}^{k+1} = \ve{x}^k + \lambda_k \frac{b_i - \langle \ve{a}_i
+    \ve{x}^k \rangle}{||\ve{a}_i||}\ve{a}_i\:,
+  \label{eq:kaczmarz_iteration}
+\end{equation}
+where $\ve{a}_i$ is the $i^{th}$ row vector of the operator and
+$\lambda_k$ is the relaxation parameter.
+
+The stochastic nature of the algorithm comes from effectively
+modifying the algorithm such that the value of $i$ is now selected at
+random. This is interesting but arguably does not serve our purposes
+at well. We have an interest in a fully stochastic process like a
+random walk sequence because of its potential benefits in terms of
+resiliency and the ability to use superposition to combine results
+from replicated problems to enhance performance. In Sabelfeld's work,
+the randomness is used to select which deterministic iteration step
+will be performed and does not meet either of these
+requirements. However, that does not mean there is nothing to take
+from this work. If one writes Eq~\ref{eq:kaczmarz_iteration} in terms
+of applications of projection operators as that outlined in
+\cite{tanabe_projection_1971} we see that this actually takes a
+similar form as the Neumann series and could therefore be used in a
+random walk sequence. Research studying random walks based on
+projection sequences is suggested as a means of moving beyond the
+spectral radius limitation.
 
@@
 \label{sec:closing}
+
+This work was performed with the goal of investigating new solution
+techniques based on Monte Carlo Synthetic Acceleration and how they
+perform when applied to real problems in nuclear engineering. These
+methods aim to improve over conventional methods when looking towards
+the advanced computer architecutures of the near future. Parallelism
+at high levels of concurrency, a potential reduction in memory
+footprint, and a natural element of resiliency deriving from the
+stochastic nature of the algorithm motivated us to research their
+strengths and weaknesses and develop solutions to further their
+applicability.
+
+Before this work was performed, MCSA had yet to be applied to the
+neutron transport problem, asymmetric systems, and nonlinear problems
+including fluid flow systems. Techniques were developed by this work
+that enabled solutions for all of these problems using MCSA. In many
+cases, it was found that the iterative performance of the MCSA-based
+methods were superior to that of conventional solution techniques
+based on Krylov methods. Furthermore, with improved implementation
+performance, future problems may be solved in less time and more
+efficiently using MCSA due to this demonstrated improved iterative
+performance.
+
+When this work began, the Neumann-Ulam sequence and MCSA in general
+had yet to be parallelized in a domain decomposed framework. This
+research developed a new parallel algorithm based on Monte Carlo
+particle transport that permitted MCSA solutions to be generated on
+$O(100,000)$ cores on a leadership class computing facility. The new
+algorithm was also demonstrated to achieve better strong and weak
+scaling than conventional Krylov methods at high levels of
+concurrency. If future work can improve the time to solution using
+MCSA through optimization, then this improved scaling demonstrated by
+the algorithm can potentially improve the time to solution for the
+engineer and more efficiently leverage high performance computing
+hardware.
+
+With this work, a new foundation for Monte Carlo Synethic Acceleration
+methods has been laid with new theory, algorithms, and numerical
+experiments that demonstrate both their potential to make great
+contributions to nuclear engineering calculations in the future and
+where the research must go in order to achieve this. By advancing
+these techniques, we look forward to future work that will enable the
+improved design and analysis of nuclear systems.

### 2013-07-24 doc/dissertation/document/chapters/introduction/introduction.tex
finished conclusion draft

@@
 system of equations for detailed neutronics models of fission
-reactor-based systems\cite{evans_simpli_2013}. By fully forming these
+reactor-based systems\cite{evans_simplified_2013}. By fully forming these
 equations and formulating them as a linear algebra problem instead of

### 2013-07-24 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
finished conclusion draft

@@
 equations, closely following the work of Evans
-\cite{evans_simpli_2013}\footnote{The $SP_N$ derivations in this
+\cite{evans_simplified_2013}\footnote{The $SP_N$ derivations in this
   chapter and the associated appendices are heavily based on those
-  presented by Evans in \cite{evans_simpli_2013}}, in order to gain
+  presented by Evans in \cite{evans_simplified_2013}}, in order to gain
 full understanding of the underlying system and its potential behavior
@@
 the diffusive term in each equation is only for a single
-'pseudo-moment' $u_n$. As noted by Evans \cite{evans_simpli_2013},
+'pseudo-moment' $u_n$. As noted by Evans \cite{evans_simplified_2013},
 lower order $SP_N$ approximations can be generated by setting higher
@@
 with the fission matrix $\mathbf{F}$ given by
-\cite{evans_simpli_2013}. We again apply the change of variables to
+\cite{evans_simplified_2013}. We again apply the change of variables to
 yield the set of multigroup pseudo-moments $\mathbb{U}_n$ where now

### 2013-07-25 doc/dissertation/document/chapters/conclusion/conclusion.tex
some cleanup in the concllusion

@@
 instead as it produces $O(1)$ complexity for sampling and $O(N)$ time
-complexity for construction \cite{smith_analysis_2005}. Profiling of
-the implementation shows that this improvement could also provide
-quality reductions in run time.
+complexity for construction \cite{smith_analysis_2005}. Sabelfeld
+outlines the use of this sampling technique in his Monte Carlo
+algorithms \cite{sabelfeld_sparsified_2009}. Profiling of the
+implementation shows that this improvement could also provide quality
+reductions in run time.
 
@@
 
-All methods required quality preconditioning. Even a Krylov method may
-be effectively useful without it. However, these methods typically
-converge eventually whereas quality preconditioning may not be enough
-to gaurantee convergence of MCSA. We saw in our work that in many
-cases very agressive preconditioning was not enough to condition the
-eigenvalues of the system into a regime where MCSA would apply. The
-limitation that the spectral radius of the physics operator iteration
-matrix must be less than one is by far the most prohibitive component
-of MCSA. The fact that the Monte Carlo sequence is bsed on the
-Richardson iteration means that without sufficient preconditioning,
-even if this condition is achieved convergence may be incredibly
-slow. Therefore, we propose two elements of future work that aim to
-move away from this restriction, both fundamentally based on sampling
-a sequence not based on the Neumann series.
+All numerical methods for solving linear problems require quality
+preconditioning. Even a Krylov method may be effectively useful
+without it. However, these methods typically converge eventually
+whereas quality preconditioning may not be enough to gaurantee
+convergence of MCSA. We saw in our work that in many cases very
+agressive preconditioning was not enough to condition the eigenvalues
+of the system into a regime where MCSA would apply. The limitation
+that the spectral radius of the physics operator iteration matrix must
+be less than one is by far the most prohibitive component of MCSA. The
+fact that the Monte Carlo sequence is based on the Richardson
+iteration means that without sufficient preconditioning, even if this
+condition is achieved convergence may be incredibly slow. Therefore,
+we propose two elements of future work that aim to move away from this
+restriction, both fundamentally based on sampling a sequence not based
+on the Neumann series.
 
@@
 methods as outlined by Hallett \cite{hughes_hallett_second-order_1984}
-that have garanteed convergence at all eigenvalue spectra. In his
+that have gauranteed convergence at all eigenvalue spectra. In his
 work, Hallett builds a second order method such that $\ve{x}^{k+1} =
@@
 
-As it turns out, Dimov has already applied such a technique to the
+As it turns out, Sabelfeld has already applied such a technique to the
 Monte Carlo method but arrived at the same conclusions in a different
-way \cite{dimov_parallel_2001,dimov_new_1998}. In his work, Dimov
-instead formulated the random walk sequence by instead building the
-Neumann series from the \textit{resolvent} of the operator. The
-resolvent operator is defined as that which spans all of the complex
-plane but has singularities at the points that are eigenvalues of the
-original operator. By building the Monte Carlo sequence from this
-operator, Dimov arrived at a Monte Carlo method which in fact was
-equivalent to a fixed point method of the second degree and utilized
-effectively the same spectral mapping procedure as Hallett. Just like
-the deterministic case, some knowledge of the eigenvalue spectrum of
-the system was required to ensure convergence. These methods should be
+way \cite{sabelfeld_sparsified_2009}\footnote{Although the technique
+  is outlined in this document, Sabelfeld cites his previous work in
+  this area in 1982.} with this work effectively re-developed by Dimov
+without citation of Sabelfeld
+\cite{dimov_parallel_2001,dimov_new_1998}. In his work, Sabelfeld
+formulated the random walk sequence by instead building the Neumann
+series from the \textit{resolvent} of the operator. The resolvent
+operator is defined as that which spans all of the complex plane but
+has singularities at the points that are eigenvalues of the original
+operator. By building the Monte Carlo sequence from this operator,
+Sabelfeld arrived at a Monte Carlo method which in fact was equivalent
+to a fixed point method of the second degree and utilized effectively
+the same spectral mapping procedure as Hallett. Just like the
+deterministic case, some knowledge of the eigenvalue spectrum of the
+system was required to ensure convergence. These methods should be
 researched and developed further in the context of MCSA as a potential
@@
 where $\ve{a}_i$ is the $i^{th}$ row vector of the operator and
-$\lambda_k$ is the relaxation parameter.
+$\lambda_k$ is a relaxation parameter.
 
@@
 will be performed and does not meet either of these
-requirements. However, that does not mean there is nothing to take
+requirements. However, that does not mean there is nothing to learn
 from this work. If one writes Eq~\ref{eq:kaczmarz_iteration} in terms
@@
 similar form as the Neumann series and could therefore be used in a
-random walk sequence. Research studying random walks based on
-projection sequences is suggested as a means of moving beyond the
-spectral radius limitation.
+random walk sequence. In addition, perhaps the above algorithm can be
+modified such that the randomness of the projection sequence permits
+improved results via the superposition principle. Research studying
+random walks based on projection sequences is therefore suggested as a
+means of moving beyond the spectral radius limitation.
 
@@
 neutron transport problem, asymmetric systems, and nonlinear problems
-including fluid flow systems. Techniques were developed by this work
+including fluid flow systems. Techniques was developed by this work
 that enabled solutions for all of these problems using MCSA. In many
@@
 engineer and more efficiently leverage high performance computing
-hardware.
+hardware for analysis.
 

### 2013-07-25 doc/dissertation/document/frontmatter/abstract.tex
adding dissertation abstract

@@
-\svnidlong{$LastChangedBy$}{$LastChangedRevision$}{$LastChangedDate$}{$HeadURL: http://freevariable.com/dissertation/branches/diss-template/frontmatter/abstract.tex $}
-\vcinfo{}
+Predictive modeling and simulation is an important and developing
+component of modern nuclear reactor analysis. New high performance
+computing hardware and algorithms that leverage this hardware are
+required to help meet the meet the goals and requirements of
+engineering analysis. Current practices in neutron transport and fluid
+flow leverage methods based on subspace iterations to achieve good
+convergence properties and quality parallel scaling on large
+machines. However, the next generation of machines is expected to
+possess both high levels of concurrency and low memory availability
+per core compared to today's machines. Therefore, we seek new
+algorithms to solve these problems in reactor analysis that have the
+potential to perform well on forthcoming hardware with the goal of
+improving both iterative performance and parallel scalability.
+
+This work researches and develops Monte Carlo Synthetic Acceleration
+(MCSA) methods as a new class of solution techniques for discrete
+neutron transport and fluid flow problems. Monte Carlo Synthetic
+Acceleration methods use a traditional Monte Carlo process to
+approximate the solution to the discrete problem as a means of
+accelerating traditional fixed-point methods. To apply these methods
+to neutronics and fluid flow and determine the feasibility of these
+methods on modern hardware, three complementary research and
+development exercises are performed.
+
+First, solutions to the $SP_N$ discretization of the linear Boltzmann
+neutron transport equation are obtained using Monte Carlo Synthetic
+Acceleration with a difficult criticality calculation for a light
+water reactor fuel assembly used as the driving problem. It was
+discovered that basic Jacobi-based preconditioning was not sufficient
+to achieve convergence due to the dominating effect of the light water
+moderator and therefore a group of modern preconditioning strategies
+are researched. The new preconditioning techniques, however, create
+unwanted growth in both memory consumption and computation time for
+these problems. Monte Carlo Synthetic Acceleration was verified to
+produce the correct solution for the fuel assembly criticality problem
+when compared to conventional Krylov methods and demonstrated improved
+iterative performance over GMRES by converging in fewer iterations
+when using the same preconditioning.
+
+Second, solutions to the full Navier-Stokes equations were obtained by
+developing the Forward-Automated Newton-MCSA (FANM) scheme for
+nonlinear systems based on Newton's method. Three difficult fluid
+benchmark problems in both convective and driven flow regimes were
+used to drive the research and development of the method. It was found
+that for all benchmarks that FANM produced identical solutions to
+modern Newton-Krylov methods. For each benchmark, performance
+comparisons between FANM and Newton-Krylov were also conducted. For 8
+out of 12 benchmark cases, it was found that FANM had better iterative
+performance than the Newton-Krylov method by converging the nonlinear
+residual in fewer linear solver iterations with the same
+preconditioning.
+
+Third, a new domain decomposed algorithm to parallelize MCSA aimed at
+leveraging leadership-class computing facilities was developed by
+utilizing parallel strategies from the radiation transport
+community. The new algorithm utilizes the Multiple-Set
+Overlapping-Domain strategy in an attempt to reduce parallel overhead
+and add a natural element of replication to the algorithm. A large
+number of parallel scaling studies using a neutron diffusion problem
+were performed on the Titan Cray XK7 machine at Oak Ridge National
+Laboratory with up to 65,356 cores used in the strong scaling studies
+and 131,072 cores used in the weak scaling studies. It was found that
+for the current implementation of MCSA, both weak and strong scaling
+were superior to production implementations of Krylov methods. In
+addition, the replication in the algorithm was able to boost parallel
+efficiencies by up to 20\% at large core counts in the strong scaling
+studies. By using the principle of superposition, replicating the
+Monte Carlo sequence was also demonstrated to enhance time to
+solution. Finally, it was discovered that MCSA exhibits the most
+scalable behavior when it is formulated as a stochastic realization of
+an additive Schwarz method.
+
+Each of these research and development activities meet the goals of
+this work by producing methods that in many cases were demonstrated to
+converge in fewer iterations than conventional methods and exhibit
+better parallel efficiencies at high levels of concurrency. All the
+components of this work are required to enable future simulation of
+nonlinear physics problems on the next generation hardware. To enable
+high performance simulations with FANM, we require the MCSA techniques
+developed in the research on the neutron transport problem and in turn
+effectively leveraging MCSA on leadership-class hardware requires the
+new parallel algorithm developed by this work. With the further
+development of these methods based on this work, future nuclear
+reactor simulations may be enabled that have an improved time to
+solution and more efficiently utilize the available computational
+resources.

### 2013-07-25 doc/dissertation/document/chapters/conclusion/conclusion.tex
pushing full draft of disseration

@@
 problems on future hardware. To do this, we carried out three research
-and developement activities. First, we applied MCSA to the $SP_N$ form
+and development activities. First, we applied MCSA to the $SP_N$ form
 of the neutron transport equation and analyzed the preconditioning
@@
 we parallelized MCSA such that it may be applied to physics problems
-in a leadership-class computing environment.
+on leadership-class computing platforms.
 
@@
 approximation, a set of $SP_N$ calculations using up to 4 energy
-groups were enabled for the fuel assembly problem. To verify MCSA, the
+groups was enabled for the fuel assembly problem. To verify MCSA, the
 fuel assembly problem was solved with different energy groups using
@@
 and generate the same k-eigenvalue in the same number of eigenvalue
-iterations.
+iterations when used in conjuction with an eigenvalue solver.
 
@@
 performance over GMRES while a Conjugate Gradient-based method
-performed better using the same preconditioining. Improved performance
+performed better using the same preconditioning. Improved performance
 over GMRES meets the goal of improving iterative performance to a
@@
 method to perform better as this structure can be more readily
-utilized than the more general GMRES and MCSA methods. In terms of CPU
-time, MCSA was observed to perform $O(100)$ times slower than the
-Krylov methods due to both lack of optimization in the random walk
-sequence and the explicit preconditioning strategy.
+leveraged to improve performance than with the more general GMRES and
+MCSA methods. In terms of CPU time, MCSA was observed to perform
+$O(100)$ times slower than the Krylov methods due to both lack of
+optimization in the random walk sequence and the explicit
+preconditioning strategy.
 
@@
 
-Iterative performance for the problems dominated by intertial driven
+Iterative performance for the problems dominated by inertial driven
 flow varied depending on the situation. For the lid driven cavity
@@
 The backward facing step problem at low Reynolds number gave better
-performance with FANM convergining in fewer linear solver
+performance with FANM converging in fewer linear solver
 iterations. However, as the Reynolds number was increased so did the
@@
 backward facing step problem. For the first two cases, the timing
-differences were observed to be qualititavely the same as those
+differences were observed to be qualitatively the same as those
 observed for the $SP_N$ performance analysis. Again, this comes from
@@
 and increase parallel efficiencies. Scaling studies showed that in the
-strong case overlap in small quanities on the order of the
+strong case overlap in small quantities on the order of the
 mean-free-path of a stochastic history in the simulation could boost
 parallel efficiencies by a few percent. However, it was found that
-this additional overal was not effective in boosting strong scaling
-efficiencies. It was found that overall overlap was not very effective
-due to the fact that the parallel communication saved during the Monte
-Carlo transport sequence is simply deferred until after transport is
-complete when it manifests itself as an overlapping parallel vector
-reduction operation.
+this additional overlap was not effective in boosting strong scaling
+efficiencies. It was found that in general overlap was not very
+effective due to the fact that the parallel communication saved during
+the Monte Carlo transport sequence is simply deferred until after
+transport is complete when it manifests itself as an overlapping
+parallel vector reduction operation.
 
@@
 transport problem, using MCSA means that we can actually make a
-physical copy of the problem on the machine and can combine seperate
+physical copy of the problem on the machine and can combine separate
 Monte Carlo solutions for each copy through superposition to enhance
@@
 parallel efficiencies, it was postulated that very little
-domain-to-domain communication of histories was occuring in the first
+domain-to-domain communication of histories was occurring in the first
 place and then subsequently verified using the analytic relationships
@@
 be less than one. This was demonstrated to prohibit convergence for
-cases where the preconditioning was considered agressive and Krylov
+cases where the preconditioning was considered aggressive and Krylov
 solvers exhibited excellent convergence properties. We will address
@@
 to translate the current global state of the random walk into its
-local state\footnote{This is a lookup in a hash table.}. What this
+local state\footnote{This is a look-up in a hash table.}. What this
 then means is that we are performing orders of magnitude more
@@
 number is generated and the cumulative distribution function is
-queried with a binary seach to find the next discrete state. This
+queried with a binary search to find the next discrete state. This
 algorithm has $O(log(N))$ time complexity for sampling and $O(N)$ time
@@
 the weak scaling analysis, it was noted that there was a growing
-parallel overhead as a function of core count. This was counter
-intuitive in that the size of the parallel group over which the set
-reductions were occuring was fixed at all core counts and therefore we
-do not expect the addition of such an operation to increase parallel
-run times. It did, however, and therefore future research should
-explore why the implementation for multiple sets generated parallel
-overhead that increased with core count instead of remaining constant.
+parallel overhead as a function of core count. This was
+counter-intuitive in that the size of the parallel group over which
+the set reductions were occurring was fixed at all core counts and
+therefore we do not expect the addition of such an operation to
+increase parallel run times. It did, however, and therefore future
+research should explore why the implementation for multiple sets
+generated parallel overhead that increased with core count instead of
+remaining constant.
 
 Nonlinear performance may also be improved by a closer relationship
-between MCSA and the Newton iteration. We saw in our reseaerch how the
+between MCSA and the Newton iteration. We saw in our research how the
 forcing term selected very tightly controls the convergence of the
@@
 in order to generate weights and probabilities with which to play the
-Monte Carlo game, we must first invert each of the preconditioners
+Monte Carlo game we must first invert each of the preconditioners
 which in many cases means applying their inverse over each row in the
-system, and then we must perform matrix-matrix multiply operations to
-build the composite physics operator.
+system and then perform matrix-matrix multiply operations to build the
+composite physics operator.
 
@@
 
-\begin{itemize}
-\item \textbf{Research and develop alternative sampling schemes for
+\paragraph{\textbf{Research and develop alternative sampling schemes for
   the general left/right preconditioned Monte Carlo method not relying
-  on composite matrix generation.} Developments that break away from
+  on composite matrix generation.}} Developments that break away from
 matrix-matrix multiplications within the preconditioned solver
@@
 strategy for the Monte Carlo game.
-\item \textbf{Develop Monte Carlo variance reduction based
+\paragraph{\textbf{Develop Monte Carlo variance reduction based
   preconditioning strategies that do not rely on an algebraic
-  preconditioner.}  Constructing the complete inverse of the
+  preconditioner.}} Constructing the complete inverse of the
   preconditioner is the core time and scalability constraint when
@@
   Carlo game as an effective means of preconditioning.
-\item \textbf{Study physics or PDE-based preconditioners for linear
+\paragraph{\textbf{Study physics or PDE-based preconditioners for linear
   systems that produce a reduced-order model of the original system to
-  which the Monte Carlo method may be applied.} As the Monte Carlo
-  method serves to construct the solution correction vector in the
-  MCSA method, the Monte Carlo method may be applied to a
-  reduced-order physics or PDE-based system that captures the essence
-  of the original system and still allows for acceleration of the
-  iterative procedure. By reducing the order of the system, we are
-  effectively generating a Monte Carlo problem within MCSA in which
-  the linear operator is either more sparse, has a better-conditioned
-  eigenvalue spectrum, or both. The reduced domain approximation has
-  shown that a purely algebraic strategy for reducing the order of the
-  Monte Carlo problem can result in improved overall timing
-  performance due to increased sparsity at the cost of iterative
-  performance as a result of loss of information. A more well informed
-  strategy for order reduction via either an intelligent
-  simplification of the PDEs that describe the system or a
-  simplification of the physics involved may prove to be a more
-  effective strategy.
-\end{itemize}
+  which the Monte Carlo method may be applied.}} As the Monte Carlo
+method serves to construct the solution correction vector in the MCSA
+method, the Monte Carlo method may be applied to a reduced-order
+physics or PDE-based system that captures the essence of the original
+system and still allows for acceleration of the iterative
+procedure. By reducing the order of the system, we are effectively
+generating a Monte Carlo problem within MCSA in which the linear
+operator is either more sparse, has a better-conditioned eigenvalue
+spectrum, or both. The reduced domain approximation has shown that a
+purely algebraic strategy for reducing the order of the Monte Carlo
+problem can result in improved overall timing performance due to
+increased sparsity at the cost of iterative performance as a result of
+loss of information. A more well informed strategy for order reduction
+via either an intelligent simplification of the PDEs that describe the
+system or a simplification of the physics involved may prove to be a
+more effective strategy.
 
@@
 without it. However, these methods typically converge eventually
-whereas quality preconditioning may not be enough to gaurantee
+whereas quality preconditioning may not be enough to guarantee
 convergence of MCSA. We saw in our work that in many cases very
-agressive preconditioning was not enough to condition the eigenvalues
+aggressive preconditioning was not enough to condition the eigenvalues
 of the system into a regime where MCSA would apply. The limitation
@@
 \end{equation}
-where $\ve{A}$ is the linear operator $\ve{b}$ is the right hand side
-as defined by Eq~(\ref{eq:linear_problem}). Richardson's iteration is
-an iterative method of the first degree such that $\ve{x}^{k+1} =
-F(\ve{A},\ve{b},\ve{x}^{k})$. To move the Monte Carlo method away from
-the spectral radius limitation and therefore we seek higher order
-methods as outlined by Hallett \cite{hughes_hallett_second-order_1984}
-that have gauranteed convergence at all eigenvalue spectra. In his
-work, Hallett builds a second order method such that $\ve{x}^{k+1} =
-F(\ve{A},\ve{b},\ve{x}^{k-1})$. Such a method results from mapping the
-original Richardson iteration into a different of eigenvalues in the
-complex plane that have a spectral radius less than one, thus
-garanteeing convergence. Derived as expansions of the iteration
-sequence in Chebyshev polynomials, Hallett arrives at simple
+where $\ve{A}$ is the linear operator and $\ve{b}$ is the right hand
+side as defined by Eq~(\ref{eq:linear_problem}). Richardson's
+iteration is an iterative method of the first degree such that
+$\ve{x}^{k+1} = F(\ve{A},\ve{b},\ve{x}^{k})$. To move the Monte Carlo
+method away from the spectral radius limitation we therefore we seek
+higher order methods as outlined by Hallett
+\cite{hughes_hallett_second-order_1984} that have guaranteed
+convergence at all eigenvalue spectra. In his work, Hallett builds a
+second order method such that $\ve{x}^{k+1} =
+F(\ve{A},\ve{b},\ve{x}^{k},\ve{x}^{k-1})$. Such a method results from
+mapping the original Richardson iteration onto a different set of
+eigenvalues in the complex plane that have a spectral radius less than
+one, thus guaranteeing convergence. Derived as expansions of the
+iteration sequence in Chebyshev polynomials, Hallett arrives at simple
 formulation for the second degree method that requires knowledge of
@@
   is outlined in this document, Sabelfeld cites his previous work in
-  this area in 1982.} with this work effectively re-developed by Dimov
-without citation of Sabelfeld
+  this area in 1982.} with this work effectively re-developed later by
+Dimov without citation of Sabelfeld
 \cite{dimov_parallel_2001,dimov_new_1998}. In his work, Sabelfeld
@@
 researched and developed further in the context of MCSA as a potential
-means of always gauranteeing convergence of the algorithm regardless
+means of always guaranteeing convergence of the algorithm regardless
 of the eigenvalue spectrum.
@@
 his work, Sabelfeld claims his method is 'beyond Markov chains and von
-Neumann-Ulam scheme'. To a certain extent this is true, his method is
+Neumann-Ulam scheme'. To a certain extent this is true; his method is
 based on a projection method and should always converge. In his work,
@@
 The stochastic nature of the algorithm comes from effectively
-modifying the algorithm such that the value of $i$ is now selected at
-random. This is interesting but arguably does not serve our purposes
-at well. We have an interest in a fully stochastic process like a
-random walk sequence because of its potential benefits in terms of
-resiliency and the ability to use superposition to combine results
-from replicated problems to enhance performance. In Sabelfeld's work,
-the randomness is used to select which deterministic iteration step
-will be performed and does not meet either of these
+modifying the iteration sequence such that the value of $i$ is now
+selected at random. This is interesting but arguably does not serve
+our purposes at well. We have an interest in a fully stochastic
+process like a random walk sequence because of its potential benefits
+in terms of resiliency and the ability to use superposition to combine
+results from replicated problems to enhance performance. In
+Sabelfeld's work, the randomness is used to select which deterministic
+iteration step will be performed and does not meet either of these
 requirements. However, that does not mean there is nothing to learn
@@
 perform when applied to real problems in nuclear engineering. These
-methods aim to improve over conventional methods when looking towards
-the advanced computer architecutures of the near future. Parallelism
-at high levels of concurrency, a potential reduction in memory
-footprint, and a natural element of resiliency deriving from the
-stochastic nature of the algorithm motivated us to research their
-strengths and weaknesses and develop solutions to further their
-applicability.
+methods aim to make improvements over conventional methods when
+looking towards the advanced computer architectures of the near
+future. Parallelism at high levels of concurrency, a potential
+reduction in memory footprint, and a natural element of resiliency
+deriving from the stochastic nature of the algorithm motivated us to
+research their strengths and weaknesses and develop solutions to
+further their applicability.
 
@@
 neutron transport problem, asymmetric systems, and nonlinear problems
-including fluid flow systems. Techniques was developed by this work
+including fluid flow systems. Techniques were developed by this work
 that enabled solutions for all of these problems using MCSA. In many
@@
 particle transport that permitted MCSA solutions to be generated on
-$O(100,000)$ cores on a leadership class computing facility. The new
+$O(100,000)$ cores on a leadership class computing platform. The new
 algorithm was also demonstrated to achieve better strong and weak
@@
 
-With this work, a new foundation for Monte Carlo Synethic Acceleration
+With this work, a new foundation for Monte Carlo Synthetic Acceleration
 methods has been laid with new theory, algorithms, and numerical

### 2013-07-25 doc/dissertation/document/chapters/introduction/introduction.tex
pushing full draft of disseration

@@
 subsystem, similar to the same components of the departure from
-nucleate boiling example above, was performed as part of analysis of
-the Department of Energy's Consortium for Advanced Simulation of Light
+nucleate boiling example, was performed as part of analysis of the
+Department of Energy's Consortium for Advanced Simulation of Light
 Water Reactors (CASL) modeling and simulation hub. CASL used the
@@
 stochastic nature of the solution method provides a natural solution
-to the issue of resiliency, and (3) is amenable to parallelization
-using modern methods developed by the transport community
-\cite{wagner_hybrid_2010}. The development of MCSA as a general
-solution technique and the development of a parallel MCSA method are
-new and unique features of this work, providing a framework with which
-other issues such as resiliency or performance may be addressed in the
-future.
+to some aspects of the issue of resiliency, and (3) is amenable to
+parallelization using modern methods developed by the transport
+community \cite{wagner_hybrid_2010}. The development of MCSA as a
+general solution technique and the development of a parallel MCSA
+method are new and unique features of this work, providing a framework
+with which other issues such as resiliency or performance may be
+addressed in the future.
 
-In addition to advancements in the solution of linear problems,
-nonlinear problems may also benefit from a general and parallel MCSA
-scheme. In the nuclear engineering community, nonlinear problems are
-often addressed by either linearizing the problem or building a
-segregated scheme and using traditionally iterative or direct methods
-to solve the resulting system \cite{pletcher_computational_1997}. In
-the mathematics community, various Newton methods have been popular
+In addition to advancements in the parallel solution of linear
+problems, nonlinear problems may also benefit from a general and
+parallel MCSA scheme. In the nuclear engineering community, nonlinear
+problems are often addressed by either linearizing the problem or
+building a segregated scheme and using traditionally iterative or
+direct methods to solve the resulting system
+\cite{pletcher_computational_1997}. In the mathematics community,
+various Newton methods have been popular
 \cite{kelley_iterative_1995}. Recently, Jacobian-Free Newton-Krylov
@@
 preconditioned systems) the Krylov subspace generated by the GMRES
-solver may become prohibitively large
-\cite{knoll_newton-krylov_1995}.
+solver may become prohibitively large \cite{knoll_newton-krylov_1995}.
 

### 2013-07-25 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
pushing full draft of disseration

@@
-\chapter{Monte Carlo Synthetic Acceleration Methods for the
-  Navier-Stokes \\Equations\ }
+\chapter{Monte Carlo Synthetic Acceleration\\ Methods for the
+  Navier-Stokes\\ Equations\ }
 \label{ch:nonlinear_problem}
@@
 methods, a variant of inexact Newton methods, will be outlined
-including with two common techniques for numerically representing the
-Jacobian most commonly used in today's applications. Following this, a
-new algorithm utilizing Monte Carlo Synthetic Acceleration will be
-presented and implementation details will be discussed. Using the new
-algorithm and a production fluid physics code, a group of challenging
-nonlinear benchmark problems for the Navier-Stokes equations will be
-used to verify the correctness of the method relative to solutions
-obtained using conventional methods in various flow regimes and
+including two techniques for numerically representing the Jacobian
+commonly used in today's applications. Following this, a new algorithm
+utilizing Monte Carlo Synthetic Acceleration will be presented and
+implementation details will be discussed. Using the new algorithm and
+a production fluid physics code, a group of challenging nonlinear
+benchmark problems for the Navier-Stokes equations will be used to
+verify the correctness of the method relative to solutions obtained
+using conventional methods in various flow regimes and
 geometries. Finally, we will compare the performance of the new method
@@
 
-\clearpage
-
 %%---------------------------------------------------------------------------%%
@@
 spectral radius of the Jacobian operators generated at each Newton
-step. This preconditioning was required to obtain good convergence
-properties for the Newton-Krylov solutions although it was not
-necessary for their convergence. In addition, the same nonlinear
+step. This preconditioning was also required to obtain good
+convergence properties for the Newton-Krylov solutions although it was
+not necessary for their convergence. In addition, the same nonlinear
 solver parameters including globalization and forcing term strategies
@@
 solution at each Newton step, a polynomial line search method was used
-as a means of globalization assisting the Newton correction
+as a means of globalization, assisting the Newton correction
 computation as outlined by Pawlowski in
@@
 a polynomial line search method was used as a means of globalization
-assisting the Newton correction computation as outlined by Pawlowski
-in \cite{pawlowski_globalization_2006} while the forcing term was
-again computed using method 2 from Shadid's work
-\cite{shadid_inexact_1997}.
+while the forcing term was again computed using method 2 from Shadid's
+work \cite{shadid_inexact_1997}.
 
@@
 difficulty even though it is preconditioned in an identical
-manner.
-
-For the \sn{1}{5} Rayleigh number case, this means that the GMRES
-solver at each Newton step is converging the linear residual to a
-smaller tolerance than the MCSA solver as the forcing terms adapt to
-the convergence of the method over the course of nonlinear
-iterations. As the Rayleigh number increases, the problem becomes more
-difficult to solve and therefore a more aggressive preconditioner
-should maintain iterative performance of the nonlinear method. This
-was observed to be true for the \sn{1}{6} Rayleigh number case. For
-this calculation, the number of multigrid cycle applications was
-bumped up to 3 from 1 and FANM converged in the same number of
-nonlinear iterations as the Newton-Krylov method.
+manner. This means that the GMRES solver at each Newton step is
+converging the linear residual to a smaller tolerance than the MCSA
+solver as the forcing terms adapt to the convergence of the method
+over the course of nonlinear iterations. For this problem, tighter
+convergence of the linear model at later Newton steps also improved
+convergence of nonlinear model, resulting in the extra FANM iteration.
+
+As the Rayleigh number increases, the problem becomes more difficult
+to solve and therefore a more aggressive preconditioner should
+maintain iterative performance of the nonlinear method. This was
+observed to be true for the \sn{1}{6} Rayleigh number case. For this
+calculation, the number of multigrid cycle applications was bumped up
+to 3 from 1 and FANM converged in the same number of nonlinear
+iterations as the Newton-Krylov method.
 
@@
 
-\clearpage
-
 \subsection{Lid Driven Cavity Results}
@@
 Table~\ref{tab:benchmark_linear_comparison}, the difference is more
-dramatic with FANM converging in fewer the nonlinear residual in fewer
-linear solver iterations for twice as many cases as the Newton-Krylov
-method keeping in mind that for all cases both GMRES and MCSA were
+dramatic with FANM converging the nonlinear residual in fewer linear
+solver iterations for twice as many cases as the Newton-Krylov method
+keeping in mind that for all cases both GMRES and MCSA were
 preconditioned with identical multigrid parameters using the same

### 2013-07-25 doc/dissertation/document/frontmatter/abstract.tex
pushing full draft of disseration

@@
 computing hardware and algorithms that leverage this hardware are
-required to help meet the meet the goals and requirements of
-engineering analysis. Current practices in neutron transport and fluid
-flow leverage methods based on subspace iterations to achieve good
+required to help meet the goals and requirements of engineering
+analysis. Current practices in neutron transport and fluid flow
+leverage methods based on subspace iterations to achieve good
 convergence properties and quality parallel scaling on large
@@
 Second, solutions to the full Navier-Stokes equations were obtained by
-developing the Forward-Automated Newton-MCSA (FANM) scheme for
+developing the Forward-Automated Newton-MCSA (FANM) method for
 nonlinear systems based on Newton's method. Three difficult fluid
@@
 used to drive the research and development of the method. It was found
-that for all benchmarks that FANM produced identical solutions to
-modern Newton-Krylov methods. For each benchmark, performance
-comparisons between FANM and Newton-Krylov were also conducted. For 8
-out of 12 benchmark cases, it was found that FANM had better iterative
+for all benchmarks that FANM produced identical solutions to modern
+Newton-Krylov methods. For each benchmark, performance comparisons
+between FANM and Newton-Krylov were also conducted. For 8 out of 12
+benchmark cases, it was found that FANM had better iterative
 performance than the Newton-Krylov method by converging the nonlinear
@@
 components of this work are required to enable future simulation of
-nonlinear physics problems on the next generation hardware. To enable
-high performance simulations with FANM, we require the MCSA techniques
-developed in the research on the neutron transport problem and in turn
-effectively leveraging MCSA on leadership-class hardware requires the
-new parallel algorithm developed by this work. With the further
-development of these methods based on this work, future nuclear
-reactor simulations may be enabled that have an improved time to
-solution and more efficiently utilize the available computational
+nonlinear physics problems on the next generation of hardware. To
+enable high performance simulations with FANM, we require the MCSA
+techniques developed in the research on the neutron transport problem
+and in turn effectively leveraging MCSA on leadership-class hardware
+requires the new parallel algorithm developed by this work. With the
+further development of these methods based on this work, future
+nuclear reactor simulations may be enabled that have an improved time
+to solution and more efficiently utilize the available computational
 resources.

### 2013-07-29 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
pushing full draft of disseration
adding to the spn performance comparison

@@
-\chapter{Monte Carlo Synthetic Acceleration Methods for the $SP_N$ Equations}
+\chapter{Monte Carlo Synthetic Acceleration\\ Methods for the $SP_N$ Equations}
 \label{ch:spn_equations}
@@
 
-\clearpage
-
 \subsection{Point Jacobi Spectral Analysis Results}
@@
 both ensure and improve convergence for Monte Carlo methods.
-
-\clearpage
 
@@
 representative of physical systems.
-\clearpage
 
@@
 
-\clearpage
-
 \subsection{Preliminary Jacobi Preconditioned Calculations}
@@
 therefore we expect the conjugate gradient-based methods to perform
-well. It also important to note that the iterative performance of each
-solver is not a strong function of the number of energy groups in the
-problem (and therefore the problem size).
+well due to the structure of the projection scheme. It important to
+note that the iterative performance of each solver is not a strong
+function of the number of energy groups in the problem (and therefore
+the problem size). Even more important, although BiCGStab produced the
+best iterative results, MCSA did perform better than GMRES in many
+cases with the Richardson iteration accelerated with Monte Carlo using
+the collision estimator performing best.
 

### 2013-07-29 doc/kaczmarz/python/rep_kaczmarz.py

@@ -1,0 +1,68 @@
+import numpy
+import math
+import random
+
+m = 3
+
+# Kaczmarz projection method
+def kaczmarz(A,x,b):
+    r = b - A*x
+    num_iters = 0
+    while numpy.linalg.norm(r,numpy.inf)/numpy.linalg.norm(b,numpy.inf) > 1.0e-2:
+        i = random.randint(0,m)
+        j = m-i-1
+        alpha = numpy.asscalar( numpy.dot(A[j],A[j].T) )
+        scalarb = numpy.asscalar(b[j])
+        scalarAjdotx = numpy.asscalar( numpy.dot(A[j],x) )
+        x = x + (1/alpha)*(scalarb - scalarAjdotx)*(A.T)[j].T
+        r = b - A*x
+        num_iters += 1
+    print "Kaczmarz iterations to converge:", num_iters
+    return x
+
+# Solve a 3x3 linear system Ax=b with initial state x_i = 0
+A = numpy.matrix([ [1.0, -0.3, -0.4],[-0.2, 1.0, -0.4],[-0.5, -0.4, 1.0] ])
+b = numpy.matrix([ [10.0], [1.0], [1.0] ])
+x_kaczmarz_1 = numpy.matrix([ [0.0], [0.0], [0.0] ])
+x_kaczmarz_2 = numpy.matrix([ [0.0], [0.0], [0.0] ])
+x_kaczmarz_3 = numpy.matrix([ [0.0], [0.0], [0.0] ])
+x_kaczmarz_4 = numpy.matrix([ [0.0], [0.0], [0.0] ])
+x_adjoint = numpy.matrix([ [0.0], [0.0], [0.0] ])
+
+# Spectral Radius of A
+eigenvalues_A, eigenvectors_A = numpy.linalg.eig(A)
+rho_A = max(eigenvalues_A)
+print 
+print "Spectral radius A =", rho_A
+print
+
+# Kaczmarz solution 1
+x_kaczmarz_1 = kaczmarz(A,x_kaczmarz_1,b)
+print
+print "Kaczmarz"
+print x_kaczmarz_1
+
+# Kaczmarz solution 2
+x_kaczmarz_2 = kaczmarz(A,x_kaczmarz_2,b)
+print
+print "Kaczmarz"
+print x_kaczmarz_2
+
+# Kaczmarz solution 3
+x_kaczmarz_3 = kaczmarz(A,x_kaczmarz_3,b)
+print
[…]

### 2013-07-29 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
cleaning up some figures in the spn section

@@
       the region of fast quadratic convergence is reached, MCSA
-      forcing terms were much larger.}}
+      forcing terms were much smaller.}}
   \label{fig:driven_re500_forcing}
@@
 therefore use fewer MCSA iterations at each Newton step to converge
-the linear model. This is indeed the case for the 400 Reynolds number
-case where the linear model is better conditioned and FANM used fewer
-total linear solver iterations to converge the nonlinear
-residual. However, at a Reynolds number of 500 FANM has larger forcing
-terms but requires many more MCSA iterations to converge the last 2
-Newton steps. Therefore, these extra iterations in this case must come
-from ill-conditioning.
+the linear model if the iterative performance is comparable. This is
+indeed the case for the 400 Reynolds number case where the linear
+model is better conditioned and FANM used fewer total linear solver
+iterations to converge the nonlinear residual. However, at a Reynolds
+number of 500 FANM has larger forcing terms but requires many more
+MCSA iterations to converge the last 2 Newton steps. Therefore, these
+extra iterations in this case must come from ill-conditioning.
 

### 2013-07-29 doc/dissertation/presentation/defense.tex
adding thesis defense presentation

\documentclass{beamer}
\usetheme[white]{Wisconsin}
\usepackage{longtable}
\usepackage{listings}
\usepackage{color}
%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
\usepackage{amsthm} 
\usepackage{amsmath} 
\usepackage{tmadd,tmath}
\usepackage[mathcal]{euscript} 
\usepackage{color}
\usepackage{textcomp}
\usepackage{algorithm,algorithmic}
\definecolor{listinggray}{gray}{0.9}
\definecolor{lbcolor}{rgb}{0.9,0.9,0.9}
\lstset{
  backgroundcolor=\color{lbcolor},
  tabsize=4,
  rulecolor=,
  language=c++,
  basicstyle=\scriptsize,
  upquote=true,
  aboveskip={1.5\baselineskip},
  columns=fixed,
  showstringspaces=false,
  extendedchars=true,
  breaklines=true,
  prebreak =
  \raisebox{0ex}[0ex][0ex]{\ensuremath{\hookleftarrow}},
  frame=single,
  showtabs=false,
  showspaces=false,
  showstringspaces=false,
  identifierstyle=\ttfamily,
  keywordstyle=\color[rgb]{0,0,1},
  commentstyle=\color[rgb]{0.133,0.545,0.133},
  stringstyle=\color[rgb]{0.627,0.126,0.941},
}

%% colors
\setbeamercolor{boxheadcolor}{fg=white,bg=UWRed}
\setbeamercolor{boxbodycolor}{fg=black,bg=white}

%%---------------------------------------------------------------------------%%
\author{Stuart R. Slattery\\ University of Wisconsin - Madison}

\date{\today} 
\title{Parallel Monte Carlo Synthetic Acceleration Methods for
  Discrete Transport Problems}
\begin{document}
\maketitle

%%---------------------------------------------------------------------------%%
\begin{frame}{Outline}

  \begin{itemize}
  \item Monte Carlo Synthetic Acceleration Methods
    \bigskip
  \item Application to Neutron Transport
    \bigskip
  \item Application to Fluid Flow
    \bigskip
  \item Parallelization of MCSA
  \end{itemize}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Acknowledgments}

  This work was performed under appointment to the Nuclear Regulatory
  Commission Fellowship program at the University of Wisconsin - Madison
  Engineering Physics Department

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Statement of Work}

The goal of this work is to improve the iterative performance and
parallel scalability of solutions to discrete linear and nonlinear
transport problems by researching and developing a new set of domain
decomposed Monte Carlo Synthetic Acceleration methods.

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Hardware-Based Motivation}

  \begin{itemize}
  \item Modern hardware is moving in two directions (Kogge,2011):
    \begin{itemize}
    \item Lightweight machines
    \item Heterogeneous machines
    \item Both characterized by low power and high concurrency
    \end{itemize}
    \medskip \medskip
  \item Some issues:
    \begin{itemize}
    \item Higher potential for both soft and hard failures (DOE,2012)
    \item Memory restrictions are expected with a continued decrease
      in memory/FLOPS
    \end{itemize}
    \medskip \medskip
  \item Potential resolution from Monte Carlo:
    \begin{itemize}
    \item Soft failures buried within the tally variance
    \item Hard failures are high variance events
    \item Memory savings over conventional methods
    \end{itemize}
  \end{itemize}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Research Outline}
  \begin{itemize}
  \item Development of a linear solver for discrete systems leveraging
    Monte Carlo Synthetic Acceleration
    \medskip
    \begin{itemize}
    \item Application to neutron transport
    \item Research is required to explore general solver development
    \item Performance is of concern
    \end{itemize}
    \bigskip
  \item Development of a nonlinear solver for discrete systems
    leveraging Monte Carlo Synthetic Acceleration
    \begin{itemize}
    \item Application to fluid flow
    \item Potential memory benefits
    \item Convergence of the linear model is of concern
    \end{itemize}
    \bigskip
  \item Parallelization of Monte Carlo Synthetic Acceleration
    \medskip
    \begin{itemize}
    \item Parallel strategies taken from modern reactor physics
      methods
    \item Research is required to explore varying parallel strategies
    \item Scalability is of concern
    \end{itemize}
  \end{itemize}
\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Outline}

  \begin{itemize}
  \item \textbf{Monte Carlo Synthetic Acceleration Methods}
    \bigskip
  \item Application to Neutron Transport
    \bigskip
  \item Application to Fluid Flow
    \bigskip
  \item Parallelization of MCSA
  \end{itemize}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Monte Carlo Methods for Discrete Linear Systems}

  \begin{itemize}
  \item First proposed by J. Von Neumann and S.M. Ulam in the 1940's
    \medskip \medskip
  \item Earliest published reference in 1950
    \medskip \medskip
  \item General lack of published work
    \medskip \medskip
  \item Modern work by Evans and others has yielded new
    applications\let\thefootnote\relax\footnote{\tiny{Thomas Evans and
        Scott Mosher, "A Monte Carlo Synthetic Acceleration method for
        the non-linear, time-dependent diffusion equation", American
        Nuclear Society - International Conference on Mathematics,
        Computational Methods and Reactor Physics, 2009.}}
  \end{itemize}
\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Monte Carlo Linear Solver Preliminaries}

  \begin{itemize}
  \item Split the linear operator
  \end{itemize}

  \[
  \ve{A}\ve{x} = \ve{b} \ \ \ \rightarrow \ \ \ \ve{x} = \ve{H} \ve{x}
  + \ve{b}
  \]

  \[
  \ve{H} = \ve{I} - \ve{A}
  \]

  \medskip
  \begin{itemize}
  \item Generate the \textit{Neumann series}
  \end{itemize}

  \[
  \ve{A}^{-1} = (\ve{I}-\ve{H})^{-1} = \sum_{k=0}^{\infty} \ve{H}^k
  \]

  \medskip
  \begin{itemize}
  \item Require $\rho(\ve{H}) < 1$ for convergence
  \end{itemize}

  \[
  \ve{A}^{-1}\ve{b} = \sum_{k=0}^{\infty} \ve{H}^k\ve{b} = \ve{x}
  \]

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Monte Carlo Linear Solver Preliminaries}

  \begin{itemize}
  \item Expand the Neumann series
  \end{itemize}

  \[
  x_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}
  \]

  \begin{itemize}
  \item Define a sequence of state transitions
  \end{itemize}

  \[
  \nu = i \rightarrow i_1 \rightarrow \cdots \rightarrow i_{k-1}
  \rightarrow i_{k}
  \]

  \begin{itemize}
  \item Use the adjoint Neumann-Ulam
    decomposition\let\thefootnote\relax\footnote{The Hadamard product
      $\ve{A} = \ve{B} \circ \ve{C}$ is defined element-wise as
      $a_{ij} = b_{ij} c_{ij}$.}
  \end{itemize}

  \[
  \ve{H}^{T} = \ve{P} \circ \ve{W}
  \]

  \[
  p_{ij} = \frac{|h_{ji}|}{\sum_j |h_{ji}|},\ w_{ij} =
  \frac{h_{ji}}{p_{ij}}
  \]

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Evolution of a Solution}

  \begin{figure}[htpb!]
    \begin{center}
      \scalebox{1.0}{ \input{heat_eq_setup.pdftex_t} }
    \end{center}
    \caption{\textbf{Poisson Problem.}
      \textit{Distributed source of 1.0 in the domain.}}
  \end{figure}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Evolution of a Solution}

  \begin{figure}[h!]
    \begin{center}
      \includegraphics[width=4in]{adjoint_1.png}
    \end{center}
    \caption{\textbf{Adjoint solution to Poisson Equation.}
      \textit{\sn{1}{0} total histories, 0.286 seconds CPU time.} }
  \end{figure}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Evolution of a Solution}

  \begin{figure}[h!]
    \begin{center}
      \includegraphics[width=4in]{adjoint_10.png}
    \end{center}
    \caption{\textbf{Adjoint solution to Poisson Equation.}
      \textit{\sn{1}{1} total histories, 0.278 seconds CPU time.} }
  \end{figure}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Evolution of a Solution}

  \begin{figure}[h!]
    \begin{center}
      \includegraphics[width=4in]{adjoint_100.png}
    \end{center}
    \caption{\textbf{Adjoint solution to Poisson Equation.}
      \textit{\sn{1}{2} total histories, 0.275 seconds CPU time.} }
  \end{figure}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Evolution of a Solution}

  \begin{figure}[h!]
    \begin{center}
      \includegraphics[width=4in]{adjoint_1000.png}
    \end{center}
    \caption{\textbf{Adjoint solution to Poisson Equation.}
      \textit{\sn{1}{3} total histories, 0.291 seconds CPU time.} }
  \end{figure}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Evolution of a Solution}

  \begin{figure}[h!]
    \begin{center}
      \includegraphics[width=4in]{adjoint_10000.png}
    \end{center}
    \caption{\textbf{Adjoint solution to Poisson Equation.}
      \textit{\sn{1}{4} total histories, 0.428 seconds CPU time.} }
  \end{figure}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Evolution of a Solution}

  \begin{figure}[h!]
    \begin{center}
      \includegraphics[width=4in]{adjoint_100000.png}
    \end{center}
    \caption{\textbf{Adjoint solution to Poisson Equation.}
      \textit{\sn{1}{5} total histories, 1.76 seconds CPU time.} }
  \end{figure}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Evolution of a Solution}

  \begin{figure}[h!]
    \begin{center}
      \includegraphics[width=4in]{adjoint_1000000.png}
    \end{center}
    \caption{\textbf{Adjoint solution to Poisson Equation.}
      \textit{\sn{1}{6} total histories, 15.1 seconds CPU time.} }
  \end{figure}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Evolution of a Solution}

  \begin{figure}[h!]
    \begin{center}
      \includegraphics[width=4in]{adjoint_10000000.png}
    \end{center}
    \caption{\textbf{Adjoint solution to Poisson Equation.}
      \textit{\sn{1}{7} total histories, 149 seconds CPU time.} }
  \end{figure}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Monte Carlo Synthetic-Acceleration}

  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
    {MCSA Iteration}

    \[
    \ve{r}^{k} = \ve{b} - \ve{A}\ve{x}^{k}
    \]
    \[
    \ve{x}^{k+1/2} = \ve{x}^k + \ve{r}^k
    \]
    \[
    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}
    \]
    \[
    \hat{\ve{A}}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}
    \]
    \[
    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}
    \]

  \end{beamerboxesrounded}

  \medskip \medskip
  \begin{itemize}
  \item Neumann-Ulam methods bound by the Central Limit Theorem
  \item Build on Halton's 1962 Sequential Monte Carlo method
  \item Neumann-Ulam Monte Carlo solver computes the correction
  \item Decouples MC error from solution error, exponential convergence
  \end{itemize}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Monte Carlo Linear Solvers Library (MCLS)}

  \begin{itemize}
  \item Preliminary development during January of 2013
    \bigskip
  \item General asynchronous MSOD MCSA implementation
    \begin{itemize}
    \item Forward and adjoint Monte Carlo with method of expected
      values
    \item Parallel row matrix/vector interface
    \item General fixed point iteration strategy
    \item Explicit algebraic preconditioner suite
    \end{itemize}
    \medskip
  \item Implemented in C++
    \bigskip
  \item Heavy use of the Trilinos scientific computing libraries
    \bigskip
  \item Open-source BSD 3-clause license
    \bigskip
  \item https://github.com/sslattery/MCLS
  \end{itemize}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Outline}

  \begin{itemize}
  \item Monte Carlo Synthetic Acceleration Methods
    \bigskip
  \item \textbf{Application to Neutron Transport}
    \bigskip
  \item Application to Fluid Flow
    \bigskip
  \item Parallelization of MCSA
  \end{itemize}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{$SP_N$ Equations}

  \begin{multline}
    \hat{\Omega} \cdot \vec{\nabla} \psi(\vec{r},\hat{\Omega},E) +
    \sigma(\vec{r},E) \psi(\vec{r},\hat{\Omega},E) = \\ \iint
    \sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}' \cdot \hat{\Omega})
    \psi(\vec{r},\hat{\Omega}',E') d\Omega' dE' +
    q(\vec{r},\hat{\Omega},E)\:,
    \label{eq:general_transport}
  \end{multline}

  \begin{multline}
    -\nabla \cdot \Bigg[\frac{n}{2n+1}\frac{1}{\Sigma_{n-1}} \nabla
      \Big(\frac{n-1}{2n-1} \phi_{n-2} + \frac{n}{2n-1}\phi_n \Big) \\+
      \frac{n+1}{2n+1}\frac{1}{\Sigma_{n+1}} \nabla
      \Big(\frac{n+1}{2n+3}\phi_n + \frac{n+2}{2n+3}\phi_{n+2}\Big)
      \Bigg] \\+ \Sigma_n \phi_n = q \delta_{n0}\ \ \ \ \ \ \ \ \ n =
    0,2,4,\cdots,N\:,
    \label{eq:spn_equations}
  \end{multline}

  \begin{equation}
    -\nabla \cdot \mathbb{D}_n \nabla \mathbb{U}_n + \sum_{m=1}^4
    \mathbb{A}_{nm} \mathbb{U}_m = \frac{1}{k} \sum_{m=1}^4
    \mathbb{F}_{nm} \mathbb{U}_n\ \ \ \ \ \ \ n = 1,2,3,4\:.
    \label{eq:spn_fission_matrix}
  \end{equation}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}[fragile]{k-eigenvalue Solutions with MCSA}

  \begin{algorithm}[H]
    \caption{Power Iteration MCSA Scheme}
    \label{alg:power_iteration}
    \begin{algorithmic}
      \STATE $k_0 =$ initial guess
      \STATE $\mathbf{\Phi}_0 =$ initial guess
      \STATE $n = 0$
      \WHILE{$|\frac{k^n - k^{n-1}}{k^n}| < \epsilon$}
      \STATE $\mathbf{M} \mathbf{\Phi}^{n+1} = \frac{1}{k^n} \mathbf{F} \mathbf{\Phi}^n$
      \COMMENT{Solve for the new flux state with MCSA}
      \STATE $k^{n+1} = k^n \frac{\int \mathbf{F} \mathbf{\Phi}^{n+1} d\mathbf{r}}{\int
        \mathbf{F} \mathbf{\Phi}^n d\mathbf{r}}$
      \STATE $n = n+1$
      \ENDWHILE
    \end{algorithmic}
  \end{algorithm}

  \begin{itemize}
  \item Swap in MCSA as the solver at each eigenvalue operation
    \medskip
  \item Use Thyra/Stratimikos interfaces for plug and play - Thanks
    Ross!
    \medskip
  \item Current Exnihilo implementation gives the full operator
  \end{itemize}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Fuel Assembly Criticality Calculations}

  \begin{columns}

    \begin{column}{0.5\textwidth}

      \begin{figure}[htpb!]
        \begin{center}
          \includegraphics[width=1.5in]{problem3_radial_mat.png}
        \end{center}
      \end{figure}

      \begin{figure}[htbp!]
        \begin{center}
          \includegraphics[width=2in]{problem3_axial_mat.png}
        \end{center}
      \end{figure}

      \begin{figure}[htpb!]
        \begin{center}
          \includegraphics[width=2in]{problem3_end.png}
        \end{center}
      \end{figure}

    \end{column}

    \begin{column}{0.5\textwidth}
      {\small
        \begin{itemize}
        \item CASL Problem 3: $17 \times 17$ quarter symmetry HZP LWR
          fuel assembly
        \item $SP_N$ criticality calculation
        \item MCLS leveraged by the Exnihilo code base (ORNL)
        \item Comparison to Trilinos Aztec Krylov solvers with ILUT
        \end{itemize}
      }
      {\tiny
        \begin{table}[htpb!]
          \begin{center}
            \begin{tabular}{ll}\hline\hline
              \multicolumn{1}{l}{\textbf{Parameter}} & 
              \multicolumn{1}{l}{\textbf{Value}} \\
              Power Level & 0 MW \\
              Inlet Temperature & 326.85C \\
              Fuel Temperature & 600C \\
              Boron Concentration & 1300 ppm \\
              Moderator Density & 0.743 g/cc \\
              Helium Density & \sn{1.79}{-4} g/cc \\
              Zirconium Density & 6.56 g/cc \\
              Stainless Steel Density & 8.0 g/cc \\
              Inconel Density & 8.19 g/cc \\
              UO2 Density & 10.257 g/cc \\
              Fuel Pin Radius (w/o clad) & 0.4096 cm \\
              %%
              \hline\hline
            \end{tabular}
          \end{center}
        \end{table}
      }
    \end{column}

  \end{columns}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Jacobi Preconditioned Calculations}

{\tiny  \begin{table}[h!]
    \begin{center}
      \begin{tabular}{cccccc}\hline\hline
        \multicolumn{1}{c}{}& 
        \multicolumn{1}{c}{}& 
        \multicolumn{1}{c}{}& 
        \multicolumn{1}{c}{$SP_N$ Order}& 
        \multicolumn{1}{c}{}& 
        \multicolumn{1}{c}{} \\
        &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
        & \textbf{0} & 0.0647 & 0.1275 & 0.1449 & 0.1514 \\
        & \textbf{1} & 0.0686 & 0.1338 & 0.1484 & 0.1547 \\
        $P_N$ Order & \textbf{3} & 0.0687 & 0.1399 & 0.1582 & 0.1625 \\
        & \textbf{5} & 0.0692 & 0.1399 & 0.1582 & 0.1657 \\
        & \textbf{7} & 0.0678 & 0.1393 & 0.1624 & 0.166 \\
        %%
        \hline\hline
      \end{tabular}
    \end{center}
    \caption{\textbf{Spectral radius results for the block Jacobi
        preconditioned iteration matrix with 10 energy groups and full
        downscatter for sample problem.}}
    \label{tab:group10dsbj}
  \end{table}
}

  \begin{columns}

    \begin{column}{0.5\textwidth}

      \begin{figure}[t!]
        \begin{center}
          \includegraphics[width=2.0in]{block_jacobi_res.pdf}
        \end{center}
      \end{figure}

    \end{column}

    \begin{column}{0.5\textwidth}
      \begin{itemize}
      \item Light water moderator creates a lot of scattering and
        $\mathbf{\rho}(\mathbf{H}) \approx 1$
        \medskip
      \item Intractable number of histories required for MCSA
        convergence 
      \end{itemize}
    \end{column}

  \end{columns}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{MCSA Breakdown}

  \begin{columns}

    \begin{column}{0.5\textwidth}

      \begin{figure}[h!]
        \begin{center}
          \includegraphics[width=1.9in]{breakdown_iterations.pdf}
        \end{center}
      \end{figure}

      \begin{figure}[h!]
        \begin{center}
          \includegraphics[width=1.9in]{breakdown_histories.pdf}
        \end{center}
      \end{figure}

    \end{column}

    \begin{column}{0.5\textwidth}

    \begin{itemize}
    \item As $\mathbf{\rho}(\mathbf{H}) \rightarrow 1$ terrible things
      happen...
      \medskip
    \item A more robust set of preconditioners is required for the
      $SP_N$ equations
    \end{itemize}

      \begin{figure}[h!]
        \begin{center}
          \includegraphics[width=1.9in]{breakdown_time.pdf}
        \end{center}
      \end{figure}

    \end{column}

  \end{columns}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Explicit Preconditioning}

  \[
  \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\ve{M}_R\ve{x} = \ve{M}_L^{-1}\ve{b}
  \ \ \ \rightarrow \ \ \ 
  \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\ve{u} = \ve{M}_L^{-1}\ve{b}
  \]

  \[
  \ve{x} = \ve{M}_R^{-1}\ve{u}
  \]

  \begin{beamerboxesrounded}[upper=boxheadcolor,lower=boxbodycolor,shadow=true]
    {Left/Right Preconditioned MCSA Iteration}

    \[
    \ve{r}^{k} = \ve{M}_L^{-1}(\ve{b}-\ve{A}\ve{M}_R^{-1}\ve{u}^{k})
    \]
    \[
    \ve{u}^{k+1/2} = \ve{u}^k + \ve{r}^k
    \]
    \[
    \ve{r}^{k+1/2} = \ve{M}_L^{-1}(\ve{b}-\ve{A}\ve{M}_R^{-1}\ve{u}^{k+1/2})
    \]
    \[
    \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\delta\ve{u}^{k+1/2} = \ve{r}^{k+1/2}
    \]
    \[
    \ve{u}^{k+1} = \ve{u}^{k+1/2} + \delta \ve{u}^{k+1/2}
    \]

  \end{beamerboxesrounded}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Fuel Assembly Results}

  \begin{columns}

    \begin{column}{0.5\textwidth}

      \begin{figure}[htpb!]
        \begin{center}
          \includegraphics[width=1.9in]{solver_iters.pdf}
        \end{center}
      \end{figure}

      \begin{figure}[htpb!]
        \begin{center}
          \includegraphics[width=1.9in]{solver_time.pdf}
        \end{center}
      \end{figure}

    \end{column}

    \begin{column}{0.5\textwidth}

      \begin{itemize}
        \item MCLS generates the same k-eigenvalue and neutron flux in
          all groups and spatial locations
          \bigskip
        \item MCSA converged in fewer iterations than GMRES, more
          iterations than BiCGStab
          \bigskip
        \item Explicit preconditioning strategy destroys sparsity and
          elevates CPU times (Ifpack ILUT)
          \bigskip
        \item Spectral radius and memory limitations combined to
          prevent solutions at finer discretizations
      \end{itemize}

    \end{column}

  \end{columns}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Outline}

  \begin{itemize}
  \item Monte Carlo Synthetic Acceleration Methods
    \bigskip
  \item Application to Neutron Transport
    \bigskip
  \item \textbf{Application to Fluid Flow}
    \bigskip
  \item Parallelization of MCSA
  \end{itemize}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Monte Carlo Synthetic Acceleration for Nonlinear Problems}

  \begin{itemize}
  \item Many physics problems of interest are nonlinear
    \bigskip
  \item Signficant research on Newton methods since the 1980's
    \bigskip
  \item Newton methods often leverage Krylov solvers
    \begin{itemize}
    \item Robust implementations
    \item No operator required
    \end{itemize}
    \bigskip
  \item Monte Carlo methods need the full operator
    \bigskip
  \item Automatic construction of the linear model is available
    \begin{itemize}
    \item Operator overloading for nonlinear residual differentiation
    \item Ideal for Monte Carlo
    \item Potential scaling improvements
    \item Resiliency benefits
    \end{itemize}
  \end{itemize}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Newton's Method}

  \begin{itemize}
  \item Seek solutions of the general nonlinear problem
  \end{itemize}

  \[
  \ve{F}(\ve{u}) = \ve{0}
  \]
  \[
  \ve{u} \in \mathbb{R}^n,\ \ve{F}:\mathbb{R}^N \rightarrow
  \mathbb{R}^N
  \]

  \begin{itemize}
  \item Interpret the exact solution $\ve{u}$ to be the roots of
    $\ve{F}(\ve{u})$
  \end{itemize}

  \[
  \ve{F}(\ve{u}^{k+1}) = \ve{F}(\ve{u}^{k}) +
  \ve{F}'(\ve{u}^{k})(\ve{u}^{k+1}-\ve{u}^{k}) +
  \frac{\ve{F}''(\ve{u}^{k})}{2}(\ve{u}^{k+1}-\ve{u}^{k})^2 + \cdots
  \]

  \begin{itemize}
  \item Form Newton's method
  \end{itemize}
  \[
  \ve{J}(\ve{u}) \delta \ve{u}^k = -\ve{F}(\ve{u}^{k})
  \]
  \[
  \ve{u}^{k+1} = \ve{u}^k + \delta \ve{u}^k
  \]

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}[fragile]{The FANM Method}

  Forward-Automated Newton-MCSA

  \begin{algorithm}[H]
    \begin{algorithmic}[1]
      \STATE $k := 0$ 
      \WHILE{$||\ve{F}(\ve{u}^{k})|| > \epsilon
        ||\ve{F}(\ve{u}^{0})||$} 
      \STATE $\ve{J}(\ve{u}^{k}) \leftarrow AD(\ve{F}(\ve{u}^k))$ 
      \COMMENT{Automatic differentiation} 
      \STATE $\ve{J}(\ve{u}^k) \delta \ve{u}^k = -\ve{F}(\ve{u}^{k})$
      \COMMENT{Solve for the Newton correction with MCSA} 
      \STATE $\ve{u}^{k+1} \leftarrow \ve{u}^k + \delta \ve{u}^k$ 
      \STATE $k \leftarrow k+1$ 
      \ENDWHILE
    \end{algorithmic}
    \caption{FANM}
  \end{algorithm}

  \begin{itemize}
  \item Robustness of Newton's method (inexact)
  \item Accuracy and convenience of FAD
  \item Potential parallelism, and resiliency benefits of MCSA
  \item Requires only nonlinear function evaluations
  \item Can utilize globalization and forcing term selection methods
  \end{itemize}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{FANM Numerical Experiments}

  \begin{itemize}
  \item Sequence of Navier-Stokes benchmarks for the nonlinear methods
    \begin{itemize}
    \item Thermal convection cavity problem (De Vahl Davis, 1983)
    \item Lid driven cavity problem (Ghia et al., 1982)
    \item Backward-Facing step problem (Gartling, 1990)
    \end{itemize}
    \medskip
  \item Tuning benchmark parameters varies the strength of
    nonlinearities
    \medskip
  \item Newton-Krylov method leveraging Aztec GMRES used for
    comparisons
    \medskip
  \item All problems preconditioned with algebraic multigrid (ML) and
    leveraged some kind of globalization (e.g. backtracking)
  \end{itemize}

  \medskip  \medskip
  \begin{columns}
    \begin{column}{0.5\textwidth}
      \[
      \rho \ve{u} \cdot \nabla \ve{u} - \nabla \cdot \ve{T} - \rho
      \ve{g} = \ve{0}
      \]
      \[
      \nabla \cdot \ve{u} = 0
      \]
      \[
      \rho C_p \ve{u} \cdot \nabla T + \nabla \cdot \ve{q} = 0
      \]
    \end{column}

    \begin{column}{0.5\textwidth}
      \[
      \ve{T} = -P \ve{I} + \mu[\nabla \ve{u} + \nabla \ve{u}^T]
      \]
      \[
      \ve{q} = - k \nabla T
      \]
    \end{column}
  \end{columns}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{FANM Numerical Experiments}

  \begin{columns}
    \begin{column}{0.5\textwidth}
      \begin{figure}[htpb!]
        \begin{center}
          \includegraphics[width=2.5in]{drekar1.png}
        \end{center}
      \end{figure}
    \end{column}

    \begin{column}{0.5\textwidth}
      \begin{figure}[htpb!]
        \begin{center}
          \includegraphics[width=2.5in]{drekar2.png}
        \end{center}
      \end{figure}
    \end{column}
  \end{columns}

  \bigskip

  \begin{itemize}
  \item Drekar is a production physics suite that leverages Newton
    methods with FAD
  \item MCLS incorporated into the Drekar nonlinear solver scheme
    using Thyra/Stratimikos to implement FANM
  \item Thanks Roger!
  \end{itemize}

  {\small \sl Images source: www.casl.gov}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{FANM Nonlinear Iteration Results}

  \begin{table}[h!]
    \begin{center}
      \begin{tabular}{lcc}\hline\hline
        \multicolumn{1}{l}{Benchmark}& 
        \multicolumn{1}{c}{Newton-Krylov}&
        \multicolumn{1}{c}{FANM}\\
        \hline
        Convection, Ra=\sn{1}{3} & 5 & 5 \\
        Convection, Ra=\sn{1}{4} & 7 & 7 \\
        Convection, Ra=\sn{1}{5} & 9 & 10 \\
        Convection, Ra=\sn{1}{6} & 11 & 11 \\
        \hline
        Lid Driven, Re=100 & 6 & 6 \\
        Lid Driven, Re=300 & 9 & 9 \\
        Lid Driven, Re=500 & 11 & 11 \\
        Lid Driven, Re=700 & 14 & 10 \\
        \hline
        Backward Step, Re=200 & 10 & 9 \\
        Backward Step, Re=300 & 15 & 14 \\
        Backward Step, Re=400 & 10 & 10 \\
        Backward Step, Re=500 & 19 & 20 \\
        %%
        \hline\hline
      \end{tabular}`
    \end{center}
    \caption{\textbf{Navier-Stokes benchmark comparison for nonlinear
        iterations.} \textit{Over all benchmarks, FANM performed better
        in terms of nonlinear iterations for 1 more case than the
        Newton-Krylov method.}}
    \label{tab:benchmark_nonlinear_comparison}

  \end{table}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{FANM Linear Solver Iteration Results}

  \begin{table}[h!]
    \begin{center}
      \begin{tabular}{lcc}\hline\hline
        \multicolumn{1}{l}{Benchmark}& 
        \multicolumn{1}{c}{Newton-Krylov}&
        \multicolumn{1}{c}{FANM}\\
        \hline
        Convection, Ra=\sn{1}{3} & 32 & 18 \\
        Convection, Ra=\sn{1}{4} & 23 & 17 \\
        Convection, Ra=\sn{1}{5} & 25 & 20 \\
        Convection, Ra=\sn{1}{6} & 39 & 25 \\
        \hline
        Lid Driven, Re=100 & 27 & 42 \\
        Lid Driven, Re=300 & 35 & 52 \\
        Lid Driven, Re=500 & 41 & 56 \\
        Lid Driven, Re=700 & 21 & 14 \\
        \hline
        Backward Step, Re=200 & 24 & 13 \\
        Backward Step, Re=300 & 23 & 17 \\
        Backward Step, Re=400 & 18 & 12 \\
        Backward Step, Re=500 & 30 & 52 \\
        %%
        \hline\hline
      \end{tabular}
    \end{center}
    \caption{\textbf{Navier-Stokes benchmark comparison for total linear
        solver iterations.} \textit{Over all benchmarks, FANM performed
        better in terms of linear solver iterations for twice as many
        cases as the Newton-Krylov method.}}
    \label{tab:benchmark_linear_comparison}
  \end{table}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{FANM CPU Time Results}

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{lc}\hline\hline
      \multicolumn{1}{l}{Benchmark}& 
      \multicolumn{1}{c}{Newton-Krylov Speedup}\\
      \hline
      Convection, Ra=\sn{1}{3} & 338 \\
      Convection, Ra=\sn{1}{4} & 336 \\
      Convection, Ra=\sn{1}{5} & 346 \\
      Convection, Ra=\sn{1}{6} & 465 \\
      \hline
      Lid Driven, Re=100 & 299 \\
      Lid Driven, Re=300 & 322 \\
      Lid Driven, Re=500 & 288 \\
      Lid Driven, Re=700 & 488 \\
      \hline
      Backward Step, Re=200 & 400 \\
      Backward Step, Re=300 & 593 \\
      Backward Step, Re=400 & 825 \\
      Backward Step, Re=500 & 1057 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Newton-Krylov speedup over FANM.}  \textit{For all
      benchmarks the explicit MCSA preconditioning strategy caused
      significantly larger CPU times for FANM when compared to the
      Newton-Krylov solutions.}}
  \label{tab:benchmark_time_comparison}
\end{table}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Outline}

  \begin{itemize}
  \item Monte Carlo Synthetic Acceleration Methods
    \bigskip
  \item Application to Neutron Transport
    \bigskip
  \item Application to Fluid Flow
    \bigskip
  \item \textbf{Parallelization of MCSA}
  \end{itemize}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Parallelization of Monte Carlo Methods}

  \begin{itemize}
  \item No literature observed for parallel Neumann-Ulam solvers
    beyond history-level parallelism
    \medskip \medskip
  \item Numerous references for modern parallel Monte Carlo methods in
    reactor physics
    \medskip \medskip
  \item Build a strategy for applying modern methods to the
    Neumann-Ulam method
    \medskip \medskip
  \item MCSA iteration-level parallelism comes from parallel
    matrix/vector operations
  \end{itemize}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Domain Decomposed Monte Carlo}

  \begin{columns}
    \begin{column}{0.5\textwidth}
      \begin{itemize}
      \item Each parallel process owns a piece of the domain (linear
        system)
        \bigskip
      \item Random walks must be transported between adjacent domains
        through parallel communication
        \bigskip
      \item Domain decomposition determined by the input system
      \end{itemize}
    \end{column}

    \begin{column}{0.5\textwidth}
      \begin{figure}[htpb!]
        \begin{center}
          \scalebox{0.75}{ \input{ddnu_example.pdftex_t} }
        \end{center}
        \caption{\small \sl Domain decomposition example illustrating
          how domain-to-domain transport creates communication costs.}
      \end{figure}
    \end{column}
  \end{columns}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Asynchronous Monte Carlo Transport Kernel}

  \begin{columns}

    \begin{column}{0.5\textwidth}

      \begin{itemize}
      \item Developed by Brunner and Brantley in 2009
      \item Asynchronous nearest neighbor communication of histories
      \item Binary asynchronous communication tree for completing
        transport
      \end{itemize}

      \begin{figure}[htpb!]
        \begin{center}
          \scalebox{0.45}{ \input{domain_to_domain.pdftex_t} }
        \end{center}
      \end{figure}

    \end{column}

    \begin{column}{0.5\textwidth}

      \begin{itemize}
      \item Extensible to problems where histories may be created
        (i.e. variance reduction)
      \end{itemize}

      \begin{figure}[htpb!]
        \begin{center}
          \scalebox{0.45}{ \input{binary_comm_tree.pdftex_t} }
        \end{center}
      \end{figure}

    \end{column}

  \end{columns}

  \let\thefootnote\relax\footnote{\tiny{Thomas A. Brunner and Patrick
      S. Brantley, "An efficient, robust, domain-decomposition
      algorithm for particle Monte Carlo", Journal of Computational
      Physics, vol. 228, pp.3882-3890, 2009.}}
\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Multiple-Set Overlapping-Domain Decomposition}

  \begin{columns}

    \begin{column}{0.5\textwidth}
      \begin{figure}[htpb!]
        \begin{center}
          \scalebox{0.75}{ \input{msod_example.pdftex_t} }
        \end{center}
        \caption{\small \sl Overlapping domain example illustrating
          how domain overlap can reduce communication costs.}
      \end{figure}
    \end{column}

    \begin{column}{0.5\textwidth}
      \begin{itemize}
      \item Developed by Wagner and colleagues in 2010
        \medskip
      \item Each set contains the full domain
        \medskip
      \item Multiple sets replicate the domain
        \medskip
      \item Domains overlap within a set
        \medskip
      \item Redundancy for resiliency (and useful work)
      \end{itemize}
    \end{column}

  \end{columns}

  \let\thefootnote\relax\footnote{\tiny{Wagner et. al., "Hybrid and
      parallel domain-decomposition methods development to enable
      Monte Carlo for reactor analysis", Joint International
      Conference on Supercomputing in Nuclear Applications and Monte
      Carl (SNA+MC 2010), 2010.}}
\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Multiple-Set Overlapping-Domain Decomposition}

  \begin{columns}

    \begin{column}{0.5\textwidth}

      \begin{figure}[htpb!]
        \begin{center}
          \scalebox{0.25}{ \input{msod_construction.pdftex_t} }
        \end{center}
        \caption{\small \sl MSOD construction.}
      \end{figure}

    \end{column}

    \begin{column}{0.5\textwidth}

      \begin{figure}[htpb!]
        \begin{center}
          \scalebox{0.25}{ \input{msod_tally.pdftex_t} }
        \end{center}
        \caption{\small \sl MSOD tally reduction.}
      \end{figure}

    \end{column}

  \end{columns}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Parallel Scaling}

  \begin{itemize}
  \item Simple 2D neutron diffusion problem for control - spectral
    radius is maintained as global problem size grows
    \medskip
  \item Comparison to Trilinos Belos Krylov solvers with Jacobi
    preconditioning - conjugate gradient and GMRES
    \medskip
  \item Strong scaling - Global size fixed at 1.6E7 DOFs
    \medskip
  \item Weak scaling - Local size fixed at 4.0E4 DOFs
    \medskip
  \item Calculations performed on the Titan Cray XK7 machine at ORNL
    (MPI only)
    \medskip
  \item Limited MCLS arithmetic optimization artificially inflates
    efficiencies
  \end{itemize}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Strong Scaling Results}

  \begin{figure}[htpb!]
    \begin{center}
      \includegraphics[width=4in]{titan_pure_strong.pdf}
    \end{center}
    \caption{\textbf{Pure domain decomposition.} \textit{Super-linear
        speed-up from memory thrashing in base case. MCLS is an order
        of magnitude slower arithmetically.}}
  \end{figure}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Weak Scaling Results}

  \begin{columns}
    \begin{column}{0.55\textwidth}

      \begin{figure}[htpb!]
        \begin{center}
          \includegraphics[width=2.5in]{titan_weak_absolute.pdf}
        \end{center}
      \end{figure}

      \begin{figure}[htpb!]
        \begin{center}
          \includegraphics[width=2.5in]{titan_weak_implementation.pdf}
        \end{center}
      \end{figure}

    \end{column}

    \begin{column}{0.45\textwidth}

      \begin{itemize}
      \item CG demonstrates poor scaling due to the cheaper iteration
        sequence
        \bigskip
      \item Implementation efficiency is effectively the parallel
        efficiency of a single iteration
        \bigskip
      \item Improved implementation efficiency means improved
        iterative performance will potentially give a better time to
        solution
      \end{itemize}

    \end{column}
  \end{columns}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Strong Scaling Results with Multiple Sets}

  \textbf{Splitting:} same global number of histories as the single
  set

  \bigskip

  \textbf{Replicating:} set-multiple of single set problem global
  histories

  \begin{columns}
    \begin{column}{0.5\textwidth}

      \begin{figure}[htpb!]
        \begin{center}
          \includegraphics[width=2.35in]{titan_strong_ms_eff.pdf}
        \end{center}
        \caption{\small{Absolute parallel efficiency relative to
            16-core 1-set base case.}}
      \end{figure}

    \end{column}

    \begin{column}{0.5\textwidth}

      \begin{figure}[htpb!]
        \begin{center}
          \includegraphics[width=2.35in]{titan_strong_ms_time.pdf}
        \end{center}
        \caption{\small{Wall time in seconds to solution for each
            case.}}
      \end{figure}

    \end{column}
  \end{columns}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Weak Scaling Results with Multiple Sets}

  \begin{itemize}
  \item Need to consider adding sets is a strong scaling exercise
  \item Modify the weak scaling efficiency computation to account for
    these extra resources
  \item Superposition of Monte Carlo results enhances time to solution!
  \end{itemize}

  \begin{columns}
    \begin{column}{0.5\textwidth}

      \begin{figure}[htpb!]
        \begin{center}
          \includegraphics[width=2.4in]{titan_weak_ms_eff.pdf}
        \end{center}
        \caption{Absolute parallel efficiency relative to 16-core
          1-set base case.}
      \end{figure}

    \end{column}

    \begin{column}{0.5\textwidth}

      \begin{figure}[htpb!]
        \begin{center}
          \includegraphics[width=2.4in]{titan_weak_ms_time.pdf}
        \end{center}
        \caption{Wall time in seconds to solution for each case.}
      \end{figure}

    \end{column}
  \end{columns}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Strong Scaling Results with Overlap}

  \begin{itemize}
  \item Overlap values selected based on average 'diffusion length'
    of a history in the system of 2.6 discrete states
    \bigskip
  \item Overlap eliminates communication in the Monte Carlo sequence
    but simply differs it to an overlapping tally vector reduction
  \end{itemize}

  \begin{columns}
    \begin{column}{0.5\textwidth}

      \begin{figure}[htpb!]
        \begin{center}
          \includegraphics[width=2.4in]{titan_strong_overlap_diff.pdf}
        \end{center}
        \caption{Strong scaling efficiency difference compared to the
          0 overlap case.}
      \end{figure}

    \end{column}

    \begin{column}{0.5\textwidth}

      \begin{figure}[htpb!]
        \begin{center}
          \includegraphics[width=2.4in]{titan_weak_overlap_diff.pdf}
        \end{center}
        \caption{Weak scaling efficiency difference compared to the 0
          overlap case.}
      \end{figure}

    \end{column}
  \end{columns}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{MCSA as a Stochastic Additive Schwarz Method}

  \begin{itemize}
  \item No domain-to-domain communication in Monte Carlo sequence
    \smallskip
  \item Fixed point iteration acts as a smoother
    \smallskip
  \item Observed to converge in the same number of iterations
    \smallskip
  \item Can add overlap to preserve iterative performance for more
    ill-conditioned problems
  \end{itemize}

  \begin{columns}
    \begin{column}{0.5\textwidth}

      \begin{figure}[t!]
        \begin{center}
          \includegraphics[width=2.35in]{titan_strong_subdomain.pdf}
        \end{center}
        \caption{Strong scaling absolute efficiency for pure domain
          decomposition.}
        \label{fig:titan_strong_subdomain}
      \end{figure}

    \end{column}

    \begin{column}{0.5\textwidth}

      \begin{figure}[t!]
        \begin{center}
          \includegraphics[width=2.35in]{titan_weak_subdomain.pdf}
        \end{center}
        \caption{Weak scaling absolute efficiency for pure domain
          decomposition.}
        \label{fig:titan_weak_subdomain}
      \end{figure}

    \end{column}
  \end{columns}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{MCSA Application to Neutronics Summary}

  \small{
    \begin{itemize}
    \item MCSA can solve the asymmetric system generated by the $SP_N$
      equations
      \medskip
    \item Light water reactor problems are difficult to solve with MCSA as
      they have large spectral radii due to the neutron scattering in the
      moderator
      \medskip
    \item Advanced algebraic preconditioning strategies were applied to
      the $SP_N$ equations to obtain convergence with ILUT chosen for
      subsequent investigations
      \medskip
    \item MCSA was verified to produce the same flux distribution and
      k-eigenvalue for the fuel assembly as production Krylov methods
      \medskip
    \item MCSA was observed to converge in fewer iterations per eigenvalue
      iteration than GMRES for the fuel assembly criticality problem and
      more than Bi-CGStab using the same preconditioning
    \end{itemize}
  }

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{MCSA Application to Fluid Flow Summary}

  \small{
    \begin{itemize}
    \item Forward-Automated Newton-MCSA (FANM) has been developed
      \medskip
    \item The FANM method has been verified to produce the same solutions
      as a production Newton-Krylov method for three difficult benchmark
      problems
      \medskip
    \item The FANM method has better iterative performance than the
      Newton-Krylov method for convection dominated problems, converging
      in fewer linear solver iterations with the same preconditioning for
      high and low Rayleigh numbers
      \medskip
    \item The spectral radius convergence restriction on MCSA was observed
      to be a significant hindrance by preventing solutions to forced flow
      problems at high Reynolds numbers
      \medskip
    \item More Monte Carlo histories at every FANM iteration can reduce
      the number of linear and nonlinear iterations required to converge
      the problem
    \end{itemize}
  }

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Parallelization of MCSA Summary}

  \small{
    \begin{itemize}
    \item The multiple-set overlapping-domain (MSOD) parallel algorithm
      for particle transport has been adapted to parallelize MCSA
      \medskip
    \item MCSA scales favorably compared to production Krylov methods for
      both strong and weak scaling cases
      \medskip
    \item Overlap in small quantities can provide parallel efficiency
      boosts of a few percent in strong scaling cases but is
      ineffective in weak scaling cases
      \medskip
    \item Multiple sets offers a means to reduce time to solution by
      solving multiple copies of the original problem and combining the
      solutions using superposition
      \medskip
    \item MCSA is most efficiently used in parallel as a stochastic
      realization of an additive Schwarz method
    \end{itemize}
  }

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Future Work}

  \begin{itemize}
  \item Shortcomings observed on real problems
    \begin{itemize}
    \item Significant optimization required to determine production
      feasibility and true scalability
    \item Explicit algebraic preconditioning methods not sufficient
    \item Spectral radius limitation is severe
    \end{itemize}
    \medskip
  \item Performance improvements
    \begin{itemize}
    \item Random walk optimizations
    \item Alias sampling implementations
    \item Multiple set reduction analysis
    \item FANM forcing term and MCSA history relationships
    \end{itemize}
    \medskip
  \item Preconditioning improvements
    \begin{itemize}
    \item Alternative left/right sampling schemes
    \item Variance reduction based strategy
    \item Reduced order physics/PDE models for acceleration
    \end{itemize}
    \medskip
  \item Breaking away from $\rho(\mathbf{H}) < 1$
    \begin{itemize}
    \item Monte Carlo methods of the second degree
    \item Stochastic projection methods
    \end{itemize}
  \end{itemize}

\end{frame}

%%---------------------------------------------------------------------------%%
\begin{frame}{Discussion}

\end{frame}

%%---------------------------------------------------------------------------%%

\end{document}

### 2013-07-30 doc/ttsp2013/ttsp_2013.tex
adding initial ttsp 2013 paper draft

--- 2013-07-29 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
@@
-\chapter{Monte Carlo Synthetic Acceleration\\ Methods for the $SP_N$ Equations}
-\label{ch:spn_equations}
-
-In this chapter, we briefly derive the simplified $P_N$ ($SP_N$)
-equations, closely following the work of Evans
-\cite{evans_simplified_2013}\footnote{The $SP_N$ derivations in this
-  chapter and the associated appendices are heavily based on those
-  presented by Evans in \cite{evans_simplified_2013}}, in order to gain
-full understanding of the underlying system and its potential behavior
-in a Monte Carlo context. From the $P_N$ equations, we apply a set of
-approximations to yield the $SP_N$ equations for fixed source and
-criticality problems. Using the fully-formed linear operator for the
-transport problem, we explore solutions to the $SP_N$ equations with
-Monte Carlo Synthetic Acceleration using a challenging light water
-reactor fuel assembly criticality calculation as the driving
-problem. Several difficulties arise when applying MCSA to these
-problems that were not observed when investigating the simpler
-transport systems.
-
-For Jacobi-based preconditioners, convergence with MCSA is difficult
-if not impossible for ill-conditioned systems with this behavior
+\documentclass[letterpaper,12pt]{article}
+\usepackage[top=1.0in,bottom=1.0in,left=1.25in,right=1.25in]{geometry}
+\usepackage{verbatim}
+\usepackage{amssymb}
+\usepackage{graphicx}
+\usepackage{longtable}
+\usepackage{amsfonts}
+\usepackage{amsmath}
+\usepackage[usenames]{color}
+\usepackage[
+naturalnames = true, 
+colorlinks = true, 
+linkcolor = black,
+anchorcolor = black,
+citecolor = black,
+menucolor = black,
+urlcolor = blue
+]{hyperref}
+\usepackage{listings}
+\usepackage{textcomp}
+\definecolor{listinggray}{gray}{0.9}
+\definecolor{lbcolor}{rgb}{0.9,0.9,0.9}
+\lstset{
+  backgroundcolor=\color{lbcolor},
+  tabsize=4,
+  rulecolor=,
+  language=c++,
+  basicstyle=\scriptsize,
+  upquote=true,
+  aboveskip={1.5\baselineskip},
+  columns=fixed,
+  showstringspaces=false,
+  extendedchars=true,
+  breaklines=true,
+  prebreak =
+  \raisebox{0ex}[0ex][0ex]{\ensuremath{\hookleftarrow}},
+  frame=single,
+  showtabs=false,
+  showspaces=false,
+  showstringspaces=false,
+  identifierstyle=\ttfamily,
+  keywordstyle=\color[rgb]{0,0,1},
+  commentstyle=\color[rgb]{0.133,0.545,0.133},
+  stringstyle=\color[rgb]{0.627,0.126,0.941},
+}
+
+%%---------------------------------------------------------------------------%%
+\author{Stuart R. Slattery
+\\ \href{mailto:sslattery@wisc.edu}{\texttt{sslattery@wisc.edu}}
+}
+
+\date{\today}
+\title{Monte Carlo Synthetic Acceleration for the Simplified $P_N$ Approximation}
+\begin{document}
+\maketitle
+
+%%---------------------------------------------------------------------------%%
+\abstract
+
+In this paper, we present Monte Carlo synthetic acceleration as a
+linear solver technique and briefly derive the simplified $P_N$
+($SP_N$) for fixed source and criticality problems. Using the
+fully-formed linear operator for the transport problem, we explore
+solutions to the $SP_N$ equations with Monte Carlo Synthetic
+Acceleration using a challenging light water reactor fuel assembly
+criticality calculation as the driving problem. Several difficulties
+arise when applying MCSA to these problems that were not observed when
+investigating the simpler transport systems and we apply additional
+algebraic preconditioning techniques to achieve convergence. For
+Jacobi-based preconditioners, convergence with MCSA is difficult if
+not impossible for ill-conditioned systems with this behavior
 demonstrated using a simple neutron diffusion problem. In order to
@@
 several additional issues are observed and alleviated to a certain
-extent by applying the reduced domain approximation. MCSA solutions
-for the fuel assembly problem are then verified by comparing the
-solutions against production Krylov solvers for problems with varying
-energy groups. Finally performance is analyzed through comparison with
-those same Krylov solvers in terms of both iterative performance and
-CPU timing using the same set of problems utilized for the
-verification.
+extent by applying the reduced domain approximation. Finally
+performance is analyzed through comparison with those same Krylov
+solvers in terms of both iterative performance and CPU timing using
+the fuel assembly problem.
 
 %%---------------------------------------------------------------------------%%
-\section{The Neutron Transport Equation}
-\label{sec:transport_eq}
-As a starting point for the $SP_N$ equations we define the
-time-independent neutron transport equation
-\cite{lewis_computational_1993}:
-\begin{multline}
-  \hat{\Omega} \cdot \vec{\nabla} \psi(\vec{r},\hat{\Omega},E) +
-  \sigma(\vec{r},E) \psi(\vec{r},\hat{\Omega},E) = \\ \iint
-  \sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}' \cdot \hat{\Omega})
-  \psi(\vec{r},\hat{\Omega}',E') d\Omega' dE' +
-  q(\vec{r},\hat{\Omega},E)\:,
-  \label{eq:general_transport}
-\end{multline}
-with the variables defined as:
-\begin{itemize}
-\item $\vec{r}$ - neutron spatial position
-\item $\hat{\Omega}$ - neutron streaming direction with radial
-  component $\mu$ and azimuthal component $\omega$
-\item $\hat{\Omega}' \cdot \hat{\Omega} = \mu_0$ is the angle of
-  scattering
-\item $E$ - neutron energy
-\item $\psi(\vec{r},\hat{\Omega},E)$ - angular flux
-\item $\sigma(\vec{r},E)$ - total interaction cross section
-\item $\sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}')$ - probability
-  of scattering from direction $\hat{\Omega}'$ into an angular domain
-  $d\hat{\Omega}'$ about the direction $\hat{\Omega}$ and from energy
-  $E'$ to an energy domain $dE'$ about energy $E$
-\item $q(\vec{r},\hat{\Omega},E)$ - external source of neutrons.
-\end{itemize}
-For this work, it is sufficient to formulate
-Eq~(\ref{eq:general_transport}) in 1-dimensional Cartesian geometry:
-\begin{multline}
-  \mu \frac{\partial}{\partial x} \psi(x,\mu,E) + \sigma(x,E)
-  \psi(x,\mu,E) = \\ \iint \sigma_s(x,E' \rightarrow E,\hat{\Omega}'
-  \cdot \hat{\Omega}) \psi(x,\hat{\Omega}',E') d\Omega' dE' +
-  \frac{q(x,E)}{4 \pi}\:,
-  \label{eq:cart_1d_transport}
-\end{multline}
-where the angular component of the solution is no longer dependent on
-the azimuthal direction of travel and an isotropic source of neutrons
-is assumed. In addition, for fission systems the eigenvalue form of
-the transport equation is:
-\begin{multline}
-  \hat{\Omega} \cdot \vec{\nabla} \psi(\vec{r},\hat{\Omega},E) +
-  \sigma(\vec{r},E) \psi(\vec{r},\hat{\Omega},E) = \\ \iint
-  \sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}' \cdot \hat{\Omega})
-  \psi(\vec{r},\hat{\Omega}',E') d\Omega' dE' + \\ \frac{1}{k} \chi(E)
-  \iint \nu \sigma_f(\hat{r},E') \psi(\vec{r},\hat{\Omega}',E')
-  d\Omega' dE' + q(\vec{r},\hat{\Omega},E) \:,
-  \label{eq:eigenvalue_transport}
-\end{multline}
-with the additional variables defined as
-\begin{itemize}
-\item $k$ - multiplication factor
-\item $\chi(E)$ - fission neutron energy spectrum
-\item $\nu$ - average number of neutrons per fission
-\item $\sigma_f(r,E')$ - fission cross section\:.
-\end{itemize}
-In 1-dimensional Cartesian geometry,
-Eq~(\ref{eq:eigenvalue_transport}) becomes:
-\begin{multline}
-  \mu \frac{\partial}{\partial x} \psi(x,\mu,E) + \sigma(x,E)
-  \psi(x,\mu,E) = \\ \iint \sigma_s(x,E' \rightarrow
-  E,\hat{\Omega}' \cdot \hat{\Omega}) \psi(x,\hat{\Omega}',E')
-  d\Omega' dE' + \\ \frac{1}{k} \chi(E)
-  \iint \nu \sigma_f(x,E') \psi(x,\hat{\Omega}',E') d\Omega'
-  dE' + \frac{q(x,E)}{4 \pi}\:.
-  \label{eq:cart_1d_eigenvalue}
-\end{multline}
+\section{Introduction}
 
 %%---------------------------------------------------------------------------%%
-\section{Derivation of the Monoenergetic $SP_N$ Equations}
-\label{sec:spn_equations}
-The $P_N$ equations as derived in \cite{lewis_computational_1993}
-give $N+1$ coupled first-order equations capturing the spatial and
-angular-dependence of the solution:
-\begin{equation}
-   \frac{1}{2n+1} \frac{\partial}{\partial x}\Big[ (n+1) \phi_{n+1} + n
-     \phi_{n-1} \Big] + \Sigma_n \phi_n = q\delta_{n0} \:,
-  \label{eq:final_pn_equations}
-\end{equation}
-where $\Sigma_n = \sigma-\sigma_{sn}$ and the summations are truncated
-at some level of approximation $N$ such that $n = 0,1,\dotsc,N$. This
-yields a set of $N+1$ equations for $N+2$ flux moments. In multiple
-dimensions, the equation set becomes large and coupled not only
-through angular moments but also through the spatial variables. As a
-simpler alternative to multidimensional $P_N$ solutions, Gelbard
-recognized in 1960 that the planar $P_N$ equations could be simplified
-and applied an ad-hoc method to extend them to multiple dimensions,
-yielding the $SP_N$ equations. These equations are not only fewer in
-number, but also take on a diffusion-like form while maintaining the
-angular character of the flux, making them amenable to solutions with
-modern diffusion methods.
-
-First, the $P_N$ equations can be simplified to $(N+1)/2$ second-order
-equations by solving for the $n^{th}$ Legendre flux moment in the
-odd-order equations:
-\begin{equation}
-  \phi_n = \frac{1}{\Sigma_n}\Bigg[ q \delta_{no} -
-    \frac{\partial}{\partial x}\Big(\frac{n}{2n+1}\phi_{n-1} +
-    \frac{n+1}{2n+1} \phi_{n+1} \Big) \Bigg]\:, 
-  \label{eq:odd_moments}
-\end{equation}
-for $n = 1,3,\cdots,N$ and $\delta_{no} = 0\ \forall n \neq 0$. We can
-insert the odd moments into Eq~(\ref{eq:final_pn_equations}) to get a
-reduced group of equations for the even moments:
-\begin{multline}
-  -\frac{\partial}{\partial x}
-  \Bigg[\frac{n}{2n+1}\frac{1}{\Sigma_{n-1}} \frac{\partial}{\partial
-      x} \Big(\frac{n-1}{2n-1} \phi_{n-2} + \frac{n}{2n-1}\phi_n \Big)
-    \\+ \frac{n+1}{2n+1}\frac{1}{\Sigma_{n+1}} \frac{\partial}{\partial
-      x} \Big(\frac{n+1}{2n+3}\phi_n + \frac{n+2}{2n+3}\phi_{n+2}\Big)
-    \Bigg] \\+ \Sigma_n \phi_n = q \delta_{n0}\ \ \ \ \ \ \ \ \ n =
-  0,2,4,\cdots,N\:.
-  \label{eq:reduced_pn}
-\end{multline}
-Immediately, we note the diffusion-like nature of
-Eq~(\ref{eq:reduced_pn}) as compared to the original $P_N$
-equations. To extend these equations to multiple dimensions, Gelbard
-simply replaced the planar spatial derivatives in the reduced set of
-equations with general multidimensional gradient operators:
-\begin{multline}
-  -\nabla \cdot \Bigg[\frac{n}{2n+1}\frac{1}{\Sigma_{n-1}} \nabla
-    \Big(\frac{n-1}{2n-1} \phi_{n-2} + \frac{n}{2n-1}\phi_n \Big) \\+
-    \frac{n+1}{2n+1}\frac{1}{\Sigma_{n+1}} \nabla
-    \Big(\frac{n+1}{2n+3}\phi_n + \frac{n+2}{2n+3}\phi_{n+2}\Big)
-    \Bigg] \\+ \Sigma_n \phi_n = q \delta_{n0}\ \ \ \ \ \ \ \ \ n =
-  0,2,4,\cdots,N\:,
-  \label{eq:spn_equations}
-\end{multline}
-yielding a multidimensional set of $(N+1)/1$ angular coupled equations
-defined as the $SP_N$ equations. As with the $P_N$ equations, we
-provide closure to this set of equations with $\phi_{N+1} = 0$. As a
-concrete example, we will consider the $SP_7$ equations:
+\section{Monte Carlo Synthetic Acceleration}
+
+Using the ideas of Halton, Evans and Mosher recently developed a Monte
+Carlo solution method that was not prohibited severely by the quality
+of the initial guess for the system \cite{evans_monte_2009} and later
+applied it more rigorously as a solution mechanism for the radiation
+diffusion equation \cite{evans_monte_2012}. With their new methods,
+they achieved identical numerical results as conventional Krylov
+solvers as well as comparable performance in both number of iterations
+and CPU time. Their approach was instead to use residual Monte Carlo
+as a synthetic acceleration for a stationary method. To derive this
+method, we begin by splitting the operator in
+Eq~(\ref{eq:linear_problem})
+\begin{equation}
+  \ve{x} = (\ve{I} - \ve{A})\ve{x} + \ve{b}\:.
+  \label{eq:linear_split}
+\end{equation}
+With this we can then define the stationary method
+\textit{Richardson's iteration} as:
+\begin{equation}
+  \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,
+  \label{eq:richardsons_iteration}
+\end{equation}
+which will converge if $\rho(\ve{I} - \ve{A}) < 1$. We then define the
+solution error at the $k^{th}$ iterate relative to the true solution:
+\begin{equation}
+  \delta \ve{x}^k = \ve{x} - \ve{x}^k\:.
+  \label{eq:mcsa_error}
+\end{equation}
+Subtracting Eq~(\ref{eq:richardsons_iteration}) from
+Eq~(\ref{eq:linear_split}) we get:
+\begin{equation}
+  \delta \ve{x}^{k+1} = (\ve{I} - \ve{A})\delta \ve{x}^k\:.
+  \label{eq:mcsa_setup_1}
+\end{equation}
+Subtracting from this $(\ve{I} - \ve{A})\delta \ve{x}^{k+1}$ yields:
+\begin{equation}
+  \begin{split}
+    \ve{A}\delta \ve{x}^{k+1} &= (\ve{I} -
+    \ve{A})(\ve{x}^{k+1}-\ve{x}^{k}) \\ &= \ve{r}^{k+1}\:.
+    \label{eq:mcsa_setup_2}
+  \end{split}
+\end{equation}
+Using this, we define the following scheme that will converge in one
+iteration if $\ve{A}$ is inverted exactly:
 \begin{subequations}
   \begin{gather}
-    -\nabla \cdot \frac{1}{3 \Sigma_1} \nabla ( \phi_0 + 2\phi_2 ) +
-    \Sigma_0 \phi_0 = q \\ 
-    -\nabla \cdot \Bigg[ \frac{2}{15 \Sigma_1} \nabla ( \phi_0 + 2\phi_2
-      ) + \frac{3}{35 \Sigma_3}\nabla( 3\phi_2 + 4\phi_4)\Bigg] +
-    \Sigma_2 \phi_2 = 0\\
-    -\nabla \cdot \Bigg[ \frac{4}{63 \Sigma_3} \nabla ( 3\phi_2 +
-      4\phi_4 ) + \frac{5}{99 \Sigma_5}\nabla( 5\phi_4 +
-      6\phi_6)\Bigg] + \Sigma_4 \phi_4 = 0\\
-    -\nabla \cdot \Bigg[ \frac{6}{143 \Sigma_5} \nabla ( 5\phi_4 +
-      6\phi_6 ) + \frac{7}{195 \Sigma_7}\nabla(7\phi_6)\Bigg] +
-    \Sigma_6 \phi_6 = 0 \:.
+    \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,\\
+    \ve{A} \delta \ve{x}^{k+1} = \ve{r}^{k+1}\:,\\
+    \ve{x} = \ve{x}^{k+1} + \delta \ve{x}^{k+1}\:.
   \end{gather}
-  \label{eq:sp7_equations}
+  \label{eq:mcsa_setup_3}
 \end{subequations}
-To further modify these equations, we can use a change of variables to
-create a new group of equations such that the gradients are operating
-on a single vector:
+However, $\ve{A}$ is only approximately inverted by our numerical
+methods and therefore we instead pose an iterative scheme in which the
+Monte Carlo solvers are used to invert the operator. The
+\textit{Fixed-Point Monte Carlo Synthetic Acceleration} (MCSA) method
+is defined as:
 \begin{subequations}
   \begin{gather}
-    u_1 = \phi_0 + 2\phi_2 \\
-    u_2 = 3\phi_2 + 4\phi_4 \\
-    u_3 = 5\phi_4 + 6\phi_6 \\
-    u_4 = 7\phi_6 \:.
+    \ve{x}^{k+1/2} = \ve{x}^k + \ve{r}^k\:,\\
+    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\
+    \ve{A}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:,\\
+    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,\\
+    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:,
   \end{gather}
-  \label{eq:spn7_subs}
+  \label{eq:mcsa}
 \end{subequations}
-When substituted into Eq~(\ref{eq:sp7_equations}), these terms give:
-\begin{subequations}
-  \begin{gather}
-    -\nabla \cdot \frac{1}{3 \Sigma_1} \nabla u_1 + \Sigma_0 \Bigg[
-    u_1 - \frac{2}{3}u_2 + \frac{8}{15}u_3 - \frac{16}{35}u_4 \Bigg]
-    = -q \\
-    -\nabla \cdot \Bigg[ \frac{2}{15 \Sigma_1} \nabla u_1 +
-    \frac{3}{35 \Sigma_3} \nabla u_2 \Bigg] + \Sigma_2 \Bigg[
-    \frac{1}{3}u_2 - \frac{4}{15}u_3 + \frac{8}{35}u_4 \Bigg] = 0 \\
-    -\nabla \cdot \Bigg[ \frac{4}{63 \Sigma_3} \nabla u_2 +
-    \frac{5}{99 \Sigma_5} \nabla u_3 \Bigg] + \Sigma_4 \Bigg[
-    \frac{1}{5}u_3 - \frac{6}{35}u_4 \Bigg] = 0 \\ 
-    -\nabla \cdot \Bigg[ \frac{6}{143 \Sigma_5} \nabla u_3 +
-    \frac{7}{195 \Sigma_7} \nabla u_4 \Bigg] + \Sigma_6 \Bigg[
-    \frac{1}{7}u_4 \Bigg] = 0 \:.
-  \end{gather}
-  \label{eq:spn7_subs_equations}
-\end{subequations}
-If we rearrange Eq~(\ref{eq:spn7_subs_equations}) such that only one
-divergence operation is present in each equation, we can formulate
-this as a matrix system of 4 equations in the case of the $SP_7$
-approximation:
-\begin{equation}
-  -\nabla \cdot D_n \nabla u_n + \sum_{m=1}^4 A_{nm} u_m =
-  q_n\ \ \ \ \ \ \ n = 1,2,3,4\:,
-  \label{eq:spn_matrix}
-\end{equation}
-with $\mathbf{u}$ the vector of solution variables:
-\begin{equation}
-  \mathbf{u} = ( u_1\ \ u_2\ \ u_3\ \ u_4 )^T \:,
-  \label{eq:spn7_solution_vector}
-\end{equation}
-$\mathbf{D}$ the vector of effective diffusion coefficients:
-\begin{equation}
-  \mathbf{D} = \Bigg( \frac{1}{3\Sigma_1}\ \ \frac{1}{7\Sigma_3}\ \
-  \frac{1}{11\Sigma_5}\ \ \frac{1}{15\Sigma_7} \Bigg)^T\:,
-  \label{eq:spn7_diffusion_coeffs}
-\end{equation}
-$\mathbf{q}$ the vector of source terms where the $0^{th}$ moment
-source has now been distributed through the system:
-\begin{equation}
-  \mathbf{q} = (
-  q\ \ -\frac{2}{3}q\ \ \frac{8}{15}q\ \ -\frac{16}{35}q )^T\:,
-  \label{eq:spn7_source_vector}
-\end{equation}
-and $\mathbf{A}$ a matrix of angular scattering terms:
-% NOTE: I copied the following matrix directly out of Tom's tech note
-% on the SPn equations which I am effectively following here because I
-% was feeling lazy. I have verified its correctness.
-\begin{equation}
-  \mathbf{A} = 
-  {\tiny \begin{bmatrix}
-    (\Sigma_0) &
-    (-\frac{2}{3}\Sigma_0) &
-    (\frac{8}{15}\Sigma_0) &
-    (-\frac{16}{35}\Sigma_0) \\
-    %%
-    &&&\\
-    %%
-    (-\frac{2}{3}\Sigma_0) &
-    (\frac{4}{9}\Sigma_0 + \frac{5}{9}\Sigma_2) &
-    (-\frac{16}{45}\Sigma_0 - \frac{4}{9}\Sigma_2) &
-    (\frac{32}{105}\Sigma_0 + \frac{8}{21}\Sigma_2) \\
-    %%
-    &&&\\
-    %%
-    (\frac{8}{15}\Sigma_0) &
-    (-\frac{16}{45}\Sigma_0 - \frac{4}{9}\Sigma_2) &
-    (\frac{64}{225}\Sigma_0 + \frac{16}{45}\Sigma_2 + \frac{9}{25}\Sigma_4) &
-    (-\frac{128}{525}\Sigma_0 - \frac{32}{105}\Sigma_2 - \frac{54}{175}\Sigma_4)
-    \\ 
-    %%
-    &&&\\
-    %%
-    (-\frac{16}{35}\Sigma_0) &
-    (\frac{32}{105}\Sigma_0 + \frac{8}{21}\Sigma_2) &
-    (-\frac{128}{525}\Sigma_0 - \frac{32}{105}\Sigma_2 - \frac{54}{175}\Sigma_4)
-    & 
-    (\frac{256}{1225}\Sigma_0 + \frac{64}{245}\Sigma_2 +
-    \frac{324}{1225}\Sigma_4 + \frac{13}{49}\Sigma_6)
-  \end{bmatrix}}\:.
-  \label{eq:A_matrix}
-\end{equation}
-Note that the term $\sum_{m=1}^4 A_{nm} u_m$ in
-Eq~(\ref{eq:spn_matrix}) couples the moments in each equation while
-the diffusive term in each equation is only for a single
-'pseudo-moment' $u_n$. As noted by Evans \cite{evans_simplified_2013},
-lower order $SP_N$ approximations can be generated by setting higher
-order even moments in this system to zero (e.g. $\phi_6 = \phi_4 = 0$
-yields the $SP_3$ equations).
-
-\subsection{Eigenvalue Form of the $SP_N$ Equations}
-\label{subsec:eigenvalue_form}
-For criticality problems, the space, angle, and energy discretizations
-for the $SP_N$ equations can be applied to the general eigenvalue form
-of the transport equation given by
-Eq~(\ref{eq:eigenvalue_transport}). We can readily replace the fixed
-source in Eq~(\ref{eq:spn_matrix}) with a fission source this time
-considering the multigroup form of the $SP_N$ equations derived in
-Appendix~\ref{chap:mg_spn_equations}:
-\begin{multline}
-  -\nabla \cdot \Bigg[\frac{n}{2n+1}\mathbf{\Sigma_{n-1}}^{-1} \nabla
-    \Big(\frac{n-1}{2n-1} \mathbf{\Phi_{n-2}} +
-    \frac{n}{2n-1}\mathbf{\Phi_n} \Big) \\+
-    \frac{n+1}{2n+1}\mathbf{\Sigma_{n+1}}^{-1} \nabla
-    \Big(\frac{n+1}{2n+3}\mathbf{\Phi_n} +
-    \frac{n+2}{2n+3}\mathbf{\Phi_{n+2}}\Big) \Bigg] \\+
-  \mathbf{\Sigma_n} \mathbf{\Phi_n} = \frac{1}{k} \mathbf{F}
-  \mathbf{\Phi_n} \delta_{n0} \ \ \ \ \ \ \ \ \ n = 0,2,4,\cdots,N\:.
-  \label{eq:multigroup_spn_eigenvalue}
-\end{multline}
-with the fission matrix $\mathbf{F}$ given by
-\cite{evans_simplified_2013}. We again apply the change of variables to
-yield the set of multigroup pseudo-moments $\mathbb{U}_n$ where now
-the application of the fission matrix to the moment vectors will be
-expanded into a set of block matrices in identical fashion to the
-scattering matrices:
-\begin{equation}
-  -\nabla \cdot \mathbb{D}_n \nabla \mathbb{U}_n + \sum_{m=1}^4
-  \mathbb{A}_{nm} \mathbb{U}_m = \frac{1}{k} \sum_{m=1}^4
-  \mathbb{F}_{nm} \mathbb{U}_n\ \ \ \ \ \ \ n = 1,2,3,4\:.
-  \label{eq:spn_fission_matrix}
-\end{equation}
-It should be noted here that with respect to a general MCSA scheme,
-the introduction of the fission in the system will only affect the
-source vector in the linear system. The linear operator, and therefore
-overall MCSA performance will be dictated by streaming and scattering
-as defined on the left-hand side.
+where a Neumann-Ulam Monte Carlo method is used to generate the
+solution correction from the residual and Richardson's iteration in
+the first step has been rewritten as a residual correction. Using
+Monte Carlo in this way achieves the same effect as Halton's method,
+decoupling its convergence rate from the overall convergence rate of
+the method. Here, the approximate Monte Carlo solution is not driven
+to a particular convergence as it merely supplies a correction for the
+initial guess generated by Richardson's iteration. Rather, only a set
+number of histories are required using the Neumann-Ulam method to
+generate the correction. In addition, the fact that the scheme in
+Eq~(\ref{eq:mcsa_setup_3}) will converge in one iteration if $\ve{A}$
+is inverted exactly means that as more and more stochastic histories
+are used to compute the correction and the error is reduced towards
+zero, the number of iterations required for MCSA to converge should
+decrease accordingly, thus accelerating the solution.
+
+In addition to the Monte Carlo solver parameters dictating the number
+of histories and weight cutoff, the outer MCSA iterations also have
+the following stopping criteria:
+\begin{equation}
+  ||\ve{r}||_\infty < \epsilon \ ||\ve{b}||_\infty\:,
+  \label{eq:mcsa_stopping_criteria}
+\end{equation}
+where $\epsilon$ is a user-defined parameter. As with any iterative
+method, other stopping criteria using other vector norms could be
+computed, however, for this work we will only use
+Eq~(\ref{eq:mcsa_stopping_criteria}). We therefore have 3 parameters
+to tune in an MCSA implementation: the number of Monte Carlo histories
+computed in the Neumann-Ulam solve during each MCSA iteration, the
+weight cutoff for those histories, and the total MCSA convergence
+tolerance as specified by $\epsilon$.
 
 %%---------------------------------------------------------------------------%%
-\section{Spectral Analysis of the $SP_N$ Equations}
-\label{sec:spn_spectral_analysis}
-Before we can move on to solving the $SP_N$ equations with both
-conventional and Monte Carlo methods, we must first understand their
-spectral character to ensure that MCSA will be applicable. As
-presented in Chapter~\ref{ch:stochastic_methods}, the spectral radius
-of the iteration matrix must be less than unity to ensure
-convergence. Therefore, we will compute the spectral radius of the
-iteration matrix formed by the $SP_N$ equations while parametrically
-varying the $SP_N$ order, the $P_N$ order, the number of energy
-groups, and upscattering/downscattering between energy groups. For
-each problem in the parameter study, the values given in
-Table~\ref{tab:spn_fixed_parameters} were fixed for each spatial
-location and energy group.
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{ll}\hline\hline
-      \multicolumn{1}{c}{Parameter}& 
-      \multicolumn{1}{c}{Value} \\\hline
-      Mesh Element Size & 1.0 \\
-      Mesh Elements & $4 \times 4 \times 4$ \\
-      Reflecting Boundaries & True \\
-      Materials & 1 \\
-      Source Strength & 1.0 \\
-      Total Cross Section & 5.0 \\
-      In-Group Cross Section & 0.25 \\
-      Downscatter Cross Section & 1.0 \\
-      Upscatter Cross Section & 0.1 \\
-      Eigenvalue Solver Tolerance & $1.0\times10^{-8}$ \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Fixed parameters for the $SP_N$ spectral radius
-      parameter study.} \textit{The parameters were fixed for each
-      spatial location and energy group.}}
-  \label{tab:spn_fixed_parameters}
-\end{table}
-Mesh element sizes were the same in all cardinal directions with
-reflecting conditions on each boundary. For the energy parameter, 1
-and 10 energy groups were used with upscatter/downscatter varied for
-the 10 group case. For the downscatter cases, all groups could scatter
-to all lower energy groups with the cross sections given in
-Table~\ref{tab:spn_fixed_parameters} while for the upscatter case all
-groups could scatter to all other energy groups with different cross
-sections for upscatter and downscatter used. A uniform isotropic
-source of neutrons is given in each group as well.
-
-The matrices generated by the discretization of this problem are
-sparse with a block-based form as dictated by
-Eq~(\ref{eq:spn_multigroup_system}). Figure~\ref{fig:group1} gives the
-sparsity pattern for a monoenergetic $SP_7$ discretization while
-Figures~\ref{fig:group10ds} and \ref{fig:group10us} give the sparsity
-patterns for the same problem for 10 energy groups without and with
-upscatter respectively. For these figures, the parameters in
-Table~\ref{tab:spn_fixed_parameters} with the number of spatial
-elements reduced to $2 \times 2 \times 2$ to show the blocks generated
-by the discretization.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=3.25in]{chapters/spn_equations/group1.png}
-  \end{center}
-  \caption{\textbf{Sparsity pattern for 1-group $SP_7$
-      discretization.} \textit{A $2\times 2 \times 2$ element mesh was
-      used to show detail of the blocks formed by the discretization.}}
-  \label{fig:group1}
-\end{figure}
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=3.25in]{chapters/spn_equations/group10ds.png}
-  \end{center}
-  \caption{\textbf{Sparsity pattern for 10-group $SP_7$ discretization
-      with downscatter only.} \textit{A $2\times 2 \times 2$ element
-      mesh was used to show detail of the blocks formed by the
-      discretization.}}
-  \label{fig:group10ds}
-\end{figure}
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=3.25in]{chapters/spn_equations/group10us.png}
-  \end{center}
-  \caption{\textbf{Sparsity pattern for 10-group $SP_7$ discretization
-      with downscatter and upscatter.} \textit{A $2\times 2 \times 2$ element
-      mesh was used to show detail of the blocks formed by the
-      discretization.}}
-  \label{fig:group10us}
-\end{figure}
-We note a few key features of the sparsity plots. The first is that
-for multigroup problems without full upscatter and downscatter
-(i.e. Figure~\ref{fig:group10ds}), the resulting matrix is asymmetric
-and therefore a linear solver that can handle asymmetric linear
-systems is required. Nearly all problems of interest will not have
-full upscattering or downscattering. Second we note the largely
-diagonal character of these systems, although the blocks from
-Eq~(\ref{eq:A_block_matrix}) are readily apparent. Our first attempt
-at preconditioning this system will be to use the point Jacobi
-preconditioning from \S\ref{subsubsec:basic_mcsa_preconditioning} due
-to this diagonal form.
-
-\subsection{Point Jacobi Spectral Analysis Results}
-\label{subsec:spn_analysis_results}
-Spectral radius computations were performed for the cases described
-above for the point Jacobi preconditioned iteration matrix $\ve{H} =
-\ve{I}-\ve{M}^{-1}\ve{A}$ with $\ve{M} =
-diag(\ve{A})$. Table~\ref{tab:group1pj} gives the results for the
-1-group case, Table~\ref{tab:group10dspj} for the 10-group case with
-full downscatter only and Table~\ref{tab:group10uspj} for the 10-group
-case with full downscatter and full upscatter.
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{cccccc}\hline\hline
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{$SP_N$ Order}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{} \\
-       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
-       & \textbf{0} & 0.0635 & 0.6722 & 1.3144 & 1.976 \\
-       & \textbf{1} & 0.0666 & 0.6728 & 1.3141 & 1.9755 \\
-      $P_N$ Order & \textbf{3} & 0.0666 & 0.6822 & 1.3141 & 1.9755 \\
-       & \textbf{5} & 0.0666 & 0.6822 & 1.3278 & 1.9847 \\
-       & \textbf{7} & 0.0666 & 0.6822 & 1.3278 & 1.9917 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Spectral radius results for the point Jacobi
-      preconditioned iteration matrix with 1 energy group.}}
-  \label{tab:group1pj}
-\end{table}
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{cccccc}\hline\hline
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{$SP_N$ Order}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{} \\
-       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
-       & \textbf{0} & 0.0655 & 0.677 & 1.32 & 1.982 \\
-       & \textbf{1} & 0.071 & 0.6777 & 1.319 & 1.982 \\
-      $P_N$ Order & \textbf{3} & 0.071 & 0.687 & 1.327 & 1.9872 \\
-       & \textbf{5} & 0.071 & 0.687 & 1.336 & 1.997 \\
-       & \textbf{7} & 0.071 & 0.687 & 1.336 & 1.9995 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Spectral radius results for the point Jacobi
-      preconditioned iteration matrix with 10 energy groups and full
-      downscatter.}}
-  \label{tab:group10dspj}
-\end{table}
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{cccccc}\hline\hline
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{$SP_N$ Order}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{} \\
-       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
-       & \textbf{0} & 0.7283 & 0.81 & 1.47 & 2.1446 \\
-       & \textbf{1} & 0.7317 & 0.8 & 1.46 & 2.1368 \\
-      $P_N$ Order & \textbf{3} & 0.7317 & 0.91 & 1.526 & 2.2274 \\
-       & \textbf{5} & 0.7317 & 0.91 & 1.5344 & 2.2562 \\
-       & \textbf{7} & 0.7317 & 0.91 & 1.5345 & 2.2842 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Spectral radius results for the point Jacobi
-      preconditioned iteration matrix with 10 energy groups, full
-      downscatter and full upscatter.}}
-  \label{tab:group10uspj}
-\end{table}
-
-It is readily apparent from the tabulated data that point Jacobi
-preconditioning is insufficient. For problems of order $SP_5$ and
-$SP_7$, the method will not converge at all and for the upscatter case
-with $SP_3$ the spectral radius is still quite large for large $P_N$
-orders and therefore convergence is expected to be slow. These
-eigenvalues signal a need for a better preconditioning strategy to
-both ensure and improve convergence for Monte Carlo methods.
-
-\subsection{Block Jacobi Preconditioning}
-\label{sec:spn_preconditioning}
-If Monte Carlo methods are to be used to solve the $SP_N$ system of
-equations, a different preconditioning strategy is required in order
-to ensure convergence for systems of all $SP_N$ and $P_N$ orders with
-arbitrary energy group structures. As another means of achieving this,
-we look back to the sparsity plots we generated in
-Figures~\ref{fig:group1}, \ref{fig:group10ds} and \ref{fig:group10us}
-as well as the multigroup $SP_N$ equations. Initially, the diagonal
-character of the system led us to try point Jacobi preconditioning
-with only marginal results. From the sparsity plots we note the block
-structure that ultimately arises from the multigroup scattering
-matrices and their insertion into Eq~(\ref{eq:A_block_matrix}). When
-full upscatter and downscatter are used the resulting blocks are
-completely dense while only downscatter gives a lower triangular
-scattering matrix and the block structure shown in
-Figure~\ref{fig:group10ds}.
-
-Based on this both block and diagonally dominant structure for
-matrices formed by the general multigroup $SP_N$ equations, we instead
-choose \textit{block Jacobi} preconditioning as a left preconditioner
-for the system. Like point Jacobi preconditioning, block Jacobi
-preconditioning extracts the diagonal elements of the matrix as the
-preconditioner where now the elements extracted are the blocks on the
-diagonal as shown on the left side of
-Figure~\ref{fig:block_jacobi_ex}. Shown on the right side of
-Figure~\ref{fig:block_jacobi_ex}, inversion of the preconditioner is
-trivial with each diagonal block inverted separately. For the $SP_N$
-equations, Eq~(\ref{eq:A_block_matrix}) gives a block size of
-$N_g\times(N+1)/2$.
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.5}{
-    \input{chapters/spn_equations/block_jacobi.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Block Jacobi preconditioning strategy used for the
-      $SP_N$ equations.} \textit{Left: The preconditioner is formed by
-      the diagonal blocks of the matrix. Right: Inversion of the
-      preconditioner is trivial and decoupled by block.}}
-  \label{fig:block_jacobi_ex}
-\end{figure}
-The inversion of this preconditioner is trivial as shown on the right
-side of Figure~\ref{fig:block_jacobi_ex}. Each block can be inverted
-individually and combined to form the inverse. In addition, in the
-limit of a block size of one, the block Jacobi method reduces to the
-point Jacobi method.
-
-For high performance implementations this has several attractive
-properties. First, the blocks in the matrix come from the energy/angle
-discretization of the transport equation as given by
-Eq~(\ref{eq:A_block_matrix}). Each block on the diagonal is bound to a
-mesh element in the system (note there are 8 blocks on the diagonal in
-each of the sparsity patterns with a mesh of $2 \times 2 \times 2$)
-and therefore we expect the matrix elements forming the block to be
-entirely local. Second, these blocks are typically dense and nearly
-lower triangular for many transport problems meaning that established
-dense matrix methods can be used for fast inversion.
-
-\subsection{Block Jacobi Spectral Analysis Results}
-\label{subsec:spn_analysis_results}
-Spectral radius computations for the block Jacobi preconditioned
-iteration matrix were performed for the same problems as the point
-Jacobi preconditioned case. Table~\ref{tab:group1bj} gives the results
-for the 1-group case, Table~\ref{tab:group10dsbj} for the 10-group
-case with full downscatter only and Table~\ref{tab:group10usbj} for
-the 10-group case with full downscatter and full upscatter. A block
-size of $N_g\times(N+1)/2$ was used for the preconditioner in all
-cases.
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{cccccc}\hline\hline
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{$SP_N$ Order}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{} \\
-       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
-       & \textbf{0} & 0.0635 & 0.1269 & 0.1444 & 0.1513 \\
-       & \textbf{1} & 0.0666 & 0.1315 & 0.1474 & 0.1534 \\
-      $P_N$ Order & \textbf{3} & 0.0666 & 0.1365 & 0.154 & 0.1592 \\
-       & \textbf{5} & 0.0666 & 0.1365 & 0.1562 & 0.163 \\
-       & \textbf{7} & 0.0666 & 0.1365 & 0.1562 & 0.164 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Spectral radius results for the block Jacobi
-      preconditioned iteration matrix with 1 energy group.}}
-  \label{tab:group1bj}
-\end{table}
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{cccccc}\hline\hline
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{$SP_N$ Order}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{} \\
-       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
-       & \textbf{0} & 0.0647 & 0.1275 & 0.1449 & 0.1514 \\
-       & \textbf{1} & 0.0686 & 0.1338 & 0.1484 & 0.1547 \\
-      $P_N$ Order & \textbf{3} & 0.0687 & 0.1399 & 0.1582 & 0.1625 \\
-       & \textbf{5} & 0.0692 & 0.1399 & 0.1582 & 0.1657 \\
-       & \textbf{7} & 0.0678 & 0.1393 & 0.1624 & 0.166 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Spectral radius results for the block Jacobi
-      preconditioned iteration matrix with 10 energy groups and full
-      downscatter.}}
-  \label{tab:group10dsbj}
-\end{table}
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{cccccc}\hline\hline
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{$SP_N$ Order}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{} \\
-       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
-       & \textbf{0} & 0.1887 & 0.2267 & 0.2285 & 0.2286 \\
-       & \textbf{1} & 0.4535 & 0.5044 & 0.5045 & 0.5045 \\
-      $P_N$ Order & \textbf{3} & 0.4535 & 0.6453 & 0.6506 & 0.6506 \\
-       & \textbf{5} & 0.4535 & 0.6453 & 0.6802 & 0.6818 \\
-       & \textbf{7} & 0.4535 & 0.6453 & 0.6802 & 0.6927 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Spectral radius results for the block Jacobi
-      preconditioned iteration matrix with 10 energy groups, full
-      downscatter and full upscatter.}}
-  \label{tab:group10usbj}
-\end{table}
-
-From the tabulated block Jacobi data it is clear that this is a viable
-preconditioning choice for all $SP_N$ problems in term of Monte Carlo
-solution methods. All cases were observed to have a spectral radius
-below unity and often significantly smaller than that, greatly
-improving convergence rates over the point Jacobi preconditioned
-problem. Based on these results, the block Jacobi method should be the
-first preconditioning strategy applied to more complicated problems
-representative of physical systems.
+\section{Simplified $P_N$ Approximation}
 
 %%---------------------------------------------------------------------------%%
-\section{Fuel Assembly Criticality Calculations}
-\label{sec:fuel_assembly_calcs}
+\section{Model Problem}
 Fuel assembly calculations are a critical piece of nuclear engineering
@@
 
+%%---------------------------------------------------------------------------%%
+\section{Preconditioned MCSA}
+
+In most cases, at least a minimal amount of \textit{preconditioning}
+of the linear system will be required in order to use the class of
+stochastic methods described. Although these methods have no symmetry
+requirements for convergence, they do require that the spectral radius
+of the iteration matrix be less than one. Preconditioning serves as a
+means of achieving this by altering the eigenvalue spectrum of the
+iteration matrix. It is possible to use general left, right, and left/right
+preconditioning with MCSA by carefully considering the underlying
+Monte Carlo problem that will be solved with the Neumann-Ulam
+method. We consider here the general left/right preconditioned method
+as the left or right preconditioned methods can be inferred from its
+formulation.
+
+We consider a left preconditioner $\ve{M_L}$ and a right
+preconditioner $\ve{M_R}$. The left/right preconditioned linear
+problem is then:
+\begin{equation}
+  \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\ve{M}_R\ve{x} = \ve{M}_L^{-1}\ve{b}\:.
+  \label{eq:left_right_linear_problem}
+\end{equation}
+To handle the right preconditioning, the system is written with a
+substitution of variables:
+\begin{equation}
+  \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\ve{u} = \ve{M}_L^{-1}\ve{b}\:,
+  \label{eq:left_right_subs_problem}
+\end{equation}
+with
+\begin{equation}
+  \ve{x} = \ve{M}_R^{-1}\ve{u}\:.
+  \label{eq:left_right_recover}
+\end{equation}
+To apply such a method to MCSA, we solve for the substituted variable
+$\ve{u}$ during the iteration sequence:
+\begin{subequations}
+  \begin{gather}
+    \ve{u}^{k+1/2} = \ve{u}^k + \ve{r}^k\:,\\
+    \ve{r}^{k+1/2} = \ve{M}_L^{-1}(\ve{b}-\ve{A}\ve{M}_R^{-1}\ve{u}^{k+1/2})\:,\\ 
+    \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\delta\ve{u}^{k+1/2} = \ve{r}^{k+1/2}\:,\\ 
+    \ve{u}^{k+1} = \ve{u}^{k+1/2} + \delta \ve{u}^{k+1/2}\:,\\
+    \ve{r}^{k+1} = \ve{M}_L^{-1}(\ve{b}-\ve{A}\ve{M}_R^{-1}\ve{u}^{k+1})\:,
+  \end{gather}
+  \label{eq:left_right_mcsa}
+\end{subequations}
+and then recover the original solution vector with
+Eq~(\ref{eq:left_right_recover}). For the Monte Carlo problem, we
+isolate the generation of the correction:
+\begin{equation}
+  \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\delta\ve{u}^{k+1/2} = \ve{r}^{k+1/2}\:,
+  \label{eq:left_right_correction}
+\end{equation}
+and note that the preconditioned residual of the substituted variable
+is now serving as the source and the new iteration matrix is:
+\begin{equation}
+  \ve{H} = \ve{I} - \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\:.
+  \label{eq:left_right_iteration_matrix}
+\end{equation}
+As we require $(i,j)$ element-wise access to the iteration matrix in
+order to construct probabilities and weights for the Monte Carlo
+procedure from the Neumann-Ulam decomposition, the \textit{composite
+  operator}, $\ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}$, must be formed via
+matrix-matrix multiplication.
+
+Several possible shortcomings of this preconditioning approach are
+readily observed. First, the matrix-matrix multiplication operation
+for sparse, parallel distributed matrices is significantly more
+expensive than a matrix-vector multiplication operation. Second, each
+preconditioner must be explicitly inverted, an operation in itself
+that may be expensive and which prohibits the use of any
+preconditioners which provide no mechanism to extract their
+inverse. Third, for many modern preconditioning methods, this
+inversion may yield dense matrices, destroying sparsity and further
+impeding the performance of a matrix-matrix multiplication
+operation. It is also interesting to note that the Monte Carlo problem
+in the general left/right preconditioned scheme given by
+Eq~(\ref{eq:left_right_correction}) is not fully left/right
+preconditioned (meaning that we do not recover $\ve{x}$), but instead
+part of a sequence for finding the substituted variable $\ve{u}$. We
+do, however, gain the benefits of this general preconditioning by
+building the iteration matrix in
+Eq~(\ref{eq:left_right_iteration_matrix}) from the fully
+preconditioned linear operator. In addition, for MCSA to apply to
+increasingly difficult problems, more advanced preconditioning
+techniques that require the generation of the composite operator may
+be necessary for convergence.
+
 \subsection{Preliminary Jacobi Preconditioned Calculations}
@@
 
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{Advanced Preconditioning Strategies}
+\subsection{Algebraic Preconditioning Strategies}
 \label{subsec:spn_advanced_preconditioning}
@@
 $\mathbf{b} \in \mathbb{R}^{N}$ where $||\mathbf{b}||_1 \neq 0$.
-
-\subsection{Sparse Approximate Inverse Preconditioning}
-\label{subsec:spn_spainv_preconditioning}
-Explicit formation of the preconditioner inverse is a requirement of
-the current MCSA preconditioning strategy. As a result, although ILUT
-preconditioning provided excellent iterative performance, a
-significant time penalty is paid to extract the inverse of the
-triangular factors through $N$ triangular solves. In addition, this
-explicit inversion yielded a composite linear operator that was dense,
-requiring modification of the system in order to achieve better
-scalability and Monte Carlo performance. Sparse approximate inverse
-(SPAINV) preconditioning is a technique that may potentially alleviate
-both of these constraints at the cost of reduced iterative
-performance. The method produces a sparse approximation of the inverse
-matrix directly by minimizing the Frobenius norm of the residual
-matrix \cite{saad_iterative_2003}:
-\begin{equation}
-  || \mathbf{I} - \mathbf{A} \mathbf{M} ||^2_F =
-  \sum_{j=1}^N ||\mathbf{e}_j - \mathbf{A} \mathbf{m}_j||^2_2 \:,
-  \label{eq:frobenius_norm_min}
-\end{equation}
-where the inverse preconditioning matrix $\mathbf{M}$ minimizes the
-norm on the right and $\mathbf{e}_j$ and $\mathbf{m}_j$ are the
-$j^{th}$ columns of the identity and preconditioning matrices
-respectively. Here, $\mathbf{M}$ is the actual inverse of the
-preconditioner and is formed directly by the algorithm. On the right
-hand side of Eq~(\ref{eq:frobenius_norm_min}), the Frobenius norm is
-represented as a effective sum of $N$ linear system residual norms.
-
-A few iterations of a subspace method or other iterative scheme can be
-used to effectively solve these systems to a loose tolerance and yield
-the columns of the approximate preconditioning matrix. At each
-iteration, threshold values are used to eliminate the components of
-each $\mathbf{m}_j$ column to maintain the desired level of
-sparsity. In addition, a sparsity pattern can be predetermined and
-enforced during this process. For the SPAINV implementation used for
-this work, the sparsity pattern was defined by levels such that a
-level of $N$ for an input operator $\mathbf{A}$ will generate a
-preconditioning matrix $\mathbf{M}$ with the same sparsity pattern as
-$\mathbf{A}^{N+1}$. By using the norm minimization strategy instead of
-a factorization such as ILUT, parallel results are reproduced
-regardless of parallel problem size, meaning that a serial computation
-will converge in the same number of iterations as a computation with
-thousands of cores.
-
-MCSA preconditioned with SPAINV was used to solve the fuel assembly
-problem as with ILUT for a 1-group $SP_1$ discretization. To study
-convergence sensitivity to SPAINV parameters, the number of levels in
-the sparsity pattern and threshold were parametrically varied with the
-number of iterations to converge a single eigenvalue iteration for the
-fuel assembly problem recorded along with the maximum number of
-non-zero entries observed in all matrix rows for the left/right
-preconditioned composite linear operator. For each calculation, the
-number of iterations required to converge reported was for a single
-eigenvalue iteration with \sn{3}{4} histories at each MCSA iteration
-to compute the Monte Carlo correction using the adjoint collision
-estimator. All SPAINV calculations reported here were performed on a
-single CPU.
-
-Figure~\ref{fig:spainv_iterations} gives the number of MCSA iterations
-needed to converge the fuel assembly problem and
-Figure~\ref{fig:spainv_size} the maximum number of non-zero entries
-per row observed in the composite linear operator as a function the
-SPAINV threshold. For this analysis, the number of levels in the
-sparsity pattern was varied from 3 to 7 with convergence of the fuel
-assembly problem not achieved for smaller level values. Compared to
-ILUT preconditioning, at higher levels we see comparable iterative
-performance with a relative invariance to the threshold value. The
-threshold did have a significant effect, however, on the time required
-to generate the preconditioner with a threshold of 0.001 over an order
-of magnitude slower than a threshold of 0.01 due to the inclusion of a
-significant number of extra values in the input operator. In addition
-to comparable iterative performance, the sparsity of the composite
-operator is greatly improved with nearly an order of magnitude
-reduction in number of non-zero entries per row for lower level
-values.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4.25in]{chapters/spn_equations/spainv_iterations.pdf}
-  \end{center}
-  \caption{\textbf{Number of MCSA iterations required to converge a
-      single eigenvalue iteration for the fuel assembly problem with
-      SPAINV preconditioning as a function of SPAINV threshold.}
-    \textit{Each colored curve represents the iteration behavior for a
-      different SPAINV level pattern. Levels of 3, 4, 5, 6, and 7 were
-      used.}}
-  \label{fig:spainv_iterations}
-\end{figure}
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4.25in]{chapters/spn_equations/spainv_size.pdf}
-  \end{center}
-  \caption{\textbf{Maximum number of non-zero entries observed for all
-      rows in the composite linear operator for the fuel assembly
-      problem with SPAINV preconditioning given as a function of
-      SPAINV threshold.} \textit{Each colored curve represents the row
-      size for a different SPAINV level pattern. Levels of 3, 4, 5, 6,
-      and 7 were used.}}
-  \label{fig:spainv_size}
-\end{figure}
-
-With the comparable iterative performance and improved sparsity,
-SPAINV preconditioning should then have a more favorable quality
-metric than ILUT preconditioning. Figure~\ref{fig:spainv_quality}
-gives the quality metric as a function of the threshold value for each
-of the sparsity levels computed. Not only do we see an improved
-quality metric over ILUT preconditioning (about an order of magnitude
-less), but we also note that the ideal sparsity level is not the largest
-nor the smallest. In fact, the largest and smallest levels (3 and 7)
-performed the worst in terms of the quality metric with a level of 4
-performing the best. When CPU time to generate the preconditioner is
-considered, 4 is the clear choice for this problem as more time is
-required to do the approximate inversion as the sparsity pattern of
-the preconditioner grows in size.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4.25in]{chapters/spn_equations/spainv_quality.pdf}
-  \end{center}
-  \caption{\textbf{SPAINV preconditioning quality metric for the fuel
-      assembly problem given as a function of SPAINV threshold.}
-    \textit{Each colored curve represents the quality metric behavior
-      for a different SPAINV level pattern. Levels of 3, 4, 5, 6, and
-      7 were used.}}
-  \label{fig:spainv_quality}
-\end{figure}
-
-\clearpage
-
-\subsection{SPAINV Improved ILUT Preconditioning}
-\label{subsec:spn_spainv_preconditioning}
-As a means to improve convergence over using only SPAINV
-preconditioning, ILUT preconditioning may be modified through an
-additional application of SPAINV
-(SPAINV-ILUT)\cite{saad_iterative_2003}. For the ILUT preconditioning
-case we have the following linear problem to solve for the substituted
-variable:
-\begin{equation}
-  \mathbf{L}^{-1} \mathbf{A} \mathbf{U}^{-1} \mathbf{u} =
-  \mathbf{L}^{-1} \mathbf{b}\:.
-  \label{eq:ilut_standalone}
-\end{equation}
-We then apply SPAINV and minimize the following problem, this time on
-the left:
-\begin{equation}
-  || \mathbf{I} - \mathbf{M} \mathbf{L}^{-1} \mathbf{A}
-  \mathbf{U}^{-1} ||^2_F\:,
-  \label{eq:ilut_minimize}
-\end{equation}
-giving a new preconditioned system:
-\begin{equation}
-  \mathbf{M} \mathbf{L}^{-1} \mathbf{A} \mathbf{U}^{-1} \mathbf{u} =
-  \mathbf{M} \mathbf{L}^{-1} \mathbf{b}\:.
-  \label{eq:ilut_spainv}
-\end{equation}
-Doing this permits a smaller fill level to be used with ILUT and a
-lower-order sparsity pattern to be used with SPAINV, giving marginally
-more sparsity in the resulting composite linear operator while
-maintaining good iterative performance.
-
-MCSA preconditioned with SPAINV-ILUT was also used to solve the same
-fuel assembly problem as with the other preconditioners for the same
-1-group $SP_1$ discretization. To study convergence sensitivity to
-SPAINV-ILUT parameters, the ILUT drop tolerance and ILUT fill level
-were parametrically varied with the number of iterations to converge a
-single eigenvalue iteration for the fuel assembly problem recorded
-along with the maximum number of non-zero entries observed in all
-matrix rows for the left/right preconditioned composite linear
-operator. SPAINV parameters for the data presented here were fixed as
-it was found that the preconditioning quality was largely insensitive
-to their variation. In addition, the fact that SPAINV is being used to
-preconditioned a problem already preconditioned with ILUT means that
-the input matrix sparsity pattern is already somewhat dense and large
-SPAINV levels would result in an even denser system. Therefore, a
-smaller SPAINV sparsity pattern level will be used for this
-preconditioning. For each calculation, the number of iterations
-required to converge reported was for a single eigenvalue iteration.
-Again, \sn{3}{4} histories are used at each MCSA iteration to compute
-the Monte Carlo correction using the adjoint collision estimator. All
-SPAINV-ILUT calculations reported here were performed on a single CPU.
-
-Figure~\ref{fig:spainv_ilut_iterations} gives the number of iterations
-required to converge as a function of ILUT drop tolerance for ILUT
-fill levels of 1.5, 2.0, 3.0, 4.0 and 5.0 with a fixed SPAINV
-threshold of 1.0 and sparsity level of 1. SPAINV threshold values
-higher than 1.0 resulted in a loss of convergence and lower values did
-not significantly affect performance. Larger sparsity levels resulted
-in a composite operator that was too dense and a prohibitive amount of
-CPU time required to compute the preconditioner. At smaller levels of
-fill, the benefits of using an additional step of SPAINV
-preconditioning are readily observed with the number of iterations
-required to converge cut in half for large drop tolerances. For
-smaller drop tolerances and higher ILUT levels of fill, the additional
-step of SPAINV preconditioning offers marginal improvement to
-iterative performance. As shown in Figure~\ref{fig:spainv_ilut_size},
-because ILUT is used in the preconditioning sequence, sparsity is
-again lost with only marginally improved non-zero entry numbers due to
-the application of the SPAINV threshold. As a result, the quality
-metrics given by Figure~\ref{fig:spainv_ilut_quality} are only
-marginally better than those observed for ILUT preconditioning and
-much larger than the SPAINV quality metric values.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4.25in]{chapters/spn_equations/psilut_iterations.pdf}
-  \end{center}
-  \caption{\textbf{Number of MCSA iterations required to converge a
-      single eigenvalue iteration for the fuel assembly problem with
-      SPAINV-ILUT preconditioning as a function of ILUT drop
-      tolerance.} \textit{Each colored curve represents the row size
-      for a different ILUT fill level. Fill levels of 1.5, 2.0, 3.0,
-      4.0, and 5.0 were used.}}
-  \label{fig:spainv_ilut_iterations}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4.25in]{chapters/spn_equations/psilut_size.pdf}
-  \end{center}
-  \caption{\textbf{Maximum number of non-zero entries observed for all
-      rows in the composite linear operator for the fuel assembly
-      problem with SPAINV-ILUT preconditioning given as a function of
-      ILUT drop tolerance.} \textit{Each colored curve represents the
-      row size for a different ILUT fill level. Fill levels of 1.5,
-      2.0, 3.0, 4.0 and 5.0 were used.}}
- \label{fig:spainv_ilut_size}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4.25in]{chapters/spn_equations/psilut_quality.pdf}
-  \end{center}
-  \caption{\textbf{SPAINV-ILUT preconditioning quality metric for the
-      fuel assembly problem given as a function of ILUT drop
-      tolerance.} \textit{Each colored curve represents the row size
-      for a different ILUT fill level. Fill levels of 1.5, 2.0, 3.0,
-      4.0 and 5.0 were used.}}
-  \label{fig:spainv_ilut_quality}
-\end{figure}
-
-\clearpage
 
@@
 
-\clearpage
-
-\subsection{Monte Carlo Estimator Comparison}
-\label{subsec:spn_estimator_comparison}
-With convergence obtained for the fuel assembly problem, the advanced
-preconditioners will next be studied with both the collision and
-expected value estimators using the best preconditioning and
-relaxation parameter combinations found by the previous
-analysis. Additionally, both the Richardson-based MCSA iteration given
-by Eq~(\ref{eq:mcsa}) and the minimal residual-based MCSA iteration given by
-Eq~(\ref{eq:mcsa_min_res}) will be used. Although the expected value
-estimator achieved the best iterative performance in the previous
-comparison and analysis for the simple Poisson problem in
-\S~\ref{sec:parameter_estimator_analysis}, it is possible that the
-collision estimator may perform better from a timing perspective due
-to the sparsity lost by explicit preconditioning. The deterministic
-component of the expected value estimator that couples the current
-state to other local states through the iteration matrix stencil
-requires application of the estimator to potentially several orders of
-magnitude more states at every transition event.
-
-For this study, the fuel assembly problem with a 1-group $SP_1$
-discretization was solved using MCSA with the adjoint Monte Carlo
-solver using both the collision and expected value estimators and both
-the Richardson and minimal residual fixed point iterations. Using the
-results from the previous section, a Neumann relaxation parameter of
-0.7 was used with both estimators and fixed point iterations. For the
-collision estimator, a Richardson relaxation parameter of 1.1 was used
-with the Richardson iteration while it was found the expected value
-estimator had the best performance when a Richardson relaxation
-parameter of 1.0 was used instead. In addition, ILUT preconditioning
-with a drop tolerance of \sn{1}{-5} and a fill level of 5 was used
-with a reduced domain approximation fill level of 100 and a threshold
-of \sn{1}{-10} as in the previous analysis. Artificial absorption was
-also introduced with a value of 0.2 for the MCSA calculations that
-used the collision estimator as CPU timing performance was improved
-without a loss in iterative performance while no absorption was used
-for calculations using the expected value estimator. For each
-estimator, the number of stochastic histories used to compute the MCSA
-correction was varied from 25 to 50,000 with the number of MCSA
-iterations required to converge to a tolerance of \sn{1}{-8} recorded
-for a single eigenvalue iteration along with the CPU time required to
-achieve convergence.
-
-Figure~\ref{fig:spn_estimator_iters} gives the iteration count results
-and Figure~\ref{fig:spn_estimator_time} gives the timing
-results. Remarkably, although the fuel assembly problem is
-significantly more complicated than the simple Poisson problem studied
-in \S~\ref{sec:parameter_estimator_analysis}, the results here are
-largely the same with Figure~\ref{fig:spn_estimator_iters} and
-Figure~\ref{fig:estimator_nh_iters} having identical qualitative
-behavior. Using the expected value estimator with both fixed point
-iterations creates a relative insensitivity to the number of histories
-used to compute the correction. Marginally better iterative and timing
-performance was observed for the expected value estimator when using
-the Richardson iteration. Better timing performance is expected as
-computing the extrapolation parameter in the minimal residual
-iteration requires several additional parallel operations. However, we
-expect that because the minimal residual iteration will converge
-faster on its own when compared to the Richardson iteration that
-better iterative performance will be observed when it is accelerated
-with MCSA. The fact that this was not observed signals that an MCSA
-scheme leveraging the expected value estimator is also effectively
-insensitive to the fixed point iteration used. For the collision
-estimator, the minimal residual iteration permits fewer stochastic
-histories to be used with each MCSA iteration while still maintaining
-good iterative performance. With respect to CPU time, there is a more
-pronounced effect due to the larger density of states created by the
-explicit ILUT preconditioning. Using a reduced domain fill level of
-100 creates a situation where the expected value estimator is
-significantly more expensive per history to compute that the collision
-estimator due to the coupling of states.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/estimator_iters.pdf}
-  \end{center}
-  \caption{\textbf{MCSA iterations required to converge the fuel
-      assembly problem to a tolerance of \sn{1}{-8}.} \textit{Both the
-      collision and expected value estimators were used with the
-      adjoint Monte Carlo solver to compute the correction at each
-      iteration. Each estimator was used with the Richardson and
-      minimal residual fixed point iterations within an MCSA
-      iteration.}}
-  \label{fig:spn_estimator_iters}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/estimator_time.pdf}
-  \end{center}
-  \caption{\textbf{Total CPU time in seconds required to converge the
-      fuel assembly problem to a tolerance of \sn{1}{-8}.}
-    \textit{Both the collision and expected value estimators were used
-      with the adjoint Monte Carlo solver to compute the correction at
-      each iteration. Each estimator was used with the Richardson and
-      minimal residual fixed point iterations within an MCSA
-      iteration.}}
-  \label{fig:spn_estimator_time}
-\end{figure}
-
-From both Figure~\ref{fig:spn_estimator_iters} and
-Figure~\ref{fig:spn_estimator_time} we see that both estimators and
-fixed point iterations have comparable iterative and timing
-performance when 1,000 histories are used at each MCSA
-iteration. Using 1,000 histories per MCSA iteration, the infinity norm
-of the residual vector is presented at every MCSA iteration for both
-estimators and both fixed point iterations in
-Figure~\ref{fig:spn_estimator_convergence} as an additional means of
-comparison. At this number of histories, all MCSA combinations
-converge the fuel assembly problem monotonically at and approximately
-the same rate, giving little reason to choose one over the other. The
-end result is that both estimators provide a viable method for getting
-good MCSA iterative performance with comparable timing
-performance. For the collision estimator, using around 20,000
-stochastic histories at every iteration gave the best iterative
-performance while 500 histories per iteration gave the best iterative
-performance when using the expected value estimator. For purely serial
-computational performance, there is not a distinguishing feature for
-the fuel assembly problem presented that would cause one to choose one
-of the estimators and fixed point iterations over the other as they
-all have similar minimum CPU times over the range of values tested.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/estimator_convergence.pdf}
-  \end{center}
-  \caption{\textbf{Residual infinity norm as a function of MCSA
-      iteration with 1,000 stochastic histories per iteration.}
-    \textit{Both the collision and expected value estimators were used
-      with the adjoint Monte Carlo solver to compute the correction at
-      each iteration. Each estimator was used with the Richardson and
-      minimal residual fixed point iterations within an MCSA
-      iteration.}}
-  \label{fig:spn_estimator_convergence}
-\end{figure}
-
-\clearpage
-
 %%---------------------------------------------------------------------------%%
-\section{MCSA Verification\ }
-\label{sec:spn_mcsa_verification}
+\section{Results}
+
 Through the numerical studies presented in this chapter, we have
@@
 
-In order to verify the correctness of the MCSA method and
-implementation used for this work, MCSA solutions to the fuel assembly
-problem with 1, 2, and 4 energy groups using $SP_1$ discretization
-will be directly compared over all group and cell fluxes computed by
-converging the criticality problem using both BiCGStab and GMRES
-production implementations from the Trilinos scientific computing
-library Aztec \cite{heroux_overview_2005}. In addition to comparing
-the cell and group fluxes, the number of eigenvalue iterations
-required to converge the problem along with the eigenvalue itself will
-be compared as a means of verification. If all values computed using
-MCSA agree with the results from the production Krylov solvers within
-the solution tolerance, then the MCSA solutions will be deemed correct
-for these problems.
-
-For the verification calculations, all solvers will be preconditioned
+For the comparison calculations, all solvers will be preconditioned
 with ILUT using a drop tolerance of \sn{1}{-5} and a fill level of
@@
 group 0).
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{ll}\hline\hline
-      \multicolumn{1}{l}{\textbf{Name}} & 
-      \multicolumn{1}{l}{\textbf{Definition}} \\
-      BiCGStab-ILUT & BiCGStab preconditioned with ILUT \\
-      GMRES-ILUT & GMRES preconditioned with ILUT \\
-      MCSA-ILUT-R-C & MCSA preconditioned with ILUT using Richardson \\ 
-                    & fixed point iteration and collision estimator \\
-      MCSA-ILUT-MR-C & MCSA preconditioned with ILUT using minimal residual \\
-                     & fixed point iteration and collision estimator \\
-      MCSA-ILUT-R-EV & MCSA preconditioned with ILUT using Richardson \\
-                     & fixed point iteration and expected value estimator \\
-      MCSA-ILUT-MR-EV & MCSA preconditioned with ILUT using minimal residual \\
-                      & fixed point iteration and expected value estimator \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Solver definitions used for MCSA verification and
-      performance analysis.} \textit{The ILUT preconditioner
-      parameters were identical for all calculations and solvers.}}
-  \label{tab:spn_solver_defs}
-\end{table}
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{cl}\hline\hline
-      \multicolumn{1}{c}{\textbf{Number of Groups}} & 
-      \multicolumn{1}{l}{\textbf{Lower Bounds (eV)}} \\
-      1 & \{ \sn{1}{-5} \} \\
-      2 & \{ \sn{1}{-1}, \sn{1}{-5} \} \\
-      4 & \{ \sn{1}{1}, \sn{1}{0}, \sn{1}{-1}, \sn{1}{-5} \} \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Energy group lower bounds for the multigroup fuel
-      assembly criticality problem in electron volts.} \textit{An
-      implicit upper bound of \sn{2}{6} eV is assumed for group zero.}}
-  \label{tab:spn_group_structure}
-\end{table}
-
-For every combination of energy group structure and solver, the
-eigenvalue problem was converged with an eigensolver tolerance of
-\sn{1}{-6} and a linear solver tolerance of \sn{1}{-8} in a serial
-calculation. Table~\ref{tab:serial_ev_results} gives the eigenvalue
-computed by each solver for each problem as well as the number of
-eigenvalue iterations required to converge. For every problem, each
-solver converged to the same eigenvalue within the tolerance of the
-eigensolver in the same number of eigenvalue iterations.
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lccc}\hline\hline
-      \multicolumn{1}{c}{\textbf{Solver}} & 
-      \multicolumn{1}{c}{\textbf{Number of Groups}} & 
-      \multicolumn{1}{c}{\textbf{Iterations}} &
-      \multicolumn{1}{c}{\textbf{Eigenvalue}} \\
-      \hline
-      BiCGStab-ILUT & 1 & 25 & 1.1059870 \\
-      GMRES-ILUT & 1 & 25 & 1.1059870 \\
-      MCSA-ILUT-R-C & 1 & 25 & 1.1059870 \\
-      MCSA-ILUT-MR-C & 1 & 25 & 1.1059870 \\
-      MCSA-ILUT-R-EV & 1 & 25 & 1.1059870 \\
-      MCSA-ILUT-MR-EV & 1 & 25 & 1.1059870 \\
-      \hline
-      BiCGStab-ILUT & 2 & 35 & 1.1552390 \\
-      GMRES-ILUT & 2 & 35 & 1.1552390 \\
-      MCSA-ILUT-R-C & 2 & 35 & 1.1552390 \\
-      MCSA-ILUT-MR-C & 2 & 35 & 1.1552390 \\
-      MCSA-ILUT-R-EV & 2 & 35 & 1.1552390 \\
-      MCSA-ILUT-MR-EV & 2 & 35 & 1.1552390 \\
-      \hline
-      BiCGStab-ILUT & 4 & 31 & 1.1758500 \\
-      GMRES-ILUT & 4 & 31 & 1.1758500 \\
-      MCSA-ILUT-R-C & 4 & 31 & 1.1758500 \\
-      MCSA-ILUT-MR-C & 4 & 31 & 1.1758500 \\
-      MCSA-ILUT-R-EV & 4 & 31 & 1.1758500 \\
-      MCSA-ILUT-MR-EV & 4 & 31 & 1.1758500 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Serial eigensolver verification results for the
-      fuel assembly problem.} \textit{Each calculation was converged
-      with the same eigensolver and linear solver
-      tolerances. Table~\ref{tab:spn_solver_defs} gives the
-      description for each solver type presented.}}
-  \label{tab:serial_ev_results}
-\end{table}
-
-For each converged eigenvalue calculation, the scalar flux in each
-energy group was compared over the entire spatial domain with the
-BiCGStab-ILUT solution used as the reference to compute an absolute
-difference of the flux in each mesh cell. For the 1 group
-calculations, these absolute differences are plotted spatially along
-the center-line of the fuel assembly in
-Figure~\ref{fig:serial_1_group_diff}. Interestingly, except for the
-MCSA-ILUT-MR-C calculation, all other 1-group calculations including
-GMRES had a maximum global flux difference of \sn{9.313}{-10} even
-though different spatial results were observed for all cases. For the
-MCSA-ILUT-MR-C calculation, a slightly larger maximum error of
-\sn{2.205}{-9} was observed for reasons that remain unexplained. All
-1-group calculations have a minimum observed absolute difference of 0
-with respect to the BiCGStab calculation.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/serial_1_group.pdf}
-  \end{center}
-  \caption{\textbf{Absolute difference of the group 0 scalar flux
-      between the Aztec BiCGStab calculations and Aztec GMRES and the
-      MCSA variations for the 1-group fuel assembly criticality
-      calculation.} \textit{Left to right: GMRES-ILUT, MCSA-ILUT-R-C,
-      MCSA-ILUT-MR-C, MCSA-ILUT-R-EV, MCSA-ILUT-MR-EV. The fluxes are
-      plotted on a plane intersecting the center-line of the
-      assembly. Table~\ref{tab:spn_solver_defs} gives the description
-      for each solver type presented.}}
-  \label{fig:serial_1_group_diff}
-\end{figure}
-
-Table~\ref{tab:serial_differences_g1},
-Table~\ref{tab:serial_differences_g2}, and
-Table~\ref{tab:serial_differences_g4} give the maximum and minimum
-absolute differences of the cell scalar fluxes for all calculations
-and all energy groups using the BiCGStab calculation as the
-reference. All comparisons were performed within a group such that the
-differences for the MCSA-ILUT-R-C results in group 3 for the 4-group
-calculation were computed with the BiCGStab-ILUT solution in group 3
-for the 4-group calculation as a reference. For all group flux results
-given in the table, the maximum differences for all cases was less
-than the linear solver tolerance used to converge the flux at each
-eigenvalue iteration meaning that all values over the entire energy
-and spatial domain are within these limits. In addition, the minimum
-error for all calculations was observed to be zero, meaning that at
-least one mesh cell in all calculations for all energy groups the
-scalar flux values computed by all solvers were identical. Based on
-the flux results and the eigenvalue results presented here, the MCSA
-results should be deemed correct for the fuel assembly problem for
-serial calculations with multiple energy groups.
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lccc}\hline\hline
-      \multicolumn{1}{c}{\textbf{Solver}} & 
-      \multicolumn{1}{c}{\textbf{Group}} &
-      \multicolumn{1}{c}{\textbf{Minimum Difference}} &
-      \multicolumn{1}{c}{\textbf{Maximum Difference}} \\
-      \hline
-      GMRES-ILUT & 0 & 0.0 & \sn{9.313}{-10} \\
-      MCSA-ILUT-R-C & 0 & 0.0 & \sn{9.313}{-10} \\
-      MCSA-ILUT-MR-C & 0 & 0.0 & \sn{2.205}{-9} \\
-      MCSA-ILUT-R-EV & 0 & 0.0 & \sn{9.313}{-10} \\
-      MCSA-ILUT-MR-EV & 0 & 0.0 & \sn{9.313}{-10} \\
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Serial scalar flux verification results for the
-      1-group fuel assembly problem.} \textit{The maximum and minimum
-      absolute differences over the entire spatial domain are
-      presented with BiCGStab used as the reference. Each calculation
-      was converged with the same eigensolver and linear solver
-      tolerances. Table~\ref{tab:spn_solver_defs} gives the
-      description for each solver type presented.}}
-  \label{tab:serial_differences_g1}
-\end{table}
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lccc}\hline\hline
-      \multicolumn{1}{c}{\textbf{Solver}} & 
-      \multicolumn{1}{c}{\textbf{Group}} &
-      \multicolumn{1}{c}{\textbf{Minimum Difference}} &
-      \multicolumn{1}{c}{\textbf{Maximum Difference}} \\
-      \hline
-      GMRES-ILUT & 0 & 0.0 & \sn{2.794}{-9} \\
-      MCSA-ILUT-R-C & 0 & 0.0 & \sn{9.313}{-10} \\
-      MCSA-ILUT-MR-C & 0 & 0.0 & \sn{9.313}{-10} \\
-      MCSA-ILUT-R-EV & 0 & 0.0 & \sn{9.313}{-10} \\
-      MCSA-ILUT-MR-EV & 0 & 0.0 & \sn{9.313}{-10} \\
-      \hline
-      GMRES-ILUT & 1 & 0.0 & \sn{4.657}{-10} \\
-      MCSA-ILUT-R-C & 1 & 0.0 & \sn{1.364}{-10} \\
-      MCSA-ILUT-MR-C & 1 & 0.0 & \sn{1.261}{-9} \\
-      MCSA-ILUT-R-EV & 1 & 0.0 & \sn{2.474}{-10} \\
-      MCSA-ILUT-MR-EV & 1 & 0.0 & \sn{5.857}{-10} \\
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Serial scalar flux verification results for the
-      2-group fuel assembly problem.} \textit{The maximum and minimum
-      absolute differences over the entire spatial domain are
-      presented with BiCGStab used as the reference. Each calculation
-      was converged with the same eigensolver and linear solver
-      tolerances. Table~\ref{tab:spn_solver_defs} gives the
-      description for each solver type presented.}}
-  \label{tab:serial_differences_g2}
-\end{table}
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lccc}\hline\hline
-      \multicolumn{1}{c}{\textbf{Solver}} & 
-      \multicolumn{1}{c}{\textbf{Group}} &
-      \multicolumn{1}{c}{\textbf{Minimum Difference}} &
-      \multicolumn{1}{c}{\textbf{Maximum Difference}} \\
-      \hline
-      GMRES-ILUT & 0 & 0.0 & \sn{9.313}{-10} \\
-      MCSA-ILUT-R-C & 0 & 0.0 & \sn{1.397}{-9} \\
-      MCSA-ILUT-MR-C & 0 & 0.0 & \sn{1.870}{-9} \\
-      MCSA-ILUT-R-EV & 0 & 0.0 & \sn{1.397}{-9} \\
-      MCSA-ILUT-MR-EV & 0 & 0.0 & \sn{1.869}{-9} \\
-      \hline
-      GMRES-ILUT & 1 & 0.0 & \sn{5.821}{-11} \\
-      MCSA-ILUT-R-C & 1 & 0.0 & \sn{1.164}{-10} \\
-      MCSA-ILUT-MR-C & 1 & 0.0 & \sn{1.164}{-10} \\
-      MCSA-ILUT-R-EV & 1 & 0.0 & \sn{1.164}{-10} \\
-      MCSA-ILUT-MR-EV & 1 & 0.0 & \sn{1.164}{-10} \\
-      \hline
-      GMRES-ILUT & 2 & 0.0 & \sn{1.164}{-10} \\
-      MCSA-ILUT-R-C & 2 & 0.0 & \sn{2.328}{-10} \\
-      MCSA-ILUT-MR-C & 2 & 0.0 & \sn{2.328}{-10} \\
-      MCSA-ILUT-R-EV & 2 & 0.0 & \sn{2.328}{-10} \\
-      MCSA-ILUT-MR-EV & 2 & 0.0 & \sn{2.328}{-10} \\
-      \hline
-      GMRES-ILUT & 3 & 0.0 & \sn{1.164}{-10} \\
-      MCSA-ILUT-R-C & 3 & 0.0 & \sn{1.746}{-10} \\
-      MCSA-ILUT-MR-C & 3 & 0.0 & \sn{2.256}{-9} \\
-      MCSA-ILUT-R-EV & 3 & 0.0 & \sn{2.801}{-10} \\
-      MCSA-ILUT-MR-EV & 3 & 0.0 & \sn{2.256}{-10} \\
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Serial scalar flux verification results for the
-      4-group fuel assembly problem.} \textit{The maximum and minimum
-      absolute differences over the entire spatial domain are
-      presented with BiCGStab used as the reference. Each calculation
-      was converged with the same eigensolver and linear solver
-      tolerances. Table~\ref{tab:spn_solver_defs} gives the
-      description for each solver type presented.}}
-  \label{tab:serial_differences_g4}
-\end{table}
-
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{MCSA Performance Comparison to Conventional Methods\ }
-\label{sec:spn_comparison}
-In the previous section, numerical results using MCSA for the fuel
-assembly calculation were directly compared with production Krylov
-methods in order to verify their correctness. Deemed correct,
-performance of the algorithm for serial computations will now be
-addressed. With the same set of solver parameters used for
-verification, MCSA will now be directly compared to the same
-production Krylov solvers in terms of both iterative performance and
-CPU timing for solutions to the fuel assembly problem.  This
-performance comparison is necessary in order to assess the feasibility
-of the method and identify areas that require further research and
-development.
 
@@
 
-\clearpage
-
 %%---------------------------------------------------------------------------%%
-\section{Summary\ }
-\label{sec:mc_summary}
-
-In this chapter, the Monte Carlo Synthetic Acceleration method has
-been applied to the $SP_N$ discretization of the neutron transport
-equation and explored in the context of a difficult nuclear fuel
-assembly criticality problem. The following are the significant
-observations and findings.
-
-\begin{itemize}
-\item MCSA can solve the asymmetric system generated by the $SP_N$
-  equations
-\item MCSA has been incorporated into the Exnihilo neutronics
-  production code base developed at Oak Ridge National Laboratory
-\item Simple Jacobi-based preconditioning reduces the spectral radius
-  of the $SP_N$ system below unity, however, this alone is not
-  sufficient for MCSA convergence
-\item As the spectral radius approaches unity, MCSA was demonstrated
-  to break down with more stochastic histories required for
-  convergence and more CPU time required per history
-\item Light water reactor problems are difficult to solve with MCSA as
-  they have large spectral radii due to the neutron scattering in the
-  moderator
-\item Advanced algebraic preconditioning strategies were applied to
-  the $SP_N$ equations to obtain convergence with ILUT chosen for
-  subsequent investigations
-\item The reduced domain approximation was required to mitigate the
-  memory and performance constraints of the explicit preconditioning
-  strategy although both remain significant obstacles
-\item MCSA relaxation parameters were developed and observed to
-  enhance the time to solution for the fuel assembly criticality
-  problem
-\item MCSA was verified to produce the same flux distribution and
-  k-eigenvalue for the fuel assembly as production Krylov methods
-\item MCSA was observed to converge in fewer iterations per eigenvalue
-  iteration than GMRES for the fuel assembly criticality problem and
-  more than Bi-CGStab using the same preconditioning
-\item The explicit preconditioning strategy required to overcome the
-  MCSA spectral radius restriction creates runtimes $O(100)$ larger
-  than the production Krylov methods for the fuel assembly criticality
-  problem
-\end{itemize}
+\section{Conclusion}
+
+%%---------------------------------------------------------------------------%%
+\pagebreak
+\bibliographystyle{ieeetr} \bibliography{references}
+\end{document}

### 2013-07-30 doc/ttsp2013/ttsp_2013.tex
adding ttsp makefile

@@
 \usepackage{amssymb}
+\usepackage{amsmath}
+\usepackage{amsthm}
 \usepackage{graphicx}
+\usepackage{tmadd,tmath}
 \usepackage{longtable}
@@
 \usepackage{amsmath}
+\usepackage{algpseudocode}
+\usepackage{algorithm}
+\usepackage[mathcal]{euscript} 
 \usepackage[usenames]{color}
@@
   \begin{center}
-    \includegraphics[width=4in]{chapters/spn_equations/problem3_radial_mat.png}
+    \includegraphics[width=4in]{problem3_radial_mat.png}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=6.0in]{chapters/spn_equations/problem3_axial_mat.png}
+    \includegraphics[width=6.0in]{problem3_axial_mat.png}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=6.0in]{chapters/spn_equations/problem3_end.png}
+    \includegraphics[width=6.0in]{problem3_end.png}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=5in]{chapters/spn_equations/block_jacobi_res.pdf}
+    \includegraphics[width=5in]{block_jacobi_res.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=5in]{chapters/spn_equations/block_jacobi_rich_res.pdf}
+    \includegraphics[width=5in]{block_jacobi_rich_res.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=4.25in]{chapters/spn_equations/breakdown_iterations.pdf}
+    \includegraphics[width=4.25in]{breakdown_iterations.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=4.25in]{chapters/spn_equations/breakdown_histories.pdf}
+    \includegraphics[width=4.25in]{breakdown_histories.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=4.25in]{chapters/spn_equations/breakdown_time.pdf}
+    \includegraphics[width=4.25in]{breakdown_time.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=4.25in]{chapters/spn_equations/ilut_iterations.pdf}
+    \includegraphics[width=4.25in]{ilut_iterations.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=4.25in]{chapters/spn_equations/ilut_size.pdf}
+    \includegraphics[width=4.25in]{ilut_size.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=4.25in]{chapters/spn_equations/ilut_quality.pdf}
+    \includegraphics[width=4.25in]{ilut_quality.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=4.25in]{chapters/spn_equations/rda_iterations.pdf}
+    \includegraphics[width=4.25in]{rda_iterations.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=4.25in]{chapters/spn_equations/rda_quality.pdf}
+    \includegraphics[width=4.25in]{rda_quality.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/relax_iters.pdf}
+    \includegraphics[width=6in]{relax_iters.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/relax_time.pdf}
+    \includegraphics[width=6in]{relax_time.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=5in]{chapters/spn_equations/solver_iters.pdf}
+    \includegraphics[width=5in]{solver_iters.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=5in]{chapters/spn_equations/solver_time.pdf}
+    \includegraphics[width=5in]{solver_time.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=5in]{chapters/spn_equations/solver_p_time.pdf}
+    \includegraphics[width=5in]{solver_p_time.pdf}
   \end{center}

### 2013-07-31 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
fixing indexing in spn fission matrix
updating spn performance comparison plots with color coded results

@@
   \mathbb{A}_{nm} \mathbb{U}_m = \frac{1}{k} \sum_{m=1}^4
-  \mathbb{F}_{nm} \mathbb{U}_n\ \ \ \ \ \ \ n = 1,2,3,4\:.
+  \mathbb{F}_{nm} \mathbb{U}_m\ \ \ \ \ \ \ n = 1,2,3,4\:.
   \label{eq:spn_fission_matrix}
@@
     \textit{Table~\ref{tab:spn_solver_defs} gives the description for
-      each solver type presented in the legend.}}
+      each solver type presented in the legend with the Krylov solvers
+      represented in black and the MCSA solvers represented in red.}}
   \label{fig:spn_comparison_iterations}
@@
       average. Table~\ref{tab:spn_solver_defs} gives the description
-      for each solver type presented in the legend.}}
+      for each solver type presented in the legend with the Krylov
+      solvers represented in black and the MCSA solvers represented in
+      red.}}
   \label{fig:spn_comparison_time}
@@
       for each solver type presented in the legend. Data labeled
-      starting with PMCSA is identical to those labeled with MCSA
-      except that they additionally include the cost of generating the
-      inverse of the preconditioners and composite linear operator.}}
+      starting with PMCSA and shown in blue is identical to those
+      labeled with MCSA except that they additionally include the cost
+      of generating the inverse of the preconditioners and composite
+      linear operator.}}
   \label{fig:spn_comparison_prec_time}

### 2013-08-06 doc/dissertation/document/chapters/introduction/introduction.tex
cleaning up the abstract and intro and formatting the references

@@
 To achieve these goals, three complementary research and development
-activities are performed. First, Monte Carlo Sythetic Acceleration is
-applied to the neutron transport problem in a light water reactor
-calculation to develop the necessary preconditioning techniques and
-understanding required to solve difficult problems. Second, using
-these new techniques, a series of benchmarks for the Navier-Stokes
-equations are solved in different flow regimes to research the
-performance of the Monte Carlo algorithm for nonlinear problems using
-the new Forward-Automated Newton-MCSA algorithm developed by this
-work. Third, a novel parallel algorithm based on domain decomposed
-Monte Carlo methods for particle transport is developed to fully
-parallelize the solution techniques developed to solve the neutron
-transport and fluid flow problems on leadership class hardware with
-scaling studies performed using a neutron diffusion problem.
+activities are performed. First, Monte Carlo Sythetic Acceleration
+(MCSA) is applied to the neutron transport problem in a light water
+reactor calculation to develop the necessary preconditioning
+techniques and understanding required to solve difficult
+problems. Second, using these new techniques, a series of benchmarks
+for the Navier-Stokes equations are solved in different flow regimes
+to research the performance of the Monte Carlo algorithm for nonlinear
+problems using the new Forward-Automated Newton-MCSA algorithm
+developed by this work. Third, a novel parallel algorithm based on
+domain decomposed Monte Carlo methods for particle transport is
+developed to fully parallelize the solution techniques developed to
+solve the neutron transport and fluid flow problems on leadership
+class hardware with scaling studies performed using a neutron
+diffusion problem.
 
@@
 Monte Carlo Synthetic Acceleration methods and studying their
-application to transport systems by providing problems of interest in
-nuclear reactor analysis. Hardware-based motivations are also provided
-by considering the potential impact of these methods on forthcoming
-computing architectures. In addition, background on the current solver
-techniques for these physics problems and a brief comparison to the
-proposed methods are provided to further motivate this work. Finally,
-the statement of work is given and the organization of the document is
-then reviewed.
+application to transport phenomena by providing problems of interest
+in nuclear reactor analysis. Hardware-based motivation is also
+provided by considering the potential impact of these methods on
+forthcoming high performance computing architectures. In addition,
+background on the current solver techniques for these physics problems
+and a brief comparison to the proposed methods are provided to further
+motivate this work. Finally, the statement of work is given and the
+organization of the document is then reviewed.
 
@@
 primarily due to Larsen and his colleagues, the $SP_N$ equations have
-been given a more rigorous treatment with both variational and
-asymptotic derivations performed as a means of verification. In
-addition, these equations have been more rigorously studied as
-solution methods to MOX fuel problems and have been shown to provide
-accurate solutions. With this mathematical literature to provide a
-solid numerical footing for the method, we look at its application to
-today's challenge problems in neutron transport for fission reactor
-analysis. The reduction in numerical complexity of current
-deterministic solution methods using the $S_N$ approximation could
-mean significant savings in both compute time and memory required. In
-addition, the characteristics of the solution to the transport problem
-for a steady state reactor core permit diffusion theory to be used; a
-staple of the nuclear industry since its inception. Therefore, if
-diffusion theory is applicable, then finer grained solutions that
-capture more of the physics contained in the transport equation should
-be possible with the $SP_N$ method. In doing so, we also expect from
-the literature to obtain computed responses on the order of accuracy
-we would expect from an appropriately discretized $S_N$ method at a
-fraction of the cost.
+been given a more rigorous treatment in many forms with both
+variational and asymptotic derivations performed as a means of
+verification \cite{olbrant_asymptotic_2013}. In addition, these
+equations have been more rigorously studied as solution methods to MOX
+fuel problems and have been shown to provide accurate solutions
+\cite{brantley_simplified_2000}. With this mathematical literature to
+provide a solid numerical footing for the method, we look at its
+application to today's challenge problems in neutron transport for
+fission reactor analysis. The reduction in numerical complexity of
+current deterministic solution methods using the $S_N$ approximation
+could mean significant savings in both compute time and memory
+required. In addition, the characteristics of the solution to the
+transport problem for a steady state reactor core permit diffusion
+theory to be used; a staple of the nuclear industry since its
+inception. Therefore, if diffusion theory is applicable, then finer
+grained solutions that capture more of the physics contained in the
+transport equation should be possible with the $SP_N$ method. In doing
+so, we also expect from the literature to obtain computed responses on
+the order of accuracy we would expect from an appropriately
+discretized $S_N$ method at a fraction of the cost.
 
@@
 system of equations for detailed neutronics models of fission
-reactor-based systems\cite{evans_simplified_2013}. By fully forming these
-equations and formulating them as a linear algebra problem instead of
-using the explicit iterative methods of the past, we now have access
-to all of the modern advancements in computational linear algebra
-including Krylov solvers for asymmetric systems and algebraic
+reactor-based systems \cite{evans_simplified_2013}. By fully forming
+these equations and formulating them as a linear algebra problem
+instead of using the explicit iterative methods of the past, we now
+have access to all of the modern advancements in computational linear
+algebra including Krylov solvers for asymmetric systems and algebraic
 preconditioning methods.
@@
 are predicted to be common, resilient solver technologies are required
-to overcome these events. With linear and nonlinear solvers based on
-Monte Carlo techniques, such issues are alleviated by statistical
-arguments. In the case of soft failures, isolated floating point
-errors in Monte Carlo simulation are absorbed within tally statistics
-while completely losing hardware during a hard failure is manifested
-as a high variance event where some portion of the Monte Carlo
-histories are lost. These stochastic methods are a paradigm shift from
-current deterministic solver techniques that will suffer greatly from
-the non-deterministic behavior expected from the next generation of
-machines.
+to overcome these events at the application level. With linear and
+nonlinear solution schemes based on Monte Carlo techniques, such
+issues are alleviated by statistical arguments. In the case of soft
+failures, isolated floating point errors in Monte Carlo simulation are
+absorbed within tally statistics while completely losing hardware
+during a hard failure is manifested as a high variance event where
+some portion of the Monte Carlo histories are lost. These stochastic
+methods are a paradigm shift from current deterministic solver
+techniques that will in their current form potentially suffer greatly
+from the non-deterministic behavior expected from the next generation
+of machines.
 
@@
 Jacobian is never formed, simplifying the implementation, and a Krylov
-solver is leveraged (typically GMRES or Conjugate Gradient), providing
-excellent convergence properties for well-conditioned and well-scaled
-systems. However, there are two potential drawbacks to these methods
-for high fidelity predictive simulations: (1) the Jacobian is
-approximated by a first-order differencing method on the order of
-machine precision such that this error can grow beyond that of those
-in a fine-grained system \cite{kelley_iterative_1995} and (2) for
-systems that are not symmetric positive-definite (which will be the
-case for most multiphysics systems and certainly for most
-preconditioned systems) the Krylov subspace generated by the GMRES
-solver may become prohibitively large \cite{knoll_newton-krylov_1995}.
+solver is leveraged (typically GMRES or a method based on Conjugate
+Gradient), providing excellent convergence properties for
+well-conditioned and well-scaled systems. However, there are two
+potential drawbacks to these methods for high fidelity predictive
+simulations: (1) the Jacobian is approximated by a first-order
+differencing method on the order of machine precision such that this
+error can grow beyond that of those in a fine-grained system
+\cite{kelley_iterative_1995} and (2) for systems that are not
+symmetric positive-definite (which will be the case for most
+multiphysics systems and certainly for most preconditioned systems)
+the Krylov subspace generated by the GMRES solver may become
+prohibitively large \cite{knoll_newton-krylov_1995}.
 
-To address these issues, the Forward-Automated Newton-MCSA (FANM)
-method is developed as new nonlinear solution method. The key features
-of FANM are: full Jacobian generation using modern Forward Automated
+In this work, the Forward-Automated Newton-MCSA (FANM) method is
+developed as new nonlinear solution method. The key features of FANM
+are: full Jacobian generation using modern Forward Automated
 Differentiation (FAD) methods \cite{bartlett_automatic_2006}, and MCSA
@@
 explicitly formed to use MCSA, for problems that take more than a few
-GMRES iterations to converge the size of the Krylov subspace will grow
-beyond that of the Jacobian. Finally, using MCSA to solve the linear
-model provides its benefits for preconditioning, potential parallelism
-and performance.
+GMRES iterations to converge the size of the Krylov subspace will
+typically grow beyond that of the Jacobian. Finally, using MCSA to
+solve the linear model provides its benefits for preconditioning,
+potential parallelism, and performance.
 
@@
 decomposed Monte Carlo Synthetic Acceleration methods. Improvements in
-performance and scalability will be demonstrated in a wide variety of
-numerical experiments using both simple model neutron transport
-problems and neutron transport and fluid flow problems generated by
-production physics codes in both simple and complex geometries.
+performance and parallel scalability will be demonstrated in a wide
+variety of numerical experiments using both simple model neutron
+transport problems and neutron transport and fluid flow problems
+generated by production physics codes in both simple and complex
+geometries.
 
@@
 neutron transport equation is developed and solved using the Monte
-Carlo methods. Using a difficult light water fission reactor fuel
-assembly criticality calculation to drive research and development for
-these Monte Carlo methods, several important issues regarding Monte
-Carlo solver applicability, performance, and preconditioning will be
+Carlo methods. Using a difficult light water reactor fuel assembly
+criticality calculation to drive research and development for these
+Monte Carlo methods, several important issues regarding Monte Carlo
+solver applicability, performance, and preconditioning will be
 discussed along with solutions to these problems. Using these

### 2013-08-06 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
cleaning up the abstract and intro and formatting the references

@@
 Carlo, this stagnation manifests itself in histories that simply do no
-terminate and therefore no solution is acquired.
+terminate and therefore no solution is acquired. For many of the
+problems presented here, it is likely that a more physics-based
+preconditioning strategy such as those presented in
+\cite{evans_development_2006,evans_enhanced_2007} will yield better
+results for more difficult problems. However, the explicit
+preconditioning strategy used here currently limits us to algebraic
+preconditioning.
 

### 2013-08-06 doc/dissertation/document/frontmatter/abstract.tex
cleaning up the abstract and intro and formatting the references

@@
 computing hardware and algorithms that leverage this hardware are
-required to help meet the goals and requirements of engineering
+needed to help meet the goals and requirements of engineering
 analysis. Current practices in neutron transport and fluid flow
@@
 water reactor fuel assembly used as the driving problem. It was
-discovered that basic Jacobi-based preconditioning was not sufficient
-to achieve convergence due to the dominating effect of the light water
-moderator and therefore a group of modern preconditioning strategies
-are researched. The new preconditioning techniques, however, create
+discovered that basic Jacobi-based preconditioning as used in previous
+work was not sufficient to achieve convergence due to the dominating
+effect of the light water moderator. To enable MCSA as a solution
+technique a group of modern preconditioning strategies are
+researched. The new preconditioning techniques, however, create
 unwanted growth in both memory consumption and computation time for
@@
 components of this work are required to enable future simulation of
-nonlinear physics problems on the next generation of hardware. To
+nonlinear transport phenomena on the next generation of hardware. To
 enable high performance simulations with FANM, we require the MCSA

### 2013-08-08 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
adding Newton-Richardson comparison

@@
 performance of the FANM method against a conventional Newton-Krylov
-method. For each benchmark, we will analyze the iterative performance
-of both the nonlinear Newton solver and the linear solver used to
-compute the Newton correction vector at each step. Limiting the number
-of nonlinear iterations required for solution is critical for
-efficiency as nonlinear function evaluations, Jacobian generation, and
-preconditioner construction are typically required at every Newton
-iteration. In addition, timing results will also be discussed.
+(NK) method. For each benchmark, we will analyze the iterative
+performance of both the nonlinear Newton solver and the linear solver
+used to compute the Newton correction vector at each step. Limiting
+the number of nonlinear iterations required for solution is critical
+for efficiency as nonlinear function evaluations, Jacobian generation,
+and preconditioner construction are typically required at every Newton
+iteration. In addition, timing results will also be discussed. To
+quantify the acceleration provided by the residual Monte Carlo
+component of MCSA, each calculation was also performed using the same
+Newton method but with a plain preconditioned Richardson iteration as
+the linear solution technique to give a Newton-Richardson (NR)
+scheme. For all cases the Richardson iteration was preconditioned
+in the same way as GMRES and MCSA.
 
@@
   \begin{center}
-    \begin{tabular}{ccc}\hline\hline
+    \begin{tabular}{cccc}\hline\hline
       \multicolumn{1}{c}{Rayleigh Number}& 
-      \multicolumn{1}{c}{Newton-Krylov Iterations}&
-      \multicolumn{1}{c}{FANM Iterations}\\
-      \hline
-      \sn{1}{3} & 5 & 5 \\
-      \sn{1}{4} & 7 & 7 \\
-      \sn{1}{5} & 9 & 10 \\
-      \sn{1}{6} & 11 & 11 \\
+      \multicolumn{1}{c}{NK Iterations}&
+      \multicolumn{1}{c}{FANM Iterations}&
+      \multicolumn{1}{c}{NR Iterations}\\
+      \hline
+      \sn{1}{3} & 5 & 5 & 5\\
+      \sn{1}{4} & 7 & 7 & 7\\
+      \sn{1}{5} & 9 & 10 & 9\\
+      \sn{1}{6} & 11 & 11 & 11\\
       %%
@@
 number case that required more Newton iterations required 5 fewer MCSA
-iterations to converge than GMRES iterations. If the performance of
+iterations to converge than GMRES iterations. In all cases, MCSA
+improved on the iterative performance of using only the Richardson
+iteration through the residual Monte Carlo acceleration. If the performance of
 the MCSA implementation used here were to be potentially optimized in
@@
   \begin{center}
-    \begin{tabular}{ccc}\hline\hline
+    \begin{tabular}{cccc}\hline\hline
       \multicolumn{1}{c}{Rayleigh Number}&
       \multicolumn{1}{c}{GMRES Iterations}&
-      \multicolumn{1}{c}{MCSA Iterations}\\
-      \hline
-      \sn{1}{3} & 23 & 18 \\
-      \sn{1}{4} & 23 & 17 \\
-      \sn{1}{5} & 25 & 20 \\
-      \sn{1}{6} & 39 & 25 \\
+      \multicolumn{1}{c}{MCSA Iterations}&
+      \multicolumn{1}{c}{Richardson Iterations}\\
+      \hline
+      \sn{1}{3} & 23 & 18 & 38 \\
+      \sn{1}{4} & 23 & 17 & 34 \\
+      \sn{1}{5} & 25 & 20 & 34 \\
+      \sn{1}{6} & 39 & 25 & 48  \\
       %%
@@
       number case.} \textit{MCSA shows better iterative performance
-      than GMRES at all but the second nonlinear iteration. The
-      forcing term was fixed at \sn{1}{-4} at all nonlinear
-      iterations.}}
+      than GMRES at all but the second nonlinear iteration and MCSA
+      always accelerated the Richardson iteration. The forcing term
+      was fixed at \sn{1}{-4} at all nonlinear iterations.}}
   \label{fig:ra1e6_linear_iters}
@@
   \begin{center}
-    \begin{tabular}{ccc}\hline\hline
+    \begin{tabular}{cccc}\hline\hline
       \multicolumn{1}{c}{Reynolds Number}& 
-      \multicolumn{1}{c}{Newton-Krylov Iterations}&
-      \multicolumn{1}{c}{FANM Iterations}\\
-      \hline
-      100 & 6 & 6 \\
-      300 & 9 & 9 \\
-      500 & 11 & 11 \\
-      700 & 14 & 10 \\
+      \multicolumn{1}{c}{NK Iterations}&
+      \multicolumn{1}{c}{FANM Iterations}&
+      \multicolumn{1}{c}{NR Iterations}\\
+      \hline
+      100 & 6 & 6 & 7\\
+      300 & 9 & 9 & 9\\
+      500 & 11 & 11 & 11\\
+      700 & 14 & 10 & 12\\
       %%
@@
 performed. The average number of linear iterations per nonlinear
-iteration was still much larger for the FANM calculations.
+iteration was still much larger for the FANM calculations. In all
+cases, MCSA improved on the iterative performance of using only the
+Richardson iteration through the residual Monte Carlo acceleration
+even considering the cases with Reynolds numbers of 100 and 700 where
+an extra Newton-Richardson iteration was required to converge the
+nonlinear model.
 
@@
   \begin{center}
-    \begin{tabular}{ccc}\hline\hline
+    \begin{tabular}{cccc}\hline\hline
       \multicolumn{1}{c}{Reynolds Number}& 
       \multicolumn{1}{c}{GMRES Iterations}&
-      \multicolumn{1}{c}{MCSA Iterations}\\
-      \hline
-      100 & 27 & 42 \\
-      300 & 35 & 52 \\
-      500 & 41 & 56 \\
-      700 & 21 & 14 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Lid driven cavity total sum of linear
-      iterations for all nonlinear iterations.} \textit{The results
-      presented here were obtained from the benchmark verification
-      calculations.}}
+      \multicolumn{1}{c}{MCSA Iterations}&
+      \multicolumn{1}{c}{Richardson Iterations}\\
+      \hline
+      100 & 27 & 42 & 151\\
+      300 & 35 & 52 & 133\\
+      500 & 41 & 56 & 154\\
+      700 & 21 & 14 & 32\\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Lid driven cavity total sum of linear iterations
+      for all nonlinear iterations.} \textit{The results presented
+      here were obtained from the benchmark verification
+      calculations. MCSA always accelerated the Richardson
+      iteration.}}
   \label{tab:driven_linear_iter_comparison}
@@
 converge the problem while the ill-conditioned case at a Reynolds
-number of 500 is observed to use many more MCSA iterations.
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{ccc}\hline\hline
+number of 500 is observed to use many more MCSA iterations. For
+Reynolds numbers of both 200 and 500, FANM also reduced the number of
+Newton steps need when compared to the Newton-Richardson scheme.
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{cccc}\hline\hline
       \multicolumn{1}{c}{Reynolds Number}& 
-      \multicolumn{1}{c}{Newton-Krylov Iterations}&
-      \multicolumn{1}{c}{FANM Iterations}\\
-      \hline
-      200 & 10 & 9 \\
-      300 & 15 & 14 \\
-      400 & 10 & 10 \\
-      500 & 19 & 20 \\
+      \multicolumn{1}{c}{NK Iterations}&
+      \multicolumn{1}{c}{FANM Iterations}&
+      \multicolumn{1}{c}{NR Iterations}\\
+      \hline
+      200 & 10 & 9 & 10\\
+      300 & 15 & 14 & 15\\
+      400 & 10 & 10 & 10\\
+      500 & 19 & 20 & 21\\
       %%
@@
   \begin{center}
-    \begin{tabular}{ccc}\hline\hline
+    \begin{tabular}{cccc}\hline\hline
       \multicolumn{1}{c}{Reynolds Number}& 
       \multicolumn{1}{c}{GMRES Iterations}&
-      \multicolumn{1}{c}{MCSA Iterations}\\
-      \hline
-      200 & 24 & 13 \\
-      300 & 23 & 17 \\
-      400 & 18 & 12 \\
-      500 & 30 & 52 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Backward facing step total sum of linear
-      iterations for all nonlinear iterations.} \textit{The results
-      presented here were obtained from the benchmark verification
-      calculations.}}
+      \multicolumn{1}{c}{MCSA Iterations}&
+      \multicolumn{1}{c}{Richardson Iterations}\\
+      \hline
+      200 & 24 & 13 & 21\\
+      300 & 23 & 17 & 21\\
+      400 & 18 & 12 & 14\\
+      500 & 30 & 52 & 98\\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Backward facing step total sum of linear iterations
+      for all nonlinear iterations.} \textit{The results presented
+      here were obtained from the benchmark verification
+      calculations.MCSA always accelerated the Richardson iteration.}}
   \label{tab:step_linear_iter_comparison}
@@
 
+We additionally see the effects of ill-conditioned Jacobians when
+looking at the acceleration provided by MCSA over using a
+preconditioned Richardson iteration. For the first 3 cases, the
+acceleration was not as significant as for the other benchmark
+problems. In this case, this means that the Richardson iteration and
+its acceleration by a stochastic realization of a Richardson iteration
+are not doing the bulk of the work in this problem. Rather, the
+algebraic multigrid preconditioning is performing more work at each
+iteration and therefore the acceleration is smaller. When the system
+becomes even more ill-conditioned in the 500 Reynolds number case
+the acceleration has more of an impact.
+
 \begin{figure}[t!]
@@
 \end{table}
-
-\clearpage
 

### 2013-08-08 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding richardson results to the spn performance comparison

@@
 production Krylov solvers in terms of both iterative performance and
-CPU timing for solutions to the fuel assembly problem.  This
-performance comparison is necessary in order to assess the feasibility
-of the method and identify areas that require further research and
-development.
+CPU timing for solutions to the fuel assembly problem. The iterative
+performance will also be compared to using a Richardson iteration
+preconditioned in the same way for all calculations in order to assess
+the acceleration provided by the residual Monte Carlo component of
+MCSA. This performance comparison is necessary in order to assess the
+feasibility of the method and identify areas that require further
+research and development.
 
@@
 and then averaged over all eigenvalue iterations for each
-solver. Figure~\ref{fig:spn_comparison_iterations} gives the average
+solver. Table~\ref{tab:spn_comparison_iterations} gives the average
 number of linear solver iterations required to converge a single
@@
 cases with the Richardson iteration accelerated with Monte Carlo using
-the collision estimator performing best.
+the collision estimator performing best. When compared to the
+preconditioned Richardson iteration, the acceleration provided by MCSA
+is clear with 3-4 times fewer iterations required for convergence.
 
@@
 
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=5in]{chapters/spn_equations/solver_iters.pdf}
-  \end{center}
-  \caption{\textbf{Average number of iterations required to converge
-      each eigenvalue iteration for the fuel assembly problem as a
-      function of energy groups.}
-    \textit{Table~\ref{tab:spn_solver_defs} gives the description for
-      each solver type presented in the legend with the Krylov solvers
-      represented in black and the MCSA solvers represented in red.}}
-  \label{fig:spn_comparison_iterations}
-\end{figure}
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lccc}\hline\hline
+      \multicolumn{1}{l}{Solver}&
+      \multicolumn{1}{c}{1 Group}&
+      \multicolumn{1}{c}{2 Groups}&
+      \multicolumn{1}{c}{4 Groups}\\
+      \hline
+      BiCGStab-ILUT & 11.6 & 11.6 & 12.4 \\ 
+      GMRES-ILUT & 18.1 & 17.9 & 18.9 \\
+      MCSA-ILUT-R-C & 14.6 & 15.4 & 17.6 \\
+      MCSA-ILUT-MR-C & 16.0 & 17.1 & 23.7 \\
+      MCSA-ILUT-R-EV & 18.3 & 19.4 & 16.8 \\
+      MCSA-ILUT-MR-EV & 19.6 & 22.4 & 17.5 \\
+      Richardson-ILUT & 60.9 & 60.4 & 63.4 \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Average number of linear solver iterations
+      iterations required to converge the fixed source $SP_N$ problem
+      at each eigenvalue iteration for the fuel assembly problem for
+      1, 2 and 4 energy groups.} \textit{Values are rounded to the
+      first decimal place. Table~\ref{tab:spn_solver_defs} gives the
+      description for each solver type presented and Richardson-ILUT
+      is a ILUT preconditioned Richardson iteration.}}
+  \label{tab:spn_comparison_iterations}
+\end{table}
 
@@
 themselves are dense, leading to exceedingly slow computation times
-even in the fixed point iteration. For the production Krylov methods,
-the ILUT preconditioners are applied to a vector with two triangular
-solves, one for each triangular factor. Also, the original linear
-operator has $O(10)$ elements in each row of the system, an order of
-magnitude less than that used in the reduced domain composite
-operator. Second, the implementation presented here has not been
-optimized and it is likely that several components of the algorithm
-can be implemented in a more desirable time complexity. The
-production Krylov solvers presented here have enjoyed decades of
-professional development and optimization. However, it is most
-important to note here that the CPU timing data presented in
-Figure~\ref{fig:spn_comparison_time} shows that all methods have the
-same time complexity; their differences are merely the manifestation
-of a large time constant due to the effects of the explicit
-preconditioning scheme. For transport systems where this is not a
-problem, the literature explicitly shows that MCSA can be competitive
-with these production Krylov methods using CPU time as a measure
-\cite{evans_monte_2012}.
+even in the fixed point iteration.
+
+For the production Krylov methods, the ILUT preconditioners are
+applied to a vector with two triangular solves, one for each
+triangular factor. Also, the original linear operator has $O(10)$
+elements in each row of the system, an order of magnitude less than
+that used in the reduced domain composite operator. Second, the
+implementation presented here has not been optimized and it is likely
+that several components of the algorithm can be implemented in a more
+desirable time complexity. The production Krylov solvers presented
+here have enjoyed decades of professional development and
+optimization. However, it is most important to note here that the CPU
+timing data presented in Figure~\ref{fig:spn_comparison_time} shows
+that all methods have the same time complexity; their differences are
+merely the manifestation of a large time constant due to the effects
+of the explicit preconditioning scheme. For transport systems where
+this is not a problem, the literature explicitly shows that MCSA can
+be competitive with these production Krylov methods using CPU time as
+a measure \cite{evans_monte_2012}.
 

### 2013-08-12 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
recovering from failed rebase

@@
 preconditioning from \S\ref{subsubsec:basic_mcsa_preconditioning} due
-to this diagonal form.
+to this diagonal form and the success achieved by applying it to MCSA
+solutions of the radiation diffusion equation as presented in
+\cite{evans_monte_2012}.
 
@@
 a criticality calculation. A cross section along the vertical axis
-showing homogenized fuel pin materials and the associated grid is
-given by Figure~\ref{fig:problem3_radial_mat} while a cross section of
-the materials configuration along the horizontal axis is given by
-Figure~\ref{fig:problem3_axial_mat}. A detailed view of the assembly
-bottom is given in Figure~\ref{fig:problem3_end}. On the top and
-bottom of the assembly, vacuum conditions are used as well as on the
-top and right boundaries in
-Figure~\ref{fig:problem3_radial_mat}. Reflecting conditions are used
-on the left and bottom boundaries of
+showing homogenized fuel pin, control rod, and moderator materials and
+the associated grid is given by Figure~\ref{fig:problem3_radial_mat}
+while a cross section of the materials configuration along the
+horizontal axis is given by Figure~\ref{fig:problem3_axial_mat}. A
+detailed view of the assembly bottom is given in
+Figure~\ref{fig:problem3_end}. On the top and bottom of the assembly,
+vacuum conditions are used as well as on the top and right boundaries
+in Figure~\ref{fig:problem3_radial_mat}. Reflecting conditions are
+used on the left and bottom boundaries of
 Figure~\ref{fig:problem3_radial_mat}, effectively giving a
@@
       geometry. Gray regions are homogenized fuel and red regions are
-      homogenized moderator. Each fuel pin is resolved by a $2 \times
-      2$ mesh.}}
+      homogenized control rods. The control rods were removed for
+      these problems. Each fuel pin is resolved by a $2 \times 2$
+      mesh.}}
   \label{fig:problem3_radial_mat}

### 2013-08-12 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
updating dissertation build

@@
 equations, closely following the work of Evans
-\cite{evans_simplified_2013}\footnote{The $SP_N$ derivations in this
-  chapter and the associated appendices are heavily based on those
-  presented by Evans in \cite{evans_simplified_2013}}, in order to gain
-full understanding of the underlying system and its potential behavior
-in a Monte Carlo context. From the $P_N$ equations, we apply a set of
-approximations to yield the $SP_N$ equations for fixed source and
-criticality problems. Using the fully-formed linear operator for the
-transport problem, we explore solutions to the $SP_N$ equations with
-Monte Carlo Synthetic Acceleration using a challenging light water
-reactor fuel assembly criticality calculation as the driving
-problem. Several difficulties arise when applying MCSA to these
-problems that were not observed when investigating the simpler
-transport systems.
+\cite{evans_simplified_2013}, in order to gain full understanding of
+the underlying system and its potential behavior in a Monte Carlo
+context. From the $P_N$ equations, we apply a set of approximations to
+yield the $SP_N$ equations for fixed source and criticality
+problems. Using the fully-formed linear operator for the transport
+problem, we explore solutions to the $SP_N$ equations with Monte Carlo
+Synthetic Acceleration using a challenging light water reactor fuel
+assembly criticality calculation as the driving problem. Several
+difficulties arise when applying MCSA to these problems that were not
+observed when investigating the simple transport system in the
+previous chapter.
 
@@
 problems with spectral radii above $\approx 0.97$, using MCSA to solve
-any problems of interest is entirely ineffective and not practical. We
-therefore require a more expansive set of preconditioning techniques
-to move the eigenvalue spectrum of the $SP_N$ problem into a regime in
-which MCSA is more applicable and in which performance is improved.
+any problems of interest is entirely ineffective and not practical. At
+iteration matrix spectral radii this large, using significant numbers
+of stochastic histories in the Monte Carlo solve are not enough to
+reduce the stochastic uncertainty in the correction, causing the MCSA
+sequence to diverge instead of accelerating convergence. We therefore
+require a more expansive set of preconditioning techniques to move the
+eigenvalue spectrum of the $SP_N$ problem into a regime in which MCSA
+is more applicable and in which performance is improved.
 \begin{figure}[t!]

### 2013-08-16 doc/dissertation/presentation/defense.tex
adding a publications slide to the defense talk

@@
   \item Parallelization of MCSA
+    \bigskip
+  \item Summary
   \end{itemize}
@@
     \begin{itemize}
+    \item Monte Carlo is a more natural fit
     \item Application to fluid flow
-    \item Potential memory benefits
     \item Convergence of the linear model is of concern
@@
   \begin{itemize}
-  \item Preliminary development during January of 2013
+  \item Designed to be easily incorporated with production physics
+    codes
     \bigskip
@@
 
-      \begin{figure}[htpb!]
-        \begin{center}
-          \includegraphics[width=1.9in]{solver_iters.pdf}
-        \end{center}
-      \end{figure}
+      {\tiny
+      \begin{table}[h!]
+        \begin{center}
+          \begin{tabular}{lccc}\hline\hline
+            \multicolumn{1}{l}{Solver}&
+            \multicolumn{1}{c}{1 Group}&
+            \multicolumn{1}{c}{2 Groups}&
+            \multicolumn{1}{c}{4 Groups}\\
+            \hline
+            BiCGStab-ILUT & 11.6 & 11.6 & 12.4 \\ 
+            GMRES-ILUT & 18.1 & 17.9 & 18.9 \\
+            MCSA-ILUT-R-C & 14.6 & 15.4 & 17.6 \\
+            MCSA-ILUT-MR-C & 16.0 & 17.1 & 23.7 \\
+            MCSA-ILUT-R-EV & 18.3 & 19.4 & 16.8 \\
+            MCSA-ILUT-MR-EV & 19.6 & 22.4 & 17.5 \\
+            Richardson-ILUT & 60.9 & 60.4 & 63.4 \\
+            %%
+            \hline\hline
+          \end{tabular}
+        \end{center}
+        \caption{Average number of linear solver iterations per
+          eigenvalue iteration.}
+      \end{table}
+      }
 
@@
     \begin{center}
-      \begin{tabular}{lcc}\hline\hline
+      \begin{tabular}{lccc}\hline\hline
         \multicolumn{1}{l}{Benchmark}& 
-        \multicolumn{1}{c}{Newton-Krylov}&
-        \multicolumn{1}{c}{FANM}\\
+        \multicolumn{1}{c}{NK}&
+        \multicolumn{1}{c}{FANM}&
+        \multicolumn{1}{c}{NR}\\
         \hline
-        Convection, Ra=\sn{1}{3} & 5 & 5 \\
-        Convection, Ra=\sn{1}{4} & 7 & 7 \\
-        Convection, Ra=\sn{1}{5} & 9 & 10 \\
-        Convection, Ra=\sn{1}{6} & 11 & 11 \\
+        Convection, Ra=\sn{1}{3} & 5 & 5 & 5\\
+        Convection, Ra=\sn{1}{4} & 7 & 7 & 7\\
+        Convection, Ra=\sn{1}{5} & 9 & 10 & 9\\
+        Convection, Ra=\sn{1}{6} & 11 & 11 & 11\\
         \hline
-        Lid Driven, Re=100 & 6 & 6 \\
-        Lid Driven, Re=300 & 9 & 9 \\
-        Lid Driven, Re=500 & 11 & 11 \\
-        Lid Driven, Re=700 & 14 & 10 \\
+        Lid Driven, Re=100 & 6 & 6 & 7\\
+        Lid Driven, Re=300 & 9 & 9 & 9\\
+        Lid Driven, Re=500 & 11 & 11 & 11\\
+        Lid Driven, Re=700 & 14 & 10 & 12\\
         \hline
-        Backward Step, Re=200 & 10 & 9 \\
-        Backward Step, Re=300 & 15 & 14 \\
-        Backward Step, Re=400 & 10 & 10 \\
-        Backward Step, Re=500 & 19 & 20 \\
+        Backward Step, Re=200 & 10 & 9 & 10\\
+        Backward Step, Re=300 & 15 & 14 & 15\\
+        Backward Step, Re=400 & 10 & 10 & 10\\
+        Backward Step, Re=500 & 19 & 20 & 21\\
         %%
@@
     \begin{center}
-      \begin{tabular}{lcc}\hline\hline
+      \begin{tabular}{lccc}\hline\hline
         \multicolumn{1}{l}{Benchmark}& 
-        \multicolumn{1}{c}{Newton-Krylov}&
-        \multicolumn{1}{c}{FANM}\\
+        \multicolumn{1}{c}{GMRES}&
+        \multicolumn{1}{c}{MCSA}&
+        \multicolumn{1}{c}{Richardson}\\
         \hline
-        Convection, Ra=\sn{1}{3} & 32 & 18 \\
-        Convection, Ra=\sn{1}{4} & 23 & 17 \\
-        Convection, Ra=\sn{1}{5} & 25 & 20 \\
-        Convection, Ra=\sn{1}{6} & 39 & 25 \\
+        Convection, Ra=\sn{1}{3} & 32 & 18 & 38\\
+        Convection, Ra=\sn{1}{4} & 23 & 17 & 34\\
+        Convection, Ra=\sn{1}{5} & 25 & 20 & 34\\
+        Convection, Ra=\sn{1}{6} & 39 & 25 & 48\\
         \hline
-        Lid Driven, Re=100 & 27 & 42 \\
-        Lid Driven, Re=300 & 35 & 52 \\
-        Lid Driven, Re=500 & 41 & 56 \\
-        Lid Driven, Re=700 & 21 & 14 \\
+        Lid Driven, Re=100 & 27 & 42 & 151\\
+        Lid Driven, Re=300 & 35 & 52 & 133\\
+        Lid Driven, Re=500 & 41 & 56 & 154\\
+        Lid Driven, Re=700 & 21 & 14 & 32\\
         \hline
-        Backward Step, Re=200 & 24 & 13 \\
-        Backward Step, Re=300 & 23 & 17 \\
-        Backward Step, Re=400 & 18 & 12 \\
-        Backward Step, Re=500 & 30 & 52 \\
+        Backward Step, Re=200 & 24 & 13 & 21\\
+        Backward Step, Re=300 & 23 & 17 & 21\\
+        Backward Step, Re=400 & 18 & 12 & 14\\
+        Backward Step, Re=500 & 30 & 52 & 98\\
         %%
@@
   \item \textbf{Parallelization of MCSA}
+    \bigskip
+  \item Summary
   \end{itemize}
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Parallel Scaling}
+\begin{frame}{Leadership-Class Parallel Scaling Studies}
 
@@
 %%---------------------------------------------------------------------------%%
+\begin{frame}{Outline}
+
+  \begin{itemize}
+  \item Monte Carlo Synthetic Acceleration Methods
+    \bigskip
+  \item Application to Neutron Transport
+    \bigskip
+  \item Application to Fluid Flow
+    \bigskip
+  \item Parallelization of MCSA
+    \bigskip
+  \item \textbf{Summary}
+  \end{itemize}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
 \begin{frame}{MCSA Application to Neutronics Summary}
@@
     \begin{itemize}
+    \item MCSA has been incorporated into the Exnihilo neutronics
+      production code base developed at Oak Ridge National Laboratory
+      \medskip
     \item MCSA can solve the asymmetric system generated by the $SP_N$
@@
       \medskip
-    \item MCSA was verified to produce the same flux distribution and
-      k-eigenvalue for the fuel assembly as production Krylov methods
-      \medskip
     \item MCSA was observed to converge in fewer iterations per eigenvalue
@@
       \medskip
-    \item The FANM method has been verified to produce the same solutions
-      as a production Newton-Krylov method for three difficult benchmark
-      problems
+    \item The FANM method has been incorporated into the Drekar
+      multiphysics production code base developed at Sandia National
+      Laboratories
       \medskip
@@
     \begin{itemize}
-    \item The multiple-set overlapping-domain (MSOD) parallel algorithm
-      for particle transport has been adapted to parallelize MCSA
+    \item The multiple-set overlapping-domain (MSOD) parallel
+      algorithm for domain decomposed particle transport has been
+      adapted to parallelize MCSA
       \medskip
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Discussion}
+\begin{frame}{Publications}
+
+  \small{
+    \begin{enumerate}
+    \item S.R. Slattery, T.M. Evans, P.P.H. Wilson, \textbf{A
+      Multiple-Set Overlapping-Domain Decomposed Monte Carlo Synthetic
+      Acceleration Method for Linear Systems}, \textit{Joint
+      International Conference on Supercomputing in Nuclear
+      Applications and Monte Carlo 2013 (SNA+MC 2013), Paris, France,
+      October 27-31, 2013. Accepted for oral presentation.}
+      \medskip
+    \item T.M. Evans, S.W. Mosher, S.R. Slattery, S.P. Hamilton,
+      \textbf{A Monte Carlo Synthetic-Acceleration Method for Solving
+        the Thermal Radiation Diffusion Equation}, \textit{Journal of
+        Computational Physics, Submitted.}
+      \medskip
+    \item S.R. Slattery, T.M. Evans, P.P.H. Wilson, \textbf{A Spectral
+      Analysis of the Domain Decomposed Monte Carlo Method for Linear
+      Systems}, \textit{International Conference on Mathematics and
+      Computational Methods Applied to Nuclear Science \& Engineering
+      (M\&C 2013), American Nuclear Society, Sun Valley, ID, May 5-9,
+      2013.}
+    \end{enumerate}
+  }
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Thank You}
+
+  \begin{itemize}
+  \item Paul
+  \item Tom
+  \item Roger
+  \item CASL, ORNL, Sandia, OLCF
+\end{itemize}
 

### 2013-08-16 doc/dissertation/presentation/defense.tex
more cleanup in the defense slides

@@
   \begin{itemize}
-  \item Development of a linear solver for discrete systems leveraging
+  \item Development of a linear scheme for discrete systems leveraging
     Monte Carlo Synthetic Acceleration
@@
     \item Application to neutron transport
-    \item Research is required to explore general solver development
+    \item Research is required to explore general MCSA development
     \item Performance is of concern
@@
     \bigskip
-  \item Development of a nonlinear solver for discrete systems
+  \item Development of a nonlinear scheme for discrete systems
     leveraging Monte Carlo Synthetic Acceleration
@@
 
+  \smallskip
+
   \begin{multline}
@@
 
+  \smallskip
+
   \begin{equation}
@@
     \mathbb{A}_{nm} \mathbb{U}_m = \frac{1}{k} \sum_{m=1}^4
-    \mathbb{F}_{nm} \mathbb{U}_n\ \ \ \ \ \ \ n = 1,2,3,4\:.
+    \mathbb{F}_{nm} \mathbb{U}_m\ \ \ \ \ \ \ n = 1,2,3,4\:.
     \label{eq:spn_fission_matrix}
@@
     \medskip
-  \item Use Thyra/Stratimikos interfaces for plug and play - Thanks
-    Ross!
+  \item Use Thyra/Stratimikos interfaces for plug and play with $SP_N$
     \medskip
@@
         \end{center}
-        \caption{Average number of linear solver iterations per
+        \caption{ Average number of linear solver iterations per
           eigenvalue iteration.}
@@
           \bigskip
-        \item Spectral radius and memory limitations combined to
+        \item Spectral radius and memory limitations combine to
           prevent solutions at finer discretizations
@@
     \item Potential scaling improvements
-    \item Resiliency benefits
+    \item Potential resiliency benefits
     \end{itemize}
@@
     methods with FAD
+  \bigskip
   \item MCLS incorporated into the Drekar nonlinear solver scheme
     using Thyra/Stratimikos to implement FANM
-  \item Thanks Roger!
   \end{itemize}
@@
       \multicolumn{1}{l}{Benchmark}& 
-      \multicolumn{1}{c}{Newton-Krylov Speedup}\\
+      \multicolumn{1}{c}{NK Speedup}\\
       \hline

### 2013-08-19 doc/dissertation/document/chapters/introduction/introduction.tex
starting to incorporate some of Pauls initial comments

@@
 nonlinear transport problems are a primary focus of study. Neutronics,
-fluid flow, and heat transfer are among the predominate physics that
+fluid flow, and heat transfer are among the predominant physics that
 fall into this category and simulation technologies to solve these
@@
 phase space and the problems of interest are often geometrically
-complex, very large, or both, requiring tremendous computational
-resources to generate an adequate solution. Modern deterministic
-methods for large scale problems are commonly variants on the discrete
-ordinates ($S_N$) method \cite{evans_denovo:_2010}. For fission
-reactor neutronics simulations, the $S_N$ method requires potentially
-trillions of unknown angular flux moments to be computed to achieve
-good accuracy for the responses of interest
+complex, contain large numbers of degrees of freedom, or both,
+requiring tremendous computational resources to generate an adequate
+solution. Modern high fidelity deterministic methods for large scale
+problems are commonly variants on the discrete ordinates ($S_N$)
+method \cite{evans_denovo:_2010}. For fission reactor neutronics
+simulations, the $S_N$ method requires potentially trillions of
+unknown angular flux moments to be computed to achieve good accuracy
+for the responses of interest
 \cite{slaybaugh_acceleration_2011}. Other forms of the transport
@@
 application to today's challenge problems in neutron transport for
-fission reactor analysis. The reduction in numerical complexity of
-current deterministic solution methods using the $S_N$ approximation
-could mean significant savings in both compute time and memory
-required. In addition, the characteristics of the solution to the
-transport problem for a steady state reactor core permit diffusion
-theory to be used; a staple of the nuclear industry since its
-inception. Therefore, if diffusion theory is applicable, then finer
-grained solutions that capture more of the physics contained in the
-transport equation should be possible with the $SP_N$ method. In doing
-so, we also expect from the literature to obtain computed responses on
-the order of accuracy we would expect from an appropriately
-discretized $S_N$ method at a fraction of the cost.
+fission reactor analysis. The reduction in numerical complexity
+compared to current deterministic solution methods using the $S_N$
+approximation could mean significant savings in both compute time and
+memory required.
 
@@
 have access to all of the modern advancements in computational linear
-algebra including Krylov solvers for asymmetric systems and algebraic
-preconditioning methods.
+algebra including the use of Krylov solvers for asymmetric systems
+without the need to implement the action of the physics operator on a
+vector and easy application of algebraic preconditioning methods for
+improved convergence properties and time to solution.
 
@@
 The computational resources required to solve the neutronics and fluid
-dynamics problems are tremendous. Recent work in modeling coupled
+dynamics problems can be tremendous. Recent work in modeling coupled
 fluid flow and solid material heat and mass transfer in a reactor

### 2013-08-19 doc/dissertation/document/chapters/mc_background/mc_background.tex
pushing full draft of disseration
fixing typo in adjoint constraints
recovering from failed rebase
starting to incorporate some of Pauls initial comments

@@
-\chapter{Monte Carlo Synthetic Acceleration Methods}
+\chapter{Monte Carlo Synthetic Acceleration\\ Methods}
 \label{ch:stochastic_methods}
@@
 to Eq~(\ref{eq:linear_problem}). For a given linear operator $\ve{A}$,
-we can use diagonal splitting in a similar manner as the stationary
-method in Eq~(\ref{eq:linear_split_equation2}) to define the following
+we can use diagonal splitting similar to the stationary method in
+Eq~(\ref{eq:linear_split_equation2}) to define the following
 operator\footnote{It should be noted that non-diagonal splittings have
@@
 that approaches, but never reaches zero. We do this by noticing that
-the factors added to Eq~(\ref{eq:direct_permutation_weight}) will
-become diminishingly small due to their definition in
-Eq~(\ref{eq:direct_weight}) and therefore their contributions to the
-solution estimate will become negligible. Using this, we choose
-terminate a random walk sequence with a \textit{weight cutoff}, $W_c$,
-that is enforced when $W_m < W_c$ for a particular random walk
-permutation.
+if the problem is convergent, the factors added to
+Eq~(\ref{eq:direct_permutation_weight}) will make $W_m$ become
+diminishingly small due to their definition in
+Eq~(\ref{eq:direct_weight}) and therefore the contribution of the
+history to the solution estimate will become negligible. Using this,
+we choose terminate a random walk sequence with a \textit{weight
+  cutoff}, $W_c$, that is enforced when $W_m < W_c$ for a particular
+random walk permutation.
 
@@
 \end{equation}
-and applying Eq~(\ref{eq:direct_permutation_contribution}):
-\begin{equation}
-  \sigma_i^2 = \sum_{\nu} P_{\nu} \Big(\sum_{m=0}^k W_{m}^2 b_{i_m}^2
-  + 2 \sum_{\substack{ j=0 \\ m<j}}^k W_m W_j b_{i_m} b_{i_j} \Big)-
+and applying Eq~(\ref{eq:direct_permutation_contribution}) and
+expanding the square of the summation:
+\begin{equation}
+  \sigma_i^2 = \sum_{\nu} P_{\nu} \Big(\sum_{m=0}^k \Big[ W_{m}^2 b_{i_m}^2
+  + 2 \sum_{j=0}^{m-1} W_m W_j b_{i_m} b_{i_j} \Big] \Big)-
   x_i^2\:.
@@
 high variance prevents a precise solution from being obtained although
-we begin to see the solution take shape as expected. At 1000 histories
-per state, enough tallies have been made to generate a reasonable
-estimate for the structure of the solution. It is interesting to note
-here that as the statistical uncertainty is reduced at each grid point
-in the domain, the solution is resolved in a certain sense, analogous
-to the convergence of a traditional iterative method.
+we begin to see the solution take shape as expected. It is interesting
+to note here that as the statistical uncertainty is reduced at each
+grid point in the domain, the solution is resolved in a certain sense,
+analogous to the convergence of a traditional iterative method.
 \clearpage
@@
 \begin{equation}
-  \langle \ve{A}^T \ve{x}, \ve{y} \rangle = \langle \ve{x}, \ve{A}
-  \ve{y} \rangle\:.
+  \langle \ve{A}^T \ve{y}, \ve{x} \rangle = \langle \ve{y}, \ve{A}
+  \ve{x} \rangle\:.
   \label{eq:adjoint_operator_product}
@@
 contributions to the tallies in those states as they are transported.
+
+\clearpage
 

### 2013-08-19 doc/dissertation/document/chapters/mc_background/mc_background.tex
finished incorporating pauls comments up to p72

@@
 \begin{equation}
-  W_{m} = w_{i_0,i_1} w_{i_1,i_2} \cdots w_{i_{m-1},i_m}\:,
+  W_{m} = W_0 w_{i_0,i_1} w_{i_1,i_2} \cdots w_{i_{m-1},i_m}\:,
   \label{eq:direct_permutation_weight}
@@
 \begin{equation}
-  P_{\nu} = p_{i,i_1} p_{i_1,i_2} \cdots p_{i_{k-1},i_k}\:.
+  P_{\nu} = P_{i_0=i} p_{i,i_1} p_{i_1,i_2} \cdots p_{i_{k-1},i_k}\:,
   \label{eq:direct_permutation_probability}
 \end{equation}
-Finally, we define the expectation value of $X$ to be the collection
-of all random walk permutations and their probabilities:
+with $P_{i_0=i} = 1$ for the direct method. Finally, we define the
+expectation value of $X$ to be the collection of all random walk
+permutations and their probabilities:
 \begin{equation}
@@
   \sigma_i^2 = \sum_{\nu} P_{\nu} \sum_{m=0}^k W_{m}^2 b_{i_m}^2 + 2
-  \sum_{\nu} P_{\nu} \sum_{\substack{ j=0 \\ m<j}}^k W_m W_j b_{i_m}
+  \sum_{\nu} P_{\nu} \sum_{m=0}^k \sum_{j=0}^{m-1} W_m W_j b_{i_m}
   b_{i_j} - x_i^2\:.
@@
 \end{equation}
-In particular, we will isolate the first summation of the variance by
+In particular, we will isolate the first term of the variance by
 expanding it:
@@
 \begin{equation}
-  \sigma^2_i = (\ve{T}\ve{b}^2)_i + 2 \sum_{\nu} P_{\nu}
-  \sum_{\substack{ j=0 \\ m<j}}^k W_m W_j b_{i_m} b_{i_j} - x_i^2\:.
+  \sigma^2_i = (\ve{T}\ve{b}^2)_i + 2 \sum_{\nu} P_{\nu} \sum_{m=0}^k
+  \sum_{j=0}^{m-1} W_m W_j b_{i_m} b_{i_j} - x_i^2\:.
   \label{eq:direct_variance_5}
@@
 \begin{equation}
-  P_{i_0=i}(\nu) = \frac{|b_i|}{||\ve{b}||_1}\:.
+  P_{i_0=i}(\nu) = \frac{|b_i|}{||\ve{b}||_1}\:,
   \label{eq:adjoint_source_probability}
 \end{equation}
-The random walk starting weight will then be $W_0 = b_{i_0}$.  As a
-result of using the adjoint system, we modify our probabilities and
-weights using the \textit{adjoint Neumann-Ulam decomposition} of
+with a random walk starting weight of:
+\begin{equation}
+  W_0 = ||\ve{b}||_1 \frac{b_i}{|b_i|}\:,
+  \label{eq:adjoint_starting_weight}
+\end{equation}
+which gives the additional useful relation:
+\begin{equation}
+  b_{i_0} = W_0 P_{i_0=i}\:.
+  \label{eq:adjoint_source_definition}
+\end{equation}
+As a result of using the adjoint system, we modify our probabilities
+and weights using the \textit{adjoint Neumann-Ulam decomposition} of
 $\ve{H}$:
@@
 have a dependency on the source state as in
-Eq~(\ref{eq:adjoint_permutation_contribution}), providing a remedy for
+Eq~(\ref{eq:direct_permutation_contribution}), providing a remedy for
 the situation in the direct method where we must start a random walk
@@
 \end{equation}
-which, if expanded in the same way as the direct method, directly
-recovers the exact solution:
+which, if expanded in the same way as the direct method and utilizing
+Eq~(\ref{eq:adjoint_source_definition}) to insert the source term,
+directly recovers the exact solution:
 \begin{equation}
@@
 \end{equation}
-therefore also providing an unbiased Monte Carlo estimate of the
+therefore, also providing an unbiased Monte Carlo estimate of the
 solution. Note that this expansion produces the effective sequence of
@@
 this estimator, the $w_{ij}$ elements that are added into the tally
-for each state are only selected if the random walk currently resides
-in that state. Much like a mesh tally in a particle transport
+for each state are only selected if/when the random walk currently
+resides in that state. Much like a mesh tally in a particle transport
 simulation, we have $N$ simultaneous tallies for $\ve{A} \in
 \mathbb{R}^{N \times N}$ that will yield the entire solution
-vector. Based on their current state in the system, the random walks
-will contribute tally $j$ in this group.
+vector.
 
@@
 \begin{equation}
-  \sigma_j^2 = \sum_{\nu} P_{\nu} \sum_{m=0}^k W_{m}^2 \delta_{i_m,j} + 2
-  \sum_{\nu} P_{\nu} \sum_{\substack{ l=0 \\ m<l}}^k W_m W_l
+  \sigma_j^2 = \sum_{\nu} P_{\nu} \sum_{m=0}^k W_{m}^2 \delta_{i_m,j}
+  + 2 \sum_{\nu} P_{\nu} \sum_{m=0}^k \sum_{l=0}^{m-1} W_m W_l
   \delta_{i_m,j} \delta_{i_l,j} - x_j^2\:.
@@
 In this case, in the second summation $m \neq l$ for all states due to
-the form of the summation and therefore the delta functions,
-$\delta_{i_m,j}$ or $\delta_{i_l,j}$, will only be nonzero for random
-walks that are in the current solution state (when $i_m = i_l = j$)
-and other states visited do not contribute to the variance of the
-current state (when $i_m \neq i_l$). This is important as it shows a
-decoupling of the variance among the solution vector states, meaning
-that the variance in solution state $i$ does not depend on the
+the form of the summation and therefore the combination of delta
+functions, $\delta_{i_m,j}$ of $\delta_{i_l,j}$, will only be nonzero
+for random walks that are in the current solution state (when $i_m =
+i_l = j$) and other states visited do not contribute to the variance
+of the current state (when $i_m \neq i_l$). This is important as it
+shows a decoupling of the variance among the solution vector states,
+meaning that the variance in solution state $i$ does not depend on the
 variance in solution state $j$.
@@
   w^2_{i_k,i_{k-1}}\ldots w^2_{i_2,i_1} w^2_{i_1,i_0}
-  b^2_{i_0}\delta_{i_k,j} + \\ 2 \sum_{\nu} P_{\nu} \sum_{\substack{
-      l=0 \\ m<l}}^k W_m W_l \delta_{i_m,j} \delta_{i_l,j} - x_j^2\:.
+  b^2_{i_0}\delta_{i_k,j} + \\ 2 \sum_{\nu} P_{\nu} \sum_{m=0}^k
+  \sum_{l=0}^{m-1} W_m W_l \delta_{i_m,j} \delta_{i_l,j} - x_j^2\:.
   \label{eq:collision_variance_2}
@@
   g_{i_2,i_1}g_{i_1,i_0}b_{i_0}\delta_{i_k,j} + \\ 2 \sum_{\nu}
-  P_{\nu} \sum_{\substack{ l=0 \\ m<l}}^k W_m W_l \delta_{i_m,j}
+  P_{\nu} \sum_{m=0}^k \sum_{l=0}^{m-1} W_m W_l \delta_{i_m,j}
   \delta_{i_l,j} - x_j^2\:,
@@
 that may be reached from the current state $i_m$. Via Okten, the
-estimator can be shown to be unbiased through a comparison to
-the expected value estimator. We can first rewrite the summation in
+estimator can be shown to be unbiased through a comparison to the
+collision estimator. We can first rewrite the summation in
 Eq~(\ref{eq:expected_value_estimator}):
@@
   \sum_{m=0}^k W_{m} h_{j,i_m} +\\ \sum_{\nu} P_{\nu} \sum_{m=0}^k
-  W_{m}^2 h_{j,i_m}^2 + 2 \sum_{\nu} P_{\nu} \sum_{\substack{ l=0
-      \\ m<l}}^k W_m W_l h_{j,i_m} h_{j,i_l} - x_j^2\:.
+  W_{m}^2 h_{j,i_m}^2 + 2 \sum_{\nu} P_{\nu} \sum_{m=0}^k
+  \sum_{l=0}^{m-1} W_m W_l h_{j,i_m} h_{j,i_l} - x_j^2\:.
   \label{eq:expected_value_variance_1}
@@
 \end{figure}
-As the number of histories used per state is increased, the
+As the number of histories used per state (or DOF) is increased, the
 statistical variance of the solutions is decreased as more tally
@@
 diffusion equation \cite{evans_monte_2012}. With their new methods,
-they achieved identical numerical results as conventional Krylov
+they achieved numerical results identical to conventional Krylov
 solvers as well as comparable performance in both number of iterations
@@
 Figure~\ref{fig:poisson_cpu_time} gives the CPU time needed for each
-case to converge in seconds and Figure~\ref{fig:poisson_iterations}
+case to converge in seconds, and Figure~\ref{fig:poisson_iterations}
 gives the number of iterations needed for each case to converge to the
-specified tolerance as a function of the problem size. All
+specified tolerance, both as a function of the problem size. All
 computations presented in this section and the remaining sections of
@@
 while the number of iterations required to converge is also reduced as
-shown in Figure~\ref{fig:poisson_iterations}. We expect this for several
-reasons. First, with an equivalent number of histories specified for
-both solvers per MCSA iteration and a system of size $N \times N$, the
-direct solver will compute a single random walk for each state in the
-system per iteration to acquire a solution in that state, regardless
-of the size of the residual in that state. This is necessary in the
-direct method to ensure a contribution from each state as the random
-walk sequence will only contribute to the starting state. For the
-adjoint method, a total of $N \times N$ random walk events will have
-their starting state determined by sampling the residual
-vector. Because the random walk sequence contributes to the state in
-which it currently resides, sampling the residual vector as the Monte
-Carlo source gives a higher density of random walk events in regions
-with a high residual, thus giving a more accurate correction in that
-region due to reduced statistical error. From an iteration
-perspective, Figure~\ref{fig:poisson_iterations} shows that using the
-direct method yields a roughly unchanging number of iterations
-required to converge as the problem size increases. Again, if we
-desire a correction value for all states in the problem, then we must
-start a random walk in each state in the system which does not reduce
-the number of iterations need as the problem size grows. Conversely,
-as the problem size grows in the adjoint method, the additional
-stochastic histories that will be computed are concentrated in regions
-with a large residual, further reducing the stochastic error in the
+shown in Figure~\ref{fig:poisson_iterations}. We expect this for
+several reasons. First, with an equivalent number of histories
+specified for both solvers per MCSA iteration and a system of size $N
+\times N$, the direct solver will compute a single random walk for
+each state in the system per iteration to acquire a solution in that
+state. This is necessary in the direct method to ensure a contribution
+from each state as the random walk sequence will only contribute to
+the starting state. For the adjoint method, a total of $N \times N$
+random walk events will have their starting state determined by
+sampling the residual vector. Because the random walk sequence
+contributes to the state in which it currently resides, sampling the
+residual vector as the Monte Carlo source gives a higher density of
+random walk events in regions with a high residual, thus giving a more
+accurate correction in that region due to reduced statistical
+error. From an iteration perspective,
+Figure~\ref{fig:poisson_iterations} shows that using the direct method
+yields a roughly unchanging number of iterations required to converge
+as the problem size increases. Again, if we desire a correction value
+for all states in the problem, then we must start a random walk in
+each state in the system without taking into account the structure of
+the residual vector. Because of this, as the problem size grows adding
+histories is ineffective as many are added in states where the error
+is smaller than in other parts of the problem, ultimately not reducing
+the number of iterations needed to converge. Conversely, as the
+problem size grows in the adjoint method, the additional stochastic
+histories that will be computed are concentrated in regions with a
+large residual, further reducing the stochastic error in the
 correction in those regions and subsequently reducing the required
@@
 iterations at a very large weight cutoff. This suggests that a fairly
-large weight cutoff can be used in practice with both estimators and
-that the preliminary components of the random walk are more important
-to MCSA convergence than those that occur later and with lower weight
-contributions. To further motivate using a larger weight cutoff,
-Figure~\ref{fig:estimator_wc_time} gives the CPU time need to converge
-as a function of weight cutoff. As expected, lowering the weight
-cutoff lengthens the random walk lengths and increases CPU time at no
-gain of iterative performance. In general, these results suggest using
-the expected value estimator over the collision estimator with MCSA as
-well as using a larger weight cutoff.
+large weight cutoff can be used in practice with both adjoint
+estimators and that the preliminary components of the random walk are
+more important to MCSA convergence than those that occur later and
+with lower weight contributions. To further motivate using a larger
+weight cutoff, Figure~\ref{fig:estimator_wc_time} gives the CPU time
+need to converge as a function of weight cutoff. As expected, lowering
+the weight cutoff lengthens the random walk lengths and increases CPU
+time at no gain of iterative performance. In general, these results
+suggest using the expected value estimator over the collision
+estimator with MCSA as well as using a larger weight cutoff.
 
@@
 weights of the Monte Carlo problem are modified to improve the
-performance of a given estimator. In many cases, such modifications
-will introduce bias into the system. However, using Monte Carlo within
-an MCSA iterative scheme provides a potential buffer for the solution
-from this bias as the correction generated by the Monte Carlo solve
-will contain large statistical error even without variance
-reduction. For this work, any method that alters the weights and
-probabilities of the Monte Carlo method with the aim of improving
-either iterative performance or time to convergence will be defined as
-a variance reduction scheme.
+performance of a given estimator. In the context of improving MCSA
+results, such modifications may potentially introduce bias into the
+system. However, using Monte Carlo within an MCSA iterative scheme
+provides a potential buffer for the solution from this bias as the
+correction generated by the Monte Carlo solve will contain large
+statistical error even without variance reduction. For this work, any
+method that alters the weights and probabilities of the Monte Carlo
+method with the aim of improving either iterative performance or time
+to convergence will be defined as a variance reduction scheme.
 
@@
   MCSA compared to the forward method
-\item MCSA has superior performance to Halton's residual Monte Carlo
-  method
+\item MCSA has superior performance when compared to Halton's residual
+  Monte Carlo method
 \item The Richardson iteration in MCSA behaves as a smoother for the

### 2013-08-19 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
finished incorporating pauls comments up to p72

@@
 solutions against production Krylov solvers for problems with varying
-energy groups. Finally performance is analyzed through comparison with
-those same Krylov solvers in terms of both iterative performance and
-CPU timing using the same set of problems utilized for the
-verification.
+numbers of energy groups. Finally performance is analyzed through
+comparison with those same Krylov solvers in terms of both iterative
+performance and CPU timing using the same set of problems utilized for
+the verification.
 
@@
 It should be noted here that with respect to a general MCSA scheme,
-the introduction of the fission in the system will only affect the
-source vector in the linear system. The linear operator, and therefore
+the introduction of fission in the system will only affect the source
+vector in the linear system. The linear operator, and therefore
 overall MCSA performance will be dictated by streaming and scattering

### 2013-08-20 doc/dissertation/document/frontmatter/acks.tex
adding proper olcf acknowledgement

@@
 
-Thank you to the Oak Ridge Leadership Computing Facility (OLCF) for
-providing computational resources and support.
+This research used resources of the Oak Ridge Leadership Computing
+Facility at the Oak Ridge National Laboratory, which is supported by
+the Office of Science of the U.S. Department of Energy under Contract
+No. DE-AC05-00OR22725.

### 2013-08-20 doc/dissertation/document/frontmatter/abstract.tex
adding abstract corrections

@@
 possess both high levels of concurrency and low memory availability
-per core compared to today's machines. Therefore, we seek new
-algorithms to solve these problems in reactor analysis that have the
-potential to perform well on forthcoming hardware with the goal of
-improving both iterative performance and parallel scalability.
+per core compared to today's machines with a potentially rate of
+faults. Therefore, we seek new algorithms to solve these problems in
+reactor analysis that have the potential to perform well on
+forthcoming hardware with the goal of improving both iterative
+performance and parallel scalability.
 
@@
 First, solutions to the $SP_N$ discretization of the linear Boltzmann
-neutron transport equation are obtained using Monte Carlo Synthetic
-Acceleration with a difficult criticality calculation for a light
-water reactor fuel assembly used as the driving problem. It was
-discovered that basic Jacobi-based preconditioning as used in previous
-work was not sufficient to achieve convergence due to the dominating
-effect of the light water moderator. To enable MCSA as a solution
-technique a group of modern preconditioning strategies are
-researched. The new preconditioning techniques, however, create
-unwanted growth in both memory consumption and computation time for
-these problems. Monte Carlo Synthetic Acceleration was verified to
-produce the correct solution for the fuel assembly criticality problem
-when compared to conventional Krylov methods and demonstrated improved
+neutron transport equation are obtained using MCSA with a difficult
+criticality calculation for a light water reactor fuel assembly used
+as the driving problem. It was discovered that basic Jacobi-based
+preconditioning as used in previous work was not sufficient to achieve
+convergence due to the dominating effect of the light water
+moderator. To enable MCSA as a solution technique a group of modern
+preconditioning strategies are researched. The new preconditioning
+techniques, however, create unwanted growth in both memory consumption
+and computation time for these problems. MCSA was verified to produce
+the correct solution for the fuel assembly criticality problem when
+compared to conventional Krylov methods and demonstrated improved
 iterative performance over GMRES by converging in fewer iterations
@@
 
-Second, solutions to the full Navier-Stokes equations were obtained by
-developing the Forward-Automated Newton-MCSA (FANM) method for
-nonlinear systems based on Newton's method. Three difficult fluid
+Second, solutions to the compressibleNavier-Stokes equations were
+obtained by developing the Forward-Automated Newton-MCSA (FANM) method
+for nonlinear systems based on Newton's method. Three difficult fluid
 benchmark problems in both convective and driven flow regimes were
 used to drive the research and development of the method. It was found
-for all benchmarks that FANM produced identical solutions to modern
-Newton-Krylov methods. For each benchmark, performance comparisons
-between FANM and Newton-Krylov were also conducted. For 8 out of 12
-benchmark cases, it was found that FANM had better iterative
-performance than the Newton-Krylov method by converging the nonlinear
-residual in fewer linear solver iterations with the same
-preconditioning.
+for all benchmarks that FANM produced identical solutions when
+compared to those generated by modern Newton-Krylov methods. For each
+benchmark, performance comparisons between FANM and Newton-Krylov were
+also conducted. For 8 out of 12 benchmark cases, it was found that
+FANM had better iterative performance than the Newton-Krylov method by
+converging the nonlinear residual in fewer linear solver iterations
+with the same preconditioning.
 
@@
 techniques developed in the research on the neutron transport problem
-and in turn effectively leveraging MCSA on leadership-class hardware
+and in turn, effectively leveraging MCSA on leadership-class hardware
 requires the new parallel algorithm developed by this work. With the
 further development of these methods based on this work, future
-nuclear reactor simulations may be enabled that have an improved time
-to solution and more efficiently utilize the available computational
-resources.
+nuclear reactor simulations may potentially be enabled that have an
+improved time to solution and more efficiently utilize the available
+computational resources.

### 2013-08-20 doc/dissertation/document/chapters/introduction/introduction.tex
adding my chapter 1 corrections

@@
 to research the performance of the Monte Carlo algorithm for nonlinear
-problems using the new Forward-Automated Newton-MCSA algorithm
+problems using the new Forward-Automated Newton-MCSA (FANM) algorithm
 developed by this work. Third, a novel parallel algorithm based on
@@
 solve the neutron transport and fluid flow problems on leadership
-class hardware with scaling studies performed using a neutron
-diffusion problem.
+class hardware with scaling studies performed in this work using a
+neutron diffusion problem.
 
@@
 components, will have strong nonlinearities in their behavior and
-would therefore benefit from fully consistent nonlinear solution
-schemes such as that developed by this work instead of fixed-point
-type iterations between physics\footnote{Fixed-point iterations
-  between physics are commonly referred to as Picard iterations in the
+would therefore benefit from fully implicit nonlinear solution schemes
+such as that developed by this work instead of fixed-point type
+iterations between physics\footnote{Fixed-point iterations between
+  physics are commonly referred to as Picard iterations in the
   multiphysics community.}. To leverage such a method for analysis
@@
 nonlinear component of the solution. In the last 30 years, fully
-consistent nonlinear methods based on Newton's method have become more
+implicit nonlinear methods based on Newton's method have become more
 popular and many advances have been made in the nuclear engineering
@@
 subsystem, similar to the same components of the departure from
-nucleate boiling example, was performed as part of analysis of the
+nucleate boiling example, was performed as part of analysis for the
 Department of Energy's Consortium for Advanced Simulation of Light
@@
 comparison to experimental data. Using Drekar, multiphysics
-simulations have been performed with fully consistent methods for the
+simulations have been performed with fully implicit methods for the
 solution of nonlinear systems using meshes of $O(\sn{1}{9})$ elements
@@
 complexity of these problems, if we aim to advance solution techniques
-in neutronics and fluid flow, then we are motivated to advance the
-solution of complex physics problems exploiting leadership class
-levels of parallelism.
+in neutronics and fluid flow for reactor analysis, then we are
+motivated to advance the solution of complex physics problems
+exploiting leadership class levels of parallelism.
 
@@
 nonlinear solution schemes based on Monte Carlo techniques, such
-issues are alleviated by statistical arguments. In the case of soft
-failures, isolated floating point errors in Monte Carlo simulation are
-absorbed within tally statistics while completely losing hardware
-during a hard failure is manifested as a high variance event where
-some portion of the Monte Carlo histories are lost. These stochastic
-methods are a paradigm shift from current deterministic solver
-techniques that will in their current form potentially suffer greatly
-from the non-deterministic behavior expected from the next generation
-of machines.
+issues are potentially alleviated by statistical arguments. In the
+case of soft failures, isolated floating point errors in Monte Carlo
+simulation are absorbed within tally statistics while completely
+losing hardware during a hard failure is manifested as a high variance
+event where some portion of the Monte Carlo histories are lost. These
+stochastic methods are a paradigm shift from current deterministic
+solver techniques that will in their current form potentially suffer
+greatly from the non-deterministic behavior expected from the next
+generation of machines.
 
@@
 typically grow beyond that of the Jacobian. Finally, using MCSA to
-solve the linear model provides its benefits for preconditioning,
-potential parallelism, and performance.
+solve the linear model provides its benefits of potential parallelism
+and iterative performance.
 
@@
 decomposed Monte Carlo Synthetic Acceleration methods. Improvements in
-performance and parallel scalability will be demonstrated in a wide
-variety of numerical experiments using both simple model neutron
-transport problems and neutron transport and fluid flow problems
-generated by production physics codes in both simple and complex
-geometries.
+iterative performance and parallel scalability will be demonstrated in
+a wide variety of numerical experiments using both simple model
+neutron transport problems and neutron transport and fluid flow
+problems generated by production physics codes in both simple and
+complex geometries.
 

### 2013-08-20 doc/dissertation/document/chapters/mc_background/mc_background.tex
adding my corrections for chapter 2

@@
 \begin{equation}
-  \rho(\ve{A}) = \max_{\lambda \in \sigma(\ve{A})} |\lambda| \:.
+  \rho(\ve{A}) = \max_{\lambda \in \sigma(\ve{A})} |\lambda| \:,
   \label{eq:spectral_radius}
 \end{equation}
+which is the radius of the largest circle in the complex plane
+centered about the origin that encapsulates the entire eigenvalue
+spectrum of the operator.
 
@@
 to Eq~(\ref{eq:linear_problem}). For a given linear operator $\ve{A}$,
-we can use diagonal splitting similar to the stationary method in
-Eq~(\ref{eq:linear_split_equation2}) to define the following
-operator\footnote{It should be noted that non-diagonal splittings have
-  been recently explored \cite{srinivasan_monte_2010} and have the
-  potential to improve efficiency. However, it was observed in this
-  work that this type of splitting did not improve performance in the
-  asymptotic limit of $\rho(\ve{H}) \rightarrow 1$ for non-trivial
-  problems.}:
+we can use diagonal splitting\footnote{It should be noted that
+  non-diagonal splittings have been recently explored
+  \cite{srinivasan_monte_2010} and have the potential to improve
+  efficiency. However, it was observed in this work that this type of
+  splitting did not improve performance in the asymptotic limit of
+  $\rho(\ve{H}) \rightarrow 1$ for non-trivial problems.} similar to
+the stationary method in Eq~(\ref{eq:linear_split_equation2}) to
+define the \textit{iteration matrix}, $\ve{H}$:
 \begin{equation}
@@
 \end{equation}
-and is therefore and unbiased estimate. Compared to the collision
+and is therefore an unbiased estimate. Compared to the collision
 estimator, the expected value estimator provides additional
@@
 selecting a more optimal search direction at each iteration,
-potentially further accelerating convergence.
+potentially further improving convergence.
 
@@
 should be considered for many classes of problems, regardless of any
-other preconditioning that is applied to the system.
+other preconditioning that is applied to the system. In addition,
+Jacobi preconditioning has been shown to be an effective
+preconditioning mechanism for MCSA when used with the thermal
+radiation diffusion equation \cite{evans_monte_2012}.
 
@@
 and then recover the original solution vector with
-Eq~(\ref{eq:left_right_recover}). For the Monte Carlo problem, we
-isolate the generation of the correction:
+Eq~(\ref{eq:left_right_recover}) after convergence. For the Monte
+Carlo problem, we isolate the generation of the correction:
 \begin{equation}
@@
 equation as a simple model transport problem\footnote{The Poisson
-  equation is simple form of the diffusion equation and has a similar
-  elliptic character in its time-dependent form to the neutron
+  equation is a simple form of the diffusion equation and has a
+  similar elliptic character in its time-dependent form to the neutron
   transport equations that will be solved in the next chapter.}:
@@
 \times N$ total random walk permutations were computed per MCSA
-iteration where $N \times N$ is the number of discrete grid points in
-the system. Solver parameters were set to a weight cutoff of
-\sn{1}{-4} for the stochastic linear solver and a convergence
-tolerance of \sn{1}{-8} for the MCSA iterative solver.
+iteration where $N \times N$ is the number of discrete grid points (or
+degrees of freedom) in the system. Solver parameters were set to a
+weight cutoff of \sn{1}{-4} for the stochastic linear solver and a
+convergence tolerance of \sn{1}{-8} for the MCSA iterative solver.
 Figure~\ref{fig:poisson_cpu_time} gives the CPU time needed for each
@@
 probability of being terminated through absorption as well as
-transition to other states in the system. This probability, $p_{abs}$,
-modifies the probabilities and weights in the adjoint random walk
-sequence as:
+transitioning to other states in the system. This probability,
+$p_{abs}$, modifies the probabilities and weights in the adjoint
+random walk sequence as:
 \begin{equation}
@@
 of iterations. At a probability of 0.6, the number of iterations
-required grows rapidly as the bias creates an MCSA correction that
-pushes the solution in the wrong direction. Convergence was observed
-to be lost at probabilities of 0.7 and higher.
+required grows rapidly as the shortened randow walks create an MCSA
+correction that has a larger statistical error. Convergence was
+observed to be lost at probabilities of 0.7 and higher.
 

### 2013-08-21 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding edits to spn chapter

@@
 order even moments in this system to zero (e.g. $\phi_6 = \phi_4 = 0$
-yields the $SP_3$ equations).
+yields the $SP_3$ equations). For these equations, the $SP_N$ order
+defines the largest order flux moment maintained in the calculation
+and the $P_N$ order defines the largest order of the cross section
+expansion maintained.
 
@@
 the introduction of fission in the system will only affect the source
-vector in the linear system. The linear operator, and therefore
-overall MCSA performance will be dictated by streaming and scattering
-as defined on the left-hand side.
+vector in the linear system for the eigenvalue schemes used in this
+work. The linear operator, and therefore overall MCSA performance will
+be dictated by streaming and scattering as defined on the left-hand
+side.
 

### 2013-08-21 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
updating the eigenvalue verification results for spn. there was one to many digits reported

@@
       \hline
-      BiCGStab-ILUT & 1 & 25 & 1.1059870 \\
-      GMRES-ILUT & 1 & 25 & 1.1059870 \\
-      MCSA-ILUT-R-C & 1 & 25 & 1.1059870 \\
-      MCSA-ILUT-MR-C & 1 & 25 & 1.1059870 \\
-      MCSA-ILUT-R-EV & 1 & 25 & 1.1059870 \\
-      MCSA-ILUT-MR-EV & 1 & 25 & 1.1059870 \\
+      BiCGStab-ILUT & 1 & 25 & 1.105987 \\
+      GMRES-ILUT & 1 & 25 & 1.105987 \\
+      MCSA-ILUT-R-C & 1 & 25 & 1.105987 \\
+      MCSA-ILUT-MR-C & 1 & 25 & 1.105987 \\
+      MCSA-ILUT-R-EV & 1 & 25 & 1.105987 \\
+      MCSA-ILUT-MR-EV & 1 & 25 & 1.105987 \\
       \hline
-      BiCGStab-ILUT & 2 & 35 & 1.1552390 \\
-      GMRES-ILUT & 2 & 35 & 1.1552390 \\
-      MCSA-ILUT-R-C & 2 & 35 & 1.1552390 \\
-      MCSA-ILUT-MR-C & 2 & 35 & 1.1552390 \\
-      MCSA-ILUT-R-EV & 2 & 35 & 1.1552390 \\
-      MCSA-ILUT-MR-EV & 2 & 35 & 1.1552390 \\
+      BiCGStab-ILUT & 2 & 35 & 1.155239 \\
+      GMRES-ILUT & 2 & 35 & 1.155239 \\
+      MCSA-ILUT-R-C & 2 & 35 & 1.155239 \\
+      MCSA-ILUT-MR-C & 2 & 35 & 1.155239 \\
+      MCSA-ILUT-R-EV & 2 & 35 & 1.155239 \\
+      MCSA-ILUT-MR-EV & 2 & 35 & 1.155239 \\
       \hline
-      BiCGStab-ILUT & 4 & 31 & 1.1758500 \\
-      GMRES-ILUT & 4 & 31 & 1.1758500 \\
-      MCSA-ILUT-R-C & 4 & 31 & 1.1758500 \\
-      MCSA-ILUT-MR-C & 4 & 31 & 1.1758500 \\
-      MCSA-ILUT-R-EV & 4 & 31 & 1.1758500 \\
-      MCSA-ILUT-MR-EV & 4 & 31 & 1.1758500 \\
+      BiCGStab-ILUT & 4 & 31 & 1.175850 \\
+      GMRES-ILUT & 4 & 31 & 1.175850 \\
+      MCSA-ILUT-R-C & 4 & 31 & 1.175850 \\
+      MCSA-ILUT-MR-C & 4 & 31 & 1.175850 \\
+      MCSA-ILUT-R-EV & 4 & 31 & 1.175850 \\
+      MCSA-ILUT-MR-EV & 4 & 31 & 1.175850 \\
       %%

### 2013-08-21 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding relative flux comparison for spn verification

@@
                       & fixed point iteration and expected value estimator \\
+      Richardson-ILUT & Richardson iteration preconditioned with ILUT \\
       %%
@@
       MCSA-ILUT-MR-EV & 1 & 25 & 1.105987 \\
+      Richardson-ILUT & 1 & 25 & 1.105987 \\
       \hline
@@
       MCSA-ILUT-MR-EV & 2 & 35 & 1.155239 \\
+      Richardson-ILUT & 2 & 35 & 1.155239 \\
       \hline
@@
       MCSA-ILUT-MR-EV & 4 & 31 & 1.175850 \\
+      Richardson-ILUT & 4 & 31 & 1.175850 \\
       %%
@@
 energy group was compared over the entire spatial domain with the
-BiCGStab-ILUT solution used as the reference to compute an absolute
+BiCGStab-ILUT solution used as the reference to compute an relative
 difference of the flux in each mesh cell. For the 1 group
-calculations, these absolute differences are plotted spatially along
+calculations, these relative differences are plotted spatially along
 the center-line of the fuel assembly in
@@
 though different spatial results were observed for all cases. For the
-MCSA-ILUT-MR-C calculation, a slightly larger maximum error of
-\sn{2.205}{-9} was observed for reasons that remain unexplained. All
-1-group calculations have a minimum observed absolute difference of 0
-with respect to the BiCGStab calculation.
+MCSA-ILUT-MR-C calculation, a slightly larger maximum relative
+difference of \sn{2.205}{-9} was observed for reasons that remain
+unexplained. All 1-group calculations have a minimum observed relative
+difference of 0 with respect to the BiCGStab calculation.
 
@@
   \end{center}
-  \caption{\textbf{Absolute difference of the group 0 scalar flux
-      between the Aztec BiCGStab calculations and Aztec GMRES and the
-      MCSA variations for the 1-group fuel assembly criticality
-      calculation.} \textit{Left to right: GMRES-ILUT, MCSA-ILUT-R-C,
-      MCSA-ILUT-MR-C, MCSA-ILUT-R-EV, MCSA-ILUT-MR-EV. The fluxes are
-      plotted on a plane intersecting the center-line of the
-      assembly. Table~\ref{tab:spn_solver_defs} gives the description
-      for each solver type presented.}}
+  \caption{\textbf{Relative difference of the group 0 scalar flux
+      between the Aztec BiCGStab calculations, Aztec GMRES, Richardson
+      iteration, and the MCSA variations for the 1-group fuel assembly
+      criticality calculation.} \textit{Left to right: GMRES-ILUT,
+      MCSA-ILUT-R-C, MCSA-ILUT-MR-C, MCSA-ILUT-R-EV,
+      MCSA-ILUT-MR-EV. The fluxes are plotted on a plane intersecting
+      the center-line of the assembly. Table~\ref{tab:spn_solver_defs}
+      gives the description for each solver type presented.}}
   \label{fig:serial_1_group_diff}
@@
 Table~\ref{tab:serial_differences_g4} give the maximum and minimum
-absolute differences of the cell scalar fluxes for all calculations
+relative differences of the cell scalar fluxes for all calculations
 and all energy groups using the BiCGStab calculation as the
@@
 and spatial domain are within these limits. In addition, the minimum
-error for all calculations was observed to be zero, meaning that at
-least one mesh cell in all calculations for all energy groups the
-scalar flux values computed by all solvers were identical. Based on
-the flux results and the eigenvalue results presented here, the MCSA
-results should be deemed correct for the fuel assembly problem for
-serial calculations with multiple energy groups.
+relative difference for all calculations was observed to be zero,
+meaning that at least one mesh cell in all calculations for all energy
+groups the scalar flux values computed by all solvers were
+identical. Based on the flux results and the eigenvalue results
+presented here, the MCSA results should be deemed correct for the fuel
+assembly problem for serial calculations with multiple energy groups.
 
@@
       \hline
-      GMRES-ILUT & 0 & 0.0 & \sn{9.313}{-10} \\
-      MCSA-ILUT-R-C & 0 & 0.0 & \sn{9.313}{-10} \\
-      MCSA-ILUT-MR-C & 0 & 0.0 & \sn{2.205}{-9} \\
-      MCSA-ILUT-R-EV & 0 & 0.0 & \sn{9.313}{-10} \\
-      MCSA-ILUT-MR-EV & 0 & 0.0 & \sn{9.313}{-10} \\
+      GMRES-ILUT & 0 & 0.0 & \sn{4.562}{-6} \\
+      MCSA-ILUT-R-C & 0 & 0.0 & \sn{2.099}{-5} \\
+      MCSA-ILUT-MR-C & 0 & 0.0 & \sn{4.812}{-5} \\
+      MCSA-ILUT-R-EV & 0 & 0.0 & \sn{2.371}{-5} \\
+      MCSA-ILUT-MR-EV & 0 & 0.0 & \sn{3.605}{-5} \\
+      Richardson-ILUT & 0 & 0.0 & \sn{1.531}{-5} \\
       \hline\hline
@@
       1-group fuel assembly problem.} \textit{The maximum and minimum
-      absolute differences over the entire spatial domain are
+      relative differences over the entire spatial domain are
       presented with BiCGStab used as the reference. Each calculation
@@
       \hline
-      GMRES-ILUT & 0 & 0.0 & \sn{2.794}{-9} \\
-      MCSA-ILUT-R-C & 0 & 0.0 & \sn{9.313}{-10} \\
-      MCSA-ILUT-MR-C & 0 & 0.0 & \sn{9.313}{-10} \\
-      MCSA-ILUT-R-EV & 0 & 0.0 & \sn{9.313}{-10} \\
-      MCSA-ILUT-MR-EV & 0 & 0.0 & \sn{9.313}{-10} \\
+      GMRES-ILUT & 0 & 0.0 & \sn{7.590}{-6} \\
+      MCSA-ILUT-R-C & 0 & 0.0 & \sn{5.308}{-6} \\
+      MCSA-ILUT-MR-C & 0 & 0.0 & \sn{6.387}{-5} \\
+      MCSA-ILUT-R-EV & 0 & 0.0 & \sn{4.197}{-5} \\
+      MCSA-ILUT-MR-EV & 0 & 0.0 & \sn{2.736}{-5} \\
+      Richardson-ILUT & 0 & 0.0 & \sn{3.976}{-6} \\
       \hline
-      GMRES-ILUT & 1 & 0.0 & \sn{4.657}{-10} \\
-      MCSA-ILUT-R-C & 1 & 0.0 & \sn{1.364}{-10} \\
-      MCSA-ILUT-MR-C & 1 & 0.0 & \sn{1.261}{-9} \\
-      MCSA-ILUT-R-EV & 1 & 0.0 & \sn{2.474}{-10} \\
-      MCSA-ILUT-MR-EV & 1 & 0.0 & \sn{5.857}{-10} \\
+      GMRES-ILUT & 1 & 0.0 & \sn{7.579}{-6} \\
+      MCSA-ILUT-R-C & 1 & 0.0 & \sn{1.645}{-5} \\
+      MCSA-ILUT-MR-C & 1 & 0.0 & \sn{2.634}{-4} \\
+      MCSA-ILUT-R-EV & 1 & 0.0 & \sn{1.797}{-4} \\
+      MCSA-ILUT-MR-EV & 1 & 0.0 & \sn{1.556}{-4} \\
+      Richardson-ILUT & 1 & 0.0 & \sn{4.031}{-6} \\
       \hline\hline
@@
       2-group fuel assembly problem.} \textit{The maximum and minimum
-      absolute differences over the entire spatial domain are
+      relative differences over the entire spatial domain are
       presented with BiCGStab used as the reference. Each calculation
@@
       \hline
-      GMRES-ILUT & 0 & 0.0 & \sn{9.313}{-10} \\
-      MCSA-ILUT-R-C & 0 & 0.0 & \sn{1.397}{-9} \\
-      MCSA-ILUT-MR-C & 0 & 0.0 & \sn{1.870}{-9} \\
-      MCSA-ILUT-R-EV & 0 & 0.0 & \sn{1.397}{-9} \\
-      MCSA-ILUT-MR-EV & 0 & 0.0 & \sn{1.869}{-9} \\
+      GMRES-ILUT & 0 & 0.0 & \sn{2.376}{-5} \\
+      MCSA-ILUT-R-C & 0 & 0.0 & \sn{2.639}{-5} \\
+      MCSA-ILUT-MR-C & 0 & 0.0 & \sn{3.958}{-5} \\
+      MCSA-ILUT-R-EV & 0 & 0.0 & \sn{2.749}{-5} \\
+      MCSA-ILUT-MR-EV & 0 & 0.0 & \sn{2.969}{-5} \\
+      Richardson-ILUT & 0 & 0.0 & \sn{2.189}{-5} \\
       \hline
-      GMRES-ILUT & 1 & 0.0 & \sn{5.821}{-11} \\
-      MCSA-ILUT-R-C & 1 & 0.0 & \sn{1.164}{-10} \\
-      MCSA-ILUT-MR-C & 1 & 0.0 & \sn{1.164}{-10} \\
-      MCSA-ILUT-R-EV & 1 & 0.0 & \sn{1.164}{-10} \\
-      MCSA-ILUT-MR-EV & 1 & 0.0 & \sn{1.164}{-10} \\
+      GMRES-ILUT & 0 & 0.0 & \sn{2.271}{-5} \\
+      MCSA-ILUT-R-C & 0 & 0.0 & \sn{3.702}{-5} \\
+      MCSA-ILUT-MR-C & 0 & 0.0 & \sn{3.803}{-5} \\
+      MCSA-ILUT-R-EV & 0 & 0.0 & \sn{4.142}{-5} \\
+      MCSA-ILUT-MR-EV & 0 & 0.0 & \sn{2.723}{-5} \\
+      Richardson-ILUT & 0 & 0.0 & \sn{2.137}{-5} \\
       \hline
-      GMRES-ILUT & 2 & 0.0 & \sn{1.164}{-10} \\
-      MCSA-ILUT-R-C & 2 & 0.0 & \sn{2.328}{-10} \\
-      MCSA-ILUT-MR-C & 2 & 0.0 & \sn{2.328}{-10} \\
-      MCSA-ILUT-R-EV & 2 & 0.0 & \sn{2.328}{-10} \\
-      MCSA-ILUT-MR-EV & 2 & 0.0 & \sn{2.328}{-10} \\
+      GMRES-ILUT & 0 & 0.0 & \sn{2.310}{-5} \\
+      MCSA-ILUT-R-C & 0 & 0.0 & \sn{2.490}{-5} \\
+      MCSA-ILUT-MR-C & 0 & 0.0 & \sn{2.214}{-5} \\
+      MCSA-ILUT-R-EV & 0 & 0.0 & \sn{4.456}{-5} \\
+      MCSA-ILUT-MR-EV & 0 & 0.0 & \sn{3.366}{-5} \\
+      Richardson-ILUT & 0 & 0.0 & \sn{2.143}{-5} \\
       \hline
-      GMRES-ILUT & 3 & 0.0 & \sn{1.164}{-10} \\
-      MCSA-ILUT-R-C & 3 & 0.0 & \sn{1.746}{-10} \\
-      MCSA-ILUT-MR-C & 3 & 0.0 & \sn{2.256}{-9} \\
-      MCSA-ILUT-R-EV & 3 & 0.0 & \sn{2.801}{-10} \\
-      MCSA-ILUT-MR-EV & 3 & 0.0 & \sn{2.256}{-10} \\
+      GMRES-ILUT & 0 & 0.0 & \sn{2.335}{-5} \\
+      MCSA-ILUT-R-C & 0 & 0.0 & \sn{3.045}{-5} \\
+      MCSA-ILUT-MR-C & 0 & 0.0 & \sn{2.696}{-5} \\
+      MCSA-ILUT-R-EV & 0 & 0.0 & \sn{1.946}{-4} \\
+      MCSA-ILUT-MR-EV & 0 & 0.0 & \sn{3.295}{-5} \\
+      Richardson-ILUT & 0 & 0.0 & \sn{2.161}{-5} \\
       \hline\hline
@@
       4-group fuel assembly problem.} \textit{The maximum and minimum
-      absolute differences over the entire spatial domain are
+      relative differences over the entire spatial domain are
       presented with BiCGStab used as the reference. Each calculation
@@
       first decimal place. Table~\ref{tab:spn_solver_defs} gives the
-      description for each solver type presented and Richardson-ILUT
-      is a ILUT preconditioned Richardson iteration.}}
+      description for each solver type presented.}}
   \label{tab:spn_comparison_iterations}

### 2013-08-21 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
removing spn point-wise verification in favor over difference vector norm

@@
 \end{multline}
-yielding a multidimensional set of $(N+1)/1$ angular coupled equations
+yielding a multidimensional set of $(N+1)/2$ angular coupled equations
 defined as the $SP_N$ equations. As with the $P_N$ equations, we
@@
 \end{multline}
-with the fission matrix $\mathbf{F}$ given by
-\cite{evans_simplified_2013}. We again apply the change of variables to
-yield the set of multigroup pseudo-moments $\mathbb{U}_n$ where now
+with $\mathbf{\Phi_n}$ the vector of multigroup flux moments and the
+fission matrix, $\mathbf{F}$, given by
+\cite{evans_simplified_2013}. We again apply the change of variables
+to yield the set of multigroup pseudo-moments $\mathbb{U}_n$ where now
 the application of the fission matrix to the moment vectors will be
@@
 to this diagonal form and the success achieved by applying it to MCSA
-solutions of the radiation diffusion equation as presented in
+solutions of the thermal radiation diffusion equation as presented in
 \cite{evans_monte_2012}.
@@
 Spectral radius computations were performed for the cases described
-above for the point Jacobi preconditioned iteration matrix $\ve{H} =
-\ve{I}-\ve{M}^{-1}\ve{A}$ with $\ve{M} =
+above for the point Jacobi preconditioned iteration matrix, $\ve{H} =
+\ve{I}-\ve{M}^{-1}\ve{A}$, with $\ve{M} =
 diag(\ve{A})$. Table~\ref{tab:group1pj} gives the results for the
@@
 as well as the multigroup $SP_N$ equations. Initially, the diagonal
-character of the system led us to try point Jacobi preconditioning
-with only marginal results. From the sparsity plots we note the block
-structure that ultimately arises from the multigroup scattering
-matrices and their insertion into Eq~(\ref{eq:A_block_matrix}). When
-full upscatter and downscatter are used the resulting blocks are
-completely dense while only downscatter gives a lower triangular
-scattering matrix and the block structure shown in
-Figure~\ref{fig:group10ds}.
+character of the system and previous success with the thermal
+radiuation difussion equation led us to try point Jacobi
+preconditioning with only marginal results. From the sparsity plots we
+note the block structure that ultimately arises from the multigroup
+scattering matrices and their insertion into
+Eq~(\ref{eq:A_block_matrix}). When full upscatter and downscatter are
+used the resulting blocks are completely dense while only downscatter
+gives a lower triangular scattering matrix and the block structure
+shown in Figure~\ref{fig:group10ds}.
 
@@
 \end{figure}
-The inversion of this preconditioner is trivial as shown on the right
-side of Figure~\ref{fig:block_jacobi_ex}. Each block can be inverted
-individually and combined to form the inverse. In addition, in the
-limit of a block size of one, the block Jacobi method reduces to the
-point Jacobi method.
+In addition, in the limit of a block size of one, the block Jacobi
+method reduces to the point Jacobi method.
 
@@
 and therefore we expect the matrix elements forming the block to be
-entirely local. Second, these blocks are typically dense and nearly
-lower triangular for many transport problems meaning that established
-dense matrix methods can be used for fast inversion.
+entirely local in domain decomposed calculations. Second, these blocks
+are typically dense and nearly lower triangular for many transport
+problems meaning that established dense matrix methods can be used for
+fast inversion.
 
@@
 From the tabulated block Jacobi data it is clear that this is a viable
-preconditioning choice for all $SP_N$ problems in term of Monte Carlo
+preconditioning choice for all $SP_N$ problems in terms of Monte Carlo
 solution methods. All cases were observed to have a spectral radius
@@
 first preconditioning strategy applied to more complicated problems
-representative of physical systems.
+representative of physical nuclear systems.
 
@@
   $\mathbf{M}$.} while the addition of fission to the system will only
-modify the source of neutrons and the multiplication factor while not
-affecting Monte Carlo transport.
+modify the source of neutrons and the multiplication factor at every
+eigenvalue iteration while not affecting Monte Carlo transport.
 
@@
 behavior at a significant computational cost and compute time. Even if
-the problem may be ill-conditioned, we do expect convergence of
-MCSA. To investigate further, a block Jacobi preconditioned Richardson
-iteration was used to solve the same
+the problem may be ill-conditioned, we do expect convergence of MCSA
+given that the spectral radius requirement was met. To investigate
+further, a block Jacobi preconditioned Richardson iteration was used
+to solve the same
 problem. Figure~\ref{fig:block_jacobi_res_richardson} gives the
@@
 \end{figure}
-Poor converge is observed for the Richardson iteration, however,
+Poor convergence is observed for the Richardson iteration, however,
 convergence is achieved meaning that the preliminary eigenvalue
@@
   \text{Quality Metric} = (\text{\# iterations}) \times (\text{maximum
-    \# of non-zero values})\:,
+    \# non-zero values})\:,
 \end{equation}
@@
 fuel assembly problem recorded along with the maximum number of
-non-zero entries observed in all matrix rows for the left/right
-preconditioned composite linear operator. For each calculation, the
-number of iterations required to converge reported was for a single
-eigenvalue iteration with \sn{3}{4} histories at each MCSA iteration
-to compute the Monte Carlo correction using the adjoint collision
-estimator. All SPAINV calculations reported here were performed on a
-single CPU.
+non-zero entries observed in all matrix rows for the preconditioned
+composite linear operator. For each calculation, the number of
+iterations required to converge reported was for a single eigenvalue
+iteration with \sn{3}{4} histories at each MCSA iteration to compute
+the Monte Carlo correction using the adjoint collision estimator. All
+SPAINV calculations reported here were performed on a single CPU.
 
@@
 \sn{1}{-10} in order to eliminate any exceedingly small values from
-the matrix (generally this is simply removing non-zero elements within
-the floating point tolerance of zero). The reduced domain fill level
-was then varied, starting with the largest non-zero entries per row
-value observed for each of the preconditioning types in order to
-assess its effects relative to the case where no reduced domain
-approximation was applied.
+the matrix (often this is simply removing non-zero elements within the
+floating point tolerance of zero). The reduced domain fill level was
+then varied, starting with the largest non-zero entries per row value
+observed for each of the preconditioning types in order to assess its
+effects relative to the case where no reduced domain approximation was
+applied.
 
@@
 this addition of fill rather than the extra SPAINV
-preconditioning. Looking at the quality metric data, we see a nice
-power-law reduction in the quality metric as a function of the reduced
-domain fill level, achieving 2 orders of magnitude reduction in the
-quality metric for the ILUT-based preconditioning methods.
+preconditioning. Looking at the quality metric data, we see a
+reduction in the quality metric as a function of the reduced domain
+fill level, achieving 2 orders of magnitude reduction in the quality
+metric for the ILUT-based preconditioning methods.
 
@@
 \label{subsec:spn_mcsa_relaxation}
-As another means of preconditioning (or variance reduction), the
-Richardson iteration upon which the Neumann-Ulam method is built may
-be implemented with a scalar relaxation parameter that can be adjusted
-to improve convergence:
+As another means of preconditioning (or variance reduction) to meet
+the goal of improving iterative performance, the Richardson iteration
+upon which the Neumann-Ulam method is built may be implemented with an
+adjustable scalar relaxation parameter:
 \begin{equation}
@@
 where $\omega$ is the relaxation parameter. This is very similar to
-point Jacobi preconditioning where the system is being scaled on the
-left by a constant value in all rows. Analogously, these relaxation
-parameter techniques can be applied to Monte Carlo to improve
-convergence as demonstrated by Dimov \cite{dimov_new_1998}. In this
-case, building an iteration matrix from
+point Jacobi preconditioning where the system is now being scaled on
+the left by a constant value in all rows. Analogously, these
+relaxation parameter techniques can be applied to Monte Carlo to
+improve convergence as demonstrated by Dimov \cite{dimov_new_1998}. In
+this case, building an iteration matrix from
 Eq~(\ref{eq:richardson_relaxation}) gives:
@@
 with the probabilities and weights for the Monte Carlo procedure
-appropriately scaled. By inspection, such a scaling is a simple form
-of preconditioning on the left where all rows in the system are scaled
-by the same scalar parameter. For MCSA, we can stage this scheme with
-two separate relaxation parameters; one for the outer Richardson
-iteration and one for the inner Monte Carlo solve:
+appropriately scaled via a modified Neumann-Ulam decomposition. For
+MCSA, we can stage this scheme with two separate relaxation
+parameters; one for the outer Richardson iteration and one for the
+inner Monte Carlo solve:
 \begin{subequations}
@@
 
-\subsection{Monte Carlo Estimator Comparison}
+\subsection{MCSA Estimator and Fixed Point Iteration Analysis}
 \label{subsec:spn_estimator_comparison}
-With convergence obtained for the fuel assembly problem, the advanced
-preconditioners will next be studied with both the collision and
+With convergence obtained for the fuel assembly problem, the iterative
+performance of MCSA will next be studied with both the collision and
 expected value estimators using the best preconditioning and
@@
 analysis. Additionally, both the Richardson-based MCSA iteration given
-by Eq~(\ref{eq:mcsa}) and the minimal residual-based MCSA iteration given by
-Eq~(\ref{eq:mcsa_min_res}) will be used. Although the expected value
-estimator achieved the best iterative performance in the previous
-comparison and analysis for the simple Poisson problem in
+by Eq~(\ref{eq:mcsa}) and the minimal residual-based MCSA iteration
+given by Eq~(\ref{eq:mcsa_min_res}) will be used. Although the
+expected value estimator achieved the best iterative performance in
+the previous comparison and analysis for the simple Poisson problem in
 \S~\ref{sec:parameter_estimator_analysis}, it is possible that the
@@
 behavior. Using the expected value estimator with both fixed point
-iterations creates a relative insensitivity to the number of histories
-used to compute the correction. Marginally better iterative and timing
-performance was observed for the expected value estimator when using
-the Richardson iteration. Better timing performance is expected as
-computing the extrapolation parameter in the minimal residual
-iteration requires several additional parallel operations. However, we
-expect that because the minimal residual iteration will converge
-faster on its own when compared to the Richardson iteration that
-better iterative performance will be observed when it is accelerated
-with MCSA. The fact that this was not observed signals that an MCSA
-scheme leveraging the expected value estimator is also effectively
-insensitive to the fixed point iteration used. For the collision
-estimator, the minimal residual iteration permits fewer stochastic
-histories to be used with each MCSA iteration while still maintaining
-good iterative performance. With respect to CPU time, there is a more
-pronounced effect due to the larger density of states created by the
-explicit ILUT preconditioning. Using a reduced domain fill level of
-100 creates a situation where the expected value estimator is
-significantly more expensive per history to compute that the collision
-estimator due to the coupling of states.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/estimator_iters.pdf}
+iterations again creates a relative insensitivity to the number of
+histories used to compute the correction. Marginally better iterative
+and timing performance was observed for the expected value estimator
+when using the Richardson iteration. Better timing performance is
+expected as computing the extrapolation parameter in the minimal
+residual iteration requires several additional parallel
+operations. Compared to the collision estimator, the minimal residual
+iteration permits fewer stochastic histories to be used with each MCSA
+iteration while still maintaining good iterative performance. With
+respect to CPU time, there is a more pronounced effect due to the
+larger density of states created by the explicit ILUT
+preconditioning. Using a reduced domain fill level of 100 creates a
+situation where the expected value estimator is significantly more
+expensive per history to compute that the collision estimator due to
+the coupling of states.
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=5in]{chapters/spn_equations/estimator_iters.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/estimator_time.pdf}
+    \includegraphics[width=5in]{chapters/spn_equations/estimator_time.pdf}
   \end{center}
@@
 comparison. At this number of histories, all MCSA combinations
-converge the fuel assembly problem monotonically at and approximately
-the same rate, giving little reason to choose one over the other. The
-end result is that both estimators provide a viable method for getting
+converge the fuel assembly problem monotonically at approximately the
+same rate, giving little reason to choose one over the other. The end
+result is that both estimators provide a viable method for getting
 good MCSA iterative performance with comparable timing
@@
   \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/estimator_convergence.pdf}
+    \includegraphics[width=5in]{chapters/spn_equations/estimator_convergence.pdf}
   \end{center}
@@
 the dense composite linear operators that arise when performing
-explicit MCSA preconditioning in order to achieve convergence. Using
-these results, we compare MCSA directly to conventional linear solvers
-that would be typically used to solve the $SP_N$ form of the transport
-equation as presented here. In particular, we will use a pair of
-production Krylov solvers with which we will compare direct numerical
-results in order to verify MCSA solutions in this section.
+explicit MCSA preconditioning necessary for achieve convergence. Using
+these results, we know compare MCSA directly to conventional linear
+solvers that would be typically used to solve the $SP_N$ form of the
+transport equation as presented here. In particular, we will use a
+pair of production Krylov solvers with which we will compare direct
+numerical results in order to verify MCSA solutions in this section.
 
@@
 problem with 1, 2, and 4 energy groups using $SP_1$ discretization
-will be directly compared over all group and cell fluxes computed by
-converging the criticality problem using both BiCGStab and GMRES
-production implementations from the Trilinos scientific computing
-library Aztec \cite{heroux_overview_2005}. In addition to comparing
-the cell and group fluxes, the number of eigenvalue iterations
-required to converge the problem along with the eigenvalue itself will
-be compared as a means of verification. If all values computed using
-MCSA agree with the results from the production Krylov solvers within
-the solution tolerance, then the MCSA solutions will be deemed correct
-for these problems.
+will be directly compared point-wise over all group and cell fluxes
+computed by converging the criticality problem using both BiCGStab and
+GMRES production implementations from the Trilinos scientific
+computing library Aztec \cite{heroux_overview_2005}. In addition to
+comparing the cell and group fluxes, the number of eigenvalue
+iterations required to converge the problem along with the eigenvalue
+itself will be compared as a means of verification. For the flux
+values, we look to the 2-norm of the absolute difference between the
+reference case and the comparison case in each energy group:
+\begin{equation}
+  || \ve{\Phi}_{ref}^g - \ve{\Phi}^g ||_2 < \epsilon_k\:,
+  \label{eq:norm_comparison}
+\end{equation}
+such that this norm is less than the eigensolver tolerance,
+$\epsilon_k$. For this work, the BiCGStab results will be taken as the
+reference case. If the MCSA flux results satisfy
+Eq~(\ref{eq:norm_comparison}), they will be deemed correct for the
+purposes of this work. For the eigenvalues, we will look for the same
+agreement within the tolerance of the eigensolver.
 
@@
 eigenvalue problem was converged with an eigensolver tolerance of
-\sn{1}{-6} and a linear solver tolerance of \sn{1}{-8} in a serial
-calculation. Table~\ref{tab:serial_ev_results} gives the eigenvalue
-computed by each solver for each problem as well as the number of
-eigenvalue iterations required to converge. For every problem, each
-solver converged to the same eigenvalue within the tolerance of the
-eigensolver in the same number of eigenvalue iterations.
+$\epsilon_k = \sn{1}{-6}$ and a linear solver tolerance of \sn{1}{-8}
+in a serial calculation. Table~\ref{tab:serial_ev_results} gives the
+eigenvalue computed by each solver for each problem as well as the
+number of eigenvalue iterations required to converge. For every
+problem, each solver converged to the same eigenvalue up to the number
+of digits reported by the eigensolver eigensolver in the same number
+of eigenvalue iterations.
 \begin{table}[h!]
@@
 
-For each converged eigenvalue calculation, the scalar flux in each
-energy group was compared over the entire spatial domain with the
-BiCGStab-ILUT solution used as the reference to compute an relative
-difference of the flux in each mesh cell. For the 1 group
-calculations, these relative differences are plotted spatially along
-the center-line of the fuel assembly in
-Figure~\ref{fig:serial_1_group_diff}. Interestingly, except for the
-MCSA-ILUT-MR-C calculation, all other 1-group calculations including
-GMRES had a maximum global flux difference of \sn{9.313}{-10} even
-though different spatial results were observed for all cases. For the
-MCSA-ILUT-MR-C calculation, a slightly larger maximum relative
-difference of \sn{2.205}{-9} was observed for reasons that remain
-unexplained. All 1-group calculations have a minimum observed relative
-difference of 0 with respect to the BiCGStab calculation.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/serial_1_group.pdf}
-  \end{center}
-  \caption{\textbf{Relative difference of the group 0 scalar flux
-      between the Aztec BiCGStab calculations, Aztec GMRES, Richardson
-      iteration, and the MCSA variations for the 1-group fuel assembly
-      criticality calculation.} \textit{Left to right: GMRES-ILUT,
-      MCSA-ILUT-R-C, MCSA-ILUT-MR-C, MCSA-ILUT-R-EV,
-      MCSA-ILUT-MR-EV. The fluxes are plotted on a plane intersecting
-      the center-line of the assembly. Table~\ref{tab:spn_solver_defs}
-      gives the description for each solver type presented.}}
-  \label{fig:serial_1_group_diff}
-\end{figure}
-
 Table~\ref{tab:serial_differences_g1},
 Table~\ref{tab:serial_differences_g2}, and
-Table~\ref{tab:serial_differences_g4} give the maximum and minimum
-relative differences of the cell scalar fluxes for all calculations
-and all energy groups using the BiCGStab calculation as the
-reference. All comparisons were performed within a group such that the
+Table~\ref{tab:serial_differences_g4} give the 2-norms of the cell
+scalar fluxes difference vectors for all calculations and all energy
+groups. All comparisons were performed within a group such that the
 differences for the MCSA-ILUT-R-C results in group 3 for the 4-group
@@
 eigenvalue iteration meaning that all values over the entire energy
-and spatial domain are within these limits. In addition, the minimum
-relative difference for all calculations was observed to be zero,
-meaning that at least one mesh cell in all calculations for all energy
-groups the scalar flux values computed by all solvers were
-identical. Based on the flux results and the eigenvalue results
-presented here, the MCSA results should be deemed correct for the fuel
-assembly problem for serial calculations with multiple energy groups.
+and spatial domain are within these limits. Based on the flux results
+and the eigenvalue results presented here, the MCSA results should be
+deemed correct for the fuel assembly problem for serial calculations
+with multiple energy groups.
 
@@
       \multicolumn{1}{c}{\textbf{Group}} &
-      \multicolumn{1}{c}{\textbf{Minimum Difference}} &
-      \multicolumn{1}{c}{\textbf{Maximum Difference}} \\
+      \multicolumn{1}{c}{\textbf{$|| \ve{\Phi}_{ref}^g - \ve{\Phi}^g ||_2$}} \\
       \hline
-      GMRES-ILUT & 0 & 0.0 & \sn{4.562}{-6} \\
-      MCSA-ILUT-R-C & 0 & 0.0 & \sn{2.099}{-5} \\
-      MCSA-ILUT-MR-C & 0 & 0.0 & \sn{4.812}{-5} \\
-      MCSA-ILUT-R-EV & 0 & 0.0 & \sn{2.371}{-5} \\
-      MCSA-ILUT-MR-EV & 0 & 0.0 & \sn{3.605}{-5} \\
-      Richardson-ILUT & 0 & 0.0 & \sn{1.531}{-5} \\
+      GMRES-ILUT & 0 & \sn{4.562}{-6} \\
+      MCSA-ILUT-R-C & 0 & \sn{2.099}{-5} \\
+      MCSA-ILUT-MR-C & 0 & \sn{4.812}{-5} \\
+      MCSA-ILUT-R-EV & 0 & \sn{2.371}{-5} \\
+      MCSA-ILUT-MR-EV & 0 & \sn{3.605}{-5} \\
+      Richardson-ILUT & 0 & \sn{1.531}{-5} \\
       \hline\hline
@@
   \caption{\textbf{Serial scalar flux verification results for the
-      1-group fuel assembly problem.} \textit{The maximum and minimum
-      relative differences over the entire spatial domain are
-      presented with BiCGStab used as the reference. Each calculation
-      was converged with the same eigensolver and linear solver
+      1-group fuel assembly problem.} \textit{BiCGStab is used as the
+      reference. Each calculation was converged with the same
+      eigensolver and linear solver
       tolerances. Table~\ref{tab:spn_solver_defs} gives the
@@
       \multicolumn{1}{c}{\textbf{Group}} &
-      \multicolumn{1}{c}{\textbf{Minimum Difference}} &
-      \multicolumn{1}{c}{\textbf{Maximum Difference}} \\
+      \multicolumn{1}{c}{\textbf{$|| \ve{\Phi}_{ref}^g - \ve{\Phi}^g ||_2$}} \\
       \hline
-      GMRES-ILUT & 0 & 0.0 & \sn{7.590}{-6} \\
-      MCSA-ILUT-R-C & 0 & 0.0 & \sn{5.308}{-6} \\
-      MCSA-ILUT-MR-C & 0 & 0.0 & \sn{6.387}{-5} \\
-      MCSA-ILUT-R-EV & 0 & 0.0 & \sn{4.197}{-5} \\
-      MCSA-ILUT-MR-EV & 0 & 0.0 & \sn{2.736}{-5} \\
-      Richardson-ILUT & 0 & 0.0 & \sn{3.976}{-6} \\
+      GMRES-ILUT & 0 & \sn{7.590}{-6} \\
+      MCSA-ILUT-R-C & 0 & \sn{5.308}{-6} \\
+      MCSA-ILUT-MR-C & 0 & \sn{6.387}{-5} \\
+      MCSA-ILUT-R-EV & 0 & \sn{4.197}{-5} \\
+      MCSA-ILUT-MR-EV & 0 & \sn{2.736}{-5} \\
+      Richardson-ILUT & 0 & \sn{3.976}{-6} \\
       \hline
-      GMRES-ILUT & 1 & 0.0 & \sn{7.579}{-6} \\
-      MCSA-ILUT-R-C & 1 & 0.0 & \sn{1.645}{-5} \\
-      MCSA-ILUT-MR-C & 1 & 0.0 & \sn{2.634}{-4} \\
-      MCSA-ILUT-R-EV & 1 & 0.0 & \sn{1.797}{-4} \\
-      MCSA-ILUT-MR-EV & 1 & 0.0 & \sn{1.556}{-4} \\
-      Richardson-ILUT & 1 & 0.0 & \sn{4.031}{-6} \\
+      GMRES-ILUT & 1 & \sn{7.579}{-6} \\
+      MCSA-ILUT-R-C & 1 & \sn{1.645}{-5} \\
+      MCSA-ILUT-MR-C & 1 & \sn{2.634}{-4} \\
+      MCSA-ILUT-R-EV & 1 & \sn{1.797}{-4} \\
+      MCSA-ILUT-MR-EV & 1 & \sn{1.556}{-4} \\
+      Richardson-ILUT & 1 & \sn{4.031}{-6} \\
       \hline\hline
@@
   \caption{\textbf{Serial scalar flux verification results for the
-      2-group fuel assembly problem.} \textit{The maximum and minimum
-      relative differences over the entire spatial domain are
-      presented with BiCGStab used as the reference. Each calculation
-      was converged with the same eigensolver and linear solver
+      2-group fuel assembly problem.} \textit{BiCGStab is used as the
+      reference. Each calculation was converged with the same
+      eigensolver and linear solver
       tolerances. Table~\ref{tab:spn_solver_defs} gives the
@@
       \multicolumn{1}{c}{\textbf{Group}} &
-      \multicolumn{1}{c}{\textbf{Minimum Difference}} &
-      \multicolumn{1}{c}{\textbf{Maximum Difference}} \\
+      \multicolumn{1}{c}{\textbf{$|| \ve{\Phi}_{ref}^g - \ve{\Phi}^g ||_2$}} \\
       \hline
-      GMRES-ILUT & 0 & 0.0 & \sn{2.376}{-5} \\
-      MCSA-ILUT-R-C & 0 & 0.0 & \sn{2.639}{-5} \\
-      MCSA-ILUT-MR-C & 0 & 0.0 & \sn{3.958}{-5} \\
-      MCSA-ILUT-R-EV & 0 & 0.0 & \sn{2.749}{-5} \\
-      MCSA-ILUT-MR-EV & 0 & 0.0 & \sn{2.969}{-5} \\
-      Richardson-ILUT & 0 & 0.0 & \sn{2.189}{-5} \\
+      GMRES-ILUT & 0 & \sn{2.376}{-5} \\
+      MCSA-ILUT-R-C & 0 & \sn{2.639}{-5} \\
+      MCSA-ILUT-MR-C & 0 & \sn{3.958}{-5} \\
+      MCSA-ILUT-R-EV & 0 & \sn{2.749}{-5} \\
+      MCSA-ILUT-MR-EV & 0 & \sn{2.969}{-5} \\
+      Richardson-ILUT & 0 & \sn{2.189}{-5} \\
       \hline
-      GMRES-ILUT & 0 & 0.0 & \sn{2.271}{-5} \\
-      MCSA-ILUT-R-C & 0 & 0.0 & \sn{3.702}{-5} \\
-      MCSA-ILUT-MR-C & 0 & 0.0 & \sn{3.803}{-5} \\
-      MCSA-ILUT-R-EV & 0 & 0.0 & \sn{4.142}{-5} \\
-      MCSA-ILUT-MR-EV & 0 & 0.0 & \sn{2.723}{-5} \\
-      Richardson-ILUT & 0 & 0.0 & \sn{2.137}{-5} \\
+      GMRES-ILUT & 1 & \sn{2.271}{-5} \\
+      MCSA-ILUT-R-C & 1 & \sn{3.702}{-5} \\
+      MCSA-ILUT-MR-C & 1 & \sn{3.803}{-5} \\
+      MCSA-ILUT-R-EV & 1 & \sn{4.142}{-5} \\
+      MCSA-ILUT-MR-EV & 1 & \sn{2.723}{-5} \\
+      Richardson-ILUT & 1 & \sn{2.137}{-5} \\
       \hline
-      GMRES-ILUT & 0 & 0.0 & \sn{2.310}{-5} \\
-      MCSA-ILUT-R-C & 0 & 0.0 & \sn{2.490}{-5} \\
-      MCSA-ILUT-MR-C & 0 & 0.0 & \sn{2.214}{-5} \\
-      MCSA-ILUT-R-EV & 0 & 0.0 & \sn{4.456}{-5} \\
-      MCSA-ILUT-MR-EV & 0 & 0.0 & \sn{3.366}{-5} \\
-      Richardson-ILUT & 0 & 0.0 & \sn{2.143}{-5} \\
+      GMRES-ILUT & 2 & \sn{2.310}{-5} \\
+      MCSA-ILUT-R-C & 2 & \sn{2.490}{-5} \\
+      MCSA-ILUT-MR-C & 2 & \sn{2.214}{-5} \\
+      MCSA-ILUT-R-EV & 2 & \sn{4.456}{-5} \\
+      MCSA-ILUT-MR-EV & 2 & \sn{3.366}{-5} \\
+      Richardson-ILUT & 2 & \sn{2.143}{-5} \\
       \hline
-      GMRES-ILUT & 0 & 0.0 & \sn{2.335}{-5} \\
-      MCSA-ILUT-R-C & 0 & 0.0 & \sn{3.045}{-5} \\
-      MCSA-ILUT-MR-C & 0 & 0.0 & \sn{2.696}{-5} \\
-      MCSA-ILUT-R-EV & 0 & 0.0 & \sn{1.946}{-4} \\
-      MCSA-ILUT-MR-EV & 0 & 0.0 & \sn{3.295}{-5} \\
-      Richardson-ILUT & 0 & 0.0 & \sn{2.161}{-5} \\
+      GMRES-ILUT & 3 & \sn{2.335}{-5} \\
+      MCSA-ILUT-R-C & 3 & \sn{3.045}{-5} \\
+      MCSA-ILUT-MR-C & 3 & \sn{2.696}{-5} \\
+      MCSA-ILUT-R-EV & 3 & \sn{1.946}{-4} \\
+      MCSA-ILUT-MR-EV & 3 & \sn{3.295}{-5} \\
+      Richardson-ILUT & 3 & \sn{2.161}{-5} \\
       \hline\hline
@@
   \caption{\textbf{Serial scalar flux verification results for the
-      4-group fuel assembly problem.} \textit{The maximum and minimum
-      relative differences over the entire spatial domain are
-      presented with BiCGStab used as the reference. Each calculation
-      was converged with the same eigensolver and linear solver
+      4-group fuel assembly problem.} \textit{BiCGStab is used as the
+      reference. Each calculation was converged with the same
+      eigensolver and linear solver
       tolerances. Table~\ref{tab:spn_solver_defs} gives the
@@
 memory restrictions arising from the explicit preconditioning strategy
-prevent larger energy groups to be analyzed using an appropriate
-preconditioner for the fuel assembly problem.
+prevent larger numbers of energy groups to be analyzed using an
+appropriate preconditioner for the fuel assembly problem.
 

### 2013-08-22 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
fixing the spn verification section. i am finally happy with this

@@
 itself will be compared as a means of verification. For the flux
-values, we look to the 2-norm of the absolute difference between the
-reference case and the comparison case in each energy group:
+values, we look to the 2-norm and infinity norm of the relative
+difference vector between the reference case and the comparison case
+in each energy group:
 \begin{equation}
-  || \ve{\Phi}_{ref}^g - \ve{\Phi}^g ||_2 < \epsilon_k\:,
-  \label{eq:norm_comparison}
+  \text{Relative Difference} = \frac{\ve{\Phi}_{ref}^g -
+    \ve{\Phi}^g}{\ve{\Phi}_{ref}^g}\:,
+  \label{eq:relative_difference_vector}
 \end{equation}
-such that this norm is less than the eigensolver tolerance,
-$\epsilon_k$. For this work, the BiCGStab results will be taken as the
-reference case. If the MCSA flux results satisfy
-Eq~(\ref{eq:norm_comparison}), they will be deemed correct for the
-purposes of this work. For the eigenvalues, we will look for the same
-agreement within the tolerance of the eigensolver.
+where the addition and subtraction operations are performed
+element-wise to construct the vector. For this work, the BiCGStab
+results will be taken as the reference case.
 
@@
 eigenvalue problem was converged with an eigensolver tolerance of
-$\epsilon_k = \sn{1}{-6}$ and a linear solver tolerance of \sn{1}{-8}
-in a serial calculation. Table~\ref{tab:serial_ev_results} gives the
-eigenvalue computed by each solver for each problem as well as the
-number of eigenvalue iterations required to converge. For every
-problem, each solver converged to the same eigenvalue up to the number
-of digits reported by the eigensolver eigensolver in the same number
-of eigenvalue iterations.
+\sn{1}{-6} and a linear solver tolerance of \sn{1}{-8} in a serial
+calculation. Table~\ref{tab:serial_ev_results} gives the eigenvalue
+computed by each solver for each problem as well as the number of
+eigenvalue iterations required to converge. For every problem, each
+solver converged to the same eigenvalue up to the number of digits
+reported by the eigensolver eigensolver in the same number of
+eigenvalue iterations.
 \begin{table}[h!]
@@
 Table~\ref{tab:serial_differences_g2}, and
-Table~\ref{tab:serial_differences_g4} give the 2-norms of the cell
-scalar fluxes difference vectors for all calculations and all energy
-groups. All comparisons were performed within a group such that the
-differences for the MCSA-ILUT-R-C results in group 3 for the 4-group
-calculation were computed with the BiCGStab-ILUT solution in group 3
-for the 4-group calculation as a reference. For all group flux results
-given in the table, the maximum differences for all cases was less
-than the linear solver tolerance used to converge the flux at each
-eigenvalue iteration meaning that all values over the entire energy
-and spatial domain are within these limits. Based on the flux results
-and the eigenvalue results presented here, the MCSA results should be
-deemed correct for the fuel assembly problem for serial calculations
-with multiple energy groups.
+Table~\ref{tab:serial_differences_g4} give the infinity norms and
+2-norms of the flux relative difference vectors given by
+Eq~(\ref{eq:relative_difference_vector}) for all calculations and all
+energy groups. In general, we see good agreement in the two norm
+between the MCSA variants and the GMRES calculation, meaning that the
+have approximately the same variation from the BiCGStab
+calculation. We do note larger infinity norm values of this vector for
+some of the MCSA variants (i.e. group 1 in the 2-group calculation),
+however this large maximum value was not a global phenomenon with good
+agreement in the 2-norm for the calculations. Based on the good
+agreement between MCSA and GMRES for the relative difference of the
+flux and the eigenvalue results presented here, the MCSA results
+should be deemed correct for the fuel assembly problem for serial
+calculations with multiple energy groups.
 
@@
       \multicolumn{1}{c}{\textbf{Group}} &
-      \multicolumn{1}{c}{\textbf{$|| \ve{\Phi}_{ref}^g - \ve{\Phi}^g ||_2$}} \\
+      \multicolumn{1}{c}{\textbf{$|| \cdot ||_{inf}$}} &
+      \multicolumn{1}{c}{\textbf{$|| \cdot ||_2$}} \\
       \hline
-      GMRES-ILUT & 0 & \sn{4.562}{-6} \\
-      MCSA-ILUT-R-C & 0 & \sn{2.099}{-5} \\
-      MCSA-ILUT-MR-C & 0 & \sn{4.812}{-5} \\
-      MCSA-ILUT-R-EV & 0 & \sn{2.371}{-5} \\
-      MCSA-ILUT-MR-EV & 0 & \sn{3.605}{-5} \\
-      Richardson-ILUT & 0 & \sn{1.531}{-5} \\
+      GMRES-ILUT & 0 & \sn{4.562}{-6} & \sn{6.746}{-5} \\
+      MCSA-ILUT-R-C & 0 & \sn{2.099}{-5} & \sn{2.541}{-4} \\
+      MCSA-ILUT-MR-C & 0 & \sn{4.812}{-5} & \sn{4.672}{-4} \\
+      MCSA-ILUT-R-EV & 0 & \sn{2.371}{-5} & \sn{2.740}{-4} \\
+      MCSA-ILUT-MR-EV & 0 & \sn{3.605}{-5} & \sn{2.554}{-4} \\
+      Richardson-ILUT & 0 & \sn{1.531}{-5} & \sn{2.771}{-4} \\
       \hline\hline
@@
   \caption{\textbf{Serial scalar flux verification results for the
-      1-group fuel assembly problem.} \textit{BiCGStab is used as the
-      reference. Each calculation was converged with the same
-      eigensolver and linear solver
+      1-group fuel assembly problem.} \textit{The infinity norm and
+      2-norm of the relative flux vector in each group is given with
+      BiCGStab is used as the reference. Each calculation was
+      converged with the same eigensolver and linear solver
       tolerances. Table~\ref{tab:spn_solver_defs} gives the
@@
       \multicolumn{1}{c}{\textbf{Group}} &
-      \multicolumn{1}{c}{\textbf{$|| \ve{\Phi}_{ref}^g - \ve{\Phi}^g ||_2$}} \\
+      \multicolumn{1}{c}{\textbf{$|| \cdot ||_{inf}$}} &
+      \multicolumn{1}{c}{\textbf{$|| \cdot ||_2$}} \\
       \hline
-      GMRES-ILUT & 0 & \sn{7.590}{-6} \\
-      MCSA-ILUT-R-C & 0 & \sn{5.308}{-6} \\
-      MCSA-ILUT-MR-C & 0 & \sn{6.387}{-5} \\
-      MCSA-ILUT-R-EV & 0 & \sn{4.197}{-5} \\
-      MCSA-ILUT-MR-EV & 0 & \sn{2.736}{-5} \\
-      Richardson-ILUT & 0 & \sn{3.976}{-6} \\
+      GMRES-ILUT & 0 & \sn{7.590}{-6} & \sn{1.234}{-4} \\
+      MCSA-ILUT-R-C & 0 & \sn{5.308}{-6} & \sn{1.066}{-4} \\
+      MCSA-ILUT-MR-C & 0 & \sn{6.388}{-5} & \sn{2.233}{-4} \\
+      MCSA-ILUT-R-EV & 0 & \sn{4.197}{-5} & \sn{2.168}{-4} \\
+      MCSA-ILUT-MR-EV & 0 & \sn{2.736}{-5} & \sn{8.160}{-5} \\
+      Richardson-ILUT & 0 & \sn{3.976}{-6} & \sn{7.059}{-5} \\
       \hline
-      GMRES-ILUT & 1 & \sn{7.579}{-6} \\
-      MCSA-ILUT-R-C & 1 & \sn{1.645}{-5} \\
-      MCSA-ILUT-MR-C & 1 & \sn{2.634}{-4} \\
-      MCSA-ILUT-R-EV & 1 & \sn{1.797}{-4} \\
-      MCSA-ILUT-MR-EV & 1 & \sn{1.556}{-4} \\
-      Richardson-ILUT & 1 & \sn{4.031}{-6} \\
+      GMRES-ILUT & 1 & \sn{7.579}{-6} & \sn{1.246}{-4} \\
+      MCSA-ILUT-R-C & 1 & \sn{1.645}{-5} & \sn{1.210}{-4} \\
+      MCSA-ILUT-MR-C & 1 & \sn{6.635}{-4} & \sn{9.435}{-4} \\
+      MCSA-ILUT-R-EV & 1 & \sn{1.797}{-4} & \sn{2.983}{-4} \\
+      MCSA-ILUT-MR-EV & 1 & \sn{1.556}{-4} & \sn{2.055}{-4} \\
+      Richardson-ILUT & 1 & \sn{4.031}{-6} & \sn{7.110}{-5} \\
       \hline\hline
@@
   \caption{\textbf{Serial scalar flux verification results for the
-      2-group fuel assembly problem.} \textit{BiCGStab is used as the
-      reference. Each calculation was converged with the same
-      eigensolver and linear solver
+      2-group fuel assembly problem.} \textit{The infinity norm and
+      2-norm of the relative flux vector in each group is given with
+      BiCGStab is used as the reference. Each calculation was
+      converged with the same eigensolver and linear solver
       tolerances. Table~\ref{tab:spn_solver_defs} gives the
@@
       \multicolumn{1}{c}{\textbf{Group}} &
-      \multicolumn{1}{c}{\textbf{$|| \ve{\Phi}_{ref}^g - \ve{\Phi}^g ||_2$}} \\
+      \multicolumn{1}{c}{\textbf{$|| \cdot ||_{inf}$}} &
+      \multicolumn{1}{c}{\textbf{$|| \cdot ||_2$}} \\
       \hline
-      GMRES-ILUT & 0 & \sn{2.376}{-5} \\
-      MCSA-ILUT-R-C & 0 & \sn{2.639}{-5} \\
-      MCSA-ILUT-MR-C & 0 & \sn{3.958}{-5} \\
-      MCSA-ILUT-R-EV & 0 & \sn{2.749}{-5} \\
-      MCSA-ILUT-MR-EV & 0 & \sn{2.969}{-5} \\
-      Richardson-ILUT & 0 & \sn{2.189}{-5} \\
+      GMRES-ILUT & 0 & \sn{2.376}{-5} & \sn{4.043}{-4} \\
+      MCSA-ILUT-R-C & 0 & \sn{2.639}{-5} & \sn{3.950}{-4} \\
+      MCSA-ILUT-MR-C & 0 & \sn{3.958}{-5} & \sn{4.635}{-4} \\
+      MCSA-ILUT-R-EV & 0 & \sn{2.749}{-5} & \sn{3.975}{-4} \\
+      MCSA-ILUT-MR-EV & 0 & \sn{2.969}{-5} & \sn{4.054}{-4} \\
+      Richardson-ILUT & 0 & \sn{2.189}{-5} & \sn{3.791}{-4} \\
       \hline
-      GMRES-ILUT & 1 & \sn{2.271}{-5} \\
-      MCSA-ILUT-R-C & 1 & \sn{3.702}{-5} \\
-      MCSA-ILUT-MR-C & 1 & \sn{3.803}{-5} \\
-      MCSA-ILUT-R-EV & 1 & \sn{4.142}{-5} \\
-      MCSA-ILUT-MR-EV & 1 & \sn{2.723}{-5} \\
-      Richardson-ILUT & 1 & \sn{2.137}{-5} \\
+      GMRES-ILUT & 1 & \sn{2.271}{-5} & \sn{3.952}{-4} \\
+      MCSA-ILUT-R-C & 1 & \sn{3.701}{-5} & \sn{3.933}{-4} \\
+      MCSA-ILUT-MR-C & 1 & \sn{3.803}{-5} & \sn{3.933}{-4} \\
+      MCSA-ILUT-R-EV & 1 & \sn{4.142}{-5} & \sn{4.011}{-4} \\
+      MCSA-ILUT-MR-EV & 1 & \sn{2.723}{-5} & \sn{4.030}{-4} \\
+      Richardson-ILUT & 1 & \sn{2.137}{-5} & \sn{3.768}{-4} \\
       \hline
-      GMRES-ILUT & 2 & \sn{2.310}{-5} \\
-      MCSA-ILUT-R-C & 2 & \sn{2.490}{-5} \\
-      MCSA-ILUT-MR-C & 2 & \sn{2.214}{-5} \\
-      MCSA-ILUT-R-EV & 2 & \sn{4.456}{-5} \\
-      MCSA-ILUT-MR-EV & 2 & \sn{3.366}{-5} \\
-      Richardson-ILUT & 2 & \sn{2.143}{-5} \\
+      GMRES-ILUT & 2 & \sn{2.310}{-5} & \sn{4.021}{-4} \\
+      MCSA-ILUT-R-C & 2 & \sn{2.490}{-5} & \sn{3.956}{-4} \\
+      MCSA-ILUT-MR-C & 2 & \sn{2.214}{-5} & \sn{3.608}{-4} \\
+      MCSA-ILUT-R-EV & 2 & \sn{4.456}{-5} & \sn{3.833}{-4} \\
+      MCSA-ILUT-MR-EV & 2 & \sn{3.366}{-5} & \sn{3.797}{-4} \\
+      Richardson-ILUT & 2 & \sn{2.143}{-5} & \sn{3.767}{-4} \\
       \hline
-      GMRES-ILUT & 3 & \sn{2.335}{-5} \\
-      MCSA-ILUT-R-C & 3 & \sn{3.045}{-5} \\
-      MCSA-ILUT-MR-C & 3 & \sn{2.696}{-5} \\
-      MCSA-ILUT-R-EV & 3 & \sn{1.946}{-4} \\
-      MCSA-ILUT-MR-EV & 3 & \sn{3.295}{-5} \\
-      Richardson-ILUT & 3 & \sn{2.161}{-5} \\
+      GMRES-ILUT & 3 & \sn{2.335}{-5} & \sn{4.029}{-4} \\
+      MCSA-ILUT-R-C & 3 & \sn{3.045}{-5} & \sn{3.963}{-4} \\
+      MCSA-ILUT-MR-C & 3 & \sn{2.696}{-5} & \sn{3.606}{-4} \\
+      MCSA-ILUT-R-EV & 3 & \sn{1.947}{-5} & \sn{4.283}{-4} \\
+      MCSA-ILUT-MR-EV & 3 & \sn{3.295}{-5} & \sn{3.837}{-4} \\
+      Richardson-ILUT & 3 & \sn{2.161}{-5} & \sn{3.772}{-4} \\
       \hline\hline
@@
   \caption{\textbf{Serial scalar flux verification results for the
-      4-group fuel assembly problem.} \textit{BiCGStab is used as the
-      reference. Each calculation was converged with the same
-      eigensolver and linear solver
+      4-group fuel assembly problem.} \textit{The infinity norm and
+      2-norm of the relative flux vector in each group is given with
+      BiCGStab is used as the reference. Each calculation was
+      converged with the same eigensolver and linear solver
       tolerances. Table~\ref{tab:spn_solver_defs} gives the

### 2013-08-22 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
removng point jacobi preconditioning section from the spn chapter

@@
 %%---------------------------------------------------------------------------%%
-\section{Spectral Analysis of the $SP_N$ Equations}
+\section{Basic Preconditioning for the $SP_N$ Equations}
 \label{sec:spn_spectral_analysis}
-Before we can move on to solving the $SP_N$ equations with both
-conventional and Monte Carlo methods, we must first understand their
-spectral character to ensure that MCSA will be applicable. As
-presented in Chapter~\ref{ch:stochastic_methods}, the spectral radius
-of the iteration matrix must be less than unity to ensure
-convergence. Therefore, we will compute the spectral radius of the
-iteration matrix formed by the $SP_N$ equations while parametrically
-varying the $SP_N$ order, the $P_N$ order, the number of energy
-groups, and upscattering/downscattering between energy groups. For
-each problem in the parameter study, the values given in
-Table~\ref{tab:spn_fixed_parameters} were fixed for each spatial
-location and energy group.
+Previous work in applying MCSA to the thermal radiation diffusion
+equations had success in using Jacobi preconditioning
+\cite{evans_monte_2012}. In this section, we will compute the spectral
+radius of the iteration matrix formed by the $SP_N$ equations while
+parametrically varying the $SP_N$ order, the $P_N$ order, the number
+of energy groups, and upscattering/downscattering between energy
+groups for Jacobi preconditioned problems to ensure that this type of
+preconditioning will be applicable. For each problem in the parameter
+study, the values given in Table~\ref{tab:spn_fixed_parameters} were
+fixed for each spatial location and energy group.
+
 \begin{table}[h!]
@@
 \end{table}
+
 Mesh element sizes were the same in all cardinal directions with
@@
 diagonal character of these systems, although the blocks from
-Eq~(\ref{eq:A_block_matrix}) are readily apparent. Our first attempt
-at preconditioning this system will be to use the point Jacobi
-preconditioning from \S\ref{subsubsec:basic_mcsa_preconditioning} due
-to this diagonal form and the success achieved by applying it to MCSA
-solutions of the thermal radiation diffusion equation as presented in
-\cite{evans_monte_2012}.
-
-\subsection{Point Jacobi Spectral Analysis Results}
-\label{subsec:spn_analysis_results}
-Spectral radius computations were performed for the cases described
-above for the point Jacobi preconditioned iteration matrix, $\ve{H} =
-\ve{I}-\ve{M}^{-1}\ve{A}$, with $\ve{M} =
-diag(\ve{A})$. Table~\ref{tab:group1pj} gives the results for the
-1-group case, Table~\ref{tab:group10dspj} for the 10-group case with
-full downscatter only and Table~\ref{tab:group10uspj} for the 10-group
-case with full downscatter and full upscatter.
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{cccccc}\hline\hline
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{$SP_N$ Order}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{} \\
-       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
-       & \textbf{0} & 0.0635 & 0.6722 & 1.3144 & 1.976 \\
-       & \textbf{1} & 0.0666 & 0.6728 & 1.3141 & 1.9755 \\
-      $P_N$ Order & \textbf{3} & 0.0666 & 0.6822 & 1.3141 & 1.9755 \\
-       & \textbf{5} & 0.0666 & 0.6822 & 1.3278 & 1.9847 \\
-       & \textbf{7} & 0.0666 & 0.6822 & 1.3278 & 1.9917 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Spectral radius results for the point Jacobi
-      preconditioned iteration matrix with 1 energy group.}}
-  \label{tab:group1pj}
-\end{table}
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{cccccc}\hline\hline
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{$SP_N$ Order}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{} \\
-       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
-       & \textbf{0} & 0.0655 & 0.677 & 1.32 & 1.982 \\
-       & \textbf{1} & 0.071 & 0.6777 & 1.319 & 1.982 \\
-      $P_N$ Order & \textbf{3} & 0.071 & 0.687 & 1.327 & 1.9872 \\
-       & \textbf{5} & 0.071 & 0.687 & 1.336 & 1.997 \\
-       & \textbf{7} & 0.071 & 0.687 & 1.336 & 1.9995 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Spectral radius results for the point Jacobi
-      preconditioned iteration matrix with 10 energy groups and full
-      downscatter.}}
-  \label{tab:group10dspj}
-\end{table}
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{cccccc}\hline\hline
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{$SP_N$ Order}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{} \\
-       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
-       & \textbf{0} & 0.7283 & 0.81 & 1.47 & 2.1446 \\
-       & \textbf{1} & 0.7317 & 0.8 & 1.46 & 2.1368 \\
-      $P_N$ Order & \textbf{3} & 0.7317 & 0.91 & 1.526 & 2.2274 \\
-       & \textbf{5} & 0.7317 & 0.91 & 1.5344 & 2.2562 \\
-       & \textbf{7} & 0.7317 & 0.91 & 1.5345 & 2.2842 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Spectral radius results for the point Jacobi
-      preconditioned iteration matrix with 10 energy groups, full
-      downscatter and full upscatter.}}
-  \label{tab:group10uspj}
-\end{table}
-
-It is readily apparent from the tabulated data that point Jacobi
-preconditioning is insufficient. For problems of order $SP_5$ and
-$SP_7$, the method will not converge at all and for the upscatter case
-with $SP_3$ the spectral radius is still quite large for large $P_N$
-orders and therefore convergence is expected to be slow. These
-eigenvalues signal a need for a better preconditioning strategy to
-both ensure and improve convergence for Monte Carlo methods.
-
-\subsection{Block Jacobi Preconditioning}
-\label{sec:spn_preconditioning}
-If Monte Carlo methods are to be used to solve the $SP_N$ system of
-equations, a different preconditioning strategy is required in order
-to ensure convergence for systems of all $SP_N$ and $P_N$ orders with
-arbitrary energy group structures. As another means of achieving this,
-we look back to the sparsity plots we generated in
-Figures~\ref{fig:group1}, \ref{fig:group10ds} and \ref{fig:group10us}
-as well as the multigroup $SP_N$ equations. Initially, the diagonal
-character of the system and previous success with the thermal
-radiuation difussion equation led us to try point Jacobi
-preconditioning with only marginal results. From the sparsity plots we
-note the block structure that ultimately arises from the multigroup
-scattering matrices and their insertion into
-Eq~(\ref{eq:A_block_matrix}). When full upscatter and downscatter are
-used the resulting blocks are completely dense while only downscatter
-gives a lower triangular scattering matrix and the block structure
-shown in Figure~\ref{fig:group10ds}.
+Eq~(\ref{eq:A_block_matrix}) are readily apparent.
 
 Based on this both block and diagonally dominant structure for
-matrices formed by the general multigroup $SP_N$ equations, we instead
-choose \textit{block Jacobi} preconditioning as a left preconditioner
-for the system. Like point Jacobi preconditioning, block Jacobi
+matrices formed by the general multigroup $SP_N$ equations, we choose
+\textit{block Jacobi} preconditioning as a left preconditioner for the
+system. Like point Jacobi preconditioning, block Jacobi
 preconditioning extracts the diagonal elements of the matrix as the
@@
 
-\subsection{Block Jacobi Spectral Analysis Results}
-\label{subsec:spn_analysis_results}
 Spectral radius computations for the block Jacobi preconditioned
@@
 solution methods. All cases were observed to have a spectral radius
-below unity and often significantly smaller than that, greatly
-improving convergence rates over the point Jacobi preconditioned
-problem. Based on these results, the block Jacobi method should be the
-first preconditioning strategy applied to more complicated problems
-representative of physical nuclear systems.
+below unity. Based on these results, the block Jacobi method should be
+the applicable to more complicated problems representative of physical
+nuclear systems.
 
@@
 transport equation as presented here. In particular, we will use a
-pair of production Krylov solvers with which we will compare direct
-numerical results in order to verify MCSA solutions in this section.
-
-In order to verify the correctness of the MCSA method and
-implementation used for this work, MCSA solutions to the fuel assembly
-problem with 1, 2, and 4 energy groups using $SP_1$ discretization
-will be directly compared point-wise over all group and cell fluxes
-computed by converging the criticality problem using both BiCGStab and
-GMRES production implementations from the Trilinos scientific
-computing library Aztec \cite{heroux_overview_2005}. In addition to
-comparing the cell and group fluxes, the number of eigenvalue
-iterations required to converge the problem along with the eigenvalue
-itself will be compared as a means of verification. For the flux
-values, we look to the 2-norm and infinity norm of the relative
-difference vector between the reference case and the comparison case
-in each energy group:
+pair of Krylov solvers with which we will compare numerical results to
+verify MCSA solutions the implementation used for this work.
+
+MCSA solutions to the fuel assembly problem with 1, 2, and 4 energy
+groups using $SP_1$ discretization will computed using both BiCGStab
+and GMRES implementations from the Trilinos scientific computing
+library Aztec \cite{heroux_overview_2005}. The neutron fluxes, the
+number of eigenvalue iterations required to converge the problem, and
+the eigenvalue itself will be compared as a means of verification. For
+the flux values, we will look at the 2-norm and infinity norm of the
+relative difference vector between the reference case and the
+comparison case in each energy group:
 \begin{equation}
-  \text{Relative Difference} = \frac{\ve{\Phi}_{ref}^g -
+  \text{Relative Group Flux Difference} = \frac{\ve{\Phi}_{ref}^g -
     \ve{\Phi}^g}{\ve{\Phi}_{ref}^g}\:,
@@
 
-For the verification calculations, all solvers will be preconditioned
-with ILUT using a drop tolerance of \sn{1}{-5} and a fill level of
-5. For GMRES, no restrictions were placed on the size of the subspace
-and therefore no restarts occurred. When the collision estimator was
-used with MCSA, \sn{2}{4} stochastic histories were used to compute
-the correction for every energy group in the problem (\sn{4}{4} and
-\sn{8}{4} histories total for the 2 and 4 group calculations
-respectively) corresponding to approximately 1 stochastic history per
-DOF. When the expected value estimator was used, \sn{5}{2} stochastic
-histories used for each energy group (\sn{1}{3} and \sn{1.5}{3}
-histories total for the 2 and 4 group calculations) giving
-approximately 1 stochastic history for every 5 DOFs. The reduced
-domain approximation was also applied in conjunction with the ILUT
-preconditioning using a fill level of 100 and a threshold value of
-\sn{1}{-10} to reduce the density of states in the Monte Carlo
-problem. For relaxation parameters, all MCSA computations used a
-Neumann relaxation parameter of 0.7 while a Richardson relaxation of
-1.1 was used with the collision estimator and 1.0 with the expected
-value estimator as determined by the previous analysis of the
-relaxation parameters. Table~\ref{tab:spn_solver_defs} gives
-definitions for the solvers used to generate the results in the
-remainder of this section. For the energy group structures,
+All solvers will be preconditioned with ILUT using a drop tolerance of
+\sn{1}{-5} and a fill level of 5. For GMRES, no restrictions were
+placed on the size of the subspace and therefore no restarts
+occurred. When the collision estimator was used with MCSA, \sn{2}{4}
+stochastic histories were used to compute the correction for every
+energy group in the problem (\sn{4}{4} and \sn{8}{4} histories total
+for the 2 and 4 group calculations respectively) corresponding to
+approximately 1 stochastic history per DOF. When the expected value
+estimator was used, \sn{5}{2} stochastic histories used for each
+energy group (\sn{1}{3} and \sn{1.5}{3} histories total for the 2 and
+4 group calculations) giving approximately 1 stochastic history for
+every 5 DOFs. The reduced domain approximation was also applied in
+conjunction with the ILUT preconditioning using a fill level of 100
+and a threshold value of \sn{1}{-10} to reduce the density of states
+in the Monte Carlo problem. For relaxation parameters, all MCSA
+computations used a Neumann relaxation parameter of 0.7 while a
+Richardson relaxation of 1.1 was used with the collision estimator and
+1.0 with the expected value estimator as determined by the previous
+analysis. Table~\ref{tab:spn_solver_defs} gives definitions for the
+solvers used to generate the results in the remainder of this
+section. For the energy group structures,
 Table~\ref{tab:spn_group_structure} gives the lower bounds of each
@@
 solver converged to the same eigenvalue up to the number of digits
-reported by the eigensolver eigensolver in the same number of
-eigenvalue iterations.
+reported by the eigensolver in the same number of eigenvalue
+iterations.
+
 \begin{table}[h!]
@@
 energy groups. In general, we see good agreement in the two norm
-between the MCSA variants and the GMRES calculation, meaning that the
+between the MCSA variants and the GMRES calculation, meaning that they
 have approximately the same variation from the BiCGStab
@@
 agreement between MCSA and GMRES for the relative difference of the
-flux and the eigenvalue results presented here, the MCSA results
-should be deemed correct for the fuel assembly problem for serial
-calculations with multiple energy groups.
+group fluxes and the eigenvalue results presented here, the MCSA
+implementation should be deemed correct for the fuel assembly problem
+for serial calculations with multiple energy groups.
 
@@
       2-norm of the relative flux vector in each group is given with
-      BiCGStab is used as the reference. Each calculation was
-      converged with the same eigensolver and linear solver
+      BiCGStab used as the reference. Each calculation was converged
+      with the same eigensolver and linear solver
       tolerances. Table~\ref{tab:spn_solver_defs} gives the
@@
 addressed. With the same set of solver parameters used for
-verification, MCSA will now be directly compared to the same
-production Krylov solvers in terms of both iterative performance and
-CPU timing for solutions to the fuel assembly problem. The iterative
-performance will also be compared to using a Richardson iteration
-preconditioned in the same way for all calculations in order to assess
-the acceleration provided by the residual Monte Carlo component of
-MCSA. This performance comparison is necessary in order to assess the
-feasibility of the method and identify areas that require further
-research and development.
+verification caluclation, MCSA will now be compared to the same Krylov
+solvers in terms of both iterative performance and CPU timing. A
+Richardson iteration solution will also be used in order to assess the
+acceleration provided by the residual Monte Carlo component of MCSA.
 
@@
 number of energy groups. In general, the iterative performance of all
-the solvers tested is comparable with BiCGStab performing the best. We
-expect this because not only are the multigroup $SP_N$ equations
-elliptic and positive-definite, but they are also nearly symmetric and
-therefore we expect the conjugate gradient-based methods to perform
-well due to the structure of the projection scheme. It important to
-note that the iterative performance of each solver is not a strong
-function of the number of energy groups in the problem (and therefore
-the problem size). Even more important, although BiCGStab produced the
-best iterative results, MCSA did perform better than GMRES in many
-cases with the Richardson iteration accelerated with Monte Carlo using
-the collision estimator performing best. When compared to the
+the solvers tested aside from the Richardson iteration is comparable
+with BiCGStab performing the best. We expect this because not only are
+the multigroup $SP_N$ equations positive-definite, but they are also
+block-symmetric and therefore we expect the conjugate gradient-based
+methods to perform well due to the structure of the projection
+scheme. It important to note that the iterative performance of each
+solver is not a strong function of the number of energy groups in the
+problem (and therefore the problem size). Even more important,
+although BiCGStab produced the best iterative results, MCSA did
+perform better than GMRES in many cases with the Richardson iteration
+clearly accelerated by the Monte Carlo solve. When compared to the
 preconditioned Richardson iteration, the acceleration provided by MCSA
@@
 
-For MCSA, we are linearly increasing the number of stochastic
-histories used to compute the correction at each iteration as the
-number of energy groups are increased and therefore the work performed
-for the problem is of $O(N)$. It is also important to note the reduced
-domain approximation parameters, particularly the fill level, were
-fixed as the number of energy groups was increased. By fixing the fill
-level, we are effectively fixing the amount of information contained
-in the composite linear operator available for the Monte Carlo
+It is also important to note the reduced domain approximation
+parameters, particularly the fill level, were fixed as the number of
+energy groups was increased. By fixing the fill level, we are
+effectively fixing the amount of information contained in the
+composite linear operator available for the Monte Carlo
 problem. Limiting this information to 100 entries per row in the
@@
 Although iterative performance for MCSA was comparable to that
-observed for production Krylov implementations using exactly the same
-preconditioning scheme, timing performance is not as competitive. For
-this comparison, the amount of time to perform each linear solver
-iteration in every eigenvalue iteration was recorded and then averaged
-over all linear solver
-iterations. Figure~\ref{fig:spn_comparison_time} gives the average CPU
-time per linear solver iteration required to converge the fuel
+observed for Krylov implementations using exactly the same
+preconditioning scheme, timing performance is not as
+competitive. Figure~\ref{fig:spn_comparison_time} gives the average
+CPU time per linear solver iteration required to converge the fuel
 assembly problem with all linear solver iterations over all eigenvalue
-iterations used to compute the average. The production Krylov solver
+iterations used to compute the average. The Krylov solver
 implementations are approximately an order of magnitude faster than
@@
 
-For the production Krylov methods, the ILUT preconditioners are
-applied to a vector with two triangular solves, one for each
-triangular factor. Also, the original linear operator has $O(10)$
-elements in each row of the system, an order of magnitude less than
-that used in the reduced domain composite operator. Second, the
-implementation presented here has not been optimized and it is likely
-that several components of the algorithm can be implemented in a more
-desirable time complexity. The production Krylov solvers presented
-here have enjoyed decades of professional development and
-optimization. However, it is most important to note here that the CPU
-timing data presented in Figure~\ref{fig:spn_comparison_time} shows
-that all methods have the same time complexity; their differences are
-merely the manifestation of a large time constant due to the effects
-of the explicit preconditioning scheme. For transport systems where
-this is not a problem, the literature explicitly shows that MCSA can
-be competitive with these production Krylov methods using CPU time as
-a measure \cite{evans_monte_2012}.
+For the Krylov methods, the ILUT preconditioners are applied to a
+vector with two triangular solves, one for each triangular
+factor. Also, the original linear operator has $O(10)$ elements in
+each row of the system, an order of magnitude less than that used in
+the reduced domain composite operator. Second, the implementation
+presented here has not been optimized and it is likely that several
+components of the algorithm can be implemented in a more desirable
+time complexity. However, it is most important to note here that the
+CPU timing data presented in Figure~\ref{fig:spn_comparison_time}
+shows that all methods have the same time complexity; their
+differences are merely the manifestation of a large time constant due
+to the effects of the explicit preconditioning scheme. For transport
+systems where this is not a problem, the literature explicitly shows
+that MCSA can be competitive with these production Krylov methods
+using CPU time as a measure \cite{evans_monte_2012}.
 
@@
 to production linear solvers, but the iterative performance is
-comparable to production Krylov methods with identical preconditioning
-that would typically be used every-day calculations. From a CPU timing
+comparable to Krylov methods with identical preconditioning that would
+typically be used every-day calculations. From a CPU timing
 perspective, the methods yield the same time complexity as a function
 of the number of energy groups in the problem when compared to the
-production Krylov methods. Here, a large time constant is generated
-due to the explicit preconditioning strategy and the generation of
-dense linear operators as a result. To improve these results and put
-general MCSA schemes into a performance regime where they are
-competitive with Krylov methods for neutron transport problems,
-significant research will be required to improve upon the explicit
-preconditioning scheme presented here.
+Krylov methods. Here, a large time constant is generated due to the
+explicit preconditioning strategy and the generation of dense linear
+operators as a result. To improve these results and put general MCSA
+schemes into a performance regime where they are competitive with
+Krylov methods for neutron transport problems, significant research
+will be required to improve upon the explicit preconditioning scheme
+presented here.
 

### 2013-08-22 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
removing spainv-ilut section

@@
 \textit{block Jacobi} preconditioning as a left preconditioner for the
-system. Like point Jacobi preconditioning, block Jacobi
-preconditioning extracts the diagonal elements of the matrix as the
-preconditioner where now the elements extracted are the blocks on the
-diagonal as shown on the left side of
+system. Block Jacobi preconditioning extracts the diagonal elements of
+the matrix as the preconditioner where now the elements extracted are
+the blocks on the diagonal as shown on the left side of
 Figure~\ref{fig:block_jacobi_ex}. Shown on the right side of
@@
 \end{figure}
-In addition, in the limit of a block size of one, the block Jacobi
-method reduces to the point Jacobi method.
 
@@
 Spectral radius computations for the block Jacobi preconditioned
-iteration matrix were performed for the same problems as the point
-Jacobi preconditioned case. Table~\ref{tab:group1bj} gives the results
-for the 1-group case, Table~\ref{tab:group10dsbj} for the 10-group
-case with full downscatter only and Table~\ref{tab:group10usbj} for
-the 10-group case with full downscatter and full upscatter. A block
-size of $N_g\times(N+1)/2$ was used for the preconditioner in all
-cases.
+iteration matrix were performed with Table~\ref{tab:group1bj} giving
+the results for the 1-group case, Table~\ref{tab:group10dsbj} for the
+10-group case with full downscatter only and
+Table~\ref{tab:group10usbj} for the 10-group case with full
+downscatter and full upscatter. A block size of $N_g\times(N+1)/2$ was
+used for the preconditioner in all cases.
 \begin{table}[h!]
@@
 iteration will dictate convergence and remain unchanged throughout the
-computation\footnote{The operator will change if, for example, physics
-  feedback through temperature or potentially burnup is
-  considered. These additional physics will modify the cross sections
-  used to assemble $\mathbf{M}$. The calculations presented here will
-  consist of a single eigenvalue calculation with a static
-  $\mathbf{M}$.} while the addition of fission to the system will only
-modify the source of neutrons and the multiplication factor at every
-eigenvalue iteration while not affecting Monte Carlo transport.
+eigenvalue computation while the addition of fission to the system
+will only modify the source of neutrons and the multiplication factor
+at every eigenvalue iteration while not affecting Monte Carlo
+transport.
 
@@
 \label{subsec:jacobi_prec_assembly_calc}
-Based on the success of block Jacobi preconditioning with the test
-problem used for the spectral radius parameter study, we use it first
-to solve the $17 \times 17$ fuel assembly problem. A single energy
-group problem was first solved with $SP_1$ discretization, effectively
-giving the one-speed neutron diffusion system for the fuel assembly
-resulting in 20,088 degrees of freedom in the
+We first use block Jacobi preconditioning to solve the $17 \times 17$
+fuel assembly problem. A single energy group problem was first
+computed with $SP_1$ discretization, effectively giving the one-speed
+neutron diffusion system for the fuel assembly resulting in 20,088
+degrees of freedom in the
 problem. Figure~\ref{fig:block_jacobi_res_mcsa} gives the residual
@@
 \label{subsec:mcsa_break_down}
-For the fuel assembly problem, the initial block Jacobi
-preconditioned\footnote{For 1-group $SP_1$ problems the block size is
-  1, giving a point Jacobi preconditioner.}  calculations used a
-one-speed $SP_1$ discretization, effectively giving a diffusion
-system. To study the breakdown of MCSA at iteration matrix spectral
-radii near one, we will use the simpler homogeneous 2-dimensional
-one-speed neutron diffusion system presented in
+For the fuel assembly problem, the initial block Jacobi preconditioned
+calculations used a one-speed $SP_1$ discretization, effectively
+giving a diffusion system. To study the breakdown of MCSA at iteration
+matrix spectral radii near one, we will use the simpler homogeneous
+2-dimensional one-speed neutron diffusion system presented in
 Appendix~\ref{chap:diffusion_problem} to isolate this behavior. In
@@
 are required beyond simple Jacobi methods that can reduce the spectral
-radius into a region of better MCSA performance. Several modern
-algebraic preconditioning strategies will be presented here and used
-within the explicit preconditioning framework given by
-Eq~(\ref{eq:left_right_mcsa}). Data showing their effects on MCSA
-solutions of the fuel assembly criticality problem will be presented.
+radius into a region of better MCSA performance. Two modern algebraic
+preconditioning strategies, ILUT and SPAINV, will be presented here
+and used within the explicit preconditioning framework given by
+Eq~(\ref{eq:left_right_mcsa}). Data
+showing their effects on MCSA solutions of the fuel assembly
+criticality problem will be presented.
 
@@
 \label{subsec:spn_ilut_preconditioning}
-Incomplete lower-upper (ILU) factorizations of the linear operator can
-be used a simple mechanism to form an approximate inverse of a
+ILUT preconditioning is chosen due to its applicability to problems
+that have a dominating elliptic compoenent such as the fuel assembly
+problem and because it is easily formulated algebraicly. Incomplete
+lower-upper (ILU) factorizations of the linear operator can be used a
+simple mechanism to form an approximate inverse of a
 preconditioner. To build the factorization, the sparse upper and lower
@@
 \label{subsec:spn_spainv_preconditioning}
-Explicit formation of the preconditioner inverse is a requirement of
-the current MCSA preconditioning strategy. As a result, although ILUT
-preconditioning provided excellent iterative performance, a
-significant time penalty is paid to extract the inverse of the
-triangular factors through $N$ triangular solves. In addition, this
-explicit inversion yielded a composite linear operator that was dense,
-requiring modification of the system in order to achieve better
-scalability and Monte Carlo performance. Sparse approximate inverse
-(SPAINV) preconditioning is a technique that may potentially alleviate
-both of these constraints at the cost of reduced iterative
-performance. The method produces a sparse approximation of the inverse
-matrix directly by minimizing the Frobenius norm of the residual
-matrix \cite{saad_iterative_2003}:
+SPAINV is chosen because it the true inverse of the preconditioner is
+byproduct of the algorithm and the algorithm seeks to maintain the
+sparsity of the inverse preconditioner, potentially alleviating some
+of the issues observed with ILUT. Explicit formation of the
+preconditioner inverse is a requirement of the current MCSA
+preconditioning strategy. As a result, although ILUT preconditioning
+provided excellent iterative performance, a significant time penalty
+is paid to extract the inverse of the triangular factors through $N$
+triangular solves. In addition, this explicit inversion yielded a
+composite linear operator that was dense, requiring modification of
+the system in order to achieve better scalability and Monte Carlo
+performance. Sparse approximate inverse (SPAINV) preconditioning is a
+technique that may potentially alleviate both of these constraints at
+the cost of reduced iterative performance. The method produces a
+sparse approximation of the inverse matrix directly by minimizing the
+Frobenius norm of the residual matrix \cite{saad_iterative_2003}:
 \begin{equation}
@@
   \label{fig:spainv_quality}
-\end{figure}
-
-\clearpage
-
-\subsection{SPAINV Improved ILUT Preconditioning}
-\label{subsec:spn_spainv_preconditioning}
-As a means to improve convergence over using only SPAINV
-preconditioning, ILUT preconditioning may be modified through an
-additional application of SPAINV
-(SPAINV-ILUT)\cite{saad_iterative_2003}. For the ILUT preconditioning
-case we have the following linear problem to solve for the substituted
-variable:
-\begin{equation}
-  \mathbf{L}^{-1} \mathbf{A} \mathbf{U}^{-1} \mathbf{u} =
-  \mathbf{L}^{-1} \mathbf{b}\:.
-  \label{eq:ilut_standalone}
-\end{equation}
-We then apply SPAINV and minimize the following problem, this time on
-the left:
-\begin{equation}
-  || \mathbf{I} - \mathbf{M} \mathbf{L}^{-1} \mathbf{A}
-  \mathbf{U}^{-1} ||^2_F\:,
-  \label{eq:ilut_minimize}
-\end{equation}
-giving a new preconditioned system:
-\begin{equation}
-  \mathbf{M} \mathbf{L}^{-1} \mathbf{A} \mathbf{U}^{-1} \mathbf{u} =
-  \mathbf{M} \mathbf{L}^{-1} \mathbf{b}\:.
-  \label{eq:ilut_spainv}
-\end{equation}
-Doing this permits a smaller fill level to be used with ILUT and a
-lower-order sparsity pattern to be used with SPAINV, giving marginally
-more sparsity in the resulting composite linear operator while
-maintaining good iterative performance.
-
-MCSA preconditioned with SPAINV-ILUT was also used to solve the same
-fuel assembly problem as with the other preconditioners for the same
-1-group $SP_1$ discretization. To study convergence sensitivity to
-SPAINV-ILUT parameters, the ILUT drop tolerance and ILUT fill level
-were parametrically varied with the number of iterations to converge a
-single eigenvalue iteration for the fuel assembly problem recorded
-along with the maximum number of non-zero entries observed in all
-matrix rows for the left/right preconditioned composite linear
-operator. SPAINV parameters for the data presented here were fixed as
-it was found that the preconditioning quality was largely insensitive
-to their variation. In addition, the fact that SPAINV is being used to
-preconditioned a problem already preconditioned with ILUT means that
-the input matrix sparsity pattern is already somewhat dense and large
-SPAINV levels would result in an even denser system. Therefore, a
-smaller SPAINV sparsity pattern level will be used for this
-preconditioning. For each calculation, the number of iterations
-required to converge reported was for a single eigenvalue iteration.
-Again, \sn{3}{4} histories are used at each MCSA iteration to compute
-the Monte Carlo correction using the adjoint collision estimator. All
-SPAINV-ILUT calculations reported here were performed on a single CPU.
-
-Figure~\ref{fig:spainv_ilut_iterations} gives the number of iterations
-required to converge as a function of ILUT drop tolerance for ILUT
-fill levels of 1.5, 2.0, 3.0, 4.0 and 5.0 with a fixed SPAINV
-threshold of 1.0 and sparsity level of 1. SPAINV threshold values
-higher than 1.0 resulted in a loss of convergence and lower values did
-not significantly affect performance. Larger sparsity levels resulted
-in a composite operator that was too dense and a prohibitive amount of
-CPU time required to compute the preconditioner. At smaller levels of
-fill, the benefits of using an additional step of SPAINV
-preconditioning are readily observed with the number of iterations
-required to converge cut in half for large drop tolerances. For
-smaller drop tolerances and higher ILUT levels of fill, the additional
-step of SPAINV preconditioning offers marginal improvement to
-iterative performance. As shown in Figure~\ref{fig:spainv_ilut_size},
-because ILUT is used in the preconditioning sequence, sparsity is
-again lost with only marginally improved non-zero entry numbers due to
-the application of the SPAINV threshold. As a result, the quality
-metrics given by Figure~\ref{fig:spainv_ilut_quality} are only
-marginally better than those observed for ILUT preconditioning and
-much larger than the SPAINV quality metric values.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4.25in]{chapters/spn_equations/psilut_iterations.pdf}
-  \end{center}
-  \caption{\textbf{Number of MCSA iterations required to converge a
-      single eigenvalue iteration for the fuel assembly problem with
-      SPAINV-ILUT preconditioning as a function of ILUT drop
-      tolerance.} \textit{Each colored curve represents the row size
-      for a different ILUT fill level. Fill levels of 1.5, 2.0, 3.0,
-      4.0, and 5.0 were used.}}
-  \label{fig:spainv_ilut_iterations}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4.25in]{chapters/spn_equations/psilut_size.pdf}
-  \end{center}
-  \caption{\textbf{Maximum number of non-zero entries observed for all
-      rows in the composite linear operator for the fuel assembly
-      problem with SPAINV-ILUT preconditioning given as a function of
-      ILUT drop tolerance.} \textit{Each colored curve represents the
-      row size for a different ILUT fill level. Fill levels of 1.5,
-      2.0, 3.0, 4.0 and 5.0 were used.}}
- \label{fig:spainv_ilut_size}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4.25in]{chapters/spn_equations/psilut_quality.pdf}
-  \end{center}
-  \caption{\textbf{SPAINV-ILUT preconditioning quality metric for the
-      fuel assembly problem given as a function of ILUT drop
-      tolerance.} \textit{Each colored curve represents the row size
-      for a different ILUT fill level. Fill levels of 1.5, 2.0, 3.0,
-      4.0 and 5.0 were used.}}
-  \label{fig:spainv_ilut_quality}
 \end{figure}

### 2013-08-22 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
removing residual norm figure from spn estimator comparison

@@
 % NOTE: I copied the following matrix directly out of Tom's tech note
-% on the SPn equations which I am effectively following here because I
-% was feeling lazy. I have verified its correctness.
+% on the SPn equations.
 \begin{equation}
@@
 groups for Jacobi preconditioned problems to ensure that this type of
-preconditioning will be applicable. For each problem in the parameter
-study, the values given in Table~\ref{tab:spn_fixed_parameters} were
-fixed for each spatial location and energy group.
+preconditioning will be applicable to the fuel assembly problem. For
+each problem in the parameter study, the values given in
+Table~\ref{tab:spn_fixed_parameters} were fixed for each spatial
+location and energy group.
 
@@
 \label{subsec:spn_ilut_preconditioning}
-ILUT preconditioning is chosen due to its applicability to problems
-that have a dominating elliptic compoenent such as the fuel assembly
-problem and because it is easily formulated algebraicly. Incomplete
-lower-upper (ILU) factorizations of the linear operator can be used a
-simple mechanism to form an approximate inverse of a
-preconditioner. To build the factorization, the sparse upper and lower
-triangular factors, $\mathbf{L}$ and $\mathbf{U}$, are computed such
-that the residual matrix formed by the factorization
+ILUT preconditioning is chosen first due to its applicability to
+problems that have a dominating elliptic component such as the fuel
+assembly problem or other problems with a high scattering to
+absorption ratio. Incomplete lower-upper (ILU) factorizations of the
+linear operator can be used a simple mechanism to form an approximate
+inverse of a preconditioner. To build the factorization, the sparse
+upper and lower triangular factors, $\mathbf{L}$ and $\mathbf{U}$, are
+computed such that the residual matrix formed by the factorization
 \cite{saad_iterative_2003}:
@@
 \label{subsec:spn_spainv_preconditioning}
-SPAINV is chosen because it the true inverse of the preconditioner is
-byproduct of the algorithm and the algorithm seeks to maintain the
-sparsity of the inverse preconditioner, potentially alleviating some
-of the issues observed with ILUT. Explicit formation of the
-preconditioner inverse is a requirement of the current MCSA
-preconditioning strategy. As a result, although ILUT preconditioning
-provided excellent iterative performance, a significant time penalty
-is paid to extract the inverse of the triangular factors through $N$
-triangular solves. In addition, this explicit inversion yielded a
-composite linear operator that was dense, requiring modification of
-the system in order to achieve better scalability and Monte Carlo
-performance. Sparse approximate inverse (SPAINV) preconditioning is a
-technique that may potentially alleviate both of these constraints at
-the cost of reduced iterative performance. The method produces a
-sparse approximation of the inverse matrix directly by minimizing the
+SPAINV is chosen next because it the true inverse of the
+preconditioner is product of the algorithm, avoiding the need for many
+inverse matrix applications. In addition, the algorithm seeks to
+maintain the sparsity of the inverse preconditioner, potentially
+alleviating some of the dense operator issues observed with
+ILUT. Explicit formation of the preconditioner inverse is a
+requirement of the current MCSA preconditioning strategy. As a result,
+although ILUT preconditioning provided excellent iterative
+performance, a significant time penalty is paid to extract the inverse
+of the triangular factors through $N$ triangular solves. In addition,
+this explicit inversion yielded a composite linear operator that was
+dense, requiring modification of the system in order to achieve better
+scalability and Monte Carlo performance.
+
+Sparse approximate inverse (SPAINV) preconditioning is a technique
+that may potentially alleviate both of these constraints at the cost
+of reduced iterative performance. The method produces a sparse
+approximation of the inverse matrix directly by minimizing the
 Frobenius norm of the residual matrix \cite{saad_iterative_2003}:
@@
 \label{subsec:spn_prec_rda}
-For each preconditioning technique presented convergence was achieved
-for the single fuel assembly problem. However, a primary concern is
-the number of non-zero states in each row of the system generated by
-the explicit preconditioning strategy. In many cases, orders of
-magnitude more matrix elements were generated resulting in poor
-scalability for domain decomposed algorithms and overall performance
-issues for Monte Carlo. As outlined in
-\S~\ref{subsec:reduced_domain_approximation}, the reduced domain
-approximation may be used as a mechanism to potentially alleviate this
-problem by filtering elements of the composite matrix in each row that
-fall below a certain threshold value or by maintaining the largest $N$
-elements in each row where $N$ is a designated fill level.
-
-For the ILUT, SPAINV, and SPAINV-ILUT preconditioning strategies the
-reduced domain approximation will be applied to reduce the density of
-the composite linear operator to more manageable levels. For each
-preconditioner, the parameters that achieved the best quality metric
-results from the previous analysis were used. These correspond to ILUT
-parameters of a fill level of 5.0 and a drop tolerance of \sn{1}{-5},
-SPAINV parameters of a sparsity level of 4 and a threshold of 0.1 and
-SPAINV-ILUT with ILUT parameters of a fill level of 4.0 and a drop
-tolerance of \sn{1}{-5} and SPAINV parameters of a sparsity level of 1
-and a threshold of 1.0. The reduced domain threshold was set to
-\sn{1}{-10} in order to eliminate any exceedingly small values from
-the matrix (often this is simply removing non-zero elements within the
-floating point tolerance of zero). The reduced domain fill level was
-then varied, starting with the largest non-zero entries per row value
+Although ILUT and SPAINV preconditioning permitted MCSA solutions for
+the fuel assembly problem, a primary concern is the number of non-zero
+states in each row of the system generated by the explicit
+preconditioning strategy. In many cases, orders of magnitude more
+matrix elements were generated resulting in poor scalability for
+domain decomposed algorithms and overall performance issues for Monte
+Carlo. As outlined in \S~\ref{subsec:reduced_domain_approximation},
+the reduced domain approximation may be used as a mechanism to
+potentially alleviate this problem by filtering elements of the
+composite matrix in each row that fall below a certain threshold value
+or by maintaining the largest $N$ elements in each row where $N$ is a
+designated fill level. In this section, we will apply the reduced
+domain approximation to both ILUT and SPAINV preconditioned system to
+reduce the density of the composite linear operator to more manageable
+levels and study the resulting iterative performance.
+
+For each preconditioner, the parameters that achieved the best quality
+metric results from the previous analysis were used. These correspond
+to ILUT parameters of a fill level of 5.0 and a drop tolerance of
+\sn{1}{-5} and SPAINV parameters of a sparsity level of 4 and a
+threshold of 0.1. The reduced domain threshold was set to \sn{1}{-10}
+in order to eliminate any exceedingly small values from the matrix
+(often this is simply removing non-zero elements within the floating
+point tolerance of zero). The reduced domain fill level was then
+varied, starting with the largest non-zero entries per row value
 observed for each of the preconditioning types in order to assess its
@@
 only 10 non-zero entries in each row (vs. 7 non-zero entries for the
-case with no preconditioning). SPAINV-ILUT iterative performance was
-marginally better than ILUT alone for all reduced domain fill
-levels. However, it should be noted that the ILUT level of fill used
-for the ILUT only calculations was set to 4 for this case while the
-SPAINV-ILUT preconditioner used an ILUT level of fill of 5 and
-therefore the marginally better performance is more likely a result of
-this addition of fill rather than the extra SPAINV
-preconditioning. Looking at the quality metric data, we see a
-reduction in the quality metric as a function of the reduced domain
-fill level, achieving 2 orders of magnitude reduction in the quality
-metric for the ILUT-based preconditioning methods.
+case with no preconditioning). Looking at the quality metric data, we
+see a reduction in the quality metric as a function of the reduced
+domain fill level, achieving 2 orders of magnitude reduction in the
+quality metric for the ILUT preconditioning.
 
@@
 its error are effectively dampened to improve iterative
-performance. Considering the CPU time required to converge a single
-eigenvalue iteration, similar results are also observed for the
-relaxation parameters as given by Figure~\ref{fig:relax_time}. For the
-base cases given by the plots on the left, it was found the a
-Richardson relaxation parameter of 1.1 and a Neumann relaxation
-parameter of 0.7 provided the fastest CPU time for convergence. Fixing
-each parameter at these new values, the calculations were repeated as
-shown for the plots on the right hand sides of both
-Figures~\ref{fig:relax_iters} and \ref{fig:relax_time}. For each
-repeated timing calculation, it was found that the same combination of
-relaxation parameters found in the base cases performed the best
-although they did not necessarily have the best iterative performance.
+performance.
+
+Considering the CPU time required to converge a single eigenvalue
+iteration, similar results are also observed for the relaxation
+parameters as given by Figure~\ref{fig:relax_time}. For the base cases
+given by the plots on the left, it was found the a Richardson
+relaxation parameter of 1.1 and a Neumann relaxation parameter of 0.7
+provided the fastest CPU time for convergence. Fixing each parameter
+at these new values, the calculations were repeated as shown for the
+plots on the right hand sides of both Figures~\ref{fig:relax_iters}
+and \ref{fig:relax_time}. For each repeated timing calculation, it was
+found that the same combination of relaxation parameters found in the
+base cases performed the best although they did not necessarily have
+the best iterative performance.
 
@@
 \label{subsec:spn_estimator_comparison}
-With convergence obtained for the fuel assembly problem, the iterative
-performance of MCSA will next be studied with both the collision and
-expected value estimators using the best preconditioning and
-relaxation parameter combinations found by the previous
+The iterative performance of MCSA will next be studied with both the
+collision and expected value estimators using the best preconditioning
+and relaxation parameter combinations found by the previous
 analysis. Additionally, both the Richardson-based MCSA iteration given
@@
 \S~\ref{sec:parameter_estimator_analysis}, it is possible that the
-collision estimator may perform better from a timing perspective due
-to the sparsity lost by explicit preconditioning. The deterministic
-component of the expected value estimator that couples the current
-state to other local states through the iteration matrix stencil
-requires application of the estimator to potentially several orders of
-magnitude more states at every transition event.
+collision estimator may perform better from a timing perspective. The
+deterministic component of the expected value estimator that couples
+the current state to other local states through the iteration matrix
+stencil requires application of the estimator to potentially several
+orders of magnitude more states at every transition event.
 
@@
 residual iteration requires several additional parallel
-operations. Compared to the collision estimator, the minimal residual
-iteration permits fewer stochastic histories to be used with each MCSA
-iteration while still maintaining good iterative performance. With
-respect to CPU time, there is a more pronounced effect due to the
-larger density of states created by the explicit ILUT
-preconditioning. Using a reduced domain fill level of 100 creates a
-situation where the expected value estimator is significantly more
-expensive per history to compute that the collision estimator due to
-the coupling of states.
+operations.
+
+Compared to the collision estimator, the minimal residual iteration
+permits fewer stochastic histories to be used with each MCSA iteration
+while still maintaining good iterative performance. With respect to
+CPU time, there is a more pronounced effect due to the larger density
+of states created by the explicit ILUT preconditioning. Using a
+reduced domain fill level of 100 creates a situation where the
+expected value estimator is significantly more expensive per history
+to compute that the collision estimator due to the coupling of states.
+For the collision estimator, using around 20,000 stochastic histories
+at every iteration gave the best iterative performance while 500
+histories per iteration gave the best iterative performance when using
+the expected value estimator. For purely serial computational
+performance, there is not a distinguishing feature for the fuel
+assembly problem presented that would cause one to choose one of the
+estimators and fixed point iterations over the other.
 
@@
   \label{fig:spn_estimator_time}
-\end{figure}
-
-From both Figure~\ref{fig:spn_estimator_iters} and
-Figure~\ref{fig:spn_estimator_time} we see that both estimators and
-fixed point iterations have comparable iterative and timing
-performance when 1,000 histories are used at each MCSA
-iteration. Using 1,000 histories per MCSA iteration, the infinity norm
-of the residual vector is presented at every MCSA iteration for both
-estimators and both fixed point iterations in
-Figure~\ref{fig:spn_estimator_convergence} as an additional means of
-comparison. At this number of histories, all MCSA combinations
-converge the fuel assembly problem monotonically at approximately the
-same rate, giving little reason to choose one over the other. The end
-result is that both estimators provide a viable method for getting
-good MCSA iterative performance with comparable timing
-performance. For the collision estimator, using around 20,000
-stochastic histories at every iteration gave the best iterative
-performance while 500 histories per iteration gave the best iterative
-performance when using the expected value estimator. For purely serial
-computational performance, there is not a distinguishing feature for
-the fuel assembly problem presented that would cause one to choose one
-of the estimators and fixed point iterations over the other as they
-all have similar minimum CPU times over the range of values tested.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=5in]{chapters/spn_equations/estimator_convergence.pdf}
-  \end{center}
-  \caption{\textbf{Residual infinity norm as a function of MCSA
-      iteration with 1,000 stochastic histories per iteration.}
-    \textit{Both the collision and expected value estimators were used
-      with the adjoint Monte Carlo solver to compute the correction at
-      each iteration. Each estimator was used with the Richardson and
-      minimal residual fixed point iterations within an MCSA
-      iteration.}}
-  \label{fig:spn_estimator_convergence}
 \end{figure}
@@
 explicit MCSA preconditioning necessary for achieve convergence. Using
-these results, we know compare MCSA directly to conventional linear
-solvers that would be typically used to solve the $SP_N$ form of the
-transport equation as presented here. In particular, we will use a
-pair of Krylov solvers with which we will compare numerical results to
-verify MCSA solutions the implementation used for this work.
+these results, we now compare MCSA to conventional Krylov solvers that
+would be typically used to solve the $SP_N$ form of the transport
+equation as presented here.
 
@@
 Eq~(\ref{eq:relative_difference_vector}) for all calculations and all
-energy groups. In general, we see good agreement in the two norm
-between the MCSA variants and the GMRES calculation, meaning that they
-have approximately the same variation from the BiCGStab
-calculation. We do note larger infinity norm values of this vector for
-some of the MCSA variants (i.e. group 1 in the 2-group calculation),
-however this large maximum value was not a global phenomenon with good
-agreement in the 2-norm for the calculations. Based on the good
-agreement between MCSA and GMRES for the relative difference of the
-group fluxes and the eigenvalue results presented here, the MCSA
-implementation should be deemed correct for the fuel assembly problem
-for serial calculations with multiple energy groups.
+energy groups. In general, we see good agreement in the 2-norm between
+the MCSA variants and the GMRES calculation, meaning that they have
+approximately the same variation from the BiCGStab calculation. We do
+note larger infinity norm values of this vector for some of the MCSA
+variants relative to the GMRES results (i.e. group 1 in the 2-group
+calculation), however, this large maximum value is apparently a local
+phenomenon with good agreement in the 2-norm for the
+calculations. Based on the good agreement between MCSA and GMRES for
+the relative difference of the group fluxes and the eigenvalue results
+presented here, the MCSA implementation should be deemed correct for
+the fuel assembly problem for serial calculations with multiple energy
+groups.
 
@@
 \label{sec:spn_comparison}
-In the previous section, numerical results using MCSA for the fuel
-assembly calculation were directly compared with production Krylov
-methods in order to verify their correctness. Deemed correct,
-performance of the algorithm for serial computations will now be
-addressed. With the same set of solver parameters used for
-verification caluclation, MCSA will now be compared to the same Krylov
-solvers in terms of both iterative performance and CPU timing. A
-Richardson iteration solution will also be used in order to assess the
+With the same set of solver parameters used for verification
+caluclation, MCSA will now be compared to the same Krylov solvers in
+terms of both iterative performance and CPU timing. A Richardson
+iteration solution will also be used in order to assess the
 acceleration provided by the residual Monte Carlo component of MCSA.
-
 To compare iterative performance, the number of linear solver
@@
 and then averaged over all eigenvalue iterations for each
-solver. Table~\ref{tab:spn_comparison_iterations} gives the average
-number of linear solver iterations required to converge a single
-eigenvalue iteration of the fuel assembly problem as a function of the
-number of energy groups. In general, the iterative performance of all
-the solvers tested aside from the Richardson iteration is comparable
-with BiCGStab performing the best. We expect this because not only are
-the multigroup $SP_N$ equations positive-definite, but they are also
+solver.
+
+Table~\ref{tab:spn_comparison_iterations} gives the average number of
+linear solver iterations required to converge a single eigenvalue
+iteration of the fuel assembly problem as a function of the number of
+energy groups. In general, the iterative performance of all the
+solvers tested aside from the Richardson iteration is comparable with
+BiCGStab performing the best. We expect this because not only are the
+multigroup $SP_N$ equations positive-definite, but they are also
 block-symmetric and therefore we expect the conjugate gradient-based
@@
   problem
-\item MCSA was verified to produce the same flux distribution and
-  k-eigenvalue for the fuel assembly as production Krylov methods
+\item MCSA was verified to produce the same k-eigenvalue for the fuel
+  assembly as production Krylov methods and gave good relative flux
+  agreement
 \item MCSA was observed to converge in fewer iterations per eigenvalue

### 2013-08-22 doc/parallel_verfication/compare_flux.cpp
updating parallel flux comparison to use relative values

@@ -1,0 +1,52 @@
+#include <iostream>
+#include <iomanip>
+#include <vector>
+#include <cmath>
+#include <cstdlib>
+#include <fstream>
+#include <algorithm>
+#include <string>
+#include <cassert>
+
+int main( int argc, char * argv[] )
+{
+    std::string file_1 = "cg_reference.dat";
+    std::string file_2 = "mcsa_1_8_10.dat";
+
+    int global_length;
+
+    std::ifstream data_file_1;
+    data_file_1.open( file_1.c_str() );
+    data_file_1 >> global_length;
+    std::vector<double> global_vector_1( global_length );
+    for ( int i = 0; i < global_length; ++i )
+    {
+        data_file_1 >> global_vector_1[i];
+    }
+    data_file_1.close();
+
+    std::ifstream data_file_2;
+    data_file_2.open( file_2.c_str() );
+    data_file_2 >> global_length;
+    std::vector<double> global_vector_2( global_length );
+    for ( int i = 0; i < global_length; ++i )
+    {
+        data_file_2 >> global_vector_2[i];
+    }
+    data_file_2.close();
+
+    std::vector<double> difference( global_length );
+    for ( int i = 0; i < global_length; ++i )
+    {
+        difference[i] = std::abs( global_vector_1[i] - global_vector_2[i] ) /
+                        std::abs( global_vector_1[i] );
+    }
+
+    std::cout << "Max Difference: " 
+              << *std::max_element( difference.begin(), difference.end() )
+              << std::endl;
+
+    std::cout << "Min Difference: " 
+              << *std::min_element( difference.begin(), difference.end() )
+              << std::endl;
+}

### 2013-08-22 doc/dissertation/document/backmatter/parallel_metrics.tex
moving parallel metrics to an appendix

--- 2013-07-24 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
@@
-\chapter{Parallel Monte Carlo Synthetic \\ Acceleration Methods\ }
-\label{ch:parallel_methods}
-
-In this chapter we briefly review particle transport methods for
-domain decomposed Monte Carlo. Considering a domain decomposed
-Neumann-Ulam method, we derive an analytic framework based on
-algebraic quantities from which to estimate aspects of its performance
-when applied to neutron transport systems. We then devise a parallel
-algorithm for the Neumann-Ulam method based on the multiple-set
-overlapping-domain decomposition algorithm and a parallel algorithm
-for the MCSA iteration that leverages the parallel Monte Carlo
-algorithm and general parallel matrix-vector operations. Using
-parallel solutions for the neutron diffusion problem, we verify the
-correctness of the parallel algorithm and implementation by comparing
-the numerical results against production Krylov methods. With the
-verified implementation of the algorithm, we perform parallel scaling
-studies to test its performance on a leadership class machine and
-compare this performance to the production Krylov methods.
-
 %%---------------------------------------------------------------------------%%
-\section{Domain Decomposed Monte Carlo\ }
-\label{sec:msod}
-Large-scale problems will for reasons typically related to memory
-restrictions or performance have their data partitioned such that each
-parallel process owns a subset of the equations in the linear
-system. Given this convention, the Neumann-Ulam Monte Carlo algorithm
-must perform random walks over a domain that is decomposed and must
-remain decomposed to avoid the same performance and memory
-restrictions that required the domain to be decomposed. To parallelize
-this algorithm, we then seek parallel Monte Carlo algorithms that
-handle domain decomposition.
-
-To motivate this problem, consider the square domain presented in
-Figure~\ref{fig:ddmc_example} and on this domain we imagine a Monte
-Carlo particle transport problem. If the domain were decomposed into 9
-subdomains as shown, each of those subdomains and their associated
-data (i.e. cross sections) could be owned by a different parallel
-process in the computation. In Figure~\ref{fig:ddmc_example}, the
-tracks of 3 particles that are born in the center subdomain are
-shown. In this example, particle A is first transported from the
-center domain to the domain directly to the left. Before the
-scattering event in the new domain may be processed, particle A must
-be communicated between between the two parallel processes that own
-those domains. For any given particle in the transport simulation,
-this communication event may hardly occur during the lifetime of the
-particle (particle B), or may occur many times (particle C) depending
-on the problem parameters and the random path in phase space taken by
-the particle. Scalable parallel algorithms for domain decomposed Monte
-Carlo are those that handle this domain-to-domain communication of
-particles effectively and balance it with the on-process computations
-for particle interactions, tallies, response calculations, and other
-simulation requirements.
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.5}{
-      \input{chapters/parallel_mc/ddmc_example.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Domain decomposed Monte Carlo transport example
-      illustrating how domain decomposition requires parallel
-      communication of particles.} \textit{If a particle crosses a
-      domain boundary it must be communicated to the next domain in
-      its path.}}
-  \label{fig:ddmc_example}
-\end{figure}
-
-For a domain decomposed Neumann-Ulam method, the situation is very
-much the same where instead the transport process is now discrete and
-the "physics" of the transport process is the Monte Carlo game with
-probabilities and weights described by
-Eq~(\ref{eq:neumann_ulam_decomposition}) for the given linear
-problem. Figure~\ref{fig:ddnu_example} gives the domain decomposed
-Neumann-Ulam analog of particle transport problem in
-Figure~\ref{fig:ddmc_example}. Imagine in this problem that we are
-solving for the monoenergetic scalar neutron flux as in the model
-diffusion problem presented in Appendix~\ref{chap:diffusion_problem}.
-In this problem, the solution has only a spatial dependence and
-therefore each point in the mesh corresponds to an equation in the
-resulting linear system. In Figure~\ref{fig:ddnu_example}, the domain
-has now been discretized by this mesh and again decomposed into 9
-subdomains, each owning a piece of the mesh and therefore the
-corresponding equations in the linear system. As we play the Monte
-Carlo game presented in Chapter~\ref{ch:stochastic_methods}, the
-discretization and resulting Neumann-Ulam decomposition of the problem
-describes how each state in the system is coupled to the other states
-in the system through the set of equations that form the linear
-system. This coupling is then responsible for the discrete random walk
-that each history takes as shown in this example. As discrete states
-(or rows in the linear system) that do not exist in the local domain
-are reached during the random walk by stochastic histories, the
-histories must be communicated to the subdomain that does own the
-discrete state in an analogous fashion to the particles in the
-previous example.
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.5}{
-      \input{chapters/parallel_mc/ddnu_example.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Domain decomposed Neumann-Ulam example illustrating
-      how domain decomposition requires parallel communication of
-      histories.} \textit{Each mesh point corresponds to an equation
-      in the linear system and the coupling among equations is
-      described by the discretization of the problem.}}
-  \label{fig:ddnu_example}
-\end{figure}
-
-To further motivate using a parallel algorithm similar to particle
-transport methods, again consider history A in
-Figure~\ref{fig:ddnu_example}. This history begins in the center
-subdomain and through the course of transitioning through states in
-the system arrives at a state that exists in the subdomain directly to
-the left of the center subdomain. History A must then be communicated
-from the center subdomain to the subdomain immediately to the left to
-continue the random walk. As with the particle transport system, we
-may find that histories in a domain decomposed Neumann-Ulam method are
-only communicated a few times (history B) if at all, or they may be
-communicated many times (history C) depending on the discretization
-and the outcome of the Monte Carlo game. In identical fashion to the
-particle transport problem, a domain decomposed Neumann-Ulam method
-has the same communication requirements with all histories in both
-examples requiring the same parallel operations. The amount of
-communication of histories from domain to domain will be the primary
-factor in the parallel scalability of the algorithm. In the next
-section, the number of histories communicated in a given problem will
-be quantified analytically for a simple model problem.
-
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{Analytic Framework for Domain-Decomposed Monte Carlo\ }
-\label{sec:analytic_framework}
-To date, parallel Neumann-Ulam methods have been limited to full
-domain replication with parallelism exploited through individual
-histories \cite{alexandrov_efficient_1998} and in this chapter we
-will exploit particle transport algorithms to alleviate this. To
-accomplish this, we recognize from the literature that stochastic
-histories must be transported from domain to domain as the simulation
-progresses and they transition to states that are not in the local
-domain. Because we have chosen a domain decomposition strategy in a
-parallel environment, this means that communication of these histories
-must occur between compute nodes owning neighboring pieces of the
-global domain. We wish to characterize this communication not only
-because communication is in general expensive, but also because these
-nearest-neighbor communication sequences, specifically, have poor
-algorithmic strong scaling \cite{gropp_high-performance_2001} in much
-the same way as a parallel matrix-vector multiply
-operation. Therefore, we desire a framework to provide a simple,
-analytic theory based on the properties of the transport system that
-will allow for estimates of the domain decomposed behavior of the
-Neumann-Ulam method in terms of the amount of information that must be
-communicated.
-
-When solving problems where the linear operator is symmetric, a host
-of analytic techniques exist based on the eigenvalue spectrum of the
-operator that characterize their behavior in the context of
-deterministic linear solvers. Using past work, these techniques are
-adapted to the domain decomposed Neumann-Ulam method using the
-one-speed, two-dimensional neutron diffusion equation and spatial
-discretization presented in Appendix~\ref{chap:diffusion_problem} as a
-model transport problem. Using the linear system generated by the
-discretization of the model problem, we use a spectral analysis to
-generate analytic relations for the eigenvalues of the operator based
-on system parameters. Using the eigenvalue spectra, we then build
-relationships to characterize the transport of stochastic histories in
-a decomposed domain and the fraction of histories that leak from a
-domain and will therefore have to be communicated. Finally, we compare
-these analytic results to numerical experiments conducted with the
-model transport problem.
-
-\clearpage
-
-\subsection{Spectral Analysis }
-\label{subsec:spectral_analysis}
-The convergence of the Neumann series in
-Eq~(\ref{eq:adjoint_neumann_series}) approximated by the Monte Carlo
-solver is dependent on the eigenvalues of the iteration matrix. We
-will compute these eigenvalues by assuming eigenfunctions of the form
-\cite{leveque_finite_2007}:
-\begin{equation}
-  \Phi_{p,q}(x,y) = e^{2 \pi \imath p x} e^{2 \pi \imath q y}\:,
-  \label{eq:eigenfunction_form}
-\end{equation}
-where different combinations of $p$ and $q$ represent the different
-eigenmodes of the solution. As these are valid forms of the solution,
-then the action of the linear operator on these eigenfunctions will
-yield the eigenvalues of the matrix as they exist on the unit circle
-in the complex plane.
-
-For the model problem, we first compute the eigenvalues for the
-diffusion operator $\ve{D}$ by applying the operator to the
-eigenfunctions and noting that $x=ih$ and $y=jh$:
-\begin{multline}
-  \ve{D}\Phi_{p,q}(x,y) = \lambda_{p,q}(\ve{D})
-  =\\ -\frac{D}{6h^2}\Big[4 e^{-2 \pi \imath p h} + 4 e^{2 \pi \imath
-      p h} + 4 e^{-2 \pi \imath q h} + 4 e^{2 \pi \imath q h} + e^{-2
-      \pi \imath p h} e^{-2 \pi \imath q h} \\ + e^{-2 \pi \imath p h}
-    e^{2 \pi \imath q h} + e^{2 \pi \imath p h} e^{-2 \pi \imath q h}
-    + e^{2 \pi \imath p h} e^{2 \pi \imath q h} - 20\Big] + \Sigma_a
-  \:.
-  \label{eq:deriv_diff_1}
-\end{multline}
-Using Euler's formula, we can collapse the exponentials to
-trigonometric functions:
-\begin{equation}
-  \lambda_{p,q}(\ve{D}) = -\frac{D}{6h^2}[ 8 \cos(\pi p h) + 8
-    \cos(\pi q h) + 4 \cos(\pi p h) \cos(\pi q h) - 20] + \Sigma_a\:.
-  \label{eq:deriv_diff_2}
-\end{equation}
-
-As Eq~(\ref{eq:diffusion_eq}) is diagonally dominant, point Jacobi
-preconditioning as outlined in
-\S~\ref{subsec:stochastic_preconditioning} is sufficient to reduce the
-spectral radius of the iteration matrix below unity and therefore
-ensure convergence of the Neumann series. Applying this
-preconditioner, we are then solving the following diffusion system:
-\begin{equation}
-  \ve{M}^{-1} \ve{D} \boldsymbol{\phi} = \ve{M}^{-1} \ve{s}\:.
-  \label{eq:precond_diffsion}
-\end{equation}
-The operator $\ve{M}^{-1} \ve{D}$ is merely the original diffusion
-operator with each row scaled by the diagonal component. As we have
-defined a homogeneous domain, the scaling factor, $\alpha$, is the
-same for all rows in the operator and defined as the $\phi_{i,j}$
-coefficient from Eq~(\ref{eq:fd_system}):
-\begin{equation}
-  \alpha = \Bigg[\frac{10 D}{3 h^2} + \Sigma_a\Bigg]^{-1}\:.
-  \label{eq:jacobi_scaling}
-\end{equation}
-Using this coefficient, we then have the following spectrum of
-preconditioned eigenvalues:
-\begin{equation}
-  \lambda_{p,q}(\ve{M}^{-1} \ve{D}) = \alpha \lambda_{p,q}(\ve{D})\:.
-  \label{eq:preconditioned_eigenvalues}
-\end{equation}
-
-The spectral radius of the iteration matrix is obtained by seeking its
-largest eigenvalue. As with the diffusion operator, we can use the
-same analysis techniques to find the eigenvalues for the iteration
-matrix. We use a few simplifications by noting that if the Jacobi
-preconditioned iteration matrix is $\ve{H} = \ve{I} -
-\ve{M}^{-1}\ve{D}$, the we except all terms on the diagonal of the
-iteration matrix to be zero such that we have the following stencil:
-\begin{multline}
-  \ve{H}\boldsymbol{\phi} = \frac{\alpha D}{6h^2}\Big[4 \phi_{i-1,j} +
-    4 \phi_{i+1,j} + 4 \phi_{i,j-1} + 4 \phi_{i,j+1}
-    +\\ \phi_{i-1,j-1} + \phi_{i-1,j+1} + \phi_{i+1,j-1} +
-    \phi_{i+1,j+1}\Big]\:.
-  \label{eq:iteration_stencil}
-\end{multline}
-Inserting the eigenfunctions defined by
-Eq~(\ref{eq:eigenfunction_form}) we get:
-\begin{multline}
-  \lambda_{p,q}(\ve{H}) = \frac{\alpha D}{6h^2}\Big[4 e^{-2 \pi \imath p
-      h} + 4 e^{2 \pi \imath p h} + 4 e^{-2 \pi \imath q h} + 4 e^{2
-      \pi \imath q h} + e^{-2 \pi \imath p h} e^{-2 \pi \imath q h}
-    \\ + e^{-2 \pi \imath p h} e^{2 \pi \imath q q} + e^{2 \pi \imath
-      p h} e^{-2 \pi \imath q h} + e^{2 \pi \imath p h} e^{2 \pi
-      \imath q h}\Big]\:,
-  \label{eq:iteration_deriv}
-\end{multline}
-which simplifies to:
-\begin{equation}
-  \lambda_{p,q}(\ve{H}) = \frac{\alpha D}{6h^2}[ 8 \cos(\pi p h) + 8
-    \cos(\pi q h) + 4 \cos(\pi p h) \cos(\pi q h)]\:,
-  \label{eq:iteration_spectrum}
-\end{equation}
-giving the eigenvalue spectrum for the Jacobi preconditioned iteration
-matrix. We find that the maximum eigenvalue exists when $p=q=0$,
-giving the following for the spectral radius of the Jacobi
-preconditioned iteration matrix:
-\begin{equation}
-  \rho(\ve{H}) = \frac{10 \alpha D}{3 h^2}\:.
-  \label{eq:iteration_radius}
-\end{equation}
-
-\clearpage
-
-\subsection{Neumann Series Convergence }
-\label{subsec:neumann_convergence}
-As outlined in \S\ref{sec:mc_preliminaries}, the adjoint Monte Carlo
-method is effectively an approximation to a stationary method. In the
-adjoint Neumann-Ulam method, $k$ iterations, equivalent to $k$
-applications of the iteration matrix, are approximated by a random
-walk of average length $k$ to yield the summation in
-Eq~(\ref{eq:adjoint_neumann_solution})
-\cite{dimov_new_1998,danilov_asymptotic_2000}. This random walk
-length, or the number of transitions before the termination of a
-history (either by the weight cutoff, absorption, or exiting the
-global domain) is therefore approximately the number of stationary
-iterations required to converge to the specified tolerance. In the
-case of the adjoint Neumann-Ulam method, no such tolerance exists,
-however, we have specified a weight cutoff, $W_c$, that determines
-when low-weight histories will be prematurely terminated as their
-contributions are deemed minute. After $k$ iterations, a stationary
-method is terminated as the error has reached some fraction,
-$\epsilon$, of the initial error:
-\begin{equation}
-  ||\ve{e}^{k}||_2 = \epsilon ||\ve{e}^0||_2\:.
-  \label{eq:linear_k_iter_norm4}
-\end{equation}
-Per Eq~(\ref{eq:linear_k_iter_norm3}), we see that this fraction is
-equivalent to $\epsilon = \rho(\ve{H})^k$. For the adjoint
-Neumann-Ulam method, if we take this fraction to be the weight cutoff,
-a measure of how accurately the contributions of a particular history
-to the solution are tallied, we then have the following relationship
-for $k$:
-\begin{equation}
-  k = \frac{ \log(W_c) }{ \log( \rho(\ve{H}) ) }\:.
-  \label{eq:analytic_k}
-\end{equation}
-This then gives us a means to estimate the length of the random walks
-that will be generated from a particular linear operator based on the
-eigenvalues of its iteration matrix (independent of the linear
-operator splitting chosen) and based on the weight cutoff parameter
-used in the Neumann-Ulam method.
-
-\subsection{Domain Leakage Approximations }
-\label{subsec:domain_leak_approx}
-In a domain decomposed situation, not all histories will remain within
-the domain they started in and must instead be communicated. This
-communication, expected to be expensive, was analyzed by Siegel and
-colleagues for idealized, load balanced situations for full nuclear
-reactor core Monte Carlo neutral particle simulations
-\cite{siegel_analysis_2012}.  To quantify the number of particles
-that leak out of the local domain they define a leakage fraction,
-$\Lambda$, as:
-\begin{equation}
-  \Lambda = \frac{average\ \#\ of\ particles\ leaving\ local\ domain}
-          {total\ of\ \#\ of\ particles\ starting\ in\ local\ domain}\:.
-          \label{eq:leakage_fraction}
-\end{equation}
-For their studies, Siegel and colleagues assumed that the value of
-$\Lambda$ was dependent on the total cross section of the system via
-the Wigner rational approximation. Outlined more thoroughly by Hwang's
-chapter in \cite{azmy_nuclear_2010}, we will use both the Wigner
-rational approximation and the mean chord approximation as a means to
-estimate the leakage fraction.
-
-In the case of domain decomposed linear systems, we can use diffusion
-theory to estimate the optical thickness of a domain in the
-decomposition and the corresponding leakage fraction in terms of
-properties of the linear operator and the discretization. To begin we
-must first calculate the mean distance a Monte Carlo history will move
-in the grid by computing the mean squared distance of its movement
-along the chord of length $l$ defined across the domain. After a
-single transition a history will have moved a mean squared distance
-of:
-\begin{equation}
-  \langle \bar{r_1^2} \rangle = (n_s h)^2\:,
-  \label{eq:step_1_length}
-\end{equation}
-where $h$ is the size of the discrete grid elements along the chord
-and $n_s$ is the number of grid elements a history will move on
-average every transition. For our diffusion model problem, $n_s$ would
-equate to the expected number of states in the $i$ (or $j$ as the problem is
-symmetric) direction that a history will move in a single
-transition and is dependent on the stencil used for the
-discretization. After $k$ transitions in the random walk, the history
-will have moved a mean squared distance of:
-\begin{equation}
-  \langle \bar{r_k^2} \rangle = k (n_s h)^2\:.
-  \label{eq:step_k_length}
-\end{equation}
-If our chord is of length $l$ and there are $n_i$ grid elements (or
-states to which a history may transition) along that chord, then $h =
-l / n_i$ giving:
-\begin{equation}
-  \langle \bar{r_k^2} \rangle = k \Bigg(\frac{n_s l}{n_i}\Bigg)^2\:.
-  \label{eq:step_k_length_sub}
-\end{equation}
-From diffusion theory, we expect the average number of interactions
-along the chord to be:
-\begin{equation}
-  \tau = \frac{l}{2 d \sqrt{\langle \bar{r_k^2} \rangle}}\:,
-  \label{eq:optical_thickness_1}
-\end{equation}
-where $d$ is the dimensionality of the problem and $\sqrt{\langle
-  \bar{r_k^2} \rangle}$ is effectively the mean free path of the Monte
-Carlo history in the domain. We can readily interpret $\tau$ to be the
-\textit{effective optical thickness} of a domain of length
-$l$. Inserting Eq~(\ref{eq:step_k_length_sub}) we arrive at:
-\begin{equation}
-  \tau = \frac{n_i}{2 d n_s \sqrt{k}}\:,
-  \label{eq:optical_thickness_2}
-\end{equation}
-which if expanded with Eq~(\ref{eq:analytic_k}) gives us the final
-relation for the effective optical thickness:
-\begin{equation}
-  \tau = \frac{n_i}{2 d n_s}
-  \sqrt{\frac{\log(\rho(\ve{H}))}{\log(W_c)}}\:.
-  \label{eq:optical_thickness_3}
-\end{equation}
-
-For optically thin domains, we expect that most histories will be
-communicated, while optically thick domains will leak the fraction of
-histories that did not interact within. Using the optical thickness
-defined in Eq~(\ref{eq:optical_thickness_3}), we can then complete the
-leakage approximations by defining the bounds of $\tau \rightarrow 0,
-\Lambda \rightarrow 1$ and $\tau \rightarrow \infty, \Lambda
-\rightarrow \tau^{-1}$.  With these bounds we can then define the
-leakage fraction out of a domain for the adjoint Neumann-Ulam method
-using the Wigner rational approximation:
-\begin{equation}
-  \Lambda = \frac{1}{1+\tau}\:,
-  \label{eq:wigner_domain_leakage}
-\end{equation}
-and using the mean-chord approximation:
-\begin{equation}
-  \Lambda = \frac{1-e^{-\tau}}{\tau}\:.
-  \label{eq:mean_chord_domain_leakage}
-\end{equation}
-Here, the leakage fraction is explicitly bound to the eigenvalues of
-the iteration matrix, the size of the domain, the content of the
-discretization stencil, and the weight cutoff selected to terminate
-low weight histories.
-
-\subsection{Numerical Experiments }
-\label{subsec:numerical_experiments}
-To test the relationships developed by the spectral analysis, we form
-two simple numerical experiments using the diffusion model problem:
-one to measure the length of the random walks as a function of the
-iteration matrix eigenvalues, and one to measure the domain leakage
-fraction as a function of the iteration matrix eigenvalues and the
-discretization properties. Before doing this, we verify our
-computation of the spectral radius of the iteration matrix by
-numerically computing the largest eigenvalue of the diffusion operator
-using an iterative eigenvalue solver. For this verification, a $100
-\times 100$ square grid with $h=0.01$, $h=0.1$, and $h=1.0$ and the
-absorption cross varied from 0 to 100 while the scattering cross
-section was fixed at unity. Figure~\ref{fig:measured_spec_rad} gives
-the measured spectral radius of the iteration matrix and the computed
-spectral radius for the preconditioned diffusion operator using
-Eq~(\ref{eq:iteration_radius}) as function of the absorption to
-scattering ratio $(\Sigma_a / \Sigma_s)$. Excellent agreement was
-observed between the analytic and numerical results with all data
-points computed within the tolerance of the iterative eigenvalue
-solver.
-\begin{figure}[t!]
-  \begin{spacing}{1.0}
-    \begin{center}
-      \includegraphics[width=4in,clip]{chapters/parallel_mc/measured_spec_rad.png}
-    \end{center}
-    \caption{\textbf{Measured and analytic preconditioned diffusion
-        operator spectral radius as a function of the absorption cross
-        section to scattering cross section ratio.} \textit{Values of
-        $h=0.01$, $h=0.1$, and $h=1.0$ were used. The red data was
-        computed numerically by an eigensolver while the black dashed
-        data was generated by Eq~(\ref{eq:iteration_radius}).}}
-    \label{fig:measured_spec_rad}
-  \end{spacing}
-\end{figure}
-
-\subsubsection{Random Walk Length }
-\label{subsubsec:walk_length}
-With the eigenvalue derivations verified, we can go about setting up
-an experiment to measure the length of the random walks generated by
-the adjoint Neumann-Ulam solver. To do this, we again use a $100
-\times 100$ square grid with $h=0.1$ and the absorption cross varied
-from 0 to 100 while the scattering cross section was fixed at
-unity. Three weight cutoff values of \sn{1}{-2}, \sn{1}{-4}, and
-\sn{1}{-8} were used with 10,000 histories generated by a point source
-of strength 1 in the center of the domain. For each of the histories,
-the number of transitions made was tallied to provide an effective
-value of $k$ for each history. This value was then averaged over all
-histories to get a measured value of $k$ for the particular
-operator. On the left, Figure~\ref{fig:measured_length} presents these
-measurements as well as the analytic result computed by
-Eq~(\ref{eq:analytic_k}) as a function of the iteration matrix
-spectral radius, $\rho(\ve{H})$. On the right,
-Figure~\ref{fig:measured_length} gives the relative difference between the
-predicted and observed results. We note good qualitative agreement
-between the measured and analytic results. However, we observe a
-larger relative difference for both long and short random walks.
-\begin{figure}[t!]
-  \begin{spacing}{1.0}
-    \begin{center}
-      \includegraphics[width=6.0in,clip]{chapters/parallel_mc/measured_length_2.pdf}
-    \end{center}
-    \caption{\textbf{Measured and analytic random walk length as a
-        function of the iteration matrix spectral radius.} \textit{The
-        weight cutoff was varied with \sn{1}{-2}, \sn{1}{-4}, and
-        \sn{1}{-8}. In the left plot, the red data was computed
-        numerically by an adjoint Neumann-Ulam implementation while
-        the black dashed data was generated by
-        Eq~(\ref{eq:analytic_k}). In the right plot, the relative
-        difference between the predicted and measured results is
-        presented for each weight cutoff.}}
-    \label{fig:measured_length}
-  \end{spacing}
-\end{figure}
-
-\subsubsection{Domain Leakage }
-\label{subsubsec:domain_leakage}
-Finally, we seek to measure the leakage from a domain in a domain
-decomposed Monte Carlo calculation and assess the quality of our
-analytic relation for the optical thickness of a domain and the
-associated leakage approximations. For this experiment, a square grid
-with $h=0.1$ was decomposed into 9 square domains, 3 in each cardinal
-direction with measurements occurring in the central domain without
-boundary grid points. For the cross sections, the absorption cross
-section was varied from 1 to 100 while the scattering cross
-section was set to zero to create a purely absorbing environment
-with weight cutoff of \sn{1}{-4}. The optical thickness of these
-domains will vary as a function of the absorption cross section if
-the other parameters are fixed.
-
-To compute the optical thickness, along with the spectral radius as
-given by Eq~(\ref{eq:iteration_radius}), we also need the parameters
-$n_i$ and $n_s$ which respectively describe the typical domain length
-and the average number of states moved along that typical length per
-history transition. For our grid above, the domains are varied in size
-with $50 \times 50$, $100 \times 100$, and $200 \times 200$ cells
-giving $n_i=50$, $n_i=100$, and $n_i=200$ grid points or states along
-the typical length of the domain respectively. Looking at the
-Laplacian stencil in Eq~(\ref{eq:nine_point_stencil}), we see that all
-history transitions will only move a single state in either the $i$ or
-$j$ directions due to the symmetry of the problem. Furthermore, if we
-choose the $i$ direction, not all states we will transition to will
-move the history in that direction. Therefore, we look to the
-definition of the iteration matrix in Eq~(\ref{eq:iteration_stencil})
-and the definition of the adjoint probability matrix in
-Eq~(\ref{eq:adjoint_probability}) to estimate the $n_s$ parameter. For
-a particular transition starting at state $(i,j)$, 6 of the 8 possible
-new states in the stencil move the history in $i$ direction with
-relative coefficients of 4 for moving in the $(\pm i,0)$ direction and
-of 1 for moving in the $(\pm i,\pm j)$. These coefficients dictate the
-frequency those states are visited relative to the others. For those 6
-states we can visit along the typical length, their sum is 12 out of
-the total 20 for the coefficients for all possible states with their
-ratio giving $n_s = \frac{3}{5}$.
-
-To compute the leakage fraction numerically, \sn{3}{5} histories were
-sampled from a uniform source of strength unity over the global
-domain. At the start of a stage of histories, the number of histories
-starting in the center domain was computed and as the stage
-progressed, the number of histories that exited that domain was
-tallied with the ratio of the two numbers providing a numerical
-measure for the leakage fraction. Figure~\ref{fig:measured_leakage}
-gives the domain leakage measurements for the domain in the center of
-the global grid as well as the analytic result computed by
-Eqs~(\ref{eq:wigner_domain_leakage}) and
-(\ref{eq:mean_chord_domain_leakage}) as a function of the iteration
-matrix spectral radius.
-\begin{figure}[t!]
-  \begin{spacing}{1.0}
-    \begin{center}
-      \includegraphics[width=4.75in,clip]{chapters/parallel_mc/leakage_variation_2.pdf}
-    \end{center}
-    \caption{\textbf{Measured and analytic domain leakage as a
-        function of the iteration matrix spectral radius.} \textit{To
-        test the behavior with respect to domain size, $n_i=50$,
-        $n_i=100$,and $n_i=200$ were used. The red data was computed
-        numerically by a domain-decomposed adjoint Neumann-Ulam
-        implementation, the black dashed data was generated by
-        Eq~(\ref{eq:mean_chord_domain_leakage}) using the mean-chord
-        approximation, and the dashed-dotted black data was generated
-        by Eq~(\ref{eq:wigner_domain_leakage}) using the Wigner
-        rational approximation.}}
-    \label{fig:measured_leakage}
-  \end{spacing}
-\end{figure}
-Again, we note good qualitative agreement between the measured and
-analytic quantities but we begin to see the limits of the leakage
-approximations.
-
-To compare the quality of the two approximations, the absolute
-difference between the computed leakage fraction and that generated by
-the Wigner rational and mean chord approximations is plotted in
-Figure~\ref{fig:leakage_error} for all domain sizes tested.
-\begin{figure}[t!]
-  \begin{spacing}{1.0}
-    \begin{center}
-      \includegraphics[width=4.75in,clip]{chapters/parallel_mc/leakage_error_2.pdf}
-    \end{center}
-    \caption{\textbf{Measured and analytic domain leakage absolute
-        difference as a function of the iteration matrix spectral radius.}
-      \textit{To test the behavior with respect to domain size,
-        $n_i=50$ (green), $n_i=100$ (blue), and $n_i=200$ (red) were
-        used. The dashed lines represent the difference using the Wigner
-        rational approximation while the solid lines represent the
-        difference using the mean-chord approximation.}}
-    \label{fig:leakage_error}
-  \end{spacing}
-\end{figure}
-From these difference results, the mean chord approximation is shown to
-have a lower difference for ill-conditioned systems as compared to the
-Wigner approximation while the Wigner approximation produces less
-difference for more well-conditioned systems. We also note that for the
-optically thick domains, the difference is likely corresponded to that
-observed in Figure~\ref{fig:measured_length} for the $k$ parameter
-while the large relative difference in $k$ for optically thin domains does
-not affect the approximation significantly. In general, the mean chord
-approximation is a better choice to estimate the leakage fraction in a
-domain from the adjoint Neumann-Ulam method and except for a single
-data point with $n_i=50$, the mean chord approximation yielded leakage
-fractions within 0.05 of the measured results. As the domain becomes
-more optically thick (with both increasing $n_i$ and decreasing
-$\rho(\ve{H})$), the approximations are more accurate.
-
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{Domain Decomposed Neumann-Ulam Algorithm\ }
-\label{sec:asynchronous_algorithm}
-In the context of radiation transport, in 2009 Brunner and colleagues
-provided a fully asynchronous domain decomposed parallel algorithm as
-implemented in production implicit Monte Carlo codes
-\cite{brunner_efficient_2009}. We will adapt their algorithm and
-directly apply it to a parallel formulation of the Neumann-Ulam
-method. Direct analogs can be derived from their works by noting that
-the primary difference between solving a linear transport system with
-Monte Carlo methods and traditional fixed source Monte Carlo transport
-problems is the content of the Markov chains that are generated.
-
-The transitions represented by these chains are bound by probabilities
-and weights and are initiated by the sampling of a discrete source. In
-the context of transport problems, those transitions represent events
-such as particle scattering and absorption with probabilities that are
-determined by physical data in the form of cross sections. For
-stochastic matrix inversion, those transitions represent moving
-between the equations of the linear system (and therefore the
-components of phase space which they represent) and their
-probabilities are defined by the coefficients of those
-equations. Ultimately, we tally the contributions to generate
-expectation values in the desired states as we progress through the
-chains. Therefore, parallel methods for Monte Carlo radiation
-transport can be abstracted and we can use those concepts that apply
-to matrix inversion methods as an initial means of developing a
-parallel Neumann-Ulam-type solver.
-
-In this section Brunner and Brantley's fully asynchronous algorithm,
-which was effectively implemented verbatim for this work, is presented
-along with its application to the Neumann-Ulam method. For
-considerably more detail, the algorithm presented can be found in
-\cite{brunner_efficient_2009}. In their work they identify two data
-sets that are required to be communicated: the sharing of particles
-that are transported from one domain to another and therefore from one
-processor to another and a global communication that signals if
-particle transport has been completed on all processors. Both of these
-communication sequences will be addressed at a high level along with
-how the Monte Carlo data structures they require are constructed in
-parallel.
-
-\subsection{Parallel Transport Domain and Source }
-\label{subsec:domain_generation}
-To utilize a parallel transport algorithm, we must generate the
-required data with the correct parallel decomposition. For the
-Neumann-Ulam method, the transport domain consists of all states in
-the system that are local, and the probabilities and weights for all
-state transitions possible in the local domain. At the core of this
-representation is the Neumann-Ulam decomposition of the linear
-operator as given by Eq~\ref{eq:neumann_ulam_decomposition}. Given an
-input decomposition for the linear operator, by definition the
-Neumann-Ulam decomposition will have the same parallel
-decomposition. In addition, any modification made to the linear
-operator through preconditioning or relaxation parameters will also
-modify the Neumann-Ulam decomposition. All possible transitions are
-described by the local graph of the sparse input matrix. In addition,
-any left preconditioning or relaxation parameters will modify the
-right hand side of the linear system and therefore the fixed source in
-the Monte Carlo calculation. This source vector will have the same
-parallel decomposition as the input matrix and therefore all of the
-birth states for the histories will exist for at least a single
-transition event within the local domain.
-
-Compared to the serial construction of the Neumann-Ulam decomposition,
-data for all states that is required to be on process must be
-collected in parallel. For the pure domain decomposition case, given
-an input parallel decomposition for the Neumann-Ulam decomposition,
-for all local states $m$ we require access to $P_{mn}$ and $W_{mn}$
-(or $P^T_{mn}$ and $W^T_{mn}$) for the adjoint method for all possible
-states $n$. Given by Figure~\ref{fig:diffusion_graph}, consider the
-adjacency graph of a single state in the neutron diffusion matrix for
-the model problem presented in Appendix~\ref{chap:diffusion_problem}.
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.5}{ \input{chapters/parallel_mc/stencil_graph.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Neutron diffusion equation state adjacency graph.}
-    \textit{The structure of the graph comes from the discretization
-      of the Laplacian operator describing the diffusion
-      physics. Adjacent boundary states are collected by traversing
-      the graph one step for all local boundary states and collecting
-      the equations for those states that are not local. For a given
-      transition in the random walk for the diffusion problem, a
-      history at mesh point $(i,j)$ may transition to all adjacent
-      mesh points including itself, corresponding to a global state
-      transition from $m$ to $n$ in the linear system.}}
-  \label{fig:diffusion_graph}
-\end{figure}
-
-In this example, we currently reside in state $m$ of the system,
-directly correlating to physical location $(i,j)$ in the mesh (the
-center node). The discretization stencil of the diffusion problem
-dictates the structure of this graph and the states to which a history
-may transition. If we play the Monte Carlo game to move to a new state
-$n$, then we may move to any of the nodes in this graph, including the
-node at which we started. If grid point $(i,j)$ in this example is on
-the right of the local domain boundary and we transition to the right
-to state $n$ corresponding to mesh point $(i+1,j)$, we are now in a
-state that is owned by the adjacent domain. We must then gather the
-data in the Neumann-Ulam decomposition from the neighboring process
-that owns grid point $(i+1,j)$. Doing this provides us with the data
-required by the estimators permitting $P_{mn}$ and $W_{mn}$ to be
-computed locally for this particular boundary transition. In addition,
-we also collect the identification number for the process that owns
-this state such that we have an address to send all histories that
-leave the local domain boundary for this state. For the parallel
-source, these adjacent states are not required as histories may only
-be born in states in the local domain. Once the local Neumann-Ulam
-decomposition has been generated with the proper data collected from
-adjacent domains, Monte Carlo transport may proceed in parallel.
-
-\clearpage
-
-\subsection{Domain Decomposed Algorithm}
-\label{subsec:parallel_mc_algorithm}
-
-Here we present in detail Brunner and Brantley's 2009 algorithm in
-detail and discuss how it is adapted to parallelize the Neumann Ulam
-method. Presented in Algorithm~\ref{alg:parallel_mc_algorithm}, the
-top level sequence performs the task of managing history transport
-through the local domain, communication of histories to adjacent
-domains, and the completion of transport. For each of these specific
-tasks, additional algorithms, shown in bold in
-Algorithm~\ref{alg:parallel_mc_algorithm}
-(e.g. \textbf{LocalHistoryTransport()}), are presented for additional
-detail in the same manner as Brunner and Brantley.
-
-\begin{algorithm}[h!]
-  \caption{Parallel Neumann-Ulam Algorithm}
-  \label{alg:parallel_mc_algorithm}
-  \begin{algorithmic}[1]
-    \State get list of neighbor processors 
-    \Comment{Each neighbor owns an adjacent subdomain}
-    \ForAll{neighbors}
-    \State post nonblocking receive for maximum history buffer size
-    \State allocate history buffer
-    \EndFor
-    \State \textit{historiesCompleted} = 0
-    \Comment{local+child finished histories}
-    \State \textit{localProcessed} = 0
-    \Comment{local trajectories computed (not necessarily finished)}
-    \State calculate parent and children processor ID numbers in
-    binary tree
-    \ForAll{child processes}
-    \State post nonblocking receive for \textit{historiesCompleted} tally
-    \EndFor
-    \State post nonblocking receive for stop message from parent
-    \While{stop flag not set}
-    \If{any local histories in source or stack}
-    \State \textbf{LocalHistoryTransport()}
-    \State ++\textit{localProcessed}
-    \EndIf
-    \If{message check period == \textit{localProcessed} || no local histories}
-    \State \textbf{ProcessMessages()}
-    \State \textit{localProcessed} = 0
-    \EndIf
-    \If{no local histories}
-    \State \textbf{ControlTermination()}
-    \EndIf
-    \EndWhile
-    \State cancel outstanding nonblocking requests
-    \State free all history buffers
-  \end{algorithmic}
-\end{algorithm}
-
-Successful execution of this algorithm requires construction of the
-parallel Neumann-Ulam decomposition as described in the preceding
-section. This data is used to begin
-Algorithm~\ref{alg:parallel_mc_algorithm} where lines 1-5 use the
-collected list of neighboring processors generated while building the
-Neumann-Ulam decomposition to setup the set of asynchronous messages
-required for domain-to-domain communication. Again, consider this
-communication pattern for the 9 subdomain example given by
-Figure~\ref{fig:nearest_neighbor_comm}. In this pattern, each boundary
-domain has two neighbors and the center domain four neighbors with
-which they will communicate parallel histories and this communication
-goes both ways as represented by the adjacent arrows in the
-figure. For each set of neighbors, a non-blocking send and receive is
-required with data buffers allocated with a user-defined size prepared
-for incoming and outgoing histories. This nonblocking structure is
-critical to the performance of the algorithm in that it permits local
-history transport to continue while new histories to transport are
-being collected in the buffers. When a given process is ready to do
-more work, it can check these data buffers for incoming histories
-requiring further transport. In this way there is a maximum amount of
-overlap between communication and transport of histories.
-
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.5}{ \input{chapters/parallel_mc/domain_to_domain.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Nearest neighbor history communication sequence.}
-    \textit{Each subdomain in the system has a set of nearest
-      neighbors determined by the parallel adjacency graph of the
-      input matrix. The subdomains are indexed by an integer.}}
-  \label{fig:nearest_neighbor_comm}
-\end{figure}
-
-Once the nearest-neighbor communication sequence has been prepared,
-the completion of transport sequence is readied in lines 6 and 8-12 in
-Algorithm~\ref{alg:parallel_mc_algorithm}. In these lines we are
-setting up an asynchronous binary communication tree as presented for
-the same 9 subdomain example in Figure~\ref{fig:binary_comm_tree}. In
-this communication pattern, each process has one parent process
-(except for the MASTER process 0) to which it will nonblocking send
-the number of histories that terminated their transport procedure by
-weight cutoff in the local domain. Equivalently, each process has up
-to two child processes from which it will receive their completed
-number of histories with a nonblocking receive operation. Setting up a
-tree in this manner lets the completed history tally
-(\textit{historiesCompleted} in the algorithm) be updated
-incrementally and funneled to the root process. Because we are
-solving fixed source problems with the Neumann-Ulam algorithm without
-any type of variance reduction that may generate more histories that
-we started with, once the root process completion sum tallies to the
-number of histories in the source, transport is complete. Once this
-occurs, the root process nonblocking sends a process to its children
-and each process nonblocking receives a stop message from its parent
-as shown in Figure~\ref{fig:binary_comm_tree}. The stop message is
-then propagated up the tree in the same manner.
-
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.25}{
-      \input{chapters/parallel_mc/binary_comm_tree.pdftex_t} }
-    \caption{\textbf{Binary communication tree for coordinating the
-        end of a parallel Neumann-Ulam solve.} \textit{Each child
-        process reports to its parents how many histories completed
-        within its domain. When the root process sums all complete
-        histories, it forwards the stop signal to its children which
-        in turn forward the message up the tree. The subdomains are
-        indexed by an integer.}}
-  \end{center}
-  \label{fig:binary_comm_tree}
-\end{figure}
-
-With these communication structures prepared, we can now enter the
-main transport loop at line 13 of
-Algorithm~\ref{alg:parallel_mc_algorithm}. In this loop, a set of
-mutually exclusive tasks enabled by the fully asynchronous
-communication patterns are executed until the stop signal is received
-from the parent process in the binary tree. In this case, mutually
-exclusivity permits all local operations to occur independently of
-other processes in the system and their current state. For each
-process, there are two data structures from which histories may be
-obtained for the transport procedure. The first is the local source
-and the second is a LIFO (last in first out) stack of histories
-transported to the local domain from the adjacent domain. In lines
-14-17, if histories exist in either of these data structures, then
-they are transported through the local domain using
-Algorithm~\ref{alg:local_history_transport}. If the history is
-terminated by weight cutoff the \textit{historiesCompleted} tally is
-updated. If it hits the boundary of the domain it is added to the
-outgoing history buffer for the neighboring domain and if the buffer
-is full, it is sent to the neighboring domain with a nonblocking
-operation and the memory associated with that buffer reallocated. In
-all instances that Algorithm~\ref{alg:local_history_transport} is
-executed, \textit{localProcessed} is incremented to account for
-histories that have been processed locally.
-
-\begin{algorithm}[h!]
-  \caption{\textbf{LocalHistoryTransport()}}
-  \label{alg:local_history_transport}
-  \begin{algorithmic}[1]
-    \State transport history through the domain until termination
-    \If{history is in neighbor boundary state}
-    \State add history to neighbor buffer
-    \If{neighbor buffer is full}
-    \State nonblocking send history buffer to neighbor
-    \State allocate new history buffer for neighbor
-    \EndIf
-    \Else
-    \If{history terminated by weight cutoff}
-    \State post-process history
-    \State ++\textit{historiesCompleted}
-    \EndIf
-    \EndIf
-  \end{algorithmic}
-\end{algorithm}
-
-Continuing in the transport loop, if there are no local histories to
-transport or the \textit{localProcessed} count has reached some
-user-defined check frequency, lines 18-21 in
-Algorithm~\ref{alg:local_history_transport} check for more histories
-to transport in the incoming data buffers by calling
-Algorithm~\ref{alg:process_messages}. For each neighboring domain in
-the problem, if there are histories in those data buffers then they
-are added to the stack from processing and the nonblocking receive
-operation re-instantiated. In addition, the terminated histories count
-is updated with those values received from the child processes.
-
-\begin{algorithm}[h!]
-  \caption{\textbf{ProcessMessages()}}
-  \label{alg:process_messages}
-  \begin{algorithmic}[1]
-    \ForAll{received history buffers from neighbor}
-    \State unpack number of histories in buffer
-    \State add the histories to the running stack
-    \State repost nonblocking receive with neighbor
-    \EndFor
-    \ForAll{\textit{historiesCompleted} messages from children}
-    \State \textit{historiesCompleted} += message value
-    \State repost nonblocking receive with child
-    \EndFor
-  \end{algorithmic}
-\end{algorithm}
-
-Finally, the transport loop finishes in lines 22-24 of
-Algorithm~\ref{alg:local_history_transport} by calling
-Algorithm~\ref{alg:control_termination} if there are no local
-histories to transport in the stack or the source. In
-Algorithm~\ref{alg:control_termination}, we continue to forward the
-terminated history tally down to the parent in the binary tree and
-send all history buffers to our neighbors, even if they are not
-full. If Algorithm~\ref{alg:control_termination} is called by the
-MASTER process, it checks for completion of the problem and if it is
-complete, forwards to stop flag onto its children as in
-Figure~\ref{fig:binary_comm_tree}. If the process is not the MASTER,
-then we check for a stop signal from the parent and forward to the
-child processes if transport is complete.
-
-\begin{algorithm}[h!]
-  \caption{\textbf{ControlTermination()}}
-  \label{alg:control_termination}
-  \begin{algorithmic}[1]
-    \State nonblocking send an partially full history buffers
-    \ForAll{\textit{historiesCompleted} messages from children}
-    \State \textit{historiesCompleted} += message \textit{historiesCompleted}
-    \State repost nonblocking receive with child
-    \EndFor
-    \If{MASTER processor}
-    \If{\textit{historiesCompleted} == global \# of source histories}
-    \State set stop flag
-    \ForAll{children}
-    \State nonblocking send stop message to child
-    \EndFor
-    \EndIf
-    \Else
-    \State nonblocking send \textit{historiesCompleted} to parent
-    \State \textit{historiesCompleted} = 0
-    \State check for stop signal from parent
-    \If{stop signal from parent}
-    \ForAll{children}
-    \State nonblocking send stop message to child
-    \EndFor
-    \EndIf
-    \EndIf
-  \end{algorithmic}
-\end{algorithm}
-
-In this manner the transport loop continues until all process have
-received the stop signal from their parents. Older versions of this
-algorithm, particularly some of those presented in
-\cite{brunner_comparison_2006}, use a master/slave approach as given
-by Figure~\ref{fig:master_comm_tree} for the completion of a transport
-stage instead of the binary tree scheme. In this approach, the root
-process still manages the final summation of the completion tally, it
-directly receives completed tally results from all other processes in
-the problem instead of from just its children. Although this
-implementation is potentially simpler to understand than the binary
-tree approach, the authors of \cite{brunner_comparison_2006} observed
-that this type of termination pattern, even when implemented in a
-fully asynchronous manner, was a major scalability bottleneck. This
-bottleneck is due to the fact that the master process falls behind the
-others, creating a load imbalance even with the addition of a few
-integer addition operations. We will explicitly demonstrate in scaling
-studies how this bottleneck also occurs within a Neumann-Ulam
-implementation of this algorithm thus requiring the implementation of
-a binary communication tree.
-
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{0.75}{
-      \input{chapters/parallel_mc/master_comm_tree.pdftex_t} }
-    \caption{\textbf{Master/slave scheme for coordinating the end of a
-        parallel Neumann-Ulam solve.} \textit{Each slave process
-        reports to the master how many histories completed within its
-        domain. When the master process sums all complete histories,
-        it sends the stop signal to all slave processes. The
-        subdomains are indexed by an integer.}}
-  \end{center}
-  \label{fig:master_comm_tree}
-\end{figure}
-
-There is an additional advantage to Brunner and Brantley's work that
-although not immediately applicable to this work has the potential to
-provide value in the future. In addition to a robust fully
-asynchronous communication pattern, this algorithm may also be modified
-to account for situations where the total number of histories in a
-given stage are not known before starting transport. From a physics
-perspective, we might expect this for situations where perhaps an
-(n,2n) interaction occurs in a neutronics problem. In this case, the
-algorithm is modified to account for both histories terminated and
-created and several mechanisms are introduced to determine completion
-of the transport stage. For future variations of this work, certain
-variance reduction techniques which create histories, such as
-splitting, have the potential to be successfully employed as a means
-of accelerating the time to solution for a given problem using
-MCSA. The parallel Neumann-Ulam algorithm presented here may be
-adapted to account for these additional techniques.
-
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{Multiple-Set Overlapping-Domain Algorithm\ }
-\label{subsec:msod}
-Although the implementation presented in the previous section was
-observed by Brunner and Brantley to be robust and allowed for scaling
-to large numbers of processors, performance issues were still noted
-with parallel efficiency improvements needed in both the weak and
-strong scaling cases for unbalanced problems. These results led them
-to conclude that a combination of domain decomposition and domain
-replication could be used to solve some of these issues. In 2010,
-Wagner and colleagues developed the \textit{multiple-set
-  overlapping-domain} (MSOD) decomposition for parallel Monte Carlo
-applications for full-core light water reactor analysis
-\cite{wagner_hybrid_2010}. In their work, an extension of Brunner's,
-their scheme employed similar parallel algorithms for particle
-transport but a certain amount of overlap between adjacent domains was
-used to decrease the number of particles leaving the local domain. In
-addition, Wagner utilized a level of replication of the domain such
-that the domain was only decomposed on $O(100)$ processors and if
-replicated $O(1,000)$ times potentially achieves efficient simulation
-on $O(100,000)$ processors, thus providing both spatial and particle
-parallelism.
-
-Each collection of processors that constitutes a representation of the
-entire domain is referred to as a set, and within a set overlap occurs
-among its sub-domains. The original motivation was to decompose the
-domain in a way that it remained in a physical cabinet in a large
-distributed machine, thus reducing latency costs during
-communication. A multiple set scheme is also motivated by the fact
-that communication during particle transport only occurs within a set,
-limiting communications during the transport procedure to a group of
-$O(100)$ processors, a number that was shown to have excellent
-parallel efficiencies in Brunner's work and therefore will scale well
-in this algorithm. The overlapping domains within each set also
-demonstrated reduced communication costs. On each processor, the
-source is sampled in the local domain that would exist if no overlap
-was used while tallies can be made over the entire overlapping domain.
-
-To demonstrate this, consider the example adapted from Mervin's work
-with Wagner and others in the same area \cite{mervin_variance_2012}
-and presented in Figure~\ref{fig:msod_example}.
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.5}{
-      \input{chapters/parallel_mc/msod_example.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Overlapping domain example illustrating how domain
-      overlap can reduce communication costs.}
-    \textit{All particles start in the blue region of interest. The
-      dashed line represents 0.5 domain overlap between domains.}}
-  \label{fig:msod_example}
-\end{figure}
-In this example, 3 particle histories are presented emanating from the
-blue region of interest. Starting with particle A, if no domain
-overlap is used then the only the blue domain exists on the starting
-processor. Particle A is then transported through 3 other domains
-before the history ends, therefore requiring three communications to
-occur in Brunner's algorithm. If a 0.5 domain overlap is permitted as
-shown by the dashed line, then the starting process owns enough of the
-domain such that no communications must occur in order to complete the
-particle A transport process. Using 0.5 domain overlap also easily
-eliminates cases such as that represented by the path of particle
-C. In this case, particle C is scattering between two adjacent
-domains, incurring a large latency cost for a single
-particle. Finally, with particle B we observe that 0.5 domain overlap
-will still not eliminate all communications. However, if 1 domain
-overlap were used, the entire geometry shown in
-Figure~\ref{fig:msod_example} would be contained on the source
-processor and therefore transport of all 3 particles without
-communication would occur.
-
-Wagner and colleagues used this methodology for a 2-dimensional
-calculation of a pressurized water reactor core and varied the domain
-overlap from 0 to 3 domain overlap (a $7 \times 7$ box in the context
-of our example) where a domain contained an entire fuel assembly. For
-the fully domain decomposed case, they observed that 76.12\% of all
-source particles leave the domain. At 1.5 domain overlap, the
-percentage of source particles born in the center assembly leaving the
-processor domain dropped to 1.05\% and even further for 0.02\% for the
-3 domain overlap. Based on their results, we hypothesize that the
-overlap approach coupled with the multiple sets paradigm that will
-enhance the scalability of the pure domain-decomposition algorithm for
-Neumann-Ulam presented in the previous section.
-
-\subsection{Generating the MSOD Decomposition }
-\label{subsec:msod_generation}
-In \S~\ref{subsec:domain_generation} we discussed how we generated the
-parallel transport domain from the Neumann-Ulam decomposition of a
-domain decomposed linear operator. We can readily adapt those data
-structures to account for the extra information required by an
-implementation of the MSOD algorithm. First, we consider the
-generation of overlap. Conveniently, this is identical to the
-neighboring states discovery problem discussed in conjunction with
-Figure~\ref{fig:diffusion_graph}. To generate the boundary for the
-local transport domain, we needed to gather all data from the
-immediately adjacent states in the system that did not already exist
-in the local domain. We solved this problem by traversing the graph
-of the matrix for each of the boundary states and determining which
-processes owned those adjacent states.
-
-To generate overlap, we use the identical algorithm but in this case
-we perform as many graph level traversals as the specified amount of
-overlap. For example, consider the analytic relations derived in
-\S~\ref{sec:analytic_framework} for the length of random walks in a
-Neumann-Ulam method and the amount of leakage from a domain
-considering a neutron diffusion problem. From these relations we might
-determine that, for our particular problem, if we can grow the domain
-by 10 additional discrete state transitions, we can reduce the number
-of histories that must be communicated by a significant fraction. We
-determine the information we must gather from neighboring domains to
-build the overlap by traversing the graph of the matrix outward from
-the boundary 10 levels. At each step, we find the data in the adjacent
-domain that we require and make that data local, incrementing the size
-of the overlap. Once the overlap has been gathered, one extra graph
-traversal on the boundary states is performed to get the new
-neighboring states and their owning processes.
-
-Once overlap has been generated, the transport domain for a single set
-is complete. However, if multiple sets are to be used, an additional
-set of parallel procedures is required to generate the necessary data
-structures. Figure~\ref{fig:msod_construction} gives a schematic
-representation of the MSOD construction process using 4 sets. For a
-problem where $P$ parallel processors are available and $S$ sets are
-to be used, each set is allocated $P/S$ processors for computation. On
-the first $P/S$ processors, the linear problem is generated. From the
-linear problem, the linear operator is used to construct the transport
-domain, the solution vector used to construct the tallies, and the
-forcing term vector used to construct the fixed source. Once these
-Monte Carlo data structures have been generated, they are broadcast to
-the remaining sets in the problem as shown in
-Figure~\ref{fig:msod_construction} such that we have $S$ replications
-of the Monte Carlo problem for a single instance of the original
-linear problem.
-
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{0.6}{ \input{chapters/parallel_mc/msod_construction.pdftex_t} }
-  \end{center}
-  \caption{\textbf{MSOD construction for 4 sets with overlap.}
-    \textit{The linear system is used to construct the Monte Carlo
-      transport domain on the primary set. The Monte Carlo data
-      structures are then broadcast among the blocks to all sets in
-      the system.}}
-  \label{fig:msod_construction}
-\end{figure}
-
-\clearpage
-
-\subsection{MSOD Neumann Ulam Algorithm }
-\label{subsec:msod_algorithm}
-
-With the ability to generate the transport domain within an MSOD
-decomposition as well as the source and tally structures, we can now
-define how to perform a Neumann-Ulam solve using MSOD. Given by
-Algorithm~\ref{alg:msod_transport}, we begin by constructing the Monte
-Carlo data structures for each set including the transport domain, the
-source, and the tallies. Next, in line 4 we perform the asynchronous
-transport algorithm presented in \S~\ref{sec:asynchronous_algorithm}
-for each of the individual sets. Adding overlap to
-Algorithm~\ref{alg:parallel_mc_algorithm} in this case simply modifies
-the set of local states and the set of neighboring states to
-additional states gathered during overlap generation.
-
-\begin{algorithm}[h!]
-  \caption{\textbf{MSOD Transport Sequence}}
-  \label{alg:msod_transport}
-  \begin{algorithmic}[1]
-    \State build MSOD domain
-    \State build Monte Carlo source
-    \State build Monte Carlo tally
-    \State perform parallel Neumann-Ulam transport in each set
-    \State combine set tallies
-    \State combine block tallies
-  \end{algorithmic}
-\end{algorithm}
-
-Once Monte Carlo transport is complete complete, the tallies for each
-individual computation must be gathered to the original set in order
-to build the correction vector within an MCSA sequence. To do this, we
-perform two parallel vector operations. First, overlap not only
-creates additional states in the local transport domain but also
-additional states in the local tally vector such that any events that
-occur in the local domain may be tallied in a local vector. Because of
-this, the tally vector in a single set will share local states with
-other domains and a parallel sum reduction operation is required to
-transform the tally into the original Neumann-Ulam parallel
-decomposition. Second, the tallies in individual sets must be combined
-and applied to the primary set in the problem that shares the
-processor space with the original linear problem. For this operation,
-we employ the concept of blocks as a subdivision of parallel space
-complementary to the concept of sets.
-
-Consider the schematic representation of the final parallel reduction
-presented in Figure~\ref{fig:msod_tally_reduction}. For this
-operation, the tally vectors in each set must be summed together and
-applied to the primary set. As each set is an exact replica of the
-others, the parallel decomposition of the tally vector is the same
-within all sets. We can take advantage of this by building a processor
-space that encapsulates each identical domain in each set. We can then
-use this processor space for the reduction. For the schematic in
-Figure~\ref{fig:msod_tally_reduction}, the upper-left domain in each
-set creates a single block, giving 9 total blocks for this example
-problem. The processor space that encapsulates each of the upper-left
-domains is then used for the parallel reduction operation. This
-reduction will happen in a mutually exclusive manner for each of the
-blocks in the system. One benefit of this approach is that this
-reduction occurs only over a subset of the processors in the problem,
-providing better scalability than a reduction operation over all
-processors in the problem.
-
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{0.6}{ \input{chapters/parallel_mc/msod_tally.pdftex_t} }
-  \end{center}
-  \caption{\textbf{MSOD tally reduction over blocks.} \textit{The
-      tally computed independently in each set is reduced across the
-      blocks to the primary set. The linear system vector is updated
-      on the primary set. In this example the tally vector in the
-      block containing upper-left subdomains is combined. Identical
-      reduction operations occur for all other blocks
-      simultaneously.}}
-  \label{fig:msod_tally_reduction}
-\end{figure}
-
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{Parallel MCSA\ }
-\label{sec:parallel_mcsa}
-With the parallel adjoint Neumann-Ulam solver implementation described
-above, the parallel implementation of an MCSA iteration is
-trivial. Recall the MCSA iteration procedure presented again here for
-clarity:
-\begin{subequations}
-  \begin{gather}
-    \label{eq:mcsa_r1}
-    \ve{x}^{k+1/2} = \ve{x}^k + \ve{r}^k\:,\\
-    \label{eq:mcsa_r2}
-    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\
-    \label{eq:mcsa_r3}
-    \ve{A}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:,\\
-    \label{eq:mcsa_r4}
-    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,\\
-    \label{eq:mcsa_r5}
-    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:.
-  \end{gather}
-\end{subequations}
-In \S\ref{sec:parallel_krylov_methods} we discuss parallel matrix and
-vector operations as utilized in conventional projection methods. We
-utilize these here for the parallel MCSA implementation for the
-computations required in Eqs~(\ref{eq:mcsa_r1}), (\ref{eq:mcsa_r2}),
-(\ref{eq:mcsa_r4}), and (\ref{eq:mcsa_r5}). In the first step, a
-parallel matrix-vector multiply is used to apply the split operator to
-the previous iterate's solution. A parallel vector update is then
-performed with the source vector to arrive at the initial iteration
-guess. In the next step, the residual is computed by the same
-operations where now the operator is applied to the solution guess
-with a parallel matrix-vector multiply and then a parallel vector
-update with the source vector is performed. Once the correction is
-computed with a parallel adjoint Neumann-Ulam solve, this correction
-is applied to the guess with a parallel vector update to get the new
-iteration solution. Additionally, as given by
-Eq~(\ref{eq:mcsa_stopping_criteria}), 2 parallel vector reductions
-will be required to check the stopping criteria: one initially to
-compute the infinity norm of the source vector, and another at every
-iteration to compute the infinity norm of the residual vector.
-
-To parallelize the solution of Eq~(\ref{eq:mcsa_r3}), we apply the
-MSOD Neumann-Ulam algorithm presented in the previous section. The
-problem is potentially replicated with overlap and the Monte Carlo
-procedure returns a correction vector, $\delta \ve{x}$, with the same
-parallel decomposition as the input linear problem. As was the case
-for generating the replications required for a multiple set
-implementation, the explicit matrix and vector operations required for
-the rest of an MCSA iteration occur only on the subset of processors
-containing the primary set. Only the Monte Carlo data structures exist
-on the rest of the processors in the problem as replication of the
-other operations does not generate any more information that could be
-used to accelerate the time to solution.
-
-\subsection{Parallel FANM Implementation}
-\label{subsec:parallel_fanm}
-Here we briefly comment on how a FANM method may be implemented in
-parallel. A parallel FANM method relies on a basic set of parallel
-matrix-vector operations outlined in
-\S~\ref{sec:parallel_krylov_methods} as well as the global residual
-and Jacobian assembly procedure described in
-\S~\ref{subsec:automatic_differentiation}. Consider the FANM iteration
-scheme in Algorithm~\ref{alg:fanm}. We must first assemble the linear
-system in parallel through the element-wise function evaluations to
-generate both the global Jacobian operator and the global residual
-vector on the right hand side. Per Bartlett's work on automatic
-differentiation, efficient and automated parallel mechanisms are
-available to do this through a sequence of scatter/gather
-operations. With these tools available for residual and Jacobian
-generation, the remainder of the parallel procedure is simple. The
-linear Newton correction system is solved using the parallel MCSA
-method as described in the previous section and the Newton correction
-applied to the previous iterate's solution through a parallel vector
-update.
-
-%%---------------------------------------------------------------------------%%
-\section{Parallel MCSA Verification\ }
-\label{sec:parallel_verification}
-Before we consider the parallel performance of the MCSA algorithm, we
-must first verify that the algorithm generates the correct solution in
-parallel with respect to reference solutions provided by production
-Krylov methods. To verify correctness, we use solutions to the neutron
-diffusion system presented in
-Appendix~\ref{chap:diffusion_problem}. As the reference computation,
-serial results will be used from the conjugate gradient solver in
-Belos, a linear solvers package in the Trilinos scientific computing
-libraries \cite{heroux_overview_2005}. In identical fashion to the
-verification for the $SP_N$ equations performed in
-\S~\ref{sec:spn_mcsa_verification}, we will solve the neutron
-diffusion problem in parallel using multiple cores on a desktop
-machine. For each parallel solution, then the solution at each grid
-point in the system will be compared to the solution at the same grid
-point in the serial reference case with the minimum and maximum
-absolute differences of these values reported.
-
-GMRES and conjugate gradient solvers from the Belos package were used
-on 4 cores as a means of both assuring the Krylov solvers produce the
-same results on multiple cores and as a means of providing an
-acceptable range of minimum and maximum values in the comparison. For
-MCSA, a range of MSOD parameters will be used such that problems with
-multiple sets and multiple blocks may be tested along with
-overlap. For all MCSA calculations, a single history was used for
-every DOF in the problem to compute the correction vector. For all
-solvers, the flux was converged to a tolerance of \sn{1}{-8} and
-therefore if the comparisons agree within this tolerance the solutions
-will be deemed
-correct. Table~\ref{tab:parallel_verification_parameters} gives the
-parameters for the verification problem and
-Table~\ref{tab:parallel_mcsa_verification} gives the results of the
-calculations. Solver parameters used for MCSA are given in
-Table~\ref{tab:parallel_mcsa_parameters}.
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lc}\hline\hline
-      \multicolumn{1}{l}{Parameter}& 
-      \multicolumn{1}{c}{Value}\\\hline
-      $dx$ & 0.1 \\
-      $dy$ & 0.1 \\
-      $N_x$ & 400 \\
-      $N_y$ & 400 \\
-      $N$ & 160,000 \\
-      $\Sigma_a$ & 5.0 \\
-      $\Sigma_s$ & 1.0 \\
-      Boundary Type & Vacuum \\
-      Linear Solver Tolerance & \sn{1}{-8} \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Parallel MCSA verification model problem parameters.}
-    \textit{The neutron diffusion equation in 2 dimensions is used for
-      the model problem.}}
-  \label{tab:parallel_verification_parameters}
-\end{table}
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lc}\hline\hline
-      \multicolumn{1}{l}{Parameter}& 
-      \multicolumn{1}{c}{Value}\\\hline
-      Histories & 1 per DOF \\
-      Weight Cutoff & \sn{1}{-2} \\
-      Fixed Point Iteration & Richardson \\
-      Estimator & Adjoint Collision \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Parallel MCSA verification solver parameters.}
-    \textit{No relaxation parameters or variance reduction techniques
-      were used with MCSA.}}
-  \label{tab:parallel_mcsa_parameters}
-\end{table}
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lccccll}\hline\hline
-      \multicolumn{1}{l}{Solver}& 
-      \multicolumn{1}{c}{Blocks}& 
-      \multicolumn{1}{c}{Sets}& 
-      \multicolumn{1}{c}{Cores}& 
-      \multicolumn{1}{c}{Overlap}& 
-      \multicolumn{1}{l}{Min Difference}& 
-      \multicolumn{1}{l}{Max Difference}\\\hline
-      Conjugate Gradient & - & - & 4 & - & 0.0 & \sn{1.11022}{-16} \\
-      GMRES & - & - & 4 & - & \sn{5.64215}{-13} & \sn{3.83348}{-9} \\
-      MCSA & 2 & 1 & 2 & 0 & \sn{8.60423}{-16} & \sn{7.6806}{-9} \\
-      MCSA & 3 & 1 & 3 & 0 & \sn{6.41154}{-15} & \sn{7.65397}{-9} \\
-      MCSA & 4 & 1 & 4 & 0 & \sn{2.66454}{-15} & \sn{7.62623}{-9} \\
-      MCSA & 2 & 2 & 4 & 0 & \sn{1.52656}{-15} & \sn{7.57108}{-9} \\
-      MCSA & 1 & 4 & 4 & 0 & \sn{5.55112}{-16} & \sn{7.61631}{-9} \\
-      MCSA & 8 & 1 & 8 & 0 & \sn{6.13398}{-15} & \sn{7.51744}{-9} \\
-      MCSA & 4 & 2 & 8 & 0 & \sn{1.03251}{-14} & \sn{7.62601}{-9} \\
-      MCSA & 2 & 4 & 8 & 0 & \sn{5.13478}{-15} & \sn{7.60777}{-9} \\
-      MCSA & 1 & 8 & 8 & 0 & \sn{1.80411}{-15} & \sn{7.60651}{-9} \\
-      MCSA & 2 & 1 & 2 & 5 & \sn{2.47025}{-15} & \sn{7.58882}{-9} \\
-      MCSA & 3 & 1 & 3 & 5 & \sn{2.44249}{-15} & \sn{7.6326}{-9} \\
-      MCSA & 4 & 1 & 4 & 5 & \sn{7.49401}{-16} & \sn{7.49014}{-9} \\
-      MCSA & 2 & 2 & 4 & 5 & \sn{1.19349}{-15} & \sn{7.61814}{-9} \\
-      MCSA & 1 & 4 & 4 & 5 & \sn{5.55112}{-16} & \sn{7.61631}{-9} \\
-      MCSA & 8 & 1 & 8 & 5 & \sn{2.05391}{-15} & \sn{7.63955}{-9} \\
-      MCSA & 4 & 2 & 8 & 5 & \sn{9.57567}{-15} & \sn{7.61313}{-9} \\
-      MCSA & 2 & 4 & 8 & 5 & \sn{5.7454}{-15} & \sn{7.60767}{-9} \\
-      MCSA & 1 & 8 & 8 & 5 & \sn{1.80411}{-15} & \sn{7.60651}{-9} \\
-      MCSA & 2 & 1 & 2 & 10 & \sn{1.08247}{-15} & \sn{7.63231}{-9} \\
-      MCSA & 3 & 1 & 3 & 10 & \sn{3.27516}{-15} & \sn{7.60211}{-9} \\
-      MCSA & 4 & 1 & 4 & 10 & \sn{4.996}{-16} & \sn{7.64756}{-9} \\
-      MCSA & 2 & 2 & 4 & 10 & \sn{4.13558}{-15} & \sn{7.60259}{-9} \\
-      MCSA & 1 & 4 & 4 & 10 & \sn{5.55112}{-16} & \sn{7.61631}{-9} \\
-      MCSA & 8 & 1 & 8 & 10 & \sn{9.71445}{-16} & \sn{7.58535}{-9} \\
-      MCSA & 4 & 2 & 8 & 10 & \sn{7.77156}{-16} & \sn{7.65978}{-9} \\
-      MCSA & 2 & 4 & 8 & 10 & \sn{4.16334}{-16} & \sn{7.63191}{-9} \\
-      MCSA & 1 & 8 & 8 & 10 & \sn{1.80411}{-15} & \sn{7.60651}{-9} \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Parallel MCSA verification results.}
-    \textit{Difference values given are absolute values. The neutron
-      diffusion equation in 2 dimensions is used for the model problem
-      with the parameters given in
-      Table~\ref{tab:parallel_verification_parameters}. A serial
-      conjugate gradient calculation is used as the reference
-      computation and neutron fluxes are compared
-      element-wise. Parameters used for the MCSA solver are given in
-      Table~\ref{tab:parallel_mcsa_parameters}. All solvers including
-      the reference computation were converged to tolerance of
-      \sn{1}{-8}.}}
-  \label{tab:parallel_mcsa_verification}
-\end{table}
-
-From the data we see that for all solvers, the parallel results agree
-with the reference serial results within the solution tolerance for
-all DOF in the system. In addition, we do not notice any systematic
-differences in the results as a function of number of blocks, number
-of sets, or amount of overlap. In addition, the parallel GMRES results
-have maximum absolute differences of the same magnitude as all MCSA
-results and a larger minimum absolute difference than all MCSA results
-by three orders of magnitude. Based on these results, MCSA
-parallelized with the MSOD algorithm, and in particular the
-implementation used here, generates the correct results with respect
-to the serial conjugate gradient reference computation.
-
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{Parallel Performance Metrics\ }
-\label{sec:parallel_performance_metrics}
+\chapter{Parallel Performance Metrics }
+\label{chap:parallel_metrics}
 
@@
 considered when preparing a parallel calculation.
-
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{Leadership-Class Parallel Scaling Studies\ }
-\label{sec:leadership_scaling_studies}
-Using the parallel performance metrics presented in the previous
-section, we will next measure performance in various scaling studies
-using an implementation of the parallel MCSA algorithm to assess its 
-quality using the neutron diffusion problem presented in
-Appendix~\ref{chap:diffusion_problem}. For these scaling studies,
-large problems at high levels of concurrency will be solved at a
-leadership-class computing facility. In this section, we will outline
-the parallel machine on which the calculations were performed and then
-present a series of numerical studies that identify both strengths and
-weaknesses of the parallel algorithm. Both pure domain decomposition
-and MSOD parallel performance will be explored and compared to
-performance results for the same problems using a set of production
-parallel Krylov solvers.
-
-For the scaling results presented here, all parallel MCSA results were
-computed using the Monte Carlo Linear Solvers\footnote{MCLS is an open
-  source C++ library built on the Trilinos framework and can be
-  downloaded at \url{https://github.com/sslattery/MCLS}} (MCLS)
-library generated specifically for this work while the conjugate
-gradient and GMRES Krylov solvers used are part of the Belos linear
-solver package within the Trilinos framework
-\cite{heroux_overview_2005}. In general, the Belos package produces
-solutions to the neutron diffusion problem nearly two orders of
-magnitude faster than the MCLS library for serial computations using
-the conjugate gradient solver. Therefore, when we directly compare
-these two codes, the MCLS parallel efficiencies will appear
-artificially inflated relative to the Belos efficiencies due to the
-fact that the MCLS ratio of work to communication will be much larger
-than that for the Belos calculations. For every scaling study
-presented, the calculations were performed 3 times with the average
-values reported for wall time and those average values used for the
-efficiency computations. In addition, it should be noted that in
-practice, symmetric positive-definite problems like the neutron
-diffusion problem solved here would never be solved using
-GMRES. Instead, a method such as conjugate gradient would typically be
-used due to the enhanced performance. We use GMRES here simply as a
-means of providing an additional comparison for MCSA performance.
-
-\subsection{Titan }
-\label{subsec:titan}
-All parallel scaling studies presented in this work were performed on
-the Titan Cray XK7 machine at the Oak Ridge Leadership Computing
-Facility located at Oak Ridge National Laboratory. Titan is a
-heterogeneous architecture machine combining both CPU and GPU
-components to achieve a theoretical peak performance of more than 20
-petaflops. The technical specifications for Titan for only the CPU
-portion of the machine are given in Table~\ref{tab:titan_hardware}.
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lc}\hline\hline
-      Processor & 16 core AMD \\
-      Nodes & 18,688 AMD Opterons \\
-      Cores/Node & 16 \\
-      Total Cores & 299,008 \\
-      Memory/Node & 32 GB \\
-      Memory/Core & 2 GB \\
-      Interconnect & Gemini \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Titan Cray XK7 hardware specifications.}
-    \textit{Specifications presented for Titan in this table were
-      gathered from the Oak Ridge Leadership Computing Facility at}
-    \url{http://www.olcf.ornl.gov/computing-resources/titan-cray-xk7/}}
-  \label{tab:titan_hardware}
-\end{table}
-
-For the scaling studies presented in this work, only the CPU
-components of Titan specified in Table~\ref{tab:titan_hardware} were
-utilized. This is due to the fact that the implementations of the
-parallel MCSA algorithm in the MCLS library utilized code written only
-with the Message Passing Interface (MPI) and therefore could not take
-advantage of the GPU accelerators. Even considering just the CPU
-portion of Titan, nearly 300,000 CPUs are available to test the
-parallel performance of the MCSA algorithm at high levels of
-concurrency.
-
-For each parallel performance metric presented in the previous
-section, a reference case is required to compute efficiencies for all
-other cases. For smaller machines, often a serial computation can be
-performed and used as the base case to get a true measure of speed-up
-for a particular problem. For our scaling studies, we will always use
-as a reference case a calculation which uses 16 cores, filling an
-entire node of the machine. We fill an entire node so that we may
-ignore the effects on scalability measures that arise from using only
-a subset of cores in a node. Typically this results in a large drop in
-efficiency from a serial computation to a calculation that fills the
-entire node. Once a node is filled, adding subsequent nodes has less
-of an effect on the scaling, providing a more stable region of
-analysis.
-
-\subsection{Communication Parameters}
-\label{subsec:comm_parameters}
-
-In our description of Brunner and Brantley's algorithm as applied to
-the Neumann-Ulam method, we noted two free parameters that are
-user-defined at runtime. The first is the static data buffer size for
-communicating histories between neighboring domains and the second is
-the frequency by which a domain checks for incoming information from
-neighboring domain, both of which may potentially affect latency and
-load balancing. Changing the static history buffer size parameter
-affects the scalability of the algorithm in two ways. If the buffer
-size is increased, the buffer takes longer to fill and is therefore
-sent less frequently, potentially causing load balancing bottlenecks
-where a neighboring domain is idle and waiting for more work to do but
-is waiting for another domain to complete work. In addition, if the
-buffer size is too large, the network may become saturated with these
-very large data packets, slowing down performance. If the buffer size
-is too small, too many history messages are sent, not providing enough
-data for a neighboring process to work on to make up for the time
-spent communicating. Considering the second parameter, when the
-message check frequency period is increased, more histories are
-processed locally before looking to do work elsewhere. If this value
-becomes too large, incoming data buffers may not be cleared frequently
-enough and cause decreased performance in the domain-to-domain
-communications. If this value is too small, too much time is spent
-looking for work instead of actually performing it.
-
-To assess how these parameters affect the performance of the parallel
-Monte Carlo algorithm on the Titan machine, a parameter study was
-performed using the neutron diffusion problem on 64 cores (4
-nodes). This problem had a reasonably well conditioned spectral radius
-of 0.78 with a global problem size of \sn{1.6}{7} DOFs with one
-stochastic history used per DOF in the MCSA Neumann-Ulam used to
-compute the correction for the acceleration. Each parameter was varied
-between 16 and 4096 and the time to solve the problem recorded. The
-results of this study are presented in
-Figure~\ref{fig:titan_comm_parameters}. Compared to the results of
-Brunner and Brantley for the same type of study, we note qualitatively
-similar results with the general observation that a larger history
-buffer check period is better than a shorter one. The larger the check
-period, the more on-process work is done before checking
-messages. Second, there was not a significant variation observed in
-the history buffer size parameter for a given check frequency.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_comm_parameters.pdf}
-  \end{center}
-  \caption{\textbf{Neumann Ulam parallel algorithm history buffer
-      check period vs. history buffer size sensitivity.} \textit{The
-      color scale represents the ratio of the runtime of a particular
-      parameter variation with the runtime from the faster parameter
-      variation. A value of 1.02 on the color scale corresponds to a
-      runtime 2\% longer than the faster time. All runtimes were
-      within 6.5\% of the fastest observed case.}}
-  \label{fig:titan_comm_parameters}
-\end{figure}
-
-Compared to the reference work, there was also not a significant
-change in the runtime over the values of parameters tested with all
-runtimes reported within 6.5\% of the fastest case observed. For this
-work, a history buffer size of 4096 and a buffer check period of 128
-performed best for the initial calculations. However, most values in
-the range tested will give good performance. Brunner and Brantley
-found that a value of 1024 was best for both parameters. For the
-results observed here, these parameters were only 0.7\% slower than
-the fastest case. It may be best practice to fix these parameters
-internally at 1024 as both this work and the reference found these to
-perform well several years apart on both different parallel machines
-and for different transport sequences. Therefore, for the scaling
-studies presented in the following sections, values of 1024 for both
-the history buffer size and message check frequency will be used.
-
-\clearpage
-
-\subsection{Pure Domain Decomposition Scaling}
-\label{subsec:pure_domain_decomp}
-
-For the first set of scaling studies, we consider the parallel MCSA
-algorithm where MSOD has not been used in the parallel Neumann-Ulam
-algorithm. We refer to this case as pure domain decomposition where no
-overlap or replication has been leveraged. Such a test will isolate
-specifically the performance of Brunner and Brantley's algorithm as
-applied to the Neumann-Ulam algorithm. For the strong scaling study,
-the global diffusion problem size was fixed at \sn{1.6}{7} DOFs for
-all solvers and the 16 core run used as the base case. MCSA was used
-again with one stochastic history used per DOF in the Neumann-Ulam
-solve to compute the correction for the acceleration with the
-remainder of the problems and solver parameters the same as those for
-the verification calculation given in
-Table~\ref{tab:parallel_verification_parameters} and
-Table~\ref{tab:parallel_mcsa_parameters}. As with all of the scaling
-studies, 3 calculations for all data points were performed with the
-average wall time reported and used for the efficiency
-calculations. Figure~\ref{fig:titan_pure_strong_time} gives the wall
-time per iteration for these calculations while
-Figure~\ref{fig:titan_pure_strong} gives the results of the this study
-with the given absolute efficiencies computed using
-Eq~(\ref{eq:strong_scaling_absolute}). We report wall time per
-iteration due to the fact that MCSA converged on the solution in 22
-iterations while the Krylov solvers converged in 19 iterations.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_pure_strong_time.pdf}
-  \end{center}
-  \caption{\textbf{Pure domain decomposition strong scaling wall time
-      per iteration.} \textit{Wall time is reported per iteration
-      because the methods converged in different numbers of
-      iterations. MCLS is over an order of magnitude slower
-      arithmetically than the Krylov solvers. GMRES executes more
-      operations than conjugate gradient and is therefore slower as
-      well.}}
-  \label{fig:titan_pure_strong_time}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_pure_strong.pdf}
-  \end{center}
-  \caption{\textbf{Pure domain decomposition strong scaling absolute
-      efficiency.}  \textit{Super-linear speed-up is from memory
-      thrashing in the base case. MCLS is over an order of magnitude
-      slower arithmetically, causing the higher efficiency. GMRES
-      executes more operations than conjugate gradient, also creating
-      a higher efficiency.}}
-  \label{fig:titan_pure_strong}
-\end{figure}
-
-There are several key features of
-Figure~\ref{fig:titan_pure_strong_time} and
-Figure~\ref{fig:titan_pure_strong} worth noting. First, up to around
-5,000 cores, all methods experience a super-linear improvement in
-parallel efficiencies. For all methods, this increase is not a
-function of the algorithmic efficiency as those computed using
-Eq~(\ref{eq:algorithmic_efficiency}) remained at unity, meaning that
-all computations converged in the same number of iterations relative
-to the base case\footnote{When the algorithmic efficiency remains at
-  unity for all calculations, the implementation efficiency is
-  equivalent to the absolute efficiency and therefore is not presented
-  here.}. Instead, what is happening here and was observed by Brunner
-and Brantley in their work is that the large problem used to scale to
-O(10,000) cores was large enough that memory thrashing was occurring
-in the 16 core base case. As the number of cores in the strong scaling
-study was increased and the local problem size decreased, more of the
-local problem fit into the cache of the system, thus causing improved
-serial performance and therefore enhanced runtimes. In this region of
-super-linear efficiencies, we also observe larger efficiencies for
-GMRES as compared to conjugate gradient. This occurs due to the fact
-that conjugate gradient has a static set of operations performed at
-every iteration while GMRES operates on a continually growing
-subspace. In addition, the orthogonalization operations required to
-generate that subspace consume much more time than the basic conjugate
-gradient operations, causing a larger serial runtime and subsequently
-higher efficiencies. For the same reasons, we also observe higher
-efficiencies for the MCSA implementation due to the serial performance
-of the implementation.
-
-Beyond around 5,000 cores, the global problem size selected for this
-strong scaling study causes the local problem size to shrink enough
-that parallel communication costs begin to overtake arithmetic costs
-on process. This takeover causes this effective strong scaling wall
-that is observed in all methods. Using any more cores to solve this
-particular global problem is not an efficient use of those parallel
-resources. Furthermore, the fact that we see a very similar trends in
-efficiencies for all methods means that they are bound by the same
-parallel performance limitations in a strong scaling environment. This
-is important as we can expect the same qualitative scaling performance
-from MCSA if arithmetic optimization has been performed to improve the
-serial performance of the implementation. In addition, these results
-are valuable in that they show MCSA, when properly parallelized, has
-the ability to be competitive on leadership-class computing platforms
-with conventional subspace methods with good strong scaling observed
-in the pure domain decomposition case to O(10,000) cores.
-
-Next, we consider the weak scaling performance of the algorithm. In
-this case, we fix the local problem size at \sn{4}{4} DOFs and
-increase the number of processors used in the
-simulation. Figure~\ref{fig:titan_pure_weak_time} gives the wall
-time per iteration and Figure~\ref{fig:titan_weak_absolute} gives the
-weak scaling absolute efficiencies for this problem as computed by
-Eq~(\ref{eq:weak_scaling_absolute}). In general, the relative trends
-when comparing MCSA to the Krylov methods are the same with MCSA and
-GMRES scaling better than the conjugate gradient method due to the
-increased amount of arithmetic work on-process. In addition, it is
-important to note that the weak scaling of MCSA when properly
-parallelized remains competitive with the Krylov methods keeping in
-mind the artificial inflation of efficiencies.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_pure_weak_time.pdf}
-  \end{center}
-  \caption{\textbf{Pure domain decomposition weak scaling wall time
-      per iteration.} \textit{Wall time is reported per iteration
-      because the methods converged in different numbers of
-      iterations. MCLS is over an order of magnitude slower
-      arithmetically than the Krylov solvers. GMRES executes more
-      operations than conjugate gradient and is therefore slower as
-      well.}}
-  \label{fig:titan_pure_weak_time}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_absolute.pdf}
-  \end{center}
-  \caption{\textbf{Pure domain decomposition weak scaling absolute
-      efficiency.} \textit{MCLS is an over order of magnitude slower
-      arithmetically, causing the higher efficiency. GMRES executes
-      more operations than conjugate gradient, also creating a higher
-      efficiency.}}
-  \label{fig:titan_weak_absolute}
-\end{figure}
-
-As an additional means of comparison between MCSA and the Krylov
-methods, we compute the algorithmic efficiencies of the calculations
-using Eq~(\ref{eq:algorithmic_efficiency}). As shown in
-Figure~\ref{fig:titan_weak_algorithmic}, the algorithm efficiency of
-the Krylov algorithm improves as a function of the number of cores in
-the problem due to the fact that the number of iterations required to
-converge decreased from 20 in the base case to 19 and then to 18 for
-the final set of weak scaling computations. When this reduction in
-iterations is used to account for the observed runtimes and computed
-absolute efficiencies for weak scaling we end up with the
-implementation efficiencies computed using
-Eq~(\ref{eq:implementation_efficiency}) given in
-Figure~\ref{fig:titan_weak_implementation}. Because MCSA maintains the
-same iterative performance as a function of global problem size in the
-weak scaling study, the implementation efficiencies compare more
-favorably to the Krylov methods. Using the implementation efficiency,
-we are effectively measuring the parallel performance of a single
-iteration for each method by looking at these implementation
-efficiencies. As the global problem size increases, MCSA is less prone
-to a drop in parallel efficiency. If optimization has been performed
-and runtimes potentially approach those of the Krylov methods, this
-comparison may not be as favorable. Overall, weak scaling performance
-of the parallel MCSA algorithm is excellent for the pure domain
-decomposition case with absolute parallel efficiencies of over 80\%
-observed at over 32,000 cores.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_alg_eff.pdf}
-  \end{center}
-  \caption{\textbf{Pure domain decomposition weak scaling algorithmic
-      efficiency.} \textit{As the global problem size was increased in
-      the weak scaling study, both conjugate gradient and GMRES
-      converged in fewer iterations, causing the rise in algorithmic
-      efficiency.}}
-  \label{fig:titan_weak_algorithmic}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_implementation.pdf}
-  \end{center}
-  \caption{\textbf{Pure domain decomposition weak scaling
-      implementation efficiency.} \textit{Considering the scalability
-      of a single iteration of the implementation, the inflated
-      efficiencies of the MCLS MCSA implementation become more
-      obvious. Implementation optimization will likely reduce the MCSA
-      values to a level more comparable to the Belos values.}}
-  \label{fig:titan_weak_implementation}
-\end{figure}
-
-In addition to comparing MCSA scaling to conventional methods, we also
-motivate avoiding any type of master/slave algorithms at these high
-levels of concurrency by comparing the binary communication tree
-implementation used in the previous results with a parallel MCSA
-implementation that uses the master/slave scheme to determine the end
-of a history transport stage as shown in
-Figure~\ref{fig:master_comm_tree}
-instead. Figure~\ref{fig:titan_strong_bvsm} provides this comparison
-using the strong scaling study while Figure~\ref{fig:titan_weak_bvsm}
-gives the results of the comparison for the weak scaling study. In
-both cases, the deficiencies caused by using the master/slave scheme
-for transport completion are obvious. At around 1,000 cores, both the
-strong and weak scaling performance seriously degrades while the
-binary tree pattern maintains excellent performance. At this number of
-cores, any extra work given to the master process, even if it amounts
-to summing a handful of integers and checking for a few extra messages
-as in this algorithm, grows enough to create serious load imbalance
-and reduce performance.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_bvsm.pdf}
-  \end{center}
-  \caption{\textbf{Binary tree vs. master/slave communication scheme
-      strong scaling absolute efficiency.} \textit{The master/slave
-      scheme will not scale beyond $O(1,000)$ cores due to load
-      imbalance.}}
-  \label{fig:titan_strong_bvsm}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_bvsm.pdf}
-  \end{center}
-  \caption{\textbf{Binary tree vs. master/slave communication scheme
-      weak scaling absolute efficiency.}  \textit{The master/slave
-      scheme will not scale beyond $O(1,000)$ cores due to load
-      imbalance.}}
-  \label{fig:titan_weak_bvsm}
-\end{figure}
-
-\clearpage
-
-\subsection{Overlapping Domain Decomposition}
-\label{subsec:overlapping_domain_decomp}
-
-In the next set of scaling studies, we will explore how adding overlap
-to the subdomains in the problem affects runtimes and parallel
-performance as compared to the case with pure domain decomposition. To
-do this, we utilize the set of analytic relations developed in
-\S~\ref{sec:analytic_framework} to design these numerical
-experiments. For the neutron diffusion problem used in these scaling
-studies, the linear operator had a spectral radius of 0.787 and a
-weight cutoff of \sn{1}{-2} was used with the Neumann-Ulam
-solver. Using Eq~(\ref{eq:analytic_k}) gives an expected random walk
-length of 19.22 transitions for any given history in the
-problem. Using the mean-chord approximation given by
-Eq~(\ref{eq:mean_chord_domain_leakage}), we then expect approximately
-5.3\% of the histories born in a particular domain to leak out after
-the first transport stage. By adding overlap, we aim to reduce this
-fraction of communicated histories and therefore boost the parallel
-performance of the algorithm.
-
-To determine the amount of overlap to test for the scaling studies, we
-use a variation of Eq~(\ref{eq:step_k_length}) to determine how many
-discrete states a history will move on average from its birth site. We
-modify Eq~(\ref{eq:step_k_length}) to only account for discrete
-movements:
-\begin{equation}
-  \sqrt{\bar{n^2_k}} = n_s \sqrt{k}\:,
-  \label{eq:discrete_distance}
-\end{equation}
-where $\sqrt{\bar{n^2_k}}$ is now the root-mean-squared number of
-discrete states a history will move from its birth site. For our
-diffusion problem used in these scaling studies, we compute a value of
-2.63 states per history for any of the dimensions of the system. This
-means that we can expect to end up on average 2.63 states in both the
-$i$ and $j$ dimensions away from the starting state. Therefore, we
-will use the overlap generation algorithm outlined in
-\S~\ref{subsec:msod_generation} such that we parametrically grow the
-local domain using 1, 2, 3, and 5 levels to measure how scalability is
-affected when this root-mean-squared number of states is
-considered. For example, using 3 levels of overlap should keep over
-50\% of the histories that would be communicated in the pure domain
-decomposition case on-process as this is greater than the
-root-mean-squared number of states we would expect a history to
-travel.
-
-For the performance metrics, we may use the same weak and strong
-scaling efficiency metrics that we used for the pure domain
-decomposition case. As we are comparing to the case without overlap
-and are looking to boost performance relative to the case without
-overlap, for both weak and strong scaling calculations we will use the
-16-core case without overlap as the reference computation for all
-efficiency measures reported.
-
-The absolute efficiency results from the strong scaling exercise are
-given in Figure~\ref{fig:titan_strong_overlap}. Aside from the
-boosting of efficiencies at 2,048 cores, little evidence of strong
-scaling improvement is evident for all levels of overlap tested. To
-further explore any potential benefits of adding overlap into the
-system, Figure~\ref{fig:titan_strong_overlap_diff} gives the
-difference in the efficiencies between all overlap cases and the
-reference case without overlap at each core count. At smaller numbers
-of cores, there is no observed benefit to using overlap to improve
-strong scaling. The improvements noted at 2,048 cores are also present
-here and account for the best efficiency improvements for all core
-counts tested. at 16,384 cores, using 2 levels of overlap did give a
-6.4\% boost in efficiency while at 1,024 cores this same amount of
-overlap actually reduced the efficiency. Beyond 2 levels of overlap,
-efficiency levels begin to fall with 5 levels of overlap performing
-the worst for all cases. From this, we don't expect adding any more
-overlap will improve the scalability of the algorithm and the best
-strong scaling performance was observed when the number of overlap
-levels is approximately equal to the root-mean-squared number of
-states we expect a history to move from its starting point as computed
-by Eq~(\ref{eq:discrete_distance}). We will defer an explanation of
-the drop in efficiencies for levels of overlap beyond this value until
-we have analyzed the weak scaling data.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_overlap.pdf}
-  \end{center}
-  \caption{\textbf{Strong scaling absolute efficiency with varying
-      levels of overlap.} \textit{Values of 0, 1, 2, 3, and 5 levels
-      of overlap were used as the problem had a root-mean-squared
-      number of states of 2.63 for each history.}}
-  \label{fig:titan_strong_overlap}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_overlap_diff.pdf}
-  \end{center}
-  \caption{\textbf{Strong scaling absolute efficiency difference with
-      respect to the 0 overlap base case.} \textit{Values of 1, 2, 3,
-      and 5 levels of overlap are compared here with the absolute
-      difference of the absolute efficiencies presented.}}
-  \label{fig:titan_strong_overlap_diff}
-\end{figure}
-
-For weak scaling, the same study as in the pure domain decomposition
-case was performed again with 1, 2, 3, and 5 levels of overlap. The
-absolute parallel efficiencies from these calculations are given in
-Figure~\ref{fig:titan_weak_overlap} and the differences in
-efficiencies from the calculations without overlap. The results here
-are much the same in that there is not a dramatic improvement in
-scalability relative to the case without overlap. Opposite the strong
-scaling calculation, overlap enhances weak scaling efficiencies at
-lower core counts and degrades the efficiencies at higher core
-counts. In addition, 1 level of overlap provided the best enhancement
-of scalability when overlap was beneficial.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_overlap.pdf}
-  \end{center}
-  \caption{\textbf{Weak scaling absolute efficiency with varying
-      levels of overlap.} \textit{Values of 0, 1, 2, 3, and 5 levels
-      of overlap were used as the problem had a root-mean-squared
-      number of states of 2.63 for each history.}}
-  \label{fig:titan_weak_overlap}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_overlap_diff.pdf}
-  \end{center}
-  \caption{\textbf{Weak scaling absolute efficiency difference with
-      respect to the 0 overlap base case.} \textit{Values of 1, 2, 3,
-      and 5 levels of overlap are compared here with the absolute
-      difference of the absolute efficiencies presented.}}
-  \label{fig:titan_weak_overlap_diff}
-\end{figure}
-
-For both the strong and weak scaling studies, the parallel performance
-of the algorithm was either marginally improved or not improved at
-all. We would always expect an improvement in performance for using
-overlap to prevent the transport of histories from domain to domain
-during execution of the Neumann-Ulam algorithm. The reduction in
-transport means a reduction in parallel communication and therefore
-should never degrade the performance of the algorithm. It is true that
-doing so in fact enhances Monte Carlo performance. However, the
-Neumann-Ulam algorithm is not representative of the entire MCSA
-algorithm. When overlap is used, an overlapping tally vector is
-constructed in parallel such that multiple cores may own multiple
-states in the system. When this occurs, the tally data associated with
-these shared states must be communicated after the Neumann-Ulam solve
-as outlined in \S~\ref{subsec:msod_algorithm}. This reduction
-operation then builds the full MCSA correction vector in the
-decomposition of the solution vector.
-
-Therefore, by building overlap we are in fact reducing parallel
-communication during the Neumann-Ulam solve but we are simply
-deferring that communication until after Monte Carlo transport when
-the overlapping tally vector must be reduced. When overlap becomes
-large enough (i.e. larger than the root-mean-squared number of
-discrete states a history will travel), then the savings in
-communication during the Neumann-Ulam solve start to become smaller
-than the communication created in the overlapping tally vector
-reduction. We see this in both the strong and weak scaling cases at
-larger levels of overlap. For the strong scaling case, we see better
-performance improvement at higher core counts because the total number
-of overlapping states on each core is smaller from the local problem
-size shrinking. For weak scaling, we see better performance
-improvement at lower numbers of cores because although the local
-overlap size is fixed, the global amount of overlap is growing with
-core count and therefore reducing the performance of the tally
-reduction operation.
-
-Based on this data, a few levels of overlap within the expected
-distance of travel for a history may offer some marginal improvement
-in scalability, however, it is not as effective as one might think
-given its improvement to reactor physics calculations as observed in
-the literature. This is largely due to the density of states in the
-system that creates the described performance bottlenecks in the
-overlapping tally vector reduction operation. In a Monte Carlo
-particle simulation, these overlapping tally reduction operations are
-not expected to be performed at the speed they are required in
-MCSA. For example, in a criticality calculation with overlapping
-domains, we would expect to perform as many parallel tally vector
-reduction in a particle transport simulation as there are particle
-stages required for convergence of the eigenvalue. However, these
-particle stages will likely be much more costly in terms of wall time
-relative to a Neumann-Ulam solve in an MCSA iteration due to the more
-extensive particle physics and logic structure that must be
-evaluated. Therefore, we can expect this overlapping tally reduction
-to be less significant to scalability in those operations. We do not
-have as complicated a logic structure in a history random walk for the
-Neumann-Ulam method and therefore overlap can inhibit scalability if
-too large.
-
-\clearpage
-
-\subsection{Multiple Set Domain Decomposition}
-\label{subsec:ms_decomposition}
-
-We next look to apply replication to the problem in the form of
-multiple sets. As the previous analysis showed overlap to be of little
-assistance in improving scalability, we only apply multiple sets to
-the problem with pure domain decomposition within the sets for these
-scaling studies. Before performing parallel scaling studies using
-multiple sets, several aspects of the replicated parallel problem must
-be considered in order to effectively craft the studies and interpret
-their results. First is the question of how many stochastic histories
-to use to generate the correction vector in the MCSA iteration. For
-this work we consider two choices; the split case and the fully
-replicated case. As an example of these cases, consider a single set
-problem where 20,000 histories are used in the Monte Carlo solve. If
-we choose the split case, then solving the same problem with 2 sets
-would give 10,000 histories in each set and 4 sets would give 5,000
-histories in each set. Splitting the histories in this way provides
-the same global number of histories used to compute the correction in
-the single set problem. We expect to reduce the time to solution by
-doing this as each individual set solve will occur faster in an almost
-linear fashion with the reduced number of histories. For the
-replicated history case, each additional set in the problem will use
-just as many histories as the single set case. Using 2 sets with fully
-replicated histories for our example problem then has each set
-computing 20,000 histories with 40,000 global histories in the
-computation. For the 4 set case, the number of global histories is
-increased to 80,000. We also expect this type of replication to reduce
-the time to solution as the extra global histories should reduce the
-stochastic uncertainty in the correction vector and reduce the overall
-number of MCSA iterations required to converge.
-
-\subsubsection{Strong Scaling with Multiple Sets}
-\label{subsubsec:ms_strong}
-
-For the strong scaling study, we apply the additional sets as
-follows. Using the same problems as the pure domain decomposition
-scaling study, the multiple sets are used to replicate those problems
-and boost the core count. For example, for the 4 set case, the data
-point at 64 cores has the same $N/P$ ratio in one set as the single
-set problem with 16 cores. This is due to the fact that $N$ is fixed
-for the strong scaling study and 4 replications gives a set size of 16
-cores. For the 2 set case, the 32 core data point is solving the same
-$N/P$ problem in a set as the 16 core base case.
-
-In addition to considering the replicated domain, we must also
-consider how to modify the efficiencies we will compute. For the
-strong scaling case, recall that the objective is to decrease the time
-to solution for a given global problem size $N$ by increasing
-$P$. Therefore, we may use Eqs~(\ref{eq:strong_scaling_absolute_ref}),
-(\ref{eq:algorithmic_efficiency}), and
-(\ref{eq:implementation_efficiency}) to compute the various strong
-scaling efficiencies, but we must modify the reference computation in
-each of these formulas to reflect our objective. For all efficiency
-computations, the 16 core single set computation will be used as the
-reference case. By adding more cores to the problem, either by
-replicating with multiple sets, or decreasing the local problem size,
-any computational resources added to the problem should aim to result
-in speedup of this reference case. If this speedup for one technique
-of adding cores is less than another, then that technique is deemed a
-less efficient use of those computational resources.
-
-For the multiple sets problem, the single set strong scaling study
-from the pure domain decomposition base case was repeated for multiple
-sets with both splitting of histories across sets and full replication
-of histories across sets using both 2 and 4
-sets. Figure~\ref{fig:titan_strong_ms_time} gives the actual wall
-times for these computations. There are two important features to note
-in the wall time data. First, splitting the histories among sets
-instead of replicating them is faster due to the fact that cutting the
-number of histories in half reduces the time of an iteration by a
-factor that is nearly linear while increasing linearly the number of
-histories through replication does not decrease the iterations
-required to converge by a linear factor. Second, until the strong
-scaling wall is hit starting at around 10,000 cores, the single set
-case actually performs the fastest with only marginal improvement in
-runtimes at higher numbers of cores with more sets. At these larger
-numbers of cores, however, the improvements are so modest that one is
-better off running the same problem using a single set at lower core
-counts to conserve computational resources while maintaining a good
-time to solution.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_ms_time.pdf}
-  \end{center}
-  \caption{\textbf{Wall time in seconds to solution for each case for
-      the strong scaling study with multiple sets.} \textit{Until the
-      strong scaling wall is reached at O(10,000) cores, the single
-      set case performs the best. For the multiple set cases, it is
-      faster to split histories among sets and maintain iterative
-      performance rather than replicate histories to improve iterative
-      performance.}}
-  \label{fig:titan_strong_ms_time}
-\end{figure}
-
-From this timing data we then compute the absolute efficiencies for
-strong scaling as given in Figure~\ref{fig:titan_strong_ms_eff} using
-the 16-core single set calculation as the reference case. As observed
-in the timing data, because the single set case had the fastest run
-times up to the strong scaling wall it also has the best observed
-efficiencies. For the 2 set case with history splitting, although the
-time to perform an iteration was nearly cut in half, the extra
-parallel overhead associated with reducing the tally across the blocks
-to combine the set results increases run times and reduces
-efficiencies. This reduction across blocks is even more obvious for
-the 4 set case. For the replicated cases, we do observe a flat region
-of parallel efficiencies at larger core counts (up to around 32,000
-for the 4 set case with history replication) showing improved scaling
-within context of not considering the single set
-calculation. Unfortunately, the additional cores do not reduce the
-runtime of the global problem as efficiently as expected. Ultimately,
-all cases effectively hit the strong scaling wall at the same time at
-around O(10,000) cores.
-
-We expect this if we consider the fact that adding sets to the problem
-is in itself effectively a form of strong scaling. For any given
-problem of fixed size $N$, adding extra sets to the computation
-decreases the global ratio of $N/P$ as $P$ will grow with the addition
-of sets (although $N/P$ is constant within a set). For the split
-history case, the splitting further exacerbates this situation by
-further decreasing the amount of histories that will be computed in
-the local domain, giving a larger ratio of communication to work in
-the Monte Carlo solve. As with with single set results, we also
-observe some efficiencies greater than unity for the multiple set
-cases also due to improved cache utilization.
-
-Although not as dramatic as one might originally expect from
-introducing replication, there are some modest gains in efficiencies
-at high core counts by using multiple sets. Of interest here is where
-the multiple set cases perform better than the single set case after
-the single set case has hit the strong scaling wall. In
-Figure~\ref{fig:titan_strong_ms_eff}, at 16,384 cores the single set
-case operates at 38\% efficiency while at the same number of cores,
-the 2 set case with history replication operates at 58\%
-efficiency. This is reasonable improvement in the use of those
-computational resources, gained by replicating the problem. Beyond
-16,384 cores, there is not a compelling reason to use more
-computational resources for the this problem.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_ms_eff.pdf}
-  \end{center}
-  \caption{\textbf{Multiple set absolute parallel efficiency.}
-    \textit{Computed relative to the 16-core 1-set base case. Using a
-      single set is the best means by which to improve strong scaling
-      up to the strong scaling wall. At higher core counts past where
-      the single set case hits the strong scaling wall adding sets can
-      provide some improvements in efficiencies.}}
-  \label{fig:titan_strong_ms_eff}
-\end{figure}
-
-Also of interest here is the effect of splitting histories among sets
-versus replicating them on the iterative performance of the
-method. Table~\ref{tab:ms_strong_alg_eff} gives the iterations
-required to converge and computed algorithmic efficiencies for each of
-the cases presented in the strong scaling data. As expected, splitting
-histories maintains a fixed global number of histories and therefore
-maintained the number of iterations required to converge relative to
-the single set base case. When histories were replicated, adding sets
-decreased the number of iterations required to converge in a nonlinear
-fashion. Therefore, increasing the number of histories per MCSA
-iteration improves algorithmic efficiency. Using these algorithmic
-efficiencies, the implementation efficiencies for the multiple set
-cases were computed and are given in
-Figure~\ref{fig:titan_strong_ms_impeff}. From these results, we see
-that the replicated history cases are even less efficient than in the
-absolute case with respect to the performance of a single
-iteration. Not only do we have the additional parallel overhead of
-doing the block level reduction, but we also have additional histories
-that increase the time to solution for a single iteration.
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lcc}\hline\hline
-      \multicolumn{1}{l}{Case}& 
-      \multicolumn{1}{c}{Iterations}&
-      \multicolumn{1}{c}{$\eta_{alg}$} \\\hline
-      1 & 22 & 1.0 \\
-      2-split & 22 & 1.0 \\
-      2-replicated & 16 & 1.375 \\
-      4-split & 22 & 1.0 \\
-      4-replicated & 13 & 1.692 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Strong scaling algorithmic efficiency using
-      multiple sets.} \textit{Replicating histories improves
-      algorithmic efficiency by converging in fewer iterations.}}
-  \label{tab:ms_strong_alg_eff}
-\end{table}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=5in]{chapters/parallel_mc/titan_strong_ms_impeff.pdf}
-  \end{center}
-  \caption{\textbf{Multiple set strong scaling implementation
-      efficiency.} \textit{Computed relative to the 16-core 1-set base
-      case. Considering iterations required to converge, the
-      performance of the split history cases at much better than the
-      cases with history replication.}}
-  \label{fig:titan_strong_ms_impeff}
-\end{figure}
-
-\clearpage
-
-\subsubsection{Weak Scaling with Multiple Sets}
-\label{subsubsec:ms_weak}
-
-Next, we continue to a weak scaling study of the multiple sets
-problem. In the same way as the strong scaling study, we will modify
-the pure domain decomposition study to account for the additional
-replication. For each calculation in the single set study, that
-problem will be replicated either 2 or 4 times with both split and
-replicated histories to assess the effects on scalability and
-runtime. Figure~\ref{fig:titan_weak_ms_time} gives the wall times from
-the multiple sets weak scaling study. Of primary importance here is
-the fact that adding sets to the problem with both split and
-replicated histories actually decreases the time to solution. This was
-not observed in the strong scaling case until much larger core counts
-as the multiple set computations hit the strong scaling wall after the
-single set computation.
-
-This result is significant in that it shows a solution technique in
-which the linear problem can actually be replicated and enhance time
-to solution, something that cannot be achieved with conventional
-methods. At larger core counts, however, this reduction in compute
-time is not as large as at lower core counts due to the fact that the
-tally reduction across blocks requires more resources and those
-resources are clearly growing with core count (most obviously in the 4
-set case with split histories). In addition, at 131,072 cores for the
-largest 4 set computation, load balance is likely becoming an issue
-due to the linear system and subsequent parallel matrix and vector
-operations only occurring on a subset of the entire processor space of
-the problem while the Monte Carlo calculation is occurring on all
-processors.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_ms_time.pdf}
-  \end{center}
-  \caption{\textbf{Wall time in seconds to solution for each case for
-      the weak scaling study with multiple sets.}}
-  \label{fig:titan_weak_ms_time}
-\end{figure}
-
-For a multiple set analysis of the weak scaling timing results, we
-again have to consider that adding sets to the problem is actually a
-form of strong scaling. Consider a single set problem ($S=1$) for a
-given $P$ and $N$. When that problem is now solved using 2 sets we
-have $S=2$, $N$ and $2P$. For any number of sets we then have a global
-problem size of $N$ while the number of processors is $SP$ where $P$
-is the number of processors in the single set case. Therefore, $N/P$
-is no longer fixed in the weak scaling study but rather $N/(SP)$ is
-fixed. As a result, we modify the weak scaling absolute efficiency
-relationship to account for this fact:
-\begin{equation}
-\eta_{weak}(M|N,P|Q,S) = \frac{T(M,Q)}{S \times T(N,P)}\:,
-  \label{eq:ms_weak_efficiency}
-\end{equation}
-where now the absolute efficiency is a function of the number of sets
-in the problem and again $M/Q = N/P$. As with the strong scaling
-study, we will use the 16-core single set case as the reference
-computation for efficiencies. Interestingly,
-Eq~(\ref{eq:ms_weak_efficiency}) has a very similar form to the strong
-scaling efficiency given by Eq~(\ref{eq:strong_scaling_absolute}),
-representative of the strong scaling nature of adding additional
-sets. For problems where adding sets improves time to solution, this
-new weak scaling formulation accounts for how efficiently those extra
-resources are incorporated into the problem. Without such a
-modification, the largest case run for the split 4 set case would have
-a computed weak scaling efficiency of nearly 100\% using the standard
-formula for absolute efficiency. However, using 4 times as many cores
-as the single set case increased the runtime for a fixed global
-problem size and therefore we would actually expect to compute a very
-low efficiency that shows this very poor use of resources.
-
-Figure~\ref{fig:titan_weak_ms_eff} gives the results of the weak
-scaling efficiency computations using the multiple set metric given by
-Eq~(\ref{eq:ms_weak_efficiency}). Although we do reduce the time to
-solution over the base cases for nearly all data points, the
-observation here is that the reduction in time to solution is still
-not large enough to efficiently use all resources. For example, for
-the 2 set case to be a perfect use of resources with either split or
-replicated histories, the runtime of that calculation must be half of
-that for the single set case. For the 4 set case, this run time should
-be a quarter of the single set case for perfect scaling. This was not
-the observation for the runtimes in
-Figure~\ref{fig:titan_weak_ms_time} and therefore the single set case
-is still the most efficient use of computational resources in this
-scaling study.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_ms_eff.pdf}
-  \end{center}
-  \caption{\textbf{Weak scaling absolute efficiency relative to
-      16-core 1-set base case.}}
-  \label{fig:titan_weak_ms_eff}
-\end{figure}
-
-As in the strong scaling case, splitting and replicating histories
-across sets had effectively the same results for the weak scaling
-study with the iterations to converge and computed algorithmic
-efficiencies given by Table~\ref{tab:ms_weak_alg_eff}. Again, when the
-implementation efficiencies are computed, we see in
-Figure~\ref{fig:titan_weak_ms_impeff} that splitting histories among
-sets is still a more efficient manner of introducing multiple sets
-rather than replicating histories in an attempt to reduce the number
-of iterations required to converge.  Here we also note that this is
-not a true weak scaling study as indicated by
-Eq~(\ref{eq:ms_weak_efficiency}) but rather an odd combination of both
-weak and strong scaling. Regardless of this, although multiple sets
-are not the most efficient use of resources in this analysis, they do
-enhance the time to solution through replication and do so
-significantly at core counts of O(1,000) and lower. In both the weak
-and strong scaling cases, due to the additional computational and
-communication overhead of adding sets to the system, one should never
-expect to seriously boost parallel efficiencies for MCSA using this
-technique except for certain instances past the strong scaling wall in
-a purely strong scaling environment. In a weak scaling environment,
-although no efficiency increase will be noted, one can expect an
-enhancement of time to convergence by replicating.
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lcc}\hline\hline
-      \multicolumn{1}{l}{Case}& 
-      \multicolumn{1}{c}{Iterations}&
-      \multicolumn{1}{c}{$\eta_{alg}$} \\\hline
-      1 & 22 & 1.0 \\
-      2-split & 22 & 1.0 \\
-      2-replicated & 16 & 1.375 \\
-      4-split & 22 & 1.0 \\
-      4-replicated & 13 & 1.692 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Weak scaling algorithmic efficiency using multiple
-      sets.} \textit{Replicating histories improves algorithmic
-      efficiency by converging in fewer iterations.}}
-  \label{tab:ms_weak_alg_eff}
-\end{table}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_ms_impeff.pdf}
-  \end{center}
-  \caption{\textbf{Implementation parallel efficiency relative to 16-core
-      1-set base case.}}
-  \label{fig:titan_weak_ms_impeff}
-\end{figure}
-
-\clearpage
-
-\subsection{Subdomain Neumann-Ulam}
-\label{subsec:full_clip}
-
-Based on the analysis of the pure domain decomposed MCSA algorithm and
-subsequently the multiple sets and overlapping domain implementation,
-one may ask then ask if there is any need to communicate histories
-from domain-to-domain at all. Overlap was not as successful because
-because the additional communication required to reduce the
-overlapping tally vector eventually becomes larger than the
-communication savings during Monte Carlo transport. In addition, for
-local domains that are large enough to get good performance
-(e.g. those used in the weak scaling studies), then our analytic
-relationships for domain leakage, and in particular
-Eq~(\ref{eq:mean_chord_domain_leakage}), tell us that only about 5\%
-of the histories in this diffusion problem will leave the domain in
-the first place.  If that is true, then perhaps without any overlap or
-communication, just doing local Monte Carlo transport such that
-histories are terminated upon reaching the domain boundary may be
-satisfactory for accelerating the MCSA iteration. In this case,
-histories near the boundaries of the domain may be truncated before
-reaching the weight cutoff if they reach the domain boundary
-first. This truncation has the potential to increase the stochastic
-uncertainty of the Neumann-Ulam result and therefore increase the
-number of MCSA iterations required to converge. In addition, the
-scalability of the method will then be purely limited by the parallel
-matrix and vector operations required to implement the remainder of
-the algorithm on the subset of processors containing the linear
-problem.
-
-Considering using MCSA in this way such that the Monte Carlo solve
-occurs only in the subdomains casts the method as more of a true
-domain decomposition as outlined by Chapter 14 of Saad's text on
-iterative methods \cite{saad_iterative_2003}. Compared to traditional
-domain decomposition methods we see many similarities including the
-potential to overlap the subdomains to improve the results. By
-building the boundary data as outlined in
-\S~\ref{subsec:domain_generation}, we are effectively gathering one
-level of the adjacent domains, albeit in this case to satisfy the data
-requirements of the Monte Carlo estimators. Furthermore, the fact that
-we are using residual Monte Carlo to compute the correction vector
-casts MCSA as a stochastic realization of an additive Schwarz method
-with a fixed point iteration used as a relaxation step\footnote{See
-  section 14.3.3 in \cite{saad_iterative_2003} for a full definition
-  of the additive Schwarz method.}. Both the fixed point iteration and
-the computation of the residual propagate the solution from individual
-subdomains to the others through the matrix-vector multiply operation
-after all subdomain Monte Carlo solves have been completed. Perhaps an
-even more important consequence of viewing MCSA as a stochastic
-additive Schwarz procedure is that it has the potential to more
-readily fit as an accelerator into other linear solver techniques
-including GMRES or potentially as a preconditioner.
-
-Using the Neumann-Ulam solver to compute the MCSA correction only in
-the subdomains, the strong and weak scaling studies were repeated to
-assess the performance gains made by eliminating communication in the
-Monte Carlo solver. For each scaling study, the 16-core base case with
-Monte Carlo communication was used as the reference point for
-efficiency computations in order to measure how eliminating
-communication boosts
-efficiencies. Figure~\ref{fig:titan_strong_subdomain} gives the
-results of the strong scaling study and
-Figure~\ref{fig:titan_weak_subdomain} gives the results of the the
-weak scaling study in terms of absolute efficiencies. In both cases,
-the benefits to parallel performance of using MCSA as a subdomain
-solver are obvious. In the strong scaling case we make the important
-observation that we hit the strong scaling wall at the same time with
-and without the additional Monte Carlo communication. Although the
-Monte Carlo communication did contribute to the efficiency decrease,
-it is clear that the remaining matrix and vector operations are the
-primary impediment to scalability. Furthermore, we see an even greater
-rise above unity in efficiencies because there is less communication
-overhead hiding the better cache utilization as the local problem size
-shrinks. At 16,384 cores, performing the subdomain solves boost the
-absolute efficiency from 38\% to 67\% relative to the reference case.
-
-In the weak scaling study, we do not have the scaling wall to contend
-with and eliminating Monte Carlo communication through subdomain
-solves provides an effectively linear boost in the weak scaling
-absolute efficiency. We still see a small reduction in weak scaling
-efficiency as a function of core count due to the communication
-overhead for the parallel matrix and vector operations. Here we
-observe super unitary efficiencies for the subdomain case as there was
-a substantial improvement in runtimes relative to the 16-core base
-case with Monte Carlo communication.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_subdomain.pdf}
-  \end{center}
-  \caption{\textbf{Strong scaling absolute efficiency for pure domain
-      decomposition.} \textit{The subdomain Monte Carlo solver
-      improves the parallel efficiency and therefore time to solution
-      by eliminating all parallel communication during the Monte Carlo
-      solve.}}
-  \label{fig:titan_strong_subdomain}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_subdomain.pdf}
-  \end{center}
-  \caption{\textbf{Weak scaling absolute efficiency for pure domain
-      decomposition.} \textit{The subdomain Monte Carlo solver
-      improves the parallel efficiency and therefore time to solution
-      by eliminating all parallel communication during the Monte Carlo
-      solve.}}
-  \label{fig:titan_weak_subdomain}
-\end{figure}
-
-Of additional concern for the subdomain method is the potential
-decrease in iterative performance due to the fact that histories near
-the boundary will potentially have their random walk truncated, thus
-decreasing the information available to compute the correction
-vector. For both the weak and strong scaling study, all calculations
-with split histories converged in 22 or in only a few cases 21
-iterations for both the subdomain solve and the solve with
-communication giving an algorithmic efficiency of 1.048 for all
-cases. This small improvement in iteration count was not observed to
-be a function of problem size, number of processors, or whether the
-subdomain method was used. For the replicated history cases, the
-algorithmic efficiency improvement was not as good as when
-communication was allowed for the Monte Carlo algorithm. For these
-calculations, the number of iterations to converge was reduced to 18
-or 19 in both the strong and weak scaling studies with lower core
-counts converging in fewer iterations.
-
-Therefore, there is no reason to expect overlap to improve the
-iterative or scaling performance of the method as iterative
-performance is maintained without it for this particular
-problem. Overlap is not as effective because histories relinquish most
-of their useful information within the first few steps (which is why
-we can use such a large weight cutoff and there is an insensitivity to
-weight cutoff as shown in
-Figure~\ref{fig:estimator_wc_iters}). Keeping them around longer with
-overlap to do extra work in the local domain will not improve
-iterative performance and thus increase compute times.
-
-As a final analysis of MCSA using subdomain Monte Carlo, the multiple
-sets scaling studies are repeated to determine if any performance
-benefits are made from replication by eliminating Monte Carlo
-communication. Figure~\ref{fig:titan_strong_subdomain_ms_time} gives
-the runtimes for the strong scaling calculation and
-Figure~\ref{fig:titan_strong_subdomain_ms} the computed absolute
-efficiencies. For the weak scaling studies,
-Figure~\ref{fig:titan_weak_subdomain_ms_time} gives the wall times and
-Figure~\ref{fig:titan_weak_subdomain_ms} the efficiencies. For both
-scaling studies, the results are effectively the same as for the case
-with Monte Carlo communication. This is again due to the fact that
-adding sets to the problem is not only an exercise in strong scaling
-but it also adds additional parallel overhead in the form of the tally
-vector reduction across blocks.
-
-From a parallel performance perspective, MCSA may then be best used in
-a mode where the Monte Carlo is performed over only the subdomains
-with no replication or overlap added to the system. In this mode,
-iterative performance is maintained relative to the solutions with
-communication even for more ill-conditioned problems with larger
-spectral radii. Furthermore, no overlap was required to maintain
-iterative performance be letting histories near the boundary take a
-few more steps in their random walks. If in the future there are
-ill-conditioned problems where using subdomain Monte Carlo reduces the
-algorithmic efficiency of MCSA due to the information lost in the
-correction vector, overlap would be a viable mechanism to potentially
-maintain that iterative performance, keeping in mind the potential
-impacts on scalability that were observed in this work. However,
-although eliminating all Monte Carlo communication was observed to be
-the best mode of use for MCSA, the parallel Monte Carlo methods
-developed for this work still provide value for cases where only a
-Monte Carlo solution is desired. In those cases, full transport of the
-Monte Carlo histories are required in order to generate the correct
-solution.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_subdomain_ms_time.pdf}
-  \end{center}
-  \caption{\textbf{Strong scaling total wall time with multiple sets
-      and subdomain Monte Carlo.} \textit{Adding multiple sets has in
-      general a negative effect on the wall time as was observed for
-      the case with full Monte Carlo transport.}}
-  \label{fig:titan_strong_subdomain_ms_time}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_subdomain_ms.pdf}
-  \end{center}
-  \caption{\textbf{Strong scaling absolute efficiency for with
-      multiple sets and subdomain Monte Carlo.} \textit{Increased wall
-      times from adding parallel reduction operations for multiple
-      sets reduces parallel scalability at lower core counts but gives
-      a higher efficiency at the strong scaling wall.}}
-  \label{fig:titan_strong_subdomain_ms}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_subdomain_ms_time.pdf}
-  \end{center}
-  \caption{\textbf{Weak scaling total wall time with multiple sets and
-      subdomain Monte Carlo.} \textit{The addition of the set tally
-      reduction is significant with cost of the reduction observed as
-      core count grows.}}
-  \label{fig:titan_weak_subdomain_ms_time}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_subdomain_ms.pdf}
-  \end{center}
-  \caption{\textbf{Weak scaling absolute efficiency for with multiple
-      sets and subdomain Monte Carlo.} \textit{Weak scaling
-      efficiencies are not improved at large core counts when multiple
-      sets are added.}}
-  \label{fig:titan_weak_subdomain_ms}
-\end{figure}
-
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{Summary\ }
-\label{sec:parallel_summary}
-
-In this chapter we have developed a domain-decomposed parallel
-algorithm for Monte Carlo Synthetic Acceleration and explored the
-qualities of the algorithm through parallel scaling studies. The
-following are the significant observations and findings.
-
-\begin{itemize}
-\item MCSA can be effectively parallelized using domain decomposition
-\item The leakage fraction of stochastic histories from a parallel
-  subdomain can be quantified analytically using algebraic quantities
-\item The multiple-set overlapping-domain (MSOD) parallel algorithm
-  for particle transport has been adapted to parallelize MCSA
-\item MCSA parallelized with MSOD was verified to produce the same
-  results for solutions to the neutron diffusion equation as
-  production Krylov methods
-\item Strong scaling studies were performed with up to 65,536 cores
-  and weak scaling studies were performed with up to 131,072 cores on
-  the Titan Cray XK7 leadership class machine at Oak Ridge National
-  Laboratory
-\item MCSA scales favorably compared to production Krylov methods for
-  both strong and weak scaling cases
-\item Overlap in small quantities determined analytically by random
-  walk length can provide parallel efficiency boosts of a few percent
-  in strong scaling cases but is ineffective in weak scaling cases
-\item Multiple sets will never boost parallel efficiencies in the weak
-  scaling case due to additional communication overhead from added
-  reduction operations but will boost efficiencies beyond the strong
-  scaling wall
-\item Multiple sets offers a means to reduce time to solution by
-  solving multiple copies of the original problem and combining the
-  solutions using superposition
-\item MCSA is most efficiently used in parallel as a stochastic
-  realization of an additive Schwarz method
-\item With future optimization, large scale reactor analyses in which
-  the transport operator is fully formed (e.g. the $SP_N$ equations)
-  could experience improved time to solution on leadership-class
-  machines with MCSA given the improved scaling performance
-\end{itemize}

### 2013-08-22 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
pushing full draft of disseration
moving parallel metrics to an appendix

@@
 model transport problem.
-
-\clearpage
 
@@
 \end{equation}
-
-\clearpage
 
@@
 adjacent domains, Monte Carlo transport may proceed in parallel.
-
-\clearpage
 
@@
 adapted to account for these additional techniques.
-
-\clearpage
 
@@
 
-From the data we see that for all solvers, the parallel results agree
+From the data we see that for all solvers the parallel results agree
 with the reference serial results within the solution tolerance for
@@
 %%---------------------------------------------------------------------------%%
-\section{Parallel Performance Metrics\ }
-\label{sec:parallel_performance_metrics}
-
-How effectively the presented algorithms parallelize MCSA will be
-determined by several performance measures. For this work, these
-measures will come directly from Keyes' primer on the scalability of
-domain decomposed methods and specifically his discussion on how to
-measure their parallel performance \cite{keyes_how_1999}. All of
-these metrics are derived from two measured quantities: the number of
-iterations required for convergence and the total wall time required
-for convergence. Each of these quantities are then measured in varying
-parameter studies by changing the number of processes in a simulation
-and by varying the local or global problem size. Using the notation of
-Keyes, we will define \textit{parallel efficiency} as $\eta(N,P)$
-where the efficiency is a function of the global problem size, $N$,
-and the number of processes in the calculation, $P$. There are several
-types of efficiencies that will be useful in our analysis of
-scalability with different definitions depending on the type of
-parametric scaling study performed. For all efficiencies, a value of
-100\% is considered perfect while a value of 0\% signals the worst
-possible performance. Often, as was observed in this work, it is the
-case that efficiencies above 100\% are computed in certain scaling
-studies. For the data presented in the following sections, an
-explanation of each observed case of super-unitary efficiencies will
-be provided.
-
-Of primary interest to the nuclear engineer is the \textit{strong
-  scaling} performance of an algorithm
-\cite{siegel_analysis_2012}. In this type of scaling study, the
-global problem size is fixed and the local problem size changes with
-the number of processes in the calculation. In this case, as $P$
-increases, $N$ is fixed, thus reducing the local problem size and
-performance by having a larger ratio of communication to computational
-work. This type of scaling is important because typically the global
-problem size is fixed due to the problem of interest. Consider the
-desire for a high fidelity neutronics simulation of the entire core of
-a nuclear reactor. In this case, the global problem is known from the
-geometry definition including the number of fuel assemblies and the
-geometry of those assemblies all the way down to the resolution of the
-space, angle, and energy discretizations required in order to capture
-physical phenomena of interest. Because the global problem size is
-known a priori, strong scaling is an important measure of how useful a
-solution technique is when larger and larger values of $P$ are used to
-solve the problem. Ideally, the time to solution for the the problem
-will decrease linearly as a function of $P$, permitting a faster time
-to solution using the maximum available resources and thus making sure
-all $P$ are positively affecting the runtime of the problem.
-
-For strong scaling, the objective is to then decrease the runtime of a
-fixed size global problem as a linear function of $P$. Based on this,
-we can define the strong scaling \textit{absolute efficiency} as:
-\begin{equation}
-  \eta_{strong}(N,P) = \frac{1}{P} \frac{T(N,1)}{T(N,P)}\:,
-  \label{eq:strong_scaling_absolute}
-\end{equation}
-where $T(N,P)$ is the wall time for a computation of global size $N$
-using $P$ processes. Note here that this particular definition is
-measured with respect to a serial computation. On leadership class
-machines, it is often the case that a problem size large enough to
-effectively execute on $O(10,000)$ cores or larger is significantly
-larger than any problem that may be solved by a single core due to
-memory restrictions. Therefore, we consider absolute scaling for a
-base case where $Q$ processes are used instead of a serial
-computation:
-\begin{equation}
-  \eta_{strong}(N,P|Q) = \frac{Q}{P} \frac{T(N,Q)}{T(N,P)}\:.
-  \label{eq:strong_scaling_absolute_ref}
-\end{equation}
-The absolute efficiency measure presented here only considers the
-total time to solution. Often, that total time to solution may be
-changing for an iterative method due to the fact that the number of
-iterations required to converge may be changing as a function of
-$P$. This will often happen in cases where modifying $N$ changes the
-spectral character of the problem, often making it stiffer and
-therefore more difficult to solve when $N$ increases. Therefore we
-also consider \textit{algorithmic efficiency} from Keyes' work which
-considers this potential change in iteration count:
-\begin{equation}
-  \eta_{alg}(N,P|Q) = \frac{I(N,Q)}{I(N,P)}\:,
-  \label{eq:algorithmic_efficiency}
-\end{equation}
-where $I(N,P)$ is the number of iterations required to converge a
-problem of global size $N$ computed using $P$ processes. We can then
-adjust the absolute efficiency using this new algorithmic efficiency
-in order to compute the parallel scalability of a single iteration
-such that algorithmic effects may be neglected. We define the
-\textit{implementation efficiency} as:
-\begin{equation}
-  \eta_{impl}(N,P|Q) = \frac{\eta}{\eta_{alg}}\:,
-  \label{eq:implementation_efficiency}
-\end{equation}
-using the absolute efficiency in the numerator, giving an effective
-measure of scalability of a single iteration of the algorithm.
-
-In addition to strong scaling, in certain cases it is useful to
-consider the \textit{weak scaling} of an iterative method. In this
-case, the ratio $N/P$ is fixed such that the local problem size
-remains the same for all values of $P$. In this case, the objective is
-to maintain a constant wall time while both $N$ and $P$ grow as a
-fixed ratio. In this case, we define the absolute efficiency as:
-\begin{equation}
-  \eta_{weak}(M|N,P|Q) = \frac{T(M,Q)}{T(N,P)}\:,
-  \label{eq:weak_scaling_absolute}
-\end{equation}
-subject to the constraint $M/Q = N/P$. The scaling is perfect if the
-runtime is static for all $P$ when $N/P$ is fixed.
-
-Often, weak scaling is of interest to the nuclear engineer due to the
-fact that the computational resources available are fixed and may not
-be able to contain the entire global problem of interest. In this
-case, an algorithm with good weak scaling performance permits all of
-the computational resources available to be used effectively to allow
-for the highest fidelity problem possible to be solved. In addition,
-for the resources available, the memory to processor ratio on a
-distributed machine is often fixed, giving an exercise in fixed $N/P$
-to solve the highest fidelity problem possible. Therefore, good weak
-scaling should not be viewed as an enhancement of time to solution for
-the engineer but rather permitting a higher fidelity problem to be
-solved within a given time constraint. For problems where increasing
-fidelity increases the stiffness of the system, there is often a
-trade-off between the absolute efficiencies computed for a given
-method and the degrading algorithmic efficiencies that arise when
-increasing the fidelity. Both time and hardware constraints should be
-considered when preparing a parallel calculation.
-
-\clearpage
-
-%%---------------------------------------------------------------------------%%
 \section{Leadership-Class Parallel Scaling Studies\ }
 \label{sec:leadership_scaling_studies}
-Using the parallel performance metrics presented in the previous
-section, we will next measure performance in various scaling studies
-using an implementation of the parallel MCSA algorithm to assess its 
-quality using the neutron diffusion problem presented in
-Appendix~\ref{chap:diffusion_problem}. For these scaling studies,
-large problems at high levels of concurrency will be solved at a
-leadership-class computing facility. In this section, we will outline
-the parallel machine on which the calculations were performed and then
-present a series of numerical studies that identify both strengths and
-weaknesses of the parallel algorithm. Both pure domain decomposition
-and MSOD parallel performance will be explored and compared to
-performance results for the same problems using a set of production
-parallel Krylov solvers.
+Using the parallel performance metrics presented in
+Appendix~\ref{chap:parallel_metrics}, we will next measure performance
+in various scaling studies using an implementation of the parallel
+MCSA algorithm to assess its quality using the neutron diffusion
+problem presented in Appendix~\ref{chap:diffusion_problem}. For these
+scaling studies, large problems at high levels of concurrency will be
+solved at a leadership-class computing facility. In this section, we
+will outline the parallel machine on which the calculations were
+performed and then present a series of numerical studies that identify
+both strengths and weaknesses of the parallel algorithm. Both pure
+domain decomposition and MSOD parallel performance will be explored
+and compared to performance results for the same problems using a set
+of production parallel Krylov solvers.
 

### 2013-08-22 doc/dissertation/document/backmatter/parallel_metrics.tex
removing redundant fanm verification data.

@@
 
-How effectively the presented algorithms parallelize MCSA will be
-determined by several performance measures. For this work, these
-measures will come directly from Keyes' primer on the scalability of
-domain decomposed methods and specifically his discussion on how to
-measure their parallel performance \cite{keyes_how_1999}. All of
-these metrics are derived from two measured quantities: the number of
-iterations required for convergence and the total wall time required
-for convergence. Each of these quantities are then measured in varying
-parameter studies by changing the number of processes in a simulation
-and by varying the local or global problem size. Using the notation of
-Keyes, we will define \textit{parallel efficiency} as $\eta(N,P)$
-where the efficiency is a function of the global problem size, $N$,
-and the number of processes in the calculation, $P$. There are several
-types of efficiencies that will be useful in our analysis of
-scalability with different definitions depending on the type of
-parametric scaling study performed. For all efficiencies, a value of
-100\% is considered perfect while a value of 0\% signals the worst
-possible performance. Often, as was observed in this work, it is the
-case that efficiencies above 100\% are computed in certain scaling
-studies. For the data presented in the following sections, an
-explanation of each observed case of super-unitary efficiencies will
-be provided.
+How effectively the algorithms presented in this work parallelize MCSA
+will be determined by several performance measures. For this work,
+these measures will come directly from Keyes' primer on the
+scalability of domain decomposed methods and specifically his
+discussion on how to measure their parallel performance
+\cite{keyes_how_1999}. All of these metrics are derived from two
+measured quantities: the number of iterations required for convergence
+and the total wall time required for convergence. Each of these
+quantities are then measured in varying parameter studies by changing
+the number of processes in a simulation and by varying the local or
+global problem size. Using the notation of Keyes, we will define
+\textit{parallel efficiency} as $\eta(N,P)$ where the efficiency is a
+function of the global problem size, $N$, and the number of processes
+in the calculation, $P$. There are several types of efficiencies that
+will be useful in our analysis of scalability with different
+definitions depending on the type of parametric scaling study
+performed. For all efficiencies, a value of 100\% is considered
+perfect while a value of 0\% signals the worst possible
+performance. Often, as was observed in this work, it is the case that
+efficiencies above 100\% are computed in certain scaling studies. For
+the data presented in the following sections, an explanation of each
+observed case of super-unitary efficiencies will be provided.
 

### 2013-08-22 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
removing redundant fanm verification data.

@@
 \label{sec:fanm}
-In production physics codes based on nonlinear equations sets,
-Newton-Krylov methods are the primary means of generating a fully
-consistent solution scheme
+In physics codes based on nonlinear equations sets, Newton-Krylov
+methods are the primary means of generating a fully consistent
+solution scheme
 \cite{evans_development_2006,evans_enhanced_2007,gaston_parallel_2009,godoy_parallel_2012}. It
@@
 For each of the benchmark problems, we will present the results of
-computations using FANM and directly compare them to results using a
+computations using FANM and compare them to results using a
 Newton-Krylov method in order to verify the correctness of FANM and
-its applicability to fluid problems. For each case, we will compare
-the global minimum, maximum, and average values for fluid velocity and
-pressure for all benchmarks and additionally temperature for the
-thermal convection cavity problem. If the values for these quantities
-computed using FANM match those computed using a production
-Newton-Krylov method, the implementation will be deemed correct and
-the method deemed valid for the purposes of this work. In addition, we
-will report the tolerance to which the nonlinear residual was
-converged as an additional means of comparison for correctness.
-
-Given the difficulty of these benchmark problems, the linear solve at
-each Newton step was preconditioned with an algebraic multigrid
-method\footnote{The ML algebraic multigrid library in the Trilinos
-  scientific computing libraries \cite{heroux_overview_2005} was used
-  for all problems.}, a common technique for preconditioning these
-equations \cite{ghia_high-re_1982,evans_enhanced_2007} with the
-preconditioner reconstructed at every Newton step for the new
-Jacobian. Without multigrid preconditioning, the residual in MCSA was
-observed to stagnate. Typically this happens because the Richardson
-iteration and Neumann series on which the Monte Carlo method is based
-will quickly damp the higher order error modes on the scale of the
-computational grid used to solve the problem but will reduce lower
-order error modes at a considerably lower rate. In terms of Monte
-Carlo, this stagnation manifests itself in histories that simply do no
-terminate and therefore no solution is acquired. For many of the
-problems presented here, it is likely that a more physics-based
-preconditioning strategy such as those presented in
+its applicability to fluid problems. Given the difficulty of these
+benchmark problems, the linear solve at each Newton step was
+preconditioned with an algebraic multigrid method\footnote{The ML
+  algebraic multigrid library in the Trilinos scientific computing
+  libraries \cite{heroux_overview_2005} was used for all problems.}, a
+common technique for preconditioning these equations
+\cite{ghia_high-re_1982,evans_enhanced_2007} with the preconditioner
+reconstructed at every Newton step for the new Jacobian. Without
+multigrid preconditioning, the residual in MCSA was observed to
+stagnate. Typically this happens because the Richardson iteration and
+Neumann series on which the Monte Carlo method is based will quickly
+damp the higher order error modes on the scale of the computational
+grid used to solve the problem but will reduce lower order error modes
+at a considerably lower rate. In terms of Monte Carlo, this stagnation
+manifests itself in histories that simply do no terminate and
+therefore no solution is acquired. For many of the problems presented
+here, it is likely that a more physics-based preconditioning strategy
+such as those presented in
 \cite{evans_development_2006,evans_enhanced_2007} will yield better
@@
 solver parameters including globalization and forcing term strategies
-were used with both methods. As with the radiation transport
-calculations in the previous chapter, algebraic preconditioning with
-the multigrid method results in dense linear systems. To mitigate
-this, the reduced domain approximation was again applied to the Monte
-Carlo solution within the MCSA solve occurring at each nonlinear
-iteration. The values used for the approximation will be given for
-each benchmark problem.
+were used with both methods with a nonlinear convergence criteria of a
+\sn{1}{-12} norm for the nonlinear residual. As with the radiation
+transport calculations in the previous chapter, algebraic
+preconditioning with the multigrid method results in dense linear
+systems. To mitigate this, the reduced domain approximation was again
+applied to the Monte Carlo solution within the MCSA solve occurring at
+each nonlinear iteration. The values used for the approximation will
+be given for each benchmark problem.
 
@@
 implementation of the FANM method. For the Newton-Krylov solutions,
-the same production implementation of GMRES from the Trilinos
-scientific computing library Aztec \cite{heroux_overview_2005} used
-for the verification of MCSA for the $SP_N$ solutions in the previous
-chapter was also used here as a means of comparison and FANM
-benchmarking.
+the same implementation of GMRES from the Trilinos scientific
+computing library Aztec \cite{heroux_overview_2005} used for the
+verification of MCSA for the $SP_N$ solutions in the previous chapter
+was also used here as a means of comparison and FANM benchmarking.
 
@@
 
-For the verification calculations, at each Rayleigh number the global
-minimum, maximum, and average values were recorded for fluid velocity
-in the x and y directions, fluid pressure, and fluid
-temperature. Tables~\ref{tab:convection_ra1e3_results},
-\ref{tab:convection_ra1e4_results},
-\ref{tab:convection_ra1e5_results}, and
-\ref{tab:convection_ra1e6_results} give the results of these
-computations. It should be noted that the negative pressures computed
-are due to a relative pressure being reported. For each of the data
-points presented, both Newton-Krylov solutions and FANM solutions
-computed identical numerical results for all of these global
-values. No variation in solution was noted for each of the
-computations larger than the tolerance of the converged nonlinear
-residual norms given in
-Table~\ref{tab:convection_residual_norm_comparison}. Variations in the
-convergence behavior and differences in performance for these
-solutions will be analyzed in the following section.
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lll}\hline\hline
-      \multicolumn{1}{l}{Parameter}& 
-      \multicolumn{1}{l}{Newton-Krylov Result}&
-      \multicolumn{1}{l}{FANM Result}\\
-      \hline
-      Average $U$ & -1.288835730E-05 & -1.288835730E-05 \\
-      Minimum $U$ & -3.677243960E+00 & -3.677243960E+00 \\
-      Maximum $U$ & 3.688908550E+00 & 3.688908550E+00 \\
-      \hline
-      Average $V$ & 3.403670530E-02 & 3.403670530E-02 \\
-      Minimum $V$ & -3.721178580E+00 & -3.721178580E+00 \\
-      Maximum $V$ & 3.791483420E+00 & 3.791483420E+00 \\
-      \hline
-      Average $P$ & 2.345554770E+02 & 2.345554770E+02 \\
-      Minimum $P$ & -5.509779180E+00 & -5.509779180E+00 \\
-      Maximum $P$ & 4.994885560E+02 & 4.994885560E+02 \\
-      \hline
-      Average $T$ & 3.595148700E-02 & 3.595148700E-02 \\
-      Minimum $T$ & 2.092448340E-04 & 2.092448340E-04 \\
-      Maximum $T$ & 7.178874510E-02 & 7.178874510E-02 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Thermal convection cavity FANM verification for a
-      Rayleigh number of \sn{1}{3}.} \textit{The values reported are
-      global minima, maxima, and averages for each variable in the
-      problem. The velocity is given in components in the $x$ and $y$
-      directions.}}
-  \label{tab:convection_ra1e3_results}
-\end{table}
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lll}\hline\hline
-      \multicolumn{1}{l}{Parameter}& 
-      \multicolumn{1}{l}{Newton-Krylov Result}&
-      \multicolumn{1}{l}{FANM Result}\\
-      \hline
-      Average $U$ & -7.489201240E-04 & -7.489201240E-04 \\
-      Minimum $U$ & -1.607318000E+01 & -1.607318000E+01 \\
-      Maximum $U$ & 1.669098270E+01 & 1.669098270E+01 \\
-      \hline
-      Average $V$ & 3.377624420E-01 & 3.377624420E-01 \\
-      Minimum $V$ & -1.932174800E+01 & -1.932174800E+01 \\
-      Maximum $V$ & 2.047326820E+01 & 2.047326820E+01 \\
-      \hline
-      Average $P$ & 1.861894950E+03 & 1.861894950E+03 \\
-      Minimum $P$ & -3.353416240E+01 & -3.353416240E+01 \\
-      Maximum $P$ & 4.440642810E+03 & 4.440642810E+03 \\
-      \hline
-      Average $T$ & 3.510568440E-01 & 3.510568440E-01 \\
-      Minimum $T$ & 2.092448340E-04 & 2.092448340E-04 \\
-      Maximum $T$ & 7.178874510E-02 & 7.178874510E-02 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Thermal convection cavity FANM verification for a
-      Rayleigh number of \sn{1}{4}.} \textit{The values reported are
-      global minima, maxima, and averages for each variable in the
-      problem. The velocity is given in components in the $x$ and $y$
-      directions.}}
-  \label{tab:convection_ra1e4_results}
-\end{table}
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lll}\hline\hline
-      \multicolumn{1}{l}{Parameter}& 
-      \multicolumn{1}{l}{Newton-Krylov Result}&
-      \multicolumn{1}{l}{FANM Result}\\
-      \hline
-      Average $U$ & -2.043639370E-02 & -2.043639370E-02 \\
-      Minimum $U$ & -3.814922470E+01 & -3.814922470E+01 \\
-      Maximum $U$ & -3.814922470E+01 & -3.814922470E+01 \\
-      \hline
-      Average $V$ & 2.774864300E+00 & 2.774864300E+00 \\
-      Minimum $V$ & -6.038682990E+01 & -6.038682990E+01 \\
-      Maximum $V$ & 7.873873560E+01 & 7.873873560E+01 \\
-      \hline
-      Average $P$ & 1.324208380E+04 & 1.324208380E+04 \\
-      Minimum $P$ & -1.390235490E+02 & -1.390235490E+02 \\
-      Maximum $P$ & 3.559215530E+04 & 3.559215530E+04 \\
-      \hline
-      Average $T$ & 3.027827770E+00 & 3.027827770E+00 \\
-      Minimum $T$ & 1.626405740E-02 & 1.626405740E-02 \\
-      Maximum $T$ & 7.162571130E+00 & 7.162571130E+00 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Thermal convection cavity FANM verification for a
-      Rayleigh number of \sn{1}{5}.} \textit{The values reported are
-      global minima, maxima, and averages for each variable in the
-      problem. The velocity is given in components in the $x$ and $y$
-      directions.}}
-  \label{tab:convection_ra1e5_results}
-\end{table}
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lll}\hline\hline
-      \multicolumn{1}{l}{Parameter}& 
-      \multicolumn{1}{l}{Newton-Krylov Result}&
-      \multicolumn{1}{l}{FANM Result}\\
-      \hline
-      Average $U$ & -1.802569840E-01 & -1.802569840E-01 \\
-      Minimum $U$ & -6.292084130E+01 & -6.292084130E+01 \\
-      Maximum $U$ & 2.951113240E+02 & 2.951113240E+02 \\
-      \hline
-      Average $V$ & 1.336398550E+01 & 1.336398550E+01 \\
-      Minimum $V$ & -1.345663290E+02 & -1.345663290E+02 \\
-      Maximum $V$ & 3.227873890E+02 & 3.227873890E+02 \\
-      \hline
-      Average $P$ & 6.354313840E+04 & 6.354313840E+04 \\
-      Minimum $P$ & -1.356309630E+02 & -1.356309630E+02 \\
-      Maximum $P$ & 2.316308730E+05 & 2.316308730E+05 \\
-      \hline
-      Average $T$ & 1.780727610E+01 & 1.780727610E+01 \\
-      Minimum $T$ & 8.265788890E-02 & 8.265788890E-02 \\
-      Maximum $T$ & 7.098873170E+01 & 7.098873170E+01 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Thermal convection cavity FANM verification for a
-      Rayleigh number of \sn{1}{6}.} \textit{The values reported are
-      global minima, maxima, and averages for each variable in the
-      problem. The velocity is given in components in the $x$ and $y$
-      directions.}}
-  \label{tab:convection_ra1e6_results}
-\end{table}
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{ccc}\hline\hline
-      \multicolumn{1}{c}{Rayleigh Number}& 
-      \multicolumn{1}{c}{Newton Krylov $||\ve{F(\ve{u})}||$}&
-      \multicolumn{1}{c}{FANM $||\ve{F(\ve{u})}||$}\\
-      \hline
-      \sn{1}{3} & \sn{4.542}{-14} & \sn{1.208}{-14} \\
-      \sn{1}{4} & \sn{1.045}{-12} & \sn{7.012}{-13} \\
-      \sn{1}{5} & \sn{1.784}{-12} & \sn{1.059}{-12} \\
-      \sn{1}{6} & \sn{3.404}{-12} & \sn{3.479}{-12} \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Thermal convection cavity nonlinear residual norm
-      achieved convergence.} \textit{The results presented here were
-      obtained from the benchmark verification calculations.}}
-  \label{tab:convection_residual_norm_comparison}
-\end{table}
-
-\clearpage
+For the verification calculations, at each Rayleigh number the fluid
+velocity in the x and y directions, fluid pressure, and fluid
+temperature values calculated Newton-Krylov were compared against the
+Newton-Krylov calculations. No variation in solution was noted for
+each of the computations larger than the floating point tolerance
+reported by Drekar with an absolute difference value of zero observed
+at all spatial locations for each quantity.
 
@@
 
-For the verification calculations, at each Reynolds number the global
-minimum, maximum, and average values were recorded for the fluid
-velocity in the x and y directions and the fluid
-pressure. Tables~\ref{tab:driven_re100_results},
-\ref{tab:driven_re300_results}, \ref{tab:driven_re500_results}, and
-\ref{tab:driven_re700_results} give the results of these
-computations. As with the thermal convection problem, the negative
-pressures computed are due to a relative pressure being reported. For
-each of the data points given, both Newton-Krylov solutions and FANM
-solutions computed identical numerical results for all of these global
-values. The norm of the converged nonlinear residual for the
-calculations is given in
-Table~\ref{tab:driven_residual_norm_comparison}. All computed global
-values were within these tolerances.
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lll}\hline\hline
-      \multicolumn{1}{l}{Parameter}& 
-      \multicolumn{1}{l}{Newton-Krylov Result}&
-      \multicolumn{1}{l}{FANM Result}\\
-      \hline
-      Average $U$ & 1.165632320E-01 & 1.165632320E-01 \\
-      Minimum $U$ & -2.423642920E+01 & -2.423642920E+01 \\
-      Maximum $U$ & 9.742670880E+01 & 9.742670880E+01 \\
-      \hline
-      Average $V$ & 2.038480420E-02 & 2.038480420E-02 \\
-      Minimum $V$ & -5.266315430E+01 & -5.266315430E+01 \\
-      Maximum $V$ & 2.720044160E+01 & 2.720044160E+01 \\
-      \hline
-      Average $P$ & -1.656193900E+02 & -1.656193900E+02 \\
-      Minimum $P$ & -9.635479370E+03 & -9.635479370E+03 \\
-      Maximum $P$ & 1.854517020E+04 & 1.854517020E+04 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Lid driven cavity FANM verification for a Reynolds
-      number of 100.} \textit{The values reported are global minima,
-      maxima, and averages for each variable in the problem. The
-      velocity is given in components in the $x$ and $y$ directions.}}
-  \label{tab:driven_re100_results}
-\end{table}
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lll}\hline\hline
-      \multicolumn{1}{l}{Parameter}& 
-      \multicolumn{1}{l}{Newton-Krylov Result}&
-      \multicolumn{1}{l}{FANM Result}\\
-      \hline
-      Average $U$ & 3.382078040E-01 & 3.382078040E-01 \\
-      Minimum $U$ & -9.467140360E+01 & -9.467140360E+01 \\
-      Maximum $U$ & 2.895885380E+02 & 2.895885380E+02 \\
-      \hline
-      Average $V$ & 8.780242440E-02 & 8.780242440E-02 \\
-      Minimum $V$ & -1.830156060E+02 & -1.830156060E+02 \\
-      Maximum $V$ & 8.827373340E+01 & 8.827373340E+01 \\
-      \hline
-      Average $P$ & -2.830380160E+03 & -2.830380160E+03 \\
-      Minimum $P$ & -2.352177540E+04 & -2.352177540E+04 \\
-      Maximum $P$ & 9.254997120E+04 & 9.254997120E+04 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Lid driven cavity FANM verification for a Reynolds
-      number of 300.} \textit{The values reported are global minima,
-      maxima, and averages for each variable in the problem. The
-      velocity is given in components in the $x$ and $y$ directions.}}
-  \label{tab:driven_re300_results}
-\end{table}
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lll}\hline\hline
-      \multicolumn{1}{l}{Parameter}& 
-      \multicolumn{1}{l}{Newton-Krylov Result}&
-      \multicolumn{1}{l}{FANM Result}\\
-      \hline
-      Average $U$ & 5.098921320E-01 & 5.098921320E-01 \\
-      Minimum $U$ & -1.731861690E+02 & -1.731861690E+02 \\
-      Maximum $U$ & 4.789146040E+02 & 4.789146040E+02 \\
-      \hline
-      Average $V$ & 1.365493140E-01 & 1.365493140E-01 \\
-      Minimum $V$ & -3.170930360E+02 & -3.170930360E+02 \\
-      Maximum $V$ & 1.655401440E+02 & 1.655401440E+02 \\
-      \hline
-      Average $P$ & -9.378317550E+03 & -9.378317550E+03 \\
-      Minimum $P$ & -4.155215280E+04 & -4.155215280E+04 \\
-      Maximum $P$ & 2.058742010E+05 & 2.058742010E+05 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Lid driven cavity FANM verification for a Reynolds
-      number of 500.} \textit{The values reported are global minima,
-      maxima, and averages for each variable in the problem. The
-      velocity is given in components in the $x$ and $y$ directions.}}
-  \label{tab:driven_re500_results}
-\end{table}
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lll}\hline\hline
-      \multicolumn{1}{l}{Parameter}& 
-      \multicolumn{1}{l}{Newton-Krylov Result}&
-      \multicolumn{1}{l}{FANM Result}\\
-      \hline
-      Average $U$ & 6.479541170E-01 & 6.479541170E-01 \\
-      Minimum $U$ & -2.550685400E+02 & -2.550685400E+02 \\
-      Maximum $U$ & 6.663054900E+02 & 6.663054900E+02 \\
-      \hline
-      Average $V$ & 1.833690840E-01 & 1.833690840E-01 \\
-      Minimum $V$ & -4.491052980E+02 & -4.491052980E+02 \\
-      Maximum $V$ & 2.474376340E+02 & 2.474376340E+02 \\
-      \hline
-      Average $P$ & 1.833690840E-01 & 1.833690840E-01 \\
-      Minimum $P$ & -6.241704500E+04 & -6.241704500E+04 \\
-      Maximum $P$ & 3.518174810E+05 & 3.518174810E+05 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Lid driven cavity FANM verification for a Reynolds
-      number of 700.} \textit{The values reported are global minima,
-      maxima, and averages for each variable in the problem. The
-      velocity is given in components in the $x$ and $y$ directions.}}
-  \label{tab:driven_re700_results}
-\end{table}
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lcc}\hline\hline
-      \multicolumn{1}{l}{Reynolds Number}& 
-      \multicolumn{1}{c}{Newton Krylov $||\ve{F(\ve{u})}||$}&
-      \multicolumn{1}{c}{FANM $||\ve{F(\ve{u})}||$}\\
-      \hline
-      100 & \sn{5.453}{-14} & \sn{6.577}{-14} \\
-      300 & \sn{2.537}{-13} & \sn{2.779}{-13} \\
-      500 & \sn{6.367}{-13} & \sn{6.252}{-13} \\
-      700 & \sn{9.159}{-13} & \sn{1.282}{-12} \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Lid driven cavity nonlinear residual norm achieved
-      convergence.} \textit{The results presented here were obtained
-      from the benchmark verification calculations.}}
-  \label{tab:driven_residual_norm_comparison}
-\end{table}
-
-\clearpage
+For the verification calculations, at each Reynolds number the fluid
+velocity in the x and y directions and fluid pressure values
+calculated Newton-Krylov were compared against the Newton-Krylov
+calculations. No variation in solution was noted for each of the
+computations larger than the floating point tolerance reported by
+Drekar with an absolute difference value of zero observed at all
+spatial locations for each quantity.
 
@@
 
-For the verification calculations, at each Reynolds number the global
-minimum, maximum, and average values were recorded for the fluid
-velocity in the x and y directions and the fluid
-pressure. Tables~\ref{tab:step_re200_results},
-\ref{tab:step_re300_results}, \ref{tab:step_re400_results}, and
-\ref{tab:step_re500_results} give the results of these computations.
-The norm of the converged nonlinear residual for the calculations is
-given in Table~\ref{tab:step_residual_norm_comparison}. The
-Newton-Krylov solutions and FANM solutions computed identical
-numerical results for all of these global values within converged
-tolerances.
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lll}\hline\hline
-      \multicolumn{1}{l}{Parameter}& 
-      \multicolumn{1}{l}{Newton-Krylov Result}&
-      \multicolumn{1}{l}{FANM Result}\\
-      \hline
-      Average $U$ & 8.776187720E+01 & 8.776187720E+01 \\
-      Minimum $U$ & -2.796828970E+01 & -2.796828970E+01 \\
-      Maximum $U$ & 2.639993990E+02 & 2.639993990E+02 \\
-      \hline
-      Average $V$ & -8.870766180E-01 & -8.870766180E-01 \\
-      Minimum $V$ & -2.636357380E+01 & -2.636357380E+01 \\
-      Maximum $V$ & 1.330345700E+01 & 1.330345700E+01 \\
-      \hline
-      Average $P$ & 1.226960220E+04 & 1.226960220E+04 \\ 
-      Minimum $P$ & 2.953808400E+01 & 2.953808400E+01 \\ 
-      Maximum $P$ & 2.202121190E+04 & 2.202121190E+04 \\ 
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Backward facing step FANM verification for a
-      Reynolds number of 200.} \textit{The values reported are global
-      minima, maxima, and averages for each variable in the
-      problem. The velocity is given in components in the $x$ and $y$
-      directions.}}
-  \label{tab:step_re200_results}
-\end{table}
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lll}\hline\hline
-      \multicolumn{1}{l}{Parameter}& 
-      \multicolumn{1}{l}{Newton-Krylov Result}&
-      \multicolumn{1}{l}{FANM Result}\\
-      \hline
-      Average $U$ & 1.316701890E+02 & 1.316701890E+02 \\
-      Minimum $U$ & -4.357431070E+01 & -4.357431070E+01 \\
-      Maximum $U$ & 3.958553050E+02 & 3.958553050E+02 \\
-      \hline
-      Average $V$ & -1.330267460E+00 & -1.330267460E+00 \\
-      Minimum $V$ & -3.467114630E+01 & -3.467114630E+01 \\
-      Maximum $V$ & 1.694243270E+01 & 1.694243270E+01 \\
-      \hline
-      Average $P$ & 1.681552980E+04 & 1.681552980E+04 \\
-      Minimum $P$ & 4.129311640E+01 & 4.129311640E+01 \\
-      Maximum $P$ & 2.940820620E+04 & 2.940820620E+04 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Backward facing step FANM verification for a
-      Reynolds number of 300.} \textit{The values reported are global
-      minima, maxima, and averages for each variable in the
-      problem. The velocity is given in components in the $x$ and $y$
-      directions.}}
-  \label{tab:step_re300_results}
-\end{table}
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lll}\hline\hline
-      \multicolumn{1}{l}{Parameter}& 
-      \multicolumn{1}{l}{Newton-Krylov Result}&
-      \multicolumn{1}{l}{FANM Result}\\
-      \hline
-      Average $U$ & 1.759720450E+02 & 1.759720450E+02 \\
-      Minimum $U$ & -5.941731260E+01 & -5.941731260E+01 \\
-      Maximum $U$ & 5.287146290E+02 & 5.287146290E+02 \\
-      \hline
-      Average $V$ & -1.776706740E+00 & -1.776706740E+00 \\
-      Minimum $V$ & -4.497362670E+01 & -4.497362670E+01 \\
-      Maximum $V$ & 1.973372840E+01 & 1.973372840E+01 \\
-      \hline
-      Average $P$ & 1.954352940E+04 & 1.954352940E+04 \\
-      Minimum $P$ & 5.294095080E+01 & 5.294095080E+01 \\
-      Maximum $P$ & 3.457969480E+04 & 3.457969480E+04 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Backward facing step FANM verification for a
-      Reynolds number of 400.} \textit{The values reported are global
-      minima, maxima, and averages for each variable in the
-      problem. The velocity is given in components in the $x$ and $y$
-      directions.}}
-  \label{tab:step_re400_results}
-\end{table}
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lll}\hline\hline
-      \multicolumn{1}{l}{Parameter}& 
-      \multicolumn{1}{l}{Newton-Krylov Result}&
-      \multicolumn{1}{l}{FANM Result}\\
-      \hline
-      Average $U$ & 2.199858290E+02 & 2.199858290E+02 \\
-      Minimum $U$ & -7.549221730E+01 & -7.549221730E+01 \\
-      Maximum $U$ & 6.605863330E+02 & 6.605863330E+02 \\
-      \hline
-      Average $V$ & -2.220108250E+00 & -2.220108250E+00 \\
-      Minimum $V$ & -5.888473080E+01 & -5.888473080E+01 \\
-      Maximum $V$ & 2.201938360E+01 & 2.201938360E+01 \\
-      \hline
-      Average $P$ & 1.983947940E+04 & 1.983947940E+04 \\
-      Minimum $P$ & -9.707913370E+03 & -9.707913370E+03 \\
-      Maximum $P$ & 3.771193860E+04 & 3.771193860E+04 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Backward facing step FANM verification for a
-      Reynolds number of 500.} \textit{The values reported are global
-      minima, maxima, and averages for each variable in the
-      problem. The velocity is given in components in the $x$ and $y$
-      directions.}}
-  \label{tab:step_re500_results}
-\end{table}
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{ccc}\hline\hline
-      \multicolumn{1}{c}{Reynolds Number}& 
-      \multicolumn{1}{c}{Newton Krylov $||\ve{F(\ve{u})}||$}&
-      \multicolumn{1}{c}{FANM $||\ve{F(\ve{u})}||$}\\
-      \hline
-      200 & 1.939E-11 & 1.975E-11 \\
-      300 & 2.786E-11 & 2.751E-11 \\
-      400 & 3.360E-11 & 3.597E-11 \\
-      500 & 5.824E-11 & 4.062E-11 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Backward facing step nonlinear residual norm
-      achieved convergence.} \textit{The results presented here were
-      obtained from the benchmark verification calculations.}}
-  \label{tab:step_residual_norm_comparison}
-\end{table}
-
-\clearpage
+For the verification calculations, at each Reynolds number the fluid
+velocity in the x and y directions and fluid pressure values
+calculated Newton-Krylov were compared against the Newton-Krylov
+calculations. No variation in solution was noted for each of the
+computations larger than the floating point tolerance reported by
+Drekar with an absolute difference value of zero observed at all
+spatial locations for each quantity.
 

### 2013-08-22 doc/dissertation/document/backmatter/mcsa_relaxation_parameters.tex
moving mcsa relaxation parameters to an appendix

--- 2013-08-22 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
@@
-\chapter{Monte Carlo Synthetic Acceleration\\ Methods for the $SP_N$ Equations}
-\label{ch:spn_equations}
-
-In this chapter, we briefly derive the simplified $P_N$ ($SP_N$)
-equations, closely following the work of Evans
-\cite{evans_simplified_2013}, in order to gain full understanding of
-the underlying system and its potential behavior in a Monte Carlo
-context. From the $P_N$ equations, we apply a set of approximations to
-yield the $SP_N$ equations for fixed source and criticality
-problems. Using the fully-formed linear operator for the transport
-problem, we explore solutions to the $SP_N$ equations with Monte Carlo
-Synthetic Acceleration using a challenging light water reactor fuel
-assembly criticality calculation as the driving problem. Several
-difficulties arise when applying MCSA to these problems that were not
-observed when investigating the simple transport system in the
-previous chapter.
-
-For Jacobi-based preconditioners, convergence with MCSA is difficult
-if not impossible for ill-conditioned systems with this behavior
-demonstrated using a simple neutron diffusion problem. In order to
-effectively solve the $SP_N$ equations for the fuel assembly problem,
-a suite of preconditioners along with a relaxation scheme is developed
-and studied within the context of MCSA. Using these preconditioners,
-several additional issues are observed and alleviated to a certain
-extent by applying the reduced domain approximation. MCSA solutions
-for the fuel assembly problem are then verified by comparing the
-solutions against production Krylov solvers for problems with varying
-numbers of energy groups. Finally performance is analyzed through
-comparison with those same Krylov solvers in terms of both iterative
-performance and CPU timing using the same set of problems utilized for
-the verification.
-
-%%---------------------------------------------------------------------------%%
-\section{The Neutron Transport Equation}
-\label{sec:transport_eq}
-As a starting point for the $SP_N$ equations we define the
-time-independent neutron transport equation
-\cite{lewis_computational_1993}:
-\begin{multline}
-  \hat{\Omega} \cdot \vec{\nabla} \psi(\vec{r},\hat{\Omega},E) +
-  \sigma(\vec{r},E) \psi(\vec{r},\hat{\Omega},E) = \\ \iint
-  \sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}' \cdot \hat{\Omega})
-  \psi(\vec{r},\hat{\Omega}',E') d\Omega' dE' +
-  q(\vec{r},\hat{\Omega},E)\:,
-  \label{eq:general_transport}
-\end{multline}
-with the variables defined as:
-\begin{itemize}
-\item $\vec{r}$ - neutron spatial position
-\item $\hat{\Omega}$ - neutron streaming direction with radial
-  component $\mu$ and azimuthal component $\omega$
-\item $\hat{\Omega}' \cdot \hat{\Omega} = \mu_0$ is the angle of
-  scattering
-\item $E$ - neutron energy
-\item $\psi(\vec{r},\hat{\Omega},E)$ - angular flux
-\item $\sigma(\vec{r},E)$ - total interaction cross section
-\item $\sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}')$ - probability
-  of scattering from direction $\hat{\Omega}'$ into an angular domain
-  $d\hat{\Omega}'$ about the direction $\hat{\Omega}$ and from energy
-  $E'$ to an energy domain $dE'$ about energy $E$
-\item $q(\vec{r},\hat{\Omega},E)$ - external source of neutrons.
-\end{itemize}
-For this work, it is sufficient to formulate
-Eq~(\ref{eq:general_transport}) in 1-dimensional Cartesian geometry:
-\begin{multline}
-  \mu \frac{\partial}{\partial x} \psi(x,\mu,E) + \sigma(x,E)
-  \psi(x,\mu,E) = \\ \iint \sigma_s(x,E' \rightarrow E,\hat{\Omega}'
-  \cdot \hat{\Omega}) \psi(x,\hat{\Omega}',E') d\Omega' dE' +
-  \frac{q(x,E)}{4 \pi}\:,
-  \label{eq:cart_1d_transport}
-\end{multline}
-where the angular component of the solution is no longer dependent on
-the azimuthal direction of travel and an isotropic source of neutrons
-is assumed. In addition, for fission systems the eigenvalue form of
-the transport equation is:
-\begin{multline}
-  \hat{\Omega} \cdot \vec{\nabla} \psi(\vec{r},\hat{\Omega},E) +
-  \sigma(\vec{r},E) \psi(\vec{r},\hat{\Omega},E) = \\ \iint
-  \sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}' \cdot \hat{\Omega})
-  \psi(\vec{r},\hat{\Omega}',E') d\Omega' dE' + \\ \frac{1}{k} \chi(E)
-  \iint \nu \sigma_f(\hat{r},E') \psi(\vec{r},\hat{\Omega}',E')
-  d\Omega' dE' + q(\vec{r},\hat{\Omega},E) \:,
-  \label{eq:eigenvalue_transport}
-\end{multline}
-with the additional variables defined as
-\begin{itemize}
-\item $k$ - multiplication factor
-\item $\chi(E)$ - fission neutron energy spectrum
-\item $\nu$ - average number of neutrons per fission
-\item $\sigma_f(r,E')$ - fission cross section\:.
-\end{itemize}
-In 1-dimensional Cartesian geometry,
-Eq~(\ref{eq:eigenvalue_transport}) becomes:
-\begin{multline}
-  \mu \frac{\partial}{\partial x} \psi(x,\mu,E) + \sigma(x,E)
-  \psi(x,\mu,E) = \\ \iint \sigma_s(x,E' \rightarrow
-  E,\hat{\Omega}' \cdot \hat{\Omega}) \psi(x,\hat{\Omega}',E')
-  d\Omega' dE' + \\ \frac{1}{k} \chi(E)
-  \iint \nu \sigma_f(x,E') \psi(x,\hat{\Omega}',E') d\Omega'
-  dE' + \frac{q(x,E)}{4 \pi}\:.
-  \label{eq:cart_1d_eigenvalue}
-\end{multline}
-
-%%---------------------------------------------------------------------------%%
-\section{Derivation of the Monoenergetic $SP_N$ Equations}
-\label{sec:spn_equations}
-The $P_N$ equations as derived in \cite{lewis_computational_1993}
-give $N+1$ coupled first-order equations capturing the spatial and
-angular-dependence of the solution:
-\begin{equation}
-   \frac{1}{2n+1} \frac{\partial}{\partial x}\Big[ (n+1) \phi_{n+1} + n
-     \phi_{n-1} \Big] + \Sigma_n \phi_n = q\delta_{n0} \:,
-  \label{eq:final_pn_equations}
-\end{equation}
-where $\Sigma_n = \sigma-\sigma_{sn}$ and the summations are truncated
-at some level of approximation $N$ such that $n = 0,1,\dotsc,N$. This
-yields a set of $N+1$ equations for $N+2$ flux moments. In multiple
-dimensions, the equation set becomes large and coupled not only
-through angular moments but also through the spatial variables. As a
-simpler alternative to multidimensional $P_N$ solutions, Gelbard
-recognized in 1960 that the planar $P_N$ equations could be simplified
-and applied an ad-hoc method to extend them to multiple dimensions,
-yielding the $SP_N$ equations. These equations are not only fewer in
-number, but also take on a diffusion-like form while maintaining the
-angular character of the flux, making them amenable to solutions with
-modern diffusion methods.
-
-First, the $P_N$ equations can be simplified to $(N+1)/2$ second-order
-equations by solving for the $n^{th}$ Legendre flux moment in the
-odd-order equations:
-\begin{equation}
-  \phi_n = \frac{1}{\Sigma_n}\Bigg[ q \delta_{no} -
-    \frac{\partial}{\partial x}\Big(\frac{n}{2n+1}\phi_{n-1} +
-    \frac{n+1}{2n+1} \phi_{n+1} \Big) \Bigg]\:, 
-  \label{eq:odd_moments}
-\end{equation}
-for $n = 1,3,\cdots,N$ and $\delta_{no} = 0\ \forall n \neq 0$. We can
-insert the odd moments into Eq~(\ref{eq:final_pn_equations}) to get a
-reduced group of equations for the even moments:
-\begin{multline}
-  -\frac{\partial}{\partial x}
-  \Bigg[\frac{n}{2n+1}\frac{1}{\Sigma_{n-1}} \frac{\partial}{\partial
-      x} \Big(\frac{n-1}{2n-1} \phi_{n-2} + \frac{n}{2n-1}\phi_n \Big)
-    \\+ \frac{n+1}{2n+1}\frac{1}{\Sigma_{n+1}} \frac{\partial}{\partial
-      x} \Big(\frac{n+1}{2n+3}\phi_n + \frac{n+2}{2n+3}\phi_{n+2}\Big)
-    \Bigg] \\+ \Sigma_n \phi_n = q \delta_{n0}\ \ \ \ \ \ \ \ \ n =
-  0,2,4,\cdots,N\:.
-  \label{eq:reduced_pn}
-\end{multline}
-Immediately, we note the diffusion-like nature of
-Eq~(\ref{eq:reduced_pn}) as compared to the original $P_N$
-equations. To extend these equations to multiple dimensions, Gelbard
-simply replaced the planar spatial derivatives in the reduced set of
-equations with general multidimensional gradient operators:
-\begin{multline}
-  -\nabla \cdot \Bigg[\frac{n}{2n+1}\frac{1}{\Sigma_{n-1}} \nabla
-    \Big(\frac{n-1}{2n-1} \phi_{n-2} + \frac{n}{2n-1}\phi_n \Big) \\+
-    \frac{n+1}{2n+1}\frac{1}{\Sigma_{n+1}} \nabla
-    \Big(\frac{n+1}{2n+3}\phi_n + \frac{n+2}{2n+3}\phi_{n+2}\Big)
-    \Bigg] \\+ \Sigma_n \phi_n = q \delta_{n0}\ \ \ \ \ \ \ \ \ n =
-  0,2,4,\cdots,N\:,
-  \label{eq:spn_equations}
-\end{multline}
-yielding a multidimensional set of $(N+1)/2$ angular coupled equations
-defined as the $SP_N$ equations. As with the $P_N$ equations, we
-provide closure to this set of equations with $\phi_{N+1} = 0$. As a
-concrete example, we will consider the $SP_7$ equations:
-\begin{subequations}
-  \begin{gather}
-    -\nabla \cdot \frac{1}{3 \Sigma_1} \nabla ( \phi_0 + 2\phi_2 ) +
-    \Sigma_0 \phi_0 = q \\ 
-    -\nabla \cdot \Bigg[ \frac{2}{15 \Sigma_1} \nabla ( \phi_0 + 2\phi_2
-      ) + \frac{3}{35 \Sigma_3}\nabla( 3\phi_2 + 4\phi_4)\Bigg] +
-    \Sigma_2 \phi_2 = 0\\
-    -\nabla \cdot \Bigg[ \frac{4}{63 \Sigma_3} \nabla ( 3\phi_2 +
-      4\phi_4 ) + \frac{5}{99 \Sigma_5}\nabla( 5\phi_4 +
-      6\phi_6)\Bigg] + \Sigma_4 \phi_4 = 0\\
-    -\nabla \cdot \Bigg[ \frac{6}{143 \Sigma_5} \nabla ( 5\phi_4 +
-      6\phi_6 ) + \frac{7}{195 \Sigma_7}\nabla(7\phi_6)\Bigg] +
-    \Sigma_6 \phi_6 = 0 \:.
-  \end{gather}
-  \label{eq:sp7_equations}
-\end{subequations}
-To further modify these equations, we can use a change of variables to
-create a new group of equations such that the gradients are operating
-on a single vector:
-\begin{subequations}
-  \begin{gather}
-    u_1 = \phi_0 + 2\phi_2 \\
-    u_2 = 3\phi_2 + 4\phi_4 \\
-    u_3 = 5\phi_4 + 6\phi_6 \\
-    u_4 = 7\phi_6 \:.
-  \end{gather}
-  \label{eq:spn7_subs}
-\end{subequations}
-When substituted into Eq~(\ref{eq:sp7_equations}), these terms give:
-\begin{subequations}
-  \begin{gather}
-    -\nabla \cdot \frac{1}{3 \Sigma_1} \nabla u_1 + \Sigma_0 \Bigg[
-    u_1 - \frac{2}{3}u_2 + \frac{8}{15}u_3 - \frac{16}{35}u_4 \Bigg]
-    = -q \\
-    -\nabla \cdot \Bigg[ \frac{2}{15 \Sigma_1} \nabla u_1 +
-    \frac{3}{35 \Sigma_3} \nabla u_2 \Bigg] + \Sigma_2 \Bigg[
-    \frac{1}{3}u_2 - \frac{4}{15}u_3 + \frac{8}{35}u_4 \Bigg] = 0 \\
-    -\nabla \cdot \Bigg[ \frac{4}{63 \Sigma_3} \nabla u_2 +
-    \frac{5}{99 \Sigma_5} \nabla u_3 \Bigg] + \Sigma_4 \Bigg[
-    \frac{1}{5}u_3 - \frac{6}{35}u_4 \Bigg] = 0 \\ 
-    -\nabla \cdot \Bigg[ \frac{6}{143 \Sigma_5} \nabla u_3 +
-    \frac{7}{195 \Sigma_7} \nabla u_4 \Bigg] + \Sigma_6 \Bigg[
-    \frac{1}{7}u_4 \Bigg] = 0 \:.
-  \end{gather}
-  \label{eq:spn7_subs_equations}
-\end{subequations}
-If we rearrange Eq~(\ref{eq:spn7_subs_equations}) such that only one
-divergence operation is present in each equation, we can formulate
-this as a matrix system of 4 equations in the case of the $SP_7$
-approximation:
-\begin{equation}
-  -\nabla \cdot D_n \nabla u_n + \sum_{m=1}^4 A_{nm} u_m =
-  q_n\ \ \ \ \ \ \ n = 1,2,3,4\:,
-  \label{eq:spn_matrix}
-\end{equation}
-with $\mathbf{u}$ the vector of solution variables:
-\begin{equation}
-  \mathbf{u} = ( u_1\ \ u_2\ \ u_3\ \ u_4 )^T \:,
-  \label{eq:spn7_solution_vector}
-\end{equation}
-$\mathbf{D}$ the vector of effective diffusion coefficients:
-\begin{equation}
-  \mathbf{D} = \Bigg( \frac{1}{3\Sigma_1}\ \ \frac{1}{7\Sigma_3}\ \
-  \frac{1}{11\Sigma_5}\ \ \frac{1}{15\Sigma_7} \Bigg)^T\:,
-  \label{eq:spn7_diffusion_coeffs}
-\end{equation}
-$\mathbf{q}$ the vector of source terms where the $0^{th}$ moment
-source has now been distributed through the system:
-\begin{equation}
-  \mathbf{q} = (
-  q\ \ -\frac{2}{3}q\ \ \frac{8}{15}q\ \ -\frac{16}{35}q )^T\:,
-  \label{eq:spn7_source_vector}
-\end{equation}
-and $\mathbf{A}$ a matrix of angular scattering terms:
-% NOTE: I copied the following matrix directly out of Tom's tech note
-% on the SPn equations.
-\begin{equation}
-  \mathbf{A} = 
-  {\tiny \begin{bmatrix}
-    (\Sigma_0) &
-    (-\frac{2}{3}\Sigma_0) &
-    (\frac{8}{15}\Sigma_0) &
-    (-\frac{16}{35}\Sigma_0) \\
-    %%
-    &&&\\
-    %%
-    (-\frac{2}{3}\Sigma_0) &
-    (\frac{4}{9}\Sigma_0 + \frac{5}{9}\Sigma_2) &
-    (-\frac{16}{45}\Sigma_0 - \frac{4}{9}\Sigma_2) &
-    (\frac{32}{105}\Sigma_0 + \frac{8}{21}\Sigma_2) \\
-    %%
-    &&&\\
-    %%
-    (\frac{8}{15}\Sigma_0) &
-    (-\frac{16}{45}\Sigma_0 - \frac{4}{9}\Sigma_2) &
-    (\frac{64}{225}\Sigma_0 + \frac{16}{45}\Sigma_2 + \frac{9}{25}\Sigma_4) &
-    (-\frac{128}{525}\Sigma_0 - \frac{32}{105}\Sigma_2 - \frac{54}{175}\Sigma_4)
-    \\ 
-    %%
-    &&&\\
-    %%
-    (-\frac{16}{35}\Sigma_0) &
-    (\frac{32}{105}\Sigma_0 + \frac{8}{21}\Sigma_2) &
-    (-\frac{128}{525}\Sigma_0 - \frac{32}{105}\Sigma_2 - \frac{54}{175}\Sigma_4)
-    & 
-    (\frac{256}{1225}\Sigma_0 + \frac{64}{245}\Sigma_2 +
-    \frac{324}{1225}\Sigma_4 + \frac{13}{49}\Sigma_6)
-  \end{bmatrix}}\:.
-  \label{eq:A_matrix}
-\end{equation}
-Note that the term $\sum_{m=1}^4 A_{nm} u_m$ in
-Eq~(\ref{eq:spn_matrix}) couples the moments in each equation while
-the diffusive term in each equation is only for a single
-'pseudo-moment' $u_n$. As noted by Evans \cite{evans_simplified_2013},
-lower order $SP_N$ approximations can be generated by setting higher
-order even moments in this system to zero (e.g. $\phi_6 = \phi_4 = 0$
-yields the $SP_3$ equations). For these equations, the $SP_N$ order
-defines the largest order flux moment maintained in the calculation
-and the $P_N$ order defines the largest order of the cross section
-expansion maintained.
-
-\subsection{Eigenvalue Form of the $SP_N$ Equations}
-\label{subsec:eigenvalue_form}
-For criticality problems, the space, angle, and energy discretizations
-for the $SP_N$ equations can be applied to the general eigenvalue form
-of the transport equation given by
-Eq~(\ref{eq:eigenvalue_transport}). We can readily replace the fixed
-source in Eq~(\ref{eq:spn_matrix}) with a fission source this time
-considering the multigroup form of the $SP_N$ equations derived in
-Appendix~\ref{chap:mg_spn_equations}:
-\begin{multline}
-  -\nabla \cdot \Bigg[\frac{n}{2n+1}\mathbf{\Sigma_{n-1}}^{-1} \nabla
-    \Big(\frac{n-1}{2n-1} \mathbf{\Phi_{n-2}} +
-    \frac{n}{2n-1}\mathbf{\Phi_n} \Big) \\+
-    \frac{n+1}{2n+1}\mathbf{\Sigma_{n+1}}^{-1} \nabla
-    \Big(\frac{n+1}{2n+3}\mathbf{\Phi_n} +
-    \frac{n+2}{2n+3}\mathbf{\Phi_{n+2}}\Big) \Bigg] \\+
-  \mathbf{\Sigma_n} \mathbf{\Phi_n} = \frac{1}{k} \mathbf{F}
-  \mathbf{\Phi_n} \delta_{n0} \ \ \ \ \ \ \ \ \ n = 0,2,4,\cdots,N\:.
-  \label{eq:multigroup_spn_eigenvalue}
-\end{multline}
-with $\mathbf{\Phi_n}$ the vector of multigroup flux moments and the
-fission matrix, $\mathbf{F}$, given by
-\cite{evans_simplified_2013}. We again apply the change of variables
-to yield the set of multigroup pseudo-moments $\mathbb{U}_n$ where now
-the application of the fission matrix to the moment vectors will be
-expanded into a set of block matrices in identical fashion to the
-scattering matrices:
-\begin{equation}
-  -\nabla \cdot \mathbb{D}_n \nabla \mathbb{U}_n + \sum_{m=1}^4
-  \mathbb{A}_{nm} \mathbb{U}_m = \frac{1}{k} \sum_{m=1}^4
-  \mathbb{F}_{nm} \mathbb{U}_m\ \ \ \ \ \ \ n = 1,2,3,4\:.
-  \label{eq:spn_fission_matrix}
-\end{equation}
-It should be noted here that with respect to a general MCSA scheme,
-the introduction of fission in the system will only affect the source
-vector in the linear system for the eigenvalue schemes used in this
-work. The linear operator, and therefore overall MCSA performance will
-be dictated by streaming and scattering as defined on the left-hand
-side.
-
-%%---------------------------------------------------------------------------%%
-\section{Basic Preconditioning for the $SP_N$ Equations}
-\label{sec:spn_spectral_analysis}
-Previous work in applying MCSA to the thermal radiation diffusion
-equations had success in using Jacobi preconditioning
-\cite{evans_monte_2012}. In this section, we will compute the spectral
-radius of the iteration matrix formed by the $SP_N$ equations while
-parametrically varying the $SP_N$ order, the $P_N$ order, the number
-of energy groups, and upscattering/downscattering between energy
-groups for Jacobi preconditioned problems to ensure that this type of
-preconditioning will be applicable to the fuel assembly problem. For
-each problem in the parameter study, the values given in
-Table~\ref{tab:spn_fixed_parameters} were fixed for each spatial
-location and energy group.
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{ll}\hline\hline
-      \multicolumn{1}{c}{Parameter}& 
-      \multicolumn{1}{c}{Value} \\\hline
-      Mesh Element Size & 1.0 \\
-      Mesh Elements & $4 \times 4 \times 4$ \\
-      Reflecting Boundaries & True \\
-      Materials & 1 \\
-      Source Strength & 1.0 \\
-      Total Cross Section & 5.0 \\
-      In-Group Cross Section & 0.25 \\
-      Downscatter Cross Section & 1.0 \\
-      Upscatter Cross Section & 0.1 \\
-      Eigenvalue Solver Tolerance & $1.0\times10^{-8}$ \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Fixed parameters for the $SP_N$ spectral radius
-      parameter study.} \textit{The parameters were fixed for each
-      spatial location and energy group.}}
-  \label{tab:spn_fixed_parameters}
-\end{table}
-
-Mesh element sizes were the same in all cardinal directions with
-reflecting conditions on each boundary. For the energy parameter, 1
-and 10 energy groups were used with upscatter/downscatter varied for
-the 10 group case. For the downscatter cases, all groups could scatter
-to all lower energy groups with the cross sections given in
-Table~\ref{tab:spn_fixed_parameters} while for the upscatter case all
-groups could scatter to all other energy groups with different cross
-sections for upscatter and downscatter used. A uniform isotropic
-source of neutrons is given in each group as well.
-
-The matrices generated by the discretization of this problem are
-sparse with a block-based form as dictated by
-Eq~(\ref{eq:spn_multigroup_system}). Figure~\ref{fig:group1} gives the
-sparsity pattern for a monoenergetic $SP_7$ discretization while
-Figures~\ref{fig:group10ds} and \ref{fig:group10us} give the sparsity
-patterns for the same problem for 10 energy groups without and with
-upscatter respectively. For these figures, the parameters in
-Table~\ref{tab:spn_fixed_parameters} with the number of spatial
-elements reduced to $2 \times 2 \times 2$ to show the blocks generated
-by the discretization.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=3.25in]{chapters/spn_equations/group1.png}
-  \end{center}
-  \caption{\textbf{Sparsity pattern for 1-group $SP_7$
-      discretization.} \textit{A $2\times 2 \times 2$ element mesh was
-      used to show detail of the blocks formed by the discretization.}}
-  \label{fig:group1}
-\end{figure}
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=3.25in]{chapters/spn_equations/group10ds.png}
-  \end{center}
-  \caption{\textbf{Sparsity pattern for 10-group $SP_7$ discretization
-      with downscatter only.} \textit{A $2\times 2 \times 2$ element
-      mesh was used to show detail of the blocks formed by the
-      discretization.}}
-  \label{fig:group10ds}
-\end{figure}
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=3.25in]{chapters/spn_equations/group10us.png}
-  \end{center}
-  \caption{\textbf{Sparsity pattern for 10-group $SP_7$ discretization
-      with downscatter and upscatter.} \textit{A $2\times 2 \times 2$ element
-      mesh was used to show detail of the blocks formed by the
-      discretization.}}
-  \label{fig:group10us}
-\end{figure}
-We note a few key features of the sparsity plots. The first is that
-for multigroup problems without full upscatter and downscatter
-(i.e. Figure~\ref{fig:group10ds}), the resulting matrix is asymmetric
-and therefore a linear solver that can handle asymmetric linear
-systems is required. Nearly all problems of interest will not have
-full upscattering or downscattering. Second we note the largely
-diagonal character of these systems, although the blocks from
-Eq~(\ref{eq:A_block_matrix}) are readily apparent.
-
-Based on this both block and diagonally dominant structure for
-matrices formed by the general multigroup $SP_N$ equations, we choose
-\textit{block Jacobi} preconditioning as a left preconditioner for the
-system. Block Jacobi preconditioning extracts the diagonal elements of
-the matrix as the preconditioner where now the elements extracted are
-the blocks on the diagonal as shown on the left side of
-Figure~\ref{fig:block_jacobi_ex}. Shown on the right side of
-Figure~\ref{fig:block_jacobi_ex}, inversion of the preconditioner is
-trivial with each diagonal block inverted separately. For the $SP_N$
-equations, Eq~(\ref{eq:A_block_matrix}) gives a block size of
-$N_g\times(N+1)/2$.
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.5}{
-    \input{chapters/spn_equations/block_jacobi.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Block Jacobi preconditioning strategy used for the
-      $SP_N$ equations.} \textit{Left: The preconditioner is formed by
-      the diagonal blocks of the matrix. Right: Inversion of the
-      preconditioner is trivial and decoupled by block.}}
-  \label{fig:block_jacobi_ex}
-\end{figure}
-
-For high performance implementations this has several attractive
-properties. First, the blocks in the matrix come from the energy/angle
-discretization of the transport equation as given by
-Eq~(\ref{eq:A_block_matrix}). Each block on the diagonal is bound to a
-mesh element in the system (note there are 8 blocks on the diagonal in
-each of the sparsity patterns with a mesh of $2 \times 2 \times 2$)
-and therefore we expect the matrix elements forming the block to be
-entirely local in domain decomposed calculations. Second, these blocks
-are typically dense and nearly lower triangular for many transport
-problems meaning that established dense matrix methods can be used for
-fast inversion.
-
-Spectral radius computations for the block Jacobi preconditioned
-iteration matrix were performed with Table~\ref{tab:group1bj} giving
-the results for the 1-group case, Table~\ref{tab:group10dsbj} for the
-10-group case with full downscatter only and
-Table~\ref{tab:group10usbj} for the 10-group case with full
-downscatter and full upscatter. A block size of $N_g\times(N+1)/2$ was
-used for the preconditioner in all cases.
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{cccccc}\hline\hline
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{$SP_N$ Order}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{} \\
-       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
-       & \textbf{0} & 0.0635 & 0.1269 & 0.1444 & 0.1513 \\
-       & \textbf{1} & 0.0666 & 0.1315 & 0.1474 & 0.1534 \\
-      $P_N$ Order & \textbf{3} & 0.0666 & 0.1365 & 0.154 & 0.1592 \\
-       & \textbf{5} & 0.0666 & 0.1365 & 0.1562 & 0.163 \\
-       & \textbf{7} & 0.0666 & 0.1365 & 0.1562 & 0.164 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Spectral radius results for the block Jacobi
-      preconditioned iteration matrix with 1 energy group.}}
-  \label{tab:group1bj}
-\end{table}
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{cccccc}\hline\hline
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{$SP_N$ Order}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{} \\
-       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
-       & \textbf{0} & 0.0647 & 0.1275 & 0.1449 & 0.1514 \\
-       & \textbf{1} & 0.0686 & 0.1338 & 0.1484 & 0.1547 \\
-      $P_N$ Order & \textbf{3} & 0.0687 & 0.1399 & 0.1582 & 0.1625 \\
-       & \textbf{5} & 0.0692 & 0.1399 & 0.1582 & 0.1657 \\
-       & \textbf{7} & 0.0678 & 0.1393 & 0.1624 & 0.166 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Spectral radius results for the block Jacobi
-      preconditioned iteration matrix with 10 energy groups and full
-      downscatter.}}
-  \label{tab:group10dsbj}
-\end{table}
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{cccccc}\hline\hline
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{$SP_N$ Order}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{} \\
-       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
-       & \textbf{0} & 0.1887 & 0.2267 & 0.2285 & 0.2286 \\
-       & \textbf{1} & 0.4535 & 0.5044 & 0.5045 & 0.5045 \\
-      $P_N$ Order & \textbf{3} & 0.4535 & 0.6453 & 0.6506 & 0.6506 \\
-       & \textbf{5} & 0.4535 & 0.6453 & 0.6802 & 0.6818 \\
-       & \textbf{7} & 0.4535 & 0.6453 & 0.6802 & 0.6927 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Spectral radius results for the block Jacobi
-      preconditioned iteration matrix with 10 energy groups, full
-      downscatter and full upscatter.}}
-  \label{tab:group10usbj}
-\end{table}
-
-From the tabulated block Jacobi data it is clear that this is a viable
-preconditioning choice for all $SP_N$ problems in terms of Monte Carlo
-solution methods. All cases were observed to have a spectral radius
-below unity. Based on these results, the block Jacobi method should be
-the applicable to more complicated problems representative of physical
-nuclear systems.
-
-%%---------------------------------------------------------------------------%%
-\section{Fuel Assembly Criticality Calculations}
-\label{sec:fuel_assembly_calcs}
-Fuel assembly calculations are a critical piece of nuclear engineering
-infrastructure for reactor core analysis and design. At this level,
-individual fuel pins may be resolved at fine resolution in a variety
-of configurations. As a sophisticated problem of interest to push the
-limits of MCSA, a hot zero-power $17 \times 17$ pin assembly will be
-used with varying energy group structure and $SP_N$ discretization in
-a criticality calculation. A cross section along the vertical axis
-showing homogenized fuel pin, control rod, and moderator materials and
-the associated grid is given by Figure~\ref{fig:problem3_radial_mat}
-while a cross section of the materials configuration along the
-horizontal axis is given by Figure~\ref{fig:problem3_axial_mat}. A
-detailed view of the assembly bottom is given in
-Figure~\ref{fig:problem3_end}. On the top and bottom of the assembly,
-vacuum conditions are used as well as on the top and right boundaries
-in Figure~\ref{fig:problem3_radial_mat}. Reflecting conditions are
-used on the left and bottom boundaries of
-Figure~\ref{fig:problem3_radial_mat}, effectively giving a
-representation of one quarter of the assembly. For the spatial
-discretization, each fuel pin (gray regions in
-Figure~\ref{fig:problem3_radial_mat}) is resolved by a $2 \times 2$
-mesh with materials and cross sections homogenized over this region.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4in]{chapters/spn_equations/problem3_radial_mat.png}
-  \end{center}
-  \caption{\textbf{Fuel assembly mesh and geometry cross section.}
-    \textit{Reflecting boundaries are used on the left and lower
-      boundaries to create a complete $17 \times 17$ assembly
-      geometry. Gray regions are homogenized fuel and red regions are
-      homogenized control rods. The control rods were removed for
-      these problems. Each fuel pin is resolved by a $2 \times 2$
-      mesh.}}
-  \label{fig:problem3_radial_mat}
-\end{figure}
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6.0in]{chapters/spn_equations/problem3_axial_mat.png}
-  \end{center}
-  \caption{\textbf{Fuel assembly geometry cross section.} \textit{The
-      geometry is subdivided along the axial direction into 50 zones
-      spaced to capture material boundaries. Important details include
-      spacer grids along the length of the fuel pins and reactor core
-      structures on the top and bottom of the assembly. Lighter purple
-      material in the center of the assembly is moderator and darker
-      purple/red material is fuel.}}
-  \label{fig:problem3_axial_mat}
-\end{figure}
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6.0in]{chapters/spn_equations/problem3_end.png}
-  \end{center}
-  \caption{\textbf{Fuel assembly geometry end detail.}
-    \textit{Reactor core structure including spacer grids and plenum
-      has been included. Lighter purple material on the right of the
-      figure is moderator and darker purple/red material is fuel.}}
-  \label{fig:problem3_end}
-\end{figure}
-
-Significant geometric details are contained in the model including
-spacer grids, fuel pins with homogenized cladding and gas gap, core
-plenum, and moderator with boron. Group cross sections and other
-discrete nuclear data are generated as needed by a cross-section
-processing module dependent on the meshing parameters used to
-discretize the geometry and single-dimension pin-cell calculations for
-initial flux spectrum generation. Table~\ref{tab:problem3_parameters}
-gives the primary design parameters for the fuel assembly
-calculations.
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{ll}\hline\hline
-      \multicolumn{1}{l}{\textbf{Parameter}} & 
-      \multicolumn{1}{l}{\textbf{Value}} \\
-      Power Level & 0 MW \\
-      Inlet Temperature & 326.85C \\
-      Fuel Temperature & 600C \\
-      Boron Concentration & 1300 ppm \\
-      Moderator Density & 0.743 g/cc \\
-      Helium Density & \sn{1.79}{-4} g/cc \\
-      Zirconium Density & 6.56 g/cc \\
-      Stainless Steel Density & 8.0 g/cc \\
-      Inconel Density & 8.19 g/cc \\
-      UO2 Density & 10.257 g/cc \\
-      Fuel Pin Radius (w/o clad) & 0.4096 cm \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Design parameters for the $17 \times 17$ pin fuel
-      assembly criticality calculation.}}
-  \label{tab:problem3_parameters}
-\end{table}
-
-To generate the multiplication factor and steady-state flux
-distribution for this problem, at every eigenvalue iteration MCSA is
-used to solve the resulting $SP_N$ problem using the provided fission
-source. Algorithm~\ref{alg:power_iteration} presents the use of MCSA
-within a power iteration strategy to find the multiplication factor.
-\begin{algorithm}[h!]
-  \caption{Power Iteration MCSA Scheme}
-  \label{alg:power_iteration}
-  \begin{algorithmic}
-    \State $k_0 =$ initial guess
-    \State $\mathbf{\Phi}_0 =$ initial guess
-    \State $n = 0$
-    \While{$|\frac{k^n - k^{n-1}}{k^n}| < \epsilon$}
-    \Comment{Iterate until convergence of the eigenvalue}
-    \State $\mathbf{M} \mathbf{\Phi}^{n+1} = \frac{1}{k^n} \mathbf{F} \mathbf{\Phi}^n$
-    \Comment{Solve for the new flux state with MCSA}
-    \State $k^{n+1} = k^n \frac{\int \mathbf{F} \mathbf{\Phi}^{n+1} d\mathbf{r}}{\int
-      \mathbf{F} \mathbf{\Phi}^n d\mathbf{r}}$
-    \Comment{Update the multiplication factor}
-    \State $n = n+1$
-    \EndWhile
-  \end{algorithmic}
-\end{algorithm}
-Here, $\mathbf{M}$ is the transport operator generated on the
-left-hand side of the $SP_N$ discretization, $\mathbf{F}$ is the
-fission matrix, and $\mathbf{\Phi}$ the multigroup neutron flux. This
-problem is significantly more complicated than the simple test problem
-used for the previous spectral analysis. Fission has been introduced
-into the set of equations and the addition of moderator into the
-system will increase the amount of scattering, creating a
-significantly more difficult problem manifesting itself in an
-iteration matrix with a larger spectral radius. When using MCSA, the
-linear operator applied to $\mathbf{\Phi}^{n+1}$ at each eigenvalue
-iteration will dictate convergence and remain unchanged throughout the
-eigenvalue computation while the addition of fission to the system
-will only modify the source of neutrons and the multiplication factor
-at every eigenvalue iteration while not affecting Monte Carlo
-transport.
-
-\subsection{Preliminary Jacobi Preconditioned Calculations}
-\label{subsec:jacobi_prec_assembly_calc}
-We first use block Jacobi preconditioning to solve the $17 \times 17$
-fuel assembly problem. A single energy group problem was first
-computed with $SP_1$ discretization, effectively giving the one-speed
-neutron diffusion system for the fuel assembly resulting in 20,088
-degrees of freedom in the
-problem. Figure~\ref{fig:block_jacobi_res_mcsa} gives the residual
-infinity norm as a function of iteration for the MCSA linear solve in
-the first eigenvalue iteration using 25,000 stochastic histories at
-every iteration for the adjoint Neumann-Ulam solve.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=5in]{chapters/spn_equations/block_jacobi_res.pdf}
-  \end{center}
-  \caption{\textbf{Residual infinity norm vs. iteration for the block
-      Jacobi preconditioned MCSA solve during the first eigenvalue
-      iteration of the 1-group $17 \times 17$ fuel assembly problem.}
-    \textit{Convergence was not achieved with the block Jacobi
-      preconditioned method.}}
-  \label{fig:block_jacobi_res_mcsa}
-\end{figure}
-Convergence was not achieved as noted by the rapid rise in the
-residual over a few iterations. Based on the spectral radius
-computations performed, these results are not in line with
-expectations for this problem. Additional computations performed with
-\sn{1}{6} histories per iteration exhibited the same divergent
-behavior at a significant computational cost and compute time. Even if
-the problem may be ill-conditioned, we do expect convergence of MCSA
-given that the spectral radius requirement was met. To investigate
-further, a block Jacobi preconditioned Richardson iteration was used
-to solve the same
-problem. Figure~\ref{fig:block_jacobi_res_richardson} gives the
-residual infinity norm as a function of iteration for the Richardson
-linear solve in the first eigenvalue iteration.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=5in]{chapters/spn_equations/block_jacobi_rich_res.pdf}
-  \end{center}
-  \caption{\textbf{Residual infinity norm vs. iteration for the block
-      Jacobi preconditioned Richardson solve during the first
-      eigenvalue iteration of the 1-group $17 \times 17$ fuel assembly
-      problem.} \textit{A spectral radius near 1 was observed for the
-      iteration matrix.}}
-  \label{fig:block_jacobi_res_richardson}
-\end{figure}
-Poor convergence is observed for the Richardson iteration, however,
-convergence is achieved meaning that the preliminary eigenvalue
-criteria needed to satisfy MCSA convergence has also been met. The
-number of iterations required for the Richardson iteration to converge
-will give an approximation for the spectral radius via
-Eq~(\ref{eq:linear_k_iter_norm3}). With 7291 iterations required to
-converge to a tolerance of $\sn{1}{-6}$, $\rho(\mathbf{H}) \approx
-0.998$, nearing the limits of MCSA applicability and well beyond the
-spectral radii generated in the initial spectral analysis.
-
-Based on these results, it then appears that even if the simple
-criteria of a spectral radius of less than one is met and the
-Richardson iteration will converge with the same preconditioning, MCSA
-still may not converge. We then expect the issue to reside with the
-Neumann-Ulam solve providing the correction as the Richardson
-iteration is known to provide the correct result. Furthermore, the
-fact that the spectral radius is less than one means that the
-stochastic histories in the block Jacobi preconditioned Neumann-Ulam
-method are eventually being terminated by the weight cutoff as no
-artificial absorption was used for these preliminary
-calculations. Based on this, the correction being generated by the
-Neumann-Ulam solve is the component of MCSA causing a divergent
-solution.
-
-\subsection{MCSA Breakdown}
-\label{subsec:mcsa_break_down}
-For the fuel assembly problem, the initial block Jacobi preconditioned
-calculations used a one-speed $SP_1$ discretization, effectively
-giving a diffusion system. To study the breakdown of MCSA at iteration
-matrix spectral radii near one, we will use the simpler homogeneous
-2-dimensional one-speed neutron diffusion system presented in
-Appendix~\ref{chap:diffusion_problem} to isolate this behavior. In
-this system, we can vary the cross sections while maintaining a fixed
-grid in order to achieve varying spectral radii. For these studies, we
-neglect fission as MCSA behavior is dictated by the transport operator
-$\mathbf{M}$ in an eigenvalue scheme with the fission matrix used to
-generate a fixed source.
-
-For each solver and estimator combination, the spectral radius of the
-iteration matrix generated by the diffusion problem was varied by
-changing the absorption cross section from 0.25 to 100 while fixing
-the grid size at $100 \times 100$ with $h = 0.1$ and a fixed
-scattering cross section of unity. For each parameter variation, a
-minimum of one stochastic history per degree of freedom (DOF) in the
-problem was used to compute the Monte Carlo correction. If the solver
-could not converge in less than 100 iterations, the number of
-histories was increased by increments of 5,000 until convergence was
-achieved in less than 100 iterations. The number of iterations
-required to converge MCSA and the time to converge was recorded as a
-means to capture the breakdown.
-
-Figure~\ref{fig:breakdown_iterations} gives the number of iterations
-required to converge for the chosen number of histories per iteration
-given by Figure~\ref{fig:breakdown_histories} using the adjoint solver
-with the collision and expected value estimators and the forward
-solver with the collision estimator. For spectral radii less than
-0.97, all MCSA problems converged with 1 history per DOF (10,000 for
-this problem) with the number of iterations required to converge
-increasing as a function of spectral radius. Near a spectral radius of
-0.97, the number of histories required to converge MCSA in less than
-100 iterations takes a dramatic rise that exhibits neither exponential
-nor power law behavior. As the spectral radius approaches 1, the
-number of histories required becomes significant and effectively
-impractical to compute. Even with this simple diffusion problem, the
-behavior is consistent with that observed for the fuel assembly
-problem with $SP_1$ discretization. In that case, we estimated a
-spectral radius of $\approx 0.998$, larger than any of the spectral
-radii that could be computed within even 90 minutes of compute time
-for this simple two dimensional problem. For that problem, even
-1,000,000 histories ($\approx 50$ per DOF) was not enough to provide
-convergence. Even if single solve times of an hour can be tolerated,
-dozens of solves are typically required to compute the multiplication
-factor and flux spectrum within the $SP_N$ eigenvalue scheme for more
-difficult problems like the fuel assembly, making the method unusable.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4.25in]{chapters/spn_equations/breakdown_iterations.pdf}
-  \end{center}
-  \caption{\textbf{Iterations required to converge as a function of
-      spectral radius for the neutron diffusion problem.} \textit{The
-      number of histories was increased to achieve convergence in less
-      than 100 iterations. At least 10,000 histories were used for
-      each calculation.}}
-  \label{fig:breakdown_iterations}
-\end{figure}
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4.25in]{chapters/spn_equations/breakdown_histories.pdf}
-  \end{center}
-  \caption{\textbf{Histories per DOF required to converge as a
-      function of spectral radius for the neutron diffusion problem.}
-    \textit{The number of histories was increased to achieve
-      convergence in less than 100 iterations. At least 10,000
-      histories were used for each calculation with 10,000 DOFs in the
-      problem.}}
-  \label{fig:breakdown_histories}
-\end{figure}
-
-In addition to the significantly larger number of histories required
-to achieve convergence for ill-conditioned problems another penalty is
-paid due to histories that take longer to
-compute. Figure~\ref{fig:breakdown_time} gives the CPU time in seconds
-required to compute a single random walk averaged over the entire set
-of histories run in the calculation over all iterations. As the
-spectral radius increases (correlating to a higher ratio of scattering
-in the system) the random walk lengths increase, using more CPU time
-to finish the computation. Compared to spectral radii of 0.5, larger
-spectral radii over 0.97 have histories that require two orders of
-magnitude more computation time. This significant increase in
-computation time per history coupled with the significant increase in
-the number of histories required to converge is evidence that for
-problems with spectral radii above $\approx 0.97$, using MCSA to solve
-any problems of interest is entirely ineffective and not practical. At
-iteration matrix spectral radii this large, using significant numbers
-of stochastic histories in the Monte Carlo solve are not enough to
-reduce the stochastic uncertainty in the correction, causing the MCSA
-sequence to diverge instead of accelerating convergence. We therefore
-require a more expansive set of preconditioning techniques to move the
-eigenvalue spectrum of the $SP_N$ problem into a regime in which MCSA
-is more applicable and in which performance is improved.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4.25in]{chapters/spn_equations/breakdown_time.pdf}
-  \end{center}
-  \caption{\textbf{CPU time per history as a function of spectral
-      radius for the neutron diffusion problem.} \textit{As the
-      spectral radius grows, so do the length of the random
-      walks. Longer random walks require more CPU time to compute.}}
-  \label{fig:breakdown_time}
-\end{figure}
-
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{Advanced Preconditioning Strategies}
-\label{subsec:spn_advanced_preconditioning}
-For the fuel assembly criticality problem presented in the previous
-section, the spectral radius was near one using Jacobi preconditioning
-and in a regime in which MCSA breakdown is observed. In this regime
-the number of stochastic histories required to converge MCSA increases
-rapidly and the resulting poor performance is compounded by those
-histories being increasingly expensive to compute. To overcome this
-difficulty, advanced preconditioning strategies for the $SP_N$ problem
-are required beyond simple Jacobi methods that can reduce the spectral
-radius into a region of better MCSA performance. Two modern algebraic
-preconditioning strategies, ILUT and SPAINV, will be presented here
-and used within the explicit preconditioning framework given by
-Eq~(\ref{eq:left_right_mcsa}). Data
-showing their effects on MCSA solutions of the fuel assembly
-criticality problem will be presented.
-
-\subsection{ILUT Preconditioning}
-\label{subsec:spn_ilut_preconditioning}
-ILUT preconditioning is chosen first due to its applicability to
-problems that have a dominating elliptic component such as the fuel
-assembly problem or other problems with a high scattering to
-absorption ratio. Incomplete lower-upper (ILU) factorizations of the
-linear operator can be used a simple mechanism to form an approximate
-inverse of a preconditioner. To build the factorization, the sparse
-upper and lower triangular factors, $\mathbf{L}$ and $\mathbf{U}$, are
-computed such that the residual matrix formed by the factorization
-\cite{saad_iterative_2003}:
-\begin{equation}
-  \mathbf{R} = \mathbf{L} \mathbf{U} - \mathbf{A} \:,
-  \label{eq:ilu_residual_matrix}
-\end{equation}
-has a specified sparsity pattern and element magnitude. When a
-magnitude threshold is used, elements generated in the factors below
-that magnitude are dropped, resulting in ILU threshold (ILUT)
-preconditioning. The sparsity pattern in this case is determined from
-the input matrix to be preconditioned and the number of elements
-maintained in the factor is specified by a fill level parameter. A
-fill level of 1 will generate the same number of elements as the
-sparsity pattern of the input matrix while a fill level of 2 will
-contain twice as many elements in the factorization, resulting in a
-better representation of the true LU factorization. The inverse of the
-lower and upper triangular factors may then be easily inverted by
-means of simple elimination to produce the preconditioner. For MCSA,
-$\mathbf{L}^{-1}$ will be used on the left and $\mathbf{U}^{-1}$ on
-the right to precondition the system.
-
-For modern subspace methods, only the action to the preconditioner on
-a vector is required for efficient implementations and therefore a
-triangular solve can be used for this purpose. For the explicit
-preconditioning scheme presented in Eq~(\ref{eq:left_right_mcsa}), the
-fully inverted operator must be generated in order to build the set of
-probabilities and weights required for Monte Carlo
-sampling. Therefore, for a linear system of size $N$, $N$ triangular
-solves will be required in order to extract the inverse matrices from
-production ILU implementations. In addition, parallel implementations
-typically generate lower and upper factors that are only triangular
-locally, providing an easily parallelizable mechanism to generate the
-action of the preconditioner inverse. A consequence of this choice for
-parallel scalability is a degradation of the preconditioning quality
-as the size of the parallel system is increased. For serial
-computations, the triangular factorization is potentially exact
-depending on the parameters chosen while at thousands of parallel
-tasks, the global triangular factors differ significantly from the true
-factorization. As a result, more iterations are required at higher
-levels of parallelism to converge the system and will ultimately
-degrade overall scalability of the system with respect to total wall
-time to converge.
-
-MCSA preconditioned with ILUT was used to solve the fuel assembly
-problem presented in the previous section for a 1-group $SP_1$
-discretization. Unlike the Jacobi preconditioned strategy, convergence
-was achieved with ILUT preconditioning. To study convergence
-sensitivity to ILUT parameters, the fill level and drop tolerance were
-parametrically varied with the number of iterations to converge an
-eigenvalue iteration for the fuel assembly problem recorded along with
-the maximum number of non-zero entries observed in all matrix rows for
-the left/right preconditioned composite linear operator. To provide
-some sparsity to the factorization, the ILUT drop tolerance was used
-to drop elements in the extracted inverse triangular factors. For each
-calculation, the number of iterations required to converge reported
-was for a single eigenvalue iteration with \sn{3}{4} histories at each
-MCSA iteration to compute the Monte Carlo correction using the adjoint
-collision estimator. All ILUT calculations reported here were
-performed on a single CPU and therefore this data does not take into
-account the aforementioned effects of parallel decomposition on the
-quality of the ILUT preconditioning.
-
-Figure~\ref{fig:ilut_iterations} gives the number of iterations to
-converge the fixed source problem to a tolerance of \sn{1}{8}. As
-expected, the higher the fill level chosen for the ILUT factorization,
-the fewer iterations are required to converge the problem (only 9 MCSA
-iterations were required to converge the problem for the fill level of
-3 and drop tolerance of \sn{1}{-5} case). For higher levels of fill,
-the iterations needed to converge were not as sensitive, signaling
-that a larger drop tolerance can perhaps be used without a significant
-degradation in iterative performance. At smaller fill levels, the
-sensitivity to the ILUT drop tolerance is more significant. For all
-fill levels used, convergence was not achieved for the fuel assembly
-problem with a drop tolerance larger than \sn{1}{-3}.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4.25in]{chapters/spn_equations/ilut_iterations.pdf}
-  \end{center}
-  \caption{\textbf{Number of MCSA iterations required to converge an
-      eigenvalue iteration for the fuel assembly problem with ILUT
-      preconditioning as a function of ILUT drop tolerance.}
-    \textit{Each colored curve represents the iteration behavior for a
-      different ILUT fill level. Fill levels of 1.5, 2.0, 2.5, and 3.0
-      were used.}}
-  \label{fig:ilut_iterations}
-\end{figure}
-
-Unfortunately, gaining convergence (and excellent iterative
-performance) with MCSA for the $SP_N$ fuel assembly problem comes at
-an immediate cost. Figure~\ref{fig:ilut_size} gives the maximum number
-of non-zero entries in the composite linear operator generated by
-preconditioning as a function ILUT drop tolerance for varying values
-of fill level. For the 1-group $SP_1$ discretization, the original
-linear operator will contain only a maximum of 7 non-zero entries per
-row in the system. As observed in Figure~\ref{fig:ilut_size},
-performing the explicit preconditioning yields composite linear
-operators with $O(1,000)$ elements in a row for all combinations of
-fill level and drop tolerance, over 10\% of the total row size for
-this particular problem. This large number of row entries, observed
-for a significant fraction of rows in the system, creates several
-problems. First, sparsity is completely destroyed with each state in
-the system now coupled to over 10\% of the total states in the system
-through the composite iteration matrix. This means that Monte Carlo
-sampling tables will be large, requiring significant memory to store
-them and a substantial overhead in the sampling procedure during
-transport. Second, the matrix-matrix multiply operations required to
-build the composite operator see significant performance losses due to
-the amount of data that must be handled. In parallel, losing sparsity
-severely inhibits performance where now parallel operations require
-communications among orders of magnitude more processors for
-nearest-neighbor type algorithms.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4.25in]{chapters/spn_equations/ilut_size.pdf}
-  \end{center}
-  \caption{\textbf{Maximum number of non-zero entries observed for all
-      rows in the composite linear operator for the fuel assembly
-      problem with ILUT preconditioning given as a function of ILUT
-      drop tolerance.} \textit{Each colored curve represents the row
-      size for a different ILUT fill level. Fill levels of 1.5, 2.0,
-      2.5, and 3.0 were used.}}
-  \label{fig:ilut_size}
-\end{figure}
-
-As a means to assess the quality of the preconditioning, a simple
-metric is developed to address these concerns and allow comparison to
-future developments. For our studies, our core performance metric will
-be iterative performance with the minimum number of MCSA iterations
-required to converge the problem desired. For Monte Carlo
-calculations, this improved performance is balanced by the creation of
-a dense composite system and we seek to reduce the number of non-zero
-entries to a minimum value in order to lower the amount of coupling
-among states in the system and reduce the amount of memory used
-along with the potential latency overhead. Given these two objectives,
-the following metric is proposed:
-\begin{equation}
-  \text{Quality Metric} = (\text{\# iterations}) \times (\text{maximum
-    \# non-zero values})\:,
-\end{equation}
-where the highest quality preconditioning is one that minimizes this
-metric. For the ILUT preconditioned data provided,
-Figure~\ref{fig:ilut_quality} provides the computed metric as a
-function of ILUT drop tolerance for each fill level used. In general,
-the metric is similar to the data observed for the number of
-iterations required to converge as the non-zero row entries changed by
-a smaller fraction over drop tolerances tested.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4.25in]{chapters/spn_equations/ilut_quality.pdf}
-  \end{center}
-  \caption{\textbf{ILUT preconditioning quality metric for the fuel
-      assembly problem given as a function of ILUT drop tolerance.}
-    \textit{Each colored curve represents the quality metric behavior
-      for a different ILUT fill level. Fill levels of 1.5, 2.0, 2.5,
-      and 3.0 were used.}}
-  \label{fig:ilut_quality}
-\end{figure}
-
-It should be noted here that although only the behavior of the linear
-solve during a single eigenvalue iteration is reported, the behavior
-of the linear solver was observed to be consistent throughout the
-eigenvalue iterations (25 total eigenvalue iterations were required to
-converge the 1-group $SP_1$ problem). We expect this as the linear
-operator (which is unchanging in the fuel assembly criticality
-problem) dictates the convergence of MCSA. At each iteration the
-fission source provided to the linear problem is changing, however, we
-expect the same convergence behavior for all source vectors
-$\mathbf{b} \in \mathbb{R}^{N}$ where $||\mathbf{b}||_1 \neq 0$.
-
-\subsection{Sparse Approximate Inverse Preconditioning}
-\label{subsec:spn_spainv_preconditioning}
-SPAINV is chosen next because it the true inverse of the
-preconditioner is product of the algorithm, avoiding the need for many
-inverse matrix applications. In addition, the algorithm seeks to
-maintain the sparsity of the inverse preconditioner, potentially
-alleviating some of the dense operator issues observed with
-ILUT. Explicit formation of the preconditioner inverse is a
-requirement of the current MCSA preconditioning strategy. As a result,
-although ILUT preconditioning provided excellent iterative
-performance, a significant time penalty is paid to extract the inverse
-of the triangular factors through $N$ triangular solves. In addition,
-this explicit inversion yielded a composite linear operator that was
-dense, requiring modification of the system in order to achieve better
-scalability and Monte Carlo performance.
-
-Sparse approximate inverse (SPAINV) preconditioning is a technique
-that may potentially alleviate both of these constraints at the cost
-of reduced iterative performance. The method produces a sparse
-approximation of the inverse matrix directly by minimizing the
-Frobenius norm of the residual matrix \cite{saad_iterative_2003}:
-\begin{equation}
-  || \mathbf{I} - \mathbf{A} \mathbf{M} ||^2_F =
-  \sum_{j=1}^N ||\mathbf{e}_j - \mathbf{A} \mathbf{m}_j||^2_2 \:,
-  \label{eq:frobenius_norm_min}
-\end{equation}
-where the inverse preconditioning matrix $\mathbf{M}$ minimizes the
-norm on the right and $\mathbf{e}_j$ and $\mathbf{m}_j$ are the
-$j^{th}$ columns of the identity and preconditioning matrices
-respectively. Here, $\mathbf{M}$ is the actual inverse of the
-preconditioner and is formed directly by the algorithm. On the right
-hand side of Eq~(\ref{eq:frobenius_norm_min}), the Frobenius norm is
-represented as a effective sum of $N$ linear system residual norms.
-
-A few iterations of a subspace method or other iterative scheme can be
-used to effectively solve these systems to a loose tolerance and yield
-the columns of the approximate preconditioning matrix. At each
-iteration, threshold values are used to eliminate the components of
-each $\mathbf{m}_j$ column to maintain the desired level of
-sparsity. In addition, a sparsity pattern can be predetermined and
-enforced during this process. For the SPAINV implementation used for
-this work, the sparsity pattern was defined by levels such that a
-level of $N$ for an input operator $\mathbf{A}$ will generate a
-preconditioning matrix $\mathbf{M}$ with the same sparsity pattern as
-$\mathbf{A}^{N+1}$. By using the norm minimization strategy instead of
-a factorization such as ILUT, parallel results are reproduced
-regardless of parallel problem size, meaning that a serial computation
-will converge in the same number of iterations as a computation with
-thousands of cores.
-
-MCSA preconditioned with SPAINV was used to solve the fuel assembly
-problem as with ILUT for a 1-group $SP_1$ discretization. To study
-convergence sensitivity to SPAINV parameters, the number of levels in
-the sparsity pattern and threshold were parametrically varied with the
-number of iterations to converge a single eigenvalue iteration for the
-fuel assembly problem recorded along with the maximum number of
-non-zero entries observed in all matrix rows for the preconditioned
-composite linear operator. For each calculation, the number of
-iterations required to converge reported was for a single eigenvalue
-iteration with \sn{3}{4} histories at each MCSA iteration to compute
-the Monte Carlo correction using the adjoint collision estimator. All
-SPAINV calculations reported here were performed on a single CPU.
-
-Figure~\ref{fig:spainv_iterations} gives the number of MCSA iterations
-needed to converge the fuel assembly problem and
-Figure~\ref{fig:spainv_size} the maximum number of non-zero entries
-per row observed in the composite linear operator as a function the
-SPAINV threshold. For this analysis, the number of levels in the
-sparsity pattern was varied from 3 to 7 with convergence of the fuel
-assembly problem not achieved for smaller level values. Compared to
-ILUT preconditioning, at higher levels we see comparable iterative
-performance with a relative invariance to the threshold value. The
-threshold did have a significant effect, however, on the time required
-to generate the preconditioner with a threshold of 0.001 over an order
-of magnitude slower than a threshold of 0.01 due to the inclusion of a
-significant number of extra values in the input operator. In addition
-to comparable iterative performance, the sparsity of the composite
-operator is greatly improved with nearly an order of magnitude
-reduction in number of non-zero entries per row for lower level
-values.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4.25in]{chapters/spn_equations/spainv_iterations.pdf}
-  \end{center}
-  \caption{\textbf{Number of MCSA iterations required to converge a
-      single eigenvalue iteration for the fuel assembly problem with
-      SPAINV preconditioning as a function of SPAINV threshold.}
-    \textit{Each colored curve represents the iteration behavior for a
-      different SPAINV level pattern. Levels of 3, 4, 5, 6, and 7 were
-      used.}}
-  \label{fig:spainv_iterations}
-\end{figure}
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4.25in]{chapters/spn_equations/spainv_size.pdf}
-  \end{center}
-  \caption{\textbf{Maximum number of non-zero entries observed for all
-      rows in the composite linear operator for the fuel assembly
-      problem with SPAINV preconditioning given as a function of
-      SPAINV threshold.} \textit{Each colored curve represents the row
-      size for a different SPAINV level pattern. Levels of 3, 4, 5, 6,
-      and 7 were used.}}
-  \label{fig:spainv_size}
-\end{figure}
-
-With the comparable iterative performance and improved sparsity,
-SPAINV preconditioning should then have a more favorable quality
-metric than ILUT preconditioning. Figure~\ref{fig:spainv_quality}
-gives the quality metric as a function of the threshold value for each
-of the sparsity levels computed. Not only do we see an improved
-quality metric over ILUT preconditioning (about an order of magnitude
-less), but we also note that the ideal sparsity level is not the largest
-nor the smallest. In fact, the largest and smallest levels (3 and 7)
-performed the worst in terms of the quality metric with a level of 4
-performing the best. When CPU time to generate the preconditioner is
-considered, 4 is the clear choice for this problem as more time is
-required to do the approximate inversion as the sparsity pattern of
-the preconditioner grows in size.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4.25in]{chapters/spn_equations/spainv_quality.pdf}
-  \end{center}
-  \caption{\textbf{SPAINV preconditioning quality metric for the fuel
-      assembly problem given as a function of SPAINV threshold.}
-    \textit{Each colored curve represents the quality metric behavior
-      for a different SPAINV level pattern. Levels of 3, 4, 5, 6, and
-      7 were used.}}
-  \label{fig:spainv_quality}
-\end{figure}
-
-\clearpage
-
-\subsection{Applying the Reduced Domain Approximation}
-\label{subsec:spn_prec_rda}
-Although ILUT and SPAINV preconditioning permitted MCSA solutions for
-the fuel assembly problem, a primary concern is the number of non-zero
-states in each row of the system generated by the explicit
-preconditioning strategy. In many cases, orders of magnitude more
-matrix elements were generated resulting in poor scalability for
-domain decomposed algorithms and overall performance issues for Monte
-Carlo. As outlined in \S~\ref{subsec:reduced_domain_approximation},
-the reduced domain approximation may be used as a mechanism to
-potentially alleviate this problem by filtering elements of the
-composite matrix in each row that fall below a certain threshold value
-or by maintaining the largest $N$ elements in each row where $N$ is a
-designated fill level. In this section, we will apply the reduced
-domain approximation to both ILUT and SPAINV preconditioned system to
-reduce the density of the composite linear operator to more manageable
-levels and study the resulting iterative performance.
-
-For each preconditioner, the parameters that achieved the best quality
-metric results from the previous analysis were used. These correspond
-to ILUT parameters of a fill level of 5.0 and a drop tolerance of
-\sn{1}{-5} and SPAINV parameters of a sparsity level of 4 and a
-threshold of 0.1. The reduced domain threshold was set to \sn{1}{-10}
-in order to eliminate any exceedingly small values from the matrix
-(often this is simply removing non-zero elements within the floating
-point tolerance of zero). The reduced domain fill level was then
-varied, starting with the largest non-zero entries per row value
-observed for each of the preconditioning types in order to assess its
-effects relative to the case where no reduced domain approximation was
-applied.
-
-Figure~\ref{fig:rda_iterations} gives the number of iterations
-required to converge for each preconditioner type as a function of
-reduced domain fill level. Figure~\ref{fig:rda_quality} gives the
-corresponding quality metric for each data point where the number of
-non-zero entries used to compute the metric is equivalent to the
-reduced domain fill level. We first note that SPAINV preconditioning
-alone is significantly more sensitive to the reduction in domain size
-over the ILUT-based methods, although the preconditioner was of
-$O(100)$ non-zero entries per row without any approximation
-applied. For the ILUT-based methods, performance was significantly
-better with convergence achieved in less than 40 MCSA iterations with
-only 10 non-zero entries in each row (vs. 7 non-zero entries for the
-case with no preconditioning). Looking at the quality metric data, we
-see a reduction in the quality metric as a function of the reduced
-domain fill level, achieving 2 orders of magnitude reduction in the
-quality metric for the ILUT preconditioning.
-
-Although applying the reduced domain approximation results in a
-successful recovery of sparsity for the Monte Carlo problem while
-maintaining good convergence properties, there is still an issue of
-forming the composite operator before applying the approximation and
-potentially generating the transpose of this operator in the case of
-the adjoint Monte Carlo method. Because of this, memory and scaling
-issues may still be observed when building the probability and weight
-matrices for the Monte Carlo problem. In addition, the expensive
-extraction of the inverse of the preconditioning operators for the
-explicit scheme creates a significant cost in overall
-performance. Future work in this area should be considerate of these
-important components of the preconditioned MCSA algorithm.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4.25in]{chapters/spn_equations/rda_iterations.pdf}
-  \end{center}
-  \caption{\textbf{Number of MCSA iterations required to converge a
-      single eigenvalue iteration for the fuel assembly problem with
-      each preconditioning as a function of reduced domain
-      approximation fill level.} \textit{The largest fill level for
-      each preconditioning presented is that using the parameters that
-      gave the best results without the approximation.}}
-  \label{fig:rda_iterations}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4.25in]{chapters/spn_equations/rda_quality.pdf}
-  \end{center}
-  \caption{\textbf{Preconditioning quality metric for the fuel
-      assembly problem given as a function of reduced domain
-      approximation fill level.} \textit{The largest fill level for
-      each preconditioning presented is that using the parameters that
-      gave the best results without the approximation.}}
-  \label{fig:rda_quality}
-\end{figure}
-
-\clearpage
-
-\subsection{MCSA Relaxation Parameters}
-\label{subsec:spn_mcsa_relaxation}
+\chapter{MCSA Relaxation Parameters}
+\label{chap:spn_mcsa_relaxation}
 As another means of preconditioning (or variance reduction) to meet
 the goal of improving iterative performance, the Richardson iteration
-upon which the Neumann-Ulam method is built may be implemented with an
-adjustable scalar relaxation parameter:
+may be implemented with an adjustable scalar relaxation parameter:
 \begin{equation}
@@
 We apply these relaxation parameters to the fuel assembly problem
-along with the reduced domain approximation as a means of studying
-their effects. For each calculation presented, the number of
-iterations required to converge reported was for a single eigenvalue
-iteration.  Again, \sn{3}{4} histories are used at each MCSA iteration
-to compute the Monte Carlo correction using the adjoint collision
-estimator. A reduced domain fill level of 100 is used with a threshold
-of \sn{1}{-10} to filter small values. ILUT preconditioning with a
-drop tolerance of \sn{1}{-5} and fill level of 5 is used as
-well.
+defined in \S~\ref{sec:fuel_assembly_calcs} along with the reduced
+domain approximation as a means of studying their effects. For each
+calculation presented, the number of iterations required to converge
+was for a single eigenvalue iteration.  Again, \sn{3}{4} histories are
+used at each MCSA iteration to compute the Monte Carlo correction
+using the adjoint collision estimator. A reduced domain fill level of
+100 is used with a threshold of \sn{1}{-10} to filter small
+values. ILUT preconditioning with a drop tolerance of \sn{1}{-5} and
+fill level of 5 is used as well.
 
 Figure~\ref{fig:relax_iters} gives the number of iterations required
-to converge the fuel assembly problem for a 1-group SP1 discretization
-using varying combinations of the relaxation parameters. First, both
-parameters were fixed at the base case of 1 while the other parameter
-was varied as shown in the left-hand side plots of
-Figure~\ref{fig:relax_iters}. For the Richardson relaxation parameter,
-using a value larger than 1 gave better iterative performance up to a
-point, effectively providing a stronger extrapolation using the
-residual at each iteration. For the Neumann relaxation parameter, a
-value of less than 1 is observed to be ideal. Although initially
-counter-intuitive, the fact that the correction computed by the Monte
-Carlo solver has a stochastic error associated with it means that by
-using a Neumann relaxation parameter less than 1, the correction and
-its error are effectively dampened to improve iterative
-performance.
+to converge an eigenvalue iteration of the fuel assembly problem for a
+1-group SP1 discretization using varying combinations of the
+relaxation parameters. First, both parameters were fixed at the base
+case of 1 while the other parameter was varied as shown in the
+left-hand side plots of Figure~\ref{fig:relax_iters}. For the
+Richardson relaxation parameter, using a value larger than 1 gave
+better iterative performance up to a point, effectively providing a
+stronger extrapolation using the residual at each iteration. For the
+Neumann relaxation parameter, a value of less than 1 is observed to be
+ideal. Although initially counter-intuitive, the fact that the
+correction computed by the Monte Carlo solver has a stochastic error
+associated with it means that by using a Neumann relaxation parameter
+less than 1 the correction and its error are effectively dampened to
+improve iterative performance.
 
@@
 \end{figure}
-
-\clearpage
-
-\subsection{MCSA Estimator and Fixed Point Iteration Analysis}
-\label{subsec:spn_estimator_comparison}
-The iterative performance of MCSA will next be studied with both the
-collision and expected value estimators using the best preconditioning
-and relaxation parameter combinations found by the previous
-analysis. Additionally, both the Richardson-based MCSA iteration given
-by Eq~(\ref{eq:mcsa}) and the minimal residual-based MCSA iteration
-given by Eq~(\ref{eq:mcsa_min_res}) will be used. Although the
-expected value estimator achieved the best iterative performance in
-the previous comparison and analysis for the simple Poisson problem in
-\S~\ref{sec:parameter_estimator_analysis}, it is possible that the
-collision estimator may perform better from a timing perspective. The
-deterministic component of the expected value estimator that couples
-the current state to other local states through the iteration matrix
-stencil requires application of the estimator to potentially several
-orders of magnitude more states at every transition event.
-
-For this study, the fuel assembly problem with a 1-group $SP_1$
-discretization was solved using MCSA with the adjoint Monte Carlo
-solver using both the collision and expected value estimators and both
-the Richardson and minimal residual fixed point iterations. Using the
-results from the previous section, a Neumann relaxation parameter of
-0.7 was used with both estimators and fixed point iterations. For the
-collision estimator, a Richardson relaxation parameter of 1.1 was used
-with the Richardson iteration while it was found the expected value
-estimator had the best performance when a Richardson relaxation
-parameter of 1.0 was used instead. In addition, ILUT preconditioning
-with a drop tolerance of \sn{1}{-5} and a fill level of 5 was used
-with a reduced domain approximation fill level of 100 and a threshold
-of \sn{1}{-10} as in the previous analysis. Artificial absorption was
-also introduced with a value of 0.2 for the MCSA calculations that
-used the collision estimator as CPU timing performance was improved
-without a loss in iterative performance while no absorption was used
-for calculations using the expected value estimator. For each
-estimator, the number of stochastic histories used to compute the MCSA
-correction was varied from 25 to 50,000 with the number of MCSA
-iterations required to converge to a tolerance of \sn{1}{-8} recorded
-for a single eigenvalue iteration along with the CPU time required to
-achieve convergence.
-
-Figure~\ref{fig:spn_estimator_iters} gives the iteration count results
-and Figure~\ref{fig:spn_estimator_time} gives the timing
-results. Remarkably, although the fuel assembly problem is
-significantly more complicated than the simple Poisson problem studied
-in \S~\ref{sec:parameter_estimator_analysis}, the results here are
-largely the same with Figure~\ref{fig:spn_estimator_iters} and
-Figure~\ref{fig:estimator_nh_iters} having identical qualitative
-behavior. Using the expected value estimator with both fixed point
-iterations again creates a relative insensitivity to the number of
-histories used to compute the correction. Marginally better iterative
-and timing performance was observed for the expected value estimator
-when using the Richardson iteration. Better timing performance is
-expected as computing the extrapolation parameter in the minimal
-residual iteration requires several additional parallel
-operations.
-
-Compared to the collision estimator, the minimal residual iteration
-permits fewer stochastic histories to be used with each MCSA iteration
-while still maintaining good iterative performance. With respect to
-CPU time, there is a more pronounced effect due to the larger density
-of states created by the explicit ILUT preconditioning. Using a
-reduced domain fill level of 100 creates a situation where the
-expected value estimator is significantly more expensive per history
-to compute that the collision estimator due to the coupling of states.
-For the collision estimator, using around 20,000 stochastic histories
-at every iteration gave the best iterative performance while 500
-histories per iteration gave the best iterative performance when using
-the expected value estimator. For purely serial computational
-performance, there is not a distinguishing feature for the fuel
-assembly problem presented that would cause one to choose one of the
-estimators and fixed point iterations over the other.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=5in]{chapters/spn_equations/estimator_iters.pdf}
-  \end{center}
-  \caption{\textbf{MCSA iterations required to converge the fuel
-      assembly problem to a tolerance of \sn{1}{-8}.} \textit{Both the
-      collision and expected value estimators were used with the
-      adjoint Monte Carlo solver to compute the correction at each
-      iteration. Each estimator was used with the Richardson and
-      minimal residual fixed point iterations within an MCSA
-      iteration.}}
-  \label{fig:spn_estimator_iters}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=5in]{chapters/spn_equations/estimator_time.pdf}
-  \end{center}
-  \caption{\textbf{Total CPU time in seconds required to converge the
-      fuel assembly problem to a tolerance of \sn{1}{-8}.}
-    \textit{Both the collision and expected value estimators were used
-      with the adjoint Monte Carlo solver to compute the correction at
-      each iteration. Each estimator was used with the Richardson and
-      minimal residual fixed point iterations within an MCSA
-      iteration.}}
-  \label{fig:spn_estimator_time}
-\end{figure}
-
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{MCSA Verification\ }
-\label{sec:spn_mcsa_verification}
-Through the numerical studies presented in this chapter, we have
-developed a set of preconditioning techniques that permit MCSA to be
-used with the $SP_N$ form of the neutron transport equation and
-applied to a difficult fuel assembly criticality problem. Along with
-these preconditioning techniques, a set a varying MCSA parameters were
-analyzed to find those that yielded the best performance for this
-particular problem. In addition, several steps were taken to mitigate
-the dense composite linear operators that arise when performing
-explicit MCSA preconditioning necessary for achieve convergence. Using
-these results, we now compare MCSA to conventional Krylov solvers that
-would be typically used to solve the $SP_N$ form of the transport
-equation as presented here.
-
-MCSA solutions to the fuel assembly problem with 1, 2, and 4 energy
-groups using $SP_1$ discretization will computed using both BiCGStab
-and GMRES implementations from the Trilinos scientific computing
-library Aztec \cite{heroux_overview_2005}. The neutron fluxes, the
-number of eigenvalue iterations required to converge the problem, and
-the eigenvalue itself will be compared as a means of verification. For
-the flux values, we will look at the 2-norm and infinity norm of the
-relative difference vector between the reference case and the
-comparison case in each energy group:
-\begin{equation}
-  \text{Relative Group Flux Difference} = \frac{\ve{\Phi}_{ref}^g -
-    \ve{\Phi}^g}{\ve{\Phi}_{ref}^g}\:,
-  \label{eq:relative_difference_vector}
-\end{equation}
-where the addition and subtraction operations are performed
-element-wise to construct the vector. For this work, the BiCGStab
-results will be taken as the reference case.
-
-All solvers will be preconditioned with ILUT using a drop tolerance of
-\sn{1}{-5} and a fill level of 5. For GMRES, no restrictions were
-placed on the size of the subspace and therefore no restarts
-occurred. When the collision estimator was used with MCSA, \sn{2}{4}
-stochastic histories were used to compute the correction for every
-energy group in the problem (\sn{4}{4} and \sn{8}{4} histories total
-for the 2 and 4 group calculations respectively) corresponding to
-approximately 1 stochastic history per DOF. When the expected value
-estimator was used, \sn{5}{2} stochastic histories used for each
-energy group (\sn{1}{3} and \sn{1.5}{3} histories total for the 2 and
-4 group calculations) giving approximately 1 stochastic history for
-every 5 DOFs. The reduced domain approximation was also applied in
-conjunction with the ILUT preconditioning using a fill level of 100
-and a threshold value of \sn{1}{-10} to reduce the density of states
-in the Monte Carlo problem. For relaxation parameters, all MCSA
-computations used a Neumann relaxation parameter of 0.7 while a
-Richardson relaxation of 1.1 was used with the collision estimator and
-1.0 with the expected value estimator as determined by the previous
-analysis. Table~\ref{tab:spn_solver_defs} gives definitions for the
-solvers used to generate the results in the remainder of this
-section. For the energy group structures,
-Table~\ref{tab:spn_group_structure} gives the lower bounds of each
-group in eV (an implicit upper bound of \sn{2}{6} eV is assumed for
-group 0).
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{ll}\hline\hline
-      \multicolumn{1}{l}{\textbf{Name}} & 
-      \multicolumn{1}{l}{\textbf{Definition}} \\
-      BiCGStab-ILUT & BiCGStab preconditioned with ILUT \\
-      GMRES-ILUT & GMRES preconditioned with ILUT \\
-      MCSA-ILUT-R-C & MCSA preconditioned with ILUT using Richardson \\ 
-                    & fixed point iteration and collision estimator \\
-      MCSA-ILUT-MR-C & MCSA preconditioned with ILUT using minimal residual \\
-                     & fixed point iteration and collision estimator \\
-      MCSA-ILUT-R-EV & MCSA preconditioned with ILUT using Richardson \\
-                     & fixed point iteration and expected value estimator \\
-      MCSA-ILUT-MR-EV & MCSA preconditioned with ILUT using minimal residual \\
-                      & fixed point iteration and expected value estimator \\
-      Richardson-ILUT & Richardson iteration preconditioned with ILUT \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Solver definitions used for MCSA verification and
-      performance analysis.} \textit{The ILUT preconditioner
-      parameters were identical for all calculations and solvers.}}
-  \label{tab:spn_solver_defs}
-\end{table}
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{cl}\hline\hline
-      \multicolumn{1}{c}{\textbf{Number of Groups}} & 
-      \multicolumn{1}{l}{\textbf{Lower Bounds (eV)}} \\
-      1 & \{ \sn{1}{-5} \} \\
-      2 & \{ \sn{1}{-1}, \sn{1}{-5} \} \\
-      4 & \{ \sn{1}{1}, \sn{1}{0}, \sn{1}{-1}, \sn{1}{-5} \} \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Energy group lower bounds for the multigroup fuel
-      assembly criticality problem in electron volts.} \textit{An
-      implicit upper bound of \sn{2}{6} eV is assumed for group zero.}}
-  \label{tab:spn_group_structure}
-\end{table}
-
-For every combination of energy group structure and solver, the
-eigenvalue problem was converged with an eigensolver tolerance of
-\sn{1}{-6} and a linear solver tolerance of \sn{1}{-8} in a serial
-calculation. Table~\ref{tab:serial_ev_results} gives the eigenvalue
-computed by each solver for each problem as well as the number of
-eigenvalue iterations required to converge. For every problem, each
-solver converged to the same eigenvalue up to the number of digits
-reported by the eigensolver in the same number of eigenvalue
-iterations.
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lccc}\hline\hline
-      \multicolumn{1}{c}{\textbf{Solver}} & 
-      \multicolumn{1}{c}{\textbf{Number of Groups}} & 
-      \multicolumn{1}{c}{\textbf{Iterations}} &
-      \multicolumn{1}{c}{\textbf{Eigenvalue}} \\
-      \hline
-      BiCGStab-ILUT & 1 & 25 & 1.105987 \\
-      GMRES-ILUT & 1 & 25 & 1.105987 \\
-      MCSA-ILUT-R-C & 1 & 25 & 1.105987 \\
-      MCSA-ILUT-MR-C & 1 & 25 & 1.105987 \\
-      MCSA-ILUT-R-EV & 1 & 25 & 1.105987 \\
-      MCSA-ILUT-MR-EV & 1 & 25 & 1.105987 \\
-      Richardson-ILUT & 1 & 25 & 1.105987 \\
-      \hline
-      BiCGStab-ILUT & 2 & 35 & 1.155239 \\
-      GMRES-ILUT & 2 & 35 & 1.155239 \\
-      MCSA-ILUT-R-C & 2 & 35 & 1.155239 \\
-      MCSA-ILUT-MR-C & 2 & 35 & 1.155239 \\
-      MCSA-ILUT-R-EV & 2 & 35 & 1.155239 \\
-      MCSA-ILUT-MR-EV & 2 & 35 & 1.155239 \\
-      Richardson-ILUT & 2 & 35 & 1.155239 \\
-      \hline
-      BiCGStab-ILUT & 4 & 31 & 1.175850 \\
-      GMRES-ILUT & 4 & 31 & 1.175850 \\
-      MCSA-ILUT-R-C & 4 & 31 & 1.175850 \\
-      MCSA-ILUT-MR-C & 4 & 31 & 1.175850 \\
-      MCSA-ILUT-R-EV & 4 & 31 & 1.175850 \\
-      MCSA-ILUT-MR-EV & 4 & 31 & 1.175850 \\
-      Richardson-ILUT & 4 & 31 & 1.175850 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Serial eigensolver verification results for the
-      fuel assembly problem.} \textit{Each calculation was converged
-      with the same eigensolver and linear solver
-      tolerances. Table~\ref{tab:spn_solver_defs} gives the
-      description for each solver type presented.}}
-  \label{tab:serial_ev_results}
-\end{table}
-
-Table~\ref{tab:serial_differences_g1},
-Table~\ref{tab:serial_differences_g2}, and
-Table~\ref{tab:serial_differences_g4} give the infinity norms and
-2-norms of the flux relative difference vectors given by
-Eq~(\ref{eq:relative_difference_vector}) for all calculations and all
-energy groups. In general, we see good agreement in the 2-norm between
-the MCSA variants and the GMRES calculation, meaning that they have
-approximately the same variation from the BiCGStab calculation. We do
-note larger infinity norm values of this vector for some of the MCSA
-variants relative to the GMRES results (i.e. group 1 in the 2-group
-calculation), however, this large maximum value is apparently a local
-phenomenon with good agreement in the 2-norm for the
-calculations. Based on the good agreement between MCSA and GMRES for
-the relative difference of the group fluxes and the eigenvalue results
-presented here, the MCSA implementation should be deemed correct for
-the fuel assembly problem for serial calculations with multiple energy
-groups.
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lccc}\hline\hline
-      \multicolumn{1}{c}{\textbf{Solver}} & 
-      \multicolumn{1}{c}{\textbf{Group}} &
-      \multicolumn{1}{c}{\textbf{$|| \cdot ||_{inf}$}} &
-      \multicolumn{1}{c}{\textbf{$|| \cdot ||_2$}} \\
-      \hline
-      GMRES-ILUT & 0 & \sn{4.562}{-6} & \sn{6.746}{-5} \\
-      MCSA-ILUT-R-C & 0 & \sn{2.099}{-5} & \sn{2.541}{-4} \\
-      MCSA-ILUT-MR-C & 0 & \sn{4.812}{-5} & \sn{4.672}{-4} \\
-      MCSA-ILUT-R-EV & 0 & \sn{2.371}{-5} & \sn{2.740}{-4} \\
-      MCSA-ILUT-MR-EV & 0 & \sn{3.605}{-5} & \sn{2.554}{-4} \\
-      Richardson-ILUT & 0 & \sn{1.531}{-5} & \sn{2.771}{-4} \\
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Serial scalar flux verification results for the
-      1-group fuel assembly problem.} \textit{The infinity norm and
-      2-norm of the relative flux vector in each group is given with
-      BiCGStab used as the reference. Each calculation was converged
-      with the same eigensolver and linear solver
-      tolerances. Table~\ref{tab:spn_solver_defs} gives the
-      description for each solver type presented.}}
-  \label{tab:serial_differences_g1}
-\end{table}
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lccc}\hline\hline
-      \multicolumn{1}{c}{\textbf{Solver}} & 
-      \multicolumn{1}{c}{\textbf{Group}} &
-      \multicolumn{1}{c}{\textbf{$|| \cdot ||_{inf}$}} &
-      \multicolumn{1}{c}{\textbf{$|| \cdot ||_2$}} \\
-      \hline
-      GMRES-ILUT & 0 & \sn{7.590}{-6} & \sn{1.234}{-4} \\
-      MCSA-ILUT-R-C & 0 & \sn{5.308}{-6} & \sn{1.066}{-4} \\
-      MCSA-ILUT-MR-C & 0 & \sn{6.388}{-5} & \sn{2.233}{-4} \\
-      MCSA-ILUT-R-EV & 0 & \sn{4.197}{-5} & \sn{2.168}{-4} \\
-      MCSA-ILUT-MR-EV & 0 & \sn{2.736}{-5} & \sn{8.160}{-5} \\
-      Richardson-ILUT & 0 & \sn{3.976}{-6} & \sn{7.059}{-5} \\
-      \hline
-      GMRES-ILUT & 1 & \sn{7.579}{-6} & \sn{1.246}{-4} \\
-      MCSA-ILUT-R-C & 1 & \sn{1.645}{-5} & \sn{1.210}{-4} \\
-      MCSA-ILUT-MR-C & 1 & \sn{6.635}{-4} & \sn{9.435}{-4} \\
-      MCSA-ILUT-R-EV & 1 & \sn{1.797}{-4} & \sn{2.983}{-4} \\
-      MCSA-ILUT-MR-EV & 1 & \sn{1.556}{-4} & \sn{2.055}{-4} \\
-      Richardson-ILUT & 1 & \sn{4.031}{-6} & \sn{7.110}{-5} \\
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Serial scalar flux verification results for the
-      2-group fuel assembly problem.} \textit{The infinity norm and
-      2-norm of the relative flux vector in each group is given with
-      BiCGStab used as the reference. Each calculation was converged
-      with the same eigensolver and linear solver
-      tolerances. Table~\ref{tab:spn_solver_defs} gives the
-      description for each solver type presented.}}
-  \label{tab:serial_differences_g2}
-\end{table}
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lccc}\hline\hline
-      \multicolumn{1}{c}{\textbf{Solver}} & 
-      \multicolumn{1}{c}{\textbf{Group}} &
-      \multicolumn{1}{c}{\textbf{$|| \cdot ||_{inf}$}} &
-      \multicolumn{1}{c}{\textbf{$|| \cdot ||_2$}} \\
-      \hline
-      GMRES-ILUT & 0 & \sn{2.376}{-5} & \sn{4.043}{-4} \\
-      MCSA-ILUT-R-C & 0 & \sn{2.639}{-5} & \sn{3.950}{-4} \\
-      MCSA-ILUT-MR-C & 0 & \sn{3.958}{-5} & \sn{4.635}{-4} \\
-      MCSA-ILUT-R-EV & 0 & \sn{2.749}{-5} & \sn{3.975}{-4} \\
-      MCSA-ILUT-MR-EV & 0 & \sn{2.969}{-5} & \sn{4.054}{-4} \\
-      Richardson-ILUT & 0 & \sn{2.189}{-5} & \sn{3.791}{-4} \\
-      \hline
-      GMRES-ILUT & 1 & \sn{2.271}{-5} & \sn{3.952}{-4} \\
-      MCSA-ILUT-R-C & 1 & \sn{3.701}{-5} & \sn{3.933}{-4} \\
-      MCSA-ILUT-MR-C & 1 & \sn{3.803}{-5} & \sn{3.933}{-4} \\
-      MCSA-ILUT-R-EV & 1 & \sn{4.142}{-5} & \sn{4.011}{-4} \\
-      MCSA-ILUT-MR-EV & 1 & \sn{2.723}{-5} & \sn{4.030}{-4} \\
-      Richardson-ILUT & 1 & \sn{2.137}{-5} & \sn{3.768}{-4} \\
-      \hline
-      GMRES-ILUT & 2 & \sn{2.310}{-5} & \sn{4.021}{-4} \\
-      MCSA-ILUT-R-C & 2 & \sn{2.490}{-5} & \sn{3.956}{-4} \\
-      MCSA-ILUT-MR-C & 2 & \sn{2.214}{-5} & \sn{3.608}{-4} \\
-      MCSA-ILUT-R-EV & 2 & \sn{4.456}{-5} & \sn{3.833}{-4} \\
-      MCSA-ILUT-MR-EV & 2 & \sn{3.366}{-5} & \sn{3.797}{-4} \\
-      Richardson-ILUT & 2 & \sn{2.143}{-5} & \sn{3.767}{-4} \\
-      \hline
-      GMRES-ILUT & 3 & \sn{2.335}{-5} & \sn{4.029}{-4} \\
-      MCSA-ILUT-R-C & 3 & \sn{3.045}{-5} & \sn{3.963}{-4} \\
-      MCSA-ILUT-MR-C & 3 & \sn{2.696}{-5} & \sn{3.606}{-4} \\
-      MCSA-ILUT-R-EV & 3 & \sn{1.947}{-5} & \sn{4.283}{-4} \\
-      MCSA-ILUT-MR-EV & 3 & \sn{3.295}{-5} & \sn{3.837}{-4} \\
-      Richardson-ILUT & 3 & \sn{2.161}{-5} & \sn{3.772}{-4} \\
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Serial scalar flux verification results for the
-      4-group fuel assembly problem.} \textit{The infinity norm and
-      2-norm of the relative flux vector in each group is given with
-      BiCGStab used as the reference. Each calculation was converged
-      with the same eigensolver and linear solver
-      tolerances. Table~\ref{tab:spn_solver_defs} gives the
-      description for each solver type presented.}}
-  \label{tab:serial_differences_g4}
-\end{table}
-
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{MCSA Performance Comparison to Conventional Methods\ }
-\label{sec:spn_comparison}
-With the same set of solver parameters used for verification
-caluclation, MCSA will now be compared to the same Krylov solvers in
-terms of both iterative performance and CPU timing. A Richardson
-iteration solution will also be used in order to assess the
-acceleration provided by the residual Monte Carlo component of MCSA.
-To compare iterative performance, the number of linear solver
-iterations required to converge each eigenvalue iteration was recorded
-and then averaged over all eigenvalue iterations for each
-solver.
-
-Table~\ref{tab:spn_comparison_iterations} gives the average number of
-linear solver iterations required to converge a single eigenvalue
-iteration of the fuel assembly problem as a function of the number of
-energy groups. In general, the iterative performance of all the
-solvers tested aside from the Richardson iteration is comparable with
-BiCGStab performing the best. We expect this because not only are the
-multigroup $SP_N$ equations positive-definite, but they are also
-block-symmetric and therefore we expect the conjugate gradient-based
-methods to perform well due to the structure of the projection
-scheme. It important to note that the iterative performance of each
-solver is not a strong function of the number of energy groups in the
-problem (and therefore the problem size). Even more important,
-although BiCGStab produced the best iterative results, MCSA did
-perform better than GMRES in many cases with the Richardson iteration
-clearly accelerated by the Monte Carlo solve. When compared to the
-preconditioned Richardson iteration, the acceleration provided by MCSA
-is clear with 3-4 times fewer iterations required for convergence.
-
-It is also important to note the reduced domain approximation
-parameters, particularly the fill level, were fixed as the number of
-energy groups was increased. By fixing the fill level, we are
-effectively fixing the amount of information contained in the
-composite linear operator available for the Monte Carlo
-problem. Limiting this information to 100 entries per row in the
-preconditioned system did not appear to have a significant effect on
-the iterative performance of the solver. For larger numbers of energy
-groups (and therefore DOFs) this may be an issue and this number may
-need to be increased to maintain iterative performance. Currently,
-memory restrictions arising from the explicit preconditioning strategy
-prevent larger numbers of energy groups to be analyzed using an
-appropriate preconditioner for the fuel assembly problem.
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lccc}\hline\hline
-      \multicolumn{1}{l}{Solver}&
-      \multicolumn{1}{c}{1 Group}&
-      \multicolumn{1}{c}{2 Groups}&
-      \multicolumn{1}{c}{4 Groups}\\
-      \hline
-      BiCGStab-ILUT & 11.6 & 11.6 & 12.4 \\ 
-      GMRES-ILUT & 18.1 & 17.9 & 18.9 \\
-      MCSA-ILUT-R-C & 14.6 & 15.4 & 17.6 \\
-      MCSA-ILUT-MR-C & 16.0 & 17.1 & 23.7 \\
-      MCSA-ILUT-R-EV & 18.3 & 19.4 & 16.8 \\
-      MCSA-ILUT-MR-EV & 19.6 & 22.4 & 17.5 \\
-      Richardson-ILUT & 60.9 & 60.4 & 63.4 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Average number of linear solver iterations
-      iterations required to converge the fixed source $SP_N$ problem
-      at each eigenvalue iteration for the fuel assembly problem for
-      1, 2 and 4 energy groups.} \textit{Values are rounded to the
-      first decimal place. Table~\ref{tab:spn_solver_defs} gives the
-      description for each solver type presented.}}
-  \label{tab:spn_comparison_iterations}
-\end{table}
-
-Although iterative performance for MCSA was comparable to that
-observed for Krylov implementations using exactly the same
-preconditioning scheme, timing performance is not as
-competitive. Figure~\ref{fig:spn_comparison_time} gives the average
-CPU time per linear solver iteration required to converge the fuel
-assembly problem with all linear solver iterations over all eigenvalue
-iterations used to compute the average. The Krylov solver
-implementations are approximately an order of magnitude faster than
-the MCSA implementations presented here. We expect these results for
-several reasons. First, even when the reduced domain approximation is
-used there are $O(100)$ elements in each row of the system and each of
-those elements must be processed during a transition in the Monte
-Carlo random walk sequence. In addition, although the preconditioned
-fixed point iterations used in the MCSA sequence do not use the
-composite linear operator but rather a sequence of matrix vector
-multiplies to achieve the same preconditioning effect, they do use the
-explicitly extracted inverse of the preconditioning matrices which
-themselves are dense, leading to exceedingly slow computation times
-even in the fixed point iteration.
-
-For the Krylov methods, the ILUT preconditioners are applied to a
-vector with two triangular solves, one for each triangular
-factor. Also, the original linear operator has $O(10)$ elements in
-each row of the system, an order of magnitude less than that used in
-the reduced domain composite operator. Second, the implementation
-presented here has not been optimized and it is likely that several
-components of the algorithm can be implemented in a more desirable
-time complexity. However, it is most important to note here that the
-CPU timing data presented in Figure~\ref{fig:spn_comparison_time}
-shows that all methods have the same time complexity; their
-differences are merely the manifestation of a large time constant due
-to the effects of the explicit preconditioning scheme. For transport
-systems where this is not a problem, the literature explicitly shows
-that MCSA can be competitive with these production Krylov methods
-using CPU time as a measure \cite{evans_monte_2012}.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=5in]{chapters/spn_equations/solver_time.pdf}
-  \end{center}
-  \caption{\textbf{Average CPU time per linear solver iteration in
-      seconds for the fuel assembly problem as a function of energy
-      groups.}  \textit{All linear solver iterations over all
-      eigenvalue iterations were used to compute the
-      average. Table~\ref{tab:spn_solver_defs} gives the description
-      for each solver type presented in the legend with the Krylov
-      solvers represented in black and the MCSA solvers represented in
-      red.}}
-  \label{fig:spn_comparison_time}
-\end{figure}
-
-Finally, not considered in Figure~\ref{fig:spn_comparison_time} is the
-time required to actually form the explicit inverse preconditioner
-matrices and the composite operator through matrix-matrix
-multiplication. The timing numbers reported were simply to perform the
-MCSA iteration procedure with the composite operator already
-formed. Figure~\ref{fig:spn_comparison_prec_time} additionally
-presents the CPU times for MCSA convergence with the time to form the
-inverse of the preconditioners and the composite linear operator
-through matrix-matrix multiplication amortized over all iterations. As
-is readily observed, including the costs of these operations increases
-the MCSA computation time by another order of magnitude.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=5in]{chapters/spn_equations/solver_p_time.pdf}
-  \end{center}
-  \caption{\textbf{Average CPU time per iteration in seconds for the
-      fuel assembly problem as a function of energy groups with
-      preconditioning time included for the MCSA methods.}
-    \textit{All linear solver iterations over all eigenvalue
-      iterations were used to compute the
-      average. Table~\ref{tab:spn_solver_defs} gives the description
-      for each solver type presented in the legend. Data labeled
-      starting with PMCSA and shown in blue is identical to those
-      labeled with MCSA except that they additionally include the cost
-      of generating the inverse of the preconditioners and composite
-      linear operator.}}
-  \label{fig:spn_comparison_prec_time}
-\end{figure}
-
-Based on the performance results in this section, MCSA shows promise
-as a competitive and perhaps even superior method for solutions to the
-neutron transport problem discretized with the $SP_N$
-approximation. Not only are the correct answers produced when compared
-to production linear solvers, but the iterative performance is
-comparable to Krylov methods with identical preconditioning that would
-typically be used every-day calculations. From a CPU timing
-perspective, the methods yield the same time complexity as a function
-of the number of energy groups in the problem when compared to the
-Krylov methods. Here, a large time constant is generated due to the
-explicit preconditioning strategy and the generation of dense linear
-operators as a result. To improve these results and put general MCSA
-schemes into a performance regime where they are competitive with
-Krylov methods for neutron transport problems, significant research
-will be required to improve upon the explicit preconditioning scheme
-presented here.
-
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{Summary\ }
-\label{sec:mc_summary}
-
-In this chapter, the Monte Carlo Synthetic Acceleration method has
-been applied to the $SP_N$ discretization of the neutron transport
-equation and explored in the context of a difficult nuclear fuel
-assembly criticality problem. The following are the significant
-observations and findings.
-
-\begin{itemize}
-\item MCSA can solve the asymmetric system generated by the $SP_N$
-  equations
-\item MCSA has been incorporated into the Exnihilo neutronics
-  production code base developed at Oak Ridge National Laboratory
-\item Simple Jacobi-based preconditioning reduces the spectral radius
-  of the $SP_N$ system below unity, however, this alone is not
-  sufficient for MCSA convergence
-\item As the spectral radius approaches unity, MCSA was demonstrated
-  to break down with more stochastic histories required for
-  convergence and more CPU time required per history
-\item Light water reactor problems are difficult to solve with MCSA as
-  they have large spectral radii due to the neutron scattering in the
-  moderator
-\item Advanced algebraic preconditioning strategies were applied to
-  the $SP_N$ equations to obtain convergence with ILUT chosen for
-  subsequent investigations
-\item The reduced domain approximation was required to mitigate the
-  memory and performance constraints of the explicit preconditioning
-  strategy although both remain significant obstacles
-\item MCSA relaxation parameters were developed and observed to
-  enhance the time to solution for the fuel assembly criticality
-  problem
-\item MCSA was verified to produce the same k-eigenvalue for the fuel
-  assembly as production Krylov methods and gave good relative flux
-  agreement
-\item MCSA was observed to converge in fewer iterations per eigenvalue
-  iteration than GMRES for the fuel assembly criticality problem and
-  more than Bi-CGStab using the same preconditioning
-\item The explicit preconditioning strategy required to overcome the
-  MCSA spectral radius restriction creates runtimes $O(100)$ larger
-  than the production Krylov methods for the fuel assembly criticality
-  problem
-\end{itemize}

### 2013-08-22 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
moving parallel metrics to an appendix
moving mcsa relaxation parameters to an appendix

@@
 at some level of approximation $N$ such that $n = 0,1,\dotsc,N$. This
-yields a set of $N+1$ equations for $N+2$ flux moments. In multiple
-dimensions, the equation set becomes large and coupled not only
-through angular moments but also through the spatial variables. As a
-simpler alternative to multidimensional $P_N$ solutions, Gelbard
-recognized in 1960 that the planar $P_N$ equations could be simplified
-and applied an ad-hoc method to extend them to multiple dimensions,
-yielding the $SP_N$ equations. These equations are not only fewer in
-number, but also take on a diffusion-like form while maintaining the
-angular character of the flux, making them amenable to solutions with
-modern diffusion methods.
+yields a set of $N+1$ equations for $N+2$ flux moments with the
+highest order moment set to zero for closure. In multiple dimensions,
+the equation set becomes large and coupled not only through angular
+moments but also through the spatial variables. As a simpler
+alternative to multidimensional $P_N$ solutions, Gelbard recognized in
+1960 that the planar $P_N$ equations could be simplified and applied
+an ad-hoc method to extend them to multiple dimensions, yielding the
+$SP_N$ equations. These equations are not only fewer in number, but
+also take on a diffusion-like form while maintaining the angular
+character of the flux, making them amenable to solutions with modern
+diffusion methods.
 
@@
 \end{multline}
-Immediately, we note the diffusion-like nature of
-Eq~(\ref{eq:reduced_pn}) as compared to the original $P_N$
-equations. To extend these equations to multiple dimensions, Gelbard
-simply replaced the planar spatial derivatives in the reduced set of
+To extend these equations to multiple dimensions, Gelbard simply
+replaced the planar spatial derivatives in the reduced set of
 equations with general multidimensional gradient operators:
@@
 \end{equation}
-It should be noted here that with respect to a general MCSA scheme,
-the introduction of fission in the system will only affect the source
-vector in the linear system for the eigenvalue schemes used in this
-work. The linear operator, and therefore overall MCSA performance will
-be dictated by streaming and scattering as defined on the left-hand
-side.
+To complete the spatial discretization, the gradient operators are
+resolved by a finite volume scheme as in
+\cite{evans_simplified_2013}. It should be noted here that with
+respect to a general MCSA scheme, the introduction of fission in the
+system will only affect the source vector in the linear system for the
+eigenvalue schemes used in this work. The linear operator, and
+therefore overall MCSA performance will be dictated by streaming and
+scattering as defined on the left-hand side.
 
@@
 Previous work in applying MCSA to the thermal radiation diffusion
-equations had success in using Jacobi preconditioning
-\cite{evans_monte_2012}. In this section, we will compute the spectral
-radius of the iteration matrix formed by the $SP_N$ equations while
-parametrically varying the $SP_N$ order, the $P_N$ order, the number
-of energy groups, and upscattering/downscattering between energy
-groups for Jacobi preconditioned problems to ensure that this type of
-preconditioning will be applicable to the fuel assembly problem. For
-each problem in the parameter study, the values given in
+equation had success in using Jacobi preconditioning
+\cite{evans_monte_2012} and therefore we apply it here to the $SP_N$
+equations given their diffusion-like form. In this section, we will
+compute the spectral radius of the iteration matrix formed by the
+$SP_N$ equations while parametrically varying the $SP_N$ order, the
+$P_N$ order, the number of energy groups, and
+upscattering/downscattering between energy groups for Jacobi
+preconditioned problems to ensure that this type of preconditioning
+will be applicable to the fuel assembly problem. For each problem in
+the parameter study, the values given in
 Table~\ref{tab:spn_fixed_parameters} were fixed for each spatial
@@
 by the discretization.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=3.25in]{chapters/spn_equations/group1.png}
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=3.3in]{chapters/spn_equations/group1.png}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=3.25in]{chapters/spn_equations/group10ds.png}
+    \includegraphics[width=3.3in]{chapters/spn_equations/group10ds.png}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=3.25in]{chapters/spn_equations/group10us.png}
+    \includegraphics[width=3.3in]{chapters/spn_equations/group10us.png}
   \end{center}
@@
 \end{figure}
+
 We note a few key features of the sparsity plots. The first is that
@@
 diagonal character of these systems, although the blocks from
-Eq~(\ref{eq:A_block_matrix}) are readily apparent.
+Eq~(\ref{eq:A_block_matrix}) are readily apparent with a block size of
+$N_g\times(N+1)/2$.
 
@@
 \textit{block Jacobi} preconditioning as a left preconditioner for the
-system. Block Jacobi preconditioning extracts the diagonal elements of
-the matrix as the preconditioner where now the elements extracted are
-the blocks on the diagonal as shown on the left side of
-Figure~\ref{fig:block_jacobi_ex}. Shown on the right side of
-Figure~\ref{fig:block_jacobi_ex}, inversion of the preconditioner is
-trivial with each diagonal block inverted separately. For the $SP_N$
-equations, Eq~(\ref{eq:A_block_matrix}) gives a block size of
-$N_g\times(N+1)/2$.
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.5}{
+system. Block Jacobi preconditioning extracts the diagonal block
+elements of the matrix to use as the preconditioner as shown on the
+left side of Figure~\ref{fig:block_jacobi_ex}. Shown on the right side
+of Figure~\ref{fig:block_jacobi_ex}, inversion of the preconditioner
+is trivial with each diagonal block inverted separately.
+
+\begin{figure}[t!]
+  \begin{center}
+    \scalebox{1.6}{
     \input{chapters/spn_equations/block_jacobi.pdftex_t} }
@@
 
-For high performance implementations this has several attractive
-properties. First, the blocks in the matrix come from the energy/angle
-discretization of the transport equation as given by
-Eq~(\ref{eq:A_block_matrix}). Each block on the diagonal is bound to a
-mesh element in the system (note there are 8 blocks on the diagonal in
-each of the sparsity patterns with a mesh of $2 \times 2 \times 2$)
-and therefore we expect the matrix elements forming the block to be
-entirely local in domain decomposed calculations. Second, these blocks
-are typically dense and nearly lower triangular for many transport
-problems meaning that established dense matrix methods can be used for
-fast inversion.
+For high performance implementations block Jacobi preconditioning has
+several attractive properties. First, the blocks in the transport
+operator come from the energy/angle discretization of the transport
+equation as given by Eq~(\ref{eq:A_block_matrix}). Each block on the
+diagonal is bound to a mesh element in the system (note there are 8
+blocks on the diagonal in each of the sparsity patterns with a mesh of
+$2 \times 2 \times 2$) and therefore we expect the matrix elements
+forming the block to be entirely local in domain decomposed
+calculations. Second, these blocks are typically dense and nearly
+lower triangular for many transport problems meaning that established
+dense matrix methods can be used for fast inversion.
 
@@
 while a cross section of the materials configuration along the
-horizontal axis is given by Figure~\ref{fig:problem3_axial_mat}. A
-detailed view of the assembly bottom is given in
+horizontal axis is given by Figure~\ref{fig:problem3_axial_mat}. The
+control rods were removed for these problems leaving only moderator in
+their place. A detailed view of the assembly bottom is given in
 Figure~\ref{fig:problem3_end}. On the top and bottom of the assembly,
@@
       geometry. Gray regions are homogenized fuel and red regions are
-      homogenized control rods. The control rods were removed for
-      these problems. Each fuel pin is resolved by a $2 \times 2$
-      mesh.}}
+      homogenized control rods. Each fuel pin is resolved by a $2
+      \times 2$ mesh.}}
   \label{fig:problem3_radial_mat}
@@
       structures on the top and bottom of the assembly. Lighter purple
-      material in the center of the assembly is moderator and darker
-      purple/red material is fuel.}}
+      material in the center of the assembly is homogenized
+      moderator/control rodes and darker purple/red material is
+      homogenized moderator/fuel.}}
   \label{fig:problem3_axial_mat}
@@
       has been included. Lighter purple material on the right of the
-      figure is moderator and darker purple/red material is fuel.}}
+      figure is homogenized moderator/control rods and darker
+      purple/red material is homogenized moderator/fuel.}}
   \label{fig:problem3_end}
@@
 left-hand side of the $SP_N$ discretization, $\mathbf{F}$ is the
-fission matrix, and $\mathbf{\Phi}$ the multigroup neutron flux. This
-problem is significantly more complicated than the simple test problem
-used for the previous spectral analysis. Fission has been introduced
-into the set of equations and the addition of moderator into the
-system will increase the amount of scattering, creating a
-significantly more difficult problem manifesting itself in an
-iteration matrix with a larger spectral radius. When using MCSA, the
-linear operator applied to $\mathbf{\Phi}^{n+1}$ at each eigenvalue
-iteration will dictate convergence and remain unchanged throughout the
-eigenvalue computation while the addition of fission to the system
-will only modify the source of neutrons and the multiplication factor
-at every eigenvalue iteration while not affecting Monte Carlo
-transport.
+fission matrix, and $\mathbf{\Phi}$ the multigroup neutron flux. Using
+MCSA in this context is significantly more complicated than the simple
+test problem used for the previous fixed-source preconditioner
+analysis. Fission has been introduced into the set of equations and
+the addition of moderator into the system will increase the amount of
+scattering, creating a significantly more difficult problem
+manifesting itself in an iteration matrix with a larger spectral
+radius. When using MCSA, the linear operator applied to
+$\mathbf{\Phi}^{n+1}$ at each eigenvalue iteration will dictate
+convergence and remain unchanged throughout the eigenvalue computation
+while the addition of fission to the system will only modify the
+source of neutrons and the multiplication factor at every eigenvalue
+iteration while not affecting Monte Carlo transport.
 
@@
 \sn{1}{6} histories per iteration exhibited the same divergent
-behavior at a significant computational cost and compute time. Even if
-the problem may be ill-conditioned, we do expect convergence of MCSA
-given that the spectral radius requirement was met. To investigate
-further, a block Jacobi preconditioned Richardson iteration was used
-to solve the same
+behavior at a significant computational cost and compute time. To
+investigate further, a block Jacobi preconditioned Richardson
+iteration was used to solve the same
 problem. Figure~\ref{fig:block_jacobi_res_richardson} gives the
@@
 Neumann-Ulam solve is the component of MCSA causing a divergent
-solution.
+solution which we study next.
 
@@
   \begin{center}
-    \includegraphics[width=4.25in]{chapters/spn_equations/breakdown_iterations.pdf}
+    \includegraphics[width=5.75in]{chapters/spn_equations/breakdown_iterations.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=4.25in]{chapters/spn_equations/breakdown_histories.pdf}
+    \includegraphics[width=5.75in]{chapters/spn_equations/breakdown_histories.pdf}
   \end{center}
@@
 iteration matrix spectral radii this large, using significant numbers
-of stochastic histories in the Monte Carlo solve are not enough to
+of stochastic histories in the Monte Carlo solve is not enough to
 reduce the stochastic uncertainty in the correction, causing the MCSA
@@
   \begin{center}
-    \includegraphics[width=4.25in]{chapters/spn_equations/breakdown_time.pdf}
+    \includegraphics[width=5.75in]{chapters/spn_equations/breakdown_time.pdf}
   \end{center}
@@
 and used within the explicit preconditioning framework given by
-Eq~(\ref{eq:left_right_mcsa}). Data
-showing their effects on MCSA solutions of the fuel assembly
-criticality problem will be presented.
+Eq~(\ref{eq:left_right_mcsa}). Data showing their effects on MCSA
+solutions of the fuel assembly criticality problem will be provided.
 
@@
 assembly problem or other problems with a high scattering to
-absorption ratio. Incomplete lower-upper (ILU) factorizations of the
-linear operator can be used a simple mechanism to form an approximate
-inverse of a preconditioner. To build the factorization, the sparse
-upper and lower triangular factors, $\mathbf{L}$ and $\mathbf{U}$, are
-computed such that the residual matrix formed by the factorization
+absorption ratio. As with MCSA, the algebraic formulation of ILUT
+requires the physics operator to be fully formed and is therefore
+easily incorporated into our solution sequence.
+
+Incomplete lower-upper (ILU) factorizations of the linear operator can
+be used as a simple mechanism to form an approximate inverse of a
+preconditioner. To build the factorization, the sparse upper and lower
+triangular factors, $\mathbf{L}$ and $\mathbf{U}$, are computed such
+that the residual matrix formed by the factorization
 \cite{saad_iterative_2003}:
@@
 lower and upper triangular factors may then be easily inverted by
-means of simple elimination to produce the preconditioner. For MCSA,
-$\mathbf{L}^{-1}$ will be used on the left and $\mathbf{U}^{-1}$ on
-the right to precondition the system.
-
-For modern subspace methods, only the action to the preconditioner on
-a vector is required for efficient implementations and therefore a
-triangular solve can be used for this purpose. For the explicit
-preconditioning scheme presented in Eq~(\ref{eq:left_right_mcsa}), the
-fully inverted operator must be generated in order to build the set of
-probabilities and weights required for Monte Carlo
-sampling. Therefore, for a linear system of size $N$, $N$ triangular
-solves will be required in order to extract the inverse matrices from
-production ILU implementations. In addition, parallel implementations
-typically generate lower and upper factors that are only triangular
-locally, providing an easily parallelizable mechanism to generate the
-action of the preconditioner inverse. A consequence of this choice for
-parallel scalability is a degradation of the preconditioning quality
-as the size of the parallel system is increased. For serial
-computations, the triangular factorization is potentially exact
-depending on the parameters chosen while at thousands of parallel
-tasks, the global triangular factors differ significantly from the true
-factorization. As a result, more iterations are required at higher
-levels of parallelism to converge the system and will ultimately
-degrade overall scalability of the system with respect to total wall
-time to converge.
+means of simple elimination procedure to produce the
+preconditioner. For MCSA, $\mathbf{L}^{-1}$ will be used on the left
+and $\mathbf{U}^{-1}$ on the right to precondition the system.
+
+For the explicit preconditioning scheme presented in
+Eq~(\ref{eq:left_right_mcsa}), the fully inverted operator must be
+generated in order to build the set of probabilities and weights
+required for Monte Carlo sampling. Therefore, for a linear system of
+size $N$, $N$ triangular solves will be required in order to extract
+the inverse matrices from production ILU implementations. In addition,
+parallel implementations typically generate lower and upper factors
+that are only triangular locally, providing an easily parallelizable
+mechanism to generate the action of the preconditioner inverse. A
+consequence of this choice for parallel scalability is a degradation
+of the preconditioning quality as the size of the parallel system is
+increased. For serial computations, the triangular factorization is
+potentially exact depending on the parameters chosen while at
+thousands of parallel tasks, the global triangular factors differ
+significantly from the true factorization. As a result, more
+iterations are required at higher levels of parallelism to converge
+the system and will ultimately degrade overall scalability of the
+system with respect to total wall time to converge.
 
@@
 expected, the higher the fill level chosen for the ILUT factorization,
-the fewer iterations are required to converge the problem (only 9 MCSA
-iterations were required to converge the problem for the fill level of
-3 and drop tolerance of \sn{1}{-5} case). For higher levels of fill,
-the iterations needed to converge were not as sensitive, signaling
-that a larger drop tolerance can perhaps be used without a significant
-degradation in iterative performance. At smaller fill levels, the
-sensitivity to the ILUT drop tolerance is more significant. For all
-fill levels used, convergence was not achieved for the fuel assembly
-problem with a drop tolerance larger than \sn{1}{-3}.
+the fewer iterations are required to converge the problem with only 9
+MCSA iterations were required to converge the problem for the fill
+level of 3 and drop tolerance of \sn{1}{-5} case. For higher levels of
+fill, the iterations needed to converge were not as sensitive,
+signaling that a larger drop tolerance can perhaps be used without a
+significant degradation in iterative performance. At smaller fill
+levels, the sensitivity to the ILUT drop tolerance is more
+significant. For all fill levels used, convergence was not achieved
+for the fuel assembly problem with a drop tolerance larger than
+\sn{1}{-3}.
 \begin{figure}[t!]
@@
 nearest-neighbor type algorithms.
+
 \begin{figure}[t!]
@@
 iterations required to converge as the non-zero row entries changed by
-a smaller fraction over drop tolerances tested.
+a smaller fraction over drop tolerances tested. For higher levels of
+fill, the quality metric is observed to be relatively insensitive to
+the drop tolerance as maintaining more entries in the preconditioner
+has a stronger impact than filtering small values.
+
 \begin{figure}[t!]
@@
 
-It should be noted here that although only the behavior of the linear
-solve during a single eigenvalue iteration is reported, the behavior
-of the linear solver was observed to be consistent throughout the
-eigenvalue iterations (25 total eigenvalue iterations were required to
-converge the 1-group $SP_1$ problem). We expect this as the linear
-operator (which is unchanging in the fuel assembly criticality
-problem) dictates the convergence of MCSA. At each iteration the
-fission source provided to the linear problem is changing, however, we
-expect the same convergence behavior for all source vectors
-$\mathbf{b} \in \mathbb{R}^{N}$ where $||\mathbf{b}||_1 \neq 0$.
-
 \subsection{Sparse Approximate Inverse Preconditioning}
 \label{subsec:spn_spainv_preconditioning}
-SPAINV is chosen next because it the true inverse of the
-preconditioner is product of the algorithm, avoiding the need for many
-inverse matrix applications. In addition, the algorithm seeks to
-maintain the sparsity of the inverse preconditioner, potentially
-alleviating some of the dense operator issues observed with
-ILUT. Explicit formation of the preconditioner inverse is a
-requirement of the current MCSA preconditioning strategy. As a result,
-although ILUT preconditioning provided excellent iterative
-performance, a significant time penalty is paid to extract the inverse
-of the triangular factors through $N$ triangular solves. In addition,
-this explicit inversion yielded a composite linear operator that was
-dense, requiring modification of the system in order to achieve better
-scalability and Monte Carlo performance.
-
-Sparse approximate inverse (SPAINV) preconditioning is a technique
-that may potentially alleviate both of these constraints at the cost
-of reduced iterative performance. The method produces a sparse
-approximation of the inverse matrix directly by minimizing the
-Frobenius norm of the residual matrix \cite{saad_iterative_2003}:
+Sparse approximate inverse (SPAINV) preconditioning is chosen next because the true inverse of the preconditioner
+is product of the algorithm, avoiding the need for many inverse matrix
+applications as required by using ILUT with MCSA. In addition, the
+algorithm seeks to maintain the sparsity of the inverse
+preconditioner, potentially alleviating some of the dense operator
+issues observed with ILUT.
+
+SPAINV produces a sparse approximation of the inverse matrix directly
+by minimizing the Frobenius norm of the residual matrix
+\cite{saad_iterative_2003}:
 \begin{equation}
@@
 MCSA preconditioned with SPAINV was used to solve the fuel assembly
-problem as with ILUT for a 1-group $SP_1$ discretization. To study
+problem again for a 1-group $SP_1$ discretization. To study
 convergence sensitivity to SPAINV parameters, the number of levels in
@@
 values.
+
 \begin{figure}[t!]
@@
 
-With the comparable iterative performance and improved sparsity,
-SPAINV preconditioning should then have a more favorable quality
-metric than ILUT preconditioning. Figure~\ref{fig:spainv_quality}
-gives the quality metric as a function of the threshold value for each
-of the sparsity levels computed. Not only do we see an improved
-quality metric over ILUT preconditioning (about an order of magnitude
-less), but we also note that the ideal sparsity level is not the largest
-nor the smallest. In fact, the largest and smallest levels (3 and 7)
-performed the worst in terms of the quality metric with a level of 4
-performing the best. When CPU time to generate the preconditioner is
-considered, 4 is the clear choice for this problem as more time is
-required to do the approximate inversion as the sparsity pattern of
-the preconditioner grows in size.
+Figure~\ref{fig:spainv_quality} gives the quality metric as a function
+of the threshold value for each of the sparsity levels computed. Not
+only do we see an improved quality metric over ILUT preconditioning
+(about an order of magnitude less), but we also note that the ideal
+sparsity level is not the largest nor the smallest. In fact, the
+largest and smallest levels (3 and 7) performed the worst in terms of
+the quality metric with a level of 4 performing the best.
+
 \begin{figure}[t!]
@@
 
-\subsection{MCSA Relaxation Parameters}
-\label{subsec:spn_mcsa_relaxation}
-As another means of preconditioning (or variance reduction) to meet
-the goal of improving iterative performance, the Richardson iteration
-upon which the Neumann-Ulam method is built may be implemented with an
-adjustable scalar relaxation parameter:
-\begin{equation}
-  \mathbf{x} = \mathbf{x} + \omega \mathbf{r}\:,
-  \label{eq:richardson_relaxation}
-\end{equation}
-where $\omega$ is the relaxation parameter. This is very similar to
-point Jacobi preconditioning where the system is now being scaled on
-the left by a constant value in all rows. Analogously, these
-relaxation parameter techniques can be applied to Monte Carlo to
-improve convergence as demonstrated by Dimov \cite{dimov_new_1998}. In
-this case, building an iteration matrix from
-Eq~(\ref{eq:richardson_relaxation}) gives:
-\begin{equation}
-  \mathbf{H} = \mathbf{I} - \omega \mathbf{A}\:,
-  \label{eq:relaxed_iteration_matrix}
-\end{equation}
-with the probabilities and weights for the Monte Carlo procedure
-appropriately scaled via a modified Neumann-Ulam decomposition. For
-MCSA, we can stage this scheme with two separate relaxation
-parameters; one for the outer Richardson iteration and one for the
-inner Monte Carlo solve:
-\begin{subequations}
-  \begin{gather}
-    \ve{x}^{k+1/2} = \ve{x}^k + \omega_R \ve{r}^k\:,\\ \ve{r}^{k+1/2}
-    = \ve{b} -
-    \ve{A}\ve{x}^{k+1/2}\:,\\ \omega_N\ve{A}\delta\ve{x}^{k+1/2} =
-    \omega_N\ve{r}^{k+1/2}\:,\\ \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta
-    \ve{x}^{k+1/2}\:,\\ \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:,
-  \end{gather}
-  \label{eq:mcsa_relaxed}
-\end{subequations}
-where $\omega_R$ is the Richardson iteration relaxation parameter and
-$\omega_N$ is the Neumann-Ulam Monte Carlo solve relaxation parameter.
-
-We apply these relaxation parameters to the fuel assembly problem
-along with the reduced domain approximation as a means of studying
-their effects. For each calculation presented, the number of
-iterations required to converge reported was for a single eigenvalue
-iteration.  Again, \sn{3}{4} histories are used at each MCSA iteration
-to compute the Monte Carlo correction using the adjoint collision
-estimator. A reduced domain fill level of 100 is used with a threshold
-of \sn{1}{-10} to filter small values. ILUT preconditioning with a
-drop tolerance of \sn{1}{-5} and fill level of 5 is used as
-well.
-
-Figure~\ref{fig:relax_iters} gives the number of iterations required
-to converge the fuel assembly problem for a 1-group SP1 discretization
-using varying combinations of the relaxation parameters. First, both
-parameters were fixed at the base case of 1 while the other parameter
-was varied as shown in the left-hand side plots of
-Figure~\ref{fig:relax_iters}. For the Richardson relaxation parameter,
-using a value larger than 1 gave better iterative performance up to a
-point, effectively providing a stronger extrapolation using the
-residual at each iteration. For the Neumann relaxation parameter, a
-value of less than 1 is observed to be ideal. Although initially
-counter-intuitive, the fact that the correction computed by the Monte
-Carlo solver has a stochastic error associated with it means that by
-using a Neumann relaxation parameter less than 1, the correction and
-its error are effectively dampened to improve iterative
-performance.
-
-Considering the CPU time required to converge a single eigenvalue
-iteration, similar results are also observed for the relaxation
-parameters as given by Figure~\ref{fig:relax_time}. For the base cases
-given by the plots on the left, it was found the a Richardson
-relaxation parameter of 1.1 and a Neumann relaxation parameter of 0.7
-provided the fastest CPU time for convergence. Fixing each parameter
-at these new values, the calculations were repeated as shown for the
-plots on the right hand sides of both Figures~\ref{fig:relax_iters}
-and \ref{fig:relax_time}. For each repeated timing calculation, it was
-found that the same combination of relaxation parameters found in the
-base cases performed the best although they did not necessarily have
-the best iterative performance.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/relax_iters.pdf}
-  \end{center}
-  \caption{\textbf{Number of iterations to converge a single
-      eigenvalue iteration of the fuel assembly problem as a function
-      of the relaxation parameters.} \textit{Starting in upper left
-      and moving counter-clockwise: Neumann relaxation parameter fixed
-      at 1.0, Richardson relaxation parameter fixed at 1.0, Neumann
-      relaxation parameter fixed at 0.7, Richardson relaxation
-      parameter fixed at 1.1. For each calculation \sn{3}{4}
-      stochastic histories were used to compute the MCSA correction at
-      each iteration.}}
-  \label{fig:relax_iters}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/spn_equations/relax_time.pdf}
-  \end{center}
-  \caption{\textbf{CPU time in seconds to converge a single eigenvalue
-      iteration of the fuel assembly problem as a function of the
-      relaxation parameters.} \textit{Starting in upper left and
-      moving counter-clockwise: Neumann relaxation parameter fixed at
-      1.0, Richardson relaxation parameter fixed at 1.0, Neumann
-      relaxation parameter fixed at 0.7, Richardson relaxation
-      parameter fixed at 1.1. For each calculation \sn{3}{4}
-      stochastic histories were used to compute the MCSA correction at
-      each iteration.}}
-  \label{fig:relax_time}
-\end{figure}
-
-\clearpage
-
 \subsection{MCSA Estimator and Fixed Point Iteration Analysis}
@@
 collision and expected value estimators using the best preconditioning
-and relaxation parameter combinations found by the previous
-analysis. Additionally, both the Richardson-based MCSA iteration given
-by Eq~(\ref{eq:mcsa}) and the minimal residual-based MCSA iteration
-given by Eq~(\ref{eq:mcsa_min_res}) will be used. Although the
-expected value estimator achieved the best iterative performance in
-the previous comparison and analysis for the simple Poisson problem in
+parameter combinations found by the previous analysis. Additionally,
+both the Richardson-based MCSA iteration given by Eq~(\ref{eq:mcsa})
+and the minimal residual-based MCSA iteration given by
+Eq~(\ref{eq:mcsa_min_res}) will be used. Although the expected value
+estimator achieved the best iterative performance in the previous
+comparison and analysis for the simple Poisson problem in
 \S~\ref{sec:parameter_estimator_analysis}, it is possible that the

### 2013-08-22 doc/dissertation/document/backmatter/mcsa_spn_estimator_analysis.tex
moving spn estimator analysis to an appendix

--- 2013-08-22 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
@@
-\chapter{Monte Carlo Synthetic Acceleration\\ Methods for the $SP_N$ Equations}
-\label{ch:spn_equations}
-
-In this chapter, we briefly derive the simplified $P_N$ ($SP_N$)
-equations, closely following the work of Evans
-\cite{evans_simplified_2013}, in order to gain full understanding of
-the underlying system and its potential behavior in a Monte Carlo
-context. From the $P_N$ equations, we apply a set of approximations to
-yield the $SP_N$ equations for fixed source and criticality
-problems. Using the fully-formed linear operator for the transport
-problem, we explore solutions to the $SP_N$ equations with Monte Carlo
-Synthetic Acceleration using a challenging light water reactor fuel
-assembly criticality calculation as the driving problem. Several
-difficulties arise when applying MCSA to these problems that were not
-observed when investigating the simple transport system in the
-previous chapter.
-
-For Jacobi-based preconditioners, convergence with MCSA is difficult
-if not impossible for ill-conditioned systems with this behavior
-demonstrated using a simple neutron diffusion problem. In order to
-effectively solve the $SP_N$ equations for the fuel assembly problem,
-a suite of preconditioners along with a relaxation scheme is developed
-and studied within the context of MCSA. Using these preconditioners,
-several additional issues are observed and alleviated to a certain
-extent by applying the reduced domain approximation. MCSA solutions
-for the fuel assembly problem are then verified by comparing the
-solutions against production Krylov solvers for problems with varying
-numbers of energy groups. Finally performance is analyzed through
-comparison with those same Krylov solvers in terms of both iterative
-performance and CPU timing using the same set of problems utilized for
-the verification.
-
-%%---------------------------------------------------------------------------%%
-\section{The Neutron Transport Equation}
-\label{sec:transport_eq}
-As a starting point for the $SP_N$ equations we define the
-time-independent neutron transport equation
-\cite{lewis_computational_1993}:
-\begin{multline}
-  \hat{\Omega} \cdot \vec{\nabla} \psi(\vec{r},\hat{\Omega},E) +
-  \sigma(\vec{r},E) \psi(\vec{r},\hat{\Omega},E) = \\ \iint
-  \sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}' \cdot \hat{\Omega})
-  \psi(\vec{r},\hat{\Omega}',E') d\Omega' dE' +
-  q(\vec{r},\hat{\Omega},E)\:,
-  \label{eq:general_transport}
-\end{multline}
-with the variables defined as:
-\begin{itemize}
-\item $\vec{r}$ - neutron spatial position
-\item $\hat{\Omega}$ - neutron streaming direction with radial
-  component $\mu$ and azimuthal component $\omega$
-\item $\hat{\Omega}' \cdot \hat{\Omega} = \mu_0$ is the angle of
-  scattering
-\item $E$ - neutron energy
-\item $\psi(\vec{r},\hat{\Omega},E)$ - angular flux
-\item $\sigma(\vec{r},E)$ - total interaction cross section
-\item $\sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}')$ - probability
-  of scattering from direction $\hat{\Omega}'$ into an angular domain
-  $d\hat{\Omega}'$ about the direction $\hat{\Omega}$ and from energy
-  $E'$ to an energy domain $dE'$ about energy $E$
-\item $q(\vec{r},\hat{\Omega},E)$ - external source of neutrons.
-\end{itemize}
-For this work, it is sufficient to formulate
-Eq~(\ref{eq:general_transport}) in 1-dimensional Cartesian geometry:
-\begin{multline}
-  \mu \frac{\partial}{\partial x} \psi(x,\mu,E) + \sigma(x,E)
-  \psi(x,\mu,E) = \\ \iint \sigma_s(x,E' \rightarrow E,\hat{\Omega}'
-  \cdot \hat{\Omega}) \psi(x,\hat{\Omega}',E') d\Omega' dE' +
-  \frac{q(x,E)}{4 \pi}\:,
-  \label{eq:cart_1d_transport}
-\end{multline}
-where the angular component of the solution is no longer dependent on
-the azimuthal direction of travel and an isotropic source of neutrons
-is assumed. In addition, for fission systems the eigenvalue form of
-the transport equation is:
-\begin{multline}
-  \hat{\Omega} \cdot \vec{\nabla} \psi(\vec{r},\hat{\Omega},E) +
-  \sigma(\vec{r},E) \psi(\vec{r},\hat{\Omega},E) = \\ \iint
-  \sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}' \cdot \hat{\Omega})
-  \psi(\vec{r},\hat{\Omega}',E') d\Omega' dE' + \\ \frac{1}{k} \chi(E)
-  \iint \nu \sigma_f(\hat{r},E') \psi(\vec{r},\hat{\Omega}',E')
-  d\Omega' dE' + q(\vec{r},\hat{\Omega},E) \:,
-  \label{eq:eigenvalue_transport}
-\end{multline}
-with the additional variables defined as
-\begin{itemize}
-\item $k$ - multiplication factor
-\item $\chi(E)$ - fission neutron energy spectrum
-\item $\nu$ - average number of neutrons per fission
-\item $\sigma_f(r,E')$ - fission cross section\:.
-\end{itemize}
-In 1-dimensional Cartesian geometry,
-Eq~(\ref{eq:eigenvalue_transport}) becomes:
-\begin{multline}
-  \mu \frac{\partial}{\partial x} \psi(x,\mu,E) + \sigma(x,E)
-  \psi(x,\mu,E) = \\ \iint \sigma_s(x,E' \rightarrow
-  E,\hat{\Omega}' \cdot \hat{\Omega}) \psi(x,\hat{\Omega}',E')
-  d\Omega' dE' + \\ \frac{1}{k} \chi(E)
-  \iint \nu \sigma_f(x,E') \psi(x,\hat{\Omega}',E') d\Omega'
-  dE' + \frac{q(x,E)}{4 \pi}\:.
-  \label{eq:cart_1d_eigenvalue}
-\end{multline}
-
-%%---------------------------------------------------------------------------%%
-\section{Derivation of the Monoenergetic $SP_N$ Equations}
-\label{sec:spn_equations}
-The $P_N$ equations as derived in \cite{lewis_computational_1993}
-give $N+1$ coupled first-order equations capturing the spatial and
-angular-dependence of the solution:
-\begin{equation}
-   \frac{1}{2n+1} \frac{\partial}{\partial x}\Big[ (n+1) \phi_{n+1} + n
-     \phi_{n-1} \Big] + \Sigma_n \phi_n = q\delta_{n0} \:,
-  \label{eq:final_pn_equations}
-\end{equation}
-where $\Sigma_n = \sigma-\sigma_{sn}$ and the summations are truncated
-at some level of approximation $N$ such that $n = 0,1,\dotsc,N$. This
-yields a set of $N+1$ equations for $N+2$ flux moments with the
-highest order moment set to zero for closure. In multiple dimensions,
-the equation set becomes large and coupled not only through angular
-moments but also through the spatial variables. As a simpler
-alternative to multidimensional $P_N$ solutions, Gelbard recognized in
-1960 that the planar $P_N$ equations could be simplified and applied
-an ad-hoc method to extend them to multiple dimensions, yielding the
-$SP_N$ equations. These equations are not only fewer in number, but
-also take on a diffusion-like form while maintaining the angular
-character of the flux, making them amenable to solutions with modern
-diffusion methods.
-
-First, the $P_N$ equations can be simplified to $(N+1)/2$ second-order
-equations by solving for the $n^{th}$ Legendre flux moment in the
-odd-order equations:
-\begin{equation}
-  \phi_n = \frac{1}{\Sigma_n}\Bigg[ q \delta_{no} -
-    \frac{\partial}{\partial x}\Big(\frac{n}{2n+1}\phi_{n-1} +
-    \frac{n+1}{2n+1} \phi_{n+1} \Big) \Bigg]\:, 
-  \label{eq:odd_moments}
-\end{equation}
-for $n = 1,3,\cdots,N$ and $\delta_{no} = 0\ \forall n \neq 0$. We can
-insert the odd moments into Eq~(\ref{eq:final_pn_equations}) to get a
-reduced group of equations for the even moments:
-\begin{multline}
-  -\frac{\partial}{\partial x}
-  \Bigg[\frac{n}{2n+1}\frac{1}{\Sigma_{n-1}} \frac{\partial}{\partial
-      x} \Big(\frac{n-1}{2n-1} \phi_{n-2} + \frac{n}{2n-1}\phi_n \Big)
-    \\+ \frac{n+1}{2n+1}\frac{1}{\Sigma_{n+1}} \frac{\partial}{\partial
-      x} \Big(\frac{n+1}{2n+3}\phi_n + \frac{n+2}{2n+3}\phi_{n+2}\Big)
-    \Bigg] \\+ \Sigma_n \phi_n = q \delta_{n0}\ \ \ \ \ \ \ \ \ n =
-  0,2,4,\cdots,N\:.
-  \label{eq:reduced_pn}
-\end{multline}
-To extend these equations to multiple dimensions, Gelbard simply
-replaced the planar spatial derivatives in the reduced set of
-equations with general multidimensional gradient operators:
-\begin{multline}
-  -\nabla \cdot \Bigg[\frac{n}{2n+1}\frac{1}{\Sigma_{n-1}} \nabla
-    \Big(\frac{n-1}{2n-1} \phi_{n-2} + \frac{n}{2n-1}\phi_n \Big) \\+
-    \frac{n+1}{2n+1}\frac{1}{\Sigma_{n+1}} \nabla
-    \Big(\frac{n+1}{2n+3}\phi_n + \frac{n+2}{2n+3}\phi_{n+2}\Big)
-    \Bigg] \\+ \Sigma_n \phi_n = q \delta_{n0}\ \ \ \ \ \ \ \ \ n =
-  0,2,4,\cdots,N\:,
-  \label{eq:spn_equations}
-\end{multline}
-yielding a multidimensional set of $(N+1)/2$ angular coupled equations
-defined as the $SP_N$ equations. As with the $P_N$ equations, we
-provide closure to this set of equations with $\phi_{N+1} = 0$. As a
-concrete example, we will consider the $SP_7$ equations:
-\begin{subequations}
-  \begin{gather}
-    -\nabla \cdot \frac{1}{3 \Sigma_1} \nabla ( \phi_0 + 2\phi_2 ) +
-    \Sigma_0 \phi_0 = q \\ 
-    -\nabla \cdot \Bigg[ \frac{2}{15 \Sigma_1} \nabla ( \phi_0 + 2\phi_2
-      ) + \frac{3}{35 \Sigma_3}\nabla( 3\phi_2 + 4\phi_4)\Bigg] +
-    \Sigma_2 \phi_2 = 0\\
-    -\nabla \cdot \Bigg[ \frac{4}{63 \Sigma_3} \nabla ( 3\phi_2 +
-      4\phi_4 ) + \frac{5}{99 \Sigma_5}\nabla( 5\phi_4 +
-      6\phi_6)\Bigg] + \Sigma_4 \phi_4 = 0\\
-    -\nabla \cdot \Bigg[ \frac{6}{143 \Sigma_5} \nabla ( 5\phi_4 +
-      6\phi_6 ) + \frac{7}{195 \Sigma_7}\nabla(7\phi_6)\Bigg] +
-    \Sigma_6 \phi_6 = 0 \:.
-  \end{gather}
-  \label{eq:sp7_equations}
-\end{subequations}
-To further modify these equations, we can use a change of variables to
-create a new group of equations such that the gradients are operating
-on a single vector:
-\begin{subequations}
-  \begin{gather}
-    u_1 = \phi_0 + 2\phi_2 \\
-    u_2 = 3\phi_2 + 4\phi_4 \\
-    u_3 = 5\phi_4 + 6\phi_6 \\
-    u_4 = 7\phi_6 \:.
-  \end{gather}
-  \label{eq:spn7_subs}
-\end{subequations}
-When substituted into Eq~(\ref{eq:sp7_equations}), these terms give:
-\begin{subequations}
-  \begin{gather}
-    -\nabla \cdot \frac{1}{3 \Sigma_1} \nabla u_1 + \Sigma_0 \Bigg[
-    u_1 - \frac{2}{3}u_2 + \frac{8}{15}u_3 - \frac{16}{35}u_4 \Bigg]
-    = -q \\
-    -\nabla \cdot \Bigg[ \frac{2}{15 \Sigma_1} \nabla u_1 +
-    \frac{3}{35 \Sigma_3} \nabla u_2 \Bigg] + \Sigma_2 \Bigg[
-    \frac{1}{3}u_2 - \frac{4}{15}u_3 + \frac{8}{35}u_4 \Bigg] = 0 \\
-    -\nabla \cdot \Bigg[ \frac{4}{63 \Sigma_3} \nabla u_2 +
-    \frac{5}{99 \Sigma_5} \nabla u_3 \Bigg] + \Sigma_4 \Bigg[
-    \frac{1}{5}u_3 - \frac{6}{35}u_4 \Bigg] = 0 \\ 
-    -\nabla \cdot \Bigg[ \frac{6}{143 \Sigma_5} \nabla u_3 +
-    \frac{7}{195 \Sigma_7} \nabla u_4 \Bigg] + \Sigma_6 \Bigg[
-    \frac{1}{7}u_4 \Bigg] = 0 \:.
-  \end{gather}
-  \label{eq:spn7_subs_equations}
-\end{subequations}
-If we rearrange Eq~(\ref{eq:spn7_subs_equations}) such that only one
-divergence operation is present in each equation, we can formulate
-this as a matrix system of 4 equations in the case of the $SP_7$
-approximation:
-\begin{equation}
-  -\nabla \cdot D_n \nabla u_n + \sum_{m=1}^4 A_{nm} u_m =
-  q_n\ \ \ \ \ \ \ n = 1,2,3,4\:,
-  \label{eq:spn_matrix}
-\end{equation}
-with $\mathbf{u}$ the vector of solution variables:
-\begin{equation}
-  \mathbf{u} = ( u_1\ \ u_2\ \ u_3\ \ u_4 )^T \:,
-  \label{eq:spn7_solution_vector}
-\end{equation}
-$\mathbf{D}$ the vector of effective diffusion coefficients:
-\begin{equation}
-  \mathbf{D} = \Bigg( \frac{1}{3\Sigma_1}\ \ \frac{1}{7\Sigma_3}\ \
-  \frac{1}{11\Sigma_5}\ \ \frac{1}{15\Sigma_7} \Bigg)^T\:,
-  \label{eq:spn7_diffusion_coeffs}
-\end{equation}
-$\mathbf{q}$ the vector of source terms where the $0^{th}$ moment
-source has now been distributed through the system:
-\begin{equation}
-  \mathbf{q} = (
-  q\ \ -\frac{2}{3}q\ \ \frac{8}{15}q\ \ -\frac{16}{35}q )^T\:,
-  \label{eq:spn7_source_vector}
-\end{equation}
-and $\mathbf{A}$ a matrix of angular scattering terms:
-% NOTE: I copied the following matrix directly out of Tom's tech note
-% on the SPn equations.
-\begin{equation}
-  \mathbf{A} = 
-  {\tiny \begin{bmatrix}
-    (\Sigma_0) &
-    (-\frac{2}{3}\Sigma_0) &
-    (\frac{8}{15}\Sigma_0) &
-    (-\frac{16}{35}\Sigma_0) \\
-    %%
-    &&&\\
-    %%
-    (-\frac{2}{3}\Sigma_0) &
-    (\frac{4}{9}\Sigma_0 + \frac{5}{9}\Sigma_2) &
-    (-\frac{16}{45}\Sigma_0 - \frac{4}{9}\Sigma_2) &
-    (\frac{32}{105}\Sigma_0 + \frac{8}{21}\Sigma_2) \\
-    %%
-    &&&\\
-    %%
-    (\frac{8}{15}\Sigma_0) &
-    (-\frac{16}{45}\Sigma_0 - \frac{4}{9}\Sigma_2) &
-    (\frac{64}{225}\Sigma_0 + \frac{16}{45}\Sigma_2 + \frac{9}{25}\Sigma_4) &
-    (-\frac{128}{525}\Sigma_0 - \frac{32}{105}\Sigma_2 - \frac{54}{175}\Sigma_4)
-    \\ 
-    %%
-    &&&\\
-    %%
-    (-\frac{16}{35}\Sigma_0) &
-    (\frac{32}{105}\Sigma_0 + \frac{8}{21}\Sigma_2) &
-    (-\frac{128}{525}\Sigma_0 - \frac{32}{105}\Sigma_2 - \frac{54}{175}\Sigma_4)
-    & 
-    (\frac{256}{1225}\Sigma_0 + \frac{64}{245}\Sigma_2 +
-    \frac{324}{1225}\Sigma_4 + \frac{13}{49}\Sigma_6)
-  \end{bmatrix}}\:.
-  \label{eq:A_matrix}
-\end{equation}
-Note that the term $\sum_{m=1}^4 A_{nm} u_m$ in
-Eq~(\ref{eq:spn_matrix}) couples the moments in each equation while
-the diffusive term in each equation is only for a single
-'pseudo-moment' $u_n$. As noted by Evans \cite{evans_simplified_2013},
-lower order $SP_N$ approximations can be generated by setting higher
-order even moments in this system to zero (e.g. $\phi_6 = \phi_4 = 0$
-yields the $SP_3$ equations). For these equations, the $SP_N$ order
-defines the largest order flux moment maintained in the calculation
-and the $P_N$ order defines the largest order of the cross section
-expansion maintained.
-
-\subsection{Eigenvalue Form of the $SP_N$ Equations}
-\label{subsec:eigenvalue_form}
-For criticality problems, the space, angle, and energy discretizations
-for the $SP_N$ equations can be applied to the general eigenvalue form
-of the transport equation given by
-Eq~(\ref{eq:eigenvalue_transport}). We can readily replace the fixed
-source in Eq~(\ref{eq:spn_matrix}) with a fission source this time
-considering the multigroup form of the $SP_N$ equations derived in
-Appendix~\ref{chap:mg_spn_equations}:
-\begin{multline}
-  -\nabla \cdot \Bigg[\frac{n}{2n+1}\mathbf{\Sigma_{n-1}}^{-1} \nabla
-    \Big(\frac{n-1}{2n-1} \mathbf{\Phi_{n-2}} +
-    \frac{n}{2n-1}\mathbf{\Phi_n} \Big) \\+
-    \frac{n+1}{2n+1}\mathbf{\Sigma_{n+1}}^{-1} \nabla
-    \Big(\frac{n+1}{2n+3}\mathbf{\Phi_n} +
-    \frac{n+2}{2n+3}\mathbf{\Phi_{n+2}}\Big) \Bigg] \\+
-  \mathbf{\Sigma_n} \mathbf{\Phi_n} = \frac{1}{k} \mathbf{F}
-  \mathbf{\Phi_n} \delta_{n0} \ \ \ \ \ \ \ \ \ n = 0,2,4,\cdots,N\:.
-  \label{eq:multigroup_spn_eigenvalue}
-\end{multline}
-with $\mathbf{\Phi_n}$ the vector of multigroup flux moments and the
-fission matrix, $\mathbf{F}$, given by
-\cite{evans_simplified_2013}. We again apply the change of variables
-to yield the set of multigroup pseudo-moments $\mathbb{U}_n$ where now
-the application of the fission matrix to the moment vectors will be
-expanded into a set of block matrices in identical fashion to the
-scattering matrices:
-\begin{equation}
-  -\nabla \cdot \mathbb{D}_n \nabla \mathbb{U}_n + \sum_{m=1}^4
-  \mathbb{A}_{nm} \mathbb{U}_m = \frac{1}{k} \sum_{m=1}^4
-  \mathbb{F}_{nm} \mathbb{U}_m\ \ \ \ \ \ \ n = 1,2,3,4\:.
-  \label{eq:spn_fission_matrix}
-\end{equation}
-To complete the spatial discretization, the gradient operators are
-resolved by a finite volume scheme as in
-\cite{evans_simplified_2013}. It should be noted here that with
-respect to a general MCSA scheme, the introduction of fission in the
-system will only affect the source vector in the linear system for the
-eigenvalue schemes used in this work. The linear operator, and
-therefore overall MCSA performance will be dictated by streaming and
-scattering as defined on the left-hand side.
-
-%%---------------------------------------------------------------------------%%
-\section{Basic Preconditioning for the $SP_N$ Equations}
-\label{sec:spn_spectral_analysis}
-Previous work in applying MCSA to the thermal radiation diffusion
-equation had success in using Jacobi preconditioning
-\cite{evans_monte_2012} and therefore we apply it here to the $SP_N$
-equations given their diffusion-like form. In this section, we will
-compute the spectral radius of the iteration matrix formed by the
-$SP_N$ equations while parametrically varying the $SP_N$ order, the
-$P_N$ order, the number of energy groups, and
-upscattering/downscattering between energy groups for Jacobi
-preconditioned problems to ensure that this type of preconditioning
-will be applicable to the fuel assembly problem. For each problem in
-the parameter study, the values given in
-Table~\ref{tab:spn_fixed_parameters} were fixed for each spatial
-location and energy group.
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{ll}\hline\hline
-      \multicolumn{1}{c}{Parameter}& 
-      \multicolumn{1}{c}{Value} \\\hline
-      Mesh Element Size & 1.0 \\
-      Mesh Elements & $4 \times 4 \times 4$ \\
-      Reflecting Boundaries & True \\
-      Materials & 1 \\
-      Source Strength & 1.0 \\
-      Total Cross Section & 5.0 \\
-      In-Group Cross Section & 0.25 \\
-      Downscatter Cross Section & 1.0 \\
-      Upscatter Cross Section & 0.1 \\
-      Eigenvalue Solver Tolerance & $1.0\times10^{-8}$ \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Fixed parameters for the $SP_N$ spectral radius
-      parameter study.} \textit{The parameters were fixed for each
-      spatial location and energy group.}}
-  \label{tab:spn_fixed_parameters}
-\end{table}
-
-Mesh element sizes were the same in all cardinal directions with
-reflecting conditions on each boundary. For the energy parameter, 1
-and 10 energy groups were used with upscatter/downscatter varied for
-the 10 group case. For the downscatter cases, all groups could scatter
-to all lower energy groups with the cross sections given in
-Table~\ref{tab:spn_fixed_parameters} while for the upscatter case all
-groups could scatter to all other energy groups with different cross
-sections for upscatter and downscatter used. A uniform isotropic
-source of neutrons is given in each group as well.
-
-The matrices generated by the discretization of this problem are
-sparse with a block-based form as dictated by
-Eq~(\ref{eq:spn_multigroup_system}). Figure~\ref{fig:group1} gives the
-sparsity pattern for a monoenergetic $SP_7$ discretization while
-Figures~\ref{fig:group10ds} and \ref{fig:group10us} give the sparsity
-patterns for the same problem for 10 energy groups without and with
-upscatter respectively. For these figures, the parameters in
-Table~\ref{tab:spn_fixed_parameters} with the number of spatial
-elements reduced to $2 \times 2 \times 2$ to show the blocks generated
-by the discretization.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=3.3in]{chapters/spn_equations/group1.png}
-  \end{center}
-  \caption{\textbf{Sparsity pattern for 1-group $SP_7$
-      discretization.} \textit{A $2\times 2 \times 2$ element mesh was
-      used to show detail of the blocks formed by the discretization.}}
-  \label{fig:group1}
-\end{figure}
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=3.3in]{chapters/spn_equations/group10ds.png}
-  \end{center}
-  \caption{\textbf{Sparsity pattern for 10-group $SP_7$ discretization
-      with downscatter only.} \textit{A $2\times 2 \times 2$ element
-      mesh was used to show detail of the blocks formed by the
-      discretization.}}
-  \label{fig:group10ds}
-\end{figure}
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=3.3in]{chapters/spn_equations/group10us.png}
-  \end{center}
-  \caption{\textbf{Sparsity pattern for 10-group $SP_7$ discretization
-      with downscatter and upscatter.} \textit{A $2\times 2 \times 2$ element
-      mesh was used to show detail of the blocks formed by the
-      discretization.}}
-  \label{fig:group10us}
-\end{figure}
-
-We note a few key features of the sparsity plots. The first is that
-for multigroup problems without full upscatter and downscatter
-(i.e. Figure~\ref{fig:group10ds}), the resulting matrix is asymmetric
-and therefore a linear solver that can handle asymmetric linear
-systems is required. Nearly all problems of interest will not have
-full upscattering or downscattering. Second we note the largely
-diagonal character of these systems, although the blocks from
-Eq~(\ref{eq:A_block_matrix}) are readily apparent with a block size of
-$N_g\times(N+1)/2$.
-
-Based on this both block and diagonally dominant structure for
-matrices formed by the general multigroup $SP_N$ equations, we choose
-\textit{block Jacobi} preconditioning as a left preconditioner for the
-system. Block Jacobi preconditioning extracts the diagonal block
-elements of the matrix to use as the preconditioner as shown on the
-left side of Figure~\ref{fig:block_jacobi_ex}. Shown on the right side
-of Figure~\ref{fig:block_jacobi_ex}, inversion of the preconditioner
-is trivial with each diagonal block inverted separately.
-
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.6}{
-    \input{chapters/spn_equations/block_jacobi.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Block Jacobi preconditioning strategy used for the
-      $SP_N$ equations.} \textit{Left: The preconditioner is formed by
-      the diagonal blocks of the matrix. Right: Inversion of the
-      preconditioner is trivial and decoupled by block.}}
-  \label{fig:block_jacobi_ex}
-\end{figure}
-
-For high performance implementations block Jacobi preconditioning has
-several attractive properties. First, the blocks in the transport
-operator come from the energy/angle discretization of the transport
-equation as given by Eq~(\ref{eq:A_block_matrix}). Each block on the
-diagonal is bound to a mesh element in the system (note there are 8
-blocks on the diagonal in each of the sparsity patterns with a mesh of
-$2 \times 2 \times 2$) and therefore we expect the matrix elements
-forming the block to be entirely local in domain decomposed
-calculations. Second, these blocks are typically dense and nearly
-lower triangular for many transport problems meaning that established
-dense matrix methods can be used for fast inversion.
-
-Spectral radius computations for the block Jacobi preconditioned
-iteration matrix were performed with Table~\ref{tab:group1bj} giving
-the results for the 1-group case, Table~\ref{tab:group10dsbj} for the
-10-group case with full downscatter only and
-Table~\ref{tab:group10usbj} for the 10-group case with full
-downscatter and full upscatter. A block size of $N_g\times(N+1)/2$ was
-used for the preconditioner in all cases.
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{cccccc}\hline\hline
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{$SP_N$ Order}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{} \\
-       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
-       & \textbf{0} & 0.0635 & 0.1269 & 0.1444 & 0.1513 \\
-       & \textbf{1} & 0.0666 & 0.1315 & 0.1474 & 0.1534 \\
-      $P_N$ Order & \textbf{3} & 0.0666 & 0.1365 & 0.154 & 0.1592 \\
-       & \textbf{5} & 0.0666 & 0.1365 & 0.1562 & 0.163 \\
-       & \textbf{7} & 0.0666 & 0.1365 & 0.1562 & 0.164 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Spectral radius results for the block Jacobi
-      preconditioned iteration matrix with 1 energy group.}}
-  \label{tab:group1bj}
-\end{table}
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{cccccc}\hline\hline
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{$SP_N$ Order}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{} \\
-       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
-       & \textbf{0} & 0.0647 & 0.1275 & 0.1449 & 0.1514 \\
-       & \textbf{1} & 0.0686 & 0.1338 & 0.1484 & 0.1547 \\
-      $P_N$ Order & \textbf{3} & 0.0687 & 0.1399 & 0.1582 & 0.1625 \\
-       & \textbf{5} & 0.0692 & 0.1399 & 0.1582 & 0.1657 \\
-       & \textbf{7} & 0.0678 & 0.1393 & 0.1624 & 0.166 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Spectral radius results for the block Jacobi
-      preconditioned iteration matrix with 10 energy groups and full
-      downscatter.}}
-  \label{tab:group10dsbj}
-\end{table}
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{cccccc}\hline\hline
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{$SP_N$ Order}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{} \\
-       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
-       & \textbf{0} & 0.1887 & 0.2267 & 0.2285 & 0.2286 \\
-       & \textbf{1} & 0.4535 & 0.5044 & 0.5045 & 0.5045 \\
-      $P_N$ Order & \textbf{3} & 0.4535 & 0.6453 & 0.6506 & 0.6506 \\
-       & \textbf{5} & 0.4535 & 0.6453 & 0.6802 & 0.6818 \\
-       & \textbf{7} & 0.4535 & 0.6453 & 0.6802 & 0.6927 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Spectral radius results for the block Jacobi
-      preconditioned iteration matrix with 10 energy groups, full
-      downscatter and full upscatter.}}
-  \label{tab:group10usbj}
-\end{table}
-
-From the tabulated block Jacobi data it is clear that this is a viable
-preconditioning choice for all $SP_N$ problems in terms of Monte Carlo
-solution methods. All cases were observed to have a spectral radius
-below unity. Based on these results, the block Jacobi method should be
-the applicable to more complicated problems representative of physical
-nuclear systems.
-
-%%---------------------------------------------------------------------------%%
-\section{Fuel Assembly Criticality Calculations}
-\label{sec:fuel_assembly_calcs}
-Fuel assembly calculations are a critical piece of nuclear engineering
-infrastructure for reactor core analysis and design. At this level,
-individual fuel pins may be resolved at fine resolution in a variety
-of configurations. As a sophisticated problem of interest to push the
-limits of MCSA, a hot zero-power $17 \times 17$ pin assembly will be
-used with varying energy group structure and $SP_N$ discretization in
-a criticality calculation. A cross section along the vertical axis
-showing homogenized fuel pin, control rod, and moderator materials and
-the associated grid is given by Figure~\ref{fig:problem3_radial_mat}
-while a cross section of the materials configuration along the
-horizontal axis is given by Figure~\ref{fig:problem3_axial_mat}. The
-control rods were removed for these problems leaving only moderator in
-their place. A detailed view of the assembly bottom is given in
-Figure~\ref{fig:problem3_end}. On the top and bottom of the assembly,
-vacuum conditions are used as well as on the top and right boundaries
-in Figure~\ref{fig:problem3_radial_mat}. Reflecting conditions are
-used on the left and bottom boundaries of
-Figure~\ref{fig:problem3_radial_mat}, effectively giving a
-representation of one quarter of the assembly. For the spatial
-discretization, each fuel pin (gray regions in
-Figure~\ref{fig:problem3_radial_mat}) is resolved by a $2 \times 2$
-mesh with materials and cross sections homogenized over this region.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4in]{chapters/spn_equations/problem3_radial_mat.png}
-  \end{center}
-  \caption{\textbf{Fuel assembly mesh and geometry cross section.}
-    \textit{Reflecting boundaries are used on the left and lower
-      boundaries to create a complete $17 \times 17$ assembly
-      geometry. Gray regions are homogenized fuel and red regions are
-      homogenized control rods. Each fuel pin is resolved by a $2
-      \times 2$ mesh.}}
-  \label{fig:problem3_radial_mat}
-\end{figure}
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6.0in]{chapters/spn_equations/problem3_axial_mat.png}
-  \end{center}
-  \caption{\textbf{Fuel assembly geometry cross section.} \textit{The
-      geometry is subdivided along the axial direction into 50 zones
-      spaced to capture material boundaries. Important details include
-      spacer grids along the length of the fuel pins and reactor core
-      structures on the top and bottom of the assembly. Lighter purple
-      material in the center of the assembly is homogenized
-      moderator/control rodes and darker purple/red material is
-      homogenized moderator/fuel.}}
-  \label{fig:problem3_axial_mat}
-\end{figure}
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6.0in]{chapters/spn_equations/problem3_end.png}
-  \end{center}
-  \caption{\textbf{Fuel assembly geometry end detail.}
-    \textit{Reactor core structure including spacer grids and plenum
-      has been included. Lighter purple material on the right of the
-      figure is homogenized moderator/control rods and darker
-      purple/red material is homogenized moderator/fuel.}}
-  \label{fig:problem3_end}
-\end{figure}
-
-Significant geometric details are contained in the model including
-spacer grids, fuel pins with homogenized cladding and gas gap, core
-plenum, and moderator with boron. Group cross sections and other
-discrete nuclear data are generated as needed by a cross-section
-processing module dependent on the meshing parameters used to
-discretize the geometry and single-dimension pin-cell calculations for
-initial flux spectrum generation. Table~\ref{tab:problem3_parameters}
-gives the primary design parameters for the fuel assembly
-calculations.
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{ll}\hline\hline
-      \multicolumn{1}{l}{\textbf{Parameter}} & 
-      \multicolumn{1}{l}{\textbf{Value}} \\
-      Power Level & 0 MW \\
-      Inlet Temperature & 326.85C \\
-      Fuel Temperature & 600C \\
-      Boron Concentration & 1300 ppm \\
-      Moderator Density & 0.743 g/cc \\
-      Helium Density & \sn{1.79}{-4} g/cc \\
-      Zirconium Density & 6.56 g/cc \\
-      Stainless Steel Density & 8.0 g/cc \\
-      Inconel Density & 8.19 g/cc \\
-      UO2 Density & 10.257 g/cc \\
-      Fuel Pin Radius (w/o clad) & 0.4096 cm \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Design parameters for the $17 \times 17$ pin fuel
-      assembly criticality calculation.}}
-  \label{tab:problem3_parameters}
-\end{table}
-
-To generate the multiplication factor and steady-state flux
-distribution for this problem, at every eigenvalue iteration MCSA is
-used to solve the resulting $SP_N$ problem using the provided fission
-source. Algorithm~\ref{alg:power_iteration} presents the use of MCSA
-within a power iteration strategy to find the multiplication factor.
-\begin{algorithm}[h!]
-  \caption{Power Iteration MCSA Scheme}
-  \label{alg:power_iteration}
-  \begin{algorithmic}
-    \State $k_0 =$ initial guess
-    \State $\mathbf{\Phi}_0 =$ initial guess
-    \State $n = 0$
-    \While{$|\frac{k^n - k^{n-1}}{k^n}| < \epsilon$}
-    \Comment{Iterate until convergence of the eigenvalue}
-    \State $\mathbf{M} \mathbf{\Phi}^{n+1} = \frac{1}{k^n} \mathbf{F} \mathbf{\Phi}^n$
-    \Comment{Solve for the new flux state with MCSA}
-    \State $k^{n+1} = k^n \frac{\int \mathbf{F} \mathbf{\Phi}^{n+1} d\mathbf{r}}{\int
-      \mathbf{F} \mathbf{\Phi}^n d\mathbf{r}}$
-    \Comment{Update the multiplication factor}
-    \State $n = n+1$
-    \EndWhile
-  \end{algorithmic}
-\end{algorithm}
-Here, $\mathbf{M}$ is the transport operator generated on the
-left-hand side of the $SP_N$ discretization, $\mathbf{F}$ is the
-fission matrix, and $\mathbf{\Phi}$ the multigroup neutron flux. Using
-MCSA in this context is significantly more complicated than the simple
-test problem used for the previous fixed-source preconditioner
-analysis. Fission has been introduced into the set of equations and
-the addition of moderator into the system will increase the amount of
-scattering, creating a significantly more difficult problem
-manifesting itself in an iteration matrix with a larger spectral
-radius. When using MCSA, the linear operator applied to
-$\mathbf{\Phi}^{n+1}$ at each eigenvalue iteration will dictate
-convergence and remain unchanged throughout the eigenvalue computation
-while the addition of fission to the system will only modify the
-source of neutrons and the multiplication factor at every eigenvalue
-iteration while not affecting Monte Carlo transport.
-
-\subsection{Preliminary Jacobi Preconditioned Calculations}
-\label{subsec:jacobi_prec_assembly_calc}
-We first use block Jacobi preconditioning to solve the $17 \times 17$
-fuel assembly problem. A single energy group problem was first
-computed with $SP_1$ discretization, effectively giving the one-speed
-neutron diffusion system for the fuel assembly resulting in 20,088
-degrees of freedom in the
-problem. Figure~\ref{fig:block_jacobi_res_mcsa} gives the residual
-infinity norm as a function of iteration for the MCSA linear solve in
-the first eigenvalue iteration using 25,000 stochastic histories at
-every iteration for the adjoint Neumann-Ulam solve.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=5in]{chapters/spn_equations/block_jacobi_res.pdf}
-  \end{center}
-  \caption{\textbf{Residual infinity norm vs. iteration for the block
-      Jacobi preconditioned MCSA solve during the first eigenvalue
-      iteration of the 1-group $17 \times 17$ fuel assembly problem.}
-    \textit{Convergence was not achieved with the block Jacobi
-      preconditioned method.}}
-  \label{fig:block_jacobi_res_mcsa}
-\end{figure}
-Convergence was not achieved as noted by the rapid rise in the
-residual over a few iterations. Based on the spectral radius
-computations performed, these results are not in line with
-expectations for this problem. Additional computations performed with
-\sn{1}{6} histories per iteration exhibited the same divergent
-behavior at a significant computational cost and compute time. To
-investigate further, a block Jacobi preconditioned Richardson
-iteration was used to solve the same
-problem. Figure~\ref{fig:block_jacobi_res_richardson} gives the
-residual infinity norm as a function of iteration for the Richardson
-linear solve in the first eigenvalue iteration.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=5in]{chapters/spn_equations/block_jacobi_rich_res.pdf}
-  \end{center}
-  \caption{\textbf{Residual infinity norm vs. iteration for the block
-      Jacobi preconditioned Richardson solve during the first
-      eigenvalue iteration of the 1-group $17 \times 17$ fuel assembly
-      problem.} \textit{A spectral radius near 1 was observed for the
-      iteration matrix.}}
-  \label{fig:block_jacobi_res_richardson}
-\end{figure}
-Poor convergence is observed for the Richardson iteration, however,
-convergence is achieved meaning that the preliminary eigenvalue
-criteria needed to satisfy MCSA convergence has also been met. The
-number of iterations required for the Richardson iteration to converge
-will give an approximation for the spectral radius via
-Eq~(\ref{eq:linear_k_iter_norm3}). With 7291 iterations required to
-converge to a tolerance of $\sn{1}{-6}$, $\rho(\mathbf{H}) \approx
-0.998$, nearing the limits of MCSA applicability and well beyond the
-spectral radii generated in the initial spectral analysis.
-
-Based on these results, it then appears that even if the simple
-criteria of a spectral radius of less than one is met and the
-Richardson iteration will converge with the same preconditioning, MCSA
-still may not converge. We then expect the issue to reside with the
-Neumann-Ulam solve providing the correction as the Richardson
-iteration is known to provide the correct result. Furthermore, the
-fact that the spectral radius is less than one means that the
-stochastic histories in the block Jacobi preconditioned Neumann-Ulam
-method are eventually being terminated by the weight cutoff as no
-artificial absorption was used for these preliminary
-calculations. Based on this, the correction being generated by the
-Neumann-Ulam solve is the component of MCSA causing a divergent
-solution which we study next.
-
-\subsection{MCSA Breakdown}
-\label{subsec:mcsa_break_down}
-For the fuel assembly problem, the initial block Jacobi preconditioned
-calculations used a one-speed $SP_1$ discretization, effectively
-giving a diffusion system. To study the breakdown of MCSA at iteration
-matrix spectral radii near one, we will use the simpler homogeneous
-2-dimensional one-speed neutron diffusion system presented in
-Appendix~\ref{chap:diffusion_problem} to isolate this behavior. In
-this system, we can vary the cross sections while maintaining a fixed
-grid in order to achieve varying spectral radii. For these studies, we
-neglect fission as MCSA behavior is dictated by the transport operator
-$\mathbf{M}$ in an eigenvalue scheme with the fission matrix used to
-generate a fixed source.
-
-For each solver and estimator combination, the spectral radius of the
-iteration matrix generated by the diffusion problem was varied by
-changing the absorption cross section from 0.25 to 100 while fixing
-the grid size at $100 \times 100$ with $h = 0.1$ and a fixed
-scattering cross section of unity. For each parameter variation, a
-minimum of one stochastic history per degree of freedom (DOF) in the
-problem was used to compute the Monte Carlo correction. If the solver
-could not converge in less than 100 iterations, the number of
-histories was increased by increments of 5,000 until convergence was
-achieved in less than 100 iterations. The number of iterations
-required to converge MCSA and the time to converge was recorded as a
-means to capture the breakdown.
-
-Figure~\ref{fig:breakdown_iterations} gives the number of iterations
-required to converge for the chosen number of histories per iteration
-given by Figure~\ref{fig:breakdown_histories} using the adjoint solver
-with the collision and expected value estimators and the forward
-solver with the collision estimator. For spectral radii less than
-0.97, all MCSA problems converged with 1 history per DOF (10,000 for
-this problem) with the number of iterations required to converge
-increasing as a function of spectral radius. Near a spectral radius of
-0.97, the number of histories required to converge MCSA in less than
-100 iterations takes a dramatic rise that exhibits neither exponential
-nor power law behavior. As the spectral radius approaches 1, the
-number of histories required becomes significant and effectively
-impractical to compute. Even with this simple diffusion problem, the
-behavior is consistent with that observed for the fuel assembly
-problem with $SP_1$ discretization. In that case, we estimated a
-spectral radius of $\approx 0.998$, larger than any of the spectral
-radii that could be computed within even 90 minutes of compute time
-for this simple two dimensional problem. For that problem, even
-1,000,000 histories ($\approx 50$ per DOF) was not enough to provide
-convergence. Even if single solve times of an hour can be tolerated,
-dozens of solves are typically required to compute the multiplication
-factor and flux spectrum within the $SP_N$ eigenvalue scheme for more
-difficult problems like the fuel assembly, making the method unusable.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=5.75in]{chapters/spn_equations/breakdown_iterations.pdf}
-  \end{center}
-  \caption{\textbf{Iterations required to converge as a function of
-      spectral radius for the neutron diffusion problem.} \textit{The
-      number of histories was increased to achieve convergence in less
-      than 100 iterations. At least 10,000 histories were used for
-      each calculation.}}
-  \label{fig:breakdown_iterations}
-\end{figure}
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=5.75in]{chapters/spn_equations/breakdown_histories.pdf}
-  \end{center}
-  \caption{\textbf{Histories per DOF required to converge as a
-      function of spectral radius for the neutron diffusion problem.}
-    \textit{The number of histories was increased to achieve
-      convergence in less than 100 iterations. At least 10,000
-      histories were used for each calculation with 10,000 DOFs in the
-      problem.}}
-  \label{fig:breakdown_histories}
-\end{figure}
-
-In addition to the significantly larger number of histories required
-to achieve convergence for ill-conditioned problems another penalty is
-paid due to histories that take longer to
-compute. Figure~\ref{fig:breakdown_time} gives the CPU time in seconds
-required to compute a single random walk averaged over the entire set
-of histories run in the calculation over all iterations. As the
-spectral radius increases (correlating to a higher ratio of scattering
-in the system) the random walk lengths increase, using more CPU time
-to finish the computation. Compared to spectral radii of 0.5, larger
-spectral radii over 0.97 have histories that require two orders of
-magnitude more computation time. This significant increase in
-computation time per history coupled with the significant increase in
-the number of histories required to converge is evidence that for
-problems with spectral radii above $\approx 0.97$, using MCSA to solve
-any problems of interest is entirely ineffective and not practical. At
-iteration matrix spectral radii this large, using significant numbers
-of stochastic histories in the Monte Carlo solve is not enough to
-reduce the stochastic uncertainty in the correction, causing the MCSA
-sequence to diverge instead of accelerating convergence. We therefore
-require a more expansive set of preconditioning techniques to move the
-eigenvalue spectrum of the $SP_N$ problem into a regime in which MCSA
-is more applicable and in which performance is improved.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=5.75in]{chapters/spn_equations/breakdown_time.pdf}
-  \end{center}
-  \caption{\textbf{CPU time per history as a function of spectral
-      radius for the neutron diffusion problem.} \textit{As the
-      spectral radius grows, so do the length of the random
-      walks. Longer random walks require more CPU time to compute.}}
-  \label{fig:breakdown_time}
-\end{figure}
-
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{Advanced Preconditioning Strategies}
-\label{subsec:spn_advanced_preconditioning}
-For the fuel assembly criticality problem presented in the previous
-section, the spectral radius was near one using Jacobi preconditioning
-and in a regime in which MCSA breakdown is observed. In this regime
-the number of stochastic histories required to converge MCSA increases
-rapidly and the resulting poor performance is compounded by those
-histories being increasingly expensive to compute. To overcome this
-difficulty, advanced preconditioning strategies for the $SP_N$ problem
-are required beyond simple Jacobi methods that can reduce the spectral
-radius into a region of better MCSA performance. Two modern algebraic
-preconditioning strategies, ILUT and SPAINV, will be presented here
-and used within the explicit preconditioning framework given by
-Eq~(\ref{eq:left_right_mcsa}). Data showing their effects on MCSA
-solutions of the fuel assembly criticality problem will be provided.
-
-\subsection{ILUT Preconditioning}
-\label{subsec:spn_ilut_preconditioning}
-ILUT preconditioning is chosen first due to its applicability to
-problems that have a dominating elliptic component such as the fuel
-assembly problem or other problems with a high scattering to
-absorption ratio. As with MCSA, the algebraic formulation of ILUT
-requires the physics operator to be fully formed and is therefore
-easily incorporated into our solution sequence.
-
-Incomplete lower-upper (ILU) factorizations of the linear operator can
-be used as a simple mechanism to form an approximate inverse of a
-preconditioner. To build the factorization, the sparse upper and lower
-triangular factors, $\mathbf{L}$ and $\mathbf{U}$, are computed such
-that the residual matrix formed by the factorization
-\cite{saad_iterative_2003}:
-\begin{equation}
-  \mathbf{R} = \mathbf{L} \mathbf{U} - \mathbf{A} \:,
-  \label{eq:ilu_residual_matrix}
-\end{equation}
-has a specified sparsity pattern and element magnitude. When a
-magnitude threshold is used, elements generated in the factors below
-that magnitude are dropped, resulting in ILU threshold (ILUT)
-preconditioning. The sparsity pattern in this case is determined from
-the input matrix to be preconditioned and the number of elements
-maintained in the factor is specified by a fill level parameter. A
-fill level of 1 will generate the same number of elements as the
-sparsity pattern of the input matrix while a fill level of 2 will
-contain twice as many elements in the factorization, resulting in a
-better representation of the true LU factorization. The inverse of the
-lower and upper triangular factors may then be easily inverted by
-means of simple elimination procedure to produce the
-preconditioner. For MCSA, $\mathbf{L}^{-1}$ will be used on the left
-and $\mathbf{U}^{-1}$ on the right to precondition the system.
-
-For the explicit preconditioning scheme presented in
-Eq~(\ref{eq:left_right_mcsa}), the fully inverted operator must be
-generated in order to build the set of probabilities and weights
-required for Monte Carlo sampling. Therefore, for a linear system of
-size $N$, $N$ triangular solves will be required in order to extract
-the inverse matrices from production ILU implementations. In addition,
-parallel implementations typically generate lower and upper factors
-that are only triangular locally, providing an easily parallelizable
-mechanism to generate the action of the preconditioner inverse. A
-consequence of this choice for parallel scalability is a degradation
-of the preconditioning quality as the size of the parallel system is
-increased. For serial computations, the triangular factorization is
-potentially exact depending on the parameters chosen while at
-thousands of parallel tasks, the global triangular factors differ
-significantly from the true factorization. As a result, more
-iterations are required at higher levels of parallelism to converge
-the system and will ultimately degrade overall scalability of the
-system with respect to total wall time to converge.
-
-MCSA preconditioned with ILUT was used to solve the fuel assembly
-problem presented in the previous section for a 1-group $SP_1$
-discretization. Unlike the Jacobi preconditioned strategy, convergence
-was achieved with ILUT preconditioning. To study convergence
-sensitivity to ILUT parameters, the fill level and drop tolerance were
-parametrically varied with the number of iterations to converge an
-eigenvalue iteration for the fuel assembly problem recorded along with
-the maximum number of non-zero entries observed in all matrix rows for
-the left/right preconditioned composite linear operator. To provide
-some sparsity to the factorization, the ILUT drop tolerance was used
-to drop elements in the extracted inverse triangular factors. For each
-calculation, the number of iterations required to converge reported
-was for a single eigenvalue iteration with \sn{3}{4} histories at each
-MCSA iteration to compute the Monte Carlo correction using the adjoint
-collision estimator. All ILUT calculations reported here were
-performed on a single CPU and therefore this data does not take into
-account the aforementioned effects of parallel decomposition on the
-quality of the ILUT preconditioning.
-
-Figure~\ref{fig:ilut_iterations} gives the number of iterations to
-converge the fixed source problem to a tolerance of \sn{1}{8}. As
-expected, the higher the fill level chosen for the ILUT factorization,
-the fewer iterations are required to converge the problem with only 9
-MCSA iterations were required to converge the problem for the fill
-level of 3 and drop tolerance of \sn{1}{-5} case. For higher levels of
-fill, the iterations needed to converge were not as sensitive,
-signaling that a larger drop tolerance can perhaps be used without a
-significant degradation in iterative performance. At smaller fill
-levels, the sensitivity to the ILUT drop tolerance is more
-significant. For all fill levels used, convergence was not achieved
-for the fuel assembly problem with a drop tolerance larger than
-\sn{1}{-3}.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4.25in]{chapters/spn_equations/ilut_iterations.pdf}
-  \end{center}
-  \caption{\textbf{Number of MCSA iterations required to converge an
-      eigenvalue iteration for the fuel assembly problem with ILUT
-      preconditioning as a function of ILUT drop tolerance.}
-    \textit{Each colored curve represents the iteration behavior for a
-      different ILUT fill level. Fill levels of 1.5, 2.0, 2.5, and 3.0
-      were used.}}
-  \label{fig:ilut_iterations}
-\end{figure}
-
-Unfortunately, gaining convergence (and excellent iterative
-performance) with MCSA for the $SP_N$ fuel assembly problem comes at
-an immediate cost. Figure~\ref{fig:ilut_size} gives the maximum number
-of non-zero entries in the composite linear operator generated by
-preconditioning as a function ILUT drop tolerance for varying values
-of fill level. For the 1-group $SP_1$ discretization, the original
-linear operator will contain only a maximum of 7 non-zero entries per
-row in the system. As observed in Figure~\ref{fig:ilut_size},
-performing the explicit preconditioning yields composite linear
-operators with $O(1,000)$ elements in a row for all combinations of
-fill level and drop tolerance, over 10\% of the total row size for
-this particular problem. This large number of row entries, observed
-for a significant fraction of rows in the system, creates several
-problems. First, sparsity is completely destroyed with each state in
-the system now coupled to over 10\% of the total states in the system
-through the composite iteration matrix. This means that Monte Carlo
-sampling tables will be large, requiring significant memory to store
-them and a substantial overhead in the sampling procedure during
-transport. Second, the matrix-matrix multiply operations required to
-build the composite operator see significant performance losses due to
-the amount of data that must be handled. In parallel, losing sparsity
-severely inhibits performance where now parallel operations require
-communications among orders of magnitude more processors for
-nearest-neighbor type algorithms.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4.25in]{chapters/spn_equations/ilut_size.pdf}
-  \end{center}
-  \caption{\textbf{Maximum number of non-zero entries observed for all
-      rows in the composite linear operator for the fuel assembly
-      problem with ILUT preconditioning given as a function of ILUT
-      drop tolerance.} \textit{Each colored curve represents the row
-      size for a different ILUT fill level. Fill levels of 1.5, 2.0,
-      2.5, and 3.0 were used.}}
-  \label{fig:ilut_size}
-\end{figure}
-
-As a means to assess the quality of the preconditioning, a simple
-metric is developed to address these concerns and allow comparison to
-future developments. For our studies, our core performance metric will
-be iterative performance with the minimum number of MCSA iterations
-required to converge the problem desired. For Monte Carlo
-calculations, this improved performance is balanced by the creation of
-a dense composite system and we seek to reduce the number of non-zero
-entries to a minimum value in order to lower the amount of coupling
-among states in the system and reduce the amount of memory used
-along with the potential latency overhead. Given these two objectives,
-the following metric is proposed:
-\begin{equation}
-  \text{Quality Metric} = (\text{\# iterations}) \times (\text{maximum
-    \# non-zero values})\:,
-\end{equation}
-where the highest quality preconditioning is one that minimizes this
-metric. For the ILUT preconditioned data provided,
-Figure~\ref{fig:ilut_quality} provides the computed metric as a
-function of ILUT drop tolerance for each fill level used. In general,
-the metric is similar to the data observed for the number of
-iterations required to converge as the non-zero row entries changed by
-a smaller fraction over drop tolerances tested. For higher levels of
-fill, the quality metric is observed to be relatively insensitive to
-the drop tolerance as maintaining more entries in the preconditioner
-has a stronger impact than filtering small values.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4.25in]{chapters/spn_equations/ilut_quality.pdf}
-  \end{center}
-  \caption{\textbf{ILUT preconditioning quality metric for the fuel
-      assembly problem given as a function of ILUT drop tolerance.}
-    \textit{Each colored curve represents the quality metric behavior
-      for a different ILUT fill level. Fill levels of 1.5, 2.0, 2.5,
-      and 3.0 were used.}}
-  \label{fig:ilut_quality}
-\end{figure}
-
-\subsection{Sparse Approximate Inverse Preconditioning}
-\label{subsec:spn_spainv_preconditioning}
-Sparse approximate inverse (SPAINV) preconditioning is chosen next because the true inverse of the preconditioner
-is product of the algorithm, avoiding the need for many inverse matrix
-applications as required by using ILUT with MCSA. In addition, the
-algorithm seeks to maintain the sparsity of the inverse
-preconditioner, potentially alleviating some of the dense operator
-issues observed with ILUT.
-
-SPAINV produces a sparse approximation of the inverse matrix directly
-by minimizing the Frobenius norm of the residual matrix
-\cite{saad_iterative_2003}:
-\begin{equation}
-  || \mathbf{I} - \mathbf{A} \mathbf{M} ||^2_F =
-  \sum_{j=1}^N ||\mathbf{e}_j - \mathbf{A} \mathbf{m}_j||^2_2 \:,
-  \label{eq:frobenius_norm_min}
-\end{equation}
-where the inverse preconditioning matrix $\mathbf{M}$ minimizes the
-norm on the right and $\mathbf{e}_j$ and $\mathbf{m}_j$ are the
-$j^{th}$ columns of the identity and preconditioning matrices
-respectively. Here, $\mathbf{M}$ is the actual inverse of the
-preconditioner and is formed directly by the algorithm. On the right
-hand side of Eq~(\ref{eq:frobenius_norm_min}), the Frobenius norm is
-represented as a effective sum of $N$ linear system residual norms.
-
-A few iterations of a subspace method or other iterative scheme can be
-used to effectively solve these systems to a loose tolerance and yield
-the columns of the approximate preconditioning matrix. At each
-iteration, threshold values are used to eliminate the components of
-each $\mathbf{m}_j$ column to maintain the desired level of
-sparsity. In addition, a sparsity pattern can be predetermined and
-enforced during this process. For the SPAINV implementation used for
-this work, the sparsity pattern was defined by levels such that a
-level of $N$ for an input operator $\mathbf{A}$ will generate a
-preconditioning matrix $\mathbf{M}$ with the same sparsity pattern as
-$\mathbf{A}^{N+1}$. By using the norm minimization strategy instead of
-a factorization such as ILUT, parallel results are reproduced
-regardless of parallel problem size, meaning that a serial computation
-will converge in the same number of iterations as a computation with
-thousands of cores.
-
-MCSA preconditioned with SPAINV was used to solve the fuel assembly
-problem again for a 1-group $SP_1$ discretization. To study
-convergence sensitivity to SPAINV parameters, the number of levels in
-the sparsity pattern and threshold were parametrically varied with the
-number of iterations to converge a single eigenvalue iteration for the
-fuel assembly problem recorded along with the maximum number of
-non-zero entries observed in all matrix rows for the preconditioned
-composite linear operator. For each calculation, the number of
-iterations required to converge reported was for a single eigenvalue
-iteration with \sn{3}{4} histories at each MCSA iteration to compute
-the Monte Carlo correction using the adjoint collision estimator. All
-SPAINV calculations reported here were performed on a single CPU.
-
-Figure~\ref{fig:spainv_iterations} gives the number of MCSA iterations
-needed to converge the fuel assembly problem and
-Figure~\ref{fig:spainv_size} the maximum number of non-zero entries
-per row observed in the composite linear operator as a function the
-SPAINV threshold. For this analysis, the number of levels in the
-sparsity pattern was varied from 3 to 7 with convergence of the fuel
-assembly problem not achieved for smaller level values. Compared to
-ILUT preconditioning, at higher levels we see comparable iterative
-performance with a relative invariance to the threshold value. The
-threshold did have a significant effect, however, on the time required
-to generate the preconditioner with a threshold of 0.001 over an order
-of magnitude slower than a threshold of 0.01 due to the inclusion of a
-significant number of extra values in the input operator. In addition
-to comparable iterative performance, the sparsity of the composite
-operator is greatly improved with nearly an order of magnitude
-reduction in number of non-zero entries per row for lower level
-values.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4.25in]{chapters/spn_equations/spainv_iterations.pdf}
-  \end{center}
-  \caption{\textbf{Number of MCSA iterations required to converge a
-      single eigenvalue iteration for the fuel assembly problem with
-      SPAINV preconditioning as a function of SPAINV threshold.}
-    \textit{Each colored curve represents the iteration behavior for a
-      different SPAINV level pattern. Levels of 3, 4, 5, 6, and 7 were
-      used.}}
-  \label{fig:spainv_iterations}
-\end{figure}
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4.25in]{chapters/spn_equations/spainv_size.pdf}
-  \end{center}
-  \caption{\textbf{Maximum number of non-zero entries observed for all
-      rows in the composite linear operator for the fuel assembly
-      problem with SPAINV preconditioning given as a function of
-      SPAINV threshold.} \textit{Each colored curve represents the row
-      size for a different SPAINV level pattern. Levels of 3, 4, 5, 6,
-      and 7 were used.}}
-  \label{fig:spainv_size}
-\end{figure}
-
-Figure~\ref{fig:spainv_quality} gives the quality metric as a function
-of the threshold value for each of the sparsity levels computed. Not
-only do we see an improved quality metric over ILUT preconditioning
-(about an order of magnitude less), but we also note that the ideal
-sparsity level is not the largest nor the smallest. In fact, the
-largest and smallest levels (3 and 7) performed the worst in terms of
-the quality metric with a level of 4 performing the best.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4.25in]{chapters/spn_equations/spainv_quality.pdf}
-  \end{center}
-  \caption{\textbf{SPAINV preconditioning quality metric for the fuel
-      assembly problem given as a function of SPAINV threshold.}
-    \textit{Each colored curve represents the quality metric behavior
-      for a different SPAINV level pattern. Levels of 3, 4, 5, 6, and
-      7 were used.}}
-  \label{fig:spainv_quality}
-\end{figure}
-
-\clearpage
-
-\subsection{Applying the Reduced Domain Approximation}
-\label{subsec:spn_prec_rda}
-Although ILUT and SPAINV preconditioning permitted MCSA solutions for
-the fuel assembly problem, a primary concern is the number of non-zero
-states in each row of the system generated by the explicit
-preconditioning strategy. In many cases, orders of magnitude more
-matrix elements were generated resulting in poor scalability for
-domain decomposed algorithms and overall performance issues for Monte
-Carlo. As outlined in \S~\ref{subsec:reduced_domain_approximation},
-the reduced domain approximation may be used as a mechanism to
-potentially alleviate this problem by filtering elements of the
-composite matrix in each row that fall below a certain threshold value
-or by maintaining the largest $N$ elements in each row where $N$ is a
-designated fill level. In this section, we will apply the reduced
-domain approximation to both ILUT and SPAINV preconditioned system to
-reduce the density of the composite linear operator to more manageable
-levels and study the resulting iterative performance.
-
-For each preconditioner, the parameters that achieved the best quality
-metric results from the previous analysis were used. These correspond
-to ILUT parameters of a fill level of 5.0 and a drop tolerance of
-\sn{1}{-5} and SPAINV parameters of a sparsity level of 4 and a
-threshold of 0.1. The reduced domain threshold was set to \sn{1}{-10}
-in order to eliminate any exceedingly small values from the matrix
-(often this is simply removing non-zero elements within the floating
-point tolerance of zero). The reduced domain fill level was then
-varied, starting with the largest non-zero entries per row value
-observed for each of the preconditioning types in order to assess its
-effects relative to the case where no reduced domain approximation was
-applied.
-
-Figure~\ref{fig:rda_iterations} gives the number of iterations
-required to converge for each preconditioner type as a function of
-reduced domain fill level. Figure~\ref{fig:rda_quality} gives the
-corresponding quality metric for each data point where the number of
-non-zero entries used to compute the metric is equivalent to the
-reduced domain fill level. We first note that SPAINV preconditioning
-alone is significantly more sensitive to the reduction in domain size
-over the ILUT-based methods, although the preconditioner was of
-$O(100)$ non-zero entries per row without any approximation
-applied. For the ILUT-based methods, performance was significantly
-better with convergence achieved in less than 40 MCSA iterations with
-only 10 non-zero entries in each row (vs. 7 non-zero entries for the
-case with no preconditioning). Looking at the quality metric data, we
-see a reduction in the quality metric as a function of the reduced
-domain fill level, achieving 2 orders of magnitude reduction in the
-quality metric for the ILUT preconditioning.
-
-Although applying the reduced domain approximation results in a
-successful recovery of sparsity for the Monte Carlo problem while
-maintaining good convergence properties, there is still an issue of
-forming the composite operator before applying the approximation and
-potentially generating the transpose of this operator in the case of
-the adjoint Monte Carlo method. Because of this, memory and scaling
-issues may still be observed when building the probability and weight
-matrices for the Monte Carlo problem. In addition, the expensive
-extraction of the inverse of the preconditioning operators for the
-explicit scheme creates a significant cost in overall
-performance. Future work in this area should be considerate of these
-important components of the preconditioned MCSA algorithm.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4.25in]{chapters/spn_equations/rda_iterations.pdf}
-  \end{center}
-  \caption{\textbf{Number of MCSA iterations required to converge a
-      single eigenvalue iteration for the fuel assembly problem with
-      each preconditioning as a function of reduced domain
-      approximation fill level.} \textit{The largest fill level for
-      each preconditioning presented is that using the parameters that
-      gave the best results without the approximation.}}
-  \label{fig:rda_iterations}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4.25in]{chapters/spn_equations/rda_quality.pdf}
-  \end{center}
-  \caption{\textbf{Preconditioning quality metric for the fuel
-      assembly problem given as a function of reduced domain
-      approximation fill level.} \textit{The largest fill level for
-      each preconditioning presented is that using the parameters that
-      gave the best results without the approximation.}}
-  \label{fig:rda_quality}
-\end{figure}
-
-\clearpage
-
 \subsection{MCSA Estimator and Fixed Point Iteration Analysis}
 \label{subsec:spn_estimator_comparison}
-The iterative performance of MCSA will next be studied with both the
-collision and expected value estimators using the best preconditioning
-parameter combinations found by the previous analysis. Additionally,
-both the Richardson-based MCSA iteration given by Eq~(\ref{eq:mcsa})
-and the minimal residual-based MCSA iteration given by
-Eq~(\ref{eq:mcsa_min_res}) will be used. Although the expected value
-estimator achieved the best iterative performance in the previous
-comparison and analysis for the simple Poisson problem in
-\S~\ref{sec:parameter_estimator_analysis}, it is possible that the
-collision estimator may perform better from a timing perspective. The
-deterministic component of the expected value estimator that couples
-the current state to other local states through the iteration matrix
-stencil requires application of the estimator to potentially several
-orders of magnitude more states at every transition event.
+The iterative performance of MCSA is studied in this appendix with
+both the collision and expected value estimators using the ILUT
+preconditioning. Although the expected value estimator achieved the
+best iterative performance in analysis for the simple Poisson problem
+in \S~\ref{sec:parameter_estimator_analysis}, it is possible that the
+collision estimator may perform better from a timing perspective for
+the $SP_N$ equations. The deterministic component of the expected
+value estimator that couples the current state to other local states
+through the iteration matrix stencil requires application of the
+estimator to potentially several orders of magnitude more states at
+every transition event.
 
@@
 the Richardson and minimal residual fixed point iterations. Using the
-results from the previous section, a Neumann relaxation parameter of
-0.7 was used with both estimators and fixed point iterations. For the
-collision estimator, a Richardson relaxation parameter of 1.1 was used
-with the Richardson iteration while it was found the expected value
-estimator had the best performance when a Richardson relaxation
-parameter of 1.0 was used instead. In addition, ILUT preconditioning
-with a drop tolerance of \sn{1}{-5} and a fill level of 5 was used
-with a reduced domain approximation fill level of 100 and a threshold
-of \sn{1}{-10} as in the previous analysis. Artificial absorption was
-also introduced with a value of 0.2 for the MCSA calculations that
-used the collision estimator as CPU timing performance was improved
-without a loss in iterative performance while no absorption was used
-for calculations using the expected value estimator. For each
-estimator, the number of stochastic histories used to compute the MCSA
-correction was varied from 25 to 50,000 with the number of MCSA
-iterations required to converge to a tolerance of \sn{1}{-8} recorded
-for a single eigenvalue iteration along with the CPU time required to
-achieve convergence.
+analysis from Appendix~\ref{chap:spn_mcsa_relaxation}, a Neumann
+relaxation parameter of 0.7 was used with both estimators and fixed
+point iterations. For the collision estimator, a Richardson relaxation
+parameter of 1.1 was used with the Richardson iteration while it was
+found the expected value estimator had the best performance when a
+Richardson relaxation parameter of 1.0 was used instead. In addition,
+ILUT preconditioning with a drop tolerance of \sn{1}{-5} and a fill
+level of 5 was used with a reduced domain approximation fill level of
+100 and a threshold of \sn{1}{-10} as in the previous
+analysis. Artificial absorption was also introduced with a value of
+0.2 for the MCSA calculations that used the collision estimator as CPU
+timing performance was improved without a loss in iterative
+performance while no absorption was used for calculations using the
+expected value estimator. For each estimator, the number of stochastic
+histories used to compute the MCSA correction was varied from 25 to
+50,000 with the number of MCSA iterations required to converge to a
+tolerance of \sn{1}{-8} recorded for a single eigenvalue iteration
+along with the CPU time required to achieve convergence.
 
@@
 \end{figure}
-
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{MCSA Verification\ }
-\label{sec:spn_mcsa_verification}
-Through the numerical studies presented in this chapter, we have
-developed a set of preconditioning techniques that permit MCSA to be
-used with the $SP_N$ form of the neutron transport equation and
-applied to a difficult fuel assembly criticality problem. Along with
-these preconditioning techniques, a set a varying MCSA parameters were
-analyzed to find those that yielded the best performance for this
-particular problem. In addition, several steps were taken to mitigate
-the dense composite linear operators that arise when performing
-explicit MCSA preconditioning necessary for achieve convergence. Using
-these results, we now compare MCSA to conventional Krylov solvers that
-would be typically used to solve the $SP_N$ form of the transport
-equation as presented here.
-
-MCSA solutions to the fuel assembly problem with 1, 2, and 4 energy
-groups using $SP_1$ discretization will computed using both BiCGStab
-and GMRES implementations from the Trilinos scientific computing
-library Aztec \cite{heroux_overview_2005}. The neutron fluxes, the
-number of eigenvalue iterations required to converge the problem, and
-the eigenvalue itself will be compared as a means of verification. For
-the flux values, we will look at the 2-norm and infinity norm of the
-relative difference vector between the reference case and the
-comparison case in each energy group:
-\begin{equation}
-  \text{Relative Group Flux Difference} = \frac{\ve{\Phi}_{ref}^g -
-    \ve{\Phi}^g}{\ve{\Phi}_{ref}^g}\:,
-  \label{eq:relative_difference_vector}
-\end{equation}
-where the addition and subtraction operations are performed
-element-wise to construct the vector. For this work, the BiCGStab
-results will be taken as the reference case.
-
-All solvers will be preconditioned with ILUT using a drop tolerance of
-\sn{1}{-5} and a fill level of 5. For GMRES, no restrictions were
-placed on the size of the subspace and therefore no restarts
-occurred. When the collision estimator was used with MCSA, \sn{2}{4}
-stochastic histories were used to compute the correction for every
-energy group in the problem (\sn{4}{4} and \sn{8}{4} histories total
-for the 2 and 4 group calculations respectively) corresponding to
-approximately 1 stochastic history per DOF. When the expected value
-estimator was used, \sn{5}{2} stochastic histories used for each
-energy group (\sn{1}{3} and \sn{1.5}{3} histories total for the 2 and
-4 group calculations) giving approximately 1 stochastic history for
-every 5 DOFs. The reduced domain approximation was also applied in
-conjunction with the ILUT preconditioning using a fill level of 100
-and a threshold value of \sn{1}{-10} to reduce the density of states
-in the Monte Carlo problem. For relaxation parameters, all MCSA
-computations used a Neumann relaxation parameter of 0.7 while a
-Richardson relaxation of 1.1 was used with the collision estimator and
-1.0 with the expected value estimator as determined by the previous
-analysis. Table~\ref{tab:spn_solver_defs} gives definitions for the
-solvers used to generate the results in the remainder of this
-section. For the energy group structures,
-Table~\ref{tab:spn_group_structure} gives the lower bounds of each
-group in eV (an implicit upper bound of \sn{2}{6} eV is assumed for
-group 0).
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{ll}\hline\hline
-      \multicolumn{1}{l}{\textbf{Name}} & 
-      \multicolumn{1}{l}{\textbf{Definition}} \\
-      BiCGStab-ILUT & BiCGStab preconditioned with ILUT \\
-      GMRES-ILUT & GMRES preconditioned with ILUT \\
-      MCSA-ILUT-R-C & MCSA preconditioned with ILUT using Richardson \\ 
-                    & fixed point iteration and collision estimator \\
-      MCSA-ILUT-MR-C & MCSA preconditioned with ILUT using minimal residual \\
-                     & fixed point iteration and collision estimator \\
-      MCSA-ILUT-R-EV & MCSA preconditioned with ILUT using Richardson \\
-                     & fixed point iteration and expected value estimator \\
-      MCSA-ILUT-MR-EV & MCSA preconditioned with ILUT using minimal residual \\
-                      & fixed point iteration and expected value estimator \\
-      Richardson-ILUT & Richardson iteration preconditioned with ILUT \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Solver definitions used for MCSA verification and
-      performance analysis.} \textit{The ILUT preconditioner
-      parameters were identical for all calculations and solvers.}}
-  \label{tab:spn_solver_defs}
-\end{table}
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{cl}\hline\hline
-      \multicolumn{1}{c}{\textbf{Number of Groups}} & 
-      \multicolumn{1}{l}{\textbf{Lower Bounds (eV)}} \\
-      1 & \{ \sn{1}{-5} \} \\
-      2 & \{ \sn{1}{-1}, \sn{1}{-5} \} \\
-      4 & \{ \sn{1}{1}, \sn{1}{0}, \sn{1}{-1}, \sn{1}{-5} \} \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Energy group lower bounds for the multigroup fuel
-      assembly criticality problem in electron volts.} \textit{An
-      implicit upper bound of \sn{2}{6} eV is assumed for group zero.}}
-  \label{tab:spn_group_structure}
-\end{table}
-
-For every combination of energy group structure and solver, the
-eigenvalue problem was converged with an eigensolver tolerance of
-\sn{1}{-6} and a linear solver tolerance of \sn{1}{-8} in a serial
-calculation. Table~\ref{tab:serial_ev_results} gives the eigenvalue
-computed by each solver for each problem as well as the number of
-eigenvalue iterations required to converge. For every problem, each
-solver converged to the same eigenvalue up to the number of digits
-reported by the eigensolver in the same number of eigenvalue
-iterations.
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lccc}\hline\hline
-      \multicolumn{1}{c}{\textbf{Solver}} & 
-      \multicolumn{1}{c}{\textbf{Number of Groups}} & 
-      \multicolumn{1}{c}{\textbf{Iterations}} &
-      \multicolumn{1}{c}{\textbf{Eigenvalue}} \\
-      \hline
-      BiCGStab-ILUT & 1 & 25 & 1.105987 \\
-      GMRES-ILUT & 1 & 25 & 1.105987 \\
-      MCSA-ILUT-R-C & 1 & 25 & 1.105987 \\
-      MCSA-ILUT-MR-C & 1 & 25 & 1.105987 \\
-      MCSA-ILUT-R-EV & 1 & 25 & 1.105987 \\
-      MCSA-ILUT-MR-EV & 1 & 25 & 1.105987 \\
-      Richardson-ILUT & 1 & 25 & 1.105987 \\
-      \hline
-      BiCGStab-ILUT & 2 & 35 & 1.155239 \\
-      GMRES-ILUT & 2 & 35 & 1.155239 \\
-      MCSA-ILUT-R-C & 2 & 35 & 1.155239 \\
-      MCSA-ILUT-MR-C & 2 & 35 & 1.155239 \\
-      MCSA-ILUT-R-EV & 2 & 35 & 1.155239 \\
-      MCSA-ILUT-MR-EV & 2 & 35 & 1.155239 \\
-      Richardson-ILUT & 2 & 35 & 1.155239 \\
-      \hline
-      BiCGStab-ILUT & 4 & 31 & 1.175850 \\
-      GMRES-ILUT & 4 & 31 & 1.175850 \\
-      MCSA-ILUT-R-C & 4 & 31 & 1.175850 \\
-      MCSA-ILUT-MR-C & 4 & 31 & 1.175850 \\
-      MCSA-ILUT-R-EV & 4 & 31 & 1.175850 \\
-      MCSA-ILUT-MR-EV & 4 & 31 & 1.175850 \\
-      Richardson-ILUT & 4 & 31 & 1.175850 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Serial eigensolver verification results for the
-      fuel assembly problem.} \textit{Each calculation was converged
-      with the same eigensolver and linear solver
-      tolerances. Table~\ref{tab:spn_solver_defs} gives the
-      description for each solver type presented.}}
-  \label{tab:serial_ev_results}
-\end{table}
-
-Table~\ref{tab:serial_differences_g1},
-Table~\ref{tab:serial_differences_g2}, and
-Table~\ref{tab:serial_differences_g4} give the infinity norms and
-2-norms of the flux relative difference vectors given by
-Eq~(\ref{eq:relative_difference_vector}) for all calculations and all
-energy groups. In general, we see good agreement in the 2-norm between
-the MCSA variants and the GMRES calculation, meaning that they have
-approximately the same variation from the BiCGStab calculation. We do
-note larger infinity norm values of this vector for some of the MCSA
-variants relative to the GMRES results (i.e. group 1 in the 2-group
-calculation), however, this large maximum value is apparently a local
-phenomenon with good agreement in the 2-norm for the
-calculations. Based on the good agreement between MCSA and GMRES for
-the relative difference of the group fluxes and the eigenvalue results
-presented here, the MCSA implementation should be deemed correct for
-the fuel assembly problem for serial calculations with multiple energy
-groups.
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lccc}\hline\hline
-      \multicolumn{1}{c}{\textbf{Solver}} & 
-      \multicolumn{1}{c}{\textbf{Group}} &
-      \multicolumn{1}{c}{\textbf{$|| \cdot ||_{inf}$}} &
-      \multicolumn{1}{c}{\textbf{$|| \cdot ||_2$}} \\
-      \hline
-      GMRES-ILUT & 0 & \sn{4.562}{-6} & \sn{6.746}{-5} \\
-      MCSA-ILUT-R-C & 0 & \sn{2.099}{-5} & \sn{2.541}{-4} \\
-      MCSA-ILUT-MR-C & 0 & \sn{4.812}{-5} & \sn{4.672}{-4} \\
-      MCSA-ILUT-R-EV & 0 & \sn{2.371}{-5} & \sn{2.740}{-4} \\
-      MCSA-ILUT-MR-EV & 0 & \sn{3.605}{-5} & \sn{2.554}{-4} \\
-      Richardson-ILUT & 0 & \sn{1.531}{-5} & \sn{2.771}{-4} \\
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Serial scalar flux verification results for the
-      1-group fuel assembly problem.} \textit{The infinity norm and
-      2-norm of the relative flux vector in each group is given with
-      BiCGStab used as the reference. Each calculation was converged
-      with the same eigensolver and linear solver
-      tolerances. Table~\ref{tab:spn_solver_defs} gives the
-      description for each solver type presented.}}
-  \label{tab:serial_differences_g1}
-\end{table}
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lccc}\hline\hline
-      \multicolumn{1}{c}{\textbf{Solver}} & 
-      \multicolumn{1}{c}{\textbf{Group}} &
-      \multicolumn{1}{c}{\textbf{$|| \cdot ||_{inf}$}} &
-      \multicolumn{1}{c}{\textbf{$|| \cdot ||_2$}} \\
-      \hline
-      GMRES-ILUT & 0 & \sn{7.590}{-6} & \sn{1.234}{-4} \\
-      MCSA-ILUT-R-C & 0 & \sn{5.308}{-6} & \sn{1.066}{-4} \\
-      MCSA-ILUT-MR-C & 0 & \sn{6.388}{-5} & \sn{2.233}{-4} \\
-      MCSA-ILUT-R-EV & 0 & \sn{4.197}{-5} & \sn{2.168}{-4} \\
-      MCSA-ILUT-MR-EV & 0 & \sn{2.736}{-5} & \sn{8.160}{-5} \\
-      Richardson-ILUT & 0 & \sn{3.976}{-6} & \sn{7.059}{-5} \\
-      \hline
-      GMRES-ILUT & 1 & \sn{7.579}{-6} & \sn{1.246}{-4} \\
-      MCSA-ILUT-R-C & 1 & \sn{1.645}{-5} & \sn{1.210}{-4} \\
-      MCSA-ILUT-MR-C & 1 & \sn{6.635}{-4} & \sn{9.435}{-4} \\
-      MCSA-ILUT-R-EV & 1 & \sn{1.797}{-4} & \sn{2.983}{-4} \\
-      MCSA-ILUT-MR-EV & 1 & \sn{1.556}{-4} & \sn{2.055}{-4} \\
-      Richardson-ILUT & 1 & \sn{4.031}{-6} & \sn{7.110}{-5} \\
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Serial scalar flux verification results for the
-      2-group fuel assembly problem.} \textit{The infinity norm and
-      2-norm of the relative flux vector in each group is given with
-      BiCGStab used as the reference. Each calculation was converged
-      with the same eigensolver and linear solver
-      tolerances. Table~\ref{tab:spn_solver_defs} gives the
-      description for each solver type presented.}}
-  \label{tab:serial_differences_g2}
-\end{table}
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lccc}\hline\hline
-      \multicolumn{1}{c}{\textbf{Solver}} & 
-      \multicolumn{1}{c}{\textbf{Group}} &
-      \multicolumn{1}{c}{\textbf{$|| \cdot ||_{inf}$}} &
-      \multicolumn{1}{c}{\textbf{$|| \cdot ||_2$}} \\
-      \hline
-      GMRES-ILUT & 0 & \sn{2.376}{-5} & \sn{4.043}{-4} \\
-      MCSA-ILUT-R-C & 0 & \sn{2.639}{-5} & \sn{3.950}{-4} \\
-      MCSA-ILUT-MR-C & 0 & \sn{3.958}{-5} & \sn{4.635}{-4} \\
-      MCSA-ILUT-R-EV & 0 & \sn{2.749}{-5} & \sn{3.975}{-4} \\
-      MCSA-ILUT-MR-EV & 0 & \sn{2.969}{-5} & \sn{4.054}{-4} \\
-      Richardson-ILUT & 0 & \sn{2.189}{-5} & \sn{3.791}{-4} \\
-      \hline
-      GMRES-ILUT & 1 & \sn{2.271}{-5} & \sn{3.952}{-4} \\
-      MCSA-ILUT-R-C & 1 & \sn{3.701}{-5} & \sn{3.933}{-4} \\
-      MCSA-ILUT-MR-C & 1 & \sn{3.803}{-5} & \sn{3.933}{-4} \\
-      MCSA-ILUT-R-EV & 1 & \sn{4.142}{-5} & \sn{4.011}{-4} \\
-      MCSA-ILUT-MR-EV & 1 & \sn{2.723}{-5} & \sn{4.030}{-4} \\
-      Richardson-ILUT & 1 & \sn{2.137}{-5} & \sn{3.768}{-4} \\
-      \hline
-      GMRES-ILUT & 2 & \sn{2.310}{-5} & \sn{4.021}{-4} \\
-      MCSA-ILUT-R-C & 2 & \sn{2.490}{-5} & \sn{3.956}{-4} \\
-      MCSA-ILUT-MR-C & 2 & \sn{2.214}{-5} & \sn{3.608}{-4} \\
-      MCSA-ILUT-R-EV & 2 & \sn{4.456}{-5} & \sn{3.833}{-4} \\
-      MCSA-ILUT-MR-EV & 2 & \sn{3.366}{-5} & \sn{3.797}{-4} \\
-      Richardson-ILUT & 2 & \sn{2.143}{-5} & \sn{3.767}{-4} \\
-      \hline
-      GMRES-ILUT & 3 & \sn{2.335}{-5} & \sn{4.029}{-4} \\
-      MCSA-ILUT-R-C & 3 & \sn{3.045}{-5} & \sn{3.963}{-4} \\
-      MCSA-ILUT-MR-C & 3 & \sn{2.696}{-5} & \sn{3.606}{-4} \\
-      MCSA-ILUT-R-EV & 3 & \sn{1.947}{-5} & \sn{4.283}{-4} \\
-      MCSA-ILUT-MR-EV & 3 & \sn{3.295}{-5} & \sn{3.837}{-4} \\
-      Richardson-ILUT & 3 & \sn{2.161}{-5} & \sn{3.772}{-4} \\
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Serial scalar flux verification results for the
-      4-group fuel assembly problem.} \textit{The infinity norm and
-      2-norm of the relative flux vector in each group is given with
-      BiCGStab used as the reference. Each calculation was converged
-      with the same eigensolver and linear solver
-      tolerances. Table~\ref{tab:spn_solver_defs} gives the
-      description for each solver type presented.}}
-  \label{tab:serial_differences_g4}
-\end{table}
-
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{MCSA Performance Comparison to Conventional Methods\ }
-\label{sec:spn_comparison}
-With the same set of solver parameters used for verification
-caluclation, MCSA will now be compared to the same Krylov solvers in
-terms of both iterative performance and CPU timing. A Richardson
-iteration solution will also be used in order to assess the
-acceleration provided by the residual Monte Carlo component of MCSA.
-To compare iterative performance, the number of linear solver
-iterations required to converge each eigenvalue iteration was recorded
-and then averaged over all eigenvalue iterations for each
-solver.
-
-Table~\ref{tab:spn_comparison_iterations} gives the average number of
-linear solver iterations required to converge a single eigenvalue
-iteration of the fuel assembly problem as a function of the number of
-energy groups. In general, the iterative performance of all the
-solvers tested aside from the Richardson iteration is comparable with
-BiCGStab performing the best. We expect this because not only are the
-multigroup $SP_N$ equations positive-definite, but they are also
-block-symmetric and therefore we expect the conjugate gradient-based
-methods to perform well due to the structure of the projection
-scheme. It important to note that the iterative performance of each
-solver is not a strong function of the number of energy groups in the
-problem (and therefore the problem size). Even more important,
-although BiCGStab produced the best iterative results, MCSA did
-perform better than GMRES in many cases with the Richardson iteration
-clearly accelerated by the Monte Carlo solve. When compared to the
-preconditioned Richardson iteration, the acceleration provided by MCSA
-is clear with 3-4 times fewer iterations required for convergence.
-
-It is also important to note the reduced domain approximation
-parameters, particularly the fill level, were fixed as the number of
-energy groups was increased. By fixing the fill level, we are
-effectively fixing the amount of information contained in the
-composite linear operator available for the Monte Carlo
-problem. Limiting this information to 100 entries per row in the
-preconditioned system did not appear to have a significant effect on
-the iterative performance of the solver. For larger numbers of energy
-groups (and therefore DOFs) this may be an issue and this number may
-need to be increased to maintain iterative performance. Currently,
-memory restrictions arising from the explicit preconditioning strategy
-prevent larger numbers of energy groups to be analyzed using an
-appropriate preconditioner for the fuel assembly problem.
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lccc}\hline\hline
-      \multicolumn{1}{l}{Solver}&
-      \multicolumn{1}{c}{1 Group}&
-      \multicolumn{1}{c}{2 Groups}&
-      \multicolumn{1}{c}{4 Groups}\\
-      \hline
-      BiCGStab-ILUT & 11.6 & 11.6 & 12.4 \\ 
-      GMRES-ILUT & 18.1 & 17.9 & 18.9 \\
-      MCSA-ILUT-R-C & 14.6 & 15.4 & 17.6 \\
-      MCSA-ILUT-MR-C & 16.0 & 17.1 & 23.7 \\
-      MCSA-ILUT-R-EV & 18.3 & 19.4 & 16.8 \\
-      MCSA-ILUT-MR-EV & 19.6 & 22.4 & 17.5 \\
-      Richardson-ILUT & 60.9 & 60.4 & 63.4 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Average number of linear solver iterations
-      iterations required to converge the fixed source $SP_N$ problem
-      at each eigenvalue iteration for the fuel assembly problem for
-      1, 2 and 4 energy groups.} \textit{Values are rounded to the
-      first decimal place. Table~\ref{tab:spn_solver_defs} gives the
-      description for each solver type presented.}}
-  \label{tab:spn_comparison_iterations}
-\end{table}
-
-Although iterative performance for MCSA was comparable to that
-observed for Krylov implementations using exactly the same
-preconditioning scheme, timing performance is not as
-competitive. Figure~\ref{fig:spn_comparison_time} gives the average
-CPU time per linear solver iteration required to converge the fuel
-assembly problem with all linear solver iterations over all eigenvalue
-iterations used to compute the average. The Krylov solver
-implementations are approximately an order of magnitude faster than
-the MCSA implementations presented here. We expect these results for
-several reasons. First, even when the reduced domain approximation is
-used there are $O(100)$ elements in each row of the system and each of
-those elements must be processed during a transition in the Monte
-Carlo random walk sequence. In addition, although the preconditioned
-fixed point iterations used in the MCSA sequence do not use the
-composite linear operator but rather a sequence of matrix vector
-multiplies to achieve the same preconditioning effect, they do use the
-explicitly extracted inverse of the preconditioning matrices which
-themselves are dense, leading to exceedingly slow computation times
-even in the fixed point iteration.
-
-For the Krylov methods, the ILUT preconditioners are applied to a
-vector with two triangular solves, one for each triangular
-factor. Also, the original linear operator has $O(10)$ elements in
-each row of the system, an order of magnitude less than that used in
-the reduced domain composite operator. Second, the implementation
-presented here has not been optimized and it is likely that several
-components of the algorithm can be implemented in a more desirable
-time complexity. However, it is most important to note here that the
-CPU timing data presented in Figure~\ref{fig:spn_comparison_time}
-shows that all methods have the same time complexity; their
-differences are merely the manifestation of a large time constant due
-to the effects of the explicit preconditioning scheme. For transport
-systems where this is not a problem, the literature explicitly shows
-that MCSA can be competitive with these production Krylov methods
-using CPU time as a measure \cite{evans_monte_2012}.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=5in]{chapters/spn_equations/solver_time.pdf}
-  \end{center}
-  \caption{\textbf{Average CPU time per linear solver iteration in
-      seconds for the fuel assembly problem as a function of energy
-      groups.}  \textit{All linear solver iterations over all
-      eigenvalue iterations were used to compute the
-      average. Table~\ref{tab:spn_solver_defs} gives the description
-      for each solver type presented in the legend with the Krylov
-      solvers represented in black and the MCSA solvers represented in
-      red.}}
-  \label{fig:spn_comparison_time}
-\end{figure}
-
-Finally, not considered in Figure~\ref{fig:spn_comparison_time} is the
-time required to actually form the explicit inverse preconditioner
-matrices and the composite operator through matrix-matrix
-multiplication. The timing numbers reported were simply to perform the
-MCSA iteration procedure with the composite operator already
-formed. Figure~\ref{fig:spn_comparison_prec_time} additionally
-presents the CPU times for MCSA convergence with the time to form the
-inverse of the preconditioners and the composite linear operator
-through matrix-matrix multiplication amortized over all iterations. As
-is readily observed, including the costs of these operations increases
-the MCSA computation time by another order of magnitude.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=5in]{chapters/spn_equations/solver_p_time.pdf}
-  \end{center}
-  \caption{\textbf{Average CPU time per iteration in seconds for the
-      fuel assembly problem as a function of energy groups with
-      preconditioning time included for the MCSA methods.}
-    \textit{All linear solver iterations over all eigenvalue
-      iterations were used to compute the
-      average. Table~\ref{tab:spn_solver_defs} gives the description
-      for each solver type presented in the legend. Data labeled
-      starting with PMCSA and shown in blue is identical to those
-      labeled with MCSA except that they additionally include the cost
-      of generating the inverse of the preconditioners and composite
-      linear operator.}}
-  \label{fig:spn_comparison_prec_time}
-\end{figure}
-
-Based on the performance results in this section, MCSA shows promise
-as a competitive and perhaps even superior method for solutions to the
-neutron transport problem discretized with the $SP_N$
-approximation. Not only are the correct answers produced when compared
-to production linear solvers, but the iterative performance is
-comparable to Krylov methods with identical preconditioning that would
-typically be used every-day calculations. From a CPU timing
-perspective, the methods yield the same time complexity as a function
-of the number of energy groups in the problem when compared to the
-Krylov methods. Here, a large time constant is generated due to the
-explicit preconditioning strategy and the generation of dense linear
-operators as a result. To improve these results and put general MCSA
-schemes into a performance regime where they are competitive with
-Krylov methods for neutron transport problems, significant research
-will be required to improve upon the explicit preconditioning scheme
-presented here.
-
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{Summary\ }
-\label{sec:mc_summary}
-
-In this chapter, the Monte Carlo Synthetic Acceleration method has
-been applied to the $SP_N$ discretization of the neutron transport
-equation and explored in the context of a difficult nuclear fuel
-assembly criticality problem. The following are the significant
-observations and findings.
-
-\begin{itemize}
-\item MCSA can solve the asymmetric system generated by the $SP_N$
-  equations
-\item MCSA has been incorporated into the Exnihilo neutronics
-  production code base developed at Oak Ridge National Laboratory
-\item Simple Jacobi-based preconditioning reduces the spectral radius
-  of the $SP_N$ system below unity, however, this alone is not
-  sufficient for MCSA convergence
-\item As the spectral radius approaches unity, MCSA was demonstrated
-  to break down with more stochastic histories required for
-  convergence and more CPU time required per history
-\item Light water reactor problems are difficult to solve with MCSA as
-  they have large spectral radii due to the neutron scattering in the
-  moderator
-\item Advanced algebraic preconditioning strategies were applied to
-  the $SP_N$ equations to obtain convergence with ILUT chosen for
-  subsequent investigations
-\item The reduced domain approximation was required to mitigate the
-  memory and performance constraints of the explicit preconditioning
-  strategy although both remain significant obstacles
-\item MCSA relaxation parameters were developed and observed to
-  enhance the time to solution for the fuel assembly criticality
-  problem
-\item MCSA was verified to produce the same k-eigenvalue for the fuel
-  assembly as production Krylov methods and gave good relative flux
-  agreement
-\item MCSA was observed to converge in fewer iterations per eigenvalue
-  iteration than GMRES for the fuel assembly criticality problem and
-  more than Bi-CGStab using the same preconditioning
-\item The explicit preconditioning strategy required to overcome the
-  MCSA spectral radius restriction creates runtimes $O(100)$ larger
-  than the production Krylov methods for the fuel assembly criticality
-  problem
-\end{itemize}

### 2013-08-22 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
moving spn estimator analysis to an appendix

@@
 
-\subsection{MCSA Estimator and Fixed Point Iteration Analysis}
-\label{subsec:spn_estimator_comparison}
-The iterative performance of MCSA will next be studied with both the
-collision and expected value estimators using the best preconditioning
-parameter combinations found by the previous analysis. Additionally,
-both the Richardson-based MCSA iteration given by Eq~(\ref{eq:mcsa})
-and the minimal residual-based MCSA iteration given by
-Eq~(\ref{eq:mcsa_min_res}) will be used. Although the expected value
-estimator achieved the best iterative performance in the previous
-comparison and analysis for the simple Poisson problem in
-\S~\ref{sec:parameter_estimator_analysis}, it is possible that the
-collision estimator may perform better from a timing perspective. The
-deterministic component of the expected value estimator that couples
-the current state to other local states through the iteration matrix
-stencil requires application of the estimator to potentially several
-orders of magnitude more states at every transition event.
-
-For this study, the fuel assembly problem with a 1-group $SP_1$
-discretization was solved using MCSA with the adjoint Monte Carlo
-solver using both the collision and expected value estimators and both
-the Richardson and minimal residual fixed point iterations. Using the
-results from the previous section, a Neumann relaxation parameter of
-0.7 was used with both estimators and fixed point iterations. For the
-collision estimator, a Richardson relaxation parameter of 1.1 was used
-with the Richardson iteration while it was found the expected value
-estimator had the best performance when a Richardson relaxation
-parameter of 1.0 was used instead. In addition, ILUT preconditioning
-with a drop tolerance of \sn{1}{-5} and a fill level of 5 was used
-with a reduced domain approximation fill level of 100 and a threshold
-of \sn{1}{-10} as in the previous analysis. Artificial absorption was
-also introduced with a value of 0.2 for the MCSA calculations that
-used the collision estimator as CPU timing performance was improved
-without a loss in iterative performance while no absorption was used
-for calculations using the expected value estimator. For each
-estimator, the number of stochastic histories used to compute the MCSA
-correction was varied from 25 to 50,000 with the number of MCSA
-iterations required to converge to a tolerance of \sn{1}{-8} recorded
-for a single eigenvalue iteration along with the CPU time required to
-achieve convergence.
-
-Figure~\ref{fig:spn_estimator_iters} gives the iteration count results
-and Figure~\ref{fig:spn_estimator_time} gives the timing
-results. Remarkably, although the fuel assembly problem is
-significantly more complicated than the simple Poisson problem studied
-in \S~\ref{sec:parameter_estimator_analysis}, the results here are
-largely the same with Figure~\ref{fig:spn_estimator_iters} and
-Figure~\ref{fig:estimator_nh_iters} having identical qualitative
-behavior. Using the expected value estimator with both fixed point
-iterations again creates a relative insensitivity to the number of
-histories used to compute the correction. Marginally better iterative
-and timing performance was observed for the expected value estimator
-when using the Richardson iteration. Better timing performance is
-expected as computing the extrapolation parameter in the minimal
-residual iteration requires several additional parallel
-operations.
-
-Compared to the collision estimator, the minimal residual iteration
-permits fewer stochastic histories to be used with each MCSA iteration
-while still maintaining good iterative performance. With respect to
-CPU time, there is a more pronounced effect due to the larger density
-of states created by the explicit ILUT preconditioning. Using a
-reduced domain fill level of 100 creates a situation where the
-expected value estimator is significantly more expensive per history
-to compute that the collision estimator due to the coupling of states.
-For the collision estimator, using around 20,000 stochastic histories
-at every iteration gave the best iterative performance while 500
-histories per iteration gave the best iterative performance when using
-the expected value estimator. For purely serial computational
-performance, there is not a distinguishing feature for the fuel
-assembly problem presented that would cause one to choose one of the
-estimators and fixed point iterations over the other.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=5in]{chapters/spn_equations/estimator_iters.pdf}
-  \end{center}
-  \caption{\textbf{MCSA iterations required to converge the fuel
-      assembly problem to a tolerance of \sn{1}{-8}.} \textit{Both the
-      collision and expected value estimators were used with the
-      adjoint Monte Carlo solver to compute the correction at each
-      iteration. Each estimator was used with the Richardson and
-      minimal residual fixed point iterations within an MCSA
-      iteration.}}
-  \label{fig:spn_estimator_iters}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=5in]{chapters/spn_equations/estimator_time.pdf}
-  \end{center}
-  \caption{\textbf{Total CPU time in seconds required to converge the
-      fuel assembly problem to a tolerance of \sn{1}{-8}.}
-    \textit{Both the collision and expected value estimators were used
-      with the adjoint Monte Carlo solver to compute the correction at
-      each iteration. Each estimator was used with the Richardson and
-      minimal residual fixed point iterations within an MCSA
-      iteration.}}
-  \label{fig:spn_estimator_time}
-\end{figure}
-
-\clearpage
-
 %%---------------------------------------------------------------------------%%
@@
 Richardson relaxation of 1.1 was used with the collision estimator and
-1.0 with the expected value estimator as determined by the previous
-analysis. Table~\ref{tab:spn_solver_defs} gives definitions for the
+1.0 with the expected value estimator. Additionally, both the
+Richardson-based MCSA iteration given by Eq~(\ref{eq:mcsa}) and the
+minimal residual-based MCSA iteration given by
+Eq~(\ref{eq:mcsa_min_res}) will be
+used. Table~\ref{tab:spn_solver_defs} gives definitions for the
 solvers used to generate the results in the remainder of this

### 2013-08-22 doc/dissertation/document/backmatter/mcsa_spn_estimator_analysis.tex
removing basic preconditioning sectiion and adding block jacobi subsection to the fuel assembly problem

@@
-\subsection{MCSA Estimator and Fixed Point Iteration Analysis}
-\label{subsec:spn_estimator_comparison}
+\chapter{MCSA Estimator and Fixed Point\\ Iteration Analysis}
+\label{chap:spn_estimator_comparison}
 The iterative performance of MCSA is studied in this appendix with

### 2013-08-22 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
removing basic preconditioning sectiion and adding block jacobi subsection to the fuel assembly problem

@@
 %%---------------------------------------------------------------------------%%
-\section{Basic Preconditioning for the $SP_N$ Equations}
-\label{sec:spn_spectral_analysis}
-Previous work in applying MCSA to the thermal radiation diffusion
-equation had success in using Jacobi preconditioning
-\cite{evans_monte_2012} and therefore we apply it here to the $SP_N$
-equations given their diffusion-like form. In this section, we will
-compute the spectral radius of the iteration matrix formed by the
-$SP_N$ equations while parametrically varying the $SP_N$ order, the
-$P_N$ order, the number of energy groups, and
-upscattering/downscattering between energy groups for Jacobi
-preconditioned problems to ensure that this type of preconditioning
-will be applicable to the fuel assembly problem. For each problem in
-the parameter study, the values given in
-Table~\ref{tab:spn_fixed_parameters} were fixed for each spatial
-location and energy group.
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{ll}\hline\hline
-      \multicolumn{1}{c}{Parameter}& 
-      \multicolumn{1}{c}{Value} \\\hline
-      Mesh Element Size & 1.0 \\
-      Mesh Elements & $4 \times 4 \times 4$ \\
-      Reflecting Boundaries & True \\
-      Materials & 1 \\
-      Source Strength & 1.0 \\
-      Total Cross Section & 5.0 \\
-      In-Group Cross Section & 0.25 \\
-      Downscatter Cross Section & 1.0 \\
-      Upscatter Cross Section & 0.1 \\
-      Eigenvalue Solver Tolerance & $1.0\times10^{-8}$ \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Fixed parameters for the $SP_N$ spectral radius
-      parameter study.} \textit{The parameters were fixed for each
-      spatial location and energy group.}}
-  \label{tab:spn_fixed_parameters}
-\end{table}
-
-Mesh element sizes were the same in all cardinal directions with
-reflecting conditions on each boundary. For the energy parameter, 1
-and 10 energy groups were used with upscatter/downscatter varied for
-the 10 group case. For the downscatter cases, all groups could scatter
-to all lower energy groups with the cross sections given in
-Table~\ref{tab:spn_fixed_parameters} while for the upscatter case all
-groups could scatter to all other energy groups with different cross
-sections for upscatter and downscatter used. A uniform isotropic
-source of neutrons is given in each group as well.
-
-The matrices generated by the discretization of this problem are
-sparse with a block-based form as dictated by
-Eq~(\ref{eq:spn_multigroup_system}). Figure~\ref{fig:group1} gives the
-sparsity pattern for a monoenergetic $SP_7$ discretization while
-Figures~\ref{fig:group10ds} and \ref{fig:group10us} give the sparsity
-patterns for the same problem for 10 energy groups without and with
-upscatter respectively. For these figures, the parameters in
-Table~\ref{tab:spn_fixed_parameters} with the number of spatial
-elements reduced to $2 \times 2 \times 2$ to show the blocks generated
-by the discretization.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=3.3in]{chapters/spn_equations/group1.png}
-  \end{center}
-  \caption{\textbf{Sparsity pattern for 1-group $SP_7$
-      discretization.} \textit{A $2\times 2 \times 2$ element mesh was
-      used to show detail of the blocks formed by the discretization.}}
-  \label{fig:group1}
-\end{figure}
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=3.3in]{chapters/spn_equations/group10ds.png}
-  \end{center}
-  \caption{\textbf{Sparsity pattern for 10-group $SP_7$ discretization
-      with downscatter only.} \textit{A $2\times 2 \times 2$ element
-      mesh was used to show detail of the blocks formed by the
-      discretization.}}
-  \label{fig:group10ds}
-\end{figure}
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=3.3in]{chapters/spn_equations/group10us.png}
-  \end{center}
-  \caption{\textbf{Sparsity pattern for 10-group $SP_7$ discretization
-      with downscatter and upscatter.} \textit{A $2\times 2 \times 2$ element
-      mesh was used to show detail of the blocks formed by the
-      discretization.}}
-  \label{fig:group10us}
-\end{figure}
-
-We note a few key features of the sparsity plots. The first is that
-for multigroup problems without full upscatter and downscatter
-(i.e. Figure~\ref{fig:group10ds}), the resulting matrix is asymmetric
-and therefore a linear solver that can handle asymmetric linear
-systems is required. Nearly all problems of interest will not have
-full upscattering or downscattering. Second we note the largely
-diagonal character of these systems, although the blocks from
-Eq~(\ref{eq:A_block_matrix}) are readily apparent with a block size of
-$N_g\times(N+1)/2$.
-
-Based on this both block and diagonally dominant structure for
-matrices formed by the general multigroup $SP_N$ equations, we choose
-\textit{block Jacobi} preconditioning as a left preconditioner for the
-system. Block Jacobi preconditioning extracts the diagonal block
-elements of the matrix to use as the preconditioner as shown on the
-left side of Figure~\ref{fig:block_jacobi_ex}. Shown on the right side
-of Figure~\ref{fig:block_jacobi_ex}, inversion of the preconditioner
-is trivial with each diagonal block inverted separately.
-
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.6}{
-    \input{chapters/spn_equations/block_jacobi.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Block Jacobi preconditioning strategy used for the
-      $SP_N$ equations.} \textit{Left: The preconditioner is formed by
-      the diagonal blocks of the matrix. Right: Inversion of the
-      preconditioner is trivial and decoupled by block.}}
-  \label{fig:block_jacobi_ex}
-\end{figure}
-
-For high performance implementations block Jacobi preconditioning has
-several attractive properties. First, the blocks in the transport
-operator come from the energy/angle discretization of the transport
-equation as given by Eq~(\ref{eq:A_block_matrix}). Each block on the
-diagonal is bound to a mesh element in the system (note there are 8
-blocks on the diagonal in each of the sparsity patterns with a mesh of
-$2 \times 2 \times 2$) and therefore we expect the matrix elements
-forming the block to be entirely local in domain decomposed
-calculations. Second, these blocks are typically dense and nearly
-lower triangular for many transport problems meaning that established
-dense matrix methods can be used for fast inversion.
-
-Spectral radius computations for the block Jacobi preconditioned
-iteration matrix were performed with Table~\ref{tab:group1bj} giving
-the results for the 1-group case, Table~\ref{tab:group10dsbj} for the
-10-group case with full downscatter only and
-Table~\ref{tab:group10usbj} for the 10-group case with full
-downscatter and full upscatter. A block size of $N_g\times(N+1)/2$ was
-used for the preconditioner in all cases.
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{cccccc}\hline\hline
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{$SP_N$ Order}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{} \\
-       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
-       & \textbf{0} & 0.0635 & 0.1269 & 0.1444 & 0.1513 \\
-       & \textbf{1} & 0.0666 & 0.1315 & 0.1474 & 0.1534 \\
-      $P_N$ Order & \textbf{3} & 0.0666 & 0.1365 & 0.154 & 0.1592 \\
-       & \textbf{5} & 0.0666 & 0.1365 & 0.1562 & 0.163 \\
-       & \textbf{7} & 0.0666 & 0.1365 & 0.1562 & 0.164 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Spectral radius results for the block Jacobi
-      preconditioned iteration matrix with 1 energy group.}}
-  \label{tab:group1bj}
-\end{table}
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{cccccc}\hline\hline
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{$SP_N$ Order}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{} \\
-       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
-       & \textbf{0} & 0.0647 & 0.1275 & 0.1449 & 0.1514 \\
-       & \textbf{1} & 0.0686 & 0.1338 & 0.1484 & 0.1547 \\
-      $P_N$ Order & \textbf{3} & 0.0687 & 0.1399 & 0.1582 & 0.1625 \\
-       & \textbf{5} & 0.0692 & 0.1399 & 0.1582 & 0.1657 \\
-       & \textbf{7} & 0.0678 & 0.1393 & 0.1624 & 0.166 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Spectral radius results for the block Jacobi
-      preconditioned iteration matrix with 10 energy groups and full
-      downscatter.}}
-  \label{tab:group10dsbj}
-\end{table}
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{cccccc}\hline\hline
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{$SP_N$ Order}& 
-      \multicolumn{1}{c}{}& 
-      \multicolumn{1}{c}{} \\
-       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
-       & \textbf{0} & 0.1887 & 0.2267 & 0.2285 & 0.2286 \\
-       & \textbf{1} & 0.4535 & 0.5044 & 0.5045 & 0.5045 \\
-      $P_N$ Order & \textbf{3} & 0.4535 & 0.6453 & 0.6506 & 0.6506 \\
-       & \textbf{5} & 0.4535 & 0.6453 & 0.6802 & 0.6818 \\
-       & \textbf{7} & 0.4535 & 0.6453 & 0.6802 & 0.6927 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Spectral radius results for the block Jacobi
-      preconditioned iteration matrix with 10 energy groups, full
-      downscatter and full upscatter.}}
-  \label{tab:group10usbj}
-\end{table}
-
-From the tabulated block Jacobi data it is clear that this is a viable
-preconditioning choice for all $SP_N$ problems in terms of Monte Carlo
-solution methods. All cases were observed to have a spectral radius
-below unity. Based on these results, the block Jacobi method should be
-the applicable to more complicated problems representative of physical
-nuclear systems.
-
-%%---------------------------------------------------------------------------%%
 \section{Fuel Assembly Criticality Calculations}
@@
 
-\subsection{Preliminary Jacobi Preconditioned Calculations}
+\subsection{Block Jacobi Preconditioning}
+\label{subsec:spn_spectral_analysis}
+Previous work in applying MCSA to the thermal radiation diffusion
+equation had success in using Jacobi preconditioning
+\cite{evans_monte_2012} and therefore we apply it here to the $SP_N$
+equations given their diffusion-like form. In this section, we will
+the structure of the transport operator formed by the $SP_N$ equations
+while parametrically varying the $SP_N$ order, the $P_N$ order, the
+number of energy groups, and upscattering/downscattering between
+energy groups. For each basic problem, the values given in
+Table~\ref{tab:spn_fixed_parameters} were fixed for each spatial
+location and energy group.
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{ll}\hline\hline
+      \multicolumn{1}{c}{Parameter}& 
+      \multicolumn{1}{c}{Value} \\\hline
+      Mesh Element Size & 1.0 \\
+      Mesh Elements & $2 \times 2 \times 2$ \\
+      Reflecting Boundaries & True \\
+      Materials & 1 \\
+      Source Strength & 1.0 \\
+      Total Cross Section & 5.0 \\
+      In-Group Cross Section & 0.25 \\
+      Downscatter Cross Section & 1.0 \\
+      Upscatter Cross Section & 0.1 \\
+      Eigenvalue Solver Tolerance & $1.0\times10^{-8}$ \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Fixed parameters for the $SP_N$ spectral radius
+      parameter study.} \textit{The parameters were fixed for each
+      spatial location and energy group.}}
+  \label{tab:spn_fixed_parameters}
+\end{table}
+
+Mesh element sizes were the same in all cardinal directions with
+reflecting conditions on each boundary. For the energy parameter, 1
+and 10 energy groups were used with upscatter/downscatter varied for
+the 10 group case. For the downscatter cases, all groups could scatter
+to all lower energy groups with the cross sections given in
+Table~\ref{tab:spn_fixed_parameters} while for the upscatter case all
+groups could scatter to all other energy groups with different cross
+sections for upscatter and downscatter used. A uniform isotropic
+source of neutrons is given in each group as well.
+
+The matrices generated by the discretization of this problem are
+sparse with a block-based form as dictated by
+Eq~(\ref{eq:spn_multigroup_system}). Figure~\ref{fig:group1} gives the
+sparsity pattern for a monoenergetic $SP_7$ discretization while
+Figures~\ref{fig:group10ds} and \ref{fig:group10us} give the sparsity
+patterns for the same problem for 10 energy groups without and with
+upscatter respectively.
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=3.3in]{chapters/spn_equations/group1.png}
+  \end{center}
+  \caption{\textbf{Sparsity pattern for 1-group $SP_7$
+      discretization.} \textit{A $2\times 2 \times 2$ element mesh was
+      used to show detail of the blocks formed by the discretization.}}
+  \label{fig:group1}
+\end{figure}
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=3.3in]{chapters/spn_equations/group10ds.png}
+  \end{center}
+  \caption{\textbf{Sparsity pattern for 10-group $SP_7$ discretization
+      with downscatter only.} \textit{A $2\times 2 \times 2$ element
+      mesh was used to show detail of the blocks formed by the
+      discretization.}}
+  \label{fig:group10ds}
+\end{figure}
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=3.3in]{chapters/spn_equations/group10us.png}
+  \end{center}
+  \caption{\textbf{Sparsity pattern for 10-group $SP_7$ discretization
+      with downscatter and upscatter.} \textit{A $2\times 2 \times 2$ element
+      mesh was used to show detail of the blocks formed by the
+      discretization.}}
+  \label{fig:group10us}
+\end{figure}
+
+We note a few key features of the sparsity plots. The first is that
+for multigroup problems without full upscatter and downscatter
+(i.e. Figure~\ref{fig:group10ds}), the resulting matrix is asymmetric
+and therefore a linear solver that can handle asymmetric linear
+systems is required. Nearly all problems of interest will not have
+full upscattering or downscattering. Second we note the largely
+diagonal character of these systems, although the blocks from
+Eq~(\ref{eq:A_block_matrix}) are readily apparent with a block size of
+$N_g\times(N+1)/2$.
+
+Based on this both block and diagonally dominant structure for
+matrices formed by the general multigroup $SP_N$ equations, we choose
+\textit{block Jacobi} preconditioning as a left preconditioner for the
+system. Block Jacobi preconditioning extracts the diagonal block
+elements of the matrix to use as the preconditioner as shown on the
+left side of Figure~\ref{fig:block_jacobi_ex}. Shown on the right side
+of Figure~\ref{fig:block_jacobi_ex}, inversion of the preconditioner
+is trivial with each diagonal block inverted separately.
+
+\begin{figure}[t!]
+  \begin{center}
+    \scalebox{1.6}{
+    \input{chapters/spn_equations/block_jacobi.pdftex_t} }
+  \end{center}
+  \caption{\textbf{Block Jacobi preconditioning strategy used for the
+      $SP_N$ equations.} \textit{Left: The preconditioner is formed by
+      the diagonal blocks of the matrix. Right: Inversion of the
+      preconditioner is trivial and decoupled by block.}}
+  \label{fig:block_jacobi_ex}
+\end{figure}
+
+For high performance implementations block Jacobi preconditioning has
+several attractive properties. First, the blocks in the transport
+operator come from the energy/angle discretization of the transport
+equation as given by Eq~(\ref{eq:A_block_matrix}). Each block on the
+diagonal is bound to a mesh element in the system (note there are 8
+blocks on the diagonal in each of the sparsity patterns with a mesh of
+$2 \times 2 \times 2$) and therefore we expect the matrix elements
+forming the block to be entirely local in domain decomposed
+calculations. Second, these blocks are typically dense and nearly
+lower triangular for many transport problems meaning that established
+dense matrix methods can be used for fast inversion.
+
+\subsection{Block Jacobi Preconditioned Results}
 \label{subsec:jacobi_prec_assembly_calc}
-We first use block Jacobi preconditioning to solve the $17 \times 17$
-fuel assembly problem. A single energy group problem was first
-computed with $SP_1$ discretization, effectively giving the one-speed
-neutron diffusion system for the fuel assembly resulting in 20,088
-degrees of freedom in the
-problem. Figure~\ref{fig:block_jacobi_res_mcsa} gives the residual
+A single energy group problem was first computed with $SP_1$
+discretization, effectively giving the one-speed neutron diffusion
+system for the fuel assembly resulting in 20,088 degrees of freedom in
+the problem. Figure~\ref{fig:block_jacobi_res_mcsa} gives the residual
 infinity norm as a function of iteration for the MCSA linear solve in
@@
 used with the $SP_N$ form of the neutron transport equation and
-applied to a difficult fuel assembly criticality problem. Along with
-these preconditioning techniques, a set a varying MCSA parameters were
-analyzed to find those that yielded the best performance for this
-particular problem. In addition, several steps were taken to mitigate
-the dense composite linear operators that arise when performing
-explicit MCSA preconditioning necessary for achieve convergence. Using
-these results, we now compare MCSA to conventional Krylov solvers that
-would be typically used to solve the $SP_N$ form of the transport
-equation as presented here.
+applied to a difficult fuel assembly criticality problem. In addition,
+several steps were taken to mitigate the dense composite linear
+operators that arise when performing explicit MCSA preconditioning
+necessary for achieve convergence. Using these results, we now compare
+MCSA to conventional Krylov solvers that would be typically used to
+solve the $SP_N$ form of the transport equation as presented here.
 
@@
 minimal residual-based MCSA iteration given by
-Eq~(\ref{eq:mcsa_min_res}) will be
-used. Table~\ref{tab:spn_solver_defs} gives definitions for the
+Eq~(\ref{eq:mcsa_min_res}) will be used. A supplementary performance
+analysis for this set of MCSA variations is presented in
+Appendix~\ref{chap:spn_estimator_comparison}.
+
+Table~\ref{tab:spn_solver_defs} gives definitions for the
 solvers used to generate the results in the remainder of this
@@
 group 0).
+
 \begin{table}[h!]
@@
 \end{table}
+
 \begin{table}[h!]

### 2013-08-22 doc/dissertation/document/backmatter/mcsa_artificial_absorption.tex
moving articial absorption into an appendix and the reduced domain approximation into the spn chapter

--- 2013-08-19 doc/dissertation/document/chapters/mc_background/mc_background.tex
@@
-\chapter{Monte Carlo Synthetic Acceleration\\ Methods}
-\label{ch:stochastic_methods}
-An alternative approach to approximate matrix inversion is to employ
-Monte Carlo methods that sample a distribution with an expectation
-value equivalent to that of the inverted operator. Such methods have
-been in existence for decades with the earliest reference noted here a
-manuscript published in 1950 by Forsythe and Leibler
-\cite{forsythe_matrix_1950}. In their outline, Forsythe and Leibler
-in fact credit the creation of this technique to J. Von Neumann and
-S.M. Ulam some years earlier than its publication. In 1952 Wasow
-provided a more formal explanation of Von Neumann and Ulam's method
-\cite{wasow_note_1952} and Hammersley and Handscomb's 1964 monograph
-\cite{hammersley_monte_1964} and Spanier and Gelbard's 1969 book
-\cite{spanier_monte_1969} present additional detail on this topic
-using a collection of references from the 1950's and early
-1960's.
-
-Following this work, Halton presented a residual Monte Carlo scheme
-that improved dramatically on the performance of these methods
-\cite{halton_sequential_1962}. Recent work in this area has utilized
-these residual Monte Carlo methods in particle transport
-\cite{evans_residual_2003}, giving accelerated convergence over
-traditional methods. As an expansion of Halton's work, Evans and
-others have leveraged these methods in a synthetic acceleration scheme
-that has been shown to be competitive with production Krylov methods
-for radiation transport problems and improve performance significantly
-over Halton's sequential method \cite{evans_monte_2012}.
-
-In this chapter, we will present the fundamentals of the Monte Carlo
-method for discrete linear systems. Both the forward and adjoint
-methods will be presented and analyzed including a brief variance
-analysis of several Monte Carlo estimators. The Sequential Monte Carlo
-method of Halton and Monte Carlo Synthetic Acceleration methods will
-then be presented as a means of leveraging the basic Monte Carlo
-methods devised by Neumann and Ulam in an iterative refinement scheme
-that accelerates their convergence. Using these iterative schemes, a
-set of modest variance reduction techniques will be introduced and
-their effects along with other parameters of the algorithms on the
-Monte Carlo solutions explored using a simple model transport problem.
-
-\section{Preliminaries}
-\label{sec:linear_preliminaries}
-We seek solutions of the general linear problem in the following form:
-\begin{equation}
-  \ve{A} \ve{x} = \ve{b}\:,
-  \label{eq:linear_problem}
-\end{equation}
-where $\ve{A} \in \mathbb{R}^{N \times N}$ is a matrix operator such
-that $\ve{A} : \mathbb{R}^{N} \rightarrow \mathbb{R}^{N}$, $\ve{x} \in
-\mathbb{R}^N$ is the solution vector, and $\ve{b} \in \mathbb{R}^N$ is
-the forcing term. The solutions to Eq~(\ref{eq:linear_problem}) will
-be generated by inverting $\ve{A}$ either directly or indirectly:
-\begin{equation}
-  \ve{x} = \ve{A}^{-1} \ve{b}
-  \label{eq:linear_problem_solution}\:.
-\end{equation}
-In addition we can define the residual:
-\begin{equation}
-  \ve{r} = \ve{b} - \ve{A}\ve{x}\:,
-  \label{eq:linear_residual}
-\end{equation}
-such that an exact solution $\ve{x}$ has been found when
-$\ve{r}=\ve{0}$.  From the statement in
-Eq~(\ref{eq:linear_problem_solution}) we can already place a
-restriction on $\ve{A}$ by requiring that it be \textit{nonsingular},
-meaning that we can in fact compute $\ve{A}^{-1}$. In this work we
-will focus our efforts on approximately inverting the operator through
-various means.
-
-In a discussion of methods for solving linear systems, several
-mathematical tools are useful in characterizing the qualities of the
-linear system. Among the most useful are the \textit{eigenvalues} of
-the matrix, $\sigma(\ve{A})$. We find these by solving the eigenvalue
-problem:
-\begin{equation}
-  \ve{A} \ve{x} = \lambda \ve{x},\ \lambda \in \sigma(\ve{A})\:.
-  \label{eq:eigenvalue_problem}
-\end{equation}
-By writing Eq~(\ref{eq:eigenvalue_problem}) in a different form,
-\begin{equation}
-  (\ve{A} - \lambda \ve{I})\ve{x} = 0 \:,
-  \label{eq:eigenvalue_problem_2}
-\end{equation}
-and demanding that non-trivial solutions for $\ve{x}$ exist, it is
-then required that $|\ve{A} - \lambda \ve{I}| = 0$. Expanding this
-determinant yields a characteristic polynomial in terms of $\lambda$
-with roots that form the set of eigenvalues, $\sigma(\ve{A})$. Each
-component of $\sigma(\ve{A})$ can then be used to solve
-Eq~(\ref{eq:eigenvalue_problem_2}) for a particular permutation of
-$\ve{x}$. The set of all permutations form the \textit{eigenvectors}
-of $\ve{A}$. A quantity of particular interest that is computable from
-the eigenvalues of a matrix $\ve{A}$ is the \textit{spectral radius},
-$\rho(\ve{A})$, defined by Saad \cite{saad_iterative_2003} as:
-\begin{equation}
-  \rho(\ve{A}) = \max_{\lambda \in \sigma(\ve{A})} |\lambda| \:.
-  \label{eq:spectral_radius}
-\end{equation}
-
-%%---------------------------------------------------------------------------%%
-\section{Neumann-Ulam Method}
-\label{sec:mc_preliminaries}
-We begin our discussion of Monte Carlo methods by seeking a solution
-to Eq~(\ref{eq:linear_problem}). For a given linear operator $\ve{A}$,
-we can use diagonal splitting similar to the stationary method in
-Eq~(\ref{eq:linear_split_equation2}) to define the following
-operator\footnote{It should be noted that non-diagonal splittings have
-  been recently explored \cite{srinivasan_monte_2010} and have the
-  potential to improve efficiency. However, it was observed in this
-  work that this type of splitting did not improve performance in the
-  asymptotic limit of $\rho(\ve{H}) \rightarrow 1$ for non-trivial
-  problems.}:
-\begin{equation}
-  \ve{H} = \ve{I} - \ve{A}\:,
-  \label{eq:linear_mc_iteration_matrix}
-\end{equation}
-such that we are solving the system:
-\begin{equation}
-  \ve{x} = \ve{H} \ve{x} + \ve{b}\:.
-  \label{eq:richardson_split}
-\end{equation}
-We can then form an alternative representation for $\ve{A}^{-1}$ by
-generating the \textit{Neumann series}:
-\begin{equation}
-  \ve{A}^{-1} = (\ve{I}-\ve{H})^{-1} = \sum_{k=0}^{\infty} \ve{H}^k\:,
-  \label{eq:neumann_series}
-\end{equation}
-which will converge if the spectral radius of $\ve{H}$ is less than
-1. If we then apply this Neumann series to the right hand side of
-Eq~(\ref{eq:linear_problem}) we acquire the solution to the linear
-problem:
-\begin{equation}
-  \ve{A}^{-1}\ve{b} = \sum_{k=0}^{\infty} \ve{H}^k\ve{b} = \ve{x}\:.
-  \label{eq:neumann_solution}
-\end{equation}
-An approximation of this summation by truncation will therefore lead
-to an approximation of the solution. If we expand the summation with a
-succession of matrix-vector multiply operations, we arrive at an
-alternative perspective of this summation by considering the $i^{th}$
-component of the solution vector:
-\begin{equation}
-  x_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-  \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}\:,
-  \label{eq:expanded_neumann_solution}
-\end{equation}
-which can interpreted as a series of transitions between states,
-\begin{equation}
- \nu = i \rightarrow i_1 \rightarrow \cdots \rightarrow i_{k-1}
- \rightarrow i_{k}\:,
-  \label{eq:mc_walk_permutation}
-\end{equation}
-in $\ve{H}$ where $\nu$ is interpreted as a particular random walk
-sequence permutation. We can generate these sequences of transitions
-through Monte Carlo random walks by assigning them both a probability
-and weight. As a reinterpretation of the iteration matrix, we then
-form the \textit{Neumann-Ulam decomposition} of \ve{H}:
-\begin{equation}
-  \ve{H} = \ve{P} \circ \ve{W}\:,
-  \label{eq:neumann_ulam_decomposition}
-\end{equation}
-where $\circ$ denotes the Hadamard product operation\footnote{The
-  Hadamard product $\ve{A} = \ve{B} \circ \ve{C}$ is defined
-  element-wise as $a_{ij} = b_{ij} c_{ij}$.}, $\ve{P}$ denotes the
-transition probability matrix, and $\ve{W}$ denotes the transition
-weight matrix. This decomposition, a generalization of Dimov's work
-\cite{dimov_new_1998}, is an extension of the original Neumann-Ulam
-scheme in that now a weight cutoff can be used to terminate a random
-walk sequence and therefore truncate the Neumann series it is
-approximating. The formulation of $\ve{P}$ and $\ve{W}$ will be
-dependent on whether we choose a direct or adjoint Monte Carlo
-sequence to estimate the state transitions in
-Eq~(\ref{eq:expanded_neumann_solution}). In the direct method, we will
-use the provided linear operator $\ve{A}$ to form the Neumann Ulam
-decomposition while the adjoint method will use the adjoint linear
-operator ($\ve{A}^T$ for real-valued systems) to form the
-decomposition.
-
-%%---------------------------------------------------------------------------%%
-\section{Direct Monte Carlo Method}
-\label{sec:direct_mc}
-In the context of matrix inversion, a direct (forward) method
-resembles an adjoint Monte Carlo method in the reactor physics
-community where the solution state is sampled and the source terms
-that contribute to it are assembled. To achieve this, we build the
-forward Neumann-Ulam decomposition per Dimov's approach by first
-choosing a probability matrix that is a row scaling of $\ve{H}$ such
-that its components are:
-\begin{equation}
-  p_{ij} = \frac{|h_{ij}|}{\sum_j |h_{ij}|}\:.
-  \label{eq:direct_probability}
-\end{equation}
-From this, we then see that the probability of transitioning from a
-state $i$ to a state $j$ is implicitly linked to the original operator
-$\ve{A}$ in that those terms with large values, and therefore those
-that make the greatest contribution to the numerical solution, will be
-sampled with a higher probability than smaller terms. In addition, the
-row scaling provides a normalization over the state to which we are
-transitioning such that $\sum_j p_{ij} = 1$, meaning that we sample
-the probabilities over the rows of the matrix. The components of
-the weight matrix are then defined by
-Eq~(\ref{eq:neumann_ulam_decomposition}) as:
-\begin{equation}
-  w_{ij} = \frac{h_{ij}}{p_{ij}}\:.
-  \label{eq:direct_weight}
-\end{equation}
-It should be noted here that if $\ve{A}$ is sparse, then $\ve{H}$,
-$\ve{P}$, and $\ve{W}$ must be sparse as well by
-definition. Additionally, we only compute $\ve{P}$ and $\ve{W}$ from
-the non-zero elements of $\ve{H}$ as those components that are zero
-will not participate in the random walk and thus produce identical
-sparsity patterns for all matrices.
-
-Using these matrices, we can then form the expectation value of the
-forward solution. For a given random walk permutation $\nu$, we define
-the weight of that permutation on the $m^{th}$ step to be:
-\begin{equation}
-  W_{m} = w_{i_0,i_1} w_{i_1,i_2} \cdots w_{i_{m-1},i_m}\:,
-  \label{eq:direct_permutation_weight}
-\end{equation}
-such that the weight of each transition event contributes to the total
-through multiplication with $W_0 = 1$ as the starting weight. The
-contribution to the solution from a particular random walk permutation
-with $k$ total events is then the \textit{forward estimator}:
-\begin{equation}
-  X_{i_0 = i}(\nu) = \sum_{m=0}^k W_{m} b_{i_m}\:,
-  \label{eq:direct_permutation_contribution}
-\end{equation}
-where $X_{i_0 = i}(\nu)$ signifies that the solution state, $i_0$, in
-which the random walk $\nu$ started is also the state, $i$, in which
-we are tallying. We then define the probability that a particular
-random walk permutation of $k$ events will occur:
-\begin{equation}
-  P_{\nu} = p_{i,i_1} p_{i_1,i_2} \cdots p_{i_{k-1},i_k}\:.
-  \label{eq:direct_permutation_probability}
-\end{equation}
-Finally, we define the expectation value of $X$ to be the collection
-of all random walk permutations and their probabilities:
-\begin{equation}
-  E\{X_i\} = \sum_{\nu} P_{\nu} X_{i}(\nu)\:,
-  \label{eq:direct_expectation_value}
-\end{equation}
-which, if expanded, directly recovers the exact solution by forming
-the Neumann series:
-\begin{equation}
-  \begin{split}
-    E\{X_i\}
-    &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-    \sum_{i_k}^{N} p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
-    w_{i,i_1}w_{i_1,i_2}\ldots w_{i_{k-1},i_k} b_{i_k}\\ &= x_i\:,
-  \end{split}
-  \label{eq:direct_expectation_expansion}
-\end{equation}
-therefore providing an unbiased Monte Carlo estimator.
-
-In cases where we seek only approximate solutions, we need only to
-perform a predetermined number of random walks in order to generate an
-approximation for $\ve{x}$. If we are only to approximate the
-solution, we also need conditions by which we may terminate a random
-walk as the Neumann Ulam decomposition defined by
-Eqs~(\ref{eq:direct_probability}) and (\ref{eq:direct_weight}) will
-create a random walk weight in Eq~(\ref{eq:direct_permutation_weight})
-that approaches, but never reaches zero. We do this by noticing that
-if the problem is convergent, the factors added to
-Eq~(\ref{eq:direct_permutation_weight}) will make $W_m$ become
-diminishingly small due to their definition in
-Eq~(\ref{eq:direct_weight}) and therefore the contribution of the
-history to the solution estimate will become negligible. Using this,
-we choose terminate a random walk sequence with a \textit{weight
-  cutoff}, $W_c$, that is enforced when $W_m < W_c$ for a particular
-random walk permutation.
-
-\subsection{Forward Estimator Variance}
-\label{subsec:forward_variance}
-We can compute the variance of the forward estimator through traditional
-methods by defining the variance, $\sigma_i$, for each component in
-the solution:
-\begin{equation}
-  {\sigma_i}^2 = E\{X_i - (\ve{A}^{-1}\ve{b})_i\}^2 = E\{X_i^2\} - x_i^2\:,
-  \label{eq:direct_variance_1}
-\end{equation}
-where the vector exponentials are computed element-wise. Inserting
-Eq~(\ref{eq:direct_expectation_value}) gives:
-\begin{equation}
-  \sigma_i^2 = \sum_{\nu} P_{\nu} X_{i}(\nu)^2 - x_i^2\:,
-  \label{eq:direct_variance_2}
-\end{equation}
-and applying Eq~(\ref{eq:direct_permutation_contribution}) and
-expanding the square of the summation:
-\begin{equation}
-  \sigma_i^2 = \sum_{\nu} P_{\nu} \Big(\sum_{m=0}^k \Big[ W_{m}^2 b_{i_m}^2
-  + 2 \sum_{j=0}^{m-1} W_m W_j b_{i_m} b_{i_j} \Big] \Big)-
-  x_i^2\:.
-  \label{eq:direct_variance_3}
-\end{equation}
-We can distribute the random walk probability to yield an expanded
-form of the variance:
-\begin{equation}
-  \sigma_i^2 = \sum_{\nu} P_{\nu} \sum_{m=0}^k W_{m}^2 b_{i_m}^2 + 2
-  \sum_{\nu} P_{\nu} \sum_{\substack{ j=0 \\ m<j}}^k W_m W_j b_{i_m}
-  b_{i_j} - x_i^2\:.
-  \label{eq:direct_variance_3_2}
-\end{equation}
-In particular, we will isolate the first summation of the variance by
-expanding it:
-\begin{equation}
-  \sum_{\nu} P_{\nu} \sum_{m=0}^k W_{m}^2 b_{i_m}^2 =
-  \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots \sum_{i_k}^{N}
-  p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
-  w^2_{i,i_1}w^2_{i_1,i_2}\ldots w^2_{i_{k-1},i_k} b_{i_k}^2\:.
-  \label{eq:direct_variance_4}
-\end{equation}
-Using this expansion, we can arrive at a more natural
-reason for enforcing $\rho(\ve{H}) < 1$ for our Monte Carlo method to
-converge. Per the Hadamard product, we can concatenate the summation
-in Eq~(\ref{eq:direct_variance_4}):
-\begin{equation}
-  (\ve{P} \circ \ve{W} \circ \ve{W})^k =
-  \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots \sum_{i_k}^{N}
-  p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
-  w^2_{i,i_1}w^2_{i_1,i_2}\ldots w^2_{i_{k-1},i_k}\:.
-  \label{eq:double_weighted_decomposition}
-\end{equation}
-If we assign $\ve{G} = \ve{P} \circ \ve{W} \circ \ve{W}$ as in
-Eq~(\ref{eq:neumann_ulam_decomposition}), we then have:
-\begin{equation}
-   (\ve{P} \circ \ve{W} \circ \ve{W})^k =
-  \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-  \sum_{i_k}^{N}g_{i,i_1}g_{i_1,i_2}\ldots g_{i_{k-1},i_k}\:,
-  \label{eq:direct_variance_6}
-\end{equation}
-which is the general Neumann series for $\ve{G}$,
-\begin{equation}
-  \ve{T} = \sum_{k=0}^{\infty} \ve{G}^k\:,
-  \label{eq:variance_neumann_series}
-\end{equation}
-where $\ve{T} = (\ve{I}-\ve{G})^{-1}$. We can then insert $T$ back
-into the variance formulation for a more concise definition:
-\begin{equation}
-  \sigma^2_i = (\ve{T}\ve{b}^2)_i + 2 \sum_{\nu} P_{\nu}
-  \sum_{\substack{ j=0 \\ m<j}}^k W_m W_j b_{i_m} b_{i_j} - x_i^2\:.
-  \label{eq:direct_variance_5}
-\end{equation}
-We can relate $\ve{G}$ to $\ve{H}$ by noting that $\ve{G}$ simply
-contains an additional Hadamard product with the weight matrix. The
-Hadamard product has the property that:
-\begin{equation}
-  |\ve{H} \circ \ve{W}| \geq |\ve{H}|\ |\ve{W}|\:.
-  \label{eq:hadamard_inequality}
-\end{equation}
-Using the norm property of the Hadamard product and
-Eq~(\ref{eq:neumann_ulam_decomposition}), we can define the norm of
-$\ve{W}$ as:
-\begin{equation}
-  \frac{|\ve{H}|}{|\ve{P}|} \geq |\ve{W}|\:.
-  \label{eq:weight_norm}
-\end{equation}
-Choosing the infinity norm of the operator, the row normalized
-probability matrix will yield a norm of 1 giving the following
-inequality for relating $\ve{G}$ and $\ve{H}$:
-\begin{equation}
-  |\ve{G}| \geq |\ve{H}|^2
-  \label{eq:variance_norm_inequality}
-\end{equation}
-Using these relations to analyze Eq~(\ref{eq:direct_variance_5}), we
-see that if $\rho(\ve{G}) > 1$, then the first sum in
-Eq~(\ref{eq:variance_neumann_series}) will not converge and an
-infinite variance will arise as the elements of $\ve{T}$ become
-infinite in Eq~(\ref{eq:direct_variance_5}). We must restrict $\ve{G}$
-to alleviate this and therefore restrict $\ve{H}$ due to
-Eq~(\ref{eq:variance_norm_inequality}) with $\rho(\ve{H}) < 1$ so that
-our expectation values for the solution may have a finite variance.
-
-\subsection{Direct Method: Evolution of a Solution}
-\label{subsec:direct_evolution}
-As a means of visually demonstrating the direct Monte Carlo method,
-consider a 2-dimensional thermal diffusion problem with sources on the
-left and right hand sides of the domain and a uniform source
-throughout the domain of 1/5 the strength of the boundary sources as
-shown in Figure~\ref{fig:heat_setup}.
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.2}{ \input{chapters/mc_background/heat_eq_setup.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Problem setup for 2D heat equation.}
-    \textit{Dirichlet conditions are set for the temperature on all 4
-      boundaries of the Cartesian grid. Background source of 1/5 the
-      value of the boundary sources present. $50 \times 50$ grid.}}
-  \label{fig:heat_setup}
-\end{figure}
-For this problem, the number of histories used to compute the solution
-at each grid point (state) in the domain was increased from 1 to 1000
-in order to demonstrate the effects on the solution and the
-statistical nature of the method. Figure~\ref{fig:direct_evolution}
-gives these results.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/mc_background/direct_evolution.png}
-  \end{center}
-  \caption{\textbf{Direct Monte Carlo solution to the heat equation
-      with varying numbers of histories.} \textit{Top left: 1 history
-      per state. Top right: 10 histories per state. Bottom left: 100
-      histories per state. Bottom right: 1000 histories per state.}}
-  \label{fig:direct_evolution}
-\end{figure}
-As the number of histories used per state is increased, the
-statistical variance of the solutions is decreased as more tallies are
-made. Starting with a single history at each state in the domain, the
-high variance prevents a precise solution from being obtained although
-we begin to see the solution take shape as expected. It is interesting
-to note here that as the statistical uncertainty is reduced at each
-grid point in the domain, the solution is resolved in a certain sense,
-analogous to the convergence of a traditional iterative method.
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{Adjoint Monte Carlo Method}
-\label{sec:adjoint_mc}
-Often a more useful form, an alternative to forward Monte Carlo matrix
-inversion is the adjoint method. We begin by defining the linear
-system adjoint to Eq~(\ref{eq:linear_problem}):
-\begin{equation}
-  \ve{A}^T \ve{y} = \ve{d}\:,
-  \label{eq:adjoint_linear_problem}
-\end{equation}
-where $\ve{y}$ and $\ve{d}$ are the adjoint solution and source
-vectors respectively and $\ve{A}^T$ is the adjoint operator for
-$\ve{A} \in \mathbb{R}^{N \times N}$. We can split this equation to
-mirror Eq~(\ref{eq:richardson_split}):
-\begin{equation}
-  \ve{y} = \ve{H}^T \ve{y} + \ve{d}\:.
-  \label{eq:adjoint_split_system}
-\end{equation}
-As was required for convergence with the direct method using
-Eq~(\ref{eq:richardson_split}), the spectral radius of $\ve{H}$ must
-remain less than 1 as $\ve{H}^T$ contains the same eigenvalues and
-therefore has the same spectral radius. By defining the following
-inner product equivalence \cite{spanier_monte_1969}:
-\begin{equation}
-  \langle \ve{A}^T \ve{y}, \ve{x} \rangle = \langle \ve{y}, \ve{A}
-  \ve{x} \rangle\:.
-  \label{eq:adjoint_operator_product}
-\end{equation}
-it follows that:
-\begin{equation}
-  \langle \ve{x}, \ve{d} \rangle = \langle \ve{y}, \ve{b} \rangle\:.
-  \label{eq:adjoint_vector_relation}
-\end{equation}
-Using these definitions, we can derive an estimator from the adjoint
-method that will also give the solution vector, $\ve{x}$. As with the
-direct method, we can acquire the adjoint solution by forming the
-Neumann series by writing Eq~(\ref{eq:adjoint_split_system}) as:
-\begin{equation}
-  \ve{y} = (\ve{I} - \ve{H}^T)^{-1} \ve{d}\:,
-  \label{eq:adjoint_split_system_2}
-\end{equation}
-which in turn yields the Neumann series using the adjoint operator:
-\begin{equation}
-  \ve{y} = \sum_{k=0}^{\infty} (\ve{H}^T)^k\ve{d}\:.
-  \label{eq:adjoint_neumann_series}
-\end{equation}
-We expand this summation to again yield a series of transitions that
-can be approximated by a Monte Carlo random walk sequence, this time
-forming the Neumann series in reverse order:
-\begin{equation}
-  y_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-  \sum_{i_k}^{N}h_{i_k,i_{k-1}}\ldots h_{i_2,i_1} h_{i_1,i} d_{i_k}\:.
-  \label{eq:adjoint_neumann_solution}
-\end{equation}
-We can readily build an estimator for the adjoint solution from this
-series expansion, but we instead desire the solution to
-Eq~(\ref{eq:linear_problem}). Here we have 2 unknowns, $\ve{y}$ and
-$\ve{d}$, and therefore we require two constraints to close the
-system. We use Eq~(\ref{eq:adjoint_vector_relation}) as the first
-constraint and as a second constraint we select:
-\begin{equation}
-  \ve{d} = \boldsymbol{\delta}_j\:,
-  \label{eq:adjoint_second_constraint}
-\end{equation}
-where $\boldsymbol{\delta}_j$ is one of a set of vectors in which the
-$j^{th}$ component is the Kronecker delta function $\delta_{i,j}$. If
-we apply Eq~(\ref{eq:adjoint_second_constraint}) to our first
-constraint Eq~(\ref{eq:adjoint_vector_relation}), we get the following
-convenient outcome:
-\begin{equation}
-  \langle \ve{y}, \ve{b} \rangle = \langle \ve{x},
-  \boldsymbol{\delta}_j \rangle = x_j \:,
-  \label{eq:inner_product_constraint}
-\end{equation}
-meaning that if we compute the inner product of the original source
-and the adjoint solution using a delta function source, we recover one
-component of the original solution.
-
-In terms of radiation transport, this adjoint method is equivalent to
-a traditional forward method where the initial state $i_0$ of the
-random walk is determined by sampling the source vector $\ve{b}$ with
-probabilities:
-\begin{equation}
-  P_{i_0=i}(\nu) = \frac{|b_i|}{||\ve{b}||_1}\:.
-  \label{eq:adjoint_source_probability}
-\end{equation}
-The random walk starting weight will then be $W_0 = b_{i_0}$.  As a
-result of using the adjoint system, we modify our probabilities and
-weights using the \textit{adjoint Neumann-Ulam decomposition} of
-$\ve{H}$:
-\begin{equation}
-  \ve{H}^{T} = \ve{P} \circ \ve{W}\:,
-  \label{eq:adjoint_neumann_ulam}
-\end{equation}
-where now we are forming the decomposition with respect to the
-transpose of $\ve{H}$. We then follow the same procedure as the direct
-method for forming the probability and weight matrices in the
-decomposition. Using the adjoint form, probabilities should instead be
-column-scaled:
-\begin{equation}
-  p_{ij} = \frac{|h_{ji}|}{\sum_j |h_{ji}|}\:,
-  \label{eq:adjoint_probability}
-\end{equation}
-such that we expect to select a new state, $j$, from the current state
-in the random walk, $i$, by sampling column-wise (or row-wise if an
-adjoint probability matrix is formed). Per
-Eq~(\ref{eq:adjoint_neumann_ulam}), the transition weight is then
-defined as:
-\begin{equation}
-  w_{ij} = \frac{h_{ji}}{p_{ij}}\:.
-  \label{eq:adjoint_weight}
-\end{equation}
-Using the decomposition we can then define an expectation value for
-the adjoint method. Given Eq~(\ref{eq:direct_permutation_weight}) as
-the weight generated for a particular random walk permutation as in
-Eq~(\ref{eq:mc_walk_permutation}) and our result from
-Eq~(\ref{eq:inner_product_constraint}) generated by applying the
-adjoint constraints, the contribution to the solution in state $j$
-from a particular random walk permutation of $k$ events is then the
-\textit{collision estimator}:
-\begin{equation}
-  X_{j}(\nu) = \sum_{m=0}^k W_{m} \delta_{i_m,j}\:,
-  \label{eq:adjoint_permutation_contribution}
-\end{equation}
-where the Kronecker delta indicates that the tally contributes only in
-the current state, $i_m$, of the random walk.  Note here that the
-estimator in Eq~(\ref{eq:adjoint_permutation_contribution}) does not
-have a dependency on the source state as in
-Eq~(\ref{eq:adjoint_permutation_contribution}), providing a remedy for
-the situation in the direct method where we must start a random walk
-in each source state for every permutation if we want to compute a
-solution estimate for that state. In the adjoint method, we instead
-tally in all states and those of lesser importance will not be visited
-as frequently by the random walk. Finally, the expectation value using
-all permutations is:
-\begin{equation}
-  E\{X_j\} = \sum_{\nu} P_{\nu} X_{j}(\nu)\:
-  \label{eq:adjoint_expectation_value}
-\end{equation}
-which, if expanded in the same way as the direct method, directly
-recovers the exact solution:
-\begin{equation}
-  \begin{split}
-    E\{X_j\} &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-    \sum_{i_k}^{N} b_{i_0} h_{i_0,i_1}h_{i_1,i_2}\ldots
-    h_{i_{k-1},i_k} \delta_{i_k,j} \\ &= x_{j}\:,
-  \end{split}
-  \label{eq:adjoint_expectation_expansion}
-\end{equation}
-therefore also providing an unbiased Monte Carlo estimate of the
-solution. Note that this expansion produces the effective sequence of
-matrix-vector multiplications with the $b_{i_0}$ component of the
-source vector which will be selected at random for each walk
-permutation and the unbiasedness of the estimator relies on an
-unbiased sampling of the source. It should also be noted here that
-Eq~(\ref{eq:adjoint_expectation_expansion}) only computes a single
-component of our desired solution vector when really what we desire is
-the entire solution vector. In an adjoint Monte Carlo simulation using
-this estimator, the $w_{ij}$ elements that are added into the tally
-for each state are only selected if the random walk currently resides
-in that state. Much like a mesh tally in a particle transport
-simulation, we have $N$ simultaneous tallies for $\ve{A} \in
-\mathbb{R}^{N \times N}$ that will yield the entire solution
-vector. Based on their current state in the system, the random walks
-will contribute tally $j$ in this group.
-
-Like the direct method, we also desire a criteria for random walk
-termination for problems where only an approximate solution is
-necessary. For the adjoint method, we utilize a \textit{relative
-  weight cutoff}:
-\begin{equation}
-  W_f = W_c b_{i_0}\:,
-  \label{eq:relative_weight_cutoff}
-\end{equation}
-where $W_c$ is defined as in the direct method. The adjoint random
-walk will then be terminated after $m$ steps if $W_m < W_f$ as tally
-contributions become increasingly small.
-
-\subsection{Collision Estimator Variance}
-\label{subsec:forward_variance}
-We can compute the variance of the collision estimator in the same way
-as the direct estimator for each component in the solution where now:
-\begin{equation}
-  \sigma_j^2 = \sum_{\nu} P_{\nu} \sum_{m=0}^k W_{m}^2 \delta_{i_m,j} + 2
-  \sum_{\nu} P_{\nu} \sum_{\substack{ l=0 \\ m<l}}^k W_m W_l
-  \delta_{i_m,j} \delta_{i_l,j} - x_j^2\:.
-  \label{eq:collision_variance_1}
-\end{equation}
-In this case, in the second summation $m \neq l$ for all states due to
-the form of the summation and therefore the delta functions,
-$\delta_{i_m,j}$ or $\delta_{i_l,j}$, will only be nonzero for random
-walks that are in the current solution state (when $i_m = i_l = j$)
-and other states visited do not contribute to the variance of the
-current state (when $i_m \neq i_l$). This is important as it shows a
-decoupling of the variance among the solution vector states, meaning
-that the variance in solution state $i$ does not depend on the
-variance in solution state $j$.
-
-Expanding the transition probabilities in the first sum again yields a
-Neumann series in the same form as that explored for the forward
-estimator, this time with the terms in reverse order and the
-introduction of the Kronecker delta:
-\begin{multline}
-  \sigma_j^2 = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-  \sum_{i_k}^{N} p_{i_k,i_{k-1}}\ldots p_{i_2,i_1} p_{i_1,i_0}
-  w^2_{i_k,i_{k-1}}\ldots w^2_{i_2,i_1} w^2_{i_1,i_0}
-  b^2_{i_0}\delta_{i_k,j} + \\ 2 \sum_{\nu} P_{\nu} \sum_{\substack{
-      l=0 \\ m<l}}^k W_m W_l \delta_{i_m,j} \delta_{i_l,j} - x_j^2\:.
-  \label{eq:collision_variance_2}
-\end{multline}
-If we again define $\ve{G} = \ve{P} \circ \ve{W} \circ \ve{W}$, we
-then have:
-\begin{multline}
-  \sigma^2_j = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-  \sum_{i_k}^{N}g_{i_{k},i_{k-1}} \ldots
-  g_{i_2,i_1}g_{i_1,i_0}b_{i_0}\delta_{i_k,j} + \\ 2 \sum_{\nu}
-  P_{\nu} \sum_{\substack{ l=0 \\ m<l}}^k W_m W_l \delta_{i_m,j}
-  \delta_{i_l,j} - x_j^2\:,
-\label{eq:collision_variance_3}
-\end{multline}
-where now $\ve{G}$ in this case is the transpose of that used in
-Eq~(\ref{eq:direct_variance_6}). The Kronecker delta again implies
-that the variance contribution from each random walk will only be in
-its current state and for many random walks, many starting weights of
-$b_{i_0}$ will contribute to the variance in the same way as they
-contribute to the solution tally. As the transpose is formed and the
-Neumann series of $\ve{G}$ is in reverse order in
-Eq~(\ref{eq:collision_variance_3}) relative to its formulation in the
-forward estimator, then Eq~(\ref{eq:collision_variance_3}) and
-Eq~(\ref{eq:direct_variance_6}) are equivalent and therefore the
-collision estimator is bound by the same restrictions on
-$\rho(\ve{G})$ and $\rho(\ve{H})$ to ensure that the variance is
-finite.
-
-\subsection{Expected Value Estimator}
-\label{subsec:expected_value_estimator}
-In addition to the collision estimator, an additional estimator is
-available due to the work of Okten \cite{okten_solving_2005} that
-uses the method of expected values as a means to improve the Monte
-Carlo estimate. As outlined by Spanier and Gelbard
-\cite{spanier_monte_1969}, the method of expected values is a
-deterministic averaging of events that may potentially occur in the
-Monte Carlo random walk sequence. Okten applied this principle
-directly to discrete Monte Carlo by forming the \textit{expected value
-  estimator} for a random walk of $k$ events:
-\begin{equation}
-  X_{j}(\nu) = b_j + \sum_{m=0}^k W_m h_{j,i_m}\,
-  \label{eq:expected_value_estimator}
-\end{equation}
-where now the contribution of the iteration matrix is
-deterministically averaged at step $m$ over all potential states $j$
-that may be reached from the current state $i_m$. Via Okten, the
-estimator can be shown to be unbiased through a comparison to
-the expected value estimator. We can first rewrite the summation in
-Eq~(\ref{eq:expected_value_estimator}):
-\begin{equation}
-  X_{j}(\nu) = b_j + \sum_{m=0}^k \sum_{i=1}^N W_m
-  \delta_{i_m,i} h_{ji}\,
-  \label{eq:unbiased_eval_1}
-\end{equation}
-where $N$ is the number of states in the system. Immediately, we see
-the collision estimator as defined by
-Eq~(\ref{eq:adjoint_permutation_contribution}) and can therefore write
-the expectation value as:
-\begin{equation}
-  E\{X_{j}\} = b_j + \sum_{i=1}^N E\{X_{i}\} h_{ji}\,
-  \label{eq:unbiased_eval_2}
-\end{equation}
-which is equivalently is the $j^{th}$ component of
-Eq~(\ref{eq:richardson_split}):
-\begin{equation}
-  E\{X_{j}\} = b_j + \sum_{i=1}^N x_{i} h_{ji}\,
-  \label{eq:unbiased_eval_2}
-\end{equation}
-and is therefore and unbiased estimate. Compared to the collision
-estimator, the expected value estimator provides additional
-information at every step of the random walk, yielding potentially
-better statistics with the same amount of transport
-work. Conveniently, even if no Monte Carlo histories are computed, the
-expected value estimator still deterministically computes the first
-term of the Neumann Series, $\ve{H}^0\ve{b}$, whereas the collision
-estimator will provide no information.
-
-\subsubsection{Expected Value Estimator Variance}
-\label{subsubsec:expected_value_estimator_variance}
-Following the collision estimator variance, the expected value
-estimator variance is given as:
-\begin{multline}
-  \sigma_j^2 = \sum_{\nu} P_{\nu} b_j^2 + 2 \sum_{\nu} P_{\nu} b_j
-  \sum_{m=0}^k W_{m} h_{j,i_m} +\\ \sum_{\nu} P_{\nu} \sum_{m=0}^k
-  W_{m}^2 h_{j,i_m}^2 + 2 \sum_{\nu} P_{\nu} \sum_{\substack{ l=0
-      \\ m<l}}^k W_m W_l h_{j,i_m} h_{j,i_l} - x_j^2\:.
-  \label{eq:expected_value_variance_1}
-\end{multline}
-As the delta functions are no longer present, the final summation in
-Eq~(\ref{eq:expected_value_variance_1}) contains contributions for all
-valid state combinations in the same row of the iteration matrix and
-therefore the variance of each state in the solution tally is coupled
-to a group of other states as defined by the sparsity pattern of the
-iteration matrix. This coupling of variances is expected due to the
-deterministic averaging used to generate the expected value estimator.
-
-\subsection{Adjoint Method: Evolution of a Solution}
-\label{subsec:adjoint_evolution}
-As a means of visually demonstrating the adjoint Monte Carlo method,
-again consider the 2-dimensional thermal diffusion problem with
-sources on the left and right hand sides of the domain and a smaller
-uniform source as shown in Figure~\ref{fig:heat_setup}. Using the
-adjoint method with the collision estimator, the number of histories
-sampled from the source was increased from 10 to 10,000,000 in order
-to show its effects on the solution and the statistical nature of the
-method. Figure~\ref{fig:adjoint_evolution} gives these results.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/mc_background/adjoint_evolution.png}
-  \end{center}
-  \caption{\textbf{Adjoint Monte Carlo solution to the heat equation
-      with varying numbers of histories.} \textit{Top left: 10
-      histories per state. Top right: 1,000 histories per
-      state. Bottom left: 100,000 histories per state. Bottom right:
-      10,000,000 histories per state.}}
-  \label{fig:adjoint_evolution}
-\end{figure}
-As the number of histories used per state is increased, the
-statistical variance of the solutions is decreased as more tally
-contributions are made. At 10,000,000 histories per state, enough
-information has been tallied to generate a reasonable estimate for the
-structure of the solution. The visual difference between
-Figures~\ref{fig:direct_evolution} and \ref{fig:adjoint_evolution} is
-precisely that determined by their mathematics. As the adjoint
-solution evolves with the addition of histories, more histories
-emanate from the boundary the smaller uniform source in the domain
-with more penetrating from the boundary into the domain and making
-contributions to the tallies in those states as they are transported.
-
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{Sequential Monte Carlo}
-\label{sec:sequential_mc}
-The direct and adjoint Neumann-Ulam methods described are limited by a
-convergence rate of $1/\sqrt{N}$ by the Central Limit Theorem where
-$N$ is the number of random walk permutations. In 1962, Halton
-presented a residual Monte Carlo method that moves towards exponential
-convergence rates \cite{halton_sequential_1962} and further refined
-his work some years later \cite{halton_sequential_1994}. Applications
-of his work by the transport community have confirmed convergence
-rates on the order of of $e^{-N}$ \cite{evans_residual_2003}. In much
-the same way as the projection methods outlined in
-Appendix~\ref{ch:linear_problem}, Halton's method, sequential Monte
-Carlo, utilizes the adjoint Monte Carlo solver as a means of directly
-reducing the elements residual vector. He proposed the following
-iterative scheme as a solution to Eq~(\ref{eq:linear_problem})\:
-\begin{subequations}
-  \begin{gather}
-    \ve{r}^k = \ve{b} - \ve{A}\ve{x}^k\:,\\  
-    \ve{A}\boldsymbol{\delta}^{k} = \ve{r}^{k}\:,\\
-    \ve{x}^{k+1} = \ve{x}^k + \boldsymbol{\delta}^{k}\:,
-  \end{gather}
-  \label{eq:sequential_monte_carlo}
-\end{subequations}
-where the correction $\boldsymbol{\delta}^k$ is computed by the
-adjoint Monte Carlo method at each iteration. The merits of Halton's
-approach are immediately visible in that we have now broken the
-binding of the convergence rate to the Central Limit Theorem. Here,
-the Monte Carlo solver is used to produce a correction from the
-residual, analogous to using the residual to extract a correction from
-the search subspace in a projection method. By doing this, the Monte
-Carlo error is bound in the correction used to update the solution and
-therefore does not explicitly manifest itself in the overall
-convergence of the solution. The downside of such a method is that if
-the solution guess is poor, then many iterations are required in order
-to reach exponential converge as the Monte Carlo error (and therefore
-the Central Limit Theorem) does dominate in this situation.
-
-%%---------------------------------------------------------------------------%%
-\section{Monte Carlo Synthetic Acceleration}
-\label{sec:mcsa}
-Using the ideas of Halton, Evans and Mosher recently developed a Monte
-Carlo solution method that was not prohibited severely by the quality
-of the initial guess for the system \cite{evans_monte_2009} and later
-applied it more rigorously as a solution mechanism for the radiation
-diffusion equation \cite{evans_monte_2012}. With their new methods,
-they achieved identical numerical results as conventional Krylov
-solvers as well as comparable performance in both number of iterations
-and CPU time. Their approach was instead to use residual Monte Carlo
-as a synthetic acceleration for a stationary method. To derive this
-method, we begin by splitting the operator in
-Eq~(\ref{eq:linear_problem})
-\begin{equation}
-  \ve{x} = (\ve{I} - \ve{A})\ve{x} + \ve{b}\:.
-  \label{eq:linear_split}
-\end{equation}
-With this we can then define the stationary method
-\textit{Richardson's iteration} as:
-\begin{equation}
-  \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,
-  \label{eq:richardsons_iteration}
-\end{equation}
-which will converge if $\rho(\ve{I} - \ve{A}) < 1$. We then define the
-solution error at the $k^{th}$ iterate relative to the true solution:
-\begin{equation}
-  \delta \ve{x}^k = \ve{x} - \ve{x}^k\:.
-  \label{eq:mcsa_error}
-\end{equation}
-Subtracting Eq~(\ref{eq:richardsons_iteration}) from
-Eq~(\ref{eq:linear_split}) we get:
-\begin{equation}
-  \delta \ve{x}^{k+1} = (\ve{I} - \ve{A})\delta \ve{x}^k\:.
-  \label{eq:mcsa_setup_1}
-\end{equation}
-Subtracting from this $(\ve{I} - \ve{A})\delta \ve{x}^{k+1}$ yields:
-\begin{equation}
-  \begin{split}
-    \ve{A}\delta \ve{x}^{k+1} &= (\ve{I} -
-    \ve{A})(\ve{x}^{k+1}-\ve{x}^{k}) \\ &= \ve{r}^{k+1}\:.
-    \label{eq:mcsa_setup_2}
-  \end{split}
-\end{equation}
-Using this, we define the following scheme that will converge in one
-iteration if $\ve{A}$ is inverted exactly:
-\begin{subequations}
-  \begin{gather}
-    \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,\\
-    \ve{A} \delta \ve{x}^{k+1} = \ve{r}^{k+1}\:,\\
-    \ve{x} = \ve{x}^{k+1} + \delta \ve{x}^{k+1}\:.
-  \end{gather}
-  \label{eq:mcsa_setup_3}
-\end{subequations}
-However, $\ve{A}$ is only approximately inverted by our numerical
-methods and therefore we instead pose an iterative scheme in which the
-Monte Carlo solvers are used to invert the operator. The
-\textit{Fixed-Point Monte Carlo Synthetic Acceleration} (MCSA) method
-is defined as:
-\begin{subequations}
-  \begin{gather}
-    \ve{x}^{k+1/2} = \ve{x}^k + \ve{r}^k\:,\\
-    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\
-    \ve{A}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:,\\
-    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,\\
-    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:,
-  \end{gather}
-  \label{eq:mcsa}
-\end{subequations}
-where a Neumann-Ulam Monte Carlo method is used to generate the
-solution correction from the residual and Richardson's iteration in
-the first step has been rewritten as a residual correction. Using
-Monte Carlo in this way achieves the same effect as Halton's method,
-decoupling its convergence rate from the overall convergence rate of
-the method. Here, the approximate Monte Carlo solution is not driven
-to a particular convergence as it merely supplies a correction for the
-initial guess generated by Richardson's iteration. Rather, only a set
-number of histories are required using the Neumann-Ulam method to
-generate the correction. In addition, the fact that the scheme in
-Eq~(\ref{eq:mcsa_setup_3}) will converge in one iteration if $\ve{A}$
-is inverted exactly means that as more and more stochastic histories
-are used to compute the correction and the error is reduced towards
-zero, the number of iterations required for MCSA to converge should
-decrease accordingly, thus accelerating the solution.
-
-In addition to the Monte Carlo solver parameters dictating the number
-of histories and weight cutoff, the outer MCSA iterations also have
-the following stopping criteria:
-\begin{equation}
-  ||\ve{r}||_\infty < \epsilon \ ||\ve{b}||_\infty\:,
-  \label{eq:mcsa_stopping_criteria}
-\end{equation}
-where $\epsilon$ is a user-defined parameter. As with any iterative
-method, other stopping criteria using other vector norms could be
-computed, however, for this work we will only use
-Eq~(\ref{eq:mcsa_stopping_criteria}). We therefore have 3 parameters
-to tune in an MCSA implementation: the number of Monte Carlo histories
-computed in the Neumann-Ulam solve during each MCSA iteration, the
-weight cutoff for those histories, and the total MCSA convergence
-tolerance as specified by $\epsilon$.
-
-\subsection{Alternative Fixed Point Iterations}
-\label{subsubsec:alternative_fixed_point}
-In addition to the basic Richardson iteration, the MCSA algorithm
-presented in Eq~(\ref{eq:mcsa}) can be used to accelerate any fixed
-point iteration which only depends on the state (typically the
-residual) of the last iteration. As outlined in
-Appendix~\ref{ch:linear_problem}, subspace methods take on a general
-form using the Petrov-Galerkin conditions as constraints for
-extracting a correction from the search subspace as given by
-Eq~(\ref{eq:linear_projection_iteration}). Those that are
-one-dimensional may be used with fixed-point MCSA as they only depend
-the previous iteration for information. As an example, for
-positive-definite but not necessarily symmetric problems, the minimal
-residual iteration \cite{saad_iterative_2003} can be used in
-conjunction with MCSA where the residual vector, $\mathbf{r}$, defines
-the search subspace and the action of the linear operator on the
-residual vector, $\mathbf{A}\mathbf{r}$, defines the constraint
-subspace. Using this, we can then define an MCSA scheme that
-accelerates the minimal residual iteration:
-\begin{subequations}
-  \begin{gather}
-    \alpha = \frac{\langle \mathbf{A}\mathbf{r}^k, \mathbf{r}^k
-      \rangle}{\langle \mathbf{A}\mathbf{r}^k, \mathbf{A}\mathbf{r}^k
-      \rangle}\:, \\
-    \ve{x}^{k+1/2} = \ve{x}^k + \alpha \ve{r}^k\:,
-    \label{eq:min_res_correction}\\
-    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\ 
-    \ve{A}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:,\\ 
-    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta\ve{x}^{k+1/2}\:,\\ 
-    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:.
-  \end{gather}
-  \label{eq:mcsa_min_res}
-\end{subequations}
-where $\alpha$ is an optimal extrapolation parameter generated from
-the constraints. Interestingly, this scheme is nearly identical to the
-Richardson iteration version in Eq~(\ref{eq:mcsa}) except now the
-additional extrapolation parameter, $\alpha$, is applied to the
-residual correction in Eq~(\ref{eq:min_res_correction}) as a means of
-selecting a more optimal search direction at each iteration,
-potentially further accelerating convergence.
-
-\subsection{Preconditioning MCSA}
-\label{subsec:stochastic_preconditioning}
-In most cases, at least a minimal amount of \textit{preconditioning}
-of the linear system will be required in order to use the class of
-stochastic methods described. Although these methods have no symmetry
-requirements for convergence, they do require that the spectral radius
-of the iteration matrix be less than one. Preconditioning serves as a
-means of achieving this by altering the eigenvalue spectrum of the
-iteration matrix.
-
-\subsubsection{Basic Preconditioning}
-\label{subsubsec:basic_mcsa_preconditioning}
-As an example of basic preconditioning, to achieve a spectral radius
-of less than one for diagonally dominant matrices point Jacobi
-preconditioning can be used such that the preconditioning matrix
-$\ve{M}$ is:
-\begin{equation}
-  \ve{M} = diag(\ve{A})\:,
-  \label{eq:jacobi_preconditioner}
-\end{equation}
-which may be trivially inverted. With the application of this
-preconditioner we are instead solving the following scaled linear
-system:
-\begin{equation}
-  \ve{M}^{-1}\ve{A}\ve{x} = \ve{M}^{-1}\ve{b}\:.
-  \label{eq:jacobi_precond_linear_problem}
-\end{equation}
-Next, we can apply MCSA to solve
-Eq~(\ref{eq:jacobi_precond_linear_problem}): 
-\begin{subequations}
-  \begin{gather}
-    \ve{x}^{k+1/2} = \ve{x}^k +
-    \ve{M}^{-1}\ve{r}^k\:,\\ \ve{r}^{k+1/2} =
-    \ve{b}-\ve{A}\ve{x}^{k+1/2}\:,\\ \ve{M}^{-1}\ve{A}\delta\ve{x}^{k+1/2}
-    = \ve{M}^{-1}\ve{r}^{k+1/2}\:,\\ \ve{x}^{k+1} = \ve{x}^{k+1/2} +
-    \delta \ve{x}^{k+1/2}\:,\\
-    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:,
-    \label{eq:jacobi_preconditioned_mcsa}
-  \end{gather}
-\end{subequations}
-where the Neumann-Ulam Monte Carlo solve now has a preconditioned
-operator from which to build weights and probabilities for transport
-and a preconditioned source vector to sample.
-
-Choosing point Jacobi preconditioning with MCSA is advantageous for
-several reasons. First, $\rho(\ve{I} - \ve{M}^{-1}\ve{A}) < 1$ is true
-for all $\ve{A}$ that is diagonally dominant and is easy to formulate
-because the inversion of $\ve{M}$ is trivial. Second, because the
-Monte Carlo method used within MCSA to compute the correction operates
-on a linear problem with the preconditioned operator, then $\ve{H}$ in
-the Neumann-Ulam solver will have a zero term in each of its diagonal
-elements, thereby eliminating all in-state transitions during the
-random walk sequence. Because of this, point Jacobi preconditioning
-should be considered for many classes of problems, regardless of any
-other preconditioning that is applied to the system.
-
-\clearpage
-
-\subsubsection{General Preconditioning Strategies}
-\label{subsubsec:general_mcsa_preconditioning}
-It is possible to use general left, right, and left/right
-preconditioning with MCSA by carefully considering the underlying
-Monte Carlo problem that will be solved with the Neumann-Ulam
-method. We consider here the general left/right preconditioned method
-as the left or right preconditioned methods can be inferred from its
-formulation. We consider a left preconditioner $\ve{M_L}$ and a right
-preconditioner $\ve{M_R}$. The left/right preconditioned linear
-problem is then:
-\begin{equation}
-  \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\ve{M}_R\ve{x} = \ve{M}_L^{-1}\ve{b}\:.
-  \label{eq:left_right_linear_problem}
-\end{equation}
-To handle the right preconditioning, the system is written with a
-substitution of variables:
-\begin{equation}
-  \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\ve{u} = \ve{M}_L^{-1}\ve{b}\:,
-  \label{eq:left_right_subs_problem}
-\end{equation}
-with
-\begin{equation}
-  \ve{x} = \ve{M}_R^{-1}\ve{u}\:.
-  \label{eq:left_right_recover}
-\end{equation}
-To apply such a method to MCSA, we solve for the substituted variable
-$\ve{u}$ during the iteration sequence:
-\begin{subequations}
-  \begin{gather}
-    \ve{u}^{k+1/2} = \ve{u}^k + \ve{r}^k\:,\\
-    \ve{r}^{k+1/2} = \ve{M}_L^{-1}(\ve{b}-\ve{A}\ve{M}_R^{-1}\ve{u}^{k+1/2})\:,\\ 
-    \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\delta\ve{u}^{k+1/2} = \ve{r}^{k+1/2}\:,\\ 
-    \ve{u}^{k+1} = \ve{u}^{k+1/2} + \delta \ve{u}^{k+1/2}\:,\\
-    \ve{r}^{k+1} = \ve{M}_L^{-1}(\ve{b}-\ve{A}\ve{M}_R^{-1}\ve{u}^{k+1})\:,
-  \end{gather}
-  \label{eq:left_right_mcsa}
-\end{subequations}
-and then recover the original solution vector with
-Eq~(\ref{eq:left_right_recover}). For the Monte Carlo problem, we
-isolate the generation of the correction:
-\begin{equation}
-  \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\delta\ve{u}^{k+1/2} = \ve{r}^{k+1/2}\:,
-  \label{eq:left_right_correction}
-\end{equation}
-and note that the preconditioned residual of the substituted variable
-is now serving as the source and the new iteration matrix is:
-\begin{equation}
-  \ve{H} = \ve{I} - \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\:.
-  \label{eq:left_right_iteration_matrix}
-\end{equation}
-As we require $(i,j)$ element-wise access to the iteration matrix in
-order to construct probabilities and weights for the Monte Carlo
-procedure from the Neumann-Ulam decomposition, the \textit{composite
-  operator}, $\ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}$, must be formed via
-matrix-matrix multiplication.
-
-Several possible shortcomings of this preconditioning approach are
-readily observed. First, the matrix-matrix multiplication operation
-for sparse, parallel distributed matrices is significantly more
-expensive than a matrix-vector multiplication operation. Second, each
-preconditioner must be explicitly inverted, an operation in itself
-that may be expensive and which prohibits the use of any
-preconditioners which provide no mechanism to extract their
-inverse. Third, for many modern preconditioning methods, this
-inversion may yield dense matrices, destroying sparsity and further
-impeding the performance of a matrix-matrix multiplication
-operation. It is also interesting to note that the Monte Carlo problem
-in the general left/right preconditioned scheme given by
-Eq~(\ref{eq:left_right_correction}) is not fully left/right
-preconditioned (meaning that we do not recover $\ve{x}$), but instead
-part of a sequence for finding the substituted variable $\ve{u}$. We
-do, however, gain the benefits of this general preconditioning by
-building the iteration matrix in
-Eq~(\ref{eq:left_right_iteration_matrix}) from the fully
-preconditioned linear operator. In addition, for MCSA to apply to
-increasingly difficult problems, more advanced preconditioning
-techniques that require the generation of the composite operator may
-be necessary for convergence.
-
-%%---------------------------------------------------------------------------%%
-\section{Monte Carlo Method Selection}
-\label{sec:mc_method_selection}
-The MCSA method defined in Eq.~(\ref{eq:mcsa}) uses the adjoint method
-to estimate the error in a residual Monte Carlo solve instead of the
-direct method outlined in \S\ref{sec:direct_mc}. To demonstrate the
-effectiveness of the adjoint method over the direct method within the
-context of MCSA, we choose the two-dimensional time-dependent Poisson
-equation as a simple model transport problem\footnote{The Poisson
-  equation is simple form of the diffusion equation and has a similar
-  elliptic character in its time-dependent form to the neutron
-  transport equations that will be solved in the next chapter.}:
-\begin{equation}
-  \frac{\partial \ve{u}}{\partial t} = \nabla^2 \ve{u}\:.
-  \label{eq:poisson_equation}
-\end{equation}
-For all comparisons, a single time step is computed with backwards Euler time
-integration. The Laplacian is differenced on a square Cartesian grid with a
-second-order five-point stencil,
-\begin{equation}
-  \nabla^2_5 = \frac{1}{\Delta^2}[u_{i-1,j} + u_{i+1,j} + u_{i,j-1} +
-    u_{i,j+1} - 4 u_{i,j}]\:,
-  \label{eq:five_point_stencil}
-\end{equation}
-and a fourth-order nine-point stencil,
-\begin{multline}
-  \nabla^2_9 = \frac{1}{6\Delta^2}[4 u_{i-1,j} + 4 u_{i+1,j} + 4
-    u_{i,j-1} + 4 u_{i,j+1} + u_{i-1,j-1}\\ + u_{i-1,j+1} +
-    u_{i+1,j-1} + u_{i+1,j+1} - 20 u_{i,j}]\:,
-  \label{eq:nine_point_stencil}
-\end{multline}
-both assuming a grid size of $\Delta$ in both the $i$ and $j$ directions. For
-a single time step solution, we then have the following sparse linear system
-to be solved with the MCSA method:
-\begin{equation}
-  \ve{A} \ve{u}^{n+1} = \ve{u}^n\:.
-  \label{eq:poisson_eq_lin_sys}
-\end{equation}
-Both the stencils will be used to vary the size and density of the sparse
-linear system in Eq.~(\ref{eq:poisson_eq_lin_sys}).
-
-A timing and convergence study is used to demonstrate the
-effectiveness of the adjoint method with the collision estimator as
-compared to the direct method. To assess both the CPU time and number
-of iterations required to converge to a solution, a problem of
-constant $\Delta$ was used with varying values of the number of mesh
-elements, fixing the spectral radius of the system at a constant value
-for each variation. Both the five-point and nine-point stencils were
-used with both the direct and adjoint solvers. For each case, $N
-\times N$ total random walk permutations were computed per MCSA
-iteration where $N \times N$ is the number of discrete grid points in
-the system. Solver parameters were set to a weight cutoff of
-\sn{1}{-4} for the stochastic linear solver and a convergence
-tolerance of \sn{1}{-8} for the MCSA iterative solver.
-Figure~\ref{fig:poisson_cpu_time} gives the CPU time needed for each
-case to converge in seconds and Figure~\ref{fig:poisson_iterations}
-gives the number of iterations needed for each case to converge to the
-specified tolerance as a function of the problem size. All
-computations presented in this section and the remaining sections of
-this chapter were completed on a 3.0 GHz Intel Core 2 Quad Q9650 CPU
-machine with 16 GB 1067 MHz DDR3 memory.
-\begin{figure}[t!]
-  \centering
-  \includegraphics[width=5in,clip]{chapters/mc_background/dir_adj_cpu.pdf}
-  \caption{\textbf{CPU Time (s) to converge vs. Problem Size ($N$ for
-      an $N \times N$ square mesh).} \textit{Both the adjoint and
-      direct solvers are used with the five point and nine point
-      stencils. A CPU time speedup is noted with the adjoint method
-      due to the higher density of random walk events in regions with
-      a large residual.}}
-  \label{fig:poisson_cpu_time}
-\end{figure}
-
-\begin{figure}[t!]
-  \centering
-  \includegraphics[width=5in,clip]{chapters/mc_background/dir_adj_iterations.pdf}
-  \caption{\textbf{Iterations to converge vs. Problem Size ($N$ for an
-      $N \times N$ square mesh).} \textit{Both the adjoint and direct
-      solvers are used with the five-point and nine-point stencils.}}
-  \label{fig:poisson_iterations}
-\end{figure}
-
-We see clearly in Figure~\ref{fig:poisson_cpu_time} that the using the
-adjoint solver with MCSA results in a speedup over the direct solver
-while the number of iterations required to converge is also reduced as
-shown in Figure~\ref{fig:poisson_iterations}. We expect this for several
-reasons. First, with an equivalent number of histories specified for
-both solvers per MCSA iteration and a system of size $N \times N$, the
-direct solver will compute a single random walk for each state in the
-system per iteration to acquire a solution in that state, regardless
-of the size of the residual in that state. This is necessary in the
-direct method to ensure a contribution from each state as the random
-walk sequence will only contribute to the starting state. For the
-adjoint method, a total of $N \times N$ random walk events will have
-their starting state determined by sampling the residual
-vector. Because the random walk sequence contributes to the state in
-which it currently resides, sampling the residual vector as the Monte
-Carlo source gives a higher density of random walk events in regions
-with a high residual, thus giving a more accurate correction in that
-region due to reduced statistical error. From an iteration
-perspective, Figure~\ref{fig:poisson_iterations} shows that using the
-direct method yields a roughly unchanging number of iterations
-required to converge as the problem size increases. Again, if we
-desire a correction value for all states in the problem, then we must
-start a random walk in each state in the system which does not reduce
-the number of iterations need as the problem size grows. Conversely,
-as the problem size grows in the adjoint method, the additional
-stochastic histories that will be computed are concentrated in regions
-with a large residual, further reducing the stochastic error in the
-correction in those regions and subsequently reducing the required
-number of iterations to converge.
-
-As an additional comparison, the convergence behavior of MCSA can be
-analyzed using both the adjoint and direct solvers to detect any
-performance benefits. To assess the convergence properties of MCSA
-using each solver and stencil, the infinity norm of the residual
-computed in Eq.~(\ref{eq:mcsa}) was collected at each iteration for a
-fixed problem size of $N=500$. Figure~\ref{fig:poisson_convergence}
-gives the results of these computations. First, it is worthy to note
-on the semi-log plot that we are indeed achieving the expected
-exponential convergence from MCSA with both Monte Carlo
-solvers. Second, we note that using the adjoint method with the same
-number of stochastic histories per MCSA iteration gives a faster rate
-of converge for the same reasons as above. We also note here that
-fewer iterations are required for convergence when the 9-point stencil
-is used to discretize the Laplacian operator (although at no gain in
-speed as given by the results in
-Figure~\ref{fig:poisson_cpu_time}). This is due to the fact that the
-smaller discretization error directly corresponds to a more well
-defined residual source generated by the Richardson extrapolation for
-the Monte Carlo calculation. In addition, the better defined source is
-transported through a domain described more accurately by the 9-point
-stencil, thus yielding a more accurate correction vector from the
-Monte Carlo calculation.
-\begin{figure}[t!]
-  \centering
-  \includegraphics[width=5in,clip]{chapters/mc_background/dir_adj_conv.pdf}
-  \caption{\textbf{Infinity norm of the solution residual
-      vs. iteration number for a problem of size $N=500$.}
-    \textit{Both the adjoint and direct solvers are used with the five
-      point and nine point stencils. A higher rate of convergence is
-      observed for MCSA using the adjoint Monte Carlo solver as
-      compared to the direct method when both solvers compute the same
-      number of random walks per iteration.}}
-  \label{fig:poisson_convergence}
-\end{figure}
-
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{MCSA Comparison to Sequential Monte Carlo}
-\label{subsec:sequential_comparison}
-To further motivate using Monte Carlo Synthetic Acceleration, we
-compare its performance to Halton's Sequential Monte Carlo method on
-which previous work in this area was based. For this comparison, we
-use the same transient Poisson problem as described in the previous
-section and choose only the 5-point stencil to discretize the
-Laplacian operator as the previous results yielded little qualitative
-difference between the discretizations. Both MCSA and Halton's method
-are used with the adjoint Monte Carlo solver and the collision
-estimator. In order to complete the same study as in the previous
-section, the number of histories computed by the Monte Carlo solver at
-each iteration had to be doubled to $2 \times N \times N$ in order to
-ensure convergence in Sequential Monte Carlo Method. For the majority
-of the problems in the previous section, the Sequential method used
-with $N \times N$ histories would not
-converge. Figure~\ref{fig:seq_cpu_time} gives the CPU time results for
-this comparison as a function of problem size while
-Figure~\ref{fig:seq_iterations} gives the number of iterations to
-converge as a function of problem size with a convergence tolerance of
-\sn{1}{-8}. In both cases, using the Monte Carlo solver as a synthetic
-acceleration rather than in a pure residual Monte Carlo scheme
-resulted in a reduction in both CPU time and iterations required to
-converge. The additional Richardson extrapolation between each Monte
-Carlo solve in the MCSA method gives a better converged residual
-source to use with the Monte Carlo calculation while the Sequential
-method requires more iterations to achieve the same level of
-convergence in the residual.
-
-\begin{figure}[t!]
-  \centering
-  \includegraphics[width=4.5in,clip]{chapters/mc_background/seq_cpu.pdf}
-  \caption{\textbf{CPU Time (s) to converge vs. Problem Size ($N$ for
-      an $N \times N$ square mesh).} \textit{Both the Sequential Monte
-      Carlo and MCSA solvers are used with the five point stencils and
-      the adjoint Monte Carlo solver. The number of random walks was
-      twice the number of discrete states in the system in order to
-      ensure convergence in the Sequential Monte Carlo method.}}
-  \label{fig:seq_cpu_time}
-\end{figure}
-
-\begin{figure}[t!]
-  \centering
-  \includegraphics[width=4.5in,clip]{chapters/mc_background/seq_iterations.pdf}
-  \caption{\textbf{Iterations to converge vs. Problem Size ($N$ for an
-      $N \times N$ square mesh).} \textit{Both the Sequential Monte
-      Carlo and MCSA solvers are used with the five point stencils and
-      the adjoint Monte Carlo solver.}}
-  \label{fig:seq_iterations}
-\end{figure}
-
-The benefits of using a synthetic acceleration scheme are also noted
-when the infinity norm of the residual computed at each iteration for
-both methods was collected at each iteration for a fixed problem sizes
-of $N=100$ and $N=500$ as shown in figures Figure~\ref{fig:seq_100}
-and \ref{fig:seq_500} respectively. In both cases, the Sequential
-method is subject to two regimes of exponential convergence with high
-frequency error modes removed in the first regime leaving lower
-frequency and slower converging error modes in the second. Using MCSA
-we a see a single rate of exponential convergence observed to be much
-higher than that computed by Halton's method due to the fact that the
-extra Richardson iteration is providing a smoothing effect to
-alleviate the error mode variations. Even with the doubling of the
-number of stochastic histories computed per time step in order to
-ensure convergence for the Sequential method, we still see robustness
-issues with a non-monotonically decreasing residual observed for the
-$N=100$ case. In both cases the MCSA solver is observed to be robust
-with a monotonically decreasing residual.
-
-\begin{figure}[t!]
-  \centering
-  \includegraphics[width=4.5in,clip]{chapters/mc_background/seq_conv_100.pdf}
-  \caption{\textbf{Infinity norm of the solution residual
-      vs. iteration number for a problem of size $N=100$.}
-    \textit{Both the Sequential Monte Carlo and MCSA solvers are used
-      with the five point stencils and the adjoint Monte Carlo
-      solver.}}
-  \label{fig:seq_100}
-\end{figure}
-
-\begin{figure}[t!]
-  \centering
-  \includegraphics[width=4.5in,clip]{chapters/mc_background/seq_conv_500.pdf}
-  \caption{\textbf{Infinity norm of the solution residual
-      vs. iteration number for a problem of size $N=500$.}
-    \textit{Both the Sequential Monte Carlo and MCSA solvers are used
-      with the five point stencils and the adjoint Monte Carlo
-      solver.}}
-  \label{fig:seq_500}
-\end{figure}
-
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{Monte Carlo Parameter and Estimator Analysis}
-\label{sec:parameter_estimator_analysis}
-With the adjoint method shown to be more effective than the direct
-method and MCSA to have better iterative and timing performance than
-sequential Monte Carlo, we now aim to study the effects of the adjoint
-Monte Carlo parameters, weight cutoff and number of histories, on MCSA
-performance with both the collision and expected value
-estimators. With the same Poisson problem, we will use a $200 \times
-200$ grid and a convergence tolerance of \sn{1}{-8} for all
-calculations. To study the effects of the number of Monte Carlo
-histories per MCSA iteration, the weight cutoff was fixed at
-\sn{1}{-4} while the number of histories per iteration was varied from
-6,000 to 100,000. It was observed that MCSA would not converge for
-this problem using the collision estimator with less than 6,000
-histories while the expected value estimator permitted convergence
-with only 2,000 histories.
-
-Figure~\ref{fig:estimator_nh_iters} gives the number of iterations
-required to converge for both estimators as a function of the number
-of histories per iteration. For smaller numbers of histories, the
-performance of the expected value estimator is significantly better
-than the collision estimator. This result is valuable in that less
-transport is required to achieve the same MCSA iterative performance
-with the expected value estimator, important for situations where
-transport is expensive (i.e. in domain decomposed
-calculations). Interestingly, as the number of histories per iteration
-are increased, the iterative performance with the collision estimator
-approaches that of the expected value estimator. However, given the
-performance of the expected value estimator at a fractional number of
-histories, one should strongly consider its use over using the
-collision estimator with more histories. The CPU time required to
-converge is presented in Figure~\ref{fig:estimator_nh_time} and
-reflects the results of the iterative performance. In general, using
-the collision estimator is slightly slower overall, but the time per
-iteration is faster due to the fact that the estimator does not have
-to cycle through multiple states during the tally procedure.
-
-\begin{figure}[t!]
-  \centering
-  \includegraphics[width=4.75in,clip]{chapters/mc_background/estimator_nh_iters.pdf}
-  \caption{\textbf{Iterations (s) to converge vs. Monte Carlo
-      histories per MCSA iteration for a $200 \times 200$ square mesh
-      and a weight cutoff of \sn{1}{-4}.} \textit{For low numbers of
-      histories, the expected value estimator performance is
-      significantly better than the collision estimator. At higher
-      numbers of histories, the estimators become roughly
-      equivalent.}}
-  \label{fig:estimator_nh_iters}
-\end{figure}
-
-\begin{figure}[t!]
-  \centering
-  \includegraphics[width=4.75in,clip]{chapters/mc_background/estimator_nh_time.pdf}
-  \caption{\textbf{CPU Time (s) to converge vs. Monte Carlo histories
-      per MCSA iteration for a $200 \times 200$ square mesh and a
-      weight cutoff of \sn{1}{-4}.} \textit{For low numbers of
-      histories, the expected value estimator performance is better
-      than the collision estimator due to a lower iteration count
-      while the actual compute time per iteration is higher.}}
-  \label{fig:estimator_nh_time}
-\end{figure}
-
-For the weight cutoff study, the number of histories per iteration was
-fixed at 40,000 and the weight cutoff varied from \sn{5}{-1} down to
-\sn{1}{-10}. Surprisingly, the number of iterations to converge given
-by Figure~\ref{fig:estimator_wc_iters} is effectively invariant to the
-weight cutoff, only seeing detrimental effects on the number of
-iterations at a very large weight cutoff. This suggests that a fairly
-large weight cutoff can be used in practice with both estimators and
-that the preliminary components of the random walk are more important
-to MCSA convergence than those that occur later and with lower weight
-contributions. To further motivate using a larger weight cutoff,
-Figure~\ref{fig:estimator_wc_time} gives the CPU time need to converge
-as a function of weight cutoff. As expected, lowering the weight
-cutoff lengthens the random walk lengths and increases CPU time at no
-gain of iterative performance. In general, these results suggest using
-the expected value estimator over the collision estimator with MCSA as
-well as using a larger weight cutoff.
-
-\begin{figure}[t!]
-  \centering
-  \includegraphics[width=4.75in,clip]{chapters/mc_background/estimator_wc_iters.pdf}
-  \caption{\textbf{Iterations (s) to converge vs. history weight
-      cutoff for a $200 \times 200$ square mesh and 40,000 histories.}
-    \textit{For low numbers of histories, the expected value estimator
-      performance is significantly better than the collision
-      estimator. At higher numbers of histories, the estimators become
-      roughly equivalent.}}
-  \label{fig:estimator_wc_iters}
-\end{figure}
-
-\begin{figure}[t!]
-  \centering
-  \includegraphics[width=4.75in,clip]{chapters/mc_background/estimator_wc_time.pdf}
-  \caption{\textbf{CPU Time (s) to converge vs. history weight cutoff
-      for a $200 \times 200$ square mesh and 40,000 histories.}
-    \textit{For low numbers of histories, the expected value estimator
-      performance is better than the collision estimator due to a
-      lower iteration count while the actual compute time per
-      iteration is higher.}}
-  \label{fig:estimator_wc_time}
-\end{figure}
-
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{Variance Reduction Techniques}
-\label{sec:variance_reduction}
+\chapter{Artificial Absorption}
+\label{chap:artificial_absorption}
 Variance reduction is a mechanism by which the probabilities and
 weights of the Monte Carlo problem are modified to improve the
-performance of a given estimator. In many cases, such modifications
-will introduce bias into the system. However, using Monte Carlo within
-an MCSA iterative scheme provides a potential buffer for the solution
-from this bias as the correction generated by the Monte Carlo solve
-will contain large statistical error even without variance
-reduction. For this work, any method that alters the weights and
-probabilities of the Monte Carlo method with the aim of improving
-either iterative performance or time to convergence will be defined as
-a variance reduction scheme.
+performance of a given estimator. In the context of improving MCSA
+results, such modifications may potentially introduce bias into the
+system. However, using Monte Carlo within an MCSA iterative scheme
+provides a potential buffer for the solution from this bias as the
+correction generated by the Monte Carlo solve will contain large
+statistical error even without variance reduction.
 
-\subsection{Artificial Absorption}
-\label{subsec:artificial_absorption}
 The random walk sequences generated by the adjoint Neumann Ulam
@@
 transition exist within the system and have a non-zero weight. A
-preliminary variance reduction scheme is to introduce
-\textit{artificial absorption} into the system such that at each
-transition event, a stochastic history will now have some finite
-probability of being terminated through absorption as well as
-transition to other states in the system. This probability, $p_{abs}$,
-modifies the probabilities and weights in the adjoint random walk
-sequence as:
+simple variance reduction scheme is to introduce \textit{artificial
+  absorption} into the system such that at each transition event, a
+stochastic history will now have some finite probability of being
+terminated through absorption as well as transitioning to other states
+in the system. This probability, $p_{abs}$, modifies the probabilities
+and weights in the adjoint random walk sequence as:
 \begin{equation}
@@
 \end{figure}
-\clearpage
-
-\subsection{Reduced Domain Approximation}
-\label{subsec:reduced_domain_approximation}
-For scalable parallel implementations and efficient Monte Carlo
-transport, the domain is required to be sparse. We may find through
-general preconditioning operations as outlined by the Monte Carlo
-problem in Eq~(\ref{eq:left_right_correction}) that this sparsity is
-lost. To alleviate this, we propose modifying the domain sampled by
-the Monte Carlo method to generate the correction within MCSA by
-removing terms in the iteration matrix through some predetermined
-criteria in order to recover sparsity. As the Neumann-Ulam scheme is a
-stochastic realization of multiple matrix-vector multiplies, the
-criteria we choose should be based on maintaining those elements that
-will make the largest contributions to the multiplication and
-therefore recover as much of the original product as possible.
-
-For each row in the preconditioned system, we can then apply the
-reduced domain approximation to eliminate those elements that either
-fall below a specified tolerance level, keep the $N$ largest elements
-in each row, or both giving:
-\begin{subequations}
-  \begin{gather}
-    \ve{x}^{k+1/2} = \ve{x}^k + \ve{r}^k\:,\\
-    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\
-    \hat{\ve{A}}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:, 
-    \label{eq:rda_correction}\\
-    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,\\
-    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:,
-  \end{gather}
-  \label{eq:rda_mcsa}
-\end{subequations}
-where $\hat{\ve{A}}$ in Eq~(\ref{eq:rda_correction}) means that the
-reduced domain approximation has been applied to the iteration matrix
-in the Monte Carlo method. If enough of the terms in the original
-iteration matrix have been maintained, the character of the original
-iteration matrix should be retained and continue to accelerate the
-iterative scheme. The effects of the reduced domain approximation on
-neutron transport systems where it is required due to the explicit
-preconditioning strategy will be presented in
-Chapter~\ref{ch:spn_equations}.
-
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{Summary}
-\label{sec:mc_summary}
-
-In this chapter, the fundamentals of Monte Carlo Synthetic
-Acceleration methods were presented and explored in the context of a
-simple model transport system. The following are the significant
-observations and findings.
-
-\begin{itemize}
-\item Monte Carlo Synthetic Acceleration (MCSA) has been generalized
-  for all discrete linear transport problems
-\item MCSA can be preconditioned using a general left/right scheme but
-  requires the explicit inversion of the preconditioners and
-  construction of the composite linear operator
-\item The adjoint Neumann-Ulam method performs better when used with
-  MCSA compared to the forward method
-\item MCSA has superior performance to Halton's residual Monte Carlo
-  method
-\item The Richardson iteration in MCSA behaves as a smoother for the
-  stochastic Monte Carlo error
-\item MCSA iterative performance when used with the expected value
-  estimator is insensitive to histories per iteration
-\item MCSA iterative performance is insensitive to the stochastic
-  history weight cutoff
-\item Variance reduction with artificial absorption can enhance time
-  to solution
-\item The reduced domain approximation was introduced to mitigate the
-  consequences of composite linear operator generation
-\end{itemize}

### 2013-08-22 doc/dissertation/document/chapters/mc_background/mc_background.tex
moving articial absorption into an appendix and the reduced domain approximation into the spn chapter

@@
 methods devised by Neumann and Ulam in an iterative refinement scheme
-that accelerates their convergence. Using these iterative schemes, a
-set of modest variance reduction techniques will be introduced and
-their effects along with other parameters of the algorithms on the
-Monte Carlo solutions explored using a simple model transport problem.
+that accelerates their convergence. Using these iterative schemes,
+the effects of the algorithm parameters on solutions to a simple model
+transport problem are explored.
 
@@
 %%---------------------------------------------------------------------------%%
-\section{Variance Reduction Techniques}
-\label{sec:variance_reduction}
-Variance reduction is a mechanism by which the probabilities and
-weights of the Monte Carlo problem are modified to improve the
-performance of a given estimator. In the context of improving MCSA
-results, such modifications may potentially introduce bias into the
-system. However, using Monte Carlo within an MCSA iterative scheme
-provides a potential buffer for the solution from this bias as the
-correction generated by the Monte Carlo solve will contain large
-statistical error even without variance reduction. For this work, any
-method that alters the weights and probabilities of the Monte Carlo
-method with the aim of improving either iterative performance or time
-to convergence will be defined as a variance reduction scheme.
-
-\subsection{Artificial Absorption}
-\label{subsec:artificial_absorption}
-The random walk sequences generated by the adjoint Neumann Ulam
-decomposition in Eqs~(\ref{eq:adjoint_probability}) and
-(\ref{eq:adjoint_weight}) will continue on indefinitely without the
-implementation of a weight cutoff procedure (which in itself is
-effectively a form of biased variance reduction). This is due to the
-fact that all states to which a stochastic history may move in a given
-transition exist within the system and have a non-zero weight. A
-preliminary variance reduction scheme is to introduce
-\textit{artificial absorption} into the system such that at each
-transition event, a stochastic history will now have some finite
-probability of being terminated through absorption as well as
-transitioning to other states in the system. This probability,
-$p_{abs}$, modifies the probabilities and weights in the adjoint
-random walk sequence as:
-\begin{equation}
-  p_{ij} = \frac{|h_{ji}|}{\sum_j|h_{ji}|}(1-p_{abs})\:,
-  \label{eq:absorption_probabilities}
-\end{equation}
-and
-\begin{equation}
-  w_{ij} = \frac{sign(h_{ji})}{(1-p_{abs})}\sum_j|h_{ji}|\:.
-  \label{eq:absorption_weights}
-\end{equation}
-At each transition step, a history at state $i$ will then sample the
-probability distribution function generated by
-Eq~(\ref{eq:absorption_probabilities}) with an additional absorption
-state added to the distribution. If the sampling procedure results in
-a transition to the absorption state, the history is immediately
-terminated. The advantages of this for residual Monte Carlo are that
-increasing the absorption probability decreases the length of each
-random walk (resulting in speedup), while the random walk has a higher
-weight contribution at every step. This higher weight means that each
-history is depositing more information near its birth site which will
-on average be located where the system residual is the largest. An
-additional advantage of this formulation is that the full Neumann-Ulam
-decomposition is maintained.
-
-To observe the effects of this variance reduction on MCSA solutions,
-the transient Poisson problem was again solved with the adjoint
-method, this time with varying values of artificial absorption. A $200
-\times 200$ grid was used with 40,000 histories at every iteration
-with the expected value estimator converged to a tolerance of
-\sn{1}{-8}. To reduce the effect of the weight cutoff on this study,
-the weight cutoff was set to \sn{1}{-12} such that it is significantly
-more likely that a history will be terminated by absorption rather
-than weight cutoff. Figure~\ref{fig:absorption_iters} gives the number
-of iterations to converge as a function of the artificial absorption
-probability. Up to a probability of roughly 0.5, the iterative
-performance is not significantly affected with the information lost
-due to shortened random walks causing a small increase in the number
-of iterations. At a probability of 0.6, the number of iterations
-required grows rapidly as the shortened randow walks create an MCSA
-correction that has a larger statistical error. Convergence was
-observed to be lost at probabilities of 0.7 and higher.
-
-The random walk length has a strong effect on CPU time as
-well. Figure~\ref{fig:absorption_time} gives the CPU time in seconds
-required to converge as a function of artificial absorption
-probability. Initially, small absorption probabilities not only
-maintain iterative performance, but also drastically reduced the time
-required to converge as absorption was more frequent than even a large
-weight cutoff but the weights themselves were not significantly
-modified. As the number of iterations grow rapidly, so does the time
-required to converge to a solution. Based on these results, using a
-small amount of artificial absorption should provide some CPU speedup
-while maintaining iterative performance. For the problem presented
-here, an absorption probability of around 0.25 may be the best choice
-as it minimizes CPU time.
-
-\begin{figure}[t!]
-  \centering
-  \includegraphics[width=4.5in,clip]{chapters/mc_background/absorption_iters.pdf}
-  \caption{\textbf{Iterations (s) to converge vs. artificial
-      absorption probability for a $200 \times 200$ square mesh and
-      40,000 histories.}  \textit{The addition of absorption is
-      detrimental to iterative performance due to both the shortening
-      of the random walks as well as the modification of the
-      transition weight.}}
-  \label{fig:absorption_iters}
-\end{figure}
-
-\begin{figure}[t!]
-  \centering
-  \includegraphics[width=4.5in,clip]{chapters/mc_background/absorption_time.pdf}
-  \caption{\textbf{CPU Time (s) to converge vs. artificial absorption
-      probability for a $200 \times 200$ square mesh and 40,000
-      histories.}  \textit{The reduced random walk length speeds up
-      the calculation with a fixed number of histories. Too much
-      artificial absorption increases iterations rapidly giving an
-      increasing CPU time.}}
-  \label{fig:absorption_time}
-\end{figure}
-\clearpage
-
-\subsection{Reduced Domain Approximation}
-\label{subsec:reduced_domain_approximation}
-For scalable parallel implementations and efficient Monte Carlo
-transport, the domain is required to be sparse. We may find through
-general preconditioning operations as outlined by the Monte Carlo
-problem in Eq~(\ref{eq:left_right_correction}) that this sparsity is
-lost. To alleviate this, we propose modifying the domain sampled by
-the Monte Carlo method to generate the correction within MCSA by
-removing terms in the iteration matrix through some predetermined
-criteria in order to recover sparsity. As the Neumann-Ulam scheme is a
-stochastic realization of multiple matrix-vector multiplies, the
-criteria we choose should be based on maintaining those elements that
-will make the largest contributions to the multiplication and
-therefore recover as much of the original product as possible.
-
-For each row in the preconditioned system, we can then apply the
-reduced domain approximation to eliminate those elements that either
-fall below a specified tolerance level, keep the $N$ largest elements
-in each row, or both giving:
-\begin{subequations}
-  \begin{gather}
-    \ve{x}^{k+1/2} = \ve{x}^k + \ve{r}^k\:,\\
-    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\
-    \hat{\ve{A}}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:, 
-    \label{eq:rda_correction}\\
-    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,\\
-    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:,
-  \end{gather}
-  \label{eq:rda_mcsa}
-\end{subequations}
-where $\hat{\ve{A}}$ in Eq~(\ref{eq:rda_correction}) means that the
-reduced domain approximation has been applied to the iteration matrix
-in the Monte Carlo method. If enough of the terms in the original
-iteration matrix have been maintained, the character of the original
-iteration matrix should be retained and continue to accelerate the
-iterative scheme. The effects of the reduced domain approximation on
-neutron transport systems where it is required due to the explicit
-preconditioning strategy will be presented in
-Chapter~\ref{ch:spn_equations}.
-
-\clearpage
-
-%%---------------------------------------------------------------------------%%
 \section{Summary}
@@
   history weight cutoff
-\item Variance reduction with artificial absorption can enhance time
-  to solution
-\item The reduced domain approximation was introduced to mitigate the
-  consequences of composite linear operator generation
 \end{itemize}

### 2013-08-22 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
moving articial absorption into an appendix and the reduced domain approximation into the spn chapter

@@
 several additional issues are observed and alleviated to a certain
-extent by applying the reduced domain approximation. MCSA solutions
-for the fuel assembly problem are then verified by comparing the
-solutions against production Krylov solvers for problems with varying
-numbers of energy groups. Finally performance is analyzed through
-comparison with those same Krylov solvers in terms of both iterative
-performance and CPU timing using the same set of problems utilized for
-the verification.
+extent by developing and applying the reduced domain
+approximation. MCSA solutions for the fuel assembly problem are then
+verified by comparing the solutions against production Krylov solvers
+for problems with varying numbers of energy groups. Finally
+performance is analyzed through comparison with those same Krylov
+solvers in terms of both iterative performance and CPU timing using
+the same set of problems utilized for the verification.
 
@@
 \label{subsec:spn_spectral_analysis}
-Previous work in applying MCSA to the thermal radiation diffusion
-equation had success in using Jacobi preconditioning
+To apply MCSA to the fuel assembly problem, we require some form of
+preconditioning to achieve an iteration matrix spectral radius of less
+than unity. Previous work in applying MCSA to the thermal radiation
+diffusion equation had success in using Jacobi preconditioning
 \cite{evans_monte_2012} and therefore we apply it here to the $SP_N$
-equations given their diffusion-like form. In this section, we will
-the structure of the transport operator formed by the $SP_N$ equations
+equations given their diffusion-like form. We first investigate the
+structure of the transport operator formed by the $SP_N$ equations
 while parametrically varying the $SP_N$ order, the $P_N$ order, the
 number of energy groups, and upscattering/downscattering between
-energy groups. For each basic problem, the values given in
-Table~\ref{tab:spn_fixed_parameters} were fixed for each spatial
-location and energy group.
+energy groups with the values given in
+Table~\ref{tab:spn_fixed_parameters} fixed for each spatial location
+and energy group.
 
@@
 patterns for the same problem for 10 energy groups without and with
-upscatter respectively.
+upscatter respectively. We note a few key features of the sparsity
+plots. The first is that for multigroup problems without full
+upscatter and downscatter (i.e. Figure~\ref{fig:group10ds}), the
+resulting matrix is asymmetric and therefore a linear solver that can
+handle asymmetric linear systems is required. Nearly all problems of
+interest will not have full upscattering or downscattering. Second we
+note the largely diagonal character of these systems, although the
+blocks from Eq~(\ref{eq:A_block_matrix}) are readily apparent with a
+block size of $N_g\times(N+1)/2$.
 
@@
 \end{figure}
-
-We note a few key features of the sparsity plots. The first is that
-for multigroup problems without full upscatter and downscatter
-(i.e. Figure~\ref{fig:group10ds}), the resulting matrix is asymmetric
-and therefore a linear solver that can handle asymmetric linear
-systems is required. Nearly all problems of interest will not have
-full upscattering or downscattering. Second we note the largely
-diagonal character of these systems, although the blocks from
-Eq~(\ref{eq:A_block_matrix}) are readily apparent with a block size of
-$N_g\times(N+1)/2$.
 
@@
 \label{subsec:jacobi_prec_assembly_calc}
-A single energy group problem was first computed with $SP_1$
-discretization, effectively giving the one-speed neutron diffusion
-system for the fuel assembly resulting in 20,088 degrees of freedom in
-the problem. Figure~\ref{fig:block_jacobi_res_mcsa} gives the residual
-infinity norm as a function of iteration for the MCSA linear solve in
-the first eigenvalue iteration using 25,000 stochastic histories at
-every iteration for the adjoint Neumann-Ulam solve.
+A single energy group problem was first computed with
+$SP_1$ discretization, effectively giving the one-speed neutron
+diffusion system for the fuel assembly resulting in 20,088 degrees of
+freedom in the problem. Figure~\ref{fig:block_jacobi_res_mcsa} gives
+the residual infinity norm as a function of iteration for the MCSA
+linear solve in the first eigenvalue iteration using 25,000 stochastic
+histories at every iteration for the adjoint Neumann-Ulam solve.
+
 \begin{figure}[t!]
@@
 \end{figure}
+
 Convergence was not achieved as noted by the rapid rise in the
@@
 linear solve in the first eigenvalue iteration.
+
 \begin{figure}[t!]
@@
 \end{figure}
+
 Poor convergence is observed for the Richardson iteration, however,
@@
 stochastic histories in the block Jacobi preconditioned Neumann-Ulam
-method are eventually being terminated by the weight cutoff as no
-artificial absorption was used for these preliminary
-calculations. Based on this, the correction being generated by the
-Neumann-Ulam solve is the component of MCSA causing a divergent
-solution which we study next.
+method are eventually being terminated by the weight cutoff. Based on
+this, the correction being generated by the Neumann-Ulam solve is the
+component of MCSA causing a divergent solution which we study next.
 
@@
 \label{subsec:mcsa_break_down}
-For the fuel assembly problem, the initial block Jacobi preconditioned
-calculations used a one-speed $SP_1$ discretization, effectively
-giving a diffusion system. To study the breakdown of MCSA at iteration
-matrix spectral radii near one, we will use the simpler homogeneous
-2-dimensional one-speed neutron diffusion system presented in
+To study the breakdown of MCSA at iteration matrix spectral radii near
+one, we will use the simpler homogeneous 2-dimensional one-speed
+neutron diffusion system presented in
 Appendix~\ref{chap:diffusion_problem} to isolate this behavior. In
@@
   \begin{center}
-    \includegraphics[width=5.75in]{chapters/spn_equations/breakdown_iterations.pdf}
+    \includegraphics[width=4.5in]{chapters/spn_equations/breakdown_iterations.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=5.75in]{chapters/spn_equations/breakdown_histories.pdf}
+    \includegraphics[width=4.5in]{chapters/spn_equations/breakdown_histories.pdf}
   \end{center}
@@
 is more applicable and in which performance is improved.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=5.75in]{chapters/spn_equations/breakdown_time.pdf}
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=4.5in]{chapters/spn_equations/breakdown_time.pdf}
   \end{center}
@@
 problems that have a dominating elliptic component such as the fuel
-assembly problem or other problems with a high scattering to
-absorption ratio. As with MCSA, the algebraic formulation of ILUT
-requires the physics operator to be fully formed and is therefore
-easily incorporated into our solution sequence.
+assembly problem or other problems with a large scattering to
+absorption ratio. The algebraic formulation of ILUT requires the
+physics operator to be fully formed, also a requirement for MCSA, and
+is therefore easily incorporated into our solution sequence.
 
@@
 
-\subsection{Applying the Reduced Domain Approximation}
+\subsection{The Reduced Domain Approximation}
 \label{subsec:spn_prec_rda}
@@
 domain decomposed algorithms and overall performance issues for Monte
-Carlo. As outlined in \S~\ref{subsec:reduced_domain_approximation},
-the reduced domain approximation may be used as a mechanism to
-potentially alleviate this problem by filtering elements of the
-composite matrix in each row that fall below a certain threshold value
-or by maintaining the largest $N$ elements in each row where $N$ is a
-designated fill level. In this section, we will apply the reduced
-domain approximation to both ILUT and SPAINV preconditioned system to
-reduce the density of the composite linear operator to more manageable
-levels and study the resulting iterative performance.
-
-For each preconditioner, the parameters that achieved the best quality
-metric results from the previous analysis were used. These correspond
-to ILUT parameters of a fill level of 5.0 and a drop tolerance of
-\sn{1}{-5} and SPAINV parameters of a sparsity level of 4 and a
-threshold of 0.1. The reduced domain threshold was set to \sn{1}{-10}
-in order to eliminate any exceedingly small values from the matrix
-(often this is simply removing non-zero elements within the floating
-point tolerance of zero). The reduced domain fill level was then
-varied, starting with the largest non-zero entries per row value
+Carlo. To alleviate this, we propose modifying the domain sampled by
+the Monte Carlo method to generate the correction within MCSA by
+removing terms in the iteration matrix through some predetermined
+criteria in order to recover sparsity. As the Neumann-Ulam scheme is a
+stochastic realization of multiple matrix-vector multiplies, the
+criteria we choose should be based on maintaining those elements that
+will make the largest contributions to the multiplication and
+therefore recover as much of the original product as possible.
+
+For each row in the preconditioned system, we can then apply the
+reduced domain approximation to eliminate those elements that either
+fall below a specified tolerance level, keep the $N$ largest elements
+in each row, or both giving:
+\begin{subequations}
+  \begin{gather}
+    \ve{x}^{k+1/2} = \ve{x}^k + \ve{r}^k\:,\\
+    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\
+    \hat{\ve{A}}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:, 
+    \label{eq:rda_correction}\\
+    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,\\
+    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:,
+  \end{gather}
+  \label{eq:rda_mcsa}
+\end{subequations}
+where $\hat{\ve{A}}$ in Eq~(\ref{eq:rda_correction}) means that the
+reduced domain approximation has been applied to the iteration matrix
+in the Monte Carlo method. If enough of the terms in the original
+iteration matrix have been maintained, the character of the original
+iteration matrix should be retained and continue to accelerate the
+iterative scheme.
+
+We next apply the reduced domain approximation to both ILUT and SPAINV
+preconditioned system to reduce the density of the composite linear
+operator to more manageable levels and study the resulting iterative
+performance. For each preconditioner, the parameters that achieved
+the best quality metric results from the previous analysis were
+used. These correspond to ILUT parameters of a fill level of 5.0 and a
+drop tolerance of \sn{1}{-5} and SPAINV parameters of a sparsity level
+of 4 and a threshold of 0.1. The reduced domain threshold was set to
+\sn{1}{-10} in order to eliminate any exceedingly small values from
+the matrix (often this is simply removing non-zero elements within the
+floating point tolerance of zero). The reduced domain fill level was
+then varied, starting with the largest non-zero entries per row value
 observed for each of the preconditioning types in order to assess its
@@
 successful recovery of sparsity for the Monte Carlo problem while
-maintaining good convergence properties, there is still an issue of
-forming the composite operator before applying the approximation and
-potentially generating the transpose of this operator in the case of
-the adjoint Monte Carlo method. Because of this, memory and scaling
-issues may still be observed when building the probability and weight
-matrices for the Monte Carlo problem. In addition, the expensive
-extraction of the inverse of the preconditioning operators for the
-explicit scheme creates a significant cost in overall
-performance. Future work in this area should be considerate of these
-important components of the preconditioned MCSA algorithm.
+maintaining good convergence properties with ILUT, there is still an
+issue of forming the composite operator before applying the
+approximation and potentially generating the transpose of this
+operator in the case of the adjoint Monte Carlo method. Because of
+this, memory and scaling issues may still be observed when building
+the probability and weight matrices for the Monte Carlo problem. In
+addition, the expensive extraction of the inverse of the
+preconditioning operators for the explicit scheme creates a
+significant cost in overall performance. Future work in this area
+should be considerate of these important components of the
+preconditioned MCSA algorithm.
 
@@
 MCSA to conventional Krylov solvers that would be typically used to
-solve the $SP_N$ form of the transport equation as presented here.
+solve the $SP_N$ form of the transport equation as a means of
+verification.
 
@@
 number of eigenvalue iterations required to converge the problem, and
-the eigenvalue itself will be compared as a means of verification. For
-the flux values, we will look at the 2-norm and infinity norm of the
-relative difference vector between the reference case and the
-comparison case in each energy group:
+the eigenvalue itself will each be evaluated. For the flux values, we
+will look at the 2-norm and infinity norm of the relative difference
+vector between the reference case and the comparison case in each
+energy group:
 \begin{equation}
-  \text{Relative Group Flux Difference} = \frac{\ve{\Phi}_{ref}^g -
-    \ve{\Phi}^g}{\ve{\Phi}_{ref}^g}\:,
+  \text{Relative Group Flux Difference} = \frac{|\ve{\Phi}_{ref}^g -
+    \ve{\Phi}^g|}{|\ve{\Phi}_{ref}^g|}\:,
   \label{eq:relative_difference_vector}
 \end{equation}
-where the addition and subtraction operations are performed
-element-wise to construct the vector. For this work, the BiCGStab
-results will be taken as the reference case.
+where the absolute value, subtraction, and division operations are
+performed element-wise to construct the vector. For this work, the
+BiCGStab results will be taken as the reference case.
 
@@
 Richardson relaxation of 1.1 was used with the collision estimator and
-1.0 with the expected value estimator. Additionally, both the
-Richardson-based MCSA iteration given by Eq~(\ref{eq:mcsa}) and the
-minimal residual-based MCSA iteration given by
-Eq~(\ref{eq:mcsa_min_res}) will be used. A supplementary performance
-analysis for this set of MCSA variations is presented in
-Appendix~\ref{chap:spn_estimator_comparison}.
-
-Table~\ref{tab:spn_solver_defs} gives definitions for the
-solvers used to generate the results in the remainder of this
-section. For the energy group structures,
-Table~\ref{tab:spn_group_structure} gives the lower bounds of each
-group in eV (an implicit upper bound of \sn{2}{6} eV is assumed for
-group 0).
+1.0 with the expected value estimator. When the collision estimator
+was used, an artificial absorption value of 0.2 as outlined in
+Appendix~\ref{chap:artificial_absorption} was also
+applied. Additionally, both the Richardson-based MCSA iteration given
+by Eq~(\ref{eq:mcsa}) and the minimal residual-based MCSA iteration
+given by Eq~(\ref{eq:mcsa_min_res}) will be used. A supplementary
+performance analysis for this set of MCSA variations is presented in
+Appendix~\ref{chap:spn_estimator_comparison}. Table~\ref{tab:spn_solver_defs}
+gives definitions for the solvers used to generate the results in the
+remainder of this section and Table~\ref{tab:spn_group_structure}
+gives the group structures with lower bounds of each group in eV.
 
@@
 energy groups. In general, we see good agreement in the 2-norm between
-the MCSA variants and the GMRES calculation, meaning that they have
-approximately the same variation from the BiCGStab calculation. We do
+the MCSA variants and the GMRES calculations, meaning that they have
+approximately the same variation from the BiCGStab calculations. We do
 note larger infinity norm values of this vector for some of the MCSA
@@
       2-norm of the relative flux vector in each group is given with
-      BiCGStab used as the reference. Each calculation was converged
-      with the same eigensolver and linear solver
-      tolerances. Table~\ref{tab:spn_solver_defs} gives the
-      description for each solver type presented.}}
+      BiCGStab used as the reference.}}
   \label{tab:serial_differences_g1}
@@
       2-norm of the relative flux vector in each group is given with
-      BiCGStab used as the reference. Each calculation was converged
-      with the same eigensolver and linear solver
-      tolerances. Table~\ref{tab:spn_solver_defs} gives the
-      description for each solver type presented.}}
+      BiCGStab used as the reference.}}
   \label{tab:serial_differences_g2}
@@
       2-norm of the relative flux vector in each group is given with
-      BiCGStab used as the reference. Each calculation was converged
-      with the same eigensolver and linear solver
-      tolerances. Table~\ref{tab:spn_solver_defs} gives the
-      description for each solver type presented.}}
+      BiCGStab used as the reference.}}
   \label{tab:serial_differences_g4}
@@
 With the same set of solver parameters used for verification
-caluclation, MCSA will now be compared to the same Krylov solvers in
+calculation, MCSA will now be compared to the same Krylov solvers in
 terms of both iterative performance and CPU timing. A Richardson
@@
   subsequent investigations
+\item The reduced domain approximation was introduced to mitigate the
+  consequences of composite linear operator generation
 \item The reduced domain approximation was required to mitigate the

### 2013-08-22 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
updating doc build with refactored spn chapter

@@
 
-In this chapter, we briefly derive the simplified $P_N$ ($SP_N$)
-equations, closely following the work of Evans
+In this chapter, we state the neutron transport equation and briefly
+derive the simplified $P_N$ ($SP_N$) equations in both a fixed source
+and eigenvalue form, closely following the work of Evans
 \cite{evans_simplified_2013}, in order to gain full understanding of
 the underlying system and its potential behavior in a Monte Carlo
-context. From the $P_N$ equations, we apply a set of approximations to
-yield the $SP_N$ equations for fixed source and criticality
-problems. Using the fully-formed linear operator for the transport
+context. Using the fully-formed linear operator for the transport
 problem, we explore solutions to the $SP_N$ equations with Monte Carlo
@@
 effectively solve the $SP_N$ equations for the fuel assembly problem,
-a suite of preconditioners along with a relaxation scheme is developed
-and studied within the context of MCSA. Using these preconditioners,
-several additional issues are observed and alleviated to a certain
-extent by developing and applying the reduced domain
-approximation. MCSA solutions for the fuel assembly problem are then
-verified by comparing the solutions against production Krylov solvers
-for problems with varying numbers of energy groups. Finally
-performance is analyzed through comparison with those same Krylov
-solvers in terms of both iterative performance and CPU timing using
-the same set of problems utilized for the verification.
+a suite of preconditioners is developed and studied within the context
+of MCSA. Using these preconditioners, several additional issues are
+observed and alleviated to a certain extent by developing and applying
+the reduced domain approximation. MCSA solutions for the fuel assembly
+problem are then verified by comparing the solutions against
+production Krylov solvers for problems with varying numbers of energy
+groups. Finally performance is analyzed through comparison with those
+same Krylov solvers in terms of both iterative performance and CPU
+timing using the same set of problems utilized for the verification.
 
@@
 
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=4.25in]{chapters/spn_equations/rda_iterations.pdf}
+  \end{center}
+  \caption{\textbf{Number of MCSA iterations required to converge a
+      single eigenvalue iteration for the fuel assembly problem with
+      each preconditioning as a function of reduced domain
+      approximation fill level.} \textit{The largest fill level for
+      each preconditioning presented is that using the parameters that
+      gave the best results without the approximation.}}
+  \label{fig:rda_iterations}
+\end{figure}
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=4.25in]{chapters/spn_equations/rda_quality.pdf}
+  \end{center}
+  \caption{\textbf{Preconditioning quality metric for the fuel
+      assembly problem given as a function of reduced domain
+      approximation fill level.} \textit{The largest fill level for
+      each preconditioning presented is that using the parameters that
+      gave the best results without the approximation.}}
+  \label{fig:rda_quality}
+\end{figure}
+
 Although applying the reduced domain approximation results in a
@@
 
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4.25in]{chapters/spn_equations/rda_iterations.pdf}
-  \end{center}
-  \caption{\textbf{Number of MCSA iterations required to converge a
-      single eigenvalue iteration for the fuel assembly problem with
-      each preconditioning as a function of reduced domain
-      approximation fill level.} \textit{The largest fill level for
-      each preconditioning presented is that using the parameters that
-      gave the best results without the approximation.}}
-  \label{fig:rda_iterations}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=4.25in]{chapters/spn_equations/rda_quality.pdf}
-  \end{center}
-  \caption{\textbf{Preconditioning quality metric for the fuel
-      assembly problem given as a function of reduced domain
-      approximation fill level.} \textit{The largest fill level for
-      each preconditioning presented is that using the parameters that
-      gave the best results without the approximation.}}
-  \label{fig:rda_quality}
-\end{figure}
-
-\clearpage
-
 %%---------------------------------------------------------------------------%%
-\section{MCSA Verification\ }
+\section{MCSA Verification}
 \label{sec:spn_mcsa_verification}
@@
 %%---------------------------------------------------------------------------%%
-\section{MCSA Performance Comparison to Conventional Methods\ }
+\section{MCSA Performance Comparison to Conventional Methods}
 \label{sec:spn_comparison}
@@
 %%---------------------------------------------------------------------------%%
-\section{Summary\ }
+\section{Summary}
 \label{sec:mc_summary}
@@
 \item The reduced domain approximation was introduced to mitigate the
-  consequences of composite linear operator generation
-\item The reduced domain approximation was required to mitigate the
   memory and performance constraints of the explicit preconditioning

### 2013-08-22 doc/dissertation/document/chapters/mc_background/mc_background.tex
cleaning up some whitespace to reduce the length

@@
 %%---------------------------------------------------------------------------%%
-\section{Neumann-Ulam Method}
+\section{Neumann-Ulam Methods}
 \label{sec:mc_preliminaries}
@@
 
-%%---------------------------------------------------------------------------%%
-\section{Direct Monte Carlo Method}
+\subsection{Forward Neumann-Ulam Method}
 \label{sec:direct_mc}
-In the context of matrix inversion, a direct (forward) method
+In the context of matrix inversion, a forward (direct) method
 resembles an adjoint Monte Carlo method in the reactor physics
@@
 
-\subsection{Forward Estimator Variance}
+\subsubsection{Forward Estimator Variance}
 \label{subsec:forward_variance}
@@
 
-\subsection{Direct Method: Evolution of a Solution}
+\subsubsection{Forward Method: Evolution of a Solution}
 \label{subsec:direct_evolution}
-As a means of visually demonstrating the direct Monte Carlo method,
+As a means of visually demonstrating the forward Monte Carlo method,
 consider a 2-dimensional thermal diffusion problem with sources on the
@@
 
-%%---------------------------------------------------------------------------%%
-\section{Adjoint Monte Carlo Method}
+\subsection{Adjoint Neumann-Ulam Method}
 \label{sec:adjoint_mc}
@@
 
-\subsection{Collision Estimator Variance}
+\subsubsection{Collision Estimator Variance}
 \label{subsec:forward_variance}
@@
 
-\subsection{Expected Value Estimator}
+\subsubsection{Expected Value Estimator}
 \label{subsec:expected_value_estimator}
@@
 
-\subsection{Adjoint Method: Evolution of a Solution}
+\subsubsection{Adjoint Method: Evolution of a Solution}
 \label{subsec:adjoint_evolution}
@@
 with $N \times N$ histories would not
-converge. Figure~\ref{fig:seq_cpu_time} gives the CPU time results for
-this comparison as a function of problem size while
+converge.
+
+Figure~\ref{fig:seq_cpu_time} gives the CPU time results for this
+comparison as a function of problem size while
 Figure~\ref{fig:seq_iterations} gives the number of iterations to

### 2013-08-22 doc/dissertation/document/frontmatter/abstract.tex
cleaning up some whitespace to reduce the length

@@
 moderator. To enable MCSA as a solution technique a group of modern
-preconditioning strategies are researched. The new preconditioning
+preconditioning strategies are researched. MCSA was verified to
+produce the correct solution for the fuel assembly criticality problem
+when compared to conventional Krylov methods and demonstrated improved
+iterative performance over GMRES by converging in fewer iterations
+when using the same preconditioning. The new preconditioning
 techniques, however, create unwanted growth in both memory consumption
-and computation time for these problems. MCSA was verified to produce
-the correct solution for the fuel assembly criticality problem when
-compared to conventional Krylov methods and demonstrated improved
-iterative performance over GMRES by converging in fewer iterations
-when using the same preconditioning.
+and computation time for these problems.
 
-Second, solutions to the compressibleNavier-Stokes equations were
+Second, solutions to the compressible Navier-Stokes equations were
 obtained by developing the Forward-Automated Newton-MCSA (FANM) method

### 2013-08-22 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
spell check for spn chapter edits

@@
 same Krylov solvers in terms of both iterative performance and CPU
-timing using the same set of problems utilized for the verification.
+timing with MCSA demonstrated to have improved iterative performance
+over GMRES.
 
@@
       material in the center of the assembly is homogenized
-      moderator/control rodes and darker purple/red material is
+      moderator/control rods and darker purple/red material is
       homogenized moderator/fuel.}}
@@
 ILUT preconditioning is chosen first due to its applicability to
-problems that have a dominating elliptic component such as the fuel
+problems that have a dominating elliptic character such as the fuel
 assembly problem or other problems with a large scattering to
@@
       \multicolumn{1}{c}{\textbf{Group}} &
-      \multicolumn{1}{c}{\textbf{$|| \cdot ||_{inf}$}} &
+      \multicolumn{1}{c}{\textbf{$|| \cdot ||_{\infty}$}} &
       \multicolumn{1}{c}{\textbf{$|| \cdot ||_2$}} \\
@@
 assembly criticality problem. The following are the significant
-observations and findings.
+observations, findings, and contributions.
 
@@
   production code base developed at Oak Ridge National Laboratory
+\item Light water reactor problems are difficult to solve with MCSA as
+  they have large spectral radii due to the neutron scattering in the
+  moderator
 \item Simple Jacobi-based preconditioning reduces the spectral radius
@@
   convergence and more CPU time required per history
-\item Light water reactor problems are difficult to solve with MCSA as
-  they have large spectral radii due to the neutron scattering in the
-  moderator
 \item Advanced algebraic preconditioning strategies were applied to
@@
   strategy although both remain significant obstacles
-\item MCSA relaxation parameters were developed and observed to
-  enhance the time to solution for the fuel assembly criticality
-  problem
 \item MCSA was verified to produce the same k-eigenvalue for the fuel

### 2013-08-22 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
updating doc build with refactored spn chapter
spell check for spn chapter edits
removing jfnk section

@@
 methods, a variant of inexact Newton methods, will be outlined
-including two techniques for numerically representing the Jacobian
-commonly used in today's applications. Following this, a new algorithm
-utilizing Monte Carlo Synthetic Acceleration will be presented and
-implementation details will be discussed. Using the new algorithm and
-a production fluid physics code, a group of challenging nonlinear
+including an automatic technique for generating the Jacobian commonly
+used in today's applications. Following this, a new nonlinear solution
+algorithm utilizing Monte Carlo Synthetic Acceleration will be
+presented and implementation details will be discussed. Using the new
+algorithm and a production fluid physics code, a group of challenging
 benchmark problems for the Navier-Stokes equations will be used to
@@
 
-Newton's method is \textit{consistent} in that all components of the
-nonlinear functions that describe the physics we are modeling are
-updated simultaneously in the iteration sequence with respect to one
-another. This is in comparison to \textit{inconsistent} strategies,
-such as a pressure correction strategy for solving the Navier-Stokes
-equations \cite{pletcher_computational_1997}, where the components of
-$\ve{u}$ are updated in a staggered fashion depending on the
-particular equations that they are associated with.
-
 \subsection{Inexact Newton Methods}
@@
 therefore used it for subsequent work in this area.
-
-\subsection{Jacobian-Free Approximation}
-\label{subsec:jacobian_free_approximation}
-In many cases, the Jacobian is difficult to form from the difference
-equations and costly to evaluate for large equation sets. For simple
-nonlinear cases such as the Navier-Stokes equations, the derivatives
-can be computed and coded, but due to the complexity of those
-derivatives and the resulting difference equations this task can be
-tedious, error prone, and must be repeated for every equation
-set. Furthermore, in their 1995 work, Knoll and McHugh also noted that
-a dominating part of their computation time was the evaluation of the
-difference equations for building the Jacobian
-\cite{knoll_newton-krylov_1995}. By recognizing that Krylov methods
-only need the action of the operator on a vector instead of the
-operator itself, the Jacobian can instead be approximated through
-various numerical methods including a difference-based Jacobian-free
-formulation.
-
-Jacobian-Free methods, and in particular \textit{Jacobian-Free
-  Newton-Krylov} (JFNK) methods \cite{knoll_jacobian-free_2004}, rely
-on forming the action of the Jacobian on a vector as required by the
-Krylov solver through a forward difference scheme. In this case, the
-action of the Jacobian on some vector $\ve{v}$ is given as:
-\begin{equation}
-  \ve{J}(\ve{u})\ve{v} = \frac{\ve{F}(\ve{u} + \epsilon \ve{v}) -
-    \ve{F}(\ve{u})}{\epsilon}\:,
-  \label{eq:jacobian_free_product}
-\end{equation}
-where $\epsilon$ is a small number typically on the order of machine
-precision. Kelley \cite{kelley_iterative_1995} points out a potential
-downfall of this formulation in that if the discretization error in
-$\ve{F}(\ve{u})$ is on the order of the perturbation parameter
-$\epsilon$, then the finite difference error from
-Eq~(\ref{eq:jacobian_free_product}) pollutes the solution.
-
-In addition, Knoll and McHugh noted that for preconditioning purposes,
-part of the Jacobian must still explicitly be formed periodically and
-that linear solver robustness issues were magnified by the matrix-free
-approach due to the first-order approximation. This formation
-frequency coupled with the numerous evaluations of the Jacobian
-approximation create a situation where after so many nonlinear
-iterations, it becomes cheaper to instead fully form the Jacobian. For
-simple equation sets, this may only take 5-10 Newton iterations to
-reach this point while over 30 may be required for larger equations
-sets and therefore larger Jacobians.
 
@@
 Acceleration methods in the context of solutions to the Navier-Stokes
-equations. The following are the significant observations and findings.
+equations. The following are the significant observations, findings,
+and contributions.
 

### 2013-08-23 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
more edits to the nonlinear chapter

@@
 \label{subsec:automatic_differentiation}
-If it is acceptable to store the actual Jacobian matrix, other methods
-are available to construct it without requiring hand-coding and
-evaluating derivatives, thus eliminating the associated issues. In
-addition, if any additional equations are added to the system or a
-higher order functional approximation is desired, it would be useful
-to avoid regenerating and coding these derivatives. Becoming more
-prominent in the 1990's, \textit{automatic differentiation} is a
-mechanism by which the derivatives of a function can be generated
-automatically by evaluating it. Automatic differentiation is built on
-the concept that all functions discretely represented in a computer
-are ultimately represented by elementary mathematical operations. If
-the chain rule is applied to those elementary operations, then the
-derivatives of those functions can be computed to the order of
-accuracy of their original discretization in a completely automated
-way \cite{averick_computing_1994}.
+If it is acceptable to store the actual Jacobian matrix, methods are
+available to construct it without requiring hand-coding and evaluating
+derivatives, thus eliminating the associated issues. In addition, if
+any additional equations are added to the system or a higher order
+functional approximation is desired, it would be useful to avoid
+regenerating and coding these derivatives. Becoming more prominent in
+the 1990's, \textit{automatic differentiation} is a mechanism by which
+the derivatives of a function can be generated automatically by
+evaluating it. Automatic differentiation is built on the concept that
+all functions discretely represented in a computer are ultimately
+represented by elementary mathematical operations. If the chain rule
+is applied to those elementary operations, then the derivatives of
+those functions can be computed to the order of accuracy of their
+original discretization in a completely automated way
+\cite{averick_computing_1994}.
 
@@
 matrix-free approach is chosen to relax memory requirements over
-directly generating the Jacobian matrix and facilitate the
-implementation. However, as we observed in previous sections, these
-matrix-free methods suffer from poorly scaled problems and the first
-order error introduced by the Jacobian approximation. In addition, it
-was observed that the savings induced by the matrix-free approach is
+directly generating the Jacobian matrix and to facilitate the
+implementation. However, these matrix-free methods suffer from the
+first order error introduced by the Jacobian approximation in poorly
+scaled problems \citep{kelley_iterative_1995}. In addition, it has
+been observed that the savings induced by the matrix-free approach is
 eventually amortized over a number of nonlinear iterations where it
-becomes more efficient computationally to instead form the Jacobian.
+becomes more efficient computationally to instead form the Jacobian
+\citep{knoll_newton-krylov_1995}.
 
@@
 of the energy, mass, and momentum equations.
-
 \begin{figure}[t!]
@@
 \end{figure}
-
 In this problem, a rectilinear grid is applied to the unit square. No
@@
 to conductive heat transfer), we can adjust the influence of the
-nonlinear convection term in Eq~(\ref{eq:ns_momentum}). In Shadid's
-work, Rayleigh numbers of up to \sn{1}{6} were used for this benchmark.
+nonlinear terms in Eq~(\ref{eq:ns_momentum}) and
+Eq~(\ref{eq:ns_energy}). In Shadid's work, Rayleigh numbers of up to
+\sn{1}{6} were used for this benchmark.
 
@@
 difficulty in obtaining a solution is dictated by the Reynolds number
-of the fluid. In Shadid's work, a Reynolds numbers up to 800 were
-used.
-
-Compared to the lid driven cavity benchmark, the flow over a backward
-step is much more representative of flow driven by inertial forces in
-a nuclear reactor that may encounter several obstacles along the flow
-path. For example, spacer grids in fuel assemblies or mixing vanes
-designed to induce turbulence and improve heat transfer will create
-recirculation zones much like those observed for this benchmark
-problem. Figure~\ref{fig:step_velocity_isocurves} gives the velocity
-magnitude contours for solutions to this problem with Reynolds numbers
-of 200, 300, 400, and 500.
+of the fluid.
+
+Figure~\ref{fig:step_velocity_isocurves} gives the velocity magnitude
+contours for solutions to this problem with Reynolds numbers of 200,
+300, 400, and 500. As the Reynolds number increases, note the
+downstream recirculation zone beginning to form in the upper right
+corner of the solution plots where the velocity isocurves are
+depressed. In Shadid's work, Reynolds numbers up to 800 were used.
 
@@
 common technique for preconditioning these equations
-\cite{ghia_high-re_1982,evans_enhanced_2007} with the preconditioner
-reconstructed at every Newton step for the new Jacobian. Without
-multigrid preconditioning, the residual in MCSA was observed to
+\cite{ghia_high-re_1982,evans_enhanced_2007}. The preconditioner is
+reconstructed at every Newton step for the new Jacobian and used on
+the right as outlined in
+\S~\ref{subsubsec:general_mcsa_preconditioning}. Without multigrid
+preconditioning, the residual in MCSA was observed to
 stagnate. Typically this happens because the Richardson iteration and
-Neumann series on which the Monte Carlo method is based will quickly
-damp the higher order error modes on the scale of the computational
-grid used to solve the problem but will reduce lower order error modes
-at a considerably lower rate. In terms of Monte Carlo, this stagnation
-manifests itself in histories that simply do no terminate and
-therefore no solution is acquired. For many of the problems presented
-here, it is likely that a more physics-based preconditioning strategy
-such as those presented in
+subsequently the Neumann series will quickly damp the higher order
+error modes on the scale of the computational grid used to solve the
+problem but will reduce lower order error modes at a considerably
+lower rate. In terms of Monte Carlo, this stagnation manifests itself
+in histories that simply do no terminate and therefore no solution is
+acquired. For many of the problems presented here, it is likely that a
+more physics-based preconditioning strategy such as those presented in
 \cite{evans_development_2006,evans_enhanced_2007} will yield better
-results for more difficult problems. However, the explicit
-preconditioning strategy used here currently limits us to algebraic
-preconditioning.
+results for more difficult problems. However, in this work we will
+restrict our attention to multigrid for simplicity.
 
@@
 solver parameters including globalization and forcing term strategies
-were used with both methods with a nonlinear convergence criteria of a
-\sn{1}{-12} norm for the nonlinear residual. As with the radiation
-transport calculations in the previous chapter, algebraic
+were used for both methods with a \sn{1}{-12} norm for the nonlinear
+residual set as the nonlinear convergence criteria. As with the
+radiation transport calculations in the previous chapter, algebraic
 preconditioning with the multigrid method results in dense linear
@@
 \cite{pawlowski_drekar_2012}. Drekar is a massively parallel physics
-framework that utilizes Newton-Krylov methods to generate a fully
-consistent solution scheme for the Navier-Stokes equations among other
-systems with Jacobian generation through automatic differentiation. To
-achieve these results, a Monte Carlo solver generated for this work
-was injected into the nonlinear tool chain within Drekar to form an
+framework that utilizes a finite element formulation of the
+Navier-Stokes equations and Newton-Krylov methods with Jacobian
+generation through automatic differentiation. To achieve these
+results, a Monte Carlo solver generated for this work was injected
+into the nonlinear solver sequence within Drekar to form an
 implementation of the FANM method. For the Newton-Krylov solutions,
-the same implementation of GMRES from the Trilinos scientific
-computing library Aztec \cite{heroux_overview_2005} used for the
-verification of MCSA for the $SP_N$ solutions in the previous chapter
-was also used here as a means of comparison and FANM benchmarking.
+the same implementation of GMRES used for the $SP_N$ verification from
+the Trilinos scientific computing library Aztec
+\cite{heroux_overview_2005} was also used here as a means of
+comparison and FANM benchmarking.
 
@@
 
-For the verification calculations, at each Rayleigh number the fluid
-velocity in the x and y directions, fluid pressure, and fluid
-temperature values calculated Newton-Krylov were compared against the
-Newton-Krylov calculations. No variation in solution was noted for
-each of the computations larger than the floating point tolerance
-reported by Drekar with an absolute difference value of zero observed
-at all spatial locations for each quantity.
+At each Rayleigh number the fluid velocity in the x and y directions,
+fluid pressure, and fluid temperature values calculated with FANM were
+verified by computing the point-wise absolute difference with the
+Newton-Krylov results. No variation in solution was noted for each of
+the computations larger than the floating point tolerance reported by
+Drekar with an absolute difference value of zero observed at all
+spatial locations for each quantity.
 
@@
 
-For the verification calculations, at each Reynolds number the fluid
-velocity in the x and y directions and fluid pressure values
-calculated Newton-Krylov were compared against the Newton-Krylov
-calculations. No variation in solution was noted for each of the
+At each Reynolds number the fluid velocity in the x and y directions
+and fluid pressure values calculated with FANM were verified by
+computing the point-wise absolute difference with the Newton-Krylov
+results. No variation in solution was noted for each of the
 computations larger than the floating point tolerance reported by
@@
 (NK) method. For each benchmark, we will analyze the iterative
-performance of both the nonlinear Newton solver and the linear solver
-used to compute the Newton correction vector at each step. Limiting
-the number of nonlinear iterations required for solution is critical
-for efficiency as nonlinear function evaluations, Jacobian generation,
-and preconditioner construction are typically required at every Newton
+performance of both the Newton solver and the linear solver used to
+compute the Newton correction vector at each step. Limiting the number
+of nonlinear iterations required for solution is critical for
+efficiency as nonlinear function evaluations, Jacobian generation, and
+preconditioner construction are typically required at every Newton
 iteration. In addition, timing results will also be discussed. To
@@
 component of MCSA, each calculation was also performed using the same
-Newton method but with a plain preconditioned Richardson iteration as
-the linear solution technique to give a Newton-Richardson (NR)
-scheme. For all cases the Richardson iteration was preconditioned
+Newton method but with a preconditioned Richardson iteration as the
+linear solution technique to give a Newton-Richardson (NR) scheme. For
+all cases the Richardson iteration was preconditioned with multigrid
 in the same way as GMRES and MCSA.

### 2013-08-23 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
spell check for nonlinear chapter

@@
 methods, a variant of inexact Newton methods, will be outlined
-including an automatic technique for generating the Jacobian commonly
+including an automated technique for generating the Jacobian commonly
 used in today's applications. Following this, a new nonlinear solution
@@
 Inexact Newton methods arise when the Jacobian operator is not exactly
-inverted, resulting in an inexact Newton correction as initially
-described by Dembo and others \cite{dembo_inexact_1982}. For common
-sparse nonlinear systems, which in turn yield a sparse Jacobian
-matrix, this situation occurs when conventional iterative methods are
-applied. In their definition, Dembo formulated inexact methods such
-that they are independent of the linear method used to solve for the
-Newton correction and therefore are amenable to use with any linear
-solver. Furthermore, they bind the convergence of the outer nonlinear
-iteration to the inner linear iteration such that:
+inverted as initially described by Dembo and others
+\cite{dembo_inexact_1982}. For common sparse nonlinear systems, which
+in turn yield a sparse Jacobian matrix, this situation occurs when
+conventional iterative methods are applied. In their definition, Dembo
+formulated inexact methods such that they are independent of the
+linear method used to solve for the Newton correction and therefore
+are amenable to use with any linear solver. Furthermore, they bind the
+convergence of the outer nonlinear iteration to the inner linear
+iteration such that:
 \begin{equation}
@@
     methods} may be used to modify the Newton correction in a more
-  desirable direction such that convergence properties can be
-  improved when the initial guess for $\ve{u}$ is poor
+  desirable direction such that nonlinear convergence properties can
+  be improved when the initial guess for $\ve{u}$ is poor
   \cite{pawlowski_globalization_2006}.
@@
 architectures. Furthermore, their lack of explicit dependence on the
-operator make them easier to implement than other
-methods. Additionally, although many iterations can become memory
-intensive due to the need to store the Krylov subspace for the
-orthogonalization procedure, at each nonlinear iteration this cost is
-reset as the Jacobian matrix will change due to its dependence on the
-solution vector. This means that for every nonlinear iteration, a
-completely new linear system is formed for generating the Newton
-correction and we can modify the Krylov solver parameters accordingly
-to accommodate this. In most nonlinear problems, the Jacobian operator
-is generally non-symmetric and therefore either Krylov methods with
-long recurrence relations that can handle non-symmetric systems must
-be considered or the Newton correction system must be preconditioned
-such that the operator is symmetric and short recurrence relation
-methods can be potentially be used.
+operator make them easier to implement than other methods. In most
+nonlinear problems, the Jacobian operator is generally non-symmetric
+and therefore either Krylov methods with long recurrence relations
+that can handle non-symmetric systems must be considered or the Newton
+correction system must be preconditioned such that the operator is
+symmetric and short recurrence relation methods can be potentially be
+used.
 
@@
 set of tokamak plasma problems with the goal of measuring solver
-performance with Newton's method. They note several trade-offs in
-using Krylov methods with the Newton solver. The first is that the
-optimization condition that results from the constraints (e.g. the
-minimization of the GMRES residual over the Krylov space) can be
-relaxed by restricting the size of the subspace such that only a fixed
-number of subspace vectors may be maintained, thus reducing memory
-requirements.
-
-We can also relax the optimization condition by instead restarting the
-recurrence relation with a new set of vectors once a certain number of
-vectors have been generated. The optimization condition is maintained
-over that particular set of vectors, however, Knoll and McHugh note
-that this ultimately slows the convergence rate as compared to keeping
-all vectors as the new set of vectors is not necessarily orthogonal to
-the previous set, and therefore not optimal over the entire iteration
-procedure. The orthogonality condition can be relaxed by using a
-recurrence relation that does not generate a strictly orthonormal
-basis for the Krylov subspace such as the Lanzcos biorthogonalization
-procedure, resulting in memory savings due to the shorter Lanzcos
-recurrence relation. Based on their numerical analysis, they observed
-GMRES to be the most robust for any given nonlinear problem and
-therefore used it for subsequent work in this area.
+performance with Newton's method. Based on their numerical analysis,
+they observed GMRES to be the most robust for any given nonlinear
+problem and therefore used it for subsequent work in this area. Using
+their analysis, this work will also use GMRES as the linear solver for
+all Newton-Krylov reference calculations.
 
@@
 all functions discretely represented in a computer are ultimately
-represented by elementary mathematical operations. If the chain rule
-is applied to those elementary operations, then the derivatives of
-those functions can be computed to the order of accuracy of their
-original discretization in a completely automated way
+formed by elementary mathematical operations. If the chain rule is
+applied to those elementary operations, then the derivatives of those
+functions can be computed to the order of accuracy of their original
+discretization in a completely automated way
 \cite{averick_computing_1994}.
@@
 where $\ve{J}_{k_i} = \partial e_{k_i} / \partial P_i u$ is the
-$k^{th}$ element function Jacobian, $\ve{Q} \in \mathbb{R}^{n_{k_i}
-  \times N}$ is a projector onto the element domain and $\ve{P} \in
-\mathbb{R}^{m_{k_i} \times N}$ a projector onto the element range for
+$k^{th}$ element function Jacobian, $\ve{Q} \in \mathbb{R}^{N \times
+  n_{k_i}}$ is a projector onto the element domain and $\ve{P} \in
+\mathbb{R}^{m_{k_i} \times N}$ a projector from the element range for
 $\ve{F}(\ve{u}) \in \mathbb{R}^{N \times N}$. The Jacobian matrix for
@@
 In physics codes based on nonlinear equations sets, Newton-Krylov
-methods are the primary means of generating a fully consistent
+methods are the primary means of generating a fully implicit
 solution scheme
@@
 MCSA to apply and therefore the current preconditioning strategy will
-destroy all potential memory benefits of FANM.
+eliminate all potential memory benefits of FANM.
 
@@
 solutions for the 2-dimensional, steady, incompressible Navier-Stokes
-equations on a rectilinear grid in much the same way as Shadid and
-Pawlowski's work on Newton-Krylov methods for the solution of these
-equations \cite{shadid_inexact_1997,pawlowski_globalization_2006}. We
-define these equations as follows:
+equations on a rectilinear grid as outlined in Shadid and Pawlowski's
+work on Newton-Krylov methods
+\cite{shadid_inexact_1997,pawlowski_globalization_2006}. We define
+these equations as follows:
 \begin{subequations}
@@
 method. Further, physical parameters within these equations can be
-tuned to enhance the nonlinearities. We will then apply these
-equations to the following three standard benchmark problems that
-cover convection dominated flow and flow dominated by inertial
-forces in two different geometries.
+tuned to enhance the nonlinearities. We next apply these equations to
+the following three standard benchmark problems that cover convection
+dominated flow and flow dominated by inertial forces in two different
+geometries.
 
@@
 as well with the fluid, given the increased rotation observed for the
-isotherms in the figure. As the Rayleigh number increases, thermal
-energy and momentum transport in the system become more dominated by
-convection rather than diffusion, causing the nonlinearities in the
-advective derivatives to grow.
+isotherms in the figure. As the Rayleigh number and rotation
+increases, thermal energy and momentum transport in the system become
+more dominated by convection rather than diffusion, causing the
+nonlinearities in the advective derivatives to grow.
 
@@
 over this step will generate a recirculating backflow under the inlet
-flow towards the step and at higher Reynolds numbers a recirculating
+flow towards the step and at higher Reynolds numbers a recirculation
 zone will form farther downstream from the step. As in the lid driven
@@
 Navier-Stokes equations and Newton-Krylov methods with Jacobian
-generation through automatic differentiation. To achieve these
-results, a Monte Carlo solver generated for this work was injected
-into the nonlinear solver sequence within Drekar to form an
+generation through automatic differentiation. To achieve the results
+presented here, a Monte Carlo solver generated for this work was
+injected into the nonlinear solver sequence within Drekar to form an
 implementation of the FANM method. For the Newton-Krylov solutions,
@@
 number of the system was achieved by keeping the parameters given in
-Table~\ref{tab:thermal_convection_parameters} constant and varying
+Table~\ref{tab:thermal_convection_parameters} constant and varying the
 temperature of the hot side of the cavity. In addition to multigrid
@@
 verified by computing the point-wise absolute difference with the
-Newton-Krylov results. No variation in solution was noted for each of
-the computations larger than the floating point tolerance reported by
-Drekar with an absolute difference value of zero observed at all
-spatial locations for each quantity.
+Newton-Krylov results. No variation in the solution was noted for each
+of the computations larger than the floating point tolerance reported
+by Drekar with an absolute difference value of zero computed at all
+spatial locations for each physical quantity.
 
@@
 computing the point-wise absolute difference with the Newton-Krylov
-results. No variation in solution was noted for each of the
+results. No variation in the solution was noted for each of the
 computations larger than the floating point tolerance reported by
-Drekar with an absolute difference value of zero observed at all
-spatial locations for each quantity.
+Drekar with an absolute difference value of zero computed at all
+spatial locations for each physical quantity.
 
@@
 observed to be true for the \sn{1}{6} Rayleigh number case. For this
-calculation, the number of multigrid cycle applications was bumped up
+calculation, the number of multigrid cycle applications was increased
 to 3 from 1 and FANM converged in the same number of nonlinear
@@
 operations. Table~\ref{tab:convection_linear_iter_comparison} gives
-the total number of GMRES iterations required to converge the problem
-for the Newton-Krylov method and the total number of MCSA iterations
-for the FANM method at each Rayleigh number used. For all cases, fewer
-MCSA iterations were required than GMRES iterations to converge the
-problem even with both solvers preconditioned with the same multigrid
-package and the same parameters. For the \sn{1}{6} Rayleigh number
-case where a constant forcing term of \sn{1}{-4} was kept, this
-difference is most drastic with FANM requiring only 64\% of the linear
-iterations required by the Newton-Krylov scheme with the linear solver
-per nonlinear iteration given for this case in
+the total number of GMRES, FANM, and Richardson iterations required to
+converge the Newton method at each Rayleigh number used. For all
+cases, fewer MCSA iterations were required than GMRES iterations to
+converge the problem even with both solvers preconditioned with the
+same multigrid package and the same parameters. For the \sn{1}{6}
+Rayleigh number case where a constant forcing term of \sn{1}{-4} was
+kept, this difference is most drastic with FANM requiring only 64\% of
+the linear iterations required by the Newton-Krylov scheme with the
+linear solver per nonlinear iteration given for this case in
 Figure~\ref{fig:ra1e6_linear_iters}. Even the \sn{1}{5} Rayleigh
@@
 improved on the iterative performance of using only the Richardson
-iteration through the residual Monte Carlo acceleration. If the performance of
-the MCSA implementation used here were to be potentially optimized in
-the future, this improved iterative performance could also mean
-improved time to solution.
+iteration through the residual Monte Carlo acceleration. If the
+performance of the MCSA implementation used here were to be
+potentially optimized in the future, this improved iterative
+performance could also mean improved time to solution.
 
@@
 Newton-Krylov over FANM meaning that the implementation of the FANM
-method used here is significantly slower. As the inexact Newton scheme
-in Drekar was modified by simply swapping out the linear solver used
-at each Newton step, this increased compute time is coming directly
-from using MCSA as the linear solver and the overhead due to
-preconditioning. When preconditioning overhead for the $SP_N$
-equations was considered as shown by the data in
+method used here is significantly slower.
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{cc}\hline\hline
+      \multicolumn{1}{c}{Rayleigh Number}& 
+      \multicolumn{1}{c}{Newton-Krylov Speedup}\\
+      \hline
+      \sn{1}{3} & 338 \\
+      \sn{1}{4} & 336 \\
+      \sn{1}{5} & 346 \\
+      \sn{1}{6} & 465 \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Thermal convection cavity Newton-Krylov speedup
+      over FANM method.} \textit{Speedup values are rounded to the
+      nearest integer. The results presented here were obtained from
+      the benchmark verification calculations.}}
+  \label{tab:convection_speedup_comparison}
+\end{table}
+
+As the inexact Newton scheme in Drekar was modified by simply swapping
+out the linear solver used at each Newton step, this increased compute
+time is coming directly from using MCSA as the linear solver and the
+overhead due to preconditioning. When preconditioning overhead for the
+$SP_N$ equations was considered as shown by the data in
 Figure~\ref{fig:spn_comparison_prec_time}, a similar speedup of
@@
 constant value, also consistent with the $SP_N$ observations.
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{cc}\hline\hline
-      \multicolumn{1}{c}{Rayleigh Number}& 
-      \multicolumn{1}{c}{Newton-Krylov Speedup}\\
-      \hline
-      \sn{1}{3} & 338 \\
-      \sn{1}{4} & 336 \\
-      \sn{1}{5} & 346 \\
-      \sn{1}{6} & 465 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Thermal convection cavity Newton-Krylov speedup
-      over FANM method.} \textit{Speedup values are rounded to the
-      nearest integer. The results presented here were obtained from
-      the benchmark verification calculations.}}
-  \label{tab:convection_speedup_comparison}
-\end{table}
 
@@
 convection cavity problem is readily apparent in the linear iteration
-counts as given by Table~\ref{tab:driven_linear_iter_comparison}. As
-the opposite of what was observed for that benchmark, lid driven
-cavity results show that the linear models were significantly more
-ill-conditioned, requiring many more MCSA and GMRES iterations. For
-all cases except the 700 Reynolds number case, many more MCSA
-iterations were required to converge with the observed GMRES iteration
-count 64\% of that observed for MCSA when looking at the 100 Reynolds
-number case. For the 700 Reynolds number case, fewer MCSA iterations
-were observed due to the fact that 4 fewer nonlinear iterations were
-performed. The average number of linear iterations per nonlinear
-iteration was still much larger for the FANM calculations. In all
-cases, MCSA improved on the iterative performance of using only the
-Richardson iteration through the residual Monte Carlo acceleration
-even considering the cases with Reynolds numbers of 100 and 700 where
-an extra Newton-Richardson iteration was required to converge the
-nonlinear model.
+counts as given by Table~\ref{tab:driven_linear_iter_comparison}. Lid
+driven cavity results show that the linear models were significantly
+more ill-conditioned, requiring many more MCSA and GMRES
+iterations. For all cases except the 700 Reynolds number case, many
+more MCSA iterations were required to converge with the observed GMRES
+iteration count 64\% of that observed for MCSA when looking at the 100
+Reynolds number case. For the 700 Reynolds number case, fewer MCSA
+iterations were observed due to the fact that 4 fewer nonlinear
+iterations were performed. The average number of linear iterations per
+nonlinear iteration was still much larger for the FANM
+calculations. In all cases, MCSA improved on the iterative performance
+of using only the Richardson iteration through the residual Monte
+Carlo acceleration even considering the cases with Reynolds numbers of
+100 and 700 where an extra Newton-Richardson iteration was required to
+converge the nonlinear model.
 
@@
 relative to the observations for the first two benchmark problems of
-O(100). However, the we expect the ill-conditioning to cause the Monte
+O(100). However, we expect the ill-conditioning to cause the Monte
 Carlo histories in the MCSA solver to have longer random walk lengths
@@
   the problem
-\item With future optimization and improved preconditioning, reactor
-  accident analyses that rely on convection dominated transport or
-  steady state calculations requiring driven flow could experience
-  improved time to solution with FANM given the improved iterative
-  performance observed for problems of these types
 \end{itemize}

### 2013-08-23 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
cleaning up some whitespace to reduce the length
spell check for spn chapter edits
spell check for nonlinear chapter

@@
   \begin{center}
-    \scalebox{0.6}{ \input{chapters/parallel_mc/msod_construction.pdftex_t} }
+    \scalebox{0.55}{ \input{chapters/parallel_mc/msod_construction.pdftex_t} }
   \end{center}
@@
   \begin{center}
-    \scalebox{0.6}{ \input{chapters/parallel_mc/msod_tally.pdftex_t} }
+    \scalebox{0.55}{ \input{chapters/parallel_mc/msod_tally.pdftex_t} }
   \end{center}
@@
 
-\clearpage
-
 \subsection{Pure Domain Decomposition Scaling}
@@
 too large.
-
-\clearpage
 
@@
 qualities of the algorithm through parallel scaling studies. The
-following are the significant observations and findings.
+following are the significant observations, findings, and
+contributions.
 

### 2013-08-23 doc/dissertation/document/backmatter/parallel_theory.tex
moving spectral analysis into an appendix

--- 2013-08-22 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
@@
-\chapter{Parallel Monte Carlo Synthetic \\ Acceleration Methods\ }
-\label{ch:parallel_methods}
-
-In this chapter we briefly review particle transport methods for
-domain decomposed Monte Carlo. Considering a domain decomposed
-Neumann-Ulam method, we derive an analytic framework based on
-algebraic quantities from which to estimate aspects of its performance
-when applied to neutron transport systems. We then devise a parallel
-algorithm for the Neumann-Ulam method based on the multiple-set
-overlapping-domain decomposition algorithm and a parallel algorithm
-for the MCSA iteration that leverages the parallel Monte Carlo
-algorithm and general parallel matrix-vector operations. Using
-parallel solutions for the neutron diffusion problem, we verify the
-correctness of the parallel algorithm and implementation by comparing
-the numerical results against production Krylov methods. With the
-verified implementation of the algorithm, we perform parallel scaling
-studies to test its performance on a leadership class machine and
-compare this performance to the production Krylov methods.
-
-%%---------------------------------------------------------------------------%%
-\section{Domain Decomposed Monte Carlo\ }
-\label{sec:msod}
-Large-scale problems will for reasons typically related to memory
-restrictions or performance have their data partitioned such that each
-parallel process owns a subset of the equations in the linear
-system. Given this convention, the Neumann-Ulam Monte Carlo algorithm
-must perform random walks over a domain that is decomposed and must
-remain decomposed to avoid the same performance and memory
-restrictions that required the domain to be decomposed. To parallelize
-this algorithm, we then seek parallel Monte Carlo algorithms that
-handle domain decomposition.
-
-To motivate this problem, consider the square domain presented in
-Figure~\ref{fig:ddmc_example} and on this domain we imagine a Monte
-Carlo particle transport problem. If the domain were decomposed into 9
-subdomains as shown, each of those subdomains and their associated
-data (i.e. cross sections) could be owned by a different parallel
-process in the computation. In Figure~\ref{fig:ddmc_example}, the
-tracks of 3 particles that are born in the center subdomain are
-shown. In this example, particle A is first transported from the
-center domain to the domain directly to the left. Before the
-scattering event in the new domain may be processed, particle A must
-be communicated between between the two parallel processes that own
-those domains. For any given particle in the transport simulation,
-this communication event may hardly occur during the lifetime of the
-particle (particle B), or may occur many times (particle C) depending
-on the problem parameters and the random path in phase space taken by
-the particle. Scalable parallel algorithms for domain decomposed Monte
-Carlo are those that handle this domain-to-domain communication of
-particles effectively and balance it with the on-process computations
-for particle interactions, tallies, response calculations, and other
-simulation requirements.
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.5}{
-      \input{chapters/parallel_mc/ddmc_example.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Domain decomposed Monte Carlo transport example
-      illustrating how domain decomposition requires parallel
-      communication of particles.} \textit{If a particle crosses a
-      domain boundary it must be communicated to the next domain in
-      its path.}}
-  \label{fig:ddmc_example}
-\end{figure}
-
-For a domain decomposed Neumann-Ulam method, the situation is very
-much the same where instead the transport process is now discrete and
-the "physics" of the transport process is the Monte Carlo game with
-probabilities and weights described by
-Eq~(\ref{eq:neumann_ulam_decomposition}) for the given linear
-problem. Figure~\ref{fig:ddnu_example} gives the domain decomposed
-Neumann-Ulam analog of particle transport problem in
-Figure~\ref{fig:ddmc_example}. Imagine in this problem that we are
-solving for the monoenergetic scalar neutron flux as in the model
-diffusion problem presented in Appendix~\ref{chap:diffusion_problem}.
-In this problem, the solution has only a spatial dependence and
-therefore each point in the mesh corresponds to an equation in the
-resulting linear system. In Figure~\ref{fig:ddnu_example}, the domain
-has now been discretized by this mesh and again decomposed into 9
-subdomains, each owning a piece of the mesh and therefore the
-corresponding equations in the linear system. As we play the Monte
-Carlo game presented in Chapter~\ref{ch:stochastic_methods}, the
-discretization and resulting Neumann-Ulam decomposition of the problem
-describes how each state in the system is coupled to the other states
-in the system through the set of equations that form the linear
-system. This coupling is then responsible for the discrete random walk
-that each history takes as shown in this example. As discrete states
-(or rows in the linear system) that do not exist in the local domain
-are reached during the random walk by stochastic histories, the
-histories must be communicated to the subdomain that does own the
-discrete state in an analogous fashion to the particles in the
-previous example.
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.5}{
-      \input{chapters/parallel_mc/ddnu_example.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Domain decomposed Neumann-Ulam example illustrating
-      how domain decomposition requires parallel communication of
-      histories.} \textit{Each mesh point corresponds to an equation
-      in the linear system and the coupling among equations is
-      described by the discretization of the problem.}}
-  \label{fig:ddnu_example}
-\end{figure}
-
-To further motivate using a parallel algorithm similar to particle
-transport methods, again consider history A in
-Figure~\ref{fig:ddnu_example}. This history begins in the center
-subdomain and through the course of transitioning through states in
-the system arrives at a state that exists in the subdomain directly to
-the left of the center subdomain. History A must then be communicated
-from the center subdomain to the subdomain immediately to the left to
-continue the random walk. As with the particle transport system, we
-may find that histories in a domain decomposed Neumann-Ulam method are
-only communicated a few times (history B) if at all, or they may be
-communicated many times (history C) depending on the discretization
-and the outcome of the Monte Carlo game. In identical fashion to the
-particle transport problem, a domain decomposed Neumann-Ulam method
-has the same communication requirements with all histories in both
-examples requiring the same parallel operations. The amount of
-communication of histories from domain to domain will be the primary
-factor in the parallel scalability of the algorithm. In the next
-section, the number of histories communicated in a given problem will
-be quantified analytically for a simple model problem.
-
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{Analytic Framework for Domain-Decomposed Monte Carlo\ }
-\label{sec:analytic_framework}
-To date, parallel Neumann-Ulam methods have been limited to full
-domain replication with parallelism exploited through individual
-histories \cite{alexandrov_efficient_1998} and in this chapter we
-will exploit particle transport algorithms to alleviate this. To
-accomplish this, we recognize from the literature that stochastic
-histories must be transported from domain to domain as the simulation
-progresses and they transition to states that are not in the local
-domain. Because we have chosen a domain decomposition strategy in a
-parallel environment, this means that communication of these histories
-must occur between compute nodes owning neighboring pieces of the
-global domain. We wish to characterize this communication not only
-because communication is in general expensive, but also because these
+\chapter{Analytic Framework for Domain-Decomposed Monte Carlo\ }
+\label{chap:parallel_theory}
+Because we have chosen a domain decomposition strategy in a parallel
+environment, this means that communication of these histories must
+occur between compute nodes owning neighboring pieces of the global
+domain. We wish to characterize this communication not only because
+communication is in general expensive, but also because these
 nearest-neighbor communication sequences, specifically, have poor
@@
 
-\subsection{Spectral Analysis }
+\section{Spectral Analysis }
 \label{subsec:spectral_analysis}
@@
 diffusion operator $\ve{D}$ by applying the operator to the
-eigenfunctions and noting that $x=ih$ and $y=jh$:
+eigenfunctions and by noting that $x=ih$ and $y=jh$:
 \begin{multline}
@@
 preconditioned iteration matrix is $\ve{H} = \ve{I} -
-\ve{M}^{-1}\ve{D}$, the we except all terms on the diagonal of the
+\ve{M}^{-1}\ve{D}$, then we expect all terms on the diagonal of the
 iteration matrix to be zero such that we have the following stencil:
@@
 matrix. We find that the maximum eigenvalue exists when $p=q=0$,
-giving the following for the spectral radius of the Jacobi
-preconditioned iteration matrix:
+giving the following spectral radius for the Jacobi preconditioned
+iteration matrix:
 \begin{equation}
@@
 
-\subsection{Domain Leakage Approximations }
+\section{Domain Leakage Approximations }
 \label{subsec:domain_leak_approx}
@@
 
-\subsection{Numerical Experiments }
+\section{Numerical Experiments }
 \label{subsec:numerical_experiments}
@@
 \times 100$ square grid with $h=0.01$, $h=0.1$, and $h=1.0$ and the
-absorption cross varied from 0 to 100 while the scattering cross
-section was fixed at unity. Figure~\ref{fig:measured_spec_rad} gives
-the measured spectral radius of the iteration matrix and the computed
-spectral radius for the preconditioned diffusion operator using
-Eq~(\ref{eq:iteration_radius}) as function of the absorption to
+absorption cross section varied from 0 to 100 while the scattering
+cross section was fixed at unity. Figure~\ref{fig:measured_spec_rad}
+gives the measured spectral radius of the iteration matrix and the
+computed spectral radius for the preconditioned diffusion operator
+using Eq~(\ref{eq:iteration_radius}) as function of the absorption to
 scattering ratio $(\Sigma_a / \Sigma_s)$. Excellent agreement was
@@
 
-\subsubsection{Random Walk Length }
+\subsection{Random Walk Length }
 \label{subsubsec:walk_length}
-With the eigenvalue derivations verified, we can go about setting up
-an experiment to measure the length of the random walks generated by
-the adjoint Neumann-Ulam solver. To do this, we again use a $100
-\times 100$ square grid with $h=0.1$ and the absorption cross varied
-from 0 to 100 while the scattering cross section was fixed at
-unity. Three weight cutoff values of \sn{1}{-2}, \sn{1}{-4}, and
-\sn{1}{-8} were used with 10,000 histories generated by a point source
-of strength 1 in the center of the domain. For each of the histories,
-the number of transitions made was tallied to provide an effective
-value of $k$ for each history. This value was then averaged over all
-histories to get a measured value of $k$ for the particular
-operator. On the left, Figure~\ref{fig:measured_length} presents these
-measurements as well as the analytic result computed by
-Eq~(\ref{eq:analytic_k}) as a function of the iteration matrix
-spectral radius, $\rho(\ve{H})$. On the right,
-Figure~\ref{fig:measured_length} gives the relative difference between the
-predicted and observed results. We note good qualitative agreement
-between the measured and analytic results. However, we observe a
-larger relative difference for both long and short random walks.
+Next we perform an experiment to measure the length of the random
+walks generated by the adjoint Neumann-Ulam solver. To do this, we
+again use a $100 \times 100$ square grid with $h=0.1$ and the
+absorption cross section varied from 0 to 100 while the scattering
+cross section was fixed at unity. Three weight cutoff values of
+\sn{1}{-2}, \sn{1}{-4}, and \sn{1}{-8} were used with 10,000 histories
+generated by a point source of strength 1 in the center of the
+domain. For each of the histories, the number of transitions made was
+tallied to provide an effective value of $k$ for each history. This
+value was then averaged over all histories to get a measured value of
+$k$ for the particular operator. On the left,
+Figure~\ref{fig:measured_length} presents these measurements as well
+as the analytic result computed by Eq~(\ref{eq:analytic_k}) as a
+function of the iteration matrix spectral radius, $\rho(\ve{H})$. On
+the right, Figure~\ref{fig:measured_length} gives the relative
+difference between the predicted and observed results. We note good
+qualitative agreement between the measured and analytic
+results. However, we observe a larger relative difference for both
+long and short random walks.
 \begin{figure}[t!]
@@
 
-\subsubsection{Domain Leakage }
+\subsection{Domain Leakage }
 \label{subsubsec:domain_leakage}
-Finally, we seek to measure the leakage from a domain in a domain
-decomposed Monte Carlo calculation and assess the quality of our
-analytic relation for the optical thickness of a domain and the
-associated leakage approximations. For this experiment, a square grid
-with $h=0.1$ was decomposed into 9 square domains, 3 in each cardinal
-direction with measurements occurring in the central domain without
-boundary grid points. For the cross sections, the absorption cross
-section was varied from 1 to 100 while the scattering cross
-section was set to zero to create a purely absorbing environment
-with weight cutoff of \sn{1}{-4}. The optical thickness of these
-domains will vary as a function of the absorption cross section if
-the other parameters are fixed.
+Finally, we seek to measure the leakage from a domain and assess the
+quality of our analytic relation for the optical thickness of a domain
+and the associated leakage approximations. For this experiment, a
+square grid with $h=0.1$ was decomposed into 9 square domains, 3 in
+each cardinal direction with measurements occurring in the central
+domain without boundary grid points. For the cross sections, the
+absorption cross section was varied from 1 to 100 while the scattering
+cross section was set to zero to create a purely absorbing environment
+with a weight cutoff of \sn{1}{-4}. The optical thickness of these
+domains will then vary as a function of the absorption cross section
+if the other parameters are fixed.
 
@@
 given by Eq~(\ref{eq:iteration_radius}), we also need the parameters
-$n_i$ and $n_s$ which respectively describe the typical domain length
+$n_i$ and $n_s$, which respectively describe the typical domain length
 and the average number of states moved along that typical length per
@@
 \end{figure}
-From these difference results, the mean chord approximation is shown to
-have a lower difference for ill-conditioned systems as compared to the
-Wigner approximation while the Wigner approximation produces less
-difference for more well-conditioned systems. We also note that for the
-optically thick domains, the difference is likely corresponded to that
-observed in Figure~\ref{fig:measured_length} for the $k$ parameter
-while the large relative difference in $k$ for optically thin domains does
-not affect the approximation significantly. In general, the mean chord
-approximation is a better choice to estimate the leakage fraction in a
-domain from the adjoint Neumann-Ulam method and except for a single
-data point with $n_i=50$, the mean chord approximation yielded leakage
-fractions within 0.05 of the measured results. As the domain becomes
-more optically thick (with both increasing $n_i$ and decreasing
-$\rho(\ve{H})$), the approximations are more accurate.
-
-%%---------------------------------------------------------------------------%%
-\section{Domain Decomposed Neumann-Ulam Algorithm\ }
-\label{sec:asynchronous_algorithm}
-In the context of radiation transport, in 2009 Brunner and colleagues
-provided a fully asynchronous domain decomposed parallel algorithm as
-implemented in production implicit Monte Carlo codes
-\cite{brunner_efficient_2009}. We will adapt their algorithm and
-directly apply it to a parallel formulation of the Neumann-Ulam
-method. Direct analogs can be derived from their works by noting that
-the primary difference between solving a linear transport system with
-Monte Carlo methods and traditional fixed source Monte Carlo transport
-problems is the content of the Markov chains that are generated.
-
-The transitions represented by these chains are bound by probabilities
-and weights and are initiated by the sampling of a discrete source. In
-the context of transport problems, those transitions represent events
-such as particle scattering and absorption with probabilities that are
-determined by physical data in the form of cross sections. For
-stochastic matrix inversion, those transitions represent moving
-between the equations of the linear system (and therefore the
-components of phase space which they represent) and their
-probabilities are defined by the coefficients of those
-equations. Ultimately, we tally the contributions to generate
-expectation values in the desired states as we progress through the
-chains. Therefore, parallel methods for Monte Carlo radiation
-transport can be abstracted and we can use those concepts that apply
-to matrix inversion methods as an initial means of developing a
-parallel Neumann-Ulam-type solver.
-
-In this section Brunner and Brantley's fully asynchronous algorithm,
-which was effectively implemented verbatim for this work, is presented
-along with its application to the Neumann-Ulam method. For
-considerably more detail, the algorithm presented can be found in
-\cite{brunner_efficient_2009}. In their work they identify two data
-sets that are required to be communicated: the sharing of particles
-that are transported from one domain to another and therefore from one
-processor to another and a global communication that signals if
-particle transport has been completed on all processors. Both of these
-communication sequences will be addressed at a high level along with
-how the Monte Carlo data structures they require are constructed in
-parallel.
-
-\subsection{Parallel Transport Domain and Source }
-\label{subsec:domain_generation}
-To utilize a parallel transport algorithm, we must generate the
-required data with the correct parallel decomposition. For the
-Neumann-Ulam method, the transport domain consists of all states in
-the system that are local, and the probabilities and weights for all
-state transitions possible in the local domain. At the core of this
-representation is the Neumann-Ulam decomposition of the linear
-operator as given by Eq~\ref{eq:neumann_ulam_decomposition}. Given an
-input decomposition for the linear operator, by definition the
-Neumann-Ulam decomposition will have the same parallel
-decomposition. In addition, any modification made to the linear
-operator through preconditioning or relaxation parameters will also
-modify the Neumann-Ulam decomposition. All possible transitions are
-described by the local graph of the sparse input matrix. In addition,
-any left preconditioning or relaxation parameters will modify the
-right hand side of the linear system and therefore the fixed source in
-the Monte Carlo calculation. This source vector will have the same
-parallel decomposition as the input matrix and therefore all of the
-birth states for the histories will exist for at least a single
-transition event within the local domain.
-
-Compared to the serial construction of the Neumann-Ulam decomposition,
-data for all states that is required to be on process must be
-collected in parallel. For the pure domain decomposition case, given
-an input parallel decomposition for the Neumann-Ulam decomposition,
-for all local states $m$ we require access to $P_{mn}$ and $W_{mn}$
-(or $P^T_{mn}$ and $W^T_{mn}$) for the adjoint method for all possible
-states $n$. Given by Figure~\ref{fig:diffusion_graph}, consider the
-adjacency graph of a single state in the neutron diffusion matrix for
-the model problem presented in Appendix~\ref{chap:diffusion_problem}.
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.5}{ \input{chapters/parallel_mc/stencil_graph.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Neutron diffusion equation state adjacency graph.}
-    \textit{The structure of the graph comes from the discretization
-      of the Laplacian operator describing the diffusion
-      physics. Adjacent boundary states are collected by traversing
-      the graph one step for all local boundary states and collecting
-      the equations for those states that are not local. For a given
-      transition in the random walk for the diffusion problem, a
-      history at mesh point $(i,j)$ may transition to all adjacent
-      mesh points including itself, corresponding to a global state
-      transition from $m$ to $n$ in the linear system.}}
-  \label{fig:diffusion_graph}
-\end{figure}
-
-In this example, we currently reside in state $m$ of the system,
-directly correlating to physical location $(i,j)$ in the mesh (the
-center node). The discretization stencil of the diffusion problem
-dictates the structure of this graph and the states to which a history
-may transition. If we play the Monte Carlo game to move to a new state
-$n$, then we may move to any of the nodes in this graph, including the
-node at which we started. If grid point $(i,j)$ in this example is on
-the right of the local domain boundary and we transition to the right
-to state $n$ corresponding to mesh point $(i+1,j)$, we are now in a
-state that is owned by the adjacent domain. We must then gather the
-data in the Neumann-Ulam decomposition from the neighboring process
-that owns grid point $(i+1,j)$. Doing this provides us with the data
-required by the estimators permitting $P_{mn}$ and $W_{mn}$ to be
-computed locally for this particular boundary transition. In addition,
-we also collect the identification number for the process that owns
-this state such that we have an address to send all histories that
-leave the local domain boundary for this state. For the parallel
-source, these adjacent states are not required as histories may only
-be born in states in the local domain. Once the local Neumann-Ulam
-decomposition has been generated with the proper data collected from
-adjacent domains, Monte Carlo transport may proceed in parallel.
-
-\subsection{Domain Decomposed Algorithm}
-\label{subsec:parallel_mc_algorithm}
-
-Here we present in detail Brunner and Brantley's 2009 algorithm in
-detail and discuss how it is adapted to parallelize the Neumann Ulam
-method. Presented in Algorithm~\ref{alg:parallel_mc_algorithm}, the
-top level sequence performs the task of managing history transport
-through the local domain, communication of histories to adjacent
-domains, and the completion of transport. For each of these specific
-tasks, additional algorithms, shown in bold in
-Algorithm~\ref{alg:parallel_mc_algorithm}
-(e.g. \textbf{LocalHistoryTransport()}), are presented for additional
-detail in the same manner as Brunner and Brantley.
-
-\begin{algorithm}[h!]
-  \caption{Parallel Neumann-Ulam Algorithm}
-  \label{alg:parallel_mc_algorithm}
-  \begin{algorithmic}[1]
-    \State get list of neighbor processors 
-    \Comment{Each neighbor owns an adjacent subdomain}
-    \ForAll{neighbors}
-    \State post nonblocking receive for maximum history buffer size
-    \State allocate history buffer
-    \EndFor
-    \State \textit{historiesCompleted} = 0
-    \Comment{local+child finished histories}
-    \State \textit{localProcessed} = 0
-    \Comment{local trajectories computed (not necessarily finished)}
-    \State calculate parent and children processor ID numbers in
-    binary tree
-    \ForAll{child processes}
-    \State post nonblocking receive for \textit{historiesCompleted} tally
-    \EndFor
-    \State post nonblocking receive for stop message from parent
-    \While{stop flag not set}
-    \If{any local histories in source or stack}
-    \State \textbf{LocalHistoryTransport()}
-    \State ++\textit{localProcessed}
-    \EndIf
-    \If{message check period == \textit{localProcessed} || no local histories}
-    \State \textbf{ProcessMessages()}
-    \State \textit{localProcessed} = 0
-    \EndIf
-    \If{no local histories}
-    \State \textbf{ControlTermination()}
-    \EndIf
-    \EndWhile
-    \State cancel outstanding nonblocking requests
-    \State free all history buffers
-  \end{algorithmic}
-\end{algorithm}
-
-Successful execution of this algorithm requires construction of the
-parallel Neumann-Ulam decomposition as described in the preceding
-section. This data is used to begin
-Algorithm~\ref{alg:parallel_mc_algorithm} where lines 1-5 use the
-collected list of neighboring processors generated while building the
-Neumann-Ulam decomposition to setup the set of asynchronous messages
-required for domain-to-domain communication. Again, consider this
-communication pattern for the 9 subdomain example given by
-Figure~\ref{fig:nearest_neighbor_comm}. In this pattern, each boundary
-domain has two neighbors and the center domain four neighbors with
-which they will communicate parallel histories and this communication
-goes both ways as represented by the adjacent arrows in the
-figure. For each set of neighbors, a non-blocking send and receive is
-required with data buffers allocated with a user-defined size prepared
-for incoming and outgoing histories. This nonblocking structure is
-critical to the performance of the algorithm in that it permits local
-history transport to continue while new histories to transport are
-being collected in the buffers. When a given process is ready to do
-more work, it can check these data buffers for incoming histories
-requiring further transport. In this way there is a maximum amount of
-overlap between communication and transport of histories.
-
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.5}{ \input{chapters/parallel_mc/domain_to_domain.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Nearest neighbor history communication sequence.}
-    \textit{Each subdomain in the system has a set of nearest
-      neighbors determined by the parallel adjacency graph of the
-      input matrix. The subdomains are indexed by an integer.}}
-  \label{fig:nearest_neighbor_comm}
-\end{figure}
-
-Once the nearest-neighbor communication sequence has been prepared,
-the completion of transport sequence is readied in lines 6 and 8-12 in
-Algorithm~\ref{alg:parallel_mc_algorithm}. In these lines we are
-setting up an asynchronous binary communication tree as presented for
-the same 9 subdomain example in Figure~\ref{fig:binary_comm_tree}. In
-this communication pattern, each process has one parent process
-(except for the MASTER process 0) to which it will nonblocking send
-the number of histories that terminated their transport procedure by
-weight cutoff in the local domain. Equivalently, each process has up
-to two child processes from which it will receive their completed
-number of histories with a nonblocking receive operation. Setting up a
-tree in this manner lets the completed history tally
-(\textit{historiesCompleted} in the algorithm) be updated
-incrementally and funneled to the root process. Because we are
-solving fixed source problems with the Neumann-Ulam algorithm without
-any type of variance reduction that may generate more histories that
-we started with, once the root process completion sum tallies to the
-number of histories in the source, transport is complete. Once this
-occurs, the root process nonblocking sends a process to its children
-and each process nonblocking receives a stop message from its parent
-as shown in Figure~\ref{fig:binary_comm_tree}. The stop message is
-then propagated up the tree in the same manner.
-
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.25}{
-      \input{chapters/parallel_mc/binary_comm_tree.pdftex_t} }
-    \caption{\textbf{Binary communication tree for coordinating the
-        end of a parallel Neumann-Ulam solve.} \textit{Each child
-        process reports to its parents how many histories completed
-        within its domain. When the root process sums all complete
-        histories, it forwards the stop signal to its children which
-        in turn forward the message up the tree. The subdomains are
-        indexed by an integer.}}
-  \end{center}
-  \label{fig:binary_comm_tree}
-\end{figure}
-
-With these communication structures prepared, we can now enter the
-main transport loop at line 13 of
-Algorithm~\ref{alg:parallel_mc_algorithm}. In this loop, a set of
-mutually exclusive tasks enabled by the fully asynchronous
-communication patterns are executed until the stop signal is received
-from the parent process in the binary tree. In this case, mutually
-exclusivity permits all local operations to occur independently of
-other processes in the system and their current state. For each
-process, there are two data structures from which histories may be
-obtained for the transport procedure. The first is the local source
-and the second is a LIFO (last in first out) stack of histories
-transported to the local domain from the adjacent domain. In lines
-14-17, if histories exist in either of these data structures, then
-they are transported through the local domain using
-Algorithm~\ref{alg:local_history_transport}. If the history is
-terminated by weight cutoff the \textit{historiesCompleted} tally is
-updated. If it hits the boundary of the domain it is added to the
-outgoing history buffer for the neighboring domain and if the buffer
-is full, it is sent to the neighboring domain with a nonblocking
-operation and the memory associated with that buffer reallocated. In
-all instances that Algorithm~\ref{alg:local_history_transport} is
-executed, \textit{localProcessed} is incremented to account for
-histories that have been processed locally.
-
-\begin{algorithm}[h!]
-  \caption{\textbf{LocalHistoryTransport()}}
-  \label{alg:local_history_transport}
-  \begin{algorithmic}[1]
-    \State transport history through the domain until termination
-    \If{history is in neighbor boundary state}
-    \State add history to neighbor buffer
-    \If{neighbor buffer is full}
-    \State nonblocking send history buffer to neighbor
-    \State allocate new history buffer for neighbor
-    \EndIf
-    \Else
-    \If{history terminated by weight cutoff}
-    \State post-process history
-    \State ++\textit{historiesCompleted}
-    \EndIf
-    \EndIf
-  \end{algorithmic}
-\end{algorithm}
-
-Continuing in the transport loop, if there are no local histories to
-transport or the \textit{localProcessed} count has reached some
-user-defined check frequency, lines 18-21 in
-Algorithm~\ref{alg:local_history_transport} check for more histories
-to transport in the incoming data buffers by calling
-Algorithm~\ref{alg:process_messages}. For each neighboring domain in
-the problem, if there are histories in those data buffers then they
-are added to the stack from processing and the nonblocking receive
-operation re-instantiated. In addition, the terminated histories count
-is updated with those values received from the child processes.
-
-\begin{algorithm}[h!]
-  \caption{\textbf{ProcessMessages()}}
-  \label{alg:process_messages}
-  \begin{algorithmic}[1]
-    \ForAll{received history buffers from neighbor}
-    \State unpack number of histories in buffer
-    \State add the histories to the running stack
-    \State repost nonblocking receive with neighbor
-    \EndFor
-    \ForAll{\textit{historiesCompleted} messages from children}
-    \State \textit{historiesCompleted} += message value
-    \State repost nonblocking receive with child
-    \EndFor
-  \end{algorithmic}
-\end{algorithm}
-
-Finally, the transport loop finishes in lines 22-24 of
-Algorithm~\ref{alg:local_history_transport} by calling
-Algorithm~\ref{alg:control_termination} if there are no local
-histories to transport in the stack or the source. In
-Algorithm~\ref{alg:control_termination}, we continue to forward the
-terminated history tally down to the parent in the binary tree and
-send all history buffers to our neighbors, even if they are not
-full. If Algorithm~\ref{alg:control_termination} is called by the
-MASTER process, it checks for completion of the problem and if it is
-complete, forwards to stop flag onto its children as in
-Figure~\ref{fig:binary_comm_tree}. If the process is not the MASTER,
-then we check for a stop signal from the parent and forward to the
-child processes if transport is complete.
-
-\begin{algorithm}[h!]
-  \caption{\textbf{ControlTermination()}}
-  \label{alg:control_termination}
-  \begin{algorithmic}[1]
-    \State nonblocking send an partially full history buffers
-    \ForAll{\textit{historiesCompleted} messages from children}
-    \State \textit{historiesCompleted} += message \textit{historiesCompleted}
-    \State repost nonblocking receive with child
-    \EndFor
-    \If{MASTER processor}
-    \If{\textit{historiesCompleted} == global \# of source histories}
-    \State set stop flag
-    \ForAll{children}
-    \State nonblocking send stop message to child
-    \EndFor
-    \EndIf
-    \Else
-    \State nonblocking send \textit{historiesCompleted} to parent
-    \State \textit{historiesCompleted} = 0
-    \State check for stop signal from parent
-    \If{stop signal from parent}
-    \ForAll{children}
-    \State nonblocking send stop message to child
-    \EndFor
-    \EndIf
-    \EndIf
-  \end{algorithmic}
-\end{algorithm}
-
-In this manner the transport loop continues until all process have
-received the stop signal from their parents. Older versions of this
-algorithm, particularly some of those presented in
-\cite{brunner_comparison_2006}, use a master/slave approach as given
-by Figure~\ref{fig:master_comm_tree} for the completion of a transport
-stage instead of the binary tree scheme. In this approach, the root
-process still manages the final summation of the completion tally, it
-directly receives completed tally results from all other processes in
-the problem instead of from just its children. Although this
-implementation is potentially simpler to understand than the binary
-tree approach, the authors of \cite{brunner_comparison_2006} observed
-that this type of termination pattern, even when implemented in a
-fully asynchronous manner, was a major scalability bottleneck. This
-bottleneck is due to the fact that the master process falls behind the
-others, creating a load imbalance even with the addition of a few
-integer addition operations. We will explicitly demonstrate in scaling
-studies how this bottleneck also occurs within a Neumann-Ulam
-implementation of this algorithm thus requiring the implementation of
-a binary communication tree.
-
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{0.75}{
-      \input{chapters/parallel_mc/master_comm_tree.pdftex_t} }
-    \caption{\textbf{Master/slave scheme for coordinating the end of a
-        parallel Neumann-Ulam solve.} \textit{Each slave process
-        reports to the master how many histories completed within its
-        domain. When the master process sums all complete histories,
-        it sends the stop signal to all slave processes. The
-        subdomains are indexed by an integer.}}
-  \end{center}
-  \label{fig:master_comm_tree}
-\end{figure}
-
-There is an additional advantage to Brunner and Brantley's work that
-although not immediately applicable to this work has the potential to
-provide value in the future. In addition to a robust fully
-asynchronous communication pattern, this algorithm may also be modified
-to account for situations where the total number of histories in a
-given stage are not known before starting transport. From a physics
-perspective, we might expect this for situations where perhaps an
-(n,2n) interaction occurs in a neutronics problem. In this case, the
-algorithm is modified to account for both histories terminated and
-created and several mechanisms are introduced to determine completion
-of the transport stage. For future variations of this work, certain
-variance reduction techniques which create histories, such as
-splitting, have the potential to be successfully employed as a means
-of accelerating the time to solution for a given problem using
-MCSA. The parallel Neumann-Ulam algorithm presented here may be
-adapted to account for these additional techniques.
-
-%%---------------------------------------------------------------------------%%
-\section{Multiple-Set Overlapping-Domain Algorithm\ }
-\label{subsec:msod}
-Although the implementation presented in the previous section was
-observed by Brunner and Brantley to be robust and allowed for scaling
-to large numbers of processors, performance issues were still noted
-with parallel efficiency improvements needed in both the weak and
-strong scaling cases for unbalanced problems. These results led them
-to conclude that a combination of domain decomposition and domain
-replication could be used to solve some of these issues. In 2010,
-Wagner and colleagues developed the \textit{multiple-set
-  overlapping-domain} (MSOD) decomposition for parallel Monte Carlo
-applications for full-core light water reactor analysis
-\cite{wagner_hybrid_2010}. In their work, an extension of Brunner's,
-their scheme employed similar parallel algorithms for particle
-transport but a certain amount of overlap between adjacent domains was
-used to decrease the number of particles leaving the local domain. In
-addition, Wagner utilized a level of replication of the domain such
-that the domain was only decomposed on $O(100)$ processors and if
-replicated $O(1,000)$ times potentially achieves efficient simulation
-on $O(100,000)$ processors, thus providing both spatial and particle
-parallelism.
-
-Each collection of processors that constitutes a representation of the
-entire domain is referred to as a set, and within a set overlap occurs
-among its sub-domains. The original motivation was to decompose the
-domain in a way that it remained in a physical cabinet in a large
-distributed machine, thus reducing latency costs during
-communication. A multiple set scheme is also motivated by the fact
-that communication during particle transport only occurs within a set,
-limiting communications during the transport procedure to a group of
-$O(100)$ processors, a number that was shown to have excellent
-parallel efficiencies in Brunner's work and therefore will scale well
-in this algorithm. The overlapping domains within each set also
-demonstrated reduced communication costs. On each processor, the
-source is sampled in the local domain that would exist if no overlap
-was used while tallies can be made over the entire overlapping domain.
-
-To demonstrate this, consider the example adapted from Mervin's work
-with Wagner and others in the same area \cite{mervin_variance_2012}
-and presented in Figure~\ref{fig:msod_example}.
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.5}{
-      \input{chapters/parallel_mc/msod_example.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Overlapping domain example illustrating how domain
-      overlap can reduce communication costs.}
-    \textit{All particles start in the blue region of interest. The
-      dashed line represents 0.5 domain overlap between domains.}}
-  \label{fig:msod_example}
-\end{figure}
-In this example, 3 particle histories are presented emanating from the
-blue region of interest. Starting with particle A, if no domain
-overlap is used then the only the blue domain exists on the starting
-processor. Particle A is then transported through 3 other domains
-before the history ends, therefore requiring three communications to
-occur in Brunner's algorithm. If a 0.5 domain overlap is permitted as
-shown by the dashed line, then the starting process owns enough of the
-domain such that no communications must occur in order to complete the
-particle A transport process. Using 0.5 domain overlap also easily
-eliminates cases such as that represented by the path of particle
-C. In this case, particle C is scattering between two adjacent
-domains, incurring a large latency cost for a single
-particle. Finally, with particle B we observe that 0.5 domain overlap
-will still not eliminate all communications. However, if 1 domain
-overlap were used, the entire geometry shown in
-Figure~\ref{fig:msod_example} would be contained on the source
-processor and therefore transport of all 3 particles without
-communication would occur.
-
-Wagner and colleagues used this methodology for a 2-dimensional
-calculation of a pressurized water reactor core and varied the domain
-overlap from 0 to 3 domain overlap (a $7 \times 7$ box in the context
-of our example) where a domain contained an entire fuel assembly. For
-the fully domain decomposed case, they observed that 76.12\% of all
-source particles leave the domain. At 1.5 domain overlap, the
-percentage of source particles born in the center assembly leaving the
-processor domain dropped to 1.05\% and even further for 0.02\% for the
-3 domain overlap. Based on their results, we hypothesize that the
-overlap approach coupled with the multiple sets paradigm that will
-enhance the scalability of the pure domain-decomposition algorithm for
-Neumann-Ulam presented in the previous section.
-
-\subsection{Generating the MSOD Decomposition }
-\label{subsec:msod_generation}
-In \S~\ref{subsec:domain_generation} we discussed how we generated the
-parallel transport domain from the Neumann-Ulam decomposition of a
-domain decomposed linear operator. We can readily adapt those data
-structures to account for the extra information required by an
-implementation of the MSOD algorithm. First, we consider the
-generation of overlap. Conveniently, this is identical to the
-neighboring states discovery problem discussed in conjunction with
-Figure~\ref{fig:diffusion_graph}. To generate the boundary for the
-local transport domain, we needed to gather all data from the
-immediately adjacent states in the system that did not already exist
-in the local domain. We solved this problem by traversing the graph
-of the matrix for each of the boundary states and determining which
-processes owned those adjacent states.
-
-To generate overlap, we use the identical algorithm but in this case
-we perform as many graph level traversals as the specified amount of
-overlap. For example, consider the analytic relations derived in
-\S~\ref{sec:analytic_framework} for the length of random walks in a
-Neumann-Ulam method and the amount of leakage from a domain
-considering a neutron diffusion problem. From these relations we might
-determine that, for our particular problem, if we can grow the domain
-by 10 additional discrete state transitions, we can reduce the number
-of histories that must be communicated by a significant fraction. We
-determine the information we must gather from neighboring domains to
-build the overlap by traversing the graph of the matrix outward from
-the boundary 10 levels. At each step, we find the data in the adjacent
-domain that we require and make that data local, incrementing the size
-of the overlap. Once the overlap has been gathered, one extra graph
-traversal on the boundary states is performed to get the new
-neighboring states and their owning processes.
-
-Once overlap has been generated, the transport domain for a single set
-is complete. However, if multiple sets are to be used, an additional
-set of parallel procedures is required to generate the necessary data
-structures. Figure~\ref{fig:msod_construction} gives a schematic
-representation of the MSOD construction process using 4 sets. For a
-problem where $P$ parallel processors are available and $S$ sets are
-to be used, each set is allocated $P/S$ processors for computation. On
-the first $P/S$ processors, the linear problem is generated. From the
-linear problem, the linear operator is used to construct the transport
-domain, the solution vector used to construct the tallies, and the
-forcing term vector used to construct the fixed source. Once these
-Monte Carlo data structures have been generated, they are broadcast to
-the remaining sets in the problem as shown in
-Figure~\ref{fig:msod_construction} such that we have $S$ replications
-of the Monte Carlo problem for a single instance of the original
-linear problem.
-
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{0.6}{ \input{chapters/parallel_mc/msod_construction.pdftex_t} }
-  \end{center}
-  \caption{\textbf{MSOD construction for 4 sets with overlap.}
-    \textit{The linear system is used to construct the Monte Carlo
-      transport domain on the primary set. The Monte Carlo data
-      structures are then broadcast among the blocks to all sets in
-      the system.}}
-  \label{fig:msod_construction}
-\end{figure}
-
-\clearpage
-
-\subsection{MSOD Neumann Ulam Algorithm }
-\label{subsec:msod_algorithm}
-
-With the ability to generate the transport domain within an MSOD
-decomposition as well as the source and tally structures, we can now
-define how to perform a Neumann-Ulam solve using MSOD. Given by
-Algorithm~\ref{alg:msod_transport}, we begin by constructing the Monte
-Carlo data structures for each set including the transport domain, the
-source, and the tallies. Next, in line 4 we perform the asynchronous
-transport algorithm presented in \S~\ref{sec:asynchronous_algorithm}
-for each of the individual sets. Adding overlap to
-Algorithm~\ref{alg:parallel_mc_algorithm} in this case simply modifies
-the set of local states and the set of neighboring states to
-additional states gathered during overlap generation.
-
-\begin{algorithm}[h!]
-  \caption{\textbf{MSOD Transport Sequence}}
-  \label{alg:msod_transport}
-  \begin{algorithmic}[1]
-    \State build MSOD domain
-    \State build Monte Carlo source
-    \State build Monte Carlo tally
-    \State perform parallel Neumann-Ulam transport in each set
-    \State combine set tallies
-    \State combine block tallies
-  \end{algorithmic}
-\end{algorithm}
-
-Once Monte Carlo transport is complete complete, the tallies for each
-individual computation must be gathered to the original set in order
-to build the correction vector within an MCSA sequence. To do this, we
-perform two parallel vector operations. First, overlap not only
-creates additional states in the local transport domain but also
-additional states in the local tally vector such that any events that
-occur in the local domain may be tallied in a local vector. Because of
-this, the tally vector in a single set will share local states with
-other domains and a parallel sum reduction operation is required to
-transform the tally into the original Neumann-Ulam parallel
-decomposition. Second, the tallies in individual sets must be combined
-and applied to the primary set in the problem that shares the
-processor space with the original linear problem. For this operation,
-we employ the concept of blocks as a subdivision of parallel space
-complementary to the concept of sets.
-
-Consider the schematic representation of the final parallel reduction
-presented in Figure~\ref{fig:msod_tally_reduction}. For this
-operation, the tally vectors in each set must be summed together and
-applied to the primary set. As each set is an exact replica of the
-others, the parallel decomposition of the tally vector is the same
-within all sets. We can take advantage of this by building a processor
-space that encapsulates each identical domain in each set. We can then
-use this processor space for the reduction. For the schematic in
-Figure~\ref{fig:msod_tally_reduction}, the upper-left domain in each
-set creates a single block, giving 9 total blocks for this example
-problem. The processor space that encapsulates each of the upper-left
-domains is then used for the parallel reduction operation. This
-reduction will happen in a mutually exclusive manner for each of the
-blocks in the system. One benefit of this approach is that this
-reduction occurs only over a subset of the processors in the problem,
-providing better scalability than a reduction operation over all
-processors in the problem.
-
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{0.6}{ \input{chapters/parallel_mc/msod_tally.pdftex_t} }
-  \end{center}
-  \caption{\textbf{MSOD tally reduction over blocks.} \textit{The
-      tally computed independently in each set is reduced across the
-      blocks to the primary set. The linear system vector is updated
-      on the primary set. In this example the tally vector in the
-      block containing upper-left subdomains is combined. Identical
-      reduction operations occur for all other blocks
-      simultaneously.}}
-  \label{fig:msod_tally_reduction}
-\end{figure}
-
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{Parallel MCSA\ }
-\label{sec:parallel_mcsa}
-With the parallel adjoint Neumann-Ulam solver implementation described
-above, the parallel implementation of an MCSA iteration is
-trivial. Recall the MCSA iteration procedure presented again here for
-clarity:
-\begin{subequations}
-  \begin{gather}
-    \label{eq:mcsa_r1}
-    \ve{x}^{k+1/2} = \ve{x}^k + \ve{r}^k\:,\\
-    \label{eq:mcsa_r2}
-    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\
-    \label{eq:mcsa_r3}
-    \ve{A}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:,\\
-    \label{eq:mcsa_r4}
-    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,\\
-    \label{eq:mcsa_r5}
-    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:.
-  \end{gather}
-\end{subequations}
-In \S\ref{sec:parallel_krylov_methods} we discuss parallel matrix and
-vector operations as utilized in conventional projection methods. We
-utilize these here for the parallel MCSA implementation for the
-computations required in Eqs~(\ref{eq:mcsa_r1}), (\ref{eq:mcsa_r2}),
-(\ref{eq:mcsa_r4}), and (\ref{eq:mcsa_r5}). In the first step, a
-parallel matrix-vector multiply is used to apply the split operator to
-the previous iterate's solution. A parallel vector update is then
-performed with the source vector to arrive at the initial iteration
-guess. In the next step, the residual is computed by the same
-operations where now the operator is applied to the solution guess
-with a parallel matrix-vector multiply and then a parallel vector
-update with the source vector is performed. Once the correction is
-computed with a parallel adjoint Neumann-Ulam solve, this correction
-is applied to the guess with a parallel vector update to get the new
-iteration solution. Additionally, as given by
-Eq~(\ref{eq:mcsa_stopping_criteria}), 2 parallel vector reductions
-will be required to check the stopping criteria: one initially to
-compute the infinity norm of the source vector, and another at every
-iteration to compute the infinity norm of the residual vector.
-
-To parallelize the solution of Eq~(\ref{eq:mcsa_r3}), we apply the
-MSOD Neumann-Ulam algorithm presented in the previous section. The
-problem is potentially replicated with overlap and the Monte Carlo
-procedure returns a correction vector, $\delta \ve{x}$, with the same
-parallel decomposition as the input linear problem. As was the case
-for generating the replications required for a multiple set
-implementation, the explicit matrix and vector operations required for
-the rest of an MCSA iteration occur only on the subset of processors
-containing the primary set. Only the Monte Carlo data structures exist
-on the rest of the processors in the problem as replication of the
-other operations does not generate any more information that could be
-used to accelerate the time to solution.
-
-\subsection{Parallel FANM Implementation}
-\label{subsec:parallel_fanm}
-Here we briefly comment on how a FANM method may be implemented in
-parallel. A parallel FANM method relies on a basic set of parallel
-matrix-vector operations outlined in
-\S~\ref{sec:parallel_krylov_methods} as well as the global residual
-and Jacobian assembly procedure described in
-\S~\ref{subsec:automatic_differentiation}. Consider the FANM iteration
-scheme in Algorithm~\ref{alg:fanm}. We must first assemble the linear
-system in parallel through the element-wise function evaluations to
-generate both the global Jacobian operator and the global residual
-vector on the right hand side. Per Bartlett's work on automatic
-differentiation, efficient and automated parallel mechanisms are
-available to do this through a sequence of scatter/gather
-operations. With these tools available for residual and Jacobian
-generation, the remainder of the parallel procedure is simple. The
-linear Newton correction system is solved using the parallel MCSA
-method as described in the previous section and the Newton correction
-applied to the previous iterate's solution through a parallel vector
-update.
-
-%%---------------------------------------------------------------------------%%
-\section{Parallel MCSA Verification\ }
-\label{sec:parallel_verification}
-Before we consider the parallel performance of the MCSA algorithm, we
-must first verify that the algorithm generates the correct solution in
-parallel with respect to reference solutions provided by production
-Krylov methods. To verify correctness, we use solutions to the neutron
-diffusion system presented in
-Appendix~\ref{chap:diffusion_problem}. As the reference computation,
-serial results will be used from the conjugate gradient solver in
-Belos, a linear solvers package in the Trilinos scientific computing
-libraries \cite{heroux_overview_2005}. In identical fashion to the
-verification for the $SP_N$ equations performed in
-\S~\ref{sec:spn_mcsa_verification}, we will solve the neutron
-diffusion problem in parallel using multiple cores on a desktop
-machine. For each parallel solution, then the solution at each grid
-point in the system will be compared to the solution at the same grid
-point in the serial reference case with the minimum and maximum
-absolute differences of these values reported.
-
-GMRES and conjugate gradient solvers from the Belos package were used
-on 4 cores as a means of both assuring the Krylov solvers produce the
-same results on multiple cores and as a means of providing an
-acceptable range of minimum and maximum values in the comparison. For
-MCSA, a range of MSOD parameters will be used such that problems with
-multiple sets and multiple blocks may be tested along with
-overlap. For all MCSA calculations, a single history was used for
-every DOF in the problem to compute the correction vector. For all
-solvers, the flux was converged to a tolerance of \sn{1}{-8} and
-therefore if the comparisons agree within this tolerance the solutions
-will be deemed
-correct. Table~\ref{tab:parallel_verification_parameters} gives the
-parameters for the verification problem and
-Table~\ref{tab:parallel_mcsa_verification} gives the results of the
-calculations. Solver parameters used for MCSA are given in
-Table~\ref{tab:parallel_mcsa_parameters}.
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lc}\hline\hline
-      \multicolumn{1}{l}{Parameter}& 
-      \multicolumn{1}{c}{Value}\\\hline
-      $dx$ & 0.1 \\
-      $dy$ & 0.1 \\
-      $N_x$ & 400 \\
-      $N_y$ & 400 \\
-      $N$ & 160,000 \\
-      $\Sigma_a$ & 5.0 \\
-      $\Sigma_s$ & 1.0 \\
-      Boundary Type & Vacuum \\
-      Linear Solver Tolerance & \sn{1}{-8} \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Parallel MCSA verification model problem parameters.}
-    \textit{The neutron diffusion equation in 2 dimensions is used for
-      the model problem.}}
-  \label{tab:parallel_verification_parameters}
-\end{table}
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lc}\hline\hline
-      \multicolumn{1}{l}{Parameter}& 
-      \multicolumn{1}{c}{Value}\\\hline
-      Histories & 1 per DOF \\
-      Weight Cutoff & \sn{1}{-2} \\
-      Fixed Point Iteration & Richardson \\
-      Estimator & Adjoint Collision \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Parallel MCSA verification solver parameters.}
-    \textit{No relaxation parameters or variance reduction techniques
-      were used with MCSA.}}
-  \label{tab:parallel_mcsa_parameters}
-\end{table}
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lccccll}\hline\hline
-      \multicolumn{1}{l}{Solver}& 
-      \multicolumn{1}{c}{Blocks}& 
-      \multicolumn{1}{c}{Sets}& 
-      \multicolumn{1}{c}{Cores}& 
-      \multicolumn{1}{c}{Overlap}& 
-      \multicolumn{1}{l}{Min Difference}& 
-      \multicolumn{1}{l}{Max Difference}\\\hline
-      Conjugate Gradient & - & - & 4 & - & 0.0 & \sn{1.11022}{-16} \\
-      GMRES & - & - & 4 & - & \sn{5.64215}{-13} & \sn{3.83348}{-9} \\
-      MCSA & 2 & 1 & 2 & 0 & \sn{8.60423}{-16} & \sn{7.6806}{-9} \\
-      MCSA & 3 & 1 & 3 & 0 & \sn{6.41154}{-15} & \sn{7.65397}{-9} \\
-      MCSA & 4 & 1 & 4 & 0 & \sn{2.66454}{-15} & \sn{7.62623}{-9} \\
-      MCSA & 2 & 2 & 4 & 0 & \sn{1.52656}{-15} & \sn{7.57108}{-9} \\
-      MCSA & 1 & 4 & 4 & 0 & \sn{5.55112}{-16} & \sn{7.61631}{-9} \\
-      MCSA & 8 & 1 & 8 & 0 & \sn{6.13398}{-15} & \sn{7.51744}{-9} \\
-      MCSA & 4 & 2 & 8 & 0 & \sn{1.03251}{-14} & \sn{7.62601}{-9} \\
-      MCSA & 2 & 4 & 8 & 0 & \sn{5.13478}{-15} & \sn{7.60777}{-9} \\
-      MCSA & 1 & 8 & 8 & 0 & \sn{1.80411}{-15} & \sn{7.60651}{-9} \\
-      MCSA & 2 & 1 & 2 & 5 & \sn{2.47025}{-15} & \sn{7.58882}{-9} \\
-      MCSA & 3 & 1 & 3 & 5 & \sn{2.44249}{-15} & \sn{7.6326}{-9} \\
-      MCSA & 4 & 1 & 4 & 5 & \sn{7.49401}{-16} & \sn{7.49014}{-9} \\
-      MCSA & 2 & 2 & 4 & 5 & \sn{1.19349}{-15} & \sn{7.61814}{-9} \\
-      MCSA & 1 & 4 & 4 & 5 & \sn{5.55112}{-16} & \sn{7.61631}{-9} \\
-      MCSA & 8 & 1 & 8 & 5 & \sn{2.05391}{-15} & \sn{7.63955}{-9} \\
-      MCSA & 4 & 2 & 8 & 5 & \sn{9.57567}{-15} & \sn{7.61313}{-9} \\
-      MCSA & 2 & 4 & 8 & 5 & \sn{5.7454}{-15} & \sn{7.60767}{-9} \\
-      MCSA & 1 & 8 & 8 & 5 & \sn{1.80411}{-15} & \sn{7.60651}{-9} \\
-      MCSA & 2 & 1 & 2 & 10 & \sn{1.08247}{-15} & \sn{7.63231}{-9} \\
-      MCSA & 3 & 1 & 3 & 10 & \sn{3.27516}{-15} & \sn{7.60211}{-9} \\
-      MCSA & 4 & 1 & 4 & 10 & \sn{4.996}{-16} & \sn{7.64756}{-9} \\
-      MCSA & 2 & 2 & 4 & 10 & \sn{4.13558}{-15} & \sn{7.60259}{-9} \\
-      MCSA & 1 & 4 & 4 & 10 & \sn{5.55112}{-16} & \sn{7.61631}{-9} \\
-      MCSA & 8 & 1 & 8 & 10 & \sn{9.71445}{-16} & \sn{7.58535}{-9} \\
-      MCSA & 4 & 2 & 8 & 10 & \sn{7.77156}{-16} & \sn{7.65978}{-9} \\
-      MCSA & 2 & 4 & 8 & 10 & \sn{4.16334}{-16} & \sn{7.63191}{-9} \\
-      MCSA & 1 & 8 & 8 & 10 & \sn{1.80411}{-15} & \sn{7.60651}{-9} \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Parallel MCSA verification results.}
-    \textit{Difference values given are absolute values. The neutron
-      diffusion equation in 2 dimensions is used for the model problem
-      with the parameters given in
-      Table~\ref{tab:parallel_verification_parameters}. A serial
-      conjugate gradient calculation is used as the reference
-      computation and neutron fluxes are compared
-      element-wise. Parameters used for the MCSA solver are given in
-      Table~\ref{tab:parallel_mcsa_parameters}. All solvers including
-      the reference computation were converged to tolerance of
-      \sn{1}{-8}.}}
-  \label{tab:parallel_mcsa_verification}
-\end{table}
-
-From the data we see that for all solvers the parallel results agree
-with the reference serial results within the solution tolerance for
-all DOF in the system. In addition, we do not notice any systematic
-differences in the results as a function of number of blocks, number
-of sets, or amount of overlap. In addition, the parallel GMRES results
-have maximum absolute differences of the same magnitude as all MCSA
-results and a larger minimum absolute difference than all MCSA results
-by three orders of magnitude. Based on these results, MCSA
-parallelized with the MSOD algorithm, and in particular the
-implementation used here, generates the correct results with respect
-to the serial conjugate gradient reference computation.
-
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{Leadership-Class Parallel Scaling Studies\ }
-\label{sec:leadership_scaling_studies}
-Using the parallel performance metrics presented in
-Appendix~\ref{chap:parallel_metrics}, we will next measure performance
-in various scaling studies using an implementation of the parallel
-MCSA algorithm to assess its quality using the neutron diffusion
-problem presented in Appendix~\ref{chap:diffusion_problem}. For these
-scaling studies, large problems at high levels of concurrency will be
-solved at a leadership-class computing facility. In this section, we
-will outline the parallel machine on which the calculations were
-performed and then present a series of numerical studies that identify
-both strengths and weaknesses of the parallel algorithm. Both pure
-domain decomposition and MSOD parallel performance will be explored
-and compared to performance results for the same problems using a set
-of production parallel Krylov solvers.
-
-For the scaling results presented here, all parallel MCSA results were
-computed using the Monte Carlo Linear Solvers\footnote{MCLS is an open
-  source C++ library built on the Trilinos framework and can be
-  downloaded at \url{https://github.com/sslattery/MCLS}} (MCLS)
-library generated specifically for this work while the conjugate
-gradient and GMRES Krylov solvers used are part of the Belos linear
-solver package within the Trilinos framework
-\cite{heroux_overview_2005}. In general, the Belos package produces
-solutions to the neutron diffusion problem nearly two orders of
-magnitude faster than the MCLS library for serial computations using
-the conjugate gradient solver. Therefore, when we directly compare
-these two codes, the MCLS parallel efficiencies will appear
-artificially inflated relative to the Belos efficiencies due to the
-fact that the MCLS ratio of work to communication will be much larger
-than that for the Belos calculations. For every scaling study
-presented, the calculations were performed 3 times with the average
-values reported for wall time and those average values used for the
-efficiency computations. In addition, it should be noted that in
-practice, symmetric positive-definite problems like the neutron
-diffusion problem solved here would never be solved using
-GMRES. Instead, a method such as conjugate gradient would typically be
-used due to the enhanced performance. We use GMRES here simply as a
-means of providing an additional comparison for MCSA performance.
-
-\subsection{Titan }
-\label{subsec:titan}
-All parallel scaling studies presented in this work were performed on
-the Titan Cray XK7 machine at the Oak Ridge Leadership Computing
-Facility located at Oak Ridge National Laboratory. Titan is a
-heterogeneous architecture machine combining both CPU and GPU
-components to achieve a theoretical peak performance of more than 20
-petaflops. The technical specifications for Titan for only the CPU
-portion of the machine are given in Table~\ref{tab:titan_hardware}.
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lc}\hline\hline
-      Processor & 16 core AMD \\
-      Nodes & 18,688 AMD Opterons \\
-      Cores/Node & 16 \\
-      Total Cores & 299,008 \\
-      Memory/Node & 32 GB \\
-      Memory/Core & 2 GB \\
-      Interconnect & Gemini \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Titan Cray XK7 hardware specifications.}
-    \textit{Specifications presented for Titan in this table were
-      gathered from the Oak Ridge Leadership Computing Facility at}
-    \url{http://www.olcf.ornl.gov/computing-resources/titan-cray-xk7/}}
-  \label{tab:titan_hardware}
-\end{table}
-
-For the scaling studies presented in this work, only the CPU
-components of Titan specified in Table~\ref{tab:titan_hardware} were
-utilized. This is due to the fact that the implementations of the
-parallel MCSA algorithm in the MCLS library utilized code written only
-with the Message Passing Interface (MPI) and therefore could not take
-advantage of the GPU accelerators. Even considering just the CPU
-portion of Titan, nearly 300,000 CPUs are available to test the
-parallel performance of the MCSA algorithm at high levels of
-concurrency.
-
-For each parallel performance metric presented in the previous
-section, a reference case is required to compute efficiencies for all
-other cases. For smaller machines, often a serial computation can be
-performed and used as the base case to get a true measure of speed-up
-for a particular problem. For our scaling studies, we will always use
-as a reference case a calculation which uses 16 cores, filling an
-entire node of the machine. We fill an entire node so that we may
-ignore the effects on scalability measures that arise from using only
-a subset of cores in a node. Typically this results in a large drop in
-efficiency from a serial computation to a calculation that fills the
-entire node. Once a node is filled, adding subsequent nodes has less
-of an effect on the scaling, providing a more stable region of
-analysis.
-
-\subsection{Communication Parameters}
-\label{subsec:comm_parameters}
-
-In our description of Brunner and Brantley's algorithm as applied to
-the Neumann-Ulam method, we noted two free parameters that are
-user-defined at runtime. The first is the static data buffer size for
-communicating histories between neighboring domains and the second is
-the frequency by which a domain checks for incoming information from
-neighboring domain, both of which may potentially affect latency and
-load balancing. Changing the static history buffer size parameter
-affects the scalability of the algorithm in two ways. If the buffer
-size is increased, the buffer takes longer to fill and is therefore
-sent less frequently, potentially causing load balancing bottlenecks
-where a neighboring domain is idle and waiting for more work to do but
-is waiting for another domain to complete work. In addition, if the
-buffer size is too large, the network may become saturated with these
-very large data packets, slowing down performance. If the buffer size
-is too small, too many history messages are sent, not providing enough
-data for a neighboring process to work on to make up for the time
-spent communicating. Considering the second parameter, when the
-message check frequency period is increased, more histories are
-processed locally before looking to do work elsewhere. If this value
-becomes too large, incoming data buffers may not be cleared frequently
-enough and cause decreased performance in the domain-to-domain
-communications. If this value is too small, too much time is spent
-looking for work instead of actually performing it.
-
-To assess how these parameters affect the performance of the parallel
-Monte Carlo algorithm on the Titan machine, a parameter study was
-performed using the neutron diffusion problem on 64 cores (4
-nodes). This problem had a reasonably well conditioned spectral radius
-of 0.78 with a global problem size of \sn{1.6}{7} DOFs with one
-stochastic history used per DOF in the MCSA Neumann-Ulam used to
-compute the correction for the acceleration. Each parameter was varied
-between 16 and 4096 and the time to solve the problem recorded. The
-results of this study are presented in
-Figure~\ref{fig:titan_comm_parameters}. Compared to the results of
-Brunner and Brantley for the same type of study, we note qualitatively
-similar results with the general observation that a larger history
-buffer check period is better than a shorter one. The larger the check
-period, the more on-process work is done before checking
-messages. Second, there was not a significant variation observed in
-the history buffer size parameter for a given check frequency.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_comm_parameters.pdf}
-  \end{center}
-  \caption{\textbf{Neumann Ulam parallel algorithm history buffer
-      check period vs. history buffer size sensitivity.} \textit{The
-      color scale represents the ratio of the runtime of a particular
-      parameter variation with the runtime from the faster parameter
-      variation. A value of 1.02 on the color scale corresponds to a
-      runtime 2\% longer than the faster time. All runtimes were
-      within 6.5\% of the fastest observed case.}}
-  \label{fig:titan_comm_parameters}
-\end{figure}
-
-Compared to the reference work, there was also not a significant
-change in the runtime over the values of parameters tested with all
-runtimes reported within 6.5\% of the fastest case observed. For this
-work, a history buffer size of 4096 and a buffer check period of 128
-performed best for the initial calculations. However, most values in
-the range tested will give good performance. Brunner and Brantley
-found that a value of 1024 was best for both parameters. For the
-results observed here, these parameters were only 0.7\% slower than
-the fastest case. It may be best practice to fix these parameters
-internally at 1024 as both this work and the reference found these to
-perform well several years apart on both different parallel machines
-and for different transport sequences. Therefore, for the scaling
-studies presented in the following sections, values of 1024 for both
-the history buffer size and message check frequency will be used.
-
-\clearpage
-
-\subsection{Pure Domain Decomposition Scaling}
-\label{subsec:pure_domain_decomp}
-
-For the first set of scaling studies, we consider the parallel MCSA
-algorithm where MSOD has not been used in the parallel Neumann-Ulam
-algorithm. We refer to this case as pure domain decomposition where no
-overlap or replication has been leveraged. Such a test will isolate
-specifically the performance of Brunner and Brantley's algorithm as
-applied to the Neumann-Ulam algorithm. For the strong scaling study,
-the global diffusion problem size was fixed at \sn{1.6}{7} DOFs for
-all solvers and the 16 core run used as the base case. MCSA was used
-again with one stochastic history used per DOF in the Neumann-Ulam
-solve to compute the correction for the acceleration with the
-remainder of the problems and solver parameters the same as those for
-the verification calculation given in
-Table~\ref{tab:parallel_verification_parameters} and
-Table~\ref{tab:parallel_mcsa_parameters}. As with all of the scaling
-studies, 3 calculations for all data points were performed with the
-average wall time reported and used for the efficiency
-calculations. Figure~\ref{fig:titan_pure_strong_time} gives the wall
-time per iteration for these calculations while
-Figure~\ref{fig:titan_pure_strong} gives the results of the this study
-with the given absolute efficiencies computed using
-Eq~(\ref{eq:strong_scaling_absolute}). We report wall time per
-iteration due to the fact that MCSA converged on the solution in 22
-iterations while the Krylov solvers converged in 19 iterations.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_pure_strong_time.pdf}
-  \end{center}
-  \caption{\textbf{Pure domain decomposition strong scaling wall time
-      per iteration.} \textit{Wall time is reported per iteration
-      because the methods converged in different numbers of
-      iterations. MCLS is over an order of magnitude slower
-      arithmetically than the Krylov solvers. GMRES executes more
-      operations than conjugate gradient and is therefore slower as
-      well.}}
-  \label{fig:titan_pure_strong_time}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_pure_strong.pdf}
-  \end{center}
-  \caption{\textbf{Pure domain decomposition strong scaling absolute
-      efficiency.}  \textit{Super-linear speed-up is from memory
-      thrashing in the base case. MCLS is over an order of magnitude
-      slower arithmetically, causing the higher efficiency. GMRES
-      executes more operations than conjugate gradient, also creating
-      a higher efficiency.}}
-  \label{fig:titan_pure_strong}
-\end{figure}
-
-There are several key features of
-Figure~\ref{fig:titan_pure_strong_time} and
-Figure~\ref{fig:titan_pure_strong} worth noting. First, up to around
-5,000 cores, all methods experience a super-linear improvement in
-parallel efficiencies. For all methods, this increase is not a
-function of the algorithmic efficiency as those computed using
-Eq~(\ref{eq:algorithmic_efficiency}) remained at unity, meaning that
-all computations converged in the same number of iterations relative
-to the base case\footnote{When the algorithmic efficiency remains at
-  unity for all calculations, the implementation efficiency is
-  equivalent to the absolute efficiency and therefore is not presented
-  here.}. Instead, what is happening here and was observed by Brunner
-and Brantley in their work is that the large problem used to scale to
-O(10,000) cores was large enough that memory thrashing was occurring
-in the 16 core base case. As the number of cores in the strong scaling
-study was increased and the local problem size decreased, more of the
-local problem fit into the cache of the system, thus causing improved
-serial performance and therefore enhanced runtimes. In this region of
-super-linear efficiencies, we also observe larger efficiencies for
-GMRES as compared to conjugate gradient. This occurs due to the fact
-that conjugate gradient has a static set of operations performed at
-every iteration while GMRES operates on a continually growing
-subspace. In addition, the orthogonalization operations required to
-generate that subspace consume much more time than the basic conjugate
-gradient operations, causing a larger serial runtime and subsequently
-higher efficiencies. For the same reasons, we also observe higher
-efficiencies for the MCSA implementation due to the serial performance
-of the implementation.
-
-Beyond around 5,000 cores, the global problem size selected for this
-strong scaling study causes the local problem size to shrink enough
-that parallel communication costs begin to overtake arithmetic costs
-on process. This takeover causes this effective strong scaling wall
-that is observed in all methods. Using any more cores to solve this
-particular global problem is not an efficient use of those parallel
-resources. Furthermore, the fact that we see a very similar trends in
-efficiencies for all methods means that they are bound by the same
-parallel performance limitations in a strong scaling environment. This
-is important as we can expect the same qualitative scaling performance
-from MCSA if arithmetic optimization has been performed to improve the
-serial performance of the implementation. In addition, these results
-are valuable in that they show MCSA, when properly parallelized, has
-the ability to be competitive on leadership-class computing platforms
-with conventional subspace methods with good strong scaling observed
-in the pure domain decomposition case to O(10,000) cores.
-
-Next, we consider the weak scaling performance of the algorithm. In
-this case, we fix the local problem size at \sn{4}{4} DOFs and
-increase the number of processors used in the
-simulation. Figure~\ref{fig:titan_pure_weak_time} gives the wall
-time per iteration and Figure~\ref{fig:titan_weak_absolute} gives the
-weak scaling absolute efficiencies for this problem as computed by
-Eq~(\ref{eq:weak_scaling_absolute}). In general, the relative trends
-when comparing MCSA to the Krylov methods are the same with MCSA and
-GMRES scaling better than the conjugate gradient method due to the
-increased amount of arithmetic work on-process. In addition, it is
-important to note that the weak scaling of MCSA when properly
-parallelized remains competitive with the Krylov methods keeping in
-mind the artificial inflation of efficiencies.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_pure_weak_time.pdf}
-  \end{center}
-  \caption{\textbf{Pure domain decomposition weak scaling wall time
-      per iteration.} \textit{Wall time is reported per iteration
-      because the methods converged in different numbers of
-      iterations. MCLS is over an order of magnitude slower
-      arithmetically than the Krylov solvers. GMRES executes more
-      operations than conjugate gradient and is therefore slower as
-      well.}}
-  \label{fig:titan_pure_weak_time}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_absolute.pdf}
-  \end{center}
-  \caption{\textbf{Pure domain decomposition weak scaling absolute
-      efficiency.} \textit{MCLS is an over order of magnitude slower
-      arithmetically, causing the higher efficiency. GMRES executes
-      more operations than conjugate gradient, also creating a higher
-      efficiency.}}
-  \label{fig:titan_weak_absolute}
-\end{figure}
-
-As an additional means of comparison between MCSA and the Krylov
-methods, we compute the algorithmic efficiencies of the calculations
-using Eq~(\ref{eq:algorithmic_efficiency}). As shown in
-Figure~\ref{fig:titan_weak_algorithmic}, the algorithm efficiency of
-the Krylov algorithm improves as a function of the number of cores in
-the problem due to the fact that the number of iterations required to
-converge decreased from 20 in the base case to 19 and then to 18 for
-the final set of weak scaling computations. When this reduction in
-iterations is used to account for the observed runtimes and computed
-absolute efficiencies for weak scaling we end up with the
-implementation efficiencies computed using
-Eq~(\ref{eq:implementation_efficiency}) given in
-Figure~\ref{fig:titan_weak_implementation}. Because MCSA maintains the
-same iterative performance as a function of global problem size in the
-weak scaling study, the implementation efficiencies compare more
-favorably to the Krylov methods. Using the implementation efficiency,
-we are effectively measuring the parallel performance of a single
-iteration for each method by looking at these implementation
-efficiencies. As the global problem size increases, MCSA is less prone
-to a drop in parallel efficiency. If optimization has been performed
-and runtimes potentially approach those of the Krylov methods, this
-comparison may not be as favorable. Overall, weak scaling performance
-of the parallel MCSA algorithm is excellent for the pure domain
-decomposition case with absolute parallel efficiencies of over 80\%
-observed at over 32,000 cores.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_alg_eff.pdf}
-  \end{center}
-  \caption{\textbf{Pure domain decomposition weak scaling algorithmic
-      efficiency.} \textit{As the global problem size was increased in
-      the weak scaling study, both conjugate gradient and GMRES
-      converged in fewer iterations, causing the rise in algorithmic
-      efficiency.}}
-  \label{fig:titan_weak_algorithmic}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_implementation.pdf}
-  \end{center}
-  \caption{\textbf{Pure domain decomposition weak scaling
-      implementation efficiency.} \textit{Considering the scalability
-      of a single iteration of the implementation, the inflated
-      efficiencies of the MCLS MCSA implementation become more
-      obvious. Implementation optimization will likely reduce the MCSA
-      values to a level more comparable to the Belos values.}}
-  \label{fig:titan_weak_implementation}
-\end{figure}
-
-In addition to comparing MCSA scaling to conventional methods, we also
-motivate avoiding any type of master/slave algorithms at these high
-levels of concurrency by comparing the binary communication tree
-implementation used in the previous results with a parallel MCSA
-implementation that uses the master/slave scheme to determine the end
-of a history transport stage as shown in
-Figure~\ref{fig:master_comm_tree}
-instead. Figure~\ref{fig:titan_strong_bvsm} provides this comparison
-using the strong scaling study while Figure~\ref{fig:titan_weak_bvsm}
-gives the results of the comparison for the weak scaling study. In
-both cases, the deficiencies caused by using the master/slave scheme
-for transport completion are obvious. At around 1,000 cores, both the
-strong and weak scaling performance seriously degrades while the
-binary tree pattern maintains excellent performance. At this number of
-cores, any extra work given to the master process, even if it amounts
-to summing a handful of integers and checking for a few extra messages
-as in this algorithm, grows enough to create serious load imbalance
-and reduce performance.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_bvsm.pdf}
-  \end{center}
-  \caption{\textbf{Binary tree vs. master/slave communication scheme
-      strong scaling absolute efficiency.} \textit{The master/slave
-      scheme will not scale beyond $O(1,000)$ cores due to load
-      imbalance.}}
-  \label{fig:titan_strong_bvsm}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_bvsm.pdf}
-  \end{center}
-  \caption{\textbf{Binary tree vs. master/slave communication scheme
-      weak scaling absolute efficiency.}  \textit{The master/slave
-      scheme will not scale beyond $O(1,000)$ cores due to load
-      imbalance.}}
-  \label{fig:titan_weak_bvsm}
-\end{figure}
-
-\clearpage
-
-\subsection{Overlapping Domain Decomposition}
-\label{subsec:overlapping_domain_decomp}
-
-In the next set of scaling studies, we will explore how adding overlap
-to the subdomains in the problem affects runtimes and parallel
-performance as compared to the case with pure domain decomposition. To
-do this, we utilize the set of analytic relations developed in
-\S~\ref{sec:analytic_framework} to design these numerical
-experiments. For the neutron diffusion problem used in these scaling
-studies, the linear operator had a spectral radius of 0.787 and a
-weight cutoff of \sn{1}{-2} was used with the Neumann-Ulam
-solver. Using Eq~(\ref{eq:analytic_k}) gives an expected random walk
-length of 19.22 transitions for any given history in the
-problem. Using the mean-chord approximation given by
-Eq~(\ref{eq:mean_chord_domain_leakage}), we then expect approximately
-5.3\% of the histories born in a particular domain to leak out after
-the first transport stage. By adding overlap, we aim to reduce this
-fraction of communicated histories and therefore boost the parallel
-performance of the algorithm.
-
-To determine the amount of overlap to test for the scaling studies, we
-use a variation of Eq~(\ref{eq:step_k_length}) to determine how many
-discrete states a history will move on average from its birth site. We
-modify Eq~(\ref{eq:step_k_length}) to only account for discrete
-movements:
-\begin{equation}
-  \sqrt{\bar{n^2_k}} = n_s \sqrt{k}\:,
-  \label{eq:discrete_distance}
-\end{equation}
-where $\sqrt{\bar{n^2_k}}$ is now the root-mean-squared number of
-discrete states a history will move from its birth site. For our
-diffusion problem used in these scaling studies, we compute a value of
-2.63 states per history for any of the dimensions of the system. This
-means that we can expect to end up on average 2.63 states in both the
-$i$ and $j$ dimensions away from the starting state. Therefore, we
-will use the overlap generation algorithm outlined in
-\S~\ref{subsec:msod_generation} such that we parametrically grow the
-local domain using 1, 2, 3, and 5 levels to measure how scalability is
-affected when this root-mean-squared number of states is
-considered. For example, using 3 levels of overlap should keep over
-50\% of the histories that would be communicated in the pure domain
-decomposition case on-process as this is greater than the
-root-mean-squared number of states we would expect a history to
-travel.
-
-For the performance metrics, we may use the same weak and strong
-scaling efficiency metrics that we used for the pure domain
-decomposition case. As we are comparing to the case without overlap
-and are looking to boost performance relative to the case without
-overlap, for both weak and strong scaling calculations we will use the
-16-core case without overlap as the reference computation for all
-efficiency measures reported.
-
-The absolute efficiency results from the strong scaling exercise are
-given in Figure~\ref{fig:titan_strong_overlap}. Aside from the
-boosting of efficiencies at 2,048 cores, little evidence of strong
-scaling improvement is evident for all levels of overlap tested. To
-further explore any potential benefits of adding overlap into the
-system, Figure~\ref{fig:titan_strong_overlap_diff} gives the
-difference in the efficiencies between all overlap cases and the
-reference case without overlap at each core count. At smaller numbers
-of cores, there is no observed benefit to using overlap to improve
-strong scaling. The improvements noted at 2,048 cores are also present
-here and account for the best efficiency improvements for all core
-counts tested. at 16,384 cores, using 2 levels of overlap did give a
-6.4\% boost in efficiency while at 1,024 cores this same amount of
-overlap actually reduced the efficiency. Beyond 2 levels of overlap,
-efficiency levels begin to fall with 5 levels of overlap performing
-the worst for all cases. From this, we don't expect adding any more
-overlap will improve the scalability of the algorithm and the best
-strong scaling performance was observed when the number of overlap
-levels is approximately equal to the root-mean-squared number of
-states we expect a history to move from its starting point as computed
-by Eq~(\ref{eq:discrete_distance}). We will defer an explanation of
-the drop in efficiencies for levels of overlap beyond this value until
-we have analyzed the weak scaling data.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_overlap.pdf}
-  \end{center}
-  \caption{\textbf{Strong scaling absolute efficiency with varying
-      levels of overlap.} \textit{Values of 0, 1, 2, 3, and 5 levels
-      of overlap were used as the problem had a root-mean-squared
-      number of states of 2.63 for each history.}}
-  \label{fig:titan_strong_overlap}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_overlap_diff.pdf}
-  \end{center}
-  \caption{\textbf{Strong scaling absolute efficiency difference with
-      respect to the 0 overlap base case.} \textit{Values of 1, 2, 3,
-      and 5 levels of overlap are compared here with the absolute
-      difference of the absolute efficiencies presented.}}
-  \label{fig:titan_strong_overlap_diff}
-\end{figure}
-
-For weak scaling, the same study as in the pure domain decomposition
-case was performed again with 1, 2, 3, and 5 levels of overlap. The
-absolute parallel efficiencies from these calculations are given in
-Figure~\ref{fig:titan_weak_overlap} and the differences in
-efficiencies from the calculations without overlap. The results here
-are much the same in that there is not a dramatic improvement in
-scalability relative to the case without overlap. Opposite the strong
-scaling calculation, overlap enhances weak scaling efficiencies at
-lower core counts and degrades the efficiencies at higher core
-counts. In addition, 1 level of overlap provided the best enhancement
-of scalability when overlap was beneficial.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_overlap.pdf}
-  \end{center}
-  \caption{\textbf{Weak scaling absolute efficiency with varying
-      levels of overlap.} \textit{Values of 0, 1, 2, 3, and 5 levels
-      of overlap were used as the problem had a root-mean-squared
-      number of states of 2.63 for each history.}}
-  \label{fig:titan_weak_overlap}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_overlap_diff.pdf}
-  \end{center}
-  \caption{\textbf{Weak scaling absolute efficiency difference with
-      respect to the 0 overlap base case.} \textit{Values of 1, 2, 3,
-      and 5 levels of overlap are compared here with the absolute
-      difference of the absolute efficiencies presented.}}
-  \label{fig:titan_weak_overlap_diff}
-\end{figure}
-
-For both the strong and weak scaling studies, the parallel performance
-of the algorithm was either marginally improved or not improved at
-all. We would always expect an improvement in performance for using
-overlap to prevent the transport of histories from domain to domain
-during execution of the Neumann-Ulam algorithm. The reduction in
-transport means a reduction in parallel communication and therefore
-should never degrade the performance of the algorithm. It is true that
-doing so in fact enhances Monte Carlo performance. However, the
-Neumann-Ulam algorithm is not representative of the entire MCSA
-algorithm. When overlap is used, an overlapping tally vector is
-constructed in parallel such that multiple cores may own multiple
-states in the system. When this occurs, the tally data associated with
-these shared states must be communicated after the Neumann-Ulam solve
-as outlined in \S~\ref{subsec:msod_algorithm}. This reduction
-operation then builds the full MCSA correction vector in the
-decomposition of the solution vector.
-
-Therefore, by building overlap we are in fact reducing parallel
-communication during the Neumann-Ulam solve but we are simply
-deferring that communication until after Monte Carlo transport when
-the overlapping tally vector must be reduced. When overlap becomes
-large enough (i.e. larger than the root-mean-squared number of
-discrete states a history will travel), then the savings in
-communication during the Neumann-Ulam solve start to become smaller
-than the communication created in the overlapping tally vector
-reduction. We see this in both the strong and weak scaling cases at
-larger levels of overlap. For the strong scaling case, we see better
-performance improvement at higher core counts because the total number
-of overlapping states on each core is smaller from the local problem
-size shrinking. For weak scaling, we see better performance
-improvement at lower numbers of cores because although the local
-overlap size is fixed, the global amount of overlap is growing with
-core count and therefore reducing the performance of the tally
-reduction operation.
-
-Based on this data, a few levels of overlap within the expected
-distance of travel for a history may offer some marginal improvement
-in scalability, however, it is not as effective as one might think
-given its improvement to reactor physics calculations as observed in
-the literature. This is largely due to the density of states in the
-system that creates the described performance bottlenecks in the
-overlapping tally vector reduction operation. In a Monte Carlo
-particle simulation, these overlapping tally reduction operations are
-not expected to be performed at the speed they are required in
-MCSA. For example, in a criticality calculation with overlapping
-domains, we would expect to perform as many parallel tally vector
-reduction in a particle transport simulation as there are particle
-stages required for convergence of the eigenvalue. However, these
-particle stages will likely be much more costly in terms of wall time
-relative to a Neumann-Ulam solve in an MCSA iteration due to the more
-extensive particle physics and logic structure that must be
-evaluated. Therefore, we can expect this overlapping tally reduction
-to be less significant to scalability in those operations. We do not
-have as complicated a logic structure in a history random walk for the
-Neumann-Ulam method and therefore overlap can inhibit scalability if
-too large.
-
-\clearpage
-
-\subsection{Multiple Set Domain Decomposition}
-\label{subsec:ms_decomposition}
-
-We next look to apply replication to the problem in the form of
-multiple sets. As the previous analysis showed overlap to be of little
-assistance in improving scalability, we only apply multiple sets to
-the problem with pure domain decomposition within the sets for these
-scaling studies. Before performing parallel scaling studies using
-multiple sets, several aspects of the replicated parallel problem must
-be considered in order to effectively craft the studies and interpret
-their results. First is the question of how many stochastic histories
-to use to generate the correction vector in the MCSA iteration. For
-this work we consider two choices; the split case and the fully
-replicated case. As an example of these cases, consider a single set
-problem where 20,000 histories are used in the Monte Carlo solve. If
-we choose the split case, then solving the same problem with 2 sets
-would give 10,000 histories in each set and 4 sets would give 5,000
-histories in each set. Splitting the histories in this way provides
-the same global number of histories used to compute the correction in
-the single set problem. We expect to reduce the time to solution by
-doing this as each individual set solve will occur faster in an almost
-linear fashion with the reduced number of histories. For the
-replicated history case, each additional set in the problem will use
-just as many histories as the single set case. Using 2 sets with fully
-replicated histories for our example problem then has each set
-computing 20,000 histories with 40,000 global histories in the
-computation. For the 4 set case, the number of global histories is
-increased to 80,000. We also expect this type of replication to reduce
-the time to solution as the extra global histories should reduce the
-stochastic uncertainty in the correction vector and reduce the overall
-number of MCSA iterations required to converge.
-
-\subsubsection{Strong Scaling with Multiple Sets}
-\label{subsubsec:ms_strong}
-
-For the strong scaling study, we apply the additional sets as
-follows. Using the same problems as the pure domain decomposition
-scaling study, the multiple sets are used to replicate those problems
-and boost the core count. For example, for the 4 set case, the data
-point at 64 cores has the same $N/P$ ratio in one set as the single
-set problem with 16 cores. This is due to the fact that $N$ is fixed
-for the strong scaling study and 4 replications gives a set size of 16
-cores. For the 2 set case, the 32 core data point is solving the same
-$N/P$ problem in a set as the 16 core base case.
-
-In addition to considering the replicated domain, we must also
-consider how to modify the efficiencies we will compute. For the
-strong scaling case, recall that the objective is to decrease the time
-to solution for a given global problem size $N$ by increasing
-$P$. Therefore, we may use Eqs~(\ref{eq:strong_scaling_absolute_ref}),
-(\ref{eq:algorithmic_efficiency}), and
-(\ref{eq:implementation_efficiency}) to compute the various strong
-scaling efficiencies, but we must modify the reference computation in
-each of these formulas to reflect our objective. For all efficiency
-computations, the 16 core single set computation will be used as the
-reference case. By adding more cores to the problem, either by
-replicating with multiple sets, or decreasing the local problem size,
-any computational resources added to the problem should aim to result
-in speedup of this reference case. If this speedup for one technique
-of adding cores is less than another, then that technique is deemed a
-less efficient use of those computational resources.
-
-For the multiple sets problem, the single set strong scaling study
-from the pure domain decomposition base case was repeated for multiple
-sets with both splitting of histories across sets and full replication
-of histories across sets using both 2 and 4
-sets. Figure~\ref{fig:titan_strong_ms_time} gives the actual wall
-times for these computations. There are two important features to note
-in the wall time data. First, splitting the histories among sets
-instead of replicating them is faster due to the fact that cutting the
-number of histories in half reduces the time of an iteration by a
-factor that is nearly linear while increasing linearly the number of
-histories through replication does not decrease the iterations
-required to converge by a linear factor. Second, until the strong
-scaling wall is hit starting at around 10,000 cores, the single set
-case actually performs the fastest with only marginal improvement in
-runtimes at higher numbers of cores with more sets. At these larger
-numbers of cores, however, the improvements are so modest that one is
-better off running the same problem using a single set at lower core
-counts to conserve computational resources while maintaining a good
-time to solution.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_ms_time.pdf}
-  \end{center}
-  \caption{\textbf{Wall time in seconds to solution for each case for
-      the strong scaling study with multiple sets.} \textit{Until the
-      strong scaling wall is reached at O(10,000) cores, the single
-      set case performs the best. For the multiple set cases, it is
-      faster to split histories among sets and maintain iterative
-      performance rather than replicate histories to improve iterative
-      performance.}}
-  \label{fig:titan_strong_ms_time}
-\end{figure}
-
-From this timing data we then compute the absolute efficiencies for
-strong scaling as given in Figure~\ref{fig:titan_strong_ms_eff} using
-the 16-core single set calculation as the reference case. As observed
-in the timing data, because the single set case had the fastest run
-times up to the strong scaling wall it also has the best observed
-efficiencies. For the 2 set case with history splitting, although the
-time to perform an iteration was nearly cut in half, the extra
-parallel overhead associated with reducing the tally across the blocks
-to combine the set results increases run times and reduces
-efficiencies. This reduction across blocks is even more obvious for
-the 4 set case. For the replicated cases, we do observe a flat region
-of parallel efficiencies at larger core counts (up to around 32,000
-for the 4 set case with history replication) showing improved scaling
-within context of not considering the single set
-calculation. Unfortunately, the additional cores do not reduce the
-runtime of the global problem as efficiently as expected. Ultimately,
-all cases effectively hit the strong scaling wall at the same time at
-around O(10,000) cores.
-
-We expect this if we consider the fact that adding sets to the problem
-is in itself effectively a form of strong scaling. For any given
-problem of fixed size $N$, adding extra sets to the computation
-decreases the global ratio of $N/P$ as $P$ will grow with the addition
-of sets (although $N/P$ is constant within a set). For the split
-history case, the splitting further exacerbates this situation by
-further decreasing the amount of histories that will be computed in
-the local domain, giving a larger ratio of communication to work in
-the Monte Carlo solve. As with with single set results, we also
-observe some efficiencies greater than unity for the multiple set
-cases also due to improved cache utilization.
-
-Although not as dramatic as one might originally expect from
-introducing replication, there are some modest gains in efficiencies
-at high core counts by using multiple sets. Of interest here is where
-the multiple set cases perform better than the single set case after
-the single set case has hit the strong scaling wall. In
-Figure~\ref{fig:titan_strong_ms_eff}, at 16,384 cores the single set
-case operates at 38\% efficiency while at the same number of cores,
-the 2 set case with history replication operates at 58\%
-efficiency. This is reasonable improvement in the use of those
-computational resources, gained by replicating the problem. Beyond
-16,384 cores, there is not a compelling reason to use more
-computational resources for the this problem.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_ms_eff.pdf}
-  \end{center}
-  \caption{\textbf{Multiple set absolute parallel efficiency.}
-    \textit{Computed relative to the 16-core 1-set base case. Using a
-      single set is the best means by which to improve strong scaling
-      up to the strong scaling wall. At higher core counts past where
-      the single set case hits the strong scaling wall adding sets can
-      provide some improvements in efficiencies.}}
-  \label{fig:titan_strong_ms_eff}
-\end{figure}
-
-Also of interest here is the effect of splitting histories among sets
-versus replicating them on the iterative performance of the
-method. Table~\ref{tab:ms_strong_alg_eff} gives the iterations
-required to converge and computed algorithmic efficiencies for each of
-the cases presented in the strong scaling data. As expected, splitting
-histories maintains a fixed global number of histories and therefore
-maintained the number of iterations required to converge relative to
-the single set base case. When histories were replicated, adding sets
-decreased the number of iterations required to converge in a nonlinear
-fashion. Therefore, increasing the number of histories per MCSA
-iteration improves algorithmic efficiency. Using these algorithmic
-efficiencies, the implementation efficiencies for the multiple set
-cases were computed and are given in
-Figure~\ref{fig:titan_strong_ms_impeff}. From these results, we see
-that the replicated history cases are even less efficient than in the
-absolute case with respect to the performance of a single
-iteration. Not only do we have the additional parallel overhead of
-doing the block level reduction, but we also have additional histories
-that increase the time to solution for a single iteration.
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lcc}\hline\hline
-      \multicolumn{1}{l}{Case}& 
-      \multicolumn{1}{c}{Iterations}&
-      \multicolumn{1}{c}{$\eta_{alg}$} \\\hline
-      1 & 22 & 1.0 \\
-      2-split & 22 & 1.0 \\
-      2-replicated & 16 & 1.375 \\
-      4-split & 22 & 1.0 \\
-      4-replicated & 13 & 1.692 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Strong scaling algorithmic efficiency using
-      multiple sets.} \textit{Replicating histories improves
-      algorithmic efficiency by converging in fewer iterations.}}
-  \label{tab:ms_strong_alg_eff}
-\end{table}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=5in]{chapters/parallel_mc/titan_strong_ms_impeff.pdf}
-  \end{center}
-  \caption{\textbf{Multiple set strong scaling implementation
-      efficiency.} \textit{Computed relative to the 16-core 1-set base
-      case. Considering iterations required to converge, the
-      performance of the split history cases at much better than the
-      cases with history replication.}}
-  \label{fig:titan_strong_ms_impeff}
-\end{figure}
-
-\clearpage
-
-\subsubsection{Weak Scaling with Multiple Sets}
-\label{subsubsec:ms_weak}
-
-Next, we continue to a weak scaling study of the multiple sets
-problem. In the same way as the strong scaling study, we will modify
-the pure domain decomposition study to account for the additional
-replication. For each calculation in the single set study, that
-problem will be replicated either 2 or 4 times with both split and
-replicated histories to assess the effects on scalability and
-runtime. Figure~\ref{fig:titan_weak_ms_time} gives the wall times from
-the multiple sets weak scaling study. Of primary importance here is
-the fact that adding sets to the problem with both split and
-replicated histories actually decreases the time to solution. This was
-not observed in the strong scaling case until much larger core counts
-as the multiple set computations hit the strong scaling wall after the
-single set computation.
-
-This result is significant in that it shows a solution technique in
-which the linear problem can actually be replicated and enhance time
-to solution, something that cannot be achieved with conventional
-methods. At larger core counts, however, this reduction in compute
-time is not as large as at lower core counts due to the fact that the
-tally reduction across blocks requires more resources and those
-resources are clearly growing with core count (most obviously in the 4
-set case with split histories). In addition, at 131,072 cores for the
-largest 4 set computation, load balance is likely becoming an issue
-due to the linear system and subsequent parallel matrix and vector
-operations only occurring on a subset of the entire processor space of
-the problem while the Monte Carlo calculation is occurring on all
-processors.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_ms_time.pdf}
-  \end{center}
-  \caption{\textbf{Wall time in seconds to solution for each case for
-      the weak scaling study with multiple sets.}}
-  \label{fig:titan_weak_ms_time}
-\end{figure}
-
-For a multiple set analysis of the weak scaling timing results, we
-again have to consider that adding sets to the problem is actually a
-form of strong scaling. Consider a single set problem ($S=1$) for a
-given $P$ and $N$. When that problem is now solved using 2 sets we
-have $S=2$, $N$ and $2P$. For any number of sets we then have a global
-problem size of $N$ while the number of processors is $SP$ where $P$
-is the number of processors in the single set case. Therefore, $N/P$
-is no longer fixed in the weak scaling study but rather $N/(SP)$ is
-fixed. As a result, we modify the weak scaling absolute efficiency
-relationship to account for this fact:
-\begin{equation}
-\eta_{weak}(M|N,P|Q,S) = \frac{T(M,Q)}{S \times T(N,P)}\:,
-  \label{eq:ms_weak_efficiency}
-\end{equation}
-where now the absolute efficiency is a function of the number of sets
-in the problem and again $M/Q = N/P$. As with the strong scaling
-study, we will use the 16-core single set case as the reference
-computation for efficiencies. Interestingly,
-Eq~(\ref{eq:ms_weak_efficiency}) has a very similar form to the strong
-scaling efficiency given by Eq~(\ref{eq:strong_scaling_absolute}),
-representative of the strong scaling nature of adding additional
-sets. For problems where adding sets improves time to solution, this
-new weak scaling formulation accounts for how efficiently those extra
-resources are incorporated into the problem. Without such a
-modification, the largest case run for the split 4 set case would have
-a computed weak scaling efficiency of nearly 100\% using the standard
-formula for absolute efficiency. However, using 4 times as many cores
-as the single set case increased the runtime for a fixed global
-problem size and therefore we would actually expect to compute a very
-low efficiency that shows this very poor use of resources.
-
-Figure~\ref{fig:titan_weak_ms_eff} gives the results of the weak
-scaling efficiency computations using the multiple set metric given by
-Eq~(\ref{eq:ms_weak_efficiency}). Although we do reduce the time to
-solution over the base cases for nearly all data points, the
-observation here is that the reduction in time to solution is still
-not large enough to efficiently use all resources. For example, for
-the 2 set case to be a perfect use of resources with either split or
-replicated histories, the runtime of that calculation must be half of
-that for the single set case. For the 4 set case, this run time should
-be a quarter of the single set case for perfect scaling. This was not
-the observation for the runtimes in
-Figure~\ref{fig:titan_weak_ms_time} and therefore the single set case
-is still the most efficient use of computational resources in this
-scaling study.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_ms_eff.pdf}
-  \end{center}
-  \caption{\textbf{Weak scaling absolute efficiency relative to
-      16-core 1-set base case.}}
-  \label{fig:titan_weak_ms_eff}
-\end{figure}
-
-As in the strong scaling case, splitting and replicating histories
-across sets had effectively the same results for the weak scaling
-study with the iterations to converge and computed algorithmic
-efficiencies given by Table~\ref{tab:ms_weak_alg_eff}. Again, when the
-implementation efficiencies are computed, we see in
-Figure~\ref{fig:titan_weak_ms_impeff} that splitting histories among
-sets is still a more efficient manner of introducing multiple sets
-rather than replicating histories in an attempt to reduce the number
-of iterations required to converge.  Here we also note that this is
-not a true weak scaling study as indicated by
-Eq~(\ref{eq:ms_weak_efficiency}) but rather an odd combination of both
-weak and strong scaling. Regardless of this, although multiple sets
-are not the most efficient use of resources in this analysis, they do
-enhance the time to solution through replication and do so
-significantly at core counts of O(1,000) and lower. In both the weak
-and strong scaling cases, due to the additional computational and
-communication overhead of adding sets to the system, one should never
-expect to seriously boost parallel efficiencies for MCSA using this
-technique except for certain instances past the strong scaling wall in
-a purely strong scaling environment. In a weak scaling environment,
-although no efficiency increase will be noted, one can expect an
-enhancement of time to convergence by replicating.
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lcc}\hline\hline
-      \multicolumn{1}{l}{Case}& 
-      \multicolumn{1}{c}{Iterations}&
-      \multicolumn{1}{c}{$\eta_{alg}$} \\\hline
-      1 & 22 & 1.0 \\
-      2-split & 22 & 1.0 \\
-      2-replicated & 16 & 1.375 \\
-      4-split & 22 & 1.0 \\
-      4-replicated & 13 & 1.692 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Weak scaling algorithmic efficiency using multiple
-      sets.} \textit{Replicating histories improves algorithmic
-      efficiency by converging in fewer iterations.}}
-  \label{tab:ms_weak_alg_eff}
-\end{table}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_ms_impeff.pdf}
-  \end{center}
-  \caption{\textbf{Implementation parallel efficiency relative to 16-core
-      1-set base case.}}
-  \label{fig:titan_weak_ms_impeff}
-\end{figure}
-
-\clearpage
-
-\subsection{Subdomain Neumann-Ulam}
-\label{subsec:full_clip}
-
-Based on the analysis of the pure domain decomposed MCSA algorithm and
-subsequently the multiple sets and overlapping domain implementation,
-one may ask then ask if there is any need to communicate histories
-from domain-to-domain at all. Overlap was not as successful because
-because the additional communication required to reduce the
-overlapping tally vector eventually becomes larger than the
-communication savings during Monte Carlo transport. In addition, for
-local domains that are large enough to get good performance
-(e.g. those used in the weak scaling studies), then our analytic
-relationships for domain leakage, and in particular
-Eq~(\ref{eq:mean_chord_domain_leakage}), tell us that only about 5\%
-of the histories in this diffusion problem will leave the domain in
-the first place.  If that is true, then perhaps without any overlap or
-communication, just doing local Monte Carlo transport such that
-histories are terminated upon reaching the domain boundary may be
-satisfactory for accelerating the MCSA iteration. In this case,
-histories near the boundaries of the domain may be truncated before
-reaching the weight cutoff if they reach the domain boundary
-first. This truncation has the potential to increase the stochastic
-uncertainty of the Neumann-Ulam result and therefore increase the
-number of MCSA iterations required to converge. In addition, the
-scalability of the method will then be purely limited by the parallel
-matrix and vector operations required to implement the remainder of
-the algorithm on the subset of processors containing the linear
-problem.
-
-Considering using MCSA in this way such that the Monte Carlo solve
-occurs only in the subdomains casts the method as more of a true
-domain decomposition as outlined by Chapter 14 of Saad's text on
-iterative methods \cite{saad_iterative_2003}. Compared to traditional
-domain decomposition methods we see many similarities including the
-potential to overlap the subdomains to improve the results. By
-building the boundary data as outlined in
-\S~\ref{subsec:domain_generation}, we are effectively gathering one
-level of the adjacent domains, albeit in this case to satisfy the data
-requirements of the Monte Carlo estimators. Furthermore, the fact that
-we are using residual Monte Carlo to compute the correction vector
-casts MCSA as a stochastic realization of an additive Schwarz method
-with a fixed point iteration used as a relaxation step\footnote{See
-  section 14.3.3 in \cite{saad_iterative_2003} for a full definition
-  of the additive Schwarz method.}. Both the fixed point iteration and
-the computation of the residual propagate the solution from individual
-subdomains to the others through the matrix-vector multiply operation
-after all subdomain Monte Carlo solves have been completed. Perhaps an
-even more important consequence of viewing MCSA as a stochastic
-additive Schwarz procedure is that it has the potential to more
-readily fit as an accelerator into other linear solver techniques
-including GMRES or potentially as a preconditioner.
-
-Using the Neumann-Ulam solver to compute the MCSA correction only in
-the subdomains, the strong and weak scaling studies were repeated to
-assess the performance gains made by eliminating communication in the
-Monte Carlo solver. For each scaling study, the 16-core base case with
-Monte Carlo communication was used as the reference point for
-efficiency computations in order to measure how eliminating
-communication boosts
-efficiencies. Figure~\ref{fig:titan_strong_subdomain} gives the
-results of the strong scaling study and
-Figure~\ref{fig:titan_weak_subdomain} gives the results of the the
-weak scaling study in terms of absolute efficiencies. In both cases,
-the benefits to parallel performance of using MCSA as a subdomain
-solver are obvious. In the strong scaling case we make the important
-observation that we hit the strong scaling wall at the same time with
-and without the additional Monte Carlo communication. Although the
-Monte Carlo communication did contribute to the efficiency decrease,
-it is clear that the remaining matrix and vector operations are the
-primary impediment to scalability. Furthermore, we see an even greater
-rise above unity in efficiencies because there is less communication
-overhead hiding the better cache utilization as the local problem size
-shrinks. At 16,384 cores, performing the subdomain solves boost the
-absolute efficiency from 38\% to 67\% relative to the reference case.
-
-In the weak scaling study, we do not have the scaling wall to contend
-with and eliminating Monte Carlo communication through subdomain
-solves provides an effectively linear boost in the weak scaling
-absolute efficiency. We still see a small reduction in weak scaling
-efficiency as a function of core count due to the communication
-overhead for the parallel matrix and vector operations. Here we
-observe super unitary efficiencies for the subdomain case as there was
-a substantial improvement in runtimes relative to the 16-core base
-case with Monte Carlo communication.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_subdomain.pdf}
-  \end{center}
-  \caption{\textbf{Strong scaling absolute efficiency for pure domain
-      decomposition.} \textit{The subdomain Monte Carlo solver
-      improves the parallel efficiency and therefore time to solution
-      by eliminating all parallel communication during the Monte Carlo
-      solve.}}
-  \label{fig:titan_strong_subdomain}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_subdomain.pdf}
-  \end{center}
-  \caption{\textbf{Weak scaling absolute efficiency for pure domain
-      decomposition.} \textit{The subdomain Monte Carlo solver
-      improves the parallel efficiency and therefore time to solution
-      by eliminating all parallel communication during the Monte Carlo
-      solve.}}
-  \label{fig:titan_weak_subdomain}
-\end{figure}
-
-Of additional concern for the subdomain method is the potential
-decrease in iterative performance due to the fact that histories near
-the boundary will potentially have their random walk truncated, thus
-decreasing the information available to compute the correction
-vector. For both the weak and strong scaling study, all calculations
-with split histories converged in 22 or in only a few cases 21
-iterations for both the subdomain solve and the solve with
-communication giving an algorithmic efficiency of 1.048 for all
-cases. This small improvement in iteration count was not observed to
-be a function of problem size, number of processors, or whether the
-subdomain method was used. For the replicated history cases, the
-algorithmic efficiency improvement was not as good as when
-communication was allowed for the Monte Carlo algorithm. For these
-calculations, the number of iterations to converge was reduced to 18
-or 19 in both the strong and weak scaling studies with lower core
-counts converging in fewer iterations.
-
-Therefore, there is no reason to expect overlap to improve the
-iterative or scaling performance of the method as iterative
-performance is maintained without it for this particular
-problem. Overlap is not as effective because histories relinquish most
-of their useful information within the first few steps (which is why
-we can use such a large weight cutoff and there is an insensitivity to
-weight cutoff as shown in
-Figure~\ref{fig:estimator_wc_iters}). Keeping them around longer with
-overlap to do extra work in the local domain will not improve
-iterative performance and thus increase compute times.
-
-As a final analysis of MCSA using subdomain Monte Carlo, the multiple
-sets scaling studies are repeated to determine if any performance
-benefits are made from replication by eliminating Monte Carlo
-communication. Figure~\ref{fig:titan_strong_subdomain_ms_time} gives
-the runtimes for the strong scaling calculation and
-Figure~\ref{fig:titan_strong_subdomain_ms} the computed absolute
-efficiencies. For the weak scaling studies,
-Figure~\ref{fig:titan_weak_subdomain_ms_time} gives the wall times and
-Figure~\ref{fig:titan_weak_subdomain_ms} the efficiencies. For both
-scaling studies, the results are effectively the same as for the case
-with Monte Carlo communication. This is again due to the fact that
-adding sets to the problem is not only an exercise in strong scaling
-but it also adds additional parallel overhead in the form of the tally
-vector reduction across blocks.
-
-From a parallel performance perspective, MCSA may then be best used in
-a mode where the Monte Carlo is performed over only the subdomains
-with no replication or overlap added to the system. In this mode,
-iterative performance is maintained relative to the solutions with
-communication even for more ill-conditioned problems with larger
-spectral radii. Furthermore, no overlap was required to maintain
-iterative performance be letting histories near the boundary take a
-few more steps in their random walks. If in the future there are
-ill-conditioned problems where using subdomain Monte Carlo reduces the
-algorithmic efficiency of MCSA due to the information lost in the
-correction vector, overlap would be a viable mechanism to potentially
-maintain that iterative performance, keeping in mind the potential
-impacts on scalability that were observed in this work. However,
-although eliminating all Monte Carlo communication was observed to be
-the best mode of use for MCSA, the parallel Monte Carlo methods
-developed for this work still provide value for cases where only a
-Monte Carlo solution is desired. In those cases, full transport of the
-Monte Carlo histories are required in order to generate the correct
-solution.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_subdomain_ms_time.pdf}
-  \end{center}
-  \caption{\textbf{Strong scaling total wall time with multiple sets
-      and subdomain Monte Carlo.} \textit{Adding multiple sets has in
-      general a negative effect on the wall time as was observed for
-      the case with full Monte Carlo transport.}}
-  \label{fig:titan_strong_subdomain_ms_time}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_subdomain_ms.pdf}
-  \end{center}
-  \caption{\textbf{Strong scaling absolute efficiency for with
-      multiple sets and subdomain Monte Carlo.} \textit{Increased wall
-      times from adding parallel reduction operations for multiple
-      sets reduces parallel scalability at lower core counts but gives
-      a higher efficiency at the strong scaling wall.}}
-  \label{fig:titan_strong_subdomain_ms}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_subdomain_ms_time.pdf}
-  \end{center}
-  \caption{\textbf{Weak scaling total wall time with multiple sets and
-      subdomain Monte Carlo.} \textit{The addition of the set tally
-      reduction is significant with cost of the reduction observed as
-      core count grows.}}
-  \label{fig:titan_weak_subdomain_ms_time}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_subdomain_ms.pdf}
-  \end{center}
-  \caption{\textbf{Weak scaling absolute efficiency for with multiple
-      sets and subdomain Monte Carlo.} \textit{Weak scaling
-      efficiencies are not improved at large core counts when multiple
-      sets are added.}}
-  \label{fig:titan_weak_subdomain_ms}
-\end{figure}
-
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{Summary\ }
-\label{sec:parallel_summary}
-
-In this chapter we have developed a domain-decomposed parallel
-algorithm for Monte Carlo Synthetic Acceleration and explored the
-qualities of the algorithm through parallel scaling studies. The
-following are the significant observations and findings.
-
-\begin{itemize}
-\item MCSA can be effectively parallelized using domain decomposition
-\item The leakage fraction of stochastic histories from a parallel
-  subdomain can be quantified analytically using algebraic quantities
-\item The multiple-set overlapping-domain (MSOD) parallel algorithm
-  for particle transport has been adapted to parallelize MCSA
-\item MCSA parallelized with MSOD was verified to produce the same
-  results for solutions to the neutron diffusion equation as
-  production Krylov methods
-\item Strong scaling studies were performed with up to 65,536 cores
-  and weak scaling studies were performed with up to 131,072 cores on
-  the Titan Cray XK7 leadership class machine at Oak Ridge National
-  Laboratory
-\item MCSA scales favorably compared to production Krylov methods for
-  both strong and weak scaling cases
-\item Overlap in small quantities determined analytically by random
-  walk length can provide parallel efficiency boosts of a few percent
-  in strong scaling cases but is ineffective in weak scaling cases
-\item Multiple sets will never boost parallel efficiencies in the weak
-  scaling case due to additional communication overhead from added
-  reduction operations but will boost efficiencies beyond the strong
-  scaling wall
-\item Multiple sets offers a means to reduce time to solution by
-  solving multiple copies of the original problem and combining the
-  solutions using superposition
-\item MCSA is most efficiently used in parallel as a stochastic
-  realization of an additive Schwarz method
-\item With future optimization, large scale reactor analyses in which
-  the transport operator is fully formed (e.g. the $SP_N$ equations)
-  could experience improved time to solution on leadership-class
-  machines with MCSA given the improved scaling performance
-\end{itemize}
+From these difference results, the mean chord approximation is shown
+to have a lower difference for ill-conditioned systems as compared to
+the Wigner approximation while the Wigner approximation produces less
+difference for more well-conditioned systems. We also note that for
+the optically thick domains, the difference is likely corresponded to
+that observed in Figure~\ref{fig:measured_length} for the $k$
+parameter while the large relative difference in $k$ for optically
+thin domains does not affect the approximation significantly. In
+general, the mean chord approximation is a better choice to estimate
+the leakage fraction in a domain from the adjoint Neumann-Ulam method
+and except for a single data point with $n_i=50$, the mean chord
+approximation yielded leakage fractions within an absoulute difference
+of 5\% of the measured results. As the domain becomes more optically
+thick (with both increasing $n_i$ and decreasing $\rho(\ve{H})$), the
+approximations are more accurate.

### 2013-08-23 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
moving spectral analysis into an appendix

@@
 
-In this chapter we briefly review particle transport methods for
-domain decomposed Monte Carlo. Considering a domain decomposed
-Neumann-Ulam method, we derive an analytic framework based on
-algebraic quantities from which to estimate aspects of its performance
-when applied to neutron transport systems. We then devise a parallel
-algorithm for the Neumann-Ulam method based on the multiple-set
-overlapping-domain decomposition algorithm and a parallel algorithm
-for the MCSA iteration that leverages the parallel Monte Carlo
-algorithm and general parallel matrix-vector operations. Using
-parallel solutions for the neutron diffusion problem, we verify the
-correctness of the parallel algorithm and implementation by comparing
-the numerical results against production Krylov methods. With the
-verified implementation of the algorithm, we perform parallel scaling
-studies to test its performance on a leadership class machine and
-compare this performance to the production Krylov methods.
+To date, parallel Neumann-Ulam methods have been limited to full
+domain replication with parallelism exploited through individual
+histories \cite{alexandrov_efficient_1998} and in this chapter we will
+exploit particle transport algorithms to alleviate this. To accomplish
+this, we recognize from the literature that stochastic histories must
+be transported from domain to domain as the simulation progresses and
+they transition to states that are not in the local domain. In this
+chapter we briefly review particle transport methods for domain
+decomposed Monte Carlo. We then devise a parallel algorithm for the
+Neumann-Ulam method based on the multiple-set overlapping-domain
+decomposition algorithm and a parallel algorithm for the MCSA
+iteration that leverages the parallel Monte Carlo algorithm and
+general parallel matrix-vector operations. Using parallel solutions
+for the neutron diffusion problem, we verify the correctness of the
+parallel algorithm and implementation by comparing the numerical
+results against production Krylov methods. With the verified
+implementation of the algorithm, we perform parallel scaling studies
+to test its performance on a leadership class machine and compare this
+performance to the production Krylov methods.
 
@@
 process in the computation. In Figure~\ref{fig:ddmc_example}, the
-tracks of 3 particles that are born in the center subdomain are
-shown. In this example, particle A is first transported from the
-center domain to the domain directly to the left. Before the
-scattering event in the new domain may be processed, particle A must
-be communicated between between the two parallel processes that own
-those domains. For any given particle in the transport simulation,
-this communication event may hardly occur during the lifetime of the
-particle (particle B), or may occur many times (particle C) depending
-on the problem parameters and the random path in phase space taken by
-the particle. Scalable parallel algorithms for domain decomposed Monte
-Carlo are those that handle this domain-to-domain communication of
-particles effectively and balance it with the on-process computations
-for particle interactions, tallies, response calculations, and other
-simulation requirements.
+tracks of 3 particles born in the center subdomain are shown. In this
+example, particle A is first transported from the center domain to the
+domain directly to the left. Before the scattering event in the new
+domain may be processed, particle A must be communicated between
+between the two parallel processes that own those domains. For any
+given particle in the transport simulation, this communication event
+may hardly occur during the lifetime of the particle (particle B), or
+may occur many times (particle C) depending on the problem parameters
+and the random path in phase space taken by the particle. Scalable
+parallel algorithms for domain decomposed Monte Carlo are those that
+handle this domain-to-domain communication of particles effectively
+and balance it with the on-process computations for particle
+interactions, tallies, response calculations, and other simulation
+requirements.
 \begin{figure}[t!]
@@
 describes how each state in the system is coupled to the other states
-in the system through the set of equations that form the linear
-system. This coupling is then responsible for the discrete random walk
-that each history takes as shown in this example. As discrete states
-(or rows in the linear system) that do not exist in the local domain
-are reached during the random walk by stochastic histories, the
-histories must be communicated to the subdomain that does own the
-discrete state in an analogous fashion to the particles in the
-previous example.
+in the system. This coupling is then responsible for the discrete
+random walk that each history takes as shown in this example. As
+discrete states (or rows in the linear system) that do not exist in
+the local domain are reached during the random walk by stochastic
+histories, the histories must be communicated to the subdomain that
+does own the discrete state in an analogous fashion to the particles
+in the previous example.
 \begin{figure}[t!]
@@
 communication of histories from domain to domain will be the primary
-factor in the parallel scalability of the algorithm. In the next
-section, the number of histories communicated in a given problem will
-be quantified analytically for a simple model problem.
+factor in the parallel scalability of the algorithm. In
+Appendix~\ref{chap:parallel_theory}, the number of histories
+communicated in a given problem are quantified analytically for a
+simple model problem.
 
 \clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{Analytic Framework for Domain-Decomposed Monte Carlo\ }
-\label{sec:analytic_framework}
-To date, parallel Neumann-Ulam methods have been limited to full
-domain replication with parallelism exploited through individual
-histories \cite{alexandrov_efficient_1998} and in this chapter we
-will exploit particle transport algorithms to alleviate this. To
-accomplish this, we recognize from the literature that stochastic
-histories must be transported from domain to domain as the simulation
-progresses and they transition to states that are not in the local
-domain. Because we have chosen a domain decomposition strategy in a
-parallel environment, this means that communication of these histories
-must occur between compute nodes owning neighboring pieces of the
-global domain. We wish to characterize this communication not only
-because communication is in general expensive, but also because these
-nearest-neighbor communication sequences, specifically, have poor
-algorithmic strong scaling \cite{gropp_high-performance_2001} in much
-the same way as a parallel matrix-vector multiply
-operation. Therefore, we desire a framework to provide a simple,
-analytic theory based on the properties of the transport system that
-will allow for estimates of the domain decomposed behavior of the
-Neumann-Ulam method in terms of the amount of information that must be
-communicated.
-
-When solving problems where the linear operator is symmetric, a host
-of analytic techniques exist based on the eigenvalue spectrum of the
-operator that characterize their behavior in the context of
-deterministic linear solvers. Using past work, these techniques are
-adapted to the domain decomposed Neumann-Ulam method using the
-one-speed, two-dimensional neutron diffusion equation and spatial
-discretization presented in Appendix~\ref{chap:diffusion_problem} as a
-model transport problem. Using the linear system generated by the
-discretization of the model problem, we use a spectral analysis to
-generate analytic relations for the eigenvalues of the operator based
-on system parameters. Using the eigenvalue spectra, we then build
-relationships to characterize the transport of stochastic histories in
-a decomposed domain and the fraction of histories that leak from a
-domain and will therefore have to be communicated. Finally, we compare
-these analytic results to numerical experiments conducted with the
-model transport problem.
-
-\subsection{Spectral Analysis }
-\label{subsec:spectral_analysis}
-The convergence of the Neumann series in
-Eq~(\ref{eq:adjoint_neumann_series}) approximated by the Monte Carlo
-solver is dependent on the eigenvalues of the iteration matrix. We
-will compute these eigenvalues by assuming eigenfunctions of the form
-\cite{leveque_finite_2007}:
-\begin{equation}
-  \Phi_{p,q}(x,y) = e^{2 \pi \imath p x} e^{2 \pi \imath q y}\:,
-  \label{eq:eigenfunction_form}
-\end{equation}
-where different combinations of $p$ and $q$ represent the different
-eigenmodes of the solution. As these are valid forms of the solution,
-then the action of the linear operator on these eigenfunctions will
-yield the eigenvalues of the matrix as they exist on the unit circle
-in the complex plane.
-
-For the model problem, we first compute the eigenvalues for the
-diffusion operator $\ve{D}$ by applying the operator to the
-eigenfunctions and noting that $x=ih$ and $y=jh$:
-\begin{multline}
-  \ve{D}\Phi_{p,q}(x,y) = \lambda_{p,q}(\ve{D})
-  =\\ -\frac{D}{6h^2}\Big[4 e^{-2 \pi \imath p h} + 4 e^{2 \pi \imath
-      p h} + 4 e^{-2 \pi \imath q h} + 4 e^{2 \pi \imath q h} + e^{-2
-      \pi \imath p h} e^{-2 \pi \imath q h} \\ + e^{-2 \pi \imath p h}
-    e^{2 \pi \imath q h} + e^{2 \pi \imath p h} e^{-2 \pi \imath q h}
-    + e^{2 \pi \imath p h} e^{2 \pi \imath q h} - 20\Big] + \Sigma_a
-  \:.
-  \label{eq:deriv_diff_1}
-\end{multline}
-Using Euler's formula, we can collapse the exponentials to
-trigonometric functions:
-\begin{equation}
-  \lambda_{p,q}(\ve{D}) = -\frac{D}{6h^2}[ 8 \cos(\pi p h) + 8
-    \cos(\pi q h) + 4 \cos(\pi p h) \cos(\pi q h) - 20] + \Sigma_a\:.
-  \label{eq:deriv_diff_2}
-\end{equation}
-
-As Eq~(\ref{eq:diffusion_eq}) is diagonally dominant, point Jacobi
-preconditioning as outlined in
-\S~\ref{subsec:stochastic_preconditioning} is sufficient to reduce the
-spectral radius of the iteration matrix below unity and therefore
-ensure convergence of the Neumann series. Applying this
-preconditioner, we are then solving the following diffusion system:
-\begin{equation}
-  \ve{M}^{-1} \ve{D} \boldsymbol{\phi} = \ve{M}^{-1} \ve{s}\:.
-  \label{eq:precond_diffsion}
-\end{equation}
-The operator $\ve{M}^{-1} \ve{D}$ is merely the original diffusion
-operator with each row scaled by the diagonal component. As we have
-defined a homogeneous domain, the scaling factor, $\alpha$, is the
-same for all rows in the operator and defined as the $\phi_{i,j}$
-coefficient from Eq~(\ref{eq:fd_system}):
-\begin{equation}
-  \alpha = \Bigg[\frac{10 D}{3 h^2} + \Sigma_a\Bigg]^{-1}\:.
-  \label{eq:jacobi_scaling}
-\end{equation}
-Using this coefficient, we then have the following spectrum of
-preconditioned eigenvalues:
-\begin{equation}
-  \lambda_{p,q}(\ve{M}^{-1} \ve{D}) = \alpha \lambda_{p,q}(\ve{D})\:.
-  \label{eq:preconditioned_eigenvalues}
-\end{equation}
-
-The spectral radius of the iteration matrix is obtained by seeking its
-largest eigenvalue. As with the diffusion operator, we can use the
-same analysis techniques to find the eigenvalues for the iteration
-matrix. We use a few simplifications by noting that if the Jacobi
-preconditioned iteration matrix is $\ve{H} = \ve{I} -
-\ve{M}^{-1}\ve{D}$, the we except all terms on the diagonal of the
-iteration matrix to be zero such that we have the following stencil:
-\begin{multline}
-  \ve{H}\boldsymbol{\phi} = \frac{\alpha D}{6h^2}\Big[4 \phi_{i-1,j} +
-    4 \phi_{i+1,j} + 4 \phi_{i,j-1} + 4 \phi_{i,j+1}
-    +\\ \phi_{i-1,j-1} + \phi_{i-1,j+1} + \phi_{i+1,j-1} +
-    \phi_{i+1,j+1}\Big]\:.
-  \label{eq:iteration_stencil}
-\end{multline}
-Inserting the eigenfunctions defined by
-Eq~(\ref{eq:eigenfunction_form}) we get:
-\begin{multline}
-  \lambda_{p,q}(\ve{H}) = \frac{\alpha D}{6h^2}\Big[4 e^{-2 \pi \imath p
-      h} + 4 e^{2 \pi \imath p h} + 4 e^{-2 \pi \imath q h} + 4 e^{2
-      \pi \imath q h} + e^{-2 \pi \imath p h} e^{-2 \pi \imath q h}
-    \\ + e^{-2 \pi \imath p h} e^{2 \pi \imath q q} + e^{2 \pi \imath
-      p h} e^{-2 \pi \imath q h} + e^{2 \pi \imath p h} e^{2 \pi
-      \imath q h}\Big]\:,
-  \label{eq:iteration_deriv}
-\end{multline}
-which simplifies to:
-\begin{equation}
-  \lambda_{p,q}(\ve{H}) = \frac{\alpha D}{6h^2}[ 8 \cos(\pi p h) + 8
-    \cos(\pi q h) + 4 \cos(\pi p h) \cos(\pi q h)]\:,
-  \label{eq:iteration_spectrum}
-\end{equation}
-giving the eigenvalue spectrum for the Jacobi preconditioned iteration
-matrix. We find that the maximum eigenvalue exists when $p=q=0$,
-giving the following for the spectral radius of the Jacobi
-preconditioned iteration matrix:
-\begin{equation}
-  \rho(\ve{H}) = \frac{10 \alpha D}{3 h^2}\:.
-  \label{eq:iteration_radius}
-\end{equation}
-
-\subsection{Neumann Series Convergence }
-\label{subsec:neumann_convergence}
-As outlined in \S\ref{sec:mc_preliminaries}, the adjoint Monte Carlo
-method is effectively an approximation to a stationary method. In the
-adjoint Neumann-Ulam method, $k$ iterations, equivalent to $k$
-applications of the iteration matrix, are approximated by a random
-walk of average length $k$ to yield the summation in
-Eq~(\ref{eq:adjoint_neumann_solution})
-\cite{dimov_new_1998,danilov_asymptotic_2000}. This random walk
-length, or the number of transitions before the termination of a
-history (either by the weight cutoff, absorption, or exiting the
-global domain) is therefore approximately the number of stationary
-iterations required to converge to the specified tolerance. In the
-case of the adjoint Neumann-Ulam method, no such tolerance exists,
-however, we have specified a weight cutoff, $W_c$, that determines
-when low-weight histories will be prematurely terminated as their
-contributions are deemed minute. After $k$ iterations, a stationary
-method is terminated as the error has reached some fraction,
-$\epsilon$, of the initial error:
-\begin{equation}
-  ||\ve{e}^{k}||_2 = \epsilon ||\ve{e}^0||_2\:.
-  \label{eq:linear_k_iter_norm4}
-\end{equation}
-Per Eq~(\ref{eq:linear_k_iter_norm3}), we see that this fraction is
-equivalent to $\epsilon = \rho(\ve{H})^k$. For the adjoint
-Neumann-Ulam method, if we take this fraction to be the weight cutoff,
-a measure of how accurately the contributions of a particular history
-to the solution are tallied, we then have the following relationship
-for $k$:
-\begin{equation}
-  k = \frac{ \log(W_c) }{ \log( \rho(\ve{H}) ) }\:.
-  \label{eq:analytic_k}
-\end{equation}
-This then gives us a means to estimate the length of the random walks
-that will be generated from a particular linear operator based on the
-eigenvalues of its iteration matrix (independent of the linear
-operator splitting chosen) and based on the weight cutoff parameter
-used in the Neumann-Ulam method.
-
-\subsection{Domain Leakage Approximations }
-\label{subsec:domain_leak_approx}
-In a domain decomposed situation, not all histories will remain within
-the domain they started in and must instead be communicated. This
-communication, expected to be expensive, was analyzed by Siegel and
-colleagues for idealized, load balanced situations for full nuclear
-reactor core Monte Carlo neutral particle simulations
-\cite{siegel_analysis_2012}.  To quantify the number of particles
-that leak out of the local domain they define a leakage fraction,
-$\Lambda$, as:
-\begin{equation}
-  \Lambda = \frac{average\ \#\ of\ particles\ leaving\ local\ domain}
-          {total\ of\ \#\ of\ particles\ starting\ in\ local\ domain}\:.
-          \label{eq:leakage_fraction}
-\end{equation}
-For their studies, Siegel and colleagues assumed that the value of
-$\Lambda$ was dependent on the total cross section of the system via
-the Wigner rational approximation. Outlined more thoroughly by Hwang's
-chapter in \cite{azmy_nuclear_2010}, we will use both the Wigner
-rational approximation and the mean chord approximation as a means to
-estimate the leakage fraction.
-
-In the case of domain decomposed linear systems, we can use diffusion
-theory to estimate the optical thickness of a domain in the
-decomposition and the corresponding leakage fraction in terms of
-properties of the linear operator and the discretization. To begin we
-must first calculate the mean distance a Monte Carlo history will move
-in the grid by computing the mean squared distance of its movement
-along the chord of length $l$ defined across the domain. After a
-single transition a history will have moved a mean squared distance
-of:
-\begin{equation}
-  \langle \bar{r_1^2} \rangle = (n_s h)^2\:,
-  \label{eq:step_1_length}
-\end{equation}
-where $h$ is the size of the discrete grid elements along the chord
-and $n_s$ is the number of grid elements a history will move on
-average every transition. For our diffusion model problem, $n_s$ would
-equate to the expected number of states in the $i$ (or $j$ as the problem is
-symmetric) direction that a history will move in a single
-transition and is dependent on the stencil used for the
-discretization. After $k$ transitions in the random walk, the history
-will have moved a mean squared distance of:
-\begin{equation}
-  \langle \bar{r_k^2} \rangle = k (n_s h)^2\:.
-  \label{eq:step_k_length}
-\end{equation}
-If our chord is of length $l$ and there are $n_i$ grid elements (or
-states to which a history may transition) along that chord, then $h =
-l / n_i$ giving:
-\begin{equation}
-  \langle \bar{r_k^2} \rangle = k \Bigg(\frac{n_s l}{n_i}\Bigg)^2\:.
-  \label{eq:step_k_length_sub}
-\end{equation}
-From diffusion theory, we expect the average number of interactions
-along the chord to be:
-\begin{equation}
-  \tau = \frac{l}{2 d \sqrt{\langle \bar{r_k^2} \rangle}}\:,
-  \label{eq:optical_thickness_1}
-\end{equation}
-where $d$ is the dimensionality of the problem and $\sqrt{\langle
-  \bar{r_k^2} \rangle}$ is effectively the mean free path of the Monte
-Carlo history in the domain. We can readily interpret $\tau$ to be the
-\textit{effective optical thickness} of a domain of length
-$l$. Inserting Eq~(\ref{eq:step_k_length_sub}) we arrive at:
-\begin{equation}
-  \tau = \frac{n_i}{2 d n_s \sqrt{k}}\:,
-  \label{eq:optical_thickness_2}
-\end{equation}
-which if expanded with Eq~(\ref{eq:analytic_k}) gives us the final
-relation for the effective optical thickness:
-\begin{equation}
-  \tau = \frac{n_i}{2 d n_s}
-  \sqrt{\frac{\log(\rho(\ve{H}))}{\log(W_c)}}\:.
-  \label{eq:optical_thickness_3}
-\end{equation}
-
-For optically thin domains, we expect that most histories will be
-communicated, while optically thick domains will leak the fraction of
-histories that did not interact within. Using the optical thickness
-defined in Eq~(\ref{eq:optical_thickness_3}), we can then complete the
-leakage approximations by defining the bounds of $\tau \rightarrow 0,
-\Lambda \rightarrow 1$ and $\tau \rightarrow \infty, \Lambda
-\rightarrow \tau^{-1}$.  With these bounds we can then define the
-leakage fraction out of a domain for the adjoint Neumann-Ulam method
-using the Wigner rational approximation:
-\begin{equation}
-  \Lambda = \frac{1}{1+\tau}\:,
-  \label{eq:wigner_domain_leakage}
-\end{equation}
-and using the mean-chord approximation:
-\begin{equation}
-  \Lambda = \frac{1-e^{-\tau}}{\tau}\:.
-  \label{eq:mean_chord_domain_leakage}
-\end{equation}
-Here, the leakage fraction is explicitly bound to the eigenvalues of
-the iteration matrix, the size of the domain, the content of the
-discretization stencil, and the weight cutoff selected to terminate
-low weight histories.
-
-\subsection{Numerical Experiments }
-\label{subsec:numerical_experiments}
-To test the relationships developed by the spectral analysis, we form
-two simple numerical experiments using the diffusion model problem:
-one to measure the length of the random walks as a function of the
-iteration matrix eigenvalues, and one to measure the domain leakage
-fraction as a function of the iteration matrix eigenvalues and the
-discretization properties. Before doing this, we verify our
-computation of the spectral radius of the iteration matrix by
-numerically computing the largest eigenvalue of the diffusion operator
-using an iterative eigenvalue solver. For this verification, a $100
-\times 100$ square grid with $h=0.01$, $h=0.1$, and $h=1.0$ and the
-absorption cross varied from 0 to 100 while the scattering cross
-section was fixed at unity. Figure~\ref{fig:measured_spec_rad} gives
-the measured spectral radius of the iteration matrix and the computed
-spectral radius for the preconditioned diffusion operator using
-Eq~(\ref{eq:iteration_radius}) as function of the absorption to
-scattering ratio $(\Sigma_a / \Sigma_s)$. Excellent agreement was
-observed between the analytic and numerical results with all data
-points computed within the tolerance of the iterative eigenvalue
-solver.
-\begin{figure}[t!]
-  \begin{spacing}{1.0}
-    \begin{center}
-      \includegraphics[width=4in,clip]{chapters/parallel_mc/measured_spec_rad.png}
-    \end{center}
-    \caption{\textbf{Measured and analytic preconditioned diffusion
-        operator spectral radius as a function of the absorption cross
-        section to scattering cross section ratio.} \textit{Values of
-        $h=0.01$, $h=0.1$, and $h=1.0$ were used. The red data was
-        computed numerically by an eigensolver while the black dashed
-        data was generated by Eq~(\ref{eq:iteration_radius}).}}
-    \label{fig:measured_spec_rad}
-  \end{spacing}
-\end{figure}
-
-\subsubsection{Random Walk Length }
-\label{subsubsec:walk_length}
-With the eigenvalue derivations verified, we can go about setting up
-an experiment to measure the length of the random walks generated by
-the adjoint Neumann-Ulam solver. To do this, we again use a $100
-\times 100$ square grid with $h=0.1$ and the absorption cross varied
-from 0 to 100 while the scattering cross section was fixed at
-unity. Three weight cutoff values of \sn{1}{-2}, \sn{1}{-4}, and
-\sn{1}{-8} were used with 10,000 histories generated by a point source
-of strength 1 in the center of the domain. For each of the histories,
-the number of transitions made was tallied to provide an effective
-value of $k$ for each history. This value was then averaged over all
-histories to get a measured value of $k$ for the particular
-operator. On the left, Figure~\ref{fig:measured_length} presents these
-measurements as well as the analytic result computed by
-Eq~(\ref{eq:analytic_k}) as a function of the iteration matrix
-spectral radius, $\rho(\ve{H})$. On the right,
-Figure~\ref{fig:measured_length} gives the relative difference between the
-predicted and observed results. We note good qualitative agreement
-between the measured and analytic results. However, we observe a
-larger relative difference for both long and short random walks.
-\begin{figure}[t!]
-  \begin{spacing}{1.0}
-    \begin{center}
-      \includegraphics[width=6.0in,clip]{chapters/parallel_mc/measured_length_2.pdf}
-    \end{center}
-    \caption{\textbf{Measured and analytic random walk length as a
-        function of the iteration matrix spectral radius.} \textit{The
-        weight cutoff was varied with \sn{1}{-2}, \sn{1}{-4}, and
-        \sn{1}{-8}. In the left plot, the red data was computed
-        numerically by an adjoint Neumann-Ulam implementation while
-        the black dashed data was generated by
-        Eq~(\ref{eq:analytic_k}). In the right plot, the relative
-        difference between the predicted and measured results is
-        presented for each weight cutoff.}}
-    \label{fig:measured_length}
-  \end{spacing}
-\end{figure}
-
-\subsubsection{Domain Leakage }
-\label{subsubsec:domain_leakage}
-Finally, we seek to measure the leakage from a domain in a domain
-decomposed Monte Carlo calculation and assess the quality of our
-analytic relation for the optical thickness of a domain and the
-associated leakage approximations. For this experiment, a square grid
-with $h=0.1$ was decomposed into 9 square domains, 3 in each cardinal
-direction with measurements occurring in the central domain without
-boundary grid points. For the cross sections, the absorption cross
-section was varied from 1 to 100 while the scattering cross
-section was set to zero to create a purely absorbing environment
-with weight cutoff of \sn{1}{-4}. The optical thickness of these
-domains will vary as a function of the absorption cross section if
-the other parameters are fixed.
-
-To compute the optical thickness, along with the spectral radius as
-given by Eq~(\ref{eq:iteration_radius}), we also need the parameters
-$n_i$ and $n_s$ which respectively describe the typical domain length
-and the average number of states moved along that typical length per
-history transition. For our grid above, the domains are varied in size
-with $50 \times 50$, $100 \times 100$, and $200 \times 200$ cells
-giving $n_i=50$, $n_i=100$, and $n_i=200$ grid points or states along
-the typical length of the domain respectively. Looking at the
-Laplacian stencil in Eq~(\ref{eq:nine_point_stencil}), we see that all
-history transitions will only move a single state in either the $i$ or
-$j$ directions due to the symmetry of the problem. Furthermore, if we
-choose the $i$ direction, not all states we will transition to will
-move the history in that direction. Therefore, we look to the
-definition of the iteration matrix in Eq~(\ref{eq:iteration_stencil})
-and the definition of the adjoint probability matrix in
-Eq~(\ref{eq:adjoint_probability}) to estimate the $n_s$ parameter. For
-a particular transition starting at state $(i,j)$, 6 of the 8 possible
-new states in the stencil move the history in $i$ direction with
-relative coefficients of 4 for moving in the $(\pm i,0)$ direction and
-of 1 for moving in the $(\pm i,\pm j)$. These coefficients dictate the
-frequency those states are visited relative to the others. For those 6
-states we can visit along the typical length, their sum is 12 out of
-the total 20 for the coefficients for all possible states with their
-ratio giving $n_s = \frac{3}{5}$.
-
-To compute the leakage fraction numerically, \sn{3}{5} histories were
-sampled from a uniform source of strength unity over the global
-domain. At the start of a stage of histories, the number of histories
-starting in the center domain was computed and as the stage
-progressed, the number of histories that exited that domain was
-tallied with the ratio of the two numbers providing a numerical
-measure for the leakage fraction. Figure~\ref{fig:measured_leakage}
-gives the domain leakage measurements for the domain in the center of
-the global grid as well as the analytic result computed by
-Eqs~(\ref{eq:wigner_domain_leakage}) and
-(\ref{eq:mean_chord_domain_leakage}) as a function of the iteration
-matrix spectral radius.
-\begin{figure}[t!]
-  \begin{spacing}{1.0}
-    \begin{center}
-      \includegraphics[width=4.75in,clip]{chapters/parallel_mc/leakage_variation_2.pdf}
-    \end{center}
-    \caption{\textbf{Measured and analytic domain leakage as a
-        function of the iteration matrix spectral radius.} \textit{To
-        test the behavior with respect to domain size, $n_i=50$,
-        $n_i=100$,and $n_i=200$ were used. The red data was computed
-        numerically by a domain-decomposed adjoint Neumann-Ulam
-        implementation, the black dashed data was generated by
-        Eq~(\ref{eq:mean_chord_domain_leakage}) using the mean-chord
-        approximation, and the dashed-dotted black data was generated
-        by Eq~(\ref{eq:wigner_domain_leakage}) using the Wigner
-        rational approximation.}}
-    \label{fig:measured_leakage}
-  \end{spacing}
-\end{figure}
-Again, we note good qualitative agreement between the measured and
-analytic quantities but we begin to see the limits of the leakage
-approximations.
-
-To compare the quality of the two approximations, the absolute
-difference between the computed leakage fraction and that generated by
-the Wigner rational and mean chord approximations is plotted in
-Figure~\ref{fig:leakage_error} for all domain sizes tested.
-\begin{figure}[t!]
-  \begin{spacing}{1.0}
-    \begin{center}
-      \includegraphics[width=4.75in,clip]{chapters/parallel_mc/leakage_error_2.pdf}
-    \end{center}
-    \caption{\textbf{Measured and analytic domain leakage absolute
-        difference as a function of the iteration matrix spectral radius.}
-      \textit{To test the behavior with respect to domain size,
-        $n_i=50$ (green), $n_i=100$ (blue), and $n_i=200$ (red) were
-        used. The dashed lines represent the difference using the Wigner
-        rational approximation while the solid lines represent the
-        difference using the mean-chord approximation.}}
-    \label{fig:leakage_error}
-  \end{spacing}
-\end{figure}
-From these difference results, the mean chord approximation is shown to
-have a lower difference for ill-conditioned systems as compared to the
-Wigner approximation while the Wigner approximation produces less
-difference for more well-conditioned systems. We also note that for the
-optically thick domains, the difference is likely corresponded to that
-observed in Figure~\ref{fig:measured_length} for the $k$ parameter
-while the large relative difference in $k$ for optically thin domains does
-not affect the approximation significantly. In general, the mean chord
-approximation is a better choice to estimate the leakage fraction in a
-domain from the adjoint Neumann-Ulam method and except for a single
-data point with $n_i=50$, the mean chord approximation yielded leakage
-fractions within 0.05 of the measured results. As the domain becomes
-more optically thick (with both increasing $n_i$ and decreasing
-$\rho(\ve{H})$), the approximations are more accurate.
 

### 2013-08-23 doc/dissertation/document/backmatter/parallel_theory.tex
fixing undefined references in parallel chapter

@@
-\chapter{Analytic Framework for Domain-Decomposed Monte Carlo\ }
+\chapter{Analytic Framework for\\ Domain-Decomposed Monte Carlo\ }
 \label{chap:parallel_theory}
-Because we have chosen a domain decomposition strategy in a parallel
-environment, this means that communication of these histories must
-occur between compute nodes owning neighboring pieces of the global
-domain. We wish to characterize this communication not only because
-communication is in general expensive, but also because these
-nearest-neighbor communication sequences, specifically, have poor
-algorithmic strong scaling \cite{gropp_high-performance_2001} in much
-the same way as a parallel matrix-vector multiply
-operation. Therefore, we desire a framework to provide a simple,
-analytic theory based on the properties of the transport system that
-will allow for estimates of the domain decomposed behavior of the
-Neumann-Ulam method in terms of the amount of information that must be
+Because we have chosen a domain decomposition strategy for the
+Neumann-Ulam method, this means that communication of the stochastic
+histories must occur between compute nodes owning neighboring pieces
+of the global domain. We wish to characterize this communication not
+only because communication is in general expensive, but also because
+these nearest-neighbor communication sequences, specifically, have
+poor algorithmic strong scaling \cite{gropp_high-performance_2001} in
+much the same way as a parallel matrix-vector multiply operation. In
+this appendix, we devise a framework to provide a simple, analytic
+theory based on the properties of the transport system that will allow
+for estimates of the domain decomposed behavior of the Neumann-Ulam
+method in terms of the amount of information that must be
 communicated.

### 2013-08-23 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
fixing undefined references in parallel chapter

@@
   \begin{center}
-    \scalebox{1.5}{ \input{chapters/parallel_mc/domain_to_domain.pdftex_t} }
+    \scalebox{1.25}{ \input{chapters/parallel_mc/domain_to_domain.pdftex_t} }
   \end{center}
@@
   \begin{center}
-    \scalebox{1.25}{
+    \scalebox{1.0}{
       \input{chapters/parallel_mc/binary_comm_tree.pdftex_t} }
@@
 overlap. For example, consider the analytic relations derived in
-\S~\ref{sec:analytic_framework} for the length of random walks in a
+Appendix~\ref{chap:parallel_theory} for the length of random walks in a
 Neumann-Ulam method and the amount of leakage from a domain
@@
 do this, we utilize the set of analytic relations developed in
-\S~\ref{sec:analytic_framework} to design these numerical
+Appendix~\ref{chap:parallel_theory} to design these numerical
 experiments. For the neutron diffusion problem used in these scaling
@@
   subdomain can be quantified analytically using algebraic quantities
+  (see Appendix~\ref{chap:parallel_theory})
 \item The multiple-set overlapping-domain (MSOD) parallel algorithm

### 2013-08-23 doc/dissertation/document/backmatter/extra_scaling.tex
moving subomain nuemann ulam results for multiple sets into an appendix

--- 2013-08-23 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
@@
-\chapter{Parallel Monte Carlo Synthetic \\ Acceleration Methods\ }
-\label{ch:parallel_methods}
-
-To date, parallel Neumann-Ulam methods have been limited to full
-domain replication with parallelism exploited through individual
-histories \cite{alexandrov_efficient_1998} and in this chapter we will
-exploit particle transport algorithms to alleviate this. To accomplish
-this, we recognize from the literature that stochastic histories must
-be transported from domain to domain as the simulation progresses and
-they transition to states that are not in the local domain. In this
-chapter we briefly review particle transport methods for domain
-decomposed Monte Carlo. We then devise a parallel algorithm for the
-Neumann-Ulam method based on the multiple-set overlapping-domain
-decomposition algorithm and a parallel algorithm for the MCSA
-iteration that leverages the parallel Monte Carlo algorithm and
-general parallel matrix-vector operations. Using parallel solutions
-for the neutron diffusion problem, we verify the correctness of the
-parallel algorithm and implementation by comparing the numerical
-results against production Krylov methods. With the verified
-implementation of the algorithm, we perform parallel scaling studies
-to test its performance on a leadership class machine and compare this
-performance to the production Krylov methods.
-
-%%---------------------------------------------------------------------------%%
-\section{Domain Decomposed Monte Carlo\ }
-\label{sec:msod}
-Large-scale problems will for reasons typically related to memory
-restrictions or performance have their data partitioned such that each
-parallel process owns a subset of the equations in the linear
-system. Given this convention, the Neumann-Ulam Monte Carlo algorithm
-must perform random walks over a domain that is decomposed and must
-remain decomposed to avoid the same performance and memory
-restrictions that required the domain to be decomposed. To parallelize
-this algorithm, we then seek parallel Monte Carlo algorithms that
-handle domain decomposition.
-
-To motivate this problem, consider the square domain presented in
-Figure~\ref{fig:ddmc_example} and on this domain we imagine a Monte
-Carlo particle transport problem. If the domain were decomposed into 9
-subdomains as shown, each of those subdomains and their associated
-data (i.e. cross sections) could be owned by a different parallel
-process in the computation. In Figure~\ref{fig:ddmc_example}, the
-tracks of 3 particles born in the center subdomain are shown. In this
-example, particle A is first transported from the center domain to the
-domain directly to the left. Before the scattering event in the new
-domain may be processed, particle A must be communicated between
-between the two parallel processes that own those domains. For any
-given particle in the transport simulation, this communication event
-may hardly occur during the lifetime of the particle (particle B), or
-may occur many times (particle C) depending on the problem parameters
-and the random path in phase space taken by the particle. Scalable
-parallel algorithms for domain decomposed Monte Carlo are those that
-handle this domain-to-domain communication of particles effectively
-and balance it with the on-process computations for particle
-interactions, tallies, response calculations, and other simulation
-requirements.
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.5}{
-      \input{chapters/parallel_mc/ddmc_example.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Domain decomposed Monte Carlo transport example
-      illustrating how domain decomposition requires parallel
-      communication of particles.} \textit{If a particle crosses a
-      domain boundary it must be communicated to the next domain in
-      its path.}}
-  \label{fig:ddmc_example}
-\end{figure}
-
-For a domain decomposed Neumann-Ulam method, the situation is very
-much the same where instead the transport process is now discrete and
-the "physics" of the transport process is the Monte Carlo game with
-probabilities and weights described by
-Eq~(\ref{eq:neumann_ulam_decomposition}) for the given linear
-problem. Figure~\ref{fig:ddnu_example} gives the domain decomposed
-Neumann-Ulam analog of particle transport problem in
-Figure~\ref{fig:ddmc_example}. Imagine in this problem that we are
-solving for the monoenergetic scalar neutron flux as in the model
-diffusion problem presented in Appendix~\ref{chap:diffusion_problem}.
-In this problem, the solution has only a spatial dependence and
-therefore each point in the mesh corresponds to an equation in the
-resulting linear system. In Figure~\ref{fig:ddnu_example}, the domain
-has now been discretized by this mesh and again decomposed into 9
-subdomains, each owning a piece of the mesh and therefore the
-corresponding equations in the linear system. As we play the Monte
-Carlo game presented in Chapter~\ref{ch:stochastic_methods}, the
-discretization and resulting Neumann-Ulam decomposition of the problem
-describes how each state in the system is coupled to the other states
-in the system. This coupling is then responsible for the discrete
-random walk that each history takes as shown in this example. As
-discrete states (or rows in the linear system) that do not exist in
-the local domain are reached during the random walk by stochastic
-histories, the histories must be communicated to the subdomain that
-does own the discrete state in an analogous fashion to the particles
-in the previous example.
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.5}{
-      \input{chapters/parallel_mc/ddnu_example.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Domain decomposed Neumann-Ulam example illustrating
-      how domain decomposition requires parallel communication of
-      histories.} \textit{Each mesh point corresponds to an equation
-      in the linear system and the coupling among equations is
-      described by the discretization of the problem.}}
-  \label{fig:ddnu_example}
-\end{figure}
-
-To further motivate using a parallel algorithm similar to particle
-transport methods, again consider history A in
-Figure~\ref{fig:ddnu_example}. This history begins in the center
-subdomain and through the course of transitioning through states in
-the system arrives at a state that exists in the subdomain directly to
-the left of the center subdomain. History A must then be communicated
-from the center subdomain to the subdomain immediately to the left to
-continue the random walk. As with the particle transport system, we
-may find that histories in a domain decomposed Neumann-Ulam method are
-only communicated a few times (history B) if at all, or they may be
-communicated many times (history C) depending on the discretization
-and the outcome of the Monte Carlo game. In identical fashion to the
-particle transport problem, a domain decomposed Neumann-Ulam method
-has the same communication requirements with all histories in both
-examples requiring the same parallel operations. The amount of
-communication of histories from domain to domain will be the primary
-factor in the parallel scalability of the algorithm. In
-Appendix~\ref{chap:parallel_theory}, the number of histories
-communicated in a given problem are quantified analytically for a
-simple model problem.
-
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{Domain Decomposed Neumann-Ulam Algorithm\ }
-\label{sec:asynchronous_algorithm}
-In the context of radiation transport, in 2009 Brunner and colleagues
-provided a fully asynchronous domain decomposed parallel algorithm as
-implemented in production implicit Monte Carlo codes
-\cite{brunner_efficient_2009}. We will adapt their algorithm and
-directly apply it to a parallel formulation of the Neumann-Ulam
-method. Direct analogs can be derived from their works by noting that
-the primary difference between solving a linear transport system with
-Monte Carlo methods and traditional fixed source Monte Carlo transport
-problems is the content of the Markov chains that are generated.
-
-The transitions represented by these chains are bound by probabilities
-and weights and are initiated by the sampling of a discrete source. In
-the context of transport problems, those transitions represent events
-such as particle scattering and absorption with probabilities that are
-determined by physical data in the form of cross sections. For
-stochastic matrix inversion, those transitions represent moving
-between the equations of the linear system (and therefore the
-components of phase space which they represent) and their
-probabilities are defined by the coefficients of those
-equations. Ultimately, we tally the contributions to generate
-expectation values in the desired states as we progress through the
-chains. Therefore, parallel methods for Monte Carlo radiation
-transport can be abstracted and we can use those concepts that apply
-to matrix inversion methods as an initial means of developing a
-parallel Neumann-Ulam-type solver.
-
-In this section Brunner and Brantley's fully asynchronous algorithm,
-which was effectively implemented verbatim for this work, is presented
-along with its application to the Neumann-Ulam method. For
-considerably more detail, the algorithm presented can be found in
-\cite{brunner_efficient_2009}. In their work they identify two data
-sets that are required to be communicated: the sharing of particles
-that are transported from one domain to another and therefore from one
-processor to another and a global communication that signals if
-particle transport has been completed on all processors. Both of these
-communication sequences will be addressed at a high level along with
-how the Monte Carlo data structures they require are constructed in
-parallel.
-
-\subsection{Parallel Transport Domain and Source }
-\label{subsec:domain_generation}
-To utilize a parallel transport algorithm, we must generate the
-required data with the correct parallel decomposition. For the
-Neumann-Ulam method, the transport domain consists of all states in
-the system that are local, and the probabilities and weights for all
-state transitions possible in the local domain. At the core of this
-representation is the Neumann-Ulam decomposition of the linear
-operator as given by Eq~\ref{eq:neumann_ulam_decomposition}. Given an
-input decomposition for the linear operator, by definition the
-Neumann-Ulam decomposition will have the same parallel
-decomposition. In addition, any modification made to the linear
-operator through preconditioning or relaxation parameters will also
-modify the Neumann-Ulam decomposition. All possible transitions are
-described by the local graph of the sparse input matrix. In addition,
-any left preconditioning or relaxation parameters will modify the
-right hand side of the linear system and therefore the fixed source in
-the Monte Carlo calculation. This source vector will have the same
-parallel decomposition as the input matrix and therefore all of the
-birth states for the histories will exist for at least a single
-transition event within the local domain.
-
-Compared to the serial construction of the Neumann-Ulam decomposition,
-data for all states that is required to be on process must be
-collected in parallel. For the pure domain decomposition case, given
-an input parallel decomposition for the Neumann-Ulam decomposition,
-for all local states $m$ we require access to $P_{mn}$ and $W_{mn}$
-(or $P^T_{mn}$ and $W^T_{mn}$) for the adjoint method for all possible
-states $n$. Given by Figure~\ref{fig:diffusion_graph}, consider the
-adjacency graph of a single state in the neutron diffusion matrix for
-the model problem presented in Appendix~\ref{chap:diffusion_problem}.
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.5}{ \input{chapters/parallel_mc/stencil_graph.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Neutron diffusion equation state adjacency graph.}
-    \textit{The structure of the graph comes from the discretization
-      of the Laplacian operator describing the diffusion
-      physics. Adjacent boundary states are collected by traversing
-      the graph one step for all local boundary states and collecting
-      the equations for those states that are not local. For a given
-      transition in the random walk for the diffusion problem, a
-      history at mesh point $(i,j)$ may transition to all adjacent
-      mesh points including itself, corresponding to a global state
-      transition from $m$ to $n$ in the linear system.}}
-  \label{fig:diffusion_graph}
-\end{figure}
-
-In this example, we currently reside in state $m$ of the system,
-directly correlating to physical location $(i,j)$ in the mesh (the
-center node). The discretization stencil of the diffusion problem
-dictates the structure of this graph and the states to which a history
-may transition. If we play the Monte Carlo game to move to a new state
-$n$, then we may move to any of the nodes in this graph, including the
-node at which we started. If grid point $(i,j)$ in this example is on
-the right of the local domain boundary and we transition to the right
-to state $n$ corresponding to mesh point $(i+1,j)$, we are now in a
-state that is owned by the adjacent domain. We must then gather the
-data in the Neumann-Ulam decomposition from the neighboring process
-that owns grid point $(i+1,j)$. Doing this provides us with the data
-required by the estimators permitting $P_{mn}$ and $W_{mn}$ to be
-computed locally for this particular boundary transition. In addition,
-we also collect the identification number for the process that owns
-this state such that we have an address to send all histories that
-leave the local domain boundary for this state. For the parallel
-source, these adjacent states are not required as histories may only
-be born in states in the local domain. Once the local Neumann-Ulam
-decomposition has been generated with the proper data collected from
-adjacent domains, Monte Carlo transport may proceed in parallel.
-
-\subsection{Domain Decomposed Algorithm}
-\label{subsec:parallel_mc_algorithm}
-
-Here we present in detail Brunner and Brantley's 2009 algorithm in
-detail and discuss how it is adapted to parallelize the Neumann Ulam
-method. Presented in Algorithm~\ref{alg:parallel_mc_algorithm}, the
-top level sequence performs the task of managing history transport
-through the local domain, communication of histories to adjacent
-domains, and the completion of transport. For each of these specific
-tasks, additional algorithms, shown in bold in
-Algorithm~\ref{alg:parallel_mc_algorithm}
-(e.g. \textbf{LocalHistoryTransport()}), are presented for additional
-detail in the same manner as Brunner and Brantley.
-
-\begin{algorithm}[h!]
-  \caption{Parallel Neumann-Ulam Algorithm}
-  \label{alg:parallel_mc_algorithm}
-  \begin{algorithmic}[1]
-    \State get list of neighbor processors 
-    \Comment{Each neighbor owns an adjacent subdomain}
-    \ForAll{neighbors}
-    \State post nonblocking receive for maximum history buffer size
-    \State allocate history buffer
-    \EndFor
-    \State \textit{historiesCompleted} = 0
-    \Comment{local+child finished histories}
-    \State \textit{localProcessed} = 0
-    \Comment{local trajectories computed (not necessarily finished)}
-    \State calculate parent and children processor ID numbers in
-    binary tree
-    \ForAll{child processes}
-    \State post nonblocking receive for \textit{historiesCompleted} tally
-    \EndFor
-    \State post nonblocking receive for stop message from parent
-    \While{stop flag not set}
-    \If{any local histories in source or stack}
-    \State \textbf{LocalHistoryTransport()}
-    \State ++\textit{localProcessed}
-    \EndIf
-    \If{message check period == \textit{localProcessed} || no local histories}
-    \State \textbf{ProcessMessages()}
-    \State \textit{localProcessed} = 0
-    \EndIf
-    \If{no local histories}
-    \State \textbf{ControlTermination()}
-    \EndIf
-    \EndWhile
-    \State cancel outstanding nonblocking requests
-    \State free all history buffers
-  \end{algorithmic}
-\end{algorithm}
-
-Successful execution of this algorithm requires construction of the
-parallel Neumann-Ulam decomposition as described in the preceding
-section. This data is used to begin
-Algorithm~\ref{alg:parallel_mc_algorithm} where lines 1-5 use the
-collected list of neighboring processors generated while building the
-Neumann-Ulam decomposition to setup the set of asynchronous messages
-required for domain-to-domain communication. Again, consider this
-communication pattern for the 9 subdomain example given by
-Figure~\ref{fig:nearest_neighbor_comm}. In this pattern, each boundary
-domain has two neighbors and the center domain four neighbors with
-which they will communicate parallel histories and this communication
-goes both ways as represented by the adjacent arrows in the
-figure. For each set of neighbors, a non-blocking send and receive is
-required with data buffers allocated with a user-defined size prepared
-for incoming and outgoing histories. This nonblocking structure is
-critical to the performance of the algorithm in that it permits local
-history transport to continue while new histories to transport are
-being collected in the buffers. When a given process is ready to do
-more work, it can check these data buffers for incoming histories
-requiring further transport. In this way there is a maximum amount of
-overlap between communication and transport of histories.
-
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.5}{ \input{chapters/parallel_mc/domain_to_domain.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Nearest neighbor history communication sequence.}
-    \textit{Each subdomain in the system has a set of nearest
-      neighbors determined by the parallel adjacency graph of the
-      input matrix. The subdomains are indexed by an integer.}}
-  \label{fig:nearest_neighbor_comm}
-\end{figure}
-
-Once the nearest-neighbor communication sequence has been prepared,
-the completion of transport sequence is readied in lines 6 and 8-12 in
-Algorithm~\ref{alg:parallel_mc_algorithm}. In these lines we are
-setting up an asynchronous binary communication tree as presented for
-the same 9 subdomain example in Figure~\ref{fig:binary_comm_tree}. In
-this communication pattern, each process has one parent process
-(except for the MASTER process 0) to which it will nonblocking send
-the number of histories that terminated their transport procedure by
-weight cutoff in the local domain. Equivalently, each process has up
-to two child processes from which it will receive their completed
-number of histories with a nonblocking receive operation. Setting up a
-tree in this manner lets the completed history tally
-(\textit{historiesCompleted} in the algorithm) be updated
-incrementally and funneled to the root process. Because we are
-solving fixed source problems with the Neumann-Ulam algorithm without
-any type of variance reduction that may generate more histories that
-we started with, once the root process completion sum tallies to the
-number of histories in the source, transport is complete. Once this
-occurs, the root process nonblocking sends a process to its children
-and each process nonblocking receives a stop message from its parent
-as shown in Figure~\ref{fig:binary_comm_tree}. The stop message is
-then propagated up the tree in the same manner.
-
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.25}{
-      \input{chapters/parallel_mc/binary_comm_tree.pdftex_t} }
-    \caption{\textbf{Binary communication tree for coordinating the
-        end of a parallel Neumann-Ulam solve.} \textit{Each child
-        process reports to its parents how many histories completed
-        within its domain. When the root process sums all complete
-        histories, it forwards the stop signal to its children which
-        in turn forward the message up the tree. The subdomains are
-        indexed by an integer.}}
-  \end{center}
-  \label{fig:binary_comm_tree}
-\end{figure}
-
-With these communication structures prepared, we can now enter the
-main transport loop at line 13 of
-Algorithm~\ref{alg:parallel_mc_algorithm}. In this loop, a set of
-mutually exclusive tasks enabled by the fully asynchronous
-communication patterns are executed until the stop signal is received
-from the parent process in the binary tree. In this case, mutually
-exclusivity permits all local operations to occur independently of
-other processes in the system and their current state. For each
-process, there are two data structures from which histories may be
-obtained for the transport procedure. The first is the local source
-and the second is a LIFO (last in first out) stack of histories
-transported to the local domain from the adjacent domain. In lines
-14-17, if histories exist in either of these data structures, then
-they are transported through the local domain using
-Algorithm~\ref{alg:local_history_transport}. If the history is
-terminated by weight cutoff the \textit{historiesCompleted} tally is
-updated. If it hits the boundary of the domain it is added to the
-outgoing history buffer for the neighboring domain and if the buffer
-is full, it is sent to the neighboring domain with a nonblocking
-operation and the memory associated with that buffer reallocated. In
-all instances that Algorithm~\ref{alg:local_history_transport} is
-executed, \textit{localProcessed} is incremented to account for
-histories that have been processed locally.
-
-\begin{algorithm}[h!]
-  \caption{\textbf{LocalHistoryTransport()}}
-  \label{alg:local_history_transport}
-  \begin{algorithmic}[1]
-    \State transport history through the domain until termination
-    \If{history is in neighbor boundary state}
-    \State add history to neighbor buffer
-    \If{neighbor buffer is full}
-    \State nonblocking send history buffer to neighbor
-    \State allocate new history buffer for neighbor
-    \EndIf
-    \Else
-    \If{history terminated by weight cutoff}
-    \State post-process history
-    \State ++\textit{historiesCompleted}
-    \EndIf
-    \EndIf
-  \end{algorithmic}
-\end{algorithm}
-
-Continuing in the transport loop, if there are no local histories to
-transport or the \textit{localProcessed} count has reached some
-user-defined check frequency, lines 18-21 in
-Algorithm~\ref{alg:local_history_transport} check for more histories
-to transport in the incoming data buffers by calling
-Algorithm~\ref{alg:process_messages}. For each neighboring domain in
-the problem, if there are histories in those data buffers then they
-are added to the stack from processing and the nonblocking receive
-operation re-instantiated. In addition, the terminated histories count
-is updated with those values received from the child processes.
-
-\begin{algorithm}[h!]
-  \caption{\textbf{ProcessMessages()}}
-  \label{alg:process_messages}
-  \begin{algorithmic}[1]
-    \ForAll{received history buffers from neighbor}
-    \State unpack number of histories in buffer
-    \State add the histories to the running stack
-    \State repost nonblocking receive with neighbor
-    \EndFor
-    \ForAll{\textit{historiesCompleted} messages from children}
-    \State \textit{historiesCompleted} += message value
-    \State repost nonblocking receive with child
-    \EndFor
-  \end{algorithmic}
-\end{algorithm}
-
-Finally, the transport loop finishes in lines 22-24 of
-Algorithm~\ref{alg:local_history_transport} by calling
-Algorithm~\ref{alg:control_termination} if there are no local
-histories to transport in the stack or the source. In
-Algorithm~\ref{alg:control_termination}, we continue to forward the
-terminated history tally down to the parent in the binary tree and
-send all history buffers to our neighbors, even if they are not
-full. If Algorithm~\ref{alg:control_termination} is called by the
-MASTER process, it checks for completion of the problem and if it is
-complete, forwards to stop flag onto its children as in
-Figure~\ref{fig:binary_comm_tree}. If the process is not the MASTER,
-then we check for a stop signal from the parent and forward to the
-child processes if transport is complete.
-
-\begin{algorithm}[h!]
-  \caption{\textbf{ControlTermination()}}
-  \label{alg:control_termination}
-  \begin{algorithmic}[1]
-    \State nonblocking send an partially full history buffers
-    \ForAll{\textit{historiesCompleted} messages from children}
-    \State \textit{historiesCompleted} += message \textit{historiesCompleted}
-    \State repost nonblocking receive with child
-    \EndFor
-    \If{MASTER processor}
-    \If{\textit{historiesCompleted} == global \# of source histories}
-    \State set stop flag
-    \ForAll{children}
-    \State nonblocking send stop message to child
-    \EndFor
-    \EndIf
-    \Else
-    \State nonblocking send \textit{historiesCompleted} to parent
-    \State \textit{historiesCompleted} = 0
-    \State check for stop signal from parent
-    \If{stop signal from parent}
-    \ForAll{children}
-    \State nonblocking send stop message to child
-    \EndFor
-    \EndIf
-    \EndIf
-  \end{algorithmic}
-\end{algorithm}
-
-In this manner the transport loop continues until all process have
-received the stop signal from their parents. Older versions of this
-algorithm, particularly some of those presented in
-\cite{brunner_comparison_2006}, use a master/slave approach as given
-by Figure~\ref{fig:master_comm_tree} for the completion of a transport
-stage instead of the binary tree scheme. In this approach, the root
-process still manages the final summation of the completion tally, it
-directly receives completed tally results from all other processes in
-the problem instead of from just its children. Although this
-implementation is potentially simpler to understand than the binary
-tree approach, the authors of \cite{brunner_comparison_2006} observed
-that this type of termination pattern, even when implemented in a
-fully asynchronous manner, was a major scalability bottleneck. This
-bottleneck is due to the fact that the master process falls behind the
-others, creating a load imbalance even with the addition of a few
-integer addition operations. We will explicitly demonstrate in scaling
-studies how this bottleneck also occurs within a Neumann-Ulam
-implementation of this algorithm thus requiring the implementation of
-a binary communication tree.
-
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{0.75}{
-      \input{chapters/parallel_mc/master_comm_tree.pdftex_t} }
-    \caption{\textbf{Master/slave scheme for coordinating the end of a
-        parallel Neumann-Ulam solve.} \textit{Each slave process
-        reports to the master how many histories completed within its
-        domain. When the master process sums all complete histories,
-        it sends the stop signal to all slave processes. The
-        subdomains are indexed by an integer.}}
-  \end{center}
-  \label{fig:master_comm_tree}
-\end{figure}
-
-There is an additional advantage to Brunner and Brantley's work that
-although not immediately applicable to this work has the potential to
-provide value in the future. In addition to a robust fully
-asynchronous communication pattern, this algorithm may also be modified
-to account for situations where the total number of histories in a
-given stage are not known before starting transport. From a physics
-perspective, we might expect this for situations where perhaps an
-(n,2n) interaction occurs in a neutronics problem. In this case, the
-algorithm is modified to account for both histories terminated and
-created and several mechanisms are introduced to determine completion
-of the transport stage. For future variations of this work, certain
-variance reduction techniques which create histories, such as
-splitting, have the potential to be successfully employed as a means
-of accelerating the time to solution for a given problem using
-MCSA. The parallel Neumann-Ulam algorithm presented here may be
-adapted to account for these additional techniques.
-
-%%---------------------------------------------------------------------------%%
-\section{Multiple-Set Overlapping-Domain Algorithm\ }
-\label{subsec:msod}
-Although the implementation presented in the previous section was
-observed by Brunner and Brantley to be robust and allowed for scaling
-to large numbers of processors, performance issues were still noted
-with parallel efficiency improvements needed in both the weak and
-strong scaling cases for unbalanced problems. These results led them
-to conclude that a combination of domain decomposition and domain
-replication could be used to solve some of these issues. In 2010,
-Wagner and colleagues developed the \textit{multiple-set
-  overlapping-domain} (MSOD) decomposition for parallel Monte Carlo
-applications for full-core light water reactor analysis
-\cite{wagner_hybrid_2010}. In their work, an extension of Brunner's,
-their scheme employed similar parallel algorithms for particle
-transport but a certain amount of overlap between adjacent domains was
-used to decrease the number of particles leaving the local domain. In
-addition, Wagner utilized a level of replication of the domain such
-that the domain was only decomposed on $O(100)$ processors and if
-replicated $O(1,000)$ times potentially achieves efficient simulation
-on $O(100,000)$ processors, thus providing both spatial and particle
-parallelism.
-
-Each collection of processors that constitutes a representation of the
-entire domain is referred to as a set, and within a set overlap occurs
-among its sub-domains. The original motivation was to decompose the
-domain in a way that it remained in a physical cabinet in a large
-distributed machine, thus reducing latency costs during
-communication. A multiple set scheme is also motivated by the fact
-that communication during particle transport only occurs within a set,
-limiting communications during the transport procedure to a group of
-$O(100)$ processors, a number that was shown to have excellent
-parallel efficiencies in Brunner's work and therefore will scale well
-in this algorithm. The overlapping domains within each set also
-demonstrated reduced communication costs. On each processor, the
-source is sampled in the local domain that would exist if no overlap
-was used while tallies can be made over the entire overlapping domain.
-
-To demonstrate this, consider the example adapted from Mervin's work
-with Wagner and others in the same area \cite{mervin_variance_2012}
-and presented in Figure~\ref{fig:msod_example}.
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.5}{
-      \input{chapters/parallel_mc/msod_example.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Overlapping domain example illustrating how domain
-      overlap can reduce communication costs.}
-    \textit{All particles start in the blue region of interest. The
-      dashed line represents 0.5 domain overlap between domains.}}
-  \label{fig:msod_example}
-\end{figure}
-In this example, 3 particle histories are presented emanating from the
-blue region of interest. Starting with particle A, if no domain
-overlap is used then the only the blue domain exists on the starting
-processor. Particle A is then transported through 3 other domains
-before the history ends, therefore requiring three communications to
-occur in Brunner's algorithm. If a 0.5 domain overlap is permitted as
-shown by the dashed line, then the starting process owns enough of the
-domain such that no communications must occur in order to complete the
-particle A transport process. Using 0.5 domain overlap also easily
-eliminates cases such as that represented by the path of particle
-C. In this case, particle C is scattering between two adjacent
-domains, incurring a large latency cost for a single
-particle. Finally, with particle B we observe that 0.5 domain overlap
-will still not eliminate all communications. However, if 1 domain
-overlap were used, the entire geometry shown in
-Figure~\ref{fig:msod_example} would be contained on the source
-processor and therefore transport of all 3 particles without
-communication would occur.
-
-Wagner and colleagues used this methodology for a 2-dimensional
-calculation of a pressurized water reactor core and varied the domain
-overlap from 0 to 3 domain overlap (a $7 \times 7$ box in the context
-of our example) where a domain contained an entire fuel assembly. For
-the fully domain decomposed case, they observed that 76.12\% of all
-source particles leave the domain. At 1.5 domain overlap, the
-percentage of source particles born in the center assembly leaving the
-processor domain dropped to 1.05\% and even further for 0.02\% for the
-3 domain overlap. Based on their results, we hypothesize that the
-overlap approach coupled with the multiple sets paradigm that will
-enhance the scalability of the pure domain-decomposition algorithm for
-Neumann-Ulam presented in the previous section.
-
-\subsection{Generating the MSOD Decomposition }
-\label{subsec:msod_generation}
-In \S~\ref{subsec:domain_generation} we discussed how we generated the
-parallel transport domain from the Neumann-Ulam decomposition of a
-domain decomposed linear operator. We can readily adapt those data
-structures to account for the extra information required by an
-implementation of the MSOD algorithm. First, we consider the
-generation of overlap. Conveniently, this is identical to the
-neighboring states discovery problem discussed in conjunction with
-Figure~\ref{fig:diffusion_graph}. To generate the boundary for the
-local transport domain, we needed to gather all data from the
-immediately adjacent states in the system that did not already exist
-in the local domain. We solved this problem by traversing the graph
-of the matrix for each of the boundary states and determining which
-processes owned those adjacent states.
-
-To generate overlap, we use the identical algorithm but in this case
-we perform as many graph level traversals as the specified amount of
-overlap. For example, consider the analytic relations derived in
-\S~\ref{sec:analytic_framework} for the length of random walks in a
-Neumann-Ulam method and the amount of leakage from a domain
-considering a neutron diffusion problem. From these relations we might
-determine that, for our particular problem, if we can grow the domain
-by 10 additional discrete state transitions, we can reduce the number
-of histories that must be communicated by a significant fraction. We
-determine the information we must gather from neighboring domains to
-build the overlap by traversing the graph of the matrix outward from
-the boundary 10 levels. At each step, we find the data in the adjacent
-domain that we require and make that data local, incrementing the size
-of the overlap. Once the overlap has been gathered, one extra graph
-traversal on the boundary states is performed to get the new
-neighboring states and their owning processes.
-
-Once overlap has been generated, the transport domain for a single set
-is complete. However, if multiple sets are to be used, an additional
-set of parallel procedures is required to generate the necessary data
-structures. Figure~\ref{fig:msod_construction} gives a schematic
-representation of the MSOD construction process using 4 sets. For a
-problem where $P$ parallel processors are available and $S$ sets are
-to be used, each set is allocated $P/S$ processors for computation. On
-the first $P/S$ processors, the linear problem is generated. From the
-linear problem, the linear operator is used to construct the transport
-domain, the solution vector used to construct the tallies, and the
-forcing term vector used to construct the fixed source. Once these
-Monte Carlo data structures have been generated, they are broadcast to
-the remaining sets in the problem as shown in
-Figure~\ref{fig:msod_construction} such that we have $S$ replications
-of the Monte Carlo problem for a single instance of the original
-linear problem.
-
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{0.55}{ \input{chapters/parallel_mc/msod_construction.pdftex_t} }
-  \end{center}
-  \caption{\textbf{MSOD construction for 4 sets with overlap.}
-    \textit{The linear system is used to construct the Monte Carlo
-      transport domain on the primary set. The Monte Carlo data
-      structures are then broadcast among the blocks to all sets in
-      the system.}}
-  \label{fig:msod_construction}
-\end{figure}
-
-\clearpage
-
-\subsection{MSOD Neumann Ulam Algorithm }
-\label{subsec:msod_algorithm}
-
-With the ability to generate the transport domain within an MSOD
-decomposition as well as the source and tally structures, we can now
-define how to perform a Neumann-Ulam solve using MSOD. Given by
-Algorithm~\ref{alg:msod_transport}, we begin by constructing the Monte
-Carlo data structures for each set including the transport domain, the
-source, and the tallies. Next, in line 4 we perform the asynchronous
-transport algorithm presented in \S~\ref{sec:asynchronous_algorithm}
-for each of the individual sets. Adding overlap to
-Algorithm~\ref{alg:parallel_mc_algorithm} in this case simply modifies
-the set of local states and the set of neighboring states to
-additional states gathered during overlap generation.
-
-\begin{algorithm}[h!]
-  \caption{\textbf{MSOD Transport Sequence}}
-  \label{alg:msod_transport}
-  \begin{algorithmic}[1]
-    \State build MSOD domain
-    \State build Monte Carlo source
-    \State build Monte Carlo tally
-    \State perform parallel Neumann-Ulam transport in each set
-    \State combine set tallies
-    \State combine block tallies
-  \end{algorithmic}
-\end{algorithm}
-
-Once Monte Carlo transport is complete complete, the tallies for each
-individual computation must be gathered to the original set in order
-to build the correction vector within an MCSA sequence. To do this, we
-perform two parallel vector operations. First, overlap not only
-creates additional states in the local transport domain but also
-additional states in the local tally vector such that any events that
-occur in the local domain may be tallied in a local vector. Because of
-this, the tally vector in a single set will share local states with
-other domains and a parallel sum reduction operation is required to
-transform the tally into the original Neumann-Ulam parallel
-decomposition. Second, the tallies in individual sets must be combined
-and applied to the primary set in the problem that shares the
-processor space with the original linear problem. For this operation,
-we employ the concept of blocks as a subdivision of parallel space
-complementary to the concept of sets.
-
-Consider the schematic representation of the final parallel reduction
-presented in Figure~\ref{fig:msod_tally_reduction}. For this
-operation, the tally vectors in each set must be summed together and
-applied to the primary set. As each set is an exact replica of the
-others, the parallel decomposition of the tally vector is the same
-within all sets. We can take advantage of this by building a processor
-space that encapsulates each identical domain in each set. We can then
-use this processor space for the reduction. For the schematic in
-Figure~\ref{fig:msod_tally_reduction}, the upper-left domain in each
-set creates a single block, giving 9 total blocks for this example
-problem. The processor space that encapsulates each of the upper-left
-domains is then used for the parallel reduction operation. This
-reduction will happen in a mutually exclusive manner for each of the
-blocks in the system. One benefit of this approach is that this
-reduction occurs only over a subset of the processors in the problem,
-providing better scalability than a reduction operation over all
-processors in the problem.
-
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{0.55}{ \input{chapters/parallel_mc/msod_tally.pdftex_t} }
-  \end{center}
-  \caption{\textbf{MSOD tally reduction over blocks.} \textit{The
-      tally computed independently in each set is reduced across the
-      blocks to the primary set. The linear system vector is updated
-      on the primary set. In this example the tally vector in the
-      block containing upper-left subdomains is combined. Identical
-      reduction operations occur for all other blocks
-      simultaneously.}}
-  \label{fig:msod_tally_reduction}
-\end{figure}
-
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{Parallel MCSA\ }
-\label{sec:parallel_mcsa}
-With the parallel adjoint Neumann-Ulam solver implementation described
-above, the parallel implementation of an MCSA iteration is
-trivial. Recall the MCSA iteration procedure presented again here for
-clarity:
-\begin{subequations}
-  \begin{gather}
-    \label{eq:mcsa_r1}
-    \ve{x}^{k+1/2} = \ve{x}^k + \ve{r}^k\:,\\
-    \label{eq:mcsa_r2}
-    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\
-    \label{eq:mcsa_r3}
-    \ve{A}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:,\\
-    \label{eq:mcsa_r4}
-    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,\\
-    \label{eq:mcsa_r5}
-    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:.
-  \end{gather}
-\end{subequations}
-In \S\ref{sec:parallel_krylov_methods} we discuss parallel matrix and
-vector operations as utilized in conventional projection methods. We
-utilize these here for the parallel MCSA implementation for the
-computations required in Eqs~(\ref{eq:mcsa_r1}), (\ref{eq:mcsa_r2}),
-(\ref{eq:mcsa_r4}), and (\ref{eq:mcsa_r5}). In the first step, a
-parallel matrix-vector multiply is used to apply the split operator to
-the previous iterate's solution. A parallel vector update is then
-performed with the source vector to arrive at the initial iteration
-guess. In the next step, the residual is computed by the same
-operations where now the operator is applied to the solution guess
-with a parallel matrix-vector multiply and then a parallel vector
-update with the source vector is performed. Once the correction is
-computed with a parallel adjoint Neumann-Ulam solve, this correction
-is applied to the guess with a parallel vector update to get the new
-iteration solution. Additionally, as given by
-Eq~(\ref{eq:mcsa_stopping_criteria}), 2 parallel vector reductions
-will be required to check the stopping criteria: one initially to
-compute the infinity norm of the source vector, and another at every
-iteration to compute the infinity norm of the residual vector.
-
-To parallelize the solution of Eq~(\ref{eq:mcsa_r3}), we apply the
-MSOD Neumann-Ulam algorithm presented in the previous section. The
-problem is potentially replicated with overlap and the Monte Carlo
-procedure returns a correction vector, $\delta \ve{x}$, with the same
-parallel decomposition as the input linear problem. As was the case
-for generating the replications required for a multiple set
-implementation, the explicit matrix and vector operations required for
-the rest of an MCSA iteration occur only on the subset of processors
-containing the primary set. Only the Monte Carlo data structures exist
-on the rest of the processors in the problem as replication of the
-other operations does not generate any more information that could be
-used to accelerate the time to solution.
-
-\subsection{Parallel FANM Implementation}
-\label{subsec:parallel_fanm}
-Here we briefly comment on how a FANM method may be implemented in
-parallel. A parallel FANM method relies on a basic set of parallel
-matrix-vector operations outlined in
-\S~\ref{sec:parallel_krylov_methods} as well as the global residual
-and Jacobian assembly procedure described in
-\S~\ref{subsec:automatic_differentiation}. Consider the FANM iteration
-scheme in Algorithm~\ref{alg:fanm}. We must first assemble the linear
-system in parallel through the element-wise function evaluations to
-generate both the global Jacobian operator and the global residual
-vector on the right hand side. Per Bartlett's work on automatic
-differentiation, efficient and automated parallel mechanisms are
-available to do this through a sequence of scatter/gather
-operations. With these tools available for residual and Jacobian
-generation, the remainder of the parallel procedure is simple. The
-linear Newton correction system is solved using the parallel MCSA
-method as described in the previous section and the Newton correction
-applied to the previous iterate's solution through a parallel vector
-update.
-
-%%---------------------------------------------------------------------------%%
-\section{Parallel MCSA Verification\ }
-\label{sec:parallel_verification}
-Before we consider the parallel performance of the MCSA algorithm, we
-must first verify that the algorithm generates the correct solution in
-parallel with respect to reference solutions provided by production
-Krylov methods. To verify correctness, we use solutions to the neutron
-diffusion system presented in
-Appendix~\ref{chap:diffusion_problem}. As the reference computation,
-serial results will be used from the conjugate gradient solver in
-Belos, a linear solvers package in the Trilinos scientific computing
-libraries \cite{heroux_overview_2005}. In identical fashion to the
-verification for the $SP_N$ equations performed in
-\S~\ref{sec:spn_mcsa_verification}, we will solve the neutron
-diffusion problem in parallel using multiple cores on a desktop
-machine. For each parallel solution, then the solution at each grid
-point in the system will be compared to the solution at the same grid
-point in the serial reference case with the minimum and maximum
-absolute differences of these values reported.
-
-GMRES and conjugate gradient solvers from the Belos package were used
-on 4 cores as a means of both assuring the Krylov solvers produce the
-same results on multiple cores and as a means of providing an
-acceptable range of minimum and maximum values in the comparison. For
-MCSA, a range of MSOD parameters will be used such that problems with
-multiple sets and multiple blocks may be tested along with
-overlap. For all MCSA calculations, a single history was used for
-every DOF in the problem to compute the correction vector. For all
-solvers, the flux was converged to a tolerance of \sn{1}{-8} and
-therefore if the comparisons agree within this tolerance the solutions
-will be deemed
-correct. Table~\ref{tab:parallel_verification_parameters} gives the
-parameters for the verification problem and
-Table~\ref{tab:parallel_mcsa_verification} gives the results of the
-calculations. Solver parameters used for MCSA are given in
-Table~\ref{tab:parallel_mcsa_parameters}.
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lc}\hline\hline
-      \multicolumn{1}{l}{Parameter}& 
-      \multicolumn{1}{c}{Value}\\\hline
-      $dx$ & 0.1 \\
-      $dy$ & 0.1 \\
-      $N_x$ & 400 \\
-      $N_y$ & 400 \\
-      $N$ & 160,000 \\
-      $\Sigma_a$ & 5.0 \\
-      $\Sigma_s$ & 1.0 \\
-      Boundary Type & Vacuum \\
-      Linear Solver Tolerance & \sn{1}{-8} \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Parallel MCSA verification model problem parameters.}
-    \textit{The neutron diffusion equation in 2 dimensions is used for
-      the model problem.}}
-  \label{tab:parallel_verification_parameters}
-\end{table}
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lc}\hline\hline
-      \multicolumn{1}{l}{Parameter}& 
-      \multicolumn{1}{c}{Value}\\\hline
-      Histories & 1 per DOF \\
-      Weight Cutoff & \sn{1}{-2} \\
-      Fixed Point Iteration & Richardson \\
-      Estimator & Adjoint Collision \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Parallel MCSA verification solver parameters.}
-    \textit{No relaxation parameters or variance reduction techniques
-      were used with MCSA.}}
-  \label{tab:parallel_mcsa_parameters}
-\end{table}
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lccccll}\hline\hline
-      \multicolumn{1}{l}{Solver}& 
-      \multicolumn{1}{c}{Blocks}& 
-      \multicolumn{1}{c}{Sets}& 
-      \multicolumn{1}{c}{Cores}& 
-      \multicolumn{1}{c}{Overlap}& 
-      \multicolumn{1}{l}{Min Difference}& 
-      \multicolumn{1}{l}{Max Difference}\\\hline
-      Conjugate Gradient & - & - & 4 & - & 0.0 & \sn{1.11022}{-16} \\
-      GMRES & - & - & 4 & - & \sn{5.64215}{-13} & \sn{3.83348}{-9} \\
-      MCSA & 2 & 1 & 2 & 0 & \sn{8.60423}{-16} & \sn{7.6806}{-9} \\
-      MCSA & 3 & 1 & 3 & 0 & \sn{6.41154}{-15} & \sn{7.65397}{-9} \\
-      MCSA & 4 & 1 & 4 & 0 & \sn{2.66454}{-15} & \sn{7.62623}{-9} \\
-      MCSA & 2 & 2 & 4 & 0 & \sn{1.52656}{-15} & \sn{7.57108}{-9} \\
-      MCSA & 1 & 4 & 4 & 0 & \sn{5.55112}{-16} & \sn{7.61631}{-9} \\
-      MCSA & 8 & 1 & 8 & 0 & \sn{6.13398}{-15} & \sn{7.51744}{-9} \\
-      MCSA & 4 & 2 & 8 & 0 & \sn{1.03251}{-14} & \sn{7.62601}{-9} \\
-      MCSA & 2 & 4 & 8 & 0 & \sn{5.13478}{-15} & \sn{7.60777}{-9} \\
-      MCSA & 1 & 8 & 8 & 0 & \sn{1.80411}{-15} & \sn{7.60651}{-9} \\
-      MCSA & 2 & 1 & 2 & 5 & \sn{2.47025}{-15} & \sn{7.58882}{-9} \\
-      MCSA & 3 & 1 & 3 & 5 & \sn{2.44249}{-15} & \sn{7.6326}{-9} \\
-      MCSA & 4 & 1 & 4 & 5 & \sn{7.49401}{-16} & \sn{7.49014}{-9} \\
-      MCSA & 2 & 2 & 4 & 5 & \sn{1.19349}{-15} & \sn{7.61814}{-9} \\
-      MCSA & 1 & 4 & 4 & 5 & \sn{5.55112}{-16} & \sn{7.61631}{-9} \\
-      MCSA & 8 & 1 & 8 & 5 & \sn{2.05391}{-15} & \sn{7.63955}{-9} \\
-      MCSA & 4 & 2 & 8 & 5 & \sn{9.57567}{-15} & \sn{7.61313}{-9} \\
-      MCSA & 2 & 4 & 8 & 5 & \sn{5.7454}{-15} & \sn{7.60767}{-9} \\
-      MCSA & 1 & 8 & 8 & 5 & \sn{1.80411}{-15} & \sn{7.60651}{-9} \\
-      MCSA & 2 & 1 & 2 & 10 & \sn{1.08247}{-15} & \sn{7.63231}{-9} \\
-      MCSA & 3 & 1 & 3 & 10 & \sn{3.27516}{-15} & \sn{7.60211}{-9} \\
-      MCSA & 4 & 1 & 4 & 10 & \sn{4.996}{-16} & \sn{7.64756}{-9} \\
-      MCSA & 2 & 2 & 4 & 10 & \sn{4.13558}{-15} & \sn{7.60259}{-9} \\
-      MCSA & 1 & 4 & 4 & 10 & \sn{5.55112}{-16} & \sn{7.61631}{-9} \\
-      MCSA & 8 & 1 & 8 & 10 & \sn{9.71445}{-16} & \sn{7.58535}{-9} \\
-      MCSA & 4 & 2 & 8 & 10 & \sn{7.77156}{-16} & \sn{7.65978}{-9} \\
-      MCSA & 2 & 4 & 8 & 10 & \sn{4.16334}{-16} & \sn{7.63191}{-9} \\
-      MCSA & 1 & 8 & 8 & 10 & \sn{1.80411}{-15} & \sn{7.60651}{-9} \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Parallel MCSA verification results.}
-    \textit{Difference values given are absolute values. The neutron
-      diffusion equation in 2 dimensions is used for the model problem
-      with the parameters given in
-      Table~\ref{tab:parallel_verification_parameters}. A serial
-      conjugate gradient calculation is used as the reference
-      computation and neutron fluxes are compared
-      element-wise. Parameters used for the MCSA solver are given in
-      Table~\ref{tab:parallel_mcsa_parameters}. All solvers including
-      the reference computation were converged to tolerance of
-      \sn{1}{-8}.}}
-  \label{tab:parallel_mcsa_verification}
-\end{table}
-
-From the data we see that for all solvers the parallel results agree
-with the reference serial results within the solution tolerance for
-all DOF in the system. In addition, we do not notice any systematic
-differences in the results as a function of number of blocks, number
-of sets, or amount of overlap. In addition, the parallel GMRES results
-have maximum absolute differences of the same magnitude as all MCSA
-results and a larger minimum absolute difference than all MCSA results
-by three orders of magnitude. Based on these results, MCSA
-parallelized with the MSOD algorithm, and in particular the
-implementation used here, generates the correct results with respect
-to the serial conjugate gradient reference computation.
-
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{Leadership-Class Parallel Scaling Studies\ }
-\label{sec:leadership_scaling_studies}
-Using the parallel performance metrics presented in
-Appendix~\ref{chap:parallel_metrics}, we will next measure performance
-in various scaling studies using an implementation of the parallel
-MCSA algorithm to assess its quality using the neutron diffusion
-problem presented in Appendix~\ref{chap:diffusion_problem}. For these
-scaling studies, large problems at high levels of concurrency will be
-solved at a leadership-class computing facility. In this section, we
-will outline the parallel machine on which the calculations were
-performed and then present a series of numerical studies that identify
-both strengths and weaknesses of the parallel algorithm. Both pure
-domain decomposition and MSOD parallel performance will be explored
-and compared to performance results for the same problems using a set
-of production parallel Krylov solvers.
-
-For the scaling results presented here, all parallel MCSA results were
-computed using the Monte Carlo Linear Solvers\footnote{MCLS is an open
-  source C++ library built on the Trilinos framework and can be
-  downloaded at \url{https://github.com/sslattery/MCLS}} (MCLS)
-library generated specifically for this work while the conjugate
-gradient and GMRES Krylov solvers used are part of the Belos linear
-solver package within the Trilinos framework
-\cite{heroux_overview_2005}. In general, the Belos package produces
-solutions to the neutron diffusion problem nearly two orders of
-magnitude faster than the MCLS library for serial computations using
-the conjugate gradient solver. Therefore, when we directly compare
-these two codes, the MCLS parallel efficiencies will appear
-artificially inflated relative to the Belos efficiencies due to the
-fact that the MCLS ratio of work to communication will be much larger
-than that for the Belos calculations. For every scaling study
-presented, the calculations were performed 3 times with the average
-values reported for wall time and those average values used for the
-efficiency computations. In addition, it should be noted that in
-practice, symmetric positive-definite problems like the neutron
-diffusion problem solved here would never be solved using
-GMRES. Instead, a method such as conjugate gradient would typically be
-used due to the enhanced performance. We use GMRES here simply as a
-means of providing an additional comparison for MCSA performance.
-
-\subsection{Titan }
-\label{subsec:titan}
-All parallel scaling studies presented in this work were performed on
-the Titan Cray XK7 machine at the Oak Ridge Leadership Computing
-Facility located at Oak Ridge National Laboratory. Titan is a
-heterogeneous architecture machine combining both CPU and GPU
-components to achieve a theoretical peak performance of more than 20
-petaflops. The technical specifications for Titan for only the CPU
-portion of the machine are given in Table~\ref{tab:titan_hardware}.
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lc}\hline\hline
-      Processor & 16 core AMD \\
-      Nodes & 18,688 AMD Opterons \\
-      Cores/Node & 16 \\
-      Total Cores & 299,008 \\
-      Memory/Node & 32 GB \\
-      Memory/Core & 2 GB \\
-      Interconnect & Gemini \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Titan Cray XK7 hardware specifications.}
-    \textit{Specifications presented for Titan in this table were
-      gathered from the Oak Ridge Leadership Computing Facility at}
-    \url{http://www.olcf.ornl.gov/computing-resources/titan-cray-xk7/}}
-  \label{tab:titan_hardware}
-\end{table}
-
-For the scaling studies presented in this work, only the CPU
-components of Titan specified in Table~\ref{tab:titan_hardware} were
-utilized. This is due to the fact that the implementations of the
-parallel MCSA algorithm in the MCLS library utilized code written only
-with the Message Passing Interface (MPI) and therefore could not take
-advantage of the GPU accelerators. Even considering just the CPU
-portion of Titan, nearly 300,000 CPUs are available to test the
-parallel performance of the MCSA algorithm at high levels of
-concurrency.
-
-For each parallel performance metric presented in the previous
-section, a reference case is required to compute efficiencies for all
-other cases. For smaller machines, often a serial computation can be
-performed and used as the base case to get a true measure of speed-up
-for a particular problem. For our scaling studies, we will always use
-as a reference case a calculation which uses 16 cores, filling an
-entire node of the machine. We fill an entire node so that we may
-ignore the effects on scalability measures that arise from using only
-a subset of cores in a node. Typically this results in a large drop in
-efficiency from a serial computation to a calculation that fills the
-entire node. Once a node is filled, adding subsequent nodes has less
-of an effect on the scaling, providing a more stable region of
-analysis.
-
-\subsection{Communication Parameters}
-\label{subsec:comm_parameters}
-
-In our description of Brunner and Brantley's algorithm as applied to
-the Neumann-Ulam method, we noted two free parameters that are
-user-defined at runtime. The first is the static data buffer size for
-communicating histories between neighboring domains and the second is
-the frequency by which a domain checks for incoming information from
-neighboring domain, both of which may potentially affect latency and
-load balancing. Changing the static history buffer size parameter
-affects the scalability of the algorithm in two ways. If the buffer
-size is increased, the buffer takes longer to fill and is therefore
-sent less frequently, potentially causing load balancing bottlenecks
-where a neighboring domain is idle and waiting for more work to do but
-is waiting for another domain to complete work. In addition, if the
-buffer size is too large, the network may become saturated with these
-very large data packets, slowing down performance. If the buffer size
-is too small, too many history messages are sent, not providing enough
-data for a neighboring process to work on to make up for the time
-spent communicating. Considering the second parameter, when the
-message check frequency period is increased, more histories are
-processed locally before looking to do work elsewhere. If this value
-becomes too large, incoming data buffers may not be cleared frequently
-enough and cause decreased performance in the domain-to-domain
-communications. If this value is too small, too much time is spent
-looking for work instead of actually performing it.
-
-To assess how these parameters affect the performance of the parallel
-Monte Carlo algorithm on the Titan machine, a parameter study was
-performed using the neutron diffusion problem on 64 cores (4
-nodes). This problem had a reasonably well conditioned spectral radius
-of 0.78 with a global problem size of \sn{1.6}{7} DOFs with one
-stochastic history used per DOF in the MCSA Neumann-Ulam used to
-compute the correction for the acceleration. Each parameter was varied
-between 16 and 4096 and the time to solve the problem recorded. The
-results of this study are presented in
-Figure~\ref{fig:titan_comm_parameters}. Compared to the results of
-Brunner and Brantley for the same type of study, we note qualitatively
-similar results with the general observation that a larger history
-buffer check period is better than a shorter one. The larger the check
-period, the more on-process work is done before checking
-messages. Second, there was not a significant variation observed in
-the history buffer size parameter for a given check frequency.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_comm_parameters.pdf}
-  \end{center}
-  \caption{\textbf{Neumann Ulam parallel algorithm history buffer
-      check period vs. history buffer size sensitivity.} \textit{The
-      color scale represents the ratio of the runtime of a particular
-      parameter variation with the runtime from the faster parameter
-      variation. A value of 1.02 on the color scale corresponds to a
-      runtime 2\% longer than the faster time. All runtimes were
-      within 6.5\% of the fastest observed case.}}
-  \label{fig:titan_comm_parameters}
-\end{figure}
-
-Compared to the reference work, there was also not a significant
-change in the runtime over the values of parameters tested with all
-runtimes reported within 6.5\% of the fastest case observed. For this
-work, a history buffer size of 4096 and a buffer check period of 128
-performed best for the initial calculations. However, most values in
-the range tested will give good performance. Brunner and Brantley
-found that a value of 1024 was best for both parameters. For the
-results observed here, these parameters were only 0.7\% slower than
-the fastest case. It may be best practice to fix these parameters
-internally at 1024 as both this work and the reference found these to
-perform well several years apart on both different parallel machines
-and for different transport sequences. Therefore, for the scaling
-studies presented in the following sections, values of 1024 for both
-the history buffer size and message check frequency will be used.
-
-\subsection{Pure Domain Decomposition Scaling}
-\label{subsec:pure_domain_decomp}
-
-For the first set of scaling studies, we consider the parallel MCSA
-algorithm where MSOD has not been used in the parallel Neumann-Ulam
-algorithm. We refer to this case as pure domain decomposition where no
-overlap or replication has been leveraged. Such a test will isolate
-specifically the performance of Brunner and Brantley's algorithm as
-applied to the Neumann-Ulam algorithm. For the strong scaling study,
-the global diffusion problem size was fixed at \sn{1.6}{7} DOFs for
-all solvers and the 16 core run used as the base case. MCSA was used
-again with one stochastic history used per DOF in the Neumann-Ulam
-solve to compute the correction for the acceleration with the
-remainder of the problems and solver parameters the same as those for
-the verification calculation given in
-Table~\ref{tab:parallel_verification_parameters} and
-Table~\ref{tab:parallel_mcsa_parameters}. As with all of the scaling
-studies, 3 calculations for all data points were performed with the
-average wall time reported and used for the efficiency
-calculations. Figure~\ref{fig:titan_pure_strong_time} gives the wall
-time per iteration for these calculations while
-Figure~\ref{fig:titan_pure_strong} gives the results of the this study
-with the given absolute efficiencies computed using
-Eq~(\ref{eq:strong_scaling_absolute}). We report wall time per
-iteration due to the fact that MCSA converged on the solution in 22
-iterations while the Krylov solvers converged in 19 iterations.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_pure_strong_time.pdf}
-  \end{center}
-  \caption{\textbf{Pure domain decomposition strong scaling wall time
-      per iteration.} \textit{Wall time is reported per iteration
-      because the methods converged in different numbers of
-      iterations. MCLS is over an order of magnitude slower
-      arithmetically than the Krylov solvers. GMRES executes more
-      operations than conjugate gradient and is therefore slower as
-      well.}}
-  \label{fig:titan_pure_strong_time}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_pure_strong.pdf}
-  \end{center}
-  \caption{\textbf{Pure domain decomposition strong scaling absolute
-      efficiency.}  \textit{Super-linear speed-up is from memory
-      thrashing in the base case. MCLS is over an order of magnitude
-      slower arithmetically, causing the higher efficiency. GMRES
-      executes more operations than conjugate gradient, also creating
-      a higher efficiency.}}
-  \label{fig:titan_pure_strong}
-\end{figure}
-
-There are several key features of
-Figure~\ref{fig:titan_pure_strong_time} and
-Figure~\ref{fig:titan_pure_strong} worth noting. First, up to around
-5,000 cores, all methods experience a super-linear improvement in
-parallel efficiencies. For all methods, this increase is not a
-function of the algorithmic efficiency as those computed using
-Eq~(\ref{eq:algorithmic_efficiency}) remained at unity, meaning that
-all computations converged in the same number of iterations relative
-to the base case\footnote{When the algorithmic efficiency remains at
-  unity for all calculations, the implementation efficiency is
-  equivalent to the absolute efficiency and therefore is not presented
-  here.}. Instead, what is happening here and was observed by Brunner
-and Brantley in their work is that the large problem used to scale to
-O(10,000) cores was large enough that memory thrashing was occurring
-in the 16 core base case. As the number of cores in the strong scaling
-study was increased and the local problem size decreased, more of the
-local problem fit into the cache of the system, thus causing improved
-serial performance and therefore enhanced runtimes. In this region of
-super-linear efficiencies, we also observe larger efficiencies for
-GMRES as compared to conjugate gradient. This occurs due to the fact
-that conjugate gradient has a static set of operations performed at
-every iteration while GMRES operates on a continually growing
-subspace. In addition, the orthogonalization operations required to
-generate that subspace consume much more time than the basic conjugate
-gradient operations, causing a larger serial runtime and subsequently
-higher efficiencies. For the same reasons, we also observe higher
-efficiencies for the MCSA implementation due to the serial performance
-of the implementation.
-
-Beyond around 5,000 cores, the global problem size selected for this
-strong scaling study causes the local problem size to shrink enough
-that parallel communication costs begin to overtake arithmetic costs
-on process. This takeover causes this effective strong scaling wall
-that is observed in all methods. Using any more cores to solve this
-particular global problem is not an efficient use of those parallel
-resources. Furthermore, the fact that we see a very similar trends in
-efficiencies for all methods means that they are bound by the same
-parallel performance limitations in a strong scaling environment. This
-is important as we can expect the same qualitative scaling performance
-from MCSA if arithmetic optimization has been performed to improve the
-serial performance of the implementation. In addition, these results
-are valuable in that they show MCSA, when properly parallelized, has
-the ability to be competitive on leadership-class computing platforms
-with conventional subspace methods with good strong scaling observed
-in the pure domain decomposition case to O(10,000) cores.
-
-Next, we consider the weak scaling performance of the algorithm. In
-this case, we fix the local problem size at \sn{4}{4} DOFs and
-increase the number of processors used in the
-simulation. Figure~\ref{fig:titan_pure_weak_time} gives the wall
-time per iteration and Figure~\ref{fig:titan_weak_absolute} gives the
-weak scaling absolute efficiencies for this problem as computed by
-Eq~(\ref{eq:weak_scaling_absolute}). In general, the relative trends
-when comparing MCSA to the Krylov methods are the same with MCSA and
-GMRES scaling better than the conjugate gradient method due to the
-increased amount of arithmetic work on-process. In addition, it is
-important to note that the weak scaling of MCSA when properly
-parallelized remains competitive with the Krylov methods keeping in
-mind the artificial inflation of efficiencies.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_pure_weak_time.pdf}
-  \end{center}
-  \caption{\textbf{Pure domain decomposition weak scaling wall time
-      per iteration.} \textit{Wall time is reported per iteration
-      because the methods converged in different numbers of
-      iterations. MCLS is over an order of magnitude slower
-      arithmetically than the Krylov solvers. GMRES executes more
-      operations than conjugate gradient and is therefore slower as
-      well.}}
-  \label{fig:titan_pure_weak_time}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_absolute.pdf}
-  \end{center}
-  \caption{\textbf{Pure domain decomposition weak scaling absolute
-      efficiency.} \textit{MCLS is an over order of magnitude slower
-      arithmetically, causing the higher efficiency. GMRES executes
-      more operations than conjugate gradient, also creating a higher
-      efficiency.}}
-  \label{fig:titan_weak_absolute}
-\end{figure}
-
-As an additional means of comparison between MCSA and the Krylov
-methods, we compute the algorithmic efficiencies of the calculations
-using Eq~(\ref{eq:algorithmic_efficiency}). As shown in
-Figure~\ref{fig:titan_weak_algorithmic}, the algorithm efficiency of
-the Krylov algorithm improves as a function of the number of cores in
-the problem due to the fact that the number of iterations required to
-converge decreased from 20 in the base case to 19 and then to 18 for
-the final set of weak scaling computations. When this reduction in
-iterations is used to account for the observed runtimes and computed
-absolute efficiencies for weak scaling we end up with the
-implementation efficiencies computed using
-Eq~(\ref{eq:implementation_efficiency}) given in
-Figure~\ref{fig:titan_weak_implementation}. Because MCSA maintains the
-same iterative performance as a function of global problem size in the
-weak scaling study, the implementation efficiencies compare more
-favorably to the Krylov methods. Using the implementation efficiency,
-we are effectively measuring the parallel performance of a single
-iteration for each method by looking at these implementation
-efficiencies. As the global problem size increases, MCSA is less prone
-to a drop in parallel efficiency. If optimization has been performed
-and runtimes potentially approach those of the Krylov methods, this
-comparison may not be as favorable. Overall, weak scaling performance
-of the parallel MCSA algorithm is excellent for the pure domain
-decomposition case with absolute parallel efficiencies of over 80\%
-observed at over 32,000 cores.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_alg_eff.pdf}
-  \end{center}
-  \caption{\textbf{Pure domain decomposition weak scaling algorithmic
-      efficiency.} \textit{As the global problem size was increased in
-      the weak scaling study, both conjugate gradient and GMRES
-      converged in fewer iterations, causing the rise in algorithmic
-      efficiency.}}
-  \label{fig:titan_weak_algorithmic}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_implementation.pdf}
-  \end{center}
-  \caption{\textbf{Pure domain decomposition weak scaling
-      implementation efficiency.} \textit{Considering the scalability
-      of a single iteration of the implementation, the inflated
-      efficiencies of the MCLS MCSA implementation become more
-      obvious. Implementation optimization will likely reduce the MCSA
-      values to a level more comparable to the Belos values.}}
-  \label{fig:titan_weak_implementation}
-\end{figure}
-
-In addition to comparing MCSA scaling to conventional methods, we also
-motivate avoiding any type of master/slave algorithms at these high
-levels of concurrency by comparing the binary communication tree
-implementation used in the previous results with a parallel MCSA
-implementation that uses the master/slave scheme to determine the end
-of a history transport stage as shown in
-Figure~\ref{fig:master_comm_tree}
-instead. Figure~\ref{fig:titan_strong_bvsm} provides this comparison
-using the strong scaling study while Figure~\ref{fig:titan_weak_bvsm}
-gives the results of the comparison for the weak scaling study. In
-both cases, the deficiencies caused by using the master/slave scheme
-for transport completion are obvious. At around 1,000 cores, both the
-strong and weak scaling performance seriously degrades while the
-binary tree pattern maintains excellent performance. At this number of
-cores, any extra work given to the master process, even if it amounts
-to summing a handful of integers and checking for a few extra messages
-as in this algorithm, grows enough to create serious load imbalance
-and reduce performance.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_bvsm.pdf}
-  \end{center}
-  \caption{\textbf{Binary tree vs. master/slave communication scheme
-      strong scaling absolute efficiency.} \textit{The master/slave
-      scheme will not scale beyond $O(1,000)$ cores due to load
-      imbalance.}}
-  \label{fig:titan_strong_bvsm}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_bvsm.pdf}
-  \end{center}
-  \caption{\textbf{Binary tree vs. master/slave communication scheme
-      weak scaling absolute efficiency.}  \textit{The master/slave
-      scheme will not scale beyond $O(1,000)$ cores due to load
-      imbalance.}}
-  \label{fig:titan_weak_bvsm}
-\end{figure}
-
-\clearpage
-
-\subsection{Overlapping Domain Decomposition}
-\label{subsec:overlapping_domain_decomp}
-
-In the next set of scaling studies, we will explore how adding overlap
-to the subdomains in the problem affects runtimes and parallel
-performance as compared to the case with pure domain decomposition. To
-do this, we utilize the set of analytic relations developed in
-\S~\ref{sec:analytic_framework} to design these numerical
-experiments. For the neutron diffusion problem used in these scaling
-studies, the linear operator had a spectral radius of 0.787 and a
-weight cutoff of \sn{1}{-2} was used with the Neumann-Ulam
-solver. Using Eq~(\ref{eq:analytic_k}) gives an expected random walk
-length of 19.22 transitions for any given history in the
-problem. Using the mean-chord approximation given by
-Eq~(\ref{eq:mean_chord_domain_leakage}), we then expect approximately
-5.3\% of the histories born in a particular domain to leak out after
-the first transport stage. By adding overlap, we aim to reduce this
-fraction of communicated histories and therefore boost the parallel
-performance of the algorithm.
-
-To determine the amount of overlap to test for the scaling studies, we
-use a variation of Eq~(\ref{eq:step_k_length}) to determine how many
-discrete states a history will move on average from its birth site. We
-modify Eq~(\ref{eq:step_k_length}) to only account for discrete
-movements:
-\begin{equation}
-  \sqrt{\bar{n^2_k}} = n_s \sqrt{k}\:,
-  \label{eq:discrete_distance}
-\end{equation}
-where $\sqrt{\bar{n^2_k}}$ is now the root-mean-squared number of
-discrete states a history will move from its birth site. For our
-diffusion problem used in these scaling studies, we compute a value of
-2.63 states per history for any of the dimensions of the system. This
-means that we can expect to end up on average 2.63 states in both the
-$i$ and $j$ dimensions away from the starting state. Therefore, we
-will use the overlap generation algorithm outlined in
-\S~\ref{subsec:msod_generation} such that we parametrically grow the
-local domain using 1, 2, 3, and 5 levels to measure how scalability is
-affected when this root-mean-squared number of states is
-considered. For example, using 3 levels of overlap should keep over
-50\% of the histories that would be communicated in the pure domain
-decomposition case on-process as this is greater than the
-root-mean-squared number of states we would expect a history to
-travel.
-
-For the performance metrics, we may use the same weak and strong
-scaling efficiency metrics that we used for the pure domain
-decomposition case. As we are comparing to the case without overlap
-and are looking to boost performance relative to the case without
-overlap, for both weak and strong scaling calculations we will use the
-16-core case without overlap as the reference computation for all
-efficiency measures reported.
-
-The absolute efficiency results from the strong scaling exercise are
-given in Figure~\ref{fig:titan_strong_overlap}. Aside from the
-boosting of efficiencies at 2,048 cores, little evidence of strong
-scaling improvement is evident for all levels of overlap tested. To
-further explore any potential benefits of adding overlap into the
-system, Figure~\ref{fig:titan_strong_overlap_diff} gives the
-difference in the efficiencies between all overlap cases and the
-reference case without overlap at each core count. At smaller numbers
-of cores, there is no observed benefit to using overlap to improve
-strong scaling. The improvements noted at 2,048 cores are also present
-here and account for the best efficiency improvements for all core
-counts tested. at 16,384 cores, using 2 levels of overlap did give a
-6.4\% boost in efficiency while at 1,024 cores this same amount of
-overlap actually reduced the efficiency. Beyond 2 levels of overlap,
-efficiency levels begin to fall with 5 levels of overlap performing
-the worst for all cases. From this, we don't expect adding any more
-overlap will improve the scalability of the algorithm and the best
-strong scaling performance was observed when the number of overlap
-levels is approximately equal to the root-mean-squared number of
-states we expect a history to move from its starting point as computed
-by Eq~(\ref{eq:discrete_distance}). We will defer an explanation of
-the drop in efficiencies for levels of overlap beyond this value until
-we have analyzed the weak scaling data.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_overlap.pdf}
-  \end{center}
-  \caption{\textbf{Strong scaling absolute efficiency with varying
-      levels of overlap.} \textit{Values of 0, 1, 2, 3, and 5 levels
-      of overlap were used as the problem had a root-mean-squared
-      number of states of 2.63 for each history.}}
-  \label{fig:titan_strong_overlap}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_overlap_diff.pdf}
-  \end{center}
-  \caption{\textbf{Strong scaling absolute efficiency difference with
-      respect to the 0 overlap base case.} \textit{Values of 1, 2, 3,
-      and 5 levels of overlap are compared here with the absolute
-      difference of the absolute efficiencies presented.}}
-  \label{fig:titan_strong_overlap_diff}
-\end{figure}
-
-For weak scaling, the same study as in the pure domain decomposition
-case was performed again with 1, 2, 3, and 5 levels of overlap. The
-absolute parallel efficiencies from these calculations are given in
-Figure~\ref{fig:titan_weak_overlap} and the differences in
-efficiencies from the calculations without overlap. The results here
-are much the same in that there is not a dramatic improvement in
-scalability relative to the case without overlap. Opposite the strong
-scaling calculation, overlap enhances weak scaling efficiencies at
-lower core counts and degrades the efficiencies at higher core
-counts. In addition, 1 level of overlap provided the best enhancement
-of scalability when overlap was beneficial.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_overlap.pdf}
-  \end{center}
-  \caption{\textbf{Weak scaling absolute efficiency with varying
-      levels of overlap.} \textit{Values of 0, 1, 2, 3, and 5 levels
-      of overlap were used as the problem had a root-mean-squared
-      number of states of 2.63 for each history.}}
-  \label{fig:titan_weak_overlap}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_overlap_diff.pdf}
-  \end{center}
-  \caption{\textbf{Weak scaling absolute efficiency difference with
-      respect to the 0 overlap base case.} \textit{Values of 1, 2, 3,
-      and 5 levels of overlap are compared here with the absolute
-      difference of the absolute efficiencies presented.}}
-  \label{fig:titan_weak_overlap_diff}
-\end{figure}
-
-For both the strong and weak scaling studies, the parallel performance
-of the algorithm was either marginally improved or not improved at
-all. We would always expect an improvement in performance for using
-overlap to prevent the transport of histories from domain to domain
-during execution of the Neumann-Ulam algorithm. The reduction in
-transport means a reduction in parallel communication and therefore
-should never degrade the performance of the algorithm. It is true that
-doing so in fact enhances Monte Carlo performance. However, the
-Neumann-Ulam algorithm is not representative of the entire MCSA
-algorithm. When overlap is used, an overlapping tally vector is
-constructed in parallel such that multiple cores may own multiple
-states in the system. When this occurs, the tally data associated with
-these shared states must be communicated after the Neumann-Ulam solve
-as outlined in \S~\ref{subsec:msod_algorithm}. This reduction
-operation then builds the full MCSA correction vector in the
-decomposition of the solution vector.
-
-Therefore, by building overlap we are in fact reducing parallel
-communication during the Neumann-Ulam solve but we are simply
-deferring that communication until after Monte Carlo transport when
-the overlapping tally vector must be reduced. When overlap becomes
-large enough (i.e. larger than the root-mean-squared number of
-discrete states a history will travel), then the savings in
-communication during the Neumann-Ulam solve start to become smaller
-than the communication created in the overlapping tally vector
-reduction. We see this in both the strong and weak scaling cases at
-larger levels of overlap. For the strong scaling case, we see better
-performance improvement at higher core counts because the total number
-of overlapping states on each core is smaller from the local problem
-size shrinking. For weak scaling, we see better performance
-improvement at lower numbers of cores because although the local
-overlap size is fixed, the global amount of overlap is growing with
-core count and therefore reducing the performance of the tally
-reduction operation.
-
-Based on this data, a few levels of overlap within the expected
-distance of travel for a history may offer some marginal improvement
-in scalability, however, it is not as effective as one might think
-given its improvement to reactor physics calculations as observed in
-the literature. This is largely due to the density of states in the
-system that creates the described performance bottlenecks in the
-overlapping tally vector reduction operation. In a Monte Carlo
-particle simulation, these overlapping tally reduction operations are
-not expected to be performed at the speed they are required in
-MCSA. For example, in a criticality calculation with overlapping
-domains, we would expect to perform as many parallel tally vector
-reduction in a particle transport simulation as there are particle
-stages required for convergence of the eigenvalue. However, these
-particle stages will likely be much more costly in terms of wall time
-relative to a Neumann-Ulam solve in an MCSA iteration due to the more
-extensive particle physics and logic structure that must be
-evaluated. Therefore, we can expect this overlapping tally reduction
-to be less significant to scalability in those operations. We do not
-have as complicated a logic structure in a history random walk for the
-Neumann-Ulam method and therefore overlap can inhibit scalability if
-too large.
-
-\subsection{Multiple Set Domain Decomposition}
-\label{subsec:ms_decomposition}
-
-We next look to apply replication to the problem in the form of
-multiple sets. As the previous analysis showed overlap to be of little
-assistance in improving scalability, we only apply multiple sets to
-the problem with pure domain decomposition within the sets for these
-scaling studies. Before performing parallel scaling studies using
-multiple sets, several aspects of the replicated parallel problem must
-be considered in order to effectively craft the studies and interpret
-their results. First is the question of how many stochastic histories
-to use to generate the correction vector in the MCSA iteration. For
-this work we consider two choices; the split case and the fully
-replicated case. As an example of these cases, consider a single set
-problem where 20,000 histories are used in the Monte Carlo solve. If
-we choose the split case, then solving the same problem with 2 sets
-would give 10,000 histories in each set and 4 sets would give 5,000
-histories in each set. Splitting the histories in this way provides
-the same global number of histories used to compute the correction in
-the single set problem. We expect to reduce the time to solution by
-doing this as each individual set solve will occur faster in an almost
-linear fashion with the reduced number of histories. For the
-replicated history case, each additional set in the problem will use
-just as many histories as the single set case. Using 2 sets with fully
-replicated histories for our example problem then has each set
-computing 20,000 histories with 40,000 global histories in the
-computation. For the 4 set case, the number of global histories is
-increased to 80,000. We also expect this type of replication to reduce
-the time to solution as the extra global histories should reduce the
-stochastic uncertainty in the correction vector and reduce the overall
-number of MCSA iterations required to converge.
-
-\subsubsection{Strong Scaling with Multiple Sets}
-\label{subsubsec:ms_strong}
-
-For the strong scaling study, we apply the additional sets as
-follows. Using the same problems as the pure domain decomposition
-scaling study, the multiple sets are used to replicate those problems
-and boost the core count. For example, for the 4 set case, the data
-point at 64 cores has the same $N/P$ ratio in one set as the single
-set problem with 16 cores. This is due to the fact that $N$ is fixed
-for the strong scaling study and 4 replications gives a set size of 16
-cores. For the 2 set case, the 32 core data point is solving the same
-$N/P$ problem in a set as the 16 core base case.
-
-In addition to considering the replicated domain, we must also
-consider how to modify the efficiencies we will compute. For the
-strong scaling case, recall that the objective is to decrease the time
-to solution for a given global problem size $N$ by increasing
-$P$. Therefore, we may use Eqs~(\ref{eq:strong_scaling_absolute_ref}),
-(\ref{eq:algorithmic_efficiency}), and
-(\ref{eq:implementation_efficiency}) to compute the various strong
-scaling efficiencies, but we must modify the reference computation in
-each of these formulas to reflect our objective. For all efficiency
-computations, the 16 core single set computation will be used as the
-reference case. By adding more cores to the problem, either by
-replicating with multiple sets, or decreasing the local problem size,
-any computational resources added to the problem should aim to result
-in speedup of this reference case. If this speedup for one technique
-of adding cores is less than another, then that technique is deemed a
-less efficient use of those computational resources.
-
-For the multiple sets problem, the single set strong scaling study
-from the pure domain decomposition base case was repeated for multiple
-sets with both splitting of histories across sets and full replication
-of histories across sets using both 2 and 4
-sets. Figure~\ref{fig:titan_strong_ms_time} gives the actual wall
-times for these computations. There are two important features to note
-in the wall time data. First, splitting the histories among sets
-instead of replicating them is faster due to the fact that cutting the
-number of histories in half reduces the time of an iteration by a
-factor that is nearly linear while increasing linearly the number of
-histories through replication does not decrease the iterations
-required to converge by a linear factor. Second, until the strong
-scaling wall is hit starting at around 10,000 cores, the single set
-case actually performs the fastest with only marginal improvement in
-runtimes at higher numbers of cores with more sets. At these larger
-numbers of cores, however, the improvements are so modest that one is
-better off running the same problem using a single set at lower core
-counts to conserve computational resources while maintaining a good
-time to solution.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_ms_time.pdf}
-  \end{center}
-  \caption{\textbf{Wall time in seconds to solution for each case for
-      the strong scaling study with multiple sets.} \textit{Until the
-      strong scaling wall is reached at O(10,000) cores, the single
-      set case performs the best. For the multiple set cases, it is
-      faster to split histories among sets and maintain iterative
-      performance rather than replicate histories to improve iterative
-      performance.}}
-  \label{fig:titan_strong_ms_time}
-\end{figure}
-
-From this timing data we then compute the absolute efficiencies for
-strong scaling as given in Figure~\ref{fig:titan_strong_ms_eff} using
-the 16-core single set calculation as the reference case. As observed
-in the timing data, because the single set case had the fastest run
-times up to the strong scaling wall it also has the best observed
-efficiencies. For the 2 set case with history splitting, although the
-time to perform an iteration was nearly cut in half, the extra
-parallel overhead associated with reducing the tally across the blocks
-to combine the set results increases run times and reduces
-efficiencies. This reduction across blocks is even more obvious for
-the 4 set case. For the replicated cases, we do observe a flat region
-of parallel efficiencies at larger core counts (up to around 32,000
-for the 4 set case with history replication) showing improved scaling
-within context of not considering the single set
-calculation. Unfortunately, the additional cores do not reduce the
-runtime of the global problem as efficiently as expected. Ultimately,
-all cases effectively hit the strong scaling wall at the same time at
-around O(10,000) cores.
-
-We expect this if we consider the fact that adding sets to the problem
-is in itself effectively a form of strong scaling. For any given
-problem of fixed size $N$, adding extra sets to the computation
-decreases the global ratio of $N/P$ as $P$ will grow with the addition
-of sets (although $N/P$ is constant within a set). For the split
-history case, the splitting further exacerbates this situation by
-further decreasing the amount of histories that will be computed in
-the local domain, giving a larger ratio of communication to work in
-the Monte Carlo solve. As with with single set results, we also
-observe some efficiencies greater than unity for the multiple set
-cases also due to improved cache utilization.
-
-Although not as dramatic as one might originally expect from
-introducing replication, there are some modest gains in efficiencies
-at high core counts by using multiple sets. Of interest here is where
-the multiple set cases perform better than the single set case after
-the single set case has hit the strong scaling wall. In
-Figure~\ref{fig:titan_strong_ms_eff}, at 16,384 cores the single set
-case operates at 38\% efficiency while at the same number of cores,
-the 2 set case with history replication operates at 58\%
-efficiency. This is reasonable improvement in the use of those
-computational resources, gained by replicating the problem. Beyond
-16,384 cores, there is not a compelling reason to use more
-computational resources for the this problem.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_ms_eff.pdf}
-  \end{center}
-  \caption{\textbf{Multiple set absolute parallel efficiency.}
-    \textit{Computed relative to the 16-core 1-set base case. Using a
-      single set is the best means by which to improve strong scaling
-      up to the strong scaling wall. At higher core counts past where
-      the single set case hits the strong scaling wall adding sets can
-      provide some improvements in efficiencies.}}
-  \label{fig:titan_strong_ms_eff}
-\end{figure}
-
-Also of interest here is the effect of splitting histories among sets
-versus replicating them on the iterative performance of the
-method. Table~\ref{tab:ms_strong_alg_eff} gives the iterations
-required to converge and computed algorithmic efficiencies for each of
-the cases presented in the strong scaling data. As expected, splitting
-histories maintains a fixed global number of histories and therefore
-maintained the number of iterations required to converge relative to
-the single set base case. When histories were replicated, adding sets
-decreased the number of iterations required to converge in a nonlinear
-fashion. Therefore, increasing the number of histories per MCSA
-iteration improves algorithmic efficiency. Using these algorithmic
-efficiencies, the implementation efficiencies for the multiple set
-cases were computed and are given in
-Figure~\ref{fig:titan_strong_ms_impeff}. From these results, we see
-that the replicated history cases are even less efficient than in the
-absolute case with respect to the performance of a single
-iteration. Not only do we have the additional parallel overhead of
-doing the block level reduction, but we also have additional histories
-that increase the time to solution for a single iteration.
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lcc}\hline\hline
-      \multicolumn{1}{l}{Case}& 
-      \multicolumn{1}{c}{Iterations}&
-      \multicolumn{1}{c}{$\eta_{alg}$} \\\hline
-      1 & 22 & 1.0 \\
-      2-split & 22 & 1.0 \\
-      2-replicated & 16 & 1.375 \\
-      4-split & 22 & 1.0 \\
-      4-replicated & 13 & 1.692 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Strong scaling algorithmic efficiency using
-      multiple sets.} \textit{Replicating histories improves
-      algorithmic efficiency by converging in fewer iterations.}}
-  \label{tab:ms_strong_alg_eff}
-\end{table}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=5in]{chapters/parallel_mc/titan_strong_ms_impeff.pdf}
-  \end{center}
-  \caption{\textbf{Multiple set strong scaling implementation
-      efficiency.} \textit{Computed relative to the 16-core 1-set base
-      case. Considering iterations required to converge, the
-      performance of the split history cases at much better than the
-      cases with history replication.}}
-  \label{fig:titan_strong_ms_impeff}
-\end{figure}
-
-\clearpage
-
-\subsubsection{Weak Scaling with Multiple Sets}
-\label{subsubsec:ms_weak}
-
-Next, we continue to a weak scaling study of the multiple sets
-problem. In the same way as the strong scaling study, we will modify
-the pure domain decomposition study to account for the additional
-replication. For each calculation in the single set study, that
-problem will be replicated either 2 or 4 times with both split and
-replicated histories to assess the effects on scalability and
-runtime. Figure~\ref{fig:titan_weak_ms_time} gives the wall times from
-the multiple sets weak scaling study. Of primary importance here is
-the fact that adding sets to the problem with both split and
-replicated histories actually decreases the time to solution. This was
-not observed in the strong scaling case until much larger core counts
-as the multiple set computations hit the strong scaling wall after the
-single set computation.
-
-This result is significant in that it shows a solution technique in
-which the linear problem can actually be replicated and enhance time
-to solution, something that cannot be achieved with conventional
-methods. At larger core counts, however, this reduction in compute
-time is not as large as at lower core counts due to the fact that the
-tally reduction across blocks requires more resources and those
-resources are clearly growing with core count (most obviously in the 4
-set case with split histories). In addition, at 131,072 cores for the
-largest 4 set computation, load balance is likely becoming an issue
-due to the linear system and subsequent parallel matrix and vector
-operations only occurring on a subset of the entire processor space of
-the problem while the Monte Carlo calculation is occurring on all
-processors.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_ms_time.pdf}
-  \end{center}
-  \caption{\textbf{Wall time in seconds to solution for each case for
-      the weak scaling study with multiple sets.}}
-  \label{fig:titan_weak_ms_time}
-\end{figure}
-
-For a multiple set analysis of the weak scaling timing results, we
-again have to consider that adding sets to the problem is actually a
-form of strong scaling. Consider a single set problem ($S=1$) for a
-given $P$ and $N$. When that problem is now solved using 2 sets we
-have $S=2$, $N$ and $2P$. For any number of sets we then have a global
-problem size of $N$ while the number of processors is $SP$ where $P$
-is the number of processors in the single set case. Therefore, $N/P$
-is no longer fixed in the weak scaling study but rather $N/(SP)$ is
-fixed. As a result, we modify the weak scaling absolute efficiency
-relationship to account for this fact:
-\begin{equation}
-\eta_{weak}(M|N,P|Q,S) = \frac{T(M,Q)}{S \times T(N,P)}\:,
-  \label{eq:ms_weak_efficiency}
-\end{equation}
-where now the absolute efficiency is a function of the number of sets
-in the problem and again $M/Q = N/P$. As with the strong scaling
-study, we will use the 16-core single set case as the reference
-computation for efficiencies. Interestingly,
-Eq~(\ref{eq:ms_weak_efficiency}) has a very similar form to the strong
-scaling efficiency given by Eq~(\ref{eq:strong_scaling_absolute}),
-representative of the strong scaling nature of adding additional
-sets. For problems where adding sets improves time to solution, this
-new weak scaling formulation accounts for how efficiently those extra
-resources are incorporated into the problem. Without such a
-modification, the largest case run for the split 4 set case would have
-a computed weak scaling efficiency of nearly 100\% using the standard
-formula for absolute efficiency. However, using 4 times as many cores
-as the single set case increased the runtime for a fixed global
-problem size and therefore we would actually expect to compute a very
-low efficiency that shows this very poor use of resources.
-
-Figure~\ref{fig:titan_weak_ms_eff} gives the results of the weak
-scaling efficiency computations using the multiple set metric given by
-Eq~(\ref{eq:ms_weak_efficiency}). Although we do reduce the time to
-solution over the base cases for nearly all data points, the
-observation here is that the reduction in time to solution is still
-not large enough to efficiently use all resources. For example, for
-the 2 set case to be a perfect use of resources with either split or
-replicated histories, the runtime of that calculation must be half of
-that for the single set case. For the 4 set case, this run time should
-be a quarter of the single set case for perfect scaling. This was not
-the observation for the runtimes in
-Figure~\ref{fig:titan_weak_ms_time} and therefore the single set case
-is still the most efficient use of computational resources in this
-scaling study.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_ms_eff.pdf}
-  \end{center}
-  \caption{\textbf{Weak scaling absolute efficiency relative to
-      16-core 1-set base case.}}
-  \label{fig:titan_weak_ms_eff}
-\end{figure}
-
-As in the strong scaling case, splitting and replicating histories
-across sets had effectively the same results for the weak scaling
-study with the iterations to converge and computed algorithmic
-efficiencies given by Table~\ref{tab:ms_weak_alg_eff}. Again, when the
-implementation efficiencies are computed, we see in
-Figure~\ref{fig:titan_weak_ms_impeff} that splitting histories among
-sets is still a more efficient manner of introducing multiple sets
-rather than replicating histories in an attempt to reduce the number
-of iterations required to converge.  Here we also note that this is
-not a true weak scaling study as indicated by
-Eq~(\ref{eq:ms_weak_efficiency}) but rather an odd combination of both
-weak and strong scaling. Regardless of this, although multiple sets
-are not the most efficient use of resources in this analysis, they do
-enhance the time to solution through replication and do so
-significantly at core counts of O(1,000) and lower. In both the weak
-and strong scaling cases, due to the additional computational and
-communication overhead of adding sets to the system, one should never
-expect to seriously boost parallel efficiencies for MCSA using this
-technique except for certain instances past the strong scaling wall in
-a purely strong scaling environment. In a weak scaling environment,
-although no efficiency increase will be noted, one can expect an
-enhancement of time to convergence by replicating.
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{lcc}\hline\hline
-      \multicolumn{1}{l}{Case}& 
-      \multicolumn{1}{c}{Iterations}&
-      \multicolumn{1}{c}{$\eta_{alg}$} \\\hline
-      1 & 22 & 1.0 \\
-      2-split & 22 & 1.0 \\
-      2-replicated & 16 & 1.375 \\
-      4-split & 22 & 1.0 \\
-      4-replicated & 13 & 1.692 \\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Weak scaling algorithmic efficiency using multiple
-      sets.} \textit{Replicating histories improves algorithmic
-      efficiency by converging in fewer iterations.}}
-  \label{tab:ms_weak_alg_eff}
-\end{table}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_ms_impeff.pdf}
-  \end{center}
-  \caption{\textbf{Implementation parallel efficiency relative to 16-core
-      1-set base case.}}
-  \label{fig:titan_weak_ms_impeff}
-\end{figure}
-
-\clearpage
-
-\subsection{Subdomain Neumann-Ulam}
-\label{subsec:full_clip}
-
-Based on the analysis of the pure domain decomposed MCSA algorithm and
-subsequently the multiple sets and overlapping domain implementation,
-one may ask then ask if there is any need to communicate histories
-from domain-to-domain at all. Overlap was not as successful because
-because the additional communication required to reduce the
-overlapping tally vector eventually becomes larger than the
-communication savings during Monte Carlo transport. In addition, for
-local domains that are large enough to get good performance
-(e.g. those used in the weak scaling studies), then our analytic
-relationships for domain leakage, and in particular
-Eq~(\ref{eq:mean_chord_domain_leakage}), tell us that only about 5\%
-of the histories in this diffusion problem will leave the domain in
-the first place.  If that is true, then perhaps without any overlap or
-communication, just doing local Monte Carlo transport such that
-histories are terminated upon reaching the domain boundary may be
-satisfactory for accelerating the MCSA iteration. In this case,
-histories near the boundaries of the domain may be truncated before
-reaching the weight cutoff if they reach the domain boundary
-first. This truncation has the potential to increase the stochastic
-uncertainty of the Neumann-Ulam result and therefore increase the
-number of MCSA iterations required to converge. In addition, the
-scalability of the method will then be purely limited by the parallel
-matrix and vector operations required to implement the remainder of
-the algorithm on the subset of processors containing the linear
-problem.
-
-Considering using MCSA in this way such that the Monte Carlo solve
-occurs only in the subdomains casts the method as more of a true
-domain decomposition as outlined by Chapter 14 of Saad's text on
-iterative methods \cite{saad_iterative_2003}. Compared to traditional
-domain decomposition methods we see many similarities including the
-potential to overlap the subdomains to improve the results. By
-building the boundary data as outlined in
-\S~\ref{subsec:domain_generation}, we are effectively gathering one
-level of the adjacent domains, albeit in this case to satisfy the data
-requirements of the Monte Carlo estimators. Furthermore, the fact that
-we are using residual Monte Carlo to compute the correction vector
-casts MCSA as a stochastic realization of an additive Schwarz method
-with a fixed point iteration used as a relaxation step\footnote{See
-  section 14.3.3 in \cite{saad_iterative_2003} for a full definition
-  of the additive Schwarz method.}. Both the fixed point iteration and
-the computation of the residual propagate the solution from individual
-subdomains to the others through the matrix-vector multiply operation
-after all subdomain Monte Carlo solves have been completed. Perhaps an
-even more important consequence of viewing MCSA as a stochastic
-additive Schwarz procedure is that it has the potential to more
-readily fit as an accelerator into other linear solver techniques
-including GMRES or potentially as a preconditioner.
-
-Using the Neumann-Ulam solver to compute the MCSA correction only in
-the subdomains, the strong and weak scaling studies were repeated to
-assess the performance gains made by eliminating communication in the
-Monte Carlo solver. For each scaling study, the 16-core base case with
-Monte Carlo communication was used as the reference point for
-efficiency computations in order to measure how eliminating
-communication boosts
-efficiencies. Figure~\ref{fig:titan_strong_subdomain} gives the
-results of the strong scaling study and
-Figure~\ref{fig:titan_weak_subdomain} gives the results of the the
-weak scaling study in terms of absolute efficiencies. In both cases,
-the benefits to parallel performance of using MCSA as a subdomain
-solver are obvious. In the strong scaling case we make the important
-observation that we hit the strong scaling wall at the same time with
-and without the additional Monte Carlo communication. Although the
-Monte Carlo communication did contribute to the efficiency decrease,
-it is clear that the remaining matrix and vector operations are the
-primary impediment to scalability. Furthermore, we see an even greater
-rise above unity in efficiencies because there is less communication
-overhead hiding the better cache utilization as the local problem size
-shrinks. At 16,384 cores, performing the subdomain solves boost the
-absolute efficiency from 38\% to 67\% relative to the reference case.
-
-In the weak scaling study, we do not have the scaling wall to contend
-with and eliminating Monte Carlo communication through subdomain
-solves provides an effectively linear boost in the weak scaling
-absolute efficiency. We still see a small reduction in weak scaling
-efficiency as a function of core count due to the communication
-overhead for the parallel matrix and vector operations. Here we
-observe super unitary efficiencies for the subdomain case as there was
-a substantial improvement in runtimes relative to the 16-core base
-case with Monte Carlo communication.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_subdomain.pdf}
-  \end{center}
-  \caption{\textbf{Strong scaling absolute efficiency for pure domain
-      decomposition.} \textit{The subdomain Monte Carlo solver
-      improves the parallel efficiency and therefore time to solution
-      by eliminating all parallel communication during the Monte Carlo
-      solve.}}
-  \label{fig:titan_strong_subdomain}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_subdomain.pdf}
-  \end{center}
-  \caption{\textbf{Weak scaling absolute efficiency for pure domain
-      decomposition.} \textit{The subdomain Monte Carlo solver
-      improves the parallel efficiency and therefore time to solution
-      by eliminating all parallel communication during the Monte Carlo
-      solve.}}
-  \label{fig:titan_weak_subdomain}
-\end{figure}
-
-Of additional concern for the subdomain method is the potential
-decrease in iterative performance due to the fact that histories near
-the boundary will potentially have their random walk truncated, thus
-decreasing the information available to compute the correction
-vector. For both the weak and strong scaling study, all calculations
-with split histories converged in 22 or in only a few cases 21
-iterations for both the subdomain solve and the solve with
-communication giving an algorithmic efficiency of 1.048 for all
-cases. This small improvement in iteration count was not observed to
-be a function of problem size, number of processors, or whether the
-subdomain method was used. For the replicated history cases, the
-algorithmic efficiency improvement was not as good as when
-communication was allowed for the Monte Carlo algorithm. For these
-calculations, the number of iterations to converge was reduced to 18
-or 19 in both the strong and weak scaling studies with lower core
-counts converging in fewer iterations.
-
-Therefore, there is no reason to expect overlap to improve the
-iterative or scaling performance of the method as iterative
-performance is maintained without it for this particular
-problem. Overlap is not as effective because histories relinquish most
-of their useful information within the first few steps (which is why
-we can use such a large weight cutoff and there is an insensitivity to
-weight cutoff as shown in
-Figure~\ref{fig:estimator_wc_iters}). Keeping them around longer with
-overlap to do extra work in the local domain will not improve
-iterative performance and thus increase compute times.
-
-As a final analysis of MCSA using subdomain Monte Carlo, the multiple
+\chapter{Subdomain Neumann-Ulam Scaling with Multiple Sets}
+\label{chap:extra_scaling}
+As an additional analysis of MCSA using subdomain Monte Carlo, the multiple
 sets scaling studies are repeated to determine if any performance
@@
 \end{figure}
-
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{Summary\ }
-\label{sec:parallel_summary}
-
-In this chapter we have developed a domain-decomposed parallel
-algorithm for Monte Carlo Synthetic Acceleration and explored the
-qualities of the algorithm through parallel scaling studies. The
-following are the significant observations, findings, and
-contributions.
-
-\begin{itemize}
-\item MCSA can be effectively parallelized using domain decomposition
-\item The leakage fraction of stochastic histories from a parallel
-  subdomain can be quantified analytically using algebraic quantities
-\item The multiple-set overlapping-domain (MSOD) parallel algorithm
-  for particle transport has been adapted to parallelize MCSA
-\item MCSA parallelized with MSOD was verified to produce the same
-  results for solutions to the neutron diffusion equation as
-  production Krylov methods
-\item Strong scaling studies were performed with up to 65,536 cores
-  and weak scaling studies were performed with up to 131,072 cores on
-  the Titan Cray XK7 leadership class machine at Oak Ridge National
-  Laboratory
-\item MCSA scales favorably compared to production Krylov methods for
-  both strong and weak scaling cases
-\item Overlap in small quantities determined analytically by random
-  walk length can provide parallel efficiency boosts of a few percent
-  in strong scaling cases but is ineffective in weak scaling cases
-\item Multiple sets will never boost parallel efficiencies in the weak
-  scaling case due to additional communication overhead from added
-  reduction operations but will boost efficiencies beyond the strong
-  scaling wall
-\item Multiple sets offers a means to reduce time to solution by
-  solving multiple copies of the original problem and combining the
-  solutions using superposition
-\item MCSA is most efficiently used in parallel as a stochastic
-  realization of an additive Schwarz method
-\item With future optimization, large scale reactor analyses in which
-  the transport operator is fully formed (e.g. the $SP_N$ equations)
-  could experience improved time to solution on leadership-class
-  machines with MCSA given the improved scaling performance
-\end{itemize}

### 2013-08-23 doc/dissertation/document/chapters/mc_background/mc_background.tex
spell check for spn chapter edits
moving subomain nuemann ulam results for multiple sets into an appendix

@@
 %%---------------------------------------------------------------------------%%
-\section{Monte Carlo Method Selection}
-\label{sec:mc_method_selection}
-The MCSA method defined in Eq.~(\ref{eq:mcsa}) uses the adjoint method
-to estimate the error in a residual Monte Carlo solve instead of the
-direct method outlined in \S\ref{sec:direct_mc}. To demonstrate the
-effectiveness of the adjoint method over the direct method within the
-context of MCSA, we choose the two-dimensional time-dependent Poisson
-equation as a simple model transport problem\footnote{The Poisson
-  equation is a simple form of the diffusion equation and has a
-  similar elliptic character in its time-dependent form to the neutron
-  transport equations that will be solved in the next chapter.}:
+\section{Monte Carlo Synthetic Acceleration Analysis}
+\label{sec:mcsa_analysis}
+In this section we analyze the effects of various MCSA parameters to
+better our understanding of the algorithm and facilitate future
+work. In addition, we demonstrate the effectiveness of MCSA as
+compared to Halton's Sequential Monte Carlo method. To do this, we
+choose the two-dimensional time-dependent Poisson equation as a simple
+model transport problem\footnote{The Poisson equation is a simple form
+  of the diffusion equation and has a similar elliptic character in
+  its time-dependent form to the neutron transport equations that will
+  be solved in the next chapter.}:
 \begin{equation}
@@
 
-A timing and convergence study is used to demonstrate the
-effectiveness of the adjoint method with the collision estimator as
-compared to the direct method. To assess both the CPU time and number
-of iterations required to converge to a solution, a problem of
+\subsection{Monte Carlo Method Selection}
+\label{sec:mc_method_selection}
+The MCSA method defined in Eq.~(\ref{eq:mcsa}) uses the adjoint method
+to estimate the error in a residual Monte Carlo solve instead of the
+direct method outlined in \S\ref{sec:direct_mc}. To demonstrate the
+effectiveness of the adjoint method over the direct method within the
+context of MCSA, we solve Poisson equation in a series of numerical
+experiments. A timing and convergence study is used to demonstrate
+the effectiveness of the adjoint method with the collision estimator
+as compared to the direct method. To assess both the CPU time and
+number of iterations required to converge to a solution, a problem of
 constant $\Delta$ was used with varying values of the number of mesh
@@
 machine with 16 GB 1067 MHz DDR3 memory.
+
 \begin{figure}[t!]
@@
 accurate correction in that region due to reduced statistical
-error. From an iteration perspective,
-Figure~\ref{fig:poisson_iterations} shows that using the direct method
-yields a roughly unchanging number of iterations required to converge
-as the problem size increases. Again, if we desire a correction value
-for all states in the problem, then we must start a random walk in
-each state in the system without taking into account the structure of
-the residual vector. Because of this, as the problem size grows adding
-histories is ineffective as many are added in states where the error
-is smaller than in other parts of the problem, ultimately not reducing
-the number of iterations needed to converge. Conversely, as the
-problem size grows in the adjoint method, the additional stochastic
-histories that will be computed are concentrated in regions with a
-large residual, further reducing the stochastic error in the
-correction in those regions and subsequently reducing the required
-number of iterations to converge.
+error.
+
+From an iteration perspective, Figure~\ref{fig:poisson_iterations}
+shows that using the direct method yields a roughly unchanging number
+of iterations required to converge as the problem size
+increases. Again, if we desire a correction value for all states in
+the problem, then we must start a random walk in each state in the
+system without taking into account the structure of the residual
+vector. Because of this, as the problem size grows adding histories is
+ineffective as many are added in states where the error is smaller
+than in other parts of the problem, ultimately not reducing the number
+of iterations needed to converge. Conversely, as the problem size
+grows in the adjoint method, the additional stochastic histories that
+will be computed are concentrated in regions with a large residual,
+further reducing the stochastic error in the correction in those
+regions and subsequently reducing the required number of iterations to
+converge.
 
@@
 fixed problem size of $N=500$. Figure~\ref{fig:poisson_convergence}
-gives the results of these computations. First, it is worthy to note
-on the semi-log plot that we are indeed achieving the expected
-exponential convergence from MCSA with both Monte Carlo
-solvers. Second, we note that using the adjoint method with the same
-number of stochastic histories per MCSA iteration gives a faster rate
-of converge for the same reasons as above. We also note here that
-fewer iterations are required for convergence when the 9-point stencil
-is used to discretize the Laplacian operator (although at no gain in
-speed as given by the results in
-Figure~\ref{fig:poisson_cpu_time}). This is due to the fact that the
-smaller discretization error directly corresponds to a more well
-defined residual source generated by the Richardson extrapolation for
-the Monte Carlo calculation. In addition, the better defined source is
-transported through a domain described more accurately by the 9-point
-stencil, thus yielding a more accurate correction vector from the
-Monte Carlo calculation.
+gives the results of these computations. We note that using the
+adjoint method with the same number of stochastic histories per MCSA
+iteration gives a faster rate of converge for the same reasons as
+above. We also note here that fewer iterations are required for
+convergence when the 9-point stencil is used to discretize the
+Laplacian operator (although at no gain in speed as given by the
+results in Figure~\ref{fig:poisson_cpu_time}). This is due to the fact
+that the smaller discretization error directly corresponds to a more
+well defined residual source generated by the Richardson extrapolation
+for the Monte Carlo calculation. In addition, the better defined
+source is transported through a domain described more accurately by
+the 9-point stencil, thus yielding a more accurate correction vector
+from the Monte Carlo calculation.
+
 \begin{figure}[t!]
@@
 
-%%---------------------------------------------------------------------------%%
-\section{MCSA Comparison to Sequential Monte Carlo}
+\subsection{MCSA Comparison to Sequential Monte Carlo}
 \label{subsec:sequential_comparison}
 To further motivate using Monte Carlo Synthetic Acceleration, we
-compare its performance to Halton's Sequential Monte Carlo method on
-which previous work in this area was based. For this comparison, we
-use the same transient Poisson problem as described in the previous
-section and choose only the 5-point stencil to discretize the
-Laplacian operator as the previous results yielded little qualitative
-difference between the discretizations. Both MCSA and Halton's method
-are used with the adjoint Monte Carlo solver and the collision
-estimator. In order to complete the same study as in the previous
-section, the number of histories computed by the Monte Carlo solver at
-each iteration had to be doubled to $2 \times N \times N$ in order to
-ensure convergence in Sequential Monte Carlo Method. For the majority
-of the problems in the previous section, the Sequential method used
-with $N \times N$ histories would not
-converge.
+compare its performance to Halton's Sequential Monte Carlo. For this
+comparison, we use the same transient Poisson problem as described in
+the previous section and choose only the 5-point stencil to discretize
+the Laplacian operator as the previous results yielded little
+qualitative difference between the discretizations. Both MCSA and
+Halton's method are used with the adjoint Monte Carlo solver and the
+collision estimator. In order to complete the same study as in the
+previous section, the number of histories computed by the Monte Carlo
+solver at each iteration had to be doubled to $2 \times N \times N$ in
+order to ensure convergence in Sequential Monte Carlo Method. For the
+majority of the problems in the previous section, the Sequential
+method used with $N \times N$ histories would not converge.
 
@@
 
-%%---------------------------------------------------------------------------%%
-\section{Monte Carlo Parameter and Estimator Analysis}
+\subsection{Monte Carlo Parameter and Estimator Analysis}
 \label{sec:parameter_estimator_analysis}
-With the adjoint method shown to be more effective than the direct
-method and MCSA to have better iterative and timing performance than
-sequential Monte Carlo, we now aim to study the effects of the adjoint
-Monte Carlo parameters, weight cutoff and number of histories, on MCSA
-performance with both the collision and expected value
-estimators. With the same Poisson problem, we will use a $200 \times
-200$ grid and a convergence tolerance of \sn{1}{-8} for all
-calculations. To study the effects of the number of Monte Carlo
-histories per MCSA iteration, the weight cutoff was fixed at
-\sn{1}{-4} while the number of histories per iteration was varied from
-6,000 to 100,000. It was observed that MCSA would not converge for
-this problem using the collision estimator with less than 6,000
-histories while the expected value estimator permitted convergence
-with only 2,000 histories.
+We now study the effects of the adjoint Monte Carlo parameters, weight
+cutoff and number of histories, on MCSA performance with both the
+collision and expected value estimators. With the same Poisson
+problem, we will use a $200 \times 200$ grid and a convergence
+tolerance of \sn{1}{-8} for all calculations. To study the effects of
+the number of Monte Carlo histories per MCSA iteration, the weight
+cutoff was fixed at \sn{1}{-4} while the number of histories per
+iteration was varied from 6,000 to 100,000. It was observed that MCSA
+would not converge for this problem using the collision estimator with
+less than 6,000 histories while the expected value estimator permitted
+convergence with only 2,000 histories.
 
@@
 simple model transport system. The following are the significant
-observations and findings.
+observations, findings, and contributions.
 

### 2013-08-23 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
moving subomain nuemann ulam results for multiple sets into an appendix

@@
 histories \cite{alexandrov_efficient_1998} and in this chapter we will
-exploit particle transport algorithms to alleviate this. To accomplish
-this, we recognize from the literature that stochastic histories must
-be transported from domain to domain as the simulation progresses and
-they transition to states that are not in the local domain. In this
+exploit particle transport algorithms to alleviate this. In this
 chapter we briefly review particle transport methods for domain
@@
 remain decomposed to avoid the same performance and memory
-restrictions that required the domain to be decomposed. To parallelize
-this algorithm, we then seek parallel Monte Carlo algorithms that
-handle domain decomposition.
+restrictions. To parallelize this algorithm, we then seek parallel
+Monte Carlo algorithms that handle domain decomposition.
 
@@
 \cite{brunner_efficient_2009}. We will adapt their algorithm and
-directly apply it to a parallel formulation of the Neumann-Ulam
-method. Direct analogs can be derived from their works by noting that
-the primary difference between solving a linear transport system with
-Monte Carlo methods and traditional fixed source Monte Carlo transport
-problems is the content of the Markov chains that are generated.
-
-The transitions represented by these chains are bound by probabilities
-and weights and are initiated by the sampling of a discrete source. In
-the context of transport problems, those transitions represent events
-such as particle scattering and absorption with probabilities that are
-determined by physical data in the form of cross sections. For
-stochastic matrix inversion, those transitions represent moving
-between the equations of the linear system (and therefore the
-components of phase space which they represent) and their
-probabilities are defined by the coefficients of those
-equations. Ultimately, we tally the contributions to generate
-expectation values in the desired states as we progress through the
-chains. Therefore, parallel methods for Monte Carlo radiation
-transport can be abstracted and we can use those concepts that apply
-to matrix inversion methods as an initial means of developing a
-parallel Neumann-Ulam-type solver.
+directly apply it to construct a parallel formulation of the
+Neumann-Ulam method. Direct analogs can be derived from their works by
+noting that the primary difference between solving a linear transport
+system with Monte Carlo methods and traditional fixed source Monte
+Carlo transport problems is the content of the Markov chains that are
+generated.
 
@@
 which was effectively implemented verbatim for this work, is presented
-along with its application to the Neumann-Ulam method. For
-considerably more detail, the algorithm presented can be found in
-\cite{brunner_efficient_2009}. In their work they identify two data
-sets that are required to be communicated: the sharing of particles
-that are transported from one domain to another and therefore from one
-processor to another and a global communication that signals if
-particle transport has been completed on all processors. Both of these
-communication sequences will be addressed at a high level along with
-how the Monte Carlo data structures they require are constructed in
-parallel.
+along with its application to the Neumann-Ulam method. In their work
+they identify two data sets that are required to be communicated: the
+sharing of particles that are transported from one domain to another
+and therefore from one processor to another and a global communication
+that signals if particle transport has been completed on all
+processors. Both of these communication sequences will be addressed at
+a high level along with how the Monte Carlo data structures they
+require are constructed in parallel.
 
@@
 operator as given by Eq~\ref{eq:neumann_ulam_decomposition}. Given an
-input decomposition for the linear operator, by definition the
-Neumann-Ulam decomposition will have the same parallel
-decomposition. In addition, any modification made to the linear
-operator through preconditioning or relaxation parameters will also
-modify the Neumann-Ulam decomposition. All possible transitions are
-described by the local graph of the sparse input matrix. In addition,
-any left preconditioning or relaxation parameters will modify the
-right hand side of the linear system and therefore the fixed source in
-the Monte Carlo calculation. This source vector will have the same
+input parallel decomposition for the linear operator, by definition
+the Neumann-Ulam decomposition will have the same parallel
+decomposition. In addition, the source vector will have the same
 parallel decomposition as the input matrix and therefore all of the
@@
 data for all states that is required to be on process must be
-collected in parallel. For the pure domain decomposition case, given
-an input parallel decomposition for the Neumann-Ulam decomposition,
-for all local states $m$ we require access to $P_{mn}$ and $W_{mn}$
-(or $P^T_{mn}$ and $W^T_{mn}$) for the adjoint method for all possible
-states $n$. Given by Figure~\ref{fig:diffusion_graph}, consider the
-adjacency graph of a single state in the neutron diffusion matrix for
-the model problem presented in Appendix~\ref{chap:diffusion_problem}.
+collected in parallel so that all possible local transitions may be
+completed. For the pure domain decomposition case, given an input
+parallel decomposition for the Neumann-Ulam decomposition, for all
+local states $m$ we require access to $P_{mn}$ and $W_{mn}$ (or
+$P^T_{mn}$ and $W^T_{mn}$ for the adjoint method) for all possible
+states $n$. As an example of this state collection,
+Figure~\ref{fig:diffusion_graph} gives the adjacency graph of a single
+state in the neutron diffusion matrix for the model problem presented
+in Appendix~\ref{chap:diffusion_problem}.
+
 \begin{figure}[t!]
@@
 node at which we started. If grid point $(i,j)$ in this example is on
-the right of the local domain boundary and we transition to the right
-to state $n$ corresponding to mesh point $(i+1,j)$, we are now in a
-state that is owned by the adjacent domain. We must then gather the
-data in the Neumann-Ulam decomposition from the neighboring process
-that owns grid point $(i+1,j)$. Doing this provides us with the data
-required by the estimators permitting $P_{mn}$ and $W_{mn}$ to be
-computed locally for this particular boundary transition. In addition,
-we also collect the identification number for the process that owns
-this state such that we have an address to send all histories that
-leave the local domain boundary for this state. For the parallel
+the right side of the local domain boundary and we transition to the
+state $n$ corresponding to mesh point $(i+1,j)$, we are now in a state
+that is owned by the adjacent domain. We must then gather the data in
+the Neumann-Ulam decomposition from the neighboring process that owns
+grid point $(i+1,j)$. Doing this provides us with the data required by
+the estimators permitting $P_{mn}$ and $W_{mn}$ to be computed locally
+for this particular boundary transition. In addition, we also collect
+the identification number for the process that owns this state such
+that we have an address to send all histories that leave the local
+domain boundary with this state as their destination. For the parallel
 source, these adjacent states are not required as histories may only
@@
 
-Here we present in detail Brunner and Brantley's 2009 algorithm in
-detail and discuss how it is adapted to parallelize the Neumann Ulam
+We next present Brunner and Brantley's 2009 algorithm in detail and
+discuss how it is adapted to parallelize the Neumann Ulam
 method. Presented in Algorithm~\ref{alg:parallel_mc_algorithm}, the
@@
 figure. For each set of neighbors, a non-blocking send and receive is
-required with data buffers allocated with a user-defined size prepared
-for incoming and outgoing histories. This nonblocking structure is
+required with data buffers allocated with a user-defined size for
+incoming and outgoing histories. This nonblocking structure is
 critical to the performance of the algorithm in that it permits local
@@
 (\textit{historiesCompleted} in the algorithm) be updated
-incrementally and funneled to the root process. Because we are
-solving fixed source problems with the Neumann-Ulam algorithm without
-any type of variance reduction that may generate more histories that
-we started with, once the root process completion sum tallies to the
-number of histories in the source, transport is complete. Once this
-occurs, the root process nonblocking sends a process to its children
-and each process nonblocking receives a stop message from its parent
-as shown in Figure~\ref{fig:binary_comm_tree}. The stop message is
-then propagated up the tree in the same manner.
+incrementally and funneled to the root process. Because we are solving
+fixed source problems with the Neumann-Ulam algorithm without any type
+of variance reduction that may generate more histories that we started
+with, once the root process completion sum tallies to the number of
+histories in the source, transport is complete. Once this occurs, the
+root process nonblocking sends the stop message to its children and
+each process nonblocking receives a stop message from its parent as
+shown in Figure~\ref{fig:binary_comm_tree}. The stop message is then
+propagated up the tree in the same manner.
 
@@
       \input{chapters/parallel_mc/binary_comm_tree.pdftex_t} }
-    \caption{\textbf{Binary communication tree for coordinating the
-        end of a parallel Neumann-Ulam solve.} \textit{Each child
-        process reports to its parents how many histories completed
-        within its domain. When the root process sums all complete
-        histories, it forwards the stop signal to its children which
-        in turn forward the message up the tree. The subdomains are
-        indexed by an integer.}}
-  \end{center}
+  \end{center}
+  \caption{\textbf{Binary communication tree for coordinating the end
+      of a parallel Neumann-Ulam solve.} \textit{Each child process
+      reports to its parents how many histories completed within its
+      domain. When the root process sums all complete histories, it
+      forwards the stop signal to its children which in turn forward
+      the message up the tree. The subdomains are indexed by an
+      integer.}}
   \label{fig:binary_comm_tree}
@@
 and the second is a LIFO (last in first out) stack of histories
-transported to the local domain from the adjacent domain. In lines
+transported to the local domain from the adjacent domains. In lines
 14-17, if histories exist in either of these data structures, then
@@
 the problem, if there are histories in those data buffers then they
-are added to the stack from processing and the nonblocking receive
+are added to the stack for processing and the nonblocking receive
 operation re-instantiated. In addition, the terminated histories count
@@
 
-In this manner the transport loop continues until all process have
+The transport loop continues in this manner until all processes have
 received the stop signal from their parents. Older versions of this
@@
 stage instead of the binary tree scheme. In this approach, the root
-process still manages the final summation of the completion tally, it
-directly receives completed tally results from all other processes in
-the problem instead of from just its children. Although this
+process still manages the final summation of the completion tally, but
+it directly receives completed tally results from all other processes
+in the problem instead of from just its children. Although this
 implementation is potentially simpler to understand than the binary
@@
 that this type of termination pattern, even when implemented in a
-fully asynchronous manner, was a major scalability bottleneck. This
-bottleneck is due to the fact that the master process falls behind the
-others, creating a load imbalance even with the addition of a few
-integer addition operations. We will explicitly demonstrate in scaling
-studies how this bottleneck also occurs within a Neumann-Ulam
-implementation of this algorithm thus requiring the implementation of
-a binary communication tree.
+fully asynchronous manner, was a significant scalability
+bottleneck. This bottleneck is due to the fact that the master process
+falls behind the others, creating a load imbalance even with the
+addition of a few integer addition operations. We will explicitly
+demonstrate in scaling studies how this bottleneck also occurs within
+a Neumann-Ulam implementation of this algorithm thus requiring the
+implementation of a binary communication tree.
 
@@
 parallel efficiencies in Brunner's work and therefore will scale well
-in this algorithm. The overlapping domains within each set also
-demonstrated reduced communication costs. On each processor, the
-source is sampled in the local domain that would exist if no overlap
-was used while tallies can be made over the entire overlapping domain.
+in this algorithm.
 
@@
 linear problem, the linear operator is used to construct the transport
-domain, the solution vector used to construct the tallies, and the
-forcing term vector used to construct the fixed source. Once these
-Monte Carlo data structures have been generated, they are broadcast to
-the remaining sets in the problem as shown in
+domain with overlap, the solution vector used to construct the
+tallies, and the forcing term vector used to construct the fixed
+source. Once these Monte Carlo data structures have been generated,
+they are broadcast to the remaining sets in the problem as shown in
 Figure~\ref{fig:msod_construction} such that we have $S$ replications
@@
 Algorithm~\ref{alg:parallel_mc_algorithm} in this case simply modifies
-the set of local states and the set of neighboring states to
+the set of local states and the set of neighboring states with the
 additional states gathered during overlap generation.
@@
 
-Once Monte Carlo transport is complete complete, the tallies for each
+Once Monte Carlo transport is complete, the tallies for each
 individual computation must be gathered to the original set in order
@@
 within all sets. We can take advantage of this by building a processor
-space that encapsulates each identical domain in each set. We can then
-use this processor space for the reduction. For the schematic in
-Figure~\ref{fig:msod_tally_reduction}, the upper-left domain in each
-set creates a single block, giving 9 total blocks for this example
+space that encapsulates each identical domain in each set and then use
+this processor space for the reduction. For the schematic in
+Figure~\ref{fig:msod_tally_reduction}, the combined upper-left domains
+in each set create a block, giving 9 total blocks for this example
 problem. The processor space that encapsulates each of the upper-left
@@
 reduction will happen in a mutually exclusive manner for each of the
-blocks in the system. One benefit of this approach is that this
-reduction occurs only over a subset of the processors in the problem,
-providing better scalability than a reduction operation over all
-processors in the problem.
+blocks in the system.
 
@@
 (\ref{eq:mcsa_r4}), and (\ref{eq:mcsa_r5}). In the first step, a
-parallel matrix-vector multiply is used to apply the split operator to
-the previous iterate's solution. A parallel vector update is then
-performed with the source vector to arrive at the initial iteration
-guess. In the next step, the residual is computed by the same
-operations where now the operator is applied to the solution guess
-with a parallel matrix-vector multiply and then a parallel vector
-update with the source vector is performed. Once the correction is
+parallel vector update is used to apply the residual to the previous
+iterate's solution in a Richardson iteration. In the next step and
+Eq~(\ref{eq:mcsa_r5}), the residual is computed by a parallel
+matrix-vector multiply and vector update. Once the correction is
 computed with a parallel adjoint Neumann-Ulam solve, this correction
-is applied to the guess with a parallel vector update to get the new
-iteration solution. Additionally, as given by
+is applied to the solution with a parallel vector
+update. Additionally, as given by
 Eq~(\ref{eq:mcsa_stopping_criteria}), 2 parallel vector reductions
@@
 the rest of an MCSA iteration occur only on the subset of processors
-containing the primary set. Only the Monte Carlo data structures exist
-on the rest of the processors in the problem as replication of the
-other operations does not generate any more information that could be
-used to accelerate the time to solution.
+containing the primary set\footnote{For applications where resiliency
+  is potentially of concern, the operations in Eqs~(\ref{eq:mcsa_r1}),
+  (\ref{eq:mcsa_r2}), (\ref{eq:mcsa_r4}), and (\ref{eq:mcsa_r5}) may
+  be replicated without significant penalty. Doing so my alleviate
+  issues where any part of the linear problem or residual is disturbed
+  as now there will be multiple copies of these data
+  structures.}. Only the Monte Carlo data structures exist on the rest
+of the processors in the problem as replication of the other
+operations does not generate any more information that could be used
+to accelerate the time to solution.
 
@@
 system in parallel through the element-wise function evaluations to
-generate both the global Jacobian operator and the global residual
-vector on the right hand side. Per Bartlett's work on automatic
-differentiation, efficient and automated parallel mechanisms are
-available to do this through a sequence of scatter/gather
+generate both the global Jacobian operator and the global nonlinear
+residual vector on the right hand side. Per Bartlett's work on
+automatic differentiation, efficient and automated parallel mechanisms
+are available to do this through a sequence of scatter/gather
 operations. With these tools available for residual and Jacobian
@@
 applied to the previous iterate's solution through a parallel vector
-update.
+update with parallell norm operations applied to check for
+convergence.
 
@@
 Belos, a linear solvers package in the Trilinos scientific computing
-libraries \cite{heroux_overview_2005}. In identical fashion to the
-verification for the $SP_N$ equations performed in
-\S~\ref{sec:spn_mcsa_verification}, we will solve the neutron
-diffusion problem in parallel using multiple cores on a desktop
-machine. For each parallel solution, then the solution at each grid
-point in the system will be compared to the solution at the same grid
-point in the serial reference case with the minimum and maximum
-absolute differences of these values reported.
+libraries \cite{heroux_overview_2005}. For each parallel solution, the
+solution at each grid point in the system will be compared to the
+solution at the same grid point in the serial reference case with the
+minimum and maximum absolute differences of these values reported.
 
@@
 every DOF in the problem to compute the correction vector. For all
-solvers, the flux was converged to a tolerance of \sn{1}{-8} and
-therefore if the comparisons agree within this tolerance the solutions
-will be deemed
-correct. Table~\ref{tab:parallel_verification_parameters} gives the
+solvers, the flux was converged to a tolerance of
+\sn{1}{-8}. Table~\ref{tab:parallel_verification_parameters} gives the
 parameters for the verification problem and
@@
 differences in the results as a function of number of blocks, number
-of sets, or amount of overlap. In addition, the parallel GMRES results
-have maximum absolute differences of the same magnitude as all MCSA
-results and a larger minimum absolute difference than all MCSA results
-by three orders of magnitude. Based on these results, MCSA
-parallelized with the MSOD algorithm, and in particular the
-implementation used here, generates the correct results with respect
-to the serial conjugate gradient reference computation.
+of sets, or amount of overlap. The parallel GMRES results have maximum
+absolute differences of the same magnitude as all MCSA results and a
+larger minimum absolute difference than all MCSA results by three
+orders of magnitude. Based on these results, MCSA parallelized with
+the MSOD algorithm, and in particular the implementation used here,
+generates the correct results with respect to the serial conjugate
+gradient reference computation.
 
@@
 scaling studies, large problems at high levels of concurrency will be
-solved at a leadership-class computing facility. In this section, we
-will outline the parallel machine on which the calculations were
+solved using a leadership-class computing facility. In this section,
+we will outline the parallel machine on which the calculations were
 performed and then present a series of numerical studies that identify
-both strengths and weaknesses of the parallel algorithm. Both pure
-domain decomposition and MSOD parallel performance will be explored
-and compared to performance results for the same problems using a set
-of production parallel Krylov solvers.
+both strengths and weaknesses of the parallel algorithm developed by
+this work. Both pure domain decomposition and MSOD parallel
+performance will be explored and compared to performance results for
+the same problems using a set of production parallel Krylov solvers.
 
@@
 the frequency by which a domain checks for incoming information from
-neighboring domain, both of which may potentially affect latency and
+neighboring domains, both of which may potentially affect latency and
 load balancing. Changing the static history buffer size parameter
@@
 of 0.78 with a global problem size of \sn{1.6}{7} DOFs with one
-stochastic history used per DOF in the MCSA Neumann-Ulam used to
-compute the correction for the acceleration. Each parameter was varied
-between 16 and 4096 and the time to solve the problem recorded. The
-results of this study are presented in
+stochastic history used per DOF in the MCSA solution. Each parameter
+was varied between 16 and 4096 and the time to solve the problem
+recorded. The results of this study are presented in
 Figure~\ref{fig:titan_comm_parameters}. Compared to the results of
@@
   \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_comm_parameters.pdf}
+    \includegraphics[width=5.0in]{chapters/parallel_mc/titan_comm_parameters.pdf}
   \end{center}
@@
 Table~\ref{tab:parallel_verification_parameters} and
-Table~\ref{tab:parallel_mcsa_parameters}. As with all of the scaling
-studies, 3 calculations for all data points were performed with the
-average wall time reported and used for the efficiency
-calculations. Figure~\ref{fig:titan_pure_strong_time} gives the wall
-time per iteration for these calculations while
-Figure~\ref{fig:titan_pure_strong} gives the results of the this study
+Table~\ref{tab:parallel_mcsa_parameters}. Figure~\ref{fig:titan_pure_strong_time}
+gives the wall time per iteration for these calculations while
+Figure~\ref{fig:titan_pure_strong} gives the results of this study
 with the given absolute efficiencies computed using
@@
 all computations converged in the same number of iterations relative
-to the base case\footnote{When the algorithmic efficiency remains at
-  unity for all calculations, the implementation efficiency is
-  equivalent to the absolute efficiency and therefore is not presented
-  here.}. Instead, what is happening here and was observed by Brunner
-and Brantley in their work is that the large problem used to scale to
-O(10,000) cores was large enough that memory thrashing was occurring
-in the 16 core base case. As the number of cores in the strong scaling
-study was increased and the local problem size decreased, more of the
-local problem fit into the cache of the system, thus causing improved
-serial performance and therefore enhanced runtimes. In this region of
-super-linear efficiencies, we also observe larger efficiencies for
-GMRES as compared to conjugate gradient. This occurs due to the fact
-that conjugate gradient has a static set of operations performed at
-every iteration while GMRES operates on a continually growing
-subspace. In addition, the orthogonalization operations required to
-generate that subspace consume much more time than the basic conjugate
-gradient operations, causing a larger serial runtime and subsequently
-higher efficiencies. For the same reasons, we also observe higher
+to the base case. Instead, what is happening here and was observed by
+Brunner and Brantley in their work is that the large problem used to
+scale to O(10,000) cores was large enough that memory thrashing was
+occurring in the 16 core base case. As the number of cores in the
+strong scaling study was increased and the local problem size
+decreased, more of the local problem fit into the cache of the system,
+thus causing improved serial performance and therefore enhanced
+runtimes. In this region of super-linear efficiencies, we also observe
+larger efficiencies for GMRES when compared to conjugate
+gradient. This occurs due to the fact that conjugate gradient has a
+static set of operations performed at every iteration while GMRES
+operates on a continually growing subspace. In addition, the
+orthogonalization operations required to generate that subspace
+consume much more time than the basic conjugate gradient operations,
+causing a larger serial runtime and subsequently higher
+efficiencies. For the same reasons, we also observe higher
 efficiencies for the MCSA implementation due to the serial performance
@@
 particular global problem is not an efficient use of those parallel
-resources. Furthermore, the fact that we see a very similar trends in
+resources. Furthermore, the fact that we see very similar trends in
 efficiencies for all methods means that they are bound by the same
@@
 increase the number of processors used in the
-simulation. Figure~\ref{fig:titan_pure_weak_time} gives the wall
-time per iteration and Figure~\ref{fig:titan_weak_absolute} gives the
-weak scaling absolute efficiencies for this problem as computed by
+simulation. Figure~\ref{fig:titan_pure_weak_time} gives the wall time
+per iteration and Figure~\ref{fig:titan_weak_absolute} gives the weak
+scaling absolute efficiencies for this problem as computed by
 Eq~(\ref{eq:weak_scaling_absolute}). In general, the relative trends
@@
 GMRES scaling better than the conjugate gradient method due to the
-increased amount of arithmetic work on-process. In addition, it is
-important to note that the weak scaling of MCSA when properly
-parallelized remains competitive with the Krylov methods keeping in
-mind the artificial inflation of efficiencies.
+increased amount of arithmetic work on-process. Again, it is important
+to note that the weak scaling of MCSA when properly parallelized
+remains competitive with the Krylov methods, keeping in mind the
+artificial inflation of efficiencies.
 
@@
 using Eq~(\ref{eq:algorithmic_efficiency}). As shown in
-Figure~\ref{fig:titan_weak_algorithmic}, the algorithm efficiency of
+Figure~\ref{fig:titan_weak_algorithmic}, the algorithmic efficiency of
 the Krylov algorithm improves as a function of the number of cores in
@@
 we are effectively measuring the parallel performance of a single
-iteration for each method by looking at these implementation
-efficiencies. As the global problem size increases, MCSA is less prone
-to a drop in parallel efficiency. If optimization has been performed
-and runtimes potentially approach those of the Krylov methods, this
+iteration for each method. If optimization has been performed and
+runtimes potentially approach those of the Krylov methods, this
 comparison may not be as favorable. Overall, weak scaling performance
@@
 cores, any extra work given to the master process, even if it amounts
-to summing a handful of integers and checking for a few extra messages
-as in this algorithm, grows enough to create serious load imbalance
-and reduce performance.
+to summing a handful of integers and checking for a few extra
+messages, grows enough to create serious load imbalance and reduce
+performance.
 
@@
 use a variation of Eq~(\ref{eq:step_k_length}) to determine how many
-discrete states a history will move on average from its birth site. We
-modify Eq~(\ref{eq:step_k_length}) to only account for discrete
-movements:
+discrete states a history will move on average from its birth site by
+modifying it to only account for discrete movements:
 \begin{equation}
@@
 decomposition case. As we are comparing to the case without overlap
-and are looking to boost performance relative to the case without
-overlap, for both weak and strong scaling calculations we will use the
-16-core case without overlap as the reference computation for all
-efficiency measures reported.
-
-The absolute efficiency results from the strong scaling exercise are
-given in Figure~\ref{fig:titan_strong_overlap}. Aside from the
-boosting of efficiencies at 2,048 cores, little evidence of strong
-scaling improvement is evident for all levels of overlap tested. To
-further explore any potential benefits of adding overlap into the
-system, Figure~\ref{fig:titan_strong_overlap_diff} gives the
-difference in the efficiencies between all overlap cases and the
-reference case without overlap at each core count. At smaller numbers
-of cores, there is no observed benefit to using overlap to improve
-strong scaling. The improvements noted at 2,048 cores are also present
-here and account for the best efficiency improvements for all core
-counts tested. at 16,384 cores, using 2 levels of overlap did give a
-6.4\% boost in efficiency while at 1,024 cores this same amount of
-overlap actually reduced the efficiency. Beyond 2 levels of overlap,
-efficiency levels begin to fall with 5 levels of overlap performing
-the worst for all cases. From this, we don't expect adding any more
-overlap will improve the scalability of the algorithm and the best
-strong scaling performance was observed when the number of overlap
-levels is approximately equal to the root-mean-squared number of
-states we expect a history to move from its starting point as computed
-by Eq~(\ref{eq:discrete_distance}). We will defer an explanation of
-the drop in efficiencies for levels of overlap beyond this value until
-we have analyzed the weak scaling data.
+and are looking to boost performance relative to this case, for both
+weak and strong scaling calculations we will use the 16-core case
+without overlap as the reference computation for all efficiency
+measures reported. The absolute efficiency results from the strong
+scaling exercise are given in
+Figure~\ref{fig:titan_strong_overlap}. Aside from the boosting of
+efficiencies at 2,048 cores, little evidence of strong scaling
+improvement is evident for all levels of overlap tested. To further
+explore any potential benefits of adding overlap into the system,
+Figure~\ref{fig:titan_strong_overlap_diff} gives the difference in the
+efficiencies between all overlap cases and the reference case without
+overlap at each core count. At smaller numbers of cores, there is no
+observed benefit to using overlap to improve strong scaling. The
+improvements noted at 2,048 cores are also present here and account
+for the best efficiency improvements for all core counts tested. At
+16,384 cores, using 2 levels of overlap did give a 6.4\% boost in
+efficiency while at 1,024 cores this same amount of overlap actually
+reduced the efficiency. Beyond 2 levels of overlap, efficiency levels
+begin to fall with 5 levels of overlap performing the worst for all
+cases. From this, we don't expect adding any more overlap will improve
+the scalability of the algorithm and the best strong scaling
+performance was observed when the number of overlap levels is
+approximately equal to the root-mean-squared number of states we
+expect a history to move from its starting point as computed by
+Eq~(\ref{eq:discrete_distance}).
 
@@
 Figure~\ref{fig:titan_weak_overlap} and the differences in
-efficiencies from the calculations without overlap. The results here
-are much the same in that there is not a dramatic improvement in
-scalability relative to the case without overlap. Opposite the strong
-scaling calculation, overlap enhances weak scaling efficiencies at
-lower core counts and degrades the efficiencies at higher core
-counts. In addition, 1 level of overlap provided the best enhancement
-of scalability when overlap was beneficial.
+efficiencies from the calculations without overlap given in
+Figure~\ref{fig:titan_weak_overlap_diff}. The results here are much
+the same in that there is not a dramatic improvement in scalability
+relative to the case without overlap. Opposite the strong scaling
+calculation, overlap enhances weak scaling efficiencies at lower core
+counts and degrades the efficiencies at higher core counts. In
+addition, 1 level of overlap provided the best enhancement of
+scalability when overlap was beneficial.
 
@@
 performance improvement at higher core counts because the total number
-of overlapping states on each core is smaller from the local problem
+of overlapping states on each core is smaller due to the local problem
 size shrinking. For weak scaling, we see better performance
@@
 
-Based on this data, a few levels of overlap within the expected
-distance of travel for a history may offer some marginal improvement
-in scalability, however, it is not as effective as one might think
-given its improvement to reactor physics calculations as observed in
-the literature. This is largely due to the density of states in the
-system that creates the described performance bottlenecks in the
-overlapping tally vector reduction operation. In a Monte Carlo
-particle simulation, these overlapping tally reduction operations are
-not expected to be performed at the speed they are required in
-MCSA. For example, in a criticality calculation with overlapping
-domains, we would expect to perform as many parallel tally vector
-reduction in a particle transport simulation as there are particle
-stages required for convergence of the eigenvalue. However, these
-particle stages will likely be much more costly in terms of wall time
-relative to a Neumann-Ulam solve in an MCSA iteration due to the more
-extensive particle physics and logic structure that must be
-evaluated. Therefore, we can expect this overlapping tally reduction
-to be less significant to scalability in those operations. We do not
-have as complicated a logic structure in a history random walk for the
-Neumann-Ulam method and therefore overlap can inhibit scalability if
-too large.
-
 \subsection{Multiple Set Domain Decomposition}
@@
 the problem with pure domain decomposition within the sets for these
-scaling studies. Before performing parallel scaling studies using
-multiple sets, several aspects of the replicated parallel problem must
-be considered in order to effectively craft the studies and interpret
-their results. First is the question of how many stochastic histories
-to use to generate the correction vector in the MCSA iteration. For
-this work we consider two choices; the split case and the fully
-replicated case. As an example of these cases, consider a single set
-problem where 20,000 histories are used in the Monte Carlo solve. If
-we choose the split case, then solving the same problem with 2 sets
-would give 10,000 histories in each set and 4 sets would give 5,000
-histories in each set. Splitting the histories in this way provides
-the same global number of histories used to compute the correction in
-the single set problem. We expect to reduce the time to solution by
-doing this as each individual set solve will occur faster in an almost
-linear fashion with the reduced number of histories. For the
-replicated history case, each additional set in the problem will use
-just as many histories as the single set case. Using 2 sets with fully
-replicated histories for our example problem then has each set
-computing 20,000 histories with 40,000 global histories in the
-computation. For the 4 set case, the number of global histories is
-increased to 80,000. We also expect this type of replication to reduce
-the time to solution as the extra global histories should reduce the
-stochastic uncertainty in the correction vector and reduce the overall
-number of MCSA iterations required to converge.
+scaling studies. When adding sets, we must decide how many stochastic
+histories to use to generate the correction vector in the MCSA
+iteration. For this work we consider two choices; the split case and
+the fully replicated case. As an example of these cases, consider a
+single set problem where 20,000 histories are used in the Monte Carlo
+solve. If we choose the split case, then solving the same problem with
+2 sets would give 10,000 histories in each set and 4 sets would give
+5,000 histories in each set. Splitting the histories in this way
+provides the same global number of histories used to compute the
+correction in the single set problem. For the replicated history case,
+each additional set in the problem will use just as many histories as
+the single set case. Using 2 sets with fully replicated histories for
+our example problem then has each set computing 20,000 histories with
+40,000 global histories in the computation. For the 4 set case, the
+number of global histories is increased to 80,000.
 
@@
 
-For the strong scaling study, we apply the additional sets as
-follows. Using the same problems as the pure domain decomposition
-scaling study, the multiple sets are used to replicate those problems
-and boost the core count. For example, for the 4 set case, the data
-point at 64 cores has the same $N/P$ ratio in one set as the single
-set problem with 16 cores. This is due to the fact that $N$ is fixed
-for the strong scaling study and 4 replications gives a set size of 16
-cores. For the 2 set case, the 32 core data point is solving the same
-$N/P$ problem in a set as the 16 core base case.
-
-In addition to considering the replicated domain, we must also
-consider how to modify the efficiencies we will compute. For the
-strong scaling case, recall that the objective is to decrease the time
-to solution for a given global problem size $N$ by increasing
+For the strong scaling case, recall that the objective is to decrease
+the time to solution for a given global problem size $N$ by increasing
 $P$. Therefore, we may use Eqs~(\ref{eq:strong_scaling_absolute_ref}),
@@
 
-From this timing data we then compute the absolute efficiencies for
-strong scaling as given in Figure~\ref{fig:titan_strong_ms_eff} using
-the 16-core single set calculation as the reference case. As observed
-in the timing data, because the single set case had the fastest run
-times up to the strong scaling wall it also has the best observed
+From this timing data we compute the absolute efficiencies for strong
+scaling as given in Figure~\ref{fig:titan_strong_ms_eff} using the
+16-core single set calculation as the reference case. As observed in
+the timing data, because the single set case had the fastest run times
+up to the strong scaling wall it also has the best observed
 efficiencies. For the 2 set case with history splitting, although the
@@
 efficiencies. This reduction across blocks is even more obvious for
-the 4 set case. For the replicated cases, we do observe a flat region
-of parallel efficiencies at larger core counts (up to around 32,000
-for the 4 set case with history replication) showing improved scaling
-within context of not considering the single set
-calculation. Unfortunately, the additional cores do not reduce the
-runtime of the global problem as efficiently as expected. Ultimately,
-all cases effectively hit the strong scaling wall at the same time at
-around O(10,000) cores.
+the 4 set case. Ultimately, all cases effectively hit the strong
+scaling wall at the same time at around O(10,000) cores.
 
@@
 16,384 cores, there is not a compelling reason to use more
-computational resources for the this problem.
+computational resources for this problem.
 
@@
 that the replicated history cases are even less efficient than in the
-absolute case with respect to the performance of a single
-iteration. Not only do we have the additional parallel overhead of
-doing the block level reduction, but we also have additional histories
-that increase the time to solution for a single iteration.
+absolute case with respect to the performance of a single iteration.
 
@@
 
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=5in]{chapters/parallel_mc/titan_strong_ms_impeff.pdf}
+\clearpage
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_ms_impeff.pdf}
   \end{center}
@@
       case. Considering iterations required to converge, the
-      performance of the split history cases at much better than the
+      performance of the split history cases is much better than the
       cases with history replication.}}
@@
 methods. At larger core counts, however, this reduction in compute
-time is not as large as at lower core counts due to the fact that the
-tally reduction across blocks requires more resources and those
-resources are clearly growing with core count (most obviously in the 4
-set case with split histories). In addition, at 131,072 cores for the
-largest 4 set computation, load balance is likely becoming an issue
-due to the linear system and subsequent parallel matrix and vector
-operations only occurring on a subset of the entire processor space of
-the problem while the Monte Carlo calculation is occurring on all
-processors.
+time is not as large due to the fact that the tally reduction across
+blocks requires more resources and those resources are clearly growing
+with core count (most obviously in the 4 set case with split
+histories). In addition, at 131,072 cores for the largest 4 set
+computation, processors that share a block may be some physical
+distatance from one another in the machine, adding to latency and
+decreasing performance.
 
@@
 given $P$ and $N$. When that problem is now solved using 2 sets we
-have $S=2$, $N$ and $2P$. For any number of sets we then have a global
-problem size of $N$ while the number of processors is $SP$ where $P$
-is the number of processors in the single set case. Therefore, $N/P$
-is no longer fixed in the weak scaling study but rather $N/(SP)$ is
-fixed. As a result, we modify the weak scaling absolute efficiency
-relationship to account for this fact:
+have $S=2$, $N$, and $2P$. For any number of sets we then have a
+global problem size of $N$ while the number of processors is $SP$
+where $P$ is the number of processors in the single set
+case. Therefore, $N/P$ is no longer fixed in the weak scaling study
+but rather $N/(SP)$ is fixed. As a result, we modify the weak scaling
+absolute efficiency relationship to account for this fact:
 \begin{equation}
@@
 new weak scaling formulation accounts for how efficiently those extra
-resources are incorporated into the problem. Without such a
-modification, the largest case run for the split 4 set case would have
-a computed weak scaling efficiency of nearly 100\% using the standard
-formula for absolute efficiency. However, using 4 times as many cores
-as the single set case increased the runtime for a fixed global
-problem size and therefore we would actually expect to compute a very
-low efficiency that shows this very poor use of resources.
+resources are incorporated into the problem.
 
@@
 replicated histories, the runtime of that calculation must be half of
-that for the single set case. For the 4 set case, this run time should
-be a quarter of the single set case for perfect scaling. This was not
-the observation for the runtimes in
-Figure~\ref{fig:titan_weak_ms_time} and therefore the single set case
-is still the most efficient use of computational resources in this
-scaling study.
+that for the single set case. This was not the observation for the
+runtimes in Figure~\ref{fig:titan_weak_ms_time} and therefore the
+single set case is still the most efficient use of computational
+resources in this scaling study.
 
@@
 rather than replicating histories in an attempt to reduce the number
-of iterations required to converge.  Here we also note that this is
-not a true weak scaling study as indicated by
+of iterations required to converge.  We also note that this is not a
+true weak scaling study as indicated by
 Eq~(\ref{eq:ms_weak_efficiency}) but rather an odd combination of both
@@
 subsequently the multiple sets and overlapping domain implementation,
-one may ask then ask if there is any need to communicate histories
-from domain-to-domain at all. Overlap was not as successful because
-because the additional communication required to reduce the
-overlapping tally vector eventually becomes larger than the
-communication savings during Monte Carlo transport. In addition, for
-local domains that are large enough to get good performance
-(e.g. those used in the weak scaling studies), then our analytic
-relationships for domain leakage, and in particular
-Eq~(\ref{eq:mean_chord_domain_leakage}), tell us that only about 5\%
-of the histories in this diffusion problem will leave the domain in
-the first place.  If that is true, then perhaps without any overlap or
-communication, just doing local Monte Carlo transport such that
+one may ask if there is any need to communicate histories from
+domain-to-domain at all. For local domains that are large enough to
+get good performance (e.g. those used in the weak scaling studies),
+our analytic relationships for domain leakage tell us that only about
+5\% of the histories in this diffusion problem will leave the domain
+in the first place.  If that is true, then perhaps without any overlap
+or communication, just doing local Monte Carlo transport such that
 histories are terminated upon reaching the domain boundary may be
@@
 occurs only in the subdomains casts the method as more of a true
-domain decomposition as outlined by Chapter 14 of Saad's text on
-iterative methods \cite{saad_iterative_2003}. Compared to traditional
-domain decomposition methods we see many similarities including the
-potential to overlap the subdomains to improve the results. By
-building the boundary data as outlined in
+domain decomposition scheme as outlined in Chapter 14 of Saad's text
+on iterative methods \cite{saad_iterative_2003}. Compared to
+traditional domain decomposition methods we see many similarities
+including the potential to overlap the subdomains to improve the
+results. By building the boundary data as outlined in
 \S~\ref{subsec:domain_generation}, we are effectively gathering one
@@
 subdomains to the others through the matrix-vector multiply operation
-after all subdomain Monte Carlo solves have been completed. Perhaps an
-even more important consequence of viewing MCSA as a stochastic
-additive Schwarz procedure is that it has the potential to more
-readily fit as an accelerator into other linear solver techniques
-including GMRES or potentially as a preconditioner.
+after all subdomain Monte Carlo solves have been completed.
 
@@
 efficiency as a function of core count due to the communication
-overhead for the parallel matrix and vector operations. Here we
-observe super unitary efficiencies for the subdomain case as there was
-a substantial improvement in runtimes relative to the 16-core base
-case with Monte Carlo communication.
+overhead for the parallel matrix and vector operations. We observe
+super unitary efficiencies for the subdomain case as there was a
+substantial improvement in runtimes relative to the 16-core base case
+with Monte Carlo communication.
 
@@
 overlap to do extra work in the local domain will not improve
-iterative performance and thus increase compute times.
-
-As a final analysis of MCSA using subdomain Monte Carlo, the multiple
-sets scaling studies are repeated to determine if any performance
-benefits are made from replication by eliminating Monte Carlo
-communication. Figure~\ref{fig:titan_strong_subdomain_ms_time} gives
-the runtimes for the strong scaling calculation and
-Figure~\ref{fig:titan_strong_subdomain_ms} the computed absolute
-efficiencies. For the weak scaling studies,
-Figure~\ref{fig:titan_weak_subdomain_ms_time} gives the wall times and
-Figure~\ref{fig:titan_weak_subdomain_ms} the efficiencies. For both
-scaling studies, the results are effectively the same as for the case
-with Monte Carlo communication. This is again due to the fact that
-adding sets to the problem is not only an exercise in strong scaling
-but it also adds additional parallel overhead in the form of the tally
-vector reduction across blocks.
-
-From a parallel performance perspective, MCSA may then be best used in
-a mode where the Monte Carlo is performed over only the subdomains
-with no replication or overlap added to the system. In this mode,
-iterative performance is maintained relative to the solutions with
-communication even for more ill-conditioned problems with larger
-spectral radii. Furthermore, no overlap was required to maintain
-iterative performance be letting histories near the boundary take a
-few more steps in their random walks. If in the future there are
-ill-conditioned problems where using subdomain Monte Carlo reduces the
-algorithmic efficiency of MCSA due to the information lost in the
-correction vector, overlap would be a viable mechanism to potentially
-maintain that iterative performance, keeping in mind the potential
-impacts on scalability that were observed in this work. However,
-although eliminating all Monte Carlo communication was observed to be
-the best mode of use for MCSA, the parallel Monte Carlo methods
-developed for this work still provide value for cases where only a
-Monte Carlo solution is desired. In those cases, full transport of the
-Monte Carlo histories are required in order to generate the correct
-solution.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_subdomain_ms_time.pdf}
-  \end{center}
-  \caption{\textbf{Strong scaling total wall time with multiple sets
-      and subdomain Monte Carlo.} \textit{Adding multiple sets has in
-      general a negative effect on the wall time as was observed for
-      the case with full Monte Carlo transport.}}
-  \label{fig:titan_strong_subdomain_ms_time}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_subdomain_ms.pdf}
-  \end{center}
-  \caption{\textbf{Strong scaling absolute efficiency for with
-      multiple sets and subdomain Monte Carlo.} \textit{Increased wall
-      times from adding parallel reduction operations for multiple
-      sets reduces parallel scalability at lower core counts but gives
-      a higher efficiency at the strong scaling wall.}}
-  \label{fig:titan_strong_subdomain_ms}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_subdomain_ms_time.pdf}
-  \end{center}
-  \caption{\textbf{Weak scaling total wall time with multiple sets and
-      subdomain Monte Carlo.} \textit{The addition of the set tally
-      reduction is significant with cost of the reduction observed as
-      core count grows.}}
-  \label{fig:titan_weak_subdomain_ms_time}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_subdomain_ms.pdf}
-  \end{center}
-  \caption{\textbf{Weak scaling absolute efficiency for with multiple
-      sets and subdomain Monte Carlo.} \textit{Weak scaling
-      efficiencies are not improved at large core counts when multiple
-      sets are added.}}
-  \label{fig:titan_weak_subdomain_ms}
-\end{figure}
+iterative performance and thus increase compute times. Supplemenatary
+scaling studies using the subdomain Neumann-Ulam method with multiple
+sets are presented in Appendix~\ref{chap:extra_scaling}.
 

### 2013-08-23 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
moving subomain nuemann ulam results for multiple sets into an appendix

@@
 %%---------------------------------------------------------------------------%%
-\section{The Neutron Transport Equation}
+\section{Derivation of the $SP_N$ Equations}
 \label{sec:transport_eq}
@@
 
-%%---------------------------------------------------------------------------%%
-\section{Derivation of the Monoenergetic $SP_N$ Equations}
-\label{sec:spn_equations}
-The $P_N$ equations as derived in \cite{lewis_computational_1993}
-give $N+1$ coupled first-order equations capturing the spatial and
+\subsection{Fixed-Source Monoenergetic $SP_N$ Equations}
+\label{subsec:fixed_source_spn}
+Using the one-dimensional Cartesian form of the transport equation,
+the $P_N$ equations as derived in \cite{lewis_computational_1993} give
+$N+1$ coupled first-order equations capturing the spatial and
 angular-dependence of the solution:
@@
 
-\subsection{Eigenvalue Form of the $SP_N$ Equations}
+\subsection{Eigenvalue Form of the Multigroup $SP_N$ Equations}
 \label{subsec:eigenvalue_form}

### 2013-08-23 doc/dissertation/document/backmatter/extra_scaling.tex
finished parallel corrections

@@
-\chapter{Subdomain Neumann-Ulam Scaling with Multiple Sets}
+\chapter{Subdomain Neumann-Ulam Scaling with Multiple Sets\ }
 \label{chap:extra_scaling}
@@
 vector reduction across blocks.
-
-From a parallel performance perspective, MCSA may then be best used in
-a mode where the Monte Carlo is performed over only the subdomains
-with no replication or overlap added to the system. In this mode,
-iterative performance is maintained relative to the solutions with
-communication even for more ill-conditioned problems with larger
-spectral radii. Furthermore, no overlap was required to maintain
-iterative performance be letting histories near the boundary take a
-few more steps in their random walks. If in the future there are
-ill-conditioned problems where using subdomain Monte Carlo reduces the
-algorithmic efficiency of MCSA due to the information lost in the
-correction vector, overlap would be a viable mechanism to potentially
-maintain that iterative performance, keeping in mind the potential
-impacts on scalability that were observed in this work. However,
-although eliminating all Monte Carlo communication was observed to be
-the best mode of use for MCSA, the parallel Monte Carlo methods
-developed for this work still provide value for cases where only a
-Monte Carlo solution is desired. In those cases, full transport of the
-Monte Carlo histories are required in order to generate the correct
-solution.
 

### 2013-08-23 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
finished parallel corrections

@@
 Figure~\ref{fig:titan_weak_subdomain} gives the results of the the
-weak scaling study in terms of absolute efficiencies. In both cases,
-the benefits to parallel performance of using MCSA as a subdomain
-solver are obvious. In the strong scaling case we make the important
-observation that we hit the strong scaling wall at the same time with
-and without the additional Monte Carlo communication. Although the
-Monte Carlo communication did contribute to the efficiency decrease,
-it is clear that the remaining matrix and vector operations are the
-primary impediment to scalability. Furthermore, we see an even greater
-rise above unity in efficiencies because there is less communication
+weak scaling study in terms of absolute efficiencies.
+
+In the strong scaling case we make the important observation that we
+hit the strong scaling wall at the same time with and without the
+additional Monte Carlo communication. Although the Monte Carlo
+communication did contribute to the efficiency decrease, it is clear
+that the remaining matrix and vector operations are the primary
+impediment to scalability. Furthermore, we see an even greater rise
+above unity in efficiencies because there is less communication
 overhead hiding the better cache utilization as the local problem size
@@
 with and eliminating Monte Carlo communication through subdomain
-solves provides an effectively linear boost in the weak scaling
+solves provides an effectively constant boost in the weak scaling
 absolute efficiency. We still see a small reduction in weak scaling
@@
 
-Of additional concern for the subdomain method is the potential
-decrease in iterative performance due to the fact that histories near
-the boundary will potentially have their random walk truncated, thus
-decreasing the information available to compute the correction
-vector. For both the weak and strong scaling study, all calculations
-with split histories converged in 22 or in only a few cases 21
-iterations for both the subdomain solve and the solve with
-communication giving an algorithmic efficiency of 1.048 for all
-cases. This small improvement in iteration count was not observed to
-be a function of problem size, number of processors, or whether the
-subdomain method was used. For the replicated history cases, the
-algorithmic efficiency improvement was not as good as when
-communication was allowed for the Monte Carlo algorithm. For these
-calculations, the number of iterations to converge was reduced to 18
-or 19 in both the strong and weak scaling studies with lower core
-counts converging in fewer iterations.
-
-Therefore, there is no reason to expect overlap to improve the
-iterative or scaling performance of the method as iterative
-performance is maintained without it for this particular
-problem. Overlap is not as effective because histories relinquish most
-of their useful information within the first few steps (which is why
-we can use such a large weight cutoff and there is an insensitivity to
-weight cutoff as shown in
-Figure~\ref{fig:estimator_wc_iters}). Keeping them around longer with
-overlap to do extra work in the local domain will not improve
-iterative performance and thus increase compute times. Supplemenatary
-scaling studies using the subdomain Neumann-Ulam method with multiple
-sets are presented in Appendix~\ref{chap:extra_scaling}.
-
-\clearpage
+From a parallel performance perspective, MCSA may then be best used in
+a mode where the Monte Carlo is performed over only the subdomains
+with no replication or overlap added to the system. In this mode,
+iterative performance is maintained relative to the solutions with
+communication even for more ill-conditioned problems with larger
+spectral radii. If in the future there are ill-conditioned problems
+where using subdomain Monte Carlo reduces the algorithmic efficiency
+of MCSA due to the information lost in the correction vector, overlap
+would be a viable mechanism to potentially maintain that iterative
+performance, keeping in mind the potential impacts on scalability that
+were observed in this work. However, although eliminating all Monte
+Carlo communication was observed to be the best mode of use for MCSA,
+the parallel Monte Carlo methods developed in this work still provide
+value for cases where only a Monte Carlo solution is desired. In those
+cases, full transport of the Monte Carlo histories are required in
+order to generate the correct solution.
 
@@
   realization of an additive Schwarz method
-\item With future optimization, large scale reactor analyses in which
-  the transport operator is fully formed (e.g. the $SP_N$ equations)
-  could experience improved time to solution on leadership-class
-  machines with MCSA given the improved scaling performance
 \end{itemize}

### 2013-08-23 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
spell check for parallel chapter

@@
 applied to the previous iterate's solution through a parallel vector
-update with parallell norm operations applied to check for
+update with parallel norm operations applied to check for
 convergence.
@@
 computation, processors that share a block may be some physical
-distatance from one another in the machine, adding to latency and
+distance from one another in the machine, adding to latency and
 decreasing performance.

### 2013-08-23 doc/dissertation/document/chapters/conclusion/conclusion.tex
spell check for nonlinear chapter
finished conclusion corrections

@@
 of requirements for both time to solution and the amount of memory
-required to contain the description of the entire problem and the
-solution. Looking forward to the next generation of machines,
-concurrency will go up and available memory will go down, putting
-pressure on physics application developers to research and develop
-solution techniques that can efficiently leverage this hardware. In
-nuclear reactor simulation, both neutron transport and fluid flow
-calculations consume a vast amount of computational time in order to
-properly characterize both steady state operational characteristics
-and transient accident scenarios. In both cases, we desire solution
-techniques that are aware of coming changes in hardware.
+required to contain the description of the entire problem. Looking
+forward to the next generation of machines, concurrency will go up and
+available memory will go down, putting pressure on physics application
+developers to research and develop solution techniques that can
+efficiently leverage this hardware. In nuclear reactor simulation,
+both neutron transport and fluid flow calculations consume a vast
+amount of computational time in order to properly characterize both
+steady state operational characteristics and transient accident
+scenarios. In both cases, we desire solution techniques that are aware
+of coming changes in hardware.
 
@@
 \label{sec:spn_conclusion}
-
 This work demonstrates the first application of MCSA to the neutron
@@
 fact that the transport operator generated in the eigenvalue
-calculation was very ill-conditioned due to the large amount of
-neutron scattering in the light water moderator. The more scattering
-present in the system, the longer a random walk will take and thus a
-spectral radius approaching unity was observed. A simpler neutron
-diffusion problem was used to demonstrate the breakdown of MCSA when
-spectral radii this large are generated.
+calculation was very ill-conditioned as a result of the large amount
+of neutron scattering in the light water moderator. The more
+scattering present in the system, the longer a random walk will take
+and thus a spectral radius approaching unity was observed. A simpler
+neutron diffusion problem was used to demonstrate the breakdown of
+MCSA when spectral radii this large are generated.
 
@@
 developed for this work generated very dense systems that consumed
-tremendous amounts of memory to store and compute time to
-generate. The memory problem was partially alleviated by applying the
-reduced domain approximation, however a significant memory overhead
-remained along with prohibitive time for construction.
+tremendous amounts of memory and compute time. The memory problem was
+partially alleviated by developing and applying the reduced domain
+approximation, however, a significant memory overhead remained along
+with prohibitive time for construction.
 
-Using the developed preconditioners and the reduced domain
-approximation, a set of $SP_N$ calculations using up to 4 energy
-groups was enabled for the fuel assembly problem. To verify MCSA, the
-fuel assembly problem was solved with different energy groups using
-both MCSA and two production Krylov methods. MCSA was observed to
-produce the same scalar flux in all groups at all spatial locations
-and generate the same k-eigenvalue in the same number of eigenvalue
-iterations when used in conjuction with an eigenvalue solver.
-
-For the same calculations, MCSA had qualitatively similar iterative
-performance to the production Krylov methods, showing improved
-performance over GMRES while a Conjugate Gradient-based method
-performed better using the same preconditioning. Improved performance
-over GMRES meets the goal of improving iterative performance to a
-certain extent but this was not the true for the Conjugate
-Gradient-based method. However, although the $SP_N$ equations are
-asymmetric for multigroup calculations, they are symmetric at the
-block level and therefore we expect the Conjugate Gradient-based
-method to perform better as this structure can be more readily
-leveraged to improve performance than with the more general GMRES and
-MCSA methods. In terms of CPU time, MCSA was observed to perform
-$O(100)$ times slower than the Krylov methods due to both lack of
-optimization in the random walk sequence and the explicit
-preconditioning strategy.
+To verify MCSA, the fuel assembly problem was solved with different
+energy groups using both MCSA and two production Krylov methods. MCSA
+was observed to produce the same k-eigenvalue in the same number of
+eigenvalue iterations when used in conjunction with an eigenvalue
+solver. For the same calculations, MCSA had qualitatively similar
+iterative performance to the production Krylov methods, showing
+improved performance over GMRES using the same preconditioning,
+meeting the goal of demonstrating improved iterative performance. In
+terms of CPU time, MCSA was observed to perform $O(100)$ times slower
+than the Krylov methods due to both lack of optimization in the random
+walk sequence and the explicit preconditioning strategy.
 
@@
 \label{sec:nonlinear_conclusions}
-
 This work presents the FANM method, a new nonlinear solution scheme
@@
 FANM solutions were observed to be numerically identical to those
-generated by the Newton-Krylov method. In addition, the benchmark
-problems also served as a vehicle to compare the performance of FANM
-with the Newton-Krylov method. For the thermal convection cavity
-problem, FANM was observed to converge in fewer linear solver
+generated by the Newton-Krylov method. For the thermal convection
+cavity problem, FANM was observed to converge in fewer linear solver
 iterations than Newton-Krylov at all Rayleigh numbers tested when the
@@
 
-The backward facing step problem at low Reynolds number gave better
+The backward facing step problem at low Reynolds numbers gave better
 performance with FANM converging in fewer linear solver
@@
 \label{sec:parallel_mc_conclusions}
-
 A new parallel MCSA algorithm was developed for this work to meet the
-goal of demonstrating the improved scalability of MCSA on
-leadership-class hardware. It was found the MCSA can indeed be
-effectively parallelized using the multiple-set overlapping-domain
-decomposition algorithm borrowed from the reactor physics
-community. Using a neutron diffusion problem, the parallel algorithm
-was verified to produce the same results as two production Krylov
-methods.
+goal of demonstrating the improved scalability on leadership-class
+hardware. It was found the MCSA can indeed be effectively parallelized
+using the multiple-set overlapping-domain decomposition algorithm
+borrowed from the reactor physics community. Using a neutron diffusion
+problem, the parallel algorithm was verified to produce the same
+results as two production Krylov methods.
 
@@
 observed to produce better parallel scaling results when compared to a
-production GMRES and Conjugate Gradient method. We note here that
-these scaling studies should be reconsidered if arithmetic
+production GMRES and Conjugate Gradient method. We note here, however,
+that these scaling studies should be reconsidered if arithmetic
 optimization of the code has been completed.
@@
 physical copy of the problem on the machine and can combine separate
-Monte Carlo solutions for each copy through superposition to enhance
-time to solution. Time to solution is then improved because fewer
-histories were run in each copy and therefore each MCSA iteration is
-faster or more global histories are computed and fewer MCSA iterations
-are required to converge.
+Monte Carlo solutions for each copy through superposition. Time to
+solution is then improved because fewer histories were run in each
+copy and therefore each MCSA iteration is faster or more global
+histories are computed and fewer MCSA iterations are required to
+converge.
 
@@
 domain-to-domain communication of histories was occurring in the first
-place and then subsequently verified using the analytic relationships
-for domain leakage. Using this idea, we implemented a subdomain
-Neumann-Ulam method with MCSA such that MCSA now takes the form of a
-stochastic realization of an additive Schwarz method. Scaling studies
-using this technique showed significant improvements in parallel
-scalability even when compared to the MCSA results when
-domain-to-domain communication was present, further enhancing the
-method and meeting the goal of improved scalability.
+place. Using this idea, we implemented a subdomain Neumann-Ulam method
+with MCSA such that MCSA now takes the form of a stochastic
+realization of an additive Schwarz method. Scaling studies using this
+technique showed significant improvements in parallel scalability even
+when compared to the MCSA results when domain-to-domain communication
+was present, further enhancing the method and meeting the goal of
+improved scalability.
 
@@
 
-Throughout this work, three salient issues were observed over all
-others. First, the performance of the MCSA implementation developed
-for this work was severely lacking in terms of time to solution when
-compared to production Krylov methods. In all cases, this difference
-was several orders of magnitude. Second, the preconditioning strategy
-developed by this work, although often successful in achieving
-convergence for difficult problems, creates many obstacles to a
-production implementation. These obstacles all primarily stem from the
-dense composite operators that are required to form the probabilities
-and weights for the Monte Carlo game. Third, perhaps the most
-restrictive and unattractive piece of MCSA as a whole is the simple
-fact that the spectral radius of the physics operator iteration matrix
-be less than one. This was demonstrated to prohibit convergence for
-cases where the preconditioning was considered aggressive and Krylov
-solvers exhibited excellent convergence properties. We will address
-each one of these problems in turn in this section and provide
-suggestions for their solution.
+Throughout this work, three salient issues were observed. First, the
+performance of the MCSA implementation developed for this work was
+severely lacking in terms of time to solution when compared to
+production Krylov methods. In all cases, this difference was several
+orders of magnitude. Second, the preconditioning strategy developed by
+this work, although often successful in achieving convergence for
+difficult problems, creates many obstacles to a production
+implementation. These obstacles all primarily stem from the dense
+composite operators that are required to form the probabilities and
+weights for the Monte Carlo game. Third, perhaps the most restrictive
+and unattractive piece of MCSA as a whole is the simple fact that the
+spectral radius of the physics operator iteration matrix be less than
+one. This was demonstrated to prohibit convergence for cases where the
+preconditioning was considered aggressive and Krylov solvers exhibited
+excellent convergence properties. We will address each one of these
+problems in turn in this section and provide suggestions for their
+solution.
 
@@
 \label{subsec:future_performance}
-
 When issues arising from preconditioning are not considered, there are
@@
 are those given by Eq~(\ref{eq:mc_walk_permutation}). For random walks
-that are traversing a domain which has been decomposed in a parallel
-environment, there are two sets of these states: one that represents
-the states over the entire global domain and one that represents them
-only over the local subdomain. Each state in the system gets a unique
-index in both of these sets that act as absolute coordinates in the
-random walk and permit the construction of the Monte Carlo estimates
-and other data needed for the solution.
+that are traversing a domain which has been decomposed, there are two
+sets of these states: one that represents the states over the entire
+global domain and one that represents them only over the local
+subdomain. Each state in the system gets a unique index in both of
+these sets that act as absolute coordinates in the random walk and
+permit the construction of the Monte Carlo estimates and other data
+needed for the solution.
 
@@
 
-A second mode of optimization to improve performance regards how
+A second mode of optimization to improve the performance of how
 discrete probability distribution functions are sampled in the
@@
 
-Nonlinear performance may also be improved by a closer relationship
-between MCSA and the Newton iteration. We saw in our research how the
-forcing term selected very tightly controls the convergence of the
-linear model in FANM. For MCSA, in addition to the residual tolerance
-that is dictated by this forcing term, convergence may also be
-controlled by adjusting the number of stochastic histories computed at
-each MCSA iteration. As the nonlinear iteration evolves, we may find
-it useful to adjust the number of histories used at each MCSA
-iteration in the same way as we adjust the convergence tolerance of
-the linear model through the forcing term. Future research is
-suggested to analyze the effects of the number of stochastic histories
-on the convergence of the linear and nonlinear models and how this may
-be controlled by the forcing term or another parameter.
-
-With these implementation improvements perhaps MCSA and FANM timing
-may approach that of production Krylov methods for equally conditioned
-problems. Next we discuss potential research in the area of
-preconditioning which may enable performance research for MCSA in the
-area of memory consumption.
-
 \subsection{Preconditioning}
 \label{subsec:future_preconditioning}
-
 In general, preconditioning developments in this work enabled much of
@@
 incurred for inverting the preconditioners. We therefore propose the
-following three developments in preconditioning for Monte Carlo
-methods:
+following two developments in preconditioning for Monte Carlo
+methods.
 
-\paragraph{\textbf{Research and develop alternative sampling schemes for
-  the general left/right preconditioned Monte Carlo method not relying
-  on composite matrix generation.}} Developments that break away from
-matrix-matrix multiplications within the preconditioned solver
-framework improve scalability. This means avoiding the generation of
-the composite operator and instead playing the Monte Carlo game with
-weights and probabilities formed by considering the original linear
-operator and the preconditioners individually. This individual
-consideration may come from an alternative algebraic formulation of
-the general preconditioned MCSA method in
-Eq~(\ref{eq:left_right_mcsa}) or from an alternative random sampling
-strategy for the Monte Carlo game.
-\paragraph{\textbf{Develop Monte Carlo variance reduction based
-  preconditioning strategies that do not rely on an algebraic
-  preconditioner.}} Constructing the complete inverse of the
-  preconditioner is the core time and scalability constraint when
-  modern algebraic preconditioners are used. Building the inverse for
+\subsubsection{Variance Reduction-Based Preconditioning}
+Traditional Monte Carlo variance reduction techniques instead of
+preconditioning strategies that rely on algebraic preconditioners
+should be investigated. Constructing the complete inverse of the
+preconditioner is the core time and scalability constraint when modern
+algebraic preconditioners are used. Building the inverse for
 preconditioners where a sparsity pattern is not enforced typically
 results in a dense matrix that destroys scalability for domain
-  decomposed parallel computations. Alleviating both of these issues
-  by stochastic preconditioning methods should be studied. Decades of
+decomposed parallel computations. Alleviating both of these issues by
+stochastic preconditioning methods should be studied. Decades of
 research and development in Monte Carlo variance reduction for
-  particle transport \cite{booth_1994} may be leveraged to form
-  similar strategies for linear systems. In the same way that this
-  work has been able to adapt modern particle transport strategies for
-  parallelism and apply them to linear systems, the same strategy may
-  possibly be applied to variance reduction techniques to improve MCSA
-  convergence by modifying the weights and probabilities of the Monte
-  Carlo game as an effective means of preconditioning.
-\paragraph{\textbf{Study physics or PDE-based preconditioners for linear
-  systems that produce a reduced-order model of the original system to
-  which the Monte Carlo method may be applied.}} As the Monte Carlo
-method serves to construct the solution correction vector in the MCSA
-method, the Monte Carlo method may be applied to a reduced-order
+particle transport \cite{booth_1994} may be leveraged to form similar
+strategies for linear systems. In the same way that this work has been
+able to adapt modern particle transport strategies for parallelism and
+apply them to linear systems, the same strategy may possibly be
+applied to variance reduction techniques to improve MCSA convergence
+by modifying the weights and probabilities of the Monte Carlo game as
+an effective means of preconditioning.
+
+\subsubsection{Physics and PDE-based Preconditioners}
+Studies should be considered for physics or PDE-based preconditioners
+for linear systems that produce a reduced-order model of the original
+system to which the Monte Carlo method may be applied. As the Monte
+Carlo method serves to construct the solution correction vector in the
+MCSA method, the Monte Carlo method may be applied to a reduced-order
 physics or PDE-based system that captures the essence of the original
@@
 \label{subsec:future_spec_rad}
-
 All numerical methods for solving linear problems require quality
@@
 condition is achieved convergence may be incredibly slow. Therefore,
-we propose two elements of future work that aim to move away from this
-restriction, both fundamentally based on sampling a sequence not based
-on the Neumann series.
+we propose future work that aims to move away from this restriction
+that is fundamentally based on sampling a sequence not based on the
+Neumann series.
 
@@
 
-\subsubsection{Stochastic Projection Methods}
-\label{subsubsec:stochastic_projection_methods}
-
-As a final means of improvement we consider stochastic projection
-methods as developed by Sabelfeld \cite{hutchison_stochastic_2011}. In
-his work, Sabelfeld claims his method is 'beyond Markov chains and von
-Neumann-Ulam scheme'. To a certain extent this is true; his method is
-based on a projection method and should always converge. In his work,
-he utilizes a Kaczmarz iteration:
-\begin{equation}
-  \ve{x}^{k+1} = \ve{x}^k + \lambda_k \frac{b_i - \langle \ve{a}_i
-    \ve{x}^k \rangle}{||\ve{a}_i||}\ve{a}_i\:,
-  \label{eq:kaczmarz_iteration}
-\end{equation}
-where $\ve{a}_i$ is the $i^{th}$ row vector of the operator and
-$\lambda_k$ is a relaxation parameter.
-
-The stochastic nature of the algorithm comes from effectively
-modifying the iteration sequence such that the value of $i$ is now
-selected at random. This is interesting but arguably does not serve
-our purposes at well. We have an interest in a fully stochastic
-process like a random walk sequence because of its potential benefits
-in terms of resiliency and the ability to use superposition to combine
-results from replicated problems to enhance performance. In
-Sabelfeld's work, the randomness is used to select which deterministic
-iteration step will be performed and does not meet either of these
-requirements. However, that does not mean there is nothing to learn
-from this work. If one writes Eq~\ref{eq:kaczmarz_iteration} in terms
-of applications of projection operators as that outlined in
-\cite{tanabe_projection_1971} we see that this actually takes a
-similar form as the Neumann series and could therefore be used in a
-random walk sequence. In addition, perhaps the above algorithm can be
-modified such that the randomness of the projection sequence permits
-improved results via the superposition principle. Research studying
-random walks based on projection sequences is therefore suggested as a
-means of moving beyond the spectral radius limitation.
-
 %%---------------------------------------------------------------------------%%

### 2013-08-23 doc/dissertation/document/frontmatter/acks.tex
updating acks

@@
+Great thanks are owed to my advisor and friend Paul Wilson for
+graciously giving me years of guidance and the opportunity to develop
+this work. He provided me with both the resources and challenges to
+grow as a researcher. Without his influence and support, this work
+would have never been possible.
+
+Tom Evans is responsible for presenting me with the seeds of this work
+and for that I thank him. His mentoring over the past few years has
+been invaluable and the skills and lessons I have learned from him
+have proven indispensable.
+
+Roger Pawlowski and other members of the Consortium for Advanced
+Simulation of Light Water Reactors and Trilinos teams as well as many
+staff members at Oak Ridge National Laboratory have provided
+tremendous resources for my professional and technical development and
+have greatly facilitated this work.
+
 This work was performed under appointment to the Nuclear Regulatory

### 2013-08-26 doc/dissertation/document/backmatter/estimator_variance.tex
moving estimator variance into an appendix

--- 2013-08-22 doc/dissertation/document/chapters/mc_background/mc_background.tex
@@
-\chapter{Monte Carlo Synthetic Acceleration\\ Methods}
-\label{ch:stochastic_methods}
-An alternative approach to approximate matrix inversion is to employ
-Monte Carlo methods that sample a distribution with an expectation
-value equivalent to that of the inverted operator. Such methods have
-been in existence for decades with the earliest reference noted here a
-manuscript published in 1950 by Forsythe and Leibler
-\cite{forsythe_matrix_1950}. In their outline, Forsythe and Leibler
-in fact credit the creation of this technique to J. Von Neumann and
-S.M. Ulam some years earlier than its publication. In 1952 Wasow
-provided a more formal explanation of Von Neumann and Ulam's method
-\cite{wasow_note_1952} and Hammersley and Handscomb's 1964 monograph
-\cite{hammersley_monte_1964} and Spanier and Gelbard's 1969 book
-\cite{spanier_monte_1969} present additional detail on this topic
-using a collection of references from the 1950's and early
-1960's.
+\chapter{Monte Carlo Estimator Variance\ }
+\label{chap:estimator_variance}
 
-Following this work, Halton presented a residual Monte Carlo scheme
-that improved dramatically on the performance of these methods
-\cite{halton_sequential_1962}. Recent work in this area has utilized
-these residual Monte Carlo methods in particle transport
-\cite{evans_residual_2003}, giving accelerated convergence over
-traditional methods. As an expansion of Halton's work, Evans and
-others have leveraged these methods in a synthetic acceleration scheme
-that has been shown to be competitive with production Krylov methods
-for radiation transport problems and improve performance significantly
-over Halton's sequential method \cite{evans_monte_2012}.
+In this appendix we briefly discuss the variance of the three primary
+Monte Carlo estimators used with the Neumann-Ulam method in this work.
 
-In this chapter, we will present the fundamentals of the Monte Carlo
-method for discrete linear systems. Both the forward and adjoint
-methods will be presented and analyzed including a brief variance
-analysis of several Monte Carlo estimators. The Sequential Monte Carlo
-method of Halton and Monte Carlo Synthetic Acceleration methods will
-then be presented as a means of leveraging the basic Monte Carlo
-methods devised by Neumann and Ulam in an iterative refinement scheme
-that accelerates their convergence. Using these iterative schemes,
-the effects of the algorithm parameters on solutions to a simple model
-transport problem are explored.
-
-\section{Preliminaries}
-\label{sec:linear_preliminaries}
-We seek solutions of the general linear problem in the following form:
-\begin{equation}
-  \ve{A} \ve{x} = \ve{b}\:,
-  \label{eq:linear_problem}
-\end{equation}
-where $\ve{A} \in \mathbb{R}^{N \times N}$ is a matrix operator such
-that $\ve{A} : \mathbb{R}^{N} \rightarrow \mathbb{R}^{N}$, $\ve{x} \in
-\mathbb{R}^N$ is the solution vector, and $\ve{b} \in \mathbb{R}^N$ is
-the forcing term. The solutions to Eq~(\ref{eq:linear_problem}) will
-be generated by inverting $\ve{A}$ either directly or indirectly:
-\begin{equation}
-  \ve{x} = \ve{A}^{-1} \ve{b}
-  \label{eq:linear_problem_solution}\:.
-\end{equation}
-In addition we can define the residual:
-\begin{equation}
-  \ve{r} = \ve{b} - \ve{A}\ve{x}\:,
-  \label{eq:linear_residual}
-\end{equation}
-such that an exact solution $\ve{x}$ has been found when
-$\ve{r}=\ve{0}$.  From the statement in
-Eq~(\ref{eq:linear_problem_solution}) we can already place a
-restriction on $\ve{A}$ by requiring that it be \textit{nonsingular},
-meaning that we can in fact compute $\ve{A}^{-1}$. In this work we
-will focus our efforts on approximately inverting the operator through
-various means.
-
-In a discussion of methods for solving linear systems, several
-mathematical tools are useful in characterizing the qualities of the
-linear system. Among the most useful are the \textit{eigenvalues} of
-the matrix, $\sigma(\ve{A})$. We find these by solving the eigenvalue
-problem:
-\begin{equation}
-  \ve{A} \ve{x} = \lambda \ve{x},\ \lambda \in \sigma(\ve{A})\:.
-  \label{eq:eigenvalue_problem}
-\end{equation}
-By writing Eq~(\ref{eq:eigenvalue_problem}) in a different form,
-\begin{equation}
-  (\ve{A} - \lambda \ve{I})\ve{x} = 0 \:,
-  \label{eq:eigenvalue_problem_2}
-\end{equation}
-and demanding that non-trivial solutions for $\ve{x}$ exist, it is
-then required that $|\ve{A} - \lambda \ve{I}| = 0$. Expanding this
-determinant yields a characteristic polynomial in terms of $\lambda$
-with roots that form the set of eigenvalues, $\sigma(\ve{A})$. Each
-component of $\sigma(\ve{A})$ can then be used to solve
-Eq~(\ref{eq:eigenvalue_problem_2}) for a particular permutation of
-$\ve{x}$. The set of all permutations form the \textit{eigenvectors}
-of $\ve{A}$. A quantity of particular interest that is computable from
-the eigenvalues of a matrix $\ve{A}$ is the \textit{spectral radius},
-$\rho(\ve{A})$, defined by Saad \cite{saad_iterative_2003} as:
-\begin{equation}
-  \rho(\ve{A}) = \max_{\lambda \in \sigma(\ve{A})} |\lambda| \:,
-  \label{eq:spectral_radius}
-\end{equation}
-which is the radius of the largest circle in the complex plane
-centered about the origin that encapsulates the entire eigenvalue
-spectrum of the operator.
-
-%%---------------------------------------------------------------------------%%
-\section{Neumann-Ulam Method}
-\label{sec:mc_preliminaries}
-We begin our discussion of Monte Carlo methods by seeking a solution
-to Eq~(\ref{eq:linear_problem}). For a given linear operator $\ve{A}$,
-we can use diagonal splitting\footnote{It should be noted that
-  non-diagonal splittings have been recently explored
-  \cite{srinivasan_monte_2010} and have the potential to improve
-  efficiency. However, it was observed in this work that this type of
-  splitting did not improve performance in the asymptotic limit of
-  $\rho(\ve{H}) \rightarrow 1$ for non-trivial problems.} similar to
-the stationary method in Eq~(\ref{eq:linear_split_equation2}) to
-define the \textit{iteration matrix}, $\ve{H}$:
-\begin{equation}
-  \ve{H} = \ve{I} - \ve{A}\:,
-  \label{eq:linear_mc_iteration_matrix}
-\end{equation}
-such that we are solving the system:
-\begin{equation}
-  \ve{x} = \ve{H} \ve{x} + \ve{b}\:.
-  \label{eq:richardson_split}
-\end{equation}
-We can then form an alternative representation for $\ve{A}^{-1}$ by
-generating the \textit{Neumann series}:
-\begin{equation}
-  \ve{A}^{-1} = (\ve{I}-\ve{H})^{-1} = \sum_{k=0}^{\infty} \ve{H}^k\:,
-  \label{eq:neumann_series}
-\end{equation}
-which will converge if the spectral radius of $\ve{H}$ is less than
-1. If we then apply this Neumann series to the right hand side of
-Eq~(\ref{eq:linear_problem}) we acquire the solution to the linear
-problem:
-\begin{equation}
-  \ve{A}^{-1}\ve{b} = \sum_{k=0}^{\infty} \ve{H}^k\ve{b} = \ve{x}\:.
-  \label{eq:neumann_solution}
-\end{equation}
-An approximation of this summation by truncation will therefore lead
-to an approximation of the solution. If we expand the summation with a
-succession of matrix-vector multiply operations, we arrive at an
-alternative perspective of this summation by considering the $i^{th}$
-component of the solution vector:
-\begin{equation}
-  x_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-  \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}\:,
-  \label{eq:expanded_neumann_solution}
-\end{equation}
-which can interpreted as a series of transitions between states,
-\begin{equation}
- \nu = i \rightarrow i_1 \rightarrow \cdots \rightarrow i_{k-1}
- \rightarrow i_{k}\:,
-  \label{eq:mc_walk_permutation}
-\end{equation}
-in $\ve{H}$ where $\nu$ is interpreted as a particular random walk
-sequence permutation. We can generate these sequences of transitions
-through Monte Carlo random walks by assigning them both a probability
-and weight. As a reinterpretation of the iteration matrix, we then
-form the \textit{Neumann-Ulam decomposition} of \ve{H}:
-\begin{equation}
-  \ve{H} = \ve{P} \circ \ve{W}\:,
-  \label{eq:neumann_ulam_decomposition}
-\end{equation}
-where $\circ$ denotes the Hadamard product operation\footnote{The
-  Hadamard product $\ve{A} = \ve{B} \circ \ve{C}$ is defined
-  element-wise as $a_{ij} = b_{ij} c_{ij}$.}, $\ve{P}$ denotes the
-transition probability matrix, and $\ve{W}$ denotes the transition
-weight matrix. This decomposition, a generalization of Dimov's work
-\cite{dimov_new_1998}, is an extension of the original Neumann-Ulam
-scheme in that now a weight cutoff can be used to terminate a random
-walk sequence and therefore truncate the Neumann series it is
-approximating. The formulation of $\ve{P}$ and $\ve{W}$ will be
-dependent on whether we choose a direct or adjoint Monte Carlo
-sequence to estimate the state transitions in
-Eq~(\ref{eq:expanded_neumann_solution}). In the direct method, we will
-use the provided linear operator $\ve{A}$ to form the Neumann Ulam
-decomposition while the adjoint method will use the adjoint linear
-operator ($\ve{A}^T$ for real-valued systems) to form the
-decomposition.
-
-%%---------------------------------------------------------------------------%%
-\section{Direct Monte Carlo Method}
-\label{sec:direct_mc}
-In the context of matrix inversion, a direct (forward) method
-resembles an adjoint Monte Carlo method in the reactor physics
-community where the solution state is sampled and the source terms
-that contribute to it are assembled. To achieve this, we build the
-forward Neumann-Ulam decomposition per Dimov's approach by first
-choosing a probability matrix that is a row scaling of $\ve{H}$ such
-that its components are:
-\begin{equation}
-  p_{ij} = \frac{|h_{ij}|}{\sum_j |h_{ij}|}\:.
-  \label{eq:direct_probability}
-\end{equation}
-From this, we then see that the probability of transitioning from a
-state $i$ to a state $j$ is implicitly linked to the original operator
-$\ve{A}$ in that those terms with large values, and therefore those
-that make the greatest contribution to the numerical solution, will be
-sampled with a higher probability than smaller terms. In addition, the
-row scaling provides a normalization over the state to which we are
-transitioning such that $\sum_j p_{ij} = 1$, meaning that we sample
-the probabilities over the rows of the matrix. The components of
-the weight matrix are then defined by
-Eq~(\ref{eq:neumann_ulam_decomposition}) as:
-\begin{equation}
-  w_{ij} = \frac{h_{ij}}{p_{ij}}\:.
-  \label{eq:direct_weight}
-\end{equation}
-It should be noted here that if $\ve{A}$ is sparse, then $\ve{H}$,
-$\ve{P}$, and $\ve{W}$ must be sparse as well by
-definition. Additionally, we only compute $\ve{P}$ and $\ve{W}$ from
-the non-zero elements of $\ve{H}$ as those components that are zero
-will not participate in the random walk and thus produce identical
-sparsity patterns for all matrices.
-
-Using these matrices, we can then form the expectation value of the
-forward solution. For a given random walk permutation $\nu$, we define
-the weight of that permutation on the $m^{th}$ step to be:
-\begin{equation}
-  W_{m} = W_0 w_{i_0,i_1} w_{i_1,i_2} \cdots w_{i_{m-1},i_m}\:,
-  \label{eq:direct_permutation_weight}
-\end{equation}
-such that the weight of each transition event contributes to the total
-through multiplication with $W_0 = 1$ as the starting weight. The
-contribution to the solution from a particular random walk permutation
-with $k$ total events is then the \textit{forward estimator}:
-\begin{equation}
-  X_{i_0 = i}(\nu) = \sum_{m=0}^k W_{m} b_{i_m}\:,
-  \label{eq:direct_permutation_contribution}
-\end{equation}
-where $X_{i_0 = i}(\nu)$ signifies that the solution state, $i_0$, in
-which the random walk $\nu$ started is also the state, $i$, in which
-we are tallying. We then define the probability that a particular
-random walk permutation of $k$ events will occur:
-\begin{equation}
-  P_{\nu} = P_{i_0=i} p_{i,i_1} p_{i_1,i_2} \cdots p_{i_{k-1},i_k}\:,
-  \label{eq:direct_permutation_probability}
-\end{equation}
-with $P_{i_0=i} = 1$ for the direct method. Finally, we define the
-expectation value of $X$ to be the collection of all random walk
-permutations and their probabilities:
-\begin{equation}
-  E\{X_i\} = \sum_{\nu} P_{\nu} X_{i}(\nu)\:,
-  \label{eq:direct_expectation_value}
-\end{equation}
-which, if expanded, directly recovers the exact solution by forming
-the Neumann series:
-\begin{equation}
-  \begin{split}
-    E\{X_i\}
-    &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-    \sum_{i_k}^{N} p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
-    w_{i,i_1}w_{i_1,i_2}\ldots w_{i_{k-1},i_k} b_{i_k}\\ &= x_i\:,
-  \end{split}
-  \label{eq:direct_expectation_expansion}
-\end{equation}
-therefore providing an unbiased Monte Carlo estimator.
-
-In cases where we seek only approximate solutions, we need only to
-perform a predetermined number of random walks in order to generate an
-approximation for $\ve{x}$. If we are only to approximate the
-solution, we also need conditions by which we may terminate a random
-walk as the Neumann Ulam decomposition defined by
-Eqs~(\ref{eq:direct_probability}) and (\ref{eq:direct_weight}) will
-create a random walk weight in Eq~(\ref{eq:direct_permutation_weight})
-that approaches, but never reaches zero. We do this by noticing that
-if the problem is convergent, the factors added to
-Eq~(\ref{eq:direct_permutation_weight}) will make $W_m$ become
-diminishingly small due to their definition in
-Eq~(\ref{eq:direct_weight}) and therefore the contribution of the
-history to the solution estimate will become negligible. Using this,
-we choose terminate a random walk sequence with a \textit{weight
-  cutoff}, $W_c$, that is enforced when $W_m < W_c$ for a particular
-random walk permutation.
-
-\subsection{Forward Estimator Variance}
+\section{Forward Estimator Variance\ }
 \label{subsec:forward_variance}
@@
 \begin{equation}
-  |\ve{H} \circ \ve{W}| \geq |\ve{H}|\ |\ve{W}|\:.
+  ||\ve{H} \circ \ve{W}|| \geq ||\ve{H}||\ ||\ve{W}||\:.
   \label{eq:hadamard_inequality}
@@
 \begin{equation}
-  \frac{|\ve{H}|}{|\ve{P}|} \geq |\ve{W}|\:.
+  \frac{||\ve{H}||}{||\ve{P}||} \geq ||\ve{W}||\:.
   \label{eq:weight_norm}
 \end{equation}
-Choosing the infinity norm of the operator, the row normalized
-probability matrix will yield a norm of 1 giving the following
-inequality for relating $\ve{G}$ and $\ve{H}$:
+Choosing the 2-norm of the operators, if we assume symmetry of
+$\ve{H}$ and therefore $\ve{P}$, $\ve{W}$, and $\ve{G}$, we have
+\cite{leveque_finite_2007}:
 \begin{equation}
-  |\ve{G}| \geq |\ve{H}|^2
+  ||\ve{H}||_2 = \rho(\ve{H})\:.
+  \label{eq:matrix_2_norm}
+\end{equation}
+The row normalized probability matrix will yield a norm of 1 giving
+the following inequality for relating $\ve{G}$ and $\ve{H}$:
+\begin{equation}
+  ||\ve{G}||_2 \geq ||\ve{H}||^2_2\:,
   \label{eq:variance_norm_inequality}
+\end{equation}
+or
+\begin{equation}
+  \rho(\ve{G}) \geq \rho(\ve{H})^2\:.
+  \label{eq:variance_norm_inequality_2}
 \end{equation}
@@
 to alleviate this and therefore restrict $\ve{H}$ due to
-Eq~(\ref{eq:variance_norm_inequality}) with $\rho(\ve{H}) < 1$ so that
+Eq~(\ref{eq:variance_norm_inequality_2}) with $\rho(\ve{H}) < 1$ so that
 our expectation values for the solution may have a finite variance.
 
-\subsection{Direct Method: Evolution of a Solution}
-\label{subsec:direct_evolution}
-As a means of visually demonstrating the direct Monte Carlo method,
-consider a 2-dimensional thermal diffusion problem with sources on the
-left and right hand sides of the domain and a uniform source
-throughout the domain of 1/5 the strength of the boundary sources as
-shown in Figure~\ref{fig:heat_setup}.
-\begin{figure}[t!]
-  \begin{center}
-    \scalebox{1.2}{ \input{chapters/mc_background/heat_eq_setup.pdftex_t} }
-  \end{center}
-  \caption{\textbf{Problem setup for 2D heat equation.}
-    \textit{Dirichlet conditions are set for the temperature on all 4
-      boundaries of the Cartesian grid. Background source of 1/5 the
-      value of the boundary sources present. $50 \times 50$ grid.}}
-  \label{fig:heat_setup}
-\end{figure}
-For this problem, the number of histories used to compute the solution
-at each grid point (state) in the domain was increased from 1 to 1000
-in order to demonstrate the effects on the solution and the
-statistical nature of the method. Figure~\ref{fig:direct_evolution}
-gives these results.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/mc_background/direct_evolution.png}
-  \end{center}
-  \caption{\textbf{Direct Monte Carlo solution to the heat equation
-      with varying numbers of histories.} \textit{Top left: 1 history
-      per state. Top right: 10 histories per state. Bottom left: 100
-      histories per state. Bottom right: 1000 histories per state.}}
-  \label{fig:direct_evolution}
-\end{figure}
-As the number of histories used per state is increased, the
-statistical variance of the solutions is decreased as more tallies are
-made. Starting with a single history at each state in the domain, the
-high variance prevents a precise solution from being obtained although
-we begin to see the solution take shape as expected. It is interesting
-to note here that as the statistical uncertainty is reduced at each
-grid point in the domain, the solution is resolved in a certain sense,
-analogous to the convergence of a traditional iterative method.
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{Adjoint Monte Carlo Method}
-\label{sec:adjoint_mc}
-Often a more useful form, an alternative to forward Monte Carlo matrix
-inversion is the adjoint method. We begin by defining the linear
-system adjoint to Eq~(\ref{eq:linear_problem}):
-\begin{equation}
-  \ve{A}^T \ve{y} = \ve{d}\:,
-  \label{eq:adjoint_linear_problem}
-\end{equation}
-where $\ve{y}$ and $\ve{d}$ are the adjoint solution and source
-vectors respectively and $\ve{A}^T$ is the adjoint operator for
-$\ve{A} \in \mathbb{R}^{N \times N}$. We can split this equation to
-mirror Eq~(\ref{eq:richardson_split}):
-\begin{equation}
-  \ve{y} = \ve{H}^T \ve{y} + \ve{d}\:.
-  \label{eq:adjoint_split_system}
-\end{equation}
-As was required for convergence with the direct method using
-Eq~(\ref{eq:richardson_split}), the spectral radius of $\ve{H}$ must
-remain less than 1 as $\ve{H}^T$ contains the same eigenvalues and
-therefore has the same spectral radius. By defining the following
-inner product equivalence \cite{spanier_monte_1969}:
-\begin{equation}
-  \langle \ve{A}^T \ve{y}, \ve{x} \rangle = \langle \ve{y}, \ve{A}
-  \ve{x} \rangle\:.
-  \label{eq:adjoint_operator_product}
-\end{equation}
-it follows that:
-\begin{equation}
-  \langle \ve{x}, \ve{d} \rangle = \langle \ve{y}, \ve{b} \rangle\:.
-  \label{eq:adjoint_vector_relation}
-\end{equation}
-Using these definitions, we can derive an estimator from the adjoint
-method that will also give the solution vector, $\ve{x}$. As with the
-direct method, we can acquire the adjoint solution by forming the
-Neumann series by writing Eq~(\ref{eq:adjoint_split_system}) as:
-\begin{equation}
-  \ve{y} = (\ve{I} - \ve{H}^T)^{-1} \ve{d}\:,
-  \label{eq:adjoint_split_system_2}
-\end{equation}
-which in turn yields the Neumann series using the adjoint operator:
-\begin{equation}
-  \ve{y} = \sum_{k=0}^{\infty} (\ve{H}^T)^k\ve{d}\:.
-  \label{eq:adjoint_neumann_series}
-\end{equation}
-We expand this summation to again yield a series of transitions that
-can be approximated by a Monte Carlo random walk sequence, this time
-forming the Neumann series in reverse order:
-\begin{equation}
-  y_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-  \sum_{i_k}^{N}h_{i_k,i_{k-1}}\ldots h_{i_2,i_1} h_{i_1,i} d_{i_k}\:.
-  \label{eq:adjoint_neumann_solution}
-\end{equation}
-We can readily build an estimator for the adjoint solution from this
-series expansion, but we instead desire the solution to
-Eq~(\ref{eq:linear_problem}). Here we have 2 unknowns, $\ve{y}$ and
-$\ve{d}$, and therefore we require two constraints to close the
-system. We use Eq~(\ref{eq:adjoint_vector_relation}) as the first
-constraint and as a second constraint we select:
-\begin{equation}
-  \ve{d} = \boldsymbol{\delta}_j\:,
-  \label{eq:adjoint_second_constraint}
-\end{equation}
-where $\boldsymbol{\delta}_j$ is one of a set of vectors in which the
-$j^{th}$ component is the Kronecker delta function $\delta_{i,j}$. If
-we apply Eq~(\ref{eq:adjoint_second_constraint}) to our first
-constraint Eq~(\ref{eq:adjoint_vector_relation}), we get the following
-convenient outcome:
-\begin{equation}
-  \langle \ve{y}, \ve{b} \rangle = \langle \ve{x},
-  \boldsymbol{\delta}_j \rangle = x_j \:,
-  \label{eq:inner_product_constraint}
-\end{equation}
-meaning that if we compute the inner product of the original source
-and the adjoint solution using a delta function source, we recover one
-component of the original solution.
-
-In terms of radiation transport, this adjoint method is equivalent to
-a traditional forward method where the initial state $i_0$ of the
-random walk is determined by sampling the source vector $\ve{b}$ with
-probabilities:
-\begin{equation}
-  P_{i_0=i}(\nu) = \frac{|b_i|}{||\ve{b}||_1}\:,
-  \label{eq:adjoint_source_probability}
-\end{equation}
-with a random walk starting weight of:
-\begin{equation}
-  W_0 = ||\ve{b}||_1 \frac{b_i}{|b_i|}\:,
-  \label{eq:adjoint_starting_weight}
-\end{equation}
-which gives the additional useful relation:
-\begin{equation}
-  b_{i_0} = W_0 P_{i_0=i}\:.
-  \label{eq:adjoint_source_definition}
-\end{equation}
-As a result of using the adjoint system, we modify our probabilities
-and weights using the \textit{adjoint Neumann-Ulam decomposition} of
-$\ve{H}$:
-\begin{equation}
-  \ve{H}^{T} = \ve{P} \circ \ve{W}\:,
-  \label{eq:adjoint_neumann_ulam}
-\end{equation}
-where now we are forming the decomposition with respect to the
-transpose of $\ve{H}$. We then follow the same procedure as the direct
-method for forming the probability and weight matrices in the
-decomposition. Using the adjoint form, probabilities should instead be
-column-scaled:
-\begin{equation}
-  p_{ij} = \frac{|h_{ji}|}{\sum_j |h_{ji}|}\:,
-  \label{eq:adjoint_probability}
-\end{equation}
-such that we expect to select a new state, $j$, from the current state
-in the random walk, $i$, by sampling column-wise (or row-wise if an
-adjoint probability matrix is formed). Per
-Eq~(\ref{eq:adjoint_neumann_ulam}), the transition weight is then
-defined as:
-\begin{equation}
-  w_{ij} = \frac{h_{ji}}{p_{ij}}\:.
-  \label{eq:adjoint_weight}
-\end{equation}
-Using the decomposition we can then define an expectation value for
-the adjoint method. Given Eq~(\ref{eq:direct_permutation_weight}) as
-the weight generated for a particular random walk permutation as in
-Eq~(\ref{eq:mc_walk_permutation}) and our result from
-Eq~(\ref{eq:inner_product_constraint}) generated by applying the
-adjoint constraints, the contribution to the solution in state $j$
-from a particular random walk permutation of $k$ events is then the
-\textit{collision estimator}:
-\begin{equation}
-  X_{j}(\nu) = \sum_{m=0}^k W_{m} \delta_{i_m,j}\:,
-  \label{eq:adjoint_permutation_contribution}
-\end{equation}
-where the Kronecker delta indicates that the tally contributes only in
-the current state, $i_m$, of the random walk.  Note here that the
-estimator in Eq~(\ref{eq:adjoint_permutation_contribution}) does not
-have a dependency on the source state as in
-Eq~(\ref{eq:direct_permutation_contribution}), providing a remedy for
-the situation in the direct method where we must start a random walk
-in each source state for every permutation if we want to compute a
-solution estimate for that state. In the adjoint method, we instead
-tally in all states and those of lesser importance will not be visited
-as frequently by the random walk. Finally, the expectation value using
-all permutations is:
-\begin{equation}
-  E\{X_j\} = \sum_{\nu} P_{\nu} X_{j}(\nu)\:
-  \label{eq:adjoint_expectation_value}
-\end{equation}
-which, if expanded in the same way as the direct method and utilizing
-Eq~(\ref{eq:adjoint_source_definition}) to insert the source term,
-directly recovers the exact solution:
-\begin{equation}
-  \begin{split}
-    E\{X_j\} &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-    \sum_{i_k}^{N} b_{i_0} h_{i_0,i_1}h_{i_1,i_2}\ldots
-    h_{i_{k-1},i_k} \delta_{i_k,j} \\ &= x_{j}\:,
-  \end{split}
-  \label{eq:adjoint_expectation_expansion}
-\end{equation}
-therefore, also providing an unbiased Monte Carlo estimate of the
-solution. Note that this expansion produces the effective sequence of
-matrix-vector multiplications with the $b_{i_0}$ component of the
-source vector which will be selected at random for each walk
-permutation and the unbiasedness of the estimator relies on an
-unbiased sampling of the source. It should also be noted here that
-Eq~(\ref{eq:adjoint_expectation_expansion}) only computes a single
-component of our desired solution vector when really what we desire is
-the entire solution vector. In an adjoint Monte Carlo simulation using
-this estimator, the $w_{ij}$ elements that are added into the tally
-for each state are only selected if/when the random walk currently
-resides in that state. Much like a mesh tally in a particle transport
-simulation, we have $N$ simultaneous tallies for $\ve{A} \in
-\mathbb{R}^{N \times N}$ that will yield the entire solution
-vector.
-
-Like the direct method, we also desire a criteria for random walk
-termination for problems where only an approximate solution is
-necessary. For the adjoint method, we utilize a \textit{relative
-  weight cutoff}:
-\begin{equation}
-  W_f = W_c b_{i_0}\:,
-  \label{eq:relative_weight_cutoff}
-\end{equation}
-where $W_c$ is defined as in the direct method. The adjoint random
-walk will then be terminated after $m$ steps if $W_m < W_f$ as tally
-contributions become increasingly small.
-
-\subsection{Collision Estimator Variance}
+\section{Collision Estimator Variance\ }
 \label{subsec:forward_variance}
@@
 
-\subsection{Expected Value Estimator}
-\label{subsec:expected_value_estimator}
-In addition to the collision estimator, an additional estimator is
-available due to the work of Okten \cite{okten_solving_2005} that
-uses the method of expected values as a means to improve the Monte
-Carlo estimate. As outlined by Spanier and Gelbard
-\cite{spanier_monte_1969}, the method of expected values is a
-deterministic averaging of events that may potentially occur in the
-Monte Carlo random walk sequence. Okten applied this principle
-directly to discrete Monte Carlo by forming the \textit{expected value
-  estimator} for a random walk of $k$ events:
-\begin{equation}
-  X_{j}(\nu) = b_j + \sum_{m=0}^k W_m h_{j,i_m}\,
-  \label{eq:expected_value_estimator}
-\end{equation}
-where now the contribution of the iteration matrix is
-deterministically averaged at step $m$ over all potential states $j$
-that may be reached from the current state $i_m$. Via Okten, the
-estimator can be shown to be unbiased through a comparison to the
-collision estimator. We can first rewrite the summation in
-Eq~(\ref{eq:expected_value_estimator}):
-\begin{equation}
-  X_{j}(\nu) = b_j + \sum_{m=0}^k \sum_{i=1}^N W_m
-  \delta_{i_m,i} h_{ji}\,
-  \label{eq:unbiased_eval_1}
-\end{equation}
-where $N$ is the number of states in the system. Immediately, we see
-the collision estimator as defined by
-Eq~(\ref{eq:adjoint_permutation_contribution}) and can therefore write
-the expectation value as:
-\begin{equation}
-  E\{X_{j}\} = b_j + \sum_{i=1}^N E\{X_{i}\} h_{ji}\,
-  \label{eq:unbiased_eval_2}
-\end{equation}
-which is equivalently is the $j^{th}$ component of
-Eq~(\ref{eq:richardson_split}):
-\begin{equation}
-  E\{X_{j}\} = b_j + \sum_{i=1}^N x_{i} h_{ji}\,
-  \label{eq:unbiased_eval_2}
-\end{equation}
-and is therefore an unbiased estimate. Compared to the collision
-estimator, the expected value estimator provides additional
-information at every step of the random walk, yielding potentially
-better statistics with the same amount of transport
-work. Conveniently, even if no Monte Carlo histories are computed, the
-expected value estimator still deterministically computes the first
-term of the Neumann Series, $\ve{H}^0\ve{b}$, whereas the collision
-estimator will provide no information.
-
-\subsubsection{Expected Value Estimator Variance}
+\section{Expected Value Estimator Variance\ }
 \label{subsubsec:expected_value_estimator_variance}
@@
 deterministic averaging used to generate the expected value estimator.
-
-\subsection{Adjoint Method: Evolution of a Solution}
-\label{subsec:adjoint_evolution}
-As a means of visually demonstrating the adjoint Monte Carlo method,
-again consider the 2-dimensional thermal diffusion problem with
-sources on the left and right hand sides of the domain and a smaller
-uniform source as shown in Figure~\ref{fig:heat_setup}. Using the
-adjoint method with the collision estimator, the number of histories
-sampled from the source was increased from 10 to 10,000,000 in order
-to show its effects on the solution and the statistical nature of the
-method. Figure~\ref{fig:adjoint_evolution} gives these results.
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/mc_background/adjoint_evolution.png}
-  \end{center}
-  \caption{\textbf{Adjoint Monte Carlo solution to the heat equation
-      with varying numbers of histories.} \textit{Top left: 10
-      histories per state. Top right: 1,000 histories per
-      state. Bottom left: 100,000 histories per state. Bottom right:
-      10,000,000 histories per state.}}
-  \label{fig:adjoint_evolution}
-\end{figure}
-As the number of histories used per state (or DOF) is increased, the
-statistical variance of the solutions is decreased as more tally
-contributions are made. At 10,000,000 histories per state, enough
-information has been tallied to generate a reasonable estimate for the
-structure of the solution. The visual difference between
-Figures~\ref{fig:direct_evolution} and \ref{fig:adjoint_evolution} is
-precisely that determined by their mathematics. As the adjoint
-solution evolves with the addition of histories, more histories
-emanate from the boundary the smaller uniform source in the domain
-with more penetrating from the boundary into the domain and making
-contributions to the tallies in those states as they are transported.
-
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{Sequential Monte Carlo}
-\label{sec:sequential_mc}
-The direct and adjoint Neumann-Ulam methods described are limited by a
-convergence rate of $1/\sqrt{N}$ by the Central Limit Theorem where
-$N$ is the number of random walk permutations. In 1962, Halton
-presented a residual Monte Carlo method that moves towards exponential
-convergence rates \cite{halton_sequential_1962} and further refined
-his work some years later \cite{halton_sequential_1994}. Applications
-of his work by the transport community have confirmed convergence
-rates on the order of of $e^{-N}$ \cite{evans_residual_2003}. In much
-the same way as the projection methods outlined in
-Appendix~\ref{ch:linear_problem}, Halton's method, sequential Monte
-Carlo, utilizes the adjoint Monte Carlo solver as a means of directly
-reducing the elements residual vector. He proposed the following
-iterative scheme as a solution to Eq~(\ref{eq:linear_problem})\:
-\begin{subequations}
-  \begin{gather}
-    \ve{r}^k = \ve{b} - \ve{A}\ve{x}^k\:,\\  
-    \ve{A}\boldsymbol{\delta}^{k} = \ve{r}^{k}\:,\\
-    \ve{x}^{k+1} = \ve{x}^k + \boldsymbol{\delta}^{k}\:,
-  \end{gather}
-  \label{eq:sequential_monte_carlo}
-\end{subequations}
-where the correction $\boldsymbol{\delta}^k$ is computed by the
-adjoint Monte Carlo method at each iteration. The merits of Halton's
-approach are immediately visible in that we have now broken the
-binding of the convergence rate to the Central Limit Theorem. Here,
-the Monte Carlo solver is used to produce a correction from the
-residual, analogous to using the residual to extract a correction from
-the search subspace in a projection method. By doing this, the Monte
-Carlo error is bound in the correction used to update the solution and
-therefore does not explicitly manifest itself in the overall
-convergence of the solution. The downside of such a method is that if
-the solution guess is poor, then many iterations are required in order
-to reach exponential converge as the Monte Carlo error (and therefore
-the Central Limit Theorem) does dominate in this situation.
-
-%%---------------------------------------------------------------------------%%
-\section{Monte Carlo Synthetic Acceleration}
-\label{sec:mcsa}
-Using the ideas of Halton, Evans and Mosher recently developed a Monte
-Carlo solution method that was not prohibited severely by the quality
-of the initial guess for the system \cite{evans_monte_2009} and later
-applied it more rigorously as a solution mechanism for the radiation
-diffusion equation \cite{evans_monte_2012}. With their new methods,
-they achieved numerical results identical to conventional Krylov
-solvers as well as comparable performance in both number of iterations
-and CPU time. Their approach was instead to use residual Monte Carlo
-as a synthetic acceleration for a stationary method. To derive this
-method, we begin by splitting the operator in
-Eq~(\ref{eq:linear_problem})
-\begin{equation}
-  \ve{x} = (\ve{I} - \ve{A})\ve{x} + \ve{b}\:.
-  \label{eq:linear_split}
-\end{equation}
-With this we can then define the stationary method
-\textit{Richardson's iteration} as:
-\begin{equation}
-  \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,
-  \label{eq:richardsons_iteration}
-\end{equation}
-which will converge if $\rho(\ve{I} - \ve{A}) < 1$. We then define the
-solution error at the $k^{th}$ iterate relative to the true solution:
-\begin{equation}
-  \delta \ve{x}^k = \ve{x} - \ve{x}^k\:.
-  \label{eq:mcsa_error}
-\end{equation}
-Subtracting Eq~(\ref{eq:richardsons_iteration}) from
-Eq~(\ref{eq:linear_split}) we get:
-\begin{equation}
-  \delta \ve{x}^{k+1} = (\ve{I} - \ve{A})\delta \ve{x}^k\:.
-  \label{eq:mcsa_setup_1}
-\end{equation}
-Subtracting from this $(\ve{I} - \ve{A})\delta \ve{x}^{k+1}$ yields:
-\begin{equation}
-  \begin{split}
-    \ve{A}\delta \ve{x}^{k+1} &= (\ve{I} -
-    \ve{A})(\ve{x}^{k+1}-\ve{x}^{k}) \\ &= \ve{r}^{k+1}\:.
-    \label{eq:mcsa_setup_2}
-  \end{split}
-\end{equation}
-Using this, we define the following scheme that will converge in one
-iteration if $\ve{A}$ is inverted exactly:
-\begin{subequations}
-  \begin{gather}
-    \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,\\
-    \ve{A} \delta \ve{x}^{k+1} = \ve{r}^{k+1}\:,\\
-    \ve{x} = \ve{x}^{k+1} + \delta \ve{x}^{k+1}\:.
-  \end{gather}
-  \label{eq:mcsa_setup_3}
-\end{subequations}
-However, $\ve{A}$ is only approximately inverted by our numerical
-methods and therefore we instead pose an iterative scheme in which the
-Monte Carlo solvers are used to invert the operator. The
-\textit{Fixed-Point Monte Carlo Synthetic Acceleration} (MCSA) method
-is defined as:
-\begin{subequations}
-  \begin{gather}
-    \ve{x}^{k+1/2} = \ve{x}^k + \ve{r}^k\:,\\
-    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\
-    \ve{A}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:,\\
-    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,\\
-    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:,
-  \end{gather}
-  \label{eq:mcsa}
-\end{subequations}
-where a Neumann-Ulam Monte Carlo method is used to generate the
-solution correction from the residual and Richardson's iteration in
-the first step has been rewritten as a residual correction. Using
-Monte Carlo in this way achieves the same effect as Halton's method,
-decoupling its convergence rate from the overall convergence rate of
-the method. Here, the approximate Monte Carlo solution is not driven
-to a particular convergence as it merely supplies a correction for the
-initial guess generated by Richardson's iteration. Rather, only a set
-number of histories are required using the Neumann-Ulam method to
-generate the correction. In addition, the fact that the scheme in
-Eq~(\ref{eq:mcsa_setup_3}) will converge in one iteration if $\ve{A}$
-is inverted exactly means that as more and more stochastic histories
-are used to compute the correction and the error is reduced towards
-zero, the number of iterations required for MCSA to converge should
-decrease accordingly, thus accelerating the solution.
-
-In addition to the Monte Carlo solver parameters dictating the number
-of histories and weight cutoff, the outer MCSA iterations also have
-the following stopping criteria:
-\begin{equation}
-  ||\ve{r}||_\infty < \epsilon \ ||\ve{b}||_\infty\:,
-  \label{eq:mcsa_stopping_criteria}
-\end{equation}
-where $\epsilon$ is a user-defined parameter. As with any iterative
-method, other stopping criteria using other vector norms could be
-computed, however, for this work we will only use
-Eq~(\ref{eq:mcsa_stopping_criteria}). We therefore have 3 parameters
-to tune in an MCSA implementation: the number of Monte Carlo histories
-computed in the Neumann-Ulam solve during each MCSA iteration, the
-weight cutoff for those histories, and the total MCSA convergence
-tolerance as specified by $\epsilon$.
-
-\subsection{Alternative Fixed Point Iterations}
-\label{subsubsec:alternative_fixed_point}
-In addition to the basic Richardson iteration, the MCSA algorithm
-presented in Eq~(\ref{eq:mcsa}) can be used to accelerate any fixed
-point iteration which only depends on the state (typically the
-residual) of the last iteration. As outlined in
-Appendix~\ref{ch:linear_problem}, subspace methods take on a general
-form using the Petrov-Galerkin conditions as constraints for
-extracting a correction from the search subspace as given by
-Eq~(\ref{eq:linear_projection_iteration}). Those that are
-one-dimensional may be used with fixed-point MCSA as they only depend
-the previous iteration for information. As an example, for
-positive-definite but not necessarily symmetric problems, the minimal
-residual iteration \cite{saad_iterative_2003} can be used in
-conjunction with MCSA where the residual vector, $\mathbf{r}$, defines
-the search subspace and the action of the linear operator on the
-residual vector, $\mathbf{A}\mathbf{r}$, defines the constraint
-subspace. Using this, we can then define an MCSA scheme that
-accelerates the minimal residual iteration:
-\begin{subequations}
-  \begin{gather}
-    \alpha = \frac{\langle \mathbf{A}\mathbf{r}^k, \mathbf{r}^k
-      \rangle}{\langle \mathbf{A}\mathbf{r}^k, \mathbf{A}\mathbf{r}^k
-      \rangle}\:, \\
-    \ve{x}^{k+1/2} = \ve{x}^k + \alpha \ve{r}^k\:,
-    \label{eq:min_res_correction}\\
-    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\ 
-    \ve{A}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:,\\ 
-    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta\ve{x}^{k+1/2}\:,\\ 
-    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:.
-  \end{gather}
-  \label{eq:mcsa_min_res}
-\end{subequations}
-where $\alpha$ is an optimal extrapolation parameter generated from
-the constraints. Interestingly, this scheme is nearly identical to the
-Richardson iteration version in Eq~(\ref{eq:mcsa}) except now the
-additional extrapolation parameter, $\alpha$, is applied to the
-residual correction in Eq~(\ref{eq:min_res_correction}) as a means of
-selecting a more optimal search direction at each iteration,
-potentially further improving convergence.
-
-\subsection{Preconditioning MCSA}
-\label{subsec:stochastic_preconditioning}
-In most cases, at least a minimal amount of \textit{preconditioning}
-of the linear system will be required in order to use the class of
-stochastic methods described. Although these methods have no symmetry
-requirements for convergence, they do require that the spectral radius
-of the iteration matrix be less than one. Preconditioning serves as a
-means of achieving this by altering the eigenvalue spectrum of the
-iteration matrix.
-
-\subsubsection{Basic Preconditioning}
-\label{subsubsec:basic_mcsa_preconditioning}
-As an example of basic preconditioning, to achieve a spectral radius
-of less than one for diagonally dominant matrices point Jacobi
-preconditioning can be used such that the preconditioning matrix
-$\ve{M}$ is:
-\begin{equation}
-  \ve{M} = diag(\ve{A})\:,
-  \label{eq:jacobi_preconditioner}
-\end{equation}
-which may be trivially inverted. With the application of this
-preconditioner we are instead solving the following scaled linear
-system:
-\begin{equation}
-  \ve{M}^{-1}\ve{A}\ve{x} = \ve{M}^{-1}\ve{b}\:.
-  \label{eq:jacobi_precond_linear_problem}
-\end{equation}
-Next, we can apply MCSA to solve
-Eq~(\ref{eq:jacobi_precond_linear_problem}): 
-\begin{subequations}
-  \begin{gather}
-    \ve{x}^{k+1/2} = \ve{x}^k +
-    \ve{M}^{-1}\ve{r}^k\:,\\ \ve{r}^{k+1/2} =
-    \ve{b}-\ve{A}\ve{x}^{k+1/2}\:,\\ \ve{M}^{-1}\ve{A}\delta\ve{x}^{k+1/2}
-    = \ve{M}^{-1}\ve{r}^{k+1/2}\:,\\ \ve{x}^{k+1} = \ve{x}^{k+1/2} +
-    \delta \ve{x}^{k+1/2}\:,\\
-    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:,
-    \label{eq:jacobi_preconditioned_mcsa}
-  \end{gather}
-\end{subequations}
-where the Neumann-Ulam Monte Carlo solve now has a preconditioned
-operator from which to build weights and probabilities for transport
-and a preconditioned source vector to sample.
-
-Choosing point Jacobi preconditioning with MCSA is advantageous for
-several reasons. First, $\rho(\ve{I} - \ve{M}^{-1}\ve{A}) < 1$ is true
-for all $\ve{A}$ that is diagonally dominant and is easy to formulate
-because the inversion of $\ve{M}$ is trivial. Second, because the
-Monte Carlo method used within MCSA to compute the correction operates
-on a linear problem with the preconditioned operator, then $\ve{H}$ in
-the Neumann-Ulam solver will have a zero term in each of its diagonal
-elements, thereby eliminating all in-state transitions during the
-random walk sequence. Because of this, point Jacobi preconditioning
-should be considered for many classes of problems, regardless of any
-other preconditioning that is applied to the system. In addition,
-Jacobi preconditioning has been shown to be an effective
-preconditioning mechanism for MCSA when used with the thermal
-radiation diffusion equation \cite{evans_monte_2012}.
-
-\clearpage
-
-\subsubsection{General Preconditioning Strategies}
-\label{subsubsec:general_mcsa_preconditioning}
-It is possible to use general left, right, and left/right
-preconditioning with MCSA by carefully considering the underlying
-Monte Carlo problem that will be solved with the Neumann-Ulam
-method. We consider here the general left/right preconditioned method
-as the left or right preconditioned methods can be inferred from its
-formulation. We consider a left preconditioner $\ve{M_L}$ and a right
-preconditioner $\ve{M_R}$. The left/right preconditioned linear
-problem is then:
-\begin{equation}
-  \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\ve{M}_R\ve{x} = \ve{M}_L^{-1}\ve{b}\:.
-  \label{eq:left_right_linear_problem}
-\end{equation}
-To handle the right preconditioning, the system is written with a
-substitution of variables:
-\begin{equation}
-  \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\ve{u} = \ve{M}_L^{-1}\ve{b}\:,
-  \label{eq:left_right_subs_problem}
-\end{equation}
-with
-\begin{equation}
-  \ve{x} = \ve{M}_R^{-1}\ve{u}\:.
-  \label{eq:left_right_recover}
-\end{equation}
-To apply such a method to MCSA, we solve for the substituted variable
-$\ve{u}$ during the iteration sequence:
-\begin{subequations}
-  \begin{gather}
-    \ve{u}^{k+1/2} = \ve{u}^k + \ve{r}^k\:,\\
-    \ve{r}^{k+1/2} = \ve{M}_L^{-1}(\ve{b}-\ve{A}\ve{M}_R^{-1}\ve{u}^{k+1/2})\:,\\ 
-    \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\delta\ve{u}^{k+1/2} = \ve{r}^{k+1/2}\:,\\ 
-    \ve{u}^{k+1} = \ve{u}^{k+1/2} + \delta \ve{u}^{k+1/2}\:,\\
-    \ve{r}^{k+1} = \ve{M}_L^{-1}(\ve{b}-\ve{A}\ve{M}_R^{-1}\ve{u}^{k+1})\:,
-  \end{gather}
-  \label{eq:left_right_mcsa}
-\end{subequations}
-and then recover the original solution vector with
-Eq~(\ref{eq:left_right_recover}) after convergence. For the Monte
-Carlo problem, we isolate the generation of the correction:
-\begin{equation}
-  \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\delta\ve{u}^{k+1/2} = \ve{r}^{k+1/2}\:,
-  \label{eq:left_right_correction}
-\end{equation}
-and note that the preconditioned residual of the substituted variable
-is now serving as the source and the new iteration matrix is:
-\begin{equation}
-  \ve{H} = \ve{I} - \ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}\:.
-  \label{eq:left_right_iteration_matrix}
-\end{equation}
-As we require $(i,j)$ element-wise access to the iteration matrix in
-order to construct probabilities and weights for the Monte Carlo
-procedure from the Neumann-Ulam decomposition, the \textit{composite
-  operator}, $\ve{M}_L^{-1}\ve{A}\ve{M}_R^{-1}$, must be formed via
-matrix-matrix multiplication.
-
-Several possible shortcomings of this preconditioning approach are
-readily observed. First, the matrix-matrix multiplication operation
-for sparse, parallel distributed matrices is significantly more
-expensive than a matrix-vector multiplication operation. Second, each
-preconditioner must be explicitly inverted, an operation in itself
-that may be expensive and which prohibits the use of any
-preconditioners which provide no mechanism to extract their
-inverse. Third, for many modern preconditioning methods, this
-inversion may yield dense matrices, destroying sparsity and further
-impeding the performance of a matrix-matrix multiplication
-operation. It is also interesting to note that the Monte Carlo problem
-in the general left/right preconditioned scheme given by
-Eq~(\ref{eq:left_right_correction}) is not fully left/right
-preconditioned (meaning that we do not recover $\ve{x}$), but instead
-part of a sequence for finding the substituted variable $\ve{u}$. We
-do, however, gain the benefits of this general preconditioning by
-building the iteration matrix in
-Eq~(\ref{eq:left_right_iteration_matrix}) from the fully
-preconditioned linear operator. In addition, for MCSA to apply to
-increasingly difficult problems, more advanced preconditioning
-techniques that require the generation of the composite operator may
-be necessary for convergence.
-
-%%---------------------------------------------------------------------------%%
-\section{Monte Carlo Method Selection}
-\label{sec:mc_method_selection}
-The MCSA method defined in Eq.~(\ref{eq:mcsa}) uses the adjoint method
-to estimate the error in a residual Monte Carlo solve instead of the
-direct method outlined in \S\ref{sec:direct_mc}. To demonstrate the
-effectiveness of the adjoint method over the direct method within the
-context of MCSA, we choose the two-dimensional time-dependent Poisson
-equation as a simple model transport problem\footnote{The Poisson
-  equation is a simple form of the diffusion equation and has a
-  similar elliptic character in its time-dependent form to the neutron
-  transport equations that will be solved in the next chapter.}:
-\begin{equation}
-  \frac{\partial \ve{u}}{\partial t} = \nabla^2 \ve{u}\:.
-  \label{eq:poisson_equation}
-\end{equation}
-For all comparisons, a single time step is computed with backwards Euler time
-integration. The Laplacian is differenced on a square Cartesian grid with a
-second-order five-point stencil,
-\begin{equation}
-  \nabla^2_5 = \frac{1}{\Delta^2}[u_{i-1,j} + u_{i+1,j} + u_{i,j-1} +
-    u_{i,j+1} - 4 u_{i,j}]\:,
-  \label{eq:five_point_stencil}
-\end{equation}
-and a fourth-order nine-point stencil,
-\begin{multline}
-  \nabla^2_9 = \frac{1}{6\Delta^2}[4 u_{i-1,j} + 4 u_{i+1,j} + 4
-    u_{i,j-1} + 4 u_{i,j+1} + u_{i-1,j-1}\\ + u_{i-1,j+1} +
-    u_{i+1,j-1} + u_{i+1,j+1} - 20 u_{i,j}]\:,
-  \label{eq:nine_point_stencil}
-\end{multline}
-both assuming a grid size of $\Delta$ in both the $i$ and $j$ directions. For
-a single time step solution, we then have the following sparse linear system
-to be solved with the MCSA method:
-\begin{equation}
-  \ve{A} \ve{u}^{n+1} = \ve{u}^n\:.
-  \label{eq:poisson_eq_lin_sys}
-\end{equation}
-Both the stencils will be used to vary the size and density of the sparse
-linear system in Eq.~(\ref{eq:poisson_eq_lin_sys}).
-
-A timing and convergence study is used to demonstrate the
-effectiveness of the adjoint method with the collision estimator as
-compared to the direct method. To assess both the CPU time and number
-of iterations required to converge to a solution, a problem of
-constant $\Delta$ was used with varying values of the number of mesh
-elements, fixing the spectral radius of the system at a constant value
-for each variation. Both the five-point and nine-point stencils were
-used with both the direct and adjoint solvers. For each case, $N
-\times N$ total random walk permutations were computed per MCSA
-iteration where $N \times N$ is the number of discrete grid points (or
-degrees of freedom) in the system. Solver parameters were set to a
-weight cutoff of \sn{1}{-4} for the stochastic linear solver and a
-convergence tolerance of \sn{1}{-8} for the MCSA iterative solver.
-Figure~\ref{fig:poisson_cpu_time} gives the CPU time needed for each
-case to converge in seconds, and Figure~\ref{fig:poisson_iterations}
-gives the number of iterations needed for each case to converge to the
-specified tolerance, both as a function of the problem size. All
-computations presented in this section and the remaining sections of
-this chapter were completed on a 3.0 GHz Intel Core 2 Quad Q9650 CPU
-machine with 16 GB 1067 MHz DDR3 memory.
-\begin{figure}[t!]
-  \centering
-  \includegraphics[width=5in,clip]{chapters/mc_background/dir_adj_cpu.pdf}
-  \caption{\textbf{CPU Time (s) to converge vs. Problem Size ($N$ for
-      an $N \times N$ square mesh).} \textit{Both the adjoint and
-      direct solvers are used with the five point and nine point
-      stencils. A CPU time speedup is noted with the adjoint method
-      due to the higher density of random walk events in regions with
-      a large residual.}}
-  \label{fig:poisson_cpu_time}
-\end{figure}
-
-\begin{figure}[t!]
-  \centering
-  \includegraphics[width=5in,clip]{chapters/mc_background/dir_adj_iterations.pdf}
-  \caption{\textbf{Iterations to converge vs. Problem Size ($N$ for an
-      $N \times N$ square mesh).} \textit{Both the adjoint and direct
-      solvers are used with the five-point and nine-point stencils.}}
-  \label{fig:poisson_iterations}
-\end{figure}
-
-We see clearly in Figure~\ref{fig:poisson_cpu_time} that the using the
-adjoint solver with MCSA results in a speedup over the direct solver
-while the number of iterations required to converge is also reduced as
-shown in Figure~\ref{fig:poisson_iterations}. We expect this for
-several reasons. First, with an equivalent number of histories
-specified for both solvers per MCSA iteration and a system of size $N
-\times N$, the direct solver will compute a single random walk for
-each state in the system per iteration to acquire a solution in that
-state. This is necessary in the direct method to ensure a contribution
-from each state as the random walk sequence will only contribute to
-the starting state. For the adjoint method, a total of $N \times N$
-random walk events will have their starting state determined by
-sampling the residual vector. Because the random walk sequence
-contributes to the state in which it currently resides, sampling the
-residual vector as the Monte Carlo source gives a higher density of
-random walk events in regions with a high residual, thus giving a more
-accurate correction in that region due to reduced statistical
-error. From an iteration perspective,
-Figure~\ref{fig:poisson_iterations} shows that using the direct method
-yields a roughly unchanging number of iterations required to converge
-as the problem size increases. Again, if we desire a correction value
-for all states in the problem, then we must start a random walk in
-each state in the system without taking into account the structure of
-the residual vector. Because of this, as the problem size grows adding
-histories is ineffective as many are added in states where the error
-is smaller than in other parts of the problem, ultimately not reducing
-the number of iterations needed to converge. Conversely, as the
-problem size grows in the adjoint method, the additional stochastic
-histories that will be computed are concentrated in regions with a
-large residual, further reducing the stochastic error in the
-correction in those regions and subsequently reducing the required
-number of iterations to converge.
-
-As an additional comparison, the convergence behavior of MCSA can be
-analyzed using both the adjoint and direct solvers to detect any
-performance benefits. To assess the convergence properties of MCSA
-using each solver and stencil, the infinity norm of the residual
-computed in Eq.~(\ref{eq:mcsa}) was collected at each iteration for a
-fixed problem size of $N=500$. Figure~\ref{fig:poisson_convergence}
-gives the results of these computations. First, it is worthy to note
-on the semi-log plot that we are indeed achieving the expected
-exponential convergence from MCSA with both Monte Carlo
-solvers. Second, we note that using the adjoint method with the same
-number of stochastic histories per MCSA iteration gives a faster rate
-of converge for the same reasons as above. We also note here that
-fewer iterations are required for convergence when the 9-point stencil
-is used to discretize the Laplacian operator (although at no gain in
-speed as given by the results in
-Figure~\ref{fig:poisson_cpu_time}). This is due to the fact that the
-smaller discretization error directly corresponds to a more well
-defined residual source generated by the Richardson extrapolation for
-the Monte Carlo calculation. In addition, the better defined source is
-transported through a domain described more accurately by the 9-point
-stencil, thus yielding a more accurate correction vector from the
-Monte Carlo calculation.
-\begin{figure}[t!]
-  \centering
-  \includegraphics[width=5in,clip]{chapters/mc_background/dir_adj_conv.pdf}
-  \caption{\textbf{Infinity norm of the solution residual
-      vs. iteration number for a problem of size $N=500$.}
-    \textit{Both the adjoint and direct solvers are used with the five
-      point and nine point stencils. A higher rate of convergence is
-      observed for MCSA using the adjoint Monte Carlo solver as
-      compared to the direct method when both solvers compute the same
-      number of random walks per iteration.}}
-  \label{fig:poisson_convergence}
-\end{figure}
-
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{MCSA Comparison to Sequential Monte Carlo}
-\label{subsec:sequential_comparison}
-To further motivate using Monte Carlo Synthetic Acceleration, we
-compare its performance to Halton's Sequential Monte Carlo method on
-which previous work in this area was based. For this comparison, we
-use the same transient Poisson problem as described in the previous
-section and choose only the 5-point stencil to discretize the
-Laplacian operator as the previous results yielded little qualitative
-difference between the discretizations. Both MCSA and Halton's method
-are used with the adjoint Monte Carlo solver and the collision
-estimator. In order to complete the same study as in the previous
-section, the number of histories computed by the Monte Carlo solver at
-each iteration had to be doubled to $2 \times N \times N$ in order to
-ensure convergence in Sequential Monte Carlo Method. For the majority
-of the problems in the previous section, the Sequential method used
-with $N \times N$ histories would not
-converge. Figure~\ref{fig:seq_cpu_time} gives the CPU time results for
-this comparison as a function of problem size while
-Figure~\ref{fig:seq_iterations} gives the number of iterations to
-converge as a function of problem size with a convergence tolerance of
-\sn{1}{-8}. In both cases, using the Monte Carlo solver as a synthetic
-acceleration rather than in a pure residual Monte Carlo scheme
-resulted in a reduction in both CPU time and iterations required to
-converge. The additional Richardson extrapolation between each Monte
-Carlo solve in the MCSA method gives a better converged residual
-source to use with the Monte Carlo calculation while the Sequential
-method requires more iterations to achieve the same level of
-convergence in the residual.
-
-\begin{figure}[t!]
-  \centering
-  \includegraphics[width=4.5in,clip]{chapters/mc_background/seq_cpu.pdf}
-  \caption{\textbf{CPU Time (s) to converge vs. Problem Size ($N$ for
-      an $N \times N$ square mesh).} \textit{Both the Sequential Monte
-      Carlo and MCSA solvers are used with the five point stencils and
-      the adjoint Monte Carlo solver. The number of random walks was
-      twice the number of discrete states in the system in order to
-      ensure convergence in the Sequential Monte Carlo method.}}
-  \label{fig:seq_cpu_time}
-\end{figure}
-
-\begin{figure}[t!]
-  \centering
-  \includegraphics[width=4.5in,clip]{chapters/mc_background/seq_iterations.pdf}
-  \caption{\textbf{Iterations to converge vs. Problem Size ($N$ for an
-      $N \times N$ square mesh).} \textit{Both the Sequential Monte
-      Carlo and MCSA solvers are used with the five point stencils and
-      the adjoint Monte Carlo solver.}}
-  \label{fig:seq_iterations}
-\end{figure}
-
-The benefits of using a synthetic acceleration scheme are also noted
-when the infinity norm of the residual computed at each iteration for
-both methods was collected at each iteration for a fixed problem sizes
-of $N=100$ and $N=500$ as shown in figures Figure~\ref{fig:seq_100}
-and \ref{fig:seq_500} respectively. In both cases, the Sequential
-method is subject to two regimes of exponential convergence with high
-frequency error modes removed in the first regime leaving lower
-frequency and slower converging error modes in the second. Using MCSA
-we a see a single rate of exponential convergence observed to be much
-higher than that computed by Halton's method due to the fact that the
-extra Richardson iteration is providing a smoothing effect to
-alleviate the error mode variations. Even with the doubling of the
-number of stochastic histories computed per time step in order to
-ensure convergence for the Sequential method, we still see robustness
-issues with a non-monotonically decreasing residual observed for the
-$N=100$ case. In both cases the MCSA solver is observed to be robust
-with a monotonically decreasing residual.
-
-\begin{figure}[t!]
-  \centering
-  \includegraphics[width=4.5in,clip]{chapters/mc_background/seq_conv_100.pdf}
-  \caption{\textbf{Infinity norm of the solution residual
-      vs. iteration number for a problem of size $N=100$.}
-    \textit{Both the Sequential Monte Carlo and MCSA solvers are used
-      with the five point stencils and the adjoint Monte Carlo
-      solver.}}
-  \label{fig:seq_100}
-\end{figure}
-
-\begin{figure}[t!]
-  \centering
-  \includegraphics[width=4.5in,clip]{chapters/mc_background/seq_conv_500.pdf}
-  \caption{\textbf{Infinity norm of the solution residual
-      vs. iteration number for a problem of size $N=500$.}
-    \textit{Both the Sequential Monte Carlo and MCSA solvers are used
-      with the five point stencils and the adjoint Monte Carlo
-      solver.}}
-  \label{fig:seq_500}
-\end{figure}
-
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{Monte Carlo Parameter and Estimator Analysis}
-\label{sec:parameter_estimator_analysis}
-With the adjoint method shown to be more effective than the direct
-method and MCSA to have better iterative and timing performance than
-sequential Monte Carlo, we now aim to study the effects of the adjoint
-Monte Carlo parameters, weight cutoff and number of histories, on MCSA
-performance with both the collision and expected value
-estimators. With the same Poisson problem, we will use a $200 \times
-200$ grid and a convergence tolerance of \sn{1}{-8} for all
-calculations. To study the effects of the number of Monte Carlo
-histories per MCSA iteration, the weight cutoff was fixed at
-\sn{1}{-4} while the number of histories per iteration was varied from
-6,000 to 100,000. It was observed that MCSA would not converge for
-this problem using the collision estimator with less than 6,000
-histories while the expected value estimator permitted convergence
-with only 2,000 histories.
-
-Figure~\ref{fig:estimator_nh_iters} gives the number of iterations
-required to converge for both estimators as a function of the number
-of histories per iteration. For smaller numbers of histories, the
-performance of the expected value estimator is significantly better
-than the collision estimator. This result is valuable in that less
-transport is required to achieve the same MCSA iterative performance
-with the expected value estimator, important for situations where
-transport is expensive (i.e. in domain decomposed
-calculations). Interestingly, as the number of histories per iteration
-are increased, the iterative performance with the collision estimator
-approaches that of the expected value estimator. However, given the
-performance of the expected value estimator at a fractional number of
-histories, one should strongly consider its use over using the
-collision estimator with more histories. The CPU time required to
-converge is presented in Figure~\ref{fig:estimator_nh_time} and
-reflects the results of the iterative performance. In general, using
-the collision estimator is slightly slower overall, but the time per
-iteration is faster due to the fact that the estimator does not have
-to cycle through multiple states during the tally procedure.
-
-\begin{figure}[t!]
-  \centering
-  \includegraphics[width=4.75in,clip]{chapters/mc_background/estimator_nh_iters.pdf}
-  \caption{\textbf{Iterations (s) to converge vs. Monte Carlo
-      histories per MCSA iteration for a $200 \times 200$ square mesh
-      and a weight cutoff of \sn{1}{-4}.} \textit{For low numbers of
-      histories, the expected value estimator performance is
-      significantly better than the collision estimator. At higher
-      numbers of histories, the estimators become roughly
-      equivalent.}}
-  \label{fig:estimator_nh_iters}
-\end{figure}
-
-\begin{figure}[t!]
-  \centering
-  \includegraphics[width=4.75in,clip]{chapters/mc_background/estimator_nh_time.pdf}
-  \caption{\textbf{CPU Time (s) to converge vs. Monte Carlo histories
-      per MCSA iteration for a $200 \times 200$ square mesh and a
-      weight cutoff of \sn{1}{-4}.} \textit{For low numbers of
-      histories, the expected value estimator performance is better
-      than the collision estimator due to a lower iteration count
-      while the actual compute time per iteration is higher.}}
-  \label{fig:estimator_nh_time}
-\end{figure}
-
-For the weight cutoff study, the number of histories per iteration was
-fixed at 40,000 and the weight cutoff varied from \sn{5}{-1} down to
-\sn{1}{-10}. Surprisingly, the number of iterations to converge given
-by Figure~\ref{fig:estimator_wc_iters} is effectively invariant to the
-weight cutoff, only seeing detrimental effects on the number of
-iterations at a very large weight cutoff. This suggests that a fairly
-large weight cutoff can be used in practice with both adjoint
-estimators and that the preliminary components of the random walk are
-more important to MCSA convergence than those that occur later and
-with lower weight contributions. To further motivate using a larger
-weight cutoff, Figure~\ref{fig:estimator_wc_time} gives the CPU time
-need to converge as a function of weight cutoff. As expected, lowering
-the weight cutoff lengthens the random walk lengths and increases CPU
-time at no gain of iterative performance. In general, these results
-suggest using the expected value estimator over the collision
-estimator with MCSA as well as using a larger weight cutoff.
-
-\begin{figure}[t!]
-  \centering
-  \includegraphics[width=4.75in,clip]{chapters/mc_background/estimator_wc_iters.pdf}
-  \caption{\textbf{Iterations (s) to converge vs. history weight
-      cutoff for a $200 \times 200$ square mesh and 40,000 histories.}
-    \textit{For low numbers of histories, the expected value estimator
-      performance is significantly better than the collision
-      estimator. At higher numbers of histories, the estimators become
-      roughly equivalent.}}
-  \label{fig:estimator_wc_iters}
-\end{figure}
-
-\begin{figure}[t!]
-  \centering
-  \includegraphics[width=4.75in,clip]{chapters/mc_background/estimator_wc_time.pdf}
-  \caption{\textbf{CPU Time (s) to converge vs. history weight cutoff
-      for a $200 \times 200$ square mesh and 40,000 histories.}
-    \textit{For low numbers of histories, the expected value estimator
-      performance is better than the collision estimator due to a
-      lower iteration count while the actual compute time per
-      iteration is higher.}}
-  \label{fig:estimator_wc_time}
-\end{figure}
-
-\clearpage
-
-%%---------------------------------------------------------------------------%%
-\section{Summary}
-\label{sec:mc_summary}
-
-In this chapter, the fundamentals of Monte Carlo Synthetic
-Acceleration methods were presented and explored in the context of a
-simple model transport system. The following are the significant
-observations and findings.
-
-\begin{itemize}
-\item Monte Carlo Synthetic Acceleration (MCSA) has been generalized
-  for all discrete linear transport problems
-\item MCSA can be preconditioned using a general left/right scheme but
-  requires the explicit inversion of the preconditioners and
-  construction of the composite linear operator
-\item The adjoint Neumann-Ulam method performs better when used with
-  MCSA compared to the forward method
-\item MCSA has superior performance when compared to Halton's residual
-  Monte Carlo method
-\item The Richardson iteration in MCSA behaves as a smoother for the
-  stochastic Monte Carlo error
-\item MCSA iterative performance when used with the expected value
-  estimator is insensitive to histories per iteration
-\item MCSA iterative performance is insensitive to the stochastic
-  history weight cutoff
-\end{itemize}

### 2013-08-26 doc/dissertation/document/chapters/conclusion/conclusion.tex
moving estimator variance into an appendix

@@
-\chapter{Conclusions and Analysis\ }
+\chapter{Conclusion\ }
 \label{ch:conclusion}
@@
 %%---------------------------------------------------------------------------%%
-\section{Monte Carlo Synthetic Acceleration Solutions for the $SP_N$ Equations\ }
+\section{Monte Carlo Synthetic Acceleration Methods for the $SP_N$ Equations\ }
 \label{sec:spn_conclusion}
@@
 %%---------------------------------------------------------------------------%%
-\section{Monte Carlo Synthetic Acceleration Solutions for Navier-Stokes Equations\ }
+\section{Monte Carlo Synthetic Acceleration Methods for the Navier-Stokes Equations\ }
 \label{sec:nonlinear_conclusions}
@@
 equations. This research was enabled by implementing the FANM method
-within the nonlinear solver sequence leveraged by the Drekar production
-multiphysics code base being developed at Sandia National
+within the nonlinear solver sequence leveraged by the Drekar
+production multiphysics code base being developed at Sandia National
 Laboratories. In both convection and driven flow regimes we solved
@@
 %%---------------------------------------------------------------------------%%
-\section{Parallel Monte Carlo Synthetic Acceleration\ }
+\section{Parallel Monte Carlo Synthetic Acceleration Methods\ }
 \label{sec:parallel_mc_conclusions}

### 2013-08-26 doc/dissertation/document/chapters/introduction/introduction.tex
moving estimator variance into an appendix

@@
 cores already coming on-line, new algorithms to solve these complex
-problems must be developed to leverage this new hardware
+problems must be developed to utilize this new hardware
 \cite{kogge_using_2011}. Issues such as resiliency to node failure,
 limited growth of memory available per node, and scaling to large
-numbers of cores will be pertinent to robust algorithms aimed at this
-new hardware.
+numbers of cores will be pertinent to robust algorithms aimed at the
+next generation of machines.
 
@@
 been developed, we look forward to parallelizing them with the goal
-that they may be leveraged competitively on leadership class computing
+that they may be leveraged competitively on leadership-class computing
 platforms.
@@
 domain decomposed Monte Carlo methods for particle transport is
-developed to fully parallelize the solution techniques developed to
-solve the neutron transport and fluid flow problems on leadership
-class hardware with scaling studies performed in this work using a
-neutron diffusion problem.
+developed to fully parallelize the new solution techniques on
+leadership class hardware with scaling studies performed in this work
+using a neutron diffusion problem.
 
@@
 solution. Modern high fidelity deterministic methods for large scale
-problems are commonly variants on the discrete ordinates ($S_N$)
-method \cite{evans_denovo:_2010}. For fission reactor neutronics
-simulations, the $S_N$ method requires potentially trillions of
-unknown angular flux moments to be computed to achieve good accuracy
-for the responses of interest
-\cite{slaybaugh_acceleration_2011}. Other forms of the transport
-problem, including the $P_N$ method, take on a simpler form than the
-more common $S_N$ methods but lack in accuracy when compared while
-still requiring considerable computational resources for solutions in
-multiple dimensions.
+problems are often variants on the discrete ordinates ($S_N$) method
+\cite{evans_denovo:_2010}. For fission reactor neutronics simulations,
+the $S_N$ method requires potentially trillions of unknown angular
+flux moments to be computed to achieve good accuracy for the responses
+of interest \cite{slaybaugh_acceleration_2011}. Other forms of the
+transport problem, including the $P_N$ method, take on a simpler form
+than the more common $S_N$ methods, but lack in accuracy when compared
+while still requiring considerable computational resources for
+solutions in multiple dimensions.
 
-In the 1960's, Gelbard developed an ad-hoc multidimensional extension
+In the 1960's, Gelbard developed an ad-hoc, multidimensional extension
 of the simple single dimension planar $P_N$ equations that created a
@@
 primarily due to Larsen and his colleagues, the $SP_N$ equations have
-been given a more rigorous treatment in many forms with both
-variational and asymptotic derivations performed as a means of
-verification \cite{olbrant_asymptotic_2013}. In addition, these
-equations have been more rigorously studied as solution methods to MOX
-fuel problems and have been shown to provide accurate solutions
+been given a more rigorous treatment with both variational and
+asymptotic derivations performed as a means of verification
+\cite{olbrant_asymptotic_2013}. In addition, these equations have been
+more rigorously studied as solution methods to MOX fuel problems and
+have been shown to provide accurate solutions
 \cite{brantley_simplified_2000}. With this mathematical literature to
@@
 Carlo algorithm, which may be significantly more scalable than current
-$S_N$ transport practices. In this work we therefore apply Monte Carlo
-Synthetic Acceleration to the $SP_N$ equations and research its
-applicability, performance, and convergence properties in the context
-of a light water reactor system.
+$S_N$ transport practices. In this work we apply Monte Carlo Synthetic
+Acceleration to the $SP_N$ equations and research its applicability,
+performance, and convergence properties in the context of a light
+water reactor system.
 
@@
 popular and many advances have been made in the nuclear engineering
-field to employ these methods.
+field to employ these methods \cite{gaston_parallel_2009}.
 
-In the context of solving standalone linear systems, Monte Carlo
+In the context of solving standalone linear problems, Monte Carlo
 methods do not provide significant merit over subspace methods due to
@@
 of the components necessary for a Monte Carlo solver to provide
-value. We therefore devise a new nonlinear method in this work based
-on the Monte Carlo Synthetic Acceleration algorithm and Newton's
+value. In this work we devise a new nonlinear method in this work
+based on the Monte Carlo Synthetic Acceleration algorithm and Newton's
 method and research its properties, applicability, and performance in
@@
 Department of Energy's Consortium for Advanced Simulation of Light
-Water Reactors (CASL) modeling and simulation hub. CASL used the
-Drekar multiphysics code developed at Sandia National Laboratories
+Water Reactors (CASL) modeling and simulation hub
+\cite{u.s._department_of_energy_casl_2011}. CASL used the Drekar
+multiphysics code developed at Sandia National Laboratories
 \cite{pawlowski_drekar_2012} for modeling goals in
@@
 literature, Monte Carlo Synthetic Acceleration has yet to be
-parallelized and the Neumann-Ulam Monte Carlo method on which it is
-built has only been parallelized through history-level parallelism
-with full domain replication \cite{alexandrov_efficient_1998}. In
-order to solve large linear physics systems with Monte Carlo Synthetic
-Acceleration, a domain decomposed parallel strategy is required. In
-the formulation of a parallel Monte Carlo Synthetic Acceleration
-algorithm, we recognize that the algorithm occurs in two stages, an
-initial step performing a fixed point iteration and a secondary step
-leveraging a Monte Carlo solver that is providing the
-acceleration. The parallel aspects of both these components must be
-considered. Therefore, in this work we will develop and implement
-parallel algorithms for the fundamental random walk sequence and Monte
-Carlo Synthetic Acceleration methods leveraging both the knowledge
-gained from the general parallel implementations of subspace methods
-reviewed in Appendix~\ref{ch:linear_problem} and modern parallel
-strategies for domain decomposed Monte Carlo as developed by the
-reactor physics community.
+parallelized and the underlying Neumann-Ulam Monte Carlo method has
+only been parallelized through history-level parallelism with full
+domain replication \cite{alexandrov_efficient_1998}. In order to solve
+large linear physics systems with Monte Carlo Synthetic Acceleration,
+a domain decomposed parallel strategy is required. In the formulation
+of a parallel Monte Carlo Synthetic Acceleration algorithm, we
+recognize that the algorithm occurs in two stages, an initial step
+performing a fixed point iteration and a secondary step leveraging a
+Monte Carlo solver that is providing the acceleration. The parallel
+aspects of both these components must be considered. Therefore, in
+this work we will develop and implement parallel algorithms for the
+fundamental random walk sequence and Monte Carlo Synthetic
+Acceleration methods leveraging both the knowledge gained from the
+general parallel implementations of subspace methods reviewed in
+Appendix~\ref{ch:linear_problem} and modern parallel strategies for
+domain decomposed Monte Carlo as developed by the reactor physics
+community.
 
@@
 
-The end result of these hardware changes is that the larger numbers of
+The end result of these hardware changes is that the larger number of
 low-powered processors will be prone to both soft failures such as bit
@@
 issues are potentially alleviated by statistical arguments. In the
-case of soft failures, isolated floating point errors in Monte Carlo
-simulation are absorbed within tally statistics while completely
-losing hardware during a hard failure is manifested as a high variance
-event where some portion of the Monte Carlo histories are lost. These
-stochastic methods are a paradigm shift from current deterministic
-solver techniques that will in their current form potentially suffer
+case of soft failures, isolated floating point errors in the Monte
+Carlo simulation are absorbed within tally statistics. Completely
+losing memory during a hard failure is treated as a high variance
+event where some portion of the Monte Carlo histories and subsequently
+the solution are lost with other histories maintained by replicating
+the problem and combining the results through superposition. These
+stochastic methods are a paradigm shift from modern deterministic
+solver techniques that will, in their current form, potentially suffer
 greatly from the non-deterministic behavior expected from the next
 generation of machines.
-
-In addition to resiliency concerns, the memory restrictions on future
-hardware will hinder modern solvers that derive their robustness from
-using large amounts of memory. Stochastic methods that are formulated
-to use less memory than conventional methods will serve to alleviate
-some of this pressure. New parallel strategies that may
-be implemented with stochastic methods could offer a new avenue for
-leveraging the expected levels of high concurrency in machines
-operating at hundreds of petaflops and beyond.
 
@@
 \label{sec:research_outline}
-For some time, the particle transport community has been utilizing
-Monte Carlo methods for the solution of transport problems
+For some time, the particle transport community has utilized Monte
+Carlo methods for the solution of transport problems
 \cite{lewis_computational_1993}. The partial differential equation
@@
 differencing method on the order of machine precision such that this
-error can grow beyond that of those in a fine-grained system
+error can grow beyond that of those in a poorly-scaled system
 \cite{kelley_iterative_1995} and (2) for systems that are not
@@
 are: full Jacobian generation using modern Forward Automated
-Differentiation (FAD) methods \cite{bartlett_automatic_2006}, and MCSA
-as the inner linear solver. This method has several attractive
-properties. First, the first-order approximation to the Jacobian used
-in JFNK type methods is eliminated by generating the Jacobian
-explicitly with the model equations through FAD. Second, the Jacobian
-need not be explicitly formed by the user but is instead automated
-through FAD; this eliminates the complexity of hand-coding derivatives
-and has also been demonstrated to be more efficient computationally
-than evaluating difference derivatives. Third, unlike GMRES, MCSA does
-not build a subspace during iterations. Although the Jacobian must be
+Differentiation (FAD) methods, and MCSA as the inner linear
+solver. This method has several attractive properties. First, the
+first-order approximation to the Jacobian used in JFNK type methods is
+eliminated with the Jacobian fully formed. Second, the Jacobian need
+not be explicitly formed by the user but is instead automated through
+FAD; this eliminates the complexity of hand-coding derivatives and has
+also been demonstrated to be more efficient computationally than
+evaluating difference derivatives
+\cite{bartlett_automatic_2006}. Third, unlike GMRES, MCSA does not
+build a subspace during iterations. Although the Jacobian must be
 explicitly formed to use MCSA, for problems that take more than a few
@@
 a wide variety of numerical experiments using both simple model
-neutron transport problems and neutron transport and fluid flow
-problems generated by production physics codes in both simple and
-complex geometries.
+transport problems and neutron transport and fluid flow problems
+generated by production physics codes in both simple and complex
+geometries.
 

### 2013-08-26 doc/dissertation/document/chapters/mc_background/mc_background.tex
moving estimator variance into an appendix

@@
 method for discrete linear systems. Both the forward and adjoint
-methods will be presented and analyzed including a brief variance
-analysis of several Monte Carlo estimators. The Sequential Monte Carlo
-method of Halton and Monte Carlo Synthetic Acceleration methods will
-then be presented as a means of leveraging the basic Monte Carlo
-methods devised by Neumann and Ulam in an iterative refinement scheme
-that accelerates their convergence. Using these iterative schemes,
-the effects of the algorithm parameters on solutions to a simple model
+methods will be presented and analyzed. The Sequential Monte Carlo
+method of Halton and Monte Carlo Synthetic Acceleration will then be
+presented as a means of leveraging the basic Monte Carlo methods
+devised by Neumann and Ulam in an iterative refinement scheme that
+accelerates their convergence. Using these iterative schemes, the
+effects of the algorithm parameters on solutions to a simple model
 transport problem are explored.
@@
 \end{equation}
-such that an exact solution $\ve{x}$ has been found when
+such that an exact solution, $\ve{x}$, has been found when
 $\ve{r}=\ve{0}$.  From the statement in
@@
 restriction on $\ve{A}$ by requiring that it be \textit{nonsingular},
-meaning that we can in fact compute $\ve{A}^{-1}$. In this work we
-will focus our efforts on approximately inverting the operator through
-various means.
-
-In a discussion of methods for solving linear systems, several
+meaning that we can in fact compute $\ve{A}^{-1}$.
+
+In a discussion of methods for solving linear problems, several
 mathematical tools are useful in characterizing the qualities of the
@@
 Eq~(\ref{eq:eigenvalue_problem_2}) for a particular permutation of
-$\ve{x}$. The set of all permutations form the \textit{eigenvectors}
-of $\ve{A}$. A quantity of particular interest that is computable from
-the eigenvalues of a matrix $\ve{A}$ is the \textit{spectral radius},
-$\rho(\ve{A})$, defined by Saad \cite{saad_iterative_2003} as:
+$\ve{x}$ and the set of all permutations forms the
+\textit{eigenvectors} of $\ve{A}$. A quantity of particular interest
+that is computable from the eigenvalues of a matrix $\ve{A}$ is the
+\textit{spectral radius}, $\rho(\ve{A})$, defined by Saad
+\cite{saad_iterative_2003} as:
 \begin{equation}
@@
 \end{equation}
-in $\ve{H}$ where $\nu$ is interpreted as a particular random walk
-sequence permutation. We can generate these sequences of transitions
-through Monte Carlo random walks by assigning them both a probability
-and weight. As a reinterpretation of the iteration matrix, we then
-form the \textit{Neumann-Ulam decomposition} of \ve{H}:
+in $\ve{H}$ where $\nu$ is interpreted as a particular sequence
+permutation. We can generate these sequences of transitions through
+Monte Carlo random walks by assigning them both a probability and
+weight. As a reinterpretation of the iteration matrix, we then form
+the \textit{Neumann-Ulam decomposition} of \ve{H}:
 \begin{equation}
@@
 approximating. The formulation of $\ve{P}$ and $\ve{W}$ will be
-dependent on whether we choose a direct or adjoint Monte Carlo
+dependent on whether we choose a forward or adjoint Monte Carlo
 sequence to estimate the state transitions in
-Eq~(\ref{eq:expanded_neumann_solution}). In the direct method, we will
-use the provided linear operator $\ve{A}$ to form the Neumann Ulam
-decomposition while the adjoint method will use the adjoint linear
-operator ($\ve{A}^T$ for real-valued systems) to form the
+Eq~(\ref{eq:expanded_neumann_solution}). In the forward method, we
+will use the provided linear operator $\ve{A}$ to form the Neumann
+Ulam decomposition while the adjoint method will use the adjoint
+linear operator ($\ve{A}^T$ for real-valued systems) to form the
 decomposition.
@@
 contribution to the solution from a particular random walk permutation
-with $k$ total events is then the \textit{forward estimator}:
+with $k$ total events is then the \textit{forward
+  estimator}\footnote{The variance for this estimator is discussed in
+  Appendix~\ref{chap:estimator_variance}.}:
 \begin{equation}
@@
 \begin{equation}
-  P_{\nu} = P_{i_0=i} p_{i,i_1} p_{i_1,i_2} \cdots p_{i_{k-1},i_k}\:,
+  P_{\nu} = P_{(i_0=i)} p_{i,i_1} p_{i_1,i_2} \cdots p_{i_{k-1},i_k}\:,
   \label{eq:direct_permutation_probability}
 \end{equation}
-with $P_{i_0=i} = 1$ for the direct method. Finally, we define the
+with $P_{(i_0=i)} = 1$ for the direct method. Finally, we define the
 expectation value of $X$ to be the collection of all random walk
@@
 which, if expanded, directly recovers the exact solution by forming
-the Neumann series:
+the Neumann series through the Hadamard product:
 \begin{equation}
@@
 perform a predetermined number of random walks in order to generate an
-approximation for $\ve{x}$. If we are only to approximate the
-solution, we also need conditions by which we may terminate a random
-walk as the Neumann Ulam decomposition defined by
-Eqs~(\ref{eq:direct_probability}) and (\ref{eq:direct_weight}) will
+approximation for $\ve{x}$. We then also need conditions by which we
+may terminate a random walk as the Neumann Ulam decomposition defined
+by Eqs~(\ref{eq:direct_probability}) and (\ref{eq:direct_weight}) will
 create a random walk weight in Eq~(\ref{eq:direct_permutation_weight})
@@
 
-\subsubsection{Forward Estimator Variance}
-\label{subsec:forward_variance}
-We can compute the variance of the forward estimator through traditional
-methods by defining the variance, $\sigma_i$, for each component in
-the solution:
-\begin{equation}
-  {\sigma_i}^2 = E\{X_i - (\ve{A}^{-1}\ve{b})_i\}^2 = E\{X_i^2\} - x_i^2\:,
-  \label{eq:direct_variance_1}
-\end{equation}
-where the vector exponentials are computed element-wise. Inserting
-Eq~(\ref{eq:direct_expectation_value}) gives:
-\begin{equation}
-  \sigma_i^2 = \sum_{\nu} P_{\nu} X_{i}(\nu)^2 - x_i^2\:,
-  \label{eq:direct_variance_2}
-\end{equation}
-and applying Eq~(\ref{eq:direct_permutation_contribution}) and
-expanding the square of the summation:
-\begin{equation}
-  \sigma_i^2 = \sum_{\nu} P_{\nu} \Big(\sum_{m=0}^k \Big[ W_{m}^2 b_{i_m}^2
-  + 2 \sum_{j=0}^{m-1} W_m W_j b_{i_m} b_{i_j} \Big] \Big)-
-  x_i^2\:.
-  \label{eq:direct_variance_3}
-\end{equation}
-We can distribute the random walk probability to yield an expanded
-form of the variance:
-\begin{equation}
-  \sigma_i^2 = \sum_{\nu} P_{\nu} \sum_{m=0}^k W_{m}^2 b_{i_m}^2 + 2
-  \sum_{\nu} P_{\nu} \sum_{m=0}^k \sum_{j=0}^{m-1} W_m W_j b_{i_m}
-  b_{i_j} - x_i^2\:.
-  \label{eq:direct_variance_3_2}
-\end{equation}
-In particular, we will isolate the first term of the variance by
-expanding it:
-\begin{equation}
-  \sum_{\nu} P_{\nu} \sum_{m=0}^k W_{m}^2 b_{i_m}^2 =
-  \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots \sum_{i_k}^{N}
-  p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
-  w^2_{i,i_1}w^2_{i_1,i_2}\ldots w^2_{i_{k-1},i_k} b_{i_k}^2\:.
-  \label{eq:direct_variance_4}
-\end{equation}
-Using this expansion, we can arrive at a more natural
-reason for enforcing $\rho(\ve{H}) < 1$ for our Monte Carlo method to
-converge. Per the Hadamard product, we can concatenate the summation
-in Eq~(\ref{eq:direct_variance_4}):
-\begin{equation}
-  (\ve{P} \circ \ve{W} \circ \ve{W})^k =
-  \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots \sum_{i_k}^{N}
-  p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
-  w^2_{i,i_1}w^2_{i_1,i_2}\ldots w^2_{i_{k-1},i_k}\:.
-  \label{eq:double_weighted_decomposition}
-\end{equation}
-If we assign $\ve{G} = \ve{P} \circ \ve{W} \circ \ve{W}$ as in
-Eq~(\ref{eq:neumann_ulam_decomposition}), we then have:
-\begin{equation}
-   (\ve{P} \circ \ve{W} \circ \ve{W})^k =
-  \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-  \sum_{i_k}^{N}g_{i,i_1}g_{i_1,i_2}\ldots g_{i_{k-1},i_k}\:,
-  \label{eq:direct_variance_6}
-\end{equation}
-which is the general Neumann series for $\ve{G}$,
-\begin{equation}
-  \ve{T} = \sum_{k=0}^{\infty} \ve{G}^k\:,
-  \label{eq:variance_neumann_series}
-\end{equation}
-where $\ve{T} = (\ve{I}-\ve{G})^{-1}$. We can then insert $T$ back
-into the variance formulation for a more concise definition:
-\begin{equation}
-  \sigma^2_i = (\ve{T}\ve{b}^2)_i + 2 \sum_{\nu} P_{\nu} \sum_{m=0}^k
-  \sum_{j=0}^{m-1} W_m W_j b_{i_m} b_{i_j} - x_i^2\:.
-  \label{eq:direct_variance_5}
-\end{equation}
-We can relate $\ve{G}$ to $\ve{H}$ by noting that $\ve{G}$ simply
-contains an additional Hadamard product with the weight matrix. The
-Hadamard product has the property that:
-\begin{equation}
-  |\ve{H} \circ \ve{W}| \geq |\ve{H}|\ |\ve{W}|\:.
-  \label{eq:hadamard_inequality}
-\end{equation}
-Using the norm property of the Hadamard product and
-Eq~(\ref{eq:neumann_ulam_decomposition}), we can define the norm of
-$\ve{W}$ as:
-\begin{equation}
-  \frac{|\ve{H}|}{|\ve{P}|} \geq |\ve{W}|\:.
-  \label{eq:weight_norm}
-\end{equation}
-Choosing the infinity norm of the operator, the row normalized
-probability matrix will yield a norm of 1 giving the following
-inequality for relating $\ve{G}$ and $\ve{H}$:
-\begin{equation}
-  |\ve{G}| \geq |\ve{H}|^2
-  \label{eq:variance_norm_inequality}
-\end{equation}
-Using these relations to analyze Eq~(\ref{eq:direct_variance_5}), we
-see that if $\rho(\ve{G}) > 1$, then the first sum in
-Eq~(\ref{eq:variance_neumann_series}) will not converge and an
-infinite variance will arise as the elements of $\ve{T}$ become
-infinite in Eq~(\ref{eq:direct_variance_5}). We must restrict $\ve{G}$
-to alleviate this and therefore restrict $\ve{H}$ due to
-Eq~(\ref{eq:variance_norm_inequality}) with $\rho(\ve{H}) < 1$ so that
-our expectation values for the solution may have a finite variance.
-
 \subsubsection{Forward Method: Evolution of a Solution}
@@
 \label{sec:adjoint_mc}
-Often a more useful form, an alternative to forward Monte Carlo matrix
-inversion is the adjoint method. We begin by defining the linear
-system adjoint to Eq~(\ref{eq:linear_problem}):
+An alternative to forward Monte Carlo matrix inversion is the adjoint
+method. We begin by defining the linear system adjoint to
+Eq~(\ref{eq:linear_problem}):
 \begin{equation}
@@
 therefore has the same spectral radius. By defining the following
-inner product equivalence \cite{spanier_monte_1969}:
+adjoint inner product equivalence \cite{spanier_monte_1969}:
 \begin{equation}
@@
 \begin{equation}
-  P_{i_0=i}(\nu) = \frac{|b_i|}{||\ve{b}||_1}\:,
+  P_{(i_0=i)}(\nu) = \frac{|b_i|}{||\ve{b}||_1}\:,
   \label{eq:adjoint_source_probability}
@@
 \begin{equation}
-  b_{i_0} = W_0 P_{i_0=i}\:.
+  b_{i_0} = W_0 P_{(i_0=i)}\:.
   \label{eq:adjoint_source_definition}
@@
 from a particular random walk permutation of $k$ events is then the
-\textit{collision estimator}:
+\textit{collision estimator}\footnote{The variance for this estimator
+  is discussed in Appendix~\ref{chap:estimator_variance}.}:
 \begin{equation}
@@
     E\{X_j\} &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-    \sum_{i_k}^{N} b_{i_0} h_{i_0,i_1}h_{i_1,i_2}\ldots
-    h_{i_{k-1},i_k} \delta_{i_k,j} \\ &= x_{j}\:,
+    \sum_{i_k}^{N} b_{i_0} p_{i_0,i_1}p_{i_1,i_2}\ldots
+    p_{i_{k-1},i_k} w_{i_0,i_1}w_{i_1,i_2}\ldots
+    w_{i_{k-1},i_k} \delta_{i_k,j} \\ &= x_{j}\:,
   \end{split}
@@
 therefore, also providing an unbiased Monte Carlo estimate of the
-solution. Note that this expansion produces the effective sequence of
-matrix-vector multiplications with the $b_{i_0}$ component of the
-source vector which will be selected at random for each walk
-permutation and the unbiasedness of the estimator relies on an
-unbiased sampling of the source. It should also be noted here that
+solution. It should be noted here that
 Eq~(\ref{eq:adjoint_expectation_expansion}) only computes a single
@@
 simulation, we have $N$ simultaneous tallies for $\ve{A} \in
-\mathbb{R}^{N \times N}$ that will yield the entire solution
-vector.
+\mathbb{R}^{N \times N}$ that will yield the entire solution vector.
 
@@
 
-\subsubsection{Collision Estimator Variance}
-\label{subsec:forward_variance}
-We can compute the variance of the collision estimator in the same way
-as the direct estimator for each component in the solution where now:
-\begin{equation}
-  \sigma_j^2 = \sum_{\nu} P_{\nu} \sum_{m=0}^k W_{m}^2 \delta_{i_m,j}
-  + 2 \sum_{\nu} P_{\nu} \sum_{m=0}^k \sum_{l=0}^{m-1} W_m W_l
-  \delta_{i_m,j} \delta_{i_l,j} - x_j^2\:.
-  \label{eq:collision_variance_1}
-\end{equation}
-In this case, in the second summation $m \neq l$ for all states due to
-the form of the summation and therefore the combination of delta
-functions, $\delta_{i_m,j}$ of $\delta_{i_l,j}$, will only be nonzero
-for random walks that are in the current solution state (when $i_m =
-i_l = j$) and other states visited do not contribute to the variance
-of the current state (when $i_m \neq i_l$). This is important as it
-shows a decoupling of the variance among the solution vector states,
-meaning that the variance in solution state $i$ does not depend on the
-variance in solution state $j$.
-
-Expanding the transition probabilities in the first sum again yields a
-Neumann series in the same form as that explored for the forward
-estimator, this time with the terms in reverse order and the
-introduction of the Kronecker delta:
-\begin{multline}
-  \sigma_j^2 = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-  \sum_{i_k}^{N} p_{i_k,i_{k-1}}\ldots p_{i_2,i_1} p_{i_1,i_0}
-  w^2_{i_k,i_{k-1}}\ldots w^2_{i_2,i_1} w^2_{i_1,i_0}
-  b^2_{i_0}\delta_{i_k,j} + \\ 2 \sum_{\nu} P_{\nu} \sum_{m=0}^k
-  \sum_{l=0}^{m-1} W_m W_l \delta_{i_m,j} \delta_{i_l,j} - x_j^2\:.
-  \label{eq:collision_variance_2}
-\end{multline}
-If we again define $\ve{G} = \ve{P} \circ \ve{W} \circ \ve{W}$, we
-then have:
-\begin{multline}
-  \sigma^2_j = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
-  \sum_{i_k}^{N}g_{i_{k},i_{k-1}} \ldots
-  g_{i_2,i_1}g_{i_1,i_0}b_{i_0}\delta_{i_k,j} + \\ 2 \sum_{\nu}
-  P_{\nu} \sum_{m=0}^k \sum_{l=0}^{m-1} W_m W_l \delta_{i_m,j}
-  \delta_{i_l,j} - x_j^2\:,
-\label{eq:collision_variance_3}
-\end{multline}
-where now $\ve{G}$ in this case is the transpose of that used in
-Eq~(\ref{eq:direct_variance_6}). The Kronecker delta again implies
-that the variance contribution from each random walk will only be in
-its current state and for many random walks, many starting weights of
-$b_{i_0}$ will contribute to the variance in the same way as they
-contribute to the solution tally. As the transpose is formed and the
-Neumann series of $\ve{G}$ is in reverse order in
-Eq~(\ref{eq:collision_variance_3}) relative to its formulation in the
-forward estimator, then Eq~(\ref{eq:collision_variance_3}) and
-Eq~(\ref{eq:direct_variance_6}) are equivalent and therefore the
-collision estimator is bound by the same restrictions on
-$\rho(\ve{G})$ and $\rho(\ve{H})$ to ensure that the variance is
-finite.
-
 \subsubsection{Expected Value Estimator}
@@
 In addition to the collision estimator, an additional estimator is
-available due to the work of Okten \cite{okten_solving_2005} that
-uses the method of expected values as a means to improve the Monte
-Carlo estimate. As outlined by Spanier and Gelbard
+available due to the work of Okten \cite{okten_solving_2005} that uses
+the method of expected values as a means to improve the Monte Carlo
+estimate. As outlined by Spanier and Gelbard
 \cite{spanier_monte_1969}, the method of expected values is a
@@
 directly to discrete Monte Carlo by forming the \textit{expected value
-  estimator} for a random walk of $k$ events:
+  estimator}\footnote{The variance for this estimator is discussed in
+  Appendix~\ref{chap:estimator_variance}.} for a random walk of $k$
+events:
 \begin{equation}
@@
 estimator will provide no information.
-
-\subsubsection{Expected Value Estimator Variance}
-\label{subsubsec:expected_value_estimator_variance}
-Following the collision estimator variance, the expected value
-estimator variance is given as:
-\begin{multline}
-  \sigma_j^2 = \sum_{\nu} P_{\nu} b_j^2 + 2 \sum_{\nu} P_{\nu} b_j
-  \sum_{m=0}^k W_{m} h_{j,i_m} +\\ \sum_{\nu} P_{\nu} \sum_{m=0}^k
-  W_{m}^2 h_{j,i_m}^2 + 2 \sum_{\nu} P_{\nu} \sum_{m=0}^k
-  \sum_{l=0}^{m-1} W_m W_l h_{j,i_m} h_{j,i_l} - x_j^2\:.
-  \label{eq:expected_value_variance_1}
-\end{multline}
-As the delta functions are no longer present, the final summation in
-Eq~(\ref{eq:expected_value_variance_1}) contains contributions for all
-valid state combinations in the same row of the iteration matrix and
-therefore the variance of each state in the solution tally is coupled
-to a group of other states as defined by the sparsity pattern of the
-iteration matrix. This coupling of variances is expected due to the
-deterministic averaging used to generate the expected value estimator.
 
@@
 convergence rates \cite{halton_sequential_1962} and further refined
-his work some years later \cite{halton_sequential_1994}. Applications
-of his work by the transport community have confirmed convergence
-rates on the order of of $e^{-N}$ \cite{evans_residual_2003}. In much
-the same way as the projection methods outlined in
-Appendix~\ref{ch:linear_problem}, Halton's method, sequential Monte
-Carlo, utilizes the adjoint Monte Carlo solver as a means of directly
-reducing the elements residual vector. He proposed the following
-iterative scheme as a solution to Eq~(\ref{eq:linear_problem})\:
+his work some years later \cite{halton_sequential_1994} with
+applications of his work by the transport community confirming
+exponential convergence rates \cite{evans_residual_2003}. Halton's
+method, sequential Monte Carlo, utilizes the adjoint Monte Carlo
+solver as a means of directly reducing the elements residual
+vector. He proposed the following iterative scheme as a solution to
+Eq~(\ref{eq:linear_problem})\:
 \begin{subequations}
@@
 applied it more rigorously as a solution mechanism for the radiation
-diffusion equation \cite{evans_monte_2012}. With their new methods,
-they achieved numerical results identical to conventional Krylov
-solvers as well as comparable performance in both number of iterations
-and CPU time. Their approach was instead to use residual Monte Carlo
-as a synthetic acceleration for a stationary method. To derive this
-method, we begin by splitting the operator in
-Eq~(\ref{eq:linear_problem})
+diffusion equation \cite{evans_monte_2012}. Their approach was instead
+to use residual Monte Carlo as a synthetic acceleration for a
+stationary method. To derive this method, we begin by splitting the
+operator in Eq~(\ref{eq:linear_problem})
 \begin{equation}

### 2013-08-26 doc/dissertation/document/frontmatter/abstract.tex
moving estimator variance into an appendix

@@
 possess both high levels of concurrency and low memory availability
-per core compared to today's machines with a potentially rate of
+per core compared to today's machines with a potentially high rate of
 faults. Therefore, we seek new algorithms to solve these problems in
@@
 for the current implementation of MCSA, both weak and strong scaling
-were superior to production implementations of Krylov methods. In
-addition, the replication in the algorithm was able to boost parallel
-efficiencies by up to 20\% at large core counts in the strong scaling
-studies. By using the principle of superposition, replicating the
-Monte Carlo sequence was also demonstrated to enhance time to
-solution. Finally, it was discovered that MCSA exhibits the most
-scalable behavior when it is formulated as a stochastic realization of
-an additive Schwarz method.
+improved on that observed for production implementations of Krylov
+methods. In addition, the replication in the algorithm was able to
+boost parallel efficiencies by up to 20\% at large core counts in the
+strong scaling studies and enhance time to solution in the weak
+scaling stduies by using the principle of superposition. Finally, it
+was discovered that MCSA exhibits the most scalable behavior when it
+is formulated as a stochastic realization of an additive Schwarz
+method.
 
@@
 requires the new parallel algorithm developed by this work. With the
-further development of these methods based on this work, future
-nuclear reactor simulations may potentially be enabled that have an
-improved time to solution and more efficiently utilize the available
+further development of methods based on this work, future nuclear
+reactor simulations may potentially be enabled that have an improved
+time to solution and more efficiently utilize the available
 computational resources.

### 2013-08-26 doc/dissertation/document/chapters/introduction/introduction.tex
adding pauls comments through chapter 3

@@
 To achieve these goals, three complementary research and development
-activities are performed. First, Monte Carlo Sythetic Acceleration
+activities are performed. First, Monte Carlo Synthetic Acceleration
 (MCSA) is applied to the neutron transport problem in a light water

### 2013-08-26 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding pauls comments through chapter 3

@@
 difficulty, advanced preconditioning strategies for the $SP_N$ problem
-are required beyond simple Jacobi methods that can reduce the spectral
-radius into a region of better MCSA performance. Two modern algebraic
-preconditioning strategies, ILUT and SPAINV, will be presented here
-and used within the explicit preconditioning framework given by
-Eq~(\ref{eq:left_right_mcsa}). Data showing their effects on MCSA
-solutions of the fuel assembly criticality problem will be provided.
+are required, beyond simple Jacobi methods, that can reduce the
+spectral radius into a region of better MCSA performance. Two modern
+algebraic preconditioning strategies, ILUT and SPAINV, will be
+presented here and used within the explicit preconditioning framework
+given by Eq~(\ref{eq:left_right_mcsa}). Data showing their effects on
+MCSA solutions of the fuel assembly criticality problem will be
+provided.
 
@@
 assembly problem or other problems with a large scattering to
-absorption ratio. The algebraic formulation of ILUT requires the
-physics operator to be fully formed, also a requirement for MCSA, and
-is therefore easily incorporated into our solution sequence.
+absorption ratio \cite{saad_iterative_2003}. The algebraic formulation
+of ILUT requires the physics operator to be fully formed, also a
+requirement for MCSA, and is therefore easily incorporated into our
+solution sequence.
 
@@
 \end{equation}
-has a specified sparsity pattern and element magnitude. When a
-magnitude threshold is used, elements generated in the factors below
+has a specified sparsity pattern and element magnitude. When a drop
+tolerance threshold is used, elements generated in the factors below
 that magnitude are dropped, resulting in ILU threshold (ILUT)
@@
 For the explicit preconditioning scheme presented in
-Eq~(\ref{eq:left_right_mcsa}), the fully inverted operator must be
-generated in order to build the set of probabilities and weights
+Eq~(\ref{eq:left_right_mcsa}), the fully inverted preconditioners must
+be generated in order to build the set of probabilities and weights
 required for Monte Carlo sampling. Therefore, for a linear system of
-size $N$, $N$ triangular solves will be required in order to extract
-the inverse matrices from production ILU implementations. In addition,
+size $N$, $N$ triangular solves with the basis of each row in the
+system as a solution vector will be required in order to extract the
+inverse matrices from production ILU implementations. In addition,
 parallel implementations typically generate lower and upper factors
@@
 parametrically varied with the number of iterations to converge an
-eigenvalue iteration for the fuel assembly problem recorded along with
-the maximum number of non-zero entries observed in all matrix rows for
-the left/right preconditioned composite linear operator. To provide
-some sparsity to the factorization, the ILUT drop tolerance was used
-to drop elements in the extracted inverse triangular factors. For each
-calculation, the number of iterations required to converge reported
-was for a single eigenvalue iteration with \sn{3}{4} histories at each
-MCSA iteration to compute the Monte Carlo correction using the adjoint
-collision estimator. All ILUT calculations reported here were
-performed on a single CPU and therefore this data does not take into
-account the aforementioned effects of parallel decomposition on the
-quality of the ILUT preconditioning.
+eigenvalue iteration for the fuel assembly problem recorded. In
+addition, the maximum number of non-zero entries observed in all rows
+for the left/right preconditioned composite linear operator was
+recorded to observe the effects of the explicit preconditioner
+inversion and matrix-matrix multiplication on sparsity. To provide
+some sparsity to the factorization, an additional step was added with
+the ILUT drop tolerance used to drop elements in the explicitly
+inverted preconditioners. For each calculation, the number of
+iterations reported was for a single eigenvalue iteration with
+\sn{3}{4} histories at each MCSA iteration to compute the Monte Carlo
+correction using the adjoint collision estimator. All ILUT
+calculations reported here were performed on a single CPU and
+therefore this data does not take into account the aforementioned
+effects of parallel decomposition on the quality of the ILUT
+preconditioning.
 
@@
 entries to a minimum value in order to lower the amount of coupling
-among states in the system and reduce the amount of memory used
-along with the potential latency overhead. Given these two objectives,
-the following metric is proposed:
+among states in the system and reduce the amount of memory used along
+with the potential latency overhead. Given these two objectives, the
+following metric is proposed:
 \begin{equation}
@@
 where the highest quality preconditioning is one that minimizes this
-metric. For the ILUT preconditioned data provided,
+metric. This metric gives equal importance to both iterative
+performance and sparsity with the purpose of simply measuring their
+combined effect rather than seeking some optimal relationship of
+importance. For the ILUT preconditioned data provided,
 Figure~\ref{fig:ilut_quality} provides the computed metric as a
@@
 \label{subsec:spn_spainv_preconditioning}
-Sparse approximate inverse (SPAINV) preconditioning is chosen next because the true inverse of the preconditioner
-is product of the algorithm, avoiding the need for many inverse matrix
-applications as required by using ILUT with MCSA. In addition, the
-algorithm seeks to maintain the sparsity of the inverse
-preconditioner, potentially alleviating some of the dense operator
-issues observed with ILUT.
+Sparse approximate inverse (SPAINV) preconditioning is chosen next
+because the true inverse of the preconditioner is the product of the
+algorithm, avoiding the need for many inverse matrix applications as
+required by using ILUT with MCSA. In addition, the algorithm seeks to
+maintain the sparsity of the inverse preconditioner, potentially
+alleviating some of the dense operator issues observed with ILUT.
 
@@
 of magnitude slower than a threshold of 0.01 due to the inclusion of a
-significant number of extra values in the input operator. In addition
-to comparable iterative performance, the sparsity of the composite
-operator is greatly improved with nearly an order of magnitude
-reduction in number of non-zero entries per row for lower level
-values.
+significant number of extra values in the input operator,
+$\mathbf{A}$. In addition to comparable iterative performance, the
+sparsity of the composite operator is greatly improved with nearly an
+order of magnitude reduction in number of non-zero entries per row for
+lower level values.
 
@@
 reduced domain approximation to eliminate those elements that either
-fall below a specified tolerance level, keep the $N$ largest elements
-in each row, or both giving:
+fall below a specified threshold level, fall within a fill level $N$
+where we keep the $N$ largest elements in each row, or both giving:
 \begin{subequations}
@@
 
-Table~\ref{tab:spn_comparison_iterations} gives the average number of
+Table~\ref{tab:spn_comparison_iterations} gives this average number of
 linear solver iterations required to converge a single eigenvalue
@@
 methods to perform well due to the structure of the projection
-scheme. It important to note that the iterative performance of each
+scheme. It is important to note that the iterative performance of each
 solver is not a strong function of the number of energy groups in the
@@
 used there are $O(100)$ elements in each row of the system and each of
-those elements must be processed during a transition in the Monte
-Carlo random walk sequence. In addition, although the preconditioned
-fixed point iterations used in the MCSA sequence do not use the
-composite linear operator but rather a sequence of matrix vector
-multiplies to achieve the same preconditioning effect, they do use the
-explicitly extracted inverse of the preconditioning matrices which
-themselves are dense, leading to exceedingly slow computation times
-even in the fixed point iteration.
+those elements must be accessed for next-event sampling and tally
+vector modification during a transition in the Monte Carlo random walk
+sequence. In addition, although the preconditioned fixed point
+iterations used in the MCSA sequence do not use the composite linear
+operator but rather a sequence of matrix vector multiplies to achieve
+the same preconditioning effect, they do use the explicitly extracted
+inverse of the preconditioning matrices which themselves are dense,
+leading to exceedingly slow computation times even in the fixed point
+iteration.
 
@@
 Based on the performance results in this section, MCSA shows promise
-as a competitive and perhaps even superior method for solutions to the
-neutron transport problem discretized with the $SP_N$
-approximation. Not only are the correct answers produced when compared
-to production linear solvers, but the iterative performance is
-comparable to Krylov methods with identical preconditioning that would
-typically be used every-day calculations. From a CPU timing
-perspective, the methods yield the same time complexity as a function
-of the number of energy groups in the problem when compared to the
-Krylov methods. Here, a large time constant is generated due to the
-explicit preconditioning strategy and the generation of dense linear
-operators as a result. To improve these results and put general MCSA
-schemes into a performance regime where they are competitive with
+as a competitive method for solutions to the neutron transport problem
+discretized with the $SP_N$ approximation and could potentially be
+superior if the preconditioning strategy were improved to reduce time
+to solution and memory usage. Not only are the correct answers
+produced when compared to production linear solvers, but the iterative
+performance is comparable to Krylov methods with identical
+preconditioning that would typically be used most calculations. From a
+CPU timing perspective, the methods yield the same time complexity as
+a function of the number of energy groups in the problem when compared
+to the Krylov methods. Here, a large time constant is generated due to
+the explicit preconditioning strategy and the generation of dense
+linear operators as a result. To improve these results and put general
+MCSA schemes into a performance regime where they are competitive with
 Krylov methods for neutron transport problems, significant research
@@
 \item The explicit preconditioning strategy required to overcome the
-  MCSA spectral radius restriction creates runtimes $O(100)$ larger
+  MCSA spectral radius restriction creates run times $O(100)$ larger
   than the production Krylov methods for the fuel assembly criticality

### 2013-08-26 doc/dissertation/document/frontmatter/abstract.tex
adding pauls comments through chapter 3

@@
 strong scaling studies and enhance time to solution in the weak
-scaling stduies by using the principle of superposition. Finally, it
+scaling studies by using the principle of superposition. Finally, it
 was discovered that MCSA exhibits the most scalable behavior when it

### 2013-08-26 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
adding pauls changes to chapter 4

@@
 where $\eta^k \in [0,1)$ is defined as the \textit{forcing term} at
-  the $k$ iterate. Eq~(\ref{eq:inexact_newton_forcing}) then states
-  that the residual generated by the linear solver is bound by the
-  nonlinear residual and how tightly it is bound is defined by the
-  forcing term. This is useful in that we can vary how tightly coupled
-  the convergence of the linear iterations used to generate the Newton
+  nonlinear iteration $k$, the right-hand side is the nonlinear
+  residual at nonlinear iteration $k$, and the left-hand side is the
+  residual for the linear model at nonlinear iteration
+  $k$. Eq~(\ref{eq:inexact_newton_forcing}) then states that the
+  residual generated by the linear solver is bound by the nonlinear
+  residual and how tightly it is bound is defined by the forcing
+  term. This is useful in that we can vary how tightly coupled the
+  convergence of the linear iterations used to generate the Newton
   correction is to the nonlinear iteration by relaxing or tightening
@@
 
-At first glance, FANM is simply a variant of an inexact Newton method
-with a specific requirement of a fully formed Jacobian. However, FANM
-has the potential to provide value in several areas. First, if
-iterative or parallel performance of MCSA can be demonstrated to
-better than that of GMRES or similar subspace methods for asymmetric
-systems, FANM may outperform a Newton-Krylov method. Second, for
-situations where memory is a constraint, especially in situations
-where the Jacobian is generated often if not at every Newton
-iteration, not building a subspace for ill-conditioned problems will
-alleviate the associated cost. However, until the preconditioning
-issues associated with MCSA as presented in the previous chapter using
-the $SP_N$ equations can be resolved, this hypothesis can
-unfortunately not be researched. Explicit algebraic preconditioning
-used in conventional solution methods will almost certainly be
-required to reduce the spectral radius of most Jacobian operators for
-MCSA to apply and therefore the current preconditioning strategy will
-eliminate all potential memory benefits of FANM.
+Upon preliminary inspection, FANM is simply a variant of an inexact
+Newton method with a specific requirement of a fully formed
+Jacobian. However, FANM has the potential to provide value in several
+areas. First, if iterative or parallel performance of MCSA can be
+demonstrated to better than that of GMRES or similar subspace methods
+for asymmetric systems, FANM may outperform a Newton-Krylov
+method. Second, for situations where memory is a constraint,
+especially in situations where the Jacobian is generated often if not
+at every Newton iteration, not building a subspace for ill-conditioned
+problems will alleviate the associated cost. However, until the
+preconditioning issues associated with MCSA as presented in the
+previous chapter using the $SP_N$ equations can be resolved, this
+hypothesis can unfortunately not be researched. Explicit algebraic
+preconditioning used in conventional solution methods will almost
+certainly be required to reduce the spectral radius of most Jacobian
+operators for MCSA to apply and therefore the current preconditioning
+strategy will eliminate all potential memory benefits of FANM.
 
@@
 stagnate. Typically this happens because the Richardson iteration and
-subsequently the Neumann series will quickly damp the higher order
+subsequently the Neumann series will quickly dampen the higher order
 error modes on the scale of the computational grid used to solve the
@@
 difficulty even though it is preconditioned in an identical
-manner. This means that the GMRES solver at each Newton step is
-converging the linear residual to a smaller tolerance than the MCSA
-solver as the forcing terms adapt to the convergence of the method
-over the course of nonlinear iterations. For this problem, tighter
-convergence of the linear model at later Newton steps also improved
-convergence of nonlinear model, resulting in the extra FANM iteration.
+manner. This means that GMRES at each Newton step is converging the
+linear residual to a smaller tolerance than MCSA as the forcing terms
+adapt to the convergence of the method over the course of nonlinear
+iterations. For this problem, tighter convergence of the linear model
+at later Newton steps also improved convergence of nonlinear model,
+resulting in the extra FANM iteration.
 
@@
 In addition to nonlinear iterative performance, performance of the
-inner linear solves is also of interest. Here, we seek to maintain a
+inner linear solves is also of interest. Here, we seek to minimize the
 minimum number of total linear solver iterations over the course of
@@
 parameters were the same for both calculations, the MCSA iteration was
-pushed harder by doubling the number of histories used for the
-computation at each nonlinear iteration. These extra histories
-provided better convergence of the linear model at each iteration and
-did not result in oversolving, thereby increasing the convergence rate
-of the nonlinear model as well. This result is significant in that it
-shows increasing the number of histories used for each MCSA iteration
-can improve the performance of both linear and nonlinear convergence.
+pushed harder at a Reynolds number of 700 by doubling the number of
+histories used for the computation at each nonlinear iteration. These
+extra histories provided better convergence of the linear model at
+each iteration and did not result in over-solving, thereby increasing
+the convergence rate of the nonlinear model as well. This result is
+significant in that it shows increasing the number of histories used
+for each MCSA iteration can improve the performance of both linear and
+nonlinear convergence. However, further research is required to study
+the relationship between the acceleration parameters and the nonlinear
+convergence to confirm this for a broader range of problems.
 
@@
 forcing terms at later Newton iterations for MCSA were much smaller
-than those for GMRES. This means that MCSA was converged to a tighter
-tolerance at these iterations and therefore required the extra
-iterations.
+than those for GMRES. Per Eq~(\ref{eq:newton_stopping_criteria}), this
+means that MCSA was converged to a smaller tolerance at these
+nonlinear iterations than GMRES and therefore required extra linear
+iterations to meet the convergence criteria.
 
@@
 number of 500 is observed to use many more MCSA iterations. For
-Reynolds numbers of both 200 and 500, FANM also reduced the number of
+Reynolds numbers of 200, 300, and 500, FANM also reduced the number of
 Newton steps need when compared to the Newton-Richardson scheme.
@@
 the linear model if the iterative performance is comparable. This is
-indeed the case for the 400 Reynolds number case where the linear
-model is better conditioned and FANM used fewer total linear solver
-iterations to converge the nonlinear residual. However, at a Reynolds
-number of 500 FANM has larger forcing terms but requires many more
-MCSA iterations to converge the last 2 Newton steps. Therefore, these
-extra iterations in this case must come from ill-conditioning.
+indeed the case for the 400 Reynolds number case where FANM used fewer
+total linear solver iterations to converge the nonlinear
+residual. However, at a Reynolds number of 500 FANM has larger forcing
+terms but requires many more MCSA iterations to converge the last 2
+Newton steps. Therefore, these extra iterations in this case must come
+from ill-conditioning.
 
@@
   high Reynolds numbers
+\item Using more Monte Carlo histories at every FANM iteration was
+  observed to reduce the number nonlinear iterations required to
+  converge the lid-driven cavity problem at high Reynolds numbers
 \item The spectral radius convergence restriction on MCSA was observed
@@
   $O(100)$-$O(1,000)$ slower than the Newton-Krylov solver
-\item More Monte Carlo histories at every FANM iteration can reduce
-  the number of linear and nonlinear iterations required to converge
-  the problem
 \end{itemize}

### 2013-08-26 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
whitespace formatting
adding pauls corrections to parallel chapter

@@
 domain directly to the left. Before the scattering event in the new
-domain may be processed, particle A must be communicated between
-between the two parallel processes that own those domains. For any
-given particle in the transport simulation, this communication event
-may hardly occur during the lifetime of the particle (particle B), or
-may occur many times (particle C) depending on the problem parameters
-and the random path in phase space taken by the particle. Scalable
-parallel algorithms for domain decomposed Monte Carlo are those that
-handle this domain-to-domain communication of particles effectively
-and balance it with the on-process computations for particle
-interactions, tallies, response calculations, and other simulation
-requirements.
+domain may be processed, particle A must be communicated between the
+two parallel processes that own those domains. For any given particle
+in the transport simulation, this communication event may hardly occur
+during the lifetime of the particle (particle B), or may occur many
+times (particle C) depending on the problem parameters and the random
+path in phase space taken by the particle. Scalable parallel
+algorithms for domain decomposed Monte Carlo are those that handle
+this domain-to-domain communication of particles effectively and
+balance it with the on-process computations for particle interactions,
+tallies, response calculations, and other simulation requirements.
 \begin{figure}[t!]
@@
 Compared to the serial construction of the Neumann-Ulam decomposition,
-data for all states that is required to be on process must be
-collected in parallel so that all possible local transitions may be
-completed. For the pure domain decomposition case, given an input
-parallel decomposition for the Neumann-Ulam decomposition, for all
-local states $m$ we require access to $P_{mn}$ and $W_{mn}$ (or
-$P^T_{mn}$ and $W^T_{mn}$ for the adjoint method) for all possible
-states $n$. As an example of this state collection,
-Figure~\ref{fig:diffusion_graph} gives the adjacency graph of a single
-state in the neutron diffusion matrix for the model problem presented
-in Appendix~\ref{chap:diffusion_problem}.
+data for all states that are required to be on process must be
+collected in parallel so that all possible transitions that leave the
+local domain may be correctly communicated. As an example of this
+state collection, Figure~\ref{fig:diffusion_graph} gives the adjacency
+graph of a single state in the neutron diffusion matrix for the model
+problem presented in Appendix~\ref{chap:diffusion_problem}. The
+discretization stencil of the diffusion problem dictates the structure
+of this graph and the states to which a history may transition.
 
@@
 directly correlating to physical location $(i,j)$ in the mesh (the
-center node). The discretization stencil of the diffusion problem
-dictates the structure of this graph and the states to which a history
-may transition. If we play the Monte Carlo game to move to a new state
+center node). If we play the Monte Carlo game to move to a new state
 $n$, then we may move to any of the nodes in this graph, including the
@@
 state $n$ corresponding to mesh point $(i+1,j)$, we are now in a state
-that is owned by the adjacent domain. We must then gather the data in
-the Neumann-Ulam decomposition from the neighboring process that owns
-grid point $(i+1,j)$. Doing this provides us with the data required by
-the estimators permitting $P_{mn}$ and $W_{mn}$ to be computed locally
-for this particular boundary transition. In addition, we also collect
-the identification number for the process that owns this state such
-that we have an address to send all histories that leave the local
-domain boundary with this state as their destination. For the parallel
-source, these adjacent states are not required as histories may only
-be born in states in the local domain. Once the local Neumann-Ulam
-decomposition has been generated with the proper data collected from
-adjacent domains, Monte Carlo transport may proceed in parallel.
+that is owned by the adjacent domain. We must then gather the
+identification number for the neighboring process that owns grid point
+$(i+1,j)$ such that we have an address to send all histories that
+leave the local domain boundary with this state as their
+destination. For the parallel source, these adjacent states are not
+required as histories may only be born in states in the local
+domain. Once the local Neumann-Ulam decomposition has been generated
+with the proper communication data collected from adjacent domains,
+Monte Carlo transport may proceed in parallel.
 
@@
     \ForAll{neighbors}
-    \State post nonblocking receive for maximum history buffer size
+    \State post non-blocking receive for maximum history buffer size
     \State allocate history buffer
@@
     \ForAll{child processes}
-    \State post nonblocking receive for \textit{historiesCompleted} tally
+    \State post non-blocking receive for \textit{historiesCompleted} tally
     \EndFor
-    \State post nonblocking receive for stop message from parent
+    \State post non-blocking receive for stop message from parent
     \While{stop flag not set}
@@
     \EndWhile
-    \State cancel outstanding nonblocking requests
+    \State cancel outstanding non-blocking requests
     \State free all history buffers
@@
 required with data buffers allocated with a user-defined size for
-incoming and outgoing histories. This nonblocking structure is
+incoming and outgoing histories. This non-blocking structure is
 critical to the performance of the algorithm in that it permits local
@@
 this communication pattern, each process has one parent process
-(except for the MASTER process 0) to which it will nonblocking send
+(except for the MASTER process 0) to which it will non-blocking send
 the number of histories that terminated their transport procedure by
@@
 to two child processes from which it will receive their completed
-number of histories with a nonblocking receive operation. Setting up a
+number of histories with a non-blocking receive operation. Setting up a
 tree in this manner lets the completed history tally
@@
 histories in the source, transport is complete. Once this occurs, the
-root process nonblocking sends the stop message to its children and
-each process nonblocking receives a stop message from its parent as
+root process non-blocking sends the stop message to its children and
+each process non-blocking receives a stop message from its parent as
 shown in Figure~\ref{fig:binary_comm_tree}. The stop message is then
@@
 outgoing history buffer for the neighboring domain and if the buffer
-is full, it is sent to the neighboring domain with a nonblocking
+is full, it is sent to the neighboring domain with a non-blocking
 operation and the memory associated with that buffer reallocated. In
@@
     \If{neighbor buffer is full}
-    \State nonblocking send history buffer to neighbor
+    \State non-blocking send history buffer to neighbor
     \State allocate new history buffer for neighbor
@@
 the problem, if there are histories in those data buffers then they
-are added to the stack for processing and the nonblocking receive
+are added to the stack for processing and the non-blocking receive
 operation re-instantiated. In addition, the terminated histories count
@@
     \State add the histories to the running stack
-    \State repost nonblocking receive with neighbor
+    \State re-post non-blocking receive with neighbor
     \EndFor
@@
     \State \textit{historiesCompleted} += message value
-    \State repost nonblocking receive with child
+    \State re-post non-blocking receive with child
     \EndFor
@@
 full. If Algorithm~\ref{alg:control_termination} is called by the
-MASTER process, it checks for completion of the problem and if it is
-complete, forwards to stop flag onto its children as in
+MASTER process, it checks for completion of the problem and if
+complete, forwards the stop flag onto its children as in
 Figure~\ref{fig:binary_comm_tree}. If the process is not the MASTER,
-then we check for a stop signal from the parent and forward to the
-child processes if transport is complete.
+then we check for a stop signal from the parent and forward it to the
+child processes.
 
@@
   \begin{algorithmic}[1]
-    \State nonblocking send an partially full history buffers
-    \ForAll{\textit{historiesCompleted} messages from children}
-    \State \textit{historiesCompleted} += message \textit{historiesCompleted}
-    \State repost nonblocking receive with child
-    \EndFor
-    \If{MASTER processor}
-    \If{\textit{historiesCompleted} == global \# of source histories}
-    \State set stop flag
-    \ForAll{children}
-    \State nonblocking send stop message to child
-    \EndFor
-    \EndIf
-    \Else
-    \State nonblocking send \textit{historiesCompleted} to parent
-    \State \textit{historiesCompleted} = 0
-    \State check for stop signal from parent
-    \If{stop signal from parent}
-    \ForAll{children}
-    \State nonblocking send stop message to child
-    \EndFor
-    \EndIf
-    \EndIf
+    \State non-blocking send all partially full history buffers
+    \ForAll{\textit{historiesCompleted} messages from children} \State
+    \textit{historiesCompleted} += message \textit{historiesCompleted}
+    \State re-post non-blocking receive with child \EndFor \If{MASTER
+      processor} \If{\textit{historiesCompleted} == global \# of
+      source histories} \State set stop flag \ForAll{children} \State
+    non-blocking send stop message to child \EndFor \EndIf \Else \State
+    non-blocking send \textit{historiesCompleted} to parent \State
+    \textit{historiesCompleted} = 0 \State check for stop signal from
+    parent \If{stop signal from parent} \ForAll{children} \State
+    non-blocking send stop message to child \EndFor \EndIf \EndIf
   \end{algorithmic}
@@
       \input{chapters/parallel_mc/master_comm_tree.pdftex_t} }
+  \end{center}
     \caption{\textbf{Master/slave scheme for coordinating the end of a
@@
         subdomains are indexed by an integer.}}
-  \end{center}
   \label{fig:master_comm_tree}
@@
 provide value in the future. In addition to a robust fully
-asynchronous communication pattern, this algorithm may also be modified
-to account for situations where the total number of histories in a
-given stage are not known before starting transport. From a physics
-perspective, we might expect this for situations where perhaps an
-(n,2n) interaction occurs in a neutronics problem. In this case, the
-algorithm is modified to account for both histories terminated and
+asynchronous communication pattern, this algorithm may also be
+modified to account for situations where the total number of histories
+in a given stage are not known before starting transport. From a
+physics perspective, we might expect this for situations where perhaps
+an (n,2n) interaction occurs in a neutronics problem. In this case,
+the algorithm is modified to account for both histories terminated and
 created and several mechanisms are introduced to determine completion
 of the transport stage. For future variations of this work, certain
-variance reduction techniques which create histories, such as
+variance reduction techniques that create histories, such as
 splitting, have the potential to be successfully employed as a means
@@
 \label{subsec:msod_generation}
-In \S~\ref{subsec:domain_generation} we discussed how we generated the
+In \S~\ref{subsec:domain_generation} we discussed how to generate the
 parallel transport domain from the Neumann-Ulam decomposition of a
@@
 Figure~\ref{fig:diffusion_graph}. To generate the boundary for the
-local transport domain, we needed to gather all data from the
-immediately adjacent states in the system that did not already exist
-in the local domain. We solved this problem by traversing the graph
-of the matrix for each of the boundary states and determining which
-processes owned those adjacent states.
+local transport domain, we needed to gather the identification numbers
+of all owning processes from the immediately adjacent states in the
+system. We solved this problem by traversing the graph of the matrix
+for each of the boundary states and determining which processes owned
+those adjacent states.
 
@@
 we perform as many graph level traversals as the specified amount of
-overlap. For example, consider the analytic relations derived in
-Appendix~\ref{chap:parallel_theory} for the length of random walks in a
-Neumann-Ulam method and the amount of leakage from a domain
+overlap and collect additional information. For the overlapping domain
+decomposition case, given an input parallel decomposition for the
+Neumann-Ulam decomposition, for all local states $m$, including those
+in the overlap, we require access to $P_{mn}$ and $W_{mn}$ (or
+$P^T_{mn}$ and $W^T_{mn}$ for the adjoint method) for all possible
+states $n$. Doing this provides us with the data required by the
+estimators permitting $P_{mn}$ and $W_{mn}$ to be computed locally for
+any boundary transition.
+
+As an example, consider the analytic relations derived in
+Appendix~\ref{chap:parallel_theory} for the length of random walks in
+a Neumann-Ulam method and the amount of leakage from a domain
 considering a neutron diffusion problem. From these relations we might
@@
 
+\clearpage
+
 \begin{figure}[t!]
@@
 within all sets. We can take advantage of this by building a processor
-space that encapsulates each identical domain in each set and then use
-this processor space for the reduction. For the schematic in
+space that encapsulates each identical domain across all sets and then
+use this processor space for the reduction. For the schematic in
 Figure~\ref{fig:msod_tally_reduction}, the combined upper-left domains
@@
       check period vs. history buffer size sensitivity.} \textit{The
-      color scale represents the ratio of the runtime of a particular
-      parameter variation with the runtime from the faster parameter
+      color scale represents the ratio of the run time of a particular
+      parameter variation with the run time from the faster parameter
       variation. A value of 1.02 on the color scale corresponds to a
-      runtime 2\% longer than the faster time. All runtimes were
+      run time 2\% longer than the faster time. All run times were
       within 6.5\% of the fastest observed case.}}
@@
 Compared to the reference work, there was also not a significant
-change in the runtime over the values of parameters tested with all
-runtimes reported within 6.5\% of the fastest case observed. For this
+change in the run time over the values of parameters tested with all
+run times reported within 6.5\% of the fastest case observed. For this
 work, a history buffer size of 4096 and a buffer check period of 128
@@
 thus causing improved serial performance and therefore enhanced
-runtimes. In this region of super-linear efficiencies, we also observe
+run times. In this region of super-linear efficiencies, we also observe
 larger efficiencies for GMRES when compared to conjugate
@@
 consume much more time than the basic conjugate gradient operations,
-causing a larger serial runtime and subsequently higher
+causing a larger serial run time and subsequently higher
 efficiencies. For the same reasons, we also observe higher
@@
 
-\begin{figure}[t!]
+\begin{figure}[p!]
   \begin{center}
@@
 the final set of weak scaling computations. When this reduction in
-iterations is used to account for the observed runtimes and computed
+iterations is used to account for the observed run times and computed
 absolute efficiencies for weak scaling we end up with the
@@
 iteration for each method. If optimization has been performed and
-runtimes potentially approach those of the Krylov methods, this
+run times potentially approach those of the Krylov methods, this
 comparison may not be as favorable. Overall, weak scaling performance
@@
   \label{fig:titan_weak_implementation}
+\end{figure}
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_bvsm.pdf}
+  \end{center}
+  \caption{\textbf{Binary tree vs. master/slave communication scheme
+      strong scaling absolute efficiency.} \textit{The master/slave
+      scheme will not scale beyond $O(1,000)$ cores due to load
+      imbalance.}}
+  \label{fig:titan_strong_bvsm}
+\end{figure}
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_bvsm.pdf}
+  \end{center}
+  \caption{\textbf{Binary tree vs. master/slave communication scheme
+      weak scaling absolute efficiency.}  \textit{The master/slave
+      scheme will not scale beyond $O(1,000)$ cores due to load
+      imbalance.}}
+  \label{fig:titan_weak_bvsm}
 \end{figure}
@@
 
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_bvsm.pdf}
-  \end{center}
-  \caption{\textbf{Binary tree vs. master/slave communication scheme
-      strong scaling absolute efficiency.} \textit{The master/slave
-      scheme will not scale beyond $O(1,000)$ cores due to load
-      imbalance.}}
-  \label{fig:titan_strong_bvsm}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_bvsm.pdf}
-  \end{center}
-  \caption{\textbf{Binary tree vs. master/slave communication scheme
-      weak scaling absolute efficiency.}  \textit{The master/slave
-      scheme will not scale beyond $O(1,000)$ cores due to load
-      imbalance.}}
-  \label{fig:titan_weak_bvsm}
-\end{figure}
+\clearpage
 
@@
 In the next set of scaling studies, we will explore how adding overlap
-to the subdomains in the problem affects runtimes and parallel
+to the subdomains in the problem affects run times and parallel
 performance as compared to the case with pure domain decomposition. To
@@
 provides the same global number of histories used to compute the
-correction in the single set problem. For the replicated history case,
-each additional set in the problem will use just as many histories as
-the single set case. Using 2 sets with fully replicated histories for
-our example problem then has each set computing 20,000 histories with
+correction in the single set problem and we should therefore expect
+the same iterative performance at a reduced time per iteration and
+thus reduced time to solution. For the replicated history case, each
+additional set in the problem will use just as many histories as the
+single set case. Using 2 sets with fully replicated histories for our
+example problem then has each set computing 20,000 histories with
 40,000 global histories in the computation. For the 4 set case, the
-number of global histories is increased to 80,000.
+number of global histories is increased to 80,000. By increasing the
+global number of histories, the replicated cases should approximately
+maintain time per iteration but should decrease the number of
+iterations required to converge and subsequently reduce time to
+solution.
 
@@
 case actually performs the fastest with only marginal improvement in
-runtimes at higher numbers of cores with more sets. At these larger
+run times at higher numbers of cores with more sets. At these larger
 numbers of cores, however, the improvements are so modest that one is
@@
 the 2 set case with history replication operates at 58\%
-efficiency. This is reasonable improvement in the use of those
-computational resources, gained by replicating the problem. Beyond
-16,384 cores, there is not a compelling reason to use more
-computational resources for this problem.
+efficiency. Although the efficiencies here are still low, this is
+reasonable improvement in the use of those computational resources,
+gained by replicating the problem. Beyond 16,384 cores, there is not a
+compelling reason to use more computational resources for this
+problem.
 
@@
 that the replicated history cases are even less efficient than in the
-absolute case with respect to the performance of a single iteration.
+absolute case with respect to the performance of a single
+iteration. This is due to the fact that doubling the number of
+histories does not halve the number of MCSA iterations required for
+convergence as the acceleration is bound by the Central Limit
+Theorem. As a result, the improved iterative performance from the
+replicated cases comes at a higher cost in terms of time to solution
+than the split cases.
 
@@
 replicated histories to assess the effects on scalability and
-runtime. Figure~\ref{fig:titan_weak_ms_time} gives the wall times from
+run time. Figure~\ref{fig:titan_weak_ms_time} gives the wall times from
 the multiple sets weak scaling study. Of primary importance here is
@@
 the 2 set case to be a perfect use of resources with either split or
-replicated histories, the runtime of that calculation must be half of
+replicated histories, the run time of that calculation must be half of
 that for the single set case. This was not the observation for the
-runtimes in Figure~\ref{fig:titan_weak_ms_time} and therefore the
+run times in Figure~\ref{fig:titan_weak_ms_time} and therefore the
 single set case is still the most efficient use of computational
@@
 study with the iterations to converge and computed algorithmic
-efficiencies given by Table~\ref{tab:ms_weak_alg_eff}. Again, when the
-implementation efficiencies are computed, we see in
+efficiencies given by Table~\ref{tab:ms_weak_alg_eff}. The iterations
+to converge these problems were not a function of problem size. Again,
+when the implementation efficiencies are computed, we see in
 Figure~\ref{fig:titan_weak_ms_impeff} that splitting histories among
@@
 first. This truncation has the potential to increase the stochastic
-uncertainty of the Neumann-Ulam result and therefore increase the
-number of MCSA iterations required to converge. In addition, the
+uncertainty and bias of the Neumann-Ulam result and therefore increase
+the number of MCSA iterations required to converge. In addition, the
 scalability of the method will then be purely limited by the parallel
@@
 super unitary efficiencies for the subdomain case as there was a
-substantial improvement in runtimes relative to the 16-core base case
+substantial improvement in run times relative to the 16-core base case
 with Monte Carlo communication.

### 2013-08-26 doc/dissertation/document/chapters/conclusion/conclusion.tex
adding pauls corrections to the conclusion

@@
 
-It was found that the leakage of histories from domain to domain in
-the parallel Monte Carlo algorithm could in fact be quantified
-analytically using the algebraic properties of the system. These
-relationships were then used to determine the amount of overlap one
-may require in the parallel algorithm to reduce communication costs
-and increase parallel efficiencies. Scaling studies showed that in the
-strong case overlap in small quantities on the order of the
-mean-free-path of a stochastic history in the simulation could boost
-parallel efficiencies by a few percent. However, it was found that
-this additional overlap was not effective in boosting strong scaling
-efficiencies. It was found that in general overlap was not very
-effective due to the fact that the parallel communication saved during
-the Monte Carlo transport sequence is simply deferred until after
-transport is complete when it manifests itself as an overlapping
-parallel vector reduction operation.
+Outlined in Appendix~\ref{chap:parallel_theory}, it was found that the
+leakage of histories from domain to domain in the parallel Monte Carlo
+algorithm could in fact be quantified analytically using the algebraic
+properties of the system. These relationships were then used to
+determine the amount of overlap one may require in the parallel
+algorithm to reduce communication costs and increase parallel
+efficiencies. Scaling studies showed that in the strong case overlap
+in small quantities on the order of the mean-free-path of a stochastic
+history in the simulation could boost parallel efficiencies by up to
+10\% in isolated cases. However, it was found that this additional
+overlap was not effective in boosting weak scaling efficiencies. In
+general, overlap was not very effective due to the fact that the
+parallel communication saved during the Monte Carlo transport sequence
+is simply deferred until after transport is complete when it manifests
+itself as an overlapping parallel vector reduction operation. As
+compared to transport calculations where this overlap procedure was
+very effective, in the context of MCSA the Monte Carlo calculations
+are significantly shorter with $O(10,000)$ histories used in this
+work. These shorter calculations and more frequent overlapping tally
+vector reductions create an overhead that is not observed in the
+literature for transport calculations.
 
@@
 domain-to-domain communication of histories was occurring in the first
-place. Using this idea, we implemented a subdomain Neumann-Ulam method
-with MCSA such that MCSA now takes the form of a stochastic
+place. Motivated by this idea, we implemented a subdomain Neumann-Ulam
+method with MCSA such that MCSA now takes the form of a stochastic
 realization of an additive Schwarz method. Scaling studies using this
@@
 
-These operations were observed to substantially increase runtimes of
+These operations were observed to substantially increase run-times of
 MCSA and consume unacceptable amounts of memory, mostly due to the
@@
 applied to variance reduction techniques to improve MCSA convergence
-by modifying the weights and probabilities of the Monte Carlo game as
-an effective means of preconditioning.
+by modifying the weights and probabilities of the Monte Carlo game,
+effectively serving as a form of preconditioning by modifying the
+Neumann-Ulam decomposition and subsequently the physics operator.
 
@@
 All numerical methods for solving linear problems require quality
-preconditioning. Even a Krylov method may be effectively useful
+preconditioning. Even a Krylov method may be effectively useless
 without it. However, these methods typically converge eventually

### 2013-08-26 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding final edits to spn chapter

@@
 limits of MCSA, a hot zero-power $17 \times 17$ pin assembly will be
-used with varying energy group structure and $SP_N$ discretization in
-a criticality calculation. A cross section along the vertical axis
-showing homogenized fuel pin, control rod, and moderator materials and
-the associated grid is given by Figure~\ref{fig:problem3_radial_mat}
-while a cross section of the materials configuration along the
-horizontal axis is given by Figure~\ref{fig:problem3_axial_mat}. The
-control rods were removed for these problems leaving only moderator in
-their place. A detailed view of the assembly bottom is given in
+used with varying energy group structure and the $SP_N$ discretization
+in a criticality calculation. A cross section perpendicular to the
+vertical axis showing homogenized fuel pin, control rod, and moderator
+materials and the associated grid is given by
+Figure~\ref{fig:problem3_radial_mat} while a cross section of the
+materials configuration perpendicular to the horizontal axis is given
+by Figure~\ref{fig:problem3_axial_mat}. The control rods were removed
+for these problems leaving only moderator in their place. A detailed
+view of the assembly bottom is given in
 Figure~\ref{fig:problem3_end}. On the top and bottom of the assembly,
@@
 Figure~\ref{fig:problem3_radial_mat}, effectively giving a
-representation of one quarter of the assembly. For the spatial
-discretization, each fuel pin (gray regions in
-Figure~\ref{fig:problem3_radial_mat}) is resolved by a $2 \times 2$
-mesh with materials and cross sections homogenized over this region.
+representation of the entire assembly. For the spatial discretization,
+each fuel pin (gray regions in Figure~\ref{fig:problem3_radial_mat})
+is resolved by a $2 \times 2$ mesh with materials and cross sections
+homogenized over this region.
 \begin{figure}[t!]
@@
 solver with the collision estimator. For spectral radii less than
-0.97, all MCSA problems converged with 1 history per DOF (10,000 for
-this problem) with the number of iterations required to converge
-increasing as a function of spectral radius. Near a spectral radius of
-0.97, the number of histories required to converge MCSA in less than
-100 iterations takes a dramatic rise that exhibits neither exponential
-nor power law behavior. As the spectral radius approaches 1, the
-number of histories required becomes significant and effectively
-impractical to compute. Even with this simple diffusion problem, the
-behavior is consistent with that observed for the fuel assembly
-problem with $SP_1$ discretization. In that case, we estimated a
-spectral radius of $\approx 0.998$, larger than any of the spectral
-radii that could be computed within even 90 minutes of compute time
-for this simple two dimensional problem. For that problem, even
-1,000,000 histories ($\approx 50$ per DOF) was not enough to provide
-convergence. Even if single solve times of an hour can be tolerated,
-dozens of solves are typically required to compute the multiplication
-factor and flux spectrum within the $SP_N$ eigenvalue scheme for more
-difficult problems like the fuel assembly, making the method unusable.
+0.97, all MCSA problems converged with 1 history per DOF (10,000 total
+histories for this problem) with the number of iterations required to
+converge increasing as a function of spectral radius. Near a spectral
+radius of 0.97, the number of histories required to converge MCSA in
+less than 100 iterations takes a dramatic rise that exhibits neither
+exponential nor power law behavior. As the spectral radius approaches
+1, the number of histories required becomes significant and
+effectively impractical to compute. Even with this simple diffusion
+problem, the behavior is consistent with that observed for the fuel
+assembly problem with $SP_1$ discretization. In that case, we
+estimated a spectral radius of $\approx 0.998$, larger than any of the
+spectral radii that could be computed within even 90 minutes of
+compute time for this simple two dimensional problem. For that large
+of a spectral radius, even 1,000,000 histories ($\approx 50$ per DOF)
+was not enough to provide convergence. Even if single solve times of
+an hour can be tolerated, dozens of solves are typically required to
+compute the multiplication factor and flux spectrum within the $SP_N$
+eigenvalue scheme for more difficult problems like the fuel assembly,
+making the method unusable.
 \begin{figure}[t!]

### 2013-08-26 doc/dissertation/document/chapters/conclusion/conclusion.tex
whiitespace tweaks

@@
 
-With this work, a new foundation for Monte Carlo Synthetic Acceleration
-methods has been laid with new theory, algorithms, and numerical
-experiments that demonstrate both their potential to make great
-contributions to nuclear engineering calculations in the future and
-where the research must go in order to achieve this. By advancing
+With this work, a new foundation for Monte Carlo Synthetic
+Acceleration methods is developed with new theory, algorithms, and
+numerical experiments that demonstrate both their potential to make
+great contributions to nuclear engineering calculations in the future
+and where the research must go in order to achieve this. By advancing
 these techniques, we look forward to future work that will enable the

### 2013-08-28 doc/dissertation/presentation/defense.tex
apreliminary updates to defense talk

@@
 %%---------------------------------------------------------------------------%%
+\begin{frame}{Hardware-Based Motivation}
+
+  \begin{itemize}
+  \item Modern hardware is moving in two directions (Kogge,2011):
+    \begin{itemize}
+    \item Lightweight machines
+    \item Heterogeneous machines
+    \item Both characterized by low power and high concurrency
+    \end{itemize}
+    \medskip \medskip
+  \item Some issues:
+    \begin{itemize}
+    \item Higher potential for both soft and hard failures (DOE,2012)
+    \item Memory restrictions are expected with a continued decrease
+      in memory/FLOPS
+    \end{itemize}
+    \medskip \medskip
+  \item Potential resolution from Monte Carlo:
+    \begin{itemize}
+    \item Soft failures buried within the tally variance
+    \item Hard failures are high variance events
+    \item Memory savings over conventional methods
+    \end{itemize}
+  \end{itemize}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Physics-Based Motivation}
+
+  \begin{itemize}
+  \item Physics-driven development
+    \begin{itemize}
+    \item Identify the physics of interest and research MCSA
+      applicability
+    \item Offer solutions or a potential path forward to the observed
+      issues
+    \end{itemize}
+  \end{itemize}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
 \begin{frame}{Statement of Work}
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Hardware-Based Motivation}
-
-  \begin{itemize}
-  \item Modern hardware is moving in two directions (Kogge,2011):
-    \begin{itemize}
-    \item Lightweight machines
-    \item Heterogeneous machines
-    \item Both characterized by low power and high concurrency
-    \end{itemize}
-    \medskip \medskip
-  \item Some issues:
-    \begin{itemize}
-    \item Higher potential for both soft and hard failures (DOE,2012)
-    \item Memory restrictions are expected with a continued decrease
-      in memory/FLOPS
-    \end{itemize}
-    \medskip \medskip
-  \item Potential resolution from Monte Carlo:
-    \begin{itemize}
-    \item Soft failures buried within the tally variance
-    \item Hard failures are high variance events
-    \item Memory savings over conventional methods
-    \end{itemize}
-  \end{itemize}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
 \begin{frame}{Research Outline}
@@
     \item Application to neutron transport
-    \item Research is required to explore general MCSA development
-    \item Performance is of concern
+    \item Research is required to study MCSA preconditioning
+    \item Iterative performance is of concern
     \end{itemize}
@@
     \item Research is required to explore varying parallel strategies
-    \item Scalability is of concern
+    \item Parallel scalability is of concern
     \end{itemize}
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{$SP_N$ Equations}
-
-  \begin{multline}
-    \hat{\Omega} \cdot \vec{\nabla} \psi(\vec{r},\hat{\Omega},E) +
-    \sigma(\vec{r},E) \psi(\vec{r},\hat{\Omega},E) = \\ \iint
-    \sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}' \cdot \hat{\Omega})
-    \psi(\vec{r},\hat{\Omega}',E') d\Omega' dE' +
-    q(\vec{r},\hat{\Omega},E)\:,
-    \label{eq:general_transport}
-  \end{multline}
-
-  \smallskip
-
-  \begin{multline}
-    -\nabla \cdot \Bigg[\frac{n}{2n+1}\frac{1}{\Sigma_{n-1}} \nabla
-      \Big(\frac{n-1}{2n-1} \phi_{n-2} + \frac{n}{2n-1}\phi_n \Big) \\+
-      \frac{n+1}{2n+1}\frac{1}{\Sigma_{n+1}} \nabla
-      \Big(\frac{n+1}{2n+3}\phi_n + \frac{n+2}{2n+3}\phi_{n+2}\Big)
-      \Bigg] \\+ \Sigma_n \phi_n = q \delta_{n0}\ \ \ \ \ \ \ \ \ n =
-    0,2,4,\cdots,N\:,
-    \label{eq:spn_equations}
-  \end{multline}
-
-  \smallskip
-
-  \begin{equation}
-    -\nabla \cdot \mathbb{D}_n \nabla \mathbb{U}_n + \sum_{m=1}^4
-    \mathbb{A}_{nm} \mathbb{U}_m = \frac{1}{k} \sum_{m=1}^4
-    \mathbb{F}_{nm} \mathbb{U}_m\ \ \ \ \ \ \ n = 1,2,3,4\:.
-    \label{eq:spn_fission_matrix}
-  \end{equation}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}[fragile]{k-eigenvalue Solutions with MCSA}
-
-  \begin{algorithm}[H]
-    \caption{Power Iteration MCSA Scheme}
-    \label{alg:power_iteration}
-    \begin{algorithmic}
-      \STATE $k_0 =$ initial guess
-      \STATE $\mathbf{\Phi}_0 =$ initial guess
-      \STATE $n = 0$
-      \WHILE{$|\frac{k^n - k^{n-1}}{k^n}| < \epsilon$}
-      \STATE $\mathbf{M} \mathbf{\Phi}^{n+1} = \frac{1}{k^n} \mathbf{F} \mathbf{\Phi}^n$
-      \COMMENT{Solve for the new flux state with MCSA}
-      \STATE $k^{n+1} = k^n \frac{\int \mathbf{F} \mathbf{\Phi}^{n+1} d\mathbf{r}}{\int
-        \mathbf{F} \mathbf{\Phi}^n d\mathbf{r}}$
-      \STATE $n = n+1$
-      \ENDWHILE
-    \end{algorithmic}
-  \end{algorithm}
-
-  \begin{itemize}
-  \item Swap in MCSA as the solver at each eigenvalue operation
-    \medskip
-  \item Use Thyra/Stratimikos interfaces for plug and play with $SP_N$
-    \medskip
-  \item Current Exnihilo implementation gives the full operator
-  \end{itemize}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
 \begin{frame}{Fuel Assembly Criticality Calculations}
@@
           fuel assembly
-        \item $SP_N$ criticality calculation
+        \item Multigroup $SP_N$ discretization
         \item MCLS leveraged by the Exnihilo code base (ORNL)
-        \item Comparison to Trilinos Aztec Krylov solvers with ILUT
+        \item Emphasize algorithm development for iterative
+          performance
         \end{itemize}
@@
 %%---------------------------------------------------------------------------%%
+\begin{frame}{$SP_N$ Equations}
+
+  \begin{multline}
+    \hat{\Omega} \cdot \vec{\nabla} \psi(\vec{r},\hat{\Omega},E) +
+    \sigma(\vec{r},E) \psi(\vec{r},\hat{\Omega},E) = \\ \iint
+    \sigma_s(\vec{r},E' \rightarrow E,\hat{\Omega}' \cdot \hat{\Omega})
+    \psi(\vec{r},\hat{\Omega}',E') d\Omega' dE' +
+    q(\vec{r},\hat{\Omega},E)\:,
+    \label{eq:general_transport}
+  \end{multline}
+
+  \smallskip
+
+  \begin{multline}
+    -\nabla \cdot \Bigg[\frac{n}{2n+1}\frac{1}{\Sigma_{n-1}} \nabla
+      \Big(\frac{n-1}{2n-1} \phi_{n-2} + \frac{n}{2n-1}\phi_n \Big) \\+
+      \frac{n+1}{2n+1}\frac{1}{\Sigma_{n+1}} \nabla
+      \Big(\frac{n+1}{2n+3}\phi_n + \frac{n+2}{2n+3}\phi_{n+2}\Big)
+      \Bigg] \\+ \Sigma_n \phi_n = q \delta_{n0}\ \ \ \ \ \ \ \ \ n =
+    0,2,4,\cdots,N\:,
+    \label{eq:spn_equations}
+  \end{multline}
+
+  \smallskip
+
+  \begin{equation}
+    -\nabla \cdot \mathbb{D}_n \nabla \mathbb{U}_n + \sum_{m=1}^4
+    \mathbb{A}_{nm} \mathbb{U}_m = \frac{1}{k} \sum_{m=1}^4
+    \mathbb{F}_{nm} \mathbb{U}_m\ \ \ \ \ \ \ n = 1,2,3,4\:.
+    \label{eq:spn_fission_matrix}
+  \end{equation}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}[fragile]{k-eigenvalue Solutions with MCSA}
+
+  \begin{algorithm}[H]
+    \caption{Power Iteration MCSA Scheme}
+    \label{alg:power_iteration}
+    \begin{algorithmic}
+      \STATE $k_0 =$ initial guess
+      \STATE $\mathbf{\Phi}_0 =$ initial guess
+      \STATE $n = 0$
+      \WHILE{$|\frac{k^n - k^{n-1}}{k^n}| < \epsilon$}
+      \STATE $\mathbf{M} \mathbf{\Phi}^{n+1} = \frac{1}{k^n} \mathbf{F} \mathbf{\Phi}^n$
+      \COMMENT{Solve for the new flux state with MCSA}
+      \STATE $k^{n+1} = k^n \frac{\int \mathbf{F} \mathbf{\Phi}^{n+1} d\mathbf{r}}{\int
+        \mathbf{F} \mathbf{\Phi}^n d\mathbf{r}}$
+      \STATE $n = n+1$
+      \ENDWHILE
+    \end{algorithmic}
+  \end{algorithm}
+
+  \begin{itemize}
+  \item Inject MCSA as the solver at each eigenvalue operation
+    \medskip
+  \item Transport operator is static - one time cost for MCSA setup
+    \medskip
+  \item Current Exnihilo implementation gives the full operator
+  \end{itemize}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
 \begin{frame}{Jacobi Preconditioned Calculations}
@@
       \begin{itemize}
+      \item Rapidly divergent results
+        \medskip
       \item Light water moderator creates a lot of scattering and
@@
         \medskip
-      \item Intractable number of histories required for MCSA
-        convergence 
+      \item Convergence not achieved with 50 histories per DOF and 90
+        minutes compute time
       \end{itemize}
@@
   \end{beamerboxesrounded}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{ILUT Preconditioning}
+
+  \begin{columns}
+
+    \begin{column}{0.5\textwidth}
+
+      \begin{figure}[t!]
+        \begin{center}
+          \includegraphics[width=2.0in]{ilut_iterations.pdf}
+        \end{center}
+      \end{figure}
+
+      \begin{figure}[t!]
+        \begin{center}
+          \includegraphics[width=2.0in]{ilut_size.pdf}
+        \end{center}
+      \end{figure}
+
+    \end{column}
+
+    \begin{column}{0.5\textwidth}
+
+      \begin{itemize}
+      \item {\small Factor transport operator into upper and lower
+        triangular parts}
+        \begin{itemize}
+        \item {\small $\mathbf{R} = \mathbf{L} \mathbf{U} -
+          \mathbf{M}$}
+        \end{itemize}
+      \item {\small Control factorization content with level-of-fill
+        and drop tolerance}
+      \item {\small Use the reduced domain approximation to recover
+        sparsity}
+      \end{itemize}
+
+      \begin{figure}[t!]
+        \begin{center}
+          \includegraphics[width=2.0in]{rda_ilut_iterations.pdf}
+        \end{center}
+      \end{figure}
+
+    \end{column}
+
+  \end{columns}
 
@@
       \begin{itemize}
+        {\small
+        \item Comparison to Trilinos Aztec Krylov solvers with ILUT
+          \medskip
         \item MCLS generates the same k-eigenvalue and neutron flux in
           all groups and spatial locations
-          \bigskip
+          \medskip
         \item MCSA converged in fewer iterations than GMRES, more
           iterations than BiCGStab
-          \bigskip
+          \medskip
         \item Explicit preconditioning strategy destroys sparsity and
           elevates CPU times (Ifpack ILUT)
-          \bigskip
+          \medskip
         \item Spectral radius and memory limitations combine to
           prevent solutions at finer discretizations
+        }
       \end{itemize}
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Monte Carlo Synthetic Acceleration for Nonlinear Problems}
-
-  \begin{itemize}
-  \item Many physics problems of interest are nonlinear
+\begin{frame}{Navier-Stokes Benchmark Problems}
+
+  \begin{itemize}
+  \item Sequence of Navier-Stokes benchmarks for the nonlinear methods
+    \begin{itemize}
+    \item Thermal convection cavity problem (De Vahl Davis, 1983)
+    \item Lid driven cavity problem (Ghia et al., 1982)
+    \item Backward-Facing step problem (Gartling, 1990)
+    \end{itemize}
+    \medskip
+  \item Tuning benchmark parameters varies the strength of
+    nonlinearities
+    \medskip
+  \end{itemize}
+
+  \medskip  \medskip
+  \begin{columns}
+    \begin{column}{0.5\textwidth}
+      \[
+      \rho \ve{u} \cdot \nabla \ve{u} - \nabla \cdot \ve{T} - \rho
+      \ve{g} = \ve{0}
+      \]
+      \[
+      \nabla \cdot \ve{u} = 0
+      \]
+      \[
+      \rho C_p \ve{u} \cdot \nabla T + \nabla \cdot \ve{q} = 0
+      \]
+    \end{column}
+
+    \begin{column}{0.5\textwidth}
+      \[
+      \ve{T} = -P \ve{I} + \mu[\nabla \ve{u} + \nabla \ve{u}^T]
+      \]
+      \[
+      \ve{q} = - k \nabla T
+      \]
+    \end{column}
+  \end{columns}
+
+\end{frame}
+
+%%----------s-----------------------------------------------------------------%%
+\begin{frame}{MCSA for Nonlinear Problems}
+
+  \begin{itemize}
+  \item Need a nonlinear solution scheme for the Navier-Stokes equations
     \bigskip
@@
     \item Ideal for Monte Carlo
-    \item Potential scaling improvements
+    \item Potential parallel scaling and iterative performance
+      improvements
     \item Potential resiliency benefits
@@
 
-  \begin{itemize}
-  \item Sequence of Navier-Stokes benchmarks for the nonlinear methods
-    \begin{itemize}
-    \item Thermal convection cavity problem (De Vahl Davis, 1983)
-    \item Lid driven cavity problem (Ghia et al., 1982)
-    \item Backward-Facing step problem (Gartling, 1990)
-    \end{itemize}
-    \medskip
-  \item Tuning benchmark parameters varies the strength of
-    nonlinearities
-    \medskip
-  \item Newton-Krylov method leveraging Aztec GMRES used for
+  \begin{columns}
+    \begin{column}{0.5\textwidth}
+      \begin{figure}[htpb!]
+        \begin{center}
+          \includegraphics[width=2.0in]{drekar1.png}
+        \end{center}
+      \end{figure}
+    \end{column}
+
+    \begin{column}{0.5\textwidth}
+      \begin{figure}[htpb!]
+        \begin{center}
+          \includegraphics[width=2.0in]{drekar2.png}
+        \end{center}
+      \end{figure}
+    \end{column}
+  \end{columns}
+
+  \medskip
+
+  \begin{itemize}
+    {\small
+    \item Drekar is a production fluid code base using Newton methods
+      with FAD
+      \medskip
+    \item MCLS incorporated into Drekar nonlinear scheme to implement
+      FANM
+      \medskip
+    \item Newton-Krylov method with Aztec GMRES used for benchmark
       comparisons
-    \medskip
-  \item All problems preconditioned with algebraic multigrid (ML) and
-    leveraged some kind of globalization (e.g. backtracking)
-  \end{itemize}
-
-  \medskip  \medskip
-  \begin{columns}
-    \begin{column}{0.5\textwidth}
-      \[
-      \rho \ve{u} \cdot \nabla \ve{u} - \nabla \cdot \ve{T} - \rho
-      \ve{g} = \ve{0}
-      \]
-      \[
-      \nabla \cdot \ve{u} = 0
-      \]
-      \[
-      \rho C_p \ve{u} \cdot \nabla T + \nabla \cdot \ve{q} = 0
-      \]
-    \end{column}
-
-    \begin{column}{0.5\textwidth}
-      \[
-      \ve{T} = -P \ve{I} + \mu[\nabla \ve{u} + \nabla \ve{u}^T]
-      \]
-      \[
-      \ve{q} = - k \nabla T
-      \]
-    \end{column}
-  \end{columns}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{FANM Numerical Experiments}
-
-  \begin{columns}
-    \begin{column}{0.5\textwidth}
-      \begin{figure}[htpb!]
-        \begin{center}
-          \includegraphics[width=2.5in]{drekar1.png}
-        \end{center}
-      \end{figure}
-    \end{column}
-
-    \begin{column}{0.5\textwidth}
-      \begin{figure}[htpb!]
-        \begin{center}
-          \includegraphics[width=2.5in]{drekar2.png}
-        \end{center}
-      \end{figure}
-    \end{column}
-  \end{columns}
-
-  \bigskip
-
-  \begin{itemize}
-  \item Drekar is a production physics suite that leverages Newton
-    methods with FAD
-  \bigskip
-  \item MCLS incorporated into the Drekar nonlinear solver scheme
-    using Thyra/Stratimikos to implement FANM
+    \item All problems preconditioned with explicit scheme using
+      algebraic multigrid (ML) and leveraged some kind of
+      globalization (e.g. backtracking) 
+    }
   \end{itemize}
@@
 
-  \begin{figure}[htpb!]
-    \begin{center}
-      \includegraphics[width=4in]{titan_pure_strong.pdf}
-    \end{center}
-    \caption{\textbf{Pure domain decomposition.} \textit{Super-linear
-        speed-up from memory thrashing in base case. MCLS is an order
-        of magnitude slower arithmetically.}}
-  \end{figure}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Weak Scaling Results}
-
   \begin{columns}
-    \begin{column}{0.55\textwidth}
-
-      \begin{figure}[htpb!]
-        \begin{center}
-          \includegraphics[width=2.5in]{titan_weak_absolute.pdf}
-        \end{center}
-      \end{figure}
-
-      \begin{figure}[htpb!]
-        \begin{center}
-          \includegraphics[width=2.5in]{titan_weak_implementation.pdf}
-        \end{center}
-      \end{figure}
-
-    \end{column}
-
-    \begin{column}{0.45\textwidth}
-
-      \begin{itemize}
+
+    \begin{column}{0.5\textwidth}
+
+      \begin{figure}[htpb!]
+        \begin{center}
+          \includegraphics[width=2.4in]{titan_pure_strong_time.pdf}
+        \end{center}
+        \caption{CPU Time.}
+      \end{figure}
+
+    \end{column}
+
+    \begin{column}{0.5\textwidth}
+
+      \begin{figure}[htpb!]
+        \begin{center}
+          \includegraphics[width=2.4in]{titan_pure_strong.pdf}
+        \end{center}
+        \caption{Absolute Efficiency.}
+      \end{figure}
+
+    \end{column}
+
+  \end{columns}
+
+  \begin{itemize}
+  \item MCLS is an order of magnitude slower arithmetically.
+  \item Super-linear speed-up from memory thrashing in base case.
   \item CG demonstrates poor scaling due to the cheaper iteration
     sequence
-        \bigskip
-      \item Implementation efficiency is effectively the parallel
-        efficiency of a single iteration
-        \bigskip
+  \end{itemize}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Weak Scaling Results}
+
+  \begin{columns}
+    \begin{column}{0.5\textwidth}
+
+      \begin{figure}[htpb!]
+        \begin{center}
+          \includegraphics[width=2.4in]{titan_pure_weak_time.pdf}
+        \end{center}
+        \caption{CPU Time.}
+      \end{figure}
+
+    \end{column}
+
+    \begin{column}{0.5\textwidth}
+
+    \begin{figure}[htpb!]
+      \begin{center}
+        \includegraphics[width=2.4in]{titan_weak_absolute.pdf}
+      \end{center}
+      \caption{Absolute Efficiency.}
+    \end{figure}
+
+    \end{column}
+  \end{columns}
+
+  \begin{itemize}
   \item Improved implementation efficiency means improved
@@
     solution
-      \end{itemize}
-
-    \end{column}
-  \end{columns}
+    \bigskip
+  \item Implementation efficiency is effectively the parallel
+    efficiency of a single iteration
+  \end{itemize}
 
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Strong Scaling Results with Overlap}
-
-  \begin{itemize}
-  \item Overlap values selected based on average 'diffusion length'
-    of a history in the system of 2.6 discrete states
+\begin{frame}{Scaling Results with Overlap}
+
+  \begin{itemize}
+  \item Overlap values selected based on average 'diffusion length' of
+    a history in the system of 2.6 discrete states
     \bigskip
@@
       \medskip
-    \item MCSA was observed to converge in fewer iterations per eigenvalue
-      iteration than GMRES for the fuel assembly criticality problem and
-      more than Bi-CGStab using the same preconditioning
+    \item MCSA with the reduced domain approximation was observed to
+      converge in fewer iterations per eigenvalue iteration than GMRES
+      for the fuel assembly criticality problem and more than
+      Bi-CGStab using the same preconditioning
     \end{itemize}
@@
       Laboratories
+      \medskip
+    \item The FANM method has been verified to produce the same solutions
+      as a production Newton-Krylov method for three difficult benchmark
+      problems for the Navier-Stokes equations in different flow regimes
+      and geometries
       \medskip
@@
       problems at high Reynolds numbers
-      \medskip
-    \item More Monte Carlo histories at every FANM iteration can reduce
-      the number of linear and nonlinear iterations required to converge
-      the problem
     \end{itemize}
@@
     \item Random walk optimizations
-    \item Alias sampling implementations
     \item Multiple set reduction analysis
@@
     \begin{itemize}
-    \item Alternative left/right sampling schemes
     \item Variance reduction based strategy
@@
     \item Monte Carlo methods of the second degree
-    \item Stochastic projection methods
     \end{itemize}
@@
       \medskip
-    \item T.M. Evans, S.W. Mosher, S.R. Slattery, S.P. Hamilton,
-      \textbf{A Monte Carlo Synthetic-Acceleration Method for Solving
-        the Thermal Radiation Diffusion Equation}, \textit{Journal of
-        Computational Physics, Submitted.}
-      \medskip
     \item S.R. Slattery, T.M. Evans, P.P.H. Wilson, \textbf{A Spectral
@@
       2013.}
+      \medskip
+    \item T.M. Evans, S.W. Mosher, S.R. Slattery, S.P. Hamilton,
+      \textbf{A Monte Carlo Synthetic-Acceleration Method for Solving
+        the Thermal Radiation Diffusion Equation}, \textit{Journal of
+        Computational Physics, Submitted.}
     \end{enumerate}
@@
   \begin{itemize}
-  \item Paul
-  \item Tom
-  \item Roger
+  \item Paul Wilson
+  \item Tom Evans
+  \item Roger Pawlowski
   \item CASL, ORNL, Sandia, OLCF

### 2013-08-28 doc/dissertation/presentation/defense.tex
updating defense presentation

@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Outline}
-
-  \begin{itemize}
-  \item Monte Carlo Synthetic Acceleration Methods
-    \bigskip
-  \item Application to Neutron Transport
-    \bigskip
-  \item Application to Fluid Flow
-    \bigskip
-  \item Parallelization of MCSA
-    \bigskip
-  \item Summary
-  \end{itemize}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
 \begin{frame}{Acknowledgments}
@@
 %%---------------------------------------------------------------------------%%
+\begin{frame}{Outline}
+
+  \begin{itemize}
+  \item Monte Carlo Synthetic Acceleration Methods
+    \bigskip
+  \item Application to Neutron Transport
+    \bigskip
+  \item Application to Fluid Flow
+    \bigskip
+  \item Parallelization of MCSA
+    \bigskip
+  \item Summary
+  \end{itemize}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
 \begin{frame}{Hardware-Based Motivation}
@@
     \item Soft failures buried within the tally variance
-    \item Hard failures are high variance events
+    \item Hard failures mitigated by replication
     \item Memory savings over conventional methods
@@
   \begin{itemize}
+  \item New algorithms required to leverage new computational
+    resources
+    \begin{itemize}
+    \item Tremendous computational resources are required with
+      $O(\sn{1}{9})$ element meshes and $O(100,000)+$ cores used today
+      for neutronics and fluid problems (Evans,2010)(Pawlowski,2012)
+    \end{itemize}
+    \bigskip
+  \item Necessary to determine applicability to reactor physics and
+    potential performance improvements before moving forward with
+    other work
+    \bigskip
   \item Physics-driven development
     \begin{itemize}
-    \item Identify the physics of interest and research MCSA
-      applicability
-    \item Offer solutions or a potential path forward to the observed
+    \item Research applicability and potential improvements to
+      neutronics and fluid flow
+    \item Offer solutions or a potential path forward for the observed
       issues
+    \item Work to improve iterative and parallel performance where
+      possible
     \end{itemize}
@@
     leveraging Monte Carlo Synthetic Acceleration
+    \medskip
     \begin{itemize}
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Jacobi Preconditioned Calculations}
+\begin{frame}{Block Jacobi Preconditioned Calculations}
 
@@
   \begin{itemize}
-  \item Sequence of Navier-Stokes benchmarks for the nonlinear methods
+  \item Sequence of Navier-Stokes benchmarks
     \begin{itemize}
@@
 
-  \medskip  \medskip
   \begin{columns}
-    \begin{column}{0.5\textwidth}
+    \begin{column}{0.4\textwidth}
+      {\small
       \[
@@
       \]
-    \end{column}
-
-    \begin{column}{0.5\textwidth}
       \[
@@
       \]
+      }
+    \end{column}
+
+    \begin{column}{0.6\textwidth}
+
+      \begin{figure}[htpb!]
+        \begin{center}
+          \includegraphics[width=2.9in]{benchmark_solutions.png}
+        \end{center}
+      \end{figure}
+
     \end{column}
@@
 %%----------s-----------------------------------------------------------------%%
-\begin{frame}{MCSA for Nonlinear Problems}
+\begin{frame}{MCSA for the Navier-Stokes Equations}
 
@@
     \bigskip
-  \item Signficant research on Newton methods since the 1980's
+  \item Significant research on Newton methods since the 1980's
     \bigskip
@@
     \item Ideal for Monte Carlo
-    \item Potential parallel scaling and iterative performance
-      improvements
-    \item Potential resiliency benefits
+    \item Similar framework properties to matrix-free methods
     \end{itemize}
@@
     {\small
-    \item Drekar is a production fluid code base using Newton methods
+    \item Drekar is a production fluid base using Newton methods
       with FAD
@@
         \end{center}
-        \caption{CPU Time.}
+        \caption{Wall time.}
       \end{figure}
@@
         \end{center}
-        \caption{Absolute Efficiency.}
+        \caption{Absolute efficiency.}
       \end{figure}
@@
         \begin{center}
-          \includegraphics[width=2.4in]{titan_pure_weak_time.pdf}
-        \end{center}
-        \caption{CPU Time.}
-      \end{figure}
-
-    \end{column}
-
-    \begin{column}{0.5\textwidth}
-
-    \begin{figure}[htpb!]
-      \begin{center}
-        \includegraphics[width=2.4in]{titan_weak_absolute.pdf}
-      \end{center}
-      \caption{Absolute Efficiency.}
-    \end{figure}
+          \includegraphics[width=2.3in]{titan_pure_weak_time.pdf}
+        \end{center}
+%        \caption{Wall Time.}
+      \end{figure}
+
+      \begin{figure}[htpb!]
+        \begin{center}
+          \includegraphics[width=2.3in]{titan_weak_implementation.pdf}
+        \end{center}
+%        \caption{Implementation Efficiency.}
+      \end{figure}
+
+    \end{column}
+
+    \begin{column}{0.5\textwidth}
+
+      \begin{figure}[htpb!]
+        \begin{center}
+          \includegraphics[width=2.3in]{titan_weak_absolute.pdf}
+        \end{center}
+%        \caption{Absolute Efficiency.}
+      \end{figure}
+
+      {\small
+        \begin{itemize}
+        \item Krylov implementation efficiencies drop with
+          reduced iteration count
+        \item MCSA maintained constant number of iterations
+        \end{itemize}
+      }
 
@@
   \end{columns}
-
-  \begin{itemize}
-  \item Improved implementation efficiency means improved
-    iterative performance will potentially give a better time to
-    solution
-    \bigskip
-  \item Implementation efficiency is effectively the parallel
-    efficiency of a single iteration
-  \end{itemize}
 
@@
         \end{center}
-        \caption{\small{Absolute parallel efficiency relative to
-            16-core 1-set base case.}}
+        \caption{\small{Absolute efficiency relative to 16-core 1-set
+            base case.}}
       \end{figure}
@@
         \end{center}
-        \caption{\small{Wall time in seconds to solution for each
-            case.}}
+        \caption{\small{Wall time}}
       \end{figure}
@@
         \end{center}
-        \caption{Absolute parallel efficiency relative to 16-core
-          1-set base case.}
+        \caption{Absolute efficiency relative to 16-core 1-set base
+          case.}
       \end{figure}
@@
         \end{center}
-        \caption{Wall time in seconds to solution for each case.}
+        \caption{Wall time.}
       \end{figure}
@@
         \end{center}
-        \caption{Strong scaling absolute efficiency for pure domain
-          decomposition.}
+        \caption{Strong scaling absolute efficiency relative to full
+          transport base case.}
         \label{fig:titan_strong_subdomain}
@@
         \end{center}
-        \caption{Weak scaling absolute efficiency for pure domain
-          decomposition.}
+        \caption{Weak scaling absolute efficiency relative to full
+          transport base case.}
         \label{fig:titan_weak_subdomain}

### 2013-08-29 doc/sna2013/paper/slattery_evans_wilson.tex
adding sna paper

--- 2013-01-30 doc/sna2013/abstract/slattery_wilson_evans.tex
@@
 \usepackage{microtype} % improves typography for PDF
+
 \usepackage[breaklinks=true,colorlinks=true,linkcolor=black,citecolor=black]{hyperref}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{A Multiple-Set Overlapping-Domain Decomposed Monte Carlo
-  Synthetic Acceleration Method for Linear Systems}
+\title{Writing a SNA+MC Submission in \LaTeX}
 
@@
 \affil[1]{University of Wisconsin - Madison, Engineering Physics
-  Department, 1500 Engineering Drive, Madison, WI 53706}
-\affil[2]{Oak Ridge National Laboratory, One Bethel Valley Road, Oak
-  Ridge, TN 37830}
+  Department, 1500 Engineering Dr., Madison, WI 53716}
+\affil[2]{Oak Ridge National Laboratory, Reactor and Nuclear Systems
+  Division, 1 Bethel Valley Rd., Oak Ridge, TN 37831}
 
@@
   of Monte Carlo solvers for linear systems to accelerate the fixed
-  point method upon which the Neumann-Ulam solvers are based. To
+  point method on which the Neumann-Ulam solvers are based. To
   effectively parallelize MCSA methods requires the parallelization of
@@
   problem. The multiple-set overlapping-domain decomposition algorithm
-  will be presented in the submission along with parallel scaling data
-  for the resulting implementation.}
+  is presented along with parallel scaling data for the resulting
+  implementation using the Titan Cray XK7 machine at Oak Ridge
+  National Laboratory.}
 
@@
 \begin{document}
-
-%%---------------------------------------------------------------------------%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Introduction}
@@
 
-%%---------------------------------------------------------------------------%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Monte Carlo Synthetic Acceleration}
@@
 
-%%---------------------------------------------------------------------------%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Parallel Neumann-Ulam Algorithm}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Multiple-Set Overlapping-Domain Decomposition}
@@
 
-%%---------------------------------------------------------------------------%%
-\section{Submission Outline}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Parallel Scaling Studies}
 
-With our submission we will present a multiple-set overlapping-domain
-decomposed implementation of the Monte Carlo Synthetic Acceleration
-method. This presentation will include an outline of the Neumann-Ulam
-and MCSA methods followed by a presentation of the MSOD scheme as
-implemented for MCSA. This implementation will be tested and the
-results of parallel scaling studies presented that demonstrate the
-achieved parallel performance.
+\subsection{Overlap Scaling}
 
-%%---------------------------------------------------------------------------%%
+\subsection{Multiple Sets Scaling}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Conclusions}
+
+It was found the MCSA can indeed be effectively parallelized using the
+multiple-set overlapping-domain decomposition algorithm borrowed from
+the reactor physics community.  The new algorithm was tested in a wide
+variety of parallel scaling studies on the Titan Cray XK7 machine at
+the Oak Ridge Leadership Computing Facility. To test the algorithm at
+high levels of concurrency, up to 65,356 cores were used in strong
+scaling exercises and 131,072 cores used in weak scaling exercises
+using the neutron diffusion problem.
+
+Scaling studies showed that in the strong case overlap in small
+quantities on the order of the mean-free-path of a stochastic history
+in the simulation could boost parallel efficiencies by up to 10\% in
+isolated cases. However, it was found that this additional overlap was
+not effective in boosting weak scaling efficiencies. In general,
+overlap was not very effective due to the fact that the parallel
+communication saved during the Monte Carlo transport sequence is
+simply deferred until after transport is complete when it manifests
+itself as an overlapping parallel vector reduction operation. As
+compared to transport calculations where this overlap procedure was
+very effective, in the context of MCSA the Monte Carlo calculations
+are significantly shorter with $O(10,000)$ histories used in this
+work. These shorter calculations and more frequent overlapping tally
+vector reductions create an overhead that is not observed in the
+literature for transport calculations.
+
+Applying multiple sets in the parallel algorithm was found to not
+enhance the weak scaling of the problem as an additional parallel
+overhead is introduced when the calculations from the set are combined
+in superposition. For the strong scaling case, improvements were not
+noted until after the strong scaling wall was hit. At this point,
+multiple sets were observed to increase parallel efficiencies from
+38\% to 58\% at 16,384 cores. Perhaps more important here is the fact
+that although the parallel efficiency was reduced, multiple sets were
+observed to actually improve the time to solution. Unlike a
+traditional Krylov method that we might apply to solve a neutron
+transport problem, using MCSA means that we can actually make a
+physical copy of the problem on the machine and can combine separate
+Monte Carlo solutions for each copy through superposition. Time to
+solution is then improved because fewer histories were run in each
+copy and therefore each MCSA iteration is faster or more global
+histories are computed and fewer MCSA iterations are required to
+converge.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Acknowledgments}
+
+This work was performed under appointment to the Nuclear Regulatory
+Commission Fellowship program at the University of Wisconsin - Madison
+Department of Engineering Physics.
+
+This research used resources of the Oak Ridge Leadership Computing
+Facility at the Oak Ridge National Laboratory, which is supported by
+the Office of Science of the U.S. Department of Energy under Contract
+No. DE-AC05-00OR22725.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \bibliographystyle{ans}

### 2013-09-03 doc/dissertation/presentation/defense.tex
updating solution evolution slide

@@
   \begin{itemize}
+  \item \textbf{Introduction}
+    \bigskip
   \item Monte Carlo Synthetic Acceleration Methods
@@
       issues
-    \item Work to improve iterative and parallel performance where
-      possible
+    \item Work to improve iterative and parallel performance
     \end{itemize}
@@
   \begin{itemize}
-  \item Development of a linear scheme for discrete systems leveraging
-    Monte Carlo Synthetic Acceleration
+  \item Development of a linear scheme for the $SP_N$ equations
+    leveraging Monte Carlo Synthetic Acceleration
     \medskip
@@
     \bigskip
-  \item Development of a nonlinear scheme for discrete systems
-    leveraging Monte Carlo Synthetic Acceleration
+  \item Development of a nonlinear scheme for the Navier-Stokes
+    equations leveraging Monte Carlo Synthetic Acceleration
     \medskip
@@
   \begin{itemize}
+  \item Introduction
+    \bigskip
   \item \textbf{Monte Carlo Synthetic Acceleration Methods}
@@
   \end{itemize}
+
 \end{frame}
@@
     \begin{center}
-      \includegraphics[width=4in]{adjoint_1.png}
-    \end{center}
-    \caption{\textbf{Adjoint solution to Poisson Equation.}
+      \includegraphics<1>[width=4in]{adjoint_1.png}
+      \includegraphics<2>[width=4in]{adjoint_10.png}
+      \includegraphics<3>[width=4in]{adjoint_100.png}
+      \includegraphics<4>[width=4in]{adjoint_1000.png}
+      \includegraphics<5>[width=4in]{adjoint_10000.png}
+      \includegraphics<6>[width=4in]{adjoint_100000.png}
+      \includegraphics<7>[width=4in]{adjoint_1000000.png}
+      \includegraphics<8>[width=4in]{adjoint_10000000.png}
+    \end{center}
+    \caption{
+      \only<1>{\textbf{Adjoint solution to Poisson Equation.}
         \textit{\sn{1}{0} total histories, 0.286 seconds CPU time.} }
-  \end{figure}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Evolution of a Solution}
-
-  \begin{figure}[h!]
-    \begin{center}
-      \includegraphics[width=4in]{adjoint_10.png}
-    \end{center}
-    \caption{\textbf{Adjoint solution to Poisson Equation.}
+      \only<2>{\textbf{Adjoint solution to Poisson Equation.}
         \textit{\sn{1}{1} total histories, 0.278 seconds CPU time.} }
-  \end{figure}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Evolution of a Solution}
-
-  \begin{figure}[h!]
-    \begin{center}
-      \includegraphics[width=4in]{adjoint_100.png}
-    \end{center}
-    \caption{\textbf{Adjoint solution to Poisson Equation.}
+      \only<3>{\textbf{Adjoint solution to Poisson Equation.}
         \textit{\sn{1}{2} total histories, 0.275 seconds CPU time.} }
-  \end{figure}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Evolution of a Solution}
-
-  \begin{figure}[h!]
-    \begin{center}
-      \includegraphics[width=4in]{adjoint_1000.png}
-    \end{center}
-    \caption{\textbf{Adjoint solution to Poisson Equation.}
+      \only<4>{\textbf{Adjoint solution to Poisson Equation.}
         \textit{\sn{1}{3} total histories, 0.291 seconds CPU time.} }
-  \end{figure}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Evolution of a Solution}
-
-  \begin{figure}[h!]
-    \begin{center}
-      \includegraphics[width=4in]{adjoint_10000.png}
-    \end{center}
-    \caption{\textbf{Adjoint solution to Poisson Equation.}
+      \only<5>{\textbf{Adjoint solution to Poisson Equation.}
         \textit{\sn{1}{4} total histories, 0.428 seconds CPU time.} }
-  \end{figure}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Evolution of a Solution}
-
-  \begin{figure}[h!]
-    \begin{center}
-      \includegraphics[width=4in]{adjoint_100000.png}
-    \end{center}
-    \caption{\textbf{Adjoint solution to Poisson Equation.}
+      \only<6>{\textbf{Adjoint solution to Poisson Equation.}
         \textit{\sn{1}{5} total histories, 1.76 seconds CPU time.} }
-  \end{figure}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Evolution of a Solution}
-
-  \begin{figure}[h!]
-    \begin{center}
-      \includegraphics[width=4in]{adjoint_1000000.png}
-    \end{center}
-    \caption{\textbf{Adjoint solution to Poisson Equation.}
+      \only<7>{\textbf{Adjoint solution to Poisson Equation.}
         \textit{\sn{1}{6} total histories, 15.1 seconds CPU time.} }
-  \end{figure}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{Evolution of a Solution}
-
-  \begin{figure}[h!]
-    \begin{center}
-      \includegraphics[width=4in]{adjoint_10000000.png}
-    \end{center}
-    \caption{\textbf{Adjoint solution to Poisson Equation.}
+      \only<8>{\textbf{Adjoint solution to Poisson Equation.}
         \textit{\sn{1}{7} total histories, 149 seconds CPU time.} } 
+    }
   \end{figure}
@@
   \begin{itemize}
+  \item Introduction
+    \bigskip
   \item Monte Carlo Synthetic Acceleration Methods

### 2013-09-03 doc/dissertation/presentation/defense.tex
restructuring nonlinear results

@@
   \begin{itemize}
-  \item Inject MCSA as the solver at each eigenvalue operation
+  \item Inject MCSA as the solver at each eigenvalue iteration
     \medskip
@@
         \begin{center}
-          \includegraphics[width=1.9in]{solver_time.pdf}
+          \includegraphics[width=2.25in]{solver_time.pdf}
         \end{center}
@@
           \medskip
-        \item MCLS generates the same k-eigenvalue and neutron flux in
-          all groups and spatial locations
+        \item MCLS verified for the k-eigenvalue and neutron flux in
+          all groups
           \medskip
@@
     nonlinearities
-    \medskip
   \end{itemize}
@@
     \begin{itemize}
-    \item Robust implementations
+    \item Simple implementation
     \item No operator required
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{FANM Nonlinear Iteration Results}
-
+\begin{frame}{Thermal Convection Cavity Results}
+
+  \begin{columns}
+    \begin{column}{0.6\textwidth}
+
+      \small{
         \begin{table}[h!]
@@
               \hline
-        Convection, Ra=\sn{1}{3} & 5 & 5 & 5\\
-        Convection, Ra=\sn{1}{4} & 7 & 7 & 7\\
-        Convection, Ra=\sn{1}{5} & 9 & 10 & 9\\
-        Convection, Ra=\sn{1}{6} & 11 & 11 & 11\\
-        \hline
-        Lid Driven, Re=100 & 6 & 6 & 7\\
-        Lid Driven, Re=300 & 9 & 9 & 9\\
-        Lid Driven, Re=500 & 11 & 11 & 11\\
-        Lid Driven, Re=700 & 14 & 10 & 12\\
-        \hline
-        Backward Step, Re=200 & 10 & 9 & 10\\
-        Backward Step, Re=300 & 15 & 14 & 15\\
-        Backward Step, Re=400 & 10 & 10 & 10\\
-        Backward Step, Re=500 & 19 & 20 & 21\\
+              Ra=\sn{1}{3} & 5 & 5 & 5\\
+              Ra=\sn{1}{4} & 7 & 7 & 7\\
+              Ra=\sn{1}{5} & 9 & 10 & 9\\
+              Ra=\sn{1}{6} & 11 & 11 & 11\\
               %%
@@
           \end{center}
-    \caption{\textbf{Navier-Stokes benchmark comparison for nonlinear
-        iterations.} \textit{Over all benchmarks, FANM performed better
-        in terms of nonlinear iterations for 1 more case than the
-        Newton-Krylov method.}}
-    \label{tab:benchmark_nonlinear_comparison}
-
+          \caption{Nonlinear iterations.}
         \end{table}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{FANM Linear Solver Iteration Results}
 
@@
               \hline
-        Convection, Ra=\sn{1}{3} & 32 & 18 & 38\\
-        Convection, Ra=\sn{1}{4} & 23 & 17 & 34\\
-        Convection, Ra=\sn{1}{5} & 25 & 20 & 34\\
-        Convection, Ra=\sn{1}{6} & 39 & 25 & 48\\
-        \hline
-        Lid Driven, Re=100 & 27 & 42 & 151\\
-        Lid Driven, Re=300 & 35 & 52 & 133\\
-        Lid Driven, Re=500 & 41 & 56 & 154\\
-        Lid Driven, Re=700 & 21 & 14 & 32\\
-        \hline
-        Backward Step, Re=200 & 24 & 13 & 21\\
-        Backward Step, Re=300 & 23 & 17 & 21\\
-        Backward Step, Re=400 & 18 & 12 & 14\\
-        Backward Step, Re=500 & 30 & 52 & 98\\
+              Ra=\sn{1}{3} & 32 & 18 & 38\\
+              Ra=\sn{1}{4} & 23 & 17 & 34\\
+              Ra=\sn{1}{5} & 25 & 20 & 34\\
+              Ra=\sn{1}{6} & 39 & 25 & 48\\
               %%
@@
           \end{center}
-    \caption{\textbf{Navier-Stokes benchmark comparison for total linear
-        solver iterations.} \textit{Over all benchmarks, FANM performed
-        better in terms of linear solver iterations for twice as many
-        cases as the Newton-Krylov method.}}
-    \label{tab:benchmark_linear_comparison}
+          \caption{Total linear solver iterations.}
         \end{table}
-
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}{FANM CPU Time Results}
-
+      }
+
+    \end{column}
+
+    \begin{column}{0.4\textwidth}
+
+      \begin{itemize}
+      \item Weaker preconditioning at \sn{1}{-5} case
+      \item Forcing term method adapts to nonlinear convergence
+      \item Constant forcing term for \sn{1}{-6} case
+      \end{itemize}
+
+      \bigskip
+
+      \tiny{
         \begin{table}[h!]
@@
               \hline
-      Convection, Ra=\sn{1}{3} & 338 \\
-      Convection, Ra=\sn{1}{4} & 336 \\
-      Convection, Ra=\sn{1}{5} & 346 \\
-      Convection, Ra=\sn{1}{6} & 465 \\
+              Ra=\sn{1}{3} & 338 \\
+              Ra=\sn{1}{4} & 336 \\
+              Ra=\sn{1}{5} & 346 \\
+              Ra=\sn{1}{6} & 465 \\
+              \hline\hline
+            \end{tabular}
+          \end{center}
+          \caption{Newton-Krylov speedup over FANM.}
+        \end{table}
+      }
+
+    \end{column}
+  \end{columns}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Lid Driven Cavity Results}
+
+  \begin{columns}
+    \begin{column}{0.6\textwidth}
+
+      \small{
+        \begin{table}[h!]
+          \begin{center}
+            \begin{tabular}{lccc}\hline\hline
+              \multicolumn{1}{l}{Benchmark}& 
+              \multicolumn{1}{c}{NK}&
+              \multicolumn{1}{c}{FANM}&
+              \multicolumn{1}{c}{NR}\\
               \hline
-      Lid Driven, Re=100 & 299 \\
-      Lid Driven, Re=300 & 322 \\
-      Lid Driven, Re=500 & 288 \\
-      Lid Driven, Re=700 & 488 \\
+              Re=100 & 6 & 6 & 7\\
+              Re=300 & 9 & 9 & 9\\
+              Re=500 & 11 & 11 & 11\\
+              Re=700 & 14 & 10 & 12\\
+              %%
+              \hline\hline
+            \end{tabular}`
+          \end{center}
+          \caption{Nonlinear iterations.}
+        \end{table}
+
+        \begin{table}[h!]
+          \begin{center}
+            \begin{tabular}{lccc}\hline\hline
+              \multicolumn{1}{l}{Benchmark}& 
+              \multicolumn{1}{c}{GMRES}&
+              \multicolumn{1}{c}{MCSA}&
+              \multicolumn{1}{c}{Richardson}\\
               \hline
-      Backward Step, Re=200 & 400 \\
-      Backward Step, Re=300 & 593 \\
-      Backward Step, Re=400 & 825 \\
-      Backward Step, Re=500 & 1057 \\
+              Re=100 & 27 & 42 & 151\\
+              Re=300 & 35 & 52 & 133\\
+              Re=500 & 41 & 56 & 154\\
+              Re=700 & 21 & 14 & 32\\
               %%
@@
           \end{center}
-  \caption{\textbf{Newton-Krylov speedup over FANM.}  \textit{For all
-      benchmarks the explicit MCSA preconditioning strategy caused
-      significantly larger CPU times for FANM when compared to the
-      Newton-Krylov solutions.}}
-  \label{tab:benchmark_time_comparison}
+          \caption{Total linear solver iterations.}
         \end{table}
+      }
+
+    \end{column}
+
+    \begin{column}{0.4\textwidth}
+
+      \begin{itemize}
+      \item Forcing term selection can significantly modify nonlinear
+        convergence
+        \medskip
+      \item $\eta_k = \gamma \Big(
+        \frac{||\ve{F}(\ve{u}_k)||}{||\ve{F}(\ve{u}_{k-1})||}
+        \Big)^{\alpha}$
+      \end{itemize}
+
+      \bigskip
+
+      \tiny{
+        \begin{table}[h!]
+          \begin{center}
+            \begin{tabular}{lc}\hline\hline
+              \multicolumn{1}{l}{Benchmark}& 
+              \multicolumn{1}{c}{NK Speedup}\\
+              \hline
+              Re=100 & 299 \\
+              Re=300 & 322 \\
+              Re=500 & 288 \\
+              Re=700 & 488 \\
+              %%
+              \hline\hline
+            \end{tabular}
+          \end{center}
+          \caption{Newton-Krylov speedup over FANM.}
+        \end{table}
+      }
+
+    \end{column}
+  \end{columns}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Backward Facing Step Results}
+
+  \begin{columns}
+    \begin{column}{0.6\textwidth}
+
+      \small{
+        \begin{table}[h!]
+          \begin{center}
+            \begin{tabular}{lccc}\hline\hline
+              \multicolumn{1}{l}{Benchmark}& 
+              \multicolumn{1}{c}{NK}&
+              \multicolumn{1}{c}{FANM}&
+              \multicolumn{1}{c}{NR}\\
+              \hline
+              Re=200 & 10 & 9 & 10\\
+              Re=300 & 15 & 14 & 15\\
+              Re=400 & 10 & 10 & 10\\
+              Re=500 & 19 & 20 & 21\\
+              %%
+              \hline\hline
+            \end{tabular}`
+          \end{center}
+          \caption{Nonlinear iterations.}
+        \end{table}
+
+        \begin{table}[h!]
+          \begin{center}
+            \begin{tabular}{lccc}\hline\hline
+              \multicolumn{1}{l}{Benchmark}& 
+              \multicolumn{1}{c}{GMRES}&
+              \multicolumn{1}{c}{MCSA}&
+              \multicolumn{1}{c}{Richardson}\\
+              \hline
+              Re=200 & 24 & 13 & 21\\
+              Re=300 & 23 & 17 & 21\\
+              Re=400 & 18 & 12 & 14\\
+              Re=500 & 30 & 52 & 98\\
+              %%
+              \hline\hline
+            \end{tabular}
+          \end{center}
+          \caption{Total linear solver iterations.}
+        \end{table}
+      }
+
+    \end{column}
+
+    \begin{column}{0.4\textwidth}
+
+      \begin{itemize}
+      \item Significantly more ill-conditioned than cavity problems
+      \item Forcing terms are not the culprit
+      \item Multigrid is doing significantly more work
+      \end{itemize}
+
+      \bigskip
+
+      \tiny{
+        \begin{table}[h!]
+          \begin{center}
+            \begin{tabular}{lc}\hline\hline
+              \multicolumn{1}{l}{Benchmark}& 
+              \multicolumn{1}{c}{NK Speedup}\\
+              \hline
+              Re=200 & 400 \\
+              Re=300 & 593 \\
+              Re=400 & 825 \\
+              Re=500 & 1057 \\
+              %%
+              \hline\hline
+            \end{tabular}
+          \end{center}
+          \caption{Newton-Krylov speedup over FANM.}
+        \end{table}
+      }
+
+    \end{column}
+  \end{columns}
+
+\end{frame}
+
+%%---------------------------------------------------------------------------%%
+\begin{frame}{Iterative Performance Summary}
+
+  {\tiny
+
+    \begin{columns}
+      \begin{column}{0.5\textwidth}
+
+        \begin{table}[h!]
+          \begin{center}
+            \begin{tabular}{lcc}\hline\hline
+              \multicolumn{1}{l}{Benchmark}& 
+              \multicolumn{1}{c}{Newton-Krylov}&
+              \multicolumn{1}{c}{FANM}\\
+              \hline
+              Convection, Ra=\sn{1}{3} & $\times$ & $\times$ \\
+              Convection, Ra=\sn{1}{4} & $\times$ & $\times$ \\
+              Convection, Ra=\sn{1}{5} & $\times$ & \\
+              Convection, Ra=\sn{1}{6} & $\times$ & $\times$ \\
+              Lid Driven, Re=100 & $\times$ & $\times$ \\
+              Lid Driven, Re=300 & $\times$ & $\times$ \\
+              Lid Driven, Re=500 & $\times$ & $\times$ \\
+              Lid Driven, Re=700 & & $\times$ \\
+              Backward Step, Re=200 & & $\times$ \\
+              Backward Step, Re=300 & & $\times$ \\
+              Backward Step, Re=400 & $\times$ & $\times$ \\
+              Backward Step, Re=500 & $\times$ & \\
+              %%
+              \hline\hline
+            \end{tabular}
+          \end{center}
+          \caption{Navier-Stokes benchmark comparison for nonlinear
+            iterations.}
+        \end{table}
+
+      \end{column}
+
+      \begin{column}{0.5\textwidth}
+
+        \begin{table}[h!]
+          \begin{center}
+            \begin{tabular}{lcc}\hline\hline
+              \multicolumn{1}{l}{Benchmark}& 
+              \multicolumn{1}{c}{Newton-Krylov}&
+              \multicolumn{1}{c}{FANM}\\
+              \hline
+              Convection, Ra=\sn{1}{3} & & $\times$ \\
+              Convection, Ra=\sn{1}{4} & & $\times$ \\
+              Convection, Ra=\sn{1}{5} & & $\times$ \\
+              Convection, Ra=\sn{1}{6} & & $\times$ \\
+              Lid Driven, Re=100 & $\times$ & \\
+              Lid Driven, Re=300 & $\times$ & \\
+              Lid Driven, Re=500 & $\times$ & \\
+              Lid Driven, Re=700 & & $\times$ \\
+              Backward Step, Re=200 & & $\times$ \\
+              Backward Step, Re=300 & & $\times$ \\
+              Backward Step, Re=400 & & $\times$ \\
+              Backward Step, Re=500 & $\times$ & \\
+              %%
+              \hline\hline
+            \end{tabular}
+          \end{center}
+          \caption{Navier-Stokes benchmark comparison for total
+            linear solver iterations.}
+        \end{table}
+
+      \end{column}
+    \end{columns}
+  }
+
+  {\small
+    \begin{itemize}
+    \item Over all benchmarks, FANM performed better in terms of nonlinear
+      iterations for 1 more case than the Newton-Krylov method
+      \medskip
+    \item Over all benchmarks, FANM performed better in terms of linear
+      solver iterations for twice as many cases as the Newton-Krylov
+      method
+    \end{itemize}
+  }
 

### 2013-09-03 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
whiitespace tweaks
fixing typo in nonlinear chapter

@@
 operations. Table~\ref{tab:convection_linear_iter_comparison} gives
-the total number of GMRES, FANM, and Richardson iterations required to
+the total number of GMRES, MCSA, and Richardson iterations required to
 converge the Newton method at each Rayleigh number used. For all
@@
 
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/nonlinear_problem/driven_re500_iters.pdf}
+  \end{center}
+  \caption{\textbf{Lid driven cavity linear solver iterations as a
+      function of Newton iteration for the 500 Reynolds number case.}
+    \textit{More MCSA iterations were required to converge the linear
+      models at later Newton iterations due to smaller forcing
+      terms.}}
+  \label{fig:driven_re500_iters}
+\end{figure}
+
+\begin{figure}[t!]
+  \begin{center}
+    \includegraphics[width=6in]{chapters/nonlinear_problem/driven_re500_forcing.pdf}
+  \end{center}
+  \caption{\textbf{Lid driven cavity forcing terms as a function of
+      Newton iteration for the 500 Reynolds number case.} \textit{Once
+      the region of fast quadratic convergence is reached, MCSA
+      forcing terms were much smaller.}}
+  \label{fig:driven_re500_forcing}
+\end{figure}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{cccc}\hline\hline
+      \multicolumn{1}{c}{Reynolds Number}& 
+      \multicolumn{1}{c}{GMRES Iterations}&
+      \multicolumn{1}{c}{MCSA Iterations}&
+      \multicolumn{1}{c}{Richardson Iterations}\\
+      \hline
+      100 & 27 & 42 & 151\\
+      300 & 35 & 52 & 133\\
+      500 & 41 & 56 & 154\\
+      700 & 21 & 14 & 32\\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Lid driven cavity total sum of linear iterations
+      for all nonlinear iterations.} \textit{The results presented
+      here were obtained from the benchmark verification
+      calculations. MCSA always accelerated the Richardson
+      iteration.}}
+  \label{tab:driven_linear_iter_comparison}
+\end{table}
+
 As an additional means of analysis for the linear solver iteration
@@
 iterations to meet the convergence criteria.
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/nonlinear_problem/driven_re500_iters.pdf}
-  \end{center}
-  \caption{\textbf{Lid driven cavity linear solver iterations as a
-      function of Newton iteration for the 500 Reynolds number case.}
-    \textit{More MCSA iterations were required to converge the linear
-      models at later Newton iterations due to smaller forcing
-      terms.}}
-  \label{fig:driven_re500_iters}
-\end{figure}
-
-\begin{figure}[t!]
-  \begin{center}
-    \includegraphics[width=6in]{chapters/nonlinear_problem/driven_re500_forcing.pdf}
-  \end{center}
-  \caption{\textbf{Lid driven cavity forcing terms as a function of
-      Newton iteration for the 500 Reynolds number case.} \textit{Once
-      the region of fast quadratic convergence is reached, MCSA
-      forcing terms were much smaller.}}
-  \label{fig:driven_re500_forcing}
-\end{figure}
-
-\begin{table}[h!]
-  \begin{center}
-    \begin{tabular}{cccc}\hline\hline
-      \multicolumn{1}{c}{Reynolds Number}& 
-      \multicolumn{1}{c}{GMRES Iterations}&
-      \multicolumn{1}{c}{MCSA Iterations}&
-      \multicolumn{1}{c}{Richardson Iterations}\\
-      \hline
-      100 & 27 & 42 & 151\\
-      300 & 35 & 52 & 133\\
-      500 & 41 & 56 & 154\\
-      700 & 21 & 14 & 32\\
-      %%
-      \hline\hline
-    \end{tabular}
-  \end{center}
-  \caption{\textbf{Lid driven cavity total sum of linear iterations
-      for all nonlinear iterations.} \textit{The results presented
-      here were obtained from the benchmark verification
-      calculations. MCSA always accelerated the Richardson
-      iteration.}}
-  \label{tab:driven_linear_iter_comparison}
-\end{table}
 

### 2013-09-04 doc/dissertation/presentation/defense.tex
adding some more corrections to slides

@@
     \end{center}
-    \caption{\textbf{Spectral radius results for the block Jacobi
+    \caption{Spectral radius results for the block Jacobi
       preconditioned iteration matrix with 10 energy groups and full
-        downscatter for sample problem.}}
+      downscatter for sample problem.}
     \label{tab:group10dsbj}
@@
         \end{center}
-        \caption{ Average number of linear solver iterations per
+        \caption{Average number of linear solver iterations per
           eigenvalue iteration.}
@@
 
-      \begin{itemize}
-      \item Weaker preconditioning at \sn{1}{-5} case
-      \item Forcing term method adapts to nonlinear convergence
-      \item Constant forcing term for \sn{1}{-6} case
-      \end{itemize}
+      \small{
+        \begin{itemize}
+        \item Weaker preconditioning at Ra=\sn{1}{-5} case
+          \medskip
+        \item $\eta_k = \gamma \Big(
+          \frac{||\ve{F}(\ve{u}_k)||}{||\ve{F}(\ve{u}_{k-1})||}
+          \Big)^{\alpha}$
+          \medskip
+        \item Constant forcing term for Ra=\sn{1}{-6} case
+          \medskip
+        \item RDA applied in all cases
+        \end{itemize}
+      }
 
@@
 
+      \small{
         \begin{itemize}
@@
           convergence
-        \medskip
-      \item $\eta_k = \gamma \Big(
-        \frac{||\ve{F}(\ve{u}_k)||}{||\ve{F}(\ve{u}_{k-1})||}
-        \Big)^{\alpha}$
-      \end{itemize}
-
-      \bigskip
+        \item Newton-Krylov had larger forcing terms for Re=100,300,500
+        \item At Re=700, MCSA histories doubled and RDA fill increased
+          from 200 to 300
+        \end{itemize}
+      }
+
+      \smallskip
 
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Iterative Performance Summary}
+\begin{frame}{FANM Iterative Performance Summary}
 
@@
         \end{center}
-        \caption{\small \sl Domain decomposition example illustrating
+        \caption{\small Domain decomposition example illustrating
           how domain-to-domain transport creates communication costs.}
@@
         \end{center}
-        \caption{\small \sl Overlapping domain example illustrating
-          how domain overlap can reduce communication costs.}
+        \caption{\small Overlapping domain example illustrating how
+          domain overlap can reduce communication costs.}
       \end{figure}
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Multiple-Set Overlapping-Domain Decomposition}
-
-  \begin{columns}
-
-    \begin{column}{0.5\textwidth}
-
-      \begin{figure}[htpb!]
-        \begin{center}
-          \scalebox{0.25}{ \input{msod_construction.pdftex_t} }
-        \end{center}
-        \caption{\small \sl MSOD construction.}
-      \end{figure}
-
-    \end{column}
-
-    \begin{column}{0.5\textwidth}
-
-      \begin{figure}[htpb!]
-        \begin{center}
-          \scalebox{0.25}{ \input{msod_tally.pdftex_t} }
-        \end{center}
-        \caption{\small \sl MSOD tally reduction.}
-      \end{figure}
-
-    \end{column}
-
-  \end{columns}
+\begin{frame}[fragile]{Multiple-Set Overlapping-Domain Decomposition}
+
+  \begin{columns}
+
+    \begin{column}{0.5\textwidth}
+
+      \begin{figure}[htpb!]
+        \begin{center}
+          \scalebox{0.2}{ \input{msod_construction.pdftex_t} }
+        \end{center}
+        \caption{\small MSOD construction.}
+      \end{figure}
+
+    \end{column}
+
+    \begin{column}{0.5\textwidth}
+
+      \begin{figure}[htpb!]
+        \begin{center}
+          \scalebox{0.2}{ \input{msod_tally.pdftex_t} }
+        \end{center}
+        \caption{\small MSOD tally reduction.}
+      \end{figure}
+
+    \end{column}
+
+  \end{columns}
+
+    \begin{algorithm}[H]
+      \begin{algorithmic}[1]
+        \tiny
+        \STATE build MSOD domain
+        \STATE build Monte Carlo source
+        \STATE build Monte Carlo tally
+        \STATE perform parallel Neumann-Ulam transport in each set
+        \STATE combine set tallies
+        \STATE combine block tallies
+      \end{algorithmic}
+      \caption{MSOD Transport Sequence}
+    \end{algorithm}
 

### 2013-09-04 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
fixing some more typos in the nonlinear chapter of the document

@@
 \end{equation}
-where $\eta^k \in [0,1)$ is defined as the \textit{forcing term} at
-  nonlinear iteration $k$, the right-hand side is the nonlinear
-  residual at nonlinear iteration $k$, and the left-hand side is the
-  residual for the linear model at nonlinear iteration
-  $k$. Eq~(\ref{eq:inexact_newton_forcing}) then states that the
-  residual generated by the linear solver is bound by the nonlinear
-  residual and how tightly it is bound is defined by the forcing
-  term. This is useful in that we can vary how tightly coupled the
-  convergence of the linear iterations used to generate the Newton
-  correction is to the nonlinear iteration by relaxing or tightening
-  the convergence properties on the linear iterative method. As a
-  result, strategies for determining the forcing term can vary
-  depending on the problem type and can greatly affect the convergence
-  of the method or even prohibit convergence
-  \cite{eisenstat_choosing_1996}. In addition, \textit{globalization
-    methods} may be used to modify the Newton correction in a more
-  desirable direction such that nonlinear convergence properties can
-  be improved when the initial guess for $\ve{u}$ is poor
-  \cite{pawlowski_globalization_2006}.
+where $\eta^k \in [0,1)$ is defined as the \textit{forcing term}, the
+  right-hand side is the nonlinear residual, and the left-hand side is
+  the residual for the linear model with all defined at nonlinear
+  iteration $k$. Eq~(\ref{eq:inexact_newton_forcing}) then states that
+  the residual generated by the linear solver is bound by the
+  nonlinear residual and how tightly it is bound is defined by the
+  forcing term. As a result, strategies for determining the forcing
+  term can vary depending on the problem type and can greatly affect
+  the convergence of the method or even prohibit convergence
+  \cite{eisenstat_choosing_1996}. In this work we will either use a
+  constant forcing term or use method 2 from Shadid's work
+  \cite{shadid_inexact_1997}:
+\begin{equation}
+  \eta_k = \gamma \Big(
+  \frac{||\ve{F}(\ve{u}_k)||}{||\ve{F}(\ve{u}_{k-1})||} \Big)^{\alpha}\:,
+  \label{eq:shadid_forcing}
+\end{equation}
+where $\gamma$ and $\alpha$ are free parameters with typical values of
+0.9 for $\gamma$ and 2.0 for $\alpha$. In addition,
+\textit{globalization methods} may be used to modify the Newton
+correction in a more desirable direction such that nonlinear
+convergence properties can be improved when the initial guess for
+$\ve{u}$ is poor \cite{pawlowski_globalization_2006}.
 
@@
 nonlinear problems, the Jacobian operator is generally non-symmetric
-and therefore either Krylov methods with long recurrence relations
-that can handle non-symmetric systems must be considered or the Newton
-correction system must be preconditioned such that the operator is
-symmetric and short recurrence relation methods can be potentially be
-used.
+and therefore either Krylov methodsthat can handle non-symmetric
+systems must be considered or the Newton correction system must be
+preconditioned such that the operator is symmetric.
 
@@
 areas. First, if iterative or parallel performance of MCSA can be
-demonstrated to better than that of GMRES or similar subspace methods
-for asymmetric systems, FANM may outperform a Newton-Krylov
+demonstrated to be better than that of GMRES or similar subspace
+methods for asymmetric systems, FANM may outperform a Newton-Krylov
 method. Second, for situations where memory is a constraint,
@@
 
-\subsubsection{Backward-Facing Step Problem}
+\subsection{Backward-Facing Step Problem}
 \label{subsubsec:backward_facing_step}
@@
 computation while the forcing term at all Rayleigh numbers but
-\sn{1}{6} was computed at each Newton step using method 2 from
-Shadid's work. In this forcing term method, nonlinear residuals from
-the two previous Newton steps are used to inform the choice of the
-forcing term for the next step. At a Rayleigh number of \sn{1}{6}, a
+\sn{1}{6} was computed at each Newton step using
+Eq~\ref{eq:shadid_forcing}. At a Rayleigh number of \sn{1}{6}, a
 constant forcing term of \sn{1}{-4} was use to give better convergence
@@
 \cite{pawlowski_globalization_2006} while the forcing term was again
-computed using method 2 from Shadid's work \cite{shadid_inexact_1997}
-at all Reynolds numbers.
+computed using Eq~\ref{eq:shadid_forcing} at all Reynolds numbers.
 
@@
 a polynomial line search method was used as a means of globalization
-while the forcing term was again computed using method 2 from Shadid's
-work \cite{shadid_inexact_1997}.
+while the forcing term was again computed using Eq~\ref{eq:shadid_forcing}.
 
@@
 inner linear solves is also of interest. Here, we seek to minimize the
-minimum number of total linear solver iterations over the course of
-all nonlinear iterations in order to limit matrix-vector multiplies
-and other costly
+number of total linear solver iterations over the course of all
+nonlinear iterations in order to limit matrix-vector multiplies and
+other costly
 operations. Table~\ref{tab:convection_linear_iter_comparison} gives
@@
 the linear iterations required by the Newton-Krylov scheme with the
-linear solver per nonlinear iteration given for this case in
-Figure~\ref{fig:ra1e6_linear_iters}. Even the \sn{1}{5} Rayleigh
-number case that required more Newton iterations required 5 fewer MCSA
-iterations to converge than GMRES iterations. In all cases, MCSA
-improved on the iterative performance of using only the Richardson
-iteration through the residual Monte Carlo acceleration. If the
-performance of the MCSA implementation used here were to be
+linear solver iterations at every nonlinear iteration given for this
+case in Figure~\ref{fig:ra1e6_linear_iters}. Even the \sn{1}{5}
+Rayleigh number case that required more Newton iterations required 5
+fewer MCSA iterations to converge than GMRES iterations. In all cases,
+MCSA improved on the iterative performance of using only the
+Richardson iteration through the residual Monte Carlo acceleration. If
+the performance of the MCSA implementation used here were to be
 potentially optimized in the future, this improved iterative
@@
 Reynolds numbers of 200, 300, and 500, FANM also reduced the number of
-Newton steps need when compared to the Newton-Richardson scheme.
+Newton steps needed when compared to the Newton-Richardson scheme.
 
@@
 dramatic with FANM converging the nonlinear residual in fewer linear
-solver iterations for twice as many cases as the Newton-Krylov method
+solver iterations for twice as many cases as the Newton-Krylov method,
 keeping in mind that for all cases both GMRES and MCSA were

### 2013-09-04 doc/dissertation/presentation/defense.tex
small updates to presentation

@@
     \bigskip
-  \item Necessary to determine applicability to reactor physics and
-    potential performance improvements before moving forward with
-    other work
     \bigskip
@@
     {\small
-    \item Drekar is a production fluid base using Newton methods
-      with FAD
+    \item Drekar is a production CFD code using Newton methods with
+      FAD
       \medskip
@@
           convergence
-        \item Newton-Krylov had larger forcing terms for Re=100,300,500
+        \item Newton-Krylov always had larger forcing terms
         \item At Re=700, MCSA histories doubled and RDA fill increased

### 2013-09-04 doc/dissertation/presentation/defense.tex
adding corrections from practice talk

@@
       \STATE $n = 0$
-      \WHILE{$|\frac{k^n - k^{n-1}}{k^n}| < \epsilon$}
+      \WHILE{$|\frac{k^n - k^{n-1}}{k^n}| > \epsilon$}
       \STATE $\mathbf{M} \mathbf{\Phi}^{n+1} = \frac{1}{k^n} \mathbf{F} \mathbf{\Phi}^n$
@@
       comparisons
-    \item All problems preconditioned with explicit scheme using
-      algebraic multigrid (ML) and leveraged some kind of
-      globalization (e.g. backtracking) 
-    }
+    \item All methods and problems preconditioned with explicit scheme
+      using algebraic multigrid (ML) and leveraged some kind of
+      globalization (e.g. backtracking) }
   \end{itemize}
@@
         \begin{itemize}
-        \item Weaker preconditioning at Ra=\sn{1}{-5} case
+        \item Weaker preconditioning at Ra=\sn{1}{5} case
           \medskip
@@
           \medskip
-        \item Constant forcing term for Ra=\sn{1}{-6} case
+        \item Constant forcing term for Ra=\sn{1}{6} case
           \medskip
@@
       Conference on Supercomputing in Nuclear Applications and Monte
-      Carl (SNA+MC 2010), 2010.}}
+      Carlo (SNA+MC 2010), 2010.}}
 \end{frame}
@@
   \item Overlap eliminates communication in the Monte Carlo sequence
-    but simply differs it to an overlapping tally vector reduction
+    but simply defers it to an overlapping tally vector reduction
   \end{itemize}

### 2013-09-04 doc/dissertation/presentation/defense.tex
reducing msod explanation to one slide

@@
       using algebraic multigrid (ML) and leveraged some kind of
-      globalization (e.g. backtracking) }
+      globalization (e.g. backtracking) with RDA
+    }
   \end{itemize}
@@
           \medskip
-        \item $\eta_k = \gamma \Big(
-          \frac{||\ve{F}(\ve{u}_k)||}{||\ve{F}(\ve{u}_{k-1})||}
-          \Big)^{\alpha}$
-          \medskip
         \item Constant forcing term for Ra=\sn{1}{6} case
-          \medskip
-        \item RDA applied in all cases
         \end{itemize}
@@
         \begin{itemize}
-        \item Forcing term selection can significantly modify nonlinear
-          convergence
+        \item $\eta_k = \gamma \Big(
+          \frac{||\ve{F}(\ve{u}_k)||}{||\ve{F}(\ve{u}_{k-1})||}
+          \Big)^{\alpha}$
         \item Newton-Krylov always had larger forcing terms
@@
 %%---------------------------------------------------------------------------%%
-\begin{frame}{Multiple-Set Overlapping-Domain Decomposition}
+\begin{frame}[fragile]{Multiple-Set Overlapping-Domain Decomposition}
 
@@
 
-    \begin{column}{0.5\textwidth}
-      \begin{figure}[htpb!]
-        \begin{center}
-          \scalebox{0.75}{ \input{msod_example.pdftex_t} }
-        \end{center}
-        \caption{\small Overlapping domain example illustrating how
-          domain overlap can reduce communication costs.}
-      \end{figure}
-    \end{column}
-
-    \begin{column}{0.5\textwidth}
-      \begin{itemize}
-      \item Developed by Wagner and colleagues in 2010
-        \medskip
+    \begin{column}{0.52\textwidth}
+
+      \begin{figure}[t!]
+        \begin{center}
+          \scalebox{0.275}{ \input{msod_construction.pdftex_t} }
+        \end{center}
+        \caption{\small MSOD construction.}
+      \end{figure}
+
+    \end{column}
+
+    \begin{column}{0.48\textwidth}
+
+      \begin{figure}[t!]
+        \begin{center}
+          \scalebox{0.2}{ \input{msod_tally.pdftex_t} }
+        \end{center}
+        \caption{\small MSOD tally reduction.}
+      \end{figure}
+
+      \begin{itemize}
+        {\small
+        \item Multiple sets replicate the domain
+        \item Domains overlap within a set
         \item Each set contains the full domain
-        \medskip
-      \item Multiple sets replicate the domain
-        \medskip
-      \item Domains overlap within a set
-        \medskip
-      \item Redundancy for resiliency (and useful work)
-      \end{itemize}
+        }
+      \end{itemize}
+
     \end{column}
@@
       Carlo (SNA+MC 2010), 2010.}}
-\end{frame}
-
-%%---------------------------------------------------------------------------%%
-\begin{frame}[fragile]{Multiple-Set Overlapping-Domain Decomposition}
-
-  \begin{columns}
-
-    \begin{column}{0.5\textwidth}
-
-      \begin{figure}[htpb!]
-        \begin{center}
-          \scalebox{0.2}{ \input{msod_construction.pdftex_t} }
-        \end{center}
-        \caption{\small MSOD construction.}
-      \end{figure}
-
-    \end{column}
-
-    \begin{column}{0.5\textwidth}
-
-      \begin{figure}[htpb!]
-        \begin{center}
-          \scalebox{0.2}{ \input{msod_tally.pdftex_t} }
-        \end{center}
-        \caption{\small MSOD tally reduction.}
-      \end{figure}
-
-    \end{column}
-
-  \end{columns}
-
-    \begin{algorithm}[H]
-      \begin{algorithmic}[1]
-        \tiny
-        \STATE build MSOD domain
-        \STATE build Monte Carlo source
-        \STATE build Monte Carlo tally
-        \STATE perform parallel Neumann-Ulam transport in each set
-        \STATE combine set tallies
-        \STATE combine block tallies
-      \end{algorithmic}
-      \caption{MSOD Transport Sequence}
-    \end{algorithm}
 

### 2013-09-04 doc/dissertation/presentation/defense.tex
updating weak scaling slide
updating slides build

@@
     leveraging Monte Carlo Synthetic Acceleration
-    \medskip
     \begin{itemize}
@@
     equations leveraging Monte Carlo Synthetic Acceleration
-    \medskip
     \begin{itemize}
@@
   \item Parallelization of Monte Carlo Synthetic Acceleration
-    \medskip
     \begin{itemize}
@@
     \item All methods and problems preconditioned with explicit scheme
-      using algebraic multigrid (ML) and leveraged some kind of
-      globalization (e.g. backtracking) with RDA
-    }
+      using algebraic multigrid (ML) with RDA and leveraged some kind
+      of globalization (e.g. backtracking) }
   \end{itemize}
@@
         \end{center}
-%        \caption{Wall Time.}
-      \end{figure}
+        \caption{Wall Time.}
+      \end{figure}
+
+      %% \begin{figure}[htpb!]
+      %%   \begin{center}
+      %%     \includegraphics[width=2.3in]{titan_weak_implementation.pdf}
+      %%   \end{center}
+      %%   \caption{Implementation Efficiency.}
+      %% \end{figure}
+
+    \end{column}
+
+    \begin{column}{0.5\textwidth}
 
@@
         \begin{center}
-          \includegraphics[width=2.3in]{titan_weak_implementation.pdf}
-        \end{center}
-%        \caption{Implementation Efficiency.}
-      \end{figure}
-
-    \end{column}
-
-    \begin{column}{0.5\textwidth}
-
-      \begin{figure}[htpb!]
-        \begin{center}
           \includegraphics[width=2.3in]{titan_weak_absolute.pdf}
         \end{center}
-%        \caption{Absolute Efficiency.}
-      \end{figure}
+        \caption{Absolute Efficiency.}
+      \end{figure}
+
+    \end{column}
+  \end{columns}
 
@@
     \begin{itemize}
-        \item Krylov implementation efficiencies drop with
-          reduced iteration count
+    \item Spectral radius maintained by growing the global boundary
     \item MCSA maintained constant number of iterations
-        \end{itemize}
-      }
-
-    \end{column}
-  \end{columns}
+    \item Krylov iteration count reduced as a function of problem size
+    \end{itemize}
+  }
 

### 2013-09-05 doc/sna2013/paper/slattery_evans_wilson.tex
adding images from parallel thesis chapter to sna dir
adding background in paper for nu methods and mcsa

@@
 \usepackage{microtype} % improves typography for PDF
+\usepackage{tabls}
+\usepackage{afterpage}
+\usepackage{amssymb}
+\usepackage{amsfonts}
+\usepackage{amsmath}
+\usepackage[mathcal]{euscript}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{Writing a SNA+MC Submission in \LaTeX}
+\title{A Multiple-Set Overlapping-Domain Decomposed Monte Carlo
+  Synthetic Acceleration Method for Linear Systems}
 
@@
 
-For some time, the particle transport community has been utilizing
-Monte Carlo methods for the solution of transport problems
+For some time, the particle transport community has utilized Monte
+Carlo methods for the solution of transport problems
 \cite{lewis_computational_1993}. The partial differential equation
 (PDE) community has focused on various deterministic methods for
-solutions to linear problems \cite{saad_iterative_2003}. In between
-these two areas are a not widely known group of Monte Carlo methods
-for solving sparse linear systems \cite{forsythe_matrix_1950,
-  hammersley_monte_1964, halton_sequential_1962,
-  halton_sequential_1994}. In recent years, these methods have been
-further developed for radiation transport problems in the form of
-Monte Carlo Synthetic-Acceleration (MCSA) \cite{evans_monte_2009,
-  evans_monte_2012} but have yet to be applied to more general sparse
-linear systems commonly generated by the computational physics
-community. Compared to other methods in this regime, MCSA offers three
-attractive qualities; (1) the linear problem operator need not be
-symmetric or positive-definite, thereby reducing preconditioning
-complexity and widening applicability, (2) the stochastic nature of
-the solution method provides a natural solution to the issue of
-resiliency, and (3) is amenable to parallelization using modern
-methods developed by the transport community
-\cite{wagner_hybrid_2010}. The development of MCSA as a general linear
-solver and the development of a parallel MCSA method will be new and
-unique features of this work, providing a framework with which other
-issues such as resiliency may be addressed in the future.
+solutions to linear problems \cite{saad_iterative_2003,
+  kelley_iterative_1995}. In between these two areas are a not widely
+known group of Monte Carlo methods for solving sparse linear systems
+\cite{forsythe_matrix_1950, hammersley_monte_1964,
+  halton_sequential_1962, halton_sequential_1994}. In recent years,
+these methods have been further developed for radiation transport
+problems in the form of Monte Carlo Synthetic Acceleration (MCSA)
+\cite{evans_monte_2009, evans_monte_2012} but have yet to be applied
+to more general sparse linear systems commonly generated by the
+computational physics community. Compared to other methods in this
+regime, MCSA offers three attractive qualities; (1) the physics
+operator need not be symmetric or positive-definite, (2) the
+stochastic nature of the solution method provides a natural solution
+to some aspects of the issue of resiliency, and (3) is amenable to
+parallelization using modern methods developed by the transport
+community \cite{wagner_hybrid_2010}. The development a parallel MCSA
+algorithm using techniques is the primary contribution of this work.
 
-As leadership class machines move towards the exascale, new algorithms
+As leadership class machines move beyond the petascale, new algorithms
 must be developed that leverage their strengths and adapt to their
-shortcomings. Looking towards machines operating at hundreds of
-petaflops peak performance and beyond, trends toward reduced energy
-consumption will require incredibly high levels of concurrency to
-achieve the desired computation rates \cite{kogge_using_2011}. The end
-result of these hardware changes is that the larger numbers of
-low-powered processors will be prone to both soft and hard failures
-with resilient solver technologies are required to overcome these
-events. With linear solvers based on Monte Carlo techniques, such
-issues are alleviated by statistical arguments. In the case of soft
-failures, isolated floating point errors in Monte Carlo simulation are
-absorbed within tally statistics while completely losing hardware
-during a hard failure is manifested as a high variance event where
-some portion of the Monte Carlo histories are lost.
+shortcomings. Basic research is required now to advance methods in
+time for these new machines to become operational. As machines begin
+to operate at hundreds of petaflops peak performance and beyond,
+trends toward reduced energy consumption will require incredibly high
+levels of concurrency to achieve the desired computation rates
+\cite{kogge_using_2011}. The end result of these hardware changes is
+that the larger number of low-powered processors will be prone to both
+soft failures such as bit errors in floating point operations and hard
+failures where the data owned by that processor cannot be
+recovered. Because these failures are predicted to be common,
+resilient solver technologies are required to overcome these events at
+the application level. With linear solvers based on Monte Carlo
+techniques, such issues are potentially alleviated by statistical
+arguments. In the case of soft failures, isolated floating point
+errors in the Monte Carlo simulation are absorbed within tally
+statistics. Completely losing memory during a hard failure is treated
+as a high variance event where some portion of the Monte Carlo
+histories and subsequently the solution are lost with other histories
+maintained by replicating the problem and combining the results
+through superposition.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Monte Carlo Synthetic Acceleration}
-
-Using the ideas of Halton's residual method
-\cite{halton_sequential_1994}, Evans and Mosher recently developed a
-Monte Carlo solution method that was not prohibited severely by the
-quality of the initial guess for the system \cite{evans_monte_2009}
-and later applied it more rigorously as a solution mechanism for the
-radiation diffusion equation \cite{evans_monte_2012}. With their new
-methods, they achieved identical numerical results as and marginally
-better performance than conventional Krylov solvers. Their approach
-was instead to use residual Monte Carlo as a synthetic acceleration
-for a stationary method. Using Monte Carlo in this way achieves the
-same effect as Halton's method, decoupling its convergence rate from
-the overall convergence rate of the method. Here, the approximate
-Monte Carlo solution is not driven to a particular convergence as it
-merely supplies a correction for the initial guess generated by
-Richardson's iteration. Rather, only a set number of histories are
-required using the adjoint method to generate the correction.
-
+\section{Monte Carlo Methods}
 Monte Carlo Synthetic Acceleration (MCSA) methods rely the
@@
 \cite{spanier_monte_1969} present additional detail on these solvers.
+In this section we briefly present the Neumann-Ulam Monte Carlo method
+for solving linear systems and then present the Monte Carlo Synthetic
+Acceleration method.
+
+\subsection{Neumann-Ulam Method}
+We seek solutions of the general linear problem in the following form:
+\begin{equation}
+  \mathbf{A} \mathbf{x} = \mathbf{b}\:,
+  \label{eq:linear_problem}
+\end{equation}
+where $\mathbf{A} \in \mathbb{R}^{N \times N}$ is a matrix operator
+such that $\mathbf{A} : \mathbb{R}^{N} \rightarrow \mathbb{R}^{N}$,
+$\mathbf{x} \in \mathbb{R}^N$ is the solution vector, and $\mathbf{b}
+\in \mathbb{R}^N$ is the forcing term.  For a given linear operator
+$\mathbf{A}$, we can use diagonal splitting to define the
+\textit{iteration matrix}, $\mathbf{H}$:
+\begin{equation}
+  \mathbf{H} = \mathbf{I} - \mathbf{A}\:,
+  \label{eq:linear_mc_iteration_matrix}
+\end{equation}
+such that we are solving the system:
+\begin{equation}
+  \mathbf{x} = \mathbf{H} \mathbf{x} + \mathbf{b}\:.
+  \label{eq:richardson_split}
+\end{equation}
+We can then form an alternative representation for $\mathbf{A}^{-1}$
+by generating the \textit{Neumann series} solution:
+\begin{equation}
+  \mathbf{A}^{-1}\mathbf{b} = \sum_{k=0}^{\infty}
+  \mathbf{H}^k\mathbf{b} = \mathbf{x}\:,
+  \label{eq:neumann_solution}
+\end{equation}
+which will converge if the spectral radius of $\mathbf{H}$ is less
+than 1. If we expand the summation with a succession of matrix-vector
+multiply operations, we arrive at an alternative perspective of this
+summation by considering the $i^{th}$ component of the solution
+vector:
+\begin{equation}
+  x_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
+  \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}\:,
+  \label{eq:expanded_neumann_solution}
+\end{equation}
+which can interpreted as a series of transitions between states,
+\begin{equation}
+ \nu = i \rightarrow i_1 \rightarrow \cdots \rightarrow i_{k-1}
+ \rightarrow i_{k}\:,
+  \label{eq:mc_walk_permutation}
+\end{equation}
+in $\mathbf{H}$ where $\nu$ is interpreted as a particular sequence
+permutation. We can generate these sequences of transitions through
+Monte Carlo random walks by assigning them both a probability and
+weight. Using the adjoint form \cite{spanier_monte_1969},
+probabilities are column-scaled:
+\begin{equation}
+  p_{ij} = \frac{|h_{ji}|}{\sum_j |h_{ji}|}\:,
+  \label{eq:adjoint_probability}
+\end{equation}
+such that we expect to select a new state, $j$, from the current state
+in the random walk, $i$, by sampling column-wise (or row-wise if an
+adjoint probability matrix is formed). The transition weight is then
+defined as:
+\begin{equation}
+  w_{ij} = \frac{h_{ji}}{p_{ij}}\:.
+  \label{eq:adjoint_weight}
+\end{equation}
+The initial state $i_0$ of the random walk is determined by sampling
+the source vector $\mathbf{b}$ with probabilities:
+\begin{equation}
+  P_{(i_0=i)}(\nu) = \frac{|b_i|}{||\mathbf{b}||_1}\:,
+  \label{eq:adjoint_source_probability}
+\end{equation}
+with a random walk starting weight of:
+\begin{equation}
+  W_0 = ||\mathbf{b}||_1 \frac{b_i}{|b_i|}\:.
+  \label{eq:adjoint_starting_weight}
+\end{equation}
+The contribution to the solution in state $j$ from a particular random
+walk permutation of $k$ events is then the \textit{collision
+  estimator}:
+\begin{equation}
+  X_{j}(\nu) = \sum_{m=0}^k W_{m} \delta_{i_m,j}\:,
+  \label{eq:adjoint_permutation_contribution}
+\end{equation}
+where the Kronecker delta indicates that the tally contributes only in
+the current state, $i_m$, of the random walk and 
+\begin{equation}
+  W_{m} = W_0 w_{i_0,i_1} w_{i_1,i_2} \cdots w_{i_{m-1},i_m}\:.
+  \label{eq:adjoint_permutation_weight}
+\end{equation}
+Finally, the expectation value using all permutations is:
+\begin{equation}
+  E\{X_j\} = \sum_{\nu} P_{\nu} X_{j}(\nu)\:,
+  \label{eq:adjoint_expectation_value}
+\end{equation}
+with
+\begin{equation}
+  P_{\nu} = P_{(i_0=i)} p_{i,i_1} p_{i_1,i_2} \cdots p_{i_{k-1},i_k}\:.
+  \label{eq:adjoint_permutation_probability}
+\end{equation}
+
+\subsection{Monte Carlo Synthetic Acceleration}
+Using the ideas of Halton's residual method
+\cite{halton_sequential_1994}, Evans and Mosher recently developed a
+Monte Carlo solution method that was not prohibited severely by the
+quality of the initial guess for the system \cite{evans_monte_2009}
+and later applied it more rigorously as a solution mechanism for the
+radiation diffusion equation \cite{evans_monte_2012}. With their new
+methods, they achieved identical numerical results as and marginally
+better performance than conventional Krylov solvers. Their approach
+was instead to use residual Monte Carlo as a synthetic acceleration
+for a stationary method.
+
+The \textit{Fixed-Point Monte Carlo Synthetic Acceleration} (MCSA)
+method is defined as:
+\begin{subequations}
+  \begin{gather}
+    \mathbf{r}^{k} = \mathbf{b} - \mathbf{A}\mathbf{x}^{k}\:,\\
+    \mathbf{x}^{k+1/2} = \mathbf{x}^k + \mathbf{r}^k\:,\\
+    \mathbf{r}^{k+1/2} = \mathbf{b} - \mathbf{A}\mathbf{x}^{k+1/2}\:,\\
+    \label{eq:mcsa_mc_solve}
+    \mathbf{A}\delta\mathbf{x}^{k+1/2} = \mathbf{r}^{k+1/2}\:,\\
+    \mathbf{x}^{k+1} = \mathbf{x}^{k+1/2} + \delta \mathbf{x}^{k+1/2}\:,
+  \end{gather}
+  \label{eq:mcsa}
+\end{subequations}
+where $\mathbf{r}$ is the residual vector, $\delta\mathbf{x}$ is the
+Monte Carlo correction vector, and $k$ indicates the iteration
+index. Here, a Neumann-Ulam Monte Carlo method is used to generate the
+solution correction in Eq~\ref{eq:mcsa_mc_solve} from the residual and
+Richardson's iteration in the first step has been rewritten as a
+residual correction. Using Monte Carlo in this way achieves the same
+effect as Halton's method, decoupling its convergence rate from the
+overall convergence rate of the method. Here, the approximate Monte
+Carlo solution is not driven to a particular convergence as it merely
+supplies a correction for the initial guess generated by Richardson's
+iteration. Rather, only a set number of random walk permuations are
+required using the Neumann-Ulam method to generate the correction. In
+addition to the Monte Carlo solver parameters dictating the number of
+histories and weight cutoff, the outer MCSA iterations also have the
+following stopping criteria:
+\begin{equation}
+  ||\mathbf{r}||_\infty < \epsilon \ ||\mathbf{b}||_\infty\:,
+  \label{eq:mcsa_stopping_criteria}
+\end{equation}
+where $\epsilon$ is a user-defined parameter. We refer the reader to
+\cite{evans_monte_2012} and \cite{evans_monte_2009} for a more
+detailed explanation and analysis of MCSA and the Neumann-Ulam
+methods.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Parallel Neumann-Ulam Algorithm}
+\section{Parallel MCSA Algorithm}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Multiple-Set Overlapping-Domain Decomposition}
-
+\subsection{Parallel Neumann-Ulam Algorithm}
 In the literature, current implementations of parallel Neumann-Ulam
@@
 
+\subsection{MSOD Algorithm}
 At the 2010 SNA+MC meeting, Wagner and colleagues presented the

### 2013-09-05 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
whiitespace tweaks
updating dissertation doc build

@@
 simple model problem.
-
-\clearpage
 
@@
 given $P$ and $N$. When that problem is now solved using 2 sets we
-have $S=2$, $N$, and $2P$. For any number of sets we then have a
-global problem size of $N$ while the number of processors is $SP$
-where $P$ is the number of processors in the single set
-case. Therefore, $N/P$ is no longer fixed in the weak scaling study
-but rather $N/(SP)$ is fixed. As a result, we modify the weak scaling
-absolute efficiency relationship to account for this fact:
+have $S=2$, $N$, and $2P_s$ where $P_s$ is the number of processors in
+a single set. For any number of sets we then have a global problem
+size of $N$ while the number of processors is $SP_s$. Therefore, $N/P$
+is no longer fixed in the weak scaling study but rather $N/(SP_s)$ is
+fixed. As a result, we modify the weak scaling absolute efficiency
+relationship to account for this fact:
 \begin{equation}

### 2013-09-05 doc/sna2013/paper/slattery_evans_wilson.tex
fixing madison zip code in sna paper
updating dissertation doc build

@@
 \usepackage[mathcal]{euscript}
+\usepackage{tmath}
 
@@
 \affil[1]{University of Wisconsin - Madison, Engineering Physics
-  Department, 1500 Engineering Dr., Madison, WI 53716}
+  Department, 1500 Engineering Dr., Madison, WI 53706}
 \affil[2]{Oak Ridge National Laboratory, Reactor and Nuclear Systems
@@
 \section{Monte Carlo Methods}
-Monte Carlo Synthetic Acceleration (MCSA) methods rely the
-Neumann-Ulam class of linear solver to accelerate the stationary
+Monte Carlo Synthetic Acceleration (MCSA) methods rely on the
+Neumann-Ulam class of linear solvers to accelerate a stationary
 method. These Monte Carlo methods work by sampling a distribution with
@@
 operator. Neumann and Ulam's work was first published in 1950 by
-Forsythe and Leibler \cite{forsythe_matrix_1950} with a time frame for
-the creation of the method some years earlier than its publication. In
-the following decade, the books of Hammersley and Handscomb
+Forsythe and Leibler \cite{forsythe_matrix_1950}. In the following
+decade, the books of Hammersley and Handscomb
 \cite{hammersley_monte_1964} and Spanier and Gelbard
@@
 such that we expect to select a new state, $j$, from the current state
-in the random walk, $i$, by sampling column-wise (or row-wise if an
-adjoint probability matrix is formed). The transition weight is then
-defined as:
+in the random walk, $i$, by sampling by sampling the probability
+distribution function formed by the $i^{th}$ of the probability
+matrix. The transition weight is then defined as:
 \begin{equation}
@@
 methods, they achieved identical numerical results as and marginally
-better performance than conventional Krylov solvers. Their approach
+better performance than conventional subspace methods. Their approach
 was instead to use residual Monte Carlo as a synthetic acceleration
@@
   \begin{gather}
+    \label{eq:mcsa_1}
     \mathbf{r}^{k} = \mathbf{b} - \mathbf{A}\mathbf{x}^{k}\:,\\
+    \label{eq:mcsa_2}
     \mathbf{x}^{k+1/2} = \mathbf{x}^k + \mathbf{r}^k\:,\\
+    \label{eq:mcsa_3}
     \mathbf{r}^{k+1/2} = \mathbf{b} - \mathbf{A}\mathbf{x}^{k+1/2}\:,\\
-    \label{eq:mcsa_mc_solve}
+    \label{eq:mcsa_4}
     \mathbf{A}\delta\mathbf{x}^{k+1/2} = \mathbf{r}^{k+1/2}\:,\\
+    \label{eq:mcsa_5}
     \mathbf{x}^{k+1} = \mathbf{x}^{k+1/2} + \delta \mathbf{x}^{k+1/2}\:,
@@
 index. Here, a Neumann-Ulam Monte Carlo method is used to generate the
-solution correction in Eq~\ref{eq:mcsa_mc_solve} from the residual and
+solution correction in Eq~\ref{eq:mcsa_4} from the residual and
 Richardson's iteration in the first step has been rewritten as a
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Parallel MCSA Algorithm}
+\section{Parallel Algorithms}
+In this section devise a parallel algorithm for the Neumann-Ulam Monte
+Carlo method based on the multiple-set overlapping-domain
+decomposition algorithm and a parallel algorithm for the MCSA
+iteration that leverages the parallel Monte Carlo algorithm and
+general parallel matrix-vector operations.
 
@@
 
+In the context of radiation transport, in 2009 Brunner and colleagues
+provided a fully asynchronous domain decomposed parallel algorithm as
+implemented in production implicit Monte Carlo codes
+\cite{brunner_efficient_2009}. In their algorithm, a set of mutually
+exclusive tasks enabled by the fully asynchronous communication
+patterns are executed. For the Neumann-Ulam method, mutually
+exclusivity permits all local operations to occur independently of
+other processes in the system and their current state to achieve an
+optimal overlap of work and communication. We will adapt their
+algorithm and directly apply it to construct a parallel formulation of
+the Neumann-Ulam method. Direct analogs can be derived from their
+works by noting that the primary difference between solving a linear
+transport system with Monte Carlo methods and traditional fixed source
+Monte Carlo transport problems is the content of the Markov chains
+that are generated.
+
+Consider this communication pattern for the 9 subdomain example given
+by Figure~\ref{fig:nearest_neighbor_comm}. In this pattern, each
+boundary domain has two neighbors and the center domain four neighbors
+with which they will communicate parallel histories and this
+communication goes both ways as represented by the adjacent arrows in
+the figure. For each set of neighbors, a non-blocking send and receive
+is required with data buffers allocated with a user-defined size for
+incoming and outgoing histories. This non-blocking structure is
+critical to the performance of the algorithm in that it permits local
+history transport to continue while new histories to transport are
+being collected in the buffers. When a given process is ready to do
+more work, it can check these data buffers for incoming histories
+requiring further transport. In this way there is a maximum amount of
+overlap between communication and transport of histories.
+
+\begin{figure}[h!]
+  \begin{center}
+    \scalebox{0.7}{ \input{domain_to_domain.pdftex_t} }
+  \end{center}
+  \caption{Nearest neighbor history communication sequence.
+    \textit{Each subdomain in the system has a set of nearest
+      neighbors determined by the parallel adjacency graph of the
+      input matrix. The subdomains are indexed by an integer.}}
+  \label{fig:nearest_neighbor_comm}
+\end{figure}
+
+Once the nearest-neighbor communication sequence has been completed
+for all histories, the completion of transport sequence is
+required. To stop transport, an asynchronous binary communication tree
+as presented for the same 9 subdomain example in
+Figure~\ref{fig:binary_comm_tree}. In this communication pattern, each
+process has one parent process (except for the MASTER process 0) to
+which it will non-blocking send the number of histories that
+terminated their transport procedure by weight cutoff in the local
+domain. Equivalently, each process has up to two child processes from
+which it will receive their completed number of histories with a
+non-blocking receive operation. Setting up a tree in this manner lets
+the completed history tally be updated incrementally and funneled to
+the root process. Because we are solving fixed source problems with
+the Neumann-Ulam algorithm without any type of variance reduction that
+may generate more histories that we started with, once the root
+process completion sum tallies to the number of histories in the
+source, transport is complete. Once this occurs, the root process
+non-blocking sends the stop message to its children and each process
+non-blocking receives a stop message from its parent as shown in
+Figure~\ref{fig:binary_comm_tree}. The stop message is then propagated
+up the tree in the same manner.
+
+\begin{figure}[h!]
+  \begin{center}
+    \scalebox{0.6}{
+      \input{binary_comm_tree.pdftex_t} }
+  \end{center}
+  \caption{Binary communication tree for coordinating the end of a
+    parallel Neumann-Ulam solve. \textit{Each child process reports to
+      its parents how many histories completed within its domain. When
+      the root process sums all complete histories, it forwards the
+      stop signal to its children which in turn forward the message up
+      the tree. The subdomains are indexed by an integer.}}
+  \label{fig:binary_comm_tree}
+\end{figure}
+
 \subsection{MSOD Algorithm}
@@
 remained in a physical cabinet in a large distributed machine, thus
-reducing latency costs during communication. A multiple set scheme is
-also motivated by the fact that communication during particle
-transport only occurs within a set, limiting communications during the
-transport procedure to a group of $O(100)$ processors, a number that
-was shown to have excellent parallel efficiencies in Brunner's work
-\cite{brunner_efficient_2009} and therefore will scale well in this
-algorithm. The overlapping domains within each set also demonstrated
-reduced communication costs.
+reducing latency costs during communication.
+
+\subsection{Parallel MCSA Algorithm}
+We utilize parallel matrix-vector multiply, vector update, and vector
+norm operations as oultined in \cite{saad_iterative_2003} for a
+parallel MCSA implementation with these computations required in
+Eqs~(\ref{eq:mcsa_1}), (\ref{eq:mcsa_2}), (\ref{eq:mcsa_4}), and
+(\ref{eq:mcsa_5}). In the first step, a parallel vector update is used
+to apply the residual to the previous iterate's solution in a
+Richardson iteration. In the next step and Eq~(\ref{eq:mcsa_5}), the
+residual is computed by a parallel matrix-vector multiply and vector
+update. Once the correction is computed with a parallel adjoint
+Neumann-Ulam solve, this correction is applied to the solution with a
+parallel vector update. Additionally, as given by
+Eq~(\ref{eq:mcsa_stopping_criteria}), 2 parallel vector reductions
+will be required to check the stopping criteria: one initially to
+compute the infinity norm of the source vector, and another at every
+iteration to compute the infinity norm of the residual vector.
+
+To parallelize the solution of Eq~(\ref{eq:mcsa_3}), we apply the
+MSOD Neumann-Ulam algorithm presented in the previous section. The
+problem is potentially replicated with overlap and the Monte Carlo
+procedure returns a correction vector, $\delta \mathbf{x}$, with the same
+parallel decomposition as the input linear problem. As was the case
+for generating the replications required for a multiple set
+implementation, the explicit matrix and vector operations required for
+the rest of an MCSA iteration occur only on the subset of processors
+containing the primary set\footnote{For applications where resiliency
+  is potentially of concern, the operations in Eqs~(\ref{eq:mcsa_1}),
+  (\ref{eq:mcsa_2}), (\ref{eq:mcsa_4}), and (\ref{eq:mcsa_5}) may
+  be replicated without significant penalty. Doing so my alleviate
+  issues where any part of the linear problem or residual is disturbed
+  as now there will be multiple copies of these data
+  structures.}. Only the Monte Carlo data structures exist on the rest
+of the processors in the problem as replication of the other
+operations does not generate any more information that could be used
+to accelerate the time to solution.
 
@@
 \section{Parallel Scaling Studies}
+We will next measure performance in various scaling studies using an
+implementation of the parallel MCSA algorithm to assess its quality
+using a simple two-dimensional neutron diffusion problem in a
+homogeneous medium. For the scaling results presented here, all
+parallel MCSA results were computed using the Monte Carlo Linear
+Solvers\footnote{MCLS is an open source C++ library built on the
+  Trilinos\cite{heroux_overview_2005} framework and can be downloaded
+  at \url{https://github.com/sslattery/MCLS}} (MCLS) library generated
+specifically for this work. For every scaling study presented, the
+calculations were performed 3 times with the average values reported
+for wall time and those average values used for the efficiency
+computations. For all cases, both strong and weak scaling studies were
+performed. For strong scaling, the objective is to then decrease the
+runtime of a fixed size global problem as a linear function of
+$P$. Based on this, we can define the strong scaling \textit{absolute
+  efficiency} as:
+\begin{equation}
+  \eta_{strong}(N,P) = \frac{1}{P} \frac{T(N,1)}{T(N,P)}\:,
+  \label{eq:strong_scaling_absolute}
+\end{equation}
+where $T(N,P)$ is the wall time for a computation of global size $N$
+using $P$ processes. Note here that this particular definition is
+measured with respect to a serial computation. On leadership class
+machines, it is often the case that a problem size large enough to
+effectively execute on $O(10,000)$ cores or larger is significantly
+larger than any problem that may be solved by a single core due to
+memory restrictions. Therefore, we consider absolute scaling for a
+base case where $Q$ processes are used instead of a serial
+computation:
+\begin{equation}
+  \eta_{strong}(N,P|Q) = \frac{Q}{P} \frac{T(N,Q)}{T(N,P)}\:.
+  \label{eq:strong_scaling_absolute_ref}
+\end{equation}
+For weak scaling, For strong scaling, the objective is to then decrease the runtime of a
+fixed size global problem as a linear function of $P$. Based on this,
+we can define the strong scaling \textit{absolute efficiency} as:
+\begin{equation}
+  \eta_{strong}(N,P) = \frac{1}{P} \frac{T(N,1)}{T(N,P)}\:,
+  \label{eq:strong_scaling_absolute}
+\end{equation}
+where $T(N,P)$ is the wall time for a computation of global size $N$
+using $P$ processes. Note here that this particular definition is
+measured with respect to a serial computation. On leadership class
+machines, it is often the case that a problem size large enough to
+effectively execute on $O(10,000)$ cores or larger is significantly
+larger than any problem that may be solved by a single core due to
+memory restrictions. Therefore, we consider absolute scaling for a
+base case where $Q$ processes are used instead of a serial
+computation:
+\begin{equation}
+  \eta_{strong}(N,P|Q) = \frac{Q}{P} \frac{T(N,Q)}{T(N,P)}\:.
+  \label{eq:strong_scaling_absolute_ref}
+\end{equation}
+All parallel scaling studies presented in this work were performed on
+the Titan Cray XK7 machine at the Oak Ridge Leadership Computing
+Facility located at Oak Ridge National Laboratory.
 
-\subsection{Overlap Scaling}
+\subsection{Pure Domain Decomposition Scaling}
+\label{subsec:pure_domain_decomp}
+For the first set of scaling studies, we consider the parallel MCSA
+algorithm where MSOD has not been used in the parallel Neumann-Ulam
+algorithm. We refer to this case as pure domain decomposition where no
+overlap or replication has been leveraged. Such a test will isolate
+specifically the performance of Brunner and Brantley's algorithm as
+applied to the Neumann-Ulam algorithm. For the strong scaling study,
+the global diffusion problem size was fixed at \sn{1.6}{7} DOFs for
+all solvers and the 16 core run used as the base
+case. Figure~\ref{fig:titan_pure_strong_time} gives the wall time per
+iteration for these calculations while
+Figure~\ref{fig:titan_pure_strong} gives the results of this study
+with the given absolute efficiencies computed using
+Eq~(\ref{eq:strong_scaling_absolute}). All MCSA calculations converged
+in 22 iterations.
 
-\subsection{Multiple Sets Scaling}
+\begin{figure}[h!]
+  \begin{center}
+    \includegraphics[width=3in]{titan_pure_strong_time.pdf}
+  \end{center}
+  \caption{Pure domain decomposition strong scaling wall time per
+    iteration.}
+  \label{fig:titan_pure_strong_time}
+\end{figure}
+
+\begin{figure}[h!]
+  \begin{center}
+    \includegraphics[width=3in]{titan_pure_strong.pdf}
+  \end{center}
+  \caption{Pure domain decomposition strong scaling absolute
+    efficiency. \textit{Super-linear speed-up is from memory thrashing
+      in the base case.}}
+  \label{fig:titan_pure_strong}
+\end{figure}
+
+There are several key features of
+Figure~\ref{fig:titan_pure_strong_time} and
+Figure~\ref{fig:titan_pure_strong} worth noting. First, up to around
+5,000 cores, all methods experience a super-linear improvement in
+parallel efficiencies. For all methods, this increase is not a
+function of the algorithmic efficiency as those computed using
+Eq~(\ref{eq:algorithmic_efficiency}) remained at unity, meaning that
+all computations converged in the same number of iterations relative
+to the base case. Instead, what is happening here and was observed by
+Brunner and Brantley in their work is that the large problem used to
+scale to O(10,000) cores was large enough that memory thrashing was
+occurring in the 16 core base case. As the number of cores in the
+strong scaling study was increased and the local problem size
+decreased, more of the local problem fit into the cache of the system,
+thus causing improved serial performance and therefore enhanced run
+times.
+
+Beyond around 5,000 cores, the global problem size selected for this
+strong scaling study causes the local problem size to shrink enough
+that parallel communication costs begin to overtake arithmetic costs
+on process. This takeover causes this effective strong scaling wall
+that is observed in all methods. Using any more cores to solve this
+particular global problem is not an efficient use of those parallel
+resources. Furthermore, the fact that we see very similar trends in
+efficiencies for all methods means that they are bound by the same
+parallel performance limitations in a strong scaling environment. This
+is important as we can expect the same qualitative scaling performance
+from MCSA if arithmetic optimization has been performed to improve the
+serial performance of the implementation. In addition, these results
+are valuable in that they show MCSA, when properly parallelized, has
+the ability to be competitive on leadership-class computing platforms
+with conventional subspace methods with good strong scaling observed
+in the pure domain decomposition case to O(10,000) cores.
+
+Next, we consider the weak scaling performance of the algorithm. In
+this case, we fix the local problem size at \sn{4}{4} DOFs and
+increase the number of processors used in the
+simulation. Figure~\ref{fig:titan_pure_weak_time} gives the wall time
+per iteration and Figure~\ref{fig:titan_weak_absolute} gives the weak
+scaling absolute efficiencies for this problem. Overall, weak scaling
+performance of the parallel MCSA algorithm is excellent for the pure
+domain decomposition case with absolute parallel efficiencies of over
+80\% observed at over 32,000 cores.
+
+\begin{figure}[p!]
+  \begin{center}
+    \includegraphics[width=3in]{titan_pure_weak_time.pdf}
+  \end{center}
+  \caption{Pure domain decomposition weak scaling wall time
+      per iteration.}
+  \label{fig:titan_pure_weak_time}
+\end{figure}
+
+\begin{figure}[h!]
+  \begin{center}
+    \includegraphics[width=3in]{titan_weak_absolute.pdf}
+  \end{center}
+  \caption{\textbf{Pure domain decomposition weak scaling absolute
+      efficiency.} \textit{MCLS is an over order of magnitude slower
+      arithmetically, causing the higher efficiency. GMRES executes
+      more operations than conjugate gradient, also creating a higher
+      efficiency.}}
+  \label{fig:titan_weak_absolute}
+\end{figure}
+
+\subsection{Overlapping Domain Decomposition}
+\label{subsec:overlapping_domain_decomp}
+
+In the next set of scaling studies, we will explore how adding overlap
+to the subdomains in the problem affects run times and parallel
+performance as compared to the case with pure domain decomposition. To
+do this, we utilize the set of analytic relations developed in
+Appendix~\ref{chap:parallel_theory} to design these numerical
+experiments. For the neutron diffusion problem used in these scaling
+studies, the linear operator had a spectral radius of 0.787 and a
+weight cutoff of \sn{1}{-2} was used with the Neumann-Ulam
+solver. Using Eq~(\ref{eq:analytic_k}) gives an expected random walk
+length of 19.22 transitions for any given history in the
+problem. Using the mean-chord approximation given by
+Eq~(\ref{eq:mean_chord_domain_leakage}), we then expect approximately
+5.3\% of the histories born in a particular domain to leak out after
+the first transport stage. By adding overlap, we aim to reduce this
+fraction of communicated histories and therefore boost the parallel
+performance of the algorithm.
+
+To determine the amount of overlap to test for the scaling studies, we
+use a variation of Eq~(\ref{eq:step_k_length}) to determine how many
+discrete states a history will move on average from its birth site by
+modifying it to only account for discrete movements:
+\begin{equation}
+  \sqrt{\bar{n^2_k}} = n_s \sqrt{k}\:,
+  \label{eq:discrete_distance}
+\end{equation}
+where $\sqrt{\bar{n^2_k}}$ is now the root-mean-squared number of
+discrete states a history will move from its birth site. For our
+diffusion problem used in these scaling studies, we compute a value of
+2.63 states per history for any of the dimensions of the system. This
+means that we can expect to end up on average 2.63 states in both the
+$i$ and $j$ dimensions away from the starting state. Therefore, we
+will use the overlap generation algorithm outlined in
+\S~\ref{subsec:msod_generation} such that we parametrically grow the
+local domain using 1, 2, 3, and 5 levels to measure how scalability is
+affected when this root-mean-squared number of states is
+considered. For example, using 3 levels of overlap should keep over
+50\% of the histories that would be communicated in the pure domain
+decomposition case on-process as this is greater than the
+root-mean-squared number of states we would expect a history to
+travel.
+
+For the performance metrics, we may use the same weak and strong
+scaling efficiency metrics that we used for the pure domain
+decomposition case. As we are comparing to the case without overlap
+and are looking to boost performance relative to this case, for both
+weak and strong scaling calculations we will use the 16-core case
+without overlap as the reference computation for all efficiency
+measures reported. The absolute efficiency results from the strong
+scaling exercise are given in
+Figure~\ref{fig:titan_strong_overlap}. Aside from the boosting of
+efficiencies at 2,048 cores, little evidence of strong scaling
+improvement is evident for all levels of overlap tested. To further
+explore any potential benefits of adding overlap into the system,
+Figure~\ref{fig:titan_strong_overlap_diff} gives the difference in the
+efficiencies between all overlap cases and the reference case without
+overlap at each core count. At smaller numbers of cores, there is no
+observed benefit to using overlap to improve strong scaling. The
+improvements noted at 2,048 cores are also present here and account
+for the best efficiency improvements for all core counts tested. At
+16,384 cores, using 2 levels of overlap did give a 6.4\% boost in
+efficiency while at 1,024 cores this same amount of overlap actually
+reduced the efficiency. Beyond 2 levels of overlap, efficiency levels
+begin to fall with 5 levels of overlap performing the worst for all
+cases. From this, we don't expect adding any more overlap will improve
+the scalability of the algorithm and the best strong scaling
+performance was observed when the number of overlap levels is
+approximately equal to the root-mean-squared number of states we
+expect a history to move from its starting point as computed by
+Eq~(\ref{eq:discrete_distance}).
+
+\begin{figure}[h!]
+  \begin{center}
+    \includegraphics[width=3in]{titan_strong_overlap.pdf}
+  \end{center}
+  \caption{\textbf{Strong scaling absolute efficiency with varying
+      levels of overlap.} \textit{Values of 0, 1, 2, 3, and 5 levels
+      of overlap were used as the problem had a root-mean-squared
+      number of states of 2.63 for each history.}}
+  \label{fig:titan_strong_overlap}
+\end{figure}
+
+\begin{figure}[h!]
+  \begin{center}
+    \includegraphics[width=3in]{titan_strong_overlap_diff.pdf}
+  \end{center}
+  \caption{\textbf{Strong scaling absolute efficiency difference with
+      respect to the 0 overlap base case.} \textit{Values of 1, 2, 3,
+      and 5 levels of overlap are compared here with the absolute
+      difference of the absolute efficiencies presented.}}
+  \label{fig:titan_strong_overlap_diff}
+\end{figure}
+
+For weak scaling, the same study as in the pure domain decomposition
+case was performed again with 1, 2, 3, and 5 levels of overlap. The
+absolute parallel efficiencies from these calculations are given in
+Figure~\ref{fig:titan_weak_overlap} and the differences in
+efficiencies from the calculations without overlap given in
+Figure~\ref{fig:titan_weak_overlap_diff}. The results here are much
+the same in that there is not a dramatic improvement in scalability
+relative to the case without overlap. Opposite the strong scaling
+calculation, overlap enhances weak scaling efficiencies at lower core
+counts and degrades the efficiencies at higher core counts. In
+addition, 1 level of overlap provided the best enhancement of
+scalability when overlap was beneficial.
+
+\begin{figure}[h!]
+  \begin{center}
+    \includegraphics[width=3in]{titan_weak_overlap.pdf}
+  \end{center}
+  \caption{\textbf{Weak scaling absolute efficiency with varying
+      levels of overlap.} \textit{Values of 0, 1, 2, 3, and 5 levels
+      of overlap were used as the problem had a root-mean-squared
+      number of states of 2.63 for each history.}}
+  \label{fig:titan_weak_overlap}
+\end{figure}
+
+\begin{figure}[h!]
+  \begin{center}
+    \includegraphics[width=3in]{titan_weak_overlap_diff.pdf}
+  \end{center}
+  \caption{\textbf{Weak scaling absolute efficiency difference with
+      respect to the 0 overlap base case.} \textit{Values of 1, 2, 3,
+      and 5 levels of overlap are compared here with the absolute
+      difference of the absolute efficiencies presented.}}
+  \label{fig:titan_weak_overlap_diff}
+\end{figure}
+
+For both the strong and weak scaling studies, the parallel performance
+of the algorithm was either marginally improved or not improved at
+all. We would always expect an improvement in performance for using
+overlap to prevent the transport of histories from domain to domain
+during execution of the Neumann-Ulam algorithm. The reduction in
+transport means a reduction in parallel communication and therefore
+should never degrade the performance of the algorithm. It is true that
+doing so in fact enhances Monte Carlo performance. However, the
+Neumann-Ulam algorithm is not representative of the entire MCSA
+algorithm. When overlap is used, an overlapping tally vector is
+constructed in parallel such that multiple cores may own multiple
+states in the system. When this occurs, the tally data associated with
+these shared states must be communicated after the Neumann-Ulam solve
+as outlined in \S~\ref{subsec:msod_algorithm}. This reduction
+operation then builds the full MCSA correction vector in the
+decomposition of the solution vector.
+
+Therefore, by building overlap we are in fact reducing parallel
+communication during the Neumann-Ulam solve but we are simply
+deferring that communication until after Monte Carlo transport when
+the overlapping tally vector must be reduced. When overlap becomes
+large enough (i.e. larger than the root-mean-squared number of
+discrete states a history will travel), then the savings in
+communication during the Neumann-Ulam solve start to become smaller
+than the communication created in the overlapping tally vector
+reduction. We see this in both the strong and weak scaling cases at
+larger levels of overlap. For the strong scaling case, we see better
+performance improvement at higher core counts because the total number
+of overlapping states on each core is smaller due to the local problem
+size shrinking. For weak scaling, we see better performance
+improvement at lower numbers of cores because although the local
+overlap size is fixed, the global amount of overlap is growing with
+core count and therefore reducing the performance of the tally
+reduction operation.
+
+\subsection{Multiple Set Domain Decomposition}
+\label{subsec:ms_decomposition}
+
+We next look to apply replication to the problem in the form of
+multiple sets. As the previous analysis showed overlap to be of little
+assistance in improving scalability, we only apply multiple sets to
+the problem with pure domain decomposition within the sets for these
+scaling studies. When adding sets, we must decide how many stochastic
+histories to use to generate the correction vector in the MCSA
+iteration. For this work we consider two choices; the split case and
+the fully replicated case. As an example of these cases, consider a
+single set problem where 20,000 histories are used in the Monte Carlo
+solve. If we choose the split case, then solving the same problem with
+2 sets would give 10,000 histories in each set and 4 sets would give
+5,000 histories in each set. Splitting the histories in this way
+provides the same global number of histories used to compute the
+correction in the single set problem and we should therefore expect
+the same iterative performance at a reduced time per iteration and
+thus reduced time to solution. For the replicated history case, each
+additional set in the problem will use just as many histories as the
+single set case. Using 2 sets with fully replicated histories for our
+example problem then has each set computing 20,000 histories with
+40,000 global histories in the computation. For the 4 set case, the
+number of global histories is increased to 80,000. By increasing the
+global number of histories, the replicated cases should approximately
+maintain time per iteration but should decrease the number of
+iterations required to converge and subsequently reduce time to
+solution.
+
+\subsubsection{Strong Scaling with Multiple Sets}
+\label{subsubsec:ms_strong}
+
+For the strong scaling case, recall that the objective is to decrease
+the time to solution for a given global problem size $N$ by increasing
+$P$. Therefore, we may use Eqs~(\ref{eq:strong_scaling_absolute_ref}),
+(\ref{eq:algorithmic_efficiency}), and
+(\ref{eq:implementation_efficiency}) to compute the various strong
+scaling efficiencies, but we must modify the reference computation in
+each of these formulas to reflect our objective. For all efficiency
+computations, the 16 core single set computation will be used as the
+reference case. By adding more cores to the problem, either by
+replicating with multiple sets, or decreasing the local problem size,
+any computational resources added to the problem should aim to result
+in speedup of this reference case. If this speedup for one technique
+of adding cores is less than another, then that technique is deemed a
+less efficient use of those computational resources.
+
+For the multiple sets problem, the single set strong scaling study
+from the pure domain decomposition base case was repeated for multiple
+sets with both splitting of histories across sets and full replication
+of histories across sets using both 2 and 4
+sets. Figure~\ref{fig:titan_strong_ms_time} gives the actual wall
+times for these computations. There are two important features to note
+in the wall time data. First, splitting the histories among sets
+instead of replicating them is faster due to the fact that cutting the
+number of histories in half reduces the time of an iteration by a
+factor that is nearly linear while increasing linearly the number of
+histories through replication does not decrease the iterations
+required to converge by a linear factor. Second, until the strong
+scaling wall is hit starting at around 10,000 cores, the single set
+case actually performs the fastest with only marginal improvement in
+run times at higher numbers of cores with more sets. At these larger
+numbers of cores, however, the improvements are so modest that one is
+better off running the same problem using a single set at lower core
+counts to conserve computational resources while maintaining a good
+time to solution.
+
+\begin{figure}[h!]
+  \begin{center}
+    \includegraphics[width=3in]{titan_strong_ms_time.pdf}
+  \end{center}
+  \caption{\textbf{Wall time in seconds to solution for each case for
+      the strong scaling study with multiple sets.} \textit{Until the
+      strong scaling wall is reached at O(10,000) cores, the single
+      set case performs the best. For the multiple set cases, it is
+      faster to split histories among sets and maintain iterative
+      performance rather than replicate histories to improve iterative
+      performance.}}
+  \label{fig:titan_strong_ms_time}
+\end{figure}
+
+From this timing data we compute the absolute efficiencies for strong
+scaling as given in Figure~\ref{fig:titan_strong_ms_eff} using the
+16-core single set calculation as the reference case. As observed in
+the timing data, because the single set case had the fastest run times
+up to the strong scaling wall it also has the best observed
+efficiencies. For the 2 set case with history splitting, although the
+time to perform an iteration was nearly cut in half, the extra
+parallel overhead associated with reducing the tally across the blocks
+to combine the set results increases run times and reduces
+efficiencies. This reduction across blocks is even more obvious for
+the 4 set case. Ultimately, all cases effectively hit the strong
+scaling wall at the same time at around O(10,000) cores.
+
+We expect this if we consider the fact that adding sets to the problem
+is in itself effectively a form of strong scaling. For any given
+problem of fixed size $N$, adding extra sets to the computation
+decreases the global ratio of $N/P$ as $P$ will grow with the addition
+of sets (although $N/P$ is constant within a set). For the split
+history case, the splitting further exacerbates this situation by
+further decreasing the amount of histories that will be computed in
+the local domain, giving a larger ratio of communication to work in
+the Monte Carlo solve. As with with single set results, we also
+observe some efficiencies greater than unity for the multiple set
+cases also due to improved cache utilization.
+
+Although not as dramatic as one might originally expect from
+introducing replication, there are some modest gains in efficiencies
+at high core counts by using multiple sets. Of interest here is where
+the multiple set cases perform better than the single set case after
+the single set case has hit the strong scaling wall. In
+Figure~\ref{fig:titan_strong_ms_eff}, at 16,384 cores the single set
+case operates at 38\% efficiency while at the same number of cores,
+the 2 set case with history replication operates at 58\%
+efficiency. Although the efficiencies here are still low, this is
+reasonable improvement in the use of those computational resources,
+gained by replicating the problem. Beyond 16,384 cores, there is not a
+compelling reason to use more computational resources for this
+problem.
+
+\begin{figure}[h!]
+  \begin{center}
+    \includegraphics[width=3in]{titan_strong_ms_eff.pdf}
+  \end{center}
+  \caption{\textbf{Multiple set absolute parallel efficiency.}
+    \textit{Computed relative to the 16-core 1-set base case. Using a
+      single set is the best means by which to improve strong scaling
+      up to the strong scaling wall. At higher core counts past where
+      the single set case hits the strong scaling wall adding sets can
+      provide some improvements in efficiencies.}}
+  \label{fig:titan_strong_ms_eff}
+\end{figure}
+
+Also of interest here is the effect of splitting histories among sets
+versus replicating them on the iterative performance of the
+method. Table~\ref{tab:ms_strong_alg_eff} gives the iterations
+required to converge and computed algorithmic efficiencies for each of
+the cases presented in the strong scaling data. As expected, splitting
+histories maintains a fixed global number of histories and therefore
+maintained the number of iterations required to converge relative to
+the single set base case. When histories were replicated, adding sets
+decreased the number of iterations required to converge in a nonlinear
+fashion. Therefore, increasing the number of histories per MCSA
+iteration improves algorithmic efficiency. Using these algorithmic
+efficiencies, the implementation efficiencies for the multiple set
+cases were computed and are given in
+Figure~\ref{fig:titan_strong_ms_impeff}. From these results, we see
+that the replicated history cases are even less efficient than in the
+absolute case with respect to the performance of a single
+iteration. This is due to the fact that doubling the number of
+histories does not halve the number of MCSA iterations required for
+convergence as the acceleration is bound by the Central Limit
+Theorem. As a result, the improved iterative performance from the
+replicated cases comes at a higher cost in terms of time to solution
+than the split cases.
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lcc}\hline\hline
+      \multicolumn{1}{l}{Case}& 
+      \multicolumn{1}{c}{Iterations}&
+      \multicolumn{1}{c}{$\eta_{alg}$} \\\hline
+      1 & 22 & 1.0 \\
+      2-split & 22 & 1.0 \\
+      2-replicated & 16 & 1.375 \\
+      4-split & 22 & 1.0 \\
+      4-replicated & 13 & 1.692 \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Strong scaling algorithmic efficiency using
+      multiple sets.} \textit{Replicating histories improves
+      algorithmic efficiency by converging in fewer iterations.}}
+  \label{tab:ms_strong_alg_eff}
+\end{table}
+
+\begin{figure}[h!]
+  \begin{center}
+    \includegraphics[width=3in]{titan_strong_ms_impeff.pdf}
+  \end{center}
+  \caption{\textbf{Multiple set strong scaling implementation
+      efficiency.} \textit{Computed relative to the 16-core 1-set base
+      case. Considering iterations required to converge, the
+      performance of the split history cases is much better than the
+      cases with history replication.}}
+  \label{fig:titan_strong_ms_impeff}
+\end{figure}
+
+\subsubsection{Weak Scaling with Multiple Sets}
+\label{subsubsec:ms_weak}
+
+Next, we continue to a weak scaling study of the multiple sets
+problem. In the same way as the strong scaling study, we will modify
+the pure domain decomposition study to account for the additional
+replication. For each calculation in the single set study, that
+problem will be replicated either 2 or 4 times with both split and
+replicated histories to assess the effects on scalability and
+run time. Figure~\ref{fig:titan_weak_ms_time} gives the wall times from
+the multiple sets weak scaling study. Of primary importance here is
+the fact that adding sets to the problem with both split and
+replicated histories actually decreases the time to solution. This was
+not observed in the strong scaling case until much larger core counts
+as the multiple set computations hit the strong scaling wall after the
+single set computation.
+
+This result is significant in that it shows a solution technique in
+which the linear problem can actually be replicated and enhance time
+to solution, something that cannot be achieved with conventional
+methods. At larger core counts, however, this reduction in compute
+time is not as large due to the fact that the tally reduction across
+blocks requires more resources and those resources are clearly growing
+with core count (most obviously in the 4 set case with split
+histories). In addition, at 131,072 cores for the largest 4 set
+computation, processors that share a block may be some physical
+distance from one another in the machine, adding to latency and
+decreasing performance.
+
+\begin{figure}[h!]
+  \begin{center}
+    \includegraphics[width=3in]{titan_weak_ms_time.pdf}
+  \end{center}
+  \caption{\textbf{Wall time in seconds to solution for each case for
+      the weak scaling study with multiple sets.}}
+  \label{fig:titan_weak_ms_time}
+\end{figure}
+
+For a multiple set analysis of the weak scaling timing results, we
+again have to consider that adding sets to the problem is actually a
+form of strong scaling. Consider a single set problem ($S=1$) for a
+given $P$ and $N$. When that problem is now solved using 2 sets we
+have $S=2$, $N$, and $2P_s$ where $P_s$ is the number of processors in
+a single set. For any number of sets we then have a global problem
+size of $N$ while the number of processors is $SP_s$. Therefore,
+$N/P$ is no longer fixed in the weak scaling study but rather
+$N/(SP_s)$ is fixed. As a result, we modify the weak scaling absolute
+efficiency relationship to account for this fact:
+\begin{equation}
+\eta_{weak}(M|N,P|Q,S) = \frac{T(M,Q)}{S \times T(N,P)}\:,
+  \label{eq:ms_weak_efficiency}
+\end{equation}
+where now the absolute efficiency is a function of the number of sets
+in the problem and again $M/Q = N/P$. As with the strong scaling
+study, we will use the 16-core single set case as the reference
+computation for efficiencies. Interestingly,
+Eq~(\ref{eq:ms_weak_efficiency}) has a very similar form to the strong
+scaling efficiency given by Eq~(\ref{eq:strong_scaling_absolute}),
+representative of the strong scaling nature of adding additional
+sets. For problems where adding sets improves time to solution, this
+new weak scaling formulation accounts for how efficiently those extra
+resources are incorporated into the problem.
+
+Figure~\ref{fig:titan_weak_ms_eff} gives the results of the weak
+scaling efficiency computations using the multiple set metric given by
+Eq~(\ref{eq:ms_weak_efficiency}). Although we do reduce the time to
+solution over the base cases for nearly all data points, the
+observation here is that the reduction in time to solution is still
+not large enough to efficiently use all resources. For example, for
+the 2 set case to be a perfect use of resources with either split or
+replicated histories, the run time of that calculation must be half of
+that for the single set case. This was not the observation for the
+run times in Figure~\ref{fig:titan_weak_ms_time} and therefore the
+single set case is still the most efficient use of computational
+resources in this scaling study.
+
+\begin{figure}[h!]
+  \begin{center}
+    \includegraphics[width=3in]{titan_weak_ms_eff.pdf}
+  \end{center}
+  \caption{\textbf{Weak scaling absolute efficiency relative to
+      16-core 1-set base case.}}
+  \label{fig:titan_weak_ms_eff}
+\end{figure}
+
+As in the strong scaling case, splitting and replicating histories
+across sets had effectively the same results for the weak scaling
+study with the iterations to converge and computed algorithmic
+efficiencies given by Table~\ref{tab:ms_weak_alg_eff}. The iterations
+to converge these problems were not a function of problem size. Again,
+when the implementation efficiencies are computed, we see in
+Figure~\ref{fig:titan_weak_ms_impeff} that splitting histories among
+sets is still a more efficient manner of introducing multiple sets
+rather than replicating histories in an attempt to reduce the number
+of iterations required to converge.  We also note that this is not a
+true weak scaling study as indicated by
+Eq~(\ref{eq:ms_weak_efficiency}) but rather an odd combination of both
+weak and strong scaling. Regardless of this, although multiple sets
+are not the most efficient use of resources in this analysis, they do
+enhance the time to solution through replication and do so
+significantly at core counts of O(1,000) and lower. In both the weak
+and strong scaling cases, due to the additional computational and
+communication overhead of adding sets to the system, one should never
+expect to seriously boost parallel efficiencies for MCSA using this
+technique except for certain instances past the strong scaling wall in
+a purely strong scaling environment. In a weak scaling environment,
+although no efficiency increase will be noted, one can expect an
+enhancement of time to convergence by replicating.
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lcc}\hline\hline
+      \multicolumn{1}{l}{Case}& 
+      \multicolumn{1}{c}{Iterations}&
+      \multicolumn{1}{c}{$\eta_{alg}$} \\\hline
+      1 & 22 & 1.0 \\
+      2-split & 22 & 1.0 \\
+      2-replicated & 16 & 1.375 \\
+      4-split & 22 & 1.0 \\
+      4-replicated & 13 & 1.692 \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Weak scaling algorithmic efficiency using multiple
+      sets.} \textit{Replicating histories improves algorithmic
+      efficiency by converging in fewer iterations.}}
+  \label{tab:ms_weak_alg_eff}
+\end{table}
+
+\begin{figure}[h!]
+  \begin{center}
+    \includegraphics[width=3in]{titan_weak_ms_impeff.pdf}
+  \end{center}
+  \caption{\textbf{Implementation parallel efficiency relative to 16-core
+      1-set base case.}}
+  \label{fig:titan_weak_ms_impeff}
+\end{figure}
 
@@
 observed to actually improve the time to solution. Unlike a
-traditional Krylov method that we might apply to solve a neutron
+traditional subspace method that we might apply to solve a neutron
 transport problem, using MCSA means that we can actually make a

### 2013-09-05 doc/sna2013/paper/slattery_evans_wilson.tex
adding parallel efficiency measures to sna paper

@@
 \usepackage{tmath}
+\usepackage[section]{placeins}
 
@@
 
-\abstract{We present a novel multiple-set overlapping-domain
-  decomposed strategy for parallelizing the Monte Carlo Synthetic
-  Acceleration (MCSA) method. MCSA methods use the Neumann-Ulam class
-  of Monte Carlo solvers for linear systems to accelerate the fixed
-  point method on which the Neumann-Ulam solvers are based. To
-  effectively parallelize MCSA methods requires the parallelization of
-  the underlying Neumann-Ulam solvers. To do this in a domain
-  decomposed environment, we borrow strategies traditionally
-  implemented in Monte Carlo particle transport to parallelize the
-  problem. The multiple-set overlapping-domain decomposition algorithm
-  is presented along with parallel scaling data for the resulting
-  implementation using the Titan Cray XK7 machine at Oak Ridge
-  National Laboratory.}
+\abstract{We present a multiple-set overlapping-domain decomposed
+  strategy for parallelizing the Monte Carlo Synthetic Acceleration
+  (MCSA) method. MCSA methods use the Neumann-Ulam class of Monte
+  Carlo solvers for linear systems to accelerate the fixed point
+  method on which the Neumann-Ulam solvers are based. To effectively
+  parallelize MCSA methods requires the parallelization of the
+  underlying Neumann-Ulam solvers. To do this in a domain decomposed
+  environment, we borrow strategies traditionally implemented in Monte
+  Carlo particle transport to parallelize the problem. The
+  multiple-set overlapping-domain decomposition algorithm is presented
+  along with parallel scaling data for the resulting implementation
+  using the Titan Cray XK7 machine at Oak Ridge National Laboratory.}
 
@@
 through superposition.
+
+In this paper, we will present the Monte Carlo methods of Neumann and
+Ulam and then outline Monte Carlo Synthetic Acceleration
+methods. Next, we present a parallel algorithm for MCSA based on the
+work of Brunner and Brantley \cite{brunner_efficient_2009} and Wagner
+et. al. \cite{wagner_hybrid_2010}. We then perform a series of
+parallel scaling studies to characterize the parallel
+algorithm. Finally, we summarize the work and offer future work that
+may improve the algorithm.
 
@@
 supplies a correction for the initial guess generated by Richardson's
-iteration. Rather, only a set number of random walk permuations are
+iteration. Rather, only a set number of random walk permutations are
 required using the Neumann-Ulam method to generate the correction. In
@@
 up the tree in the same manner.
-
 \begin{figure}[h!]
@@
 
+To demonstrate the algorithm, consider the example adapted from
+Mervin's work with Wagner and others in the same area
+\cite{mervin_variance_2012} and presented in
+Figure~\ref{fig:msod_example}.
+\begin{figure}[h!]
+  \begin{center}
+    \scalebox{0.8}{ \input{msod_example.pdftex_t} }
+  \end{center}
+  \caption{\textbf{Overlapping domain example illustrating how domain
+      overlap can reduce communication costs.}
+    \textit{All particles start in the blue region of interest. The
+      dashed line represents 0.5 domain overlap between domains.}}
+  \label{fig:msod_example}
+\end{figure}
+In this example, 3 particle histories are presented emanating from the
+blue region of interest. Starting with particle A, if no domain
+overlap is used then the only the blue domain exists on the starting
+processor. Particle A is then transported through 3 other domains
+before the history ends, therefore requiring three communications to
+occur in Brunner's algorithm. If a 0.5 domain overlap is permitted as
+shown by the dashed line, then the starting process owns enough of the
+domain such that no communications must occur in order to complete the
+particle A transport process. Using 0.5 domain overlap also easily
+eliminates cases such as that represented by the path of particle
+C. In this case, particle C is scattering between two adjacent
+domains, incurring a large latency cost for a single
+particle. Finally, with particle B we observe that 0.5 domain overlap
+will still not eliminate all communications. However, if 1 domain
+overlap were used, the entire geometry shown in
+Figure~\ref{fig:msod_example} would be contained on the source
+processor and therefore transport of all 3 particles without
+communication would occur.
+
 \subsection{Parallel MCSA Algorithm}
 We utilize parallel matrix-vector multiply, vector update, and vector
-norm operations as oultined in \cite{saad_iterative_2003} for a
+norm operations as outlined in \cite{saad_iterative_2003} for a
 parallel MCSA implementation with these computations required in
@@
 for wall time and those average values used for the efficiency
-computations. For all cases, both strong and weak scaling studies were
+computations.
+
+For all cases, both strong and weak scaling studies were
 performed. For strong scaling, the objective is to then decrease the
@@
 $P$. Based on this, we can define the strong scaling \textit{absolute
-  efficiency} as:
+  efficiency} as \cite{keyes_how_1999}:
 \begin{equation}
@@
 \end{equation}
-For weak scaling, For strong scaling, the objective is to then decrease the runtime of a
-fixed size global problem as a linear function of $P$. Based on this,
-we can define the strong scaling \textit{absolute efficiency} as:
-\begin{equation}
-  \eta_{strong}(N,P) = \frac{1}{P} \frac{T(N,1)}{T(N,P)}\:,
-  \label{eq:strong_scaling_absolute}
-\end{equation}
-where $T(N,P)$ is the wall time for a computation of global size $N$
-using $P$ processes. Note here that this particular definition is
-measured with respect to a serial computation. On leadership class
-machines, it is often the case that a problem size large enough to
-effectively execute on $O(10,000)$ cores or larger is significantly
-larger than any problem that may be solved by a single core due to
-memory restrictions. Therefore, we consider absolute scaling for a
-base case where $Q$ processes are used instead of a serial
-computation:
-\begin{equation}
-  \eta_{strong}(N,P|Q) = \frac{Q}{P} \frac{T(N,Q)}{T(N,P)}\:.
-  \label{eq:strong_scaling_absolute_ref}
-\end{equation}
-All parallel scaling studies presented in this work were performed on
-the Titan Cray XK7 machine at the Oak Ridge Leadership Computing
-Facility located at Oak Ridge National Laboratory.
+For weak scaling, the ratio $N/P$ is fixed such that the local problem
+size remains the same for all values of $P$. In this case, the
+objective is to maintain a constant wall time while both $N$ and $P$
+grow as a fixed ratio. In this case, we define the absolute efficiency
+as \cite{keyes_how_1999}:
+\begin{equation}
+  \eta_{weak}(M|N,P|Q) = \frac{T(M,Q)}{T(N,P)}\:,
+  \label{eq:weak_scaling_absolute}
+\end{equation}
+subject to the constraint $M/Q = N/P$. The scaling is perfect if the
+runtime is static for all $P$ when $N/P$ is fixed.
+
+The absolute efficiency measures presented here only considers the
+total time to solution. Often, that total time to solution may be
+changing for an iterative method due to the fact that the number of
+iterations required to converge may be changing as a function of
+$P$. This will often happen in cases where modifying $N$ changes the
+spectral character of the problem, often making it stiffer and
+therefore more difficult to solve when $N$ increases. Therefore we
+also consider \textit{algorithmic efficiency} which considers this
+potential change in iteration count \cite{keyes_how_1999}:
+\begin{equation}
+  \eta_{alg}(N,P|Q) = \frac{I(N,Q)}{I(N,P)}\:,
+  \label{eq:algorithmic_efficiency}
+\end{equation}
+where $I(N,P)$ is the number of iterations required to converge a
+problem of global size $N$ computed using $P$ processes.  All parallel
+scaling studies presented in this work were performed on the Titan
+Cray XK7 machine at the Oak Ridge Leadership Computing Facility
+located at Oak Ridge National Laboratory.
 
@@
 \label{subsec:overlapping_domain_decomp}
-
 In the next set of scaling studies, we will explore how adding overlap
@@
 efficiencies at 2,048 cores, little evidence of strong scaling
-improvement is evident for all levels of overlap tested. To further
-explore any potential benefits of adding overlap into the system,
-Figure~\ref{fig:titan_strong_overlap_diff} gives the difference in the
-efficiencies between all overlap cases and the reference case without
-overlap at each core count. At smaller numbers of cores, there is no
-observed benefit to using overlap to improve strong scaling. The
-improvements noted at 2,048 cores are also present here and account
-for the best efficiency improvements for all core counts tested. At
-16,384 cores, using 2 levels of overlap did give a 6.4\% boost in
-efficiency while at 1,024 cores this same amount of overlap actually
-reduced the efficiency. Beyond 2 levels of overlap, efficiency levels
-begin to fall with 5 levels of overlap performing the worst for all
-cases. From this, we don't expect adding any more overlap will improve
-the scalability of the algorithm and the best strong scaling
-performance was observed when the number of overlap levels is
-approximately equal to the root-mean-squared number of states we
-expect a history to move from its starting point as computed by
-Eq~(\ref{eq:discrete_distance}).
+improvement is evident for all levels of overlap tested.
+
+To further explore any potential benefits of adding overlap into the
+system, Figure~\ref{fig:titan_strong_overlap_diff} gives the
+difference in the efficiencies between all overlap cases and the
+reference case without overlap at each core count. At smaller numbers
+of cores, there is no observed benefit to using overlap to improve
+strong scaling. The improvements noted at 2,048 cores are also present
+here and account for the best efficiency improvements for all core
+counts tested. At 16,384 cores, using 2 levels of overlap did give a
+6.4\% boost in efficiency while at 1,024 cores this same amount of
+overlap actually reduced the efficiency. Beyond 2 levels of overlap,
+efficiency levels begin to fall with 5 levels of overlap performing
+the worst for all cases. From this, we don't expect adding any more
+overlap will improve the scalability of the algorithm and the best
+strong scaling performance was observed when the number of overlap
+levels is approximately equal to the root-mean-squared number of
+states we expect a history to move from its starting point as computed
+by Eq~(\ref{eq:discrete_distance}).
 
@@
 \label{subsec:ms_decomposition}
-
 We next look to apply replication to the problem in the form of
@@
 \label{subsubsec:ms_strong}
-
 For the strong scaling case, recall that the objective is to decrease
@@
 \label{subsubsec:ms_weak}
-
 Next, we continue to a weak scaling study of the multiple sets
@@
 given $P$ and $N$. When that problem is now solved using 2 sets we
-have $S=2$, $N$, and $2P_s$ where $P_s$ is the number of processors in
+have $S=2$, $N$, and $P=2P_s$ where $P_s$ is the number of processors in
 a single set. For any number of sets we then have a global problem
@@
 scaling exercises and 131,072 cores used in weak scaling exercises
-using the neutron diffusion problem.
+using a neutron diffusion problem.
 
@@
 
+For future work, overall parallel scaling results indicated that the
+implementation was of good quality in both the strong and weak scaling
+cases. However, for the multiple set cases in the weak scaling
+analysis, it was noted that there was a growing parallel overhead as a
+function of core count. This was counter-intuitive in that the size of
+the parallel group over which the set reductions were occurring was
+fixed at all core counts and therefore we do not expect the addition
+of such an operation to increase parallel run times. It did, however,
+and therefore future research should explore why the implementation
+for multiple sets generated parallel overhead that increased with core
+count instead of remaining constant.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2013-09-06 doc/sna2013/paper/slattery_evans_wilson.tex
fixing relative weight cutoff definition in dissertation

@@
   strategy for parallelizing the Monte Carlo Synthetic Acceleration
-  (MCSA) method. MCSA methods use the Neumann-Ulam class of Monte
-  Carlo solvers for linear systems to accelerate the fixed point
-  method on which the Neumann-Ulam solvers are based. To effectively
-  parallelize MCSA methods requires the parallelization of the
+  method. Monte Carlo Synthetic Acceleration methods use the
+  Neumann-Ulam class of Monte Carlo solvers for linear systems to
+  accelerate a fixed-point iteration sequence. Effective parallel
+  algorithms for these methods require the parallelization of the
   underlying Neumann-Ulam solvers. To do this in a domain decomposed
   environment, we borrow strategies traditionally implemented in Monte
-  Carlo particle transport to parallelize the problem. The
-  multiple-set overlapping-domain decomposition algorithm is presented
-  along with parallel scaling data for the resulting implementation
-  using the Titan Cray XK7 machine at Oak Ridge National Laboratory.}
+  Carlo particle transport to parallelize the problem. The parallel
+  Neumann-Ulam and multiple-set overlapping-domain decomposition
+  algorithm is presented along with parallel scaling data for the
+  resulting implementation using the Titan Cray XK7 machine at Oak
+  Ridge National Laboratory.}
 
@@
 operator need not be symmetric or positive-definite, (2) the
-stochastic nature of the solution method provides a natural solution
-to some aspects of the issue of resiliency, and (3) is amenable to
-parallelization using modern methods developed by the transport
-community \cite{wagner_hybrid_2010}. The development a parallel MCSA
-algorithm using techniques is the primary contribution of this work.
+stochastic nature of the solution method potentially provides a
+natural solution to some aspects of the issue of resiliency, and (3)
+is amenable to parallelization using modern methods developed by the
+transport community \cite{wagner_hybrid_2010}. The development of a
+parallel MCSA algorithm using these techniques is the primary
+contribution of this work.
 
@@
 soft failures such as bit errors in floating point operations and hard
-failures where the data owned by that processor cannot be
-recovered. Because these failures are predicted to be common,
-resilient solver technologies are required to overcome these events at
-the application level. With linear solvers based on Monte Carlo
-techniques, such issues are potentially alleviated by statistical
-arguments. In the case of soft failures, isolated floating point
-errors in the Monte Carlo simulation are absorbed within tally
-statistics. Completely losing memory during a hard failure is treated
-as a high variance event where some portion of the Monte Carlo
-histories and subsequently the solution are lost with other histories
-maintained by replicating the problem and combining the results
-through superposition.
+failures where data cannot be recovered. Because these failures are
+predicted to be common, resilient solver technologies are required to
+overcome these events at the application level. With linear solvers
+based on Monte Carlo techniques, such issues are potentially
+alleviated by statistical arguments. In the case of soft failures,
+isolated floating point errors in the Monte Carlo simulation are
+absorbed within tally statistics. Completely losing memory during a
+hard failure is treated as a high variance event where some portion of
+the Monte Carlo histories and subsequently the solution are lost with
+other histories maintained by replicating the problem and combining
+the results through superposition.
 
 In this paper, we will present the Monte Carlo methods of Neumann and
-Ulam and then outline Monte Carlo Synthetic Acceleration
-methods. Next, we present a parallel algorithm for MCSA based on the
-work of Brunner and Brantley \cite{brunner_efficient_2009} and Wagner
-et. al. \cite{wagner_hybrid_2010}. We then perform a series of
-parallel scaling studies to characterize the parallel
-algorithm. Finally, we summarize the work and offer future work that
-may improve the algorithm.
+Ulam and then outline Monte Carlo Synthetic Acceleration methods. We
+then present a parallel algorithm for MCSA based on the work of
+Brunner and Brantley \cite{brunner_efficient_2009} and Wagner
+et. al. \cite{wagner_hybrid_2010}. Using this algorithm, we perform a
+series of scaling studies to characterize parallel
+performance. Finally, we summarize the results and offer future work
+that may improve the algorithm.
 
@@
 \section{Monte Carlo Methods}
-Monte Carlo Synthetic Acceleration (MCSA) methods rely on the
-Neumann-Ulam class of linear solvers to accelerate a stationary
-method. These Monte Carlo methods work by sampling a distribution with
-an expectation value equivalent to that of the inverted
-operator. Neumann and Ulam's work was first published in 1950 by
-Forsythe and Leibler \cite{forsythe_matrix_1950}. In the following
-decade, the books of Hammersley and Handscomb
-\cite{hammersley_monte_1964} and Spanier and Gelbard
-\cite{spanier_monte_1969} present additional detail on these solvers.
-In this section we briefly present the Neumann-Ulam Monte Carlo method
-for solving linear systems and then present the Monte Carlo Synthetic
-Acceleration method.
+Monte Carlo Synthetic Acceleration methods rely on the Neumann-Ulam
+class of linear solvers to accelerate a stationary method. These Monte
+Carlo methods work by sampling a distribution with an expectation
+value equivalent to that of the inverted operator. Neumann and Ulam's
+work on these solvers was first published in 1950 by Forsythe and
+Leibler \cite{forsythe_matrix_1950}. In the following decade, the
+books of Hammersley and Handscomb \cite{hammersley_monte_1964} and
+Spanier and Gelbard \cite{spanier_monte_1969} present additional
+detail on the methods. In this section we briefly present the
+Neumann-Ulam Monte Carlo method for solving linear systems and then
+present Monte Carlo Synthetic Acceleration. We refer the reader to
+\cite{evans_monte_2009} and \cite{evans_monte_2012} for a more
+detailed explanation and analysis of MCSA and the Neumann-Ulam
+methods.
 
@@
 such that we expect to select a new state, $j$, from the current state
-in the random walk, $i$, by sampling by sampling the probability
-distribution function formed by the $i^{th}$ of the probability
-matrix. The transition weight is then defined as:
+in the random walk, $i$, by sampling the probability distribution
+function formed by the $i^{th}$ row of the probability matrix. The
+transition weight is then defined as:
 \begin{equation}
@@
 The initial state $i_0$ of the random walk is determined by sampling
-the source vector $\mathbf{b}$ with probabilities:
-\begin{equation}
-  P_{(i_0=i)}(\nu) = \frac{|b_i|}{||\mathbf{b}||_1}\:,
-  \label{eq:adjoint_source_probability}
-\end{equation}
-with a random walk starting weight of:
-\begin{equation}
-  W_0 = ||\mathbf{b}||_1 \frac{b_i}{|b_i|}\:.
-  \label{eq:adjoint_starting_weight}
+the source vector $\mathbf{b}$ with initial probabilities and weights:
+\begin{equation}
+  P_{(i_0=i)}(\nu) = \frac{|b_i|}{||\mathbf{b}||_1},\ \ \ W_0 =
+  ||\mathbf{b}||_1 \frac{b_i}{|b_i|}\:.
+  \label{eq:adjoint_starting_params}
 \end{equation}
@@
 \end{equation}
+Finally, we set a criteria for random walk termination by setting a
+\textit{relative weight cutoff}:
+\begin{equation}
+  W_f = W_c W_0\:,
+  \label{eq:relative_weight_cutoff}
+\end{equation}
+where $W_c$ is a user defined cutoff. The random walk will then be
+terminated after $m$ steps if $W_m < W_f$.
 
@@
 radiation diffusion equation \cite{evans_monte_2012}. With their new
-methods, they achieved identical numerical results as and marginally
-better performance than conventional subspace methods. Their approach
-was instead to use residual Monte Carlo as a synthetic acceleration
-for a stationary method.
-
-The \textit{Fixed-Point Monte Carlo Synthetic Acceleration} (MCSA)
-method is defined as:
+methods, they achieved identical numerical results when compared to
+conventional subspace methods and observed better performance with
+respect to time to solution. Their approach was instead to use
+residual Monte Carlo as a synthetic acceleration for a stationary
+method.
+
+The \textit{Fixed-Point Monte Carlo Synthetic Acceleration} method is
+defined as:
 \begin{subequations}
@@
 index. Here, a Neumann-Ulam Monte Carlo method is used to generate the
-solution correction in Eq~\ref{eq:mcsa_4} from the residual and
-Richardson's iteration in the first step has been rewritten as a
-residual correction. Using Monte Carlo in this way achieves the same
-effect as Halton's method, decoupling its convergence rate from the
-overall convergence rate of the method. Here, the approximate Monte
-Carlo solution is not driven to a particular convergence as it merely
-supplies a correction for the initial guess generated by Richardson's
-iteration. Rather, only a set number of random walk permutations are
-required using the Neumann-Ulam method to generate the correction. In
-addition to the Monte Carlo solver parameters dictating the number of
-histories and weight cutoff, the outer MCSA iterations also have the
-following stopping criteria:
+solution correction in Eq~(\ref{eq:mcsa_4}) using the residual as
+source and Richardson's iteration in Eq~(\ref{eq:mcsa_2}) has been
+rewritten as a residual correction. Using Monte Carlo in this way
+achieves the same effect as Halton's method, decoupling its
+convergence rate from the overall convergence rate of the method. For
+the Neumann-Ulam method, a set number of random walk permutations is
+specified to generate the correction at each MCSA iteration. Adding
+more histories to the Neumann-Ulam sequence reduces the statistical
+uncertainty in $\delta\mathbf{x}$ and further accelerates the
+fixed-point iteration. In addition to the Monte Carlo solver
+parameters dictating the number of histories and weight cutoff, the
+outer MCSA iterations also have the following stopping criteria:
 \begin{equation}
@@
 \end{equation}
-where $\epsilon$ is a user-defined parameter. We refer the reader to
-\cite{evans_monte_2012} and \cite{evans_monte_2009} for a more
-detailed explanation and analysis of MCSA and the Neumann-Ulam
-methods.
+where $\epsilon$ is a user-defined parameter.
 
@@
 do this, we utilize the set of analytic relations developed in
-Appendix~\ref{chap:parallel_theory} to design these numerical
+\cite{slattery_spectral_2013} to design these numerical
 experiments. For the neutron diffusion problem used in these scaling

### 2013-09-06 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
adding parallel efficiency measures to sna paper
adding sna corrections through section 3

@@
 containing the primary set\footnote{For applications where resiliency
-  is potentially of concern, the operations in Eqs~(\ref{eq:mcsa_r1}),
-  (\ref{eq:mcsa_r2}), (\ref{eq:mcsa_r4}), and (\ref{eq:mcsa_r5}) may
-  be replicated without significant penalty. Doing so my alleviate
-  issues where any part of the linear problem or residual is disturbed
-  as now there will be multiple copies of these data
-  structures.}. Only the Monte Carlo data structures exist on the rest
-of the processors in the problem as replication of the other
-operations does not generate any more information that could be used
-to accelerate the time to solution.
+  is potentially of concern, the linear problem and subsequently the
+  operations in Eqs~(\ref{eq:mcsa_r1}), (\ref{eq:mcsa_r2}),
+  (\ref{eq:mcsa_r4}), and (\ref{eq:mcsa_r5}) may be replicated with
+  the penalty of broadcasting the data. Doing so may alleviate issues
+  where any part of the linear problem or residual is disturbed as now
+  there will be multiple copies of these data structures.}. Only the
+Monte Carlo data structures exist on the rest of the processors in the
+problem as replication of the other operations does not generate any
+more information that could be used to accelerate the time to
+solution.
 

### 2013-09-06 doc/sna2013/paper/slattery_evans_wilson.tex
adding sna corrections through section 3

@@
 \end{equation}
-where $W_c$ is a user defined cutoff. The random walk will then be
-terminated after $m$ steps if $W_m < W_f$.
+where $W_c$ is a user-defined cutoff value. The random walk will then
+be terminated after $m$ steps if $W_m < W_f$.
 
@@
 \end{equation}
-where $\epsilon$ is a user-defined parameter.
+where $\epsilon$ is a user-defined tolerance.
 
@@
 \section{Parallel Algorithms}
-In this section devise a parallel algorithm for the Neumann-Ulam Monte
-Carlo method based on the multiple-set overlapping-domain
+In this section we devise a parallel algorithm for the Neumann-Ulam
+Monte Carlo method based on the multiple-set overlapping-domain
 decomposition algorithm and a parallel algorithm for the MCSA
@@
 decomposed Monte Carlo implementation, stochastic histories must be
-communicated between domains as the simulation progresses and they
-transition to locations outside of the local domain. In order to
-implement domain decomposition strategy for Monte Carlo solvers in a
-parallel environment, communication of these histories must occur
-between compute nodes owning neighboring pieces of the global domain.
-
-In the context of radiation transport, in 2009 Brunner and colleagues
+communicated between adjacent domains and the owning compute nodes as
+the simulation progresses and they transition to locations outside of
+the local domain.
+
+In the context of radiation transport, in 2009 Brunner and Brantley
 provided a fully asynchronous domain decomposed parallel algorithm as
@@
 exclusivity permits all local operations to occur independently of
-other processes in the system and their current state to achieve an
-optimal overlap of work and communication. We will adapt their
-algorithm and directly apply it to construct a parallel formulation of
-the Neumann-Ulam method. Direct analogs can be derived from their
-works by noting that the primary difference between solving a linear
-transport system with Monte Carlo methods and traditional fixed source
-Monte Carlo transport problems is the content of the Markov chains
-that are generated.
-
-Consider this communication pattern for the 9 subdomain example given
+other processes in the system to achieve an optimal overlap of work
+and communication. We will adapt their algorithm and directly apply it
+to construct a parallel formulation of the Neumann-Ulam method. Direct
+analogs can be derived from their work by noting that the primary
+difference between solving a linear system with a Neumann-Ulam method
+and traditional fixed source Monte Carlo transport problems is the
+content of the Markov chains that are generated.
+
+Consider their communication pattern for the 9 subdomain example given
 by Figure~\ref{fig:nearest_neighbor_comm}. In this pattern, each
 boundary domain has two neighbors and the center domain four neighbors
-with which they will communicate parallel histories and this
-communication goes both ways as represented by the adjacent arrows in
-the figure. For each set of neighbors, a non-blocking send and receive
-is required with data buffers allocated with a user-defined size for
-incoming and outgoing histories. This non-blocking structure is
-critical to the performance of the algorithm in that it permits local
-history transport to continue while new histories to transport are
-being collected in the buffers. When a given process is ready to do
-more work, it can check these data buffers for incoming histories
-requiring further transport. In this way there is a maximum amount of
-overlap between communication and transport of histories.
+with which they will communicate parallel histories. For each set of
+neighbors, a non-blocking send and receive is initiated and data
+buffers allocated with a user-defined size for incoming and outgoing
+histories. This non-blocking structure is critical to the performance
+of the algorithm in that it permits local history transport to
+continue while new histories to transport are being collected in the
+buffers. When a given process is ready to do more work, it can check
+these data buffers for incoming histories requiring further
+transport.
 
@@
 
-Once the nearest-neighbor communication sequence has been completed
-for all histories, the completion of transport sequence is
-required. To stop transport, an asynchronous binary communication tree
-as presented for the same 9 subdomain example in
+Given the mutual exclusivity of the transport algorithm, an additional
+communcation sequence is required to efficiently stop the algorithm
+after all histories have been computed. To stop transport, Brunner and
+Brantley use an asynchronous binary communication tree as presented
+for the same 9 subdomain example in
 Figure~\ref{fig:binary_comm_tree}. In this communication pattern, each
-process has one parent process (except for the MASTER process 0) to
-which it will non-blocking send the number of histories that
-terminated their transport procedure by weight cutoff in the local
+process has one parent process (except for process 0) to which it will
+non-blocking send the number of histories that terminated their
+transport procedure by weight cutoff in the local
 domain. Equivalently, each process has up to two child processes from
@@
 the completed history tally be updated incrementally and funneled to
-the root process. Because we are solving fixed source problems with
-the Neumann-Ulam algorithm without any type of variance reduction that
-may generate more histories that we started with, once the root
-process completion sum tallies to the number of histories in the
-source, transport is complete. Once this occurs, the root process
-non-blocking sends the stop message to its children and each process
-non-blocking receives a stop message from its parent as shown in
-Figure~\ref{fig:binary_comm_tree}. The stop message is then propagated
-up the tree in the same manner.
+the root process in a mutually exclusive manner. Once the root process
+has tallied a completed number of histories equal to the number of
+histories in the source, transport is complete. Once this occurs, the
+root process non-blocking sends the stop message to its children and
+each child process non-blocking receives a stop message from the root
+as shown in Figure~\ref{fig:binary_comm_tree}. The stop message is
+then propagated up the tree in the same manner between parents and
+children.
 \begin{figure}[h!]
@@
 number of particles leaving the local domain. In addition, Wagner
-utilized a level of replication of the domain such that the domain was
-only decomposed on $O(100)$ processors and if replicated $O(1,000)$
-times achieves simulation on $O(100,000)$ processors, thus providing
-spatial and particle parallelism. Each collection of processors that
+utilized a level of replication such that the decomposition was only
+generated on $O(100)$ processors and if replicated $O(1,000)$ times
+achieves simulation on $O(100,000)$ processors, thus providing spatial
+and particle parallelism. Each collection of processors that
 constitutes a representation of the entire domain is referred to as a
-set, and within a set overlap occurs among its sub-domains. The
-original motivation was to decompose the domain in a way that it
-remained in a physical cabinet in a large distributed machine, thus
-reducing latency costs during communication.
-
-To demonstrate the algorithm, consider the example adapted from
-Mervin's work with Wagner and others in the same area
-\cite{mervin_variance_2012} and presented in
-Figure~\ref{fig:msod_example}.
+set, and within a set overlap occurs among its sub-domains.
+
+To demonstrate the algorithm, consider the example presented in
+Figure~\ref{fig:msod_example} adapted from Mervin's work with Wagner
+and others in the same area \cite{mervin_variance_2012}.
 \begin{figure}[h!]
@@
 parallel MCSA implementation with these computations required in
-Eqs~(\ref{eq:mcsa_1}), (\ref{eq:mcsa_2}), (\ref{eq:mcsa_4}), and
-(\ref{eq:mcsa_5}). In the first step, a parallel vector update is used
-to apply the residual to the previous iterate's solution in a
-Richardson iteration. In the next step and Eq~(\ref{eq:mcsa_5}), the
+Eqs~(\ref{eq:mcsa_1}), (\ref{eq:mcsa_2}), (\ref{eq:mcsa_3}), and
+(\ref{eq:mcsa_4}). In Eqs~(\ref{eq:mcsa_1}) and (\ref{eq:mcsa_3}), the
 residual is computed by a parallel matrix-vector multiply and vector
-update. Once the correction is computed with a parallel adjoint
+update. In Eq~(\ref{eq:mcsa_2}) a parallel vector update is used to
+apply the residual to the previous iterate's solution in a Richardson
+iteration. Once the correction is computed with a parallel adjoint
 Neumann-Ulam solve, this correction is applied to the solution with a
-parallel vector update. Additionally, as given by
-Eq~(\ref{eq:mcsa_stopping_criteria}), 2 parallel vector reductions
-will be required to check the stopping criteria: one initially to
-compute the infinity norm of the source vector, and another at every
-iteration to compute the infinity norm of the residual vector.
-
-To parallelize the solution of Eq~(\ref{eq:mcsa_3}), we apply the
-MSOD Neumann-Ulam algorithm presented in the previous section. The
-problem is potentially replicated with overlap and the Monte Carlo
-procedure returns a correction vector, $\delta \mathbf{x}$, with the same
-parallel decomposition as the input linear problem. As was the case
-for generating the replications required for a multiple set
+parallel vector update in Eq~(\ref{eq:mcsa_5}). Additionally, as given
+by Eq~(\ref{eq:mcsa_stopping_criteria}), 2 parallel vector norm
+operations will be required to check the stopping criteria: one
+initially to compute the infinity norm of the source vector, and
+another at every iteration to compute the infinity norm of the
+residual vector.
+
+To parallelize the solution of Eq~(\ref{eq:mcsa_4}), we apply the MSOD
+Neumann-Ulam algorithm presented in the previous section. The problem
+potentially has both replication and overlap and the Monte Carlo
+procedure returns the correction vector, $\delta \mathbf{x}$, with the
+same parallel decomposition as the input linear problem. Although
+replication of the Monte Carlo problem is required for a multiple set
 implementation, the explicit matrix and vector operations required for
 the rest of an MCSA iteration occur only on the subset of processors
-containing the primary set\footnote{For applications where resiliency
-  is potentially of concern, the operations in Eqs~(\ref{eq:mcsa_1}),
-  (\ref{eq:mcsa_2}), (\ref{eq:mcsa_4}), and (\ref{eq:mcsa_5}) may
-  be replicated without significant penalty. Doing so my alleviate
-  issues where any part of the linear problem or residual is disturbed
-  as now there will be multiple copies of these data
+containing the description of the linear problem\footnote{For
+  applications where resiliency is potentially of concern, the linear
+  problem and subsequently the operations in Eqs~(\ref{eq:mcsa_1}),
+  (\ref{eq:mcsa_2}), (\ref{eq:mcsa_4}), and (\ref{eq:mcsa_5}) may be
+  replicated with the penalty of broadcasting the data. Doing so may
+  alleviate issues where any part of the linear problem or residual is
+  disturbed as now there will be multiple copies of these data
   structures.}. Only the Monte Carlo data structures exist on the rest
@@
 \section{Parallel Scaling Studies}
-We will next measure performance in various scaling studies using an
-implementation of the parallel MCSA algorithm to assess its quality
-using a simple two-dimensional neutron diffusion problem in a
-homogeneous medium. For the scaling results presented here, all
-parallel MCSA results were computed using the Monte Carlo Linear
-Solvers\footnote{MCLS is an open source C++ library built on the
-  Trilinos\cite{heroux_overview_2005} framework and can be downloaded
-  at \url{https://github.com/sslattery/MCLS}} (MCLS) library generated
+We will next measure performance of the parallel MCSA algorithm in
+various scaling studies to assess its quality using a simple
+two-dimensional neutron diffusion problem in a homogeneous medium. For
+the scaling results presented here, all parallel MCSA results were
+computed using the Monte Carlo Linear Solvers\footnote{MCLS is an open
+  source C++ library built on the Trilinos\cite{heroux_overview_2005}
+  framework and can be downloaded at
+  \url{https://github.com/sslattery/MCLS}} (MCLS) library generated
 specifically for this work. For every scaling study presented, the

### 2013-09-06 doc/sna2013/paper/slattery_evans_wilson.tex
fixing figure layout in sna paper

@@
 \usepackage{tmath}
-\usepackage[section]{placeins}
-
+\usepackage{float}
 \usepackage[breaklinks=true,colorlinks=true,linkcolor=black,citecolor=black]{hyperref}
@@
 hard failure is treated as a high variance event where some portion of
-the Monte Carlo histories and subsequently the solution are lost with
-other histories maintained by replicating the problem and combining
-the results through superposition.
-
-In this paper, we will present the Monte Carlo methods of Neumann and
-Ulam and then outline Monte Carlo Synthetic Acceleration methods. We
-then present a parallel algorithm for MCSA based on the work of
-Brunner and Brantley \cite{brunner_efficient_2009} and Wagner
-et. al. \cite{wagner_hybrid_2010}. Using this algorithm, we perform a
-series of scaling studies to characterize parallel
+the Monte Carlo histories and subsequently of the solution are lost
+with other histories maintained by replicating the problem and
+combining the results through superposition.
+
+In this paper, we will present the Monte Carlo method of Neumann and
+Ulam for solving linear systems and then outline Monte Carlo Synthetic
+Acceleration. We then present a parallel algorithm for MCSA based on
+the work of Brunner and Brantley \cite{brunner_efficient_2009} and
+Wagner et. al. \cite{wagner_hybrid_2010}. Using this algorithm, we
+perform a series of scaling studies to characterize parallel
 performance. Finally, we summarize the results and offer future work
@@
 \end{figure}
-
 Given the mutual exclusivity of the transport algorithm, an additional
@@
 and others in the same area \cite{mervin_variance_2012}.
-\begin{figure}[h!]
+\begin{figure}[H]
   \begin{center}
@@
   \end{center}
-  \caption{\textbf{Overlapping domain example illustrating how domain
-      overlap can reduce communication costs.}
-    \textit{All particles start in the blue region of interest. The
-      dashed line represents 0.5 domain overlap between domains.}}
+  \caption{Overlapping domain example illustrating how domain overlap
+    can reduce communication costs. \textit{All particles start in the
+      blue region of interest. The dashed line represents 0.5 domain
+      overlap between domains.}}
   \label{fig:msod_example}
@@
   framework and can be downloaded at
-  \url{https://github.com/sslattery/MCLS}} (MCLS) library generated
-specifically for this work. For every scaling study presented, the
-calculations were performed 3 times with the average values reported
-for wall time and those average values used for the efficiency
-computations.
+  \url{https://github.com/sslattery/MCLS}} (MCLS) library which
+contains implementations of the parallel Neumann-Ulam and MCSA
+algorithms. For every scaling study presented, the calculations were
+performed 3 times with the average values reported for wall time and
+those average values used for the efficiency computations. All
+parallel scaling studies presented in this work were performed on the
+Titan Cray XK7 machine at the Oak Ridge Leadership Computing Facility
+located at Oak Ridge National Laboratory.
 
 For all cases, both strong and weak scaling studies were
-performed. For strong scaling, the objective is to then decrease the
-runtime of a fixed size global problem as a linear function of
-$P$. Based on this, we can define the strong scaling \textit{absolute
-  efficiency} as \cite{keyes_how_1999}:
+performed. For strong scaling, the objective is to decrease the
+runtime of a fixed size global problem as a linear function of $P$,
+where $P$ is the number of processors being used. Based on this, we
+can define the strong scaling \textit{absolute efficiency} as
+\cite{keyes_how_1999}:
 \begin{equation}
@@
 using $P$ processes. Note here that this particular definition is
-measured with respect to a serial computation. On leadership class
-machines, it is often the case that a problem size large enough to
-effectively execute on $O(10,000)$ cores or larger is significantly
-larger than any problem that may be solved by a single core due to
-memory restrictions. Therefore, we consider absolute scaling for a
-base case where $Q$ processes are used instead of a serial
-computation:
+measured with respect to a serial computation. On leadership-class
+machines, it is often the case that we use a reference case that
+occupies an entire node in the system. Therefore, we consider absolute
+scaling for a reference case where $Q$ processes are used instead of a
+serial computation:
 \begin{equation}
@@
 
-The absolute efficiency measures presented here only considers the
+The absolute efficiency measures presented here only consider the
 total time to solution. Often, that total time to solution may be
@@
 iterations required to converge may be changing as a function of
-$P$. This will often happen in cases where modifying $N$ changes the
+$N$. This will often happen in cases where modifying $N$ changes the
 spectral character of the problem, often making it stiffer and
-therefore more difficult to solve when $N$ increases. Therefore we
-also consider \textit{algorithmic efficiency} which considers this
-potential change in iteration count \cite{keyes_how_1999}:
+therefore more difficult to solve when $N$ increases. Therefore, we
+also consider \textit{algorithmic efficiency} which quantifies this
+change in iteration count \cite{keyes_how_1999}:
 \begin{equation}
@@
 where $I(N,P)$ is the number of iterations required to converge a
-problem of global size $N$ computed using $P$ processes.  All parallel
-scaling studies presented in this work were performed on the Titan
-Cray XK7 machine at the Oak Ridge Leadership Computing Facility
-located at Oak Ridge National Laboratory.
+problem of global size $N$ computed using $P$ processes.
 
@@
 algorithm. We refer to this case as pure domain decomposition where no
-overlap or replication has been leveraged. Such a test will isolate
-specifically the performance of Brunner and Brantley's algorithm as
-applied to the Neumann-Ulam algorithm. For the strong scaling study,
-the global diffusion problem size was fixed at \sn{1.6}{7} DOFs for
-all solvers and the 16 core run used as the base
-case. Figure~\ref{fig:titan_pure_strong_time} gives the wall time per
-iteration for these calculations while
-Figure~\ref{fig:titan_pure_strong} gives the results of this study
-with the given absolute efficiencies computed using
-Eq~(\ref{eq:strong_scaling_absolute}). All MCSA calculations converged
-in 22 iterations.
-
-\begin{figure}[h!]
+overlap or replication has been leveraged. Such a test will
+specifically isolate the performance of Brunner and Brantley's
+algorithm as applied to the random walk sequence. For the strong
+scaling study, the global diffusion problem size was fixed at a
+relatively small \sn{1.6}{7} DOFs and the 16 core run used as the
+reference case. Figure~\ref{fig:titan_pure_strong_time} gives the wall
+time per iteration for these calculations while
+Figure~\ref{fig:titan_pure_strong} gives the absolute efficiencies
+computed using Eq~(\ref{eq:strong_scaling_absolute}). All MCSA
+calculations converged the diffusion problem in 22 iterations.
+
+\begin{figure}[H]
   \begin{center}
@@
 
-\begin{figure}[h!]
+\begin{figure}[H]
   \begin{center}
@@
     efficiency. \textit{Super-linear speed-up is from memory thrashing
-      in the base case.}}
+      in the reference case.}}
   \label{fig:titan_pure_strong}
@@
 
-There are several key features of
-Figure~\ref{fig:titan_pure_strong_time} and
-Figure~\ref{fig:titan_pure_strong} worth noting. First, up to around
-5,000 cores, all methods experience a super-linear improvement in
-parallel efficiencies. For all methods, this increase is not a
-function of the algorithmic efficiency as those computed using
-Eq~(\ref{eq:algorithmic_efficiency}) remained at unity, meaning that
-all computations converged in the same number of iterations relative
-to the base case. Instead, what is happening here and was observed by
-Brunner and Brantley in their work is that the large problem used to
-scale to O(10,000) cores was large enough that memory thrashing was
-occurring in the 16 core base case. As the number of cores in the
-strong scaling study was increased and the local problem size
-decreased, more of the local problem fit into the cache of the system,
-thus causing improved serial performance and therefore enhanced run
-times.
-
-Beyond around 5,000 cores, the global problem size selected for this
-strong scaling study causes the local problem size to shrink enough
-that parallel communication costs begin to overtake arithmetic costs
-on process. This takeover causes this effective strong scaling wall
-that is observed in all methods. Using any more cores to solve this
+There are several features of Figure~\ref{fig:titan_pure_strong_time}
+and Figure~\ref{fig:titan_pure_strong} worth noting. First, up to
+around 5,000 cores we observe a super-linear improvement in parallel
+efficiencies. This increase was also observed by Brunner and Brantley
+in their work and is a result of using a problem large enough to scale
+to O(10,000) cores from 16, generating a large enough memory footprint
+that memory thrashing was occurring in the 16 core reference case. As
+the number of cores in the strong scaling study was increased and the
+local problem size decreased, more of the local problem fit into the
+cache of the system, thus causing improved serial performance and
+therefore enhanced run times and efficiencies. Beyond around 5,000
+cores, the global problem size selected for this strong scaling study
+causes the local problem size to shrink enough that parallel
+communication costs begin to overtake serial work. This takeover
+causes the strong scaling wall that is observed in
+Figure~\ref{fig:titan_pure_strong}. Using any more cores to solve this
 particular global problem is not an efficient use of those parallel
-resources. Furthermore, the fact that we see very similar trends in
-efficiencies for all methods means that they are bound by the same
-parallel performance limitations in a strong scaling environment. This
-is important as we can expect the same qualitative scaling performance
-from MCSA if arithmetic optimization has been performed to improve the
-serial performance of the implementation. In addition, these results
-are valuable in that they show MCSA, when properly parallelized, has
-the ability to be competitive on leadership-class computing platforms
-with conventional subspace methods with good strong scaling observed
-in the pure domain decomposition case to O(10,000) cores.
+resources.
 
@@
 per iteration and Figure~\ref{fig:titan_weak_absolute} gives the weak
-scaling absolute efficiencies for this problem. Overall, weak scaling
-performance of the parallel MCSA algorithm is excellent for the pure
-domain decomposition case with absolute parallel efficiencies of over
-80\% observed at over 32,000 cores.
-
-\begin{figure}[p!]
+scaling absolute efficiencies. Overall, weak scaling performance of
+the parallel MCSA algorithm is excellent for the pure domain
+decomposition case with absolute parallel efficiencies of over 80\%
+observed at over 32,000 cores.
+
+\begin{figure}[h!]
   \begin{center}
@@
   \end{center}
-  \caption{\textbf{Pure domain decomposition weak scaling absolute
-      efficiency.} \textit{MCLS is an over order of magnitude slower
+  \caption{Pure domain decomposition weak scaling absolute
+    efficiency. \textit{MCLS is an over order of magnitude slower
       arithmetically, causing the higher efficiency. GMRES executes
@@
   \end{center}
-  \caption{\textbf{Strong scaling absolute efficiency with varying
-      levels of overlap.} \textit{Values of 0, 1, 2, 3, and 5 levels
-      of overlap were used as the problem had a root-mean-squared
-      number of states of 2.63 for each history.}}
+  \caption{Strong scaling absolute efficiency with varying levels of
+    overlap. \textit{Values of 0, 1, 2, 3, and 5 levels of overlap
+      were used as the problem had a root-mean-squared number of
+      states of 2.63 for each history.}}
   \label{fig:titan_strong_overlap}
@@
   \end{center}
-  \caption{\textbf{Strong scaling absolute efficiency difference with
-      respect to the 0 overlap base case.} \textit{Values of 1, 2, 3,
+  \caption{Strong scaling absolute efficiency difference with
+      respect to the 0 overlap reference case. \textit{Values of 1, 2, 3,
       and 5 levels of overlap are compared here with the absolute
@@
   \end{center}
-  \caption{\textbf{Weak scaling absolute efficiency with varying
-      levels of overlap.} \textit{Values of 0, 1, 2, 3, and 5 levels
-      of overlap were used as the problem had a root-mean-squared
-      number of states of 2.63 for each history.}}
+  \caption{Weak scaling absolute efficiency with varying levels of
+    overlap. \textit{Values of 0, 1, 2, 3, and 5 levels of overlap
+      were used as the problem had a root-mean-squared number of
+      states of 2.63 for each history.}}
   \label{fig:titan_weak_overlap}
@@
   \end{center}
-  \caption{\textbf{Weak scaling absolute efficiency difference with
-      respect to the 0 overlap base case.} \textit{Values of 1, 2, 3,
-      and 5 levels of overlap are compared here with the absolute
-      difference of the absolute efficiencies presented.}}
+  \caption{Weak scaling absolute efficiency difference with respect to
+    the 0 overlap reference case. \textit{Values of 1, 2, 3, and 5
+      levels of overlap are compared here with the absolute difference
+      of the absolute efficiencies presented.}}
   \label{fig:titan_weak_overlap_diff}
@@
 For the multiple sets problem, the single set strong scaling study
-from the pure domain decomposition base case was repeated for multiple
-sets with both splitting of histories across sets and full replication
-of histories across sets using both 2 and 4
+from the pure domain decomposition reference case was repeated for
+multiple sets with both splitting of histories across sets and full
+replication of histories across sets using both 2 and 4
 sets. Figure~\ref{fig:titan_strong_ms_time} gives the actual wall
@@
   \end{center}
-  \caption{\textbf{Wall time in seconds to solution for each case for
-      the strong scaling study with multiple sets.} \textit{Until the
-      strong scaling wall is reached at O(10,000) cores, the single
-      set case performs the best. For the multiple set cases, it is
-      faster to split histories among sets and maintain iterative
-      performance rather than replicate histories to improve iterative
+  \caption{Wall time in seconds to solution for each case for the
+    strong scaling study with multiple sets. \textit{Until the strong
+      scaling wall is reached at O(10,000) cores, the single set case
+      performs the best. For the multiple set cases, it is faster to
+      split histories among sets and maintain iterative performance
+      rather than replicate histories to improve iterative
       performance.}}
@@
   \end{center}
-  \caption{\textbf{Multiple set absolute parallel efficiency.}
-    \textit{Computed relative to the 16-core 1-set base case. Using a
-      single set is the best means by which to improve strong scaling
-      up to the strong scaling wall. At higher core counts past where
-      the single set case hits the strong scaling wall adding sets can
-      provide some improvements in efficiencies.}}
+  \caption{Multiple set absolute parallel efficiency.
+    \textit{Computed relative to the 16-core 1-set reference
+      case. Using a single set is the best means by which to improve
+      strong scaling up to the strong scaling wall. At higher core
+      counts past where the single set case hits the strong scaling
+      wall adding sets can provide some improvements in
+      efficiencies.}}
   \label{fig:titan_strong_ms_eff}
@@
 maintained the number of iterations required to converge relative to
-the single set base case. When histories were replicated, adding sets
-decreased the number of iterations required to converge in a nonlinear
-fashion. Therefore, increasing the number of histories per MCSA
-iteration improves algorithmic efficiency. Using these algorithmic
-efficiencies, the implementation efficiencies for the multiple set
-cases were computed and are given in
+the single set reference case. When histories were replicated, adding
+sets decreased the number of iterations required to converge in a
+nonlinear fashion. Therefore, increasing the number of histories per
+MCSA iteration improves algorithmic efficiency. Using these
+algorithmic efficiencies, the implementation efficiencies for the
+multiple set cases were computed and are given in
 Figure~\ref{fig:titan_strong_ms_impeff}. From these results, we see
@@
 
-\begin{table}[h!]
+\begin{table}[H]
   \begin{center}
@@
   \end{center}
-  \caption{\textbf{Strong scaling algorithmic efficiency using
-      multiple sets.} \textit{Replicating histories improves
-      algorithmic efficiency by converging in fewer iterations.}}
+  \caption{Strong scaling algorithmic efficiency using multiple
+    sets. \textit{Replicating histories improves algorithmic
+      efficiency by converging in fewer iterations.}}
   \label{tab:ms_strong_alg_eff}
@@
   \end{center}
-  \caption{\textbf{Multiple set strong scaling implementation
-      efficiency.} \textit{Computed relative to the 16-core 1-set base
-      case. Considering iterations required to converge, the
+  \caption{Multiple set strong scaling implementation
+    efficiency. \textit{Computed relative to the 16-core 1-set
+      reference case. Considering iterations required to converge, the
       performance of the split history cases is much better than the
@@
   \end{center}
-  \caption{\textbf{Wall time in seconds to solution for each case for
-      the weak scaling study with multiple sets.}}
+  \caption{Wall time in seconds to solution for each case for the weak
+    scaling study with multiple sets.}
   \label{fig:titan_weak_ms_time}
@@
 Eq~(\ref{eq:ms_weak_efficiency}). Although we do reduce the time to
-solution over the base cases for nearly all data points, the
+solution over the reference cases for nearly all data points, the
 observation here is that the reduction in time to solution is still
@@
 replicated histories, the run time of that calculation must be half of
-that for the single set case. This was not the observation for the
-run times in Figure~\ref{fig:titan_weak_ms_time} and therefore the
-single set case is still the most efficient use of computational
-resources in this scaling study.
+that for the single set case. This was not the observation for the run
+times in Figure~\ref{fig:titan_weak_ms_time} and therefore the single
+set case is still the most efficient use of computational resources in
+this scaling study.
 
@@
   \end{center}
-  \caption{\textbf{Weak scaling absolute efficiency relative to
-      16-core 1-set base case.}}
+  \caption{Weak scaling absolute efficiency relative to 16-core 1-set
+    reference case.}
   \label{fig:titan_weak_ms_eff}
@@
   \end{center}
-  \caption{\textbf{Weak scaling algorithmic efficiency using multiple
-      sets.} \textit{Replicating histories improves algorithmic
+  \caption{Weak scaling algorithmic efficiency using multiple
+    sets. \textit{Replicating histories improves algorithmic
       efficiency by converging in fewer iterations.}}
@@
   \end{center}
-  \caption{\textbf{Implementation parallel efficiency relative to 16-core
-      1-set base case.}}
+  \caption{Implementation parallel efficiency relative to 16-core
+    1-set reference case.}
   \label{fig:titan_weak_ms_impeff}

### 2013-09-08 doc/dissertation/document/chapters/conclusion/conclusion.tex
finished sna draft

@@
 goal of demonstrating the improved scalability on leadership-class
-hardware. It was found the MCSA can indeed be effectively parallelized
+hardware. It was found that MCSA can indeed be effectively parallelized
 using the multiple-set overlapping-domain decomposition algorithm

### 2013-09-08 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
finished sna draft

@@
       respect to the 0 overlap base case.} \textit{Values of 1, 2, 3,
-      and 5 levels of overlap are compared here with the absolute
-      difference of the absolute efficiencies presented.}}
+      and 5 levels of overlap are compared here with the difference of
+      the absolute efficiencies presented.}}
   \label{fig:titan_strong_overlap_diff}
@@
       respect to the 0 overlap base case.} \textit{Values of 1, 2, 3,
-      and 5 levels of overlap are compared here with the absolute
-      difference of the absolute efficiencies presented.}}
+      and 5 levels of overlap are compared here with the difference of
+      the absolute efficiencies presented.}}
   \label{fig:titan_weak_overlap_diff}

### 2013-09-08 doc/sna2013/paper/slattery_evans_wilson.tex
finished sna draft

@@
   Neumann-Ulam and multiple-set overlapping-domain decomposition
-  algorithm is presented along with parallel scaling data for the
+  algorithms are presented along with parallel scaling data for the
   resulting implementation using the Titan Cray XK7 machine at Oak
@@
 \end{equation}
-Finally, the expectation value using all permutations is:
+The expectation value using all permutations is:
 \begin{equation}
@@
 methods, they achieved identical numerical results when compared to
-conventional subspace methods and observed better performance with
-respect to time to solution. Their approach was instead to use
-residual Monte Carlo as a synthetic acceleration for a stationary
-method.
+conventional subspace methods (e.g. Krylov solvers) and observed
+better performance with respect to time to solution. Their approach
+was instead to use residual Monte Carlo as a synthetic acceleration
+for a stationary method.
 
@@
       its parents how many histories completed within its domain. When
-      the root process sums all complete histories, it forwards the
-      stop signal to its children which in turn forward the message up
-      the tree. The subdomains are indexed by an integer.}}
+      the root process history sum equals the total number of
+      histories, it forwards the stop signal to its children which in
+      turn forward the message up the tree. The subdomains are indexed
+      by an integer.}}
   \label{fig:binary_comm_tree}
@@
 communication would occur.
+
+To implement the MSOD algorithm, we physically copy all data
+structures required to implement the Neumann-Ulam algorithm to each of
+the sets in the problem. Within each set, the subdomains detect their
+neighboring subdomains and gather the number of adjacent states
+specified for overlap. For each of these overlap states, the
+probablities, weights, tallies, and other information required to
+implement the Monte Carlo sequence are gathered. When transport
+begins, within each individual set the Neumann-Ulam sequence is
+executed with parallelism provided Brunner and Brantely's algorithm as
+outlined in the previous section. Here, transport occurs both in the
+local domain and the overlap region before histories are
+communicated. Once transport in all sets is compelete, the tallies
+from all overlapping states are gathered to their original owning
+processors within each set and then all tallies are summed onto the
+set of processors occupied by the linear problem to construct the
+final tally vector.
 
@@
 \section{Parallel Scaling Studies}
-We will next measure performance of the parallel MCSA algorithm in
+We will next measure the performance of the parallel MCSA algorithm in
 various scaling studies to assess its quality using a simple
 two-dimensional neutron diffusion problem in a homogeneous medium. For
-the scaling results presented here, all parallel MCSA results were
+MCSA, a range of MSOD parameters will be used such that problems with
+multiple sets may be tested along with overlap. For all MCSA
+calculations, a single history was used for every degree of freedom in
+the problem to compute the correction vector. For all calculations,
+the flux was converged to a tolerance of
+$\epsilon = \sn{1}{-8}$. Table~\ref{tab:parallel_mcsa_parameters} gives the
+parameters for the diffusion problem.
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{lc}\hline\hline
+      \multicolumn{1}{l}{Parameter}& 
+      \multicolumn{1}{c}{Value}\\\hline
+      $dx$ & 0.1 \\
+      $dy$ & 0.1 \\
+      $\Sigma_a$ & 5.0 \\
+      $\Sigma_s$ & 1.0 \\
+      Boundary Type & Vacuum \\
+      MCSA Solver Tolerance & \sn{1}{-8} \\
+      Histories & 1 per DOF \\
+      Weight Cutoff & \sn{1}{-2} \\
+      %%
+      \hline\hline
+    \end{tabular}
+  \end{center}
+  \caption{\textbf{Parallel MCSA verification model problem parameters.}
+    \textit{The neutron diffusion equation in 2 dimensions is used for
+      the model problem.}}
+  \label{tab:parallel_mcsa_parameters}
+\end{table}
+For the scaling results presented here, all parallel MCSA results were
 computed using the Monte Carlo Linear Solvers\footnote{MCLS is an open
@@
 where $P$ is the number of processors being used. Based on this, we
-can define the strong scaling \textit{absolute efficiency} as
+define the strong scaling \textit{absolute efficiency} as
 \cite{keyes_how_1999}:
@@
 For the first set of scaling studies, we consider the parallel MCSA
-algorithm where MSOD has not been used in the parallel Neumann-Ulam
+algorithm where MSOD has not been used in the parallel Monte Carlo
 algorithm. We refer to this case as pure domain decomposition where no
@@
 specifically isolate the performance of Brunner and Brantley's
-algorithm as applied to the random walk sequence. For the strong
+algorithm as applied to Neumann and Ulam's solver. For the strong
 scaling study, the global diffusion problem size was fixed at a
@@
   \begin{center}
-    \includegraphics[width=3in]{titan_pure_strong_time.pdf}
+    \includegraphics[width=3.25in]{titan_pure_strong_time.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=3in]{titan_pure_strong.pdf}
+    \includegraphics[width=3.25in]{titan_pure_strong.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=3in]{titan_pure_weak_time.pdf}
+    \includegraphics[width=3.25in]{titan_pure_weak_time.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=3in]{titan_weak_absolute.pdf}
+    \includegraphics[width=3.25in]{titan_weak_absolute.pdf}
   \end{center}
   \caption{Pure domain decomposition weak scaling absolute
-    efficiency. \textit{MCLS is an over order of magnitude slower
-      arithmetically, causing the higher efficiency. GMRES executes
-      more operations than conjugate gradient, also creating a higher
-      efficiency.}}
+    efficiency.}
   \label{fig:titan_weak_absolute}
@@
 do this, we utilize the set of analytic relations developed in
-\cite{slattery_spectral_2013} to design these numerical
-experiments. For the neutron diffusion problem used in these scaling
-studies, the linear operator had a spectral radius of 0.787 and a
-weight cutoff of \sn{1}{-2} was used with the Neumann-Ulam
-solver. Using Eq~(\ref{eq:analytic_k}) gives an expected random walk
-length of 19.22 transitions for any given history in the
-problem. Using the mean-chord approximation given by
-Eq~(\ref{eq:mean_chord_domain_leakage}), we then expect approximately
-5.3\% of the histories born in a particular domain to leak out after
-the first transport stage. By adding overlap, we aim to reduce this
-fraction of communicated histories and therefore boost the parallel
-performance of the algorithm.
-
-To determine the amount of overlap to test for the scaling studies, we
-use a variation of Eq~(\ref{eq:step_k_length}) to determine how many
-discrete states a history will move on average from its birth site by
-modifying it to only account for discrete movements:
+\cite{slattery_spectral_2013} that describe the distances a history in
+a Neumann-Ulam random walk will travel for the neutron diffusion
+problem. To determine the amount of overlap to test for the scaling
+studies, we calculate how many discrete states a history will move on
+average from its birth site using:
 \begin{equation}
@@
 \end{equation}
-where $\sqrt{\bar{n^2_k}}$ is now the root-mean-squared number of
-discrete states a history will move from its birth site. For our
-diffusion problem used in these scaling studies, we compute a value of
-2.63 states per history for any of the dimensions of the system. This
-means that we can expect to end up on average 2.63 states in both the
-$i$ and $j$ dimensions away from the starting state. Therefore, we
-will use the overlap generation algorithm outlined in
-\S~\ref{subsec:msod_generation} such that we parametrically grow the
-local domain using 1, 2, 3, and 5 levels to measure how scalability is
-affected when this root-mean-squared number of states is
-considered. For example, using 3 levels of overlap should keep over
-50\% of the histories that would be communicated in the pure domain
-decomposition case on-process as this is greater than the
-root-mean-squared number of states we would expect a history to
+where $\sqrt{\bar{n^2_k}}$ is the root-mean-squared number of discrete
+states a history will move from its birth site, $k$ the average number
+of transitions a history will make in a random walk, and $n_s$ the
+average number of states a history will move per transition in a
+particular direction. Using \cite{slattery_spectral_2013} we compute
+$k=19.22$ as the average random walk length and $n_s = 3/5$ states per
+random walk step for the neutron diffusion problem giving
+$\sqrt{\bar{n^2_k}} = 2.63$. This means that for every random walk, we
+can expect the sequence to terminate on average 2.63 states in both
+the $i$ and $j$ dimensions away from the starting state. Therefore, we
+will parametrically grow the local domain using 1, 2, 3, and 5 levels
+to measure how scalability is affected when this root-mean-squared
+number of states is considered. For example, using 3 levels of overlap
+should keep over 50\% of the histories that would be communicated in
+the pure domain decomposition case on-process as this is greater than
+the root-mean-squared number of states we would expect a history to
 travel.
@@
 efficiencies at 2,048 cores, little evidence of strong scaling
-improvement is evident for all levels of overlap tested.
+improvement is observed for all levels of overlap tested.
 
@@
   \begin{center}
-    \includegraphics[width=3in]{titan_strong_overlap.pdf}
+    \includegraphics[width=3.25in]{titan_strong_overlap.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=3in]{titan_strong_overlap_diff.pdf}
-  \end{center}
-  \caption{Strong scaling absolute efficiency difference with
-      respect to the 0 overlap reference case. \textit{Values of 1, 2, 3,
-      and 5 levels of overlap are compared here with the absolute
-      difference of the absolute efficiencies presented.}}
+    \includegraphics[width=3.25in]{titan_strong_overlap_diff.pdf}
+  \end{center}
+  \caption{Strong scaling absolute efficiency difference with respect
+    to the 0 overlap reference case. \textit{Values of 1, 2, 3, and 5
+      levels of overlap are compared here with the difference of the
+      absolute efficiencies presented.}}
   \label{fig:titan_strong_overlap_diff}
@@
 efficiencies from the calculations without overlap given in
-Figure~\ref{fig:titan_weak_overlap_diff}. The results here are much
-the same in that there is not a dramatic improvement in scalability
-relative to the case without overlap. Opposite the strong scaling
-calculation, overlap enhances weak scaling efficiencies at lower core
-counts and degrades the efficiencies at higher core counts. In
-addition, 1 level of overlap provided the best enhancement of
-scalability when overlap was beneficial.
+Figure~\ref{fig:titan_weak_overlap_diff}. The results here are similar
+in that there is not a dramatic improvement in scalability relative to
+the case without overlap. Opposite the strong scaling calculation,
+overlap enhances weak scaling efficiencies at lower core counts and
+degrades the efficiencies at higher core counts. In addition, 1 level
+of overlap provided the best enhancement of scalability when overlap
+was beneficial.
 
@@
   \begin{center}
-    \includegraphics[width=3in]{titan_weak_overlap.pdf}
+    \includegraphics[width=3.25in]{titan_weak_overlap.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=3in]{titan_weak_overlap_diff.pdf}
+    \includegraphics[width=3.25in]{titan_weak_overlap_diff.pdf}
   \end{center}
@@
     the 0 overlap reference case. \textit{Values of 1, 2, 3, and 5
-      levels of overlap are compared here with the absolute difference
-      of the absolute efficiencies presented.}}
+      levels of overlap are compared here with the difference of the
+      absolute efficiencies presented.}}
   \label{fig:titan_weak_overlap_diff}
@@
 For both the strong and weak scaling studies, the parallel performance
-of the algorithm was either marginally improved or not improved at
-all. We would always expect an improvement in performance for using
-overlap to prevent the transport of histories from domain to domain
-during execution of the Neumann-Ulam algorithm. The reduction in
-transport means a reduction in parallel communication and therefore
-should never degrade the performance of the algorithm. It is true that
-doing so in fact enhances Monte Carlo performance. However, the
-Neumann-Ulam algorithm is not representative of the entire MCSA
-algorithm. When overlap is used, an overlapping tally vector is
-constructed in parallel such that multiple cores may own multiple
-states in the system. When this occurs, the tally data associated with
-these shared states must be communicated after the Neumann-Ulam solve
-as outlined in \S~\ref{subsec:msod_algorithm}. This reduction
-operation then builds the full MCSA correction vector in the
-decomposition of the solution vector.
+of the algorithm was either marginally improved or not improved at all
+when overlap was added. From the literature, we would always expect an
+improvement in performance when using overlap to prevent the transport
+of histories from domain to domain during execution of the
+Neumann-Ulam algorithm. The reduction in transport means a reduction
+in parallel communication and therefore should never degrade the
+performance of the algorithm. It is true that doing so in fact
+enhances Monte Carlo performance. However, the Neumann-Ulam algorithm
+is not representative of the entire MCSA algorithm. When overlap is
+used, an overlapping tally vector is constructed in parallel such that
+multiple cores may own multiple states in the system. When this
+occurs, the tally data associated with these shared states must be
+communicated after the Neumann-Ulam solve. This reduction operation
+then builds the full MCSA correction vector in the decomposition of
+the solution vector.
 
@@
 \label{subsec:ms_decomposition}
-We next look to apply replication to the problem in the form of
-multiple sets. As the previous analysis showed overlap to be of little
+We next apply replication to the problem in the form of multiple
+sets. As the previous analysis showed overlap to be of little
 assistance in improving scalability, we only apply multiple sets to
@@
 the fully replicated case. As an example of these cases, consider a
-single set problem where 20,000 histories are used in the Monte Carlo
-solve. If we choose the split case, then solving the same problem with
-2 sets would give 10,000 histories in each set and 4 sets would give
-5,000 histories in each set. Splitting the histories in this way
+single set problem where 20,000 histories are used in the Neumann-Ulam
+sequence. If we choose the split case, then solving the same problem
+with 2 sets would give 10,000 histories in each set and 4 sets would
+give 5,000 histories in each set. Splitting the histories in this way
 provides the same global number of histories used to compute the
@@
 the time to solution for a given global problem size $N$ by increasing
-$P$. Therefore, we may use Eqs~(\ref{eq:strong_scaling_absolute_ref}),
-(\ref{eq:algorithmic_efficiency}), and
-(\ref{eq:implementation_efficiency}) to compute the various strong
+$P$. Therefore, we may use Eqs~(\ref{eq:strong_scaling_absolute_ref})
+and (\ref{eq:algorithmic_efficiency}) to compute the various strong
 scaling efficiencies, but we must modify the reference computation in
@@
 any computational resources added to the problem should aim to result
-in speedup of this reference case. If this speedup for one technique
-of adding cores is less than another, then that technique is deemed a
-less efficient use of those computational resources.
+in speedup of this reference case.
 
@@
 number of histories in half reduces the time of an iteration by a
-factor that is nearly linear while increasing linearly the number of
+factor that is nearly linear while linearly increasing the number of
 histories through replication does not decrease the iterations
@@
   \begin{center}
-    \includegraphics[width=3in]{titan_strong_ms_time.pdf}
+    \includegraphics[width=3.25in]{titan_strong_ms_time.pdf}
   \end{center}
@@
 to combine the set results increases run times and reduces
-efficiencies. This reduction across blocks is even more obvious for
-the 4 set case. Ultimately, all cases effectively hit the strong
+efficiencies. The overhead from this reduction is even more obvious
+for the 4 set case. Ultimately, all cases effectively hit the strong
 scaling wall at the same time at around O(10,000) cores.
@@
 the local domain, giving a larger ratio of communication to work in
-the Monte Carlo solve. As with with single set results, we also
+the Neumann-Ulam solve. As with with single set results, we also
 observe some efficiencies greater than unity for the multiple set
@@
   \begin{center}
-    \includegraphics[width=3in]{titan_strong_ms_eff.pdf}
+    \includegraphics[width=3.25in]{titan_strong_ms_eff.pdf}
   \end{center}
@@
 sets decreased the number of iterations required to converge in a
-nonlinear fashion. Therefore, increasing the number of histories per
-MCSA iteration improves algorithmic efficiency. Using these
-algorithmic efficiencies, the implementation efficiencies for the
-multiple set cases were computed and are given in
-Figure~\ref{fig:titan_strong_ms_impeff}. From these results, we see
-that the replicated history cases are even less efficient than in the
-absolute case with respect to the performance of a single
-iteration. This is due to the fact that doubling the number of
+nonlinear fashion, thus improving algorithmic efficiency. The
+nonlinear relationship is due to the fact that doubling the number of
 histories does not halve the number of MCSA iterations required for
-convergence as the acceleration is bound by the Central Limit
-Theorem. As a result, the improved iterative performance from the
-replicated cases comes at a higher cost in terms of time to solution
-than the split cases.
+convergence as the accuracy of the acceleration provided by the
+Neumann-Ulam method is bound by the Central Limit Theorem. As a
+result, the improved iterative performance from the replicated cases
+comes at a higher cost in terms of time to solution than the split
+cases.
 
@@
 
-\begin{figure}[H]
-  \begin{center}
-    \includegraphics[width=3in]{titan_strong_ms_impeff.pdf}
-  \end{center}
-  \caption{Multiple set strong scaling implementation
-    efficiency. \textit{Computed relative to the 16-core 1-set
-      reference case. Considering iterations required to converge, the
-      performance of the split history cases is much better than the
-      cases with history replication.}}
-  \label{fig:titan_strong_ms_impeff}
-\end{figure}
-
 \subsubsection{Weak Scaling with Multiple Sets}
@@
 Next, we continue to a weak scaling study of the multiple sets
-problem. In the same way as the strong scaling study, we will modify
-the pure domain decomposition study to account for the additional
-replication. For each calculation in the single set study, that
-problem will be replicated either 2 or 4 times with both split and
-replicated histories to assess the effects on scalability and
-run time. Figure~\ref{fig:titan_weak_ms_time} gives the wall times from
-the multiple sets weak scaling study. Of primary importance here is
+problem. We will again modify the pure domain decomposition study to
+account for the additional replication. For each calculation in the
+single set study, the problem will be replicated with either 2 or 4
+sets with both split and replicated histories to assess the effects on
+scalability and run time. Figure~\ref{fig:titan_weak_ms_time} gives
+the wall times for these calculations. Of primary importance here is
 the fact that adding sets to the problem with both split and
@@
 which the linear problem can actually be replicated and enhance time
-to solution, something that cannot be achieved with conventional
-methods. At larger core counts, however, this reduction in compute
-time is not as large due to the fact that the tally reduction across
-blocks requires more resources and those resources are clearly growing
-with core count (most obviously in the 4 set case with split
-histories). In addition, at 131,072 cores for the largest 4 set
-computation, processors that share a block may be some physical
-distance from one another in the machine, adding to latency and
-decreasing performance.
+to solution, something that cannot be achieved with, for example,
+conventional subspace methods. At larger core counts, however, this
+reduction in compute time is not as large due to the fact that the
+tally reduction across sets requires more resources and those
+resources are clearly growing with core count (most obviously in the 4
+set case with split histories). In addition, at 131,072 cores for the
+largest 4 set computation, processors that share the same domain in
+their respective sets may be some physical distance from one another
+in the machine, adding to latency and decreasing performance.
 
@@
   \begin{center}
-    \includegraphics[width=3in]{titan_weak_ms_time.pdf}
+    \includegraphics[width=3.25in]{titan_weak_ms_time.pdf}
   \end{center}
@@
   \begin{center}
-    \includegraphics[width=3in]{titan_weak_ms_eff.pdf}
+    \includegraphics[width=3.25in]{titan_weak_ms_eff.pdf}
   \end{center}
@@
 efficiencies given by Table~\ref{tab:ms_weak_alg_eff}. The iterations
-to converge these problems were not a function of problem size. Again,
-when the implementation efficiencies are computed, we see in
-Figure~\ref{fig:titan_weak_ms_impeff} that splitting histories among
-sets is still a more efficient manner of introducing multiple sets
-rather than replicating histories in an attempt to reduce the number
-of iterations required to converge.  We also note that this is not a
-true weak scaling study as indicated by
+to converge these problems were not a function of problem size.  We
+also note that this is not a true weak scaling study as indicated by
 Eq~(\ref{eq:ms_weak_efficiency}) but rather an odd combination of both
@@
 
-\begin{figure}[h!]
-  \begin{center}
-    \includegraphics[width=3in]{titan_weak_ms_impeff.pdf}
-  \end{center}
-  \caption{Implementation parallel efficiency relative to 16-core
-    1-set reference case.}
-  \label{fig:titan_weak_ms_impeff}
-\end{figure}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-It was found the MCSA can indeed be effectively parallelized using the
+It was found that MCSA can be effectively parallelized using the
 multiple-set overlapping-domain decomposition algorithm borrowed from
@@
 vector reductions create an overhead that is not observed in the
-literature for transport calculations.
+literature for particle transport calculations.
 
@@
 observed to actually improve the time to solution. Unlike a
-traditional subspace method that we might apply to solve a neutron
-transport problem, using MCSA means that we can actually make a
-physical copy of the problem on the machine and can combine separate
-Monte Carlo solutions for each copy through superposition. Time to
-solution is then improved because fewer histories were run in each
-copy and therefore each MCSA iteration is faster or more global
-histories are computed and fewer MCSA iterations are required to
-converge.
+traditional subspace method, using MCSA means that we can actually
+make a physical copy of the problem on the machine and can combine
+separate Monte Carlo solutions for each copy through
+superposition. Time to solution is then improved because fewer
+histories were run in each copy and therefore each MCSA iteration is
+faster or more global histories are computed and fewer MCSA iterations
+are required to converge. Enabling improved time to solution through
+solving multiple copies of the linear problem then has implications
+for future resilient solver techniques where replication may
+potentially mitigate hard errors at the application level.
 
@@
 of such an operation to increase parallel run times. It did, however,
-and therefore future research should explore why the implementation
-for multiple sets generated parallel overhead that increased with core
-count instead of remaining constant.
+and our current hypothesis is that the processors involved in these
+reductions are physically some distance from each other in the
+machine. Future research should explore why the implementation for
+multiple sets generated this parallel overhead that increased with
+core count instead of remaining constant in order to improve the
+scalability of the algorithm when using replication at high core
+counts. In addition, future work should consider optimization of the
+solution algorithm and a performance comparison to conventional
+subspace methods.
 

### 2013-09-08 doc/sna2013/paper/slattery_evans_wilson.tex
fixing typos in sna

@@
 Given the mutual exclusivity of the transport algorithm, an additional
-communcation sequence is required to efficiently stop the algorithm
+communication sequence is required to efficiently stop the algorithm
 after all histories have been computed. To stop transport, Brunner and
@@
 specified for overlap. For each of these overlap states, the
-probablities, weights, tallies, and other information required to
+probabilities, weights, tallies, and other information required to
 implement the Monte Carlo sequence are gathered. When transport
 begins, within each individual set the Neumann-Ulam sequence is
-executed with parallelism provided Brunner and Brantely's algorithm as
+executed with parallelism provided Brunner and Brantley's algorithm as
 outlined in the previous section. Here, transport occurs both in the
 local domain and the overlap region before histories are
-communicated. Once transport in all sets is compelete, the tallies
+communicated. Once transport in all sets is complete, the tallies
 from all overlapping states are gathered to their original owning

### 2013-09-09 doc/dissertation/presentation/defense.tex
updating date in defense talk

@@
 
-\date{\today} 
+\date{September 9, 2013} 
 \title{Parallel Monte Carlo Synthetic Acceleration Methods for
@@
     \caption{Spectral radius results for the block Jacobi
-      preconditioned iteration matrix with 10 energy groups and full
-      downscatter for sample problem.}
+      preconditioned $SP_N$ iteration matrix with 10 energy groups and
+      full downscatter.}
     \label{tab:group10dsbj}
@@
         \begin{center}
-          \includegraphics[width=2.0in]{block_jacobi_res.pdf}
+          \includegraphics[width=2.25in]{block_jacobi_res.pdf}
         \end{center}
@@
       \item Rapidly divergent results
+        \medskip
+      \item Convergence not achieved with 50 histories per DOF and 90
+        minutes compute time
         \medskip
@@
         $\mathbf{\rho}(\mathbf{H}) \approx 1$
-        \medskip
-      \item Convergence not achieved with 50 histories per DOF and 90
-        minutes compute time
       \end{itemize}
@@
         \item Explicit preconditioning strategy destroys sparsity and
-          elevates CPU times (Ifpack ILUT)
+          elevates CPU times
           \medskip

### 2013-09-09 doc/dissertation/document/chapters/conclusion/conclusion.tex
adding Prof Sovinec corrections

@@
 operator is either more sparse, has a better-conditioned eigenvalue
-spectrum, or both. The reduced domain approximation has shown that a
-purely algebraic strategy for reducing the order of the Monte Carlo
-problem can result in improved overall timing performance due to
-increased sparsity at the cost of iterative performance as a result of
-loss of information. A more well informed strategy for order reduction
-via either an intelligent simplification of the PDEs that describe the
-system or a simplification of the physics involved may prove to be a
-more effective strategy.
+spectrum, or both. In addition, the reduced domain approximation has
+shown that a purely algebraic strategy for reducing the order of the
+Monte Carlo problem can result in improved overall timing performance
+due to increased sparsity at the cost of iterative performance as a
+result of loss of information. A more well informed strategy for order
+reduction via either an intelligent simplification of the PDEs that
+describe the system or a simplification of the physics involved may
+prove to be a more effective strategy. In turn, this reduction of
+order strategy for the Monte Carlo problem also potentially simplifies
+the preconditioning strategy with less sophisticated algebraic
+techniques needed for better conditioned approximations of the Monte
+Carlo problem.
 

### 2013-09-09 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
adding Prof Sovinec corrections

@@
   \caption{\textbf{Isotherms from solution to the thermal convection
-        cavity problem.} \textit{Top left: Ra = \sn{1}{3}; Top right:
-        Ra = \sn{1}{4}; Bottom left: Ra = \sn{1}{5}; Bottom left: Ra =
-        \sn{1}{6}. Isotherms on unit temperature scale from 0 to 1
-        with 10 uniformly spaced curves for each calculation.}}
+      cavity problem.} \textit{Top left: Ra = \sn{1}{3}; Top right: Ra
+      = \sn{1}{4}; Bottom left: Ra = \sn{1}{5}; Bottom left: Ra =
+      \sn{1}{6}. Isotherms on unit temperature scale from 0 to 1 with
+      10 uniformly spaced curves for each calculation. These results
+      were computed with the Drekar fluid code and FANM.}}
   \label{fig:convection_isotherms}
@@
       700. Isocurves on unit velocity scale from 0 to 1 with 20
-      uniformly spaced curves for each calculation.}}
+      uniformly spaced curves for each calculation. These results were
+      computed with the Drekar fluid code and FANM.}}
   \label{fig:driven_velocity_isocurves}
@@
       calculation. The depression forming in the upper right portion
-      of visible flow is the second vortex.}}
+      of visible flow is the second vortex. These results were
+      computed with the Drekar fluid code and FANM.}}
   \label{fig:step_velocity_isocurves}

### 2013-09-09 doc/dissertation/umi_abstract/umi_abstract.tex
adding umi abstract

--- 2013-08-26 doc/dissertation/document/frontmatter/abstract.tex
@@
-Predictive modeling and simulation is an important and developing
-component of modern nuclear reactor analysis. New high performance
-computing hardware and algorithms that leverage this hardware are
-needed to help meet the goals and requirements of engineering
-analysis. Current practices in neutron transport and fluid flow
-leverage methods based on subspace iterations to achieve good
-convergence properties and quality parallel scaling on large
-machines. However, the next generation of machines is expected to
-possess both high levels of concurrency and low memory availability
-per core compared to today's machines with a potentially high rate of
-faults. Therefore, we seek new algorithms to solve these problems in
-reactor analysis that have the potential to perform well on
-forthcoming hardware with the goal of improving both iterative
-performance and parallel scalability.
-
 This work researches and develops Monte Carlo Synthetic Acceleration
@@
 criticality calculation for a light water reactor fuel assembly used
-as the driving problem. It was discovered that basic Jacobi-based
-preconditioning as used in previous work was not sufficient to achieve
-convergence due to the dominating effect of the light water
-moderator. To enable MCSA as a solution technique a group of modern
-preconditioning strategies are researched. MCSA was verified to
-produce the correct solution for the fuel assembly criticality problem
-when compared to conventional Krylov methods and demonstrated improved
+as the driving problem. To enable MCSA as a solution technique a group
+of modern preconditioning strategies are researched. MCSA when
+compared to conventional Krylov methods demonstrated improved
 iterative performance over GMRES by converging in fewer iterations
-when using the same preconditioning. The new preconditioning
-techniques, however, create unwanted growth in both memory consumption
-and computation time for these problems.
+when using the same preconditioning.
 
@@
 benchmark problems in both convective and driven flow regimes were
-used to drive the research and development of the method. It was found
-for all benchmarks that FANM produced identical solutions when
-compared to those generated by modern Newton-Krylov methods. For each
-benchmark, performance comparisons between FANM and Newton-Krylov were
-also conducted. For 8 out of 12 benchmark cases, it was found that
-FANM had better iterative performance than the Newton-Krylov method by
-converging the nonlinear residual in fewer linear solver iterations
-with the same preconditioning.
+used to drive the research and development of the method. For 8 out of
+12 benchmark cases, it was found that FANM had better iterative
+performance than the Newton-Krylov method by converging the nonlinear
+residual in fewer linear solver iterations with the same
+preconditioning.
 
@@
 Overlapping-Domain strategy in an attempt to reduce parallel overhead
-and add a natural element of replication to the algorithm. A large
-number of parallel scaling studies using a neutron diffusion problem
-were performed on the Titan Cray XK7 machine at Oak Ridge National
-Laboratory with up to 65,356 cores used in the strong scaling studies
-and 131,072 cores used in the weak scaling studies. It was found that
-for the current implementation of MCSA, both weak and strong scaling
-improved on that observed for production implementations of Krylov
-methods. In addition, the replication in the algorithm was able to
-boost parallel efficiencies by up to 20\% at large core counts in the
-strong scaling studies and enhance time to solution in the weak
-scaling studies by using the principle of superposition. Finally, it
-was discovered that MCSA exhibits the most scalable behavior when it
-is formulated as a stochastic realization of an additive Schwarz
-method.
-
-Each of these research and development activities meet the goals of
-this work by producing methods that in many cases were demonstrated to
-converge in fewer iterations than conventional methods and exhibit
-better parallel efficiencies at high levels of concurrency. All the
-components of this work are required to enable future simulation of
-nonlinear transport phenomena on the next generation of hardware. To
-enable high performance simulations with FANM, we require the MCSA
-techniques developed in the research on the neutron transport problem
-and in turn, effectively leveraging MCSA on leadership-class hardware
-requires the new parallel algorithm developed by this work. With the
-further development of methods based on this work, future nuclear
-reactor simulations may potentially be enabled that have an improved
-time to solution and more efficiently utilize the available
-computational resources.
+and add a natural element of replication to the algorithm. It was
+found that for the current implementation of MCSA, both weak and
+strong scaling improved on that observed for production
+implementations of Krylov methods.

### 2013-09-09 doc/dissertation/document/chapters/conclusion/conclusion.tex
fixed typo on cover page

@@
 MCSA method, the Monte Carlo method may be applied to a reduced-order
-physics or PDE-based system that captures the essence of the original
-system and still allows for acceleration of the iterative
-procedure. By reducing the order of the system, we are effectively
-generating a Monte Carlo problem within MCSA in which the linear
-operator is either more sparse, has a better-conditioned eigenvalue
-spectrum, or both. In addition, the reduced domain approximation has
-shown that a purely algebraic strategy for reducing the order of the
-Monte Carlo problem can result in improved overall timing performance
-due to increased sparsity at the cost of iterative performance as a
-result of loss of information. A more well informed strategy for order
-reduction via either an intelligent simplification of the PDEs that
-describe the system or a simplification of the physics involved may
-prove to be a more effective strategy. In turn, this reduction of
-order strategy for the Monte Carlo problem also potentially simplifies
-the preconditioning strategy with less sophisticated algebraic
-techniques needed for better conditioned approximations of the Monte
-Carlo problem.
+physics or PDE-based system instead of the original system that
+captures the essence of the original system and still allows for
+acceleration of the iterative procedure. By reducing the order of the
+system, we are effectively generating a Monte Carlo problem within
+MCSA in which the linear operator is either more sparse, has a
+better-conditioned eigenvalue spectrum, or both. In addition, the
+reduced domain approximation has shown that a purely algebraic
+strategy for reducing the order of the Monte Carlo problem can result
+in improved overall timing performance due to increased sparsity at
+the cost of iterative performance as a result of loss of
+information. A more well informed strategy for order reduction via
+either an intelligent simplification of the PDEs that describe the
+system or a simplification of the physics involved may prove to be a
+more effective strategy for the acceleration step. In turn, this
+reduction of order strategy for the Monte Carlo problem also
+potentially simplifies the preconditioning strategy with less
+sophisticated algebraic techniques needed for better conditioned
+approximations of the Monte Carlo problem and therefore less
+difficulty with composite operators.
 

### 2013-10-18 doc/jcomp2012/jcomp2012.tex
Final, final edits.
Final response to editors.

@@
 and Leibler \cite{forsythe}.  Their paper credits the idea to unpublished work
-by J.v. Neumann and S.M. Ulam dating back to the 1940's.  The basic principles
-of Monte Carlo matrix inversion were further elucidated in Hammersley and
-Handscomb's 1964 text \cite{hammersley_1964}.  These early methods are
-distinguished by very slow, statistically noisy convergence properties; thus,
-they have not made any significant impact in the physics community.
+by J. von Neumann and S.M. Ulam dating back to the 1940's.  The basic
+principles of Monte Carlo matrix inversion were further elucidated in
+Hammersley and Handscomb's 1964 text \cite{hammersley_1964}.  These early
+methods are distinguished by very slow, statistically noisy convergence
+properties; thus, they have not made any significant impact in the physics
+community.
 
@@
 \end{equation}
-Using this series, we can write an iterative method that solves
-Eq.~(\ref{eq:Ax=b}) (Richardson's Iteration):
+Using this series, we can write an iterative method, known as Richardson's
+Iteration, that solves Eq.~(\ref{eq:Ax=b}):
 \begin{equation}
@@
 \end{equation}
-Here, $m$ is the random walk step index and the $h$ terms are the matrix
+Here, $k$ is the random walk step index and the $h$ terms are the matrix
 elements of $\vH$.  The terms in the Neumann series can be interpreted as a
-series of transitions from $i_{m-1}\rightarrow i_m$ that can be simulated by a
+series of transitions from $i_{k-1}\rightarrow i_k$ that can be simulated by a
 random walk.  Let $X$ be a random variable sampled from a random walk with $k$
@@
 
-An alternative approach to the one just described is to calculate
-contributions to every component of $\vx$ during the random walk. In
-this method, the weight change from state $i\rightarrow j$ is
+An alternative approach to the direct method is to calculate contributions to
+every component of $\vx$ during the random walk. In this method, the weight
+change from state $i\rightarrow j$ is
 \begin{equation}
@@
 \end{equation}
-in which $\vA^{-1}$ indicates the approximate inversion of $\vA$ via the
+in which $\hat{\vA}^{-1}$ indicates the approximate inversion of $\vA$ via the
 adjoint Monte Carlo method.  MCSA, on the other hand, is identical to a
@@
 
-A timing and convergence study is used to demonstrate the effectiveness of the
-adjoint method as compared to the direct method. To assess both the CPU time
-and number of iterations required to converge to a solution, a problem of
-constant $\Delta$ was used with varying number of grid points, fixing the
-spectral radius of the system at a constant value for each variation. Both the
-five-point and nine-point stencils were used with both the direct and adjoint
-solvers. For each case, $N \times N$ total random walk permutations were
-computed per MCSA iteration where $N \times N$ is the number of discrete grid
-points in the system. Solver parameters were set to a weight cutoff of
-\sn{1}{-4} for the stochastic linear solver and a convergence tolerance of
-\sn{1}{-8} for the MCSA iterative solver.  Figure~\ref{fig:poisson_cpu_time}
-gives the CPU time needed for each case to converge in seconds and
-Figure~\ref{fig:poisson_iterations} gives the number of iterations needed for
-each case to converge to the specified tolerance as a function of the problem
-size. All computations presented in this section and
+A timing and convergence study is used to demonstrate the improved
+effectiveness of the adjoint method as compared to the direct method. To
+assess both the CPU time and number of iterations required to converge to a
+solution, a problem of constant $\Delta$ was used with varying number of grid
+points, fixing the spectral radius of the system at a constant value for each
+variation. Both the five-point and nine-point stencils were used with both the
+direct and adjoint solvers. For each case, $N \times N$ total random walk
+permutations were computed per MCSA iteration where $N \times N$ is the number
+of discrete grid points in the system. Solver parameters were set to a weight
+cutoff of \sn{1}{-4} for the stochastic linear solver and a convergence
+tolerance of \sn{1}{-8} for the MCSA iterative solver.
+Figure~\ref{fig:poisson_cpu_time} gives the CPU time needed for each case to
+converge in seconds and Figure~\ref{fig:poisson_iterations} gives the number
+of iterations needed for each case to converge to the specified tolerance as a
+function of the problem size. All computations presented in this section and
 \S~\ref{subsec:sequential_comparison} were completed on a 3.0 GHz Intel Core 2
@@
     used with the five point and nine point stencils. A CPU time
-    speedup is noted with the adjoint method due to the higher density
+    speedup is achieved with the adjoint method due to the higher density
     of random walk events in regions with a large residual.}
@@
 
-We see clearly in Figure~\ref{fig:poisson_cpu_time} that the using the
-adjoint solver with MCSA results in a speedup over the direct solver
-while the number of iterations required to converge is also reduced as
-shown in Figure~\ref{fig:poisson_iterations}. We expect this for several
-reasons. First, with an equivalent number of histories specified for
-both solvers per MCSA iteration and a system of size $N \times N$, the
-direct solver will compute a single random walk for each state in the
-system per iteration to acquire a solution in that state, regardless
-of the size of the residual in that state. This is necessary in the
-direct method to ensure a contribution from each state as the random
-walk sequence will only contribute to the starting state. For the
-adjoint method, a total of $N \times N$ random walk events will have
-their starting state determined by sampling the residual
-vector. Because the random walk sequence contributes to the state in
-which it currently resides, sampling the residual vector as the Monte
-Carlo source gives a higher density of random walk events in regions
-with a high residual, thus giving a more accurate correction in that
-region due to reduced statistical error. From an iteration
-perspective, Figure~\ref{fig:poisson_iterations} shows that using the
-direct method yields a roughly unchanging number of iterations
-required to converge as the problem size increases. Again, if we
-desire a correction value for all states in the problem, then we must
-start a random walk in each state in the system which does not reduce
-the number of iterations need as the problem size grows. Conversely,
-as the problem size grows in the adjoint method, the additional
-stochastic histories that will be computed are concentrated in regions
-with a large residual, further reducing the stochastic error in the
-correction in those regions and subsequently reducing the required
+We see in Figures~\ref{fig:poisson_cpu_time} and \ref{fig:poisson_iterations}
+that the adjoint solver with MCSA is faster and requires fewer iterations than
+the direct method. Several reasons explain this result. First, with an
+equivalent number of histories specified for both solvers per MCSA iteration
+and a system of size $N \times N$, the direct solver will compute a single
+random walk for each state in the system per iteration to acquire a solution
+in that state, regardless of the size of the residual in that state. This is
+necessary in the direct method to ensure a contribution from each state as the
+random walk sequence will only contribute to the starting state. For the
+adjoint method, a total of $N \times N$ random walk events will have their
+starting state determined by sampling the residual vector. Because the random
+walk sequence contributes to the state in which it currently resides, sampling
+the residual vector as the Monte Carlo source gives a higher density of random
+walk events in regions with a high residual, thus giving a more effective
+correction in those regions due to reduced statistical error. From an
+iteration perspective, Figure~\ref{fig:poisson_iterations} shows that using
+the direct method yields a roughly unchanging number of iterations required to
+converge as the problem size increases. Again, if we desire a correction value
+for all states in the problem, then we must start a random walk in each state
+in the system which does not reduce the number of iterations needed as the
+problem size grows. Conversely, as the problem size grows in the adjoint
+method, the additional stochastic histories that will be computed are
+concentrated in regions with a large residual, further reducing the stochastic
+error in the correction in those regions and ultimately reducing the required
 number of iterations to converge.
@@
 Poisson problem as described in the previous section and choose only the
-5-point stencil to discretize the Laplacian operator as the previous results
-yielded little qualitative difference between the discretizations. Both MCSA
-and Halton's method are used with the adjoint Monte Carlo solver. In order to
-complete the same study as in the previous section, the number of histories
-computed by the Monte Carlo solver at each iteration had to be doubled to $2
-\times N \times N$ in order to ensure convergence when using the sequential
-Monte Carlo Method. Figure~\ref{fig:seq_cpu_time} gives the CPU time results
-for this comparison as a function of problem size while
+5-point stencil to discretize the Laplacian operator because the previous
+results yielded little qualitative difference between the
+discretizations. Both MCSA and Halton's method are used with the adjoint Monte
+Carlo solver. In order to complete the same study as in the previous section,
+the number of histories computed by the Monte Carlo solver at each iteration
+had to be doubled to $2 \times N \times N$ in order to ensure convergence when
+using the sequential Monte Carlo Method. Figure~\ref{fig:seq_cpu_time} gives
+the CPU time results for this comparison as a function of problem size, and
 Figure~\ref{fig:seq_iterations} gives the number of iterations to converge as
@@
 
-The benefits of using a synthetic acceleration scheme are also noted when the
-infinity norm of the residual computed at each iteration for both methods was
-collected at each iteration for a fixed problem sizes of $N=100$ and $N=500$
-as shown in figures Figure~\ref{fig:seq_100} and \ref{fig:seq_500}
+The benefits of using a synthetic acceleration scheme are also illustrated
+when comparing the residuals at each iteration.  Figures~\ref{fig:seq_100} and
+\ref{fig:seq_500} show the infinity norm of the residual computed at each
+iteration for both methods for fixed problems sizes of $N=100$ and $500$,
 respectively.  In both cases, the sequential method is characterized by a slow
@@
 that this discrete form is amenable to interpretation using Monte Carlo
-concepts and can be written in a way that a random walk sequence can be
-generated.
+concepts and can be written in a way that generates a random walk sequence.
 
@@
 
-We choose to use the angle-energy integrated radiation intensity,
+We choose to use the angle- and energy-integrated radiation intensity,
 $\phi\equiv\phi(\ve{r},t)$
@@
   -\nabla\cdot \Dn\grad{\phi} + \sign\phi = \qn\:,
+  \label{eq:backwards_euler}
 \end{equation}
@@
 negative and symmetric.  In other words, the leakage out of a cell must be
-equivalent to the source into the adjacent cell.
+equivalent to its corresponding source into the adjacent cell.
 
@@
 \end{center}
-The problem is run to $100$~nanoseconds.
+The problem is run to $100$~nanoseconds.  The variable timestep control
+started with \sn{1.25}{-9}~ns and increased to 0.01~ns using a maximum change
+in $\Delta t$ of 10\% per cycle.  This maximum timestep was selected based on
+numerical experiments that showed good agreement with analytic solutions of
+the Marshak wave problem for the time-discretization employed in
+Eq.~(\ref{eq:backwards_euler}).
 
@@
 number of particles per iteration for a relative weight cutoff of
-$\sn{1}{-4}$.  For less than five particles per iteration, the method does not
-converge.  Figure~\ref{fig:CPU_Ns} shows the variation in the maximum number
-of iterations per cycle with number of particles.  Analysis of these two
-figures shows that increasing the number of particles per iteration reduces
-the number of iterations required to converge the solution.  However, the cost
-of the Monte Carlo transport is high; thus, better performance is attained by
-running more iterations with fewer particles per iteration \cite{evans_2003}.
+$\sn{1}{-4}$.  For fewer than five particles per iteration, the method does
+not converge.  Figure~\ref{fig:CPU_Ns} shows the variation in the maximum
+number of iterations per cycle with number of particles.  Analysis of these
+two figures shows that increasing the number of particles per iteration
+reduces the number of iterations required to converge the solution.  However,
+the cost of the Monte Carlo transport is high; thus, better performance is
+attained by running more iterations with fewer particles per iteration
+\cite{evans_2003}.
 \begin{figure}[ht!]
@@
 \end{center}
+The timestep controls from the Marshak problem were used here resulting in a
+maximum timestep of 0.01~ns.
 

### 2013-10-18 doc/jcomp2012/jcomp2012.tex
Last edits.

@@
 \end{equation}
-Using this series, we can write an iterative method, known as Richardson's
-Iteration, that solves Eq.~(\ref{eq:Ax=b}):
+Using this series, we can write an iterative method, known as Richardson
+iteration, that solves Eq.~(\ref{eq:Ax=b}):
 \begin{equation}
@@
 The MCSA method defined in Eq.~(\ref{eq:MCSA-iteration}) uses the adjoint
-method to estimate the error in Eq.~(\ref{eq:MCSA-residual_solve}) instead of
-the direct method outlined in \S~\ref{sec:direct-method}. To demonstrate the
+method to estimate the error in Eq.~(\ref{eq:MCSA-residual_solve}).
+To demonstrate the
 effectiveness of the adjoint method over the direct method within the context
